### 2012-02-10 p.tex
init

\documentclass[preprint]{sigplanconf}

\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{xspace}
\usepackage{multicol}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newenvironment{proof}[1][Proof.]{\begin{trivlist}\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\arrow}[1]{\mathtt{#1}}

\begin{document}

\conferenceinfo{ICFP'12}{}
\CopyrightYear{}
\copyrightdata{}
\titlebanner{}
\preprintfooter{}

\title{Continuations via Negative Information Flow}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\category{D.3.1}{Formal Definitions and Theory}{}
\category{F.3.2}{Semantics of Programming Languages}{}
\category{F.3.3}{Studies of Program Constructs}{Type structure}

\terms
Languages, Theory

\keywords Arrows, Linear logic, Quantum computing, Reversible logic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Blah\cite{infeffects}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

\acks This project was partially funded by Indiana University's Office of the
Vice President for Research and the Office of the Vice Provost for Research
through its Faculty Research Support Program.  We also acknowledge support
from Indiana University's Institute for Advanced Study.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{small}
\bibliographystyle{abbrvnat}
\bibliography{cites}
\end{small}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-02-12 dual.tex
added omake, hgignore etc

\documentclass[preprint]{sigplanconf}

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\newcommand{\xcomment}[2]{\textbf{#1:~\textsl{#2}}}
\newcommand{\amr}[1]{\xcomment{Amr}{#1}}
\newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}

\newcommand{\ie}{\textit{i.e.}\xspace}
\newcommand{\eg}{\textit{e.g.}\xspace}

\newcommand{\code}[1]{\lstinline[basicstyle=\small]{#1}\xspace}
\newcommand{\name}[1]{\code{#1}}

\def\newblock{}

\newenvironment{floatrule}
    {\hrule width \hsize height .33pt \vspace{.5pc}}
    {\par\addvspace{.5pc}}

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = &::=&
%! /= = \neq
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! theseus = Theseus

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\conferenceinfo{ICFP'12}{}
\CopyrightYear{}
\copyrightdata{}
\titlebanner{}
\preprintfooter{}

\title{Continuations via Negative Information Flow}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
TODO
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\category{D.3.1}{Formal Definitions and Theory}{}
\category{F.3.2}{Semantics of Programming Languages}{}
\category{F.3.3}{Studies of Program Constructs}{Type structure}

\terms
Languages, Theory

\keywords Arrows, Linear logic, Quantum computing, Reversible logic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

We introduced this thesis that computation should be based on isomorphisms
that preserve information~\cite{infeffects}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Value Producers and Consumers}

Consider the following ways of purchasing an item that costs \$20.00:

\begin{enumerate}
\item You produce a \$20.00 bill and give it to the seller.
\item You use a credit card. 
\end{enumerate}
In both cases the seller receives the money immediately but there is a subtle
difference. In the second transation, nobody has yet produced the money that
the seller has received: instead a \emph{debt} is generated and this debt
travels ``backwards'' towards the bank and (hopefully) reconciled there.

Computationally we model the first transaction using essentially the identity
function which receives a \$20.00 bill as input from the buyer and propagates
it on its output to the seller. The second transaction is more
complicated. We model it as shown in the circuit below:

FIGURE

There are two ways to understand this circuit that are both quite
instructive. Let's first examine the type structure of each of the
combinators that comprise the circuit. The first combinator on the right
outputs \$20.00 to the seller. This \$20.00 is produced from nothing so to
speak by generating an equivalent debt that travels backwards. COMPLETE BASED
ON THE FIGURE.

The othe way is to follow the execution. It goes forward in time so to speak,
comes back, and then goes forwards again.

\paragraph*{Continuations.} The idea of using negative types to model
information flowing backwards, demand for values, continuations, etc. goes
back to at Filinski's thesis. We recall these connections below but we first
note that all these systems are complicated because in all these systems
information can be ignored, destroyed, or duplicated. Clearly the possibility
of erasure of information would mean that our credit card transaction is
incorrect. In our work, information is maintained and hence we have a
guarantee that, in a closed program, the debt must be accounted and paid for.

\paragraph*{Filinski.}

\paragraph*{Ariola, Herbelin, and Sabry.} We use subtractive types. Things
are constrained so that the positive and negative values must be used
together. In our new work we have no such constrained, the negative value can
flow anywhere. The entire system guarantees that any closed program would
have to account for it. We don't have to introduce special constraints to
achieve that.

\paragraph*{Focusing, Polarization, etc.}

\paragraph*{Int Construction.}

\paragraph*{Recursion.} In our previous work we introduce recursive types and
trace operators. This is dangerous here because infinite loops allow us to
prolong paying the debt for as long as we want.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Additive Duality in {{langRev}} }

%%%%%%%%%%%%%%%
\subsection{Syntax}

%subcode{bnf} include main
% Values, v = () | (v, v) | L v | R v
% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
%
% Sequential Contexts, P, F = [] | P:c
% Parallel Contexts, D = [] 
%                     &|& (P, [](+)c, F):D 
%                     &|&  (P, c(+)[], F):D
%                     &|& (P, c v, F):D 
%                     &|&  (P, v c, F):D
%
%
% Machine States  = <P, F, c v, D> | <P, F, v c, D>
% Start State  = <[], [], c v, []>
% Stop State  = <P, [], v c, []>

Combinator reconstruction, {{P[c]}}: 
%subcode{opsem} include main
% [][c] '= c
% (P:c')[c] '= P[c'(;)c]

%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

The small step semantics present for {{langRev}} below work
symmetrically for forward and backward evaluation.

\begin{itemize}
\item 
Basic reduction of an isomorphism. Note that the evaluation leaves the
adjoint of the combinator behind. This will become important when we
reverse the direction of computation.
%subcode{opsem} include main
% <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>

\item
Sequencing involves pushing and popping from the Future and Past
continuations:
%subcode{opsem} include main
% <P, F, c1(;)c2 v , D>  &|-->& <P, F:c2, c1 v, D>
% <P, F:c2, v c1, D> &|-->& <P:c1, F, c2 v, D>

\item
Parallel composition, captures the current Future and Past and extends
a Dump/meta-continuation.
%subcode{opsem} include main
% <P, F, c1 (+) c2 (L v), D> &|-->& <[], [], c1 v, (P,[](+)c2,F):D>
% <P, F, c1 (+) c2 (R v), D> &|-->& <[], [], c2 v, (P,c1(+)[],F):D> 
% <P', [], v c1, (P,[](+)c2,F):D> &|-->& < P, F, (L v) (P'[c1](+)c2{dagger}), D > 
% <P', [], v c2, (P,c1(+)[],F):D> &|-->& <P, F, (R v) (c1{dagger}(+)P'[c2]), D>

\item
Similarly for products:
%subcode{opsem} include main
% <P, F, c1 (*) c2 (v1, v2), D> &|-->& <[], [], c1 v1, (P, [] (*) c2 v2,F):D>
% <P', [], v1 c1, (P,[] (*) c2 v2,F):D> &|-->& <[], [], c2 v2, (P,v1 P'[c1] (*) [],F):D>
% <P', [], v2 c2, (P,v1 c1 (*) [],F):D> &|-->& <P, F, (v1, v2) (c1(*)P'[c2]):D>
and symmetrical rules for evaluation along the second branch. 
%subcode{opsem} include main
% <P', [], v2 c2, (P,c1 v1 (*) [],F):D> &|-->& <[], [], c1 v1, (P,[] (*) v2 P'[c2],F):D>
% <P', [], v1 c1, (P,[] (*) v2 c2,F):D> &|-->& <P, F, (v1, v2) (P'[c1] (*) c2):D>

The later two rules will be relevant only for reverse execution.

\end{itemize}

%%%%%%%%%%%%%%%%%
\section{Rules for {{eta}} and {{eps}} }

The operation {{eps}} reverses the direction of a particle by reversing the world.

\begin{itemize}
\item Grammar
%subcode{bnf} include main
% Values, v = () | (v, v) | L v | R v | -v
% Isomorphisms, iso &=& ... | eta | eps
% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c

\item
Type judgement.
%subcode{opsem} include main
% eta &: 0 <-> (-b) + b :& eps

%subcode{proof} include main
%@ |- v : b
%@@ |- -v : -b

\item
Operational Semantics.
%subcode{opsem} include main
% <P; F; eps (R v); D>      &|-->& <F; P; (L (-v))~ eps; D{dagger}>

Note: there is NO reduction rule for {{eta}}.

\item
The adjoint of a Dump is defined to be:
%subcode{opsem} include main
% []{dagger} '= []
% ((P, [](+)c, F):D){dagger} '= (F, [](+)c{dagger}, P):D{dagger}
% ((P, c(+)[], F):D){dagger} '= (F, c{dagger}(+)[], P):D{dagger}
% ((P, [](*)c v, F):D){dagger} '= (F, [](*)v c, P):D{dagger}
% ((P, [](*)v c, F):D){dagger} '= (F, [](*)c v, P):D{dagger}
% ((P, c v(*)[], F):D){dagger} '= (F, v c(*)[], P):D{dagger}
% ((P, v c(*)[], F):D){dagger} '= (F, c v(*)[], P):D{dagger}

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

\acks This project was partially funded by Indiana University's Office of the
Vice President for Research and the Office of the Vice Provost for Research
through its Faculty Research Support Program.  We also acknowledge support
from Indiana University's Institute for Advanced Study.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{small}
\bibliographystyle{abbrvnat}
\bibliography{cites}
\end{small}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-02-13 dual.tex
new formatting for the semantics
dgms
more diagrams and notes

@@
 \usepackage{multicol}
+\usepackage{bm}
 
@@
 \newcommand{\eg}{\textit{e.g.}\xspace}
+
+\newcommand{\lcal}{\ensuremath{\lambda}-calculus}
+\newcommand{\G}{\ensuremath{\mathcal{G}}\xspace}
 
@@
 %! ~> = \leadsto
+%! -o = \multimap
 %! ::= = &::=&
@@
 %! langRev = \Pi
+%! langRevT = \Pi^{o}
+%! langRevEE = \Pi^{\eta\epsilon}
 %! theseus = Theseus
+%! * = \times
 
@@
 \begin{abstract}
-TODO
+We show a deep symmetry between functions and delimited continuations,
+values and continuations that arises in {{langRev}} in a manner that
+is reminiscent of Filinksi's Symmetric \lcal
+~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
+extending {{langRev}} with a notion of additive duality over the
+monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
+Compact Closed Categories. The resulting dual types, which we denote
+{{-b}}, have a time traveling ``backward information flow''
+interpretation and allow for the encoding of higher-order function and
+iteration via the construction of {{trace}} operators, thereby making
+the extended language {{langRevEE}} a Turing-complete reversible
+programming language with higher-order functions and first-class
+delimited continuations.
 \end{abstract}
@@
 \paragraph*{Int Construction.}
+For a traced monoidal category {{C}} the Int construction produces a
+Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
+Further we know that the target of the Int construction is isomorphic
+to the target of \G construction of Abramsky \cite{Abramsky96:0} from
+Haghverdi. However, note that the {{langRevEE}} is not the same as the
+image of the Int construction on {{langRevT}}, since the the later
+lacks a multiplicative tensor that distrbutes over the additive tensor
+in Int {{langRevT}}.
 
@@
 prolong paying the debt for as long as we want.
+
+\paragraph*{GoI machines} 
+We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
 
@@
 % Parallel Contexts, D = [] 
-%                     &|& (P, [](+)c, F):D 
-%                     &|&  (P, c(+)[], F):D
-%                     &|& (P, c v, F):D 
-%                     &|&  (P, v c, F):D
+%                     &|& D:(P| [](+)c| F)
+%                     &|& D:(P| c(+)[]| F)
+%                     &|& D:(P| [] (*) c v| F)
+%                     &|& D:(P| [] (*) v c| F)
+%                     &|& D:(P| c v (*)  []| F)
+%                     &|& D:(P| v c (*) []| F)
 %
 %
-% Machine States  = <P, F, c v, D> | <P, F, v c, D>
-% Start State  = <[], [], c v, []>
-% Stop State  = <P, [], v c, []>
+% Machine States  = D[P| c v| F] | D[P| v c| F]
+% Start State  = [][[]| v c| []]
+% Stop State  = [][P| c v| []]
 
@@
 %subcode{opsem} include main
-% [][c] '= c
-% (P:c')[c] '= P[c'(;)c]
+% [](c) '= c
+% P:c'(c) '= P(c'(;)c)
 
@@
 reverse the direction of computation.
+%% %subcode{opsem} include main
+%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
+
 %subcode{opsem} include main
-% <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
+% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
 
@@
 %subcode{opsem} include main
-% <P, F, c1(;)c2 v , D>  &|-->& <P, F:c2, c1 v, D>
-% <P, F:c2, v c1, D> &|-->& <P:c1, F, c2 v, D>
+% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
+% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
 
@@
 Parallel composition, captures the current Future and Past and extends
-a Dump/meta-continuation.
+the parallel context.
 %subcode{opsem} include main
-% <P, F, c1 (+) c2 (L v), D> &|-->& <[], [], c1 v, (P,[](+)c2,F):D>
-% <P, F, c1 (+) c2 (R v), D> &|-->& <[], [], c2 v, (P,c1(+)[],F):D> 
-% <P', [], v c1, (P,[](+)c2,F):D> &|-->& < P, F, (L v) (P'[c1](+)c2{dagger}), D > 
-% <P', [], v c2, (P,c1(+)[],F):D> &|-->& <P, F, (R v) (c1{dagger}(+)P'[c2]), D>
+% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
+% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
+% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
+% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
 
@@
 %subcode{opsem} include main
-% <P, F, c1 (*) c2 (v1, v2), D> &|-->& <[], [], c1 v1, (P, [] (*) c2 v2,F):D>
-% <P', [], v1 c1, (P,[] (*) c2 v2,F):D> &|-->& <[], [], c2 v2, (P,v1 P'[c1] (*) [],F):D>
-% <P', [], v2 c2, (P,v1 c1 (*) [],F):D> &|-->& <P, F, (v1, v2) (c1(*)P'[c2]):D>
+% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
+% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
+% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
+
 and symmetrical rules for evaluation along the second branch. 
 %subcode{opsem} include main
-% <P', [], v2 c2, (P,c1 v1 (*) [],F):D> &|-->& <[], [], c1 v1, (P,[] (*) v2 P'[c2],F):D>
-% <P', [], v1 c1, (P,[] (*) v2 c2,F):D> &|-->& <P, F, (v1, v2) (P'[c1] (*) c2):D>
+% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
+% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
 
@@
 
-The operation {{eps}} reverses the direction of a particle by reversing the world.
+The operation {{eps}} reverses the direction of a particle by
+reversing the world.
+
+Note that we deviate from the categorical definition of {{eta}} and
+{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
+choice of {{eta}}. This however does not affect us, because we deal
+with a symmetric category.
 
@@
 %subcode{opsem} include main
-% <P; F; eps (R v); D>      &|-->& <F; P; (L (-v))~ eps; D{dagger}>
+% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
 
@@
 \item
-The adjoint of a Dump is defined to be:
+The adjoint of a parallel context is defined to be:
 %subcode{opsem} include main
 % []{dagger} '= []
-% ((P, [](+)c, F):D){dagger} '= (F, [](+)c{dagger}, P):D{dagger}
-% ((P, c(+)[], F):D){dagger} '= (F, c{dagger}(+)[], P):D{dagger}
-% ((P, [](*)c v, F):D){dagger} '= (F, [](*)v c, P):D{dagger}
-% ((P, [](*)v c, F):D){dagger} '= (F, [](*)c v, P):D{dagger}
-% ((P, c v(*)[], F):D){dagger} '= (F, v c(*)[], P):D{dagger}
-% ((P, v c(*)[], F):D){dagger} '= (F, c v(*)[], P):D{dagger}
+% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
+% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
+% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
+% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
+% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
+% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
+
+\end{itemize}
+
+\section{Diagrams}
+
+{{eta}}
+
+\begin{center}
+  \includegraphics{diagrams/eta.pdf}
+\end{center}
+
+Note that the connective is a {{+}}, hence only one of the branches
+may be inhabited at any time. Thus the action of {{eta}} is to
+transfer a backward flowing value on one wire to a forward flowing
+value on the other wire.
+
+{{eps}}
+
+\begin{center}
+  \includegraphics{diagrams/eps.pdf}
+\end{center}
+
+Coherence condition
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+Function
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
+
+Function application
+
+\begin{center}
+  \includegraphics{diagrams/apply1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/apply2.pdf}
+\end{center}
+
+Function composition
+
+\begin{center}
+  \includegraphics{diagrams/compose1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/compose.pdf}
+\end{center}
+
+Delimited continuation
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
+input {{v1 : b1}} gives us a value {{v2 : b2}} on
+termination. Execution is well defined only if {{b1}} and {{b2}} are
+entirely positive types. Consider the program that
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The ouputs of
+programs that ouput mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
+
+\subsection{To think about}
+
+\begin{itemize}
+
+\item {{(-b1) * b2}} is not the same as {{b1 * (-b2)}} as there is no
+  isomorphism to translate from one to the other.
+
+\item {{-(b1 * b2)}} is not the same as {{b1 * b2}} and is also not
+  the same as {{(-b1) * (-b2)}} as there is no isomorphism that goes
+  from one to the other.
+
+\item {{-(-b)}} is not the same as {{b}} as there is no isomorphism
+  that goes from one to the other.
 
@@
 
-\acks This project was partially funded by Indiana University's Office of the
-Vice President for Research and the Office of the Vice Provost for Research
-through its Faculty Research Support Program.  We also acknowledge support
-from Indiana University's Institute for Advanced Study.
+\acks This project was partially funded by Indiana University's Office
+of the Vice President for Research and the Office of the Vice Provost
+for Research through its Faculty Research Support Program.  We also
+acknowledge support from Indiana University's Institute for Advanced
+Study.
 

### 2012-02-13 dual.tex
trace
_

@@
 
+Trace
+
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
 \paragraph*{Observability.} 
@@
 
+\item If we built an effect over {{langRevEE}}, say {{create}} and
+  {{erase}}, are effects structured by an arrow or a monad now?
+
+\item These functions aren't really values. There is no value one can
+  produce that denotes a function. These functions are the ability to
+  transform a value - the possibility of transforming a value.  In the
+  product encoding of environments, there is no value one can assign
+  to a variable such that it denotes a function.
+
 \end{itemize}

### 2012-02-13 dual.tex
added neg circuits
added function dispatch

@@
 
+Actually this is possible. Consider two functions {{f : b1 <-> b2}}
+and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
+discriminate them. Hence the {{boo}} is the first class representation
+of the functions and can be thought of as the address of the function.
+
+\begin{center}
+  \includegraphics{diagrams/dispatch.pdf}
+\end{center}
+
+\item It is not fair to say that negative types flow backwards. The
+  following circuits are valid in {{langRevEE}}. It is however proper
+  to say that for any type {{b}} that flows in a direction, the type
+  {{-b}} flows in the reverse direction.
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit2.pdf}
+\end{center}
+
 \end{itemize}

### 2012-02-13 p.tex
paper

@@
 
+\usepackage{graphicx}
+\usepackage{longtable}
+\usepackage{comment}
+\usepackage{amsmath}
+\usepackage{mdwlist}
+\usepackage{txfonts}
+\usepackage{xspace}
+\usepackage{amstext}
 \usepackage{amssymb}
-\usepackage{graphicx}
+\usepackage{stmaryrd}
 \usepackage{proof}
-\usepackage{stmaryrd}
-\usepackage{url}
-\usepackage{xspace}
+\usepackage{multicol}
+\usepackage[nodayofweek]{datetime}
+\usepackage{etex}
+\usepackage[all, cmtip]{xy}
+\usepackage{xcolor}
+\usepackage{listings}
 \usepackage{multicol}
@@
 \newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{definition}[theorem]{Definition}
 \newtheorem{proposition}[theorem]{Proposition}
-\newenvironment{proof}[1][Proof.]{\begin{trivlist}\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+\newtheorem{corollary}[theorem]{Corollary}
 
-\newcommand{\arrow}[1]{\mathtt{#1}}
+\newcommand{\xcomment}[2]{\textbf{#1:~\textsl{#2}}}
+\newcommand{\amr}[1]{\xcomment{Amr}{#1}}
+\newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}
+
+\newcommand{\ie}{\textit{i.e.}\xspace}
+\newcommand{\eg}{\textit{e.g.}\xspace}
+
+\newcommand{\code}[1]{\lstinline[basicstyle=\small]{#1}\xspace}
+\newcommand{\name}[1]{\code{#1}}
+
+\def\newblock{}
+
+\newenvironment{floatrule}
+    {\hrule width \hsize height .33pt \vspace{.5pc}}
+    {\par\addvspace{.5pc}}
+
+%subcode-inline{bnf-inline} name langRev
+%! swap+ = \mathit{swap}^+
+%! swap* = \mathit{swap}^*
+%! dagger =  ^{\dagger}
+%! assocl+ = \mathit{assocl}^+
+%! assocr+ = \mathit{assocr}^+
+%! assocl* = \mathit{assocl}^*
+%! assocr* = \mathit{assocr}^*
+%! identr* = \mathit{uniti}
+%! identl* = \mathit{unite}
+%! dist = \mathit{distrib}
+%! factor = \mathit{factor}
+%! (o) = \fatsemi
+%! (;) = \fatsemi
+%! (*) = \times
+%! (+) = +
+
+%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
+%! Gx = \Gamma^{\times}
+%! G = \Gamma
+%! [] = \Box
+%! |-->* = \mapsto^{*}
+%! |-->> = \mapsto_{\ggg}
+%! |--> = \mapsto
+%! |- = \vdash
+%! ==> = \Longrightarrow
+%! <== = \Longleftarrow
+%! <=> = \Longleftrightarrow
+%! <-> = \leftrightarrow
+%! ~> = \leadsto
+%! ::= = &::=&
+%! /= = \neq
+%! forall = \forall
+%! exists = \exists
+%! empty = \epsilon
+%! langRev = \Pi
+%! theseus = Theseus
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{abstract}
-
+TODO
 \end{abstract}
@@
 
-Blah\cite{infeffects}
+We introduced this thesis that computation should be based on isomorphisms
+that preserve information~\cite{infeffects}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Value Producers and Consumers}
+
+Consider the following ways of purchasing an item that costs \$20.00:
+
+\begin{enumerate}
+\item You produce a \$20.00 bill and give it to the seller.
+\item You use a credit card. 
+\end{enumerate}
+In both cases the seller receives the money immediately but there is a subtle
+difference. In the second transaction, nobody has yet produced the money that
+the seller has received: instead a \emph{debt} is generated and this debt
+travels ``backwards'' towards the bank and (hopefully) reconciled there.
+
+Computationally we model the first transaction using essentially the identity
+function which receives a \$20.00 bill as input from the buyer and propagates
+it on its output to the seller. The second transaction is more
+complicated. We model it as shown in the circuit below:
+
+FIGURE
+
+There are two ways to understand this circuit that are both quite
+instructive. Let's first examine the type structure of each of the
+combinators that comprise the circuit. The first combinator on the right
+outputs \$20.00 to the seller. This \$20.00 is produced from nothing so to
+speak by generating an equivalent debt that travels backwards. COMPLETE BASED
+ON THE FIGURE.
+
+The other way is to follow the execution. It goes forward in time so to speak,
+comes back, and then goes forwards again.
+
+\paragraph*{Continuations.} The idea of using negative types to model
+information flowing backwards, demand for values, continuations, etc. goes
+back to at Filinski's thesis. We recall these connections below but we first
+note that all these systems are complicated because in all these systems
+information can be ignored, destroyed, or duplicated. Clearly the possibility
+of erasure of information would mean that our credit card transaction is
+incorrect. In our work, information is maintained and hence we have a
+guarantee that, in a closed program, the debt must be accounted and paid for.
+
+\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
+In his Masters thesis, Filinski proposes that continuations are a
+\emph{declarative} concept. He, furthermore, introduces a symmetric extension
+of the $\lambda$-calculus in which values and continuations are treated as
+opposites. This is essentially what we are proposing with one fundamental
+difference: our underlying language is not the $\lambda$-calculus but a
+language of pure isomorphisms in which information is preserved. This shift
+of perspective enables us to distill and generalize the duality of values and
+continuations: in particular, in the conventional $\lambda$-calculus setting
+values and continuations can be erased and duplicated which makes it
+difficult to maintain the correspondence between a value and its negative
+counterpart. In contrast, in our setting, one can start from the empty type
+$0$, introduce a positive value and its negative counterpart, and let each of
+these flow in arbitrary ways. The entire framework guarantees that neither
+the value nor its negative counterpart will be deleted or duplicated and
+hence that, in any closed program, the ``debt'' corresponding to the negative
+value is paid off exactly once.
+
+The forward and backward executions in our framework correspond to
+call-by-value and call-by-name. This duality was observed by Filinski and
+others following him but it is particularly clean in our framework.
+
+\paragraph*{Ariola, Herbelin, and Sabry.} We use subtractive types. Things
+are constrained so that the positive and negative values must be used
+together. In our new work we have no such constrained, the negative value can
+flow anywhere. The entire system guarantees that any closed program would
+have to account for it. We don't have to introduce special constraints to
+achieve that.
+
+\paragraph*{Focusing, Polarization, etc.}
+
+\paragraph*{Int Construction.}
+
+\paragraph*{Recursion.} In our previous work we introduce recursive types and
+trace operators. This is dangerous here because infinite loops allow us to
+prolong paying the debt for as long as we want.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Additive Duality in {{langRev}} }
+
+%%%%%%%%%%%%%%%
+\subsection{Syntax}
+
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
+%
+% Sequential Contexts, P, F = [] | P:c
+% Parallel Contexts, D = [] 
+%                     &|& (P, [](+)c, F):D 
+%                     &|&  (P, c(+)[], F):D
+%                     &|& (P, c v, F):D 
+%                     &|&  (P, v c, F):D
+%
+%
+% Machine States  = <P, F, c v, D> | <P, F, v c, D>
+% Start State  = <[], [], c v, []>
+% Stop State  = <P, [], v c, []>
+
+Combinator reconstruction, {{P[c]}}: 
+%subcode{opsem} include main
+% [][c] '= c
+% (P:c')[c] '= P[c'(;)c]
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection{Operational Semantics}
+
+The small step semantics present for {{langRev}} below work
+symmetrically for forward and backward evaluation.
+
+\begin{itemize}
+\item 
+Basic reduction of an isomorphism. Note that the evaluation leaves the
+adjoint of the combinator behind. This will become important when we
+reverse the direction of computation.
+%subcode{opsem} include main
+% <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
+
+\item
+Sequencing involves pushing and popping from the Future and Past
+continuations:
+%subcode{opsem} include main
+% <P, F, c1(;)c2 v , D>  &|-->& <P, F:c2, c1 v, D>
+% <P, F:c2, v c1, D> &|-->& <P:c1, F, c2 v, D>
+
+\item
+Parallel composition, captures the current Future and Past and extends
+a Dump/meta-continuation.
+%subcode{opsem} include main
+% <P, F, c1 (+) c2 (L v), D> &|-->& <[], [], c1 v, (P,[](+)c2,F):D>
+% <P, F, c1 (+) c2 (R v), D> &|-->& <[], [], c2 v, (P,c1(+)[],F):D> 
+% <P', [], v c1, (P,[](+)c2,F):D> &|-->& < P, F, (L v) (P'[c1](+)c2{dagger}), D > 
+% <P', [], v c2, (P,c1(+)[],F):D> &|-->& <P, F, (R v) (c1{dagger}(+)P'[c2]), D>
+
+\item
+Similarly for products:
+%subcode{opsem} include main
+% <P, F, c1 (*) c2 (v1, v2), D> &|-->& <[], [], c1 v1, (P, [] (*) c2 v2,F):D>
+% <P', [], v1 c1, (P,[] (*) c2 v2,F):D> &|-->& <[], [], c2 v2, (P,v1 P'[c1] (*) [],F):D>
+% <P', [], v2 c2, (P,v1 c1 (*) [],F):D> &|-->& <P, F, (v1, v2) (c1(*)P'[c2]):D>
+and symmetrical rules for evaluation along the second branch. 
+%subcode{opsem} include main
+% <P', [], v2 c2, (P,c1 v1 (*) [],F):D> &|-->& <[], [], c1 v1, (P,[] (*) v2 P'[c2],F):D>
+% <P', [], v1 c1, (P,[] (*) v2 c2,F):D> &|-->& <P, F, (v1, v2) (P'[c1] (*) c2):D>
+
+The later two rules will be relevant only for reverse execution.
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%
+\section{Rules for {{eta}} and {{eps}} }
+
+The operation {{eps}} reverses the direction of a particle by reversing the world.
+
+\begin{itemize}
+\item Grammar
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v | -v
+% Isomorphisms, iso &=& ... | eta | eps
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
+
+\item
+Type judgement.
+%subcode{opsem} include main
+% eta &: 0 <-> (-b) + b :& eps
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+\item
+Operational Semantics.
+%subcode{opsem} include main
+% <P; F; eps (R v); D>      &|-->& <F; P; (L (-v))~ eps; D{dagger}>
+
+Note: there is NO reduction rule for {{eta}}.
+
+\item
+The adjoint of a Dump is defined to be:
+%subcode{opsem} include main
+% []{dagger} '= []
+% ((P, [](+)c, F):D){dagger} '= (F, [](+)c{dagger}, P):D{dagger}
+% ((P, c(+)[], F):D){dagger} '= (F, c{dagger}(+)[], P):D{dagger}
+% ((P, [](*)c v, F):D){dagger} '= (F, [](*)v c, P):D{dagger}
+% ((P, [](*)v c, F):D){dagger} '= (F, [](*)c v, P):D{dagger}
+% ((P, c v(*)[], F):D){dagger} '= (F, v c(*)[], P):D{dagger}
+% ((P, v c(*)[], F):D){dagger} '= (F, c v(*)[], P):D{dagger}
+
+\end{itemize}
 

### 2012-02-13 dual.tex
_

@@
 In both cases the seller receives the money immediately but there is a subtle
-difference. In the second transation, nobody has yet produced the money that
+difference. In the second transaction, nobody has yet produced the money that
 the seller has received: instead a \emph{debt} is generated and this debt
@@
 
-The othe way is to follow the execution. It goes forward in time so to speak,
+The other way is to follow the execution. It goes forward in time so to speak,
 comes back, and then goes forwards again.
@@
 
-\paragraph*{Filinski.}
+\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
+In his Masters thesis, Filinski proposes that continuations are a
+\emph{declarative} concept. He, furthermore, introduces a symmetric extension
+of the $\lambda$-calculus in which values and continuations are treated as
+opposites. This is essentially what we are proposing with one fundamental
+difference: our underlying language is not the $\lambda$-calculus but a
+language of pure isomorphisms in which information is preserved. This shift
+of perspective enables us to distill and generalize the duality of values and
+continuations: in particular, in the conventional $\lambda$-calculus setting
+values and continuations can be erased and duplicated which makes it
+difficult to maintain the correspondence between a value and its negative
+counterpart. In contrast, in our setting, one can start from the empty type
+$0$, introduce a positive value and its negative counterpart, and let each of
+these flow in arbitrary ways. The entire framework guarantees that neither
+the value nor its negative counterpart will be deleted or duplicated and
+hence that, in any closed program, the ``debt'' corresponding to the negative
+value is paid off exactly once.
+
+The forward and backward executions in our framework correspond to
+call-by-value and call-by-name. This duality was observed by Filinski and
+others following him but it is particularly clean in our framework.
 
@@
 
+\item Which operations can be lifted to work on negative types?
+
 \item These functions aren't really values. There is no value one can

### 2012-02-14 dual.tex
more constructions
_

--- 2012-02-13 dual.tex
@@
 
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
 \begin{center}
   \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+Double Negation
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+Negation distributes over {{+}}.
+
+{{-(b1 + b2) <-> (b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+Lifting negation out of {{*}}.
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/mult_neg.pdf}
 \end{center}
@@
 
-\item {{(-b1) * b2}} is not the same as {{b1 * (-b2)}} as there is no
-  isomorphism to translate from one to the other.
-
 \item {{-(b1 * b2)}} is not the same as {{b1 * b2}} and is also not
@@
 
-\item {{-(-b)}} is not the same as {{b}} as there is no isomorphism
-  that goes from one to the other.
-
 \item If we built an effect over {{langRevEE}}, say {{create}} and
   {{erase}}, are effects structured by an arrow or a monad now?
+
+\item Which operations can be lifted to work on negative types?
+
+\item Can the operational semantics for {{langRevEE}} be an
+  interpreter that is implemented in {{langRevT}} (similar to how the
+  the tree traversal interpreter was implemented). This would imply
+  the existence of a more powerful construction than Int, wherein the
+  products would also be preserved. This is possibly worth a paper in
+  itself.
 

### 2012-02-14 dual.tex
added composition and eps_fst
semantics fix

--- 2012-02-13 dual.tex
@@
 % D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
+% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
 
@@
 
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
 Delimited continuation
@@
 
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
 \begin{center}
@@
 \end{center}
+
+Double Negation
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+Negation distributes over {{+}}.
+
+{{-(b1 + b2) <-> (b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+{{eps_{fst} }}
+
+{{(-b1)*b2 + b1*b2 <-> 0}}
+
+\begin{center}
+  \includegraphics{diagrams/eps_fst.pdf}
+\end{center}
+
+Lifting negation out of {{*}}.
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/mult_neg.pdf}
+\end{center}
+
+The following isomoprhism can be constructed similarly
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
 
@@
 
-\item {{(-b1) * b2}} is not the same as {{b1 * (-b2)}} as there is no
-  isomorphism to translate from one to the other.
-
-\item {{-(b1 * b2)}} is not the same as {{b1 * b2}} and is also not
-  the same as {{(-b1) * (-b2)}} as there is no isomorphism that goes
-  from one to the other.
-
-\item {{-(-b)}} is not the same as {{b}} as there is no isomorphism
-  that goes from one to the other.
-
 \item If we built an effect over {{langRevEE}}, say {{create}} and
@@
 \item Which operations can be lifted to work on negative types?
+
+\item Can the operational semantics for {{langRevEE}} be an
+  interpreter that is implemented in {{langRevT}} (similar to how the
+  the tree traversal interpreter was implemented). This would imply
+  the existence of a more powerful construction than Int, wherein the
+  products would also be preserved. This is possibly worth a paper in
+  itself.
 

### 2012-02-14 dual.tex
added a couple of sentences in intro

@@
 that preserve information~\cite{infeffects}.
+
+For a while we didn't know how to interpret the pair $(2,-3)$. I was thinking
+that it corresponds to having two things: the value $2$ and the continuation
+$3$. This makes no sense because $(2,-3)$ is isomorphic to $-(2,3)$. Instead
+the right interpretation is that $(2,-3)$ corresponds to having 2
+$3$-continuations which is the same as having a $6$-continuations or 3
+$2$-continuations and so on.
+
+Since Filinski, we've had the idea that values and continuations are like
+mirror images. In a conventional language, the negative (continuation) side
+is implicit and we introduce information effects on the positive. Trying to
+recover the duality from this distorted positive side has always been
+messy. Now it looks clean because we have kept the positive side pure.
 

### 2012-02-15 dual.tex
_

--- 2012-02-14 dual.tex
@@
 
+{{eps_{fst} }}
+
+{{(-b1)*b2 + b1*b2 <-> 0}}
+
+\begin{center}
+  \includegraphics{diagrams/eps_fst.pdf}
+\end{center}
+
 Lifting negation out of {{*}}.
@@
   \includegraphics{diagrams/mult_neg.pdf}
+\end{center}
+
+The following isomoprhism can be constructed similarly
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
+
+Lifting a operation of postive types to negated types:
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
 \end{center}
@@
 \begin{itemize}
-
-\item {{-(b1 * b2)}} is not the same as {{b1 * b2}} and is also not
-  the same as {{(-b1) * (-b2)}} as there is no isomorphism that goes
-  from one to the other.
 

### 2012-02-15 dual.tex
intro

--- 2012-02-14 dual.tex
@@
 We introduced this thesis that computation should be based on isomorphisms
-that preserve information~\cite{infeffects}.
+that preserve information~\cite{infeffects}. Since Filinski, we've had the
+idea that values and continuations are like mirror images. In a conventional
+language, the negative (continuation) side is implicit and we introduce
+information effects on the positive. Trying to recover the duality from this
+distorted positive side has always been messy. Now it looks clean because we
+have kept the positive side pure.
+
+The way to think about something of type $A$ is that it is a value we have
+produced. The way to think about something of type $-A$ is that is a value we
+have already consumed.
+
+Other interpretations of the types of think about. The first one is
+arithmetic obviously. Another one is languages consisting of sets of
+string. The type 0 is the empty set, the type 1 is the set containing the
+empty word, the $+$ constructor corresponds to union, and the $*$ constructor
+corresponds to concatenation. The constructor $-$ would not correspond to set
+difference however. It would correspond to marking the elements in the set as
+``consumed'' so that if we take the union and a ``consumed'' element appears
+in the other set, the two cancel. This makes it clear that concatenating a
+produced $a$ and a consumed $b$ is not the same as concatenating a consumed
+$a$ and a produced $b$. They really need to be kept separate. Incidentally,
+division would be defined as follows:
+\[
+L_1 / L_2 = \{ x ~|~ xy \in L_1 \mbox{~for~some~} y \in L_2 \}
+\]
 
@@
 
+Lifting a operation of postive types to negated types:
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
 \paragraph*{Observability.} 

### 2012-02-15 iso-int/iso-int.tex
isomorphic interpreters

\documentclass{llncs}
\usepackage{url}
\usepackage{proof}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{graphicx}

\newcommand{\todo}[1]{\textbf{TODO: #1}}

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! |- = \vdash
%! in = \!\!\in\!\!
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! trace+ = trace
%! trace* = trace_{\times}
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! bullet = \bullet
%! * = \times

\urldef{\mails}\path|{rpjames, sabry}@indiana.edu|

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\title{On the Construction of Isomorphic Interpreters} 
\titlerunning{On the construction of Isomorphic Interpreters} 
\author{Roshan P. James \and Amr Sabry}
\institute{School of Informatics and Computing, Indiana University\\
\mails}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The constructions in this paper apply only to a class of interpreters
that can be represented by logically reversible small-step operational
semantics. While, in a formal sense, they are related to Abramsky's
``bi-orthogonal automata'' , we will try to give them a more direct
characterization purely in terms of logical reversibility of the
small-step semantics.

\section{Simple Bounded Iterator}
Consider definition for the following small-step abstract machine:

%subcode{bnf} include main
% Expressions, e = n
% Numbers, n, m = 0 | n + 1
%
% Machine states = <e, e>
% Start state = <n, 0>
% Stop State = <0, n>

Small-step Operational semantics:

%subcode{opsem} include main
% <n+1, m> |--> <n, m+1>

This machine start with a number {{n}} in the first position and at
each reduction step decrements the first number and increments the
second. It stops when the first number reaches 0, thereby taking
exactly {{n}} steps. While the machine does nothing interesting,
it useful to illustrate the general idea.

We can turn this abstract machine into a an isomorphic interpreter
implemented in {{langRevT}}, in a systematic way. To start with we
realize that the interpreter should implement the following map.

\begin{center}
  \includegraphics{diagrams/nat-nat1.pdf}
\end{center}

Now numbers can be implemented by the recursive type {{nat = @@x.1+x}}
in {{langRevT}}. This gives us the isomorphisms

\begin{center}
  {{unfold : nat <-> 1 + nat : fold}}
\end{center}

We start by examining the first value.

\begin{center}
  \includegraphics{diagrams/nat-nat2.pdf}
\end{center}

The triangle denotes the {{unfold}} isomorphism, which works as
follows: If the number {{n}} is zero, the output is the top branch
which has type {{1}}. If the number was non-zero, the output is on the
bottom branch and has value {{n-1}}.

Since the abstract machine increments second component, we know that
we need a {{fold}} on the second component to accomplish this. So we
can fill in the output part of the interpreter dually.

\begin{center}
  \includegraphics{diagrams/nat-nat3.pdf}
\end{center}

Let us examine all the possible input states, by distributing on the
input. Dually let us do the same with the output, but this time also
be a little bit explicit about the swap operations required so the
right types are in place.

\begin{center}
  \includegraphics{diagrams/nat-nat4.pdf}
\end{center}

The construction above already gives us some insight into how we can
implement our one reduction step. If we chose {{m = m'-1}} and we
chose {{n-1 =n'}} then we have essentially encoded the required
rewrite. Lets state this in words - the new value of {{n}}, namely
{{n'}}, is equal to {{n-1}}. And further - the new value of {{m}},
namely {{m'}}, is {{m+1}} (because {{m'-1=m}}).

We can thus connect the lower branches with the appropriate
isomorphism to reflect this relationship. In this case the isomorphism
is just a swap of the two wires involved.

\begin{center}
  \includegraphics{diagrams/nat-nat5.pdf}
\end{center}

We are a few steps away from an interpreter at this point and the
final steps rely on the answers to a few questions.

\begin{enumerate}
\item How do relate the outputs of the combinator to the inputs, such
  that the machine correctly takes multiple steps?

\item What do we do about the two inner branches?

\end{enumerate}

The branch labelled {{((), m)}} looks like the stop state of the
machine. The branch labelled {{((), n')}} looks like the start state
of the program. To complete the interpreter, we flip our current
construction inside out.

\begin{center}
  \includegraphics{diagrams/nat-nat6.pdf}
\end{center}

Which gives us

\begin{center}
  \includegraphics{diagrams/nat-nat7.pdf}
\end{center}

This basically completely our construction of the interpreter. We
basically need to use a {{trace}} to feedback related machine states
to related machine states.

\begin{center}
  \includegraphics{diagrams/nat-nat8.pdf}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Abstracting out the Construction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Adder and Multiplier}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tree Traversal}

\bibliographystyle{splncs03} 
\bibliography{cites}
\end{document}

### 2012-02-16 iso-int/iso-int.tex
_

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{What are the high level steps in the tranlsation?}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{What abstract machines can be turned into isomorphic interpreters?}
+
+If we had only the reduction relation {{<n+1> |--> <n>}}, can we
+construct an isomorphic interpreter? After all, every rewrite rule is
+logically reversible. In fact, this is not possible becuase there is
+no distinguished start state for such an interpreter and hence one
+does not know when to stop during reverse execution.
+
+Conditions that must be satisfied for constructing a isomorphic
+interpreter:
+
+\begin{enumerate}
+
+\item Distinguishable start and stop states. There may be more than one
+  valid start state and more than one valid stop state.
+
+\item Each valid machine state must match a unique reduction LHS. Each
+  valid machine state must machine a unique reduction RHS.
+
+\item Every reduction step must be computable (in {{langRevT}}) and
+  must be logically reversible -- i.e. it must be possible to reduce
+  from right to left.
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Adder and Multiplier}

### 2012-02-16 iso-int/iso-int.tex
adder

@@
 semantics. While, in a formal sense, they are related to Abramsky's
-``bi-orthogonal automata'' , we will try to give them a more direct
-characterization purely in terms of logical reversibility of the
-small-step semantics.
+``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
+to give them a more direct characterization purely in terms of logical
+reversibility of the small-step semantics.
 
@@
 
+\begin{enumerate}
+\item 
 We can turn this abstract machine into a an isomorphic interpreter
@@
 
+\item
 The triangle denotes the {{unfold}} isomorphism, which works as
@@
 
+\item
 Let us examine all the possible input states, by distributing on the
@@
 
+\item
 The construction above already gives us some insight into how we can
@@
 
+\item
 We are a few steps away from an interpreter at this point and the
@@
 
-The branch labelled {{((), m)}} looks like the stop state of the
-machine. The branch labelled {{((), n')}} looks like the start state
+The branch labeled {{((), m)}} looks like the stop state of the
+machine. The branch label-led {{((), n')}} looks like the start state
 of the program. To complete the interpreter, we flip our current
@@
 
+\end{enumerate}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{What are the high level steps in the tranlsation?}
+\subsection{What are the high level steps in the translation?}
+\label{sec:steps}
+
+TODO
+
+\begin{enumerate}
+\item Expand the input such that every input term that the LHS of the
+  interpreter pattern matches on is exposed.
+
+\item Expand the output such that every output term that it
+  reconstructed by the interpreter is exposed.
+
+\item Shuffle matching LHS term to matching RHS terms, inserting any
+  appropriate mediating computations. In the previous example, this
+  was only a swap operation.
+
+\end{enumerate}
 
@@
 
+Logically reversible abstract machines with distinguishable start and
+stop states can be converted into isomorphic interpreters in
+{{langRevT}}.  To develop an understanding for what sorts of machines
+these are, let us look at some examples.
+
+Consider a machine with the following two reduction steps
+{{<n+1,m>|--> <n, m+1>}} and {{<n+1,m>|--><n,m>}}. Can we create an
+isomorphic interpreter for such a machine? We can't, because for any
+given machine state, it is not obvious what reduction rule to
+apply. We could reverse this machine if there was a clear
+non-ambiguity in which machine rule was chosen.
+
+Consider a machine, with the following rewrite rules
+{{<n+1,m,p>|--><n,m,p+1>}} and {{<0,m+1,p>|--><0,m,p+1>}}. Can we
+create an isomorphic interpreter for this machine? It turns out we
+can't because for any given output state {{<0,m,p+1>}} it is not
+obvious which input state it came from - i.e. there are two
+possibilities {{<1,m,p>}} or there is {{<0,m+1,p>}}.
+
+Consider a machine that has the addition rule {{<''x+y''>|--><''z''>}}
+where {{z}} is the sum of {{x}} and {{y}}. This machine again cannot
+be turned into an isomorphic interpreter because the one reduction
+rule does not have a logically inverse -- in other words, given a
+{{z}}, we are unable to determine which {{x}} and {{y}} it originated
+from.
+
 If we had only the reduction relation {{<n+1> |--> <n>}}, can we
 construct an isomorphic interpreter? After all, every rewrite rule is
-logically reversible. In fact, this is not possible becuase there is
+logically reversible. In fact, this is not possible because there is
 no distinguished start state for such an interpreter and hence one
@@
 
+\noindent
 Conditions that must be satisfied for constructing a isomorphic
@@
 
+\subsection{Logically Reversible Adder}
+
+%subcode{bnf} include main
+% Expressions, e = n
+% Numbers, n, m,p = 0 | n + 1
+%
+% Machine states = <e, e, e>
+% Start state = <n, p, 0>
+% Stop State = <0, p', m>
+
+Small-step Operational semantics:
+
+%subcode{opsem} include main
+% <n+1, p, m> |--> <n, p+1, m+1>
+
+This is an ostensibly useful machine in the sense that it computes the
+sum of its two inputs.  This construction proceeds exactly like the
+previous one till we have the following situation:
+
+\begin{center}
+  \includegraphics{diagrams/adder1.pdf}
+\end{center}
+
+The equalities we would like to express at this point are {{n-1=n'}},
+{{m=m'-1}} and that {{p+1=p'}}. Fortunately {{langRevT}} provides the
+{{add1: nat <-> nat}} combinator and so we could just use that to
+mediate between {{p}} and {{p'}}. This is one solution and this gives
+us:
+
+\begin{center}
+  \includegraphics{diagrams/adder2.pdf}
+\end{center}
+
+This can be inverted as before, to give us an interpreter. While this
+is a correct implementation of the small-step semantics, it is
+somewhat unsatisyfing for the reason that it does not see very
+systematic. After all we chose to use an {{add1}} for {{p}} -- why
+didn't we do the same for {{m}}? More importantly, the fact that we
+chose used {{add1}} is also hiding something -- namely that {{add1}}
+is a partial isomorphism. Its inverse is not defined when applied to
+{{0}} i.e. it will go into an infinite loop.
+
+In a sense this implicit assymetry of the abstract machine was being
+hidden by the {{add1}} operation. So instead of chosing to use
+{{add1}} let us chose to expand out {{p'}}, as recommended by step 2
+in section \ref{sec:steps}. This gives us the following:
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/adder3.pdf}
+}
+\end{center}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Tree Traversal}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A {{langRev}} Interpreter}
 

### 2012-02-16 iso-int/iso-int.tex
adder

@@
 
+We can very that the connected wires do indeed give us teh desired
+equalities between the inputs and outputs. Also we recognize our
+expected start state and expected stop state. What is the dangling
+state?
+
+The dangling state was what has hidden by the use of {{add1}} and
+denotes the fact that the abstract machine did not define any valid
+backward reduction for the following valid machine state
+{{<n,0,m+1>}}. To take a backward step, both {{m}} and {{p}} have to
+be simulatenously decremented and this is not possible if {{p}} is
+{{0}}.  At this point we have two choices:
+
+\begin{enumerate}
+\item We can hide the use the combinator {{omega : b <-> 0}}
+  expressible in {{langRevT}}. This combinator is basically an
+  infinote loop on all inputs and hence would allow us to entirely
+  discard one possible branch. In other words, we are explicitely
+  admitting that our machine is undefined on some inputs.
+
+\item We can admit that there are two possible start states to this
+  abstract machine {{<n, m, 0>}} and {{<n,0,m>}}. One of this is an
+  artifact of the fact that backward execution encounters stuck
+  states. Dually, if there were stuck states for forward execution,
+  they would be reflected in the fact that there were multiple stop
+  states -- only one of which denotes a desirable or correct execution
+  of the machine.
+
+\end{enumerate}
+
+As we did before, we flip inside out the construction of the map and
+apply trace to give us the interpreter:
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/adder4.pdf}
+}
+\end{center}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-02-16 iso-int/iso-int.tex
spell check

@@
 is a correct implementation of the small-step semantics, it is
-somewhat unsatisyfing for the reason that it does not see very
+somewhat unsatisfying for the reason that it does not see very
 systematic. After all we chose to use an {{add1}} for {{p}} -- why
@@
 
-In a sense this implicit assymetry of the abstract machine was being
-hidden by the {{add1}} operation. So instead of chosing to use
+In a sense this implicit asymmetry of the abstract machine was being
+hidden by the {{add1}} operation. So instead of choosing to use
 {{add1}} let us chose to expand out {{p'}}, as recommended by step 2
@@
 
-We can very that the connected wires do indeed give us teh desired
+We can very that the connected wires do indeed give us the desired
 equalities between the inputs and outputs. Also we recognize our
@@
 {{<n,0,m+1>}}. To take a backward step, both {{m}} and {{p}} have to
-be simulatenously decremented and this is not possible if {{p}} is
+be simultaneously decremented and this is not possible if {{p}} is
 {{0}}.  At this point we have two choices:
@@
   expressible in {{langRevT}}. This combinator is basically an
-  infinote loop on all inputs and hence would allow us to entirely
-  discard one possible branch. In other words, we are explicitely
+  infinite loop on all inputs and hence would allow us to entirely
+  discard one possible branch. In other words, we are explicitly
   admitting that our machine is undefined on some inputs.
@@
 
-As we did before, we flip inside out the construction of the map and
-apply trace to give us the interpreter:
+As we did before, we flip inside out the construction so far and
+apply {{trace}} to get the isomorphic interpreter:
 

### 2012-02-16 iso-int/iso-int.tex
diagrams - added boxes to indicate what is swapped
added section on tree traversal

@@
 %subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
+%! [^ = \ulcorner
+%! ^] = \urcorner
+%! [v = \llcorner
+%! v] = \lrcorner
 %! |-->* = \mapsto^{*}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Logically Reversible Multiplier}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Tree Traversal}
+
+Let us take the ideas developed so far and apply them to developing a
+tree traversal interpreter. We define a typical binary tree, one that
+has numbers as leaf values and also define a notion of tree
+contexts. The tree contexts serve to track which subtree is currently
+being explored and are similar to tree zippers CITE.
+
+%subcode{bnf} include main
+% Tree, t = L n | N t t
+% Tree Contexts, c = Lft c t | Rgt t c | []
+%
+% Machine states = <t, c> | {[c, t]}
+% Start state = <t, []>
+% Stop State = {[[], t]}
+
+We define an abstract machine whose task is to traverse a given tree
+and increment every leaf value. Small-step Operational semantics:
+
+%subcode{opsem} include main
+% <L n, c> &|-->& {[c, L (n+1)]}
+% <N t1 t2, c> &|-->& <t1, Lft c t2>
+% {[Lft c t2, t1]} &|-->& <t2, Rgt t1 c>
+% {[Rgt t1 c, t2]} &|-->& {[N t1 t2, c]}
+
+The machine here is a little richer than the ones dealt with
+previously. In particular, we have a two types of machine states --
+{{<t,c>}} and {{ {[t, c]} }}. One of these corresponds to walkking
+down a tree, building up the context in the process. The other
+corresponds to reconstructing the tree from the context and also
+switching focus to any unexplored subtrees in the process. There are
+also two syntactic categories to deal with -- trees and tree context
+-- where previously we only had numbers. The {{fold}} and {{unfold}}
+isomorphisms that we need for trees and tree contexts are:
+
+%subcode{bnf} include main
+%! columnStyle = rrcll
+% unfold_t :& ~~~~t &<->& n + t * t &: fold_t
+% unfold_c :& c &<->& 1 + c * t + t * c ~~~~~~&: fold_c
+
+However we will see that the previously developed systematic approach
+developing {{langRevT}} interpreters carries over to this abstract
+machine as well.  As the first step, we quickly check that the
+abstract machine however has disstinguished start and stop states, has
+no ambiguities in reduction and that each reduction step is logically
+reversible.
+
+\subsection{New Notation}
+
+Before we start on the actual construction, let us introduce a
+syntactic convenience. In the past interpreter we explicitely drew out
+every {{fold}} and {{unfold}} followed/preceded immediately by any
+required swaps and then a {{distribute}} or {{factor}}. In this
+presentation we will combine all of these into one construct which we
+will draw as a thin vertical rectangle. Also we will introduce the
+convention that the component that is being expanded or collapsed will
+be marked by using {{[^ ^]}} and the components that are being
+generated will be marked by {{[v v]}}.
+
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/syntax1.pdf}
+}
+\end{center}
+
+The above shows the expansion of a {{tree*nat}} type. First the
+{{tree}} component is expanded and in one of the generated branches
+the {{nat}} component is expanded.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Isomorphic Interpreter}
+
+The first step to developing teh isomorphic interpreter is to
+recognize that the two possible kinds of machine states simply hide an
+implicit {{bool}}. We make this explicit.
+
+%subcode{opsem} include main
+% columnStyle = rclr
+% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
+% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
+% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
+% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
+
+And so machine states have the form:
+
+%subcode{bnf} include main
+% Machine states = <bool, t, c>
+% Start state = <T, t, []>
+% Stop state =  <F, t, []>
+
+And we start examining the machine components as before:
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/tree1.pdf}
+}
+\end{center}
+
+On the LHS, for {{T}} states, we have expanded the tree component and
+for {{F}} states we have expanded the tree contexts. Similarly we have
+done the opposite on the RHS side exactly matching up what the
+abstract machine does. One thing to note is that we dropped the {{1}}
+introduced by expanding the {{bool}} and instead just labelled the
+{{true}} and {{false}} branches.
+
+We are ready to start connecting the machine states corresponding to
+the reductions that we would like.
+
+\begin{enumerate}
+
+\item 
+When we encounter a leaf we would like to increment its value and move
+to the corresponding {{F}} machine state. For the sake of simplicity
+in this interpreter we won't be concerned with exposing stuck states
+on reverse execution caused by {{sub1 0}}. So we are free to use the
+{{add1}} combinator.
+
+\item
+For all the other reduction rules, it is a straight forward mapping of
+related states following the reduction rules.
+
+\end{enumerate}
+
+The required reduction rules have been labeled on the construction
+belong. We have also included subscripts on the various {{t}}s to
+indicate any swaps that may need to be inserted.
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/tree2.pdf}
+}
+\end{center}
+
+This essentially completes the construction of the (partially)
+isomorphic tree-traversal interpreter. (Partially because of the
+implicit non-termination introduced by the {{add1}}).
 

### 2012-02-16 iso-int/iso-int.tex
_

@@
 
+Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
+properties and various constructions in it. However, for a
+computational model, there has been little focus on actually writing
+programs directly in {{langRevT}}. This paper aims to remedy that by
+presenting a systematic technique of expressing logically reversible
+small-step abstract machines in {{langRevT}}. This paper shows that
+once we have a logically reversible machine in a notation of our
+choice, expressing the machine as an isomorphic interpreter in
+{{langRevT}} can be done systematically and does not present any
+significant conceptual difficulties.
+
 The constructions in this paper apply only to a class of interpreters
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Simple Bounded Iterator}

### 2012-02-17 iso-int/iso-int.tex
pi

@@
 
+%subcode{bnf} include main
+% Combinators, c = iso | c (;) c | c (*) c | c (+) c
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+% Values, v = () | (v, v) | L v | R v
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+%subcode{opsem} include main
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+
+TODO : The real difficulty with this interpreter is to argue why it is
+not full of undefined states. For instance, why the machines will
+never reach stuck states such as {{<c1(+)c2, (), ..>}}.
+
+The right way out of that is to make the type system of the abstract
+machine rich enough that the structure of the abstract automatically
+rules out this sort of thing.
+
 \bibliographystyle{splncs03} 

### 2012-02-17 iso-int/iso-int.tex
doing pi is difficult - for the wrong reasons

@@
 not full of undefined states. For instance, why the machines will
-never reach stuck states such as {{<c1(+)c2, (), ..>}}.
+never reach stuck states such as {{<c1(+)c2, (), ..>}}.  The right way
+out of that is to make the type system of the abstract machine rich
+enough that the structure of the abstract automatically rules out this
+sort of thing.
 
-The right way out of that is to make the type system of the abstract
-machine rich enough that the structure of the abstract automatically
-rules out this sort of thing.
+%subcode{bnf} include main
+% Combinators, c(t1<->t2) = 
+%             &|& c(t<->t) = iso 
+%             &|& c(t1<->t2) = c(t1<->t3) (;) c(t3<->t2) 
+%             &|& c(t1*t2<->t3*t4) = c(t1<->t3) (*) c(t2<->t4) 
+%             &|& c(t1+t2<->t3+t4) = c(t1<->t3) (+) c(t2<->t4)
+%
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+%
+% Type Indexed Values, v(t) =
+%  &|& v(1) = () 
+%  &|& v(t1*t2) = (v(t1), v(t2)) 
+%  &|& v(t1+t2) = L v(t1) | R v(t2)
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+However this begs the question - does {{langRevT}} need all these GADT
+types to reason about all these things? Something to think about. If
+we do, then thereare two ways out
+
+(1) add GADTs to {{langRevT}} and keep the structure of things
+simple. They can always be added opaquely for any given types -- we
+only need to add the proper fold/unflod morphisms for those types
+(maybe).
+
+(2) compile down to {{langRevT}} as it is. In this case, {{langRevT}}
+is indeed a relatively less typed language than the source language
+and will hence have to treat as stuck-states the non-type correct
+machine states. We could apply omega to them and make them vanish --
+for example. This is similar to the handling of the {{sub1}} in the
+previous example.
 

### 2012-02-18 dual.tex

@@
 
-\title{Continuations via Negative Information Flow}
+\title{Functional Pearl: \\ Programming with Negative and Fractional Types} 
 \authorinfo{Roshan P. James}
@@
 \begin{abstract}
-We show a deep symmetry between functions and delimited continuations,
-values and continuations that arises in {{langRev}} in a manner that
-is reminiscent of Filinksi's Symmetric \lcal
-~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
-extending {{langRev}} with a notion of additive duality over the
-monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
-Compact Closed Categories. The resulting dual types, which we denote
-{{-b}}, have a time traveling ``backward information flow''
-interpretation and allow for the encoding of higher-order function and
-iteration via the construction of {{trace}} operators, thereby making
-the extended language {{langRevEE}} a Turing-complete reversible
-programming language with higher-order functions and first-class
-delimited continuations.
+  Every functional programmer knows about sum and product types, $a+b$ and
+  $a*b$ respectively. Negative and fractional types, $a-b$ and $a/b$
+  respectively, are much less known and their computational interpretation is
+  unfamiliar and often complicated. We show that in a programming model in
+  which information is preserved (such as the model introduced in our recent
+  paper on \emph{Information Effects}), negative and fractional types have
+  particularly intuitive and natural computational
+  interpretations. Intuitively, values of negative types are values that flow
+  ``backwards'' to satisfy demands, and values of fractional types are values
+  that represent first-class pattern clauses. The combination of negative and
+  fractional types allow a programmer to express a wide range of programming
+  idioms, including higher-order functions, delimited continuations,
+  speculative computation, logic programming, and more. 
 \end{abstract}
@@
 \section{Introduction}
+
+We show a deep symmetry between functions and delimited continuations,
+values and continuations that arises in {{langRev}} in a manner that
+is reminiscent of Filinksi's Symmetric \lcal
+~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
+extending {{langRev}} with a notion of additive duality over the
+monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
+Compact Closed Categories. The resulting dual types, which we denote
+{{-b}}, have a time traveling ``backward information flow''
+interpretation and allow for the encoding of higher-order function and
+iteration via the construction of {{trace}} operators, thereby making
+the extended language {{langRevEE}} a Turing-complete reversible
+programming language with higher-order functions and first-class
+delimited continuations.
 
@@
 
-\paragraph*{Ariola, Herbelin, and Sabry.} We use subtractive types. Things
-are constrained so that the positive and negative values must be used
-together. In our new work we have no such constrained, the negative value can
-flow anywhere. The entire system guarantees that any closed program would
-have to account for it. We don't have to introduce special constraints to
-achieve that.
+\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
+subtraction type because its ``computational interpretation is not
+familiar.'' Curien and Herbelin study duality in classical logic, show that
+it exchanges call-by-value and call-by-name. They extend classical natural
+deduction with subtraction but give it no computational meaning. Crolard (in
+the formulae-as-types interpretation of subtractive logic) address the
+computational interpretation of substraction. He explains the type $A-B$ as
+the of \emph{coroutines} with a local environment of type $A$ and a
+continuation of type $B$. The description is complicated by what is
+essentially the desire to enforce linearity constraints so that coroutines
+cannot access the local environment of other coroutines. Must cite Selinger
+control categories in this context of duality but I am not sure what to say:
+it assumes cartesian closed categories for one thing and not symmetric
+monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
+explain delimited continuations.
 
-\paragraph*{Focusing, Polarization, etc.}
+\paragraph*{Linear Logic} In accounts that are linear, the value and
+continuation that comprise the substractive type need to be constrained to
+``stay together.'' This can be achieved by various restrictions. In this work
+we have no such constraints, the negative value can flow anywhere. The entire
+system guarantees that any closed program would have to account for it. We
+don't have to introduce special constraints to achieve that. Zeilberger in
+the paper on polarity and the logic of delimited continuations uses polarized
+logic: he shows that if positive and negative values are completely symmetric
+except that answer types are positive, then the framework accommodates
+delimited continuations. But he interprets negative values are control
+operators, or as values defined by the shape of their continuations. We
+simply interpret values of negative type as values flowing in the ``other''
+direction.
 

### 2012-02-18 PiNF.hs
some useless code

@@ -1,8 +1,19 @@
-{-# OPTIONS_GHC -XGADTs -XTypeOperators -XEmptyDataDecls #-}
+{-# OPTIONS_GHC 
+  -XGADTs 
+  -XTypeOperators 
+  -XEmptyDataDecls 
+  -XRankNTypes 
+  -XNamedFieldPuns 
+#-}
 
 -- Pi with negative and fractional types
 
 module PiNF where
+
+import Control.Monad
+import qualified Control.Monad.State as ST
+import Data.Dynamic
+import Data.Maybe
 
 -----------------------------------------------------------------------
 -- Morphisms for symmetric monoidal structure extended with trace
@@ -10,6 +21,7 @@ module PiNF where
 data Zero 
 data Minus a = Minus a
 data Recip a = Recip a
+data LVar = LVar Integer
 
 absurd :: Zero -> a
 absurd _ = undefined
@@ -123,9 +135,162 @@ adjoint EtaTimes = EpsilonTimes
 adjoint EpsilonTimes = EtaTimes
 
 -- Semantics
--- forward
 
-(@@>) :: (a :<=> b) -> a -> b
+-- applying basic combinators
+(@@) :: (a :<=> b) -> a -> b
+PlusZeroL @@ (Right a) = a
+PlusZeroR @@ a = Right a
+CommutePlus @@ (Left a) = Right a
+CommutePlus @@ (Right b) = Left b 
+AssocPlusL @@ (Left a) = Left (Left a) 
+AssocPlusL @@ (Right (Left b)) = Left (Right b) 
+AssocPlusL @@ (Right (Right c)) = Right c
+AssocPlusR @@ (Left (Left a)) = Left a
+AssocPlusR @@ (Left (Right b)) = Right (Left b)
+AssocPlusR @@ (Right c) = Right (Right c)
+TimesOneL @@ ((), a) = a
+TimesOneR @@ a = ((), a)
+CommuteTimes @@ (a,b) = (b,a) 
+AssocTimesL @@ (a,(b,c)) = ((a,b),c) 
+AssocTimesR @@ ((a,b),c)  = (a,(b,c))
+Distribute @@ (Left b, a) = Left (b, a) 
+Distribute @@ (Right c, a) = Right (c, a) 
+Factor @@ (Left (b, a)) = (Left b, a) 
+Factor @@ (Right (c, a)) = (Right c, a) 
+Id @@ a = a
[…]

### 2012-02-19 PiNF.hs
removed code that i started because roshan has done it already

@@ -1,476 +1,0 @@
-{-# OPTIONS_GHC 
-  -XGADTs 
-  -XTypeOperators 
-  -XEmptyDataDecls 
-  -XRankNTypes 
-  -XNamedFieldPuns 
-#-}
-
--- Pi with negative and fractional types
-
-module PiNF where
-
-import Control.Monad
-import qualified Control.Monad.State as ST
-import Data.Dynamic
-import Data.Maybe
-
------------------------------------------------------------------------
--- Morphisms for symmetric monoidal structure extended with trace
-
-data Zero 
-data Minus a = Minus a
-data Recip a = Recip a
-data LVar = LVar Integer
-
-absurd :: Zero -> a
-absurd _ = undefined
-
-data a :<=> b where 
--- (+) is associative, commutative, and has a unit
-  PlusZeroL   :: Either Zero a :<=> a
-  PlusZeroR   :: a :<=> Either Zero a
-  CommutePlus :: Either a b :<=> Either b a
-  AssocPlusL  :: Either a (Either b c) :<=> Either (Either a b) c 
-  AssocPlusR  :: Either (Either a b) c :<=> Either a (Either b c) 
--- (*) is associative, commutative, and has a unit
-  TimesOneL    :: ((), a) :<=> a
-  TimesOneR    :: a :<=> ((), a)
-  CommuteTimes :: (a,b) :<=> (b,a) 
-  AssocTimesL  :: (a,(b,c)) :<=> ((a,b),c)
-  AssocTimesR  :: ((a,b),c) :<=> (a,(b,c))
--- (*) distributes over (+) 
-  TimesZeroL  :: (Zero, a) :<=> Zero
-  TimesZeroR  :: Zero :<=> (Zero, a)
-  Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
-  Factor      :: Either (b, a) (c, a) :<=> (Either b c, a)
--- Congruence
-  Id    :: a :<=> a
-  Sym   :: (a :<=> b) -> (b :<=> a) 
-  (:.:) :: (b :<=> c) -> (a :<=> b) -> (a :<=> c)
-  (:+:) :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
-  (:*:) :: (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
--- Trace
[…]

### 2012-02-19 dual.tex
intuition

@@
 
-\title{Functional Pearl: \\ Programming with Negative and Fractional Types} 
+\title{Functional Pearl \\ Programming with Negative and Fractional Types} 
 \authorinfo{Roshan P. James}
@@
   ``backwards'' to satisfy demands, and values of fractional types are values
-  that represent first-class pattern clauses. The combination of negative and
+  that represent first-class constraints. The combination of negative and
   fractional types allow a programmer to express a wide range of programming
@@
 
-\keywords Arrows, Linear logic, Quantum computing, Reversible logic.
+\keywords continuations, information flow, linear logic, logic programming,
+quantum computing, reversible logic, symmetric monoidal categories, compact
+closed categories.
 
@@
 
+Connections to type logical grammars and Lambek calculus
+
+Some background on {{langRev}} because it is mentioned in the next section.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Value Producers and Consumers}
+\section{Negative Types: Intuition}
 
-Consider the following ways of purchasing an item that costs \$20.00:
+Consider the following two ways of purchasing an item that costs \$20.00:
+\begin{enumerate}
+\item You give the seller a \$20.00 bill.
+\item You use a credit card to pay the seller.
+\end{enumerate}
+In both cases, the seller receives the money immediately but there is a
+subtle difference. In the first transaction, the money received by the seller
+corresponds to a value that has been produced earlier in the computation. In
+the second transaction, the money may or may not exist yet: computationally,
+a \emph{debt} is generated to compensate for the money received by the seller
+and this debt travels ``backwards'' towards the bank and is (hopefully)
+reconciled there.
 
-\begin{enumerate}
-\item You produce a \$20.00 bill and give it to the seller.
-\item You use a credit card. 
-\end{enumerate}
-In both cases the seller receives the money immediately but there is a subtle
-difference. In the second transaction, nobody has yet produced the money that
-the seller has received: instead a \emph{debt} is generated and this debt
-travels ``backwards'' towards the bank and (hopefully) reconciled there.
+The example suggests that the ability to consider values flowing backwards
+enriches our computational model. This observation goes back to Filinski's
+Masters thesis where continuations are identified with these negative values.
+We discuss the connections to Filinski's work and others in more detail in
+Sec.~\ref{sec:related}. For now, we note that in a conventional programming
+language in which values can be copied and deleted at will, extreme care is
+needed to keep track of negative values. Indeed, in the example above, if it
+were possible to simply delete the variable corresponding to the debt, we
+would have produced money out of nothing.
 
-Computationally we model the first transaction using essentially the identity
-function which receives a \$20.00 bill as input from the buyer and propagates
-it on its output to the seller. The second transaction is more
-complicated. We model it as shown in the circuit below:
+For this reason, our treatment of negative values in the context of
+{{langRev}} will be particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative value) will eventually have to match the negative value
+with some corresponding value, effectively using the isomorphism in the
+right-to-left direction. In more detail, we can model the credit card
+transaction above using the following program (shown diagrammatically):
 
@@
 
-There are two ways to understand this circuit that are both quite
-instructive. Let's first examine the type structure of each of the
-combinators that comprise the circuit. The first combinator on the right
-outputs \$20.00 to the seller. This \$20.00 is produced from nothing so to
-speak by generating an equivalent debt that travels backwards. COMPLETE BASED
-ON THE FIGURE.
+CLEAN UP the following based on the figure. In contrast, in our setting, one
+can start from the empty type $0$, introduce a positive value and its
+negative counterpart, and let each of these flow in arbitrary ways. The
+entire framework guarantees that neither the value nor its negative
+counterpart will be deleted or duplicated and hence that, in any closed
+program, the ``debt'' corresponding to the negative value is paid off exactly
+once. Computationally we model the first transaction using essentially the
+identity function which receives a \$20.00 bill as input from the buyer and
+propagates it on its output to the seller. The second transaction is more
+complicated. We model it as shown in the circuit below: There are two ways to
+understand this circuit that are both quite instructive. Let's first examine
+the type structure of each of the combinators that comprise the circuit. The
+first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
+produced from nothing so to speak by generating an equivalent debt that
+travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
+the execution. It goes forward in time so to speak, comes back, and then goes
+forwards again.
 
-The other way is to follow the execution. It goes forward in time so to speak,
-comes back, and then goes forwards again.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Fractional Types: Intuition}
 
-\paragraph*{Continuations.} The idea of using negative types to model
-information flowing backwards, demand for values, continuations, etc. goes
-back to at Filinski's thesis. We recall these connections below but we first
-note that all these systems are complicated because in all these systems
-information can be ignored, destroyed, or duplicated. Clearly the possibility
-of erasure of information would mean that our credit card transaction is
-incorrect. In our work, information is maintained and hence we have a
-guarantee that, in a closed program, the debt must be accounted and paid for.
+The type $a/b$ is prominent in the Lambek-Grishin calculus which is
+extensively used in computational linguistics. In the common interpretation,
+a value of type $a/b$ is a value of some type $c$ such that when put in a
+context of type $b$, the result is a value of type $a$. For example, assuming
+contexts are represented as functions, a value of type $\texttt{bool} /
+(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
+$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
+in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
+type $\texttt{bool}$. More generally, one can think of a value of type $a/b$
+as a value that imposes a constraint on its context of use: it is a value
+that requires its context to be of type $b$ and only if that condition is
+satisfied, can the value be considered as a value of type $a$. In the
+simplest case, the type $1/b$ just constraints the context to be of type $b$.
 
-\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
-In his Masters thesis, Filinski proposes that continuations are a
-\emph{declarative} concept. He, furthermore, introduces a symmetric extension
-of the $\lambda$-calculus in which values and continuations are treated as
-opposites. This is essentially what we are proposing with one fundamental
-difference: our underlying language is not the $\lambda$-calculus but a
-language of pure isomorphisms in which information is preserved. This shift
-of perspective enables us to distill and generalize the duality of values and
-continuations: in particular, in the conventional $\lambda$-calculus setting
-values and continuations can be erased and duplicated which makes it
-difficult to maintain the correspondence between a value and its negative
-counterpart. In contrast, in our setting, one can start from the empty type
-$0$, introduce a positive value and its negative counterpart, and let each of
-these flow in arbitrary ways. The entire framework guarantees that neither
-the value nor its negative counterpart will be deleted or duplicated and
-hence that, in any closed program, the ``debt'' corresponding to the negative
-value is paid off exactly once.
+By allowing the fractional types to be first-class, we can separate the
+generation of constraints from their use. For example, consider a variation
+on the example in the previous where the parent gives a child the \$20.00
+bill but requires that the child only spends the money on buying books. This
+can achieved by producing a constraint that can only be resolved if the child
+indeed buys a book. Technically, this is achieved by adding a new type
+isomorphism {{ 1 <-> a * (1/a) }} which when used in the left-to-right
+direction allows the creation of a value and a corresponding constraint. Both
+the value and the constraint can flow in arbitrary ways during the
+computation but eventually in a closed program with no ``dangling''
+constraints, the isomorphism should be used in the right-to-left direction to
+resolve the constraints.
 
-The forward and backward executions in our framework correspond to
-call-by-value and call-by-name. This duality was observed by Filinski and
-others following him but it is particularly clean in our framework.
-
-\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
-subtraction type because its ``computational interpretation is not
-familiar.'' Curien and Herbelin study duality in classical logic, show that
-it exchanges call-by-value and call-by-name. They extend classical natural
-deduction with subtraction but give it no computational meaning. Crolard (in
-the formulae-as-types interpretation of subtractive logic) address the
-computational interpretation of substraction. He explains the type $A-B$ as
-the of \emph{coroutines} with a local environment of type $A$ and a
-continuation of type $B$. The description is complicated by what is
-essentially the desire to enforce linearity constraints so that coroutines
-cannot access the local environment of other coroutines. Must cite Selinger
-control categories in this context of duality but I am not sure what to say:
-it assumes cartesian closed categories for one thing and not symmetric
-monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
-explain delimited continuations.
-
-\paragraph*{Linear Logic} In accounts that are linear, the value and
-continuation that comprise the substractive type need to be constrained to
-``stay together.'' This can be achieved by various restrictions. In this work
-we have no such constraints, the negative value can flow anywhere. The entire
-system guarantees that any closed program would have to account for it. We
-don't have to introduce special constraints to achieve that. Zeilberger in
-the paper on polarity and the logic of delimited continuations uses polarized
-logic: he shows that if positive and negative values are completely symmetric
-except that answer types are positive, then the framework accommodates
-delimited continuations. But he interprets negative values are control
-operators, or as values defined by the shape of their continuations. We
-simply interpret values of negative type as values flowing in the ``other''
-direction.
-
-\paragraph*{Int Construction.}
-For a traced monoidal category {{C}} the Int construction produces a
-Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
-Further we know that the target of the Int construction is isomorphic
-to the target of \G construction of Abramsky \cite{Abramsky96:0} from
-Haghverdi. However, note that the {{langRevEE}} is not the same as the
-image of the Int construction on {{langRevT}}, since the the later
-lacks a multiplicative tensor that distrbutes over the additive tensor
-in Int {{langRevT}}.
-
-\paragraph*{Recursion.} In our previous work we introduce recursive types and
-trace operators. This is dangerous here because infinite loops allow us to
-prolong paying the debt for as long as we want.
-
-\paragraph*{GoI machines} 
-We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
+NEED to work out the example in detail (or perhaps another example that's
+better. Not sure if this one makes sense because we have to create the book
+and the context at once.???
 
@@
 %%%%%%%%%%%%%%%%%
-\section{Rules for {{eta}} and {{eps}} }
+\subsection{Rules for {{eta}} and {{eps}} }
 
@@
 
-\section{Diagrams}
+%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Diagrams}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Multiplicative Duality in {{langRev}} }
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related Work} 
+\label{sec:related}
+
+Filinski proposes that continuations are a \emph{declarative} concept. He,
+furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+which values and continuations are treated as opposites. This is essentially
+what we are proposing with one fundamental difference: our underlying
+language is not the $\lambda$-calculus but a language of pure isomorphisms in
+which information is preserved. This shift of perspective enables us to
+distill and generalize the duality of values and continuations: in
+particular, in the conventional $\lambda$-calculus setting values and
+continuations can be erased and duplicated which makes it difficult to
+maintain the correspondence between a value and its negative counterpart.
+
+\paragraph*{Continuations.} The idea of using negative types to model
+information flowing backwards, demand for values, continuations, etc. goes
+back to at Filinski's thesis. We recall these connections below but we first
+note that all these systems are complicated because in all these systems
+information can be ignored, destroyed, or duplicated. Clearly the possibility
+of erasure of information would mean that our credit card transaction is
+incorrect. In our work, information is maintained and hence we have a
+guarantee that, in a closed program, the debt must be accounted and paid for.
+
+\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
+In his Masters thesis, Filinski proposes that continuations are a
+\emph{declarative} concept. He, furthermore, introduces a symmetric extension
+of the $\lambda$-calculus in which values and continuations are treated as
+opposites. This is essentially what we are proposing with one fundamental
+difference: our underlying language is not the $\lambda$-calculus but a
+language of pure isomorphisms in which information is preserved. This shift
+of perspective enables us to distill and generalize the duality of values and
+continuations: in particular, in the conventional $\lambda$-calculus setting
+values and continuations can be erased and duplicated which makes it
+difficult to maintain the correspondence between a value and its negative
+counterpart. In contrast, in our setting, one can start from the empty type
+$0$, introduce a positive value and its negative counterpart, and let each of
+these flow in arbitrary ways. The entire framework guarantees that neither
+the value nor its negative counterpart will be deleted or duplicated and
+hence that, in any closed program, the ``debt'' corresponding to the negative
+value is paid off exactly once. The forward and backward executions in our
+framework correspond to call-by-value and call-by-name. This duality was
+observed by Filinski and others following him but it is particularly clean in
+our framework.
+
+\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
+subtraction type because its ``computational interpretation is not
+familiar.'' Curien and Herbelin study duality in classical logic, show that
+it exchanges call-by-value and call-by-name. They extend classical natural
+deduction with subtraction but give it no computational meaning. Crolard (in
+the formulae-as-types interpretation of subtractive logic) address the
+computational interpretation of substraction. He explains the type $A-B$ as
+the of \emph{coroutines} with a local environment of type $A$ and a
+continuation of type $B$. The description is complicated by what is
+essentially the desire to enforce linearity constraints so that coroutines
+cannot access the local environment of other coroutines. Must cite Selinger
+control categories in this context of duality but I am not sure what to say:
+it assumes cartesian closed categories for one thing and not symmetric
+monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
+explain delimited continuations.
+
+\paragraph*{Linear Logic} In accounts that are linear, the value and
+continuation that comprise the substractive type need to be constrained to
+``stay together.'' This can be achieved by various restrictions. In this work
+we have no such constraints, the negative value can flow anywhere. The entire
+system guarantees that any closed program would have to account for it. We
+don't have to introduce special constraints to achieve that. Zeilberger in
+the paper on polarity and the logic of delimited continuations uses polarized
+logic: he shows that if positive and negative values are completely symmetric
+except that answer types are positive, then the framework accommodates
+delimited continuations. But he interprets negative values are control
+operators, or as values defined by the shape of their continuations. We
+simply interpret values of negative type as values flowing in the ``other''
+direction.
+
+\paragraph*{Int Construction.}
+For a traced monoidal category {{C}} the Int construction produces a
+Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
+Further we know that the target of the Int construction is isomorphic
+to the target of \G construction of Abramsky \cite{Abramsky96:0} from
+Haghverdi. However, note that the {{langRevEE}} is not the same as the
+image of the Int construction on {{langRevT}}, since the the later
+lacks a multiplicative tensor that distrbutes over the additive tensor
+in Int {{langRevT}}.
+
+\paragraph*{Recursion.} In our previous work we introduce recursive types and
+trace operators. This is dangerous here because infinite loops allow us to
+prolong paying the debt for as long as we want.
+
+\paragraph*{GoI machines} 
+We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}

### 2012-02-19 dual.tex
small change

@@
 
+It is critical that the framework in which the negative types are introduced
+is a framework in which all information is preserved, with no duplication or
+erasure. This guarantees that the generated debts are paid once and exactly
+once.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 and the context at once.???
+
+Again it is critical that the framework in which the fractional types are
+introduced is a framework in which all information is preserved, with no
+duplication or erasure. This guarantees that the generated constraints are
+satisfied once and exactly once.
 

### 2012-02-19 dual.tex
reworded fractional discussion

@@
 a \emph{debt} is generated to compensate for the money received by the seller
-and this debt travels ``backwards'' towards the bank and is (hopefully)
-reconciled there.
+and this debt travels ``backwards'' towards the bank where it is (hopefully)
+reconciled.
 
@@
 For this reason, our treatment of negative values in the context of
-{{langRev}} will be particularly simple. We will have a type $0$ and an
+{{langRev}} is particularly simple. We will have a type $0$ and an
 isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
@@
 information is preserved, a closed program (which does not have any
-``dangling'' negative value) will eventually have to match the negative value
-with some corresponding value, effectively using the isomorphism in the
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
 right-to-left direction. In more detail, we can model the credit card
@@
 in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
-type $\texttt{bool}$. More generally, one can think of a value of type $a/b$
-as a value that imposes a constraint on its context of use: it is a value
-that requires its context to be of type $b$ and only if that condition is
-satisfied, can the value be considered as a value of type $a$. In the
-simplest case, the type $1/b$ just constraints the context to be of type $b$.
+type $\texttt{bool}$.
 
-By allowing the fractional types to be first-class, we can separate the
-generation of constraints from their use. For example, consider a variation
-on the example in the previous where the parent gives a child the \$20.00
-bill but requires that the child only spends the money on buying books. This
-can achieved by producing a constraint that can only be resolved if the child
-indeed buys a book. Technically, this is achieved by adding a new type
-isomorphism {{ 1 <-> a * (1/a) }} which when used in the left-to-right
-direction allows the creation of a value and a corresponding constraint. Both
-the value and the constraint can flow in arbitrary ways during the
+In our case, the situation is simpler. We have an isomorphism 
+{{ 1 <-> a * (1/a) }} which when used in the left-to-right direction allows the
+creation of a value and a corresponding constraint on the context. 
+Both the value and the constraint can flow in arbitrary ways during the computation 
+but eventually in a closed program with no ``dangling'' constraints, the 
+isomorphism should be used in the right-to-left direction to resolve the constraints. 
+To understand the computational interpretation, consider the following example.
+
+The goal is to produce a value of type $(\texttt{bool} \times
+\texttt{int})$. One part of the program, $e_1$, knows how to produce the
+value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
+to produce the value of type $\texttt{bool}$ (say {{true}}). Computationally,
+we model this as follows. The first subcomputation $e_1$ produces {{((),3)}}
+and uses the isomorphism to convert {{()}} to {{(alpha,1/alpha)}} where
+{{alpha}} is a yet-unknown boolean value. Reshuffling the components, $e_1$
+produces {{((alpha,3), 1/alpha)}}. Similarly, $e_2$ can independently produce
+{{((true,beta),1/beta)}} where {{beta}} is a yet-unknown integer value.
+When the two values meet, we can group the components as follows:
+\[
+{{(alpha,beta)}}  {{(3,1/beta)}}  {{(true,1/alpha)}}
+\]
+Using the isomorphism in the right-to-left direction on the last two pairs,
+forces {{alpha}} to be resolved to {{true}} and forces {{beta}} to be
+resolved to {{3}}. Both of these pairs then become {{()}} and be absorbed. We
+are left with the pair {{(true,3)}} as desired.
+
+To summarize, one can think of a value of type $a/b$ as a value that imposes
+a constraint on its context of use: it is a value that requires its context
+to be of type $b$ and only if that condition is satisfied, can the value be
+considered as a value of type $a$. In the simplest case, the type $1/b$ just
+constraints the context to be of type $b$. By allowing the fractional types
+to be first-class, we can separate the generation of constraints from their
+use. Both the value and the constraint can flow in arbitrary ways during the
 computation but eventually in a closed program with no ``dangling''
@@
 NEED to work out the example in detail (or perhaps another example that's
-better. Not sure if this one makes sense because we have to create the book
-and the context at once.???
+better.
 

### 2012-02-19 dual.tex
reworded fractional discussion

@@
 
-In our case, the situation is simpler. We have an isomorphism 
-{{ 1 <-> a * (1/a) }} which when used in the left-to-right direction allows the
-creation of a value and a corresponding constraint on the context. 
-Both the value and the constraint can flow in arbitrary ways during the computation 
-but eventually in a closed program with no ``dangling'' constraints, the 
-isomorphism should be used in the right-to-left direction to resolve the constraints. 
-To understand the computational interpretation, consider the following example.
+In our case, the situation is simpler. If the goal is to produce a value of
+type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
+have type {{(a * b) / b}}. In most settings, this type would be equivalent to
+{{a}}. However in our setting, the constraint that this value must eventually
+fit in a larger value that supplies the {{b}} is explicitly recorded and must
+be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
+which when used in the left-to-right direction allows the creation of a value
+and a corresponding constraint on the context.  Both the value and the
+constraint can flow in arbitrary ways during the computation but eventually
+in a closed program with no ``dangling'' constraints, the isomorphism should
+be used in the right-to-left direction to resolve the constraints.  To
+understand the computational interpretation, consider the following example.
 

### 2012-02-20 dual.tex
imaginary

@@
 
-\title{Functional Pearl \\ Programming with Negative and Fractional Types} 
+\title{Functional Pearl \\ Programming with Negative, Fractional, and Imaginary Types} 
 \authorinfo{Roshan P. James}
@@
   Every functional programmer knows about sum and product types, $a+b$ and
-  $a*b$ respectively. Negative and fractional types, $a-b$ and $a/b$
-  respectively, are much less known and their computational interpretation is
-  unfamiliar and often complicated. We show that in a programming model in
-  which information is preserved (such as the model introduced in our recent
-  paper on \emph{Information Effects}), negative and fractional types have
-  particularly intuitive and natural computational
-  interpretations. Intuitively, values of negative types are values that flow
-  ``backwards'' to satisfy demands, and values of fractional types are values
-  that represent first-class constraints. The combination of negative and
-  fractional types allow a programmer to express a wide range of programming
-  idioms, including higher-order functions, delimited continuations,
-  speculative computation, logic programming, and more.
+  $a*b$ respectively. Negative, fractional, and imaginary types, $a-b$,
+  $a/b$, and $a+ib$ respectively, are much less known and their computational
+  interpretation is unfamiliar and often complicated. We show that in a
+  programming model in which information is preserved (such as the model
+  introduced in our recent paper on \emph{Information Effects}), negative,
+  fractional, and imaginary types have particularly intuitive and natural
+  computational interpretations. Intuitively, values of negative types are
+  values that flow ``backwards'' to satisfy demands, values of fractional
+  types are values that represent first-class constraints, and values of
+  imaginary types are values that capture the essence of recursive datatype
+  declarations. The combination of negative, fractional, and imaginary types
+  allow a programmer to express a wide range of programming idioms, including
+  higher-order functions, delimited continuations, speculative computation,
+  logic programming, and more.
 \end{abstract}
@@
 satisfied once and exactly once.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Imaginary Types: Intuition}
+
+Imaginary types have also appeared in the literature: we relegate the
+connections to Sec.~\ref{related} and proceed with a simple explanation. We
+have so far extended the set of types to be the rational numbers. Now we will
+push this and extend the set of types to algebraic numbers. In other words,
+we will allow datatypes defined by arbitrary polynomials and allow the roots
+of such polynomials to be types. Consider the type of binary trees:
+\begin{verbatim}
+data B = 1 + B * B
+\end{verbatim}
+Rearranging the terms, this declaration introduces the polynomial $b^2 - b +
+1 = 0$ whose roots are $(1 \pm i\sqrt{3}) / 2$.
 

### 2012-02-20 dual.tex
imaginary

@@
 \end{verbatim}
-Rearranging the terms, this declaration introduces the polynomial $b^2 - b +
-1 = 0$ whose roots are $(1 \pm i\sqrt{3}) / 2$.
+In conventional settings, this declaration introduces an isomorphism between
+$b$ and $1+b^2$ (whose two sides are commonly known as {{fold}} and
+{{unfold}}). In our setting, this induces an isomorphism between $0$ and $b^2
+- b + 1$. The latter polynomial has roots $(1 \pm i\sqrt{3}) / 2$ and hence
+we have two isomorphisms between $b - (1 + i\sqrt{3})$ and $0$ and also
+between $b - (1 - i\sqrt{3})$ and $0$.
+
+Now say I want to produce a value of type $b$. We can use the isomorphisms
+above to break this task in many ways. Essentially exploit that $b$ is
+isomorphic to $b^7$, let the various program components produce the various
+parts of $b^7$ and merge them.
 

### 2012-02-20 dual.tex
imaginary

@@
 parts of $b^7$ and merge them.
+
+Consider some simple examples first.
+
+\begin{itemize}
+
+\item $b=1+1$. Boring.
+
+\item $b=b+1$. That introduces the natural numbers. No solution for this
+  polynomial over the algebraic numbers. We must extend the numbers with
+  $\omega$ to get a solution. We reject this in this paper and prefer to
+  stick to algebraic numbers. The advantage is that all the isomorphisms are
+  valid numerically. With the above type we could subtract $b$ from both
+  sides to show that $0=1$ which is nonsense. We can however have infinite
+  types as long as they have algebraic solutions.
+
+\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
+  boolean! If we had superpositions, we could then write a function that
+  performs the square root of negation in the sense that applying it twice
+  would be equivalent to boolean negation.
+
+\end{itemize}
 

### 2012-02-20 dual.tex
added stuff on imaginary and square root types

@@
 
-\title{Functional Pearl \\ Programming with Negative, Fractional, and Imaginary Types} 
+\title{Functional Pearl --- Programming with \\ 
+  Negative, Fractional, Square Root, and Imaginary Types} 
 \authorinfo{Roshan P. James}
@@
   Every functional programmer knows about sum and product types, $a+b$ and
-  $a*b$ respectively. Negative, fractional, and imaginary types, $a-b$,
-  $a/b$, and $a+ib$ respectively, are much less known and their computational
-  interpretation is unfamiliar and often complicated. We show that in a
-  programming model in which information is preserved (such as the model
-  introduced in our recent paper on \emph{Information Effects}), negative,
-  fractional, and imaginary types have particularly intuitive and natural
-  computational interpretations. Intuitively, values of negative types are
-  values that flow ``backwards'' to satisfy demands, values of fractional
-  types are values that represent first-class constraints, and values of
-  imaginary types are values that capture the essence of recursive datatype
-  declarations. The combination of negative, fractional, and imaginary types
-  allow a programmer to express a wide range of programming idioms, including
+  $a*b$ respectively. Negative, fractional, square root, and imaginary types,
+  $a-b$, $a/b$, $\sqrt{a}$, and $a+ib$ respectively, are much less known and
+  their computational interpretation is unfamiliar and often complicated. We
+  show that in a programming model in which information is preserved (such as
+  the model introduced in our recent paper on \emph{Information Effects}),
+  these types have particularly intuitive and natural computational
+  interpretations. Intuitively, values of negative types are values that flow
+  ``backwards'' to satisfy demands, values of fractional types are values
+  that represent first-class ``structural'' constraints, and values of square
+  root and imaginary types are values that can be related by arbitrary
+  algebraic constraints. The combination of these negative, fractional,
+  square root, and imaginary types enable greater flexibility in programming
+  by breaking complicated invariants into local ones that can be
+  independently satisfied by a subcomputation. Furthermore, they allow a
+  programmer to express a wide range of programming idioms, including
   higher-order functions, delimited continuations, speculative computation,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Imaginary Types: Intuition}
+\section{Algebraic Types: Intuition}
 
-Imaginary types have also appeared in the literature: we relegate the
-connections to Sec.~\ref{related} and proceed with a simple explanation. We
-have so far extended the set of types to be the rational numbers. Now we will
-push this and extend the set of types to algebraic numbers. In other words,
-we will allow datatypes defined by arbitrary polynomials and allow the roots
-of such polynomials to be types. Consider the type of binary trees:
-\begin{verbatim}
-data B = 1 + B * B
-\end{verbatim}
-In conventional settings, this declaration introduces an isomorphism between
-$b$ and $1+b^2$ (whose two sides are commonly known as {{fold}} and
-{{unfold}}). In our setting, this induces an isomorphism between $0$ and $b^2
-- b + 1$. The latter polynomial has roots $(1 \pm i\sqrt{3}) / 2$ and hence
-we have two isomorphisms between $b - (1 + i\sqrt{3})$ and $0$ and also
-between $b - (1 - i\sqrt{3})$ and $0$.
+Square root and imaginary types have also appeared in the literature: we
+relegate the connections to Sec.~\ref{related} and proceed with a simple
+explanation. We have so far extended the set of types to be the rational
+numbers. Now we will push this and extend the set of types to algebraic
+numbers. In other words, we will allow datatypes defined by arbitrary
+polynomials and allow the roots of such polynomials to be types.
 
-Now say I want to produce a value of type $b$. We can use the isomorphisms
-above to break this task in many ways. Essentially exploit that $b$ is
-isomorphic to $b^7$, let the various program components produce the various
-parts of $b^7$ and merge them.
+Consider first an example in which we want to compute with the sides of a
+rectangle whose area is 91 and whose length is longer than its width by 6
+units. One can solve the quadratic equation to determine that the sides are 7
+and 13 and proceed. This however prematuraly forces us to globally solve the
+constraint. Instead we can let the two sides of the rectangle be $x$ and
+$x+6$ and use the following equation to capture the desired constraint:
+\[
+x^2 + 6x - 91 = 0
+\]
+The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
+the type $0$. We can now proceed to compute with the unknown $x$, being
+assured that in a closed program, our computation will eventually be
+consistent with the solution of the algebraic equation.
 
-Consider some simple examples first.
+Previously the most famous example of a similar nature is the puzzling
+isomorphism that one can establish between seven binary trees and one.
+A binary tree is defined by the datatype
+\[
+x = 1 + x * x 
+\]
+which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
+algebraic manipulation we can reason as follows:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+Fiore poses the question of why such algebraic manipulation would make sense
+type theoretically but states that even though some of the intermediate steps
+make no sense, the final equivalence is valid and can be used to actually
+construct an isomorphism between $x^7$ the type of seven binary trees 
+and $x$ the type of binary trees.
 
+Discussion of possible polynomials:
 \begin{itemize}
-
 \item $b=1+1$. Boring.
-
 \item $b=b+1$. That introduces the natural numbers. No solution for this
@@
   types as long as they have algebraic solutions.
-
 \item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
@@
   would be equivalent to boolean negation.
+\end{itemize}
 
-\end{itemize}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Summary}
+
+To summarize negative, fractional, square root, and imaginary types all make
+sense. What they help you accomplish as a programmer is to disassociate
+global invariants into local ones that can be satisfied independently by
+subcomputations with no synchronization or communication. A computation
+producing something of type $a/b$ does not need to concern itself with who is
+going to supply the missing $b$: it just does its part. Conversely faced with
+a complicated task, a computation might decide to break it into pieces and
+demand these pieces using negative types.
+
+It is no surprise that these types are closely related to quantum mechanics
+and that they give us the feel that this is how nature computes. This is
+speculation however.
+
+In any case, in a framework where information can be copied and deleted, none
+of this makes much sense. It is critical that these constraints and demands
+can neither be duplicated nor erased.  This gives us the maximum
+``parallelism'' possible.
 

### 2012-02-22 iso-int/iso-int.tex
pi diagrams
updated pi diagram
PiEE interpreter
operational semantics
_
fiore

@@
 %! |--> = \mapsto
+%! <--| = \mapsfrom
 %! |- = \vdash
@@
 %! langRevT = \Pi^{o}
+%! langRevEE = \Pi^{\eta\epsilon}_{+}
 %! bullet = \bullet
@@
 \begin{document}
-\title{On the Construction of Isomorphic Interpreters} 
+\title{Isomorphic Interpreters in {{langRevT}} (Journal Version)} 
 \titlerunning{On the construction of Isomorphic Interpreters} 
@@
 %subcode{opsem} include main
-% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
-% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
-% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
-% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
-% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
-% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
-% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
-% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
-% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
-% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
-% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/pi1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/pi2.pdf}
+}
+\end{center}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A {{langRevT}} Interpreter}
+
+%subcode{bnf} include main
+% Combinators, c = iso | c (;) c | c (*) c | c (+) c | trace c
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc | Trace cc
+% Values, v = () | (v, v) | L v | R v
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
+% < trace c, v, cc> &|-->& <c, R v, Trace cc> &~~~~~~~~~~~~ rule 12
+% {[c, L v, Trace cc]} &|-->& <c, L v, Trace cc> &~~~~~~~~~~~~ rule 13
+% {[c, R v, Trace cc]} &|-->& {[trace c, R v, cc]} &~~~~~~~~~~~~ rule 14
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/pi0-1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/pi0-2.pdf}
+}
+\end{center}
+
+TODO : The real difficulty with this interpreter is to argue why it is
+not full of undefined states. For instance, why the machines will
+never reach stuck states such as {{<c1(+)c2, (), ..>}}.  The right way
+out of that is to make the type system of the abstract machine rich
+enough that the structure of the abstract automatically rules out this
+sort of thing.
+
+%subcode{bnf} include main
+% Combinators, c(t1<->t2) = 
+%             &|& c(t<->t) = iso 
+%             &|& c(t1<->t2) = c(t1<->t3) (;) c(t3<->t2) 
+%             &|& c(t1*t2<->t3*t4) = c(t1<->t3) (*) c(t2<->t4) 
+%             &|& c(t1+t2<->t3+t4) = c(t1<->t3) (+) c(t2<->t4)
+%
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+%
+% Type Indexed Values, v(t) =
+%  &|& v(1) = () 
+%  &|& v(t1*t2) = (v(t1), v(t2)) 
+%  &|& v(t1+t2) = L v(t1) | R v(t2)
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+However this begs the question - does {{langRevT}} need all these GADT
+types to reason about all these things? Something to think about. If
+we do, then thereare two ways out
+
+(1) add GADTs to {{langRevT}} and keep the structure of things
+simple. They can always be added opaquely for any given types -- we
+only need to add the proper fold/unflod morphisms for those types
+(maybe).
+
+(2) compile down to {{langRevT}} as it is. In this case, {{langRevT}}
+is indeed a relatively less typed language than the source language
+and will hence have to treat as stuck-states the non-type correct
+machine states. We could apply omega to them and make them vanish --
+for example. This is similar to the handling of the {{sub1}} in the
+previous example.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A {{langRevEE}} interpreter}
+
+%subcode{bnf} include main
+% Combinators, c = iso | eta | eps | c (;) c | c (*) c | c (+) c
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+% Values, v = () | (v, v) | L v | R v
+%
+% Machine states = <c, v, cc> | {[c, v, cc]} | <c, v, cc>^{dagger} | {[c, v, cc]}^{dagger}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+%
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
+% <eps, R v, cc> &|-->& <eps, L (-v), cc>^{dagger} &~~~~~~~~~~~~ rule 12
+% <eps, L (-v), cc> &|-->& <eps, R v, cc>^{dagger} &~~~~~~~~~~~~ rule 13
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, cc>^{dagger} &<--|& {[iso, iso(v), cc]}^{dagger} &~~~~~~~~~~~~ rule 1'
+% <c1(;)c2, v, cc>^{dagger} &<--|& <c1, v, Fst cc c2>^{dagger} &~~~~~~~~~~~~ rule 2'
+% {[c1, v, Fst cc c2]}^{dagger} &<--|& <c2, v, Snd c1 cc>^{dagger} &~~~~~~~~~~~~ rule 3'
+% {[c2, v, Snd c1 cc]}^{dagger} &<--|& {[ c1(;)c2, v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 4'
+% <c1(+)c2, L v, cc>^{dagger} &<--|& <c1, v, LeftPlus cc c2>^{dagger} &~~~~~~~~~~~~ rule 5'
+% {[ c1, v, LeftPlus cc c2 ]}^{dagger} &<--|& {[c1 (+) c2, L v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 6'
+% <c1(+)c2, R v, cc>^{dagger} &<--|& <c2, v, RightPlus c1 cc>^{dagger} &~~~~~~~~~~~~ rule 7'
+% {[ c2, v, RightPlus c1 cc ]}^{dagger} &<--|& {[c1 (+) c2, R v, cc ]}^{dagger} &~~~~~~~~~~~~ rule 8'
+% <c1(*)c2, (v1, v2), cc>^{dagger} &<--|& <c1, v1, LeftTimes cc c2 v2>^{dagger} &~~~~~~~~~~~~ rule 9'
+% {[ c1, v1, LeftTimes cc c2 v2 ]}^{dagger} &<--|& <c2, v2, RightTimes c1 v1 cc>^{dagger} &~~~~~~~~~~~~ rule 10'
+% {[ c2, v2, RightTimes c1 v1 cc ]}^{dagger} &<--|& {[ c1 (*) c2, (v1, v2), cc ]}^{dagger} &~~~~~~~~~~~~ rule 11'
+% {[eps, L (-v), cc]}  &<--|& {[eps, R v, cc]}^{dagger}  &~~~~~~~~~~~~ rule 12'
+% {[eps, R v, cc]} &<--|&  {[eps, L (-v), cc]}^{dagger} &~~~~~~~~~~~~ rule 13'
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/piee1.pdf}
+}
+\end{center}
+
+Let us summarize it as follows.
+
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/piee2.pdf}
+}
+\end{center}
+
+Adding backward execution using the adjoint.
+
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/piee3.pdf}
+}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{CEK machine}
+
+%% %subcode{bnf} include main
+%% % e = x | \x.e | e e
+%% % rho = [] | x=v:rho
+%% % E = [] | (rho, e):E | v:E
+%% % v = (\x.e, rho)
+%% %
+%% % Machine states = <e, rho, E>
+%% % Start state = <e, [], []>
+%% % Stop state = <v, rho, []>
+
+%% %subcode{opsem} include main
+%% % <e1 e2, rho, E > &|-->& <e1, rho, (rho, e2):E >
+%% % <\x.e, rho, (rho', e'):E > &|-->& <e', rho', (\x.e,rho):E >
+%% % <\x.e, rho, (\x'.e',rho'):E > &|-->& <e', x' =(\x.e,rho):rho', E >
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\section{Isomorphic Interpreters from Algebraic manipulation}
+
+aka. ultra-constructivism.
+
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/fiore_t3.pdf}
+}
+\end{center}
+
+The following diagram expresses the isomorphism {{t^3 <-> -1}} from
+Fiore's paper \cite{Fiore:2004}. The computation corresponds to the
+following algebraic manipulation:
+
+{{t^3 = t*(t*t)}}
+
+{{t^3 = t*(t-1)}} (because {{t = t*t + 1}}  and hence {{t*t = t - 1}}).
+
+{{t^3 = t*t-t}}
+
+{{t^3 = -1}} (because {{t = t*t + 1}}  and hence {{t*t - t = -1}}).
+
+However this computation is always non-terminating. So it is
+computationally useless. Fiore on the other hand says that {{t^7 = t}}
+has a useful computational interpretation. Which gives rise to the
+conclusion that there are algebraic proofs of equalities that have
+interesting computational content and once that are nonsense. What
+does this mean? If these programs are proof, why are there two sorts
+of proofs here? Are one set of proofs more ``constructive'' than the
+other? If so, how do we characterize them?
+
+%% data Tree = Node Tree Tree 
+%%           | Leaf
+
+%% t = t * t + 1
+
 \bibliographystyle{splncs03} 

### 2012-02-23 iso-int/iso-int.tex
_

@@
 \begin{document}
-\title{Isomorphic Interpreters in {{langRevT}} (Journal Version)} 
+\title{On the Construction of Isomorphic Interpreters (Journal Version)} 
 \titlerunning{On the construction of Isomorphic Interpreters} 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Simple Bounded Iterator}
+\section{Isomorphic Interpreters from Logically Reversible Abstract Machines}
+\subsection{Simple Bounded Iterator}
+
 Consider definition for the following small-step abstract machine:
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Generating Inductive Walks for Algebraic Data Types}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage
-\section{Isomorphic Interpreters from Algebraic manipulation}
+\section{Isomorphic Interpreters from Algebraic Manipulation}
 
@@
 
+- Control theory
+
+- Fiore's conditions
+
+- b7 = b (two derivations)
+
 \bibliographystyle{splncs03} 

### 2012-02-23 dual.tex
intro ideas

@@
 \section{Introduction}
+
+The world of computation we are describing has:
+\begin{itemize}
+\item suppliers, 
+\item consumers, and
+\item bi-directional transformations
+\end{itemize}
+This is the same world described by the papers on the duality of computation
+but that work only scratched the surface! We have the following features:
+\begin{itemize}
+\item we can start from the supplier and push the values towards the
+  consumer (call-by-value in the duality of computation papers)
+\item we can start from the consumer and pull the values from the suppliers
+  (call-by-name in the duality of computation papers)
+\item we can combine the pushing and pulling and values using eta/epsilon for
+  sum types; these allow us to at any point in the middle of the computation
+  create out of nothing a value to send to the consumers and a demand to send
+  to the suppliers.
+\item we can break a big data structure into fragments described by
+  fractional types; the suppliers and consumers can produce and consume the
+  pieces completely independently of each other. Eventually the pieces will
+  fit together at the consumer to produce the desired output.
+\item we can break a bi-directional transformation into pieces using square
+  roots
+\item we can take into account that values have phase (complex numbers),
+  i.e., it is not that they flow towards the consumer or just towards the
+  suppliers; they can be flowing in direction that ``30 degrees'' towards the
+  consumer for example.
+\end{itemize}
+So it is all about breaking dependencies in some sense to allow for maximum
+autonomy (parallelism) of subcomputations. It is probably the case that to
+make full use of square root types and imaginary types, we have to move to a
+vector space. If that's the case, we should probably leave this stuff out and
+focus on negative and fractional types and only have a short discussion of
+the polynomials restricted to seven trees in one and similar issues.
 

### 2012-02-23 iso-int/rc-iso-int.tex
this will be the rc2012 paper
fiore
_
outline

--- 2012-02-17 iso-int/iso-int.tex
@@
 \begin{document}
-\title{On the Construction of Isomorphic Interpreters} 
+\title{Isomorphic Interpreters from Logically Reversible Abstract Machines} 
 \titlerunning{On the construction of Isomorphic Interpreters} 
@@
 \section{Introduction}
+
+Main points: 
+\begin{itemize}
+\item Make the case for reversibility via type isomorphisms
+\item Type isomorphisms between finite types are straightforward (but
+  tedious)
+\item Type isomorphisms between recursive types are quite difficult. This is
+  what we focus on.
+\end{itemize}
+To understand the problem in more detail, consider the problem of converting
+a reversible recursive function into a partial isomorphism. In general such a
+function takes an input, makes several recursive calls, and produces an
+answer. To implement this in a reversible way, the structure of the recursion
+and the number of recursive calls need to be exposed. The key trick is to
+express tease the function into two components: an outer driver that performs
+iteration and an inner state transition function that realizes one
+``unfolding'' of the recursion. In program transformation world, this is
+called, light-weight fission. Using this idea, one can realize any recursive
+function as a partial isomorphism. We show a few examples of the construction
+culminating with the most challenging recursive function: a full interpreter.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Logically Reversible Multiplier}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Tree Traversal}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{A {{langRev}} Interpreter}
+\section{A {{langRevT}} Interpreter}
 
 %subcode{bnf} include main
-% Combinators, c = iso | c (;) c | c (*) c | c (+) c
+% Combinators, c = iso | c (;) c | c (*) c | c (+) c | trace c
 % Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
 %                  &|& LeftTimes cc c v | RightTimes c v cc 
-%                  &|& LeftPlus cc c | RightPlus c cc 
+%                  &|& LeftPlus cc c | RightPlus c cc | Trace cc
 % Values, v = () | (v, v) | L v | R v
@@
 %subcode{opsem} include main
-% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
-% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
-% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
-% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
-% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
-% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
-% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
-% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
-% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
-% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
-% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
+% < trace c, v, cc> &|-->& <c, R v, Trace cc> &~~~~~~~~~~~~ rule 12
+% {[c, L v, Trace cc]} &|-->& <c, L v, Trace cc> &~~~~~~~~~~~~ rule 13
+% {[c, R v, Trace cc]} &|-->& {[trace c, R v, cc]} &~~~~~~~~~~~~ rule 14
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/pi0-1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/pi0-2.pdf}
+}
+\end{center}
 

### 2012-02-24 iso-int/rc-iso-int.tex
started rc cleanup

@@
 \section{Simple Bounded Iterator}
-Consider definition for the following small-step abstract machine:
+We illustrate the main ideas underlying the derivation of isomorphic
+interpreters using a simple abstract machine.  Consider the definition
+of the following small-step abstract machine:
+
+\begin{multicols}{2}
+%subcode{bnf} include main
+% Numbers, n, m = 0 | n + 1
+% Machine states = <n, n>
 
 %subcode{bnf} include main
-% Expressions, e = n
-% Numbers, n, m = 0 | n + 1
-%
-% Machine states = <e, e>
 % Start state = <n, 0>
 % Stop State = <0, n>
-
-Small-step Operational semantics:
+\end{multicols}
+% Small-step operational semantics:
 
@@
 
-This machine start with a number {{n}} in the first position and at
-each reduction step decrements the first number and increments the
-second. It stops when the first number reaches 0, thereby taking
-exactly {{n}} steps. While the machine does nothing interesting,
-it useful to illustrate the general idea.
-
-\begin{enumerate}
-\item 
-We can turn this abstract machine into a an isomorphic interpreter
-implemented in {{langRevT}}, in a systematic way. To start with we
-realize that the interpreter should implement the following map.
+This machine starts with a number {{n}} in the first position and
+{{0}} in the second. Each reduction step decrements the first number
+and increments the second. The machine stops when the first number
+reaches 0, thereby taking exactly {{n}} steps.  
+%% We can turn this
+%% abstract machine into a an isomorphic interpreter implemented in
+%% {{langRevT}}, in a systematic way. 
+Broadly, the interpreter should implement the map on the left:
 
@@
 
-Now numbers can be implemented by the recursive type {{nat = @@x.1+x}}
-in {{langRevT}}. This gives us the isomorphisms
+%% \begin{center}
+%%   \includegraphics{diagrams/nat-nat2.pdf}
+%% \end{center}  
+%% \end{multicols}
+
+In {{langRevT}} we have {{nat}} given by the recursive definition
+{{@@x.1+x}} along with isomorphisms {{unfold:nat <->1+nat:fold}}. The
+only reduction step of the abstract machine examines the first value
+of the input and reconstructs the second.
 
 \begin{center}
-  {{unfold : nat <-> 1 + nat : fold}}
+  \includegraphics{diagrams/nat-nat3.pdf}
 \end{center}
 
-We start by examining the first value.
-
-\begin{center}
-  \includegraphics{diagrams/nat-nat2.pdf}
-\end{center}
-
-\item
 The triangle denotes the {{unfold}} isomorphism, which works as
@@
 
-Since the abstract machine increments second component, we know that
-we need a {{fold}} on the second component to accomplish this. So we
-can fill in the output part of the interpreter dually.
+%% Now numbers can be implemented by the recursive type {{nat = @@x.1+x}}
+%% in {{langRevT}}. This gives us the isomorphisms
 
-\begin{center}
-  \includegraphics{diagrams/nat-nat3.pdf}
-\end{center}
+%% \begin{center}
 
-\item
+%% \end{center}
+
+%% Since the abstract machine increments second component, we know that
+%% we need a {{fold}} on the second component to accomplish this. So we
+%% can fill in the output part of the interpreter dually.
+
 Let us examine all the possible input states, by distributing on the
@@
 
-\item
 The construction above already gives us some insight into how we can
@@
 
-\item
 We are a few steps away from an interpreter at this point and the
@@
 
-\item What do we do about the two inner branches?
+\item
+
+ What do we do about the two inner branches?
 
@@
 \end{center}
-
-\end{enumerate}
 

### 2012-02-24 dual.tex
added some long term thoughts

@@
 the polynomials restricted to seven trees in one and similar issues.
+
+The conventional idea is to divide the world into a ``real'' one and a
+``virtual'' one. In the ``real'' world, we can define datatypes like
+\verb|t=t^2+1| but we don't have additive inverses so it makes no sense to
+talk of negative types and we can't rearrange the terms in the datatype
+definition. However the observation is that we can map these datatypes to a
+virtual world that has more structure (a ring that provides additive inverses
+or a field that also provides multiplicative inverses) and then perform
+computations in the ring/field. If we perform computations in the ring, then
+some of these will use additive inverses in ways that cannot be mapped back
+to the ``real'' world. Much of current research attempts to characterize
+which computations done in the ring are valid isomorphisms between datatypes
+in the ``real'' world. This is nice but is not what I am after. In fact I am
+not interested in the ring or the semiring at all. I am interested in the
+field and I want this field to be \textbf{the real world.} This is partly
+motivated by the fact that Quantum Mechanics seems to demand an underlying
+field and more generally that the field provides the maximum generality in
+slicing and dicing computations. So assuming I live in a field and that the
+negative, fractional, square root, and imaginary types are all ``real,'' how
+do I compute in this field? Clearly there will be constraints on
+``measurement'' in the sense that a full program cannot produce any of the
+crazy types but that's done outside the formalism in some sense just as in
+Quantum Computing. The main question I am after is how to compute in this
+field with first-class negative, fractional, etc. types. As I mentioned in my
+previous email, we can produce programs that have types \verb|t^3 <-> -1| and
+they ``run'' (but only to give infinite loops).
+
+So when a programmer writes the datatype declaration \verb|t = t^2+1|, if we
+allow negative etc. then this is effectively writing 
+\verb|t = cubicroot{-1}|. If we are in the field then computations that 
+manipulate these trees can be sliced and diced even at interfaces that
+expose the cubic root and the imaginary types.
+
+Future work: develop a type system for a ``normal language'' that has
+negative, fractional, etc. types as first-class types. More long term,
+instead of adding one polynomial at a time, we can go to an algebraically
+closed field. The complex numbers is an obvious choice but I would rather go
+to something computable like the field of algebraic numbers. Is the adele
+ring or the p-adics relevant here?
 

### 2012-02-24 iso-int/rc-iso-int.tex
sec 2 done

@@
 of the following small-step abstract machine:
-
+%\vspace{-30px}
 \begin{multicols}{2}
@@
 
-The triangle denotes the {{unfold}} isomorphism, which works as
-follows: If the number {{n}} is zero, the output is the top branch
-which has type {{1}}. If the number was non-zero, the output is on the
-bottom branch and has value {{n-1}}.
+The triangle on the left denotes the {{unfold}} isomorphism, which
+works as follows: If the number {{n}} is zero, the output is the top
+branch which has type {{1}}. If the number was non-zero, the output is
+on the bottom branch and has value {{n-1}}. The triangle on the right
+is the dual {{fold}} isomorphism.  Let us examine all the possible
+machine states by introducing distribution operations and any required
+swaps.
 
-%% Now numbers can be implemented by the recursive type {{nat = @@x.1+x}}
-%% in {{langRevT}}. This gives us the isomorphisms
-
-%% \begin{center}
-
-%% \end{center}
-
-%% Since the abstract machine increments second component, we know that
-%% we need a {{fold}} on the second component to accomplish this. So we
-%% can fill in the output part of the interpreter dually.
-
-Let us examine all the possible input states, by distributing on the
-input. Dually let us do the same with the output, but this time also
-be a little bit explicit about the swap operations required so the
-right types are in place.
+%% by distributing on the
+%% input. Dually let us do the same with the output, but this time also
+%% be a little bit explicit about the swap operations required so the
+%% right types are in place.
 
@@
 The construction above already gives us some insight into how we can
-implement our one reduction step. If we chose {{m = m'-1}} and we
-chose {{n-1 =n'}} then we have essentially encoded the required
-rewrite. Lets state this in words - the new value of {{n}}, namely
-{{n'}}, is equal to {{n-1}}. And further - the new value of {{m}},
-namely {{m'}}, is {{m+1}} (because {{m'-1=m}}).
-
-We can thus connect the lower branches with the appropriate
-isomorphism to reflect this relationship. In this case the isomorphism
-is just a swap of the two wires involved.
+implement our machine's one reduction step. We want the output {{n'}}
+to be {{n+1}} where {{n}} is the first component of the input and
+similarly the output {{m'}} to be {{m+1}} where {{m}} is the second
+component of the input.  We can express these equalities by connecting
+wires -- by connecting the {{n-1}} wire to {{n'}} we express the
+equality {{n' =n-1}}. Similarly we connect {{m}} to {{m'-1}} to
+express {{m' = m+1}}.
 
@@
 
-We are a few steps away from an interpreter at this point and the
-final steps rely on the answers to a few questions.
+%% We are a few steps away from an interpreter at this point and the
+%% final steps rely on the answers to a few questions.
 
-\begin{enumerate}
-\item How do relate the outputs of the combinator to the inputs, such
-  that the machine correctly takes multiple steps?
+%% \begin{enumerate}
+%% \item How do relate the outputs of the combinator to the inputs, such
+%%   that the machine correctly takes multiple steps?
 
-\item
+%% \item
 
- What do we do about the two inner branches?
+%%  What do we do about the two inner branches?
 
-\end{enumerate}
+%% \end{enumerate}
 
-The branch labeled {{((), m)}} looks like the stop state of the
-machine. The branch label-led {{((), n')}} looks like the start state
-of the program. To complete the interpreter, we flip our current
-construction inside out.
+The branch labeled {{((), m)}} corresponds to the machine state
+{{<0,m>}} and hence looks like the stop state of the
+machine. Similarly the branch labeled {{((), n')}} corresponds to the
+start state of the machine.
+
+We are close to the completion of the interpreter at this point. The
+last step in the construction involves introducing a {{trace}]
+operation for iteration of reduction steps and exposing the start and
+stop states. We do this by `flipping' the construction so far
+inside-out.
+
+\vspace{-20px}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.5}{
+  \includegraphics{diagrams/nat-nat6.pdf}
+}
+\end{center}
+\columnbreak
+\begin{center}
+\scalebox{0.6}{
+  \includegraphics{diagrams/nat-nat7.pdf}
+}
+\end{center}  
+\end{multicols}
+
+The introduced {{trace}} feeds back related machine states. The
+exposed start state essentially requires the initial value of {{n}} to
+start the machine.
 
 \begin{center}
-  \includegraphics{diagrams/nat-nat6.pdf}
-\end{center}
-
-Which gives us
-
-\begin{center}
-  \includegraphics{diagrams/nat-nat7.pdf}
-\end{center}
-
-This basically completely our construction of the interpreter. We
-basically need to use a {{trace}} to feedback related machine states
-to related machine states.
-
-\begin{center}
-  \includegraphics{diagrams/nat-nat8.pdf}
+\scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
 \end{center}

### 2012-02-24 iso-int/rc-iso-int.tex

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{What are the high level steps in the translation?}
-\label{sec:steps}
-
-TODO
-
-\begin{enumerate}
-\item Expand the input such that every input term that the LHS of the
-  interpreter pattern matches on is exposed.
-
-\item Expand the output such that every output term that it
-  reconstructed by the interpreter is exposed.
-
-\item Shuffle matching LHS term to matching RHS terms, inserting any
-  appropriate mediating computations. In the previous example, this
-  was only a swap operation.
-
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{What abstract machines can be turned into isomorphic interpreters?}
-
-Logically reversible abstract machines with distinguishable start and
-stop states can be converted into isomorphic interpreters in
-{{langRevT}}.  To develop an understanding for what sorts of machines
-these are, let us look at some examples.
+The construction of the isomorphic interpreter essentially exploits
+the structure of most abstract machines -- the LHS of the transition
+rules determines a machine state by pattern machine. The RHS does the
+dual namely term reconstruction. The fact that the machine is
+logically reversible ensures that the computation performed by each
+reduction is essentially captured by an isomorphism. 
+%% To develop an
+%% understanding for what sort of machines can be treated this way, let
+%% us look at some examples.
 
@@
 rule does not have a logically inverse -- in other words, given a
-{{z}}, we are unable to determine which {{x}} and {{y}} it originated
-from.
+{{z}}, we are unable to determine the original {{x}} and {{y}}.
 
@@
 
-\noindent
-Conditions that must be satisfied for constructing a isomorphic
-interpreter:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \subsection{What abstract machines can be turned into isomorphic interpreters?}
+
+\paragraph*{Conditions that must be satisfied for constructing a isomorphic
+interpreter}:
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Adder and Multiplier}
+\paragraph*{High level steps in the translation:}
+\label{sec:steps}
 
-\subsection{Logically Reversible Adder}
+\begin{enumerate}
+\item Expand the input such that every input term that the LHS of the
+  interpreter pattern matches on is exposed.
+
+\item Expand the output such that every output term that it
+  reconstructed by the interpreter is exposed.
+
+\item Shuffle matching LHS term to matching RHS terms, inserting any
+  appropriate mediating computations. In the previous example, this
+  was only a swap operation.
+
+\item Flip the construction inside out introducing a {{trace}} and
+  exposing the start and stop states.
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Adder and Multiplier}
+
+\section{Logically Reversible Adder}
+
+Consider
+
+\begin{multicols}{2}  
+%subcode{bnf} include main
+% Numbers, n, m,p = 0 | n + 1
+% Machine states = <e, e, e>
 
 %subcode{bnf} include main
-% Expressions, e = n
-% Numbers, n, m,p = 0 | n + 1
-%
-% Machine states = <e, e, e>
 % Start state = <n, p, 0>
 % Stop State = <0, p', m>
+\end{multicols}
 
-Small-step Operational semantics:
+%Small-step Operational semantics:
 

### 2012-02-26 rational.tex
copied out the rationals paper
lots of restructuring

--- 2012-02-24 dual.tex
@@
 %! /= = \neq
+%! @@ = \mu
 %! forall = \forall
@@
 
-\title{Functional Pearl --- Programming with \\ 
-  Negative, Fractional, Square Root, and Imaginary Types} 
+\title{Computing in the Field of Rationals}
 \authorinfo{Roshan P. James}
@@
 \begin{abstract}
-  Every functional programmer knows about sum and product types, $a+b$ and
-  $a*b$ respectively. Negative, fractional, square root, and imaginary types,
-  $a-b$, $a/b$, $\sqrt{a}$, and $a+ib$ respectively, are much less known and
-  their computational interpretation is unfamiliar and often complicated. We
-  show that in a programming model in which information is preserved (such as
-  the model introduced in our recent paper on \emph{Information Effects}),
+  Every functional programmer knows about sum and product types, $a+b$
+  and $a*b$ respectively. Negative and fractional types, $a-b$ and
+  $a/b$ respectively, are much less known and their computational
+  interpretation is unfamiliar and often complicated. We show that in
+  a programming model in which information is preserved (such as the
+  model introduced in our recent paper on \emph{Information Effects}),
   these types have particularly intuitive and natural computational
-  interpretations. Intuitively, values of negative types are values that flow
-  ``backwards'' to satisfy demands, values of fractional types are values
-  that represent first-class ``structural'' constraints, and values of square
-  root and imaginary types are values that can be related by arbitrary
-  algebraic constraints. The combination of these negative, fractional,
-  square root, and imaginary types enable greater flexibility in programming
-  by breaking complicated invariants into local ones that can be
-  independently satisfied by a subcomputation. Furthermore, they allow a
-  programmer to express a wide range of programming idioms, including
-  higher-order functions, delimited continuations, speculative computation,
-  logic programming, and more.
+  interpretations. Intuitively, values of negative types are values
+  that flow ``backwards'' to satisfy demands, values of fractional
+  types are values that represent first-class ``structural''
+  constraints.
+  %% , and values of square root and imaginary types are values that can
+  %% be related by arbitrary algebraic constraints.
+  The combination of these negative, fractional, square root, and
+  imaginary types enable greater flexibility in programming by
+  breaking complicated invariants into local ones that can be
+  independently satisfied by a subcomputation. Furthermore, they allow
+  a programmer to express a wide range of programming idioms,
+  including higher-order functions, delimited continuations,
+  speculative computation, logic programming, and more.
 \end{abstract}
@@
 \section{Introduction}
+
+In a technical sense, this paper extends the language of isomorphisms
+{{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
+other systems which have one notion of duality over additive and
+multiplicative componenents, {{langRevEE}} has two notions of duality
+-- an additive duality over the monoid {{(0, +)}} and a multiplicative
+duality over the monoid {{(1, *)}}. Each axis of duality also give us
+a function space and hence {{langRevEE}} has an additive function
+space correspoding to a notion of control or backtracking and a
+multiplicative function space corresponding to a notion of unification
+or constraint satisfaction.
 
@@
 Some background on {{langRev}} because it is mentioned in the next section.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related Work} 
+\label{sec:related}
+
+(NOTE: Lets use this section here to establish a long of ideas and
+intuition. None of these connections have to be formal, but they can
+be concrete.)
+
+Filinski proposes that continuations are a \emph{declarative} concept. He,
+furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+which values and continuations are treated as opposites. This is essentially
+what we are proposing with one fundamental difference: our underlying
+language is not the $\lambda$-calculus but a language of pure isomorphisms in
+which information is preserved. This shift of perspective enables us to
+distill and generalize the duality of values and continuations: in
+particular, in the conventional $\lambda$-calculus setting values and
+continuations can be erased and duplicated which makes it difficult to
+maintain the correspondence between a value and its negative counterpart.
+
+\paragraph*{Continuations.} The idea of using negative types to model
+information flowing backwards, demand for values, continuations, etc. goes
+back to at Filinski's thesis. We recall these connections below but we first
+note that all these systems are complicated because in all these systems
+information can be ignored, destroyed, or duplicated. Clearly the possibility
+of erasure of information would mean that our credit card transaction is
+incorrect. In our work, information is maintained and hence we have a
+guarantee that, in a closed program, the debt must be accounted and paid for.
+
+\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
+In his Masters thesis, Filinski proposes that continuations are a
+\emph{declarative} concept. He, furthermore, introduces a symmetric extension
+of the $\lambda$-calculus in which values and continuations are treated as
+opposites. This is essentially what we are proposing with one fundamental
+difference: our underlying language is not the $\lambda$-calculus but a
+language of pure isomorphisms in which information is preserved. This shift
+of perspective enables us to distill and generalize the duality of values and
+continuations: in particular, in the conventional $\lambda$-calculus setting
+values and continuations can be erased and duplicated which makes it
+difficult to maintain the correspondence between a value and its negative
+counterpart. In contrast, in our setting, one can start from the empty type
+$0$, introduce a positive value and its negative counterpart, and let each of
+these flow in arbitrary ways. The entire framework guarantees that neither
+the value nor its negative counterpart will be deleted or duplicated and
+hence that, in any closed program, the ``debt'' corresponding to the negative
+value is paid off exactly once. The forward and backward executions in our
+framework correspond to call-by-value and call-by-name. This duality was
+observed by Filinski and others following him but it is particularly clean in
+our framework.
+
+\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
+subtraction type because its ``computational interpretation is not
+familiar.'' Curien and Herbelin study duality in classical logic, show that
+it exchanges call-by-value and call-by-name. They extend classical natural
+deduction with subtraction but give it no computational meaning. Crolard (in
+the formulae-as-types interpretation of subtractive logic) address the
+computational interpretation of substraction. He explains the type $A-B$ as
+the of \emph{coroutines} with a local environment of type $A$ and a
+continuation of type $B$. The description is complicated by what is
+essentially the desire to enforce linearity constraints so that coroutines
+cannot access the local environment of other coroutines. Must cite Selinger
+control categories in this context of duality but I am not sure what to say:
+it assumes cartesian closed categories for one thing and not symmetric
+monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
+explain delimited continuations.
+
+\paragraph*{Linear Logic} In accounts that are linear, the value and
+continuation that comprise the substractive type need to be constrained to
+``stay together.'' This can be achieved by various restrictions. In this work
+we have no such constraints, the negative value can flow anywhere. The entire
+system guarantees that any closed program would have to account for it. We
+don't have to introduce special constraints to achieve that. Zeilberger in
+the paper on polarity and the logic of delimited continuations uses polarized
+logic: he shows that if positive and negative values are completely symmetric
+except that answer types are positive, then the framework accommodates
+delimited continuations. But he interprets negative values are control
+operators, or as values defined by the shape of their continuations. We
+simply interpret values of negative type as values flowing in the ``other''
+direction.
+
+\paragraph*{Int Construction.}
+For a traced monoidal category {{C}} the Int construction produces a
+Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
+Further we know that the target of the Int construction is isomorphic
+to the target of \G construction of Abramsky \cite{Abramsky96:0} from
+Haghverdi. However, note that the {{langRevEE}} is not the same as the
+image of the Int construction on {{langRevT}}, since the the later
+lacks a multiplicative tensor that distrbutes over the additive tensor
+in Int {{langRevT}}.
+
+\paragraph*{Recursion.} In our previous work we introduce recursive types and
+trace operators. This is dangerous here because infinite loops allow us to
+prolong paying the debt for as long as we want.
+
+\paragraph*{GoI machines} 
+We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Algebraic Types: Intuition}
-
-Square root and imaginary types have also appeared in the literature: we
-relegate the connections to Sec.~\ref{related} and proceed with a simple
-explanation. We have so far extended the set of types to be the rational
-numbers. Now we will push this and extend the set of types to algebraic
-numbers. In other words, we will allow datatypes defined by arbitrary
-polynomials and allow the roots of such polynomials to be types.
-
-Consider first an example in which we want to compute with the sides of a
-rectangle whose area is 91 and whose length is longer than its width by 6
-units. One can solve the quadratic equation to determine that the sides are 7
-and 13 and proceed. This however prematuraly forces us to globally solve the
-constraint. Instead we can let the two sides of the rectangle be $x$ and
-$x+6$ and use the following equation to capture the desired constraint:
-\[
-x^2 + 6x - 91 = 0
-\]
-The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
-the type $0$. We can now proceed to compute with the unknown $x$, being
-assured that in a closed program, our computation will eventually be
-consistent with the solution of the algebraic equation.
-
-Previously the most famous example of a similar nature is the puzzling
-isomorphism that one can establish between seven binary trees and one.
-A binary tree is defined by the datatype
-\[
-x = 1 + x * x 
-\]
-which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
-algebraic manipulation we can reason as follows:
-\[\begin{array}{rclclclcl}
-x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
-x^6 &=& 1 \\
-x^7 &=& x^6 x &=& x
-\end{array}\]
-Fiore poses the question of why such algebraic manipulation would make sense
-type theoretically but states that even though some of the intermediate steps
-make no sense, the final equivalence is valid and can be used to actually
-construct an isomorphism between $x^7$ the type of seven binary trees 
-and $x$ the type of binary trees.
-
-Discussion of possible polynomials:
-\begin{itemize}
-\item $b=1+1$. Boring.
-\item $b=b+1$. That introduces the natural numbers. No solution for this
-  polynomial over the algebraic numbers. We must extend the numbers with
-  $\omega$ to get a solution. We reject this in this paper and prefer to
-  stick to algebraic numbers. The advantage is that all the isomorphisms are
-  valid numerically. With the above type we could subtract $b$ from both
-  sides to show that $0=1$ which is nonsense. We can however have infinite
-  types as long as they have algebraic solutions.
-\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
-  boolean! If we had superpositions, we could then write a function that
-  performs the square root of negation in the sense that applying it twice
-  would be equivalent to boolean negation.
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Summary}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Additive Duality in {{langRev}} }
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRevEE}} }
 
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
+input {{v1 : b1}} gives us a value {{v2 : b2}} on
+termination. Execution is well defined only if {{b1}} and {{b2}} are
+entirely positive types. Consider the program that
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The ouputs of
+programs that ouput mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Additive Duality in {{langRev}} }
+
+{{eta}}
+
+\begin{center}
+  \includegraphics{diagrams/eta.pdf}
+\end{center}
+
+Note that the connective is a {{+}}, hence only one of the branches
+may be inhabited at any time. Thus the action of {{eta}} is to
+transfer a backward flowing value on one wire to a forward flowing
+value on the other wire.
+
+{{eps}}
+
+\begin{center}
+  \includegraphics{diagrams/eps.pdf}
+\end{center}
+
+Coherence condition
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Multiplicative Duality in {{langRev}} }
+
+\section{Semantics}
 %%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Rules for {{eta}} and {{eps}} }
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Diagrams}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions in {{langRevEE}} }
 
-{{eta}}
+These constructions are applicable to the additive and the
+multiplicative monoid.
 
-\begin{center}
-  \includegraphics{diagrams/eta.pdf}
-\end{center}
-
-Note that the connective is a {{+}}, hence only one of the branches
-may be inhabited at any time. Thus the action of {{eta}} is to
-transfer a backward flowing value on one wire to a forward flowing
-value on the other wire.
-
-{{eps}}
-
-\begin{center}
-  \includegraphics{diagrams/eps.pdf}
-\end{center}
-
-Coherence condition
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-Function
+\paragraph*{Functions}
 
@@
 
-Function application
+\paragraph*{Function application}
 
@@
 
-Function composition
+\paragraph*{Function composition}
 
@@
 
-Delimited continuation
+\paragraph*{Delimited continuation}
 
@@
 
-Trace
+\paragraph*{Trace }
 
@@
 
-Double Negation
+\paragraph*{Principium Contradictiones}
 
@@
 \end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Negation}
 
@@
 
-\paragraph*{Observability.} 
-Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
-input {{v1 : b1}} gives us a value {{v2 : b2}} on
-termination. Execution is well defined only if {{b1}} and {{b2}} are
-entirely positive types. Consider the program that
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Multiplication}
 
-\begin{center}
-  \includegraphics{diagrams/shuffle.pdf}
-\end{center}
-
-We define observables to be only positive types. The ouputs of
-programs that ouput mixed positive and negative types are not
-observable.  Also, programs that input mixed positive and negatives
-types are not executable.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{To think about}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Multiplicative Duality in {{langRev}} }
+\section{Advanced Programming Examples}
+
+\subsection{SAT solver}
+
+\subsection{GoI machine}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Related Work} 
-\label{sec:related}
+\section{The Field of Algebraic Numbers}
 
-Filinski proposes that continuations are a \emph{declarative} concept. He,
-furthermore, introduces a symmetric extension of the $\lambda$-calculus in
-which values and continuations are treated as opposites. This is essentially
-what we are proposing with one fundamental difference: our underlying
-language is not the $\lambda$-calculus but a language of pure isomorphisms in
-which information is preserved. This shift of perspective enables us to
-distill and generalize the duality of values and continuations: in
-particular, in the conventional $\lambda$-calculus setting values and
-continuations can be erased and duplicated which makes it difficult to
-maintain the correspondence between a value and its negative counterpart.
+The field of algebraic numbers consist of the rationals extended with
+solutions to polynomials. In a limited sense {{langRevEE}} with
+recursive types already include polynomials such as the definition of
+binary tree {{@@x.(1+x*x)}} implies the equality {{t = 1 + t*t}} and
+hence {{t^2-t+1=0}}. While this equation does indeed have an
+irrational imaginary root, as we will see -- {{langRevEE}} does not
+have the means of expressing meaningful computation that uses the
+root.
 
-\paragraph*{Continuations.} The idea of using negative types to model
-information flowing backwards, demand for values, continuations, etc. goes
-back to at Filinski's thesis. We recall these connections below but we first
-note that all these systems are complicated because in all these systems
-information can be ignored, destroyed, or duplicated. Clearly the possibility
-of erasure of information would mean that our credit card transaction is
-incorrect. In our work, information is maintained and hence we have a
-guarantee that, in a closed program, the debt must be accounted and paid for.
+TODO.
 
-\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
-In his Masters thesis, Filinski proposes that continuations are a
-\emph{declarative} concept. He, furthermore, introduces a symmetric extension
-of the $\lambda$-calculus in which values and continuations are treated as
-opposites. This is essentially what we are proposing with one fundamental
-difference: our underlying language is not the $\lambda$-calculus but a
-language of pure isomorphisms in which information is preserved. This shift
-of perspective enables us to distill and generalize the duality of values and
-continuations: in particular, in the conventional $\lambda$-calculus setting
-values and continuations can be erased and duplicated which makes it
-difficult to maintain the correspondence between a value and its negative
-counterpart. In contrast, in our setting, one can start from the empty type
-$0$, introduce a positive value and its negative counterpart, and let each of
-these flow in arbitrary ways. The entire framework guarantees that neither
-the value nor its negative counterpart will be deleted or duplicated and
-hence that, in any closed program, the ``debt'' corresponding to the negative
-value is paid off exactly once. The forward and backward executions in our
-framework correspond to call-by-value and call-by-name. This duality was
-observed by Filinski and others following him but it is particularly clean in
-our framework.
+Square root and imaginary types have also appeared in the literature: we
+relegate the connections to Sec.~\ref{related} and proceed with a simple
+explanation. We have so far extended the set of types to be the rational
+numbers. Now we will push this and extend the set of types to algebraic
+numbers. In other words, we will allow datatypes defined by arbitrary
+polynomials and allow the roots of such polynomials to be types.
 
-\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
-subtraction type because its ``computational interpretation is not
-familiar.'' Curien and Herbelin study duality in classical logic, show that
-it exchanges call-by-value and call-by-name. They extend classical natural
-deduction with subtraction but give it no computational meaning. Crolard (in
-the formulae-as-types interpretation of subtractive logic) address the
-computational interpretation of substraction. He explains the type $A-B$ as
-the of \emph{coroutines} with a local environment of type $A$ and a
-continuation of type $B$. The description is complicated by what is
-essentially the desire to enforce linearity constraints so that coroutines
-cannot access the local environment of other coroutines. Must cite Selinger
-control categories in this context of duality but I am not sure what to say:
-it assumes cartesian closed categories for one thing and not symmetric
-monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
-explain delimited continuations.
+Consider first an example in which we want to compute with the sides of a
+rectangle whose area is 91 and whose length is longer than its width by 6
+units. One can solve the quadratic equation to determine that the sides are 7
+and 13 and proceed. This however prematuraly forces us to globally solve the
+constraint. Instead we can let the two sides of the rectangle be $x$ and
+$x+6$ and use the following equation to capture the desired constraint:
+\[
+x^2 + 6x - 91 = 0
+\]
+The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
+the type $0$. We can now proceed to compute with the unknown $x$, being
+assured that in a closed program, our computation will eventually be
+consistent with the solution of the algebraic equation.
 
-\paragraph*{Linear Logic} In accounts that are linear, the value and
-continuation that comprise the substractive type need to be constrained to
-``stay together.'' This can be achieved by various restrictions. In this work
-we have no such constraints, the negative value can flow anywhere. The entire
-system guarantees that any closed program would have to account for it. We
-don't have to introduce special constraints to achieve that. Zeilberger in
-the paper on polarity and the logic of delimited continuations uses polarized
-logic: he shows that if positive and negative values are completely symmetric
-except that answer types are positive, then the framework accommodates
-delimited continuations. But he interprets negative values are control
-operators, or as values defined by the shape of their continuations. We
-simply interpret values of negative type as values flowing in the ``other''
-direction.
+Previously the most famous example of a similar nature is the puzzling
+isomorphism that one can establish between seven binary trees and one.
+A binary tree is defined by the datatype
+\[
+x = 1 + x * x 
+\]
+which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
+algebraic manipulation we can reason as follows:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+Fiore poses the question of why such algebraic manipulation would make sense
+type theoretically but states that even though some of the intermediate steps
+make no sense, the final equivalence is valid and can be used to actually
+construct an isomorphism between $x^7$ the type of seven binary trees 
+and $x$ the type of binary trees.
 
-\paragraph*{Int Construction.}
-For a traced monoidal category {{C}} the Int construction produces a
-Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
-Further we know that the target of the Int construction is isomorphic
-to the target of \G construction of Abramsky \cite{Abramsky96:0} from
-Haghverdi. However, note that the {{langRevEE}} is not the same as the
-image of the Int construction on {{langRevT}}, since the the later
-lacks a multiplicative tensor that distrbutes over the additive tensor
-in Int {{langRevT}}.
-
-\paragraph*{Recursion.} In our previous work we introduce recursive types and
-trace operators. This is dangerous here because infinite loops allow us to
-prolong paying the debt for as long as we want.
-
-\paragraph*{GoI machines} 
-We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
+Discussion of possible polynomials:
+\begin{itemize}
+\item $b=1+1$. Boring.
+\item $b=b+1$. That introduces the natural numbers. No solution for this
+  polynomial over the algebraic numbers. We must extend the numbers with
+  $\omega$ to get a solution. We reject this in this paper and prefer to
+  stick to algebraic numbers. The advantage is that all the isomorphisms are
+  valid numerically. With the above type we could subtract $b$ from both
+  sides to show that $0=1$ which is nonsense. We can however have infinite
+  types as long as they have algebraic solutions.
+\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
+  boolean! If we had superpositions, we could then write a function that
+  performs the square root of negation in the sense that applying it twice
+  would be equivalent to boolean negation.
+\end{itemize}
 

### 2012-02-27 rational.tex
lots of edits

@@
 %! factor = \mathit{factor}
+%! eta+ = \eta^+
+%! eps+ = \epsilon^+
+%! eta* = \eta^*
+%! eps* = \epsilon^*
 %! (o) = \fatsemi
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRev}} }
+
+The presentation here differs from \cite{infeffects} in the sense that
+we have introduced the {{0}} type here, thereby completeing the monoid
+{{(0,+)}} and we present a small-step operational semantics.
+
+%subcode{bnf} include main
+% Combinators, c = iso | c (;) c | c (*) c | c (+) c
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+% Values, v = () | (v, v) | L v | R v
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{ {{langRevEE}} }
@@
 
-{{eta}}
+Additive duality is added to {{langRev}} by addition of negative types
+denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
+extend the language as follows:
 
+%subcode{bnf} include main
+% Base Types, b = ... | -b
+% Values, v = ... | -v
+%
+% Isomorphisms, iso &=& ... | eta+ | eps+
+
+\noindent
+with the types judgements
+
+%subcode{opsem} include main
+% eta+ &: 0 <-> (-b) + b :& eps+
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
+the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
+dualy on the right is {{eps+}} showing the map from {{-b+b}} to
+0. Even though the diagrams below show a {{0}} wire for completeness,
+later diagrams will always drop them in contexts where we can
+implicitely introduce/eliminate then using {{zeroi}}/{{zeroe}}.
+
+\begin{multicols}{2}
 \begin{center}
@@
 
-Note that the connective is a {{+}}, hence only one of the branches
-may be inhabited at any time. Thus the action of {{eta}} is to
-transfer a backward flowing value on one wire to a forward flowing
-value on the other wire.
-
-{{eps}}
-
 \begin{center}
@@
 \end{center}
+\end{multicols}
 
-Coherence condition
+The usual interpretation of the type {{b1+b2}} is that we either have
+an appropriately tagged value of type {{b1}} or an appropriately
+tagged value of type {{b2}}. This interpretation is mainted in the
+precense of {{eta+}} and {{eps+}} in the following sense: a value of
+type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
+switches into a value {{L (-v): -b + b}} that flows backwards on the
+upper wire. The inversion of direction is captured by the negative
+sign on the value.
+
+\paragraph*{Negative Information Flow.} 
+Hence, the intuitive way to read the digrams is that they reverse the
+flow of information -- in normal {{langRev}} circuits information
+flows from left-to-right and {{eps+}} operations cause information to
+flow from right-to-left.  The interpretation of {{eta+}} is precisely
+the opposite: it turns a right-to-left flow of information in a
+cicruit into a left-to-right flow.
+
+\paragraph*{Time Travelling.}
+A time travelling intuition is also applicable here: if view the
+left-to-right direction of information flow as the canonical forward
+direction of information, then the action of {{eps+}} allows values,
+aka information particles, to flow backwards in time. Particles that
+flow backward in time get to see and interact with the history of
+particles that they coexist with i.e. values that they exist
+multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
+(whose construction is shown in Section \ref{sec:neg-constructions}).
+
+\paragraph*{Coherence.}
+The appropriate coherence condition for Compact Closed Categories,
+expressed as the traingle axiom, is satisfied by {{eta+}}/{{eps+}} and
+may be visualized as explained by Selinger
+\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
 
@@
 
+\paragraph*{Non-termination.}
+As is well known in category theory, Compact Closed Categories contain
+trace operations.  In the context of {{langRev}} this implies that
+{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
+(detailed in Section \ref{sec:monoidal-constructions}) hence
+introducing \emph{non-termination} into the language
+\cite{infeffects}.
+
+\paragraph*{Zero Information.}
+The fact that {{eta+}} (and hence {{eps+}}) introduces no information
+can be understood is by viewing {{-b+b}} as a type whose value is
+observable only if you `pay' {{b}} amount of information to the {{-b}}
+branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
+amount of information is returned to you. (TODO. more needs to be said
+here.)
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Semantics}
-%%%%%%%%%%%%%%%
+
 \subsection{Syntax}
@@
 \subsection{Constructions in {{langRevEE}} }
+\label{sec:monoidal-constructions}
 
@@
 \subsection{Constructions Specific to Negation}
+\label{sec:neg-constructions}
 

### 2012-02-27 rational.tex
pi semantics and more

@@
 
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
+%% \newtheorem{theorem}{Theorem}[section]
+%% \newtheorem{lemma}[theorem]{Lemma}
+%% \newtheorem{proposition}[theorem]{Proposition}
+%% \newtheorem{corollary}[theorem]{Corollary}
 
@@
     {\par\addvspace{.5pc}}
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{proposition}[theorem]{Proposition}
+\newenvironment{proof}[1][Proof.]{\begin{trivlist}\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+
+\newcommand{\arrow}[1]{\mathtt{#1}}
 
@@
 %subcode{bnf} include main
+% value types, b ::= 0 | 1 | b + b | b * b 
+% values, v ::= () | left v | right v | (v, v)
+
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%zeroe :&  0 + b <-> b &: zeroi
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist0 :& 0 * b <-> 0 &: factor0
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+
+%subcode{proof} include main
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+
+\begin{definition}{(Syntax of {{langRev}})}
+\label{def:langRev}
+We collect our types, values, and combinators, to get the full language
+definition.
+%subcode{bnf} include main
+% value types, b ::= 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% comb.~types, t ::= b <-> b
+% iso ::= swap+ | assocl+ | assocr+ 
+%     &|& identl* | identr* 
+%     &|& swap* | assocl* | assocr* 
+%     &|& dist | factor 
+% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+\end{definition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Semantics}
+
+%subcode{bnf} include main
 % Combinators, c = iso | c (;) c | c (*) c | c (+) c
@@
 % Stop State = {[c, v, []]}
+
+\begin{scriptsize}
+%subcode{opsem} include main
+%! columnStyle = rlcl
+% swap+ & (left v) &|-->& right v
+% swap+ & (right v) &|-->& left v 
+% assocl+ & (left v1) &|-->& left (left v1)
+% assocl+ & (right (left v2)) &|-->& left (right v2)
+% assocl+ & (right (right v3)) &|-->& right v3 
+% assocr+ & (left (left v1)) &|-->& left v1
+% assocr+ & (left (right v2)) &|-->& right (left v2)
+% assocr+ & (right v3) &|-->& right (right v3)
+% identl* & ((), v) &|-->& v 
+% identr* & v &|-->& ((), v) 
+% swap* & (v1, v2) &|-->& (v2, v1) 
+% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
+% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
+% dist & (left v1, v3) &|-->& left (v1, v3)
+% dist & (right v2, v3) &|-->& right (v2, v3)
+% factor & (left (v1, v3)) &|-->& (left v1, v3) 
+% factor & (right (v2, v3)) &|-->& (right v2, v3)   
+\end{scriptsize}
 
@@
 
+(TODO: So this is a really bad explanation. But this section is really
+important and we need to explain carefully. Maybe an analogy can be
+drawn to Cayley diagrams for groups to point out that the notion of
+`i' is entirely artificial, but it is essential to the discourse.)
+
 \begin{center}
@@
 %subcode{bnf} include main
-% Base Types, b = ... | -b
+% Value Types, b = ... | -b
 % Values, v = ... | -v
@@
 
-Negation distributes over {{+}}.
+\paragraph*{Negation distributes over {{+}}. }
 
@@
 
-{{eps_{fst} }}
+\paragraph*{ {{eps_{fst} }} }
 
@@
 
-Lifting negation out of {{*}}.
+\paragraph*{Lifting negation out of {{*}}. }
 
@@
 
-Lifting a operation of postive types to negated types:
+\paragraph*{Lifting a operation of postive types to negated types}
 

### 2012-02-27 rational.tex
multiplicative

@@
 %! eta* = \eta^*
+%! ^^^ = ^{-1}
 %! eps* = \epsilon^*
@@
 
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoid}
+{{langRev}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:groupoid}
+{{langRev}} is a dagger symmetric monoidal category. 
+\end{proposition}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Multiplicative Duality in {{langRev}} }
+
+%subcode{bnf} include main
+% Value Types, b = ... | b^^^
+% Values, v = ... | v^^^
+%
+% Isomorphisms, iso &=& ... | eta* | eps*
+
+\noindent
+with the types judgements
+
+%subcode{opsem} include main
+% eta* &: 1 <-> b^^^ * b :& eps*
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- v^^^ : b^^^
+
+Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
+{{a/b}}.
+
+\paragraph*{Unification.}
+
+\paragraph*{Entanglement.}
+
+\paragraph*{Coherence.}
+
+\paragraph*{Annihilation.}
+
+\paragraph*{Zero Information.}
 
@@
 \end{itemize}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoid}
+{{langRevEE}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:groupoid}
+{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
+\end{proposition}
+
+Note: We use `Monoidal Closed' in the sense of Compact Closed. While
+the category has an multiplicative tensor that distributes over an
+additive tensor, it does not have injections or projections and hence
+does not have categorical sums and products.
 

### 2012-02-27 rational.tex
_

@@
 \begin{abstract}
-  Every functional programmer knows about sum and product types, $a+b$
-  and $a*b$ respectively. Negative and fractional types, $a-b$ and
-  $a/b$ respectively, are much less known and their computational
-  interpretation is unfamiliar and often complicated. We show that in
-  a programming model in which information is preserved (such as the
-  model introduced in our recent paper on \emph{Information Effects}),
-  these types have particularly intuitive and natural computational
-  interpretations. Intuitively, values of negative types are values
-  that flow ``backwards'' to satisfy demands, values of fractional
-  types are values that represent first-class ``structural''
-  constraints.
-  %% , and values of square root and imaginary types are values that can
-  %% be related by arbitrary algebraic constraints.
-  The combination of these negative, fractional, square root, and
-  imaginary types enable greater flexibility in programming by
-  breaking complicated invariants into local ones that can be
-  independently satisfied by a subcomputation. Furthermore, they allow
-  a programmer to express a wide range of programming idioms,
-  including higher-order functions, delimited continuations,
+  Every functional programmer knows about sum and product types,
+  {{a+b}} and {{a*b}} respectively. Negative and fractional types,
+  $a-b$ and $a/b$ respectively, are much less known and their
+  computational interpretation is unfamiliar and often complicated. We
+  show that in a programming model in which information is preserved
+  (such as the model introduced in our recent paper on
+  \emph{Information Effects}), these types have particularly intuitive
+  and natural computational interpretations. Intuitively, values of
+  negative types are values that flow ``backwards'' to satisfy
+  demands, values of fractional types are values that represent
+  first-class ``structural'' constraints.  The combination of these
+  negative and fractional types enable greater flexibility in
+  programming by breaking complicated invariants into local ones that
+  can be independently satisfied by a subcomputation. Furthermore,
+  they allow a programmer to express a wide range of programming
+  idioms, including higher-order functions, delimited continuations,
   speculative computation, logic programming, and more.
+
 \end{abstract}

### 2012-02-28 rational.tex
organization and abstract

--- 2012-02-27 rational.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{document}
@@
 
-\title{Computing in the Field of Rationals}
+\title{The Two Dualities of Computation: Negative and Fractional Types}
 \authorinfo{Roshan P. James}
@@
 \begin{abstract}
-  Every functional programmer knows about sum and product types, $a+b$
-  and $a*b$ respectively. Negative and fractional types, $a-b$ and
-  $a/b$ respectively, are much less known and their computational
-  interpretation is unfamiliar and often complicated. We show that in
-  a programming model in which information is preserved (such as the
-  model introduced in our recent paper on \emph{Information Effects}),
-  these types have particularly intuitive and natural computational
-  interpretations. Intuitively, values of negative types are values
-  that flow ``backwards'' to satisfy demands, values of fractional
-  types are values that represent first-class ``structural''
-  constraints.
-  %% , and values of square root and imaginary types are values that can
-  %% be related by arbitrary algebraic constraints.
-  The combination of these negative, fractional, square root, and
-  imaginary types enable greater flexibility in programming by
-  breaking complicated invariants into local ones that can be
-  independently satisfied by a subcomputation. Furthermore, they allow
-  a programmer to express a wide range of programming idioms,
-  including higher-order functions, delimited continuations,
-  speculative computation, logic programming, and more.
+  Every functional programmer knows about sum and product types, {{a+b}} and
+  {{a*b}} respectively. Negative and fractional types, $a-b$ and $a/b$
+  respectively, are much less known and their computational interpretation is
+  unfamiliar and often complicated. We show that in a programming model in
+  which information is preserved (such as the model introduced in our recent
+  paper on \emph{Information Effects}), these types have particularly
+  intuitive and natural computational interpretations. Intuitively, values of
+  negative types are values that flow ``backwards'' to satisfy demands and
+  values of fractional types are values that impose constraints on their
+  context.  The combination of these negative and fractional types enables
+  greater flexibility in programming by breaking global invariants into local
+  ones that can be autonomously satisfied by a subcomputation. Furthermore,
+  they allow a programmer to express a wide range of programming idioms,
+  including higher-order functions, delimited continuations, speculative
+  computation, logic programming, and more.
 \end{abstract}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \category{D.3.1}{Formal Definitions and Theory}{}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Field of Rational Types: Intuitions}
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Negative Types}
+
+Consider the following two ways of purchasing an item that costs \$20.00:
+\begin{enumerate}
+\item You give the seller a \$20.00 bill.
+\item You use a credit card to pay the seller.
+\end{enumerate}
+In both cases, the seller receives the money immediately but there is a
+subtle difference. In the first transaction, the money received by the seller
+corresponds to a value that has been produced earlier in the computation. In
+the second transaction, the money may or may not exist yet: computationally,
+a \emph{debt} is generated to compensate for the money received by the seller
+and this debt travels ``backwards'' towards the bank where it is (hopefully)
+reconciled.
+
+The example suggests that the ability to consider values flowing backwards
+enriches our computational model. This observation goes back to Filinski's
+Masters thesis where continuations are identified with these negative values.
+We discuss the connections to Filinski's work and others in more detail in
+Sec.~\ref{sec:related}. For now, we note that in a conventional programming
+language in which values can be copied and deleted at will, extreme care is
+needed to keep track of negative values. Indeed, in the example above, if it
+were possible to simply delete the variable corresponding to the debt, we
+would have produced money out of nothing.
+
+For this reason, our treatment of negative values in the context of
+{{langRev}} is particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
+right-to-left direction. In more detail, we can model the credit card
+transaction above using the following program (shown diagrammatically):
+
+FIGURE
+
+CLEAN UP the following based on the figure. In contrast, in our setting, one
+can start from the empty type $0$, introduce a positive value and its
+negative counterpart, and let each of these flow in arbitrary ways. The
+entire framework guarantees that neither the value nor its negative
+counterpart will be deleted or duplicated and hence that, in any closed
+program, the ``debt'' corresponding to the negative value is paid off exactly
+once. Computationally we model the first transaction using essentially the
+identity function which receives a \$20.00 bill as input from the buyer and
+propagates it on its output to the seller. The second transaction is more
+complicated. We model it as shown in the circuit below: There are two ways to
+understand this circuit that are both quite instructive. Let's first examine
+the type structure of each of the combinators that comprise the circuit. The
+first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
+produced from nothing so to speak by generating an equivalent debt that
+travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
+the execution. It goes forward in time so to speak, comes back, and then goes
+forwards again.
+
+It is critical that the framework in which the negative types are introduced
+is a framework in which all information is preserved, with no duplication or
+erasure. This guarantees that the generated debts are paid once and exactly
+once.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Fractional Types}
+
+The type $a/b$ is prominent in the Lambek-Grishin calculus which is
+extensively used in computational linguistics. In the common interpretation,
+a value of type $a/b$ is a value of some type $c$ such that when put in a
+context of type $b$, the result is a value of type $a$. For example, assuming
+contexts are represented as functions, a value of type $\texttt{bool} /
+(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
+$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
+in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
+type $\texttt{bool}$.
+
+In our case, the situation is simpler. If the goal is to produce a value of
+type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
+have type {{(a * b) / b}}. In most settings, this type would be equivalent to
+{{a}}. However in our setting, the constraint that this value must eventually
+fit in a larger value that supplies the {{b}} is explicitly recorded and must
+be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
+which when used in the left-to-right direction allows the creation of a value
+and a corresponding constraint on the context.  Both the value and the
+constraint can flow in arbitrary ways during the computation but eventually
+in a closed program with no ``dangling'' constraints, the isomorphism should
+be used in the right-to-left direction to resolve the constraints.  To
+understand the computational interpretation, consider the following example.
+
+The goal is to produce a value of type $(\texttt{bool} \times
+\texttt{int})$. One part of the program, $e_1$, knows how to produce the
+value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
+to produce the value of type $\texttt{bool}$ (say {{true}}). Computationally,
+we model this as follows. The first subcomputation $e_1$ produces {{((),3)}}
+and uses the isomorphism to convert {{()}} to {{(alpha,1/alpha)}} where
+{{alpha}} is a yet-unknown boolean value. Reshuffling the components, $e_1$
+produces {{((alpha,3), 1/alpha)}}. Similarly, $e_2$ can independently produce
+{{((true,beta),1/beta)}} where {{beta}} is a yet-unknown integer value.
+When the two values meet, we can group the components as follows:
+\[
+{{(alpha,beta)}}  {{(3,1/beta)}}  {{(true,1/alpha)}}
+\]
+Using the isomorphism in the right-to-left direction on the last two pairs,
+forces {{alpha}} to be resolved to {{true}} and forces {{beta}} to be
+resolved to {{3}}. Both of these pairs then become {{()}} and be absorbed. We
+are left with the pair {{(true,3)}} as desired.
+
+To summarize, one can think of a value of type $a/b$ as a value that imposes
+a constraint on its context of use: it is a value that requires its context
+to be of type $b$ and only if that condition is satisfied, can the value be
+considered as a value of type $a$. In the simplest case, the type $1/b$ just
+constraints the context to be of type $b$. By allowing the fractional types
+to be first-class, we can separate the generation of constraints from their
+use. Both the value and the constraint can flow in arbitrary ways during the
+computation but eventually in a closed program with no ``dangling''
+constraints, the isomorphism should be used in the right-to-left direction to
+resolve the constraints.
+
+NEED to work out the example in detail (or perhaps another example that's
+better.
+
+Again it is critical that the framework in which the fractional types are
+introduced is a framework in which all information is preserved, with no
+duplication or erasure. This guarantees that the generated constraints are
+satisfied once and exactly once.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Summary}
+
+To summarize negative, fractional, square root, and imaginary types all make
+sense. What they help you accomplish as a programmer is to disassociate
+global invariants into local ones that can be satisfied independently by
+subcomputations with no synchronization or communication. A computation
+producing something of type $a/b$ does not need to concern itself with who is
+going to supply the missing $b$: it just does its part. Conversely faced with
+a complicated task, a computation might decide to break it into pieces and
+demand these pieces using negative types.
+
+It is no surprise that these types are closely related to quantum mechanics
+and that they give us the feel that this is how nature computes. This is
+speculation however.
+
+In any case, in a framework where information can be copied and deleted, none
+of this makes much sense. It is critical that these constraints and demands
+can neither be duplicated nor erased.  This gives us the maximum
+``parallelism'' possible.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRev}} }
+
+The presentation here differs from \cite{infeffects} in the sense that
+we have introduced the {{0}} type here, thereby completeing the monoid
+{{(0,+)}} and we present a small-step operational semantics.
+
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b + b | b * b 
+% values, v ::= () | left v | right v | (v, v)
+
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%zeroe :&  0 + b <-> b &: zeroi
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist0 :& 0 * b <-> 0 &: factor0
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+
+%subcode{proof} include main
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+
+\begin{definition}{(Syntax of {{langRev}})}
+\label{def:langRev}
+We collect our types, values, and combinators, to get the full language
+definition.
+%subcode{bnf} include main
+% value types, b ::= 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% comb.~types, t ::= b <-> b
+% iso ::= swap+ | assocl+ | assocr+ 
+%     &|& identl* | identr* 
+%     &|& swap* | assocl* | assocr* 
+%     &|& dist | factor 
+% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+\end{definition}
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Semantics}
+
+%subcode{bnf} include main
+% Combinators, c = iso | c (;) c | c (*) c | c (+) c
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+% Values, v = () | (v, v) | L v | R v
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+\begin{scriptsize}
+%subcode{opsem} include main
+%! columnStyle = rlcl
+% swap+ & (left v) &|-->& right v
+% swap+ & (right v) &|-->& left v 
+% assocl+ & (left v1) &|-->& left (left v1)
+% assocl+ & (right (left v2)) &|-->& left (right v2)
+% assocl+ & (right (right v3)) &|-->& right v3 
+% assocr+ & (left (left v1)) &|-->& left v1
+% assocr+ & (left (right v2)) &|-->& right (left v2)
+% assocr+ & (right v3) &|-->& right (right v3)
+% identl* & ((), v) &|-->& v 
+% identr* & v &|-->& ((), v) 
+% swap* & (v1, v2) &|-->& (v2, v1) 
+% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
+% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
+% dist & (left v1, v3) &|-->& left (v1, v3)
+% dist & (right v2, v3) &|-->& right (v2, v3)
+% factor & (left (v1, v3)) &|-->& (left v1, v3) 
+% factor & (right (v2, v3)) &|-->& (right v2, v3)   
+\end{scriptsize}
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoid}
+{{langRev}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:groupoid}
+{{langRev}} is a dagger symmetric monoidal category. 
+\end{proposition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRevEE}} }
+
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
+input {{v1 : b1}} gives us a value {{v2 : b2}} on
+termination. Execution is well defined only if {{b1}} and {{b2}} are
+entirely positive types. Consider the program that
+
+(TODO: So this is a really bad explanation. But this section is really
+important and we need to explain carefully. Maybe an analogy can be
+drawn to Cayley diagrams for groups to point out that the notion of
+`i' is entirely artificial, but it is essential to the discourse.)
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The ouputs of
+programs that ouput mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Additive Duality in {{langRev}} }
+
+Additive duality is added to {{langRev}} by addition of negative types
+denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
+extend the language as follows:
+
+%subcode{bnf} include main
+% Value Types, b = ... | -b
+% Values, v = ... | -v
+%
+% Isomorphisms, iso &=& ... | eta+ | eps+
+
+\noindent
+with the types judgements
+
+%subcode{opsem} include main
+% eta+ &: 0 <-> (-b) + b :& eps+
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
+the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
+dualy on the right is {{eps+}} showing the map from {{-b+b}} to
+0. Even though the diagrams below show a {{0}} wire for completeness,
+later diagrams will always drop them in contexts where we can
+implicitely introduce/eliminate then using {{zeroi}}/{{zeroe}}.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps.pdf}
+\end{center}
+\end{multicols}
+
+The usual interpretation of the type {{b1+b2}} is that we either have
+an appropriately tagged value of type {{b1}} or an appropriately
+tagged value of type {{b2}}. This interpretation is mainted in the
+precense of {{eta+}} and {{eps+}} in the following sense: a value of
+type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
+switches into a value {{L (-v): -b + b}} that flows backwards on the
+upper wire. The inversion of direction is captured by the negative
+sign on the value.
+
+\paragraph*{Negative Information Flow.} 
+Hence, the intuitive way to read the digrams is that they reverse the
+flow of information -- in normal {{langRev}} circuits information
+flows from left-to-right and {{eps+}} operations cause information to
+flow from right-to-left.  The interpretation of {{eta+}} is precisely
+the opposite: it turns a right-to-left flow of information in a
+cicruit into a left-to-right flow.
+
+\paragraph*{Time Travelling.}
+A time travelling intuition is also applicable here: if view the
+left-to-right direction of information flow as the canonical forward
+direction of information, then the action of {{eps+}} allows values,
+aka information particles, to flow backwards in time. Particles that
+flow backward in time get to see and interact with the history of
+particles that they coexist with i.e. values that they exist
+multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
+(whose construction is shown in Section \ref{sec:neg-constructions}).
+
+\paragraph*{Coherence.}
+The appropriate coherence condition for Compact Closed Categories,
+expressed as the traingle axiom, is satisfied by {{eta+}}/{{eps+}} and
+may be visualized as explained by Selinger
+\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+\paragraph*{Non-termination.}
+As is well known in category theory, Compact Closed Categories contain
+trace operations.  In the context of {{langRev}} this implies that
+{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
+(detailed in Section \ref{sec:monoidal-constructions}) hence
+introducing \emph{non-termination} into the language
+\cite{infeffects}.
+
+\paragraph*{Zero Information.}
+The fact that {{eta+}} (and hence {{eps+}}) introduces no information
+can be understood is by viewing {{-b+b}} as a type whose value is
+observable only if you `pay' {{b}} amount of information to the {{-b}}
+branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
+amount of information is returned to you. (TODO. more needs to be said
+here.)
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection{Multiplicative Duality in {{langRev}} }
+
+%subcode{bnf} include main
+% Value Types, b = ... | b^^^
+% Values, v = ... | v^^^
+%
+% Isomorphisms, iso &=& ... | eta* | eps*
+
+\noindent
+with the types judgements
+
+%subcode{opsem} include main
+% eta* &: 1 <-> b^^^ * b :& eps*
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- v^^^ : b^^^
+
+Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
+{{a/b}}.
+
+\paragraph*{Unification.}
+
+\paragraph*{Entanglement.}
+
+\paragraph*{Coherence.}
+
+\paragraph*{Annihilation.}
+
+\paragraph*{Zero Information.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Semantics}
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Syntax}
+
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
+%
+% Sequential Contexts, P, F = [] | P:c
+% Parallel Contexts, D = [] 
+%                     &|& D:(P| [](+)c| F)
+%                     &|& D:(P| c(+)[]| F)
+%                     &|& D:(P| [] (*) c v| F)
+%                     &|& D:(P| [] (*) v c| F)
+%                     &|& D:(P| c v (*)  []| F)
+%                     &|& D:(P| v c (*) []| F)
+%
+%
+% Machine States  = D[P| c v| F] | D[P| v c| F]
+% Start State  = [][[]| v c| []]
+% Stop State  = [][P| c v| []]
+
+Combinator reconstruction, {{P[c]}}:
+%subcode{opsem} include main
+% [](c) '= c
+% P:c'(c) '= P(c'(;)c)
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection{Operational Semantics}
+
+The small step semantics present for {{langRev}} below work
+symmetrically for forward and backward evaluation.
+
+\begin{itemize}
+\item 
+Basic reduction of an isomorphism. Note that the evaluation leaves the
+adjoint of the combinator behind. This will become important when we
+reverse the direction of computation.
+%% %subcode{opsem} include main
+%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
+
+%subcode{opsem} include main
+% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
+
+\item
+Sequencing involves pushing and popping from the Future and Past
+continuations:
+%subcode{opsem} include main
+% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
+% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
+
+\item
+Parallel composition, captures the current Future and Past and extends
+the parallel context.
+%subcode{opsem} include main
+% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
+% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
+% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
+% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
+
+\item
+Similarly for products:
+%subcode{opsem} include main
+% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
+% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
+% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
+
+and symmetrical rules for evaluation along the second branch. 
+%subcode{opsem} include main
+% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
+% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
+
+The later two rules will be relevant only for reverse execution.
+
+\end{itemize}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoid}
+{{langRevEE}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:groupoid}
+{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
+\end{proposition}
+
+Note: We use `Monoidal Closed' in the sense of Compact Closed. While
+the category has an multiplicative tensor that distributes over an
+additive tensor, it does not have injections or projections and hence
+does not have categorical sums and products.
+
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Rules for {{eta}} and {{eps}} }
+
+The operation {{eps}} reverses the direction of a particle by
+reversing the world.
+
+Note that we deviate from the categorical definition of {{eta}} and
+{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
+choice of {{eta}}. This however does not affect us, because we deal
+with a symmetric category.
+
+\begin{itemize}
+\item Grammar
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v | -v
+% Isomorphisms, iso &=& ... | eta | eps
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
+
+\item
+Type judgement.
+%subcode{opsem} include main
+% eta &: 0 <-> (-b) + b :& eps
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+\item
+Operational Semantics.
+%subcode{opsem} include main
+% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
+% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
+
+Note: there is NO reduction rule for {{eta}}.
+
+\item
+The adjoint of a parallel context is defined to be:
+%subcode{opsem} include main
+% []{dagger} '= []
+% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
+% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
+% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
+% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
+% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
+% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions in {{langRevEE}} }
+\label{sec:monoidal-constructions}
+
+These constructions are applicable to the additive and the
+multiplicative monoid.
+
+\paragraph*{Functions}
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
+
+\paragraph*{Function application}
+
+\begin{center}
+  \includegraphics{diagrams/apply1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/apply2.pdf}
+\end{center}
+
+\paragraph*{Function composition}
+
+\begin{center}
+  \includegraphics{diagrams/compose1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/compose.pdf}
+\end{center}
+
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+\paragraph*{Delimited continuation}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+\paragraph*{Trace }
+
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+\paragraph*{Principium Contradictiones}
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Negation}
+\label{sec:neg-constructions}
+
+\paragraph*{Negation distributes over {{+}}. }
+
+{{-(b1 + b2) <-> (b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+\paragraph*{ {{eps_{fst} }} }
+
+{{(-b1)*b2 + b1*b2 <-> 0}}
+
+\begin{center}
+  \includegraphics{diagrams/eps_fst.pdf}
+\end{center}
+
+\paragraph*{Lifting negation out of {{*}}. }
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/mult_neg.pdf}
+\end{center}
+
+The following isomoprhism can be constructed similarly
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
+
+\paragraph*{Lifting a operation of postive types to negated types}
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Multiplication}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{To think about}
+
+\begin{itemize}
+
+\item If we built an effect over {{langRevEE}}, say {{create}} and
+  {{erase}}, are effects structured by an arrow or a monad now?
+
+\item Which operations can be lifted to work on negative types?
+
+\item Can the operational semantics for {{langRevEE}} be an
+  interpreter that is implemented in {{langRevT}} (similar to how the
+  the tree traversal interpreter was implemented). This would imply
+  the existence of a more powerful construction than Int, wherein the
+  products would also be preserved. This is possibly worth a paper in
+  itself.
+
+\item These functions aren't really values. There is no value one can
+  produce that denotes a function. These functions are the ability to
+  transform a value - the possibility of transforming a value.  In the
+  product encoding of environments, there is no value one can assign
+  to a variable such that it denotes a function.
+
+Actually this is possible. Consider two functions {{f : b1 <-> b2}}
+and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
+discriminate them. Hence the {{boo}} is the first class representation
+of the functions and can be thought of as the address of the function.
+
+\begin{center}
+  \includegraphics{diagrams/dispatch.pdf}
+\end{center}
+
+\item It is not fair to say that negative types flow backwards. The
+  following circuits are valid in {{langRevEE}}. It is however proper
+  to say that for any type {{b}} that flows in a direction, the type
+  {{-b}} flows in the reverse direction.
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit2.pdf}
+\end{center}
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Advanced Programming Examples}
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{SAT solver}
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{GoI machine}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Related Work} 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Negative Types: Intuition}
-
-Consider the following two ways of purchasing an item that costs \$20.00:
-\begin{enumerate}
-\item You give the seller a \$20.00 bill.
-\item You use a credit card to pay the seller.
-\end{enumerate}
-In both cases, the seller receives the money immediately but there is a
-subtle difference. In the first transaction, the money received by the seller
-corresponds to a value that has been produced earlier in the computation. In
-the second transaction, the money may or may not exist yet: computationally,
-a \emph{debt} is generated to compensate for the money received by the seller
-and this debt travels ``backwards'' towards the bank where it is (hopefully)
-reconciled.
-
-The example suggests that the ability to consider values flowing backwards
-enriches our computational model. This observation goes back to Filinski's
-Masters thesis where continuations are identified with these negative values.
-We discuss the connections to Filinski's work and others in more detail in
-Sec.~\ref{sec:related}. For now, we note that in a conventional programming
-language in which values can be copied and deleted at will, extreme care is
-needed to keep track of negative values. Indeed, in the example above, if it
-were possible to simply delete the variable corresponding to the debt, we
-would have produced money out of nothing.
-
-For this reason, our treatment of negative values in the context of
-{{langRev}} is particularly simple. We will have a type $0$ and an
-isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-allows us to create a value and a corresponding debt out of nothing. Both the
-value and its negative counterpart can flow anywhere in the system. Because
-information is preserved, a closed program (which does not have any
-``dangling'' negative values) will eventually have to match the negative
-value with some corresponding value, effectively using the isomorphism in the
-right-to-left direction. In more detail, we can model the credit card
-transaction above using the following program (shown diagrammatically):
-
-FIGURE
-
-CLEAN UP the following based on the figure. In contrast, in our setting, one
-can start from the empty type $0$, introduce a positive value and its
-negative counterpart, and let each of these flow in arbitrary ways. The
-entire framework guarantees that neither the value nor its negative
-counterpart will be deleted or duplicated and hence that, in any closed
-program, the ``debt'' corresponding to the negative value is paid off exactly
-once. Computationally we model the first transaction using essentially the
-identity function which receives a \$20.00 bill as input from the buyer and
-propagates it on its output to the seller. The second transaction is more
-complicated. We model it as shown in the circuit below: There are two ways to
-understand this circuit that are both quite instructive. Let's first examine
-the type structure of each of the combinators that comprise the circuit. The
-first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
-produced from nothing so to speak by generating an equivalent debt that
-travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
-the execution. It goes forward in time so to speak, comes back, and then goes
-forwards again.
-
-It is critical that the framework in which the negative types are introduced
-is a framework in which all information is preserved, with no duplication or
-erasure. This guarantees that the generated debts are paid once and exactly
-once.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Fractional Types: Intuition}
-
-The type $a/b$ is prominent in the Lambek-Grishin calculus which is
-extensively used in computational linguistics. In the common interpretation,
-a value of type $a/b$ is a value of some type $c$ such that when put in a
-context of type $b$, the result is a value of type $a$. For example, assuming
-contexts are represented as functions, a value of type $\texttt{bool} /
-(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
-$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
-in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
-type $\texttt{bool}$.
-
-In our case, the situation is simpler. If the goal is to produce a value of
-type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
-have type {{(a * b) / b}}. In most settings, this type would be equivalent to
-{{a}}. However in our setting, the constraint that this value must eventually
-fit in a larger value that supplies the {{b}} is explicitly recorded and must
-be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
-which when used in the left-to-right direction allows the creation of a value
-and a corresponding constraint on the context.  Both the value and the
-constraint can flow in arbitrary ways during the computation but eventually
-in a closed program with no ``dangling'' constraints, the isomorphism should
-be used in the right-to-left direction to resolve the constraints.  To
-understand the computational interpretation, consider the following example.
-
-The goal is to produce a value of type $(\texttt{bool} \times
-\texttt{int})$. One part of the program, $e_1$, knows how to produce the
-value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
-to produce the value of type $\texttt{bool}$ (say {{true}}). Computationally,
-we model this as follows. The first subcomputation $e_1$ produces {{((),3)}}
-and uses the isomorphism to convert {{()}} to {{(alpha,1/alpha)}} where
-{{alpha}} is a yet-unknown boolean value. Reshuffling the components, $e_1$
-produces {{((alpha,3), 1/alpha)}}. Similarly, $e_2$ can independently produce
-{{((true,beta),1/beta)}} where {{beta}} is a yet-unknown integer value.
-When the two values meet, we can group the components as follows:
-\[
-{{(alpha,beta)}}  {{(3,1/beta)}}  {{(true,1/alpha)}}
-\]
-Using the isomorphism in the right-to-left direction on the last two pairs,
-forces {{alpha}} to be resolved to {{true}} and forces {{beta}} to be
-resolved to {{3}}. Both of these pairs then become {{()}} and be absorbed. We
-are left with the pair {{(true,3)}} as desired.
-
-To summarize, one can think of a value of type $a/b$ as a value that imposes
-a constraint on its context of use: it is a value that requires its context
-to be of type $b$ and only if that condition is satisfied, can the value be
-considered as a value of type $a$. In the simplest case, the type $1/b$ just
-constraints the context to be of type $b$. By allowing the fractional types
-to be first-class, we can separate the generation of constraints from their
-use. Both the value and the constraint can flow in arbitrary ways during the
-computation but eventually in a closed program with no ``dangling''
-constraints, the isomorphism should be used in the right-to-left direction to
-resolve the constraints.
-
-NEED to work out the example in detail (or perhaps another example that's
-better.
-
-Again it is critical that the framework in which the fractional types are
-introduced is a framework in which all information is preserved, with no
-duplication or erasure. This guarantees that the generated constraints are
-satisfied once and exactly once.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary}
-
-To summarize negative, fractional, square root, and imaginary types all make
-sense. What they help you accomplish as a programmer is to disassociate
-global invariants into local ones that can be satisfied independently by
-subcomputations with no synchronization or communication. A computation
-producing something of type $a/b$ does not need to concern itself with who is
-going to supply the missing $b$: it just does its part. Conversely faced with
-a complicated task, a computation might decide to break it into pieces and
-demand these pieces using negative types.
-
-It is no surprise that these types are closely related to quantum mechanics
-and that they give us the feel that this is how nature computes. This is
-speculation however.
-
-In any case, in a framework where information can be copied and deleted, none
-of this makes much sense. It is critical that these constraints and demands
-can neither be duplicated nor erased.  This gives us the maximum
-``parallelism'' possible.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ {{langRev}} }
-
-The presentation here differs from \cite{infeffects} in the sense that
-we have introduced the {{0}} type here, thereby completeing the monoid
-{{(0,+)}} and we present a small-step operational semantics.
-
-%subcode{bnf} include main
-% value types, b ::= 0 | 1 | b + b | b * b 
-% values, v ::= () | left v | right v | (v, v)
-
-%subcode{bnf} include main
-%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
-%zeroe :&  0 + b <-> b &: zeroi
-%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
-%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
-%identl* :&  1 * b <-> b &: identr*
-%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
-%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
-%dist0 :& 0 * b <-> 0 &: factor0
-%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-
-%subcode{proof} include main
-%@  ~
-%@@ id : b <-> b 
-%
-%@ c : b1 <-> b2
-%@@ sym c : b2 <-> b1
-%
-%@ c1 : b1 <-> b2
-%@ c2 : b2 <-> b3
-%@@ c1(;)c2 : b1 <-> b3
-%---
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
-%
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
-
-\begin{definition}{(Syntax of {{langRev}})}
-\label{def:langRev}
-We collect our types, values, and combinators, to get the full language
-definition.
-%subcode{bnf} include main
-% value types, b ::= 1 | b+b | b*b 
-% values, v ::= () | left v | right v | (v,v) 
-%
-% comb.~types, t ::= b <-> b
-% iso ::= swap+ | assocl+ | assocr+ 
-%     &|& identl* | identr* 
-%     &|& swap* | assocl* | assocr* 
-%     &|& dist | factor 
-% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-\end{definition}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Semantics}
-
-%subcode{bnf} include main
-% Combinators, c = iso | c (;) c | c (*) c | c (+) c
-% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
-%                  &|& LeftTimes cc c v | RightTimes c v cc 
-%                  &|& LeftPlus cc c | RightPlus c cc 
-% Values, v = () | (v, v) | L v | R v
-%
-% Machine states = <c, v, cc> | {[c, v, cc]}
-% Start state = <c, v, []> 
-% Stop State = {[c, v, []]}
-
-\begin{scriptsize}
-%subcode{opsem} include main
-%! columnStyle = rlcl
-% swap+ & (left v) &|-->& right v
-% swap+ & (right v) &|-->& left v 
-% assocl+ & (left v1) &|-->& left (left v1)
-% assocl+ & (right (left v2)) &|-->& left (right v2)
-% assocl+ & (right (right v3)) &|-->& right v3 
-% assocr+ & (left (left v1)) &|-->& left v1
-% assocr+ & (left (right v2)) &|-->& right (left v2)
-% assocr+ & (right v3) &|-->& right (right v3)
-% identl* & ((), v) &|-->& v 
-% identr* & v &|-->& ((), v) 
-% swap* & (v1, v2) &|-->& (v2, v1) 
-% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
-% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
-% dist & (left v1, v3) &|-->& left (v1, v3)
-% dist & (right v2, v3) &|-->& right (v2, v3)
-% factor & (left (v1, v3)) &|-->& (left v1, v3) 
-% factor & (right (v2, v3)) &|-->& (right v2, v3)   
-\end{scriptsize}
-
-%subcode{opsem} include main
-%! columnStyle = rclr
-% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
-% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
-% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
-% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
-% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
-% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
-% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
-% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
-% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
-% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
-% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrev}
-{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
-\end{proposition}
-
-\begin{proposition}[Groupoid]
-\label{prop:groupoid}
-{{langRev}} is a groupoid. 
-\end{proposition}
-
-\begin{proposition}
-\label{prop:groupoid}
-{{langRev}} is a dagger symmetric monoidal category. 
-\end{proposition}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ {{langRevEE}} }
-
-\paragraph*{Observability.} 
-Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
-input {{v1 : b1}} gives us a value {{v2 : b2}} on
-termination. Execution is well defined only if {{b1}} and {{b2}} are
-entirely positive types. Consider the program that
-
-(TODO: So this is a really bad explanation. But this section is really
-important and we need to explain carefully. Maybe an analogy can be
-drawn to Cayley diagrams for groups to point out that the notion of
-`i' is entirely artificial, but it is essential to the discourse.)
-
-\begin{center}
-  \includegraphics{diagrams/shuffle.pdf}
-\end{center}
-
-We define observables to be only positive types. The ouputs of
-programs that ouput mixed positive and negative types are not
-observable.  Also, programs that input mixed positive and negatives
-types are not executable.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Additive Duality in {{langRev}} }
-
-Additive duality is added to {{langRev}} by addition of negative types
-denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
-extend the language as follows:
-
-%subcode{bnf} include main
-% Value Types, b = ... | -b
-% Values, v = ... | -v
-%
-% Isomorphisms, iso &=& ... | eta+ | eps+
-
-\noindent
-with the types judgements
-
-%subcode{opsem} include main
-% eta+ &: 0 <-> (-b) + b :& eps+
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-
-We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
-the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
-dualy on the right is {{eps+}} showing the map from {{-b+b}} to
-0. Even though the diagrams below show a {{0}} wire for completeness,
-later diagrams will always drop them in contexts where we can
-implicitely introduce/eliminate then using {{zeroi}}/{{zeroe}}.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/eta.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/eps.pdf}
-\end{center}
-\end{multicols}
-
-The usual interpretation of the type {{b1+b2}} is that we either have
-an appropriately tagged value of type {{b1}} or an appropriately
-tagged value of type {{b2}}. This interpretation is mainted in the
-precense of {{eta+}} and {{eps+}} in the following sense: a value of
-type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
-switches into a value {{L (-v): -b + b}} that flows backwards on the
-upper wire. The inversion of direction is captured by the negative
-sign on the value.
-
-\paragraph*{Negative Information Flow.} 
-Hence, the intuitive way to read the digrams is that they reverse the
-flow of information -- in normal {{langRev}} circuits information
-flows from left-to-right and {{eps+}} operations cause information to
-flow from right-to-left.  The interpretation of {{eta+}} is precisely
-the opposite: it turns a right-to-left flow of information in a
-cicruit into a left-to-right flow.
-
-\paragraph*{Time Travelling.}
-A time travelling intuition is also applicable here: if view the
-left-to-right direction of information flow as the canonical forward
-direction of information, then the action of {{eps+}} allows values,
-aka information particles, to flow backwards in time. Particles that
-flow backward in time get to see and interact with the history of
-particles that they coexist with i.e. values that they exist
-multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
-(whose construction is shown in Section \ref{sec:neg-constructions}).
-
-\paragraph*{Coherence.}
-The appropriate coherence condition for Compact Closed Categories,
-expressed as the traingle axiom, is satisfied by {{eta+}}/{{eps+}} and
-may be visualized as explained by Selinger
-\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-\paragraph*{Non-termination.}
-As is well known in category theory, Compact Closed Categories contain
-trace operations.  In the context of {{langRev}} this implies that
-{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
-(detailed in Section \ref{sec:monoidal-constructions}) hence
-introducing \emph{non-termination} into the language
-\cite{infeffects}.
-
-\paragraph*{Zero Information.}
-The fact that {{eta+}} (and hence {{eps+}}) introduces no information
-can be understood is by viewing {{-b+b}} as a type whose value is
-observable only if you `pay' {{b}} amount of information to the {{-b}}
-branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
-amount of information is returned to you. (TODO. more needs to be said
-here.)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multiplicative Duality in {{langRev}} }
-
-%subcode{bnf} include main
-% Value Types, b = ... | b^^^
-% Values, v = ... | v^^^
-%
-% Isomorphisms, iso &=& ... | eta* | eps*
-
-\noindent
-with the types judgements
-
-%subcode{opsem} include main
-% eta* &: 1 <-> b^^^ * b :& eps*
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- v^^^ : b^^^
-
-Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
-{{a/b}}.
-
-\paragraph*{Unification.}
-
-\paragraph*{Entanglement.}
-
-\paragraph*{Coherence.}
-
-\paragraph*{Annihilation.}
-
-\paragraph*{Zero Information.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semantics}
-
-\subsection{Syntax}
-
-%subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
-%
-% Sequential Contexts, P, F = [] | P:c
-% Parallel Contexts, D = [] 
-%                     &|& D:(P| [](+)c| F)
-%                     &|& D:(P| c(+)[]| F)
-%                     &|& D:(P| [] (*) c v| F)
-%                     &|& D:(P| [] (*) v c| F)
-%                     &|& D:(P| c v (*)  []| F)
-%                     &|& D:(P| v c (*) []| F)
-%
-%
-% Machine States  = D[P| c v| F] | D[P| v c| F]
-% Start State  = [][[]| v c| []]
-% Stop State  = [][P| c v| []]
-
-Combinator reconstruction, {{P[c]}}:
-%subcode{opsem} include main
-% [](c) '= c
-% P:c'(c) '= P(c'(;)c)
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection{Operational Semantics}
-
-The small step semantics present for {{langRev}} below work
-symmetrically for forward and backward evaluation.
-
-\begin{itemize}
-\item 
-Basic reduction of an isomorphism. Note that the evaluation leaves the
-adjoint of the combinator behind. This will become important when we
-reverse the direction of computation.
-%% %subcode{opsem} include main
-%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
-
-%subcode{opsem} include main
-% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
-
-\item
-Sequencing involves pushing and popping from the Future and Past
-continuations:
-%subcode{opsem} include main
-% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
-% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
-
-\item
-Parallel composition, captures the current Future and Past and extends
-the parallel context.
-%subcode{opsem} include main
-% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
-% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
-% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
-% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
-
-\item
-Similarly for products:
-%subcode{opsem} include main
-% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
-% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
-% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
-
-and symmetrical rules for evaluation along the second branch. 
-%subcode{opsem} include main
-% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
-% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
-
-The later two rules will be relevant only for reverse execution.
-
-\end{itemize}
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrev}
-{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
-\end{proposition}
-
-\begin{proposition}[Groupoid]
-\label{prop:groupoid}
-{{langRevEE}} is a groupoid. 
-\end{proposition}
-
-\begin{proposition}
-\label{prop:groupoid}
-{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
-\end{proposition}
-
-Note: We use `Monoidal Closed' in the sense of Compact Closed. While
-the category has an multiplicative tensor that distributes over an
-additive tensor, it does not have injections or projections and hence
-does not have categorical sums and products.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Rules for {{eta}} and {{eps}} }
-
-The operation {{eps}} reverses the direction of a particle by
-reversing the world.
-
-Note that we deviate from the categorical definition of {{eta}} and
-{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
-choice of {{eta}}. This however does not affect us, because we deal
-with a symmetric category.
-
-\begin{itemize}
-\item Grammar
-%subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v | -v
-% Isomorphisms, iso &=& ... | eta | eps
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
-
-\item
-Type judgement.
-%subcode{opsem} include main
-% eta &: 0 <-> (-b) + b :& eps
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-
-\item
-Operational Semantics.
-%subcode{opsem} include main
-% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
-% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
-
-Note: there is NO reduction rule for {{eta}}.
-
-\item
-The adjoint of a parallel context is defined to be:
-%subcode{opsem} include main
-% []{dagger} '= []
-% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
-% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
-% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
-% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
-% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
-% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions in {{langRevEE}} }
-\label{sec:monoidal-constructions}
-
-These constructions are applicable to the additive and the
-multiplicative monoid.
-
-\paragraph*{Functions}
-
-\begin{center}
-  \includegraphics{diagrams/function.pdf}
-\end{center}
-
-Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
-
-\paragraph*{Function application}
-
-\begin{center}
-  \includegraphics{diagrams/apply1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/apply2.pdf}
-\end{center}
-
-\paragraph*{Function composition}
-
-\begin{center}
-  \includegraphics{diagrams/compose1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/compose.pdf}
-\end{center}
-
-This is also equivalent to sequencing both the computation blocks.
-
-\begin{center}
-  \includegraphics{diagrams/compose2.pdf}
-\end{center}
-
-\paragraph*{Delimited continuation}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}
-
-\paragraph*{Trace }
-
-%subcode{proof} include main
-%@ c : b2 + b1 <-> b2 + b3
-%@@ trace c : b1 <-> b3
-
-\begin{center}
-  \includegraphics{diagrams/trace.pdf}
-\end{center}
-
-\paragraph*{Principium Contradictiones}
-
-{{b <-> -(-b)}}
-
-\begin{center}
-  \includegraphics{diagrams/double_neg.pdf}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions Specific to Negation}
-\label{sec:neg-constructions}
-
-\paragraph*{Negation distributes over {{+}}. }
-
-{{-(b1 + b2) <-> (b1) + (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/dist_neg_plus.pdf}
-\end{center}
-
-\paragraph*{ {{eps_{fst} }} }
-
-{{(-b1)*b2 + b1*b2 <-> 0}}
-
-\begin{center}
-  \includegraphics{diagrams/eps_fst.pdf}
-\end{center}
-
-\paragraph*{Lifting negation out of {{*}}. }
-
-{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/mult_neg.pdf}
-\end{center}
-
-The following isomoprhism can be constructed similarly
-
-{{b1 * b2 <-> (-b1)*(-b2)}}
-
-\paragraph*{Lifting a operation of postive types to negated types}
-
-Given {{c : b1 <-> b2}}
-
-\begin{center}
-  \includegraphics{diagrams/neg_lift.pdf}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions Specific to Multiplication}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{To think about}
-
-\begin{itemize}
-
-\item If we built an effect over {{langRevEE}}, say {{create}} and
-  {{erase}}, are effects structured by an arrow or a monad now?
-
-\item Which operations can be lifted to work on negative types?
-
-\item Can the operational semantics for {{langRevEE}} be an
-  interpreter that is implemented in {{langRevT}} (similar to how the
-  the tree traversal interpreter was implemented). This would imply
-  the existence of a more powerful construction than Int, wherein the
-  products would also be preserved. This is possibly worth a paper in
-  itself.
-
-\item These functions aren't really values. There is no value one can
-  produce that denotes a function. These functions are the ability to
-  transform a value - the possibility of transforming a value.  In the
-  product encoding of environments, there is no value one can assign
-  to a variable such that it denotes a function.
-
-Actually this is possible. Consider two functions {{f : b1 <-> b2}}
-and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
-discriminate them. Hence the {{boo}} is the first class representation
-of the functions and can be thought of as the address of the function.
-
-\begin{center}
-  \includegraphics{diagrams/dispatch.pdf}
-\end{center}
-
-\item It is not fair to say that negative types flow backwards. The
-  following circuits are valid in {{langRevEE}}. It is however proper
-  to say that for any type {{b}} that flows in a direction, the type
-  {{-b}} flows in the reverse direction.
-
-\begin{center}
-  \includegraphics{diagrams/neg_circuit1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/neg_circuit2.pdf}
-\end{center}
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Advanced Programming Examples}
-
-\subsection{SAT solver}
-
-\subsection{GoI machine}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{The Field of Algebraic Numbers}
@@
 
+%% Jacques
+
 \acks This project was partially funded by Indiana University's Office

### 2012-02-28 rational.tex
para in intro
diagrams

@@
 \begin{abstract}
+
   Every functional programmer knows about sum and product types, {{a+b}} and
@@
   computation, logic programming, and more.
+
+  %% Every functional programmer knows about sum and product types,
+  %% {{a+b}} and {{a*b}} respectively. Negative and fractional types,
+  %% $a-b$ and $a/b$ respectively, are much less known and their
+  %% computational interpretation is unfamiliar and often complicated. We
+  %% show that in a programming model in which information is preserved
+  %% (such as the model introduced in our recent paper on
+  %% \emph{Information Effects}), these types have particularly intuitive
+  %% and natural computational interpretations. Intuitively, values of
+  %% negative types are values that flow ``backwards'' to satisfy
+  %% demands, values of fractional types are values that represent
+  %% first-class ``structural'' constraints.  The combination of these
+  %% negative and fractional types enable greater flexibility in
+  %% programming by breaking complicated invariants into local ones that
+  %% can be independently satisfied by a subcomputation. Furthermore,
+  %% they allow a programmer to express a wide range of programming
+  %% idioms, including higher-order functions, delimited continuations,
+  %% speculative computation, logic programming, and more.
+
 \end{abstract}
@@
 \section{Introduction}
+
+In 1989, \cite{Filinski:1989:DCI:648332.755574} observed that values and
+continuations are dual notions. This observation was followed by numerous
+papers (\cite{Griffin:1989:FNC:96709.96714, curien00duality, Wadler:2003,
+  DBLP:conf/rta/Wadler05}) which explored and refined this duality.
+
+We introduced this thesis that computation should be based on isomorphisms
+that preserve information~\cite{infeffects}. Since Filinski, we've had the
+idea that values and continuations are like mirror images. In a conventional
+language, the negative (continuation) side is implicit and we introduce
+information effects on the positive. Trying to recover the duality from this
+distorted positive side has always been messy. Now it looks clean because we
+have kept the positive side pure.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Negative Types: Intuition}
+
+Consider the following two ways of purchasing an item that costs \$20.00:
+\begin{enumerate}
+\item You give the seller a \$20.00 bill.
+\item You use a credit card to pay the seller.
+\end{enumerate}
+In both cases, the seller receives the money immediately but there is a
+subtle difference. In the first transaction, the money received by the seller
+corresponds to a value that has been produced earlier in the computation. In
+the second transaction, the money may or may not exist yet: computationally,
+a \emph{debt} is generated to compensate for the money received by the seller
+and this debt travels ``backwards'' towards the bank where it is (hopefully)
+reconciled.
+
+The example suggests that the ability to consider values flowing backwards
+enriches our computational model. This observation goes back to Filinski's
+Masters thesis where continuations are identified with these negative values.
+We discuss the connections to Filinski's work and others in more detail in
+Sec.~\ref{sec:related}. For now, we note that in a conventional programming
+language in which values can be copied and deleted at will, extreme care is
+needed to keep track of negative values. Indeed, in the example above, if it
+were possible to simply delete the variable corresponding to the debt, we
+would have produced money out of nothing.
+
+For this reason, our treatment of negative values in the context of
+{{langRev}} is particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
+right-to-left direction. In more detail, we can model the credit card
+transaction above using the following program (shown diagrammatically):
+
+FIGURE
+
+CLEAN UP the following based on the figure. In contrast, in our setting, one
+can start from the empty type $0$, introduce a positive value and its
+negative counterpart, and let each of these flow in arbitrary ways. The
+entire framework guarantees that neither the value nor its negative
+counterpart will be deleted or duplicated and hence that, in any closed
+program, the ``debt'' corresponding to the negative value is paid off exactly
+once. Computationally we model the first transaction using essentially the
+identity function which receives a \$20.00 bill as input from the buyer and
+propagates it on its output to the seller. The second transaction is more
+complicated. We model it as shown in the circuit below: There are two ways to
+understand this circuit that are both quite instructive. Let's first examine
+the type structure of each of the combinators that comprise the circuit. The
+first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
+produced from nothing so to speak by generating an equivalent debt that
+travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
+the execution. It goes forward in time so to speak, comes back, and then goes
+forwards again.
+
+It is critical that the framework in which the negative types are introduced
+is a framework in which all information is preserved, with no duplication or
+erasure. This guarantees that the generated debts are paid once and exactly
+once.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Fractional Types: Intuition}
+
+The type $a/b$ is prominent in the Lambek-Grishin calculus which is
+extensively used in computational linguistics. In the common interpretation,
+a value of type $a/b$ is a value of some type $c$ such that when put in a
+context of type $b$, the result is a value of type $a$. For example, assuming
+contexts are represented as functions, a value of type $\texttt{bool} /
+(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
+$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
+in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
+type $\texttt{bool}$.
+
+In our case, the situation is simpler. If the goal is to produce a value of
+type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
+have type {{(a * b) / b}}. In most settings, this type would be equivalent to
+{{a}}. However in our setting, the constraint that this value must eventually
+fit in a larger value that supplies the {{b}} is explicitly recorded and must
+be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
+which when used in the left-to-right direction allows the creation of a value
+and a corresponding constraint on the context.  Both the value and the
+constraint can flow in arbitrary ways during the computation but eventually
+in a closed program with no ``dangling'' constraints, the isomorphism should
+be used in the right-to-left direction to resolve the constraints.  To
+understand the computational interpretation, consider the following example.
+
+The goal is to produce a value of type $(\texttt{bool} \times
+\texttt{int})$. One part of the program, $e_1$, knows how to produce the
+value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
+to produce the value of type $\texttt{bool}$ (say {{true}}). Computationally,
+we model this as follows. The first subcomputation $e_1$ produces {{((),3)}}
+and uses the isomorphism to convert {{()}} to {{(alpha,1/alpha)}} where
+{{alpha}} is a yet-unknown boolean value. Reshuffling the components, $e_1$
+produces {{((alpha,3), 1/alpha)}}. Similarly, $e_2$ can independently produce
+{{((true,beta),1/beta)}} where {{beta}} is a yet-unknown integer value.
+When the two values meet, we can group the components as follows:
+\[
+{{(alpha,beta)}}  {{(3,1/beta)}}  {{(true,1/alpha)}}
+\]
+Using the isomorphism in the right-to-left direction on the last two pairs,
+forces {{alpha}} to be resolved to {{true}} and forces {{beta}} to be
+resolved to {{3}}. Both of these pairs then become {{()}} and be absorbed. We
+are left with the pair {{(true,3)}} as desired.
+
+To summarize, one can think of a value of type $a/b$ as a value that imposes
+a constraint on its context of use: it is a value that requires its context
+to be of type $b$ and only if that condition is satisfied, can the value be
+considered as a value of type $a$. In the simplest case, the type $1/b$ just
+constraints the context to be of type $b$. By allowing the fractional types
+to be first-class, we can separate the generation of constraints from their
+use. Both the value and the constraint can flow in arbitrary ways during the
+computation but eventually in a closed program with no ``dangling''
+constraints, the isomorphism should be used in the right-to-left direction to
+resolve the constraints.
+
+NEED to work out the example in detail (or perhaps another example that's
+better.
+
+Again it is critical that the framework in which the fractional types are
+introduced is a framework in which all information is preserved, with no
+duplication or erasure. This guarantees that the generated constraints are
+satisfied once and exactly once.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Summary}
+
+To summarize negative, fractional, square root, and imaginary types all make
+sense. What they help you accomplish as a programmer is to disassociate
+global invariants into local ones that can be satisfied independently by
+subcomputations with no synchronization or communication. A computation
+producing something of type $a/b$ does not need to concern itself with who is
+going to supply the missing $b$: it just does its part. Conversely faced with
+a complicated task, a computation might decide to break it into pieces and
+demand these pieces using negative types.
+
+It is no surprise that these types are closely related to quantum mechanics
+and that they give us the feel that this is how nature computes. This is
+speculation however.
+
+In any case, in a framework where information can be copied and deleted, none
+of this makes much sense. It is critical that these constraints and demands
+can neither be duplicated nor erased.  This gives us the maximum
+``parallelism'' possible.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRev}} }
+
+The presentation here differs from \cite{infeffects} in the sense that
+we have introduced the {{0}} type here, thereby completeing the monoid
+{{(0,+)}} and we present a small-step operational semantics.
+
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b + b | b * b 
+% values, v ::= () | left v | right v | (v, v)
+
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%zeroe :&  0 + b <-> b &: zeroi
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist0 :& 0 * b <-> 0 &: factor0
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+
+%subcode{proof} include main
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+
+\begin{definition}{(Syntax of {{langRev}})}
+\label{def:langRev}
+We collect our types, values, and combinators, to get the full language
+definition.
+%subcode{bnf} include main
+% value types, b ::= 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% comb.~types, t ::= b <-> b
+% iso ::= swap+ | assocl+ | assocr+ 
+%     &|& identl* | identr* 
+%     &|& swap* | assocl* | assocr* 
+%     &|& dist | factor 
+% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+\end{definition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Semantics}
+
+%subcode{bnf} include main
+% Combinators, c = iso | c (;) c | c (*) c | c (+) c
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+% Values, v = () | (v, v) | L v | R v
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+\begin{scriptsize}
+%subcode{opsem} include main
+%! columnStyle = rlcl
+% swap+ & (left v) &|-->& right v
+% swap+ & (right v) &|-->& left v 
+% assocl+ & (left v1) &|-->& left (left v1)
+% assocl+ & (right (left v2)) &|-->& left (right v2)
+% assocl+ & (right (right v3)) &|-->& right v3 
+% assocr+ & (left (left v1)) &|-->& left v1
+% assocr+ & (left (right v2)) &|-->& right (left v2)
+% assocr+ & (right v3) &|-->& right (right v3)
+% identl* & ((), v) &|-->& v 
+% identr* & v &|-->& ((), v) 
+% swap* & (v1, v2) &|-->& (v2, v1) 
+% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
+% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
+% dist & (left v1, v3) &|-->& left (v1, v3)
+% dist & (right v2, v3) &|-->& right (v2, v3)
+% factor & (left (v1, v3)) &|-->& (left v1, v3) 
+% factor & (right (v2, v3)) &|-->& (right v2, v3)   
+\end{scriptsize}
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoid}
+{{langRev}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:groupoid}
+{{langRev}} is a dagger symmetric monoidal category. 
+\end{proposition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRevEE}} }
+
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
+input {{v1 : b1}} gives us a value {{v2 : b2}} on
+termination. Execution is well defined only if {{b1}} and {{b2}} are
+entirely positive types. Consider the program that
+
+(TODO: So this is a really bad explanation. But this section is really
+important and we need to explain carefully. Maybe an analogy can be
+drawn to Cayley diagrams for groups to point out that the notion of
+`i' is entirely artificial, but it is essential to the discourse.)
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The ouputs of
+programs that ouput mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Additive Duality in {{langRev}} }
+
+Additive duality is added to {{langRev}} by addition of negative types
+denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
+extend the language as follows:
+
+%subcode{bnf} include main
+% Value Types, b = ... | -b
+% Values, v = ... | -v
+%
+% Isomorphisms, iso &=& ... | eta+ | eps+
+
+\noindent
+with the types judgements
+
+%subcode{opsem} include main
+% eta+ &: 0 <-> (-b) + b :& eps+
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
+the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
+dualy on the right is {{eps+}} showing the map from {{-b+b}} to
+0. Even though the diagrams below show a {{0}} wire for completeness,
+later diagrams will always drop them in contexts where we can
+implicitely introduce/eliminate then using {{zeroi}}/{{zeroe}}.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps.pdf}
+\end{center}
+\end{multicols}
+
+The usual interpretation of the type {{b1+b2}} is that we either have
+an appropriately tagged value of type {{b1}} or an appropriately
+tagged value of type {{b2}}. This interpretation is mainted in the
+precense of {{eta+}} and {{eps+}} in the following sense: a value of
+type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
+switches into a value {{L (-v): -b + b}} that flows backwards on the
+upper wire. The inversion of direction is captured by the negative
+sign on the value.
+
+\paragraph*{Negative Information Flow.} 
+Hence, the intuitive way to read the digrams is that they reverse the
+flow of information -- in normal {{langRev}} circuits information
+flows from left-to-right and {{eps+}} operations cause information to
+flow from right-to-left.  The interpretation of {{eta+}} is precisely
+the opposite: it turns a right-to-left flow of information in a
+cicruit into a left-to-right flow.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus2.pdf}
+}
+\end{center}  
+\end{multicols}
+
+\paragraph*{Time Travelling.}
+A time travelling intuition is also applicable here: if view the
+left-to-right direction of information flow as the canonical forward
+direction of information, then the action of {{eps+}} allows values,
+aka information particles, to flow backwards in time. Particles that
+flow backward in time get to see and interact with the history of
+particles that they coexist with i.e. values that they exist
+multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
+(whose construction is shown in Section \ref{sec:neg-constructions}).
+
+\paragraph*{Coherence.}
+The appropriate coherence condition for Compact Closed Categories,
+expressed as the traingle axiom, is satisfied by {{eta+}}/{{eps+}} and
+may be visualized as explained by Selinger
+\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+\paragraph*{Non-termination.}
+As is well known in category theory, Compact Closed Categories contain
+trace operations.  In the context of {{langRev}} this implies that
+{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
+(detailed in Section \ref{sec:monoidal-constructions}) hence
+introducing \emph{non-termination} into the language
+\cite{infeffects}.
+
+\paragraph*{Zero Information.}
+The fact that {{eta+}} (and hence {{eps+}}) introduces no information
+can be understood is by viewing {{-b+b}} as a type whose value is
+observable only if you `pay' {{b}} amount of information to the {{-b}}
+branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
+amount of information is returned to you. (TODO. more needs to be said
+here.)
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Multiplicative Duality in {{langRev}} }
+
+%subcode{bnf} include main
+% Value Types, b = ... | b^^^
+% Values, v = ... | v^^^
+%
+% Isomorphisms, iso &=& ... | eta* | eps*
+
+\noindent
+with the types judgements
+
+%subcode{opsem} include main
+% eta* &: 1 <-> b^^^ * b :& eps*
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- v^^^ : b^^^
+
+Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
+{{a/b}}.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta_times.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps_times.pdf}
+\end{center}
+\end{multicols}
+
+\paragraph*{Unification.}
+
+test test test
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eta_times1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_times1.pdf}
+}
+\end{center}  
+\end{multicols}
+
+\paragraph*{Entanglement.}
+
+\paragraph*{Coherence.}
+
+\paragraph*{Annihilation.}
+
+\paragraph*{Zero Information.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Semantics}
+
+\subsection{Syntax}
+
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
+%
+% Sequential Contexts, P, F = [] | P:c
+% Parallel Contexts, D = [] 
+%                     &|& D:(P| [](+)c| F)
+%                     &|& D:(P| c(+)[]| F)
+%                     &|& D:(P| [] (*) c v| F)
+%                     &|& D:(P| [] (*) v c| F)
+%                     &|& D:(P| c v (*)  []| F)
+%                     &|& D:(P| v c (*) []| F)
+%
+%
+% Machine States  = D[P| c v| F] | D[P| v c| F]
+% Start State  = [][[]| v c| []]
+% Stop State  = [][P| c v| []]
+
+Combinator reconstruction, {{P[c]}}:
+%subcode{opsem} include main
+% [](c) '= c
+% P:c'(c) '= P(c'(;)c)
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection{Operational Semantics}
+
+The small step semantics present for {{langRev}} below work
+symmetrically for forward and backward evaluation.
+
+\begin{itemize}
+\item 
+Basic reduction of an isomorphism. Note that the evaluation leaves the
+adjoint of the combinator behind. This will become important when we
+reverse the direction of computation.
+%% %subcode{opsem} include main
+%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
+
+%subcode{opsem} include main
+% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
+
+\item
+Sequencing involves pushing and popping from the Future and Past
+continuations:
+%subcode{opsem} include main
+% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
+% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
+
+\item
+Parallel composition, captures the current Future and Past and extends
+the parallel context.
+%subcode{opsem} include main
+% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
+% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
+% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
+% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
+
+\item
+Similarly for products:
+%subcode{opsem} include main
+% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
+% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
+% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
+
+and symmetrical rules for evaluation along the second branch. 
+%subcode{opsem} include main
+% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
+% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
+
+The later two rules will be relevant only for reverse execution.
+
+\end{itemize}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoid}
+{{langRevEE}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:groupoid}
+{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
+\end{proposition}
+
+Note: We use `Monoidal Closed' in the sense of Compact Closed. While
+the category has an multiplicative tensor that distributes over an
+additive tensor, it does not have injections or projections and hence
+does not have categorical sums and products.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Rules for {{eta}} and {{eps}} }
+
+The operation {{eps}} reverses the direction of a particle by
+reversing the world.
+
+Note that we deviate from the categorical definition of {{eta}} and
+{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
+choice of {{eta}}. This however does not affect us, because we deal
+with a symmetric category.
+
+\begin{itemize}
+\item Grammar
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v | -v
+% Isomorphisms, iso &=& ... | eta | eps
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
+
+\item
+Type judgement.
+%subcode{opsem} include main
+% eta &: 0 <-> (-b) + b :& eps
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+\item
+Operational Semantics.
+%subcode{opsem} include main
+% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
+% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
+
+Note: there is NO reduction rule for {{eta}}.
+
+\item
+The adjoint of a parallel context is defined to be:
+%subcode{opsem} include main
+% []{dagger} '= []
+% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
+% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
+% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
+% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
+% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
+% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions in {{langRevEE}} }
+\label{sec:monoidal-constructions}
+
+These constructions are applicable to the additive and the
+multiplicative monoid.
+
+\paragraph*{Functions}
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
+
+\paragraph*{Function application}
+
+\begin{center}
+  \includegraphics{diagrams/apply1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/apply2.pdf}
+\end{center}
+
+\paragraph*{Function composition}
+
+\begin{center}
+  \includegraphics{diagrams/compose1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/compose.pdf}
+\end{center}
+
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+\paragraph*{Delimited continuation}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+\paragraph*{Trace }
+
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+\paragraph*{Principium Contradictiones}
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Negation}
+\label{sec:neg-constructions}
+
+\paragraph*{Negation distributes over {{+}}. }
+
+{{-(b1 + b2) <-> (b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+\paragraph*{ {{eps_{fst} }} }
+
+{{(-b1)*b2 + b1*b2 <-> 0}}
+
+\begin{center}
+  \includegraphics{diagrams/eps_fst.pdf}
+\end{center}
+
+\paragraph*{Lifting negation out of {{*}}. }
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/mult_neg.pdf}
+\end{center}
+
+The following isomoprhism can be constructed similarly
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
+
+\paragraph*{Lifting a operation of postive types to negated types}
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Multiplication}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{To think about}
+
+\begin{itemize}
+
+\item If we built an effect over {{langRevEE}}, say {{create}} and
+  {{erase}}, are effects structured by an arrow or a monad now?
+
+\item Which operations can be lifted to work on negative types?
+
+\item Can the operational semantics for {{langRevEE}} be an
+  interpreter that is implemented in {{langRevT}} (similar to how the
+  the tree traversal interpreter was implemented). This would imply
+  the existence of a more powerful construction than Int, wherein the
+  products would also be preserved. This is possibly worth a paper in
+  itself.
+
+\item These functions aren't really values. There is no value one can
+  produce that denotes a function. These functions are the ability to
+  transform a value - the possibility of transforming a value.  In the
+  product encoding of environments, there is no value one can assign
+  to a variable such that it denotes a function.
+
+Actually this is possible. Consider two functions {{f : b1 <-> b2}}
+and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
+discriminate them. Hence the {{boo}} is the first class representation
+of the functions and can be thought of as the address of the function.
+
+\begin{center}
+  \includegraphics{diagrams/dispatch.pdf}
+\end{center}
+
+\item It is not fair to say that negative types flow backwards. The
+  following circuits are valid in {{langRevEE}}. It is however proper
+  to say that for any type {{b}} that flows in a direction, the type
+  {{-b}} flows in the reverse direction.
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit2.pdf}
+\end{center}
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Advanced Programming Examples}
+
+\subsection{SAT solver}
+
+\subsection{GoI machine}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{The Field of Algebraic Numbers}

### 2012-02-28 rational.tex
hopefully undid some damage
abstract and intro

--- 2012-02-28 rational.tex
@@
 \usepackage{bm}
+\usepackage{cmll}
 
@@
   Every functional programmer knows about sum and product types, {{a+b}} and
-  {{a*b}} respectively. Negative and fractional types, $a-b$ and $a/b$
-  respectively, are much less known and their computational interpretation is
-  unfamiliar and often complicated. We show that in a programming model in
-  which information is preserved (such as the model introduced in our recent
-  paper on \emph{Information Effects}), these types have particularly
-  intuitive and natural computational interpretations. Intuitively, values of
-  negative types are values that flow ``backwards'' to satisfy demands and
-  values of fractional types are values that impose constraints on their
-  context.  The combination of these negative and fractional types enables
-  greater flexibility in programming by breaking global invariants into local
-  ones that can be autonomously satisfied by a subcomputation. Furthermore,
-  they allow a programmer to express a wide range of programming idioms,
-  including higher-order functions, delimited continuations, speculative
-  computation, logic programming, and more.
+  {{a*b}}. Negative and fractional types, $a-b$ and $a/b$ respectively, are
+  much less known and their computational interpretation is unfamiliar and
+  often complicated. We show that in a programming model in which information
+  is preserved (such as the model introduced in our recent paper on
+  \emph{Information Effects}), these types have particularly natural
+  computational interpretations. Intuitively, values of negative types are
+  values that flow ``backwards'' to satisfy demands and values of fractional
+  types are values that impose constraints on their context.  The combination
+  of these negative and fractional types enables greater flexibility in
+  programming by breaking global invariants into local ones that can be
+  autonomously satisfied by a subcomputation. Furthermore, these types and
+  their properties suggest that the previously observed duality of
+  computation conflated two orthogonal notions: an additive duality that
+  corresponds to backtracking and a multiplicative duality that corresponds
+  to unification.
 \end{abstract}
@@
 \section{Introduction}
+
+In 1989, Filinski~\cite{Filinski:1989:DCI:648332.755574} observed that values
+and continuations are dual notions. This observation was followed by numerous
+papers~\cite{Griffin:1989:FNC:96709.96714, curien00duality, Wadler:2003,
+  DBLP:conf/rta/Wadler05} which explored and refined this duality. Even
+earlier, Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2}
+introduced a logic which contains a dual to implication. Her work has been
+distilled in the form of \emph{subtractive logic}~\cite{Crolard01} which has
+recently been related to coroutines~\cite{Crolard01082004} and delimited
+continuations~\cite{Ariola:2009:TFD:1743339.1743381}.
+
+In a seemingly unrelated development, Girard in 1987 introduced linear
+logic~\cite{Girard87tcs} which, among other contributions, exposed an
+additive/multiplicative distinction in logical connectives and rules. In
+particular, linear logic includes additive disjunctions $\oplus$ and
+conjunctions $\with$ as well as multiplicative disjunctions $\parr$ and
+conjuctions $\otimes$. Duality is also prominent in linear logic: it relates
+the additive connectives to each other (the dual of $\oplus$ is $\with$ and
+vice-versa) and the multiplicative connectives to each other (the dual of
+$\otimes$ is $\parr$ and vice-versa).
+
+We introduced this thesis that computation should be based on isomorphisms
+that preserve information~\cite{infeffects}. Since Filinski, we've had the
+idea that values and continuations are like mirror images. In a conventional
+language, the negative (continuation) side is implicit and we introduce
+information effects on the positive. Trying to recover the duality from this
+distorted positive side has always been messy. Now it looks clean because we
+have kept the positive side pure.
 
@@
 
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus2.pdf}
+}
+\end{center}  
+\end{multicols}
+
 \paragraph*{Time Travelling.}
@@
 
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta_times.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps_times.pdf}
+\end{center}
+\end{multicols}
+
 \paragraph*{Unification.}
+
+test test test
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eta_times1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_times1.pdf}
+}
+\end{center}  
+\end{multicols}
 

### 2012-02-28 rational.tex
spell check and notes on algebraic types

@@
 other systems which have one notion of duality over additive and
-multiplicative componenents, {{langRevEE}} has two notions of duality
+multiplicative components, {{langRevEE}} has two notions of duality
 -- an additive duality over the monoid {{(0, +)}} and a multiplicative
@@
 a function space and hence {{langRevEE}} has an additive function
-space correspoding to a notion of control or backtracking and a
+space corresponding to a notion of control or backtracking and a
 multiplicative function space corresponding to a notion of unification
@@
 The presentation here differs from \cite{infeffects} in the sense that
-we have introduced the {{0}} type here, thereby completeing the monoid
+we have introduced the {{0}} type here, thereby completing the monoid
 {{(0,+)}} and we present a small-step operational semantics.
@@
 
-We define observables to be only positive types. The ouputs of
-programs that ouput mixed positive and negative types are not
+We define observables to be only positive types. The outputs of
+programs that output mixed positive and negative types are not
 observable.  Also, programs that input mixed positive and negatives
@@
 \noindent
-with the types judgements
+with the types judgments
 
@@
 the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
-dualy on the right is {{eps+}} showing the map from {{-b+b}} to
+dually on the right is {{eps+}} showing the map from {{-b+b}} to
 0. Even though the diagrams below show a {{0}} wire for completeness,
 later diagrams will always drop them in contexts where we can
-implicitely introduce/eliminate then using {{zeroi}}/{{zeroe}}.
+implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.
 
@@
 an appropriately tagged value of type {{b1}} or an appropriately
-tagged value of type {{b2}}. This interpretation is mainted in the
-precense of {{eta+}} and {{eps+}} in the following sense: a value of
+tagged value of type {{b2}}. This interpretation is maintained in the
+presence of {{eta+}} and {{eps+}} in the following sense: a value of
 type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
@@
 \paragraph*{Negative Information Flow.} 
-Hence, the intuitive way to read the digrams is that they reverse the
+Hence, the intuitive way to read the diagrams is that they reverse the
 flow of information -- in normal {{langRev}} circuits information
@@
 the opposite: it turns a right-to-left flow of information in a
-cicruit into a left-to-right flow.
+circuit into a left-to-right flow.
 
@@
 
-\paragraph*{Time Travelling.}
-A time travelling intuition is also applicable here: if view the
+\paragraph*{Time Traveling.}
+A time traveling intuition is also applicable here: if view the
 left-to-right direction of information flow as the canonical forward
@@
 The appropriate coherence condition for Compact Closed Categories,
-expressed as the traingle axiom, is satisfied by {{eta+}}/{{eps+}} and
+expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
 may be visualized as explained by Selinger
@@
 \item
-Type judgement.
+Type judgment.
 %subcode{opsem} include main
@@
 
-The following isomoprhism can be constructed similarly
+The following isomorphism can be constructed similarly
 
@@
 
-\paragraph*{Lifting a operation of postive types to negated types}
+\paragraph*{Lifting a operation of positive types to negated types}
 
@@
 the formulae-as-types interpretation of subtractive logic) address the
-computational interpretation of substraction. He explains the type $A-B$ as
+computational interpretation of subtraction. He explains the type $A-B$ as
 the of \emph{coroutines} with a local environment of type $A$ and a
@@
 \paragraph*{Linear Logic} In accounts that are linear, the value and
-continuation that comprise the substractive type need to be constrained to
+continuation that comprise the subtractive type need to be constrained to
 ``stay together.'' This can be achieved by various restrictions. In this work
@@
 image of the Int construction on {{langRevT}}, since the the later
-lacks a multiplicative tensor that distrbutes over the additive tensor
+lacks a multiplicative tensor that distributes over the additive tensor
 in Int {{langRevT}}.
@@
 
+Here is one way to view computations involving irrationals that might
+shed some insight into how we should view recursive data types
+also. For finite datatypes say of size {{n}} we can say that an
+element specifies a choice of {{1/n}} and hence contains as much
+information. For finite things, each value is computationally equal to
+any other value. Any permutation is possible among the values. Hence
+there is no metric possible on the values.  Recursive data types, such
+as binary trees, have an infinite number of elements and hence the
+argument that says that value specifies {{1/n}} units of information
+is not longer valid. Instead to talk meaningfully about information
+content, we should talk about a probability distribution on the values
+of the type. However what shall we base this distribution on? Is there
+a natural order beyond the unfolding order of the values of a type?
+
+In this context, let us go back to the thought that recursive such as
+{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
+interesting solution of trees is precisely that they involve
+irrational numbers. Our usual notation for irrationals such as
+{{2^{1/2} }} can actually be viewed as a thunk representing the
+infinite computation {{1.4142..}}. Irrationals in this sense have a
+very different information theoretic interpretation from full real
+numbers. With Chaitin's Omega numbers he postulates the existence of
+real numbers that have an infinite amount of information -- ones for
+example are the solution to the question about the halting properties
+of every Turning machine.
+
+Real numbers in that sense really are a series of bits of unknown
+information theoretic interpretation -- a sort of {{top}}. Numbers
+such as {{2^{1/2} }} are very different in the sense that even though
+they contain an infinite sequence of bits, there is a finite program
+that generates those bits and its only the cost of computing bits that
+is infinite.  The consistency of arithmetic isn't affected by
+extending the rational field by adding any specific computable
+irrational. So it might be possible to design a model of computing
+where we can indeed deal with them. Such a model will have to explicit
+thunks for the delayed computation represented by trees and never
+equate and cancel two infinite computations unless the represent
+exactly the same infinite computation.
+
 TODO.
@@
 units. One can solve the quadratic equation to determine that the sides are 7
-and 13 and proceed. This however prematuraly forces us to globally solve the
+and 13 and proceed. This however prematurely forces us to globally solve the
 constraint. Instead we can let the two sides of the rectangle be $x$ and

### 2012-02-28 rational.tex
a bit of random stuff for the intro

--- 2012-02-28 rational.tex
@@
 
+%% I am getting an error about too many math packages used!!!
+%% I commented the ones we don't seem to be using.
+
 \usepackage{graphicx}
-\usepackage{longtable}
-\usepackage{comment}
+%%\usepackage{longtable}
+%%\usepackage{comment}
 \usepackage{amsmath}
-\usepackage{mdwlist}
-\usepackage{txfonts}
+%%\usepackage{mdwlist}
+%%\usepackage{txfonts}
 \usepackage{xspace}
-\usepackage{amstext}
+%%\usepackage{amstext}
 \usepackage{amssymb}
@@
 \usepackage{multicol}
-\usepackage[nodayofweek]{datetime}
-\usepackage{etex}
-\usepackage[all, cmtip]{xy}
-\usepackage{xcolor}
-\usepackage{listings}
-\usepackage{multicol}
-\usepackage{bm}
+%%\usepackage[nodayofweek]{datetime}
+%%\usepackage{etex}
+%%\usepackage[all, cmtip]{xy}
+%%\usepackage{xcolor}
+%%\usepackage{listings}
+%%\usepackage{bm}
 \usepackage{cmll}
@@
 %! factor = \mathit{factor}
+%! eta = \eta
+%! eps = \epsilon
 %! eta+ = \eta^+
@@
 
+Continuations made their introduction to the world of programming language
+semantics as a mathematical device to model first-class labels and
+jumps~\cite{springerlink:10.1023/A:1010026413531}. In a remarkable
+development, Filinski~\cite{Filinski:1989:DCI:648332.755574} observed that
+--- with the right perspective --- this highly imperative concept was
+actually the symmetric dual of values. The heart of the observation is that
+values represent entities that flow from producers to consumers while
+continuations represent \emph{demands} for such entities, that flow from
+consumers to producers. In more detail, Filinski describes continuations as
+representing ``the \emph{lack} or \emph{absence} of a value, just as having a
+negative amount of money represents a debt.'' He then proceeds to construct a
+language where values and continuations are treated truly symmetrically. To
+that end, he abandons the $\lambda$-calculus amalgamation of functions and
+values and distinguishes between three different syntactic classes:
+functions, values, and continuations, with the property that any function can
+be used either as a value transformer or as a continuation transformer.
+
+This highly intuitive and appealing idea was further explored and refined by
+many authors~\cite{Griffin:1989:FNC:96709.96714, curien00duality,
+  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet despite of its appeal, the idea
+that continuations of this
+
+appeared as functional
+representation of labels and jumps,
+Filinski, in 1989, observed that they
+are the categorical introduced the symmetric $\lambda$-calculus
+
+Computationally, symmetry exhibits itself as a duality between
+two concepts.
+
+Symmetry is pervasive in both natural and man-made environments.
+
 In 1989, Filinski~\cite{Filinski:1989:DCI:648332.755574} observed that values
 and continuations are dual notions. This observation was followed by numerous
-papers~\cite{Griffin:1989:FNC:96709.96714, curien00duality, Wadler:2003,
-  DBLP:conf/rta/Wadler05} which explored and refined this duality. Even
-earlier, Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2}
-introduced a logic which contains a dual to implication. Her work has been
-distilled in the form of \emph{subtractive logic}~\cite{Crolard01} which has
-recently been related to coroutines~\cite{Crolard01082004} and delimited
+
+Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
+logic which contains a dual to implication. Her work has been distilled in
+the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
+related to coroutines~\cite{Crolard01082004} and delimited
 continuations~\cite{Ariola:2009:TFD:1743339.1743381}.
@@
 
-We show a deep symmetry between functions and delimited continuations,
-values and continuations that arises in {{langRev}} in a manner that
-is reminiscent of Filinksi's Symmetric \lcal
-~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
-extending {{langRev}} with a notion of additive duality over the
-monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
-Compact Closed Categories. The resulting dual types, which we denote
-{{-b}}, have a time traveling ``backward information flow''
-interpretation and allow for the encoding of higher-order function and
-iteration via the construction of {{trace}} operators, thereby making
-the extended language {{langRevEE}} a Turing-complete reversible
-programming language with higher-order functions and first-class
-delimited continuations.
+We show a deep symmetry between functions and delimited continuations, values
+and continuations that arises in {{langRev}} in a manner that is reminiscent
+of Filinksi's Symmetric \lcal ~\cite{Filinski:1989:DCI:648332.755574}. The
+symmetry arises by extending {{langRev}} with a notion of additive duality
+over the monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
+Compact Closed Categories. The resulting dual types, which we denote {{-b}},
+have a time traveling ``backward information flow'' interpretation and allow
+for the encoding of higher-order function and iteration via the construction
+of trace operators, thereby making the extended language {{langRevEE}} a
+Turing-complete reversible programming language with higher-order functions
+and first-class delimited continuations.
 
@@
 value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
-to produce the value of type $\texttt{bool}$ (say {{true}}). Computationally,
-we model this as follows. The first subcomputation $e_1$ produces {{((),3)}}
-and uses the isomorphism to convert {{()}} to {{(alpha,1/alpha)}} where
-{{alpha}} is a yet-unknown boolean value. Reshuffling the components, $e_1$
-produces {{((alpha,3), 1/alpha)}}. Similarly, $e_2$ can independently produce
-{{((true,beta),1/beta)}} where {{beta}} is a yet-unknown integer value.
-When the two values meet, we can group the components as follows:
+to produce the value of type $\texttt{bool}$ (say true). Computationally, we
+model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
+uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
+is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
+$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
+$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
+the two values meet, we can group the components as follows:
 \[
-{{(alpha,beta)}}  {{(3,1/beta)}}  {{(true,1/alpha)}}
+(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
 \]
 Using the isomorphism in the right-to-left direction on the last two pairs,
-forces {{alpha}} to be resolved to {{true}} and forces {{beta}} to be
-resolved to {{3}}. Both of these pairs then become {{()}} and be absorbed. We
-are left with the pair {{(true,3)}} as desired.
+forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
+{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
+with the pair $(true,3)$ as desired.
 

### 2012-02-28 rational.tex
fixed some latex issues and small edits

--- 2012-02-28 rational.tex
@@
 \usepackage{multicol}
+\newcommand\hmmax{0} % default 3
+\newcommand\bmmax{0} % default 4
 \usepackage{bm}
@@
 %! eps+ = \epsilon^+
-%! eta* = \eta^*
+%! eta* = \eta^{\times}
+%! eps* = \epsilon^{\times}
+%! trace+ = trace^+
+%! trace* = trace^{\times}
 %! ^^^ = ^{-1}
-%! eps* = \epsilon^*
 %! (o) = \fatsemi
@@
 
-So when a programmer writes the datatype declaration \verb|t = t^2+1|, if we
-allow negative etc. then this is effectively writing 
-\verb|t = cubicroot{-1}|. If we are in the field then computations that 
-manipulate these trees can be sliced and diced even at interfaces that
-expose the cubic root and the imaginary types.
+So when a programmer writes the datatype declaration \verb|t = t^2+1|,
+if we allow negative etc. then this is effectively writing
+\verb|t = cubicroot{-1}|. If we are in the field then computations
+that manipulate these trees can be sliced and diced even at interfaces
+that expose the cubic root and the imaginary types.
 
@@
 extending {{langRev}} with a notion of additive duality over the
-monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
+monoid {{(+, 0)}} by including {{eta+}} and {{eps+}} operators of
 Compact Closed Categories. The resulting dual types, which we denote
@@
 \begin{proposition}
-\label{prop:groupoid}
+\label{prop:category}
 {{langRev}} is a dagger symmetric monoidal category. 
@@
 
+Multiplicative duality is added to {{langRev}} by adding types
+{{b^^^}}, values {{v^^^}} and isomorphisms {{eta*}} and {{eps*}}.
+Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
+{{a/b}}.
+
 %subcode{bnf} include main
@@
 \noindent
-with the types judgments
+These have the type judgments
 
@@
 %@@ |- v^^^ : b^^^
-
-Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
-{{a/b}}.
 
@@
 \begin{proposition}[Logical Reversibility]
-\label{prop:logrev}
+\label{prop:logrevEE}
 {{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
@@
 \begin{proposition}[Groupoid]
-\label{prop:groupoid}
+\label{prop:groupoidEE}
 {{langRevEE}} is a groupoid. 
@@
 \begin{proposition}
-\label{prop:groupoid}
+\label{prop:categoryEE}
 {{langRev}} is a Dagger Symmetric Monoidal Closed Category.
@@
 Square root and imaginary types have also appeared in the literature: we
-relegate the connections to Sec.~\ref{related} and proceed with a simple
+relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
 explanation. We have so far extended the set of types to be the rational

### 2012-02-28 rational.tex

--- 2012-02-28 rational.tex
@@
 \usepackage{multicol}
-%%\usepackage[nodayofweek]{datetime}
-%%\usepackage{etex}
-%%\usepackage[all, cmtip]{xy}
-%%\usepackage{xcolor}
-%%\usepackage{listings}
-%%\usepackage{bm}
+\usepackage[nodayofweek]{datetime}
+\usepackage{etex}
+\usepackage[all, cmtip]{xy}
+\usepackage{xcolor}
+\usepackage{listings}
+\usepackage{multicol}
+\newcommand\hmmax{0} % default 3
+\newcommand\bmmax{0} % default 4
+\usepackage{bm}
 \usepackage{cmll}
@@
 symmetry arises by extending {{langRev}} with a notion of additive duality
-over the monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
+over the monoid {{(+, 0)}} by including {{eta+}} and {{eps+}} operators of
 Compact Closed Categories. The resulting dual types, which we denote {{-b}},

### 2012-02-28 rational.tex
notes on multiplicative duals

@@
 
-%%%%%%%%%%%%%%%%%%%%%
+ADD MATH.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Multiplicative Duality in {{langRev}} }
@@
 
+As with additive duals, we represent {{eta*}} and {{eps*}] using
+`U'-shaped connectors. In the general case the wires for {{1}} will be
+dropped as it can freely introduced or eliminated using {{uniti}} and
+{{unite}}.
+
 \begin{multicols}{2}
@@
 
-\paragraph*{Unification.}
-
-test test test
+The usual interpretation of {{b1*b2}} we have both a value of type
+{{b1}} and a value of type {{b2}}. This interpretation is maintained
+in the precense of fractionals. Hence an {{eta* : 1 <-> b^^^ * b}} is
+to be viewed as a fission point for a value of type {{b}} and its
+multiplicative inverse {{b^^^}}. This naturally raises the question --
+which value of type {{b}} is produced? The isomorphism {{eta*}} does
+not produce any specific value of type {{b}}, instead it produces a
+place holder for a value such that the value of type {{b^^^}} is
+always the dual of the chosen one.
+
 \begin{multicols}{2}
@@
 
-\paragraph*{Entanglement.}
+\paragraph*{Unification.}
+A computational interpretation of this is that {{langRevEE}} acts like
+a logic programming system and {{eta*}} acts as the site of creation
+of a new typed logic variable of type {{b}} and its corresponding dual
+of type {{b^^^}}. These two values are free to move independantly
+through a circuit, but any operations on value affect what operations
+are possible on the other since they are both related by an underlying
+logic variable. Dually, {{eps*}} acts as a unification site for values
+flowing along both branches. If the values are not exactly duals, then
+the system fails to unify and hence the program is stuck. If the
+values are exactly duals, then unification succeeds and {{eta*}}
+returns {{1}}.
+
+\paragraph*{Entanglement and Anti-Particles.}
+A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
+quantum mechanics. The {{eta*}} is a site of fission that creates an
+information particle and its anti-particle. Both flow in the same
+direction in time. The produce particles are entangles in that actions
+on one (such as transformation by application of an isomorphism)
+affects the other. Further the fission site does not create a particle
+in any specific state, but in a superposition of all possible states
+as dictated by its type. In each possible world that the particle
+exists, it takes on a specific value inhibiting its type and
+correspondingly its entangled anti-particle takes the dual of the
+specific value.
+
+This also leads to the exiciting point of view that entangled
+particle/anti-particle pairs act as functions in a physical
+first-class sense. We can ask the fuction to be applied to a value by
+unifying the anti-particle with a particle in a well-known state and
+examining is pair. We shall use this point of view in the development
+our SAT solver (see Sec. \ref{sec:sat-solver}).
 
 \paragraph*{Coherence.} 
+As before, the requisite coherence condition on Compact Closed
+Categories is applicable and is expressed below in terms of
+{{eta*}}/{{eps*}} and the multiplicative monoid.
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
 
 \paragraph*{Annihilation.}
+As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
+additive {{trace+}}, {{trace*}} gives us the ability to find the
+fixpoint of the contraint (expressed with combinators) that is
+traced. If the constraint has not satisfying values, there is not
+fixpoint possible -- we use the term `annihilation' to describe the
+corresponding undefined state of the program. Annihilation and a
+non-termination both describe undefined program execution -- while
+non-termination characterizes an iteration that fails to terminate,
+annihilation characterizes a constraint that has no fixpoint.
 
 \paragraph*{Zero Information.}
+As with {{eta+}}, we argue that {{eta*}} constructs no new
+information, instead it gives us the ability to explore as many worlds
+as permitted by the type {{b}}. So situate ourselves in any specific
+world, we have to supply additional information in teh form of
+unifying the particle or anti-particle with a particle representing
+known information. Hence only by supplying information to a an
+entangled pair can we read information from it -- and always exactly
+as much information as we supplied.
+
+ADD MATH.
 
@@
 \subsection{SAT solver}
+\label{sec:sat-solver}
 

### 2012-02-29 rational.tex
sat

@@
 
+The construction presented here is a novel SAT-solver that relies on
+{{trace*}} for its solution. It varies from previous classical SAT
+solvers in that there is not explicit search operation on the boolean
+space. It also varies from previous quantum SAT solvers in that it
+does uses a {{trace}} operation to compute a fixpoint which is the
+solution following which an isomorphic clone operation allows us to
+examine the result.
+
+The construction here is not limited to boolean satisfiability, but
+generalises to any constraint satisfaction problem whose search space
+can be represented by a recursive type and for which an isomorphic
+clone operation can be constructed. That said, we will focus on the
+limited case of boolean satisfiability here.
+
+Consider the circuit below that has no observable output -- i.e. for
+all inputs it is annihilating. This happens because it constructs a
+boolean and its dual, negates one of them and attempts to collapse
+them.
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/not_trace.pdf}
+}
+\end{center}
+
+We leverage the basic idea of the above circuit, to collapse all
+values that are not solutions to the specified isomorphic constraint
+{{f}}. The operation {{f}} acts on some inputs and its adjoint
+reconstructs the inputs from the outputs. While the output of {{f}}
+may contain many components, its must contain a boolean indicating if
+the input satisfy {{f}}. The special output wire is used to `control'
+the negation of a boolean `control wire'. If the inputs do not satisfy
+{{f}}, the control wire is negated.
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/sat1.pdf}
+}
+\end{center}
+
+If this construction is traced using {{trace*}}, all the values that
+do not satisfy {{f}} get annihilated because the control wire acts
+like the closed-loop `not' in the previous construction.
+
 %%%%%%%%%%%%%%%%%%%

### 2012-02-29 rational.tex
graphical language

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{A Graphical Language for {{langRev}} }
+
+Combinators of {{langRev}} can be written using the the operators
+described in definition \ref{def:langRev} or via a graphical language
+similar in spirit to those developed for GoI
+\cite{DBLP:conf/popl/Mackie95} and for category theory
+\cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
+conventions and shorthands we will describe here, the wiring diagrams
+are a complete and equivalent to the operator based (syntactic)
+description of programs.  {{langRev}} combinators expressed in this
+graphical language looks like a ``wiring diagrams''. Value take the
+form of ``particles'' that flow along the wires. Computation is
+expressed by the flow of particles.
+
+\begin{itemize}
+\item 
+The simplest sort of diagram is the {{id : b <-> b}} combinator which
+is simply represented as a wire labeled by its type {{b}}. In more
+complex diagrams, if the type of a wire is obvious from the context,
+it may be omitted.
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire.pdf}
+}
+\end{center}
+
+\item
+The product type {{b1*b2}} may be represented both as one wire
+labeled {{b1*b2}} or by two parallel wires labeled {{b1}} and
+{{b2}}. Both representations maybe used interchangeably.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
+\includegraphics{diagrams/thesis/product-one-wire.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
+\includegraphics{diagrams/thesis/product-two-wires.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item
+Sum types may similarly be represented using using parallel wires with
+a {{+}} operator between them.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
+\includegraphics{diagrams/thesis/sum-one-wire.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
+\includegraphics{diagrams/thesis/sum-two-wires.pdf}
+}
+\end{center}
+\end{multicols}
+
+%% \item
+%% When representing complex types like {{(b1*b2)+b3}} some visual
+%% grouping of the wires may be done to aid readability. The exact type
+%% however will always be clarified by the context of the diagram.
+
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
+%% }
+%% \end{center}
+
+\item
+Associativity is entirely skipped in the graphical language. Hence
+three parallel wires maybe inferred as {{b1*(b2*b3)}} or
+{{(b1*b2)*b3}}, based on the context. This is much like handling of
+associativity in the graphical representations of categories
+\cite{springerlink:10.1007/978-3-642-12821-94},
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
+\includegraphics{diagrams/thesis/assoc.pdf}
+}
+\end{center}
+
+\item
+Commutativity is represented by crisscrossing wires.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
+\includegraphics{diagrams/thesis/swap_times.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
+\includegraphics{diagrams/thesis/swap_plus.pdf}
+}
+\end{center}
+
+\end{multicols}
+
+\item
+Identity morphisms and their duals are represented as follows:
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
+\includegraphics{diagrams/thesis/uniti.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
+\includegraphics{diagrams/thesis/unite.pdf}
+}
+\end{center}  
+\end{multicols}
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
+\includegraphics{diagrams/thesis/zeroi.pdf}
+}
+\end{center}
+
+\columnbreak
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
+\includegraphics{diagrams/thesis/zeroe.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item
+Finally, distributivity and factoring are represented using the dual
+boxes shown below:
+
+\begin{multicols}{2}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/distrib-crop.pdf]
+\includegraphics{diagrams/thesis/dist.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/factor-crop.pdf]
+\includegraphics{diagrams/thesis/factor.pdf}
+}
+\end{center}
+\end{multicols}
+
+\end{itemize}
+
+\noindent 
+\textit{Example.} As an example consider the wiring diagram
+of the combinator {{c}} below:
+
+{{c : b * (1+1) <-> b + b}}
+
+{{c = swap* (;) dist (;) (identl* (+) identl*)}}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
+\includegraphics{diagrams/thesis/example1.pdf}
+}
+\end{center}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-02-29 rational.tex
intro

--- 2012-02-28 rational.tex
@@
 
+In a recent paper~\cite{infeffects}, we argued that, by encompassing
+irreversible physical primitives, conventional abstract models of computation
+have also inadvertently included some \emph{implicit} computational effects
+which we called \emph{information effects}. We then developed a pure
+reversible model of computation that is obtained from the type isomorphisms
+and categorical structures that underlie models of linear logic and quantum
+computing and that treats information as a linear resource that can neither
+be erased nor duplicated. In this paper, we show that our pure reversible
+model unveils deeper and more elegant symmetries of computation than have
+previously been reported. In particular, we expose two notions of duality of
+computation: an additive duality and a multiplicative duality that give rise
+to negative types and fractional types respectively. Although these types
+have previously appeared in the literature (see Sec.~\ref{sec:related}), they
+have typically appeared in the context of conventional languages with
+information effects, which obscured their appeal and murked their properties.
+
+To get a feeling for the main idea, consider the following algebraic
+manipulation relating a natural number $a$ to itself:
+\[\begin{array}{rclclclclcl}
+a &=& a + 0 &=& a + ((-a) + a) &=& (a + (-a)) + a &=& 0 + a &=& a
+\end{array}\]
+Although seemingly pointless, this isomorphism has a non-trivial and
+interesting computational interpretation (see Sec.~\ref{sub:neg}).
+
+corresponds to a type isomorphism between the object $a$ and itself.
+
+Furthermore, we illustrate their true appeal and expressiveness is brought
+forth by viewing them in the context of an information-preserving
+computational model.
+
+In addition, we show how these types enrich our computational model, they
+obey the same laws as the rational numbers. In other words, every algebraic
+identity that holds for the rational numbers corresponds to a type
+isomorphism in our model.
+
+Specifically, 
+isomorphisms enrich our computational model with have an interesting
+computational interpretation
+
+ in our
+model. For example,
+
+On the left we have a producer
+that can supply a value of type $a$ and on the right we have a consumer that
+demands a value of type $a$. For the purposes of the example, think of the
+producer as a buyer who has \$20.00 and the consumer as a seller who expects
+\$20.00 for the merchandise. In a conventional computational model, one might
+realize this situation by simply writing the identity function: the buyer
+hands the money to the seller to finish the transaction. The above sequence
+of isomorphisms implements this identity function in a much more interesting
+way, however. 
+of the above series of isomorphisms is that money is ``created'' on the
+consumer side together with a corresponding debt that travels ``backwards''
+until it is paid for by the money at the supplier. In other words, this
+computation models a credit card transaction. It shares the same benefits of
+such transactions: allowing transactions to occur without necessarily
+demanding that the resources be present at the time of the transaction.
+
+On the producer side, the debt is paid for by the money computation with
+the identity function. the producer and consumer must somehow share an
+explicit dependency that allows the value. However in our model, the presence
+of negative types allows the produced value to satisfy the demand without the
+producer or consumer even knowing about each other. As is explained in detail
+in
+
+Indeed, in 
+the above example, the computation on the consumer side creates
+
+In particular, linear logic~\cite{Girard87tcs}, among other contributions,
+exposed an additive/multiplicative distinction in logical connectives and
+rules. In particular, linear logic includes additive disjunctions $\oplus$
+and conjunctions $\with$ as well as multiplicative disjunctions $\parr$ and
+conjuctions $\otimes$. Duality is also prominent in linear logic: it relates
+the additive connectives to each other (the dual of $\oplus$ is $\with$ and
+vice-versa) and the multiplicative connectives to each other (the dual of
+$\otimes$ is $\parr$ and vice-versa).
+
+We furthermore demonstrate that, in our model,
+programming with these negative and fractional types, is a new ``revolution''
+breaking dependencies...
+
+Since Filinski, we've had the
+idea that values and continuations are like mirror images. In a conventional
+language, the negative (continuation) side is implicit and we introduce
+information effects on the positive. Trying to recover the duality from this
+distorted positive side has always been messy. Now it looks clean because we
+have kept the positive side pure.
+
+\newpage
+
 Continuations made their introduction to the world of programming language
@@
 many authors~\cite{Griffin:1989:FNC:96709.96714, curien00duality,
-  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet despite of its appeal, the idea
-that continuations of this
-
-appeared as functional
-representation of labels and jumps,
-Filinski, in 1989, observed that they
-are the categorical introduced the symmetric $\lambda$-calculus
-
-Computationally, symmetry exhibits itself as a duality between
-two concepts.
+  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet, despite its appeal, the duality
+between values and continuations
+
+Computationally, symmetry exhibits itself as a duality between two concepts.
 
@@
 In a seemingly unrelated development, Girard in 1987 introduced linear
-logic~\cite{Girard87tcs} which, among other contributions, exposed an
-additive/multiplicative distinction in logical connectives and rules. In
-particular, linear logic includes additive disjunctions $\oplus$ and
-conjunctions $\with$ as well as multiplicative disjunctions $\parr$ and
-conjuctions $\otimes$. Duality is also prominent in linear logic: it relates
-the additive connectives to each other (the dual of $\oplus$ is $\with$ and
-vice-versa) and the multiplicative connectives to each other (the dual of
-$\otimes$ is $\parr$ and vice-versa).
 
@@
 \subsection{Negative Types}
+\label{sub:neg}
 
@@
 
-ADD MATH.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%
 \subsection{Multiplicative Duality in {{langRev}} }
@@
 
-As with additive duals, we represent {{eta*}} and {{eps*}] using
-`U'-shaped connectors. In the general case the wires for {{1}} will be
-dropped as it can freely introduced or eliminated using {{uniti}} and
-{{unite}}.
-
 \begin{multicols}{2}
@@
 
-The usual interpretation of {{b1*b2}} we have both a value of type
-{{b1}} and a value of type {{b2}}. This interpretation is maintained
-in the precense of fractionals. Hence an {{eta* : 1 <-> b^^^ * b}} is
-to be viewed as a fission point for a value of type {{b}} and its
-multiplicative inverse {{b^^^}}. This naturally raises the question --
-which value of type {{b}} is produced? The isomorphism {{eta*}} does
-not produce any specific value of type {{b}}, instead it produces a
-place holder for a value such that the value of type {{b^^^}} is
-always the dual of the chosen one.
-
+\paragraph*{Unification.}
+
+test test test
 \begin{multicols}{2}
@@
 
-\paragraph*{Unification.}
-A computational interpretation of this is that {{langRevEE}} acts like
-a logic programming system and {{eta*}} acts as the site of creation
-of a new typed logic variable of type {{b}} and its corresponding dual
-of type {{b^^^}}. These two values are free to move independantly
-through a circuit, but any operations on value affect what operations
-are possible on the other since they are both related by an underlying
-logic variable. Dually, {{eps*}} acts as a unification site for values
-flowing along both branches. If the values are not exactly duals, then
-the system fails to unify and hence the program is stuck. If the
-values are exactly duals, then unification succeeds and {{eta*}}
-returns {{1}}.
-
-\paragraph*{Entanglement and Anti-Particles.}
-A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
-quantum mechanics. The {{eta*}} is a site of fission that creates an
-information particle and its anti-particle. Both flow in the same
-direction in time. The produce particles are entangles in that actions
-on one (such as transformation by application of an isomorphism)
-affects the other. Further the fission site does not create a particle
-in any specific state, but in a superposition of all possible states
-as dictated by its type. In each possible world that the particle
-exists, it takes on a specific value inhibiting its type and
-correspondingly its entangled anti-particle takes the dual of the
-specific value.
-
-This also leads to the exiciting point of view that entangled
-particle/anti-particle pairs act as functions in a physical
-first-class sense. We can ask the fuction to be applied to a value by
-unifying the anti-particle with a particle in a well-known state and
-examining is pair. We shall use this point of view in the development
-our SAT solver (see Sec. \ref{sec:sat-solver}).
+\paragraph*{Entanglement.}
 
 \paragraph*{Coherence.}
-As before, the requisite coherence condition on Compact Closed
-Categories is applicable and is expressed below in terms of
-{{eta*}}/{{eps*}} and the multiplicative monoid.
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
 
 \paragraph*{Annihilation.}
-As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
-additive {{trace+}}, {{trace*}} gives us the ability to find the
-fixpoint of the contraint (expressed with combinators) that is
-traced. If the constraint has not satisfying values, there is not
-fixpoint possible -- we use the term `annihilation' to describe the
-corresponding undefined state of the program. Annihilation and a
-non-termination both describe undefined program execution -- while
-non-termination characterizes an iteration that fails to terminate,
-annihilation characterizes a constraint that has no fixpoint.
 
 \paragraph*{Zero Information.}
-As with {{eta+}}, we argue that {{eta*}} constructs no new
-information, instead it gives us the ability to explore as many worlds
-as permitted by the type {{b}}. So situate ourselves in any specific
-world, we have to supply additional information in teh form of
-unifying the particle or anti-particle with a particle representing
-known information. Hence only by supplying information to a an
-entangled pair can we read information from it -- and always exactly
-as much information as we supplied.
-
-ADD MATH.
 
@@
 \subsection{SAT solver}
-\label{sec:sat-solver}
 

### 2012-02-29 rational.tex
intro

@@
   Every functional programmer knows about sum and product types, {{a+b}} and
-  {{a*b}}. Negative and fractional types, $a-b$ and $a/b$ respectively, are
-  much less known and their computational interpretation is unfamiliar and
-  often complicated. We show that in a programming model in which information
-  is preserved (such as the model introduced in our recent paper on
-  \emph{Information Effects}), these types have particularly natural
+  {{a*b}} respectively. Negative and fractional types, $a-b$ and $a/b$
+  respectively, are much less known and their computational interpretation is
+  unfamiliar and often complicated. We show that in a programming model in
+  which information is preserved (such as the model introduced in our recent
+  paper on \emph{Information Effects}), these types have particularly natural
   computational interpretations. Intuitively, values of negative types are
@@
 \[\begin{array}{rclclclclcl}
-a &=& a + 0 &=& a + ((-a) + a) &=& (a + (-a)) + a &=& 0 + a &=& a
+a &=& a + 0 &=& a + ((-a) + a) \\
+  &=& (a + (-a)) + a &=& 0 + a &=& a
 \end{array}\]
-Although seemingly pointless, this isomorphism has a non-trivial and
-interesting computational interpretation (see Sec.~\ref{sub:neg}).
-
-corresponds to a type isomorphism between the object $a$ and itself.
+
+Although seemingly pointless, this algebraic proof corresponds, in our model,
+to an isomorphism with a non-trivial and interesting computational
+interpretation.  On the left we have a producer that can supply a value of
+type $a$ and on the right we have a consumer that demands a value of type
+$a$. For the purposes of the example, think of the producer as someone having
+\$20.00 and the consumer as someone demanding \$20.00. It is certainly
+possible to directly connect the producer and consumer: such a connection
+would be modeled by the identity function. The proof above implements this
+connection in a much more interesting way, however. A transformation $0 =
+(-a) + a$ is used at the consumer site to create money and a corresponding
+debt. The consumer gets the money immediately and the debt travels
+``backwards'' towards the producer where it is paid.  In other words, this
+computation models something similar to a credit card transaction. We discuss
+this example in detail in Sec.~\ref{sub:neg} and only focus on two important
+points here:
+\begin{itemize}
+\item As the example illustrates, negative types corresponds to ``debts.''
+  It would clearly be disastrous if debts could be deleted or
+  duplicated. This simple observation explains why negative types are much
+  simpler and much more appealing in a framework where information is
+  guaranteed to be preserved. In previous work that used negative types,
+  complicated mechanisms are typically needed to constrain the propagation
+  and use of negative values because the surrounding computational framework
+  is, generally speaking, careless in its treatment of information.
+\item The appeal of negative types can already be gleaned from the credit
+  card analogy. The main reason credit card transactions are convenient is
+  because they disentangle the propagation of the money from the propagation
+  of the services. Not every transaction needs both the resources and
+  services to be brought together: it is sufficient to have a promise that
+  the money will be paid at some point, as long as the infrastructure can be
+  trusted that such promises will be fulfilled. 
+\end{itemize}
+
+Similar observations hold for fractional types but we defer these to
+Sec.~\ref{sub:frac}. Before proceeding to the main part of the paper, we
+briefly summarize the our technical contributions:
+\begin{itemize}
+\item 
+\item 
+\item 
+\item 
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Field of Rational Types: Intuitions}
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Negative Types}
+\label{sub:neg}
+
+Consider the following two ways of purchasing an item that costs \$20.00:
+\begin{enumerate}
+\item You give the seller a \$20.00 bill.
+\item You use a credit card to pay the seller.
+\end{enumerate}
+In both cases, the seller receives the money immediately but there is a
+subtle difference. In the first transaction, the money received by the seller
+corresponds to a value that has been produced earlier in the computation. In
+the second transaction, the money may or may not exist yet: computationally,
+a \emph{debt} is generated to compensate for the money received by the seller
+and this debt travels ``backwards'' towards the bank where it is (hopefully)
+reconciled.
+
+The example suggests that the ability to consider values flowing backwards
+enriches our computational model. This observation goes back to Filinski's
+Masters thesis where continuations are identified with these negative values.
+We discuss the connections to Filinski's work and others in more detail in
+Sec.~\ref{sec:related}. For now, we note that in a conventional programming
+language in which values can be copied and deleted at will, extreme care is
+needed to keep track of negative values. Indeed, in the example above, if it
+were possible to simply delete the variable corresponding to the debt, we
+would have produced money out of nothing.
+
+For this reason, our treatment of negative values in the context of
+{{langRev}} is particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
+right-to-left direction. In more detail, we can model the credit card
+transaction above using the following program (shown diagrammatically):
+
+FIGURE
+
+CLEAN UP the following based on the figure. In contrast, in our setting, one
+can start from the empty type $0$, introduce a positive value and its
+negative counterpart, and let each of these flow in arbitrary ways. The
+entire framework guarantees that neither the value nor its negative
+counterpart will be deleted or duplicated and hence that, in any closed
+program, the ``debt'' corresponding to the negative value is paid off exactly
+once. Computationally we model the first transaction using essentially the
+identity function which receives a \$20.00 bill as input from the buyer and
+propagates it on its output to the seller. The second transaction is more
+complicated. We model it as shown in the circuit below: There are two ways to
+understand this circuit that are both quite instructive. Let's first examine
+the type structure of each of the combinators that comprise the circuit. The
+first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
+produced from nothing so to speak by generating an equivalent debt that
+travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
+the execution. It goes forward in time so to speak, comes back, and then goes
+forwards again.
+
+It is critical that the framework in which the negative types are introduced
+is a framework in which all information is preserved, with no duplication or
+erasure. This guarantees that the generated debts are paid once and exactly
+once.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Fractional Types}
+
+The type $a/b$ is prominent in the Lambek-Grishin calculus which is
+extensively used in computational linguistics. In the common interpretation,
+a value of type $a/b$ is a value of some type $c$ such that when put in a
+context of type $b$, the result is a value of type $a$. For example, assuming
+contexts are represented as functions, a value of type $\texttt{bool} /
+(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
+$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
+in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
+type $\texttt{bool}$.
+
+In our case, the situation is simpler. If the goal is to produce a value of
+type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
+have type {{(a * b) / b}}. In most settings, this type would be equivalent to
+{{a}}. However in our setting, the constraint that this value must eventually
+fit in a larger value that supplies the {{b}} is explicitly recorded and must
+be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
+which when used in the left-to-right direction allows the creation of a value
+and a corresponding constraint on the context.  Both the value and the
+constraint can flow in arbitrary ways during the computation but eventually
+in a closed program with no ``dangling'' constraints, the isomorphism should
+be used in the right-to-left direction to resolve the constraints.  To
+understand the computational interpretation, consider the following example.
+
+The goal is to produce a value of type $(\texttt{bool} \times
+\texttt{int})$. One part of the program, $e_1$, knows how to produce the
+value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
+to produce the value of type $\texttt{bool}$ (say true). Computationally, we
+model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
+uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
+is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
+$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
+$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
+the two values meet, we can group the components as follows:
+\[
+(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
+\]
+Using the isomorphism in the right-to-left direction on the last two pairs,
+forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
+{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
+with the pair $(true,3)$ as desired.
+
+To summarize, one can think of a value of type $a/b$ as a value that imposes
+a constraint on its context of use: it is a value that requires its context
+to be of type $b$ and only if that condition is satisfied, can the value be
+considered as a value of type $a$. In the simplest case, the type $1/b$ just
+constraints the context to be of type $b$. By allowing the fractional types
+to be first-class, we can separate the generation of constraints from their
+use. Both the value and the constraint can flow in arbitrary ways during the
+computation but eventually in a closed program with no ``dangling''
+constraints, the isomorphism should be used in the right-to-left direction to
+resolve the constraints.
+
+NEED to work out the example in detail (or perhaps another example that's
+better.
+
+Again it is critical that the framework in which the fractional types are
+introduced is a framework in which all information is preserved, with no
+duplication or erasure. This guarantees that the generated constraints are
+satisfied once and exactly once.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Summary}
+
+To summarize negative, fractional, square root, and imaginary types all make
+sense. What they help you accomplish as a programmer is to disassociate
+global invariants into local ones that can be satisfied independently by
+subcomputations with no synchronization or communication. A computation
+producing something of type $a/b$ does not need to concern itself with who is
+going to supply the missing $b$: it just does its part. Conversely faced with
+a complicated task, a computation might decide to break it into pieces and
+demand these pieces using negative types.
+
+It is no surprise that these types are closely related to quantum mechanics
+and that they give us the feel that this is how nature computes. This is
+speculation however.
+
+In any case, in a framework where information can be copied and deleted, none
+of this makes much sense. It is critical that these constraints and demands
+can neither be duplicated nor erased.  This gives us the maximum
+``parallelism'' possible.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRev}} }
+
+The presentation here differs from \cite{infeffects} in the sense that
+we have introduced the {{0}} type here, thereby completing the monoid
+{{(0,+)}} and we present a small-step operational semantics.
+
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b + b | b * b 
+% values, v ::= () | left v | right v | (v, v)
+
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%zeroe :&  0 + b <-> b &: zeroi
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist0 :& 0 * b <-> 0 &: factor0
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+
+%subcode{proof} include main
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+
+\begin{definition}{(Syntax of {{langRev}})}
+\label{def:langRev}
+We collect our types, values, and combinators, to get the full language
+definition.
+%subcode{bnf} include main
+% value types, b ::= 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% comb.~types, t ::= b <-> b
+% iso ::= swap+ | assocl+ | assocr+ 
+%     &|& identl* | identr* 
+%     &|& swap* | assocl* | assocr* 
+%     &|& dist | factor 
+% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+\end{definition}
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Semantics}
+
+%subcode{bnf} include main
+% Combinators, c = iso | c (;) c | c (*) c | c (+) c
+% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+%                  &|& LeftTimes cc c v | RightTimes c v cc 
+%                  &|& LeftPlus cc c | RightPlus c cc 
+% Values, v = () | (v, v) | L v | R v
+%
+% Machine states = <c, v, cc> | {[c, v, cc]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
+\begin{scriptsize}
+%subcode{opsem} include main
+%! columnStyle = rlcl
+% swap+ & (left v) &|-->& right v
+% swap+ & (right v) &|-->& left v 
+% assocl+ & (left v1) &|-->& left (left v1)
+% assocl+ & (right (left v2)) &|-->& left (right v2)
+% assocl+ & (right (right v3)) &|-->& right v3 
+% assocr+ & (left (left v1)) &|-->& left v1
+% assocr+ & (left (right v2)) &|-->& right (left v2)
+% assocr+ & (right v3) &|-->& right (right v3)
+% identl* & ((), v) &|-->& v 
+% identr* & v &|-->& ((), v) 
+% swap* & (v1, v2) &|-->& (v2, v1) 
+% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
+% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
+% dist & (left v1, v3) &|-->& left (v1, v3)
+% dist & (right v2, v3) &|-->& right (v2, v3)
+% factor & (left (v1, v3)) &|-->& (left v1, v3) 
+% factor & (right (v2, v3)) &|-->& (right v2, v3)   
+\end{scriptsize}
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
+% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
+% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
+% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
+% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
+% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
+% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
+% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
+% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
+% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
+% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoid}
+{{langRev}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:category}
+{{langRev}} is a dagger symmetric monoidal category. 
+\end{proposition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{A Graphical Language for {{langRev}} }
+
+Combinators of {{langRev}} can be written using the the operators
+described in definition \ref{def:langRev} or via a graphical language
+similar in spirit to those developed for GoI
+\cite{DBLP:conf/popl/Mackie95} and for category theory
+\cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
+conventions and shorthands we will describe here, the wiring diagrams
+are a complete and equivalent to the operator based (syntactic)
+description of programs.  {{langRev}} combinators expressed in this
+graphical language looks like a ``wiring diagrams''. Value take the
+form of ``particles'' that flow along the wires. Computation is
+expressed by the flow of particles.
+
+\begin{itemize}
+\item 
+The simplest sort of diagram is the {{id : b <-> b}} combinator which
+is simply represented as a wire labeled by its type {{b}}. In more
+complex diagrams, if the type of a wire is obvious from the context,
+it may be omitted.
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire.pdf}
+}
+\end{center}
+
+\item
+The product type {{b1*b2}} may be represented both as one wire
+labeled {{b1*b2}} or by two parallel wires labeled {{b1}} and
+{{b2}}. Both representations maybe used interchangeably.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
+\includegraphics{diagrams/thesis/product-one-wire.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
+\includegraphics{diagrams/thesis/product-two-wires.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item
+Sum types may similarly be represented using using parallel wires with
+a {{+}} operator between them.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
+\includegraphics{diagrams/thesis/sum-one-wire.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
+\includegraphics{diagrams/thesis/sum-two-wires.pdf}
+}
+\end{center}
+\end{multicols}
+
+%% \item
+%% When representing complex types like {{(b1*b2)+b3}} some visual
+%% grouping of the wires may be done to aid readability. The exact type
+%% however will always be clarified by the context of the diagram.
+
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
+%% }
+%% \end{center}
+
+\item
+Associativity is entirely skipped in the graphical language. Hence
+three parallel wires maybe inferred as {{b1*(b2*b3)}} or
+{{(b1*b2)*b3}}, based on the context. This is much like handling of
+associativity in the graphical representations of categories
+\cite{springerlink:10.1007/978-3-642-12821-94},
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
+\includegraphics{diagrams/thesis/assoc.pdf}
+}
+\end{center}
+
+\item
+Commutativity is represented by crisscrossing wires.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
+\includegraphics{diagrams/thesis/swap_times.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
+\includegraphics{diagrams/thesis/swap_plus.pdf}
+}
+\end{center}
+
+\end{multicols}
+
+\item
+Identity morphisms and their duals are represented as follows:
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
+\includegraphics{diagrams/thesis/uniti.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
+\includegraphics{diagrams/thesis/unite.pdf}
+}
+\end{center}  
+\end{multicols}
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
+\includegraphics{diagrams/thesis/zeroi.pdf}
+}
+\end{center}
+
+\columnbreak
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
+\includegraphics{diagrams/thesis/zeroe.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item
+Finally, distributivity and factoring are represented using the dual
+boxes shown below:
+
+\begin{multicols}{2}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/distrib-crop.pdf]
+\includegraphics{diagrams/thesis/dist.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/factor-crop.pdf]
+\includegraphics{diagrams/thesis/factor.pdf}
+}
+\end{center}
+\end{multicols}
+
+\end{itemize}
+
+\noindent 
+\textit{Example.} As an example consider the wiring diagram
+of the combinator {{c}} below:
+
+{{c : b * (1+1) <-> b + b}}
+
+{{c = swap* (;) dist (;) (identl* (+) identl*)}}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
+\includegraphics{diagrams/thesis/example1.pdf}
+}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ {{langRevEE}} }
+
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
+input {{v1 : b1}} gives us a value {{v2 : b2}} on
+termination. Execution is well defined only if {{b1}} and {{b2}} are
+entirely positive types. Consider the program that
+
+(TODO: So this is a really bad explanation. But this section is really
+important and we need to explain carefully. Maybe an analogy can be
+drawn to Cayley diagrams for groups to point out that the notion of
+`i' is entirely artificial, but it is essential to the discourse.)
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The outputs of
+programs that output mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Additive Duality in {{langRev}} }
+
+Additive duality is added to {{langRev}} by addition of negative types
+denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
+extend the language as follows:
+
+%subcode{bnf} include main
+% Value Types, b = ... | -b
+% Values, v = ... | -v
+%
+% Isomorphisms, iso &=& ... | eta+ | eps+
+
+\noindent
+with the types judgments
+
+%subcode{opsem} include main
+% eta+ &: 0 <-> (-b) + b :& eps+
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
+the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
+dually on the right is {{eps+}} showing the map from {{-b+b}} to
+0. Even though the diagrams below show a {{0}} wire for completeness,
+later diagrams will always drop them in contexts where we can
+implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps.pdf}
+\end{center}
+\end{multicols}
+
+The usual interpretation of the type {{b1+b2}} is that we either have
+an appropriately tagged value of type {{b1}} or an appropriately
+tagged value of type {{b2}}. This interpretation is maintained in the
+presence of {{eta+}} and {{eps+}} in the following sense: a value of
+type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
+switches into a value {{L (-v): -b + b}} that flows backwards on the
+upper wire. The inversion of direction is captured by the negative
+sign on the value.
+
+\paragraph*{Negative Information Flow.} 
+Hence, the intuitive way to read the diagrams is that they reverse the
+flow of information -- in normal {{langRev}} circuits information
+flows from left-to-right and {{eps+}} operations cause information to
+flow from right-to-left.  The interpretation of {{eta+}} is precisely
+the opposite: it turns a right-to-left flow of information in a
+circuit into a left-to-right flow.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus2.pdf}
+}
+\end{center}  
+\end{multicols}
+
+\paragraph*{Time Traveling.}
+A time traveling intuition is also applicable here: if view the
+left-to-right direction of information flow as the canonical forward
+direction of information, then the action of {{eps+}} allows values,
+aka information particles, to flow backwards in time. Particles that
+flow backward in time get to see and interact with the history of
+particles that they coexist with i.e. values that they exist
+multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
+(whose construction is shown in Section \ref{sec:neg-constructions}).
+
+\paragraph*{Coherence.}
+The appropriate coherence condition for Compact Closed Categories,
+expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
+may be visualized as explained by Selinger
+\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+\paragraph*{Non-termination.}
+As is well known in category theory, Compact Closed Categories contain
+trace operations.  In the context of {{langRev}} this implies that
+{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
+(detailed in Section \ref{sec:monoidal-constructions}) hence
+introducing \emph{non-termination} into the language
+\cite{infeffects}.
+
+\paragraph*{Zero Information.}
+The fact that {{eta+}} (and hence {{eps+}}) introduces no information
+can be understood is by viewing {{-b+b}} as a type whose value is
+observable only if you `pay' {{b}} amount of information to the {{-b}}
+branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
+amount of information is returned to you. (TODO. more needs to be said
+here.)
+
+ADD MATH.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Multiplicative Duality in {{langRev}} }
+
+Multiplicative duality is added to {{langRev}} by adding types
+{{b^^^}}, values {{v^^^}} and isomorphisms {{eta*}} and {{eps*}}.
+Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
+{{a/b}}.
+
+%subcode{bnf} include main
+% Value Types, b = ... | b^^^
+% Values, v = ... | v^^^
+%
+% Isomorphisms, iso &=& ... | eta* | eps*
+
+\noindent
+These have the type judgments
+
+%subcode{opsem} include main
+% eta* &: 1 <-> b^^^ * b :& eps*
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- v^^^ : b^^^
+
+As with additive duals, we represent {{eta*}} and {{eps*}] using
+`U'-shaped connectors. In the general case the wires for {{1}} will be
+dropped as it can freely introduced or eliminated using {{uniti}} and
+{{unite}}.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta_times.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps_times.pdf}
+\end{center}
+\end{multicols}
+
+The usual interpretation of {{b1*b2}} we have both a value of type
+{{b1}} and a value of type {{b2}}. This interpretation is maintained
+in the precense of fractionals. Hence an {{eta* : 1 <-> b^^^ * b}} is
+to be viewed as a fission point for a value of type {{b}} and its
+multiplicative inverse {{b^^^}}. This naturally raises the question --
+which value of type {{b}} is produced? The isomorphism {{eta*}} does
+not produce any specific value of type {{b}}, instead it produces a
+place holder for a value such that the value of type {{b^^^}} is
+always the dual of the chosen one.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eta_times1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_times1.pdf}
+}
+\end{center}  
+\end{multicols}
+
+\paragraph*{Unification.}
+A computational interpretation of this is that {{langRevEE}} acts like
+a logic programming system and {{eta*}} acts as the site of creation
+of a new typed logic variable of type {{b}} and its corresponding dual
+of type {{b^^^}}. These two values are free to move independantly
+through a circuit, but any operations on value affect what operations
+are possible on the other since they are both related by an underlying
+logic variable. Dually, {{eps*}} acts as a unification site for values
+flowing along both branches. If the values are not exactly duals, then
+the system fails to unify and hence the program is stuck. If the
+values are exactly duals, then unification succeeds and {{eta*}}
+returns {{1}}.
+
+\paragraph*{Entanglement and Anti-Particles.}
+A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
+quantum mechanics. The {{eta*}} is a site of fission that creates an
+information particle and its anti-particle. Both flow in the same
+direction in time. The produce particles are entangles in that actions
+on one (such as transformation by application of an isomorphism)
+affects the other. Further the fission site does not create a particle
+in any specific state, but in a superposition of all possible states
+as dictated by its type. In each possible world that the particle
+exists, it takes on a specific value inhibiting its type and
+correspondingly its entangled anti-particle takes the dual of the
+specific value.
+
+This also leads to the exiciting point of view that entangled
+particle/anti-particle pairs act as functions in a physical
+first-class sense. We can ask the fuction to be applied to a value by
+unifying the anti-particle with a particle in a well-known state and
+examining is pair. We shall use this point of view in the development
+our SAT solver (see Sec. \ref{sec:sat-solver}).
+
+\paragraph*{Coherence.} 
+As before, the requisite coherence condition on Compact Closed
+Categories is applicable and is expressed below in terms of
+{{eta*}}/{{eps*}} and the multiplicative monoid.
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+\paragraph*{Annihilation.}
+As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
+additive {{trace+}}, {{trace*}} gives us the ability to find the
+fixpoint of the contraint (expressed with combinators) that is
+traced. If the constraint has not satisfying values, there is not
+fixpoint possible -- we use the term `annihilation' to describe the
+corresponding undefined state of the program. Annihilation and a
+non-termination both describe undefined program execution -- while
+non-termination characterizes an iteration that fails to terminate,
+annihilation characterizes a constraint that has no fixpoint.
+
+\paragraph*{Zero Information.}
+As with {{eta+}}, we argue that {{eta*}} constructs no new
+information, instead it gives us the ability to explore as many worlds
+as permitted by the type {{b}}. So situate ourselves in any specific
+world, we have to supply additional information in teh form of
+unifying the particle or anti-particle with a particle representing
+known information. Hence only by supplying information to a an
+entangled pair can we read information from it -- and always exactly
+as much information as we supplied.
+
+ADD MATH.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Semantics}
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Syntax}
+
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
+%
+% Sequential Contexts, P, F = [] | P:c
+% Parallel Contexts, D = [] 
+%                     &|& D:(P| [](+)c| F)
+%                     &|& D:(P| c(+)[]| F)
+%                     &|& D:(P| [] (*) c v| F)
+%                     &|& D:(P| [] (*) v c| F)
+%                     &|& D:(P| c v (*)  []| F)
+%                     &|& D:(P| v c (*) []| F)
+%
+%
+% Machine States  = D[P| c v| F] | D[P| v c| F]
+% Start State  = [][[]| v c| []]
+% Stop State  = [][P| c v| []]
+
+Combinator reconstruction, {{P[c]}}:
+%subcode{opsem} include main
+% [](c) '= c
+% P:c'(c) '= P(c'(;)c)
+
+%%%%%%%%%%%%%%%%%%%%%
+\subsection{Operational Semantics}
+
+The small step semantics present for {{langRev}} below work
+symmetrically for forward and backward evaluation.
+
+\begin{itemize}
+\item 
+Basic reduction of an isomorphism. Note that the evaluation leaves the
+adjoint of the combinator behind. This will become important when we
+reverse the direction of computation.
+%% %subcode{opsem} include main
+%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
+
+%subcode{opsem} include main
+% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
+
+\item
+Sequencing involves pushing and popping from the Future and Past
+continuations:
+%subcode{opsem} include main
+% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
+% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
+
+\item
+Parallel composition, captures the current Future and Past and extends
+the parallel context.
+%subcode{opsem} include main
+% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
+% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
+% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
+% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
+
+\item
+Similarly for products:
+%subcode{opsem} include main
+% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
+% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
+% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
+
+and symmetrical rules for evaluation along the second branch. 
+%subcode{opsem} include main
+% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
+% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
+
+The later two rules will be relevant only for reverse execution.
+
+\end{itemize}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrevEE}
+{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+\end{proposition}
+
+\begin{proposition}[Groupoid]
+\label{prop:groupoidEE}
+{{langRevEE}} is a groupoid. 
+\end{proposition}
+
+\begin{proposition}
+\label{prop:categoryEE}
+{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
+\end{proposition}
+
+Note: We use `Monoidal Closed' in the sense of Compact Closed. While
+the category has an multiplicative tensor that distributes over an
+additive tensor, it does not have injections or projections and hence
+does not have categorical sums and products.
+
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Rules for {{eta}} and {{eps}} }
+
+The operation {{eps}} reverses the direction of a particle by
+reversing the world.
+
+Note that we deviate from the categorical definition of {{eta}} and
+{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
+choice of {{eta}}. This however does not affect us, because we deal
+with a symmetric category.
+
+\begin{itemize}
+\item Grammar
+%subcode{bnf} include main
+% Values, v = () | (v, v) | L v | R v | -v
+% Isomorphisms, iso &=& ... | eta | eps
+% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
+
+\item
+Type judgment.
+%subcode{opsem} include main
+% eta &: 0 <-> (-b) + b :& eps
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+\item
+Operational Semantics.
+%subcode{opsem} include main
+% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
+% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
+
+Note: there is NO reduction rule for {{eta}}.
+
+\item
+The adjoint of a parallel context is defined to be:
+%subcode{opsem} include main
+% []{dagger} '= []
+% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
+% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
+% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
+% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
+% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
+% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions in {{langRevEE}} }
+\label{sec:monoidal-constructions}
+
+These constructions are applicable to the additive and the
+multiplicative monoid.
+
+\paragraph*{Functions}
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
+
+\paragraph*{Function application}
+
+\begin{center}
+  \includegraphics{diagrams/apply1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/apply2.pdf}
+\end{center}
+
+\paragraph*{Function composition}
+
+\begin{center}
+  \includegraphics{diagrams/compose1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/compose.pdf}
+\end{center}
+
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+\paragraph*{Delimited continuation}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+\paragraph*{Trace }
+
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+\paragraph*{Principium Contradictiones}
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Negation}
+\label{sec:neg-constructions}
+
+\paragraph*{Negation distributes over {{+}}. }
+
+{{-(b1 + b2) <-> (b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+\paragraph*{ {{eps_{fst} }} }
+
+{{(-b1)*b2 + b1*b2 <-> 0}}
+
+\begin{center}
+  \includegraphics{diagrams/eps_fst.pdf}
+\end{center}
+
+\paragraph*{Lifting negation out of {{*}}. }
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/mult_neg.pdf}
+\end{center}
+
+The following isomorphism can be constructed similarly
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
+
+\paragraph*{Lifting a operation of positive types to negated types}
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions Specific to Multiplication}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{To think about}
+
+\begin{itemize}
+
+\item If we built an effect over {{langRevEE}}, say {{create}} and
+  {{erase}}, are effects structured by an arrow or a monad now?
+
+\item Which operations can be lifted to work on negative types?
+
+\item Can the operational semantics for {{langRevEE}} be an
+  interpreter that is implemented in {{langRevT}} (similar to how the
+  the tree traversal interpreter was implemented). This would imply
+  the existence of a more powerful construction than Int, wherein the
+  products would also be preserved. This is possibly worth a paper in
+  itself.
+
+\item These functions aren't really values. There is no value one can
+  produce that denotes a function. These functions are the ability to
+  transform a value - the possibility of transforming a value.  In the
+  product encoding of environments, there is no value one can assign
+  to a variable such that it denotes a function.
+
+Actually this is possible. Consider two functions {{f : b1 <-> b2}}
+and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
+discriminate them. Hence the {{boo}} is the first class representation
+of the functions and can be thought of as the address of the function.
+
+\begin{center}
+  \includegraphics{diagrams/dispatch.pdf}
+\end{center}
+
+\item It is not fair to say that negative types flow backwards. The
+  following circuits are valid in {{langRevEE}}. It is however proper
+  to say that for any type {{b}} that flows in a direction, the type
+  {{-b}} flows in the reverse direction.
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/neg_circuit2.pdf}
+\end{center}
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Advanced Programming Examples}
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{SAT solver}
+\label{sec:sat-solver}
+
+The construction presented here is a novel SAT-solver that relies on
+{{trace*}} for its solution. It varies from previous classical SAT
+solvers in that there is not explicit search operation on the boolean
+space. It also varies from previous quantum SAT solvers in that it
+does uses a {{trace}} operation to compute a fixpoint which is the
+solution following which an isomorphic clone operation allows us to
+examine the result.
+
+The construction here is not limited to boolean satisfiability, but
+generalises to any constraint satisfaction problem whose search space
+can be represented by a recursive type and for which an isomorphic
+clone operation can be constructed. That said, we will focus on the
+limited case of boolean satisfiability here.
+
+Consider the circuit below that has no observable output -- i.e. for
+all inputs it is annihilating. This happens because it constructs a
+boolean and its dual, negates one of them and attempts to collapse
+them.
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/not_trace.pdf}
+}
+\end{center}
+
+We leverage the basic idea of the above circuit, to collapse all
+values that are not solutions to the specified isomorphic constraint
+{{f}}. The operation {{f}} acts on some inputs and its adjoint
+reconstructs the inputs from the outputs. While the output of {{f}}
+may contain many components, its must contain a boolean indicating if
+the input satisfy {{f}}. The special output wire is used to `control'
+the negation of a boolean `control wire'. If the inputs do not satisfy
+{{f}}, the control wire is negated.
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/sat1.pdf}
+}
+\end{center}
+
+If this construction is traced using {{trace*}}, all the values that
+do not satisfy {{f}} get annihilated because the control wire acts
+like the closed-loop `not' in the previous construction.
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{GoI machine}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related Work} 
+\label{sec:related}
+
+(NOTE: Lets use this section here to establish a long of ideas and
+intuition. None of these connections have to be formal, but they can
+be concrete.)
+
+Filinski proposes that continuations are a \emph{declarative} concept. He,
+furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+which values and continuations are treated as opposites. This is essentially
+what we are proposing with one fundamental difference: our underlying
+language is not the $\lambda$-calculus but a language of pure isomorphisms in
+which information is preserved. This shift of perspective enables us to
+distill and generalize the duality of values and continuations: in
+particular, in the conventional $\lambda$-calculus setting values and
+continuations can be erased and duplicated which makes it difficult to
+maintain the correspondence between a value and its negative counterpart.
+
+\paragraph*{Continuations.} The idea of using negative types to model
+information flowing backwards, demand for values, continuations, etc. goes
+back to at Filinski's thesis. We recall these connections below but we first
+note that all these systems are complicated because in all these systems
+information can be ignored, destroyed, or duplicated. Clearly the possibility
+of erasure of information would mean that our credit card transaction is
+incorrect. In our work, information is maintained and hence we have a
+guarantee that, in a closed program, the debt must be accounted and paid for.
+
+\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
+In his Masters thesis, Filinski proposes that continuations are a
+\emph{declarative} concept. He, furthermore, introduces a symmetric extension
+of the $\lambda$-calculus in which values and continuations are treated as
+opposites. This is essentially what we are proposing with one fundamental
+difference: our underlying language is not the $\lambda$-calculus but a
+language of pure isomorphisms in which information is preserved. This shift
+of perspective enables us to distill and generalize the duality of values and
+continuations: in particular, in the conventional $\lambda$-calculus setting
+values and continuations can be erased and duplicated which makes it
+difficult to maintain the correspondence between a value and its negative
+counterpart. In contrast, in our setting, one can start from the empty type
+$0$, introduce a positive value and its negative counterpart, and let each of
+these flow in arbitrary ways. The entire framework guarantees that neither
+the value nor its negative counterpart will be deleted or duplicated and
+hence that, in any closed program, the ``debt'' corresponding to the negative
+value is paid off exactly once. The forward and backward executions in our
+framework correspond to call-by-value and call-by-name. This duality was
+observed by Filinski and others following him but it is particularly clean in
+our framework.
+
+\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
+subtraction type because its ``computational interpretation is not
+familiar.'' Curien and Herbelin study duality in classical logic, show that
+it exchanges call-by-value and call-by-name. They extend classical natural
+deduction with subtraction but give it no computational meaning. Crolard (in
+the formulae-as-types interpretation of subtractive logic) address the
+computational interpretation of subtraction. He explains the type $A-B$ as
+the of \emph{coroutines} with a local environment of type $A$ and a
+continuation of type $B$. The description is complicated by what is
+essentially the desire to enforce linearity constraints so that coroutines
+cannot access the local environment of other coroutines. Must cite Selinger
+control categories in this context of duality but I am not sure what to say:
+it assumes cartesian closed categories for one thing and not symmetric
+monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
+explain delimited continuations.
+
+\paragraph*{Linear Logic} In accounts that are linear, the value and
+continuation that comprise the subtractive type need to be constrained to
+``stay together.'' This can be achieved by various restrictions. In this work
+we have no such constraints, the negative value can flow anywhere. The entire
+system guarantees that any closed program would have to account for it. We
+don't have to introduce special constraints to achieve that. Zeilberger in
+the paper on polarity and the logic of delimited continuations uses polarized
+logic: he shows that if positive and negative values are completely symmetric
+except that answer types are positive, then the framework accommodates
+delimited continuations. But he interprets negative values are control
+operators, or as values defined by the shape of their continuations. We
+simply interpret values of negative type as values flowing in the ``other''
+direction.
+
+\paragraph*{Int Construction.}
+For a traced monoidal category {{C}} the Int construction produces a
+Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
+Further we know that the target of the Int construction is isomorphic
+to the target of \G construction of Abramsky \cite{Abramsky96:0} from
+Haghverdi. However, note that the {{langRevEE}} is not the same as the
+image of the Int construction on {{langRevT}}, since the the later
+lacks a multiplicative tensor that distributes over the additive tensor
+in Int {{langRevT}}.
+
+\paragraph*{Recursion.} In our previous work we introduce recursive types and
+trace operators. This is dangerous here because infinite loops allow us to
+prolong paying the debt for as long as we want.
+
+\paragraph*{GoI machines} 
+We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Field of Algebraic Numbers}
+
+The field of algebraic numbers consist of the rationals extended with
+solutions to polynomials. In a limited sense {{langRevEE}} with
+recursive types already include polynomials such as the definition of
+binary tree {{@@x.(1+x*x)}} implies the equality {{t = 1 + t*t}} and
+hence {{t^2-t+1=0}}. While this equation does indeed have an
+irrational imaginary root, as we will see -- {{langRevEE}} does not
+have the means of expressing meaningful computation that uses the
+root.
+
+Here is one way to view computations involving irrationals that might
+shed some insight into how we should view recursive data types
+also. For finite datatypes say of size {{n}} we can say that an
+element specifies a choice of {{1/n}} and hence contains as much
+information. For finite things, each value is computationally equal to
+any other value. Any permutation is possible among the values. Hence
+there is no metric possible on the values.  Recursive data types, such
+as binary trees, have an infinite number of elements and hence the
+argument that says that value specifies {{1/n}} units of information
+is not longer valid. Instead to talk meaningfully about information
+content, we should talk about a probability distribution on the values
+of the type. However what shall we base this distribution on? Is there
+a natural order beyond the unfolding order of the values of a type?
+
+In this context, let us go back to the thought that recursive such as
+{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
+interesting solution of trees is precisely that they involve
+irrational numbers. Our usual notation for irrationals such as
+{{2^{1/2} }} can actually be viewed as a thunk representing the
+infinite computation {{1.4142..}}. Irrationals in this sense have a
+very different information theoretic interpretation from full real
+numbers. With Chaitin's Omega numbers he postulates the existence of
+real numbers that have an infinite amount of information -- ones for
+example are the solution to the question about the halting properties
+of every Turning machine.
+
+Real numbers in that sense really are a series of bits of unknown
+information theoretic interpretation -- a sort of {{top}}. Numbers
+such as {{2^{1/2} }} are very different in the sense that even though
+they contain an infinite sequence of bits, there is a finite program
+that generates those bits and its only the cost of computing bits that
+is infinite.  The consistency of arithmetic isn't affected by
+extending the rational field by adding any specific computable
+irrational. So it might be possible to design a model of computing
+where we can indeed deal with them. Such a model will have to explicit
+thunks for the delayed computation represented by trees and never
+equate and cancel two infinite computations unless the represent
+exactly the same infinite computation.
+
+TODO.
+
+Square root and imaginary types have also appeared in the literature: we
+relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
+explanation. We have so far extended the set of types to be the rational
+numbers. Now we will push this and extend the set of types to algebraic
+numbers. In other words, we will allow datatypes defined by arbitrary
+polynomials and allow the roots of such polynomials to be types.
+
+Consider first an example in which we want to compute with the sides of a
+rectangle whose area is 91 and whose length is longer than its width by 6
+units. One can solve the quadratic equation to determine that the sides are 7
+and 13 and proceed. This however prematurely forces us to globally solve the
+constraint. Instead we can let the two sides of the rectangle be $x$ and
+$x+6$ and use the following equation to capture the desired constraint:
+\[
+x^2 + 6x - 91 = 0
+\]
+The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
+the type $0$. We can now proceed to compute with the unknown $x$, being
+assured that in a closed program, our computation will eventually be
+consistent with the solution of the algebraic equation.
+
+Previously the most famous example of a similar nature is the puzzling
+isomorphism that one can establish between seven binary trees and one.
+A binary tree is defined by the datatype
+\[
+x = 1 + x * x 
+\]
+which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
+algebraic manipulation we can reason as follows:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+Fiore poses the question of why such algebraic manipulation would make sense
+type theoretically but states that even though some of the intermediate steps
+make no sense, the final equivalence is valid and can be used to actually
+construct an isomorphism between $x^7$ the type of seven binary trees 
+and $x$ the type of binary trees.
+
+Discussion of possible polynomials:
+\begin{itemize}
+\item $b=1+1$. Boring.
+\item $b=b+1$. That introduces the natural numbers. No solution for this
+  polynomial over the algebraic numbers. We must extend the numbers with
+  $\omega$ to get a solution. We reject this in this paper and prefer to
+  stick to algebraic numbers. The advantage is that all the isomorphisms are
+  valid numerically. With the above type we could subtract $b$ from both
+  sides to show that $0=1$ which is nonsense. We can however have infinite
+  types as long as they have algebraic solutions.
+\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
+  boolean! If we had superpositions, we could then write a function that
+  performs the square root of negation in the sense that applying it twice
+  would be equivalent to boolean negation.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+The simplest way to connect the In a conventional computational model, one
+might realize this situation by simply writing the identity function: the
+buyer hands the money to the seller to finish the transaction. The above
+sequence of isomorphisms implements this identity function in a much more
+interesting way, however.  of the above series of
+
+On the producer side, the debt is paid for by the money computation with the
+identity function. the producer and consumer must somehow share an explicit
+dependency that allows the value. However in our model, the presence of
+negative types allows the produced value to satisfy the demand without the
+producer or consumer even knowing about each other. As is explained in detail
+in
 
@@
 
-Specifically, 
-isomorphisms enrich our computational model with have an interesting
-computational interpretation
-
- in our
-model. For example,
-
-On the left we have a producer
-that can supply a value of type $a$ and on the right we have a consumer that
-demands a value of type $a$. For the purposes of the example, think of the
-producer as a buyer who has \$20.00 and the consumer as a seller who expects
-\$20.00 for the merchandise. In a conventional computational model, one might
-realize this situation by simply writing the identity function: the buyer
-hands the money to the seller to finish the transaction. The above sequence
-of isomorphisms implements this identity function in a much more interesting
-way, however. 
-of the above series of isomorphisms is that money is ``created'' on the
-consumer side together with a corresponding debt that travels ``backwards''
-until it is paid for by the money at the supplier. In other words, this
-computation models a credit card transaction. It shares the same benefits of
-such transactions: allowing transactions to occur without necessarily
-demanding that the resources be present at the time of the transaction.
-
-On the producer side, the debt is paid for by the money computation with
-the identity function. the producer and consumer must somehow share an
-explicit dependency that allows the value. However in our model, the presence
-of negative types allows the produced value to satisfy the demand without the
-producer or consumer even knowing about each other. As is explained in detail
-in
-
-Indeed, in 
-the above example, the computation on the consumer side creates
+Specifically, isomorphisms enrich our computational model with have an
+interesting computational interpretation
 
@@
 have kept the positive side pure.
-
-\newpage
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Field of Rational Types: Intuitions}
-
-%%%%%%%%%%%%%%%%%%
-\subsection{Negative Types}
-\label{sub:neg}
-
-Consider the following two ways of purchasing an item that costs \$20.00:
-\begin{enumerate}
-\item You give the seller a \$20.00 bill.
-\item You use a credit card to pay the seller.
-\end{enumerate}
-In both cases, the seller receives the money immediately but there is a
-subtle difference. In the first transaction, the money received by the seller
-corresponds to a value that has been produced earlier in the computation. In
-the second transaction, the money may or may not exist yet: computationally,
-a \emph{debt} is generated to compensate for the money received by the seller
-and this debt travels ``backwards'' towards the bank where it is (hopefully)
-reconciled.
-
-The example suggests that the ability to consider values flowing backwards
-enriches our computational model. This observation goes back to Filinski's
-Masters thesis where continuations are identified with these negative values.
-We discuss the connections to Filinski's work and others in more detail in
-Sec.~\ref{sec:related}. For now, we note that in a conventional programming
-language in which values can be copied and deleted at will, extreme care is
-needed to keep track of negative values. Indeed, in the example above, if it
-were possible to simply delete the variable corresponding to the debt, we
-would have produced money out of nothing.
-
-For this reason, our treatment of negative values in the context of
-{{langRev}} is particularly simple. We will have a type $0$ and an
-isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-allows us to create a value and a corresponding debt out of nothing. Both the
-value and its negative counterpart can flow anywhere in the system. Because
-information is preserved, a closed program (which does not have any
-``dangling'' negative values) will eventually have to match the negative
-value with some corresponding value, effectively using the isomorphism in the
-right-to-left direction. In more detail, we can model the credit card
-transaction above using the following program (shown diagrammatically):
-
-FIGURE
-
-CLEAN UP the following based on the figure. In contrast, in our setting, one
-can start from the empty type $0$, introduce a positive value and its
-negative counterpart, and let each of these flow in arbitrary ways. The
-entire framework guarantees that neither the value nor its negative
-counterpart will be deleted or duplicated and hence that, in any closed
-program, the ``debt'' corresponding to the negative value is paid off exactly
-once. Computationally we model the first transaction using essentially the
-identity function which receives a \$20.00 bill as input from the buyer and
-propagates it on its output to the seller. The second transaction is more
-complicated. We model it as shown in the circuit below: There are two ways to
-understand this circuit that are both quite instructive. Let's first examine
-the type structure of each of the combinators that comprise the circuit. The
-first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
-produced from nothing so to speak by generating an equivalent debt that
-travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
-the execution. It goes forward in time so to speak, comes back, and then goes
-forwards again.
-
-It is critical that the framework in which the negative types are introduced
-is a framework in which all information is preserved, with no duplication or
-erasure. This guarantees that the generated debts are paid once and exactly
-once.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section{Fractional Types}
-
-The type $a/b$ is prominent in the Lambek-Grishin calculus which is
-extensively used in computational linguistics. In the common interpretation,
-a value of type $a/b$ is a value of some type $c$ such that when put in a
-context of type $b$, the result is a value of type $a$. For example, assuming
-contexts are represented as functions, a value of type $\texttt{bool} /
-(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
-$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
-in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
-type $\texttt{bool}$.
-
-In our case, the situation is simpler. If the goal is to produce a value of
-type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
-have type {{(a * b) / b}}. In most settings, this type would be equivalent to
-{{a}}. However in our setting, the constraint that this value must eventually
-fit in a larger value that supplies the {{b}} is explicitly recorded and must
-be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
-which when used in the left-to-right direction allows the creation of a value
-and a corresponding constraint on the context.  Both the value and the
-constraint can flow in arbitrary ways during the computation but eventually
-in a closed program with no ``dangling'' constraints, the isomorphism should
-be used in the right-to-left direction to resolve the constraints.  To
-understand the computational interpretation, consider the following example.
-
-The goal is to produce a value of type $(\texttt{bool} \times
-\texttt{int})$. One part of the program, $e_1$, knows how to produce the
-value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
-to produce the value of type $\texttt{bool}$ (say true). Computationally, we
-model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
-uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
-is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
-$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
-$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
-the two values meet, we can group the components as follows:
-\[
-(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
-\]
-Using the isomorphism in the right-to-left direction on the last two pairs,
-forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
-{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
-with the pair $(true,3)$ as desired.
-
-To summarize, one can think of a value of type $a/b$ as a value that imposes
-a constraint on its context of use: it is a value that requires its context
-to be of type $b$ and only if that condition is satisfied, can the value be
-considered as a value of type $a$. In the simplest case, the type $1/b$ just
-constraints the context to be of type $b$. By allowing the fractional types
-to be first-class, we can separate the generation of constraints from their
-use. Both the value and the constraint can flow in arbitrary ways during the
-computation but eventually in a closed program with no ``dangling''
-constraints, the isomorphism should be used in the right-to-left direction to
-resolve the constraints.
-
-NEED to work out the example in detail (or perhaps another example that's
-better.
-
-Again it is critical that the framework in which the fractional types are
-introduced is a framework in which all information is preserved, with no
-duplication or erasure. This guarantees that the generated constraints are
-satisfied once and exactly once.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary}
-
-To summarize negative, fractional, square root, and imaginary types all make
-sense. What they help you accomplish as a programmer is to disassociate
-global invariants into local ones that can be satisfied independently by
-subcomputations with no synchronization or communication. A computation
-producing something of type $a/b$ does not need to concern itself with who is
-going to supply the missing $b$: it just does its part. Conversely faced with
-a complicated task, a computation might decide to break it into pieces and
-demand these pieces using negative types.
-
-It is no surprise that these types are closely related to quantum mechanics
-and that they give us the feel that this is how nature computes. This is
-speculation however.
-
-In any case, in a framework where information can be copied and deleted, none
-of this makes much sense. It is critical that these constraints and demands
-can neither be duplicated nor erased.  This gives us the maximum
-``parallelism'' possible.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ {{langRev}} }
-
-The presentation here differs from \cite{infeffects} in the sense that
-we have introduced the {{0}} type here, thereby completing the monoid
-{{(0,+)}} and we present a small-step operational semantics.
-
-%subcode{bnf} include main
-% value types, b ::= 0 | 1 | b + b | b * b 
-% values, v ::= () | left v | right v | (v, v)
-
-%subcode{bnf} include main
-%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
-%zeroe :&  0 + b <-> b &: zeroi
-%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
-%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
-%identl* :&  1 * b <-> b &: identr*
-%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
-%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
-%dist0 :& 0 * b <-> 0 &: factor0
-%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-
-%subcode{proof} include main
-%@  ~
-%@@ id : b <-> b 
-%
-%@ c : b1 <-> b2
-%@@ sym c : b2 <-> b1
-%
-%@ c1 : b1 <-> b2
-%@ c2 : b2 <-> b3
-%@@ c1(;)c2 : b1 <-> b3
-%---
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
-%
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
-
-\begin{definition}{(Syntax of {{langRev}})}
-\label{def:langRev}
-We collect our types, values, and combinators, to get the full language
-definition.
-%subcode{bnf} include main
-% value types, b ::= 1 | b+b | b*b 
-% values, v ::= () | left v | right v | (v,v) 
-%
-% comb.~types, t ::= b <-> b
-% iso ::= swap+ | assocl+ | assocr+ 
-%     &|& identl* | identr* 
-%     &|& swap* | assocl* | assocr* 
-%     &|& dist | factor 
-% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-\end{definition}
-
-%%%%%%%%%%%%%%%%%%%
-\subsection{Semantics}
-
-%subcode{bnf} include main
-% Combinators, c = iso | c (;) c | c (*) c | c (+) c
-% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
-%                  &|& LeftTimes cc c v | RightTimes c v cc 
-%                  &|& LeftPlus cc c | RightPlus c cc 
-% Values, v = () | (v, v) | L v | R v
-%
-% Machine states = <c, v, cc> | {[c, v, cc]}
-% Start state = <c, v, []> 
-% Stop State = {[c, v, []]}
-
-\begin{scriptsize}
-%subcode{opsem} include main
-%! columnStyle = rlcl
-% swap+ & (left v) &|-->& right v
-% swap+ & (right v) &|-->& left v 
-% assocl+ & (left v1) &|-->& left (left v1)
-% assocl+ & (right (left v2)) &|-->& left (right v2)
-% assocl+ & (right (right v3)) &|-->& right v3 
-% assocr+ & (left (left v1)) &|-->& left v1
-% assocr+ & (left (right v2)) &|-->& right (left v2)
-% assocr+ & (right v3) &|-->& right (right v3)
-% identl* & ((), v) &|-->& v 
-% identr* & v &|-->& ((), v) 
-% swap* & (v1, v2) &|-->& (v2, v1) 
-% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
-% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
-% dist & (left v1, v3) &|-->& left (v1, v3)
-% dist & (right v2, v3) &|-->& right (v2, v3)
-% factor & (left (v1, v3)) &|-->& (left v1, v3) 
-% factor & (right (v2, v3)) &|-->& (right v2, v3)   
-\end{scriptsize}
-
-%subcode{opsem} include main
-%! columnStyle = rclr
-% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
-% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
-% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
-% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
-% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
-% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
-% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
-% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
-% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
-% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
-% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrev}
-{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
-\end{proposition}
-
-\begin{proposition}[Groupoid]
-\label{prop:groupoid}
-{{langRev}} is a groupoid. 
-\end{proposition}
-
-\begin{proposition}
-\label{prop:category}
-{{langRev}} is a dagger symmetric monoidal category. 
-\end{proposition}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{A Graphical Language for {{langRev}} }
-
-Combinators of {{langRev}} can be written using the the operators
-described in definition \ref{def:langRev} or via a graphical language
-similar in spirit to those developed for GoI
-\cite{DBLP:conf/popl/Mackie95} and for category theory
-\cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
-conventions and shorthands we will describe here, the wiring diagrams
-are a complete and equivalent to the operator based (syntactic)
-description of programs.  {{langRev}} combinators expressed in this
-graphical language looks like a ``wiring diagrams''. Value take the
-form of ``particles'' that flow along the wires. Computation is
-expressed by the flow of particles.
-
-\begin{itemize}
-\item 
-The simplest sort of diagram is the {{id : b <-> b}} combinator which
-is simply represented as a wire labeled by its type {{b}}. In more
-complex diagrams, if the type of a wire is obvious from the context,
-it may be omitted.
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
-\includegraphics{diagrams/thesis/b-wire.pdf}
-}
-\end{center}
-
-\item
-The product type {{b1*b2}} may be represented both as one wire
-labeled {{b1*b2}} or by two parallel wires labeled {{b1}} and
-{{b2}}. Both representations maybe used interchangeably.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
-\includegraphics{diagrams/thesis/product-one-wire.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
-\includegraphics{diagrams/thesis/product-two-wires.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item
-Sum types may similarly be represented using using parallel wires with
-a {{+}} operator between them.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
-\includegraphics{diagrams/thesis/sum-one-wire.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
-\includegraphics{diagrams/thesis/sum-two-wires.pdf}
-}
-\end{center}
-\end{multicols}
-
-%% \item
-%% When representing complex types like {{(b1*b2)+b3}} some visual
-%% grouping of the wires may be done to aid readability. The exact type
-%% however will always be clarified by the context of the diagram.
-
-%% \begin{center}
-%% \scalebox{0.95}{
-%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
-%% }
-%% \end{center}
-
-\item
-Associativity is entirely skipped in the graphical language. Hence
-three parallel wires maybe inferred as {{b1*(b2*b3)}} or
-{{(b1*b2)*b3}}, based on the context. This is much like handling of
-associativity in the graphical representations of categories
-\cite{springerlink:10.1007/978-3-642-12821-94},
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
-\includegraphics{diagrams/thesis/assoc.pdf}
-}
-\end{center}
-
-\item
-Commutativity is represented by crisscrossing wires.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
-\includegraphics{diagrams/thesis/swap_times.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
-\includegraphics{diagrams/thesis/swap_plus.pdf}
-}
-\end{center}
-
-\end{multicols}
-
-\item
-Identity morphisms and their duals are represented as follows:
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
-\includegraphics{diagrams/thesis/uniti.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
-\includegraphics{diagrams/thesis/unite.pdf}
-}
-\end{center}  
-\end{multicols}
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
-\includegraphics{diagrams/thesis/zeroi.pdf}
-}
-\end{center}
-
-\columnbreak
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
-\includegraphics{diagrams/thesis/zeroe.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item
-Finally, distributivity and factoring are represented using the dual
-boxes shown below:
-
-\begin{multicols}{2}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/distrib-crop.pdf]
-\includegraphics{diagrams/thesis/dist.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/factor-crop.pdf]
-\includegraphics{diagrams/thesis/factor.pdf}
-}
-\end{center}
-\end{multicols}
-
-\end{itemize}
-
-\noindent 
-\textit{Example.} As an example consider the wiring diagram
-of the combinator {{c}} below:
-
-{{c : b * (1+1) <-> b + b}}
-
-{{c = swap* (;) dist (;) (identl* (+) identl*)}}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
-\includegraphics{diagrams/thesis/example1.pdf}
-}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ {{langRevEE}} }
-
-\paragraph*{Observability.} 
-Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
-input {{v1 : b1}} gives us a value {{v2 : b2}} on
-termination. Execution is well defined only if {{b1}} and {{b2}} are
-entirely positive types. Consider the program that
-
-(TODO: So this is a really bad explanation. But this section is really
-important and we need to explain carefully. Maybe an analogy can be
-drawn to Cayley diagrams for groups to point out that the notion of
-`i' is entirely artificial, but it is essential to the discourse.)
-
-\begin{center}
-  \includegraphics{diagrams/shuffle.pdf}
-\end{center}
-
-We define observables to be only positive types. The outputs of
-programs that output mixed positive and negative types are not
-observable.  Also, programs that input mixed positive and negatives
-types are not executable.
-
-%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Additive Duality in {{langRev}} }
-
-Additive duality is added to {{langRev}} by addition of negative types
-denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
-extend the language as follows:
-
-%subcode{bnf} include main
-% Value Types, b = ... | -b
-% Values, v = ... | -v
-%
-% Isomorphisms, iso &=& ... | eta+ | eps+
-
-\noindent
-with the types judgments
-
-%subcode{opsem} include main
-% eta+ &: 0 <-> (-b) + b :& eps+
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-
-We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
-the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
-dually on the right is {{eps+}} showing the map from {{-b+b}} to
-0. Even though the diagrams below show a {{0}} wire for completeness,
-later diagrams will always drop them in contexts where we can
-implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/eta.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/eps.pdf}
-\end{center}
-\end{multicols}
-
-The usual interpretation of the type {{b1+b2}} is that we either have
-an appropriately tagged value of type {{b1}} or an appropriately
-tagged value of type {{b2}}. This interpretation is maintained in the
-presence of {{eta+}} and {{eps+}} in the following sense: a value of
-type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
-switches into a value {{L (-v): -b + b}} that flows backwards on the
-upper wire. The inversion of direction is captured by the negative
-sign on the value.
-
-\paragraph*{Negative Information Flow.} 
-Hence, the intuitive way to read the diagrams is that they reverse the
-flow of information -- in normal {{langRev}} circuits information
-flows from left-to-right and {{eps+}} operations cause information to
-flow from right-to-left.  The interpretation of {{eta+}} is precisely
-the opposite: it turns a right-to-left flow of information in a
-circuit into a left-to-right flow.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus1.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus2.pdf}
-}
-\end{center}  
-\end{multicols}
-
-\paragraph*{Time Traveling.}
-A time traveling intuition is also applicable here: if view the
-left-to-right direction of information flow as the canonical forward
-direction of information, then the action of {{eps+}} allows values,
-aka information particles, to flow backwards in time. Particles that
-flow backward in time get to see and interact with the history of
-particles that they coexist with i.e. values that they exist
-multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
-(whose construction is shown in Section \ref{sec:neg-constructions}).
-
-\paragraph*{Coherence.}
-The appropriate coherence condition for Compact Closed Categories,
-expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-may be visualized as explained by Selinger
-\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-\paragraph*{Non-termination.}
-As is well known in category theory, Compact Closed Categories contain
-trace operations.  In the context of {{langRev}} this implies that
-{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
-(detailed in Section \ref{sec:monoidal-constructions}) hence
-introducing \emph{non-termination} into the language
-\cite{infeffects}.
-
-\paragraph*{Zero Information.}
-The fact that {{eta+}} (and hence {{eps+}}) introduces no information
-can be understood is by viewing {{-b+b}} as a type whose value is
-observable only if you `pay' {{b}} amount of information to the {{-b}}
-branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
-amount of information is returned to you. (TODO. more needs to be said
-here.)
-
-ADD MATH.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multiplicative Duality in {{langRev}} }
-
-Multiplicative duality is added to {{langRev}} by adding types
-{{b^^^}}, values {{v^^^}} and isomorphisms {{eta*}} and {{eps*}}.
-Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
-{{a/b}}.
-
-%subcode{bnf} include main
-% Value Types, b = ... | b^^^
-% Values, v = ... | v^^^
-%
-% Isomorphisms, iso &=& ... | eta* | eps*
-
-\noindent
-These have the type judgments
-
-%subcode{opsem} include main
-% eta* &: 1 <-> b^^^ * b :& eps*
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- v^^^ : b^^^
-
-As with additive duals, we represent {{eta*}} and {{eps*}] using
-`U'-shaped connectors. In the general case the wires for {{1}} will be
-dropped as it can freely introduced or eliminated using {{uniti}} and
-{{unite}}.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/eta_times.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/eps_times.pdf}
-\end{center}
-\end{multicols}
-
-The usual interpretation of {{b1*b2}} we have both a value of type
-{{b1}} and a value of type {{b2}}. This interpretation is maintained
-in the precense of fractionals. Hence an {{eta* : 1 <-> b^^^ * b}} is
-to be viewed as a fission point for a value of type {{b}} and its
-multiplicative inverse {{b^^^}}. This naturally raises the question --
-which value of type {{b}} is produced? The isomorphism {{eta*}} does
-not produce any specific value of type {{b}}, instead it produces a
-place holder for a value such that the value of type {{b^^^}} is
-always the dual of the chosen one.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eta_times1.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_times1.pdf}
-}
-\end{center}  
-\end{multicols}
-
-\paragraph*{Unification.}
-A computational interpretation of this is that {{langRevEE}} acts like
-a logic programming system and {{eta*}} acts as the site of creation
-of a new typed logic variable of type {{b}} and its corresponding dual
-of type {{b^^^}}. These two values are free to move independantly
-through a circuit, but any operations on value affect what operations
-are possible on the other since they are both related by an underlying
-logic variable. Dually, {{eps*}} acts as a unification site for values
-flowing along both branches. If the values are not exactly duals, then
-the system fails to unify and hence the program is stuck. If the
-values are exactly duals, then unification succeeds and {{eta*}}
-returns {{1}}.
-
-\paragraph*{Entanglement and Anti-Particles.}
-A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
-quantum mechanics. The {{eta*}} is a site of fission that creates an
-information particle and its anti-particle. Both flow in the same
-direction in time. The produce particles are entangles in that actions
-on one (such as transformation by application of an isomorphism)
-affects the other. Further the fission site does not create a particle
-in any specific state, but in a superposition of all possible states
-as dictated by its type. In each possible world that the particle
-exists, it takes on a specific value inhibiting its type and
-correspondingly its entangled anti-particle takes the dual of the
-specific value.
-
-This also leads to the exiciting point of view that entangled
-particle/anti-particle pairs act as functions in a physical
-first-class sense. We can ask the fuction to be applied to a value by
-unifying the anti-particle with a particle in a well-known state and
-examining is pair. We shall use this point of view in the development
-our SAT solver (see Sec. \ref{sec:sat-solver}).
-
-\paragraph*{Coherence.} 
-As before, the requisite coherence condition on Compact Closed
-Categories is applicable and is expressed below in terms of
-{{eta*}}/{{eps*}} and the multiplicative monoid.
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-\paragraph*{Annihilation.}
-As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
-additive {{trace+}}, {{trace*}} gives us the ability to find the
-fixpoint of the contraint (expressed with combinators) that is
-traced. If the constraint has not satisfying values, there is not
-fixpoint possible -- we use the term `annihilation' to describe the
-corresponding undefined state of the program. Annihilation and a
-non-termination both describe undefined program execution -- while
-non-termination characterizes an iteration that fails to terminate,
-annihilation characterizes a constraint that has no fixpoint.
-
-\paragraph*{Zero Information.}
-As with {{eta+}}, we argue that {{eta*}} constructs no new
-information, instead it gives us the ability to explore as many worlds
-as permitted by the type {{b}}. So situate ourselves in any specific
-world, we have to supply additional information in teh form of
-unifying the particle or anti-particle with a particle representing
-known information. Hence only by supplying information to a an
-entangled pair can we read information from it -- and always exactly
-as much information as we supplied.
-
-ADD MATH.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semantics}
-
-%%%%%%%%%%%%%%%%%%
-\subsection{Syntax}
-
-%subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
-%
-% Sequential Contexts, P, F = [] | P:c
-% Parallel Contexts, D = [] 
-%                     &|& D:(P| [](+)c| F)
-%                     &|& D:(P| c(+)[]| F)
-%                     &|& D:(P| [] (*) c v| F)
-%                     &|& D:(P| [] (*) v c| F)
-%                     &|& D:(P| c v (*)  []| F)
-%                     &|& D:(P| v c (*) []| F)
-%
-%
-% Machine States  = D[P| c v| F] | D[P| v c| F]
-% Start State  = [][[]| v c| []]
-% Stop State  = [][P| c v| []]
-
-Combinator reconstruction, {{P[c]}}:
-%subcode{opsem} include main
-% [](c) '= c
-% P:c'(c) '= P(c'(;)c)
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection{Operational Semantics}
-
-The small step semantics present for {{langRev}} below work
-symmetrically for forward and backward evaluation.
-
-\begin{itemize}
-\item 
-Basic reduction of an isomorphism. Note that the evaluation leaves the
-adjoint of the combinator behind. This will become important when we
-reverse the direction of computation.
-%% %subcode{opsem} include main
-%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
-
-%subcode{opsem} include main
-% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
-
-\item
-Sequencing involves pushing and popping from the Future and Past
-continuations:
-%subcode{opsem} include main
-% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
-% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
-
-\item
-Parallel composition, captures the current Future and Past and extends
-the parallel context.
-%subcode{opsem} include main
-% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
-% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
-% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
-% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
-
-\item
-Similarly for products:
-%subcode{opsem} include main
-% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
-% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
-% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
-
-and symmetrical rules for evaluation along the second branch. 
-%subcode{opsem} include main
-% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
-% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
-
-The later two rules will be relevant only for reverse execution.
-
-\end{itemize}
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrevEE}
-{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
-\end{proposition}
-
-\begin{proposition}[Groupoid]
-\label{prop:groupoidEE}
-{{langRevEE}} is a groupoid. 
-\end{proposition}
-
-\begin{proposition}
-\label{prop:categoryEE}
-{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
-\end{proposition}
-
-Note: We use `Monoidal Closed' in the sense of Compact Closed. While
-the category has an multiplicative tensor that distributes over an
-additive tensor, it does not have injections or projections and hence
-does not have categorical sums and products.
-
-%%%%%%%%%%%%%%%%%%%%
-\subsection{Rules for {{eta}} and {{eps}} }
-
-The operation {{eps}} reverses the direction of a particle by
-reversing the world.
-
-Note that we deviate from the categorical definition of {{eta}} and
-{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
-choice of {{eta}}. This however does not affect us, because we deal
-with a symmetric category.
-
-\begin{itemize}
-\item Grammar
-%subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v | -v
-% Isomorphisms, iso &=& ... | eta | eps
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
-
-\item
-Type judgment.
-%subcode{opsem} include main
-% eta &: 0 <-> (-b) + b :& eps
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-
-\item
-Operational Semantics.
-%subcode{opsem} include main
-% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
-% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
-
-Note: there is NO reduction rule for {{eta}}.
-
-\item
-The adjoint of a parallel context is defined to be:
-%subcode{opsem} include main
-% []{dagger} '= []
-% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
-% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
-% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
-% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
-% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
-% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions in {{langRevEE}} }
-\label{sec:monoidal-constructions}
-
-These constructions are applicable to the additive and the
-multiplicative monoid.
-
-\paragraph*{Functions}
-
-\begin{center}
-  \includegraphics{diagrams/function.pdf}
-\end{center}
-
-Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
-
-\paragraph*{Function application}
-
-\begin{center}
-  \includegraphics{diagrams/apply1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/apply2.pdf}
-\end{center}
-
-\paragraph*{Function composition}
-
-\begin{center}
-  \includegraphics{diagrams/compose1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/compose.pdf}
-\end{center}
-
-This is also equivalent to sequencing both the computation blocks.
-
-\begin{center}
-  \includegraphics{diagrams/compose2.pdf}
-\end{center}
-
-\paragraph*{Delimited continuation}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}
-
-\paragraph*{Trace }
-
-%subcode{proof} include main
-%@ c : b2 + b1 <-> b2 + b3
-%@@ trace c : b1 <-> b3
-
-\begin{center}
-  \includegraphics{diagrams/trace.pdf}
-\end{center}
-
-\paragraph*{Principium Contradictiones}
-
-{{b <-> -(-b)}}
-
-\begin{center}
-  \includegraphics{diagrams/double_neg.pdf}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions Specific to Negation}
-\label{sec:neg-constructions}
-
-\paragraph*{Negation distributes over {{+}}. }
-
-{{-(b1 + b2) <-> (b1) + (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/dist_neg_plus.pdf}
-\end{center}
-
-\paragraph*{ {{eps_{fst} }} }
-
-{{(-b1)*b2 + b1*b2 <-> 0}}
-
-\begin{center}
-  \includegraphics{diagrams/eps_fst.pdf}
-\end{center}
-
-\paragraph*{Lifting negation out of {{*}}. }
-
-{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/mult_neg.pdf}
-\end{center}
-
-The following isomorphism can be constructed similarly
-
-{{b1 * b2 <-> (-b1)*(-b2)}}
-
-\paragraph*{Lifting a operation of positive types to negated types}
-
-Given {{c : b1 <-> b2}}
-
-\begin{center}
-  \includegraphics{diagrams/neg_lift.pdf}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions Specific to Multiplication}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{To think about}
-
-\begin{itemize}
-
-\item If we built an effect over {{langRevEE}}, say {{create}} and
-  {{erase}}, are effects structured by an arrow or a monad now?
-
-\item Which operations can be lifted to work on negative types?
-
-\item Can the operational semantics for {{langRevEE}} be an
-  interpreter that is implemented in {{langRevT}} (similar to how the
-  the tree traversal interpreter was implemented). This would imply
-  the existence of a more powerful construction than Int, wherein the
-  products would also be preserved. This is possibly worth a paper in
-  itself.
-
-\item These functions aren't really values. There is no value one can
-  produce that denotes a function. These functions are the ability to
-  transform a value - the possibility of transforming a value.  In the
-  product encoding of environments, there is no value one can assign
-  to a variable such that it denotes a function.
-
-Actually this is possible. Consider two functions {{f : b1 <-> b2}}
-and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
-discriminate them. Hence the {{boo}} is the first class representation
-of the functions and can be thought of as the address of the function.
-
-\begin{center}
-  \includegraphics{diagrams/dispatch.pdf}
-\end{center}
-
-\item It is not fair to say that negative types flow backwards. The
-  following circuits are valid in {{langRevEE}}. It is however proper
-  to say that for any type {{b}} that flows in a direction, the type
-  {{-b}} flows in the reverse direction.
-
-\begin{center}
-  \includegraphics{diagrams/neg_circuit1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/neg_circuit2.pdf}
-\end{center}
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Advanced Programming Examples}
-
-%%%%%%%%%%%%%%%%%%%
-\subsection{SAT solver}
-\label{sec:sat-solver}
-
-The construction presented here is a novel SAT-solver that relies on
-{{trace*}} for its solution. It varies from previous classical SAT
-solvers in that there is not explicit search operation on the boolean
-space. It also varies from previous quantum SAT solvers in that it
-does uses a {{trace}} operation to compute a fixpoint which is the
-solution following which an isomorphic clone operation allows us to
-examine the result.
-
-The construction here is not limited to boolean satisfiability, but
-generalises to any constraint satisfaction problem whose search space
-can be represented by a recursive type and for which an isomorphic
-clone operation can be constructed. That said, we will focus on the
-limited case of boolean satisfiability here.
-
-Consider the circuit below that has no observable output -- i.e. for
-all inputs it is annihilating. This happens because it constructs a
-boolean and its dual, negates one of them and attempts to collapse
-them.
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/not_trace.pdf}
-}
-\end{center}
-
-We leverage the basic idea of the above circuit, to collapse all
-values that are not solutions to the specified isomorphic constraint
-{{f}}. The operation {{f}} acts on some inputs and its adjoint
-reconstructs the inputs from the outputs. While the output of {{f}}
-may contain many components, its must contain a boolean indicating if
-the input satisfy {{f}}. The special output wire is used to `control'
-the negation of a boolean `control wire'. If the inputs do not satisfy
-{{f}}, the control wire is negated.
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/sat1.pdf}
-}
-\end{center}
-
-If this construction is traced using {{trace*}}, all the values that
-do not satisfy {{f}} get annihilated because the control wire acts
-like the closed-loop `not' in the previous construction.
-
-%%%%%%%%%%%%%%%%%%%
-\subsection{GoI machine}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Related Work} 
-\label{sec:related}
-
-(NOTE: Lets use this section here to establish a long of ideas and
-intuition. None of these connections have to be formal, but they can
-be concrete.)
-
-Filinski proposes that continuations are a \emph{declarative} concept. He,
-furthermore, introduces a symmetric extension of the $\lambda$-calculus in
-which values and continuations are treated as opposites. This is essentially
-what we are proposing with one fundamental difference: our underlying
-language is not the $\lambda$-calculus but a language of pure isomorphisms in
-which information is preserved. This shift of perspective enables us to
-distill and generalize the duality of values and continuations: in
-particular, in the conventional $\lambda$-calculus setting values and
-continuations can be erased and duplicated which makes it difficult to
-maintain the correspondence between a value and its negative counterpart.
-
-\paragraph*{Continuations.} The idea of using negative types to model
-information flowing backwards, demand for values, continuations, etc. goes
-back to at Filinski's thesis. We recall these connections below but we first
-note that all these systems are complicated because in all these systems
-information can be ignored, destroyed, or duplicated. Clearly the possibility
-of erasure of information would mean that our credit card transaction is
-incorrect. In our work, information is maintained and hence we have a
-guarantee that, in a closed program, the debt must be accounted and paid for.
-
-\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
-In his Masters thesis, Filinski proposes that continuations are a
-\emph{declarative} concept. He, furthermore, introduces a symmetric extension
-of the $\lambda$-calculus in which values and continuations are treated as
-opposites. This is essentially what we are proposing with one fundamental
-difference: our underlying language is not the $\lambda$-calculus but a
-language of pure isomorphisms in which information is preserved. This shift
-of perspective enables us to distill and generalize the duality of values and
-continuations: in particular, in the conventional $\lambda$-calculus setting
-values and continuations can be erased and duplicated which makes it
-difficult to maintain the correspondence between a value and its negative
-counterpart. In contrast, in our setting, one can start from the empty type
-$0$, introduce a positive value and its negative counterpart, and let each of
-these flow in arbitrary ways. The entire framework guarantees that neither
-the value nor its negative counterpart will be deleted or duplicated and
-hence that, in any closed program, the ``debt'' corresponding to the negative
-value is paid off exactly once. The forward and backward executions in our
-framework correspond to call-by-value and call-by-name. This duality was
-observed by Filinski and others following him but it is particularly clean in
-our framework.
-
-\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
-subtraction type because its ``computational interpretation is not
-familiar.'' Curien and Herbelin study duality in classical logic, show that
-it exchanges call-by-value and call-by-name. They extend classical natural
-deduction with subtraction but give it no computational meaning. Crolard (in
-the formulae-as-types interpretation of subtractive logic) address the
-computational interpretation of subtraction. He explains the type $A-B$ as
-the of \emph{coroutines} with a local environment of type $A$ and a
-continuation of type $B$. The description is complicated by what is
-essentially the desire to enforce linearity constraints so that coroutines
-cannot access the local environment of other coroutines. Must cite Selinger
-control categories in this context of duality but I am not sure what to say:
-it assumes cartesian closed categories for one thing and not symmetric
-monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
-explain delimited continuations.
-
-\paragraph*{Linear Logic} In accounts that are linear, the value and
-continuation that comprise the subtractive type need to be constrained to
-``stay together.'' This can be achieved by various restrictions. In this work
-we have no such constraints, the negative value can flow anywhere. The entire
-system guarantees that any closed program would have to account for it. We
-don't have to introduce special constraints to achieve that. Zeilberger in
-the paper on polarity and the logic of delimited continuations uses polarized
-logic: he shows that if positive and negative values are completely symmetric
-except that answer types are positive, then the framework accommodates
-delimited continuations. But he interprets negative values are control
-operators, or as values defined by the shape of their continuations. We
-simply interpret values of negative type as values flowing in the ``other''
-direction.
-
-\paragraph*{Int Construction.}
-For a traced monoidal category {{C}} the Int construction produces a
-Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
-Further we know that the target of the Int construction is isomorphic
-to the target of \G construction of Abramsky \cite{Abramsky96:0} from
-Haghverdi. However, note that the {{langRevEE}} is not the same as the
-image of the Int construction on {{langRevT}}, since the the later
-lacks a multiplicative tensor that distributes over the additive tensor
-in Int {{langRevT}}.
-
-\paragraph*{Recursion.} In our previous work we introduce recursive types and
-trace operators. This is dangerous here because infinite loops allow us to
-prolong paying the debt for as long as we want.
-
-\paragraph*{GoI machines} 
-We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Field of Algebraic Numbers}
-
-The field of algebraic numbers consist of the rationals extended with
-solutions to polynomials. In a limited sense {{langRevEE}} with
-recursive types already include polynomials such as the definition of
-binary tree {{@@x.(1+x*x)}} implies the equality {{t = 1 + t*t}} and
-hence {{t^2-t+1=0}}. While this equation does indeed have an
-irrational imaginary root, as we will see -- {{langRevEE}} does not
-have the means of expressing meaningful computation that uses the
-root.
-
-Here is one way to view computations involving irrationals that might
-shed some insight into how we should view recursive data types
-also. For finite datatypes say of size {{n}} we can say that an
-element specifies a choice of {{1/n}} and hence contains as much
-information. For finite things, each value is computationally equal to
-any other value. Any permutation is possible among the values. Hence
-there is no metric possible on the values.  Recursive data types, such
-as binary trees, have an infinite number of elements and hence the
-argument that says that value specifies {{1/n}} units of information
-is not longer valid. Instead to talk meaningfully about information
-content, we should talk about a probability distribution on the values
-of the type. However what shall we base this distribution on? Is there
-a natural order beyond the unfolding order of the values of a type?
-
-In this context, let us go back to the thought that recursive such as
-{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
-interesting solution of trees is precisely that they involve
-irrational numbers. Our usual notation for irrationals such as
-{{2^{1/2} }} can actually be viewed as a thunk representing the
-infinite computation {{1.4142..}}. Irrationals in this sense have a
-very different information theoretic interpretation from full real
-numbers. With Chaitin's Omega numbers he postulates the existence of
-real numbers that have an infinite amount of information -- ones for
-example are the solution to the question about the halting properties
-of every Turning machine.
-
-Real numbers in that sense really are a series of bits of unknown
-information theoretic interpretation -- a sort of {{top}}. Numbers
-such as {{2^{1/2} }} are very different in the sense that even though
-they contain an infinite sequence of bits, there is a finite program
-that generates those bits and its only the cost of computing bits that
-is infinite.  The consistency of arithmetic isn't affected by
-extending the rational field by adding any specific computable
-irrational. So it might be possible to design a model of computing
-where we can indeed deal with them. Such a model will have to explicit
-thunks for the delayed computation represented by trees and never
-equate and cancel two infinite computations unless the represent
-exactly the same infinite computation.
-
-TODO.
-
-Square root and imaginary types have also appeared in the literature: we
-relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
-explanation. We have so far extended the set of types to be the rational
-numbers. Now we will push this and extend the set of types to algebraic
-numbers. In other words, we will allow datatypes defined by arbitrary
-polynomials and allow the roots of such polynomials to be types.
-
-Consider first an example in which we want to compute with the sides of a
-rectangle whose area is 91 and whose length is longer than its width by 6
-units. One can solve the quadratic equation to determine that the sides are 7
-and 13 and proceed. This however prematurely forces us to globally solve the
-constraint. Instead we can let the two sides of the rectangle be $x$ and
-$x+6$ and use the following equation to capture the desired constraint:
-\[
-x^2 + 6x - 91 = 0
-\]
-The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
-the type $0$. We can now proceed to compute with the unknown $x$, being
-assured that in a closed program, our computation will eventually be
-consistent with the solution of the algebraic equation.
-
-Previously the most famous example of a similar nature is the puzzling
-isomorphism that one can establish between seven binary trees and one.
-A binary tree is defined by the datatype
-\[
-x = 1 + x * x 
-\]
-which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
-algebraic manipulation we can reason as follows:
-\[\begin{array}{rclclclcl}
-x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
-x^6 &=& 1 \\
-x^7 &=& x^6 x &=& x
-\end{array}\]
-Fiore poses the question of why such algebraic manipulation would make sense
-type theoretically but states that even though some of the intermediate steps
-make no sense, the final equivalence is valid and can be used to actually
-construct an isomorphism between $x^7$ the type of seven binary trees 
-and $x$ the type of binary trees.
-
-Discussion of possible polynomials:
-\begin{itemize}
-\item $b=1+1$. Boring.
-\item $b=b+1$. That introduces the natural numbers. No solution for this
-  polynomial over the algebraic numbers. We must extend the numbers with
-  $\omega$ to get a solution. We reject this in this paper and prefer to
-  stick to algebraic numbers. The advantage is that all the isomorphisms are
-  valid numerically. With the above type we could subtract $b$ from both
-  sides to show that $0=1$ which is nonsense. We can however have infinite
-  types as long as they have algebraic solutions.
-\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
-  boolean! If we had superpositions, we could then write a function that
-  performs the square root of negation in the sense that applying it twice
-  would be equivalent to boolean negation.
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
 %% Jacques

### 2012-02-29 rational.tex
changed to fractions

@@
 %! theseus = Theseus
+%! sqrt(x) = \sqrt{#x}
+%! surd(p,x) = \sqrt[#p]{#x}
+%! inv(x) = \frac{1}{#x}
+%! frac(x,y) = \frac{#x}{#y}
 %! * = \times
@@
 %subcode{bnf} include main
-% Value Types, b = ... | b^^^
-% Values, v = ... | v^^^
+% Value Types, b = ... | inv(b)
+% Values, v = ... | inv(v)
 %
@@
 %subcode{opsem} include main
-% eta* &: 1 <-> b^^^ * b :& eps*
+% eta* &: 1 <-> inv(b) * b :& eps*
 
@@
 %@ |- v : b
-%@@ |- v^^^ : b^^^
-
-As with additive duals, we represent {{eta*}} and {{eps*}] using
+%@@ |- inv(v) : inv(b)
+
+As with additive duals, we represent {{eta*}} and {{eps*}} using
 `U'-shaped connectors. In the general case the wires for {{1}} will be
@@
 {{b1}} and a value of type {{b2}}. This interpretation is maintained
-in the precense of fractionals. Hence an {{eta* : 1 <-> b^^^ * b}} is
+in the precense of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
 to be viewed as a fission point for a value of type {{b}} and its
-multiplicative inverse {{b^^^}}. This naturally raises the question --
+multiplicative inverse {{inv(b)}}. This naturally raises the question --
 which value of type {{b}} is produced? The isomorphism {{eta*}} does
 not produce any specific value of type {{b}}, instead it produces a
-place holder for a value such that the value of type {{b^^^}} is
+place holder for a value such that the value of type {{inv(b)}} is
 always the dual of the chosen one.
@@
 of a new typed logic variable of type {{b}} and its corresponding dual
-of type {{b^^^}}. These two values are free to move independantly
+of type {{inv(b)}}. These two values are free to move independantly
 through a circuit, but any operations on value affect what operations
@@
 irrational numbers. Our usual notation for irrationals such as
-{{2^{1/2} }} can actually be viewed as a thunk representing the
+{{sqrt(2) }} can actually be viewed as a thunk representing the
 infinite computation {{1.4142..}}. Irrationals in this sense have a
@@
 information theoretic interpretation -- a sort of {{top}}. Numbers
-such as {{2^{1/2} }} are very different in the sense that even though
+such as {{sqrt(2) }} are very different in the sense that even though
 they contain an infinite sequence of bits, there is a finite program

### 2012-03-02 rational.tex
_

@@
 %%\usepackage{longtable}
-%%\usepackage{comment}
+\usepackage{comment}
 \usepackage{amsmath}
@@
 \item
-Associativity is entirely skipped in the graphical language. Hence
-three parallel wires maybe inferred as {{b1*(b2*b3)}} or
-{{(b1*b2)*b3}}, based on the context. This is much like handling of
-associativity in the graphical representations of categories
-\cite{springerlink:10.1007/978-3-642-12821-94},
+Associativity is entirely implicit in the graphical language. Three
+parallel wires maybe inferred as {{b1*(b2*b3)}} or {{(b1*b2)*b3}},
+based on the context, much like handling of associativity in the
+graphical representations of categories
+\cite{springerlink:10.1007/978-3-642-12821-94}.
 
@@
 \item
-Identity morphisms and their duals are represented as follows:
+Identity morphisms and their duals ({{uniti, unite, zeroi}} and
+{{zeroe}}) are represented as follows:
 
@@
 \end{multicols}
+
+Since the monoidal units cxan be freely introduced and eliminated, in
+many diagrams they are omitted and dealt with explicitely when they
+are of special interest.
 
@@
 
+\begin{comment}
 %%%%%%%%%%%%%%%%%%
@@
 \end{itemize}
+
+\end{comment}
 

### 2012-03-02 iso-int/rc-iso-int.tex
added abstract

@@
 \begin{abstract}
+Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
+properties and various constructions in it. However, for a
+computational model, there has been little focus on actually writing
+programs directly in {{langRevT}}. This paper aims to remedy that by
+presenting a systematic technique of expressing logically reversible
+small-step abstract machines in {{langRevT}}. In other words, this
+paper shows that once we have a logically reversible machine in a
+notation of our choice, expressing the machine as an isomorphic
+interpreter in {{langRevT}} can be done systematically and does not
+present any significant conceptual difficulties. This gives us a means
+of developing large {{langRevT}} programs with the ease of reasoning
+at the level of a conventional small-step semantics.  We develop
+several simple interpreters over numbers and addition, move onto tree
+traversal and finish with an interpreter for {{langRevT}} expressed in
+{{langRevT}}.
 \end{abstract}
@@
 previously. In particular, we have a two types of machine states --
-{{<t,c>}} and {{ {[t, c]} }}. One of these corresponds to walkking
+{{<t,c>}} and {{ {[t, c]} }}. One of these corresponds to walking
 down a tree, building up the context in the process. The other
@@
 machine as well.  As the first step, we quickly check that the
-abstract machine however has disstinguished start and stop states, has
+abstract machine however has distinguished start and stop states, has
 no ambiguities in reduction and that each reduction step is logically
@@
 Before we start on the actual construction, let us introduce a
-syntactic convenience. In the past interpreter we explicitely drew out
+syntactic convenience. In the past interpreter we explicitly drew out
 every {{fold}} and {{unfold}} followed/preceded immediately by any
@@
 
-The first step to developing teh isomorphic interpreter is to
+The first step to developing the isomorphic interpreter is to
 recognize that the two possible kinds of machine states simply hide an
@@
 abstract machine does. One thing to note is that we dropped the {{1}}
-introduced by expanding the {{bool}} and instead just labelled the
+introduced by expanding the {{bool}} and instead just labeled the
 {{true}} and {{false}} branches.
@@
 types to reason about all these things? Something to think about. If
-we do, then thereare two ways out
+we do, then there are two ways out
 
@@
 simple. They can always be added opaquely for any given types -- we
-only need to add the proper fold/unflod morphisms for those types
+only need to add the proper fold/unfold morphisms for those types
 (maybe).

### 2012-03-02 iso-int/rc-iso-int.tex
abstract

--- 2012-02-24 iso-int/rc-iso-int.tex
@@
 \begin{abstract}
+  In our previous work [RC'11, POPL'12], we developed a reversible
+  programming language and established that every computation in it is a
+  (partial) isomorphism that is reversible and that preserves
+  information. The language is founded on type isomorphisms that have a clear
+  categorical semantics but that are awkward as a notation for writing actual
+  programs, especially recursive programs. This paper remedies this aspect by
+  presenting a systematic technique for developing a large and expressive
+  class of reversible recursive programs, that of logically reversible
+  small-step abstract machines. In other words, this paper shows that once we
+  have a logically reversible machine in a notation of our choice, expressing
+  the machine as an isomorphic interpreter can be done systematically and
+  does not present any significant conceptual difficulties. Concretely, we
+  develop several simple interpreters over numbers and addition, move on to
+  tree traversals, and finish with a meta-circular interpreter for our
+  reversible language. Generally, this gives us a means of developing large
+  reversible programs with the ease of reasoning at the level of a
+  conventional small-step semantics.
 \end{abstract}

### 2012-03-05 rational.tex
intro a bit

@@
 
-In a recent paper~\cite{infeffects}, we argued that, by encompassing
+In a recent paper~\cite{infeffects}, we argued that, because they include
 irreversible physical primitives, conventional abstract models of computation
-have also inadvertently included some \emph{implicit} computational effects
-which we called \emph{information effects}. We then developed a pure
-reversible model of computation that is obtained from the type isomorphisms
-and categorical structures that underlie models of linear logic and quantum
+have inadvertently included some \emph{implicit} computational effects which
+we called \emph{information effects}. We then developed a pure reversible
+model of computation that is obtained from the type isomorphisms and
+categorical structures that underlie models of linear logic and quantum
 computing and that treats information as a linear resource that can neither
@@
 \$20.00 and the consumer as someone demanding \$20.00. It is certainly
-possible to directly connect the producer and consumer: such a connection
-would be modeled by the identity function. The proof above implements this
-connection in a much more interesting way, however. A transformation $0 =
-(-a) + a$ is used at the consumer site to create money and a corresponding
+possible to directly connect the producer and consumer using the algebraic
+identity $a=a$: such a connection would be computationally modeled by the
+identity function as expected. However, the algebraic proof above implements
+this connection in a much more interesting way: the transformation $0 = (-a)
++ a$ is used at the consumer site to create money and a corresponding
 debt. The consumer gets the money immediately and the debt travels
@@
   simpler and much more appealing in a framework where information is
-  guaranteed to be preserved. In previous work that used negative types,
-  complicated mechanisms are typically needed to constrain the propagation
-  and use of negative values because the surrounding computational framework
-  is, generally speaking, careless in its treatment of information.
+  guaranteed to be preserved. In previous work that used negative types (see
+  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
+  constrain the propagation and use of negative values because the
+  surrounding computational framework is, generally speaking, careless in its
+  treatment of information.
 \item The appeal of negative types can already be gleaned from the credit
   card analogy. The main reason credit card transactions are convenient is
-  because they disentangle the propagation of the money from the propagation
-  of the services. Not every transaction needs both the resources and
-  services to be brought together: it is sufficient to have a promise that
-  the money will be paid at some point, as long as the infrastructure can be
-  trusted that such promises will be fulfilled. 
+  because they disentangle the propagation of the resources (money) from the
+  propagation of the services. Not every transaction needs both the resources
+  and services to be brought together: it is sufficient to have a promise
+  that the money will be paid at some point, as long as the infrastructure
+  can be trusted that such promises will be fulfilled. Computationally, this
+  means that negative types allow us to break the dependency between the
+  producer and consumer and let each compute independently.
 \end{itemize}
 
-Similar observations hold for fractional types but we defer these to
-Sec.~\ref{sub:frac}. Before proceeding to the main part of the paper, we
-briefly summarize the our technical contributions:
+Instead of also discussing fractional types and related work in this
+introduction, we briefly summarize our main technical contributions and move
+on to the heart of the paper where these topics can be explained in their
+proper context. Our main contributions are:
 \begin{itemize}
-\item 
-\item 
-\item 
-\item 
+\item We extend {{langRev}} our reversible programming language of type
+  isomorphisms~\cite{rc2011,infeffects} with a notion of negative types, that
+  satisfies the isomorphism $a + (-a) = 0$. The semantics of this extension
+  is expressed by having a ``dual'' evaluator that reverses the flow of
+  execution for negative values. 
+\item We independently extend {{langRev}} with a notion of fractional types,
+  that satisfies the isomorphism $a * (1/a) = 1$. The semantics of this
+  extension is expressed by introducing logical variables and a unification
+  mechanism to model and resolve the constraints introduced by the fractional
+  types.
+\item We combine the above two extensions into one extension that includes
+  both negative and fractional types. The resulting language allows any
+  rational number to be used as a type. The types satisfy the same familiar
+  and intuitive isomorphisms that are satisfied in the mathematical field of
+  rational numbers.
+\item We develop programming intuition and argue that negative and fractional
+  types ought to be part of the vocabulary of every programmer.
+\item We relate our notions of negative and fractional types to previous work
+  on continuations. We analyze and relate two apparently unrelated lines of
+  work: the first pioneered by
+  Filinski~\cite{Filinski:1989:DCI:648332.755574} relates continuations to
+  negative types; the second~\cite{Bernardi:2010:CSL:1749618.1749689} relates
+  continuations to the fractional types of the Lambek-Grishin calculus.
 \end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Field of Rational Types: Intuitions}
-
-%%%%%%%%%%%%%%%%%%
-\subsection{Negative Types}
-\label{sub:neg}
-
-Consider the following two ways of purchasing an item that costs \$20.00:
-\begin{enumerate}
-\item You give the seller a \$20.00 bill.
-\item You use a credit card to pay the seller.
-\end{enumerate}
-In both cases, the seller receives the money immediately but there is a
-subtle difference. In the first transaction, the money received by the seller
-corresponds to a value that has been produced earlier in the computation. In
-the second transaction, the money may or may not exist yet: computationally,
-a \emph{debt} is generated to compensate for the money received by the seller
-and this debt travels ``backwards'' towards the bank where it is (hopefully)
-reconciled.
-
-The example suggests that the ability to consider values flowing backwards
-enriches our computational model. This observation goes back to Filinski's
-Masters thesis where continuations are identified with these negative values.
-We discuss the connections to Filinski's work and others in more detail in
-Sec.~\ref{sec:related}. For now, we note that in a conventional programming
-language in which values can be copied and deleted at will, extreme care is
-needed to keep track of negative values. Indeed, in the example above, if it
-were possible to simply delete the variable corresponding to the debt, we
-would have produced money out of nothing.
-
-For this reason, our treatment of negative values in the context of
-{{langRev}} is particularly simple. We will have a type $0$ and an
-isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-allows us to create a value and a corresponding debt out of nothing. Both the
-value and its negative counterpart can flow anywhere in the system. Because
-information is preserved, a closed program (which does not have any
-``dangling'' negative values) will eventually have to match the negative
-value with some corresponding value, effectively using the isomorphism in the
-right-to-left direction. In more detail, we can model the credit card
-transaction above using the following program (shown diagrammatically):
-
-FIGURE
-
-CLEAN UP the following based on the figure. In contrast, in our setting, one
-can start from the empty type $0$, introduce a positive value and its
-negative counterpart, and let each of these flow in arbitrary ways. The
-entire framework guarantees that neither the value nor its negative
-counterpart will be deleted or duplicated and hence that, in any closed
-program, the ``debt'' corresponding to the negative value is paid off exactly
-once. Computationally we model the first transaction using essentially the
-identity function which receives a \$20.00 bill as input from the buyer and
-propagates it on its output to the seller. The second transaction is more
-complicated. We model it as shown in the circuit below: There are two ways to
-understand this circuit that are both quite instructive. Let's first examine
-the type structure of each of the combinators that comprise the circuit. The
-first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
-produced from nothing so to speak by generating an equivalent debt that
-travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
-the execution. It goes forward in time so to speak, comes back, and then goes
-forwards again.
-
-It is critical that the framework in which the negative types are introduced
-is a framework in which all information is preserved, with no duplication or
-erasure. This guarantees that the generated debts are paid once and exactly
-once.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section{Fractional Types}
-
-The type $a/b$ is prominent in the Lambek-Grishin calculus which is
-extensively used in computational linguistics. In the common interpretation,
-a value of type $a/b$ is a value of some type $c$ such that when put in a
-context of type $b$, the result is a value of type $a$. For example, assuming
-contexts are represented as functions, a value of type $\texttt{bool} /
-(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
-$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
-in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
-type $\texttt{bool}$.
-
-In our case, the situation is simpler. If the goal is to produce a value of
-type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
-have type {{(a * b) / b}}. In most settings, this type would be equivalent to
-{{a}}. However in our setting, the constraint that this value must eventually
-fit in a larger value that supplies the {{b}} is explicitly recorded and must
-be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
-which when used in the left-to-right direction allows the creation of a value
-and a corresponding constraint on the context.  Both the value and the
-constraint can flow in arbitrary ways during the computation but eventually
-in a closed program with no ``dangling'' constraints, the isomorphism should
-be used in the right-to-left direction to resolve the constraints.  To
-understand the computational interpretation, consider the following example.
-
-The goal is to produce a value of type $(\texttt{bool} \times
-\texttt{int})$. One part of the program, $e_1$, knows how to produce the
-value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
-to produce the value of type $\texttt{bool}$ (say true). Computationally, we
-model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
-uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
-is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
-$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
-$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
-the two values meet, we can group the components as follows:
-\[
-(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
-\]
-Using the isomorphism in the right-to-left direction on the last two pairs,
-forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
-{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
-with the pair $(true,3)$ as desired.
-
-To summarize, one can think of a value of type $a/b$ as a value that imposes
-a constraint on its context of use: it is a value that requires its context
-to be of type $b$ and only if that condition is satisfied, can the value be
-considered as a value of type $a$. In the simplest case, the type $1/b$ just
-constraints the context to be of type $b$. By allowing the fractional types
-to be first-class, we can separate the generation of constraints from their
-use. Both the value and the constraint can flow in arbitrary ways during the
-computation but eventually in a closed program with no ``dangling''
-constraints, the isomorphism should be used in the right-to-left direction to
-resolve the constraints.
-
-NEED to work out the example in detail (or perhaps another example that's
-better.
-
-Again it is critical that the framework in which the fractional types are
-introduced is a framework in which all information is preserved, with no
-duplication or erasure. This guarantees that the generated constraints are
-satisfied once and exactly once.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary}
-
-To summarize negative, fractional, square root, and imaginary types all make
-sense. What they help you accomplish as a programmer is to disassociate
-global invariants into local ones that can be satisfied independently by
-subcomputations with no synchronization or communication. A computation
-producing something of type $a/b$ does not need to concern itself with who is
-going to supply the missing $b$: it just does its part. Conversely faced with
-a complicated task, a computation might decide to break it into pieces and
-demand these pieces using negative types.
-
-It is no surprise that these types are closely related to quantum mechanics
-and that they give us the feel that this is how nature computes. This is
-speculation however.
-
-In any case, in a framework where information can be copied and deleted, none
-of this makes much sense. It is critical that these constraints and demands
-can neither be duplicated nor erased.  This gives us the maximum
-``parallelism'' possible.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{A Graphical Language for {{langRev}} }
@@
 
-Since the monoidal units cxan be freely introduced and eliminated, in
-many diagrams they are omitted and dealt with explicitely when they
+Since the monoidal units can be freely introduced and eliminated, in
+many diagrams they are omitted and dealt with explicitly when they
 are of special interest.
@@
 \end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Field of Rational Types: Intuitions}
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Negative Types}
+\label{sub:neg}
+
+Consider the following two ways of purchasing an item that costs \$20.00:
+\begin{enumerate}
+\item You give the seller a \$20.00 bill.
+\item You use a credit card to pay the seller.
+\end{enumerate}
+In both cases, the seller receives the money immediately but there is a
+subtle difference. In the first transaction, the money received by the seller
+corresponds to a value that has been produced earlier in the computation. In
+the second transaction, the money may or may not exist yet: computationally,
+a \emph{debt} is generated to compensate for the money received by the seller
+and this debt travels ``backwards'' towards the bank where it is (hopefully)
+reconciled.
+
+The example suggests that the ability to consider values flowing backwards
+enriches our computational model. This observation goes back to Filinski's
+Masters thesis where continuations are identified with these negative values.
+We discuss the connections to Filinski's work and others in more detail in
+Sec.~\ref{sec:related}. For now, we note that in a conventional programming
+language in which values can be copied and deleted at will, extreme care is
+needed to keep track of negative values. Indeed, in the example above, if it
+were possible to simply delete the variable corresponding to the debt, we
+would have produced money out of nothing.
+
+For this reason, our treatment of negative values in the context of
+{{langRev}} is particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
+right-to-left direction. In more detail, we can model the credit card
+transaction above using the following program (shown diagrammatically):
+
+FIGURE
+
+CLEAN UP the following based on the figure. In contrast, in our setting, one
+can start from the empty type $0$, introduce a positive value and its
+negative counterpart, and let each of these flow in arbitrary ways. The
+entire framework guarantees that neither the value nor its negative
+counterpart will be deleted or duplicated and hence that, in any closed
+program, the ``debt'' corresponding to the negative value is paid off exactly
+once. Computationally we model the first transaction using essentially the
+identity function which receives a \$20.00 bill as input from the buyer and
+propagates it on its output to the seller. The second transaction is more
+complicated. We model it as shown in the circuit below: There are two ways to
+understand this circuit that are both quite instructive. Let's first examine
+the type structure of each of the combinators that comprise the circuit. The
+first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
+produced from nothing so to speak by generating an equivalent debt that
+travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
+the execution. It goes forward in time so to speak, comes back, and then goes
+forwards again.
+
+It is critical that the framework in which the negative types are introduced
+is a framework in which all information is preserved, with no duplication or
+erasure. This guarantees that the generated debts are paid once and exactly
+once.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Fractional Types}
+
+The type $a/b$ is prominent in the Lambek-Grishin calculus which is
+extensively used in computational linguistics. In the common interpretation,
+a value of type $a/b$ is a value of some type $c$ such that when put in a
+context of type $b$, the result is a value of type $a$. For example, assuming
+contexts are represented as functions, a value of type $\texttt{bool} /
+(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
+$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
+in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
+type $\texttt{bool}$.
+
+In our case, the situation is simpler. If the goal is to produce a value of
+type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
+have type {{(a * b) / b}}. In most settings, this type would be equivalent to
+{{a}}. However in our setting, the constraint that this value must eventually
+fit in a larger value that supplies the {{b}} is explicitly recorded and must
+be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
+which when used in the left-to-right direction allows the creation of a value
+and a corresponding constraint on the context.  Both the value and the
+constraint can flow in arbitrary ways during the computation but eventually
+in a closed program with no ``dangling'' constraints, the isomorphism should
+be used in the right-to-left direction to resolve the constraints.  To
+understand the computational interpretation, consider the following example.
+
+The goal is to produce a value of type $(\texttt{bool} \times
+\texttt{int})$. One part of the program, $e_1$, knows how to produce the
+value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
+to produce the value of type $\texttt{bool}$ (say true). Computationally, we
+model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
+uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
+is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
+$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
+$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
+the two values meet, we can group the components as follows:
+\[
+(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
+\]
+Using the isomorphism in the right-to-left direction on the last two pairs,
+forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
+{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
+with the pair $(true,3)$ as desired.
+
+To summarize, one can think of a value of type $a/b$ as a value that imposes
+a constraint on its context of use: it is a value that requires its context
+to be of type $b$ and only if that condition is satisfied, can the value be
+considered as a value of type $a$. In the simplest case, the type $1/b$ just
+constraints the context to be of type $b$. By allowing the fractional types
+to be first-class, we can separate the generation of constraints from their
+use. Both the value and the constraint can flow in arbitrary ways during the
+computation but eventually in a closed program with no ``dangling''
+constraints, the isomorphism should be used in the right-to-left direction to
+resolve the constraints.
+
+NEED to work out the example in detail (or perhaps another example that's
+better.
+
+Again it is critical that the framework in which the fractional types are
+introduced is a framework in which all information is preserved, with no
+duplication or erasure. This guarantees that the generated constraints are
+satisfied once and exactly once.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Summary}
+
+To summarize negative, fractional, square root, and imaginary types all make
+sense. What they help you accomplish as a programmer is to disassociate
+global invariants into local ones that can be satisfied independently by
+subcomputations with no synchronization or communication. A computation
+producing something of type $a/b$ does not need to concern itself with who is
+going to supply the missing $b$: it just does its part. Conversely faced with
+a complicated task, a computation might decide to break it into pieces and
+demand these pieces using negative types.
+
+It is no surprise that these types are closely related to quantum mechanics
+and that they give us the feel that this is how nature computes. This is
+speculation however.
+
+In any case, in a framework where information can be copied and deleted, none
+of this makes much sense. It is critical that these constraints and demands
+can neither be duplicated nor erased.  This gives us the maximum
+``parallelism'' possible.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%
 \subsection{Multiplicative Duality in {{langRev}} }
@@
 {{b1}} and a value of type {{b2}}. This interpretation is maintained
-in the precense of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
+in the presence of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
 to be viewed as a fission point for a value of type {{b}} and its
@@
 of a new typed logic variable of type {{b}} and its corresponding dual
-of type {{inv(b)}}. These two values are free to move independantly
+of type {{inv(b)}}. These two values are free to move independently
 through a circuit, but any operations on value affect what operations
@@
 
-This also leads to the exiciting point of view that entangled
+This also leads to the exciting point of view that entangled
 particle/anti-particle pairs act as functions in a physical
-first-class sense. We can ask the fuction to be applied to a value by
+first-class sense. We can ask the function to be applied to a value by
 unifying the anti-particle with a particle in a well-known state and
@@
 additive {{trace+}}, {{trace*}} gives us the ability to find the
-fixpoint of the contraint (expressed with combinators) that is
+fixpoint of the constraint (expressed with combinators) that is
 traced. If the constraint has not satisfying values, there is not
@@
 as permitted by the type {{b}}. So situate ourselves in any specific
-world, we have to supply additional information in teh form of
+world, we have to supply additional information in the form of
 unifying the particle or anti-particle with a particle representing
@@
 The construction here is not limited to boolean satisfiability, but
-generalises to any constraint satisfaction problem whose search space
+generalizes to any constraint satisfaction problem whose search space
 can be represented by a recursive type and for which an isomorphic

### 2012-03-06 rational.tex
organization
_
related work

@@
 
-Combinators of {{langRev}} can be written using the the operators
-described in definition \ref{def:langRev} or via a graphical language
-similar in spirit to those developed for GoI
-\cite{DBLP:conf/popl/Mackie95} and for category theory
-\cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
-conventions and shorthands we will describe here, the wiring diagrams
-are a complete and equivalent to the operator based (syntactic)
-description of programs.  {{langRev}} combinators expressed in this
-graphical language looks like a ``wiring diagrams''. Value take the
-form of ``particles'' that flow along the wires. Computation is
-expressed by the flow of particles.
+Combinators of {{langRev}} can be written using the the operators described
+in definition \ref{def:langRev} or via a graphical language similar in spirit
+to those developed for GoI~\cite{DBLP:conf/popl/Mackie95} and for category
+theory \cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
+conventions and shorthands we will describe here, the wiring diagrams are a
+complete and equivalent to the operator based (syntactic) description of
+programs.  {{langRev}} combinators expressed in this graphical language looks
+like a ``wiring diagrams''. Value take the form of ``particles'' that flow
+along the wires. Computation is expressed by the flow of particles.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Field of Rational Types: Intuitions}
-
-%%%%%%%%%%%%%%%%%%
-\subsection{Negative Types}
+\section{Negative Types}
 \label{sub:neg}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Fractional Types}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
+{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
+defined only if {{b1}} and {{b2}} are entirely positive types. This is
+similar to the constraint that Zeilberger imposes to explain intuitionistic
+polarity and delimited control~\cite{10.1109/LICS.2010.23}.
+
+Consider the program that
+
+(TODO: So this is a really bad explanation. But this section is really
+important and we need to explain carefully. Maybe an analogy can be
+drawn to Cayley diagrams for groups to point out that the notion of
+`i' is entirely artificial, but it is essential to the discourse.)
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The outputs of
+programs that output mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Additive Duality in {{langRev}} }
+
+Additive duality is added to {{langRev}} by addition of negative types
+denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
+extend the language as follows:
+
+%subcode{bnf} include main
+% Value Types, b = ... | -b
+% Values, v = ... | -v
+%
+% Isomorphisms, iso &=& ... | eta+ | eps+
+
+\noindent
+with the types judgments
+
+%subcode{opsem} include main
+% eta+ &: 0 <-> (-b) + b :& eps+
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
+
+We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
+the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
+dually on the right is {{eps+}} showing the map from {{-b+b}} to
+0. Even though the diagrams below show a {{0}} wire for completeness,
+later diagrams will always drop them in contexts where we can
+implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps.pdf}
+\end{center}
+\end{multicols}
+
+The usual interpretation of the type {{b1+b2}} is that we either have
+an appropriately tagged value of type {{b1}} or an appropriately
+tagged value of type {{b2}}. This interpretation is maintained in the
+presence of {{eta+}} and {{eps+}} in the following sense: a value of
+type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
+switches into a value {{L (-v): -b + b}} that flows backwards on the
+upper wire. The inversion of direction is captured by the negative
+sign on the value.
+
+\paragraph*{Negative Information Flow.} 
+Hence, the intuitive way to read the diagrams is that they reverse the
+flow of information -- in normal {{langRev}} circuits information
+flows from left-to-right and {{eps+}} operations cause information to
+flow from right-to-left.  The interpretation of {{eta+}} is precisely
+the opposite: it turns a right-to-left flow of information in a
+circuit into a left-to-right flow.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus2.pdf}
+}
+\end{center}  
+\end{multicols}
+
+\paragraph*{Time Traveling.}
+A time traveling intuition is also applicable here: if view the
+left-to-right direction of information flow as the canonical forward
+direction of information, then the action of {{eps+}} allows values,
+aka information particles, to flow backwards in time. Particles that
+flow backward in time get to see and interact with the history of
+particles that they coexist with i.e. values that they exist
+multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
+(whose construction is shown in Section \ref{sec:neg-constructions}).
+
+\paragraph*{Coherence.}
+The appropriate coherence condition for Compact Closed Categories,
+expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
+may be visualized as explained by Selinger
+\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+\paragraph*{Non-termination.}
+As is well known in category theory, Compact Closed Categories contain
+trace operations.  In the context of {{langRev}} this implies that
+{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
+(detailed in Section \ref{sec:monoidal-constructions}) hence
+introducing \emph{non-termination} into the language
+\cite{infeffects}.
+
+\paragraph*{Zero Information.}
+The fact that {{eta+}} (and hence {{eps+}}) introduces no information
+can be understood is by viewing {{-b+b}} as a type whose value is
+observable only if you `pay' {{b}} amount of information to the {{-b}}
+branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
+amount of information is returned to you. (TODO. more needs to be said
+here.)
+
+ADD MATH.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Fractional Types}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Summary}
-
-To summarize negative, fractional, square root, and imaginary types all make
-sense. What they help you accomplish as a programmer is to disassociate
-global invariants into local ones that can be satisfied independently by
-subcomputations with no synchronization or communication. A computation
-producing something of type $a/b$ does not need to concern itself with who is
-going to supply the missing $b$: it just does its part. Conversely faced with
-a complicated task, a computation might decide to break it into pieces and
-demand these pieces using negative types.
-
-It is no surprise that these types are closely related to quantum mechanics
-and that they give us the feel that this is how nature computes. This is
-speculation however.
-
-In any case, in a framework where information can be copied and deleted, none
-of this makes much sense. It is critical that these constraints and demands
-can neither be duplicated nor erased.  This gives us the maximum
-``parallelism'' possible.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ {{langRevEE}} }
-
-\paragraph*{Observability.} 
-Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
-input {{v1 : b1}} gives us a value {{v2 : b2}} on
-termination. Execution is well defined only if {{b1}} and {{b2}} are
-entirely positive types. Consider the program that
-
-(TODO: So this is a really bad explanation. But this section is really
-important and we need to explain carefully. Maybe an analogy can be
-drawn to Cayley diagrams for groups to point out that the notion of
-`i' is entirely artificial, but it is essential to the discourse.)
-
-\begin{center}
-  \includegraphics{diagrams/shuffle.pdf}
-\end{center}
-
-We define observables to be only positive types. The outputs of
-programs that output mixed positive and negative types are not
-observable.  Also, programs that input mixed positive and negatives
-types are not executable.
-
-%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Additive Duality in {{langRev}} }
-
-Additive duality is added to {{langRev}} by addition of negative types
-denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
-extend the language as follows:
-
-%subcode{bnf} include main
-% Value Types, b = ... | -b
-% Values, v = ... | -v
-%
-% Isomorphisms, iso &=& ... | eta+ | eps+
-
-\noindent
-with the types judgments
-
-%subcode{opsem} include main
-% eta+ &: 0 <-> (-b) + b :& eps+
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-
-We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
-the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
-dually on the right is {{eps+}} showing the map from {{-b+b}} to
-0. Even though the diagrams below show a {{0}} wire for completeness,
-later diagrams will always drop them in contexts where we can
-implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/eta.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/eps.pdf}
-\end{center}
-\end{multicols}
-
-The usual interpretation of the type {{b1+b2}} is that we either have
-an appropriately tagged value of type {{b1}} or an appropriately
-tagged value of type {{b2}}. This interpretation is maintained in the
-presence of {{eta+}} and {{eps+}} in the following sense: a value of
-type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
-switches into a value {{L (-v): -b + b}} that flows backwards on the
-upper wire. The inversion of direction is captured by the negative
-sign on the value.
-
-\paragraph*{Negative Information Flow.} 
-Hence, the intuitive way to read the diagrams is that they reverse the
-flow of information -- in normal {{langRev}} circuits information
-flows from left-to-right and {{eps+}} operations cause information to
-flow from right-to-left.  The interpretation of {{eta+}} is precisely
-the opposite: it turns a right-to-left flow of information in a
-circuit into a left-to-right flow.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus1.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus2.pdf}
-}
-\end{center}  
-\end{multicols}
-
-\paragraph*{Time Traveling.}
-A time traveling intuition is also applicable here: if view the
-left-to-right direction of information flow as the canonical forward
-direction of information, then the action of {{eps+}} allows values,
-aka information particles, to flow backwards in time. Particles that
-flow backward in time get to see and interact with the history of
-particles that they coexist with i.e. values that they exist
-multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
-(whose construction is shown in Section \ref{sec:neg-constructions}).
-
-\paragraph*{Coherence.}
-The appropriate coherence condition for Compact Closed Categories,
-expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-may be visualized as explained by Selinger
-\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-\paragraph*{Non-termination.}
-As is well known in category theory, Compact Closed Categories contain
-trace operations.  In the context of {{langRev}} this implies that
-{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
-(detailed in Section \ref{sec:monoidal-constructions}) hence
-introducing \emph{non-termination} into the language
-\cite{infeffects}.
-
-\paragraph*{Zero Information.}
-The fact that {{eta+}} (and hence {{eps+}}) introduces no information
-can be understood is by viewing {{-b+b}} as a type whose value is
-observable only if you `pay' {{b}} amount of information to the {{-b}}
-branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
-amount of information is returned to you. (TODO. more needs to be said
-here.)
-
-ADD MATH.
-
 %%%%%%%%%%%%%%%%%%%%%
@@
 
+For a traced monoidal category {{C}} the Int construction produces a
+Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
+Further we know that the target of the Int construction is isomorphic
+to the target of \G construction of Abramsky \cite{Abramsky96:0} from
+Haghverdi. However, note that the {{langRevEE}} is not the same as the
+image of the Int construction on {{langRevT}}, since the the later
+lacks a multiplicative tensor that distributes over the additive tensor
+in Int {{langRevT}}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+We can now encode the GoI machine of
+Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. \amr{The machine uses bang
+  which allows arbitrary duplication. Can we really do that???}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-(NOTE: Lets use this section here to establish a long of ideas and
-intuition. None of these connections have to be formal, but they can
-be concrete.)
+\paragraph*{Negation in Classical Linear Logic} 
+Filinski~\cite{Filinski92} uses the negative types of linear logic to model
+continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
+negative types of linear logic as \emph{acceptors}, which are like
+continuations in the sense that they take an input and return no
+output. Acceptors however are also similar in flavor to logical variables:
+they can created and instantiated later once their context of use is
+determined. It is clear that, at an intuitive level, acceptors are entities
+that combine elements of our negative and fractional types.
+
+\paragraph*{Declarative Continuations.} 
+In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
+proposes that continuations are a \emph{declarative} concept. He,
+furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+which values and continuations are treated as opposites. This is essentially
+what we are proposing with one fundamental difference: our underlying
+language is not the $\lambda$-calculus but a language of pure isomorphisms in
+which information is preserved. This shift of perspective enables us to
+distill and generalize the duality of values and continuations: in
+particular, in the conventional $\lambda$-calculus setting values and
+continuations can be erased and duplicated which makes it difficult to
+maintain the correspondence between a value and its negative counterpart. In
+contrast, in our setting, one can start from the empty type $0$, introduce a
+positive value and its negative counterpart, and let each of these flow in
+arbitrary ways. The entire framework guarantees that neither the value nor
+its negative counterpart will be deleted or duplicated and hence that, in any
+closed program, the ``debt'' corresponding to the negative value is paid off
+exactly once. The forward and backward executions in our framework correspond
+to call-by-value and call-by-name. This duality was observed by Filinski and
+others following him but it is particularly clean in our framework.
+
+\paragraph*{The Duality of Computation.}
+Wadler the reloaded paper, does not consider the subtraction type because its
+``computational interpretation is not familiar.'' Curien and Herbelin study
+duality in classical logic, show that it exchanges call-by-value and
+call-by-name. They extend classical natural deduction with subtraction but
+give it no computational meaning. Crolard (in the formulae-as-types
+interpretation of subtractive logic) address the computational interpretation
+of subtraction. He explains the type $A-B$ as the of \emph{coroutines} with a
+local environment of type $A$ and a continuation of type $B$. The description
+is complicated by what is essentially the desire to enforce linearity
+constraints so that coroutines cannot access the local environment of other
+coroutines. Must cite Selinger control categories in this context of duality
+but I am not sure what to say: it assumes cartesian closed categories for one
+thing and not symmetric monoidal ones. Ariola, Herbelin, and Sabry also use
+subtractive types to explain delimited continuations.
+
+\paragraph*{Polarized Logic.} Polarized logic is the logic of values and
+continuations: each connective can be given a positive or negative polarity,
+which can be related to evaluation
+order~\cite{10.1109/LICS.2010.23}. Negative types in this work are a mix of
+our negative and fractional types: they include a notion of demand which
+corresponds to our negative types and a notion of pattern-matching on
+continuations (i.e., a notion of constraints on the context) which
+corresponds to our fractional types.
+
+\paragraph*{The Lambek-Grishin Calculus.} The ``parsing-as-deduction'' style
+of linguistic analysis uses the Lambek-Grishin calculus with the following
+types: product, left division, right division, coproduct, right difference,
+and left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division
+and difference types are similar to our types but because the calculus lacks
+commutativity and associativity and only has limited notions of
+distributivity, each connective needs a left and right version. The
+Lambek-Grishin exhibits two notions of symmetry but they are unrelated to our
+notions. In particular, the first notion of symmetry expresses commutativity
+and the second relates products to sums and divisions to subtractions. In
+contract, our two symmetries relate sums to subtractions and products to
+divisions.
+
+\newpage
+
+Curien and Herbelin~\cite{Curien:2000} introduce a calculus that exhibits
+symmetries between values and continuations and between call-by-value and
+call-by-name. The calculus includes the type $A-B$ which is the dual of
+implication, i.e., a value of type $A-B$ is a context expecting a function of
+type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
+as a \emph{pair} consisting of a value of type $A$ and a continuation of type
+$B$. which whose computational interpretation is left for future work. Curien
+and Herbelin informally explain values of types $A-B$ either as a context
+expecting a function $A \rightarrow B$ or as a \emph{pair} containing a value
+and a continuation. In our interpretation, a value of type $A-B$ is
+\emph{either} a value of $A$ or a demand for a value of type $B$.
 
@@
 
-\paragraph*{Continuations.} The idea of using negative types to model
-information flowing backwards, demand for values, continuations, etc. goes
-back to at Filinski's thesis. We recall these connections below but we first
-note that all these systems are complicated because in all these systems
-information can be ignored, destroyed, or duplicated. Clearly the possibility
-of erasure of information would mean that our credit card transaction is
-incorrect. In our work, information is maintained and hence we have a
-guarantee that, in a closed program, the debt must be accounted and paid for.
-
-\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
-In his Masters thesis, Filinski proposes that continuations are a
-\emph{declarative} concept. He, furthermore, introduces a symmetric extension
-of the $\lambda$-calculus in which values and continuations are treated as
-opposites. This is essentially what we are proposing with one fundamental
-difference: our underlying language is not the $\lambda$-calculus but a
-language of pure isomorphisms in which information is preserved. This shift
-of perspective enables us to distill and generalize the duality of values and
-continuations: in particular, in the conventional $\lambda$-calculus setting
-values and continuations can be erased and duplicated which makes it
-difficult to maintain the correspondence between a value and its negative
-counterpart. In contrast, in our setting, one can start from the empty type
-$0$, introduce a positive value and its negative counterpart, and let each of
-these flow in arbitrary ways. The entire framework guarantees that neither
-the value nor its negative counterpart will be deleted or duplicated and
-hence that, in any closed program, the ``debt'' corresponding to the negative
-value is paid off exactly once. The forward and backward executions in our
-framework correspond to call-by-value and call-by-name. This duality was
-observed by Filinski and others following him but it is particularly clean in
-our framework.
-
-\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
-subtraction type because its ``computational interpretation is not
-familiar.'' Curien and Herbelin study duality in classical logic, show that
-it exchanges call-by-value and call-by-name. They extend classical natural
-deduction with subtraction but give it no computational meaning. Crolard (in
-the formulae-as-types interpretation of subtractive logic) address the
-computational interpretation of subtraction. He explains the type $A-B$ as
-the of \emph{coroutines} with a local environment of type $A$ and a
-continuation of type $B$. The description is complicated by what is
-essentially the desire to enforce linearity constraints so that coroutines
-cannot access the local environment of other coroutines. Must cite Selinger
-control categories in this context of duality but I am not sure what to say:
-it assumes cartesian closed categories for one thing and not symmetric
-monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
-explain delimited continuations.
-
-\paragraph*{Linear Logic} In accounts that are linear, the value and
-continuation that comprise the subtractive type need to be constrained to
-``stay together.'' This can be achieved by various restrictions. In this work
-we have no such constraints, the negative value can flow anywhere. The entire
-system guarantees that any closed program would have to account for it. We
-don't have to introduce special constraints to achieve that. Zeilberger in
-the paper on polarity and the logic of delimited continuations uses polarized
-logic: he shows that if positive and negative values are completely symmetric
-except that answer types are positive, then the framework accommodates
-delimited continuations. But he interprets negative values are control
-operators, or as values defined by the shape of their continuations. We
-simply interpret values of negative type as values flowing in the ``other''
-direction.
-
-\paragraph*{Int Construction.}
-For a traced monoidal category {{C}} the Int construction produces a
-Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
-Further we know that the target of the Int construction is isomorphic
-to the target of \G construction of Abramsky \cite{Abramsky96:0} from
-Haghverdi. However, note that the {{langRevEE}} is not the same as the
-image of the Int construction on {{langRevT}}, since the the later
-lacks a multiplicative tensor that distributes over the additive tensor
-in Int {{langRevT}}.
-
-\paragraph*{Recursion.} In our previous work we introduce recursive types and
-trace operators. This is dangerous here because infinite loops allow us to
-prolong paying the debt for as long as we want.
-
-\paragraph*{GoI machines} 
-We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.
+The idea of using negative types to model information flowing backwards,
+demand for values, continuations, etc. goes back to at Filinski's thesis. We
+recall these connections below but we first note that all these systems are
+complicated because in all these systems information can be ignored,
+destroyed, or duplicated. Clearly the possibility of erasure of information
+would mean that our credit card transaction is incorrect. In our work,
+information is maintained and hence we have a guarantee that, in a closed
+program, the debt must be accounted and paid for.
+
+Much of previous work builds on the idea that there is one duality in
+computation between values and continuations which manifests itself as a
+duality between the call-by-value and call-by-name parameter-passing
+mechanisms. The former mechanism focuses on evaluating expressions to values
+even if these values are not demanded by the context; the latter focuses on
+evaluating expressions to continuations even if these continuations might be
+aborted. The idea that a continuation is dual to a value is intuitive but
+then one would expect that the sum of two values naturally corresponds to the
+sum of two continuations and that the product of two values naturally
+corresponds to the product of two continuations.
+
+check and say that our work teases the continuations into a negative part
+(which simply demands a value) and a fractional part (which imposes
+constraints on how this value will be used). So something like $-1/c$ is
+needed to express a conventional continuation. Having two dualities makes the
+whole calculus natural and symmetric.
+
+Consider a continuation that takes $x$ and $y$ and swaps them. It can't be
+expressed using two conventional continuations because the demand and the way
+it is used are entangled together.
+
+In accounts that are linear, the value and continuation that comprise the
+subtractive type need to be constrained to ``stay together.'' This can be
+achieved by various restrictions. In this work we have no such constraints,
+the negative value can flow anywhere. The entire system guarantees that any
+closed program would have to account for it. We don't have to introduce
+special constraints to achieve that. Zeilberger in the paper on polarity and
+the logic of delimited continuations uses polarized logic: he shows that if
+positive and negative values are completely symmetric except that answer
+types are positive, then the framework accommodates delimited
+continuations. But he interprets negative values are control operators, or as
+values defined by the shape of their continuations. We simply interpret
+values of negative type as values flowing in the ``other'' direction.
 
@@
 
+To summarize negative, fractional, square root, and imaginary types all make
+sense. What they help you accomplish as a programmer is to disassociate
+global invariants into local ones that can be satisfied independently by
+subcomputations with no synchronization or communication. A computation
+producing something of type $a/b$ does not need to concern itself with who is
+going to supply the missing $b$: it just does its part. Conversely faced with
+a complicated task, a computation might decide to break it into pieces and
+demand these pieces using negative types.
+
+It is no surprise that these types are closely related to quantum mechanics
+and that they give us the feel that this is how nature computes. This is
+speculation however.
+
+In any case, in a framework where information can be copied and deleted, none
+of this makes much sense. It is critical that these constraints and demands
+can neither be duplicated nor erased.  This gives us the maximum
+``parallelism'' possible.
+
+In our previous work we introduce recursive types and trace operators. This
+is dangerous here because infinite loops allow us to prolong paying the debt
+for as long as we want. The algebraic
+$\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that extreme
+care is needed when combining negative and fractional types with the full
+$\lambda$-calculus (including recursion).
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 This highly intuitive and appealing idea was further explored and refined by
-many authors~\cite{Griffin:1989:FNC:96709.96714, curien00duality,
+many authors~\cite{Griffin:1989:FNC:96709.96714, Curien:2000,
   Wadler:2003, DBLP:conf/rta/Wadler05}. Yet, despite its appeal, the duality

### 2012-03-06 rational.tex
related work

@@
 
+\paragraph*{Declarative Continuations.} 
+In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
+proposes that continuations are a \emph{declarative} concept. He,
+furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+which values and continuations are treated as opposites. In more detail, the
+symmetric calculus contains a ``value'' fragment and a ``continuation''
+fragment which are mirror images. Pairs and sums are treated as duals in the
+sense that the ``value'' fragment includes pairs whose mirror image in the
+``continuation'' fragment are sums. In contrast, our language includes pairs
+and sums in the value fragment and two symmetries: one that maps the pairs to
+fractions and another that maps the sums to subtractions.
+
+\paragraph*{The Duality of Computation.}
+The precise duality between call-by-name and call-by-value was further
+investigated by Selinger~\cite{Selinger:2001:CCD:966910.966911}. Curien and
+Herbelin~\cite{Curien:2000} introduce a calculus that exhibits symmetries
+between values and continuations and between call-by-value and
+call-by-name. The calculus includes the type $A-B$ which is the dual of
+implication, i.e., a value of type $A-B$ is a context expecting a function of
+type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
+as a \emph{pair} consisting of a value of type $A$ and a continuation of type
+$B$. This is to be contrasted with our interpretation of a value of that type
+as \emph{either} a value of $A$ or a demand for a value of type $B$. This
+calculus was further analyzed and extended by
+Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
+interpretation to the subtraction connective and like the original symmetric
+calculus of Filinksi, introduces a duality that relates sums to products and
+vice-versa.
+
+\paragraph*{Subtractive Logic.} 
+Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
+logic which contains a dual to implication. Her work has been distilled in
+the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
+related to coroutines~\cite{Crolard01082004} and delimited
+continuations~\cite{Ariola:2009:TFD:1743339.1743381}.  In more detail,
+Crolard explains the type $A-B$ as the of \emph{coroutines} with a local
+environment of type $A$ and a continuation of type $B$. The description is
+complicated by what is essentially the desire to enforce linearity
+constraints so that coroutines cannot access the local environment of other
+coroutines.
+
 \paragraph*{Negation in Classical Linear Logic} 
@@
 that combine elements of our negative and fractional types.
-
-\paragraph*{Declarative Continuations.} 
-In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
-proposes that continuations are a \emph{declarative} concept. He,
-furthermore, introduces a symmetric extension of the $\lambda$-calculus in
-which values and continuations are treated as opposites. This is essentially
-what we are proposing with one fundamental difference: our underlying
-language is not the $\lambda$-calculus but a language of pure isomorphisms in
-which information is preserved. This shift of perspective enables us to
-distill and generalize the duality of values and continuations: in
-particular, in the conventional $\lambda$-calculus setting values and
-continuations can be erased and duplicated which makes it difficult to
-maintain the correspondence between a value and its negative counterpart. In
-contrast, in our setting, one can start from the empty type $0$, introduce a
-positive value and its negative counterpart, and let each of these flow in
-arbitrary ways. The entire framework guarantees that neither the value nor
-its negative counterpart will be deleted or duplicated and hence that, in any
-closed program, the ``debt'' corresponding to the negative value is paid off
-exactly once. The forward and backward executions in our framework correspond
-to call-by-value and call-by-name. This duality was observed by Filinski and
-others following him but it is particularly clean in our framework.
-
-\paragraph*{The Duality of Computation.}
-Wadler the reloaded paper, does not consider the subtraction type because its
-``computational interpretation is not familiar.'' Curien and Herbelin study
-duality in classical logic, show that it exchanges call-by-value and
-call-by-name. They extend classical natural deduction with subtraction but
-give it no computational meaning. Crolard (in the formulae-as-types
-interpretation of subtractive logic) address the computational interpretation
-of subtraction. He explains the type $A-B$ as the of \emph{coroutines} with a
-local environment of type $A$ and a continuation of type $B$. The description
-is complicated by what is essentially the desire to enforce linearity
-constraints so that coroutines cannot access the local environment of other
-coroutines. Must cite Selinger control categories in this context of duality
-but I am not sure what to say: it assumes cartesian closed categories for one
-thing and not symmetric monoidal ones. Ariola, Herbelin, and Sabry also use
-subtractive types to explain delimited continuations.
-
-\paragraph*{Polarized Logic.} Polarized logic is the logic of values and
-continuations: each connective can be given a positive or negative polarity,
-which can be related to evaluation
-order~\cite{10.1109/LICS.2010.23}. Negative types in this work are a mix of
-our negative and fractional types: they include a notion of demand which
-corresponds to our negative types and a notion of pattern-matching on
-continuations (i.e., a notion of constraints on the context) which
-corresponds to our fractional types.
 
@@
 divisions.
-
-\newpage
-
-Curien and Herbelin~\cite{Curien:2000} introduce a calculus that exhibits
-symmetries between values and continuations and between call-by-value and
-call-by-name. The calculus includes the type $A-B$ which is the dual of
-implication, i.e., a value of type $A-B$ is a context expecting a function of
-type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
-as a \emph{pair} consisting of a value of type $A$ and a continuation of type
-$B$. which whose computational interpretation is left for future work. Curien
-and Herbelin informally explain values of types $A-B$ either as a context
-expecting a function $A \rightarrow B$ or as a \emph{pair} containing a value
-and a continuation. In our interpretation, a value of type $A-B$ is
-\emph{either} a value of $A$ or a demand for a value of type $B$.
-
-Filinski proposes that continuations are a \emph{declarative} concept. He,
-furthermore, introduces a symmetric extension of the $\lambda$-calculus in
-which values and continuations are treated as opposites. This is essentially
-what we are proposing with one fundamental difference: our underlying
-language is not the $\lambda$-calculus but a language of pure isomorphisms in
-which information is preserved. This shift of perspective enables us to
-distill and generalize the duality of values and continuations: in
-particular, in the conventional $\lambda$-calculus setting values and
-continuations can be erased and duplicated which makes it difficult to
-maintain the correspondence between a value and its negative counterpart.
-
-The idea of using negative types to model information flowing backwards,
-demand for values, continuations, etc. goes back to at Filinski's thesis. We
-recall these connections below but we first note that all these systems are
-complicated because in all these systems information can be ignored,
-destroyed, or duplicated. Clearly the possibility of erasure of information
-would mean that our credit card transaction is incorrect. In our work,
-information is maintained and hence we have a guarantee that, in a closed
-program, the debt must be accounted and paid for.
-
-Much of previous work builds on the idea that there is one duality in
-computation between values and continuations which manifests itself as a
-duality between the call-by-value and call-by-name parameter-passing
-mechanisms. The former mechanism focuses on evaluating expressions to values
-even if these values are not demanded by the context; the latter focuses on
-evaluating expressions to continuations even if these continuations might be
-aborted. The idea that a continuation is dual to a value is intuitive but
-then one would expect that the sum of two values naturally corresponds to the
-sum of two continuations and that the product of two values naturally
-corresponds to the product of two continuations.
-
-check and say that our work teases the continuations into a negative part
-(which simply demands a value) and a fractional part (which imposes
-constraints on how this value will be used). So something like $-1/c$ is
-needed to express a conventional continuation. Having two dualities makes the
-whole calculus natural and symmetric.
-
-Consider a continuation that takes $x$ and $y$ and swaps them. It can't be
-expressed using two conventional continuations because the demand and the way
-it is used are entangled together.
-
-In accounts that are linear, the value and continuation that comprise the
-subtractive type need to be constrained to ``stay together.'' This can be
-achieved by various restrictions. In this work we have no such constraints,
-the negative value can flow anywhere. The entire system guarantees that any
-closed program would have to account for it. We don't have to introduce
-special constraints to achieve that. Zeilberger in the paper on polarity and
-the logic of delimited continuations uses polarized logic: he shows that if
-positive and negative values are completely symmetric except that answer
-types are positive, then the framework accommodates delimited
-continuations. But he interprets negative values are control operators, or as
-values defined by the shape of their continuations. We simply interpret
-values of negative type as values flowing in the ``other'' direction.
 
@@
 and continuations are dual notions. This observation was followed by numerous
-
-Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
-logic which contains a dual to implication. Her work has been distilled in
-the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
-related to coroutines~\cite{Crolard01082004} and delimited
-continuations~\cite{Ariola:2009:TFD:1743339.1743381}.
-
-In a seemingly unrelated development, Girard in 1987 introduced linear
 
@@
 
-Connections to type logical grammars and Lambek calculus
-
-Some background on {{langRev}} because it is mentioned in the next section.
+Filinski proposes that continuations are a \emph{declarative} concept. He,
+furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+which values and continuations are treated as opposites. This is essentially
+what we are proposing with one fundamental difference: our underlying
+language is not the $\lambda$-calculus but a language of pure isomorphisms in
+which information is preserved. This shift of perspective enables us to
+distill and generalize the duality of values and continuations: in
+particular, in the conventional $\lambda$-calculus setting values and
+continuations can be erased and duplicated which makes it difficult to
+maintain the correspondence between a value and its negative counterpart.
+
+The idea of using negative types to model information flowing backwards,
+demand for values, continuations, etc. goes back to at Filinski's thesis. We
+recall these connections below but we first note that all these systems are
+complicated because in all these systems information can be ignored,
+destroyed, or duplicated. Clearly the possibility of erasure of information
+would mean that our credit card transaction is incorrect. In our work,
+information is maintained and hence we have a guarantee that, in a closed
+program, the debt must be accounted and paid for.
+
+Much of previous work builds on the idea that there is one duality in
+computation between values and continuations which manifests itself as a
+duality between the call-by-value and call-by-name parameter-passing
+mechanisms. The former mechanism focuses on evaluating expressions to values
+even if these values are not demanded by the context; the latter focuses on
+evaluating expressions to continuations even if these continuations might be
+aborted. The idea that a continuation is dual to a value is intuitive but
+then one would expect that the sum of two values naturally corresponds to the
+sum of two continuations and that the product of two values naturally
+corresponds to the product of two continuations.
+
+check and say that our work teases the continuations into a negative part
+(which simply demands a value) and a fractional part (which imposes
+constraints on how this value will be used). So something like $-1/c$ is
+needed to express a conventional continuation. Having two dualities makes the
+whole calculus natural and symmetric.
+
+Consider a continuation that takes $x$ and $y$ and swaps them. It can't be
+expressed using two conventional continuations because the demand and the way
+it is used are entangled together.
+
+In accounts that are linear, the value and continuation that comprise the
+subtractive type need to be constrained to ``stay together.'' This can be
+achieved by various restrictions. In this work we have no such constraints,
+the negative value can flow anywhere. The entire system guarantees that any
+closed program would have to account for it. We don't have to introduce
+special constraints to achieve that. Zeilberger in the paper on polarity and
+the logic of delimited continuations uses polarized logic: he shows that if
+positive and negative values are completely symmetric except that answer
+types are positive, then the framework accommodates delimited
+continuations. But he interprets negative values are control operators, or as
+values defined by the shape of their continuations. We simply interpret
+values of negative type as values flowing in the ``other'' direction.
+
+This is essentially what we are proposing with one fundamental difference:
+our underlying language is not the $\lambda$-calculus but a language of pure
+isomorphisms in which information is preserved. This shift of perspective
+enables us to distill and generalize the duality of values and continuations:
+in particular, in the conventional $\lambda$-calculus setting values and
+continuations can be erased and duplicated which makes it difficult to
+maintain the correspondence between a value and its negative counterpart. In
+contrast, in our setting, one can start from the empty type $0$, introduce a
+positive value and its negative counterpart, and let each of these flow in
+arbitrary ways. The entire framework guarantees that neither the value nor
+its negative counterpart will be deleted or duplicated and hence that, in any
+closed program, the ``debt'' corresponding to the negative value is paid off
+exactly once. The forward and backward executions in our framework correspond
+to call-by-value and call-by-name. This duality was observed by Filinski and
+others following him but it is particularly clean in our framework.
 

### 2012-03-06 rational.tex
related work

@@
 
+The idea of ``negative types'' has appeared many times in the literature and
+has often been related to some form of continuations. Fractional types are
+less common but have also appeared in relation to parsing natural
+languages. Although each of these previous occurrences of negative and
+fractional types is somewhat related to our work, our results are
+subtantially different. To clarify this point, we start by reviewing the
+salient point of the major pieces of related work and conclude this section
+with a summary contrasting our approach to previous work.
+
 \paragraph*{Declarative Continuations.} 
@@
 furthermore, introduces a symmetric extension of the $\lambda$-calculus in
-which values and continuations are treated as opposites. In more detail, the
-symmetric calculus contains a ``value'' fragment and a ``continuation''
-fragment which are mirror images. Pairs and sums are treated as duals in the
-sense that the ``value'' fragment includes pairs whose mirror image in the
-``continuation'' fragment are sums. In contrast, our language includes pairs
-and sums in the value fragment and two symmetries: one that maps the pairs to
-fractions and another that maps the sums to subtractions.
+which call-by-value is dual to call-by-name and values are dual to
+continuations. In more detail, the symmetric calculus contains a ``value''
+fragment and a ``continuation'' fragment which are mirror images. Pairs and
+sums are treated as duals in the sense that the ``value'' fragment includes
+pairs whose mirror image in the ``continuation'' fragment are sums. In
+contrast, our language includes pairs and sums in the value fragment and two
+symmetries: one that maps the pairs to fractions and another that maps the
+sums to subtractions.
 
 \paragraph*{The Duality of Computation.}
-The precise duality between call-by-name and call-by-value was further
-investigated by Selinger~\cite{Selinger:2001:CCD:966910.966911}. Curien and
-Herbelin~\cite{Curien:2000} introduce a calculus that exhibits symmetries
-between values and continuations and between call-by-value and
+The duality between call-by-name and call-by-value was further investigated
+by Selinger using control
+categories~\cite{Selinger:2001:CCD:966910.966911}. Curien and
+Herbelin~\cite{Curien:2000} also introduce a calculus that exhibits
+symmetries between values and continuations and between call-by-value and
 call-by-name. The calculus includes the type $A-B$ which is the dual of
@@
 continuations~\cite{Ariola:2009:TFD:1743339.1743381}.  In more detail,
-Crolard explains the type $A-B$ as the of \emph{coroutines} with a local
+Crolard explains the type $A-B$ as the type of \emph{coroutines} with a local
 environment of type $A$ and a continuation of type $B$. The description is
@@
 they can created and instantiated later once their context of use is
-determined. It is clear that, at an intuitive level, acceptors are entities
-that combine elements of our negative and fractional types.
+determined. Although a formal connection is lacking, it is clear that, at an
+intuitive level, acceptors are entities that combine elements of our negative
+and fractional types.
 
@@
 of linguistic analysis uses the Lambek-Grishin calculus with the following
-types: product, left division, right division, coproduct, right difference,
-and left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division
-and difference types are similar to our types but because the calculus lacks
+types: product, left division, right division, sum, right difference, and
+left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division and
+difference types are similar to our types but because the calculus lacks
 commutativity and associativity and only has limited notions of
@@
 and the second relates products to sums and divisions to subtractions. In
-contract, our two symmetries relate sums to subtractions and products to
+contrast, our two symmetries relate sums to subtractions and products to
 divisions.
+
+\paragraph*{Our Approach.} The salient aspects of our approach are the following:
+\begin{itemize}
+\item Negative and fractional types have an elementary and familiar
+  interpretation borrowed from the algebra of rational numbers. One can write
+  any algebraic identity that is valid for the rational numbers and interpret
+  it as an isomorphism with a clear computational interpretation: negative
+  values flow backwards and fractional values represent fragments of
+  aggregate data structures. None of the systems above has such a natural
+  interpretation of negative and fractional types. 
+\item Because we are \emph{not} in the context of the full
+  $\lambda$-calculus, which allows arbitrary duplication and erasure of
+  information, values of negative and fractional types are first-class values
+  that can flow anywhere. The information-preserving computational
+  infrastructure guarantees that, in a complete program, every negative
+  demand will be satisfied exactly once, and every constraint imposed by a
+  fractional value will also be satisfied exactly once. This property is
+  shared with systems that are based on linear logic; other systems must
+  impose ad hoc constraints to ensure negative values are used exactly
+  once. 
+\item In contrast to all the work that takes continuations as primitive
+  entities of negative types, we view continuations as a derived notion that
+  combines a demand for a value with constraints on how this value will be
+  used to proceed with the evaluation (to the closest delimiter or to the end
+  of the program). In other words, we view a continuation as a non-elementary
+  notion that combines the negative types to demand a value and the
+  fractional types to explain how this value will be used to continue the
+  evaluation. As a consequence, the previously observed duality between
+  values and continuations can be teased into two dualities: a duality
+  between values flowing in one direction or the other and a duality between
+  aggregate values composing and decomposing into smaller values. Arguably
+  each of the dualities is more natural than a duality that maps regular
+  values to a conflated notion of negative and fractional types, and hence
+  requires notions like ``additive pairs'' and ``multiplicative sums.''
+\end{itemize}
 

### 2012-03-07 rational.tex
small sentence
intro and reorganize

@@
 \end{array}\]
-
 Although seemingly pointless, this algebraic proof corresponds, in our model,
@@
   and services to be brought together: it is sufficient to have a promise
-  that the money will be paid at some point, as long as the infrastructure
-  can be trusted that such promises will be fulfilled. Computationally, this
-  means that negative types allow us to break the dependency between the
-  producer and consumer and let each compute independently.
+  that the demand for resources will be somehow satisfied, as long as the
+  infrastructure can be trusted with such promises. Computationally, this
+  means that negative types allow us to break the dependency between
+  producers and consumers and let each compute independently.
 \end{itemize}
@@
 \item We extend {{langRev}} our reversible programming language of type
-  isomorphisms~\cite{rc2011,infeffects} with a notion of negative types, that
-  satisfies the isomorphism $a + (-a) = 0$. The semantics of this extension
-  is expressed by having a ``dual'' evaluator that reverses the flow of
-  execution for negative values. 
+  isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
+  of negative types, that satisfies the isomorphism $a + (-a) = 0$. The
+  semantics of this extension is expressed by having a ``dual'' evaluator
+  that reverses the flow of execution for negative
+  values. (Sec.~\ref{sec:neg})
 \item We independently extend {{langRev}} with a notion of fractional types,
@@
   mechanism to model and resolve the constraints introduced by the fractional
-  types.
+  types. (Sec.~\ref{sec:frac})
 \item We combine the above two extensions into one extension that includes
@@
   and intuitive isomorphisms that are satisfied in the mathematical field of
-  rational numbers.
+  rational numbers. (Sec.~\ref{sec:rat})
 \item We develop programming intuition and argue that negative and fractional
-  types ought to be part of the vocabulary of every programmer.
+  types ought to be part of the vocabulary of every
+  programmer. (Sec.~\ref{sec:prog})
 \item We relate our notions of negative and fractional types to previous work
-  on continuations. We analyze and relate two apparently unrelated lines of
-  work: the first pioneered by
-  Filinski~\cite{Filinski:1989:DCI:648332.755574} relates continuations to
-  negative types; the second~\cite{Bernardi:2010:CSL:1749618.1749689} relates
-  continuations to the fractional types of the Lambek-Grishin calculus.
+  on continuations. Briefly, we argue that conventional continuations
+  conflate negative and fractional components. This observation allows us to
+  relate two apparently unrelated lines of work: the first pioneered by
+  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
+  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
+  relating continuations to the fractional types of the Lambek-Grishin
+  calculus. (Sec.~\ref{sec:related})
 \end{itemize}
@@
 \section{ {{langRev}} }
+\label{sec:pi}
 
@@
 \section{Fractional Types}
+\label{sec:frac}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semantics}
+\section{Rational Types}
+\label{sec:rat}
 
@@
 \section{Advanced Programming Examples}
+\label{sec:prog}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Field of Algebraic Numbers}
+\section{Conclusion}
+\label{sec:conc}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+Say we have not considered recursion in this paper.
 

### 2012-03-07 rational.tex
sec2

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ {{langRev}} }
+\section{The Core Reversible Language: {{langRev}} }
 \label{sec:pi}
 
-The presentation here differs from \cite{infeffects} in the sense that
-we have introduced the {{0}} type here, thereby completing the monoid
-{{(0,+)}} and we present a small-step operational semantics.
-
+We review our reversible language {{langRev}}: the presentation in this
+section differs from the one in our previous paper~\cite{infeffects} in two
+aspects. First, we add the empty type $0$ which is necessary to express the
+additive duality. Second, instead of explaining evaluation using a natural
+semantics, we give a small-step operational semantics that is more
+appropriate for the connections with continuations explored in this paper.
+
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Syntax}
+
+The set of types includes the empty type {{0}}, the unit type {{1}}, sum
+types {{b1+b2}}, and products types {{b1*b2}}. The set of values {{v}}
+includes {{()}} which is the only value of type {{1}}, {{left v}} and {{right v}} 
+which inject {{v}} into a sum type, and {{(v1,v2)}} which builds a
+value of product type. There are no values of type {{0}}:
 %subcode{bnf} include main
@@
 
+The combinators of {{langRev}} are witnesses to the following type
+isomorphisms: 
 %subcode{bnf} include main
@@
 %dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-
+Each line of the above table introduces one or two combinators that witness
+the isomorphism in the middle. Collectively the isomorphisms state that the
+structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
+of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
+multiplication distributes over addition. The isomorphisms are extended to
+form a congruence relation by adding the following constructors that witness
+the equivalence and compatible closure:
 %subcode{proof} include main
@@
 
+To summarize, the syntax of {{langRev}} is given as follows.
+
 \begin{definition}{(Syntax of {{langRev}})}
@@
 %subcode{bnf} include main
-% value types, b ::= 1 | b+b | b*b 
+% value types, b ::= 0 | 1 | b+b | b*b 
 % values, v ::= () | left v | right v | (v,v) 
@@
 % comb.~types, t ::= b <-> b
-% iso ::= swap+ | assocl+ | assocr+ 
+% iso ::= zeroe | zeroi 
+%     &|& swap+ | assocl+ | assocr+ 
 %     &|& identl* | identr* 
 %     &|& swap* | assocl* | assocr* 
-%     &|& dist | factor 
+%     &|& dist0 | factor0 | dist | factor 
 % comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Graphical Language}
+
+The syntactic notation described above is often obscure. Following the
+tradition established for monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
+graphical language that conveys the intuitive semantics of the language
+(which is formalized in the next section). The general idea of the graphical
+notation is that combinators are modeled by ``wiring diagrams'' or
+``circuits''; values are modeled as ``particles'' that may appear on the
+wires; and evaluation is modeled by the flow of particles along the wires. We
+now proceed to give the graphical notation for each combinator:
+
+\begin{itemize}
+\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
+  is simply represented as a wire labeled by its type {{b}}. In more complex
+  diagrams, if the type of a wire is obvious from the context, it may be
+  omitted.
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire.pdf}
+}
+\end{center}
+
+\item The product type {{b1*b2}} may be represented both as one wire labeled
+  {{b1*b2}} or by two parallel wires labeled {{b1}} and {{b2}}. Both
+  representations maybe used interchangeably.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
+\includegraphics{diagrams/thesis/product-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
+\includegraphics{diagrams/thesis/product-two-wires.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Sum types may similarly be represented by one wire or using using
+  parallel wires with a {{+}} operator between them.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
+\includegraphics{diagrams/thesis/sum-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
+\includegraphics{diagrams/thesis/sum-two-wires.pdf}
+}
+\end{center}
+\end{multicols}
+
+%% \item
+%% When representing complex types like {{(b1*b2)+b3}} some visual
+%% grouping of the wires may be done to aid readability. The exact type
+%% however will always be clarified by the context of the diagram.
+
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
+%% }
+%% \end{center}
+
+\item Associativity is implicit in the graphical language. Three parallel
+  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
+\includegraphics{diagrams/thesis/assoc.pdf}
+}
+\end{center}
+
+\item Commutativity is represented by crisscrossing wires.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
+\includegraphics{diagrams/thesis/swap_times.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
+\includegraphics{diagrams/thesis/swap_plus.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Identity morphisms and their duals ({{uniti, unite, zeroi}} and
+  {{zeroe}}) are represented as follows:
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
+\includegraphics{diagrams/thesis/uniti.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
+\includegraphics{diagrams/thesis/unite.pdf}
+}
+\end{center}  
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
+\includegraphics{diagrams/thesis/zeroi.pdf}
+}
+\end{center}
+\columnbreak
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
+\includegraphics{diagrams/thesis/zeroe.pdf}
+}
+\end{center}
+\end{multicols}
+Since the monoidal units can be freely introduced and eliminated, in
+many diagrams they are omitted and dealt with explicitly when they
+are of special interest.
+
+\item Finally, distributivity and factoring are represented using the dual
+  boxes shown below:
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/distrib-crop.pdf]
+\includegraphics{diagrams/thesis/dist.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/factor-crop.pdf]
+\includegraphics{diagrams/thesis/factor.pdf}
+}
+\end{center}
+\end{multicols}
+
+\end{itemize}
+
+\noindent 
+\textit{Example.} As an example, the following combinator is represented by
+the given diagram:
+
+{{c : b * (1+1) <-> b + b}}
+
+{{c = swap* (;) dist (;) (identl* (+) identl*)}}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
+\includegraphics{diagrams/thesis/example1.pdf}
+}
+\end{center}
+
 %%%%%%%%%%%%%%%%%%%
@@
 
-%subcode{bnf} include main
-% Combinators, c = iso | c (;) c | c (*) c | c (+) c
-% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
-%                  &|& LeftTimes cc c v | RightTimes c v cc 
-%                  &|& LeftPlus cc c | RightPlus c cc 
-% Values, v = () | (v, v) | L v | R v
-%
-% Machine states = <c, v, cc> | {[c, v, cc]}
-% Start state = <c, v, []> 
-% Stop State = {[c, v, []]}
-
+The semantics of the primitive combinators is given by the following
+single-step reductions. Since there are no values of type {{0}}, the rules
+omit some of the impossible cases:
 \begin{scriptsize}
@@
 
+For the non-primitive combinators, we express the semantics using evaluation
+contexts which are used to represent the ``rest of the computation.'' More
+precisely, to calculate the result of applying combinator $c$ to value $v$,
+we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
+transitions of the machine (see below) are unusual in the sense that they
+always maintain full information about {{c}}, although this information may
+be split in the first and last component of the machine. The {{v}} component
+of the machine is the one that captures the current value as it is flowing in
+the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
+transitions based on the structure of {{c}}. When the evaluation of {{c}} is
+complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
+restored.
+
+This is a mess. Perhaps give a small example and explain things based on the
+example before giving the full transitions.
+
+%subcode{bnf} include main
+% Combinator Contexts, C = [] | Fst C c | Snd c C 
+%                  &|& LeftT C c v | RightT c v C 
+%                  &|& LeftP C c | RightP c C 
+% Machine states = <c, v, C> | {[c, v, C]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+
 %subcode{opsem} include main
 %! columnStyle = rclr
-% <iso, v, cc> &|-->& {[iso, iso(v), cc]}
-% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2>
-% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc>
-% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]}
-% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2>
-% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]}
-% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc>
-% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]}
-% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2>
-% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc>
-% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]}
+% <iso, v, C> &|-->& {[iso, iso(v), C]}
+% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2>
+% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C>
+% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]}
+% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2>
+% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]}
+% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C>
+% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]}
+% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2>
+% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C>
+% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{A Graphical Language for {{langRev}} }
-
-Combinators of {{langRev}} can be written using the the operators described
-in definition \ref{def:langRev} or via a graphical language similar in spirit
-to those developed for GoI~\cite{DBLP:conf/popl/Mackie95} and for category
-theory \cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
-conventions and shorthands we will describe here, the wiring diagrams are a
-complete and equivalent to the operator based (syntactic) description of
-programs.  {{langRev}} combinators expressed in this graphical language looks
-like a ``wiring diagrams''. Value take the form of ``particles'' that flow
-along the wires. Computation is expressed by the flow of particles.
-
-\begin{itemize}
-\item 
-The simplest sort of diagram is the {{id : b <-> b}} combinator which
-is simply represented as a wire labeled by its type {{b}}. In more
-complex diagrams, if the type of a wire is obvious from the context,
-it may be omitted.
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
-\includegraphics{diagrams/thesis/b-wire.pdf}
-}
-\end{center}
-
-\item
-The product type {{b1*b2}} may be represented both as one wire
-labeled {{b1*b2}} or by two parallel wires labeled {{b1}} and
-{{b2}}. Both representations maybe used interchangeably.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
-\includegraphics{diagrams/thesis/product-one-wire.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
-\includegraphics{diagrams/thesis/product-two-wires.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item
-Sum types may similarly be represented using using parallel wires with
-a {{+}} operator between them.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
-\includegraphics{diagrams/thesis/sum-one-wire.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
-\includegraphics{diagrams/thesis/sum-two-wires.pdf}
-}
-\end{center}
-\end{multicols}
-
-%% \item
-%% When representing complex types like {{(b1*b2)+b3}} some visual
-%% grouping of the wires may be done to aid readability. The exact type
-%% however will always be clarified by the context of the diagram.
-
-%% \begin{center}
-%% \scalebox{0.95}{
-%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
-%% }
-%% \end{center}
-
-\item
-Associativity is entirely implicit in the graphical language. Three
-parallel wires maybe inferred as {{b1*(b2*b3)}} or {{(b1*b2)*b3}},
-based on the context, much like handling of associativity in the
-graphical representations of categories
-\cite{springerlink:10.1007/978-3-642-12821-94}.
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
-\includegraphics{diagrams/thesis/assoc.pdf}
-}
-\end{center}
-
-\item
-Commutativity is represented by crisscrossing wires.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
-\includegraphics{diagrams/thesis/swap_times.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
-\includegraphics{diagrams/thesis/swap_plus.pdf}
-}
-\end{center}
-
-\end{multicols}
-
-\item
-Identity morphisms and their duals ({{uniti, unite, zeroi}} and
-{{zeroe}}) are represented as follows:
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
-\includegraphics{diagrams/thesis/uniti.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
-\includegraphics{diagrams/thesis/unite.pdf}
-}
-\end{center}  
-\end{multicols}
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
-\includegraphics{diagrams/thesis/zeroi.pdf}
-}
-\end{center}
-
-\columnbreak
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
-\includegraphics{diagrams/thesis/zeroe.pdf}
-}
-\end{center}
-\end{multicols}
-
-Since the monoidal units can be freely introduced and eliminated, in
-many diagrams they are omitted and dealt with explicitly when they
-are of special interest.
-
-\item
-Finally, distributivity and factoring are represented using the dual
-boxes shown below:
-
-\begin{multicols}{2}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/distrib-crop.pdf]
-\includegraphics{diagrams/thesis/dist.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/factor-crop.pdf]
-\includegraphics{diagrams/thesis/factor.pdf}
-}
-\end{center}
-\end{multicols}
-
-\end{itemize}
-
-\noindent 
-\textit{Example.} As an example consider the wiring diagram
-of the combinator {{c}} below:
-
-{{c : b * (1+1) <-> b + b}}
-
-{{c = swap* (;) dist (;) (identl* (+) identl*)}}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
-\includegraphics{diagrams/thesis/example1.pdf}
-}
-\end{center}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Negative Types}
-\label{sub:neg}
+\label{sec:neg}
 
@@
 presence of {{eta+}} and {{eps+}} in the following sense: a value of
-type {{R v : -b + b}} flowing into an {{eps+}} on the lower wire
-switches into a value {{L (-v): -b + b}} that flows backwards on the
+type {{right v : -b + b}} flowing into an {{eps+}} on the lower wire
+switches into a value {{left (-v): -b + b}} that flows backwards on the
 upper wire. The inversion of direction is captured by the negative
@@
 %subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v
+% Values, v = () | (v, v) | left v | right v
 % Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
@@
 %subcode{opsem} include main
-% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
-% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
-% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
-% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
+% D[P|(left v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
+% D[P| (right v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
+% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (left v)| F] 
+% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (right v)| F]
 
@@
 %subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v | -v
+% Values, v = () | (v, v) | left v | right v | -v
 % Isomorphisms, iso &=& ... | eta | eps
@@
 %subcode{opsem} include main
-% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
-% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
+% D[P| (right v) eps| F]      &|-->&  D{dagger}[F | eps (left (-v))| P]
+% D[P| (left (-v)) eps| F]      &|-->&  D{dagger}[F | eps (right v)| P]
 

### 2012-03-07 rational.tex
semantics

--- 2012-03-07 rational.tex
@@
 %! (+) = +
+%! LeftPlus = L^+
+%! RightPlus = R^+
+%! LeftTimes = L^{\times}
+%! RightTimes = R^{\times}
 
@@
 %! |- = \vdash
+%! <><> = \approx
 %! ==> = \Longrightarrow
@@
 
+\paragraph{Primitive Isomorphsims.}
+Previously the reductions of the primitive isomorphisms were specified
+in the following form.
+
+{{ iso v_{input pattern} |--> v_{output pattern} }}
+
+\noindent
+In the precense of unification, we rewrite them to accept the incoming
+substitution and extend it with the appropriate unification.
+
+{{ iso s v' |--> (v_{output pattern}, s[v' <><> v_{input pattern}]) }}
+
+\paragraph{Extending the evaulator with unification.}
+%subcode{opsem} include main
+%! columnStyle = rclr
+%! fwd = \triangleright
+% <iso, v, cc, s>_{fwd} &|-->& {[iso, v', cc, s']}_{fwd}
+% & & where iso s v |--> (v', s')
+% <c1(;)c2, v, cc, s>_{fwd} &|-->& <c1, v, Fst cc c2, s>_{fwd}
+% {[c1, v, Fst cc c2, s]}_{fwd} &|-->& <c2, v, Snd c1 cc, s>_{fwd}
+% {[c2, v, Snd c1 cc, s]}_{fwd} &|-->& {[ c1(;)c2, v, cc,s ]}_{fwd}
+% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c1, v, LeftPlus cc c2, s[v' <><> left v]>_{fwd}
+% {[ c1, v, LeftPlus cc c2, s ]}_{fwd} &|-->& {[c1 (+) c2, left v, cc,s ]}_{fwd}
+% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c2, v, RightPlus c1 cc, s[v' <><> right v]>_{fwd}
+% {[ c2, v, RightPlus c1 cc,s ]}_{fwd} &|-->& {[c1 (+) c2, right v, cc,s ]}_{fwd}
+% <c1(*)c2, v', cc, s>_{fwd} &|-->& <c1, v1, LeftTimes cc c2 v2, s'>_{fwd}
+% & & where s' = s[v <><> (v1, v2)]
+% {[ c1, v1, LeftTimes cc c2 v2,s ]}_{fwd} &|-->& <c2, v2, RightTimes c1 v1 cc, s>_{fwd}
+% {[ c2, v2, RightTimes c1 v1 cc,s ]}_{fwd} &|-->& {[ c1 (*) c2, (v1, v2), cc,s ]}_{fwd}
+
+\paragraph{Backward evaluator.}
+%subcode{opsem} include main
+%! columnStyle = rclr
+%! bck = \triangleleft
+% {[iso, v, cc, s]}_{bck} &|-->& <iso, v', cc, s'>_{bck}
+% & & where iso^{dagger} s v |--> (v', s')
+% <c1, v, Fst cc c2, s>_{bck} &|-->& <c1(;)c2, v, cc, s>_{bck}
+% <c2, v, Snd c1 cc, s>_{bck} &|-->& {[c1, v, Fst cc c2, s]}_{bck}
+% {[ c1(;)c2, v, cc, s ]}_{bck} &|-->& {[c2, v, Snd c1 cc, s]}_{bck}
+% <c1, v, LeftPlus cc c2,s >_{bck} &|-->& <c1(+)c2,left v, cc>_{bck}
+% {[c1 (+) c2, v', cc, s ]}_{bck} &|-->& {[ c1, v, LeftPlus cc c2, s[v' <><> left v] ]}_{bck}
+% <c2, v, RightPlus c1 cc, s>_{bck} &|-->& <c1(+)c2, right v, cc>_{bck}
+% {[c1 (+) c2, v', cc,s ]}_{bck} &|-->& {[ c2, v, RightPlus c1 cc, s[v' <><> right v] ]}_{bck}
+% <c1, v1, LeftTimes cc c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), cc, s>_{bck}
+% <c2, v2, RightTimes c1 v1 cc, s>_{bck} &|-->& {[ c1, v1, LeftTimes cc c2 v2, s ]}_{bck}
+% {[ c1 (*) c2, v, cc, s ]}_{bck} &|-->& {[ c2, v2, RightTimes c1 v1 cc, S' ]}_{bck}
+% && where s' = s[v <><> (v1, v2)]
+
 \begin{comment}
@@
 \end{center}
+
+MUST TALK ABOUT AND CITE ABRAMSKY AND COECKE'S TR ABOUT QUANTUM
+COMPUTING.
 

### 2012-03-07 rational.tex
semantics

@@
 
+We start with the semantics for {{langRev}} defined in Sec
+\ref{sec:pi}, which we will systematically rewrite with unification
+and then present a ``reverse'' evaluator written in the same style. We
+shall then combine these into the semantics for {{langRevEE}}.
+
+TODO. define substitutions and unification.
+
 \paragraph{Primitive Isomorphsims.}
@@
 \paragraph{Extending the evaulator with unification.} 
+We have rewriten the semantics of Sec. \ref{sec:pi} with
+unification. The behaviour of the intepreter has not changed.
+
 %subcode{opsem} include main
@@
 \paragraph{Backward evaluator.}
+Here we have written a ``backward'' evaluator in the same style --
+i.e. evaluation {{c v}} in this evaluator is equivalent to evaluating
+{{C^{dagger} v}} in the forward evaluator.
+
 %subcode{opsem} include main
@@
 
-\begin{comment}
-%%%%%%%%%%%%%%%%%%
-\subsection{Syntax}
-
-%subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
-%
-% Sequential Contexts, P, F = [] | P:c
-% Parallel Contexts, D = [] 
-%                     &|& D:(P| [](+)c| F)
-%                     &|& D:(P| c(+)[]| F)
-%                     &|& D:(P| [] (*) c v| F)
-%                     &|& D:(P| [] (*) v c| F)
-%                     &|& D:(P| c v (*)  []| F)
-%                     &|& D:(P| v c (*) []| F)
-%
-%
-% Machine States  = D[P| c v| F] | D[P| v c| F]
-% Start State  = [][[]| v c| []]
-% Stop State  = [][P| c v| []]
-
-Combinator reconstruction, {{P[c]}}:
+\subsection{Semantics of {{langRevEE}} }
+
+\begin{enumerate}
+\item 
+To add multiplicative duality, we add the following rules to the set
+of primitive isomorphisms:
+
+\begin{enumerate}
+\item {{eta* s () |--> ((inv(v), v), s)}}
+  where {{v}} is a fresh logic variable.
+
+\item {{eps* s v |--> (1, s[v <><> (inv(v'),v') ])}}
+  where {{v'}} is a fresh logic variable.
+
+\end{enumerate}
+
+\item To add negative types we add the following rules to the
+  reductions above:
+
+  \begin{enumerate}
+  \item
+
 %subcode{opsem} include main
-% [](c) '= c
-% P:c'(c) '= P(c'(;)c)
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection{Operational Semantics}
-
-The small step semantics present for {{langRev}} below work
-symmetrically for forward and backward evaluation.
-
-\begin{itemize}
+%! columnStyle = rclr
+%! fwd = \triangleright
+%! bck = \triangleleft
+% <eps+, v, cc, s>_{fwd} |--> <eps+, left (-v'), cc, s[v <><> right v]>_{bck}
+% <eps+, v, cc, s>_{fwd} |--> <eps+, right v', cc, s[v <><> left (-v')]>_{bck}
+
     \item
-Basic reduction of an isomorphism. Note that the evaluation leaves the
-adjoint of the combinator behind. This will become important when we
-reverse the direction of computation.
-%% %subcode{opsem} include main
-%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
-
 %subcode{opsem} include main
-% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
-
-\item
-Sequencing involves pushing and popping from the Future and Past
-continuations:
-%subcode{opsem} include main
-% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
-% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
-
-\item
-Parallel composition, captures the current Future and Past and extends
-the parallel context.
-%subcode{opsem} include main
-% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
-% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
-% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
-% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]
-
-\item
-Similarly for products:
-%subcode{opsem} include main
-% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
-% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
-% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
-
-and symmetrical rules for evaluation along the second branch. 
-%subcode{opsem} include main
-% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
-% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
-
-The later two rules will be relevant only for reverse execution.
-
-\end{itemize}
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrevEE}
-{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
-\end{proposition}
-
-\begin{proposition}[Groupoid]
-\label{prop:groupoidEE}
-{{langRevEE}} is a groupoid. 
-\end{proposition}
-
-\begin{proposition}
-\label{prop:categoryEE}
-{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
-\end{proposition}
-
-Note: We use `Monoidal Closed' in the sense of Compact Closed. While
-the category has an multiplicative tensor that distributes over an
-additive tensor, it does not have injections or projections and hence
-does not have categorical sums and products.
-
-%%%%%%%%%%%%%%%%%%%%
-\subsection{Rules for {{eta}} and {{eps}} }
-
-The operation {{eps}} reverses the direction of a particle by
-reversing the world.
-
-Note that we deviate from the categorical definition of {{eta}} and
-{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
-choice of {{eta}}. This however does not affect us, because we deal
-with a symmetric category.
-
-\begin{itemize}
-\item Grammar
-%subcode{bnf} include main
-% Values, v = () | (v, v) | L v | R v | -v
-% Isomorphisms, iso &=& ... | eta | eps
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
-
-\item
-Type judgment.
-%subcode{opsem} include main
-% eta &: 0 <-> (-b) + b :& eps
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-
-\item
-Operational Semantics.
-%subcode{opsem} include main
-% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]
-% D[P| (L (-v)) eps| F]      &|-->&  D{dagger}[F | eps (R v)| P]
-
-Note: there is NO reduction rule for {{eta}}.
-
-\item
-The adjoint of a parallel context is defined to be:
-%subcode{opsem} include main
-% []{dagger} '= []
-% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
-% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
-% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
-% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
-% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
-% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
-
-\end{itemize}
-
-\end{comment}
+%! columnStyle = rclr
+%! fwd = \triangleright
+%! bck = \triangleleft
+% <eta+, v, cc, s>_{bck} |--> <eta+, left (-v'), cc, s[v <><> right v]>_{fwd}
+% <eta+, v, cc, s>_{bck} |--> <eta+, right v', cc, s[v <><> left (-v')]>_{fwd}
+
+  \end{enumerate}
+
+\end{enumerate}
 

### 2012-03-07 rational.tex

--- 2012-03-07 rational.tex
@@
 
-\begin{comment}
-%%%%%%%%%%%%%%%%%%
-\subsection{Syntax}
-
-%subcode{bnf} include main
-% Values, v = () | (v, v) | left v | right v
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
-%
-% Sequential Contexts, P, F = [] | P:c
-% Parallel Contexts, D = [] 
-%                     &|& D:(P| [](+)c| F)
-%                     &|& D:(P| c(+)[]| F)
-%                     &|& D:(P| [] (*) c v| F)
-%                     &|& D:(P| [] (*) v c| F)
-%                     &|& D:(P| c v (*)  []| F)
-%                     &|& D:(P| v c (*) []| F)
-%
-%
-% Machine States  = D[P| c v| F] | D[P| v c| F]
-% Start State  = [][[]| v c| []]
-% Stop State  = [][P| c v| []]
-
-Combinator reconstruction, {{P[c]}}:
+We start with the semantics for {{langRev}} defined in Sec
+\ref{sec:pi}, which we will systematically rewrite with unification
+and then present a ``reverse'' evaluator written in the same style. We
+shall then combine these into the semantics for {{langRevEE}}.
+
+\paragraph{Primitive Isomorphsims.}
+Previously the reductions of the primitive isomorphisms were specified
+in the following form.
+
+{{ iso v_{input pattern} |--> v_{output pattern} }}
+
+\noindent
+In the precense of unification, we rewrite them to accept the incoming
+substitution and extend it with the appropriate unification.
+
+{{ iso s v' |--> (v_{output pattern}, s[v' <><> v_{input pattern}]) }}
+
+\paragraph{Extending the evaulator with unification.} 
+We have rewriten the semantics of Sec. \ref{sec:pi} with
+unification. The behaviour of the intepreter has not changed.
+
 %subcode{opsem} include main
-% [](c) '= c
-% P:c'(c) '= P(c'(;)c)
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection{Operational Semantics}
-
-The small step semantics present for {{langRev}} below work
-symmetrically for forward and backward evaluation.
-
-\begin{itemize}
+%! columnStyle = rclr
+%! fwd = \triangleright
+% <iso, v, cc, s>_{fwd} &|-->& {[iso, v', cc, s']}_{fwd}
+% & & where iso s v |--> (v', s')
+% <c1(;)c2, v, cc, s>_{fwd} &|-->& <c1, v, Fst cc c2, s>_{fwd}
+% {[c1, v, Fst cc c2, s]}_{fwd} &|-->& <c2, v, Snd c1 cc, s>_{fwd}
+% {[c2, v, Snd c1 cc, s]}_{fwd} &|-->& {[ c1(;)c2, v, cc,s ]}_{fwd}
+% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c1, v, LeftPlus cc c2, s[v' <><> left v]>_{fwd}
+% {[ c1, v, LeftPlus cc c2, s ]}_{fwd} &|-->& {[c1 (+) c2, left v, cc,s ]}_{fwd}
+% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c2, v, RightPlus c1 cc, s[v' <><> right v]>_{fwd}
+% {[ c2, v, RightPlus c1 cc,s ]}_{fwd} &|-->& {[c1 (+) c2, right v, cc,s ]}_{fwd}
+% <c1(*)c2, v', cc, s>_{fwd} &|-->& <c1, v1, LeftTimes cc c2 v2, s'>_{fwd}
+% & & where s' = s[v <><> (v1, v2)]
+% {[ c1, v1, LeftTimes cc c2 v2,s ]}_{fwd} &|-->& <c2, v2, RightTimes c1 v1 cc, s>_{fwd}
+% {[ c2, v2, RightTimes c1 v1 cc,s ]}_{fwd} &|-->& {[ c1 (*) c2, (v1, v2), cc,s ]}_{fwd}
+
+\paragraph{Backward evaluator.}
+Here we have written a ``backward'' evaluator in the same style --
+i.e. evaluation {{c v}} in this evaluator is equivalent to evaluating
+{{C^{dagger} v}} in the forward evaluator.
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+%! bck = \triangleleft
+% {[iso, v, cc, s]}_{bck} &|-->& <iso, v', cc, s'>_{bck}
+% & & where iso^{dagger} s v |--> (v', s')
+% <c1, v, Fst cc c2, s>_{bck} &|-->& <c1(;)c2, v, cc, s>_{bck}
+% <c2, v, Snd c1 cc, s>_{bck} &|-->& {[c1, v, Fst cc c2, s]}_{bck}
+% {[ c1(;)c2, v, cc, s ]}_{bck} &|-->& {[c2, v, Snd c1 cc, s]}_{bck}
+% <c1, v, LeftPlus cc c2,s >_{bck} &|-->& <c1(+)c2,left v, cc>_{bck}
+% {[c1 (+) c2, v', cc, s ]}_{bck} &|-->& {[ c1, v, LeftPlus cc c2, s[v' <><> left v] ]}_{bck}
+% <c2, v, RightPlus c1 cc, s>_{bck} &|-->& <c1(+)c2, right v, cc>_{bck}
+% {[c1 (+) c2, v', cc,s ]}_{bck} &|-->& {[ c2, v, RightPlus c1 cc, s[v' <><> right v] ]}_{bck}
+% <c1, v1, LeftTimes cc c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), cc, s>_{bck}
+% <c2, v2, RightTimes c1 v1 cc, s>_{bck} &|-->& {[ c1, v1, LeftTimes cc c2 v2, s ]}_{bck}
+% {[ c1 (*) c2, v, cc, s ]}_{bck} &|-->& {[ c2, v2, RightTimes c1 v1 cc, S' ]}_{bck}
+% && where s' = s[v <><> (v1, v2)]
+
+\subsection{Semantics of {{langRevEE}} }
+
+\begin{enumerate}
 \item 
-Basic reduction of an isomorphism. Note that the evaluation leaves the
-adjoint of the combinator behind. This will become important when we
-reverse the direction of computation.
-%% %subcode{opsem} include main
-%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>
+To add multiplicative duality, we add the following rules to the set
+of primitive isomorphisms:
+
+\begin{enumerate}
+\item {{eta* s () |--> ((inv(v), v), s)}}
+  where {{v}} is a fresh logic variable.
+
+\item {{eps* s v |--> (1, s[v <><> (inv(v'),v') ])}}
+  where {{v'}} is a fresh logic variable.
+
+\end{enumerate}
+
+\item To add negative types we add the following rules to the
+  reductions above:
+
+  \begin{enumerate}
+  \item
 
 %subcode{opsem} include main
-% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]
+%! columnStyle = rclr
+%! fwd = \triangleright
+%! bck = \triangleleft
+% <eps+, v, cc, s>_{fwd} |--> <eps+, left (-v'), cc, s[v <><> right v]>_{bck}
+% <eps+, v, cc, s>_{fwd} |--> <eps+, right v', cc, s[v <><> left (-v')]>_{bck}
 
     \item
-Sequencing involves pushing and popping from the Future and Past
-continuations:
 %subcode{opsem} include main
-% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
-% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]
-
-\item
-Parallel composition, captures the current Future and Past and extends
-the parallel context.
-%subcode{opsem} include main
-% D[P|(left v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
-% D[P| (right v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
-% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (left v)| F] 
-% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (right v)| F]
-
-\item
-Similarly for products:
-%subcode{opsem} include main
-% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
-% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
-% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]
-
-and symmetrical rules for evaluation along the second branch. 
-%subcode{opsem} include main
-% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
-% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]
-
-The later two rules will be relevant only for reverse execution.
-
-\end{itemize}
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrevEE}
-{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
-\end{proposition}
-
-\begin{proposition}[Groupoid]
-\label{prop:groupoidEE}
-{{langRevEE}} is a groupoid. 
-\end{proposition}
-
-\begin{proposition}
-\label{prop:categoryEE}
-{{langRev}} is a Dagger Symmetric Monoidal Closed Category.
-\end{proposition}
-
-Note: We use `Monoidal Closed' in the sense of Compact Closed. While
-the category has an multiplicative tensor that distributes over an
-additive tensor, it does not have injections or projections and hence
-does not have categorical sums and products.
-
-%%%%%%%%%%%%%%%%%%%%
-\subsection{Rules for {{eta}} and {{eps}} }
-
-The operation {{eps}} reverses the direction of a particle by
-reversing the world.
-
-Note that we deviate from the categorical definition of {{eta}} and
-{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
-choice of {{eta}}. This however does not affect us, because we deal
-with a symmetric category.
-
-\begin{itemize}
-\item Grammar
-%subcode{bnf} include main
-% Values, v = () | (v, v) | left v | right v | -v
-% Isomorphisms, iso &=& ... | eta | eps
-% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c
-
-\item
-Type judgment.
-%subcode{opsem} include main
-% eta &: 0 <-> (-b) + b :& eps
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-
-\item
-Operational Semantics.
-%subcode{opsem} include main
-% D[P| (right v) eps| F]      &|-->&  D{dagger}[F | eps (left (-v))| P]
-% D[P| (left (-v)) eps| F]      &|-->&  D{dagger}[F | eps (right v)| P]
-
-Note: there is NO reduction rule for {{eta}}.
-
-\item
-The adjoint of a parallel context is defined to be:
-%subcode{opsem} include main
-% []{dagger} '= []
-% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
-% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
-% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
-% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
-% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
-% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)
-
-\end{itemize}
-
-\end{comment}
+%! columnStyle = rclr
+%! fwd = \triangleright
+%! bck = \triangleleft
+% <eta+, v, cc, s>_{bck} |--> <eta+, left (-v'), cc, s[v <><> right v]>_{fwd}
+% <eta+, v, cc, s>_{bck} |--> <eta+, right v', cc, s[v <><> left (-v')]>_{fwd}
+
+  \end{enumerate}
+
+\end{enumerate}
 

### 2012-03-07 rational.tex
_

@@
 %! (+) = +
-%! LeftPlus = L^+
-%! RightPlus = R^+
-%! LeftTimes = L^{\times}
-%! RightTimes = R^{\times}
+%! LeftP = L^+
+%! RightP = R^+
+%! LeftT = L^{\times}
+%! RightT = R^{\times}
 
@@
 
-We start with the semantics for {{langRev}} defined in Sec
-\ref{sec:pi}, which we will systematically rewrite with unification
-and then present a ``reverse'' evaluator written in the same style. We
-shall then combine these into the semantics for {{langRevEE}}.
+In this section we develop the semantics for {{langRev}} extended with
+negative and fractional types, {{langRevEE}}.  We start with the
+semantics for {{langRev}} defined in Sec \ref{sec:pi}, which we will
+systematically rewrite until we have the desired {{langRevEE}}
+sematics. Broadly this construction takes two steps
+
+\begin{enumerate}
+\item We rewrite the semantics in Sec \ref{sec:pi} using unification
+  of the values instead of direct pattern matching. This gives us the
+  neccessary infrastructure for fractional types, as we will see.
+
+\item We write a reverse interpreter by reversing the direction of the
+  `{{|-->}}' reductions. This gives us the neccessary infrastructure for
+  negative types.
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Unification and Logic Programming}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Extending {{langRev}} Semantics}
 
@@
 %! fwd = \triangleright
-% <iso, v, cc, s>_{fwd} &|-->& {[iso, v', cc, s']}_{fwd}
+% <iso, v, C, s>_{fwd} &|-->& {[iso, v', C, s']}_{fwd}
 % & & where iso s v |--> (v', s')
-% <c1(;)c2, v, cc, s>_{fwd} &|-->& <c1, v, Fst cc c2, s>_{fwd}
-% {[c1, v, Fst cc c2, s]}_{fwd} &|-->& <c2, v, Snd c1 cc, s>_{fwd}
-% {[c2, v, Snd c1 cc, s]}_{fwd} &|-->& {[ c1(;)c2, v, cc,s ]}_{fwd}
-% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c1, v, LeftPlus cc c2, s[v' <><> left v]>_{fwd}
-% {[ c1, v, LeftPlus cc c2, s ]}_{fwd} &|-->& {[c1 (+) c2, left v, cc,s ]}_{fwd}
-% <c1(+)c2, v', cc, s>_{fwd} &|-->& <c2, v, RightPlus c1 cc, s[v' <><> right v]>_{fwd}
-% {[ c2, v, RightPlus c1 cc,s ]}_{fwd} &|-->& {[c1 (+) c2, right v, cc,s ]}_{fwd}
-% <c1(*)c2, v', cc, s>_{fwd} &|-->& <c1, v1, LeftTimes cc c2 v2, s'>_{fwd}
+% <c1(;)c2, v, C, s>_{fwd} &|-->& <c1, v, Fst C c2, s>_{fwd}
+% {[c1, v, Fst C c2, s]}_{fwd} &|-->& <c2, v, Snd c1 C, s>_{fwd}
+% {[c2, v, Snd c1 C, s]}_{fwd} &|-->& {[ c1(;)c2, v, C,s ]}_{fwd}
+% <c1(+)c2, v', C, s>_{fwd} &|-->& <c1, v, LeftP C c2, s[v' <><> left v]>_{fwd}
+% {[ c1, v, LeftP C c2, s ]}_{fwd} &|-->& {[c1 (+) c2, left v, C,s ]}_{fwd}
+% <c1(+)c2, v', C, s>_{fwd} &|-->& <c2, v, RightP c1 C, s[v' <><> right v]>_{fwd}
+% {[ c2, v, RightP c1 C,s ]}_{fwd} &|-->& {[c1 (+) c2, right v, C,s ]}_{fwd}
+% <c1(*)c2, v', C, s>_{fwd} &|-->& <c1, v1, LeftT C c2 v2, s'>_{fwd}
 % & & where s' = s[v <><> (v1, v2)]
-% {[ c1, v1, LeftTimes cc c2 v2,s ]}_{fwd} &|-->& <c2, v2, RightTimes c1 v1 cc, s>_{fwd}
-% {[ c2, v2, RightTimes c1 v1 cc,s ]}_{fwd} &|-->& {[ c1 (*) c2, (v1, v2), cc,s ]}_{fwd}
+% {[ c1, v1, LeftT C c2 v2,s ]}_{fwd} &|-->& <c2, v2, RightT c1 v1 C, s>_{fwd}
+% {[ c2, v2, RightT c1 v1 C,s ]}_{fwd} &|-->& {[ c1 (*) c2, (v1, v2), C,s ]}_{fwd}
 
@@
 %! bck = \triangleleft
-% {[iso, v, cc, s]}_{bck} &|-->& <iso, v', cc, s'>_{bck}
+% {[iso, v, C, s]}_{bck} &|-->& <iso, v', C, s'>_{bck}
 % & & where iso^{dagger} s v |--> (v', s')
-% <c1, v, Fst cc c2, s>_{bck} &|-->& <c1(;)c2, v, cc, s>_{bck}
-% <c2, v, Snd c1 cc, s>_{bck} &|-->& {[c1, v, Fst cc c2, s]}_{bck}
-% {[ c1(;)c2, v, cc, s ]}_{bck} &|-->& {[c2, v, Snd c1 cc, s]}_{bck}
-% <c1, v, LeftPlus cc c2,s >_{bck} &|-->& <c1(+)c2,left v, cc>_{bck}
-% {[c1 (+) c2, v', cc, s ]}_{bck} &|-->& {[ c1, v, LeftPlus cc c2, s[v' <><> left v] ]}_{bck}
-% <c2, v, RightPlus c1 cc, s>_{bck} &|-->& <c1(+)c2, right v, cc>_{bck}
-% {[c1 (+) c2, v', cc,s ]}_{bck} &|-->& {[ c2, v, RightPlus c1 cc, s[v' <><> right v] ]}_{bck}
-% <c1, v1, LeftTimes cc c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), cc, s>_{bck}
-% <c2, v2, RightTimes c1 v1 cc, s>_{bck} &|-->& {[ c1, v1, LeftTimes cc c2 v2, s ]}_{bck}
-% {[ c1 (*) c2, v, cc, s ]}_{bck} &|-->& {[ c2, v2, RightTimes c1 v1 cc, S' ]}_{bck}
+% <c1, v, Fst C c2, s>_{bck} &|-->& <c1(;)c2, v, C, s>_{bck}
+% <c2, v, Snd c1 C, s>_{bck} &|-->& {[c1, v, Fst C c2, s]}_{bck}
+% {[ c1(;)c2, v, C, s ]}_{bck} &|-->& {[c2, v, Snd c1 C, s]}_{bck}
+% <c1, v, LeftP C c2,s >_{bck} &|-->& <c1(+)c2,left v, C>_{bck}
+% {[c1 (+) c2, v', C, s ]}_{bck} &|-->& {[ c1, v, LeftP C c2, s[v' <><> left v] ]}_{bck}
+% <c2, v, RightP c1 C, s>_{bck} &|-->& <c1(+)c2, right v, C>_{bck}
+% {[c1 (+) c2, v', C,s ]}_{bck} &|-->& {[ c2, v, RightP c1 C, s[v' <><> right v] ]}_{bck}
+% <c1, v1, LeftT C c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), C, s>_{bck}
+% <c2, v2, RightT c1 v1 C, s>_{bck} &|-->& {[ c1, v1, LeftT C c2 v2, s ]}_{bck}
+% {[ c1 (*) c2, v, C, s ]}_{bck} &|-->& {[ c2, v2, RightT c1 v1 C, S' ]}_{bck}
 % && where s' = s[v <><> (v1, v2)]
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics of {{langRevEE}} }
+
+The previous section has defined all the pieces needed to define the
+semantics of {{langRevEE}} and we piece them together here.
 
@@
 %! bck = \triangleleft
-% <eps+, v, cc, s>_{fwd} |--> <eps+, left (-v'), cc, s[v <><> right v]>_{bck}
-% <eps+, v, cc, s>_{fwd} |--> <eps+, right v', cc, s[v <><> left (-v')]>_{bck}
+% <eps+, v, C, s>_{fwd} |--> <eps+, left (-v'), C, s[v <><> right v]>_{bck}
+% <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}
 
@@
 %! bck = \triangleleft
-% <eta+, v, cc, s>_{bck} |--> <eta+, left (-v'), cc, s[v <><> right v]>_{fwd}
-% <eta+, v, cc, s>_{bck} |--> <eta+, right v', cc, s[v <><> left (-v')]>_{fwd}
+% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v]>_{fwd}
+% <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}
 

### 2012-03-07 rational.tex
intro

@@
 manipulation relating a natural number $a$ to itself:
-\[\begin{array}{rclclclclcl}
-a &=& a + 0 &=& a + ((-a) + a) \\
-  &=& (a + (-a)) + a &=& 0 + a &=& a
+\[\begin{array}{rclll}
+&& a & \mbox{initial~$a$} & (0) \\
+&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
+&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
+&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
+&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
+&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
 \end{array}\]
@@
 to an isomorphism with a non-trivial and interesting computational
-interpretation.  On the left we have a producer that can supply a value of
-type $a$ and on the right we have a consumer that demands a value of type
-$a$. For the purposes of the example, think of the producer as someone having
-\$20.00 and the consumer as someone demanding \$20.00. It is certainly
-possible to directly connect the producer and consumer using the algebraic
-identity $a=a$: such a connection would be computationally modeled by the
-identity function as expected. However, the algebraic proof above implements
-this connection in a much more interesting way: the transformation $0 = (-a)
-+ a$ is used at the consumer site to create money and a corresponding
-debt. The consumer gets the money immediately and the debt travels
-``backwards'' towards the producer where it is paid.  In other words, this
-computation models something similar to a credit card transaction. We discuss
-this example in detail in Sec.~\ref{sub:neg} and only focus on two important
-points here:
+interpretation. Consider the starting $a$ as being supplied by a producer of
+values and the final $a$ as being demanded by a consumer. For the purposes of
+the example, think of the producer as someone having \$20.00 and the consumer
+as someone demanding \$20.00. It is certainly possible to directly connect
+the producer and consumer using the algebraic identity $a=a$: such a
+connection would be computationally modeled by the identity function as
+expected. However, the algebraic proof above implements this connection in a
+much more interesting way. As Sec.~\ref{sec:neg} explains in detail, the
+flow of the computation is as follows:
+\begin{itemize}
+\item We start at line (0) with \$20.00; 
+\item We proceed to line (1) with the same \$20.00 but tagged as being in the
+  left summand of the type $a+0$; we indicate this value as {{left 20}}
+\item We continue to line (2) with the same value {{left 20}};
+\item At line (3), the tag on the \$20.00 changes to indicate that it is
+  in the left-left summand, i.e., the value is now {{left (left 20)}};
+\item At line (4), we find ourselves needing to produce a value of type 0
+  which is impossible; this signals the beginning of a reverse execution
+  which sends us back to line (3) with a value {{left (right 20)}};
+\item Execution continues in reverse to line (2) with the value 
+  {{right (left 20)}};
+\item At line (1) we find ourselves again facing an empty type so we reverse
+  execution again; we go to line (2) with a value {{right (right 20)}};
+\item We proceed to line (3) and (4) with the value {{right 20}};
+\item We finally reach line 5 with the value {{20}}.
+\end{itemize}
+
+Collecting the various events that happen at the same lines, we can
+reorganize the above computation as follows:
+\begin{itemize}
+\item At line 0, we have the value {{20}} flowing forward;
+\item At line 1, we have the value {{left 20}} flowing forward;
+\item At line 2, we have the value {{left 20}} flowing forward or the value
+  {{right (left 20)}} flowing backwards or the value {{right (right 20)}}
+  flowing forward;
+\item At line 3, we have the value {{left (left 20)}} flowing forward or the
+  value {{left (right 20)}} flowing backward or the value {{right 20}}
+  flowing forward;
+\item At line 4, we have the value {{right 20}} flowing forward;
+\item At line 5, we have the value {{20}} flowing forward.
+\end{itemize}
+
+Use the above to argue that we have something like a credit card transaction!
+
+the transformation $0 = (-a) + a$ is used at the
+consumer site to create money and a corresponding debt. The consumer gets the
+money immediately and the debt travels ``backwards'' towards the producer
+where it is paid.  In other words, this computation models something similar
+to a credit card transaction. We discuss this example in detail in
+Sec.~\ref{sub:neg} and only focus on two important points here:
 \begin{itemize}

### 2012-03-07 rational.tex
intro

@@
 Although seemingly pointless, this algebraic proof corresponds, in our model,
-to an isomorphism with a non-trivial and interesting computational
-interpretation. Consider the starting $a$ as being supplied by a producer of
-values and the final $a$ as being demanded by a consumer. For the purposes of
-the example, think of the producer as someone having \$20.00 and the consumer
-as someone demanding \$20.00. It is certainly possible to directly connect
-the producer and consumer using the algebraic identity $a=a$: such a
-connection would be computationally modeled by the identity function as
-expected. However, the algebraic proof above implements this connection in a
-much more interesting way. As Sec.~\ref{sec:neg} explains in detail, the
-flow of the computation is as follows:
+to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
+computational interpretation. Consider the starting value of type $a$ as
+being supplied by a producer of values and the final value of type $a$ as
+being demanded by a consumer. For the purposes of the example, think of the
+producer as someone having \$20.00 and the consumer as someone demanding
+\$20.00. It is certainly possible to directly connect the producer and
+consumer using the algebraic identity $a=a$: such a connection would be
+computationally modeled by the identity function as expected. However, the
+algebraic proof above implements this connection in a much more interesting
+way. As the semantics of Sec.~\ref{sec:rat} formalizes, the flow of 
+computation is as follows:
 \begin{itemize}
@@
 \end{itemize}
-
-Collecting the various events that happen at the same lines, we can
-reorganize the above computation as follows:
-\begin{itemize}
-\item At line 0, we have the value {{20}} flowing forward;
-\item At line 1, we have the value {{left 20}} flowing forward;
-\item At line 2, we have the value {{left 20}} flowing forward or the value
-  {{right (left 20)}} flowing backwards or the value {{right (right 20)}}
-  flowing forward;
-\item At line 3, we have the value {{left (left 20)}} flowing forward or the
-  value {{left (right 20)}} flowing backward or the value {{right 20}}
-  flowing forward;
-\item At line 4, we have the value {{right 20}} flowing forward;
-\item At line 5, we have the value {{20}} flowing forward.
-\end{itemize}
-
-Use the above to argue that we have something like a credit card transaction!
-
-the transformation $0 = (-a) + a$ is used at the
-consumer site to create money and a corresponding debt. The consumer gets the
-money immediately and the debt travels ``backwards'' towards the producer
-where it is paid.  In other words, this computation models something similar
-to a credit card transaction. We discuss this example in detail in
-Sec.~\ref{sub:neg} and only focus on two important points here:
+The beauty of the above computation is that it can be realized with a
+completely different concurrent schedule as follows. \emph{Simultaneously}
+the execution begins at line (0) with the value {{20}}, at line (2) with the
+value {{right (right 20)}} flowing forward and the value {{right (left 20)}}
+flowing backwards, and at line (3) with the value {{left (left 20)}} flowing
+forward and the value {{left (right 20)}} flowing backward. We now have three
+independent subcomputations proceeding in parallel: the positive value
+created at line (2) which can flow to the consumer; the negative value
+created at line (3) which can pay for the debt created at line (2); and the
+initial value from line (0) which can flow to the required forward-flowing
+value needed at line (3). At some intuitive level, one might understand the
+situation as modeling a credit card payment in which money is created at line
+(2) for the consumer together with a corresponding debt. The producer's bank
+pays for the debt at line (3) and waits for the actual money to arrive from
+the producer.
+
+There are at least two fundamental points about this example that must be
+emphasized:
 \begin{itemize}

### 2012-03-08 rational.tex
_

@@
 values that are not solutions to the specified isomorphic constraint
-{{f}}. The operation {{f}} acts on some inputs and its adjoint
-reconstructs the inputs from the outputs. While the output of {{f}}
-may contain many components, its must contain a boolean indicating if
-the input satisfy {{f}}. The special output wire is used to `control'
-the negation of a boolean `control wire'. If the inputs do not satisfy
-{{f}}, the control wire is negated.
+{{f}}. The contraint {{f}} checks the inputs to see if they match a
+constraint -- in the case of the SAT solver, {{f}} would be the
+circuit expressing the boolean expression to be satisfied.  While the
+output of {{f}} may contain many components, its must contain a
+boolean indicating if the input satisfy {{f}}. The operation {{f}}
+acts on some inputs and its adjoint, {{f^{dagger} }}, reconstructs the
+inputs from the outputs.  The `trick' is the following: the special
+output wire is used to `control' the negation of a boolean `control
+wire'.  If the inputs do not satisfy {{f}}, the control wire is
+negated.
 

### 2012-03-08 rational.tex
sat

@@
 \begin{center}
-\scalebox{1.5}{
+\scalebox{1.2}{
   \includegraphics{diagrams/not_trace.pdf}
@@
 like the closed-loop `not' in the previous construction.
+
+The final detail we have to be concerned with is the handling of the
+heap and garbage of {{f}}. Any boolean expression {{bool^n -> bool}}
+can be compiled into the isomorphism {{f:bool^h*bool^n<->bool^g*bool}}
+where the extra bits {{bool^h}} and {{bool^g}} are the heap and
+garbage. Constructing such an isomorphism has been detailed before
+\cite{Toffoli:1980,infeffects}.  However the computation of {{f}} is
+meaningful {{iff}} the heap {{bool^h}} is initialized appropriately. We
+can ensure that the heap has the appropriate initial value by checking
+the heap and negating a second control wire, if the values do not
+match.
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/sat2.pdf}
+}
+\end{center}
 

### 2012-03-08 rational.tex
logical rev. prop.

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Graphical Language}
@@
 
-%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics}
@@
 \end{proposition}
+
+\roshan{We really have to rethink this prop in the precense of
+  negatives and fractionals. With negatives, the program can actually
+  end at the begining of the circuit -- i.e. the program {{c:b1<->b2}}
+  can stop with a value of type {{b1}}. With fractionals, there can be
+  several values of type {{b2}} produced for a value of type {{b1}}. }
 

### 2012-03-08 rational.tex
diagram1 for negatives
diagrams
sat

@@
 
-FIGURE
+\begin{center}
+  \includegraphics{diagrams/algebra1.pdf}
+\end{center}
+
+\roshan{I chose to draw the igure with b instead of a, since b is what
+  we use for the types. Should this be a? Also I chose to make the
+  swap and 0 wires, explicit.Is this what you intended? }
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%
+\begin{center}
+  \includegraphics{diagrams/algebra2.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Multiplicative Duality in {{langRev}} }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Advanced Programming Examples}
+\section{SAT Solver in {{langRevEE}} }
 \label{sec:prog}
@@
 %%%%%%%%%%%%%%%%%%%
-\subsection{SAT solver}
+% \subsection{SAT solver}
 \label{sec:sat-solver}
@@
   \includegraphics{diagrams/sat2.pdf}
+}
+\end{center}
+
+\roshan{Explain how the dotted part is implemented.}
+
+Let us call this construction which maps inputs, heap and control
+wires to input, heap and control wires as {{sat_f}}. The required
+contraint solver is then constructed as shown:
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/sat3.pdf}
 }

### 2012-03-08 rational.tex
particles

@@
 \subsection{Syntax} 
+\label{sec:pi-syntax}
 
@@
 
-The syntactic notation described above is often obscure. Following the
-tradition established for monoidal
-categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
-graphical language that conveys the intuitive semantics of the language
-(which is formalized in the next section). The general idea of the graphical
-notation is that combinators are modeled by ``wiring diagrams'' or
-``circuits''; values are modeled as ``particles'' that may appear on the
-wires; and evaluation is modeled by the flow of particles along the wires. We
-now proceed to give the graphical notation for each combinator:
+The syntactic notation described in Sec \ref{sec:pi-syntax} is often
+obscure and hard to read.  Following the tradition established for
+monoidal categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we
+present a graphical language that conveys the intuitive semantics of
+the language (which is formalized in the next section).  For the rest
+of the paper {{langRev}} programs will be presented graphically.
+
+The general idea of the graphical notation is that combinators are
+modeled by ``wiring diagrams'' or ``circuits''; values are modeled as
+``particles'' that may appear on the wires; and evaluation is modeled
+by the flow of particles along the wires.
+
+%% We now proceed to give the graphical notation for each combinator:
 
 \begin{itemize}
-\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
-  is simply represented as a wire labeled by its type {{b}}. In more complex
-  diagrams, if the type of a wire is obvious from the context, it may be
-  omitted.
+\item The simplest sort of diagram is the {{id : b <-> b}} combinator
+  which is simply represented as a wire labeled by its type {{b}}, as
+  shown on the left. In more complex diagrams, if the type of a wire
+  is obvious from the context, it may be omitted.
+
+  \begin{multicols}{2}
 \begin{center}
@@
 \end{center}
+
+DIAGRAM    
+  \end{multicols}
+
+When tracing a computation, one might imagine a value {{v}} of of type
+{{b}} on the wire, as shown on the right.
 
@@
 
+In the case of products represented by a pair of wires, when tracing
+execution using particles, one should think of one value/particle on
+each wire.
+
 \item Sum types may similarly be represented by one wire or using using
@@
 \end{multicols}
+
+When tracing the execution of two additive wires, a value can reside
+on only one of the two wires.
 
@@
 
+The crisscrossing wires have the right semantics in that a value on
+the top wire, ends up on the bottom wire and vice versa.
+
 \item Identity morphisms and their duals ({{uniti, unite, zeroi}} and
@@
 \end{multicols}
-Since the monoidal units can be freely introduced and eliminated, in
-many diagrams they are omitted and dealt with explicitly when they
-are of special interest.
+Note that since there is no value of type 0, there can be no particle
+on a wire of type {{0}}.  Since the monoidal units can be freely
+introduced and eliminated, in many diagrams they are omitted and dealt
+with explicitly when they are of special interest.
 

### 2012-03-08 rational.tex
intro

@@
 %! RightT = R^{\times}
+%! alpha = \alpha
 
@@
 to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
-computational interpretation. Consider the starting value of type $a$ as
-being supplied by a producer of values and the final value of type $a$ as
-being demanded by a consumer. For the purposes of the example, think of the
-producer as someone having \$20.00 and the consumer as someone demanding
-\$20.00. It is certainly possible to directly connect the producer and
-consumer using the algebraic identity $a=a$: such a connection would be
-computationally modeled by the identity function as expected. However, the
-algebraic proof above implements this connection in a much more interesting
-way. As the semantics of Sec.~\ref{sec:rat} formalizes, the flow of 
-computation is as follows:
+computational interpretation. The witness for this isomorphism is a
+computation that takes a value of type $a$ (say \$20.00) and eventually
+produces the same \$20.00 value as its output. As the semantics of
+Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
+proceed as follows:
 \begin{itemize}
@@
 \item We continue to line (2) with the same value {{left 20}};
-\item At line (3), the tag on the \$20.00 changes to indicate that it is
-  in the left-left summand, i.e., the value is now {{left (left 20)}};
+\item At line (3), as a result of re-association the tag on the \$20.00
+  changes to indicate that it is in the left-left summand, i.e., the value is
+  now {{left (left 20)}};
 \item At line (4), we find ourselves needing to produce a value of type 0
@@
 \end{itemize}
+
+Now consider the following algebraic manipulation (assuming $a \neq 0$):
+\[\begin{array}{rclll}
+&& a & \mbox{initial~$a$} & (0) \\
+&=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
+&=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
+&=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
+&=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
+&=& a & \mbox{$1$~is~identity~for~$*$} & (5)
+\end{array}\]
+As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
+{{a <-> a}} isomorphism commputes as follows:
+\begin{itemize}
+\item We start at line (0) with the value {{20}};
+\item We proceed to line (1) with the value {{(20,())}};
+\item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
+  $\alpha$ is essentially a logic variable;
+\item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
+\item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
+  forces the unification $\alpha=20$, i.e., the value becomes
+  {{((20,1/20),20)}} which simplifies to {{((),20)}};
+\item We reach line (5) with the value {{20}}.
+\end{itemize}
+
+\amr{the above is a boring trace; we need to convey why it is exciting but we
+  are not agreeing on the interpretation.}
+
 The beauty of the above computation is that it can be realized with a

### 2012-03-08 rational.tex
semantics of pi

@@
 
-For the non-primitive combinators, we express the semantics using evaluation
-contexts which are used to represent the ``rest of the computation.'' More
-precisely, to calculate the result of applying combinator $c$ to value $v$,
-we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
-transitions of the machine (see below) are unusual in the sense that they
-always maintain full information about {{c}}, although this information may
-be split in the first and last component of the machine. The {{v}} component
-of the machine is the one that captures the current value as it is flowing in
-the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
-transitions based on the structure of {{c}}. When the evaluation of {{c}} is
-complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
-restored.
-
-This is a mess. Perhaps give a small example and explain things based on the
-example before giving the full transitions.
+The reductions for the primitive isomorphisms are exactly the same as
+have been persented before \cite{infeffects}. The reductions for the
+closure combinators are however presented in a small-setp operational
+style. The following are the definition for evaluation contexts and
+machine states:
 
@@
 
+The reduction rules below essentially track the flow of particles
+through their circuit. The start machine state, {{<c,v,[]>}}, denotes
+the particle/value {{v}} about to be evluated by the circuit
+{{c}}. The end machine state, {{[c, v, [] ]}}, denotes the situation
+where the particle {{v}} has exited the circuit {{c}}.
+
+%% For the non-primitive combinators, we express the semantics using evaluation
+%% contexts which are used to represent the ``rest of the computation.'' More
+%% precisely, to calculate the result of applying combinator $c$ to value $v$,
+%% we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
+%% transitions of the machine (see below) are unusual in the sense that they
+%% always maintain full information about {{c}}, although this information may
+%% be split in the first and last component of the machine. The {{v}} component
+%% of the machine is the one that captures the current value as it is flowing in
+%% the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
+%% transitions based on the structure of {{c}}. When the evaluation of {{c}} is
+%% complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
+%% restored.
+
+%% This is a mess. Perhaps give a small example and explain things based on the
+%% example before giving the full transitions.
+
 %subcode{opsem} include main
 %! columnStyle = rclr
-% <iso, v, C> &|-->& {[iso, iso(v), C]}
-% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2>
-% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C>
-% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]}
-% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2>
-% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]}
-% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C>
-% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]}
-% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2>
-% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C>
-% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]}
+% <iso, v, C> &|-->& {[iso, v', C]} & (1)
+% & & where iso v |--> v' & (2) 
+% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2> & (3) 
+% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C> & (4) 
+% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]} & (5) 
+% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2> & (6) 
+% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]} & (7)
+% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C> & (8)
+% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]} & (9)
+% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2> & (10)
+% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C> & (11) 
+% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (12)
+
+The rule (1) describes evaluation by a primitive isomorphism. Rules
+(2), (3) and (4) deal with sequential evaluation. Rule (2) says that
+for the value to flow through the sequence {{c1 (;) c2}} we should
+first flow through {{c1}} with {{c2}} pending in the context and rule
+(3) says the value that exits from {{c1}} should proceed to flow
+through {{c2}}. Rule (4) says that if the value {{v}} has exited
+{{c2}} as well, then it is equivalent to having exited from
+{{c1(;)c2}}.
+
+Rules (5) to (9) deal with {{c1 (+) c2}} in the same way. In the case
+of sums, the shape of the value {{left v}}/{{right v}} determines
+whether path {{c1}} or path {{c2}} is taken. Rules (10), (11) and (12)
+deal with {{c1 (*) c2}} similarly. In the case of products the value
+should have the form {{(v1, v2)}} where {{v1}} flows through {{c1}}
+and {{v2}} flows through {{c2}}. Both these paths are entirely
+independant of each other and we could chosen to evaluate either
+first, or evaluate both in parallel. In this presentation we have
+chosen to follow {{c1}] first, but this choice is entirely artificial.
+
+Finally, we end by noting the abstract machine here is interesting is
+that it is a reversible machine. We could change reductions `{{|-->}}'
+to run backwards `{{<--|}}', thereby computing the start state from the
+stop state. We will take advantange of this property in
+Sec. \ref{sec:rat}.
 

### 2012-03-08 rational.tex
_

@@
 The reductions for the primitive isomorphisms are exactly the same as
-have been persented before \cite{infeffects}. The reductions for the
-closure combinators are however presented in a small-setp operational
+have been presented before \cite{infeffects}. The reductions for the
+closure combinators are however presented in a small-step operational
 style. The following are the definition for evaluation contexts and
@@
 through their circuit. The start machine state, {{<c,v,[]>}}, denotes
-the particle/value {{v}} about to be evluated by the circuit
+the particle/value {{v}} about to be evaluated by the circuit
 {{c}}. The end machine state, {{[c, v, [] ]}}, denotes the situation
@@
 and {{v2}} flows through {{c2}}. Both these paths are entirely
-independant of each other and we could chosen to evaluate either
+independent of each other and we could chosen to evaluate either
 first, or evaluate both in parallel. In this presentation we have
@@
 that it is a reversible machine. We could change reductions `{{|-->}}'
-to run backwards `{{<--|}}', thereby computing the start state from the
-stop state. We will take advantange of this property in
+to run backwards `{{<--|}}', thereby computing the start state from
+the stop state. We will take advantage of this property in
 Sec. \ref{sec:rat}.
@@
 
-\roshan{We really have to rethink this prop in the precense of
+\roshan{We really have to rethink this prop in the presence of
   negatives and fractionals. With negatives, the program can actually
-  end at the begining of the circuit -- i.e. the program {{c:b1<->b2}}
+  end at the beginning of the circuit -- i.e. the program {{c:b1<->b2}}
   can stop with a value of type {{b1}}. With fractionals, there can be
@@
 
-\roshan{I chose to draw the igure with b instead of a, since b is what
+\roshan{I chose to draw the figure with b instead of a, since b is what
   we use for the types. Should this be a? Also I chose to make the

### 2012-03-08 rational.tex
_

@@
 % <iso, v, C> &|-->& {[iso, v', C]} & (1)
-% & & where iso v |--> v' & (2) 
-% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2> & (3) 
-% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C> & (4) 
-% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]} & (5) 
-% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2> & (6) 
-% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]} & (7)
-% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C> & (8)
-% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]} & (9)
-% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2> & (10)
-% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C> & (11) 
-% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (12)
+% & & where iso v |--> v' &
+% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2> & (2) 
+% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C> & (3) 
+% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]} & (4) 
+% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2> & (5) 
+% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]} & (6)
+% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C> & (7)
+% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]} & (8)
+% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2> & (9)
+% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C> & (10) 
+% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (11)
 
@@
 (2), (3) and (4) deal with sequential evaluation. Rule (2) says that
-for the value to flow through the sequence {{c1 (;) c2}} we should
-first flow through {{c1}} with {{c2}} pending in the context and rule
-(3) says the value that exits from {{c1}} should proceed to flow
-through {{c2}}. Rule (4) says that if the value {{v}} has exited
-{{c2}} as well, then it is equivalent to having exited from
-{{c1(;)c2}}.
-
-Rules (5) to (9) deal with {{c1 (+) c2}} in the same way. In the case
+for the value {{v}} to flow through the sequence {{c1 (;) c2}}, it
+should first flow through {{c1}} with {{c2}} pending in the context
+({{Fst C c2}}). Rule (3) says the value {{v}} that exits from {{c1}}
+should proceed to flow through {{c2}}. Rule (4) says that if the value
+{{v}} has exited {{c2}} as well, then it is equivalent to having
+exited from {{c1(;)c2}}.
+
+Rules (5) to (8) deal with {{c1 (+) c2}} in the same way. In the case
 of sums, the shape of the value {{left v}}/{{right v}} determines
-whether path {{c1}} or path {{c2}} is taken. Rules (10), (11) and (12)
+whether path {{c1}} or path {{c2}} is taken. Rules (9), (10) and (11)
 deal with {{c1 (*) c2}} similarly. In the case of products the value

### 2012-03-08 rational.tex
_

@@
 Rules (5) to (8) deal with {{c1 (+) c2}} in the same way. In the case
-of sums, the shape of the value {{left v}}/{{right v}} determines
-whether path {{c1}} or path {{c2}} is taken. Rules (9), (10) and (11)
-deal with {{c1 (*) c2}} similarly. In the case of products the value
-should have the form {{(v1, v2)}} where {{v1}} flows through {{c1}}
-and {{v2}} flows through {{c2}}. Both these paths are entirely
-independent of each other and we could chosen to evaluate either
-first, or evaluate both in parallel. In this presentation we have
-chosen to follow {{c1}] first, but this choice is entirely artificial.
-
-Finally, we end by noting the abstract machine here is interesting is
-that it is a reversible machine. We could change reductions `{{|-->}}'
-to run backwards `{{<--|}}', thereby computing the start state from
-the stop state. We will take advantage of this property in
+of sums, the shape of the value,i.e. is it is tagged with {{left}} or
+{{right}}, determines whether path {{c1}} or path {{c2}} is
+taken. Rules (9), (10) and (11) deal with {{c1 (*) c2}} similarly. In
+the case of products the value should have the form {{(v1, v2)}} where
+{{v1}} flows through {{c1}} and {{v2}} flows through {{c2}}. Both
+these paths are entirely independent of each other and we could
+evaluate either first, or evaluate both in parallel. In this
+presentation we have chosen to follow {{c1}} first, but this choice is
+entirely arbitrary.
+
+The interesting thing about the semantics is that it represents a
+reversible abstract machine. In other words, we can compute the start
+state from the stop state by changing the reductions `{{|-->}}' to run
+backwards `{{<--|}}'. We will take advantage of this property in
 Sec. \ref{sec:rat}.

### 2012-03-08 rational.tex
_
more particle based tracing diagrams

@@
 %%%%%%%%%%%%%%%%%%%%
-\subsection{Syntax} 
+\subsection{Syntax and Types} 
 \label{sec:pi-syntax}
@@
 present a graphical language that conveys the intuitive semantics of
-the language (which is formalized in the next section).  For the rest
-of the paper {{langRev}} programs will be presented graphically.
+the language (which is formalized in the next section).  Since, the
+graphical notation is easier to read an trace, for the rest of the
+paper {{langRev}} programs will be presented graphically.
 
@@
 
-DIAGRAM    
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire-value.pdf}
+}
+\end{center}
   \end{multicols}
@@
 
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
+}
+\end{center}
+\end{multicols}
+
 In the case of products represented by a pair of wires, when tracing
@@
 on only one of the two wires.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
+}
+\end{center}
+\end{multicols}
 

### 2012-03-08 rational.tex
_

@@
 As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
-{{a <-> a}} isomorphism commputes as follows:
+{{a <-> a}} isomorphism computes as follows:
 \begin{itemize}
@@
 ``particles'' that may appear on the wires; and evaluation is modeled
-by the flow of particles along the wires.
+by the flow of particles along the wires. In fact, when talking about
+{{langRev}} circuits, we will often use the words {{value}} and
+{{particle}] interchangeably.
 

### 2012-03-08 rational.tex
intro:

--- 2012-03-08 rational.tex
@@
 %! |--> = \mapsto
-%! <--| = \mapsfrom
 %! |- = \vdash
@@
 
+start with int times bool divided by bool: it is like an int but it can only
+be used as the second component of a pair whose first component is an int;
+gives you the power to say here is 20 but you must use it to buy books
+
 To get a feeling for the main idea, consider the following algebraic
@@
 computational interpretation. The witness for this isomorphism is a
-computation that takes a value of type $a$ (say \$20.00) and eventually
+computation that takes a value of type~$a$ (say \$20.00) and eventually
 produces the same \$20.00 value as its output. As the semantics of
@@
 
-Now consider the following algebraic manipulation (assuming $a \neq 0$):
-\[\begin{array}{rclll}
-&& a & \mbox{initial~$a$} & (0) \\
-&=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
-&=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
-&=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
-&=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
-&=& a & \mbox{$1$~is~identity~for~$*$} & (5)
-\end{array}\]
-As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
-{{a <-> a}} isomorphism commputes as follows:
-\begin{itemize}
-\item We start at line (0) with the value {{20}};
-\item We proceed to line (1) with the value {{(20,())}};
-\item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
-  $\alpha$ is essentially a logic variable;
-\item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
-\item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
-  forces the unification $\alpha=20$, i.e., the value becomes
-  {{((20,1/20),20)}} which simplifies to {{((),20)}};
-\item We reach line (5) with the value {{20}}.
-\end{itemize}
-
-\amr{the above is a boring trace; we need to convey why it is exciting but we
-  are not agreeing on the interpretation.}
+% Now consider the following algebraic manipulation (assuming $a \neq 0$):
+% \[\begin{array}{rclll}
+% && a & \mbox{initial~$a$} & (0) \\
+% &=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
+% &=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
+% &=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
+% &=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
+% &=& a & \mbox{$1$~is~identity~for~$*$} & (5)
+% \end{array}\]
+% As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
+% {{a <-> a}} isomorphism commputes as follows:
+% \begin{itemize}
+% \item We start at line (0) with the value {{20}};
+% \item We proceed to line (1) with the value {{(20,())}};
+% \item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
+%   $\alpha$ is essentially a logic variable;
+% \item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
+% \item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
+%   forces the unification $\alpha=20$, i.e., the value becomes
+%   {{((20,1/20),20)}} which simplifies to {{((),20)}};
+% \item We reach line (5) with the value {{20}}.
+% \end{itemize}
+
+% \amr{the above is a boring trace; we need to convey why it is exciting but we
+%   are not agreeing on the interpretation.}
 
@@
 
-The reductions for the primitive isomorphisms are exactly the same as
-have been presented before \cite{infeffects}. The reductions for the
-closure combinators are however presented in a small-step operational
-style. The following are the definition for evaluation contexts and
-machine states:
+For the non-primitive combinators, we express the semantics using evaluation
+contexts which are used to represent the ``rest of the computation.'' More
+precisely, to calculate the result of applying combinator $c$ to value $v$,
+we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
+transitions of the machine (see below) are unusual in the sense that they
+always maintain full information about {{c}}, although this information may
+be split in the first and last component of the machine. The {{v}} component
+of the machine is the one that captures the current value as it is flowing in
+the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
+transitions based on the structure of {{c}}. When the evaluation of {{c}} is
+complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
+restored.
+
+This is a mess. Perhaps give a small example and explain things based on the
+example before giving the full transitions.
 
@@
 
-The reduction rules below essentially track the flow of particles
-through their circuit. The start machine state, {{<c,v,[]>}}, denotes
-the particle/value {{v}} about to be evaluated by the circuit
-{{c}}. The end machine state, {{[c, v, [] ]}}, denotes the situation
-where the particle {{v}} has exited the circuit {{c}}.
-
-%% For the non-primitive combinators, we express the semantics using evaluation
-%% contexts which are used to represent the ``rest of the computation.'' More
-%% precisely, to calculate the result of applying combinator $c$ to value $v$,
-%% we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
-%% transitions of the machine (see below) are unusual in the sense that they
-%% always maintain full information about {{c}}, although this information may
-%% be split in the first and last component of the machine. The {{v}} component
-%% of the machine is the one that captures the current value as it is flowing in
-%% the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
-%% transitions based on the structure of {{c}}. When the evaluation of {{c}} is
-%% complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
-%% restored.
-
-%% This is a mess. Perhaps give a small example and explain things based on the
-%% example before giving the full transitions.
-
 %subcode{opsem} include main
 %! columnStyle = rclr
-% <iso, v, C> &|-->& {[iso, v', C]} & (1)
-% & & where iso v |--> v' &
-% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2> & (2) 
-% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C> & (3) 
-% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]} & (4) 
-% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2> & (5) 
-% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]} & (6)
-% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C> & (7)
-% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]} & (8)
-% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2> & (9)
-% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C> & (10) 
-% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (11)
-
-The rule (1) describes evaluation by a primitive isomorphism. Rules
-(2), (3) and (4) deal with sequential evaluation. Rule (2) says that
-for the value {{v}} to flow through the sequence {{c1 (;) c2}}, it
-should first flow through {{c1}} with {{c2}} pending in the context
-({{Fst C c2}}). Rule (3) says the value {{v}} that exits from {{c1}}
-should proceed to flow through {{c2}}. Rule (4) says that if the value
-{{v}} has exited {{c2}} as well, then it is equivalent to having
-exited from {{c1(;)c2}}.
-
-Rules (5) to (8) deal with {{c1 (+) c2}} in the same way. In the case
-of sums, the shape of the value,i.e. is it is tagged with {{left}} or
-{{right}}, determines whether path {{c1}} or path {{c2}} is
-taken. Rules (9), (10) and (11) deal with {{c1 (*) c2}} similarly. In
-the case of products the value should have the form {{(v1, v2)}} where
-{{v1}} flows through {{c1}} and {{v2}} flows through {{c2}}. Both
-these paths are entirely independent of each other and we could
-evaluate either first, or evaluate both in parallel. In this
-presentation we have chosen to follow {{c1}} first, but this choice is
-entirely arbitrary.
-
-The interesting thing about the semantics is that it represents a
-reversible abstract machine. In other words, we can compute the start
-state from the stop state by changing the reductions `{{|-->}}' to run
-backwards `{{<--|}}'. We will take advantage of this property in
-Sec. \ref{sec:rat}.
+% <iso, v, C> &|-->& {[iso, iso(v), C]}
+% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2>
+% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C>
+% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]}
+% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2>
+% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]}
+% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C>
+% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]}
+% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2>
+% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C>
+% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]}
 
@@
 
-\roshan{We really have to rethink this prop in the presence of
+\roshan{We really have to rethink this prop in the precense of
   negatives and fractionals. With negatives, the program can actually
-  end at the beginning of the circuit -- i.e. the program {{c:b1<->b2}}
+  end at the begining of the circuit -- i.e. the program {{c:b1<->b2}}
   can stop with a value of type {{b1}}. With fractionals, there can be
@@
 
-\roshan{I chose to draw the figure with b instead of a, since b is what
+\roshan{I chose to draw the igure with b instead of a, since b is what
   we use for the types. Should this be a? Also I chose to make the

### 2012-03-08 rational.tex

--- 2012-03-08 rational.tex
@@
 
-Now consider the following algebraic manipulation (assuming $a \neq 0$):
-\[\begin{array}{rclll}
-&& a & \mbox{initial~$a$} & (0) \\
-&=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
-&=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
-&=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
-&=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
-&=& a & \mbox{$1$~is~identity~for~$*$} & (5)
-\end{array}\]
-As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
-{{a <-> a}} isomorphism computes as follows:
-\begin{itemize}
-\item We start at line (0) with the value {{20}};
-\item We proceed to line (1) with the value {{(20,())}};
-\item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
-  $\alpha$ is essentially a logic variable;
-\item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
-\item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
-  forces the unification $\alpha=20$, i.e., the value becomes
-  {{((20,1/20),20)}} which simplifies to {{((),20)}};
-\item We reach line (5) with the value {{20}}.
-\end{itemize}
-
-\amr{the above is a boring trace; we need to convey why it is exciting but we
-  are not agreeing on the interpretation.}
+% Now consider the following algebraic manipulation (assuming $a \neq 0$):
+% \[\begin{array}{rclll}
+% && a & \mbox{initial~$a$} & (0) \\
+% &=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
+% &=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
+% &=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
+% &=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
+% &=& a & \mbox{$1$~is~identity~for~$*$} & (5)
+% \end{array}\]
+% As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
+% {{a <-> a}} isomorphism commputes as follows:
+% \begin{itemize}
+% \item We start at line (0) with the value {{20}};
+% \item We proceed to line (1) with the value {{(20,())}};
+% \item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
+%   $\alpha$ is essentially a logic variable;
+% \item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
+% \item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
+%   forces the unification $\alpha=20$, i.e., the value becomes
+%   {{((20,1/20),20)}} which simplifies to {{((),20)}};
+% \item We reach line (5) with the value {{20}}.
+% \end{itemize}
+% \amr{the above is a boring trace; we need to convey why it is exciting but we
+%   are not agreeing on the interpretation.}
 

### 2012-03-08 rational.tex
_

--- 2012-03-08 rational.tex
@@
 \end{array}\]
+\[\begin{array}{rlllllllll}
+ &  a & &       & &    & \mbox{initial~$a$} & (0) \\
+=&  a &+&       &0&    & \mbox{$0$~is~identity~for~$+$} & (1) \\
+=&  a &+& (-a   &+& a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
+=& (a &+& (-a)) &+& a  & \mbox{$+$~is~associative} & (3) \\
+=&    &0&       &+& a  & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
+=&    & &       & & a  & \mbox{$0$~is~identity~for~$+$} & (5)
+\end{array}\]
+
 Although seemingly pointless, this algebraic proof corresponds, in our model,

### 2012-03-08 rational.tex
negative space...

@@
 \[\begin{array}{rlllllllll}
- &  a & &       & &    & \mbox{initial~$a$} & (0) \\
-=&  a &+&       &0&    & \mbox{$0$~is~identity~for~$+$} & (1) \\
-=&  a &+& (-a   &+& a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
-=& (a &+& (-a)) &+& a  & \mbox{$+$~is~associative} & (3) \\
-=&    &0&       &+& a  & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
-=&    & &       & & a  & \mbox{$0$~is~identity~for~$+$} & (5)
+ \!&\!\!  a \!&\! \!&\!       \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
+=\!&\!\!  a \!&\!+\!&\!       \!&\!0\!&\!    \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (1) \\
+=\!&\!\!  a \!&\!+\!&\! (-a   \!&\!+\!&\! a) \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (2) \\
+=\!&\!\! (a \!&\!+\!&\! (-a)) \!&\!+\!&\! a  \!&\! \mbox{$+$~is~associative} \!&\! (3) \\
+=\!&\!\!    \!&\!0\!&\!       \!&\!+\!&\! a  \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (4) \\
+=\!&\!\!    \!&\! \!&\!       \!&\! \!&\! a  \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (5)
 \end{array}\]

### 2012-03-08 rational.tex

--- 2012-03-08 rational.tex
@@
 \end{array}\]
+\[\begin{array}{rlllllllll}
+ \!&\!\!  a \!&\! \!&\!       \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
+=\!&\!\!  a \!&\!+\!&\!       \!&\!0\!&\!    \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (1) \\
+=\!&\!\!  a \!&\!+\!&\! (-a   \!&\!+\!&\! a) \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (2) \\
+=\!&\!\! (a \!&\!+\!&\! (-a)) \!&\!+\!&\! a  \!&\! \mbox{$+$~is~associative} \!&\! (3) \\
+=\!&\!\!    \!&\!0\!&\!       \!&\!+\!&\! a  \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (4) \\
+=\!&\!\!    \!&\! \!&\!       \!&\! \!&\! a  \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (5)
+\end{array}\]
+
 Although seemingly pointless, this algebraic proof corresponds, in our model,

### 2012-03-09 rational.tex
algebra with wire superimposed

--- 2012-03-08 rational.tex
@@
 manipulation relating a natural number $a$ to itself:
-\[\begin{array}{rclll}
-&& a & \mbox{initial~$a$} & (0) \\
-&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
-&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
-&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
-&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
-&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
-\end{array}\]
-\[\begin{array}{rlllllllll}
- \!&\!\!  a \!&\! \!&\!       \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
-=\!&\!\!  a \!&\!+\!&\!       \!&\!0\!&\!    \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (1) \\
-=\!&\!\!  a \!&\!+\!&\! (-a   \!&\!+\!&\! a) \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (2) \\
-=\!&\!\! (a \!&\!+\!&\! (-a)) \!&\!+\!&\! a  \!&\! \mbox{$+$~is~associative} \!&\! (3) \\
-=\!&\!\!    \!&\!0\!&\!       \!&\!+\!&\! a  \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (4) \\
-=\!&\!\!    \!&\! \!&\!       \!&\! \!&\! a  \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (5)
-\end{array}\]
+%% \[\begin{array}{rclll}
+%% && a & \mbox{initial~$a$} & (0) \\
+%% &=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
+%% &=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
+%% &=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
+%% &=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
+%% &=& a & \mbox{$0$~is~identity~for~$+$} & (5)
+%% \end{array}\]
+%% \[\begin{array}{rlllllllll}
+%%  \!&\!\!  a \!&\! \!&\!       \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
+%% =\!&\!\!  a \!&\!+\!&\!       \!&\!0\!&\!    \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (1) \\
+%% =\!&\!\!  a \!&\!+\!&\! (-a   \!&\!+\!&\! a) \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (2) \\
+%% =\!&\!\! (a \!&\!+\!&\! (-a)) \!&\!+\!&\! a  \!&\! \mbox{$+$~is~associative} \!&\! (3) \\
+%% =\!&\!\!    \!&\!0\!&\!       \!&\!+\!&\! a  \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (4) \\
+%% =\!&\!\!    \!&\! \!&\!       \!&\! \!&\! a  \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (5)
+%% \end{array}\]
+
+\begin{center}
+\scalebox{1.1}{
+\includegraphics{diagrams/thesis/algebra-wire1.pdf}
+}
+\end{center}
 

### 2012-03-09 rational.tex
added wire over the multiplicative algebra

--- 2012-03-09 rational.tex
@@
 Now consider the following algebraic manipulation (assuming $a \neq 0$):
-\[\begin{array}{rclll}
-&& a & \mbox{initial~$a$} & (0) \\
-&=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
-&=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
-&=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
-&=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
-&=& a & \mbox{$1$~is~identity~for~$*$} & (5)
-\end{array}\]
+%% \[\begin{array}{rclll}
+%% && a & \mbox{initial~$a$} & (0) \\
+%% &=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
+%% &=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
+%% &=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
+%% &=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
+%% &=& a & \mbox{$1$~is~identity~for~$*$} & (5)
+%% \end{array}\]
+
+%% \[\begin{array}{rcllllll}
+%%  \!&\! a \!&\! \!&\!        \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
+%% =\!&\! a \!&\!*\!&\!        \!&\!1\!&\!    \!&\! \mbox{$1$~is~identity~for~$*$} \!&\! (1) \\
+%% =\!&\! a \!&\!*\!&\! ((1/a) \!&\!*\!&\! a) \!&\! \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} \!&\! (2) \\
+%% =\!&\!(a \!&\!*\!&\! (1/a)) \!&\!*\!&\! a  \!&\! \mbox{$*$~is~associative} \!&\! (3) \\
+%% =\!&\!   \!&\!1\!&\!        \!&\!*\!&\! a  \!&\! \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} \!&\! (4) \\
+%% =\!&\!   \!&\! \!&\!        \!&\! \!&\! a  \!&\! \mbox{$1$~is~identity~for~$*$} \!&\! (5)
+%% \end{array}\]
+
+\begin{center}
+\scalebox{1.1}{
+\includegraphics{diagrams/thesis/algebra-wire2.pdf}
+}
+\end{center}
+
 As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 

### 2012-03-09 rational.tex

@@
 \end{itemize}
-
-\amr{the above is a boring trace; we need to convey why it is exciting but we
-  are not agreeing on the interpretation.}
 

### 2012-03-09 rational.tex
error
mult constructions
_

@@
 appropriate for the connections with continuations explored in this paper.
+
+\roshan{Say somewhere that {{langRev}} corresponds to computing in the
+  commutative rig of whole numbers. More bluntly {{langRev}} is whole
+  numbers with addition and multiplication, while {{langRevEE}} is the
+  rational with addition and multiplication.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Rational Types}
+\section{Computing in the Field of Rationals : {{langRevEE}} }
 \label{sec:rat}
@@
 
-{{-(b1 + b2) <-> (b1) + (-b2)}}
+{{-(b1 + b2) <-> (-b1) + (-b2)}}
 
@@
 \subsection{Constructions Specific to Multiplication}
+
+\paragraph*{Distributing fractions over {{*}} }
+
+{{1/(b1*b2) <-> (1/b1) * (1/b2)}}
+
+\paragraph*{Multiplying Fractions}
+
+{{b1/b2 * b3/b4 <-> (b1*b3)/(b2*b4)}}
+
+\paragraph*{Lifting Computation intro fractions}
+
+{{b1 <-> b2 ==> 1/b1 <-> 1/b2}}
+
+\paragraph*{Adding the numerator}
+
+{{b1/b + b2/b <-> (b1+b2)/b}}
+
+\paragraph*{Fraction Addition}
+
+{{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
 

### 2012-03-09 rational.tex
some reminders and re-org
more shuffling and notes

@@
 
+\roshan{Can we cite away unification, reificatio, handling of
+  substitutions etc?}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 {{ iso s v' |--> (v_{output pattern}, s[v' <><> v_{input pattern}]) }}
+
+\roshan{Talk about non-determinism, determinism and our handling of
+  choices. Also, in what sense are we logically reversible?}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions in {{langRevEE}} }
@@
 
+MUST TALK ABOUT AND CITE ABRAMSKY AND COECKE'S TR ABOUT QUANTUM
+COMPUTING.
+
 These constructions are applicable to the additive and the
@@
 
+Mention LL and {{a -o b}}.
+
 \begin{center}
@@
 Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
+
+\roshan{Also: additive functions aren't values - they are
+  possibilities. Multiplicative functions really are values.}
 
@@
 
-MUST TALK ABOUT AND CITE ABRAMSKY AND COECKE'S TR ABOUT QUANTUM
-COMPUTING.
-
-%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions Specific to Negation}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions Specific to Multiplication}
@@
 \end{itemize}
-
-For a traced monoidal category {{C}} the Int construction produces a
-Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
-Further we know that the target of the Int construction is isomorphic
-to the target of \G construction of Abramsky \cite{Abramsky96:0} from
-Haghverdi. However, note that the {{langRevEE}} is not the same as the
-image of the Int construction on {{langRevT}}, since the the later
-lacks a multiplicative tensor that distributes over the additive tensor
-in Int {{langRevT}}.
 
@@
 
+\paragraph*{Logic Programming and Backtracking.} 
+This is a constrained form of backtracking and a constrained form of
+logic programming.
+
+\paragraph*{Linear Logic and GoI.} 
+Say something.
+
+\paragraph*{Int Construction.}
+For a traced monoidal category {{C}} the Int construction produces a
+Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
+Further we know that the target of the Int construction is isomorphic
+to the target of \G construction of Abramsky \cite{Abramsky96:0} from
+Haghverdi. However, note that the {{langRevEE}} is not the same as the
+image of the Int construction on {{langRevT}}, since the the later
+lacks a multiplicative tensor that distributes over the additive
+tensor in Int {{langRevT}}.
+
 \paragraph*{Our Approach.} The salient aspects of our approach are the following:
@@
 \item Negative and fractional types have an elementary and familiar
-  interpretation borrowed from the algebra of rational numbers. One can write
-  any algebraic identity that is valid for the rational numbers and interpret
-  it as an isomorphism with a clear computational interpretation: negative
-  values flow backwards and fractional values represent fragments of
-  aggregate data structures. None of the systems above has such a natural
+  interpretation borrowed from the algebra of rational numbers. One
+  can write any algebraic identity that is valid for the rational
+  numbers and interpret it as an isomorphism with a clear
+  computational interpretation: negative values flow backwards and
+  fractional values represent fragments of aggregate data
+  structures. None of the systems above has such a natural
   interpretation of negative and fractional types.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-\label{sec:conc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Computing in the Field of Algebraic Numbers}
 
@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+\label{sec:conc}
 

### 2012-03-09 rational.tex
dumping more thoughts into the conclusion

@@
 
-\item If we built an effect over {{langRevEE}}, say {{create}} and
-  {{erase}}, are effects structured by an arrow or a monad now?
-
-\item Which operations can be lifted to work on negative types?
-
 \item Can the operational semantics for {{langRevEE}} be an
@@
 
+Several things we don't know how to do, or are sloppy about. Some of
+these are deep questions to which there is no immeidate answer.
+
+\begin{itemize}
+
+\item Is the language without recursive types always terminating? For
+  the right symmetry with annihilation, it is desirable to have
+  non-termination.
+
+\item What is the correct lemma for logical reversibility now?
+
+\item We should say things about information preservation. At least
+  more so than what is being said now.
+
+\item Do we really have HOF and continuations? What is the exact
+  relationship? Do we have linearly used functions? Do we have linear
+  functions (in that they use their argument linearly)?
+
+\item Can we complete the construction in \cite{infeffects} to be full
+  STLC? This is again a HOF question in another guise.
+
+\item What are information effects in {{langRevEE}}? Depending on what
+  we think about HOF, do we have a monad or is it still an arrow?
+
+\item Multiplicative duality gives us a means of talking about
+  equivalence classes of values. This should correspond to all sorts
+  of other things -- groupoids, Lawvere theories etc. We barely even
+  understand these connections.
+
+\item How does this connect to QC? Is there any strong connection such
+  that the SAT solver here is an alternate QC algorithm to Shor's?
+
+\end{itemize}
+
 To summarize negative, fractional, square root, and imaginary types all make

### 2012-03-09 rational.tex
the best intro ever written in the entire world

@@
 %! alpha = \alpha
+%! bool = \textit{bool}
+%! color = \textit{color}
+%! Gr = G
 
@@
   programming by breaking global invariants into local ones that can be
-  autonomously satisfied by a subcomputation. Furthermore, these types and
-  their properties suggest that the previously observed duality of
+  autonomously satisfied by a subcomputation. Theoretically, these types give
+  rise to \emph{two} function spaces and to \emph{two} notions of
+  continuations, suggesting that the previously observed duality of
   computation conflated two orthogonal notions: an additive duality that
@@
 
-start with int times bool divided by bool: it is like an int but it can only
-be used as the second component of a pair whose first component is an int;
-gives you the power to say here is 20 but you must use it to buy books
-
-To get a feeling for the main idea, consider the following algebraic
-manipulation relating a natural number $a$ to itself:
-%% \[\begin{array}{rclll}
-%% && a & \mbox{initial~$a$} & (0) \\
-%% &=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
-%% &=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
-%% &=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
-%% &=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
-%% &=& a & \mbox{$0$~is~identity~for~$+$} & (5)
-%% \end{array}\]
-%% \[\begin{array}{rlllllllll}
-%%  \!&\!\!  a \!&\! \!&\!       \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
-%% =\!&\!\!  a \!&\!+\!&\!       \!&\!0\!&\!    \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (1) \\
-%% =\!&\!\!  a \!&\!+\!&\! (-a   \!&\!+\!&\! a) \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (2) \\
-%% =\!&\!\! (a \!&\!+\!&\! (-a)) \!&\!+\!&\! a  \!&\! \mbox{$+$~is~associative} \!&\! (3) \\
-%% =\!&\!\!    \!&\!0\!&\!       \!&\!+\!&\! a  \!&\! \mbox{$-a$~is~the~additive~inverse~of~$a$} \!&\! (4) \\
-%% =\!&\!\!    \!&\! \!&\!       \!&\! \!&\! a  \!&\! \mbox{$0$~is~identity~for~$+$} \!&\! (5)
-%% \end{array}\]
-
+\paragraph*{Negative Types.} 
+Consider the following algebraic manipulation relating a natural number $a$
+to itself (ignoring the dotted line for a moment):
 \begin{center}
@@
 \end{center}
-
 Although seemingly pointless, this algebraic proof corresponds, in our model,
@@
 produces the same \$20.00 value as its output. As the semantics of
-Sec.~\ref{sec:rat} formalizes, the intermediate steps of the computation
-proceed as follows:
+Sec.~\ref{sec:rat} formalizes, this computation flows along the dotted
+line with the following intermediate steps:
 \begin{itemize}
@@
 \end{itemize}
-
-Now consider the following algebraic manipulation (assuming $a \neq 0$):
-%% \[\begin{array}{rclll}
-%% && a & \mbox{initial~$a$} & (0) \\
-%% &=& a * 1 & \mbox{$1$~is~identity~for~$*$} & (1) \\
-%% &=& a * ((1/a) * a) & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (2) \\
-%% &=& (a * (1/a)) * a & \mbox{$*$~is~associative} & (3) \\
-%% &=& 1 * a & \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} & (4) \\
-%% &=& a & \mbox{$1$~is~identity~for~$*$} & (5)
-%% \end{array}\]
-
-%% \[\begin{array}{rcllllll}
-%%  \!&\! a \!&\! \!&\!        \!&\! \!&\!    \!&\! \mbox{initial~$a$} \!&\! (0) \\
-%% =\!&\! a \!&\!*\!&\!        \!&\!1\!&\!    \!&\! \mbox{$1$~is~identity~for~$*$} \!&\! (1) \\
-%% =\!&\! a \!&\!*\!&\! ((1/a) \!&\!*\!&\! a) \!&\! \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} \!&\! (2) \\
-%% =\!&\!(a \!&\!*\!&\! (1/a)) \!&\!*\!&\! a  \!&\! \mbox{$*$~is~associative} \!&\! (3) \\
-%% =\!&\!   \!&\!1\!&\!        \!&\!*\!&\! a  \!&\! \mbox{$1/a$~is~the~multiplicative~inverse~of~$a$} \!&\! (4) \\
-%% =\!&\!   \!&\! \!&\!        \!&\! \!&\! a  \!&\! \mbox{$1$~is~identity~for~$*$} \!&\! (5)
-%% \end{array}\]
-
-\begin{center}
-\scalebox{1.1}{
-\includegraphics{diagrams/thesis/algebra-wire2.pdf}
-}
-\end{center}
-
-As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
-{{a <-> a}} isomorphism computes as follows:
-\begin{itemize}
-\item We start at line (0) with the value {{20}};
-\item We proceed to line (1) with the value {{(20,())}};
-\item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
-  $\alpha$ is essentially a logic variable;
-\item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
-\item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
-  forces the unification $\alpha=20$, i.e., the value becomes
-  {{((20,1/20),20)}} which simplifies to {{((),20)}};
-\item We reach line (5) with the value {{20}}.
-\end{itemize}
-
-The beauty of the above computation is that it can be realized with a
-completely different concurrent schedule as follows. \emph{Simultaneously}
-the execution begins at line (0) with the value {{20}}, at line (2) with the
-value {{right (right 20)}} flowing forward and the value {{right (left 20)}}
-flowing backwards, and at line (3) with the value {{left (left 20)}} flowing
-forward and the value {{left (right 20)}} flowing backward. We now have three
-independent subcomputations proceeding in parallel: the positive value
-created at line (2) which can flow to the consumer; the negative value
-created at line (3) which can pay for the debt created at line (2); and the
-initial value from line (0) which can flow to the required forward-flowing
-value needed at line (3). At some intuitive level, one might understand the
-situation as modeling a credit card payment in which money is created at line
-(2) for the consumer together with a corresponding debt. The producer's bank
-pays for the debt at line (3) and waits for the actual money to arrive from
-the producer.
-
-There are at least two fundamental points about this example that must be
+The example illustrates that the empty type and negative types have a
+fascinating computational interpretation with some flavor of continuations.
+What is even more fascinating is that the above computation can be realized
+with a completely different concurrent schedule as
+follows. \emph{Simultaneously} the execution begins at line~(0) with the
+value {{20}}, at line (2) with the value {{right (right 20)}} flowing forward
+and the value {{right (left 20)}} flowing backwards, and at line (3) with the
+value {{left (left 20)}} flowing forward and the value {{left (right 20)}}
+flowing backward. We now have three independent subcomputations proceeding in
+parallel: the positive value created at line (2) which can flow to the
+consumer; the negative value created at line (3) which can pay for the debt
+created at line (2); and the initial value from line (0) which can flow to
+the required forward-flowing value needed at line (3). At some intuitive
+level, one might understand the situation as modeling a credit card payment
+in which money is created at line (2) for the consumer together with a
+corresponding debt. The producer's bank pays for the debt at line (3) and
+waits for the actual money to arrive from the producer.
+
+There are at least three fundamental points about this example that must be
 emphasized:
@@
   producers and consumers and let each compute independently.
+\item Consider a value of type {{-a + b}}. The above discussion suggests that
+  this value demands {{a}} and produces {{b}}, i.e., that it is a
+  higher-order funtion mapping {{a}} to {{b}}. Dually and as the underlying
+  categorical semantics justifies (see Sec.~\ref{sec:neg}), the same value is
+  a continuation which produces {{b}} and then demands {{a}}, which
+  introduces a twist on how the function can be used: get the goods {{b}} now
+  and pay {{a}} later.
 \end{itemize}
 
-Instead of also discussing fractional types and related work in this
-introduction, we briefly summarize our main technical contributions and move
-on to the heart of the paper where these topics can be explained in their
-proper context. Our main contributions are:
+\paragraph*{Fractional Types.} 
+Consider the following realization of boolean negation written using a
+pattern-matching style:
+\begin{verbatim}
+notB b = case b of 
+  False -> True
+  True -> False
+\end{verbatim}
+In a functional language, we are not used to treating the pattern-matching
+clauses \verb|False -> True| and \verb|True -> False| as first-class entities
+but fractional types allow us to do just that! Informally, we can treat each
+of the pattern-matching clauses as a first-class value of type 
+{{(1/bool) * bool}}. 
+A small but complete example will illustrate the main aspects of this idea.
+
+Let \verb|color| be the datatype with three constructors \verb|R|, \verb|G|,
+and \verb|B| and let's try to construct a function that rotates the
+colors. In other words, we would like to build the following three
+pattern-matching clauses \verb|R -> G|, \verb|G -> B|, and \verb|B -> R|.  In
+the algebra of rational numbers we have that for each {{a}} that is not equal
+to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic identity
+corresponds to a type isomorphism {{1 <-> (1/a) * a}}. For the datatype
+\verb|color|, there are three possible witnesses for this isomorphism mapping
+{{()}} to either {{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the
+particular witness determined by further constraints encountered during the
+computation. The idea of variables that capture constraints that are
+satisfied later is at the heart of logic programming and hence the witness for
+the isomorphisms above is generally {{(1/alpha,alpha)}} where~{{alpha}} is a
+fresh logic variable.
+
+Returning to our goal of creating the function that rotates the colors, we
+proceed as follows:
+\begin{itemize}
+\item Using the fact that {{1}} is the multiplicative unit, generate from the
+  input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
+\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
+  components of the above tuple. The resulting value is 
+{{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
+{{alpha2}}, and {{alpha3}} are fresh logic variables.
+\item Using the fact that {{*}} is associative and commutative, we can
+  rearrange the above tuple to produce the value:
+
+{{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
+\item At this point, we have produced the skeleton for our desired function.
+  Assume now that a client uses the first clause of the function in a context
+  that contains the value \verb|R|, i.e., the client has the value
+  {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
+  and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
+  At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
+  the first component of the pair which has the following effects:
+  \begin{itemize}
+    \item the logic variable {{alpha1}} is unified with {{R}};
+    \item the first component of the pair then reduces to () and can be
+      absorbed by the fact that {{1}} is the multiplicative unit. In other
+      words, the value reduces to simply {{alpha2}} reflecting that the
+      pattern-matching was successful and the result of the clause is now
+      available;
+    \item the first clause of the function has now disappeared and the the
+      unification adjusts the remaining two clauses to
+    {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+  \item The same client or another client can now use the second clause
+    applying it to {{G}} which would produce the result {{alpha3}} for the
+    second client and ground the result of the first client to {{G}}.
+  \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
+    which grounds the result of the second client to {{B}}.
+  \end{itemize}
+\end{itemize}
+It should be evident that in a universe in which information is not
+guaranteed to be preserved by the computational infrastructure, the above
+slicing and dicing of functions as a collection of first-class pattern
+clauses would make no sense. However, in a computational infrastructure that
+is guaranteed to preserve information, computations such as the one above are
+guaranteed to ``make sense.'' If one branch uses a clause that matches one
+part of the input to produce one part of the output, all other clauses are
+automatically adjusted to account for this to ensure that this input is not
+matched again and that this output is not produced again. Furthermore, just
+like the case for negative types, a dual reading of the the pattern-matching
+clauses as continuations allows the context to use the result of the clause
+and provide the matching argument later.
+
+\paragraph*{Contributions and Outline.} 
+To summarize, in a computational framework that guarantees that information
+is preserved, negative and fractional types provide fascinating mechanisms in
+which computations can be sliced and diced, decomposed and recomposed,
+proceeding forward and backwards, in arbitrary ways. The remainder of the
+paper formalizes these informal observations. Specifically our main
+contributions are:
 \begin{itemize}
@@
 \end{itemize}
+
+% Now consider the following algebraic manipulation (assuming $a \neq 0$):
+% \begin{center}
+% \scalebox{1.1}{
+% \includegraphics{diagrams/thesis/algebra-wire2.pdf}
+% }
+% \end{center}
+% As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
+% {{a <-> a}} isomorphism computes as follows:
+% \begin{itemize}
+% \item We start at line (0) with the value {{20}};
+% \item We proceed to line (1) with the value {{(20,())}};
+% \item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
+%   $\alpha$ is essentially a logic variable;
+% \item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
+% \item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
+%   forces the unification $\alpha=20$, i.e., the value becomes
+%   {{((20,1/20),20)}} which simplifies to {{((),20)}};
+% \item We reach line (5) with the value {{20}}.
+% \end{itemize}
 

### 2012-03-09 rational.tex
sec 2

@@
   Every functional programmer knows about sum and product types, {{a+b}} and
-  {{a*b}} respectively. Negative and fractional types, $a-b$ and $a/b$
+  {{a*b}} respectively. Negative and fractional types, {{a-b}} and {{a/b}}
   respectively, are much less known and their computational interpretation is
@@
 \paragraph*{Negative Types.} 
-Consider the following algebraic manipulation relating a natural number $a$
+Consider the following algebraic manipulation relating a natural number {{a}}
 to itself (ignoring the dotted line for a moment):
@@
 computational interpretation. The witness for this isomorphism is a
-computation that takes a value of type~$a$ (say \$20.00) and eventually
+computation that takes a value of type~{{a}} (say \$20.00) and eventually
 produces the same \$20.00 value as its output. As the semantics of
@@
 \item We proceed to line (1) with the same \$20.00 but tagged as being in the
-  left summand of the type $a+0$; we indicate this value as {{left 20}}
+  left summand of the type {{a+0}}; we indicate this value as {{left 20}}
 \item We continue to line (2) with the same value {{left 20}};
@@
 follows. \emph{Simultaneously} the execution begins at line~(0) with the
-value {{20}}, at line (2) with the value {{right (right 20)}} flowing forward
+value {{20}}, at line (2) with the value {{right (right 20)}} flowing forwards
 and the value {{right (left 20)}} flowing backwards, and at line (3) with the
-value {{left (left 20)}} flowing forward and the value {{left (right 20)}}
-flowing backward. We now have three independent subcomputations proceeding in
+value {{left (left 20)}} flowing forwards and the value {{left (right 20)}}
+flowing backwards. We now have three independent subcomputations proceeding in
 parallel: the positive value created at line (2) which can flow to the
@@
       available;
-    \item the first clause of the function has now disappeared and the the
-      unification adjusts the remaining two clauses to
-    {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+    \item the first clause of the function has now disappeared and the
+      remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
   \item The same client or another client can now use the second clause
-    applying it to {{G}} which would produce the result {{alpha3}} for the
-    second client and ground the result of the first client to {{G}}.
+    applying it to {{Gr}} which would produce the result {{alpha3}} for the
+    second client and ground the result of the first client to {{Gr}}.
   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
@@
 matched again and that this output is not produced again. Furthermore, just
-like the case for negative types, a dual reading of the the pattern-matching
+like the case for negative types, a dual reading of the pattern-matching
 clauses as continuations allows the context to use the result of the clause
@@
 is preserved, negative and fractional types provide fascinating mechanisms in
-which computations can be sliced and diced, decomposed and recomposed,
-proceeding forward and backwards, in arbitrary ways. The remainder of the
-paper formalizes these informal observations. Specifically our main
-contributions are:
+which computations can be sliced and diced, decomposed and recomposed, ran
+forwards and backwards, in arbitrary ways. The remainder of the paper
+formalizes these informal observations. Specifically our main contributions
+are:
 \begin{itemize}
@@
   isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
-  of negative types, that satisfies the isomorphism $a + (-a) = 0$. The
+  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
   semantics of this extension is expressed by having a ``dual'' evaluator
@@
 \item We independently extend {{langRev}} with a notion of fractional types,
-  that satisfies the isomorphism $a * (1/a) = 1$. The semantics of this
+  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
   extension is expressed by introducing logical variables and a unification
@@
 
-% Now consider the following algebraic manipulation (assuming $a \neq 0$):
-% \begin{center}
-% \scalebox{1.1}{
-% \includegraphics{diagrams/thesis/algebra-wire2.pdf}
-% }
-% \end{center}
-% As the semantics in Sec.~\ref{sec:rat} formalizes, the witness to this 
-% {{a <-> a}} isomorphism computes as follows:
-% \begin{itemize}
-% \item We start at line (0) with the value {{20}};
-% \item We proceed to line (1) with the value {{(20,())}};
-% \item We proceed to line (2) with the value {{(20,((1/alpha),alpha))}} where
-%   $\alpha$ is essentially a logic variable;
-% \item We proceed to line (3) with the value {{((20,1/alpha),alpha)}};
-% \item At line (4) we require that $20$ and $1/\alpha$ cancel each other which
-%   forces the unification $\alpha=20$, i.e., the value becomes
-%   {{((20,1/20),20)}} which simplifies to {{((),20)}};
-% \item We reach line (5) with the value {{20}}.
-% \end{itemize}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 section differs from the one in our previous paper~\cite{infeffects} in two
-aspects. First, we add the empty type $0$ which is necessary to express the
+aspects. First, we add the empty type {{0}} which is necessary to express the
 additive duality. Second, instead of explaining evaluation using a natural
@@
 appropriate for the connections with continuations explored in this paper.
-
-\roshan{Say somewhere that {{langRev}} corresponds to computing in the
-  commutative rig of whole numbers. More bluntly {{langRev}} is whole
-  numbers with addition and multiplication, while {{langRevEE}} is the
-  rational with addition and multiplication.}
 
@@
 form a congruence relation by adding the following constructors that witness
-the equivalence and compatible closure:
+equivalence and compatible closure:
 %subcode{proof} include main
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%
 \subsection{Graphical Language}
 
-The syntactic notation described in Sec \ref{sec:pi-syntax} is often
-obscure and hard to read.  Following the tradition established for
-monoidal categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we
-present a graphical language that conveys the intuitive semantics of
-the language (which is formalized in the next section).  Since, the
-graphical notation is easier to read an trace, for the rest of the
-paper {{langRev}} programs will be presented graphically.
-
-The general idea of the graphical notation is that combinators are
-modeled by ``wiring diagrams'' or ``circuits''; values are modeled as
-``particles'' that may appear on the wires; and evaluation is modeled
-by the flow of particles along the wires. In fact, when talking about
-{{langRev}} circuits, we will often use the words {{value}} and
-{{particle}] interchangeably.
-
-%% We now proceed to give the graphical notation for each combinator:
+The syntactic notation above is often obscure and hard to read.  Following
+the tradition established for monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
+graphical language that conveys the intuitive semantics of the language
+(which is formalized in the next section).
+
+The general idea of the graphical notation is that combinators are modeled by
+``wiring diagrams'' or ``circuits'' and that values are modeled as
+``particles'' that may appear on the wires. Evaluation therefore is modeled
+by the flow of particles along the wires.
+
+%% In fact, when talking about {{langRev}} circuits, we will often use the
+%% words {{value}} and {{particle}} interchangeably.
 
 \begin{itemize}
-\item The simplest sort of diagram is the {{id : b <-> b}} combinator
-  which is simply represented as a wire labeled by its type {{b}}, as
-  shown on the left. In more complex diagrams, if the type of a wire
-  is obvious from the context, it may be omitted.
+\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
+  is simply represented as a wire labeled by its type {{b}}, as shown on the
+  left. In more complex diagrams, if the type of a wire is obvious from the
+  context, it may be omitted. When tracing a computation, one might imagine a
+  value {{v}} of type {{b}} on the wire, as shown on the right.
 
@@
 \end{center}
-
 \begin{center}
@@
 
-When tracing a computation, one might imagine a value {{v}} of of type
-{{b}} on the wire, as shown on the right.
-
-\item The product type {{b1*b2}} may be represented both as one wire labeled
-  {{b1*b2}} or by two parallel wires labeled {{b1}} and {{b2}}. Both
-  representations maybe used interchangeably.
+\item The product type {{b1*b2}} may be represented using either one wire
+  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
+  case of products represented by a pair of wires, when tracing execution
+  using particles, one should think of one value/particle on each wire.
 \begin{multicols}{2}
@@
 \scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
 %%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
@@
 \end{center}
-\end{multicols}
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
-}
-\end{center}
 \begin{center}
@@
 
-In the case of products represented by a pair of wires, when tracing
-execution using particles, one should think of one value/particle on
-each wire.
-
 \item Sum types may similarly be represented by one wire or using using
-  parallel wires with a {{+}} operator between them.
+  parallel wires with a {{+}} operator between them. When tracing the
+  execution of two additive wires, a value can reside on only one of the two
+  wires.
 \begin{multicols}{2}
@@
 \scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
 %%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
 \includegraphics{diagrams/thesis/sum-two-wires.pdf}
-}
-\end{center}
-\end{multicols}
-
-When tracing the execution of two additive wires, a value can reside
-on only one of the two wires.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
 }
@@
 
-The crisscrossing wires have the right semantics in that a value on
-the top wire, ends up on the bottom wire and vice versa.
-
-\item Identity morphisms and their duals ({{uniti, unite, zeroi}} and
-  {{zeroe}}) are represented as follows:
+\item The morphisms that witness that {{0}} and {{1}} are the additive and
+  multiplicative unites are represented as shown below. Note that since there
+  is no value of type 0, there can be no particle on a wire of type {{0}}.
+  Also since the monoidal units can be freely introduced and eliminated, in
+  many diagrams they are omitted and dealt with explicitly only when they are
+  of special interest.
 \begin{multicols}{2}
@@
 \end{multicols}
-Note that since there is no value of type 0, there can be no particle
-on a wire of type {{0}}.  Since the monoidal units can be freely
-introduced and eliminated, in many diagrams they are omitted and dealt
-with explicitly when they are of special interest.
 
@@
 \noindent 
-\textit{Example.} As an example, the following combinator is represented by
-the given diagram:
+\textit{Example.} The following combinator is represented by the given
+diagram:
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics}
@@
 single-step reductions. Since there are no values of type {{0}}, the rules
-omit some of the impossible cases:
+omit the impossible cases:
 \begin{scriptsize}
@@
 \end{scriptsize}
-
-The reductions for the primitive isomorphisms are exactly the same as
-have been presented before \cite{infeffects}. The reductions for the
-closure combinators are however presented in a small-step operational
-style. The following are the definition for evaluation contexts and
-machine states:
-
+The reductions for the primitive isomorphisms above are exactly the same as
+have been presented before~\cite{infeffects}. The reductions for the closure
+combinators are however presented in a small-step operational style using the
+following definitions of evaluation contexts and machine states:
+
+\begin{scriptsize}
 %subcode{bnf} include main
@@
 % Stop State = {[c, v, []]}
-
-The reduction rules below essentially track the flow of particles
-through their circuit. The start machine state, {{<c,v,[]>}}, denotes
-the particle/value {{v}} about to be evaluated by the circuit
-{{c}}. The end machine state, {{[c, v, [] ]}}, denotes the situation
-where the particle {{v}} has exited the circuit {{c}}.
-
-%% For the non-primitive combinators, we express the semantics using evaluation
-%% contexts which are used to represent the ``rest of the computation.'' More
-%% precisely, to calculate the result of applying combinator $c$ to value $v$,
-%% we start a small-step abstract machine in the configuration {{<c,v,[]>}}. The
-%% transitions of the machine (see below) are unusual in the sense that they
-%% always maintain full information about {{c}}, although this information may
-%% be split in the first and last component of the machine. The {{v}} component
-%% of the machine is the one that captures the current value as it is flowing in
-%% the combinator {{c}}. When  scattered in the machine will examine {{c}} and make
-%% transitions based on the structure of {{c}}. When the evaluation of {{c}} is
-%% complete, the configuration should be {{[c,v,[]]}} where the {{c}} is
-%% restored.
-
-%% This is a mess. Perhaps give a small example and explain things based on the
-%% example before giving the full transitions.
-
+\end{scriptsize}
+The machine transitions below track the flow of particles through a
+circuit. The start machine state, {{<c,v,[]>}}, denotes the particle/value
+{{v}} about to be evaluated by the circuit {{c}}. The end machine state,
+{{[c, v, [] ]}}, denotes the situation where the particle {{v}} has exited
+the circuit {{c}}.
+
+\begin{scriptsize}
 %subcode{opsem} include main
@@
 % {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (11)
-
-The rule (1) describes evaluation by a primitive isomorphism. Rules
-(2), (3) and (4) deal with sequential evaluation. Rule (2) says that
-for the value {{v}} to flow through the sequence {{c1 (;) c2}}, it
-should first flow through {{c1}} with {{c2}} pending in the context
-({{Fst C c2}}). Rule (3) says the value {{v}} that exits from {{c1}}
-should proceed to flow through {{c2}}. Rule (4) says that if the value
-{{v}} has exited {{c2}} as well, then it is equivalent to having
-exited from {{c1(;)c2}}.
-
-Rules (5) to (8) deal with {{c1 (+) c2}} in the same way. In the case
-of sums, the shape of the value,i.e. is it is tagged with {{left}} or
-{{right}}, determines whether path {{c1}} or path {{c2}} is
-taken. Rules (9), (10) and (11) deal with {{c1 (*) c2}} similarly. In
-the case of products the value should have the form {{(v1, v2)}} where
-{{v1}} flows through {{c1}} and {{v2}} flows through {{c2}}. Both
-these paths are entirely independent of each other and we could
-evaluate either first, or evaluate both in parallel. In this
-presentation we have chosen to follow {{c1}} first, but this choice is
-entirely arbitrary.
-
-The interesting thing about the semantics is that it represents a
-reversible abstract machine. In other words, we can compute the start
-state from the stop state by changing the reductions `{{|-->}}' to run
-backwards `{{<--|}}'. We will take advantage of this property in
-Sec. \ref{sec:rat}.
+\end{scriptsize}
+Rule (1) describes evaluation by a primitive isomorphism. Rules (2), (3) and
+(4) deal with sequential evaluation. Rule (2) says that for the value {{v}}
+to flow through the sequence {{c1 (;) c2}}, it should first flow through
+{{c1}} with {{c2}} pending in the context ({{Fst C c2}}). Rule (3) says the
+value {{v}} that exits from {{c1}} should proceed to flow
+through~{{c2}}. Rule (4) says that when the value {{v}} exits {{c2}}, it also
+exits the sequential composition {{c1(;)c2}}. Rules (5) to (8) deal with 
+{{c1 (+) c2}} in the same way. In the case of sums, the shape of the value,
+i.e., whether it is tagged with {{left}} or {{right}}, determines whether
+path {{c1}} or path {{c2}} is taken. Rules (9), (10) and (11) deal with 
+{{c1 (*) c2}} similarly. In the case of products the value should have the
+form {{(v1, v2)}} where {{v1}} flows through {{c1}} and {{v2}} flows through
+{{c2}}. Both these paths are entirely independent of each other and we could
+evaluate either first, or evaluate both in parallel. In this presentation we
+have chosen to follow {{c1}} first, but this choice is entirely arbitrary.
+
+The interesting thing about the semantics is that it represents a reversible
+abstract machine. In other words, we can compute the start state from the
+stop state by changing the reductions {{|-->}} to run backwards
+{{<--|}}. When running backwards, we use the isomorphism represented by a
+combinator {{c}} in the reverse direction which we indicate by writing
+{{c{dagger}}}. For example, {{distrib{dagger} = factor}} and so on.
 
@@
 \label{prop:logrev}
-{{c v |--> v'}} iff {{c{dagger} v' |--> v}}.
+{{<c,v,[]> |--> [c,v',[]]}} iff 
+{{<c{dagger},v',[]> <--| [c{dagger},v,[]]}} 
 \end{proposition}
 
-\roshan{We really have to rethink this prop in the presence of
-  negatives and fractionals. With negatives, the program can actually
-  end at the beginning of the circuit -- i.e. the program {{c:b1<->b2}}
-  can stop with a value of type {{b1}}. With fractionals, there can be
-  several values of type {{b2}} produced for a value of type {{b1}}. }
-
-\begin{proposition}[Groupoid]
-\label{prop:groupoid}
-{{langRev}} is a groupoid. 
-\end{proposition}
-
-\begin{proposition}
-\label{prop:category}
-{{langRev}} is a dagger symmetric monoidal category. 
-\end{proposition}
+% \roshan{We really have to rethink this prop in the presence of
+%   negatives and fractionals. With negatives, the program can actually
+%   end at the beginning of the circuit -- i.e. the program {{c:b1<->b2}}
+%   can stop with a value of type {{b1}}. With fractionals, there can be
+%   several values of type {{b2}} produced for a value of type {{b1}}. }
+
+% \begin{proposition}[Groupoid]
+% \label{prop:groupoid}
+% {{langRev}} is a groupoid. 
+% \end{proposition}
+
+% \begin{proposition}
+% \label{prop:category}
+% {{langRev}} is a dagger symmetric monoidal category. 
+% \end{proposition}
 
@@
 travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
-the execution. It goes forward in time so to speak, comes back, and then goes
+the execution. It goes forwards in time so to speak, comes back, and then goes
 forwards again.
@@
 aka information particles, to flow backwards in time. Particles that
-flow backward in time get to see and interact with the history of
+flow backwards in time get to see and interact with the history of
 particles that they coexist with i.e. values that they exist
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Fractional Types}
@@
 
-The type $a/b$ is prominent in the Lambek-Grishin calculus which is
+The type {{a/b}} is prominent in the Lambek-Grishin calculus which is
 extensively used in computational linguistics. In the common interpretation,
-a value of type $a/b$ is a value of some type $c$ such that when put in a
+a value of type {{a/b}} is a value of some type $c$ such that when put in a
 context of type $b$, the result is a value of type $a$. For example, assuming
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Multiplicative Duality in {{langRev}} }
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%
 \subsection{Unification and Logic Programming}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%
 \subsection{Extending {{langRev}} Semantics}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%
 \subsection{Semantics of {{langRevEE}} }
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions in {{langRevEE}} }
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%
 \subsection{Constructions Specific to Negation}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions Specific to Multiplication}
@@
 output. Acceptors however are also similar in flavor to logical variables:
-they can created and instantiated later once their context of use is
+they can be created and instantiated later once their context of use is
 determined. Although a formal connection is lacking, it is clear that, at an
@@
 
-\paragraph*{Logic Programming and Backtracking.} 
-This is a constrained form of backtracking and a constrained form of
-logic programming.
-
-\paragraph*{Linear Logic and GoI.} 
-Say something.
-
-\paragraph*{Int Construction.}
-For a traced monoidal category {{C}} the Int construction produces a
-Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
-Further we know that the target of the Int construction is isomorphic
-to the target of \G construction of Abramsky \cite{Abramsky96:0} from
-Haghverdi. However, note that the {{langRevEE}} is not the same as the
-image of the Int construction on {{langRevT}}, since the the later
-lacks a multiplicative tensor that distributes over the additive
-tensor in Int {{langRevT}}.
-
-\paragraph*{Our Approach.} The salient aspects of our approach are the following:
+\paragraph*{Our Approach.} The salient aspects of our approach are the
+following:
 \begin{itemize}
 \item Negative and fractional types have an elementary and familiar
-  interpretation borrowed from the algebra of rational numbers. One
-  can write any algebraic identity that is valid for the rational
-  numbers and interpret it as an isomorphism with a clear
-  computational interpretation: negative values flow backwards and
-  fractional values represent fragments of aggregate data
-  structures. None of the systems above has such a natural
-  interpretation of negative and fractional types.
+  interpretation borrowed from the algebra of rational numbers. One can write
+  any algebraic identity that is valid for the rational numbers and interpret
+  it as an isomorphism with a clear computational interpretation: negative
+  values flow backwards and fractional values represent constraints on the
+  context. None of the systems above has such a natural interpretation of
+  negative and fractional types.
 \item Because we are \emph{not} in the context of the full
@@
   shared with systems that are based on linear logic; other systems must
-  impose ad hoc constraints to ensure negative values are used exactly
-  once. 
+  impose ad hoc constraints to ensure negative and fractional values are used
+  exactly once.
 \item In contrast to all the work that takes continuations as primitive
@@
 
+\paragraph*{Logic Programming and Backtracking.} 
+This is a constrained form of backtracking and a constrained form of
+logic programming.
+
+\paragraph*{Linear Logic and GoI.} 
+Say something.
+
+\paragraph*{Int Construction.}
+For a traced monoidal category {{C}} the Int construction produces a Compact
+Closed Category called Int {{C}} \cite{joyal1996traced}.  Further we know
+that the target of the Int construction is isomorphic to the target of \G
+construction of Abramsky \cite{Abramsky96:0} from Haghverdi. However, note
+that the {{langRevEE}} is not the same as the image of the Int construction
+on {{langRevT}}, since the later lacks a multiplicative tensor that
+distributes over the additive tensor in Int {{langRevT}}.
+
 %% Jacques

### 2012-03-09 rational.tex
_
nothing really
lots of stuff

@@
   that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
-  extension is expressed by introducing logical variables and a unification
+  extension is expressed by introducing logic variables and a unification
   mechanism to model and resolve the constraints introduced by the fractional
@@
 {{<c,v,[]> |--> [c,v',[]]}} iff 
-{{<c{dagger},v',[]> <--| [c{dagger},v,[]]}} 
+{{<c{dagger},v',[]> |--> [c{dagger},v,[]]}} 
 \end{proposition}
@@
 
-Consider the following two ways of purchasing an item that costs \$20.00:
-\begin{enumerate}
-\item You give the seller a \$20.00 bill.
-\item You use a credit card to pay the seller.
-\end{enumerate}
-In both cases, the seller receives the money immediately but there is a
-subtle difference. In the first transaction, the money received by the seller
-corresponds to a value that has been produced earlier in the computation. In
-the second transaction, the money may or may not exist yet: computationally,
-a \emph{debt} is generated to compensate for the money received by the seller
-and this debt travels ``backwards'' towards the bank where it is (hopefully)
-reconciled.
-
-The example suggests that the ability to consider values flowing backwards
-enriches our computational model. This observation goes back to Filinski's
-Masters thesis where continuations are identified with these negative values.
-We discuss the connections to Filinski's work and others in more detail in
-Sec.~\ref{sec:related}. For now, we note that in a conventional programming
-language in which values can be copied and deleted at will, extreme care is
-needed to keep track of negative values. Indeed, in the example above, if it
-were possible to simply delete the variable corresponding to the debt, we
-would have produced money out of nothing.
-
-For this reason, our treatment of negative values in the context of
-{{langRev}} is particularly simple. We will have a type $0$ and an
-isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-allows us to create a value and a corresponding debt out of nothing. Both the
-value and its negative counterpart can flow anywhere in the system. Because
-information is preserved, a closed program (which does not have any
-``dangling'' negative values) will eventually have to match the negative
-value with some corresponding value, effectively using the isomorphism in the
-right-to-left direction. In more detail, we can model the credit card
-transaction above using the following program (shown diagrammatically):
-
-\begin{center}
-  \includegraphics{diagrams/algebra1.pdf}
-\end{center}
-
-\roshan{I chose to draw the figure with b instead of a, since b is what
-  we use for the types. Should this be a? Also I chose to make the
-  swap and 0 wires, explicit.Is this what you intended? }
-
-CLEAN UP the following based on the figure. In contrast, in our setting, one
-can start from the empty type $0$, introduce a positive value and its
-negative counterpart, and let each of these flow in arbitrary ways. The
-entire framework guarantees that neither the value nor its negative
-counterpart will be deleted or duplicated and hence that, in any closed
-program, the ``debt'' corresponding to the negative value is paid off exactly
-once. Computationally we model the first transaction using essentially the
-identity function which receives a \$20.00 bill as input from the buyer and
-propagates it on its output to the seller. The second transaction is more
-complicated. We model it as shown in the circuit below: There are two ways to
-understand this circuit that are both quite instructive. Let's first examine
-the type structure of each of the combinators that comprise the circuit. The
-first combinator on the right outputs \$20.00 to the seller. This \$20.00 is
-produced from nothing so to speak by generating an equivalent debt that
-travels backwards. COMPLETE BASED ON THE FIGURE. The other way is to follow
-the execution. It goes forwards in time so to speak, comes back, and then goes
-forwards again.
-
-It is critical that the framework in which the negative types are introduced
-is a framework in which all information is preserved, with no duplication or
-erasure. This guarantees that the generated debts are paid once and exactly
-once.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Observability.} 
-Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
-{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
-defined only if {{b1}} and {{b2}} are entirely positive types. This is
-similar to the constraint that Zeilberger imposes to explain intuitionistic
-polarity and delimited control~\cite{10.1109/LICS.2010.23}.
-
-Consider the program that
-
-(TODO: So this is a really bad explanation. But this section is really
-important and we need to explain carefully. Maybe an analogy can be
-drawn to Cayley diagrams for groups to point out that the notion of
-`i' is entirely artificial, but it is essential to the discourse.)
-
-\begin{center}
-  \includegraphics{diagrams/shuffle.pdf}
-\end{center}
-
-We define observables to be only positive types. The outputs of
-programs that output mixed positive and negative types are not
-observable.  Also, programs that input mixed positive and negatives
-types are not executable.
+The structure of this section and the next are the same. The general outline
+is as follows:\footnote{We refer the reader to Selinger's excellent survey
+  article of monoidal
+  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
+  definitions.}
+\begin{itemize}
+\item As established in our earlier work~\cite{rc2011,infeffects}, the
+  underlying categorical semantics of our core reversible language
+  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
+  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
+  tensor.
+\item We extend each underlying symmetric monoidal structure to a
+  \emph{compact closed} structure by adding a dual for each object and two
+  special morphims traditionally called {{eta}} and {{eps}}. 
+\end{itemize}
+After presenting the extension at the syntactic level, we discuss the
+categorical semantics informally via the graphical language. The operational
+semantics is presented in Sec.~\ref{sec:rat}. As will be detailed in this
+section and the next, the compact closed structure provides several
+properties of interest: (i) morphisms or wires are allowed to run from right
+to left, (ii) the structure admits a trace that, depending on the situation,
+can be used to implement various notions of loops and
+recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
+(iii) the structure includes an isomorphism showing that the dual operator is
+an involution, and (iv) the structure is equipped with objects representing
+higher-order functions and a morphism \textit{eval} that applies these
+functional objects. Interestingly each monoidal structure provides the same
+ingredients, resulting in two dualities, two traces, two involutions, and two
+notions of higher-order functions.
 
@@
 
-Additive duality is added to {{langRev}} by addition of negative types
-denoted {{-b}} and the isomorphisms {{eta+}} and {{eps+}}. Hence we
-extend the language as follows:
+Describing the syntax and types of an extension to {{langRev}} with additive
+duality is fairly straightforward. Basically, we extend the language with
+negative types denoted {{-b}}, negative values denoted {{-v}}, and two
+isomorphisms {{eta+}} and {{eps+}}:
 
@@
 
-\noindent
-with the types judgments
-
+The isomorphisms {{eta+}} and {{eps+}} witness the following
+isomorphism:
 %subcode{opsem} include main
 % eta+ &: 0 <-> (-b) + b :& eps+
-
+And as expected we have the following typing rule for negative value:
 %subcode{proof} include main
@@
 
-We visually present {{eta+}} and {{eps+}} as `U'-shaped connectors. On
-the left below is {{eta+}} showing the map from {{0}} to {{-b+b}} and
-dually on the right is {{eps+}} showing the map from {{-b+b}} to
-0. Even though the diagrams below show a {{0}} wire for completeness,
-later diagrams will always drop them in contexts where we can
-implicitly introduce/eliminate then using {{zeroi}}/{{zeroe}}.
-
+For the graphical language, we visually present {{eta+}} and {{eps+}} as
+U-shaped connectors. On the left below is {{eta+}} showing the map from {{0}}
+to {{-b+b}} and dually on the right is {{eps+}} showing the map from {{-b+b}}
+to 0. Even though the diagrams below show a {{0}} wire for completeness,
+later diagrams will always drop them in contexts where we can implicitly
+introduce/eliminate then using {{zeroi}}/{{zeroe}}:
 \begin{multicols}{2}
@@
 
-The usual interpretation of the type {{b1+b2}} is that we either have
-an appropriately tagged value of type {{b1}} or an appropriately
-tagged value of type {{b2}}. This interpretation is maintained in the
-presence of {{eta+}} and {{eps+}} in the following sense: a value of
-type {{right v : -b + b}} flowing into an {{eps+}} on the lower wire
-switches into a value {{left (-v): -b + b}} that flows backwards on the
-upper wire. The inversion of direction is captured by the negative
-sign on the value.
+The usual interpretation of the type {{b1+b2}} is that we either have an
+appropriately tagged value of type {{b1}} or an appropriately tagged value of
+type {{b2}}. This interpretation is maintained in the presence of {{eta+}}
+and {{eps+}} in the following sense: a value of type {{right v : -b + b}}
+flowing into an {{eps+}} on the lower wire switches to a value 
+{{left (-v): -b + b}} that flows backwards on the upper wire. The inversion of
+direction is captured by the negative sign on the value. 
+As an example, the diagram representing the isomorphism from the introduction
+is:
+\begin{center}
+  \includegraphics{diagrams/algebra1.pdf}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Compact Closure with Additive Inverses}
+
+\paragraph*{Coherence.}
+The appropriate coherence condition for Compact Closed Categories,
+expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
+may be visualized as explained by Selinger
+\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+\paragraph*{Trace}
+
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+\paragraph*{Involution (Principium Contradictiones)}
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+\paragraph*{Higher-Order Functions}
+
+Mention LL and {{a -o b}}.
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
+
+\roshan{Also: additive functions aren't values - they are
+  possibilities. Multiplicative functions really are values.}
+
+\paragraph*{Function application}
+
+\begin{center}
+  \includegraphics{diagrams/apply1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/apply2.pdf}
+\end{center}
+
+\paragraph*{Function composition}
+
+\begin{center}
+  \includegraphics{diagrams/compose1.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/compose.pdf}
+\end{center}
+
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+\paragraph*{Delimited continuation}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+%%%%%%%%%%%%%
 
@@
 
-\paragraph*{Coherence.}
-The appropriate coherence condition for Compact Closed Categories,
-expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-may be visualized as explained by Selinger
-\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
 \paragraph*{Non-termination.}
@@
 
-ADD MATH.
+For this reason, our treatment of negative values in the context of
+{{langRev}} is particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
+right-to-left direction.
+
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
+{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
+defined only if {{b1}} and {{b2}} are entirely positive types. This is
+similar to the constraint that Zeilberger imposes to explain intuitionistic
+polarity and delimited control~\cite{10.1109/LICS.2010.23}.
+
+Consider the program that
+
+(TODO: So this is a really bad explanation. But this section is really
+important and we need to explain carefully. Maybe an analogy can be
+drawn to Cayley diagrams for groups to point out that the notion of
+`i' is entirely artificial, but it is essential to the discourse.)
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The outputs of
+programs that output mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
 
@@
 
-The type {{a/b}} is prominent in the Lambek-Grishin calculus which is
-extensively used in computational linguistics. In the common interpretation,
-a value of type {{a/b}} is a value of some type $c$ such that when put in a
-context of type $b$, the result is a value of type $a$. For example, assuming
-contexts are represented as functions, a value of type $\texttt{bool} /
-(\texttt{int} \rightarrow \texttt{bool})$ is simply a value of type
-$\texttt{int}$. Indeed in this case, putting a value of type $\texttt{int}$
-in the context $(\texttt{int} \rightarrow \texttt{bool})$ produces a value of
-type $\texttt{bool}$.
-
-In our case, the situation is simpler. If the goal is to produce a value of
-type {{a * b}} and a subcomputation can only produce the {{a}}-part, it would
-have type {{(a * b) / b}}. In most settings, this type would be equivalent to
-{{a}}. However in our setting, the constraint that this value must eventually
-fit in a larger value that supplies the {{b}} is explicitly recorded and must
-be resolved. In more detail, we have an isomorphism {{ 1 <-> a * (1/a) }}
-which when used in the left-to-right direction allows the creation of a value
-and a corresponding constraint on the context.  Both the value and the
-constraint can flow in arbitrary ways during the computation but eventually
-in a closed program with no ``dangling'' constraints, the isomorphism should
-be used in the right-to-left direction to resolve the constraints.  To
-understand the computational interpretation, consider the following example.
-
-The goal is to produce a value of type $(\texttt{bool} \times
-\texttt{int})$. One part of the program, $e_1$, knows how to produce the
-value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
-to produce the value of type $\texttt{bool}$ (say true). Computationally, we
-model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
-uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
-is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
-$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
-$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
-the two values meet, we can group the components as follows:
-\[
-(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
-\]
-Using the isomorphism in the right-to-left direction on the last two pairs,
-forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
-{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
-with the pair $(true,3)$ as desired.
-
-To summarize, one can think of a value of type $a/b$ as a value that imposes
-a constraint on its context of use: it is a value that requires its context
-to be of type $b$ and only if that condition is satisfied, can the value be
-considered as a value of type $a$. In the simplest case, the type $1/b$ just
-constraints the context to be of type $b$. By allowing the fractional types
-to be first-class, we can separate the generation of constraints from their
-use. Both the value and the constraint can flow in arbitrary ways during the
-computation but eventually in a closed program with no ``dangling''
-constraints, the isomorphism should be used in the right-to-left direction to
-resolve the constraints.
-
-NEED to work out the example in detail (or perhaps another example that's
-better.
-
-Again it is critical that the framework in which the fractional types are
-introduced is a framework in which all information is preserved, with no
-duplication or erasure. This guarantees that the generated constraints are
-satisfied once and exactly once.
-
-\begin{center}
-  \includegraphics{diagrams/algebra2.pdf}
-\end{center}
+The section follows the outline of the previous one, first presenting the new
+syntax, and discussing the graphical language and the properites implied by
+the underlying categorical structure.
 
@@
 Multiplicative duality is added to {{langRev}} by adding types
-{{b^^^}}, values {{v^^^}} and isomorphisms {{eta*}} and {{eps*}}.
-Note that we will freely write {{b^^^}} as {{1/b}} and {{a*b^^^}} as
-{{a/b}}.
+{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.
 
 %subcode{bnf} include main
-% Value Types, b = ... | inv(b)
-% Values, v = ... | inv(v)
+% Value Types, b = ... | 1/b
+% Values, v = ... | 1/v
 %
@@
 %subcode{opsem} include main
-% eta* &: 1 <-> inv(b) * b :& eps*
+% eta* &: 1 <-> (1/b) * b :& eps*
 
@@
 %@ |- v : b
-%@@ |- inv(v) : inv(b)
+%@@ |- 1/v : 1/b
 
 As with additive duals, we represent {{eta*}} and {{eps*}} using
-`U'-shaped connectors. In the general case the wires for {{1}} will be
+U-shaped connectors. In the general case the wires for {{1}} will be
 dropped as it can freely introduced or eliminated using {{uniti}} and
@@
 {{b1}} and a value of type {{b2}}. This interpretation is maintained
-in the presence of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
+in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
 to be viewed as a fission point for a value of type {{b}} and its
-multiplicative inverse {{inv(b)}}. This naturally raises the question --
+multiplicative inverse {{1/b}}. This naturally raises the question --
 which value of type {{b}} is produced? The isomorphism {{eta*}} does
 not produce any specific value of type {{b}}, instead it produces a
-place holder for a value such that the value of type {{inv(b)}} is
+place holder for a value such that the value of type {{1/b}} is
 always the dual of the chosen one.
@@
 \end{multicols}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Compact Closure with Multiplicative Inverses}
+
+\paragraph*{Coherence.} 
+As before, the requisite coherence condition on Compact Closed
+Categories is applicable and is expressed below in terms of
+{{eta*}}/{{eps*}} and the multiplicative monoid.
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
+The constructions for functions, involution, and trace are identical,
+replacing {{+}} by {{*}} mutatis mutandis.
+
+%%%%%
 
@@
 of a new typed logic variable of type {{b}} and its corresponding dual
-of type {{inv(b)}}. These two values are free to move independently
+of type {{1/b}}. These two values are free to move independently
 through a circuit, but any operations on value affect what operations
@@
 
-\paragraph*{Coherence.} 
-As before, the requisite coherence condition on Compact Closed
-Categories is applicable and is expressed below in terms of
-{{eta*}}/{{eps*}} and the multiplicative monoid.
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
 \paragraph*{Annihilation.}
@@
 \label{sec:rat}
+
+once we combine the two structures, we seem to retain the monoidal
+structure. How!?
 
@@
 \begin{enumerate}
-\item {{eta* s () |--> ((inv(v), v), s)}}
+\item {{eta* s () |--> ((1/v, v), s)}}
   where {{v}} is a fresh logic variable.
@@
 \label{sec:monoidal-constructions}
-
-MUST TALK ABOUT AND CITE ABRAMSKY AND COECKE'S TR ABOUT QUANTUM
-COMPUTING.
-
-These constructions are applicable to the additive and the
-multiplicative monoid.
-
-\paragraph*{Functions}
-
-Mention LL and {{a -o b}}.
-
-\begin{center}
-  \includegraphics{diagrams/function.pdf}
-\end{center}
-
-Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
-
-\roshan{Also: additive functions aren't values - they are
-  possibilities. Multiplicative functions really are values.}
-
-\paragraph*{Function application}
-
-\begin{center}
-  \includegraphics{diagrams/apply1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/apply2.pdf}
-\end{center}
-
-\paragraph*{Function composition}
-
-\begin{center}
-  \includegraphics{diagrams/compose1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/compose.pdf}
-\end{center}
-
-This is also equivalent to sequencing both the computation blocks.
-
-\begin{center}
-  \includegraphics{diagrams/compose2.pdf}
-\end{center}
-
-\paragraph*{Delimited continuation}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}
-
-\paragraph*{Trace }
-
-%subcode{proof} include main
-%@ c : b2 + b1 <-> b2 + b3
-%@@ trace c : b1 <-> b3
-
-\begin{center}
-  \includegraphics{diagrams/trace.pdf}
-\end{center}
-
-\paragraph*{Principium Contradictiones}
-
-{{b <-> -(-b)}}
-
-\begin{center}
-  \includegraphics{diagrams/double_neg.pdf}
-\end{center}
 
@@
 continuations in the sense that they take an input and return no
-output. Acceptors however are also similar in flavor to logical variables:
+output. Acceptors however are also similar in flavor to logic variables:
 they can be created and instantiated later once their context of use is
@@
 
-%% Jacques
+%% Jacques Carette, Michael Adams
 

### 2012-03-10 rational.tex
intro.

--- 2012-03-09 rational.tex
@@
 produces the same \$20.00 value as its output. As the semantics of
-Sec.~\ref{sec:rat} formalizes, this computation flows along the dotted
-line with the following intermediate steps:
+Sec.~\ref{sec:rat} formalizes, this computation `flows along the dotted
+line' with the following intermediate steps:
 \begin{itemize}
 \item We start at line (0) with \$20.00; 
-\item We proceed to line (1) with the same \$20.00 but tagged as being in the
-  left summand of the type {{a+0}}; we indicate this value as {{left 20}}
+\item We proceed to line (1) with the same \$20.00 but tagged as being
+  in the left summand of the sum type `{{a+0}}'; we indicate this value
+  as {{left 20}}
 \item We continue to line (2) with the same value {{left 20}};
@@
 The example illustrates that the empty type and negative types have a
-fascinating computational interpretation with some flavor of continuations.
-What is even more fascinating is that the above computation can be realized
-with a completely different concurrent schedule as
-follows. \emph{Simultaneously} the execution begins at line~(0) with the
-value {{20}}, at line (2) with the value {{right (right 20)}} flowing forwards
-and the value {{right (left 20)}} flowing backwards, and at line (3) with the
-value {{left (left 20)}} flowing forwards and the value {{left (right 20)}}
-flowing backwards. We now have three independent subcomputations proceeding in
-parallel: the positive value created at line (2) which can flow to the
-consumer; the negative value created at line (3) which can pay for the debt
-created at line (2); and the initial value from line (0) which can flow to
-the required forward-flowing value needed at line (3). At some intuitive
-level, one might understand the situation as modeling a credit card payment
-in which money is created at line (2) for the consumer together with a
-corresponding debt. The producer's bank pays for the debt at line (3) and
-waits for the actual money to arrive from the producer.
-
-There are at least three fundamental points about this example that must be
+fascinating computational interpretation as values that backtrack to
+satisfy dependancies and have some flavor of continuations.
+
+\paragraph*{Fractional Types.} What is even more fascinating is that
+the above computation can be realized with a completely different
+concurrent schedule as follows, with fractional types.
+
+\begin{center}
+\scalebox{1.1}{
+\includegraphics{diagrams/thesis/algebra-wire2.pdf}
+}
+\end{center}
+
+%% The execution begins at line~(0) with the value {{20}}, at line (2)
+%% with the value {{right (right 20)}} flowing forwards and the value
+%% {{right (left 20)}} flowing backwards, and at line (3) with the value
+%% {{left (left 20)}} flowing forwards and the value {{left (right 20)}}
+%% flowing backwards.
+
+%\emph{Simultaneously} 
+The execution begins at line~(0) with the value {{20}}. At line~(1)
+values {{20}} and {{()}} flow forward, corresponding to the type
+`{{a*1}}'. At line~(2) three independent subcomputations \emph{proceed
+  concurrently}: the intial value {{20}} flowing from line (1) and two
+entangled values, {{1/x}} and {{x}}, such that specifying one
+implicitly specifies the dependant other. These three subcomputations
+exist \emph{at the same time}. At line (3), the dependant {{1/x}} is
+fixed to be the value {{20}}, thereby making fixing values of all
+three branches.
+%% the positive value created at line (2) which can flow to the
+%% consumer; the negative value created at line (3) which can pay for the
+%% debt created at line (2); and the initial value from line (0) which
+%% can flow to the required forward-flowing value needed at line (3). 
+At some intuitive level, one might understand the situation as
+modeling a credit card payment in which money is created at line (2)
+for the consumer together with a corresponding debt. The producer's
+bank pays for the debt at line (3) and waits for the actual money to
+arrive from the producer.
+
+There are at least two fundamental points about this example that must be
 emphasized:
 \begin{itemize}
-\item As the example illustrates, negative types corresponds to ``debts.''
-  It would clearly be disastrous if debts could be deleted or
-  duplicated. This simple observation explains why negative types are much
-  simpler and much more appealing in a framework where information is
-  guaranteed to be preserved. In previous work that used negative types (see
-  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
-  constrain the propagation and use of negative values because the
-  surrounding computational framework is, generally speaking, careless in its
-  treatment of information.
-\item The appeal of negative types can already be gleaned from the credit
-  card analogy. The main reason credit card transactions are convenient is
-  because they disentangle the propagation of the resources (money) from the
-  propagation of the services. Not every transaction needs both the resources
-  and services to be brought together: it is sufficient to have a promise
-  that the demand for resources will be somehow satisfied, as long as the
-  infrastructure can be trusted with such promises. Computationally, this
-  means that negative types allow us to break the dependency between
-  producers and consumers and let each compute independently.
-\item Consider a value of type {{-a + b}}. The above discussion suggests that
-  this value demands {{a}} and produces {{b}}, i.e., that it is a
-  higher-order funtion mapping {{a}} to {{b}}. Dually and as the underlying
-  categorical semantics justifies (see Sec.~\ref{sec:neg}), the same value is
-  a continuation which produces {{b}} and then demands {{a}}, which
-  introduces a twist on how the function can be used: get the goods {{b}} now
-  and pay {{a}} later.
+\item As the example illustrates, negative types corresponds to
+  ``debts'' satified by backtracking and fractional types correspond
+  to debts satisfied by constraint propagation.  It would clearly be
+  disastrous if debts could be deleted or duplicated. This simple
+  observation explains why these types are much simpler and much more
+  appealing in a framework where information is guaranteed to be
+  preserved. In previous work that used negative types (see
+  Sec.~\ref{sec:related}), complicated mechanisms are typically needed
+  to constrain the propagation and use of negative values because the
+  surrounding computational framework is, generally speaking, careless
+  in its treatment of information.
+\item The appeal of negative and fractional types can already be
+  gleaned from the credit card analogy. The main reason credit card
+  transactions are convenient is because they disentangle the
+  propagation of the resources (money) from the propagation of the
+  services. Not every transaction needs both the resources and
+  services to be brought together: it is sufficient to have a promise
+  that the demand for resources will be somehow satisfied, as long as
+  the infrastructure can be trusted with such
+  promises. 
+%% Computationally, this means that negative types allow us
+%%   to break the dependency between producers and consumers and let each
+%%   compute independently.
+
+%% \item Consider a value of type {{-a + b}}. The above discussion suggests that
+%%   this value demands {{a}} and produces {{b}}, i.e., that it is a
+%%   higher-order funtion mapping {{a}} to {{b}}. Dually and as the underlying
+%%   categorical semantics justifies (see Sec.~\ref{sec:neg}), the same value is
+%%   a continuation which produces {{b}} and then demands {{a}}, which
+%%   introduces a twist on how the function can be used: get the goods {{b}} now
+%%   and pay {{a}} later.
 \end{itemize}
 
-\paragraph*{Fractional Types.} 
+\paragraph*{Functions and Dependancy Splitting.} 
 Consider the following realization of boolean negation written using a

### 2012-03-10 rational.tex
_

@@
   concurrently}: the intial value {{20}} flowing from line (1) and two
-entangled values, {{1/x}} and {{x}}, such that specifying one
-implicitly specifies the dependant other. These three subcomputations
-exist \emph{at the same time}. At line (3), the dependant {{1/x}} is
-fixed to be the value {{20}}, thereby making fixing values of all
-three branches.
+entangled values, let us call them {{1/x}} and {{x}}, such that
+specifying one implicitly specifies the dependant other. These three
+subcomputations exist \emph{at the same time}. At line (3), the
+dependant {{1/x}} is fixed to be the value {{20}}, thereby making
+fixing values of all three branches.
 %% the positive value created at line (2) which can flow to the

### 2012-03-10 rational.tex
intro.

@@
 specifying one implicitly specifies the dependant other. These three
-subcomputations exist \emph{at the same time}. At line (3), the
+subcomputations \emph{exist at the same time}. At line (3), the
 dependant {{1/x}} is fixed to be the value {{20}}, thereby making
@@
 bank pays for the debt at line (3) and waits for the actual money to
-arrive from the producer.
-
-There are at least two fundamental points about this example that must be
-emphasized:
+arrive from the producer.  There are at least two fundamental points
+about this example that must be emphasized:
 \begin{itemize}
@@
 \paragraph*{Functions and Dependancy Splitting.} 
-Consider the following realization of boolean negation written using a
-pattern-matching style:
-\begin{verbatim}
-notB b = case b of 
-  False -> True
-  True -> False
-\end{verbatim}
-In a functional language, we are not used to treating the pattern-matching
-clauses \verb|False -> True| and \verb|True -> False| as first-class entities
-but fractional types allow us to do just that! Informally, we can treat each
-of the pattern-matching clauses as a first-class value of type 
-{{(1/bool) * bool}}. 
-A small but complete example will illustrate the main aspects of this idea.
-
-Let \verb|color| be the datatype with three constructors \verb|R|, \verb|G|,
-and \verb|B| and let's try to construct a function that rotates the
-colors. In other words, we would like to build the following three
-pattern-matching clauses \verb|R -> G|, \verb|G -> B|, and \verb|B -> R|.  In
-the algebra of rational numbers we have that for each {{a}} that is not equal
-to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic identity
-corresponds to a type isomorphism {{1 <-> (1/a) * a}}. For the datatype
-\verb|color|, there are three possible witnesses for this isomorphism mapping
-{{()}} to either {{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the
-particular witness determined by further constraints encountered during the
+%% Consider the following realization of boolean negation written using a
+%% pattern-matching style:
+%% \begin{verbatim}
+%% notB b = case b of 
+%%   False -> True
+%%   True -> False
+%% \end{verbatim}
+%% In a functional language, we are not used to treating the pattern-matching
+%% clauses \verb|False -> True| and \verb|True -> False| as first-class entities
+%% but fractional types allow us to do just that! Informally, we can treat each
+%% of the pattern-matching clauses as a first-class value of type 
+%% {{(1/bool) * bool}}. 
+%% A small but complete example will illustrate the main aspects of this idea.
+
+%% Let \verb|color| be the datatype with three constructors \verb|R|, \verb|G|,
+%% and \verb|B| and let's try to construct a function that rotates the
+%% colors. In other words, we would like to build the following three
+%% pattern-matching clauses \verb|R -> G|, \verb|G -> B|, and \verb|B -> R|.
+
+In the algebra of rational numbers we have that for each {{a}} that is
+not equal to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic
+identity corresponds to a type isomorphism {{1 <-> (1/a) * a}}, whose
+witness maybe thought of as the `identity function' at {{a}}.  For
+example, for the datatype \verb#color = R|G|B#, there are three
+possible witnesses for this isomorphism mapping {{()}} to either
+{{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular
+witness determined by further constraints encountered during the
 computation.  The idea of variables that capture constraints that are
-satisfied later is at the heart of logic programming and hence the witness for
-the isomorphisms above is generally {{(1/alpha,alpha)}} where~{{alpha}} is a
-fresh logic variable.
-
-Returning to our goal of creating the function that rotates the colors, we
-proceed as follows:
+satisfied later is at the heart of logic programming and hence the
+witness for the isomorphisms above is generally {{(1/alpha,alpha)}}
+where~{{alpha}} is a fresh logic variable.
+
+%% Returning to our goal of creating the function that rotates the colors, we
+%% proceed as follows:
+
 \begin{itemize}
-\item Using the fact that {{1}} is the multiplicative unit, generate from the
-  input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
-\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
-  components of the above tuple. The resulting value is 
-{{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
-{{alpha2}}, and {{alpha3}} are fresh logic variables.
+\item Using the fact that {{1}} is the multiplicative unit, generate
+  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
+\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
+  the components of the above tuple. The resulting value is
+  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}}, and
+  {{alpha2}} are fresh logic variables.
 \item Using the fact that {{*}} is associative and commutative, we can
@@
 
-{{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
-\item At this point, we have produced the skeleton for our desired function.
-  Assume now that a client uses the first clause of the function in a context
-  that contains the value \verb|R|, i.e., the client has the value
-  {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
-  and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
-  At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
-  the first component of the pair which has the following effects:
-  \begin{itemize}
-    \item the logic variable {{alpha1}} is unified with {{R}};
-    \item the first component of the pair then reduces to () and can be
-      absorbed by the fact that {{1}} is the multiplicative unit. In other
-      words, the value reduces to simply {{alpha2}} reflecting that the
-      pattern-matching was successful and the result of the clause is now
-      available;
-    \item the first clause of the function has now disappeared and the
-      remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
-  \item The same client or another client can now use the second clause
-    applying it to {{Gr}} which would produce the result {{alpha3}} for the
-    second client and ground the result of the first client to {{Gr}}.
-  \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
-    which grounds the result of the second client to {{B}}.
+{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.
+
+%% \begin{itemize}
+%% \item Using the fact that {{1}} is the multiplicative unit, generate from the
+%%   input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
+%% \item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
+%%   components of the above tuple. The resulting value is 
+%% {{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
+%% {{alpha2}}, and {{alpha3}} are fresh logic variables.
+%% \item Using the fact that {{*}} is associative and commutative, we can
+%%   rearrange the above tuple to produce the value:
+
+%% {{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
+
+%% \item At this point, we have produced the skeleton for our desired function.
+%%   Assume now that a client uses the first clause of the function in a context
+%%   that contains the value \verb|R|, i.e., the client has the value
+%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
+%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
+%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
+%%   the first component of the pair which has the following effects:
+%%   \begin{itemize}
+%%     \item the logic variable {{alpha1}} is unified with {{R}};
+%%     \item the first component of the pair then reduces to () and can be
+%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
+%%       words, the value reduces to simply {{alpha2}} reflecting that the
+%%       pattern-matching was successful and the result of the clause is now
+%%       available;
+%%     \item the first clause of the function has now disappeared and the
+%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+%%   \item The same client or another client can now use the second clause
+%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
+%%     second client and ground the result of the first client to {{Gr}}.
+%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
+%%     which grounds the result of the second client to {{B}}.
+%%   \end{itemize}
 \end{itemize}
-\end{itemize}
+
+At this point we have constructed a strange mixing of two functions;
+`inputs' of one form `outputs' of the other. If {{(1/alpha1,alpha2)}}
+is held by one subcomputation and {{1/alpha2,alpha1)}} is held by
+another subcomputation, these `remixed functions' form a communication
+channel between the subcomputations -- unifying {{1/alpha1}} with
+{{R}} in one subcomputation fixes it in the other. Depending on the
+choice of the type {{a}}, an arbitrary number of bits may be
+communicated. The dual reading of the above, with additive functions
+of the form {{0<-> ((-a)+b)}}, correspond to control flow and
+corotuining between two subcomputations -- only one of which `exists'
+at any point.
+
 It should be evident that in a universe in which information is not
-guaranteed to be preserved by the computational infrastructure, the above
-slicing and dicing of functions as a collection of first-class pattern
-clauses would make no sense. However, in a computational infrastructure that
-is guaranteed to preserve information, computations such as the one above are
-guaranteed to ``make sense.'' If one branch uses a clause that matches one
-part of the input to produce one part of the output, all other clauses are
-automatically adjusted to account for this to ensure that this input is not
-matched again and that this output is not produced again. Furthermore, just
-like the case for negative types, a dual reading of the pattern-matching
-clauses as continuations allows the context to use the result of the clause
-and provide the matching argument later.
+guaranteed to be preserved by the computational infrastructure, the
+above slicing and dicing of functions would make no sense. However, in
+a computational infrastructure that is guaranteed to preserve
+information, computations such as the one above are guaranteed to
+``make sense.''
+
+%% If one branch uses a clause that matches one part of the input to
+%% produce one part of the output, all other clauses are automatically
+%% adjusted to account for this to ensure that this input is not matched
+%% again and that this output is not produced again.
+
+%% \item At this point, we have produced the skeleton for our desired function.
+%%   Assume now that a client uses the first clause of the function in a context
+%%   that contains the value \verb|R|, i.e., the client has the value
+%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
+%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
+%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
+%%   the first component of the pair which has the following effects:
+%%   \begin{itemize}
+%%     \item the logic variable {{alpha1}} is unified with {{R}};
+%%     \item the first component of the pair then reduces to () and can be
+%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
+%%       words, the value reduces to simply {{alpha2}} reflecting that the
+%%       pattern-matching was successful and the result of the clause is now
+%%       available;
+%%     \item the first clause of the function has now disappeared and the
+%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+%%   \item The same client or another client can now use the second clause
+%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
+%%     second client and ground the result of the first client to {{Gr}}.
+%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
+%%     which grounds the result of the second client to {{B}}.
+%%   \end{itemize}
+
+%% Furthermore, just like the case for negative types, a dual reading of
+%% the pattern-matching clauses as continuations allows the context to
+%% use the result of the clause and provide the matching argument later.
 

### 2012-03-10 rational.tex
intro

@@
 `inputs' of one form `outputs' of the other. If {{(1/alpha1,alpha2)}}
-is held by one subcomputation and {{1/alpha2,alpha1)}} is held by
+is held by one subcomputation and {{(1/alpha2,alpha1)}} is held by
 another subcomputation, these `remixed functions' form a communication
 channel between the subcomputations -- unifying {{1/alpha1}} with
-{{R}} in one subcomputation fixes it in the other. Depending on the
-choice of the type {{a}}, an arbitrary number of bits may be
-communicated. The dual reading of the above, with additive functions
-of the form {{0<-> ((-a)+b)}}, correspond to control flow and
-corotuining between two subcomputations -- only one of which `exists'
-at any point.
+\verb|color| {{R}} in one subcomputation, fixes it in the
+other. Depending on the choice of the type {{a}}, an arbitrary number
+of bits may be communicated. The dual reading of the above, with
+additive functions of the form {{0<-> (-a)+b}}, correspond to
+control flow and corotuining between two subcomputations -- only one
+of which `exists' at any point.
 

### 2012-03-10 rational.tex
its flaky, but SAT is done.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{SAT Solver in {{langRevEE}} }
+\section{Advanced Example: SAT Solver in {{langRevEE}} }
 \label{sec:prog}
-
-%%%%%%%%%%%%%%%%%%%
-% \subsection{SAT solver}
 \label{sec:sat-solver}
+
+Additive traces correspond to iteration and multiplicative traces
+correspond to fixpoint of constraints. Addtive traces have been
+studied in the past \cite{infeffects} and here we focus on the
+multiplicative trace.
 
@@
 solution following which an isomorphic clone operation allows us to
-examine the result.
-
-The construction here is not limited to boolean satisfiability, but
-generalizes to any constraint satisfaction problem whose search space
-can be represented by a recursive type and for which an isomorphic
-clone operation can be constructed. That said, we will focus on the
-limited case of boolean satisfiability here.
-
-Consider the circuit below that has no observable output -- i.e. for
-all inputs it is annihilating. This happens because it constructs a
-boolean and its dual, negates one of them and attempts to collapse
-them.
+examine the result.  Though we focus on SAT here, this construction
+can be generalized to any constraint satisfaction problem whose search
+space can be represented by a recursive type and for which an
+isomorphic clone operation can be constructed. Before we start, let us
+recall two constructions:
+
+\begin{enumerate}
+\item Recall from previous work that for the types {{bool}} and, by
+  extension, any tuple of {{bool}}, an isomorphic clone operation can
+  be constructed (See Lemma 7.2 \cite{infeffects}).
+
+\item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
+  {{if_{not} }}, which when generalized to two bits was
+  {{toffoli}}. This construction maybe generalised to {{n}} bits as
+  {{if^n_{not} }}.
+\end{enumerate}
+
+%% That said, we will focus on the limited case of boolean satisfiability
+%% here.
+
+\emph{A Total Annihilation Circuit.}  Consider the circuit below that
+has no observable output -- i.e. for all inputs it is
+annihilating. This happens because it constructs a boolean and its
+dual, negates one of them and attempts to collapse them.
+
 \begin{center}
@@
 
+\emph{Controlled Annihilation.}
 We leverage the basic idea of the above circuit, to collapse all
@@
 \begin{center}
-\scalebox{1.5}{
+\scalebox{1.2}{
   \includegraphics{diagrams/sat1.pdf}
@@
 
-The final detail we have to be concerned with is the handling of the
-heap and garbage of {{f}}. Any boolean expression {{bool^n -> bool}}
-can be compiled into the isomorphism {{f:bool^h*bool^n<->bool^g*bool}}
-where the extra bits {{bool^h}} and {{bool^g}} are the heap and
-garbage. Constructing such an isomorphism has been detailed before
-\cite{Toffoli:1980,infeffects}.  However the computation of {{f}} is
-meaningful {{iff}} the heap {{bool^h}} is initialized appropriately. We
-can ensure that the heap has the appropriate initial value by checking
-the heap and negating a second control wire, if the values do not
-match.
-
-\begin{center}
-\scalebox{1.5}{
+\emph{Controlling the heap.}  The final detail we have to be concerned
+with is the handling of the heap and garbage of {{f}}. Any boolean
+expression {{bool^n -> bool}} can be compiled into the isomorphism
+{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
+{{bool^g}} are the heap and garbage. Constructing such an isomorphism
+has been detailed before \cite{Toffoli:1980,infeffects}.  However the
+computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
+initialized appropriately. We can ensure that the heap has the
+appropriate initial value by checking the heap and negating a second
+control wire, if the values do not match -- i.e. the dotted part in
+the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
+`heap control wire'.
+
+\begin{center}
+\scalebox{1.2}{
   \includegraphics{diagrams/sat2.pdf}
@@
 
-\roshan{Explain how the dotted part is implemented.}
-
-Let us call this construction which maps inputs, heap and control
-wires to input, heap and control wires as {{sat_f}}. The required
-contraint solver is then constructed as shown:
+% \roshan{Explain how the dotted part is implemented.}
+
+\emph{Cloning the Solution.}.  Let us call the above construction
+which maps inputs, heap and control wires to input, heap and control
+wires as {{sat_f}}. The SAT-solver is is completed by tracing the
+{{sat_f}} and cloning the `inputs'.
 
@@
 
-%%%%%%%%%%%%%%%%%%%
-\subsection{GoI machine}
-
-We can now encode the GoI machine of
-Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. \amr{The machine uses bang
-  which allows arbitrary duplication. Can we really do that???}
+When the solver is fed inputs initialized to {{true}}, the {{clone}}
+will clone any inputs to {{sat_f}} that satisfies the constraints
+imposed by {{f}} and the constraints on the {{heap}}. In the case of
+unique-SAT the solver will produce exactly 1 or 0 solutions. In the
+case of general SAT the solver will produce solutions as determined by
+the top-level (Sec. \ref{sec:rat}).
+
+%% %%%%%%%%%%%%%%%%%%%
+%% \subsection{GoI machine}
+
+%% We can now encode the GoI machine of
+%% Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. \amr{The machine uses bang
+%%   which allows arbitrary duplication. Can we really do that???}
 

### 2012-03-10 rational.tex
moved two constructions to teh section on generic constructions
constructions
negative tpes and constructions

@@
 negative types denoted {{-b}}, negative values denoted {{-v}}, and two
-isomorphisms {{eta+}} and {{eps+}}:
+isomorphisms {{eta+}} and {{eps+}} with the following type judgements:
 
@@
 
-The isomorphisms {{eta+}} and {{eps+}} witness the following
-isomorphism:
+\vspace{-15pt}
+\begin{multicols}{2}  
 %subcode{opsem} include main
 % eta+ &: 0 <-> (-b) + b :& eps+
-And as expected we have the following typing rule for negative value:
+
 %subcode{proof} include main
@@
 %@@ |- -v : -b
+\end{multicols}
 
@@
 \begin{center}
+\scalebox{0.8}{
   \includegraphics{diagrams/eta.pdf}
-\end{center}
-
-\begin{center}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.8}{
  \includegraphics{diagrams/eps.pdf}
+}
 \end{center}
@@
 direction is captured by the negative sign on the value.
-As an example, the diagram representing the isomorphism from the introduction
-is:
-\begin{center}
-  \includegraphics{diagrams/algebra1.pdf}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Compact Closure with Additive Inverses}
-
-\paragraph*{Coherence.}
-The appropriate coherence condition for Compact Closed Categories,
-expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-may be visualized as explained by Selinger
-\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-\paragraph*{Trace}
-
-%subcode{proof} include main
-%@ c : b2 + b1 <-> b2 + b3
-%@@ trace c : b1 <-> b3
-
-\begin{center}
-  \includegraphics{diagrams/trace.pdf}
-\end{center}
-
-\paragraph*{Involution (Principium Contradictiones)}
-
-{{b <-> -(-b)}}
-
-\begin{center}
-  \includegraphics{diagrams/double_neg.pdf}
-\end{center}
-
-\paragraph*{Higher-Order Functions}
-
-Mention LL and {{a -o b}}.
-
-\begin{center}
-  \includegraphics{diagrams/function.pdf}
-\end{center}
-
-Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
-
-\roshan{Also: additive functions aren't values - they are
-  possibilities. Multiplicative functions really are values.}
-
-\paragraph*{Function application}
-
-\begin{center}
-  \includegraphics{diagrams/apply1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/apply2.pdf}
-\end{center}
-
-\paragraph*{Function composition}
-
-\begin{center}
-  \includegraphics{diagrams/compose1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/compose.pdf}
-\end{center}
-
-This is also equivalent to sequencing both the computation blocks.
-
-\begin{center}
-  \includegraphics{diagrams/compose2.pdf}
-\end{center}
-
-\paragraph*{Delimited continuation}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}
-
-%%%%%%%%%%%%%
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.1}{
+  \includegraphics{diagrams/eps_plus1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.1}{
+  \includegraphics{diagrams/eps_plus2.pdf}
+}
+\end{center}  
+\end{multicols}
 
@@
 circuit into a left-to-right flow.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus1.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus2.pdf}
-}
-\end{center}  
-\end{multicols}
+As an example, the diagram representing the isomorphism from the
+introduction is:
+\begin{center}
+  \includegraphics{diagrams/algebra1.pdf}
+\end{center}
 
@@
 
+%\subsection{Compact Closure with Additive Inverses}
+\paragraph*{Coherence for Compact Closure with Additive Inverses.}
+The appropriate coherence condition for Compact Closed Categories,
+expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
+may be visualized as explained by Selinger
+\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
 \paragraph*{Non-termination.}
@@
 types are not executable.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions in the Additive Monoid}
+
+\begin{enumerate}
+\item 
+\emph{Trace}
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
+\end{multicols}
+
+\item
+\emph{Involution (Principium Contradictiones)}
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+\item
+\emph{Functions}
+
+Mention LL and {{a -o b}}.
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
+
+\roshan{Also: additive functions aren't values - they are
+  possibilities. Multiplicative functions really are values.}
+
+\item
+\emph{Function composition.}
+
+\begin{multicols}{2}
+    \includegraphics{diagrams/compose1.pdf}
+
+    \includegraphics{diagrams/compose.pdf}
+\end{multicols}
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+\item
+\emph{Function application.}
+test.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/apply1.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/apply2.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item
+\emph{Delimited continuation}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+\item
+\emph{Negation distributes over {{+}}. }
+
+{{-(b1 + b2) <-> (-b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+\item
+\emph{Lifting a operation of positive types to negated types}
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
+\end{enumerate}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions in {{langRevEE}} }
-\label{sec:monoidal-constructions}
+%% %%%%%%%%%%%%%%%%%%%%%%
+%% \subsection{Constructions in {{langRevEE}} }
+%% \label{sec:monoidal-constructions}
 
 %%%%%%%%%%%%%%%%%%
-\subsection{Constructions Specific to Negation}
+\subsection{Constructions in {{langRevEE}} specific to Addition and Multiplication}
 \label{sec:neg-constructions}
 
-\paragraph*{Negation distributes over {{+}}. }
-
-{{-(b1 + b2) <-> (-b1) + (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/dist_neg_plus.pdf}
-\end{center}
-
-\paragraph*{ {{eps_{fst} }} }
+\begin{enumerate}
+
+\item 
+\emph{ {{eps_{fst} }} }
 
@@
 
-\paragraph*{Lifting negation out of {{*}}. }
+\item 
+\emph{Lifting negation out of {{*}}. }
 
@@
 
-\paragraph*{Lifting a operation of positive types to negated types}
-
-Given {{c : b1 <-> b2}}
-
-\begin{center}
-  \includegraphics{diagrams/neg_lift.pdf}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions Specific to Multiplication}
-
-\paragraph*{Distributing fractions over {{*}} }
-
-{{1/(b1*b2) <-> (1/b1) * (1/b2)}}
-
-\paragraph*{Multiplying Fractions}
+%% \emph{Distributing fractions over {{*}} }
+%% {{1/(b1*b2) <-> (1/b1) * (1/b2)}}
+
+%% \emph{Lifting Computation intro fractions}
+%% {{b1 <-> b2 ==> 1/b1 <-> 1/b2}}
+
+\item 
+\emph{Multiplying Fractions}
 
@@
 
-\paragraph*{Lifting Computation intro fractions}
-
-{{b1 <-> b2 ==> 1/b1 <-> 1/b2}}
-
-\paragraph*{Adding the numerator}
+\item 
+\emph{Adding the numerator}
 
@@
 
-\paragraph*{Fraction Addition}
+\item 
+\emph{Fraction Addition}
 
 {{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
+
+\end{enumerate}
 

### 2012-03-10 rational.tex
_
HOF subsection

@@
 
+It is not fair to say that negative types flow backwards. The
+following circuits are valid in {{langRevEE}}. It is however proper to
+say that for any type {{b}} that flows in a direction, the type {{-b}}
+flows in the reverse direction.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+test.
+
 \begin{multicols}{2}
@@
 
+test.
+
 \item
@@
 \begin{multicols}{2}
+\scalebox{1.1}{
     \includegraphics{diagrams/compose1.pdf}
-
+}
+
+\scalebox{0.8}{
     \includegraphics{diagrams/compose.pdf}
+}
 \end{multicols}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{To think about}
-
-\begin{itemize}
-
-\item Can the operational semantics for {{langRevEE}} be an
-  interpreter that is implemented in {{langRevT}} (similar to how the
-  the tree traversal interpreter was implemented). This would imply
-  the existence of a more powerful construction than Int, wherein the
-  products would also be preserved. This is possibly worth a paper in
-  itself.
-
-\item These functions aren't really values. There is no value one can
-  produce that denotes a function. These functions are the ability to
-  transform a value - the possibility of transforming a value.  In the
-  product encoding of environments, there is no value one can assign
-  to a variable such that it denotes a function.
-
-Actually this is possible. Consider two functions {{f : b1 <-> b2}}
-and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
-discriminate them. Hence the {{boo}} is the first class representation
-of the functions and can be thought of as the address of the function.
-
-\begin{center}
-  \includegraphics{diagrams/dispatch.pdf}
-\end{center}
-
-\item It is not fair to say that negative types flow backwards. The
-  following circuits are valid in {{langRevEE}}. It is however proper
-  to say that for any type {{b}} that flows in a direction, the type
-  {{-b}} flows in the reverse direction.
-
-\begin{center}
-  \includegraphics{diagrams/neg_circuit1.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/neg_circuit2.pdf}
-\end{center}
-
-\end{itemize}
+%% %%%%%%%%%%%%%%%%%%%%%%
+%% \subsection{To think about}
+
+%% \begin{itemize}
+
+%% \item Can the operational semantics for {{langRevEE}} be an
+%%   interpreter that is implemented in {{langRevT}} (similar to how the
+%%   the tree traversal interpreter was implemented). This would imply
+%%   the existence of a more powerful construction than Int, wherein the
+%%   products would also be preserved. This is possibly worth a paper in
+%%   itself.
+
+%% \item 
+%% \item
+
+%% \begin{center}
+%%   \includegraphics{diagrams/neg_circuit1.pdf}
+%% \end{center}
+
+%% \begin{center}
+%%   \includegraphics{diagrams/neg_circuit2.pdf}
+%% \end{center}
+
+%% \end{itemize}
 
@@
 
+\subsection{Higher Order Functions}
+
+These functions aren't really values. There is no value one can
+produce that denotes a function. These functions are the ability to
+transform a value - the possibility of transforming a value.  In the
+product encoding of environments, there is no value one can assign to
+a variable such that it denotes a function.
+
+Actually this is possible. Consider two functions {{f : b1 <-> b2}}
+and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
+discriminate them. Hence the {{boo}} is the first class representation
+of the functions and can be thought of as the address of the function.
+
+%% \begin{center}
+%%   \includegraphics{diagrams/dispatch.pdf}
+%% \end{center}
+
+\subsection{Other}
 To summarize negative, fractional, square root, and imaginary types all make

### 2012-03-10 rational.tex
reorg

@@
 Multiplicative duality is added to {{langRev}} by adding types
-{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.
+{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}} with
+the following type judgements.
 
@@
 
-\noindent
-These have the type judgments
-
+\vspace{-15pt}
+\begin{multicols}{2}
 %subcode{opsem} include main
@@
 %@@ |- 1/v : 1/b  
+\end{multicols}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Compact Closure with Multiplicative Inverses}
-
-\paragraph*{Coherence.} 
+\paragraph*{Coherence for Compact Closure with Multiplicative Inverses} 
 As before, the requisite coherence condition on Compact Closed
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions in {{langRevEE}} specific to Addition and Multiplication}
+\label{sec:neg-constructions}
+
+\begin{enumerate}
+
+\item 
+\emph{ {{eps_{fst} }} }
+
+{{(-b1)*b2 + b1*b2 <-> 0}}
+
+\begin{center}
+  \includegraphics{diagrams/eps_fst.pdf}
+\end{center}
+
+\item 
+\emph{Lifting negation out of {{*}}. }
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/mult_neg.pdf}
+\end{center}
+
+The following isomorphism can be constructed similarly
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
+
+%% \emph{Distributing fractions over {{*}} }
+%% {{1/(b1*b2) <-> (1/b1) * (1/b2)}}
+
+%% \emph{Lifting Computation intro fractions}
+%% {{b1 <-> b2 ==> 1/b1 <-> 1/b2}}
+
+\item 
+\emph{Multiplying Fractions}
+
+{{b1/b2 * b3/b4 <-> (b1*b3)/(b2*b4)}}
+
+\item 
+\emph{Adding the numerator}
+
+{{b1/b + b2/b <-> (b1+b2)/b}}
+
+\item 
+\emph{Fraction Addition}
+
+{{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Computing in the Field of Rationals : {{langRevEE}} }
@@
 
-%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Unification and Logic Programming}
@@
 %% \label{sec:monoidal-constructions}
-
-%%%%%%%%%%%%%%%%%%
-\subsection{Constructions in {{langRevEE}} specific to Addition and Multiplication}
-\label{sec:neg-constructions}
-
-\begin{enumerate}
-
-\item 
-\emph{ {{eps_{fst} }} }
-
-{{(-b1)*b2 + b1*b2 <-> 0}}
-
-\begin{center}
-  \includegraphics{diagrams/eps_fst.pdf}
-\end{center}
-
-\item 
-\emph{Lifting negation out of {{*}}. }
-
-{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/mult_neg.pdf}
-\end{center}
-
-The following isomorphism can be constructed similarly
-
-{{b1 * b2 <-> (-b1)*(-b2)}}
-
-%% \emph{Distributing fractions over {{*}} }
-%% {{1/(b1*b2) <-> (1/b1) * (1/b2)}}
-
-%% \emph{Lifting Computation intro fractions}
-%% {{b1 <-> b2 ==> 1/b1 <-> 1/b2}}
-
-\item 
-\emph{Multiplying Fractions}
-
-{{b1/b2 * b3/b4 <-> (b1*b3)/(b2*b4)}}
-
-\item 
-\emph{Adding the numerator}
-
-{{b1/b + b2/b <-> (b1+b2)/b}}
-
-\item 
-\emph{Fraction Addition}
-
-{{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
-
-\end{enumerate}
-
-%% %%%%%%%%%%%%%%%%%%%%%%
-%% \subsection{To think about}
-
-%% \begin{itemize}
-
-%% \item Can the operational semantics for {{langRevEE}} be an
-%%   interpreter that is implemented in {{langRevT}} (similar to how the
-%%   the tree traversal interpreter was implemented). This would imply
-%%   the existence of a more powerful construction than Int, wherein the
-%%   products would also be preserved. This is possibly worth a paper in
-%%   itself.
-
-%% \item 
-%% \item
-
-%% \begin{center}
-%%   \includegraphics{diagrams/neg_circuit1.pdf}
-%% \end{center}
-
-%% \begin{center}
-%%   \includegraphics{diagrams/neg_circuit2.pdf}
-%% \end{center}
-
-%% \end{itemize}
 

### 2012-03-10 rational.tex
HOF

@@
 
-\newcommand{\lcal}{\ensuremath{\lambda}-calculus}
+\newcommand{\lcal}{\ensuremath{\lambda}-calculus\xspace}
 \newcommand{\G}{\ensuremath{\mathcal{G}}\xspace}
@@
 %! ~> = \leadsto
+%! -o+ = \multimap^{+}
+%! -o* = \multimap^{\times}
 %! -o = \multimap
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Computing in the Field of Algebraic Numbers}
+\section{Future Work}
+
+Several things we don't know how to do, or are sloppy about. Some of
+these are deep questions to which there is no immeidate answer.
+
+\begin{itemize}
+
+\item Is the language without recursive types always terminating? For
+  the right symmetry with annihilation, it is desirable to have
+  non-termination.
+
+\item What is the correct lemma for logical reversibility now?
+
+\item We should say things about information preservation. At least
+  more so than what is being said now.
+
+\item Multiplicative duality gives us a means of talking about
+  equivalence classes of values. This should correspond to all sorts
+  of other things -- groupoids, Lawvere theories etc. We barely even
+  understand these connections.
+
+\item How does this connect to QC? Is there any strong connection such
+  that the SAT solver here is an alternate QC algorithm to Shor's?
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Conventional Higher Order Functions}
+
+In a conventional language, higher-order functions capture scope and
+can be represented as first class values. On the other hand,
+{{langRevEE}} gives us two notions of functions both of which share
+similarities with conventional HOFs.
+
+\begin{itemize}
+\item 
+Additive functions are not really values, but are possible paths of
+execution. However, they can be encoded as values as follows: Consider
+two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
+{{bool}} can discriminate them and hence a value of type {{bool}} acts
+as the `address' of the function and is a first class value
+representing one of the {{b1 -o+ b2}} functions.  This is similar to
+how first classes functions a'la \lcal are represented by their
+addresses when compiled to conventional hardware.
+
+\item
+Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+are indeed first class values. However these functions can only be
+linearly used -- i.e. they represent a contraint that can be satisfied
+at most once in one `computational world'. It may be possible that to
+repeatedly invoke multiplicative functions with an appropriate
+recursive type whose unfoldings allow for repeated unification.
+\end{itemize}
+
+The exact encoding on HOF in {{langRevEE}} should shed light on
+several problems, including the connections to continuations discussed
+in Sec. \ref{sec:related}. They should also allow for a natural
+encoding of HOFs in the translations of \lcal to {{langRevT}} and
+indicate a monadic encapsulation of of information effects which are
+currrently structured as an arrow.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Computing in the Field of Algebraic Numbers}
 
@@
 
-Several things we don't know how to do, or are sloppy about. Some of
-these are deep questions to which there is no immeidate answer.
-
-\begin{itemize}
-
-\item Is the language without recursive types always terminating? For
-  the right symmetry with annihilation, it is desirable to have
-  non-termination.
-
-\item What is the correct lemma for logical reversibility now?
-
-\item We should say things about information preservation. At least
-  more so than what is being said now.
-
-\item Do we really have HOF and continuations? What is the exact
-  relationship? Do we have linearly used functions? Do we have linear
-  functions (in that they use their argument linearly)?
-
-\item Can we complete the construction in \cite{infeffects} to be full
-  STLC? This is again a HOF question in another guise.
-
-\item What are information effects in {{langRevEE}}? Depending on what
-  we think about HOF, do we have a monad or is it still an arrow?
-
-\item Multiplicative duality gives us a means of talking about
-  equivalence classes of values. This should correspond to all sorts
-  of other things -- groupoids, Lawvere theories etc. We barely even
-  understand these connections.
-
-\item How does this connect to QC? Is there any strong connection such
-  that the SAT solver here is an alternate QC algorithm to Shor's?
-
-\end{itemize}
-
-\subsection{Higher Order Functions}
-
-These functions aren't really values. There is no value one can
-produce that denotes a function. These functions are the ability to
-transform a value - the possibility of transforming a value.  In the
-product encoding of environments, there is no value one can assign to
-a variable such that it denotes a function.
-
-Actually this is possible. Consider two functions {{f : b1 <-> b2}}
-and {{g : b1 <-> b2}}. A value of type {{bool}} is sufficient to
-discriminate them. Hence the {{boo}} is the first class representation
-of the functions and can be thought of as the address of the function.
-
 %% \begin{center}
@@
 
-\subsection{Other}
+% \subsection{Other}
+
 To summarize negative, fractional, square root, and imaginary types all make

### 2012-03-10 rational.tex
more on functions

@@
 (detailed in Section \ref{sec:monoidal-constructions}) hence
-introducing \emph{non-termination} into the language
-\cite{infeffects}.
+introducing \emph{non-termination} into the language (in the presence
+of recursive types) \cite{infeffects}.
 
@@
 
-Let us use the shorthand {{b1 -o b2 = -b1 + b2}}
-
-\roshan{Also: additive functions aren't values - they are
-  possibilities. Multiplicative functions really are values.}
+Let us use the shorthand {{b1 -o+ b2 = 0 <-> -b1 + b2}}. See
+Sec. \ref{sec:hof} for a discussion of the additive functions
+vs. multiplicative functions.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions in {{langRevEE}} specific to Addition and Multiplication}
-\label{sec:neg-constructions}
+\subsection{Constructions specific to Negatives and Fractions}
+\label{sec:specific-constructions}
+
+While the constructions in Sec \ref{sec:monoidal-constructions} apply
+to both the {{(0,+)}} and {{(1,*)}} monoids, each construction
+presented here applies to a specific monoid.
 
@@
 \subsection{Conventional Higher Order Functions}
+\label{sec:hof}
 

### 2012-03-10 rational.tex
reorg
algebraics

@@
 
-once we combine the two structures, we seem to retain the monoidal
-structure. How!?
-
 In this section we develop the semantics for {{langRev}} extended with
@@
 %%   which allows arbitrary duplication. Can we really do that???}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion}
+
+-- WAVEFRONT START --
+
+Several things we don't know how to do, or are sloppy about. Some of
+these are deep questions to which there is no immeidate answer.
+
+\begin{itemize}
+
+\item Is the language without recursive types always terminating? For
+  the right symmetry with annihilation, it is desirable to have
+  non-termination.
+
+\item What is the correct lemma for logical reversibility now?
+
+\item We should say things about information preservation. At least
+  more so than what is being said now.
+
+\item Multiplicative duality gives us a means of talking about
+  equivalence classes of values. This should correspond to all sorts
+  of other things -- groupoids, Lawvere theories etc. We barely even
+  understand these connections.
+
+\item How does this connect to QC? Is there any strong connection such
+  that the SAT solver here is an alternate QC algorithm to Shor's?
+
+\end{itemize}
+
+-- WAVEFRONT STOP --
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Conventional Higher Order Functions}
+\label{sec:hof}
+
+In a conventional language, higher-order functions capture scope and
+can be represented as first class values. On the other hand,
+{{langRevEE}} gives us two notions of functions both of which share
+similarities with conventional HOFs.
+
+\begin{itemize}
+\item 
+Additive functions are not really values, but are possible paths of
+execution. However, they can be encoded as values as follows: Consider
+two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
+{{bool}} can discriminate them and hence a value of type {{bool}} acts
+as the `address' of the function and is a first class value
+representing one of the {{b1 -o+ b2}} functions.  This is similar to
+how first classes functions a'la \lcal are represented by their
+addresses when compiled to conventional hardware.
+
+\item
+Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+are indeed first class values. However these functions can only be
+linearly used -- i.e. they represent a contraint that can be satisfied
+at most once in one `computational world'. It may be possible that to
+repeatedly invoke multiplicative functions with an appropriate
+recursive type whose unfoldings allow for repeated unification.
+\end{itemize}
+
+The exact encoding on HOF in {{langRevEE}} should shed light on
+several problems, including the connections to continuations discussed
+in Sec. \ref{sec:related}. They should also allow for a natural
+encoding of HOFs in the translations of \lcal to {{langRevT}} and
+indicate a monadic encapsulation of of information effects which are
+currrently structured as an arrow.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Computing in the Field of Algebraic Numbers}
+
+The field of algebraic numbers consist of the rationals extended with
+solutions to polynomials. In a limited sense {{langRevEE}} extended
+with recursive types already include polynomials such as the
+definition of binary tree {{@@x.(1+x*x)}} implies the equality
+{{x=1+x*x}} and hence {{x^2-x+1=0}}.
+
+%% Partial isomorphisms algebraics
+%% do have computational
+
+%% While this equation does indeed have an irrational imaginary root,
+%% as we will see -- {{langRevEE}} does not have the means of
+%% expressing meaningful computation that uses the root.
+
+The famous example of isomorphism on algebraics is the embedding of
+seven binary trees in one. In Figure 1 of Fiore's paper
+\cite{Fiore:2004}, he demonstrates this isomorphism by the
+corrsponding algebraic manipulation of the types.  The algebraic
+manipulation can be viewed as a {{langRevEE}} combinator and it indeed
+does constructively witness the isomorphism -- i.e. you can run it and
+view one data structure encoded by the other. In this context,
+consider the following algebraically valid proof of the same
+isomorphism:
+
+%% Previously the most famous example of a similar nature is the puzzling
+%% isomorphism that one can establish between seven binary trees and one.
+%% A binary tree is defined by the datatype
+%% \[
+%% x = 1 + x * x 
+%% \]
+%% which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
+%% algebraic manipulation we can reason as follows:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+Fiore poses the question of why such algebraic manipulation would make
+sense type theoretically, even though some of the intermediate steps
+make no sense.  Stated in terms of {{langRevEE}} computations, this
+manipulation has no meaningful computational content -- it is always
+non-terminating. It suggests that some algebraic manipulations are
+somehow more `constructive' than others or dually that {{langRevEE}}
+with recursive types somehow lacks the ability to express all
+computations that are otherwise algebraically meaningful.
+
+-- WAVEFRONT START --
+
+But the interesting solution of trees is precisely that they involve
+irrational numbers. Our usual notation for irrationals such as
+{{sqrt(2) }} can actually be viewed as a thunk representing the
+infinite computation {{1.4142..}}.  Irrationals in this sense have a
+very different information theoretic interpretation from full real
+numbers.  With Chaitin's $\Omega$ numbers
+\cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
+real numbers that have an infinite amount of information -- ones for
+example are the solution to the question about the halting properties
+of every Turning machine.
+
+That introduces the natural numbers. No solution for this polynomial
+over the algebraic numbers. We must extend the numbers with $\omega$
+to get a solution. We reject this in this paper and prefer to stick to
+algebraic numbers. The advantage is that all the isomorphisms are
+valid numerically. With the above type we could subtract $b$ from both
+sides to show that $0=1$ which is nonsense. We can however have
+infinite types as long as they have algebraic solutions.
+
+-- WAVEFRONT STOP --
+
+%% the final equivalence is valid and can be used to actually
+%% construct an isomorphism between $x^7$ the type of seven binary
+%% trees and $x$ the type of binary trees.
+
+%% Here is one way to view computations involving irrationals that might
+%% shed some insight into how we should view recursive data types
+%% also. For finite datatypes say of size {{n}} we can say that an
+%% element specifies a choice of {{1/n}} and hence contains as much
+%% information. For finite things, each value is computationally equal to
+%% any other value. Any permutation is possible among the values. Hence
+%% there is no metric possible on the values.  Recursive data types, such
+%% as binary trees, have an infinite number of elements and hence the
+%% argument that says that value specifies {{1/n}} units of information
+%% is not longer valid. Instead to talk meaningfully about information
+%% content, we should talk about a probability distribution on the values
+%% of the type. However what shall we base this distribution on? Is there
+%% a natural order beyond the unfolding order of the values of a type?
+
+%% In this context, let us go back to the thought that recursive such as
+%% {{nat=nat+1}} have no meaningful solutions, whereas trees do.
+
+%% Real numbers in that sense really are a series of bits of unknown
+%% information theoretic interpretation -- a sort of {{top}}. Numbers
+%% such as {{sqrt(2) }} are very different in the sense that even though
+%% they contain an infinite sequence of bits, there is a finite program
+%% that generates those bits and its only the cost of computing bits that
+%% is infinite.  The consistency of arithmetic isn't affected by
+%% extending the rational field by adding any specific computable
+%% irrational. So it might be possible to design a model of computing
+%% where we can indeed deal with them. Such a model will have to explicit
+%% thunks for the delayed computation represented by trees and never
+%% equate and cancel two infinite computations unless the represent
+%% exactly the same infinite computation.
+
+%% TODO.
+
+%% Square root and imaginary types have also appeared in the literature: we
+%% relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
+%% explanation. We have so far extended the set of types to be the rational
+%% numbers. Now we will push this and extend the set of types to algebraic
+%% numbers. In other words, we will allow datatypes defined by arbitrary
+%% polynomials and allow the roots of such polynomials to be types.
+
+%% Consider first an example in which we want to compute with the sides of a
+%% rectangle whose area is 91 and whose length is longer than its width by 6
+%% units. One can solve the quadratic equation to determine that the sides are 7
+%% and 13 and proceed. This however prematurely forces us to globally solve the
+%% constraint. Instead we can let the two sides of the rectangle be $x$ and
+%% $x+6$ and use the following equation to capture the desired constraint:
+%% \[
+%% x^2 + 6x - 91 = 0
+%% \]
+%% The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
+%% the type $0$. We can now proceed to compute with the unknown $x$, being
+%% assured that in a closed program, our computation will eventually be
+%% consistent with the solution of the algebraic equation.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future Work}
-
-Several things we don't know how to do, or are sloppy about. Some of
-these are deep questions to which there is no immeidate answer.
-
-\begin{itemize}
-
-\item Is the language without recursive types always terminating? For
-  the right symmetry with annihilation, it is desirable to have
-  non-termination.
-
-\item What is the correct lemma for logical reversibility now?
-
-\item We should say things about information preservation. At least
-  more so than what is being said now.
-
-\item Multiplicative duality gives us a means of talking about
-  equivalence classes of values. This should correspond to all sorts
-  of other things -- groupoids, Lawvere theories etc. We barely even
-  understand these connections.
-
-\item How does this connect to QC? Is there any strong connection such
-  that the SAT solver here is an alternate QC algorithm to Shor's?
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Conventional Higher Order Functions}
-\label{sec:hof}
-
-In a conventional language, higher-order functions capture scope and
-can be represented as first class values. On the other hand,
-{{langRevEE}} gives us two notions of functions both of which share
-similarities with conventional HOFs.
-
-\begin{itemize}
-\item 
-Additive functions are not really values, but are possible paths of
-execution. However, they can be encoded as values as follows: Consider
-two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
-{{bool}} can discriminate them and hence a value of type {{bool}} acts
-as the `address' of the function and is a first class value
-representing one of the {{b1 -o+ b2}} functions.  This is similar to
-how first classes functions a'la \lcal are represented by their
-addresses when compiled to conventional hardware.
-
-\item
-Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-are indeed first class values. However these functions can only be
-linearly used -- i.e. they represent a contraint that can be satisfied
-at most once in one `computational world'. It may be possible that to
-repeatedly invoke multiplicative functions with an appropriate
-recursive type whose unfoldings allow for repeated unification.
-\end{itemize}
-
-The exact encoding on HOF in {{langRevEE}} should shed light on
-several problems, including the connections to continuations discussed
-in Sec. \ref{sec:related}. They should also allow for a natural
-encoding of HOFs in the translations of \lcal to {{langRevT}} and
-indicate a monadic encapsulation of of information effects which are
-currrently structured as an arrow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computing in the Field of Algebraic Numbers}
-
-The field of algebraic numbers consist of the rationals extended with
-solutions to polynomials. In a limited sense {{langRevEE}} with
-recursive types already include polynomials such as the definition of
-binary tree {{@@x.(1+x*x)}} implies the equality {{t = 1 + t*t}} and
-hence {{t^2-t+1=0}}. While this equation does indeed have an
-irrational imaginary root, as we will see -- {{langRevEE}} does not
-have the means of expressing meaningful computation that uses the
-root.
-
-Here is one way to view computations involving irrationals that might
-shed some insight into how we should view recursive data types
-also. For finite datatypes say of size {{n}} we can say that an
-element specifies a choice of {{1/n}} and hence contains as much
-information. For finite things, each value is computationally equal to
-any other value. Any permutation is possible among the values. Hence
-there is no metric possible on the values.  Recursive data types, such
-as binary trees, have an infinite number of elements and hence the
-argument that says that value specifies {{1/n}} units of information
-is not longer valid. Instead to talk meaningfully about information
-content, we should talk about a probability distribution on the values
-of the type. However what shall we base this distribution on? Is there
-a natural order beyond the unfolding order of the values of a type?
-
-In this context, let us go back to the thought that recursive such as
-{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
-interesting solution of trees is precisely that they involve
-irrational numbers. Our usual notation for irrationals such as
-{{sqrt(2) }} can actually be viewed as a thunk representing the
-infinite computation {{1.4142..}}. Irrationals in this sense have a
-very different information theoretic interpretation from full real
-numbers. With Chaitin's Omega numbers he postulates the existence of
-real numbers that have an infinite amount of information -- ones for
-example are the solution to the question about the halting properties
-of every Turning machine.
-
-Real numbers in that sense really are a series of bits of unknown
-information theoretic interpretation -- a sort of {{top}}. Numbers
-such as {{sqrt(2) }} are very different in the sense that even though
-they contain an infinite sequence of bits, there is a finite program
-that generates those bits and its only the cost of computing bits that
-is infinite.  The consistency of arithmetic isn't affected by
-extending the rational field by adding any specific computable
-irrational. So it might be possible to design a model of computing
-where we can indeed deal with them. Such a model will have to explicit
-thunks for the delayed computation represented by trees and never
-equate and cancel two infinite computations unless the represent
-exactly the same infinite computation.
-
-TODO.
-
-Square root and imaginary types have also appeared in the literature: we
-relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
-explanation. We have so far extended the set of types to be the rational
-numbers. Now we will push this and extend the set of types to algebraic
-numbers. In other words, we will allow datatypes defined by arbitrary
-polynomials and allow the roots of such polynomials to be types.
-
-Consider first an example in which we want to compute with the sides of a
-rectangle whose area is 91 and whose length is longer than its width by 6
-units. One can solve the quadratic equation to determine that the sides are 7
-and 13 and proceed. This however prematurely forces us to globally solve the
-constraint. Instead we can let the two sides of the rectangle be $x$ and
-$x+6$ and use the following equation to capture the desired constraint:
-\[
-x^2 + 6x - 91 = 0
-\]
-The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
-the type $0$. We can now proceed to compute with the unknown $x$, being
-assured that in a closed program, our computation will eventually be
-consistent with the solution of the algebraic equation.
-
-Previously the most famous example of a similar nature is the puzzling
-isomorphism that one can establish between seven binary trees and one.
-A binary tree is defined by the datatype
-\[
-x = 1 + x * x 
-\]
-which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
-algebraic manipulation we can reason as follows:
-\[\begin{array}{rclclclcl}
-x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
-x^6 &=& 1 \\
-x^7 &=& x^6 x &=& x
-\end{array}\]
-Fiore poses the question of why such algebraic manipulation would make sense
-type theoretically but states that even though some of the intermediate steps
-make no sense, the final equivalence is valid and can be used to actually
-construct an isomorphism between $x^7$ the type of seven binary trees 
-and $x$ the type of binary trees.
-
-Discussion of possible polynomials:
-\begin{itemize}
-\item $b=1+1$. Boring.
-\item $b=b+1$. That introduces the natural numbers. No solution for this
-  polynomial over the algebraic numbers. We must extend the numbers with
-  $\omega$ to get a solution. We reject this in this paper and prefer to
-  stick to algebraic numbers. The advantage is that all the isomorphisms are
-  valid numerically. With the above type we could subtract $b$ from both
-  sides to show that $0=1$ which is nonsense. We can however have infinite
-  types as long as they have algebraic solutions.
-\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
-  boolean! If we had superpositions, we could then write a function that
-  performs the square root of negation in the sense that applying it twice
-  would be equivalent to boolean negation.
-\end{itemize}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+once we combine the two structures, we seem to retain the monoidal
+structure. How!?
+
 %% Jacques Carette, Michael Adams

### 2012-03-10 rational.tex
_

@@
   corresponds to backtracking and a multiplicative duality that corresponds
-  to unification.
+  to constraint propagation.
 \end{abstract}
@@
 \subsection{Constructions in the Additive Monoid}
+\label{sec:monoidal-constructions}
 

### 2012-03-10 rational.tex
conclusion

@@
 
+In our previous work we introduce recursive types and trace
+operators. This is dangerous here because infinite loops allow us to
+prolong paying the debt for as long as we want. The algebraic
+$\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that
+extreme care is needed when combining negative and fractional types
+with the full $\lambda$-calculus (including recursion).
+
 -- WAVEFRONT STOP --
@@
 
+We have extended the language {{langRev}} that expressed computation
+in the commutative semiring of whole numbers to {{langRevEE}} that
+expresses computation in the fields of rationals.  Every algebraic
+identity that holds for the rational numbers corresponds to a type
+isomorphism in our model.  In the process, we have examined the two
+function spaces that these give rise to an have seen non-trivial
+constructions such as a novel SAT-solver that relies on a
+multiplicative trace.
+
+In another sense however, this paper is about the nature of duality in
+computation. The concept of duality is deep and significant and in
+many cases -- probably most famously in the divide between classical
+and intuinionistic mathematics -- contain non-trivial assumptions
+about our world-view.
+
+The problem of duality in computation and logic, has many facets of
+which `continuation' and `higher order functions' are only one.
+
+LL, GoI
+
+Int Construction
+
+QC
+
+Parallel computing with constraints and backtracking -- breaking
+dependancies,
+
+We have contributed an elegant and natural notion of duality that
+arises from the monoidal structure of sums and pairs. This notion of
+duality has a crisp semantics, clear computational interpretations and
+an information theoretic basis.
+
+--- END --
+
 %% \begin{center}
@@
 
-To summarize negative, fractional, square root, and imaginary types all make
-sense. What they help you accomplish as a programmer is to disassociate
-global invariants into local ones that can be satisfied independently by
-subcomputations with no synchronization or communication. A computation
-producing something of type $a/b$ does not need to concern itself with who is
-going to supply the missing $b$: it just does its part. Conversely faced with
-a complicated task, a computation might decide to break it into pieces and
-demand these pieces using negative types.
-
-It is no surprise that these types are closely related to quantum mechanics
-and that they give us the feel that this is how nature computes. This is
-speculation however.
-
-In any case, in a framework where information can be copied and deleted, none
-of this makes much sense. It is critical that these constraints and demands
-can neither be duplicated nor erased.  This gives us the maximum
-``parallelism'' possible.
-
-In our previous work we introduce recursive types and trace operators. This
-is dangerous here because infinite loops allow us to prolong paying the debt
-for as long as we want. The algebraic
-$\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that extreme
-care is needed when combining negative and fractional types with the full
-$\lambda$-calculus (including recursion).
-
-Say we have not considered recursion in this paper.
-
-The simplest way to connect the In a conventional computational model, one
-might realize this situation by simply writing the identity function: the
-buyer hands the money to the seller to finish the transaction. The above
-sequence of isomorphisms implements this identity function in a much more
-interesting way, however.  of the above series of
-
-On the producer side, the debt is paid for by the money computation with the
-identity function. the producer and consumer must somehow share an explicit
-dependency that allows the value. However in our model, the presence of
-negative types allows the produced value to satisfy the demand without the
-producer or consumer even knowing about each other. As is explained in detail
-in
-
-Furthermore, we illustrate their true appeal and expressiveness is brought
-forth by viewing them in the context of an information-preserving
-computational model.
-
-In addition, we show how these types enrich our computational model, they
-obey the same laws as the rational numbers. In other words, every algebraic
-identity that holds for the rational numbers corresponds to a type
-isomorphism in our model.
-
-Specifically, isomorphisms enrich our computational model with have an
-interesting computational interpretation
+%% To summarize negative, fractional, square root, and imaginary types all make
+%% sense. What they help you accomplish as a programmer is to disassociate
+%% global invariants into local ones that can be satisfied independently by
+%% subcomputations with no synchronization or communication. A computation
+%% producing something of type $a/b$ does not need to concern itself with who is
+%% going to supply the missing $b$: it just does its part. Conversely faced with
+%% a complicated task, a computation might decide to break it into pieces and
+%% demand these pieces using negative types.
+
+%% It is no surprise that these types are closely related to quantum mechanics
+%% and that they give us the feel that this is how nature computes. This is
+%% speculation however.
+
+%% In any case, in a framework where information can be copied and deleted, none
+%% of this makes much sense. It is critical that these constraints and demands
+%% can neither be duplicated nor erased.  This gives us the maximum
+%% ``parallelism'' possible.
+
+%% Say we have not considered recursion in this paper.
+
+%% The simplest way to connect the In a conventional computational model, one
+%% might realize this situation by simply writing the identity function: the
+%% buyer hands the money to the seller to finish the transaction. The above
+%% sequence of isomorphisms implements this identity function in a much more
+%% interesting way, however.  of the above series of
+
+%% On the producer side, the debt is paid for by the money computation with the
+%% identity function. the producer and consumer must somehow share an explicit
+%% dependency that allows the value. However in our model, the presence of
+%% negative types allows the produced value to satisfy the demand without the
+%% producer or consumer even knowing about each other. As is explained in detail
+%% in
+
+%% Furthermore, we illustrate their true appeal and expressiveness is brought
+%% forth by viewing them in the context of an information-preserving
+%% computational model.
+
+%% In addition, we show how these types enrich our computational model, they
+%% obey the same laws as the rational numbers.
+
+%% Specifically, isomorphisms enrich our computational model with have an
+%% interesting computational interpretation
 
@@
 
-We furthermore demonstrate that, in our model,
-programming with these negative and fractional types, is a new ``revolution''
-breaking dependencies...
+%% We furthermore demonstrate that, in our model,
+%% programming with these negative and fractional types, is a new ``revolution''
+%% breaking dependencies...
 
@@
 
-Computationally, symmetry exhibits itself as a duality between two concepts.
-
-Symmetry is pervasive in both natural and man-made environments.
+%% Computationally, symmetry exhibits itself as a duality between two concepts.
+
+%% Symmetry is pervasive in both natural and man-made environments.
 
@@
 
-We introduced this thesis that computation should be based on isomorphisms
-that preserve information~\cite{infeffects}. Since Filinski, we've had the
-idea that values and continuations are like mirror images. In a conventional
-language, the negative (continuation) side is implicit and we introduce
-information effects on the positive. Trying to recover the duality from this
-distorted positive side has always been messy. Now it looks clean because we
-have kept the positive side pure.
-
-In a technical sense, this paper extends the language of isomorphisms
-{{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
-other systems which have one notion of duality over additive and
-multiplicative components, {{langRevEE}} has two notions of duality
--- an additive duality over the monoid {{(0, +)}} and a multiplicative
-duality over the monoid {{(1, *)}}. Each axis of duality also give us
-a function space and hence {{langRevEE}} has an additive function
-space corresponding to a notion of control or backtracking and a
-multiplicative function space corresponding to a notion of unification
-or constraint satisfaction.
+%% We introduced this thesis that computation should be based on isomorphisms
+%% that preserve information~\cite{infeffects}. Since Filinski, we've had the
+%% idea that values and continuations are like mirror images. In a conventional
+%% language, the negative (continuation) side is implicit and we introduce
+%% information effects on the positive. Trying to recover the duality from this
+%% distorted positive side has always been messy. Now it looks clean because we
+%% have kept the positive side pure.
+
+%% In a technical sense, this paper extends the language of isomorphisms
+%% {{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
+%% other systems which have one notion of duality over additive and
+%% multiplicative components, {{langRevEE}} has two notions of duality
+%% -- an additive duality over the monoid {{(0, +)}} and a multiplicative
+%% duality over the monoid {{(1, *)}}. Each axis of duality also give us
+%% a function space and hence {{langRevEE}} has an additive function
+%% space corresponding to a notion of control or backtracking and a
+%% multiplicative function space corresponding to a notion of unification
+%% or constraint satisfaction.
 
@@
 \end{itemize}
-So it is all about breaking dependencies in some sense to allow for maximum
-autonomy (parallelism) of subcomputations. It is probably the case that to
-make full use of square root types and imaginary types, we have to move to a
-vector space. If that's the case, we should probably leave this stuff out and
-focus on negative and fractional types and only have a short discussion of
-the polynomials restricted to seven trees in one and similar issues.
-
-The conventional idea is to divide the world into a ``real'' one and a
-``virtual'' one. In the ``real'' world, we can define datatypes like
-\verb|t=t^2+1| but we don't have additive inverses so it makes no sense to
-talk of negative types and we can't rearrange the terms in the datatype
-definition. However the observation is that we can map these datatypes to a
-virtual world that has more structure (a ring that provides additive inverses
-or a field that also provides multiplicative inverses) and then perform
-computations in the ring/field. If we perform computations in the ring, then
-some of these will use additive inverses in ways that cannot be mapped back
-to the ``real'' world. Much of current research attempts to characterize
-which computations done in the ring are valid isomorphisms between datatypes
-in the ``real'' world. This is nice but is not what I am after. In fact I am
-not interested in the ring or the semiring at all. I am interested in the
-field and I want this field to be \textbf{the real world.} This is partly
-motivated by the fact that Quantum Mechanics seems to demand an underlying
-field and more generally that the field provides the maximum generality in
-slicing and dicing computations. So assuming I live in a field and that the
-negative, fractional, square root, and imaginary types are all ``real,'' how
-do I compute in this field? Clearly there will be constraints on
-``measurement'' in the sense that a full program cannot produce any of the
-crazy types but that's done outside the formalism in some sense just as in
-Quantum Computing. The main question I am after is how to compute in this
-field with first-class negative, fractional, etc. types. As I mentioned in my
-previous email, we can produce programs that have types \verb|t^3 <-> -1| and
-they ``run'' (but only to give infinite loops).
-
-So when a programmer writes the datatype declaration \verb|t = t^2+1|,
-if we allow negative etc. then this is effectively writing
-\verb|t = cubicroot{-1}|. If we are in the field then computations
-that manipulate these trees can be sliced and diced even at interfaces
-that expose the cubic root and the imaginary types.
-
-Future work: develop a type system for a ``normal language'' that has
-negative, fractional, etc. types as first-class types. More long term,
-instead of adding one polynomial at a time, we can go to an algebraically
-closed field. The complex numbers is an obvious choice but I would rather go
-to something computable like the field of algebraic numbers. Is the adele
-ring or the p-adics relevant here?
+
+%% So it is all about breaking dependencies in some sense to allow for maximum
+%% autonomy (parallelism) of subcomputations. It is probably the case that to
+%% make full use of square root types and imaginary types, we have to move to a
+%% vector space. If that's the case, we should probably leave this stuff out and
+%% focus on negative and fractional types and only have a short discussion of
+%% the polynomials restricted to seven trees in one and similar issues.
+
+%% The conventional idea is to divide the world into a ``real'' one and a
+%% ``virtual'' one. In the ``real'' world, we can define datatypes like
+%% \verb|t=t^2+1| but we don't have additive inverses so it makes no sense to
+%% talk of negative types and we can't rearrange the terms in the datatype
+%% definition. However the observation is that we can map these datatypes to a
+%% virtual world that has more structure (a ring that provides additive inverses
+%% or a field that also provides multiplicative inverses) and then perform
+%% computations in the ring/field. If we perform computations in the ring, then
+%% some of these will use additive inverses in ways that cannot be mapped back
+%% to the ``real'' world. Much of current research attempts to characterize
+%% which computations done in the ring are valid isomorphisms between datatypes
+%% in the ``real'' world. This is nice but is not what I am after. In fact I am
+%% not interested in the ring or the semiring at all. I am interested in the
+%% field and I want this field to be \textbf{the real world.} This is partly
+%% motivated by the fact that Quantum Mechanics seems to demand an underlying
+%% field and more generally that the field provides the maximum generality in
+%% slicing and dicing computations. So assuming I live in a field and that the
+%% negative, fractional, square root, and imaginary types are all ``real,'' how
+%% do I compute in this field? Clearly there will be constraints on
+%% ``measurement'' in the sense that a full program cannot produce any of the
+%% crazy types but that's done outside the formalism in some sense just as in
+%% Quantum Computing. The main question I am after is how to compute in this
+%% field with first-class negative, fractional, etc. types. As I mentioned in my
+%% previous email, we can produce programs that have types \verb|t^3 <-> -1| and
+%% they ``run'' (but only to give infinite loops).
+
+%% So when a programmer writes the datatype declaration \verb|t = t^2+1|,
+%% if we allow negative etc. then this is effectively writing
+%% \verb|t = cubicroot{-1}|. If we are in the field then computations
+%% that manipulate these trees can be sliced and diced even at interfaces
+%% that expose the cubic root and the imaginary types.
+
+%% Future work: develop a type system for a ``normal language'' that has
+%% negative, fractional, etc. types as first-class types. More long term,
+%% instead of adding one polynomial at a time, we can go to an algebraically
+%% closed field. The complex numbers is an obvious choice but I would rather go
+%% to something computable like the field of algebraic numbers. Is the adele
+%% ring or the p-adics relevant here?
 
@@
 
-We introduced this thesis that computation should be based on isomorphisms
-that preserve information~\cite{infeffects}. Since Filinski, we've had the
-idea that values and continuations are like mirror images. In a conventional
-language, the negative (continuation) side is implicit and we introduce
-information effects on the positive. Trying to recover the duality from this
-distorted positive side has always been messy. Now it looks clean because we
-have kept the positive side pure.
-
-The way to think about something of type $A$ is that it is a value we have
-produced. The way to think about something of type $-A$ is that is a value we
-have already consumed.
+%% We introduced this thesis that computation should be based on isomorphisms
+%% that preserve information~\cite{infeffects}. Since Filinski, we've had the
+%% idea that values and continuations are like mirror images. In a conventional
+%% language, the negative (continuation) side is implicit and we introduce
+%% information effects on the positive. Trying to recover the duality from this
+%% distorted positive side has always been messy. Now it looks clean because we
+%% have kept the positive side pure.
+
+%% The way to think about something of type $A$ is that it is a value we have
+%% produced. The way to think about something of type $-A$ is that is a value we
+%% have already consumed.
 

### 2012-03-10 rational.tex
conclusion
_
intro

@@
 The example illustrates that the empty type and negative types have a
-fascinating computational interpretation as values that backtrack to
-satisfy dependancies and have some flavor of continuations.
-
-\paragraph*{Fractional Types.} What is even more fascinating is that
-the above computation can be realized with a completely different
-concurrent schedule as follows, with fractional types.
+fascinating computational interpretation that has some flavor of
+continuations: negative types denote values that `backtrack to satisfy
+dependencies', or in other words act as `debts' that are satisfied by
+the backward flow of information.
+
+\paragraph*{Fractional Types.} 
+Consider a similar algebraic manipulation involving fractional types.
 
@@
 
-%% The execution begins at line~(0) with the value {{20}}, at line (2)
-%% with the value {{right (right 20)}} flowing forwards and the value
-%% {{right (left 20)}} flowing backwards, and at line (3) with the value
-%% {{left (left 20)}} flowing forwards and the value {{left (right 20)}}
-%% flowing backwards.
-
-%\emph{Simultaneously} 
-The execution begins at line~(0) with the value {{20}}. At line~(1)
-values {{20}} and {{()}} flow forward, corresponding to the type
-`{{a*1}}'. At line~(2) three independent subcomputations \emph{proceed
-  concurrently}: the intial value {{20}} flowing from line (1) and two
-entangled values, let us call them {{1/x}} and {{x}}, such that
-specifying one implicitly specifies the dependant other. These three
-subcomputations \emph{exist at the same time}. At line (3), the
-dependant {{1/x}} is fixed to be the value {{20}}, thereby making
-fixing values of all three branches.
-%% the positive value created at line (2) which can flow to the
-%% consumer; the negative value created at line (3) which can pay for the
-%% debt created at line (2); and the initial value from line (0) which
-%% can flow to the required forward-flowing value needed at line (3). 
-At some intuitive level, one might understand the situation as
-modeling a credit card payment in which money is created at line (2)
-for the consumer together with a corresponding debt. The producer's
-bank pays for the debt at line (3) and waits for the actual money to
-arrive from the producer.  There are at least two fundamental points
-about this example that must be emphasized:
+In the case of negatives, the dotted line indicated the `flow of
+control' whereas for fractionals it indicates the `flow of
+constraints'. At the heart of logic programming is the idea of
+variables that capture constraints and hence it useful to trace this
+computation with the analogy to logic variables in mind:
+
+As before, the execution begins at line~(0) with the value {{20}}. At
+line~(1) two values, {{20}} and {{()}}, flow forward. One can think of
+the value `{{()}}' (of type {{1}}) as `having a credit card': the
+`credit card' isn`t money, nor is it debt, but is the option to
+generate a `credit-debt' constraint.  At line~(2) we exercise this
+option and hence have three values: the initial value {{20}} flowing
+from line (1) and two entangled values, {{1/alpha}} and {{alpha}}. The
+{{alpha}} and {{1/alpha}} are unspecified values -- i.e. we don't yet
+know how much money we need to borrow, but we do know that what is
+borrowed must be what is returned. Hence {{alpha}} denotes the
+presence of an unknown quantity and dually {{1/alpha}} should be
+thought of as the absence of an unknown quantity.
+
+Unlike with negative types, wherein only one value existed at a time
+and the computation backtracked, what is fascinating here is that each
+of the three values \emph{exist at the same time}. In other words, the
+computation with fractions is realized with a completely different
+concurrent schedule in which every value independently and
+concurrently proceeds through its subcomputation.
+
+At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}},
+thereby fixing values of all three branches i.e. we have paid our debt
+of the missing information and the third subcomputation has the known
+value {{20}}. After payment of our debt, as line (4) shows, we have
+use of our `credit card' again.
+
+%% At some intuitive level, one might understand the situation as
+%% modeling a credit card payment in which money is created at line (2)
+%% for the consumer together with a corresponding debt. The producer's
+%% bank pays for the debt at line (3) and waits for the actual money to
+%% arrive from the producer.
+
+The example illustrates that fractional types also have a fascinating
+computational interpretation that have some flavor of continuations:
+the fractional types denote values ({{1/alpha}}) that represent
+`missing information which must be supplied' in much the same sense
+that continuations denote evaluation contexts with `holes that must be
+filled'.  There are at least three fundamental points about the credit
+card examples that must be emphasized:
+
 \begin{itemize}
-\item As the example illustrates, negative types corresponds to
-  ``debts'' satified by backtracking and fractional types correspond
-  to debts satisfied by constraint propagation.  It would clearly be
-  disastrous if debts could be deleted or duplicated. This simple
-  observation explains why these types are much simpler and much more
-  appealing in a framework where information is guaranteed to be
-  preserved. In previous work that used negative types (see
-  Sec.~\ref{sec:related}), complicated mechanisms are typically needed
-  to constrain the propagation and use of negative values because the
-  surrounding computational framework is, generally speaking, careless
-  in its treatment of information.
-\item The appeal of negative and fractional types can already be
-  gleaned from the credit card analogy. The main reason credit card
-  transactions are convenient is because they disentangle the
-  propagation of the resources (money) from the propagation of the
-  services. Not every transaction needs both the resources and
-  services to be brought together: it is sufficient to have a promise
-  that the demand for resources will be somehow satisfied, as long as
-  the infrastructure can be trusted with such
-  promises. 
+\item 
+As the example illustrates, both negative types and fractional types
+corresponds to ``debts'' but in different ways: negatives are satisfied
+by backtracking and fractionals are satisfied by constraint
+propagation.
+
+\item 
+It would clearly be disastrous if debts could be deleted or
+duplicated. This simple observation explains why these types are much
+simpler and much more appealing in a framework where information is
+guaranteed to be preserved. In previous work that used negative types
+(see Sec.~\ref{sec:related}), complicated mechanisms are typically
+needed to constrain the propagation and use of negative values because
+the surrounding computational framework is, generally speaking,
+careless in its treatment of information.
+
+\item 
+The main reason credit card transactions are convenient is because
+they disentangle the propagation of the resources (money) from the
+propagation of the services. Not every transaction needs both the
+resources and services to be brought together: it is sufficient to
+have a promise that the demand for resources will be somehow
+satisfied, as long as the infrastructure can be trusted with such
+promises. This idea that dependencies can be freely decoupled and
+propagated can be a powerful programming tool and we leverage this in
+the construction of our novel SAT-solver (Sec. \ref{sec:sat-solver}).
+
 %% Computationally, this means that negative types allow us
@@
 
-\paragraph*{Functions and Dependancy Splitting.} 
+\paragraph*{Functions and Dependency Splitting.} 
 %% Consider the following realization of boolean negation written using a
@@
 witness determined by further constraints encountered during the
-computation.  The idea of variables that capture constraints that are
-satisfied later is at the heart of logic programming and hence the
+computation.
+
+ and hence the
 witness for the isomorphisms above is generally {{(1/alpha,alpha)}}
@@
 additive functions of the form {{0<-> (-a)+b}}, correspond to
-control flow and corotuining between two subcomputations -- only one
+control flow and coroutining between two subcomputations -- only one
 of which `exists' at any point.
@@
 
+\begin{comment}
+
 --- END --
@@
 
-%% Jacques Carette, Michael Adams
+\end{comment}
+
+%% Jacques Carette, Michael Adams, Khaled
 

### 2012-03-10 rational.tex
_

@@
 computational interpretation. The witness for this isomorphism is a
-computation that takes a value of type~{{a}} (say \$20.00) and eventually
+computation that takes a value of type~{{a}}, say \$20.00, and eventually
 produces the same \$20.00 value as its output. As the semantics of
@@
   in the left summand of the sum type `{{a+0}}'; we indicate this value
-  as {{left 20}}
+  as {{left 20}};
 \item We continue to line (2) with the same value {{left 20}};
@@
 Consider a similar algebraic manipulation involving fractional types.
-
 \begin{center}
@@
 \end{center}
-
 In the case of negatives, the dotted line indicated the `flow of
@@
 constraints'. At the heart of logic programming is the idea of
-variables that capture constraints and hence it useful to trace this
-computation with the analogy to logic variables in mind:
+variables that capture constraints. Hence it is useful to trace the
+computation corresponding to the algebraic proof above with the
+analogy to logic variables in mind.
 
@@
 line~(1) two values, {{20}} and {{()}}, flow forward. One can think of
-the value `{{()}}' (of type {{1}}) as `having a credit card': the
-`credit card' isn`t money, nor is it debt, but is the option to
+the value `{{()}}' (of type {{1}}) as `having a credit card'. The
+`credit card' isn't money, nor is it debt, but is the option to
 generate a `credit-debt' constraint.  At line~(2) we exercise this

### 2012-03-10 rational.tex
intro part 2

@@
 variables that capture constraints. Hence it is useful to trace the
-computation corresponding to the algebraic proof above with the
+computation corresponding to the algebraic proof above, with the
 analogy to logic variables in mind.
@@
 
-In the algebra of rational numbers we have that for each {{a}} that is
-not equal to {{0}}, {{1 = (1/a) * a}}. In our framework this algebraic
-identity corresponds to a type isomorphism {{1 <-> (1/a) * a}}, whose
-witness maybe thought of as the `identity function' at {{a}}.  For
-example, for the datatype \verb#color = R|G|B#, there are three
-possible witnesses for this isomorphism mapping {{()}} to either
-{{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular
-witness determined by further constraints encountered during the
-computation.
-
- and hence the
-witness for the isomorphisms above is generally {{(1/alpha,alpha)}}
-where~{{alpha}} is a fresh logic variable.
+In the algebra of rational numbers {{1 = (1/a) * a}} if {{a}} is
+nonzero.  In our framework this algebraic identity corresponds to a
+type isomorphism {{1 <-> (1/a) * a}} (written alternately as
+{{a-o*a}}) whose witness maybe thought of as the `identity function'
+at the type {{a}}.  For example, for the datatype {{color = R|Gr|B}},
+there are three possible witnesses for this isomorphism mapping {{()}}
+to either {{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the
+particular witness determined by further constraints encountered
+during the computation. More generally, the witness for the
+isomorphism above may be represented as {{(1/alpha,alpha)}}
+where~{{alpha}} is a fresh logic variable of type {{a}}. With this
+view of {{a -o* a}} functions, let us consider the following
+manipulations:
 
@@
 
-At this point we have constructed a strange mixing of two functions;
-`inputs' of one form `outputs' of the other. If {{(1/alpha1,alpha2)}}
-is held by one subcomputation and {{(1/alpha2,alpha1)}} is held by
-another subcomputation, these `remixed functions' form a communication
-channel between the subcomputations -- unifying {{1/alpha1}} with
-\verb|color| {{R}} in one subcomputation, fixes it in the
-other. Depending on the choice of the type {{a}}, an arbitrary number
-of bits may be communicated. The dual reading of the above, with
-additive functions of the form {{0<-> (-a)+b}}, correspond to
-control flow and coroutining between two subcomputations -- only one
-of which `exists' at any point.
+At this point we have constructed a strange mixing of two `{{a-o*a}}'
+functions; `inputs' of one form `outputs' of the other. If
+{{(1/alpha1,alpha2)}} is held by one subcomputation and
+{{(1/alpha2,alpha1)}} is held by another subcomputation, these
+`remixed functions' form a communication channel between the two
+concurrent subcomputations -- unifying {{1/alpha1}} with {{color}}
+`{{R}}' in one subcomputation, fixes {{alpha1}} in the other. The type
+{{a}} thus takes the role of the type of the communication channel,
+indicating how much information can be communicated between the two
+subcomputations.  Depending on the choice of the type {{a}}, an
+arbitrary number of bits may be communicated.
+
+Dually, the additive reading of the above manipulations correspond to
+functions of the form {{a-o+a}}, witnessing isomorphisms of the form
+{{0<->(-a)+a}}). The `remixed' additive functions express control flow
+transfer between two subcomputations, \emph{only one of which exists}
+at any point -- i.e. they capture the essence of coroutines. The
+behaviour of additive and multiplicative functions are further
+discussed in Sec. \ref{sec:rat}.
 
@@
 is preserved, negative and fractional types provide fascinating mechanisms in
-which computations can be sliced and diced, decomposed and recomposed, ran
+which computations can be sliced and diced, decomposed and recomposed, run
 forwards and backwards, in arbitrary ways. The remainder of the paper

### 2012-03-10 rational.tex
intro part 2

@@
 where~{{alpha}} is a fresh logic variable of type {{a}}. With this
-view of {{a -o* a}} functions, let us consider the following
+view of `{{a -o* a}}' functions, let us consider the following
 manipulations:
@@
   the components of the above tuple. The resulting value is
-  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}}, and
-  {{alpha2}} are fresh logic variables.
+  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
+  {{alpha2}} are fresh logic variables;
 \item Using the fact that {{*}} is associative and commutative, we can
@@
 
-At this point we have constructed a strange mixing of two `{{a-o*a}}'
-functions; `inputs' of one form `outputs' of the other. If
-{{(1/alpha1,alpha2)}} is held by one subcomputation and
+At this point we have constructed a strange mix of two `{{a-o*a}}'
+functions; `inputs' of one function manifests as `outputs' of the
+other. If {{(1/alpha1,alpha2)}} is held by one subcomputation and
 {{(1/alpha2,alpha1)}} is held by another subcomputation, these
 `remixed functions' form a communication channel between the two
-concurrent subcomputations -- unifying {{1/alpha1}} with {{color}}
-`{{R}}' in one subcomputation, fixes {{alpha1}} in the other. The type
-{{a}} thus takes the role of the type of the communication channel,
-indicating how much information can be communicated between the two
-subcomputations.  Depending on the choice of the type {{a}}, an
-arbitrary number of bits may be communicated.
+concurrent subcomputations. Unifying {{1/alpha1}} with {{color}}
+`{{R}}' in one subcomputation, fixes {{alpha1}} to be {{R}} in the
+other. The type {{a}} thus takes the role of the type of the
+communication channel, indicating how much information can be
+communicated between the two subcomputations.  Depending on the choice
+of the type {{a}}, an arbitrary number of bits may be communicated.
 

### 2012-03-10 rational.tex
reorg

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Additive Duality in {{langRev}} }
@@
 right-to-left direction.
-
-\paragraph*{Observability.} 
-Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
-{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
-defined only if {{b1}} and {{b2}} are entirely positive types. This is
-similar to the constraint that Zeilberger imposes to explain intuitionistic
-polarity and delimited control~\cite{10.1109/LICS.2010.23}.
-
-Consider the program that
-
-(TODO: So this is a really bad explanation. But this section is really
-important and we need to explain carefully. Maybe an analogy can be
-drawn to Cayley diagrams for groups to point out that the notion of
-`i' is entirely artificial, but it is essential to the discourse.)
-
-\begin{center}
-  \includegraphics{diagrams/shuffle.pdf}
-\end{center}
-
-We define observables to be only positive types. The outputs of
-programs that output mixed positive and negative types are not
-observable.  Also, programs that input mixed positive and negatives
-types are not executable.
-
-It is not fair to say that negative types flow backwards. The
-following circuits are valid in {{langRevEE}}. It is however proper to
-say that for any type {{b}} that flows in a direction, the type {{-b}}
-flows in the reverse direction.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions in the Additive Monoid}
-\label{sec:monoidal-constructions}
-
-\begin{enumerate}
-\item 
-\emph{Trace}
-
-test.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/trace.pdf}
-\end{center}
-
-%subcode{proof} include main
-%@ c : b2 + b1 <-> b2 + b3
-%@@ trace c : b1 <-> b3
-
-\end{multicols}
-
-test.
-
-\item
-\emph{Involution (Principium Contradictiones)}
-
-{{b <-> -(-b)}}
-
-\begin{center}
-  \includegraphics{diagrams/double_neg.pdf}
-\end{center}
-
-\item
-\emph{Functions}
-
-Mention LL and {{a -o b}}.
-
-\begin{center}
-  \includegraphics{diagrams/function.pdf}
-\end{center}
-
-Let us use the shorthand {{b1 -o+ b2 = 0 <-> -b1 + b2}}. See
-Sec. \ref{sec:hof} for a discussion of the additive functions
-vs. multiplicative functions.
-
-\item
-\emph{Function composition.}
-
-\begin{multicols}{2}
-\scalebox{1.1}{
-    \includegraphics{diagrams/compose1.pdf}
-}
-
-\scalebox{0.8}{
-    \includegraphics{diagrams/compose.pdf}
-}
-\end{multicols}
-This is also equivalent to sequencing both the computation blocks.
-
-\begin{center}
-  \includegraphics{diagrams/compose2.pdf}
-\end{center}
-
-\item
-\emph{Function application.}
-test.
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.0}{
-  \includegraphics{diagrams/apply1.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.8}{
-  \includegraphics{diagrams/apply2.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item
-\emph{Delimited continuation}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}
-
-\item
-\emph{Negation distributes over {{+}}. }
-
-{{-(b1 + b2) <-> (-b1) + (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/dist_neg_plus.pdf}
-\end{center}
-
-\item
-\emph{Lifting a operation of positive types to negated types}
-
-Given {{c : b1 <-> b2}}
-
-\begin{center}
-  \includegraphics{diagrams/neg_lift.pdf}
-\end{center}
-
-\end{enumerate}
 
@@
 ADD MATH.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions with Negatives and Fractions}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions independant of the  Monoid}
+\label{sec:monoidal-constructions}
+
+\begin{enumerate}
+\item 
+\emph{Trace}
+
+test.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
+\end{multicols}
+
+test.
+
+\item
+\emph{Involution (Principium Contradictiones)}
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+\item
+\emph{Functions}
+
+Mention LL and {{a -o b}}.
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o+ b2 = 0 <-> -b1 + b2}}. See
+Sec. \ref{sec:hof} for a discussion of the additive functions
+vs. multiplicative functions.
+
+\item
+\emph{Function composition.}
+
+\begin{multicols}{2}
+\scalebox{1.1}{
+    \includegraphics{diagrams/compose1.pdf}
+}
+
+\scalebox{0.8}{
+    \includegraphics{diagrams/compose.pdf}
+}
+\end{multicols}
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+\item
+\emph{Function application.}
+test.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/apply1.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/apply2.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item
+\emph{Delimited continuation}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+\item
+\emph{Negation distributes over {{+}}. }
+
+{{-(b1 + b2) <-> (-b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+\item
+\emph{Lifting a operation of positive types to negated types}
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Conventional Higher Order Functions}
+\label{sec:hof}
+
+In a conventional language, higher-order functions capture scope and
+can be represented as first class values. On the other hand,
+{{langRevEE}} gives us two notions of functions both of which share
+similarities with conventional HOFs.
+
+\begin{itemize}
+\item 
+Additive functions are not really values, but are possible paths of
+execution. However, they can be encoded as values as follows: Consider
+two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
+{{bool}} can discriminate them and hence a value of type {{bool}} acts
+as the `address' of the function and is a first class value
+representing one of the {{b1 -o+ b2}} functions.  This is similar to
+how first classes functions a'la \lcal are represented by their
+addresses when compiled to conventional hardware.
+
+\item
+Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+are indeed first class values. However these functions can only be
+linearly used -- i.e. they represent a contraint that can be satisfied
+at most once in one `computational world'. It may be possible that to
+repeatedly invoke multiplicative functions with an appropriate
+recursive type whose unfoldings allow for repeated unification.
+\end{itemize}
+
+The exact encoding on HOF in {{langRevEE}} should shed light on
+several problems, including the connections to continuations discussed
+in Sec. \ref{sec:related}. They should also allow for a natural
+encoding of HOFs in the translations of \lcal to {{langRevT}} and
+indicate a monadic encapsulation of of information effects which are
+currrently structured as an arrow.
 
@@
 
-%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics of {{langRevEE}} }
@@
 %% \label{sec:monoidal-constructions}
+
+\paragraph*{Observability.} 
+Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
+{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
+defined only if {{b1}} and {{b2}} are entirely positive types. This is
+similar to the constraint that Zeilberger imposes to explain intuitionistic
+polarity and delimited control~\cite{10.1109/LICS.2010.23}.
+
+Consider the program that
+
+(TODO: So this is a really bad explanation. But this section is really
+important and we need to explain carefully. Maybe an analogy can be
+drawn to Cayley diagrams for groups to point out that the notion of
+`i' is entirely artificial, but it is essential to the discourse.)
+
+\begin{center}
+  \includegraphics{diagrams/shuffle.pdf}
+\end{center}
+
+We define observables to be only positive types. The outputs of
+programs that output mixed positive and negative types are not
+observable.  Also, programs that input mixed positive and negatives
+types are not executable.
+
+It is not fair to say that negative types flow backwards. The
+following circuits are valid in {{langRevEE}}. It is however proper to
+say that for any type {{b}} that flows in a direction, the type {{-b}}
+flows in the reverse direction.
+
+Several things we don't know how to do, or are sloppy about. Some of
+these are deep questions to which there is no immeidate answer.
+
+\begin{itemize}
+
+\item Is the language without recursive types always terminating? For
+  the right symmetry with annihilation, it is desirable to have
+  non-termination.
+
+\item What is the correct lemma for logical reversibility now?
+
+\item We should say things about information preservation. At least
+  more so than what is being said now.
+
+\item Multiplicative duality gives us a means of talking about
+  equivalence classes of values. This should correspond to all sorts
+  of other things -- groupoids, Lawvere theories etc. We barely even
+  understand these connections.
+
+\item How does this connect to QC? Is there any strong connection such
+  that the SAT solver here is an alternate QC algorithm to Shor's?
+
+\end{itemize}
 
@@
 %%   which allows arbitrary duplication. Can we really do that???}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-
--- WAVEFRONT START --
-
-Several things we don't know how to do, or are sloppy about. Some of
-these are deep questions to which there is no immeidate answer.
-
-\begin{itemize}
-
-\item Is the language without recursive types always terminating? For
-  the right symmetry with annihilation, it is desirable to have
-  non-termination.
-
-\item What is the correct lemma for logical reversibility now?
-
-\item We should say things about information preservation. At least
-  more so than what is being said now.
-
-\item Multiplicative duality gives us a means of talking about
-  equivalence classes of values. This should correspond to all sorts
-  of other things -- groupoids, Lawvere theories etc. We barely even
-  understand these connections.
-
-\item How does this connect to QC? Is there any strong connection such
-  that the SAT solver here is an alternate QC algorithm to Shor's?
-
-\end{itemize}
-
--- WAVEFRONT STOP --
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Conventional Higher Order Functions}
-\label{sec:hof}
-
-In a conventional language, higher-order functions capture scope and
-can be represented as first class values. On the other hand,
-{{langRevEE}} gives us two notions of functions both of which share
-similarities with conventional HOFs.
-
-\begin{itemize}
-\item 
-Additive functions are not really values, but are possible paths of
-execution. However, they can be encoded as values as follows: Consider
-two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
-{{bool}} can discriminate them and hence a value of type {{bool}} acts
-as the `address' of the function and is a first class value
-representing one of the {{b1 -o+ b2}} functions.  This is similar to
-how first classes functions a'la \lcal are represented by their
-addresses when compiled to conventional hardware.
-
-\item
-Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-are indeed first class values. However these functions can only be
-linearly used -- i.e. they represent a contraint that can be satisfied
-at most once in one `computational world'. It may be possible that to
-repeatedly invoke multiplicative functions with an appropriate
-recursive type whose unfoldings allow for repeated unification.
-\end{itemize}
-
-The exact encoding on HOF in {{langRevEE}} should shed light on
-several problems, including the connections to continuations discussed
-in Sec. \ref{sec:related}. They should also allow for a natural
-encoding of HOFs in the translations of \lcal to {{langRevT}} and
-indicate a monadic encapsulation of of information effects which are
-currrently structured as an arrow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computing in the Field of Algebraic Numbers}
-
-The field of algebraic numbers consist of the rationals extended with
-solutions to polynomials. In a limited sense {{langRevEE}} extended
-with recursive types already include polynomials such as the
-definition of binary tree {{@@x.(1+x*x)}} implies the equality
-{{x=1+x*x}} and hence {{x^2-x+1=0}}.
-
-%% Partial isomorphisms algebraics
-%% do have computational
-
-%% While this equation does indeed have an irrational imaginary root,
-%% as we will see -- {{langRevEE}} does not have the means of
-%% expressing meaningful computation that uses the root.
-
-The famous example of isomorphism on algebraics is the embedding of
-seven binary trees in one. In Figure 1 of Fiore's paper
-\cite{Fiore:2004}, he demonstrates this isomorphism by the
-corrsponding algebraic manipulation of the types.  The algebraic
-manipulation can be viewed as a {{langRevEE}} combinator and it indeed
-does constructively witness the isomorphism -- i.e. you can run it and
-view one data structure encoded by the other. In this context,
-consider the following algebraically valid proof of the same
-isomorphism:
-
-%% Previously the most famous example of a similar nature is the puzzling
-%% isomorphism that one can establish between seven binary trees and one.
-%% A binary tree is defined by the datatype
-%% \[
-%% x = 1 + x * x 
-%% \]
-%% which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
-%% algebraic manipulation we can reason as follows:
-\[\begin{array}{rclclclcl}
-x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
-x^6 &=& 1 \\
-x^7 &=& x^6 x &=& x
-\end{array}\]
-Fiore poses the question of why such algebraic manipulation would make
-sense type theoretically, even though some of the intermediate steps
-make no sense.  Stated in terms of {{langRevEE}} computations, this
-manipulation has no meaningful computational content -- it is always
-non-terminating. It suggests that some algebraic manipulations are
-somehow more `constructive' than others or dually that {{langRevEE}}
-with recursive types somehow lacks the ability to express all
-computations that are otherwise algebraically meaningful.
-
--- WAVEFRONT START --
-
-But the interesting solution of trees is precisely that they involve
-irrational numbers. Our usual notation for irrationals such as
-{{sqrt(2) }} can actually be viewed as a thunk representing the
-infinite computation {{1.4142..}}.  Irrationals in this sense have a
-very different information theoretic interpretation from full real
-numbers.  With Chaitin's $\Omega$ numbers
-\cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
-real numbers that have an infinite amount of information -- ones for
-example are the solution to the question about the halting properties
-of every Turning machine.
-
-That introduces the natural numbers. No solution for this polynomial
-over the algebraic numbers. We must extend the numbers with $\omega$
-to get a solution. We reject this in this paper and prefer to stick to
-algebraic numbers. The advantage is that all the isomorphisms are
-valid numerically. With the above type we could subtract $b$ from both
-sides to show that $0=1$ which is nonsense. We can however have
-infinite types as long as they have algebraic solutions.
-
-In our previous work we introduce recursive types and trace
-operators. This is dangerous here because infinite loops allow us to
-prolong paying the debt for as long as we want. The algebraic
-$\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that
-extreme care is needed when combining negative and fractional types
-with the full $\lambda$-calculus (including recursion).
-
--- WAVEFRONT STOP --
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+\section{Future work and Conclusion}
 \label{sec:conc}
@@
 dependancies,
+
+\paragraph*{Computing in the Field of Algebraic Numbers.}
+\label{sec:algebraic-field}
+The field of algebraic numbers consist of the rationals extended with
+solutions to polynomials. In a limited sense {{langRevEE}} extended
+with recursive types already include polynomials such as the
+definition of binary tree {{@@x.(1+x*x)}} implies the equality
+{{x=1+x*x}} and hence {{x^2-x+1=0}}.
+
+%% Partial isomorphisms algebraics
+%% do have computational
+
+%% While this equation does indeed have an irrational imaginary root,
+%% as we will see -- {{langRevEE}} does not have the means of
+%% expressing meaningful computation that uses the root.
+
+The famous example of isomorphism on algebraics is the embedding of
+seven binary trees in one. In Figure 1 of Fiore's paper
+\cite{Fiore:2004}, he demonstrates this isomorphism by the
+corrsponding algebraic manipulation of the types.  The algebraic
+manipulation can be viewed as a {{langRevEE}} combinator and it indeed
+does constructively witness the isomorphism -- i.e. you can run it and
+view one data structure encoded by the other. In this context,
+consider the following algebraically valid proof of the same
+isomorphism:
+
+%% Previously the most famous example of a similar nature is the puzzling
+%% isomorphism that one can establish between seven binary trees and one.
+%% A binary tree is defined by the datatype
+%% \[
+%% x = 1 + x * x 
+%% \]
+%% which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
+%% algebraic manipulation we can reason as follows:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+Fiore poses the question of why such algebraic manipulation would make
+sense type theoretically, even though some of the intermediate steps
+make no sense.  Stated in terms of {{langRevEE}} computations, this
+manipulation has no meaningful computational content -- it is always
+non-terminating. It suggests that some algebraic manipulations are
+somehow more `constructive' than others or dually that {{langRevEE}}
+with recursive types somehow lacks the ability to express all
+computations that are otherwise algebraically meaningful.
+
+-- WAVEFRONT START --
+
+But the interesting solution of trees is precisely that they involve
+irrational numbers. Our usual notation for irrationals such as
+{{sqrt(2) }} can actually be viewed as a thunk representing the
+infinite computation {{1.4142..}}.  Irrationals in this sense have a
+very different information theoretic interpretation from full real
+numbers.  With Chaitin's $\Omega$ numbers
+\cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
+real numbers that have an infinite amount of information -- ones for
+example are the solution to the question about the halting properties
+of every Turning machine.
+
+That introduces the natural numbers. No solution for this polynomial
+over the algebraic numbers. We must extend the numbers with $\omega$
+to get a solution. We reject this in this paper and prefer to stick to
+algebraic numbers. The advantage is that all the isomorphisms are
+valid numerically. With the above type we could subtract $b$ from both
+sides to show that $0=1$ which is nonsense. We can however have
+infinite types as long as they have algebraic solutions.
+
+In our previous work we introduce recursive types and trace
+operators. This is dangerous here because infinite loops allow us to
+prolong paying the debt for as long as we want. The algebraic
+$\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that
+extreme care is needed when combining negative and fractional types
+with the full $\lambda$-calculus (including recursion).
+
+-- WAVEFRONT STOP --
 

### 2012-03-11 rational.tex
intro.
_

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions with Negatives and Fractions}
+\section{Constructions with Negatives and Fractions}
 
@@
 
-LL, GoI
-
-Int Construction
-
-QC
-
-Parallel computing with constraints and backtracking -- breaking
-dependancies,
+\paragraph*{GoI}
+The other popular model of computing with values traveling along wire
+is GoI -- specifically Mackie's work. How are we related to this
+model? Preliminary investogationb suggests that many of the no-cloning
+based constructions can be simulated in {{langRevEE}} as well. If
+there is a concrete connectoion, then by Curry-Howard the logical
+interpretation of {{langRevEE}} should a logic much like LL with a
+strong notion of resource preservation, but also with a strong
+computational basis.
+
+\paragraph*{Int Construction}
+How are we preserving two compact closed structures -- what does this
+imply about the Int construction.
+
+\paragraph*{QC} 
+Are constructions are very similar to those developed in the context
+of QC. Is this how nature computes? Are the analogies to time
+traveling particles and anti-particles more concrete in some way?
+
+%% Parallel computing with constraints and backtracking -- breaking
+%% dependancies,
 
@@
 
--- WAVEFRONT START --
-
-But the interesting solution of trees is precisely that they involve
-irrational numbers. Our usual notation for irrationals such as
-{{sqrt(2) }} can actually be viewed as a thunk representing the
-infinite computation {{1.4142..}}.  Irrationals in this sense have a
-very different information theoretic interpretation from full real
-numbers.  With Chaitin's $\Omega$ numbers
-\cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
-real numbers that have an infinite amount of information -- ones for
-example are the solution to the question about the halting properties
-of every Turning machine.
-
-That introduces the natural numbers. No solution for this polynomial
-over the algebraic numbers. We must extend the numbers with $\omega$
-to get a solution. We reject this in this paper and prefer to stick to
-algebraic numbers. The advantage is that all the isomorphisms are
-valid numerically. With the above type we could subtract $b$ from both
-sides to show that $0=1$ which is nonsense. We can however have
-infinite types as long as they have algebraic solutions.
-
-In our previous work we introduce recursive types and trace
-operators. This is dangerous here because infinite loops allow us to
-prolong paying the debt for as long as we want. The algebraic
-$\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that
-extreme care is needed when combining negative and fractional types
-with the full $\lambda$-calculus (including recursion).
-
--- WAVEFRONT STOP --
-
+\begin{itemize}
+\item Maybe the problem is that there is cancellation in the second
+  case and cancellation with recursive types is problematic. However
+  not all cancellation is. So what gives?
+
+\item One idea is that irrationals correspond to an infinite amount of
+  computational work. If what we are cancelling is not the same
+  infinity, then things go wrong.
+
+\item Chaitin has already shown that reals have inifite amount of
+  information which is paradoxical and problematic. This is a
+  different issue from that of the rational being infinite
+  computations.
+
+\item Finally, not all meaningful types are meaningful polynomials --
+  ex. nat. So this is not the whole story.
+
+\end{itemize}
+
+%% -- WAVEFRONT START --
+
+%% But the interesting solution of trees is precisely that they involve
+%% irrational numbers. Our usual notation for irrationals such as
+%% {{sqrt(2) }} can actually be viewed as a thunk representing the
+%% infinite computation {{1.4142..}}.  Irrationals in this sense have a
+%% very different information theoretic interpretation from full real
+%% numbers.  With Chaitin's $\Omega$ numbers
+%% \cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
+%% real numbers that have an infinite amount of information -- ones for
+%% example are the solution to the question about the halting properties
+%% of every Turning machine.
+
+%% That introduces the natural numbers. No solution for this polynomial
+%% over the algebraic numbers. We must extend the numbers with $\omega$
+%% to get a solution. We reject this in this paper and prefer to stick to
+%% algebraic numbers. The advantage is that all the isomorphisms are
+%% valid numerically. With the above type we could subtract $b$ from both
+%% sides to show that $0=1$ which is nonsense. We can however have
+%% infinite types as long as they have algebraic solutions.
+
+%% In our previous work we introduce recursive types and trace
+%% operators. This is dangerous here because infinite loops allow us to
+%% prolong paying the debt for as long as we want. The algebraic
+%% $\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that
+%% extreme care is needed when combining negative and fractional types
+%% with the full $\lambda$-calculus (including recursion).
+
+%% -- WAVEFRONT STOP --
+
+\paragraph*{Summary}
 We have contributed an elegant and natural notion of duality that

### 2012-03-11 rational.tex
reorg
sec negative types

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Negative Types}
-\label{sec:neg}
-
-The structure of this section and the next are the same. The general outline
-is as follows:\footnote{We refer the reader to Selinger's excellent survey
+\section{Negative and Fractional Types}
+
+% The structure of this section and the next are the same.
+
+This section introduces the syntax and graphical languages for
+negative types and fractionals.The general outline is as follows:
+\footnote{We refer the reader to Selinger's excellent survey
   article of monoidal
@@
 After presenting the extension at the syntactic level, we discuss the
-categorical semantics informally via the graphical language. The operational
-semantics is presented in Sec.~\ref{sec:rat}. As will be detailed in this
-section and the next, the compact closed structure provides several
-properties of interest: (i) morphisms or wires are allowed to run from right
-to left, (ii) the structure admits a trace that, depending on the situation,
-can be used to implement various notions of loops and
+categorical semantics informally via the graphical language. The
+operational semantics is presented in Sec.~\ref{sec:rat}. As will be
+detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
+compact closed structure provides several properties of interest: (i)
+morphisms or wires are allowed to run from right to left, (ii) the
+structure admits a trace that, depending on the situation, can be used
+to implement various notions of loops and
 recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
-(iii) the structure includes an isomorphism showing that the dual operator is
-an involution, and (iv) the structure is equipped with objects representing
-higher-order functions and a morphism \textit{eval} that applies these
-functional objects. Interestingly each monoidal structure provides the same
-ingredients, resulting in two dualities, two traces, two involutions, and two
-notions of higher-order functions.
+(iii) the structure includes an isomorphism showing that the dual
+operator is an involution, and (iv) the structure is equipped with
+objects representing higher-order functions and a morphism
+\textit{eval} that applies these functional objects. Interestingly
+each monoidal structure provides the same ingredients, resulting in
+two dualities, two traces, two involutions, and two notions of
+higher-order functions.
 
@@
 \subsection{Additive Duality in {{langRev}} }
+\label{sec:neg}
 
@@
 
-\paragraph*{Negative Information Flow.} 
 Hence, the intuitive way to read the diagrams is that they reverse the
-flow of information -- in normal {{langRev}} circuits information
-flows from left-to-right and {{eps+}} operations cause information to
-flow from right-to-left.  The interpretation of {{eta+}} is precisely
-the opposite: it turns a right-to-left flow of information in a
-circuit into a left-to-right flow.
-As an example, the diagram representing the isomorphism from the
-introduction is:
-\begin{center}
+flow of information -- i.e. they capture a notion of \emph{negative
+  information flow}. In normal {{langRev}} circuits, information flows
+from left-to-right. The operator {{eps+}} causes information to flow
+from right-to-left.  The interpretation of {{eta+}} is the opposite:
+it turns a right-to-left flow of information in a circuit into a
+left-to-right flow.  The diagram below is an example of using {{eta+}}
+and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
+from the introduction. (Here the {{0}} wires have been shown for
+completeness.)
+\begin{center}
+\scalebox{1.2}{
   \includegraphics{diagrams/algebra1.pdf}
-\end{center}
-
-\paragraph*{Time Traveling.}
-A time traveling intuition is also applicable here: if view the
-left-to-right direction of information flow as the canonical forward
-direction of information, then the action of {{eps+}} allows values,
-aka information particles, to flow backwards in time. Particles that
-flow backwards in time get to see and interact with the history of
-particles that they coexist with i.e. values that they exist
-multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
-(whose construction is shown in Section \ref{sec:neg-constructions}).
+}
+\end{center}
+
+% \paragraph*{Time Traveling.}
+A \emph{time traveling} intuition is also applicable here.  In a
+normal cicruit, as computational steps are taken values flow from left
+to right, i.e. computational time progresses from left to right.  The
+action of {{eps+}} causes values, aka information particles, to flow
+from right to left i.e. backwards in computational time.  The
+interesting thing about this interpretation is that particles that
+travel backwards in time get to see and interact with the history of
+particles that they coexist with. This gives us an intuitive
+interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
+Sec. \ref{sec:specific-constructions}) where it would otherwise seem
+that {{-b1}} and {{b2}} move in opposite directions. The backward flow
+of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
+equivalent to both particles moving backward in time.
 
 %\subsection{Compact Closure with Additive Inverses}
-\paragraph*{Coherence for Compact Closure with Additive Inverses.}
-The appropriate coherence condition for Compact Closed Categories,
+\emph{Coherence for Compact Closure with Additive Duality.}  The
+appropriate coherence condition for Compact Closed Categories,
 expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-may be visualized as explained by Selinger
-\cite{springerlink:10.1007/978-3-642-12821-94} as follows:
+may be visualized (modulo {{0}} wires, swapping etc.) as explained by
+Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
+operational equivalence:
 
@@
 
-\paragraph*{Non-termination.}
-As is well known in category theory, Compact Closed Categories contain
-trace operations.  In the context of {{langRev}} this implies that
-{{eta+}}/{{eps+}} allow the construction of a {{trace+}} operator
-(detailed in Section \ref{sec:monoidal-constructions}) hence
-introducing \emph{non-termination} into the language (in the presence
-of recursive types) \cite{infeffects}.
-
-\paragraph*{Zero Information.}
-The fact that {{eta+}} (and hence {{eps+}}) introduces no information
-can be understood is by viewing {{-b+b}} as a type whose value is
-observable only if you `pay' {{b}} amount of information to the {{-b}}
-branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
-amount of information is returned to you. (TODO. more needs to be said
-here.)
-
-For this reason, our treatment of negative values in the context of
-{{langRev}} is particularly simple. We will have a type $0$ and an
-isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-allows us to create a value and a corresponding debt out of nothing. Both the
-value and its negative counterpart can flow anywhere in the system. Because
-information is preserved, a closed program (which does not have any
-``dangling'' negative values) will eventually have to match the negative
-value with some corresponding value, effectively using the isomorphism in the
-right-to-left direction.
+\emph{Non-termination.}  Compact closed categories contain {{trace}}
+operations and this implies that {{eta+}}/{{eps+}} allow the
+construction of a {{trace+}} operator (shown in Section
+\ref{sec:monoidal-constructions}).  As shown in previous work
+\cite{infeffects}, non-terminating computation can be constructed from
+     {{trace+}} and recursive types.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Fractional Types}
+% \section{Fractional Types}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Multiplicative Duality in {{langRev}} }
 \label{sec:frac}
@@
 the underlying categorical structure.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multiplicative Duality in {{langRev}} }
 
@@
 ADD MATH.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Constructions with Negatives and Fractions}
 
@@
 %% \label{sec:monoidal-constructions}
+
+\emph{Zero Information.}  The intuitive explanation that {{eta+}} and
+hence {{eps+}} introduces no information effects is that {{-b+b}} as a
+type whose value is observable only if you `pay' {{b}} amount of
+information to the {{-b}} branch. When {{b}} amount of information is
+supplied {{(-b+b)+b=b}} amount of information is returned to
+you. (TODO. more needs to be said here.)
+
+For this reason, our treatment of negative values in the context of
+{{langRev}} is particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
+right-to-left direction.
 

### 2012-03-11 rational.tex
sec negative types

@@
 \emph{Coherence for Compact Closure with Additive Duality.}  The
-appropriate coherence condition for Compact Closed Categories,
+appropriate coherence condition for compact closed categories,
 expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and

### 2012-03-11 rational.tex
sec fractional types

@@
 
-The section follows the outline of the previous one, first presenting the new
-syntax, and discussing the graphical language and the properites implied by
-the underlying categorical structure.
+%% The section follows the outline of the previous one, first presenting
+%% the new syntax, and discussing the graphical language and the
+%% properites implied by the underlying categorical structure.
 
 Multiplicative duality is added to {{langRev}} by adding types
-{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}} with
-the following type judgements.
+{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.
 
@@
 
-\vspace{-15pt}
+These have type judgements, similar to those of their additive
+couterparts.
 \begin{multicols}{2}
@@
 As with additive duals, we represent {{eta*}} and {{eps*}} using
-U-shaped connectors. In the general case the wires for {{1}} will be
-dropped as it can freely introduced or eliminated using {{uniti}} and
-{{unite}}.
+U-shaped connectors. In later diagrams, the wires for {{1}} will be
+dropped as they can freely introduced or eliminated using {{uniti}}
+and {{unite}}.
 
@@
 not produce any specific value of type {{b}}, instead it produces a
-place holder for a value such that the value of type {{1/b}} is
-always the dual of the chosen one.
+placeholder for an unspecified value and dually, {{1/b}} is inhabited
+by the placeholder that indicates the absence of the unspecified
+value.
 
@@
 \end{multicols}
-
-\paragraph*{Coherence for Compact Closure with Multiplicative Inverses} 
-As before, the requisite coherence condition on Compact Closed
-Categories is applicable and is expressed below in terms of
-{{eta*}}/{{eps*}} and the multiplicative monoid.
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-The constructions for functions, involution, and trace are identical,
-replacing {{+}} by {{*}} mutatis mutandis.
 
@@
 
+\paragraph*{Coherence for Compact Closure with Multiplicative Inverses} 
+As before, the requisite coherence condition on Compact Closed
+Categories is applicable and is expressed below in terms of
+{{eta*}}/{{eps*}} and the multiplicative monoid.
+
+\begin{center}
+  \includegraphics{diagrams/coherence.pdf}
+\end{center}
+
 \paragraph*{Annihilation.}
@@
 
-\paragraph*{Zero Information.}
-As with {{eta+}}, we argue that {{eta*}} constructs no new
-information, instead it gives us the ability to explore as many worlds
-as permitted by the type {{b}}. So situate ourselves in any specific
-world, we have to supply additional information in the form of
-unifying the particle or anti-particle with a particle representing
-known information. Hence only by supplying information to a an
-entangled pair can we read information from it -- and always exactly
-as much information as we supplied.
-
-ADD MATH.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:monoidal-constructions}
+
+The constructions for functions, involution, and trace are identical,
+replacing {{+}} by {{*}}, \textit{mutatis mutandis}.
 
@@
 %% \label{sec:monoidal-constructions}
+
+\paragraph*{Zero Information.}
+As with {{eta+}}, we argue that {{eta*}} constructs no new
+information, instead it gives us the ability to explore as many worlds
+as permitted by the type {{b}}. So situate ourselves in any specific
+world, we have to supply additional information in the form of
+unifying the particle or anti-particle with a particle representing
+known information. Hence only by supplying information to a an
+entangled pair can we read information from it -- and always exactly
+as much information as we supplied.
+
+ADD MATH.
 

### 2012-03-11 rational.tex
blah

--- 2012-03-11 rational.tex
@@
 computation that takes a value of type~{{a}}, say \$20.00, and eventually
-produces the same \$20.00 value as its output. As the semantics of
-Sec.~\ref{sec:rat} formalizes, this computation `flows along the dotted
-line' with the following intermediate steps:
+produces another \$20.00 value as its output. As the semantics of
+Sec.~\ref{sec:rat} formalizes, this computation flows along the dotted line
+with the following intermediate steps:
 \begin{itemize}
@@
 \item We proceed to line (1) with the same \$20.00 but tagged as being
-  in the left summand of the sum type `{{a+0}}'; we indicate this value
+  in the left summand of the sum type {{a+0}}; we indicate this value
   as {{left 20}};
@@
 \end{itemize}
-The example illustrates that the empty type and negative types have a
-fascinating computational interpretation that has some flavor of
-continuations: negative types denote values that `backtrack to satisfy
-dependencies', or in other words act as `debts' that are satisfied by
-the backward flow of information.
+%% The example illustrates that the empty type and negative types have a
+%% computational interpretation related to continuations: negative types denote
+%% values that backtrack to satisfy dependencies, or in other words act as debts
+%% that are satisfied by the backward flow of information.
 
@@
 \end{center}
-In the case of negatives, the dotted line indicated the `flow of
-control' whereas for fractionals it indicates the `flow of
-constraints'. At the heart of logic programming is the idea of
-variables that capture constraints. Hence it is useful to trace the
-computation corresponding to the algebraic proof above, with the
-analogy to logic variables in mind.
+In the case of negatives, the dotted line indicated the flow of control
+whereas for fractionals it indicates the flow of constraints. At the heart of
+logic programming is the idea of variables that capture constraints. Hence it
+is useful to trace the computation corresponding to the algebraic proof
+above, with the analogy to logic variables in mind.
 
 As before, the execution begins at line~(0) with the value {{20}}. At
-line~(1) two values, {{20}} and {{()}}, flow forward. One can think of
-the value `{{()}}' (of type {{1}}) as `having a credit card'. The
-`credit card' isn't money, nor is it debt, but is the option to
-generate a `credit-debt' constraint.  At line~(2) we exercise this
-option and hence have three values: the initial value {{20}} flowing
-from line (1) and two entangled values, {{1/alpha}} and {{alpha}}. The
-{{alpha}} and {{1/alpha}} are unspecified values -- i.e. we don't yet
-know how much money we need to borrow, but we do know that what is
-borrowed must be what is returned. Hence {{alpha}} denotes the
-presence of an unknown quantity and dually {{1/alpha}} should be
-thought of as the absence of an unknown quantity.
-
-Unlike with negative types, wherein only one value existed at a time
-and the computation backtracked, what is fascinating here is that each
-of the three values \emph{exist at the same time}. In other words, the
-computation with fractions is realized with a completely different
-concurrent schedule in which every value independently and
-concurrently proceeds through its subcomputation.
-
-At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}},
-thereby fixing values of all three branches i.e. we have paid our debt
-of the missing information and the third subcomputation has the known
-value {{20}}. After payment of our debt, as line (4) shows, we have
-use of our `credit card' again.
+line~(1) two values, {{20}} and {{()}}, flow forward. One can think of the
+value {{()}} (of type {{1}}) as ``having a credit card.'' The credit card
+isn't money, nor is it debt, but is the option to generate a credit-debt
+constraint.  At line~(2) we exercise this option and hence have three values:
+the initial value {{20}} flowing from line (1) and two entangled values,
+{{1/alpha}} and {{alpha}}. The {{alpha}} and {{1/alpha}} are unspecified
+values, i.e., we don't yet know how much money we need to borrow, but we do
+know that what is borrowed must be what is returned. Hence~{{alpha}} denotes
+the presence of an unknown quantity and dually {{1/alpha}} should be thought
+of as the absence of an unknown quantity.
+% Unlike with negative types, wherein only one value existed at a time and the
+% computation backtracked, here we have three values that \emph{exist at the
+%   same time}. In other words, the computation with fractions is realized with
+% a schedule in which every value independently and concurrently proceeds
+% through its subcomputation.
+At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}}, thereby
+fixing values of all three branches, i.e., we have paid our debt of the missing
+information and the third subcomputation has the known value {{20}}. After
+payment of our debt, as line (4) shows, we have use of our credit card again.
 
@@
 
-The example illustrates that fractional types also have a fascinating
-computational interpretation that have some flavor of continuations:
-the fractional types denote values ({{1/alpha}}) that represent
-`missing information which must be supplied' in much the same sense
-that continuations denote evaluation contexts with `holes that must be
-filled'.  There are at least three fundamental points about the credit
-card examples that must be emphasized:
+The example illustrates that fractional types also have a computational
+interpretation that have some flavor of continuations: the fractional types
+denote values ({{1/alpha}}) that represent missing information that must be
+supplied in much the same sense that continuations denote evaluation contexts
+with holes that must be filled.  There are at least three fundamental points
+about the examples above that must be emphasized:
 
@@
 
-\item 
-The main reason credit card transactions are convenient is because
-they disentangle the propagation of the resources (money) from the
-propagation of the services. Not every transaction needs both the
-resources and services to be brought together: it is sufficient to
-have a promise that the demand for resources will be somehow
-satisfied, as long as the infrastructure can be trusted with such
-promises. This idea that dependencies can be freely decoupled and
-propagated can be a powerful programming tool and we leverage this in
-the construction of our novel SAT-solver (Sec. \ref{sec:sat-solver}).
+\item The main reason credit card transactions are convenient is because they
+  disentangle the propagation of the resources (money) from the propagation
+  of the services. Not every transaction needs both the resources and
+  services to be brought together: it is sufficient to have a promise that
+  the demand for resources will be somehow satisfied, as long as the
+  infrastructure can be trusted with such promises. This idea that
+  dependencies can be freely decoupled and propagated can be a powerful
+  programming tool and we leverage this in the construction of a novel
+  SAT-solver (see Sec.~\ref{sec:sat-solver}).
 
@@
 
-In the algebra of rational numbers {{1 = (1/a) * a}} if {{a}} is
-nonzero.  In our framework this algebraic identity corresponds to a
-type isomorphism {{1 <-> (1/a) * a}} (written alternately as
-{{a-o*a}}) whose witness maybe thought of as the `identity function'
-at the type {{a}}.  For example, for the datatype {{color = R|Gr|B}},
-there are three possible witnesses for this isomorphism mapping {{()}}
-to either {{(1/R,R)}} or {{(1/Gr,Gr)}} or {{(1/B,B)}} with the
-particular witness determined by further constraints encountered
-during the computation. More generally, the witness for the
-isomorphism above may be represented as {{(1/alpha,alpha)}}
-where~{{alpha}} is a fresh logic variable of type {{a}}. With this
-view of `{{a -o* a}}' functions, let us consider the following
-manipulations:
-
+The pair {{(1/a) * a}} corresponds to a function type written {{a-o*a}})
+whose witness maybe thought of as the identity function at the type {{a}}.
+For example, consider a datatype {{color = R|Gr|B}}, and let us consider the
+following manipulations:
+
+%% for the
+%% there are three possible
+%% witnesses for this isomorphism mapping {{()}} to either {{(1/R,R)}} or
+%% {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular witness determined by
+%% further constraints encountered during the computation. More generally, the
+%% witness for the isomorphism above may be represented as {{(1/alpha,alpha)}}
+%% where~{{alpha}} is a fresh logic variable of type {{a}}. With this view of
+%% {{a -o* a}} functions, 
+%% 
 %% Returning to our goal of creating the function that rotates the colors, we
@@
 
-At this point we have constructed a strange mix of two `{{a-o*a}}'
-functions; `inputs' of one function manifests as `outputs' of the
-other. If {{(1/alpha1,alpha2)}} is held by one subcomputation and
-{{(1/alpha2,alpha1)}} is held by another subcomputation, these
-`remixed functions' form a communication channel between the two
-concurrent subcomputations. Unifying {{1/alpha1}} with {{color}}
-`{{R}}' in one subcomputation, fixes {{alpha1}} to be {{R}} in the
-other. The type {{a}} thus takes the role of the type of the
-communication channel, indicating how much information can be
-communicated between the two subcomputations.  Depending on the choice
-of the type {{a}}, an arbitrary number of bits may be communicated.
+At this point we have constructed a strange mix of two {{a-o*a}} functions;
+inputs of one function manifest themselves as outputs of the other. If
+{{(1/alpha1,alpha2)}} is held by one subcomputation and {{(1/alpha2,alpha1)}}
+is held by another subcomputation, these remixed functions form a
+communication channel between the two concurrent subcomputations. Unifying
+{{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
+be {{R}} in the other. The type {{a}} thus takes the role of the type of the
+communication channel, indicating how much information can be communicated
+between the two subcomputations.  Depending on the choice of the type {{a}},
+an arbitrary number of bits may be communicated.
 
@@
 functions of the form {{a-o+a}}, witnessing isomorphisms of the form
-{{0<->(-a)+a}}). The `remixed' additive functions express control flow
-transfer between two subcomputations, \emph{only one of which exists}
-at any point -- i.e. they capture the essence of coroutines. The
-behaviour of additive and multiplicative functions are further
-discussed in Sec. \ref{sec:rat}.
-
-It should be evident that in a universe in which information is not
-guaranteed to be preserved by the computational infrastructure, the
-above slicing and dicing of functions would make no sense. However, in
-a computational infrastructure that is guaranteed to preserve
-information, computations such as the one above are guaranteed to
-``make sense.''
+{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
+between two subcomputations, \emph{only one of which exists} at any point,
+i.e., they capture the essence of coroutines. Additive and multiplicative
+functions are further discussed in Sec.~\ref{sec:rat}.  It should be evident
+that in a universe in which information is not guaranteed to be preserved by
+the computational infrastructure, the above slicing and dicing of functions
+would make no sense. Once the computational framework guarantees that
+information is maintained, it becomes possible to break dependencies,
+generate debts and constraints, and let these flow as first-class values,
+because they \emph{must} eventually be satisfied.
 
@@
 \item We extend {{langRev}} our reversible programming language of type
-  isomorphisms~\cite{rc2011,infeffects} (see Sec.~\ref{sec:pi}) with a notion
+  isomorphisms~\cite{rc2011,infeffects} (reviewed in Sec.~\ref{sec:pi}) with a notion
   of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
-  semantics of this extension is expressed by having a ``dual'' evaluator
+  semantics of this extension is expressed by having a \emph{dual} evaluator
   that reverses the flow of execution for negative
@@
 \item The morphisms that witness that {{0}} and {{1}} are the additive and
-  multiplicative unites are represented as shown below. Note that since there
+  multiplicative units are represented as shown below. Note that since there
   is no value of type 0, there can be no particle on a wire of type {{0}}.
@@
 The machine transitions below track the flow of particles through a
-circuit. The start machine state, {{<c,v,[]>}}, denotes the particle/value
-{{v}} about to be evaluated by the circuit {{c}}. The end machine state,
-{{[c, v, [] ]}}, denotes the situation where the particle {{v}} has exited
-the circuit {{c}}.
+circuit. The start machine state, {{<c,v,[]>}}, denotes the
+particle/value~{{v}} about to be evaluated by the circuit {{c}}. The end
+machine state, {{[c, v, [] ]}}, denotes the situation where the particle
+{{v}} has exited the circuit {{c}}.
 
@@
 \section{Negative and Fractional Types}
-
-% The structure of this section and the next are the same.
-
-This section introduces the syntax and graphical languages for
-negative types and fractionals.The general outline is as follows:
-\footnote{We refer the reader to Selinger's excellent survey
+\label{sec:neg}
+
+The structure of this section and the next are the same. The general outline
+is as follows:\footnote{We refer the reader to Selinger's excellent survey
   article of monoidal
@@
 After presenting the extension at the syntactic level, we discuss the
-categorical semantics informally via the graphical language. The
-operational semantics is presented in Sec.~\ref{sec:rat}. As will be
-detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
-compact closed structure provides several properties of interest: (i)
-morphisms or wires are allowed to run from right to left, (ii) the
-structure admits a trace that, depending on the situation, can be used
-to implement various notions of loops and
+categorical semantics informally via the graphical language. The operational
+semantics is presented in Sec.~\ref{sec:rat}.
+
+As will be detailed in the remainder of this section, the compact closed
+structure provides several properties of interest: (i) morphisms or wires are
+allowed to run from right to left, (ii) the structure admits a trace that,
+depending on the situation, can be used to implement various notions of loops
+and
 recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
-(iii) the structure includes an isomorphism showing that the dual
-operator is an involution, and (iv) the structure is equipped with
-objects representing higher-order functions and a morphism
-\textit{eval} that applies these functional objects. Interestingly
-each monoidal structure provides the same ingredients, resulting in
-two dualities, two traces, two involutions, and two notions of
-higher-order functions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+(iii) the structure includes an isomorphism showing that the dual operator is
+an involution, and (iv) the structure is equipped with objects representing
+higher-order functions and a morphism \textit{eval} that applies these
+functional objects. Interestingly each monoidal structure provides the same
+ingredients, resulting in two dualities, two traces, two involutions, and two
+notions of higher-order functions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Additive Duality in {{langRev}} }
-\label{sec:neg}
 
@@
 
-Hence, the intuitive way to read the diagrams is that they reverse the
-flow of information -- i.e. they capture a notion of \emph{negative
-  information flow}. In normal {{langRev}} circuits, information flows
-from left-to-right. The operator {{eps+}} causes information to flow
-from right-to-left.  The interpretation of {{eta+}} is the opposite:
-it turns a right-to-left flow of information in a circuit into a
-left-to-right flow.  The diagram below is an example of using {{eta+}}
-and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
-from the introduction. (Here the {{0}} wires have been shown for
-completeness.)
-\begin{center}
-\scalebox{1.2}{
+\paragraph*{Negative Information Flow.} 
+The intuitive way to read the diagrams is that they reverse the flow of
+information. In previous {{langRev}} circuits, information was always flowing
+from left to right. The presence of {{eps+}} operations cause information to
+flow from right to left which can be reversed again using {{eta+}}.  As an
+example, the diagram representing the isomorphism from the introduction is:
+\begin{center}
   \includegraphics{diagrams/algebra1.pdf}
-}
 \end{center}
@@
 % \paragraph*{Time Traveling.}
-A \emph{time traveling} intuition is also applicable here.  In a
-normal cicruit, as computational steps are taken values flow from left
-to right, i.e. computational time progresses from left to right.  The
-action of {{eps+}} causes values, aka information particles, to flow
-from right to left i.e. backwards in computational time.  The
-interesting thing about this interpretation is that particles that
-travel backwards in time get to see and interact with the history of
-particles that they coexist with. This gives us an intuitive
-interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
-Sec. \ref{sec:specific-constructions}) where it would otherwise seem
-that {{-b1}} and {{b2}} move in opposite directions. The backward flow
-of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
-equivalent to both particles moving backward in time.
+% A time traveling intuition is also applicable here: if view the
+% left-to-right direction of information flow as the canonical forward
+% direction of information, then the action of {{eps+}} allows values,
+% aka information particles, to flow backwards in time. Particles that
+% flow backwards in time get to see and interact with the history of
+% particles that they coexist with, i.e., values that they exist
+% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
+% (whose construction is shown in Section \ref{sec:neg-constructions}).
 
 %\subsection{Compact Closure with Additive Inverses}
-\emph{Coherence for Compact Closure with Additive Duality.}  The
-appropriate coherence condition for compact closed categories,
-expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-may be visualized (modulo {{0}} wires, swapping etc.) as explained by
-Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
-operational equivalence:
-
+\paragraph*{Coherence}
+The coherence condition for compact closed categories may be visualized as
+follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
 \begin{center}
@@
 \end{center}
-
-\emph{Non-termination.}  Compact closed categories contain {{trace}}
-operations and this implies that {{eta+}}/{{eps+}} allow the
-construction of a {{trace+}} operator (shown in Section
-\ref{sec:monoidal-constructions}).  As shown in previous work
-\cite{infeffects}, non-terminating computation can be constructed from
-     {{trace+}} and recursive types.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Fractional Types}
+It is intuitively clear that if the interpretations of {{eta+}} and {{eps+}}
+are to reverse the flow of execution, then this coherence condition is
+satisfied.
+
+\paragraph*{Trace.}
+Every compact closed categoy admits a trace. In the context of {{langRev}}
+with {{eta+}}/{{eps+}}, we get the following construction:
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/trace.pdf}
+\end{center}
+
+%subcode{proof} include main
+%@ c : b2 + b1 <-> b2 + b3
+%@@ trace c : b1 <-> b3
+
+\end{multicols}
+
+% \paragraph*{Zero Information.}
+% The fact that {{eta+}} (and hence {{eps+}}) introduces no information
+% can be understood is by viewing {{-b+b}} as a type whose value is
+% observable only if you `pay' {{b}} amount of information to the {{-b}}
+% branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
+% amount of information is returned to you. (TODO. more needs to be said
+% here.)
+
+% For this reason, our treatment of negative values in the context of
+% {{langRev}} is particularly simple. We will have a type $0$ and an
+% isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+% allows us to create a value and a corresponding debt out of nothing. Both the
+% value and its negative counterpart can flow anywhere in the system. Because
+% information is preserved, a closed program (which does not have any
+% ``dangling'' negative values) will eventually have to match the negative
+% value with some corresponding value, effectively using the isomorphism in the
+% right-to-left direction.
+
+\emph{Involution (Principium Contradictiones)}
+
+{{b <-> -(-b)}}
+
+\begin{center}
+  \includegraphics{diagrams/double_neg.pdf}
+\end{center}
+
+\emph{Functions}
+
+Mention LL and {{a -o b}}.
+
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+Let us use the shorthand {{b1 -o+ b2 = 0 <-> -b1 + b2}}. See
+Sec. \ref{sec:hof} for a discussion of the additive functions
+vs. multiplicative functions.
+
+\emph{Function composition.}
+
+\begin{multicols}{2}
+\scalebox{1.1}{
+    \includegraphics{diagrams/compose1.pdf}
+}
+
+\scalebox{0.8}{
+    \includegraphics{diagrams/compose.pdf}
+}
+\end{multicols}
+This is also equivalent to sequencing both the computation blocks.
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+\emph{Function application.}
+test.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/apply1.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/apply2.pdf}
+}
+\end{center}
+\end{multicols}
+
+\emph{Delimited continuation}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+\emph{Negation distributes over {{+}}. }
+
+{{-(b1 + b2) <-> (-b1) + (-b2)}}
+
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+\emph{Lifting a operation of positive types to negated types}
+
+Given {{c : b1 <-> b2}}
+
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
 
@@
 \label{sec:frac}
-
-The section follows the outline of the previous one, first presenting the new
-syntax, and discussing the graphical language and the properites implied by
-the underlying categorical structure.
 
@@
 
-The usual interpretation of {{b1*b2}} we have both a value of type
-{{b1}} and a value of type {{b2}}. This interpretation is maintained
-in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
-to be viewed as a fission point for a value of type {{b}} and its
-multiplicative inverse {{1/b}}. This naturally raises the question --
-which value of type {{b}} is produced? The isomorphism {{eta*}} does
-not produce any specific value of type {{b}}, instead it produces a
-place holder for a value such that the value of type {{1/b}} is
-always the dual of the chosen one.
+In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
+and a value of type {{b2}}. This interpretation is maintained in the presence
+of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be viewed as a
+fission point for a value of type {{b}} and its multiplicative inverse
+{{1/b}}. This naturally raises the question -- which value of type {{b}} is
+produced? The isomorphism {{eta*}} does not produce any specific value of
+type {{b}}, instead it produces a place holder for a value such that the
+value of type {{1/b}} is always the dual of the chosen one.
 
@@
 
-\paragraph*{Coherence for Compact Closure with Multiplicative Inverses} 
-As before, the requisite coherence condition on Compact Closed
-Categories is applicable and is expressed below in terms of
-{{eta*}}/{{eps*}} and the multiplicative monoid.
+\paragraph*{Coherence.}
+The following coherence condition is satisfied as before.
 
@@
 
-%%%%%
-
-\paragraph*{Unification.}
-A computational interpretation of this is that {{langRevEE}} acts like
-a logic programming system and {{eta*}} acts as the site of creation
-of a new typed logic variable of type {{b}} and its corresponding dual
-of type {{1/b}}. These two values are free to move independently
-through a circuit, but any operations on value affect what operations
-are possible on the other since they are both related by an underlying
-logic variable. Dually, {{eps*}} acts as a unification site for values
-flowing along both branches. If the values are not exactly duals, then
-the system fails to unify and hence the program is stuck. If the
-values are exactly duals, then unification succeeds and {{eta*}}
-returns {{1}}.
-
-\paragraph*{Entanglement and Anti-Particles.}
-A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
-quantum mechanics. The {{eta*}} is a site of fission that creates an
-information particle and its anti-particle. Both flow in the same
-direction in time. The produce particles are entangles in that actions
-on one (such as transformation by application of an isomorphism)
-affects the other. Further the fission site does not create a particle
-in any specific state, but in a superposition of all possible states
-as dictated by its type. In each possible world that the particle
-exists, it takes on a specific value inhibiting its type and
-correspondingly its entangled anti-particle takes the dual of the
-specific value.
-
-This also leads to the exciting point of view that entangled
-particle/anti-particle pairs act as functions in a physical
-first-class sense. We can ask the function to be applied to a value by
-unifying the anti-particle with a particle in a well-known state and
-examining is pair. We shall use this point of view in the development
-our SAT solver (see Sec. \ref{sec:sat-solver}).
-
-\paragraph*{Annihilation.}
-As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
-additive {{trace+}}, {{trace*}} gives us the ability to find the
-fixpoint of the constraint (expressed with combinators) that is
-traced. If the constraint has not satisfying values, there is not
-fixpoint possible -- we use the term `annihilation' to describe the
-corresponding undefined state of the program. Annihilation and a
-non-termination both describe undefined program execution -- while
-non-termination characterizes an iteration that fails to terminate,
-annihilation characterizes a constraint that has no fixpoint.
-
-\paragraph*{Zero Information.}
-As with {{eta+}}, we argue that {{eta*}} constructs no new
-information, instead it gives us the ability to explore as many worlds
-as permitted by the type {{b}}. So situate ourselves in any specific
-world, we have to supply additional information in the form of
-unifying the particle or anti-particle with a particle representing
-known information. Hence only by supplying information to a an
-entangled pair can we read information from it -- and always exactly
-as much information as we supplied.
-
-ADD MATH.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions independant of the  Monoid}
-\label{sec:monoidal-constructions}
-
-\begin{enumerate}
-\item 
-\emph{Trace}
-
-test.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/trace.pdf}
-\end{center}
-
-%subcode{proof} include main
-%@ c : b2 + b1 <-> b2 + b3
-%@@ trace c : b1 <-> b3
-
-\end{multicols}
-
-test.
-
-\item
-\emph{Involution (Principium Contradictiones)}
-
-{{b <-> -(-b)}}
-
-\begin{center}
-  \includegraphics{diagrams/double_neg.pdf}
-\end{center}
-
-\item
-\emph{Functions}
-
-Mention LL and {{a -o b}}.
-
-\begin{center}
-  \includegraphics{diagrams/function.pdf}
-\end{center}
-
-Let us use the shorthand {{b1 -o+ b2 = 0 <-> -b1 + b2}}. See
-Sec. \ref{sec:hof} for a discussion of the additive functions
-vs. multiplicative functions.
-
-\item
-\emph{Function composition.}
-
-\begin{multicols}{2}
-\scalebox{1.1}{
-    \includegraphics{diagrams/compose1.pdf}
-}
-
-\scalebox{0.8}{
-    \includegraphics{diagrams/compose.pdf}
-}
-\end{multicols}
-This is also equivalent to sequencing both the computation blocks.
-
-\begin{center}
-  \includegraphics{diagrams/compose2.pdf}
-\end{center}
-
-\item
-\emph{Function application.}
-test.
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.0}{
-  \includegraphics{diagrams/apply1.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.8}{
-  \includegraphics{diagrams/apply2.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item
-\emph{Delimited continuation}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}
-
-\item
-\emph{Negation distributes over {{+}}. }
-
-{{-(b1 + b2) <-> (-b1) + (-b2)}}
-
-\begin{center}
-  \includegraphics{diagrams/dist_neg_plus.pdf}
-\end{center}
-
-\item
-\emph{Lifting a operation of positive types to negated types}
-
-Given {{c : b1 <-> b2}}
-
-\begin{center}
-  \includegraphics{diagrams/neg_lift.pdf}
-\end{center}
-
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Conventional Higher Order Functions}
-\label{sec:hof}
-
-In a conventional language, higher-order functions capture scope and
-can be represented as first class values. On the other hand,
-{{langRevEE}} gives us two notions of functions both of which share
-similarities with conventional HOFs.
-
-\begin{itemize}
-\item 
-Additive functions are not really values, but are possible paths of
-execution. However, they can be encoded as values as follows: Consider
-two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
-{{bool}} can discriminate them and hence a value of type {{bool}} acts
-as the `address' of the function and is a first class value
-representing one of the {{b1 -o+ b2}} functions.  This is similar to
-how first classes functions a'la \lcal are represented by their
-addresses when compiled to conventional hardware.
-
-\item
-Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-are indeed first class values. However these functions can only be
-linearly used -- i.e. they represent a contraint that can be satisfied
-at most once in one `computational world'. It may be possible that to
-repeatedly invoke multiplicative functions with an appropriate
-recursive type whose unfoldings allow for repeated unification.
-\end{itemize}
-
-The exact encoding on HOF in {{langRevEE}} should shed light on
-several problems, including the connections to continuations discussed
-in Sec. \ref{sec:related}. They should also allow for a natural
-encoding of HOFs in the translations of \lcal to {{langRevT}} and
-indicate a monadic encapsulation of of information effects which are
-currrently structured as an arrow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \paragraph*{Unification.}
+% A computational interpretation of this is that {{langRevEE}} acts like
+% a logic programming system and {{eta*}} acts as the site of creation
+% of a new typed logic variable of type {{b}} and its corresponding dual
+% of type {{1/b}}. These two values are free to move independently
+% through a circuit, but any operations on value affect what operations
+% are possible on the other since they are both related by an underlying
+% logic variable. Dually, {{eps*}} acts as a unification site for values
+% flowing along both branches. If the values are not exactly duals, then
+% the system fails to unify and hence the program is stuck. If the
+% values are exactly duals, then unification succeeds and {{eta*}}
+% returns {{1}}.
+
+% \paragraph*{Entanglement and Anti-Particles.}
+% A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
+% quantum mechanics. The {{eta*}} is a site of fission that creates an
+% information particle and its anti-particle. Both flow in the same
+% direction in time. The produce particles are entangles in that actions
+% on one (such as transformation by application of an isomorphism)
+% affects the other. Further the fission site does not create a particle
+% in any specific state, but in a superposition of all possible states
+% as dictated by its type. In each possible world that the particle
+% exists, it takes on a specific value inhibiting its type and
+% correspondingly its entangled anti-particle takes the dual of the
+% specific value.
+
+% This also leads to the exciting point of view that entangled
+% particle/anti-particle pairs act as functions in a physical
+% first-class sense. We can ask the function to be applied to a value by
+% unifying the anti-particle with a particle in a well-known state and
+% examining is pair. We shall use this point of view in the development
+% our SAT solver (see Sec. \ref{sec:sat-solver}).
+
+% \paragraph*{Annihilation.}
+% As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
+% additive {{trace+}}, {{trace*}} gives us the ability to find the
+% fixpoint of the constraint (expressed with combinators) that is
+% traced. If the constraint has not satisfying values, there is not
+% fixpoint possible -- we use the term `annihilation' to describe the
+% corresponding undefined state of the program. Annihilation and a
+% non-termination both describe undefined program execution -- while
+% non-termination characterizes an iteration that fails to terminate,
+% annihilation characterizes a constraint that has no fixpoint.
+
+% \paragraph*{Zero Information.}
+% As with {{eta+}}, we argue that {{eta*}} constructs no new
+% information, instead it gives us the ability to explore as many worlds
+% as permitted by the type {{b}}. So situate ourselves in any specific
+% world, we have to supply additional information in the form of
+% unifying the particle or anti-particle with a particle representing
+% known information. Hence only by supplying information to a an
+% entangled pair can we read information from it -- and always exactly
+% as much information as we supplied.
+
+% ADD MATH.
+
+% %%%%%%%%%%%%%%%%%%%%%%%%
+% \subsection{Conventional Higher-Order Functions}
+% \label{sec:hof}
+
+% In a conventional language, higher-order functions capture scope and can be
+% represented as first-class values. In {{langRevEE}}, we have two notions of
+% functions both of which share similarities with conventional higher-order
+% functions.
+
+% \begin{itemize}
+% \item 
+% Additive functions are not really values, but are possible paths of
+% execution. However, they can be encoded as values as follows: Consider
+% two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
+% {{bool}} can discriminate them and hence a value of type {{bool}} acts
+% as the address of the function and is a first class value
+% representing one of the {{b1 -o+ b2}} functions.  %% This is similar to
+% %% how first classes functions a'la \lcal are represented by their
+% %% addresses when compiled to conventional hardware.
+
+% \item
+% Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+% are indeed first-class values. However these functions can only be
+% linearly used, i.e., they represent a constraint that can be satisfied
+% at most once.
+% %%  in one `computational world'. It may be possible that to
+% %% repeatedly invoke multiplicative functions with an appropriate
+% %% recursive type whose unfoldings allow for repeated unification.
+% \end{itemize}
+
+% The exact encoding on HOF in {{langRevEE}} should shed light on
+% several problems, including the connections to continuations discussed
+% in Sec. \ref{sec:related}. They should also allow for a natural
+% encoding of HOFs in the translations of \lcal to {{langRevT}} and
+% indicate a monadic encapsulation of of information effects which are
+% currrently structured as an arrow.
+
+%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions specific to Negatives and Fractions}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%
 \subsection{Unification and Logic Programming}
@@
 \paragraph{Backward evaluator.}
-Here we have written a ``backward'' evaluator in the same style --
-i.e. evaluation {{c v}} in this evaluator is equivalent to evaluating
+Here we have written a ``backward'' evaluator in the same style, 
+i.e., evaluation {{c v}} in this evaluator is equivalent to evaluating
 {{C^{dagger} v}} in the forward evaluator.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics of {{langRevEE}} }
@@
 
-\emph{Zero Information.}  The intuitive explanation that {{eta+}} and
-hence {{eps+}} introduces no information effects is that {{-b+b}} as a
-type whose value is observable only if you `pay' {{b}} amount of
-information to the {{-b}} branch. When {{b}} amount of information is
-supplied {{(-b+b)+b=b}} amount of information is returned to
-you. (TODO. more needs to be said here.)
-
-For this reason, our treatment of negative values in the context of
-{{langRev}} is particularly simple. We will have a type $0$ and an
-isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-allows us to create a value and a corresponding debt out of nothing. Both the
-value and its negative counterpart can flow anywhere in the system. Because
-information is preserved, a closed program (which does not have any
-``dangling'' negative values) will eventually have to match the negative
-value with some corresponding value, effectively using the isomorphism in the
-right-to-left direction.
-
 \paragraph*{Observability.} 
@@
 
-Additive traces correspond to iteration and multiplicative traces
-correspond to fixpoint of constraints. Addtive traces have been
-studied in the past \cite{infeffects} and here we focus on the
-multiplicative trace.
+% Additive traces correspond to iteration and multiplicative traces
+% correspond to fixpoint of constraints. Addtive traces have been
+% studied in the past \cite{infeffects} and here we focus on the
+% multiplicative trace.
 
 The construction presented here is a novel SAT-solver that relies on
-{{trace*}} for its solution. It varies from previous classical SAT
-solvers in that there is not explicit search operation on the boolean
-space. It also varies from previous quantum SAT solvers in that it
-does uses a {{trace}} operation to compute a fixpoint which is the
-solution following which an isomorphic clone operation allows us to
-examine the result.  Though we focus on SAT here, this construction
-can be generalized to any constraint satisfaction problem whose search
-space can be represented by a recursive type and for which an
-isomorphic clone operation can be constructed. Before we start, let us
-recall two constructions:
+{{trace*}} for its solution. It varies from previous classical SAT solvers
+\amr{must cite ???} in that there is no explicit search operation on the
+boolean space. It also varies from previous quantum SAT solvers \amr{must
+  cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
+which is the solution following which an isomorphic clone operation allows us
+to examine the result.  Though we focus on SAT here, this construction can be
+generalized to any constraint satisfaction problem whose search space can be
+represented by a recursive type and for which an isomorphic clone operation
+can be constructed. Before we start, let us recall two constructions:
 
@@
 \emph{A Total Annihilation Circuit.}  Consider the circuit below that
-has no observable output -- i.e. for all inputs it is
+has no observable output, i.e., for all inputs it is
 annihilating. This happens because it constructs a boolean and its
@@
 appropriate initial value by checking the heap and negating a second
-control wire, if the values do not match -- i.e. the dotted part in
+control wire, if the values do not match, i.e., the dotted part in
 the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
@@
 manipulation can be viewed as a {{langRevEE}} combinator and it indeed
-does constructively witness the isomorphism -- i.e. you can run it and
+does constructively witness the isomorphism, i.e., you can run it and
 view one data structure encoded by the other. In this context,

### 2012-03-11 rational.tex
sec fractional types

--- 2012-03-11 rational.tex
@@
 
-%%%%%
-
-\paragraph*{Unification.}
-A computational interpretation of this is that {{langRevEE}} acts like
-a logic programming system and {{eta*}} acts as the site of creation
-of a new typed logic variable of type {{b}} and its corresponding dual
-of type {{1/b}}. These two values are free to move independently
-through a circuit, but any operations on value affect what operations
-are possible on the other since they are both related by an underlying
-logic variable. Dually, {{eps*}} acts as a unification site for values
-flowing along both branches. If the values are not exactly duals, then
-the system fails to unify and hence the program is stuck. If the
-values are exactly duals, then unification succeeds and {{eta*}}
-returns {{1}}.
-
-\paragraph*{Entanglement and Anti-Particles.}
-A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
-quantum mechanics. The {{eta*}} is a site of fission that creates an
-information particle and its anti-particle. Both flow in the same
-direction in time. The produce particles are entangles in that actions
-on one (such as transformation by application of an isomorphism)
-affects the other. Further the fission site does not create a particle
-in any specific state, but in a superposition of all possible states
-as dictated by its type. In each possible world that the particle
-exists, it takes on a specific value inhibiting its type and
+This can be understood in terms of the \emph{logic programming} as
+follows. Operator {{eta* : 1 <-> (1/b)*b}} acts as the site of
+creation of a new typed logic variable and constructs from it a value
+of type {{b}} and its dual of type {{1/b}}. These two values are free
+to move independently through the circuit. Operations on one value
+affect what operations are possible on the other since they are both
+related by an underlying logic variable. Dually, {{eps*}} acts as a
+unification site for values of type {{b}} and {{1/b}} flowing along
+both branches. If the values are not unifiable duals, then the
+specific program sequence fails. Like with logic programming, one can
+has many worlds where some might fail, while others make
+progress. Sec. \ref{sec:rat} has more details.  If the values are
+exactly duals, then unification succeeds and {{eta*}} returns {{1}}.
+
+% \paragraph*{Entanglement and Anti-Particles.}
+
+An analogy inspired by quantum mechanics is also applicable here.  The
+operator {{eta*}} is a site of fission that creates \emph{an entangled
+  particle and anti-particle}. These particles can be thought to be in
+a \emph{superposition} of states determined by their type. Both flow
+in the same direction in time.  In each possible world that the
+particle exists, it takes on a specific value inhibiting its type and
 correspondingly its entangled anti-particle takes the dual of the
-specific value.
-
-This also leads to the exciting point of view that entangled
-particle/anti-particle pairs act as functions in a physical
-first-class sense. We can ask the function to be applied to a value by
-unifying the anti-particle with a particle in a well-known state and
-examining is pair. We shall use this point of view in the development
-our SAT solver (see Sec. \ref{sec:sat-solver}).
-
-\paragraph*{Coherence for Compact Closure with Multiplicative Inverses} 
-As before, the requisite coherence condition on Compact Closed
-Categories is applicable and is expressed below in terms of
+specific value. Since the particles are entangled, actions on one
+(such as transformation by application of an isomorphism) affects the
+other in much the same sense as \emph{action at a distance}. The
+`{{a-o*b}}' functions mentioned in the introduction, are first-class
+values and they correspond to these entagled pairs. Finally {{eps*}}
+operations are sites where corresponding particle and anti-particles
+annihilate each other. While appealing as an analogy, this description
+does not imply (or preclude) any formal connection with Physics.
+
+%% We can ask the function to be applied to a value by
+%% unifying the anti-particle with a particle in a well-known state and
+%% examining is pair. We shall use this point of view in the development
+%% our SAT solver (see Sec. \ref{sec:sat-solver}).
+
+%%  This also leads to the exciting point of view that
+%% entangled such particle/anti-particle pairs, which are first class
+%% values, act as functions in a physical first-class sense.
+
+%% Further the fission site does not create a particle in any specific
+%% state, but in a superposition of all possible states as dictated by
+%% its type.
+
+% Since such pairs are first-class entities, they correspond to
+
+\emph{Coherence for Compact Closure with Multiplicative Duality.}  As
+before, the triangle axiom which is the coherence condition on compact
+closed categories is applicable and is expressed below in terms of
 {{eta*}}/{{eps*}} and the multiplicative monoid.
-
 \begin{center}
@@
 
-\paragraph*{Annihilation.}
-As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
-additive {{trace+}}, {{trace*}} gives us the ability to find the
-fixpoint of the constraint (expressed with combinators) that is
-traced. If the constraint has not satisfying values, there is not
-fixpoint possible -- we use the term `annihilation' to describe the
-corresponding undefined state of the program. Annihilation and a
-non-termination both describe undefined program execution -- while
-non-termination characterizes an iteration that fails to terminate,
-annihilation characterizes a constraint that has no fixpoint.
+\emph{Annihilation.}  
+With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
+which gives us the ability to find the fixpoint of a constraint.  If
+the constraint has no satisfying values, there is no fixpoint possible
+-- we use the term \emph{annihilation} to describe the corresponding
+undefined state of the program.
+
+Annihilation and non-termination describe undefined program
+execution. While non-termination characterizes an iteration that fails
+to terminate, annihilation characterizes a constraint that has no
+fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
+annihilating only the program states that do not match the boolean
+satisfiability constraint.
 

### 2012-03-11 rational.tex

--- 2012-03-11 rational.tex
@@
 \section{Negative and Fractional Types}
-\label{sec:neg}
-
-The structure of this section and the next are the same. The general outline
-is as follows:\footnote{We refer the reader to Selinger's excellent survey
+
+% The structure of this section and the next are the same.
+
+This section introduces the syntax and graphical languages for
+negative types and fractionals.The general outline is as follows:
+\footnote{We refer the reader to Selinger's excellent survey
   article of monoidal
@@
 After presenting the extension at the syntactic level, we discuss the
+%% <<<<<<< local
+%% categorical semantics informally via the graphical language. The
+%% operational semantics is presented in Sec.~\ref{sec:rat}. As will be
+%% detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
+%% compact closed structure provides several properties of interest: (i)
+%% morphisms or wires are allowed to run from right to left, (ii) the
+%% structure admits a trace that, depending on the situation, can be used
+%% to implement various notions of loops and
+%% =======
 categorical semantics informally via the graphical language. The operational
@@
 and
+%% >>>>>>> other
 recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
-(iii) the structure includes an isomorphism showing that the dual operator is
-an involution, and (iv) the structure is equipped with objects representing
-higher-order functions and a morphism \textit{eval} that applies these
-functional objects. Interestingly each monoidal structure provides the same
-ingredients, resulting in two dualities, two traces, two involutions, and two
-notions of higher-order functions.
+(iii) the structure includes an isomorphism showing that the dual
+operator is an involution, and (iv) the structure is equipped with
+objects representing higher-order functions and a morphism
+\textit{eval} that applies these functional objects. Interestingly
+each monoidal structure provides the same ingredients, resulting in
+two dualities, two traces, two involutions, and two notions of
+higher-order functions.
 
@@
 
+%% <<<<<<< local
+%% Hence, the intuitive way to read the diagrams is that they reverse the
+%% flow of information -- i.e. they capture a notion of \emph{negative
+%%   information flow}. In normal {{langRev}} circuits, information flows
+%% from left-to-right. The operator {{eps+}} causes information to flow
+%% from right-to-left.  The interpretation of {{eta+}} is the opposite:
+%% it turns a right-to-left flow of information in a circuit into a
+%% left-to-right flow.  The diagram below is an example of using {{eta+}}
+%% and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
+%% from the introduction. (Here the {{0}} wires have been shown for
+%% completeness.)
+%% =======
 \paragraph*{Negative Information Flow.} 
@@
 example, the diagram representing the isomorphism from the introduction is:
-\begin{center}
+%% >>>>>>> other
+\begin{center}
+\scalebox{1.2}{
   \includegraphics{diagrams/algebra1.pdf}
+}
 \end{center}
@@
 % \paragraph*{Time Traveling.}
+%% <<<<<<< local
+%% A \emph{time traveling} intuition is also applicable here.  In a
+%% normal cicruit, as computational steps are taken values flow from left
+%% to right, i.e. computational time progresses from left to right.  The
+%% action of {{eps+}} causes values, aka information particles, to flow
+%% from right to left i.e. backwards in computational time.  The
+%% interesting thing about this interpretation is that particles that
+%% travel backwards in time get to see and interact with the history of
+%% particles that they coexist with. This gives us an intuitive
+%% interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
+%% Sec. \ref{sec:specific-constructions}) where it would otherwise seem
+%% that {{-b1}} and {{b2}} move in opposite directions. The backward flow
+%% of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
+%% equivalent to both particles moving backward in time.
+%% =======
 % A time traveling intuition is also applicable here: if view the
@@
 % (whose construction is shown in Section \ref{sec:neg-constructions}).
+%% >>>>>>> other
 
 %\subsection{Compact Closure with Additive Inverses}
+%% <<<<<<< local
+%% \emph{Coherence for Compact Closure with Additive Duality.}  The
+%% appropriate coherence condition for compact closed categories,
+%% expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
+%% may be visualized (modulo {{0}} wires, swapping etc.) as explained by
+%% Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
+%% operational equivalence:
+
+%% =======
 \paragraph*{Coherence}
@@
 follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
+%% >>>>>>> other
 \begin{center}
@@
 
+%% <<<<<<< local
+%% \emph{Non-termination.}  Compact closed categories contain {{trace}}
+%% operations and this implies that {{eta+}}/{{eps+}} allow the
+%% construction of a {{trace+}} operator (shown in Section
+%% \ref{sec:monoidal-constructions}).  As shown in previous work
+%% \cite{infeffects}, non-terminating computation can be constructed from
+%%      {{trace+}} and recursive types.
+%% =======
 \begin{multicols}{2}
@@
 \end{multicols}
+%% >>>>>>> other
 
@@
 
+%% <<<<<<< local
+%% %% The section follows the outline of the previous one, first presenting
+%% %% the new syntax, and discussing the graphical language and the
+%% %% properites implied by the underlying categorical structure.
+
+%% =======
+%% >>>>>>> other
 Multiplicative duality is added to {{langRev}} by adding types
-{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}} with
-the following type judgements.
+{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.
 
@@
 
+%% <<<<<<< local
+%% The usual interpretation of {{b1*b2}} we have both a value of type
+%% {{b1}} and a value of type {{b2}}. This interpretation is maintained
+%% in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
+%% to be viewed as a fission point for a value of type {{b}} and its
+%% multiplicative inverse {{1/b}}. This naturally raises the question --
+%% which value of type {{b}} is produced? The isomorphism {{eta*}} does
+%% not produce any specific value of type {{b}}, instead it produces a
+%% placeholder for an unspecified value and dually, {{1/b}} is inhabited
+%% by the placeholder that indicates the absence of the unspecified
+%% value.
+%% =======
 In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
@@
 value of type {{1/b}} is always the dual of the chosen one.
+%% >>>>>>> other
 
@@
 
+%% <<<<<<< local
+%% =======
+%% >>>>>>> other
+
+%% <<<<<<< local
+This can be understood in terms of the \emph{logic programming} as
+follows. Operator {{eta* : 1 <-> (1/b)*b}} acts as the site of
+creation of a new typed logic variable and constructs from it a value
+of type {{b}} and its dual of type {{1/b}}. These two values are free
+to move independently through the circuit. Operations on one value
+affect what operations are possible on the other since they are both
+related by an underlying logic variable. Dually, {{eps*}} acts as a
+unification site for values of type {{b}} and {{1/b}} flowing along
+both branches. If the values are not unifiable duals, then the
+specific program sequence fails. Like with logic programming, one can
+has many worlds where some might fail, while others make
+progress. Sec. \ref{sec:rat} has more details.  If the values are
+exactly duals, then unification succeeds and {{eta*}} returns {{1}}.
+
+% \paragraph*{Entanglement and Anti-Particles.}
+
+%% An analogy inspired by quantum mechanics is also applicable here.  The
+%% operator {{eta*}} is a site of fission that creates \emph{an entangled
+%%   particle and anti-particle}. These particles can be thought to be in
+%% a \emph{superposition} of states determined by their type. Both flow
+%% in the same direction in time.  In each possible world that the
+%% particle exists, it takes on a specific value inhibiting its type and
+%% correspondingly its entangled anti-particle takes the dual of the
+%% specific value. Since the particles are entangled, actions on one
+%% (such as transformation by application of an isomorphism) affects the
+%% other in much the same sense as \emph{action at a distance}. The
+%% `{{a-o*b}}' functions mentioned in the introduction, are first-class
+%% values and they correspond to these entagled pairs. Finally {{eps*}}
+%% operations are sites where corresponding particle and anti-particles
+%% annihilate each other. While appealing as an analogy, this description
+%% does not imply (or preclude) any formal connection with Physics.
+
+%% We can ask the function to be applied to a value by
+%% unifying the anti-particle with a particle in a well-known state and
+%% examining is pair. We shall use this point of view in the development
+%% our SAT solver (see Sec. \ref{sec:sat-solver}).
+
+%%  This also leads to the exciting point of view that
+%% entangled such particle/anti-particle pairs, which are first class
+%% values, act as functions in a physical first-class sense.
+
+%% Further the fission site does not create a particle in any specific
+%% state, but in a superposition of all possible states as dictated by
+%% its type.
+
+% Since such pairs are first-class entities, they correspond to
+
+%% \emph{Coherence for Compact Closure with Multiplicative Duality.}  As
+%% before, the triangle axiom which is the coherence condition on compact
+%% closed categories is applicable and is expressed below in terms of
+%% {{eta*}}/{{eps*}} and the multiplicative monoid.
+%% =======
+%% >>>>>>> other
+
 \paragraph*{Coherence.}
@@
 
-The constructions for functions, involution, and trace are identical,
-replacing {{+}} by {{*}} mutatis mutandis.
-
+\emph{Annihilation.}  
+With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
+which gives us the ability to find the fixpoint of a constraint.  If
+the constraint has no satisfying values, there is no fixpoint possible
+-- we use the term \emph{annihilation} to describe the corresponding
+undefined state of the program.
+
+%% <<<<<<< local
+Annihilation and non-termination describe undefined program
+execution. While non-termination characterizes an iteration that fails
+to terminate, annihilation characterizes a constraint that has no
+fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
+annihilating only the program states that do not match the boolean
+satisfiability constraint.
+%% =======
 % \paragraph*{Unification.}
@@
 % functions.
+%% >>>>>>> other
 
@@
 
-% \item
-% Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-% are indeed first-class values. However these functions can only be
-% linearly used, i.e., they represent a constraint that can be satisfied
-% at most once.
-% %%  in one `computational world'. It may be possible that to
-% %% repeatedly invoke multiplicative functions with an appropriate
-% %% recursive type whose unfoldings allow for repeated unification.
-% \end{itemize}
+%% <<<<<<< local
+The constructions for functions, involution, and trace are identical,
+replacing {{+}} by {{*}}, \textit{mutatis mutandis}.
+
+%% \begin{enumerate}
+%% \item 
+%% \emph{Trace}
+%% =======
+%% % \item
+%% % Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+%% % are indeed first-class values. However these functions can only be
+%% % linearly used, i.e., they represent a constraint that can be satisfied
+%% % at most once.
+%% % %%  in one `computational world'. It may be possible that to
+%% % %% repeatedly invoke multiplicative functions with an appropriate
+%% % %% recursive type whose unfoldings allow for repeated unification.
+%% % \end{itemize}
+%% >>>>>>> other
 
@@
 %% \label{sec:monoidal-constructions}
+
+\paragraph*{Zero Information.}
+As with {{eta+}}, we argue that {{eta*}} constructs no new
+information, instead it gives us the ability to explore as many worlds
+as permitted by the type {{b}}. So situate ourselves in any specific
+world, we have to supply additional information in the form of
+unifying the particle or anti-particle with a particle representing
+known information. Hence only by supplying information to a an
+entangled pair can we read information from it -- and always exactly
+as much information as we supplied.
+
+ADD MATH.
+
+\emph{Zero Information.}  The intuitive explanation that {{eta+}} and
+hence {{eps+}} introduces no information effects is that {{-b+b}} as a
+type whose value is observable only if you `pay' {{b}} amount of
+information to the {{-b}} branch. When {{b}} amount of information is
+supplied {{(-b+b)+b=b}} amount of information is returned to
+you. (TODO. more needs to be said here.)
+
+For this reason, our treatment of negative values in the context of
+{{langRev}} is particularly simple. We will have a type $0$ and an
+isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+allows us to create a value and a corresponding debt out of nothing. Both the
+value and its negative counterpart can flow anywhere in the system. Because
+information is preserved, a closed program (which does not have any
+``dangling'' negative values) will eventually have to match the negative
+value with some corresponding value, effectively using the isomorphism in the
+right-to-left direction.
 

### 2012-03-11 rational.tex
resurrected some paras

@@
 \end{itemize}
-%% The example illustrates that the empty type and negative types have a
-%% computational interpretation related to continuations: negative types denote
-%% values that backtrack to satisfy dependencies, or in other words act as debts
-%% that are satisfied by the backward flow of information.
+\roshan{Why was this deleted?}
+The example illustrates that the empty type and negative types have a
+computational interpretation related to continuations: negative types denote
+values that backtrack to satisfy dependencies, or in other words act as debts
+that are satisfied by the backward flow of information.
 
@@
 of as the absence of an unknown quantity.
-% Unlike with negative types, wherein only one value existed at a time and the
-% computation backtracked, here we have three values that \emph{exist at the
-%   same time}. In other words, the computation with fractions is realized with
-% a schedule in which every value independently and concurrently proceeds
-% through its subcomputation.
 At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}}, thereby
@@
 payment of our debt, as line (4) shows, we have use of our credit card again.
+
+\roshan{Again, why was this deleted?}
+Unlike with negative types, wherein only one value existed at a time and the
+computation backtracked, here we have three values that \emph{exist at the
+  same time}. In other words, the computation with fractions is realized with
+a schedule in which every value independently and concurrently proceeds
+through its subcomputation.
 

### 2012-03-11 rational.tex
small edit

@@
 
-\roshan{Again, why was this deleted?}
+\roshan{Also, why was this deleted?}
 Unlike with negative types, wherein only one value existed at a time and the
@@
 %% <<<<<<< local
-The constructions for functions, involution, and trace are identical,
+\emph{Constructions.} The constructions for functions, involution, and trace are identical,
 replacing {{+}} by {{*}}, \textit{mutatis mutandis}.

### 2012-03-11 rational.tex
small fixes

@@
 
-Given {{c : b1 <-> b2}}
-
+\begin{multicols}{2}
 \begin{center}
@@
 \end{center}
+
+%subcode{proof} include main
+%@ c : b1 <-> b2
+%@@ neg c : (-b1) <-> (-b2)
+\end{multicols}
 
@@
 %% <<<<<<< local
-This can be understood in terms of the \emph{logic programming} as
-follows. Operator {{eta* : 1 <-> (1/b)*b}} acts as the site of
-creation of a new typed logic variable and constructs from it a value
-of type {{b}} and its dual of type {{1/b}}. These two values are free
-to move independently through the circuit. Operations on one value
-affect what operations are possible on the other since they are both
-related by an underlying logic variable. Dually, {{eps*}} acts as a
-unification site for values of type {{b}} and {{1/b}} flowing along
-both branches. If the values are not unifiable duals, then the
-specific program sequence fails. Like with logic programming, one can
-has many worlds where some might fail, while others make
-progress. Sec. \ref{sec:rat} has more details.  If the values are
-exactly duals, then unification succeeds and {{eta*}} returns {{1}}.
+In terms of \emph{logic programming}, {{eta* : 1 <-> (1/b)*b}} acts as
+the site of creation of a new typed logic variable and constructs two
+values using it: a value of type {{b}} and its dual of type
+{{1/b}}. These two values are free to move independently through the
+computation. Operations on one value affect what operations are possible
+on the other since they are both related by an underlying logic
+variable. The operator {{eps*}} acts as a unification site for values of
+type {{b}} and {{1/b}} flowing along both branches. If the values are
+not unifiable duals, then the specific program sequence fails. Like
+with logic programming, one can has many worlds where some might fail,
+while others make progress. Sec. \ref{sec:rat} has more details.  If
+the values are exactly duals, then unification succeeds and {{eta*}}
+returns {{1}}.
 
@@
 
-\item 
-\emph{ {{eps_{fst} }} }
-
-{{(-b1)*b2 + b1*b2 <-> 0}}
-
-\begin{center}
-  \includegraphics{diagrams/eps_fst.pdf}
-\end{center}
+%% \item 
+%% \emph{ {{eps_{fst} }} }
+
+%% {{(-b1)*b2 + b1*b2 <-> 0}}
 
@@
 \begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/eps_fst.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.9}{
   \includegraphics{diagrams/mult_neg.pdf}
-\end{center}
-
-The following isomorphism can be constructed similarly
+}
+\end{center}
+
+\item
+\emph{Multiplying Negatives}
 

### 2012-03-11 rational.tex
some work on constructions

@@
 \noindent 
-\textit{Example.} The following combinator is represented by the given
+\textit{Example.}  We use the type {{bool}} as a shorthand to denote
+the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
+to be {{false}}. The following combinator is represented by the given
 diagram:
 
-{{c : b * (1+1) <-> b + b}}
+{{c : b * bool <-> b + b}}
 
@@
 %% <<<<<<< local
-%% A \emph{time traveling} intuition is also applicable here.  In a
-%% normal cicruit, as computational steps are taken values flow from left
-%% to right, i.e. computational time progresses from left to right.  The
-%% action of {{eps+}} causes values, aka information particles, to flow
-%% from right to left i.e. backwards in computational time.  The
-%% interesting thing about this interpretation is that particles that
-%% travel backwards in time get to see and interact with the history of
-%% particles that they coexist with. This gives us an intuitive
-%% interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
-%% Sec. \ref{sec:specific-constructions}) where it would otherwise seem
-%% that {{-b1}} and {{b2}} move in opposite directions. The backward flow
-%% of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
-%% equivalent to both particles moving backward in time.
 %% =======
@@
 
-As with additive duals, we represent {{eta*}} and {{eps*}} using
-U-shaped connectors. In later diagrams, the wires for {{1}} will be
-dropped as they can freely introduced or eliminated using {{uniti}}
-and {{unite}}.
+For convenience we sometime use the notation {{b1/b2}} to indicate the
+pair {{b1 * (1/b2)}}.  As with additive duals, we represent {{eta*}}
+and {{eps*}} using U-shaped connectors. In later diagrams, the wires
+for {{1}} will be dropped as they can freely introduced or eliminated
+using {{uniti}} and {{unite}}.
 
@@
 
-%% {{(-b1)*b2 + b1*b2 <-> 0}}
+%% {{(
 
@@
 
+To build these isomorphisms, we first build an intermeidate
+construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
 \begin{center}
@@
 
+The isomorphism {{(-b1) * b2 <-> -(b1 * b2)}} can be constructed in
+terms of {{eps_{fst} }} as shown below.
+
 \begin{center}
@@
 
+The second isomorphism can be built in the same way by merely swapping
+the arguments.
+
 \item
@@
 {{b1 * b2 <-> (-b1)*(-b2)}}
+
+This isomorphism can be built using \emph{involution} constructed in
+Sec \ref{sec:neg} and the isomorphisms above, and correspond to the
+algebraic manipulation:
+
+%subcode{opsem} include main
+%! columnStyle = rc
+%   & b1 * b2
+% = & -(-(b1*b2)) 
+% = & -((-b1)*b2)
+% = & (-b1)*(-b2)
 
@@
 \paragraph*{QC} 
+A \emph{time traveling} intuition is also applicable here.  In a
+normal cicruit, as computational steps are taken values flow from left
+to right, i.e. computational time progresses from left to right.  The
+action of {{eps+}} causes values, aka information particles, to flow
+from right to left i.e. backwards in computational time.  The
+interesting thing about this interpretation is that particles that
+travel backwards in time get to see and interact with the history of
+particles that they coexist with. This gives us an intuitive
+interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
+Sec. \ref{sec:specific-constructions}) where it would otherwise seem
+that {{-b1}} and {{b2}} move in opposite directions. The backward flow
+of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
+equivalent to both particles moving backward in time.
+
+An analogy inspired by quantum mechanics is also applicable here.  The
+operator {{eta*}} is a site of fission that creates \emph{an entangled
+  particle and anti-particle}. These particles can be thought to be in
+a \emph{superposition} of states determined by their type. Both flow
+in the same direction in time.  In each possible world that the
+particle exists, it takes on a specific value inhibiting its type and
+correspondingly its entangled anti-particle takes the dual of the
+specific value. Since the particles are entangled, actions on one
+(such as transformation by application of an isomorphism) affects the
+other in much the same sense as \emph{action at a distance}. The
+`{{a-o*b}}' functions mentioned in the introduction, are first-class
+values and they correspond to these entagled pairs. Finally {{eps*}}
+operations are sites where corresponding particle and anti-particles
+annihilate each other. While appealing as an analogy, this description
+does not imply (or preclude) any formal connection with Physics.
+
 Are constructions are very similar to those developed in the context

### 2012-03-11 rational.tex
sat

@@
 represented by a recursive type and for which an isomorphic clone operation
-can be constructed. Before we start, let us recall two constructions:
-
-\begin{enumerate}
-\item Recall from previous work that for the types {{bool}} and, by
-  extension, any tuple of {{bool}}, an isomorphic clone operation can
-  be constructed (See Lemma 7.2 \cite{infeffects}).
-
-\item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
-  {{if_{not} }}, which when generalized to two bits was
-  {{toffoli}}. This construction maybe generalised to {{n}} bits as
-  {{if^n_{not} }}.
-\end{enumerate}
+can be constructed. Before we start, let us recall a constructions from 
+prior work \cite{infeffects}.
+
+\emph{Conditionals.}
+Given any combinator {{c : b <-> b}} we can construct a combinator called
+{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
+one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
+the combinator {{c}} is used to transform the value of type~{{b}}. If the
+boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
+write down the combinator for {{if_c}} in terms of {{c}} as 
+{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
+
+% Let us look at the combinator pictorially as if it were a circuit and values
+% are like particles that flow through this circuit.
+
+The diagram below shows the input value of type {{(1+1)*b}} processed
+by the distribute operator {{dist}}, which converts it into a value of
+type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
+case when the boolean is {{true}} (i.e. the value was {{left ()}}),
+the combinator~{{c}} is applied to the value of type~{{b}}. The right
+branch which corresponds to the boolean being {{false}} passes along
+the value of type {{b}} unchanged.
+
+\begin{center}
+\scalebox{1.0}{
+%%subcode-line{pdfimage}[diagrams/if_c.pdf]
+\includegraphics{diagrams/thesis/cnot.pdf}
+}
+\end{center}
+
+% We will be seeing many more such wiring diagrams in this paper and it is
+% useful to note some conventions about them. Wires indicate a
+% value that can exist in the program. Each wire, whenever possible, is
+% annotated with its type and sometimes additional information to help clarify
+% its role. When multiple wires run in parallel, it means that those values
+% exist in the system at the same time, indicating pair types. When there is a
+% disjunction, we put a {{+}} between the wires. 
+% Combinators for distribution {{dist}} and factoring {{factor}}
+% are represented as triangles with their operator symbols in them. Other
+% triangles may be used and, in each case, types or labels will be used to
+% clarify their roles. Finally, we don't draw boxes for combinators such as
+% {{id}}, commutativity, and associativity, but instead just shuffle the wires
+% as appropriate.
+
+\emph{Boolean Operation based on {{if_c}}.}
+The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
+negates its second argument if the first argument is {{true}}. This
+gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
+equivalent construction that is useful is {{else_{not} }} where we
+negate the second argument only is the first is {{false}}.
+
+Similarly, we can iterate the construction of {{if_c}} to check
+several bits. The gate {{if_{cnot} }}, which we may also write as
+{{if^2_{not} }}, checks two booleans and negates the result wire only
+if they are both {{true}}. The gate {{if^2_{not} }} is well known as
+the universal reversible primitive gate called the `Toffoli gate'
+\cite{Toffoli:1980}.  We can generalize this construction to
+     {{if^n_{not} }} which checks {{n}} bits and negates the result
+     wire only if they are all {{true}}.
+
+\emph{Cloning Bits.} The gate {{else_{not} }} has the interesting
+property that as long as the second argument is fixed to be {{true}},
+output of the second wire is always the same as teh value of the first
+wire. In other words {{else_{not} }} can be thought of as cloning the
+value of the first wire. A circuit of {{n}} parallel {{else_{not} }}
+gates can hence clone {{n}} bits. They also consume {{n}} {{true}}
+inputs in the process. Let us call this construction 
+{{clone^n_{bool} }}.
+
+%% \begin{enumerate}
+%% \item Recall from previous work that for the types {{bool}} and, by
+%%   extension, any tuple of {{bool}}, an isomorphic clone operation can
+%%   be constructed (See Lemma 7.2 \cite{infeffects}).
+
+%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% \item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
+%%   {{if_{not} }}, which when generalized to two bits was
+%%   {{toffoli}}. This construction maybe generalised to {{n}} bits as
+%%   {{if^n_{not} }}.
+%% \end{enumerate}
 
@@
 appropriate initial value by checking the heap and negating a second
-control wire, if the values do not match, i.e., the dotted part in
-the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
-`heap control wire'.
+control wire, if the values do not match, i.e., the dotted part in the
+diagram below and is {{if^h_{not} }} followed by a {{not}} operator on
+the control wire.
 
@@
 wires as {{sat_f}}. The SAT-solver is is completed by tracing the
-{{sat_f}} and cloning the `inputs'.
+{{sat_f}} and cloning the `inputs' using {{clone^n_{bool} }}.
 

### 2012-03-11 rational.tex
sat

@@
 represented by a recursive type and for which an isomorphic clone operation
-can be constructed. Before we start, let us recall a constructions from 
-prior work \cite{infeffects}.
-
-\emph{Conditionals.}
-Given any combinator {{c : b <-> b}} we can construct a combinator called
-{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
-one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
-the combinator {{c}} is used to transform the value of type~{{b}}. If the
-boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
-write down the combinator for {{if_c}} in terms of {{c}} as 
-{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
+can be constructed.
+
+\subsection{Constructions with Booleans}
+
+Before we start, let us recall a construction of conditionals from
+prior work \cite{infeffects}.  Given any combinator {{c : b <-> b}} we
+can construct a combinator called {{if_c : bool*b <->bool*b}} in terms
+of {{c}}, where {{if_c}} behaves like a one-armed
+$\mathit{if}$-expression. If the supplied boolean is {{true}} then the
+combinator {{c}} is used to transform the value of type~{{b}}. If the
+boolean is {{false}}, then the value of type {{b}} remains
+unchanged. We can write down the combinator for {{if_c}} in terms of
+{{c}} as {{ dist (;) ((id (*) c) (+) id) (;) factor }}.
 
@@
 
-%% \begin{enumerate}
-%% \item Recall from previous work that for the types {{bool}} and, by
-%%   extension, any tuple of {{bool}}, an isomorphic clone operation can
-%%   be constructed (See Lemma 7.2 \cite{infeffects}).
-
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% \item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
-%%   {{if_{not} }}, which when generalized to two bits was
-%%   {{toffoli}}. This construction maybe generalised to {{n}} bits as
-%%   {{if^n_{not} }}.
-%% \end{enumerate}
-
-%% That said, we will focus on the limited case of boolean satisfiability
-%% here.
-
-\emph{A Total Annihilation Circuit.}  Consider the circuit below that
-has no observable output, i.e., for all inputs it is
+DIAGRAM clone
+
+DIAGRAM clone n bits
+
+\subsection{Construction of the Solver}
+The key insight underlying the construction comes from the fact that we
+can build \emph{annihilation circuits}.  Consider the circuit below
+that has no observable output, i.e., for all inputs it is
 annihilating. This happens because it constructs a boolean and its
@@
 
-\emph{Controlled Annihilation.}
-We leverage the basic idea of the above circuit, to collapse all
-values that are not solutions to the specified isomorphic constraint
-{{f}}. The contraint {{f}} checks the inputs to see if they match a
-constraint -- in the case of the SAT solver, {{f}} would be the
-circuit expressing the boolean expression to be satisfied.  While the
-output of {{f}} may contain many components, its must contain a
-boolean indicating if the input satisfy {{f}}. The operation {{f}}
-acts on some inputs and its adjoint, {{f^{dagger} }}, reconstructs the
-inputs from the outputs.  The `trick' is the following: the special
-output wire is used to `control' the negation of a boolean `control
-wire'.  If the inputs do not satisfy {{f}}, the control wire is
-negated.
+\emph{Controlled Annihilation.}  We leverage the basic idea of the
+above circuit, to collapse all values that are not solutions to the
+specified isomorphic constraint {{f}}.  The contraint {{f}} checks the
+inputs to see if they match a constraint -- in the case of the SAT
+solver, {{f}} would be the circuit expressing the boolean expression
+to be satisfied.  While the output of {{f}} may contain many
+components, its must contain a boolean indicating if the input satisfy
+{{f}}. The operation {{f}} acts on some inputs and its adjoint,
+{{f^{dagger} }}, reconstructs the inputs from the outputs.  The
+`trick' is the following: the special output wire is used to `control'
+the negation of a boolean `control wire'.  If the inputs do not
+satisfy {{f}}, the control wire is negated.
 

### 2012-03-11 rational.tex
_

@@
 Sec. \ref{sec:hof} for a discussion of the additive functions
-vs. multiplicative functions.
+vs. multiplicative functions. \roshan{Where is the HOF discussion?}
 
@@
 
-While the constructions in Sec \ref{sec:monoidal-constructions} apply
-to both the {{(0,+)}} and {{(1,*)}} monoids, each construction
-presented here applies to a specific monoid.
+While the constructions in Sec \ref{sec:neg} apply to both the
+{{(0,+)}} and {{(1,*)}} monoids, each construction presented here
+applies to a specific monoid.
+
+\roshan{Each of these isomorphisms has baffled us -- either we
+  believed they were not constructible. Or we have struggled to
+  comprehend what they mean. Hence it is worth saying a sentence or
+  two about each beyond how they are constructed.}
 
@@
 \subsection{Construction of the Solver}
-The key insight underlying the construction comes from the fact that we
-can build \emph{annihilation circuits}.  Consider the circuit below
+The key insight underlying the construction comes from the fact that
+we can build \emph{annihilation circuits}.  Consider the circuit below
 that has no observable output, i.e., for all inputs it is

### 2012-03-11 rational.tex

--- 2012-03-11 rational.tex
@@
 \end{itemize}
-\roshan{Why was this deleted?}
-The example illustrates that the empty type and negative types have a
-computational interpretation related to continuations: negative types denote
-values that backtrack to satisfy dependencies, or in other words act as debts
-that are satisfied by the backward flow of information.
+%% The example illustrates that the empty type and negative types have a
+%% computational interpretation related to continuations: negative types denote
+%% values that backtrack to satisfy dependencies, or in other words act as debts
+%% that are satisfied by the backward flow of information.
 
@@
 of as the absence of an unknown quantity.
+% Unlike with negative types, wherein only one value existed at a time and the
+% computation backtracked, here we have three values that \emph{exist at the
+%   same time}. In other words, the computation with fractions is realized with
+% a schedule in which every value independently and concurrently proceeds
+% through its subcomputation.
 At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}}, thereby
@@
 payment of our debt, as line (4) shows, we have use of our credit card again.
-
-\roshan{Also, why was this deleted?}
-Unlike with negative types, wherein only one value existed at a time and the
-computation backtracked, here we have three values that \emph{exist at the
-  same time}. In other words, the computation with fractions is realized with
-a schedule in which every value independently and concurrently proceeds
-through its subcomputation.
 
@@
 \noindent 
-\textit{Example.}  We use the type {{bool}} as a shorthand to denote
-the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
-to be {{false}}. The following combinator is represented by the given
+\textit{Example.} The following combinator is represented by the given
 diagram:
 
-{{c : b * bool <-> b + b}}
+{{c : b * (1+1) <-> b + b}}
 
@@
 After presenting the extension at the syntactic level, we discuss the
-%% <<<<<<< local
-%% categorical semantics informally via the graphical language. The
-%% operational semantics is presented in Sec.~\ref{sec:rat}. As will be
-%% detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
-%% compact closed structure provides several properties of interest: (i)
-%% morphisms or wires are allowed to run from right to left, (ii) the
-%% structure admits a trace that, depending on the situation, can be used
-%% to implement various notions of loops and
-%% =======
 categorical semantics informally via the graphical language. The operational
@@
 and
-%% >>>>>>> other
 recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
-(iii) the structure includes an isomorphism showing that the dual
-operator is an involution, and (iv) the structure is equipped with
-objects representing higher-order functions and a morphism
-\textit{eval} that applies these functional objects. Interestingly
-each monoidal structure provides the same ingredients, resulting in
-two dualities, two traces, two involutions, and two notions of
-higher-order functions.
+(iii) the structure includes an isomorphism showing that the dual operator is
+an involution, and (iv) the structure is equipped with objects representing
+higher-order functions and a morphism \textit{eval} that applies these
+functional objects. Interestingly each monoidal structure provides the same
+ingredients, resulting in two dualities, two traces, two involutions, and two
+notions of higher-order functions.
 
@@
 
-%% <<<<<<< local
-%% Hence, the intuitive way to read the diagrams is that they reverse the
-%% flow of information -- i.e. they capture a notion of \emph{negative
-%%   information flow}. In normal {{langRev}} circuits, information flows
-%% from left-to-right. The operator {{eps+}} causes information to flow
-%% from right-to-left.  The interpretation of {{eta+}} is the opposite:
-%% it turns a right-to-left flow of information in a circuit into a
-%% left-to-right flow.  The diagram below is an example of using {{eta+}}
-%% and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
-%% from the introduction. (Here the {{0}} wires have been shown for
-%% completeness.)
-%% =======
 \paragraph*{Negative Information Flow.} 
@@
 example, the diagram representing the isomorphism from the introduction is:
-%% >>>>>>> other
 \begin{center}
@@
 % \paragraph*{Time Traveling.}
-%% <<<<<<< local
-%% =======
 % A time traveling intuition is also applicable here: if view the
@@
 % (whose construction is shown in Section \ref{sec:neg-constructions}).
-%% >>>>>>> other
 
 %\subsection{Compact Closure with Additive Inverses}
-%% <<<<<<< local
-%% \emph{Coherence for Compact Closure with Additive Duality.}  The
-%% appropriate coherence condition for compact closed categories,
-%% expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-%% may be visualized (modulo {{0}} wires, swapping etc.) as explained by
-%% Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
-%% operational equivalence:
-
-%% =======
 \paragraph*{Coherence}
@@
 follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
-%% >>>>>>> other
 \begin{center}
@@
 
-%% <<<<<<< local
-%% \emph{Non-termination.}  Compact closed categories contain {{trace}}
-%% operations and this implies that {{eta+}}/{{eps+}} allow the
-%% construction of a {{trace+}} operator (shown in Section
-%% \ref{sec:monoidal-constructions}).  As shown in previous work
-%% \cite{infeffects}, non-terminating computation can be constructed from
-%%      {{trace+}} and recursive types.
-%% =======
 \begin{multicols}{2}
@@
 \end{multicols}
-%% >>>>>>> other
 
@@
 
-\begin{multicols}{2}
+Given {{c : b1 <-> b2}}
+
 \begin{center}
@@
 \end{center}
-
-%subcode{proof} include main
-%@ c : b1 <-> b2
-%@@ neg c : (-b1) <-> (-b2)
-\end{multicols}
 
@@
 
-%% <<<<<<< local
-%% %% The section follows the outline of the previous one, first presenting
-%% %% the new syntax, and discussing the graphical language and the
-%% %% properites implied by the underlying categorical structure.
-
-%% =======
-%% >>>>>>> other
 Multiplicative duality is added to {{langRev}} by adding types
@@
 
-For convenience we sometime use the notation {{b1/b2}} to indicate the
-pair {{b1 * (1/b2)}}.  As with additive duals, we represent {{eta*}}
-and {{eps*}} using U-shaped connectors. In later diagrams, the wires
-for {{1}} will be dropped as they can freely introduced or eliminated
-using {{uniti}} and {{unite}}.
+As with additive duals, we represent {{eta*}} and {{eps*}} using
+U-shaped connectors. In later diagrams, the wires for {{1}} will be
+dropped as they can freely introduced or eliminated using {{uniti}}
+and {{unite}}.
 
@@
 
-%% <<<<<<< local
-%% The usual interpretation of {{b1*b2}} we have both a value of type
-%% {{b1}} and a value of type {{b2}}. This interpretation is maintained
-%% in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
-%% to be viewed as a fission point for a value of type {{b}} and its
-%% multiplicative inverse {{1/b}}. This naturally raises the question --
-%% which value of type {{b}} is produced? The isomorphism {{eta*}} does
-%% not produce any specific value of type {{b}}, instead it produces a
-%% placeholder for an unspecified value and dually, {{1/b}} is inhabited
-%% by the placeholder that indicates the absence of the unspecified
-%% value.
-%% =======
 In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
@@
 value of type {{1/b}} is always the dual of the chosen one.
-%% >>>>>>> other
 
@@
 
-%% <<<<<<< local
-%% =======
-%% >>>>>>> other
-
-%% <<<<<<< local
-In terms of \emph{logic programming}, {{eta* : 1 <-> (1/b)*b}} acts as
-the site of creation of a new typed logic variable and constructs two
-values using it: a value of type {{b}} and its dual of type
-{{1/b}}. These two values are free to move independently through the
-computation. Operations on one value affect what operations are possible
-on the other since they are both related by an underlying logic
-variable. The operator {{eps*}} acts as a unification site for values of
-type {{b}} and {{1/b}} flowing along both branches. If the values are
-not unifiable duals, then the specific program sequence fails. Like
-with logic programming, one can has many worlds where some might fail,
-while others make progress. Sec. \ref{sec:rat} has more details.  If
-the values are exactly duals, then unification succeeds and {{eta*}}
-returns {{1}}.
-
-% \paragraph*{Entanglement and Anti-Particles.}
-
-%% An analogy inspired by quantum mechanics is also applicable here.  The
-%% operator {{eta*}} is a site of fission that creates \emph{an entangled
-%%   particle and anti-particle}. These particles can be thought to be in
-%% a \emph{superposition} of states determined by their type. Both flow
-%% in the same direction in time.  In each possible world that the
-%% particle exists, it takes on a specific value inhibiting its type and
-%% correspondingly its entangled anti-particle takes the dual of the
-%% specific value. Since the particles are entangled, actions on one
-%% (such as transformation by application of an isomorphism) affects the
-%% other in much the same sense as \emph{action at a distance}. The
-%% `{{a-o*b}}' functions mentioned in the introduction, are first-class
-%% values and they correspond to these entagled pairs. Finally {{eps*}}
-%% operations are sites where corresponding particle and anti-particles
-%% annihilate each other. While appealing as an analogy, this description
-%% does not imply (or preclude) any formal connection with Physics.
-
-%% We can ask the function to be applied to a value by
-%% unifying the anti-particle with a particle in a well-known state and
-%% examining is pair. We shall use this point of view in the development
-%% our SAT solver (see Sec. \ref{sec:sat-solver}).
-
-%%  This also leads to the exciting point of view that
-%% entangled such particle/anti-particle pairs, which are first class
-%% values, act as functions in a physical first-class sense.
-
-%% Further the fission site does not create a particle in any specific
-%% state, but in a superposition of all possible states as dictated by
-%% its type.
-
-% Since such pairs are first-class entities, they correspond to
-
-%% \emph{Coherence for Compact Closure with Multiplicative Duality.}  As
-%% before, the triangle axiom which is the coherence condition on compact
-%% closed categories is applicable and is expressed below in terms of
-%% {{eta*}}/{{eps*}} and the multiplicative monoid.
-%% =======
-%% >>>>>>> other
-
 \paragraph*{Coherence.}
@@
 
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-
-\emph{Annihilation.}  
-With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
-which gives us the ability to find the fixpoint of a constraint.  If
-the constraint has no satisfying values, there is no fixpoint possible
--- we use the term \emph{annihilation} to describe the corresponding
-undefined state of the program.
-
-%% <<<<<<< local
-Annihilation and non-termination describe undefined program
-execution. While non-termination characterizes an iteration that fails
-to terminate, annihilation characterizes a constraint that has no
-fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
-annihilating only the program states that do not match the boolean
-satisfiability constraint.
-%% =======
 % \paragraph*{Unification.}
@@
 % functions.
-%% >>>>>>> other
 
@@
 
-%% <<<<<<< local
-\emph{Constructions.} The constructions for functions, involution, and trace are identical,
-replacing {{+}} by {{*}}, \textit{mutatis mutandis}.
-
-%% \begin{enumerate}
-%% \item 
-%% \emph{Trace}
-%% =======
-%% % \item
-%% % Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-%% % are indeed first-class values. However these functions can only be
-%% % linearly used, i.e., they represent a constraint that can be satisfied
-%% % at most once.
-%% % %%  in one `computational world'. It may be possible that to
-%% % %% repeatedly invoke multiplicative functions with an appropriate
-%% % %% recursive type whose unfoldings allow for repeated unification.
-%% % \end{itemize}
-%% >>>>>>> other
+% \item
+% Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+% are indeed first-class values. However these functions can only be
+% linearly used, i.e., they represent a constraint that can be satisfied
+% at most once.
+% %%  in one `computational world'. It may be possible that to
+% %% repeatedly invoke multiplicative functions with an appropriate
+% %% recursive type whose unfoldings allow for repeated unification.
+% \end{itemize}
 
@@
 
-%% \item 
-%% \emph{ {{eps_{fst} }} }
-
-%% {{(
+\item 
+\emph{ {{eps_{fst} }} }
+
+{{(-b1)*b2 + b1*b2 <-> 0}}
+
+\begin{center}
+  \includegraphics{diagrams/eps_fst.pdf}
+\end{center}
 
@@
 
-To build these isomorphisms, we first build an intermeidate
-construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
-\begin{center}
-\scalebox{0.8}{
-  \includegraphics{diagrams/eps_fst.pdf}
-}
-\end{center}
-
-The isomorphism {{(-b1) * b2 <-> -(b1 * b2)}} can be constructed in
-terms of {{eps_{fst} }} as shown below.
-
-\begin{center}
-\scalebox{0.9}{
+\begin{center}
   \includegraphics{diagrams/mult_neg.pdf}
-}
-\end{center}
-
-The second isomorphism can be built in the same way by merely swapping
-the arguments.
-
-\item
-\emph{Multiplying Negatives}
+\end{center}
+
+The following isomorphism can be constructed similarly
 
 {{b1 * b2 <-> (-b1)*(-b2)}}
-
-This isomorphism can be built using \emph{involution} constructed in
-Sec \ref{sec:neg} and the isomorphisms above, and correspond to the
-algebraic manipulation:
-
-%subcode{opsem} include main
-%! columnStyle = rc
-%   & b1 * b2
-% = & -(-(b1*b2)) 
-% = & -((-b1)*b2)
-% = & (-b1)*(-b2)
 
@@
 represented by a recursive type and for which an isomorphic clone operation
-can be constructed. Before we start, let us recall a constructions from 
-prior work \cite{infeffects}.
-
-\emph{Conditionals.}
-Given any combinator {{c : b <-> b}} we can construct a combinator called
-{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
-one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
-the combinator {{c}} is used to transform the value of type~{{b}}. If the
-boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
-write down the combinator for {{if_c}} in terms of {{c}} as 
-{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
-
-% Let us look at the combinator pictorially as if it were a circuit and values
-% are like particles that flow through this circuit.
-
-The diagram below shows the input value of type {{(1+1)*b}} processed
-by the distribute operator {{dist}}, which converts it into a value of
-type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
-case when the boolean is {{true}} (i.e. the value was {{left ()}}),
-the combinator~{{c}} is applied to the value of type~{{b}}. The right
-branch which corresponds to the boolean being {{false}} passes along
-the value of type {{b}} unchanged.
-
-\begin{center}
-\scalebox{1.0}{
-%%subcode-line{pdfimage}[diagrams/if_c.pdf]
-\includegraphics{diagrams/thesis/cnot.pdf}
-}
-\end{center}
-
-% We will be seeing many more such wiring diagrams in this paper and it is
-% useful to note some conventions about them. Wires indicate a
-% value that can exist in the program. Each wire, whenever possible, is
-% annotated with its type and sometimes additional information to help clarify
-% its role. When multiple wires run in parallel, it means that those values
-% exist in the system at the same time, indicating pair types. When there is a
-% disjunction, we put a {{+}} between the wires. 
-% Combinators for distribution {{dist}} and factoring {{factor}}
-% are represented as triangles with their operator symbols in them. Other
-% triangles may be used and, in each case, types or labels will be used to
-% clarify their roles. Finally, we don't draw boxes for combinators such as
-% {{id}}, commutativity, and associativity, but instead just shuffle the wires
-% as appropriate.
-
-\emph{Boolean Operation based on {{if_c}}.}
-The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
-negates its second argument if the first argument is {{true}}. This
-gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
-equivalent construction that is useful is {{else_{not} }} where we
-negate the second argument only is the first is {{false}}.
-
-Similarly, we can iterate the construction of {{if_c}} to check
-several bits. The gate {{if_{cnot} }}, which we may also write as
-{{if^2_{not} }}, checks two booleans and negates the result wire only
-if they are both {{true}}. The gate {{if^2_{not} }} is well known as
-the universal reversible primitive gate called the `Toffoli gate'
-\cite{Toffoli:1980}.  We can generalize this construction to
-     {{if^n_{not} }} which checks {{n}} bits and negates the result
-     wire only if they are all {{true}}.
-
-\emph{Cloning Bits.} The gate {{else_{not} }} has the interesting
-property that as long as the second argument is fixed to be {{true}},
-output of the second wire is always the same as teh value of the first
-wire. In other words {{else_{not} }} can be thought of as cloning the
-value of the first wire. A circuit of {{n}} parallel {{else_{not} }}
-gates can hence clone {{n}} bits. They also consume {{n}} {{true}}
-inputs in the process. Let us call this construction 
-{{clone^n_{bool} }}.
-
-%% \begin{enumerate}
-%% \item Recall from previous work that for the types {{bool}} and, by
-%%   extension, any tuple of {{bool}}, an isomorphic clone operation can
-%%   be constructed (See Lemma 7.2 \cite{infeffects}).
-
-%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% \item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
-%%   {{if_{not} }}, which when generalized to two bits was
-%%   {{toffoli}}. This construction maybe generalised to {{n}} bits as
-%%   {{if^n_{not} }}.
-%% \end{enumerate}
+can be constructed. Before we start, let us recall two constructions:
+
+\begin{enumerate}
+\item Recall from previous work that for the types {{bool}} and, by
+  extension, any tuple of {{bool}}, an isomorphic clone operation can
+  be constructed (See Lemma 7.2 \cite{infeffects}).
+
+\item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
+  {{if_{not} }}, which when generalized to two bits was
+  {{toffoli}}. This construction maybe generalised to {{n}} bits as
+  {{if^n_{not} }}.
+\end{enumerate}
 
@@
 appropriate initial value by checking the heap and negating a second
-control wire, if the values do not match, i.e., the dotted part in the
-diagram below and is {{if^h_{not} }} followed by a {{not}} operator on
-the control wire.
+control wire, if the values do not match, i.e., the dotted part in
+the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
+`heap control wire'.
 
@@
 wires as {{sat_f}}. The SAT-solver is is completed by tracing the
-{{sat_f}} and cloning the `inputs' using {{clone^n_{bool} }}.
+{{sat_f}} and cloning the `inputs'.
 
@@
 \paragraph*{QC} 
-A \emph{time traveling} intuition is also applicable here.  In a
-normal cicruit, as computational steps are taken values flow from left
-to right, i.e. computational time progresses from left to right.  The
-action of {{eps+}} causes values, aka information particles, to flow
-from right to left i.e. backwards in computational time.  The
-interesting thing about this interpretation is that particles that
-travel backwards in time get to see and interact with the history of
-particles that they coexist with. This gives us an intuitive
-interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
-Sec. \ref{sec:specific-constructions}) where it would otherwise seem
-that {{-b1}} and {{b2}} move in opposite directions. The backward flow
-of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
-equivalent to both particles moving backward in time.
-
-An analogy inspired by quantum mechanics is also applicable here.  The
-operator {{eta*}} is a site of fission that creates \emph{an entangled
-  particle and anti-particle}. These particles can be thought to be in
-a \emph{superposition} of states determined by their type. Both flow
-in the same direction in time.  In each possible world that the
-particle exists, it takes on a specific value inhibiting its type and
-correspondingly its entangled anti-particle takes the dual of the
-specific value. Since the particles are entangled, actions on one
-(such as transformation by application of an isomorphism) affects the
-other in much the same sense as \emph{action at a distance}. The
-`{{a-o*b}}' functions mentioned in the introduction, are first-class
-values and they correspond to these entagled pairs. Finally {{eps*}}
-operations are sites where corresponding particle and anti-particles
-annihilate each other. While appealing as an analogy, this description
-does not imply (or preclude) any formal connection with Physics.
-
 Are constructions are very similar to those developed in the context

### 2012-03-11 rational.tex
conclusion

--- 2012-03-11 rational.tex
@@
 identity that holds for the rational numbers corresponds to a type
-isomorphism in our model.  In the process, we have examined the two
-function spaces that these give rise to an have seen non-trivial
-constructions such as a novel SAT-solver that relies on a
-multiplicative trace.
+isomorphism with a computational interpretation in our model. We have
+examined the two function spaces that arise in this model and
+developed non-trivial constructions such as a SAT-solver that relies
+on a multiplicative trace.
 
@@
 and intuinionistic mathematics -- contain non-trivial assumptions
-about our world-view.
-
-The problem of duality in computation and logic, has many facets of
-which `continuation' and `higher order functions' are only one.
-
-\paragraph*{GoI}
-The other popular model of computing with values traveling along wire
-is GoI -- specifically Mackie's work. How are we related to this
-model? Preliminary investogationb suggests that many of the no-cloning
-based constructions can be simulated in {{langRevEE}} as well. If
-there is a concrete connectoion, then by Curry-Howard the logical
-interpretation of {{langRevEE}} should a logic much like LL with a
-strong notion of resource preservation, but also with a strong
+about our world-view.  The problem of duality in computation and
+logic, has many facets of which `continuation' and `higher order
+functions' are only one.  We have contributed an elegant and natural
+notion of duality that arises from the monoidal structure of sums and
+pairs. This notion of duality has a crisp semantics, clear
+computational interpretations and an information theoretic basis.
+Before we conclude, we briefly mention some closely related ideas that
+warrant further study.
+
+%% \paragraph*{Int Construction} How are we preserving two compact
+%% closed structures -- what does this imply about the Int
+%% construction.
+
+\paragraph*{Geometry of Interaction.}
+GoI was developed by Girard \cite{girard1989geometry}as basis for the
+proof theory of Linear Logic.  GoI however has been developed into a
+reversible model of computing
+\cite{Mackie2011,DBLP:conf/popl/Mackie95,Abramsky:1994:NFG:184662.184664}
+in which values traveling along wires, much like {{langRevEE}}.
+Preliminary investigations suggest that many GoI machine constructions
+can be simulated in {{langRevEE}} by treating Mackie's bi-directional
+wires as pairs of wires in {{langRevEE}} and replacing the machine's
+global state with a typed value on the wire that captures the
+appropriate state.  However a concrete connection between GoI machines
+and GoI in general requires further work. This connection is exciting
+because when viewed through a Curry-Howard lens it suggests that the
+logical interpretation of {{langRevEE}} should a logic much like LL
+with a strong notion of resource preservation, but also with a strong
 computational basis.
 
-\paragraph*{Int Construction}
-How are we preserving two compact closed structures -- what does this
-imply about the Int construction.
-
-\paragraph*{QC} 
+\paragraph*{Computing in the Field of Algebraic Numbers.}
+\label{sec:algebraic-field}
+The field of algebraic numbers consist of the rationals extended with
+solutions to polynomials. In a limited sense {{langRevEE}} extended
+with recursive types already include polynomials such as the
+definition of binary tree {{@@x.(1+x*x)}} implies the equality
+{{x=1+x*x}} and hence {{x^2-x+1=0}}.
+
+The famous example of isomorphism on algebraics is the embedding of
+seven binary trees in one. In Figure 1 of Fiore's paper
+\cite{Fiore:2004}, he demonstrates this isomorphism by the
+corrsponding algebraic manipulation of the types.  The algebraic
+manipulation can be viewed as a {{langRevEE}} combinator and it indeed
+does constructively witness the isomorphism, i.e., you can run it and
+view one data structure encoded by the other. In this context,
+consider the following algebraically valid proof of the same
+isomorphism:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+Fiore poses the question of why such algebraic manipulation would make
+sense type theoretically, even though some of the intermediate steps
+make no sense.  Stated in terms of {{langRevEE}} computations, this
+manipulation has no meaningful computational content -- it is always
+non-terminating. It suggests that some algebraic manipulations are
+somehow more `constructive' than others or dually that {{langRevEE}}
+with recursive types somehow lacks the ability to express all
+computations that are otherwise algebraically meaningful.
+
+\begin{itemize}
+\item Maybe the problem is that there is cancellation in the second
+  case and cancellation with recursive types is problematic. However
+  not all cancellation is. So what gives?
+
+\item One idea is that irrationals correspond to an infinite amount of
+  computational work. If what we are cancelling is not the same
+  infinity, then things go wrong.
+
+\item Chaitin has already shown that reals have inifite amount of
+  information which is paradoxical and problematic. This is a
+  different issue from that of the rational being infinite
+  computations.
+
+\item Finally, not all meaningful types are meaningful polynomials --
+  ex. nat. So this is not the whole story.
+
+\end{itemize}
+
+\paragraph*{Quantum Computing} 
 A \emph{time traveling} intuition is also applicable here.  In a
@@
 traveling particles and anti-particles more concrete in some way?
-
-%% Parallel computing with constraints and backtracking -- breaking
-%% dependancies,
-
-\paragraph*{Computing in the Field of Algebraic Numbers.}
-\label{sec:algebraic-field}
-The field of algebraic numbers consist of the rationals extended with
-solutions to polynomials. In a limited sense {{langRevEE}} extended
-with recursive types already include polynomials such as the
-definition of binary tree {{@@x.(1+x*x)}} implies the equality
-{{x=1+x*x}} and hence {{x^2-x+1=0}}.
-
-%% Partial isomorphisms algebraics
-%% do have computational
-
-%% While this equation does indeed have an irrational imaginary root,
-%% as we will see -- {{langRevEE}} does not have the means of
-%% expressing meaningful computation that uses the root.
-
-The famous example of isomorphism on algebraics is the embedding of
-seven binary trees in one. In Figure 1 of Fiore's paper
-\cite{Fiore:2004}, he demonstrates this isomorphism by the
-corrsponding algebraic manipulation of the types.  The algebraic
-manipulation can be viewed as a {{langRevEE}} combinator and it indeed
-does constructively witness the isomorphism, i.e., you can run it and
-view one data structure encoded by the other. In this context,
-consider the following algebraically valid proof of the same
-isomorphism:
-
-%% Previously the most famous example of a similar nature is the puzzling
-%% isomorphism that one can establish between seven binary trees and one.
-%% A binary tree is defined by the datatype
-%% \[
-%% x = 1 + x * x 
-%% \]
-%% which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
-%% algebraic manipulation we can reason as follows:
-\[\begin{array}{rclclclcl}
-x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
-x^6 &=& 1 \\
-x^7 &=& x^6 x &=& x
-\end{array}\]
-Fiore poses the question of why such algebraic manipulation would make
-sense type theoretically, even though some of the intermediate steps
-make no sense.  Stated in terms of {{langRevEE}} computations, this
-manipulation has no meaningful computational content -- it is always
-non-terminating. It suggests that some algebraic manipulations are
-somehow more `constructive' than others or dually that {{langRevEE}}
-with recursive types somehow lacks the ability to express all
-computations that are otherwise algebraically meaningful.
-
-\begin{itemize}
-\item Maybe the problem is that there is cancellation in the second
-  case and cancellation with recursive types is problematic. However
-  not all cancellation is. So what gives?
-
-\item One idea is that irrationals correspond to an infinite amount of
-  computational work. If what we are cancelling is not the same
-  infinity, then things go wrong.
-
-\item Chaitin has already shown that reals have inifite amount of
-  information which is paradoxical and problematic. This is a
-  different issue from that of the rational being infinite
-  computations.
-
-\item Finally, not all meaningful types are meaningful polynomials --
-  ex. nat. So this is not the whole story.
-
-\end{itemize}
-
-%% -- WAVEFRONT START --
-
-%% But the interesting solution of trees is precisely that they involve
-%% irrational numbers. Our usual notation for irrationals such as
-%% {{sqrt(2) }} can actually be viewed as a thunk representing the
-%% infinite computation {{1.4142..}}.  Irrationals in this sense have a
-%% very different information theoretic interpretation from full real
-%% numbers.  With Chaitin's $\Omega$ numbers
-%% \cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
-%% real numbers that have an infinite amount of information -- ones for
-%% example are the solution to the question about the halting properties
-%% of every Turning machine.
-
-%% That introduces the natural numbers. No solution for this polynomial
-%% over the algebraic numbers. We must extend the numbers with $\omega$
-%% to get a solution. We reject this in this paper and prefer to stick to
-%% algebraic numbers. The advantage is that all the isomorphisms are
-%% valid numerically. With the above type we could subtract $b$ from both
-%% sides to show that $0=1$ which is nonsense. We can however have
-%% infinite types as long as they have algebraic solutions.
-
-%% In our previous work we introduce recursive types and trace
-%% operators. This is dangerous here because infinite loops allow us to
-%% prolong paying the debt for as long as we want. The algebraic
-%% $\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that
-%% extreme care is needed when combining negative and fractional types
-%% with the full $\lambda$-calculus (including recursion).
-
-%% -- WAVEFRONT STOP --
-
-\paragraph*{Summary}
-We have contributed an elegant and natural notion of duality that
-arises from the monoidal structure of sums and pairs. This notion of
-duality has a crisp semantics, clear computational interpretations and
-an information theoretic basis.
 
@@
 
-%% Jacques Carette, Michael Adams, Khaled
+%% Jacques Carette, Michael Adams, Khaled, Will Byrd, Yin Wang
 

### 2012-03-11 rational.tex

@@
 \roshan{Why was this deleted?}
+\amr{because it says nothing new}
 The example illustrates that the empty type and negative types have a
@@
 of as the absence of an unknown quantity.
+
+% Unlike with negative types, wherein only one value existed at a time and the
+% computation backtracked, here we have three values that \emph{exist at the
+%   same time}. In other words, the computation with fractions is realized with
+% a schedule in which every value independently and concurrently proceeds
+% through its subcomputation.
+
 At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}}, thereby
@@
 \roshan{Also, why was this deleted?}
+\amr{no real info is communicated; you're saying that pairs are different 
+than sums}
 Unlike with negative types, wherein only one value existed at a time and the
@@
 After presenting the extension at the syntactic level, we discuss the
-%% <<<<<<< local
 %% categorical semantics informally via the graphical language. The
@@
 %% to implement various notions of loops and
-%% =======
 categorical semantics informally via the graphical language. The operational
@@
 
-%% <<<<<<< local
 %% Hence, the intuitive way to read the diagrams is that they reverse the
@@
 %% completeness.)
-%% =======
 \paragraph*{Negative Information Flow.} 
@@
 
-% \paragraph*{Time Traveling.}
-%% <<<<<<< local
-%% =======
 % A time traveling intuition is also applicable here: if view the
@@
 % (whose construction is shown in Section \ref{sec:neg-constructions}).
-%% >>>>>>> other
 
 %\subsection{Compact Closure with Additive Inverses}
-%% <<<<<<< local
 %% \emph{Coherence for Compact Closure with Additive Duality.}  The
@@
 
-%% =======
 \paragraph*{Coherence}
@@
 
-%% <<<<<<< local
 %% \emph{Non-termination.}  Compact closed categories contain {{trace}}
@@
 %%      {{trace+}} and recursive types.
-%% =======
 \begin{multicols}{2}
@@
 Sec. \ref{sec:hof} for a discussion of the additive functions
-vs. multiplicative functions. \roshan{Where is the HOF discussion?}
+vs. multiplicative functions. 
+\roshan{Where is the HOF discussion?}
+\amr{deleted because it was weak and unfocused}
 
@@
 \begin{multicols}{2}
+Given {{c : b1 <-> b2}}
+
 \begin{center}
@@
 \end{multicols}
+\end{center}
 
@@
 
-%% <<<<<<< local
 %% %% The section follows the outline of the previous one, first presenting
@@
 
-%% =======
-%% >>>>>>> other
 Multiplicative duality is added to {{langRev}} by adding types
@@
 
-%% <<<<<<< local
 %% The usual interpretation of {{b1*b2}} we have both a value of type
@@
 %% value.
-%% =======
 In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
@@
 
-%% <<<<<<< local
-%% =======
-%% >>>>>>> other
-
-%% <<<<<<< local
+\paragraph*{Coherence.}
+The following coherence condition is satisfied as before.
+
 In terms of \emph{logic programming}, {{eta* : 1 <-> (1/b)*b}} acts as
@@
 returns {{1}}.
+% \paragraph*{Unification.}
+% A computational interpretation of this is that {{langRevEE}} acts like
+% a logic programming system and {{eta*}} acts as the site of creation
+% of a new typed logic variable of type {{b}} and its corresponding dual
+% of type {{1/b}}. These two values are free to move independently
+% through a circuit, but any operations on value affect what operations
+% are possible on the other since they are both related by an underlying
+% logic variable. Dually, {{eps*}} acts as a unification site for values
+% flowing along both branches. If the values are not exactly duals, then
+% the system fails to unify and hence the program is stuck. If the
+% values are exactly duals, then unification succeeds and {{eta*}}
+% returns {{1}}.
 
 % \paragraph*{Entanglement and Anti-Particles.}
-
+% A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
+% quantum mechanics. The {{eta*}} is a site of fission that creates an
+% information particle and its anti-particle. Both flow in the same
+% direction in time. The produce particles are entangles in that actions
+% on one (such as transformation by application of an isomorphism)
+% affects the other. Further the fission site does not create a particle
+% in any specific state, but in a superposition of all possible states
+% as dictated by its type. In each possible world that the particle
+% exists, it takes on a specific value inhibiting its type and
+% correspondingly its entangled anti-particle takes the dual of the
+% specific value.
+
+% This also leads to the exciting point of view that entangled
+% particle/anti-particle pairs act as functions in a physical
+% first-class sense. We can ask the function to be applied to a value by
+% unifying the anti-particle with a particle in a well-known state and
+% examining is pair. We shall use this point of view in the development
+% our SAT solver (see Sec. \ref{sec:sat-solver}).
+
+% \paragraph*{Entanglement and Anti-Particles.}
+% \paragraph*{Annihilation.}
+% As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
+% additive {{trace+}}, {{trace*}} gives us the ability to find the
+% fixpoint of the constraint (expressed with combinators) that is
+% traced. If the constraint has not satisfying values, there is not
+% fixpoint possible -- we use the term `annihilation' to describe the
+% corresponding undefined state of the program. Annihilation and a
+% non-termination both describe undefined program execution -- while
+% non-termination characterizes an iteration that fails to terminate,
+% annihilation characterizes a constraint that has no fixpoint. 
 %% An analogy inspired by quantum mechanics is also applicable here.  The
@@
 %% does not imply (or preclude) any formal connection with Physics.
+% \paragraph*{Zero Information.}
+% As with {{eta+}}, we argue that {{eta*}} constructs no new
+% information, instead it gives us the ability to explore as many worlds
+% as permitted by the type {{b}}. So situate ourselves in any specific
+% world, we have to supply additional information in the form of
+% unifying the particle or anti-particle with a particle representing
+% known information. Hence only by supplying information to a an
+% entangled pair can we read information from it -- and always exactly
+% as much information as we supplied.
 
@@
 %% {{eta*}}/{{eps*}} and the multiplicative monoid.
-%% =======
-%% >>>>>>> other
 
@@
 The following coherence condition is satisfied as before.
-
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
+% ADD MATH.
+
+% %%%%%%%%%%%%%%%%%%%%%%%%
+% \subsection{Conventional Higher-Order Functions}
+% \label{sec:hof}
 
@@
 undefined state of the program.
-
-%% <<<<<<< local
+% In a conventional language, higher-order functions capture scope and can be
+% represented as first-class values. In {{langRevEE}}, we have two notions of
+% functions both of which share similarities with conventional higher-order
+% functions.
+
 Annihilation and non-termination describe undefined program
@@
 satisfiability constraint.
-%% =======
 % \paragraph*{Unification.}
@@
 % returns {{1}}.
+% \begin{itemize}
+% \item 
+% Additive functions are not really values, but are possible paths of
+% execution. However, they can be encoded as values as follows: Consider
+% two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
+% {{bool}} can discriminate them and hence a value of type {{bool}} acts
+% as the address of the function and is a first class value
+% representing one of the {{b1 -o+ b2}} functions.  %% This is similar to
+% %% how first classes functions a'la \lcal are represented by their
+% %% addresses when compiled to conventional hardware.
 
@@
 % specific value.
+% \item
+% Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+% are indeed first-class values. However these functions can only be
+% linearly used, i.e., they represent a constraint that can be satisfied
+% at most once.
+% %%  in one `computational world'. It may be possible that to
+% %% repeatedly invoke multiplicative functions with an appropriate
+% %% recursive type whose unfoldings allow for repeated unification.
+% \end{itemize}
 
@@
 % our SAT solver (see Sec. \ref{sec:sat-solver}).
+% The exact encoding on HOF in {{langRevEE}} should shed light on
+% several problems, including the connections to continuations discussed
+% in Sec. \ref{sec:related}. They should also allow for a natural
+% encoding of HOFs in the translations of \lcal to {{langRevT}} and
+% indicate a monadic encapsulation of of information effects which are
+% currrently structured as an arrow.
 
@@
 
-%% <<<<<<< local
 \emph{Constructions.} The constructions for functions, involution, and trace are identical,
@@
 %% \emph{Trace}
-%% =======
 %% % \item
@@
 %% % \end{itemize}
-%% >>>>>>> other
 
@@
 can be constructed. 
+can be constructed. Before we start, let us recall two constructions:
 
@@
 that has no observable output, i.e., for all inputs it is
+\emph{A Total Annihilation Circuit.}  Consider the circuit below that
+has no observable output, i.e., for all inputs it is
 annihilating. This happens because it constructs a boolean and its
@@
 the control wire.
+control wire, if the values do not match, i.e., the dotted part in
+the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
+`heap control wire'.
 

### 2012-03-11 rational.tex
just fetching in case

@@
 
-\emph{Lifting a operation of positive types to negated types}
+\emph{Lifting an operation of positive types to negated types}
 
@@
 \end{multicols}
-\end{center}
 
@@
 
-The construction presented here is a novel SAT-solver that relies on
-{{trace*}} for its solution. It varies from previous classical SAT solvers
-\amr{must cite ???} in that there is no explicit search operation on the
-boolean space. It also varies from previous quantum SAT solvers \amr{must
-  cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
-which is the solution following which an isomorphic clone operation allows us
-to examine the result.  Though we focus on SAT here, this construction can be
-generalized to any constraint satisfaction problem whose search space can be
-represented by a recursive type and for which an isomorphic clone operation
-can be constructed. 
-can be constructed. Before we start, let us recall two constructions:
-
+% The construction presented here is a novel SAT-solver that relies on
+% {{trace*}} for its solution. It varies from previous classical SAT solvers
+% \amr{must cite ???} in that there is no explicit search operation on the
+% boolean space. It also varies from previous quantum SAT solvers \amr{must
+%   cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
+% which is the solution following which an isomorphic clone operation allows us
+% to examine the result.  Though we focus on SAT here, this construction can be
+% generalized to any constraint satisfaction problem whose search space can be
+% represented by a recursive type and for which an isomorphic clone operation
+% can be constructed. 
+% can be constructed. Before we start, let us recall two constructions:
+
+We illustrate the expressive power of first-class constraints represented by
+fractional types
+
+%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions with Booleans}

### 2012-03-11 rational.tex
sat

@@
 Every compact closed categoy admits a trace. In the context of {{langRev}}
-with {{eta+}}/{{eps+}}, we get the following construction:
+with {{eta+}}/{{eps+}}, we get the following definition. 
+Given {{f : a+c <-> b+c }}, define {{trace+ f : a <-> b}} as:
+
+{{ trace+ f = zeroi (;) (id (+) eta+) (;) (f (+) id) (;) (id (+) eps+) (;) zeroe }}
+
+For readability, we have omitted some the commutativity and
+associativity shuffling.
 
@@
 We illustrate the expressive power of first-class constraints represented by
-fractional types
+fractional types. To understand the intuition, recall the definition of
+{{trace}} for the multiplicative fragment of {{langRevEE}}. 
+Given {{f : a*c <-> b*c }}, we have {{trace* f : a <-> b}}:
+
+{{ trace* f = uniti (;) (id (*) eta*) (;) (f (*) id) (;) (id (*) eps*) (;) unite }}
+
+\noindent This circuit uses {{eta*}} to generate all possible {{c}}-values
+together with an associated {{(1/c)}}-constraint. It then applies {{f}} to the
+pair {{(a,c)}}. The function {{f}} must produce an output {{(b,c')}} for each
+such input. If the input {{c}} and the output {{c'}} are the same they can be
+annihilated by {{eps*}}; otherwise the execution gets stuck and this
+particular choice of {{c}} is pruned.
+
+This construction can be used to solve a large class of constraint
+satisfaction problems. We illustrate the main ideas with the implementation
+of a SAT-solver. Although the main idea is the use of {{trace*}}, the
+construction is involved so we proceed in small steps, reviewing some of the
+necessary constructions presented in our earlier paper~\cite{infeffects}.
 
 %%%%%%%%%%%%%%%%%%%
-\subsection{Constructions with Booleans}
-
-Before we start, let us recall a construction of conditionals from
-prior work \cite{infeffects}.  Given any combinator {{c : b <-> b}} we
-can construct a combinator called {{if_c : bool*b <->bool*b}} in terms
-of {{c}}, where {{if_c}} behaves like a one-armed
-$\mathit{if}$-expression. If the supplied boolean is {{true}} then the
-combinator {{c}} is used to transform the value of type~{{b}}. If the
-boolean is {{false}}, then the value of type {{b}} remains
-unchanged. We can write down the combinator for {{if_c}} in terms of
-{{c}} as {{ dist (;) ((id (*) c) (+) id) (;) factor }}.
-
-% Let us look at the combinator pictorially as if it were a circuit and values
-% are like particles that flow through this circuit.
-
-The diagram below shows the input value of type {{(1+1)*b}} processed
-by the distribute operator {{dist}}, which converts it into a value of
-type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
-case when the boolean is {{true}} (i.e. the value was {{left ()}}),
-the combinator~{{c}} is applied to the value of type~{{b}}. The right
-branch which corresponds to the boolean being {{false}} passes along
-the value of type {{b}} unchanged.
+\subsection{Booleans and Conditionals}
+
+Given any combinator {{c : b <-> b}} we can construct a combinator called
+{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
+one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
+the combinator {{c}} is used to transform the value of type~{{b}}. If the
+boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
+write down the combinator for {{if_c}} in terms of {{c}} as 
+{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
+
+\noindent The diagram below shows the input value of type {{(1+1)*b}}
+processed by the distribute operator {{dist}}, which converts it into a value
+of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
+case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
+combinator~{{c}} is applied to the value of type~{{b}}. The right branch
+which corresponds to the boolean being {{false}} passes along the value of
+type {{b}} unchanged.
 
@@
 
-\emph{Boolean Operation based on {{if_c}}.}
 The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
@@
 
-Similarly, we can iterate the construction of {{if_c}} to check
-several bits. The gate {{if_{cnot} }}, which we may also write as
-{{if^2_{not} }}, checks two booleans and negates the result wire only
-if they are both {{true}}. The gate {{if^2_{not} }} is well known as
-the universal reversible primitive gate called the `Toffoli gate'
-\cite{Toffoli:1980}.  We can generalize this construction to
-     {{if^n_{not} }} which checks {{n}} bits and negates the result
-     wire only if they are all {{true}}.
-
-\emph{Cloning Bits.} The gate {{else_{not} }} has the interesting
-property that as long as the second argument is fixed to be {{true}},
-output of the second wire is always the same as teh value of the first
-wire. In other words {{else_{not} }} can be thought of as cloning the
-value of the first wire. A circuit of {{n}} parallel {{else_{not} }}
-gates can hence clone {{n}} bits. They also consume {{n}} {{true}}
-inputs in the process. Let us call this construction 
+Similarly, we can iterate the construction of {{if_c}} to check several
+bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
+checks two booleans and negates the result wire only if they are both
+{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
+universal reversible gate. We can generalize this construction to
+{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
+they are all {{true}}.
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Cloning}
+
+Although cloning is generally not allowed in reversible languages, it is
+possible at the cost of having additional constant inputs. For example,
+consider the gate {{else_{not} }}. Generally, the gate gates maps
+{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
+on the cases in which the second input is {{true}}, we get that the gate maps
+{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
+i.e., the gate clones the first input. A circuit of {{n}} parallel
+{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
+{{true}} inputs in the process.  Let us call this construction
 {{clone^n_{bool} }}.
 
-DIAGRAM clone
-
-DIAGRAM clone n bits
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Construction of the Solver}
-The key insight underlying the construction comes from the fact that
-we can build \emph{annihilation circuits}.  Consider the circuit below
-that has no observable output, i.e., for all inputs it is
-\emph{A Total Annihilation Circuit.}  Consider the circuit below that
-has no observable output, i.e., for all inputs it is
-annihilating. This happens because it constructs a boolean and its
-dual, negates one of them and attempts to collapse them.
+
+The key insight underlying the construction comes from the fact that we can
+build \emph{annihilation circuits} such as the one below:
 
@@
 \end{center}
-
-\emph{Controlled Annihilation.}  We leverage the basic idea of the
-above circuit, to collapse all values that are not solutions to the
-specified isomorphic constraint {{f}}.  The contraint {{f}} checks the
-inputs to see if they match a constraint -- in the case of the SAT
-solver, {{f}} would be the circuit expressing the boolean expression
-to be satisfied.  While the output of {{f}} may contain many
-components, its must contain a boolean indicating if the input satisfy
-{{f}}. The operation {{f}} acts on some inputs and its adjoint,
-{{f^{dagger} }}, reconstructs the inputs from the outputs.  The
-`trick' is the following: the special output wire is used to `control'
-the negation of a boolean `control wire'.  If the inputs do not
-satisfy {{f}}, the control wire is negated.
+The circuit constructs a boolean {{b}} and its dual {{1/b}}, negates one of
+them and attempts to satisfy the constraint that they are equal which
+evidently fails.
+
+With a little work, we can modify this circuit to only annihilate values that
+fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
+detail, an instance of SAT is a function/circuit {{f}} that given some
+boolean inputs returns {{true}} or {{false}} which we interpret as whether
+the inputs satisfy the constraints imposed by the structure of {{f}}. Because
+we are in a reversible world, our instance of SAT must be expressed as an
+isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
+below. Assuming that {{f}} is expressed as an isomorphism, we have enough
+information to reconstruct the input from the output. This can be done by
+using the adjoint of {{f}}, denoted {{f^{dagger}}}. At this point we have,
+the top half of the construction below:
 
@@
 
-If this construction is traced using {{trace*}}, all the values that
-do not satisfy {{f}} get annihilated because the control wire acts
-like the closed-loop `not' in the previous construction.
-
-\emph{Controlling the heap.}  The final detail we have to be concerned
-with is the handling of the heap and garbage of {{f}}. Any boolean
-expression {{bool^n -> bool}} can be compiled into the isomorphism
-{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
-{{bool^g}} are the heap and garbage. Constructing such an isomorphism
-has been detailed before \cite{Toffoli:1980,infeffects}.  However the
-computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
-initialized appropriately. We can ensure that the heap has the
-appropriate initial value by checking the heap and negating a second
-control wire, if the values do not match, i.e., the dotted part in the
-diagram below and is {{if^h_{not} }} followed by a {{not}} operator on
-the control wire.
-control wire, if the values do not match, i.e., the dotted part in
-the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
-`heap control wire'.
+To summarize, the top half of the circuit is the identity function except
+that we have also managed to produce a boolean wire labeled
+\textsf{satisfied?} that tells us if the inputs satisfy the desired
+constraints. We can take this boolean value and use it to decide whether to
+negate the control wire or not. Thus, the circuit achieves the following
+goal: if the inputs do not satisfy {{f}}, the control wire is negated.  We
+can now use {{trace*}} to annihilate all these bad values because the control
+wire acts like the closed-loop {{not}} in the previous construction.
+
+%%%%%%%%%
+\subsection{Final Details}
+\label{sub:f}
+
+Any boolean expression {{f : bool^n -> bool}} can be compiled into the
+isomorphism {{iso : bool^h*bool^n<->bool^g*bool}} where the extra bits
+{{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
+such an isomorphism has been detailed
+before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
+applying {{iso}} to some special heap values and an input \textit{bs}
+produces some bits that can be ignored and the same output that {{f}} would
+have produced on {{bs}}. We can ensure that the heap has the appropriate
+initial values by checking the heap and negating a second control wire, if
+the values do not match, i.e., the dotted part in the diagram below.
 
@@
 
-\emph{Cloning the Solution.}.  Let us call the above construction
-which maps inputs, heap and control wires to input, heap and control
-wires as {{sat_f}}. The SAT-solver is is completed by tracing the
-{{sat_f}} and cloning the `inputs' using {{clone^n_{bool} }}.
+Let us call the above construction which maps inputs, heap, and control wires
+to input, heap, and control wires as {{sat_f}}. The SAT-solver is is
+completed by tracing the {{sat_f}} and cloning the inputs using
+{{clone^n_{bool} }}.
 
@@
 
-When the solver is fed inputs initialized to {{true}}, the {{clone}}
-will clone any inputs to {{sat_f}} that satisfies the constraints
-imposed by {{f}} and the constraints on the {{heap}}. In the case of
-unique-SAT the solver will produce exactly 1 or 0 solutions. In the
-case of general SAT the solver will produce solutions as determined by
-the top-level (Sec. \ref{sec:rat}).
+When the solver is fed inputs initialized to {{true}}, the {{clone}} will
+clone any inputs to {{sat_f}} that satisfy the constraints imposed by {{f}}
+and the constraints on the {{heap}}. In the case of unique-SAT the solver
+will produce exactly 0 or 1 solutions. In the case of general SAT, the solver
+will produce solutions as determined by the semantics of the top-level
+interaction as discussed in Sec. \ref{sec:rat}.
 

### 2012-03-11 rational.tex
intro

@@
 \end{itemize}
-\roshan{Why was this deleted?}
-\amr{because it says nothing new}
 The example illustrates that the empty type and negative types have a
@@
 the presence of an unknown quantity and dually {{1/alpha}} should be thought
-of as the absence of an unknown quantity.
-
-% Unlike with negative types, wherein only one value existed at a time and the
-% computation backtracked, here we have three values that \emph{exist at the
-%   same time}. In other words, the computation with fractions is realized with
-% a schedule in which every value independently and concurrently proceeds
-% through its subcomputation.
-
-At line (3), the missing unknown {{1/alpha}} is fixed to be {{20}}, thereby
-fixing values of all three branches, i.e., we have paid our debt of the missing
-information and the third subcomputation has the known value {{20}}. After
-payment of our debt, as line (4) shows, we have use of our credit card again.
-
-\roshan{Also, why was this deleted?}
-\amr{no real info is communicated; you're saying that pairs are different 
-than sums}
+of as the absence of an unknown quantity. At line (3), the missing unknown
+{{1/alpha}} is brought together with a value {{20}} and at line (4) we use
+the {{20}} to satisfy the constraint {{1/alpha}}. In other words, this branch
+of the computation succeeded in borrowing {{20}} which immediately
+communicates the {{20}} to the rightmost branch.
+
 Unlike with negative types, wherein only one value existed at a time and the
@@
 a schedule in which every value independently and concurrently proceeds
-through its subcomputation.
-
-%% At some intuitive level, one might understand the situation as
-%% modeling a credit card payment in which money is created at line (2)
-%% for the consumer together with a corresponding debt. The producer's
-%% bank pays for the debt at line (3) and waits for the actual money to
-%% arrive from the producer.
-
-The example illustrates that fractional types also have a computational
-interpretation that have some flavor of continuations: the fractional types
-denote values ({{1/alpha}}) that represent missing information that must be
-supplied in much the same sense that continuations denote evaluation contexts
-with holes that must be filled.  There are at least three fundamental points
-about the examples above that must be emphasized:
+through its subcomputation. The example illustrates that fractional types
+also have a computational interpretation that have some flavor of
+continuations: the fractional types denote values ({{1/alpha}}) that
+represent missing information that must be supplied in much the same sense
+that continuations denote evaluation contexts with holes that must be filled.
+
+There are at least four fundamental points about the examples above that must
+be emphasized:
 
 \begin{itemize}
-\item 
-As the example illustrates, both negative types and fractional types
-corresponds to ``debts'' but in different ways: negatives are satisfied
-by backtracking and fractionals are satisfied by constraint
-propagation.
-
-\item 
-It would clearly be disastrous if debts could be deleted or
+\item As the examples illustrate, both negative types and fractional types
+  corresponds to ``debts'' but in different ways: negatives are satisfied by
+  backtracking and fractionals are satisfied by constraint propagation.
+
+\item It would clearly be disastrous if debts could be deleted or
   duplicated. This simple observation explains why these types are much
   simpler and much more appealing in a framework where information is
-guaranteed to be preserved. In previous work that used negative types
-(see Sec.~\ref{sec:related}), complicated mechanisms are typically
-needed to constrain the propagation and use of negative values because
-the surrounding computational framework is, generally speaking,
-careless in its treatment of information.
+  guaranteed to be preserved. In previous work that used negative types (see
+  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
+  constrain the propagation and use of negative values because the
+  surrounding computational framework is, generally speaking, careless in its
+  treatment of information.
+
+\item Each of the values {{-a + b}} and {{(1/a) * b}} can be viewed as a
+  function that asks for an {{a}} and then produces a {{b}}. When viewed as
+  functions, we write these types as {{a -o+ b}} and {{a -o* b}}
+  respectively. Alternatively we can view these values as first producing a
+  {{b}} and then demanding an {{a}} and in that perspective they correspond
+  to delimited continuations. Evidently, as the discussion above suggests,
+  these two notions of functions are not the same at all and should not be
+  conflated. Sec.~\ref{sub:hof} discusses this point in detail.
 
@@
   SAT-solver (see Sec.~\ref{sec:sat-solver}).
-
-%% Computationally, this means that negative types allow us
-%%   to break the dependency between producers and consumers and let each
-%%   compute independently.
-
-%% \item Consider a value of type {{-a + b}}. The above discussion suggests that
-%%   this value demands {{a}} and produces {{b}}, i.e., that it is a
-%%   higher-order funtion mapping {{a}} to {{b}}. Dually and as the underlying
-%%   categorical semantics justifies (see Sec.~\ref{sec:neg}), the same value is
-%%   a continuation which produces {{b}} and then demands {{a}}, which
-%%   introduces a twist on how the function can be used: get the goods {{b}} now
-%%   and pay {{a}} later.
 \end{itemize}
-
-\paragraph*{Functions and Dependency Splitting.} 
-%% Consider the following realization of boolean negation written using a
-%% pattern-matching style:
-%% \begin{verbatim}
-%% notB b = case b of 
-%%   False -> True
-%%   True -> False
-%% \end{verbatim}
-%% In a functional language, we are not used to treating the pattern-matching
-%% clauses \verb|False -> True| and \verb|True -> False| as first-class entities
-%% but fractional types allow us to do just that! Informally, we can treat each
-%% of the pattern-matching clauses as a first-class value of type 
-%% {{(1/bool) * bool}}. 
-%% A small but complete example will illustrate the main aspects of this idea.
-
-%% Let \verb|color| be the datatype with three constructors \verb|R|, \verb|G|,
-%% and \verb|B| and let's try to construct a function that rotates the
-%% colors. In other words, we would like to build the following three
-%% pattern-matching clauses \verb|R -> G|, \verb|G -> B|, and \verb|B -> R|.
-
-The pair {{(1/a) * a}} corresponds to a function type written {{a-o*a}})
-whose witness maybe thought of as the identity function at the type {{a}}.
-For example, consider a datatype {{color = R|Gr|B}}, and let us consider the
-following manipulations:
-
-%% for the
-%% there are three possible
-%% witnesses for this isomorphism mapping {{()}} to either {{(1/R,R)}} or
-%% {{(1/Gr,Gr)}} or {{(1/B,B)}} with the particular witness determined by
-%% further constraints encountered during the computation. More generally, the
-%% witness for the isomorphism above may be represented as {{(1/alpha,alpha)}}
-%% where~{{alpha}} is a fresh logic variable of type {{a}}. With this view of
-%% {{a -o* a}} functions, 
-%% 
-%% Returning to our goal of creating the function that rotates the colors, we
-%% proceed as follows:
-
-\begin{itemize}
-\item Using the fact that {{1}} is the multiplicative unit, generate
-  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
-\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
-  the components of the above tuple. The resulting value is
-  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
-  {{alpha2}} are fresh logic variables;
-\item Using the fact that {{*}} is associative and commutative, we can
-  rearrange the above tuple to produce the value:
-
-{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.
-
-%% \begin{itemize}
-%% \item Using the fact that {{1}} is the multiplicative unit, generate from the
-%%   input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
-%% \item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
-%%   components of the above tuple. The resulting value is 
-%% {{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
-%% {{alpha2}}, and {{alpha3}} are fresh logic variables.
-%% \item Using the fact that {{*}} is associative and commutative, we can
-%%   rearrange the above tuple to produce the value:
-
-%% {{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
-
-%% \item At this point, we have produced the skeleton for our desired function.
-%%   Assume now that a client uses the first clause of the function in a context
-%%   that contains the value \verb|R|, i.e., the client has the value
-%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
-%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
-%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
-%%   the first component of the pair which has the following effects:
-%%   \begin{itemize}
-%%     \item the logic variable {{alpha1}} is unified with {{R}};
-%%     \item the first component of the pair then reduces to () and can be
-%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
-%%       words, the value reduces to simply {{alpha2}} reflecting that the
-%%       pattern-matching was successful and the result of the clause is now
-%%       available;
-%%     \item the first clause of the function has now disappeared and the
-%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
-%%   \item The same client or another client can now use the second clause
-%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
-%%     second client and ground the result of the first client to {{Gr}}.
-%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
-%%     which grounds the result of the second client to {{B}}.
-%%   \end{itemize}
-\end{itemize}
-
-At this point we have constructed a strange mix of two {{a-o*a}} functions;
-inputs of one function manifest themselves as outputs of the other. If
-{{(1/alpha1,alpha2)}} is held by one subcomputation and {{(1/alpha2,alpha1)}}
-is held by another subcomputation, these remixed functions form a
-communication channel between the two concurrent subcomputations. Unifying
-{{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
-be {{R}} in the other. The type {{a}} thus takes the role of the type of the
-communication channel, indicating how much information can be communicated
-between the two subcomputations.  Depending on the choice of the type {{a}},
-an arbitrary number of bits may be communicated.
-
-Dually, the additive reading of the above manipulations correspond to
-functions of the form {{a-o+a}}, witnessing isomorphisms of the form
-{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
-between two subcomputations, \emph{only one of which exists} at any point,
-i.e., they capture the essence of coroutines. Additive and multiplicative
-functions are further discussed in Sec.~\ref{sec:rat}.  It should be evident
-that in a universe in which information is not guaranteed to be preserved by
-the computational infrastructure, the above slicing and dicing of functions
-would make no sense. Once the computational framework guarantees that
-information is maintained, it becomes possible to break dependencies,
-generate debts and constraints, and let these flow as first-class values,
-because they \emph{must} eventually be satisfied.
-
-%% If one branch uses a clause that matches one part of the input to
-%% produce one part of the output, all other clauses are automatically
-%% adjusted to account for this to ensure that this input is not matched
-%% again and that this output is not produced again.
-
-%% \item At this point, we have produced the skeleton for our desired function.
-%%   Assume now that a client uses the first clause of the function in a context
-%%   that contains the value \verb|R|, i.e., the client has the value
-%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
-%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
-%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
-%%   the first component of the pair which has the following effects:
-%%   \begin{itemize}
-%%     \item the logic variable {{alpha1}} is unified with {{R}};
-%%     \item the first component of the pair then reduces to () and can be
-%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
-%%       words, the value reduces to simply {{alpha2}} reflecting that the
-%%       pattern-matching was successful and the result of the clause is now
-%%       available;
-%%     \item the first clause of the function has now disappeared and the
-%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
-%%   \item The same client or another client can now use the second clause
-%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
-%%     second client and ground the result of the first client to {{Gr}}.
-%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
-%%     which grounds the result of the second client to {{B}}.
-%%   \end{itemize}
-
-%% Furthermore, just like the case for negative types, a dual reading of
-%% the pattern-matching clauses as continuations allows the context to
-%% use the result of the clause and provide the matching argument later.
 
@@
   types. (Sec.~\ref{sec:frac})
-\item We combine the above two extensions into one extension that includes
-  both negative and fractional types. The resulting language allows any
-  rational number to be used as a type. The types satisfy the same familiar
-  and intuitive isomorphisms that are satisfied in the mathematical field of
-  rational numbers. (Sec.~\ref{sec:rat})
+\item We combine the above two extensions into a language whose type system
+  allows any rational number to be used as a type. Moreover the types satisfy
+  the same familiar and intuitive isomorphisms that are satisfied in the
+  mathematical field of rational numbers. (Sec.~\ref{sec:rat})
 \item We develop programming intuition and argue that negative and fractional
@@
 
+\paragraph*{Functions and Dependency Splitting.} 
+\label{sub:hof}
+
+We write these type
+written {{a-o*a}} whose witness may be thought of as the identity function at
+the type {{a}}.  For example, consider a datatype {{color = R|Gr|B}}, and let
+us consider the following manipulations:
+
+\begin{itemize}
+\item Using the fact that {{1}} is the multiplicative unit, generate
+  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
+\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
+  the components of the above tuple. The resulting value is
+  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
+  {{alpha2}} are fresh logic variables;
+\item Using the fact that {{*}} is associative and commutative, we can
+  rearrange the above tuple to produce the value:
+
+{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.
+
+%% \begin{itemize}
+%% \item Using the fact that {{1}} is the multiplicative unit, generate from the
+%%   input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
+%% \item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
+%%   components of the above tuple. The resulting value is 
+%% {{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
+%% {{alpha2}}, and {{alpha3}} are fresh logic variables.
+%% \item Using the fact that {{*}} is associative and commutative, we can
+%%   rearrange the above tuple to produce the value:
+
+%% {{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
+
+%% \item At this point, we have produced the skeleton for our desired function.
+%%   Assume now that a client uses the first clause of the function in a context
+%%   that contains the value \verb|R|, i.e., the client has the value
+%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
+%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
+%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
+%%   the first component of the pair which has the following effects:
+%%   \begin{itemize}
+%%     \item the logic variable {{alpha1}} is unified with {{R}};
+%%     \item the first component of the pair then reduces to () and can be
+%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
+%%       words, the value reduces to simply {{alpha2}} reflecting that the
+%%       pattern-matching was successful and the result of the clause is now
+%%       available;
+%%     \item the first clause of the function has now disappeared and the
+%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+%%   \item The same client or another client can now use the second clause
+%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
+%%     second client and ground the result of the first client to {{Gr}}.
+%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
+%%     which grounds the result of the second client to {{B}}.
+%%   \end{itemize}
+\end{itemize}
+
+At this point we have constructed a strange mix of two {{a-o*a}} functions;
+inputs of one function manifest themselves as outputs of the other. If
+{{(1/alpha1,alpha2)}} is held by one subcomputation and {{(1/alpha2,alpha1)}}
+is held by another subcomputation, these remixed functions form a
+communication channel between the two concurrent subcomputations. Unifying
+{{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
+be {{R}} in the other. The type {{a}} thus takes the role of the type of the
+communication channel, indicating how much information can be communicated
+between the two subcomputations.  Depending on the choice of the type {{a}},
+an arbitrary number of bits may be communicated.
+
+Dually, the additive reading of the above manipulations correspond to
+functions of the form {{a-o+a}}, witnessing isomorphisms of the form
+{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
+between two subcomputations, \emph{only one of which exists} at any point,
+i.e., they capture the essence of coroutines. Additive and multiplicative
+functions are further discussed in Sec.~\ref{sec:rat}.  It should be evident
+that in a universe in which information is not guaranteed to be preserved by
+the computational infrastructure, the above slicing and dicing of functions
+would make no sense. Once the computational framework guarantees that
+information is maintained, it becomes possible to break dependencies,
+generate debts and constraints, and let these flow as first-class values,
+because they \emph{must} eventually be satisfied.
+
+%% If one branch uses a clause that matches one part of the input to
+%% produce one part of the output, all other clauses are automatically
+%% adjusted to account for this to ensure that this input is not matched
+%% again and that this output is not produced again.
+
+%% \item At this point, we have produced the skeleton for our desired function.
+%%   Assume now that a client uses the first clause of the function in a context
+%%   that contains the value \verb|R|, i.e., the client has the value
+%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
+%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
+%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
+%%   the first component of the pair which has the following effects:
+%%   \begin{itemize}
+%%     \item the logic variable {{alpha1}} is unified with {{R}};
+%%     \item the first component of the pair then reduces to () and can be
+%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
+%%       words, the value reduces to simply {{alpha2}} reflecting that the
+%%       pattern-matching was successful and the result of the clause is now
+%%       available;
+%%     \item the first clause of the function has now disappeared and the
+%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+%%   \item The same client or another client can now use the second clause
+%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
+%%     second client and ground the result of the first client to {{Gr}}.
+%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
+%%     which grounds the result of the second client to {{B}}.
+%%   \end{itemize}
+
+%% Furthermore, just like the case for negative types, a dual reading of
+%% the pattern-matching clauses as continuations allows the context to
+%% use the result of the clause and provide the matching argument later.
+
 \emph{Constructions.} The constructions for functions, involution, and trace are identical,

### 2012-03-11 rational.tex
working on sec. 3

@@
 circuit. The start machine state, {{<c,v,[]>}}, denotes the
-particle/value~{{v}} about to be evaluated by the circuit {{c}}. The end
+particle~{{v}} about to be evaluated by the circuit {{c}}. The end
 machine state, {{[c, v, [] ]}}, denotes the situation where the particle
@@
 \section{Negative and Fractional Types}
-
-% The structure of this section and the next are the same.
+\label{sec:neg}
+\label{sec:frac}
 
@@
 After presenting the extension at the syntactic level, we discuss the
-%% categorical semantics informally via the graphical language. The
-%% operational semantics is presented in Sec.~\ref{sec:rat}. As will be
-%% detailed in this sections \ref{sec:neg} and \ref{sec:frac}, the
-%% compact closed structure provides several properties of interest: (i)
-%% morphisms or wires are allowed to run from right to left, (ii) the
-%% structure admits a trace that, depending on the situation, can be used
-%% to implement various notions of loops and
 categorical semantics informally via the graphical language. The operational
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Additive Duality in {{langRev}} }
-\label{sec:neg}
+\subsection{Syntax and Types}
 
 Describing the syntax and types of an extension to {{langRev}} with additive
-duality is fairly straightforward. Basically, we extend the language with
-negative types denoted {{-b}}, negative values denoted {{-v}}, and two
-isomorphisms {{eta+}} and {{eps+}} with the following type judgements:
+and multiplicative duality is fairly straightforward. Basically, for the
+additive case, we extend the language with negative types denoted {{-b}},
+negative values denoted {{-v}}, and two isomorphisms {{eta+}} and
+{{eps+}}. Similarly, for the multiplicative case, we extend the language with
+fractional types denoted {{1/b}}, fractional values denoted {{1/v}}, and two
+isomorphisms {{eta*}} and {{eps*}}.
 
 %subcode{bnf} include main
-% Value Types, b = ... | -b
-% Values, v = ... | -v
+% Value Types, b = ... | -b | 1/b
+% Values, v = ... | -v | 1/v
 %
-% Isomorphisms, iso &=& ... | eta+ | eps+
-
+% Isomorphisms, iso &=& ... | eta+ | eps+ | eta* | eps*
+
+For convenience we sometime use the notations {{b1 - b2}} and {{b1/b2}} to
+indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}} respectively.  The
+types of the new constructs are: 
 \vspace{-15pt}
@@
 % eta+ &: 0 <-> (-b) + b :& eps+
+% eta* &: 1 <-> (1/b) * b :& eps*
 
@@
 %@@ |- -v : -b
+%
+%@ |- v : b
+%@@ |- 1/v : 1/b
 \end{multicols}
 
-For the graphical language, we visually present {{eta+}} and {{eps+}} as
-U-shaped connectors. On the left below is {{eta+}} showing the map from {{0}}
-to {{-b+b}} and dually on the right is {{eps+}} showing the map from {{-b+b}}
-to 0. Even though the diagrams below show a {{0}} wire for completeness,
-later diagrams will always drop them in contexts where we can implicitly
-introduce/eliminate then using {{zeroi}}/{{zeroe}}:
+For the graphical language, we visually represent {{eta+}}, {{eps+}},
+{{eta*}}, and {{eps*}} as U-shaped connectors. On the left below are {{eta+}}
+and {{eta*}} showing the maps from {{0}} to {{-b+b}} and {{1}} to {{1/b * b}}. 
+On the right are {{eps+}} and {{eps*}} showing the maps from {{-b+b}} to~0 and 
+{{1/b * b}} to~1. Even though the diagrams below show the {{0}} and {{1}} 
+wires for completeness, later diagrams will always drop them in contexts 
+where they can be implicitly introduced and eliminated. 
 \begin{multicols}{2}
@@
 }
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/eta_times.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps_times.pdf}
 \end{center}
@@
 \end{multicols}
-
-%% Hence, the intuitive way to read the diagrams is that they reverse the
-%% flow of information -- i.e. they capture a notion of \emph{negative
-%%   information flow}. In normal {{langRev}} circuits, information flows
-%% from left-to-right. The operator {{eps+}} causes information to flow
-%% from right-to-left.  The interpretation of {{eta+}} is the opposite:
-%% it turns a right-to-left flow of information in a circuit into a
-%% left-to-right flow.  The diagram below is an example of using {{eta+}}
-%% and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
-%% from the introduction. (Here the {{0}} wires have been shown for
-%% completeness.)
-\paragraph*{Negative Information Flow.} 
-The intuitive way to read the diagrams is that they reverse the flow of
-information. In previous {{langRev}} circuits, information was always flowing
-from left to right. The presence of {{eps+}} operations cause information to
-flow from right to left which can be reversed again using {{eta+}}.  As an
-example, the diagram representing the isomorphism from the introduction is:
+As an example, the diagram representing the first isomorphism in the
+introduction is:
 \begin{center}
@@
 
-% A time traveling intuition is also applicable here: if view the
-% left-to-right direction of information flow as the canonical forward
-% direction of information, then the action of {{eps+}} allows values,
-% aka information particles, to flow backwards in time. Particles that
-% flow backwards in time get to see and interact with the history of
-% particles that they coexist with, i.e., values that they exist
-% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
-% (whose construction is shown in Section \ref{sec:neg-constructions}).
-
-%\subsection{Compact Closure with Additive Inverses}
-%% \emph{Coherence for Compact Closure with Additive Duality.}  The
-%% appropriate coherence condition for compact closed categories,
-%% expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-%% may be visualized (modulo {{0}} wires, swapping etc.) as explained by
-%% Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
-%% operational equivalence:
-
-\paragraph*{Coherence}
-The coherence condition for compact closed categories may be visualized as
-follows~\cite{springerlink:10.1007/978-3-642-12821-94}:
+Similarly, in the usual interpretation of {{b1*b2}} we have both a value of
+type {{b1}} and a value of type {{b2}}. This interpretation is maintained in
+the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be
+viewed as a fission point for a value of type {{b}} and its multiplicative
+inverse {{1/b}}. Operationally, this corresponds to the creation of two
+values {{alpha}} and {{1/alpha}} where {{alpha}} is a fresh logic
+variable. The operator {{eps*}} then becomes a unification site for these
+logic variables:
+\begin{multicols}{2}
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eta_times1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_times1.pdf}
+}
+\end{center}  
+\end{multicols}
+
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Categorical Constructions}
+
+All the constructions below are standard: they are collected from Selinger's
+survey paper on monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94} and presented in
+the context of our language.
+
+For a monoidal category to be compact closed, the maps {{eta}} and {{eps}}
+must satisfy a coherence condition that is usually visualized as follows
 \begin{center}
@@
 \end{center}
-It is intuitively clear that if the interpretations of {{eta+}} and {{eps+}}
-are to reverse the flow of execution, then this coherence condition is
-satisfied.
+In the case of negatives, the condition amounts to checking that reversing
+direction twice is a no-op. In the case of fractionals, the condition amounts
+to checking that creating values {{alpha}} and {{1/alpha}} and immediately
+unifying them is also a no-op. Both checks are straightforward.
+
+We now review several interesting constructions related to looping,
+involution, and higher order functions.
 
 \paragraph*{Trace.}
-Every compact closed categoy admits a trace. In the context of {{langRev}}
-with {{eta+}}/{{eps+}}, we get the following definition. 
-Given {{f : a+c <-> b+c }}, define {{trace+ f : a <-> b}} as:
+Every compact closed categoy admits a trace. For the additive case, we get
+the following definition.  Given {{f : a+c <-> b+c }}, 
+define {{trace+ f : a <-> b}} as:
 
@@
 
-For readability, we have omitted some the commutativity and
-associativity shuffling.
-
-%% \emph{Non-termination.}  Compact closed categories contain {{trace}}
-%% operations and this implies that {{eta+}}/{{eps+}} allow the
-%% construction of a {{trace+}} operator (shown in Section
-%% \ref{sec:monoidal-constructions}).  As shown in previous work
-%% \cite{infeffects}, non-terminating computation can be constructed from
-%%      {{trace+}} and recursive types.
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/trace.pdf}
-\end{center}
-
-%subcode{proof} include main
-%@ c : b2 + b1 <-> b2 + b3
-%@@ trace c : b1 <-> b3
-
-\end{multicols}
-
-% \paragraph*{Zero Information.}
-% The fact that {{eta+}} (and hence {{eps+}}) introduces no information
-% can be understood is by viewing {{-b+b}} as a type whose value is
-% observable only if you `pay' {{b}} amount of information to the {{-b}}
-% branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
-% amount of information is returned to you. (TODO. more needs to be said
-% here.)
-
-% For this reason, our treatment of negative values in the context of
-% {{langRev}} is particularly simple. We will have a type $0$ and an
-% isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-% allows us to create a value and a corresponding debt out of nothing. Both the
-% value and its negative counterpart can flow anywhere in the system. Because
-% information is preserved, a closed program (which does not have any
-% ``dangling'' negative values) will eventually have to match the negative
-% value with some corresponding value, effectively using the isomorphism in the
-% right-to-left direction.
-
-\emph{Involution (Principium Contradictiones)}
+We have omitted some the commutativity and associativity shuffling to
+communicate the main idea. We are given a value of type {{a}} which we embed
+into {{a+0}} and then {{a + (c - c)}}. This can be reassociated into 
+{{(a+c) - c}}. The component {{a+c}} which until now is just an appropriately 
+tagged value of type {{a}} can be given to {{f}} which transforms it to
+something of type {{b+c}}. If the result is in the {{b}}-summand, it is produced
+as the answer; otherwise the result is in the {{c}}-summand; {{eps+}} is used to
+make it flow backwards to be fed to the {{eta+}} located at the beginning of the 
+sequence. Iteration continues until a {{b}} is produced.
+
+%%%%%
+\paragraph*{Involution (Principium Contradictiones)}
 
@@
 
-\emph{Functions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Functions and Delimited Continuations}
+\label{sub:hof}
 
@@
 vs. multiplicative functions.
-\roshan{Where is the HOF discussion?}
-\amr{deleted because it was weak and unfocused}
-
-\emph{Function composition.}
 
@@
 
-\emph{Function application.}
-test.
 \begin{multicols}{2}
@@
 
-\emph{Delimited continuation}
+%%%%%
+\paragraph*{Delimited continuation}
 
@@
 \end{center}
+
+In a conventional language, higher-order functions capture scope and can be
+represented as first-class values. In {{langRevEE}}, we have two notions of
+functions both of which share similarities with conventional higher-order
+functions.
+
+We write these type
+written {{a-o*a}} whose witness may be thought of as the identity function at
+the type {{a}}.  For example, consider a datatype {{color = R|Gr|B}}, and let
+us consider the following manipulations:
+
+\begin{itemize}
+\item Using the fact that {{1}} is the multiplicative unit, generate
+  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
+\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
+  the components of the above tuple. The resulting value is
+  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
+  {{alpha2}} are fresh logic variables;
+\item Using the fact that {{*}} is associative and commutative, we can
+  rearrange the above tuple to produce the value:
+
+{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.
+
+\end{itemize}
+
+At this point we have constructed a strange mix of two {{a-o*a}} functions;
+inputs of one function manifest themselves as outputs of the other. If
+{{(1/alpha1,alpha2)}} is held by one subcomputation and {{(1/alpha2,alpha1)}}
+is held by another subcomputation, these remixed functions form a
+communication channel between the two concurrent subcomputations. Unifying
+{{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
+be {{R}} in the other. The type {{a}} thus takes the role of the type of the
+communication channel, indicating how much information can be communicated
+between the two subcomputations.  Depending on the choice of the type {{a}},
+an arbitrary number of bits may be communicated.
+
+Dually, the additive reading of the above manipulations correspond to
+functions of the form {{a-o+a}}, witnessing isomorphisms of the form
+{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
+between two subcomputations, \emph{only one of which exists} at any point,
+i.e., they capture the essence of coroutines. Additive and multiplicative
+functions are further discussed in Sec.~\ref{sec:rat}.  It should be evident
+that in a universe in which information is not guaranteed to be preserved by
+the computational infrastructure, the above slicing and dicing of functions
+would make no sense. Once the computational framework guarantees that
+information is maintained, it becomes possible to break dependencies,
+generate debts and constraints, and let these flow as first-class values,
+because they \emph{must} eventually be satisfied.
+
+%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions specific to Negatives and Fractions}
+\label{sec:specific-constructions}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multiplicative Duality in {{langRev}} }
-\label{sec:frac}
-
-%% %% The section follows the outline of the previous one, first presenting
-%% %% the new syntax, and discussing the graphical language and the
-%% %% properites implied by the underlying categorical structure.
-
-Multiplicative duality is added to {{langRev}} by adding types
-{{1/b}}, values {{1/v}} and isomorphisms {{eta*}} and {{eps*}}.
-
-%subcode{bnf} include main
-% Value Types, b = ... | 1/b
-% Values, v = ... | 1/v
-%
-% Isomorphisms, iso &=& ... | eta* | eps*
-
-These have type judgements, similar to those of their additive
-couterparts.
-\begin{multicols}{2}
+Annihilation and non-termination describe undefined program
+execution. While non-termination characterizes an iteration that fails
+to terminate, annihilation characterizes a constraint that has no
+fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
+annihilating only the program states that do not match the boolean
+satisfiability constraint.
+
+\emph{Annihilation.}  
+With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
+which gives us the ability to find the fixpoint of a constraint.  If
+the constraint has no satisfying values, there is no fixpoint possible
+-- we use the term \emph{annihilation} to describe the corresponding
+undefined state of the program.
+
+While the constructions in Sec \ref{sec:neg} apply to both the
+{{(0,+)}} and {{(1,*)}} monoids, each construction presented here
+applies to a specific monoid.
+
+\begin{enumerate}
+
+\item 
+\emph{Lifting negation out of {{*}}. }
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+To build these isomorphisms, we first build an intermeidate
+construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/eps_fst.pdf}
+}
+\end{center}
+
+The isomorphism {{(-b1) * b2 <-> -(b1 * b2)}} can be constructed in
+terms of {{eps_{fst} }} as shown below.
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/mult_neg.pdf}
+}
+\end{center}
+
+The second isomorphism can be built in the same way by merely swapping
+the arguments.
+
+\item
+\emph{Multiplying Negatives}
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
+
+This isomorphism can be built using \emph{involution} constructed in
+Sec~\ref{sec:neg} and the isomorphisms above, and correspond to the
+algebraic manipulation:
+
 %subcode{opsem} include main
-% eta* &: 1 <-> (1/b) * b :& eps*
-
-%subcode{proof} include main
-%@ |- v : b
-%@@ |- 1/v : 1/b  
-\end{multicols}
-
-For convenience we sometime use the notation {{b1/b2}} to indicate the
-pair {{b1 * (1/b2)}}.  As with additive duals, we represent {{eta*}}
-and {{eps*}} using U-shaped connectors. In later diagrams, the wires
-for {{1}} will be dropped as they can freely introduced or eliminated
-using {{uniti}} and {{unite}}.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/eta_times.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/eps_times.pdf}
-\end{center}
-\end{multicols}
-
-%% The usual interpretation of {{b1*b2}} we have both a value of type
-%% {{b1}} and a value of type {{b2}}. This interpretation is maintained
-%% in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
-%% to be viewed as a fission point for a value of type {{b}} and its
-%% multiplicative inverse {{1/b}}. This naturally raises the question --
-%% which value of type {{b}} is produced? The isomorphism {{eta*}} does
-%% not produce any specific value of type {{b}}, instead it produces a
-%% placeholder for an unspecified value and dually, {{1/b}} is inhabited
-%% by the placeholder that indicates the absence of the unspecified
-%% value.
-In the usual interpretation of {{b1*b2}} we have both a value of type {{b1}}
-and a value of type {{b2}}. This interpretation is maintained in the presence
-of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be viewed as a
-fission point for a value of type {{b}} and its multiplicative inverse
-{{1/b}}. This naturally raises the question -- which value of type {{b}} is
-produced? The isomorphism {{eta*}} does not produce any specific value of
-type {{b}}, instead it produces a place holder for a value such that the
-value of type {{1/b}} is always the dual of the chosen one.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eta_times1.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_times1.pdf}
-}
-\end{center}  
-\end{multicols}
-
-\paragraph*{Coherence.}
-The following coherence condition is satisfied as before.
-
-In terms of \emph{logic programming}, {{eta* : 1 <-> (1/b)*b}} acts as
-the site of creation of a new typed logic variable and constructs two
-values using it: a value of type {{b}} and its dual of type
-{{1/b}}. These two values are free to move independently through the
-computation. Operations on one value affect what operations are possible
-on the other since they are both related by an underlying logic
-variable. The operator {{eps*}} acts as a unification site for values of
-type {{b}} and {{1/b}} flowing along both branches. If the values are
-not unifiable duals, then the specific program sequence fails. Like
-with logic programming, one can has many worlds where some might fail,
-while others make progress. Sec. \ref{sec:rat} has more details.  If
-the values are exactly duals, then unification succeeds and {{eta*}}
-returns {{1}}.
+%! columnStyle = rc
+%   & b1 * b2
+% = & -(-(b1*b2)) 
+% = & -((-b1)*b2)
+% = & (-b1)*(-b2)
+
+\item 
+\emph{Multiplying Fractions}
+
+{{b1/b2 * b3/b4 <-> (b1*b3)/(b2*b4)}}
+
+\item 
+\emph{Adding the numerator}
+
+{{b1/b + b2/b <-> (b1+b2)/b}}
+
+\item 
+\emph{Fraction Addition}
+
+{{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
+
+\end{enumerate}
+
 % \paragraph*{Unification.}
@@
 
-\paragraph*{Coherence.}
-The following coherence condition is satisfied as before.
-% ADD MATH.
-
-% %%%%%%%%%%%%%%%%%%%%%%%%
-% \subsection{Conventional Higher-Order Functions}
-% \label{sec:hof}
-
-\emph{Annihilation.}  
-With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
-which gives us the ability to find the fixpoint of a constraint.  If
-the constraint has no satisfying values, there is no fixpoint possible
--- we use the term \emph{annihilation} to describe the corresponding
-undefined state of the program.
-% In a conventional language, higher-order functions capture scope and can be
-% represented as first-class values. In {{langRevEE}}, we have two notions of
-% functions both of which share similarities with conventional higher-order
-% functions.
-
-Annihilation and non-termination describe undefined program
-execution. While non-termination characterizes an iteration that fails
-to terminate, annihilation characterizes a constraint that has no
-fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
-annihilating only the program states that do not match the boolean
-satisfiability constraint.
+%% %% The section follows the outline of the previous one, first presenting
+%% %% the new syntax, and discussing the graphical language and the
+%% %% properites implied by the underlying categorical structure.
+%% The usual interpretation of {{b1*b2}} we have both a value of type
+%% {{b1}} and a value of type {{b2}}. This interpretation is maintained
+%% in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
+%% to be viewed as a fission point for a value of type {{b}} and its
+%% multiplicative inverse {{1/b}}. This naturally raises the question --
+%% which value of type {{b}} is produced? The isomorphism {{eta*}} does
+%% not produce any specific value of type {{b}}, instead it produces a
+%% placeholder for an unspecified value and dually, {{1/b}} is inhabited
+%% by the placeholder that indicates the absence of the unspecified
+%% value.
+
+%% \begin{enumerate}
+%% \item 
+%% \emph{Trace}
+%% % \item
+%% % Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
+%% % are indeed first-class values. However these functions can only be
+%% % linearly used, i.e., they represent a constraint that can be satisfied
+%% % at most once.
+%% % %%  in one `computational world'. It may be possible that to
+%% % %% repeatedly invoke multiplicative functions with an appropriate
+%% % %% recursive type whose unfoldings allow for repeated unification.
+%% % \end{itemize}
+
+% The exact encoding on HOF in {{langRevEE}} should shed light on
+% several problems, including the connections to continuations discussed
+% in Sec. \ref{sec:related}. They should also allow for a natural
+% encoding of HOFs in the translations of \lcal to {{langRevT}} and
+% indicate a monadic encapsulation of of information effects which are
+% currrently structured as an arrow.
+
+%% If one branch uses a clause that matches one part of the input to
+%% produce one part of the output, all other clauses are automatically
+%% adjusted to account for this to ensure that this input is not matched
+%% again and that this output is not produced again.
+
+%% \item At this point, we have produced the skeleton for our desired function.
+%%   Assume now that a client uses the first clause of the function in a context
+%%   that contains the value \verb|R|, i.e., the client has the value
+%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
+%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
+%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
+%%   the first component of the pair which has the following effects:
+%%   \begin{itemize}
+%%     \item the logic variable {{alpha1}} is unified with {{R}};
+%%     \item the first component of the pair then reduces to () and can be
+%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
+%%       words, the value reduces to simply {{alpha2}} reflecting that the
+%%       pattern-matching was successful and the result of the clause is now
+%%       available;
+%%     \item the first clause of the function has now disappeared and the
+%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+%%   \item The same client or another client can now use the second clause
+%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
+%%     second client and ground the result of the first client to {{Gr}}.
+%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
+%%     which grounds the result of the second client to {{B}}.
+%%   \end{itemize}
+
+%% Furthermore, just like the case for negative types, a dual reading of
+%% the pattern-matching clauses as continuations allows the context to
+%% use the result of the clause and provide the matching argument later.
+
+%% \begin{itemize}
+%% \item Using the fact that {{1}} is the multiplicative unit, generate from the
+%%   input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
+%% \item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
+%%   components of the above tuple. The resulting value is 
+%% {{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
+%% {{alpha2}}, and {{alpha3}} are fresh logic variables.
+%% \item Using the fact that {{*}} is associative and commutative, we can
+%%   rearrange the above tuple to produce the value:
+
+%% {{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
+
+%% \item At this point, we have produced the skeleton for our desired function.
+%%   Assume now that a client uses the first clause of the function in a context
+%%   that contains the value \verb|R|, i.e., the client has the value
+%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
+%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
+%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
+%%   the first component of the pair which has the following effects:
+%%   \begin{itemize}
+%%     \item the logic variable {{alpha1}} is unified with {{R}};
+%%     \item the first component of the pair then reduces to () and can be
+%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
+%%       words, the value reduces to simply {{alpha2}} reflecting that the
+%%       pattern-matching was successful and the result of the clause is now
+%%       available;
+%%     \item the first clause of the function has now disappeared and the
+%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
+%%   \item The same client or another client can now use the second clause
+%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
+%%     second client and ground the result of the first client to {{Gr}}.
+%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
+%%     which grounds the result of the second client to {{B}}.
+%%   \end{itemize}
 % \paragraph*{Unification.}
@@
 
-\paragraph*{Functions and Dependency Splitting.} 
-\label{sub:hof}
-
-We write these type
-written {{a-o*a}} whose witness may be thought of as the identity function at
-the type {{a}}.  For example, consider a datatype {{color = R|Gr|B}}, and let
-us consider the following manipulations:
-
-\begin{itemize}
-\item Using the fact that {{1}} is the multiplicative unit, generate
-  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
-\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
-  the components of the above tuple. The resulting value is
-  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
-  {{alpha2}} are fresh logic variables;
-\item Using the fact that {{*}} is associative and commutative, we can
-  rearrange the above tuple to produce the value:
-
-{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.
-
-%% \begin{itemize}
-%% \item Using the fact that {{1}} is the multiplicative unit, generate from the
-%%   input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
-%% \item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
-%%   components of the above tuple. The resulting value is 
-%% {{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
-%% {{alpha2}}, and {{alpha3}} are fresh logic variables.
-%% \item Using the fact that {{*}} is associative and commutative, we can
-%%   rearrange the above tuple to produce the value:
-
-%% {{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
-
-%% \item At this point, we have produced the skeleton for our desired function.
-%%   Assume now that a client uses the first clause of the function in a context
-%%   that contains the value \verb|R|, i.e., the client has the value
-%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
-%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
-%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
-%%   the first component of the pair which has the following effects:
-%%   \begin{itemize}
-%%     \item the logic variable {{alpha1}} is unified with {{R}};
-%%     \item the first component of the pair then reduces to () and can be
-%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
-%%       words, the value reduces to simply {{alpha2}} reflecting that the
-%%       pattern-matching was successful and the result of the clause is now
-%%       available;
-%%     \item the first clause of the function has now disappeared and the
-%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
-%%   \item The same client or another client can now use the second clause
-%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
-%%     second client and ground the result of the first client to {{Gr}}.
-%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
-%%     which grounds the result of the second client to {{B}}.
-%%   \end{itemize}
-\end{itemize}
-
-At this point we have constructed a strange mix of two {{a-o*a}} functions;
-inputs of one function manifest themselves as outputs of the other. If
-{{(1/alpha1,alpha2)}} is held by one subcomputation and {{(1/alpha2,alpha1)}}
-is held by another subcomputation, these remixed functions form a
-communication channel between the two concurrent subcomputations. Unifying
-{{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
-be {{R}} in the other. The type {{a}} thus takes the role of the type of the
-communication channel, indicating how much information can be communicated
-between the two subcomputations.  Depending on the choice of the type {{a}},
-an arbitrary number of bits may be communicated.
-
-Dually, the additive reading of the above manipulations correspond to
-functions of the form {{a-o+a}}, witnessing isomorphisms of the form
-{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
-between two subcomputations, \emph{only one of which exists} at any point,
-i.e., they capture the essence of coroutines. Additive and multiplicative
-functions are further discussed in Sec.~\ref{sec:rat}.  It should be evident
-that in a universe in which information is not guaranteed to be preserved by
-the computational infrastructure, the above slicing and dicing of functions
-would make no sense. Once the computational framework guarantees that
-information is maintained, it becomes possible to break dependencies,
-generate debts and constraints, and let these flow as first-class values,
-because they \emph{must} eventually be satisfied.
-
-%% If one branch uses a clause that matches one part of the input to
-%% produce one part of the output, all other clauses are automatically
-%% adjusted to account for this to ensure that this input is not matched
-%% again and that this output is not produced again.
-
-%% \item At this point, we have produced the skeleton for our desired function.
-%%   Assume now that a client uses the first clause of the function in a context
-%%   that contains the value \verb|R|, i.e., the client has the value
-%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
-%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
-%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
-%%   the first component of the pair which has the following effects:
-%%   \begin{itemize}
-%%     \item the logic variable {{alpha1}} is unified with {{R}};
-%%     \item the first component of the pair then reduces to () and can be
-%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
-%%       words, the value reduces to simply {{alpha2}} reflecting that the
-%%       pattern-matching was successful and the result of the clause is now
-%%       available;
-%%     \item the first clause of the function has now disappeared and the
-%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
-%%   \item The same client or another client can now use the second clause
-%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
-%%     second client and ground the result of the first client to {{Gr}}.
-%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
-%%     which grounds the result of the second client to {{B}}.
-%%   \end{itemize}
-
-%% Furthermore, just like the case for negative types, a dual reading of
-%% the pattern-matching clauses as continuations allows the context to
-%% use the result of the clause and provide the matching argument later.
-
-\emph{Constructions.} The constructions for functions, involution, and trace are identical,
-replacing {{+}} by {{*}}, \textit{mutatis mutandis}.
-
-%% \begin{enumerate}
-%% \item 
-%% \emph{Trace}
-%% % \item
-%% % Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-%% % are indeed first-class values. However these functions can only be
-%% % linearly used, i.e., they represent a constraint that can be satisfied
-%% % at most once.
-%% % %%  in one `computational world'. It may be possible that to
-%% % %% repeatedly invoke multiplicative functions with an appropriate
-%% % %% recursive type whose unfoldings allow for repeated unification.
-%% % \end{itemize}
-
-% The exact encoding on HOF in {{langRevEE}} should shed light on
-% several problems, including the connections to continuations discussed
-% in Sec. \ref{sec:related}. They should also allow for a natural
-% encoding of HOFs in the translations of \lcal to {{langRevT}} and
-% indicate a monadic encapsulation of of information effects which are
-% currrently structured as an arrow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions specific to Negatives and Fractions}
-\label{sec:specific-constructions}
-
-While the constructions in Sec \ref{sec:neg} apply to both the
-{{(0,+)}} and {{(1,*)}} monoids, each construction presented here
-applies to a specific monoid.
-
-\roshan{Each of these isomorphisms has baffled us -- either we
-  believed they were not constructible. Or we have struggled to
-  comprehend what they mean. Hence it is worth saying a sentence or
-  two about each beyond how they are constructed.}
-
-\begin{enumerate}
-
-%% \item 
-%% \emph{ {{eps_{fst} }} }
-
-%% {{(
-
-\item 
-\emph{Lifting negation out of {{*}}. }
-
-{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
-
-To build these isomorphisms, we first build an intermeidate
-construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
-\begin{center}
-\scalebox{0.8}{
-  \includegraphics{diagrams/eps_fst.pdf}
-}
-\end{center}
-
-The isomorphism {{(-b1) * b2 <-> -(b1 * b2)}} can be constructed in
-terms of {{eps_{fst} }} as shown below.
-
-\begin{center}
-\scalebox{0.9}{
-  \includegraphics{diagrams/mult_neg.pdf}
-}
-\end{center}
-
-The second isomorphism can be built in the same way by merely swapping
-the arguments.
-
-\item
-\emph{Multiplying Negatives}
-
-{{b1 * b2 <-> (-b1)*(-b2)}}
-
-This isomorphism can be built using \emph{involution} constructed in
-Sec \ref{sec:neg} and the isomorphisms above, and correspond to the
-algebraic manipulation:
-
-%subcode{opsem} include main
-%! columnStyle = rc
-%   & b1 * b2
-% = & -(-(b1*b2)) 
-% = & -((-b1)*b2)
-% = & (-b1)*(-b2)
-
-%% \emph{Distributing fractions over {{*}} }
-%% {{1/(b1*b2) <-> (1/b1) * (1/b2)}}
-
-%% \emph{Lifting Computation intro fractions}
-%% {{b1 <-> b2 ==> 1/b1 <-> 1/b2}}
-
-\item 
-\emph{Multiplying Fractions}
-
-{{b1/b2 * b3/b4 <-> (b1*b3)/(b2*b4)}}
-
-\item 
-\emph{Adding the numerator}
-
-{{b1/b + b2/b <-> (b1+b2)/b}}
-
-\item 
-\emph{Fraction Addition}
-
-{{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
-
-\end{enumerate}
+%% \emph{Non-termination.}  Compact closed categories contain {{trace}}
+%% operations and this implies that {{eta+}}/{{eps+}} allow the
+%% construction of a {{trace+}} operator (shown in Section
+%% \ref{sec:monoidal-constructions}).  As shown in previous work
+%% \cite{infeffects}, non-terminating computation can be constructed from
+%%      {{trace+}} and recursive types.
+% \begin{multicols}{2}
+% \begin{center}
+%   \includegraphics{diagrams/trace.pdf}
+% \end{center}
+% %subcode{proof} include main
+% %@ c : b2 + b1 <-> b2 + b3
+% %@@ trace c : b1 <-> b3
+% \end{multicols}
+
+% \paragraph*{Zero Information.}
+% The fact that {{eta+}} (and hence {{eps+}}) introduces no information
+% can be understood is by viewing {{-b+b}} as a type whose value is
+% observable only if you `pay' {{b}} amount of information to the {{-b}}
+% branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
+% amount of information is returned to you. (TODO. more needs to be said
+% here.)
+
+% For this reason, our treatment of negative values in the context of
+% {{langRev}} is particularly simple. We will have a type $0$ and an
+% isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+% allows us to create a value and a corresponding debt out of nothing. Both the
+% value and its negative counterpart can flow anywhere in the system. Because
+% information is preserved, a closed program (which does not have any
+% ``dangling'' negative values) will eventually have to match the negative
+% value with some corresponding value, effectively using the isomorphism in the
+% right-to-left direction.
+
+%% Hence, the intuitive way to read the diagrams is that they reverse the
+%% flow of information -- i.e. they capture a notion of \emph{negative
+%%   information flow}. In normal {{langRev}} circuits, information flows
+%% from left-to-right. The operator {{eps+}} causes information to flow
+%% from right-to-left.  The interpretation of {{eta+}} is the opposite:
+%% it turns a right-to-left flow of information in a circuit into a
+%% left-to-right flow.  The diagram below is an example of using {{eta+}}
+%% and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
+%% from the introduction. (Here the {{0}} wires have been shown for
+%% completeness.)
+%% \paragraph*{Negative Information Flow.} 
+% A time traveling intuition is also applicable here: if view the
+% left-to-right direction of information flow as the canonical forward
+% direction of information, then the action of {{eps+}} allows values,
+% aka information particles, to flow backwards in time. Particles that
+% flow backwards in time get to see and interact with the history of
+% particles that they coexist with, i.e., values that they exist
+% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
+% (whose construction is shown in Section \ref{sec:neg-constructions}).
+%\subsection{Compact Closure with Additive Inverses}
+%% \emph{Coherence for Compact Closure with Additive Duality.}  The
+%% appropriate coherence condition for compact closed categories,
+%% expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
+%% may be visualized (modulo {{0}} wires, swapping etc.) as explained by
+%% Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
+%% operational equivalence:
 

### 2012-03-11 rational.tex
done with 3

@@
 \usepackage{cmll}
+
+\newcommand{\fname}[1]{\ulcorner #1 \urcorner}
+\newcommand{\fconame}[1]{\llcorner #1 \lrcorner}
 
@@
   respectively. Alternatively we can view these values as first producing a
-  {{b}} and then demanding an {{a}} and in that perspective they correspond
-  to delimited continuations. Evidently, as the discussion above suggests,
-  these two notions of functions are not the same at all and should not be
-  conflated. Sec.~\ref{sub:hof} discusses this point in detail.
+  value of type {{b}} and then demanding an {{a}} and in that perspective
+  they correspond to delimited continuations. Evidently, as the discussion
+  above suggests, these two notions of functions are not the same at all and
+  should not be conflated. Sec.~\ref{sub:hof} discusses this point in detail.
 
@@
 
+\paragraph*{Adjoint.} 
+An important property of the language is that every combinator {{c}} has an
+adjoint {{c{dagger}}} that reverses the action of {{c}}. This is evident by
+construction for the primitive isomorphisms. For the closure combinators, the
+adjoint is homomorphic except for the case of sequencing in which the order
+is reversed, i.e., {{(c1 (;) c2){dagger} = (c2{dagger}) (;) (c1{dagger}) }}.
+
 %%%%%%%%%%%%%%%
@@
 {{<--|}}. When running backwards, we use the isomorphism represented by a
-combinator {{c}} in the reverse direction which we indicate by writing
-{{c{dagger}}}. For example, {{distrib{dagger} = factor}} and so on.
+combinator {{c}} in the reverse direction, i.e., we use the adjoint
+{{c{dagger}}}.
 
@@
 
-For a monoidal category to be compact closed, the maps {{eta}} and {{eps}}
-must satisfy a coherence condition that is usually visualized as follows
+For a monoidal category to be compact closed, the maps~{{eta}} and~{{eps}}
+must satisfy a coherence condition that is usually visualized as follows:
 \begin{center}
@@
 
-We have omitted some the commutativity and associativity shuffling to
+\noindent We have omitted some the commutativity and associativity shuffling to
 communicate the main idea. We are given a value of type {{a}} which we embed
@@
 {{(a+c) - c}}. The component {{a+c}} which until now is just an appropriately 
-tagged value of type {{a}} can be given to {{f}} which transforms it to
+tagged value of type {{a}} can be given to~{{f}} which transforms it to
 something of type {{b+c}}. If the result is in the {{b}}-summand, it is produced
@@
 
-{{b <-> -(-b)}}
-
-\begin{center}
-  \includegraphics{diagrams/double_neg.pdf}
-\end{center}
+In a symmetric compact closed category, we can build isomorphisms that the
+dual operation is an involution. Specifically, we get the isomorphisms 
+{{b <-> -(-b)}} and {{b <-> (1/(1/b))}}. For the additive case, the 
+isomorphism is defined as follows:
+
+{{ (id (+) eta+) (;) (swap+ (+) id) (;) (id (+) eps+) }}
+
+\noindent where we have omitted the 0 introduction and elimination. The idea
+is as follows: we start with a value of type {{b}}, embed it into {{b+0}} and
+use {{eta}} to create something of type {{b + (-(-b) + (-b))}}. This is possible
+because {{eta}} has the polymorphic type {{-a + a}} which can be instantiated
+to {{-b}}. We then reshuffle the type to produce {{-(-b) + (-b + b)}} and cancel
+the right hand side using {{eps+}}.  The construction for the multiplicative 
+case is identical and omitted.
+
+%% {{b <-> -(-b)}}
+%% 
+%% this is the wrong diagram: see lemma 4.17 in selinger's paper
+%% \begin{center}
+%%   \includegraphics{diagrams/double_neg.pdf}
+%% \end{center}
 
@@
 
-Mention LL and {{a -o b}}.
+Although these constructions are also standard, they are less known and they
+are particularly important in our context: we devote a little more time to
+explain them. Our discussion is mostly based on Abramsky and Coecke's article
+on categorical quantum mechanics~\cite{abramsky-2008}.
+
+In a compact closed category, each morphism {{f : a <-> b }} can be given a
+\emph{name} and a \emph{coname}. For the additive fragment, the name
+$\fname{f}$ has type {{0 <-> (-a + b)}} and the coname $\fconame{f}$ has type
+{{a + (-b) <-> 0}}. For the multiplicative fragment, the name $\fname{f}$ has
+type {{1 <-> ((1/a) * b))}} and the coname $\fconame{f}$ has type 
+{{(a * (1/b)) <-> 1}}. Intuitively, this means that for each morphism, 
+it is possible to construct, from ``nothing,'' an object in the category that 
+denotes this morphism, and dually it is possible to eliminate this object.
+The construction of the name and coname of {{c : b1 <-> b2}} in the additive case 
+can be visualized as follows:
 
@@
 \end{center}
-
-Let us use the shorthand {{b1 -o+ b2 = 0 <-> -b1 + b2}}. See
-Sec. \ref{sec:hof} for a discussion of the additive functions
-vs. multiplicative functions.
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}
+
+Intuitively the name consists of viewing {{c}} as a function and the coname
+consists of viewing {{c}} as a delimited continuation.
+
+In addition to being able to represent morphisms, it is possible to express
+function composition. For the additive case, the composition is depicted
+below:
 
@@
 \end{multicols}
-This is also equivalent to sequencing both the computation blocks.
+which is essentially equivalent to sequencing both the computation blocks as
+shown below:
 
@@
 
+Applying a function consists of making it flow backwards to satisfy the demand 
+of the function:
 \begin{multicols}{2}
@@
 
-%%%%%
-\paragraph*{Delimited continuation}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}
-
-In a conventional language, higher-order functions capture scope and can be
-represented as first-class values. In {{langRevEE}}, we have two notions of
-functions both of which share similarities with conventional higher-order
-functions.
-
-We write these type
-written {{a-o*a}} whose witness may be thought of as the identity function at
-the type {{a}}.  For example, consider a datatype {{color = R|Gr|B}}, and let
-us consider the following manipulations:
-
+Having reviewed the representation of functions, we now discuss the
+similarities and differences between the two notions of functions and their
+relation to conventional (linear) functions which mix additive and
+multiplicative components. For that purpose, we use a small example.
+Consider a datatype {{color = R|Gr|B}}, and let us consider the following
+manipulations:
 \begin{itemize}
@@
 between two subcomputations, \emph{only one of which exists} at any point,
-i.e., they capture the essence of coroutines. Additive and multiplicative
-functions are further discussed in Sec.~\ref{sec:rat}.  It should be evident
-that in a universe in which information is not guaranteed to be preserved by
-the computational infrastructure, the above slicing and dicing of functions
-would make no sense. Once the computational framework guarantees that
-information is maintained, it becomes possible to break dependencies,
-generate debts and constraints, and let these flow as first-class values,
-because they \emph{must} eventually be satisfied.
+i.e., they capture the essence of coroutines.
+
+It should be evident that in a universe in which information is not
+guaranteed to be preserved by the computational infrastructure, the above
+slicing and dicing of functions would make no sense. But linearity is not
+sufficient: one must also recognize that the additive and multiplicative
+spaces are different.
 
 %%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructions specific to Negatives and Fractions}
+\subsection{Additional Constructions}
 \label{sec:specific-constructions}
 
-\emph{Negation distributes over {{+}}. }
-
+The additional constructions below (presented with minimal commentary)
+confirm that conventional algebraic manipulations in the mathematical field
+of rationals do indeed correspond to realizable type isomorphisms in our
+setting. The constructions involving both negative and fractional types are
+novel.
+
+\paragraph*{Negation distributes over {{+}}. }
+One certainly expects that:
 {{-(b1 + b2) <-> (-b1) + (-b2)}}
-
+which can be realized as follows:
 \begin{center}
@@
 
-\emph{Lifting an operation of positive types to negated types}
-
+\paragraph*{Lifting an operation of positive types to negated types.}
+Any operation on types can be applied to the negative versions of these
+types. The idea is simply to reverse the flow of values and use the
+adjoint of the operation:
 \begin{multicols}{2}
@@
 
-Annihilation and non-termination describe undefined program
-execution. While non-termination characterizes an iteration that fails
-to terminate, annihilation characterizes a constraint that has no
-fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
-annihilating only the program states that do not match the boolean
-satisfiability constraint.
-
-\emph{Annihilation.}  
-With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
-which gives us the ability to find the fixpoint of a constraint.  If
-the constraint has no satisfying values, there is no fixpoint possible
--- we use the term \emph{annihilation} to describe the corresponding
-undefined state of the program.
-
-While the constructions in Sec \ref{sec:neg} apply to both the
-{{(0,+)}} and {{(1,*)}} monoids, each construction presented here
-applies to a specific monoid.
-
-\begin{enumerate}
-
-\item 
-\emph{Lifting negation out of {{*}}. }
+\paragraph*{Lifting negation out of {{*}}.}
+The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
+and {{b2}} are flowing opposite to each other then it doesn't matter which is
+direction is forwards and which is backwards.
 
@@
 
-\item
-\emph{Multiplying Negatives}
+\paragraph*{Multiplying Negatives.}
 
@@
 
-This isomorphism can be built using \emph{involution} constructed in
-Sec~\ref{sec:neg} and the isomorphisms above, and correspond to the
-algebraic manipulation:
+This isomorphism is a consequence of the fact that $-$ is an involution: it
+corresponds to the algebraic manipulation:
 
@@
 
-\item 
-\emph{Multiplying Fractions}
+\paragraph*{Multiplying and Adding Fractions.}
+An isomorphism witnessing:
 
@@
 
-\item 
-\emph{Adding the numerator}
+is straightforward. More surprisingly, it is also possible to construct
+isomorphisms witnessing:
 
@@
 
-\item 
-\emph{Fraction Addition}
-
 {{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
-
-\end{enumerate}
 
@@
 
+\amr{to merge}
+Annihilation and non-termination describe undefined program
+execution. While non-termination characterizes an iteration that fails
+to terminate, annihilation characterizes a constraint that has no
+fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
+annihilating only the program states that do not match the boolean
+satisfiability constraint.
+\emph{Annihilation.}  
+With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
+which gives us the ability to find the fixpoint of a constraint.  If
+the constraint has no satisfying values, there is no fixpoint possible
+-- we use the term \emph{annihilation} to describe the corresponding
+undefined state of the program.
+
 % Additive traces correspond to iteration and multiplicative traces
@@
 information to reconstruct the input from the output. This can be done by
-using the adjoint of {{f}}, denoted {{f^{dagger}}}. At this point we have,
-the top half of the construction below:
+using the adjoint of {{f}}. At this point we have, the top half of the
+construction below:
 

### 2012-03-11 rational.tex
all but conclusion

@@
 \end{itemize}
+
+\textbf{Note:} All the constructions, semantics, and examples in this paper
+have been implemented and tested in Haskell. We will make the URL available
+once the code is organized for better presentation.
 
@@
 ``wiring diagrams'' or ``circuits'' and that values are modeled as
-``particles'' that may appear on the wires. Evaluation therefore is modeled
-by the flow of particles along the wires.
+``particles'' or ``waves'' that may appear on the wires. Evaluation therefore
+is modeled by the flow of waves and particles along the wires.
 
@@
   case of products represented by a pair of wires, when tracing execution
-  using particles, one should think of one value/particle on each wire.
+  using particles, one should think of one particle on each wire or
+  alternatively as in folklore in the literature on monoidal categories as a
+  ``wave.''
 \begin{multicols}{2}
@@
 
-In this section we develop the semantics for {{langRev}} extended with
-negative and fractional types, {{langRevEE}}.  We start with the
-semantics for {{langRev}} defined in Sec \ref{sec:pi}, which we will
-systematically rewrite until we have the desired {{langRevEE}}
-sematics. Broadly this construction takes two steps
+In this section we develop an operational semantics for {{langRev}} extended
+with negative and fractional types, {{langRevEE}}. The operational semantics
+takes the abstract categorical diagrams which were previously known but gives
+them a computational interpretation based on reverse execution for negative
+types and unification for fractional types. Even though this computational
+interpretation appears straightforward in retrospect, it constitutes a
+breakthrough because it breaks the preconceived idea that there is only one
+notion of duality and hence that both negative and fractional types should be
+implemented using the same underlying mechanism. In some sense, the folkore
+literature on monoidal categories describing the evaluations for additive and
+multiplicative fragments are ``particle-style'' and ``wave-style'' can be
+seen to suggest either a unified or separate implementations. But, since de
+Morgan, we have been predisposed to think of only one notion of duality which
+persisted to even linear logic. (See Sec.~\ref{sec:related} for further
+discussion of duality.)
+
+For the purposes of this section, we start with the semantics for {{langRev}}
+defined in Sec~\ref{sec:pi}, and we will systematically rewrite it to achieve
+the desired {{langRevEE}} semantics. This rewriting consists of two main steps:
 
 \begin{enumerate}
-\item We rewrite the semantics in Sec \ref{sec:pi} using unification
-  of the values instead of direct pattern matching. This gives us the
-  neccessary infrastructure for fractional types, as we will see.
-
+\item We rewrite the semantics in Sec~\ref{sec:pi} using unification of the
+  values instead of direct pattern matching. This gives us the neccessary
+  infrastructure for fractional types. We assume the reader is familiar with
+  the idea of unification as realized using logic variables, substitutions,
+  and reification as presented in any standard text on logic programming.
 \item We write a reverse interpreter by reversing the direction of the
-  `{{|-->}}' reductions. This gives us the neccessary infrastructure for
+  {{|-->}} reductions. This gives us the neccessary infrastructure for
   negative types.
@@
 
-%%%%%%%%%%%%%%%%%%%
-\subsection{Unification and Logic Programming}
-
-\roshan{Can we cite away unification, reificatio, handling of
-  substitutions etc?}
-
 %%%%%%%%%%%%%%%%%%%%%
-\subsection{Extending {{langRev}} Semantics}
-
-\paragraph{Primitive Isomorphsims.}
-Previously the reductions of the primitive isomorphisms were specified
-in the following form.
+\subsection{Unification}
+
+Previously the reductions of the primitive isomorphisms were specified in the
+following form:
 
@@
 \noindent
-In the precense of unification, we rewrite them to accept the incoming
-substitution and extend it with the appropriate unification.
+Instead of relying on pattern-matching, we rewrite the rules to accept an
+incoming substitution to which the required pattern-matching rules are added
+as constraints. The general case is:
 
@@
 
-\roshan{Talk about non-determinism, determinism and our handling of
-  choices. Also, in what sense are we logically reversible?}
-
-\paragraph{Extending the evaulator with unification.} 
-We have rewriten the semantics of Sec. \ref{sec:pi} with
-unification. The behaviour of the intepreter has not changed.
+Using the same the entire semantics can be extended to thread the
+substitution as shown below:
 
@@
 
-\paragraph{Backward evaluator.}
-Here we have written a ``backward'' evaluator in the same style, 
-i.e., evaluation {{c v}} in this evaluator is equivalent to evaluating
-{{C^{dagger} v}} in the forward evaluator.
+Most of the rules look obviously correct but there is a subtle
+point. Consider the case for {{c1 (+) c2}}. Previously the shape of the value
+determined which of the combinators {{c1}} or {{c2}} to use. Now the incoming
+value could be a fresh logical variable, and indeed we have two rules with
+the same left hand side, and only depending on the success or failure of some
+further unification, can we decide which of them applies. The situation is
+common in logic programming in the sense that it is considered a
+non-deterministic process that keeps searching for the right substitution if
+any. Typical implementations of logic programming languages further provide
+top-level mechanisms to manipulate this non-deterministic search, for example
+by returning one answer and giving the user to ask for more answers. We
+abstract from this top-level semantics and view the rules above as being
+applied non-deterministically.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Reverse Execution}
+
+Given that our language is reversible (Proposition~\ref{prop:logrev}), a
+backward evaluator is relatively straightforward to implement: it simply
+corresponds to. Indeed using the backward evaluator to calculate{{c v}} is
+equivalent {{c^{dagger} v}} in the forward evaluator.
 
@@
 
-The previous section has defined all the pieces needed to define the
-semantics of {{langRevEE}} and we piece them together here.
+Combining the two constructions above, we get the semantics of the full
+{{langRevEE}} language by adding the rules for the two variants of~{{eta}}
+and~{{eps}}.
 
@@
 
+The additions formalize our previous discussions and should not be surprising
+at this point.
+
 %% %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\paragraph*{Zero Information.}
-As with {{eta+}}, we argue that {{eta*}} constructs no new
-information, instead it gives us the ability to explore as many worlds
-as permitted by the type {{b}}. So situate ourselves in any specific
-world, we have to supply additional information in the form of
-unifying the particle or anti-particle with a particle representing
-known information. Hence only by supplying information to a an
-entangled pair can we read information from it -- and always exactly
-as much information as we supplied.
-
-ADD MATH.
-
-\emph{Zero Information.}  The intuitive explanation that {{eta+}} and
-hence {{eps+}} introduces no information effects is that {{-b+b}} as a
-type whose value is observable only if you `pay' {{b}} amount of
-information to the {{-b}} branch. When {{b}} amount of information is
-supplied {{(-b+b)+b=b}} amount of information is returned to
-you. (TODO. more needs to be said here.)
-
-For this reason, our treatment of negative values in the context of
-{{langRev}} is particularly simple. We will have a type $0$ and an
-isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-allows us to create a value and a corresponding debt out of nothing. Both the
-value and its negative counterpart can flow anywhere in the system. Because
-information is preserved, a closed program (which does not have any
-``dangling'' negative values) will eventually have to match the negative
-value with some corresponding value, effectively using the isomorphism in the
-right-to-left direction.
+% \paragraph*{Zero Information.}
+% As with {{eta+}}, we argue that {{eta*}} constructs no new
+% information, instead it gives us the ability to explore as many worlds
+% as permitted by the type {{b}}. So situate ourselves in any specific
+% world, we have to supply additional information in the form of
+% unifying the particle or anti-particle with a particle representing
+% known information. Hence only by supplying information to a an
+% entangled pair can we read information from it -- and always exactly
+% as much information as we supplied.
+
+% ADD MATH.
+
+% \emph{Zero Information.}  The intuitive explanation that {{eta+}} and
+% hence {{eps+}} introduces no information effects is that {{-b+b}} as a
+% type whose value is observable only if you `pay' {{b}} amount of
+% information to the {{-b}} branch. When {{b}} amount of information is
+% supplied {{(-b+b)+b=b}} amount of information is returned to
+% you. (TODO. more needs to be said here.)
+
+% For this reason, our treatment of negative values in the context of
+% {{langRev}} is particularly simple. We will have a type $0$ and an
+% isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
+% allows us to create a value and a corresponding debt out of nothing. Both the
+% value and its negative counterpart can flow anywhere in the system. Because
+% information is preserved, a closed program (which does not have any
+% ``dangling'' negative values) will eventually have to match the negative
+% value with some corresponding value, effectively using the isomorphism in the
+% right-to-left direction.
 
 \paragraph*{Observability.}
-Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
-{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
-defined only if {{b1}} and {{b2}} are entirely positive types. This is
+
+The reductions above allow us to apply a program {{c : b1 <-> b2}} to an
+input {{v1 : b1}} to produce a result {{v2 : b2}} on termination. Execution
+is well defined only if {{b1}} and {{b2}} are entirely positive types. If
+either {{b1}} or {{b2}} is a negative or fractional type, the system has
+``dangling'' unsatisfied demands or constraints. For this reason, we
+constrain entire programs to have positive non-fractional types. This is
 similar to the constraint that Zeilberger imposes to explain intuitionistic
@@
 
-Consider the program that
-
-(TODO: So this is a really bad explanation. But this section is really
-important and we need to explain carefully. Maybe an analogy can be
-drawn to Cayley diagrams for groups to point out that the notion of
-`i' is entirely artificial, but it is essential to the discourse.)
-
-\begin{center}
-  \includegraphics{diagrams/shuffle.pdf}
-\end{center}
-
-We define observables to be only positive types. The outputs of
-programs that output mixed positive and negative types are not
-observable.  Also, programs that input mixed positive and negatives
-types are not executable.
-
-It is not fair to say that negative types flow backwards. The
-following circuits are valid in {{langRevEE}}. It is however proper to
-say that for any type {{b}} that flows in a direction, the type {{-b}}
-flows in the reverse direction.
-
-Several things we don't know how to do, or are sloppy about. Some of
-these are deep questions to which there is no immeidate answer.
-
-\begin{itemize}
-
-\item Is the language without recursive types always terminating? For
-  the right symmetry with annihilation, it is desirable to have
-  non-termination.
-
-\item What is the correct lemma for logical reversibility now?
-
-\item We should say things about information preservation. At least
-  more so than what is being said now.
-
-\item Multiplicative duality gives us a means of talking about
-  equivalence classes of values. This should correspond to all sorts
-  of other things -- groupoids, Lawvere theories etc. We barely even
-  understand these connections.
-
-\item How does this connect to QC? Is there any strong connection such
-  that the SAT solver here is an alternate QC algorithm to Shor's?
-
-\end{itemize}
+% (TODO: So this is a really bad explanation. But this section is really
+% important and we need to explain carefully. Maybe an analogy can be
+% drawn to Cayley diagrams for groups to point out that the notion of
+% `i' is entirely artificial, but it is essential to the discourse.)
+
+% \begin{center}
+%   \includegraphics{diagrams/shuffle.pdf}
+% \end{center}
+
+% We define observables to be only positive types. The outputs of
+% programs that output mixed positive and negative types are not
+% observable.  Also, programs that input mixed positive and negatives
+% types are not executable.
+
+% It is not fair to say that negative types flow backwards. The
+% following circuits are valid in {{langRevEE}}. It is however proper to
+% say that for any type {{b}} that flows in a direction, the type {{-b}}
+% flows in the reverse direction.
+
+% Several things we don't know how to do, or are sloppy about. Some of
+% these are deep questions to which there is no immeidate answer.
+
+% \begin{itemize}
+
+% \item Is the language without recursive types always terminating? For
+%   the right symmetry with annihilation, it is desirable to have
+%   non-termination.
+
+% \item What is the correct lemma for logical reversibility now?
+
+% \item We should say things about information preservation. At least
+%   more so than what is being said now.
+
+% \item Multiplicative duality gives us a means of talking about
+%   equivalence classes of values. This should correspond to all sorts
+%   of other things -- groupoids, Lawvere theories etc. We barely even
+%   understand these connections.
+
+% \item How does this connect to QC? Is there any strong connection such
+%   that the SAT solver here is an alternate QC algorithm to Shor's?
+
+% \end{itemize}
 
@@
 \label{sec:sat-solver}
-
-\amr{to merge}
-Annihilation and non-termination describe undefined program
-execution. While non-termination characterizes an iteration that fails
-to terminate, annihilation characterizes a constraint that has no
-fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
-annihilating only the program states that do not match the boolean
-satisfiability constraint.
-\emph{Annihilation.}  
-With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
-which gives us the ability to find the fixpoint of a constraint.  If
-the constraint has no satisfying values, there is no fixpoint possible
--- we use the term \emph{annihilation} to describe the corresponding
-undefined state of the program.
 
@@
 particular choice of {{c}} is pruned.
+
+% Annihilation and non-termination describe undefined program
+% execution. While non-termination characterizes an iteration that fails
+% to terminate, annihilation characterizes a constraint that has no
+% fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
+% annihilating only the program states that do not match the boolean
+% satisfiability constraint.
+% \emph{Annihilation.}  
+% With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
+% which gives us the ability to find the fixpoint of a constraint.  If
+% the constraint has no satisfying values, there is no fixpoint possible
+% -- we use the term \emph{annihilation} to describe the corresponding
+% undefined state of the program.
 

### 2012-03-11 rational.tex
done

@@
 and {{b2}} are flowing opposite to each other then it doesn't matter which is
-direction is forwards and which is backwards.
+direction is forwards and which is backwards. More interestingly as {{b1}} is
+moving backwards, it can ``see the past'' of {{b2}} which is equivalent to
+both particles moving backwards.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Future work and Conclusion}
+\section{Conclusion and Future Work}
 \label{sec:conc}
@@
 In another sense however, this paper is about the nature of duality in
-computation. The concept of duality is deep and significant and in
-many cases -- probably most famously in the divide between classical
-and intuinionistic mathematics -- contain non-trivial assumptions
-about our world-view.  The problem of duality in computation and
-logic, has many facets of which `continuation' and `higher order
-functions' are only one.  We have contributed an elegant and natural
-notion of duality that arises from the monoidal structure of sums and
-pairs. This notion of duality has a crisp semantics, clear
-computational interpretations and an information theoretic basis.
-Before we conclude, we briefly mention some closely related ideas that
-warrant further study.
+computation. 
+% The concept of duality is deep and significant and in many cases ---
+% probably most famously in the divide between classical and intuitionistic
+% mathematics --- contain non-trivial assumptions about our worldview.  The
+% problem of duality in computation and logic, has many facets of which
+% `continuation' and `higher order functions' are only one.
+The concept of duality is deep and significant: we have opened the door for
+us to consider, not one but two notions of duality. Surprisingly this makes
+things significantly simpler. In particular, instead of conflating
+considering pairs as dual to sums, the tradition in mathematics has long been
+to consider fields with two notions of duality: one for sums and one for
+pairs. This double notion of duality has a crisp semantics, clear
+computational interpretation, and an information theoretic basis.
+
+Our work has barely scratched the surface of an area of computing which has
+been explored in depth before but without the combined reversible
+information-preserving framework and the two notions of duality. The new
+insights point to further new areas of investigation of which we mention
+the three most significant ones (in our opinion).
 
@@
 
-\paragraph*{Geometry of Interaction.}
-GoI was developed by Girard \cite{girard1989geometry}as basis for the
-proof theory of Linear Logic.  GoI however has been developed into a
-reversible model of computing
-\cite{Mackie2011,DBLP:conf/popl/Mackie95,Abramsky:1994:NFG:184662.184664}
-in which values traveling along wires, much like {{langRevEE}}.
-Preliminary investigations suggest that many GoI machine constructions
-can be simulated in {{langRevEE}} by treating Mackie's bi-directional
-wires as pairs of wires in {{langRevEE}} and replacing the machine's
-global state with a typed value on the wire that captures the
-appropriate state.  However a concrete connection between GoI machines
-and GoI in general requires further work. This connection is exciting
-because when viewed through a Curry-Howard lens it suggests that the
-logical interpretation of {{langRevEE}} should a logic much like LL
-with a strong notion of resource preservation, but also with a strong
-computational basis.
+\paragraph*{Geometry of Interaction (GoI.}
+This geometry was developed by Girard~\cite{girard1989geometry} as part of
+the development of linear logic. It was given a computational interpretation
+by Abramsky and Jagadeesan~\cite{Abramsky:1994:NFG:184662.184664}, and was
+developed into a reversible model of computing by
+Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. Preliminary investigations
+suggest that many of the GoI machine constructions can be simulated in
+{{langRevEE}} by treating Mackie's bi-directional wires as pairs of wires in
+{{langRevEE}} and replacing the machine's global state with a typed value on
+the wire that captures the appropriate state. This connection is exciting
+because when viewed through a Curry-Howard lens it suggests that the logical
+interpretation of {{langRevEE}} would be a linear-like logic with a notion of
+resource preservation and with a natural computational interpretation.
 
@@
 \label{sec:algebraic-field}
-The field of algebraic numbers consist of the rationals extended with
-solutions to polynomials. In a limited sense {{langRevEE}} extended
-with recursive types already include polynomials such as the
-definition of binary tree {{@@x.(1+x*x)}} implies the equality
-{{x=1+x*x}} and hence {{x^2-x+1=0}}.
-
-The famous example of isomorphism on algebraics is the embedding of
-seven binary trees in one. In Figure 1 of Fiore's paper
-\cite{Fiore:2004}, he demonstrates this isomorphism by the
-corrsponding algebraic manipulation of the types.  The algebraic
-manipulation can be viewed as a {{langRevEE}} combinator and it indeed
-does constructively witness the isomorphism, i.e., you can run it and
-view one data structure encoded by the other. In this context,
-consider the following algebraically valid proof of the same
-isomorphism:
+Algebraically, the move from {{langRev}} to {{langRevEE}} corresponds to a
+move from a ring-like structure to a full field. Our language {{langRevEE}}
+captures the structure of one particular field: that of the rational
+numbers. As we have seen, computation in this field is quite expressive and
+interesting and yet, it has two fundamental limitations. First it cannot
+express any recursive type, and second it cannot express any datatype
+definitions. We believe these to be two orthogonal extensions: recursive
+types were considered in our previous paper~\cite{infeffects}; arbitrary
+datatypes are however even more exciting that plain rationals as each
+datatype definition can be viewed as a polynomial (see below) which
+essentially means that we start computing in the field of algebraic numbers,
+which includes square roots and imaginary numbers. As crazy as it might seem,
+the type $\sqrt{2}$ and even the type $(1/2)+i(\sqrt{3}/2)$ ``make sense.''
+In fact the latter type is the solution to the polynomial {{x^2-x+1=0}} which
+if re-arranged looks like {{x=1+x*x}} and perhaps more familiarly as the
+datatype of binary trees {{@@x.(1+x*x)}}. These types happen to have been
+studied extensively following a paper by Blass~\cite{seventrees} which used
+the above datatype of trees to infer an isomorphism between seven binary
+trees and one!
+
+We have confirmed that we can extended {{langRevEE}} with the datatype
+declaration for binary trees and build a witness for this isomorphism that
+works as expected. To understand the issue in more detail, consider the
+following algebraically valid proof of the isomorphism in question:
 \[\begin{array}{rclclclcl}
@@
 \end{array}\]
-Fiore poses the question of why such algebraic manipulation would make
-sense type theoretically, even though some of the intermediate steps
-make no sense.  Stated in terms of {{langRevEE}} computations, this
-manipulation has no meaningful computational content -- it is always
-non-terminating. It suggests that some algebraic manipulations are
-somehow more `constructive' than others or dually that {{langRevEE}}
-with recursive types somehow lacks the ability to express all
-computations that are otherwise algebraically meaningful.
-
-\begin{itemize}
-\item Maybe the problem is that there is cancellation in the second
-  case and cancellation with recursive types is problematic. However
-  not all cancellation is. So what gives?
-
-\item One idea is that irrationals correspond to an infinite amount of
-  computational work. If what we are cancelling is not the same
-  infinity, then things go wrong.
-
-\item Chaitin has already shown that reals have inifite amount of
-  information which is paradoxical and problematic. This is a
-  different issue from that of the rational being infinite
-  computations.
-
-\item Finally, not all meaningful types are meaningful polynomials --
-  ex. nat. So this is not the whole story.
-
-\end{itemize}
-
-\paragraph*{Quantum Computing} 
-A \emph{time traveling} intuition is also applicable here.  In a
-normal cicruit, as computational steps are taken values flow from left
-to right, i.e. computational time progresses from left to right.  The
-action of {{eps+}} causes values, aka information particles, to flow
-from right to left i.e. backwards in computational time.  The
-interesting thing about this interpretation is that particles that
-travel backwards in time get to see and interact with the history of
-particles that they coexist with. This gives us an intuitive
-interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
-Sec. \ref{sec:specific-constructions}) where it would otherwise seem
-that {{-b1}} and {{b2}} move in opposite directions. The backward flow
-of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
-equivalent to both particles moving backward in time.
-
-An analogy inspired by quantum mechanics is also applicable here.  The
-operator {{eta*}} is a site of fission that creates \emph{an entangled
-  particle and anti-particle}. These particles can be thought to be in
-a \emph{superposition} of states determined by their type. Both flow
-in the same direction in time.  In each possible world that the
-particle exists, it takes on a specific value inhibiting its type and
-correspondingly its entangled anti-particle takes the dual of the
-specific value. Since the particles are entangled, actions on one
-(such as transformation by application of an isomorphism) affects the
-other in much the same sense as \emph{action at a distance}. The
-`{{a-o*b}}' functions mentioned in the introduction, are first-class
-values and they correspond to these entagled pairs. Finally {{eps*}}
-operations are sites where corresponding particle and anti-particles
-annihilate each other. While appealing as an analogy, this description
-does not imply (or preclude) any formal connection with Physics.
-
-Are constructions are very similar to those developed in the context
-of QC. Is this how nature computes? Are the analogies to time
-traveling particles and anti-particles more concrete in some way?
+The question is why such an algebraic manipulation makes sense type
+theoretically, even though the intermediate step asking for an isomorphism
+between $x^6$ and $1$ has no computational context. In the setting of
+{{langRevEE}}, this isomorphism can be constructed but it diverges on all
+inputs (in both ways). This suggests that, in the field of algebraic numbers,
+some algebraic manipulations are somehow 
+``more constructive'' than others.
+%% or dually that {{langRevEE}} with recursive types
+%%somehow lacks the ability to express all computations that are otherwise
+%%algebraically meaningful.
+
+% \begin{itemize}
+% \item Maybe the problem is that there is cancellation in the second
+%   case and cancellation with recursive types is problematic. However
+%   not all cancellation is. So what gives?
+
+% \item One idea is that irrationals correspond to an infinite amount of
+%   computational work. If what we are cancelling is not the same
+%   infinity, then things go wrong.
+
+% \item Chaitin has already shown that reals have inifite amount of
+%   information which is paradoxical and problematic. This is a
+%   different issue from that of the rational being infinite
+%   computations.
+
+% \item Finally, not all meaningful types are meaningful polynomials --
+%   ex. nat. So this is not the whole story.
+
+% \end{itemize}
+
+\paragraph*{Quantum Computing.} 
+One understand of quantum computing is that it exploits the laws of physics
+to build faster machines (perhaps). Another more foundational understanding
+is that it provides a computational interpretation of physics, and in
+particular directly address the question of interpretation of quantum
+mechanics. In a little known document, Rozas~\cite{Rozas:1987:CMO:889539}
+uses continuations to implement the transactional interpretation of quantum
+mechanics~\cite{transactional} which includes as its main ingredient a
+fixpoint calculation between waves or particles traveling forwards and
+backwards in time. Our work shed no light on whether this interpretation is
+the ``right one'' but it is interesting that we can directly realize it using
+the primitives of {{langRevEE}}.
+
+% A \emph{time
+%   traveling} intuition is also applicable here.  In a normal cicruit, as
+% computational steps are taken values flow from left to right,
+% i.e. computational time progresses from left to right.  The action of
+% {{eps+}} causes values, aka information particles, to flow from right to left
+% i.e. backwards in computational time.  The interesting thing about this
+% interpretation is that particles that travel backwards in time get to see and
+% interact with the history of particles that they coexist with. This gives us
+% an intuitive interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
+% Sec. \ref{sec:specific-constructions}) where it would otherwise seem that
+% {{-b1}} and {{b2}} move in opposite directions. The backward flow of the
+% {{-b1}} allows it to `see the past' of {{b2}} and is thus equivalent to both
+% particles moving backward in time.
+
+The multiplicative structure of {{langRevEE}} also has a direct connection to
+entangled quantum particles, or perhaps entangled particles and
+anti-particles.  The idea of entanglement is one action on one particle is
+``instanteneously'' communicated to the other. Again our model shed no light
+on whether this is related to how nature computes but it is again interesting
+that we can directly realize the idea using the primitives of {{langRevEE}}.
+
+% natural An analogy
+% inspired by quantum mechanics is also applicable here.  The operator {{eta*}}
+% is a site of fission that creates \emph{an entangled particle and
+%   anti-particle}. These particles can be thought to be in a
+% \emph{superposition} of states determined by their type. Both flow in the
+% same direction in time.  In each possible world that the particle exists, it
+% takes on a specific value inhibiting its type and correspondingly its
+% entangled anti-particle takes the dual of the specific value. Since the
+% particles are entangled, actions on one (such as transformation by
+% application of an isomorphism) affects the other in much the same sense as
+% \emph{action at a distance}. The `{{a-o*b}}' functions mentioned in the
+% introduction, are first-class values and they correspond to these entagled
+% pairs. Finally {{eps*}} operations are sites where corresponding particle and
+% anti-particles annihilate each other. While appealing as an analogy, this
+% description does not imply (or preclude) any formal connection with Physics.
+
+% Are constructions are very similar to those developed in the context
+% of QC. Is this how nature computes? Are the analogies to time
+% traveling particles and anti-particles more concrete in some way?
 
@@
 
-%% Jacques Carette, Michael Adams, Khaled, Will Byrd, Yin Wang
-
-\acks This project was partially funded by Indiana University's Office
-of the Vice President for Research and the Office of the Vice Provost
-for Research through its Faculty Research Support Program.  We also
-acknowledge support from Indiana University's Institute for Advanced
-Study.
+\acks We thank Jacques Carette for stimulating discussion, and Michael Adams,
+Will Byrd, Lindsey Kuper, and Yin Wang for helpful comments and
+questions. This project was partially funded by Indiana University's Office
+of the Vice President for Research and the Office of the Vice Provost for
+Research through its Faculty Research Support Program.  We also acknowledge
+support from Indiana University's Institute for Advanced Study.
 

### 2012-03-11 rational.tex
functions arrows

@@
 \begin{center}
-\scalebox{0.8}{
   \includegraphics{diagrams/eta.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.8}{
+\end{center}
+
+\begin{center}
  \includegraphics{diagrams/eps.pdf}
-}
 \end{center}
@@
 \begin{center}
-\scalebox{1.1}{
+\scalebox{1.5}{
   \includegraphics{diagrams/eps_plus1.pdf}
@@
 \begin{center}
-\scalebox{1.1}{
+\scalebox{1.5}{
   \includegraphics{diagrams/eps_plus2.pdf}
@@
 
-For a monoidal category to be compact closed, the maps~{{eta}} and~{{eps}}
-must satisfy a coherence condition that is usually visualized as follows:
+For a monoidal category to be compact closed with a dual objects
+presented as $b^*$, the maps~{{eta}} and~{{eps}} must satisfy a
+coherence condition that is usually visualized as follows:
 \begin{center}
@@
 \end{center}
-In the case of negatives, the condition amounts to checking that reversing
-direction twice is a no-op. In the case of fractionals, the condition amounts
-to checking that creating values {{alpha}} and {{1/alpha}} and immediately
-unifying them is also a no-op. Both checks are straightforward.
+In the case of negatives, the condition amounts to checking that
+reversing direction twice is a no-op. In the case of fractionals, the
+condition amounts to checking that creating values {{alpha}} and
+{{1/alpha}} and immediately unifying them is also a no-op. Both checks
+are straightforward and are essentially the constructions in the
+introduction.
 
@@
 
-In a compact closed category, each morphism {{f : a <-> b }} can be given a
+In a compact closed category, each morphism {{f : b1 <-> b2 }} can be given a
 \emph{name} and a \emph{coname}. For the additive fragment, the name
-$\fname{f}$ has type {{0 <-> (-a + b)}} and the coname $\fconame{f}$ has type
-{{a + (-b) <-> 0}}. For the multiplicative fragment, the name $\fname{f}$ has
-type {{1 <-> ((1/a) * b))}} and the coname $\fconame{f}$ has type 
-{{(a * (1/b)) <-> 1}}. Intuitively, this means that for each morphism, 
+$\fname{f}$ has type {{0 <-> (-b1 + b2)}} and the coname $\fconame{f}$ has type
+{{b1 + (-b2) <-> 0}}. For the multiplicative fragment, the name $\fname{f}$ has
+type {{1 <-> ((1/b1) * b2))}} and the coname $\fconame{f}$ has type 
+{{(b1 * (1/b2)) <-> 1}}. Intuitively, this means that for each morphism, 
 it is possible to construct, from ``nothing,'' an object in the category that 
@@
 
+\begin{multicols}{2}
 \begin{center}
@@
 \end{center}
+
 \begin{center}
@@
 \end{center}  
+\end{multicols}
 
@@
   from the input {{()}} the value {{((),())}} of type {{1 * 1}};
-\item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of
+\item Apply the isomorphism {{1 <-> (1/b) * b}} in parallel to each of
   the components of the above tuple. The resulting value is
@@
 
-At this point we have constructed a strange mix of two {{a-o*a}} functions;
+At this point we have constructed a strange mix of two {{b-o*b}} functions;
 inputs of one function manifest themselves as outputs of the other. If
@@
 {{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
-be {{R}} in the other. The type {{a}} thus takes the role of the type of the
+be {{R}} in the other. The type {{b}} thus takes the role of the type of the
 communication channel, indicating how much information can be communicated
-between the two subcomputations.  Depending on the choice of the type {{a}},
+between the two subcomputations.  Depending on the choice of the type {{b}},
 an arbitrary number of bits may be communicated.
@@
 Dually, the additive reading of the above manipulations correspond to
-functions of the form {{a-o+a}}, witnessing isomorphisms of the form
-{{0<->(-a)+a}}). The remixed additive functions express control flow transfer
+functions of the form {{b-o+b}}, witnessing isomorphisms of the form
+{{0<->(-b)+b}}). The remixed additive functions express control flow transfer
 between two subcomputations, \emph{only one of which exists} at any point,

### 2012-03-11 rational.tex
diagrams

@@
 \begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/distrib-crop.pdf]
   \includegraphics{diagrams/thesis/dist.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/factor-crop.pdf]
+\end{center}
+\begin{center}
   \includegraphics{diagrams/thesis/factor.pdf}
-}
+\end{center}
+\end{multicols}
+
+Distributivity and factoring are interesting because they represent
+interactions between sum and pair types. Distributivity should
+essentially be thought of as a multiplexer that redirects the flow of
+{{v:b}} depending on what value inhabits the type {{b1+b2}}, as shown
+below. Factoring is the corresponding adjoint operation.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
 \end{center}

### 2012-03-11 rational.tex
_
diagrams

@@
 Every compact closed categoy admits a trace. For the additive case, we get
-the following definition.  Given {{f : a+c <-> b+c }}, 
-define {{trace+ f : a <-> b}} as:
+the following definition.  Given {{f : b1+b2 <-> b1+b3 }}, 
+define {{trace+ f : b2 <-> b3}} as:
 
@@
 
+\begin{center}
+  \includegraphics{diagrams/thesis/trace_plus.pdf}
+\end{center}
+
 \noindent We have omitted some the commutativity and associativity shuffling to
-communicate the main idea. We are given a value of type {{a}} which we embed
-into {{a+0}} and then {{a + (c - c)}}. This can be reassociated into 
-{{(a+c) - c}}. The component {{a+c}} which until now is just an appropriately 
-tagged value of type {{a}} can be given to~{{f}} which transforms it to
-something of type {{b+c}}. If the result is in the {{b}}-summand, it is produced
-as the answer; otherwise the result is in the {{c}}-summand; {{eps+}} is used to
+communicate the main idea. We are given a value of type {{b2}} which we embed
+into {{0+b2}} and then {{(-b1+b1)+b2)}}. This can be reassociated into 
+{{-b1+(b1+b2)}}. The component {{b1+b2}} which until now is just an appropriately 
+tagged value of type {{b2}} can be given to~{{f}} which transforms it to
+something of type {{b1+b3}}. If the result is in the {{b3}}-summand, it is produced
+as the answer; otherwise the result is in the {{b1}}-summand; {{eps+}} is used to
 make it flow backwards to be fed to the {{eta+}} located at the beginning of the 
-sequence. Iteration continues until a {{b}} is produced.
+sequence. Iteration continues until a {{b3}} is produced.
 
@@
 backward evaluator is relatively straightforward to implement: it simply
-corresponds to. Indeed using the backward evaluator to calculate{{c v}} is
+corresponds to. Indeed using the backward evaluator to calculate {{c v}} is
 equivalent {{c^{dagger} v}} in the forward evaluator.

### 2012-03-11 rational.tex
_

@@
 
-\paragraph*{Geometry of Interaction (GoI.}
-This geometry was developed by Girard~\cite{girard1989geometry} as part of
+\paragraph*{Geometry of Interaction (GoI).}
+Geometry of Interaction was developed by Girard~\cite{girard1989geometry} as part of
 the development of linear logic. It was given a computational interpretation

### 2012-03-12 rational.tex
diagrams
_
fixes

@@
 \includegraphics{diagrams/thesis/swap_plus.pdf}
+}
+\end{center}
+\end{multicols}
+
+By visually tracking the flow of particles on the wires, one can
+verify that the expected types for commutativity are satisfied.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_times_value.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_plus_value.pdf}
 }
@@
 
-\item {{eps* s v |--> (1, s[v <><> (inv(v'),v') ])}}
+\item {{eps* s v |--> (1, s[v <><> (1/v',v') ])}}
   where {{v'}} is a fresh logic variable.
@@
 %! bck = \triangleleft
-% <eps+, v, C, s>_{fwd} |--> <eps+, left (-v'), C, s[v <><> right v]>_{bck}
+% <eps+, v, C, s>_{fwd} |--> <eps+, left (-v'), C, s[v <><> right v']>_{bck}
 % <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}
@@
 %! bck = \triangleleft
-% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v]>_{fwd}
+% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v']>_{fwd}
 % <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}

### 2012-03-12 rational.tex
constructions

@@
 
+\paragraph*{Duality preserves the monoidal tensor. }
+As with compact closed categories, the dual on the objects distributes
+over the tensor. In terms of {{langRevEE}} we have that {{-(b1+b2)}}
+can be mapped to {{(-b1)+(-b2)}} and that {{1/(b1*b2)}} can be mapped
+to {{(1/b1)*(1/b2)}}. The isomorphism {{-(b1 + b2) <-> (-b1) + (-b2)}}
+can be realized as follows:
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+The multiplicative construction is similar. This is a distinguishing
+difference from *-autonomous categories that are models for Linear
+Logic, where the dual does not preserve the tensor but maps to a dual
+tensor \cite{curien2009interactive}.
+
+\paragraph*{Duality is a functor.}
+Duality in {{langRevEE}} can map objects to their duals and morphisms
+to act on dual objects. In other {{c : b1 <-> b2}} to
+{{neg~c:-b1<->-b2}} in the additive monoid and to
+{{inv~c:1/b1<->1/b2}} in the multiplicative monoid. 
+%% Any operation on types can be applied to the negative versions of these
+%% types. 
+The idea is simply to reverse the flow of values and use the
+adjoint of the operation:
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
+%subcode{proof} include main
+%@ c : b1 <-> b2
+%@@ neg c : (-b1) <-> (-b2)
+\end{multicols}
+
+We are able to do this because in {{langRevEE}} every morphism is an
+isomorphism and hence has an inverse morphism, whereas in compact
+closed categories with no underlying dagger structure we can only
+construct an op-functor.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 novel.
-
-\paragraph*{Negation distributes over {{+}}. }
-One certainly expects that:
-{{-(b1 + b2) <-> (-b1) + (-b2)}}
-which can be realized as follows:
-\begin{center}
-  \includegraphics{diagrams/dist_neg_plus.pdf}
-\end{center}
-
-\paragraph*{Lifting an operation of positive types to negated types.}
-Any operation on types can be applied to the negative versions of these
-types. The idea is simply to reverse the flow of values and use the
-adjoint of the operation:
-\begin{multicols}{2}
-Given {{c : b1 <-> b2}}
-
-\begin{center}
-  \includegraphics{diagrams/neg_lift.pdf}
-\end{center}
-
-%subcode{proof} include main
-%@ c : b1 <-> b2
-%@@ neg c : (-b1) <-> (-b2)
-\end{multicols}
 

### 2012-03-12 rational.tex
_
explain the semantics

@@
 
-We are able to do this because in {{langRevEE}} every morphism is an
-isomorphism and hence has an inverse morphism, whereas in compact
-closed categories with no underlying dagger structure we can only
-construct an op-functor.
+The {{inv}} construction is similar.  We are able to do this because
+in {{langRevEE}} every morphism is an isomorphism and hence has an
+inverse morphism, whereas in compact closed categories with no
+underlying dagger structure we can only construct an op-functor.
 
@@
 
-%subcode{opsem} include main
-%! columnStyle = rc
-%   & b1 * b2
-% = & -(-(b1*b2)) 
-% = & -((-b1)*b2)
-% = & (-b1)*(-b2)
+%% %subcode{opsem} include main
+%% %! columnStyle = rc
+%% %   & b1 * b2
+%% % = & -(-(b1*b2)) 
+%% % = & -((-b1)*b2)
+%% % = & (-b1)*(-b2)
+
+{{b1 * b2 = -(-(b1*b2)) = -((-b1)*b2) = (-b1)*(-b2)}}
 
@@
 
+As explained previously {{eta*}} creates two values {{1/v}} and {{1/v}}
+from a single logic variable.
+
 \item {{eps* s v |--> (1, s[v <><> (1/v',v') ])}}
@@
 
+Similarly, {{eps*}} unifies the values on incoming wires. The incoming
+value {{v}} represents the values of both wires and hence unifying
+them is accomplished in terms of an intermediate logic variable
+{{v'}}.
+
 \end{enumerate}
@@
 \item To add negative types we add the following rules to the
-  reductions above:
+  reductions above. The additions formalize our previous discussions
+  and should not be surprising at this point.
 
   \begin{enumerate}
-  \item
-
+  \item The rules for {{eps+}} essentially transfer control from the
+    forward evaluator (whose states are tagged by $\triangleright$) to
+    the backward evaluator (whose states are tagged by
+    $\triangleleft$). In other words, after an {{eps+}} the direction
+    of the world is reversed. The pattern matching done by the
+    unification ensures that a value on the {{right}} wire is tagged
+    to be negative and transfered to the {{left}} wire, and vice versa.
 %subcode{opsem} include main
@@
 % <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}
-
-    \item
+   %
+    Note that there is no evaluation rule for {{eta+}} in the forward
+    evaluator. This corresponds to the fact that there is no value of
+    type {{0}} and hence the forward evaluator can never execute an
+    {{eta+}}.
+
+    \item The rules for {{eta+}} are added to backward evaluator. A
+      program executing backward starts executing forward after the
+      execution of the {{eta+}}. Dual to the previous case, there is
+      no rule for {{eps+}} in the backward evaluator since the output
+      type of {{eps+}} is {{0}}.
 %subcode{opsem} include main
@@
 \end{enumerate}
-
-The additions formalize our previous discussions and should not be surprising
-at this point.
 

### 2012-03-12 rational.tex
whatever

@@
 
-Applying a function consists of making it flow backwards to satisfy the demand 
-of the function:
+Applying a function to an argument consists of making the argument flow
+backwards to satisfy the demand of the function:
 \begin{multicols}{2}
@@
 top-level mechanisms to manipulate this non-deterministic search, for example
-by returning one answer and giving the user to ask for more answers. We
-abstract from this top-level semantics and view the rules above as being
-applied non-deterministically.
+by returning one answer and giving the user the option to ask for more
+answers. We abstract from this top-level semantics and view the rules above
+as being applied non-deterministically.
 
@@
 
-Given that our language is reversible (Proposition~\ref{prop:logrev}), a
-backward evaluator is relatively straightforward to implement: it simply
-corresponds to. Indeed using the backward evaluator to calculate {{c v}} is
-equivalent {{c^{dagger} v}} in the forward evaluator.
+Given that our language is reversible (Prop.~\ref{prop:logrev}), a backward
+evaluator is relatively straightforward to implement: using the backward
+evaluator to calculate {{c v}} is equivalent {{c^{dagger} v}} in the forward
+evaluator.
 
@@
 
-    \item The rules for {{eta+}} are added to backward evaluator. A
-      program executing backward starts executing forward after the
+    \item The rules for {{eta+}} are added to the backward evaluator. A
+      program executing backward starts executing forwards after the
       execution of the {{eta+}}. Dual to the previous case, there is
@@
 fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
-detail, an instance of SAT is a function/circuit {{f}} that given some
+detail, an instance of SAT is a function/circuit~{{f}} that given some
 boolean inputs returns {{true}} or {{false}} which we interpret as whether

### 2012-03-12 rational.tex

@@
 have typically appeared in the context of conventional languages with
-information effects, which obscured their appeal and murked their properties.
+information effects, which limited their appeal and obscured their properties.
 
@@
   types. (Sec.~\ref{sec:frac})
-\item We combine the above two extensions into a language whose type system
-  allows any rational number to be used as a type. Moreover the types satisfy
-  the same familiar and intuitive isomorphisms that are satisfied in the
-  mathematical field of rational numbers. (Sec.~\ref{sec:rat})
+\item We combine the above two extensions into a language, which we
+  call {{langRevEE}}, whose type system allows any rational number to
+  be used as a type. Moreover the types satisfy the same familiar and
+  intuitive isomorphisms that are satisfied in the mathematical field
+  of rational numbers. (Sec.~\ref{sec:rat})
 \item We develop programming intuition and argue that negative and fractional
@@
 
-\item Sum types may similarly be represented by one wire or using using
+\item Sum types may similarly be represented by one wire or using
   parallel wires with a {{+}} operator between them. When tracing the
@@
   \emph{compact closed} structure by adding a dual for each object and two
-  special morphims traditionally called {{eta}} and {{eps}}. 
+  special morphisms traditionally called {{eta}} and {{eps}}. 
 \end{itemize}
@@
 
-For convenience we sometime use the notations {{b1 - b2}} and {{b1/b2}} to
+For convenience, we sometimes use the notations {{b1 - b2}} and {{b1/b2}} to
 indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}} respectively.  The
@@
 
-For a monoidal category to be compact closed with a dual objects
-presented as $b^*$, the maps~{{eta}} and~{{eps}} must satisfy a
-coherence condition that is usually visualized as follows:
+For a monoidal category to be compact closed the maps~{{eta}}
+and~{{eps}} must satisfy a coherence condition that is usually
+visualized as follows:
 \begin{center}
@@
 \end{center}
-In the case of negatives, the condition amounts to checking that
-reversing direction twice is a no-op. In the case of fractionals, the
-condition amounts to checking that creating values {{alpha}} and
-{{1/alpha}} and immediately unifying them is also a no-op. Both checks
-are straightforward and are essentially the constructions in the
-introduction.
+where $b^*$ represents the dual of {{b}}.  In the case of negatives,
+the condition amounts to checking that reversing direction twice is a
+no-op. In the case of fractionals, the condition amounts to checking
+that creating values {{alpha}} and {{1/alpha}} and immediately
+unifying them is also a no-op. Both checks are straightforward and are
+essentially the constructions in the introduction.
 
@@
 \paragraph*{Trace.}
-Every compact closed categoy admits a trace. For the additive case, we get
+Every compact closed category admits a trace. For the additive case, we get
 the following definition.  Given {{f : b1+b2 <-> b1+b3 }}, 
@@
 
-\noindent We have omitted some the commutativity and associativity shuffling to
-communicate the main idea. We are given a value of type {{b2}} which we embed
-into {{0+b2}} and then {{(-b1+b1)+b2)}}. This can be reassociated into 
-{{-b1+(b1+b2)}}. The component {{b1+b2}} which until now is just an appropriately 
-tagged value of type {{b2}} can be given to~{{f}} which transforms it to
-something of type {{b1+b3}}. If the result is in the {{b3}}-summand, it is produced
-as the answer; otherwise the result is in the {{b1}}-summand; {{eps+}} is used to
-make it flow backwards to be fed to the {{eta+}} located at the beginning of the 
-sequence. Iteration continues until a {{b3}} is produced.
+\noindent We have omitted some of the commutativity and associativity
+shuffling to communicate the main idea. We are given a value of type
+{{b2}} which we embed into {{0+b2}} and then {{(-b1+b1)+b2)}}. This
+can be re-associated into {{-b1+(b1+b2)}}. The component {{b1+b2}},
+which until now is just an appropriately tagged value of type {{b2}},
+is transformed to a value of type {{b1+b3}} by {{f}}. 
+If the result is in the {{b3}}-summand, it is produced as
+the answer; otherwise the result is in the {{b1}}-summand; {{eps+}} is
+used to make it flow backwards to be fed to the {{eta+}} located at
+the beginning of the sequence. Iteration continues until a {{b3}} is
+produced.
 
@@
 
-The multiplicative construction is similar. This is a distinguishing
-difference from *-autonomous categories that are models for Linear
-Logic, where the dual does not preserve the tensor but maps to a dual
-tensor \cite{curien2009interactive}.
+\noindent
+The multiplicative construction is similar.
+
+%% This is a distinguishing difference from *-autonomous categories that
+%% are models for Linear Logic, where the dual does not preserve the
+%% tensor but maps to a dual tensor \cite{curien2009interactive}.
 
@@
 
-The {{inv}} construction is similar.  We are able to do this because
-in {{langRevEE}} every morphism is an isomorphism and hence has an
-inverse morphism, whereas in compact closed categories with no
-underlying dagger structure we can only construct an op-functor.
+This construction relies on the fact that every {{langRevEE}} morphism
+has an adjoint.  The {{inv}} construction is similar.
+
+%% inverse morphism, whereas in compact closed categories with no
+%% underlying dagger structure we can only construct an op-functor.
 
@@
 The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
-and {{b2}} are flowing opposite to each other then it doesn't matter which is
+and {{b2}} are flowing opposite to each other then it doesn't matter which
 direction is forwards and which is backwards. More interestingly as {{b1}} is
@@
 
-To build these isomorphisms, we first build an intermeidate
+To build these isomorphisms, we first build an intermediate
 construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
@@
 notion of duality and hence that both negative and fractional types should be
-implemented using the same underlying mechanism. In some sense, the folkore
+implemented using the same underlying mechanism. In some sense, the folklore
 literature on monoidal categories describing the evaluations for additive and
@@
 \item We rewrite the semantics in Sec~\ref{sec:pi} using unification of the
-  values instead of direct pattern matching. This gives us the neccessary
+  values instead of direct pattern matching. This gives us the necessary
   infrastructure for fractional types. We assume the reader is familiar with
@@
 \item We write a reverse interpreter by reversing the direction of the
-  {{|-->}} reductions. This gives us the neccessary infrastructure for
+  {{|-->}} reductions. This gives us the necessary infrastructure for
   negative types.
@@
 
-Using the same the entire semantics can be extended to thread the
-substitution as shown below:
+Using the same idea, the entire semantics can be extended to thread
+the substitution as shown below:
 
@@
     unification ensures that a value on the {{right}} wire is tagged
-    to be negative and transfered to the {{left}} wire, and vice versa.
+    to be negative and transferred to the {{left}} wire, and vice versa.
 %subcode{opsem} include main
@@
     \item The rules for {{eta+}} are added to the backward evaluator. A
-      program executing backward starts executing forwards after the
+      program executing backwards starts executing forwards after the
       execution of the {{eta+}}. Dual to the previous case, there is
@@
 
-This construction can be used to solve a large class of constraint
-satisfaction problems. We illustrate the main ideas with the implementation
-of a SAT-solver. Although the main idea is the use of {{trace*}}, the
-construction is involved so we proceed in small steps, reviewing some of the
-necessary constructions presented in our earlier paper~\cite{infeffects}.
+A large class of constraint satisfaction problems can be expressed
+using {{trace*}}. We illustrate the main ideas with the implementation
+of a SAT-solver.  We proceed in small steps, reviewing some of the
+necessary constructions presented in our earlier
+paper~\cite{infeffects}.
 
@@
 equivalent construction that is useful is {{else_{not} }} where we
-negate the second argument only is the first is {{false}}.
+negate the second argument only if the first is {{false}}.
 
@@
 possible at the cost of having additional constant inputs. For example,
-consider the gate {{else_{not} }}. Generally, the gate gates maps
+consider the gate {{else_{not} }}. Generally, the gate maps
 {{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
@@
 With a little work, we can modify this circuit to only annihilate values that
-fail to satisfy the constraints reprsented by a SAT-instance {{f}}. In more
+fail to satisfy the constraints represented by a SAT-instance {{f}}. In more
 detail, an instance of SAT is a function/circuit~{{f}} that given some
@@
 Any boolean expression {{f : bool^n -> bool}} can be compiled into the
-isomorphism {{iso : bool^h*bool^n<->bool^g*bool}} where the extra bits
+isomorphism {{iso_f : bool^h*bool^n<->bool^g*bool}} where the extra bits
 {{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
@@
 before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
-applying {{iso}} to some special heap values and an input \textit{bs}
+applying {{iso_f}} to some special heap values and an input \textit{bs}
 produces some bits that can be ignored and the same output that {{f}} would
@@
 
-Let us call the above construction which maps inputs, heap, and control wires
-to input, heap, and control wires as {{sat_f}}. The SAT-solver is is
-completed by tracing the {{sat_f}} and cloning the inputs using
-{{clone^n_{bool} }}.
+Let us call the above construction which maps inputs, heap, and
+control wires to inputs, heap, and control wires as {{sat_f}}. The
+SAT-solver is is completed by tracing the {{sat_f}} and cloning the
+inputs using {{clone^n_{bool} }}.
 
@@
 
-When the solver is fed inputs initialized to {{true}}, the {{clone}} will
-clone any inputs to {{sat_f}} that satisfy the constraints imposed by {{f}}
-and the constraints on the {{heap}}. In the case of unique-SAT the solver
-will produce exactly 0 or 1 solutions. In the case of general SAT, the solver
-will produce solutions as determined by the semantics of the top-level
-interaction as discussed in Sec. \ref{sec:rat}.
+When the solver is fed inputs initialized to {{true}}, it clones only
+those inputs to {{sat_f}} that satisfy {{f}} and the {{heap}}
+constraints. In the case of unique-SAT the solver will produce exactly
+0 or 1 solutions. In the case of general SAT, the solver will produce
+solutions as determined by the semantics of the top-level interaction
+(see discussion in Sec. \ref{sec:rat}).
 
@@
 fractional types is somewhat related to our work, our results are
-subtantially different. To clarify this point, we start by reviewing the
+substantially different. To clarify this point, we start by reviewing the
 salient point of the major pieces of related work and conclude this section
@@
 interpretation to the subtraction connective and like the original symmetric
-calculus of Filinksi, introduces a duality that relates sums to products and
+calculus of Filinski, introduces a duality that relates sums to products and
 vice-versa.
@@
 in the commutative semiring of whole numbers to {{langRevEE}} that
-expresses computation in the fields of rationals.  Every algebraic
+expresses computation in the field of rationals.  Every algebraic
 identity that holds for the rational numbers corresponds to a type
@@
 % `continuation' and `higher order functions' are only one.
-The concept of duality is deep and significant: we have opened the door for
-us to consider, not one but two notions of duality. Surprisingly this makes
-things significantly simpler. In particular, instead of conflating
-considering pairs as dual to sums, the tradition in mathematics has long been
-to consider fields with two notions of duality: one for sums and one for
-pairs. This double notion of duality has a crisp semantics, clear
-computational interpretation, and an information theoretic basis.
+The concept of duality is deep and significant: we have opened the
+door for us to consider, not one but two notions of
+duality. Surprisingly this makes things substantially simpler. In
+particular, instead of conflating pairs as dual to sums, the tradition
+in mathematics has long been to consider fields with two notions of
+duality: one for sums and one for pairs. This double notion of duality
+has a crisp semantics, clear computational interpretation, and an
+information theoretic basis.
 
@@
 information-preserving framework and the two notions of duality. The new
-insights point to further new areas of investigation of which we mention
+insights point to further new areas of investigation, of which we mention
 the three most significant ones (in our opinion).
@@
 We have confirmed that we can extended {{langRevEE}} with the datatype
-declaration for binary trees and build a witness for this isomorphism that
-works as expected. To understand the issue in more detail, consider the
-following algebraically valid proof of the isomorphism in question:
+declaration for binary trees and build a witness for this isomorphism
+that works as expected. However not every isomorphism constructed from
+algebraic manipulation is computationally meaningful. To understand the
+issue in more detail, consider the following algebraically valid proof
+of the isomorphism in question:
 \[\begin{array}{rclclclcl}
@@
 ``more constructive'' than others.
+
+A related issue is that not all meaningful recursive types are
+meaningful polynomials. For instance {{nat=@@x.(1+x)}} implies the
+polynomial {{x=1+x}} which has no algebraic solutions without appeal
+to more complex structures with limits etc.
+
 %% or dually that {{langRevEE}} with recursive types
@@
 \paragraph*{Quantum Computing.} 
-One understand of quantum computing is that it exploits the laws of physics
+One understanding of quantum computing is that it exploits the laws of physics
 to build faster machines (perhaps). Another more foundational understanding
 is that it provides a computational interpretation of physics, and in
-particular directly address the question of interpretation of quantum
+particular directly addresses the question of interpretation of quantum
 mechanics. In a little known document, Rozas~\cite{Rozas:1987:CMO:889539}
@@
 fixpoint calculation between waves or particles traveling forwards and
-backwards in time. Our work shed no light on whether this interpretation is
+backwards in time. Our work sheds no light on whether this interpretation is
 the ``right one'' but it is interesting that we can directly realize it using
@@
 
-The multiplicative structure of {{langRevEE}} also has a direct connection to
-entangled quantum particles, or perhaps entangled particles and
-anti-particles.  The idea of entanglement is one action on one particle is
-``instanteneously'' communicated to the other. Again our model shed no light
-on whether this is related to how nature computes but it is again interesting
-that we can directly realize the idea using the primitives of {{langRevEE}}.
+The multiplicative structure of {{langRevEE}} also has a direct
+connection to entangled quantum particles, or perhaps entangled
+particles and anti-particles.  The idea of entanglement, that an
+action on one particle is ``instantaneously'' communicated to the
+other, is analogous to how unifying one value affects its dual pair
+which is possibly in another part of the computation.  Again our model
+sheds no light on whether this is related to how nature computes but
+it is again interesting that we can directly realize the idea using
+the primitives of {{langRevEE}}.
 
@@
 and conjunctions $\with$ as well as multiplicative disjunctions $\parr$ and
-conjuctions $\otimes$. Duality is also prominent in linear logic: it relates
+conjunctions $\otimes$. Duality is also prominent in linear logic: it relates
 the additive connectives to each other (the dual of $\oplus$ is $\with$ and
@@
 and continuations that arises in {{langRev}} in a manner that is reminiscent
-of Filinksi's Symmetric \lcal ~\cite{Filinski:1989:DCI:648332.755574}. The
+of Filinski's Symmetric \lcal ~\cite{Filinski:1989:DCI:648332.755574}. The
 symmetry arises by extending {{langRev}} with a notion of additive duality

### 2012-03-12 rational.tex
_

@@
 This section introduces the syntax and graphical languages for
-negative types and fractionals.The general outline is as follows:
+negatives and fractionals.  The general outline is as follows:
 \footnote{We refer the reader to Selinger's excellent survey
@@
 \end{itemize}
-After presenting the extension at the syntactic level, we discuss the
-categorical semantics informally via the graphical language. The operational
-semantics is presented in Sec.~\ref{sec:rat}.
+The extended language is referred to as {{langRevEE}}.  After
+presenting the extension at the syntactic level, we discuss the
+categorical semantics informally via the graphical language. The
+operational semantics is presented in Sec.~\ref{sec:rat}.
 
@@
 
-For convenience, we sometimes use the notations {{b1 - b2}} and {{b1/b2}} to
-indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}} respectively.  The
-types of the new constructs are: 
+For convenience, we sometimes use the notations {{b1 - b2}} and
+{{b1/b2}} to indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}}
+respectively.  The types of the new constructs are:
 \vspace{-15pt}
@@
 When the solver is fed inputs initialized to {{true}}, it clones only
-those inputs to {{sat_f}} that satisfy {{f}} and the {{heap}}
+those inputs to {{sat_f}} that satisfy {{f}} and the heap
 constraints. In the case of unique-SAT the solver will produce exactly

### 2012-03-13 rational_ext.tex
created an _ext version for more notes

--- 2012-03-12 rational.tex
@@
 
-\title{The Two Dualities of Computation: Negative and Fractional Types}
+\title{(With Extensions)The Two Dualities of Computation: Negative and Fractional Types}
 \authorinfo{Roshan P. James}
@@
 \end{multicols}
+
+It is useful to contrast the categorical structure of {{langRevEE}},
+which has two distinct \emph{compact closed} structures, to the
+well-known and closed related \emph{*-autonomous} categories that form
+a model for linear logic. The distnguishing feature that seperates
+*-autonomous structures from compact closed structures is that the
+dualizing action in *-autonomous categories does not preserve the
+tensor. In {{langRev}}, as we will see , we can construct the
+isomorphism {{-(b1+b2)<-> (-b1)+(-b2)}}, i.e. the tensor {{+}} has
+been preserved by the negation.  Whereas in linear logic we have that
+$(a\oplus b)^\bot$ is $a^\bot \parr b^\bot$, i.e. the connective
+$\oplus$ is not preserved.
+
+In linear logic terms `additive pairs' and `multiplicative sums' are
+used to indicate this notion that some connectives has both additive
+and multiplicative properties. The usual intuitive interpretations for
+these involve suggestions like one has a dollar to spend and the both
+cigarettes and soda are available for purchase, but the fact that you
+have only a dollar to spend affords you only one of these. The
+intuitions talk about a sort of dependacy underlying the fact that
+both cigarettes and soda are available but you can have only one.
+
+In {{langRevEE}}, operators {{eta}} and {{eps}} capture a similar
+notion.The pair of values {{v}} and {{1/v}} created by {{eta*}} are
+not fully independant in the way one would think of normal pairs. The
+value of one, affects the value of the other. Similarly, the {{v}} and
+{{-v}} created by {{eta+}} do really present a strict choice of
+{{left}} or {{right}} in the sense that usual sums do, but instead can
+have a \emph{both} {{left}} and {{right}} values at different points
+during the computation. However, in our opinion, the philisophical
+difference the additiveness of the additive pairs and the
+multiplicativesness of the multiplicative pair aren't a property of
+the connective between the values, but a property intrinsic to the
+values themselves. Hence {{v}} and {{1/v}} share secretely a
+relationship that their combinting {{*}} constructor does not and need
+not know about.
 

### 2012-03-13 code/dual-pair/PiPairEE.hs
added code

@@ -1,0 +1,532 @@
+{-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
+-- {-# OPTIONS_GHC -XGADTs -XTypeOperators #-} -- 7.0.1
+
+-- Dont know: This code tested with GHC version 6.12.3, and version 7.0.1
+
+module PiCont where
+
+import Data.Typeable
+import Unsafe.Coerce -- needed for polymorphic lookup. 
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+data Zero 
+data Show a => Neg a = Negative a
+             deriving Show
+
+data a :<=> b where 
+-- Congruence
+  Id    :: a :<=> a
+  Adj   :: (a :<=> b) -> (Neg b :<=> Neg a) 
+  (:.:) :: (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  PlusZeroL   :: Either Zero a :<=> a
+  PlusZeroR   :: a :<=> Either Zero a
+  CommutePlus :: Either a b :<=> Either b a
+  AssocPlusL  :: Either a (Either b c) :<=> Either (Either a b) c 
+  AssocPlusR  :: Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  TimesOneL    :: ((), a) :<=> a
+  TimesOneR    :: a :<=> ((), a)
+  CommuteTimes :: (a,b) :<=> (b,a) 
+  AssocTimesL  :: (a,(b,c)) :<=> ((a,b),c)
+  AssocTimesR  :: ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  TimesZeroL  :: (Zero, a) :<=> Zero
+  TimesZeroR  :: Zero :<=> (Zero, a)
+  Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
+  Factor      :: Either (b, a) (c, a) :<=> (Either b c, a)
+-- Eta and Eps over the monoid (*, 1)
+  Eta :: () :<=> (Neg a, a)
+  Eps :: (a, Neg a) :<=> ()
+
[…]

### 2012-03-13 code/dual-sum/PiSumEE.hs

@@ -1,0 +1,574 @@
+-- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
+{-# OPTIONS_GHC -XGADTs -XTypeOperators #-} -- 7.0.1
+
+
+module PiCont where
+
+import Data.Typeable
+-- needed for polymorphic lookup. 
+import qualified Unsafe.Coerce as Coerce 
+import qualified Debug.Trace as Trace
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+data Zero 
+data Neg a = Neg a
+
+instance Show a => Show (Neg a) where 
+    show (Neg a) = "(Neg " ++ (show a) ++ ")"
+
+data Three = One 
+           | Two 
+           | Three
+             deriving Show
+
+data Tree = Leaf 
+          | Node Tree Tree
+            deriving Show
+
+data a :<=> b where 
+-- Congruence
+  Id    :: a :<=> a
+  IdTrace :: a :<=> a
+  Adj   :: (a :<=> b) -> (Neg b :<=> Neg a) 
+  (:.:) :: (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  ZeroE       :: Either Zero a :<=> a
+  ZeroI       :: a :<=> Either Zero a
+  CommutePlus :: Either a b :<=> Either b a
+  AssocPlusL  :: Either a (Either b c) :<=> Either (Either a b) c 
+  AssocPlusR  :: Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE        :: ((), a) :<=> a
+  UnitI        :: a :<=> ((), a)
+  CommuteTimes :: (a,b) :<=> (b,a) 
+  AssocTimesL  :: (a,(b,c)) :<=> ((a,b),c)
+  AssocTimesR  :: ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  TimesZeroL  :: (Zero, a) :<=> Zero
+  TimesZeroR  :: Zero :<=> (Zero, a)
+  Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
[…]

### 2012-03-13 code/dual/Dual.hs

@@ -1,0 +1,608 @@
+-- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1
+
+-- Dont know: This code tested with GHC version 6.12.3, and version 7.0.1
+
+module Dual where
+
+import Data.Typeable
+import Unsafe.Coerce -- needed for polymorphic lookup. 
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+data Zero 
+data Show a => Neg a = Ng a
+             deriving Show
+data Show a => Inv a = In a
+             deriving Show
+data Three = One 
+           | Two 
+           | Three
+             deriving (Eq, Show)
+
+type (:+:) a b = Either a b
+type (:*:) a b = (a, b)
+
+data a :<=> b where 
+-- Congruence
+  Id    :: a :<=> a
+  Adj   :: (a :<=> b) -> (Neg b :<=> Neg a) 
+  (:.:) :: (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  ZeroE   :: Either Zero a :<=> a
+  ZeroI   :: a :<=> Either Zero a
+  CommutePlus :: Either a b :<=> Either b a
+  AssocPlusL  :: Either a (Either b c) :<=> Either (Either a b) c 
+  AssocPlusR  :: Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE    :: ((), a) :<=> a
+  UnitI    :: a :<=> ((), a)
+  CommuteTimes :: (a,b) :<=> (b,a) 
+  AssocTimesL  :: (a,(b,c)) :<=> ((a,b),c)
+  AssocTimesR  :: ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  TimesZeroL  :: (Zero, a) :<=> Zero
+  TimesZeroR  :: Zero :<=> (Zero, a)
+  Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
[…]

### 2012-03-13 code/dual/Test.hs

@@ -1,0 +1,242 @@
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1
+
+import Dual 
+
+------------------------------------------------------------------------------------
+------------------------------------------------------------------------------------
+-- Constructions
+
+-- Some shorthands 
+type (:=>>) a b = (Inv a, b)
+type (:<<=) a b = (a, Inv b)
+type P a = () :<=> a
+type N a = a  :<=> ()
+
+makeFunc :: (a :<=> b) -> P (a :=>> b)
+makeFunc c = EtaTimes:.: (Id :*: c)
+
+-- Many apply's to be written here based on what interface we want:
+-- apply : P (a :=>> b) -> P a -> P b
+-- apply : P (a :=>> b) -> (c :<=> a) -> (c :<=> b)
+-- apply : ((a :=>> b), a) :<=> b
+
+applyFunc :: ((a :=>> b), a) :<=> b 
+applyFunc = CommuteTimes 
+            :.: AssocTimesL 
+            :.: ((CommuteTimes :.: EpsTimes) :*: Id) 
+            :.: UnitE
+
+-- These 
+makeDC :: (a :<=> b) -> N (a :<<= b)
+makeDC c = (c :*: Id) :.: CommuteTimes :.:  EpsTimes
+
+
+-- A similar apply is possible, but I dont know what it means. 
+
+-- Trace
+traceTimes :: (a, b) :<=> (a, c) -> b :<=> c
+traceTimes c = UnitI 
+               :.: (EtaTimes:*: Id) 
+               :.: AssocTimesR 
+               :.: (Id :*: c) 
+               :.: AssocTimesL 
+               :.: (EpsTimes :*: Id)
+               :.: UnitE
+
+-- This is the yanking lemma for trace. 
+yankTimes :: a :<=> a
+yankTimes = traceTimes CommuteTimes
+
+
+z_shapeTimes :: a :<=> a
+z_shapeTimes = UnitI
+               :.: (EtaTimes:*: Id) 
+               :.: (CommuteTimes :*: Id)
+               :.: AssocTimesR 
+               :.: (Id :*: EpsTimes)
+               :.: CommuteTimes
+               :.: UnitE
+               
+
+-- Not works by first converting the Haskell boolean via a Unfold to
[…]

### 2012-03-14 iso-int/rc-iso-int.tex
abstract rc

--- 2012-03-02 iso-int/rc-iso-int.tex
@@
 \begin{abstract}
+
 Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
@@
 {{langRevT}}.
+
+  In our previous work [RC'11, POPL'12], we developed a reversible
+  programming language and established that every computation in it is a
+  (partial) isomorphism that is reversible and that preserves
+  information. The language is founded on type isomorphisms that have a clear
+  categorical semantics but that are awkward as a notation for writing actual
+  programs, especially recursive programs. This paper remedies this aspect by
+  presenting a systematic technique for developing a large and expressive
+  class of reversible recursive programs, that of logically reversible
+  small-step abstract machines. In other words, this paper shows that once we
+  have a logically reversible machine in a notation of our choice, expressing
+  the machine as an isomorphic interpreter can be done systematically and
+  does not present any significant conceptual difficulties. Concretely, we
+  develop several simple interpreters over numbers and addition, move on to
+  tree traversals, and finish with a meta-circular interpreter for our
+  reversible language. Generally, this gives us a means of developing large
+  reversible programs with the ease of reasoning at the level of a
+  conventional small-step semantics.
+
 \end{abstract}

### 2012-03-14 iso-int/rc-iso-int.tex
worked on a few pieces

@@
 %! bullet = \bullet
-%! * = \times
 
@@
 \begin{abstract}
-
-Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
-properties and various constructions in it. However, for a
-computational model, there has been little focus on actually writing
-programs directly in {{langRevT}}. This paper aims to remedy that by
-presenting a systematic technique of expressing logically reversible
-small-step abstract machines in {{langRevT}}. In other words, this
-paper shows that once we have a logically reversible machine in a
-notation of our choice, expressing the machine as an isomorphic
-interpreter in {{langRevT}} can be done systematically and does not
-present any significant conceptual difficulties. This gives us a means
-of developing large {{langRevT}} programs with the ease of reasoning
-at the level of a conventional small-step semantics.  We develop
-several simple interpreters over numbers and addition, move onto tree
-traversal and finish with an interpreter for {{langRevT}} expressed in
-{{langRevT}}.
-
+% Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
+% properties and various constructions in it. However, for a
+% computational model, there has been little focus on actually writing
+% programs directly in {{langRevT}}. This paper aims to remedy that by
+% presenting a systematic technique of expressing logically reversible
+% small-step abstract machines in {{langRevT}}. In other words, this
+% paper shows that once we have a logically reversible machine in a
+% notation of our choice, expressing the machine as an isomorphic
+% interpreter in {{langRevT}} can be done systematically and does not
+% present any significant conceptual difficulties. This gives us a means
+% of developing large {{langRevT}} programs with the ease of reasoning
+% at the level of a conventional small-step semantics.  We develop
+% several simple interpreters over numbers and addition, move onto tree
+% traversal and finish with an interpreter for {{langRevT}} expressed in
+% {{langRevT}}.
   In our previous work [RC'11, POPL'12], we developed a reversible
@@
   conventional small-step semantics.
-
 \end{abstract}
@@
 \section{Introduction}
+
+Reversible languages
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Review of the Reversible Language: {{langRevT}} }
+\label{sec:pi}
+
+We briefly review the reversible language {{langRevT}} introduced in our
+previous work~\cite{rc2011,infeffects}. The set of types includes the empty
+type {{0}}, the unit type {{1}}, sum types {{b1+b2}}, product types
+{{b1*b2}}, and recursive types {{@@x.b}}. The set of values {{v}} includes
+{{()}} which is the only value of type {{1}}, {{left v}} and {{right v}}
+which inject {{v}} into a sum type, {{(v1,v2)}} which builds a value of
+product type, and {{<v>}} which is used to construct recursive values. There
+are no values of type {{0}}:
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b + b | b * b | x | @@x.b
+% values, v ::= () | left v | right v | (v, v) | <v>
+
+The expressions of {{langRevT}} are witnesses to the following type
+isomorphisms:
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%zeroe :&  0 + b <-> b &: zeroi
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist0 :& 0 * b <-> 0 &: factor0
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+%fold :&~ b[@@x.b/x] <-> @@x.b ~&: unfold
+Each line of the above table introduces one or two combinators that witness
+the isomorphism in the middle. Collectively the isomorphisms state that the
+structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
+of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
+multiplication distributes over addition. The last isomorphism witnesses the
+equivalence of a value of recursive type with all its ``unrollings.'' The
+isomorphisms are extended to form a congruence relation by adding the
+following constructors that witness equivalence and compatible closure. In
+addition, the language includes a {{trace}} operator to express looping:
+%subcode{proof} include main
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+%
+%@ c : b1 + b2 <-> b1 + b3
+%@@ trace c : b2 <-> b3
+
+\paragraph*{Adjoint.} 
+An important property of the language is that every combinator {{c}} has an
+adjoint {{c{dagger}}} that reverses the action of {{c}}. This is evident by
+construction for the primitive isomorphisms. For the closure combinators, the
+adjoint is homomorphic except for the case of sequencing in which the order
+is reversed, i.e., {{(c1 (;) c2){dagger} = (c2{dagger}) (;) (c1{dagger}) }}.
+
+\paragraph*{Graphical Language.}
+The syntactic notation above is often obscure and hard to read.  Following
+the tradition established for monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
+graphical language that conveys the semantics of the language.  The general
+idea of the graphical notation is that combinators are modeled by ``wiring
+diagrams'' or ``circuits'' and that values are modeled as ``particles'' or
+``waves'' that may appear on the wires. Evaluation therefore is modeled by
+the flow of waves and particles along the wires.
+
+\begin{itemize}
+\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
+  is simply represented as a wire labeled by its type {{b}}, as shown on the
+  left. In more complex diagrams, if the type of a wire is obvious from the
+  context, it may be omitted. When tracing a computation, one might imagine a
+  value {{v}} of type {{b}} on the wire, as shown on the right.
+
+  \begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire-value.pdf}
+}
+\end{center}
+  \end{multicols}
+
+\item The product type {{b1*b2}} may be represented using either one wire
+  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
+  case of products represented by a pair of wires, when tracing execution
+  using particles, one should think of one particle on each wire or
+  alternatively as in folklore in the literature on monoidal categories as a
+  ``wave.''
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
+\includegraphics{diagrams/thesis/product-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
+\includegraphics{diagrams/thesis/product-two-wires.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Sum types may similarly be represented by one wire or using
+  parallel wires with a {{+}} operator between them. When tracing the
+  execution of two additive wires, a value can reside on only one of the two
+  wires.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
+\includegraphics{diagrams/thesis/sum-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
+\includegraphics{diagrams/thesis/sum-two-wires.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
+}
+\end{center}
+\end{multicols}
+
+%% \item
+%% When representing complex types like {{(b1*b2)+b3}} some visual
+%% grouping of the wires may be done to aid readability. The exact type
+%% however will always be clarified by the context of the diagram.
+
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
+%% }
+%% \end{center}
+
+\item Associativity is implicit in the graphical language. Three parallel
+  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
+\includegraphics{diagrams/thesis/assoc.pdf}
+}
+\end{center}
+
+\item Commutativity is represented by crisscrossing wires.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
+\includegraphics{diagrams/thesis/swap_times.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
+\includegraphics{diagrams/thesis/swap_plus.pdf}
+}
+\end{center}
+\end{multicols}
+
+By visually tracking the flow of particles on the wires, one can
+verify that the expected types for commutativity are satisfied.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_times_value.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_plus_value.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item The morphisms that witness that {{0}} and {{1}} are the additive and
+  multiplicative units are represented as shown below. Note that since there
+  is no value of type 0, there can be no particle on a wire of type {{0}}.
+  Also since the monoidal units can be freely introduced and eliminated, in
+  many diagrams they are omitted and dealt with explicitly only when they are
+  of special interest.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
+\includegraphics{diagrams/thesis/uniti.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
+\includegraphics{diagrams/thesis/unite.pdf}
+}
+\end{center}  
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
+\includegraphics{diagrams/thesis/zeroi.pdf}
+}
+\end{center}
+\columnbreak
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
+\includegraphics{diagrams/thesis/zeroe.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Distributivity and factoring are represented using the dual
+  boxes shown below:
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/factor.pdf}
+\end{center}
+\end{multicols}
+
+Distributivity and factoring are interesting because they represent
+interactions between sum and pair types. Distributivity should
+essentially be thought of as a multiplexer that redirects the flow of
+{{v:b}} depending on what value inhabits the type {{b1+b2}}, as shown
+below. Factoring is the corresponding adjoint operation.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
+\end{center}
+\end{multicols}
+
+\item \todo{fold, unfold, and trace}
+\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Simple Bounded Iterator}
-We illustrate the main ideas underlying the derivation of isomorphic
-interpreters using a simple abstract machine.  Consider the definition
-of the following small-step abstract machine:
-%\vspace{-30px}
+\section{Simple Bounded Number Iteration}
+
+We illustrate the main concepts and constructions using two simple examples
+that essentially count $n$ steps. The first machine does nothing else; the
+second uses this counting ability to add two numbers.
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Counting}
+
+The first machine is defined as follows:
+
 \begin{multicols}{2}
@@
 
-This machine starts with a number {{n}} in the first position and
-{{0}} in the second. Each reduction step decrements the first number
-and increments the second. The machine stops when the first number
-reaches 0, thereby taking exactly {{n}} steps.  
-%% We can turn this
-%% abstract machine into a an isomorphic interpreter implemented in
-%% {{langRevT}}, in a systematic way. 
-Broadly, the interpreter should implement the map on the left:
+\noindent The machine is started with a number {{n}} in the first position
+and {{0}} in the second. Each reduction step decrements the first number and
+increments the second. The machine stops when the first number reaches 0,
+thereby taking exactly {{n}} steps. For example, if the machine is started in
+the configuration {{<3,0>}}, it would take exactly 3 steps to reach the final
+configuration: {{ <3,0> |--> <2,1> |--> <1,2> |--> <0,3> }}. Dually, since
+the machine transition is clearly reversible, the machine can execute
+backwards from the final configuration to reach the initial configuration in
+also 3 steps.
+
+Our goal is to implement this abstract machine in {{langRevT}}, i.e., our
+goal is to write a combinator {{c : nat * nat <-> nat * nat}} where {{nat}}
+is an abbreviation for the recursive type {{@@x.1+x}}. The combinator should
+map {{(n,0)}} to {{(0,n)}} by precisely mimicking the steps of the abstract
+machine. Let us analyze this desired combinator {{c}} in detail starting from
+its interface:
 
@@
 
-%% \begin{center}
-%%   \includegraphics{diagrams/nat-nat2.pdf}
-%% \end{center}  
-%% \end{multicols}
-
-In {{langRevT}} we have {{nat}} given by the recursive definition
-{{@@x.1+x}} along with isomorphisms {{unfold:nat <->1+nat:fold}}. The
-only reduction step of the abstract machine examines the first value
-of the input and reconstructs the second.
+The reduction step of the machine examines the first {{nat}} and reconstructs
+the second {{nat}}. In other words, the first {{nat}} is unfolded to examine
+its structure and the second {{nat}} is folded to reconstruct it:
 
@@
 machine states by introducing distribution operations and any required
-swaps.
+swaps:
 
@@
 The construction above already gives us some insight into how we can
-implement our machine's one reduction step. We want the output {{n'}}
-to be {{n+1}} where {{n}} is the first component of the input and
-similarly the output {{m'}} to be {{m+1}} where {{m}} is the second
-component of the input.  We can express these equalities by connecting
-wires -- by connecting the {{n-1}} wire to {{n'}} we express the
-equality {{n' =n-1}}. Similarly we connect {{m}} to {{m'-1}} to
-express {{m' = m+1}}.
+implement our machine's one reduction step. We want the output {{n'}} to be
+{{n+1}} where~{{n}} is the first component of the input and similarly the
+output {{m'}} to be {{m+1}} where {{m}} is the second component of the input.
+We can express these equalities by connecting wires, i.e., by connecting the
+{{n-1}} wire to {{n'}} and the {{m}} wire to {{m'-1}}:
 
@@
 
-The branch labeled {{((), m)}} corresponds to the machine state
-{{<0,m>}} and hence looks like the stop state of the
-machine. Similarly the branch labeled {{((), n')}} corresponds to the
-start state of the machine.
-
-We are close to the completion of the interpreter at this point. The
-last step in the construction involves introducing a {{trace}]
-operation for iteration of reduction steps and exposing the start and
-stop states. We do this by `flipping' the construction so far
-inside-out.
+The branch labeled {{((), m)}} corresponds to the machine state {{<0,m>}} and
+hence looks like the stop state of the machine. Similarly the branch labeled
+{{((), n')}} corresponds to the start state of the machine. We are close to
+the completion of the interpreter at this point. The last step in the
+construction involves introducing a {{trace}] operation for iteration of
+reduction steps and exposing the start and stop states. We do this by
+``sliding'' the two sections of the diagram as shown below:
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Abstracting out the Construction}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Abstracting out the Construction}
 
-The construction of the isomorphic interpreter essentially exploits
-the structure of most abstract machines -- the LHS of the transition
-rules determines a machine state by pattern machine. The RHS does the
-dual namely term reconstruction. The fact that the machine is
-logically reversible ensures that the computation performed by each
-reduction is essentially captured by an isomorphism. 
-%% To develop an
-%% understanding for what sort of machines can be treated this way, let
-%% us look at some examples.
+% The construction of the isomorphic interpreter essentially exploits
+% the structure of most abstract machines -- the LHS of the transition
+% rules determines a machine state by pattern machine. The RHS does the
+% dual namely term reconstruction. The fact that the machine is
+% logically reversible ensures that the computation performed by each
+% reduction is essentially captured by an isomorphism.
 
-Consider a machine with the following two reduction steps
-{{<n+1,m>|--> <n, m+1>}} and {{<n+1,m>|--><n,m>}}. Can we create an
-isomorphic interpreter for such a machine? We can't, because for any
-given machine state, it is not obvious what reduction rule to
-apply. We could reverse this machine if there was a clear
-non-ambiguity in which machine rule was chosen.
+% Consider a machine with the following two reduction steps
+% {{<n+1,m>|--> <n, m+1>}} and {{<n+1,m>|--><n,m>}}. Can we create an
+% isomorphic interpreter for such a machine? We can't, because for any
+% given machine state, it is not obvious what reduction rule to
+% apply. We could reverse this machine if there was a clear
+% non-ambiguity in which machine rule was chosen.
 
-Consider a machine, with the following rewrite rules
-{{<n+1,m,p>|--><n,m,p+1>}} and {{<0,m+1,p>|--><0,m,p+1>}}. Can we
-create an isomorphic interpreter for this machine? It turns out we
-can't because for any given output state {{<0,m,p+1>}} it is not
-obvious which input state it came from - i.e. there are two
-possibilities {{<1,m,p>}} or there is {{<0,m+1,p>}}.
+% Consider a machine, with the following rewrite rules
+% {{<n+1,m,p>|--><n,m,p+1>}} and {{<0,m+1,p>|--><0,m,p+1>}}. Can we
+% create an isomorphic interpreter for this machine? It turns out we
+% can't because for any given output state {{<0,m,p+1>}} it is not
+% obvious which input state it came from - i.e. there are two
+% possibilities {{<1,m,p>}} or there is {{<0,m+1,p>}}.
 
-Consider a machine that has the addition rule {{<''x+y''>|--><''z''>}}
-where {{z}} is the sum of {{x}} and {{y}}. This machine again cannot
-be turned into an isomorphic interpreter because the one reduction
-rule does not have a logically inverse -- in other words, given a
-{{z}}, we are unable to determine the original {{x}} and {{y}}.
+% Consider a machine that has the addition rule {{<''x+y''>|--><''z''>}}
+% where {{z}} is the sum of {{x}} and {{y}}. This machine again cannot
+% be turned into an isomorphic interpreter because the one reduction
+% rule does not have a logically inverse -- in other words, given a
+% {{z}}, we are unable to determine the original {{x}} and {{y}}.
 
-If we had only the reduction relation {{<n+1> |--> <n>}}, can we
-construct an isomorphic interpreter? After all, every rewrite rule is
-logically reversible. In fact, this is not possible because there is
-no distinguished start state for such an interpreter and hence one
-does not know when to stop during reverse execution.
+% If we had only the reduction relation {{<n+1> |--> <n>}}, can we
+% construct an isomorphic interpreter? After all, every rewrite rule is
+% logically reversible. In fact, this is not possible because there is
+% no distinguished start state for such an interpreter and hence one
+% does not know when to stop during reverse execution.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \subsection{What abstract machines can be turned into isomorphic interpreters?}
+% \paragraph*{Conditions that must be satisfied for constructing a isomorphic
+% interpreter}:
+% \begin{enumerate}
+% \item Distinguishable start and stop states. There may be more than one
+%   valid start state and more than one valid stop state.
+% \item Each valid machine state must match a unique reduction LHS. Each
+%   valid machine state must machine a unique reduction RHS.
+% \item Every reduction step must be computable (in {{langRevT}}) and
+%   must be logically reversible -- i.e. it must be possible to reduce
+%   from right to left.
+% \end{enumerate}
 
-\paragraph*{Conditions that must be satisfied for constructing a isomorphic
-interpreter}:
+% \paragraph*{High level steps in the translation:}
+% \label{sec:steps}
+% \begin{enumerate}
+% \item Expand the input such that every input term that the LHS of the
+%   interpreter pattern matches on is exposed.
+% \item Expand the output such that every output term that it
+%   reconstructed by the interpreter is exposed.
+% \item Shuffle matching LHS term to matching RHS terms, inserting any
+%   appropriate mediating computations. In the previous example, this
+%   was only a swap operation.
+% \item Flip the construction inside out introducing a {{trace}} and
+%   exposing the start and stop states. 
+% \end{enumerate}
 
+\subsection{Adder}
+
+Although trivial, the previous example captures the fundamental steps of our
+construction. For reference, these steps are:
 \begin{enumerate}
-
-\item Distinguishable start and stop states. There may be more than one
-  valid start state and more than one valid stop state.
-
-\item Each valid machine state must match a unique reduction LHS. Each
-  valid machine state must machine a unique reduction RHS.
-
-\item Every reduction step must be computable (in {{langRevT}}) and
-  must be logically reversible -- i.e. it must be possible to reduce
-  from right to left.
-
+\item Expand the input to expose enough structure to distinguish the left
+  hand side of each machine state; 
+\item Expand the output to expose enough structure to distinguish the right
+  hand side of each machine state;
+\item Shuffle matching input terms to output terms, inserting any appropriate
+  mediating computations. 
+\item Slide the two sections to expose the start and stop state and introduce
+  a {{trace}} to iterate the construction.
 \end{enumerate}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{High level steps in the translation:}
-\label{sec:steps}
-
-\begin{enumerate}
-\item Expand the input such that every input term that the LHS of the
-  interpreter pattern matches on is exposed.
-
-\item Expand the output such that every output term that it
-  reconstructed by the interpreter is exposed.
-
-\item Shuffle matching LHS term to matching RHS terms, inserting any
-  appropriate mediating computations. In the previous example, this
-  was only a swap operation.
-
-\item Flip the construction inside out introducing a {{trace}} and
-  exposing the start and stop states.
-
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Adder and Multiplier}
-
-\section{Logically Reversible Adder}
-
-Consider
-
+Let us apply these steps to the slightly more interesting example of an
+adder:
 \begin{multicols}{2}  
@@
 %subcode{bnf} include main
-% Start state = <n, p, 0>
-% Stop State = <0, p', m>
+% Start state = <n, n, 0>
+% Stop State = <0, n, n>
 \end{multicols}
-
-%Small-step Operational semantics:
-
 %subcode{opsem} include main
@@
 
-This is an ostensibly useful machine in the sense that it computes the
-sum of its two inputs.  This construction proceeds exactly like the
-previous one till we have the following situation:
+The idea of the machine is to start with 3 numbers: the two numbers to add
+and an accumulator initialized to 0. Each step of the machine, decrements one
+of the numbers and increments the second number and the accumulator. For
+example, {{<3,4,0> |--> <2,5,1> |--> <1,6,2> |--> <0,7,3>}}. In general, the
+sum of the two numbers will be in the second component, and the last
+component is supposed to record enough information to make the machine
+reversible.
 
-\begin{center}
-  \includegraphics{diagrams/adder1.pdf}
-\end{center}
+A closer look however reveals that the machine only defines a \emph{partial}
+isomorphism: not all valid final states can be mapped to valid start
+states. Indeed consider the configuration {{<0,2,3>}} which is a valid final
+state. Going backwards, the transitions start as follows
+{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine 
+gets stuck at a state that is not a valid start state. 
+There are several ways to make this machine reversible. The simplest is to
+add another possible kind of start state {{<n,0,n>}}. Our construction now
+proceeds as expected. After expanding the input to expose $n$ and the output
+to expose both $p$ and $m$, we reach the following configuration:
 
-The equalities we would like to express at this point are {{n-1=n'}},
-{{m=m'-1}} and that {{p+1=p'}}. Fortunately {{langRevT}} provides the
-{{add1: nat <-> nat}} combinator and so we could just use that to
-mediate between {{p}} and {{p'}}. This is one solution and this gives
-us:
+% \begin{center}
+%   \includegraphics{diagrams/adder1.pdf}
+% \end{center}
 
-\begin{center}
-  \includegraphics{diagrams/adder2.pdf}
-\end{center}
+% The equalities we would like to express at this point are {{n-1=n'}},
+% {{m=m'-1}} and that {{p+1=p'}}. Fortunately {{langRevT}} provides the
+% {{add1: nat <-> nat}} combinator and so we could just use that to
+% mediate between {{p}} and {{p'}}. This is one solution and this gives
+% us:
 
-This can be inverted as before, to give us an interpreter. While this
-is a correct implementation of the small-step semantics, it is
-somewhat unsatisfying for the reason that it does not see very
-systematic. After all we chose to use an {{add1}} for {{p}} -- why
-didn't we do the same for {{m}}? More importantly, the fact that we
-chose used {{add1}} is also hiding something -- namely that {{add1}}
-is a partial isomorphism. Its inverse is not defined when applied to
-{{0}} i.e. it will go into an infinite loop.
+% \begin{center}
+%   \includegraphics{diagrams/adder2.pdf}
+% \end{center}
 
-In a sense this implicit asymmetry of the abstract machine was being
-hidden by the {{add1}} operation. So instead of choosing to use
-{{add1}} let us chose to expand out {{p'}}, as recommended by step 2
-in section \ref{sec:steps}. This gives us the following:
+% This can be inverted as before, to give us an interpreter. While this
+% is a correct implementation of the small-step semantics, it is
+% somewhat unsatisfying for the reason that it does not see very
+% systematic. After all we chose to use an {{add1}} for {{p}} -- why
+% didn't we do the same for {{m}}? More importantly, the fact that we
+% chose used {{add1}} is also hiding something -- namely that {{add1}}
+% is a partial isomorphism. Its inverse is not defined when applied to
+% {{0}} i.e. it will go into an infinite loop.
+
+% In a sense this implicit asymmetry of the abstract machine was being
+% hidden by the {{add1}} operation. So instead of choosing to use
+% {{add1}} let us chose to expand out {{p'}}, as recommended by step 2
+% in section \ref{sec:steps}. This gives us the following: 
+%
+% We can very that the connected wires do indeed give us the desired
+% equalities between the inputs and outputs. Also we recognize our
+% expected start state and expected stop state. What is the dangling
+% state?
+
+% The dangling state was what has hidden by the use of {{add1}} and
+% denotes the fact that the abstract machine did not define any valid
+% backward reduction for the following valid machine state
+% {{<n,0,m+1>}}. To take a backward step, both {{m}} and {{p}} have to
+% be simultaneously decremented and this is not possible if {{p}} is
+% {{0}}.  At this point we have two choices:
 
@@
 
-We can very that the connected wires do indeed give us the desired
-equalities between the inputs and outputs. Also we recognize our
-expected start state and expected stop state. What is the dangling
-state?
-
-The dangling state was what has hidden by the use of {{add1}} and
-denotes the fact that the abstract machine did not define any valid
-backward reduction for the following valid machine state
-{{<n,0,m+1>}}. To take a backward step, both {{m}} and {{p}} have to
-be simultaneously decremented and this is not possible if {{p}} is
-{{0}}.  At this point we have two choices:
-
-\begin{enumerate}
-\item We can hide the use the combinator {{omega : b <-> 0}}
-  expressible in {{langRevT}}. This combinator is basically an
-  infinite loop on all inputs and hence would allow us to entirely
-  discard one possible branch. In other words, we are explicitly
-  admitting that our machine is undefined on some inputs.
-
-\item We can admit that there are two possible start states to this
-  abstract machine {{<n, m, 0>}} and {{<n,0,m>}}. One of this is an
-  artifact of the fact that backward execution encounters stuck
-  states. Dually, if there were stuck states for forward execution,
-  they would be reflected in the fact that there were multiple stop
-  states -- only one of which denotes a desirable or correct execution
-  of the machine.
-
-\end{enumerate}
-
-As we did before, we flip inside out the construction so far and
-apply {{trace}} to get the isomorphic interpreter:
+We have also connected the matching input and output wires. At this point, we
+have exposed both start states and the stop state. 
+% \begin{enumerate}
+% \item We can hide the use the combinator {{omega : b <-> 0}}
+%   expressible in {{langRevT}}. This combinator is basically an
+%   infinite loop on all inputs and hence would allow us to entirely
+%   discard one possible branch. In other words, we are explicitly
+%   admitting that our machine is undefined on some inputs.
+% \item We can admit that there are two possible start states to this
+%   abstract machine {{<n, m, 0>}} and {{<n,0,m>}}. One of this is an
+%   artifact of the fact that backward execution encounters stuck
+%   states. Dually, if there were stuck states for forward execution,
+%   they would be reflected in the fact that there were multiple stop
+%   states -- only one of which denotes a desirable or correct execution
+%   of the machine.
+% \end{enumerate}
+We can slide the two sides of the diagram and tie the knot using {{trace}} to
+get the desired combinator:
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tree Traversal}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Advanced Examples}
 
-Let us take the ideas developed so far and apply them to developing a
-tree traversal interpreter. We define a typical binary tree, one that
-has numbers as leaf values and also define a notion of tree
-contexts. The tree contexts serve to track which subtree is currently
-being explored and are similar to tree zippers CITE.
+We show the generality of our construction by applying it to two non-trivial
+examples: a tree traversal and a meta-circulator interpreter for
+{{langRevT}}.
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Tree Traversal}
+
+The type of binary trees we use is {{@@x.nat+x*x}}, i.e., binary trees with
+no information at the nodes and with natural numbers at the leaves. To define
+the abstract machine, we need a notion of \emph{tree contexts} to track which
+subtree is currently being explored. The definitions are:
 
@@
 
-We define an abstract machine whose task is to traverse a given tree
-and increment every leaf value. Small-step Operational semantics:
+Our example machine below traverses a given tree and increments every leaf
+value:
 
@@
 previously. In particular, we have a two types of machine states --
-{{<t,c>}} and {{ {[t, c]} }}. One of these corresponds to walking
-down a tree, building up the context in the process. The other
+{{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds to walking
+down a tree, building up the context in the process. The second
 corresponds to reconstructing the tree from the context and also
 switching focus to any unexplored subtrees in the process. There are
-also two syntactic categories to deal with -- trees and tree context
--- where previously we only had numbers. The {{fold}} and {{unfold}}
+also two syntactic categories to deal with (trees and tree contexts)
+where previously we only had numbers. The {{fold}} and {{unfold}}
 isomorphisms that we need for trees and tree contexts are:
@@
 
-However we will see that the previously developed systematic approach
-developing {{langRevT}} interpreters carries over to this abstract
-machine as well.  As the first step, we quickly check that the
-abstract machine however has distinguished start and stop states, has
-no ambiguities in reduction and that each reduction step is logically
-reversible.
-
-\subsection{New Notation}
-
-Before we start on the actual construction, let us introduce a
-syntactic convenience. In the past interpreter we explicitly drew out
-every {{fold}} and {{unfold}} followed/preceded immediately by any
-required swaps and then a {{distribute}} or {{factor}}. In this
-presentation we will combine all of these into one construct which we
-will draw as a thin vertical rectangle. Also we will introduce the
-convention that the component that is being expanded or collapsed will
-be marked by using {{[^ ^]}} and the components that are being
-generated will be marked by {{[v v]}}.
+\paragraph*{New Notation.}
+To make the diagrams easier to understand, we introduce a syntactic
+convenience which combines the first steps of the construction that consist
+of {{fold}} / {{unfold}} and {{distribute}} / {{factor}}. We collectively
+represent these steps using thin vertical rectangle. Also we will introduce
+the convention that the component that is being expanded or collapsed will be
+marked by using {{[^ ^]}} and the components that are being generated will be
+marked by {{[v v]}}. For example, given a value of type {{tree*nat}}, the
+diagram below shows how to first expand the {{tree}} component and then in
+one of the generated branches, expand the {{nat}} component:
 
@@
 
-The above shows the expansion of a {{tree*nat}} type. First the
-{{tree}} component is expanded and in one of the generated branches
-the {{nat}} component is expanded.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Isomorphic Interpreter}
-
-The first step to developing the isomorphic interpreter is to
-recognize that the two possible kinds of machine states simply hide an
-implicit {{bool}}. We make this explicit.
+We can now apply our construction. The first step to developing the
+isomorphic interpreter is to recognize that the two possible kinds of machine
+states simply hide an implicit {{bool}}. We make this explicit:
 
 %subcode{opsem} include main
-% columnStyle = rclr
+%! columnStyle = rclr
 % <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
@@
 
-On the LHS, for {{T}} states, we have expanded the tree component and
-for {{F}} states we have expanded the tree contexts. Similarly we have
-done the opposite on the RHS side exactly matching up what the
-abstract machine does. One thing to note is that we dropped the {{1}}
-introduced by expanding the {{bool}} and instead just labeled the
-{{true}} and {{false}} branches.
+On the input side, for {{T}} states, we have expanded the tree component and
+for {{F}} states we have expanded the tree contexts. Similarly we have done
+the opposite on the output side exactly matching up what the abstract machine
+does. One thing to note is that we dropped the {{1}} introduced by expanding
+the {{bool}} and instead just labeled the {{true}} and {{false}} branches.
 
 We are ready to start connecting the machine states corresponding to
-the reductions that we would like.
+the reductions that we would like:
+\begin{enumerate}
+\item When we encounter a leaf we would like to increment its value and move
+  to the corresponding {{F}} machine state. For the sake of simplicity in
+  this interpreter we won't be concerned with exposing stuck states on
+  reverse execution caused by {{sub1 0}}. So we are free to use the {{add1}}
+  combinator.
 
-\begin{enumerate}
-
-\item 
-When we encounter a leaf we would like to increment its value and move
-to the corresponding {{F}} machine state. For the sake of simplicity
-in this interpreter we won't be concerned with exposing stuck states
-on reverse execution caused by {{sub1 0}}. So we are free to use the
-{{add1}} combinator.
-
-\item
-For all the other reduction rules, it is a straight forward mapping of
+\item For all the other reduction rules, it is a straightforward mapping of
   related states following the reduction rules.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{A {{langRevT}} Interpreter}
+%%%%%%%%%%%%%%%%%%%%%
+\subsection{A {{langRevT}} Interpreter}
 
@@
 
-TODO : The real difficulty with this interpreter is to argue why it is
-not full of undefined states. For instance, why the machines will
-never reach stuck states such as {{<c1(+)c2, (), ..>}}.  The right way
-out of that is to make the type system of the abstract machine rich
-enough that the structure of the abstract automatically rules out this
-sort of thing.
+% TODO : The real difficulty with this interpreter is to argue why it is
+% not full of undefined states. For instance, why the machines will
+% never reach stuck states such as {{<c1(+)c2, (), ..>}}.  The right way
+% out of that is to make the type system of the abstract machine rich
+% enough that the structure of the abstract automatically rules out this
+% sort of thing.
 
-%subcode{bnf} include main
-% Combinators, c(t1<->t2) = 
-%             &|& c(t<->t) = iso 
-%             &|& c(t1<->t2) = c(t1<->t3) (;) c(t3<->t2) 
-%             &|& c(t1*t2<->t3*t4) = c(t1<->t3) (*) c(t2<->t4) 
-%             &|& c(t1+t2<->t3+t4) = c(t1<->t3) (+) c(t2<->t4)
-%
-% Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
-%                  &|& LeftTimes cc c v | RightTimes c v cc 
-%                  &|& LeftPlus cc c | RightPlus c cc 
-%
-% Type Indexed Values, v(t) =
-%  &|& v(1) = () 
-%  &|& v(t1*t2) = (v(t1), v(t2)) 
-%  &|& v(t1+t2) = L v(t1) | R v(t2)
-%
-% Machine states = <c, v, cc> | {[c, v, cc]}
-% Start state = <c, v, []> 
-% Stop State = {[c, v, []]}
+% %subcode{bnf} include main
+% % Combinators, c(t1<->t2) = 
+% %             &|& c(t<->t) = iso 
+% %             &|& c(t1<->t2) = c(t1<->t3) (;) c(t3<->t2) 
+% %             &|& c(t1*t2<->t3*t4) = c(t1<->t3) (*) c(t2<->t4) 
+% %             &|& c(t1+t2<->t3+t4) = c(t1<->t3) (+) c(t2<->t4)
+% %
+% % Combinator Contexts, cc = [] | Fst cc c | Snd c cc 
+% %                  &|& LeftTimes cc c v | RightTimes c v cc 
+% %                  &|& LeftPlus cc c | RightPlus c cc 
+% %
+% % Type Indexed Values, v(t) =
+% %  &|& v(1) = () 
+% %  &|& v(t1*t2) = (v(t1), v(t2)) 
+% %  &|& v(t1+t2) = L v(t1) | R v(t2)
+% %
+% % Machine states = <c, v, cc> | {[c, v, cc]}
+% % Start state = <c, v, []> 
+% % Stop State = {[c, v, []]}
 
-However this begs the question - does {{langRevT}} need all these GADT
-types to reason about all these things? Something to think about. If
-we do, then there are two ways out
+% However this begs the question - does {{langRevT}} need all these GADT
+% types to reason about all these things? Something to think about. If
+% we do, then there are two ways out
 
-(1) add GADTs to {{langRevT}} and keep the structure of things
-simple. They can always be added opaquely for any given types -- we
-only need to add the proper fold/unfold morphisms for those types
-(maybe).
+% (1) add GADTs to {{langRevT}} and keep the structure of things
+% simple. They can always be added opaquely for any given types -- we
+% only need to add the proper fold/unfold morphisms for those types
+% (maybe).
 
-(2) compile down to {{langRevT}} as it is. In this case, {{langRevT}}
-is indeed a relatively less typed language than the source language
-and will hence have to treat as stuck-states the non-type correct
-machine states. We could apply omega to them and make them vanish --
-for example. This is similar to the handling of the {{sub1}} in the
-previous example.
+% (2) compile down to {{langRevT}} as it is. In this case, {{langRevT}}
+% is indeed a relatively less typed language than the source language
+% and will hence have to treat as stuck-states the non-type correct
+% machine states. We could apply omega to them and make them vanish --
+% for example. This is similar to the handling of the {{sub1}} in the
+% previous example.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2012-03-15 iso-int/rc-iso-int.tex
citations and collecting things for intro

@@
 
-Reversible languages
+Landscape:
+
+- write program; derive inverse
+program~\cite{Gluck:2005,Mu:2002:IFF:648086.757330,conf/aplas/GluckK03,Gluck:2005};
+we're not interested in this; we want to directly write partial isomorphisms
+
+ - design reversible abstract machines by hand (e.g.,
+ SE(M)CD~\cite{Kluge:1999:SEMCD} or SECD-H~\cite{lorenz}) and then implement
+ them; this is what we are doing; our implementation is in this language in
+ which every computation is a partial isomorphism so the implementation will
+ be checked to be reversible by construction
+
+ - generally one can take any reversible algorithm; transform it to a
+ small-step reversible abstract machine, and then implement the machine in
+ our
+ language~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
+
+ \cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176} argue that the
+ clean simulation of reversible Turing machines is a criterion for computing
+ strength of reversible languages. This is kind of what we show.
+
+ \cite{Yokoyama:2007:RPL:1244381.1244404} also show a reversible
+ self-interpreter which is what we show
+
+ we argue and so does \cite{Rendel:2010:ISD:1863523.1863525} that partial iso
+ are a good foundation for reversible programming
 
@@
 
-Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
-properties and various constructions in it. However, for a
-computational model, there has been little focus on actually writing
-programs directly in {{langRevT}}. This paper aims to remedy that by
-presenting a systematic technique of expressing logically reversible
-small-step abstract machines in {{langRevT}}. This paper shows that
-once we have a logically reversible machine in a notation of our
-choice, expressing the machine as an isomorphic interpreter in
-{{langRevT}} can be done systematically and does not present any
-significant conceptual difficulties.
+Previous work on {{langRevT}} \cite{rc2011,James:2012:IE:2103656.2103667}
+studied its properties and various constructions in it. However, for a
+computational model, there has been little focus on actually writing programs
+directly in {{langRevT}}. This paper aims to remedy that by presenting a
+systematic technique of expressing logically reversible small-step abstract
+machines in {{langRevT}}. This paper shows that once we have a logically
+reversible machine in a notation of our choice, expressing the machine as an
+isomorphic interpreter in {{langRevT}} can be done systematically and does
+not present any significant conceptual difficulties.
 
@@
 We briefly review the reversible language {{langRevT}} introduced in our
-previous work~\cite{rc2011,infeffects}. The set of types includes the empty
-type {{0}}, the unit type {{1}}, sum types {{b1+b2}}, product types
-{{b1*b2}}, and recursive types {{@@x.b}}. The set of values {{v}} includes
-{{()}} which is the only value of type {{1}}, {{left v}} and {{right v}}
-which inject {{v}} into a sum type, {{(v1,v2)}} which builds a value of
-product type, and {{<v>}} which is used to construct recursive values. There
-are no values of type {{0}}:
+previous work~\cite{rc2011,James:2012:IE:2103656.2103667}. The set of types
+includes the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
+product types {{b1*b2}}, and recursive types {{@@x.b}}. The set of values
+{{v}} includes {{()}} which is the only value of type {{1}}, {{left v}} and
+{{right v}} which inject {{v}} into a sum type, {{(v1,v2)}} which builds a
+value of product type, and {{<v>}} which is used to construct recursive
+values. There are no values of type {{0}}:
 %subcode{bnf} include main

### 2012-03-15 iso-int/rc-iso-int.tex
intro

--- 2012-03-14 iso-int/rc-iso-int.tex
@@
 
-Reversible languages
+In recent papers~\cite{rc2011,James:2012:IE:2103656.2103667}, we developed a
+pure reversible model of computation that is obtained from the type
+isomorphisms and categorical structures that underlie models of linear logic
+and quantum computing and that treats information as a linear resource that
+can neither be erased nor duplicated. From a programming perspective, our
+model gives rise to a pure (with no embedded computational effects such as
+assignments) reversible programming language {{langRevT}} based on
+\emph{partial isomorphisms}. In more detail, in the recursion-free fragment
+of {{langRevT}}, every program witnesses an isomorphism between two finite
+types; in the full language with recursion, some of these isomorphisms may be
+partial, i.e., may diverge on some inputs.
 
-Main points: 
+In this paper, we investigate the practicality of writing recursive programs
+in {{langRevT}}. Specifically, we assume that we are giving some reversible
+recursive algorithm expressed as a \emph{reversible abstract machine}, and we
+show via a number of systematic steps, how to develop a {{langRevT}} program
+from that abstract machine. Our choice of starting from reversible abstract
+machines is supported by the following observations:
 \begin{itemize}
-\item Make the case for reversibility via type isomorphisms
-\item Type isomorphisms between finite types are straightforward (but
-  tedious)
-\item Type isomorphisms between recursive types are quite difficult. This is
-  what we focus on.
+\item it is an interesting enough class of reversible programs: researchers
+  have invested the effort in manually designing such machines, e.g., the
+  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and theSECD-H~\cite{lorenz});
+\item every reversible interpreter can be realized as such a machine: this
+  means that our class of programs includes self-interpreters which are
+  arguably a measure of the strength of any reversible
+  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
+\item general recursive programs can be systematically transformed to
+  abstract machines: the technique is independent of reversible programs and
+  consists of tranforming general recursion to tail recursion and then
+  applying \emph{fission} to split the program into a driver and a small-step
+  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
 \end{itemize}
-To understand the problem in more detail, consider the problem of converting
-a reversible recursive function into a partial isomorphism. In general such a
-function takes an input, makes several recursive calls, and produces an
-answer. To implement this in a reversible way, the structure of the recursion
-and the number of recursive calls need to be exposed. The key trick is to
-express tease the function into two components: an outer driver that performs
-iteration and an inner state transition function that realizes one
-``unfolding'' of the recursion. In program transformation world, this is
-called, light-weight fission. Using this idea, one can realize any recursive
-function as a partial isomorphism. We show a few examples of the construction
-culminating with the most challenging recursive function: a full interpreter.
 
-Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
-properties and various constructions in it. However, for a
-computational model, there has been little focus on actually writing
-programs directly in {{langRevT}}. This paper aims to remedy that by
-presenting a systematic technique of expressing logically reversible
-small-step abstract machines in {{langRevT}}. This paper shows that
-once we have a logically reversible machine in a notation of our
-choice, expressing the machine as an isomorphic interpreter in
-{{langRevT}} can be done systematically and does not present any
-significant conceptual difficulties.
+To summarize, we assume we are given some reversible abstract machine and we
+show how to derive a {{langRevT}} program that implements the semantics of
+the machine. Our derivation is systematic and expressive. In particular,
+{{langRevT}} can handle machines with stuck states because it is based on
+partial isomorphisms. We illustrate our techniques with simple machines that
+do bounded iteration on numbers, tree traversals, and a meta-circular
+interpreter for {{langRevT}}.
 
-The constructions in this paper apply only to a class of interpreters
-that can be represented by logically reversible small-step operational
-semantics. While, in a formal sense, they are related to Abramsky's
-``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
-to give them a more direct characterization purely in terms of logical
-reversibility of the small-step semantics.
+% To understand the problem in more detail, consider the problem of converting
+% a reversible recursive function into a partial isomorphism. In general such a
+% function takes an input, makes several recursive calls, and produces an
+% answer. To implement this in a reversible way, the structure of the recursion
+% and the number of recursive calls need to be exposed. The key trick is to
+% express tease the function into two components: an outer driver that performs
+% iteration and an inner state transition function that realizes one
+% ``unfolding'' of the recursion. In program transformation world, this is
+% called, light-weight fission. Using this idea, one can realize any recursive
+% function as a partial isomorphism. We show a few examples of the construction
+% culminating with the most challenging recursive function: a full interpreter.
+
+% Previous work on {{langRevT}} \cite{rc2011,James:2012:IE:2103656.2103667}
+% studied its properties and various constructions in it. However, for a
+% computational model, there has been little focus on actually writing programs
+% directly in {{langRevT}}. This paper aims to remedy that by presenting a
+% systematic technique of expressing logically reversible small-step abstract
+% machines in {{langRevT}}. This paper shows that once we have a logically
+% reversible machine in a notation of our choice, expressing the machine as an
+% isomorphic interpreter in {{langRevT}} can be done systematically and does
+% not present any significant conceptual difficulties.
+
+% The constructions in this paper apply only to a class of interpreters
+% that can be represented by logically reversible small-step operational
+% semantics. While, in a formal sense, they are related to Abramsky's
+% ``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
+% to give them a more direct characterization purely in terms of logical
+% reversibility of the small-step semantics.
+
+%% - write program; derive inverse
+%% program~\cite{Gluck:2005,Mu:2002:IFF:648086.757330,conf/aplas/GluckK03,Gluck:2005};
+%% we're not interested in this; we want to directly write partial isomorphisms
 

### 2012-03-15 iso-int/rc-iso-int.tex
minor

@@
 % {{langRevT}}.
-  In our previous work [RC'11, POPL'12], we developed a reversible
-  programming language and established that every computation in it is a
-  (partial) isomorphism that is reversible and that preserves
-  information. The language is founded on type isomorphisms that have a clear
-  categorical semantics but that are awkward as a notation for writing actual
-  programs, especially recursive programs. This paper remedies this aspect by
-  presenting a systematic technique for developing a large and expressive
-  class of reversible recursive programs, that of logically reversible
-  small-step abstract machines. In other words, this paper shows that once we
-  have a logically reversible machine in a notation of our choice, expressing
-  the machine as an isomorphic interpreter can be done systematically and
-  does not present any significant conceptual difficulties. Concretely, we
-  develop several simple interpreters over numbers and addition, move on to
-  tree traversals, and finish with a meta-circular interpreter for our
-  reversible language. Generally, this gives us a means of developing large
-  reversible programs with the ease of reasoning at the level of a
-  conventional small-step semantics.
+  In our previous work, we developed a reversible programming language and
+  established that every computation in it is a (partial) isomorphism that is
+  reversible and that preserves information. The language is founded on type
+  isomorphisms that have a clear categorical semantics but that are awkward
+  as a notation for writing actual programs, especially recursive
+  programs. This paper remedies this aspect by presenting a systematic
+  technique for developing a large and expressive class of reversible
+  recursive programs, that of logically reversible small-step abstract
+  machines. In other words, this paper shows that once we have a logically
+  reversible machine in a notation of our choice, expressing the machine as
+  an isomorphic interpreter can be done systematically and does not present
+  any significant conceptual difficulties. Concretely, we develop several
+  simple interpreters over numbers and addition, move on to tree traversals,
+  and finish with a meta-circular interpreter for our reversible
+  language. Generally, this gives us a means of developing large reversible
+  programs with the ease of reasoning at the level of a conventional
+  small-step semantics.
 \end{abstract}
@@
   have invested the effort in manually designing such machines, e.g., the
-  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and theSECD-H~\cite{lorenz});
+  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz});
 \item every reversible interpreter can be realized as such a machine: this

### 2012-03-15 iso-int/rc-iso-int.tex
Pi

--- 2012-03-15 iso-int/rc-iso-int.tex
@@
 %! @ = \,@\,
+%! * = \times
 %! langRev = \Pi
@@
   ``wave.''
-\begin{multicols}{2}
+\begin{multicols}{4}
 \begin{center}
@@
 \end{center}
+\columnbreak
 \begin{center}
@@
 \end{center}
-\end{multicols}
-\begin{multicols}{2}
+%% \end{multicols}
+%% \begin{multicols}{2}
 \begin{center}
@@
   wires.
-\begin{multicols}{2}
+\begin{multicols}{4}
 \begin{center}
@@
 \end{center}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
-}
-\end{center}
-\end{multicols}
-\begin{multicols}{2}
+%% \end{multicols}
+%% \begin{multicols}{2}
 \begin{center}
@@
 \includegraphics{diagrams/thesis/sum-two-wires.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
 }
@@
   wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
-\includegraphics{diagrams/thesis/assoc.pdf}
-}
-\end{center}
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
+%% \includegraphics{diagrams/thesis/assoc.pdf}
+%% }
+%% \end{center}
 
 \item Commutativity is represented by crisscrossing wires.
-\begin{multicols}{2}
+\begin{multicols}{4}
 \begin{center}
@@
 \end{center}
+%% \end{multicols}
+%% \begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_times_value.pdf}
+}
+\end{center}
+
 \begin{center}
@@
 \end{center}
+
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_plus_value.pdf}
+}
+\end{center}
 \end{multicols}
@@
 verify that the expected types for commutativity are satisfied.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/swap_times_value.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/swap_plus_value.pdf}
-}
-\end{center}
-\end{multicols}
 
@@
   of special interest.
-\begin{multicols}{2}
+\begin{multicols}{4}
 \begin{center}
@@
 \end{center}  
-\end{multicols}
-\begin{multicols}{2}
+%% \end{multicols}
+%% \begin{multicols}{2}
 \begin{center}
@@
   boxes shown below:
-\begin{multicols}{2}
+\begin{multicols}{4}
 \begin{center}
@@
   \includegraphics{diagrams/thesis/factor.pdf}
+\end{center}
+%% \end{multicols}
+%% \begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
 \end{center}
@@
 
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
-\end{center}
-\begin{center}
-  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
-\end{center}
-\end{multicols}
-
-\item \todo{fold, unfold, and trace}
+\item {{fold}} and {{unfold}} are basically drawn as triangular boxes
+  that map the a recursive type to its unfolding.
+
+\item {{trace+}} is drawn as a looped circuit, where the traced value
+  is shown as flowing backward. The semantics and interpretation of
+  {{trace}} are dealt with in detail in previous work (see Sec. 4 of
+  \cite{infeffects}).
+
+\begin{center}
+  \includegraphics{diagrams/thesis/trace_plus.pdf}
+\end{center}
+
 \end{itemize}

### 2012-03-15 iso-int/rc-iso-int.tex
Pi

@@
 
-\item Commutativity is represented by crisscrossing wires.
+\item Commutativity is represented by crisscrossing wires.  By
+  visually tracking the flow of particles on the wires, one can verify
+  that the expected types for commutativity are satisfied.
+
 \begin{multicols}{4}
@@
 \end{multicols}
-
-By visually tracking the flow of particles on the wires, one can
-verify that the expected types for commutativity are satisfied.
 
@@
 
-\item Distributivity and factoring are represented using the dual
-  boxes shown below:
+\item Distributivity and factoring are interesting because they
+  represent interactions between sum and pair types. Distributivity
+  should essentially be thought of as a multiplexer that redirects the
+  flow of {{v:b}} depending on what value inhabits the type {{b1+b2}},
+  as shown below. Factoring is the corresponding adjoint operation.
+
 \begin{multicols}{4}
@@
 \end{multicols}
-
-Distributivity and factoring are interesting because they represent
-interactions between sum and pair types. Distributivity should
-essentially be thought of as a multiplexer that redirects the flow of
-{{v:b}} depending on what value inhabits the type {{b1+b2}}, as shown
-below. Factoring is the corresponding adjoint operation.
 

### 2012-03-16 iso-int/rc-iso-int.tex
redrew the tree diagrams
shorter in pages now
space fitting
_
small corrections

@@
 the completion of the interpreter at this point. The last step in the
-construction involves introducing a {{trace}] operation for iteration of
+construction involves introducing a {{trace}} operation for iteration of
 reduction steps and exposing the start and stop states. We do this by
@@
 
-A closer look however reveals that the machine only defines a \emph{partial}
+A closer look however reveals that the machine defines a \emph{partial}
 isomorphism: not all valid final states can be mapped to valid start
@@
 state. Going backwards, the transitions start as follows 
-{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine 
-gets stuck at a state that is not a valid start state. 
-There are several ways to make this machine reversible. The simplest is to
-add another possible kind of start state {{<n,0,n>}}. Our construction now
-proceeds as expected. After expanding the input to expose $n$ and the output
-to expose both $p$ and $m$, we reach the following configuration:
+{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine gets stuck 
+at a state that is not a valid start state.  There are several ways to deal with this
+situation. The first is to use a partial isomorphism that diverges on the stuck state.
+The second is to add another
+possible kind of start state {{<n,0,n>}} to make the isomorphism total. 
+Assuming the seond solution, our construction now proceeds as
+expected. After expanding the input to expose $n$ and the output to expose
+both $p$ and $m$, we reach the following configuration:
 
@@
 the abstract machine, we need a notion of \emph{tree contexts} to track which
-subtree is currently being explored. The definitions are:
-
+subtree is currently being explored. The definitions are shown on the left:
+
+\begin{multicols}{2}
 %subcode{bnf} include main
@@
 
-Our example machine below traverses a given tree and increments every leaf
-value:
-
 %subcode{opsem} include main
@@
 
-The machine here is a little richer than the ones dealt with
-previously. In particular, we have a two types of machine states --
-{{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds to walking
-down a tree, building up the context in the process. The second
-corresponds to reconstructing the tree from the context and also
-switching focus to any unexplored subtrees in the process. There are
-also two syntactic categories to deal with (trees and tree contexts)
-where previously we only had numbers. The {{fold}} and {{unfold}}
-isomorphisms that we need for trees and tree contexts are:
+\end{multicols}
+
+The reduction rules on the right traverses a given tree and increments
+every leaf value.  The machine here is a little richer than the ones
+dealt with previously. In particular, we have a two types of machine
+states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
+to walking down a tree, building up the context in the process. The
+second corresponds to reconstructing the tree from the context and
+also switching focus to any unexplored subtrees in the process. There
+are also two syntactic categories to deal with (trees and tree
+contexts) where previously we only had numbers. The {{fold}} and
+{{unfold}} isomorphisms that we need for trees and tree contexts are:
 
@@
 
+\begin{multicols}{2}
+%subcode{bnf} include main
+% Machine states = <bool, t, c>
+%
+% Start state = <T, t, []>
+% Stop state =  <F, t, []>
+
 %subcode{opsem} include main
@@
 % <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
-
-And so machine states have the form:
-
-%subcode{bnf} include main
-% Machine states = <bool, t, c>
-% Start state = <T, t, []>
-% Stop state =  <F, t, []>
+\end{multicols}
 
@@
 \begin{center}
-\scalebox{0.9}{
+\scalebox{1.0}{
   \includegraphics{diagrams/tree1.pdf}
@@
 \begin{center}
-\scalebox{0.9}{
+\scalebox{1.1}{
   \includegraphics{diagrams/tree2.pdf}
@@
 
-\begin{center}
-\scalebox{0.9}{
-  \includegraphics{diagrams/pi0-1.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{0.9}{
+%% \begin{center}
+%% \scalebox{0.85}{
+%%   \includegraphics{diagrams/pi0-1.pdf}
+%% }
+%% \end{center}
+
+\begin{center}
+\scalebox{0.85}{
   \includegraphics{diagrams/pi0-2.pdf}

### 2012-03-16 iso-int/rc-iso-int.tex
last section and withim page limit

@@
 adder:
+
+\vspace{-20pt}
+\begin{small}
 \begin{multicols}{2}  
@@
 \end{multicols}
+\vspace{-15pt}
 %subcode{opsem} include main
 % <n+1, p, m> |--> <n, p+1, m+1>
+\end{small}
 
@@
 \begin{center}
-\scalebox{0.9}{
+\scalebox{0.7}{
   \includegraphics{diagrams/adder3.pdf}
@@
 \begin{center}
-\scalebox{0.9}{
+\scalebox{0.7}{
   \includegraphics{diagrams/adder4.pdf}
@@
 
+\vspace{-15pt}
+\begin{small}
 \begin{multicols}{2}
@@
 % {[Rgt t1 c, t2]} &|-->& {[N t1 t2, c]}
-
 \end{multicols}
+\end{small}
 
@@
 
+\begin{small}
 \begin{multicols}{2}
@@
 \end{multicols}
+\end{small}
 
@@
 \subsection{A {{langRevT}} Interpreter}
+
+In the final construction we present a metacircular interpreter for
+{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
+machine with several cases, but the deriving the corresponding
+{{langRevT}} interpreter follows in exactly the same way as
+before. Here is a logically reversible small-step abstract machine for
+{{langRevT}}.
+
+\begin{small}
 
@@
 
+\end{small}
+
+The isomorphic interpreter thus derived is shown below, just before
+the `flipping' step (which makes it harder to read). Each wire has
+been annotated with the reduction rule it corresponds to making it
+easier to compare with the abstract machine.
+
+\begin{center}
+\scalebox{0.85}{
+  \includegraphics{diagrams/pi0-2.pdf}
+}
+\end{center}
+
+\noindent
+A couple of important points:
+
+\begin{enumerate}
+
+\item The definition here shows only the composition combinators;
+  which are the intersting cases. All the primitive isomorphisms are
+  hidden in the `{{iso(v)}}' application in {{rule 1}}. This should be
+  read as `tranformation {{v}} according to the isomorphism
+  {{iso}}'. Implementing this is straightformard and is hence skipped.
+
+\item The abstract machine definition is not very strongly typed in
+  teh following sense: we know that every {{v:b1*b2}} must have the
+  form {{(v1,v2)}} however is cases like {{rule 8}} if we allow any
+  {{v}} to appear the machine will be in an undefined state. The usual
+  way this is handled in any interpreter is to either test to ensure
+  that the runtime value has teh required form or to user s stronger
+  typing discipline such as GADTs.We abstract away from this choice in
+  this presentation.
+\end{enumerate}
+
 %% \begin{center}
@@
 %% \end{center}
-
-\begin{center}
-\scalebox{0.85}{
-  \includegraphics{diagrams/pi0-2.pdf}
-}
-\end{center}
 

### 2012-03-16 iso-int/rc-iso-int.tex
sizes
many small fixes

@@
 
-\item {{fold}} and {{unfold}} are basically drawn as triangular boxes
-  that map the a recursive type to its unfolding.
-
-\item {{trace+}} is drawn as a looped circuit, where the traced value
-  is shown as flowing backward. The semantics and interpretation of
-  {{trace}} are dealt with in detail in previous work (see Sec. 4 of
-  \cite{infeffects}).
+\item Operations {{fold}} and {{unfold}} are specific to the recursive
+  data type and are drawn as triangular boxes. Sec. \ref{sec:bounded}
+  for instance shows the {{fold}}/{{unfold}} for recursive type of
+  natural numbers, {{nat=@@x.(1+x)}}.
+
+\item The {{trace+}} operation is drawn as a looped circuit, where the
+  traced value is shown as flowing backward. The semantics and
+  interpretation of {{trace}} are dealt with in detail in previous
+  work (see Sec. 4 of \cite{infeffects}).
 
@@
 \section{Simple Bounded Number Iteration}
+\label{sec:bounded}
 
@@
 
+\vspace{-15pt}
 \begin{multicols}{2}
@@
 \end{multicols}
-% Small-step operational semantics:
-
+\vspace{-15pt}
 %subcode{opsem} include main
@@
 
-A closer look however reveals that the machine defines a \emph{partial}
-isomorphism: not all valid final states can be mapped to valid start
-states. Indeed consider the configuration {{<0,2,3>}} which is a valid final
-state. Going backwards, the transitions start as follows 
-{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine gets stuck 
-at a state that is not a valid start state.  There are several ways to deal with this
-situation. The first is to use a partial isomorphism that diverges on the stuck state.
-The second is to add another
-possible kind of start state {{<n,0,n>}} to make the isomorphism total. 
-Assuming the seond solution, our construction now proceeds as
-expected. After expanding the input to expose $n$ and the output to expose
-both $p$ and $m$, we reach the following configuration:
+A closer look however reveals that the machine defines a
+\emph{partial} isomorphism: not all valid final states can be mapped
+to valid start states. Indeed consider the configuration {{<0,2,3>}}
+which is a valid final state. Going backwards, the transitions start
+as follows {{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the
+machine gets stuck at a state that is not a valid start state.
+
+\paragraph*{Stuck States.}
+In a more general sense, all abstract machines have to deal with stuch
+states and there are several standard ways of delaing with them:
+\begin{itemize}
+\item \emph{Use better types.} One could augment {{langRevT}} with a
+  richer type system that distinguishes non-zero numbers from those
+  that can be zero, thus emilinating this situation entirely. In the
+  intepreter in Sec. \ref{sec:langrevt-int} for instance one could use
+  a GADT to eliminate stuck states.
+
+\item \emph{Diverge.} Another standard approach in dealing with stuck
+  states is to make the diverge or leave the output
+  undefined/unobservable in some way. For instance {{add1}} defined in
+  {{langRevT}} (see Sec. 4.2 in \cite{James:2012:IE:2103656.2103667})
+  has teh property that {{add1^{dagger} 0}} diverges.
+
+\item \emph{Stop the machine.} In the case of our machine this
+  corresponds to another possible kind of start state {{<n,0,n>}} to
+  make the isomorphism total.  Let us choose this path in the case of
+  this machine.  After expanding the input to expose $n$ and the
+  output to expose both $p$ and $m$, we wire the machine corresponding
+  to the reduction rule.
+
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/adder3.pdf}
+}
+\end{center}
+
+\end{itemize}
+
+Note that there are indeed two start states in the interpreter. We can
+slide the two sides of the diagram and tie the knot using {{trace}} to
+get the desired interpreter:
+
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/adder4.pdf}
+}
+\end{center}
+
+%% We have also connected the matching input and output wires. At this
+%% point, we have exposed both start states and the stop state.
+
+%% There are several ways to deal with this situation. The first is to
+%% use a partial isomorphism that diverges on the stuck state.
 
@@
 
-\begin{center}
-\scalebox{0.7}{
-  \includegraphics{diagrams/adder3.pdf}
-}
-\end{center}
-
-We have also connected the matching input and output wires. At this point, we
-have exposed both start states and the stop state. 
 % \begin{enumerate}
@@
 % \end{enumerate}
-We can slide the two sides of the diagram and tie the knot using {{trace}} to
-get the desired combinator:
-
-\begin{center}
-\scalebox{0.7}{
-  \includegraphics{diagrams/adder4.pdf}
-}
-\end{center}
 
@@
 \subsection{A {{langRevT}} Interpreter}
+\label{sec:langrevt-int}
 
@@
 
-\begin{small}
+\begin{scriptsize}
 
@@
 
-\end{small}
+\end{scriptsize}
+
+\begin{center}
+\scalebox{0.85}{
+  \includegraphics{diagrams/pi0-2.pdf}
+}
+\end{center}
 
@@
 been annotated with the reduction rule it corresponds to making it
-easier to compare with the abstract machine.
-
-\begin{center}
-\scalebox{0.85}{
-  \includegraphics{diagrams/pi0-2.pdf}
-}
-\end{center}
-
-\noindent
-A couple of important points:
+easier to compare with the abstract machine.  A couple of important
+points:
 
@@
 
-\item The abstract machine definition is not very strongly typed in
-  teh following sense: we know that every {{v:b1*b2}} must have the
-  form {{(v1,v2)}} however is cases like {{rule 8}} if we allow any
-  {{v}} to appear the machine will be in an undefined state. The usual
-  way this is handled in any interpreter is to either test to ensure
-  that the runtime value has teh required form or to user s stronger
-  typing discipline such as GADTs.We abstract away from this choice in
-  this presentation.
+\item Stuck states in this machine correspond to values not matching
+  the shape expected by their types. They can be handled as decribed
+  in Sec. \ref{sec:adder} using GADTs (which eliminates them
+  entirely), divergence, or adding extra halting states.
+
 \end{enumerate}
@@
 
+\begin{scriptsize}
 \bibliographystyle{splncs03} 
 \bibliography{cites}  
+\end{scriptsize}
 \end{document}

### 2012-03-16 iso-int/rc-iso-int.tex
small fixes
man ixes

--- 2012-03-16 iso-int/rc-iso-int.tex
@@
   have invested the effort in manually designing such machines, e.g., the
-  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz});
+  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
 \item every reversible interpreter can be realized as such a machine: this
@@
 \begin{multicols}{4}
-\begin{center}
+%\begin{center}
 \scalebox{0.95}{
@@
 }
-\end{center}
+
+%\end{center}
 %% \end{multicols}
 %% \begin{multicols}{2}
-\begin{center}
+%\begin{center}
 \scalebox{0.95}{
@@
 }
-\end{center}
-\begin{center}
+
+%\end{center}
+%\begin{center}
 \scalebox{0.95}{
@@
 }
-\end{center}
+%\end{center}
 \end{multicols}
@@
 
-\item {{fold}} and {{unfold}} are basically drawn as triangular boxes
-  that map the a recursive type to its unfolding.
-
-\item {{trace+}} is drawn as a looped circuit, where the traced value
-  is shown as flowing backward. The semantics and interpretation of
-  {{trace}} are dealt with in detail in previous work (see Sec. 4 of
-  \cite{infeffects}).
+\item Operations {{fold}} and {{unfold}} are specific to the recursive
+  data type and are drawn as triangular boxes. For instance,
+  Sec. \ref{sec:bounded} shows {{fold}}/{{unfold}} isomorphisms for
+  numbers, {{nat=@@x.(1+x)}}.
+
+\item The {{trace+}} operation is drawn as a looped circuit, where the
+  traced type {{b1}} is shown as flowing backward. The semantics and
+  interpretation of {{trace}} are dealt with in detail in previous
+  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).
 
@@
 \subsection{Adder} 
+\label{sec:adder}
 
@@
   mediating computations. 
-\item Slide the two sections to expose the start and stop state and introduce
-  a {{trace}} to iterate the construction.
+\item Slide the two sections to expose the start and stop state and
+  introduce a {{trace}} to iterate the construction. This is always a
+  mechanical process completely independant of the logic of the machine.
 \end{enumerate}
@@
 
-A closer look however reveals that the machine defines a \emph{partial}
-isomorphism: not all valid final states can be mapped to valid start
-states. Indeed consider the configuration {{<0,2,3>}} which is a valid final
-state. Going backwards, the transitions start as follows 
-{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine gets stuck 
-at a state that is not a valid start state.  There are several ways to deal with this
-situation. The first is to use a partial isomorphism that diverges on the stuck state.
-The second is to add another
-possible kind of start state {{<n,0,n>}} to make the isomorphism total. 
-Assuming the seond solution, our construction now proceeds as
-expected. After expanding the input to expose $n$ and the output to expose
-both $p$ and $m$, we reach the following configuration:
+A closer look however reveals that the machine defines a
+\emph{partial} isomorphism: not all valid final states can be mapped
+to valid start states. Indeed consider the configuration {{<0,2,3>}}
+which is a valid final state. Going backwards, the transitions start
+as follows {{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the
+machine gets stuck at a state that is not a valid start state.
+
+\paragraph*{Stuck States.}
+All abstract machines, not just those written in {{langRevT}}, have to
+deal with stuck states. There are several standard ways of dealing
+with them:
+\begin{itemize}
+\item \emph{Use better types.}  One could augment {{langRevT}} with a
+  richer type system that distinguishes non-zero numbers from those
+  that can be zero, thereby eliminating the {{sub1 0}} situation
+  entirely. Similarly, in the meta-circular intepreter in
+  Sec. \ref{sec:langrevt-int} one could use GADTs to eliminate stuck
+  states.
+
+\item \emph{Divergence.}  Another standard approach in dealing with
+  stuck states is to make the machine diverge or leave the output
+  undefined/unobservable in some way. The {{add1}} combinator defined
+  in Sec. 4.2 of our related work \cite{James:2012:IE:2103656.2103667}
+  has the property that its dual {{sub1}} diverges on input 0. Hence
+  we could just use {{add1}} here.
+
+\item \emph{Stop the machine.}  If we chose to stop the machine at a
+  stuck state, we are essentially treating states of the form
+  {{<n,0,n>}} as valid stop states for reverse execution. This also
+  has the consequence of making our isomorphism total and also gives
+  us two valid start states in the case of forward execution. If we
+  choose this approach, the machine would look as shown below by the
+  end of {{step 3}}:
+
+%% In the case of our machine this corresponds to another possible kind
+%% of start state {{<n,0,n>}} to make the isomorphism total.  Let us
+%% choose this path in the case of this machine.  After expanding the
+%% input to expose $n$ and the output to expose both $p$ and $m$, we wire
+%% the machine corresponding to the reduction rule.
+
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/adder3.pdf}
+}
+\end{center}
+
+\end{itemize}
+
+Note that there are indeed two ``start states'' in the interpreter. As before, we can
+slide the two sides of the diagram and tie the knot using {{trace}} to
+get the desired interpreter:
+
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/adder4.pdf}
+}
+\end{center}
+
+More generally, how the programmer handles stuck states depends on the
+specific program. For some machines it may be feasible to use a
+stronger type system while for others, such as our tree example in
+Sec. \ref{sec:tree}, it is acceptable to diverge.
+
+%% We have also connected the matching input and output wires. At this
+%% point, we have exposed both start states and the stop state.
+
+%% There are several ways to deal with this situation. The first is to
+%% use a partial isomorphism that diverges on the stuck state.
 
@@
 
-\begin{center}
-\scalebox{0.9}{
-  \includegraphics{diagrams/adder3.pdf}
-}
-\end{center}
-
-We have also connected the matching input and output wires. At this point, we
-have exposed both start states and the stop state. 
 % \begin{enumerate}
@@
 % \end{enumerate}
-We can slide the two sides of the diagram and tie the knot using {{trace}} to
-get the desired combinator:
-
-\begin{center}
-\scalebox{0.9}{
-  \includegraphics{diagrams/adder4.pdf}
-}
-\end{center}
 
@@
 \subsection{Tree Traversal}
-
-The type of binary trees we use is {{@@x.nat+x*x}}, i.e., binary trees with
+\label{sec:tree}
+
+The type of binary trees we use is {{@@x.(nat+x*x)}}, i.e., binary trees with
 no information at the nodes and with natural numbers at the leaves. To define
@@
 % Tree, t = L n | N t t
-% Tree Contexts, c = Lft c t | Rgt t c | []
+% Tree Contexts, c = [] | Lft c t | Rgt t c
 %
@@
 
+\begin{small}
+
 %subcode{bnf} include main
@@
 % unfold_c :& c &<->& 1 + c * t + t * c ~~~~~~&: fold_c
+\end{small}
 
@@
 represent these steps using thin vertical rectangle. Also we will introduce
-the convention that the component that is being expanded or collapsed will be
-marked by using {{[^ ^]}} and the components that are being generated will be
+the convention that the component that is being expanded (or constructed) will be
+marked by using {{[^ ^]}} and the components that are being generated (or consumed) will be
 marked by {{[v v]}}. For example, given a value of type {{tree*nat}}, the
@@
 isomorphic interpreter is to recognize that the two possible kinds of machine
-states simply hide an implicit {{bool}}. We make this explicit:
-
+states simply hide an implicit {{bool=T~|~F}}. We make this explicit:
+
+\begin{small}
 \begin{multicols}{2}
@@
 \item When we encounter a leaf we would like to increment its value and move
-  to the corresponding {{F}} machine state. For the sake of simplicity in
+  to the corresponding {{F}} machine state. For the sake of simplicity, in
   this interpreter we won't be concerned with exposing stuck states on
@@
 This essentially completes the construction of the (partially)
-isomorphic tree-traversal interpreter. (Partially because of the
-implicit non-termination introduced by the {{add1}}).
+isomorphic tree-traversal interpreter, except for the boring final
+step. (Partially because of the implicit non-termination introduced by
+the {{add1}}).
 
@@
 \subsection{A {{langRevT}} Interpreter}
+\label{sec:langrevt-int}
+
+In the final construction we present a meta-circular interpreter for
+{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
+machine with several cases, but the deriving the corresponding
+{{langRevT}} interpreter follows in exactly the same way as
+before. Here is a logically reversible small-step abstract machine for
+{{langRevT}} and the derived isomorphic interpreter with the
+reductions labelled.
+
+\begin{scriptsize}
 
@@
 
+\vspace{-15pt}
 %subcode{opsem} include main
@@
 
+\end{scriptsize}
+
+\begin{center}
+\scalebox{0.85}{
+  \includegraphics{diagrams/pi0-2.pdf}
+}
+\end{center}
+
+%% The isomorphic interpreter thus derived is shown below, just before
+%% the `flipping' step (which makes it harder to read). Each wire has
+%% been annotated with the reduction rule it corresponds to making it
+%% easier to compare with the abstract machine.
+
+The derivation of the {{langRevT}} interpreter, while tedious, is
+entirely straightforward.  A couple of points however need clarification:
+
+\begin{enumerate}
+
+\item The definition here shows only the handling of the composition
+  combinators, which are the interesting cases. All the primitive
+  isomorphisms are hidden in the `{{iso(v)}}' application in
+  {{rule~1}}. This should be read as `tranform {{v}} according to the
+  primitive isomorphism {{iso}}'. Implementing this is straightformard
+  and is hence skipped.
+
+\item Stuck states in this machine correspond to runtime values not
+  matching their expected types. They can be handled as decribed in
+  Sec. \ref{sec:adder} using GADTs (which eliminates them entirely),
+  divergence, or adding extra halting states. We have abtracted from
+  this choice and marked the relevant cases with combinators labelled
+  `{{cast}}'.
+
+\end{enumerate}
+
 %% \begin{center}
@@
 
+We have developed a technique for the systematic development of
+{{langRevT}} programs from logically reversible small-step abstract
+machines. Since techniques for developing small-step interpreters are
+well developed this allows for the direct development of a large class
+of {{langRevT}} programs. We have demonstarted the effectiveness of
+this approach by deriving meta-circular interpreter for {{langRevT}}
+in this style.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\begin{scriptsize}
 \bibliographystyle{splncs03} 
 \bibliography{cites}  
+\end{scriptsize}
 \end{document}

### 2012-03-16 iso-int/rc-iso-int.tex
spell check

@@
   abstract machines: the technique is independent of reversible programs and
-  consists of tranforming general recursion to tail recursion and then
+  consists of transforming general recursion to tail recursion and then
   applying \emph{fission} to split the program into a driver and a small-step
@@
   introduce a {{trace}} to iterate the construction. This is always a
-  mechanical process completely independant of the logic of the machine.
+  mechanical process completely independent of the logic of the machine.
 \end{enumerate}
@@
   that can be zero, thereby eliminating the {{sub1 0}} situation
-  entirely. Similarly, in the meta-circular intepreter in
+  entirely. Similarly, in the meta-circular interpreter in
   Sec. \ref{sec:langrevt-int} one could use GADTs to eliminate stuck
@@
 We show the generality of our construction by applying it to two non-trivial
-examples: a tree traversal and a meta-circulator interpreter for
+examples: a tree traversal and a meta-circular interpreter for
 {{langRevT}}.
@@
 the abstract machine, we need a notion of \emph{tree contexts} to track which
-subtree is currently being explored. The definitions are shown on the left:
+sub-tree is currently being explored. The definitions are shown on the left:
 
@@
 second corresponds to reconstructing the tree from the context and
-also switching focus to any unexplored subtrees in the process. There
+also switching focus to any unexplored sub-trees in the process. There
 are also two syntactic categories to deal with (trees and tree
@@
 {{langRevT}} and the derived isomorphic interpreter with the
-reductions labelled.
+reductions labeled.
 
@@
   isomorphisms are hidden in the `{{iso(v)}}' application in
-  {{rule~1}}. This should be read as `tranform {{v}} according to the
-  primitive isomorphism {{iso}}'. Implementing this is straightformard
+  {{rule~1}}. This should be read as `transform {{v}} according to the
+  primitive isomorphism {{iso}}'. Implementing this is straightforward
   and is hence skipped.
@@
 \item Stuck states in this machine correspond to runtime values not
-  matching their expected types. They can be handled as decribed in
+  matching their expected types. They can be handled as described in
   Sec. \ref{sec:adder} using GADTs (which eliminates them entirely),
-  divergence, or adding extra halting states. We have abtracted from
-  this choice and marked the relevant cases with combinators labelled
+  divergence, or adding extra halting states. We have abstracted from
+  this choice and marked the relevant cases with combinators labeled
   `{{cast}}'.
@@
 well developed this allows for the direct development of a large class
-of {{langRevT}} programs. We have demonstarted the effectiveness of
+of {{langRevT}} programs. We have demonstrated the effectiveness of
 this approach by deriving meta-circular interpreter for {{langRevT}}

### 2012-03-16 iso-int/rc-iso-int.tex
one more fix

@@
 
-Our goal is to implement this abstract machine in {{langRevT}}, i.e., our
-goal is to write a combinator {{c : nat * nat <-> nat * nat}} where {{nat}}
-is an abbreviation for the recursive type {{@@x.1+x}}. The combinator should
-map {{(n,0)}} to {{(0,n)}} by precisely mimicking the steps of the abstract
-machine. Let us analyze this desired combinator {{c}} in detail starting from
-its interface:
+Our goal is to implement this abstract machine in {{langRevT}}: We
+start by writing a combinator {{c : nat * nat <-> nat * nat}} that
+executes a `small step'. Here {{nat}} is an abbreviation for the
+recursive type {{@@x.1+x}}. The combinator, when iterated, should map
+{{(n,0)}} to {{(0,n)}} by precisely mimicking the steps of the
+abstract machine. Let us analyze this desired combinator {{c}} in
+detail starting from its interface:
 
@@
 exposed start state essentially requires the initial value of {{n}} to
-start the machine.
+start the machine and the exposed stop state returns any result values
+that have been computed.
 

### 2012-03-16 iso-int/rc-iso-int.tex
revised

@@
   isomorphisms that have a clear categorical semantics but that are awkward
-  as a notation for writing actual programs, especially recursive
-  programs. This paper remedies this aspect by presenting a systematic
-  technique for developing a large and expressive class of reversible
-  recursive programs, that of logically reversible small-step abstract
-  machines. In other words, this paper shows that once we have a logically
-  reversible machine in a notation of our choice, expressing the machine as
-  an isomorphic interpreter can be done systematically and does not present
-  any significant conceptual difficulties. Concretely, we develop several
-  simple interpreters over numbers and addition, move on to tree traversals,
-  and finish with a meta-circular interpreter for our reversible
-  language. Generally, this gives us a means of developing large reversible
-  programs with the ease of reasoning at the level of a conventional
-  small-step semantics.
+  as a notation for writing actual programs, especially recursive ones. This
+  paper remedies this aspect by presenting a systematic technique for
+  developing a large and expressive class of reversible recursive programs,
+  that of logically reversible small-step abstract machines. In other words,
+  this paper shows that once we have a logically reversible machine in a
+  notation of our choice, expressing the machine as an isomorphic interpreter
+  can be done systematically and does not present any significant conceptual
+  difficulties. Concretely, we develop several simple interpreters over
+  numbers and addition, move on to tree traversals, and finish with a
+  meta-circular interpreter for our reversible language. Generally, this
+  gives us a means of developing large reversible programs with the ease of
+  reasoning at the level of a conventional small-step semantics.
 \end{abstract}
@@
   arguably a measure of the strength of any reversible
-  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
+  language~\cite{Axelsen:2011:RPC:1987171.1987176,Yokoyama:2008:PRP,Yokoyama:2007:RPL:1244381.1244404};
 \item general recursive programs can be systematically transformed to
@@
 \paragraph*{Graphical Language.}
-The syntactic notation above is often obscure and hard to read.  Following
-the tradition established for monoidal
+Following the tradition established for monoidal
 categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
@@
 
-\item Operations {{fold}} and {{unfold}} are specific to the recursive
+\item Operations {{fold}} and {{unfold}} are specific to each recursive
   data type and are drawn as triangular boxes. For instance,
@@
 \item The {{trace+}} operation is drawn as a looped circuit, where the
-  traced type {{b1}} is shown as flowing backward. The semantics and
-  interpretation of {{trace}} are dealt with in detail in previous
-  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).
+  traced type {{b1}} is shown as flowing backward. 
+%% The semantics and
+%%  interpretation of {{trace}} are dealt with in detail in previous
+%%  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).
 
@@
 
-Our goal is to implement this abstract machine in {{langRevT}}: We
-start by writing a combinator {{c : nat * nat <-> nat * nat}} that
-executes a `small step'. Here {{nat}} is an abbreviation for the
-recursive type {{@@x.1+x}}. The combinator, when iterated, should map
-{{(n,0)}} to {{(0,n)}} by precisely mimicking the steps of the
-abstract machine. Let us analyze this desired combinator {{c}} in
-detail starting from its interface:
+Our goal is to implement this abstract machine in {{langRevT}}: We start by
+writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
+of the machine transition. Here {{nat}} is an abbreviation for the recursive
+type {{@@x.1+x}}. The combinator, when iterated, should map {{(n,0)}} to
+{{(0,n)}} by precisely mimicking the steps of the abstract machine. Let us
+analyze this desired combinator {{c}} in detail starting from its interface:
 
@@
 
-The triangle on the left denotes the {{unfold}} isomorphism, which
-works as follows: If the number {{n}} is zero, the output is the top
-branch which has type {{1}}. If the number was non-zero, the output is
-on the bottom branch and has value {{n-1}}. The triangle on the right
-is the dual {{fold}} isomorphism.  Let us examine all the possible
-machine states by introducing distribution operations and any required
-swaps:
+The triangle on the left denotes the {{unfold}} isomorphism, which works as
+follows: If the number {{n}} is zero, the output is the top branch which has
+type {{1}}. If the number was non-zero, the output is on the bottom branch
+and has value {{n-1}}. The triangle on the right is the dual {{fold}}
+isomorphism.  We now use distribution to isolate each possible machine state:
 
@@
 
-The construction above already gives us some insight into how we can
-implement our machine's one reduction step. We want the output {{n'}} to be
-{{n+1}} where~{{n}} is the first component of the input and similarly the
-output {{m'}} to be {{m+1}} where {{m}} is the second component of the input.
-We can express these equalities by connecting wires, i.e., by connecting the
-{{n-1}} wire to {{n'}} and the {{m}} wire to {{m'-1}}:
+At this point, it is clear that the machine's transition consists connecting
+the {{n-1}} input wire to {{n'}} on the output side and the {{m}} input wire
+to {{m'-1}} on the output side:
 
@@
 
-The branch labeled {{((), m)}} corresponds to the machine state {{<0,m>}} and
-hence looks like the stop state of the machine. Similarly the branch labeled
-{{((), n')}} corresponds to the start state of the machine. We are close to
-the completion of the interpreter at this point. The last step in the
-construction involves introducing a {{trace}} operation for iteration of
-reduction steps and exposing the start and stop states. We do this by
-``sliding'' the two sections of the diagram as shown below:
+We are now close to the completion of the interpreter.  The branch labeled
+{{((), m)}} corresponds to the machine state {{<0,m>}} which is the stop
+state of the machine. Similarly the branch labeled {{((), n')}} corresponds
+to the start state of the machine.  The last step in the construction
+involves introducing a {{trace}} operation for iterating the small step
+realized so far. We do this by first ``sliding'' the two sections of the
+diagram as shown below:
 
@@
 
-The introduced {{trace}} feeds back related machine states. The
-exposed start state essentially requires the initial value of {{n}} to
-start the machine and the exposed stop state returns any result values
-that have been computed.
+This rearrangement exposes that the middle connections were really recursive
+calls to the machine transition. The final figure is given below:
+
+%% The introduced {{trace}} feeds back related machine states. The exposed start
+%% state essentially requires the initial value of {{n}} to start the machine
+%% and the exposed stop state returns any result values that have been computed.
 
@@
 as follows {{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the
-machine gets stuck at a state that is not a valid start state.
+machine gets stuck at a state that is not a valid start state. This 
+is a general problem that we discuss below in detail.
 
 \paragraph*{Stuck States.}
-All abstract machines, not just those written in {{langRevT}}, have to
-deal with stuck states. There are several standard ways of dealing
-with them:
+The type systems of most languages are not expressive enough to encode the
+precise domain and range of a function. For example, in most typed languages,
+division by zero is considered type-correct and the runtime system is
+required to deal with such an error. A stuck state of an abstract machine is
+just a manifestation of this general problem. The common solutions are:
 \begin{itemize}
-\item \emph{Use better types.}  One could augment {{langRevT}} with a
-  richer type system that distinguishes non-zero numbers from those
-  that can be zero, thereby eliminating the {{sub1 0}} situation
+\item \emph{Use a more expressive type system.}  One could augment
+  {{langRevT}} with a richer type system that distinguishes non-zero numbers
+  from those that can be zero, thereby eliminating the {{sub1 0}} situation
   entirely. Similarly, in the meta-circular interpreter in
-  Sec. \ref{sec:langrevt-int} one could use GADTs to eliminate stuck
+  Sec.~\ref{sec:langrevt-int}, one could use generalized abstract data types
+  (GADTs)~\cite{phantom,Xi:2003:GRD:604131.604150}) to eliminate the stuck
   states.
 
-\item \emph{Divergence.}  Another standard approach in dealing with
-  stuck states is to make the machine diverge or leave the output
-  undefined/unobservable in some way. The {{add1}} combinator defined
-  in Sec. 4.2 of our related work \cite{James:2012:IE:2103656.2103667}
-  has the property that its dual {{sub1}} diverges on input 0. Hence
-  we could just use {{add1}} here.
-
-\item \emph{Stop the machine.}  If we chose to stop the machine at a
-  stuck state, we are essentially treating states of the form
-  {{<n,0,n>}} as valid stop states for reverse execution. This also
-  has the consequence of making our isomorphism total and also gives
-  us two valid start states in the case of forward execution. If we
-  choose this approach, the machine would look as shown below by the
-  end of {{step 3}}:
+\item \emph{Diverge.}  Another standard approach in dealing with stuck states
+  is to make the machine diverge or leave the output undefined or
+  unobservable in some way. If the specific case of the machine above, we can
+  use a primitive {{add1}} whose dual {{sub1}} is undefined when applied to
+  {{0}}. (Such a primitive is definable in
+  {{langRevT}}~\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}.)
+
+\item \emph{Stop the machine.}  Alternatively, we can consider the stuck
+  state as a valid final state. In the case of the example above, we would
+  treat states of the form {{<n,0,n>}} as valid stop states for reverse
+  execution. This gives us two valid start states in the case of forward
+  execution and makes the isomorphism total. If we chose this approach, the
+  machine would look as shown below by the end of {{step 3}}:
 
@@
 
-More generally, how the programmer handles stuck states depends on the
-specific program. For some machines it may be feasible to use a
-stronger type system while for others, such as our tree example in
-Sec. \ref{sec:tree}, it is acceptable to diverge.
+% More generally, how the programmer handles stuck states depends on the
+% specific program. For some machines it may be feasible to use a
+% stronger type system while for others, such as our tree example in
+% Sec. \ref{sec:tree}, it is acceptable to diverge.
 
@@
 the abstract machine, we need a notion of \emph{tree contexts} to track which
-sub-tree is currently being explored. The definitions are shown on the left:
+subtree is currently being explored. The definitions are shown on the left:
 
@@
 
-The reduction rules on the right traverses a given tree and increments
+The reduction rules on the right traverse a given tree and increment
 every leaf value.  The machine here is a little richer than the ones
@@
 second corresponds to reconstructing the tree from the context and
-also switching focus to any unexplored sub-trees in the process. There
+also switching focus to any unexplored subtrees in the process. There
 are also two syntactic categories to deal with (trees and tree
@@
 On the input side, for {{T}} states, we have expanded the tree component and
-for {{F}} states we have expanded the tree contexts. Similarly we have done
+for {{F}}~states we have expanded the tree contexts. Similarly we have done
 the opposite on the output side exactly matching up what the abstract machine
@@
   to the corresponding {{F}} machine state. For the sake of simplicity, in
-  this interpreter we won't be concerned with exposing stuck states on
-  reverse execution caused by {{sub1 0}}. So we are free to use the {{add1}}
-  combinator.
+  this interpreter we won't be concerned with exposing stuck states: we
+  simply use {{add1}} whose adjoint {{sub1}} diverges when applied to {{0}}.
 
 \item For all the other reduction rules, it is a straightforward mapping of
-  related states following the reduction rules.
+  related states following the reduction rules. For readability, we have
+  annotated the diagram below with the names of the reduction rules and we
+  have included subscripts on the various {{t}}s to indicate any implicit
+  swaps that should be inserted.
 
 \end{enumerate}
-
-The required reduction rules have been labeled on the construction
-belong. We have also included subscripts on the various {{t}}s to
-indicate any swaps that may need to be inserted.
 
@@
 
-This essentially completes the construction of the (partially)
-isomorphic tree-traversal interpreter, except for the boring final
-step. (Partially because of the implicit non-termination introduced by
-the {{add1}}).
+This essentially completes the construction of the (partially) isomorphic
+tree-traversal interpreter, except for the final step which slides the input
+and output sides.
 
@@
 In the final construction we present a meta-circular interpreter for
-{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
-machine with several cases, but the deriving the corresponding
-{{langRevT}} interpreter follows in exactly the same way as
-before. Here is a logically reversible small-step abstract machine for
-{{langRevT}} and the derived isomorphic interpreter with the
-reductions labeled.
+{{langRevT}} written in~{{langRevT}}. This is a a non-trivial abstract
+machine with several cases, but the derivation follows in exactly the same
+way as before. Here is a logically reversible small-step abstract machine for
+{{langRevT}} and the derived isomorphic interpreter with the reductions
+labeled.
 
@@
   combinators, which are the interesting cases. All the primitive
-  isomorphisms are hidden in the `{{iso(v)}}' application in
-  {{rule~1}}. This should be read as `transform {{v}} according to the
-  primitive isomorphism {{iso}}'. Implementing this is straightforward
-  and is hence skipped.
+  isomorphisms are hidden in the {{iso(v)}} application in {{rule~1}}. This
+  should be read as ``transform {{v}} according to the primitive isomorphism
+  {{iso}}.'' Implementing this is straightforward and is hence skipped.
 
@@
   this choice and marked the relevant cases with combinators labeled
-  `{{cast}}'.
+  {{cast}}.
 

### 2012-03-22 iso-int/rc-iso-int.tex

@@
 \begin{center}
-\scalebox{0.85}{
+\scalebox{0.9}{
   \includegraphics{diagrams/pi0-2.pdf}
@@
 
-We have developed a technique for the systematic development of
+We have developed a technique for the systematic derivation of
 {{langRevT}} programs from logically reversible small-step abstract
-machines. Since techniques for developing small-step interpreters are
-well developed this allows for the direct development of a large class
-of {{langRevT}} programs. We have demonstrated the effectiveness of
-this approach by deriving meta-circular interpreter for {{langRevT}}
-in this style.
+machines. Since techniques for devising small-step interpreters are
+well known, this allows for the direct development of a large class of
+{{langRevT}} programs.  We have demonstrated the effectiveness of this
+approach by deriving a meta-circular interpreter for {{langRevT}}.
 

### 2012-03-31 code/dual/Test.hs
added infinite stream generator

@@ -239,4 +239,29 @@ addThirds = Factor
             :.: CommuteTimes
             :.: EpsTimes
 
-
+-- Algebra
+-- 1 
+-- (-1 + 1) + 1 
+-- -1 + (1 + 1)
+-- -1 + 1 * 2
+-- -1 + (1/2 * 2) * 2
+-- -1 + (1/2 * (2 * 2))
+-- -1 + (1/2 * ((1+1) + 2) -- swap/not
+-- -1 + (1/2 * (2 * 2))
+-- -1 + (1/2 * 2) * 2
+-- -1 + 1 * 2
+-- -1 + (1 + 1)
+-- (-1 + 1) + 1 
+-- 1 
+inf_stream :: () :<=> ()
+inf_stream = trace (FoldB :.: inner :.: UnfoldB)
+    where 
+      inner :: Bool :<=> Bool 
+      inner = 
+          UnitI
+          :.: (EtaTimes :*: Id)
+          :.: AssocTimesR
+          :.: (Id :*: cnot)
+          :.: AssocTimesL
+          :.: (EpsTimes :*: Id)
+          :.: UnitE

### 2012-04-01 code/dual/Test.hs
add inf_three

@@ -1,4 +1,4 @@
-{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts -XScopedTypeVariables #-} -- 7.0.1
 
 import Dual 
 
@@ -246,7 +246,7 @@ addThirds = Factor
 -- -1 + 1 * 2
 -- -1 + (1/2 * 2) * 2
 -- -1 + (1/2 * (2 * 2))
--- -1 + (1/2 * ((1+1) + 2) -- swap/not
+-- -1 + (1/2 * ((1+1) * 2) -- swap/not
 -- -1 + (1/2 * (2 * 2))
 -- -1 + (1/2 * 2) * 2
 -- -1 + 1 * 2
@@ -262,6 +262,18 @@ inf_stream = trace (FoldB :.: inner :.: UnfoldB)
           :.: (EtaTimes :*: Id)
           :.: AssocTimesR
           :.: (Id :*: cnot)
+          :.: AssocTimesL
+          :.: (EpsTimes :*: Id)
+          :.: UnitE
+
+inf_three :: (Three :<=> Three) -> Either () () :<=> Either () ()
+inf_three f = trace (FoldThree :.: inner :.: UnfoldThree)
+    where 
+      inner = 
+          UnitI
+          :.: (EtaTimes :*: Id)
+          :.: AssocTimesR
+          :.: (Id :*: controlled f)
           :.: AssocTimesL
           :.: (EpsTimes :*: Id)
           :.: UnitE

### 2012-04-03 code/modular/Lang.hs
added jacques code

@@ -1,0 +1,785 @@
+-- {-# OPTIONS_GHC -XGADTs -XTypeOperators #-} -- 7.0.1
+{-# LANGUAGE ScopedTypeVariables #-} -- 7.0.3, 6.12.1
+
+-- Based on code
+-- Copyright (c) 2011, William J. Bowman, Roshan P. James, and Amr
+-- Sabry. The code is released under the MIT license.
+--
+-- Original code used GADTs.  Now do 'tagless final' using classes.
+--   (Jacques Carette)
+-- This code tested with GHC version 7.0.3
+
+module Lang where
+
+import Prelude hiding (id)
+
+class Pi eq where
+-- Congruence
+  id :: eq a a  -- refl
+  (%.) :: eq a b -> eq b c -> eq a c -- trans
+-- defn of (+) and (*)
+  (%*) :: eq a b -> eq c d -> eq (a,c) (b,d)
+  (%+) :: eq a b -> eq c d -> eq (Either a c) (Either b d)
+-- (+) is associative, commutative, and has a unit
+  plusZeroL :: eq (Either Zero a) a
+  plusZeroR :: eq a (Either Zero a)
+  commutePlus :: eq (Either a b) (Either b a)
+  assocPlusL :: eq (Either a (Either b c)) (Either (Either a b) c)
+  assocPlusR :: eq (Either (Either a b) c) (Either a (Either b c))
+-- (*) is associative, commutative, and has a unit
+  timesOneL    :: eq ((), a)  a
+  timesOneR    :: eq a ((), a)
+  commuteTimes :: eq (a,b) (b,a) 
+  assocTimesL  :: eq (a,(b,c)) ((a,b),c)
+  assocTimesR  :: eq ((a,b),c) (a,(b,c))
+-- (*) distributes over (+) 
+  timesZeroL  :: eq (Zero, a) Zero
+  timesZeroR  :: eq Zero (Zero, a)
+  distribute  :: eq (Either b c, a) (Either (b, a) (c, a))
+  factor      :: eq (Either (b, a) (c, a)) (Either b c, a)
+-- Encoding of booleans
+  foldB   :: eq (Either () ()) Bool
+  unfoldB :: eq Bool (Either () ())
+-- Encoding of natural numbers using the isorecursive type (mu x.1+x)
+  foldN   :: eq (Either () Int) Int
+  unfoldN :: eq Int (Either () Int)
[…]

### 2012-04-03 code/modular/Model.hs

@@ -1,0 +1,187 @@
+-- {-# OPTIONS_GHC -XGADTs -XTypeOperators #-} -- 7.0.1
+{-# LANGUAGE GADTs, TypeOperators, ScopedTypeVariables, EmptyDataDecls #-} -- 7.0.3, 6.12.1
+
+-- Based on code
+-- Copyright (c) 2011, William J. Bowman, Roshan P. James, and Amr
+-- Sabry. The code is released under the MIT license.
+--
+-- Original code used GADTs.  Now do 'tagless final' using classes.
+-- 
+-- This code tested with GHC version 7.0.3
+
+module Model where
+
+import Prelude hiding (id)
+import Lang
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+data a :<=> b where 
+-- Congruence
+  Id    :: a :<=> a
+  Adj   :: (a :<=> b) -> (Neg b :<=> Neg a)
+  (:.:) :: (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  PlusZeroL   :: Either Zero a :<=> a
+  PlusZeroR   :: a :<=> Either Zero a
+  CommutePlus :: Either a b :<=> Either b a
+  AssocPlusL  :: Either a (Either b c) :<=> Either (Either a b) c 
+  AssocPlusR  :: Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  TimesOneL    :: ((), a) :<=> a
+  TimesOneR    :: a :<=> ((), a)
+  CommuteTimes :: (a,b) :<=> (b,a) 
+  AssocTimesL  :: (a,(b,c)) :<=> ((a,b),c)
+  AssocTimesR  :: ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  TimesZeroL  :: (Zero, a) :<=> Zero
+  TimesZeroR  :: Zero :<=> (Zero, a)
+  Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
+  Factor      :: Either (b, a) (c, a) :<=> (Either b c, a)
+-- Eta and Eps over the monoid (*, 1)
[…]

### 2012-04-03 code/modular/Model2.hs

@@ -1,0 +1,392 @@
+-- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
+{-# LANGUAGE GADTs, TypeOperators, ExistentialQuantification,
+             TypeFamilies #-} -- 7.0.1
+
+--  This code tested with GHC version 7.0.3
+
+module Model2 where
+
+import Model hiding ((@@))
+import Lang (Neg,Unpack(..))
+import Control.Monad
+import Unsafe.Coerce -- needed for polymorphic lookup. 
+
+-- Structure of Pi Values that may contain non-grounded fresh variables.
+type Var a = Int
+data V a where
+    Unit :: V () 
+    Pair :: V a -> V b -> V (a, b) 
+    L :: V a -> V (Either a b) 
+    R :: V b -> V (Either a b) 
+    Neg :: V a -> V (Neg a)
+    Fresh :: Var a -> V a
+
+instance Show (V a) where 
+    show Unit = "()"
+    show (Pair a b) = "(" ++ show a ++ ", " ++ show b ++ ")"
+    show (L a) = "L " ++ (show a) 
+    show (R a) = "R " ++ (show a)
+    show (Neg a) = "Neg " ++ (show a)
+    show (Fresh n) = show n
+
+instance Unpack V where
+    unLeft (L a) = a
+    unRight (R a) = a
+    proj1 (Pair a b) = a
+    proj2 (Pair a b) = b
+    unNeg (Neg a) = a
+
+---------------------------------------------------------------------------
+-- Unification
+
+-- This class says that, the u a's can be unified (as an effect) and generated
+class Unifyable u where
+  type W :: * -> *
+  unify :: u a -> u a -> W ()
+  fresh :: W (u a)
+
+data Binding = forall a. Binding (Var a) (V a)
+type World = ([Binding], Int)
+
+instance Show Binding where 
+    show (Binding n v) = (show n) ++ " -> " 
+
+instance Unifyable V where
+  type W = M
+  unify = unifyM
+  fresh = freshM
+
+extendW :: Var a -> V a -> [Binding] -> [Binding]
+extendW x v w = (Binding x v):w
+
+lookupW :: V a -> [Binding] -> V a
+lookupW (Fresh n) w = find n w 
+    where 
[…]

### 2012-05-23 code/dual/Dual.hs
cleanup dual.hs

@@ -1,6 +1,6 @@
--- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
-{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1, 7.0.3
 
+-- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
 -- Dont know: This code tested with GHC version 6.12.3, and version 7.0.1
 
 module Dual where
@@ -8,21 +8,85 @@ module Dual where
 import Data.Typeable
 import Unsafe.Coerce -- needed for polymorphic lookup. 
 
+
 -----------------------------------------------------------------------
--- Isomorphisms 
+-- Base types 
 
 data Zero 
 data Show a => Neg a = Ng a
              deriving Show
 data Show a => Inv a = In a
              deriving Show
+type (:+:) a b = Either a b
+type (:*:) a b = (a, b)
+
 data Three = One 
            | Two 
            | Three
              deriving (Eq, Show)
 
-type (:+:) a b = Either a b
-type (:*:) a b = (a, b)
+-----------------------------------------------------------------------
+-- Values 
+-- 
+-- Values of the form 'V a' are values used by the interpreter. 
+
+
+-- Values can contain non-grounded fresh variables
+type Var a = Int
+data V a where
+    Unit :: V () 
+    Pair :: V a -> V b -> V (a, b) 
+    L :: V a -> V (Either a b) 
+    R :: V b -> V (Either a b) 
+    Neg :: V a -> V (Neg a)
+    Inv :: V a -> V (Inv a)
+    Lift :: (Eq a, Show a) => a -> V a
+    Fresh :: Var a -> V a
+
+
+instance Show (V a) where 
+    show Unit = "()"
+    show (Pair a b) = "(" ++ show a ++ ", " ++ show b ++ ")"
+    show (L a) = "L " ++ (show a) 
+    show (R a) = "R " ++ (show a)
+    show (Neg a) = "Neg " ++ (show a)
+    show (Inv a) = "Inv " ++ (show a)
+    show (Lift a) = show a
+    show (Fresh n) = show n
+
+------------------------------------------------------------------------------------
[…]

### 2012-05-23 code/dual/Test.hs
cleanup test.hs

@@ -1,4 +1,4 @@
-{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts -XScopedTypeVariables #-} -- 7.0.1
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts -XScopedTypeVariables #-} -- 7.0.1, 7.0.3
 
 import Dual 
 
@@ -6,47 +6,17 @@ import Dual
 ------------------------------------------------------------------------------------
 -- Constructions
 
--- Some shorthands 
-type (:=>>) a b = (Inv a, b)
-type (:<<=) a b = (a, Inv b)
-type P a = () :<=> a
-type N a = a  :<=> ()
-
-makeFunc :: (a :<=> b) -> P (a :=>> b)
-makeFunc c = EtaTimes:.: (Id :*: c)
-
--- Many apply's to be written here based on what interface we want:
--- apply : P (a :=>> b) -> P a -> P b
--- apply : P (a :=>> b) -> (c :<=> a) -> (c :<=> b)
--- apply : ((a :=>> b), a) :<=> b
-
-applyFunc :: ((a :=>> b), a) :<=> b 
-applyFunc = CommuteTimes 
-            :.: AssocTimesL 
-            :.: ((CommuteTimes :.: EpsTimes) :*: Id) 
-            :.: UnitE
-
--- These 
-makeDC :: (a :<=> b) -> N (a :<<= b)
-makeDC c = (c :*: Id) :.: CommuteTimes :.:  EpsTimes
-
-
--- A similar apply is possible, but I dont know what it means. 
-
--- Trace
 traceTimes :: (a, b) :<=> (a, c) -> b :<=> c
 traceTimes c = UnitI 
-               :.: (EtaTimes:*: Id) 
+               :.: (EtaTimes :*: Id) 
                :.: AssocTimesR 
                :.: (Id :*: c) 
                :.: AssocTimesL 
                :.: (EpsTimes :*: Id)
                :.: UnitE
 
--- This is the yanking lemma for trace. 
 yankTimes :: a :<=> a
 yankTimes = traceTimes CommuteTimes
-
 
 z_shapeTimes :: a :<=> a
 z_shapeTimes = UnitI
@@ -58,14 +28,133 @@ z_shapeTimes = UnitI
                :.: UnitE
                
 
[…]

### 2012-05-29 iso-int/rc-iso-int.tex
fixes

@@
 
-% The constructions in this paper apply only to a class of interpreters
-% that can be represented by logically reversible small-step operational
-% semantics. While, in a formal sense, they are related to Abramsky's
-% ``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
-% to give them a more direct characterization purely in terms of logical
-% reversibility of the small-step semantics.
-
 %% - write program; derive inverse
@@
 
+\paragraph*{Universality.} 
+As an example, consider the implementation of the Toffoli gate below,
+which establishes that {{langRevT}} is complete for combinational
+circuits. This encoding uses the type {{1+1}} as {{bool}} and values
+{{left ()}} and {{right()}} as {{true}} and {{false}}. Previous
+work\cite{James:2012:IE:2103656.2103667} (see Sec 5) also establishes
+that {{langRevT}} is Turing complete by showing the compilation of a
+Turing complete language -- a first-order language with numbers and
+iteration -- to {{langRevT}}.
+
+\begin{center}
+  \includegraphics{diagrams/toffoli.pdf}
+\end{center}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We illustrate the main concepts and constructions using two simple examples
-that essentially count $n$ steps. The first machine does nothing else; the
-second uses this counting ability to add two numbers.
+We illustrate the main concepts and constructions using two simple
+examples that essentially count $n$ steps. The first machine does
+nothing else; the second uses this counting ability to add two
+numbers.
 
@@
 \vspace{-20px}
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.5}{
+% \begin{multicols}{2}
+\begin{center}
+\scalebox{1.0}{
   \includegraphics{diagrams/nat-nat6.pdf}
@@
 \end{center}
-\columnbreak
-\begin{center}
-\scalebox{0.6}{
+
+\noindent
+we exchange sections A and B while maintaining the connectivity of the wires:
+
+%\columnbreak
+\begin{center}
+\scalebox{1.0}{
   \includegraphics{diagrams/nat-nat7.pdf}
@@
 \end{center}  
-\end{multicols}
+%\end{multicols}
 
@@
 % does not know when to stop during reverse execution.
-
-% \paragraph*{Conditions that must be satisfied for constructing a isomorphic
-% interpreter}:
-% \begin{enumerate}
-% \item Distinguishable start and stop states. There may be more than one
-%   valid start state and more than one valid stop state.
-% \item Each valid machine state must match a unique reduction LHS. Each
-%   valid machine state must machine a unique reduction RHS.
-% \item Every reduction step must be computable (in {{langRevT}}) and
-%   must be logically reversible -- i.e. it must be possible to reduce
-%   from right to left.
-% \end{enumerate}
 
@@
 
-\subsection{Adder} 
-\label{sec:adder}
-
-Although trivial, the previous example captures the fundamental steps of our
-construction. For reference, these steps are:
+%% In the absence of a formal classification for styles of reversible
+%% abstract machines, we will present only an informal characterisation
+%% of our input language.
+
+\paragraph*{Input Machines.}
+
+Our translation apply to machines presented as small-step operational
+semantics, i.e. machines that are step-wise rewrites of tuples of
+algebraic datatypes, wherein each rewrite step must be logically
+reversible. The formal criterion logical reversibility in this setting
+are captured in Abramsky's presentation of \emph{bi-orthogonal
+  automata} \cite{abramsky2005structural} and may be summarized as:
+\begin{enumerate}
+\item
+  Machines must have distinguishable start and stop states. There may
+  be more than one valid start state and more than one valid stop
+  state.
+\item Each valid machine state must match a unique pattern on the
+  left-hand-side and right-hand-side of the `{{|-->}}' rewrite.
+\item Every reduction step must be computable (in {{langRevT}}) and
+  must be logically reversible -- i.e. it must be possible to reduce
+  from right to left.
+\end{enumerate}
+
+%% The constructions in this paper apply only to a class of interpreters
+%% that can be represented by logically reversible small-step operational
+%% semantics. While, in a formal sense, they are related to Abramsky's
+%% ``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
+%% to give them a more direct characterization purely in terms of logical
+%% reversibility of the small-step semantics.
+
+\paragraph*{Steps of the Translation.}
+
+Although trivial, the previous example captures the fundamental steps
+of our construction. For reference, these steps are:
 \begin{enumerate}
@@
 \end{enumerate}
+
+\subsection{Adder} 
+\label{sec:adder}
 
@@
 isomorphic interpreter is to recognize that the two possible kinds of machine
-states simply hide an implicit {{bool=T~|~F}}. We make this explicit:
+states simply hide an implicit {{bool}}. We make this explicit:
 
@@
 %
-% Start state = <T, t, []>
-% Stop state =  <F, t, []>
+% Start state = <true, t, []>
+% Stop state =  <false, t, []>
 
@@
 %! columnStyle = rclr
-% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
-% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
-% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
-% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
+% <true, L n, c> &|-->& <false, L (n+1), c> &~~~~~ rule 1
+% <true, N t1 t2, c> &|-->& <true, t1, Lft c t2> & rule 2 
+% <false, t1, Lft c t2> &|-->& <true, t2, Rgt t1 c> & rule 3
+% <false, t2, Rgt t1 c> &|-->& <false, c, N t1 t2> & rule 4
 \end{multicols}
@@
 
-On the input side, for {{T}} states, we have expanded the tree component and
-for {{F}}~states we have expanded the tree contexts. Similarly we have done
+On the input side, for {{true}} states, we have expanded the tree component and
+for {{false}}~states we have expanded the tree contexts. Similarly we have done
 the opposite on the output side exactly matching up what the abstract machine
@@
 \item When we encounter a leaf we would like to increment its value and move
-  to the corresponding {{F}} machine state. For the sake of simplicity, in
+  to the corresponding {{false}} machine state. For the sake of simplicity, in
   this interpreter we won't be concerned with exposing stuck states: we

### 2012-05-29 iso-int/rc-iso-int.tex
fixes

@@
 are captured in Abramsky's presentation of \emph{bi-orthogonal
-  automata} \cite{abramsky2005structural} and may be summarized as:
+  automata} \cite{abramsky2005structural} (see Sec. 2) and may be
+summarized as:
 \begin{enumerate}

### 2012-05-29 code/dual/Dual.hs
?

@@ -1,97 +1,76 @@
-{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1, 7.0.3
-
--- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
--- Dont know: This code tested with GHC version 6.12.3, and version 7.0.1
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification #-}
+{-# OPTIONS_GHC -XFlexibleContexts -XEmptyDataDecls #-} 
 
 module Dual where
 
-import Data.Typeable
-import Unsafe.Coerce -- needed for polymorphic lookup. 
+import Unsafe.Coerce -- needed for polymorphic lookup (lookupW)
 
-
------------------------------------------------------------------------
+------------------------------------------------------------------------------
 -- Base types 
 
 data Zero 
+
 data Show a => Neg a = Ng a
-             deriving Show
+  deriving Show                       
+
 data Show a => Inv a = In a
-             deriving Show
-type (:+:) a b = Either a b
-type (:*:) a b = (a, b)
+  deriving Show
 
-data Three = One 
-           | Two 
-           | Three
-             deriving (Eq, Show)
-
------------------------------------------------------------------------
+------------------------------------------------------------------------------
 -- Values 
 -- 
 -- Values of the form 'V a' are values used by the interpreter. 
-
-
 -- Values can contain non-grounded fresh variables
-type Var a = Int
-data V a where
-    Unit :: V () 
-    Pair :: V a -> V b -> V (a, b) 
-    L :: V a -> V (Either a b) 
-    R :: V b -> V (Either a b) 
-    Neg :: V a -> V (Neg a)
-    Inv :: V a -> V (Inv a)
-    Lift :: (Eq a, Show a) => a -> V a
-    Fresh :: Var a -> V a
 
+type Var a = Int
+
+data V a where
+  Unit :: V () 
+  Pair :: V a -> V b -> V (a, b) 
+  L :: V a -> V (Either a b) 
+  R :: V b -> V (Either a b) 
+  Neg :: V a -> V (Neg a)
+  Inv :: V a -> V (Inv a)
+  Fresh :: Var a -> V a
 
 instance Show (V a) where 
-    show Unit = "()"
[…]

### 2012-05-30 iso-int/rc-iso-int.tex
fixes
amr done

--- 2012-05-29 iso-int/rc-iso-int.tex
@@
 \usepackage{graphicx}
-
-\newcommand{\todo}[1]{\textbf{TODO: #1}}
 
@@
 \title{Isomorphic Interpreters from Logically Reversible Abstract Machines} 
-\titlerunning{On the construction of Isomorphic Interpreters} 
+\titlerunning{On the Construction of Isomorphic Interpreters} 
 \author{Roshan P. James \and Amr Sabry}
@@
 In this paper, we investigate the practicality of writing recursive programs
-in {{langRevT}}. Specifically, we assume that we are giving some reversible
+in {{langRevT}}. Specifically, we assume that we are given some reversible
 recursive algorithm expressed as a \emph{reversible abstract machine}, and we
@@
 \paragraph*{Graphical Language.}
-Following the tradition established for monoidal
+Following the tradition for computations in monoidal
 categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
-graphical language that conveys the semantics of the language.  The general
+graphical notation that conveys the semantics of {{langRevT}}.  The general
 idea of the graphical notation is that combinators are modeled by ``wiring
@@
 \item The simplest sort of diagram is the {{id : b <-> b}} combinator which
-  is simply represented as a wire labeled by its type {{b}}, as shown on the
-  left. In more complex diagrams, if the type of a wire is obvious from the
-  context, it may be omitted. When tracing a computation, one might imagine a
-  value {{v}} of type {{b}} on the wire, as shown on the right.
+  is simply represented as a wire labeled by its type {{b}}, as shown below
+  on the left. In more complex diagrams, if the type of a wire is obvious
+  from the context, it may be omitted. When tracing a computation, one might
+  imagine a value {{v}} of type {{b}} on the wire, as shown below on the
+  right:
 
@@
   alternatively as in folklore in the literature on monoidal categories as a
-  ``wave.''
+  ``wave:''
 \begin{multicols}{4}
@@
 
-\item Sum types may similarly be represented by one wire or using
-  parallel wires with a {{+}} operator between them. When tracing the
-  execution of two additive wires, a value can reside on only one of the two
-  wires.
+\item Sum types may similarly be represented by one wire or using parallel
+  wires with a {{+}} operator between them. When tracing the execution of two
+  additive wires, a value can reside on only one of the two wires:
 \begin{multicols}{4}
@@
 \item Associativity is implicit in the graphical language. Three parallel
-  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
+  wires may represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}.
 %% \begin{center}
@@
 
-\item Commutativity is represented by crisscrossing wires.  By
-  visually tracking the flow of particles on the wires, one can verify
-  that the expected types for commutativity are satisfied.
+\item Commutativity is represented by crisscrossing wires:
 
@@
   many diagrams they are omitted and dealt with explicitly only when they are
-  of special interest.
+  of special interest:
 \begin{multicols}{4}
@@
   flow of {{v:b}} depending on what value inhabits the type {{b1+b2}},
-  as shown below. Factoring is the corresponding adjoint operation.
+  as shown below. Factoring is the corresponding adjoint operation:
 
@@
 \item The {{trace+}} operation is drawn as a looped circuit, where the
-  traced type {{b1}} is shown as flowing backward. 
+  traced type {{b1}} is shown as flowing backwards:
 %% The semantics and
@@
 \paragraph*{Universality.} 
-As an example, consider the implementation of the Toffoli gate below,
-which establishes that {{langRevT}} is complete for combinational
-circuits. This encoding uses the type {{1+1}} as {{bool}} and values
-{{left ()}} and {{right()}} as {{true}} and {{false}}. Previous
-work\cite{James:2012:IE:2103656.2103667} (see Sec 5) also establishes
-that {{langRevT}} is Turing complete by showing the compilation of a
-Turing complete language -- a first-order language with numbers and
-iteration -- to {{langRevT}}.
-
+As an example, consider the implementation of the Toffoli gate below, which
+establishes that {{langRevT}} is complete for combinational circuits. This
+encoding uses the type {{1+1}} as {{bool}} and values {{left ()}} and
+{{right()}} as {{true}} and {{false}}:
 \begin{center}
@@
 \end{center}
+\noindent Previous work~\cite[Sec.~5]{James:2012:IE:2103656.2103667}
+establishes that {{langRevT}} is Turing complete by showing the compilation
+of a Turing complete language --- a first-order language with numbers and
+iteration --- to {{langRevT}}.
 
@@
 backwards from the final configuration to reach the initial configuration in
-also 3 steps.
-
-Our goal is to implement this abstract machine in {{langRevT}}: We start by
+also~3 steps.
+
+Our goal is to implement this abstract machine in {{langRevT}}. We start by
 writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
@@
 follows: If the number {{n}} is zero, the output is the top branch which has
-type {{1}}. If the number was non-zero, the output is on the bottom branch
+type~{{1}}. If the number is non-zero, the output is on the bottom branch
 and has value {{n-1}}. The triangle on the right is the dual {{fold}}
@@
 involves introducing a {{trace}} operation for iterating the small step
-realized so far. We do this by first ``sliding'' the two sections of the
-diagram as shown below:
+realized so far:
 
@@
 \noindent
-we exchange sections A and B while maintaining the connectivity of the wires:
+Sliding sections A and B past each other while maintaining the connectivity
+of the wires, exposes that the middle connections were really recursive calls
+to the machine transition:
 
@@
 
-This rearrangement exposes that the middle connections were really recursive
-calls to the machine transition. The final figure is given below:
+%% The final figure is given below:
 
@@
 
-\begin{center}
-\scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
-\end{center}
+%% \begin{center}
+%% \scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
+%% \end{center}
 
@@
 
-\paragraph*{Input Machines.}
-
-Our translation apply to machines presented as small-step operational
-semantics, i.e. machines that are step-wise rewrites of tuples of
-algebraic datatypes, wherein each rewrite step must be logically
-reversible. The formal criterion logical reversibility in this setting
-are captured in Abramsky's presentation of \emph{bi-orthogonal
-  automata} \cite{abramsky2005structural} and may be summarized as:
-\begin{enumerate}
-\item
-  Machines must have distinguishable start and stop states. There may
-  be more than one valid start state and more than one valid stop
-  state.
-\item Each valid machine state must match a unique pattern on the
-  left-hand-side and right-hand-side of the `{{|-->}}' rewrite.
-\item Every reduction step must be computable (in {{langRevT}}) and
-  must be logically reversible -- i.e. it must be possible to reduce
-  from right to left.
-\end{enumerate}
-
 %% The constructions in this paper apply only to a class of interpreters
@@
 
-\paragraph*{Steps of the Translation.}
-
-Although trivial, the previous example captures the fundamental steps
-of our construction. For reference, these steps are:
+\paragraph*{Steps of the Construction.}
+
+Although trivial, the previous example captures the fundamental aspects of
+our construction. In general, our starting point is an abstract machine in
+which every rewrite step is logically reversible. The formal criterion of
+logical reversibility in this setting is captured by Abramsky's
+\emph{bi-orthogonal automata}~\cite[Sec.~2]{abramsky2005structural} which is
+summarized as:
 \begin{enumerate}
-\item Expand the input to expose enough structure to distinguish the left
-  hand side of each machine state; 
-\item Expand the output to expose enough structure to distinguish the right
-  hand side of each machine state;
+\item
+  Machines must have distinguishable start and stop states. There may
+  be more than one valid start state and more than one valid stop
+  state, however.
+\item Each valid machine state must match a unique pattern on the
+  left-hand side and right-hand side of the `{{|-->}}' relation.
+\item Every reduction step must be computable (in {{langRevT}}) and
+  must be logically reversible --- i.e. it must be possible to reduce
+  from right to left.
+\end{enumerate}
+Given such a machine the process of encoding the machine in {{langRevT}}
+consists of the following steps:
+\begin{enumerate}
+\item Expand the input to expose enough structure to distinguish the
+  left-hand side of each machine state;
+\item Expand the output to expose enough structure to distinguish the
+  right-hand side of each machine state;
 \item Shuffle matching input terms to output terms, inserting any appropriate
@@
 \item Slide the two sections to expose the start and stop state and
-  introduce a {{trace}} to iterate the construction. This is always a
-  mechanical process completely independent of the logic of the machine.
+  introduce a {{trace}} to iterate the construction. 
 \end{enumerate}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Adder} 
@@
 dealt with previously. In particular, we have a two types of machine
-states -- {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
+states {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
 to walking down a tree, building up the context in the process. The
@@
 marked by using {{[^ ^]}} and the components that are being generated (or consumed) will be
-marked by {{[v v]}}. For example, given a value of type {{tree*nat}}, the
+marked by~{{[v v]}}. For example, given a value of type {{tree*nat}}, the
 diagram below shows how to first expand the {{tree}} component and then in
@@
 %! columnStyle = rclr
-% <true, L n, c> &|-->& <false, L (n+1), c> &~~~~~ rule 1
-% <true, N t1 t2, c> &|-->& <true, t1, Lft c t2> & rule 2 
-% <false, t1, Lft c t2> &|-->& <true, t2, Rgt t1 c> & rule 3
-% <false, t2, Rgt t1 c> &|-->& <false, c, N t1 t2> & rule 4
+% <true, L n, c> &|-->& <false, L (n+1), c> 
+% <true, N t1 t2, c> &|-->& <true, t1, Lft c t2> 
+% <false, t1, Lft c t2> &|-->& <true, t2, Rgt t1 c> 
+% <false, t2, Rgt t1 c> &|-->& <false, c, N t1 t2> 
 \end{multicols}
@@
 
-And we start examining the machine components as before:
+We start examining the machine components as before:
 
@@
 
-On the input side, for {{true}} states, we have expanded the tree component and
-for {{false}}~states we have expanded the tree contexts. Similarly we have done
-the opposite on the output side exactly matching up what the abstract machine
+On the input side, for {{true}} states, we have expanded the tree component
+and for {{false}}~states we have expanded the tree contexts. We have done the
+opposite on the output side exactly matching up what the abstract machine
 does. One thing to note is that we dropped the {{1}} introduced by expanding
@@
   should be read as ``transform {{v}} according to the primitive isomorphism
-  {{iso}}.'' Implementing this is straightforward and is hence skipped.
+  {{iso}}.''
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph*{Acknowledgments.} 
+This material is based upon work supported by the National Science Foundation
+under Grant No. 1116725.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2012-05-31 code/dual/ND.hs
code

@@ -1,7 +1,7 @@
 {-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification #-}
 {-# OPTIONS_GHC -XFlexibleContexts -XEmptyDataDecls #-} 
 
-module Dual where
+module ND where
 
 import Control.Monad
 
@@ -11,7 +11,9 @@ import Control.Monad
 -- 0, -t, 1/t are modeled using the following types
 
 data Zero -- empty type
+
 data Inv a = Inv a deriving Eq -- fractional types (a should not be zero) 
+
 data Neg a = Neg a deriving Eq -- negative types
 
 instance Show Zero where 
@@ -19,6 +21,8 @@ instance Show Zero where
 
 instance Eq Zero where 
   _ == _ = error "Cannot use == at type 0"
+
+-- 
 
 class Eq a => V a where
   elems :: [a]
@@ -103,24 +107,38 @@ adjoint EtaTimes = EpsTimes
 adjoint EpsTimes = EtaTimes
 
 ------------------------------------------------------------------------------
--- Non-determisnism monad
+-- Non-determisnism monad (implemented using []) 
 -- 
--- We have non-determinism because of eta*/epsilon*. In particular eta* maps
--- () non-deterministically to (1/v,v) to where v is a value of the
+-- We have non-determinism because of eta*/epsilon*. In particular eta*
+-- non-deterministically maps () to (1/v,v) to where v is a value of the
 -- approriate type; epsilon* takes a pair (1/v',v) and checks whether
 -- v==v'. If so we return () and otherwise we fail.
 
-eval_iso :: (V a, V b) => (a :<=> b) -> [a] -> [b] 
-eval_iso c xs = xs >>= eval_iso1 c
-
 eval_iso1 :: (V a, V b) => (a :<=> b) -> a -> [b]
+
 eval_iso1 Id a = return a
+
+eval_iso1 (Sym f) b = eval_iso1 (adjoint f) b
+
+eval_iso1 (f :.: g) a = do b <- eval_iso1 f a 
+                           eval_iso1 g b
+
+eval_iso1 (f :*: g) (a,c) = do b <- eval_iso1 f a 
+                               d <- eval_iso1 g c
+                               return (b,d) 
+
[…]

### 2012-06-01 iso-int/rc-iso-int.tex
small edits

@@
   numbers and addition, move on to tree traversals, and finish with a
-  meta-circular interpreter for our reversible language. Generally, this
+  meta-circular interpreter for our reversible language. This
   gives us a means of developing large reversible programs with the ease of
@@
 
-In this paper, we investigate the practicality of writing recursive programs
+In this paper, we investigate the practicality of writing large recursive programs
 in {{langRevT}}. Specifically, we assume that we are given some reversible
@@
 
-\item Associativity is implicit in the graphical language. Three parallel
+\item Associativity is implicit in the graphical language. Thus, three parallel
   wires may represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}.
@@
 \begin{center}
-\scalebox{1.0}{
+\scalebox{0.8}{
   \includegraphics{diagrams/nat-nat6.pdf}
@@
 \begin{center}
-\scalebox{1.0}{
+\scalebox{0.8}{
   \includegraphics{diagrams/nat-nat7.pdf}
@@
 
-%% The final figure is given below:
+\noindent
+The completed interpreter is given below:
 
@@
 
-%% \begin{center}
-%% \scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
-%% \end{center}
+\begin{center}
+\scalebox{1.0}{ \includegraphics{diagrams/nat-nat8.pdf} }
+\end{center}
 
@@
   left-hand side and right-hand side of the `{{|-->}}' relation.
-\item Every reduction step must be computable (in {{langRevT}}) and
-  must be logically reversible --- i.e. it must be possible to reduce
-  from right to left.
+\item Every reduction step must be computable (in {{langRevT}}), must
+  not drop or duplicate pattern variables and must be logically
+  reversible --- i.e. it must be possible to reduce from right to
+  left.
 \end{enumerate}
@@
 every leaf value.  The machine here is a little richer than the ones
-dealt with previously. In particular, we have a two types of machine
+dealt with previously. In particular, we have two types of machine
 states {{<t,c>}} and {{ {[t, c]} }}. The first of these corresponds
@@
 %subcode{bnf} include main
-%! columnStyle = rrcll
-% unfold_t :& ~~~~t &<->& n + t * t &: fold_t
-% unfold_c :& c &<->& 1 + c * t + t * c ~~~~~~&: fold_c
+%! columnStyle = rcl
+% unfold_t :& t <-> n + t * t &: fold_t
+% unfold_c :& ~~~~~ c <-> 1 + c * t + t * c ~~~~ &: fold_c
 \end{small}

### 2012-06-01 code/dual/ND.hs
code

@@ -1,5 +1,8 @@
-{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification #-}
-{-# OPTIONS_GHC -XFlexibleContexts -XEmptyDataDecls #-} 
+{-# OPTIONS_GHC -XTypeOperators  #-}
+{-# OPTIONS_GHC -XGADTs          #-}
+{-# OPTIONS_GHC -XEmptyDataDecls #-} 
+
+-- eta+/epsilon+ undefined in this model
 
 module ND where
 
@@ -22,7 +25,7 @@ instance Show Zero where
 instance Eq Zero where 
   _ == _ = error "Cannot use == at type 0"
 
--- 
+-- Returns a list of all the elements of the type
 
 class Eq a => V a where
   elems :: [a]
@@ -51,18 +54,19 @@ instance V a => V (Inv a) where
 data a :<=> b where 
 -- Congruence
   Id    :: V a => a :<=> a
-  Sym   :: (V a, V b) => (a :<=> b) -> (b :<=> a)
   (:.:) :: (V a, V b, V c) => (a :<=> b) -> (b :<=> c) -> (a :<=> c)
-  (:*:) :: (V a, V b, V c, V d) => (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
-  (:+:) :: (V a, V b, V c, V d) => (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
-  -- (a :<=> b) -> (Neg a :<=> Neg b) ??
-  -- (a :<=> b) -> (Inv a :<=> Inv b) ??
+  (:*:) :: (V a, V b, V c, V d) => 
+             (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (V a, V b, V c, V d) => 
+             (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
 -- (+) is associative, commutative, and has a unit
   ZeroE       :: V a => Either Zero a :<=> a
   ZeroI       :: V a => a :<=> Either Zero a
   CommutePlus :: (V a, V b) => Either a b :<=> Either b a
-  AssocPlusL  :: (V a, V b, V c) => Either a (Either b c) :<=> Either (Either a b) c 
-  AssocPlusR  :: (V a, V b, V c) => Either (Either a b) c :<=> Either a (Either b c) 
[…]

### 2012-06-01 rational_ext.tex
_

@@
 \usepackage{multicol}
-\newcommand\hmmax{0} % default 3
-\newcommand\bmmax{0} % default 4
+\newcommand\hmmax{0} % default \newcommand\bmmax{0} % default 4
 \usepackage{bm}
@@
 
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-pi} 
+{{langRev}} computation always terminate.
+
+{{forall. c:b1<->b2, v:b1, exists v':b2.}}  
+{{<c,v,[]> |-->  [c,v',[]]}}
+\end{proposition}
+
 \begin{proposition}[Logical Reversibility]
@@
 not know about.
+
+\roshan{We need to prove this. I am not sure how to start.}
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-piee-plus} 
+Computations in {{langRev}} extended with {{eta+}} and {{eps+}} are
+always terminating.
+\end{proposition}
+
+\roshan{This is straightforward.}
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-piee-times} 
+Computations in {{langRev}} extended with {{eta*}} and {{eps*}} are
+always terminating.
+\end{proposition}
+
+\roshan{We can present our construction here.}
+
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-piee} 
+Computations in full {{langRevEE}} are weak normalizing.
+\end{proposition}
+
+\subsection{Logical Reversibility}
+
+\begin{proposition}[Logical Reversibility]
+{{langRev}} with {{eta+}} and {{eps+}} is logically reversible.
+\end{proposition}
+\roshan{Proof follows from compilation to {{langRevT}} }
+
+\begin{proposition}[Logical Reversibility]
+{{langRev}} with {{eta*}} and {{eps*}} is logically reversible.
+\end{proposition}
+
+This is an awkward one. I dont thnk that the above statement is
+right.. At least not in the usualy sense. We can construct the
+following combinators.
+
+{{c1 : bool <-> bool}}
+
+{{c1 true |--> {[true, false ]} }}
+
+{{c1 false |--> {[true, false ]} }}
+
+{{c1{dagger} true |--> {[true, false ]} }}
+
+{{c1{dagger} false |--> {[true, false ]} }}
+
+\noindent
+And
+
+{{c2 : bool <-> bool}}
+
+{{c2 true |--> {[true, false ]} }}
+
+{{c2 false |--> undefined }}
+
+{{c2{dagger} true |--> true }}
+
+{{c2{dagger} false |--> true }}
+
+I don't think this issue will go away by any sort of choice of
+top-level. I tried to design a top-level that can fix this -- I dont
+think it is possible.
 

### 2012-06-18 rational_ext.tex
_

--- 2012-03-13 rational_ext.tex
@@
 
-\title{(With Extensions)The Two Dualities of Computation: Negative and Fractional Types}
-\authorinfo{Roshan P. James}
-           {Indiana University}
-           {rpjames@indiana.edu}
-\authorinfo{Amr Sabry}
-           {Indiana University}
-           {sabry@indiana.edu}
-\maketitle
+\title{Computing in the Field of Rationals}
+%% \authorinfo{Roshan P. James}
+%%            {Indiana University}
+%%            {rpjames@indiana.edu}
+%% \authorinfo{Amr Sabry}
+%%            {Indiana University}
+%%            {sabry@indiana.edu}
+%% \maketitle
 
 \begin{abstract}
-  Every functional programmer knows about sum and product types, {{a+b}} and
-  {{a*b}} respectively. Negative and fractional types, {{a-b}} and {{a/b}}
-  respectively, are much less known and their computational interpretation is
-  unfamiliar and often complicated. We show that in a programming model in
-  which information is preserved (such as the model introduced in our recent
-  paper on \emph{Information Effects}), these types have particularly natural
-  computational interpretations. Intuitively, values of negative types are
-  values that flow ``backwards'' to satisfy demands and values of fractional
-  types are values that impose constraints on their context.  The combination
-  of these negative and fractional types enables greater flexibility in
-  programming by breaking global invariants into local ones that can be
-  autonomously satisfied by a subcomputation. Theoretically, these types give
-  rise to \emph{two} function spaces and to \emph{two} notions of
-  continuations, suggesting that the previously observed duality of
-  computation conflated two orthogonal notions: an additive duality that
-  corresponds to backtracking and a multiplicative duality that corresponds
-  to constraint propagation.
+
+  Previous work on \emph{information effects} established the notion
+  of information preserving computation
+  \cite{infeffects,rc2011,rc2012} and the core calculus for computing
+  with the isomorphisms of finite types, {{langRev}}. Given the
+  first-order nature of {{langRev}} however it was unclear how one may
+  express first-class functions.  Andrej Filinski's masters thesis
+  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
+  symmetry underlying computation in the \lcal. Filinski showed that
+  values are dual to continuations and that functions are dual to
+  delimited continuations.
+
+  We present a computational model whose types are \emph{the field of
+    rational numbers}. This computational model is derived
+  systemmatically from the {{langRev}} by considering a symmetric
+  notion of duality. Unlike the \lcal which shows a classical demorgan
+  dulaity, here we have two axis of dualization and hence two
+  dualities -- an additive duality, namely negative types and a
+  multiplicative duality, namely fractional types.
+
+  %% Every functional programmer knows about sum and product types, {{a+b}} and
+  %% {{a*b}} respectively. Negative and fractional types, {{a-b}} and {{a/b}}
+  %% respectively, are much less known and their computational interpretation is
+  %% unfamiliar and often complicated. We show that in a programming model in
+  %% which information is preserved (such as the model introduced in our recent
+  %% paper on \emph{Information Effects}), these types have particularly natural
+  %% computational interpretations. Intuitively, values of negative types are
+  %% values that flow ``backwards'' to satisfy demands and values of fractional
+  %% types are values that impose constraints on their context.  The combination
+  %% of these negative and fractional types enables greater flexibility in
+  %% programming by breaking global invariants into local ones that can be
+  %% autonomously satisfied by a subcomputation. Theoretically, these types give
+  %% rise to \emph{two} function spaces and to \emph{two} notions of
+  %% continuations, suggesting that the previously observed duality of
+  %% computation conflated two orthogonal notions: an additive duality that
+  %% corresponds to backtracking and a multiplicative duality that corresponds
+  %% to constraint propagation.
 \end{abstract}
@@
 \section{Introduction}
+
+Andrej Filinksi's masters thesis
+\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
+underlying computation is one embraces the symmetry hidden in
+continuations. Filinski showed that values are dual to continuations,
+functions are dual to delimited continuations. Filinksi also shows a
+de-morgan style duality, by treating value pairs as the dual of
+continuation sums and value sums as dual to continuation pairs.
 
@@
 % \end{proposition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{First class functions from the permutation space.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Traces and the Int-Construction.}
 

### 2012-06-18 rational_ext.tex

@@
 not know about.
+
+\roshan{We need to prove this. I am not sure how to start.}
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-piee-plus} 
+Computations in {{langRev}} extended with {{eta+}} and {{eps+}} are
+always terminating.
+\end{proposition}
+
+\roshan{This is straightforward.}
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-piee-times} 
+Computations in {{langRev}} extended with {{eta*}} and {{eps*}} are
+always terminating.
+\end{proposition}
+
+\begin{proposition}[Weak Normalizing]
+\label{prop:termination-piee} 
+Computations in full {{langRevEE}} are weak normalizing.
+\end{proposition}
+
+\begin{center}
+  \includegraphics{diagrams/inf1.pdf}
+\end{center}
+
+\subsection{Logical Reversibility}
+
+\begin{proposition}[Logical Reversibility]
+{{langRev}} with {{eta+}} and {{eps+}} is logically reversible.
+\end{proposition}
+\roshan{Proof follows from compilation to {{langRevT}} }
+
+\begin{proposition}[Logical Reversibility]
+{{langRev}} with {{eta*}} and {{eps*}} is logically reversible.
+\end{proposition}
+
+This is an awkward one. I dont thnk that the above statement is
+right.. At least not in the usualy sense. We can construct the
+following combinators.
+
+{{c1 : bool <-> bool}}
+
+{{c1 true |--> {[true, false ]} }}
+
+{{c1 false |--> {[true, false ]} }}
+
+{{c1{dagger} true |--> {[true, false ]} }}
+
+{{c1{dagger} false |--> {[true, false ]} }}
+
+\noindent
+And
+
+{{c2 : bool <-> bool}}
+
+{{c2 true |--> {[true, false ]} }}
+
+{{c2 false |--> undefined }}
+
+{{c2{dagger} true |--> true }}
+
+{{c2{dagger} false |--> true }}
+
+Probably the right way to approach this is to see what they do in
+physics. If fractionals do correspond to measurement and the
+categorical structures of Coecke and Abramsky, then the ideas of
+reversibility in that setting should also apply here.
+
+I don't think this issue will go away by any sort of choice of
+top-level. I tried to design a top-level that can fix this -- I dont
+think it is possible.
 

### 2012-06-19 rational_ext.tex
_

--- 2012-06-18 rational_ext.tex
@@
 \newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}
+
+\newcommand{\asterix}[0]{*}
 
@@
 %! [] = \Box
-%! |-->* = \mapsto^{*}
+%! |-->* = \mapsto^{\asterix}
 %! |-->> = \mapsto_{\ggg}
@@
 %! empty = \epsilon
-%! langRev = \Pi
-%! langRevT = \Pi^{o}
-%! langRevEE = \Pi^{\eta\epsilon}
+%! Pi = \Pi
+%! Pi0 = \Pi^{o}
+%! PiEE* = \Pi^{\eta\epsilon}_{*}
+%! PiEE+ = \Pi^{\eta\epsilon}_{+}
+%! PiEE = \Pi^{\eta\epsilon}
 %! theseus = Theseus
@@
 \title{Computing in the Field of Rationals}
-%% \authorinfo{Roshan P. James}
-%%            {Indiana University}
-%%            {rpjames@indiana.edu}
-%% \authorinfo{Amr Sabry}
-%%            {Indiana University}
-%%            {sabry@indiana.edu}
-%% \maketitle
+
+\authorinfo{Roshan P. James}
+           {Indiana University}
+           {rpjames@indiana.edu}
+\authorinfo{Amr Sabry}
+           {Indiana University}
+           {sabry@indiana.edu}
+
+\maketitle
 
 \begin{abstract}
+
+  Andrzej Filinski's masters thesis
+  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
+  symmetry underlying computation in the \lcal. Filinski showed that
+  values are dual to continuations and that functions are dual to
+  delimited continuations, thereby suggesting that
+  values-and-functions and continuations-and-delimited-continuations
+  are mirror images of the same phenomena.
 
@@
   \cite{infeffects,rc2011,rc2012} and the core calculus for computing
-  with the isomorphisms of finite types, {{langRev}}. Given the
-  first-order nature of {{langRev}} however it was unclear how one may
-  express first-class functions.  Andrej Filinski's masters thesis
-  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
-  symmetry underlying computation in the \lcal. Filinski showed that
-  values are dual to continuations and that functions are dual to
-  delimited continuations.
+  with the isomorphisms of finite types, {{Pi}}. Given the
+  first-order nature of {{Pi}} however it was unclear how one may
+  express first-class functions and duality in the sense of Filinski.
 
@@
     rational numbers}. This computational model is derived
-  systemmatically from the {{langRev}} by considering a symmetric
+  systemmatically from the {{Pi}} by considering a symmetric
   notion of duality. Unlike the \lcal which shows a classical demorgan
-  dulaity, here we have two axis of dualization and hence two
+  duality, here we have two axis of dualization and hence two
   dualities -- an additive duality, namely negative types and a
@@
 
-Andrej Filinksi's masters thesis
+Andrzej Filinksi's masters thesis
 \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
-underlying computation is one embraces the symmetry hidden in
-continuations. Filinski showed that values are dual to continuations,
-functions are dual to delimited continuations. Filinksi also shows a
-de-morgan style duality, by treating value pairs as the dual of
-continuation sums and value sums as dual to continuation pairs.
+underlying computation in the \lcal. Filinski showed that values are
+dual to continuations, functions are dual to delimited
+continuations. Filinksi also shows a De Morgan style duality, by
+treating value pairs as the dual of continuation sums and value sums
+as dual to continuation pairs.
+
+\begin{comment}
 
@@
 \begin{itemize}
-\item We extend {{langRev}} our reversible programming language of type
+\item We extend {{Pi}} our reversible programming language of type
   isomorphisms~\cite{rc2011,infeffects} (reviewed in Sec.~\ref{sec:pi}) with a notion
@@
   values. (Sec.~\ref{sec:neg})
-\item We independently extend {{langRev}} with a notion of fractional types,
+\item We independently extend {{Pi}} with a notion of fractional types,
   that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
@@
 \item We combine the above two extensions into a language, which we
-  call {{langRevEE}}, whose type system allows any rational number to
+  call {{PiEE}}, whose type system allows any rational number to
   be used as a type. Moreover the types satisfy the same familiar and
@@
 
+\end{comment}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Core Reversible Language: {{langRev}} }
+\section{The Core Reversible Language: {{Pi}} }
 \label{sec:pi}
 
-We review our reversible language {{langRev}}: the presentation in this
+We review our reversible language {{Pi}}: the presentation in this
 section differs from the one in our previous paper~\cite{infeffects} in two
@@
 
-The combinators of {{langRev}} are witnesses to the following type
+The combinators of {{Pi}} are witnesses to the following type
 isomorphisms: 
@@
 
-To summarize, the syntax of {{langRev}} is given as follows.
-
-\begin{definition}{(Syntax of {{langRev}})}
-\label{def:langRev}
+To summarize, the syntax of {{Pi}} is given as follows.
+
+\begin{definition}{(Syntax of {{Pi}})}
+\label{def:Pi}
 We collect our types, values, and combinators, to get the full language
@@
 
-%% In fact, when talking about {{langRev}} circuits, we will often use the
+%% In fact, when talking about {{Pi}} circuits, we will often use the
 %% words {{value}} and {{particle}} interchangeably.
@@
 % factor & (right (v2, v3)) &|-->& (right v2, v3)
+
+%subcode{proof} include main
+%@ ~
+%@@ id v |--> v 
+%
+%@ c{dagger} v1 |--> v2
+%@@ (sym c) v1 |--> v2
+%
+%@ c1  v1 |--> v
+%@ c2  v |--> v2
+%@@ (c1(;)c2)  v1 |--> v2
+%---
+%@ c1  v1 |--> v2
+%@@ (c1 (+) c2)  (left v1) |--> left v2
+%
+%@ c2  v1 |--> v2
+%@@ (c1 (+) c2)  (right v1) |--> right v2
+%---
+%@ c1  v1 |--> v3
+%@ c2  v2 |--> v4
+%@@ (c1 (*) c2)  (v1, v2) |--> (v3, v4)
 \end{scriptsize}
+
+\begin{proposition}[Type Safety]
+
+\end{proposition}
+
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-pi} 
+{{Pi}} computation always terminate.
+
+{{forall. c:b1<->b2, v:b1, exists v':b2.}}  
+{{c v |--> v'}}
+\end{proposition}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff 
+{{ c{dagger} v' |--> v}}
+\end{proposition}
+
+\subsection{Small Step Semantics}
+
 The reductions for the primitive isomorphisms above are exactly the same as
@@
 
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrev}
-{{<c,v,[]> |--> [c,v',[]]}} iff 
-{{<c{dagger},v',[]> |--> [c{dagger},v,[]]}} 
+\begin{proposition}[Correspondence]
+  Evalution in the small-step evaulator corresponds to evaluation in
+  the natural semantics.
+
+  {{c v |--> v'}} iff {{<c, v, []> |-->* [c, v', [] ]}}
+
+\end{proposition}
+
+Consequently the small step semantics is logically reversible and
+strong normalizing.
+
+\begin{proposition}[Type Safety]
+
 \end{proposition}
@@
 % \label{prop:groupoid}
-% {{langRev}} is a groupoid. 
+% {{Pi}} is a groupoid. 
 % \end{proposition}
@@
 % \label{prop:category}
-% {{langRev}} is a dagger symmetric monoidal category. 
+% {{Pi}} is a dagger symmetric monoidal category. 
 % \end{proposition}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{First class functions from the permutation space.}
+% \section{First class functions from the permutation space.}
 
@@
 
+For a traced symmetric monoidal catgeory, the Int-construction of
+Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
+constructing a compact closed category. This is the obvious first
+thing to try for a duality in this setting.
+
+%subcode{proof} include main
+%@ c : b1 + b2 <-> b1 + b3
+%@@ trace c : b2 <-> b3
+
+%% \begin{center}
+%%   \includegraphics{diagrams/thesis/trace_plus.pdf}
+%% \end{center}
+
+Show that the int construction does not preserve the {{*}} tensor. We
+need a duality that preserves both tensors.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Negative and Fractional Types}
-\label{sec:neg}
-\label{sec:frac}
-
-This section introduces the syntax and graphical languages for
-negatives and fractionals.  The general outline is as follows:
-\footnote{We refer the reader to Selinger's excellent survey
-  article of monoidal
-  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
-  definitions.}
-\begin{itemize}
-\item As established in our earlier work~\cite{rc2011,infeffects}, the
-  underlying categorical semantics of our core reversible language
-  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
-  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
-  tensor.
-\item We extend each underlying symmetric monoidal structure to a
-  \emph{compact closed} structure by adding a dual for each object and two
-  special morphisms traditionally called {{eta}} and {{eps}}. 
-\end{itemize}
-The extended language is referred to as {{langRevEE}}.  After
-presenting the extension at the syntactic level, we discuss the
-categorical semantics informally via the graphical language. The
-operational semantics is presented in Sec.~\ref{sec:rat}.
-
-As will be detailed in the remainder of this section, the compact closed
-structure provides several properties of interest: (i) morphisms or wires are
-allowed to run from right to left, (ii) the structure admits a trace that,
-depending on the situation, can be used to implement various notions of loops
-and
-recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
-(iii) the structure includes an isomorphism showing that the dual operator is
-an involution, and (iv) the structure is equipped with objects representing
-higher-order functions and a morphism \textit{eval} that applies these
-functional objects. Interestingly each monoidal structure provides the same
-ingredients, resulting in two dualities, two traces, two involutions, and two
-notions of higher-order functions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Syntax and Types}
-
-Describing the syntax and types of an extension to {{langRev}} with additive
-and multiplicative duality is fairly straightforward. Basically, for the
-additive case, we extend the language with negative types denoted {{-b}},
-negative values denoted {{-v}}, and two isomorphisms {{eta+}} and
-{{eps+}}. Similarly, for the multiplicative case, we extend the language with
-fractional types denoted {{1/b}}, fractional values denoted {{1/v}}, and two
-isomorphisms {{eta*}} and {{eps*}}.
+\section{Fractional Types : {{PiEE*}} }
 
 %subcode{bnf} include main
-% Value Types, b = ... | -b | 1/b
-% Values, v = ... | -v | 1/v
+% Value Types, b = 0 | 1 | b + b | b * b  | 1/b
+% Values, v = () | left v | right v | (v, v) | 1/v
 %
-% Isomorphisms, iso &=& ... | eta+ | eps+ | eta* | eps*
-
-For convenience, we sometimes use the notations {{b1 - b2}} and
-{{b1/b2}} to indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}}
-respectively.  The types of the new constructs are:
-\vspace{-15pt}
+% Isomorphisms, iso &=& ... | eta* | eps*
+
+\noindent
+No division by zero.
+
 \begin{multicols}{2}  
 %subcode{opsem} include main
-% eta+ &: 0 <-> (-b) + b :& eps+
 % eta* &: 1 <-> (1/b) * b :& eps*
@@
 %subcode{proof} include main
-%@ |- v : b
-%@@ |- -v : -b
-%
 %@ |- v : b
@@
 
-For the graphical language, we visually represent {{eta+}}, {{eps+}},
-{{eta*}}, and {{eps*}} as U-shaped connectors. On the left below are {{eta+}}
-and {{eta*}} showing the maps from {{0}} to {{-b+b}} and {{1}} to {{1/b * b}}. 
-On the right are {{eps+}} and {{eps*}} showing the maps from {{-b+b}} to~0 and 
-{{1/b * b}} to~1. Even though the diagrams below show the {{0}} and {{1}} 
-wires for completeness, later diagrams will always drop them in contexts 
-where they can be implicitly introduced and eliminated. 
+For the graphical language, we visually represent {{eta*}}, and
+{{eps*}} as U-shaped connectors. On the left below is {{eta*}} showing
+the map and {{1}} to {{1/b * b}}.  On the right is {{eps*}} showing
+the map from {{1/b * b}} to~1. Even though the diagrams below show the
+{{1}} wires for completeness, later diagrams will always drop them in
+contexts where they can be implicitly introduced and eliminated.
 \begin{multicols}{2}
 \begin{center}
-  \includegraphics{diagrams/eta.pdf}
-\end{center}
-
-\begin{center}
- \includegraphics{diagrams/eps.pdf}
+  \includegraphics{diagrams/eta_times.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/eps_times.pdf}
 \end{center}
 \end{multicols}
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/eta_times.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/eps_times.pdf}
-\end{center}
-\end{multicols}
-
-The usual interpretation of the type {{b1+b2}} is that we either have an
-appropriately tagged value of type {{b1}} or an appropriately tagged value of
-type {{b2}}. This interpretation is maintained in the presence of {{eta+}}
-and {{eps+}} in the following sense: a value of type {{right v : -b + b}}
-flowing into an {{eps+}} on the lower wire switches to a value 
-{{left (-v): -b + b}} that flows backwards on the upper wire. The inversion of
-direction is captured by the negative sign on the value.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus1.pdf}
-}
-\end{center}
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/eps_plus2.pdf}
-}
-\end{center}  
-\end{multicols}
-As an example, the diagram representing the first isomorphism in the
-introduction is:
-\begin{center}
-\scalebox{1.2}{
-  \includegraphics{diagrams/algebra1.pdf}
-}
-\end{center}
-
-Similarly, in the usual interpretation of {{b1*b2}} we have both a value of
-type {{b1}} and a value of type {{b2}}. This interpretation is maintained in
-the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be
-viewed as a fission point for a value of type {{b}} and its multiplicative
-inverse {{1/b}}. Operationally, this corresponds to the creation of two
-values {{alpha}} and {{1/alpha}} where {{alpha}} is a fresh logic
-variable. The operator {{eps*}} then becomes a unification site for these
-logic variables:
+
+The usual interpretation of {{b1*b2}} we have both a value of type
+{{b1}} and a value of type {{b2}}. This interpretation is maintained
+in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
+to be viewed as a fission point for a value of type {{b}} and its
+multiplicative inverse {{1/b}}. Operationally, this corresponds to the
+creation of two values {{alpha}} and {{1/alpha}} where {{alpha}} is a
+fresh logic variable. The operator {{eps*}} then becomes a unification
+site for these logic variables:
+
 \begin{multicols}{2}
@@
 
-It is useful to contrast the categorical structure of {{langRevEE}},
-which has two distinct \emph{compact closed} structures, to the
-well-known and closed related \emph{*-autonomous} categories that form
-a model for linear logic. The distnguishing feature that seperates
-*-autonomous structures from compact closed structures is that the
-dualizing action in *-autonomous categories does not preserve the
-tensor. In {{langRev}}, as we will see , we can construct the
-isomorphism {{-(b1+b2)<-> (-b1)+(-b2)}}, i.e. the tensor {{+}} has
-been preserved by the negation.  Whereas in linear logic we have that
-$(a\oplus b)^\bot$ is $a^\bot \parr b^\bot$, i.e. the connective
-$\oplus$ is not preserved.
-
-In linear logic terms `additive pairs' and `multiplicative sums' are
-used to indicate this notion that some connectives has both additive
-and multiplicative properties. The usual intuitive interpretations for
-these involve suggestions like one has a dollar to spend and the both
-cigarettes and soda are available for purchase, but the fact that you
-have only a dollar to spend affords you only one of these. The
-intuitions talk about a sort of dependacy underlying the fact that
-both cigarettes and soda are available but you can have only one.
-
-In {{langRevEE}}, operators {{eta}} and {{eps}} capture a similar
-notion.The pair of values {{v}} and {{1/v}} created by {{eta*}} are
-not fully independant in the way one would think of normal pairs. The
-value of one, affects the value of the other. Similarly, the {{v}} and
-{{-v}} created by {{eta+}} do really present a strict choice of
-{{left}} or {{right}} in the sense that usual sums do, but instead can
-have a \emph{both} {{left}} and {{right}} values at different points
-during the computation. However, in our opinion, the philisophical
-difference the additiveness of the additive pairs and the
-multiplicativesness of the multiplicative pair aren't a property of
-the connective between the values, but a property intrinsic to the
-values themselves. Hence {{v}} and {{1/v}} share secretely a
-relationship that their combinting {{*}} constructor does not and need
-not know about.
-
-%%%%%%%%%%%%%%%%%%%%
-\subsection{Categorical Constructions}
-
-All the constructions below are standard: they are collected from Selinger's
-survey paper on monoidal
-categories~\cite{springerlink:10.1007/978-3-642-12821-94} and presented in
-the context of our language.
+Negative Information and Measurements.
+
+Compact Closed Structure (modulo 0).
+
+Show that ``no division by zero'' is supported by the semantics.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Semantics}
+
+Provide a semantics with the nondeterminism monad and define an
+\emph{eval}.
+
+\begin{proposition}[Type Safety]
+
+\end{proposition}
+
+\begin{proposition}[Strong Normalizing]
+
+\end{proposition}
+
+Note that this language is not logically reversible.
 
@@
 essentially the constructions in the introduction.
+
+\begin{proposition}[Coherence]
+
+\end{proposition}
+
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Categorical Constructions}
+
+All the constructions below are standard: they are collected from Selinger's
+survey paper on monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94} and presented in
+the context of our language.
 
@@
 As with compact closed categories, the dual on the objects distributes
-over the tensor. In terms of {{langRevEE}} we have that {{-(b1+b2)}}
+over the tensor. In terms of {{PiEE}} we have that {{-(b1+b2)}}
 can be mapped to {{(-b1)+(-b2)}} and that {{1/(b1*b2)}} can be mapped
@@
 \paragraph*{Duality is a functor.}
-Duality in {{langRevEE}} can map objects to their duals and morphisms
+Duality in {{PiEE}} can map objects to their duals and morphisms
 to act on dual objects. In other {{c : b1 <-> b2}} to
@@
 
-This construction relies on the fact that every {{langRevEE}} morphism
+This construction relies on the fact that every {{PiEE}} morphism
 has an adjoint.  The {{inv}} construction is similar.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Functions and Delimited Continuations}
-\label{sub:hof}
-
-Although these constructions are also standard, they are less known and they
-are particularly important in our context: we devote a little more time to
-explain them. Our discussion is mostly based on Abramsky and Coecke's article
-on categorical quantum mechanics~\cite{abramsky-2008}.
-
-In a compact closed category, each morphism {{f : b1 <-> b2 }} can be given a
-\emph{name} and a \emph{coname}. For the additive fragment, the name
-$\fname{f}$ has type {{0 <-> (-b1 + b2)}} and the coname $\fconame{f}$ has type
-{{b1 + (-b2) <-> 0}}. For the multiplicative fragment, the name $\fname{f}$ has
-type {{1 <-> ((1/b1) * b2))}} and the coname $\fconame{f}$ has type 
-{{(b1 * (1/b2)) <-> 1}}. Intuitively, this means that for each morphism, 
-it is possible to construct, from ``nothing,'' an object in the category that 
-denotes this morphism, and dually it is possible to eliminate this object.
-The construction of the name and coname of {{c : b1 <-> b2}} in the additive case 
-can be visualized as follows:
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{SAT Solver }
+\label{sec:prog}
+\label{sec:sat-solver}
+
+% Additive traces correspond to iteration and multiplicative traces
+% correspond to fixpoint of constraints. Addtive traces have been
+% studied in the past \cite{infeffects} and here we focus on the
+% multiplicative trace.
+
+% The construction presented here is a novel SAT-solver that relies on
+% {{trace*}} for its solution. It varies from previous classical SAT solvers
+% \amr{must cite ???} in that there is no explicit search operation on the
+% boolean space. It also varies from previous quantum SAT solvers \amr{must
+%   cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
+% which is the solution following which an isomorphic clone operation allows us
+% to examine the result.  Though we focus on SAT here, this construction can be
+% generalized to any constraint satisfaction problem whose search space can be
+% represented by a recursive type and for which an isomorphic clone operation
+% can be constructed. 
+% can be constructed. Before we start, let us recall two constructions:
+
+We illustrate the expressive power of first-class constraints represented by
+fractional types. To understand the intuition, recall the definition of
+{{trace}} for the multiplicative fragment of {{PiEE}}. 
+Given {{f : a*c <-> b*c }}, we have {{trace* f : a <-> b}}:
+
+{{ trace* f = uniti (;) (id (*) eta*) (;) (f (*) id) (;) (id (*) eps*) (;) unite }}
+
+\noindent This circuit uses {{eta*}} to generate all possible {{c}}-values
+together with an associated {{(1/c)}}-constraint. It then applies {{f}} to the
+pair {{(a,c)}}. The function {{f}} must produce an output {{(b,c')}} for each
+such input. If the input {{c}} and the output {{c'}} are the same they can be
+annihilated by {{eps*}}; otherwise the execution gets stuck and this
+particular choice of {{c}} is pruned.
+
+% Annihilation and non-termination describe undefined program
+% execution. While non-termination characterizes an iteration that fails
+% to terminate, annihilation characterizes a constraint that has no
+% fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
+% annihilating only the program states that do not match the boolean
+% satisfiability constraint.
+% \emph{Annihilation.}  
+% With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
+% which gives us the ability to find the fixpoint of a constraint.  If
+% the constraint has no satisfying values, there is no fixpoint possible
+% -- we use the term \emph{annihilation} to describe the corresponding
+% undefined state of the program.
+
+A large class of constraint satisfaction problems can be expressed
+using {{trace*}}. We illustrate the main ideas with the implementation
+of a SAT-solver.  We proceed in small steps, reviewing some of the
+necessary constructions presented in our earlier
+paper~\cite{infeffects}.
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Booleans and Conditionals}
+
+Given any combinator {{c : b <-> b}} we can construct a combinator called
+{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
+one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
+the combinator {{c}} is used to transform the value of type~{{b}}. If the
+boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
+write down the combinator for {{if_c}} in terms of {{c}} as 
+{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
+
+\noindent The diagram below shows the input value of type {{(1+1)*b}}
+processed by the distribute operator {{dist}}, which converts it into a value
+of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
+case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
+combinator~{{c}} is applied to the value of type~{{b}}. The right branch
+which corresponds to the boolean being {{false}} passes along the value of
+type {{b}} unchanged.
+
+\begin{center}
+\scalebox{1.0}{
+%%subcode-line{pdfimage}[diagrams/if_c.pdf]
+\includegraphics{diagrams/thesis/cnot.pdf}
+}
+\end{center}
+
+% We will be seeing many more such wiring diagrams in this paper and it is
+% useful to note some conventions about them. Wires indicate a
+% value that can exist in the program. Each wire, whenever possible, is
+% annotated with its type and sometimes additional information to help clarify
+% its role. When multiple wires run in parallel, it means that those values
+% exist in the system at the same time, indicating pair types. When there is a
+% disjunction, we put a {{+}} between the wires. 
+% Combinators for distribution {{dist}} and factoring {{factor}}
+% are represented as triangles with their operator symbols in them. Other
+% triangles may be used and, in each case, types or labels will be used to
+% clarify their roles. Finally, we don't draw boxes for combinators such as
+% {{id}}, commutativity, and associativity, but instead just shuffle the wires
+% as appropriate.
+
+The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
+negates its second argument if the first argument is {{true}}. This
+gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
+equivalent construction that is useful is {{else_{not} }} where we
+negate the second argument only if the first is {{false}}.
+
+Similarly, we can iterate the construction of {{if_c}} to check several
+bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
+checks two booleans and negates the result wire only if they are both
+{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
+universal reversible gate. We can generalize this construction to
+{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
+they are all {{true}}.
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Cloning}
+
+Although cloning is generally not allowed in reversible languages, it is
+possible at the cost of having additional constant inputs. For example,
+consider the gate {{else_{not} }}. Generally, the gate maps
+{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
+on the cases in which the second input is {{true}}, we get that the gate maps
+{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
+i.e., the gate clones the first input. A circuit of {{n}} parallel
+{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
+{{true}} inputs in the process.  Let us call this construction
+{{clone^n_{bool} }}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Construction of the Solver}
+
+The key insight underlying the construction comes from the fact that we can
+build \emph{annihilation circuits} such as the one below:
+
+\begin{center}
+\scalebox{1.2}{
+  \includegraphics{diagrams/not_trace.pdf}
+}
+\end{center}
+The circuit constructs a boolean {{b}} and its dual {{1/b}}, negates one of
+them and attempts to satisfy the constraint that they are equal which
+evidently fails.
+
+With a little work, we can modify this circuit to only annihilate values that
+fail to satisfy the constraints represented by a SAT-instance {{f}}. In more
+detail, an instance of SAT is a function/circuit~{{f}} that given some
+boolean inputs returns {{true}} or {{false}} which we interpret as whether
+the inputs satisfy the constraints imposed by the structure of {{f}}. Because
+we are in a reversible world, our instance of SAT must be expressed as an
+isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
+below. Assuming that {{f}} is expressed as an isomorphism, we have enough
+information to reconstruct the input from the output. This can be done by
+using the adjoint of {{f}}. At this point we have, the top half of the
+construction below:
+
+\begin{center}
+\scalebox{1.2}{
+  \includegraphics{diagrams/sat1.pdf}
+}
+\end{center}
+
+To summarize, the top half of the circuit is the identity function except
+that we have also managed to produce a boolean wire labeled
+\textsf{satisfied?} that tells us if the inputs satisfy the desired
+constraints. We can take this boolean value and use it to decide whether to
+negate the control wire or not. Thus, the circuit achieves the following
+goal: if the inputs do not satisfy {{f}}, the control wire is negated.  We
+can now use {{trace*}} to annihilate all these bad values because the control
+wire acts like the closed-loop {{not}} in the previous construction.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Final Details}
+\label{sub:f}
+
+Any boolean expression {{f : bool^n -> bool}} can be compiled into the
+isomorphism {{iso_f : bool^h*bool^n<->bool^g*bool}} where the extra bits
+{{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
+such an isomorphism has been detailed
+before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
+applying {{iso_f}} to some special heap values and an input \textit{bs}
+produces some bits that can be ignored and the same output that {{f}} would
+have produced on {{bs}}. We can ensure that the heap has the appropriate
+initial values by checking the heap and negating a second control wire, if
+the values do not match, i.e., the dotted part in the diagram below.
+
+\begin{center}
+\scalebox{1.2}{
+  \includegraphics{diagrams/sat2.pdf}
+}
+\end{center}
+
+% \roshan{Explain how the dotted part is implemented.}
+
+Let us call the above construction which maps inputs, heap, and
+control wires to inputs, heap, and control wires as {{sat_f}}. The
+SAT-solver is is completed by tracing the {{sat_f}} and cloning the
+inputs using {{clone^n_{bool} }}.
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/sat3.pdf}
+}
+\end{center}
+
+When the solver is fed inputs initialized to {{true}}, it clones only
+those inputs to {{sat_f}} that satisfy {{f}} and the heap
+constraints. In the case of unique-SAT the solver will produce exactly
+0 or 1 solutions. In the case of general SAT, the solver will produce
+solutions as determined by the semantics of the top-level interaction
+(see discussion in Sec. \ref{sec:rat}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Finite Dimensional Hilbert Spaces}
+
+Halmos \cite{halmos1958finite}, Abramsky and Coecke
+\cite{abramsky2004categorical}. Despite the fact that this fragment
+seem badly behaved (no logical reversibility) we have a close
+connection with finite dimensional hilbert spaces.
+
+The technical goal of this section is to show that the categorical
+structure of finite dimensionl hilbert spaces with unitary maps and
+inner product has the same categorical structure as {{PiEE*}}. Thus
+constructions in {{PiEE*}} can be transported to FdHilb-unitary. If we
+allow all linear maps, then {{+}} becomes a biproduct.  This has
+connections to the Quantum Picturisms of Abramsky and Coecke and hence
+to Qunatum Physics.
+
+%subcode{bnf} include main
+% v = 0 | 1 | v + v | v * v| 1/v
+
+There is much future work to be done in this line. The logical
+consequence of this line of research is either (1) there is no way to
+accomodate proabailities etc in this setting and this is all
+meaningless or (2) constrcutions in {{PiEE*}} correspond to physical
+phenomena.
+
+Also show the what the irreversible constructions can be interpreted
+in Dirac style.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Negative Types : {{PiEE+}} }
+
+What is a negative type anyway?
+
+%subcode{bnf} include main
+% Value Types, b = 0 | 1 | b + b | b * b | -b
+% Values, v = () | left v | right v | (v, v) | -v 
+%
+% Isomorphisms, iso &=& ... | eta+ | eps+
+
+For convenience, we sometimes use the notations {{b1 - b2}} and
+{{b1/b2}} to indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}}
+respectively.  The types of the new constructs are:
+\vspace{-15pt}
 \begin{multicols}{2}  
-\begin{center}
-  \includegraphics{diagrams/function.pdf}
-\end{center}
-
-\begin{center}
-  \includegraphics{diagrams/delimc.pdf}
-\end{center}  
+%subcode{opsem} include main
+% eta+ &: 0 <-> (-b) + b :& eps+
+
+%subcode{proof} include main
+%@ |- v : b
+%@@ |- -v : -b
 \end{multicols}
 
-Intuitively the name consists of viewing {{c}} as a function and the coname
-consists of viewing {{c}} as a delimited continuation.
-
-In addition to being able to represent morphisms, it is possible to express
-function composition. For the additive case, the composition is depicted
-below:
-
 \begin{multicols}{2}
-\scalebox{1.1}{
-    \includegraphics{diagrams/compose1.pdf}
-}
-
-\scalebox{0.8}{
-    \includegraphics{diagrams/compose.pdf}
-}
+\begin{center}
+  \includegraphics{diagrams/eta.pdf}
+\end{center}
+
+\begin{center}
+ \includegraphics{diagrams/eps.pdf}
+\end{center}
 \end{multicols}
-which is essentially equivalent to sequencing both the computation blocks as
-shown below:
-
-\begin{center}
-  \includegraphics{diagrams/compose2.pdf}
-\end{center}
-
-Applying a function to an argument consists of making the argument flow
-backwards to satisfy the demand of the function:
+
 \begin{multicols}{2}
 \begin{center}
-\scalebox{1.0}{
-  \includegraphics{diagrams/apply1.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.8}{
-  \includegraphics{diagrams/apply2.pdf}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus1.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{1.5}{
+  \includegraphics{diagrams/eps_plus2.pdf}
 }
@@
 
-Having reviewed the representation of functions, we now discuss the
-similarities and differences between the two notions of functions and their
-relation to conventional (linear) functions which mix additive and
-multiplicative components. For that purpose, we use a small example.
-Consider a datatype {{color = R|Gr|B}}, and let us consider the following
-manipulations:
-\begin{itemize}
-\item Using the fact that {{1}} is the multiplicative unit, generate
-  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
-\item Apply the isomorphism {{1 <-> (1/b) * b}} in parallel to each of
-  the components of the above tuple. The resulting value is
-  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
-  {{alpha2}} are fresh logic variables;
-\item Using the fact that {{*}} is associative and commutative, we can
-  rearrange the above tuple to produce the value:
-
-{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.
-
-\end{itemize}
-
-At this point we have constructed a strange mix of two {{b-o*b}} functions;
-inputs of one function manifest themselves as outputs of the other. If
-{{(1/alpha1,alpha2)}} is held by one subcomputation and {{(1/alpha2,alpha1)}}
-is held by another subcomputation, these remixed functions form a
-communication channel between the two concurrent subcomputations. Unifying
-{{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
-be {{R}} in the other. The type {{b}} thus takes the role of the type of the
-communication channel, indicating how much information can be communicated
-between the two subcomputations.  Depending on the choice of the type {{b}},
-an arbitrary number of bits may be communicated.
-
-Dually, the additive reading of the above manipulations correspond to
-functions of the form {{b-o+b}}, witnessing isomorphisms of the form
-{{0<->(-b)+b}}). The remixed additive functions express control flow transfer
-between two subcomputations, \emph{only one of which exists} at any point,
-i.e., they capture the essence of coroutines.
-
-It should be evident that in a universe in which information is not
-guaranteed to be preserved by the computational infrastructure, the above
-slicing and dicing of functions would make no sense. But linearity is not
-sufficient: one must also recognize that the additive and multiplicative
-spaces are different.
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Additional Constructions}
-\label{sec:specific-constructions}
-
-The additional constructions below (presented with minimal commentary)
-confirm that conventional algebraic manipulations in the mathematical field
-of rationals do indeed correspond to realizable type isomorphisms in our
-setting. The constructions involving both negative and fractional types are
-novel.
-
-\paragraph*{Lifting negation out of {{*}}.}
-The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
-and {{b2}} are flowing opposite to each other then it doesn't matter which
-direction is forwards and which is backwards. More interestingly as {{b1}} is
-moving backwards, it can ``see the past'' of {{b2}} which is equivalent to
-both particles moving backwards.
-
-{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
-
-To build these isomorphisms, we first build an intermediate
-construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
-\begin{center}
-\scalebox{0.8}{
-  \includegraphics{diagrams/eps_fst.pdf}
-}
-\end{center}
-
-The isomorphism {{(-b1) * b2 <-> -(b1 * b2)}} can be constructed in
-terms of {{eps_{fst} }} as shown below.
-
-\begin{center}
-\scalebox{0.9}{
-  \includegraphics{diagrams/mult_neg.pdf}
-}
-\end{center}
-
-The second isomorphism can be built in the same way by merely swapping
-the arguments.
-
-\paragraph*{Multiplying Negatives.}
-
-{{b1 * b2 <-> (-b1)*(-b2)}}
-
-This isomorphism is a consequence of the fact that $-$ is an involution: it
-corresponds to the algebraic manipulation:
-
-%% %subcode{opsem} include main
-%% %! columnStyle = rc
-%% %   & b1 * b2
-%% % = & -(-(b1*b2)) 
-%% % = & -((-b1)*b2)
-%% % = & (-b1)*(-b2)
-
-{{b1 * b2 = -(-(b1*b2)) = -((-b1)*b2) = (-b1)*(-b2)}}
-
-\paragraph*{Multiplying and Adding Fractions.}
-An isomorphism witnessing:
-
-{{b1/b2 * b3/b4 <-> (b1*b3)/(b2*b4)}}
-
-is straightforward. More surprisingly, it is also possible to construct
-isomorphisms witnessing:
-
-{{b1/b + b2/b <-> (b1+b2)/b}}
-
-{{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
-
-% \paragraph*{Unification.}
-% A computational interpretation of this is that {{langRevEE}} acts like
-% a logic programming system and {{eta*}} acts as the site of creation
-% of a new typed logic variable of type {{b}} and its corresponding dual
-% of type {{1/b}}. These two values are free to move independently
-% through a circuit, but any operations on value affect what operations
-% are possible on the other since they are both related by an underlying
-% logic variable. Dually, {{eps*}} acts as a unification site for values
-% flowing along both branches. If the values are not exactly duals, then
-% the system fails to unify and hence the program is stuck. If the
-% values are exactly duals, then unification succeeds and {{eta*}}
-% returns {{1}}.
-
-% \paragraph*{Entanglement and Anti-Particles.}
-% A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
-% quantum mechanics. The {{eta*}} is a site of fission that creates an
-% information particle and its anti-particle. Both flow in the same
-% direction in time. The produce particles are entangles in that actions
-% on one (such as transformation by application of an isomorphism)
-% affects the other. Further the fission site does not create a particle
-% in any specific state, but in a superposition of all possible states
-% as dictated by its type. In each possible world that the particle
-% exists, it takes on a specific value inhibiting its type and
-% correspondingly its entangled anti-particle takes the dual of the
-% specific value.
-
-% This also leads to the exciting point of view that entangled
-% particle/anti-particle pairs act as functions in a physical
-% first-class sense. We can ask the function to be applied to a value by
-% unifying the anti-particle with a particle in a well-known state and
-% examining is pair. We shall use this point of view in the development
-% our SAT solver (see Sec. \ref{sec:sat-solver}).
-
-% \paragraph*{Entanglement and Anti-Particles.}
-% \paragraph*{Annihilation.}
-% As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
-% additive {{trace+}}, {{trace*}} gives us the ability to find the
-% fixpoint of the constraint (expressed with combinators) that is
-% traced. If the constraint has not satisfying values, there is not
-% fixpoint possible -- we use the term `annihilation' to describe the
-% corresponding undefined state of the program. Annihilation and a
-% non-termination both describe undefined program execution -- while
-% non-termination characterizes an iteration that fails to terminate,
-% annihilation characterizes a constraint that has no fixpoint. 
-%% An analogy inspired by quantum mechanics is also applicable here.  The
-%% operator {{eta*}} is a site of fission that creates \emph{an entangled
-%%   particle and anti-particle}. These particles can be thought to be in
-%% a \emph{superposition} of states determined by their type. Both flow
-%% in the same direction in time.  In each possible world that the
-%% particle exists, it takes on a specific value inhibiting its type and
-%% correspondingly its entangled anti-particle takes the dual of the
-%% specific value. Since the particles are entangled, actions on one
-%% (such as transformation by application of an isomorphism) affects the
-%% other in much the same sense as \emph{action at a distance}. The
-%% `{{a-o*b}}' functions mentioned in the introduction, are first-class
-%% values and they correspond to these entagled pairs. Finally {{eps*}}
-%% operations are sites where corresponding particle and anti-particles
-%% annihilate each other. While appealing as an analogy, this description
-%% does not imply (or preclude) any formal connection with Physics.
-% \paragraph*{Zero Information.}
-% As with {{eta+}}, we argue that {{eta*}} constructs no new
-% information, instead it gives us the ability to explore as many worlds
-% as permitted by the type {{b}}. So situate ourselves in any specific
-% world, we have to supply additional information in the form of
-% unifying the particle or anti-particle with a particle representing
-% known information. Hence only by supplying information to a an
-% entangled pair can we read information from it -- and always exactly
-% as much information as we supplied.
-
-%% We can ask the function to be applied to a value by
-%% unifying the anti-particle with a particle in a well-known state and
-%% examining is pair. We shall use this point of view in the development
-%% our SAT solver (see Sec. \ref{sec:sat-solver}).
-
-%%  This also leads to the exciting point of view that
-%% entangled such particle/anti-particle pairs, which are first class
-%% values, act as functions in a physical first-class sense.
-
-%% Further the fission site does not create a particle in any specific
-%% state, but in a superposition of all possible states as dictated by
-%% its type.
-
-% Since such pairs are first-class entities, they correspond to
-
-%% \emph{Coherence for Compact Closure with Multiplicative Duality.}  As
-%% before, the triangle axiom which is the coherence condition on compact
-%% closed categories is applicable and is expressed below in terms of
-%% {{eta*}}/{{eps*}} and the multiplicative monoid.
-
-%% %% The section follows the outline of the previous one, first presenting
-%% %% the new syntax, and discussing the graphical language and the
-%% %% properites implied by the underlying categorical structure.
-%% The usual interpretation of {{b1*b2}} we have both a value of type
-%% {{b1}} and a value of type {{b2}}. This interpretation is maintained
-%% in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
-%% to be viewed as a fission point for a value of type {{b}} and its
-%% multiplicative inverse {{1/b}}. This naturally raises the question --
-%% which value of type {{b}} is produced? The isomorphism {{eta*}} does
-%% not produce any specific value of type {{b}}, instead it produces a
-%% placeholder for an unspecified value and dually, {{1/b}} is inhabited
-%% by the placeholder that indicates the absence of the unspecified
-%% value.
-
-%% \begin{enumerate}
-%% \item 
-%% \emph{Trace}
-%% % \item
-%% % Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-%% % are indeed first-class values. However these functions can only be
-%% % linearly used, i.e., they represent a constraint that can be satisfied
-%% % at most once.
-%% % %%  in one `computational world'. It may be possible that to
-%% % %% repeatedly invoke multiplicative functions with an appropriate
-%% % %% recursive type whose unfoldings allow for repeated unification.
-%% % \end{itemize}
-
-% The exact encoding on HOF in {{langRevEE}} should shed light on
-% several problems, including the connections to continuations discussed
-% in Sec. \ref{sec:related}. They should also allow for a natural
-% encoding of HOFs in the translations of \lcal to {{langRevT}} and
-% indicate a monadic encapsulation of of information effects which are
-% currrently structured as an arrow.
-
-%% If one branch uses a clause that matches one part of the input to
-%% produce one part of the output, all other clauses are automatically
-%% adjusted to account for this to ensure that this input is not matched
-%% again and that this output is not produced again.
-
-%% \item At this point, we have produced the skeleton for our desired function.
-%%   Assume now that a client uses the first clause of the function in a context
-%%   that contains the value \verb|R|, i.e., the client has the value
-%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
-%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
-%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
-%%   the first component of the pair which has the following effects:
-%%   \begin{itemize}
-%%     \item the logic variable {{alpha1}} is unified with {{R}};
-%%     \item the first component of the pair then reduces to () and can be
-%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
-%%       words, the value reduces to simply {{alpha2}} reflecting that the
-%%       pattern-matching was successful and the result of the clause is now
-%%       available;
-%%     \item the first clause of the function has now disappeared and the
-%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
-%%   \item The same client or another client can now use the second clause
-%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
-%%     second client and ground the result of the first client to {{Gr}}.
-%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
-%%     which grounds the result of the second client to {{B}}.
-%%   \end{itemize}
-
-%% Furthermore, just like the case for negative types, a dual reading of
-%% the pattern-matching clauses as continuations allows the context to
-%% use the result of the clause and provide the matching argument later.
-
-%% \begin{itemize}
-%% \item Using the fact that {{1}} is the multiplicative unit, generate from the
-%%   input {{()}} the value {{((),((),()))}} of type {{1 * (1 * 1)}};
-%% \item Apply the isomorphism {{1 <-> (1/a) * a}} in parallel to each of the
-%%   components of the above tuple. The resulting value is 
-%% {{((1/alpha1,alpha1),((1/alpha2,alpha2),(1/alpha3,alpha3)))}} where {{alpha1}},
-%% {{alpha2}}, and {{alpha3}} are fresh logic variables.
-%% \item Using the fact that {{*}} is associative and commutative, we can
-%%   rearrange the above tuple to produce the value:
-
-%% {{((1/alpha1,alpha2),((1/alpha2,alpha3),(1/alpha3,alpha1)))}}.
-
-%% \item At this point, we have produced the skeleton for our desired function.
-%%   Assume now that a client uses the first clause of the function in a context
-%%   that contains the value \verb|R|, i.e., the client has the value
-%%   {{((1/alpha1,alpha2),R)}} consiting of one of the clauses of the function
-%%   and an argument. This value can be rearranged to {{((1/alpha1,R),alpha2)}}.
-%%   At this point, we can use the isomorphism {{1 <-> (1/a) * a}} in reverse on
-%%   the first component of the pair which has the following effects:
-%%   \begin{itemize}
-%%     \item the logic variable {{alpha1}} is unified with {{R}};
-%%     \item the first component of the pair then reduces to () and can be
-%%       absorbed by the fact that {{1}} is the multiplicative unit. In other
-%%       words, the value reduces to simply {{alpha2}} reflecting that the
-%%       pattern-matching was successful and the result of the clause is now
-%%       available;
-%%     \item the first clause of the function has now disappeared and the
-%%       remaining two clauses become {{((1/alpha2,alpha3),(1/alpha3,R))}}.
-%%   \item The same client or another client can now use the second clause
-%%     applying it to {{Gr}} which would produce the result {{alpha3}} for the
-%%     second client and ground the result of the first client to {{Gr}}.
-%%   \item Finally the remaining clause {{(1/alpha3,R)}} can be applied to {{B}}
-%%     which grounds the result of the second client to {{B}}.
-%%   \end{itemize}
-% \paragraph*{Unification.}
-% A computational interpretation of this is that {{langRevEE}} acts like
-% a logic programming system and {{eta*}} acts as the site of creation
-% of a new typed logic variable of type {{b}} and its corresponding dual
-% of type {{1/b}}. These two values are free to move independently
-% through a circuit, but any operations on value affect what operations
-% are possible on the other since they are both related by an underlying
-% logic variable. Dually, {{eps*}} acts as a unification site for values
-% flowing along both branches. If the values are not exactly duals, then
-% the system fails to unify and hence the program is stuck. If the
-% values are exactly duals, then unification succeeds and {{eta*}}
-% returns {{1}}.
-% \begin{itemize}
-% \item 
-% Additive functions are not really values, but are possible paths of
-% execution. However, they can be encoded as values as follows: Consider
-% two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
-% {{bool}} can discriminate them and hence a value of type {{bool}} acts
-% as the address of the function and is a first class value
-% representing one of the {{b1 -o+ b2}} functions.  %% This is similar to
-% %% how first classes functions a'la \lcal are represented by their
-% %% addresses when compiled to conventional hardware.
-
-% \paragraph*{Entanglement and Anti-Particles.}
-% A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
-% quantum mechanics. The {{eta*}} is a site of fission that creates an
-% information particle and its anti-particle. Both flow in the same
-% direction in time. The produce particles are entangles in that actions
-% on one (such as transformation by application of an isomorphism)
-% affects the other. Further the fission site does not create a particle
-% in any specific state, but in a superposition of all possible states
-% as dictated by its type. In each possible world that the particle
-% exists, it takes on a specific value inhibiting its type and
-% correspondingly its entangled anti-particle takes the dual of the
-% specific value.
-% \item
-% Multiplicative functions, of the form {{b1 -o* b2}}, on the other hand
-% are indeed first-class values. However these functions can only be
-% linearly used, i.e., they represent a constraint that can be satisfied
-% at most once.
-% %%  in one `computational world'. It may be possible that to
-% %% repeatedly invoke multiplicative functions with an appropriate
-% %% recursive type whose unfoldings allow for repeated unification.
-% \end{itemize}
-
-% This also leads to the exciting point of view that entangled
-% particle/anti-particle pairs act as functions in a physical
-% first-class sense. We can ask the function to be applied to a value by
-% unifying the anti-particle with a particle in a well-known state and
-% examining is pair. We shall use this point of view in the development
-% our SAT solver (see Sec. \ref{sec:sat-solver}).
-% The exact encoding on HOF in {{langRevEE}} should shed light on
-% several problems, including the connections to continuations discussed
-% in Sec. \ref{sec:related}. They should also allow for a natural
-% encoding of HOFs in the translations of \lcal to {{langRevT}} and
-% indicate a monadic encapsulation of of information effects which are
-% currrently structured as an arrow.
-
-% \paragraph*{Annihilation.}
-% As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
-% additive {{trace+}}, {{trace*}} gives us the ability to find the
-% fixpoint of the constraint (expressed with combinators) that is
-% traced. If the constraint has not satisfying values, there is not
-% fixpoint possible -- we use the term `annihilation' to describe the
-% corresponding undefined state of the program. Annihilation and a
-% non-termination both describe undefined program execution -- while
-% non-termination characterizes an iteration that fails to terminate,
-% annihilation characterizes a constraint that has no fixpoint.
-
-% \paragraph*{Zero Information.}
-% As with {{eta+}}, we argue that {{eta*}} constructs no new
-% information, instead it gives us the ability to explore as many worlds
-% as permitted by the type {{b}}. So situate ourselves in any specific
-% world, we have to supply additional information in the form of
-% unifying the particle or anti-particle with a particle representing
-% known information. Hence only by supplying information to a an
-% entangled pair can we read information from it -- and always exactly
-% as much information as we supplied.
-
-% ADD MATH.
-
-% %%%%%%%%%%%%%%%%%%%%%%%%
-% \subsection{Conventional Higher-Order Functions}
-% \label{sec:hof}
-
-% In a conventional language, higher-order functions capture scope and can be
-% represented as first-class values. In {{langRevEE}}, we have two notions of
-% functions both of which share similarities with conventional higher-order
-% functions.
-
-% \begin{itemize}
-% \item 
-% Additive functions are not really values, but are possible paths of
-% execution. However, they can be encoded as values as follows: Consider
-% two functions {{f:b1 -o+ b2}} and {{g:b1-o+b2}}. A value of type
-% {{bool}} can discriminate them and hence a value of type {{bool}} acts
-% as the address of the function and is a first class value
-% representing one of the {{b1 -o+ b2}} functions.  %% This is similar to
-% %% how first classes functions a'la \lcal are represented by their
-% %% addresses when compiled to conventional hardware.
-
-%% \emph{Non-termination.}  Compact closed categories contain {{trace}}
-%% operations and this implies that {{eta+}}/{{eps+}} allow the
-%% construction of a {{trace+}} operator (shown in Section
-%% \ref{sec:monoidal-constructions}).  As shown in previous work
-%% \cite{infeffects}, non-terminating computation can be constructed from
-%%      {{trace+}} and recursive types.
-% \begin{multicols}{2}
-% \begin{center}
-%   \includegraphics{diagrams/trace.pdf}
-% \end{center}
-% %subcode{proof} include main
-% %@ c : b2 + b1 <-> b2 + b3
-% %@@ trace c : b1 <-> b3
-% \end{multicols}
-
-% \paragraph*{Zero Information.}
-% The fact that {{eta+}} (and hence {{eps+}}) introduces no information
-% can be understood is by viewing {{-b+b}} as a type whose value is
-% observable only if you `pay' {{b}} amount of information to the {{-b}}
-% branch. When {{b}} amount of information is supplied {{(-b+b)+b=b}}
-% amount of information is returned to you. (TODO. more needs to be said
-% here.)
-
-% For this reason, our treatment of negative values in the context of
-% {{langRev}} is particularly simple. We will have a type $0$ and an
-% isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-% allows us to create a value and a corresponding debt out of nothing. Both the
-% value and its negative counterpart can flow anywhere in the system. Because
-% information is preserved, a closed program (which does not have any
-% ``dangling'' negative values) will eventually have to match the negative
-% value with some corresponding value, effectively using the isomorphism in the
-% right-to-left direction.
-
-%% Hence, the intuitive way to read the diagrams is that they reverse the
-%% flow of information -- i.e. they capture a notion of \emph{negative
-%%   information flow}. In normal {{langRev}} circuits, information flows
-%% from left-to-right. The operator {{eps+}} causes information to flow
-%% from right-to-left.  The interpretation of {{eta+}} is the opposite:
-%% it turns a right-to-left flow of information in a circuit into a
-%% left-to-right flow.  The diagram below is an example of using {{eta+}}
-%% and {{eps+}} and corresponds to the additive {{a <-> a}} isomorphism
-%% from the introduction. (Here the {{0}} wires have been shown for
-%% completeness.)
-%% \paragraph*{Negative Information Flow.} 
-% A time traveling intuition is also applicable here: if view the
-% left-to-right direction of information flow as the canonical forward
-% direction of information, then the action of {{eps+}} allows values,
-% aka information particles, to flow backwards in time. Particles that
-% flow backwards in time get to see and interact with the history of
-% particles that they coexist with, i.e., values that they exist
-% multiplicatively with. Hence the isomorphism {{(-b1)*b2<->-(b1*b2)}}
-% (whose construction is shown in Section \ref{sec:neg-constructions}).
-%\subsection{Compact Closure with Additive Inverses}
-%% \emph{Coherence for Compact Closure with Additive Duality.}  The
-%% appropriate coherence condition for compact closed categories,
-%% expressed as the triangle axiom, is satisfied by {{eta+}}/{{eps+}} and
-%% may be visualized (modulo {{0}} wires, swapping etc.) as explained by
-%% Selinger \cite{springerlink:10.1007/978-3-642-12821-94} as the
-%% operational equivalence:
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Computing in the Field of Rationals : {{langRevEE}} }
-\label{sec:rat}
-
-In this section we develop an operational semantics for {{langRev}} extended
-with negative and fractional types, {{langRevEE}}. The operational semantics
-takes the abstract categorical diagrams which were previously known but gives
-them a computational interpretation based on reverse execution for negative
-types and unification for fractional types. Even though this computational
-interpretation appears straightforward in retrospect, it constitutes a
-breakthrough because it breaks the preconceived idea that there is only one
-notion of duality and hence that both negative and fractional types should be
-implemented using the same underlying mechanism. In some sense, the folklore
-literature on monoidal categories describing the evaluations for additive and
-multiplicative fragments are ``particle-style'' and ``wave-style'' can be
-seen to suggest either a unified or separate implementations. But, since de
-Morgan, we have been predisposed to think of only one notion of duality which
-persisted to even linear logic. (See Sec.~\ref{sec:related} for further
-discussion of duality.)
-
-For the purposes of this section, we start with the semantics for {{langRev}}
-defined in Sec~\ref{sec:pi}, and we will systematically rewrite it to achieve
-the desired {{langRevEE}} semantics. This rewriting consists of two main steps:
-
-\begin{enumerate}
-\item We rewrite the semantics in Sec~\ref{sec:pi} using unification of the
-  values instead of direct pattern matching. This gives us the necessary
-  infrastructure for fractional types. We assume the reader is familiar with
-  the idea of unification as realized using logic variables, substitutions,
-  and reification as presented in any standard text on logic programming.
-\item We write a reverse interpreter by reversing the direction of the
-  {{|-->}} reductions. This gives us the necessary infrastructure for
-  negative types.
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection{Unification}
-
-Previously the reductions of the primitive isomorphisms were specified in the
-following form:
-
-{{ iso v_{input pattern} |--> v_{output pattern} }}
-
-\noindent
-Instead of relying on pattern-matching, we rewrite the rules to accept an
-incoming substitution to which the required pattern-matching rules are added
-as constraints. The general case is:
-
-{{ iso s v' |--> (v_{output pattern}, s[v' <><> v_{input pattern}]) }}
-
-Using the same idea, the entire semantics can be extended to thread
-the substitution as shown below:
+\subsection{Semantics}
 
@@
 
-Most of the rules look obviously correct but there is a subtle
-point. Consider the case for {{c1 (+) c2}}. Previously the shape of the value
-determined which of the combinators {{c1}} or {{c2}} to use. Now the incoming
-value could be a fresh logical variable, and indeed we have two rules with
-the same left hand side, and only depending on the success or failure of some
-further unification, can we decide which of them applies. The situation is
-common in logic programming in the sense that it is considered a
-non-deterministic process that keeps searching for the right substitution if
-any. Typical implementations of logic programming languages further provide
-top-level mechanisms to manipulate this non-deterministic search, for example
-by returning one answer and giving the user the option to ask for more
-answers. We abstract from this top-level semantics and view the rules above
-as being applied non-deterministically.
-
-%%%%%%%%%%%%%%%%%
-\subsection{Reverse Execution}
-
 Given that our language is reversible (Prop.~\ref{prop:logrev}), a backward
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Semantics of {{langRevEE}} }
-
-Combining the two constructions above, we get the semantics of the full
-{{langRevEE}} language by adding the rules for the two variants of~{{eta}}
-and~{{eps}}.
-
-\begin{enumerate}
-\item 
-To add multiplicative duality, we add the following rules to the set
-of primitive isomorphisms:
-
-\begin{enumerate}
-\item {{eta* s () |--> ((1/v, v), s)}}
-  where {{v}} is a fresh logic variable.
-
-As explained previously {{eta*}} creates two values {{1/v}} and {{1/v}}
-from a single logic variable.
-
-\item {{eps* s v |--> (1, s[v <><> (1/v',v') ])}}
-  where {{v'}} is a fresh logic variable.
-
-Similarly, {{eps*}} unifies the values on incoming wires. The incoming
-value {{v}} represents the values of both wires and hence unifying
-them is accomplished in terms of an intermediate logic variable
-{{v'}}.
-
-\end{enumerate}
-
-\item To add negative types we add the following rules to the
+To add negative types we add the following rules to the
   reductions above. The additions formalize our previous discussions
@@
     to be negative and transferred to the {{left}} wire, and vice versa.
+
 %subcode{opsem} include main
@@
 % <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}
-   %
+
     Note that there is no evaluation rule for {{eta+}} in the forward
@@
       type of {{eps+}} is {{0}}.
+
 %subcode{opsem} include main
@@
 %! bck = \triangleleft
-% <eta+, v, C, s>_{bck} |--> <eta+, left (-v'), C, s[v <><> right v']>_{fwd}
-% <eta+, v, C, s>_{bck} |--> <eta+, right v', C, s[v <><> left (-v')]>_{fwd}
+% {[eta+, v, C, s]}_{bck} |--> {[eta+, left (-v'), C, s[v <><> right v'] ]}_{fwd}
+% {[eta+, v, C, s]}_{bck} |--> {[eta+, right v', C, s[v <><> left (-v')] ]}_{fwd}
 
@@
 
-\end{enumerate}
-
-%% %%%%%%%%%%%%%%%%%%%%%%
-%% \subsection{Constructions in {{langRevEE}} }
-%% \label{sec:monoidal-constructions}
-
-% \paragraph*{Zero Information.}
-% As with {{eta+}}, we argue that {{eta*}} constructs no new
-% information, instead it gives us the ability to explore as many worlds
-% as permitted by the type {{b}}. So situate ourselves in any specific
-% world, we have to supply additional information in the form of
-% unifying the particle or anti-particle with a particle representing
-% known information. Hence only by supplying information to a an
-% entangled pair can we read information from it -- and always exactly
-% as much information as we supplied.
-
-% ADD MATH.
-
-% \emph{Zero Information.}  The intuitive explanation that {{eta+}} and
-% hence {{eps+}} introduces no information effects is that {{-b+b}} as a
-% type whose value is observable only if you `pay' {{b}} amount of
-% information to the {{-b}} branch. When {{b}} amount of information is
-% supplied {{(-b+b)+b=b}} amount of information is returned to
-% you. (TODO. more needs to be said here.)
-
-% For this reason, our treatment of negative values in the context of
-% {{langRev}} is particularly simple. We will have a type $0$ and an
-% isomorphism {{0 <-> a + (-a)}} that when used in the left-to-right direction
-% allows us to create a value and a corresponding debt out of nothing. Both the
-% value and its negative counterpart can flow anywhere in the system. Because
-% information is preserved, a closed program (which does not have any
-% ``dangling'' negative values) will eventually have to match the negative
-% value with some corresponding value, effectively using the isomorphism in the
-% right-to-left direction.
+\begin{proposition}[Logical Reversibility]
+
+\end{proposition}
+
+\begin{proposition}[Strong Normalization]
+
+\end{proposition}
+
+\begin{proposition}[Type Safety]
+
+\end{proposition}
+
+\begin{proposition}[Coherence]
+
+\end{proposition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Computing in the Field of Rationals : {{PiEE}} }
+
+Not logically reversible and we can express infinite loops.
 
@@
 
-% (TODO: So this is a really bad explanation. But this section is really
-% important and we need to explain carefully. Maybe an analogy can be
-% drawn to Cayley diagrams for groups to point out that the notion of
-% `i' is entirely artificial, but it is essential to the discourse.)
-
-% \begin{center}
-%   \includegraphics{diagrams/shuffle.pdf}
-% \end{center}
-
-% We define observables to be only positive types. The outputs of
-% programs that output mixed positive and negative types are not
-% observable.  Also, programs that input mixed positive and negatives
-% types are not executable.
-
-% It is not fair to say that negative types flow backwards. The
-% following circuits are valid in {{langRevEE}}. It is however proper to
-% say that for any type {{b}} that flows in a direction, the type {{-b}}
-% flows in the reverse direction.
-
-% Several things we don't know how to do, or are sloppy about. Some of
-% these are deep questions to which there is no immeidate answer.
-
-% \begin{itemize}
-
-% \item Is the language without recursive types always terminating? For
-%   the right symmetry with annihilation, it is desirable to have
-%   non-termination.
-
-% \item What is the correct lemma for logical reversibility now?
-
-% \item We should say things about information preservation. At least
-%   more so than what is being said now.
-
-% \item Multiplicative duality gives us a means of talking about
-%   equivalence classes of values. This should correspond to all sorts
-%   of other things -- groupoids, Lawvere theories etc. We barely even
-%   understand these connections.
-
-% \item How does this connect to QC? Is there any strong connection such
-%   that the SAT solver here is an alternate QC algorithm to Shor's?
-
-% \end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Advanced Example: SAT Solver in {{langRevEE}} }
-\label{sec:prog}
-\label{sec:sat-solver}
-
-% Additive traces correspond to iteration and multiplicative traces
-% correspond to fixpoint of constraints. Addtive traces have been
-% studied in the past \cite{infeffects} and here we focus on the
-% multiplicative trace.
-
-% The construction presented here is a novel SAT-solver that relies on
-% {{trace*}} for its solution. It varies from previous classical SAT solvers
-% \amr{must cite ???} in that there is no explicit search operation on the
-% boolean space. It also varies from previous quantum SAT solvers \amr{must
-%   cite ???} in that it does uses a {{trace}} operation to compute a fixpoint
-% which is the solution following which an isomorphic clone operation allows us
-% to examine the result.  Though we focus on SAT here, this construction can be
-% generalized to any constraint satisfaction problem whose search space can be
-% represented by a recursive type and for which an isomorphic clone operation
-% can be constructed. 
-% can be constructed. Before we start, let us recall two constructions:
-
-We illustrate the expressive power of first-class constraints represented by
-fractional types. To understand the intuition, recall the definition of
-{{trace}} for the multiplicative fragment of {{langRevEE}}. 
-Given {{f : a*c <-> b*c }}, we have {{trace* f : a <-> b}}:
-
-{{ trace* f = uniti (;) (id (*) eta*) (;) (f (*) id) (;) (id (*) eps*) (;) unite }}
-
-\noindent This circuit uses {{eta*}} to generate all possible {{c}}-values
-together with an associated {{(1/c)}}-constraint. It then applies {{f}} to the
-pair {{(a,c)}}. The function {{f}} must produce an output {{(b,c')}} for each
-such input. If the input {{c}} and the output {{c'}} are the same they can be
-annihilated by {{eps*}}; otherwise the execution gets stuck and this
-particular choice of {{c}} is pruned.
-
-% Annihilation and non-termination describe undefined program
-% execution. While non-termination characterizes an iteration that fails
-% to terminate, annihilation characterizes a constraint that has no
-% fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
-% annihilating only the program states that do not match the boolean
-% satisfiability constraint.
-% \emph{Annihilation.}  
-% With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
-% which gives us the ability to find the fixpoint of a constraint.  If
-% the constraint has no satisfying values, there is no fixpoint possible
-% -- we use the term \emph{annihilation} to describe the corresponding
-% undefined state of the program.
-
-A large class of constraint satisfaction problems can be expressed
-using {{trace*}}. We illustrate the main ideas with the implementation
-of a SAT-solver.  We proceed in small steps, reviewing some of the
-necessary constructions presented in our earlier
-paper~\cite{infeffects}.
-
-%%%%%%%%%%%%%%%%%%%
-\subsection{Booleans and Conditionals}
-
-Given any combinator {{c : b <-> b}} we can construct a combinator called
-{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
-one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
-the combinator {{c}} is used to transform the value of type~{{b}}. If the
-boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
-write down the combinator for {{if_c}} in terms of {{c}} as 
-{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
-
-\noindent The diagram below shows the input value of type {{(1+1)*b}}
-processed by the distribute operator {{dist}}, which converts it into a value
-of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
-case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
-combinator~{{c}} is applied to the value of type~{{b}}. The right branch
-which corresponds to the boolean being {{false}} passes along the value of
-type {{b}} unchanged.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions with Negatives and Fractionals}
+\label{sec:specific-constructions}
+
+The additional constructions below (presented with minimal commentary)
+confirm that conventional algebraic manipulations in the mathematical field
+of rationals do indeed correspond to realizable type isomorphisms in our
+setting. The constructions involving both negative and fractional types are
+novel.
+
+\paragraph*{Lifting negation out of {{*}}.}
+The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
+and {{b2}} are flowing opposite to each other then it doesn't matter which
+direction is forwards and which is backwards. More interestingly as {{b1}} is
+moving backwards, it can ``see the past'' of {{b2}} which is equivalent to
+both particles moving backwards.
+
+{{(-b1) * b2 <-> -(b1 * b2) <-> b1 * (-b2)}}
+
+To build these isomorphisms, we first build an intermediate
+construction which we call {{eps_{fst} : (-b1)*b2 + b1*b2 <-> 0}}. 
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/eps_fst.pdf}
+}
+\end{center}
+
+The isomorphism {{(-b1) * b2 <-> -(b1 * b2)}} can be constructed in
+terms of {{eps_{fst} }} as shown below.
+
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{diagrams/mult_neg.pdf}
+}
+\end{center}
+
+The second isomorphism can be built in the same way by merely swapping
+the arguments.
+
+\paragraph*{Multiplying Negatives.}
+
+{{b1 * b2 <-> (-b1)*(-b2)}}
+
+This isomorphism is a consequence of the fact that $-$ is an involution: it
+corresponds to the algebraic manipulation:
+
+%% %subcode{opsem} include main
+%% %! columnStyle = rc
+%% %   & b1 * b2
+%% % = & -(-(b1*b2)) 
+%% % = & -((-b1)*b2)
+%% % = & (-b1)*(-b2)
+
+{{b1 * b2 = -(-(b1*b2)) = -((-b1)*b2) = (-b1)*(-b2)}}
+
+\paragraph*{Multiplying and Adding Fractions.}
+An isomorphism witnessing:
+
+{{b1/b2 * b3/b4 <-> (b1*b3)/(b2*b4)}}
+
+is straightforward. More surprisingly, it is also possible to construct
+isomorphisms witnessing:
+
+{{b1/b + b2/b <-> (b1+b2)/b}}
+
+{{b1/b2 + b3/b4 <-> (b1*b4+b3*b2)/(b2*b4) }}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Two Dualities of Computation}
+\label{sub:hof}
+
+Information Effects, Filinksi, De Morgan
+
+\subsection{Function Spaces}
+
+Although these constructions are also standard, they are less known and they
+are particularly important in our context: we devote a little more time to
+explain them. Our discussion is mostly based on Abramsky and Coecke's article
+on categorical quantum mechanics~\cite{abramsky-2008}.
+
+In a compact closed category, each morphism {{f : b1 <-> b2 }} can be given a
+\emph{name} and a \emph{coname}. For the additive fragment, the name
+$\fname{f}$ has type {{0 <-> (-b1 + b2)}} and the coname $\fconame{f}$ has type
+{{b1 + (-b2) <-> 0}}. For the multiplicative fragment, the name $\fname{f}$ has
+type {{1 <-> ((1/b1) * b2))}} and the coname $\fconame{f}$ has type 
+{{(b1 * (1/b2)) <-> 1}}. Intuitively, this means that for each morphism, 
+it is possible to construct, from ``nothing,'' an object in the category that 
+denotes this morphism, and dually it is possible to eliminate this object.
+The construction of the name and coname of {{c : b1 <-> b2}} in the additive case 
+can be visualized as follows:
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/function.pdf}
+\end{center}
+
+\begin{center}
+  \includegraphics{diagrams/delimc.pdf}
+\end{center}  
+\end{multicols}
+
+Intuitively the name consists of viewing {{c}} as a function and the coname
+consists of viewing {{c}} as a delimited continuation.
+
+In addition to being able to represent morphisms, it is possible to express
+function composition. For the additive case, the composition is depicted
+below:
+
+\begin{multicols}{2}
+\scalebox{1.1}{
+    \includegraphics{diagrams/compose1.pdf}
+}
+
+\scalebox{0.8}{
+    \includegraphics{diagrams/compose.pdf}
+}
+\end{multicols}
+which is essentially equivalent to sequencing both the computation blocks as
+shown below:
+
+\begin{center}
+  \includegraphics{diagrams/compose2.pdf}
+\end{center}
+
+Applying a function to an argument consists of making the argument flow
+backwards to satisfy the demand of the function:
+\begin{multicols}{2}
 \begin{center}
 \scalebox{1.0}{
-%%subcode-line{pdfimage}[diagrams/if_c.pdf]
-\includegraphics{diagrams/thesis/cnot.pdf}
-}
-\end{center}
-
-% We will be seeing many more such wiring diagrams in this paper and it is
-% useful to note some conventions about them. Wires indicate a
-% value that can exist in the program. Each wire, whenever possible, is
-% annotated with its type and sometimes additional information to help clarify
-% its role. When multiple wires run in parallel, it means that those values
-% exist in the system at the same time, indicating pair types. When there is a
-% disjunction, we put a {{+}} between the wires. 
-% Combinators for distribution {{dist}} and factoring {{factor}}
-% are represented as triangles with their operator symbols in them. Other
-% triangles may be used and, in each case, types or labels will be used to
-% clarify their roles. Finally, we don't draw boxes for combinators such as
-% {{id}}, commutativity, and associativity, but instead just shuffle the wires
-% as appropriate.
-
-The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
-negates its second argument if the first argument is {{true}}. This
-gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
-equivalent construction that is useful is {{else_{not} }} where we
-negate the second argument only if the first is {{false}}.
-
-Similarly, we can iterate the construction of {{if_c}} to check several
-bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
-checks two booleans and negates the result wire only if they are both
-{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
-universal reversible gate. We can generalize this construction to
-{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
-they are all {{true}}.
-
-%%%%%%%%%%%%%%%%%%
-\subsection{Cloning}
-
-Although cloning is generally not allowed in reversible languages, it is
-possible at the cost of having additional constant inputs. For example,
-consider the gate {{else_{not} }}. Generally, the gate maps
-{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
-on the cases in which the second input is {{true}}, we get that the gate maps
-{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
-i.e., the gate clones the first input. A circuit of {{n}} parallel
-{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
-{{true}} inputs in the process.  Let us call this construction
-{{clone^n_{bool} }}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Construction of the Solver}
-
-The key insight underlying the construction comes from the fact that we can
-build \emph{annihilation circuits} such as the one below:
-
-\begin{center}
-\scalebox{1.2}{
-  \includegraphics{diagrams/not_trace.pdf}
-}
-\end{center}
-The circuit constructs a boolean {{b}} and its dual {{1/b}}, negates one of
-them and attempts to satisfy the constraint that they are equal which
-evidently fails.
-
-With a little work, we can modify this circuit to only annihilate values that
-fail to satisfy the constraints represented by a SAT-instance {{f}}. In more
-detail, an instance of SAT is a function/circuit~{{f}} that given some
-boolean inputs returns {{true}} or {{false}} which we interpret as whether
-the inputs satisfy the constraints imposed by the structure of {{f}}. Because
-we are in a reversible world, our instance of SAT must be expressed as an
-isomorphism: this is easily achieved as shown in Sec.~\ref{sub:f}
-below. Assuming that {{f}} is expressed as an isomorphism, we have enough
-information to reconstruct the input from the output. This can be done by
-using the adjoint of {{f}}. At this point we have, the top half of the
-construction below:
-
-\begin{center}
-\scalebox{1.2}{
-  \includegraphics{diagrams/sat1.pdf}
-}
-\end{center}
-
-To summarize, the top half of the circuit is the identity function except
-that we have also managed to produce a boolean wire labeled
-\textsf{satisfied?} that tells us if the inputs satisfy the desired
-constraints. We can take this boolean value and use it to decide whether to
-negate the control wire or not. Thus, the circuit achieves the following
-goal: if the inputs do not satisfy {{f}}, the control wire is negated.  We
-can now use {{trace*}} to annihilate all these bad values because the control
-wire acts like the closed-loop {{not}} in the previous construction.
-
-%%%%%%%%%
-\subsection{Final Details}
-\label{sub:f}
-
-Any boolean expression {{f : bool^n -> bool}} can be compiled into the
-isomorphism {{iso_f : bool^h*bool^n<->bool^g*bool}} where the extra bits
-{{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
-such an isomorphism has been detailed
-before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
-applying {{iso_f}} to some special heap values and an input \textit{bs}
-produces some bits that can be ignored and the same output that {{f}} would
-have produced on {{bs}}. We can ensure that the heap has the appropriate
-initial values by checking the heap and negating a second control wire, if
-the values do not match, i.e., the dotted part in the diagram below.
-
-\begin{center}
-\scalebox{1.2}{
-  \includegraphics{diagrams/sat2.pdf}
-}
-\end{center}
-
-% \roshan{Explain how the dotted part is implemented.}
-
-Let us call the above construction which maps inputs, heap, and
-control wires to inputs, heap, and control wires as {{sat_f}}. The
-SAT-solver is is completed by tracing the {{sat_f}} and cloning the
-inputs using {{clone^n_{bool} }}.
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/sat3.pdf}
-}
-\end{center}
-
-When the solver is fed inputs initialized to {{true}}, it clones only
-those inputs to {{sat_f}} that satisfy {{f}} and the heap
-constraints. In the case of unique-SAT the solver will produce exactly
-0 or 1 solutions. In the case of general SAT, the solver will produce
-solutions as determined by the semantics of the top-level interaction
-(see discussion in Sec. \ref{sec:rat}).
-
-%% %%%%%%%%%%%%%%%%%%%
-%% \subsection{GoI machine}
-
-%% We can now encode the GoI machine of
-%% Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. \amr{The machine uses bang
-%%   which allows arbitrary duplication. Can we really do that???}
-
-%% the final equivalence is valid and can be used to actually
-%% construct an isomorphism between $x^7$ the type of seven binary
-%% trees and $x$ the type of binary trees.
-
-%% Here is one way to view computations involving irrationals that might
-%% shed some insight into how we should view recursive data types
-%% also. For finite datatypes say of size {{n}} we can say that an
-%% element specifies a choice of {{1/n}} and hence contains as much
-%% information. For finite things, each value is computationally equal to
-%% any other value. Any permutation is possible among the values. Hence
-%% there is no metric possible on the values.  Recursive data types, such
-%% as binary trees, have an infinite number of elements and hence the
-%% argument that says that value specifies {{1/n}} units of information
-%% is not longer valid. Instead to talk meaningfully about information
-%% content, we should talk about a probability distribution on the values
-%% of the type. However what shall we base this distribution on? Is there
-%% a natural order beyond the unfolding order of the values of a type?
-
-%% In this context, let us go back to the thought that recursive such as
-%% {{nat=nat+1}} have no meaningful solutions, whereas trees do.
-
-%% Real numbers in that sense really are a series of bits of unknown
-%% information theoretic interpretation -- a sort of {{top}}. Numbers
-%% such as {{sqrt(2) }} are very different in the sense that even though
-%% they contain an infinite sequence of bits, there is a finite program
-%% that generates those bits and its only the cost of computing bits that
-%% is infinite.  The consistency of arithmetic isn't affected by
-%% extending the rational field by adding any specific computable
-%% irrational. So it might be possible to design a model of computing
-%% where we can indeed deal with them. Such a model will have to explicit
-%% thunks for the delayed computation represented by trees and never
-%% equate and cancel two infinite computations unless the represent
-%% exactly the same infinite computation.
-
-%% TODO.
-
-%% Square root and imaginary types have also appeared in the literature: we
-%% relegate the connections to Sec.~\ref{sec:related} and proceed with a simple
-%% explanation. We have so far extended the set of types to be the rational
-%% numbers. Now we will push this and extend the set of types to algebraic
-%% numbers. In other words, we will allow datatypes defined by arbitrary
-%% polynomials and allow the roots of such polynomials to be types.
-
-%% Consider first an example in which we want to compute with the sides of a
-%% rectangle whose area is 91 and whose length is longer than its width by 6
-%% units. One can solve the quadratic equation to determine that the sides are 7
-%% and 13 and proceed. This however prematurely forces us to globally solve the
-%% constraint. Instead we can let the two sides of the rectangle be $x$ and
-%% $x+6$ and use the following equation to capture the desired constraint:
-%% \[
-%% x^2 + 6x - 91 = 0
-%% \]
-%% The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
-%% the type $0$. We can now proceed to compute with the unknown $x$, being
-%% assured that in a closed program, our computation will eventually be
-%% consistent with the solution of the algebraic equation.
+  \includegraphics{diagrams/apply1.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/apply2.pdf}
+}
+\end{center}
+\end{multicols}
+
+Having reviewed the representation of functions, we now discuss the
+similarities and differences between the two notions of functions and their
+relation to conventional (linear) functions which mix additive and
+multiplicative components. For that purpose, we use a small example.
+Consider a datatype {{color = R|Gr|B}}, and let us consider the following
+manipulations:
+\begin{itemize}
+\item Using the fact that {{1}} is the multiplicative unit, generate
+  from the input {{()}} the value {{((),())}} of type {{1 * 1}};
+\item Apply the isomorphism {{1 <-> (1/b) * b}} in parallel to each of
+  the components of the above tuple. The resulting value is
+  {{((1/alpha1,alpha1),(1/alpha2,alpha2))}} where {{alpha1}} and
+  {{alpha2}} are fresh logic variables;
+\item Using the fact that {{*}} is associative and commutative, we can
+  rearrange the above tuple to produce the value:
+
+{{((1/alpha1,alpha2),(1/alpha2,alpha1))}}.
+
+\end{itemize}
+
+At this point we have constructed a strange mix of two {{b-o*b}} functions;
+inputs of one function manifest themselves as outputs of the other. If
+{{(1/alpha1,alpha2)}} is held by one subcomputation and {{(1/alpha2,alpha1)}}
+is held by another subcomputation, these remixed functions form a
+communication channel between the two concurrent subcomputations. Unifying
+{{1/alpha1}} with {{color}} {{R}} in one subcomputation, fixes {{alpha1}} to
+be {{R}} in the other. The type {{b}} thus takes the role of the type of the
+communication channel, indicating how much information can be communicated
+between the two subcomputations.  Depending on the choice of the type {{b}},
+an arbitrary number of bits may be communicated.
+
+Dually, the additive reading of the above manipulations correspond to
+functions of the form {{b-o+b}}, witnessing isomorphisms of the form
+{{0<->(-b)+b}}). The remixed additive functions express control flow transfer
+between two subcomputations, \emph{only one of which exists} at any point,
+i.e., they capture the essence of coroutines.
+
+It should be evident that in a universe in which information is not
+guaranteed to be preserved by the computational infrastructure, the above
+slicing and dicing of functions would make no sense. But linearity is not
+sufficient: one must also recognize that the additive and multiplicative
+spaces are different.
 
@@
 
-We have extended the language {{langRev}} that expressed computation
-in the commutative semiring of whole numbers to {{langRevEE}} that
+We have extended the language {{Pi}} that expressed computation
+in the commutative semiring of whole numbers to {{PiEE}} that
 expresses computation in the field of rationals.  Every algebraic
@@
 suggest that many of the GoI machine constructions can be simulated in
-{{langRevEE}} by treating Mackie's bi-directional wires as pairs of wires in
-{{langRevEE}} and replacing the machine's global state with a typed value on
+{{PiEE}} by treating Mackie's bi-directional wires as pairs of wires in
+{{PiEE}} and replacing the machine's global state with a typed value on
 the wire that captures the appropriate state. This connection is exciting
 because when viewed through a Curry-Howard lens it suggests that the logical
-interpretation of {{langRevEE}} would be a linear-like logic with a notion of
+interpretation of {{PiEE}} would be a linear-like logic with a notion of
 resource preservation and with a natural computational interpretation.
@@
 \label{sec:algebraic-field}
-Algebraically, the move from {{langRev}} to {{langRevEE}} corresponds to a
-move from a ring-like structure to a full field. Our language {{langRevEE}}
+Algebraically, the move from {{Pi}} to {{PiEE}} corresponds to a
+move from a ring-like structure to a full field. Our language {{PiEE}}
 captures the structure of one particular field: that of the rational
@@
 
-We have confirmed that we can extended {{langRevEE}} with the datatype
+We have confirmed that we can extended {{PiEE}} with the datatype
 declaration for binary trees and build a witness for this isomorphism
@@
 between $x^6$ and $1$ has no computational context. In the setting of
-{{langRevEE}}, this isomorphism can be constructed but it diverges on all
+{{PiEE}}, this isomorphism can be constructed but it diverges on all
 inputs (in both ways). This suggests that, in the field of algebraic numbers,
@@
 
-%% or dually that {{langRevEE}} with recursive types
+%% or dually that {{PiEE}} with recursive types
 %%somehow lacks the ability to express all computations that are otherwise
@@
 the ``right one'' but it is interesting that we can directly realize it using
-the primitives of {{langRevEE}}.
+the primitives of {{PiEE}}.
 
@@
 
-The multiplicative structure of {{langRevEE}} also has a direct
+The multiplicative structure of {{PiEE}} also has a direct
 connection to entangled quantum particles, or perhaps entangled
@@
 it is again interesting that we can directly realize the idea using
-the primitives of {{langRevEE}}.
+the primitives of {{PiEE}}.
 
@@
 %% In a technical sense, this paper extends the language of isomorphisms
-%% {{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
+%% {{Pi}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
 %% other systems which have one notion of duality over additive and
-%% multiplicative components, {{langRevEE}} has two notions of duality
+%% multiplicative components, {{PiEE}} has two notions of duality
 %% -- an additive duality over the monoid {{(0, +)}} and a multiplicative
 %% duality over the monoid {{(1, *)}}. Each axis of duality also give us
-%% a function space and hence {{langRevEE}} has an additive function
+%% a function space and hence {{PiEE}} has an additive function
 %% space corresponding to a notion of control or backtracking and a
@@
 We show a deep symmetry between functions and delimited continuations, values
-and continuations that arises in {{langRev}} in a manner that is reminiscent
+and continuations that arises in {{Pi}} in a manner that is reminiscent
 of Filinski's Symmetric \lcal ~\cite{Filinski:1989:DCI:648332.755574}. The
-symmetry arises by extending {{langRev}} with a notion of additive duality
+symmetry arises by extending {{Pi}} with a notion of additive duality
 over the monoid {{(+, 0)}} by including {{eta+}} and {{eps+}} operators of
@@
 for the encoding of higher-order function and iteration via the construction
-of trace operators, thereby making the extended language {{langRevEE}} a
+of trace operators, thereby making the extended language {{PiEE}} a
 Turing-complete reversible programming language with higher-order functions
@@
 construction of Abramsky \cite{Abramsky96:0} from Haghverdi. However, note
-that the {{langRevEE}} is not the same as the image of the Int construction
-on {{langRevT}}, since the later lacks a multiplicative tensor that
-distributes over the additive tensor in Int {{langRevT}}.
+that the {{PiEE}} is not the same as the image of the Int construction
+on {{Pi0}}, since the later lacks a multiplicative tensor that
+distributes over the additive tensor in Int {{Pi0}}.
 

### 2012-06-19 rational_ext.tex
_

@@
 
-\authorinfo{Roshan P. James}
-           {Indiana University}
-           {rpjames@indiana.edu}
-\authorinfo{Amr Sabry}
-           {Indiana University}
-           {sabry@indiana.edu}
+\authorinfo{Nicolas Bourbaki}
+
+%% \authorinfo{Roshan P. James}
+%%            {Indiana University}
+%%            {rpjames@indiana.edu}
+%% \authorinfo{Amr Sabry}
+%%            {Indiana University}
+%%            {sabry@indiana.edu}
 
@@
 
+  This paper connects the line of work starting with
+  Filinski\cite{Filinski:1989:DCI:648332.755574} on the duality of
+  computation \cite{Curien:2000,DBLP:conf/rta/Wadler05} with work on
+  information preserving computation \cite{infeffects,rc2011}.
   Andrzej Filinski's masters thesis
@@
   Previous work on \emph{information effects} established the notion
-  of information preserving computation
-  \cite{infeffects,rc2011,rc2012} and the core calculus for computing
-  with the isomorphisms of finite types, {{Pi}}. Given the
-  first-order nature of {{Pi}} however it was unclear how one may
-  express first-class functions and duality in the sense of Filinski.
+  of information preserving computation \cite{infeffects} and the core
+  calculus for computing with the isomorphisms of finite types,
+  {{Pi}}. Given the first-order nature of {{Pi}} however it was
+  unclear how one may express first-class functions and duality in the
+  sense of Filinski.
 
@@
     rational numbers}. This computational model is derived
-  systemmatically from the {{Pi}} by considering a symmetric
-  notion of duality. Unlike the \lcal which shows a classical demorgan
-  duality, here we have two axis of dualization and hence two
-  dualities -- an additive duality, namely negative types and a
-  multiplicative duality, namely fractional types.
+  systemmatically from the {{Pi}} by considering a symmetric notion of
+  duality. Unlike the \lcal which shows a classical De Morgan duality,
+  here we have two axis of dualization and hence two dualities -- an
+  additive duality, namely negative types and a multiplicative
+  duality, namely fractional types.
 

### 2012-06-19 rational_ext.tex
_

@@
 
-Andrzej Filinksi's masters thesis
+Andrzej Filinski's masters thesis
 \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
@@
 \begin{proposition}[Correspondence]
-  Evalution in the small-step evaulator corresponds to evaluation in
+  Evaluation in the small-step evaluator corresponds to evaluation in
   the natural semantics.
@@
 
-For a traced symmetric monoidal catgeory, the Int-construction of
+For a traced symmetric monoidal category, the Int-construction of
 Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
@@
 
-Provide a semantics with the nondeterminism monad and define an
+Provide a semantics with the non-determinism monad and define an
 \emph{eval}.
@@
 seem badly behaved (no logical reversibility) we have a close
-connection with finite dimensional hilbert spaces.
+connection with finite dimensional Hilbert spaces.
 
 The technical goal of this section is to show that the categorical
-structure of finite dimensionl hilbert spaces with unitary maps and
+structure of finite dimensional Hilbert spaces with unitary maps and
 inner product has the same categorical structure as {{PiEE*}}. Thus
@@
 connections to the Quantum Picturisms of Abramsky and Coecke and hence
-to Qunatum Physics.
+to Quantum Physics.
 
@@
 consequence of this line of research is either (1) there is no way to
-accomodate proabailities etc in this setting and this is all
-meaningless or (2) constrcutions in {{PiEE*}} correspond to physical
+accommodate probabilities etc in this setting and this is all
+meaningless or (2) constructions in {{PiEE*}} correspond to physical
 phenomena.

### 2012-06-21 rational_ext.tex
_

@@
 
-The semantics of the primitive combinators is given by the following
-single-step reductions. Since there are no values of type {{0}}, the rules
-omit the impossible cases:
+The operational semantics of {{Pi}} is summarized below (also see
+\cite{infeffects}).  The semantics of the primitive combinators is
+given by the following single-step reductions below. Since there are
+no values of type {{0}}, the rules omit the impossible cases:
 \begin{scriptsize}
@@
 % factor & (right (v2, v3)) &|-->& (right v2, v3)
+
+The operational semantics of the closure conditions are presented in
+the usual big-step style.
 
@@
 
-\begin{proposition}[Type Safety]
-
-\end{proposition}
+The type safety of {{Pi}} follows directly from the corresondence
+between inductive definitions of the types and the big-step
+semantics. Previous work also established the following properties:
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Categorical Structure}
+
+For the purpose of establishing equivalence of morphisms in the
+categorical presentation we use the following notion of equality:
+
+\begin{lemma}[{{c1  = c2}}]
+We say {{c1 = c2}} if {{c1:b1<->b2}} and {{c2:b1<->b2}} and for all
+{{v1:b}}, we have {{c1 v1 |--> v2}} iff {{c2 v1 |--> v2}} and {{v2:b2}}.
+\end{lemma}
+
+\begin{lemma}[{{Pi}} is a category]
+  We have as objects the types {{b}} and as morphisms equivalence
+  classes of combinators {{b1<->b2}}. One can check:
+  \begin{enumerate}
+  \item Every object {{b}} has an identity morphism {{b <->b}}.
+  \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
+    {{g:b2<->b3}} is given by sequencing {{f (;) g}}. 
+  \item Associativity of composition follows from operational
+    equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
+    {{h:b3<->b4}}). One can check:
+
+%subcode{proof} include main
+%@@ f v1 |--> v2
+%@ g v2 |--> v3
+%@ h v3 |--> v4
+%@@ g (;) h  v2 |--> v4
+%@@@ f (;) (g (;) h) v1 |--> v4
+
+and
+
+%subcode{proof} include main
+%@ f v1 |--> v2
+%@ g v2 |--> v3
+%@@ f (;) g v1 |--> v3
+%@@ h v3 |--> v4
+%@@@ (f (;) g) (;) h v1 |--> v4
+
+  \item Composition respects identity: {{id (;) f = f}} and {{g(;)id=g}}.
+  \end{enumerate}
+\end{lemma}
+
+\begin{lemma}[Symmetric Monoidal (+, 0)]
+  {{Pi}} is a symmetric monoidal category with tensor {{+}} and
+  monoidal unit {{0}}. For any object {{b1+b2}} the required symmetry
+  morphism (called in categorical literature) is
+  {{swap+:b1+b2<->b2+b1}}.
+\end{lemma}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Small Step Semantics}

### 2012-06-21 rational_ext.tex
_

@@
 
+\begin{lemma}[Dagger]
+  {{Pi}} is a dagger category, where every morphism {{f : b1<->b2}}
+  has the adjoint {{f^{dagger}:b2 <->b1}}. The following properties
+  hold (where {{g : b2 <-> b3}}): 
+  \begin{enumerate}
+  \item {{id^{dagger} = id : b <-> b}}
+  \item {{(f (;) g)^{dagger} = f^{dagger} (;) g^{dagger}: b1 <-> b2}}
+  \item {{f^{dagger dagger} = f : b1 <-> b2}}
+  \end{enumerate}
+\end{lemma}
+
 \begin{lemma}[Symmetric Monoidal (+, 0)]
   {{Pi}} is a symmetric monoidal category with tensor {{+}} and
-  monoidal unit {{0}}. For any object {{b1+b2}} the required symmetry
-  morphism (called in categorical literature) is
-  {{swap+:b1+b2<->b2+b1}}.
+  monoidal unit {{0}}. The monoidal operation on morphisms is the
+  additive composition of combinators {{c1 (+) c2}}.
+
+  \begin{proof}
+    To establish that category is monoidal one must show isomorphisms
+    \begin{enumerate}
+    \item {{b1 + (b2 + b3) <-> (b1 +b2) + b3}} given by {{assoc+}}
+    \item {{0 + b <-> b}} given by {{zeroe}}
+    \item {{b + 0 <-> b}} given by {{swap+ (;) zeroe}}
+    \item We need to check that {{+}} is a bifunctor. 
+      \begin{enumerate}
+      \item {{id_{b1}(+)id_{b2} = id_{b1+b2} }}
+      \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}
+      \end{enumerate}
+    \item We need to check the naturality of {{assoc+}}, {{zeroe}} and
+      {{swap+(;)zeroe}}.
+      \begin{enumerate}
+      \item {{assoc+ (;) ((f (+)g) (+) h) = (f (+)(g (+) h)) (;) assoc+}}
+      \item {{zeroe (;) f = (id (+) f) (;) zeroe}}
+      \item {{(swap+ (;) zeroe) (;) f = (f (+) id) (;) (swap+ (;) zeroe)}}
+      \end{enumerate}
+    \item Satisfy certain coherence conditions which are usually
+      called the pentagon and traingle axioms.
+    \end{enumerate}
+
+    The last three points require checking equality of combinators by
+    writing out their derivation trees as we did in the case of
+    associativity of sequential composition. To show symmetry, we need
+    a braiding operation {{b1+b2 <-> b2+b1}} which is given by
+    {{swap+}}.
+
+    \begin{enumerate}
+    \item The braiding must satisfy two ``hexagon axioms''. 
+    \item The braiding is self inverse {{swap+_{b1+b2} = (swap+_{b2+b1})^{dagger} }}.
+    \end{enumerate}
+
+  \end{proof}
+
 \end{lemma}
+
+\begin{lemma}[Symmetric Monoidal (*, 1)]
+  {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
+  unit {{1}}. The details here follow those of the {{(0, +)}} monoid
+  and are skipped.
+\end{lemma}
+
+Some technical and pedantic comments are due at this point.
+
+\begin{itemize}
+\item We have established the categorical structure of {{Pi}} as a
+  dagger symmetric monoidal category with two monoidal structures,
+  {{(0, +)}} and {{(*, 1)}}. In Sec. \ref{sec:int} we will see that a
+  {{trace}} operator can be admitted in this category without any
+  change of expressiveness.
+
+\item By definition, in {{Pi}} every morphism is an isomorphism --
+  this makes {{Pi}} a groupoid.
+
+\item The category {{Pi}} has no initial and terminal objects. The
+  objects {{0}} and {{1}} would be initial and terminal if we admitted
+  all functions (is in the category {{CatSet}}).
+
+\item The category {{Pi}} has neither categorical products, nor
+  categorical co-products -- {{*}} and {{+}} are merely monoidal
+  tensors. This follows from the fact that injection and projection
+  maps are not isomorphisms.
+\end{itemize}
+
+While we don't do so in this work, if we extend {{Pi}} with products
+and co-products (say through the addition of information effects) it
+is concievable that part of its structure collapses. The sort of
+collapse, while catastrophic for algebraic structures (effectively
+trivializing them) still retains some interest in computing, because in
+computing we are interested in the specific operational nature (the
+computation content, so to speak) of the morphisms. Anectodal evidence
+follows from the fact several real-world programming language have
+inconsistent type systems. blah blah blah...
 

### 2012-06-21 rational_ext.tex
_

@@
 %! ^^^ = ^{-1}
-%! (o) = \fatsemi
+%! (o) = \circ
 %! (;) = \fatsemi
@@
 %! PiEE = \Pi^{\eta\epsilon}
+%! CatSet = \textbf{Set}
 %! theseus = Theseus
@@
 
+We present the categorical structure with minimum commentry. More
+details may be found in excellent references such as Barr and Wells
+[CITE] and Selinger[CITE].
+
 \begin{lemma}[{{Pi}} is a category]
-  We have as objects the types {{b}} and as morphisms equivalence
-  classes of combinators {{b1<->b2}}. One can check:
+  The category {{Pi}} has the types {{b}} as objects and equivalence
+  classes of well-typed combinators {{b1<->b2}} as morphisms. One can
+  check:
   \begin{enumerate}
-  \item Every object {{b}} has an identity morphism {{b <->b}}.
+  \item Every object {{b}} has an identity morphism {{id : b <->b}}.
   \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
@@
   \item {{id^{dagger} = id : b <-> b}}
-  \item {{(f (;) g)^{dagger} = f^{dagger} (;) g^{dagger}: b1 <-> b2}}
+  \item {{(f (;) g)^{dagger} = g^{dagger} (;) f^{dagger}: b3 <-> b1}}
   \item {{f^{dagger dagger} = f : b1 <-> b2}}
@@
 
-\begin{lemma}[Symmetric Monoidal (*, 1)]
+\begin{lemma}[Symmetric Monoidal {{(*, 1)}}]
   {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
-  unit {{1}}. The details here follow those of the {{(0, +)}} monoid
-  and are skipped.
+  unit {{1}}. The details mirror those of the {{(0, +)}} monoid.
 \end{lemma}
 
+\noindent
 Some technical and pedantic comments are due at this point.
@@
   objects {{0}} and {{1}} would be initial and terminal if we admitted
-  all functions (is in the category {{CatSet}}).
+  all functions (as in the category {{CatSet}}).
 
 \item The category {{Pi}} has neither categorical products, nor
-  categorical co-products -- {{*}} and {{+}} are merely monoidal
+  categorical co-products, i.e. {{*}} and {{+}} are merely monoidal
   tensors. This follows from the fact that injection and projection
@@
 and co-products (say through the addition of information effects) it
-is concievable that part of its structure collapses. The sort of
+is concievable that part of its structure collapses. This sort of
 collapse, while catastrophic for algebraic structures (effectively
-trivializing them) still retains some interest in computing, because in
-computing we are interested in the specific operational nature (the
-computation content, so to speak) of the morphisms. Anectodal evidence
+trivializing them), still retains some interest in computing, because
+in computing we are interested in the specific operational nature (the
+computational content, so to speak) of the morphisms. Anectodal evidence
 follows from the fact several real-world programming language have

### 2012-06-21 rational_ext.tex
_

@@
 % factor & (right (v2, v3)) &|-->& (right v2, v3)   
-
+\end{scriptsize}
 The operational semantics of the closure conditions are presented in
 the usual big-step style.
-
+\begin{scriptsize}
 %subcode{proof} include main
@@
       \begin{enumerate}
-      \item {{id_{b1}(+)id_{b2} = id_{b1+b2} }}
+      \item {{id_{b1<->b1}(+)id_{b2<->b2} = id_{b1+b2<->b1+b2} }}
       \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}
@@
     \item Satisfy certain coherence conditions which are usually
-      called the pentagon and traingle axioms.
+      called the ``pentagon'' and ``triangle'' axioms.
     \end{enumerate}

### 2012-06-21 rational_ext.tex
_

@@
 \label{prop:termination-pi} 
-{{Pi}} computation always terminate.
+{{Pi}} computations always terminate.
 
@@
 details may be found in excellent references such as Barr and Wells
-[CITE] and Selinger[CITE].
+[CITE] and Selinger \cite{springerlink:10.1007/978-3-642-12821-94}.
 
@@
     \item Satisfy certain coherence conditions which are usually
-      called the ``pentagon'' and ``triangle'' axioms.
+      called the ``pentagon'' and ``triangle'' axioms (see Sec 3.1
+      \cite{springerlink:10.1007/978-3-642-12821-94})
     \end{enumerate}
@@
     a braiding operation {{b1+b2 <-> b2+b1}} which is given by
-    {{swap+}}.
+    {{swap+}} (see Sec. 3.3 and 3.5
+    \cite{springerlink:10.1007/978-3-642-12821-94}).
 
@@
     \item The braiding must satisfy two ``hexagon axioms''.
-    \item The braiding is self inverse {{swap+_{b1+b2} = (swap+_{b2+b1})^{dagger} }}.
+    \item The braiding is self inverse
+      {{swap+_{b1+b2}=(swap+_{b2+b1})^{dagger} }}.
     \end{enumerate}
@@
   change of expressiveness.
+
+\item To be pedantic, what we have shown is that the ``term model'' of
+  {{Pi}} that follows from the extensional operational equality of
+  combinators has the requisite categorical structure. A consequence
+  is that the ``wiring diagrams'' of {{Pi}} correspond closely with ``string
+  diagrams'' developed for categories.
+
+  To establish the later rigorously, we will need to show when it is
+  valid to slide one wire over the other and that equivalent diagrams
+  for syntactically different combinators such as
+  {{(f(+)g)(;)(j(+)k)}} and {{(f(;)j)(+)(g(;)k)}} do respect
+  operational equivalence.  We don't formalize the graphical notation
+  in this work. Joyal et. al's work on ``planar isotopy'' and
+  Selinger's survey \cite{springerlink:10.1007/978-3-642-12821-94}
+  show how this has been addressed before in the categorical
+  setting. In the absense of any prior knowledge of category theory
+  however, our wiring diagrams may be read as the ``flow of types'' in
+  a combinator-circuit.
 
@@
 \section{Traces and the Int-Construction.}
+\label{sec:int}
 

### 2012-06-21 rational_ext.tex
_

@@
 %! @@ = \mu
+%! [^ = \lceil
+%! ^] = \rceil
 %! forall = \forall
@@
 
+\noindent
 To summarize, the syntax of {{Pi}} is given as follows.
@@
 
-%%%%%%%%%%%%%%%
+\begin{definition}[Size of a type]
+\label{def:size}
+The size of a type {{b}} is a number defined to be:
+\vspace{-20pt}
+\begin{multicols}{2}
+%subcode{opsem} include main
+% [^ b1 + b2 ^] '= [^ b1 ^] + [^ b2 ^]
+% [^ b1 * b2 ^] '= [^ b1 ^] * [^ b2 ^]
+
+%subcode{opsem} include main
+% [^ 0 ^] '= 0
+% [^ 1 ^] '= 1
+\end{multicols}
+
+where {{+}} is numeric addition and {{*}} is numeric multiplication.
+
+\end{definition}
+\noindent
+In the setting of {{Pi}} the \emph{size}, {{[^ b ^]}}, may be simply thought of as
+the number of inhabitants (an arity) of the type. This intuition will
+however become tenuous in the presence of negative and fractional
+types.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Graphical Language}
@@
 
+\item Composition TODO.
+
 \end{itemize}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics}
@@
 
+\vspace{-20pt}
+    \begin{multicols}{2}
+      \begin{scriptsize}
+
 %subcode{proof} include main
@@
 %@@@ f (;) (g (;) h) v1 |--> v4
-
-and
-
+~
 %subcode{proof} include main
@@
 
+      \end{scriptsize}
+    \end{multicols}
+
   \item Composition respects identity: {{id (;) f = f}} and {{g(;)id=g}}.
@@
   \begin{enumerate}
-  \item {{id^{dagger} = id : b <-> b}}
-  \item {{(f (;) g)^{dagger} = g^{dagger} (;) f^{dagger}: b3 <-> b1}}
-  \item {{f^{dagger dagger} = f : b1 <-> b2}}
+  \item {{id^{dagger} = id : b <-> b}}.
+  \item {{(f (;) g)^{dagger} = g^{dagger} (;) f^{dagger}: b3 <-> b1}}.
+  \item {{f^{dagger dagger} = f : b1 <-> b2}}.
   \end{enumerate}
@@
     \begin{enumerate}
-    \item {{b1 + (b2 + b3) <-> (b1 +b2) + b3}} given by {{assoc+}}
-    \item {{0 + b <-> b}} given by {{zeroe}}
-    \item {{b + 0 <-> b}} given by {{swap+ (;) zeroe}}
+    \item {{b1 + (b2 + b3) <-> (b1 +b2) + b3}} given by {{assocl+}}.
+    \item {{0 + b <-> b}} given by {{zeroe}}.
+    \item {{b + 0 <-> b}} given by {{swap+ (;) zeroe}}.
     \item We need to check that {{+}} is a bifunctor. 
       \begin{enumerate}
-      \item {{id_{b1<->b1}(+)id_{b2<->b2} = id_{b1+b2<->b1+b2} }}
-      \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}
+      \item {{id_{b1<->b1}(+)id_{b2<->b2} = id_{b1+b2<->b1+b2} }}.
+      \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}.
       \end{enumerate}
-    \item We need to check the naturality of {{assoc+}}, {{zeroe}} and
+    \item We need to check the naturality of {{assocl+}}, {{zeroe}} and
       {{swap+(;)zeroe}}.
       \begin{enumerate}
-      \item {{assoc+ (;) ((f (+)g) (+) h) = (f (+)(g (+) h)) (;) assoc+}}
-      \item {{zeroe (;) f = (id (+) f) (;) zeroe}}
-      \item {{(swap+ (;) zeroe) (;) f = (f (+) id) (;) (swap+ (;) zeroe)}}
+      \item {{assocl+ (;) ((f (+)g) (+) h) = (f (+)(g (+) h)) (;) assocl+}}.
+      \item {{zeroe (;) f = (id (+) f) (;) zeroe}}.
+      \item {{(swap+ (;) zeroe) (;) f = (f (+) id) (;) (swap+ (;) zeroe)}}.
       \end{enumerate}
@@
     \item The braiding must satisfy two ``hexagon axioms''.
-    \item The braiding is self inverse
+    \item The braiding is self inverse,
       {{swap+_{b1+b2}=(swap+_{b2+b1})^{dagger} }}.

### 2012-06-21 code/modular/Lang.hs
split things into pieces; add ND

@@ -1,4 +1,5 @@
 -- {-# OPTIONS_GHC -XGADTs -XTypeOperators #-} -- 7.0.1
+{-# LANGUAGE TypeFamilies, FlexibleContexts, GADTs #-}
 
 -- Based on code
 -- Copyright (c) 2011, William J. Bowman, Roshan P. James, and Amr
@@ -11,6 +12,8 @@
 module Lang where
 
 import Prelude hiding (id)
+import qualified Prelude as P
+import Control.Monad
 
 class Pi eq where
 -- Congruence
@@ -36,12 +39,18 @@ class Pi eq where
   timesZeroR  :: eq Zero (Zero, a)
   distribute  :: eq (Either b c, a) (Either (b, a) (c, a))
   factor      :: eq (Either (b, a) (c, a)) (Either b c, a)
+
+class Pi eq => PiBool eq where
 -- Encoding of booleans
   foldB   :: eq (Either () ()) Bool
   unfoldB :: eq Bool (Either () ())
+
+class Pi eq => PiNat eq where
 -- Encoding of natural numbers using the isorecursive type (mu x.1+x)
   foldN   :: eq (Either () Int) Int
   unfoldN :: eq Int (Either () Int)
+
+class PiNat eq => PiLNat eq where
 -- Encoding of lists of natural numbers using the isorecursive type
 -- (mu x.1+nat*x)
   foldLN :: eq (Either () (Int, [Int])) [Int]
@@ -50,7 +59,9 @@ class Pi eq where
 -- (mu x.1+nat*x)
   foldL :: eq (Either () (a, [a])) [a]
   unfoldL :: eq [a] (Either () (a, [a]))
--- Trace operators for looping/recursion
+
+class Pi eq => PiTracePlus eq where
+-- Trace operator for looping/recursion
   tracePlus :: eq (Either a b1) (Either a b2) -> eq b1 b2
 
 class Pi eq => PiN eq where
@@ -62,10 +73,25 @@ class Pi eq => PiN eq where
 class Pi eq => PiInv eq where
 -- etaTimes, epsTimes
   etaTimes :: eq () (Inv a,a)
-  epsTimes :: eq (a, Inv a) ()
+  epsTimes :: Eq a => eq (Inv a, a) ()
+
+-- This class says that, the u a's can be unified (as an effect) and generated
[…]

### 2012-06-21 code/modular/Model.hs

@@ -83,14 +83,22 @@ instance Pi (:<=>) where
   timesZeroR = TimesZeroR
   distribute = Distribute
   factor = Factor
+
+instance PiBool (:<=>) where
   foldB = FoldB
   unfoldB = UnfoldB
+
+instance PiNat (:<=>) where
   foldN = FoldN
   unfoldN = UnfoldN
+
+instance PiLNat (:<=>) where
   foldLN = FoldLN
   unfoldLN = UnfoldLN
   foldL = FoldL
   unfoldL = UnfoldL
+
+instance PiTracePlus (:<=>) where
   tracePlus = TracePlus
 
 instance PiN (:<=>) where
@@ -169,38 +177,38 @@ unList _ g (h:t) = g h t
 
 -- (simple) Semantics.  4 cases missing
 -- 0*a = 0, 0 = 0*a, 1 = 1/a*a, 1/a*a = 1
-instance Extract (:<=>) where
-  Id           @! v = v
-  (Adj f)      @! v = Negative ((adjoint f) @@ (unN v))
-  (f :.: g)    @! v = ((g @!) . (f @!)) v
-  (f :*: g)    @! v = (,) (f @! projL v) (g @! projR v) 
-  (f :+: g)    @! v = either (Left . (f @!)) (Right . (g @!)) v
-  PlusZeroL    @! v = unR v
-  PlusZeroR    @! v = Right v
-  CommutePlus  @! v = either Right Left v
-  AssocPlusL   @! v = either (Left . Left) (either (Left . Right) Right) v
-  AssocPlusR   @! v = either (either Left (Right . Left)) (Right . Right) v
-  TimesOneL    @! v = projR v
-  TimesOneR    @! v = ((), v)
-  CommuteTimes @! v = (projR v, projL v)
-  AssocTimesL  @! v = ((projL v, projL . projR $ v), projR . projR $ v)
-  AssocTimesR  @! v = (projL . projL $ v, (projR . projL $ v, projR v))
-  Distribute   @! v = let w = projR v in 
-                      either (\z -> Left (z, w)) (\z -> Right (z,w)) (projL v)
-  Factor       @! v = either (\z -> (Left $ projL z, projR z))
-                             (\z -> (Right $ projL z, projR z)) v
-  FoldB        @! v = either (\() -> True) (\() -> False) v
-  UnfoldB      @! v = if v then Left () else Right ()
-  FoldN        @! v = either (const 0) (1+) v
[…]

### 2012-06-21 code/modular/Model2.hs

@@ -6,8 +6,9 @@
 
 module Model2 where
 
-import Model hiding ((@@))
-import Lang (Neg,Inv,Unpack(..))
+import Model
+import Lang (Neg,Inv,Unpack(..),Unifyable(..),Extract(..),DS(..),
+    mr,mr2,mar,fresh1,fresh2,fresh3,orV)
 import Control.Monad
 import Unsafe.Coerce -- needed for polymorphic lookup. 
 
@@ -38,14 +39,15 @@ instance Unpack V where
     proj2 (Pair _ b) = b
     unNeg (Neg a) = a
 
+instance DS V where
+    unit = Unit
+    pair = pair
+    left = L
+    right = R
+    neg = Neg
+    inv = inv
 ---------------------------------------------------------------------------
 -- Unification
-
--- This class says that, the u a's can be unified (as an effect) and generated
-class Unifyable u where
-  type W :: * -> *
-  unify :: u a -> u a -> W ()
-  fresh :: W (u a)
 
 data Binding = forall a. Binding (Var a) (V a)
 type World = ([Binding], Int)
@@ -137,9 +139,6 @@ instance MonadPlus M where
   mzero = zeroM
   mplus = orM
 
-orV :: MonadPlus m => (v a -> m b) -> (v a -> m b) -> (v a -> m b)
-orV f g = \v -> f v `mplus` g v
-
 ----------------------------------------------------------------------
 -- Run the computation
 run :: M (V a) -> [V a]
@@ -147,44 +146,6 @@ run e = map reify (e `app` ([], 100))
 
 -- Note: I start with the 100 as the first system generated fresh
 -- variable thus reserving fresh variables 0..99 for user input. 
-
-----------
--- Helper fn
-fresh1 :: (Unifyable v) => (v a -> v b) -> v b -> W (v a)
-fresh1 f v =
-  do x <- fresh
-     unify (f x) v
-     return x
-
-fresh2 :: (Unifyable v) => (v a1 -> v a2 -> v b) -> v b -> W (v a1, v a2)
-fresh2 f v =
-  do x1 <- fresh
-     x2 <- fresh
-     unify (f x1 x2) v
[…]

### 2012-06-21 code/modular/ND.hs

@@ -1,0 +1,236 @@
+{-# OPTIONS_GHC -XTypeOperators  #-}
+{-# OPTIONS_GHC -XGADTs          #-}
+
+-- Fractional types introduce non-determinism; instead of having one value we
+-- have a sequence of possible values
+
+module ND where
+
+import Prelude hiding (id)
+import Control.Monad
+import Lang hiding (PiInv(..))
+
+-- Returns a list of all the elements of the type
+
+class Eq a => V a where
+  elems :: [a]
+
+instance V Zero where
+  elems = []
+
+instance V () where
+  elems = [()] 
+
+instance (V a, V b) => V (Either a b) where
+  elems = map Left elems ++ map Right elems
+
+instance (V a, V b) => V (a,b) where
+  elems = [(a,b) | a <- elems, b <- elems] 
+
+instance V a => V (Inv a) where
+  elems = map Inv elems
+
+data a :<=> b where 
+-- Congruence
+  Id    :: a :<=> a
+  (:.:) :: (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  ZeroE       :: Either Zero a :<=> a
+  ZeroI       :: a :<=> Either Zero a
+  CommutePlus :: Either a b :<=> Either b a
+  AssocPlusL  :: Either a (Either b c) :<=> Either (Either a b) c 
+  AssocPlusR  :: Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE        :: ((), a) :<=> a
+  UnitI        :: a :<=> ((), a)
+  CommuteTimes :: (a,b) :<=> (b,a) 
+  AssocTimesL  :: (a,(b,c)) :<=> ((a,b),c)
+  AssocTimesR  :: ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  TimesZeroL  :: (Zero, a) :<=> Zero
+  TimesZeroR  :: Zero :<=> (Zero, a)
+  Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
+  Factor      :: Either (b, a) (c, a) :<=> (Either b c, a)
+-- Multiplicative inverses (a cannot be 0) 
[…]

### 2012-06-21 code/modular/Test.hs

@@ -24,14 +24,14 @@ import Prelude hiding (id)
 -- *Pi> inot @@ True
 -- False
 -- *Pi> inot @@ False
-inot :: Pi eq => eq Bool Bool
+inot :: PiBool eq => eq Bool Bool
 inot = unfoldB %. commutePlus %. foldB
 
 -- Cond takes two isomorphisms from some type a to some type b, and 
 -- creates an isomorphism between a pair of (Bool, a) which will apply
 -- the first isomorphism if the Bool is True, and the second if the Bool
 -- is False.
-cond :: Pi eq => eq a b -> eq a b -> eq (Bool, a) (Bool, b)
+cond :: PiBool eq => eq a b -> eq a b -> eq (Bool, a) (Bool, b)
 cond f g = (unfoldB %* id) 
            %. distribute 
            %. ((id %* f) %+ (id %* g))
@@ -42,7 +42,7 @@ cond f g = (unfoldB %* id)
 -- creates an isomorphism (using cond) that will apply the isomorphism to
 -- the second value of the pair, if the first value if True, and apply
 -- Id otherwise.
-controlled :: Pi eq => eq a a -> eq (Bool, a) (Bool, a)
+controlled :: PiBool eq => eq a a -> eq (Bool, a) (Bool, a)
 controlled f = cond f id
 
 -- cnot is Controlled Not, as found in reversible computing papers such
@@ -56,7 +56,7 @@ controlled f = cond f id
 -- (False, True)
 -- *Pi> cnot @@ (True, False)
 -- (True, False)
-cnot :: Pi eq => eq (Bool, Bool) (Bool, Bool)
+cnot :: PiBool eq => eq (Bool, Bool) (Bool, Bool)
 cnot = controlled inot
 
 -- Toffoli is the universal nand/and gate presented in Reversible
@@ -72,18 +72,18 @@ cnot = controlled inot
 -- ((True, True), False)
 -- *Pi> toffoli @@ ((False, True), False)
 -- ((False, True), False)
-toffoli :: Pi eq => eq ((Bool,Bool),Bool) ((Bool,Bool),Bool)
+toffoli :: PiBool eq => eq ((Bool,Bool),Bool) ((Bool,Bool),Bool)
 toffoli = assocTimesR %. controlled cnot %. assocTimesL
 
 -- The Fredkin gate is a well known universal gate.
[…]

### 2012-06-21 rational_ext.tex
_

@@
 \label{def:size}
-The size of a type {{b}} is a number defined to be:
+The size of a type {{b}}, denoted by {{[^ b ^]}}, is a numeric value
+and is defined to be:
 \vspace{-20pt}
@@
     equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
-    {{h:b3<->b4}}). One can check:
+    {{h:b3<->b4}}).
+
+    Assuming {{v1:b1}}, {{v2:b2}}, {{v3:b3}}, {{v4:b4}},
+    {{f~v1|-->v2}}, {{g v2 |--> v3}} and {{h~v3 |--> v4}}, one can
+    check:
 
@@
     \begin{enumerate}
-    \item The braiding must satisfy two ``hexagon axioms''.
+    \item The braiding must satisfy two ``hexagon'' axioms.
     \item The braiding is self inverse,
@@
 
-\noindent
-No division by zero.
+\paragraph*{No division by zero.} 
+The important side-condition on the types not captured by the syntax
+above is that there are no types of the form {{1/b}}, where
+{{[^b^]=0}} (see Defn. \ref{def:size}). This eliminates syntaxes of
+the form {{1/0}}, {{1/(0+0)}}, {{1/(0*b)}} etc, from being treated as
+valid {{b}} types. Intuitively, this translates into the familiar
+restriction that there is no division by 0 in arithmetic or the
+definition of fields in abstract algebra.
 
@@
 
-%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Cloning}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Construction of the Solver}

### 2012-06-22 rational_ext.tex
_

@@
 
+\begin{proposition}[Size 0 is uninhabited]. 
+\label{prop:no-zero-values}
+There are no values {{v:b}} when {{[^b^]=0}}.
+\end{proposition}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The syntactic notation above is often obscure and hard to read.  Following
-the tradition established for monoidal
-categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present a
-graphical language that conveys the intuitive semantics of the language
-(which is formalized in the next section).
+The syntactic notation above is often obscure and hard to read.
+Following the tradition established for monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present
+a graphical language that conveys the intuitive semantics of the
+language (which is formalized in the next section).
 
@@
 
-\begin{proposition}[Coherence]
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Categorical Structure}
+
+As with {{Pi}}, {{PiEE*}} is a dagger symmetric monoidal category with
+two monoidal structures. In this section we establish that it is
+``locally compact closed'' in the sense that for non-zero objects,
+namely types where {{[^ b ^]=0}}, {{PiEE*}} satisfies the conditions
+for compact closed categories.
+
+\begin{proposition}[Locally Compact Closed]
+{{PiEE*}} is a locally compact closed category in the sense that for
+all non-zero objects {{b}}, we have an inverse object {{1/b}} in the
+symmetric monoidal structure of {{(*, 1)}}, with the morphisms:
+
+\begin{enumerate}
+\item {{1 <-> 1/b * b}} given by {{eta*}}, called the `unit map'.
+\item {{b * 1/b <-> 1}} given by {{swap* (;) eps*}}, called the `counit map'.
+\end{enumerate}
+
+The following coherence conditions hold (see 2.1 in
+\cite{selinger2007dagger}):
+
+\begin{enumerate}
+\item {{uniti (;) (eta* (;) swap(*)id) (;)(swap* (*) id) (;) assocr* (;) (id (*) eps*) (;) swap* (;) unite = id}}
+\end{enumerate}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Categorical Constructions}

### 2012-06-22 code/dual/OldDual.hs

@@ -1,77 +1,96 @@
-{-# OPTIONS_GHC -XGADTs -XFlexibleContexts -XTypeOperators -XExistentialQuantification #-}
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1, 7.0.3
 
--- {-# OPTIONS_GHC -XFlexibleContexts -XEmptyDataDecls #-} 
--- 
+-- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
+-- Dont know: This code tested with GHC version 6.12.3, and version 7.0.1
 
-module Dual where
+module OldDual where
 
-import Unsafe.Coerce -- needed for polymorphic lookup (lookupW)
+import Data.Typeable
+import Unsafe.Coerce -- needed for polymorphic lookup. 
 
-------------------------------------------------------------------------------
+
+-----------------------------------------------------------------------
 -- Base types 
 
 data Zero 
-
 data Neg a = Ng a
-
 data Inv a = In a
 
+type (:+:) a b = Either a b
+type (:*:) a b = (a, b)
 
-------------------------------------------------------------------------------
+data Three = One 
+           | Two 
+           | Three
+             deriving (Eq, Show)
+
+-----------------------------------------------------------------------
 -- Values 
 -- 
 -- Values of the form 'V a' are values used by the interpreter. 
+
+
 -- Values can contain non-grounded fresh variables
-
 type Var a = Int
-
 data V a where
-  Unit :: V () 
-  Pair :: V a -> V b -> V (a, b) 
-  L :: V a -> V (Either a b) 
-  R :: V b -> V (Either a b) 
-  Neg :: V a -> V (Neg a)
-  Inv :: V a -> V (Inv a)
-  Fresh :: Var a -> V a
+    Unit :: V () 
+    Pair :: V a -> V b -> V (a, b) 
+    L :: V a -> V (Either a b) 
+    R :: V b -> V (Either a b) 
+    Neg :: V a -> V (Neg a)
+    Inv :: V a -> V (Inv a)
+    Lift :: (Eq a, Show a) => a -> V a
+    Fresh :: Var a -> V a
+
 
 instance Show (V a) where 
-  show Unit = "()"
-  show (Pair a b) = "(" ++ show a ++ "," ++ show b ++ ")"
-  show (L a) = "L " ++ (show a) 
-  show (R a) = "R " ++ (show a)
[…]

### 2012-06-22 code/dual/Test.hs

@@ -1,6 +1,6 @@
 {-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts -XScopedTypeVariables #-} -- 7.0.1, 7.0.3
 
-import Dual 
+import OldDual 
 
 ------------------------------------------------------------------------------------
 ------------------------------------------------------------------------------------
@@ -18,9 +18,18 @@ traceTimes c = UnitI
 yankTimes :: a :<=> a
 yankTimes = traceTimes CommuteTimes
 
+-- z_shapeTimes :: a :<=> a
+-- z_shapeTimes = UnitI
+--                :.: (EtaTimes:*: Id) 
+--                :.: (CommuteTimes :*: Id)
+--                :.: AssocTimesR 
+--                :.: (Id :*: EpsTimes)
+--                :.: CommuteTimes
+--                :.: UnitE
+               
 z_shapeTimes :: a :<=> a
 z_shapeTimes = UnitI
-               :.: (EtaTimes:*: Id) 
+               :.: (EtaTimes :*: Id) 
                :.: (CommuteTimes :*: Id)
                :.: AssocTimesR 
                :.: (Id :*: EpsTimes)

### 2012-06-22 rational_ext.tex
changes to the test code and first few sections of the paper

@@
 
-Compact Closed Structure (modulo 0).
-
-Show that ``no division by zero'' is supported by the semantics.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+Show that ``no division by zero'' is supported by the semantics.
+
 \begin{proposition}[Type Safety]
@@
 Note that this language is not logically reversible.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Categorical Structure}
+
+As with {{Pi}}, {{PiEE*}} is a dagger symmetric monoidal category with
+two monoidal structures. In this section we establish that it is
+``locally compact closed'' in the sense that for non-zero objects,
+namely types where {{[^ b ^]=0}}, {{PiEE*}} satisfies the conditions
+for compact closed categories.
+
+\begin{proposition}[Locally Compact Closed]
+{{PiEE*}} is a locally compact closed category in the sense that for
+all non-zero objects {{b}}, we have an inverse object {{1/b}} in the
+symmetric monoidal structure of {{(*, 1)}}, with the morphisms:
+
+\begin{enumerate}
+\item {{1 <-> 1/b * b}} given by {{eta*}}, called the `unit map'.
+\item {{b * 1/b <-> 1}} given by {{swap* (;) eps*}}, called the `counit map'.
+\end{enumerate}
+
+\noindent
+such that the following coherence conditions hold (see 2.1 in
+\cite{selinger2007dagger}):
+
+\begin{enumerate}
+\item {{uniti (;) swap* (;) (id (*) eta*) (;) assocl* (;) ((swap* (;) eps*) (*) id) (;) unite = id : b <-> b}}
+\item {{uniti (;) (eta* (*) id) (;) assocr* (;) (id (*) (swap* (;) eps*)) (;) swap* (;) unite = id : 1/b <-> 1/b}}
+
+%% z_shape1Times :: a :<=> a
+%% z_shape1Times = UnitI
+%%                 :.: CommuteTimes
+%%                 :.: (Id :*: EtaTimes) 
+%%                 :.: AssocTimesL
+%%                 :.: ((CommuteTimes :.: EpsTimes) :*: Id)
+%%                 :.: UnitE
+
+%% z_shape2Times :: Inv a :<=> Inv a
+%% z_shape2Times = UnitI
+%%                 :.: (EtaTimes :*: Id)
+%%                 :.: AssocTimesR
+%%                 :.: (Id :*: (CommuteTimes :.: EpsTimes))
+%%                 :.: CommuteTimes
+%%                 :.: UnitE
+
+\end{enumerate}
+
+\end{proposition}
+
+TODO: once the semantics are written, write out the proof of one of
+the coherence conditions in full.
 
@@
 essentially the constructions in the introduction.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Categorical Structure}
-
-As with {{Pi}}, {{PiEE*}} is a dagger symmetric monoidal category with
-two monoidal structures. In this section we establish that it is
-``locally compact closed'' in the sense that for non-zero objects,
-namely types where {{[^ b ^]=0}}, {{PiEE*}} satisfies the conditions
-for compact closed categories.
-
-\begin{proposition}[Locally Compact Closed]
-{{PiEE*}} is a locally compact closed category in the sense that for
-all non-zero objects {{b}}, we have an inverse object {{1/b}} in the
-symmetric monoidal structure of {{(*, 1)}}, with the morphisms:
-
-\begin{enumerate}
-\item {{1 <-> 1/b * b}} given by {{eta*}}, called the `unit map'.
-\item {{b * 1/b <-> 1}} given by {{swap* (;) eps*}}, called the `counit map'.
-\end{enumerate}
-
-The following coherence conditions hold (see 2.1 in
-\cite{selinger2007dagger}):
-
-\begin{enumerate}
-\item {{uniti (;) (eta* (;) swap(*)id) (;)(swap* (*) id) (;) assocr* (;) (id (*) eps*) (;) swap* (;) unite = id}}
-\end{enumerate}
-
-\end{proposition}
 

### 2012-06-22 code/dual/Test.hs

@@ -27,14 +27,21 @@ yankTimes = traceTimes CommuteTimes
 --                :.: CommuteTimes
 --                :.: UnitE
                
-z_shapeTimes :: a :<=> a
-z_shapeTimes = UnitI
-               :.: (EtaTimes :*: Id) 
-               :.: (CommuteTimes :*: Id)
-               :.: AssocTimesR 
-               :.: (Id :*: EpsTimes)
-               :.: CommuteTimes
-               :.: UnitE
+z_shape1Times :: a :<=> a
+z_shape1Times = UnitI
+                :.: CommuteTimes
+                :.: (Id :*: EtaTimes) 
+                :.: AssocTimesL
+                :.: ((CommuteTimes :.: EpsTimes) :*: Id)
+                :.: UnitE
+               
+z_shape2Times :: Inv a :<=> Inv a
+z_shape2Times = UnitI
+                :.: (EtaTimes :*: Id)
+                :.: AssocTimesR
+                :.: (Id :*: (CommuteTimes :.: EpsTimes))
+                :.: CommuteTimes
+                :.: UnitE
                
 
 ----------------------------------------------------------------
@@ -49,7 +56,7 @@ tracePlus c = ZeroI
           :.: ZeroE
 
 yankPlus :: b :<=> b
-yankPlus = trace CommutePlus
+yankPlus = tracePlus CommutePlus
 
 -- Teleportation (??)
 -- Also the coherence condition for eta/eps for compact closed categories. 
@@ -76,7 +83,7 @@ z_shapePlus = ZeroI
 -- *PiCont> eval yank_not False
 -- True
 yank_not :: Bool :<=> Bool
-yank_not = trace (CommutePlus :.: (inot :+: Id))
+yank_not = tracePlus (CommutePlus :.: (inot :+: Id))
 
 ----------------------------------------------------------------
 -- eta and eps over products
@@ -230,7 +237,7 @@ fredkin = controlled CommuteTimes
 -- (-1 + 1) + 1 
 -- 1 
 inf_stream :: () :<=> ()
-inf_stream = trace (FoldB :.: inner :.: UnfoldB)
+inf_stream = tracePlus (FoldB :.: inner :.: UnfoldB)
     where 
       inner :: Bool :<=> Bool 
       inner = 
@@ -243,7 +250,7 @@ inf_stream = trace (FoldB :.: inner :.: UnfoldB)
           :.: UnitE
 
[…]

### 2012-06-23 rational_ext.tex
frac semantics

@@
 
-Provide a semantics with the non-determinism monad and define an
-\emph{eval}.
-
-Show that ``no division by zero'' is supported by the semantics.
+The operational semantics for {{PiEE*}} involve the addition of the
+following two rules to the small-step operational semantics of {{Pi}}.
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <eta*, (), C> &|-->& forall v:b.{[eta*, (1/v, v), C]} & (eta*)
+% <eps*, (1/v,v), C> &|-->& {[eps*, (), C]} & (eps*)
+
+The rule {{(eta*)}} is different from other rules encountered before
+due to the {{forall v:b}} quaitification and is to be understood as
+follows: when the machine encounters an {{eta*}} reduction it
+non-deterministically replicates into several machine states, each of
+which is seeded by a unique value {{v}} that inhabits the type {{b}}
+that indexes {{eta* : 1 <-> 1/b * b}}. As an example at the type
+{{bool}}, we would have two possible machine states:
+
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <eta*, (), C> &|-->& {[eta*, (1/true, true), C]}
+% and  &|-->& {[eta*, (1/false, false), C]}
+
+The rule {{(eps*)}} is straightforward : when given a tuple
+{{(1/v,v):1/b * b}} where {{eta*:1/b*b <-> 1}} the result is
+{{()}}. The machine is undefined for cases where where the input is
+{{(1/v',v):1/b*b}} and {{v' /= v}}.
 

### 2012-06-23 rational_ext.tex
_

@@
 
+\paragraph*{Annihilation.} 
+A consequence of the semantics is that they admit computations that
+are undefined. For instance consider, where {{not : bool <-> bool}}.
+
+{{eta* (;) (id (*) not) (;) eps* : 1 <-> 1}}
+
+Tracing this circuit gives us:
+
+%subcode{opsem} include main
+%! columnStyle = rllr
+% & & <eta* (;) (id (*) not) (;) eps*, (), []>
+% & |-->* & <eta*, (), Fst [] (id (*) not) (;) eps*>
+% (1) & |--> & [eta*, (1/true, true), Fst [] (id (*) not) (;) eps*] 
+% & |-->* & <not, true, RightT id 1/true (Fst (Snd~eta* []) ~eps*)> 
+% & |-->* & [not, false, RightT id 1/true (Fst (Snd~eta* []) ~eps*)] 
+% & |-->* & [eps*, (1/true, false), Snd (id (*) not) (Snd~eta* [])]
+% & |--> & undefined
+% (2) & |--> & [eta*, (1/false, false), Fst [] (id (*) not) (;) eps*] 
+% & |-->* & <not, false, RightT id 1/false (Fst (Snd~eta* []) ~eps*)> 
+% & |-->* & [not, true, RightT id 1/false (Fst (Snd~eta* []) ~eps*)] 
+% & |-->* & [eps*, (1/false, true), Snd (id (*) not) (Snd~eta* [])]
+% & |--> & undefined
+
 \begin{proposition}[Type Safety]
@@
 \end{enumerate}
+
+\begin{proof}
+  We briefly summarize the execution of the first combinator (let us
+  call it {{c}}) combinator to show that the coherence conditions
+  hold.
+\item
+
+%subcode{opsem} include main
+%! columnStyle = rllr
+% & & <c, v:b, []>
+% & |-->* & <eta*, (), C1>
+% & |-->  & <eta*, (1/v',v'), C2> & some v':b
+% & |-->* & <eps*, (1/v', v), C3> & when v' = v
+% (1) & |-->~ & [eps*, (), C3]
+% & |-->* & [c', v, [] ]
+% (2) & |-->* & <eps*, (1/v', v), C3> & when v' /= v
+% & |-->~ & undefined
+
+Thus for {{v:b}}, the evaluation of {{<c, v, []>}} always produces {{v:b}} as
+output, i.e. {{c = id}}.
+
+\end{proof}
 

### 2012-06-23 rational_ext.tex
_

@@
 
-\begin{proposition}[Size 0 is uninhabited]. 
-\label{prop:no-zero-values}
-There are no values {{v:b}} when {{[^b^]=0}}.
-\end{proposition}
+%% \begin{proposition}[Size 0 is uninhabited]. 
+%% \label{prop:no-zero-values}
+%% There are no values {{v:b}} when {{[^b^]=0}}.
+%% \end{proposition}
 
@@
 above is that there are no types of the form {{1/b}}, where
-{{[^b^]=0}} (see Defn. \ref{def:size}). This eliminates syntaxes of
-the form {{1/0}}, {{1/(0+0)}}, {{1/(0*b)}} etc, from being treated as
-valid {{b}} types. Intuitively, this translates into the familiar
-restriction that there is no division by 0 in arithmetic or the
-definition of fields in abstract algebra.
+{{[^b^]=0}} (see Defn. \ref{def:size} and \ref{def:size-frac}
+below). This eliminates syntaxes of the form {{1/0}}, {{1/(0+0)}},
+{{1/(0*b)}} etc, from being treated as valid {{b}} types. Intuitively,
+this translates into the familiar restriction that there is no
+division by 0 in arithmetic or the definition of fields in abstract
+algebra.
 
@@
 
+\begin{definition}[Size of a type.] 
+\label{def:size-frac}
+The definition of size, {{[^ b ^]}}, is extended to include
+fractionals as {{[^1 / b ^] = 1/[^ b ^]}}.
+\end{definition}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The rule {{(eta*)}} is different from other rules encountered before
-due to the {{forall v:b}} quaitification and is to be understood as
+due to the {{forall v:b}} quantification and is to be understood as
 follows: when the machine encounters an {{eta*}} reduction it
@@
 
-Tracing this circuit gives us:
+\noindent
+Tracing the execution of this circuit gives us:
 
@@
 
-Note that this language is not logically reversible.
+% Note that this language is not logically reversible.
 
@@
 \end{enumerate}
-
+\end{proposition}
 \begin{proof}
@@
   call it {{c}}) combinator to show that the coherence conditions
-  hold.
-\item
+  hold. The second combinator simply mirrors {{c}}. To show that
+  {{c=id}}, we must show for {{v:b}}, the evaluation of {{<c, v, []>}}
+  always produces {{v:b}} as output.
 
@@
 % & |-->  & <eta*, (1/v',v'), C2> & some v':b
-% & |-->* & <eps*, (1/v', v), C3> & when v' = v
-% (1) & |-->~ & [eps*, (), C3]
+% (1) & |-->* & <eps*, (1/v', v), C3> & when v' = v
+% & |-->~ & [eps*, (), C3]
 % & |-->* & [c', v, [] ]
@@
 
-Thus for {{v:b}}, the evaluation of {{<c, v, []>}} always produces {{v:b}} as
-output, i.e. {{c = id}}.
-
 \end{proof}
-
-\end{proposition}
-
-TODO: once the semantics are written, write out the proof of one of
-the coherence conditions in full.
 

### 2012-06-23 rational_ext.tex
_

@@
 
-The type safety of {{Pi}} follows directly from the corresondence
+The type safety of {{Pi}} follows directly from the correspondence
 between inductive definitions of the types and the big-step
@@
 
-We present the categorical structure with minimum commentry. More
+We present the categorical structure with minimum commentary. More
 details may be found in excellent references such as Barr and Wells
@@
   show how this has been addressed before in the categorical
-  setting. In the absense of any prior knowledge of category theory
+  setting. In the absence of any prior knowledge of category theory
   however, our wiring diagrams may be read as the ``flow of types'' in
@@
 and co-products (say through the addition of information effects) it
-is concievable that part of its structure collapses. This sort of
+is conceivable that part of its structure collapses. This sort of
 collapse, while catastrophic for algebraic structures (effectively
@@
 in computing we are interested in the specific operational nature (the
-computational content, so to speak) of the morphisms. Anectodal evidence
+computational content, so to speak) of the morphisms. Anecdotal evidence
 follows from the fact several real-world programming language have
@@
 above is that there are no types of the form {{1/b}}, where
-{{[^b^]=0}} (see Defn. \ref{def:size} and \ref{def:size-frac}
-below). This eliminates syntaxes of the form {{1/0}}, {{1/(0+0)}},
+{{[^b^]=0}} (see Def. \ref{def:size} and \ref{def:size-frac}
+below). This eliminates terms of the form {{1/0}}, {{1/(0+0)}},
 {{1/(0*b)}} etc, from being treated as valid {{b}} types. Intuitively,

### 2012-06-24 code/dual/ND.hs
?

@@ -2,7 +2,8 @@
 {-# OPTIONS_GHC -XGADTs          #-}
 {-# OPTIONS_GHC -XEmptyDataDecls #-} 
 
--- eta+/epsilon+ undefined in this model
+-- Fractional types introduce non-determinism; instead of having one value we
+-- have a sequence of possible values
 
 module ND where
 
@@ -17,13 +18,14 @@ data Zero -- empty type
 
 data Inv a = Inv a deriving Eq -- fractional types (a should not be zero) 
 
-data Neg a = Neg a deriving Eq -- negative types
-
 instance Show Zero where 
   show _ = error "Empty type has no values to show"
 
 instance Eq Zero where 
   _ == _ = error "Cannot use == at type 0"
+
+instance Show a => Show (Inv a) where
+  show (Inv a) = "1/" ++ show a
 
 -- Returns a list of all the elements of the type
 
@@ -41,9 +43,6 @@ instance (V a, V b) => V (Either a b) where
 
 instance (V a, V b) => V (a,b) where
   elems = [(a,b) | a <- elems, b <- elems] 
-
-instance V a => V (Neg a) where
-  elems = map Neg elems
 
 instance V a => V (Inv a) where
   elems = map Inv elems
@@ -80,9 +79,6 @@ data a :<=> b where
                    (Either b c, a) :<=> Either (b, a) (c, a)
   Factor      :: (V a, V b, V c) => 
                    Either (b, a) (c, a) :<=> (Either b c, a)
--- Additive inverses
-  EtaPlus :: V a => Zero :<=> (Either (Neg a) a)
-  EpsPlus :: V a => (Either (Neg a) a) :<=> Zero
 -- Multiplicative inverses (a cannot be 0) 
   EtaTimes :: V a => () :<=> (Inv a, a)
   EpsTimes :: V a => (Inv a, a) :<=> ()
@@ -106,8 +102,6 @@ adjoint TimesZeroL = TimesZeroR
 adjoint TimesZeroR = TimesZeroL
 adjoint Distribute = Factor
 adjoint Factor = Distribute
-adjoint EtaPlus = EpsPlus
-adjoint EpsPlus = EtaPlus
 adjoint EtaTimes = EpsTimes
 adjoint EpsTimes = EtaTimes
 
@@ -206,18 +200,14 @@ eval_iso1 Factor v =
     Left (b,a) -> return (Left b, a) 
     Right (c,a) -> return (Right c, a) 
 
[…]

### 2012-06-24 rational_ext.tex
_

@@
 \newcommand{\arrow}[1]{\mathtt{#1}}
+
+\newcommand{\dgm}[2][0.95]{
+\begin{center}
+\scalebox{#1}{
+\includegraphics{diagrams/#2.pdf}
+}
+\end{center}
+}
 
@@
 \begin{proof}
-  We briefly summarize the execution of the first combinator (let us
-  call it {{c}}) combinator to show that the coherence conditions
-  hold. The second combinator simply mirrors {{c}}. To show that
-  {{c=id}}, we must show for {{v:b}}, the evaluation of {{<c, v, []>}}
-  always produces {{v:b}} as output.
+  For convenience let us draw {{swap* (;) eps*}} shown on the left
+  instead as shown on the right.
+
+\begin{multicols}{2}
+\dgm[0.8]{swap_eps_times}
+
+\dgm[0.8]{cat_eps_times}    
+\end{multicols}
+
+The two coherence conditions are more intuitive when presented
+graphically. They require that the combinators below be equal to
+{{id}}, at the appropriate type (see Sec. 4
+\cite{springerlink:10.1007/978-3-642-12821-94}).
+
+\begin{multicols}{2}
+\dgm{z_times1}
+
+\dgm{z_times2}
+\end{multicols}
+
+  We briefly trace the execution of the first combinator (let us call
+  it {{c}}. To show that {{c=id}}, we must show for {{v:b}}, the
+  evaluation of {{<c,v,[]>}} always produces {{v:b}} as
+  output. Verification for the second combinator is similar.
 
@@
 
-For a monoidal category to be compact closed the maps~{{eta}}
-and~{{eps}} must satisfy a coherence condition that is usually
-visualized as follows:
-\begin{center}
-  \includegraphics{diagrams/coherence.pdf}
-\end{center}
-where $b^*$ represents the dual of {{b}}.  In the case of negatives,
-the condition amounts to checking that reversing direction twice is a
-no-op. In the case of fractionals, the condition amounts to checking
-that creating values {{alpha}} and {{1/alpha}} and immediately
-unifying them is also a no-op. Both checks are straightforward and are
-essentially the constructions in the introduction.
+It is important to note that in a compact closed category, the
+dualizing action, usually denoted $(-)^\asterix$, is a contravariant
+functor. However {{PiEE*}} is not compact closed and {{1/(-)}} is not
+a functor specifically because divsion is not defined on 0 sized
+types.
+
+Without the important side condition on division by zero, we would
+have {{1<->1/0*0}} which implies {{1 <-> 0}} and consequently
+collapses the entire structure. Structures that are ``zero totalized
+fields'' such as meadows \cite{Bergstra:2009:MES:1507774.1508000}
+deserve investigation as alternative models for {{Pi}} with
+fractionals (and negatives).
+
+%% For a monoidal category to be compact closed the maps~{{eta}}
+%% and~{{eps}} must satisfy a coherence condition that is usually
+%% visualized as follows:
+%% \begin{center}
+%%   \includegraphics{diagrams/coherence.pdf}
+%% \end{center}
+%% where $b^*$ represents the dual of {{b}}.  In the case of negatives,
+%% the condition amounts to checking that reversing direction twice is a
+%% no-op. In the case of fractionals, the condition amounts to checking
+%% that creating values {{alpha}} and {{1/alpha}} and immediately
+%% unifying them is also a no-op. Both checks are straightforward and are
+%% essentially the constructions in the introduction.
 

### 2012-06-24 rational_ext.tex
_

@@
 
+%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions}
+
+\paragraph*{Booleans and Conditionals.} 
+Given any combinator {{c : b <-> b}} we can construct a combinator called
+{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
+one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
+the combinator {{c}} is used to transform the value of type~{{b}}. If the
+boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
+write down the combinator for {{if_c}} in terms of {{c}} as 
+{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
+
+\noindent The diagram below shows the input value of type {{(1+1)*b}}
+processed by the distribute operator {{dist}}, which converts it into a value
+of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
+case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
+combinator~{{c}} is applied to the value of type~{{b}}. The right branch
+which corresponds to the boolean being {{false}} passes along the value of
+type {{b}} unchanged.
+
+\begin{center}
+\scalebox{1.0}{
+%%subcode-line{pdfimage}[diagrams/if_c.pdf]
+\includegraphics{diagrams/thesis/cnot.pdf}
+}
+\end{center}
+
+% We will be seeing many more such wiring diagrams in this paper and it is
+% useful to note some conventions about them. Wires indicate a
+% value that can exist in the program. Each wire, whenever possible, is
+% annotated with its type and sometimes additional information to help clarify
+% its role. When multiple wires run in parallel, it means that those values
+% exist in the system at the same time, indicating pair types. When there is a
+% disjunction, we put a {{+}} between the wires. 
+% Combinators for distribution {{dist}} and factoring {{factor}}
+% are represented as triangles with their operator symbols in them. Other
+% triangles may be used and, in each case, types or labels will be used to
+% clarify their roles. Finally, we don't draw boxes for combinators such as
+% {{id}}, commutativity, and associativity, but instead just shuffle the wires
+% as appropriate.
+
+The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
+negates its second argument if the first argument is {{true}}. This
+gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
+equivalent construction that is useful is {{else_{not} }} where we
+negate the second argument only if the first is {{false}}.
+
+Similarly, we can iterate the construction of {{if_c}} to check several
+bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
+checks two booleans and negates the result wire only if they are both
+{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
+universal reversible gate. We can generalize this construction to
+{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
+they are all {{true}}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\paragraph*{Cloning.}
+Although cloning is generally not allowed in reversible languages, it is
+possible at the cost of having additional constant inputs. For example,
+consider the gate {{else_{not} }}. Generally, the gate maps
+{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
+on the cases in which the second input is {{true}}, we get that the gate maps
+{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
+i.e., the gate clones the first input. A circuit of {{n}} parallel
+{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
+{{true}} inputs in the process.  Let us call this construction
+{{clone^n_{bool} }}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-  {{c v |--> v'}} iff {{<c, v, []> |-->* [c, v', [] ]}}
-
+  {{c v |--> v'}} iff {{<c, v, []> |-->* [c', v', [] ]}}
 \end{proposition}
+\begin{proof}
+  To prove the above, we first show that a more general lemma holds,
+  namely that: {{c v |--> v'}} iff {{<c, v, C> |-->* [c', v', C ]}}.
+
+  To show the left-to-right direction we proceed by induction on the
+  derivation of {{c v |--> v'}}. In the case of primitive isomorphisms
+  the condition holds trivially. In the case of composition, we work
+  out the case of {{c1+c2}} as an example. Given {{c1 +c2 v |--> v'}}
+  we have to show that there is a small step derivation sequence that
+  matches it. Here {{v:b1+b2}} can be of the form {{left v1}} or
+  {{right v2}}. Assuming {{left v1}}, we have:
+
+%subcode{proof} include main
+%@ c1 v1 |--> v1' ==> <c1, v1, C'> |-->* {[c1',v1',C']}
+%@@ c1+c2 (left v1) |--> left v1' ==> ?
+
+Choosing {{C' =Lft c2 C}}, we have the required derivation sequence: 
+{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. The proof follows similarly for the {{right v2}} case.
+
+To show the right-to-left direction we proceed by induction of the
+sequence of {{|-->*}} derivations. Again the case for primtive
+isomorphisms follows trivially. Taking the case of {{c1+c2}} we are
+required to show that given a sequence 
+{{<c1+c2,v,C> |-->* {[c', v', C]} }} there is a derviation tree for
+{{c1+c2 v |--> v'}}. In the case that {{v:v1+b2}} has the form
+{{left~v1}}, this follows by observing that the given sequence must have the form
+{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. 
+For the strictly smaller inner subsequence by induction we have
+{{c1~v1 |--> v1'}} which lets us complete the derivation of
+{{c1+c2~(left v1) |--> left v1'}}. The {{right v2}} follows similarly.
+
+\end{proof}
 
@@
 \begin{proof}
-  For convenience let us draw {{swap* (;) eps*}} shown on the left
-  instead as shown on the right.
+For convenience, we draw the combinator {{swap* (;) eps*}} (shown on
+the left) as shown on the right.
 
@@
 
-The two coherence conditions are more intuitive when presented
+\noindent
+The coherence conditions are more intuitive when presented
 graphically. They require that the combinators below be equal to
@@
 paper~\cite{infeffects}.
-
-%%%%%%%%%%%%%%%%%%%
-\subsection{Booleans and Conditionals}
-
-Given any combinator {{c : b <-> b}} we can construct a combinator called
-{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
-one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
-the combinator {{c}} is used to transform the value of type~{{b}}. If the
-boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
-write down the combinator for {{if_c}} in terms of {{c}} as 
-{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
-
-\noindent The diagram below shows the input value of type {{(1+1)*b}}
-processed by the distribute operator {{dist}}, which converts it into a value
-of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
-case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
-combinator~{{c}} is applied to the value of type~{{b}}. The right branch
-which corresponds to the boolean being {{false}} passes along the value of
-type {{b}} unchanged.
-
-\begin{center}
-\scalebox{1.0}{
-%%subcode-line{pdfimage}[diagrams/if_c.pdf]
-\includegraphics{diagrams/thesis/cnot.pdf}
-}
-\end{center}
-
-% We will be seeing many more such wiring diagrams in this paper and it is
-% useful to note some conventions about them. Wires indicate a
-% value that can exist in the program. Each wire, whenever possible, is
-% annotated with its type and sometimes additional information to help clarify
-% its role. When multiple wires run in parallel, it means that those values
-% exist in the system at the same time, indicating pair types. When there is a
-% disjunction, we put a {{+}} between the wires. 
-% Combinators for distribution {{dist}} and factoring {{factor}}
-% are represented as triangles with their operator symbols in them. Other
-% triangles may be used and, in each case, types or labels will be used to
-% clarify their roles. Finally, we don't draw boxes for combinators such as
-% {{id}}, commutativity, and associativity, but instead just shuffle the wires
-% as appropriate.
-
-The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
-negates its second argument if the first argument is {{true}}. This
-gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
-equivalent construction that is useful is {{else_{not} }} where we
-negate the second argument only if the first is {{false}}.
-
-Similarly, we can iterate the construction of {{if_c}} to check several
-bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
-checks two booleans and negates the result wire only if they are both
-{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
-universal reversible gate. We can generalize this construction to
-{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
-they are all {{true}}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Cloning}
-
-Although cloning is generally not allowed in reversible languages, it is
-possible at the cost of having additional constant inputs. For example,
-consider the gate {{else_{not} }}. Generally, the gate maps
-{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
-on the cases in which the second input is {{true}}, we get that the gate maps
-{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
-i.e., the gate clones the first input. A circuit of {{n}} parallel
-{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
-{{true}} inputs in the process.  Let us call this construction
-{{clone^n_{bool} }}.
 

### 2012-06-24 p.tex
p.tex

--- 2012-02-10 p.tex
@@
 
+\usepackage{graphicx}
+\usepackage{comment}
+\usepackage{amsmath}
+\usepackage{xspace}
 \usepackage{amssymb}
-\usepackage{graphicx}
+\usepackage{stmaryrd}
 \usepackage{proof}
-\usepackage{stmaryrd}
-\usepackage{url}
-\usepackage{xspace}
 \usepackage{multicol}
+\usepackage[nodayofweek]{datetime}
+\usepackage{etex}
+\usepackage[all, cmtip]{xy}
+\usepackage{xcolor}
+\usepackage{listings}
+\usepackage{multicol}
+\newcommand\hmmax{0} % default \newcommand\bmmax{0} % default 4
+\usepackage{bm}
+\usepackage{cmll}
+
+\newcommand{\fname}[1]{\ulcorner #1 \urcorner}
+\newcommand{\fconame}[1]{\llcorner #1 \lrcorner}
+
+\newcommand{\xcomment}[2]{\textbf{#1:~\textsl{#2}}}
+\newcommand{\amr}[1]{\xcomment{Amr}{#1}}
+\newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}
+
+\newcommand{\asterix}[0]{*}
+
+\newcommand{\ie}{\textit{i.e.}\xspace}
+\newcommand{\eg}{\textit{e.g.}\xspace}
+
+\newcommand{\lcal}{\ensuremath{\lambda}-calculus\xspace}
+\newcommand{\G}{\ensuremath{\mathcal{G}}\xspace}
+
+\newcommand{\code}[1]{\lstinline[basicstyle=\small]{#1}\xspace}
+\newcommand{\name}[1]{\code{#1}}
+
+\def\newblock{}
+
+\newenvironment{floatrule}
+    {\hrule width \hsize height .33pt \vspace{.5pc}}
+    {\par\addvspace{.5pc}}
 
@@
 
+\newcommand{\dgm}[2][0.95]{
+\begin{center}
+\scalebox{#1}{
+\includegraphics{diagrams/#2.pdf}
+}
+\end{center}
+}
+
+%subcode-inline{bnf-inline} name langRev
+%! swap+ = \mathit{swap}^+
+%! swap* = \mathit{swap}^*
+%! dagger =  ^{\dagger}
+%! assocl+ = \mathit{assocl}^+
+%! assocr+ = \mathit{assocr}^+
+%! assocl* = \mathit{assocl}^*
+%! assocr* = \mathit{assocr}^*
+%! identr* = \mathit{uniti}
+%! identl* = \mathit{unite}
+%! dist = \mathit{distrib}
+%! factor = \mathit{factor}
+%! eta = \eta
+%! eps = \epsilon
+%! eta+ = \eta^+
+%! eps+ = \epsilon^+
+%! eta* = \eta^{\times}
+%! eps* = \epsilon^{\times}
+%! trace+ = trace^+
+%! trace* = trace^{\times}
+%! ^^^ = ^{-1}
+%! (o) = \circ
+%! (;) = \fatsemi
+%! (*) = \times
+%! (+) = +
+%! LeftP = L^+
+%! RightP = R^+
+%! LeftT = L^{\times}
+%! RightT = R^{\times}
+%! alpha = \alpha
+%! bool = \textit{bool}
+%! color = \textit{color}
+%! Gr = G
+
+%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
+%! Gx = \Gamma^{\times}
+%! G = \Gamma
+%! [] = \Box
+%! |-->* = \mapsto^{\asterix}
+%! |-->> = \mapsto_{\ggg}
+%! |--> = \mapsto
+%! <--| = \mapsfrom
+%! |- = \vdash
+%! <><> = \approx
+%! ==> = \Longrightarrow
+%! <== = \Longleftarrow
+%! <=> = \Longleftrightarrow
+%! <-> = \leftrightarrow
+%! ~> = \leadsto
+%! -o+ = \multimap^{+}
+%! -o* = \multimap^{\times}
+%! -o = \multimap
+%! ::= = &::=&
+%! /= = \neq
+%! @@ = \mu
+%! [^ = \lceil
+%! ^] = \rceil
+%! forall = \forall
+%! exists = \exists
+%! empty = \epsilon
+%! Pi = \Pi
+%! Pi0 = \Pi^{o}
+%! PiEE* = \Pi^{\eta\epsilon}_{*}
+%! PiEE+ = \Pi^{\eta\epsilon}_{+}
+%! PiEE = \Pi^{\eta\epsilon}
+%! CatSet = \textbf{Set}
+%! theseus = Theseus
+%! sqrt(x) = \sqrt{#x}
+%! surd(p,x) = \sqrt[#p]{#x}
+%! inv(x) = \frac{1}{#x}
+%! frac(x,y) = \frac{#x}{#y}
+%! * = \times
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-\conferenceinfo{ICFP'12}{}
+\conferenceinfo{POPL'13}{}
 \CopyrightYear{}
@@
 
-\title{Continuations via Negative Information Flow}
+\title{Negative and Fractional Types}
+
 \authorinfo{Roshan P. James}
@@
            {rpjames@indiana.edu}
+\authorinfo{Zachary Sparks} 
+           {Indiana University}
+           {zasparks@indiana.edu}
+\authorinfo{Jacques Carette} 
+           {McMaster University}
+           {carette@mcmaster.ca}
 \authorinfo{Amr Sabry}
@@
            {sabry@indiana.edu}
+
 \maketitle
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \category{D.3.1}{Formal Definitions and Theory}{}
@@
 
-Blah\cite{infeffects}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Core Reversible Language: {{Pi}} }
+\label{sec:pi}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+We review our reversible language {{Pi}}: the presentation in this
+section differs from the one in our previous paper~\cite{infeffects} in two
+aspects. First, we add the empty type {{0}} which is necessary to express the
+additive duality. Second, instead of explaining evaluation using a natural
+semantics, we give a small-step operational semantics that is more
+appropriate for the connections with continuations explored in this paper.
 
-\acks This project was partially funded by Indiana University's Office of the
-Vice President for Research and the Office of the Vice Provost for Research
-through its Faculty Research Support Program.  We also acknowledge support
-from Indiana University's Institute for Advanced Study.
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Syntax and Types} 
+\label{sec:pi-syntax}
+
+Unlike the traditional situation with the $\lambda$-calculus, in {{Pi}},
+there is a sharp distinction between data and programs.
+
+\paragraph*{Data.} The sets of values and their types include:
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b + b | b * b 
+% values, v ::= () | left v | right v | (v, v)
+Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
+and products types {{b1*b2}}.  Values includes {{()}} which is the only value
+of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
+and {{(v1,v2)}} which builds a value of product type. There are no values of
+type {{0}}.
+
+\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
+{{v \rightarrow v}}. Specifically, they are witnesses to the following type
+isomorphisms:
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%zeroe :&  0 + b <-> b &: zeroi
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist0 :& 0 * b <-> 0 &: factor0
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+Each line of the above table introduces one or two combinators that witness
+the isomorphism in the middle. Collectively the isomorphisms state that the
+structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
+of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
+multiplication distributes over addition. The isomorphisms are extended to
+form a congruence relation by adding the following constructors that witness
+equivalence and compatible closure:
+%subcode{proof} include main
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+
+\noindent
+To summarize, the syntax of {{Pi}} is given as follows.
+
+\begin{definition}{(Syntax of {{Pi}})}
+\label{def:Pi}
+We collect our types, values, and combinators, to get the full language
+definition.
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% comb.~types, t ::= b <-> b
+% iso ::= zeroe | zeroi 
+%     &|& swap+ | assocl+ | assocr+ 
+%     &|& identl* | identr* 
+%     &|& swap* | assocl* | assocr* 
+%     &|& dist0 | factor0 | dist | factor 
+% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+\end{definition}
+
+\paragraph*{Adjoint.} 
+An important property of the language is that every combinator {{c}} has an
+adjoint {{c{dagger}}} that reverses the action of {{c}}. This is evident by
+construction for the primitive isomorphisms. For the closure combinators, the
+adjoint is homomorphic except for the case of sequencing in which the order
+is reversed, i.e., {{(c1 (;) c2){dagger} = (c2{dagger}) (;) (c1{dagger}) }}.
+
+\begin{definition}[Size of a type]
+\label{def:size}
+The size of a type {{b}}, denoted by {{[^ b ^]}}, is a numeric value
+and is defined to be:
+\vspace{-20pt}
+\begin{multicols}{2}
+%subcode{opsem} include main
+% [^ b1 + b2 ^] '= [^ b1 ^] + [^ b2 ^]
+% [^ b1 * b2 ^] '= [^ b1 ^] * [^ b2 ^]
+
+%subcode{opsem} include main
+% [^ 0 ^] '= 0
+% [^ 1 ^] '= 1
+\end{multicols}
+
+where {{+}} is numeric addition and {{*}} is numeric multiplication.
+
+\end{definition}
+\noindent
+In the setting of {{Pi}} the \emph{size}, {{[^ b ^]}}, may be simply thought of as
+the number of inhabitants (an arity) of the type. This intuition will
+however become tenuous in the presence of negative and fractional
+types.
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Graphical Language}
+
+The syntactic notation above is often obscure and hard to read.
+Following the tradition established for monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present
+a graphical language that conveys the intuitive semantics of the
+language (which is formalized in the next section).
+
+The general idea of the graphical notation is that combinators are modeled by
+``wiring diagrams'' or ``circuits'' and that values are modeled as
+``particles'' or ``waves'' that may appear on the wires. Evaluation therefore
+is modeled by the flow of waves and particles along the wires.
+
+\begin{itemize}
+\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
+  is simply represented as a wire labeled by its type {{b}}, as shown on the
+  left. In more complex diagrams, if the type of a wire is obvious from the
+  context, it may be omitted. When tracing a computation, one might imagine a
+  value {{v}} of type {{b}} on the wire, as shown on the right.
+
+  \begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/b-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/b-wire-value.pdf}
+}
+\end{center}
+  \end{multicols}
+
+\item The product type {{b1*b2}} may be represented using either one wire
+  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
+  case of products represented by a pair of wires, when tracing execution
+  using particles, one should think of one particle on each wire or
+  alternatively as in folklore in the literature on monoidal categories as a
+  ``wave.''
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
+\includegraphics{diagrams/thesis/product-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
+\includegraphics{diagrams/thesis/product-two-wires.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Sum types may similarly be represented by one wire or using
+  parallel wires with a {{+}} operator between them. When tracing the
+  execution of two additive wires, a value can reside on only one of the two
+  wires.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
+\includegraphics{diagrams/thesis/sum-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
+\includegraphics{diagrams/thesis/sum-two-wires.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Associativity is implicit in the graphical language. Three parallel
+  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
+\includegraphics{diagrams/thesis/assoc.pdf}
+}
+\end{center}
+
+\item Commutativity is represented by crisscrossing wires.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
+\includegraphics{diagrams/thesis/swap_times.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
+\includegraphics{diagrams/thesis/swap_plus.pdf}
+}
+\end{center}
+\end{multicols}
+
+By visually tracking the flow of particles on the wires, one can
+verify that the expected types for commutativity are satisfied.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_times_value.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_plus_value.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item The morphisms that witness that {{0}} and {{1}} are the additive and
+  multiplicative units are represented as shown below. Note that since there
+  is no value of type 0, there can be no particle on a wire of type {{0}}.
+  Also since the monoidal units can be freely introduced and eliminated, in
+  many diagrams they are omitted and dealt with explicitly only when they are
+  of special interest.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
+\includegraphics{diagrams/thesis/uniti.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
+\includegraphics{diagrams/thesis/unite.pdf}
+}
+\end{center}  
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
+\includegraphics{diagrams/thesis/zeroi.pdf}
+}
+\end{center}
+\columnbreak
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
+\includegraphics{diagrams/thesis/zeroe.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Finally, distributivity and factoring are represented using the dual
+  boxes shown below:
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/factor.pdf}
+\end{center}
+\end{multicols}
+
+Distributivity and factoring are interesting because they represent
+interactions between sum and pair types. Distributivity should
+essentially be thought of as a multiplexer that redirects the flow of
+{{v:b}} depending on what value inhabits the type {{b1+b2}}, as shown
+below. Factoring is the corresponding adjoint operation.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
+\end{center}
+\end{multicols}
+
+\item Composition TODO.
+
+  \begin{multicols}{2}
+\dgm{c1c2_par_sum}
+\dgm{c1c2_par_times}
+  \end{multicols}
+\dgm{c1c2_seq}
+
+\end{itemize}
+
+\noindent 
+\textit{Example.}  We use the type {{bool}} as a shorthand to denote
+the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
+to be {{false}}. The following combinator is represented by the given
+diagram:
+
+{{c : b * bool <-> b + b}}
+
+{{c = swap* (;) dist (;) (identl* (+) identl*)}}
+
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
+\includegraphics{diagrams/thesis/example1.pdf}
+}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Semantics}
+
+The operational semantics of {{Pi}} is summarized below (also see
+\cite{infeffects}).  The semantics of the primitive combinators is
+given by the following single-step reductions below. Since there are
+no values of type {{0}}, the rules omit the impossible cases:
+\begin{scriptsize}
+%subcode{opsem} include main
+%! columnStyle = rlcl
+% swap+ & (left v) &|-->& right v
+% swap+ & (right v) &|-->& left v 
+% assocl+ & (left v1) &|-->& left (left v1)
+% assocl+ & (right (left v2)) &|-->& left (right v2)
+% assocl+ & (right (right v3)) &|-->& right v3 
+% assocr+ & (left (left v1)) &|-->& left v1
+% assocr+ & (left (right v2)) &|-->& right (left v2)
+% assocr+ & (right v3) &|-->& right (right v3)
+% identl* & ((), v) &|-->& v 
+% identr* & v &|-->& ((), v) 
+% swap* & (v1, v2) &|-->& (v2, v1) 
+% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
+% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
+% dist & (left v1, v3) &|-->& left (v1, v3)
+% dist & (right v2, v3) &|-->& right (v2, v3)
+% factor & (left (v1, v3)) &|-->& (left v1, v3) 
+% factor & (right (v2, v3)) &|-->& (right v2, v3)   
+\end{scriptsize}
+The operational semantics of the closure conditions are presented in
+the usual big-step style.
+\begin{scriptsize}
+%subcode{proof} include main
+%@ ~
+%@@ id v |--> v 
+%
+%@ c{dagger} v1 |--> v2
+%@@ (sym c) v1 |--> v2
+%
+%@ c1  v1 |--> v
+%@ c2  v |--> v2
+%@@ (c1(;)c2)  v1 |--> v2
+%---
+%@ c1  v1 |--> v2
+%@@ (c1 (+) c2)  (left v1) |--> left v2
+%
+%@ c2  v1 |--> v2
+%@@ (c1 (+) c2)  (right v1) |--> right v2
+%---
+%@ c1  v1 |--> v3
+%@ c2  v2 |--> v4
+%@@ (c1 (*) c2)  (v1, v2) |--> (v3, v4)
+\end{scriptsize}
+
+The type safety of {{Pi}} follows directly from the correspondence
+between inductive definitions of the types and the big-step
+semantics. Previous work also established the following properties:
+
+\begin{proposition}[Strong Normalizing]
+\label{prop:termination-pi} 
+{{Pi}} computations always terminate.
+
+{{forall. c:b1<->b2, v:b1, exists v':b2.}}  
+{{c v |--> v'}}
+\end{proposition}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev}
+{{c v |--> v'}} iff 
+{{ c{dagger} v' |--> v}}
+\end{proposition}
+
+%%%%%%%%%%%%%%%%%%%
+\subsection{Constructions}
+
+\paragraph*{Booleans and Conditionals.} 
+Given any combinator {{c : b <-> b}} we can construct a combinator called
+{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
+one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
+the combinator {{c}} is used to transform the value of type~{{b}}. If the
+boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
+write down the combinator for {{if_c}} in terms of {{c}} as 
+{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
+
+\noindent The diagram below shows the input value of type {{(1+1)*b}}
+processed by the distribute operator {{dist}}, which converts it into a value
+of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
+case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
+combinator~{{c}} is applied to the value of type~{{b}}. The right branch
+which corresponds to the boolean being {{false}} passes along the value of
+type {{b}} unchanged.
+
+\begin{center}
+\scalebox{1.0}{
+%%subcode-line{pdfimage}[diagrams/if_c.pdf]
+\includegraphics{diagrams/thesis/cnot.pdf}
+}
+\end{center}
+
+The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
+negates its second argument if the first argument is {{true}}. This
+gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
+equivalent construction that is useful is {{else_{not} }} where we
+negate the second argument only if the first is {{false}}.
+
+Similarly, we can iterate the construction of {{if_c}} to check several
+bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
+checks two booleans and negates the result wire only if they are both
+{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
+universal reversible gate. We can generalize this construction to
+{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
+they are all {{true}}.
+
+\paragraph*{Cloning.}
+Although cloning is generally not allowed in reversible languages, it is
+possible at the cost of having additional constant inputs. For example,
+consider the gate {{else_{not} }}. Generally, the gate maps
+{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
+on the cases in which the second input is {{true}}, we get that the gate maps
+{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
+i.e., the gate clones the first input. A circuit of {{n}} parallel
+{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
+{{true}} inputs in the process.  Let us call this construction
+{{clone^n_{bool} }}.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Categorical Structure}
+
+For the purpose of establishing equivalence of morphisms in the
+categorical presentation we use the following notion of equality:
+
+\begin{lemma}[{{c1  = c2}}]
+We say {{c1 = c2}} if {{c1:b1<->b2}} and {{c2:b1<->b2}} and for all
+{{v1:b}}, we have {{c1 v1 |--> v2}} iff {{c2 v1 |--> v2}} and {{v2:b2}}.
+\end{lemma}
+
+We present the categorical structure with minimum commentary. More
+details may be found in excellent references such as Barr and Wells
+[CITE] and Selinger \cite{springerlink:10.1007/978-3-642-12821-94}.
+
+\begin{lemma}[{{Pi}} is a category]
+  The category {{Pi}} has the types {{b}} as objects and equivalence
+  classes of well-typed combinators {{b1<->b2}} as morphisms. One can
+  check:
+  \begin{enumerate}
+  \item Every object {{b}} has an identity morphism {{id : b <->b}}.
+  \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
+    {{g:b2<->b3}} is given by sequencing {{f (;) g}}. 
+  \item Associativity of composition follows from operational
+    equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
+    {{h:b3<->b4}}).
+
+    Assuming {{v1:b1}}, {{v2:b2}}, {{v3:b3}}, {{v4:b4}},
+    {{f~v1|-->v2}}, {{g v2 |--> v3}} and {{h~v3 |--> v4}}, one can
+    check:
+
+\vspace{-20pt}
+    \begin{multicols}{2}
+      \begin{scriptsize}
+
+%subcode{proof} include main
+%@@ f v1 |--> v2
+%@ g v2 |--> v3
+%@ h v3 |--> v4
+%@@ g (;) h  v2 |--> v4
+%@@@ f (;) (g (;) h) v1 |--> v4
+~
+%subcode{proof} include main
+%@ f v1 |--> v2
+%@ g v2 |--> v3
+%@@ f (;) g v1 |--> v3
+%@@ h v3 |--> v4
+%@@@ (f (;) g) (;) h v1 |--> v4
+
+      \end{scriptsize}
+    \end{multicols}
+
+  \item Composition respects identity: {{id (;) f = f}} and {{g(;)id=g}}.
+  \end{enumerate}
+\end{lemma}
+
+\begin{lemma}[Dagger]
+  {{Pi}} is a dagger category, where every morphism {{f : b1<->b2}}
+  has the adjoint {{f^{dagger}:b2 <->b1}}. The following properties
+  hold (where {{g : b2 <-> b3}}): 
+  \begin{enumerate}
+  \item {{id^{dagger} = id : b <-> b}}.
+  \item {{(f (;) g)^{dagger} = g^{dagger} (;) f^{dagger}: b3 <-> b1}}.
+  \item {{f^{dagger dagger} = f : b1 <-> b2}}.
+  \end{enumerate}
+\end{lemma}
+
+\begin{lemma}[Symmetric Monoidal (+, 0)]
+  {{Pi}} is a symmetric monoidal category with tensor {{+}} and
+  monoidal unit {{0}}. The monoidal operation on morphisms is the
+  additive composition of combinators {{c1 (+) c2}}.
+
+  \begin{proof}
+    To establish that category is monoidal one must show isomorphisms
+    \begin{enumerate}
+    \item {{b1 + (b2 + b3) <-> (b1 +b2) + b3}} given by {{assocl+}}.
+    \item {{0 + b <-> b}} given by {{zeroe}}.
+    \item {{b + 0 <-> b}} given by {{swap+ (;) zeroe}}.
+    \item We need to check that {{+}} is a bifunctor. 
+      \begin{enumerate}
+      \item {{id_{b1<->b1}(+)id_{b2<->b2} = id_{b1+b2<->b1+b2} }}.
+      \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}.
+      \end{enumerate}
+    \item We need to check the naturality of {{assocl+}}, {{zeroe}} and
+      {{swap+(;)zeroe}}.
+      \begin{enumerate}
+      \item {{assocl+ (;) ((f (+)g) (+) h) = (f (+)(g (+) h)) (;) assocl+}}.
+      \item {{zeroe (;) f = (id (+) f) (;) zeroe}}.
+      \item {{(swap+ (;) zeroe) (;) f = (f (+) id) (;) (swap+ (;) zeroe)}}.
+      \end{enumerate}
+    \item Satisfy certain coherence conditions which are usually
+      called the ``pentagon'' and ``triangle'' axioms (see Sec 3.1
+      \cite{springerlink:10.1007/978-3-642-12821-94})
+    \end{enumerate}
+
+    The last three points require checking equality of combinators by
+    writing out their derivation trees as we did in the case of
+    associativity of sequential composition. To show symmetry, we need
+    a braiding operation {{b1+b2 <-> b2+b1}} which is given by
+    {{swap+}} (see Sec. 3.3 and 3.5
+    \cite{springerlink:10.1007/978-3-642-12821-94}).
+
+    \begin{enumerate}
+    \item The braiding must satisfy two ``hexagon'' axioms.
+    \item The braiding is self inverse,
+      {{swap+_{b1+b2}=(swap+_{b2+b1})^{dagger} }}.
+    \end{enumerate}
+
+  \end{proof}
+
+\end{lemma}
+
+\begin{lemma}[Symmetric Monoidal {{(*, 1)}}]
+  {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
+  unit {{1}}. The details mirror those of the {{(0, +)}} monoid.
+\end{lemma}
+
+\noindent
+Some technical and pedantic comments are due at this point.
+
+\begin{itemize}
+\item We have established the categorical structure of {{Pi}} as a
+  dagger symmetric monoidal category with two monoidal structures,
+  {{(0, +)}} and {{(*, 1)}}. In Sec. \ref{sec:int} we will see that a
+  {{trace}} operator can be admitted in this category without any
+  change of expressiveness.
+
+\item To be pedantic, what we have shown is that the ``term model'' of
+  {{Pi}} that follows from the extensional operational equality of
+  combinators has the requisite categorical structure. A consequence
+  is that the ``wiring diagrams'' of {{Pi}} correspond closely with ``string
+  diagrams'' developed for categories.
+
+  To establish the later rigorously, we will need to show when it is
+  valid to slide one wire over the other and that equivalent diagrams
+  for syntactically different combinators such as
+  {{(f(+)g)(;)(j(+)k)}} and {{(f(;)j)(+)(g(;)k)}} do respect
+  operational equivalence.  We don't formalize the graphical notation
+  in this work. Joyal et. al's work on ``planar isotopy'' and
+  Selinger's survey \cite{springerlink:10.1007/978-3-642-12821-94}
+  show how this has been addressed before in the categorical
+  setting. In the absence of any prior knowledge of category theory
+  however, our wiring diagrams may be read as the ``flow of types'' in
+  a combinator-circuit.
+
+\item By definition, in {{Pi}} every morphism is an isomorphism --
+  this makes {{Pi}} a groupoid.
+
+\item The category {{Pi}} has no initial and terminal objects. The
+  objects {{0}} and {{1}} would be initial and terminal if we admitted
+  all functions (as in the category {{CatSet}}).
+
+\item The category {{Pi}} has neither categorical products, nor
+  categorical co-products, i.e. {{*}} and {{+}} are merely monoidal
+  tensors. This follows from the fact that injection and projection
+  maps are not isomorphisms.
+\end{itemize}
+
+While we don't do so in this work, if we extend {{Pi}} with products
+and co-products (say through the addition of information effects) it
+is conceivable that part of its structure collapses. This sort of
+collapse, while catastrophic for algebraic structures (effectively
+trivializing them), still retains some interest in computing, because
+in computing we are interested in the specific operational nature (the
+computational content, so to speak) of the morphisms. Anecdotal evidence
+follows from the fact several real-world programming language have
+inconsistent type systems. blah blah blah...
+
+%%%%%%%%%%%%%
+\subsection{Small Step Semantics}
+
+The reductions for the primitive isomorphisms above are exactly the same as
+have been presented before~\cite{infeffects}. The reductions for the closure
+combinators are however presented in a small-step operational style using the
+following definitions of evaluation contexts and machine states:
+
+\begin{scriptsize}
+%subcode{bnf} include main
+% Combinator Contexts, C = [] | Fst C c | Snd c C 
+%                  &|& LeftT C c v | RightT c v C 
+%                  &|& LeftP C c | RightP c C 
+% Machine states = <c, v, C> | {[c, v, C]}
+% Start state = <c, v, []> 
+% Stop State = {[c, v, []]}
+\end{scriptsize}
+The machine transitions below track the flow of particles through a
+circuit. The start machine state, {{<c,v,[]>}}, denotes the
+particle~{{v}} about to be evaluated by the circuit {{c}}. The end
+machine state, {{[c, v, [] ]}}, denotes the situation where the particle
+{{v}} has exited the circuit {{c}}.
+
+\begin{scriptsize}
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <iso, v, C> &|-->& {[iso, v', C]} & (1)
+% & & where iso v |--> v' &
+% <c1(;)c2, v, C> &|-->& <c1, v, Fst C c2> & (2) 
+% {[c1, v, Fst C c2]} &|-->& <c2, v, Snd c1 C> & (3) 
+% {[c2, v, Snd c1 C]} &|-->& {[ c1(;)c2, v, C ]} & (4) 
+% <c1(+)c2, left v, C> &|-->& <c1, v, LeftP C c2> & (5) 
+% {[ c1, v, LeftP C c2 ]} &|-->& {[c1 (+) c2, left v, C ]} & (6)
+% <c1(+)c2, right v, C> &|-->& <c2, v, RightP c1 C> & (7)
+% {[ c2, v, RightP c1 C ]} &|-->& {[c1 (+) c2, right v, C ]} & (8)
+% <c1(*)c2, (v1, v2), C> &|-->& <c1, v1, LeftT C c2 v2> & (9)
+% {[ c1, v1, LeftT C c2 v2 ]} &|-->& <c2, v2, RightT c1 v1 C> & (10) 
+% {[ c2, v2, RightT c1 v1 C ]} &|-->& {[ c1 (*) c2, (v1, v2), C ]} & (11)
+\end{scriptsize}
+Rule (1) describes evaluation by a primitive isomorphism. Rules (2), (3) and
+(4) deal with sequential evaluation. Rule (2) says that for the value {{v}}
+to flow through the sequence {{c1 (;) c2}}, it should first flow through
+{{c1}} with {{c2}} pending in the context ({{Fst C c2}}). Rule (3) says the
+value {{v}} that exits from {{c1}} should proceed to flow
+through~{{c2}}. Rule (4) says that when the value {{v}} exits {{c2}}, it also
+exits the sequential composition {{c1(;)c2}}. Rules (5) to (8) deal with 
+{{c1 (+) c2}} in the same way. In the case of sums, the shape of the value,
+i.e., whether it is tagged with {{left}} or {{right}}, determines whether
+path {{c1}} or path {{c2}} is taken. Rules (9), (10) and (11) deal with 
+{{c1 (*) c2}} similarly. In the case of products the value should have the
+form {{(v1, v2)}} where {{v1}} flows through {{c1}} and {{v2}} flows through
+{{c2}}. Both these paths are entirely independent of each other and we could
+evaluate either first, or evaluate both in parallel. In this presentation we
+have chosen to follow {{c1}} first, but this choice is entirely arbitrary.
+
+The interesting thing about the semantics is that it represents a reversible
+abstract machine. In other words, we can compute the start state from the
+stop state by changing the reductions {{|-->}} to run backwards
+{{<--|}}. When running backwards, we use the isomorphism represented by a
+combinator {{c}} in the reverse direction, i.e., we use the adjoint
+{{c{dagger}}}.
+
+\begin{proposition}[Correspondence]
+  Evaluation in the small-step evaluator corresponds to evaluation in
+  the natural semantics.
+
+  {{c v |--> v'}} iff {{<c, v, []> |-->* [c', v', [] ]}}
+\end{proposition}
+\begin{proof}
+  To prove the above, we first show that a more general lemma holds,
+  namely that: {{c v |--> v'}} iff {{<c, v, C> |-->* [c', v', C ]}}.
+
+  To show the left-to-right direction we proceed by induction on the
+  derivation of {{c v |--> v'}}. In the case of primitive isomorphisms
+  the condition holds trivially. In the case of composition, we work
+  out the case of {{c1+c2}} as an example. Given {{c1 +c2 v |--> v'}}
+  we have to show that there is a small step derivation sequence that
+  matches it. Here {{v:b1+b2}} can be of the form {{left v1}} or
+  {{right v2}}. Assuming {{left v1}}, we have:
+
+%subcode{proof} include main
+%@ c1 v1 |--> v1' ==> <c1, v1, C'> |-->* {[c1',v1',C']}
+%@@ c1+c2 (left v1) |--> left v1' ==> ?
+
+Choosing {{C' =Lft c2 C}}, we have the required derivation sequence: 
+{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. The proof follows similarly for the {{right v2}} case.
+
+To show the right-to-left direction we proceed by induction of the
+sequence of {{|-->*}} derivations. Again the case for primtive
+isomorphisms follows trivially. Taking the case of {{c1+c2}} we are
+required to show that given a sequence 
+{{<c1+c2,v,C> |-->* {[c', v', C]} }} there is a derviation tree for
+{{c1+c2 v |--> v'}}. In the case that {{v:v1+b2}} has the form
+{{left~v1}}, this follows by observing that the given sequence must have the form
+{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. 
+For the strictly smaller inner subsequence by induction we have
+{{c1~v1 |--> v1'}} which lets us complete the derivation of
+{{c1+c2~(left v1) |--> left v1'}}. The {{right v2}} follows similarly.
+
+\end{proof}
 

### 2012-06-25 rational_ext.tex
_

@@
   information preserving computation \cite{infeffects,rc2011}.
-  Andrzej Filinski's masters thesis
+  Andrzej Filinski's Masters thesis
   \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
@@
     rational numbers}. This computational model is derived
-  systemmatically from the {{Pi}} by considering a symmetric notion of
+  systematically from the {{Pi}} by considering a symmetric notion of
   duality. Unlike the \lcal which shows a classical De Morgan duality,
@@
 
-Andrzej Filinski's masters thesis
+Andrzej Filinski's Masters thesis
 \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
@@
 dual to continuations, functions are dual to delimited
-continuations. Filinksi also shows a De Morgan style duality, by
-treating value pairs as the dual of continuation sums and value sums
-as dual to continuation pairs.
+continuations. Filinski also shows a De Morgan style duality by
+treating pairs of values as dual to sums (choice) of continuation and
+sums of values as dual to pairs of continuations. He, furthermore,
+introduces a symmetric extension of the $\lambda$-calculus in which
+call-by-value is dual to call-by-name.
+
+This view of symmetry between the value fragment of the language and
+the continuation fragment was refined by Curien and Herbelin
+\cite{Curien:2000} who introduce the sequent-style
+$\mu\tilde{\mu}$-calculus, that exhibits symmetries between values and
+continuations and between call-by-value and call-by-name. The calculus
+includes the type $A-B$ which is the dual of implication, i.e., a
+value of type $A-B$ is a context expecting a function of type $A
+\rightarrow B$. Alternatively a value of type $A-B$ is also explained
+as a \emph{pair} consisting of a value of type $A$ and a continuation
+of type $B$. This is to be contrasted with our interpretation of a
+value of that type as \emph{either} a value of $A$ or a demand for a
+value of type $B$. The duality between call-by-name and call-by-value
+was further investigated by Selinger using control
+categories~\cite{Selinger:2001:CCD:966910.966911}. The
+$\mu\tilde{\mu}$-calculus was further analyzed and extended by
+Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives
+no interpretation to the subtraction connective and like the original
+symmetric calculus of Filinski, introduces a duality that relates sums
+to products and vice-versa.
+
+Previous work on \emph{information effects} established the logically
+reversible computational model {{Pi}} (and its extension {{Pi0}})
+wherein computation is information preserving. The computational model
+is derived from the isomorphisms of finite-types, every computation
+admits an adjoint that expresses the logical ``inverse'' of a
+computation and is complete for combinational circuits. The term model
+of {{Pi}} is a so-called commutative bimonoidal category, a term used
+for the categorification of a commutative semiring.
+
+The language {{Pi}} however lacked a natural representation for
+first-class functions due to first-order nature of all of its
+values. In the light of the work of duality of computation it was felt
+that ``right model'' for expressing functions in {{Pi}} would also
+express symmetries similar to those in the calculi of Filinski, Curien
+and Herbelin and Wadler. The technical goal of this paper is to
+present the duality of computation in the setting of {{Pi}}.
+
+The extended language, which we call {{PiEE}}, does not have one De
+Morgan style dualizing action $(\_)^\bot$, but two distinct dualizing
+actions -- an additive duality characterized by negative types $-(\_)$
+and a multiplicative duality characterized by fractional types
+$1/(\_)$. These negative and fractional dualities must not confused
+with the duality in the additive and multiplicative fragment of linear
+logic with the dualizing action $(\-)^\bot$. Each dualizing operation
+preserves the underlying connective {{-(a+b)<->(-a)+(-b)}} and
+{{1/(a*b)<->1/a*1/b}}.
+
+The types of {{PiEE}} thus correspond to the field of rational numbers
+(with the important restriction of fields that there is no division by
+zero). A direct consequence of this is that algebraic manipulations of
+rationals can be given an operational interpretation and \emph{have
+  computational content}.
+
+Models of linear logic rely on *-autonomous categories which change
+the monoidal tensor when dualizing. Compact closed categories preserve
+the monoidal tensor on dualizing.  The term model of {{PiEE}}
+corresponds to the \emph{categorification of a field}. The category is
+compact closed over the additive monoid. In the multiplicative monoid,
+since there is no division by zero, respects compact closed structure
+only when a dual object is defined.
+
+Say more: fractionals, entaglements, backtracking, superposition,
+constrants, negative information. Negative types, backward information
+flows, sets with negative types, iteration.. etc.
+
+We present {{Pi}} in the big-step style presented in the previous
+work, develop the categorical semantics of its term model and present
+a small-step semantics useful for modeling fractionals and
+negatives. We proceed to present the fractional fragment of the
+language {{PiEE*}}, work out its properties and categorical
+semantics. We further use the multiplicative {{trace}} to express a
+SAT solver. We then present the negative fragment of the language
+{{PiEE+}} and finally present the unified language {{PiEE}}.
+
+%% Traditionally, dealing with fields in a categorical setting has proven
+%% hard and hence we hope that the term model of {{PiEE}} will prove
+%% useful. Further, it is as yet unclear why there are two different
+%% dualities in the setting of {{Pi}}.
+
+%% \paragraph*{Declarative Continuations.} 
+%% In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
+%% proposes that continuations are a \emph{declarative} concept. He,
+%% furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+%% which call-by-value is dual to call-by-name and values are dual to
+%% continuations. In more detail, the symmetric calculus contains a ``value''
+%% fragment and a ``continuation'' fragment which are mirror images. Pairs and
+%% sums are treated as duals in the sense that the ``value'' fragment includes
+%% pairs whose mirror image in the ``continuation'' fragment are sums. In
+%% contrast, our language includes pairs and sums in the value fragment and two
+%% symmetries: one that maps the pairs to fractions and another that maps the
+%% sums to subtractions.
 
@@
 \item Composition TODO.
+
+  \begin{multicols}{2}
+\dgm{c1c2_par_sum}
+\dgm{c1c2_par_times}
+  \end{multicols}
+\dgm{c1c2_seq}
 
@@
 required to show that given a sequence 
-{{<c1+c2,v,C> |-->* {[c', v', C]} }} there is a derviation tree for
+{{<c1+c2,v,C> |-->* {[c', v', C]} }} there is a derivation tree for
 {{c1+c2 v |--> v'}}. In the case that {{v:v1+b2}} has the form
@@
 functor. However {{PiEE*}} is not compact closed and {{1/(-)}} is not
-a functor specifically because divsion is not defined on 0 sized
+a functor specifically because division is not defined on 0 sized
 types.
@@
 
-Information Effects, Filinksi, De Morgan
+Information Effects, Filinski, De Morgan
 

### 2012-06-25 rational_ext.tex
_

@@
 
-This view of symmetry between the value fragment of the language and
-the continuation fragment was refined by Curien and Herbelin
-\cite{Curien:2000} who introduce the sequent-style
+Symmetry between the value fragment of the language and the
+continuation fragment was refined by Curien and Herbelin
+\cite{Curien:2000} who introduced the sequent-style
 $\mu\tilde{\mu}$-calculus, that exhibits symmetries between values and
@@
 reversible computational model {{Pi}} (and its extension {{Pi0}})
-wherein computation is information preserving. The computational model
-is derived from the isomorphisms of finite-types, every computation
-admits an adjoint that expresses the logical ``inverse'' of a
-computation and is complete for combinational circuits. The term model
-of {{Pi}} is a so-called commutative bimonoidal category, a term used
-for the categorification of a commutative semiring.
-
-The language {{Pi}} however lacked a natural representation for
+wherein computation is inherently information preserving. The
+computational model is derived from the isomorphisms of finite-types,
+every computation admits an adjoint that expresses the logical
+``inverse'' of a computation and is complete for combinational
+circuits. The term model of {{Pi}} is a so-called commutative
+bimonoidal category, i.e. it corresponds to the categorification of a
+commutative semiring with two commutative monoidal structures
+{{(0,+)}} and {{(*,1)}} with {{*}} distributing over {{+}}.
+
+The language {{Pi}} however lacks a natural representation for
 first-class functions due to first-order nature of all of its
@@
 and Herbelin and Wadler. The technical goal of this paper is to
-present the duality of computation in the setting of {{Pi}}.
-
-The extended language, which we call {{PiEE}}, does not have one De
-Morgan style dualizing action $(\_)^\bot$, but two distinct dualizing
-actions -- an additive duality characterized by negative types $-(\_)$
-and a multiplicative duality characterized by fractional types
-$1/(\_)$. These negative and fractional dualities must not confused
-with the duality in the additive and multiplicative fragment of linear
-logic with the dualizing action $(\-)^\bot$. Each dualizing operation
-preserves the underlying connective {{-(a+b)<->(-a)+(-b)}} and
-{{1/(a*b)<->1/a*1/b}}.
-
-The types of {{PiEE}} thus correspond to the field of rational numbers
-(with the important restriction of fields that there is no division by
-zero). A direct consequence of this is that algebraic manipulations of
-rationals can be given an operational interpretation and \emph{have
+present such a symmetric dual extension in the setting of {{Pi}}.
+
+The extended language, which we call {{PiEE}}, is interesting because
+it does not have one De Morgan style dualizing action $(\_)^\bot$, but
+two distinct dualizing actions -- an additive duality characterized by
+negative types $-(\_)$ and a multiplicative duality characterized by
+fractional types $1/(\_)$. Each dualizing operation preserves the
+underlying connective {{-(a+b)<->(-a)+(-b)}} and
+{{1/(a*b)<->1/a*1/b}}. Negative and fractional types have a clear and
+intuitive semantics in the setting of {{Pi}}.
+
+These negative and fractional dualities must not confused with
+$(\-)^\bot$ duality in the additive and multiplicative fragment of
+linear logic.  Models of linear logic rely on *-autonomous categories
+which change the monoidal tensor when dualizing. Compact closed
+categories preserve the monoidal tensor on dualizing.  The term model
+of {{PiEE}} corresponds to the \emph{categorification of a field}. The
+category is compact closed over the additive monoid. In the
+multiplicative monoid, since there is no division by zero, respects
+compact closed structure only when a dual object is defined.
+
+The types of {{PiEE}} correspond to the \emph{field of rational
+  numbers} (with the important restriction of fields that there is no
+division by zero). Consequently algebraic manipulations of rationals
+can be given an operational interpretation and \emph{have
   computational content}.
-
-Models of linear logic rely on *-autonomous categories which change
-the monoidal tensor when dualizing. Compact closed categories preserve
-the monoidal tensor on dualizing.  The term model of {{PiEE}}
-corresponds to the \emph{categorification of a field}. The category is
-compact closed over the additive monoid. In the multiplicative monoid,
-since there is no division by zero, respects compact closed structure
-only when a dual object is defined.
 

### 2012-06-25 rational_ext.tex
_

@@
 
-These negative and fractional dualities must not confused with
-$(\-)^\bot$ duality in the additive and multiplicative fragment of
-linear logic.  Models of linear logic rely on *-autonomous categories
-which change the monoidal tensor when dualizing. Compact closed
-categories preserve the monoidal tensor on dualizing.  The term model
-of {{PiEE}} corresponds to the \emph{categorification of a field}. The
-category is compact closed over the additive monoid. In the
-multiplicative monoid, since there is no division by zero, respects
-compact closed structure only when a dual object is defined.
-
 The types of {{PiEE}} correspond to the \emph{field of rational
@@
 can be given an operational interpretation and \emph{have
-  computational content}.
+  computational content}.  These negative and fractional dualities
+must not confused with $(\-)^\bot$ duality in the additive and
+multiplicative fragment of linear logic.  Models of linear logic rely
+on *-autonomous categories which change the monoidal tensor when
+dualizing. Compact closed categories preserve the monoidal tensor on
+dualizing.  The term model of {{PiEE}} corresponds to the
+\emph{categorification of a field}. The category is compact closed
+over the additive monoid. In the multiplicative monoid (since there is
+no division by zero) respects compact closed structure only when a
+dual object is defined.
 

### 2012-06-25 code/modular/ND.hs
add some more instances for testing and generalize the type of eval

@@ -1,5 +1,6 @@
 {-# OPTIONS_GHC -XTypeOperators  #-}
 {-# OPTIONS_GHC -XGADTs          #-}
+{-# OPTIONS_GHC -XNoMonomorphismRestriction #-}
 
 -- Fractional types introduce non-determinism; instead of having one value we
 -- have a sequence of possible values
@@ -30,6 +31,14 @@ instance (V a, V b) => V (a,b) where
 instance V a => V (Inv a) where
   elems = map Inv elems
 
+-- for further testing
+data Bob = Uncle | Aunt | Other | Unrelated deriving (Eq, Show)
+instance V Bob where
+  elems = [Uncle, Aunt, Other, Unrelated]
+
+instance V Char where
+  elems = ['a'..'z']
+
 data a :<=> b where 
 -- Congruence
   Id    :: a :<=> a
@@ -54,8 +63,8 @@ data a :<=> b where
   Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
   Factor      :: Either (b, a) (c, a) :<=> (Either b c, a)
 -- Multiplicative inverses (a cannot be 0) 
-  EtaTimes :: V a => () :<=> (Inv a, a)
-  EpsTimes :: Eq a => (Inv a, a) :<=> ()
+  EtaTimes :: V c => () :<=> (Inv c, c)
+  EpsTimes :: Eq c => (Inv c, c) :<=> ()
 
 instance Pi (:<=>) where
   id = Id
@@ -95,7 +104,7 @@ instance PiInv (:<=>) where
 -- v==v'. If so we return () and otherwise we fail.
 
 class Pi eq => EvIso eq where
-  eval :: eq a b -> a -> [b]
+  eval :: (MonadPlus m, Monad m) => eq a b -> a -> m b
 
 instance EvIso (:<=>) where
   eval Id a = return a
@@ -190,7 +199,7 @@ instance EvIso (:<=>) where
 
 -- If the input is non-deterministic use:
 
-eval_iso :: (V a, V b) => (a :<=> b) -> [a] -> [b] 
+eval_iso :: (V a, V b, Monad m, MonadPlus m) => (a :<=> b) -> m a -> m b
 eval_iso c xs = xs >>= eval c
 
 ------------------------------------------------------------------------------
@@ -198,7 +207,7 @@ eval_iso c xs = xs >>= eval c
 
 type B = Either () ()
 
[…]

### 2012-06-25 rational_ext.tex
_

@@
 underlying computation in the \lcal. Filinski showed that values are
-dual to continuations, functions are dual to delimited
-continuations. Filinski also shows a De Morgan style duality by
-treating pairs of values as dual to sums (choice) of continuation and
-sums of values as dual to pairs of continuations. He, furthermore,
-introduces a symmetric extension of the $\lambda$-calculus in which
-call-by-value is dual to call-by-name.
-
-Symmetry between the value fragment of the language and the
-continuation fragment was refined by Curien and Herbelin
-\cite{Curien:2000} who introduced the sequent-style
-$\mu\tilde{\mu}$-calculus, that exhibits symmetries between values and
-continuations and between call-by-value and call-by-name. The calculus
-includes the type $A-B$ which is the dual of implication, i.e., a
-value of type $A-B$ is a context expecting a function of type $A
-\rightarrow B$. Alternatively a value of type $A-B$ is also explained
-as a \emph{pair} consisting of a value of type $A$ and a continuation
-of type $B$. This is to be contrasted with our interpretation of a
-value of that type as \emph{either} a value of $A$ or a demand for a
-value of type $B$. The duality between call-by-name and call-by-value
-was further investigated by Selinger using control
+dual to continuations, functions are dual to delimited continuations
+and a De Morgan style duality by treating pairs of values as dual to
+sums (choice) of continuation and sums of values as dual to pairs of
+continuations. He, furthermore, introduced a symmetric extension of
+the $\lambda$-calculus in which call-by-value is dual to call-by-name.
+
+Symmetry between the value fragment and the continuation fragment was
+refined by Curien and Herbelin \cite{Curien:2000} who introduced the
+sequent-style $\mu\tilde{\mu}$-calculus, that exhibits symmetries
+between values and continuations and between call-by-value and
+call-by-name. The calculus includes the type $A-B$ which is the dual
+of implication, i.e., a value of type $A-B$ is a context expecting a
+function of type $A \rightarrow B$. Alternatively a value of type
+$A-B$ is also explained as a \emph{pair} consisting of a value of type
+$A$ and a continuation of type $B$. This is to be contrasted with our
+interpretation of a value of that type as \emph{either} a value of $A$
+or a demand for a value of type $B$. The duality between call-by-name
+and call-by-value was further investigated by Selinger using control
 categories~\cite{Selinger:2001:CCD:966910.966911}. The
@@
 The language {{Pi}} however lacks a natural representation for
-first-class functions due to first-order nature of all of its
-values. In the light of the work of duality of computation it was felt
-that ``right model'' for expressing functions in {{Pi}} would also
-express symmetries similar to those in the calculi of Filinski, Curien
-and Herbelin and Wadler. The technical goal of this paper is to
-present such a symmetric dual extension in the setting of {{Pi}}.
+first-class functions due to the first-order nature of its values. In
+the light of the work of duality of computation it was felt that
+``right model'' for expressing functions in {{Pi}} would also express
+symmetries similar to those in the calculi of Filinski, Curien and
+Herbelin and Wadler. The technical goal of this paper is to present
+such a symmetric dual extension of {{Pi}}.
 

### 2012-06-25 rational_ext.tex
_

--- 2012-06-25 rational_ext.tex
@@
 The language {{Pi}} however lacks a natural representation for
-first-class functions due to first-order nature of all of its
+first-class functions due to the first-order nature of its
 values. In the light of the work of duality of computation it was felt
@@
 
-\item Composition TODO.
+\item Combinators can be composed is series ({{c1 (;) c2}}) or
+  parallel. There are two forms of parallel composition -- combinators
+  can be combined additively {{c1(+)c2}} (shown on the left) or
+  multiplicatively {{c1(*)c2}} (shown on the right).
 
@@
 
-\noindent 
-\textit{Example.}  We use the type {{bool}} as a shorthand to denote
-the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
-to be {{false}}. The following combinator is represented by the given
-diagram:
-
-{{c : b * bool <-> b + b}}
-
-{{c = swap* (;) dist (;) (identl* (+) identl*)}}
-
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
-\includegraphics{diagrams/thesis/example1.pdf}
-}
-\end{center}
+%% \noindent 
+%% \textit{Example.}  We use the type {{bool}} as a shorthand to denote
+%% the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
+%% to be {{false}}. The following combinator is represented by the given
+%% diagram:
+
+%% {{c : b * bool <-> b + b}}
+
+%% {{c = swap* (;) dist (;) (identl* (+) identl*)}}
+
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
+%% \includegraphics{diagrams/thesis/example1.pdf}
+%% }
+%% \end{center}
 
@@
 {{Pi}} computations always terminate.  
-
 {{forall. c:b1<->b2, v:b1, exists v':b2.}}  
@@
 \end{proposition}
+
+\begin{definition}[{{c1  = c2}}]
+We say {{c1 = c2}} if {{c1:b1<->b2}} and {{c2:b1<->b2}} and for all
+{{v1:b}}, we have {{c1 v1 |--> v2}} iff {{c2 v1 |--> v2}} and {{v2:b2}}.
+\end{definition}
 
@@
 \subsection{Categorical Structure}
-
-For the purpose of establishing equivalence of morphisms in the
-categorical presentation we use the following notion of equality:
-
-\begin{lemma}[{{c1  = c2}}]
-We say {{c1 = c2}} if {{c1:b1<->b2}} and {{c2:b1<->b2}} and for all
-{{v1:b}}, we have {{c1 v1 |--> v2}} iff {{c2 v1 |--> v2}} and {{v2:b2}}.
-\end{lemma}
 

### 2012-06-26 rational_ext.tex
_

@@
 
-%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions}
@@
 
-% Note that this language is not logically reversible.
+\begin{definition}[eval(c,v)]
+
+Given a combinator {{c:b1<->b2}} and a value {{v:b1}} we can define
+{{eval(c, v)}} to be the set of values {{v':b}} such that there is
+small-step reduction sequence {{c v |-->* v'}}.
+
+$eval(c, v) = \{v'~|~ c~ v\mapsto^{\asterix} v' \} $
+\end{definition}
+
+The definition of equality between combinators has to be extended to
+reflect the fact that combinators can have multiple outputs for an
+input. More generally, combinators denotes relations.
+
+\begin{definition}[c1 = c2]
+We say that {{c1 = c2}}, for combinators {{c1:b1<->b2}} and
+{{c2:b1<->b2}} iff the denote the same relation between sets {{b1}}
+and {{b2}}. Equivalently,
+
+{{forall v:b1. eval(c1, v) = eval(c2, v)}}
+\end{definition}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev2}
+
+{{c v |-->* v'}} iff {{ c{dagger} v' |-->* v}}
+\end{proposition}
+
+The logical reversibility is extended to relations and should be
+understood as follows. The combinator {{c:b1<->b2}} denotes a relation
+from {{b1}} to {{b2}} and its adjoint denotes the inverse relation
+from {{b2}} to {{b1}}. Operationally, for any execution path from
+{{v}} to {{v'}} in {{c}}, {{c{dagger} }} has an execution path from
+{{v'}} to {{v}}.
+
+\roshan{This is an important question -- is {{PiEE*}} still
+  bimonoidal? i.e. so all the coherence conditions on isomorphisms
+  naturally extend to relations. I suspect yes.}
 

### 2012-06-26 rational_ext.tex
_

--- 2012-06-25 rational_ext.tex
@@
 %! trace* = trace^{\times}
-%! ^^^ = ^{-1}
 %! (o) = \circ
@@
 %! ^] = \rceil
+%! {| = \{
+%! |} = \}
+%! ^^^ = ^{\dagger}
+%! elem = \in
 %! forall = \forall
@@
 to be viewed as a fission point for a value of type {{b}} and its
-multiplicative inverse {{1/b}}. Operationally, this corresponds to the
-creation of two values {{alpha}} and {{1/alpha}} where {{alpha}} is a
-fresh logic variable. The operator {{eps*}} then becomes a unification
-site for these logic variables:
+multiplicative inverse {{1/b}}. 
+%% Operationally, this corresponds to the
+%% creation of two values {{alpha}} and {{1/alpha}} where {{alpha}} is a
+%% fresh logic variable. The operator {{eps*}} then becomes a unification
+%% site for these logic variables:
 
@@
 
+\roshan{Maybe replace the annihilation example with union and
+  difference.}
+
 \begin{proposition}[Type Safety]
@@
 
-% Note that this language is not logically reversible.
+In the discourse below we use {{c v |-->* v'}} as a shorthand for the
+execution sequence {{<c,v,[]>|-->*{[c,v',[] ]} }}.
+
+\begin{definition}[eval(c,v)]
+
+Given a combinator {{c:b1<->b2}} and a value {{v:b1}} we define
+{{eval(c, v)}} to be the set of values {{v':b}} such that there is
+small-step reduction sequence {{c v |-->* v'}}.
+
+$eval(c, v) = \{v'~|~ c~ v\mapsto^{\asterix} v' \} $
+\end{definition}
+
+The above definition implies that combinators {{c:b1<->b2}} can be
+indetified by relations between the sets {{b1}} and {{b2}}, i.e. 
+{{c:b1<->b2 = {| (v, v') | v:b2, v':b2 and c v |-->* v'|} }}.
+The definition of equality between combinators is extended reflect the
+fact that combinators denote relations.
+
+\begin{definition}[c1 = c2]
+We say that {{c1 = c2}}, for combinators {{c1:b1<->b2}} and
+{{c2:b1<->b2}} iff they denote the same relation between sets {{b1}}
+and {{b2}}. Equivalently,
+
+{{forall v:b1. eval(c1, v) = eval(c2, v)}}
+\end{definition}
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev2} ~
+
+{{c v |-->* v'}} iff {{ c{dagger} v' |-->* v}}
+\end{proposition}
+
+Logical reversibility of {{PiEE*}} implies that {{c^{dagger} }}
+denotes the inverse relation that {{c}} denotes, i.e.
+{{c^^^={|(v',v)~|~(v,v') elem c|} }}.  Operationally, for any
+execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }} has an
+execution path from {{v'}} to {{v}}.
+
+\roshan{This is an important question -- is {{PiEE*}} still
+  bimonoidal? i.e. do all the coherence conditions on isomorphisms
+  naturally extend to relations. I suspect yes.}
 

### 2012-06-26 rational_ext.tex
_

@@
 
-  {{c v |--> v'}} iff {{<c, v, []> |-->* [c', v', [] ]}}
+  {{c v |--> v'}} iff {{<c, v, []> |-->* [c, v', [] ]}}
 \end{proposition}
@@
   To prove the above, we first show that a more general lemma holds,
-  namely that: {{c v |--> v'}} iff {{<c, v, C> |-->* [c', v', C ]}}.
+  namely that: {{c v |--> v'}} iff {{<c, v, C> |-->* [c, v', C ]}}.
 
@@
 %subcode{proof} include main
-%@ c1 v1 |--> v1' ==> <c1, v1, C'> |-->* {[c1',v1',C']}
+%@ c1 v1 |--> v1' ==> <c1, v1, C'> |-->* {[c1,v1',C']}
 %@@ c1+c2 (left v1) |--> left v1' ==> ?
@@
 Choosing {{C' =Lft c2 C}}, we have the required derivation sequence: 
-{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. The proof follows similarly for the {{right v2}} case.
+{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1,v1', LeftP c2 C]} |--> {[c1+c2,left v1',C]} }}. The proof follows similarly for the {{right v2}} case.
 
@@
 required to show that given a sequence 
-{{<c1+c2,v,C> |-->* {[c', v', C]} }} there is a derivation tree for
+{{<c1+c2,v,C> |-->* {[c1+c2, v', C]} }} there is a derivation tree for
 {{c1+c2 v |--> v'}}. In the case that {{v:v1+b2}} has the form
 {{left~v1}}, this follows by observing that the given sequence must have the form
-{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1',v1', LeftP c2 C]} |--> {[c1'+c2,left v1',C]} }}. 
+{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1,v1', LeftP c2 C]} |--> {[c1+c2,left v1',C]} }}. 
 For the strictly smaller inner subsequence by induction we have

### 2012-06-26 rational_ext.tex
_

@@
 follows: when the machine encounters an {{eta*}} reduction it
-non-deterministically replicates into several machine states, each of
-which is seeded by a unique value {{v}} that inhabits the type {{b}}
-that indexes {{eta* : 1 <-> 1/b * b}}. As an example at the type
-{{bool}}, we would have two possible machine states:
+non-deterministically replicates into several machine states or
+worlds, each of which is seeded by a unique value {{v}} that inhabits
+the type {{b}} that indexes {{eta* : 1 <-> 1/b * b}}. Each world has
+an independant existence. As an example at the type {{bool}}, we would
+have two possible machine states:
 
@@
 execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }} has an
-execution path from {{v'}} to {{v}}.
+execution path from {{v'}} to {{v}}.  Intuitively, logical
+reversibility holds because for every {{eta*}} reduction in a world
+that transforms {{()}} into {{(1/v,v)}}, on reverse execution the
+adjoint {{eps*}} will transform {{(1/v,v)}} back to {{()}}.
 
@@
 
-As with {{Pi}}, {{PiEE*}} is a dagger symmetric monoidal category with
-two monoidal structures. In this section we establish that it is
-``locally compact closed'' in the sense that for non-zero objects,
-namely types where {{[^ b ^]=0}}, {{PiEE*}} satisfies the conditions
-for compact closed categories.
+Like {{Pi}}, {{PiEE*}} is a commutative bimonoidal catgeory.
+Symmetric monoidal categories where every object had a dual object and
+where dualizing preserves the monodal tensor are referred to as
+compact closed categories. Since divison is defined only for types of
+non-zero size, not all objects have dual objects.  In this section
+establish that {{PiEE*}} is ``locally compact closed'' in the sense
+that for non-zero objects, namely types where {{[^ b ^]=0}}, it
+satisfies the conditions for compact closed categories.
 

### 2012-06-26 rational_ext.tex
_

@@
 [CITE] and Selinger \cite{springerlink:10.1007/978-3-642-12821-94}.
+For brevity, we don't re-state standard categorical definitions and
+subsequently present their {{Pi}} equivalents. Instead, we directly
+state categorical definitions in terms of {{Pi}} types and
+combinators, and provide appropriate references to the former. For
+example, associativity is denoted by the greek letter $\alpha$ (and
+$\alpha^{-1}$) in standard definitions, whereas we use {{assocl*}}
+(and {{assocr*}}).
 
@@
   naturally extend to relations. I suspect yes.}
+
+\roshan{Should we show completeness -- i.e. that every relation is
+  expressible? We can. But should we?}
 

### 2012-06-26 rational_ext.tex
_

@@
 this translates into the familiar restriction that there is no
-division by 0 in arithmetic or the definition of fields in abstract
-algebra.
+division by 0 in arithmetic, nor in algebraic fields.
 
@@
 The rule {{(eta*)}} is different from other rules encountered before
-due to the {{forall v:b}} quantification and is to be understood as
+due to the {{forall v:b}} quantification. It is to be understood as
 follows: when the machine encounters an {{eta*}} reduction it
-non-deterministically replicates into several machine states or
-worlds, each of which is seeded by a unique value {{v}} that inhabits
-the type {{b}} that indexes {{eta* : 1 <-> 1/b * b}}. Each world has
-an independant existence. As an example at the type {{bool}}, we would
-have two possible machine states:
+non-deterministically replicates into several machine states (or
+worlds). Each world is seeded by a unique value {{v:b}}, where {{b}}
+comes from {{eta*:1 <-> 1/b * b}}.  For example, at
+{{eta*:1<->1/bool*bool}}, we would have two possible machine
+states:
 
@@
 The rule {{(eps*)}} is straightforward : when given a tuple
-{{(1/v,v):1/b * b}} where {{eta*:1/b*b <-> 1}} the result is
+{{(1/v,v):1/b * b}} where {{eps*:1/b*b <-> 1}} the result is
 {{()}}. The machine is undefined for cases where where the input is
@@
 
-In the discourse below we use {{c v |-->* v'}} as a shorthand for the
-execution sequence {{<c,v,[]>|-->*{[c,v',[] ]} }}.
+\noindent
+We use {{c v |-->* v'}} as a shorthand for the reduction sequence
+{{<c,v,[]>|-->*{[c,v',[] ]} }}.
 
@@
 
-The above definition implies that combinators {{c:b1<->b2}} can be
-indetified by relations between the sets {{b1}} and {{b2}}, i.e. 
-{{c:b1<->b2 = {| (v, v') | v:b2, v':b2 and c v |-->* v'|} }}.
-The definition of equality between combinators is extended reflect the
-fact that combinators denote relations.
+The above definition implies that combinators {{c:b1<->b2}} denote
+relations between the sets {{b1}} and {{b2}}, i.e.
+{{c:b1<->b2={|(v,v') | v:b2, v':b2 and c v |-->* v'|} }}.  The
+definition of combinator equality reflects the fact that combinators
+denote relations.
 
@@
 We say that {{c1 = c2}}, for combinators {{c1:b1<->b2}} and
-{{c2:b1<->b2}} iff they denote the same relation between sets {{b1}}
-and {{b2}}. Equivalently,
-
-{{forall v:b1. eval(c1, v) = eval(c2, v)}}
+{{c2:b1<->b2}} iff they denote the same relation. Equivalently,
+{{forall v:b1. eval(c1, v) = eval(c2, v)}}.
 \end{definition}
@@
 
-Logical reversibility of {{PiEE*}} implies that {{c^{dagger} }}
-denotes the inverse relation that {{c}} denotes, i.e.
-{{c^^^={|(v',v)~|~(v,v') elem c|} }}.  Operationally, for any
-execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }} has an
-execution path from {{v'}} to {{v}}.  Intuitively, logical
-reversibility holds because for every {{eta*}} reduction in a world
-that transforms {{()}} into {{(1/v,v)}}, on reverse execution the
-adjoint {{eps*}} will transform {{(1/v,v)}} back to {{()}}.
+For any execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }}
+has an execution path from {{v'}} to {{v}}.  Stated differently,
+logical reversibility implies that {{c^{dagger} }} denotes the inverse
+relation that {{c}} denotes, i.e.  {{c^^^={|(v',v)~|~(v,v') elem c|} }}.  
+Intuitively logical reversibility holds because for every
+{{eta*}} reduction in a world that transforms {{()}} into {{(1/v,v)}},
+on reverse execution the adjoint {{eps*}} will transform {{(1/v,v)}}
+back to {{()}}.
+
+\roshan{Should we show completeness -- i.e. that every relation is
+  expressible? We can. But should we?}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Categorical Structure}
+
+Like {{Pi}}, {{PiEE*}} is a commutative bimonoidal catgeory.
+Symmetric monoidal categories where every object had a dual object and
+where dualizing preserves the monoidal tensor are referred to as
+compact closed categories. Since divison is defined only for types of
+non-zero size, not all objects have duals.  We show that {{PiEE*}}
+respects compact closed structure for non-zero objects. We introduce
+the term ``locally compact closed'' to refer to this notion.
+
+It is important to note that in a compact closed category, the
+dualizing action, usually denoted $(-)^\asterix$, is a contravariant
+functor. However {{PiEE*}} is not compact closed and {{1/(-)}} is not
+a functor specifically because division is not defined on 0 sized
+types.
+
+Without the important side condition on division by zero, we would
+have {{1<->1/0*0}} which implies {{1 <-> 0}} consequently collapsing
+the entire structure. In this context, an important area for future
+investigation is to consider alternatives to fields as a model for
+fractionals and negatives -- particularly algebraic structures that
+are ``zero totalized fields'' such as meadows
+\cite{Bergstra:2009:MES:1507774.1508000}.
 
@@
 
-\roshan{Should we show completeness -- i.e. that every relation is
-  expressible? We can. But should we?}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Categorical Structure}
-
-Like {{Pi}}, {{PiEE*}} is a commutative bimonoidal catgeory.
-Symmetric monoidal categories where every object had a dual object and
-where dualizing preserves the monodal tensor are referred to as
-compact closed categories. Since divison is defined only for types of
-non-zero size, not all objects have dual objects.  In this section
-establish that {{PiEE*}} is ``locally compact closed'' in the sense
-that for non-zero objects, namely types where {{[^ b ^]=0}}, it
-satisfies the conditions for compact closed categories.
-
 \begin{proposition}[Locally Compact Closed]
-{{PiEE*}} is a locally compact closed category in the sense that for
-all non-zero objects {{b}}, we have an inverse object {{1/b}} in the
-symmetric monoidal structure of {{(*, 1)}}, with the morphisms:
+For every non-zero object {{b}} of {{PiEE*}}, there exists an inverse
+object {{1/b}} in the {{(*, 1)}} symmetric monoid with the morphisms:
 
@@
 \noindent
-The coherence conditions are more intuitive when presented
+The coherence conditions are intuitive when presented
 graphically. They require that the combinators below be equal to
@@
   We briefly trace the execution of the first combinator (let us call
-  it {{c}}. To show that {{c=id}}, we must show for {{v:b}}, the
+  it {{c}}). To show that {{c=id}}, we must show for {{v:b}}, the
   evaluation of {{<c,v,[]>}} always produces {{v:b}} as
@@
 % & |-->~ & [eps*, (), C3]
-% & |-->* & [c', v, [] ]
+% & |-->* & [c, v, [] ]
 % (2) & |-->* & <eps*, (1/v', v), C3> & when v' /= v
@@
 \end{proof}
-
-It is important to note that in a compact closed category, the
-dualizing action, usually denoted $(-)^\asterix$, is a contravariant
-functor. However {{PiEE*}} is not compact closed and {{1/(-)}} is not
-a functor specifically because division is not defined on 0 sized
-types.
-
-Without the important side condition on division by zero, we would
-have {{1<->1/0*0}} which implies {{1 <-> 0}} and consequently
-collapses the entire structure. Structures that are ``zero totalized
-fields'' such as meadows \cite{Bergstra:2009:MES:1507774.1508000}
-deserve investigation as alternative models for {{Pi}} with
-fractionals (and negatives).
 

### 2012-06-26 rational_ext.tex
_

@@
 
-Say more: fractionals, entaglements, backtracking, superposition,
-constrants, negative information. Negative types, backward information
+Say more: fractionals, entanglements, backtracking, superposition,
+constraints, negative information. Negative types, backward information
 flows, sets with negative types, iteration.. etc.
@@
 combinators, and provide appropriate references to the former. For
-example, associativity is denoted by the greek letter $\alpha$ (and
+example, associativity is denoted by the Greek letter $\alpha$ (and
 $\alpha^{-1}$) in standard definitions, whereas we use {{assocl*}}
@@
 To show the right-to-left direction we proceed by induction of the
-sequence of {{|-->*}} derivations. Again the case for primtive
+sequence of {{|-->*}} derivations. Again the case for primitive
 isomorphisms follows trivially. Taking the case of {{c1+c2}} we are
@@
 
-For any execution path from {{v}} to {{v'}} in {{c}}, {{c{dagger} }}
-has an execution path from {{v'}} to {{v}}.  Stated differently,
-logical reversibility implies that {{c^{dagger} }} denotes the inverse
-relation that {{c}} denotes, i.e.  {{c^^^={|(v',v)~|~(v,v') elem c|} }}.  
-Intuitively logical reversibility holds because for every
-{{eta*}} reduction in a world that transforms {{()}} into {{(1/v,v)}},
-on reverse execution the adjoint {{eps*}} will transform {{(1/v,v)}}
-back to {{()}}.
+In other words, for any execution path from {{v}} to {{v'}} in {{c}},
+{{c{dagger} }} has an execution path from {{v'}} to {{v}}.
+Intuitively this maybe understood as follows: For every particular
+execution sequence {{eta*}} transforms {{()}} into particular
+{{(1/v,v)}}. On reverse execution the adjoint {{eps*}} will transform
+{{(1/v,v)}} back to {{()}}.  Logical reversibility implies that
+{{c^^^}} denotes the inverse relation that {{c}} denotes, i.e.
+{{c^^^={|(v',v)~|~(v,v') elem c|} }}.
+
+\roshan{This probably needs rigorous proof.}
 
@@
 
-Like {{Pi}}, {{PiEE*}} is a commutative bimonoidal catgeory.
+Like {{Pi}}, {{PiEE*}} is a commutative bimonoidal category.
 Symmetric monoidal categories where every object had a dual object and
 where dualizing preserves the monoidal tensor are referred to as
-compact closed categories. Since divison is defined only for types of
+compact closed categories. Since division is defined only for types of
 non-zero size, not all objects have duals.  We show that {{PiEE*}}

### 2012-06-26 rational_ext.tex
_

@@
 
-\begin{definition}[{{c1  = c2}}]
+\begin{definition}[ {{c1  = c2}} ]
 We say {{c1 = c2}} if {{c1:b1<->b2}} and {{c2:b1<->b2}} and for all
@@
 
-\begin{definition}[c1 = c2]
+\begin{definition}[{{c1 = c2}}]
 We say that {{c1 = c2}}, for combinators {{c1:b1<->b2}} and

### 2012-06-26 rational_ext.tex
small changes

--- 2012-06-26 rational_ext.tex
@@
 \documentclass[preprint]{sigplanconf}
-
-%% I am getting an error about too many math packages used!!!
-%% I commented the ones we don't seem to be using.
 
@@
 \newcommand{\fconame}[1]{\llcorner #1 \lrcorner}
-
-%% \newtheorem{theorem}{Theorem}[section]
-%% \newtheorem{lemma}[theorem]{Lemma}
-%% \newtheorem{proposition}[theorem]{Proposition}
-%% \newtheorem{corollary}[theorem]{Corollary}
 
@@
 %! |} = \}
+%! dots = \ldots
 %! ^^^ = ^{\dagger}
@@
 
-\title{Computing in the Field of Rationals}
-
-\authorinfo{Nicolas Bourbaki}
-
-%% \authorinfo{Roshan P. James}
-%%            {Indiana University}
-%%            {rpjames@indiana.edu}
-%% \authorinfo{Amr Sabry}
-%%            {Indiana University}
-%%            {sabry@indiana.edu}
+\title{Negative and Fractional Types}
+
+\authorinfo{Roshan P. James}
+           {Indiana University}
+           {rpjames@indiana.edu}
+\authorinfo{Zachary Sparks} 
+           {Indiana University}
+           {zasparks@indiana.edu}
+\authorinfo{Jacques Carette} 
+           {McMaster University}
+           {carette@mcmaster.ca}
+\authorinfo{Amr Sabry}
+           {Indiana University}
+           {sabry@indiana.edu}
 
@@
   %% to constraint propagation.
+
 \end{abstract}
@@
 
+\begin{verbatim}
+Restructure this section to emphasize the following
+
+In plain Pi: 
+------------
+
+types b ::= 0 | 1 | b+b | b*b
+values v ::= () | L v | R v | (v,v)
+typing v : b
+
+Semantics of values: 
+ - a type b is interpreted as a set
+ - a value v : b is an element of the set denoted by b
+
+types b <-> b
+combinators c ::= primitiveIso
+               | c + c | c * c | c ; c 
+typing c : b1 <-> b2
+
+Semantics of combinators:
+ - types b1 <-> b2 are interpreted as bijections 
+   between the sets denoted by b1 and b2
+ - a combinator c : b1 <-> b2 is interpeted as a 
+   bijection between the appropriate sets
+
+Evaluation is function application, i.e., 
+  eval c v applies the function denoted by c to 
+  the value denoted by v.
+
+[If we want to use categorical language, 
+we can express the above semantics using the category 
+of finite sets and bijections.]
+
+\end{verbatim}
+
 We review our reversible language {{Pi}}: the presentation in this
@@
 
-The set of types includes the empty type {{0}}, the unit type {{1}}, sum
-types {{b1+b2}}, and products types {{b1*b2}}. The set of values {{v}}
-includes {{()}} which is the only value of type {{1}}, {{left v}} and {{right v}} 
-which inject {{v}} into a sum type, and {{(v1,v2)}} which builds a
-value of product type. There are no values of type {{0}}:
+Unlike the traditional situation with the $\lambda$-calculus, in {{Pi}},
+there is a sharp distinction between data and programs.
+
+\paragraph*{Data.} The sets of values and their types include:
 %subcode{bnf} include main
@@
 % values, v ::= () | left v | right v | (v, v)
-
-The combinators of {{Pi}} are witnesses to the following type
+Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
+and products types {{b1*b2}}.  Values includes {{()}} which is the only value
+of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
+and {{(v1,v2)}} which builds a value of product type. There are no values of
+type {{0}}.
+
+\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
+{{v \rightarrow v}}. Specifically, they are witnesses to the following type
 isomorphisms:
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%
 \subsection{Graphical Language}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%
 \subsection{Semantics}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%
 \subsection{Constructions}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+%%
 \paragraph*{Cloning.}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%
 \subsection{Categorical Structure}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%
 \subsection{Small Step Semantics}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{First class functions from the permutation space.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Traces and the Int-Construction.}
+\section{Fractional Types : {{PiEE*}} }
 \label{sec:int}
 
-For a traced symmetric monoidal category, the Int-construction of
-Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
-constructing a compact closed category. This is the obvious first
-thing to try for a duality in this setting.
-
+In arithmetic the main identity satisfied by fractional numbers is that for
+non-zero {{b}}, we have {{b * 1/b = 1}}. Translating this identity to our
+world requires us to find an isomorphism between the types {{1}} and 
+{{b * 1/b}} (where {{b}} is not equivalent to {{0}}). Formally, we move 
+from the setting of symmetric monoidal categories to that of compact
+closed categories.
+
+%%%%%%%%%%%%%%%%%
+\subsection{The Powerset Monad}
+
+In order to interpret fractional types, we define a monad {{S}} over the
+{{Pi}}-category of finite sets and bijections. The functor part of the monad
+\[\begin{array}{rcl}
+
+\end{array}\]
+
+\paragraph*{Data.} We extend values types with types {{S b}} denoting sets of
+values {{ {| v1, v2, dots |} }}:
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b + b | b * b | S b
+% values, v ::= () | left v | right v | (v, v) | {| v1, v2, dots |}
+
+Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
+and products types {{b1*b2}}.  Values includes {{()}} which is the only value
+of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
+and {{(v1,v2)}} which builds a value of product type. There are no values of
+type {{0}}.
+
+\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
+{{v -> v}}. Specifically, they are witnesses to the following type
+isomorphisms:
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%zeroe :&  0 + b <-> b &: zeroi
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist0 :& 0 * b <-> 0 &: factor0
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+Each line of the above table introduces one or two combinators that witness
+the isomorphism in the middle. Collectively the isomorphisms state that the
+structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
+of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
+multiplication distributes over addition. The isomorphisms are extended to
+form a congruence relation by adding the following constructors that witness
+equivalence and compatible closure:
 %subcode{proof} include main
-%@ c : b1 + b2 <-> b1 + b3
-%@@ trace c : b2 <-> b3
-
-%% \begin{center}
-%%   \includegraphics{diagrams/thesis/trace_plus.pdf}
-%% \end{center}
-
-Show that the int construction does not preserve the {{*}} tensor. We
-need a duality that preserves both tensors.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Fractional Types : {{PiEE*}} }
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+
+\noindent
+To summarize, the syntax of {{Pi}} is given as follows.
+
+\begin{definition}{(Syntax of {{Pi}})}
+\label{def:Pi}
+We collect our types, values, and combinators, to get the full language
+definition.
+%subcode{bnf} include main
+% value types, b ::= 0 | 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% comb.~types, t ::= b <-> b
+% iso ::= zeroe | zeroi 
+%     &|& swap+ | assocl+ | assocr+ 
+%     &|& identl* | identr* 
+%     &|& swap* | assocl* | assocr* 
+%     &|& dist0 | factor0 | dist | factor 
+% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+\end{definition}
+
+%subcode{opsem} include main
+% eta* &: 1 <-> (1/b) * b :& eps*
+
+IntuitivelyThe main challenge Adding fractional types to
+{{Pi}} takes us from a symmetric monoidal category to a compact closed
+category.
 
@@
 % Isomorphisms, iso &=& ... | eta* | eps*
+
+\paragraph*{Traces and the Int-Construction.}
+For a traced symmetric monoidal category, the Int-construction of
+Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
+constructing a compact closed category. This is the obvious first
+thing to try for a duality in this setting.
+
+%subcode{proof} include main
+%@ c : b1 + b2 <-> b1 + b3
+%@@ trace c : b2 <-> b3
+
+%% \begin{center}
+%%   \includegraphics{diagrams/thesis/trace_plus.pdf}
+%% \end{center}
+
+Show that the int construction does not preserve the {{*}} tensor. We
+need a duality that preserves both tensors.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%
 \subsection{Semantics}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%
 \subsection{Categorical Structure}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%
 \subsection{Categorical Constructions}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%
 \subsection{Construction of the Solver}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%
 \subsection{Final Details}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%
 \subsection{Semantics}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%
 \subsection{Constructions with Negatives and Fractionals}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Two Dualities of Computation}
@@
 
+%%%%%%%%%%%%%
 \subsection{Function Spaces}

### 2012-06-26 rational_ext.tex
nothing really

@@
 
-In order to interpret fractional types, we define a monad {{S}} over the
-{{Pi}}-category of finite sets and bijections. The functor part of the monad
-\[\begin{array}{rcl}
-
-\end{array}\]
-
-\paragraph*{Data.} We extend values types with types {{S b}} denoting sets of
-values {{ {| v1, v2, dots |} }}:
+In order to interpret fractional types, we first define a monad {{S}} over
+the {{Pi}}-category of finite sets and bijections.
+
+\paragraph*{Data.} The syntax of {{Pi}} is extended with the type {{S b}}
+which denotes sets {{ {| v1, v2, dots |} }} of values of type {{b}}:
 %subcode{bnf} include main

### 2012-06-26 rational_ext.tex
start of new intro to section 2

@@
 \documentclass[preprint]{sigplanconf}
+
+%% I am getting an error about too many math packages used!!!
+%% I commented the ones we don't seem to be using.
 
@@
 
+%% \newtheorem{theorem}{Theorem}[section]
+%% \newtheorem{lemma}[theorem]{Lemma}
+%% \newtheorem{proposition}[theorem]{Proposition}
+%% \newtheorem{corollary}[theorem]{Corollary}
+
 \newcommand{\xcomment}[2]{\textbf{#1:~\textsl{#2}}}
@@
 \newcommand{\arrow}[1]{\mathtt{#1}}
+\newcommand{\PiL}{\ensuremath{\Pi}}
 
@@
 %! |} = \}
-%! dots = \ldots
 %! ^^^ = ^{\dagger}
@@
 
-\title{Negative and Fractional Types}
-
-\authorinfo{Roshan P. James}
-           {Indiana University}
-           {rpjames@indiana.edu}
-\authorinfo{Zachary Sparks} 
-           {Indiana University}
-           {zasparks@indiana.edu}
-\authorinfo{Jacques Carette} 
-           {McMaster University}
-           {carette@mcmaster.ca}
-\authorinfo{Amr Sabry}
-           {Indiana University}
-           {sabry@indiana.edu}
+\title{Computing in the Field of Rationals}
+
+\authorinfo{Nicolas Bourbaki}
+
+%% \authorinfo{Roshan P. James}
+%%            {Indiana University}
+%%            {rpjames@indiana.edu}
+%% \authorinfo{Amr Sabry}
+%%            {Indiana University}
+%%            {sabry@indiana.edu}
 
@@
   %% to constraint propagation.
-
 \end{abstract}
@@
 
-\begin{verbatim}
-Restructure this section to emphasize the following
-
-In plain Pi: 
-------------
-
-types b ::= 0 | 1 | b+b | b*b
-values v ::= () | L v | R v | (v,v)
-typing v : b
-
-Semantics of values: 
- - a type b is interpreted as a set
- - a value v : b is an element of the set denoted by b
-
-types b <-> b
-combinators c ::= primitiveIso
-               | c + c | c * c | c ; c 
-typing c : b1 <-> b2
-
-Semantics of combinators:
- - types b1 <-> b2 are interpreted as bijections 
-   between the sets denoted by b1 and b2
- - a combinator c : b1 <-> b2 is interpeted as a 
-   bijection between the appropriate sets
-
-Evaluation is function application, i.e., 
-  eval c v applies the function denoted by c to 
-  the value denoted by v.
-
-[If we want to use categorical language, 
-we can express the above semantics using the category 
-of finite sets and bijections.]
-
-\end{verbatim}
-
 We review our reversible language {{Pi}}: the presentation in this
-section differs from the one in our previous paper~\cite{infeffects} in two
+section differs from the one in~\cite{infeffects} in two technical
 aspects. First, we add the empty type {{0}} which is necessary to express the
@@
 semantics, we give a small-step operational semantics that is more
-appropriate for the connections with continuations explored in this paper.
+appropriate for later parts of this paper.
+
+There is also a notable shift in emphasis.  The terms of {{Pi}} are not
+classical values and functions; rather, the terms are isomorphism 
+witnesses.  In other words, the terms of {{Pi}} are proofs that certain
+``shapes of values'' are isomorphic.  And, in classical Curry-Howard
+fashion, our operational semantics shows how these proofs can be
+directly interpreted as actions on ordinary values which effect this
+shape transformation.
+
+Of course, ``shapes of values'' are very familiar already: they are usually
+called \emph{types}.  But usually one designs a type system as a method
+of classifying terms, with the eventual purpose to show that certain
+properties of well-typed terms hold, such as safety.  Our eventual goal
+is different: we start from a type system, and are striving to discover
+a term language which naturally inhabits these types, along with an
+appropriate operational semantics.
+
+Generally, such an approach would be rather vacuous.  What makes this
+pursuit different is that ... (will continue from here -- Jacques)
 
@@
 
-Unlike the traditional situation with the $\lambda$-calculus, in {{Pi}},
-there is a sharp distinction between data and programs.
-
-\paragraph*{Data.} The sets of values and their types include:
+The set of types includes the empty type {{0}}, the unit type {{1}}, sum
+types {{b1+b2}}, and products types {{b1*b2}}. The set of values {{v}}
+includes {{()}} which is the only value of type {{1}}, {{left v}} and {{right v}} 
+which inject {{v}} into a sum type, and {{(v1,v2)}} which builds a
+value of product type. There are no values of type {{0}}:
 %subcode{bnf} include main
@@
 % values, v ::= () | left v | right v | (v, v)
-Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
-and products types {{b1*b2}}.  Values includes {{()}} which is the only value
-of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
-and {{(v1,v2)}} which builds a value of product type. There are no values of
-type {{0}}.
-
-\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
-{{v \rightarrow v}}. Specifically, they are witnesses to the following type
+
+The combinators of {{Pi}} are witnesses to the following type
 isomorphisms: 
@@
 
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Graphical Language}
@@
 
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics}
@@
 
-%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions}
@@
 
-%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \paragraph*{Cloning.}
@@
 
-%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Categorical Structure}
@@
 
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Small Step Semantics}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{First class functions from the permutation space.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Traces and the Int-Construction.}
+\label{sec:int}
+
+For a traced symmetric monoidal category, the Int-construction of
+Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
+constructing a compact closed category. This is the obvious first
+thing to try for a duality in this setting.
+
+%subcode{proof} include main
+%@ c : b1 + b2 <-> b1 + b3
+%@@ trace c : b2 <-> b3
+
+%% \begin{center}
+%%   \includegraphics{diagrams/thesis/trace_plus.pdf}
+%% \end{center}
+
+Show that the int construction does not preserve the {{*}} tensor. We
+need a duality that preserves both tensors.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Fractional Types : {{PiEE*}} }
-\label{sec:int}
-
-In arithmetic the main identity satisfied by fractional numbers is that for
-non-zero {{b}}, we have {{b * 1/b = 1}}. Translating this identity to our
-world requires us to find an isomorphism between the types {{1}} and 
-{{b * 1/b}} (where {{b}} is not equivalent to {{0}}). Formally, we move 
-from the setting of symmetric monoidal categories to that of compact
-closed categories.
-
-%%%%%%%%%%%%%%%%%
-\subsection{The Powerset Monad}
-
-In order to interpret fractional types, we first define a monad {{S}} over
-the {{Pi}}-category of finite sets and bijections.
-
-\paragraph*{Data.} The syntax of {{Pi}} is extended with the type {{S b}}
-which denotes sets {{ {| v1, v2, dots |} }} of values of type {{b}}:
-%subcode{bnf} include main
-% value types, b ::= 0 | 1 | b + b | b * b | S b
-% values, v ::= () | left v | right v | (v, v) | {| v1, v2, dots |}
-
-Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
-and products types {{b1*b2}}.  Values includes {{()}} which is the only value
-of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
-and {{(v1,v2)}} which builds a value of product type. There are no values of
-type {{0}}.
-
-\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
-{{v -> v}}. Specifically, they are witnesses to the following type
-isomorphisms:
-%subcode{bnf} include main
-%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
-%zeroe :&  0 + b <-> b &: zeroi
-%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
-%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
-%identl* :&  1 * b <-> b &: identr*
-%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
-%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
-%dist0 :& 0 * b <-> 0 &: factor0
-%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-Each line of the above table introduces one or two combinators that witness
-the isomorphism in the middle. Collectively the isomorphisms state that the
-structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
-of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
-multiplication distributes over addition. The isomorphisms are extended to
-form a congruence relation by adding the following constructors that witness
-equivalence and compatible closure:
-%subcode{proof} include main
-%@  ~
-%@@ id : b <-> b 
-%
-%@ c : b1 <-> b2
-%@@ sym c : b2 <-> b1
-%
-%@ c1 : b1 <-> b2
-%@ c2 : b2 <-> b3
-%@@ c1(;)c2 : b1 <-> b3
-%---
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
-%
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
-
-\noindent
-To summarize, the syntax of {{Pi}} is given as follows.
-
-\begin{definition}{(Syntax of {{Pi}})}
-\label{def:Pi}
-We collect our types, values, and combinators, to get the full language
-definition.
-%subcode{bnf} include main
-% value types, b ::= 0 | 1 | b+b | b*b 
-% values, v ::= () | left v | right v | (v,v) 
-%
-% comb.~types, t ::= b <-> b
-% iso ::= zeroe | zeroi 
-%     &|& swap+ | assocl+ | assocr+ 
-%     &|& identl* | identr* 
-%     &|& swap* | assocl* | assocr* 
-%     &|& dist0 | factor0 | dist | factor 
-% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-\end{definition}
-
-%subcode{opsem} include main
-% eta* &: 1 <-> (1/b) * b :& eps*
-
-IntuitivelyThe main challenge Adding fractional types to
-{{Pi}} takes us from a symmetric monoidal category to a compact closed
-category.
 
@@
 % Isomorphisms, iso &=& ... | eta* | eps*
-
-\paragraph*{Traces and the Int-Construction.}
-For a traced symmetric monoidal category, the Int-construction of
-Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
-constructing a compact closed category. This is the obvious first
-thing to try for a duality in this setting.
-
-%subcode{proof} include main
-%@ c : b1 + b2 <-> b1 + b3
-%@@ trace c : b2 <-> b3
-
-%% \begin{center}
-%%   \includegraphics{diagrams/thesis/trace_plus.pdf}
-%% \end{center}
-
-Show that the int construction does not preserve the {{*}} tensor. We
-need a duality that preserves both tensors.
 
@@
 
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics}
@@
 
-%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Categorical Structure}
@@
 
-%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Categorical Constructions}
@@
 
-%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Construction of the Solver}
@@
 
-%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Final Details}
@@
 
-%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Semantics}
@@
 
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructions with Negatives and Fractionals}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Two Dualities of Computation}
@@
 
-%%%%%%%%%%%%%
 \subsection{Function Spaces}

### 2012-06-26 rational_ext.tex

--- 2012-06-26 rational_ext.tex
@@
 We review our reversible language {{Pi}}: the presentation in this
-section differs from the one in our previous paper~\cite{infeffects} in two
+section differs from the one in~\cite{infeffects} in two technical
 aspects. First, we add the empty type {{0}} which is necessary to express the

### 2012-06-27 rational_ext.tex
constructions

@@
 
-\paragraph*{Booleans and Conditionals.} 
-Given any combinator {{c : b <-> b}} we can construct a combinator called
-{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
-one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
-the combinator {{c}} is used to transform the value of type~{{b}}. If the
-boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
-write down the combinator for {{if_c}} in terms of {{c}} as 
-{{ dist (;) ((id (*) c) (+) id) (;) factor }}.
-
-\noindent The diagram below shows the input value of type {{(1+1)*b}}
-processed by the distribute operator {{dist}}, which converts it into a value
-of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
-case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
-combinator~{{c}} is applied to the value of type~{{b}}. The right branch
-which corresponds to the boolean being {{false}} passes along the value of
-type {{b}} unchanged.
-
-\begin{center}
-\scalebox{1.0}{
-%%subcode-line{pdfimage}[diagrams/if_c.pdf]
-\includegraphics{diagrams/thesis/cnot.pdf}
-}
-\end{center}
-
-% We will be seeing many more such wiring diagrams in this paper and it is
-% useful to note some conventions about them. Wires indicate a
-% value that can exist in the program. Each wire, whenever possible, is
-% annotated with its type and sometimes additional information to help clarify
-% its role. When multiple wires run in parallel, it means that those values
-% exist in the system at the same time, indicating pair types. When there is a
-% disjunction, we put a {{+}} between the wires. 
-% Combinators for distribution {{dist}} and factoring {{factor}}
-% are represented as triangles with their operator symbols in them. Other
-% triangles may be used and, in each case, types or labels will be used to
-% clarify their roles. Finally, we don't draw boxes for combinators such as
-% {{id}}, commutativity, and associativity, but instead just shuffle the wires
-% as appropriate.
-
-The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
-negates its second argument if the first argument is {{true}}. This
-gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
-equivalent construction that is useful is {{else_{not} }} where we
-negate the second argument only if the first is {{false}}.
-
-Similarly, we can iterate the construction of {{if_c}} to check several
-bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
-checks two booleans and negates the result wire only if they are both
-{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
-universal reversible gate. We can generalize this construction to
-{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
-they are all {{true}}.
-
-%%
-\paragraph*{Cloning.}
-Although cloning is generally not allowed in reversible languages, it is
-possible at the cost of having additional constant inputs. For example,
-consider the gate {{else_{not} }}. Generally, the gate maps
-{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
-on the cases in which the second input is {{true}}, we get that the gate maps
-{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
-i.e., the gate clones the first input. A circuit of {{n}} parallel
-{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
-{{true}} inputs in the process.  Let us call this construction
-{{clone^n_{bool} }}.
+The constructions in this section provide a brief overview to writing
+programs in {{Pi}}. More details maybe found in previous work
+(Sec. 3.3 \cite{infeffects} and the constructions in \cite{rc2012}).
+
+\paragraph*{Booleans, Conditionals and Cloning.} 
+We use the type {{1+1}} to denote {{bool}}, with {{left ()}} as
+{{true}} and {{right ()}} as {{false}}. The combinator
+{{not:bool<->bool}} can be expressed by {{swap+}} at the type
+{{bool}}. Further, given any two combinators {{c1:b1<->b2}} and
+{{c2:b1<->b2}}, we can write the conditional combinator
+{{if_{c1,c2}:bool*b1<->bool*b2}} textually as 
+{{dist(;)((id(*)c1)(+)(id(*)c2))(;)factor}} and graphically as:
+
+\dgm{if-c1-c2}
+
+\noindent
+If {{c1 v |--> v'}} and {{c2 v |--> v''}}, we can verify that the
+conditional maps {{if_{c1,c2} (true, v) |--> (true, v')}} and
+{{if_{c1,c2} (false, v) |--> (false, v'')}}.  Several useful
+combinators follow from this basic construction:
+
+\begin{enumerate}
+\item The combinator {{cnot:bool*bool<->bool*bool}} that negates the
+  second input if the first input (called the control wire) is
+  {{true}} may be defined as {{if_{not,id} }}.
+\item The universal reversible combinator, the Toffoli gate, may be
+  defined as
+  {{if_{cnot,id}:bool*(bool*bool)<->bool*(bool*bool)}}. This can be
+  extended to test {{n}} control wires and we call this combinator
+  {{cnot^n:bool^{n+1}<->bool^{n+1} }}.
+\item The combinator {{if_{id,not}:bool*bool<->bool*bool}} has the
+  property that it maps {{(x, true)|-->(x, x)}} i.e. if the second bit
+  is fixed to {{true}} it clones the first bit. We can generalize this
+  to clone {{n}} bits and we write
+  {{clone^n:bool^n*bool^n<->bool^n*bool^n}}. The second argument
+  {{bool^n}} must be supplied {{true}} values to correctly simulate
+  cloning.
+\end{enumerate}
+
+%% Given any combinator {{c : b <-> b}} we can construct a combinator called
+%% {{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
+%% one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
+%% the combinator {{c}} is used to transform the value of type~{{b}}. If the
+%% boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
+%% write down the combinator for {{if_c}} in terms of {{c}} as 
+%% {{ dist (;) ((id (*) c) (+) id) (;) factor }}.
+
+%% \noindent The diagram below shows the input value of type {{(1+1)*b}}
+%% processed by the distribute operator {{dist}}, which converts it into a value
+%% of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
+%% case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
+%% combinator~{{c}} is applied to the value of type~{{b}}. The right branch
+%% which corresponds to the boolean being {{false}} passes along the value of
+%% type {{b}} unchanged.
+
+%% \begin{center}
+%% \scalebox{1.0}{
+%% %%subcode-line{pdfimage}[diagrams/if_c.pdf]
+%% \includegraphics{diagrams/thesis/cnot.pdf}
+%% }
+%% \end{center}
+
+%% % We will be seeing many more such wiring diagrams in this paper and it is
+%% % useful to note some conventions about them. Wires indicate a
+%% % value that can exist in the program. Each wire, whenever possible, is
+%% % annotated with its type and sometimes additional information to help clarify
+%% % its role. When multiple wires run in parallel, it means that those values
+%% % exist in the system at the same time, indicating pair types. When there is a
+%% % disjunction, we put a {{+}} between the wires. 
+%% % Combinators for distribution {{dist}} and factoring {{factor}}
+%% % are represented as triangles with their operator symbols in them. Other
+%% % triangles may be used and, in each case, types or labels will be used to
+%% % clarify their roles. Finally, we don't draw boxes for combinators such as
+%% % {{id}}, commutativity, and associativity, but instead just shuffle the wires
+%% % as appropriate.
+
+%% The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
+%% negates its second argument if the first argument is {{true}}. This
+%% gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
+%% equivalent construction that is useful is {{else_{not} }} where we
+%% negate the second argument only if the first is {{false}}.
+
+%% Similarly, we can iterate the construction of {{if_c}} to check several
+%% bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
+%% checks two booleans and negates the result wire only if they are both
+%% {{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
+%% universal reversible gate. We can generalize this construction to
+%% {{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
+%% they are all {{true}}.
+
+%% %%
+%% \paragraph*{Cloning.}
+%% Although cloning is generally not allowed in reversible languages, it is
+%% possible at the cost of having additional constant inputs. For example,
+%% consider the gate {{else_{not} }}. Generally, the gate maps
+%% {{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
+%% on the cases in which the second input is {{true}}, we get that the gate maps
+%% {{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
+%% i.e., the gate clones the first input. A circuit of {{n}} parallel
+%% {{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
+%% {{true}} inputs in the process.  Let us call this construction
+%% {{clone^n_{bool} }}.
 

### 2012-06-27 code/dual/Test.hs

@@ -42,7 +42,15 @@ z_shape2Times = UnitI
                 :.: (Id :*: (CommuteTimes :.: EpsTimes))
                 :.: CommuteTimes
                 :.: UnitE
-               
+              
+type SBool = Either () () 
+
+ifc1c2 :: b :<=> b 
+          -> b :<=> b
+          -> (SBool :*: b) :<=> (SBool :*: b)
+ifc1c2 c1 c2 = Distribute
+               :.: ((Id :*: c1) :+: (Id :*: c2))
+               :.: Factor
 
 ----------------------------------------------------------------
 

### 2012-06-27 rational_ext.tex
dgm
_

@@
 \subsection{Constructions}
+\label{sec:pi-constructions}
 
@@
 
-\dgm{if-c1-c2}
+% \dgm{if-c1-c2}
 
@@
 
-\roshan{Maybe replace the annihilation example with union and
-  difference.}
+%% \roshan{Maybe replace the annihilation example with union and
+%%   difference.}
 
@@
 
-\roshan{This probably needs rigorous proof.}
-
-\roshan{Should we show completeness -- i.e. that every relation is
-  expressible? We can. But should we?}
+% \roshan{This probably needs rigorous proof.}
+
+\paragraph*{Union.}
+The union of the two relations {{c1:b1<->b2}} and {{c2:b1<->b2}} is
+expressed by {{trace* if_{c1,c2} :b1<->b2}} (see Sec.
+\ref{sec:pi-constructions} for {{if_{c1,c2} }} and
+Sec. \ref{sec:frac-constructions} for {{trace*}}). Intersection can be
+expressed by ``controlled-annihilation'' detailed in
+Sec. \ref{sec:sat-solver}.
+
+%% \roshan{Should we show completeness -- i.e. that every relation is
+%%   expressible? We can. But should we?}
 
@@
 \subsection{Categorical Constructions}
+\label{sec:frac-constructions}
 
@@
 \section{SAT Solver }
-\label{sec:prog}
 \label{sec:sat-solver}
@@
 Any boolean expression {{f : bool^n -> bool}} can be compiled into the
-isomorphism {{iso_f : bool^h*bool^n<->bool^g*bool}} where the extra bits
-{{bool^h}} and {{bool^g}} are considered as heap and garbage. Constructing
-such an isomorphism has been detailed
-before~\cite{Toffoli:1980,infeffects}. The important relation to note is that
-applying {{iso_f}} to some special heap values and an input \textit{bs}
-produces some bits that can be ignored and the same output that {{f}} would
-have produced on {{bs}}. We can ensure that the heap has the appropriate
-initial values by checking the heap and negating a second control wire, if
-the values do not match, i.e., the dotted part in the diagram below.
+isomorphism {{iso_f : bool^h*bool^n<->bool^g*bool}} where the extra
+bits {{bool^h}} and {{bool^g}} are considered as heap and
+garbage. Constructing such an isomorphism has been detailed
+before~\cite{Toffoli:1980,infeffects}. The important relation to note
+is that applying {{iso_f}} to some special heap values and an input
+\textit{bs} produces some bits that can be ignored and the same output
+that {{f}} would have produced on {{bs}}. We can ensure that the heap
+has the appropriate initial values by checking the heap and negating a
+second control wire, if the values do not match using, i.e., the
+dotted part in the diagram below. This is achieved by first applying
+{{not}} to the control wire and then testing that all the heap values
+are true using {{cnot^n}}, undoing the {{not}} only if they are.
 
@@
 0 or 1 solutions. In the case of general SAT, the solver will produce
-solutions as determined by the semantics of the top-level interaction
-(see discussion in Sec. \ref{sec:rat}).
+all satisfying solution.
 

### 2012-06-27 code/dual/Test.hs

@@ -45,12 +45,28 @@ z_shape2Times = UnitI
               
 type SBool = Either () () 
 
-ifc1c2 :: b :<=> b 
-          -> b :<=> b
-          -> (SBool :*: b) :<=> (SBool :*: b)
+ifc1c2 :: b1 :<=> b2 
+          -> b1 :<=> b2
+          -> (SBool :*: b1) :<=> (SBool :*: b2)
 ifc1c2 c1 c2 = Distribute
                :.: ((Id :*: c1) :+: (Id :*: c2))
                :.: Factor
+
+
+unionc1c2 :: b1 :<=> b2 -> b1 :<=> b2 -> b1 :<=> b2
+unionc1c2 c1 c2 = UnitI 
+                  :.: (EtaTimes :*: Id)
+                  :.: AssocTimesR
+                  :.: (Id :*: (ifc1c2 c1 c2))
+                  :.: AssocTimesL
+                  :.: (EpsTimes :*: Id)
+                  :.: UnitE
+
+union c1 c2 = traceTimes (ifc1c2 c1 c2)
+
+notSBool :: SBool :<=> SBool
+notSBool = CommutePlus
+
 
 ----------------------------------------------------------------
 
@@ -292,3 +308,5 @@ makeDC c = (c :*: Id) :.: CommuteTimes :.:  EpsTimes
 
 
 
+-- dualPlus :: Neg (a :+: b) :<=> (Neg a :+: Neg b)
+-- dualPlus = 

### 2012-06-27 rational_ext.tex
_
started on negatives

@@
 
-% \dgm{if-c1-c2}
+\dgm{if-c1-c2}
 
@@
 
-%%%%%%%%%
-\subsection{Categorical Constructions}
-\label{sec:frac-constructions}
-
-All the constructions below are standard: they are collected from Selinger's
-survey paper on monoidal
-categories~\cite{springerlink:10.1007/978-3-642-12821-94} and presented in
-the context of our language.
-
-We now review several interesting constructions related to looping,
-involution, and higher order functions.
-
-\paragraph*{Trace.}
-Every compact closed category admits a trace. For the additive case, we get
-the following definition.  Given {{f : b1+b2 <-> b1+b3 }}, 
-define {{trace+ f : b2 <-> b3}} as:
-
-{{ trace+ f = zeroi (;) (id (+) eta+) (;) (f (+) id) (;) (id (+) eps+) (;) zeroe }}
-
-\begin{center}
-  \includegraphics{diagrams/thesis/trace_plus.pdf}
-\end{center}
-
-\noindent We have omitted some of the commutativity and associativity
-shuffling to communicate the main idea. We are given a value of type
-{{b2}} which we embed into {{0+b2}} and then {{(-b1+b1)+b2)}}. This
-can be re-associated into {{-b1+(b1+b2)}}. The component {{b1+b2}},
-which until now is just an appropriately tagged value of type {{b2}},
-is transformed to a value of type {{b1+b3}} by {{f}}. 
-If the result is in the {{b3}}-summand, it is produced as
-the answer; otherwise the result is in the {{b1}}-summand; {{eps+}} is
-used to make it flow backwards to be fed to the {{eta+}} located at
-the beginning of the sequence. Iteration continues until a {{b3}} is
-produced.
-
-%%%%%
-\paragraph*{Involution (Principium Contradictiones)}
-
-In a symmetric compact closed category, we can build isomorphisms that the
-dual operation is an involution. Specifically, we get the isomorphisms 
-{{b <-> -(-b)}} and {{b <-> (1/(1/b))}}. For the additive case, the 
-isomorphism is defined as follows:
-
-{{ (id (+) eta+) (;) (swap+ (+) id) (;) (id (+) eps+) }}
-
-\noindent where we have omitted the 0 introduction and elimination. The idea
-is as follows: we start with a value of type {{b}}, embed it into {{b+0}} and
-use {{eta}} to create something of type {{b + (-(-b) + (-b))}}. This is possible
-because {{eta}} has the polymorphic type {{-a + a}} which can be instantiated
-to {{-b}}. We then reshuffle the type to produce {{-(-b) + (-b + b)}} and cancel
-the right hand side using {{eps+}}.  The construction for the multiplicative 
-case is identical and omitted.
-
-%% {{b <-> -(-b)}}
-%% 
-%% this is the wrong diagram: see lemma 4.17 in selinger's paper
-%% \begin{center}
-%%   \includegraphics{diagrams/double_neg.pdf}
-%% \end{center}
-
-\paragraph*{Duality preserves the monoidal tensor. }
-As with compact closed categories, the dual on the objects distributes
-over the tensor. In terms of {{PiEE}} we have that {{-(b1+b2)}}
-can be mapped to {{(-b1)+(-b2)}} and that {{1/(b1*b2)}} can be mapped
-to {{(1/b1)*(1/b2)}}. The isomorphism {{-(b1 + b2) <-> (-b1) + (-b2)}}
-can be realized as follows:
-\begin{center}
-  \includegraphics{diagrams/dist_neg_plus.pdf}
-\end{center}
-
-\noindent
-The multiplicative construction is similar.
-
-%% This is a distinguishing difference from *-autonomous categories that
-%% are models for Linear Logic, where the dual does not preserve the
-%% tensor but maps to a dual tensor \cite{curien2009interactive}.
-
-\paragraph*{Duality is a functor.}
-Duality in {{PiEE}} can map objects to their duals and morphisms
-to act on dual objects. In other {{c : b1 <-> b2}} to
-{{neg~c:-b1<->-b2}} in the additive monoid and to
-{{inv~c:1/b1<->1/b2}} in the multiplicative monoid. 
-%% Any operation on types can be applied to the negative versions of these
-%% types. 
-The idea is simply to reverse the flow of values and use the
-adjoint of the operation:
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/neg_lift.pdf}
-\end{center}
-
-%subcode{proof} include main
-%@ c : b1 <-> b2
-%@@ neg c : (-b1) <-> (-b2)
-\end{multicols}
-
-This construction relies on the fact that every {{PiEE}} morphism
-has an adjoint.  The {{inv}} construction is similar.
-
-%% inverse morphism, whereas in compact closed categories with no
-%% underlying dagger structure we can only construct an op-functor.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We illustrate the expressive power of first-class constraints represented by
-fractional types. To understand the intuition, recall the definition of
-{{trace}} for the multiplicative fragment of {{PiEE}}. 
+As we will see if Sec. \ref{sec:neg-constructions}, every compact
+closed category admits a {{trace}} operator. In the case of {{PiEE*}},
+the operator {{trace*}} is defined only on the non-zero fragment.
 Given {{f : a*c <-> b*c }}, we have {{trace* f : a <-> b}}:
@@
 using {{trace*}}. We illustrate the main ideas with the implementation
-of a SAT-solver.  We proceed in small steps, reviewing some of the
-necessary constructions presented in our earlier
-paper~\cite{infeffects}.
+of a SAT-solver.
+
+%% We proceed in small steps, reviewing some of the
+%% necessary constructions presented in our earlier
+%% paper~\cite{infeffects}.
 
@@
 
+\begin{comment}
   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\end{comment}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{definition}[Size of a type]
+The defintion of size is extended to include negative types.
+{{[^-b^] = -[^b^]}}
+\end{definition}
+
 %%%%%%%%%
 \subsection{Semantics}
+
+\begin{comment}
 
@@
 
+\end{comment}
+
 Given that our language is reversible (Prop.~\ref{prop:logrev}), a backward
@@
 %! columnStyle = rclr
-%! bck = \triangleleft
-% {[iso, v, C, s]}_{bck} &|-->& <iso, v', C, s'>_{bck}
-% & & where iso^{dagger} s v |--> (v', s')
-% <c1, v, Fst C c2, s>_{bck} &|-->& <c1(;)c2, v, C, s>_{bck}
-% <c2, v, Snd c1 C, s>_{bck} &|-->& {[c1, v, Fst C c2, s]}_{bck}
-% {[ c1(;)c2, v, C, s ]}_{bck} &|-->& {[c2, v, Snd c1 C, s]}_{bck}
-% <c1, v, LeftP C c2,s >_{bck} &|-->& <c1(+)c2,left v, C>_{bck}
-% {[c1 (+) c2, v', C, s ]}_{bck} &|-->& {[ c1, v, LeftP C c2, s[v' <><> left v] ]}_{bck}
-% <c2, v, RightP c1 C, s>_{bck} &|-->& <c1(+)c2, right v, C>_{bck}
-% {[c1 (+) c2, v', C,s ]}_{bck} &|-->& {[ c2, v, RightP c1 C, s[v' <><> right v] ]}_{bck}
-% <c1, v1, LeftT C c2 v2, s>_{bck} &|-->& <c1(*)c2, (v1, v2), C, s>_{bck}
-% <c2, v2, RightT c1 v1 C, s>_{bck} &|-->& {[ c1, v1, LeftT C c2 v2, s ]}_{bck}
-% {[ c1 (*) c2, v, C, s ]}_{bck} &|-->& {[ c2, v2, RightT c1 v1 C, S' ]}_{bck}
-% && where s' = s[v <><> (v1, v2)]
+% {[iso, v, C]}^^^ &|-->& <iso, v', C>^^^
+% & & where iso^{dagger} v |--> v'
+% <c1, v, Fst C c2>^^^ &|-->& <c1(;)c2, v, C>^^^
+% <c2, v, Snd c1 C>^^^ &|-->& {[c1, v, Fst C c2]}^^^
+% {[ c1(;)c2, v, C]}^^^ &|-->& {[c2, v, Snd c1 C]}^^^
+% <c1, v, LeftP C c2>^^^ &|-->& <c1(+)c2,left v, C>^^^
+% {[c1 (+) c2, left v, C]}^^^ &|-->& {[ c1, v, LeftP C c2 ]}^^^
+% <c2, v, RightP c1 C, s>^^^ &|-->& <c1(+)c2, right v, C>^^^
+% {[c1 (+) c2, right v, C ]}^^^ &|-->& {[ c2, v, RightP c1 C ]}^^^
+% <c1, v1, LeftT C c2 v2>^^^ &|-->& <c1(*)c2, (v1, v2), C>^^^
+% <c2, v2, RightT c1 v1 C >^^^ &|-->& {[ c1, v1, LeftT C c2 v2 ]}^^^
+% {[ c1 (*) c2, (v1, v2), C ]}^^^ &|-->& {[ c2, v2, RightT c1 v1 C ]}^^^
 
@@
 %! bck = \triangleleft
-% <eps+, v, C, s>_{fwd} |--> <eps+, left (-v'), C, s[v <><> right v']>_{bck}
-% <eps+, v, C, s>_{fwd} |--> <eps+, right v', C, s[v <><> left (-v')]>_{bck}
+% <eps+, right v, C, s> &|-->& <eps+, left (-v), C]>^^^
+% <eps+, left (-v), C, s> &|-->& <eps+, right v, C>^^^
 
@@
 %! bck = \triangleleft
-% {[eta+, v, C, s]}_{bck} |--> {[eta+, left (-v'), C, s[v <><> right v'] ]}_{fwd}
-% {[eta+, v, C, s]}_{bck} |--> {[eta+, right v', C, s[v <><> left (-v')] ]}_{fwd}
+% {[eta+, right v, C, s]}^^^ &|-->& {[eta+, left (-v), C ]}
+% {[eta+, left (-v), C, s]}^^^ &|-->& {[eta+, right v, C ]}
 
@@
 \end{proposition}
+
+%%%%%%%%%
+\subsection{Categorical Constructions}
+\label{sec:neg-constructions}
+
+\
+All the constructions below are standard: they are collected from Selinger's
+survey paper on monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94} and presented in
+the context of our language.
+
+We now review several interesting constructions related to looping,
+involution, and higher order functions.
+
+\paragraph*{Trace.}
+Every compact closed category admits a trace. For the additive case, we get
+the following definition.  Given {{f : b1+b2 <-> b1+b3 }}, 
+define {{trace+ f : b2 <-> b3}} as:
+
+{{ trace+ f = zeroi (;) (id (+) eta+) (;) (f (+) id) (;) (id (+) eps+) (;) zeroe }}
+
+\begin{center}
+  \includegraphics{diagrams/thesis/trace_plus.pdf}
+\end{center}
+
+\noindent We have omitted some of the commutativity and associativity
+shuffling to communicate the main idea. We are given a value of type
+{{b2}} which we embed into {{0+b2}} and then {{(-b1+b1)+b2)}}. This
+can be re-associated into {{-b1+(b1+b2)}}. The component {{b1+b2}},
+which until now is just an appropriately tagged value of type {{b2}},
+is transformed to a value of type {{b1+b3}} by {{f}}. 
+If the result is in the {{b3}}-summand, it is produced as
+the answer; otherwise the result is in the {{b1}}-summand; {{eps+}} is
+used to make it flow backwards to be fed to the {{eta+}} located at
+the beginning of the sequence. Iteration continues until a {{b3}} is
+produced.
+
+%%%%%
+\paragraph*{Involution (Principium Contradictiones)}
+
+In a symmetric compact closed category, we can build isomorphisms that the
+dual operation is an involution. Specifically, we get the isomorphisms 
+{{b <-> -(-b)}} and {{b <-> (1/(1/b))}}. For the additive case, the 
+isomorphism is defined as follows:
+
+{{ (id (+) eta+) (;) (swap+ (+) id) (;) (id (+) eps+) }}
+
+\noindent where we have omitted the 0 introduction and elimination. The idea
+is as follows: we start with a value of type {{b}}, embed it into {{b+0}} and
+use {{eta}} to create something of type {{b + (-(-b) + (-b))}}. This is possible
+because {{eta}} has the polymorphic type {{-a + a}} which can be instantiated
+to {{-b}}. We then reshuffle the type to produce {{-(-b) + (-b + b)}} and cancel
+the right hand side using {{eps+}}.  The construction for the multiplicative 
+case is identical and omitted.
+
+%% {{b <-> -(-b)}}
+%% 
+%% this is the wrong diagram: see lemma 4.17 in selinger's paper
+%% \begin{center}
+%%   \includegraphics{diagrams/double_neg.pdf}
+%% \end{center}
+
+\paragraph*{Duality preserves the monoidal tensor. }
+As with compact closed categories, the dual on the objects distributes
+over the tensor. In terms of {{PiEE}} we have that {{-(b1+b2)}}
+can be mapped to {{(-b1)+(-b2)}} and that {{1/(b1*b2)}} can be mapped
+to {{(1/b1)*(1/b2)}}. The isomorphism {{-(b1 + b2) <-> (-b1) + (-b2)}}
+can be realized as follows:
+\begin{center}
+  \includegraphics{diagrams/dist_neg_plus.pdf}
+\end{center}
+
+\noindent
+The multiplicative construction is similar.
+
+%% This is a distinguishing difference from *-autonomous categories that
+%% are models for Linear Logic, where the dual does not preserve the
+%% tensor but maps to a dual tensor \cite{curien2009interactive}.
+
+\paragraph*{Duality is a functor.}
+Duality in {{PiEE}} can map objects to their duals and morphisms
+to act on dual objects. In other {{c : b1 <-> b2}} to
+{{neg~c:-b1<->-b2}} in the additive monoid and to
+{{inv~c:1/b1<->1/b2}} in the multiplicative monoid. 
+%% Any operation on types can be applied to the negative versions of these
+%% types. 
+The idea is simply to reverse the flow of values and use the
+adjoint of the operation:
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/neg_lift.pdf}
+\end{center}
+
+%subcode{proof} include main
+%@ c : b1 <-> b2
+%@@ neg c : (-b1) <-> (-b2)
+\end{multicols}
+
+This construction relies on the fact that every {{PiEE}} morphism
+has an adjoint.  The {{inv}} construction is similar.
+
+%% inverse morphism, whereas in compact closed categories with no
+%% underlying dagger structure we can only construct an op-functor.
 

### 2012-06-27 rational_ext.tex
vector spaces

@@
 
+\renewcommand{\vec}[2]{\begin{bmatrix} #1 \\ #2 \end{bmatrix}}
+\newcommand{\Fpx}[1]{\mathbf{F}_{#1}}
+\newcommand{\modulusp}[1]{\left | #1 \right |}
 \newcommand{\fname}[1]{\ulcorner #1 \urcorner}
@@
 %%%%%%%%%%%%%%%%%
-\subsection{The Powerset Monad}
-
-In order to interpret fractional types, we first define a monad {{S}} over
-the {{Pi}}-category of finite sets and bijections.
+\subsection{Vector Spaces}
+
+In order to interpret fractional types, we first lift the values and
+combinators of {{Pi}} to the world of vector spaces. Intuitively:
+\begin{itemize}
+\item each type {{b}} is now interpreted as a vector space of dimension
+  $\modulusp{b}$ over the field $\Fpx{2}$; the elements of such a vector
+  space are column vectors
+\item each value {{v_i : b}} is interpreted as column vector:
+\[\begin{array}{r@{~}l}
+\begin{array}{r} 
+v_0 \\ v_1 \\ \vdots \\ v_i \\ \vdots \\ v_n 
+\end{array} & 
+\begin{bmatrix}
+0 \\ 0 \\ \vdots \\ 1 \\ \vdots \\ 0
+\end{bmatrix}
+\end{array}\]
+\item each combinator {{c : b1 <-> b2}} is interpreted as a matrix whose
+  columns are indexed by the values of type {{b1}} and whose rows are indexed
+  by the values of type {{b2}}. The matrix is \emph{orthogonal} with each row
+  and column containing exactly one occurrence of the value 1.
+\end{itemize}
+
+Once we are in the world of vector spaces, the fractional type {{1/b}} is
+interpreted as the \emph{dual} to the vector space denoted by {{b}}. The
+elements of this dual vector space can be represented as \emph{row vectors}.
+
+Multiplying a row vector by a column gives a scalar: this is eta and epsilon.
+
+\newpage
 

### 2012-06-27 rational_ext.tex

@@
 
+\newcommand{\vecspace}[1]{\mathbb{V}_{#1}}
 \renewcommand{\vec}[2]{\begin{bmatrix} #1 \\ #2 \end{bmatrix}}
-\newcommand{\Fpx}[1]{\mathbf{F}_{#1}}
+\newcommand{\Fpx}[1]{\mathbb{F}_{#1}}
 \newcommand{\modulusp}[1]{\left | #1 \right |}
@@
 \begin{itemize}
-\item each type {{b}} is now interpreted as a vector space of dimension
-  $\modulusp{b}$ over the field $\Fpx{2}$; the elements of such a vector
-  space are column vectors
-\item each value {{v_i : b}} is interpreted as column vector:
+\item instead of interpreting a type {{b}} as a set of cardinality
+  $\modulusp{b}$, we now interpret each type {{b}} as a vector space
+  $\vecspace{b}$ of dimension $\modulusp{b}$ over the field of booleans
+  $\Fpx{2}$;
+\item the elements of such a vector space are column vectors whose elements
+  are 0 or 1 and which indexed by the values of type {{b}}, i.e., 
+  a value {{v_i : b}} is interpreted as the vector:
 \[\begin{array}{r@{~}l}
@@
 
-Once we are in the world of vector spaces, the fractional type {{1/b}} is
-interpreted as the \emph{dual} to the vector space denoted by {{b}}. The
-elements of this dual vector space can be represented as \emph{row vectors}.
-
-Multiplying a row vector by a column gives a scalar: this is eta and epsilon.
+Now that the types are interpreted as vector spaces, it is possible to add a
+new type {{1/b}} which corresponds to the \emph{dual} of the vector space
+denoting {{b}}. The elements of this dual vector space are \emph{linear
+  functionals} mapping vectors in $V_b$ to a scalar (0 or 1). The isomorphism
+between the types {{1}} and {{b * 1/b}} is now a consequence of that fact
+that the pairing of a functional in the dual space $\vecspace{1/b}$ and a
+vector in the space $\vecspace{b}$ produces a scalar.
+
+The alert co-author should have noticed that using $\Fpx{2}$ as the underlying
+field is actually nonsense. To get things to work properly we need a field of
+one element, which apparently exists and has some nice properties.
 

### 2012-06-27 rational_ext.tex
pointed sets

--- 2012-06-27 rational_ext.tex
@@
 
+\newcommand{\vecspace}[1]{\mathbb{V}_{#1}}
+\renewcommand{\vec}[2]{\begin{bmatrix} #1 \\ #2 \end{bmatrix}}
+\newcommand{\Fpx}[1]{\mathbb{F}_{#1}}
+\newcommand{\modulusp}[1]{\left | #1 \right |}
 \newcommand{\fname}[1]{\ulcorner #1 \urcorner}
@@
 %%%%%%%%%%%%%%%%%
-\subsection{The Powerset Monad}
-
-In order to interpret fractional types, we first define a monad {{S}} over
-the {{Pi}}-category of finite sets and bijections.
+\subsection{Pointed Sets}
+
+In order to interpret fractional types, we first lift the values and
+combinators of {{Pi}} to the world of \emph{pointed sets}. Intuitively:
+\begin{itemize}
+\item instead of interpreting a type {{b}} as a plain set of cardinality
+  $\modulusp{b}$, we now interpret each type {{b}} as a pointed set $P_b$
+  with a distinguished \emph{basepoint} $\bot_b$;
+\item a combinator {{c : b1 <-> b2}} is interpreted as a function as before
+  with the restriction that it must map the basepoint $\bot_{b_1}$ to the
+  basepoint $\bot_{b_2}$.
+\end{itemize}
+
+Intuitively a pointed set may be seen as a vector space over the field with
+one element. This view allows us to add the type {{1/b}} and interpret it as
+the \emph{dual} of the vector space denoting {{b}}. Concretely the elements
+of the pointed set denoted by {{1/b}} are \emph{linear functionals} $f : P_b
+\rightarrow P_1$. Note that $P_1$ is the pointed set containing {{()}} and
+$\bot_1$. The isomorphism between the types {{1}} and {{b * 1/b}} is now a
+consequence of that fact that the pairing of a vector and a functional in the
+dual space produces a scalar.
+
+\newpage
 

### 2012-06-27 rational_ext.tex
partial rewrite of 2.1 to emphasize the 2 levels more

--- 2012-06-27 rational_ext.tex
@@
 \newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}
+\newcommand{\jacques}[1]{\xcomment{Jacques}{#1}}
 
@@
 \newcommand{\arrow}[1]{\mathtt{#1}}
-\newcommand{\PiL}{\ensuremath{\Pi}}
 
@@
 \begin{verbatim}
-Restructure this section to emphasize the following
+(parts of rewrite that are done are removed from here)
 
@@
 
-types b ::= 0 | 1 | b+b | b*b
-values v ::= () | L v | R v | (v,v)
-typing v : b
-
-Semantics of values: 
- - a type b is interpreted as a set
- - a value v : b is an element of the set denoted by b
-
-types b <-> b
-combinators c ::= primitiveIso
-               | c + c | c * c | c ; c 
-typing c : b1 <-> b2
-
 Semantics of combinators:
- - types b1 <-> b2 are interpreted as bijections 
+ - iso-types b1 <-> b2 are interpreted as bijections 
    between the sets denoted by b1 and b2
  - a combinator c : b1 <-> b2 is interpeted as a 
-   bijection between the appropriate sets
+   bijection from set b1 to set b2.
+ - there is a natural adjoint to c : b1 <-> b2 
+   corresponding to the bijection c^* : b2 <-> b1
 
 Evaluation is function application, i.e., 
-  eval c v applies the function denoted by c to 
+  eval c v applies the function (bijection) denoted by c to 
   the value denoted by v.
@@
 of finite sets and bijections.]
-
 \end{verbatim}
@@
 
-Generally, such an approach would be rather vacuous.  What makes this
-pursuit different is that ... (will continue from here -- Jacques)
-
 %%%%%%%%%%%%%%%%%%%%
@@
 
-Unlike the traditional situation with the $\lambda$-calculus, in {{Pi}},
-there is a sharp distinction between data and programs.
-
-\paragraph*{Data.} The sets of values and their types include:
+\paragraph*{Data.}
+We view {{Pi}} as having two levels:  it has traditional values, given by
+%subcode{bnf} include main
+% values, v ::= () | left v | right v | (v, v)
+and these are classified by ordinary types
 %subcode{bnf} include main
 % value types, b ::= 0 | 1 | b + b | b * b
-% values, v ::= () | left v | right v | (v, v)
+
 Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
@@
 
-\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
-{{v \rightarrow v}}. Specifically, they are witnesses to the following type
-isomorphisms:
+But these should be regarded as largely ancilliary.  In particular we
+do not treat the above values as first-class citizens.  They only occur
+when we want to observe the effect of an isomorphism.
+
+Nevertheless, these have a precise denotation:  a type {{b}} is 
+interpreted as a set, and {{v : b}}  denotes that {{v}} is a member
+of the set denoted by {{b}}.  As our (traditional) values are first-order,
+and non-polymorphic, such a set interpretation is sound.
+
+\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
+isomorphisms.  They have (iso) types {{v \rightarrow v}}.  
+Specifically, they are witnesses to the following type isomorphisms:
 %subcode{bnf} include main
@@
 %dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-Each line of the above table introduces one or two combinators that witness
-the isomorphism in the middle. Collectively the isomorphisms state that the
-structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
-of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
-multiplication distributes over addition. The isomorphisms are extended to
-form a congruence relation by adding the following constructors that witness
-equivalence and compatible closure:
+Each line of the above table introduces a pair of dual constants%
+\footnote{where {{swap*}} and {{swap+}} are self-dual}
+that witness the type isomorphism in the middle.  Collectively the isomorphisms
+state that the structure {{(b,+,0,*,1)}} is a \emph{commutative semiring},
+i.e., that each of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
+multiplication distributes over addition.  These are the base 
+(non-reducible) terms of the second, principal level of {{Pi}}.
+
+It is worthwhile to remember that the above correspond to the \emph{axioms}
+of commutative semirings.  However, seen as universally quantified,
+orientable statements, these axioms also induce transformations of the
+(traditional) values. \jacques{include code here to be explicit}
+
+The isomorphisms are extended to form a congruence relation by adding the
+following constructors that witness equivalence and compatible closure:
 %subcode{proof} include main
@@
 
+The syntax is overloaded: we use the same symbol at the value-type level
+and at the isomorphism-type level for denoting sums and products.  Hopefully
+this will not cause undue confusion.
+
 \noindent
@@
 \label{def:Pi}
-We collect our types, values, and combinators, to get the full language
+We collect our types, values, and isomorphisms, to get the full language
 definition.
@@
 %
-% comb.~types, t ::= b <-> b
-% iso ::= zeroe | zeroi 
+% iso.~types, t ::= b <-> b
+% base iso ::= zeroe | zeroi 
 %     &|& swap+ | assocl+ | assocr+ 
@@
 %     &|& dist0 | factor0 | dist | factor 
-% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
 \end{definition}
@@
 \paragraph*{Adjoint.} 
-An important property of the language is that every combinator {{c}} has an
-adjoint {{c{dagger}}} that reverses the action of {{c}}. This is evident by
+An important property of the language is that every term {{c}} has an
+adjoint {{c{dagger}}} that reverses the isomorphisms {{c}}.  This is evident by
 construction for the primitive isomorphisms. For the closure combinators, the
@@
 \end{multicols}
-
 where {{+}} is numeric addition and {{*}} is numeric multiplication.
@@
 \noindent
-In the setting of {{Pi}} the \emph{size}, {{[^ b ^]}}, may be simply thought of as
-the number of inhabitants (an arity) of the type. This intuition will
+In the setting of {{Pi}} the \emph{size}, {{[^ b ^]}}, may be simply thought of
+as the number of inhabitants (an arity) of the type. This intuition will
 however become tenuous in the presence of negative and fractional

### 2012-06-27 rational_ext.tex

--- 2012-06-27 rational_ext.tex
@@
 %%%%%%%%%%%%%%%%%
-\subsection{Vector Spaces}
+\subsection{Pointed Sets}
 
 In order to interpret fractional types, we first lift the values and
-combinators of {{Pi}} to the world of vector spaces. Intuitively:
+combinators of {{Pi}} to the world of \emph{pointed sets}. Intuitively:
 \begin{itemize}
-\item each type {{b}} is now interpreted as a vector space of dimension
-  $\modulusp{b}$ over the field $\Fpx{2}$; the elements of such a vector
-  space are column vectors
-\item each value {{v_i : b}} is interpreted as column vector:
-\[\begin{array}{r@{~}l}
-\begin{array}{r} 
-v_0 \\ v_1 \\ \vdots \\ v_i \\ \vdots \\ v_n 
-\end{array} & 
-\begin{bmatrix}
-0 \\ 0 \\ \vdots \\ 1 \\ \vdots \\ 0
-\end{bmatrix}
-\end{array}\]
-\item each combinator {{c : b1 <-> b2}} is interpreted as a matrix whose
-  columns are indexed by the values of type {{b1}} and whose rows are indexed
-  by the values of type {{b2}}. The matrix is \emph{orthogonal} with each row
-  and column containing exactly one occurrence of the value 1.
+\item instead of interpreting a type {{b}} as a plain set of cardinality
+  $\modulusp{b}$, we now interpret each type {{b}} as a pointed set $P_b$
+  with a distinguished \emph{basepoint} $\bot_b$;
+\item a combinator {{c : b1 <-> b2}} is interpreted as a function as before
+  with the restriction that it must map the basepoint $\bot_{b_1}$ to the
+  basepoint $\bot_{b_2}$.
 \end{itemize}
 
-Once we are in the world of vector spaces, the fractional type {{1/b}} is
-interpreted as the \emph{dual} to the vector space denoted by {{b}}. The
-elements of this dual vector space can be represented as \emph{row vectors}.
-
-Multiplying a row vector by a column gives a scalar: this is eta and epsilon.
+Intuitively a pointed set may be seen as a vector space over the field with
+one element. This view allows us to add the type {{1/b}} and interpret it as
+the \emph{dual} of the vector space denoting {{b}}. Concretely the elements
+of the pointed set denoted by {{1/b}} are \emph{linear functionals} $f : P_b
+\rightarrow P_1$. Note that $P_1$ is the pointed set containing {{()}} and
+$\bot_1$. The isomorphism between the types {{1}} and {{b * 1/b}} is now a
+consequence of that fact that the pairing of a vector and a functional in the
+dual space produces a scalar.
 

### 2012-06-28 rational_ext.tex
relations

@@
 %! |} = \}
+%! bot = \bot
+%! bot1 = \bot_1
 %! dots = \ldots
@@
 
+\begin{verbatim}
+revise the paragraph below to 
+say something that's consistent with the 
+presentation in the previous section; the
+previous type system was the natural numbers;
+we presented a natural language of terms
+for that type system and a semantics in 
+the rig category of finite
+sets and bijections (two symmetric
+monoidal structures and distributivity).
+now we extend the type system to correspond
+to the non-negative rational numbers;
+we want to find the corresponding language
+and semantic interpretation
+\end{verbatim}
+
 In arithmetic the main identity satisfied by fractional numbers is that for
-non-zero {{b}}, we have {{b * 1/b = 1}}. Translating this identity to our
-world requires us to find an isomorphism between the types {{1}} and 
-{{b * 1/b}} (where {{b}} is not equivalent to {{0}}). Formally, we move 
-from the setting of symmetric monoidal categories to that of compact
-closed categories.
+non-zero {{b}}, we have {{b * 1/b = 1}}. Accommodating this identity in the
+setting of {{Pi}} requires us to find an isomorphism between the types {{1}}
+and {{b * 1/b}} (where {{b}} is not equivalent to {{0}}). Formally, we move
+from the setting of symmetric monoidal categories to that of compact closed
+categories.
 
 %%%%%%%%%%%%%%%%%
-\subsection{Pointed Sets}
+\subsection{Non-determinism}
 
 In order to interpret fractional types, we first lift the values and
-combinators of {{Pi}} to the world of \emph{pointed sets}. Intuitively:
+combinators of {{Pi}} to the world of sets and relations. The lifting is
+realized using a monad whose functor is powerset. Intuitively:
 \begin{itemize}
-\item instead of interpreting a type {{b}} as a plain set of cardinality
-  $\modulusp{b}$, we now interpret each type {{b}} as a pointed set $P_b$
-  with a distinguished \emph{basepoint} $\bot_b$;
-\item a combinator {{c : b1 <-> b2}} is interpreted as a function as before
-  with the restriction that it must map the basepoint $\bot_{b_1}$ to the
-  basepoint $\bot_{b_2}$.
+\item instead of interpreting a type {{b}} as a finite set of values $\{ v_1,
+  v_2, \ldots, v_{\modulusp{b}}\}$, we now interpret each type {{b}} as the
+  powerset of such values; note that the type {{0}} now denotes
+  $\{\emptyset\}$, i.e., it has one element which is the empty set;
+  \item a {{Pi}}-value {{v}} of type {{b}} is lifted to a singleton set 
+  $\{ v \}$;
+\item each combinator {{c : b1 <-> b2}} is lifted to a \emph{symmetric}
+  relation $R_c$ defined as follows:
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\{ v_1, v_2, \ldots \} &R_c& \bigcup \{ c(v_i) \} 
+\end{array}\]
+For example, the combinator {{swap+ : 1+1 <-> 1+1}} is lifted to the relation $R$:
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\emptyset &R& \emptyset \\
+\{ \mathit{left}~() \} &R& \{ \mathit{right}~() \} \\
+\{ \mathit{right}~() \} &R& \{ \mathit{left}~() \} \\
+\{ \mathit{left}~(), \mathit{right}~() \} &R& \{ \mathit{left}~(), \mathit{right}~() \} 
+\end{array}\]
 \end{itemize}
-
-Intuitively a pointed set may be seen as a vector space over the field with
-one element. This view allows us to add the type {{1/b}} and interpret it as
-the \emph{dual} of the vector space denoting {{b}}. Concretely the elements
-of the pointed set denoted by {{1/b}} are \emph{linear functionals} $f : P_b
-\rightarrow P_1$. Note that $P_1$ is the pointed set containing {{()}} and
-$\bot_1$. The isomorphism between the types {{1}} and {{b * 1/b}} is now a
-consequence of that fact that the pairing of a vector and a functional in the
-dual space produces a scalar.
-
-\newpage
-
-\paragraph*{Data.} The syntax of {{Pi}} is extended with the type {{S b}}
-which denotes sets {{ {| v1, v2, dots |} }} of values of type {{b}}:
-%subcode{bnf} include main
-% value types, b ::= 0 | 1 | b + b | b * b | S b
-% values, v ::= () | left v | right v | (v, v) | {| v1, v2, dots |}
-
-Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
-and products types {{b1*b2}}.  Values includes {{()}} which is the only value
-of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
-and {{(v1,v2)}} which builds a value of product type. There are no values of
-type {{0}}.
-
-\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
-{{v -> v}}. Specifically, they are witnesses to the following type
-isomorphisms:
-%subcode{bnf} include main
-%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
-%zeroe :&  0 + b <-> b &: zeroi
-%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
-%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
-%identl* :&  1 * b <-> b &: identr*
-%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
-%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
-%dist0 :& 0 * b <-> 0 &: factor0
-%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-Each line of the above table introduces one or two combinators that witness
-the isomorphism in the middle. Collectively the isomorphisms state that the
-structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
-of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
-multiplication distributes over addition. The isomorphisms are extended to
-form a congruence relation by adding the following constructors that witness
-equivalence and compatible closure:
-%subcode{proof} include main
-%@  ~
-%@@ id : b <-> b 
-%
-%@ c : b1 <-> b2
-%@@ sym c : b2 <-> b1
-%
-%@ c1 : b1 <-> b2
-%@ c2 : b2 <-> b3
-%@@ c1(;)c2 : b1 <-> b3
-%---
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
-%
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
-
-\noindent
-To summarize, the syntax of {{Pi}} is given as follows.
-
-\begin{definition}{(Syntax of {{Pi}})}
-\label{def:Pi}
-We collect our types, values, and combinators, to get the full language
-definition.
-%subcode{bnf} include main
-% value types, b ::= 0 | 1 | b+b | b*b 
-% values, v ::= () | left v | right v | (v,v) 
-%
-% comb.~types, t ::= b <-> b
-% iso ::= zeroe | zeroi 
-%     &|& swap+ | assocl+ | assocr+ 
-%     &|& identl* | identr* 
-%     &|& swap* | assocl* | assocr* 
-%     &|& dist0 | factor0 | dist | factor 
-% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-\end{definition}
-
+Because every combinator corresponds to a bijection, the lifted relation is
+symmetric.  Formally we have moved from the category of finite sets and
+bijections to the category of finite sets and \emph{symmetric relations}.
+
+Now that we have introduced the potential for non-determinism, it is possible
+to introduce a type {{1/b}}, values {{1/v}}, and combinators {{eta*}} and
+{{eps*}} that witness the following isomorphism:
 %subcode{opsem} include main
 % eta* &: 1 <-> (1/b) * b :& eps*
-
-IntuitivelyThe main challenge Adding fractional types to
-{{Pi}} takes us from a symmetric monoidal category to a compact closed
-category.
+where {{b}} is not ``equivalent'' to {{0}} as detailed next.
+
+Intuitively, we view a value {{1/v}} as a \emph{constraint} that is only
+satisfied if it is matched with a value {{v'}} that is identical to
+{{v}}. The type {{1/b}} is interpreted as the powerset of all the values
+{{1/v}} with {{v}} being a {{Pi}}-value of type {{b}}. For example, the type
+{{1/(1+1)}} denotes:
+\[
+\{ 
+\emptyset, 
+\{ 1/\mathit{left}~() \}, 
+\{ 1/\mathit{right}~() \},
+\{ 1/\mathit{left}~(), 1/\mathit{right}~() \} 
+\}
+\]
+Of particular interest is that both the type {{0}} and the type {{1/0}}
+denote $\{\emptyset\}$.
+
+Given this interpretation of types and values, the realization of 
+{{eta* : 1 <-> (1/b) * b}} is the relation $R$: 
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\emptyset &R& \emptyset \\
+\{ () \} &R& \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \}
+\end{array}\]
+For example, at type $1+1$, the interpretation of {{eta*}} is the relation:
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\emptyset &R& \emptyset \\
+\{ () \} &R& \{ (1/\mathit{left}~(),\mathit{left}~()), 
+                (1/\mathit{right}~(),\mathit{right}~()) \} 
+\end{array}\]
+At type $0$, the interpretation of {{eta*}} is:
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\emptyset &R& \emptyset \\
+\{ () \} &R& \emptyset 
+\end{array}\]
+which is not a symmetric relation. Hence the presence of {{eta*}} introduces
+some ``non-reversible'' effect and moves us to the category of finite sets 
+and (not necessarily symmetric) relations. The interpretation of 
+{{eps* : (1/b) * b <-> 1}} is the relation:
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\emptyset &R& \emptyset \\
+\{ \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \} &R& \{ () \} \\
+\{ \{ (1/v,v') ~|~ v \mbox{~is~not~identical~to~} v' \} &R& emptyset 
+\end{array}\]
+For example, at type $0$, we have:
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\emptyset &R& \emptyset 
+\end{array}\]
+At type $1+1$, we have:
+\[\begin{array}{r@{\qquad}c@{\qquad}l}
+\emptyset &R& \emptyset \\
+\{ \{ (1/\mathit{left}~(),\mathit{left}~()) \} &R& \{ () \} \\
+\{ \{ (1/\mathit{left}~(),\mathit{right}~()) \} &R& \emptyset \\
+\{ \{ (1/\mathit{right}~(),\mathit{left}~()) \} &R& \emptyset \\
+\{ \{ (1/\mathit{right}~(),\mathit{right}~()) \} &R& \{ () \} \\
+\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
+  \{ () \} \} \\
+\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
+  \{ () \} \} \\
+\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
+  \emptyset \} \\
+\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
+  \emptyset \} 
+\end{array}\]
+
+The category of finite sets and relations is a compact closed category. Each
+object has a dual which is itself. The main coherence condition requires us
+to check that:
+\[\begin{array}{rcl}
+b &\rightarrow& b * 1 \\
+  &\rightarrow_{\eta^*}& b * (1/b * b) \\
+  &\rightarrow& (b * 1/b) * b \\
+  &\rightarrow_{\epsilon^*}& 1 * b \\
+  &\rightarrow& b
+\end{array}\]
+For example, let's check one instance of this identity at the type $1+1$ and
+the value {{left ()}}:
+\[\begin{array}{rcl}
+\{ \mathit{left}~() \} &\rightarrow& \{ (\mathit{left}~(), ()) \} \\
+  &\rightarrow_{\eta^*}& \{ (\mathit{left}~(),(1/\mathit{left}~(),\mathit{left}~())), 
+                            (\mathit{left}~(),(1/\mathit{right}~(),\mathit{right}~())) \} \\
+  &\rightarrow& \{ ((\mathit{left}~(),1/\mathit{left}~()),\mathit{left}~())), 
+                   ((\mathit{left}~(),1/\mathit{right}~()),\mathit{right}~())) \} \\
+  &\rightarrow_{\epsilon^*}& \{ ((),\mathit{left}~()) \} \\
+  &\rightarrow& \{ \mathit{left}~() \} 
+\end{array}\]
+At type $0$, we would get:
+\[\begin{array}{rcl}
+\emptyset &\rightarrow& \emptyset \\
+  &\rightarrow_{\eta^*}& \emptyset \\
+  &\rightarrow& \emptyset \\
+  &\rightarrow_{\epsilon^*}& \emptyset \\
+  &\rightarrow& \emptyset
+\end{array}\]
+
+%%%%%%%%%%%%%%%%
+\subsection{Syntax}
 

### 2012-06-28 code/modular/ND.hs
better version of Amr's test6

@@ -245,5 +245,8 @@ test5 = eval justTimes2 (Left (), Right ())
 -- inv :: (PiInv eq) => eq a a
 -- inv = timesOneR %. (etaTimes %* id) %. (epsTimes %* id) %. timesOneL
 
-
+-- to Show this, need to fix a monad; that's what test6s is for.
+test6 :: MonadPlus m => m Zero
+test6 = eval cohTimes (undefined::Zero)
+test6s = test6 :: [Zero]
 ------------------------------------------------------------------------------

### 2012-06-28 rational_ext.tex
reorder subsections of 2
add some text in 2.1 and fix opsem to add missing rules

@@
 \paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
-isomorphisms.  They have (iso) types {{v \rightarrow v}}.  
+isomorphisms.  They have (iso) types {{v <-> v}}.  
 Specifically, they are witnesses to the following type isomorphisms:
@@
 Each line of the above table introduces a pair of dual constants%
-\footnote{where {{swap*}} and {{swap+}} are self-dual}
+\footnote{where {{swap*}} and {{swap+}} are self-dual} (and its typing)
 that witness the type isomorphism in the middle.  Collectively the isomorphisms
@@
 
-It is worthwhile to remember that the above correspond to the \emph{axioms}
-of commutative semirings.  However, seen as universally quantified,
-orientable statements, these axioms also induce transformations of the
-(traditional) values. \jacques{include code here to be explicit}
-
+Note how the above has two readings: first as a set of typing relations for
+a set of constants, as well as giving names to the \emph{axioms} of
+commutative semirings.  However, if these axioms are seen as universally
+quantified, orientable statements, they also induce transformations of
+the (traditional) values.  The (categorical) intuition here is that 
+these axioms have computational content because they witness isomorphisms
+rather than merely stating an extensional equality.
+
+For example {{ swap* }} induces a function 
+%subcode{code}
+%! language = Haskell
+% swapProd (a,b) = (b,a)
 The isomorphisms are extended to form a congruence relation by adding the
@@
 %%%%%%%%%%%%
-\subsection{Graphical Language}
-
-The syntactic notation above is often obscure and hard to read.
-Following the tradition established for monoidal
-categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present
-a graphical language that conveys the intuitive semantics of the
-language (which is formalized in the next section).
-
-The general idea of the graphical notation is that combinators are modeled by
-``wiring diagrams'' or ``circuits'' and that values are modeled as
-``particles'' or ``waves'' that may appear on the wires. Evaluation therefore
-is modeled by the flow of waves and particles along the wires.
-
-%% In fact, when talking about {{Pi}} circuits, we will often use the
-%% words {{value}} and {{particle}} interchangeably.
-
-\begin{itemize}
-\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
-  is simply represented as a wire labeled by its type {{b}}, as shown on the
-  left. In more complex diagrams, if the type of a wire is obvious from the
-  context, it may be omitted. When tracing a computation, one might imagine a
-  value {{v}} of type {{b}} on the wire, as shown on the right.
-
-  \begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
-\includegraphics{diagrams/thesis/b-wire.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
-\includegraphics{diagrams/thesis/b-wire-value.pdf}
-}
-\end{center}
-  \end{multicols}
-
-\item The product type {{b1*b2}} may be represented using either one wire
-  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
-  case of products represented by a pair of wires, when tracing execution
-  using particles, one should think of one particle on each wire or
-  alternatively as in folklore in the literature on monoidal categories as a
-  ``wave.''
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
-\includegraphics{diagrams/thesis/product-one-wire.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
-}
-\end{center}
-\end{multicols}
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
-\includegraphics{diagrams/thesis/product-two-wires.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item Sum types may similarly be represented by one wire or using
-  parallel wires with a {{+}} operator between them. When tracing the
-  execution of two additive wires, a value can reside on only one of the two
-  wires.
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
-\includegraphics{diagrams/thesis/sum-one-wire.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
-}
-\end{center}
-\end{multicols}
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
-\includegraphics{diagrams/thesis/sum-two-wires.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
-}
-\end{center}
-\end{multicols}
-
-%% \item
-%% When representing complex types like {{(b1*b2)+b3}} some visual
-%% grouping of the wires may be done to aid readability. The exact type
-%% however will always be clarified by the context of the diagram.
-
-%% \begin{center}
-%% \scalebox{0.95}{
-%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
-%% }
-%% \end{center}
-
-\item Associativity is implicit in the graphical language. Three parallel
-  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
-\includegraphics{diagrams/thesis/assoc.pdf}
-}
-\end{center}
-
-\item Commutativity is represented by crisscrossing wires.
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
-\includegraphics{diagrams/thesis/swap_times.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
-\includegraphics{diagrams/thesis/swap_plus.pdf}
-}
-\end{center}
-\end{multicols}
-
-By visually tracking the flow of particles on the wires, one can
-verify that the expected types for commutativity are satisfied.
-
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/swap_times_value.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-\includegraphics{diagrams/thesis/swap_plus_value.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item The morphisms that witness that {{0}} and {{1}} are the additive and
-  multiplicative units are represented as shown below. Note that since there
-  is no value of type 0, there can be no particle on a wire of type {{0}}.
-  Also since the monoidal units can be freely introduced and eliminated, in
-  many diagrams they are omitted and dealt with explicitly only when they are
-  of special interest.
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
-\includegraphics{diagrams/thesis/uniti.pdf}
-}
-\end{center}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
-\includegraphics{diagrams/thesis/unite.pdf}
-}
-\end{center}  
-\end{multicols}
-\begin{multicols}{2}
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
-\includegraphics{diagrams/thesis/zeroi.pdf}
-}
-\end{center}
-\columnbreak
-\begin{center}
-\scalebox{0.95}{
-%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
-\includegraphics{diagrams/thesis/zeroe.pdf}
-}
-\end{center}
-\end{multicols}
-
-\item Finally, distributivity and factoring are represented using the dual
-  boxes shown below:
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/thesis/dist.pdf}
-\end{center}
-\begin{center}
-  \includegraphics{diagrams/thesis/factor.pdf}
-\end{center}
-\end{multicols}
-
-Distributivity and factoring are interesting because they represent
-interactions between sum and pair types. Distributivity should
-essentially be thought of as a multiplexer that redirects the flow of
-{{v:b}} depending on what value inhabits the type {{b1+b2}}, as shown
-below. Factoring is the corresponding adjoint operation.
-
-\begin{multicols}{2}
-\begin{center}
-  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
-\end{center}
-\begin{center}
-  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
-\end{center}
-\end{multicols}
-
-\item Combinators can be composed is series ({{c1 (;) c2}}) or
-  parallel. There are two forms of parallel composition -- combinators
-  can be combined additively {{c1(+)c2}} (shown on the left) or
-  multiplicatively {{c1(*)c2}} (shown on the right).
-
-  \begin{multicols}{2}
-\dgm{c1c2_par_sum}
-\dgm{c1c2_par_times}
-  \end{multicols}
-\dgm{c1c2_seq}
-
-\end{itemize}
-
-%% \noindent 
-%% \textit{Example.}  We use the type {{bool}} as a shorthand to denote
-%% the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
-%% to be {{false}}. The following combinator is represented by the given
-%% diagram:
-
-%% {{c : b * bool <-> b + b}}
-
-%% {{c = swap* (;) dist (;) (identl* (+) identl*)}}
-
-%% \begin{center}
-%% \scalebox{0.95}{
-%% %%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
-%% \includegraphics{diagrams/thesis/example1.pdf}
-%% }
-%% \end{center}
-
-%%%%%%%%%%%%
 \subsection{Semantics}
@@
 given by the following single-step reductions below. Since there are
-no values of type {{0}}, the rules omit the impossible cases:
+no values of type {{0}}, there are no rules for the impossible cases:
 \begin{scriptsize}
@@
 %! columnStyle = rlcl
+% zeroi & v & |-->& right v
+% zeroe & (right v) &|-->& v
 % swap+ & (left v) &|-->& right v
@@
 
-%%%%%%%%%%%%%%%%
-\subsection{Categorical Structure}
-
-We present the categorical structure with minimum commentary. More
-details may be found in excellent references such as Barr and Wells
-[CITE] and Selinger \cite{springerlink:10.1007/978-3-642-12821-94}.
-For brevity, we don't re-state standard categorical definitions and
-subsequently present their {{Pi}} equivalents. Instead, we directly
-state categorical definitions in terms of {{Pi}} types and
-combinators, and provide appropriate references to the former. For
-example, associativity is denoted by the Greek letter $\alpha$ (and
-$\alpha^{-1}$) in standard definitions, whereas we use {{assocl*}}
-(and {{assocr*}}).
-
-\begin{lemma}[{{Pi}} is a category]
-  The category {{Pi}} has the types {{b}} as objects and equivalence
-  classes of well-typed combinators {{b1<->b2}} as morphisms. One can
-  check:
-  \begin{enumerate}
-  \item Every object {{b}} has an identity morphism {{id : b <->b}}.
-  \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
-    {{g:b2<->b3}} is given by sequencing {{f (;) g}}. 
-  \item Associativity of composition follows from operational
-    equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
-    {{h:b3<->b4}}).
-
-    Assuming {{v1:b1}}, {{v2:b2}}, {{v3:b3}}, {{v4:b4}},
-    {{f~v1|-->v2}}, {{g v2 |--> v3}} and {{h~v3 |--> v4}}, one can
-    check:
-
-\vspace{-20pt}
-    \begin{multicols}{2}
-      \begin{scriptsize}
-
-%subcode{proof} include main
-%@@ f v1 |--> v2
-%@ g v2 |--> v3
-%@ h v3 |--> v4
-%@@ g (;) h  v2 |--> v4
-%@@@ f (;) (g (;) h) v1 |--> v4
-~
-%subcode{proof} include main
-%@ f v1 |--> v2
-%@ g v2 |--> v3
-%@@ f (;) g v1 |--> v3
-%@@ h v3 |--> v4
-%@@@ (f (;) g) (;) h v1 |--> v4
-
-      \end{scriptsize}
-    \end{multicols}
-
-  \item Composition respects identity: {{id (;) f = f}} and {{g(;)id=g}}.
-  \end{enumerate}
-\end{lemma}
-
-\begin{lemma}[Dagger]
-  {{Pi}} is a dagger category, where every morphism {{f : b1<->b2}}
-  has the adjoint {{f^{dagger}:b2 <->b1}}. The following properties
-  hold (where {{g : b2 <-> b3}}): 
-  \begin{enumerate}
-  \item {{id^{dagger} = id : b <-> b}}.
-  \item {{(f (;) g)^{dagger} = g^{dagger} (;) f^{dagger}: b3 <-> b1}}.
-  \item {{f^{dagger dagger} = f : b1 <-> b2}}.
-  \end{enumerate}
-\end{lemma}
-
-\begin{lemma}[Symmetric Monoidal (+, 0)]
-  {{Pi}} is a symmetric monoidal category with tensor {{+}} and
-  monoidal unit {{0}}. The monoidal operation on morphisms is the
-  additive composition of combinators {{c1 (+) c2}}.
-
-  \begin{proof}
-    To establish that category is monoidal one must show isomorphisms
-    \begin{enumerate}
-    \item {{b1 + (b2 + b3) <-> (b1 +b2) + b3}} given by {{assocl+}}.
-    \item {{0 + b <-> b}} given by {{zeroe}}.
-    \item {{b + 0 <-> b}} given by {{swap+ (;) zeroe}}.
-    \item We need to check that {{+}} is a bifunctor. 
-      \begin{enumerate}
-      \item {{id_{b1<->b1}(+)id_{b2<->b2} = id_{b1+b2<->b1+b2} }}.
-      \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}.
-      \end{enumerate}
-    \item We need to check the naturality of {{assocl+}}, {{zeroe}} and
-      {{swap+(;)zeroe}}.
-      \begin{enumerate}
-      \item {{assocl+ (;) ((f (+)g) (+) h) = (f (+)(g (+) h)) (;) assocl+}}.
-      \item {{zeroe (;) f = (id (+) f) (;) zeroe}}.
-      \item {{(swap+ (;) zeroe) (;) f = (f (+) id) (;) (swap+ (;) zeroe)}}.
-      \end{enumerate}
-    \item Satisfy certain coherence conditions which are usually
-      called the ``pentagon'' and ``triangle'' axioms (see Sec 3.1
-      \cite{springerlink:10.1007/978-3-642-12821-94})
-    \end{enumerate}
-
-    The last three points require checking equality of combinators by
-    writing out their derivation trees as we did in the case of
-    associativity of sequential composition. To show symmetry, we need
-    a braiding operation {{b1+b2 <-> b2+b1}} which is given by
-    {{swap+}} (see Sec. 3.3 and 3.5
-    \cite{springerlink:10.1007/978-3-642-12821-94}).
-
-    \begin{enumerate}
-    \item The braiding must satisfy two ``hexagon'' axioms.
-    \item The braiding is self inverse,
-      {{swap+_{b1+b2}=(swap+_{b2+b1})^{dagger} }}.
-    \end{enumerate}
-
-  \end{proof}
-
-\end{lemma}
-
-\begin{lemma}[Symmetric Monoidal {{(*, 1)}}]
-  {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
-  unit {{1}}. The details mirror those of the {{(0, +)}} monoid.
-\end{lemma}
-
-\noindent
-Some technical and pedantic comments are due at this point.
-
-\begin{itemize}
-\item We have established the categorical structure of {{Pi}} as a
-  dagger symmetric monoidal category with two monoidal structures,
-  {{(0, +)}} and {{(*, 1)}}. In Sec. \ref{sec:int} we will see that a
-  {{trace}} operator can be admitted in this category without any
-  change of expressiveness.
-
-\item To be pedantic, what we have shown is that the ``term model'' of
-  {{Pi}} that follows from the extensional operational equality of
-  combinators has the requisite categorical structure. A consequence
-  is that the ``wiring diagrams'' of {{Pi}} correspond closely with ``string
-  diagrams'' developed for categories.
-
-  To establish the later rigorously, we will need to show when it is
-  valid to slide one wire over the other and that equivalent diagrams
-  for syntactically different combinators such as
-  {{(f(+)g)(;)(j(+)k)}} and {{(f(;)j)(+)(g(;)k)}} do respect
-  operational equivalence.  We don't formalize the graphical notation
-  in this work. Joyal et. al's work on ``planar isotopy'' and
-  Selinger's survey \cite{springerlink:10.1007/978-3-642-12821-94}
-  show how this has been addressed before in the categorical
-  setting. In the absence of any prior knowledge of category theory
-  however, our wiring diagrams may be read as the ``flow of types'' in
-  a combinator-circuit.
-
-\item By definition, in {{Pi}} every morphism is an isomorphism --
-  this makes {{Pi}} a groupoid.
-
-\item The category {{Pi}} has no initial and terminal objects. The
-  objects {{0}} and {{1}} would be initial and terminal if we admitted
-  all functions (as in the category {{CatSet}}).
-
-\item The category {{Pi}} has neither categorical products, nor
-  categorical co-products, i.e. {{*}} and {{+}} are merely monoidal
-  tensors. This follows from the fact that injection and projection
-  maps are not isomorphisms.
-\end{itemize}
-
-While we don't do so in this work, if we extend {{Pi}} with products
-and co-products (say through the addition of information effects) it
-is conceivable that part of its structure collapses. This sort of
-collapse, while catastrophic for algebraic structures (effectively
-trivializing them), still retains some interest in computing, because
-in computing we are interested in the specific operational nature (the
-computational content, so to speak) of the morphisms. Anecdotal evidence
-follows from the fact several real-world programming language have
-inconsistent type systems. blah blah blah...
-
 %%%%%%%%%%%%
@@
 % \end{proposition}
+
+%%%%%%%%%%%%
+\subsection{Graphical Language}
+
+The syntactic notation above is often obscure and hard to read.
+Following the tradition established for monoidal
+categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present
+a graphical language that conveys the intuitive semantics of the
+language (which is formalized in the next section).
+
+The general idea of the graphical notation is that combinators are modeled by
+``wiring diagrams'' or ``circuits'' and that values are modeled as
+``particles'' or ``waves'' that may appear on the wires. Evaluation therefore
+is modeled by the flow of waves and particles along the wires.
+
+%% In fact, when talking about {{Pi}} circuits, we will often use the
+%% words {{value}} and {{particle}} interchangeably.
+
+\begin{itemize}
+\item The simplest sort of diagram is the {{id : b <-> b}} combinator which
+  is simply represented as a wire labeled by its type {{b}}, as shown on the
+  left. In more complex diagrams, if the type of a wire is obvious from the
+  context, it may be omitted. When tracing a computation, one might imagine a
+  value {{v}} of type {{b}} on the wire, as shown on the right.
+
+  \begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/b-wire.pdf]
+\includegraphics{diagrams/thesis/b-wire-value.pdf}
+}
+\end{center}
+  \end{multicols}
+
+\item The product type {{b1*b2}} may be represented using either one wire
+  labeled {{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. In the
+  case of products represented by a pair of wires, when tracing execution
+  using particles, one should think of one particle on each wire or
+  alternatively as in folklore in the literature on monoidal categories as a
+  ``wave.''
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/pair-one-wire.pdf]
+\includegraphics{diagrams/thesis/product-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%%subcode-line{pdfimage}[diagrams/thesis/pair-of-wires.pdf]
+\includegraphics{diagrams/thesis/product-two-wires.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Sum types may similarly be represented by one wire or using
+  parallel wires with a {{+}} operator between them. When tracing the
+  execution of two additive wires, a value can reside on only one of the two
+  wires.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
+\includegraphics{diagrams/thesis/sum-one-wire.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
+}
+\end{center}
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
+\includegraphics{diagrams/thesis/sum-two-wires.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
+}
+\end{center}
+\end{multicols}
+
+%% \item
+%% When representing complex types like {{(b1*b2)+b3}} some visual
+%% grouping of the wires may be done to aid readability. The exact type
+%% however will always be clarified by the context of the diagram.
+
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %subcode-line{pdfimage}[diagrams/thesis/complex-type-crop.pdf]
+%% }
+%% \end{center}
+
+\item Associativity is implicit in the graphical language. Three parallel
+  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/associate.pdf]
+\includegraphics{diagrams/thesis/assoc.pdf}
+}
+\end{center}
+
+\item Commutativity is represented by crisscrossing wires.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-pair.pdf]
+\includegraphics{diagrams/thesis/swap_times.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/swap-sum.pdf]
+\includegraphics{diagrams/thesis/swap_plus.pdf}
+}
+\end{center}
+\end{multicols}
+
+By visually tracking the flow of particles on the wires, one can
+verify that the expected types for commutativity are satisfied.
+
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_times_value.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+\includegraphics{diagrams/thesis/swap_plus_value.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item The morphisms that witness that {{0}} and {{1}} are the additive and
+  multiplicative units are represented as shown below. Note that since there
+  is no value of type 0, there can be no particle on a wire of type {{0}}.
+  Also since the monoidal units can be freely introduced and eliminated, in
+  many diagrams they are omitted and dealt with explicitly only when they are
+  of special interest.
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr1.pdf]
+\includegraphics{diagrams/thesis/uniti.pdf}
+}
+\end{center}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl1.pdf]
+\includegraphics{diagrams/thesis/unite.pdf}
+}
+\end{center}  
+\end{multicols}
+\begin{multicols}{2}
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identr0.pdf]
+\includegraphics{diagrams/thesis/zeroi.pdf}
+}
+\end{center}
+\columnbreak
+\begin{center}
+\scalebox{0.95}{
+%%subcode-line{pdfimage}[diagrams/thesis/identl0.pdf]
+\includegraphics{diagrams/thesis/zeroe.pdf}
+}
+\end{center}
+\end{multicols}
+
+\item Finally, distributivity and factoring are represented using the dual
+  boxes shown below:
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/factor.pdf}
+\end{center}
+\end{multicols}
+
+Distributivity and factoring are interesting because they represent
+interactions between sum and pair types. Distributivity should
+essentially be thought of as a multiplexer that redirects the flow of
+{{v:b}} depending on what value inhabits the type {{b1+b2}}, as shown
+below. Factoring is the corresponding adjoint operation.
+
+\begin{multicols}{2}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
+\end{center}
+\begin{center}
+  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
+\end{center}
+\end{multicols}
+
+\item Combinators can be composed is series ({{c1 (;) c2}}) or
+  parallel. There are two forms of parallel composition -- combinators
+  can be combined additively {{c1(+)c2}} (shown on the left) or
+  multiplicatively {{c1(*)c2}} (shown on the right).
+
+  \begin{multicols}{2}
+\dgm{c1c2_par_sum}
+\dgm{c1c2_par_times}
+  \end{multicols}
+\dgm{c1c2_seq}
+
+\end{itemize}
+
+%% \noindent 
+%% \textit{Example.}  We use the type {{bool}} as a shorthand to denote
+%% the type {{1+1}} and use {{left ()}} to be {{true}} and {{right ()}}
+%% to be {{false}}. The following combinator is represented by the given
+%% diagram:
+
+%% {{c : b * bool <-> b + b}}
+
+%% {{c = swap* (;) dist (;) (identl* (+) identl*)}}
+
+%% \begin{center}
+%% \scalebox{0.95}{
+%% %%subcode-line{pdfimage}[diagrams/thesis/example1-crop.pdf]
+%% \includegraphics{diagrams/thesis/example1.pdf}
+%% }
+%% \end{center}
+
+%%%%%%%%%%%%%%%%
+\subsection{Categorical Structure}
+
+We present the categorical structure with minimum commentary. More
+details may be found in excellent references such as Barr and Wells
+[CITE] and Selinger \cite{springerlink:10.1007/978-3-642-12821-94}.
+For brevity, we don't re-state standard categorical definitions and
+subsequently present their {{Pi}} equivalents. Instead, we directly
+state categorical definitions in terms of {{Pi}} types and
+combinators, and provide appropriate references to the former. For
+example, associativity is denoted by the Greek letter $\alpha$ (and
+$\alpha^{-1}$) in standard definitions, whereas we use {{assocl*}}
+(and {{assocr*}}).
+
+\begin{lemma}[{{Pi}} is a category]
+  The category {{Pi}} has the types {{b}} as objects and equivalence
+  classes of well-typed combinators {{b1<->b2}} as morphisms. One can
+  check:
+  \begin{enumerate}
+  \item Every object {{b}} has an identity morphism {{id : b <->b}}.
+  \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
+    {{g:b2<->b3}} is given by sequencing {{f (;) g}}. 
+  \item Associativity of composition follows from operational
+    equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
+    {{h:b3<->b4}}).
+
+    Assuming {{v1:b1}}, {{v2:b2}}, {{v3:b3}}, {{v4:b4}},
+    {{f~v1|-->v2}}, {{g v2 |--> v3}} and {{h~v3 |--> v4}}, one can
+    check:
+
+\vspace{-20pt}
+    \begin{multicols}{2}
+      \begin{scriptsize}
+
+%subcode{proof} include main
+%@@ f v1 |--> v2
+%@ g v2 |--> v3
+%@ h v3 |--> v4
+%@@ g (;) h  v2 |--> v4
+%@@@ f (;) (g (;) h) v1 |--> v4
+~
+%subcode{proof} include main
+%@ f v1 |--> v2
+%@ g v2 |--> v3
+%@@ f (;) g v1 |--> v3
+%@@ h v3 |--> v4
+%@@@ (f (;) g) (;) h v1 |--> v4
+
+      \end{scriptsize}
+    \end{multicols}
+
+  \item Composition respects identity: {{id (;) f = f}} and {{g(;)id=g}}.
+  \end{enumerate}
+\end{lemma}
+
+\begin{lemma}[Dagger]
+  {{Pi}} is a dagger category, where every morphism {{f : b1<->b2}}
+  has the adjoint {{f^{dagger}:b2 <->b1}}. The following properties
+  hold (where {{g : b2 <-> b3}}): 
+  \begin{enumerate}
+  \item {{id^{dagger} = id : b <-> b}}.
+  \item {{(f (;) g)^{dagger} = g^{dagger} (;) f^{dagger}: b3 <-> b1}}.
+  \item {{f^{dagger dagger} = f : b1 <-> b2}}.
+  \end{enumerate}
+\end{lemma}
+
+\begin{lemma}[Symmetric Monoidal (+, 0)]
+  {{Pi}} is a symmetric monoidal category with tensor {{+}} and
+  monoidal unit {{0}}. The monoidal operation on morphisms is the
+  additive composition of combinators {{c1 (+) c2}}.
+
+  \begin{proof}
+    To establish that category is monoidal one must show isomorphisms
+    \begin{enumerate}
+    \item {{b1 + (b2 + b3) <-> (b1 +b2) + b3}} given by {{assocl+}}.
+    \item {{0 + b <-> b}} given by {{zeroe}}.
+    \item {{b + 0 <-> b}} given by {{swap+ (;) zeroe}}.
+    \item We need to check that {{+}} is a bifunctor. 
+      \begin{enumerate}
+      \item {{id_{b1<->b1}(+)id_{b2<->b2} = id_{b1+b2<->b1+b2} }}.
+      \item {{(f(+)g)(;)(j(+)k) = (f (;) j) (+) (g (;) k)}}.
+      \end{enumerate}
+    \item We need to check the naturality of {{assocl+}}, {{zeroe}} and
+      {{swap+(;)zeroe}}.
+      \begin{enumerate}
+      \item {{assocl+ (;) ((f (+)g) (+) h) = (f (+)(g (+) h)) (;) assocl+}}.
+      \item {{zeroe (;) f = (id (+) f) (;) zeroe}}.
+      \item {{(swap+ (;) zeroe) (;) f = (f (+) id) (;) (swap+ (;) zeroe)}}.
+      \end{enumerate}
+    \item Satisfy certain coherence conditions which are usually
+      called the ``pentagon'' and ``triangle'' axioms (see Sec 3.1
+      \cite{springerlink:10.1007/978-3-642-12821-94})
+    \end{enumerate}
+
+    The last three points require checking equality of combinators by
+    writing out their derivation trees as we did in the case of
+    associativity of sequential composition. To show symmetry, we need
+    a braiding operation {{b1+b2 <-> b2+b1}} which is given by
+    {{swap+}} (see Sec. 3.3 and 3.5
+    \cite{springerlink:10.1007/978-3-642-12821-94}).
+
+    \begin{enumerate}
+    \item The braiding must satisfy two ``hexagon'' axioms.
+    \item The braiding is self inverse,
+      {{swap+_{b1+b2}=(swap+_{b2+b1})^{dagger} }}.
+    \end{enumerate}
+
+  \end{proof}
+
+\end{lemma}
+
+\begin{lemma}[Symmetric Monoidal {{(*, 1)}}]
+  {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
+  unit {{1}}. The details mirror those of the {{(0, +)}} monoid.
+\end{lemma}
+
+\noindent
+Some technical and pedantic comments are due at this point.
+
+\begin{itemize}
+\item We have established the categorical structure of {{Pi}} as a
+  dagger symmetric monoidal category with two monoidal structures,
+  {{(0, +)}} and {{(*, 1)}}. In Sec. \ref{sec:int} we will see that a
+  {{trace}} operator can be admitted in this category without any
+  change of expressiveness.
+
+\item To be pedantic, what we have shown is that the ``term model'' of
+  {{Pi}} that follows from the extensional operational equality of
+  combinators has the requisite categorical structure. A consequence
+  is that the ``wiring diagrams'' of {{Pi}} correspond closely with ``string
+  diagrams'' developed for categories.
+
+  To establish the later rigorously, we will need to show when it is
+  valid to slide one wire over the other and that equivalent diagrams
+  for syntactically different combinators such as
+  {{(f(+)g)(;)(j(+)k)}} and {{(f(;)j)(+)(g(;)k)}} do respect
+  operational equivalence.  We don't formalize the graphical notation
+  in this work. Joyal et. al's work on ``planar isotopy'' and
+  Selinger's survey \cite{springerlink:10.1007/978-3-642-12821-94}
+  show how this has been addressed before in the categorical
+  setting. In the absence of any prior knowledge of category theory
+  however, our wiring diagrams may be read as the ``flow of types'' in
+  a combinator-circuit.
+
+\item By definition, in {{Pi}} every morphism is an isomorphism --
+  this makes {{Pi}} a groupoid.
+
+\item The category {{Pi}} has no initial and terminal objects. The
+  objects {{0}} and {{1}} would be initial and terminal if we admitted
+  all functions (as in the category {{CatSet}}).
+
+\item The category {{Pi}} has neither categorical products, nor
+  categorical co-products, i.e. {{*}} and {{+}} are merely monoidal
+  tensors. This follows from the fact that injection and projection
+  maps are not isomorphisms.
+\end{itemize}
+
+While we don't do so in this work, if we extend {{Pi}} with products
+and co-products (say through the addition of information effects) it
+is conceivable that part of its structure collapses. This sort of
+collapse, while catastrophic for algebraic structures (effectively
+trivializing them), still retains some interest in computing, because
+in computing we are interested in the specific operational nature (the
+computational content, so to speak) of the morphisms. Anecdotal evidence
+follows from the fact several real-world programming language have
+inconsistent type systems. blah blah blah...
 

### 2012-06-28 code/modular/ND.hs
?

@@ -245,5 +245,4 @@ test5 = eval justTimes2 (Left (), Right ())
 -- inv :: (PiInv eq) => eq a a
 -- inv = timesOneR %. (etaTimes %* id) %. (epsTimes %* id) %. timesOneL
 
-
 ------------------------------------------------------------------------------

### 2012-06-28 rational_ext.tex
small tweaks in section 2 to make things more precise

@@
 
-For example {{ swap* }} induces a function 
+For example {{ swap* }} induces a function (akin to)
 %subcode{code}
@@
 % swapProd (a,b) = (b,a)
+
 The isomorphisms are extended to form a congruence relation by adding the
@@
 
+It is important to note that ``values'' and ``isomorphisms'' are 
+completely separate syntactic categories which do not intermix.  The 
+semantics of the language come when these are made to interact at
+the ``top level'', a new syntactic category:
+%subcode{bnf} include main
+% top-level term, l ::= c v
+We refer to this as \emph{application}.
+
 \noindent
@@
 % iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+% top-level term, l ::= c v
 \end{definition}
@@
 
-The operational semantics of {{Pi}} is summarized below (also see
-\cite{infeffects}).  The semantics of the primitive combinators is
-given by the following single-step reductions below. Since there are
-no values of type {{0}}, there are no rules for the impossible cases:
+The operational semantics of top-level terms of {{Pi}} is summarized below
+(also see \cite{infeffects}).  The semantics of applying the primitive
+combinators to a value is given by the following single-step reductions below.
+Since there are no values of type {{0}}, there are no rules for the impossible
+cases:
 \begin{scriptsize}
@@
 
-\begin{proposition}[Strong Normalizing]
+\begin{proposition}[Strongly Normalizing]
 \label{prop:termination-pi} 
@@
 {{c2:b1<->b2}}, we can write the conditional combinator
-{{if_{c1,c2}:bool*b1<->bool*b2}} textually as 
-{{dist(;)((id(*)c1)(+)(id(*)c2))(;)factor}} and graphically as:
-
-\dgm{if-c1-c2}
+{{if_{c1,c2}:bool*b1<->bool*b2}} as 
+{{dist(;)((id(*)c1)(+)(id(*)c2))(;)factor}}.
+
+% \dgm{if-c1-c2}
 
@@
 
-Choosing {{C' =Lft c2 C}}, we have the required derivation sequence: 
+Choosing {{C' =LeftP c2 C}}, we have the required derivation sequence: 
 {{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1,v1', LeftP c2 C]} |--> {[c1+c2,left v1',C]} }}. The proof follows similarly for the {{right v2}} case.
@@
 a graphical language that conveys the intuitive semantics of the
-language (which is formalized in the next section).
+language.
 

### 2012-06-29 code/dual/Field.hs
meadow

@@ -1,0 +1,575 @@
+{-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts #-} -- 7.0.1, 7.0.3
+
+-- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
+-- Dont know: This code tested with GHC version 6.12.3, and version 7.0.1
+
+module Field where
+
+import Data.Typeable
+import Unsafe.Coerce -- needed for polymorphic lookup. 
+
+
+-----------------------------------------------------------------------
+-- Base types 
+
+data Zero 
+data Neg a = Ng a
+data Inv a = In a
+
+type (:+:) a b = Either a b
+type (:*:) a b = (a, b)
+
+data Three = One 
+           | Two 
+           | Three
+             deriving (Eq, Show)
+
+-----------------------------------------------------------------------
+-- Values 
+-- 
+-- Values of the form 'V a' are values used by the interpreter. 
+
+
+-- Values can contain non-grounded fresh variables
+type Var a = Int
+data V a where
+    Unit :: V () 
+    Pair :: V a -> V b -> V (a, b) 
+    L :: V a -> V (Either a b) 
+    R :: V b -> V (Either a b) 
+    Neg :: V a -> V (Neg a)
+    Inv :: V a -> V (Inv a)
+    Lift :: (Eq a, Show a) => a -> V a
+    Fresh :: Var a -> V a
+
+
+instance Show (V a) where 
+    show Unit = "()"
+    show (Pair a b) = "(" ++ show a ++ ", " ++ show b ++ ")"
+    show (L a) = "L " ++ (show a) 
+    show (R a) = "R " ++ (show a)
+    show (Neg a) = "Neg " ++ (show a)
+    show (Inv a) = "Inv " ++ (show a)
+    show (Lift a) = show a
+    show (Fresh n) = show n
+
+------------------------------------------------------------------------------------
+-- Translate a normal value to value usable by the interpreter.
+-- a -> V a
+
+class ToV a where 
+    makeV :: a -> V a
+
+instance ToV () where 
+    makeV () = Unit
+
+instance (ToV a, ToV b) => ToV (Either a  b) where 
[…]

### 2012-06-29 code/dual/Meadow1.hs

@@ -3,7 +3,28 @@
 -- {-# OPTIONS_GHC -fglasgow-exts #-} -- 6.12.3
 -- Dont know: This code tested with GHC version 6.12.3, and version 7.0.1
 
-module OldDual where
+module Meadow where
+
+-- These meadows are 
+-- 
+-- CR axioms
+-- (x+y)+z = x+(y+z)
+-- x+y = y+x
+-- x+0 = x
+-- x+(-x)=0
+-- (x.y).z = x.(y.z)
+-- x.y = y.x
+-- x.1 = x
+-- x.(y+z) = x.y + x.z
+-- 
+-- Along with 
+-- 
+-- Axiom Ref (reflection)
+-- 1/1/x = x
+-- Axiom Ril (restricted inverse)
+-- x.(x.1/x) = x
+
+
 
 import Data.Typeable
 import Unsafe.Coerce -- needed for polymorphic lookup. 
@@ -107,13 +128,13 @@ data a :<=> b where
   AssocTimesL  :: (a,(b,c)) :<=> ((a,b),c)
   AssocTimesR  :: ((a,b),c) :<=> (a,(b,c))
 -- (*) distributes over (+) 
-  TimesZeroL  :: (Zero, a) :<=> Zero
-  TimesZeroR  :: Zero :<=> (Zero, a)
+--  TimesZeroL  :: (Zero, a) :<=> Zero
+--  TimesZeroR  :: Zero :<=> (Zero, a)
   Distribute  :: (Either b c, a) :<=> Either (b, a) (c, a)
   Factor      :: Either (b, a) (c, a) :<=> (Either b c, a)
 -- Eta and Eps over the monoid (*, 1)
-  EtaTimes:: () :<=> (Inv a, a)
-  EpsTimes :: (Inv a, a) :<=> ()
+--  EtaTimes:: () :<=> (Inv a, a)
+--  EpsTimes :: (Inv a, a) :<=> ()
 -- EtaTimesand EpsTimes over the monoid (+, 0)
   EtaPlus :: Zero :<=> (Neg a :+: a)
   EpsPlus :: (Neg a :+: a) :<=> Zero
@@ -124,6 +145,13 @@ data a :<=> b where
 -- Encoding of Three
   FoldThree   :: Either () (Either () ()) :<=> Three
   UnfoldThree :: Three :<=> Either () (Either () ())
+-- Axioms Ref and Ril
[…]

### 2012-06-29 code/dual/Test.hs

@@ -1,6 +1,6 @@
 {-# OPTIONS_GHC -XGADTs -XTypeOperators -XExistentialQuantification -XFlexibleContexts -XScopedTypeVariables #-} -- 7.0.1, 7.0.3
 
-import OldDual 
+import Field
 
 ------------------------------------------------------------------------------------
 ------------------------------------------------------------------------------------

### 2012-06-29 code/dual/Meadow.hs
meadow

@@ -5,7 +5,8 @@
 
 module Meadow where
 
--- These meadows are 
+-- Meadows are a sort of zero totalized fields. 
+-- http://arxiv.org/abs/0901.0823
 -- 
 -- CR axioms
 -- (x+y)+z = x+(y+z)

### 2012-06-29 rational_ext.tex
negatives

@@
 
-What is a negative type anyway?
+Charles Pinter in his book on Abstract Algebra talks about how
+negatives were used by \emph{logisticians} of a 1000 years ago used
+negative numbers as a ``useful fiction''.  The concept of negative
+numbers were ``discovered''much later than fractionals. Like Joan Baez
+says ``half an apple is easier to undertand than a negative apple''.
+
+Negative types should correspond to sets with negative arity. But what
+does that mean? Various approaches have been proposed -- Loeb uses the
+notion of multiplicity functions to express negative sized sets
+[CITE]. These notions have been extended to Euler characteristic of
+categories by Leinster, Joyal and Schanuel -- negative values are
+explained sometimes as eqaulities relating otherwise indepedant
+objects -- Baez tells a story about ``island'' and positive and
+``bridges connecting islands'' as negatives (since the effectively
+reduce trhe number of disconnected elements.)
+
+We use the use the syntax {{-b}} for negative types and {{-v}} for
+negative values and the axiom {{0 <-> (-b)+b}} as the basis for
+negatives. In {{Pi}}/{{PiEE*}}, values flowed from left-to-right in
+circuits. Negative types are negative in the sense that they flow in
+the oppostite direction.
 
@@
 
-For convenience, we sometimes use the notations {{b1 - b2}} and
-{{b1/b2}} to indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}}
-respectively.  The types of the new constructs are:
+For convenience, we sometimes use the notations {{b1 - b2}} to
+indicate the types {{b1 + (-b2)}}.  The types of the new constructs
+are:
 \vspace{-15pt}
@@
 % {[eta+, left (-v), C, s]}^^^ &|-->& {[eta+, right v, C ]}
-
   \end{enumerate}
 
+Consider the combinator {{c:(-1+1)+1<->1}} defined below. On providing
+the value {{right ():(-1+1)+1}} we get the output {{():1}}. But on
+supply input {{left (right ())}} evaluation stop on the reverse
+interpreter with value {{left (left ()):(-1+1)+1}}.
+
+By definition, since negative values flow in the opposite direction of
+positive values, this is not surprising. Combinators can terminate in
+the forward interpreter (with result of type {{b2}}) or in the reverse
+interpreter (with result of type {{b1}}). To define an {{eval}} in a
+consistent manner without worrying about state of the evaluator, we
+transform every combinator of the form {{c:b1<->b2}} to one of the
+form {{c:-b2+b1<->0}} by the construction {{action(c)=id(+)c(;)eps+}}.
+
+\begin{definition}[{{eval(c, v)}}]
+For {{c:b1<->b2}} and {{v:b1}} we define {{eval(c, v)=v'}} such that
+\begin{enumerate}
+\item {{v':-b2+b1}} and
+\item {{<action(c), right v, []> |-->* <action(c),v',[]>^^^}}
+\end{enumerate}
+\end{definition}
+
+\begin{definition}[{{c1=c2}}]
+  For {{c1:b1<->b2}} and {{c2:b1<->b2}}, we define {{c1=c2}} if
+  {{~forall v:b1.eval(c1,v) = eval(c2,v)}}.
+\end{definition}
+
 \begin{proposition}[Logical Reversibility]
@@
 
-\begin{proposition}[Coherence]
+%%%%%%%%%
+\subsection{Categorical Structure}
+% \label{sec:neg-constructions}
+
+As before, we can show that {{PiEE+}} is a bimonoidal
+category. Additionally it has a compact closed structure in the
+{{(0,+)}} monoid.
+
+\begin{proposition}[Compact Closed {{(0,+,-)}}]
+  {{PiEE+}} is a compact closed category with over the symmetric
+  monoid {{(0,+)}} with the dual of objects {{b}} give by {{-b}} with
+  morphisms:
+
+\begin{enumerate}
+\item {{0 <-> -b + b}} given by {{eta+}}, called the `unit map'.
+\item {{b + -b <-> 0}} given by {{swap+ (;) eps+}}, called the `counit map'.
+\end{enumerate}
+
+\noindent
+such that the following coherence conditions hold (see 2.1 in
+\cite{selinger2007dagger}):
+
+\begin{enumerate}
+\item {{zeroi (;) swap+ (;) (id (+) eta+) (;) assocl+ (;) ((swap+ (;) eps+) (+) id) (;) zeroe = id : b <-> b}}
+\item {{zeroi (;) (eta+ (+) id) (;) assocr+ (;) (id (+) (swap+ (;) eps+)) (;) swap+ (;) zeroe = id : -b <-> -b}}
+\end{enumerate}
 

### 2012-06-29 rational_ext.tex
meadows

@@
 
-Without the important side condition on division by zero, we would
-have {{1<->1/0*0}} which implies {{1 <-> 0}} consequently collapsing
-the entire structure. In this context, an important area for future
-investigation is to consider alternatives to fields as a model for
-fractionals and negatives -- particularly algebraic structures that
-are ``zero totalized fields'' such as meadows
-\cite{Bergstra:2009:MES:1507774.1508000}.
-
 \roshan{This is an important question -- is {{PiEE*}} still
@@
 %% essentially the constructions in the introduction.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Zero Totalized Fields}
+
+Without the important side condition on division by zero, we would
+have {{1<->1/0*0}} which implies {{1 <-> 0}} consequently collapsing
+the entire structure.
+
+While fields require the ``no division by zero'' constraint, alternate
+algebraic structures such as meadows
+\cite{Bergstra:2009:MES:1507774.1508000} have been studied which admit
+division and negative without this constraint. Such structures are
+called ``zero totalized fields'' and define multiplicative inverse of
+{{0}}. The axiomatic presentation of meadows involve removing the
+combinators corresponding the axiom {{0*b<->0}} (i.e. {{dist0}} and
+{{factor0}}) and {{1<->1/b*b}} (i.e. {{eta*}} and {{eps*}}). Instead
+meadows include axioms \emph{reflexivity}, {{refl}} and {{refr}}, and
+\emph{restricted inverse}, {{ril}} and {{rir}}, which are:
+%subcode{opsem} include main
+% refl :& 1/1/b <-> b :& refr
+% ril :& b*(b*1/b) <-> b :& rir
+
+\noindent
+Combinators {{refl}} and {{refr}} are primitive and can be defined as:
+%subcode{opsem} include main
+% refl 1/(1/v) &|-->& v
+% refr v &|-->& 1/(1/v)
+
+\noindent
+Combinators {{ril}} and {{rir}} are defined as machine transitions
+similar to {{eta*}}/{{eps*}}:
+%subcode{opsem} include main
+% <ril, (v,(v',1/v')), C> &|-->& {[ril, v, C]}
+% <rir, v, C> &|-->& forall v':b.{[rir, (v, (v',1/v')), C]}
+
+While not as commonly used as fields, zero totalization is useful not
+just because we are spared from continuously checking the division
+constraint, but also because many other structures such as the
+category of vectors spaces do admit a dual to the 0 vector
+space. However their categorization is not well developed and we
+however lose the nice correspondence that {{eta*}}/{{eps*}} have with
+compact closed structures.
 
@@
 numbers were ``discovered''much later than fractionals. Like Joan Baez
-says ``half an apple is easier to undertand than a negative apple''.
+says ``half an apple is easier to understand than a negative apple''.
 
@@
 categories by Leinster, Joyal and Schanuel -- negative values are
-explained sometimes as eqaulities relating otherwise indepedant
+explained sometimes as equalities relating otherwise independent
 objects -- Baez tells a story about ``island'' and positive and
 ``bridges connecting islands'' as negatives (since the effectively
-reduce trhe number of disconnected elements.)
+reduce the number of disconnected elements.)
 
@@
 circuits. Negative types are negative in the sense that they flow in
-the oppostite direction.
+the opposite direction.
 
@@
 \begin{definition}[Size of a type]
-The defintion of size is extended to include negative types.
+The definition of size is extended to include negative types.
 {{[^-b^] = -[^b^]}}

### 2012-06-29 fields.tex
omake

--- 2012-06-28 rational_ext.tex
@@
 
-\title{Negative and Fractional Types}
+\title{Computing in the Field of Rationals}
 
@@
 
-\begin{verbatim}
-revise the paragraph below to 
-say something that's consistent with the 
-presentation in the previous section; the
-previous type system was the natural numbers;
-we presented a natural language of terms
-for that type system and a semantics in 
-the rig category of finite
-sets and bijections (two symmetric
-monoidal structures and distributivity).
-now we extend the type system to correspond
-to the non-negative rational numbers;
-we want to find the corresponding language
-and semantic interpretation
-\end{verbatim}
-
 In arithmetic the main identity satisfied by fractional numbers is that for
@@
 categories.
-
-%%%%%%%%%%%%%%%%%
-\subsection{Non-determinism}
-
-In order to interpret fractional types, we first lift the values and
-combinators of {{Pi}} to the world of sets and relations. The lifting is
-realized using a monad whose functor is powerset. Intuitively:
-\begin{itemize}
-\item instead of interpreting a type {{b}} as a finite set of values $\{ v_1,
-  v_2, \ldots, v_{\modulusp{b}}\}$, we now interpret each type {{b}} as the
-  powerset of such values; note that the type {{0}} now denotes
-  $\{\emptyset\}$, i.e., it has one element which is the empty set;
-  \item a {{Pi}}-value {{v}} of type {{b}} is lifted to a singleton set 
-  $\{ v \}$;
-\item each combinator {{c : b1 <-> b2}} is lifted to a \emph{symmetric}
-  relation $R_c$ defined as follows:
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\{ v_1, v_2, \ldots \} &R_c& \bigcup \{ c(v_i) \} 
-\end{array}\]
-For example, the combinator {{swap+ : 1+1 <-> 1+1}} is lifted to the relation $R$:
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\emptyset &R& \emptyset \\
-\{ \mathit{left}~() \} &R& \{ \mathit{right}~() \} \\
-\{ \mathit{right}~() \} &R& \{ \mathit{left}~() \} \\
-\{ \mathit{left}~(), \mathit{right}~() \} &R& \{ \mathit{left}~(), \mathit{right}~() \} 
-\end{array}\]
-\end{itemize}
-Because every combinator corresponds to a bijection, the lifted relation is
-symmetric.  Formally we have moved from the category of finite sets and
-bijections to the category of finite sets and \emph{symmetric relations}.
-
-Now that we have introduced the potential for non-determinism, it is possible
-to introduce a type {{1/b}}, values {{1/v}}, and combinators {{eta*}} and
-{{eps*}} that witness the following isomorphism:
-%subcode{opsem} include main
-% eta* &: 1 <-> (1/b) * b :& eps*
-where {{b}} is not ``equivalent'' to {{0}} as detailed next.
-
-Intuitively, we view a value {{1/v}} as a \emph{constraint} that is only
-satisfied if it is matched with a value {{v'}} that is identical to
-{{v}}. The type {{1/b}} is interpreted as the powerset of all the values
-{{1/v}} with {{v}} being a {{Pi}}-value of type {{b}}. For example, the type
-{{1/(1+1)}} denotes:
-\[
-\{ 
-\emptyset, 
-\{ 1/\mathit{left}~() \}, 
-\{ 1/\mathit{right}~() \},
-\{ 1/\mathit{left}~(), 1/\mathit{right}~() \} 
-\}
-\]
-Of particular interest is that both the type {{0}} and the type {{1/0}}
-denote $\{\emptyset\}$.
-
-Given this interpretation of types and values, the realization of 
-{{eta* : 1 <-> (1/b) * b}} is the relation $R$: 
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\emptyset &R& \emptyset \\
-\{ () \} &R& \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \}
-\end{array}\]
-For example, at type $1+1$, the interpretation of {{eta*}} is the relation:
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\emptyset &R& \emptyset \\
-\{ () \} &R& \{ (1/\mathit{left}~(),\mathit{left}~()), 
-                (1/\mathit{right}~(),\mathit{right}~()) \} 
-\end{array}\]
-At type $0$, the interpretation of {{eta*}} is:
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\emptyset &R& \emptyset \\
-\{ () \} &R& \emptyset 
-\end{array}\]
-which is not a symmetric relation. Hence the presence of {{eta*}} introduces
-some ``non-reversible'' effect and moves us to the category of finite sets 
-and (not necessarily symmetric) relations. The interpretation of 
-{{eps* : (1/b) * b <-> 1}} is the relation:
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\emptyset &R& \emptyset \\
-\{ \{ (1/v,v) ~|~ v \mbox{~is~a~} \Pi\mbox{-value~of~type~} b \} &R& \{ () \} \\
-\{ \{ (1/v,v') ~|~ v \mbox{~is~not~identical~to~} v' \} &R& emptyset 
-\end{array}\]
-For example, at type $0$, we have:
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\emptyset &R& \emptyset 
-\end{array}\]
-At type $1+1$, we have:
-\[\begin{array}{r@{\qquad}c@{\qquad}l}
-\emptyset &R& \emptyset \\
-\{ \{ (1/\mathit{left}~(),\mathit{left}~()) \} &R& \{ () \} \\
-\{ \{ (1/\mathit{left}~(),\mathit{right}~()) \} &R& \emptyset \\
-\{ \{ (1/\mathit{right}~(),\mathit{left}~()) \} &R& \emptyset \\
-\{ \{ (1/\mathit{right}~(),\mathit{right}~()) \} &R& \{ () \} \\
-\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{left}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
-  \{ () \} \} \\
-\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{left}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{right}~(),\mathit{left}~()), (1/\mathit{right}~(),\mathit{right}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{left}~()) \} &R&
-  \{ () \} \} \\
-\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{left}~(),\mathit{right}~()) \} &R&
-  \emptyset \} \\
-\{ \{ (1/\mathit{right}~(),\mathit{right}~()), (1/\mathit{right}~(),\mathit{left}~()) \} &R&
-  \emptyset \} 
-\end{array}\]
-
-The category of finite sets and relations is a compact closed category. Each
-object has a dual which is itself. The main coherence condition requires us
-to check that:
-\[\begin{array}{rcl}
-b &\rightarrow& b * 1 \\
-  &\rightarrow_{\eta^*}& b * (1/b * b) \\
-  &\rightarrow& (b * 1/b) * b \\
-  &\rightarrow_{\epsilon^*}& 1 * b \\
-  &\rightarrow& b
-\end{array}\]
-For example, let's check one instance of this identity at the type $1+1$ and
-the value {{left ()}}:
-\[\begin{array}{rcl}
-\{ \mathit{left}~() \} &\rightarrow& \{ (\mathit{left}~(), ()) \} \\
-  &\rightarrow_{\eta^*}& \{ (\mathit{left}~(),(1/\mathit{left}~(),\mathit{left}~())), 
-                            (\mathit{left}~(),(1/\mathit{right}~(),\mathit{right}~())) \} \\
-  &\rightarrow& \{ ((\mathit{left}~(),1/\mathit{left}~()),\mathit{left}~())), 
-                   ((\mathit{left}~(),1/\mathit{right}~()),\mathit{right}~())) \} \\
-  &\rightarrow_{\epsilon^*}& \{ ((),\mathit{left}~()) \} \\
-  &\rightarrow& \{ \mathit{left}~() \} 
-\end{array}\]
-At type $0$, we would get:
-\[\begin{array}{rcl}
-\emptyset &\rightarrow& \emptyset \\
-  &\rightarrow_{\eta^*}& \emptyset \\
-  &\rightarrow& \emptyset \\
-  &\rightarrow_{\epsilon^*}& \emptyset \\
-  &\rightarrow& \emptyset
-\end{array}\]
-
-%%%%%%%%%%%%%%%%
-\subsection{Syntax}
 

### 2012-06-29 fields.tex

@@
 
-What is a negative type anyway?
+Charles Pinter in his book on Abstract Algebra talks about how
+negatives were used by \emph{logisticians} of a 1000 years ago used
+negative numbers as a ``useful fiction''.  The concept of negative
+numbers were ``discovered''much later than fractionals. Like Joan Baez
+says ``half an apple is easier to understand than a negative apple''.
+
+Negative types should correspond to sets with negative arity. But what
+does that mean? Various approaches have been proposed -- Loeb uses the
+notion of multiplicity functions to express negative sized sets
+[CITE]. These notions have been extended to Euler characteristic and
+Homotopy cardinality of categories by Leinster, Joyal and Schanuel --
+negative values are explained sometimes as equalities relating
+otherwise independent objects -- Baez tells a story about ``island''
+and positive and ``bridges connecting islands'' as negatives (since
+the effectively reduce the number of disconnected elements.)
+
+We use the use the syntax {{-b}} for negative types and {{-v}} for
+negative values and the axiom {{0 <-> (-b)+b}} as the basis for
+negatives. In {{Pi}}/{{PiEE*}}, values flowed from left-to-right in
+circuits. Negative types are negative in the sense that they flow in
+the opposite direction.
 

### 2012-07-05 fields.tex
_

@@
 
+  Previous work on information effects introduced the computational
+  model {{Pi}} which lacked a natural notion of first-class functions.
   This paper connects the line of work starting with
@@
   information preserving computation \cite{infeffects,rc2011}.
-  Andrzej Filinski's Masters thesis
-  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
-  symmetry underlying computation in the \lcal. Filinski showed that
-  values are dual to continuations and that functions are dual to
-  delimited continuations, thereby suggesting that
-  values-and-functions and continuations-and-delimited-continuations
-  are mirror images of the same phenomena.
-
-  Previous work on \emph{information effects} established the notion
-  of information preserving computation \cite{infeffects} and the core
-  calculus for computing with the isomorphisms of finite types,
-  {{Pi}}. Given the first-order nature of {{Pi}} however it was
-  unclear how one may express first-class functions and duality in the
-  sense of Filinski.
+
+  %% Andrzej Filinski's Masters thesis
+  %% \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
+  %% symmetry underlying computation in the \lcal. Filinski showed that
+  %% values are dual to continuations and that functions are dual to
+  %% delimited continuations, thereby suggesting that
+  %% values-and-functions and continuations-and-delimited-continuations
+  %% are mirror images of the same phenomena.
+
+  %% Previous work on \emph{information effects} established the notion
+  %% of information preserving computation \cite{infeffects} and the core
+  %% calculus for computing with the isomorphisms of finite types,
+  %% {{Pi}}. Given the first-order nature of {{Pi}} however it was
+  %% unclear how one may express first-class functions and duality in the
+  %% sense of Filinski.
 
@@
   additive duality, namely negative types and a multiplicative
-  duality, namely fractional types.
+  duality, namely fractional types. Intuitively, values of negative
+  types are values that flow ``backwards'' to satisfy demands and
+  values of fractional types are values that impose constraints on
+  their context.
 
@@
 \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
-underlying computation in the \lcal. Filinski showed that values are
-dual to continuations, functions are dual to delimited continuations
-and a De Morgan style duality by treating pairs of values as dual to
-sums (choice) of continuation and sums of values as dual to pairs of
-continuations. He, furthermore, introduced a symmetric extension of
-the $\lambda$-calculus in which call-by-value is dual to call-by-name.
-
-Symmetry between the value fragment and the continuation fragment was
-refined by Curien and Herbelin \cite{Curien:2000} who introduced the
+underlying computation in the \lcal. Filinski inroduced a symmetric
+extension of the \lcal in which he showed that values are dual to
+continuations, functions are dual to delimited continuations, pairs of
+values as dual to sums (choice) of continuations
+({{(a*b)^bot<->a^bot+b^bot}}), sums of values are dual to pairs of
+continuations ({{(a+b)^bot<->a^bot*b^bot}}) and that call-by-value is dual to call-by-name.
+
+The symmetries exposed by Filinski suggested that values/continuations
+and functions/delimited continuations are different aspects of the same
+computational phenomenon and led to much work in this area.  Symmetry
+between the value fragment and the continuation fragment was refined
+by Curien and Herbelin \cite{Curien:2000} who introduced the
 sequent-style $\mu\tilde{\mu}$-calculus, that exhibits symmetries
 between values and continuations and between call-by-value and
-call-by-name. The calculus includes the type $A-B$ which is the dual
-of implication, i.e., a value of type $A-B$ is a context expecting a
-function of type $A \rightarrow B$. Alternatively a value of type
-$A-B$ is also explained as a \emph{pair} consisting of a value of type
-$A$ and a continuation of type $B$. This is to be contrasted with our
-interpretation of a value of that type as \emph{either} a value of $A$
-or a demand for a value of type $B$. The duality between call-by-name
-and call-by-value was further investigated by Selinger using control
-categories~\cite{Selinger:2001:CCD:966910.966911}. The
-$\mu\tilde{\mu}$-calculus was further analyzed and extended by
-Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives
-no interpretation to the subtraction connective and like the original
-symmetric calculus of Filinski, introduces a duality that relates sums
-to products and vice-versa.
+call-by-name.  The duality between call-by-name and call-by-value was
+further investigated by Selinger using control
+categories~\cite{Selinger:2001:CCD:966910.966911} and by
+Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}.
+
+%% The calculus includes the type $A-B$ which is the dual
+%% of implication, i.e., a value of type $A-B$ is a context expecting a
+%% function of type $A \rightarrow B$. Alternatively a value of type
+%% $A-B$ is also explained as a \emph{pair} consisting of a value of type
+%% $A$ and a continuation of type $B$. This is to be contrasted with our
+%% interpretation of a value of that type as \emph{either} a value of $A$
+%% or a demand for a value of type $B$.
+
+%% The extension gives no interpretation to the subtraction connective
+%% and like the original symmetric calculus of Filinski, introduces a
+%% duality that relates sums to products and vice-versa.
 
@@
 reversible computational model {{Pi}} (and its extension {{Pi0}})
-wherein computation is inherently information preserving. The
-computational model is derived from the isomorphisms of finite-types,
-every computation admits an adjoint that expresses the logical
-``inverse'' of a computation and is complete for combinational
-circuits. The term model of {{Pi}} is a so-called commutative
-bimonoidal category, i.e. it corresponds to the categorification of a
-commutative semiring with two commutative monoidal structures
-{{(0,+)}} and {{(*,1)}} with {{*}} distributing over {{+}}.
-
-The language {{Pi}} however lacks a natural representation for
-first-class functions due to the first-order nature of its values. In
-the light of the work of duality of computation it was felt that
-``right model'' for expressing functions in {{Pi}} would also express
-symmetries similar to those in the calculi of Filinski, Curien and
-Herbelin and Wadler. The technical goal of this paper is to present
-such a symmetric dual extension of {{Pi}}.
-
-The extended language, which we call {{PiEE}}, is interesting because
-it does not have one De Morgan style dualizing action $(\_)^\bot$, but
+wherein computation is information/entropy preserving. The computational model
+is derived from the isomorphisms of finite-types, is complete for
+combinational circuits and every computation admits an adjoint which
+is the ``inverse of the computation''. The term model of {{Pi}}
+corresponds to the categorification of a commutative semiring with two
+commutative monoidal structures {{(0,+)}} and {{(*,1)}} with {{*}}
+distributing over {{+}} and is sometimes called a commutative bimonoidal
+category.
+
+%% The language {{Pi}} however lacks a natural representation for
+%% first-class functions due to the first-order nature of its values. In
+%% the light of the work of duality of computation it was felt that
+%% ``right model'' for expressing functions in {{Pi}} would also express
+%% symmetries similar to those in the calculi of Filinski, Curien and
+%% Herbelin and Wadler.
+
+The technical goal of this paper is to present the compuational model
+{{PiEE}}, a Filinski-style symmetric dual extension of {{Pi}}.
+
+The
+extended language, which we call {{PiEE}}, is interesting because it
+does not have one De Morgan style dualizing action $(\_)^\bot$, but
 two distinct dualizing actions -- an additive duality characterized by
@@
 \begin{definition}[{{eval(c, v)}}]
-For {{c:b1<->b2}} and {{v:b1}} we define {{eval(c, v)=v'}} such that
+Given {{c:b1<->b2}} and {{v:b1}} we define {{eval(c, v)=v'}} such that
 \begin{enumerate}
-\item {{v':-b2+b1}} and
+\item {{right v:-b2+b1}}, {{v':-b2+b1}} and
 \item {{<action(c), right v, []> |-->* <action(c),v',[]>^^^}}

### 2012-07-05 fields.tex
fields

@@
 reversible computational model {{Pi}} (and its extension {{Pi0}})
-wherein computation is information/entropy preserving. The computational model
-is derived from the isomorphisms of finite-types, is complete for
-combinational circuits and every computation admits an adjoint which
-is the ``inverse of the computation''. The term model of {{Pi}}
-corresponds to the categorification of a commutative semiring with two
-commutative monoidal structures {{(0,+)}} and {{(*,1)}} with {{*}}
-distributing over {{+}} and is sometimes called a commutative bimonoidal
-category.
-
-%% The language {{Pi}} however lacks a natural representation for
-%% first-class functions due to the first-order nature of its values. In
-%% the light of the work of duality of computation it was felt that
-%% ``right model'' for expressing functions in {{Pi}} would also express
-%% symmetries similar to those in the calculi of Filinski, Curien and
-%% Herbelin and Wadler.
+wherein computation is information/entropy preserving. The
+computational model is derived from the isomorphisms of finite-types,
+is complete for combinational circuits and every computation admits an
+adjoint which is the ``inverse of the computation''. The term model of
+{{Pi}} corresponds to the categorification of a commutative semiring
+with two commutative monoidal structures {{(0,+)}} and {{(*,1)}} with
+{{*}} distributing over {{+}} and is sometimes called a commutative
+bimonoidal category.
+
+The language {{Pi}} however lacks a natural representation for
+first-class functions due to the first-order nature of its values. The
+first approach to try is to re-use Filinski's duality (or one of its
+later variants, such duality in the linear logic setting).  The De
+Morgan style duality in Filinski's symmetric \lcal (and later works)
+is computationally interpreted as follows: if one has a pair of values
+{{(v1, v2)}}, then the context that consumes the value has a choice of
+comsuming either {{v1}} or {{v2}}. Hence the dual of {{(v1,v2)}} maybe
+thought of as the continuation that does {{fst}} or {{snd}}. Such a De
+Morgan style computational interpretation is inherently based on the
+deletion of values and completely collapses the logical reversibility
+and information preservation properties of {{Pi}}. Hence classical De
+Morgan duality and linear logic style De Morgan duality (which appears
+in later works on polarised focalized $\mu\tilde{\mu}$) do not apply
+directly in the context of {{Pi}}.
+
+The other obvious approach to try and build functions is to construct
+a ``closed'' category. It is well know that the Int-construction of
+Joyal et al \cite{joyal1996traced} and the $\mathcal{G}$ construction
+of Abramsky allow the construction of compact closed category
+(sometimes also called monoidal closed) from a traced symmetric
+monoidal category. The traced extension of {{Pi}}, called {{Pi0}} has
+been studied before and indeed the Int-construction gives us a compact
+closed structure in this setting. However, the Int-construction does
+not preserve the bimonoidal structure of {{Pi}}, i.e. if we use an
+additive trace, we lose the {{(1, *)}} monoid and distributivity.
+
+%% In the light of the work of duality of computation it was felt that
+%% ``right model'' for expressing functions in {{Pi}} would also
+%% express symmetries similar to those in the calculi of Filinski,
+%% Curien and Herbelin and Wadler.
 
 The technical goal of this paper is to present the compuational model
-{{PiEE}}, a Filinski-style symmetric dual extension of {{Pi}}.
-
-The
-extended language, which we call {{PiEE}}, is interesting because it
-does not have one De Morgan style dualizing action $(\_)^\bot$, but
-two distinct dualizing actions -- an additive duality characterized by
-negative types $-(\_)$ and a multiplicative duality characterized by
-fractional types $1/(\_)$. Each dualizing operation preserves the
-underlying connective {{-(a+b)<->(-a)+(-b)}} and
-{{1/(a*b)<->1/a*1/b}}. Negative and fractional types have a clear and
-intuitive semantics in the setting of {{Pi}}.
-
-The types of {{PiEE}} correspond to the \emph{field of rational
-  numbers} (with the important restriction of fields that there is no
-division by zero). Consequently algebraic manipulations of rationals
-can be given an operational interpretation and \emph{have
-  computational content}.  These negative and fractional dualities
-must not confused with $(\-)^\bot$ duality in the additive and
-multiplicative fragment of linear logic.  Models of linear logic rely
-on *-autonomous categories which change the monoidal tensor when
-dualizing. Compact closed categories preserve the monoidal tensor on
-dualizing.  The term model of {{PiEE}} corresponds to the
-\emph{categorification of a field}. The category is compact closed
-over the additive monoid. In the multiplicative monoid (since there is
-no division by zero) respects compact closed structure only when a
-dual object is defined.
-
-Say more: fractionals, entanglements, backtracking, superposition,
-constraints, negative information. Negative types, backward information
-flows, sets with negative types, iteration.. etc.
+{{PiEE}}, a Filinski-style symmetric dual extension of {{Pi}}.  The
+extended language has two distinct dualizing actions -- an additive
+duality characterized by negative types $-(\_)$ and a multiplicative
+duality characterized by fractional types $1/(\_)$. Each dualizing
+operation preserves the underlying connective {{-(a+b)<->(-a)+(-b)}}
+and {{1/(a*b)<->1/a*1/b}}.  The types of {{PiEE}} correspond to the
+\emph{field of rational numbers}. Consequently algebraic manipulations
+of rationals can be given an operational interpretation and \emph{have
+  computational content}.
+
+%% Negative and fractional types have an
+%% intuitive semantics in the setting of {{Pi}}.
+
+The term model of {{PiEE}} corresponds to the \emph{categorification
+  of a field}. The category is compact closed over the additive
+monoid. In the multiplicative monoid (since there is no division by
+zero) respects compact closed structure on non-zero objects.  Compact
+closed categories preserve the monoidal tensor on dualizing. In other
+words, negative and fractional dualities must not confused with
+$(\-)^\bot$ duality in the additive and multiplicative fragment of
+linear logic.  Models of linear logic rely on *-autonomous categories
+which change the monoidal tensor when dualizing.
+
+%% Say more: fractionals, entanglements, backtracking, superposition,
+%% constraints, negative information. Negative types, backward information
+%% flows, sets with negative types, iteration.. etc.
 
@@
 
-\begin{comment}
-
-In a recent paper~\cite{infeffects}, we argued that, because they include
-irreversible physical primitives, conventional abstract models of computation
-have inadvertently included some \emph{implicit} computational effects which
-we called \emph{information effects}. We then developed a pure reversible
-model of computation that is obtained from the type isomorphisms and
-categorical structures that underlie models of linear logic and quantum
-computing and that treats information as a linear resource that can neither
-be erased nor duplicated. In this paper, we show that our pure reversible
-model unveils deeper and more elegant symmetries of computation than have
-previously been reported. In particular, we expose two notions of duality of
-computation: an additive duality and a multiplicative duality that give rise
-to negative types and fractional types respectively. Although these types
-have previously appeared in the literature (see Sec.~\ref{sec:related}), they
-have typically appeared in the context of conventional languages with
-information effects, which limited their appeal and obscured their properties.
+% \begin{comment}
+
+%% In a recent paper~\cite{infeffects}, we argued that, because they
+%% include irreversible physical primitives, conventional abstract
+%% models of computation have inadvertently included some
+%% \emph{implicit} computational effects which we called
+%% \emph{information effects}. We then developed a pure reversible
+%% model of computation that is obtained from the type isomorphisms
+%% and categorical structures that underlie models of linear logic and
+%% quantum computing and that treats information as a linear resource
+%% that can neither be erased nor duplicated. In this paper, we show
+%% that our pure reversible model unveils deeper and more elegant
+%% symmetries of computation than have previously been reported. In
+%% particular, we expose two notions of duality of computation: an
+%% additive duality and a multiplicative duality that give rise to
+%% negative types and fractional types respectively. Although these
+%% types have previously appeared in the literature (see
+%% Sec.~\ref{sec:related}), they have typically appeared in the
+%% context of conventional languages with information effects, which
+%% limited their appeal and obscured their properties.
 
@@
 
-\end{comment}
+% \end{comment}
 

### 2012-09-14 iso-int/rc-iso-int.tex
_

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Parity Translation }
+\label{sec:parity-int}
+
+While the previous machines expressed traversals of the values with
+input values reflected in the outputs, the following machine
+translates one recursive type to another. Consider the type {{nat}}
+(defined to be {{@@x.1+x}}) and its parity encoded version {{parity}}
+(defined to be {{@@x.1+(1+x)}}). Algebraically, we may write:
+
+%subcode{bnf} include main
+% Numbers, n, m = 0 | n + 1
+% Parity, par = even | odd | A par
+
+While it is apparent that there is a natural one to one correspondence
+between their values, it is non-obvious how this can be expressed as a
+{{langRevT}} isomorphism. This map can be expressed in terms of the
+algebraic type definitions as shown below:
+
+%subcode{opsem} include main
+%! columnStyle = ||c|r||
+% 0 & even
+% 1 & odd
+% 2 & A even
+% 3 & A odd
+% 4 & A (A even)
+% ... &  ...
+
+\indent
+and in terms of their {{langRevT}} type encodings as shown below:
+
+%subcode{opsem} include main
+%! columnStyle = ||c|c||
+% <left ()> & <left ()>
+% <right <left ()>> & <right left ()>
+% <right <right <left ()> > > & <right right <left ()>>
+% <right <right <right <left ()>> >> & <right right <right left ()>>
+% <right <right <right <right <left ()>> >> > & <right right <right right <left ()> > >
+% ... &  ...
+
+\noindent
+Instead of directly programming in {{langRevT}}, one can derive a
+{{langRevT}} combinator for this purpose by first constructing an
+abstract machine that maps {{nat}} to {{parity}}.  A logically
+reversible abstract machine expressing the translation may be defined
+as follows:
+
+%subcode{bnf} include main
+% Machine states = <nat, nat> | {[parity, nat]}
+% Start state = <n , 0>
+% Stop state =  {[par, 0]}
+
+\noindent
+Machine Reductions:
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <0, m> &|-->& {[even, m]}
+% <1, m> &|-->& {[odd, m]}
+% <(n+1)+1, m> &|-->& <n, m+1>
+% {[par, m+1]} &|-->& {[A par, m]}
+
+The construction of the {{langRevT}} combinator is straightforward and
+proceeds as before. The partial combinator representing the wiring of
+machine reductions is shown below.
+
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/nat-parity.pdf}
+}
+\end{center}
+
+\noindent
+We thank Fritz Henglein for the motivation underlying this example.
+
 %%%%%%%%%%%%%%%%%%%%%

### 2012-09-14 iso-int/rc-iso-int.tex
_

--- 2012-06-01 iso-int/rc-iso-int.tex
@@
 
-\paragraph*{Steps of the Construction.}
+\subsection{Steps of the Construction.}
 
@@
 \item 
+ Machine states are described as tuples whose components are algebraic
+ data types. In the above example we used only the {{nat}} data type
+ and tuples of the form {{<n, m>}}.
+\item
   Machines must have distinguishable start and stop states. There may
@@
   state, however.
+
 \item Each valid machine state must match a unique pattern on the
   left-hand side and right-hand side of the `{{|-->}}' relation.
+
 \item Every reduction step must be computable (in {{langRevT}}), must
@@
 and output sides.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Parity Translation }
+\label{sec:parity-int}
+
+Deriving {{langRevT}} combinators from abstract machines can sometimes
+be used as an efficient indirect way of programming in {{langRevT}}.
+Say we wished to build a {{langRevT}} combinator to map {{nat}} to a
+parity encoded version {{par = @@x.1+(1+x)}}. Since {{par}} is not a
+fixed unfolding of {{nat}}, it is not immediately apparent how such a
+combinator can be constructed.
+
+Instead of directly programming in {{langRevT}}, one can derive the
+required {{nat<->par}} combinator by first constructing an abstract
+machine that maps {{nat}} to {{par}}. We first express {{nat}} and
+{{par}} algebraically and then develop a logically reversible abstract
+machine expressing the translation:
+
+% Algebraically, we may write:
+
+%% %subcode{bnf} include main
+%% % Numbers, n, m = 0 | n + 1
+%% % Parity, par = zero | one | plus2 par
+
+%% While it is apparent that there is a natural one to one correspondence
+%% between their values, it is non-obvious how this can be expressed as a
+%% {{langRevT}} isomorphism. This map can be expressed in terms of the
+%% algebraic type definitions as shown below:
+
+%% %subcode{opsem} include main
+%% %! columnStyle = ||c|r||
+%% % 0 & even
+%% % 1 & odd
+%% % 2 & A even
+%% % 3 & A odd
+%% % 4 & A (A even)
+%% % ... &  ...
+
+%% \indent
+%% and in terms of their {{langRevT}} type encodings as shown below:
+
+%% %subcode{opsem} include main
+%% %! columnStyle = ||c|c||
+%% % <left ()> & <left ()>
+%% % <right <left ()>> & <right left ()>
+%% % <right <right <left ()> > > & <right right <left ()>>
+%% % <right <right <right <left ()>> >> & <right right <right left ()>>
+%% % <right <right <right <right <left ()>> >> > & <right right <right right <left ()> > >
+%% % ... &  ...
+
+% \noindent
+
+%subcode{bnf} include main
+% Numbers, n, m = 0 | n + 1
+% Parity, par = even | odd | A par
+%
+
+\noindent
+Machine States:
+
+%subcode{bnf} include main
+% Machine states = <nat, nat> | {[parity, nat]}
+% Start state = <n , 0>
+% Stop state =  {[par, 0]}
+
+\noindent
+Machine Reductions:
+%subcode{opsem} include main
+%! columnStyle = rclr
+% <0, m> &|-->& {[even, m]}
+% <1, m> &|-->& {[odd, m]}
+% <(n+1)+1, m> &|-->& <n, m+1>
+% {[par, m+1]} &|-->& {[A par, m]}
+
+The construction of the {{langRevT}} combinator is
+straightforward and proceeds as before. The partial combinator
+representing the wiring of machine reductions is shown below.
+
+\begin{center}
+\scalebox{1.0}{
+  \includegraphics{diagrams/nat-parity.pdf}
+}
+\end{center}
+
+\noindent
+We thank Fritz Henglein for the motivation underlying this example.
 

### 2012-09-15 iso-int/rc-iso-int.tex
several corrections

@@
 \paragraph*{Universality.} 
-As an example, consider the implementation of the Toffoli gate below, which
-establishes that {{langRevT}} is complete for combinational circuits. This
-encoding uses the type {{1+1}} as {{bool}} and values {{left ()}} and
-{{right()}} as {{true}} and {{false}}:
-\begin{center}
+As an example, consider the implementation of the Toffoli gate below,
+which establishes that {{langRevT}} is complete for combinational
+circuits. This encoding uses the type {{1+1}} as {{bool}} and values
+{{left ()}} and {{right()}} as {{true}} and {{false}}.  Boolean
+negation not is simply {{swap+}}. The Toffoli gate takes three boolean
+inputs: if the first two inputs are {{true}} then the third is
+negated.
+
+Even though {{langRevT}} lacks conditional expressions, they are
+expressible using the distributivity laws as we demonstrate.  Given
+any combinator {{c : b <-> b}} we can construct a combinator called
+{{if_c : bool*b <-> bool*b}} in terms of {{c}}, where {{if_c}}
+behaves like a one-armed if-expression. If the supplied boolean is
+true then the combinator {{c}} is used to transform the value of type
+{{b}}. If the boolean is {{false}}, then the value of type {{b}}
+remains unchanged. We can write down the combinator for {{if_c}} in
+terms of {{c}} as {{distrib (;) ((id (*) c) (+) id) (;) factor}}.
+
+Given {{if_c}}, constructing the Toffoli gate is straightforward. If
+we choose {{c}} to be {{not}} we get {{if_{not} }} which is
+{{cnot:bool*bool<->bool*bool}} and {{if_{cnot} }} is the required
+{{toffoli : bool * (bool * bool) <-> bool * (bool * bool)}}, drawn
+below.
+\begin{center}
+\scalebox{1.2}{
   \includegraphics{diagrams/toffoli.pdf}
+}
 \end{center}
@@
 %subcode{bnf} include main
-% Numbers, n, m,p = 0 | n + 1
-% Machine states = <e, e, e>
+% Numbers, n, m, p = 0 | n + 1
+% Machine states = <n, n, n>
 
@@
   Sec.~\ref{sec:langrevt-int}, one could use generalized abstract data types
-  (GADTs)~\cite{phantom,Xi:2003:GRD:604131.604150}) to eliminate the stuck
+  (GADTs~\cite{phantom,Xi:2003:GRD:604131.604150}) to eliminate the stuck
   states.
@@
 
-We show the generality of our construction by applying it to two non-trivial
-examples: a tree traversal and a meta-circular interpreter for
-{{langRevT}}.
+We show the generality of our construction by applying it to three
+non-trivial examples: a tree traversal, parity translation of numbers
+and a meta-circular interpreter for {{langRevT}}.
 
@@
 be used as an efficient indirect way of programming in {{langRevT}}.
-Say we wished to build a {{langRevT}} combinator to map {{nat}} to a
-parity encoded version {{par = @@x.1+(1+x)}}. Since {{par}} is not a
-fixed unfolding of {{nat}}, it is not immediately apparent how such a
-combinator can be constructed.
+It is well known that every number {{n}} can be represented as
+{{2a+0}} or {{2a+1}} depending on whether it is odd or even.  The
+later can be viewed as the algebraic data type
+{{par=odd~|~even~|~A~par}} such that 
+{{0 = even, 1 = odd, 2 = A even, 3 = A odd, 4 = A (A even)}} etc where
+the nesting of {{A}} constructor indicates the value of {{a}}.
+
+Say we wish to build a {{langRevT}} combinator to map {{nat}} to its
+parity encoded version {{par}}, expressible as the {{langRevT}} type
+{{@@x.(1+1)+x}}. Since {{par}} is not a fixed unfolding of {{nat}}
+({{@@x.1+x}}), it is not immediately apparent how such a combinator
+can be constructed.
 
@@
 \scalebox{1.0}{
-  \includegraphics{diagrams/nat-parity.pdf}
+  \includegraphics{diagrams/nat-parity2.pdf}
 }

### 2012-09-15 iso-int/rc-iso-int.tex
several corrections

@@
   data type and are drawn as triangular boxes. For instance,
-  Sec. \ref{sec:bounded} shows {{fold}}/{{unfold}} isomorphisms for
+  Sec. \ref{sec:numeric-ops} shows {{fold}}/{{unfold}} isomorphisms for
   numbers, {{nat=@@x.(1+x)}}.
@@
 
-\paragraph*{Universality.} 
+\subsection{Universality.} 
 As an example, consider the implementation of the Toffoli gate below,
@@
 
+\subsection{Numeric Operations.}
+\label{sec:numeric-ops}
+
+In {{langRevT}} we encode numbers, {{nat = 0 ~|~ n + 1}}, using the
+recursive type {{@@x.1+x}}. The {{fold}}/{{unfold}} isomorphisms for
+{{nat}} are {{unfold : nat <-> 1+nat : fold}}.
+\begin{multicols}{3}
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/unfold_nat.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/fold_nat.pdf}
+}
+\end{center}
+
+\begin{center}
+\scalebox{0.8}{
+  \includegraphics{diagrams/add1.pdf}
+}
+\end{center}
+\end{multicols}
+The triangle on the left denotes the {{unfold}} isomorphism, which
+works as follows: If the number {{n}} is zero, the output is the top
+branch which has type~{{1}}. If the number is non-zero, the output is
+on the bottom branch and has value {{n-1}}. The triangle in the middle
+is the dual {{fold}} isomorphism.
+
+As explained in previous work
+\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}, numeric addition and
+subtraction can be expressed in {{langRevT}} as partial
+isomorphisms. The combinator on the right above is
+{{add1:nat<->nat}}. It acts like the successor function returning
+{{n+1}} for all inputs {{n}}. Its adjoint {{sub1:nat<->nat}}, which we
+get by flipping the diagram right to left, diverges on input {{0}} and
+returns {{n-1}} for all other inputs {{n}}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 writing a combinator {{c : nat * nat <-> nat * nat}} that executes one step
-of the machine transition. Here {{nat}} is an abbreviation for the recursive
-type {{@@x.1+x}}. The combinator, when iterated, should map {{(n,0)}} to
+of the machine transition. 
+%% Here {{nat}} is an abbreviation for the recursive
+%% type {{@@x.1+x}}. 
+The combinator, when iterated, should map {{(n,0)}} to
 {{(0,n)}} by precisely mimicking the steps of the abstract machine. Let us
@@
 
-The triangle on the left denotes the {{unfold}} isomorphism, which works as
-follows: If the number {{n}} is zero, the output is the top branch which has
-type~{{1}}. If the number is non-zero, the output is on the bottom branch
-and has value {{n-1}}. The triangle on the right is the dual {{fold}}
-isomorphism.  We now use distribution to isolate each possible machine state:
+We now use distribution to isolate each possible machine state:
 
@@
   unobservable in some way. If the specific case of the machine above, we can
-  use a primitive {{add1}} whose dual {{sub1}} is undefined when applied to
-  {{0}}. (Such a primitive is definable in
-  {{langRevT}}~\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}.)
+  use the primitive {{add1}} whose dual {{sub1}} is undefined when applied to
+  {{0}} (see Sec.~\ref{sec:numeric-ops}).
 

### 2012-09-15 iso-int/rc-iso-int.tex
many small edits

@@
 which establishes that {{langRevT}} is complete for combinational
-circuits. This encoding uses the type {{1+1}} as {{bool}} and values
-{{left ()}} and {{right()}} as {{true}} and {{false}}.  Boolean
-negation not is simply {{swap+}}. The Toffoli gate takes three boolean
-inputs: if the first two inputs are {{true}} then the third is
-negated.
+circuits. The Toffoli gate takes three boolean inputs: if the first
+two inputs are {{true}} then the third is negated. This encoding uses
+the type {{1+1}} as {{bool}} and values {{left ()}} and {{right()}} as
+{{true}} and {{false}}.  Boolean negation, {{not : bool<->bool}}, is
+simply {{swap+}}.
 
 Even though {{langRevT}} lacks conditional expressions, they are
-expressible using the distributivity laws as we demonstrate.  Given
-any combinator {{c : b <-> b}} we can construct a combinator called
-{{if_c : bool*b <-> bool*b}} in terms of {{c}}, where {{if_c}}
-behaves like a one-armed if-expression. If the supplied boolean is
-true then the combinator {{c}} is used to transform the value of type
-{{b}}. If the boolean is {{false}}, then the value of type {{b}}
-remains unchanged. We can write down the combinator for {{if_c}} in
-terms of {{c}} as {{distrib (;) ((id (*) c) (+) id) (;) factor}}.
+expressible using the distributivity laws.  Given any combinator
+{{c:b<->b}}, we can construct a combinator called
+{{if_c:bool*b<->bool*b}} in terms of {{c}}, which behaves
+like a one-armed if-expression. If the supplied boolean is {{true}} then
+the combinator {{c}} is used to transform the value of type {{b}}. If
+the boolean is {{false}} then the value of type {{b}} remains
+unchanged. We can write down the combinator for {{if_c}} in terms of
+{{c}} as {{distrib (;) ((id (*) c) (+) id) (;) factor}}.
 
 Given {{if_c}}, constructing the Toffoli gate is straightforward. If
-we choose {{c}} to be {{not}} we get {{if_{not} }} which is
-{{cnot:bool*bool<->bool*bool}} and {{if_{cnot} }} is the required
-{{toffoli : bool * (bool * bool) <-> bool * (bool * bool)}}, drawn
+we choose {{c}} to be {{not}} we get {{if_{not} }} which is controlled-not,
+{{cnot:bool*bool<->bool*bool}}. Further, {{if_{cnot} }} is the required
+Toffoli gate, 
+{{toffoli : bool * (bool * bool) <-> bool * (bool * bool)}}, and is drawn
 below.
@@
 
-In {{langRevT}} we encode numbers, {{nat = 0 ~|~ n + 1}}, using the
-recursive type {{@@x.1+x}}. The {{fold}}/{{unfold}} isomorphisms for
+We encode numbers, {{nat = 0 ~|~ n + 1}}, using the
+recursive {{langRevT}} type {{@@x.1+x}}. The {{fold}}/{{unfold}} isomorphisms for
 {{nat}} are {{unfold : nat <-> 1+nat : fold}}.
@@
 \begin{center}
-\scalebox{0.8}{
+\scalebox{0.9}{
   \includegraphics{diagrams/unfold_nat.pdf}
@@
 \begin{center}
-\scalebox{0.8}{
+\scalebox{0.9}{
   \includegraphics{diagrams/fold_nat.pdf}
@@
 \begin{center}
-\scalebox{0.8}{
+\scalebox{0.9}{
   \includegraphics{diagrams/add1.pdf}
@@
 \end{multicols}
-The triangle on the left denotes the {{unfold}} isomorphism, which
+The combinator on the left denotes the {{unfold}} isomorphism, which
 works as follows: If the number {{n}} is zero, the output is the top
 branch which has type~{{1}}. If the number is non-zero, the output is
-on the bottom branch and has value {{n-1}}. The triangle in the middle
-is the dual {{fold}} isomorphism.
+on the bottom branch and has value {{n-1}}. The combinator in the middle
+is its dual {{fold}} isomorphism.
 
@@
 subtraction can be expressed in {{langRevT}} as partial
-isomorphisms. The combinator on the right above is
-{{add1:nat<->nat}}. It acts like the successor function returning
-{{n+1}} for all inputs {{n}}. Its adjoint {{sub1:nat<->nat}}, which we
-get by flipping the diagram right to left, diverges on input {{0}} and
-returns {{n-1}} for all other inputs {{n}}.
+isomorphisms: The combinator on the right, {{add1:nat<->nat}}, acts
+like the successor function returning {{n+1}} for all inputs
+{{n}}. Its adjoint, {{sub1}}, obtained by flipping the
+diagram right to left, is a partial map that diverges on input {{0}}
+and returns {{n-1}} for all other inputs {{n}}.
 
@@
 Deriving {{langRevT}} combinators from abstract machines can sometimes
-be used as an efficient indirect way of programming in {{langRevT}}.
-It is well known that every number {{n}} can be represented as
-{{2a+0}} or {{2a+1}} depending on whether it is odd or even.  The
-later can be viewed as the algebraic data type
-{{par=odd~|~even~|~A~par}} such that 
-{{0 = even, 1 = odd, 2 = A even, 3 = A odd, 4 = A (A even)}} etc where
-the nesting of {{A}} constructor indicates the value of {{a}}.
+be used as an efficient indirect way to program in {{langRevT}}.  It
+is well known that every number {{n}} can be represented as {{2a+0}}
+or {{2a+1}}, depending on whether it is odd or even.  The later can be
+represented by the algebraic data type {{par=odd~|~even~|~A~par}}
+where the nesting of {{A}} constructor indicates the value of
+{{a}}. For example 
+{{0 = even, 1 = odd, 2 = A even, 3 = A odd, 4 = A (A even)}} etc.
 
 Say we wish to build a {{langRevT}} combinator to map {{nat}} to its
-parity encoded version {{par}}, expressible as the {{langRevT}} type
-{{@@x.(1+1)+x}}. Since {{par}} is not a fixed unfolding of {{nat}}
-({{@@x.1+x}}), it is not immediately apparent how such a combinator
-can be constructed.
-
-Instead of directly programming in {{langRevT}}, one can derive the
-required {{nat<->par}} combinator by first constructing an abstract
-machine that maps {{nat}} to {{par}}. We first express {{nat}} and
-{{par}} algebraically and then develop a logically reversible abstract
-machine expressing the translation:
+parity encoded version {{par}}. While the type {{par}} is expressible
+as {{@@x.(1+1)+x}} in {{langRevT}}, it is not a fixed unfolding of
+{{nat = @@x.1+x}} and hence it is not immediately apparent how such a
+combinator can be constructed.
+
+Instead of directly programming in {{langRevT}} however, one can
+derive the required {{nat<->par}} combinator by first constructing an
+abstract machine that maps {{nat}} to {{par}} and then translating it
+to {{langRevT}}. We first express {{nat}} and {{par}} algebraically
+and then develop a logically reversible abstract machine:
 
@@
 
-The construction of the {{langRevT}} combinator is
+The derivation of the {{langRevT}} combinator is
 straightforward and proceeds as before. The partial combinator

### 2012-09-29 twolevel.tex
partial write-up of ideas on two levels of our calculi

There are two ways to read types and Curry-Howard: 
\begin{enumerate}
\item One first concentrates on the \emph{types}, as propositions.  We can come up with a ``type system'' which 
reflects a logic.  Then, as a second step, choose the proofs which inhabit these types/propositions, and call these
proofs \emph{terms}.
\item Alternatively, one first concentrates on the \emph{terms}, focusing on their operational semantics as primary.
Then, as a second step, one designs a ``type system'' to classify the terms, so as to be able to prove progress and
preservations theorems, to ensure that well-typed terms ``don't go wrong''.
\end{enumerate}

Under the first interpretation, coming up with an operational interpretation of the terms can be challenging:
witness the ongoing effort to understand aspects of classical logic as a type system for a (programming) calculus.
Similarly, various modal logics as well as linear logic are still being given different interpretations as type
systems for calculi, to different degrees of effectiveness.

The second interpretation is not straightforward either.  The definition of ``doing wrong'' frequently gets 
expanded to covering effects, permissions, memory access, and so forth, which can seriously affect the decidability
properties of the type system (not to mention efficiency when decidability is not an issue).

Even in the first interpretation, the understanding is that terms represented \emph{values}, which are irreducible
terms (proof terms reduce principally via cut elimination, but there are other reduction rules for projections, etc).

We wish to shift the focus and consider \emph{equivalences} as our types.  This will mean, under the first 
interpretation, that our terms will be witnesses for these equivalences.  The underlying reason is that we wish
to examine \emph{reversible computation}.  From an information-theoretic point of view, this means that information
\emph{must be preserved}, although it can certainly be rearranged in many different ways.

In other words, we wish to regard types with shape $T_1\equiv T_2$, where $T_1$ and $T_2$ are usual value-types.
In fact, we want more: we wish to regard these as \emph{oriented}: the eventual induced operational interpretation
will regard equivalence witnesses as inducing a transformation from values of types $T_1$ to values of type $T_2$.
Of course, as these are equivalences, there will always be a term which inhabits $T_2\equiv T_1$ whenever there
is one inhabiting $T_1 \equiv T_2$.  This will be clearly visible when we give rules for terms, as these will always
by symmetric (and will provably induce invertible transformations).

- equational theories
- equations of equational theories
- axioms as constants
- various operators to create ``larger'' equiations from smaller
- (speak about lens as imperfect equivalences)
- algebraic structures as classifiers
- Curry-Howard up a level too -- sums and products of equivalences

### 2012-10-15 twolevel.tex

@@
 
-Even in the first interpretation, the understanding is that terms represented \emph{values}, which are irreducible
+Even in the first interpretation, the understanding is that terms represent \emph{values}, which are irreducible
 terms (proof terms reduce principally via cut elimination, but there are other reduction rules for projections, etc).
@@
 In other words, we wish to regard types with shape $T_1\equiv T_2$, where $T_1$ and $T_2$ are usual value-types.
-In fact, we want more: we wish to regard these as \emph{oriented}: the eventual induced operational interpretation
-will regard equivalence witnesses as inducing a transformation from values of types $T_1$ to values of type $T_2$.
+In fact, we want more: we wish to regard these as \emph{oriented}: the eventual operational interpretation
+will regard equivalence witnesses as \emph{inducing} a transformation from values of types $T_1$ to values of type $T_2$.
 Of course, as these are equivalences, there will always be a term which inhabits $T_2\equiv T_1$ whenever there
 is one inhabiting $T_1 \equiv T_2$.  This will be clearly visible when we give rules for terms, as these will always
-by symmetric (and will provably induce invertible transformations).
+be symmetric (and will provably induce invertible transformations).
 
-- equational theories
-- equations of equational theories
-- axioms as constants
-- various operators to create ``larger'' equiations from smaller
-- (speak about lens as imperfect equivalences)
-- algebraic structures as classifiers
-- Curry-Howard up a level too -- sums and products of equivalences
+So we start with a type theory $\mathcal{T}$, and move to another type theory whose only shape is $T_1\equiv T_2$.
+The constants of this new type theory (denoted $\mathcal{T}^{\equiv}$) will be
+proofs of type equivalences of the original theory.  Note that this is a very
+general set-up: given any equational theory, we can ``reify'' the axioms as
+constants.  [Insert the example
+of doing this for Monoid here].  If some of the constructors of $\mathcal{T}$ respect equivalences, then this induces
+combinators on $\mathcal{T}^{\equiv}$; the rules of equivalences (reflexivity, transitivity, symmetry) induce more.
+
+Note that it is very important that the theories be equational: the only thing we can talk about are total equivalences.  
+One of the most natural theory to consider, at least in the usual setting of type theories $\mathcal{T}$ with 
+sums and products, is that of \emph{semirings} (sometimes jokingly called \emph{rigs}), as the rules of semirings
+turns out to exactly reflect the judgments of such a type theory.
+
+What happens in equational theories is that we declare a number of values to be \emph{equal}.  Moving to the categorical
+setting, we are no longer interested in equalities, but just equivalences.  In equational theories, the equality
+witnesses are not particularly interesting, as they have no computational content  -- if $a = b$, there is no way we can
+distinguish $a$ and $b$, whereas if we only have $a \equiv b$, that may not be the case.  [The difference is that
+between syntax and semantics, where we have equivalences on the syntax which reflect equilities in the semantics.  In
+other words, plus(one,one) and two are easily distinguished, but 1+1 and 2 are not.]
+
+Furthermore, we can use these equivalences in an effective way: we can create a (total!) function
+$ev : T_1 \equiv T_2 \rightarrow T_1 \rightarrow T_2$.  In other words, given a proof of an equivalence between terms,
+and a value of the first type, we can get a value of the second type.
+
+It is important to understand that $ev$ \emph{mixes} two levels: that of (proofs of) equivalence and that of values.
+$ev$ is an \emph{induced action}.
+
+One further remark: our language of equivalences is actually a \emph{free} theory.  There are absolutely no
+equalities (or equivalences) given between its terms.  Of course, some of its terms are ``observationally equivalent'',
+in the sense that $ev$ will not be able to witness any difference.  Nevertheless, we will not regard such terms as
+equal.
+
+What needs to be shown then is the relationship between the equational theory of semirings and the first-order type
+theory with sums and products.
+
+Then we can investigate how adding further axioms to our equational theory (and corresponding witnesses to our type theory)
+changes things.

### 2012-10-24 r.tex

--- 2012-10-15 twolevel.tex
@@
+\documentclass[preprint]{sigplanconf}
+
+\usepackage{graphicx}
+\usepackage{comment}
+\usepackage{amsmath}
+\usepackage{xspace}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\usepackage{proof}
+\usepackage{multicol}
+\usepackage[nodayofweek]{datetime}
+\usepackage{etex}
+\usepackage[all, cmtip]{xy}
+\usepackage{xcolor}
+\usepackage{listings}
+\usepackage{multicol}
+\newcommand\hmmax{0} % default \newcommand\bmmax{0} % default 4
+\usepackage{bm}
+\usepackage{cmll}
+
+\newcommand{\vecspace}[1]{\mathbb{V}_{#1}}
+\renewcommand{\vec}[2]{\begin{bmatrix} #1 \\ #2 \end{bmatrix}}
+\newcommand{\Fpx}[1]{\mathbb{F}_{#1}}
+\newcommand{\modulusp}[1]{\left | #1 \right |}
+\newcommand{\fname}[1]{\ulcorner #1 \urcorner}
+\newcommand{\fconame}[1]{\llcorner #1 \lrcorner}
+
+\newcommand{\xcomment}[2]{\textbf{#1:~\textsl{#2}}}
+\newcommand{\amr}[1]{\xcomment{Amr}{#1}}
+\newcommand{\roshan}[1]{\xcomment{Roshan}{#1}}
+\newcommand{\jacques}[1]{\xcomment{Jacques}{#1}}
+
+\newcommand{\asterix}[0]{*}
+
+\newcommand{\ie}{\textit{i.e.}\xspace}
+\newcommand{\eg}{\textit{e.g.}\xspace}
+
+\newcommand{\lcal}{\ensuremath{\lambda}-calculus\xspace}
+\newcommand{\G}{\ensuremath{\mathcal{G}}\xspace}
+
+\newcommand{\code}[1]{\lstinline[basicstyle=\small]{#1}\xspace}
+\newcommand{\name}[1]{\code{#1}}
+
+\def\newblock{}
+
+\newenvironment{floatrule}
+    {\hrule width \hsize height .33pt \vspace{.5pc}}
+    {\par\addvspace{.5pc}}
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{proposition}[theorem]{Proposition}
+\newenvironment{proof}[1][Proof.]{\begin{trivlist}\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+
+\newcommand{\arrow}[1]{\mathtt{#1}}
+
+\newcommand{\dgm}[2][0.95]{
+\begin{center}
+\scalebox{#1}{
+\includegraphics{diagrams/#2.pdf}
+}
+\end{center}
+}
+
+%subcode-inline{bnf-inline} name langRev
+%! swap+ = \mathit{swap}^+
+%! swap* = \mathit{swap}^*
+%! dagger =  ^{\dagger}
+%! assocl+ = \mathit{assocl}^+
+%! assocr+ = \mathit{assocr}^+
+%! assocl* = \mathit{assocl}^*
+%! assocr* = \mathit{assocr}^*
+%! identr* = \mathit{uniti}
+%! identl* = \mathit{unite}
+%! dist = \mathit{distrib}
+%! factor = \mathit{factor}
+%! eta = \eta
+%! eps = \epsilon
+%! eta+ = \eta^+
+%! eps+ = \epsilon^+
+%! eta* = \eta^{\times}
+%! eps* = \epsilon^{\times}
+%! trace+ = trace^+
+%! trace* = trace^{\times}
+%! (o) = \circ
+%! (;) = \fatsemi
+%! (*) = \times
+%! (+) = +
+%! LeftP = L^+
+%! RightP = R^+
+%! LeftT = L^{\times}
+%! RightT = R^{\times}
+%! alpha = \alpha
+%! bool = \textit{bool}
+%! color = \textit{color}
+%! Gr = G
+
+%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
+%! Gx = \Gamma^{\times}
+%! G = \Gamma
+%! [] = \Box
+%! |-->* = \mapsto^{\asterix}
+%! |-->> = \mapsto_{\ggg}
+%! |--> = \mapsto
+%! <--| = \mapsfrom
+%! |- = \vdash
+%! <><> = \approx
+%! ==> = \Longrightarrow
+%! <== = \Longleftarrow
+%! <=> = \Longleftrightarrow
+%! <-> = \leftrightarrow
+%! ~> = \leadsto
+%! -o+ = \multimap^{+}
+%! -o* = \multimap^{\times}
+%! -o = \multimap
+%! ::= = &::=&
+%! /= = \neq
+%! @@ = \mu
+%! [^ = \lceil
+%! ^] = \rceil
+%! {| = \{
+%! |} = \}
+%! bot = \bot
+%! bot1 = \bot_1
+%! dots = \ldots
+%! ^^^ = ^{\dagger}
+%! elem = \in
+%! forall = \forall
+%! exists = \exists
+%! empty = \epsilon
+%! Pi = \Pi
+%! Pi0 = \Pi^{o}
+%! PiEE* = \Pi^{\eta\epsilon}_{*}
+%! PiEE+ = \Pi^{\eta\epsilon}_{+}
+%! PiEE = \Pi^{\eta\epsilon}
+%! CatSet = \textbf{Set}
+%! theseus = Theseus
+%! sqrt(x) = \sqrt{#x}
+%! surd(p,x) = \sqrt[#p]{#x}
+%! inv(x) = \frac{1}{#x}
+%! frac(x,y) = \frac{#x}{#y}
+%! * = \times
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+
+\conferenceinfo{ICFP'13}{}
+\CopyrightYear{}
+\copyrightdata{}
+\titlebanner{}
+\preprintfooter{}
+
+\title{Negative and Fractional Types}
+
+\authorinfo{Roshan P. James}
+           {Indiana University}
+           {rpjames@indiana.edu}
+\authorinfo{Zachary Sparks} 
+           {Indiana University}
+           {zasparks@indiana.edu}
+\authorinfo{Jacques Carette} 
+           {McMaster University}
+           {carette@mcmaster.ca}
+\authorinfo{Amr Sabry}
+           {Indiana University}
+           {sabry@indiana.edu}
+
+\maketitle
+
+\begin{abstract}
+
+\end{abstract}
+
+\category{D.3.1}{Formal Definitions and Theory}{}
+\category{F.3.2}{Semantics of Programming Languages}{}
+\category{F.3.3}{Studies of Program Constructs}{Type structure}
+
+\terms
+Languages, Theory
+
+\keywords
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}
+
+\nocite{infeffects}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Core Reversible Language: {{Pi}} }
+\label{sec:pi}
+
 There are two ways to read types and Curry-Howard: 
 \begin{enumerate}
-\item One first concentrates on the \emph{types}, as propositions.  We can come up with a ``type system'' which 
-reflects a logic.  Then, as a second step, choose the proofs which inhabit these types/propositions, and call these
-proofs \emph{terms}.
-\item Alternatively, one first concentrates on the \emph{terms}, focusing on their operational semantics as primary.
-Then, as a second step, one designs a ``type system'' to classify the terms, so as to be able to prove progress and
-preservations theorems, to ensure that well-typed terms ``don't go wrong''.
+\item One first concentrates on the \emph{types}, as propositions.  We can
+  come up with a ``type system'' which reflects a logic.  Then, as a second
+  step, choose the proofs which inhabit these types/propositions, and call
+  these proofs \emph{terms}.
+\item Alternatively, one first concentrates on the \emph{terms}, focusing on
+  their operational semantics as primary.  Then, as a second step, one
+  designs a ``type system'' to classify the terms, so as to be able to prove
+  progress and preservations theorems, to ensure that well-typed terms
+  ``don't go wrong''.
 \end{enumerate}
 
-Under the first interpretation, coming up with an operational interpretation of the terms can be challenging:
-witness the ongoing effort to understand aspects of classical logic as a type system for a (programming) calculus.
-Similarly, various modal logics as well as linear logic are still being given different interpretations as type
-systems for calculi, to different degrees of effectiveness.
+Under the first interpretation, coming up with an operational interpretation
+of the terms can be challenging: witness the ongoing effort to understand
+aspects of classical logic as a type system for a (programming) calculus.
+Similarly, various modal logics as well as linear logic are still being given
+different interpretations as type systems for calculi, to different degrees
+of effectiveness.
 
-The second interpretation is not straightforward either.  The definition of ``doing wrong'' frequently gets 
-expanded to covering effects, permissions, memory access, and so forth, which can seriously affect the decidability
-properties of the type system (not to mention efficiency when decidability is not an issue).
+The second interpretation is not straightforward either.  The definition of
+``doing wrong'' frequently gets expanded to covering effects, permissions,
+memory access, and so forth, which can seriously affect the decidability
+properties of the type system (not to mention efficiency when decidability is
+not an issue).
 
-Even in the first interpretation, the understanding is that terms represent \emph{values}, which are irreducible
-terms (proof terms reduce principally via cut elimination, but there are other reduction rules for projections, etc).
+Even in the first interpretation, the understanding is that terms represent
+\emph{values}, which are irreducible terms (proof terms reduce principally
+via cut elimination, but there are other reduction rules for projections,
+etc).
 
-We wish to shift the focus and consider \emph{equivalences} as our types.  This will mean, under the first 
-interpretation, that our terms will be witnesses for these equivalences.  The underlying reason is that we wish
-to examine \emph{reversible computation}.  From an information-theoretic point of view, this means that information
-\emph{must be preserved}, although it can certainly be rearranged in many different ways.
+We wish to shift the focus and consider \emph{equivalences} as our types.
+This will mean, under the first interpretation, that our terms will be
+witnesses for these equivalences.  The underlying reason is that we wish to
+examine \emph{reversible computation}.  From an information-theoretic point
+of view, this means that information \emph{must be preserved}, although it
+can certainly be rearranged in many different ways.
 
-In other words, we wish to regard types with shape $T_1\equiv T_2$, where $T_1$ and $T_2$ are usual value-types.
-In fact, we want more: we wish to regard these as \emph{oriented}: the eventual operational interpretation
-will regard equivalence witnesses as \emph{inducing} a transformation from values of types $T_1$ to values of type $T_2$.
-Of course, as these are equivalences, there will always be a term which inhabits $T_2\equiv T_1$ whenever there
-is one inhabiting $T_1 \equiv T_2$.  This will be clearly visible when we give rules for terms, as these will always
-be symmetric (and will provably induce invertible transformations).
+In other words, we wish to regard types with shape $T_1\equiv T_2$, where
+$T_1$ and $T_2$ are usual value-types.  In fact, we want more: we wish to
+regard these as \emph{oriented}: the eventual operational interpretation will
+regard equivalence witnesses as \emph{inducing} a transformation from values
+of types $T_1$ to values of type $T_2$.  Of course, as these are
+equivalences, there will always be a term which inhabits $T_2\equiv T_1$
+whenever there is one inhabiting $T_1 \equiv T_2$.  This will be clearly
+visible when we give rules for terms, as these will always be symmetric (and
+will provably induce invertible transformations).
 
-So we start with a type theory $\mathcal{T}$, and move to another type theory whose only shape is $T_1\equiv T_2$.
-The constants of this new type theory (denoted $\mathcal{T}^{\equiv}$) will be
-proofs of type equivalences of the original theory.  Note that this is a very
-general set-up: given any equational theory, we can ``reify'' the axioms as
-constants.  [Insert the example
-of doing this for Monoid here].  If some of the constructors of $\mathcal{T}$ respect equivalences, then this induces
-combinators on $\mathcal{T}^{\equiv}$; the rules of equivalences (reflexivity, transitivity, symmetry) induce more.
+So we start with a type theory $\mathcal{T}$, and move to another type theory
+whose only shape is $T_1\equiv T_2$.  The constants of this new type theory
+(denoted $\mathcal{T}^{\equiv}$) will be proofs of type equivalences of the
+original theory.  Note that this is a very general set-up: given any
+equational theory, we can ``reify'' the axioms as constants.  [Insert the
+example of doing this for Monoid here].  If some of the constructors of
+$\mathcal{T}$ respect equivalences, then this induces combinators on
+$\mathcal{T}^{\equiv}$; the rules of equivalences (reflexivity, transitivity,
+symmetry) induce more.
 
-Note that it is very important that the theories be equational: the only thing we can talk about are total equivalences.  
-One of the most natural theory to consider, at least in the usual setting of type theories $\mathcal{T}$ with 
-sums and products, is that of \emph{semirings} (sometimes jokingly called \emph{rigs}), as the rules of semirings
-turns out to exactly reflect the judgments of such a type theory.
+Note that it is very important that the theories be equational: the only
+thing we can talk about are total equivalences.  One of the most natural
+theory to consider, at least in the usual setting of type theories
+$\mathcal{T}$ with sums and products, is that of \emph{semirings} (sometimes
+jokingly called \emph{rigs}), as the rules of semirings turns out to exactly
+reflect the judgments of such a type theory.
 
-What happens in equational theories is that we declare a number of values to be \emph{equal}.  Moving to the categorical
-setting, we are no longer interested in equalities, but just equivalences.  In equational theories, the equality
-witnesses are not particularly interesting, as they have no computational content  -- if $a = b$, there is no way we can
-distinguish $a$ and $b$, whereas if we only have $a \equiv b$, that may not be the case.  [The difference is that
-between syntax and semantics, where we have equivalences on the syntax which reflect equilities in the semantics.  In
-other words, plus(one,one) and two are easily distinguished, but 1+1 and 2 are not.]
+What happens in equational theories is that we declare a number of values to
+be \emph{equal}.  Moving to the categorical setting, we are no longer
+interested in equalities, but just equivalences.  In equational theories, the
+equality witnesses are not particularly interesting, as they have no
+computational content -- if $a = b$, there is no way we can distinguish $a$
+and $b$, whereas if we only have $a \equiv b$, that may not be the case.
+[The difference is that between syntax and semantics, where we have
+equivalences on the syntax which reflect equilities in the semantics.  In
+other words, plus(one,one) and two are easily distinguished, but 1+1 and 2
+are not.]
 
-Furthermore, we can use these equivalences in an effective way: we can create a (total!) function
-$ev : T_1 \equiv T_2 \rightarrow T_1 \rightarrow T_2$.  In other words, given a proof of an equivalence between terms,
-and a value of the first type, we can get a value of the second type.
+Furthermore, we can use these equivalences in an effective way: we can create
+a (total!) function $ev : T_1 \equiv T_2 \rightarrow T_1 \rightarrow T_2$.
+In other words, given a proof of an equivalence between terms, and a value of
+the first type, we can get a value of the second type.
 
-It is important to understand that $ev$ \emph{mixes} two levels: that of (proofs of) equivalence and that of values.
-$ev$ is an \emph{induced action}.
+It is important to understand that $ev$ \emph{mixes} two levels: that of
+(proofs of) equivalence and that of values.  $ev$ is an \emph{induced
+  action}.
 
-One further remark: our language of equivalences is actually a \emph{free} theory.  There are absolutely no
-equalities (or equivalences) given between its terms.  Of course, some of its terms are ``observationally equivalent'',
-in the sense that $ev$ will not be able to witness any difference.  Nevertheless, we will not regard such terms as
-equal.
+One further remark: our language of equivalences is actually a \emph{free}
+theory.  There are absolutely no equalities (or equivalences) given between
+its terms.  Of course, some of its terms are ``observationally equivalent'',
+in the sense that $ev$ will not be able to witness any difference.
+Nevertheless, we will not regard such terms as equal.
 
-What needs to be shown then is the relationship between the equational theory of semirings and the first-order type
-theory with sums and products.
+What needs to be shown then is the relationship between the equational theory
+of semirings and the first-order type theory with sums and products.
 
-Then we can investigate how adding further axioms to our equational theory (and corresponding witnesses to our type theory)
-changes things.
+Then we can investigate how adding further axioms to our equational theory
+(and corresponding witnesses to our type theory) changes things.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\acks We thank Michael Adams, Will Byrd, Lindsey Kuper, and Yin Wang for
+helpful comments and questions. This project is supported by NSF grant
+1217454 on ``Information Effects.'' Initial stages of the research were
+partially funded by Indiana University's Office of the Vice President for
+Research and the Office of the Vice Provost for Research through its Faculty
+Research Support Program.  We also acknowledge support from Indiana
+University's Institute for Advanced Study.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{small}
+\bibliographystyle{abbrvnat}
+\bibliography{cites}
+\end{small}
+
+\end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-01-22 frac.tex
frac

\documentclass{llncs}
\usepackage{url}
\usepackage{proof}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{graphicx}

\newcommand{\inkscape}[2][1.5]{
\begin{center}
\scalebox{#1}{
\includegraphics{diagrams/#2}
}
\end{center}
}

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! in = \!\!\in\!\!
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! trace+ = \mathit{trace}^{+}
%! trace* = \mathit{trace}^{\times}
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\title{Fractional Types}
\author{}
\institute{}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
?
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax and Semantics}

We can turn the above system of isomorphisms into a (reversible)
programming language by simply providing a syntax for each of the
isomorphisms. For each isomorphism, we introduce two
\emph{combinators} corresponding to the left-to-right and
right-to-left directions.

\begin{definition}[Primitive {{langRev}} Combinators]
\label{def:primitives-langRev}

%subcode{bnf} include main
%! columnStyle = rrcll
%  zeroe :&  0 + b &<->& b &: zeroi
%  swap+ :&  b1 + b2 &<->& b2 + b1 &: swap+
%  assocl+ :&  b1 + (b2 + b3) &<->& (b1 + b2) + b3 &: assocr+
%  unite :&  1 * b &<->& b &: uniti
%  swap* :&  b1 * b2 &<->& b2 * b1 &: swap*
%  assocl* :&  b1 * (b2 * b3) &<->& (b1 * b2) * b3 &: assocr*
%  dist0 :&  0 *b &<->& 0  &: factor0
%  dist :& (b1 + b2) * b3 &<->& (b1 * b3) + (b2 * b3) &: factor

\end{definition}

Each line of this table is to be read as the definition of one or two
combinators. For example, corresponding to the \textit{identity of
  multiplication} ({{1*b<->b}}) we have the two combinators
{{unite:1*b<->b}} (reading the isomorphism from left to right) and
{{uniti:b<->1*b}} (reading the isomorphism from right to left). These
combinators are inverses of each other.  Each of the two cases of
commutativity defines one combinator that is its own inverse.

Now that we have primitive combinators we need some means of composing
them. We construct the composition combinators out of the congruence
closure of basic isomorphisms. Thus we have program constructs that
witness reflexivity {{id}}, symmetry {{sym}}, transitivity {{(;)}},
and two parallel composition combinators, one for sums {{(+)}} and one
for pairs {{(*)}}.

\begin{definition}[Composition in {{langRev}}]
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%----
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
\end{definition}

\noindent To summarize the syntax of our language {{langRev}} is as follows.

\begin{definition}[Syntax of {{langRev}}]
\label{def:langRev}
We collect our types, values, and combinators, to get the full language
definition:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% combinator types, t ::= b <-> b
% primitive combinators, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist0 | factor0 | dist | factor 
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

By design, for each combinator {{c : b1 <-> b2}}, there is an
\emph{adjoint} {{c^dagger : b2 <-> b1}} acting in the reverse direction.

\begin{definition}[Adjoints {{c^dagger}}]
 The adjoint of a combinator {{c}} is defined by {{c^dagger}} as
 follows. For primitive combinators {{c}}, {{c^dagger}} is given by
 the inverse in Def.~\ref{def:primitives-langRev}. For the remaining
 combinators, we have:
%subcode{opsem} include main
% id^dagger &=& id
% (sym c)^dagger &=& c
% (c1(;)c2)^dagger &=& c2^dagger (;) c1^dagger
% (c1 (+) c2)^dagger &=& c1^dagger (+) c2^dagger
% (c1 (*) c2)^dagger &=& c1^dagger (*) c2^dagger
which essentially amounts to reversing the order of sequential composition.
\end{definition}

The semantics of {{langRev}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of
{{sym}} switches control from one evaluator to the other.

\begin{definition}[Operational Semantics of {{langRev}}]
\label{def:operational-langRev}
Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
forward direction by supplying it with a value {{ v1 : b1 }} and in
the backwards direction by supplying it with a value {{v2 : b2}}. The
forward evaluator {{evalF}} and the backwards evaluator {{evalB}} are
given below. Since there are no values that have the type {{0}}, the
clauses for the combinators {{zeroe}}, {{zeroi}}, {{dist0}} and
{{factor0}} omit the impossible cases.

\paragraph*{Forward Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalF zeroe (right v) &=& v
% evalF zeroi v &=& right v
% evalF swap+ (left v) &=& right v
% evalF swap+ (right v) &=& left v 
% evalF assocl+ (left v) &=& left (left v)
% evalF assocl+ (right (left v)) &=& left (right v)
% evalF assocl+ (right (right v)) &=& right v
% evalF assocr+ (left (left v)) &=& left v
% evalF assocr+ (left (right v)) &=& right (left v)
% evalF assocr+ (right v) &=& right (right v)
% evalF unite ((), v) &=& v 
% evalF uniti v &=& ((), v) 
% evalF swap* (v1, v2) &=& (v2, v1) 
% evalF assocl* (v1, (v2, v3)) &=& ((v1, v2), v3) 
% evalF assocr* ((v1, v2), v3) &=& (v1, (v2, v3)) 
% evalF dist (left v1, v3) &=& left (v1, v3)
% evalF dist (right v2, v3) &=& right (v2, v3)
% evalF factor (left (v1, v3)) &=& (left v1, v3) 
% evalF factor (right (v2, v3)) &=& (right v2, v3)   
% evalF id v &=& v
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& evalF c2 (evalF c1 v)
% evalF (c1 (+) c2) (left v1) &=& left (evalF c1 v1)
% evalF (c1 (+) c2) (right v2) &=& right (evalF c2 v2)
% evalF (c1 (*) c2) (v1,v2) &=& (evalF c1 v1, evalF c2 v2)

\paragraph*{Backwards Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalB zeroe v &=& right v 
% evalB zeroi (right v) &=& v
% evalB swap+ (left v) &=& right v
% evalB swap+ (right v) &=& left v
% evalB assocl+ (left (left v)) &=& left v
% evalB assocl+ (left (right v)) &=& right (left v)
% evalB assocl+ (right v) &=& right (right v)
% evalB assocr+ (left v) &=& left (left v)
% evalB assocr+ (right (left v)) &=& left (right v) 
% evalB assocr+ (right (right v)) &=& right v
% evalB unite v &=& ((), v) 
% evalB uniti ((), v) &=& v
% evalB swap* (v1, v2) &=& (v2, v1) 
% evalB assocl* ((v1, v2), v3) &=& (v1, (v2, v3)) 
% evalB assocr* (v1, (v2, v3)) &=& ((v1, v2), v3) 
% evalB dist (left (v1, v3)) &=& (left v1, v3)
% evalB dist (right (v2, v3)) &=& (right v2, v3) 
% evalB factor (left v1, v3) &=& left (v1, v3) 
% evalB factor (right v2, v3) &=& right (v2, v3) 
% evalB id v &=& v
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& evalB c1 (evalB c2 v)
% evalB (c1 (+) c2) (left v1) &=& left (evalB c1 v1)
% evalB (c1 (+) c2) (right v2) &=& right (evalB c2 v2)
% evalB (c1 (*) c2) (v1,v2) &=& (evalB c1 v1, evalB c2 v2)
\end{definition}

\begin{proposition}[Termination] Evaluation of well-typed combinators
  always terminates. In other words, for {{c : b1 <-> b2}} and 
  {{v1 : b1}} there exists {{v2 : b2}} such that {{evalF c v1 = v2}}. 
  Conversely, if {{v2 : b2}} there exists {{v1 : b1}} such that 
  {{evalB c v2 = v1}}.
\end{proposition}

\begin{proposition}[Logical Reversibility]
\label{prop:logrev}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have: {{evalF c v1 = v2}} iff {{evalB c v2 = v1}}.
\end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Trace Operators: Intuition}

The language {{langRev}} is not yet universal: one of the missing
ingredients is a form of recursion or iteration. Mathematically
speaking, recursion and iteration can be expressed using categorical
trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}. We
will work out the precise categorical connections later: for now we
pursue our operational approach.

\paragraph*{Additive Case.} 
The intuitive idea of trace operators is as follows. We consider first
the additive case. Given a computation {{c:b1+b2<->b1+b3}}, we can
\emph{cancel} the common type {{b1}} from both the input and the
output by feeding back to itself, thereby getting a new combinator
{{trace+~c:b2<->b3}}. Diagrammatically, we have:

\inkscape{trace_plus.pdf}

Operationally, the execution of such a combinator proceeds as
follows. We are given a value {{v2 : b2}} which is injected into the
type {{b1 + b2}} by tagging it as {{right v2}}. This tagged value is
passed to the combinator {{f : b1 + b2 <-> b1 + b3}}. The resulting
value will either be {{left v1}} for some {{v1 : b1}} or {{right v3}}
for some {{v3 : b3}}. In the former case, the value {{left v1}} is fed
back to {{f}}. In the latter case, the entire {{trace+}}-combinator
terminates with value {{v3}}. An example should make this more
concrete.

\begin{example}
Consider the combinator {{c:bool+1<->bool+1}} below:

\inkscape{trace-example0.pdf}

\noindent
When we trace {{c}} we have the circuit {{trace+ c: 1 <-> 1}} shown below.

\inkscape{trace-example1.pdf}

\noindent
The following diagrams show snapshots the evaluation of the
circuit applied to {{()}}:
\begin{multicols}{2}
  \inkscape{trace-example2.pdf}
  \inkscape{trace-example3.pdf}
\end{multicols}

\begin{multicols}{2}
  \inkscape{trace-example4.pdf}
  \inkscape{trace-example5.pdf}
\end{multicols}

\begin{multicols}{2}
  \inkscape{trace-example6.pdf}
  \inkscape{trace-example7.pdf}
\end{multicols}

\noindent
The value starts at the input to the circuit, flows out on the
{{left}} branch of {{c}} and is traced back into {{c}}.  The feedback
loop is taken exactly two times in the execution of this circuit.
\end{example}

\paragraph*{Multiplicative Case.} 
The situation with the multiplicative trace is a bit more subtle. In
this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
want to cancel the common type {{b1}} as before, to produce a new
combinator {{trace*~c:b2<->b3}}. For the evaluation of such a
combinator, we are only given a value {{v2 : b2}} but to evaluate
{{c}}, we must provide a pair consisting of some value {{v1 : b1}}
together with the given {{v2 : b2}}. The needed value {{v1 : b1}}
cannot be arbitrary, however. We do have a constraint that the value
{{v1 : b1}} must be such that it is also produced by as the first
component of the result, i.e., that {{evalF c (v1,v2) = (v1,v3)}}. In
general, there may be several such values fixed-point values or
none. A few examples should help understand the subtleties.

\begin{example}
Consider the combinator {{c}}:

{{c : bool * 1 <-> bool * 1}}

{{c = id}}

\noindent Using the multiplicative trace operator, we can construct the
combinator {{trace* c : 1 <-> 1}}. Applying this combinator to 
{{() : 1}} requires us to find a value {{b : bool}} such that 
{{evalF c (b,()) = (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
has two values, there are two values {{false}} and {{true}} that satisfy the 
constraint. In other words, expect that {{evalF (trace* c) () = ()}}. 
\end{example}

\begin{example}
\label{ch3:ex:annihilate}
Consider a small variation on the combinator {{c}} above:

{{c : bool * 1 <-> bool * 1}}

{{c = swap+ (*) id }}

\noindent which negates the boolean component of its incoming
pair. Using the multiplicative trace operator, we can construct the
combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
combinator to {{() : 1}} requires us to find a value {{b : bool}} such
that {{evalF c (b,()) = (b,())}} which is impossible since {{c}} negates
{{b}}. Operationally, we would expect the evaluation of such a
combinator to produce no value. 
\end{example}

\begin{example}
A more fundamental problem occurs if we attempt to apply the trace
operator to cancel the empty type. Consider a combinator:

{{c : 0 * 1 <-> 0 * bool}}

{{c = distrib0 (;) factor0}}

\noindent Using the multiplicative trace, we can construct the combinator
{{trace* c : 1 <-> bool}} which relates two types of different sizes! Fortunately,
because {{0}} is the empty type, 
when attempting to evaluate such a combinator, it is impossible to find a value
{{v : 0}} to satisfy the constraint imposed by {{trace*}} and hence 
we would also expect the evaluation of such a combinator to produce no value.
\end{example}

\begin{example}
One might conclude from the examples above that the multiplicative
trace is non-sensical. There are however some excellent and legitimate
uses illustrated in the remainder of this chapter. A simple
illustrative example takes a combinator {{c : b <-> b}} and detects
whether it has a fixed point, i.e., whether there exists a value 
{{v :  b}} such that {{evalF c v = v}}:

{{hasFixedPoint c = trace* (swap* (;) unite (;) c (;) uniti (;) swap*)}}

\noindent The idea is to introduce a unit value before and after the evaluation of {{c}}
effectively changing the type of {{c}} to {{b * 1 <-> b * 1}}
and then use {{trace*}} to
cancel the value of type~{{b}}. The result of evaluating
{{(hasFixedPoint c) ()}} is {{()}} if {{c}} has a fixed point and returns 
no value otherwise. 
\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Multiplicative Trace Operators}
\label{sec:trace}

For the multiplicative trace operator, we similarly add a new family
of combinators {{trace*_{b} c}} indexed by types {{b}} to our
syntax. The type rule, can be viewed as a cancellation law that
eliminates the type~{{b}} from both sides of a product.

\begin{definition}[Type Judgment for {{trace*_{b}}}]
%subcode{proof} include main
%@ c : b * b2 <-> b * b3
%@@ trace*_{b} c : b2 <-> b3  
\end{definition}

The adjoint of {{trace*_{b} c}} is also the trace of the adjoint of
{{c}}, i.e., {{(trace*_{b} c)^dagger = trace*_{b} c^dagger}}.

As hinted in the first section, the semantics of {{trace*}} introduces
a form of non-determinism in the sense that there could zero, one, or
several values that satisfy the constraint imposed by {{trace*}}. To
model such non-determinism, we specify the semantics using
\emph{relations}. In other words, the results of {{evalF c v}} 
or {{evalB c v}} are no longer guaranteed to be a single value: they 
could be \emph{sets} of values.

The introduction of relations makes sequential composition of
combinators more involved than simply passing the value produced by
the first to the second. The following example illustrates our
notation for the composition of relations which is used in the formal
semantics.

\begin{example}[Composition of Relations]
Consider the two relations:
\[\begin{array}{rcl}                                                                       
R_1 &=& \{ (1,2),(1,3),(2,2) \} \\                                                       
R_2 &=& \{ (2,1),(2,3),(3,1) \}                                                          
\end{array}\]
These relations can be expressed in a functional notation as follows:
\[\begin{array}{rcl}                                                                       
R_1 (1) &=& \{ 2,3 \} \\                                                                 
R_1 (2) &=& \{ 2 \} \\                                                                   
R_1 (3) &=& \emptyset \\
\\ 
R_2 (1) &=& \emptyset \\                                                                 
R_2 (2) &=& \{ 1,3 \} \\                                                                 
R_2 (3) &=& \{ 1 \}                                                                      
\end{array}\]
Using the functional notation, the composition $(R_2 \circ R_1)$ is defined as
follows:
\[\begin{array}{rcl}                                                                   
(R_2 \circ R_1)~(a) &=& \bigcup_{b \in R_1(a)} R_2(b)                                  
\end{array}\]
We calculate the composition:
\[\begin{array}[t]{rcl}                                                                  
(R_2 \circ R_1)~(1) &=& \bigcup_{b \in R_1(1)} R_2(b) \\                               
         &=& R_2(2) \cup R_2(3) \\                                                     
         &=& \{1,3\} \cup \{ 1 \} \\                                                   
         &=& \{ 1,3 \} \\                                                              
(R_2 \circ R_1)~(2) &=& \{ 1,3 \} \\                                                   
(R_2 \circ R_1)~(3) &=& \emptyset                                                      
\end{array}\]
In other words, 
\[
(R_2 \circ R_1) = \{(1,1), (1,3), (2,1), (2,3)\}
\]
\end{example}

\begin{definition}[Operational Semantics for {{trace*_{b}}}]
The formal semantics is expressed by lifting the forward and backward
evaluators to relations and then adding the new cases for {{trace*}}.

\paragraph*{Forward Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalF zeroe (right v) &=& {[ v ]}
% evalF zeroi v &=& {[ right v ]}
% evalF swap+ (left v) &=& {[ right v ]}
% evalF swap+ (right v) &=& {[ left v ]}
% evalF assocl+ (left v) &=& {[ left (left v) ]}
% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
% evalF assocl+ (right (right v)) &=& {[ right v ]}
% evalF assocr+ (left (left v)) &=& {[ left v ]}
% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
% evalF assocr+ (right v) &=& {[ right (right v) ]}
% evalF unite ((), v) &=& {[ v ]}
% evalF uniti v &=& {[ ((), v) ]}
% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalF id v &=& {[ v ]}
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
% evalF (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalF c (v,v2)) ]}

\paragraph*{Backwards Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalB zeroe v &=& {[ right v ]}
% evalB zeroi (right v) &=& {[ v ]}
% evalB swap+ (left v) &=& {[ right v ]}
% evalB swap+ (right v) &=& {[ left v ]}
% evalB assocl+ (left (left v)) &=& {[ left v ]}
% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
% evalB assocl+ (right v) &=& {[ right (right v) ]}
% evalB assocr+ (left v) &=& {[ left (left v) ]}
% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
% evalB assocr+ (right (right v)) &=& {[ right v ]}
% evalB unite v &=& {[ ((), v) ]}
% evalB uniti ((), v) &=& {[ v ]}
% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
% evalB id v &=& {[ v ]}
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
% evalB (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalB c (v,v2)) ]}
\end{definition}

In the cases for {{trace*_b}}, we check every value {{v : b}} in the
denotation of the type {{b}}. Specifically, given {{v2}}, we produce the pair
{{(v,v2)}} and check if the result of evaluating {{c}} on this pair produces
the same {{v}} as the first component of its result. In that case, we include
the second component in the final answer, and otherwise, we reject that
second component.

Now that we have the formal semantics, we consider a generalized
version of Ex.~\ref{ch3:ex:annihilate}.

\begin{example}[Annihilation]
\label{ch3:ex;annihilation}
Consider the circuit below:
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}
As motivated earlier, it is clearly impossible for the constraint
implied by {{trace*}} to be satisfied as a boolean value {{v}} can
never be the same as its negation. Thus there is no incoming value of
type {{b}} that can ever go through the circuit. Indeed, if we call
the circuit above {{c}}, then we can calculate that 
{{ evalF c v = emptyset}}. We call such a circuit an 
annihilation circuit.
\end{example}

Every relation is clearly reversible by simply swapping the input and
output components of every pair, i.e., if {{r = {[ (a,b), ... ]} }}
then the inverse of {{r}} is simply {{ {[ (b,a), ... ]} }}.

\begin{proposition}[Logical Reversibility]
\label{prop:logrev-tracet}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have if {{v2 in (evalF c v1)}} then 
{{v1 in (evalB c v2)}}.
\end{proposition}

An interesting point to ponder is what happens to \emph{information}
in circuits like {{annihilate}}? We return to this point later.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{(Finite) Relational Programming}
\label{ch3:sec:lp}

Relational programming leverages set-theoretic relations and their
composition to express computational tasks in a declarative way. For
instance, consider this example:

{{parent = {[ (A,B), (B,C), (A,D) ]} }}

{{grandparent = parent (o) parent}}

\noindent The example defines a relation {{parent}} specified using a
set of tuples and another relation {{grandparent}} specified using the
composition of two parent relations. If we wrote this example in a
relational language (e.g., Prolog) and we executed the query
{{grandparent(A)}}, we would get the answer {{ {[C]} }}.

It turns out that with the addition of the multiplicative trace, and
the move to relations motivated in the previous section, we can
express relational programming. We illustrate the idea using a
complete example.

Consider the relation {{R}} on booleans given by:

{{ {[(false,false), (false,true), (true,false) ]}. }}

We show how to define a combinator {{c_R}} whose semantics is such that:

%subcode{opsem} include main
%! columnStyle = rcl
% evalF c_R false &=& {[ false, true ]} 
% evalF c_R true &=& {[ false ]}

The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
for some type {{a}} such that {{trace* cInner}} is the desired combinator
{{c_R}}. A little experimentation shows that we get the desired behavior
if {{cInner}} is chosen to behave as follows:

%subcode{opsem} include main
%! columnStyle = rcl
% evalF cInner ((false,false),false) &=& ((false,false),false) 
% evalF cInner ((false,true),false) &=& ((false,true),true) 
% evalF cInner ((false,true),true) &=& ((false,true),false) 
% evalF cInner ((false,false),true) &=& ((true,false),true) 
% evalF cInner ((true,false),false) &=& ((true,true),false) 
% evalF cInner ((true,false),true) &=& ((true,true),true) 
% evalF cInner ((true,true),false) &=& ((true,false),false) 
% evalF cInner ((true,true),true) &=& ((false,false),true)

In the first three lines, the first argument is a fixed point and hence the
multiplicative trace would map the second input to the second output
producing the desired relation. In the remaining five cases, the first
argument is not a fixed point and hence all these cases would be rejected as
solutions to the constraint imposed by the multiplicative trace. It simply
remains to find the actual realization of {{cInner}} that would produce the
behavior above. One can infer the following implementation:

%subcode{opsem} include main
%! columnStyle = rcl
% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assoct* (;) swap* (;) 
%        &&  toffoli (;) ((swap* (;) cnot (;) swap*) (*) id) (;)
%        &&  toffoli (;) (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
%        &&  toffoli (;) (cnot (*) id))
%     
% c_R &:& bool <-> bool
% c_R &=& trace* cInner

The above example should convince the reader that the language {{langRev}}
with multiplicative trace is expressive enough to model finite relational
programming. We will not prove this claim. We develop an even more
substantial example: a SAT solver.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solving Constraints}
\label{ch3:sec:constraints}

A large class of constraint satisfaction problems can be expressed
using multiplicative traces. We illustrate the main ideas with the
implementation of a SAT solver.

In the usual setting, an instance of SAT is a function~{{f}} which,
when given some boolean inputs, returns {{true}} or {{false}}. The
function returns {{true}} when the inputs satisfy the constraints
imposed by the structure of {{f}} and a solution to the SAT problem is
the set of all inputs on which {{f}} produces {{true}}. The basic idea
of our construction is to generalize the annihilation circuit from
Ex.~\ref{ch3:ex;annihilation} to only annihilate values that fail to
satisfy the constraints represented by the SAT instance {{f}}. To
achieve this goal, we must however deal with several important
details.

First, because we are in a reversible world, our instance of SAT must be
expressed as an isomorphism: this is easily achieved by the construction
which embeds any boolean function {{f}} into a reversible one {{iso_f}} with
a larger domain and range.  Given such a reversible function {{iso_f}} which
represents a SAT instance, we first construct the circuit below:

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/sat2.pdf}
}
\end{center}

As shown in the circuit, the reversible SAT instance {{iso_f}} takes
two sets of values and produces two outputs. The incoming values
labeled \textsf{inputs} are the inputs we need to test for
satisfiability. The other incoming values labeled \textsf{heap} are
the additional inputs needed to embed the original SAT instance {{f}}
into a reversible function. If these \textsf{heap} values are all
initialized to {{false}}, the output wire \textsf{satisfied?}
corresponds to the output that {{f}} would have produced on
\textsf{inputs}. The other outputs labeled \textsf{garbage} are not
needed for their own sake but they are important because they are used
as inputs to the adjoint of {{iso_f}} to reproduce the inputs exactly,
in anticipation of closing the loop with {{trace*}}.

To summarize, the top half of the circuit is the identity function
except that we have also managed to produce a boolean wire labeled
\textsf{satisfied?} that tells us if the inputs satisfy the desired
constraints. We can take this boolean value and use it to decide
whether to negate the bottom wire (labeled \textsf{control
  wire}). Specifically, if the inputs do \emph{not} satisfy {{f}}, the
control wire is negated. The last wire labeled \textsf{heap control
  wire} is negated if the heap values do not have the right initial
values, i.e., are not all {{false}}.

Let us call the above construction {{sat_f}}. If we now close the loop
using {{trace*}}, two things should happen:
\begin{itemize}
\item configurations in which the \textsf{heap} values are not all
  {{false}} will be annihilated;
\item configurations in which the \textsf{inputs} do not satisfy {{f}}
  will cause the \textsf{satisfied?} wire to be negated and hence will
  also be annihilated.
\end{itemize}
In other words, the only configurations that will survive are the ones in
which the \textsf{inputs} satisfy {{f}}. We simply need to arrange to
\emph{clone} these values and produce them as the output of the whole
circuit. The final construction is therefore:

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/sat3.pdf}
}
\end{center}

To make the previous discussion concrete, we present a small, but
complete, example. In our example, the SAT instance {{f}} is tiny: it
takes two inputs. This function is embedded into a reversible function
{{iso_f}} of type 
{{((bool * bool) * bool) <-> ((bool * bool) * bool)}} where the last
input represents the heap and the first two outputs represent the garbage. 
The realization of {{sat_f}} given below is parametrized by such 
a function {{iso_f}}. The inputs to {{sat_f}} are 
\textsf{heap control}, \textsf{control}, \textsf{heap}, \textsf{input-1}, and 
\textsf{input-2}. Its operation is simple: if the \textsf{heap} is {{true}}, 
\textsf{heap control} is negated, and if the last output of {{iso_f}} 
is {{false}}, \textsf{control} is negated:

%subcode{opsem} include main
%! columnStyle = rcl
% sat_f &:& ((((bool * bool) * bool) * bool) * bool) <-> ((((bool * bool) * bool) * bool) * bool)
% sat_f &=& ((swap* (*) id) (*) id) (;) 
% && ((assocl* (*) id) (*) id) (;) 
% &&   (((cnot (*) id) (*) id) (*) id) (;) 
% &&   assocr* (;) 
% &&   (assocr* (*) id) (;) 
% &&   (swap* (*) id) (;) 
% &&   assocr* (;) 
% &&   (id (*) assocl*) (;) 
% &&   (id (*) isof) (;) 
% &&   swap* (;) 
% &&   assocr* (;) 
% &&   (id (*) (id (*) swap*)) (;) 
% &&   (id (*) assocl*) (;) 
% &&   (id (*) ((inot (*) id) (*) id)) (;)
% &&   (id (*) (cnot (*) id)) (;) 
% &&   (id (*) ((inot (*) id) (*) id)) (;)
% &&   (id (*) assocr*) (;) 
% &&   (id (*) (id (*) swap*)) (;) 
% &&   assocl* (;)
% &&   swap* (;)
% &&   (id (*) Sym isof) (;) 
% &&   (id (*) assocr*) (;) 
% &&   assocl* (;) 
% &&   assocl*

Given the construction of {{sat_f}} we can build the full solver as
follows. The overall input is the cloning heap. The combinator given
to {{trace*}} takes the cloning heap and the inputs flowing around the
loop and produces two copies of these inputs. One copy is produced as
the overall output and another is fed back around the loop.

%subcode{opsem} include main
%! columnStyle = rcl
% solve_f &:& bool * bool <-> bool * Bool
% solve_f &=& trace* (
% && (assocr* (*) id) (;) 
% && assocr* (;)
% && (id (*) swap*) (;)
% && assocl* (;)
% && (assocr* (*) id) (;)
% && (clone2 (*) id) (;)
% && swap* (;)
% && (swap* (*) id) (;)
% && ((swap* (*) id) (*) id) (;)
% && assocl* (;)
% && assocl* (;)
% && (sat_f (*) id) (;)
% && (assocr* (*) id) (;)
% && (swap* (*) id) (;) 
% && ((id (*) swap*) (*) id) (;)
% && (assocl* (*) id) (;)
% && ((id (*) swap*) (*) id))

We can test our {{solve_f}} combinator using several SAT
instances. Here are two possible instances. The first instance is
satisfied by {{(false,false)}} and the
second is satisfied by {{(false,true)}} and {{(true,true)}}.

%subcode{opsem} include main
%! columnStyle = rcl
% iso_{f_1} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
% iso_{f_1} &=& (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
% && (((swap+ (*) id) (*) id) (;) toffoli (;) ((swap+ (*) id) (*) id)) (;) 
% && (id (*) swap+)
% 
% iso_{f_2} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
% iso_{f_2} &=& toffoli

It can indeed be verified using the semantics that the {{solve_f}}
combinators instantiated with the SAT instances produce the expected
results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Idea}

Our development so far has been fairly traditional in the sense that
we have applied well-known type constructions and semantic techniques
to a language of isomorphisms. In this chapter, we begin to explore
more novel ideas.

One of the benefits of having a reversible language in which
information is preserved is that essentially every construction can be
considered in reverse. We begin in this chapter by considering the
idea of \emph{fractional types} which are dual to product types. We
call the resulting language {{langRevTF}}.

We have already seen in Sec.~\ref{sec:trace} that the multiplicative
trace can be viewed as performing some kind of ``division'' on
types. Recall that given a combinator {{c : b * b1 <-> b * b2}}, we
have {{trace* c : b1 <-> b2}} which \emph{cancels} the type {{b}}.
Our plan is to make this ``division operation'' a first-class one. In
other words, we introduce a new type construction --- a fractional
type {{1/b}} --- that allows us to cancel any type {{b}} in a
multiplicative context, i.e., that satisfies the isomorphism:

%subcode{bnf} include main
%  eps*_b :& 1/b * b ~~<->~~ 1 &: eta*_b

Given this fractional type and its associated isomorphism, we can
perform the cancellation implicit in {{trace*}} as follows. 
Let {{ c :  b * b1 <-> b * b2 }} be given and consider the 
following sequence of type isomorphisms:

%subcode{bnf} include main
%  b1 &<->& 1 * b1 & (uniti)
%     &<->& (1/b * b) * b1 & (eta*_b)
%     &<->& 1/b * (b * b1) & (assocr*)
%     &<->& 1/b * (b * b2) & (c)
%     &<->& ((1/b * b) * b2) & (assocl*)
%     &<->& (1 * b2) & (eps*_b)
%     &<->& b2 & (unite*)

\noindent This sequence of isomorphisms, at least at the type level,
seems to produce the desired {{trace*}}. We will confirm this in the
next section after we introduce the formal syntax and semantics. For
now, we provide an intuitive understanding of the type {{1/b}}.

If {{v : b}}, we define the value {{1/v : 1/b}} and these are the only
values of type~{{1/b}}. Since there are no values of type {{0}}, this
means that {{1/0}} is also an empty type. Generally, we think of a
value like {{1/false}} as a \emph{first-class constraint} that can
only be satisfied if it is matched with an actual value {{false}}. So
intuitively, {{trace*}} allowed us to implicitly introduce constraints
in our model which we now make explicit and first-class entities. This
makes the language much more expressive: as we show in this chapter,
it essentially gains the expressive power to model higher-order
relations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Language: {{langRevTF}} }

We extend {{langRevT}} with fractional types, fractional values, and
two combinators {{eta*_b}} and {{eps*_b}} motivated above.  We provide the
full syntactic and semantic definitions for the language below.

\begin{definition}[Syntax of {{langRevT}}]
\label{def:langRevT}
The syntax of types, values, and combinators, is:
%subcode{bnf} include main
% natural numbers, n 
% value types, b ::= 0 | 1 | b+b | b*b | 1/b | bool | nat 
% values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
%
% combinator types, t ::= b <-> b
% isomorphisms, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
%                   &|& unite | uniti | swap* | assocl* | assocr* 
%                   &|& dist0 | factor0 | dist | factor 
%                   &|& foldB | unfoldB | foldN | unfoldN
%                   &|& eta*_b | eps*_b
% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
%                   &|& trace+ c | trace* c
\end{definition}

The denotation of types is extended with a clause for fractional types.

\begin{definition}[Denotation of Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 0 \rrbracket &=& \emptyset \\
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/v \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket \mathit{nat} \rrbracket &=& \{ 0, 1, 2, \ldots \} 
\end{array}\]
\end{definition}

As usual, the semantics of {{langRevTF}} is given using two mutually
recursive interpreters: one going forward and one going backwards. As
motivated and explained in Sec.~\ref{sec:trace}, each combinator denotes
a \emph{relation} and hence the result of applying a combinator to a
value generally returns a \emph{set} of values. We only show the new
cases for {{eta*_b}} and {{eps*_b}}.

\begin{definition}[Operational semantics for {{langRevTF}}]
\label{def:operational-langRevTF}
Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
forward direction by supplying it with a value {{ v1 : b1 }} and in
the backwards direction by supplying it with a value {{v2 : b2}}.

\paragraph*{Forward Evaluator (clauses for {{eta*_b}} and {{eps*_b}}): }

%subcode{opsem} include main
%! columnStyle = rcl
% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2

\paragraph*{Backwards Evaluator (clauses for {{eta*_b}} and {{eps*_b}}): }

%subcode{opsem} include main
%! columnStyle = rcl
% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
\end{definition}

The language {{langRevT}} is still logically reversible.

\begin{proposition}[Logical Reversibility]
\label{chx:prop:logrev-tracep}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have {{v2 in (evalF c v1)}} iff
{{v1 in (evalB c v2)}}.
\end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Expressiveness}

Fractional types and values add considerable expressiveness to our
language.

%%%%%%%%%%%%
\subsection{Trace}

With fractionals, the trace operator becomes expressible:

%subcode{opsem} include main
%! columnStyle = rcl
% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite

Let's use the semantics to calculate the result of applying
{{trace*_{bool} (not (*) id)}} to {{false}}. We start
with the set {{ {[ false ]} }} which gets transformed as follows:
%subcode{opsem} include main
%! columnStyle = ll
% {[ ((),false) ]}  & (uniti)
% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool})
% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (not (*) id)
% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
% emptyset & eps*_{bool}

And indeed this is the annhiliation circuit from
Ex.~\ref{ch3:ex;annihilation}.

%%%%%%%%%%%%
\subsection{First-Class Relations}

With fractional types, we can turn define a value that denotes a
first-class function or relation. Indeed a value of type {{1/b1 * b2}}
is a pair of a \emph{constraint} that can only be satisfied by some
{{v1 : b1}} and a value {{v2 : b2}}. In other words, it corresponds to
a function or relation which when given matched with a value {{v1 : b1}} 
``releases'' the value {{v2 : b2}}. In the remainder of the book, we 
introduce this abbreviation:

%subcode{bnf} include main
%  b1 :-* b2 ::= 1/b1 * b2

What is remarkable is that we can systematically turn any combinator
{{c : b1 <-> b2}} into a (constant) value of type {{b1 :-* b2}} as
shown below:

%subcode{opsem} include main
%! columnStyle = rcl
% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
% name c &=& eta*_{b1} (;) (id (*) c)

As a simple example, consider the combinator 
{{name not : 1 <-> (bool :-* bool)}}. Applying this 
combinator to {{ () }} evaluates as follows:

%subcode{opsem} include main
%! columnStyle = ll
% {[ () ]}  & 
% {[ (1/false,false), (1/true,true) ]} & eta*_{bool}
% {[ (1/false,true), (1/true,false) ]} & id (*) not

\noindent In other words, the result of applying the combinator to {{ () }}
is a value that represents the boolean negation function.

%%%%%%%%%%%%
\subsection{Higher-Order Relations}

We are now a small step to implementing various higher-order
combinators that manipulate functions or relations. In particular, we
can \emph{apply} and \emph{compose} values representing relations as
shown below:

%subcode{opsem} include main
%! columnStyle = rcl
% apply &:& (b1 :-* b2) * b1 <-> b2
% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite

\noindent At the type level, the computation proceeds as follows:
%subcode{bnf} include main
%  (b1 :-* b2) * b1 &=& (1/b1 * b2) * b1 & 
%     &<->& b1 * (1/b1 * b2) & (swap*)
%     &<->& (b1 * 1/b1) * b2) & (assocl*)
%     &<->& (1/b1 * b1) * b2) & (swap* (*) id)
%     &<->& 1 * b2 & (eps*{b1} (*) id)
%     &<->& b2 & (unite)

\noindent Intuitively, we simply match the incoming argument of type {{b1}} with
the constraint encoded by the function. If they match, they cancel
each other and the value of type {{b2}} is exposed with no
constraints. Otherwise, the mismatched values are \emph{annihilated}.

Function or relation composition is only slightly more involved:

%subcode{opsem} include main
%! columnStyle = rcl
% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
% compose &=& assocr* (;) (id (*) (assocl* (;) (swap* (*) id) (;) (eps*_{b2} (*) id) (;) unite))

\noindent Again, at the type level, the computation proceeds as follows:

%subcode{bnf} include main
%  (b1 :-* b2) * (b2 :-* b3) &=& (1/b1 * b2) * (1/b2 * b3) & 
%     &<->& (1/b1 * (b2 * (1/b2 * b3)) & (assocr*)
%     &<->& (1/b1 * ((b2 * 1/b2) * b3) & (assocl*)
%     &<->& (1/b1 * ((1/b2 * b2) * b3) & (assocl*)
%     &<->& (1/b1 * (1 * b3) & (eps*_{b2})
%     &<->& (1/b1 * b3) & (unite)
%     &=& b1 :-* b3

%%%%%%%%%%%%
\subsection{Duality}

The final remarkable property of fractional types is that they
constitute, a form of self-inverse negation. In particular, we have:

%subcode{opsem} include main
%! columnStyle = rcl
% doubleNeg &:& b <-> 1/(1/b)
% doubleNeg &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite

\noindent At the type level, this computation proceeds as follows:

%subcode{bnf} include main
%  b &<->& (1 * b) & (uniti)
%    &<->& (1/(1/b) * 1/b) * b & (eta*_{1/b} (*) id)
%    &<->& 1/(1/b) * (1/b * b) & (assocr*)
%    &<->& 1/(1/b) * 1 & (id (*) eps*_b)
%    &<->& 1 * 1/(1/b) & (swap*)
%    &<->& 1/(1/b) & (unite)

What is interesting about this isomorphism is that it holds for all
types {{b}}, and in particular we have {{0 <-> 1/(1/0)}}.

\bibliographystyle{splncs03} 
\bibliography{cites}
\end{document}

### 2013-01-22 code/frac/Pi.hs
added code in frac dir

@@ -1,0 +1,215 @@
+{-# LANGUAGE TypeOperators, GADTs #-}
+
+module Pi where
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+data Zero 
+
+data a :<=> b where 
+-- Congruence
+  Id    :: a :<=> a
+  Sym   :: (a :<=> b) -> (b :<=> a) 
+  (:.:) :: (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  ZeroE   :: Either Zero a :<=> a
+  ZeroI   :: a :<=> Either Zero a
+  SwapP :: Either a b :<=> Either b a
+  AssocLP  :: Either a (Either b c) :<=> Either (Either a b) c 
+  AssocRP  :: Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE    :: ((), a) :<=> a
+  UnitI    :: a :<=> ((), a)
+  SwapT :: (a,b) :<=> (b,a) 
+  AssocLT  :: (a,(b,c)) :<=> ((a,b),c)
+  AssocRT  :: ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  DistribZ  :: (Zero, a) :<=> Zero
+  FactorZ  :: Zero :<=> (Zero, a)
+  Distrib  :: (Either b c, a) :<=> Either (b, a) (c, a)
+  Factor      :: Either (b, a) (c, a) :<=> (Either b c, a)
+-- Encoding of booleans
+  FoldB   :: Either () () :<=> Bool
+  UnfoldB :: Bool :<=> Either () ()
+
+-- Adjoint
+
+adjoint :: (a :<=> b) -> (b :<=> a)
+adjoint Id = Id
+adjoint (Sym f) = f 
+adjoint (f :.: g) = adjoint g :.: adjoint f
+adjoint (f :*: g) = adjoint f :*: adjoint g
+adjoint (f :+: g) = adjoint f :+: adjoint g
+adjoint ZeroE = ZeroI
+adjoint ZeroI = ZeroE
+adjoint SwapP = SwapP
[…]

### 2013-01-22 code/frac/PiFrac.hs

@@ -1,0 +1,772 @@
+{-# LANGUAGE TypeOperators, GADTs, TypeSynonymInstances #-}
+
+module PiFrac where
+
+import Data.List
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+type Nat = Int
+
+data Zero 
+data Recip a = Recip a deriving (Show, Eq)
+
+class Eq a => B a where
+  elems :: [a]
+
+instance Eq Zero where
+  _ == _ = undefined
+  
+instance B Zero where
+  elems = []
+  
+instance B () where  
+  elems = [()]
+  
+instance (B a, B b) => B (Either a b) where
+  elems = map Left elems ++ map Right elems
+  
+instance B Bool where
+  elems = [False, True]
+
+instance B Nat where
+  elems = [0..]
+
+instance (B a, B b) => B (a,b) where
+  elems = [(a,b) | a <- elems, b <- elems]
+
+instance B a => B (Recip a) where
+  elems = map Recip elems
+
+data a :<=> b where 
+-- Congruence
+  Id    :: B a => a :<=> a
+  Sym   :: (B a, B b) => (a :<=> b) -> (b :<=> a) 
+  (:.:) :: (B a, B b, B c) => (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (B a, B b, B c, B d) => (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (B a, B b, B c, B d) => (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  ZeroE   :: B a => Either Zero a :<=> a
+  ZeroI   :: B a => a :<=> Either Zero a
+  SwapP :: (B a, B b) => Either a b :<=> Either b a
+  AssocLP  :: (B a, B b, B c) => Either a (Either b c) :<=> Either (Either a b) c 
+  AssocRP  :: (B a, B b, B c) => Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE    :: B a => ((), a) :<=> a
+  UnitI    :: B a => a :<=> ((), a)
+  SwapT :: (B a, B b) => (a,b) :<=> (b,a) 
[…]

### 2013-01-22 code/frac/PiNeg.hs

@@ -1,0 +1,772 @@
+{-# LANGUAGE TypeOperators, GADTs, TypeSynonymInstances #-}
+
+module PiNeg where
+
+import Data.List
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+type Nat = Int
+
+data Zero 
+data Neg a = Neg a deriving (Show, Eq)
+
+class Eq a => B a where
+  elems :: [a]
+
+instance Eq Zero where
+  _ == _ = undefined
+  
+instance B Zero where
+  elems = []
+  
+instance B () where  
+  elems = [()]
+  
+instance (B a, B b) => B (Either a b) where
+  elems = map Left elems ++ map Right elems
+  
+instance B Bool where
+  elems = [False, True]
+
+instance B Nat where
+  elems = [0..]
+
+instance (B a, B b) => B (a,b) where
+  elems = [(a,b) | a <- elems, b <- elems]
+
+instance B a => B (Neg a) where
+  elems = map Neg elems
+
+data a :<=> b where 
+-- Congruence
+  Id    :: B a => a :<=> a
+  Sym   :: (B a, B b) => (a :<=> b) -> (b :<=> a) 
+  (:.:) :: (B a, B b, B c) => (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (B a, B b, B c, B d) => (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (B a, B b, B c, B d) => (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  ZeroE   :: B a => Either Zero a :<=> a
+  ZeroI   :: B a => a :<=> Either Zero a
+  SwapP :: (B a, B b) => Either a b :<=> Either b a
+  AssocLP  :: (B a, B b, B c) => Either a (Either b c) :<=> Either (Either a b) c 
+  AssocRP  :: (B a, B b, B c) => Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE    :: B a => ((), a) :<=> a
+  UnitI    :: B a => a :<=> ((), a)
+  SwapT :: (B a, B b) => (a,b) :<=> (b,a) 
[…]

### 2013-01-29 frac.tex
frac update
fixed macros
fixed typo
reorganization

@@
 }
+
+\newcommand{\todo}[1]{\textbf{TODO:} #1}
 
@@
 %! (+) = +
+%! foldB = fold_B
+%! unfoldB = unfold_B
+%! foldN = fold_N
+%! unfoldN = unfold_N
+%! trace+ = \mathit{trace}^{+}
+%! trace* = \mathit{trace}^{\times}
+%! :-* = \multimap^{\times}
+%! :-+ = \multimap^{+}
+%! emptyset = \emptyset
 
@@
 %! v] = \lrcorner
+%! eta* = \eta^{\times}
+%! eps* = \epsilon^{\times}
+%! Union = \bigcup
+%! in = \in
 %! |-->* = \mapsto^{*}
@@
 %! |- = \vdash
-%! in = \!\!\in\!\!
 %! <=> = \Longleftrightarrow
@@
 %! least = \phi
-%! trace+ = \mathit{trace}^{+}
-%! trace* = \mathit{trace}^{\times}
 %! loop+ = loop_{+}
@@
 %! * = \times
+%! langRevT = \Pi^{o}
+%! langRevTF = \Pi^{o/}
 
@@
 \title{Fractional Types}
-\author{}
+\author{Roshan P. James \and Zachary Sparks \and Jacques Carette \and Amr Sabry}
 \institute{}
@@
 \begin{abstract}
-?
+  In previous work, we developed a \emph{first-order},
+  information-preserving, and reversible programming language {{langRev}}
+  founded on type isomorphisms. Being first-order limits the expressiveness
+  of the language: it is not possible, for example, to abstract program
+  fragments into a higher-level combinator. More importantly, however, the
+  limitation to first-order types means that the class of isomorphisms
+  between two types {{b1}} and {{b2}} cannot be itself represented as a type
+  and hence that the natural notion of equivalence between two {{langRev}}
+  computations is no longer based on isomorphisms. In this paper, we
+  introduce a higher-order extension of {{langRev}} based on the novel
+  concept of \emph{fractional types}. Intuitively, a value of a fractional
+  type represents \emph{negative} information: such fractional values can be
+  freely propagated and transformed but must ultimately --- in a complete
+  program --- be offset by the corresponding amount of positive information.
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Syntax and Semantics}
+\section{Background: {{langRev}} }
 
-We can turn the above system of isomorphisms into a (reversible)
-programming language by simply providing a syntax for each of the
-isomorphisms. For each isomorphism, we introduce two
-\emph{combinators} corresponding to the left-to-right and
-right-to-left directions.
+\todo{brief explanation}
 
@@
 
-\noindent To summarize the syntax of our language {{langRev}} is as follows.
-
-\begin{definition}[Syntax of {{langRev}}]
-\label{def:langRev}
-We collect our types, values, and combinators, to get the full language
-definition:
-%subcode{bnf} include main
-% value types, b ::= 0 | 1 | b+b | b*b 
-% values, v ::= () | left v | right v | (v,v) 
-%
-% combinator types, t ::= b <-> b
-% primitive combinators, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
-%                   &|& unite | uniti | swap* | assocl* | assocr* 
-%                   &|& dist0 | factor0 | dist | factor 
-% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-\end{definition}
-
-By design, for each combinator {{c : b1 <-> b2}}, there is an
-\emph{adjoint} {{c^dagger : b2 <-> b1}} acting in the reverse direction.
-
-\begin{definition}[Adjoints {{c^dagger}}]
- The adjoint of a combinator {{c}} is defined by {{c^dagger}} as
- follows. For primitive combinators {{c}}, {{c^dagger}} is given by
- the inverse in Def.~\ref{def:primitives-langRev}. For the remaining
- combinators, we have:
-%subcode{opsem} include main
-% id^dagger &=& id
-% (sym c)^dagger &=& c
-% (c1(;)c2)^dagger &=& c2^dagger (;) c1^dagger
-% (c1 (+) c2)^dagger &=& c1^dagger (+) c2^dagger
-% (c1 (*) c2)^dagger &=& c1^dagger (*) c2^dagger
-which essentially amounts to reversing the order of sequential composition.
-\end{definition}
-
 The semantics of {{langRev}} is given using two mutually recursive
-interpreters: one going forward and one going backwards. The use of
-{{sym}} switches control from one evaluator to the other.
-
-\begin{definition}[Operational Semantics of {{langRev}}]
-\label{def:operational-langRev}
-Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
-forward direction by supplying it with a value {{ v1 : b1 }} and in
-the backwards direction by supplying it with a value {{v2 : b2}}. The
-forward evaluator {{evalF}} and the backwards evaluator {{evalB}} are
-given below. Since there are no values that have the type {{0}}, the
-clauses for the combinators {{zeroe}}, {{zeroi}}, {{dist0}} and
-{{factor0}} omit the impossible cases.
-
-\paragraph*{Forward Evaluator:}
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalF zeroe (right v) &=& v
-% evalF zeroi v &=& right v
-% evalF swap+ (left v) &=& right v
-% evalF swap+ (right v) &=& left v 
-% evalF assocl+ (left v) &=& left (left v)
-% evalF assocl+ (right (left v)) &=& left (right v)
-% evalF assocl+ (right (right v)) &=& right v
-% evalF assocr+ (left (left v)) &=& left v
-% evalF assocr+ (left (right v)) &=& right (left v)
-% evalF assocr+ (right v) &=& right (right v)
-% evalF unite ((), v) &=& v 
-% evalF uniti v &=& ((), v) 
-% evalF swap* (v1, v2) &=& (v2, v1) 
-% evalF assocl* (v1, (v2, v3)) &=& ((v1, v2), v3) 
-% evalF assocr* ((v1, v2), v3) &=& (v1, (v2, v3)) 
-% evalF dist (left v1, v3) &=& left (v1, v3)
-% evalF dist (right v2, v3) &=& right (v2, v3)
-% evalF factor (left (v1, v3)) &=& (left v1, v3) 
-% evalF factor (right (v2, v3)) &=& (right v2, v3)   
-% evalF id v &=& v
-% evalF (sym c) v &=& evalB c v
-% evalF (c1(;)c2) v &=& evalF c2 (evalF c1 v)
-% evalF (c1 (+) c2) (left v1) &=& left (evalF c1 v1)
-% evalF (c1 (+) c2) (right v2) &=& right (evalF c2 v2)
-% evalF (c1 (*) c2) (v1,v2) &=& (evalF c1 v1, evalF c2 v2)
-
-\paragraph*{Backwards Evaluator:}
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalB zeroe v &=& right v 
-% evalB zeroi (right v) &=& v
-% evalB swap+ (left v) &=& right v
-% evalB swap+ (right v) &=& left v
-% evalB assocl+ (left (left v)) &=& left v
-% evalB assocl+ (left (right v)) &=& right (left v)
-% evalB assocl+ (right v) &=& right (right v)
-% evalB assocr+ (left v) &=& left (left v)
-% evalB assocr+ (right (left v)) &=& left (right v) 
-% evalB assocr+ (right (right v)) &=& right v
-% evalB unite v &=& ((), v) 
-% evalB uniti ((), v) &=& v
-% evalB swap* (v1, v2) &=& (v2, v1) 
-% evalB assocl* ((v1, v2), v3) &=& (v1, (v2, v3)) 
-% evalB assocr* (v1, (v2, v3)) &=& ((v1, v2), v3) 
-% evalB dist (left (v1, v3)) &=& (left v1, v3)
-% evalB dist (right (v2, v3)) &=& (right v2, v3) 
-% evalB factor (left v1, v3) &=& left (v1, v3) 
-% evalB factor (right v2, v3) &=& right (v2, v3) 
-% evalB id v &=& v
-% evalB (sym c) v &=& evalF c v
-% evalB (c1(;)c2) v &=& evalB c1 (evalB c2 v)
-% evalB (c1 (+) c2) (left v1) &=& left (evalB c1 v1)
-% evalB (c1 (+) c2) (right v2) &=& right (evalB c2 v2)
-% evalB (c1 (*) c2) (v1,v2) &=& (evalB c1 v1, evalB c2 v2)
-\end{definition}
+interpreters: one going forward and one going backwards. The use of {{sym}}
+switches control from one evaluator to the other. We will not present the
+semantics because we generalize in the next section.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Trace Operators: Intuition}
+\todo{mention that we have extended this with trace and recursive types but
+  that we are going to explore something orthogonal}
 
-The language {{langRev}} is not yet universal: one of the missing
-ingredients is a form of recursion or iteration. Mathematically
-speaking, recursion and iteration can be expressed using categorical
-trace
-operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}. We
-will work out the precise categorical connections later: for now we
-pursue our operational approach.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Language: {{langRevTF}} }
 
-\paragraph*{Additive Case.} 
-The intuitive idea of trace operators is as follows. We consider first
-the additive case. Given a computation {{c:b1+b2<->b1+b3}}, we can
-\emph{cancel} the common type {{b1}} from both the input and the
-output by feeding back to itself, thereby getting a new combinator
-{{trace+~c:b2<->b3}}. Diagrammatically, we have:
+As hinted to in the first section, the semantics of fractional types introduces
+a form of non-determinism in the sense that there could zero, one, or several
+values that satisfy the constraints imposed by fractional values. To model
+such non-determinism, we specify the semantics using \emph{relations}. In
+other words, the results of {{evalF c v}} or {{evalB c v}} are no longer
+guaranteed to be a single value: they could be \emph{sets} of values.
 
-\inkscape{trace_plus.pdf}
+The introduction of relations makes sequential composition of
+combinators more involved than simply passing the value produced by
+the first to the second. The following example illustrates our
+notation for the composition of relations which is used in the formal
+semantics.
 
-Operationally, the execution of such a combinator proceeds as
-follows. We are given a value {{v2 : b2}} which is injected into the
-type {{b1 + b2}} by tagging it as {{right v2}}. This tagged value is
-passed to the combinator {{f : b1 + b2 <-> b1 + b3}}. The resulting
-value will either be {{left v1}} for some {{v1 : b1}} or {{right v3}}
-for some {{v3 : b3}}. In the former case, the value {{left v1}} is fed
-back to {{f}}. In the latter case, the entire {{trace+}}-combinator
-terminates with value {{v3}}. An example should make this more
-concrete.
-
-\begin{example}
-Consider the combinator {{c:bool+1<->bool+1}} below:
-
-\inkscape{trace-example0.pdf}
-
-\noindent
-When we trace {{c}} we have the circuit {{trace+ c: 1 <-> 1}} shown below.
-
-\inkscape{trace-example1.pdf}
-
-\noindent
-The following diagrams show snapshots the evaluation of the
-circuit applied to {{()}}:
-\begin{multicols}{2}
-  \inkscape{trace-example2.pdf}
-  \inkscape{trace-example3.pdf}
-\end{multicols}
-
-\begin{multicols}{2}
-  \inkscape{trace-example4.pdf}
-  \inkscape{trace-example5.pdf}
-\end{multicols}
-
-\begin{multicols}{2}
-  \inkscape{trace-example6.pdf}
-  \inkscape{trace-example7.pdf}
-\end{multicols}
-
-\noindent
-The value starts at the input to the circuit, flows out on the
-{{left}} branch of {{c}} and is traced back into {{c}}.  The feedback
-loop is taken exactly two times in the execution of this circuit.
+\begin{example}[Composition of Relations]
+Consider the two relations:
+\[\begin{array}{rcl}                                                                       
+R_1 &=& \{ (1,2),(1,3),(2,2) \} \\                                                       
+R_2 &=& \{ (2,1),(2,3),(3,1) \}                                                          
+\end{array}\]
+These relations can be expressed in a functional notation as follows:
+\[\begin{array}{rcl}                                                                       
+R_1 (1) &=& \{ 2,3 \} \\                                                                 
+R_1 (2) &=& \{ 2 \} \\                                                                   
+R_1 (3) &=& \emptyset \\
+\\ 
+R_2 (1) &=& \emptyset \\                                                                 
+R_2 (2) &=& \{ 1,3 \} \\                                                                 
+R_2 (3) &=& \{ 1 \}                                                                      
+\end{array}\]
+Using the functional notation, the composition $(R_2 \circ R_1)$ is defined as
+follows:
+\[\begin{array}{rcl}                                                                   
+(R_2 \circ R_1)~(a) &=& \bigcup_{b \in R_1(a)} R_2(b)                                  
+\end{array}\]
+We calculate the composition:
+\[\begin{array}[t]{rcl}                                                                  
+(R_2 \circ R_1)~(1) &=& \bigcup_{b \in R_1(1)} R_2(b) \\                               
+         &=& R_2(2) \cup R_2(3) \\                                                     
+         &=& \{1,3\} \cup \{ 1 \} \\                                                   
+         &=& \{ 1,3 \} \\                                                              
+(R_2 \circ R_1)~(2) &=& \{ 1,3 \} \\                                                   
+(R_2 \circ R_1)~(3) &=& \emptyset                                                      
+\end{array}\]
+In other words, 
+\[
+(R_2 \circ R_1) = \{(1,1), (1,3), (2,1), (2,3)\}
+\]
 \end{example}
 
-\paragraph*{Multiplicative Case.} 
-The situation with the multiplicative trace is a bit more subtle. In
-this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
-want to cancel the common type {{b1}} as before, to produce a new
-combinator {{trace*~c:b2<->b3}}. For the evaluation of such a
-combinator, we are only given a value {{v2 : b2}} but to evaluate
-{{c}}, we must provide a pair consisting of some value {{v1 : b1}}
-together with the given {{v2 : b2}}. The needed value {{v1 : b1}}
-cannot be arbitrary, however. We do have a constraint that the value
-{{v1 : b1}} must be such that it is also produced by as the first
-component of the result, i.e., that {{evalF c (v1,v2) = (v1,v3)}}. In
-general, there may be several such values fixed-point values or
-none. A few examples should help understand the subtleties.
+We extend {{langRevT}} with fractional types, fractional values, and two
+combinators {{eta*_b}} and {{eps*_b}}.  We provide the full syntactic and
+semantic definitions for the language below.
+
+\begin{definition}[Syntax of {{langRevT}}]
+\label{def:langRevT}
+The syntax of types, values, and combinators, is:
+%subcode{bnf} include main
+% natural numbers, n 
+% value types, b ::= 0 | 1 | b+b | b*b | 1/b | bool | nat 
+% values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
+%
+% combinator types, t ::= b <-> b
+% isomorphisms, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
+%                   &|& unite | uniti | swap* | assocl* | assocr* 
+%                   &|& dist0 | factor0 | dist | factor 
+%                   &|& foldB | unfoldB | foldN | unfoldN
+%                   &|& eta*_b | eps*_b
+% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+%                   &|& trace+ c | trace* c
+\end{definition}
+
+If {{v : b}}, we define the value {{1/v : 1/b}} and these are the only
+values of type~{{1/b}}. Since there are no values of type {{0}}, this
+means that {{1/0}} is also an empty type. Generally, we think of a
+value like {{1/false}} as a \emph{first-class constraint} that can
+only be satisfied if it is matched with an actual value {{false}}.
+
+\begin{definition}[Denotation of Types {{ [[b]] }}]
+\label{chx:def:denot}
+Each type denotes a set of values as follows:
+\[\begin{array}{rcl}
+\llbracket 0 \rrbracket &=& \emptyset \\
+\llbracket 1 \rrbracket &=& \{ () \} \\
+\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
+           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
+\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
+           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
+\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} \\
+\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
+\llbracket \mathit{nat} \rrbracket &=& \{ 0, 1, 2, \ldots \} 
+\end{array}\]
+\end{definition}
+
+The semantics of {{langRevTF}} is given using two mutually recursive
+interpreters: one going forward and one going backwards. Unlike the case for
+{{langRev}}, each combinator denotes a \emph{relation} and hence the result
+of applying a combinator to a value generally returns a \emph{set} of
+values.
+
+\begin{definition}[Operational semantics for {{langRevTF}}]
+\label{def:operational-langRevTF}
+Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
+forward direction by supplying it with a value {{ v1 : b1 }} and in
+the backwards direction by supplying it with a value {{v2 : b2}}.
+
+\paragraph*{Forward Evaluator:}
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% evalF zeroe (right v) &=& {[ v ]}
+% evalF zeroi v &=& {[ right v ]}
+% evalF swap+ (left v) &=& {[ right v ]}
+% evalF swap+ (right v) &=& {[ left v ]}
+% evalF assocl+ (left v) &=& {[ left (left v) ]}
+% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
+% evalF assocl+ (right (right v)) &=& {[ right v ]}
+% evalF assocr+ (left (left v)) &=& {[ left v ]}
+% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
+% evalF assocr+ (right v) &=& {[ right (right v) ]}
+% evalF unite ((), v) &=& {[ v ]}
+% evalF uniti v &=& {[ ((), v) ]}
+% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
+% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
+% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
+% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
+% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
+% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
+% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
+% evalF id v &=& {[ v ]}
+% evalF (sym c) v &=& evalB c v
+% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
+% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
+% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
+% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
+% evalF (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalF c (v,v2)) ]}
+% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
+% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
+% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2
+
+\paragraph*{Backwards Evaluator:}
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% evalB zeroe v &=& {[ right v ]}
+% evalB zeroi (right v) &=& {[ v ]}
+% evalB swap+ (left v) &=& {[ right v ]}
+% evalB swap+ (right v) &=& {[ left v ]}
+% evalB assocl+ (left (left v)) &=& {[ left v ]}
+% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
+% evalB assocl+ (right v) &=& {[ right (right v) ]}
+% evalB assocr+ (left v) &=& {[ left (left v) ]}
+% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
+% evalB assocr+ (right (right v)) &=& {[ right v ]}
+% evalB unite v &=& {[ ((), v) ]}
+% evalB uniti ((), v) &=& {[ v ]}
+% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
+% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
+% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
+% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
+% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
+% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
+% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
+% evalB id v &=& {[ v ]}
+% evalB (sym c) v &=& evalF c v
+% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
+% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
+% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
+% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
+% evalB (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalB c (v,v2)) ]}
+% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
+% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
+% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
+\end{definition}
+
+The language {{langRevT}} is still logically reversible.
+
+\begin{proposition}[Logical Reversibility]
+\label{chx:prop:logrev-tracep}
+For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
+{{v2 : b2}} we have {{v2 in (evalF c v1)}} iff
+{{v1 in (evalB c v2)}}.
+\end{proposition}
+
+Every relation is clearly reversible by simply swapping the input and
+output components of every pair, i.e., if {{r = {[ (a,b), ... ]} }}
+then the inverse of {{r}} is simply {{ {[ (b,a), ... ]} }}.
+
+\begin{proposition}[Logical Reversibility]
+\label{prop:logrev-tracet}
+For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
+{{v2 : b2}} we have if {{v2 in (evalF c v1)}} then 
+{{v1 in (evalB c v2)}}.
+\end{proposition}
+
+An interesting point to ponder is what happens to \emph{information}
+in circuits like {{annihilate}}? We return to this point later.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Expressiveness}
+
+Fractional types and values add considerable expressiveness to our
+language.
+
+%%%%%%%%%%%%
+\subsection{First-Class Relations}
+
+With fractional types, we can define a value that denotes a first-class
+function or relation. Indeed a value of type {{1/b1 * b2}} is a pair of a
+\emph{constraint} that can only be satisfied by some {{v1 : b1}} and a value
+{{v2 : b2}}. In other words, it corresponds to a function or relation which
+when given matched with a value {{v1 : b1}} ``releases'' the value 
+{{v2 : b2}}. In the remainder of the paper, we introduce this abbreviation:
+
+%subcode{bnf} include main
+%  b1 :-* b2 ::= 1/b1 * b2
+
+What is remarkable is that we can systematically turn any combinator
+{{c : b1 <-> b2}} into a (constant) value of type {{b1 :-* b2}} as
+shown below:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
+% name c &=& eta*_{b1} (;) (id (*) c)
+
+As a simple example, consider the combinator 
+{{name not : 1 <-> (bool :-* bool)}}. Applying this 
+combinator to {{ () }} evaluates as follows:
+
+%subcode{opsem} include main
+%! columnStyle = ll
+% {[ () ]}  & 
+% {[ (1/false,false), (1/true,true) ]} & eta*_{bool}
+% {[ (1/false,true), (1/true,false) ]} & id (*) not
+
+\noindent In other words, the result of applying the combinator to {{ () }}
+is a value that represents the boolean negation function.
+
+%%%%%%%%%%%%
+\subsection{Higher-Order Relations}
+
+We are now a small step to implementing various higher-order
+combinators that manipulate functions or relations. In particular, we
+can \emph{apply} and \emph{compose} values representing relations as
+shown below:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% apply &:& (b1 :-* b2) * b1 <-> b2
+% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
+
+\noindent At the type level, the computation proceeds as follows:
+%subcode{bnf} include main
+%  (b1 :-* b2) * b1 &=& (1/b1 * b2) * b1 & 
+%     &<->& b1 * (1/b1 * b2) & (swap*)
+%     &<->& (b1 * 1/b1) * b2) & (assocl*)
+%     &<->& (1/b1 * b1) * b2) & (swap* (*) id)
+%     &<->& 1 * b2 & (eps*{b1} (*) id)
+%     &<->& b2 & (unite)
+
+\noindent Intuitively, we simply match the incoming argument of type {{b1}} with
+the constraint encoded by the function. If they match, they cancel
+each other and the value of type {{b2}} is exposed with no
+constraints. Otherwise, the mismatched values are \emph{annihilated}.
+
+Function or relation composition is only slightly more involved:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
+% compose &=& assocr* (;) (id (*) (assocl* (;) (swap* (*) id) (;) (eps*_{b2} (*) id) (;) unite))
+
+\noindent Again, at the type level, the computation proceeds as follows:
+
+%subcode{bnf} include main
+%  (b1 :-* b2) * (b2 :-* b3) &=& (1/b1 * b2) * (1/b2 * b3) & 
+%     &<->& (1/b1 * (b2 * (1/b2 * b3)) & (assocr*)
+%     &<->& (1/b1 * ((b2 * 1/b2) * b3) & (assocl*)
+%     &<->& (1/b1 * ((1/b2 * b2) * b3) & (assocl*)
+%     &<->& (1/b1 * (1 * b3) & (eps*_{b2})
+%     &<->& (1/b1 * b3) & (unite)
+%     &=& b1 :-* b3
+
+%%%%%%%%%%%%
+\subsection{Duality}
+
+The final remarkable property of fractional types is that they
+constitute, a form of self-inverse negation. In particular, we have:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% doubleNeg &:& b <-> 1/(1/b)
+% doubleNeg &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
+
+\noindent At the type level, this computation proceeds as follows:
+
+%subcode{bnf} include main
+%  b &<->& (1 * b) & (uniti)
+%    &<->& (1/(1/b) * 1/b) * b & (eta*_{1/b} (*) id)
+%    &<->& 1/(1/b) * (1/b * b) & (assocr*)
+%    &<->& 1/(1/b) * 1 & (id (*) eps*_b)
+%    &<->& 1 * 1/(1/b) & (swap*)
+%    &<->& 1/(1/b) & (unite)
+
+What is interesting about this isomorphism is that it holds for all
+types {{b}}, and in particular we have {{0 <-> 1/(1/0)}}.
+
+%%%%%%%%%%%%
+\subsection{Feedback, Iteration, and Trace Operators}
+
+Mathematically speaking, recursion and iteration can be expressed using
+categorical trace
+operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
+language like {{langRev}} there are two natural families of trace operators
+that can be defined, an additive family (explored in detail in our previous
+work) and a multiplicative family (which is the focus of this section).
+
+In this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
+want to cancel the common type {{b1}}, to produce a new combinator
+{{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
+given a value {{v2 : b2}} but to evaluate {{c}}, we must provide a pair
+consisting of some value {{v1 : b1}} together with the given {{v2 : b2}}. The
+needed value {{v1 : b1}} cannot be arbitrary, however. We do have a
+constraint that the value {{v1 : b1}} must be such that it is also produced
+by as the first component of the result, i.e., 
+that {{evalF c (v1,v2) = (v1,v3)}}. In general, there may be several 
+such values fixed-point values or none. 
+A few examples should help understand the subtleties.
 
@@
 
-\begin{example}
-One might conclude from the examples above that the multiplicative
-trace is non-sensical. There are however some excellent and legitimate
-uses illustrated in the remainder of this chapter. A simple
-illustrative example takes a combinator {{c : b <-> b}} and detects
-whether it has a fixed point, i.e., whether there exists a value 
-{{v :  b}} such that {{evalF c v = v}}:
-
-{{hasFixedPoint c = trace* (swap* (;) unite (;) c (;) uniti (;) swap*)}}
-
-\noindent The idea is to introduce a unit value before and after the evaluation of {{c}}
-effectively changing the type of {{c}} to {{b * 1 <-> b * 1}}
-and then use {{trace*}} to
-cancel the value of type~{{b}}. The result of evaluating
-{{(hasFixedPoint c) ()}} is {{()}} if {{c}} has a fixed point and returns 
-no value otherwise. 
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Multiplicative Trace Operators}
-\label{sec:trace}
-
-For the multiplicative trace operator, we similarly add a new family
-of combinators {{trace*_{b} c}} indexed by types {{b}} to our
-syntax. The type rule, can be viewed as a cancellation law that
-eliminates the type~{{b}} from both sides of a product.
-
-\begin{definition}[Type Judgment for {{trace*_{b}}}]
-%subcode{proof} include main
-%@ c : b * b2 <-> b * b3
-%@@ trace*_{b} c : b2 <-> b3  
-\end{definition}
-
-The adjoint of {{trace*_{b} c}} is also the trace of the adjoint of
-{{c}}, i.e., {{(trace*_{b} c)^dagger = trace*_{b} c^dagger}}.
-
-As hinted in the first section, the semantics of {{trace*}} introduces
-a form of non-determinism in the sense that there could zero, one, or
-several values that satisfy the constraint imposed by {{trace*}}. To
-model such non-determinism, we specify the semantics using
-\emph{relations}. In other words, the results of {{evalF c v}} 
-or {{evalB c v}} are no longer guaranteed to be a single value: they 
-could be \emph{sets} of values.
-
-The introduction of relations makes sequential composition of
-combinators more involved than simply passing the value produced by
-the first to the second. The following example illustrates our
-notation for the composition of relations which is used in the formal
-semantics.
-
-\begin{example}[Composition of Relations]
-Consider the two relations:
-\[\begin{array}{rcl}                                                                       
-R_1 &=& \{ (1,2),(1,3),(2,2) \} \\                                                       
-R_2 &=& \{ (2,1),(2,3),(3,1) \}                                                          
-\end{array}\]
-These relations can be expressed in a functional notation as follows:
-\[\begin{array}{rcl}                                                                       
-R_1 (1) &=& \{ 2,3 \} \\                                                                 
-R_1 (2) &=& \{ 2 \} \\                                                                   
-R_1 (3) &=& \emptyset \\
-\\ 
-R_2 (1) &=& \emptyset \\                                                                 
-R_2 (2) &=& \{ 1,3 \} \\                                                                 
-R_2 (3) &=& \{ 1 \}                                                                      
-\end{array}\]
-Using the functional notation, the composition $(R_2 \circ R_1)$ is defined as
-follows:
-\[\begin{array}{rcl}                                                                   
-(R_2 \circ R_1)~(a) &=& \bigcup_{b \in R_1(a)} R_2(b)                                  
-\end{array}\]
-We calculate the composition:
-\[\begin{array}[t]{rcl}                                                                  
-(R_2 \circ R_1)~(1) &=& \bigcup_{b \in R_1(1)} R_2(b) \\                               
-         &=& R_2(2) \cup R_2(3) \\                                                     
-         &=& \{1,3\} \cup \{ 1 \} \\                                                   
-         &=& \{ 1,3 \} \\                                                              
-(R_2 \circ R_1)~(2) &=& \{ 1,3 \} \\                                                   
-(R_2 \circ R_1)~(3) &=& \emptyset                                                      
-\end{array}\]
-In other words, 
-\[
-(R_2 \circ R_1) = \{(1,1), (1,3), (2,1), (2,3)\}
-\]
-\end{example}
-
-\begin{definition}[Operational Semantics for {{trace*_{b}}}]
-The formal semantics is expressed by lifting the forward and backward
-evaluators to relations and then adding the new cases for {{trace*}}.
-
-\paragraph*{Forward Evaluator:}
+With fractionals, the multiplicative trace operator becomes expressible:
 
@@
 %! columnStyle = rcl
-% evalF zeroe (right v) &=& {[ v ]}
-% evalF zeroi v &=& {[ right v ]}
-% evalF swap+ (left v) &=& {[ right v ]}
-% evalF swap+ (right v) &=& {[ left v ]}
-% evalF assocl+ (left v) &=& {[ left (left v) ]}
-% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
-% evalF assocl+ (right (right v)) &=& {[ right v ]}
-% evalF assocr+ (left (left v)) &=& {[ left v ]}
-% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
-% evalF assocr+ (right v) &=& {[ right (right v) ]}
-% evalF unite ((), v) &=& {[ v ]}
-% evalF uniti v &=& {[ ((), v) ]}
-% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
-% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
-% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
-% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
-% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
-% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
-% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
-% evalF id v &=& {[ v ]}
-% evalF (sym c) v &=& evalB c v
-% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
-% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
-% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
-% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
-% evalF (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalF c (v,v2)) ]}
+% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
+% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
+%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite
 
-\paragraph*{Backwards Evaluator:}
+Let's use the semantics to calculate the result of applying
+{{trace*_{bool} (not (*) id)}} to {{false}}. We start
+with the set {{ {[ false ]} }} which gets transformed as follows:
+%subcode{opsem} include main
+%! columnStyle = ll
+% {[ ((),false) ]}  & (uniti)
+% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool})
+% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
+% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (not (*) id)
+% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
+% emptyset & eps*_{bool}
 
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalB zeroe v &=& {[ right v ]}
-% evalB zeroi (right v) &=& {[ v ]}
-% evalB swap+ (left v) &=& {[ right v ]}
-% evalB swap+ (right v) &=& {[ left v ]}
-% evalB assocl+ (left (left v)) &=& {[ left v ]}
-% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
-% evalB assocl+ (right v) &=& {[ right (right v) ]}
-% evalB assocr+ (left v) &=& {[ left (left v) ]}
-% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
-% evalB assocr+ (right (right v)) &=& {[ right v ]}
-% evalB unite v &=& {[ ((), v) ]}
-% evalB uniti ((), v) &=& {[ v ]}
-% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
-% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
-% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
-% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
-% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
-% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
-% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
-% evalB id v &=& {[ v ]}
-% evalB (sym c) v &=& evalF c v
-% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
-% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
-% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
-% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
-% evalB (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalB c (v,v2)) ]}
-\end{definition}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Examples}
 
-In the cases for {{trace*_b}}, we check every value {{v : b}} in the
-denotation of the type {{b}}. Specifically, given {{v2}}, we produce the pair
-{{(v,v2)}} and check if the result of evaluating {{c}} on this pair produces
-the same {{v}} as the first component of its result. In that case, we include
-the second component in the final answer, and otherwise, we reject that
-second component.
-
-Now that we have the formal semantics, we consider a generalized
-version of Ex.~\ref{ch3:ex:annihilate}.
+Now that we have the formal semantics, we consider a generalized version of
+Ex.~\ref{ch3:ex:annihilate}.
 
@@
 
-Every relation is clearly reversible by simply swapping the input and
-output components of every pair, i.e., if {{r = {[ (a,b), ... ]} }}
-then the inverse of {{r}} is simply {{ {[ (b,a), ... ]} }}.
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrev-tracet}
-For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
-{{v2 : b2}} we have if {{v2 in (evalF c v1)}} then 
-{{v1 in (evalB c v2)}}.
-\end{proposition}
-
-An interesting point to ponder is what happens to \emph{information}
-in circuits like {{annihilate}}? We return to this point later.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{(Finite) Relational Programming}
+%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{(Finite) Relational Programming}
 \label{ch3:sec:lp}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Solving Constraints}
+%%%%%%%%%%%%%%%
+\subsection{Solving Constraints}
 \label{ch3:sec:constraints}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Idea}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
 
-Our development so far has been fairly traditional in the sense that
-we have applied well-known type constructions and semantic techniques
-to a language of isomorphisms. In this chapter, we begin to explore
-more novel ideas.
-
-One of the benefits of having a reversible language in which
-information is preserved is that essentially every construction can be
-considered in reverse. We begin in this chapter by considering the
-idea of \emph{fractional types} which are dual to product types. We
-call the resulting language {{langRevTF}}.
-
-We have already seen in Sec.~\ref{sec:trace} that the multiplicative
-trace can be viewed as performing some kind of ``division'' on
-types. Recall that given a combinator {{c : b * b1 <-> b * b2}}, we
-have {{trace* c : b1 <-> b2}} which \emph{cancels} the type {{b}}.
-Our plan is to make this ``division operation'' a first-class one. In
-other words, we introduce a new type construction --- a fractional
-type {{1/b}} --- that allows us to cancel any type {{b}} in a
-multiplicative context, i.e., that satisfies the isomorphism:
-
-%subcode{bnf} include main
-%  eps*_b :& 1/b * b ~~<->~~ 1 &: eta*_b
-
-Given this fractional type and its associated isomorphism, we can
-perform the cancellation implicit in {{trace*}} as follows. 
-Let {{ c :  b * b1 <-> b * b2 }} be given and consider the 
-following sequence of type isomorphisms:
-
-%subcode{bnf} include main
-%  b1 &<->& 1 * b1 & (uniti)
-%     &<->& (1/b * b) * b1 & (eta*_b)
-%     &<->& 1/b * (b * b1) & (assocr*)
-%     &<->& 1/b * (b * b2) & (c)
-%     &<->& ((1/b * b) * b2) & (assocl*)
-%     &<->& (1 * b2) & (eps*_b)
-%     &<->& b2 & (unite*)
-
-\noindent This sequence of isomorphisms, at least at the type level,
-seems to produce the desired {{trace*}}. We will confirm this in the
-next section after we introduce the formal syntax and semantics. For
-now, we provide an intuitive understanding of the type {{1/b}}.
-
-If {{v : b}}, we define the value {{1/v : 1/b}} and these are the only
-values of type~{{1/b}}. Since there are no values of type {{0}}, this
-means that {{1/0}} is also an empty type. Generally, we think of a
-value like {{1/false}} as a \emph{first-class constraint} that can
-only be satisfied if it is matched with an actual value {{false}}. So
-intuitively, {{trace*}} allowed us to implicitly introduce constraints
-in our model which we now make explicit and first-class entities. This
-makes the language much more expressive: as we show in this chapter,
-it essentially gains the expressive power to model higher-order
-relations.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Language: {{langRevTF}} }
-
-We extend {{langRevT}} with fractional types, fractional values, and
-two combinators {{eta*_b}} and {{eps*_b}} motivated above.  We provide the
-full syntactic and semantic definitions for the language below.
-
-\begin{definition}[Syntax of {{langRevT}}]
-\label{def:langRevT}
-The syntax of types, values, and combinators, is:
-%subcode{bnf} include main
-% natural numbers, n 
-% value types, b ::= 0 | 1 | b+b | b*b | 1/b | bool | nat 
-% values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
-%
-% combinator types, t ::= b <-> b
-% isomorphisms, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
-%                   &|& unite | uniti | swap* | assocl* | assocr* 
-%                   &|& dist0 | factor0 | dist | factor 
-%                   &|& foldB | unfoldB | foldN | unfoldN
-%                   &|& eta*_b | eps*_b
-% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-%                   &|& trace+ c | trace* c
-\end{definition}
-
-The denotation of types is extended with a clause for fractional types.
-
-\begin{definition}[Denotation of Types {{ [[b]] }}]
-\label{chx:def:denot}
-Each type denotes a set of values as follows:
-\[\begin{array}{rcl}
-\llbracket 0 \rrbracket &=& \emptyset \\
-\llbracket 1 \rrbracket &=& \{ () \} \\
-\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
-           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
-\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
-           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
-\llbracket 1/v \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} \\
-\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
-\llbracket \mathit{nat} \rrbracket &=& \{ 0, 1, 2, \ldots \} 
-\end{array}\]
-\end{definition}
-
-As usual, the semantics of {{langRevTF}} is given using two mutually
-recursive interpreters: one going forward and one going backwards. As
-motivated and explained in Sec.~\ref{sec:trace}, each combinator denotes
-a \emph{relation} and hence the result of applying a combinator to a
-value generally returns a \emph{set} of values. We only show the new
-cases for {{eta*_b}} and {{eps*_b}}.
-
-\begin{definition}[Operational semantics for {{langRevTF}}]
-\label{def:operational-langRevTF}
-Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
-forward direction by supplying it with a value {{ v1 : b1 }} and in
-the backwards direction by supplying it with a value {{v2 : b2}}.
-
-\paragraph*{Forward Evaluator (clauses for {{eta*_b}} and {{eps*_b}}): }
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
-% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
-% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2
-
-\paragraph*{Backwards Evaluator (clauses for {{eta*_b}} and {{eps*_b}}): }
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
-% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
-% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
-\end{definition}
-
-The language {{langRevT}} is still logically reversible.
-
-\begin{proposition}[Logical Reversibility]
-\label{chx:prop:logrev-tracep}
-For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
-{{v2 : b2}} we have {{v2 in (evalF c v1)}} iff
-{{v1 in (evalB c v2)}}.
-\end{proposition}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Expressiveness}
-
-Fractional types and values add considerable expressiveness to our
-language.
-
-%%%%%%%%%%%%
-\subsection{Trace}
-
-With fractionals, the trace operator becomes expressible:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
-% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
-%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite
-
-Let's use the semantics to calculate the result of applying
-{{trace*_{bool} (not (*) id)}} to {{false}}. We start
-with the set {{ {[ false ]} }} which gets transformed as follows:
-%subcode{opsem} include main
-%! columnStyle = ll
-% {[ ((),false) ]}  & (uniti)
-% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool})
-% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
-% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (not (*) id)
-% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
-% emptyset & eps*_{bool}
-
-And indeed this is the annhiliation circuit from
-Ex.~\ref{ch3:ex;annihilation}.
-
-%%%%%%%%%%%%
-\subsection{First-Class Relations}
-
-With fractional types, we can turn define a value that denotes a
-first-class function or relation. Indeed a value of type {{1/b1 * b2}}
-is a pair of a \emph{constraint} that can only be satisfied by some
-{{v1 : b1}} and a value {{v2 : b2}}. In other words, it corresponds to
-a function or relation which when given matched with a value {{v1 : b1}} 
-``releases'' the value {{v2 : b2}}. In the remainder of the book, we 
-introduce this abbreviation:
-
-%subcode{bnf} include main
-%  b1 :-* b2 ::= 1/b1 * b2
-
-What is remarkable is that we can systematically turn any combinator
-{{c : b1 <-> b2}} into a (constant) value of type {{b1 :-* b2}} as
-shown below:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
-% name c &=& eta*_{b1} (;) (id (*) c)
-
-As a simple example, consider the combinator 
-{{name not : 1 <-> (bool :-* bool)}}. Applying this 
-combinator to {{ () }} evaluates as follows:
-
-%subcode{opsem} include main
-%! columnStyle = ll
-% {[ () ]}  & 
-% {[ (1/false,false), (1/true,true) ]} & eta*_{bool}
-% {[ (1/false,true), (1/true,false) ]} & id (*) not
-
-\noindent In other words, the result of applying the combinator to {{ () }}
-is a value that represents the boolean negation function.
-
-%%%%%%%%%%%%
-\subsection{Higher-Order Relations}
-
-We are now a small step to implementing various higher-order
-combinators that manipulate functions or relations. In particular, we
-can \emph{apply} and \emph{compose} values representing relations as
-shown below:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% apply &:& (b1 :-* b2) * b1 <-> b2
-% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
-
-\noindent At the type level, the computation proceeds as follows:
-%subcode{bnf} include main
-%  (b1 :-* b2) * b1 &=& (1/b1 * b2) * b1 & 
-%     &<->& b1 * (1/b1 * b2) & (swap*)
-%     &<->& (b1 * 1/b1) * b2) & (assocl*)
-%     &<->& (1/b1 * b1) * b2) & (swap* (*) id)
-%     &<->& 1 * b2 & (eps*{b1} (*) id)
-%     &<->& b2 & (unite)
-
-\noindent Intuitively, we simply match the incoming argument of type {{b1}} with
-the constraint encoded by the function. If they match, they cancel
-each other and the value of type {{b2}} is exposed with no
-constraints. Otherwise, the mismatched values are \emph{annihilated}.
-
-Function or relation composition is only slightly more involved:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
-% compose &=& assocr* (;) (id (*) (assocl* (;) (swap* (*) id) (;) (eps*_{b2} (*) id) (;) unite))
-
-\noindent Again, at the type level, the computation proceeds as follows:
-
-%subcode{bnf} include main
-%  (b1 :-* b2) * (b2 :-* b3) &=& (1/b1 * b2) * (1/b2 * b3) & 
-%     &<->& (1/b1 * (b2 * (1/b2 * b3)) & (assocr*)
-%     &<->& (1/b1 * ((b2 * 1/b2) * b3) & (assocl*)
-%     &<->& (1/b1 * ((1/b2 * b2) * b3) & (assocl*)
-%     &<->& (1/b1 * (1 * b3) & (eps*_{b2})
-%     &<->& (1/b1 * b3) & (unite)
-%     &=& b1 :-* b3
-
-%%%%%%%%%%%%
-\subsection{Duality}
-
-The final remarkable property of fractional types is that they
-constitute, a form of self-inverse negation. In particular, we have:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% doubleNeg &:& b <-> 1/(1/b)
-% doubleNeg &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
-
-\noindent At the type level, this computation proceeds as follows:
-
-%subcode{bnf} include main
-%  b &<->& (1 * b) & (uniti)
-%    &<->& (1/(1/b) * 1/b) * b & (eta*_{1/b} (*) id)
-%    &<->& 1/(1/b) * (1/b * b) & (assocr*)
-%    &<->& 1/(1/b) * 1 & (id (*) eps*_b)
-%    &<->& 1 * 1/(1/b) & (swap*)
-%    &<->& 1/(1/b) & (unite)
-
-What is interesting about this isomorphism is that it holds for all
-types {{b}}, and in particular we have {{0 <-> 1/(1/0)}}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2013-02-02 frac.tex
intro

@@
   information-preserving, and reversible programming language {{langRev}}
-  founded on type isomorphisms. Being first-order limits the expressiveness
-  of the language: it is not possible, for example, to abstract program
-  fragments into a higher-level combinator. More importantly, however, the
-  limitation to first-order types means that the class of isomorphisms
-  between two types {{b1}} and {{b2}} cannot be itself represented as a type
-  and hence that the natural notion of equivalence between two {{langRev}}
-  computations is no longer based on isomorphisms. In this paper, we
-  introduce a higher-order extension of {{langRev}} based on the novel
-  concept of \emph{fractional types}. Intuitively, a value of a fractional
-  type represents \emph{negative} information: such fractional values can be
-  freely propagated and transformed but must ultimately --- in a complete
-  program --- be offset by the corresponding amount of positive information.
+  founded on type isomorphisms. Being restricted to first-order types limits
+  the expressiveness of the language: it is not possible, for example, to
+  abstract common program fragments into a higher-level combinator. In this
+  paper, we introduce a higher-order extension of {{langRev}} based on the
+  novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
+  fractional type {{1/v}} represents \emph{negative} information. A
+  higher-order function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
+  representing the needed argument and {{v2}} representing the
+  result. Fractional values can be freely propagated and transformed but must
+  ultimately --- in a complete program --- be offset by the corresponding
+  amount of positive information.
 \end{abstract}
@@
 \section{Introduction}
+
+We are witnessing a convergence of ideas from several distinct research
+communities (physics, mathematics, and computer science) towards replacing
+\emph{equalities} by \emph{isomorphisms}. The combined programme originating
+from these research areas has sparked a huge amount of research that unveiled
+new and surprising connections between geometry, algebra, logic, and
+computation.
+
+In the physics community, Landauer, Feynman, and others viewed the laws of
+physics as fundamentally related to computation. In more detail, the great
+majority of the laws of physics are formulated as equalities between
+different physical observables. For instance, Newton's second law of
+classical mechanics equates the force acting on a system to its rate of
+change of momentum. The insight of Landauer, Feynman, etc. is that
+\emph{different} physical observables should not be related by an
+\emph{equality} but more appropriately by an \emph{isomorphism} that
+witnesses, explains, and simulates the process of transforming one observable
+to the other.
+
+In the mathematics and logic community, Martin-L\"of developed an extension
+of the simply typed $\lambda$-calculus originally intended to provide a
+rigorous framework for constructive mathematics. It is now natural to extend
+this type theory with a new \emph{identity type} representing the proposition
+that two terms are ``equal.'' Specifically, given two terms $a$ and $b$ of
+the same type $A$, one forms the type $\texttt{Id}_A(a,b)$ representing the
+proposition that $a$ and $b$ are equal: in other words, a term of type
+$\texttt{Id}_A(a,b)$ witnesses, explains, and simulates the process of
+transforming $a$ to $b$ and vice-versa.
+
+In the computer science community, the theory and practice of type
+isomorphisms is well-established. Originally, such type isomorphisms were
+motivated the pragmatic concern of searching large libraries of functions by
+providing one of the many possible isomorphic types for the desired
+function. More recently, type isomorphisms have taken a more central role as
+\emph{the} fundamental computational mechanism from which more conventional,
+i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
+in our own previous work, we started with the notion of type isomorphism as
+the ``right'' starting point for a theory of computation. The central
+technical contribution was the development of a family of programming
+languages, {{langRev}} with various superscripts, in which every computation
+is an isomorphism. Other major results are that all such computations
+preserve the information-theoretic entropy, that the language is logically
+reversible, and that it has a model in symmetric monoidal categories (which
+are the backbone of models for linear logic and quantum computation). We have
+also invested a significant amount of time in using {{langRev}} as an actual
+programming language and established several idioms for programming large
+classes of interesting (recursive) programs in {{langRev}}.
+
+A major open problem remains, however. This concerns a higher-order extension
+of {{langRev}}. This extension is of fundamental importance in all the
+originating research areas. In physics, it allows the process itself or the
+observer to be treated as ``data'' that can be transformed by higher-level
+processes or observed by meta-level observers. In mathematics and logic, it
+allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
+to be itself expressed as an isomorphism as a higher type
+$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
+higher-order types allow code to abstract over other code fragments as well
+as the manipulation of code as data and data as code which is central to many
+algorithms and application domains.
+
+We address this higher-order extension in our paper using a simple novel type
+construction, that of a fractional type. The plan of the paper is therefore
+as follows: \todo{roadmap}
 

### 2013-02-03 frac.tex
sec 2

@@
 
-\todo{brief explanation}
+We review our language {{langRev}} providing the necessary background and
+context for our higher-order extension.\footnote{The presentation in this
+  section focuses on the simplest version of {{langRev}}. Other versions
+  include the empty type, recursive types, and trace opeators but these
+  extensions are orthogonal to the higher-order extension emphasized in this
+  paper.} The terms of {{Pi}} are not classical values and functions; rather,
+the terms are isomorphism witnesses.  In other words, the terms of {{Pi}} are
+proofs that certain ``shapes of values'' are isomorphic.  And, in classical
+Curry-Howard fashion, our operational semantics shows how these proofs can be
+directly interpreted as actions on ordinary values which effect this shape
+transformation. Of course, ``shapes of values'' are very familiar already:
+they are usually called \emph{types}.  But usually one designs a type system
+as a method of classifying terms, with the eventual purpose to show that
+certain properties of well-typed terms hold, such as safety.  Our approach is
+different: we start from a type system, and then present a term language
+which naturally inhabits these types, along with an appropriate operational
+semantics.
 
-\begin{definition}[Primitive {{langRev}} Combinators]
-\label{def:primitives-langRev}
+\paragraph*{Data.}
+We view {{Pi}} as having two levels:  it has traditional values, given by
+%subcode{bnf} include main
+% values, v ::= () | left v | right v | (v, v)
+and these are classified by ordinary types
+%subcode{bnf} include main
+% value types, b ::= 1 | b + b | b * b
 
+Types include the unit type {{1}}, sum types {{b1+b2}}, and products types
+{{b1*b2}}.  Values includes {{()}} which is the only value of type {{1}},
+{{left v}} and {{right v}} which inject {{v}} into a sum type, and
+{{(v1,v2)}} which builds a value of product type. But these values should be
+regarded as largely ancilliary.  In particular we do not treat the above
+values as first-class citizens.  They only occur when we want to observe the
+effect of an isomorphism.
+
+\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
+isomorphisms.  They have (iso) types {{b <-> b}}.  
+Specifically, they are witnesses to the following type isomorphisms:
 %subcode{bnf} include main
-%! columnStyle = rrcll
-%  zeroe :&  0 + b &<->& b &: zeroi
-%  swap+ :&  b1 + b2 &<->& b2 + b1 &: swap+
-%  assocl+ :&  b1 + (b2 + b3) &<->& (b1 + b2) + b3 &: assocr+
-%  unite :&  1 * b &<->& b &: uniti
-%  swap* :&  b1 * b2 &<->& b2 * b1 &: swap*
-%  assocl* :&  b1 * (b2 * b3) &<->& (b1 * b2) * b3 &: assocr*
-%  dist0 :&  0 *b &<->& 0  &: factor0
-%  dist :& (b1 + b2) * b3 &<->& (b1 * b3) + (b2 * b3) &: factor
+%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
+%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b <-> b &: identr*
+%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
+%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+Each line of the above table introduces a pair of dual constants%
+\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
+isomorphism in the middle.  These are the base (non-reducible) terms of the
+second, principal level of {{Pi}}.
 
-\end{definition}
+Note how the above has two readings: first as a set of typing relations for a
+set of constants. If these axioms are seen as universally quantified,
+orientable statements, they also induce transformations of the (traditional)
+values.  The (categorical) intuition here is that these axioms have
+computational content because they witness isomorphisms rather than merely
+stating an extensional equality.
 
-Each line of this table is to be read as the definition of one or two
-combinators. For example, corresponding to the \textit{identity of
-  multiplication} ({{1*b<->b}}) we have the two combinators
-{{unite:1*b<->b}} (reading the isomorphism from left to right) and
-{{uniti:b<->1*b}} (reading the isomorphism from right to left). These
-combinators are inverses of each other.  Each of the two cases of
-commutativity defines one combinator that is its own inverse.
+For example {{ swap* }} induces a function (akin to):
+%subcode{code}
+%! language = Haskell
+% swapProd (a,b) = (b,a)
 
-Now that we have primitive combinators we need some means of composing
-them. We construct the composition combinators out of the congruence
-closure of basic isomorphisms. Thus we have program constructs that
-witness reflexivity {{id}}, symmetry {{sym}}, transitivity {{(;)}},
-and two parallel composition combinators, one for sums {{(+)}} and one
-for pairs {{(*)}}.
-
-\begin{definition}[Composition in {{langRev}}]
+The isomorphisms are extended to form a congruence relation by adding the
+following constructors that witness equivalence and compatible closure:
 %subcode{proof} include main
@@
 %@@ c1(;)c2 : b1 <-> b3
-%----
+%---
 %@ c1 : b1 <-> b3
@@
 %@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+
+The syntax is overloaded: we use the same symbol at the value-type level
+and at the isomorphism-type level for denoting sums and products.  Hopefully
+this will not cause undue confusion.
+
+It is important to note that ``values'' and ``isomorphisms'' are 
+completely separate syntactic categories which do not intermix.  The 
+semantics of the language come when these are made to interact at
+the ``top level'', a new syntactic category:
+%subcode{bnf} include main
+% top level term, l ::= c v
+We refer to this as \emph{application}.
+
+\noindent
+To summarize, the syntax of {{Pi}} is given as follows.
+
+\begin{definition}{(Syntax of {{Pi}})}
+\label{def:Pi}
+We collect our types, values, and isomorphisms, to get the full language
+definition.
+%subcode{bnf} include main
+% value types, b ::= 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% iso.~types, t ::= b <-> b
+% base iso ::= swap+ | assocl+ | assocr+ 
+%     &|& identl* | identr* 
+%     &|& swap* | assocl* | assocr* 
+%     &|& dist | factor 
+% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+% top level term, l ::= c v
 \end{definition}
@@
 switches control from one evaluator to the other. We will not present the
-semantics because we generalize in the next section.
-
-\begin{proposition}[Termination] Evaluation of well-typed combinators
-  always terminates. In other words, for {{c : b1 <-> b2}} and 
-  {{v1 : b1}} there exists {{v2 : b2}} such that {{evalF c v1 = v2}}. 
-  Conversely, if {{v2 : b2}} there exists {{v1 : b1}} such that 
-  {{evalB c v2 = v1}}.
-\end{proposition}
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrev}
-For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
-{{v2 : b2}} we have: {{evalF c v1 = v2}} iff {{evalB c v2 = v1}}.
-\end{proposition}
-
-\todo{mention that we have extended this with trace and recursive types but
-  that we are going to explore something orthogonal}
+semantics because we present a more general version in the next section. We
+state without proof that the evaluation of well-typed combinators always
+terminates and that {{langRev}} is logically reversible, i.e., that for all
+combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have
+the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
+evaluation of {{c v2}} produces {{v1}}.
 
@@
 % natural numbers, n 
-% value types, b ::= 0 | 1 | b+b | b*b | 1/b | bool | nat 
+% value types, b ::= 1 | b+b | b*b | 1/b | bool 
 % values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
@@
 % combinator types, t ::= b <-> b
-% isomorphisms, iso ::= zeroe | zeroi | swap+ | assocl+ | assocr+ 
+% isomorphisms, iso ::= swap+ | assocl+ | assocr+ 
 %                   &|& unite | uniti | swap* | assocl* | assocr* 
-%                   &|& dist0 | factor0 | dist | factor 
-%                   &|& foldB | unfoldB | foldN | unfoldN
+%                   &|& dist | factor 
+%                   &|& foldB | unfoldB 
 %                   &|& eta*_b | eps*_b
 % combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-%                   &|& trace+ c | trace* c
 \end{definition}
@@
 \[\begin{array}{rcl}
-\llbracket 0 \rrbracket &=& \emptyset \\
 \llbracket 1 \rrbracket &=& \{ () \} \\
+\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
 \llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
@@
            v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
-\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} \\
-\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
-\llbracket \mathit{nat} \rrbracket &=& \{ 0, 1, 2, \ldots \} 
+\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
 \end{array}\]
@@
 %! columnStyle = rcl
-% evalF zeroe (right v) &=& {[ v ]}
-% evalF zeroi v &=& {[ right v ]}
 % evalF swap+ (left v) &=& {[ right v ]}
@@
 % evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
-% evalF (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalF c (v,v2)) ]}
 % evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
@@
 % evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2
-
 \paragraph*{Backwards Evaluator:}
@@
 %! columnStyle = rcl
-% evalB zeroe v &=& {[ right v ]}
-% evalB zeroi (right v) &=& {[ v ]}
 % evalB swap+ (left v) &=& {[ right v ]}
@@
 % evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
-% evalB (trace*_{b} c) v2 &=& Union_{v in [[b]]} ~{[ v3 ~|~ (v,v3) in (evalB c (v,v2)) ]}
 % evalB ~eta*_b (1/v,v) &=& {[ () ]} 

### 2013-02-04 frac.tex
small change to sec3

@@
   section focuses on the simplest version of {{langRev}}. Other versions
-  include the empty type, recursive types, and trace opeators but these
+  include the empty type, recursive types, and trace operators but these
   extensions are orthogonal to the higher-order extension emphasized in this
@@
 {{(v1,v2)}} which builds a value of product type. But these values should be
-regarded as largely ancilliary.  In particular we do not treat the above
+regarded as largely ancillary.  In particular we do not treat the above
 values as first-class citizens.  They only occur when we want to observe the
@@
 stating an extensional equality.
-
-For example {{ swap* }} induces a function (akin to):
-%subcode{code}
-%! language = Haskell
-% swapProd (a,b) = (b,a)
 
@@
 % base iso ::= swap+ | assocl+ | assocr+ 
-%     &|& identl* | identr* 
-%     &|& swap* | assocl* | assocr* 
+%     &|& unite | uniti | swap* | assocl* | assocr* 
 %     &|& dist | factor 
@@
 switches control from one evaluator to the other. We will not present the
-semantics because we present a more general version in the next section. We
-state without proof that the evaluation of well-typed combinators always
-terminates and that {{langRev}} is logically reversible, i.e., that for all
-combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have
-the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
+semantics until the next section in the context of the extended language.
+For now, we state without proof that the evaluation of well-typed combinators
+always terminates and that {{langRev}} is logically reversible, i.e., that
+for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}}
+we have the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
 evaluation of {{c v2}} produces {{v1}}.
@@
 
-As hinted to in the first section, the semantics of fractional types introduces
-a form of non-determinism in the sense that there could zero, one, or several
-values that satisfy the constraints imposed by fractional values. To model
-such non-determinism, we specify the semantics using \emph{relations}. In
-other words, the results of {{evalF c v}} or {{evalB c v}} are no longer
-guaranteed to be a single value: they could be \emph{sets} of values.
+Syntactically, the extension with fractional types is simple.
 
-The introduction of relations makes sequential composition of
-combinators more involved than simply passing the value produced by
-the first to the second. The following example illustrates our
-notation for the composition of relations which is used in the formal
-semantics.
-
-\begin{example}[Composition of Relations]
-Consider the two relations:
-\[\begin{array}{rcl}                                                                       
-R_1 &=& \{ (1,2),(1,3),(2,2) \} \\                                                       
-R_2 &=& \{ (2,1),(2,3),(3,1) \}                                                          
-\end{array}\]
-These relations can be expressed in a functional notation as follows:
-\[\begin{array}{rcl}                                                                       
-R_1 (1) &=& \{ 2,3 \} \\                                                                 
-R_1 (2) &=& \{ 2 \} \\                                                                   
-R_1 (3) &=& \emptyset \\
-\\ 
-R_2 (1) &=& \emptyset \\                                                                 
-R_2 (2) &=& \{ 1,3 \} \\                                                                 
-R_2 (3) &=& \{ 1 \}                                                                      
-\end{array}\]
-Using the functional notation, the composition $(R_2 \circ R_1)$ is defined as
-follows:
-\[\begin{array}{rcl}                                                                   
-(R_2 \circ R_1)~(a) &=& \bigcup_{b \in R_1(a)} R_2(b)                                  
-\end{array}\]
-We calculate the composition:
-\[\begin{array}[t]{rcl}                                                                  
-(R_2 \circ R_1)~(1) &=& \bigcup_{b \in R_1(1)} R_2(b) \\                               
-         &=& R_2(2) \cup R_2(3) \\                                                     
-         &=& \{1,3\} \cup \{ 1 \} \\                                                   
-         &=& \{ 1,3 \} \\                                                              
-(R_2 \circ R_1)~(2) &=& \{ 1,3 \} \\                                                   
-(R_2 \circ R_1)~(3) &=& \emptyset                                                      
-\end{array}\]
-In other words, 
-\[
-(R_2 \circ R_1) = \{(1,1), (1,3), (2,1), (2,3)\}
-\]
-\end{example}
-
-We extend {{langRevT}} with fractional types, fractional values, and two
-combinators {{eta*_b}} and {{eps*_b}}.  We provide the full syntactic and
-semantic definitions for the language below.
-
-\begin{definition}[Syntax of {{langRevT}}]
+\begin{definition}[Syntax of {{langRevTF}}]
 \label{def:langRevT}
@@
 %subcode{bnf} include main
-% natural numbers, n 
-% value types, b ::= 1 | b+b | b*b | 1/b | bool 
-% values, v ::= () | left v | right v | (v,v) | 1/v | true | false | n
+% value types, b ::= 1 | b+b | b*b | 1/b 
+% values, v ::= () | left v | right v | (v,v) | 1/v 
 %
-% combinator types, t ::= b <-> b
-% isomorphisms, iso ::= swap+ | assocl+ | assocr+ 
+% iso.~types, t ::= b <-> b
+% base iso ::= swap+ | assocl+ | assocr+ 
 %                   &|& unite | uniti | swap* | assocl* | assocr* 
 %                   &|& dist | factor 
-%                   &|& foldB | unfoldB 
 %                   &|& eta*_b | eps*_b
@@
 
-If {{v : b}}, we define the value {{1/v : 1/b}} and these are the only
-values of type~{{1/b}}. Since there are no values of type {{0}}, this
-means that {{1/0}} is also an empty type. Generally, we think of a
-value like {{1/false}} as a \emph{first-class constraint} that can
-only be satisfied if it is matched with an actual value {{false}}.
+For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
+where {{v : b}}. We think of a value like {{1/left ()}} as a
+\emph{first-class constraint} that can only be satisfied if it is matched
+with an actual value {{left ()}} or as representing the absence of some
+information (i.e., \emph{negative information}), that can only be supplied by
+{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
+point which creates --- out of nothing --- an equal amount of negative and
+positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}}
+consumes an equal amount of negative and positive information producing 
+no residual information.
+
+give some intuition; example with eta/epsilon
+
+motivate the need for relations
+
+present formal semantics
 
@@
 in circuits like {{annihilate}}? We return to this point later.
+
+As hinted to in the first section, the semantics of fractional types introduces
+a form of non-determinism in the sense that there could zero, one, or several
+values that satisfy the constraints imposed by fractional values. To model
+such non-determinism, we specify the semantics using \emph{relations}. In
+other words, the results of {{evalF c v}} or {{evalB c v}} are no longer
+guaranteed to be a single value: they could be \emph{sets} of values.
+
+The introduction of relations makes sequential composition of
+combinators more involved than simply passing the value produced by
+the first to the second. The following example illustrates our
+notation for the composition of relations which is used in the formal
+semantics.
+
+\begin{example}[Composition of Relations]
+Consider the two relations:
+\[\begin{array}{rcl}                                                                       
+R_1 &=& \{ (1,2),(1,3),(2,2) \} \\                                                       
+R_2 &=& \{ (2,1),(2,3),(3,1) \}                                                          
+\end{array}\]
+These relations can be expressed in a functional notation as follows:
+\[\begin{array}{rcl}                                                                       
+R_1 (1) &=& \{ 2,3 \} \\                                                                 
+R_1 (2) &=& \{ 2 \} \\                                                                   
+R_1 (3) &=& \emptyset \\
+\\ 
+R_2 (1) &=& \emptyset \\                                                                 
+R_2 (2) &=& \{ 1,3 \} \\                                                                 
+R_2 (3) &=& \{ 1 \}                                                                      
+\end{array}\]
+Using the functional notation, the composition $(R_2 \circ R_1)$ is defined as
+follows:
+\[\begin{array}{rcl}                                                                   
+(R_2 \circ R_1)~(a) &=& \bigcup_{b \in R_1(a)} R_2(b)                                  
+\end{array}\]
+We calculate the composition:
+\[\begin{array}[t]{rcl}                                                                  
+(R_2 \circ R_1)~(1) &=& \bigcup_{b \in R_1(1)} R_2(b) \\                               
+         &=& R_2(2) \cup R_2(3) \\                                                     
+         &=& \{1,3\} \cup \{ 1 \} \\                                                   
+         &=& \{ 1,3 \} \\                                                              
+(R_2 \circ R_1)~(2) &=& \{ 1,3 \} \\                                                   
+(R_2 \circ R_1)~(3) &=& \emptyset                                                      
+\end{array}\]
+In other words, 
+\[
+(R_2 \circ R_1) = \{(1,1), (1,3), (2,1), (2,3)\}
+\]
+\end{example}
 

### 2013-02-04 frac.tex
sec3

@@
 isomorphisms is well-established. Originally, such type isomorphisms were
-motivated the pragmatic concern of searching large libraries of functions by
-providing one of the many possible isomorphic types for the desired
+motivated by the pragmatic concern of searching large libraries of functions
+by providing one of the many possible isomorphic types for the desired
 function. More recently, type isomorphisms have taken a more central role as
@@
   extensions are orthogonal to the higher-order extension emphasized in this
-  paper.} The terms of {{Pi}} are not classical values and functions; rather,
-the terms are isomorphism witnesses.  In other words, the terms of {{Pi}} are
-proofs that certain ``shapes of values'' are isomorphic.  And, in classical
-Curry-Howard fashion, our operational semantics shows how these proofs can be
-directly interpreted as actions on ordinary values which effect this shape
-transformation. Of course, ``shapes of values'' are very familiar already:
-they are usually called \emph{types}.  But usually one designs a type system
-as a method of classifying terms, with the eventual purpose to show that
-certain properties of well-typed terms hold, such as safety.  Our approach is
-different: we start from a type system, and then present a term language
-which naturally inhabits these types, along with an appropriate operational
-semantics.
+  paper.} The terms of {{langRev}} are not classical values and functions;
+rather, the terms are isomorphism witnesses.  In other words, the terms of
+{{langRev}} are proofs that certain ``shapes of values'' are isomorphic.
+And, in classical Curry-Howard fashion, our operational semantics shows how
+these proofs can be directly interpreted as actions on ordinary values which
+effect this shape transformation. Of course, ``shapes of values'' are very
+familiar already: they are usually called \emph{types}.  But usually one
+designs a type system as a method of classifying terms, with the eventual
+purpose to show that certain properties of well-typed terms hold, such as
+safety.  Our approach is different: we start from a type system, and then
+present a term language which naturally inhabits these types, along with an
+appropriate operational semantics.
 
 \paragraph*{Data.}
-We view {{Pi}} as having two levels:  it has traditional values, given by
+We view {{langRev}} as having two levels:  it has traditional values, given by
 %subcode{bnf} include main
@@
 
-\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
+\paragraph*{Isomorphisms.} The terms of {{langRev}} are witnesses to type
 isomorphisms.  They have (iso) types {{b <-> b}}.  
@@
 isomorphism in the middle.  These are the base (non-reducible) terms of the
-second, principal level of {{Pi}}.
+second, principal level of {{langRev}}.
 
@@
 \noindent
-To summarize, the syntax of {{Pi}} is given as follows.
+To summarize, the syntax of {{langRev}} is given as follows.
 
-\begin{definition}{(Syntax of {{Pi}})}
+\begin{definition}{(Syntax of {{langRev}})}
 \label{def:Pi}
@@
 
-The semantics of {{langRev}} is given using two mutually recursive
-interpreters: one going forward and one going backwards. The use of {{sym}}
-switches control from one evaluator to the other. We will not present the
-semantics until the next section in the context of the extended language.
-For now, we state without proof that the evaluation of well-typed combinators
-always terminates and that {{langRev}} is logically reversible, i.e., that
-for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}}
-we have the forward evaluation of {{c v1}} produces {{v2}} iff the backwards
-evaluation of {{c v2}} produces {{v1}}.
+Mathematically, {{langRev}} can be interpreted in the category of finite sets
+and bijections: each type denotes a finite set of the corresponding size and
+each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
+denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{langRev}} is
+given using two mutually recursive interpreters: one going forward and one
+going backwards. The use of {{sym}} switches control from one evaluator to
+the other. We will not present the operational semantics until the next
+section in the context of the extended language. For now, we state without
+proof that the evaluation of well-typed combinators always terminates and
+that {{langRev}} is logically reversible, i.e., that for all combinators 
+{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
+evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
+{{c v2}} produces {{v1}}.
 
@@
 \label{def:langRevT}
-The syntax of types, values, and combinators, is:
+We extend Def.~\ref{def:Pi} as follows:
 %subcode{bnf} include main
-% value types, b ::= 1 | b+b | b*b | 1/b 
-% values, v ::= () | left v | right v | (v,v) | 1/v 
+% value types, b ::= ... | 1/b
+% values, v ::= ... | 1/v
 %
-% iso.~types, t ::= b <-> b
-% base iso ::= swap+ | assocl+ | assocr+ 
-%                   &|& unite | uniti | swap* | assocl* | assocr* 
-%                   &|& dist | factor 
-%                   &|& eta*_b | eps*_b
-% combinators, c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+% base iso ::= ... | eta*_b | eps*_b
 \end{definition}
@@
 positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}}
-consumes an equal amount of negative and positive information producing 
+matches up an equal amount of negative and positive information producing 
 no residual information.
+
+%%%%%%%%%%%%%%%%%%%%%5
+\subsection{Relational Semantics}
+
+Even though the categorical semantics of {{langRevTF}} is not the main focus
+of this paper, we find it helpful to briefly outline an interpretation of
+{{langRevTF}} in \emph{dagger compact closed categories}. These categories
+were introduced by Abramsky and
+Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and then by
+Selinger~\cite{Selinger:2007:DCC:1229185.1229207}. The simplest example of
+such categories is the category of finite sets and relations on which we
+focus.
+
+We recall that a \emph{symmetric monoidal category} is a category together
+with a bifunctor $\otimes$, a distinguished object $I$, and natural
+isomorphisms $\alpha_{A,B,C} : (A \otimes B) \otimes C \rightarrow A \otimes
+(B \otimes C)$, $\lambda_A : A \rightarrow I \otimes A$, and $\sigma_{A,B} :
+A \otimes B \rightarrow B \otimes A$ subject to standard coherence
+conditions~\cite{nla.cat-vn1051288}. Following common practice
+(e.g.,~\cite{Selinger:2007:DCC:1229185.1229207}), we write $\rho_A =
+\sigma_{I,A} \circ \lambda_A : A \rightarrow A \otimes I$.
+
+A \emph{compact closed category} is a symmetric monoidal category where each
+object $A$ is assigned a dual object $A^*$, together with a unit map $\eta_A
+: I \rightarrow A^* \otimes A$ and a counit map $\epsilon_A : A \otimes A^*
+\rightarrow I$, such that:
+\[\begin{array}{rcl}
+\lambda_A^{-1} \circ (\epsilon_A \otimes A) \circ \alpha_{A,A^*,A}^{-1} \circ (A \otimes \eta_A) \circ \rho_A &=& \mathit{id}_A \\
+\rho_{A^*}^{-1} \circ (A^* \otimes \epsilon_A) \circ \alpha_{A^*,A,A^*} \circ (\eta_A \otimes A^*) \circ \lambda_A &=& \mathit{id}_{A^*}
+\end{array}\]
+
+the category of finite sets and relations.
 
@@
 present formal semantics
+
+model using vector spaces; category of finite sets and relations; homotopy
+equivalence???
 

### 2013-02-04 frac.tex
fix some stuff up, and add some comments here and there

@@
 \newcommand{\todo}[1]{\textbf{TODO:} #1}
+\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
+
+\hyphenation{a-reas}
 
@@
 new and surprising connections between geometry, algebra, logic, and
-computation.
-
-In the physics community, Landauer, Feynman, and others viewed the laws of
+computation (see \cite{baez2011physics} for an overview).
+
+In the physics community, Landauer~\cite{Landauer:1961}, Feynman, and 
+others have interpreted the laws of
 physics as fundamentally related to computation. In more detail, the great
@@
 \emph{equality} but more appropriately by an \emph{isomorphism} that
-witnesses, explains, and simulates the process of transforming one observable
+witnesses, explains, and models the process of transforming one observable
 to the other.
@@
 of the simply typed $\lambda$-calculus originally intended to provide a
-rigorous framework for constructive mathematics. It is now natural to extend
-this type theory with a new \emph{identity type} representing the proposition
+rigorous framework for constructive mathematics. This has been further
+extended (first by Streicher and Awodey, and subsequently by many others)
+with a new \emph{identity type} representing the proposition
 that two terms are ``equal.'' Specifically, given two terms $a$ and $b$ of
@@
 proposition that $a$ and $b$ are equal: in other words, a term of type
-$\texttt{Id}_A(a,b)$ witnesses, explains, and simulates the process of
+$\texttt{Id}_A(a,b)$ witnesses, explains, and models the process of
 transforming $a$ to $b$ and vice-versa.
@@
 
-We address this higher-order extension in our paper using a simple novel type
-construction, that of a fractional type. The plan of the paper is therefore
+This paper addresses this higher-order extension.  It does this via a novel
+type construction, which we dub a \emph{fractional type} because of its
+interaction with product types.  By adding a formal (linear) dual to
+(type theoretic) products, we can internalize the notion that ``taking an
+input'' corresponds to negative information, which turns out to be the 
+basis of first-class functions.
+
+The plan of the paper is therefore
 as follows: \todo{roadmap}
@@
 Note how the above has two readings: first as a set of typing relations for a
-set of constants. If these axioms are seen as universally quantified,
+set of constants. Second, if these axioms are seen as universally quantified,
 orientable statements, they also induce transformations of the (traditional)
@@
 \label{def:Pi}
-We collect our types, values, and isomorphisms, to get the full language
-definition.
 %subcode{bnf} include main
@@
 Mathematically, {{langRev}} can be interpreted in the category of finite sets
-and bijections: each type denotes a finite set of the corresponding size and
-each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
+and bijections: each value type denotes a finite set of the corresponding size
+and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
 denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{langRev}} is
@@
 the other. We will not present the operational semantics until the next
-section in the context of the extended language. For now, we state without
+section, in the context of the extended language. For now, we state without
 proof that the evaluation of well-typed combinators always terminates and
@@
 
+\jacques{Since we have Agda code for the above, should we refer to it
+here, and make it available?}
+
+Note that a small modification of {{lanRev}} (the addition of an
+empty type {{0}} with its usual laws) would exactly capture, at the type
+level, the notion of \emph{semiring} (occasionally called a \emph{rig}) where
+we replace equality by isomorphism.  This can also be also compared to
+\emph{bimonoidal categories}.  As is usual with with type theories, the
+coherence conditions (pentagonal and triangle identity) are not explicit,
+but would correspond to certain identity types being trivial.  {{lanRev}}
+instead, at the type leve, models a commutative ringoid where the
+multiplicative structure forms a commutative monoid, but the additive structure
+is just a commutative magma.  We chose not to include {{0}} as a technical
+convenience for the next section.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+{{langRev}} models isomorphisms of values rather well.  However, purposefully,
+it has no notion of \emph{input} or \emph{output}; more precisely, because
+of strict preservation of information, these two concepts coincide.  This
+is the root of reversibility.
+
+But if we want to model functions of any flavor, we need to differentiate
+between these notions.  Our idea, inspired by computational dualities and
+polarity, is to introduce a \emph{formal dual} for our types.  We thus
+consider a type in a negative position in a type to possess negative
+information, as it is really a request for information.  Since information
+is logarithmic, this means that a request for information should behave
+like a \emph{fractional type}.
+
 Syntactically, the extension with fractional types is simple.
@@
 \label{def:langRevT}
-We extend Def.~\ref{def:Pi} as follows:
+We extend Def.~\ref{def:Pi} with:
 %subcode{bnf} include main
@@
 For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
-where {{v : b}}. We think of a value like {{1/left ()}} as a
+where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.
+However, we think of a value like {{1/left ()}} as a
 \emph{first-class constraint} that can only be satisfied if it is matched
@@
 {{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
-point which creates --- out of nothing --- an equal amount of negative and
-positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}}
-matches up an equal amount of negative and positive information producing 
-no residual information.
+point which creates --- out of no information --- an equal amount of negative
+and positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <->
+1}} matches up an equal amount of negative and positive information producing 
+no residual information.  The best analogy is probably that of vectors and
+covectors, also know as column vectors and row vectors (respectively): in
+the finite dimensional case, they are isomorphic, but they are nevertheless
+quite different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In
+other words, since we are explicitly interested in \emph{isomorphisms}, 
+we should not conflate negative-information values with positive-information
+values just because they happen to be isomorphic.
 

### 2013-02-05 frac.tex
more additions
typos

@@
   representing the needed argument and {{v2}} representing the
-  result. Fractional values can be freely propagated and transformed but must
-  ultimately --- in a complete program --- be offset by the corresponding
-  amount of positive information.
+  result. Fractional values are first-class: they can be freely propagated
+  and transformed but must ultimately --- in a complete program --- be offset
+  by the corresponding amount of positive information.
 \end{abstract}
@@
 new and surprising connections between geometry, algebra, logic, and
-computation (see \cite{baez2011physics} for an overview).
-
-In the physics community, Landauer~\cite{Landauer:1961}, Feynman, and 
-others have interpreted the laws of
-physics as fundamentally related to computation. In more detail, the great
-majority of the laws of physics are formulated as equalities between
-different physical observables. For instance, Newton's second law of
-classical mechanics equates the force acting on a system to its rate of
-change of momentum. The insight of Landauer, Feynman, etc. is that
-\emph{different} physical observables should not be related by an
-\emph{equality} but more appropriately by an \emph{isomorphism} that
-witnesses, explains, and models the process of transforming one observable
-to the other.
+computation (see~\cite{baez2011physics} for an overview).
+
+In the physics community, Landauer~\cite{Landauer:1961}, Feynman, and others
+have interpreted the laws of physics as fundamentally related to
+computation. In more detail, the great majority of the laws of physics are
+formulated as equalities between different physical observables. For
+instance, Newton's second law of classical mechanics equates the force acting
+on a system to its rate of change of momentum. The insight of Landauer,
+Feynman, etc. is that \emph{different} physical observables should not be
+related by an \emph{equality} but more appropriately by an \emph{isomorphism}
+that witnesses, explains, and models the process of transforming one
+observable to the other.
 
@@
 
-Note that a small modification of {{lanRev}} (the addition of an
+Note that a small modification of {{langRev}} (the addition of an
 empty type {{0}} with its usual laws) would exactly capture, at the type
@@
 coherence conditions (pentagonal and triangle identity) are not explicit,
-but would correspond to certain identity types being trivial.  {{lanRev}}
-instead, at the type leve, models a commutative ringoid where the
+but would correspond to certain identity types being trivial.  {{langRev}}
+instead, at the type level, models a commutative ringoid where the
 multiplicative structure forms a commutative monoid, but the additive structure
@@
 point which creates --- out of no information --- an equal amount of negative
-and positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <->
-1}} matches up an equal amount of negative and positive information producing 
-no residual information.  The best analogy is probably that of vectors and
-covectors, also know as column vectors and row vectors (respectively): in
+and positive information. Symmetrically, the combinator 
+{{eps*_b : 1/b * b <-> 1}} matches up an equal amount of negative and positive
+information producing no residual information.  The best analogy is probably
+that of vectors and
+co-vectors, also know as column vectors and row vectors (respectively): in
 the finite dimensional case, they are isomorphic, but they are nevertheless
@@
 
-\begin{definition}[Denotation of Types {{ [[b]] }}]
+\begin{definition}[Denotation of Value Types {{ [[b]] }}]
 \label{chx:def:denot}
@@
 
-The semantics of {{langRevTF}} is given using two mutually recursive
-interpreters: one going forward and one going backwards. Unlike the case for
-{{langRev}}, each combinator denotes a \emph{relation} and hence the result
-of applying a combinator to a value generally returns a \emph{set} of
+The operational semantics of {{langRevTF}} is given using two mutually
+recursive interpreters: one going forward and one going backwards. Unlike the
+case for {{langRev}}, each combinator denotes a \emph{relation} and hence the
+result of applying a combinator to a value generally returns a \emph{set} of
 values.
@@
 
-The language {{langRevT}} is still logically reversible.
+The language {{langRevTF}} is still logically reversible.
 

### 2013-02-05 frac.tex
cat background

@@
 of this paper, we find it helpful to briefly outline an interpretation of
-{{langRevTF}} in \emph{dagger compact closed categories}. These categories
-were introduced by Abramsky and
+{{langRevTF}} in \emph{biproduct dagger compact closed categories}. These
+categories were introduced by Abramsky and
 Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and then by
-Selinger~\cite{Selinger:2007:DCC:1229185.1229207}. The simplest example of
-such categories is the category of finite sets and relations on which we
-focus.
-
-We recall that a \emph{symmetric monoidal category} is a category together
-with a bifunctor $\otimes$, a distinguished object $I$, and natural
-isomorphisms $\alpha_{A,B,C} : (A \otimes B) \otimes C \rightarrow A \otimes
-(B \otimes C)$, $\lambda_A : A \rightarrow I \otimes A$, and $\sigma_{A,B} :
-A \otimes B \rightarrow B \otimes A$ subject to standard coherence
-conditions~\cite{nla.cat-vn1051288}. Following common practice
-(e.g.,~\cite{Selinger:2007:DCC:1229185.1229207}), we write $\rho_A =
-\sigma_{I,A} \circ \lambda_A : A \rightarrow A \otimes I$.
-
-A \emph{compact closed category} is a symmetric monoidal category where each
-object $A$ is assigned a dual object $A^*$, together with a unit map $\eta_A
-: I \rightarrow A^* \otimes A$ and a counit map $\epsilon_A : A \otimes A^*
-\rightarrow I$, such that:
-\[\begin{array}{rcl}
-\lambda_A^{-1} \circ (\epsilon_A \otimes A) \circ \alpha_{A,A^*,A}^{-1} \circ (A \otimes \eta_A) \circ \rho_A &=& \mathit{id}_A \\
-\rho_{A^*}^{-1} \circ (A^* \otimes \epsilon_A) \circ \alpha_{A^*,A,A^*} \circ (\eta_A \otimes A^*) \circ \lambda_A &=& \mathit{id}_{A^*}
-\end{array}\]
-
-the category of finite sets and relations.
+Selinger~\cite{Selinger:2007:DCC:1229185.1229207}. The bare definitions are
+collected in Appendix~\ref{app:cat}.
+
+The simplest example of such categories is the category of finite sets and
+relations on which we focus. To interpret {{langRevTF}} in this category, we
+associate a set with each type. For {{1}}, we associate a set with one
+element and we interpret {{*}} as the cartesian product and {{+}} as the
+disjoint union. For {{1/b}}, we associate the same set associated with {{b}}
+and hence, in our denotations, we identify {{b}} and {{1/b}} as well as {{v}}
+and {{1/v}}. It is intuitively clear that this category is compact closed as
+any relation between $A$ and $B$ can be represented as a finite set of type
+{{A * B}}. For example, consider the ``boolean negation'' relation:
+
+{{ {[ (1/left(), right()), (1/right (), left ()) ]} }}
+
+This is clearly isomorphic to the corresponding set of pairs. The map
+{{eta_b}} constructs the identity relation at type {{b}}. For example, {{eta_{1+1} }}
+is the relation:
+
+{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
 
@@
 
+\appendix
+\section{Categorical Background}
+\label{app:cat}
+
+We recall that a \emph{symmetric monoidal category} is a category together
+with a bifunctor $\otimes$, a distinguished object $I$, and natural
+isomorphisms $\alpha_{A,B,C} : (A \otimes B) \otimes C \rightarrow A \otimes
+(B \otimes C)$, $\lambda_A : A \rightarrow I \otimes A$, and $\sigma_{A,B} :
+A \otimes B \rightarrow B \otimes A$ subject to standard coherence
+conditions~\cite{nla.cat-vn1051288}. Following common practice
+(e.g.,~\cite{Selinger:2007:DCC:1229185.1229207}), we write $\rho_A =
+\sigma_{I,A} \circ \lambda_A : A \rightarrow A \otimes I$.
+
+A \emph{compact closed category} is a symmetric monoidal category where each
+object $A$ is assigned a dual object $A^*$, together with a unit map $\eta_A
+: I \rightarrow A^* \otimes A$ and a counit map $\epsilon_A : A \otimes A^*
+\rightarrow I$, such that:
+\[\begin{array}{rcl}
+\lambda_A^{-1} \circ (\epsilon_A \otimes A) \circ \alpha_{A,A^*,A}^{-1} \circ (A \otimes \eta_A) \circ \rho_A &=& \mathit{id}_A \\
+\rho_{A^*}^{-1} \circ (A^* \otimes \epsilon_A) \circ \alpha_{A^*,A,A^*} \circ (\eta_A \otimes A^*) \circ \lambda_A &=& \mathit{id}_{A^*}
+\end{array}\]
+
+A \emph{dagger category} is a category together with an involutive,
+identity-on-objects, contravariant functor $\dagger$. Concretely, this means
+that to every morphism $F : A \rightarrow B$, one associates a morphism
+$f^{\dagger} : B \rightarrow A$, called the \emph{adjoint} of $f$, such that
+for all $f : A \rightarrow B$ and $g : B \rightarrow C$, we have:
+\[\begin{array}{rcl}
+\mathit{id}^\dagger_A &=& \mathit{id}_A \\
+(g \circ f)^\dagger &=& f^\dagger \circ g^\dagger \\
+(f^dagger)^\dagger &=& f
+\end{array}\]
+
+A \emph{dagger symmetric monoidal category} is a symmetric monoidal category
+with a dagger structure such that the contravariant functor $\dagger$
+coherently preserves the symmetric monoidal structure. Concretely, this
+requirement means that for all $f : A \rightarrow B$ and $g : C \rightarrow
+D$, we have:
+\[\begin{array}{rcl}
+(f \otimes g)^\dagger &=& f^\dagger \otimes g^\dagger \\
+\alpha^\dagger_{A,B,C} &=& \alpha^{-1}_{A,B,C} \\
+\lambda^\dagger_A &=& \lambda^{-1}_A \\
+\sigma^\dagger_{A,B} &=& \sigma^{-1}{A,B}
+\end{array}\]
+
+\begin{definition}[Dagger Compact Closed Category]
+\label{def:cat}
+A \emph{dagger compact closed category} is a dagger symmetric monoidal
+category that is also compact closed and such that for all $A$:
+\[
+\eta_A = \sigma_{A,A^*} \circ \epsilon^\dagger_A
+\]
+\end{definition}
+
+\todo{biproducts}
+
 \bibliographystyle{splncs03} 

### 2013-02-05 frac.tex
a bit on sec3

@@
 of this paper, we find it helpful to briefly outline an interpretation of
-{{langRevTF}} in \emph{biproduct dagger compact closed categories}. These
-categories were introduced by Abramsky and
-Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and then by
-Selinger~\cite{Selinger:2007:DCC:1229185.1229207}. The bare definitions are
-collected in Appendix~\ref{app:cat}.
-
-The simplest example of such categories is the category of finite sets and
-relations on which we focus. To interpret {{langRevTF}} in this category, we
-associate a set with each type. For {{1}}, we associate a set with one
-element and we interpret {{*}} as the cartesian product and {{+}} as the
-disjoint union. For {{1/b}}, we associate the same set associated with {{b}}
-and hence, in our denotations, we identify {{b}} and {{1/b}} as well as {{v}}
-and {{1/v}}. It is intuitively clear that this category is compact closed as
-any relation between $A$ and $B$ can be represented as a finite set of type
-{{A * B}}. For example, consider the ``boolean negation'' relation:
-
-{{ {[ (1/left(), right()), (1/right (), left ()) ]} }}
-
-This is clearly isomorphic to the corresponding set of pairs. The map
-{{eta_b}} constructs the identity relation at type {{b}}. For example, {{eta_{1+1} }}
-is the relation:
-
-{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
-
-give some intuition; example with eta/epsilon
-
-motivate the need for relations
-
-present formal semantics
-
-model using vector spaces; category of finite sets and relations; homotopy
-equivalence???
+{{langRevTF}} in the category of finite sets and relations. This category is
+an example of a \emph{biproduct dagger compact closed category} which are
+categories introduced by Abramsky and
+Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
+Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
+computation.\footnote{The other common example of these categories is that of
+  finite dimensional Hilbert spaces.} The category intuitively supports
+higher-order relations as any relation between $A$ and $B$ can be represented
+as a subset of {{A * B}}, i.e., as a value. More generally, the two main
+features of these biproduct dagger compact closed categories are that they
+support an adjoint for each morphism (to model reversibility) and a notion of
+higher-order relation (built from the dagger which is modeled by our
+fractional types).
+
+To interpret {{langRevTF}} in this category, we associate a finite set
+$\llbracket b \rrbracket$ with each type $b$ and a relation $\llbracket c
+\rrbracket$ with each combinator $c$. The denotation of types is
+straighforward: the denotation of {{1/b}} is an isomorphic copy of the
+denotation of {{b}}.
 
@@
 \label{chx:def:denot}
-Each type denotes a set of values as follows:
+Each type denotes a finite set of values as follows:
 \[\begin{array}{rcl}
 \llbracket 1 \rrbracket &=& \{ () \} \\
-\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
-\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
-           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
-\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
-           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
+\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \} \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
+\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
 \llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
@@
 \end{definition}
+
+Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
+the relation:
+
+{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
+
+Perhaps get rid of the semantics below and replace by a semantics which maps
+each combinator to a relation. This is probably clearer than the operational
+style.
+
+give some intuition; example with eta/epsilon
+
+motivate the need for relations
+
+present formal semantics
+
+model using vector spaces; category of finite sets and relations; homotopy
+equivalence???
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\bibliographystyle{splncs03} 
+\bibliography{cites}
+\end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 (g \circ f)^\dagger &=& f^\dagger \circ g^\dagger \\
-(f^dagger)^\dagger &=& f
+(f^\dagger)^\dagger &=& f
 \end{array}\]
@@
 \todo{biproducts}
-
-\bibliographystyle{splncs03} 
-\bibliography{cites}
-\end{document}

### 2013-02-06 frac.tex
added relational semantics for the interpreter

@@
 
+%subcode{proof} include main
+%@ ~
+%@@ (left v) swap+ (right v)
+%
+%@ ~
+%@@ (right v) swap+ (left v)
+%----
+%@ ~
+%@@ (left v) assocl+ (left (left v))
+%
+%@ ~
+%@@ (right (left v)) assocl+ (left (right v))
+%
+%@ ~
+%@@ (right (right v)) assocl+ (right v)
+%----
+%@ ~
+%@@ (left (left v)) assocr+ (left v)
+%
+%@ ~
+%@@ (left (right v)) assocr+ (right (left v))
+%
+%@ ~
+%@@ (right v) assocr+ (right (right v))
+%----
+%@ ~
+%@@ ((), v) unite v
+%
+%@ ~
+%@@ v uniti ((), v)
+%----
+%@ ~
+%@@ (v1, v2) swap* (v2, v1)
+%----
+%@ ~
+%@@ (v1, (v2, v3)) assocl* ((v1, v2), v3)
+%
+%@ ~
+%@@ ((v1, v2), v3) assocr* (v1, (v2, v3))
+%----
+%@ ~
+%@@ (left v1, v3) dist (left (v1, v3))
+%
+%@ ~
+%@@ (right v2, v3) dist (right (v2, v3))
+%----
+%@ ~
+%@@ (left (v1, v3)) factor (left v1, v3)
+%
+%@ ~
+%@@ (right (v2, v3)) factor (right v2, v3)
+%----
+%@ ~
+%@@ v id v
+%
+%@ v' c v
+%@@ v (sym c) v'
+%----
+%@ v1 c1 v2
+%@ v2 c2 v3
+%@@ v1 (c1(;)c2) v3
+%
+%@ v c1 v'
+%@@ (left v) (c1 (+) c2) (left v')
+%
+%@ v c2 v'
+%@@ (right v) (c1 (+) c2) (right v')
+%----
+%@ v1 c1 v1'
+%@ v2 c2 v2'
+%@@ (v1, v2) (c1 (*) c2) (v1', v2')
+%----
+%@ v in [[b]]
+%@@ () ~eta*_b (1/v,v)
+%
+%@ ~
+%@@ (1/v,v) ~eps*_b ()
+
 %subcode{opsem} include main
@@
 As hinted to in the first section, the semantics of fractional types introduces
-a form of non-determinism in the sense that there could zero, one, or several
+a form of non-determinism in the sense that there could be zero, one, or several
 values that satisfy the constraints imposed by fractional values. To model

### 2013-02-07 frac.tex
relational semantics confirmed opeerational semantics deleted

--- 2013-02-05 frac.tex
@@
 %! v] = \lrcorner
+%| [[ = \llbracket
+%| ]] = \rrbracket
 %! eta* = \eta^{\times}
@@
 %! langRevT = \Pi^{o}
-%! langRevTF = \Pi^{o/}
+%! langRevTF = \Pi^{/}
 
@@
 
-{{langRev}} models isomorphisms of values rather well.  However, purposefully,
-it has no notion of \emph{input} or \emph{output}; more precisely, because
-of strict preservation of information, these two concepts coincide.  This
-is the root of reversibility.
+The language {{langRev}} models isomorphisms of values rather well.  However,
+purposefully, it has no notion of \emph{input} or \emph{output}; more
+precisely, because of strict preservation of information, these two concepts
+coincide.  This is the root of reversibility.
 
@@
 polarity, is to introduce a \emph{formal dual} for our types.  We thus
-consider a type in a negative position in a type to possess negative
-information, as it is really a request for information.  Since information
-is logarithmic, this means that a request for information should behave
-like a \emph{fractional type}.
+consider a type in a negative position to possess negative information, as it
+is really a request for information.  Since information is logarithmic, this
+means that a request for information should behave like a \emph{fractional
+  type}.
 
@@
 For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
-where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.
-However, we think of a value like {{1/left ()}} as a
-\emph{first-class constraint} that can only be satisfied if it is matched
-with an actual value {{left ()}} or as representing the absence of some
-information (i.e., \emph{negative information}), that can only be supplied by
-{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
-point which creates --- out of no information --- an equal amount of negative
-and positive information. Symmetrically, the combinator 
-{{eps*_b : 1/b * b <-> 1}} matches up an equal amount of negative and positive
-information producing no residual information.  The best analogy is probably
-that of vectors and
-co-vectors, also know as column vectors and row vectors (respectively): in
-the finite dimensional case, they are isomorphic, but they are nevertheless
-quite different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In
-other words, since we are explicitly interested in \emph{isomorphisms}, 
-we should not conflate negative-information values with positive-information
-values just because they happen to be isomorphic.
+where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.  However,
+we think of a value like {{1/left ()}} as a \emph{first-class constraint}
+that can only be satisfied if it is matched with an actual value {{left ()}}
+or as representing the absence of some information (i.e., \emph{negative
+  information}), that can only be supplied by {{left ()}}. The combinator
+{{eta*_b : 1 <-> 1/b * b}} represents a fission point which creates --- out
+of no information --- an equal amount of negative and positive
+information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
+up an equal amount of negative and positive information producing no residual
+information.  The best analogy is probably that of vectors and co-vectors,
+also know as column vectors and row vectors (respectively): in the finite
+dimensional case, they are isomorphic, but they are nevertheless quite
+different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
+since we are explicitly interested in \emph{isomorphisms}, we should not
+conflate negative-information values with positive-information values just
+because they happen to be isomorphic.
 
@@
 \subsection{Relational Semantics}
+
+motivate the need for relations
 
@@
 computation.\footnote{The other common example of these categories is that of
-  finite dimensional Hilbert spaces.} The category intuitively supports
-higher-order relations as any relation between $A$ and $B$ can be represented
-as a subset of {{A * B}}, i.e., as a value. More generally, the two main
-features of these biproduct dagger compact closed categories are that they
-support an adjoint for each morphism (to model reversibility) and a notion of
-higher-order relation (built from the dagger which is modeled by our
-fractional types).
+  finite dimensional Hilbert spaces.} The two main features of these
+biproduct dagger compact closed categories are that they support an adjoint
+for each morphism (to model reversibility) and that they support higher-order
+relation (built from the categorical dagger which is modeled by our
+fractional types). The latter is intuitively clear as any relation between
+$A$ and $B$ can be represented as a subset of {{A * B}}, i.e., as a
+value.
 
 To interpret {{langRevTF}} in this category, we associate a finite set
-$\llbracket b \rrbracket$ with each type $b$ and a relation $\llbracket c
-\rrbracket$ with each combinator $c$. The denotation of types is
-straighforward: the denotation of {{1/b}} is an isomorphic copy of the
-denotation of {{b}}.
+{{ [[b]] }} with each type $b$ and a relation {{ [[c]] }} over
+{{ [[b1]] }} and {{ [[b2]] }} for each combinator {{ c : b1 <-> b2 }}.
+The denotation of types is straighforward: the denotation of {{1/b}} 
+is an isomorphic copy of the denotation of {{b}}.
 
@@
 
+The denotation of combinators as relations is given below. We specify
+relations using a deductive system whose judgments are of the form 
+{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
+denoted by {{c}}.
+
+\begin{definition}[Relational semantics]
+\label{def:relational-langRevTF}
+Each combinator {{c : b1 <-> b2}} in {{langRevTF}} denotes a relation
+as specified below.
+
+%subcode{proof} include main
+%@ ~
+%@@ (left v) ~~swap+~~ (right v)
+%
+%@ ~
+%@@ (right v) ~~swap+~~ (left v)
+%----
+%@ ~
+%@@ (left v) ~~assocl+~~ (left (left v))
+%
+%@ ~
+%@@ (right (left v)) ~~assocl+~~ (left (right v))
+%
+%@ ~
+%@@ (right (right v)) ~~assocl+~~ (right v)
+%----
+%@ ~
+%@@ (left (left v)) ~~assocr+~~ (left v)
+%
+%@ ~
+%@@ (left (right v)) ~~assocr+~~ (right (left v))
+%
+%@ ~
+%@@ (right v) ~~assocr+~~ (right (right v))
+%----
+%@ ~
+%@@ ((), v) ~~unite~~ v
+%
+%@ ~
+%@@ v ~~uniti~~ ((), v)
+%----
+%@ ~
+%@@ (v1, v2) ~~swap*~~ (v2, v1)
+%----
+%@ ~
+%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
+%
+%@ ~
+%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
+%----
+%@ ~
+%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
+%
+%@ ~
+%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
+%----
+%@ ~
+%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
+%
+%@ ~
+%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
+%----
+%@ ~
+%@@ v ~~id~~ v
+%
+%@ v' ~~c~~ v
+%@@ v ~~(sym c)~~ v'
+%----
+%@ v1 ~~c1~~ v2
+%@ v2 ~~c2~~ v3
+%@@ v1 ~~(c1(;)c2)~~ v3
+%
+%@ v ~~c1~~ v'
+%@@ (left v) ~~(c1 (+) c2)~~ (left v')
+%
+%@ v ~~c2~~ v'
+%@@ (right v) ~~(c1 (+) c2)~~ (right v')
+%----
+%@ v1 ~~c1~~ v1'
+%@ v2 ~~c2~~ v2'
+%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
+%----
+%@ v in [[b]]
+%@@ () ~~eta*_b~~ (1/v,v)
+%
+%@ ~
+%@@ (1/v,v) ~~eps*_b~~ ()
+
+\newpage
+
 Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
@@
 
-Perhaps get rid of the semantics below and replace by a semantics which maps
-each combinator to a relation. This is probably clearer than the operational
-style.
-
 give some intuition; example with eta/epsilon
 
-motivate the need for relations
-
-present formal semantics
-
-model using vector spaces; category of finite sets and relations; homotopy
-equivalence???
-
-The operational semantics of {{langRevTF}} is given using two mutually
-recursive interpreters: one going forward and one going backwards. Unlike the
-case for {{langRev}}, each combinator denotes a \emph{relation} and hence the
-result of applying a combinator to a value generally returns a \emph{set} of
-values.
-
-\begin{definition}[Operational semantics for {{langRevTF}}]
-\label{def:operational-langRevTF}
-Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
-forward direction by supplying it with a value {{ v1 : b1 }} and in
-the backwards direction by supplying it with a value {{v2 : b2}}.
-
-\paragraph*{Forward Evaluator:}
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalF swap+ (left v) &=& {[ right v ]}
-% evalF swap+ (right v) &=& {[ left v ]}
-% evalF assocl+ (left v) &=& {[ left (left v) ]}
-% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
-% evalF assocl+ (right (right v)) &=& {[ right v ]}
-% evalF assocr+ (left (left v)) &=& {[ left v ]}
-% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
-% evalF assocr+ (right v) &=& {[ right (right v) ]}
-% evalF unite ((), v) &=& {[ v ]}
-% evalF uniti v &=& {[ ((), v) ]}
-% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
-% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
-% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
-% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
-% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
-% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
-% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
-% evalF id v &=& {[ v ]}
-% evalF (sym c) v &=& evalB c v
-% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
-% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
-% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
-% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
-% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
-% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
-% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2
-\paragraph*{Backwards Evaluator:}
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalB swap+ (left v) &=& {[ right v ]}
-% evalB swap+ (right v) &=& {[ left v ]}
-% evalB assocl+ (left (left v)) &=& {[ left v ]}
-% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
-% evalB assocl+ (right v) &=& {[ right (right v) ]}
-% evalB assocr+ (left v) &=& {[ left (left v) ]}
-% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
-% evalB assocr+ (right (right v)) &=& {[ right v ]}
-% evalB unite v &=& {[ ((), v) ]}
-% evalB uniti ((), v) &=& {[ v ]}
-% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
-% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
-% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
-% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
-% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
-% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
-% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
-% evalB id v &=& {[ v ]}
-% evalB (sym c) v &=& evalF c v
-% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
-% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
-% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
-% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
-% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
-% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
-% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
-\end{definition}
+homotopy equivalence???
 
@@
 
-What is interesting about this isomorphism is that it holds for all
-types {{b}}, and in particular we have {{0 <-> 1/(1/0)}}.
-
 %%%%%%%%%%%%
@@
 combinator to produce no value. 
-\end{example}
-
-\begin{example}
-A more fundamental problem occurs if we attempt to apply the trace
-operator to cancel the empty type. Consider a combinator:
-
-{{c : 0 * 1 <-> 0 * bool}}
-
-{{c = distrib0 (;) factor0}}
-
-\noindent Using the multiplicative trace, we can construct the combinator
-{{trace* c : 1 <-> bool}} which relates two types of different sizes! Fortunately,
-because {{0}} is the empty type, 
-when attempting to evaluate such a combinator, it is impossible to find a value
-{{v : 0}} to satisfy the constraint imposed by {{trace*}} and hence 
-we would also expect the evaluation of such a combinator to produce no value.
 \end{example}

### 2013-02-07 frac.tex
fixed latex
sec3 more or less done

@@
 \section{Background: {{langRev}} }
+\label{sec:pi}
 
@@
 
-motivate the need for relations
-
-Even though the categorical semantics of {{langRevTF}} is not the main focus
-of this paper, we find it helpful to briefly outline an interpretation of
-{{langRevTF}} in the category of finite sets and relations. This category is
-an example of a \emph{biproduct dagger compact closed category} which are
-categories introduced by Abramsky and
-Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
-Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
+In {{langRev}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
+the simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
+richer languages, e.g., in a Turing complete language that admits
+non-termination, isomorphisms necessarily denote \emph{partial
+  bijections}~\cite{rc2011,infeffects}. As we demonstrate in this section,
+once we make the extension to higher-order functions, we must interpret the
+syntactic isomorphisms as \emph{relations}. The intuitive idea, is that the
+fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be able to map the
+element {{() : 1}} to any pair of the form {{(1/v,v)}} with
+{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
+accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
+not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
+undefined, i.e., it does not produce {{():1}}. In other words, some of the
+combinators are not partial and some produce more than one result for a
+particular input, and hence we move to the setting of general binary
+relations to interpret our combinators.
+
+This setting is not only consistent with logical reversibility but is
+arguably the \emph{right} framework for considering expressive reversible
+languages (including the interesting examples of quantum languages).  Indeed,
+even though the categorical semantics of {{langRevTF}} is not the main focus
+of this paper, we find it helpful to state that the category of finite sets
+and relations in which we are going to interpret our language, is an example
+of a \emph{biproduct dagger compact closed category} which are categories
+introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
+and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
 computation.\footnote{The other common example of these categories is that of
-  finite dimensional Hilbert spaces.} The two main features of these
-biproduct dagger compact closed categories are that they support an adjoint
-for each morphism (to model reversibility) and that they support higher-order
-relation (built from the categorical dagger which is modeled by our
-fractional types). The latter is intuitively clear as any relation between
-$A$ and $B$ can be represented as a subset of {{A * B}}, i.e., as a
-value.
+  finite dimensional Hilbert spaces.} Two of the most relevant features of
+these categories for our purposes are that they support an adjoint for each
+morphism (to model reversibility) and that they support higher-order relation
+(built from the categorical dagger which is modeled by our fractional
+types). The latter is intuitively clear as any relation between $A$ and $B$
+can be represented as a subset of {{A * B}}, i.e., as a value.
 
 To interpret {{langRevTF}} in this category, we associate a finite set
-{{ [[b]] }} with each type $b$ and a relation {{ [[c]] }} over
+{{ [[b]] }} with each type {{b}} and a relation {{ [[c]] }} over
 {{ [[b1]] }} and {{ [[b2]] }} for each combinator {{ c : b1 <-> b2 }}.
-The denotation of types is straighforward: the denotation of {{1/b}} 
-is an isomorphic copy of the denotation of {{b}}.
 
@@
 %@@ (1/v,v) ~~eps*_b~~ ()
-
-\newpage
-
-Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
-the relation:
-
-{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
-
-give some intuition; example with eta/epsilon
-
-homotopy equivalence???
-
-The language {{langRevTF}} is still logically reversible.
+\end{definition}
+
+The language {{langRevTF}} is still logically reversible. This is intuitively
+clear as the relation denoted by a combinator can be used in the reverse
+direction by swapping inputs and outputs.
 
@@
 For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
-{{v2 : b2}} we have {{v2 in (evalF c v1)}} iff
-{{v1 in (evalB c v2)}}.
+{{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
 \end{proposition}
-
-Every relation is clearly reversible by simply swapping the input and
-output components of every pair, i.e., if {{r = {[ (a,b), ... ]} }}
-then the inverse of {{r}} is simply {{ {[ (b,a), ... ]} }}.
-
-\begin{proposition}[Logical Reversibility]
-\label{prop:logrev-tracet}
-For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
-{{v2 : b2}} we have if {{v2 in (evalF c v1)}} then 
-{{v1 in (evalB c v2)}}.
-\end{proposition}
-
-An interesting point to ponder is what happens to \emph{information}
-in circuits like {{annihilate}}? We return to this point later.
-
-As hinted to in the first section, the semantics of fractional types introduces
-a form of non-determinism in the sense that there could be zero, one, or several
-values that satisfy the constraints imposed by fractional values. To model
-such non-determinism, we specify the semantics using \emph{relations}. In
-other words, the results of {{evalF c v}} or {{evalB c v}} are no longer
-guaranteed to be a single value: they could be \emph{sets} of values.
-
-The introduction of relations makes sequential composition of
-combinators more involved than simply passing the value produced by
-the first to the second. The following example illustrates our
-notation for the composition of relations which is used in the formal
-semantics.
-
-\begin{example}[Composition of Relations]
-Consider the two relations:
-\[\begin{array}{rcl}                                                                       
-R_1 &=& \{ (1,2),(1,3),(2,2) \} \\                                                       
-R_2 &=& \{ (2,1),(2,3),(3,1) \}                                                          
-\end{array}\]
-These relations can be expressed in a functional notation as follows:
-\[\begin{array}{rcl}                                                                       
-R_1 (1) &=& \{ 2,3 \} \\                                                                 
-R_1 (2) &=& \{ 2 \} \\                                                                   
-R_1 (3) &=& \emptyset \\
-\\ 
-R_2 (1) &=& \emptyset \\                                                                 
-R_2 (2) &=& \{ 1,3 \} \\                                                                 
-R_2 (3) &=& \{ 1 \}                                                                      
-\end{array}\]
-Using the functional notation, the composition $(R_2 \circ R_1)$ is defined as
-follows:
-\[\begin{array}{rcl}                                                                   
-(R_2 \circ R_1)~(a) &=& \bigcup_{b \in R_1(a)} R_2(b)                                  
-\end{array}\]
-We calculate the composition:
-\[\begin{array}[t]{rcl}                                                                  
-(R_2 \circ R_1)~(1) &=& \bigcup_{b \in R_1(1)} R_2(b) \\                               
-         &=& R_2(2) \cup R_2(3) \\                                                     
-         &=& \{1,3\} \cup \{ 1 \} \\                                                   
-         &=& \{ 1,3 \} \\                                                              
-(R_2 \circ R_1)~(2) &=& \{ 1,3 \} \\                                                   
-(R_2 \circ R_1)~(3) &=& \emptyset                                                      
-\end{array}\]
-In other words, 
-\[
-(R_2 \circ R_1) = \{(1,1), (1,3), (2,1), (2,3)\}
-\]
-\end{example}
 
@@
 \section{Expressiveness}
+
+Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
+the relation:
+
+{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
+
+give some intuition; example with eta/epsilon
+
+homotopy equivalence???
 

### 2013-02-07 frac.tex
fixed latex problems

@@
 %! trace* = \mathit{trace}^{\times}
-%! :-* = \multimap^{\times}
+%! :-* = \multimap
 %! :-+ = \multimap^{+}
@@
 %! v] = \lrcorner
-%| [[ = \llbracket
-%| ]] = \rrbracket
-%! eta* = \eta^{\times}
-%! eps* = \epsilon^{\times}
+%! [[ = \llbracket
+%! ]] = \rrbracket
+%! eta* = \eta
+%! eps* = \epsilon
 %! Union = \bigcup
@@
 
-%%%%%%%%%%%%%%%%%%%%%5
-\subsection{Relational Semantics}
-
 In {{langRev}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
@@
 To interpret {{langRevTF}} in this category, we associate a finite set
-{{ [[b]] }} with each type {{b}} and a relation {{ [[c]] }} over
-{{ [[b1]] }} and {{ [[b2]] }} for each combinator {{ c : b1 <-> b2 }}.
-
-\begin{definition}[Denotation of Value Types {{ [[b]] }}]
+{{ [[ b ]] }} with each type {{b}} and a relation {{ [[ c ]] }} over
+{{ [[ b1 ]] }} and {{ [[ b2 ]] }} for each combinator {{ c : b1 <-> b2 }}.
+
+\begin{definition}[Denotation of Value Types {{ [[ b ]] }}]
 \label{chx:def:denot}
@@
 %@@ (right (left v)) ~~assocl+~~ (left (right v))
-%
+%----
 %@ ~
 %@@ (right (right v)) ~~assocl+~~ (right v)
-%----
+%
 %@ ~
 %@@ (left (left v)) ~~assocr+~~ (left v)
-%
+%----
 %@ ~
@@
 %@@ v ~~(sym c)~~ v'
-%----
+%
 %@ v1 ~~c1~~ v2
@@
 %@@ v1 ~~(c1(;)c2)~~ v3
-%
+%----
 %@ v ~~c1~~ v'
@@
 %----
-%@ v in [[b]]
+%@ v in [[ b ]]  
 %@@ () ~~eta*_b~~ (1/v,v)

### 2013-02-07 frac.tex
got rid of old evalf evalb

@@
 % values, v ::= () | left v | right v | (v, v)
-and these are classified by ordinary types
+\noindent and these are classified by ordinary types
 %subcode{bnf} include main
@@
 %dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-Each line of the above table introduces a pair of dual constants%
+
+\noindent Each line of the above table introduces a pair of dual constants%
 \footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
@@
 
-The syntax is overloaded: we use the same symbol at the value-type level
+\noindent The syntax is overloaded: we use the same symbol at the value-type level
 and at the isomorphism-type level for denoting sums and products.  Hopefully
@@
 % top level term, l ::= c v
-We refer to this as \emph{application}.
+\noindent We refer to this as \emph{application}.
 
@@
 
-Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
-the relation:
-
-{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
-
-give some intuition; example with eta/epsilon
-
-homotopy equivalence???
-
 Fractional types and values add considerable expressiveness to our
-language.
+language. In the following, we use {{bool}} as an abbreviation of {{1+1}} 
+with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
+abbreviation for {{right ()}}.
 
@@
 
-With fractional types, we can define a value that denotes a first-class
-function or relation. Indeed a value of type {{1/b1 * b2}} is a pair of a
-\emph{constraint} that can only be satisfied by some {{v1 : b1}} and a value
-{{v2 : b2}}. In other words, it corresponds to a function or relation which
-when given matched with a value {{v1 : b1}} ``releases'' the value 
-{{v2 : b2}}. In the remainder of the paper, we introduce this abbreviation:
+As motivated, with fractional types, we can define a value that denotes a
+first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
+pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
+and a value {{v2 : b2}}. In other words, it corresponds to a function or
+relation which when given matched with a value {{v1 : b1}} ``releases'' the
+value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:
 
 %subcode{bnf} include main
-%  b1 :-* b2 ::= 1/b1 * b2
-
-What is remarkable is that we can systematically turn any combinator
-{{c : b1 <-> b2}} into a (constant) value of type {{b1 :-* b2}} as
-shown below:
+%  b1 :-* b2 &::=& 1/b1 * b2
+
+\noindent which suggests a function-like behavior for the pair of a fractional value
+and a regular value. What is remarkable is that we can almost trivially turn
+any combinator {{c : b1 <-> b2}} into a (constant) value of 
+type {{b1 :-* b2}} as shown below:
 
@@
 
-As a simple example, consider the combinator 
-{{name not : 1 <-> (bool :-* bool)}}. Applying this 
-combinator to {{ () }} evaluates as follows:
-
-%subcode{opsem} include main
-%! columnStyle = ll
-% {[ () ]}  & 
-% {[ (1/false,false), (1/true,true) ]} & eta*_{bool}
-% {[ (1/false,true), (1/true,false) ]} & id (*) not
-
-\noindent In other words, the result of applying the combinator to {{ () }}
-is a value that represents the boolean negation function.
+As a simple example, consider combinator {{not}} denoting boolean negation.
+It is straightforward to check that the combinator 
+{{name not : 1 <-> (bool :-* bool)}} denotes the following
+relation:
+
+%subcode{proof} include main
+%@ ~
+%@@ () ~~(name not)~~ (1/false,true)
+%
+%@ ~
+%@@ () ~~(name not)~~ (1/true,false)
+
+\noindent The relation {{name not}} can therefore be used to convert any
+{{false}} to {{true}} and vice-versa. We formalize this ``apply'' combinator
+below.
 
@@
 that can be defined, an additive family (explored in detail in our previous
-work) and a multiplicative family (which is the focus of this section).
+work~\cite{rc2011}) and a multiplicative family (which is the focus of this
+section).
 
@@
 constraint that the value {{v1 : b1}} must be such that it is also produced
-by as the first component of the result, i.e., 
-that {{evalF c (v1,v2) = (v1,v3)}}. In general, there may be several 
-such values fixed-point values or none. 
-A few examples should help understand the subtleties.
+by as the first component of the result. In general, there may be several
+such values fixed-point values or none.  A few examples should help
+understand the subtleties.
 
@@
 {{() : 1}} requires us to find a value {{b : bool}} such that 
-{{evalF c (b,()) = (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
+{{(b,()) ~~c~~ (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
 has two values, there are two values {{false}} and {{true}} that satisfy the 
-constraint. In other words, expect that {{evalF (trace* c) () = ()}}. 
+constraint. In other words, expect that {{() ~~(trace* c)~~() }}.
 \end{example}
@@
 combinator to {{() : 1}} requires us to find a value {{b : bool}} such
-that {{evalF c (b,()) = (b,())}} which is impossible since {{c}} negates
+that {{(b,()) ~~c~~ (b,())}} which is impossible since {{c}} negates
 {{b}}. Operationally, we would expect the evaluation of such a
@@
 \end{center}
-As motivated earlier, it is clearly impossible for the constraint
-implied by {{trace*}} to be satisfied as a boolean value {{v}} can
-never be the same as its negation. Thus there is no incoming value of
-type {{b}} that can ever go through the circuit. Indeed, if we call
-the circuit above {{c}}, then we can calculate that 
-{{ evalF c v = emptyset}}. We call such a circuit an 
-annihilation circuit.
+As motivated earlier, it is clearly impossible for the constraint implied by
+{{trace*}} to be satisfied as a boolean value {{v}} can never be the same as
+its negation. Thus there is no incoming value of type {{b}} that can ever go
+through the circuit. Indeed, if we call the circuit above {{c}}, then we can
+calculate that {{c}} does not relate {{v}} to anything. We call such a
+circuit an annihilation circuit.
 \end{example}
@@
 
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalF c_R false &=& {[ false, true ]} 
-% evalF c_R true &=& {[ false ]}
+%subcode{proof} include main
+%@ ~
+%@@ false ~~c_R~~ false
+%
+%@ ~
+%@@ false ~~c_R~~ true
+%
+%@ ~
+%@@ true ~~c_R~~ false
 
@@
 
-%subcode{opsem} include main
-%! columnStyle = rcl
-% evalF cInner ((false,false),false) &=& ((false,false),false) 
-% evalF cInner ((false,true),false) &=& ((false,true),true) 
-% evalF cInner ((false,true),true) &=& ((false,true),false) 
-% evalF cInner ((false,false),true) &=& ((true,false),true) 
-% evalF cInner ((true,false),false) &=& ((true,true),false) 
-% evalF cInner ((true,false),true) &=& ((true,true),true) 
-% evalF cInner ((true,true),false) &=& ((true,false),false) 
-% evalF cInner ((true,true),true) &=& ((false,false),true)
+%subcode{proof} include main
+%@ ~
+%@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
+%----
+%@ ~
+%@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
+%----
+%@ ~
+%@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
+%----
+%@ ~
+%@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
+%----
+%@ ~
+%@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
+%----
+%@ ~
+%@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
+%----
+%@ ~
+%@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
+%----
+%@ ~
+%@@ ((true,true),true) ~~cInner~~ ((false,false),true)
 
@@
 
+Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
+the relation:
+
+{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
+
+give some intuition; example with eta/epsilon
+
+homotopy equivalence???
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-10 frac.tex
added huge to do list

@@
 
+axioms for field or meadow; actually we probably a semifield
+
+definition of logical reversibility with relations
+
+definition of information preservation in the case of relations; fanout as an
+example which we can write
+
+explain in detail the size of 2 * 1/2; it should have exactly one element;
+there are two elemens in 2 but the 1/2 identifies them somehow; be precise
+
+we can get the empty relation (eta ; id * swap; eps) what does that mean?
+
+connection to vector spaces; projectors; inner product. Given dual vectors,
+we can have a standalone |0><0| which is a piece of a function or a projector
+if viewed by itself; we can also have |v><v'| which produces the scalars YES
+or NO. We have an isomorphism between matrices and vectors by viewing 
+|0><0|+|1><1| as |00>+|11>
+
+Perhaps if restrict the ``top-level'' to non-fractional types, we can never
+observe a function built from from fractionals; we must apply it and in that
+case, the system overall behaves like Pi. (no empty relation for example)???
+Idea suggested by Zach is to change eps to (1/a) * a * a <-> a
+
+The code implementing the 16 relations... all 16 relations have the same
+information content as () ??? So we should treat (1/bool * bool) as opaque
+things; we can only extract information from them by applying them.
+
+Include dneg, name (inv cnot) as example of programs which when fed
+non-fractional values produce fractional values
+
+Also include recipT which shows that 1/ distributes over products and tens
+which shows that we can manipulate fractional types in interesting ways
+
+Using relational semantics is good because it's clear; but it's also bad
+unless we spend time to explain the operational model
+
+Explain 'name' in detail; it IS a faithful representation of the input
+relation as a value; we don't get to see the full relation though; we
+interact with it by giving it inputs and observing the related ouptut come
+out. If we make two copies and use each in a different context (applying each
+to a different value, we can ``see'' that it's really a full relation and not
+just one branch.)
+
+total functions have size 1; partial functions have fractional sizes!!!
+
+denotation of type in Pi is a set; in Pi/ we get a set with an equivalence
+relation (Id by default) but fractionals introduce first-class equivalence
+relations that can be used to restrict the sets.
+
+---
+
+Having re-read Baez-Dolan, I now feel better.
+
+On 13-02-08 08:53 AM, Amr Sabry wrote:
+> I think all the observations in this email are due to the fact that we
+> haven't formalized the size of fractionals using homotopy equivalence. I
+> started doing this but abandoned it for now but the bottom line is that
+> the size of b * 1/b is 1.
+
+Again, agree - more deeply so now.  In more detail, my current thinking:
+- v \in b has size 1.  So |b| counts how many elements it has, whenever 
+b is a type of Pi.  The sum and product rules apply.
+- for b * 1/b to have size 1, then 1/b must have size 1/|b| (assuming 
+these are independent types, which we have been assuming all along)
+
+Attempt #1
+- there are |b| elements in 1/b, and if size is additive, ( 1/|b| = 
+sum_{1/v \in 1/b} |1/v| ) should hold; since |1/v| should be constant, 
+this sum is |b|*c, so c = 1/|b|^2.  Weird.
+
+Let us look at a single 'function' in bool -o bool, namely the identity 
+function.  It should be 'the same' as { (1/true, true), (1/false, false) 
+}.  The cardinality of that set is (1/(2^2)*1 + 1/(2^2)*1) = 1/4 + 1/4 = 
+1/2.   From weird to probably-wrong.
+
+Attempt #2
+- the elements of 1/b are not distinguishable from each other (i.e. they 
+are all isomorphic).  So each element 1/v has b automorphisms, so |1/b| 
+= sum_{aut classes of 1/v \in 1/b} 1/|aut 1/v| = sum_{singleton} 1/|b| = 
+1/|b|.  [as per p.14 of Baez-Dolan]
+
+The size of the set { (1/true, true), (1/false, false) } is now 1/2*1 + 
+1/2*1 = 1.  So this 'set' indeed represents a single "concept".
+
+Note that in attempt #2, we get a new phenomenon.  Consider the 
+(partial) relation {(1/true, false)}, also of type bool -o bool.  It has 
+size 1/2 !  So, if I have not erred somewhere, only total functions have 
+size 1.  Which, I must admit, I rather like.  It's not that partial 
+functions are outlawed, they just don't pull as much weight [pun intended].
+
+So 1/b is very much like a collection of |b| constraints, all of which 
+are "the same" (externally).  While they may have internal structure, it 
+is not visible to any of our combinators, so up to iso, there is only a 
+single element of 1/b, of size 1/|b|.
+
+Yet another way to look at it: let's assume we are looking at FinSet_0, 
+where in fact our universe U has finitely (say N) things in it.  Then 
+the singleton set 1 is actually the isomorphism class of {u_1, ..., 
+u_N}.  There are N of them, but they have N automorphisms in a single 
+orbit, so the quotient has size 1 (i.e. |S| = N but |G| = N too, so |S 
+// G| = 1 ).
+
+So an element of 1/b is closer to an action of b on b; we call it a 
+constraint, or a pattern-match.  I think thinking of it as 'actions' is 
+definitely the right way to go, as we will be able to have different 
+kinds of actions on a type b that just those that come from the elements.
+
+> I don't know how much of a priority this is but if it is we should try
+> to work it out in detail. The good and bad news is that we will have a
+> model much richer and much more complicated than relations. My thought
+> when writing sec. 3 was that we would use the simple but slightly
+> inadequate model of relations (which for one thing confuses v and 1/v)
+> and simply mention that the right abstraction is that of 'biproduct
+> dagger compact closed categories' which presumably would include the
+> richer model.
+
+I am as convinced that the right model is 'biproduct dagger compact 
+closed categories' as I am that the right model for the type level is 
+'categorified field'.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-11 frac.tex
_
relational semantics

@@
 
-%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
+%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
 %! [^ = \ulcorner
@@
 %! @ = \,@\,
-%! langRev = \Pi
-%! langRevT = \Pi^{o}
-%! langRevEE = \Pi^{\eta\epsilon}_{+}
+%! Pi = \Pi
+%! PiT = \Pi^{o}
+%! PiEE = \Pi^{\eta\epsilon}_{+}
 %! bullet = \bullet
 %! * = \times
-%! langRevT = \Pi^{o}
-%! langRevTF = \Pi^{/}
+%! PiT = \Pi^{o}
+%! PiTF = \Pi^{/}
 
@@
   In previous work, we developed a \emph{first-order},
-  information-preserving, and reversible programming language {{langRev}}
+  information-preserving, and reversible programming language {{Pi}}
   founded on type isomorphisms. Being restricted to first-order types limits
@@
   abstract common program fragments into a higher-level combinator. In this
-  paper, we introduce a higher-order extension of {{langRev}} based on the
+  paper, we introduce a higher-order extension of {{Pi}} based on the
   novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
@@
 technical contribution was the development of a family of programming
-languages, {{langRev}} with various superscripts, in which every computation
+languages, {{Pi}} with various superscripts, in which every computation
 is an isomorphism. Other major results are that all such computations
@@
 are the backbone of models for linear logic and quantum computation). We have
-also invested a significant amount of time in using {{langRev}} as an actual
+also invested a significant amount of time in using {{Pi}} as an actual
 programming language and established several idioms for programming large
-classes of interesting (recursive) programs in {{langRev}}.
+classes of interesting (recursive) programs in {{Pi}}.
 
 A major open problem remains, however. This concerns a higher-order extension
-of {{langRev}}. This extension is of fundamental importance in all the
+of {{Pi}}. This extension is of fundamental importance in all the
 originating research areas. In physics, it allows the process itself or the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background: {{langRev}} }
+\section{Background: {{Pi}} }
 \label{sec:pi}
 
-We review our language {{langRev}} providing the necessary background and
+We review our language {{Pi}} providing the necessary background and
 context for our higher-order extension.\footnote{The presentation in this
-  section focuses on the simplest version of {{langRev}}. Other versions
+  section focuses on the simplest version of {{Pi}}. Other versions
   include the empty type, recursive types, and trace operators but these
   extensions are orthogonal to the higher-order extension emphasized in this
-  paper.} The terms of {{langRev}} are not classical values and functions;
+  paper.} The terms of {{Pi}} are not classical values and functions;
 rather, the terms are isomorphism witnesses.  In other words, the terms of
-{{langRev}} are proofs that certain ``shapes of values'' are isomorphic.
+{{Pi}} are proofs that certain ``shapes of values'' are isomorphic.
 And, in classical Curry-Howard fashion, our operational semantics shows how
@@
 \paragraph*{Data.}
-We view {{langRev}} as having two levels:  it has traditional values, given by
+We view {{Pi}} as having two levels:  it has traditional values, given by
 %subcode{bnf} include main
@@
 
-\paragraph*{Isomorphisms.} The terms of {{langRev}} are witnesses to type
+\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
 isomorphisms.  They have (iso) types {{b <-> b}}.  
@@
 isomorphism in the middle.  These are the base (non-reducible) terms of the
-second, principal level of {{langRev}}.
+second, principal level of {{Pi}}.
 
@@
 \noindent
-To summarize, the syntax of {{langRev}} is given as follows.
-
-\begin{definition}{(Syntax of {{langRev}})}
+To summarize, the syntax of {{Pi}} is given as follows.
+
+\begin{definition}{(Syntax of {{Pi}})}
 \label{def:Pi}
@@
 
-Mathematically, {{langRev}} can be interpreted in the category of finite sets
+Mathematically, {{Pi}} can be interpreted in the category of finite sets
 and bijections: each value type denotes a finite set of the corresponding size
 and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
-denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{langRev}} is
+denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{Pi}} is
 given using two mutually recursive interpreters: one going forward and one
@@
 proof that the evaluation of well-typed combinators always terminates and
-that {{langRev}} is logically reversible, i.e., that for all combinators 
+that {{Pi}} is logically reversible, i.e., that for all combinators 
 {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
@@
 
-Note that a small modification of {{langRev}} (the addition of an
+Note that a small modification of {{Pi}} (the addition of an
 empty type {{0}} with its usual laws) would exactly capture, at the type
@@
 coherence conditions (pentagonal and triangle identity) are not explicit,
-but would correspond to certain identity types being trivial.  {{langRev}}
+but would correspond to certain identity types being trivial.  {{Pi}}
 instead, at the type level, models a commutative ringoid where the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Language: {{langRevTF}} }
-
-The language {{langRev}} models isomorphisms of values rather well.  However,
+\section{The Language: {{PiTF}} }
+
+The language {{Pi}} models isomorphisms of values rather well.  However,
 purposefully, it has no notion of \emph{input} or \emph{output}; more
@@
 
-\begin{definition}[Syntax of {{langRevTF}}]
-\label{def:langRevT}
+\begin{definition}[Syntax of {{PiTF}}]
+\label{def:PiT}
 We extend Def.~\ref{def:Pi} with:
@@
 
-In {{langRev}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
+In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
 the simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
@@
 languages (including the interesting examples of quantum languages).  Indeed,
-even though the categorical semantics of {{langRevTF}} is not the main focus
+even though the categorical semantics of {{PiTF}} is not the main focus
 of this paper, we find it helpful to state that the category of finite sets
@@
 
-To interpret {{langRevTF}} in this category, we associate a finite set
+To interpret {{PiTF}} in this category, we associate a finite set
 {{ [[ b ]] }} with each type {{b}} and a relation {{ [[ c ]] }} over
@@
 \begin{definition}[Relational semantics]
-\label{def:relational-langRevTF}
-Each combinator {{c : b1 <-> b2}} in {{langRevTF}} denotes a relation
+\label{def:relational-PiTF}
+Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
 as specified below.
@@
 
-The language {{langRevTF}} is still logically reversible. This is intuitively
+The language {{PiTF}} is still logically reversible. This is intuitively
 clear as the relation denoted by a combinator can be used in the reverse
 direction by swapping inputs and outputs.
+
+\begin{verbatim}
+In the rules above it is unclear as to which v is chosen in the rule
+ for eta. It is unclear that these rules are logically reversible.
+\end{verbatim}
 
@@
 \end{proposition}
+
+\subsection{Relational Semantics}
+
+Some thoughts on a relational semantics. We define {{ a <-> b }}
+$\equiv$ {{a R b}}
+
+\noindent
+We now have primitives:
+%subcode{opsem} include main
+% swap* '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
+% swap+ '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) ~|~ v1 in b1, v2 in b2]}
+% assocl* '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) ~|~ v1 in b1, v2 in b2, v3 in b3]}
+etc..
+
+\noindent
+closure combinators:
+%subcode{opsem} include main
+% sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
+% c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
+% c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
+etc..
+
+\noindent
+Now we have 
+%subcode{opsem} include main
+% eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
+% eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}
+
+Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
+have {{c^{dagger}: b2 <-> b1}} such that 
+%
+{{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
+straightforward from the definitions above.
+
+We can define evaluation as follows: 
+%
+{{ c v = {[ v' ~|~ (v, v') in c ]} }}.
+
+However this is unpleasant since it is unclear how to evaluate if
+{{c~v}} and {{c^{dagger} v'}} are dual by evaluation. This ties into a
+deeper issue of interpreting relations as programs.
+
+Q) If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
+{{b2}}? If yes, then we have a model for {{PiTF}} where the objects
+are sets {{b}} and the morphisms are the relations between objects of
+equal size.
 
@@
 operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
-language like {{langRev}} there are two natural families of trace operators
+language like {{Pi}} there are two natural families of trace operators
 that can be defined, an additive family (explored in detail in our previous
@@
 
-The above example should convince the reader that the language {{langRev}}
+The above example should convince the reader that the language {{Pi}}
 with multiplicative trace is expressive enough to model finite relational
@@
 
+\begin{verbatim}
 homotopy equivalence???
@@
 'categorified field'.
+
+\end{verbatim}
 

### 2013-02-12 frac.tex
several comments

@@
 \newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
+\newcommand{\roshan}[1]{\textsc{Roshan says:} 
+  \textit{#1}
+}
 
@@
 
-%subcode-inline{bnf-inline} name langRev
+%subcode-inline{bnf-inline} name Pi
 %! swap+ = \mathit{swap}^+
@@
 %! ]] = \rrbracket
+%! ^^^ = ^{\dagger}
 %! eta* = \eta
@@
 %! @ = \,@\,
+%! Pow = \mathcal{P}
 %! Pi = \Pi
@@
 %! PiEE = \Pi^{\eta\epsilon}_{+}
+%! PiT = \Pi^{o}
+%! PiTF = \Pi^{/}
 %! bullet = \bullet
 %! * = \times
-%! PiT = \Pi^{o}
-%! PiTF = \Pi^{/}
 
@@
 
+\roshan{It is unclear what the Id types refer to. We should provide
+  some more context here}
+
 This paper addresses this higher-order extension.  It does this via a novel
@@
 
+\roshan{It is unclear how fractions help differentiate between inputs
+  and outputs. On first reading this seemed promising, but nothing
+  in the rest of the paper substantiates this point of view.}
+
+\roshan{I think we should arrive at relational programming as a means
+  of giving semantics to fractionals, rather than the other way
+  round. To show that relational programming is reversible, one does
+  not need to resort to fractionals. Everyone understand that
+  relations can be reversed and none of this setup with {{Pi}} is
+  required for that. Our key point should be to communicate that
+  fractionals have a clean semantics in the setting of relational
+  programming rather than the other way round.}
+
 Syntactically, the extension with fractional types is simple.
@@
 Each type denotes a finite set of values as follows:
-\[\begin{array}{rcl}
-\llbracket 1 \rrbracket &=& \{ () \} \\
-\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \} \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
-\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
-\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
-\end{array}\]
+%subcode{opsem} include main
+%! <- = \leftarrow
+%! union = \cup
+% [[1]] '= {[ () ]}
+% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
+% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
+% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
 \end{definition}
@@
 
+\roshan{In the judgments below it is misleading that only {{eta*}}
+  appears indexed by a type. To construct each family of relations,
+  the corresponding definition must be indexed by their types. In
+  other words, there is nothing special about {{eta*}}. Further, I am
+  unsure of how to format this so that the combinators are correctly
+  indexed by their types for both primitives and for composition
+  combinators. We could do the following, but it seems to be a
+  overkill.}
+
+%subcode{proof} include main
+%@ v1 in [[b1]]
+%@ v2 in [[b2]]
+%@ v3 in [[b3]]
+%@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)
+
 %subcode{proof} include main
@@
 
-\begin{verbatim}
-In the rules above it is unclear as to which v is chosen in the rule
- for eta. It is unclear that these rules are logically reversible.
-\end{verbatim}
+\roshan{ In the rules above it is unclear as to which v is chosen in
+  the rule for eta. It is unclear that these rules are logically
+  reversible.  }
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Relational Semantics}
@@
 %subcode{opsem} include main
-% swap* '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
-% swap+ '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) ~|~ v1 in b1, v2 in b2]}
-% assocl* '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) ~|~ v1 in b1, v2 in b2, v3 in b3]}
+% swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
+% swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
+%                     && ~~~| v1 in b1, v2 in b2]}
+% assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
+%                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
 etc..
@@
 
-We can define evaluation as follows: 
-%
-{{ c v = {[ v' ~|~ (v, v') in c ]} }}.
-
-However this is unpleasant since it is unclear how to evaluate if
-{{c~v}} and {{c^{dagger} v'}} are dual by evaluation. This ties into a
-deeper issue of interpreting relations as programs.
-
-Q) If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Evaluation}
+
+I think defining eval is an open problem.  Consider a few relations:
+
+\begin{enumerate}
+\item 
+{{empty = {[ ]} }}
+
+\item 
+{{full =  {[ R, G, B ]} * {[ R, G, B ]} }}
+
+\item 
+{{id  =  {[ (R, R), (G, G), (B, B) ]} }}
+
+\item 
+{{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
+\end{enumerate}
+
+Can we define an {{eval}} that meaningfully distinguishes these? What
+can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
+several possible {{eval}} definitions:
+
+\begin{enumerate}
+\item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
+  eval maps {{b1 -> b2}}.
+
+\item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.
+
+\item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
+  {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.
+
+\end{enumerate}
+
+For the relations above, none of the suggested eval definitions give a
+logically reversible semantics.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \subsection{Model}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Completeness}
+
+\roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
 {{b2}}? If yes, then we have a model for {{PiTF}} where the objects
 are sets {{b}} and the morphisms are the relations between objects of
-equal size.
+equal size.}
+
+\roshan{ I suspect that all relations can be constructed, but we have
+  to show this. This is not obvious.}
 

### 2013-02-12 frac.tex
groupoid

@@
 \newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
+\newcommand{\amr}[1]{\textsc{Amr says:} #1}
 \newcommand{\roshan}[1]{\textsc{Roshan says:} 
@@
 %! <-> = \leftrightarrow
+%| -> = \rightarrow
 %! ~> = \leadsto
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%
 \subsection{Relational Semantics}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%
 \subsection{Evaluation}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \subsection{Model}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+%% \subsection{Model}
+
+%%%%%%%%%%%%%%%%%
 \subsection{Completeness}
@@
   to show this. This is not obvious.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Semantics}
+
+Each type in {{PiTF}} is interpreted as a \emph{groupoid}. A groupoid can be
+equivalently viewed as an oriented graph, a generalization of a group, or a
+special category. We present the categorical view below.
+
+A groupoid is defined by:
+\begin{itemize}
+\item a set of objects $x,y,\ldots$;
+\item for each pair of objects, a (possibly empty) set $G(x,y)$ of morphisms
+  $x \rightarrow y$; 
+\item for each object $x$, there exists an identity morphism {{id_x in G(x,x)}};
+\item for each triple of objects $x,y,z$, we have a function 
+{{comp_{x,y,z} : G(x,y) * G(y,z) -> G(x,z)}}
+\item there exists a function {{inv_{x,y} : G(x,y) -> G(y,x)}}
+\end{itemize}
+such that for all {{f : x -> y}}, {{g : y -> z}}, and {{h : z -> w}}, we have:
+%subcode{bnf} include main
+% comp_{x,x,y}(id_x,f) &=& f \\
+% comp_{x,y,y}(f,id_y) &=& f \\
+% comp_{x,y,z}(f, comp_{y,z,w}(g,h)) &=& comp_{x,z,w}(comp_{x,y,z}(f,g),h) \\
+% comp_{y,x,y}(inv_{x,y}(f),f) &=& id_y \\
+% comp_{x,y,x}(f,inv_{x,y}(f)) &=& id_x
+
+We now proceed to explain how each {{PiTF}} type denotes a groupoid. First,
+each {{Pi}} type denotes a set as follows.
+
+\begin{definition}[Denotation of {{Pi}} Value Types {{ [[ b ]] }}]
+\label{chx:def:denot}
+Each {{Pi}}type denotes a finite set of values as follows:
+%subcode{opsem} include main
+%! <- = \leftarrow
+%! union = \cup
+% [[1]] '= {[ () ]}
+% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
+% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
+\end{definition}
+
+For each of the types {{b}} above, the corresponding groupoid consists of the
+set {{ [[b]] }} of elements with only the trivial identity morphisms. The
+interesting groupoid is the one associated with {{1/b}}. It is defined as
+follows.
+
+\amr{stuck}
 

### 2013-02-16 talk.tex
subcoded going crazy

\documentclass[11pt]{beamer}

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{listings}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Macros

\newcommand{\arrow}[1]{{\color{blue}{#1}}}
\newcommand{\red}[1]{{\color{red}{#1}}}
\newcommand{\blue}[1]{{\color{blue}{#1}}}
\newcommand{\timesarr}{\otimes}
\newcommand{\plusarr}{\oplus}
\newcommand{\lcal}{\ensuremath{\lambda}-calculus\xspace}

\def\newblock{}

\newenvironment{floatrule}
    {\hrule width \hsize height .33pt \vspace{.5pc}}
    {\par\addvspace{.5pc}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! identr+ = \mathit{zeroi}
%! identl+ = \mathit{zeroe}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! trace+ = \mathit{trace}^+
%! trace* = \mathit{trace}^*
%! :-* = \multimap
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta
%! eps* = \epsilon
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

%subcode-inline{bnf-inline} name langArr
%! arr = \arrow{arr}
%! >>> = ~\arrow{\ggg\xspace}~
%! *** = ~\arrow{\otimes}~
%! +++ = ~\arrow{\oplus}~
%! create = \arrow{create}
%! erase = \arrow{erase}
%! afirst = \arrow{first}
%! asecond = \arrow{second}
%! aleft = \arrow{left}
%! atrace = \arrow{traceA}
%! fstA = \arrow{fstA}
%! sndA = \arrow{sndA}
%! leftA = \arrow{leftA}
%! rightA = \arrow{rightA}
%! |-->>* = \mapsto_{\mathsf{ML}}^{\ast}
%! |-->> = \mapsto_{\mathsf{ML}}

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev, langArr
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! |-->* = \mapsto^{\ast}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <=> = \Longleftrightarrow
%! <--> = \rightleftharpoons
%! <-> = \leftrightarrow
%! ~~> = \rightharpoonup
%! ~> = \leadsto
%! ::= = &::=&
%! /= = \neq
%! trans = \mathcal{T}
%! trans1 = \mathcal{T}_1
%! trans2 = \mathcal{T}_2
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! least = \phi
%! {[ = \{
%! ]} = \}
%! bottom = \bot
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! dagger = ^\dagger
%! @@ = \mu
%! STLC = \lambda^{\rightarrow}
%! STLClet = \textsf{LET}
%! STLCfor = \textsf{LET}^{o}
%! langArr = \textsf{ML}_{\Pi}
%! langArrT = \textsf{ML}_{\Pi^o}
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! if = \mathbf{if}
%! do = \mathbf{do}
%! then = \mathbf{then}
%! else = \mathbf{else}
%! * = \times

\usetheme{Boadilla}

\title{Fractional Types}
\author{Roshan P. James \and Zachary Sparks \\ ~ \\ Jacques Carette (McMaster Univ.) 
\\ ~ \\ Amr Sabry}
\institute{}

\date{February 15, 2013}

\begin{document}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{}

\vfill
{\Large \centerline{The Evil Equal Sign}}

\vfill

\pause

{\Large \centerline{Yes this one \qquad $=$}}
\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Equality in Physics}

\begin{itemize}

\vfill\item Take your favorite physical law

\vfill\item $F = m a$

\vfill\item $F$ is the force, obviously. It is one physical \emph{observable}

\vfill\item $m a$ is the rate of change of momentum. It is \red{another}
physical \emph{observable}

\vfill\item Now why and how could two \red{different} physical observables be
\red{equal}

\vfill\item Or, what does it even mean to say that two \red{different} things
are \red{the same}?

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Feynman, Landauer, Fredkin, and others}

\begin{itemize}

\vfill\item For centuries, it seemed nobody cared too much

\vfill\item Then computer science came and many people wanted to interpret
the laws of physics computationally, even if only to simulate them.

\vfill\item Argument was made that it is more appropriate to think that
\red{different} observables are related by an \red{isomorphism} and \red{not}
an equality.

\vfill\item We have an isomorphism that \red{witnesses}, \red{explains}, and
\red{models} the \red{process} of transforming one observable to another.

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Equality in Mathematics and Logic (\blue{Prehistory})}

\begin{itemize}

\vfill\item Curry-Howard isomorphism: logic, propositions, and
  proofs have computational content.

\vfill\item The statement $e_1 = e_2$ requires proof.

\vfill\item The statement $e_1 = e_2$ is a computation.

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Equality in Mathematics and Logic (\blue{History})}

\begin{itemize}

\vfill\item Martin-L\"of type theory as a foundation for
  constructive mathematics: a \red{computable set theory} if you wish

\vfill\item A crucial component of constructive type theory is \red{equality
  types}: given two terms $e_1, e_2 : A$, we have the \red{type}
$\textit{Id}(e_1,e_2)$ whose canonical inhabitant is $\mathit{refl} :
\textit{Id}(e,e)$.

\vfill\item A term of type $\textit{Id}(e_1,e_2)$ \red{witnesses},
\red{explains}, and \red{models} the process of transforming $e_1$ to $e_2$
and vice-versa.

\vfill\item If you've used Coq or Agda, you know what I am talking about

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Equality in Mathematics and Logic (\blue{New Program})}

\begin{itemize}

  \vfill\item \red{Homotopy version of Martin-L\"of Type Theory} or
  \red{Groupoid Interpretation of Martin-L\"of Type Theory}

\vfill\item Classical definition of homotopy: ``two functions are homotopic
if one can be continuously deformed into the other''

\vfill\item Not only do we care about the existence of a proof to show that
two things are equal but we want this proof to be ``smooth'' and
``elementary'' in some sense

\vfill\item Isomorphisms upon isomorphisms upon isomorphisms. Let me explain
using \pause category theory

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Category Theory}

A category consists of:

\begin{itemize}

\vfill\item a class of objects;

\vfill\item a class of morphisms; each morphism has a source object and a
target object

\vfill\item a binary operation $\circ$ called composition of morphisms

\end{itemize}

such that \ldots

\vfill Before we get to the axioms, note that \red{isomorphism} is a central
notion in category theory; most categorical constructions cannot distinguish
between isomorphic objects

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Axioms of a Category}

\begin{itemize}

\vfill\item \red{Associativity}: if $f : A \rightarrow B$, $g : B
  \rightarrow C$, and $h : C \rightarrow D$, then $h \circ (g \circ f) = 
  (h \circ g) \circ f$.

\vfill\item \red{Identity}: for every object $A$, there exists a morphism 
  $\mathit{id}_A : A \rightarrow A$ such that for every morphism 
  $f : A \rightarrow B$, we have 
  $\mathit{id}_B \circ f = f = f \circ \mathit{id}_A$.

\end{itemize}

\vfill

\pause Any questions, comments, or observations?

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{$n$-categories}

\begin{itemize}

\vfill\item The equality signs in the axioms should have really really
  really annoyed you. What do they even mean? Equality where and how?

\vfill\item Get rid of these equalities and replace them by isomorphisms.

\vfill\item The morphisms $\mathit{id}_A$, $\mathit{id}_B$, $f$, $g$, $h$
above become \red{objects} in a second-level category and the equality is now
expressed as morphisms between these objects.

\pause

\vfill\item In English, we now have: $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$

\vfill\item Of course we can iterate this idea

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Goal: Develop a Computational Model}

\begin{itemize}

\vfill\item You might ask: why?

\vfill\item How is this related to anything ``\red{useful}''?

\vfill\item Reason I: most important for me personally: \red{Aesthetics!}
\blue{(Symmetry)}

\vfill\item Meta-reason: ``being in tune with the universe'' is likely to
produce long, lasting, and significant contributions to science

\vfill\item ``Practical'' reasons, coming next

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Why Computing with Isomorphisms?}

\begin{itemize}

\vfill\item Programs execution conserves ``resources''; a better linear logic

\vfill\item Programs obey the principle of \red{conservation of information}
(experimental validation of Landauer's principle last year); accidental
information leaks impossible

\vfill\item Program semantics is consistent with the laws of physics:
possible models of quantum, biological, or other emerging ideas

\vfill\item A theory of observation, observational equivalence, that is
itself computational and that does not arbitrarily select a particular set of
equivalences as canonical (e.g., observe assignments to memory but do not
observe memory allocation)

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Why Computing with Isomorphisms?}

\begin{itemize}

\vfill\item Type isomorphisms have a long history in PL

\vfill\item First used to \blue{search} libraries: say you are looking for a
function to format a real number to a given number of digits. Does it have type:

{{ Real -> Int -> String}}

{{ Int -> Real -> String}}

{{ Real * Int -> String}}

{{ Int * Real -> String}}

\vfill\item Use equivalence relation on types to query the library with one
of the possible types and get results matching any equivalent type

\vfill\item Related ideas: type isomorphisms allow program reuse, proof
reuse, etc.

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{}

\vfill
{\Large \centerline{Isomorphisms between Finite Types}}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Finite Types}

%subcode{bnf} include main
% base types, b ::= 0 | 1 | b+b | b*b

\begin{itemize}

  \vfill\item 0: no wire

  \vfill\item 1: wire carrying a token {{()}}

  \vfill\item {{b1+b2}}: two wires, one of which carries a value 
  \red{(particle)}

  \vfill\item {{b1*b2}}: two wires, both of which carry a value 
  \red{(wave)}

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Graphical notation: Wires}

\begin{itemize}
\item The simplest sort of diagram is the {{id : b <-> b}} 
\end{itemize}

\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/b-wire.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/b-wire-value.pdf}
}
\end{center}
\end{multicols}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Graphical notation: Products}

\begin{itemize}
\item The product type {{b1*b2}} may be represented using either one wire labeled
{{b1*b2}} or two parallel wires labeled {{b1}} and {{b2}}. 
\end{itemize}

\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-one-wire.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-one-wire-value.pdf}
}
\end{center}
\end{multicols}
\vfill
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/product-two-wires-value.pdf}
}
\end{center}
\end{multicols}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Graphical notation: Sums}

\begin{itemize}
\item  Sum types may similarly be represented by one wire or using
  parallel wires with a {{+}} operator between them. 
\end{itemize}

\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-one-wire.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
}
\end{center}
\end{multicols}
\vfill
\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
}
\end{center}
\end{multicols}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Isomorphisms for Finite Types}

\begin{itemize}

  \vfill\item What are the sound and complete type isomorphisms for
  this system?

  \vfill\item Folklore that the axioms of a \red{commutative semiring}
  constitute a sound and complete set of type isomorphisms.

  \vfill\item Proof in LICS 2002 (Balat, Di Cosmo, and Fiore)

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Sound and Complete Isomorphisms}

{\scriptsize
%subcode{bnf} include main
% 0 + b &<->& b // identity for~ +
% b1 + b2 &<->& b2 + b1 // commutativity for~ +
% b1 + (b2 + b3) &<->& (b1 + b2) + b3 // associativity for~ +
}
\pause

{\scriptsize
%subcode{bnf} include main
% 1 * b &<->& b // identity for~ *
% b1 * b2 &<->& b2 * b1 // commutativity for~ *
% b1 * (b2 * b3) &<->& (b1 * b2) * b3 //associativity for~ *
}
\pause

{\scriptsize
%subcode{bnf} include main
% 0 * b &<->& 0 // 0 annihilates
% (b1 + b2) * b3 &<->& (b1 * b3) + (b2 * b3) // distribute over~ +
}

\pause

{\scriptsize
%subcode{proof} include main
%@ ~
%@@ b1 <-> b1
%
%@ b1 <-> b2
%@@ b2 <-> b1
%
%@ b1 <-> b2
%@ b2 <-> b3
%@@ b1 <-> b3
%
%@ equivalence
%
%---
%@ b1 <-> b3
%@ b2 <-> b4
%@@ (b1 + b2) <-> (b3 + b4) 
%
%@ b1 <-> b3
%@ b2 <-> b4
%@@ (b1 * b2) <-> (b3 * b4)
%
%@ congruence
%
}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Graphical notation: Associativity}

\begin{itemize}
\item Associativity is implicit in the graphical language. Three parallel
  wires represent {{b1*(b2*b3)}} or {{(b1*b2)*b3}}, based on the context.
\end{itemize}

\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/assoc.pdf}
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Graphical notation: Commutativity}

\begin{itemize}
\item Commutativity is represented by crisscrossing wires.
\end{itemize}

\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/swap_times.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/swap_plus.pdf}
}
\end{center}
\end{multicols}

\vfill

\begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/swap_times_value.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/swap_plus_value.pdf}
}
\end{center}
\end{multicols}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Graphical notation: Distributivity and Factoring}

\begin{itemize}
\item   Distributivity and factoring are represented using the dual boxes shown
  below:
\end{itemize}

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/thesis/dist.pdf}
\end{center}
\begin{center}
  \includegraphics{diagrams/thesis/factor.pdf}
\end{center}
\end{multicols}

\vfill

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/thesis/dist-wire-value1.pdf}
\end{center}
\begin{center}
  \includegraphics{diagrams/thesis/dist-wire-value2.pdf}
\end{center}
\end{multicols}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Categorical Interlude}

\begin{itemize}

  \vfill\item A symmetric bimonoidal category

  \vfill\item A symmetric monoidal category $(0,\oplus)$

  \vfill\item A symmetric monoidal category $(1,\otimes)$

  \vfill\item The tensor $\otimes$ distributes over $\oplus$ (\red{but
    not vice-versa}).

  \vfill\item Categorical models of linear logic, quantum computing,
  knot theory, etc. are based on such categories.

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{A Reversible Language {{langRev}} }

\begin{itemize}

  \vfill\item A textual notation for the previous diagrams

  \vfill\item Can express all functions on booleans (by embedding them
  in reversible functions)

  \vfill\item With recursion, can express any Turing computable function

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Programming in {{langRev}}: Booleans, Negation}

\vfill

\begin{itemize}
\item We encode booleans by \red{ {{bool = 1 + 1}} }

{{~~~~~~~~~~~~true = left ()}}

{{~~~~~~~~~~~~false = right ()}}

\vfill

\item Negation:

\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/thesis/swap_plus.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/thesis/swap_plus_value.pdf}
\end{center}

\end{multicols}

\vfill

    {{~~~~~~ not : bool <-> bool }}

    {{~~~~~~ not = swap+}} 
\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Programming in {{langRev}}: Conditionals}

\begin{itemize}

\item Reversible conditional:
{{ if_c (flag, b) = if flag then (flag, c(b)) else (flag, b) }}

\vfill

\item Given {{c : b <-> b}}, then:

{{if_c : bool * b <-> bool * b }}

{{ if_c = dist (;) ((id (*) c) (+) id) (;) factor}}

\vfill

\begin{center}
  \includegraphics{diagrams/thesis/cnot.pdf}
\end{center}

\vfill
\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Recursion}

\begin{itemize}

\vfill\item Add \red{recursive types} (so that we can now define
natural numbers, lists, trees, etc.)

\vfill\item Add categorical \red{trace} (the categorical abstraction
of \red{feedback}, \red{looping}, and \red{recursion})

\vfill

\begin{center}
  \includegraphics{diagrams/trace_plus.pdf}
\end{center}

\vfill

\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{}

\vfill
{\Large \centerline{Higher-Order Extension}}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Extending the language {{langRevEE}} }

\begin{itemize}
\vfill\item Processes can observe other processes

\vfill\item Proofs of equalities can themselves be compared for equality

\vfill\item Generally useful to have functions as values; code-as-data;
data-as-code

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Negative Information}

\begin{itemize}
  \vfill\item Introduce the idea of \red{negative information} or
  \red{request} for information;

  \vfill\item Information is logarithmic: entropy of a random variable of
  drawn from a set of size {{N}} is {{ {@1@ log N @} }}.

  \vfill\item To get {{ {@1@ -log N @} }} information, we need to set of
  ``size'' {{1/N}}.

  \vfill\item Assume these sets exist for now \ldots and let's add
  \red{fractional} types and values.

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Extending the language {{langRevEE}} }

\begin{itemize}
\vfill\item Types
%subcode{bnf} include main
% base types, b ::= 0 | 1 | b+b | b*b | {@1@ 1/b @}
\vfill\item Isomorphisms
%subcode{opsem} include main
% eta* &: 1 <-> (1/b) * b :& eps*
\end{itemize}

\vfill

Let's remove the {{0}} to avoid some potential subtle problems with division
by {{0}} (whatever that means at the level of types)

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Multiplicative Duality}

\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eta_times1.pdf}
}
\end{center}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eps_times1.pdf}
}
\end{center}  
\end{multicols}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Sanity Checks}

\begin{itemize}
  \vfill\item Idea of negative information makes sense in a world in which
  information is conserved;

\vfill\item Can ``borrow'' because the infrastructure guarantees that all
debts must be accounted for;

\vfill\item Sets with fractional size do exist! \red{Intuition:} Consider an
object with a symmetry group of size {{N}}. If you are counting naively you
will count this object {{N}} times but all these counts are ``the
same''. Divide by {{N}} to get the ``real count''.

\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Homotopy Cardinality or Groupoid Cardinality}

\begin{itemize}
\vfill\item Formally this idea is known as \red{homotopy cardinality} or
  \red{groupoid} cardinality

\vfill\item That exactly what the ``new program'' in type theory is all
  about

\vfill\item In type theory, the intuition is that a statement may have
  several proofs but we don't want to count all the proofs as different, so 
  we divide by the size of the equivalence class of all proofs

  \vfill\item Anyway, it seems that we are on a sound footing. Lots of
  details to work out tough! For now, let's have fun and program with these
  fractional types. What can we do with them?

\end{itemize}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Programming with Fractional Types and Values}

\begin{itemize}
\vfill\item Introduce the abbreviation: {{ b1 :-* b2 = 1/b1 * b2}}

\vfill\item We can turn an isomorphism {{b1 <-> b2}} into a value:

%subcode{opsem} include main
%! columnStyle = rcl
% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
% name c &=& eta*_{b1} (;) (id (*) c)

\vfill\item We can express recursion:

%subcode{opsem} include main
%! columnStyle = rcl
% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Programming with Fractional Types and Values}

\begin{itemize}

\vfill\item We can apply and compose these higher-order relations:

%subcode{opsem} include main
%! columnStyle = rcl
% apply &:& (b1 :-* b2) * b1 <-> b2
% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
% 
% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
% compose &=& assocr* (;) 
%   && (id (*) (assocl* (;) (swap* (*) id) (;) (eps*_{b2} (*) id) (;) unite))

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{More Intuitively: Duality}

\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/coherence.pdf} 
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{More Intuitively: Functions}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/function.pdf}
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{More Intuitively}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/compose1.pdf}
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{More Intuitively}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/compose.pdf}
}
\pause
\scalebox{1.5}{
  \includegraphics{diagrams/compose2.pdf}
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{More Intuitively}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/apply1.pdf}
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{More Intuitively}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/apply2.pdf}
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{SAT Solver}

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/sat2.pdf}
}
\end{center}

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/sat3.pdf}
}
\end{center}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{}

{\Large \centerline{Conclusions --- The Bigger Picture}}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Physical Perspective on Computation}

\begin{itemize}

  \vfill\item Computer science abstractions are based on \blue{old}
  physics

  \vfill\item Computer applications are more and more ``physical''

  \vfill\item Physical principles such as \red{conservation of
    information} should be part of our foundational abstractions

  \vfill\item Other principles?

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Computational Perspective on Physics}

\begin{itemize}

  \vfill\item Towards a computational understanding of quantum mechanics

  \vfill\item Creation and annihilation of ``entangled'' particles
  appears fundamental; modeled by computational symmetries and
  dualities

  \vfill\item Negative information flow and negative entropy are
  existing concepts in quantum information

  \vfill\item Discrete Quantum Computing: Take the Hilbert space
  formalism and replace the underlying field of \red{uncomputable}
  complex numbers with a finite field! In the simplest case, we get
  {{langRev}} with one addition: a \red{disjoint union} operation.

\end{itemize}

\vfill

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2013-02-16 frac.tex
subcoded going crazy
another pass at intro with citations etc

@@
 \author{Roshan P. James \and Zachary Sparks \and Jacques Carette \and Amr Sabry}
-\institute{}
+\institute{Indiana University and McMaster University}
 \maketitle
@@
   In previous work, we developed a \emph{first-order},
-  information-preserving, and reversible programming language {{Pi}}
-  founded on type isomorphisms. Being restricted to first-order types limits
-  the expressiveness of the language: it is not possible, for example, to
+  information-preserving, and reversible programming language {{Pi}} founded
+  on type isomorphisms. Being restricted to first-order types limits the
+  expressiveness of the language: it is not possible, for example, to
   abstract common program fragments into a higher-level combinator. In this
-  paper, we introduce a higher-order extension of {{Pi}} based on the
-  novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
+  paper, we introduce a higher-order extension of {{Pi}} based on the novel
+  concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
   fractional type {{1/v}} represents \emph{negative} information. A
@@
 
-In the physics community, Landauer~\cite{Landauer:1961}, Feynman, and others
-have interpreted the laws of physics as fundamentally related to
-computation. In more detail, the great majority of the laws of physics are
-formulated as equalities between different physical observables. For
-instance, Newton's second law of classical mechanics equates the force acting
-on a system to its rate of change of momentum. The insight of Landauer,
-Feynman, etc. is that \emph{different} physical observables should not be
-related by an \emph{equality} but more appropriately by an \emph{isomorphism}
-that witnesses, explains, and models the process of transforming one
-observable to the other.
+In the physics community, Landauer~\cite{Landauer,Landauer:1961},
+Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
+the laws of physics as fundamentally related to computation. In more detail,
+the great majority of the laws of physics are formulated as equalities
+between different physical observables. For instance, Newton's second law of
+classical mechanics equates the force acting on a system to its rate of
+change of momentum. The insight of Landauer, Feynman, etc. is that
+\emph{different} physical observables should not be related by an
+\emph{equality} but more appropriately by an \emph{isomorphism} that
+witnesses, explains, and models the process of transforming one observable to
+the other.
 
@@
 of the simply typed $\lambda$-calculus originally intended to provide a
-rigorous framework for constructive mathematics. This has been further
-extended (first by Streicher and Awodey, and subsequently by many others)
-with a new \emph{identity type} representing the proposition
-that two terms are ``equal.'' Specifically, given two terms $a$ and $b$ of
-the same type $A$, one forms the type $\texttt{Id}_A(a,b)$ representing the
-proposition that $a$ and $b$ are equal: in other words, a term of type
-$\texttt{Id}_A(a,b)$ witnesses, explains, and models the process of
-transforming $a$ to $b$ and vice-versa.
+rigorous framework for constructive
+mathematics~\cite{citeulike:7374951}. This theory has been further extended
+with \emph{identity types} representing the proposition that two terms are
+``equal.'' (See~\cite{warren,streicher} for a survey.) Briefly speaking,
+given two terms $a$ and $b$ of the same type $A$, one forms the type
+$\texttt{Id}_A(a,b)$ representing the proposition that $a$ and $b$ are equal:
+in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
+models the process of transforming $a$ to $b$ and vice-versa.
 
@@
 by providing one of the many possible isomorphic types for the desired
-function. More recently, type isomorphisms have taken a more central role as
-\emph{the} fundamental computational mechanism from which more conventional,
-i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
-in our own previous work, we started with the notion of type isomorphism as
-the ``right'' starting point for a theory of computation. The central
-technical contribution was the development of a family of programming
-languages, {{Pi}} with various superscripts, in which every computation
-is an isomorphism. Other major results are that all such computations
-preserve the information-theoretic entropy, that the language is logically
-reversible, and that it has a model in symmetric monoidal categories (which
-are the backbone of models for linear logic and quantum computation). We have
-also invested a significant amount of time in using {{Pi}} as an actual
-programming language and established several idioms for programming large
-classes of interesting (recursive) programs in {{Pi}}.
+function~\cite{Rittri:1989:UTS:99370.99384}. More recently, type isomorphisms
+have taken a more central role as \emph{the} fundamental computational
+mechanism from which more conventional, i.e., irreversible computation, is
+derived. In more detail, in our own previous
+work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
+notion of type isomorphism as the ``right'' starting point for a theory of
+computation. The central technical contribution was the development of a
+family of programming languages, {{Pi}} with various superscripts, in which
+every computation is an isomorphism. Other major results are that all such
+computations preserve the information-theoretic entropy, that the languages
+are logically reversible, and that they have models in various flavors of
+symmetric monoidal categories (which are the backbone of models for linear
+logic and quantum computation). We have also invested a significant amount of
+time in using {{Pi}} as an actual programming language and established
+several idioms for programming large classes of interesting (recursive)
+programs in {{Pi}}.
 
@@
 
-\roshan{It is unclear what the Id types refer to. We should provide
-  some more context here}
-
-This paper addresses this higher-order extension.  It does this via a novel
-type construction, which we dub a \emph{fractional type} because of its
-interaction with product types.  By adding a formal (linear) dual to
-(type theoretic) products, we can internalize the notion that ``taking an
-input'' corresponds to negative information, which turns out to be the 
-basis of first-class functions.
-
-The plan of the paper is therefore
-as follows: \todo{roadmap}
+Technically speaking, obtaining a higher-order extension requires the
+construction of a \emph{compact closed category} from the underlying monoidal
+category for {{Pi}}. Although several such constructions that work in special
+cases are known in the literature (surveyed below), there are many subtleties
+in adapting these ideas to {{Pi}}. Our main novel technical device to
+achieving the higher-order extension is a \emph{fractional type}, which we so
+named because of its duality with conventional product types.  By adding a
+formal (linear) dual to (type theoretic) products, we can internalize the
+notion that ``taking an input'' corresponds to negative information, which
+turns out to be the basis of first-class functions. The remainder of the
+paper reviews {{Pi}} and then introduces the syntax and semantics of the
+extension with fractional types. We then study the properties of the extended
+language and establish its expressiveness via several constructions and
+examples that exploit its ability to model higher-order computations.
+
+%% \roshan{It is unclear what the Id types refer to. We should provide
+%%   some more context here}
+%% I think the previous paragraph in the intro that talks about Martin-Lof type 
+%% theory provides the context. Anything else??
 
@@
 
-In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
-the simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
+In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain the
+simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
 richer languages, e.g., in a Turing complete language that admits
 non-termination, isomorphisms necessarily denote \emph{partial
-  bijections}~\cite{rc2011,infeffects}. As we demonstrate in this section,
-once we make the extension to higher-order functions, we must interpret the
-syntactic isomorphisms as \emph{relations}. The intuitive idea, is that the
-fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be able to map the
-element {{() : 1}} to any pair of the form {{(1/v,v)}} with
+  bijections}~\cite{rc2011,James:2012:IE:2103656.2103667}. As we demonstrate
+in this section, once we make the extension to higher-order functions, we
+must interpret the syntactic isomorphisms as \emph{relations}. The intuitive
+idea, is that the fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be
+able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
 {{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to

### 2013-02-16 frac.tex
latex
revised sec 2

@@
 %! <-> = \leftrightarrow
-%| -> = \rightarrow
+%! -> = \rightarrow
 %! ~> = \leadsto
@@
 
+%% AMR:
 %% \roshan{It is unclear what the Id types refer to. We should provide
@@
 \paragraph*{Data.}
-We view {{Pi}} as having two levels:  it has traditional values, given by
+We view {{Pi}} as having two levels:  it has traditional values, given by:
 %subcode{bnf} include main
 % values, v ::= () | left v | right v | (v, v)
-\noindent and these are classified by ordinary types
+\noindent and these are classified by ordinary types:
 %subcode{bnf} include main
 % value types, b ::= 1 | b + b | b * b 
-
-Types include the unit type {{1}}, sum types {{b1+b2}}, and products types
-{{b1*b2}}.  Values includes {{()}} which is the only value of type {{1}},
-{{left v}} and {{right v}} which inject {{v}} into a sum type, and
-{{(v1,v2)}} which builds a value of product type. But these values should be
-regarded as largely ancillary.  In particular we do not treat the above
+\noindent Types include the unit type {{1}}, sum types {{b1+b2}}, and
+products types {{b1*b2}}.  Values includes {{()}} which is the only value of
+type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
+and {{(v1,v2)}} which builds a value of product type. But these values should
+be regarded as largely ancillary.  In particular we do not treat the above
 values as first-class citizens.  They only occur when we want to observe the
@@
 %dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
-
-\noindent Each line of the above table introduces a pair of dual constants%
-\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
-isomorphism in the middle.  These are the base (non-reducible) terms of the
-second, principal level of {{Pi}}.
-
-Note how the above has two readings: first as a set of typing relations for a
-set of constants. Second, if these axioms are seen as universally quantified,
-orientable statements, they also induce transformations of the (traditional)
-values.  The (categorical) intuition here is that these axioms have
-computational content because they witness isomorphisms rather than merely
-stating an extensional equality.
+\noindent Each line of the above table introduces a pair of dual
+constants~\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
+the type isomorphism in the middle.  These are the base (non-reducible) terms
+of the second, principal level of {{Pi}}. Note how the above has two
+readings: first as a set of typing relations for a set of constants. Second,
+if these axioms are seen as universally quantified, orientable statements,
+they also induce transformations of the (traditional) values. The
+(categorical) intuition here is that these axioms have computational content
+because they witness isomorphisms rather than merely stating an extensional
+equality.
 
@@
 %@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
-
 \noindent The syntax is overloaded: we use the same symbol at the value-type level
@@
 
-\noindent
-To summarize, the syntax of {{Pi}} is given as follows.
-
-\begin{definition}{(Syntax of {{Pi}})}
-\label{def:Pi}
-%subcode{bnf} include main
-% value types, b ::= 1 | b+b | b*b 
-% values, v ::= () | left v | right v | (v,v) 
-%
-% iso.~types, t ::= b <-> b
-% base iso ::= swap+ | assocl+ | assocr+ 
-%     &|& unite | uniti | swap* | assocl* | assocr* 
-%     &|& dist | factor 
-% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-% top level term, l ::= c v
-\end{definition}
-
-Mathematically, {{Pi}} can be interpreted in the category of finite sets
-and bijections: each value type denotes a finite set of the corresponding size
+% \noindent
+% To summarize, the syntax of {{Pi}} is given as follows.
+
+% \begin{definition}{(Syntax of {{Pi}})}
+% \label{def:Pi}
+% %subcode{bnf} include main
+% % value types, b ::= 1 | b+b | b*b 
+% % values, v ::= () | left v | right v | (v,v) 
+% %
+% % iso.~types, t ::= b <-> b
+% % base iso ::= swap+ | assocl+ | assocr+ 
+% %     &|& unite | uniti | swap* | assocl* | assocr* 
+% %     &|& dist | factor 
+% % iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+% % top level term, l ::= c v
+% \end{definition}
+
+Mathematically, {{Pi}} can be interpreted in the category of finite sets and
+bijections: each value type denotes a finite set of the corresponding size
 and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
-denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{Pi}} is
-given using two mutually recursive interpreters: one going forward and one
-going backwards. The use of {{sym}} switches control from one evaluator to
-the other. We will not present the operational semantics until the next
-section, in the context of the extended language. For now, we state without
-proof that the evaluation of well-typed combinators always terminates and
-that {{Pi}} is logically reversible, i.e., that for all combinators 
+denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{Pi}} is given
+using two mutually recursive interpreters: one going forward and one going
+backwards. The use of {{sym}} switches control from one evaluator to the
+other. We will present the operational semantics in the next section along
+with the extension with fractional types and values. For now, we state
+without proof that the evaluation of well-typed combinators always terminates
+and that {{Pi}} is logically reversible, i.e., that for all combinators 
 {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
@@
 
-\jacques{Since we have Agda code for the above, should we refer to it
-here, and make it available?}
-
-Note that a small modification of {{Pi}} (the addition of an
-empty type {{0}} with its usual laws) would exactly capture, at the type
-level, the notion of \emph{semiring} (occasionally called a \emph{rig}) where
-we replace equality by isomorphism.  This can also be also compared to
-\emph{bimonoidal categories}.  As is usual with with type theories, the
-coherence conditions (pentagonal and triangle identity) are not explicit,
-but would correspond to certain identity types being trivial.  {{Pi}}
-instead, at the type level, models a commutative ringoid where the
-multiplicative structure forms a commutative monoid, but the additive structure
-is just a commutative magma.  We chose not to include {{0}} as a technical
-convenience for the next section.
+%% AMR:
+%% \jacques{Since we have Agda code for the above, should we refer to it
+%% here, and make it available?}
+%% That is a great idea but it would take a significant amount of time to
+%% clean up the code and make it presentable. If someone is willing to do
+%% that, then we should definitely add a reference to the code and make it
+%% available.
+
+The language presented above, at the type level, models a commutative ringoid
+where the multiplicative structure forms a commutative monoid, but the
+additive structure is just a commutative magma.  Note that the version of
+{{Pi}} that include the empty type with its usual laws exactly captures, at
+the type level, the notion of a \emph{semiring} (occasionally called a
+\emph{rig}) where we replace equality by isomorphism.  This can also be also
+compared to \emph{bimonoidal categories}.  As is usual with type theories,
+the coherence conditions (pentagonal and triangle identity) are not explicit,
+but would correspond to certain identity types being trivial.
 

### 2013-02-17 frac.tex
beginning of sec3

@@
 The language {{Pi}} models isomorphisms of values rather well.  However,
-purposefully, it has no notion of \emph{input} or \emph{output}; more
-precisely, because of strict preservation of information, these two concepts
-coincide.  This is the root of reversibility.
-
-But if we want to model functions of any flavor, we need to differentiate
+purposefully, it has no distinguished notion of \emph{input} or
+\emph{output}; more precisely, because of strict preservation of information,
+these two concepts coincide.  This is the root of reversibility. But if we
+want to model functions of any flavor as values, we need to differentiate
 between these notions.  Our idea, inspired by computational dualities and
@@
 
-\roshan{It is unclear how fractions help differentiate between inputs
-  and outputs. On first reading this seemed promising, but nothing
-  in the rest of the paper substantiates this point of view.}
-
-\roshan{I think we should arrive at relational programming as a means
-  of giving semantics to fractionals, rather than the other way
-  round. To show that relational programming is reversible, one does
-  not need to resort to fractionals. Everyone understand that
-  relations can be reversed and none of this setup with {{Pi}} is
-  required for that. Our key point should be to communicate that
-  fractionals have a clean semantics in the setting of relational
-  programming rather than the other way round.}
+% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
+% \roshan{It is unclear how fractions help differentiate between inputs
+%   and outputs. On first reading this seemed promising, but nothing
+%   in the rest of the paper substantiates this point of view.}
+%
+
+% AMR: relations are indeed only presented as a way to explain the semantics
+% \roshan{I think we should arrive at relational programming as a means
+%   of giving semantics to fractionals, rather than the other way
+%   round. To show that relational programming is reversible, one does
+%   not need to resort to fractionals. Everyone understand that
+%   relations can be reversed and none of this setup with {{Pi}} is
+%   required for that. Our key point should be to communicate that
+%   fractionals have a clean semantics in the setting of relational
+%   programming rather than the other way round.}
 
@@
 % base iso ::= ... | eta*_b | eps*_b
+where:
+{{eta*_b : 1 <-> 1/b * b : eps*_b}}
 \end{definition}
@@
 conflate negative-information values with positive-information values just
-because they happen to be isomorphic.
-
-In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain the
-simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
-richer languages, e.g., in a Turing complete language that admits
-non-termination, isomorphisms necessarily denote \emph{partial
+because they happen to be isomorphic. (See Sec.~\ref{sec:cat} for a more
+thorough discussion about the meaning of fractional types and values in an
+algebraic and categorical sense.)
+
+For now, we proceed with an operational understanding of the language. The
+semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
+interpreted as a \emph{relation} between the sets denoted by {{b1}} and
+{{b2}} respectively.
+
+\begin{definition}[Denotation of Value Types {{ [[ b ]] }}]
+\label{chx:def:denot}
+Each type denotes a finite set of values as follows:
+%subcode{opsem} include main
+%! <- = \leftarrow
+%! union = \cup
+% [[1]] '= {[ () ]}
+% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
+% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
+% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
+\end{definition}
+
+The denotation of combinators as relations is given below. We specify
+relations using a deductive system whose judgments are of the form 
+{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
+denoted by {{c}}.
+
+\begin{definition}[Relational semantics]
+\label{def:relational-PiTF}
+Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
+as specified below. 
+%subcode{proof} include main
+%@ ~
+%@@ (left v) ~~swap+~~ (right v)
+%
+%@ ~
+%@@ (right v) ~~swap+~~ (left v)
+%----
+%@ ~
+%@@ (left v) ~~assocl+~~ (left (left v))
+%
+%@ ~
+%@@ (right (left v)) ~~assocl+~~ (left (right v))
+%----
+%@ ~
+%@@ (right (right v)) ~~assocl+~~ (right v)
+%
+%@ ~
+%@@ (left (left v)) ~~assocr+~~ (left v)
+%----
+%@ ~
+%@@ (left (right v)) ~~assocr+~~ (right (left v))
+%
+%@ ~
+%@@ (right v) ~~assocr+~~ (right (right v))
+%----
+%@ ~
+%@@ ((), v) ~~unite~~ v
+%
+%@ ~
+%@@ v ~~uniti~~ ((), v)
+%----
+%@ ~
+%@@ (v1, v2) ~~swap*~~ (v2, v1)
+%----
+%@ ~
+%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
+%
+%@ ~
+%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
+%----
+%@ ~
+%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
+%
+%@ ~
+%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
+%----
+%@ ~
+%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
+%
+%@ ~
+%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
+%----
+%@ ~
+%@@ v ~~id~~ v
+%
+%@ v' ~~c~~ v
+%@@ v ~~(sym c)~~ v'
+%
+%@ v1 ~~c1~~ v2
+%@ v2 ~~c2~~ v3
+%@@ v1 ~~(c1(;)c2)~~ v3
+%----
+%@ v ~~c1~~ v'
+%@@ (left v) ~~(c1 (+) c2)~~ (left v')
+%
+%@ v ~~c2~~ v'
+%@@ (right v) ~~(c1 (+) c2)~~ (right v')
+%----
+%@ v1 ~~c1~~ v1'
+%@ v2 ~~c2~~ v2'
+%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
+%----
+%@ v in [[ b ]]  
+%@@ () ~~eta*_b~~ (1/v,v)
+%
+%@ ~
+%@@ (1/v,v) ~~eps*_b~~ ()
+\end{definition}
+
+\newpage
+~
+\newpage
+
+\roshan{In the judgments below it is misleading that only {{eta*}}
+  appears indexed by a type. To construct each family of relations,
+  the corresponding definition must be indexed by their types. In
+  other words, there is nothing special about {{eta*}}. Further, I am
+  unsure of how to format this so that the combinators are correctly
+  indexed by their types for both primitives and for composition
+  combinators. We could do the following, but it seems to be a
+  overkill.}
+
+%subcode{proof} include main
+%@ v1 in [[b1]]
+%@ v2 in [[b2]]
+%@ v3 in [[b3]]
+%@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)
+
+In the previous section, we have been able to maintain the simple idea that a
+syntactic isomorphism denotes a \emph{bijection}. In richer languages, e.g.,
+in a Turing complete language that admits non-termination, isomorphisms
+necessarily denote \emph{partial
   bijections}~\cite{rc2011,James:2012:IE:2103656.2103667}. As we demonstrate
@@
 
-\begin{definition}[Denotation of Value Types {{ [[ b ]] }}]
-\label{chx:def:denot}
-Each type denotes a finite set of values as follows:
-%subcode{opsem} include main
-%! <- = \leftarrow
-%! union = \cup
-% [[1]] '= {[ () ]}
-% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
-% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
-% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
-\end{definition}
-
-The denotation of combinators as relations is given below. We specify
-relations using a deductive system whose judgments are of the form 
-{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
-denoted by {{c}}.
-
-\begin{definition}[Relational semantics]
-\label{def:relational-PiTF}
-Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
-as specified below.
-
-\roshan{In the judgments below it is misleading that only {{eta*}}
-  appears indexed by a type. To construct each family of relations,
-  the corresponding definition must be indexed by their types. In
-  other words, there is nothing special about {{eta*}}. Further, I am
-  unsure of how to format this so that the combinators are correctly
-  indexed by their types for both primitives and for composition
-  combinators. We could do the following, but it seems to be a
-  overkill.}
-
-%subcode{proof} include main
-%@ v1 in [[b1]]
-%@ v2 in [[b2]]
-%@ v3 in [[b3]]
-%@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)
-
-%subcode{proof} include main
-%@ ~
-%@@ (left v) ~~swap+~~ (right v)
-%
-%@ ~
-%@@ (right v) ~~swap+~~ (left v)
-%----
-%@ ~
-%@@ (left v) ~~assocl+~~ (left (left v))
-%
-%@ ~
-%@@ (right (left v)) ~~assocl+~~ (left (right v))
-%----
-%@ ~
-%@@ (right (right v)) ~~assocl+~~ (right v)
-%
-%@ ~
-%@@ (left (left v)) ~~assocr+~~ (left v)
-%----
-%@ ~
-%@@ (left (right v)) ~~assocr+~~ (right (left v))
-%
-%@ ~
-%@@ (right v) ~~assocr+~~ (right (right v))
-%----
-%@ ~
-%@@ ((), v) ~~unite~~ v
-%
-%@ ~
-%@@ v ~~uniti~~ ((), v)
-%----
-%@ ~
-%@@ (v1, v2) ~~swap*~~ (v2, v1)
-%----
-%@ ~
-%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
-%
-%@ ~
-%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
-%----
-%@ ~
-%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
-%
-%@ ~
-%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
-%----
-%@ ~
-%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
-%
-%@ ~
-%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
-%----
-%@ ~
-%@@ v ~~id~~ v
-%
-%@ v' ~~c~~ v
-%@@ v ~~(sym c)~~ v'
-%
-%@ v1 ~~c1~~ v2
-%@ v2 ~~c2~~ v3
-%@@ v1 ~~(c1(;)c2)~~ v3
-%----
-%@ v ~~c1~~ v'
-%@@ (left v) ~~(c1 (+) c2)~~ (left v')
-%
-%@ v ~~c2~~ v'
-%@@ (right v) ~~(c1 (+) c2)~~ (right v')
-%----
-%@ v1 ~~c1~~ v1'
-%@ v2 ~~c2~~ v2'
-%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
-%----
-%@ v in [[ b ]]  
-%@@ () ~~eta*_b~~ (1/v,v)
-%
-%@ ~
-%@@ (1/v,v) ~~eps*_b~~ ()
-\end{definition}
-
 The language {{PiTF}} is still logically reversible. This is intuitively
@@
 \end{proposition}
-
-%%%%%%%%%%%%%
-\subsection{Relational Semantics}
 
@@
 
-%%%%%%%%%%%%%%%%
-\subsection{Evaluation}
-
 I think defining eval is an open problem.  Consider a few relations:
@@
 logically reversible semantics.
-
-%%%%%%%%%%%%%%%%%
-%% \subsection{Model}
-
-%%%%%%%%%%%%%%%%%
-\subsection{Completeness}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semantics}
+\section{Expressiveness and Examples}
 
@@
 
-\amr{stuck}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Expressiveness}
-
 Fractional types and values add considerable expressiveness to our
@@
 % emptyset & eps*_{bool}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Examples}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion}
+\label{sec:cat}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}

### 2013-02-17 frac.tex
sec 3 kind of ok

@@
 want to model functions of any flavor as values, we need to differentiate
-between these notions.  Our idea, inspired by computational dualities and
-polarity, is to introduce a \emph{formal dual} for our types.  We thus
-consider a type in a negative position to possess negative information, as it
-is really a request for information.  Since information is logarithmic, this
-means that a request for information should behave like a \emph{fractional
-  type}.
+between these notions.  Our idea, inspired by computational
+dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
+polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
+of \emph{compact
+  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
+is to introduce a \emph{formal dual} for our types.  We thus consider a type
+in a negative position to possess negative information, as it is really a
+request for information.  Since information is logarithmic, this means that a
+request for information should behave like a \emph{fractional type}.
 
@@
 or as representing the absence of some information (i.e., \emph{negative
-  information}), that can only be supplied by {{left ()}}. The combinator
-{{eta*_b : 1 <-> 1/b * b}} represents a fission point which creates --- out
-of no information --- an equal amount of negative and positive
-information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
-up an equal amount of negative and positive information producing no residual
-information.  The best analogy is probably that of vectors and co-vectors,
-also know as column vectors and row vectors (respectively): in the finite
-dimensional case, they are isomorphic, but they are nevertheless quite
+  information}), that can only be reconciliated by {{left ()}}. The
+combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission point which
+creates --- out of no information --- an equal amount of negative and
+positive information. Symmetrically, the combinator 
+{{eps*_b : 1/b * b <-> 1}} 
+matches up an equal amount of negative and positive information producing no
+residual information.  The best analogy is probably that of vectors and
+co-vectors, also know as column vectors and row vectors (respectively) where
+the multiplication of the row vector by the column vector is a scalar. In the
+finite dimensional case, they are isomorphic, but they are nevertheless quite
 different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
@@
 because they happen to be isomorphic. (See Sec.~\ref{sec:cat} for a more
-thorough discussion about the meaning of fractional types and values in an
-algebraic and categorical sense.)
-
-For now, we proceed with an operational understanding of the language. The
-semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
+thorough discussion about the meaning of fractional types and values in both an
+algebraic and categorical sense.) For now, we proceed with an operational 
+understanding of the language.
+
+The semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
 interpreted as a \emph{relation} between the sets denoted by {{b1}} and
@@
 
-\newpage
-~
-\newpage
-
-\roshan{In the judgments below it is misleading that only {{eta*}}
-  appears indexed by a type. To construct each family of relations,
-  the corresponding definition must be indexed by their types. In
-  other words, there is nothing special about {{eta*}}. Further, I am
-  unsure of how to format this so that the combinators are correctly
-  indexed by their types for both primitives and for composition
-  combinators. We could do the following, but it seems to be a
-  overkill.}
-
-%subcode{proof} include main
-%@ v1 in [[b1]]
-%@ v2 in [[b2]]
-%@ v3 in [[b3]]
-%@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)
-
-In the previous section, we have been able to maintain the simple idea that a
-syntactic isomorphism denotes a \emph{bijection}. In richer languages, e.g.,
-in a Turing complete language that admits non-termination, isomorphisms
-necessarily denote \emph{partial
-  bijections}~\cite{rc2011,James:2012:IE:2103656.2103667}. As we demonstrate
-in this section, once we make the extension to higher-order functions, we
-must interpret the syntactic isomorphisms as \emph{relations}. The intuitive
-idea, is that the fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be
+If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 
+{{c : b1 <-> b2}} would define a total one-to-one function between the sets 
+{{ [[ b1 ]] }} and {{ [[ b2 ]] }}. 
+This is indeed the situation for the language {{Pi}} 
+defined in the previous section. In the presence of fractionals, this simple 
+interpretation of combinators as one-to-one functions is not sufficient.
+\footnote{The interpretation of combinators as one-to-one functions is also 
+not valid in the presence of recursive 
+types~\cite{rc2011,James:2012:IE:2103656.2103667}.}
+The intuitive reason, is that the fission point combinator 
+{{eta*_b : 1 <-> 1/b * b}} must be
 able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
@@
 not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
-undefined, i.e., it does not produce {{():1}}. In other words, some of the
-combinators are not partial and some produce more than one result for a
-particular input, and hence we move to the setting of general binary
-relations to interpret our combinators.
-
-This setting is not only consistent with logical reversibility but is
-arguably the \emph{right} framework for considering expressive reversible
-languages (including the interesting examples of quantum languages).  Indeed,
-even though the categorical semantics of {{PiTF}} is not the main focus
-of this paper, we find it helpful to state that the category of finite sets
-and relations in which we are going to interpret our language, is an example
-of a \emph{biproduct dagger compact closed category} which are categories
+undefined, i.e., it does not produce {{():1}}.
+
+The relational semantics can be given an operational interpretation by
+defining two functions: a forward evaluator and a backwards evaluator that
+exercise the relation in opposite directions. It follows that if the forward
+evaluation of {{c v1}} produces {{v2}} as a possible answer then the
+backwards evaluation of {{c v2}} produces {{v1}} as a possible answer. More
+generally, the semantics of {{PiTF}} using sets and relations, i.e., in the
+category of finite sets and relations suggests a few additional immediate
+properties.  The category of finite sets and relations is an example of a
+\emph{biproduct dagger compact closed category} which are the categories
 introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
@@
 types). The latter is intuitively clear as any relation between $A$ and $B$
-can be represented as a subset of {{A * B}}, i.e., as a value.
-
-To interpret {{PiTF}} in this category, we associate a finite set
-{{ [[ b ]] }} with each type {{b}} and a relation {{ [[ c ]] }} over
-{{ [[ b1 ]] }} and {{ [[ b2 ]] }} for each combinator {{ c : b1 <-> b2 }}.
-
-The language {{PiTF}} is still logically reversible. This is intuitively
-clear as the relation denoted by a combinator can be used in the reverse
-direction by swapping inputs and outputs.
-
-\roshan{ In the rules above it is unclear as to which v is chosen in
-  the rule for eta. It is unclear that these rules are logically
-  reversible.  }
-
-\begin{proposition}[Logical Reversibility]
-\label{chx:prop:logrev-tracep}
-For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
-{{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
-\end{proposition}
-
-Some thoughts on a relational semantics. We define {{ a <-> b }}
-$\equiv$ {{a R b}}
-
-\noindent
-We now have primitives:
-%subcode{opsem} include main
-% swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
-% swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
-%                     && ~~~| v1 in b1, v2 in b2]}
-% assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
-%                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
-etc..
-
-\noindent
-closure combinators:
-%subcode{opsem} include main
-% sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
-% c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
-% c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
-etc..
-
-\noindent
-Now we have 
-%subcode{opsem} include main
-% eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
-% eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}
-
-Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
-have {{c^{dagger}: b2 <-> b1}} such that 
-%
-{{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
-straightforward from the definitions above.
-
-I think defining eval is an open problem.  Consider a few relations:
-
-\begin{enumerate}
-\item 
-{{empty = {[ ]} }}
-
-\item 
-{{full =  {[ R, G, B ]} * {[ R, G, B ]} }}
-
-\item 
-{{id  =  {[ (R, R), (G, G), (B, B) ]} }}
-
-\item 
-{{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
-\end{enumerate}
-
-Can we define an {{eval}} that meaningfully distinguishes these? What
-can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
-several possible {{eval}} definitions:
-
-\begin{enumerate}
-\item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
-  eval maps {{b1 -> b2}}.
-
-\item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.
-
-\item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
-  {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.
-
-\end{enumerate}
-
-For the relations above, none of the suggested eval definitions give a
-logically reversible semantics.
-
-\roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
-{{b2}}? If yes, then we have a model for {{PiTF}} where the objects
-are sets {{b}} and the morphisms are the relations between objects of
-equal size.}
-
-\roshan{ I suspect that all relations can be constructed, but we have
-  to show this. This is not obvious.}
+can be represented as a subset of {{A * B}}, i.e., as a value. We return to
+such properties of {{PiTF}} in Sec.~\ref{sec:cat} after we present
+programming examples that illustrate {{PiTF}} and its ability to express
+various higher-order notions.
+
+% \roshan{ In the rules above it is unclear as to which v is chosen in
+%   the rule for eta. It is unclear that these rules are logically
+%   reversible.  }
+
+% \begin{proposition}[Logical Reversibility]
+% \label{chx:prop:logrev-tracep}
+% For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
+% {{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
+% \end{proposition}
+
+% Some thoughts on a relational semantics. We define {{ a <-> b }}
+% $\equiv$ {{a R b}}
+
+% \noindent
+% We now have primitives:
+% %subcode{opsem} include main
+% % swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
+% % swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
+% %                     && ~~~| v1 in b1, v2 in b2]}
+% % assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
+% %                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
+% etc..
+
+% \noindent
+% closure combinators:
+% %subcode{opsem} include main
+% % sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
+% % c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
+% % c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
+% etc..
+
+% \noindent
+% Now we have 
+% %subcode{opsem} include main
+% % eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
+% % eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}
+
+% Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
+% have {{c^{dagger}: b2 <-> b1}} such that 
+% %
+% {{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
+% straightforward from the definitions above.
+
+% I think defining eval is an open problem.  Consider a few relations:
+
+% \begin{enumerate}
+% \item 
+% {{empty = {[ ]} }}
+
+% \item 
+% {{full =  {[ R, G, B ]} * {[ R, G, B ]} }}
+
+% \item 
+% {{id  =  {[ (R, R), (G, G), (B, B) ]} }}
+
+% \item 
+% {{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
+% \end{enumerate}
+
+% Can we define an {{eval}} that meaningfully distinguishes these? What
+% can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
+% several possible {{eval}} definitions:
+
+% \begin{enumerate}
+% \item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
+%   eval maps {{b1 -> b2}}.
+
+% \item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.
+
+% \item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
+%   {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.
+
+% \end{enumerate}
+
+% For the relations above, none of the suggested eval definitions give a
+% logically reversible semantics.
+
+% \roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
+% {{b2}}? If yes, then we have a model for {{PiTF}} where the objects
+% are sets {{b}} and the morphisms are the relations between objects of
+% equal size.}
+
+% \roshan{ I suspect that all relations can be constructed, but we have
+%   to show this. This is not obvious.}
+
+% \roshan{In the judgments below it is misleading that only {{eta*}}
+%   appears indexed by a type. To construct each family of relations,
+%   the corresponding definition must be indexed by their types. In
+%   other words, there is nothing special about {{eta*}}. Further, I am
+%   unsure of how to format this so that the combinators are correctly
+%   indexed by their types for both primitives and for composition
+%   combinators. We could do the following, but it seems to be a
+%   overkill.}
+
+% %subcode{proof} include main
+% %@ v1 in [[b1]]
+% %@ v2 in [[b2]]
+% %@ v3 in [[b3]]
+% %@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)
 

### 2013-02-17 frac.tex
changes to section 1-3

@@
 communities (physics, mathematics, and computer science) towards replacing
-\emph{equalities} by \emph{isomorphisms}. The combined programme originating
-from these research areas has sparked a huge amount of research that unveiled
+\emph{equalities} by \emph{isomorphisms}. The combined programme 
+has sparked a huge amount of research that unveiled
 new and surprising connections between geometry, algebra, logic, and
@@
 
-In the physics community, Landauer~\cite{Landauer,Landauer:1961},
+In the physics community, Landauer~\cite{Landauer:1961,Landauer},
 Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
@@
 the great majority of the laws of physics are formulated as equalities
-between different physical observables. For instance, Newton's second law of
-classical mechanics equates the force acting on a system to its rate of
-change of momentum. The insight of Landauer, Feynman, etc. is that
+between different physical observables, such as Newton's second law of
+classical mechanics ($F=\dot{p}$).
+The insight of Landauer, Feynman, etc. is that
 \emph{different} physical observables should not be related by an
@@
 with \emph{identity types} representing the proposition that two terms are
-``equal.'' (See~\cite{warren,streicher} for a survey.) Briefly speaking,
+``equal.'' (See~\cite{streicher,warren} for a survey.) Briefly speaking,
 given two terms $a$ and $b$ of the same type $A$, one forms the type
@@
 
-A major open problem remains, however. This concerns a higher-order extension
+A major open problem remains, however: a higher-order extension
 of {{Pi}}. This extension is of fundamental importance in all the
@@
 allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
-to be itself expressed as an isomorphism as a higher type
+to itself be expressed as an isomorphism (of a higher type)
 $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
 higher-order types allow code to abstract over other code fragments as well
-as the manipulation of code as data and data as code which is central to many
-algorithms and application domains.
+as the manipulation of code as data and data as code.
 
@@
 effect this shape transformation. Of course, ``shapes of values'' are very
-familiar already: they are usually called \emph{types}.  But usually one
+familiar already: they are usually called \emph{types}.  But frequently one
 designs a type system as a method of classifying terms, with the eventual
@@
 
-\paragraph*{Isomorphisms.} The terms of {{Pi}} are witnesses to type
-isomorphisms.  They have (iso) types {{b <-> b}}.  
-Specifically, they are witnesses to the following type isomorphisms:
+\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
+(value) type isomorphisms.  They are also typed, by the shape of the (value)
+type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
+the following type isomorphisms:
 %subcode{bnf} include main
@@
 \noindent Each line of the above table introduces a pair of dual
-constants~\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
+constants\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
 the type isomorphism in the middle.  These are the base (non-reducible) terms
@@
 %% available.
+%% Jacques:  Right.  I'll definitely work on cleaning up the paper first,
+%% then see if there is time left for the code.
 
@@
 where the multiplicative structure forms a commutative monoid, but the
-additive structure is just a commutative magma.  Note that the version of
+additive structure is just a commutative semigroup.  Note that the version of
 {{Pi}} that include the empty type with its usual laws exactly captures, at
@@
 
-% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
 % \roshan{It is unclear how fractions help differentiate between inputs
@@
 %   in the rest of the paper substantiates this point of view.}
-%
-
-% AMR: relations are indeed only presented as a way to explain the semantics
+% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
+% Jacques: Exactly!  1/a is negative information is input, b is positive is
+%    output.
+%
+
 % \roshan{I think we should arrive at relational programming as a means
@@
 %   programming rather than the other way round.}
+% AMR: relations are indeed only presented as a way to explain the semantics
 
@@
 \label{def:PiT}
-We extend Def.~\ref{def:Pi} with:
+We extend our previous definition with:
 %subcode{bnf} include main
@@
 %----
-%@ v in [[ b ]]  
+%@ ~
 %@@ () ~~eta*_b~~ (1/v,v)

### 2013-02-17 frac.tex
added explanation of implicit premises

@@
 
+The operational semantics define a relation rather than a function, now, but the
+goal is for them to still be reversible; in particular, we wish for the relation
+to mode check so that for any combinator {{c}} we can treat either its left or
+its right side as an input; this allows us to define {{sym}} as just an argument
+swap. In the interest of brevity, we have treated the type annotations in the
+base isomorphisms as implicit; for example, the rule for {{assocr*}} should
+really read
+
+%subcode{proof} include main
+%@ v1 in b1
+%@ v2 in b2
+%@ v3 in b3
+%@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))
+
 If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 

### 2013-02-19 code/frac/PiFrac.hs
type error for eps

@@ -70,8 +70,8 @@ data a :<=> b where
   FoldN   :: Either () Nat :<=> Nat
   UnfoldN :: Nat :<=> Either () Nat
 -- Trace operators for looping/recursion
-  TraceP :: (B a, B b1, B b2) => (Either a b1 :<=> Either a b2) -> (b1 :<=> b2)
-  TraceT :: (B a, B b1, B b2) => ((a, b1) :<=> (a, b2)) -> (b1 :<=> b2)
+--  TraceP :: (B a, B b1, B b2) => (Either a b1 :<=> Either a b2) -> (b1 :<=> b2)
+--  TraceT :: (B a, B b1, B b2) => ((a, b1) :<=> (a, b2)) -> (b1 :<=> b2)
 -- Eta/Psi
   EtaT :: (B a) => () :<=> (Recip a, a)
   EpsT :: (B a) => (Recip a, a) :<=> ()
@@ -102,8 +102,8 @@ adjoint FoldB = UnfoldB
 adjoint UnfoldB = FoldB
 adjoint FoldN = UnfoldN
 adjoint UnfoldN = FoldN
-adjoint (TraceP c) = TraceP (adjoint c)
-adjoint (TraceT c) = TraceT (adjoint c)
+--adjoint (TraceP c) = TraceP (adjoint c)
+--adjoint (TraceT c) = TraceT (adjoint c)
 adjoint EtaT = EpsT
 adjoint EpsT = EtaT
 
@@ -147,6 +147,7 @@ eval FoldN (Left ()) = return 0
 eval FoldN (Right n) = return (n+1)
 eval UnfoldN 0 = return (Left ())
 eval UnfoldN n = return (Right (n-1))
+{--
 eval (TraceP c) v = do v' <- eval c (Right v)
                        loop c v'
     where
@@ -159,9 +160,10 @@ eval (TraceT c) v = try elems
                             if a == a' 
                               then v' : try as
                               else try as
+--}
 eval EtaT () = [(Recip v, v) | v <- elems]
 eval EpsT (Recip v1, v2) | v1 == v2 = [()]
-                         | otherwise = []
+                         | otherwise = error "Type error"
 
 evalR :: (a :<=> b) -> b -> [a]
 evalR Id a = return a
@@ -202,6 +204,7 @@ evalR FoldN 0 = return (Left ())
 evalR FoldN n = return (Right (n-1))
 evalR UnfoldN (Left ()) = return 0
[…]

### 2013-02-19 code/frac/PiFrac2.hs
strange code

@@ -1,0 +1,802 @@
+{-# LANGUAGE TypeOperators, GADTs, TypeSynonymInstances, NamedFieldPuns #-}
+
+module PiFrac2 where
+
+import Data.List
+
+-----------------------------------------------------------------------
+-- Isomorphisms between VALUES
+
+data Recip a = Recip a deriving (Show, Eq)
+
+-- the individual clauses relating values to values
+
+data v1 :<-> v2 where
+  IdUnit :: () :<-> ()
+  IdBool1   :: Bool :<-> Bool
+  IdBool2   :: Bool :<-> Bool
+  SwapBool1 :: Bool :<-> Bool
+  SwapBool2 :: Bool :<-> Bool
+  EtaBool1  :: () :<-> (Recip Bool,Bool)
+  EtaBool2  :: () :<-> (Recip Bool,Bool)
+  EpsilonBool1 :: (Recip Bool,Bool) :<-> ()
+  EpsilonBool2 :: (Recip Bool,Bool) :<-> ()
+
+eval :: (v1 :<-> v2) -> v1 -> v2
+eval IdBool1 False = False
+eval IdBool2 True = True
+eval SwapBool1 False = True
+eval SwapBool2 True = False
+eval EtaBool1 () = (Recip False,False)
+eval EtaBool2 () = (Recip True,True)
+eval EpsilonBool1 (Recip False,False) = ()
+eval EpsilonBool2 (Recip True,True) = ()
+
+-- a type is a set values {v1, v2, ...} TOGETHER with the equalities 
+-- v1 :<-> v2 which witness how various values are related to each other
+
+data Typ a = Typ { elems :: [a], eq :: [a :<-> a] }
+
+unitTyp :: Typ ()
+unitTyp = Typ { elems = [()],
+                eq = [IdUnit] }
+
+boolTyp :: Typ Bool 
+boolTyp = Typ { elems = [False,True],
+                eq = [IdBool1,IdBool2,SwapBool1,SwapBool2] }
+
+-- Now we have isomorphisms between TYPES; this is the level of isomorphisms
+-- we've always had
+
+data b1 :<=> b2 where
+  Swap :: Typ Bool :<=> Typ Bool
+
+evalB :: (b1 :<=> b2) -> b1 -> b2
+evalB Swap (Typ{elems,eq}) = undefined -- ??
+
+{--
+
+  EtaT :: (B a) => () :<=> (Recip a, a)
[…]

### 2013-02-19 frac.tex
minor tweaks to section 4

@@
 \usepackage{graphicx}
+\usepackage{comment}
 
@@
 defined in the previous section. In the presence of fractionals, this simple 
-interpretation of combinators as one-to-one functions is not sufficient.
+interpretation of combinators as one-to-one functions is not sufficient%
 \footnote{The interpretation of combinators as one-to-one functions is also 
 not valid in the presence of recursive 
-types~\cite{rc2011,James:2012:IE:2103656.2103667}.}
+types~\cite{rc2011,James:2012:IE:2103656.2103667}.}.
 The intuitive reason, is that the fission point combinator 
@@
 
+\begin{comment}
+% We are not doing groupoids in this paper
 Each type in {{PiTF}} is interpreted as a \emph{groupoid}. A groupoid can be
@@
 Fractional types and values add considerable expressiveness to our
-language. In the following, we use {{bool}} as an abbreviation of {{1+1}} 
+language. 
+\end{comment}
+In the following, we use {{bool}} as an abbreviation of {{1+1}} 
 with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
@@
 %     &<->& (1/b1 * b1) * b2) & (swap* (*) id)
-%     &<->& 1 * b2 & (eps*{b1} (*) id)
+%     &<->& 1 * b2 & (eps*_{b1} (*) id)
 %     &<->& b2 & (unite)
@@
 each other and the value of type {{b2}} is exposed with no
-constraints. Otherwise, the mismatched values are \emph{annihilated}.
-
-Function or relation composition is only slightly more involved:
+constraints. Otherwise, the result is undefined.
+
+\noindent A flipped variant is also useful:
+%subcode{opsem} include main
+%! columnStyle = rcl
+% apply' &:& b1 * (b1 :-* b2) <-> b2
+% apply' &=& assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
+
+\noindent Function or relation composition is now straightforward:
 
@@
 % compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
-% compose &=& assocr* (;) (id (*) (assocl* (;) (swap* (*) id) (;) (eps*_{b2} (*) id) (;) unite))
+% compose &=& assocr* (;) (id (*) apply')
 
@@
 %     &<->& (1/b1 * (b2 * (1/b2 * b3)) & (assocr*)
-%     &<->& (1/b1 * ((b2 * 1/b2) * b3) & (assocl*)
-%     &<->& (1/b1 * ((1/b2 * b2) * b3) & (assocl*)
-%     &<->& (1/b1 * (1 * b3) & (eps*_{b2})
-%     &<->& (1/b1 * b3) & (unite)
+%     &<->& (1/b1 * b3) & (id * apply')
 %     &=& b1 :-* b3
@@
 The final remarkable property of fractional types is that they
-constitute, a form of self-inverse negation. In particular, we have:
+are involutive.  In particular, we have:
 
@@
 %! columnStyle = rcl
-% doubleNeg &:& b <-> 1/(1/b)
-% doubleNeg &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
+% doubleDiv &:& b <-> 1/(1/b)
+% doubleDiv &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
 
@@
 {{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
-given a value {{v2 : b2}} but to evaluate {{c}}, we must provide a pair
-consisting of some value {{v1 : b1}} together with the given {{v2 : b2}}. The
-needed value {{v1 : b1}} cannot be arbitrary, however. We do have a
-constraint that the value {{v1 : b1}} must be such that it is also produced
-by as the first component of the result. In general, there may be several
-such values fixed-point values or none.  A few examples should help
-understand the subtleties.
+given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
+consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
+The needed value {{v1 : b1}} cannot be arbitrary: it must be such that is
+is the value produced as the first component of the result.  In general, there
+may be several such fixed-point values or none.  A few examples should help
+explain the subtleties.
 
 \begin{example}
-Consider the combinator {{c}}:
+Consider the combinator
 
@@
 
-\noindent Using the multiplicative trace operator, we can construct the
-combinator {{trace* c : 1 <-> 1}}. Applying this combinator to 
+\noindent Using {{trace*}}, we construct the
+combinator {{trace* c : 1 <-> 1}}. Applying this to 
 {{() : 1}} requires us to find a value {{b : bool}} such that 
@@
 that {{(b,()) ~~c~~ (b,())}} which is impossible since {{c}} negates
-{{b}}. Operationally, we would expect the evaluation of such a
-combinator to produce no value. 
+{{b}}, i.e. {{c}} has no fixed points. Operationally, the evaluation of
+such a combinator should produce no value (if it even terminates).
 \end{example}
@@
 % {[ ((),false) ]}  & (uniti)
-% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool})
+% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
 % {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
-% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (not (*) id)
+% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
 % {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
-% emptyset & eps*_{bool}
-
-Now that we have the formal semantics, we consider a generalized version of
-Ex.~\ref{ch3:ex:annihilate}.
-
+% emptyset & (eps*_{bool} * id)
+
+\jacques{in the current paper, this seems to no longer make sense}
 \begin{example}[Annihilation]

### 2013-02-20 frac.tex
major changes

@@
 given two terms $a$ and $b$ of the same type $A$, one forms the type
-$\texttt{Id}_A(a,b)$ representing the proposition that $a$ and $b$ are equal:
+$\texttt{Id}_A(a,b)$ representing the proposition that~$a$ and~$b$ are equal:
 in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
@@
 
-A major open problem remains, however: a higher-order extension
-of {{Pi}}. This extension is of fundamental importance in all the
-originating research areas. In physics, it allows the process itself or the
-observer to be treated as ``data'' that can be transformed by higher-level
-processes or observed by meta-level observers. In mathematics and logic, it
-allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
-to itself be expressed as an isomorphism (of a higher type)
+A major open problem remains, however: a higher-order extension of
+{{Pi}}. This extension is of fundamental importance in all the originating
+research areas. In physics, it allows a process or observer to be treated as
+``data'' that can be transformed by higher-level processes or observed by
+meta-level observers. In mathematics and logic, it allows the equivalence
+between different proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed
+as an isomorphism (of a higher type)
 $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
@@
 construction of a \emph{compact closed category} from the underlying monoidal
-category for {{Pi}}. Although several such constructions that work in special
-cases are known in the literature (surveyed below), there are many subtleties
-in adapting these ideas to {{Pi}}. Our main novel technical device to
-achieving the higher-order extension is a \emph{fractional type}, which we so
-named because of its duality with conventional product types.  By adding a
-formal (linear) dual to (type theoretic) products, we can internalize the
-notion that ``taking an input'' corresponds to negative information, which
-turns out to be the basis of first-class functions. The remainder of the
-paper reviews {{Pi}} and then introduces the syntax and semantics of the
-extension with fractional types. We then study the properties of the extended
-language and establish its expressiveness via several constructions and
-examples that exploit its ability to model higher-order computations.
+category for~{{Pi}}. Although the general idea of such a construction is
+well-understood, the details of adapting it to an actual programming language
+as subtle (as we explain in the main body of the paper).  Our main novel
+technical device to achieving the higher-order extension is a
+\emph{fractional type}, which we so named because of its duality with
+conventional product types.  By adding a formal (linear) dual to (type
+theoretic) products, we can internalize the notion that ``taking an input''
+corresponds to negative information, which turns out to be the basis of
+first-class functions. The remainder of the paper reviews {{Pi}} and then
+introduces the syntax and semantics of the extension with fractional
+types. We then study the properties of the extended language and establish
+its expressiveness via several constructions and examples that exploit its
+ability to model higher-order computations.
 
@@
 % value types, b ::= 1 | b + b | b * b 
-\noindent Types include the unit type {{1}}, sum types {{b1+b2}}, and
+\noindent Types include the unit type {{1}}, sum types {{b1 + b2}}, and
 products types {{b1*b2}}.  Values includes {{()}} which is the only value of
@@
 type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
-the following type isomorphisms:
+the following isomorphisms:
 %subcode{bnf} include main
@@
 Mathematically, {{Pi}} can be interpreted in the category of finite sets and
-bijections: each value type denotes a finite set of the corresponding size
-and each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
-denoted by {{b1}} and {{b2}}. Operationally, the semantics of {{Pi}} is given
-using two mutually recursive interpreters: one going forward and one going
+bijections: each value type denotes a finite set of a size calculated by
+viewing the types as natural numbers and each combinator {{c : b1 <-> b2}}
+denotes a bijection between the sets denoted by~{{b1}}
+and~{{b2}}. Operationally, the semantics of {{Pi}} is given using two
+mutually recursive interpreters: one going forward and one going
 backwards. The use of {{sym}} switches control from one evaluator to the
@@
 evaluation of {{c v1}} produces {{v2}} as a possible answer then the
-backwards evaluation of {{c v2}} produces {{v1}} as a possible answer. More
-generally, the semantics of {{PiTF}} using sets and relations, i.e., in the
-category of finite sets and relations suggests a few additional immediate
+backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.
+
+More generally, the semantics of {{PiTF}} using sets and relations, i.e., in
+the category of finite sets and relations suggests a few additional immediate
 properties.  The category of finite sets and relations is an example of a
@@
 
+In the following, we use {{bool}} as an abbreviation of {{1+1}} 
+with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
+abbreviation for {{right ()}}.
+
+%%%%%%%%%%%%
+\subsection{First-Class Relations}
+
+As motivated, with fractional types, we can define a value that denotes a
+first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
+pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
+and a value {{v2 : b2}}. In other words, it corresponds to a function or
+relation which when ``given'' a value {{v1 : b1}} ``releases'' the value 
+{{v2 : b2}}. To emphasize this view, we introduce the abbreviation:
+
+%subcode{bnf} include main
+%  b1 :-* b2 &::=& 1/b1 * b2
+
+\noindent which suggests a function-like behavior for the pair of a fractional value
+and a regular value. What is remarkable is that we can almost trivially turn
+any combinator {{c : b1 <-> b2}} into a (constant) value of 
+type {{b1 :-* b2}} as shown below:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
+% name c &=& eta*_{b1} (;) (id (*) c)
+
+As a simple example, consider the combinator {{swap+}} denoting boolean
+negation.  It is straightforward to check that the combinator 
+{{name swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:
+
+%subcode{proof} include main
+%@ ~
+%@@ () ~~(name swap+)~~ (1/false,true)
+%
+%@ ~
+%@@ () ~~(name swap+)~~ (1/true,false)
+
+\noindent The relation {{name swap+}} maps {{()}} to a value which can be used
+to convert any {{false}} to {{true}} and vice-versa. We formalize this
+``apply'' combinator below.
+
+%%%%%%%%%%%%
+\subsection{Higher-Order Relations}
+
+We are now a small step from implementing various higher-order
+combinators that manipulate functions or relations. In particular, we
+can \emph{apply} and \emph{compose} values representing relations as
+shown below:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% apply &:& (b1 :-* b2) * b1 <-> b2
+% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
+
+\noindent At the type level, the computation proceeds as follows:
+%subcode{bnf} include main
+%  (b1 :-* b2) * b1 &=& (1/b1 * b2) * b1 & 
+%     &<->& b1 * (1/b1 * b2) & (swap*)
+%     &<->& (b1 * 1/b1) * b2) & (assocl*)
+%     &<->& (1/b1 * b1) * b2) & (swap* (*) id)
+%     &<->& 1 * b2 & (eps*_{b1} (*) id)
+%     &<->& b2 & (unite)
+
+\noindent Intuitively, we simply match the incoming argument of type {{b1}} with
+the constraint encoded by the function. If they match, they cancel
+each other and the value of type {{b2}} is exposed with no
+constraints. Otherwise, the result is undefined.
+
+\noindent A flipped variant is also useful:
+%subcode{opsem} include main
+%! columnStyle = rcl
+% apply' &:& b1 * (b1 :-* b2) <-> b2
+% apply' &=& assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
+
+\noindent Function or relation composition is now straightforward:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
+% compose &=& assocr* (;) (id (*) apply')
+
+\noindent Again, at the type level, the computation proceeds as follows:
+
+%subcode{bnf} include main
+%  (b1 :-* b2) * (b2 :-* b3) &=& (1/b1 * b2) * (1/b2 * b3) & 
+%     &<->& (1/b1 * (b2 * (1/b2 * b3)) & (assocr*)
+%     &<->& (1/b1 * b3) & (id * apply')
+%     &=& b1 :-* b3
+
+%%%%%%%%%%%%
+\subsection{Duality}
+
+The final remarkable property of fractional types is that they
+are involutive.  In particular, we have:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% doubleDiv &:& b <-> 1/(1/b)
+% doubleDiv &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
+
+\noindent At the type level, this computation proceeds as follows:
+
+%subcode{bnf} include main
+%  b &<->& (1 * b) & (uniti)
+%    &<->& (1/(1/b) * 1/b) * b & (eta*_{1/b} (*) id)
+%    &<->& 1/(1/b) * (1/b * b) & (assocr*)
+%    &<->& 1/(1/b) * 1 & (id (*) eps*_b)
+%    &<->& 1 * 1/(1/b) & (swap*)
+%    &<->& 1/(1/b) & (unite)
+
+%%%%%%%%%%%%
+\subsection{Feedback, Iteration, and Trace Operators}
+
+Mathematically speaking, recursion and iteration can be expressed using
+categorical trace
+operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
+language like {{Pi}} there are two natural families of trace operators
+that can be defined, an additive family (explored in detail in our previous
+work~\cite{rc2011}) and a multiplicative family (which is the focus of this
+section).
+
+In this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
+want to cancel the common type {{b1}}, to produce a new combinator
+{{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
+given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
+consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
+The needed value {{v1 : b1}} cannot be arbitrary: it must be such that is
+is the value produced as the first component of the result.  In general, there
+may be several such fixed-point values or none.  A few examples should help
+explain the subtleties.
+
+\begin{example}
+Consider the combinator
+
+{{c : bool * 1 <-> bool * 1}}
+
+{{c = id}}
+
+\noindent Using {{trace*}}, we construct the
+combinator {{trace* c : 1 <-> 1}}. Applying this to 
+{{() : 1}} requires us to find a value {{b : bool}} such that 
+{{(b,()) ~~c~~ (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
+has two values, there are two values {{false}} and {{true}} that satisfy the 
+constraint. In other words, expect that {{() ~~(trace* c)~~() }}.
+\end{example}
+
+\begin{example}
+\label{ch3:ex:annihilate}
+Consider a small variation on the combinator {{c}} above:
+
+{{c : bool * 1 <-> bool * 1}}
+
+{{c = swap+ (*) id }}
+
+\noindent which negates the boolean component of its incoming
+pair. Using the multiplicative trace operator, we can construct the
+combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
+combinator to {{() : 1}} requires us to find a value {{b : bool}} such
+that {{(b,()) ~~c~~ (b,())}} which is impossible since {{c}} negates
+{{b}}, i.e. {{c}} has no fixed points. Operationally, the evaluation of
+such a combinator should produce no value (if it even terminates).
+\end{example}
+
+With fractionals, the multiplicative trace operator becomes expressible:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
+% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
+%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite
+
+Let's use the semantics to calculate the result of applying
+{{trace*_{bool} (not (*) id)}} to {{false}}. We start
+with the set {{ {[ false ]} }} which gets transformed as follows:
+%subcode{opsem} include main
+%! columnStyle = ll
+% {[ ((),false) ]}  & (uniti)
+% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
+% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
+% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
+% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
+% emptyset & (eps*_{bool} * id)
+
+%% \jacques{in the current paper, this seems to no longer make sense}
+%% Yes I know, but I think we should leave it to explain the problem in Sec. 5
+%% and motivate the different semantics
+
+\begin{example}[Annihilation]
+\label{ch3:ex;annihilation}
+Consider the circuit below:
+\begin{center}
+\scalebox{1.2}{
+  \includegraphics{diagrams/not_trace.pdf}
+}
+\end{center}
+As motivated earlier, it is clearly impossible for the constraint implied by
+{{trace*}} to be satisfied as a boolean value {{v}} can never be the same as
+its negation. Thus there is no incoming value of type {{b}} that can ever go
+through the circuit. Indeed, if we call the circuit above {{c}}, then we can
+calculate that {{c}} does not relate {{v}} to anything. We call such a
+circuit an annihilation circuit.
+\end{example}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{(Finite) Relational Programming}
+\label{ch3:sec:lp}
+
+Relational programming leverages set-theoretic relations and their
+composition to express computational tasks in a declarative way. 
+% For instance, consider this example:
+% {{parent = {[ (A,B), (B,C), (A,D) ]} }}
+% {{grandparent = parent (o) parent}}
+% \noindent The example defines a relation {{parent}} specified using a
+% set of tuples and another relation {{grandparent}} specified using the
+% composition of two parent relations. If we wrote this example in a
+% relational language (e.g., Prolog) and we executed the query
+% {{grandparent(A)}}, we would get the answer {{ {[C]} }}.
+It turns out that with the addition of the multiplicative trace, and
+the move to relations motivated in the previous section, we can
+express relational programming. We illustrate the idea using a
+complete example.
+
+Consider the relation {{R}} on booleans given by:
+
+{{ {[(false,false), (false,true), (true,false) ]}. }}
+
+We show how to define a combinator {{c_R}} whose semantics is such that:
+
+%subcode{proof} include main
+%@ ~
+%@@ false ~~c_R~~ false
+%
+%@ ~
+%@@ false ~~c_R~~ true
+%
+%@ ~
+%@@ true ~~c_R~~ false
+
+The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
+for some type {{a}} such that {{trace* cInner}} is the desired combinator
+{{c_R}}.  A little experimentation shows that we get the desired behavior if
+as follows:
+% %subcode{proof} include main
+% %@ ~
+% %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
+% %----
+% %@ ~
+% %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
+% %----
+% %@ ~
+% %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
+% %----
+% %@ ~
+% %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
+% %----
+% %@ ~
+% %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
+% %----
+% %@ ~
+% %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
+% %----
+% %@ ~
+% %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
+% %----
+% %@ ~
+% %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
+% In the first three lines, the first argument is a fixed point and hence the
+% multiplicative trace would map the second input to the second output
+% producing the desired relation. In the remaining five cases, the first
+% argument is not a fixed point and hence all these cases would be rejected as
+% solutions to the constraint imposed by the multiplicative trace. It simply
+% remains to find the actual realization of {{cInner}} that would produce the
+% behavior above.
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
+% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assoct* (;) swap* (;) 
+%        &&  toffoli (;) ((swap* (;) cnot (;) swap*) (*) id) (;)
+%        &&  toffoli (;) (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
+%        &&  toffoli (;) (cnot (*) id))
+%     
+% c_R &:& bool <-> bool
+% c_R &=& trace* cInner
+
+% The above example should convince the reader that the language {{Pi}}
+% with multiplicative trace is expressive enough to model finite relational
+% programming. 
+% %%%%%%%%%%%%%%%
+% \subsection{Solving Constraints}
+% \label{ch3:sec:constraints}
+
+% A large class of constraint satisfaction problems can be expressed
+% using multiplicative traces. We illustrate the main ideas with the
+% implementation of a SAT solver.
+
+% In the usual setting, an instance of SAT is a function~{{f}} which,
+% when given some boolean inputs, returns {{true}} or {{false}}. The
+% function returns {{true}} when the inputs satisfy the constraints
+% imposed by the structure of {{f}} and a solution to the SAT problem is
+% the set of all inputs on which {{f}} produces {{true}}. The basic idea
+% of our construction is to generalize the annihilation circuit from
+% Ex.~\ref{ch3:ex;annihilation} to only annihilate values that fail to
+% satisfy the constraints represented by the SAT instance {{f}}. To
+% achieve this goal, we must however deal with several important
+% details.
+
+% First, because we are in a reversible world, our instance of SAT must be
+% expressed as an isomorphism: this is easily achieved by the construction
+% which embeds any boolean function {{f}} into a reversible one {{iso_f}} with
+% a larger domain and range.  Given such a reversible function {{iso_f}} which
+% represents a SAT instance, we first construct the circuit below:
+
+% \begin{center}
+% \scalebox{1.2}{
+%   \includegraphics{diagrams/sat2.pdf}
+% }
+% \end{center}
+
+% As shown in the circuit, the reversible SAT instance {{iso_f}} takes
+% two sets of values and produces two outputs. The incoming values
+% labeled \textsf{inputs} are the inputs we need to test for
+% satisfiability. The other incoming values labeled \textsf{heap} are
+% the additional inputs needed to embed the original SAT instance {{f}}
+% into a reversible function. If these \textsf{heap} values are all
+% initialized to {{false}}, the output wire \textsf{satisfied?}
+% corresponds to the output that {{f}} would have produced on
+% \textsf{inputs}. The other outputs labeled \textsf{garbage} are not
+% needed for their own sake but they are important because they are used
+% as inputs to the adjoint of {{iso_f}} to reproduce the inputs exactly,
+% in anticipation of closing the loop with {{trace*}}.
+
+% To summarize, the top half of the circuit is the identity function
+% except that we have also managed to produce a boolean wire labeled
+% \textsf{satisfied?} that tells us if the inputs satisfy the desired
+% constraints. We can take this boolean value and use it to decide
+% whether to negate the bottom wire (labeled \textsf{control
+%   wire}). Specifically, if the inputs do \emph{not} satisfy {{f}}, the
+% control wire is negated. The last wire labeled \textsf{heap control
+%   wire} is negated if the heap values do not have the right initial
+% values, i.e., are not all {{false}}.
+
+% Let us call the above construction {{sat_f}}. If we now close the loop
+% using {{trace*}}, two things should happen:
+% \begin{itemize}
+% \item configurations in which the \textsf{heap} values are not all
+%   {{false}} will be annihilated;
+% \item configurations in which the \textsf{inputs} do not satisfy {{f}}
+%   will cause the \textsf{satisfied?} wire to be negated and hence will
+%   also be annihilated.
+% \end{itemize}
+% In other words, the only configurations that will survive are the ones in
+% which the \textsf{inputs} satisfy {{f}}. We simply need to arrange to
+% \emph{clone} these values and produce them as the output of the whole
+% circuit. The final construction is therefore:
+
+% \begin{center}
+% \scalebox{1.5}{
+%   \includegraphics{diagrams/sat3.pdf}
+% }
+% \end{center}
+
+% To make the previous discussion concrete, we present a small, but
+% complete, example. In our example, the SAT instance {{f}} is tiny: it
+% takes two inputs. This function is embedded into a reversible function
+% {{iso_f}} of type 
+% {{((bool * bool) * bool) <-> ((bool * bool) * bool)}} where the last
+% input represents the heap and the first two outputs represent the garbage. 
+% The realization of {{sat_f}} given below is parametrized by such 
+% a function {{iso_f}}. The inputs to {{sat_f}} are 
+% \textsf{heap control}, \textsf{control}, \textsf{heap}, \textsf{input-1}, and 
+% \textsf{input-2}. Its operation is simple: if the \textsf{heap} is {{true}}, 
+% \textsf{heap control} is negated, and if the last output of {{iso_f}} 
+% is {{false}}, \textsf{control} is negated:
+
+% %subcode{opsem} include main
+% %! columnStyle = rcl
+% % sat_f &:& ((((bool * bool) * bool) * bool) * bool) <-> ((((bool * bool) * bool) * bool) * bool)
+% % sat_f &=& ((swap* (*) id) (*) id) (;) 
+% % && ((assocl* (*) id) (*) id) (;) 
+% % &&   (((cnot (*) id) (*) id) (*) id) (;) 
+% % &&   assocr* (;) 
+% % &&   (assocr* (*) id) (;) 
+% % &&   (swap* (*) id) (;) 
+% % &&   assocr* (;) 
+% % &&   (id (*) assocl*) (;) 
+% % &&   (id (*) isof) (;) 
+% % &&   swap* (;) 
+% % &&   assocr* (;) 
+% % &&   (id (*) (id (*) swap*)) (;) 
+% % &&   (id (*) assocl*) (;) 
+% % &&   (id (*) ((inot (*) id) (*) id)) (;)
+% % &&   (id (*) (cnot (*) id)) (;) 
+% % &&   (id (*) ((inot (*) id) (*) id)) (;)
+% % &&   (id (*) assocr*) (;) 
+% % &&   (id (*) (id (*) swap*)) (;) 
+% % &&   assocl* (;)
+% % &&   swap* (;)
+% % &&   (id (*) Sym isof) (;) 
+% % &&   (id (*) assocr*) (;) 
+% % &&   assocl* (;) 
+% % &&   assocl*
+
+% Given the construction of {{sat_f}} we can build the full solver as
+% follows. The overall input is the cloning heap. The combinator given
+% to {{trace*}} takes the cloning heap and the inputs flowing around the
+% loop and produces two copies of these inputs. One copy is produced as
+% the overall output and another is fed back around the loop.
+
+% %subcode{opsem} include main
+% %! columnStyle = rcl
+% % solve_f &:& bool * bool <-> bool * Bool
+% % solve_f &=& trace* (
+% % && (assocr* (*) id) (;) 
+% % && assocr* (;)
+% % && (id (*) swap*) (;)
+% % && assocl* (;)
+% % && (assocr* (*) id) (;)
+% % && (clone2 (*) id) (;)
+% % && swap* (;)
+% % && (swap* (*) id) (;)
+% % && ((swap* (*) id) (*) id) (;)
+% % && assocl* (;)
+% % && assocl* (;)
+% % && (sat_f (*) id) (;)
+% % && (assocr* (*) id) (;)
+% % && (swap* (*) id) (;) 
+% % && ((id (*) swap*) (*) id) (;)
+% % && (assocl* (*) id) (;)
+% % && ((id (*) swap*) (*) id))
+
+% We can test our {{solve_f}} combinator using several SAT
+% instances. Here are two possible instances. The first instance is
+% satisfied by {{(false,false)}} and the
+% second is satisfied by {{(false,true)}} and {{(true,true)}}.
+
+% %subcode{opsem} include main
+% %! columnStyle = rcl
+% % iso_{f_1} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
+% % iso_{f_1} &=& (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
+% % && (((swap+ (*) id) (*) id) (;) toffoli (;) ((swap+ (*) id) (*) id)) (;) 
+% % && (id (*) swap+)
+% % 
+% % iso_{f_2} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
+% % iso_{f_2} &=& toffoli
+
+% It can indeed be verified using the semantics that the {{solve_f}}
+% combinators instantiated with the SAT instances produce the expected
+% results.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion}
+\label{sec:cat}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
+the relation:
+
+{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
+
+give some intuition; example with eta/epsilon
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\bibliographystyle{splncs03} 
+\bibliography{cites}
+\end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\appendix
+\section{Categorical Background}
+\label{app:cat}
+
+We recall that a \emph{symmetric monoidal category} is a category together
+with a bifunctor $\otimes$, a distinguished object $I$, and natural
+isomorphisms $\alpha_{A,B,C} : (A \otimes B) \otimes C \rightarrow A \otimes
+(B \otimes C)$, $\lambda_A : A \rightarrow I \otimes A$, and $\sigma_{A,B} :
+A \otimes B \rightarrow B \otimes A$ subject to standard coherence
+conditions~\cite{nla.cat-vn1051288}. Following common practice
+(e.g.,~\cite{Selinger:2007:DCC:1229185.1229207}), we write $\rho_A =
+\sigma_{I,A} \circ \lambda_A : A \rightarrow A \otimes I$.
+
+A \emph{compact closed category} is a symmetric monoidal category where each
+object $A$ is assigned a dual object $A^*$, together with a unit map $\eta_A
+: I \rightarrow A^* \otimes A$ and a counit map $\epsilon_A : A \otimes A^*
+\rightarrow I$, such that:
+\[\begin{array}{rcl}
+\lambda_A^{-1} \circ (\epsilon_A \otimes A) \circ \alpha_{A,A^*,A}^{-1} \circ (A \otimes \eta_A) \circ \rho_A &=& \mathit{id}_A \\
+\rho_{A^*}^{-1} \circ (A^* \otimes \epsilon_A) \circ \alpha_{A^*,A,A^*} \circ (\eta_A \otimes A^*) \circ \lambda_A &=& \mathit{id}_{A^*}
+\end{array}\]
+
+A \emph{dagger category} is a category together with an involutive,
+identity-on-objects, contravariant functor $\dagger$. Concretely, this means
+that to every morphism $F : A \rightarrow B$, one associates a morphism
+$f^{\dagger} : B \rightarrow A$, called the \emph{adjoint} of $f$, such that
+for all $f : A \rightarrow B$ and $g : B \rightarrow C$, we have:
+\[\begin{array}{rcl}
+\mathit{id}^\dagger_A &=& \mathit{id}_A \\
+(g \circ f)^\dagger &=& f^\dagger \circ g^\dagger \\
+(f^\dagger)^\dagger &=& f
+\end{array}\]
+
+A \emph{dagger symmetric monoidal category} is a symmetric monoidal category
+with a dagger structure such that the contravariant functor $\dagger$
+coherently preserves the symmetric monoidal structure. Concretely, this
+requirement means that for all $f : A \rightarrow B$ and $g : C \rightarrow
+D$, we have:
+\[\begin{array}{rcl}
+(f \otimes g)^\dagger &=& f^\dagger \otimes g^\dagger \\
+\alpha^\dagger_{A,B,C} &=& \alpha^{-1}_{A,B,C} \\
+\lambda^\dagger_A &=& \lambda^{-1}_A \\
+\sigma^\dagger_{A,B} &=& \sigma^{-1}{A,B}
+\end{array}\]
+
+\begin{definition}[Dagger Compact Closed Category]
+\label{def:cat}
+A \emph{dagger compact closed category} is a dagger symmetric monoidal
+category that is also compact closed and such that for all $A$:
+\[
+\eta_A = \sigma_{A,A^*} \circ \epsilon^\dagger_A
+\]
+\end{definition}
+
+\todo{biproducts}
+
+\begin{verbatim}
+homotopy equivalence???
+
+axioms for field or meadow; actually we probably a semifield
+
+definition of logical reversibility with relations
+
+definition of information preservation in the case of relations; fanout as an
+example which we can write
+
+explain in detail the size of 2 * 1/2; it should have exactly one element;
+there are two elemens in 2 but the 1/2 identifies them somehow; be precise
+
+we can get the empty relation (eta ; id * swap; eps) what does that mean?
+
+connection to vector spaces; projectors; inner product. Given dual vectors,
+we can have a standalone |0><0| which is a piece of a function or a projector
+if viewed by itself; we can also have |v><v'| which produces the scalars YES
+or NO. We have an isomorphism between matrices and vectors by viewing 
+|0><0|+|1><1| as |00>+|11>
+
+Perhaps if restrict the ``top-level'' to non-fractional types, we can never
+observe a function built from from fractionals; we must apply it and in that
+case, the system overall behaves like Pi. (no empty relation for example)???
+Idea suggested by Zach is to change eps to (1/a) * a * a <-> a
+
+The code implementing the 16 relations... all 16 relations have the same
+information content as () ??? So we should treat (1/bool * bool) as opaque
+things; we can only extract information from them by applying them.
+
+Include dneg, name (inv cnot) as example of programs which when fed
+non-fractional values produce fractional values
+
+Also include recipT which shows that 1/ distributes over products and tens
+which shows that we can manipulate fractional types in interesting ways
+
+Using relational semantics is good because it's clear; but it's also bad
+unless we spend time to explain the operational model
+
+Explain 'name' in detail; it IS a faithful representation of the input
+relation as a value; we don't get to see the full relation though; we
+interact with it by giving it inputs and observing the related ouptut come
+out. If we make two copies and use each in a different context (applying each
+to a different value, we can ``see'' that it's really a full relation and not
+just one branch.)
+
+total functions have size 1; partial functions have fractional sizes!!!
+
+denotation of type in Pi is a set; in Pi/ we get a set with an equivalence
+relation (Id by default) but fractionals introduce first-class equivalence
+relations that can be used to restrict the sets.
+
+---
+
+Having re-read Baez-Dolan, I now feel better.
+
+On 13-02-08 08:53 AM, Amr Sabry wrote:
+> I think all the observations in this email are due to the fact that we
+> haven't formalized the size of fractionals using homotopy equivalence. I
+> started doing this but abandoned it for now but the bottom line is that
+> the size of b * 1/b is 1.
+
+Again, agree - more deeply so now.  In more detail, my current thinking:
+- v \in b has size 1.  So |b| counts how many elements it has, whenever 
+b is a type of Pi.  The sum and product rules apply.
+- for b * 1/b to have size 1, then 1/b must have size 1/|b| (assuming 
+these are independent types, which we have been assuming all along)
+
+Attempt #1
+- there are |b| elements in 1/b, and if size is additive, ( 1/|b| = 
+sum_{1/v \in 1/b} |1/v| ) should hold; since |1/v| should be constant, 
+this sum is |b|*c, so c = 1/|b|^2.  Weird.
+
+Let us look at a single 'function' in bool -o bool, namely the identity 
+function.  It should be 'the same' as { (1/true, true), (1/false, false) 
+}.  The cardinality of that set is (1/(2^2)*1 + 1/(2^2)*1) = 1/4 + 1/4 = 
+1/2.   From weird to probably-wrong.
+
+Attempt #2
+- the elements of 1/b are not distinguishable from each other (i.e. they 
+are all isomorphic).  So each element 1/v has b automorphisms, so |1/b| 
+= sum_{aut classes of 1/v \in 1/b} 1/|aut 1/v| = sum_{singleton} 1/|b| = 
+1/|b|.  [as per p.14 of Baez-Dolan]
+
+The size of the set { (1/true, true), (1/false, false) } is now 1/2*1 + 
+1/2*1 = 1.  So this 'set' indeed represents a single "concept".
+
+Note that in attempt #2, we get a new phenomenon.  Consider the 
+(partial) relation {(1/true, false)}, also of type bool -o bool.  It has 
+size 1/2 !  So, if I have not erred somewhere, only total functions have 
+size 1.  Which, I must admit, I rather like.  It's not that partial 
+functions are outlawed, they just don't pull as much weight [pun intended].
+
+So 1/b is very much like a collection of |b| constraints, all of which 
+are "the same" (externally).  While they may have internal structure, it 
+is not visible to any of our combinators, so up to iso, there is only a 
+single element of 1/b, of size 1/|b|.
+
+Yet another way to look at it: let's assume we are looking at FinSet_0, 
+where in fact our universe U has finitely (say N) things in it.  Then 
+the singleton set 1 is actually the isomorphism class of {u_1, ..., 
+u_N}.  There are N of them, but they have N automorphisms in a single 
+orbit, so the quotient has size 1 (i.e. |S| = N but |G| = N too, so |S 
+// G| = 1 ).
+
+So an element of 1/b is closer to an action of b on b; we call it a 
+constraint, or a pattern-match.  I think thinking of it as 'actions' is 
+definitely the right way to go, as we will be able to have different 
+kinds of actions on a type b that just those that come from the elements.
+
+> I don't know how much of a priority this is but if it is we should try
+> to work it out in detail. The good and bad news is that we will have a
+> model much richer and much more complicated than relations. My thought
+> when writing sec. 3 was that we would use the simple but slightly
+> inadequate model of relations (which for one thing confuses v and 1/v)
+> and simply mention that the right abstraction is that of 'biproduct
+> dagger compact closed categories' which presumably would include the
+> richer model.
+
+I am as convinced that the right model is 'biproduct dagger compact 
+closed categories' as I am that the right model for the type level is 
+'categorified field'.
+
+\end{verbatim}
+
 \begin{comment}
-% We are not doing groupoids in this paper
 Each type in {{PiTF}} is interpreted as a \emph{groupoid}. A groupoid can be
@@
 \end{comment}
-In the following, we use {{bool}} as an abbreviation of {{1+1}} 
-with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
-abbreviation for {{right ()}}.
-
-%%%%%%%%%%%%
-\subsection{First-Class Relations}
-
-As motivated, with fractional types, we can define a value that denotes a
-first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
-pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
-and a value {{v2 : b2}}. In other words, it corresponds to a function or
-relation which when given matched with a value {{v1 : b1}} ``releases'' the
-value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:
-
-%subcode{bnf} include main
-%  b1 :-* b2 &::=& 1/b1 * b2
-
-\noindent which suggests a function-like behavior for the pair of a fractional value
-and a regular value. What is remarkable is that we can almost trivially turn
-any combinator {{c : b1 <-> b2}} into a (constant) value of 
-type {{b1 :-* b2}} as shown below:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
-% name c &=& eta*_{b1} (;) (id (*) c)
-
-As a simple example, consider combinator {{not}} denoting boolean negation.
-It is straightforward to check that the combinator 
-{{name not : 1 <-> (bool :-* bool)}} denotes the following
-relation:
-
-%subcode{proof} include main
-%@ ~
-%@@ () ~~(name not)~~ (1/false,true)
-%
-%@ ~
-%@@ () ~~(name not)~~ (1/true,false)
-
-\noindent The relation {{name not}} can therefore be used to convert any
-{{false}} to {{true}} and vice-versa. We formalize this ``apply'' combinator
-below.
-
-%%%%%%%%%%%%
-\subsection{Higher-Order Relations}
-
-We are now a small step to implementing various higher-order
-combinators that manipulate functions or relations. In particular, we
-can \emph{apply} and \emph{compose} values representing relations as
-shown below:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% apply &:& (b1 :-* b2) * b1 <-> b2
-% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
-
-\noindent At the type level, the computation proceeds as follows:
-%subcode{bnf} include main
-%  (b1 :-* b2) * b1 &=& (1/b1 * b2) * b1 & 
-%     &<->& b1 * (1/b1 * b2) & (swap*)
-%     &<->& (b1 * 1/b1) * b2) & (assocl*)
-%     &<->& (1/b1 * b1) * b2) & (swap* (*) id)
-%     &<->& 1 * b2 & (eps*_{b1} (*) id)
-%     &<->& b2 & (unite)
-
-\noindent Intuitively, we simply match the incoming argument of type {{b1}} with
-the constraint encoded by the function. If they match, they cancel
-each other and the value of type {{b2}} is exposed with no
-constraints. Otherwise, the result is undefined.
-
-\noindent A flipped variant is also useful:
-%subcode{opsem} include main
-%! columnStyle = rcl
-% apply' &:& b1 * (b1 :-* b2) <-> b2
-% apply' &=& assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
-
-\noindent Function or relation composition is now straightforward:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
-% compose &=& assocr* (;) (id (*) apply')
-
-\noindent Again, at the type level, the computation proceeds as follows:
-
-%subcode{bnf} include main
-%  (b1 :-* b2) * (b2 :-* b3) &=& (1/b1 * b2) * (1/b2 * b3) & 
-%     &<->& (1/b1 * (b2 * (1/b2 * b3)) & (assocr*)
-%     &<->& (1/b1 * b3) & (id * apply')
-%     &=& b1 :-* b3
-
-%%%%%%%%%%%%
-\subsection{Duality}
-
-The final remarkable property of fractional types is that they
-are involutive.  In particular, we have:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% doubleDiv &:& b <-> 1/(1/b)
-% doubleDiv &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
-
-\noindent At the type level, this computation proceeds as follows:
-
-%subcode{bnf} include main
-%  b &<->& (1 * b) & (uniti)
-%    &<->& (1/(1/b) * 1/b) * b & (eta*_{1/b} (*) id)
-%    &<->& 1/(1/b) * (1/b * b) & (assocr*)
-%    &<->& 1/(1/b) * 1 & (id (*) eps*_b)
-%    &<->& 1 * 1/(1/b) & (swap*)
-%    &<->& 1/(1/b) & (unite)
-
-%%%%%%%%%%%%
-\subsection{Feedback, Iteration, and Trace Operators}
-
-Mathematically speaking, recursion and iteration can be expressed using
-categorical trace
-operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
-language like {{Pi}} there are two natural families of trace operators
-that can be defined, an additive family (explored in detail in our previous
-work~\cite{rc2011}) and a multiplicative family (which is the focus of this
-section).
-
-In this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
-want to cancel the common type {{b1}}, to produce a new combinator
-{{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
-given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
-consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
-The needed value {{v1 : b1}} cannot be arbitrary: it must be such that is
-is the value produced as the first component of the result.  In general, there
-may be several such fixed-point values or none.  A few examples should help
-explain the subtleties.
-
-\begin{example}
-Consider the combinator
-
-{{c : bool * 1 <-> bool * 1}}
-
-{{c = id}}
-
-\noindent Using {{trace*}}, we construct the
-combinator {{trace* c : 1 <-> 1}}. Applying this to 
-{{() : 1}} requires us to find a value {{b : bool}} such that 
-{{(b,()) ~~c~~ (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
-has two values, there are two values {{false}} and {{true}} that satisfy the 
-constraint. In other words, expect that {{() ~~(trace* c)~~() }}.
-\end{example}
-
-\begin{example}
-\label{ch3:ex:annihilate}
-Consider a small variation on the combinator {{c}} above:
-
-{{c : bool * 1 <-> bool * 1}}
-
-{{c = swap+ (*) id }}
-
-\noindent which negates the boolean component of its incoming
-pair. Using the multiplicative trace operator, we can construct the
-combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
-combinator to {{() : 1}} requires us to find a value {{b : bool}} such
-that {{(b,()) ~~c~~ (b,())}} which is impossible since {{c}} negates
-{{b}}, i.e. {{c}} has no fixed points. Operationally, the evaluation of
-such a combinator should produce no value (if it even terminates).
-\end{example}
-
-With fractionals, the multiplicative trace operator becomes expressible:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
-% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
-%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite
-
-Let's use the semantics to calculate the result of applying
-{{trace*_{bool} (not (*) id)}} to {{false}}. We start
-with the set {{ {[ false ]} }} which gets transformed as follows:
-%subcode{opsem} include main
-%! columnStyle = ll
-% {[ ((),false) ]}  & (uniti)
-% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
-% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
-% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
-% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
-% emptyset & (eps*_{bool} * id)
-
-\jacques{in the current paper, this seems to no longer make sense}
-\begin{example}[Annihilation]
-\label{ch3:ex;annihilation}
-Consider the circuit below:
-\begin{center}
-\scalebox{1.2}{
-  \includegraphics{diagrams/not_trace.pdf}
-}
-\end{center}
-As motivated earlier, it is clearly impossible for the constraint implied by
-{{trace*}} to be satisfied as a boolean value {{v}} can never be the same as
-its negation. Thus there is no incoming value of type {{b}} that can ever go
-through the circuit. Indeed, if we call the circuit above {{c}}, then we can
-calculate that {{c}} does not relate {{v}} to anything. We call such a
-circuit an annihilation circuit.
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{(Finite) Relational Programming}
-\label{ch3:sec:lp}
-
-Relational programming leverages set-theoretic relations and their
-composition to express computational tasks in a declarative way. For
-instance, consider this example:
-
-{{parent = {[ (A,B), (B,C), (A,D) ]} }}
-
-{{grandparent = parent (o) parent}}
-
-\noindent The example defines a relation {{parent}} specified using a
-set of tuples and another relation {{grandparent}} specified using the
-composition of two parent relations. If we wrote this example in a
-relational language (e.g., Prolog) and we executed the query
-{{grandparent(A)}}, we would get the answer {{ {[C]} }}.
-
-It turns out that with the addition of the multiplicative trace, and
-the move to relations motivated in the previous section, we can
-express relational programming. We illustrate the idea using a
-complete example.
-
-Consider the relation {{R}} on booleans given by:
-
-{{ {[(false,false), (false,true), (true,false) ]}. }}
-
-We show how to define a combinator {{c_R}} whose semantics is such that:
-
-%subcode{proof} include main
-%@ ~
-%@@ false ~~c_R~~ false
-%
-%@ ~
-%@@ false ~~c_R~~ true
-%
-%@ ~
-%@@ true ~~c_R~~ false
-
-The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
-for some type {{a}} such that {{trace* cInner}} is the desired combinator
-{{c_R}}. A little experimentation shows that we get the desired behavior
-if {{cInner}} is chosen to behave as follows:
-
-%subcode{proof} include main
-%@ ~
-%@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
-%----
-%@ ~
-%@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
-%----
-%@ ~
-%@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
-%----
-%@ ~
-%@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
-%----
-%@ ~
-%@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
-%----
-%@ ~
-%@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
-%----
-%@ ~
-%@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
-%----
-%@ ~
-%@@ ((true,true),true) ~~cInner~~ ((false,false),true)
-
-In the first three lines, the first argument is a fixed point and hence the
-multiplicative trace would map the second input to the second output
-producing the desired relation. In the remaining five cases, the first
-argument is not a fixed point and hence all these cases would be rejected as
-solutions to the constraint imposed by the multiplicative trace. It simply
-remains to find the actual realization of {{cInner}} that would produce the
-behavior above. One can infer the following implementation:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
-% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assoct* (;) swap* (;) 
-%        &&  toffoli (;) ((swap* (;) cnot (;) swap*) (*) id) (;)
-%        &&  toffoli (;) (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
-%        &&  toffoli (;) (cnot (*) id))
-%     
-% c_R &:& bool <-> bool
-% c_R &=& trace* cInner
-
-The above example should convince the reader that the language {{Pi}}
-with multiplicative trace is expressive enough to model finite relational
-programming. We will not prove this claim. We develop an even more
-substantial example: a SAT solver.
-
-%%%%%%%%%%%%%%%
-\subsection{Solving Constraints}
-\label{ch3:sec:constraints}
-
-A large class of constraint satisfaction problems can be expressed
-using multiplicative traces. We illustrate the main ideas with the
-implementation of a SAT solver.
-
-In the usual setting, an instance of SAT is a function~{{f}} which,
-when given some boolean inputs, returns {{true}} or {{false}}. The
-function returns {{true}} when the inputs satisfy the constraints
-imposed by the structure of {{f}} and a solution to the SAT problem is
-the set of all inputs on which {{f}} produces {{true}}. The basic idea
-of our construction is to generalize the annihilation circuit from
-Ex.~\ref{ch3:ex;annihilation} to only annihilate values that fail to
-satisfy the constraints represented by the SAT instance {{f}}. To
-achieve this goal, we must however deal with several important
-details.
-
-First, because we are in a reversible world, our instance of SAT must be
-expressed as an isomorphism: this is easily achieved by the construction
-which embeds any boolean function {{f}} into a reversible one {{iso_f}} with
-a larger domain and range.  Given such a reversible function {{iso_f}} which
-represents a SAT instance, we first construct the circuit below:
-
-\begin{center}
-\scalebox{1.2}{
-  \includegraphics{diagrams/sat2.pdf}
-}
-\end{center}
-
-As shown in the circuit, the reversible SAT instance {{iso_f}} takes
-two sets of values and produces two outputs. The incoming values
-labeled \textsf{inputs} are the inputs we need to test for
-satisfiability. The other incoming values labeled \textsf{heap} are
-the additional inputs needed to embed the original SAT instance {{f}}
-into a reversible function. If these \textsf{heap} values are all
-initialized to {{false}}, the output wire \textsf{satisfied?}
-corresponds to the output that {{f}} would have produced on
-\textsf{inputs}. The other outputs labeled \textsf{garbage} are not
-needed for their own sake but they are important because they are used
-as inputs to the adjoint of {{iso_f}} to reproduce the inputs exactly,
-in anticipation of closing the loop with {{trace*}}.
-
-To summarize, the top half of the circuit is the identity function
-except that we have also managed to produce a boolean wire labeled
-\textsf{satisfied?} that tells us if the inputs satisfy the desired
-constraints. We can take this boolean value and use it to decide
-whether to negate the bottom wire (labeled \textsf{control
-  wire}). Specifically, if the inputs do \emph{not} satisfy {{f}}, the
-control wire is negated. The last wire labeled \textsf{heap control
-  wire} is negated if the heap values do not have the right initial
-values, i.e., are not all {{false}}.
-
-Let us call the above construction {{sat_f}}. If we now close the loop
-using {{trace*}}, two things should happen:
-\begin{itemize}
-\item configurations in which the \textsf{heap} values are not all
-  {{false}} will be annihilated;
-\item configurations in which the \textsf{inputs} do not satisfy {{f}}
-  will cause the \textsf{satisfied?} wire to be negated and hence will
-  also be annihilated.
-\end{itemize}
-In other words, the only configurations that will survive are the ones in
-which the \textsf{inputs} satisfy {{f}}. We simply need to arrange to
-\emph{clone} these values and produce them as the output of the whole
-circuit. The final construction is therefore:
-
-\begin{center}
-\scalebox{1.5}{
-  \includegraphics{diagrams/sat3.pdf}
-}
-\end{center}
-
-To make the previous discussion concrete, we present a small, but
-complete, example. In our example, the SAT instance {{f}} is tiny: it
-takes two inputs. This function is embedded into a reversible function
-{{iso_f}} of type 
-{{((bool * bool) * bool) <-> ((bool * bool) * bool)}} where the last
-input represents the heap and the first two outputs represent the garbage. 
-The realization of {{sat_f}} given below is parametrized by such 
-a function {{iso_f}}. The inputs to {{sat_f}} are 
-\textsf{heap control}, \textsf{control}, \textsf{heap}, \textsf{input-1}, and 
-\textsf{input-2}. Its operation is simple: if the \textsf{heap} is {{true}}, 
-\textsf{heap control} is negated, and if the last output of {{iso_f}} 
-is {{false}}, \textsf{control} is negated:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% sat_f &:& ((((bool * bool) * bool) * bool) * bool) <-> ((((bool * bool) * bool) * bool) * bool)
-% sat_f &=& ((swap* (*) id) (*) id) (;) 
-% && ((assocl* (*) id) (*) id) (;) 
-% &&   (((cnot (*) id) (*) id) (*) id) (;) 
-% &&   assocr* (;) 
-% &&   (assocr* (*) id) (;) 
-% &&   (swap* (*) id) (;) 
-% &&   assocr* (;) 
-% &&   (id (*) assocl*) (;) 
-% &&   (id (*) isof) (;) 
-% &&   swap* (;) 
-% &&   assocr* (;) 
-% &&   (id (*) (id (*) swap*)) (;) 
-% &&   (id (*) assocl*) (;) 
-% &&   (id (*) ((inot (*) id) (*) id)) (;)
-% &&   (id (*) (cnot (*) id)) (;) 
-% &&   (id (*) ((inot (*) id) (*) id)) (;)
-% &&   (id (*) assocr*) (;) 
-% &&   (id (*) (id (*) swap*)) (;) 
-% &&   assocl* (;)
-% &&   swap* (;)
-% &&   (id (*) Sym isof) (;) 
-% &&   (id (*) assocr*) (;) 
-% &&   assocl* (;) 
-% &&   assocl*
-
-Given the construction of {{sat_f}} we can build the full solver as
-follows. The overall input is the cloning heap. The combinator given
-to {{trace*}} takes the cloning heap and the inputs flowing around the
-loop and produces two copies of these inputs. One copy is produced as
-the overall output and another is fed back around the loop.
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% solve_f &:& bool * bool <-> bool * Bool
-% solve_f &=& trace* (
-% && (assocr* (*) id) (;) 
-% && assocr* (;)
-% && (id (*) swap*) (;)
-% && assocl* (;)
-% && (assocr* (*) id) (;)
-% && (clone2 (*) id) (;)
-% && swap* (;)
-% && (swap* (*) id) (;)
-% && ((swap* (*) id) (*) id) (;)
-% && assocl* (;)
-% && assocl* (;)
-% && (sat_f (*) id) (;)
-% && (assocr* (*) id) (;)
-% && (swap* (*) id) (;) 
-% && ((id (*) swap*) (*) id) (;)
-% && (assocl* (*) id) (;)
-% && ((id (*) swap*) (*) id))
-
-We can test our {{solve_f}} combinator using several SAT
-instances. Here are two possible instances. The first instance is
-satisfied by {{(false,false)}} and the
-second is satisfied by {{(false,true)}} and {{(true,true)}}.
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% iso_{f_1} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
-% iso_{f_1} &=& (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
-% && (((swap+ (*) id) (*) id) (;) toffoli (;) ((swap+ (*) id) (*) id)) (;) 
-% && (id (*) swap+)
-% 
-% iso_{f_2} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
-% iso_{f_2} &=& toffoli
-
-It can indeed be verified using the semantics that the {{solve_f}}
-combinators instantiated with the SAT instances produce the expected
-results.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-\label{sec:cat}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
-Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
-the relation:
-
-{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
-
-give some intuition; example with eta/epsilon
-
-\begin{verbatim}
-homotopy equivalence???
-
-axioms for field or meadow; actually we probably a semifield
-
-definition of logical reversibility with relations
-
-definition of information preservation in the case of relations; fanout as an
-example which we can write
-
-explain in detail the size of 2 * 1/2; it should have exactly one element;
-there are two elemens in 2 but the 1/2 identifies them somehow; be precise
-
-we can get the empty relation (eta ; id * swap; eps) what does that mean?
-
-connection to vector spaces; projectors; inner product. Given dual vectors,
-we can have a standalone |0><0| which is a piece of a function or a projector
-if viewed by itself; we can also have |v><v'| which produces the scalars YES
-or NO. We have an isomorphism between matrices and vectors by viewing 
-|0><0|+|1><1| as |00>+|11>
-
-Perhaps if restrict the ``top-level'' to non-fractional types, we can never
-observe a function built from from fractionals; we must apply it and in that
-case, the system overall behaves like Pi. (no empty relation for example)???
-Idea suggested by Zach is to change eps to (1/a) * a * a <-> a
-
-The code implementing the 16 relations... all 16 relations have the same
-information content as () ??? So we should treat (1/bool * bool) as opaque
-things; we can only extract information from them by applying them.
-
-Include dneg, name (inv cnot) as example of programs which when fed
-non-fractional values produce fractional values
-
-Also include recipT which shows that 1/ distributes over products and tens
-which shows that we can manipulate fractional types in interesting ways
-
-Using relational semantics is good because it's clear; but it's also bad
-unless we spend time to explain the operational model
-
-Explain 'name' in detail; it IS a faithful representation of the input
-relation as a value; we don't get to see the full relation though; we
-interact with it by giving it inputs and observing the related ouptut come
-out. If we make two copies and use each in a different context (applying each
-to a different value, we can ``see'' that it's really a full relation and not
-just one branch.)
-
-total functions have size 1; partial functions have fractional sizes!!!
-
-denotation of type in Pi is a set; in Pi/ we get a set with an equivalence
-relation (Id by default) but fractionals introduce first-class equivalence
-relations that can be used to restrict the sets.
-
----
-
-Having re-read Baez-Dolan, I now feel better.
-
-On 13-02-08 08:53 AM, Amr Sabry wrote:
-> I think all the observations in this email are due to the fact that we
-> haven't formalized the size of fractionals using homotopy equivalence. I
-> started doing this but abandoned it for now but the bottom line is that
-> the size of b * 1/b is 1.
-
-Again, agree - more deeply so now.  In more detail, my current thinking:
-- v \in b has size 1.  So |b| counts how many elements it has, whenever 
-b is a type of Pi.  The sum and product rules apply.
-- for b * 1/b to have size 1, then 1/b must have size 1/|b| (assuming 
-these are independent types, which we have been assuming all along)
-
-Attempt #1
-- there are |b| elements in 1/b, and if size is additive, ( 1/|b| = 
-sum_{1/v \in 1/b} |1/v| ) should hold; since |1/v| should be constant, 
-this sum is |b|*c, so c = 1/|b|^2.  Weird.
-
-Let us look at a single 'function' in bool -o bool, namely the identity 
-function.  It should be 'the same' as { (1/true, true), (1/false, false) 
-}.  The cardinality of that set is (1/(2^2)*1 + 1/(2^2)*1) = 1/4 + 1/4 = 
-1/2.   From weird to probably-wrong.
-
-Attempt #2
-- the elements of 1/b are not distinguishable from each other (i.e. they 
-are all isomorphic).  So each element 1/v has b automorphisms, so |1/b| 
-= sum_{aut classes of 1/v \in 1/b} 1/|aut 1/v| = sum_{singleton} 1/|b| = 
-1/|b|.  [as per p.14 of Baez-Dolan]
-
-The size of the set { (1/true, true), (1/false, false) } is now 1/2*1 + 
-1/2*1 = 1.  So this 'set' indeed represents a single "concept".
-
-Note that in attempt #2, we get a new phenomenon.  Consider the 
-(partial) relation {(1/true, false)}, also of type bool -o bool.  It has 
-size 1/2 !  So, if I have not erred somewhere, only total functions have 
-size 1.  Which, I must admit, I rather like.  It's not that partial 
-functions are outlawed, they just don't pull as much weight [pun intended].
-
-So 1/b is very much like a collection of |b| constraints, all of which 
-are "the same" (externally).  While they may have internal structure, it 
-is not visible to any of our combinators, so up to iso, there is only a 
-single element of 1/b, of size 1/|b|.
-
-Yet another way to look at it: let's assume we are looking at FinSet_0, 
-where in fact our universe U has finitely (say N) things in it.  Then 
-the singleton set 1 is actually the isomorphism class of {u_1, ..., 
-u_N}.  There are N of them, but they have N automorphisms in a single 
-orbit, so the quotient has size 1 (i.e. |S| = N but |G| = N too, so |S 
-// G| = 1 ).
-
-So an element of 1/b is closer to an action of b on b; we call it a 
-constraint, or a pattern-match.  I think thinking of it as 'actions' is 
-definitely the right way to go, as we will be able to have different 
-kinds of actions on a type b that just those that come from the elements.
-
-> I don't know how much of a priority this is but if it is we should try
-> to work it out in detail. The good and bad news is that we will have a
-> model much richer and much more complicated than relations. My thought
-> when writing sec. 3 was that we would use the simple but slightly
-> inadequate model of relations (which for one thing confuses v and 1/v)
-> and simply mention that the right abstraction is that of 'biproduct
-> dagger compact closed categories' which presumably would include the
-> richer model.
-
-I am as convinced that the right model is 'biproduct dagger compact 
-closed categories' as I am that the right model for the type level is 
-'categorified field'.
-
-\end{verbatim}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliographystyle{splncs03} 
-\bibliography{cites}
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-\section{Categorical Background}
-\label{app:cat}
-
-We recall that a \emph{symmetric monoidal category} is a category together
-with a bifunctor $\otimes$, a distinguished object $I$, and natural
-isomorphisms $\alpha_{A,B,C} : (A \otimes B) \otimes C \rightarrow A \otimes
-(B \otimes C)$, $\lambda_A : A \rightarrow I \otimes A$, and $\sigma_{A,B} :
-A \otimes B \rightarrow B \otimes A$ subject to standard coherence
-conditions~\cite{nla.cat-vn1051288}. Following common practice
-(e.g.,~\cite{Selinger:2007:DCC:1229185.1229207}), we write $\rho_A =
-\sigma_{I,A} \circ \lambda_A : A \rightarrow A \otimes I$.
-
-A \emph{compact closed category} is a symmetric monoidal category where each
-object $A$ is assigned a dual object $A^*$, together with a unit map $\eta_A
-: I \rightarrow A^* \otimes A$ and a counit map $\epsilon_A : A \otimes A^*
-\rightarrow I$, such that:
-\[\begin{array}{rcl}
-\lambda_A^{-1} \circ (\epsilon_A \otimes A) \circ \alpha_{A,A^*,A}^{-1} \circ (A \otimes \eta_A) \circ \rho_A &=& \mathit{id}_A \\
-\rho_{A^*}^{-1} \circ (A^* \otimes \epsilon_A) \circ \alpha_{A^*,A,A^*} \circ (\eta_A \otimes A^*) \circ \lambda_A &=& \mathit{id}_{A^*}
-\end{array}\]
-
-A \emph{dagger category} is a category together with an involutive,
-identity-on-objects, contravariant functor $\dagger$. Concretely, this means
-that to every morphism $F : A \rightarrow B$, one associates a morphism
-$f^{\dagger} : B \rightarrow A$, called the \emph{adjoint} of $f$, such that
-for all $f : A \rightarrow B$ and $g : B \rightarrow C$, we have:
-\[\begin{array}{rcl}
-\mathit{id}^\dagger_A &=& \mathit{id}_A \\
-(g \circ f)^\dagger &=& f^\dagger \circ g^\dagger \\
-(f^\dagger)^\dagger &=& f
-\end{array}\]
-
-A \emph{dagger symmetric monoidal category} is a symmetric monoidal category
-with a dagger structure such that the contravariant functor $\dagger$
-coherently preserves the symmetric monoidal structure. Concretely, this
-requirement means that for all $f : A \rightarrow B$ and $g : C \rightarrow
-D$, we have:
-\[\begin{array}{rcl}
-(f \otimes g)^\dagger &=& f^\dagger \otimes g^\dagger \\
-\alpha^\dagger_{A,B,C} &=& \alpha^{-1}_{A,B,C} \\
-\lambda^\dagger_A &=& \lambda^{-1}_A \\
-\sigma^\dagger_{A,B} &=& \sigma^{-1}{A,B}
-\end{array}\]
-
-\begin{definition}[Dagger Compact Closed Category]
-\label{def:cat}
-A \emph{dagger compact closed category} is a dagger symmetric monoidal
-category that is also compact closed and such that for all $A$:
-\[
-\eta_A = \sigma_{A,A^*} \circ \epsilon^\dagger_A
-\]
-\end{definition}
-
-\todo{biproducts}

### 2013-02-20 frac.tex
intro to sec3

@@
 
-It is important to note that ``values'' and ``isomorphisms'' are 
-completely separate syntactic categories which do not intermix.  The 
-semantics of the language come when these are made to interact at
-the ``top level'', a new syntactic category:
+It is important to note that ``values'' and ``isomorphisms'' are completely
+separate syntactic categories which do not intermix. The semantics of the
+language come when these are made to interact at the ``top level'', a new
+syntactic category:
 %subcode{bnf} include main
@@
 want to model functions of any flavor as values, we need to differentiate
-between these notions.  Our idea, inspired by computational
-dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
+between these notions. We also clearly need to relax the strict separation
+between values and isomorphisms as the notion of ``higher-order'' requires
+that isomorphisms be represented as values. Our idea, inspired by
+computational dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
 polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
@@
 
-Syntactically, the extension with fractional types is simple.
-
-\begin{definition}[Syntax of {{PiTF}}]
-\label{def:PiT}
-We extend our previous definition with:
+The extension of the sets of types and values is simple:
+
 %subcode{bnf} include main
@@
 % values, v ::= ... | 1/v
-%
-% base iso ::= ... | eta*_b | eps*_b
-where:
+
+For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
+where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. There is
+however have some informal intuition about the \emph{meaning} of such
+fractional types and values. (See Sec.~\ref{sec:cat} for more discussion
+about this point.) From both the programming perspective and from the
+categorical perspective we expect the fractional types to satisfy the
+following isomorphism witnessed by two new combinators {{eta*_b}} and
+{{eps*_b}}:
+
 {{eta*_b : 1 <-> 1/b * b : eps*_b}}
-\end{definition}
-
-For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
-where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.  However,
-we think of a value like {{1/left ()}} as a \emph{first-class constraint}
-that can only be satisfied if it is matched with an actual value {{left ()}}
-or as representing the absence of some information (i.e., \emph{negative
-  information}), that can only be reconciliated by {{left ()}}. The
-combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission point which
-creates --- out of no information --- an equal amount of negative and
-positive information. Symmetrically, the combinator 
-{{eps*_b : 1/b * b <-> 1}} 
-matches up an equal amount of negative and positive information producing no
-residual information.  The best analogy is probably that of vectors and
-co-vectors, also know as column vectors and row vectors (respectively) where
-the multiplication of the row vector by the column vector is a scalar. In the
-finite dimensional case, they are isomorphic, but they are nevertheless quite
-different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
-since we are explicitly interested in \emph{isomorphisms}, we should not
-conflate negative-information values with positive-information values just
-because they happen to be isomorphic. (See Sec.~\ref{sec:cat} for a more
-thorough discussion about the meaning of fractional types and values in both an
-algebraic and categorical sense.) For now, we proceed with an operational 
-understanding of the language.
-
-The semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
-interpreted as a \emph{relation} between the sets denoted by {{b1}} and
-{{b2}} respectively.
+
+From a programming perspective, we think of a value {{1/v}} as a
+\emph{first-class constraint} that can only be satisfied if it is matched
+with an actual value {{v}} or as a \emph{pattern} representing the absence of
+some information of a specific shapre (i.e., \emph{negative information}),
+that can only be reconciliated by an actual value {{v}}. The combinator
+{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
+out of no information --- an equal amount of negative and positive
+information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
+up an equal amount of negative and positive information producing no residual
+information. From the categorical perspective, the extended language {{PiTF}}
+needs to have a semantics in a \emph{compact closed category} which are the
+categories introduced by Abramsky and
+Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
+Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
+computation. Two of the most relevant features of these categories for our
+purposes are that they support an adjoint for each morphism (to model
+reversibility) and that they support higher-order relation (built from the
+categorical dagger which is modeled by our fractional types). The main
+requirement on the dagger operation is the isomorphism above witnessed by
+{{eta*_b}} and {{eps*_b}}.
+
+One can think of several concrete compact closed categories: the first and
+most intuitive is the category of finite sets and relations. In that case,
+the dagger operation (i.e., the fractional type) collapses by defining
+{{1/b}} to be {{b}}. This is still an interesting category as any relation is
+trivially reversible and more importantly any relation between {{b1}} and
+{{b2}} can be represented as a subset of {{b1 * b2}}, i.e., as a value. A
+more interesting example is the category of finite dimensional vector spaces
+and linear maps (or even the category of finite dimensional Hilbert
+spaces). In such categories the dagger operation takes a vector space
+(consisting of ``column vectors'') to the dual vector space (consisting of
+``row vectors''). These vector spaces are isomorphic, but they are
+nevertheless quite different, being $(1,0)$ and $(0,1)$ tensors
+(respectively).  In other words, the category provides a more refined model
+in which isomorphic negative-information values and positive-information are
+not identified.
+
+In this section, we provide a simple semantics of {{PiTF}} in the category of
+finite sets and relations. In other words, we interpret each combinator 
+{{c :  b1 <-> b2}} as a \emph{relation} between the sets denoted
+by {{b1}} and {{b2}} respectively. As we illustrate, this semantics allows
+\emph{arbitrary} relations (including the empty relation and
+relations that are not isomorphisms) to be represented as values in {{PiTF}}.
+In Sec.~\ref{sec:cat} we discuss a more refined semantics that is
+more suitable for richer categories and which only allows relations
+representing \emph{isomorphisms} to be represented as values in~{{PiTF}}.
 
@@
 backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.
-
-More generally, the semantics of {{PiTF}} using sets and relations, i.e., in
-the category of finite sets and relations suggests a few additional immediate
-properties.  The category of finite sets and relations is an example of a
-\emph{biproduct dagger compact closed category} which are the categories
-introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
-and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
-computation.\footnote{The other common example of these categories is that of
-  finite dimensional Hilbert spaces.} Two of the most relevant features of
-these categories for our purposes are that they support an adjoint for each
-morphism (to model reversibility) and that they support higher-order relation
-(built from the categorical dagger which is modeled by our fractional
-types). The latter is intuitively clear as any relation between $A$ and $B$
-can be represented as a subset of {{A * B}}, i.e., as a value. We return to
-such properties of {{PiTF}} in Sec.~\ref{sec:cat} after we present
-programming examples that illustrate {{PiTF}} and its ability to express
-various higher-order notions.
 

### 2013-02-20 frac.tex
cleaned up a few small things

@@
 
-The extension of the sets of types and values is simple:
+The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
+simple:
 
@@
 For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
-where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. There is
-however have some informal intuition about the \emph{meaning} of such
-fractional types and values. (See Sec.~\ref{sec:cat} for more discussion
-about this point.) From both the programming perspective and from the
-categorical perspective we expect the fractional types to satisfy the
+where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
+however some informal intuition about the \emph{meaning} of such fractional
+types and values. (See Sec.~\ref{sec:cat} for more discussion about this
+point.) In more detail, given a symmetric monoidal category modeling a
+certain language (like {{Pi}}), one can obtain a higher-order extension of
+the language by extending the category to a \emph{compact closed} one, i.e.,
+a category in which morphisms are representable as objects. The key to this
+extension is to add dual objects (in our case the fractionals) satisfying the
 following isomorphism witnessed by two new combinators {{eta*_b}} and
@@
 
-From a programming perspective, we think of a value {{1/v}} as a
+\noindent From a programming perspective, we think of a value {{1/v}} as a
 \emph{first-class constraint} that can only be satisfied if it is matched
 with an actual value {{v}} or as a \emph{pattern} representing the absence of
-some information of a specific shapre (i.e., \emph{negative information}),
+some information of a specific shape (i.e., \emph{negative information}),
 that can only be reconciliated by an actual value {{v}}. The combinator
@@
 up an equal amount of negative and positive information producing no residual
-information. From the categorical perspective, the extended language {{PiTF}}
-needs to have a semantics in a \emph{compact closed category} which are the
-categories introduced by Abramsky and
-Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
-Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
-computation. Two of the most relevant features of these categories for our
-purposes are that they support an adjoint for each morphism (to model
-reversibility) and that they support higher-order relation (built from the
-categorical dagger which is modeled by our fractional types). The main
-requirement on the dagger operation is the isomorphism above witnessed by
-{{eta*_b}} and {{eps*_b}}.
-
-One can think of several concrete compact closed categories: the first and
-most intuitive is the category of finite sets and relations. In that case,
-the dagger operation (i.e., the fractional type) collapses by defining
-{{1/b}} to be {{b}}. This is still an interesting category as any relation is
-trivially reversible and more importantly any relation between {{b1}} and
-{{b2}} can be represented as a subset of {{b1 * b2}}, i.e., as a value. A
-more interesting example is the category of finite dimensional vector spaces
-and linear maps (or even the category of finite dimensional Hilbert
-spaces). In such categories the dagger operation takes a vector space
-(consisting of ``column vectors'') to the dual vector space (consisting of
-``row vectors''). These vector spaces are isomorphic, but they are
-nevertheless quite different, being $(1,0)$ and $(0,1)$ tensors
-(respectively).  In other words, the category provides a more refined model
-in which isomorphic negative-information values and positive-information are
-not identified.
+information.
+
+From the categorical perspective, versions of compact closed categories were
+introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
+and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the
+generalization of monoidal categories to model quantum computation.  The
+first and most intuitive example of such categories is the category of finite
+sets and relations. In that case, the dual operation (i.e., the fractional
+type) collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
+interpretation of fractionals, this is still an interesting category as it
+provides a model for a reversible higher-order language. Indeed, any relation
+is trivially reversible and any relation between {{b1}} and {{b2}} can be
+represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
+example is the category of finite dimensional vector spaces and linear maps
+(or even the category of finite dimensional Hilbert spaces). In such
+categories the dual operation takes a vector space (consisting of ``column
+vectors'') to the dual vector space (consisting of ``row vectors''). These
+vector spaces are isomorphic, but they are nevertheless quite different,
+being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the
+category provides a more refined model in which isomorphic
+negative-information values and positive-information are not identified.
 
 In this section, we provide a simple semantics of {{PiTF}} in the category of
-finite sets and relations. In other words, we interpret each combinator 
-{{c :  b1 <-> b2}} as a \emph{relation} between the sets denoted
-by {{b1}} and {{b2}} respectively. As we illustrate, this semantics allows
-\emph{arbitrary} relations (including the empty relation and
-relations that are not isomorphisms) to be represented as values in {{PiTF}}.
-In Sec.~\ref{sec:cat} we discuss a more refined semantics that is
-more suitable for richer categories and which only allows relations
-representing \emph{isomorphisms} to be represented as values in~{{PiTF}}.
+finite sets and relations. In other words, we interpret each combinator {{c :
+    b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
+{{b2}} respectively. As we illustrate, this semantics allows \emph{arbitrary}
+relations (including the empty relation and relations that are not
+isomorphisms) to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat}
+we discuss more refined semantics that are more suitable for richer
+categories.
 
@@
 
-The operational semantics define a relation rather than a function, now, but the
-goal is for them to still be reversible; in particular, we wish for the relation
-to mode check so that for any combinator {{c}} we can treat either its left or
-its right side as an input; this allows us to define {{sym}} as just an argument
-swap. In the interest of brevity, we have treated the type annotations in the
-base isomorphisms as implicit; for example, the rule for {{assocr*}} should
-really read
+The operational semantics define a relation rather than a function, but the
+goal is for them to still be reversible; in particular, we wish for the
+relation to mode check so that for any combinator {{c}} we can treat either
+its left or its right side as an input; this allows us to define {{sym}} as
+just an argument swap. We can then give an operational interpretation to the
+semantic relation by defining two functions: a forward evaluator and a
+backwards evaluator that exercise the relation in opposite directions. It is
+straightforward to check that if the forward evaluation of {{c v1}} produces
+{{v2}} as a possible answer then the backwards evaluation of {{c v2}}
+produces {{v1}} as a possible answer.
+
+In the interest of brevity, we have treated the type annotations in the base
+isomorphisms as implicit; for example, the rule for {{assocr*}} should really
+read:
 
@@
 %@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))
-
-If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 
-{{c : b1 <-> b2}} would define a total one-to-one function between the sets 
-{{ [[ b1 ]] }} and {{ [[ b2 ]] }}. 
-This is indeed the situation for the language {{Pi}} 
-defined in the previous section. In the presence of fractionals, this simple 
-interpretation of combinators as one-to-one functions is not sufficient%
-\footnote{The interpretation of combinators as one-to-one functions is also 
-not valid in the presence of recursive 
-types~\cite{rc2011,James:2012:IE:2103656.2103667}.}.
-The intuitive reason, is that the fission point combinator 
-{{eta*_b : 1 <-> 1/b * b}} must be
-able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
-{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
-accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
-not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
-undefined, i.e., it does not produce {{():1}}.
-
-The relational semantics can be given an operational interpretation by
-defining two functions: a forward evaluator and a backwards evaluator that
-exercise the relation in opposite directions. It follows that if the forward
-evaluation of {{c v1}} produces {{v2}} as a possible answer then the
-backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.
 
@@
 
+If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 
+{{c : b1 <-> b2}} would define a total one-to-one function between the sets 
+{{ [[ b1 ]] }} and {{ [[ b2 ]] }}. 
+This is indeed the situation for the language {{Pi}} 
+defined in the previous section. In the presence of fractionals, this simple 
+interpretation of combinators as one-to-one functions is not sufficient%
+\footnote{The interpretation of combinators as one-to-one functions is also 
+not valid in the presence of recursive 
+types~\cite{rc2011,James:2012:IE:2103656.2103667}.}.
+The intuitive reason, is that the fission point combinator 
+{{eta*_b : 1 <-> 1/b * b}} must be
+able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
+{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
+accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
+not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
+undefined, i.e., it does not produce {{():1}}.
+
+discuss two generalizations of the semantics
+
+vector spaces (more refined because it doesn't collpase 1/b to b; also each
+vector space must have a zero vector; so we have an explicit notion of
+``error'' which we can use and track at run time. Adding this error
+eliminates at RUN TIME relations which are not isomorphisms I think.
+
+a potentionally better and more semantics is based on groupoids; mention that
+we can have sets (with stucture) who cardinality is fractionals and that this
+would formalize nicely the negative information
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-21 frac.tex
a little in sec 5

@@
 
-If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 
-{{c : b1 <-> b2}} would define a total one-to-one function between the sets 
-{{ [[ b1 ]] }} and {{ [[ b2 ]] }}. 
-This is indeed the situation for the language {{Pi}} 
-defined in the previous section. In the presence of fractionals, this simple 
-interpretation of combinators as one-to-one functions is not sufficient%
-\footnote{The interpretation of combinators as one-to-one functions is also 
-not valid in the presence of recursive 
-types~\cite{rc2011,James:2012:IE:2103656.2103667}.}.
-The intuitive reason, is that the fission point combinator 
-{{eta*_b : 1 <-> 1/b * b}} must be
-able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
-{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
-accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
-not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
-undefined, i.e., it does not produce {{():1}}.
-
-discuss two generalizations of the semantics
-
-vector spaces (more refined because it doesn't collpase 1/b to b; also each
-vector space must have a zero vector; so we have an explicit notion of
-``error'' which we can use and track at run time. Adding this error
-eliminates at RUN TIME relations which are not isomorphisms I think.
-
-a potentionally better and more semantics is based on groupoids; mention that
-we can have sets (with stucture) who cardinality is fractionals and that this
-would formalize nicely the negative information
+The interpretation of {{PiTF}} in the category of sets and relations is
+adequate in the sense that it produces a language in which every program is a
+reversible relation and in which relations are first-class values. This
+relational model is however, unsatisfactory, for several reasons:
+
+\begin{itemize}
+\item the type {{1/b}} is interpreted in the same way as {{b}} which gives no
+  insight into the ``true meaning'' of fractionals;
+\item the interpretation is inconsistent with the view of types as algebraic
+  structures; for example, {{Pi}} with the empty type, is the
+  categorificationi of a semiring but although fractional types in {{PiTF}}
+  syntactically ``look like'' rational numbers, there is no such formal
+  connection to the rational numbers;
+\item finally, we have lost some delicate structure moving from {{Pi}} to
+  {{PiTF}} as we can express arbitrary relations between types and not just
+  isomorphisms.
+\end{itemize}
+
+For these reasons, it is interesting to consider other possible semantic
+interpretations of fractionals. A natural alternative is to use the
+``canonical'' compact closed category, that is finite dimensional vector
+spaces and linear maps. We outline the main features of such an
+interpretation:
+
+\begin{itemize}
+\item before we talk about vector spaces, we must fix an underlying
+  field. For simplicity in the following discussion we choose the underlying
+  field to be the field of booleans with two elements labeled \texttt{F} and
+  \texttt{T}
+\item each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
+  vector space $\mathbf{V}_b$ over the field of booleans. Of particular
+  interest:
+  \begin{itemize}
+  \item every vector space contains a zero vector which means that the type
+    {{0}} (if included in {{PiTF}} would not be the ``empty'' type --- it
+    would contain the empty vector. Furthermore all combinators would be
+    partial as they would have to map the zero vector to itself. 
+  \item the type {{1}} is interpreted as a 1-dimensional vector space and hence
+    is isomorphic to the underlying field.
+  \item the fractional type {{1/b}} is interpreted as the dual vector space
+    of the vector space $\mathbf{V}_b$ representing {{b}} consisting of all
+    the \emph{linear functionals} on $\mathbf{V}_b$.
+  \item one can then validate certain desirable properties: {{1/(1/b)}} is
+    isomorphic to {{b}}; and {{eps*_b}} corresponds to a bilinear form which
+    maps a dual vector and a vector to a field element.
+  \end{itemize}
+\end{itemize}
+
+Although this semantics appears to have ``better'' properties than the
+relational one, we argue that it is not \emph{the} ``perfect'' semantics. By
+including the zero vector, the language has again additional partial
+morphisms that do not correspond to isomorphisms and it is difficult to
+reconcile the interpretation with the view that the types are rational
+numbers. What would really be a ``perfect'' interpretation is one in which we
+can only express {{Pi}}-isomorphisms as first class values and in which the
+types are interpreted in a way that is consistent with the rational
+numbers. Fortunately, there is promising significant work on the groupoid
+interpretation of type theory and on the categorification of the rational
+numbers that we outline next.
+
+The fundamental idea of both lines of previous work is that it is possible to
+build sets (with additional structure) with a \emph{fractional cardinality}.
+
+%% EXPAND on this
 
@@
 \end{comment}
+
+If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 
+{{c : b1 <-> b2}} would define a total one-to-one function between the sets 
+{{ [[ b1 ]] }} and {{ [[ b2 ]] }}. 
+This is indeed the situation for the language {{Pi}} 
+defined in the previous section. In the presence of fractionals, this simple 
+interpretation of combinators as one-to-one functions is not sufficient%
+\footnote{The interpretation of combinators as one-to-one functions is also 
+not valid in the presence of recursive 
+types~\cite{rc2011,James:2012:IE:2103656.2103667}.}.
+The intuitive reason, is that the fission point combinator 
+{{eta*_b : 1 <-> 1/b * b}} must be
+able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
+{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
+accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
+not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
+undefined, i.e., it does not produce {{():1}}.
+
+discuss two generalizations of the semantics
+
+vector spaces (more refined because it doesn't collapse 1/b to b; also each
+vector space must have a zero vector; so we have an explicit notion of
+``error'' which we can use and track at run time. Adding this error
+eliminates at RUN TIME relations which are not isomorphisms I think.
+
+a potentially better and more semantics is based on groupoids; mention that
+we can have sets (with structure) who cardinality is fractionals and that this
+would formalize nicely the negative information

### 2013-02-21 frac.tex
included diagrams

@@
 
-\newcommand{\inkscape}[2][1.5]{
+\newcommand{\dgm}[2][1.5]{
 \begin{center}
 \scalebox{#1}{
-\includegraphics{diagrams/#2}
+\includegraphics{diagrams/#2.pdf}
 }
@@
 
+\begin{multicols}{2}
+  \dgm{eta_times2}
+
+  \dgm{eps_times2}  
+\end{multicols}
+
 The operational semantics define a relation rather than a function, but the
@@
 
+\begin{multicols}{2}
+  \dgm{function-times}
+
+  \dgm{delimc-times}
+\end{multicols}
+
 \noindent which suggests a function-like behavior for the pair of a fractional value
@@
 % apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
+
+\begin{multicols}{2}
+  \dgm[1.2]{apply-times1}
+
+  \dgm{apply-times2}
+\end{multicols}
 
@@
 % compose &=& assocr* (;) (id (*) apply')
+
+\begin{multicols}{2}
+  \dgm[1.0]{compose-times1}
+
+  \dgm{compose-times2}
+\end{multicols}
+
+\dgm{compose-times3}
 
@@
 Consider the circuit below:
-\begin{center}
-\scalebox{1.2}{
-  \includegraphics{diagrams/not_trace.pdf}
-}
-\end{center}
+\dgm{not_trace}
 As motivated earlier, it is clearly impossible for the constraint implied by

### 2013-02-21 frac.tex
added involution diagram
do a pass through the whole paper

@@
   fractional type {{1/v}} represents \emph{negative} information. A
-  higher-order function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
+  function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
   representing the needed argument and {{v2}} representing the
@@
 new and surprising connections between geometry, algebra, logic, and
-computation (see~\cite{baez2011physics} for an overview).
+computation (see~\cite{baez2011physics} for an overview of some of the 
+connections).
 
@@
 well-understood, the details of adapting it to an actual programming language
-as subtle (as we explain in the main body of the paper).  Our main novel
+are subtle (as we explain in the main body of the paper).  Our main novel
 technical device to achieving the higher-order extension is a
@@
 %subcode{bnf} include main
-%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
-%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
-%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
-%identl* :&  1 * b <-> b &: identr*
-%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
-%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
-%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
+%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
+%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b & <-> & b &: identr*
+%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
+%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor
 \noindent Each line of the above table introduces a pair of dual
@@
 extension is to add dual objects (in our case the fractionals) satisfying the
-following isomorphism witnessed by two new combinators {{eta*_b}} and
-{{eps*_b}}:
-
-{{eta*_b : 1 <-> 1/b * b : eps*_b}}
+following isomorphism witnessed by two new type indexed combinators 
+{{eta*_b}} and {{eps*_b}}:
+
+%subcode{bnf} include main
+%! columnStyle = r r c l l
+%eta*_b :&  1 & <-> & 1/b * b &: eps*_b
+
+%% {{eta*_b : 1 <-> 1/b * b : eps*_b}}
 
@@
 provides a model for a reversible higher-order language. Indeed, any relation
-is trivially reversible and any relation between {{b1}} and {{b2}} can be
+is reversible via the standard \emph{converse} operation on a relation.
+Furthermore, any relation between {{b1}} and {{b2}} can be
 represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
@@
 In this section, we provide a simple semantics of {{PiTF}} in the category of
-finite sets and relations. In other words, we interpret each combinator {{c :
-    b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
+finite sets and relations. In other words, we interpret each combinator 
+{{c : b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
 {{b2}} respectively. As we illustrate, this semantics allows \emph{arbitrary}
@@
 its left or its right side as an input; this allows us to define {{sym}} as
-just an argument swap. We can then give an operational interpretation to the
+just relational converse, which amounts to
+an argument swap. We can then give an operational interpretation to the
 semantic relation by defining two functions: a forward evaluator and a
-backwards evaluator that exercise the relation in opposite directions. It is
+backwards evaluator that exercise the relation in opposite directions,
+but which now can return a \emph{set} of results. It is
 straightforward to check that if the forward evaluation of {{c v1}} produces
 {{v2}} as a possible answer then the backwards evaluation of {{c v2}}
-produces {{v1}} as a possible answer.
+produces {{v1}} as a possible answer.  One can check that all relations
+for {{Pi}} are total functional relations whose converse are also total 
+functional relations -- aka isomorphisms.  However {{eta*_b}} relates
+{{()}} to multiple values (one for each value of {{b}}), and is thus
+not functional; {{eps*_b}} is exactly its relational converse, as expected.
 
@@
 
-\begin{multicols}{2}
-  \dgm{function-times}
-
-  \dgm{delimc-times}
-\end{multicols}
-
 \noindent which suggests a function-like behavior for the pair of a fractional value
@@
 
-As a simple example, consider the combinator {{swap+}} denoting boolean
+\begin{multicols}{2}
+  \dgm{function-times}
+
+  \dgm{delimc-times}
+\end{multicols}
+\noindent  Dually, as illustrated above, we also have a {{coname}} combinator
+which when given an answer {{v2 : b2}}, and a request for the corresponding
+input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this as a un-needed 
+computation.
+
+As a simple example, consider {{swap+}} denoting boolean
 negation.  It is straightforward to check that the combinator 
-{{name swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:
+{{name~swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:
 
@@
 %@ ~
-%@@ () ~~(name swap+)~~ (1/false,true)
-%
-%@ ~
-%@@ () ~~(name swap+)~~ (1/true,false)
-
-\noindent The relation {{name swap+}} maps {{()}} to a value which can be used
+%@@ () ~~(name~swap+)~~ (1/false,true)
+%
+%@ ~
+%@@ () ~~(name~swap+)~~ (1/true,false)
+
+\noindent The relation {{name~swap+}} maps {{()}} to a value which can be used
 to convert any {{false}} to {{true}} and vice-versa. We formalize this
@@
 
+\dgm{involution-times}
+
 \noindent At the type level, this computation proceeds as follows:
@@
 
-In this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
+Given a computation {{c:b1 * b2 <-> b1 * b3}} and we
 want to cancel the common type {{b1}}, to produce a new combinator
@@
 is the value produced as the first component of the result.  In general, there
-may be several such fixed-point values or none.  A few examples should help
+may be several such fixed-point values, or none.  A few examples should help
 explain the subtleties.
@@
 {{() : 1}} requires us to find a value {{b : bool}} such that 
-{{(b,()) ~~c~~ (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
+{{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
 has two values, there are two values {{false}} and {{true}} that satisfy the 
-constraint. In other words, expect that {{() ~~(trace* c)~~() }}.
+constraint. In other words, {{() ~~(trace* c)~~() }}.
 \end{example}
@@
 combinator to {{() : 1}} requires us to find a value {{b : bool}} such
-that {{(b,()) ~~c~~ (b,())}} which is impossible since {{c}} negates
-{{b}}, i.e. {{c}} has no fixed points. Operationally, the evaluation of
-such a combinator should produce no value (if it even terminates).
+that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
+has no fixed points.  Operationally, the evaluation of
+such a combinator should produce no value (were it to even terminate).
 \end{example}
 
-With fractionals, the multiplicative trace operator becomes expressible:
+With fractionals, {{trace*}} becomes directly expressible:
 
@@
 
-Let's use the semantics to calculate the result of applying
-{{trace*_{bool} (not (*) id)}} to {{false}}. We start
-with the set {{ {[ false ]} }} which gets transformed as follows:
+We can use the relational semantics of definition~\ref{def:relational-PiTF}
+to calculate the result of applying
+{{trace*_{bool} (swap+ (*) id)}} to {{false}}.  We use the usual conversion
+of a relation from a subset of {{A x B}} to a function from {{A}} to the
+powerset of {{B}} (and Kleisli composition in the powerset monad) for
+the computation:
 %subcode{opsem} include main
@@
 
-%% \jacques{in the current paper, this seems to no longer make sense}
-%% Yes I know, but I think we should leave it to explain the problem in Sec. 5
-%% and motivate the different semantics
-
-\begin{example}[Annihilation]
-\label{ch3:ex;annihilation}
-Consider the circuit below:
-\dgm{not_trace}
-As motivated earlier, it is clearly impossible for the constraint implied by
-{{trace*}} to be satisfied as a boolean value {{v}} can never be the same as
-its negation. Thus there is no incoming value of type {{b}} that can ever go
-through the circuit. Indeed, if we call the circuit above {{c}}, then we can
-calculate that {{c}} does not relate {{v}} to anything. We call such a
-circuit an annihilation circuit.
-\end{example}
+\noindent  The computation with {{true}} gives the same result.  This
+implies that although our semantics is formally reversible, it 
+strays from the kind of reversibility that has been achieved with {{Pi}}.
+Nevertheless, the next subsection illustrates why this is still useful.
 
@@
 the move to relations motivated in the previous section, we can
-express relational programming. We illustrate the idea using a
-complete example.
+express relational programming.
 
@@
 
-We show how to define a combinator {{c_R}} whose semantics is such that:
-
-%subcode{proof} include main
-%@ ~
-%@@ false ~~c_R~~ false
-%
-%@ ~
-%@@ false ~~c_R~~ true
-%
-%@ ~
-%@@ true ~~c_R~~ false
-
+\noindent We can define a combinator {{c_R}} whose denotation is {{R}}.
 The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
-for some type {{a}} such that {{trace* cInner}} is the desired combinator
-{{c_R}}.  A little experimentation shows that we get the desired behavior if
+for some type {{a}} such that {{c_R}} == {{trace* cInner}}.
+A little experimentation shows that we get the desired behavior 
 as follows:
@@
 % cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
-% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assoct* (;) swap* (;) 
+% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;) 
 %        &&  toffoli (;) ((swap* (;) cnot (;) swap*) (*) id) (;)
@@
 %        &&  toffoli (;) (cnot (*) id))
-%     
-% c_R &:& bool <-> bool
-% c_R &=& trace* cInner
+\noindent where {{toffoli}} and {{cnot}} were defined (for {{Pi}}) in
+\cite{rc2011}.
+% %     
+% % c_R &:& bool <-> bool
+% % c_R &=& trace* cInner
 
@@
 \begin{itemize}
-\item the type {{1/b}} is interpreted in the same way as {{b}} which gives no
+\item the type {{1/b}} is interpreted in the same way as {{b}}, which gives no
   insight into the ``true meaning'' of fractionals;
@@
   structures; for example, {{Pi}} with the empty type, is the
-  categorificationi of a semiring but although fractional types in {{PiTF}}
+  categorification of a semiring but although fractional types in {{PiTF}}
   syntactically ``look like'' rational numbers, there is no such formal
@@
   {{PiTF}} as we can express arbitrary relations between types and not just
-  isomorphisms.
+  \emph{isomorphisms}.
 \end{itemize}
@@
 ``canonical'' compact closed category, that is finite dimensional vector
-spaces and linear maps. We outline the main features of such an
-interpretation:
+spaces and linear maps\cite{Selinger:2011:FDH:1942319.1942398,Hasegawa:2008:FDV:1805839.1805859} over fields of characteristic $0$. Let us fix an arbitrary
+field $k$ of characteristic $0$.
+Such an interpretation has the following features:
 
 \begin{itemize}
-\item before we talk about vector spaces, we must fix an underlying
-  field. For simplicity in the following discussion we choose the underlying
-  field to be the field of booleans with two elements labeled \texttt{F} and
-  \texttt{T}
 \item each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
-  vector space $\mathbf{V}_b$ over the field of booleans. Of particular
+  vector space $\mathbf{V}_b$ over the field $k$. Of particular
   interest:
@@
   \item every vector space contains a zero vector which means that the type
-    {{0}} (if included in {{PiTF}} would not be the ``empty'' type --- it
-    would contain the empty vector. Furthermore all combinators would be
-    partial as they would have to map the zero vector to itself. 
+    {{0}} (if included in {{PiTF}} would not be the ``empty'' type.
+    Furthermore all combinators would be
+    strict as they would have to map the zero vector to itself. 
   \item the type {{1}} is interpreted as a 1-dimensional vector space and hence
@@
   \item the fractional type {{1/b}} is interpreted as the dual vector space
-    of the vector space $\mathbf{V}_b$ representing {{b}} consisting of all
+    to the vector space $\mathbf{V}_b$ representing {{b}} consisting of all
     the \emph{linear functionals} on $\mathbf{V}_b$.
@@
 relational one, we argue that it is not \emph{the} ``perfect'' semantics. By
-including the zero vector, the language has again additional partial
+including the zero vector, the language has
 morphisms that do not correspond to isomorphisms and it is difficult to

### 2013-02-22 frac.tex
added proof of curry

@@
 
+%% XXX: diagrams need unit wires(?)
 \begin{multicols}{2}
@@
 with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
-abbreviation for {{right ()}}.
+abbreviation for {{right ()}}. In addition, for the sake of readability, the
+circuit diagrams in this section will elide trivial unit wires, such as in
+{{eta}} and {{eps}}.
 
@@
 %     &=& b1 :-* b3
+
+We can also derive currying (and dually, uncurrying) combinators. Observe that
+the type of {{curry}} needs to be {{b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)}},
+which can be written in mathematical notation as {{b3/(b1 * b2) = (b3 * 1/b2) *
+(1/b1)}}. This means {{curry}} can be written using {{recip}}, the
+implementation of the mathematical identity {{1/(b1 * b2) = 1/b1 * 1/b2}}:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% recip &:& 1 / (b1 * b2) <-> 1 / b1 * 1 / b2
+% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta* (*) eta*) (*) id)
+%     &&    (;) (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*)
+%     &&    (;) swap* (;) (unite)
+
+\noindent
+where {{reorder}} is the obvious combinator of type {{b1 * (b2 * b3) * b4 <-> b1
+* (b3 * b2) * b4}}. At the type level, the computation proceeds as follows:
+
+%subcode{opsem} include main
+%! columnstyle = rlr
+%        & 1/(b1 * b2) & 
+%     <->& 1 * 1/(b1 * b2) & uniti
+%     <->& 1 * (1 * 1/(b1 * b2)) & uniti
+%     <->& (1 * 1) * 1/(b1 * b2) & assocl*
+%     <->& ((1/b1 * b1) * (1/b2 * b2)) * 1/(b1 * b2) &(eta*_{b1} (*) eta*_{b2}) (*) id
+%     <->& ((1/b1 * 1/b2) * (b1 * b2)) * 1/(b1 * b2) & reoder (*) id
+%     <->& (1/b1 * 1/b2) * ((b1 * b2) * 1/(b1 * b2)) & assocr*
+%     <->& (1/b1 * 1/b2) * (1/(b1 * b2) * (b1 * b2)) & id * swap*
+%     <->& (1/b1 * 1/b2) * 1 & id * eps*
+%     <->& 1 * (1/b1 * 1/b2) & swap*
+%     <->& (1/b1 * 1/b2) & unite
+
+With {{recip}} out of the way, we can easily write {{curry}}:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% curry &:& b1 (x) b2 :-* b3 <-> b1 :-* (b2 :-* b3)
+% curry &=& (id * recip) (;) (id * swap*) (;) assocl*
 

### 2013-02-22 frac.tex
cleaned up currying stuff

@@
 combinators that manipulate functions or relations. In particular, we
-can \emph{apply} and \emph{compose} values representing relations as
-shown below:
+can \emph{apply} and \emph{compose} values representing relations.
+\emph{apply} is shown below:
 
@@
 
+\noindent
 We can also derive currying (and dually, uncurrying) combinators. Observe that
@@
 % recip &:& 1 / (b1 * b2) <-> 1 / b1 * 1 / b2
-% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta* (*) eta*) (*) id)
+% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta*_{b1} (*) eta*_{b2}) (*) id)
 %     &&    (;) (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*)
@@
 
+%% XXX: why won't the columnstyle declaration here work? (Z)
 %subcode{opsem} include main
-%! columnstyle = rlr
+%! columnStyle = rlr
 %        & 1/(b1 * b2) & 
-%     <->& 1 * 1/(b1 * b2) & uniti
-%     <->& 1 * (1 * 1/(b1 * b2)) & uniti
-%     <->& (1 * 1) * 1/(b1 * b2) & assocl*
-%     <->& ((1/b1 * b1) * (1/b2 * b2)) * 1/(b1 * b2) &(eta*_{b1} (*) eta*_{b2}) (*) id
-%     <->& ((1/b1 * 1/b2) * (b1 * b2)) * 1/(b1 * b2) & reoder (*) id
-%     <->& (1/b1 * 1/b2) * ((b1 * b2) * 1/(b1 * b2)) & assocr*
-%     <->& (1/b1 * 1/b2) * (1/(b1 * b2) * (b1 * b2)) & id * swap*
-%     <->& (1/b1 * 1/b2) * 1 & id * eps*
-%     <->& 1 * (1/b1 * 1/b2) & swap*
-%     <->& (1/b1 * 1/b2) & unite
-
+%     <->& 1 * 1/(b1 * b2) & (uniti)
+%     <->& 1 * (1 * 1/(b1 * b2)) & (uniti)
+%     <->& (1 * 1) * 1/(b1 * b2) & (assocl*)
+%     <->& ((1/b1 * b1) * (1/b2 * b2)) * 1/(b1 * b2) &((eta*_{b1} (*) eta*_{b2}) (*) id)
+%     <->& ((1/b1 * 1/b2) * (b1 * b2)) * 1/(b1 * b2) & (reoder (*) id)
+%     <->& (1/b1 * 1/b2) * ((b1 * b2) * 1/(b1 * b2)) & (assocr*)
+%     <->& (1/b1 * 1/b2) * (1/(b1 * b2) * (b1 * b2)) & (id * swap*)
+%     <->& (1/b1 * 1/b2) * 1 & (id * eps*)
+%     <->& 1 * (1/b1 * 1/b2) & (swap*)
+%     <->& (1/b1 * 1/b2) & (unite)
+
+\noindent
 With {{recip}} out of the way, we can easily write {{curry}}:
@@
 
-Given a computation {{c:b1 * b2 <-> b1 * b3}} and we
+Given a computation {{c:b1 * b2 <-> b1 * b3}},we
 want to cancel the common type {{b1}}, to produce a new combinator
@@
 consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
-The needed value {{v1 : b1}} cannot be arbitrary: it must be such that is
+The needed value {{v1 : b1}} cannot be arbitrary: it must be such that it
 is the value produced as the first component of the result.  In general, there

### 2013-02-23 frac.tex
removed type diagrams from section 4

@@
 
-\noindent At the type level, the computation proceeds as follows:
-%subcode{bnf} include main
-%  (b1 :-* b2) * b1 &=& (1/b1 * b2) * b1 & 
-%     &<->& b1 * (1/b1 * b2) & (swap*)
-%     &<->& (b1 * 1/b1) * b2) & (assocl*)
-%     &<->& (1/b1 * b1) * b2) & (swap* (*) id)
-%     &<->& 1 * b2 & (eps*_{b1} (*) id)
-%     &<->& b2 & (unite)
-
 \noindent Intuitively, we simply match the incoming argument of type {{b1}} with
@@
 \dgm{compose-times3}
-
-\noindent Again, at the type level, the computation proceeds as follows:
-
-%subcode{bnf} include main
-%  (b1 :-* b2) * (b2 :-* b3) &=& (1/b1 * b2) * (1/b2 * b3) & 
-%     &<->& (1/b1 * (b2 * (1/b2 * b3)) & (assocr*)
-%     &<->& (1/b1 * b3) & (id * apply')
-%     &=& b1 :-* b3
 
@@
 \noindent
-where {{reorder}} is the obvious combinator of type {{b1 * (b2 * b3) * b4 <-> b1
-* (b3 * b2) * b4}}. At the type level, the computation proceeds as follows:
-
-%% XXX: why won't the columnstyle declaration here work? (Z)
-%subcode{opsem} include main
-%! columnStyle = rlr
-%        & 1/(b1 * b2) & 
-%     <->& 1 * 1/(b1 * b2) & (uniti)
-%     <->& 1 * (1 * 1/(b1 * b2)) & (uniti)
-%     <->& (1 * 1) * 1/(b1 * b2) & (assocl*)
-%     <->& ((1/b1 * b1) * (1/b2 * b2)) * 1/(b1 * b2) &((eta*_{b1} (*) eta*_{b2}) (*) id)
-%     <->& ((1/b1 * 1/b2) * (b1 * b2)) * 1/(b1 * b2) & (reoder (*) id)
-%     <->& (1/b1 * 1/b2) * ((b1 * b2) * 1/(b1 * b2)) & (assocr*)
-%     <->& (1/b1 * 1/b2) * (1/(b1 * b2) * (b1 * b2)) & (id * swap*)
-%     <->& (1/b1 * 1/b2) * 1 & (id * eps*)
-%     <->& 1 * (1/b1 * 1/b2) & (swap*)
-%     <->& (1/b1 * 1/b2) & (unite)
-
-\noindent
+where {{reorder}} is the obvious combinator of type {{b1 * (b2 * b3) * b4 <-> b1 * (b3 * b2) * b4}}.
 With {{recip}} out of the way, we can easily write {{curry}}:
@@
 \dgm{involution-times}
-
-\noindent At the type level, this computation proceeds as follows:
-
-%subcode{bnf} include main
-%  b &<->& (1 * b) & (uniti)
-%    &<->& (1/(1/b) * 1/b) * b & (eta*_{1/b} (*) id)
-%    &<->& 1/(1/b) * (1/b * b) & (assocr*)
-%    &<->& 1/(1/b) * 1 & (id (*) eps*_b)
-%    &<->& 1 * 1/(1/b) & (swap*)
-%    &<->& 1/(1/b) & (unite)
 

### 2013-02-23 frac.tex
conclusion

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section{Conclusion}
 \label{sec:cat}
 
-The interpretation of {{PiTF}} in the category of sets and relations is
-adequate in the sense that it produces a language in which every program is a
-reversible relation and in which relations are first-class values. This
-relational model is however, unsatisfactory, for several reasons:
+We have introduced the idea of \emph{fractional types} in the context of a
+reversible language founded on type isomorphisms and preservation of
+information. Values of fractional types represent \emph{negative
+  information}, a concept which is difficult to introduce in a conventional
+language that allows arbitrary creation and deletion of information but which
+is much simpler to deal with when the surrounding language infrastructure
+guarantees preservation of information. Fractional types and values can be
+used to express a simple and elegant notion of higher-order functions: a
+function {{b1 :-* b2}} is a first-class value consisting of negative
+information drawn from {{b1}} and positive information drawn from {{b2}}.
+
+The interpretation of our language {{PiTF}} in the category of sets and
+relations is adequate in the sense that it produces a language in which every
+program is a reversible relation and in which relations are first-class
+values. This relational model is however, unsatisfactory, for several
+reasons:
 
@@
 relational one, we argue that it is not \emph{the} ``perfect'' semantics. By
-including the zero vector, the language has
-morphisms that do not correspond to isomorphisms and it is difficult to
-reconcile the interpretation with the view that the types are rational
-numbers. What would really be a ``perfect'' interpretation is one in which we
-can only express {{Pi}}-isomorphisms as first class values and in which the
-types are interpreted in a way that is consistent with the rational
-numbers. Fortunately, there is promising significant work on the groupoid
-interpretation of type theory and on the categorification of the rational
-numbers that we outline next.
-
-The fundamental idea of both lines of previous work is that it is possible to
-build sets (with additional structure) with a \emph{fractional cardinality}.
-
-%% EXPAND on this
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
-Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
-the relation:
-
-{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
-
-give some intuition; example with eta/epsilon
+including the zero vector, the language has morphisms that do not correspond
+to isomorphisms and it is difficult to reconcile the interpretation with the
+view that the types are rational numbers. What would really be a ``perfect''
+interpretation is one in which we can only express {{Pi}}-isomorphisms as
+first class values and in which the types are interpreted in a way that is
+consistent with the rational numbers. Fortunately, there is promising
+significant work on the groupoid interpretation of type theory and on the
+categorification of the rational numbers that we outline next. The
+fundamental idea of both lines of previous work is that it is possible to
+build sets (with additional structure) with a \emph{fractional cardinality}
+--- a topic we are currently investigating in detail.
 

### 2013-02-23 frac.tex
edits for length, style and precision.  Add a longer paragraph for the groupoid ideas

--- 2013-02-23 frac.tex
@@
 well-understood, the details of adapting it to an actual programming language
-are subtle (as we explain in the main body of the paper).  Our main novel
+are subtle.  Our main novel
 technical device to achieving the higher-order extension is a
@@
 and {{(v1,v2)}} which builds a value of product type. But these values should
-be regarded as largely ancillary.  In particular we do not treat the above
-values as first-class citizens.  They only occur when we want to observe the
-effect of an isomorphism.
+be regarded as largely ancillary: we do not treat them as first-class citizens,
+and they only occur when observing the effect of an isomorphism.
 
@@
 following constructors that witness equivalence and compatible closure:
+
 %subcode{proof} include main
@@
 simple:
-
 %subcode{bnf} include main
@@
 % values, v ::= ... | 1/v
-
 For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
@@
 {{eta*_b}} and {{eps*_b}}:
-
 %subcode{bnf} include main
@@
 %eta*_b :&  1 & <-> & 1/b * b &: eps*_b
-
-%% {{eta*_b : 1 <-> 1/b * b : eps*_b}}
+%% XXX: diagrams need unit wires(?)
+\begin{multicols}{2}
+  \dgm{eta_times2}
+
+  \dgm{eps_times2}  
+\end{multicols}
 
@@
 is reversible via the standard \emph{converse} operation on a relation.
-Furthermore, any relation between {{b1}} and {{b2}} can be
+Furthermore, as any relation between {{b1}} and {{b2}} can be
 represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
@@
 categories the dual operation takes a vector space (consisting of ``column
-vectors'') to the dual vector space (consisting of ``row vectors''). These
-vector spaces are isomorphic, but they are nevertheless quite different,
+vectors'') to the dual vector space (consisting of ``row vectors''). While
+these are isomorphic, but they are nevertheless quite different,
 being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the
@@
 In this section, we provide a simple semantics of {{PiTF}} in the category of
-finite sets and relations. In other words, we interpret each combinator 
+finite sets and relations. We interpret each combinator 
 {{c : b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
@@
 
-The denotation of combinators as relations is given below. We specify
+We specify
 relations using a deductive system whose judgments are of the form 
@@
 \end{definition}
-
-%% XXX: diagrams need unit wires(?)
-\begin{multicols}{2}
-  \dgm{eta_times2}
-
-  \dgm{eps_times2}  
-\end{multicols}
 
@@
 circuit diagrams in this section will elide trivial unit wires, such as in
-{{eta}} and {{eps}}.
+{{eta}} and {{epsilon}}.
 
@@
 
-\dgm{compose-times3}
+% \dgm{compose-times3}
 
@@
 We can also derive currying (and dually, uncurrying) combinators. Observe that
-the type of {{curry}} needs to be {{b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)}},
-which can be written in mathematical notation as {{b3/(b1 * b2) = (b3 * 1/b2) *
-(1/b1)}}. This means {{curry}} can be written using {{recip}}, the
-implementation of the mathematical identity {{1/(b1 * b2) = 1/b1 * 1/b2}}:
+the type of {{curry}} needs to be {{(b1 * b2 :-* b3) <-> (b1 :-* (b2 :-* b3))}},
+which can be written in mathematical notation as 
+{{1/(b1 * b2) * b3 = (1/b1) * ((1/b1) * b3)}}. This means {{curry}} can be
+written using {{recip}}, the implementation of the mathematical identity
+{{1/(b1 * b2) = 1/b1 * 1/b2}}:
 
@@
 % recip &:& 1 / (b1 * b2) <-> 1 / b1 * 1 / b2
-% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta*_{b1} (*) eta*_{b2}) (*) id)
-%     &&    (;) (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*)
-%     &&    (;) swap* (;) (unite)
+% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta*_{b1} (*) eta*_{b2}) (*) id) (;)
+%     && (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*) (;) swap* (;) (unite)
 
@@
 %! columnStyle = rcl
-% curry &:& b1 (x) b2 :-* b3 <-> b1 :-* (b2 :-* b3)
+% curry &:& b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)
 % curry &=& (id * recip) (;) (id * swap*) (;) assocl*
-
+\noindent That {{recip}} is the heart of currying seems quite remarkable.
 %%%%%%%%%%%%
@@
 The final remarkable property of fractional types is that they
-are involutive.  In particular, we have:
+are involutive:
 
@@
 
-\dgm{involution-times}
+% jacques: elide for now, to save space, as this diagram is backwards.
+% \dgm{involution-times}
 
@@
 that can be defined, an additive family (explored in detail in our previous
-work~\cite{rc2011}) and a multiplicative family (which is the focus of this
-section).
-
-Given a computation {{c:b1 * b2 <-> b1 * b3}},we
+work~\cite{rc2011}) and a multiplicative family, which we will focus on now.
+
+Given a computation {{c:b1 * b2 <-> b1 * b3}}, we
 want to cancel the common type {{b1}}, to produce a new combinator
-{{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
+{{trace*~c:b2<->b3}}. For its evaluation, we are only
 given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
@@
 \begin{example}
-Consider the combinator
-
-{{c : bool * 1 <-> bool * 1}}
-
-{{c = id}}
+Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.
 
@@
 {{trace*_{bool} (swap+ (*) id)}} to {{false}}.  We use the usual conversion
-of a relation from a subset of {{A x B}} to a function from {{A}} to the
+of a relation from a subset of {{A * B}} to a function from {{A}} to the
 powerset of {{B}} (and Kleisli composition in the powerset monad) for
@@
 implies that although our semantics is formally reversible, it 
-strays from the kind of reversibility that has been achieved with {{Pi}}.
+does not result in isomorphisms.
 Nevertheless, the next subsection illustrates why this is still useful.
@@
 
-\noindent We can define a combinator {{c_R}} whose denotation is {{R}}.
-The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
-for some type {{a}} such that {{c_R}} == {{trace* cInner}}.
-A little experimentation shows that we get the desired behavior 
-as follows:
+\noindent We can define a combinator {{c_R}} whose denotation is {{R}},
+by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
+for some type {{a}} such that {{c_R = trace* cInner}}:
+
 % %subcode{proof} include main
@@
 % cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
-% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;) 
-%        &&  toffoli (;) ((swap* (;) cnot (;) swap*) (*) id) (;)
-%        &&  toffoli (;) (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
-%        &&  toffoli (;) (cnot (*) id))
-\noindent where {{toffoli}} and {{cnot}} were defined (for {{Pi}}) in
+% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
+%        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
+%        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))
+
+\noindent where {{toffoli}} and {{cnot}} were defined for {{Pi}} in
 \cite{rc2011}.
@@
 field $k$ of characteristic $0$.
-Such an interpretation has the following features:
-
-\begin{itemize}
-\item each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
-  vector space $\mathbf{V}_b$ over the field $k$. Of particular
-  interest:
+Then each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
+  vector space $\mathbf{V}_b$ over the field $k$. In particular:
 \begin{itemize}
   \item every vector space contains a zero vector which means that the type
-    {{0}} (if included in {{PiTF}} would not be the ``empty'' type.
+    {{0}} (if included in {{PiTF}}) would not be the ``empty'' type.
     Furthermore all combinators would be
@@
 \end{itemize}
-\end{itemize}
 
@@
 including the zero vector, the language has
-morphisms that do not correspond to isomorphisms and it is difficult to
-reconcile the interpretation with the view that the types are rational
-numbers. What would really be a ``perfect'' interpretation is one in which we
+morphisms that do not correspond to isomorphisms (in particular it allows
+partial morphisms by treating the {{0}} element of the zero-dimensional 
+vector space as a canonical ``undefined'' value).  It is also difficult to
+reconcile the interpretation with the view that the types correspond
+to the (positive) rational numbers, something we are actively seeking. What 
+would really be a ``perfect'' interpretation is one in which we
 can only express {{Pi}}-isomorphisms as first class values and in which the
 types are interpreted in a way that is consistent with the rational
-numbers. Fortunately, there is promising significant work on the groupoid
-interpretation of type theory and on the categorification of the rational
-numbers that we outline next.
-
-The fundamental idea of both lines of previous work is that it is possible to
-build sets (with additional structure) with a \emph{fractional cardinality}.
-
-%% EXPAND on this
+numbers.
+
+Fortunately, there is promising significant work on the groupoid
+interpretation of type theory~\cite{Hofmann96thegroupoid} and on the 
+categorification of the rational numbers~\cite{math/9802029} that may well give
+us the model we desire.  The fundamental idea in both cases is that groupoids
+(viewed as sets with explicit isomorphisms as morphisms) naturally have a
+\emph{fractional cardinality}.  Types would be interpreted as groupoids, and
+terms would be (invertible) groupoid actions.  The remaining challenge is to
+identify those groupoid actions which are proper generalizations of
+isomorphisms \emph{and} can be represented as groupoids, so as to obtain a
+proper interpretation for a higher-order language.  As the category of
+groupoids is cartesian closed, this appears eminently feasible.
 
@@
 \section{Conclusion}
-
-Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
-the relation:
-
-{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}
-
-give some intuition; example with eta/epsilon
 

### 2013-02-23 frac.tex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2013-02-23 frac.tex
cleaned up examples in 4.4

@@
 
-\begin{example}
-Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.
-
-\noindent Using {{trace*}}, we construct the
-combinator {{trace* c : 1 <-> 1}}. Applying this to 
-{{() : 1}} requires us to find a value {{b : bool}} such that 
-{{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
-has two values, there are two values {{false}} and {{true}} that satisfy the 
-constraint. In other words, {{() ~~(trace* c)~~() }}.
-\end{example}
-
-\begin{example}
-\label{ch3:ex:annihilate}
-Consider a small variation on the combinator {{c}} above:
-
-{{c : bool * 1 <-> bool * 1}}
-
-{{c = swap+ (*) id }}
-
-\noindent which negates the boolean component of its incoming
+Consider the combinator {{id : bool * 1 <-> bool * 1}}.
+Using {{trace*}}, we construct the combinator {{trace* c : 1 <-> 1}}.
+Applying this to {{() : 1}} requires us to find a value {{b : bool}} such
+that {{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} has two values,
+there are two values {{false}} and {{true}} that satisfy the constraint.
+In other words, {{()~~(trace* id)~~() }}.
+
+%% \label{ch3:ex:annihilate}
+Consider another combinator that also has the type {{bool * 1 <-. bool * 1}},
+but is instead defined to be {{c = swap+ (*) id }}, which negates the boolean
+component of its incoming
 pair. Using the multiplicative trace operator, we can construct the
@@
 combinator to {{() : 1}} requires us to find a value {{b : bool}} such
-that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
+that {{(b,()) ~~swap+ (*) id~~ (b,())}} which is impossible since {{swap+}} 
 has no fixed points.  Operationally, the evaluation of
 such a combinator should produce no value (were it to even terminate).
-\end{example}
 

### 2013-02-23 frac.tex
collapsed section 4.5 into 4.4

@@
 %%%%%%%%%%%%
-\subsection{Feedback, Iteration, and Trace Operators}
+\subsection{Trace Operators and (Finite) Relational Programming}
 
@@
 
+%% XXX: do we want to cite something for relational programming?
 \noindent  The computation with {{true}} gives the same result.  This
@@
 does not result in isomorphisms.
-Nevertheless, the next subsection illustrates why this is still useful.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{(Finite) Relational Programming}
-\label{ch3:sec:lp}
-
-Relational programming leverages set-theoretic relations and their
-composition to express computational tasks in a declarative way. 
+Nevertheless, this can still be useful, as it gives us access to
+\emph{relational programming},
+which uses set-theoretic relations and their
+composition to express computational tasks in a declarative way. Their encoding
+in {{PiTF}} uses fractional types (and specifically, their ability to encode the
+multiplicative trace operator).
+%% XXX: this falls kind of flat without some kind of example. do we want to
+%% release code with the paper and mention it here?
+
 % For instance, consider this example:
@@
 % {{grandparent(A)}}, we would get the answer {{ {[C]} }}.
-It turns out that with the addition of the multiplicative trace, and
-the move to relations motivated in the previous section, we can
-express relational programming.
-
-Consider the relation {{R}} on booleans given by:
-
-{{ {[(false,false), (false,true), (true,false) ]}. }}
-
-\noindent We can define a combinator {{c_R}} whose denotation is {{R}},
-by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
-for some type {{a}} such that {{c_R = trace* cInner}}:
-
-% %subcode{proof} include main
-% %@ ~
-% %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
-% %----
-% %@ ~
-% %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
-% %----
-% %@ ~
-% %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
-% %----
-% %@ ~
-% %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
-% %----
-% %@ ~
-% %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
-% %----
-% %@ ~
-% %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
-% %----
-% %@ ~
-% %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
-% %----
-% %@ ~
-% %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
-% In the first three lines, the first argument is a fixed point and hence the
-% multiplicative trace would map the second input to the second output
-% producing the desired relation. In the remaining five cases, the first
-% argument is not a fixed point and hence all these cases would be rejected as
-% solutions to the constraint imposed by the multiplicative trace. It simply
-% remains to find the actual realization of {{cInner}} that would produce the
-% behavior above.
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
-% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
-%        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
-%        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))
-
-\noindent where {{toffoli}} and {{cnot}} were defined for {{Pi}} in
-\cite{rc2011}.
+
+% Consider the relation {{R}} on booleans given by:
+% 
+% {{ {[(false,false), (false,true), (true,false) ]}. }}
+% 
+% \noindent We can define a combinator {{c_R}} whose denotation is {{R}},
+% by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
+% for some type {{a}} such that {{c_R = trace* cInner}}:
+% 
+% % %subcode{proof} include main
+% % %@ ~
+% % %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
+% % %----
+% % %@ ~
+% % %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
+% % %----
+% % %@ ~
+% % %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
+% % %----
+% % %@ ~
+% % %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
+% % %----
+% % %@ ~
+% % %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
+% % %----
+% % %@ ~
+% % %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
+% % %----
+% % %@ ~
+% % %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
+% % %----
+% % %@ ~
+% % %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
+% % In the first three lines, the first argument is a fixed point and hence the
+% % multiplicative trace would map the second input to the second output
+% % producing the desired relation. In the remaining five cases, the first
+% % argument is not a fixed point and hence all these cases would be rejected as
+% % solutions to the constraint imposed by the multiplicative trace. It simply
+% % remains to find the actual realization of {{cInner}} that would produce the
+% % behavior above. 
+% 
+% %subcode{opsem} include main
+% %! columnStyle = rcl
+% % cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
+% % cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
+% %        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
+% %        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))
+% 
+% \noindent where {{toffoli}} and {{cnot}} were defined for {{Pi}} in
+% \cite{rc2011}.
 % %     

### 2013-02-23 frac.tex
some notes

--- 2013-02-23 frac.tex
@@
 
+\roshan{The above para is too technical in a weird way. Its unclear
+  what it contributes. If one wants to make a point of computational
+  witnesses of isomorphisms its easier to explain and cite in terms of
+  Andreas Blass and Fiore -- and more generally by Barry Mazur. I say
+  this also because it seems that HOTT is more concerned with
+  topological structure of equalities, even though they tend to start
+  their literature with some talk of intentional specifications.}
+
 In the computer science community, the theory and practice of type
@@
 mechanism from which more conventional, i.e., irreversible computation, is
-derived. In more detail, in our own previous
+derived. In our own previous
 work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
@@
 symmetric monoidal categories (which are the backbone of models for linear
-logic and quantum computation). We have also invested a significant amount of
-time in using {{Pi}} as an actual programming language and established
-several idioms for programming large classes of interesting (recursive)
-programs in {{Pi}}.
+logic and quantum computation). 
+%% We have also invested a significant amount of time in using {{Pi}} as
+%% an actual programming language and established several idioms for
+%% programming large classes of interesting (recursive) programs in
+%% {{Pi}}.
 
@@
 {{Pi}}. This extension is of fundamental importance in all the originating
-research areas. In physics, it allows a process or observer to be treated as
-``data'' that can be transformed by higher-level processes or observed by
-meta-level observers. In mathematics and logic, it allows the equivalence
-between different proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed
-as an isomorphism (of a higher type)
+research areas. 
+%% In physics, it allows a process or observer to be treated as
+%% ``data'' that can be transformed by higher-level processes or observed by
+%% meta-level observers. 
+In physics, it allows for quantum states to be viewed as processes and
+processes to be viewed as states, such as with the Choi-Jamiolkowski
+isomorphism \cite{choi1975completely,jamiolkowski1972linear}.  In
+mathematics and logic, it allows the equivalence between different
+proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed as an
+isomorphism (of a higher type)
 $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
-higher-order types allow code to abstract over other code fragments as well
-as the manipulation of code as data and data as code.
+higher-order types allow code to abstract over other code fragments as
+well as the manipulation of code as data and data as code.
+
+\roshan{Again, HOTT seems jarring. Saying something about logic in
+  this way however seems useful. Maybe one can use one of the examples
+  of equivalence between proofs that Baez's Rosetta stone uses. }
 

### 2013-02-23 frac.tex
_

@@
 Technically speaking, obtaining a higher-order extension requires the
-construction of a \emph{compact closed category} from the underlying monoidal
+construction of a 
+%\emph{compact closed category}
+\emph{closed category} 
+%
+from the underlying monoidal
 category for~{{Pi}}. Although the general idea of such a construction is
@@
 corresponds to negative information, which turns out to be the basis of
-first-class functions. The remainder of the paper reviews {{Pi}} and then
+first-class functions. \roshan{The last sentence is terrible.} 
+The remainder of the paper reviews {{Pi}} and then
 introduces the syntax and semantics of the extension with fractional
@@
 % value types, b ::= 1 | b + b | b * b 
-\noindent Types include the unit type {{1}}, sum types {{b1 + b2}}, and
+\noindent 
+Types include the unit type {{1}}, sum types {{b1+b2}}, and
 products types {{b1*b2}}.  Values includes {{()}} which is the only value of
@@
 {{eta*_b}} and {{eps*_b}}:
+\begin{multicols}{3}
+  \dgm{eta_times2}
+
+\columnbreak
+
 %subcode{bnf} include main
@@
 %% XXX: diagrams need unit wires(?)
-\begin{multicols}{2}
-  \dgm{eta_times2}
+%% Roshan: We should just be clear that unit wires are dropped. Dgms
+%% including them dont communicate any additional information and
+%% appear crowded.
+
+\columnbreak
 
@@
 
+\vspace{-10pt}
 \noindent From a programming perspective, we think of a value {{1/v}} as a

### 2013-02-24 frac.tex
_
shortened citations

--- 2013-02-23 frac.tex
@@
 \newcommand{\amr}[1]{\textsc{Amr says:} #1}
+
+%% \newcommand{\roshan}[1]{} 
 \newcommand{\roshan}[1]{\textsc{Roshan says:} 
@@
 {{Pi}}. This extension is of fundamental importance in all the originating
-research areas. In physics, it allows a process or observer to be treated as
-``data'' that can be transformed by higher-level processes or observed by
-meta-level observers. In mathematics and logic, it allows the equivalence
-between different proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed
-as an isomorphism (of a higher type)
+research areas. 
+%% In physics, it allows a process or observer to be treated as
+%% ``data'' that can be transformed by higher-level processes or observed by
+%% meta-level observers. 
+In physics, it allows for quantum states to be viewed as processes and
+processes to be viewed as states, such as with the Choi-Jamiolkowski
+isomorphism \cite{choi1975completely,jamiolkowski1972linear}.  In
+mathematics and logic, it allows the equivalence between different
+proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed as an
+isomorphism (of a higher type)
 $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
-higher-order types allow code to abstract over other code fragments as well
-as the manipulation of code as data and data as code.
+higher-order types allow code to abstract over other code fragments as
+well as the manipulation of code as data and data as code.
+
+\roshan{Again, HOTT seems jarring. Saying something about logic in
+  this way however seems useful. Maybe one can use one of the examples
+  of equivalence between proofs that Baez's Rosetta stone uses. }
 
 Technically speaking, obtaining a higher-order extension requires the
-construction of a \emph{compact closed category} from the underlying monoidal
+construction of a 
+%\emph{compact closed category}
+\emph{closed category} 
+%
+from the underlying monoidal
 category for~{{Pi}}. Although the general idea of such a construction is
@@
 
+{\small
 \bibliographystyle{splncs03} 
 \bibliography{cites}
+}
 \end{document}

### 2013-02-24 frac.tex
intro

--- 2013-02-23 frac.tex
@@
 \newcommand{\amr}[1]{\textsc{Amr says:} #1}
-\newcommand{\roshan}[1]{\textsc{Roshan says:} 
-  \textit{#1}
-}
+
+\newcommand{\roshan}[1]{} 
+%% \newcommand{\roshan}[1]{\textsc{Roshan says:} 
+%%   \textit{#1}
+%% }
 
@@
 \title{Fractional Types}
-\author{Roshan P. James \and Zachary Sparks \and Jacques Carette \and Amr Sabry}
-\institute{Indiana University and McMaster University}
+\author{Roshan P. James$^{1}$ \and Zachary Sparks$^{1}$ \and Jacques Carette$^{2}$ \and Amr Sabry$^{1}$}
+\institute{$^{(1)}$~Indiana University \qquad $^{(2)}$~McMaster University}
 \maketitle
@@
 communities (physics, mathematics, and computer science) towards replacing
-\emph{equalities} by \emph{isomorphisms}. The combined programme 
-has sparked a huge amount of research that unveiled
-new and surprising connections between geometry, algebra, logic, and
-computation (see~\cite{baez2011physics} for an overview of some of the 
-connections).
+\emph{equalities} by \emph{isomorphisms}. The combined programme has sparked
+a significant amount of research that unveiled new and surprising connections
+between geometry, algebra, logic, and computation (see~\cite{baez2011physics}
+for an overview of some of the connections).
 
@@
 Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
-the laws of physics as fundamentally related to computation. In more detail,
-the great majority of the laws of physics are formulated as equalities
-between different physical observables, such as Newton's second law of
-classical mechanics ($F=\dot{p}$).
-The insight of Landauer, Feynman, etc. is that
-\emph{different} physical observables should not be related by an
-\emph{equality} but more appropriately by an \emph{isomorphism} that
-witnesses, explains, and models the process of transforming one observable to
-the other.
+the laws of physics as fundamentally related to computation. The great
+majority of these laws are formulated as equalities between different
+physical observables which is unsatisfying: \emph{different} physical
+observables should not be related by an \emph{equality}. It is more
+appropriate to relate them by an \emph{isomorphism} that witnesses, explains,
+and models the process of transforming one observable to the other.
 
@@
 
+%% AMR: I am not sure how to take this comment into account. I think we
+%% should make a specific connection to Martin-Lof type theory to at least
+%% justify the mention of groupoids in the conclusion. I agree though that
+%% this is a very technical point and I do find it hard to express the point
+%% in a couple of sentences. Suggestions welcome.
+
 In the computer science community, the theory and practice of type
@@
 work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
-notion of type isomorphism as the ``right'' starting point for a theory of
-computation. The central technical contribution was the development of a
-family of programming languages, {{Pi}} with various superscripts, in which
-every computation is an isomorphism. Other major results are that all such
-computations preserve the information-theoretic entropy, that the languages
-are logically reversible, and that they have models in various flavors of
-symmetric monoidal categories (which are the backbone of models for linear
-logic and quantum computation). 
+notion of type isomorphism and developed from it a family of programming
+languages, {{Pi}} with various superscripts, in which computation is an
+isomorphism preserving the information-theoretic entropy.
+%% Moreover these languages have models in various flavors of symmetric
+%% monoidal categories (which are the backbone of models for linear logic and
+%% quantum computation).
 %% We have also invested a significant amount of time in using {{Pi}} as
@@
 processes to be viewed as states, such as with the Choi-Jamiolkowski
-isomorphism \cite{choi1975completely,jamiolkowski1972linear}.  In
-mathematics and logic, it allows the equivalence between different
-proofs of type $\texttt{Id}_A(a,b)$ to itself be expressed as an
-isomorphism (of a higher type)
-$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
-higher-order types allow code to abstract over other code fragments as
-well as the manipulation of code as data and data as code.
+isomorphism~\cite{choi1975completely,jamiolkowski1972linear}.  In mathematics
+and logic, it allows the equivalence between different proofs of type
+$\texttt{Id}_A(a,b)$ to itself be expressed as an isomorphism (of a higher
+type) $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
+higher-order types allow code to abstract over other code fragments as well
+as the manipulation of code as data and data as code.
 
@@
 Technically speaking, obtaining a higher-order extension requires the
-construction of a 
-%\emph{compact closed category}
-\emph{closed category} 
-%
-from the underlying monoidal
+construction of a \emph{closed category} from the underlying monoidal
 category for~{{Pi}}. Although the general idea of such a construction is
 well-understood, the details of adapting it to an actual programming language
-are subtle.  Our main novel
-technical device to achieving the higher-order extension is a
-\emph{fractional type}, which we so named because of its duality with
-conventional product types.  By adding a formal (linear) dual to (type
-theoretic) products, we can internalize the notion that ``taking an input''
-corresponds to negative information, which turns out to be the basis of
-first-class functions. \roshan{The last sentence is terrible.} 
-The remainder of the paper reviews {{Pi}} and then
-introduces the syntax and semantics of the extension with fractional
-types. We then study the properties of the extended language and establish
-its expressiveness via several constructions and examples that exploit its
-ability to model higher-order computations.
+are subtle.  Our main novel technical device to achieving the higher-order
+extension is a \emph{fractional type} which represents \emph{negative
+  information} and which is so named because of its duality with conventional
+product types.  
+%% By adding a formal (linear) dual to (type theoretic) products, we can
+%% internalize the notion that ``taking an input'' corresponds
+%% to negative information, which turns out to be the basis of first-class
+%% functions.  
+%% \roshan{The last sentence is terrible.}
+The remainder of the paper reviews {{Pi}} and then introduces the syntax and
+semantics of the extension with fractional types. We then study the
+properties of the extended language and establish its expressiveness via
+several constructions and examples that exploit its ability to model
+higher-order computations.
 

### 2013-02-24 frac.tex
global revision

@@
 % values, v ::= () | left v | right v | (v, v)
+
 \noindent and these are classified by ordinary types:
@@
 % value types, b ::= 1 | b + b | b * b
+
 \noindent 
@@
 products types {{b1*b2}}.  Values includes {{()}} which is the only value of
-type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
+type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,
 and {{(v1,v2)}} which builds a value of product type. But these values should
@@
 %dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor
+
 \noindent Each line of the above table introduces a pair of dual
-constants\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
+constants\footnote{where {{swap*}} and {{swap+}} are self-dual.} that witness
 the type isomorphism in the middle.  These are the base (non-reducible) terms
@@
 separate syntactic categories which do not intermix. The semantics of the
-language come when these are made to interact at the ``top level'', a new
-syntactic category:
+language come when these are made to interact at the ``top level'' via
+\emph{application}:
 %subcode{bnf} include main
 % top level term, l ::= c v
-\noindent We refer to this as \emph{application}.
 
@@
 
-Mathematically, {{Pi}} can be interpreted in the category of finite sets and
-bijections: each value type denotes a finite set of a size calculated by
-viewing the types as natural numbers and each combinator {{c : b1 <-> b2}}
-denotes a bijection between the sets denoted by~{{b1}}
-and~{{b2}}. Operationally, the semantics of {{Pi}} is given using two
-mutually recursive interpreters: one going forward and one going
-backwards. The use of {{sym}} switches control from one evaluator to the
-other. We will present the operational semantics in the next section along
-with the extension with fractional types and values. For now, we state
-without proof that the evaluation of well-typed combinators always terminates
-and that {{Pi}} is logically reversible, i.e., that for all combinators 
-{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
-evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
-{{c v2}} produces {{v1}}.
+The language presented above, at the type level, models a commutative ringoid
+where the multiplicative structure forms a commutative monoid, but the
+additive structure is just a commutative semigroup.  Note that the version of
+{{Pi}} that include the empty type with its usual laws exactly captures, at
+the type level, the notion of a \emph{semiring} (occasionally called a
+\emph{rig}) where we replace equality by isomorphism.  Semantically, {{Pi}}
+models a \emph{bimonoidal category} whose simplest example is the category of
+finite sets and bijections. In that interpretation, each value type denotes a
+finite set of a size calculated by viewing the types as natural numbers and
+each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
+denoted by~{{b1}} and~{{b2}}. (We discuss the operational semantics in
+conjuction with our extension in the next section.)
+
+% Operationally, the
+% semantics of {{Pi}} is given using two mutually recursive interpreters: one
+% going forward and one going backwards. The use of {{sym}} switches control
+% from one evaluator to the other. We will present the operational semantics in
+% the next section along with the extension with fractional types and
+% values. For now, we state without proof that the evaluation of well-typed
+% combinators always terminates and that {{Pi}} is logically reversible, i.e.,
+% that for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 :
+%     b2}} we have the forward evaluation of {{c v1}} produces {{v2}} iff the
+% backwards evaluation of {{c v2}} produces {{v1}}.
 
@@
 
-The language presented above, at the type level, models a commutative ringoid
-where the multiplicative structure forms a commutative monoid, but the
-additive structure is just a commutative semigroup.  Note that the version of
-{{Pi}} that include the empty type with its usual laws exactly captures, at
-the type level, the notion of a \emph{semiring} (occasionally called a
-\emph{rig}) where we replace equality by isomorphism.  This can also be also
-compared to \emph{bimonoidal categories}.  As is usual with type theories,
-the coherence conditions (pentagonal and triangle identity) are not explicit,
-but would correspond to certain identity types being trivial.
+%% As is usual with type theories,
+%% the coherence conditions (pentagonal and triangle identity) are not explicit,
+%% but would correspond to certain identity types being trivial.
 
@@
 
-The language {{Pi}} models isomorphisms of values rather well.  However,
-purposefully, it has no distinguished notion of \emph{input} or
-\emph{output}; more precisely, because of strict preservation of information,
-these two concepts coincide.  This is the root of reversibility. But if we
-want to model functions of any flavor as values, we need to differentiate
-between these notions. We also clearly need to relax the strict separation
-between values and isomorphisms as the notion of ``higher-order'' requires
-that isomorphisms be represented as values. Our idea, inspired by
+The language {{Pi}} models isomorphisms of values rather well, and as we
+established before~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012}, it is
+logically reversible and each computation it expresses preserves the
+information-theoretic entropy. However, purposefully, it has no distinguished
+notion of \emph{input} or \emph{output}; more precisely, because of strict
+preservation of information, these two concepts coincide.  This is the root
+of reversibility. But if we want to model functions of any flavor as values,
+we need to differentiate between these notions. Our idea, inspired by
 computational dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
@@
 % values, v ::= ... | 1/v
-For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
-where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
-however some informal intuition about the \emph{meaning} of such fractional
-types and values. (See Sec.~\ref{sec:cat} for more discussion about this
-point.) In more detail, given a symmetric monoidal category modeling a
-certain language (like {{Pi}}), one can obtain a higher-order extension of
-the language by extending the category to a \emph{compact closed} one, i.e.,
-a category in which morphisms are representable as objects. The key to this
-extension is to add dual objects (in our case the fractionals) satisfying the
-following isomorphism witnessed by two new type indexed combinators 
-{{eta*_b}} and {{eps*_b}}:
+
+\noindent For a given type {{b}}, the values of type {{1/b}} are of the form
+{{1/v}} where {{v : b}}. Note that {{1/v}} is purely formal, as is
+{{1/b}}.
+
+Semantically, we are extending the symmetric monoidal category modeling
+{{Pi}} to a \emph{compact closed} one, i.e., a category in which morphisms
+are representable as objects. In such a setting, the new dual objects (i.e.,
+the fractionals) must satisfy the following isomorphism witnessed by two new
+type indexed combinators {{eta*_b}} and {{eps*_b}}:
 \begin{multicols}{3}
@@
 
-From the categorical perspective, versions of compact closed categories were
-introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
-and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the
-generalization of monoidal categories to model quantum computation.  The
-first and most intuitive example of such categories is the category of finite
-sets and relations. In that case, the dual operation (i.e., the fractional
-type) collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
+Historically, versions of compact closed categories were introduced by
+Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
+Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the generalization of
+monoidal categories to model quantum computation.  The first and most
+intuitive example of such categories is the category of finite sets and
+relations. In that case, the dual operation (i.e., the fractional type)
+collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
 interpretation of fractionals, this is still an interesting category as it
@@
 is reversible via the standard \emph{converse} operation on a relation.
-Furthermore, as any relation between {{b1}} and {{b2}} can be
-represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
-example is the category of finite dimensional vector spaces and linear maps
-(or even the category of finite dimensional Hilbert spaces). In such
-categories the dual operation takes a vector space (consisting of ``column
-vectors'') to the dual vector space (consisting of ``row vectors''). While
-these are isomorphic, but they are nevertheless quite different,
-being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the
-category provides a more refined model in which isomorphic
-negative-information values and positive-information are not identified.
+Furthermore, as any relation between {{b1}} and {{b2}} can be represented as
+a subset of {{b1 * b2}}, i.e., as a value.
 
 In this section, we provide a simple semantics of {{PiTF}} in the category of
-finite sets and relations. We interpret each combinator 
-{{c : b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
-{{b2}} respectively. As we illustrate, this semantics allows \emph{arbitrary}
+finite sets and relations. We interpret each combinator {{c : b1 <-> b2}} as
+a \emph{relation} between the sets denoted by {{b1}} and {{b2}}
+respectively. As we illustrate, this semantics allows \emph{arbitrary}
 relations (including the empty relation and relations that are not
@@
 
-\begin{definition}[Denotation of Value Types {{ [[ b ]] }}]
+\begin{definition}[Denotation of Value Types]
 \label{chx:def:denot}
@@
 Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
-as specified below. 
+as specified below.~\footnote{
+In the interest of brevity, we have treated the type annotations in the base
+isomorphisms as implicit; for example, the rule for {{assocr*}} should really
+read:
+
+%subcode{proof} include main
+%@ v1 in b1
+%@ v2 in b2
+%@ v3 in b3
+%@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))
+}
 %subcode{proof} include main
@@
 
-The operational semantics define a relation rather than a function, but the
-goal is for them to still be reversible; in particular, we wish for the
-relation to mode check so that for any combinator {{c}} we can treat either
-its left or its right side as an input; this allows us to define {{sym}} as
-just relational converse, which amounts to
-an argument swap. We can then give an operational interpretation to the
+The semantics above defines a relation rather than a function. It is still
+reversible in the sense that it ``mode-checks''. In other words, for any
+combinator {{c}} we can treat either its left or its right side as an input;
+this allows us to define {{sym}} as just relational converse, which amounts
+to an argument swap. We can then give an operational interpretation to the
 semantic relation by defining two functions: a forward evaluator and a
-backwards evaluator that exercise the relation in opposite directions,
-but which now can return a \emph{set} of results. It is
-straightforward to check that if the forward evaluation of {{c v1}} produces
-{{v2}} as a possible answer then the backwards evaluation of {{c v2}}
-produces {{v1}} as a possible answer.  One can check that all relations
-for {{Pi}} are total functional relations whose converse are also total 
-functional relations -- aka isomorphisms.  However {{eta*_b}} relates
-{{()}} to multiple values (one for each value of {{b}}), and is thus
-not functional; {{eps*_b}} is exactly its relational converse, as expected.
-
-In the interest of brevity, we have treated the type annotations in the base
-isomorphisms as implicit; for example, the rule for {{assocr*}} should really
-read:
-
-%subcode{proof} include main
-%@ v1 in b1
-%@ v2 in b2
-%@ v3 in b3
-%@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))
+backwards evaluator that exercise the relation in opposite directions, but
+which now can return a \emph{set} of results. This operational view exploits
+the usual conversion of a relation from a subset of {{A * B}} to a function
+from {{A}} to the powerset of {{B}} whose composition is given by the Kleisli
+composition in the powerset monad. It is straightforward to check that if the
+forward evaluation of {{c v1}} produces {{v2}} as a possible answer then the
+backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.  One
+can check that all relations for {{Pi}} are total functional relations whose
+converse are also total functional relations -- aka isomorphisms.  However
+{{eta*_b}} relates {{()}} to multiple values (one for each value of {{b}}),
+and is thus not functional; {{eps*_b}} is exactly its relational converse, as
+expected.
 
@@
 
-In the following, we use {{bool}} as an abbreviation of {{1+1}} 
-with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
-abbreviation for {{right ()}}. In addition, for the sake of readability, the
-circuit diagrams in this section will elide trivial unit wires, such as in
-{{eta}} and {{epsilon}}.
+Having introduced the syntax of our extended language, its semantics in the
+category of sets and relations, and its operational semantics using forward
+and backwards interpreters, we now illustrate its expressiveness as a
+higher-order programming language. In the following presentation consisting
+of numerous programming examples, we use {{bool}} as an abbreviation of
+{{1+1}} with {{true}} as an abbreviation for {{left ()}} and {{false}} as an
+abbreviation for {{right ()}}. In addition, instead of presenting the code
+using the syntax of {{Pi}}, we use circuit diagrams that are hopefully more
+intuitive. Each diagrams represents a combinator whose evaluation consists of
+propagating values along the wires.  For the sake of readability, we omit
+obvious re-shuffling circuitry and elide trivial unit wires which carry no
+information.\footnote{The full code is available upon request. We plan on
+  releasing it as soon as possible.}
 
@@
 
-As motivated, with fractional types, we can define a value that denotes a
-first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
-pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
-and a value {{v2 : b2}}. In other words, it corresponds to a function or
-relation which when ``given'' a value {{v1 : b1}} ``releases'' the value 
-{{v2 : b2}}. To emphasize this view, we introduce the abbreviation:
+The most basic additional expressiveness of {{PiTF}} over {{Pi}} is the
+ability to express relations as values. Indeed a value of type {{1/b1 * b2}}
+is a pair of a \emph{constraint} that can only be satisfied by some 
+{{v1 : b1}} and a value {{v2 : b2}}. In other words, it corresponds to a
+function or relation which when ``given'' a value {{v1 : b1}} ``releases''
+the value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:
 
@@
 any combinator {{c : b1 <-> b2}} into a (constant) value of 
-type {{b1 :-* b2}} as shown below:
+type {{b1 :-* b2}} as shown below on the left:
 
@@
 \end{multicols}
-\noindent  Dually, as illustrated above, we also have a {{coname}} combinator
-which when given an answer {{v2 : b2}}, and a request for the corresponding
-input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this as a un-needed 
-computation.
-
-As a simple example, consider {{swap+}} denoting boolean
-negation.  It is straightforward to check that the combinator 
-{{name~swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:
-
-%subcode{proof} include main
-%@ ~
-%@@ () ~~(name~swap+)~~ (1/false,true)
-%
-%@ ~
-%@@ () ~~(name~swap+)~~ (1/true,false)
-
-\noindent The relation {{name~swap+}} maps {{()}} to a value which can be used
-to convert any {{false}} to {{true}} and vice-versa. We formalize this
-``apply'' combinator below.
+\noindent Dually, as illustrated above on the right, we also have a
+{{coname}} combinator which when given an answer {{v2 : b2}}, and a request
+for the corresponding input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this
+as a un-needed computation.
+
+More generally any combinator manipulating values of type {{b}} can be turned
+into a combinator manipulating values of type {{1/b}} and vice-versa. This is
+due to the fact that fractionals types satisfy a self-dual involution
+relating {{b}} and {{1/(1/b)}}:
+
+%subcode{opsem} include main
+%! columnStyle = rcl
+% doubleDiv &:& b <-> 1/(1/b)
+% doubleDiv &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
+
+% jacques: elide for now, to save space, as this diagram is backwards.
+% \dgm{involution-times}
+
+% As a simple example, consider {{swap+}} denoting boolean
+% negation.  It is straightforward to check that the combinator 
+% {{name~swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:
+
+% %subcode{proof} include main
+% %@ ~
+% %@@ () ~~(name~swap+)~~ (1/false,true)
+% %
+% %@ ~
+% %@@ () ~~(name~swap+)~~ (1/true,false)
+
+% \noindent The relation {{name~swap+}} maps {{()}} to a value which can be used
+% to convert any {{false}} to {{true}} and vice-versa. We formalize this
+% ``apply'' combinator below.
 
@@
 
-We are now a small step from implementing various higher-order
-combinators that manipulate functions or relations. In particular, we
-can \emph{apply} and \emph{compose} values representing relations.
-\emph{apply} is shown below:
+We are now a small step from implementing various higher-order combinators
+that manipulate functions or relations. In particular, we can \emph{apply},
+\emph{compose}, \emph{curry}, and \emph{uncurry} values representing
+relations.  We first show the realization of \emph{apply}:
 
@@
 % curry &=& (id * recip) (;) (id * swap*) (;) assocl*
+
 \noindent That {{recip}} is the heart of currying seems quite remarkable.
-%%%%%%%%%%%%
-\subsection{Duality}
-
-The final remarkable property of fractional types is that they
-are involutive:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% doubleDiv &:& b <-> 1/(1/b)
-% doubleDiv &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
-
-% jacques: elide for now, to save space, as this diagram is backwards.
-% \dgm{involution-times}
 
@@
 operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
-language like {{Pi}} there are two natural families of trace operators
-that can be defined, an additive family (explored in detail in our previous
-work~\cite{rc2011}) and a multiplicative family, which we will focus on now.
-
-Given a computation {{c:b1 * b2 <-> b1 * b3}}, we
-want to cancel the common type {{b1}}, to produce a new combinator
-{{trace*~c:b2<->b3}}. For its evaluation, we are only
-given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
-consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
-The needed value {{v1 : b1}} cannot be arbitrary: it must be such that it
-is the value produced as the first component of the result.  In general, there
-may be several such fixed-point values, or none.  A few examples should help
-explain the subtleties.
-
-\begin{example}
-Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.
-
-\noindent Using {{trace*}}, we construct the
-combinator {{trace* c : 1 <-> 1}}. Applying this to 
-{{() : 1}} requires us to find a value {{b : bool}} such that 
-{{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
-has two values, there are two values {{false}} and {{true}} that satisfy the 
-constraint. In other words, {{() ~~(trace* c)~~() }}.
-\end{example}
-
-\begin{example}
-\label{ch3:ex:annihilate}
-Consider a small variation on the combinator {{c}} above:
-
-{{c : bool * 1 <-> bool * 1}}
-
-{{c = swap+ (*) id }}
-
-\noindent which negates the boolean component of its incoming
-pair. Using the multiplicative trace operator, we can construct the
-combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
-combinator to {{() : 1}} requires us to find a value {{b : bool}} such
-that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
-has no fixed points.  Operationally, the evaluation of
-such a combinator should produce no value (were it to even terminate).
-\end{example}
+language like {{Pi}} there are two natural families of trace operators that
+can be defined, an additive family (explored in detail in our previous
+work~\cite{rc2011}) and a multiplicative family, which is expressible using
+fractionals.
+
+The idea of the multiplicative trace operator is as follows. We are given a
+computation {{c : b1 * b2 <-> b1 * b3}}, and we build a ``looping'' version
+which feeds the output value of type {{b1}} back as an input. Effectively,
+this construction cancels the common type {{b1}}, to produce a new combinator
+{{trace*~c:b2<->b3}}.
 
@@
 
-We can use the relational semantics of definition~\ref{def:relational-PiTF}
-to calculate the result of applying
-{{trace*_{bool} (swap+ (*) id)}} to {{false}}.  We use the usual conversion
-of a relation from a subset of {{A * B}} to a function from {{A}} to the
-powerset of {{B}} (and Kleisli composition in the powerset monad) for
-the computation:
+As an example, we can use the operational semantics (outlined in the previous
+section) to calculate the result of applying {{trace*_{bool} (swap+ (*) id)}}
+to {{false}}.
 %subcode{opsem} include main
@@
 
-\noindent  The computation with {{true}} gives the same result.  This
-implies that although our semantics is formally reversible, it 
-does not result in isomorphisms.
-Nevertheless, the next subsection illustrates why this is still useful.
+\noindent The computation with {{true}} gives the same result. This confirms
+that although our semantics is formally reversible, it does not result in
+isomorphisms.
+
+More abstractly, the evaluation of {{trace*~c : b2 <-> b3}} must ``guess'' a
+value of type {{b1}} to be provided to to the inner combinator 
+{{c : b1 * b2 <-> b1 * b3}}. 
+This value {{v1 : b1}} cannot be arbitrary: it must be such that it is the value
+produced as the first component of the result.  In general, there may be
+several such fixed-point values, or none. Indeed with a little ingenuity 
+(see the next section), one can 
+express any desired relation by devising a circuit which keeps the desired pairs as 
+valid fixed-points.
+
+% \begin{example}
+% Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.
+
+% \noindent Using {{trace*}}, we construct the
+% combinator {{trace* c : 1 <-> 1}}. Applying this to 
+% {{() : 1}} requires us to find a value {{b : bool}} such that 
+% {{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
+% has two values, there are two values {{false}} and {{true}} that satisfy the 
+% constraint. In other words, {{() ~~(trace* c)~~() }}.
+% \end{example}
+
+% \begin{example}
+% \label{ch3:ex:annihilate}
+% Consider a small variation on the combinator {{c}} above:
+
+% {{c : bool * 1 <-> bool * 1}}
+
+% {{c = swap+ (*) id }}
+
+% \noindent which negates the boolean component of its incoming
+% pair. Using the multiplicative trace operator, we can construct the
+% combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
+% combinator to {{() : 1}} requires us to find a value {{b : bool}} such
+% that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
+% has no fixed points.  Operationally, the evaluation of
+% such a combinator should produce no value (were it to even terminate).
+% \end{example}
 
@@
 
-\noindent We can define a combinator {{c_R}} whose denotation is {{R}},
-by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
-for some type {{a}} such that {{c_R = trace* cInner}}:
+\noindent We can define a combinator {{c_R}} whose denotation is {{R}}, by
+defining a combinator {{cInner : (a * bool) <-> (a * bool)}} for some type
+{{a}} such that {{c_R = trace* cInner}}. The basic requirement of {{cInner}}
+is that for each desired pair {{(v1,v2)}} in {{R}}, it maps {{(a0,v1)}} to
+{{(a0,v2)}} for some value {{a0}} and that for each pair {{(v1,v2)}} that is
+\emph{not} in {{R}}, it maps {{(a1,v1)}} to {{(a2,v2)}} for \emph{different}
+values {{a1}} and {{a2}}.
 
@@
 
-%subcode{opsem} include main
-%! columnStyle = rcl
-% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
-% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
-%        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
-%        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))
-
-\noindent where {{toffoli}} and {{cnot}} were defined for {{Pi}} in
-\cite{rc2011}.
+% %subcode{opsem} include main
+% %! columnStyle = rcl
+% % cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
+% % cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
+% %        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
+% %        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))
+
+% \noindent where {{toffoli}} and {{cnot}} are the conventional reversible gates (their definition in {{Pi}} is in~\cite{rc2011}).
 % %     
@@
 % implementation of a SAT solver.
+
+For small examples, finding such combinators by trial and error is a
+relatively straightforward but tedious task. It is, however, possible to
+automate this task by expressing what is essentially a reversible SAT solver
+as follows. In the usual setting, an instance of SAT is a function~{{f}}
+which, when given some boolean inputs, returns {{true}} or {{false}}. The
+function returns {{true}} when the inputs satisfy the constraints imposed by
+the structure of {{f}} and a solution to the SAT problem is the set of all
+inputs on which {{f}} produces {{true}}. The basic idea of our construction
+is to use {{trace*}} to annihilate values that fail to satisfy the
+constraints represented by the SAT instance {{f}}.
 
@@
 ``canonical'' compact closed category, that is finite dimensional vector
-spaces and linear maps\cite{Selinger:2011:FDH:1942319.1942398,Hasegawa:2008:FDV:1805839.1805859} over fields of characteristic $0$. Let us fix an arbitrary
-field $k$ of characteristic $0$.
+spaces and linear
+maps~\cite{Selinger:2011:FDH:1942319.1942398,Hasegawa:2008:FDV:1805839.1805859}
+over fields of characteristic $0$ (or even the category of finite dimensional
+Hilbert spaces). Let us fix an arbitrary field $k$ of characteristic $0$.
 Then each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
@@
   \item one can then validate certain desirable properties: {{1/(1/b)}} is
-    isomorphic to {{b}}; and {{eps*_b}} corresponds to a bilinear form which
+    isomorphic to~{{b}}; and {{eps*_b}} corresponds to a bilinear form which
     maps a dual vector and a vector to a field element.
 \end{itemize}
+In such categories, the fractional type is given a non-trivial
+interpretation. Indeed, while the space of column vectors is isomorphic to
+the space of row vectors, they are nevertheless quite different, being
+$(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the category
+provides a more refined model in which isomorphic negative-information values
+and positive-information are not identified.
 
 Although this semantics appears to have ``better'' properties than the
-relational one, we argue that it is not \emph{the} ``perfect'' semantics. By
-including the zero vector, the language has
-morphisms that do not correspond to isomorphisms (in particular it allows
-partial morphisms by treating the {{0}} element of the zero-dimensional 
-vector space as a canonical ``undefined'' value).  It is also difficult to
-reconcile the interpretation with the view that the types correspond
-to the (positive) rational numbers, something we are actively seeking. What 
-would really be a ``perfect'' interpretation is one in which we
-can only express {{Pi}}-isomorphisms as first class values and in which the
-types are interpreted in a way that is consistent with the rational
-numbers.
+relational one, we argue that it is not yet \emph{the} ``perfect''
+semantics. By including the zero vector, the language has morphisms that do
+not correspond to isomorphisms (in particular it allows partial morphisms by
+treating the {{0}} element of the zero-dimensional vector space as a
+canonical ``undefined'' value).  It is also difficult to reconcile the
+interpretation with the view that the types correspond to the (positive)
+rational numbers, something we are actively seeking. What would really be a
+``perfect'' interpretation is one in which we can only express
+{{Pi}}-isomorphisms as first class values and in which the types are
+interpreted in a way that is consistent with the rational numbers.
 
@@
 
-We now proceed to explain how each {{PiTF}} type denotes a groupoid. First,
+\noindent We now proceed to explain how each {{PiTF}} type denotes a groupoid. First,
 each {{Pi}} type denotes a set as follows.

### 2013-02-24 frac.tex
12 pages
add a short paragraph on dependent types

@@
 %! empty = \emptyset
+%! Sigma = \Sigma
 %! eta = \eta
@@
 %@@ v ~~uniti~~ ((), v)
-%----
+%
 %@ ~
@@
 %@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
-%----
+%
 %@ ~
@@
 
+Another promising approach is the use of dependent types for {{eps*_b}} and
+{{eta*_b}}; more precisely, {{eps*_b}} would have type 
+{{Sigma (v:b) (1/v,v) <-> 1}} where {{(1/v,v)}} here denotes a singleton type.
+This extra precision appears to restrict combinators of {{PiTF}} back to
+denoting only isomorphisms.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-24 frac.tex
minor things

@@
 is reversible via the standard \emph{converse} operation on a relation.
-Furthermore, as any relation between {{b1}} and {{b2}} can be represented as
+Furthermore, any relation between {{b1}} and {{b2}} can be represented as
 a subset of {{b1 * b2}}, i.e., as a value.
@@
 reversible in the sense that it ``mode-checks''. In other words, for any
-combinator {{c}} we can treat either its left or its right side as an input;
+combinator~{{c}} we can treat either its left or its right side as an input;
 this allows us to define {{sym}} as just relational converse, which amounts
 to an argument swap. We can then give an operational interpretation to the
-semantic relation by defining two functions: a forward evaluator and a
+semantic relation by defining two interpreters: a forward evaluator and a
 backwards evaluator that exercise the relation in opposite directions, but
@@
 is to use {{trace*}} to annihilate values that fail to satisfy the
-constraints represented by the SAT instance {{f}}.
+constraints represented by the SAT instance {{f}}. (The accompanying code
+details the construction of such a solver.)
 

### 2013-02-24 frac.tex
hasegawa

@@
 spaces and linear
-maps~\cite{Selinger:2011:FDH:1942319.1942398,Hasegawa:2008:FDV:1805839.1805859}
+maps~\cite{Selinger:2011:FDH:1942319.1942398,hasegawa2008finite}
 over fields of characteristic $0$ (or even the category of finite dimensional

### 2013-02-24 frac.tex
_

@@
 
+\roshan{In the above the, when we say that the terms of {{Pi}} are
+  proofs that certain ``shapes of values'' are isomorphic, it is
+  misleading. In {{langRev}}, the value {{left ()}} may or may not be
+  isomorphic to the value {{right ()}} or {{left ()}} itself, or any
+  other value of our choice. In particular the actual syntactic
+  structure of the value has nothing to do with values being
+  isomorphic. Values are just place-holders for the situations that
+  their type represents. Their shapes have no intrinsic content. }
+
+\roshan{Further, this is not to say that we may not be able to use the
+  syntactic structure of a value for some purpose.  The prime use for
+  such a thing might be, for example, to define a notion of distance
+  between values -- such that values live in a metric space. Time and
+  time again, we have felt the need for developing such an idea. Such
+  a thing doesn't exist for the scope of this paper however. }
+
 \paragraph*{Data.}

### 2013-02-24 frac.tex
_

@@
 
+\roshan{The use of the word ``orthogonal'' is inappropriate in the
+  footnote. Adding 0 truly affects how fractionals are developed,
+  whereas when we say ``orthogonal'' we give the impression that it does
+  not really matter. }
+
 \paragraph*{Data.}
@@
 
-\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
-(value) type isomorphisms.  They are also typed, by the shape of the (value)
-type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
-the following isomorphisms:
+\roshan{It is unclear what the usage ``first-class'' and ``effect of
+  an isomorphism'' indicate. Maybe we should rephrase. }
+
+%% \paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
+%% (value) type isomorphisms.  They are also typed, by the shape of the (value)
+%% type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
+%% the following isomorphisms:
+\paragraph*{Isomorphisms.} The terms of {{Pi}} witness
+type isomorphisms of the form {{b <-> b}}. They consist of base
+isomorphisms, as defined below, and their composition.
 %subcode{bnf} include main
@@
 additive structure is just a commutative semigroup.  Note that the version of
-{{Pi}} that include the empty type with its usual laws exactly captures, at
+{{Pi}} that includes the empty type with its usual laws exactly captures, at
 the type level, the notion of a \emph{semiring} (occasionally called a

### 2013-02-24 frac.tex
_

@@
 language come when these are made to interact at the ``top level'' via
-\emph{application}:
+\emph{application}: \roshan{I understand its intent, but this last sentence is unclear.}
 %subcode{bnf} include main
@@
 conjuction with our extension in the next section.)
+
+\roshan{Suggestion : This para is a good one to delete if we run out
+  of space. This is largely general knowledge about universal algebra
+  and maybe one useful remark about the categorical structure. However
+  we have a whole paper about the categorical structure of Pi at RC
+  itself. My point is this sort of this can be a remark inserted at
+  other places in passing and it doesn't deserve a para on its own. It
+  teaches the reader nothing new about Pi. }
 
@@
 %
+% Roshan: I completely understand this and this is not my objection. This
+% is not what the above sentence is saying however. Its talking about
+% Pi not having a notion of input and output which is simply
+% confusing.
 

### 2013-02-25 frac.tex
_
small dgm fix
fixed products types typo

@@
 Types include the unit type {{1}}, sum types {{b1+b2}}, and
-products types {{b1*b2}}.  Values includes {{()}} which is the only value of
+product types {{b1*b2}}.  Values includes {{()}} which is the only value of
 type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,
-and {{(v1,v2)}} which builds a value of product type. But these values should
-be regarded as largely ancillary: we do not treat them as first-class citizens,
-and they only occur when observing the effect of an isomorphism.
+and {{(v1,v2)}} which builds a value of product type. 
+%% But these values should be regarded as largely ancillary: we do not
+%% treat them as first-class citizens, and they only occur when observing
+%% the effect of an isomorphism.
 
@@
 % AMR: relations are indeed only presented as a way to explain the semantics
+% Roshan: it reads much better now.
 

### 2013-02-25 frac.tex
fixed another typo (going one at a time here)

@@
 Types include the unit type {{1}}, sum types {{b1+b2}}, and
-product types {{b1*b2}}.  Values includes {{()}} which is the only value of
+product types {{b1*b2}}.  Values include {{()}} which is the only value of
 type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,

### 2013-02-25 frac.tex
small clarification of piTF modelled by sets and relations

--- 2013-02-25 frac.tex
@@
 Types include the unit type {{1}}, sum types {{b1+b2}}, and
-product types {{b1*b2}}.  Values includes {{()}} which is the only value of
+products types {{b1*b2}}.  Values includes {{()}} which is the only value of
 type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,
@@
 \emph{first-class constraint} that can only be satisfied if it is matched
-with an actual value {{v}} or as a \emph{pattern} representing the absence of
+with an actual value {{v}}. In other words, {{1/v}} is a \emph{pattern} representing the absence of
 some information of a specific shape (i.e., \emph{negative information}),
@@
 
-In this section, we provide a simple semantics of {{PiTF}} in the category of
-finite sets and relations. We interpret each combinator {{c : b1 <-> b2}} as
-a \emph{relation} between the sets denoted by {{b1}} and {{b2}}
-respectively. As we illustrate, this semantics allows \emph{arbitrary}
-relations (including the empty relation and relations that are not
-isomorphisms) to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat}
-we discuss more refined semantics that are more suitable for richer
+In this section, we provide a simple semantics of {{PiTF}} in the
+category of finite sets and relations, in the following way: We
+interpret each combinator {{c : b1 <-> b2}} as a \emph{relation}
+between the sets denoted by {{b1}} and {{b2}} respectively. Sets
+{{b1}} and {{b2}} can be related only if the same information measure.
+As we illustrate, this semantics allows \emph{arbitrary} relations
+(including the empty relation and relations that are not isomorphisms)
+to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat} we
+discuss more refined semantics that are more suitable for richer
 categories.

### 2013-02-25 frac.tex

@@
 Types include the unit type {{1}}, sum types {{b1+b2}}, and
-products types {{b1*b2}}.  Values includes {{()}} which is the only value of
+product types {{b1*b2}}.  Values include {{()}} which is the only value of
 type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,

### 2013-02-25 frac.tex
slight rewording

@@
 converse are also total functional relations -- aka isomorphisms.  However
-{{eta*_b}} relates {{()}} to multiple values (one for each value of {{b}}),
+{{eta*_b}} relates {{()}} to multiple values (one for each value of type {{b}}),
 and is thus not functional; {{eps*_b}} is exactly its relational converse, as

### 2013-02-25 frac.tex
_

@@
 for the corresponding input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this
-as a un-needed computation.
+as a un-needed computation (see also Sec. 3, \cite{abramsky-2008}).
 
@@
 
+\roshan{Either we should have explain the SAT solver well or we
+  shouldn't have at all. There are some original ideas in the
+  construction of that solver and the Oxford QF group for instance
+  were very interested in its details. This presentation neither
+  leaves the idea prime for future publication, and neither does it
+  explain it well.}
+
 % In the usual setting, an instance of SAT is a function~{{f}} which,
@@
 \end{itemize}
+
+\roshan{Again, if we have lost zero, we cannot talk about vector
+  spaces with the zero vector. We should alternately included it in
+  our presentation to start with. }
+
 In such categories, the fractional type is given a non-trivial
@@
 
+\roshan{The above para should simply be deleted.}
+
 Fortunately, there is promising significant work on the groupoid
@@
 groupoids is cartesian closed, this appears eminently feasible.
+
+\roshan{Again, the above para is just lose speculation. Further, the
+  fact that we say that its cartesian closed instead of monoidal
+  closed should *immediately* tell us that its the wrong model.}
 

### 2013-02-25 frac.tex
add URL for code

@@
 obvious re-shuffling circuitry and elide trivial unit wires which carry no
-information.\footnote{The full code is available upon request. We plan on
-  releasing it as soon as possible.}
+information.\footnote{The full code is available at
+\url{www.cas.mcmaster.ca/~carette/PiFractional}.}
 

### 2013-02-25 code/frac/PiFracClean.hs
partial clean up of code

@@ -1,0 +1,670 @@
+{-# LANGUAGE TypeOperators, GADTs, TypeSynonymInstances #-}
+
+module PiFracClean where
+
+import Data.List
+
+-----------------------------------------------------------------------
+-- Isomorphisms 
+
+data Recip a = Recip a deriving (Show, Eq)
+
+class Eq a => B a where
+  elems :: [a]
+
+instance B () where  
+  elems = [()]
+  
+instance (B a, B b) => B (Either a b) where
+  elems = map Left elems ++ map Right elems
+  
+instance B Bool where
+  elems = [False, True]
+
+instance (B a, B b) => B (a,b) where
+  elems = [(a,b) | a <- elems, b <- elems]
+
+instance B a => B (Recip a) where
+  elems = map Recip elems
+
+data a :<=> b where 
+-- Congruence
+  Id    :: B a => a :<=> a
+  Sym   :: (B a, B b) => (a :<=> b) -> (b :<=> a) 
+  (:.:) :: (B a, B b, B c) => (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (B a, B b, B c, B d) => (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (B a, B b, B c, B d) => (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative and commutative
+  SwapP :: (B a, B b) => Either a b :<=> Either b a
+  AssocLP  :: (B a, B b, B c) => Either a (Either b c) :<=> Either (Either a b) c 
+  AssocRP  :: (B a, B b, B c) => Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE    :: B a => ((), a) :<=> a
+  UnitI    :: B a => a :<=> ((), a)
+  SwapT :: (B a, B b) => (a,b) :<=> (b,a) 
+  AssocLT  :: (B a, B b, B c) => (a,(b,c)) :<=> ((a,b),c)
+  AssocRT  :: (B a, B b, B c) => ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  Distrib  :: (B a, B b, B c) => (Either b c, a) :<=> Either (b, a) (c, a)
[…]

### 2013-02-25 code/frac/PiFracClean.hs
further clean up

@@ -50,9 +50,6 @@ data a :<=> b where
 -- Encoding of booleans
   FoldB   :: Either () () :<=> Bool
   UnfoldB :: Bool :<=> Either () ()
--- Trace operators for looping/recursion
---  TraceP :: (B a, B b1, B b2) => (Either a b1 :<=> Either a b2) -> (b1 :<=> b2)
---  TraceT :: (B a, B b1, B b2) => ((a, b1) :<=> (a, b2)) -> (b1 :<=> b2)
 -- Eta/Psi
   EtaT :: (B a) => () :<=> (Recip a, a)
   EpsT :: (B a) => (Recip a, a) :<=> ()
@@ -77,8 +74,6 @@ adjoint Distrib = Factor
 adjoint Factor = Distrib
 adjoint FoldB = UnfoldB
 adjoint UnfoldB = FoldB
---adjoint (TraceP c) = TraceP (adjoint c)
---adjoint (TraceT c) = TraceT (adjoint c)
 adjoint EtaT = EpsT
 adjoint EpsT = EtaT
 
@@ -116,20 +111,6 @@ eval FoldB (Left ()) = return True
 eval FoldB (Right ()) = return False
 eval UnfoldB True = return (Left ())
 eval UnfoldB False = return (Right ())
-{--
-eval (TraceP c) v = do v' <- eval c (Right v)
-                       loop c v'
-    where
-      loop c (Left v) = do v' <- eval c (Left v)
-                           loop c v'
-      loop c (Right v) = return v
-eval (TraceT c) v = try elems
-    where try [] = []
-          try (a : as) = do (a',v') <- eval c (a,v)
-                            if a == a' 
-                              then v' : try as
-                              else try as
---}
 eval EtaT () = [(Recip v, v) | v <- elems]
 eval EpsT (Recip v1, v2) | v1 == v2 = [()]
                          | otherwise = error "Type error"
@@ -214,20 +195,17 @@ name f = EtaT :.: (Id :*: f)
 coname :: (B b1, B b2) => (b1 :<=> b2) -> ((b1 :-* b2) :<=> ())
 coname f = (Id :*: Sym f) :.: EpsT
 
-compose :: (B b1, B b2, B b3) => (b1 :-* b2, b2 :-* b3) :<=> (b1 :-* b3)
-compose = -- ((1/b1,b2),(1/b2,b3))
[…]

### 2013-02-25 code/frac/PiFracClean.hs
clean version of code

@@ -1,7 +1,8 @@
-{-# LANGUAGE TypeOperators, GADTs, TypeSynonymInstances #-}
+{-# LANGUAGE TypeOperators, GADTs #-}
 
 module PiFracClean where
 
+import Prelude hiding (curry,uncurry)
 import Data.List
 
 -----------------------------------------------------------------------
@@ -207,49 +208,44 @@ apply = SwapT :.: apply'
 compose :: (B b1, B b2, B b3) => (b1 :-* b2, b2 :-* b3) :<=> (b1 :-* b3)
 compose = AssocRT :.: (Id :*: apply')
 
-dneg :: B b => b :<=> Recip (Recip b) 
-dneg = -- b 
-  UnitI :.: -- ((),b)
+doubleDiv :: B b => b :<=> Recip (Recip b) 
+doubleDiv = -- b 
+  UnitI :.:         -- ((),b)
   (EtaT :*: Id) :.: -- ((1/(1/b),1/b),b)
-  AssocRT :.: -- (1/(1/b), ((1/b),b))
+  AssocRT :.:       -- (1/(1/b), ((1/b),b))
   (Id :*: EpsT) :.: SwapT :.: UnitE
 
-tens :: (B b1, B b2, B b3, B b4) => (b2 :-* b1, b4 :-* b3) :<=> ((b2,b4) :-* (b1,b3))
-tens = -- ((1/b2,b1),(1/b4,b3))
-  reorder :.: -- ((1/b2,1/b4),(b1,b3))
-  (Sym recipT :*: Id) 
-  where reorder :: (B a, B b, B c, B d) => ((a,b),(c,d)) :<=> ((a,c),(b,d))
-        reorder = AssocLT :.: -- (((a,b),c),d)
-                  (AssocRT :*: Id) :.: -- ((a,(b,c)),d)
-                  ((Id :*: SwapT) :*: Id) :.: -- ((a,(c,b)),d)
-                  (AssocLT :*: Id) :.: -- (((a,c),b),d)
-                  AssocRT 
-
 recipT :: (B b1, B b2) => Recip (b1,b2) :<=> (Recip b1, Recip b2)
-recipT = -- 1/(b1,b2)
-  UnitI :.: -- ((),1/(b1,b2))
-  UnitI :.: -- ((),((),1/(b1,b2)))
-  AssocLT :.: -- (((),()), 1/(b1,b2))
[…]

### 2013-02-25 code/frac/PiFracClean.hs
remove error and replace with []

@@ -114,7 +114,7 @@ eval UnfoldB True = return (Left ())
 eval UnfoldB False = return (Right ())
 eval EtaT () = [(Recip v, v) | v <- elems]
 eval EpsT (Recip v1, v2) | v1 == v2 = [()]
-                         | otherwise = error "Type error"
+                         | otherwise = []
 
 evalR :: (a :<=> b) -> b -> [a]
 evalR Id a = return a
@@ -150,7 +150,7 @@ evalR FoldB False = return (Right ())
 evalR UnfoldB (Left ()) = return True
 evalR UnfoldB (Right ()) = return False
 evalR EtaT (Recip v1, v2) | v1 == v2 = [()]
-                          | otherwise = error "Type error"
+                          | otherwise = []
 evalR EpsT () = [(Recip v, v) | v <- elems]
 
 -- using nub; could use exclusive union to get modal QC or nothing to get duplicates

### 2013-03-10 rationals/rational_ext.tex
reorg

--- 2012-06-29 rational_ext.tex
@@
 notion of multiplicity functions to express negative sized sets
-[CITE]. These notions have been extended to Euler characteristic of
-categories by Leinster, Joyal and Schanuel -- negative values are
-explained sometimes as equalities relating otherwise independent
-objects -- Baez tells a story about ``island'' and positive and
-``bridges connecting islands'' as negatives (since the effectively
-reduce the number of disconnected elements.)
+[CITE]. These notions have been extended to Euler characteristic and
+Homotopy cardinality of categories by Leinster, Joyal and Schanuel --
+negative values are explained sometimes as equalities relating
+otherwise independent objects -- Baez tells a story about ``island''
+and positive and ``bridges connecting islands'' as negatives (since
+the effectively reduce the number of disconnected elements.)
 

### 2013-03-10 rc2013/frac.tex
reorg

--- 2013-02-25 frac.tex
@@
 \newcommand{\amr}[1]{\textsc{Amr says:} #1}
-
-\newcommand{\roshan}[1]{} 
-%% \newcommand{\roshan}[1]{\textsc{Roshan says:} 
-%%   \textit{#1}
-%% }
 
@@
 
-\roshan{The above para is too technical in a weird way. Its unclear
-  what it contributes. If one wants to make a point of computational
-  witnesses of isomorphisms its easier to explain and cite in terms of
-  Andreas Blass and Fiore -- and more generally by Barry Mazur. I say
-  this also because it seems that HOTT is more concerned with
-  topological structure of equalities, even though they tend to start
-  their literature with some talk of intentional specifications.}
-
 %% AMR: I am not sure how to take this comment into account. I think we
@@
 
-\roshan{Again, HOTT seems jarring. Saying something about logic in
-  this way however seems useful. Maybe one can use one of the examples
-  of equivalence between proofs that Baez's Rosetta stone uses. }
-
 Technically speaking, obtaining a higher-order extension requires the
@@
 %% functions.  
-%% \roshan{The last sentence is terrible.}
 The remainder of the paper reviews {{Pi}} and then introduces the syntax and
@@
 %% AMR:
-%% \roshan{It is unclear what the Id types refer to. We should provide
-%%   some more context here}
 %% I think the previous paragraph in the intro that talks about Martin-Lof type 
@@
 
-\roshan{In the above the, when we say that the terms of {{Pi}} are
-  proofs that certain ``shapes of values'' are isomorphic, it is
-  misleading. In {{langRev}}, the value {{left ()}} may or may not be
-  isomorphic to the value {{right ()}} or {{left ()}} itself, or any
-  other value of our choice. In particular the actual syntactic
-  structure of the value has nothing to do with values being
-  isomorphic. Values are just place-holders for the situations that
-  their type represents. Their shapes have no intrinsic content. }
-
-\roshan{Further, this is not to say that we may not be able to use the
-  syntactic structure of a value for some purpose.  The prime use for
-  such a thing might be, for example, to define a notion of distance
-  between values -- such that values live in a metric space. Time and
-  time again, we have felt the need for developing such an idea. Such
-  a thing doesn't exist for the scope of this paper however. }
-
-\roshan{The use of the word ``orthogonal'' is inappropriate in the
-  footnote. Adding 0 truly affects how fractionals are developed,
-  whereas when we say ``orthogonal'' we give the impression that it does
-  not really matter. }
-
 \paragraph*{Data.}
@@
 %% the effect of an isomorphism.
-
-\roshan{It is unclear what the usage ``first-class'' and ``effect of
-  an isomorphism'' indicate. Maybe we should rephrase. }
 
@@
 language come when these are made to interact at the ``top level'' via
-\emph{application}: \roshan{I understand its intent, but this last sentence is unclear.}
+\emph{application}: 
 %subcode{bnf} include main
@@
 conjuction with our extension in the next section.)
-
-\roshan{Suggestion : This para is a good one to delete if we run out
-  of space. This is largely general knowledge about universal algebra
-  and maybe one useful remark about the categorical structure. However
-  we have a whole paper about the categorical structure of Pi at RC
-  itself. My point is this sort of this can be a remark inserted at
-  other places in passing and it doesn't deserve a para on its own. It
-  teaches the reader nothing new about Pi. }
 
@@
 
-% \roshan{It is unclear how fractions help differentiate between inputs
-%   and outputs. On first reading this seemed promising, but nothing
-%   in the rest of the paper substantiates this point of view.}
 % AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
@@
 %    output.
-%
-% Roshan: I completely understand this and this is not my objection. This
-% is not what the above sentence is saying however. Its talking about
-% Pi not having a notion of input and output which is simply
-% confusing.
-
-% \roshan{I think we should arrive at relational programming as a means
-%   of giving semantics to fractionals, rather than the other way
-%   round. To show that relational programming is reversible, one does
-%   not need to resort to fractionals. Everyone understand that
-%   relations can be reversed and none of this setup with {{Pi}} is
-%   required for that. Our key point should be to communicate that
-%   fractionals have a clean semantics in the setting of relational
-%   programming rather than the other way round.}
+
 % AMR: relations are indeed only presented as a way to explain the semantics
-% Roshan: it reads much better now.
 
@@
 
-% \roshan{ In the rules above it is unclear as to which v is chosen in
-%   the rule for eta. It is unclear that these rules are logically
-%   reversible.  }
-
-% \begin{proposition}[Logical Reversibility]
-% \label{chx:prop:logrev-tracep}
-% For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
-% {{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
-% \end{proposition}
-
-% Some thoughts on a relational semantics. We define {{ a <-> b }}
-% $\equiv$ {{a R b}}
-
-% \noindent
-% We now have primitives:
-% %subcode{opsem} include main
-% % swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
-% % swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
-% %                     && ~~~| v1 in b1, v2 in b2]}
-% % assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
-% %                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
-% etc..
-
-% \noindent
-% closure combinators:
-% %subcode{opsem} include main
-% % sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
-% % c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
-% % c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
-% etc..
-
-% \noindent
-% Now we have 
-% %subcode{opsem} include main
-% % eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
-% % eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}
-
-% Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
-% have {{c^{dagger}: b2 <-> b1}} such that 
-% %
-% {{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
-% straightforward from the definitions above.
-
-% I think defining eval is an open problem.  Consider a few relations:
-
-% \begin{enumerate}
-% \item 
-% {{empty = {[ ]} }}
-
-% \item 
-% {{full =  {[ R, G, B ]} * {[ R, G, B ]} }}
-
-% \item 
-% {{id  =  {[ (R, R), (G, G), (B, B) ]} }}
-
-% \item 
-% {{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
-% \end{enumerate}
-
-% Can we define an {{eval}} that meaningfully distinguishes these? What
-% can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
-% several possible {{eval}} definitions:
-
-% \begin{enumerate}
-% \item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
-%   eval maps {{b1 -> b2}}.
-
-% \item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.
-
-% \item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
-%   {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.
-
-% \end{enumerate}
-
-% For the relations above, none of the suggested eval definitions give a
-% logically reversible semantics.
-
-% \roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
-% {{b2}}? If yes, then we have a model for {{PiTF}} where the objects
-% are sets {{b}} and the morphisms are the relations between objects of
-% equal size.}
-
-% \roshan{ I suspect that all relations can be constructed, but we have
-%   to show this. This is not obvious.}
-
-% \roshan{In the judgments below it is misleading that only {{eta*}}
-%   appears indexed by a type. To construct each family of relations,
-%   the corresponding definition must be indexed by their types. In
-%   other words, there is nothing special about {{eta*}}. Further, I am
-%   unsure of how to format this so that the combinators are correctly
-%   indexed by their types for both primitives and for composition
-%   combinators. We could do the following, but it seems to be a
-%   overkill.}
-
-% %subcode{proof} include main
-% %@ v1 in [[b1]]
-% %@ v2 in [[b2]]
-% %@ v3 in [[b3]]
-% %@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % curry &:& b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)
-% curry &=& (id * recip) (;) (id * swap*) (;) assocl*
+% curry &=& (recip * id) (;) assocr*
 
@@
 
-\roshan{Either we should have explain the SAT solver well or we
-  shouldn't have at all. There are some original ideas in the
-  construction of that solver and the Oxford QF group for instance
-  were very interested in its details. This presentation neither
-  leaves the idea prime for future publication, and neither does it
-  explain it well.}
-
 % In the usual setting, an instance of SAT is a function~{{f}} which,
@@
 
-\roshan{Again, if we have lost zero, we cannot talk about vector
-  spaces with the zero vector. We should alternately included it in
-  our presentation to start with. }
-
 In such categories, the fractional type is given a non-trivial
@@
 
-\roshan{The above para should simply be deleted.}
-
 Fortunately, there is promising significant work on the groupoid
@@
 groupoids is cartesian closed, this appears eminently feasible.
-
-\roshan{Again, the above para is just lose speculation. Further, the
-  fact that we say that its cartesian closed instead of monoidal
-  closed should *immediately* tell us that its the wrong model.}
 

### 2013-03-10 rc2012/iso-int.tex
reorg

--- 2012-02-23 iso-int/iso-int.tex
@@
 
-aka. ultra-constructivism.
+(Maybe this section should be a seperate paper altogether)
+
+This section introduces another class of isomorphic interpreters
+{{langRevEE}} that arise from algebraic manipulations. Recursive type
+definitions are viewed as algebraic polynomials characterizing the
+shape of the type. Thus binary trees {{@@x.1+x*x}} are represented as
+the polynomial {{x=1+x^2}} and hence {{x^2-x+1=0}}.
+
+TODO: find a few more algebraic manipulations that give useful
+combinators. Depending on which language we are working in, there may
+be many. Suggestions:
+
+{{ a * b <-> (-a) * (-b)}}
+
+{{ -(-a) <-> a }}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Deriving the {{just : a <-> a + 1}} combinator}
+
+In the spirit of the above discussion, the type {{nat}} defined by
+{{@@x.1+x}} induces the polynomial {{nat = 1 + nat}}. Consider the
+following manipulation:
+
+{{a = a}}
+
+{{a + nat = a + nat}}
+
+{{a + nat = a + nat + 1}} because we know that {{nat = nat + 1}}.
+
+{{a = a + 1}} by cancelling nat.
+
+\noindent
+These algebraic manipulations can be treated as a computation in
+{{langRevT}} by witnessing each transformation with an appropriate
+combinator.
+
+\begin{itemize}
+\item Thus we start with the the line corresponding to {{a=a}}:
+
+DIAGRAM
+
+\item And then we construct {{a * nat = a * nat}} by parallel
+  composing with {{nat = nat}}.
+
+DIAGRAM
+
+\item And then we {{unfold}} on the RHS
+
+DIAGRAM
+
+\item And finally we use {{trace}} to do the appropriate cancellation,
+  thereby getting the final circcuit, which we call {{just}}:
+
+DIAGRAM
+
+\end{itemize}
+
+\noindent
+Computationally the {{langRevT}} circuit does the following:
+
+\begin{enumerate}
+\item {{forall (v:a).just v |--> right v}} 
+\item {{forall (v:a).just^{dagger} (right v) |--> v}} 
+\item {{just (left ())}} diverges. 
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Isomorphisms of Trees}
+
+Algebraic manipulations like the above have questionable mathematical
+validitiy induced by the fact that the polynomial {{nat = nat + 1}} is
+not well defined. Hence axiomatically adding such a polynomial may be
+suspected as the underlying reason for the divergence.
+
+In this context let us examine the isomorphism from one binary tree to
+seven binary trees that is examined in the Fiore's paper
+\cite{Fiore:2004}. Binary trees imply the polynomial {{t^2-t+1=0}}
+which does have a valid root, namely {{t = 1/2 + i*sqrt(3)/2}}.
+
+ALGEBRA
+
+etc.
+
+Which gives us the circuit:
+
+DIAGRAM
+
+Indeed this circuit is proof of the embedding of trees in a strong
+constructive sense i.e. we can supply a binary tree and examine the
+resulting seven binary trees. Further the adjoint of the circuit
+allows us to supply seven binary trees and get one resultant tree.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Non-constructive Isomorphism of Trees}
+However not all valid algebraic manipulations of a binary trees also
+have meanignful computational interpretations. Consider the same
+isomophism explained with the algebraic manipulation:
+
+{{t^3 = t*(t*t)}}
+
+{{t^3 = t*(t-1)}} (because {{t = t*t + 1}}  and hence {{t*t = t - 1}}).
+
+{{t^3 = t*t-t}}
+
+{{t^3 = -1}} (because {{t = t*t + 1}}  and hence {{t*t - t = -1}}).
+
+{{t^6 = 1}}
+
+{{t^7 = t}}
+
+The construction underlying this isomorphism is {{t^3 = -1}}, which
+can be expressed in {{langRevEE}} as the cicruit:
 

### 2013-03-25 rc13/frac2.tex
frac2

--- 2013-03-10 rc2013/frac.tex
@@
-\documentclass{llncs}
+\documentclass[preprint]{sigplanconf}
 \usepackage{url}
@@
 \usepackage{comment}
+\usepackage{multicol}
 
@@
 \newcommand{\amr}[1]{\textsc{Amr says:} #1}
+\newtheorem{definition}{Definition}
 
@@
 \begin{document}
+
+\conferenceinfo{ICFP '13}{date, City.} 
+\copyrightyear{2013}
+\copyrightdata{[to be supplied]}
+
+\titlebanner{banner above paper title}        % These are ignored unless
+\preprintfooter{short description of paper}   % 'preprint' option specified.
+
 \title{Fractional Types}
-\author{Roshan P. James$^{1}$ \and Zachary Sparks$^{1}$ \and Jacques Carette$^{2}$ \and Amr Sabry$^{1}$}
-\institute{$^{(1)}$~Indiana University \qquad $^{(2)}$~McMaster University}
+\authorinfo{Jacques Carette \and Amr Sabry}
+           {McMaster University \and Indiana University}
+           {carette@mcmaster.ca \and sabry@indiana.edu}
+
 \maketitle
@@
 
-{\small
-\bibliographystyle{splncs03} 
+\bibliographystyle{abbrvnat}
 \bibliography{cites}
-}
+
 \end{document}

### 2013-03-26 rc13/frac2.tex
abstract

@@
 
-\newcommand{\todo}[1]{\textbf{TODO:} #1}
-\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
-\newcommand{\amr}[1]{\textsc{Amr says:} #1}
 \newtheorem{definition}{Definition}
-
-\hyphenation{a-reas}
 
@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Fractional Types}
+\titlebanner{}
+\preprintfooter{}
+
+\title{Fractional Types: \\
+First-Class Pattern Matching Clauses and Higher-Order Reversible Functions}
 \authorinfo{Jacques Carette \and Amr Sabry}
@@
   information-preserving, and reversible programming language {{Pi}} founded
-  on type isomorphisms. Being restricted to first-order types limits the
-  expressiveness of the language: it is not possible, for example, to
-  abstract common program fragments into a higher-level combinator. In this
-  paper, we introduce a higher-order extension of {{Pi}} based on the novel
-  concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
-  fractional type {{1/v}} represents \emph{negative} information. A
-  function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
-  representing the needed argument and {{v2}} representing the
-  result. Fractional values are first-class: they can be freely propagated
-  and transformed but must ultimately --- in a complete program --- be offset
-  by the corresponding amount of positive information.
+  on type isomorphisms. In subsequent work, we extended this language with
+  \emph{fractional types}, the dual of conventional product types which, from
+  an information-theoretic point of view, corresponds to the type of values
+  with \emph{negative information}. The extended language was given an
+  operational interpretation that hinted at two facts: fractional types can
+  be used to define first-class pattern-matching clauses as well as
+  higher-order reversible functions. Neither of these observations was backed
+  by an appropriate denotational model, however. In this paper, we develop
+  such a model by interpreting fractional types using functions mapping
+  singleton types to the unit type. This interpretation requires that nearly
+  all of the type constructs in {{Pi}} be re-interpreted as \emph{dependent
+    types}. The framework, formalized in Agda, takes steps towards developing
+  a computational interpretation of homotopy type theory using higher-order
+  reversible languages.
 \end{abstract}

### 2013-03-27 rc13/frac2.tex

@@
 \newtheorem{definition}{Definition}
+\newcommand{\bra}[1]{\langle #1 |}
+\newcommand{\singleton}[1]{\textsc{Sing}~#1}
 
@@
 %! |- = \vdash
+%! bra = \bra
+%! singleton = \singleton
 %! <=> = \Longleftrightarrow
@@
 
-\title{Fractional Types: \\
-First-Class Pattern Matching Clauses and Higher-Order Reversible Functions}
+\title{Reversible Higher-Order Programs \\ (Functional Pearl)}
 \authorinfo{Jacques Carette \and Amr Sabry}
@@
 
-We are witnessing a convergence of ideas from several distinct research
-communities (physics, mathematics, and computer science) towards replacing
-\emph{equalities} by \emph{isomorphisms}. The combined programme has sparked
-a significant amount of research that unveiled new and surprising connections
-between geometry, algebra, logic, and computation (see~\cite{baez2011physics}
-for an overview of some of the connections).
-
-In the physics community, Landauer~\cite{Landauer:1961,Landauer},
-Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
-the laws of physics as fundamentally related to computation. The great
-majority of these laws are formulated as equalities between different
-physical observables which is unsatisfying: \emph{different} physical
-observables should not be related by an \emph{equality}. It is more
-appropriate to relate them by an \emph{isomorphism} that witnesses, explains,
-and models the process of transforming one observable to the other.
-
-In the mathematics and logic community, Martin-L\"of developed an extension
-of the simply typed $\lambda$-calculus originally intended to provide a
-rigorous framework for constructive
-mathematics~\cite{citeulike:7374951}. This theory has been further extended
-with \emph{identity types} representing the proposition that two terms are
-``equal.'' (See~\cite{streicher,warren} for a survey.) Briefly speaking,
-given two terms $a$ and $b$ of the same type $A$, one forms the type
-$\texttt{Id}_A(a,b)$ representing the proposition that~$a$ and~$b$ are equal:
-in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
-models the process of transforming $a$ to $b$ and vice-versa.
-
-%% AMR: I am not sure how to take this comment into account. I think we
-%% should make a specific connection to Martin-Lof type theory to at least
-%% justify the mention of groupoids in the conclusion. I agree though that
-%% this is a very technical point and I do find it hard to express the point
-%% in a couple of sentences. Suggestions welcome.
-
-In the computer science community, the theory and practice of type
-isomorphisms is well-established. Originally, such type isomorphisms were
-motivated by the pragmatic concern of searching large libraries of functions
-by providing one of the many possible isomorphic types for the desired
-function~\cite{Rittri:1989:UTS:99370.99384}. More recently, type isomorphisms
-have taken a more central role as \emph{the} fundamental computational
-mechanism from which more conventional, i.e., irreversible computation, is
-derived. In our own previous
-work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
-notion of type isomorphism and developed from it a family of programming
-languages, {{Pi}} with various superscripts, in which computation is an
-isomorphism preserving the information-theoretic entropy.
-%% Moreover these languages have models in various flavors of symmetric
-%% monoidal categories (which are the backbone of models for linear logic and
-%% quantum computation).
-%% We have also invested a significant amount of time in using {{Pi}} as
-%% an actual programming language and established several idioms for
-%% programming large classes of interesting (recursive) programs in
-%% {{Pi}}.
-
-A major open problem remains, however: a higher-order extension of
-{{Pi}}. This extension is of fundamental importance in all the originating
-research areas.
-%% In physics, it allows a process or observer to be treated as
-%% ``data'' that can be transformed by higher-level processes or observed by
-%% meta-level observers. 
-In physics, it allows for quantum states to be viewed as processes and
-processes to be viewed as states, such as with the Choi-Jamiolkowski
-isomorphism~\cite{choi1975completely,jamiolkowski1972linear}.  In mathematics
-and logic, it allows the equivalence between different proofs of type
-$\texttt{Id}_A(a,b)$ to itself be expressed as an isomorphism (of a higher
-type) $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
-higher-order types allow code to abstract over other code fragments as well
-as the manipulation of code as data and data as code.
-
-Technically speaking, obtaining a higher-order extension requires the
-construction of a \emph{closed category} from the underlying monoidal
-category for~{{Pi}}. Although the general idea of such a construction is
-well-understood, the details of adapting it to an actual programming language
-are subtle.  Our main novel technical device to achieving the higher-order
-extension is a \emph{fractional type} which represents \emph{negative
-  information} and which is so named because of its duality with conventional
-product types.  
-%% By adding a formal (linear) dual to (type theoretic) products, we can
-%% internalize the notion that ``taking an input'' corresponds
-%% to negative information, which turns out to be the basis of first-class
-%% functions.  
-The remainder of the paper reviews {{Pi}} and then introduces the syntax and
-semantics of the extension with fractional types. We then study the
-properties of the extended language and establish its expressiveness via
-several constructions and examples that exploit its ability to model
-higher-order computations.
-
-%% AMR:
-%% I think the previous paragraph in the intro that talks about Martin-Lof type 
-%% theory provides the context. Anything else??
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background: {{Pi}} }
+\section{Background: {{Pi}} and {{PiTF}} }
 \label{sec:pi}
@@
 \noindent 
-Types include the unit type {{1}}, sum types {{b1+b2}}, and
-product types {{b1*b2}}.  Values include {{()}} which is the only value of
-type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,
-and {{(v1,v2)}} which builds a value of product type. 
-%% But these values should be regarded as largely ancillary: we do not
-%% treat them as first-class citizens, and they only occur when observing
-%% the effect of an isomorphism.
+Types include the unit type {{1}}, sum types {{b1+b2}}, and product types
+{{b1*b2}}.  Values include {{()}} which is the only value of type {{1}},
+{{left v}} and {{right v}} which inject~{{v}} into a sum type, and
+{{(v1,v2)}} which builds a value of product type.  But these values should be
+regarded as largely ancillary: we do not treat them as first-class citizens,
+and they only occur when observing the effect of an isomorphism.
 
-%% \paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
-%% (value) type isomorphisms.  They are also typed, by the shape of the (value)
-%% type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
-%% the following isomorphisms:
 \paragraph*{Isomorphisms.} The terms of {{Pi}} witness
@@
 \emph{application}: 
+subcode{bnf} include main
+top level term, l ::= c v
+
+\noindent
+To summarize, the syntax of {{Pi}} is given as follows.
+
+\begin{definition}{(Syntax of {{Pi}})}
+\label{def:Pi}
 %subcode{bnf} include main
+% value types, b ::= 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% iso.~types, t ::= b <-> b
+% base iso ::= swap+ | assocl+ | assocr+ 
+%     &|& unite | uniti | swap* | assocl* | assocr* 
+%     &|& dist | factor 
+% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
 % top level term, l ::= c v
-
-% \noindent
-% To summarize, the syntax of {{Pi}} is given as follows.
-
-% \begin{definition}{(Syntax of {{Pi}})}
-% \label{def:Pi}
-% %subcode{bnf} include main
-% % value types, b ::= 1 | b+b | b*b 
-% % values, v ::= () | left v | right v | (v,v) 
-% %
-% % iso.~types, t ::= b <-> b
-% % base iso ::= swap+ | assocl+ | assocr+ 
-% %     &|& unite | uniti | swap* | assocl* | assocr* 
-% %     &|& dist | factor 
-% % iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-% % top level term, l ::= c v
-% \end{definition}
+\end{definition}
 
@@
 
-% Operationally, the
-% semantics of {{Pi}} is given using two mutually recursive interpreters: one
-% going forward and one going backwards. The use of {{sym}} switches control
-% from one evaluator to the other. We will present the operational semantics in
-% the next section along with the extension with fractional types and
-% values. For now, we state without proof that the evaluation of well-typed
-% combinators always terminates and that {{Pi}} is logically reversible, i.e.,
-% that for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 :
-%     b2}} we have the forward evaluation of {{c v1}} produces {{v2}} iff the
-% backwards evaluation of {{c v2}} produces {{v1}}.
+Operationally, the semantics of {{Pi}} is given using two mutually recursive
+interpreters: one going forward and one going backwards. The use of {{sym}}
+switches control from one evaluator to the other.
 
-%% AMR:
-%% \jacques{Since we have Agda code for the above, should we refer to it
-%% here, and make it available?}
-%% That is a great idea but it would take a significant amount of time to
-%% clean up the code and make it presentable. If someone is willing to do
-%% that, then we should definitely add a reference to the code and make it
-%% available.
-%% Jacques:  Right.  I'll definitely work on cleaning up the paper first,
-%% then see if there is time left for the code.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Fractional Types}
 
-%% As is usual with type theories,
-%% the coherence conditions (pentagonal and triangle identity) are not explicit,
-%% but would correspond to certain identity types being trivial.
+For every value {{v : b}}, we introduce the \emph{singleton} type 
+{{ singleton{v} }} whose only element is the value {{v}}.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Language: {{PiTF}} }
-
-The language {{Pi}} models isomorphisms of values rather well, and as we
-established before~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012}, it is
-logically reversible and each computation it expresses preserves the
-information-theoretic entropy. However, purposefully, it has no distinguished
-notion of \emph{input} or \emph{output}; more precisely, because of strict
-preservation of information, these two concepts coincide.  This is the root
-of reversibility. But if we want to model functions of any flavor as values,
-we need to differentiate between these notions. Our idea, inspired by
-computational dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
-polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
-of \emph{compact
-  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
-is to introduce a \emph{formal dual} for our types.  We thus consider a type
-in a negative position to possess negative information, as it is really a
-request for information.  Since information is logarithmic, this means that a
-request for information should behave like a \emph{fractional type}.
-
-% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
-% Jacques: Exactly!  1/a is negative information is input, b is positive is
-%    output.
-
-% AMR: relations are indeed only presented as a way to explain the semantics
-
-The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
-simple:
-%subcode{bnf} include main
-% value types, b ::= ... | 1/b
-% values, v ::= ... | 1/v
-
-\noindent For a given type {{b}}, the values of type {{1/b}} are of the form
-{{1/v}} where {{v : b}}. Note that {{1/v}} is purely formal, as is
-{{1/b}}.
-
-Semantically, we are extending the symmetric monoidal category modeling
-{{Pi}} to a \emph{compact closed} one, i.e., a category in which morphisms
-are representable as objects. In such a setting, the new dual objects (i.e.,
-the fractionals) must satisfy the following isomorphism witnessed by two new
-type indexed combinators {{eta*_b}} and {{eps*_b}}:
-\begin{multicols}{3}
-  \dgm{eta_times2}
-
-\columnbreak
-
-%subcode{bnf} include main
-%! columnStyle = r r c l l
-%eta*_b :&  1 & <-> & 1/b * b &: eps*_b
-%% XXX: diagrams need unit wires(?)
-%% Roshan: We should just be clear that unit wires are dropped. Dgms
-%% including them dont communicate any additional information and
-%% appear crowded.
-
-\columnbreak
-
-  \dgm{eps_times2}  
-\end{multicols}
-
-\vspace{-10pt}
-\noindent From a programming perspective, we think of a value {{1/v}} as a
-\emph{first-class constraint} that can only be satisfied if it is matched
-with an actual value {{v}}. In other words, {{1/v}} is a \emph{pattern} representing the absence of
-some information of a specific shape (i.e., \emph{negative information}),
-that can only be reconciliated by an actual value {{v}}. The combinator
-{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
-out of no information --- an equal amount of negative and positive
-information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
-up an equal amount of negative and positive information producing no residual
-information.
-
-Historically, versions of compact closed categories were introduced by
-Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
-Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the generalization of
-monoidal categories to model quantum computation.  The first and most
-intuitive example of such categories is the category of finite sets and
-relations. In that case, the dual operation (i.e., the fractional type)
-collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
-interpretation of fractionals, this is still an interesting category as it
-provides a model for a reversible higher-order language. Indeed, any relation
-is reversible via the standard \emph{converse} operation on a relation.
-Furthermore, any relation between {{b1}} and {{b2}} can be represented as 
-a subset of {{b1 * b2}}, i.e., as a value.
-
-In this section, we provide a simple semantics of {{PiTF}} in the
-category of finite sets and relations, in the following way: We
-interpret each combinator {{c : b1 <-> b2}} as a \emph{relation}
-between the sets denoted by {{b1}} and {{b2}} respectively. Sets
-{{b1}} and {{b2}} can be related only if the same information measure.
-As we illustrate, this semantics allows \emph{arbitrary} relations
-(including the empty relation and relations that are not isomorphisms)
-to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat} we
-discuss more refined semantics that are more suitable for richer
-categories.
-
-\begin{definition}[Denotation of Value Types]
-\label{chx:def:denot}
-Each type denotes a finite set of values as follows:
-%subcode{opsem} include main
-%! <- = \leftarrow
-%! union = \cup
-% [[1]] '= {[ () ]}
-% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
-% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
-% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
-\end{definition}
-
-We specify
-relations using a deductive system whose judgments are of the form 
-{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
-denoted by {{c}}.
-
-\begin{definition}[Relational semantics]
-\label{def:relational-PiTF}
-Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
-as specified below.~\footnote{
-In the interest of brevity, we have treated the type annotations in the base
-isomorphisms as implicit; for example, the rule for {{assocr*}} should really
-read:
-
-%subcode{proof} include main
-%@ v1 in b1
-%@ v2 in b2
-%@ v3 in b3
-%@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))
-}
-%subcode{proof} include main
-%@ ~
-%@@ (left v) ~~swap+~~ (right v)
-%
-%@ ~
-%@@ (right v) ~~swap+~~ (left v)
-%----
-%@ ~
-%@@ (left v) ~~assocl+~~ (left (left v))
-%
-%@ ~
-%@@ (right (left v)) ~~assocl+~~ (left (right v))
-%----
-%@ ~
-%@@ (right (right v)) ~~assocl+~~ (right v)
-%
-%@ ~
-%@@ (left (left v)) ~~assocr+~~ (left v)
-%----
-%@ ~
-%@@ (left (right v)) ~~assocr+~~ (right (left v))
-%
-%@ ~
-%@@ (right v) ~~assocr+~~ (right (right v))
-%----
-%@ ~
-%@@ ((), v) ~~unite~~ v
-%
-%@ ~
-%@@ v ~~uniti~~ ((), v)
-%
-%@ ~
-%@@ (v1, v2) ~~swap*~~ (v2, v1)
-%----
-%@ ~
-%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
-%
-%@ ~
-%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
-%----
-%@ ~
-%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
-%
-%@ ~
-%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
-%----
-%@ ~
-%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
-%
-%@ ~
-%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
-%----
-%@ ~
-%@@ v ~~id~~ v
-%
-%@ v' ~~c~~ v
-%@@ v ~~(sym c)~~ v'
-%
-%@ v1 ~~c1~~ v2
-%@ v2 ~~c2~~ v3
-%@@ v1 ~~(c1(;)c2)~~ v3
-%----
-%@ v ~~c1~~ v'
-%@@ (left v) ~~(c1 (+) c2)~~ (left v')
-%
-%@ v ~~c2~~ v'
-%@@ (right v) ~~(c1 (+) c2)~~ (right v')
-%----
-%@ v1 ~~c1~~ v1'
-%@ v2 ~~c2~~ v2'
-%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
-%
-%@ ~
-%@@ () ~~eta*_b~~ (1/v,v)
-%
-%@ ~
-%@@ (1/v,v) ~~eps*_b~~ ()
-\end{definition}
-
-The semantics above defines a relation rather than a function. It is still
-reversible in the sense that it ``mode-checks''. In other words, for any
-combinator~{{c}} we can treat either its left or its right side as an input;
-this allows us to define {{sym}} as just relational converse, which amounts
-to an argument swap. We can then give an operational interpretation to the
-semantic relation by defining two interpreters: a forward evaluator and a
-backwards evaluator that exercise the relation in opposite directions, but
-which now can return a \emph{set} of results. This operational view exploits
-the usual conversion of a relation from a subset of {{A * B}} to a function
-from {{A}} to the powerset of {{B}} whose composition is given by the Kleisli
-composition in the powerset monad. It is straightforward to check that if the
-forward evaluation of {{c v1}} produces {{v2}} as a possible answer then the
-backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.  One
-can check that all relations for {{Pi}} are total functional relations whose
-converse are also total functional relations -- aka isomorphisms.  However
-{{eta*_b}} relates {{()}} to multiple values (one for each value of type {{b}}),
-and is thus not functional; {{eps*_b}} is exactly its relational converse, as
-expected.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Expressiveness and Examples}
-
-Having introduced the syntax of our extended language, its semantics in the
-category of sets and relations, and its operational semantics using forward
-and backwards interpreters, we now illustrate its expressiveness as a
-higher-order programming language. In the following presentation consisting
-of numerous programming examples, we use {{bool}} as an abbreviation of
-{{1+1}} with {{true}} as an abbreviation for {{left ()}} and {{false}} as an
-abbreviation for {{right ()}}. In addition, instead of presenting the code
-using the syntax of {{Pi}}, we use circuit diagrams that are hopefully more
-intuitive. Each diagrams represents a combinator whose evaluation consists of
-propagating values along the wires.  For the sake of readability, we omit
-obvious re-shuffling circuitry and elide trivial unit wires which carry no
-information.\footnote{The full code is available at
-\url{www.cas.mcmaster.ca/~carette/PiFractional}.}
-
-%%%%%%%%%%%%
-\subsection{First-Class Relations}
-
-The most basic additional expressiveness of {{PiTF}} over {{Pi}} is the
-ability to express relations as values. Indeed a value of type {{1/b1 * b2}}
-is a pair of a \emph{constraint} that can only be satisfied by some 
-{{v1 : b1}} and a value {{v2 : b2}}. In other words, it corresponds to a
-function or relation which when ``given'' a value {{v1 : b1}} ``releases''
-the value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:
-
-%subcode{bnf} include main
-%  b1 :-* b2 &::=& 1/b1 * b2
-
-\noindent which suggests a function-like behavior for the pair of a fractional value
-and a regular value. What is remarkable is that we can almost trivially turn
-any combinator {{c : b1 <-> b2}} into a (constant) value of 
-type {{b1 :-* b2}} as shown below on the left:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
-% name c &=& eta*_{b1} (;) (id (*) c)
-
-\begin{multicols}{2}
-  \dgm{function-times}
-
-  \dgm{delimc-times}
-\end{multicols}
-\noindent Dually, as illustrated above on the right, we also have a
-{{coname}} combinator which when given an answer {{v2 : b2}}, and a request
-for the corresponding input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this
-as a un-needed computation (see also Sec. 3, \cite{abramsky-2008}).
-
-More generally any combinator manipulating values of type {{b}} can be turned
-into a combinator manipulating values of type {{1/b}} and vice-versa. This is
-due to the fact that fractionals types satisfy a self-dual involution
-relating {{b}} and {{1/(1/b)}}:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% doubleDiv &:& b <-> 1/(1/b)
-% doubleDiv &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite
-
-% jacques: elide for now, to save space, as this diagram is backwards.
-% \dgm{involution-times}
-
-% As a simple example, consider {{swap+}} denoting boolean
-% negation.  It is straightforward to check that the combinator 
-% {{name~swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:
-
-% %subcode{proof} include main
-% %@ ~
-% %@@ () ~~(name~swap+)~~ (1/false,true)
-% %
-% %@ ~
-% %@@ () ~~(name~swap+)~~ (1/true,false)
-
-% \noindent The relation {{name~swap+}} maps {{()}} to a value which can be used
-% to convert any {{false}} to {{true}} and vice-versa. We formalize this
-% ``apply'' combinator below.
-
-%%%%%%%%%%%%
-\subsection{Higher-Order Relations}
-
-We are now a small step from implementing various higher-order combinators
-that manipulate functions or relations. In particular, we can \emph{apply},
-\emph{compose}, \emph{curry}, and \emph{uncurry} values representing
-relations.  We first show the realization of \emph{apply}:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% apply &:& (b1 :-* b2) * b1 <-> b2
-% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
-
-\begin{multicols}{2}
-  \dgm[1.2]{apply-times1}
-
-  \dgm{apply-times2}
-\end{multicols}
-
-\noindent Intuitively, we simply match the incoming argument of type {{b1}} with
-the constraint encoded by the function. If they match, they cancel
-each other and the value of type {{b2}} is exposed with no
-constraints. Otherwise, the result is undefined.
-
-\noindent A flipped variant is also useful:
-%subcode{opsem} include main
-%! columnStyle = rcl
-% apply' &:& b1 * (b1 :-* b2) <-> b2
-% apply' &=& assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite
-
-\noindent Function or relation composition is now straightforward:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
-% compose &=& assocr* (;) (id (*) apply')
-
-\begin{multicols}{2}
-  \dgm[1.0]{compose-times1}
-
-  \dgm{compose-times2}
-\end{multicols}
-
-% \dgm{compose-times3}
-
-\noindent
-We can also derive currying (and dually, uncurrying) combinators. Observe that
-the type of {{curry}} needs to be {{(b1 * b2 :-* b3) <-> (b1 :-* (b2 :-* b3))}},
-which can be written in mathematical notation as 
-{{1/(b1 * b2) * b3 = (1/b1) * ((1/b1) * b3)}}. This means {{curry}} can be
-written using {{recip}}, the implementation of the mathematical identity
-{{1/(b1 * b2) = 1/b1 * 1/b2}}:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% recip &:& 1 / (b1 * b2) <-> 1 / b1 * 1 / b2
-% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta*_{b1} (*) eta*_{b2}) (*) id) (;)
-%     && (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*) (;) swap* (;) (unite)
-
-\noindent
-where {{reorder}} is the obvious combinator of type {{b1 * (b2 * b3) * b4 <-> b1 * (b3 * b2) * b4}}.
-With {{recip}} out of the way, we can easily write {{curry}}:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% curry &:& b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)
-% curry &=& (recip * id) (;) assocr*
-
-\noindent That {{recip}} is the heart of currying seems quite remarkable.
-
-%%%%%%%%%%%%
-\subsection{Feedback, Iteration, and Trace Operators}
-
-Mathematically speaking, recursion and iteration can be expressed using
-categorical trace
-operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
-language like {{Pi}} there are two natural families of trace operators that
-can be defined, an additive family (explored in detail in our previous
-work~\cite{rc2011}) and a multiplicative family, which is expressible using
-fractionals.
-
-The idea of the multiplicative trace operator is as follows. We are given a
-computation {{c : b1 * b2 <-> b1 * b3}}, and we build a ``looping'' version
-which feeds the output value of type {{b1}} back as an input. Effectively,
-this construction cancels the common type {{b1}}, to produce a new combinator
-{{trace*~c:b2<->b3}}.
-
-With fractionals, {{trace*}} becomes directly expressible:
-
-%subcode{opsem} include main
-%! columnStyle = rcl
-% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
-% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
-%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite
-
-As an example, we can use the operational semantics (outlined in the previous
-section) to calculate the result of applying {{trace*_{bool} (swap+ (*) id)}}
-to {{false}}.
-%subcode{opsem} include main
-%! columnStyle = ll
-% {[ ((),false) ]}  & (uniti)
-% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
-% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
-% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
-% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
-% emptyset & (eps*_{bool} * id)
-
-\noindent The computation with {{true}} gives the same result. This confirms
-that although our semantics is formally reversible, it does not result in
-isomorphisms.
-
-More abstractly, the evaluation of {{trace*~c : b2 <-> b3}} must ``guess'' a
-value of type {{b1}} to be provided to to the inner combinator 
-{{c : b1 * b2 <-> b1 * b3}}. 
-This value {{v1 : b1}} cannot be arbitrary: it must be such that it is the value
-produced as the first component of the result.  In general, there may be
-several such fixed-point values, or none. Indeed with a little ingenuity 
-(see the next section), one can 
-express any desired relation by devising a circuit which keeps the desired pairs as 
-valid fixed-points.
-
-% \begin{example}
-% Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.
-
-% \noindent Using {{trace*}}, we construct the
-% combinator {{trace* c : 1 <-> 1}}. Applying this to 
-% {{() : 1}} requires us to find a value {{b : bool}} such that 
-% {{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
-% has two values, there are two values {{false}} and {{true}} that satisfy the 
-% constraint. In other words, {{() ~~(trace* c)~~() }}.
-% \end{example}
-
-% \begin{example}
-% \label{ch3:ex:annihilate}
-% Consider a small variation on the combinator {{c}} above:
-
-% {{c : bool * 1 <-> bool * 1}}
-
-% {{c = swap+ (*) id }}
-
-% \noindent which negates the boolean component of its incoming
-% pair. Using the multiplicative trace operator, we can construct the
-% combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
-% combinator to {{() : 1}} requires us to find a value {{b : bool}} such
-% that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
-% has no fixed points.  Operationally, the evaluation of
-% such a combinator should produce no value (were it to even terminate).
-% \end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{(Finite) Relational Programming}
-\label{ch3:sec:lp}
-
-Relational programming leverages set-theoretic relations and their
-composition to express computational tasks in a declarative way. 
-% For instance, consider this example:
-% {{parent = {[ (A,B), (B,C), (A,D) ]} }}
-% {{grandparent = parent (o) parent}}
-% \noindent The example defines a relation {{parent}} specified using a
-% set of tuples and another relation {{grandparent}} specified using the
-% composition of two parent relations. If we wrote this example in a
-% relational language (e.g., Prolog) and we executed the query
-% {{grandparent(A)}}, we would get the answer {{ {[C]} }}.
-It turns out that with the addition of the multiplicative trace, and
-the move to relations motivated in the previous section, we can
-express relational programming.
-
-Consider the relation {{R}} on booleans given by:
-
-{{ {[(false,false), (false,true), (true,false) ]}. }}
-
-\noindent We can define a combinator {{c_R}} whose denotation is {{R}}, by
-defining a combinator {{cInner : (a * bool) <-> (a * bool)}} for some type
-{{a}} such that {{c_R = trace* cInner}}. The basic requirement of {{cInner}}
-is that for each desired pair {{(v1,v2)}} in {{R}}, it maps {{(a0,v1)}} to
-{{(a0,v2)}} for some value {{a0}} and that for each pair {{(v1,v2)}} that is
-\emph{not} in {{R}}, it maps {{(a1,v1)}} to {{(a2,v2)}} for \emph{different}
-values {{a1}} and {{a2}}.
-
-% %subcode{proof} include main
-% %@ ~
-% %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
-% %----
-% %@ ~
-% %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
-% %----
-% %@ ~
-% %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
-% %----
-% %@ ~
-% %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
-% %----
-% %@ ~
-% %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
-% %----
-% %@ ~
-% %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
-% %----
-% %@ ~
-% %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
-% %----
-% %@ ~
-% %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
-% In the first three lines, the first argument is a fixed point and hence the
-% multiplicative trace would map the second input to the second output
-% producing the desired relation. In the remaining five cases, the first
-% argument is not a fixed point and hence all these cases would be rejected as
-% solutions to the constraint imposed by the multiplicative trace. It simply
-% remains to find the actual realization of {{cInner}} that would produce the
-% behavior above.
-
-% %subcode{opsem} include main
-% %! columnStyle = rcl
-% % cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
-% % cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
-% %        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
-% %        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))
-
-% \noindent where {{toffoli}} and {{cnot}} are the conventional reversible gates (their definition in {{Pi}} is in~\cite{rc2011}).
-% %     
-% % c_R &:& bool <-> bool
-% % c_R &=& trace* cInner
-
-% The above example should convince the reader that the language {{Pi}}
-% with multiplicative trace is expressive enough to model finite relational
-% programming. 
-% %%%%%%%%%%%%%%%
-% \subsection{Solving Constraints}
-% \label{ch3:sec:constraints}
-
-% A large class of constraint satisfaction problems can be expressed
-% using multiplicative traces. We illustrate the main ideas with the
-% implementation of a SAT solver.
-
-For small examples, finding such combinators by trial and error is a
-relatively straightforward but tedious task. It is, however, possible to
-automate this task by expressing what is essentially a reversible SAT solver
-as follows. In the usual setting, an instance of SAT is a function~{{f}}
-which, when given some boolean inputs, returns {{true}} or {{false}}. The
-function returns {{true}} when the inputs satisfy the constraints imposed by
-the structure of {{f}} and a solution to the SAT problem is the set of all
-inputs on which {{f}} produces {{true}}. The basic idea of our construction
-is to use {{trace*}} to annihilate values that fail to satisfy the
-constraints represented by the SAT instance {{f}}. (The accompanying code
-details the construction of such a solver.)
-
-% In the usual setting, an instance of SAT is a function~{{f}} which,
-% when given some boolean inputs, returns {{true}} or {{false}}. The
-% function returns {{true}} when the inputs satisfy the constraints
-% imposed by the structure of {{f}} and a solution to the SAT problem is
-% the set of all inputs on which {{f}} produces {{true}}. The basic idea
-% of our construction is to generalize the annihilation circuit from
-% Ex.~\ref{ch3:ex;annihilation} to only annihilate values that fail to
-% satisfy the constraints represented by the SAT instance {{f}}. To
-% achieve this goal, we must however deal with several important
-% details.
-
-% First, because we are in a reversible world, our instance of SAT must be
-% expressed as an isomorphism: this is easily achieved by the construction
-% which embeds any boolean function {{f}} into a reversible one {{iso_f}} with
-% a larger domain and range.  Given such a reversible function {{iso_f}} which
-% represents a SAT instance, we first construct the circuit below:
-
-% \begin{center}
-% \scalebox{1.2}{
-%   \includegraphics{diagrams/sat2.pdf}
-% }
-% \end{center}
-
-% As shown in the circuit, the reversible SAT instance {{iso_f}} takes
-% two sets of values and produces two outputs. The incoming values
-% labeled \textsf{inputs} are the inputs we need to test for
-% satisfiability. The other incoming values labeled \textsf{heap} are
-% the additional inputs needed to embed the original SAT instance {{f}}
-% into a reversible function. If these \textsf{heap} values are all
-% initialized to {{false}}, the output wire \textsf{satisfied?}
-% corresponds to the output that {{f}} would have produced on
-% \textsf{inputs}. The other outputs labeled \textsf{garbage} are not
-% needed for their own sake but they are important because they are used
-% as inputs to the adjoint of {{iso_f}} to reproduce the inputs exactly,
-% in anticipation of closing the loop with {{trace*}}.
-
-% To summarize, the top half of the circuit is the identity function
-% except that we have also managed to produce a boolean wire labeled
-% \textsf{satisfied?} that tells us if the inputs satisfy the desired
-% constraints. We can take this boolean value and use it to decide
-% whether to negate the bottom wire (labeled \textsf{control
-%   wire}). Specifically, if the inputs do \emph{not} satisfy {{f}}, the
-% control wire is negated. The last wire labeled \textsf{heap control
-%   wire} is negated if the heap values do not have the right initial
-% values, i.e., are not all {{false}}.
-
-% Let us call the above construction {{sat_f}}. If we now close the loop
-% using {{trace*}}, two things should happen:
-% \begin{itemize}
-% \item configurations in which the \textsf{heap} values are not all
-%   {{false}} will be annihilated;
-% \item configurations in which the \textsf{inputs} do not satisfy {{f}}
-%   will cause the \textsf{satisfied?} wire to be negated and hence will
-%   also be annihilated.
-% \end{itemize}
-% In other words, the only configurations that will survive are the ones in
-% which the \textsf{inputs} satisfy {{f}}. We simply need to arrange to
-% \emph{clone} these values and produce them as the output of the whole
-% circuit. The final construction is therefore:
-
-% \begin{center}
-% \scalebox{1.5}{
-%   \includegraphics{diagrams/sat3.pdf}
-% }
-% \end{center}
-
-% To make the previous discussion concrete, we present a small, but
-% complete, example. In our example, the SAT instance {{f}} is tiny: it
-% takes two inputs. This function is embedded into a reversible function
-% {{iso_f}} of type 
-% {{((bool * bool) * bool) <-> ((bool * bool) * bool)}} where the last
-% input represents the heap and the first two outputs represent the garbage. 
-% The realization of {{sat_f}} given below is parametrized by such 
-% a function {{iso_f}}. The inputs to {{sat_f}} are 
-% \textsf{heap control}, \textsf{control}, \textsf{heap}, \textsf{input-1}, and 
-% \textsf{input-2}. Its operation is simple: if the \textsf{heap} is {{true}}, 
-% \textsf{heap control} is negated, and if the last output of {{iso_f}} 
-% is {{false}}, \textsf{control} is negated:
-
-% %subcode{opsem} include main
-% %! columnStyle = rcl
-% % sat_f &:& ((((bool * bool) * bool) * bool) * bool) <-> ((((bool * bool) * bool) * bool) * bool)
-% % sat_f &=& ((swap* (*) id) (*) id) (;) 
-% % && ((assocl* (*) id) (*) id) (;) 
-% % &&   (((cnot (*) id) (*) id) (*) id) (;) 
-% % &&   assocr* (;) 
-% % &&   (assocr* (*) id) (;) 
-% % &&   (swap* (*) id) (;) 
-% % &&   assocr* (;) 
-% % &&   (id (*) assocl*) (;) 
-% % &&   (id (*) isof) (;) 
-% % &&   swap* (;) 
-% % &&   assocr* (;) 
-% % &&   (id (*) (id (*) swap*)) (;) 
-% % &&   (id (*) assocl*) (;) 
-% % &&   (id (*) ((inot (*) id) (*) id)) (;)
-% % &&   (id (*) (cnot (*) id)) (;) 
-% % &&   (id (*) ((inot (*) id) (*) id)) (;)
-% % &&   (id (*) assocr*) (;) 
-% % &&   (id (*) (id (*) swap*)) (;) 
-% % &&   assocl* (;)
-% % &&   swap* (;)
-% % &&   (id (*) Sym isof) (;) 
-% % &&   (id (*) assocr*) (;) 
-% % &&   assocl* (;) 
-% % &&   assocl*
-
-% Given the construction of {{sat_f}} we can build the full solver as
-% follows. The overall input is the cloning heap. The combinator given
-% to {{trace*}} takes the cloning heap and the inputs flowing around the
-% loop and produces two copies of these inputs. One copy is produced as
-% the overall output and another is fed back around the loop.
-
-% %subcode{opsem} include main
-% %! columnStyle = rcl
-% % solve_f &:& bool * bool <-> bool * Bool
-% % solve_f &=& trace* (
-% % && (assocr* (*) id) (;) 
-% % && assocr* (;)
-% % && (id (*) swap*) (;)
-% % && assocl* (;)
-% % && (assocr* (*) id) (;)
-% % && (clone2 (*) id) (;)
-% % && swap* (;)
-% % && (swap* (*) id) (;)
-% % && ((swap* (*) id) (*) id) (;)
-% % && assocl* (;)
-% % && assocl* (;)
-% % && (sat_f (*) id) (;)
-% % && (assocr* (*) id) (;)
-% % && (swap* (*) id) (;) 
-% % && ((id (*) swap*) (*) id) (;)
-% % && (assocl* (*) id) (;)
-% % && ((id (*) swap*) (*) id))
-
-% We can test our {{solve_f}} combinator using several SAT
-% instances. Here are two possible instances. The first instance is
-% satisfied by {{(false,false)}} and the
-% second is satisfied by {{(false,true)}} and {{(true,true)}}.
-
-% %subcode{opsem} include main
-% %! columnStyle = rcl
-% % iso_{f_1} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
-% % iso_{f_1} &=& (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
-% % && (((swap+ (*) id) (*) id) (;) toffoli (;) ((swap+ (*) id) (*) id)) (;) 
-% % && (id (*) swap+)
-% % 
-% % iso_{f_2} &:& ((bool * bool) * bool) <-> ((bool * bool) * bool)
-% % iso_{f_2} &=& toffoli
-
-% It can indeed be verified using the semantics that the {{solve_f}}
-% combinators instantiated with the SAT instances produce the expected
-% results.
+For every value {{v : b}}, we define the \emph{pattern} 
+{{ bra{v} : singleton{v}  1}} which
+matches exactly {{v}} and nothing else. The type of this pattern is a function
+which
 
@@
 \section{Conclusion}
-\label{sec:cat}
+\label{sec:conc}
 
-We have introduced the idea of \emph{fractional types} in the context of a
-reversible language founded on type isomorphisms and preservation of
-information. Values of fractional types represent \emph{negative
-  information}, a concept which is difficult to introduce in a conventional
-language that allows arbitrary creation and deletion of information but which
-is much simpler to deal with when the surrounding language infrastructure
-guarantees preservation of information. Fractional types and values can be
-used to express a simple and elegant notion of higher-order functions: a
-function {{b1 :-* b2}} is a first-class value consisting of negative
-information drawn from {{b1}} and positive information drawn from {{b2}}.
-
-The interpretation of our language {{PiTF}} in the category of sets and
-relations is adequate in the sense that it produces a language in which every
-program is a reversible relation and in which relations are first-class
-values. This relational model is however, unsatisfactory, for several
-reasons:
-
-\begin{itemize}
-\item the type {{1/b}} is interpreted in the same way as {{b}}, which gives no
-  insight into the ``true meaning'' of fractionals;
-\item the interpretation is inconsistent with the view of types as algebraic
-  structures; for example, {{Pi}} with the empty type, is the
-  categorification of a semiring but although fractional types in {{PiTF}}
-  syntactically ``look like'' rational numbers, there is no such formal
-  connection to the rational numbers;
-\item finally, we have lost some delicate structure moving from {{Pi}} to
-  {{PiTF}} as we can express arbitrary relations between types and not just
-  \emph{isomorphisms}.
-\end{itemize}
-
-For these reasons, it is interesting to consider other possible semantic
-interpretations of fractionals. A natural alternative is to use the
-``canonical'' compact closed category, that is finite dimensional vector
-spaces and linear
-maps~\cite{Selinger:2011:FDH:1942319.1942398,hasegawa2008finite}
-over fields of characteristic $0$ (or even the category of finite dimensional
-Hilbert spaces). Let us fix an arbitrary field $k$ of characteristic $0$.
-Then each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
-vector space $\mathbf{V}_b$ over the field $k$. In particular:
-\begin{itemize}
-  \item every vector space contains a zero vector which means that the type
-    {{0}} (if included in {{PiTF}}) would not be the ``empty'' type.
-    Furthermore all combinators would be
-    strict as they would have to map the zero vector to itself. 
-  \item the type {{1}} is interpreted as a 1-dimensional vector space and hence
-    is isomorphic to the underlying field.
-  \item the fractional type {{1/b}} is interpreted as the dual vector space
-    to the vector space $\mathbf{V}_b$ representing {{b}} consisting of all
-    the \emph{linear functionals} on $\mathbf{V}_b$.
-  \item one can then validate certain desirable properties: {{1/(1/b)}} is
-    isomorphic to~{{b}}; and {{eps*_b}} corresponds to a bilinear form which
-    maps a dual vector and a vector to a field element.
-\end{itemize}
-
-In such categories, the fractional type is given a non-trivial
-interpretation. Indeed, while the space of column vectors is isomorphic to
-the space of row vectors, they are nevertheless quite different, being
-$(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the category
-provides a more refined model in which isomorphic negative-information values
-and positive-information are not identified.
-
-Although this semantics appears to have ``better'' properties than the
-relational one, we argue that it is not yet \emph{the} ``perfect''
-semantics. By including the zero vector, the language has morphisms that do
-not correspond to isomorphisms (in particular it allows partial morphisms by
-treating the {{0}} element of the zero-dimensional vector space as a
-canonical ``undefined'' value).  It is also difficult to reconcile the
-interpretation with the view that the types correspond to the (positive)
-rational numbers, something we are actively seeking. What would really be a
-``perfect'' interpretation is one in which we can only express
-{{Pi}}-isomorphisms as first class values and in which the types are
-interpreted in a way that is consistent with the rational numbers.
-
-Fortunately, there is promising significant work on the groupoid
-interpretation of type theory~\cite{Hofmann96thegroupoid} and on the 
-categorification of the rational numbers~\cite{math/9802029} that may well give
-us the model we desire.  The fundamental idea in both cases is that groupoids
-(viewed as sets with explicit isomorphisms as morphisms) naturally have a
-\emph{fractional cardinality}.  Types would be interpreted as groupoids, and
-terms would be (invertible) groupoid actions.  The remaining challenge is to
-identify those groupoid actions which are proper generalizations of
-isomorphisms \emph{and} can be represented as groupoids, so as to obtain a
-proper interpretation for a higher-order language.  As the category of
-groupoids is cartesian closed, this appears eminently feasible.
-
-Another promising approach is the use of dependent types for {{eps*_b}} and
-{{eta*_b}}; more precisely, {{eps*_b}} would have type 
-{{Sigma (v:b) (1/v,v) <-> 1}} where {{(1/v,v)}} here denotes a singleton type.
-This extra precision appears to restrict combinators of {{PiTF}} back to
-denoting only isomorphisms.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-\section{Categorical Background}
-\label{app:cat}
-
-We recall that a \emph{symmetric monoidal category} is a category together
-with a bifunctor $\otimes$, a distinguished object $I$, and natural
-isomorphisms $\alpha_{A,B,C} : (A \otimes B) \otimes C \rightarrow A \otimes
-(B \otimes C)$, $\lambda_A : A \rightarrow I \otimes A$, and $\sigma_{A,B} :
-A \otimes B \rightarrow B \otimes A$ subject to standard coherence
-conditions~\cite{nla.cat-vn1051288}. Following common practice
-(e.g.,~\cite{Selinger:2007:DCC:1229185.1229207}), we write $\rho_A =
-\sigma_{I,A} \circ \lambda_A : A \rightarrow A \otimes I$.
-
-A \emph{compact closed category} is a symmetric monoidal category where each
-object $A$ is assigned a dual object $A^*$, together with a unit map $\eta_A
-: I \rightarrow A^* \otimes A$ and a counit map $\epsilon_A : A \otimes A^*
-\rightarrow I$, such that:
-\[\begin{array}{rcl}
-\lambda_A^{-1} \circ (\epsilon_A \otimes A) \circ \alpha_{A,A^*,A}^{-1} \circ (A \otimes \eta_A) \circ \rho_A &=& \mathit{id}_A \\
-\rho_{A^*}^{-1} \circ (A^* \otimes \epsilon_A) \circ \alpha_{A^*,A,A^*} \circ (\eta_A \otimes A^*) \circ \lambda_A &=& \mathit{id}_{A^*}
-\end{array}\]
-
-A \emph{dagger category} is a category together with an involutive,
-identity-on-objects, contravariant functor $\dagger$. Concretely, this means
-that to every morphism $F : A \rightarrow B$, one associates a morphism
-$f^{\dagger} : B \rightarrow A$, called the \emph{adjoint} of $f$, such that
-for all $f : A \rightarrow B$ and $g : B \rightarrow C$, we have:
-\[\begin{array}{rcl}
-\mathit{id}^\dagger_A &=& \mathit{id}_A \\
-(g \circ f)^\dagger &=& f^\dagger \circ g^\dagger \\
-(f^\dagger)^\dagger &=& f
-\end{array}\]
-
-A \emph{dagger symmetric monoidal category} is a symmetric monoidal category
-with a dagger structure such that the contravariant functor $\dagger$
-coherently preserves the symmetric monoidal structure. Concretely, this
-requirement means that for all $f : A \rightarrow B$ and $g : C \rightarrow
-D$, we have:
-\[\begin{array}{rcl}
-(f \otimes g)^\dagger &=& f^\dagger \otimes g^\dagger \\
-\alpha^\dagger_{A,B,C} &=& \alpha^{-1}_{A,B,C} \\
-\lambda^\dagger_A &=& \lambda^{-1}_A \\
-\sigma^\dagger_{A,B} &=& \sigma^{-1}{A,B}
-\end{array}\]
-
-\begin{definition}[Dagger Compact Closed Category]
-\label{def:cat}
-A \emph{dagger compact closed category} is a dagger symmetric monoidal
-category that is also compact closed and such that for all $A$:
-\[
-\eta_A = \sigma_{A,A^*} \circ \epsilon^\dagger_A
-\]
-\end{definition}
-
-\todo{biproducts}
-
-\begin{verbatim}
-homotopy equivalence???
-
-axioms for field or meadow; actually we probably a semifield
-
-definition of logical reversibility with relations
-
-definition of information preservation in the case of relations; fanout as an
-example which we can write
-
-explain in detail the size of 2 * 1/2; it should have exactly one element;
-there are two elemens in 2 but the 1/2 identifies them somehow; be precise
-
-we can get the empty relation (eta ; id * swap; eps) what does that mean?
-
-connection to vector spaces; projectors; inner product. Given dual vectors,
-we can have a standalone |0><0| which is a piece of a function or a projector
-if viewed by itself; we can also have |v><v'| which produces the scalars YES
-or NO. We have an isomorphism between matrices and vectors by viewing 
-|0><0|+|1><1| as |00>+|11>
-
-Perhaps if restrict the ``top-level'' to non-fractional types, we can never
-observe a function built from from fractionals; we must apply it and in that
-case, the system overall behaves like Pi. (no empty relation for example)???
-Idea suggested by Zach is to change eps to (1/a) * a * a <-> a
-
-The code implementing the 16 relations... all 16 relations have the same
-information content as () ??? So we should treat (1/bool * bool) as opaque
-things; we can only extract information from them by applying them.
-
-Include dneg, name (inv cnot) as example of programs which when fed
-non-fractional values produce fractional values
-
-Also include recipT which shows that 1/ distributes over products and tens
-which shows that we can manipulate fractional types in interesting ways
-
-Using relational semantics is good because it's clear; but it's also bad
-unless we spend time to explain the operational model
-
-Explain 'name' in detail; it IS a faithful representation of the input
-relation as a value; we don't get to see the full relation though; we
-interact with it by giving it inputs and observing the related ouptut come
-out. If we make two copies and use each in a different context (applying each
-to a different value, we can ``see'' that it's really a full relation and not
-just one branch.)
-
-total functions have size 1; partial functions have fractional sizes!!!
-
-denotation of type in Pi is a set; in Pi/ we get a set with an equivalence
-relation (Id by default) but fractionals introduce first-class equivalence
-relations that can be used to restrict the sets.
-
----
-
-Having re-read Baez-Dolan, I now feel better.
-
-On 13-02-08 08:53 AM, Amr Sabry wrote:
-> I think all the observations in this email are due to the fact that we
-> haven't formalized the size of fractionals using homotopy equivalence. I
-> started doing this but abandoned it for now but the bottom line is that
-> the size of b * 1/b is 1.
-
-Again, agree - more deeply so now.  In more detail, my current thinking:
-- v \in b has size 1.  So |b| counts how many elements it has, whenever 
-b is a type of Pi.  The sum and product rules apply.
-- for b * 1/b to have size 1, then 1/b must have size 1/|b| (assuming 
-these are independent types, which we have been assuming all along)
-
-Attempt #1
-- there are |b| elements in 1/b, and if size is additive, ( 1/|b| = 
-sum_{1/v \in 1/b} |1/v| ) should hold; since |1/v| should be constant, 
-this sum is |b|*c, so c = 1/|b|^2.  Weird.
-
-Let us look at a single 'function' in bool -o bool, namely the identity 
-function.  It should be 'the same' as { (1/true, true), (1/false, false) 
-}.  The cardinality of that set is (1/(2^2)*1 + 1/(2^2)*1) = 1/4 + 1/4 = 
-1/2.   From weird to probably-wrong.
-
-Attempt #2
-- the elements of 1/b are not distinguishable from each other (i.e. they 
-are all isomorphic).  So each element 1/v has b automorphisms, so |1/b| 
-= sum_{aut classes of 1/v \in 1/b} 1/|aut 1/v| = sum_{singleton} 1/|b| = 
-1/|b|.  [as per p.14 of Baez-Dolan]
-
-The size of the set { (1/true, true), (1/false, false) } is now 1/2*1 + 
-1/2*1 = 1.  So this 'set' indeed represents a single "concept".
-
-Note that in attempt #2, we get a new phenomenon.  Consider the 
-(partial) relation {(1/true, false)}, also of type bool -o bool.  It has 
-size 1/2 !  So, if I have not erred somewhere, only total functions have 
-size 1.  Which, I must admit, I rather like.  It's not that partial 
-functions are outlawed, they just don't pull as much weight [pun intended].
-
-So 1/b is very much like a collection of |b| constraints, all of which 
-are "the same" (externally).  While they may have internal structure, it 
-is not visible to any of our combinators, so up to iso, there is only a 
-single element of 1/b, of size 1/|b|.
-
-Yet another way to look at it: let's assume we are looking at FinSet_0, 
-where in fact our universe U has finitely (say N) things in it.  Then 
-the singleton set 1 is actually the isomorphism class of {u_1, ..., 
-u_N}.  There are N of them, but they have N automorphisms in a single 
-orbit, so the quotient has size 1 (i.e. |S| = N but |G| = N too, so |S 
-// G| = 1 ).
-
-So an element of 1/b is closer to an action of b on b; we call it a 
-constraint, or a pattern-match.  I think thinking of it as 'actions' is 
-definitely the right way to go, as we will be able to have different 
-kinds of actions on a type b that just those that come from the elements.
-
-> I don't know how much of a priority this is but if it is we should try
-> to work it out in detail. The good and bad news is that we will have a
-> model much richer and much more complicated than relations. My thought
-> when writing sec. 3 was that we would use the simple but slightly
-> inadequate model of relations (which for one thing confuses v and 1/v)
-> and simply mention that the right abstraction is that of 'biproduct
-> dagger compact closed categories' which presumably would include the
-> richer model.
-
-I am as convinced that the right model is 'biproduct dagger compact 
-closed categories' as I am that the right model for the type level is 
-'categorified field'.
-
-\end{verbatim}
-
-\begin{comment}
-Each type in {{PiTF}} is interpreted as a \emph{groupoid}. A groupoid can be
-equivalently viewed as an oriented graph, a generalization of a group, or a
-special category. We present the categorical view below.
-
-A groupoid is defined by:
-\begin{itemize}
-\item a set of objects $x,y,\ldots$;
-\item for each pair of objects, a (possibly empty) set $G(x,y)$ of morphisms
-  $x \rightarrow y$; 
-\item for each object $x$, there exists an identity morphism {{id_x in G(x,x)}};
-\item for each triple of objects $x,y,z$, we have a function 
-{{comp_{x,y,z} : G(x,y) * G(y,z) -> G(x,z)}}
-\item there exists a function {{inv_{x,y} : G(x,y) -> G(y,x)}}
-\end{itemize}
-such that for all {{f : x -> y}}, {{g : y -> z}}, and {{h : z -> w}}, we have:
-%subcode{bnf} include main
-% comp_{x,x,y}(id_x,f) &=& f \\
-% comp_{x,y,y}(f,id_y) &=& f \\
-% comp_{x,y,z}(f, comp_{y,z,w}(g,h)) &=& comp_{x,z,w}(comp_{x,y,z}(f,g),h) \\
-% comp_{y,x,y}(inv_{x,y}(f),f) &=& id_y \\
-% comp_{x,y,x}(f,inv_{x,y}(f)) &=& id_x
-
-\noindent We now proceed to explain how each {{PiTF}} type denotes a groupoid. First,
-each {{Pi}} type denotes a set as follows.
-
-\begin{definition}[Denotation of {{Pi}} Value Types {{ [[ b ]] }}]
-\label{chx:def:denot}
-Each {{Pi}}type denotes a finite set of values as follows:
-%subcode{opsem} include main
-%! <- = \leftarrow
-%! union = \cup
-% [[1]] '= {[ () ]}
-% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
-% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
-\end{definition}
-
-For each of the types {{b}} above, the corresponding groupoid consists of the
-set {{ [[b]] }} of elements with only the trivial identity morphisms. The
-interesting groupoid is the one associated with {{1/b}}. It is defined as
-follows.
-
-Fractional types and values add considerable expressiveness to our
-language. 
-\end{comment}
-
-If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 
-{{c : b1 <-> b2}} would define a total one-to-one function between the sets 
-{{ [[ b1 ]] }} and {{ [[ b2 ]] }}. 
-This is indeed the situation for the language {{Pi}} 
-defined in the previous section. In the presence of fractionals, this simple 
-interpretation of combinators as one-to-one functions is not sufficient%
-\footnote{The interpretation of combinators as one-to-one functions is also 
-not valid in the presence of recursive 
-types~\cite{rc2011,James:2012:IE:2103656.2103667}.}.
-The intuitive reason, is that the fission point combinator 
-{{eta*_b : 1 <-> 1/b * b}} must be
-able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
-{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
-accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
-not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
-undefined, i.e., it does not produce {{():1}}.
-
-discuss two generalizations of the semantics
-
-vector spaces (more refined because it doesn't collapse 1/b to b; also each
-vector space must have a zero vector; so we have an explicit notion of
-``error'' which we can use and track at run time. Adding this error
-eliminates at RUN TIME relations which are not isomorphisms I think.
-
-a potentially better and more semantics is based on groupoids; mention that
-we can have sets (with structure) who cardinality is fractionals and that this
-would formalize nicely the negative information

### 2013-03-27 rc13/frac2.tex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background: {{Pi}} and {{PiTF}} }
-\label{sec:pi}
-
-We review our language {{Pi}} providing the necessary background and
-context for our higher-order extension.\footnote{The presentation in this
-  section focuses on the simplest version of {{Pi}}. Other versions
-  include the empty type, recursive types, and trace operators but these
-  extensions are orthogonal to the higher-order extension emphasized in this
-  paper.} The terms of {{Pi}} are not classical values and functions;
-rather, the terms are isomorphism witnesses.  In other words, the terms of
-{{Pi}} are proofs that certain ``shapes of values'' are isomorphic.
-And, in classical Curry-Howard fashion, our operational semantics shows how
-these proofs can be directly interpreted as actions on ordinary values which
-effect this shape transformation. Of course, ``shapes of values'' are very
-familiar already: they are usually called \emph{types}.  But frequently one
-designs a type system as a method of classifying terms, with the eventual
-purpose to show that certain properties of well-typed terms hold, such as
-safety.  Our approach is different: we start from a type system, and then
-present a term language which naturally inhabits these types, along with an
-appropriate operational semantics.
-
-\paragraph*{Data.}
-We view {{Pi}} as having two levels:  it has traditional values, given by:
-%subcode{bnf} include main
-% values, v ::= () | left v | right v | (v, v)
-
-\noindent and these are classified by ordinary types:
-%subcode{bnf} include main
-% value types, b ::= 1 | b + b | b * b
-
-\noindent 
-Types include the unit type {{1}}, sum types {{b1+b2}}, and product types
-{{b1*b2}}.  Values include {{()}} which is the only value of type {{1}},
-{{left v}} and {{right v}} which inject~{{v}} into a sum type, and
-{{(v1,v2)}} which builds a value of product type.  But these values should be
-regarded as largely ancillary: we do not treat them as first-class citizens,
-and they only occur when observing the effect of an isomorphism.
-
-\paragraph*{Isomorphisms.} The terms of {{Pi}} witness
-type isomorphisms of the form {{b <-> b}}. They consist of base
-isomorphisms, as defined below, and their composition.
-%subcode{bnf} include main
-%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
-%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
-%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
-%identl* :&  1 * b & <-> & b &: identr*
-%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
-%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
-%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor
-
-\noindent Each line of the above table introduces a pair of dual
-constants\footnote{where {{swap*}} and {{swap+}} are self-dual.} that witness
-the type isomorphism in the middle.  These are the base (non-reducible) terms
-of the second, principal level of {{Pi}}. Note how the above has two
-readings: first as a set of typing relations for a set of constants. Second,
-if these axioms are seen as universally quantified, orientable statements,
-they also induce transformations of the (traditional) values. The
-(categorical) intuition here is that these axioms have computational content
-because they witness isomorphisms rather than merely stating an extensional
-equality.
-
-The isomorphisms are extended to form a congruence relation by adding the
-following constructors that witness equivalence and compatible closure:
-
-%subcode{proof} include main
-%@  ~
-%@@ id : b <-> b 
-%
-%@ c : b1 <-> b2
-%@@ sym c : b2 <-> b1
-%
-%@ c1 : b1 <-> b2
-%@ c2 : b2 <-> b3
-%@@ c1(;)c2 : b1 <-> b3
-%---
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
-%
-%@ c1 : b1 <-> b3
-%@ c2 : b2 <-> b4
-%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
-\noindent The syntax is overloaded: we use the same symbol at the value-type level
-and at the isomorphism-type level for denoting sums and products.  Hopefully
-this will not cause undue confusion.
-
-It is important to note that ``values'' and ``isomorphisms'' are completely
-separate syntactic categories which do not intermix. The semantics of the
-language come when these are made to interact at the ``top level'' via
-\emph{application}: 
-subcode{bnf} include main
-top level term, l ::= c v
-
-\noindent
-To summarize, the syntax of {{Pi}} is given as follows.
-
-\begin{definition}{(Syntax of {{Pi}})}
-\label{def:Pi}
-%subcode{bnf} include main
-% value types, b ::= 1 | b+b | b*b 
-% values, v ::= () | left v | right v | (v,v) 
-%
-% iso.~types, t ::= b <-> b
-% base iso ::= swap+ | assocl+ | assocr+ 
-%     &|& unite | uniti | swap* | assocl* | assocr* 
-%     &|& dist | factor 
-% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
-% top level term, l ::= c v
-\end{definition}
-
-The language presented above, at the type level, models a commutative ringoid
-where the multiplicative structure forms a commutative monoid, but the
-additive structure is just a commutative semigroup.  Note that the version of
-{{Pi}} that includes the empty type with its usual laws exactly captures, at
-the type level, the notion of a \emph{semiring} (occasionally called a
-\emph{rig}) where we replace equality by isomorphism.  Semantically, {{Pi}}
-models a \emph{bimonoidal category} whose simplest example is the category of
-finite sets and bijections. In that interpretation, each value type denotes a
-finite set of a size calculated by viewing the types as natural numbers and
-each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
-denoted by~{{b1}} and~{{b2}}. (We discuss the operational semantics in
-conjuction with our extension in the next section.)
-
-Operationally, the semantics of {{Pi}} is given using two mutually recursive
-interpreters: one going forward and one going backwards. The use of {{sym}}
-switches control from one evaluator to the other.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Fractional Types}
+\section{X}
 
@@
 
-For every value {{v : b}}, we define the \emph{pattern} 
-{{ bra{v} : singleton{v}  1}} which
-matches exactly {{v}} and nothing else. The type of this pattern is a function
-which
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-\label{sec:conc}
+For every value {{v : b}}, we define the \emph{pattern} {{ bra{v} :
+    singleton{v} 1}} which matches exactly {{v}} and nothing else. The type
+of this pattern is a function which
 

### 2013-06-01 summer13/frac.tex
added summer stuff

\documentclass{llncs}
\usepackage{url}
\usepackage{proof}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{comment}

\newcommand{\dgm}[2][1.5]{
\begin{center}
\scalebox{#1}{
\includegraphics{diagrams/#2.pdf}
}
\end{center}
}

\newcommand{\todo}[1]{\textbf{TODO:} #1}
\newcommand{\jacques}[1]{\textsc{Jacques says:} #1}
\newcommand{\amr}[1]{\textsc{Amr says:} #1}

\hyphenation{a-reas}

%subcode-inline{bnf-inline} name Pi
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! foldB = fold_B
%! unfoldB = unfold_B
%! foldN = fold_N
%! unfoldN = unfold_N
%! trace+ = \mathit{trace}^{+}
%! trace* = \mathit{trace}^{\times}
%! :-* = \multimap
%! :-+ = \multimap^{+}
%! emptyset = \emptyset

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! ^^^ = ^{\dagger}
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! -> = \rightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! Sigma = \Sigma
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pow = \mathcal{P}
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}
%! bullet = \bullet
%! * = \times

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{Higher-Order Reversible Functions}

To get from {{Pi}} types and combinators to a higher order setting, we hope to
add \emph{fractional types} to the language. Recall that to construct the
(positive) rational numbers {{Q}}, we can take pairs of positive integers
{{(m, n)}} and quotient by the equivalence relation that says {{(m1, n1)}} is
equal to {{(m2, n2)}} if {{m1 * n2 = m2 * n1}}. We can write this as an
inference rule as follows:

%subcode{proof} include main
%@ m1 * n2 = m2 * n1
%@@ (m1, n1) = (m2, n2)

\noindent
Adding in computational content, pi-style, suggests that we want something like:

%subcode{proof} include main
%@ c : A1 * B2 <-> B1 * A2
%@@ \hat{c} : (B1 |--> A1) <=> (B2 |--> A2)

\noindent
However, this does not quite work. To see why, recall that we want to treat the
pair {{(m, n)}} (written here as {{A |--> B}}) as an internalization of the
reversible function space. Thus, the type {{(B1 |--> A1) <=> (B2 |--> A2)}}
should be a transformation of a {{Pi}} combinator of type {{B1 <-> A1}} into a
{{Pi}} combinator of type {{B2 <-> A2}}. In order to do this, we need a trace
combinator (TODO: draw a diagram that demonstrates this) with the type

%subcode{proof} include main
%@ c : A * C <-> B * C
%@@ trace c : A <-> B

\noindent
This is an issue, since prior attempts at defining an adequate trace operator
failed: we got a language that could express all relations, but was no longer
reversible in a meaningful sense.

\subsection{Defining trace}

The trace operator had two problems: it allowed arbitrary function definition,
and it would not work in a language that contained the empty type {{0}}. The
former problem arose due to the logic programming-style interpretation we
assigned to trace: it would create *all* possible values of type {{C}}, then run
the given combinator on them and throw away the branches that did not return the
same {{C}} value.

We can solve this problem by parameterizing the trace operator with a value {{v
: C}}. This results in the following new typing rule for trace:

%subcode{proof} include main
%@ c : A * C <-> B * C
%@ v : C
%@@ trace_v c : A <-> B

\noindent
This ends up solving both problems: we give the trace a default value, and
prevent it from running on the empty type (since {{C}} now must be inhabited).
However, we still must assign dynamic semantics to trace. The na\"ive solution
of running {{c}} on the pair of the input value and {{v}} does not work, as this
is trivially irreversible---consider, for example, {{trace_v swap*}}, which
takes any value of the same type as {{v}} and returns {{v}}. Thus, the semantics
for {{trace}} must be slightly more complicated.

Recall that in {{Pi}}, combinators can be viewed as permutations on finite sets.
A combinator of type {{A * C <-> B * C}} can be seen as two separate
permutations: one on {{A}} (or equivalently, B, since they must have the same
size and thus the same denotation as a finite set) and one on {{C}}. If we
apply the fragment of the combinator that acts on {{C}} enough times, we will
eventually get out the same value that we started with\footnote{I want to prove
that this is actually true; having the extra {{A}} there seems like it might
complicate things.}. Thus, the semantics that we want is for {{(trace_v c) v'}}
to run {{c}} on the pair of {{v'}} and the last {{C}} value (starting with
{{v}}) until the {{C}} value matches {{v}}, at which point it returns the {{B}}
value returned by the combinator. In pseudocode:

\begin{verbatim}
TODO: figure out how to do the code environment in subcode

trace v c vin =
  v0 = v
  <vout, v'> = c <vin, v>

  while (v' != v)
    <vout, v'> := c <vin, v'>

  return vout
\end{verbatim}

A more concise formulation as an inference rule is as follows:

%subcode{proof} include main
%@ \exists v0 : C. c <v_{in}, v> = <v_{out}, v>
%@@ (trace_v c) v_{in} = v_{out}

This is easily reversible: {{(trace_v c)dagger = trace_v c
dagger}}\footnote{Also need to prove this.}.

\subsection{Higher-Order Combinators}

Now that we have the trace combinator defined, we can proceed with defining
higher-order combinators. Note that this parameterizes our definition of
{{\hat{c}}} by {{v}}, the value of type {{A}} (or equivalently, type {{B}}) used
in the trace:

%subcode{proof} include main
%@ c : A1 * B2 <-> B1 * A2
%@@ \hat{c}_v : (B1 |--> A1) <=> (B2 |--> A2)

\noindent
We want this to be a transformation on combinators, taking a combinator of type
{{B1 <-> A1}} to a combinator of type {{B2 <-> A2}}. This can be achieved by
letting the combinator {{\hat{c}_v c'}} be {{trace_v (c ; c' * id)}}, where {{c
: B1 <-> A1}} and {{v : A1}}.

From here, we want to define bifunctors {{/}}, {{*}}, and {{+}}, to make sure
that our definition of functions is truly higher-order.

\begin{itemize}

\item {{(B |--> A) / (D |--> C) = (A * D) |--> (B * C)}}, with {{{[c1]}_{v1} :
((B |--> A) <=> (D |--> C)) / {[c2]}_{v_2} : ((B' |--> A') <=> (D' |--> C')) =
{[shuffle ; c1 * c2 dagger ; shuffle]}_{<v1, c2 v2>} : B1 * A2 |--> A1 * B2 <=>
D1 * C2 |--> C1 * D2}}, where {{shuffle : (A * B) * (C * D) <-> (A * C) * (B *
D)}}.

\item {{(B |--> A) * (D |--> C) = (B * D) |--> (A * C)}}, with

\end{itemize}

\end{document}

### 2013-06-01 summer13/frac2.tex
added summer stuff

--- 2013-06-01 summer13/frac.tex
@@
 %@ c : A1 * B2 <-> B1 * A2
-%@@ \hat{c} : (B1 |--> A1) <=> (B2 |--> A2)
+%@@ {[c]} : (B1 |--> A1) <=> (B2 |--> A2)
 
@@
 %subcode{proof} include main
-%@ \exists v0 : C. c <v_{in}, v> = <v_{out}, v>
+%@ exists v0 : C. c <v_{in}, v> = <v_{out}, v>
 %@@ (trace_v c) v_{in} = v_{out}
 
-This is easily reversible: {{(trace_v c)dagger = trace_v c
-dagger}}\footnote{Also need to prove this.}.
+This is easily reversible: {{(trace_v c)dagger = trace_v (c dagger)}}
+\footnote{Also need to prove this.}.
 
@@
 higher-order combinators. Note that this parameterizes our definition of
-{{\hat{c}}} by {{v}}, the value of type {{A}} (or equivalently, type {{B}}) used
+$\{c\}$ by {{v}}, the value of type {{A}} (or equivalently, type {{B}}) used
 in the trace:
@@
 %@ c : A1 * B2 <-> B1 * A2
-%@@ \hat{c}_v : (B1 |--> A1) <=> (B2 |--> A2)
+%@@ {[c]}_v : (B1 |--> A1) <=> (B2 |--> A2)
 
@@
 {{B1 <-> A1}} to a combinator of type {{B2 <-> A2}}. This can be achieved by
-letting the combinator {{\hat{c}_v c'}} be {{trace_v (c ; c' * id)}}, where {{c
-: B1 <-> A1}} and {{v : A1}}.
+letting the combinator {{{[c]}_v c'}} be {{trace_v (c ; c' * id)}}, where
+{{c : B1 <-> A1}} and {{v : A1}}.
 
@@
 
-\item {{(B |--> A) / (D |--> C) = (A * D) |--> (B * C)}}, with {{{[c1]}_{v1} :
-((B |--> A) <=> (D |--> C)) / {[c2]}_{v_2} : ((B' |--> A') <=> (D' |--> C')) =
-{[shuffle ; c1 * c2 dagger ; shuffle]}_{<v1, c2 v2>} : B1 * A2 |--> A1 * B2 <=>
-D1 * C2 |--> C1 * D2}}, where {{shuffle : (A * B) * (C * D) <-> (A * C) * (B *
-D)}}.
+\item {{(B |--> A) / (D |--> C) = (A * D) |--> (B * C)}}, with
+
+{{{[c1]}_{v1} : ((B |--> A) <=> (D |--> C)) / {[c2]}_{v_2} : ((B' |--> A') <=> (D' |--> C')) = {[shuffle ; c1 * c2dagger ; shuffle]}_{<v1, c2 v2>}}}
+%%: B1 * A2 |--> A1 * B2 <=> D1 * C2 |--> C1 * D2}},
+where
+{{shuffle : (A * B) * (C * D) <-> (A * C) * (B * D)}}.
 

### 2013-06-03 summer13/frac2.tex
added some background and started working on syntax of pi with fractions

@@
-\documentclass{llncs}
+\documentclass{sigplanconf}
 \usepackage{url}
@@
 }
-
 \newcommand{\todo}[1]{\textbf{TODO:} #1}
@@
 
-\hyphenation{a-reas}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{proposition}[theorem]{Proposition}
 
@@
 %! ::= = ::=
+%! amp = \&
 %! /= = \neq
@@
 
+\begin{document}
+\conferenceinfo{POPL '14}{date, City.} 
+\copyrightyear{2014}
+\copyrightdata{[to be supplied]}
+
+\title{\ldots}
+\authorinfo{\ldots}
+           {\ldots}
+           {\ldots}
+
+\maketitle
+
+\begin{abstract}
+\ldots
+\end{abstract}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Background: {{Pi}} }
+\label{sec:pi}
 
-\begin{document}
+We review our language {{Pi}} providing the necessary background and
+context for our higher-order extension.\footnote{The presentation in this
+  section focuses on the simplest version of {{Pi}}. Other versions
+  include the empty type, recursive types, and trace operators but these
+  extensions are orthogonal to the higher-order extension emphasized in this
+  paper.} The terms of {{Pi}} are not classical values and functions;
+rather, the terms are isomorphism witnesses.  In other words, the terms of
+{{Pi}} are proofs that certain ``shapes of values'' are isomorphic.
+And, in classical Curry-Howard fashion, our operational semantics shows how
+these proofs can be directly interpreted as actions on ordinary values which
+effect this shape transformation. Of course, ``shapes of values'' are very
+familiar already: they are usually called \emph{types}.  But frequently one
+designs a type system as a method of classifying terms, with the eventual
+purpose to show that certain properties of well-typed terms hold, such as
+safety.  Our approach is different: we start from a type system, and then
+present a term language which naturally inhabits these types, along with an
+appropriate operational semantics.
 
+\paragraph*{Data.}
+We view {{Pi}} as having two levels:  it has traditional values, given by:
+%subcode{bnf} include main
+% values, v ::= () | left v | right v | (v, v)
+
+\noindent and these are classified by ordinary types:
+%subcode{bnf} include main
+% value types, b ::= 1 | b + b | b * b
+
+\noindent 
+Types include the unit type {{1}}, sum types {{b1+b2}}, and product types
+{{b1*b2}}.  Values include {{()}} which is the only value of type {{1}},
+{{left v}} and {{right v}} which inject~{{v}} into a sum type, and
+{{(v1,v2)}} which builds a value of product type.  These values should be
+regarded as largely ancillary: we do not treat them as first-class citizens,
+and they only occur when observing the effect of an isomorphism.
+
+\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
+(value) type isomorphisms.  They are also typed, by the shape of the (value)
+type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses
+to the following isomorphisms:
+%subcode{bnf} include main
+%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
+%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
+%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
+%identl* :&  1 * b & <-> & b &: identr*
+%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
+%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
+%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor
+
+\noindent Each line of the above table introduces a pair of dual
+constants\footnote{where {{swap*}} and {{swap+}} are self-dual.} that witness
+the type isomorphism in the middle.  These are the base (non-reducible) terms
+of the second, principal level of {{Pi}}. Note how the above has two
+readings: first as a set of typing relations for a set of constants. Second,
+if these axioms are seen as universally quantified, orientable statements,
+they also induce transformations of the (traditional) values. The
+(categorical) intuition here is that these axioms have computational content
+because they witness isomorphisms rather than merely stating an extensional
+equality.
+
+The isomorphisms are extended to form a congruence relation by adding the
+following constructors that witness equivalence and compatible closure:
+
+%subcode{proof} include main
+%@  ~
+%@@ id : b <-> b 
+%
+%@ c : b1 <-> b2
+%@@ sym c : b2 <-> b1
+%
+%@ c1 : b1 <-> b2
+%@ c2 : b2 <-> b3
+%@@ c1(;)c2 : b1 <-> b3
+%---
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
+%
+%@ c1 : b1 <-> b3
+%@ c2 : b2 <-> b4
+%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
+\noindent The syntax is overloaded: we use the same symbol at the value-type level
+and at the isomorphism-type level for denoting sums and products.  Hopefully
+this will not cause undue confusion.
+
+It is important to note that ``values'' and ``isomorphisms'' are completely
+separate syntactic categories which do not intermix. The semantics of the
+language come when these are made to interact at the ``top level'' via
+\emph{application}: 
+%subcode{bnf} include main
+% top level term, l ::= c v
+
+\noindent
+To summarize, the syntax of {{Pi}} is given as follows.
+
+\begin{definition}{(Syntax of {{Pi}})}
+\label{def:Pi}
+%subcode{bnf} include main
+% value types, b ::= 1 | b+b | b*b 
+% values, v ::= () | left v | right v | (v,v) 
+%
+% iso.~types, t ::= b <-> b
+% base iso ::= swap+ | assocl+ | assocr+ 
+%     &|& unite | uniti | swap* | assocl* | assocr* 
+%     &|& dist | factor 
+% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+% top level term, l ::= c v
+\end{definition}
+
+The language presented above, at the type level, models a commutative ringoid
+where the multiplicative structure forms a commutative monoid, but the
+additive structure is just a commutative semigroup.  Note that the version of
+{{Pi}} that includes the empty type with its usual laws exactly captures, at
+the type level, the notion of a \emph{semiring} (occasionally called a
+\emph{rig}) where we replace equality by isomorphism.  Semantically, {{Pi}}
+models a \emph{bimonoidal category} whose simplest example is the category of
+finite sets and bijections. In that interpretation, each value type denotes a
+finite set of a size calculated by viewing the types as natural numbers and
+each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
+denoted by~{{b1}} and~{{b2}}.
+
+Operationally, the semantics of {{Pi}} is given using two mutually recursive
+interpreters: one going forward and one going backwards. The use of {{sym}}
+switches control from one evaluator to the other. We will present the
+operational semantics in the next section along with the extension with
+fractional types and values. For now, we state without proof that the
+evaluation of well-typed combinators always terminates and that {{Pi}} is
+logically reversible, i.e., that for all combinators {{c : b1 <-> b2}} and
+values {{v1 : b1}} and {{v2 : b2}} we have the forward evaluation of {{c v1}}
+produces {{v2}} iff the backwards evaluation of {{c v2}} produces {{v1}}.
+
+As is usual with type theories, the coherence conditions (pentagonal and
+triangle identity) are not explicit, but would correspond to certain identity
+types being trivial.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Syntax of Fractionals}
+
+The types of {{Pi}} correspond to the natural numbers. Our goal is to have a
+type system that corresponds to the (positive) rational numbers.
+
+\begin{definition}{(Syntax of {{PiTF}})}
+\label{def:PiTF}
+%subcode{bnf} include main
+% value types, r ::= b / b
+% values, f ::= (v |--> v amp c)
+%
+% iso.~types, t ::= r <=> r
+% base iso ::= swap+ | assocl+ | assocr+ 
+%     &|& unite | uniti | swap* | assocl* | assocr* 
+%     &|& dist | factor 
+%     &|& eta | epsilon
+% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
+% top level term, l ::= c v
+\end{definition}
+
+If we restrict our attention to types {{b / 1}} and values {{ () |--> v amp id}},
+we recover {{Pi}}.
+
+In algebra, the rational number {{a/b}} denotes 
+$\{ (a','b) ~|~ a * b' = a' * b \}$. We essentially use the same denotation but
+add computational content, i.e., the type {{a/b}} denotes
+$\{ (a',b',c) ~|~ c : a * b' \leftrightarrow a' * b \}$ where $c$ is a {{Pi}}
+combinator witnessing the equivalence of {{a * b'}} and {{a' * b}}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Higher-Order Reversible Functions}
 
-To get from {{Pi}} types and combinators to a higher order setting, we hope to
-add \emph{fractional types} to the language. Recall that to construct the
-(positive) rational numbers {{Q}}, we can take pairs of positive integers
-{{(m, n)}} and quotient by the equivalence relation that says {{(m1, n1)}} is
-equal to {{(m2, n2)}} if {{m1 * n2 = m2 * n1}}. We can write this as an
-inference rule as follows:
+To get from {{Pi}}~\cite{James:2012:IE:2103656.2103667} types and combinators
+to a higher order setting, we hope to add \emph{fractional types} to the
+language. Recall that to construct the (positive) rational numbers {{Q}}, we
+can take pairs of positive integers {{(m, n)}} and quotient by the
+equivalence relation that says {{(m1, n1)}} is equal to {{(m2, n2)}} if 
+{{m1 * n2 = m2 * n1}}. We can write this as an inference rule as follows:
 
@@
 
+%%%%%%%%%%%%%%%%%%%
 \subsection{Defining trace}
@@
 \begin{verbatim}
-TODO: figure out how to do the code environment in subcode
+TODO: figure out how to do the code 
+environment in subcode
 
@@
 
+%%%%%%%%%%%%%%%%%%%
 \subsection{Higher-Order Combinators}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\bibliographystyle{abbrvnat}
+\bibliography{cites}
 \end{document}

### 2013-06-05 summer13/R.hs
R.hs

@@ -1,0 +1,177 @@
+{-# LANGUAGE TypeOperators, GADTs, RankNTypes #-}
+
+module R where
+
+-----------------------------------------------------------------------
+-- Pi 
+
+data Zero 
+
+class V a where
+instance V Zero
+instance V ()
+instance V Bool
+instance (V a, V b) => V (Either a b) 
+instance (V a, V b) => V (a,b) 
+
+data a :<=> b where 
+-- Congruence
+  Id    :: V a => a :<=> a
+  Sym   :: (V a, V b) => (a :<=> b) -> (b :<=> a) 
+  (:.:) :: (V a, V b, V c) => (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:*:) :: (V a, V b, V c, V d) => 
+           (a :<=> b) -> (c :<=> d) -> ((a,c) :<=> (b,d))
+  (:+:) :: (V a, V b, V c, V d) => 
+           (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+-- (+) is associative, commutative, and has a unit
+  ZeroE   :: V a => Either Zero a :<=> a
+  ZeroI   :: V a => a :<=> Either Zero a
+  SwapP   :: (V a, V b) => Either a b :<=> Either b a
+  AssocLP  :: (V a, V b, V c) => 
+              Either a (Either b c) :<=> Either (Either a b) c 
+  AssocRP  :: (V a, V b, V c) => 
+              Either (Either a b) c :<=> Either a (Either b c) 
+-- (*) is associative, commutative, and has a unit
+  UnitE    :: V a => ((), a) :<=> a
+  UnitI    :: V a => a :<=> ((), a)
+  SwapT    :: (V a, V b) => (a,b) :<=> (b,a) 
+  AssocLT  :: (V a, V b, V c) => (a,(b,c)) :<=> ((a,b),c)
+  AssocRT  :: (V a, V b, V c) => ((a,b),c) :<=> (a,(b,c))
+-- (*) distributes over (+) 
+  DistribZ :: V a => (Zero, a) :<=> Zero
+  FactorZ  :: V a => Zero :<=> (Zero, a)
+  Distrib  :: (V a, V b, V c) => (Either b c, a) :<=> Either (b, a) (c, a)
[…]

### 2013-06-05 summer13/R.hs
more in R.hs mostly questions

@@ -7,12 +7,23 @@ module R where
 
 data Zero 
 
-class V a where
-instance V Zero
-instance V ()
-instance V Bool
-instance (V a, V b) => V (Either a b) 
-instance (V a, V b) => V (a,b) 
+class V a where 
+  elems :: [a]
+
+instance V Zero where
+  elems = []
+
+instance V () where
+  elems = [()]
+
+instance V Bool where
+  elems = [False,True]
+
+instance (V a, V b) => V (Either a b) where
+  elems = map Left elems ++ map Right elems
+
+instance (V a, V b) => V (a,b) where
+  elems = [(a,b) | a <- elems, b <- elems] 
 
 data a :<=> b where 
 -- Congruence
@@ -140,6 +151,30 @@ evalR UnfoldB (Right ()) = False
 data Frac a b =
   forall a' b'. (V a', V b') => Frac { iso :: (a,b') :<=> (a',b) }
 
+instance (V a, V b) => V (Frac a b) where
+  elems = undefined
+
+instance (V a, V b, Show a, Show b) => Show (Frac a b) where
+  show (Frac f) = 
+      show [(a, snd (eval f (a,b'))) | a <- elems, b' <- elems]
+
+-- Should recover natural number and basic Pi with types (a/1)
+nat :: V a => Frac a () 
+nat = Frac Id
+
+oneF :: Frac () ()
+oneF = nat
+
+boolF :: Frac Bool ()
+boolF = nat
+
+boolFRecip :: Frac () Bool
+boolFRecip = Frac Id
+
+-- I can use `eval Id boolF' etc.
+test1 = eval Id boolF 
+
+--
 recip :: (V a, V b) => Frac a b -> Frac b a
 recip (Frac f) = Frac (SwapT :.: adjoint f :.: SwapT) 
 
@@ -162,16 +197,29 @@ times (Frac f) (Frac g) =
                   (Id :*: AssocRT) :.: -- (a,(b,(c,d)))
                   AssocLT -- ((a,b),(c,d))
 
+eta :: V a => Frac () () -> Frac a a
+eta (Frac f) = undefined
+
+epsilon :: V a => Frac a a -> Frac () ()
+epsilon (Frac f) = undefined
+
+{--
 plus :: (V a, V b, V c, V d) => 
-        Frac a b -> Frac c d -> Frac (Either a c) (Either b d) 
+        Frac a b -> Frac c d -> Frac (Either (a,d) (c,b)) (b,d)
[…]

### 2013-06-05 summer13/R.hs
more in R.hs mostly questions

@@ -171,6 +171,9 @@ boolF = nat
 boolFRecip :: Frac () Bool
 boolFRecip = Frac Id
 
+boolUnit :: Frac Bool Bool
+boolUnit = Frac (SwapT :: (Bool,()) :<=> ((),Bool))
+
 -- I can use `eval Id boolF' etc.
 test1 = eval Id boolF 
 

### 2013-06-05 summer13/R.hs
make existentials accessible

@@ -1,4 +1,4 @@
-{-# LANGUAGE TypeOperators, GADTs, RankNTypes #-}
+{-# LANGUAGE TypeOperators, GADTs, RankNTypes, NamedFieldPuns #-}
 
 module R where
 
@@ -149,18 +149,26 @@ evalR UnfoldB (Right ()) = False
 -- Pi-frac
 
 data Frac a b =
-  forall a' b'. (V a', V b') => Frac { iso :: (a,b') :<=> (a',b) }
+  forall a' b'. (V a', V b') => Frac { 
+               prepare :: a -> (a,b'), 
+               iso :: (a,b') :<=> (a',b), 
+               measure :: (a',b) -> b
+             }
 
 instance (V a, V b) => V (Frac a b) where
   elems = undefined
 
 instance (V a, V b, Show a, Show b) => Show (Frac a b) where
-  show (Frac f) = 
-      show [(a, snd (eval f (a,b'))) | a <- elems, b' <- elems]
+  show (Frac { prepare, iso, measure }) = 
+      show [(a, measure (eval iso (prepare a))) | a <- elems]
 
 -- Should recover natural number and basic Pi with types (a/1)
 nat :: V a => Frac a () 
-nat = Frac Id
+nat = Frac { 
+        prepare = \a -> (a,()),
+        iso = Id, 
+        measure = \(a,()) -> ()
+      }
 
 oneF :: Frac () ()
 oneF = nat
@@ -168,18 +176,29 @@ oneF = nat
 boolF :: Frac Bool ()
 boolF = nat
 
-boolFRecip :: Frac () Bool
-boolFRecip = Frac Id
+boolFRecip :: Bool -> Frac () Bool
+boolFRecip b = Frac {
+                 prepare = \() -> ((),b),
+                 iso = Id,
+                 measure = \((),b) -> b
+               }
 
 boolUnit :: Frac Bool Bool
-boolUnit = Frac (SwapT :: (Bool,()) :<=> ((),Bool))
+boolUnit = Frac {
+             prepare = \b -> (b,()),
+             iso = SwapT,
+             measure = \((),b) -> b
+           }
 
 -- I can use `eval Id boolF' etc.
 test1 = eval Id boolF 
 
 --
+
+{--
 recip :: (V a, V b) => Frac a b -> Frac b a
-recip (Frac f) = Frac (SwapT :.: adjoint f :.: SwapT) 
+recip (Frac { prepare, iso, measure }) = 
[…]

### 2013-06-13 summer13/frac2.tex
nothing really
added blurb about trace

@@
 
+\section{ {{Pi}} with trace}
+
+In order to take a step toward being able to write the Int construction in
+{{Pi}} to give it fractional types, we want to add a trace combinator to it,
+with the following type:
+
+%subcode{proof} include main
+%@  c : b1 * b <-> b2 * b
+%@@ trace c : b1 <-> b2
+
+This kind of combinator allows us to cancel out a common factor. When the type
+{{0}} is removed from the language, it is always possible to do this, but adding
+computational content becomes tricky. For example, consider permutations on the
+finite set of size {{6}}. If we wish to trace one of these to get a permutation
+on the set of size {{3}}, we are able to, but which set of size {{3}}? There
+are several different ways to do this, but there is no obvious canonical answer.
+
+More concretely, suppose we have a combinator {{c : 3 * 2 <-> 3 * 2}}. The
+combinator {{trace c : 3 <-> 3}} can be constructed in two obvious ways: let
+{{(trace c) v}} run {{c}} on {{(v, true)}} or on {{(v, false)}}. It is possible
+to choose {{c}} so that depending on which default value of type {{2}} is used,
+a different combinator results.
+
+To fix this, we parameterize {{trace}} with a value {{v : b}}, as follows:
+
+%subcode{proof} include main
+%@  c : b1 * b <-> b2 * b
+%@  v : b
+%@@ trace_v c : b1 <-> b2
+
+It remains to show that this is ``well-behaved''; that is, we want it to obey
+some (ideally all) of the trace properties, or at the very least some subset of
+them that is sufficient to prove some necessary properties for {{Pi}} extended
+with fractional types, as presented in the next few sections.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %subcode{bnf} include main
-% value types, r ::= b / b
-% values, f ::= (v |--> v amp c)
+% value types, r ::= b1 / b2
+% values, f ::= (v1 |--> v2, c)
 %

### 2013-06-13 summer13/frac2.tex
added alternatives

@@
 
+\subsection{Alternatives}
+
+If this formulation of trace does not work, we have a few alternatives:
+
+\begin{itemize}
+
+\item Don't bother choosing a default element at all. We did this previously
+with the relational semantics, without much success, but perhaps there's a
+better way of looking at it.
+
+\item Choose a canonical representative of the type {{b}} that's being traced
+out. This seems like it's a strictly worse option than parameterizing it, but
+it's still on the table, at least for now.
+
+\item Focus on the combinator. Jacques pointed out that our old version of trace
+``worked'' whenever the part of {{c}} acting on {{b}} was equivalent to the
+identity. This seems promising, but I'm not quite sure how to think about it
+yet, so I'm ignoring it for now. It's definitely something we want to look at,
+though, especialy if the current approach doesn't work.
+
+\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-10-13 f.tex
started a new monster paper

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{color, colortbl}
\usepackage{courier}

\begin{document}
\begin{titlepage}
\begin{center}
\vfill
{\LARGE Isomorphisms as Groupoids: \\ 
  Homotopy Type Theory with Fractional Types} \\[1.5cm]
{\Large Jacques Carette${}^*$ \qquad Amr Sabry${}^{\dagger}$}\\[0.5cm]
{\Large $({}^*)$~McMaster University \qquad $({}^\dagger)$~Indiana University}
\vfill
\end{center}
\vfill
\noindent\textbf{Abstract.}

\vfill
\vfill
\end{titlepage}

\title{Isomorphisms as Groupoids: \\ 
  Homotopy Type Theory with Fractional Types}
\author{Jacques Carette \and Amr Sabry}
\date{\today}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

HoTT cannot really deal with functions. The starting point is that paths
explain and witness identities. This works for all the usual types but not
for functions or universes. For functions, the HoTT treatment is to allow
arbitrary functions, single out the isos via classical extensional methods,
and then assert via an axiom that the class of singled out functions is
equivalent to paths. This is a convoluted way. Why not start with functions
that are, by construction, isos. This was the programme we started earlier,
inspired by physical considerations. Because physics requires various
conservation principles (including conservation of information) and because
computation is fundamentally a physical process, the argument (originally due
to Feynman etc.) was that computation should be based on reversible
processes, or in abstract terms, on isomorphisms. That fits very well with
the HoTT philosophy that makes isos first-class. Except that HoTT does not
deal exclusively with isos, the functions are generally not isos. So let's
investigate the radical idea that all functions are isos in the HoTT
framework. What we get is a notion of first-class functions that by
construction are isos and obey the usual path rules etc. The technical device
is the notion of fractional types which are dual to products. A function
(iso) $A \rightarrow B$ is represented as an element of the type
$(\frac{1}{A} \times B)$. The construction shows the essential nature of the
$\infty$-groupoid structure. The more higher-order functions we have the
higher we have to go in the groupoid structure. We also have something
significant in the realm of reversible computing. We finally have
higher-order functions. Of course we need to show that we have constructed a
monoidal closed category (a concrete one) and then we can inherit all the
fancy constructions. Finally we have a very interesting operational
interpretation of paths and we outline various applications.

Need a clear explanation of why we can't deal with the usual functions that
are not isomorphisms. Regular function spaces obey $1^A = 1$ for all $A$. In
terms of fractionals this gives $\frac{1}{A} = 1$ for all $A$ which is not
true. However if we restrict to isomorphisms then the only functions we can
write are $A \rightarrow B$ where $A$ and $B$ are isomorphic types and hence
the only functions we can write in the space $1^A$ are to types $A$ that are
isomorphic to 1 and in that case we do indeed get that $\frac{1}{A} =
1$. See~\cite{fiore-remarks} for more details. The benefit is that we fit
squarely in the physically motivated notion that isomorphisms are the
foundational computational mechanism as well as the HoTT approach that is
essentially all about isomorphisms. Of course we don't lose anything by
restricting our attention to isomorphisms because we know that general
(irreversible) functions can be obtained from isomorphisms via
\emph{information effects}~\cite{James:2012:IE:2103656.2103667}.

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme has sparked
a significant amount of research that unveiled new and surprising connections
between geometry, algebra, logic, and computation (see~\cite{baez2011physics}
for an overview of some of the connections).

In the physics community, Landauer~\cite{Landauer:1961,Landauer},
Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
the laws of physics as fundamentally related to computation. The great
majority of these laws are formulated as equalities between different
physical observables which is unsatisfying: \emph{different} physical
observables should not be related by an \emph{equality}. It is more
appropriate to relate them by an \emph{isomorphism} that witnesses, explains,
and models the process of transforming one observable to the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive
mathematics~\cite{citeulike:7374951}. This theory has been further extended
with \emph{identity types} representing the proposition that two terms are
``equal.'' (See~\cite{streicher,warren} for a survey.) Briefly speaking,
given two terms $a$ and $b$ of the same type $A$, one forms the type
$\texttt{Id}_A(a,b)$ representing the proposition that~$a$ and~$b$ are equal:
in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
models the process of transforming $a$ to $b$ and vice-versa.

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function~\cite{Rittri:1989:UTS:99370.99384}. More recently, type isomorphisms
have taken a more central role as \emph{the} fundamental computational
mechanism from which more conventional, i.e., irreversible computation, is
derived. In our own previous
work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
notion of type isomorphism and developed from it a family of programming
languages, $\Pi$ with various superscripts, in which computation is an
isomorphism preserving the information-theoretic entropy.

A major open problem remains, however: a higher-order extension of
$\Pi$. This extension is of fundamental importance in all the originating
research areas. In physics, it allows for quantum states to be viewed as
processes and processes to be viewed as states, such as with the
Choi-Jamiolkowski
isomorphism~\cite{choi1975completely,jamiolkowski1972linear}.  In mathematics
and logic, it allows the equivalence between different proofs of type
$\texttt{Id}_A(a,b)$ to itself be expressed as an isomorphism (of a higher
type) $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code.

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{closed category} from the underlying monoidal
category for~$\Pi$. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle.  Our main novel technical device to achieving the higher-order
extension is a \emph{fractional type} which represents \emph{negative
information} and which is so named because of its duality with conventional
product types.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Intuitive Ideas and Examples} 
\label{sec:intuition}

Talk about groupoid cardinality and give an example of how the cardinality of
a group of order $n$ is $\frac{1}{n}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finite Types as Discrete (0-) Groupoids}

A review of pi and slight reformulation of the semantics to use the groupoid
terminology. In more detail, motivate and introduce pointed types here, add
an element $\bot_A$ to every type $A$ and express the groupoids using a
pointed type and an explicit collection of discrete paths (not indexed by
anything at this point).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types as Groupoids}

In the previous section, every type was modeled by a collection of values
(including a special ``unusable'' value $\bot$) and a collection of trivial
paths. In this section, non-trivial paths will emerge to model ``fractional
values.'' The intuition is that two values that that are connected by a path
should not be counted twice as the path induces an equivalence between these
two values. As illustrated in Sec.~\ref{sec:intuition}, this idea generalizes
to conclude that a value connected to iself using several paths corresponds
to a ``fraction.''

We begin by investigating the special situation of a type $\frac{1}{A}$ where
$A$ is a 0-type (i.e., where $A$ does not contain any fractional components).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Rank-1 Isomorphisms as 1-Groupoids}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Rank-$n$ Isomorphisms as $n$-Groupoids}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Operational Semantics via Backtracking Constraints}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Type Theory}

\begin{quote} 
As usual, something funny happens at the left of the arrow. John Reynolds,
Aarhus, June 27, 2000.~\cite{danvypage} 
\end{quote}

No more! We finally have an inductive characterization of functions
(isomorphisms). No negative occurrences of types. No positivity
checks. Nothing funny.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Semantics}

Now that functions are a proper inductive type, we can use conventional
induction to reason about functions. There are also probably connections to
logical relations and parametricity that are essentially some kind of
isomorphism which might be expressible as a higher-order path.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Logic}

Similar to double negation proof of $A \vee \neg A$. Make a guess and if
wrong backtrack and choose the other branch.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Programming Practice}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Resource Interpretation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Category Theory}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Homotopy Theory}

The connection between type theory and homotopy theory raises some
interesting questions. What do the various shapes (spheres, tori, etc.) have
to go with programming? Our answer is that they represent a class
of \emph{functions}: isomorphisms.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Quantum Computing}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Connections to Physics, Information Theory, and Reversible Computing}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Future Work}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgments}

This material is based upon work supported by the National Science Foundation
under Grant No. 1217454. We would like to specially thank Roshan James and
Zachary Sparks for their contributions to early stages of this work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plain}
\bibliography{cites}
\end{document}

### 2013-10-13 f.tex
updated f.tex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Finite Types as Discrete (0-) Groupoids}
+\section{Finite Types as Discrete (0-)Groupoids}
 
@@
 paths. In this section, non-trivial paths will emerge to model ``fractional
-values.'' The intuition is that two values that that are connected by a path
+values.'' The intuition is that two values that are connected by a path
 should not be counted twice as the path induces an equivalence between these
-two values. As illustrated in Sec.~\ref{sec:intuition}, this idea generalizes
-to conclude that a value connected to iself using several paths corresponds
-to a ``fraction.''
+two values: ``they are the same.'' As illustrated in
+Sec.~\ref{sec:intuition}, this idea generalizes to conclude that a value
+connected to itself using several paths corresponds to a ``fraction.''
 
@@
 
+There are also deep connections with algebraic types, the seven trees in one,
+and the Fiore papers.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+Our construction can be interpreted in the context of speculative concurrent
+programming. A thread can at any time create, out of nothing, a value of type
+$A$ and proceed with it, generating in the process a value of type
+$\frac{1}{A}$ that a concurrent process can be manipulate. When an actual
+value of type $A$ is produced, the fractional value is used to coordinate a
+backtracking mechanism that communicates the actual value of type $A$ to the
+original process. This is very similar to how we use logical variables and
+unification.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Connections to Reversible Computing}
+
+We have higher-order reversible computations.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Connections to Category Theory}
+
+We have constructed a concrete closed monoidal category. We did not use the
+Int-construction which requires an existing trace operator. We inherit a lot
+of constructions for free: name, coname, trace, etc.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Connections to Category Theory}
+We can interpret $\times$ as parallel computations (i.e., as computations
+occurring simultaneously in different regions in space). So a value of type
+$A \times B$ consists of two values occupying two different space locations,
+e.g., two different memory locations. With that perspective, a value of type
+$\frac{1}{A}$ consists of a small dictionary that stores all the values of
+type $A$ and can shuffle them around.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Connections to Quantum Computing}
+\section{Connections to Physics and Quantum Computing}
+
+The dual vector space in quantum mechanics and its role in quantum
+computing. The fact that $\eta$ and $\epsilon$ are dual is essentially the
+teleportation algorithm. We can give it a computational interpretation in
+terms of backtracking.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Connections to Physics, Information Theory, and Reversible Computing}
+\section{Connections to Information Theory}
+
+We have given classic computational interpretations to negative
+information~\cite{negative-information} and negative
+entropy~\cite{negative-entropy}.
 
@@
 \section{Conclusions and Future Work}
+
+This is cool and deep.
 

### 2013-10-13 f.tex
small updates to f.tex

@@
 \usepackage{courier}
+\usepackage{graphicx}
 
@@
 
+Because we are \emph{not} in the context of the full $\lambda$-calculus,
+which allows arbitrary duplication and erasure of information, values of
+fractional types are first-class values that can flow anywhere. The
+information-preserving computational infrastructure guarantees that, in a
+complete program, every constraint imposed by a fractional value will also be
+satisfied exactly once. This property is shared with systems that are based
+on linear logic.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We have higher-order reversible computations.
+
+Consider a similar algebraic manipulation involving fractional types.
+\begin{center}
+\scalebox{1.1}{
+\includegraphics{diagrams/thesis/algebra-wire2.pdf}
+}
+\end{center}
 
@@
 
+One understanding of quantum computing is that it exploits the laws of physics
+to build faster machines (perhaps). Another more foundational understanding
+is that it provides a computational interpretation of physics, and in
+particular directly addresses the question of interpretation of quantum
+mechanics. In a little known document, Rozas~\cite{Rozas:1987:CMO:889539}
+uses continuations to implement the transactional interpretation of quantum
+mechanics~\cite{transactional} which includes as its main ingredient a
+fixpoint calculation between waves or particles traveling forwards and
+backwards in time. Our work sheds no light on whether this interpretation is
+the ``right one'' but it is interesting that we can directly realize it using
+the primitives of our language.
+
+The multiplicative structure has a direct connection to entangled quantum
+particles, or perhaps entangled particles and anti-particles.  The idea of
+entanglement, that an action on one particle is ``instantaneously''
+communicated to the other, is analogous to how unifying one value affects its
+dual pair which is possibly in another part of the computation.  Again our
+model sheds no light on whether this is related to how nature computes but it
+is again interesting that we can directly realize the idea.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Speculative Connections}
+
+Negative types, square root types, and imaginary types have also appeared in
+the literature: we proceed with a simple explanation. We have so far extended
+the set of types to be the positive rational numbers. Now we will push this
+and extend the set of types to algebraic numbers. In other words, we will
+allow datatypes defined by arbitrary polynomials and allow the roots of such
+polynomials to be types.
+
+Consider first an example in which we want to compute with the sides of a
+rectangle whose area is 91 and whose length is longer than its width by 6
+units. One can solve the quadratic equation to determine that the sides are 7
+and 13 and proceed. This however prematuraly forces us to globally solve the
+constraint. Instead we can let the two sides of the rectangle be $x$ and
+$x+6$ and use the following equation to capture the desired constraint:
+\[
+x^2 + 6x - 91 = 0
+\]
+The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
+the type $0$. We can now proceed to compute with the unknown $x$, being
+assured that in a closed program, our computation will eventually be
+consistent with the solution of the algebraic equation.
+
+Previously the most famous example of a similar nature is the puzzling
+isomorphism that one can establish between seven binary trees and one.
+A binary tree is defined by the datatype
+\[
+x = 1 + x * x 
+\]
+which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
+algebraic manipulation we can reason as follows:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+Fiore poses the question of why such algebraic manipulation would make sense
+type theoretically but states that even though some of the intermediate steps
+make no sense, the final equivalence is valid and can be used to actually
+construct an isomorphism between $x^7$ the type of seven binary trees 
+and $x$ the type of binary trees.
+
+Discussion of possible polynomials:
+\begin{itemize}
+\item $b=1+1$. Boring.
+\item $b=b+1$. That introduces the natural numbers. No solution for this
+  polynomial over the algebraic numbers. We must extend the numbers with
+  $\omega$ to get a solution. We reject this in this paper and prefer to
+  stick to algebraic numbers. The advantage is that all the isomorphisms are
+  valid numerically. With the above type we could subtract $b$ from both
+  sides to show that $0=1$ which is nonsense. We can however have infinite
+  types as long as they have algebraic solutions.
+\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
+  boolean! If we had superpositions, we could then write a function that
+  performs the square root of negation in the sense that applying it twice
+  would be equivalent to boolean negation.
+\end{itemize}
+
+\paragraph*{Geometry of Interaction (GoI).}
+Geometry of Interaction was developed by Girard~\cite{girard1989geometry} as
+part of the development of linear logic. It was given a computational
+interpretation by Abramsky and
+Jagadeesan~\cite{Abramsky:1994:NFG:184662.184664}, and was developed into a
+reversible model of computing by
+Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. Preliminary investigations
+suggest that many of the GoI machine constructions can be simulated by
+treating Mackie's bi-directional wires as pairs of wires and replacing the
+machine's global state with a typed value on the wire that captures the
+appropriate state. This connection is exciting because when viewed through a
+Curry-Howard lens it suggests that the logical interpretation of would be a
+linear-like logic with a notion of resource preservation and with a natural
+computational interpretation.
+
+\paragraph*{Computing in the Field of Algebraic Numbers.}
+\label{sec:algebraic-field}
+Algebraically, the addition of fractional types corresponds to a move from a
+ring-like structure to a full field. Our language captures the structure of
+one particular field: that of the rational numbers. As we have seen,
+computation in this field is quite expressive and interesting and yet, it has
+two fundamental limitations. First it cannot express any recursive type, and
+second it cannot express any datatype definitions. We believe these to be two
+orthogonal extensions: recursive types were considered in our previous
+paper~\cite{James:2012:IE:2103656.2103667}; arbitrary datatypes are however
+even more exciting that plain rationals as each datatype definition can be
+viewed as a polynomial (see below) which essentially means that we start
+computing in the field of algebraic numbers, which includes square roots and
+imaginary numbers. As crazy as it might seem, the type $\sqrt{2}$ and even
+the type $(1/2)+i(\sqrt{3}/2)$ ``make sense.''  In fact the latter type is
+the solution to the polynomial $x^2-x+1=0$ which if re-arranged looks like
+$x=1+x*x$ and perhaps more familiarly as the datatype of binary trees $\mu
+x.(1+x*x)$. These types happen to have been studied extensively following a
+paper by Blass~\cite{seventrees} which used the above datatype of trees to
+infer an isomorphism between seven binary trees and one!
+
+We have confirmed that we can extend our language with the datatype
+declaration for binary trees and build a witness for this isomorphism that
+works as expected. However not every isomorphism constructed from algebraic
+manipulation is computationally meaningful. To understand the issue in more
+detail, consider the following algebraically valid proof of the isomorphism
+in question:
+\[\begin{array}{rclclclcl}
+x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
+x^6 &=& 1 \\
+x^7 &=& x^6 x &=& x
+\end{array}\]
+The question is why such an algebraic manipulation makes sense type
+theoretically, even though the intermediate step asking for an isomorphism
+between $x^6$ and $1$ has no computational context. In our setting, this
+isomorphism can be constructed but it diverges on all inputs (in both
+ways). This suggests that, in the field of algebraic numbers, some algebraic
+manipulations are somehow ``more constructive'' than others.
+
+A related issue is that not all meaningful recursive types are meaningful
+polynomials. For instance $\textsf{nat}=\mu x.(1+x)$ implies the polynomial
+$x=1+x$ which has no algebraic solutions without appeal to more complex
+structures with limits etc.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions and Future Work}
@@
 This is cool and deep.
+
+\paragraph*{Declarative Continuations.} 
+In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
+proposes that continuations are a \emph{declarative} concept. He,
+furthermore, introduces a symmetric extension of the $\lambda$-calculus in
+which call-by-value is dual to call-by-name and values are dual to
+continuations. In more detail, the symmetric calculus contains a ``value''
+fragment and a ``continuation'' fragment which are mirror images. Pairs and
+sums are treated as duals in the sense that the ``value'' fragment includes
+pairs whose mirror image in the ``continuation'' fragment are sums. In
+contrast, our language includes pairs and sums in the value fragment and two
+symmetries: one that maps the pairs to fractions and another that maps the
+sums to subtractions.
+
+\paragraph*{The Duality of Computation.}
+The duality between call-by-name and call-by-value was further investigated
+by Selinger using control
+categories~\cite{Selinger:2001:CCD:966910.966911}. Curien and
+Herbelin~\cite{Curien:2000} also introduce a calculus that exhibits
+symmetries between values and continuations and between call-by-value and
+call-by-name. The calculus includes the type $A-B$ which is the dual of
+implication, i.e., a value of type $A-B$ is a context expecting a function of
+type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
+as a \emph{pair} consisting of a value of type $A$ and a continuation of type
+$B$. This is to be contrasted with our interpretation of a value of that type
+as \emph{either} a value of $A$ or a demand for a value of type $B$. This
+calculus was further analyzed and extended by
+Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
+interpretation to the subtraction connective and like the original symmetric
+calculus of Filinski, introduces a duality that relates sums to products and
+vice-versa.
+
+\paragraph*{Subtractive Logic.} 
+Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
+logic which contains a dual to implication. Her work has been distilled in
+the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
+related to coroutines~\cite{Crolard01082004} and delimited
+continuations~\cite{Ariola:2009:TFD:1743339.1743381}.  In more detail,
+Crolard explains the type $A-B$ as the type of \emph{coroutines} with a local
+environment of type $A$ and a continuation of type $B$. The description is
+complicated by what is essentially the desire to enforce linearity
+constraints so that coroutines cannot access the local environment of other
+coroutines.
+
+\paragraph*{Negation in Classical Linear Logic} 
+Filinski~\cite{Filinski92} uses the negative types of linear logic to model
+continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
+negative types of linear logic as \emph{acceptors}, which are like
+continuations in the sense that they take an input and return no
+output. Acceptors however are also similar in flavor to logic variables:
+they can be created and instantiated later once their context of use is
+determined. Although a formal connection is lacking, it is clear that, at an
+intuitive level, acceptors are entities that combine elements of our negative
+and fractional types.
+
+\paragraph*{The Lambek-Grishin Calculus.} The ``parsing-as-deduction'' style
+of linguistic analysis uses the Lambek-Grishin calculus with the following
+types: product, left division, right division, sum, right difference, and
+left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division and
+difference types are similar to our types but because the calculus lacks
+commutativity and associativity and only has limited notions of
+distributivity, each connective needs a left and right version. The
+Lambek-Grishin exhibits two notions of symmetry but they are unrelated to our
+notions. In particular, the first notion of symmetry expresses commutativity
+and the second relates products to sums and divisions to subtractions. In
+contrast, our two symmetries relate sums to subtractions and products to
+divisions.
 

### 2013-10-14 f.tex
0 types

@@
 \usepackage{url}
+\usepackage{stmaryrd}
 \usepackage{color, colortbl}
@@
 \usepackage{graphicx}
+
+\newcommand{\alt}{~|~}
+\newcommand{\leftv}[1]{\textsf{left}~#1}
+\newcommand{\rightv}[1]{\textsf{right}~#1}
+\newcommand{\iso}{\leftrightarrow}
+\newcommand{\identlp}{\mathit{identl}_+}
+\newcommand{\identrp}{\mathit{identr}_+}
+\newcommand{\swapp}{\mathit{swap}_+}
+\newcommand{\assoclp}{\mathit{assocl}_+}
+\newcommand{\assocrp}{\mathit{assocr}_+}
+\newcommand{\identlt}{\mathit{identl}_*}
+\newcommand{\identrt}{\mathit{identr}_*}
+\newcommand{\swapt}{\mathit{swap}_*}
+\newcommand{\assoclt}{\mathit{assocl}_*}
+\newcommand{\assocrt}{\mathit{assocr}_*}
+\newcommand{\distz}{\mathit{dist}_0}
+\newcommand{\factorz}{\mathit{factor}_0}
+\newcommand{\dist}{\mathit{dist}}
+\newcommand{\factor}{\mathit{factor}}
+\newcommand{\idc}{\mathit{id}}
+\newcommand{\symc}[1]{\mathit{sym}~#1}
+\newcommand{\proves}{\vdash}
+\newcommand{\Rule}[4]{
+\makebox{{\rm #1}
+$\displaystyle
+\frac{\begin{array}{l}#2\\\end{array}}
+{\begin{array}{l}#3\\\end{array}}$
+ #4}}
+\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
 
@@
 
-A review of pi and slight reformulation of the semantics to use the groupoid
-terminology. In more detail, motivate and introduce pointed types here, add
+We review our language $\Pi$ providing the necessary background and context
+for our higher-order extension.\footnote{The presentation in this section
+  focuses on the simplest version of $\Pi$. Other versions include recursive
+  types and trace operators but these extensions are orthogonal to the
+  higher-order extension emphasized in this document.} Even though the
+semantics of $\Pi$ could be explained using plain sets, we express it using
+the groupoid terminology to faciliate the transition to fractionals and
+higher-order functions.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Syntax and Types}
+
+The terms of $\Pi$ are not classical values and functions; rather, the terms
+are isomorphism witnesses.  In other words, the terms of $\Pi$ are proofs
+that certain ``shapes of values'' are isomorphic.  And, in classical
+Curry-Howard fashion, our operational semantics shows how these proofs can be
+directly interpreted as actions on ordinary values which effect this shape
+transformation. Of course, ``shapes of values'' are very familiar already:
+they are usually called \emph{types}.  But frequently one designs a type
+system as a method of classifying terms, with the eventual purpose to show
+that certain properties of well-typed terms hold, such as safety.  Our
+approach is different: we start from a type system, and then present a term
+language which naturally inhabits these types, along with an appropriate
+operational and denotational semantics.
+
+\paragraph*{Data.}
+We view $\Pi$ as having two levels: it has traditional values classified by
+ordinary types:
+\[\begin{array}{lrcl} 
+\textit{values}, v &::=& () \alt \leftv{v} \alt \rightv{v} \alt (v,v) \\
+\textit{value types}, b &::=& 0 \alt 1 \alt b+b \alt b * b
+\end{array}\]
+
+Types include the empty type $0$, the unit type $1$, sum types $b_1+b_2$, and
+product types $b_1*b_2$.  Values include $()$ which is the only value of type
+$1$, $\leftv{v}$ and $\rightv{v}$ which inject $v$ into a sum type, and
+$(v_1,v_2)$ which builds a value of product type.
+
+\paragraph*{Isomorphisms.} The terms of $\Pi$ witness
+type isomorphisms of the form $b \iso b$. They consist of base isomorphisms,
+as defined below, and their composition.
+\[\begin{array}{rrcll}
+\identlp :&  0 + b & \iso & b &: \identrp \\
+\swapp :&  b_1 + b_2 & \iso & b_2 + b_1 &: \swapp \\
+\assoclp :&  b_1 + (b_2 + b_3) & \iso & (b_1 + b_2) + b_3 &: \assocrp \\
+\identlt :&  1 * b & \iso & b &: \identrt \\
+\swapt :&  b_1 * b_2 & \iso & b_2 * b_1 &: \swapt \\
+\assoclt :&  b_1 * (b_2 * b_3) & \iso & (b_1 * b_2) * b_3 &: \assocrt \\
+\distz :&~ 0 * b & \iso & 0 &: \factorz \\
+\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor
+\end{array}\]
+
+Each line of the above table introduces a pair of dual
+constants\footnote{where $\swapp$ and $swapt$ are self-dual.} that witness
+the type isomorphism in the middle.  These are the base (non-reducible) terms
+of the second, principal level of $\Pi$. Note how the above has two readings:
+first as a set of typing relations for a set of constants. Second, if these
+axioms are seen as universally quantified, orientable statements, they also
+induce transformations of the (traditional) values. The (categorical or
+homotopical) intuition here is that these axioms have computational content
+because they witness isomorphisms rather than merely stating an extensional
+equality.
+
+The isomorphisms are extended to form a congruence relation by adding the
+following constructors that witness equivalence and compatible closure:
+
+\begin{table}[t]
+\hrule
+\smallskip
+\begin{center} 
+\Rule{}
+{}
+{\jdg{}{}{\idc : b \iso b}}
+{}
+\quad 
+\Rule{}
+{\jdg{}{}{c : b_1 \iso b_2}}
+{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
+{}
+\quad
+\Rule{}
+{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
+{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
+{}
+\medskip 
+\Rule{}
+{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
+{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
+{}
+\quad 
+\Rule{}
+{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
+{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
+{}
+\end{center}
+\caption{Combinators\label{pi-combinators}}
+\smallskip
+\hrule
+\end{table}
+
+It is important to note that ``values'' and ``isomorphisms'' are completely
+separate syntactic categories which do not intermix. The semantics of the
+language come when these are made to interact at the ``top level'' via
+\emph{application}: 
+\[\begin{array}{lrcl}
+\textit{top level term}, l &::=& c v
+\end{array}\]
+
+The language presented above, at the type level, models a \emph{semiring}
+(occasionally called a \emph{rig}) where we replace equality by isomorphism.
+Semantically, $\Pi$ models a \emph{bimonoidal category} whose simplest
+example is the category of finite sets and bijections. In that
+interpretation, each value type denotes a finite set of a size calculated by
+viewing the types as natural numbers and each combinator $c : b_1 \iso b_2$
+denotes a bijection between the sets denoted by $b_1$ and $b_2$.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Operational Semantics}
+
+Operationally, the semantics of $\Pi$ is given using two mutually recursive
+interpreters: one going forward and one going backwards. The use of $\symc{}$
+switches control from one evaluator to the other.  We state without proof
+that the evaluation of well-typed combinators always terminates and that
+$\Pi$ is logically reversible, i.e., that for all combinators $c : b_1 \iso
+b_2$ and values $v_1 : b_1$ and $v_2 : b_2$ we have the forward evaluation of
+$c v_1$ produces $v_2$ iff the backwards evaluation of $c v_2$ produces
+$v_1$.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Groupoid Semantics}
+
+In more detail, motivate and introduce pointed types here, add
 an element $\bot_A$ to every type $A$ and express the groupoids using a

### 2013-10-14 f.tex
f.tex with lots of interesting questions about pointed spaces

@@
 \usepackage{stmaryrd}
+\usepackage{amssymb}
 \usepackage{color, colortbl}
@@
 
+\newcommand{\amr}[1]{\fbox{Amr says:} \textbf{#1}}
+
+\newcommand{\union}{\cup}
+\newtheorem{definition}{Definition}
+\newcommand{\refl}{\mathit{refl}}
+\renewcommand{\path}{\leadsto}
+\renewcommand{\circ}{\circlearrowleft}
 \newcommand{\alt}{~|~}
+\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
 \newcommand{\leftv}[1]{\textsf{left}~#1}
@@
 \newcommand{\idc}{\mathit{id}}
+\newcommand{\wedgesum}{\vee}
+\newcommand{\smashproduct}{\wedge}
 \newcommand{\symc}[1]{\mathit{sym}~#1}
@@
 semantics of $\Pi$ could be explained using plain sets, we express it using
-the groupoid terminology to faciliate the transition to fractionals and
+the groupoid terminology to facilitate the transition to fractionals and
 higher-order functions.
@@
 ordinary types:
-\[\begin{array}{lrcl} 
+\[\begin{array}{rcl} 
 \textit{values}, v &::=& () \alt \leftv{v} \alt \rightv{v} \alt (v,v) \\
@@
 \begin{table}[t]
-\hrule
-\smallskip
+\hrule\medskip
 \begin{center} 
@@
 {}
-\medskip 
+\\ \bigskip
 \Rule{}
@@
 \caption{Combinators\label{pi-combinators}}
-\smallskip
-\hrule
+\medskip\hrule
 \end{table}
@@
 
-The language presented above, at the type level, models a \emph{semiring}
-(occasionally called a \emph{rig}) where we replace equality by isomorphism.
-Semantically, $\Pi$ models a \emph{bimonoidal category} whose simplest
-example is the category of finite sets and bijections. In that
-interpretation, each value type denotes a finite set of a size calculated by
-viewing the types as natural numbers and each combinator $c : b_1 \iso b_2$
-denotes a bijection between the sets denoted by $b_1$ and $b_2$.
-
-%%%%%%%%%%%%%%%%%
-\subsection{Operational Semantics}
-
-Operationally, the semantics of $\Pi$ is given using two mutually recursive
-interpreters: one going forward and one going backwards. The use of $\symc{}$
-switches control from one evaluator to the other.  We state without proof
-that the evaluation of well-typed combinators always terminates and that
-$\Pi$ is logically reversible, i.e., that for all combinators $c : b_1 \iso
-b_2$ and values $v_1 : b_1$ and $v_2 : b_2$ we have the forward evaluation of
-$c v_1$ produces $v_2$ iff the backwards evaluation of $c v_2$ produces
-$v_1$.
-
 %%%%%%%%%%%%%%%%%
@@
 
-In more detail, motivate and introduce pointed types here, add
-an element $\bot_A$ to every type $A$ and express the groupoids using a
-pointed type and an explicit collection of discrete paths (not indexed by
-anything at this point).
+The language presented above, at the type level, models a \emph{commutative
+  semiring} (occasionally called a \emph{commutative rig}) where we replace
+equality by isomorphism.  Semantically, $\Pi$ models a \emph{bimonoidal
+  category} whose simplest example is the category of finite sets and
+bijections. In that interpretation, each value type denotes a finite set of a
+size calculated by viewing the types as natural numbers and each combinator
+$c : b_1 \iso b_2$ denotes a bijection between the sets denoted by $b_1$ and
+$b_2$. Instead of presenting such a semantics, we present a slightly more
+involved semantics using discrete groupoids. In that semantics, the
+denotation of each type will consist of (i) a pointed set, and (ii) a trivial
+equivalence relation relating each element to itself by a \emph{path}. The
+pointed sets are constructed by adjoining, to each ``regular'' set, an
+element $\circ$ which topologically corresponds to the
+\emph{circle}. Significantly, using the Euler characteristic as a generalized
+notion of cardinality, a set containing $n$ ``regular'' elements and a circle
+$\circ$ has cardinality $n$ and not $n+1$.\footnote{Generally speaking, this
+  generalized cardinality based on the Euler characteristic counts the number
+  of ``connected components'' of a set. Consider a set with $n$ points and
+  various paths connecting them. We start by counting all the points giving
+  us cardinality $n$. Then for every path, we decrease the cardinality by one
+  to avoid double-counting connected elements. Then we add one again for
+  every path that forms a loop because the path forming the loop did not
+  decrease the number of connected components. It is easy to see that a point
+  with a path that connects it to itself would have generalized cardinality
+  0. Moreover the generalized cardinality of the disjoint union of two sets
+  is their sum and the generalized cardinality of the product of two sets is
+  their product.}
+
+\begin{definition}[Denotation of types]
+Each type $b$ is mapped to a groupoid $(\dt{b}, \circ_b, \mathit{paths}_b)$
+where $\dt{b}$ is a pointed set with distinguished point $\circ_b$ and
+$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$, between elements
+$v_1$ and $v_2$ of $\dt{b}$. The map is defined as follows:
+\[\begin{array}{rcl@{\qquad}l}
+\dt{0} &=& (\{ \circ_0 \}, \circ_0, \{ \circ_0 \path \circ_0 \}) &
+          (\mbox{circle}) \\
+\dt{1} &=& (\{ \bullet, \circ_1 \}, \circ_1, 
+           \{ \circ_1 \path \circ_1, \bullet \path \bullet \}) &
+          (\mbox{0-sphere})\\
+\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
+          (\mbox{wedge~sum}) \\
+\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
+          (\mbox{smash~product})
+\end{array}\]
+The empty type is mapped to the topological circle which is a point that is
+connected to itself. The unit type is mapped to the topological 0-sphere
+which is a pair of points with the discrete topology. The operation
+$\wedgesum$ is the \emph{wedge sum} which is a regular disjoint union with a
+new distinguished point $\circ_{b_1+b_2}$ and additional paths consisting of
+the equivalence closure of the identifications $\circ_{b_1} \path
+\circ_{b_1+b_2}$ and $\circ_{b_2} \path \circ_{b_1+b_2}$. The operation
+$\smashproduct$ is the \emph{smash product} which is a regular product with a
+new distinguished point $\circ_{b_1*b_2}$ and additional paths consisting of
+the equivalence closure of the identifications $(x,\circ_{b_2}) \path
+\circ_{b_1*b_2}$ and $(\circ_{b_1},y) \path \circ_{b_1*b_2}$ for all $x \in
+\dt{b_1}$ and $y \in \dt{b_2}$.
+\end{definition}
+
+We quickly check that the desired isomorphisms hold at the level of types,
+e.g., the isomorphism $b * 0 \iso 0$ requires that the smash product of any
+groupoid with the circle be indistinguishable from the circle.
+
+The semantics of $\Pi$ is given using two mutually recursive interpreters:
+one going forward and one going backwards. The use of $\symc{}$ switches
+control from one evaluator to the other.
+
+\[\begin{array}{rcl}
+\mathit{eval} &:& (a \iso b) \rightarrow \dt{a} \rightarrow \dt{b} \\
+\mathit{eval}~\idc~a &=& a \\
+\mathit{eval}~(\symc{f})~b &=& \mathit{evalR}~f~b \\
+\mathit{eval}~(f \fatsemi g)~a &=& \mathit{eval}~g~(\mathit{eval}~f~a) \\
+\mathit{eval}~(f \otimes g)~(a,b) &=& (\mathit{eval}~f~a, \mathit{eval}~g~b) \\
+\mathit{eval}~(f \oplus g)~(\leftv{a}) &=& \leftv{(\mathit{eval}~f~a)} \\
+\mathit{eval}~(f \oplus g)~(\rightv{b}) &=& \rightv{(\mathit{eval}~g~b)} \\
+\mathit{eval}~\identlp~(\leftv{\circ}) &=& \circ \\
+\mathit{eval}~\identlp~(\rightv{a}) &=& a \\
+\mathit{eval}~\identrp~a &=& \rightv{a} \\
+\mathit{eval}~\swapp~(\leftv{a}) &=& \rightv{a} \\
+\mathit{eval}~\swapp~(\rightv{b}) &=& \leftv{b} \\
+\mathit{eval}~\assoclp~(\leftv{a}) &=& \leftv{(\leftv{a})} \\
+\mathit{eval}~\assoclp~(\rightv{(\leftv{b})}) &=& \leftv{(\rightv{b})} \\
+\mathit{eval}~\assoclp~(\rightv{(\rightv{c})}) &=& \rightv{c} \\
+\mathit{eval}~\assocrp~(\leftv{(\leftv{a})}) &=& \leftv{a} \\
+\mathit{eval}~\assocrp~(\leftv{(\rightv{b})}) &=& \rightv{(\leftv{b})} \\
+\mathit{eval}~\assocrp~(\rightv{c}) &=& \rightv{(\rightv{c})} \\
+\mathit{eval}~\identlt~((), a) &=& a \\
+\mathit{eval}~\identrt~a &=& ((), a) \\
+\mathit{eval}~\swapt~(a,b) &=& (b,a) \\
+\mathit{eval}~\assoclt~(a,(b,c)) &=& ((a,b),c) \\
+\mathit{eval}~\assocrt~((a,b),c)  &=& (a,(b,c)) \\
+\mathit{eval}~\distz~(\rightv{c}, a) &=& \rightv{(c, a)} \\
+\mathit{eval}~\factorz~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
+\mathit{eval}~\dist~(\leftv{b}, a) &=& \leftv{(b, a)} \\
+\mathit{eval}~\dist~(\rightv{c}, a) &=& \rightv{(c, a)} \\
+\mathit{eval}~\factor~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
+\mathit{eval}~\factor~(\rightv{(c, a)}) &=& (\rightv{c}, a) \\
+\end{array}\]
+\[\begin{array}{rcl}
+\mathit{evalR} &:& (a \iso ) \rightarrow b \rightarrow a \\
+\mathit{evalR}~\idc~a &=& a \\
+\mathit{evalR}~(\symc{f}) b &=& \mathit{eval}~f b \\
+\mathit{evalR}~(f \fatsemi g) a &=& \mathit{evalR}~f (\mathit{evalR}~g a) \\
+\mathit{evalR}~(f \otimes g) (a,b) &=& 
+  (\mathit{evalR}~f a, \mathit{evalR}~g b) \\
+\mathit{evalR}~(f \oplus g) (\leftv{a}) &=& \leftv{(\mathit{evalR}~f a)} \\
+\mathit{evalR}~(f \oplus g) (\rightv{b}) &=& \rightv{(\mathit{evalR}~g b)} \\
+\mathit{evalR}~\identlp~a &=& \rightv{a} \\
+\mathit{evalR}~\identrp~(\rightv{a}) &=& a \\
+\mathit{evalR}~\swapp~(\leftv{a}) &=& \rightv{a} \\
+\mathit{evalR}~\swapp~(\rightv{b}) &=& \leftv{b} \\ 
+\mathit{evalR}~\assoclp~(\leftv{(\leftv{a})}) &=& \leftv{a} \\
+\mathit{evalR}~\assoclp~(\leftv{(\rightv{b})}) &=& \rightv{(\leftv{b})} \\
+\mathit{evalR}~\assoclp~(\rightv{c}) &=& \rightv{(\rightv{c})} \\
+\mathit{evalR}~\assocrp~(\leftv{a}) &=& \leftv{(\leftv{a})} \\
+\mathit{evalR}~\assocrp~(\rightv{(\leftv{b})}) &=& \leftv{(\rightv{b})} \\
+\mathit{evalR}~\assocrp~(\rightv{(\rightv{c})}) &=& \rightv{c} \\
+\mathit{evalR}~\identlt~a &=& ((), a) \\
+\mathit{evalR}~\identrt~((), a) &=& a \\
+\mathit{evalR}~\swapt~(a,b) &=& (b,a) \\
+\mathit{evalR}~\assoclt~((a,b),c)  &=& (a,(b,c)) \\
+\mathit{evalR}~\assocrt~(a,(b,c)) &=& ((a,b),c) \\
+\mathit{eval}~\distz~(\rightv{c}, a) &=& \rightv{(c, a)} \\
+\mathit{eval}~\factorz~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
+\mathit{evalR}~\dist~(\leftv{(b, a)}) &=& (\leftv{b}, a) \\
+\mathit{evalR}~\dist~(\rightv{(c, a)}) &=& (\rightv{c}, a) \\
+\mathit{evalR}~\factor~(\leftv{b}, a) &=& \leftv{(b, a)} \\
+\mathit{evalR}~\factor~(\rightv{c}, a) &=& \rightv{(c, a)}
+\end{array}\]
+
+We state without proof that the evaluation of well-typed combinators always
+terminates and that $\Pi$ is logically reversible, i.e., that for all
+combinators $c : b_1 \iso b_2$ and values $v_1 : b_1$ and $v_2 : b_2$ we have
+the forward evaluation of $c v_1$ produces $v_2$ iff the backwards evaluation
+of $c v_2$ produces $v_1$.
 

### 2013-10-15 f.tex
denotations of types

@@
 Each line of the above table introduces a pair of dual
-constants\footnote{where $\swapp$ and $swapt$ are self-dual.} that witness
+constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that witness
 the type isomorphism in the middle.  These are the base (non-reducible) terms
@@
 \[\begin{array}{lrcl}
-\textit{top level term}, l &::=& c v
+\textit{top level term}, l &::=& c~v
 \end{array}\]
@@
 pointed sets are constructed by adjoining, to each ``regular'' set, an
-element $\circ$ which topologically corresponds to the
-\emph{circle}. Significantly, using the Euler characteristic as a generalized
-notion of cardinality, a set containing $n$ ``regular'' elements and a circle
-$\circ$ has cardinality $n$ and not $n+1$.\footnote{Generally speaking, this
-  generalized cardinality based on the Euler characteristic counts the number
-  of ``connected components'' of a set. Consider a set with $n$ points and
+element $\bot$ with a non-trivial path to itself which, topologically,
+corresponds to the \emph{circle}. Significantly, using the Euler
+characteristic as a generalized notion of cardinality, a set containing $n$
+discrete elements (with only trivial reflexivity paths) and a circle has
+cardinality $n$ and not $n+1$.\footnote{Generally speaking, this generalized
+  cardinality based on the Euler characteristic counts the number of
+  ``connected components'' of a set. Consider a set with $n$ points and
   various paths connecting them. We start by counting all the points giving
@@
 \begin{definition}[Denotation of types]
-Each type $b$ is mapped to a groupoid $(\dt{b}, \circ_b, \mathit{paths}_b)$
-where $\dt{b}$ is a pointed set with distinguished point $\circ_b$ and
-$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$, between elements
-$v_1$ and $v_2$ of $\dt{b}$. The map is defined as follows:
+Each type $b$ is mapped to a groupoid $(\dt{b}, \bot_b, \mathit{paths}_b)$
+where~$\dt{b}$ is a pointed set with distinguished point $\bot_b$ and
+$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$,
+between elements $v_1$ and $v_2$ of $\dt{b}$. (We omit the trivial paths for
+reflexivity and only include the non-trivial ``generating'' paths as
+suggested by the notation of \emph{higher-order inductive types}.)
 \[\begin{array}{rcl@{\qquad}l}
-\dt{0} &=& (\{ \circ_0 \}, \circ_0, \{ \circ_0 \path \circ_0 \}) &
+\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \bot_0 \path \bot_0 \}) &
            (\mbox{circle}) \\
-\dt{1} &=& (\{ \bullet, \circ_1 \}, \circ_1, 
-           \{ \circ_1 \path \circ_1, \bullet \path \bullet \}) &
-          (\mbox{0-sphere})\\
+\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, \{ \bot_1 \path \bot_1 \}) &\\
 \dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
@@
 The empty type is mapped to the topological circle which is a point that is
-connected to itself. The unit type is mapped to the topological 0-sphere
-which is a pair of points with the discrete topology. The operation
-$\wedgesum$ is the \emph{wedge sum} which is a regular disjoint union with a
-new distinguished point $\circ_{b_1+b_2}$ and additional paths consisting of
-the equivalence closure of the identifications $\circ_{b_1} \path
-\circ_{b_1+b_2}$ and $\circ_{b_2} \path \circ_{b_1+b_2}$. The operation
-$\smashproduct$ is the \emph{smash product} which is a regular product with a
-new distinguished point $\circ_{b_1*b_2}$ and additional paths consisting of
-the equivalence closure of the identifications $(x,\circ_{b_2}) \path
-\circ_{b_1*b_2}$ and $(\circ_{b_1},y) \path \circ_{b_1*b_2}$ for all $x \in
-\dt{b_1}$ and $y \in \dt{b_2}$.
+connected to itself. The unit type is mapped to a pointed singleton with a
+distinguished element connected to itself with a non-trivial loop. The
+operation $\wedgesum$ is the \emph{wedge sum} which corresponds to the
+disjoint union with an additional path $\leftv{\bot_{b_1}} \path
+\rightv{\bot_{b_2}}$ identifying the two distinguished elements of the
+component spaces. The new cluster is the distinguished element of the
+composite space. The operation $\smashproduct$ is the \emph{smash product}
+which is the regular set product with additional paths connecting
+$(x,\bot_{b_2})$ and $(\bot_{b_1},y)$ for all $x \in \dt{b_1}$ and $y \in
+\dt{b_2}$. The new cluster of identified distinguished elements is the new
+distinguished element of the composite space.
 \end{definition}

### 2013-10-16 f.tex
groupoids

@@
 \newcommand{\union}{\cup}
+\newcommand{\transport}[2]{\overline{#1}(#2)}
 \newtheorem{definition}{Definition}
@@
 \renewcommand{\path}{\leadsto}
-\renewcommand{\circ}{\circlearrowleft}
+%% \renewcommand{\circ}{\circlearrowleft}
 \newcommand{\alt}{~|~}
@@
 %%%%%%%%%%%%%%%%%
-\subsection{Groupoid Semantics}
+\subsection{Types as Groupoids}
 
@@
 $c : b_1 \iso b_2$ denotes a bijection between the sets denoted by $b_1$ and
-$b_2$. Instead of presenting such a semantics, we present a slightly more
-involved semantics using discrete groupoids. In that semantics, the
-denotation of each type will consist of (i) a pointed set, and (ii) a trivial
-equivalence relation relating each element to itself by a \emph{path}. The
-pointed sets are constructed by adjoining, to each ``regular'' set, an
-element $\bot$ with a non-trivial path to itself which, topologically,
-corresponds to the \emph{circle}. Significantly, using the Euler
-characteristic as a generalized notion of cardinality, a set containing $n$
-discrete elements (with only trivial reflexivity paths) and a circle has
-cardinality $n$ and not $n+1$.\footnote{Generally speaking, this generalized
-  cardinality based on the Euler characteristic counts the number of
-  ``connected components'' of a set. Consider a set with $n$ points and
-  various paths connecting them. We start by counting all the points giving
-  us cardinality $n$. Then for every path, we decrease the cardinality by one
-  to avoid double-counting connected elements. Then we add one again for
-  every path that forms a loop because the path forming the loop did not
-  decrease the number of connected components. It is easy to see that a point
-  with a path that connects it to itself would have generalized cardinality
-  0. Moreover the generalized cardinality of the disjoint union of two sets
-  is their sum and the generalized cardinality of the product of two sets is
-  their product.}
+$b_2$. Instead of presenting such a semantics, we present a more involved
+semantics using groupoids.
+
+\begin{definition}[Path]
+A \emph{path} $p : x \path y$ between two points $x$ and $y$ in some space
+$A$ is a directed edge $p$ from $x$ to $y$. The notation $\transport{p}{x}$
+denotes the action of \emph{transporting} the point $x$ to the point~$y$
+along $p$.
+\end{definition}
+
+\begin{definition}[Disjoint union of paths]
+Given two sets of paths $P$ and $Q$, the disjoint union $P \uplus^p Q$ is
+defined as:
+\[ 
+  \{ \leftv{p} : \leftv{x} \path \leftv{y} ~|~ (p : x \path y) \in P \}
+  \union
+  \{ \rightv{q} : \leftv{x} \path \leftv{y} ~|~ (q : x \path y) \in Q \}
+\]
+\end{definition}
+We will sometimes write this as $\leftv{P} \union \rightv{Q}$.
+
+\begin{definition}[Product of paths]
+Given two sets of paths $P$ and $Q$, the product $P \times^p Q$ is
+defined as:
+\[ 
+\{ (p,q) : (x,z) \path (y,w) ~|~ (p : x \path y) \in P, (q : z \path w) \in Q \}
+\]
+\end{definition}
+We will sometimes write this as $(P,Q)$.
+
+\begin{definition}[Groupoid]
+A \emph{groupoid} $G = (G_0, \mathit{paths}_{G_0})$ is a set $G_0$ together
+with a collection of paths $x \path y$, between elements $x$ and $y$ of $G_0$
+that satisfies the following conditions:
+\begin{itemize}
+\item for every $x \in G_0$, there is a path $\refl_x : x \path x$,
+\item every path $p : x \path y$, there is an inverse path $! p : y \path x$, 
+\item for every paths $p : x \path y$ and $q : y \path z$ there is a path $p
+  \circ q : x \path z$, and
+\item for every paths $p : x \path y$, $q : y \path z$, and $r : z \path w$,
+  we have:
+  \begin{itemize}
+  \item $\refl_x \circ p = p$;
+  \item $p \circ \refl_y = p$;
+  \item $(p \circ q) \circ r = (p \circ (q \circ r))$;
+  \item $p \,\circ\, !p = \refl_x$;
+  \item $!p \,\circ\, p = \refl_y$.
+  \end{itemize}
+\end{itemize}
+\end{definition}
+When introducing a groupoid, we only explicitly mention a collection of
+non-trivial ``generating'' paths. The equivalence closure under $\refl$, $!$,
+and $\circ$ with the appropriate conditions is implicitly
+assumed.\footnote{For higher groupoids, the five equivalences in the last
+  condition will be witnessed by paths between paths and not by the
+  extensional equivalence relation $=$.}
+
+\begin{definition}[Pointed groupoid]
+A pointed groupoid $G_{\bullet} = (G_0, \bot_{G_0}, \mathit{paths}_{G_0})$ is
+a groupoid $(G_0, \mathit{paths}_{G_0})$ with a distinguished element
+$\bot_{G_0} \in G_0$.
+\end{definition}
+
+\begin{definition}[Wedge sum]
+The \emph{wedge sum} of two pointed groupoids $G_\bullet$ and $H_\bullet$,
+written $G_\bullet \wedgesum H_\bullet$ is defined as follows:
+\[\begin{array}{l}
+(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
+(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
+((G_0 \uplus H_0) \union \{ \bot \},
+ \bot,
+ (\mathit{paths}_{G_0} \uplus^p \mathit{paths}_{H_0}) \union
+ \{ L: \leftv{\bot_{G_0}} \path \bot, 
+    R: \rightv{\bot_{H_0}} \path \bot \})
+\end{array}\]
+where $\uplus$ is the disjoint union operation on sets and $\bot$ is a new
+distinguished element that is related by new paths to the distinguished
+elements of $G_\bullet$ and $H_\bullet$.
+\end{definition}
+
+\begin{definition}[Smash product]
+The \emph{smash product} of two pointed groupoids $G_\bullet$ and
+$H_\bullet$, written $G_\bullet \smashproduct H_\bullet$ is defined as
+follows:
+\[\begin{array}{l}
+(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \smashproduct
+(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
+((G_0 \times H_0) \union \{ \bot \},
+ \bot,
+ (\mathit{paths}_{G_0} \times^p \mathit{paths}_{H_0}) \union
+ \{ F_h: (\bot_{G_0},h) \path \bot \} \union 
+ \{ S_g: (g,\bot_{H_0}) \path \bot \} 
+\end{array}\]
+where $\times$ is the cartesian product operation on sets and $\bot$ is a new
+distinguished element that is related by new paths to every pair of elements
+that includes either $\bot_{G_0}$ or $\bot_{H_0}$. 
+\end{definition}
 
 \begin{definition}[Denotation of types]
-Each type $b$ is mapped to a groupoid $(\dt{b}, \bot_b, \mathit{paths}_b)$
-where~$\dt{b}$ is a pointed set with distinguished point $\bot_b$ and
-$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$,
-between elements $v_1$ and $v_2$ of $\dt{b}$. (We omit the trivial paths for
-reflexivity and only include the non-trivial ``generating'' paths as
-suggested by the notation of \emph{higher-order inductive types}.)
+Each type $b$ is mapped to a \emph{pointed groupoid} $(\dt{b}, \bot_b,
+\mathit{paths}_b)$ as follows:
 \[\begin{array}{rcl@{\qquad}l}
-\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \bot_0 \path \bot_0 \}) &
+\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) &
            (\mbox{circle}) \\
-\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, \{ \bot_1 \path \bot_1 \}) &\\
+\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, 
+           \{ \mathit{loop}_1 : \bot_1 \path \bot_1 \}) &\\
 \dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
@@
 \end{array}\]
-The empty type is mapped to the topological circle which is a point that is
-connected to itself. The unit type is mapped to a pointed singleton with a
-distinguished element connected to itself with a non-trivial loop. The
-operation $\wedgesum$ is the \emph{wedge sum} which corresponds to the
-disjoint union with an additional path $\leftv{\bot_{b_1}} \path
-\rightv{\bot_{b_2}}$ identifying the two distinguished elements of the
-component spaces. The new cluster is the distinguished element of the
-composite space. The operation $\smashproduct$ is the \emph{smash product}
-which is the regular set product with additional paths connecting
-$(x,\bot_{b_2})$ and $(\bot_{b_1},y)$ for all $x \in \dt{b_1}$ and $y \in
-\dt{b_2}$. The new cluster of identified distinguished elements is the new
-distinguished element of the composite space.
 \end{definition}
 
-We quickly check that the desired isomorphisms hold at the level of types,
-e.g., the isomorphism $b * 0 \iso 0$ requires that the smash product of any
-groupoid with the circle be indistinguishable from the circle.
+Topologically $\dt{0}$ corresponds to the circle $S^1$ which has Euler
+characteristic (which generalizes the notion of set cardinality) of 0. As a
+groupoid, the space $\dt{0}$ includes this infinite collection of paths from
+$\bot_0 \path \bot_0$:
+\[\begin{array}{l}
+\refl_{\bot_0} \\
+\mathit{loop}_0 \\
+!~\mathit{loop}_0 \\
+\mathit{loop}_0 \circ \mathit{loop}_0 \\
+!~(\mathit{loop}_0 \circ \mathit{loop}_0) \\
+\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0 \\
+!~(\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0) \\
+\vdots
+\end{array}\]
+and hence, using the definition introduced by Baez and
+Dolan~\cite{groupoidcard}, has groupoid cardinality 0. We will abbreviate
+this collection of paths as $\mathit{loop}^n_0$ for integer $n$, where
+$\mathit{loop}^0_0$ is $\refl$, and $\mathit{loop}^n_0$ abbreviates the
+$n$-fold composition in one direction for positive $n$ and in the opposite
+direction for negative $n$.
+
+Note that all our pointed spaces have a similar structure: a collection of
+``regular elements'' forming several connected components and a separate
+isolated connected component that includes the distinguished element
+$\bot$. The distinguished element $\bot$ has the infinite collection of loops
+on it and, in general, may be connected to the distinguished elements of the
+subspaces used in the construction. That entire cluster including $\bot$ has
+Euler characteristic and groupoid cardinality 0.
+
+We now check that the wedge sum of any of our pointed groupoids $G_\bullet$
+with $\dt{0}$ is ``essentially'' $G_\bullet$ itself, i.e., that $\dt{0}$ is
+the additive unit in our space of pointed groupoids. Consider:
+\[\begin{array}{l}
+G_\bullet \wedgesum \dt{0} = \\
+(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
+  (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) = \\
+((\leftv{G_0} \union \{ \rightv{\bot_0}, \bot \}, 
+ \bot,
+ (\leftv{\mathit{paths}_{G_0}} \union \{ \rightv{\mathit{loop}^n_0} , 
+ \leftv{\bot_{G_0}} \path \bot, \rightv{\bot_0} \path \bot \})
+\end{array}\]
+One can see that this space consists of $\leftv{G_0}$ with whatever structure
+is on it and another connected component consisting of $\leftv{\bot_{G_0}}$,
+$\rightv{\bot_0}$, and the new $\bot$ for the entire space. This cluster has
+cardinality 0 and hence the new space is ``equivalent'' to the original
+$G\bullet$ as we have essentially only changed the representation of 0.
+
+Therefore, since the wedge sum is evidently commutative and associative, it
+is at least plausible that the structure $(\wedgesum, \dt{0}$ is a symmetric
+monoidal category. All of that will be formalized later. The categorical
+morphisms must obviously map distinguished elements to distinguished elements
+and preserve the structure of the path. It is also reasonably easy to see
+that the structure $(\smashproduct, \dt{1})$ is another symmetric monoidal
+category. Even better one can check that for any pointed groupoid
+$G_\bullet$, we have $G_\bullet \smashproduct \dt{0}$ gives us a very
+complicated 0 but still a 0. I didn't check distributivity but I suspect it
+holds too.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Groupoid Semantics}
 
@@
 
+We have constructed a good candidate for pointed homotopy category.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 of \emph{functions}: isomorphisms.
+
+We have constructed a good candidate for pointed homotopy category.
 

### 2013-10-16 f.tex
paths

@@
 \usepackage{amssymb}
+\usepackage[sans]{dsfont}
 \usepackage{color, colortbl}
@@
 \renewcommand{\path}{\leadsto}
+\newcommand{\paths}[1]{\mathds{P}(#1)}
 %% \renewcommand{\circ}{\circlearrowleft}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Finite Types as Discrete (0-)Groupoids}
+\section{The Category of Pointed Groupoids}
+
+%%%%%%%%%%%%%%%%%
+\subsection{Paths}
+
+\begin{definition}[Path]
+A \emph{path} $p : x \path y$ between two points $x$ and $y$ in some space
+$A$ is a directed edge $p$ from $x$ to $y$. We denote the set of paths of the
+space $A$ as $\paths{A}$.
+\end{definition}
+
+\begin{definition}[Disjoint union of paths]
+Given $\paths{A}$ and $\paths{B}$ in spaces $A$ and $B$ respectively, we
+construct $\paths{A} \uplus^p \paths{B}$ in space $A \uplus B$, the disjoint
+union of $A$ and $B$, as follows:
+\[ 
+  \{ \leftv{p} : \leftv{x} \path \leftv{y} ~|~ 
+    (p : x \path y) \in \paths{A} \}
+  \union
+  \{ \rightv{q} : \rightv{x} \path \rightv{y} ~|~ 
+    (q : x \path y) \in \paths{B} \}
+\]
+\end{definition}
+We will sometimes write this as $\leftv{(\paths{A})} \union
+\rightv{(\paths{B})}$.
+
+\begin{definition}[Product of paths]
+Given $\paths{A}$ and $\paths{B}$ in spaces $A$ and $B$
+respectively, we construct $\paths{A} \times^p \paths{B}$ in
+space $A \times B$, the cartesian product of $A$ and $B$, as follows: 
+\[ 
+\{ (p,q) : (x,z) \path (y,w) ~|~ (p : x \path y) \in \paths{A}, 
+   (q : z \path w) \in \paths{B} \}
+\]
+\end{definition}
+We will sometimes write this as $(\paths{A},\paths{B})$.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Pointed Groupoids}
+
+\begin{definition}[Groupoid]
+A \emph{groupoid} $G = (G_0, \paths{G_0})$ is a set $G_0$ together
+with a collection of paths $x \path y$, between elements $x$ and $y$ of $G_0$
+that satisfies the following conditions:
+\begin{itemize}
+\item for every $x \in G_0$, there is a path $\refl_x : x \path x$,
+\item every path $p : x \path y$, there is an inverse path $! p : y \path x$, 
+\item for every paths $p : x \path y$ and $q : y \path z$ there is a path $p
+  \circ q : x \path z$, and
+\item for every paths $p : x \path y$, $q : y \path z$, and $r : z \path w$,
+  we have:
+  \begin{itemize}
+  \item $\refl_x \circ p = p$;
+  \item $p \circ \refl_y = p$;
+  \item $(p \circ q) \circ r = (p \circ (q \circ r))$;
+  \item $p \,\circ\, !p = \refl_x$;
+  \item $!p \,\circ\, p = \refl_y$.
+  \end{itemize}
+\end{itemize}
+\end{definition}
+When introducing a groupoid, we only explicitly mention a collection of
+non-trivial ``generating'' paths. The equivalence closure under $\refl$, $!$,
+and $\circ$ with the appropriate conditions is implicitly
+assumed.\footnote{For higher groupoids, the five equivalences in the last
+  condition will be witnessed by paths between paths and not by the
+  extensional equivalence relation $=$.}
+
+\begin{definition}[Pointed groupoid]
+A pointed groupoid $G_{\bullet} = (G_0, \bot_{G_0}, \paths{G_0})$ is
+a groupoid $(G_0, \paths{G_0})$ with a distinguished element
+$\bot_{G_0} \in G_0$.
+\end{definition}
+
+\begin{definition}[Wedge sum]
+The \emph{wedge sum} of two pointed groupoids $G_\bullet$ and $H_\bullet$,
+written $G_\bullet \wedgesum H_\bullet$ is defined as follows:
+\[\begin{array}{l}
+(G_0, \bot_{G_0}, \paths{G_0}) \wedgesum 
+(H_0, \bot_{H_0}, \paths{H_0}) = \\
+((G_0 \uplus H_0) \union \{ \bot \},
+ \bot,
+ (\paths{G_0} \uplus^p \paths{H_0}) \union
+ \{ L: \leftv{\bot_{G_0}} \path \bot, 
+    R: \rightv{\bot_{H_0}} \path \bot \})
+\end{array}\]
+where $\uplus$ is the disjoint union operation on sets and $\bot$ is a new
+distinguished element that is related by new paths to the distinguished
+elements of $G_\bullet$ and $H_\bullet$.
+\end{definition}
+
+\begin{definition}[Smash product]
+The \emph{smash product} of two pointed groupoids $G_\bullet$ and
+$H_\bullet$, written $G_\bullet \smashproduct H_\bullet$ is defined as
+follows:
+\[\begin{array}{l}
+(G_0, \bot_{G_0}, \paths{G_0}) \smashproduct
+(H_0, \bot_{H_0}, \paths{H_0}) = \\
+((G_0 \times H_0) \union \{ \bot \},
+ \bot,
+ (\paths{G_0} \times^p \paths{H_0}) \union
+ \{ F_h: (\bot_{G_0},h) \path \bot \} \union 
+ \{ S_g: (g,\bot_{H_0}) \path \bot \} 
+\end{array}\]
+where $\times$ is the cartesian product operation on sets and $\bot$ is a new
+distinguished element that is related by new paths to every pair of elements
+that includes either $\bot_{G_0}$ or $\bot_{H_0}$. 
+\end{definition}
+
+\begin{definition}[Denotation of types]
+Each type $b$ is mapped to a \emph{pointed groupoid} $(\dt{b}, \bot_b,
+\paths{b})$ as follows:
+\[\begin{array}{rcl@{\qquad}l}
+\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) &
+           (\mbox{circle}) \\
+\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, 
+           \{ \mathit{loop}_1 : \bot_1 \path \bot_1 \}) &\\
+\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
+           (\mbox{wedge~sum}) \\
+\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
+           (\mbox{smash~product})
+\end{array}\]
+\end{definition}
+
+Topologically $\dt{0}$ corresponds to the circle $S^1$ which has Euler
+characteristic (which generalizes the notion of set cardinality) of 0. As a
+groupoid, the space $\dt{0}$ includes this infinite collection of paths from
+$\bot_0 \path \bot_0$:
+\[\begin{array}{l}
+\refl_{\bot_0} \\
+\mathit{loop}_0 \\
+!~\mathit{loop}_0 \\
+\mathit{loop}_0 \circ \mathit{loop}_0 \\
+!~(\mathit{loop}_0 \circ \mathit{loop}_0) \\
+\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0 \\
+!~(\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0) \\
+\vdots
+\end{array}\]
+and hence, using the definition introduced by Baez and
+Dolan~\cite{groupoidcard}, has groupoid cardinality 0. We will abbreviate
+this collection of paths as $\mathit{loop}^n_0$ for integer $n$, where
+$\mathit{loop}^0_0$ is $\refl$, and $\mathit{loop}^n_0$ abbreviates the
+$n$-fold composition in one direction for positive $n$ and in the opposite
+direction for negative $n$.
+
+Note that all our pointed spaces have a similar structure: a collection of
+``regular elements'' forming several connected components and a separate
+isolated connected component that includes the distinguished element
+$\bot$. The distinguished element $\bot$ has the infinite collection of loops
+on it and, in general, may be connected to the distinguished elements of the
+subspaces used in the construction. That entire cluster including $\bot$ has
+Euler characteristic and groupoid cardinality 0.
+
+We now check that the wedge sum of any of our pointed groupoids $G_\bullet$
+with $\dt{0}$ is ``essentially'' $G_\bullet$ itself, i.e., that $\dt{0}$ is
+the additive unit in our space of pointed groupoids. Consider:
+\[\begin{array}{l}
+G_\bullet \wedgesum \dt{0} = \\
+(G_0, \bot_{G_0}, \paths{G_0}) \wedgesum 
+  (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) = \\
+((\leftv{G_0} \union \{ \rightv{\bot_0}, \bot \}, 
+ \bot,
+ (\leftv{\paths{G_0}} \union \{ \rightv{\mathit{loop}^n_0} , 
+ \leftv{\bot_{G_0}} \path \bot, \rightv{\bot_0} \path \bot \})
+\end{array}\]
+One can see that this space consists of $\leftv{G_0}$ with whatever structure
+is on it and another connected component consisting of $\leftv{\bot_{G_0}}$,
+$\rightv{\bot_0}$, and the new $\bot$ for the entire space. This cluster has
+cardinality 0 and hence the new space is ``equivalent'' to the original
+$G\bullet$ as we have essentially only changed the representation of 0.
+
+Therefore, since the wedge sum is evidently commutative and associative, it
+is at least plausible that the structure $(\wedgesum, \dt{0}$ is a symmetric
+monoidal category. All of that will be formalized later. The categorical
+morphisms must obviously map distinguished elements to distinguished elements
+and preserve the structure of the path. It is also reasonably easy to see
+that the structure $(\smashproduct, \dt{1})$ is another symmetric monoidal
+category. Even better one can check that for any pointed groupoid
+$G_\bullet$, we have $G_\bullet \smashproduct \dt{0}$ gives us a very
+complicated 0 but still a 0. I didn't check distributivity but I suspect it
+holds too.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Finite Types as Pointed Groupoids}
 
@@
 %%%%%%%%%%%%%%%%%
-\subsection{Types as Groupoids}
+\subsection{Groupoid Semantics}
 
@@
 semantics using groupoids.
-
-\begin{definition}[Path]
-A \emph{path} $p : x \path y$ between two points $x$ and $y$ in some space
-$A$ is a directed edge $p$ from $x$ to $y$. The notation $\transport{p}{x}$
-denotes the action of \emph{transporting} the point $x$ to the point~$y$
-along $p$.
-\end{definition}
-
-\begin{definition}[Disjoint union of paths]
-Given two sets of paths $P$ and $Q$, the disjoint union $P \uplus^p Q$ is
-defined as:
-\[ 
-  \{ \leftv{p} : \leftv{x} \path \leftv{y} ~|~ (p : x \path y) \in P \}
-  \union
-  \{ \rightv{q} : \leftv{x} \path \leftv{y} ~|~ (q : x \path y) \in Q \}
-\]
-\end{definition}
-We will sometimes write this as $\leftv{P} \union \rightv{Q}$.
-
-\begin{definition}[Product of paths]
-Given two sets of paths $P$ and $Q$, the product $P \times^p Q$ is
-defined as:
-\[ 
-\{ (p,q) : (x,z) \path (y,w) ~|~ (p : x \path y) \in P, (q : z \path w) \in Q \}
-\]
-\end{definition}
-We will sometimes write this as $(P,Q)$.
-
-\begin{definition}[Groupoid]
-A \emph{groupoid} $G = (G_0, \mathit{paths}_{G_0})$ is a set $G_0$ together
-with a collection of paths $x \path y$, between elements $x$ and $y$ of $G_0$
-that satisfies the following conditions:
-\begin{itemize}
-\item for every $x \in G_0$, there is a path $\refl_x : x \path x$,
-\item every path $p : x \path y$, there is an inverse path $! p : y \path x$, 
-\item for every paths $p : x \path y$ and $q : y \path z$ there is a path $p
-  \circ q : x \path z$, and
-\item for every paths $p : x \path y$, $q : y \path z$, and $r : z \path w$,
-  we have:
-  \begin{itemize}
-  \item $\refl_x \circ p = p$;
-  \item $p \circ \refl_y = p$;
-  \item $(p \circ q) \circ r = (p \circ (q \circ r))$;
-  \item $p \,\circ\, !p = \refl_x$;
-  \item $!p \,\circ\, p = \refl_y$.
-  \end{itemize}
-\end{itemize}
-\end{definition}
-When introducing a groupoid, we only explicitly mention a collection of
-non-trivial ``generating'' paths. The equivalence closure under $\refl$, $!$,
-and $\circ$ with the appropriate conditions is implicitly
-assumed.\footnote{For higher groupoids, the five equivalences in the last
-  condition will be witnessed by paths between paths and not by the
-  extensional equivalence relation $=$.}
-
-\begin{definition}[Pointed groupoid]
-A pointed groupoid $G_{\bullet} = (G_0, \bot_{G_0}, \mathit{paths}_{G_0})$ is
-a groupoid $(G_0, \mathit{paths}_{G_0})$ with a distinguished element
-$\bot_{G_0} \in G_0$.
-\end{definition}
-
-\begin{definition}[Wedge sum]
-The \emph{wedge sum} of two pointed groupoids $G_\bullet$ and $H_\bullet$,
-written $G_\bullet \wedgesum H_\bullet$ is defined as follows:
-\[\begin{array}{l}
-(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
-(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
-((G_0 \uplus H_0) \union \{ \bot \},
- \bot,
- (\mathit{paths}_{G_0} \uplus^p \mathit{paths}_{H_0}) \union
- \{ L: \leftv{\bot_{G_0}} \path \bot, 
-    R: \rightv{\bot_{H_0}} \path \bot \})
-\end{array}\]
-where $\uplus$ is the disjoint union operation on sets and $\bot$ is a new
-distinguished element that is related by new paths to the distinguished
-elements of $G_\bullet$ and $H_\bullet$.
-\end{definition}
-
-\begin{definition}[Smash product]
-The \emph{smash product} of two pointed groupoids $G_\bullet$ and
-$H_\bullet$, written $G_\bullet \smashproduct H_\bullet$ is defined as
-follows:
-\[\begin{array}{l}
-(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \smashproduct
-(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
-((G_0 \times H_0) \union \{ \bot \},
- \bot,
- (\mathit{paths}_{G_0} \times^p \mathit{paths}_{H_0}) \union
- \{ F_h: (\bot_{G_0},h) \path \bot \} \union 
- \{ S_g: (g,\bot_{H_0}) \path \bot \} 
-\end{array}\]
-where $\times$ is the cartesian product operation on sets and $\bot$ is a new
-distinguished element that is related by new paths to every pair of elements
-that includes either $\bot_{G_0}$ or $\bot_{H_0}$. 
-\end{definition}
-
-\begin{definition}[Denotation of types]
-Each type $b$ is mapped to a \emph{pointed groupoid} $(\dt{b}, \bot_b,
-\mathit{paths}_b)$ as follows:
-\[\begin{array}{rcl@{\qquad}l}
-\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) &
-           (\mbox{circle}) \\
-\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, 
-           \{ \mathit{loop}_1 : \bot_1 \path \bot_1 \}) &\\
-\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
-           (\mbox{wedge~sum}) \\
-\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
-           (\mbox{smash~product})
-\end{array}\]
-\end{definition}
-
-Topologically $\dt{0}$ corresponds to the circle $S^1$ which has Euler
-characteristic (which generalizes the notion of set cardinality) of 0. As a
-groupoid, the space $\dt{0}$ includes this infinite collection of paths from
-$\bot_0 \path \bot_0$:
-\[\begin{array}{l}
-\refl_{\bot_0} \\
-\mathit{loop}_0 \\
-!~\mathit{loop}_0 \\
-\mathit{loop}_0 \circ \mathit{loop}_0 \\
-!~(\mathit{loop}_0 \circ \mathit{loop}_0) \\
-\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0 \\
-!~(\mathit{loop}_0 \circ \mathit{loop}_0 \circ \mathit{loop}_0) \\
-\vdots
-\end{array}\]
-and hence, using the definition introduced by Baez and
-Dolan~\cite{groupoidcard}, has groupoid cardinality 0. We will abbreviate
-this collection of paths as $\mathit{loop}^n_0$ for integer $n$, where
-$\mathit{loop}^0_0$ is $\refl$, and $\mathit{loop}^n_0$ abbreviates the
-$n$-fold composition in one direction for positive $n$ and in the opposite
-direction for negative $n$.
-
-Note that all our pointed spaces have a similar structure: a collection of
-``regular elements'' forming several connected components and a separate
-isolated connected component that includes the distinguished element
-$\bot$. The distinguished element $\bot$ has the infinite collection of loops
-on it and, in general, may be connected to the distinguished elements of the
-subspaces used in the construction. That entire cluster including $\bot$ has
-Euler characteristic and groupoid cardinality 0.
-
-We now check that the wedge sum of any of our pointed groupoids $G_\bullet$
-with $\dt{0}$ is ``essentially'' $G_\bullet$ itself, i.e., that $\dt{0}$ is
-the additive unit in our space of pointed groupoids. Consider:
-\[\begin{array}{l}
-G_\bullet \wedgesum \dt{0} = \\
-(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
-  (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) = \\
-((\leftv{G_0} \union \{ \rightv{\bot_0}, \bot \}, 
- \bot,
- (\leftv{\mathit{paths}_{G_0}} \union \{ \rightv{\mathit{loop}^n_0} , 
- \leftv{\bot_{G_0}} \path \bot, \rightv{\bot_0} \path \bot \})
-\end{array}\]
-One can see that this space consists of $\leftv{G_0}$ with whatever structure
-is on it and another connected component consisting of $\leftv{\bot_{G_0}}$,
-$\rightv{\bot_0}$, and the new $\bot$ for the entire space. This cluster has
-cardinality 0 and hence the new space is ``equivalent'' to the original
-$G\bullet$ as we have essentially only changed the representation of 0.
-
-Therefore, since the wedge sum is evidently commutative and associative, it
-is at least plausible that the structure $(\wedgesum, \dt{0}$ is a symmetric
-monoidal category. All of that will be formalized later. The categorical
-morphisms must obviously map distinguished elements to distinguished elements
-and preserve the structure of the path. It is also reasonably easy to see
-that the structure $(\smashproduct, \dt{1})$ is another symmetric monoidal
-category. Even better one can check that for any pointed groupoid
-$G_\bullet$, we have $G_\bullet \smashproduct \dt{0}$ gives us a very
-complicated 0 but still a 0. I didn't check distributivity but I suspect it
-holds too.
-
-%%%%%%%%%%%%%%%%%
-\subsection{Groupoid Semantics}
 

### 2013-10-19 f.tex
f.tex

@@
 \usepackage{graphicx}
+\usepackage[all]{xy}
+\xyoption{line}
 
@@
 
+% ----------------------------------------------------------------------
+% XY-Pic definitions for making ``wire charts''
+
+\def\wirechart#1#2{\let\labelstyle\objectstyle\xy*[*1.0]\xybox{\small%
+\xymatrix@C=10mm@R=4mm#1{#2}}\endxy}
+\def\wire#1#2{\ar@{-}[#1]^<>(.5){#2}}
+\def\wireright#1#2{\wire{#1}{#2}\ar@{}[#1]|<>(.5)/.6ex/{\dir2{>}}}
+\def\wireleft#1#2{\wire{#1}{#2}\ar@{}[#1]|<>(.5)/-.6ex/{\dir2{<}}}
+\def\wwblank#1{*=<#1,0mm>{~}}
+\def\wblank{\wwblank{11mm}}
+\def\blank{\wwblank{8mm}}
+\def\vublank{*=<0mm,2.3mm>!D{}}
+\def\vdblank{*=<0mm,2.3mm>!U{}}
+\def\vsblank{*=<0mm,5.6mm>{}}
+\def\wirecross#1{\save[]!L;[#1]!R**@{-}\restore}
+\def\wirebraid#1#2{\save[]!L;[#1]!R**{}?(#2)**@{-}\restore\save[#1]!R;[]!L**{}?(#2)**@{-}\restore}
+\def\wireopen#1{\save[]!R;[#1]!R**\crv{[]!C&[#1]!C}\restore}
+\def\wireclose#1{\save[]!L;[#1]!L**\crv{[]!C&[#1]!C}\restore}
+\def\wireopenlabel#1#2{\save[]!R;[#1]!R**\crv{[]!C&[#1]!C}?<>(.5)*^+!R{#2}\restore}
+\def\wirecloselabel#1#2{\save[]!L;[#1]!L**\crv{[]!C&[#1]!C}?<>(.5)*^+!L{#2}\restore}
+\def\wireid{\blank\save[]!L*{};[]!R*{}**@{-}\restore}
+\def\wwireid{\wblank\save[]!L*{};[]!R*{}**@{-}\restore}
+\def\wwwireid#1{\wwblank{#1}\save[]!L*{};[]!R*{}**@{-}\restore}
+\newcommand{\corner}{\rule{2.5mm}{2.5mm}}
+\def\minheight{8mm}
+\def\addheight{8mm}
+\def\cornerbox#1#2#3{  % draw a box with a marked corner around
+                   % the object #1. #2=label
+                   % usage:   \ulbox{[ll].[].[ul]}{f}
+  \save#1="box";
+  "box"!C*+<0mm,\addheight>+=<0mm,\minheight>[|(5)]\frm{-}="box1";
+  "box1"*{#2};
+  ={"box1"!LU*!LU[]{\corner}}"ul";
+  ={"box1"!RU*!RU[]{\corner}}"ur";
+  ={"box1"!LD*!LD[]{\corner}}"dl";
+  ={"box1"!RD*!RD[]{\corner}}"dr";
+  #3
+  \restore
+}
+\def\nnbox#1#2{\cornerbox{#1}{#2}{}}
+\def\ulbox#1#2{\cornerbox{#1}{#2}{"ul"}}
+\def\urbox#1#2{\cornerbox{#1}{#2}{"ur"}}
+\def\dlbox#1#2{\cornerbox{#1}{#2}{"dl"}}
+\def\drbox#1#2{\cornerbox{#1}{#2}{"dr"}}
+\def\ubox#1#2{\cornerbox{#1}{#2}{"ul","ur"}}
+\def\dbox#1#2{\cornerbox{#1}{#2}{"dl","dr"}}
+\def\lbox#1#2{\cornerbox{#1}{#2}{"ul","dl"}}
+\def\rbox#1#2{\cornerbox{#1}{#2}{"ur","dr"}}
+\def\circbox#1{*+[o][F-]{#1}}
+
+\newtheorem{definition}{Definition}
+\newtheorem{n-lemma}{Lemma}[section]
+\newtheorem{n-example}[n-lemma]{Example}
+\newenvironment{example}{\begin{n-example}}{\end{n-example}}
+
+\newcommand{\ssep}{\hspace{1em}} 
+\newcommand{\sym}{c}       % braid: could be \sigma or c or \gamma
+\newcommand{\cp}{\circ}                 % composition
+\newcommand{\symi}{\sym\inv}
+\newcommand{\inv}{^{-1}}
+\newcommand{\id}{\textrm{\rm id}}       % identity
+\newcommand{\x}{\otimes}
+\newcommand{\boolt}{\mathit{bool}}
+\newcommand{\ffv}{\mathit{false}}
+\newcommand{\ttv}{\mathit{true}}
+\newcommand{\notb}{\mathit{not}}
 \newcommand{\union}{\cup}
 \newcommand{\transport}[2]{\overline{#1}(#2)}
-\newtheorem{definition}{Definition}
 \newcommand{\refl}{\mathit{refl}}
@@
 {\LARGE Isomorphisms as Groupoids: \\ 
-  Homotopy Type Theory with Fractional Types} \\[1.5cm]
+  Homotopy Type Theory and Fractional Types} \\[1.5cm]
 {\Large Jacques Carette${}^*$ \qquad Amr Sabry${}^{\dagger}$}\\[0.5cm]
@@
 \title{Isomorphisms as Groupoids: \\ 
-  Homotopy Type Theory with Fractional Types}
+  Homotopy Type Theory and Fractional Types}
 \author{Jacques Carette \and Amr Sabry}
@@
 
+From the wikipedia page on homeomorphism: 
+\begin{quote}
+The intuitive criterion of stretching, bending, cutting and gluing back
+together takes a certain amount of practice to apply correctly—it may not be
+obvious from the description above that deforming a line segment to a point
+is impermissible, for instance. It is thus important to realize that it is
+the formal definition given above that counts.
+
+This characterization of a homeomorphism often leads to confusion with the
+concept of homotopy, which is actually defined as a continuous deformation,
+but from one function to another, rather than one space to another. In the
+case of a homeomorphism, envisioning a continuous deformation is a mental
+tool for keeping track of which points on space X correspond to which points
+on Y—one just follows them as X deforms. In the case of homotopy, the
+continuous deformation from one map to the other is of the essence, and it is
+also less restrictive, since none of the maps involved need to be one-to-one
+or onto. Homotopy does lead to a relation on spaces: homotopy equivalence.
+
+There is a name for the kind of deformation involved in visualizing a
+homeomorphism. It is (except when cutting and regluing are required) an
+isotopy between the identity map on X and the homeomorphism from X to Y.
+\end{quote}
+
+It is strange for homotopy type theory to be all about isomorphisms and then
+not insist on isomorphisms between spaces.
+
+If we focus on homeomorphisms, the higher-order space is a \emph{torsor}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:intuition}
+
+%% the next 7 figures are taken from Selinger's paper on graphical languages
+%% for monoidal categories; if we use them make sure it's ok with him and
+%% acknowledge him
+
+In homotopy type theory, there is a focus on isomomorphisms represented by
+paths (proofs of identities) and yet arbitrary functions are considered. We
+propose to only consider functions that are themselves isomorphisms: this
+will allow us to view the function itself as a collection of paths and it
+doesn't lose anything in terms of computability because we know that we could
+do everything with reversible functions if we wanted to.
+
+Consider the function $\notb$ from the space $\boolt$ to itself. We can
+represent this function as a pair of paths. Since we are in a topological
+world, we should care whether the path top pass crosses over the bottom one
+or vice-versa. In the diagram below, we chose for the top wire to cross over
+the bottom one:
+
+\begin{center}
+  \begin{tabular}{@{}llc@{}}
+    $\notb : \boolt \rightarrow \boolt$ & & 
+    $\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
+        *{}\wireright{r}{\ffv}&\blank\wirecross{d}\wireright{r}{\ffv}&\\
+        *{}\wireright{r}{\ttv}&\blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
+        }}$ 
+  \end{tabular}
+\end{center}
+
+It is important also to note that the two paths must be taken together
+because there is a constraint between them. If the top path takes $\ffv$ to
+$\ttv$ then the bottom path cannot take $\ttv$ to $\ttv$: it \emph{has} to
+lead to $\ffv$. We will return to this point later.
+
+For now, consider the functions:
+\[\begin{array}{rcl}
+\mathit{id}^1 &=& \lambda x.x \\
+\mathit{id}^2 &=& \lambda x. \notb~(\notb~x)
+\end{array}\]
+These two functions are evidently extensionally equivalent and hence in the
+conventional treatment of homotopy type theory, it is acceptable to assume
+they are also propositionally equivalent, i.e., it is acceptable to postulate
+a path $\mathit{id}^1 \path \mathit{id}^2$ in the space $\boolt \rightarrow
+\boolt$.
+
+When represented as paths, we get the following situation for the second
+function:
+\[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
+    *{}\wireright{r}{\ffv}&
+    \blank\wirecross{d}\wireright{r}{\ttv}&
+    \blank\wirecross{d}\wireright{r}{\ffv}&
+    \\
+    *{}\wireright{r}{\ttv}&
+    \blank\wirebraid{u}{.3}\wireright{r}{\ffv}&
+    \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
+    \\
+    }}
+\]
+Repeating the $\notb$ operation results in a twisted path that externally
+behaves like the identity but that evidently not identical to the straight
+path with no twists. Thus although one can justify that the two functions
+should have a path between them, the function extensionality axiom does not
+give insights into \emph{what} this path is actually doing topologically.
+
+%% \begin{center}
+%%   \begin{tabular}{@{}llc@{}}
+%%     Braiding & $\sym_{A,B}$ &
+%%     $\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
+%%         *{}\wireright{r}{B}&\blank\wirecross{d}\wireright{r}{A}&\\
+%%         *{}\wireright{r}{A}&\blank\wirebraid{u}{.3}\wireright{r}{B}&
+%%         }}$ \\
+%%   \end{tabular}
+%% \end{center}
+
+%% Note that the braiding satisfies
+%% $\sym_{A,B}\cp\symi_{A,B}=\id_{A\x B}$, but not
+%% $\sym_{A,B}\cp\sym_{B,A}=\id_{A\x B}$. Graphically:
+
+%% \[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
+%%     *{}\wireright{r}{B}&
+%%     \blank\wirecross{d}\wireright{r}{A}&
+%%     \blank\wirebraid{d}{.3}\wireright{r}{B}&
+%%     \\
+%%     *{}\wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
+%%     \blank\wirecross{u}\wireright{r}{A}&
+%%     \\
+%%     }}
+%% = \id_{A\x B},
+%% \]
+%% \[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
+%%     *{}\wireright{r}{B}&
+%%     \blank\wirecross{d}\wireright{r}{A}&
+%%     \blank\wirecross{d}\wireright{r}{B}&
+%%     \\
+%%     *{}\wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{A}&
+%%     \\
+%%     }}
+%% \neq \id_{A\x B}.
+%% \]
+
+%% \begin{example}
+%%   The hexagon axiom translates into the following in the graphical
+%%   language:
+
+%%   \[ (\id_B\x\sym_{A,C}) \cp \alpha_{B,A,C} \cp (\sym_{A,B}\x\id_C)
+%%   = \alpha_{B,C,A} \cp (\sym_{B,C\x A}) \cp \alpha_{A,B,C}
+%%   \]
+%%   \[\vcenter{\wirechart{@R=6mm}{
+%%     \wireright{r}{C}&
+%%     \wireid\wireright{r}{C}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
+%%     \\
+%%     \wireright{r}{B}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
+%%     \\
+%%     \wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
+%%     \wireid\wireright{r}{B}&
+%%     \\
+%%     }}
+%%   \ssep=\ssep
+%%   \vcenter{\wirechart{@R=6mm}{
+%%     \wireright{r}{C}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
+%%     \\
+%%     \wireright{r}{B}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{C}&
+%%     \\
+%%     \wireright{r}{A}&
+%%     \blank\wirebraid{uu}{.2}\wireright{r}{B}&
+%%     \\
+%%     }}
+%%   \]
+%% \end{example}
+
+%% \begin{example}
+%%   The {\em Yang-Baxter equation} is the following equation, which is a
+%%   consequence of the hexagon axiom and naturality:
+%%   \[ (\sym_{B,C}\x\id_A)\cp(\id_B\x\sym_{A,C})\cp(\sym_{A,B}\x\id_C) =
+%%   (\id_C\x\sym_{A,B})\cp(\sym_{A,C}\x\id_B)\cp(\id_A\x\sym_{B,C}).
+%%   \]
+%%   In the graphical language, it becomes:
+%%   \[\vcenter{\wirechart{@C-4mm@R=6mm}{
+%%     \wireright{r}{C}&
+%%     \wireid\wireright{r}{C}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
+%%     \wireid\wireright{r}{A}&
+%%     \\
+%%     \wireright{r}{B}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{B}&
+%%     \\
+%%     \wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
+%%     \wireid\wireright{r}{B}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
+%%     \\
+%%     }}
+%%   \ssep=\ssep
+%%   \vcenter{\wirechart{@C-4mm@R=6mm}{
+%%     \wireright{r}{C}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{B}&
+%%     \wireid\wireright{r}{B}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
+%%     \\
+%%     \wireright{r}{B}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
+%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
+%%     \\
+%%     \wireright{r}{A}&
+%%     \wireid\wireright{r}{A}&
+%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
+%%     \wireid\wireright{r}{C}&
+%%     \\
+%%     }}
+%%   \]
+%% \end{example}
 

### 2013-10-19 f.tex
discussion about treating functions as collections of paths if these functions are isos
braids etc

@@
 
+\newcommand{\br}{\mathit{Br}}
 \newcommand{\ssep}{\hspace{1em}} 
@@
 \newcommand{\inv}{^{-1}}
+\newcommand{\lolli}{\multimap} 
 \newcommand{\id}{\textrm{\rm id}}       % identity
@@
 
-%% the next 7 figures are taken from Selinger's paper on graphical languages
+%% the next figures are taken from Selinger's paper on graphical languages
 %% for monoidal categories; if we use them make sure it's ok with him and
@@
 
-In homotopy type theory, there is a focus on isomomorphisms represented by
-paths (proofs of identities) and yet arbitrary functions are considered. We
-propose to only consider functions that are themselves isomorphisms: this
-will allow us to view the function itself as a collection of paths and it
+In homotopy type theory, there is a focus on isomorphisms: for most of the
+spaces an isomorphism between $A$ and $B$ is represented by a path denoting a
+particular proof of the identify $A \equiv B$. Different proofs yield
+different paths and these proofs can be related by 2-paths to reason about
+proof equivalence and so on. Yet in the middle of all these identities and
+isomorphisms, conventional homotopy type theory deals with arbitrary
+functions, then defines what it means \emph{extensionally} for functions to
+be isomorphic, and \emph{then} postulates an axiom that extensionally
+equivalent functions are identical in the sense of having a path between
+them. This treatment gives no insight about the nature paths between
+functions and is undesirable for many reasons.
+
+We propose to just consider functions are isomorphisms by construction. This
 doesn't lose anything in terms of computability because we know that we could
-do everything with reversible functions if we wanted to.
+do everything with reversible functions if we wanted to. There are actually
+many advantages of doing so as we and others have argued at length
+before. Anyway, the point is that if we have isomorphisms-by-construction
+functions then the function itself becomes naturally a collection of
+paths. We illustrate this point with examples.
 
@@
 represent this function as a pair of paths. Since we are in a topological
-world, we should care whether the path top pass crosses over the bottom one
-or vice-versa. In the diagram below, we chose for the top wire to cross over
-the bottom one:
-
+world, we should care about how the paths cross over each other.  In the
+diagram below, we chose for the top path to cross over the bottom one:
 \begin{center}
@@
 \end{center}
-
-It is important also to note that the two paths must be taken together
+It is important to note that the two paths must be taken together
 because there is a constraint between them. If the top path takes $\ffv$ to
 $\ttv$ then the bottom path cannot take $\ttv$ to $\ttv$: it \emph{has} to
-lead to $\ffv$. We will return to this point later.
-
-For now, consider the functions:
+lead to $\ffv$. In other words, it is incorrect to focus only on the top path
+and conclude that there is an identity between $\ffv$ and $\ttv$. What is
+correct is that, collectively, there is an identity between $\boolt$ and
+$\boolt$ that permutes the elements of $\boolt$. Technically we have a
+\emph{braid}.
+
+We will return to this point later. But for now, consider the functions:
 \[\begin{array}{rcl}
-\mathit{id}^1 &=& \lambda x.x \\
-\mathit{id}^2 &=& \lambda x. \notb~(\notb~x)
+\idc^1 &=& \lambda x.x \\
+\idc^2 &=& \lambda x. \notb~(\notb~x)
 \end{array}\]
 These two functions are evidently extensionally equivalent and hence in the
-conventional treatment of homotopy type theory, it is acceptable to assume
-they are also propositionally equivalent, i.e., it is acceptable to postulate
-a path $\mathit{id}^1 \path \mathit{id}^2$ in the space $\boolt \rightarrow
-\boolt$.
-
-When represented as paths, we get the following situation for the second
-function:
-\[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
+conventional treatment of homotopy type theory, they are also treated as
+propositionally equivalent by the function extensionality axiom. When the
+functions themselves are represented as paths, however, we get the following
+situation for the second function:
+\[
+\idc^2 = \lambda x.\notb~(\notb~x) \qquad
+\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
     *{}\wireright{r}{\ffv}&
-    \blank\wirecross{d}\wireright{r}{\ttv}&
+    \blank\wirecross{d}\wireright{r}{\ffv}&
     \blank\wirecross{d}\wireright{r}{\ffv}&
@@
     *{}\wireright{r}{\ttv}&
-    \blank\wirebraid{u}{.3}\wireright{r}{\ffv}&
+    \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
     \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
@@
 \]
-Repeating the $\notb$ operation results in a twisted path that externally
-behaves like the identity but that evidently not identical to the straight
-path with no twists. Thus although one can justify that the two functions
-should have a path between them, the function extensionality axiom does not
-give insights into \emph{what} this path is actually doing topologically.
+Repeating the $\notb$ operation results in a twisted braid that extensionally
+behaves like the trivial braid but that is evidently not identical to
+it. Formally to find a (2-)path between the functions requires us to find an
+\emph{isotopy} between the two braids, where loosely speaking an isotopy is
+essentially homotopy in the case of bijections. The two braids for $\idc^1$
+and $\idc^2$ are \emph{not} isotopic so there is no path between them in our
+world. What is interesting is that the braid isotopy problem has efficient
+solutions and hence we are in a situation where function equivalence has
+efficient algorithms.
+
+Note that one can define another version of boolean negation $\notb'$ in
+which the bottom path is the one that crosses over the top one. Using this
+version, the function $\idc^3 = \lambda x.\notb'~(\notb~x)$ would look like:
+\[
+\idc^3 = \lambda x.\notb'~(\notb~x) \qquad
+\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
+     *{}\wireright{r}{\ffv}&
+     \blank\wirecross{d}\wireright{r}{\ffv}&
+     \blank\wirebraid{d}{.3}\wireright{r}{\ffv}&
+     \\
+     *{}\wireright{r}{\ttv}&
+     \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
+     \blank\wirecross{u}\wireright{r}{\ttv}&
+     \\
+     }}
+\]
+And this version is isotopic to the trivial braid, and hence we would have a
+(2-)path between $\idc^1$ and $\idc^3$.
+
+To summarize, if we restrict our attention to a certain class of functions,
+these functions are naturally described as braids and function equivalence
+(i.e, paths between functions) are generated by braid isotopy. This is
+already quite neat but note that we have only talked about functions
+\emph{between} spaces and modeled these as braids. What we have not done yet
+is describe the space of functions itself. In our running example, we need to
+describe some space $\boolt \lolli \boolt$ that represents the various
+functions we discussed from $\boolt$ to $\boolt$.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Pointed Spaces}
+
+We will construct this in steps. First we note that a function is a ``large''
+object that can be applied to many possible arguments. Each application is
+however associated with exactly one argument. Thus, in this first step, we do
+not model functions in their entirety; we only model functions \emph{at one
+  particular point}. We will return to the issue of modeling the entire
+function object later. For now, we have technically moved from \emph{spaces}
+to \emph{pointed spaces}. Let's revisit our previous discussion from the
+perspective of pointed spaces.
+
+The type $\boolt$ is now split in two types: the type $(\boolt,\ffv)$ with
+$\ffv$ as the distinguished point and $(\boolt,\ttv)$ with $\ttv$ as the
+distinguished point. We now have two copies of the function $\notb$, a
+version $\notb_{\ffv} : (\boolt,\ffv) \rightarrow (\boolt,\ttv)$ and a
+version $\notb_{\ttv} : (\boolt,\ttv) \rightarrow (\boolt,\ffv)$ depending on
+the starting basepoint. Indeed functions must respect the basepoints, i.e.,
+map basepoints to basepoints. The function is still a braid on the complete
+$\boolt$ space but in the first case, the distinguished point $\ffv$ tells us
+that the function is ``applied'' to $\ffv$ and hence that the top branch is
+the one in focus.
+
+Concretely then we want spaces (i.e., groupoids) $(\boolt,\ffv) \lolli
+(\boolt,\ffv)$, $(\boolt,\ffv) \lolli (\boolt,\ttv)$, $(\boolt,\ttv) \lolli
+(\boolt,\ffv)$, and $(\boolt,\ttv) \lolli (\boolt,\ttv)$. As examples, the
+first space will contain $\idc^1_{\ffv}$ and $\idc^3{\ffv}$; the second space
+will contain $\notb_{\ffv}$ and $\notb'_{\ffv}$; the third space will contain
+$\notb_{\ttv}$, and the last space will contain $\idc^1_{\ttv}$. In the first
+space, there will be a path between $\idc^1_{\ffv}$ and $\idc^3{\ffv}$ as
+there is an isotopy between them etc. To describe these spaces fully, we need
+introduction rules, elimination rules, computation rules, and induction
+principles, etc. It should be clear that we can't just use
+$\lambda$-expressions as introduction rules: they are not refined enough to
+capture all the invariants we want.
+
+Let's start with some basic observations. How many points (i.e., functions or
+braids) are there in the space $(\boolt,\ffv) \lolli (\boolt,\ffv)$. These
+are all the twists on two elements with the paths starting and ending at
+$\ffv$ highlighted: by giving a twist in one direction the value $+1$ and a
+twist in the other direction the value $-1$, we see that this is isomorphic
+to the integers under addition. We would like to argue that this space has
+groupoid cardinality 1 and if we succeed then it is natural to think of it as
+the product of a space with fractional cardinality $\frac{1}{2}$ and the
+space $\boolt$ with cardinality 2. Think of all the words over the alphabet
+$\{ 0, 1, -1\}$ with rules $0x\rightarrow x$, $x0\rightarrow x$,
+$1(-1)\rightarrow\epsilon$ and $(-1)1\rightarrow\epsilon$. This produces
+equivalence classes which describe the connected components (isotopy
+equivalences) of the space $\boolt\lolli\boolt$. Let's denote by $V_n$ the
+size of the equivalence class of words whose value is $n$ where the value is
+the sum of the elements. The groupoid cardinality of the space is therefore
+$\Sigma_{n\in\mathbb{Z}} \frac{1}{A_n}$. Does this thing have value 1???
+Plausible. Assume yes for now and proceed under the assumption that it is 1
+and hence that it makes sense to decompose the space into the product of
+something of cardinality $\frac{1}{2}$ and something of cardinality $2$. So
+what kind of groupoid would have cardinality $\frac{1}{2}$. This actually is
+something well studied. Talk about groupoid cardinality and give an example
+of how the cardinality of a group of order $n$ is $\frac{1}{n}$. Some
+intuition about the paths reducing the number of elements and the Euler
+characteristic.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Fractionals}
+
+As an example, let's focus now on building the type
+$\frac{1}{(\boolt,\ffv)}$: it will correspond to the braid group $\br_2$ on
+two strands. This type will have an infinite number of points each
+representing one element of the braid group $\br_2$. Mathematically, for $n$
+strands, the braid group $\br_n$ is generated by $\sigma_1, \ldots,
+\sigma_{n-1}$ and the following relations:
+\begin{itemize}
+\item $\sigma_i\sigma_{i+1}\sigma_i = \sigma_{i+1}\sigma_{i}\sigma_{i+1}$, 
+\item $\sigma_i\sigma_j = \sigma_j\sigma_i$ if $|i-j| \geq 2$.
+\end{itemize}
+So for the particular case of $\br_2$, we have one generator $\sigma$ and no
+relations. The generator corresponds to one non-trivial twist in one
+direction. The group elements (i.e., the elements of
+$\frac{1}{(\boolt,\ffv)}$) are the trivial braid, the $n$-fold compositions
+of $\sigma$ for all $n$, and the $n$-fold compositions of $\sigma^{-1}$ for
+all $n$.
+
+So generally speaking $\eta_n$ will produce a product of $\br_n$ and $n$ and
+$\epsilon_n$ will be the group action of the group on $\br_n$ on the
+particular strand $n$.
+
+%%%%%%%%%%%%%%%%%
+\subsection{Negatives?}
+
+Need operations to change the basepoint. Would these be the negatives?
 
@@
 %% \end{example}
-
-Talk about groupoid cardinality and give an example of how the cardinality of
-a group of order $n$ is $\frac{1}{n}$.
 

### 2013-10-20 f.tex

@@
 $\epsilon_n$ will be the group action of the group on $\br_n$ on the
-particular strand $n$.
+particular strand $n$. A function will have a particular wiring; this wiring
+is reflected as a first class value using the group generators. For booleans,
+the group generators can be represented as id, not, and not'; every wiring is
+a sequence of these. So values of type $1/bool$ are (id), (not o not'), (not'
+o not'), (not o not o not), etc. or in other words the code for the circuits
+becomes represented as a value. (Is that related to the encode/decode method
+in the HoTT book?) How about functions: T1 -> T2. The types T1 and T2 must be
+connected by paths so we might as well think about T1 -> T1.
+
+So B has value f and t
+1/B has values id; not; not'; and their compositions (semantically
+represented as elements of the group Br2.
+
+eta : (1,()) -> ((B,1/B),(f,id))
+eta () = (f , id)
+
+eta : (1,()) -> ((B,1/B),(f,not))
+eta () = (t , not)
+
+...
+
+epsilon : ((B,1/B),(f,not)) -> (1,())
+epsilon (t , not) = ()
+
+...
 

### 2013-11-03 f.tex
talk
f

@@
 %% acknowledge him
+
+A permutation is represented as
+
+\begin{verbatim}
+(1 2 3 4 5)
+(2 5 4 3 1)
+\end{verbatim}
+
+We can decompose it in:
+
+\begin{verbatim}
+2
+
+and
+
+1 with paths (2 3 4 5)
+             (5 4 3 1)
+\end{verbatim}
+
+This is the decomposition of the function \verb|A -> A| into
+\verb|A * 1/A|. A problem still with cardinality: there is exactly one
+element into the latter type but presumably several functions in the
+former. Perhaps all these functions are equivalent somewho up to homotopy?
 
@@
 \end{document}
+
+{--
+
+-- In our perspective isomorphisms are represented by paths; there are
+-- 1! isomorphims from ? to itself so we expect exactly one path in
+-- the space ? o-> ?, i.e., one path from ? to itself, i.e., the space
+-- (? ? ?) should be isomorphic to ?. However we can't witness this
+-- isomorphism yet until we postulate the ? and ? paths.
+--
+-- unitPath' : (? ? ?) ? ?
+-- unitPath' = ?
+
+-- functions and types should have their equivalences as paths NOT
+-- extensional isomorphisms which are later postulated to be paths!!!
+
+-- need some kind of ``vertical'' path that allows us to move one level up.
+
+-- explore this: 1 -> 1/x * x goes to point x which has a higher
+-- groupoid structure ``above it'', gives you the higher groupoid
+-- structure as 1/x and the naked point as x. Epsilon can then
+-- recombine a higher groupoid collection of paths with a
+-- point. Recall the 3/4 example: the paths not attached to any points
+-- are 3/4: they take 4 points and return 3 points.
+
+-- to get 1/5
+-- start with 5 points * * * * *
+-- and paths that connect them so that they are identified to be 1 point
+-- Separate the points from the paths: the paths represent 1/5: something that takes
+-- 5 points and collapses them to 1 and so we get 1_{5} --> 1/5 * 5
+-- The problem is that 5 now has no paths over it so it's a different 5 and 1/5 has
+-- no points so it's not a groupoid or is it?
+
+-- 1/x can switch the two components of [_,_] so that it is an involution
+
+We don't have pointed types. We have HITs. Each type is a collection
+of points and a collection of non-trivial paths. (We get refl, !, ?,
+associativity, etc. for tree and we also get the higher groupoid
+structure for free which witnesses for example the pentagon and
+triangle axioms, i.e., the coherence of assoc and the nice interplay
+with !).
+
+Start with the type 1 at 3 [{a,b,c) , {(a==b),(b==c)}]. This type
+represents ``1'' but uses three identical elements to express this fact.
+
+The paths by themselves are some sort of function that takes three
+points and identifies them; this can be represented by the type
+1/3. The three points by themselves can be represented by the type 3.
+
+So 3 is [{a,b,c}, ?]
+
+and 1/3 is [{? a b c ? (a==b,b==c)}, ?] 
+the ? is a mini one step function to express that the paths are not
+attached to any end points. You give me the end points and I will connect them.
+
+The only value of type 1/3 is that function. We can pass around it and
+eventually we must apply to 3 points which become identified. So
+eventually we recover 1 back.
+
+So now what is 1/(1/3). Remember that to express 1/3. We started with
+a version of 1 that had 3 elements and paths between them. So to do
+1/(1/3) we need to start with a version of 1 that has 1/3 elements and
+paths between them.
+
+1 at 1/3 is [{? a b c ? (a==b,b==c)}, 3 paths from this point to itself ]
+
+so 1 at 1/3 can be split into 1/3 and and 1/(1/3) as follows:
+1/3 = [{? a b c ? (a==b,b==c)}, ?]
+1/(1/3) = [ ? a ? 3 paths from a to itself , ? ]
+
+--
+
+swapP : {A B : Set} ? A × B ? B × A
+swapP {A} {B} = ?
+
+It looks like I will need an axiom (not univalence necessarily but
+something like it. We want to say that type isos from Pi correspond to
+paths. So instead of saying (A ? B) ? (A ? B) we would say (A ? B) ?
+(A ? B) where ? is the type of isos from Pi (making the types into a
+field like structure).
+
+Ok so there is an external map between A × B and B × A but because we
+are monoidal closed there is also an internal map (1/(A × B)) × (B × A)
+or more readably (A × B) -o (B × A). This map is defined as follows.
+
+First let's start with 1 at A × B. This is:
+
+[ {(a0,b0),(a0,b1),(a1,b0),(a1,b1)} , {(a0,b0)?(a0,b1), (a0,b0)?(a1,b0), ... }]
+
+Anyway the point is that we can build 1/(A×B). Now we know that every
+(a0,b0)?(a1,b1) comes from (a0?a1) and (b0?b1) and hence we can
+assemble them in the other direction to get (b0,a0)?(b1,a1).
+
+Now we can split using ? to 1/(A×B) and (A×B) where:
+
+A×B = [ {(a0,b0),(a0,b1),(a1,b0),(a1,b1)} , ?]
+and
+1/(A×B) = [? (a0,b0) (a0,b1) (a1,b0) (a1,b1) ? 
+             {(a0,b0)?(a0,b1), (a0,b0)?(a1,b0), ... } , 
+           ?]
+
+which is the same as:
+
+1/(A×B) = [? a0 a1 b0 b1 ? {a0?a1, b0?b1} , ?]
+
+Now the map (A×B)-o(B×A) is represented by the type:
+
+[{((b0,a0),? a0 a1 b0 b1 ? {a0?a1, b0?b1}),
+   ((b0,a1),? a0 a1 b0 b1 ? {a0?a1, b0?b1}),
+   ((b1,a0),? a0 a1 b0 b1 ? {a0?a1, b0?b1}),
+   ((b1,a1),? a0 a1 b0 b1 ? {a0?a1, b0?b1})}
+  ?]
+
+Now applying the function to (a0,b0) means pairing the above with (a0,b0):
+
+[{((b0,a0),(a0,b0),? a0 a1 b0 b1 ? {a0?a1, b0?b1}),
+   ((b0,a1),(a0,b0),? a0 a1 b0 b1 ? {a0?a1, b0?b1}),
+   ((b1,a0),(a0,b0),? a0 a1 b0 b1 ? {a0?a1, b0?b1}),
+   ((b1,a1),(a0,b0),? a0 a1 b0 b1 ? {a0?a1, b0?b1})}
+  ?]
+
+--}
+
+------------------------------------------------------------------------------

### 2013-11-04 f.tex

@@
 
-A permutation is represented as
-
+What can we say in pi about these two combinators:
 \begin{verbatim}
-(1 2 3 4 5)
-(2 5 4 3 1)
+Let f : A -> C and G : B -> D
+
+(id x g) o (f x id)  :  A x B -> C x D
+
+(f x id) o (id x g)  :  A x B -> C x D
 \end{verbatim}
-
-We can decompose it in:
-
-\begin{verbatim}
-2
-
-and
-
-1 with paths (2 3 4 5)
-             (5 4 3 1)
-\end{verbatim}
-
-This is the decomposition of the function \verb|A -> A| into
-\verb|A * 1/A|. A problem still with cardinality: there is exactly one
-element into the latter type but presumably several functions in the
-former. Perhaps all these functions are equivalent somewho up to homotopy?
+Are they the ``same''? We can reason extensionally to check that for all
+inputs these two combinators produce the same results. In other words, we
+reason from the axioms of category theory which includes things like
+\verb|f o id = f| which are extensional equalities. (If we had weak
+$n$-categories here what would happen, perhaps this is where we are leading
+with the next sentence.) Better is to reason topologically exploiting the
+original connections established by Joyal and
+Street~\cite{planardiagrams,geometrytensor}. One can in this case show the
+equivalence of these combinators by using the graphical language and noticing
+that there is a \emph{planar isotopy} between the two diagrams, that moves
+things around without crossing, cutting, or gluing any wires.
+
+More formally, we view each type as a topological space and each function as
+a deformation of the space expressed as a (sequential, parallel, etc.)
+composition of primitive deformations (id, swap, unit, assoc, distribute,
+factor, etc.). If we have a planar monoidal category, then there isn't much
+you can do: the types are collections of lines grouped in bundles. The only
+deformations that functions can do are to re-associate the bundles and get
+rid of unit wires. (If the ambient space is a plane, i.e., if we restrict our
+``hardware model'' to be a 2D model in which wires can't cross over each
+other, then we can't even equate the functions that apply id to different
+bundles. If the ambient space is 3D then we can. Eq. 3.2 in Selinger's survey
+of graphical languages for monoidal categories.) An example of two functions
+we can show to be the same in the planar model is the one above in the
+verbatim block. An isotopy exists between the two combinators because we can
+move things around in the plane as long as we don't cross or cut any wires.
+
+If we allow braiding (i.e, commutativity) we must move to ambient isotopies
+in 3D. In particular, the two versions of not that cross true over false or
+vice versa are not isotopic.
+
+If we also want, the composition of not and not to be expressible directly as
+a twist, we must move to \emph{framed} isotopies in 3 dimensions, where the
+lines are represented as flat ribbons.
+
+A symmetric monoidal category requires ambient isotopy in 4 dimensions so
+we're not going to consider it.
+
+If we are going to have higher-order functions, we need to move up, i.e., we
+need to build a space in which the types are the functions of the lower
+level, and in which the functions are the isotopies of the lower level. In
+other words, we need to be able to express the lower level functions as
+topological spaces and we need to be able to express the lower level
+isotopies as functions.
+
+Think of a function as a map connecting points from one space to points in
+the other space (respecting paths etc.). This map gives a geometric shape. We
+will think of one line in that shape as solid with all the other lines dashed.
+
+Another unrelated observation is that Pi gives us an inductive
+characterization of functions! If that whole thing with fractionals
+works out we will have an inductive characterization of higher-order
+functions!
 
@@
 
-We propose to just consider functions are isomorphisms by construction. This
-doesn't lose anything in terms of computability because we know that we could
-do everything with reversible functions if we wanted to. There are actually
-many advantages of doing so as we and others have argued at length
-before. Anyway, the point is that if we have isomorphisms-by-construction
-functions then the function itself becomes naturally a collection of
-paths. We illustrate this point with examples.
+We propose to just consider functions that are isomorphisms by
+construction. This doesn't lose anything in terms of computability because we
+know that we could do everything with reversible functions if we wanted
+to. There are actually many advantages of doing so as we and others have
+argued at length before. Anyway, the point is that if we have
+isomorphisms-by-construction functions then the function itself becomes
+naturally a collection of paths. We illustrate this point with examples.
 
@@
 
+Are higher-order inductive types something like ADTs where we have equations
+or something like graphs where we can build interesting shapes inductively or
+perhaps both somehow?
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 wrong backtrack and choose the other branch.
+
+Double negation vs. braids and twists. Any insights from category theory or
+topology that can be used to understand classical proofs?
 
@@
 \section{Connections to Physics and Quantum Computing}
+
+Connections between standard circuit model and topological quantum computing.
 
@@
 ------------------------------------------------------------------------------
+
+CS: Give computational content of science
+  Logic: Curry-Howard
+  Classical logic: Continuations
+  Physics: Quantum information; computation
+  cognition
+  linguistics
+  Construction of cellos
+
+Mathematical proofs have computational content
+Physical equations have computational content
+
+Now at the heart of it all: EQUALITY has computational content and is the
+most fundamental relation we use implicitly
+
+Theory at the center of many many things: topology, algebra, geometry, type
+theory, physics, etc. ==> Exciting
+
+Conservation of information, energy, etc. 
+Symmetry
+F = m a
+or E = m C^2
+are not EQUALITIES. They are isomorphisms: go from one to the other
+preserving various entitites. 
+Get rid of equality ==> use isomorphisms exclusively
+
+Turtles all the way down (or up in that case)
+
+--
+
+when are two structures "the same"
+if their elements are the same
+
+what is the structures are infinite? (i.e., described by a process); go
+right, then right, then right, then right is the same as stay here but not
+quite the same
+
+E=mc2, F=ma
+
+Leibnitz
+
+--
+
+ - topology of computation/evaluation: in addition to loops/circle,
+   bernstein conditions for parallelism; or more accurately two
+   computations are homotopic if one is sequential and is parallel and I
+   can massage one to the other
+
+ - univalence: (A \== B) isEquiv (A isEquiv B) fun. ext. (f \== g) -> (x
+   : A) f(x) \== g(x) is an equivalence paths between complex types iff
+   there are paths between the points: works for simple finite types;
+   for infinite things can't be proved; need these two additional axioms
+
+ - for my intro: what is hott; article about hott in magazine; 
+   steve awodey's talk
+   "http://www.andrew.cmu.edu/user/awodey/hott/CMUslides.pdf"
+   formalize mathematics if you care about that: eckmann-hilton
+   "http://blogs.scientificamerican.com/guest-blog/2013/10/01/voevodskys-mathematical-revolution/"
+    Logic level -1
+    Set theory level 0
+    groupoid level 1
+    ...
+    The Propositions-as-Types conception of Martin-Löf type theory leads
+   to a system of logic that is different from both classical and
+   intuitionistic logic with respect to the statements that hold in the
+   fragment of logical language that these three share, namely the
+   quantifiers \Pi , \Sigma , equality \mathsf{Id} , conjunction \times
+   , disjunction + , implication \rightarrow and negation \neg .  For
+   example, it validates the axiom of choice but not the law
+   of excluded middle  a combination that is impossible
+   classically and intuitionistically.  Let us call this conception
+   constructive validity following a paper by Dana Scott in
+   which it was first introduced (Constructive Validity, Symposium on
+   Automated Deduction, LNM 125, Springer (1970), pp. 237-275).
+
+ - for class "http://math.ucr.edu/home/baez/eckmann_hilton.jpg"
+
+ - circle ==> recursion: homotopy between path p and path q means
+   if we can move from p to q without getting snagged in a loop, i.e.,
+   without crossing over an infinite loop!!
+
+ - Agda and homotopy type theory idea: Ulf showed us how to fill in
+   "small" program fragments automatically. If we had more precise types
+   can the compiler fill in different algorithms for us. Here is an
+   idea: in HoTT we can control the isomorphisms and what we mean by
+   them. For example, we can add paths between A and B to mean that B is
+   a sorted version of A. Then different arrows would correspond to
+   different sorting algorithms. If we maintain proof relevance then we
+   can do actions depending on the particular sorting algorithm that is
+   used to go from A to B. This might be doable in Pi actually.
+
+- different levels correspond to fractal equality or perhaps each level gives
+  you more precision
+
+- Don't you think that these:
+  "http://en.wikipedia.org/wiki/Loop_optimization"
+  are homotopies. They do make some explicit connections to
+  "geometry/topology". I think we have discovered a gold mine
+
+So I thought more about it and I think it holds more water. Here is the story:
+
+- paths in HoTT represent computations; this is true in the conventional
+  HoTT but even more or so in our context where all computations are
+  isomorphisms;
+
+- paths that loop are therefore looping computations;
+
+- the loop in a 2groupoid terminates after two iterations;
+
+- the loop in S1 is an infinite loop;
+
+- the loop in S2 is a double infinite loop (on two natural
+  numbers);
+
+- the various geometric objects capture various patterns of iteration!
+
+- Filinski's paper below shows that if you are given one simple infinite
+  loop and continuations, then you can expression recursion.  So if we
+  include S1 in our system and we figure out how to do higher-order
+  functions (and their duals as delimited continuations), then we should
+  be able to express recursion. A small hope would be that we can get
+  away without full continuations and that we can get recursion from S1
+  using just the groupoid structure but that's speculative.
+
+I like it! --Amr
+
+Check this out: "http://iml.univ-mrs.fr/geocal06/"=20
+
+In particular this workshop:
+
+Algebraic Topology, Concurrency and Rewriting
+
+Organizers: =C3=89ric Goubault, Yves Lafont
+
+The objective of the course is to give a general overview of the recent
+developments of ``geometrical'' approaches to the theory of computation,
+in particular concurrent and distributed computations, and rewriting.
+
+The course will start by a general introduction to homotopy and
+homology, classical on topological spaces, and generalize through
+abstract homotopy theory and generalized homology theories. Then two
+different applications will be developped.
+
+The first one concerns concurrency theory. In this application, a
+refined form of homotopy (dihomotopy) has to be designed, that takes
+into account the direction of time. Similar methods as the ones on
+ordinary topological spaces can be defined, in order to find
+``invariants'' under dihomotopy, i.e. a way to classify directed shapes
+under (directed) deformation. These invariants can in turn be used for
+purposes of proofs or static analysis of algorithms or programs. We will
+show such applications in the course. Other applications, of more
+classical algebraic topological invariants will also be derived, in the
+field of fault-tolerant distributed systems.
+
+The second one concerns the rewriting theory. Since the pioneering work
+by Squier, we know how to use homology to characterize the algebraic
+objects presented by rewriting systems. Basically, words are seen as
+objects of dimension one, rewriting rules are seen as of dimension two,
+confluence diagrams as of dimension three etc. providing resolutions to
+the algebraic objects they present. Extensions to term rewriting has
+been recently solved, along similar lines, using more general homology
+theories. Also higher-dimensional rewriting has gained wide interest and
+its theory will be fully developped, in conjunction with the necessary
+n-categorical (or polygraphic) theory.
+
+mfsp paper by Fajstrup, Goubault, and Raussen
+
+http://www.pps.univ-paris-diderot.fr/~laurent/geocal/slides/mellies.pdf‎
+
+-- Repeat for monoids without using equivalences!
+
+-- Let's build a monoidal category whose objects are ∞-groupoids using
+-- Thorsten's trick perhaps. Or perhaps we only need 1-groupoids or
+-- 2-groupoids... Or better use HITs: explicit constructors for points
+-- and paths and let the whole ∞-groupoid structure be implicit and
+-- generated by J on demand as needed. A morphim between such HITs
+-- would map points to points and paths to paths LYING OVER the
+-- original paths (using transport).
+
+-- multiplicative monoid:
+-- bifunctor ⊗ maps two types A and B to the type (A ⊗ B):
+--   map a point in A and a point in B to a point in (A ⊗ B)
+--   map a path in A and a path in B to a path in (A ⊗ B)
+--   if paths p in A and q in B are respectively equal to paths p' in
+--     A and q' in B then their images in (A ⊗ B) are also equal
+-- associative up to ...
+-- an object I which is both a left and right identity for ⊗ 
+-- coherence conditions: pentagon and triangle
+-- 
+-- Now what is the structure on 1 and ⊗ in terms of paths on the
+-- components
+
+-- map points to points
+-- map paths to paths
+-- preserve ≡
+
+apTensor : {A B C : Set} {x y : A × B} → (f : A × B → C) → (x ≡ y) → (f x ≡ f y)
+apTensor {A} {B} {C} {x} {y} f p = 
+  pathInd 
+    (λ {x} {y} p → f x ≡ f y) 
+    (λ x → refl (f x))
+    {x} {y} p
+
+-- monoid 0 and ⊕
+
+-- do all these sections in Ch. 2 with monoidal categories instead of
+-- the usual type constructors
+
+-- imagine we had tensor products instead would be it possible to
+-- convert a path : A × B → C × D to two sequential paths A → C with
+-- refl on B followed by the path B → D and refl on C. So the
+-- topological view of computation would express which dependencies
+-- allow you to move computations around to make them more or less
+-- parallel. Delooping sounds like it would allow you to make a loop
+-- (sequential) into something that operates in parallel (say every
+-- two iterations happen in parallel) but that can only happen if the
+-- dependencies are "right".
+
+loops/spaces; loop optimizations; cata/epi/etc-morphisms are also
+patterns of recursion; probably even parametricity and logical
+relations; all these things must be homotopies.
+
+1/x gives you the pattern to define a clause
++ gives you the ability to write all the clauses
+- gives you the ability to retract a clause
+all functions are constructible from the identity by massaging it
+functors between types internalized inside a type using 1/., +, and -
+
+PI
+
+cardinality of these infinity groupoids
+
+delooping: 1/Bool = 1/(delooping false) + 1/(deloopoing true)
+
+≡ {u+1} {Set u : Set (u+1)} (T1 : Set u) (T2 : Set u) -> Set (u+1)
+
+CC : ∀ {u} → (A B : Set u) → _≡_ {L.suc u} {Set u} A B → Set (L.suc u)
+CC {u} A B p = _≡_ {L.suc u} {Set u} (A × B) (B × A)
+
+cc : ∀ {u} → (A : Set u) → _≡_ {L.suc u} {Set u} (A × A) (A × A)
+cc A = refl
+
+swapP : ∀ {u} → (A B : Set u) → _≡_ {L.suc u} {Set u} (A × B) (B × A) 
+swapP {u} = pathIndU {u} {Set u} CC cc
+
+t : ℕ × Bool ≡ Bool × ℕ
+t = swapP ℕ Bool
+
+How many elements are in this set:
+
+{ Egyptian, American, Chinese }
+
+It depends on the equivalence relation. If you think the trivial
+relation, there are 3 elements. If you think the relation that
+identifies Egyptian/American (Amr) then you have two elements only in
+the set. We can make this equivalence relation depend on the
+particular element in question and we can make it first-class. When
+viewed by itself, it is like the fraction 2/3: it takes 3 elements and
+identifies two of them giving you back two elements.
+
+https://inf.ug.edu.pl/~stefan/AlgTop/
+
+--
+
+All programming only builds trees. 
+Graphs require assignments and ugly things
+HoTT allows us to build datatypes with "paths", "shapes", etc.
+
+CS -> Topology: study shapes with inductive types
+Topology -> CS: richer datatypes with cycles
+
+Can I build types that solve polynomials ???
+
+Define pointed types (A,a)
+
+Reciprocal of a pointed type (A,a) is delooping (a=a,refl)
+
+delooping of A is (B,b) such that loopspace(B,a) = A
+
+our construction of fractionals is delooping: take n points connected by a
+group stucture; turn them into a point with the group structure turned into
+path composition.
+
+A permutation is represented as
+
+\begin{verbatim}
+(1 2 3 4 5)
+(2 5 4 3 1)
+\end{verbatim}
+
+We can decompose it in:
+
+\begin{verbatim}
+2
+
+and
+
+1 with paths (2 3 4 5)
+             (5 4 3 1)
+\end{verbatim}
+
+This is the decomposition of the function \verb|A -> A| into
+\verb|A * 1/A|. A problem still with cardinality: there is exactly one
+element into the latter type but presumably several functions in the
+former. Perhaps all these functions are equivalent somewho up to homotopy?

### 2013-11-07 f.tex
new versions of agda code compatible with 3.3

@@
 \usepackage{fullpage}
+\usepackage{tikz}
 \usepackage{url}
@@
 \newcommand{\cp}{\circ}                 % composition
+\newcommand{\pc}{\fatsemi}                 % path composition
 \newcommand{\symi}{\sym\inv}
@@
 \newcommand{\proves}{\vdash}
+\newcommand{\ap}[2]{\mathit{ap}~#1~#2}
 \newcommand{\Rule}[4]{
@@
 \newcommand{\jdg}[3]{#2 \proves_{#1} #3}
+\newcommand{\adjoint}[1]{#1^{\dagger}}
 
@@
 %% acknowledge him
+
+Univalence says:
+\[ 
+(A \simeq B) \quad\simeq\quad (A \equiv B)
+\]
+where $\simeq$ is a strong form of isomorphism witnessed by functions from
+$A$ to $B$ and vice versa that are inverses and $\equiv$ is the propositional
+equality in type theoretic terms or a path (homotopy) in topological
+terms. The axiom relates two notions that are a priori different: functions
+and paths. However if we define functions from the outset as paths then there
+is no need for this axiom.
+
+So what would it take to have each function correspond to a path and
+higher-order functions correspond to paths between paths etc.? Since paths
+form an equivalence relation, the first immediate observation is that all
+functions must be reversible. Formally we need the following properties of
+paths to be satisfied by functions. (We write $\path$ for paths instead of
+$\equiv$.)
+
+\begin{itemize}
+\item For each $A$, there is function $\id : A \path A$,
+
+\item for each function $f: A \path B$, there exists a function $\adjoint{f}
+  : B \path A$, and
+
+\item for each functions $f : A \path B$ and $g : B \path C$, there exists a
+  function $g \cp f : A \path C$, such that:
+
+\begin{itemize}
+\item $f \path f \cp \id$;
+\item $f \path \id \cp f$;
+\item $\adjoint{f} \cp f \path \id$;
+\item $f \cp \, \adjoint{f} \path \id$;
+\item $\adjoint{(\adjoint{f})} \path f$;
+\item $f \cp (g \cp h) \path (f \cp g) \cp h$
+\end{itemize}
+\end{itemize}
+Similar equivalences hold are higher-levels. A good starting point for this
+notion of functions are the $\Pi$-combinators we previously developed. Each
+combinator in that setting had an \emph{adjoint} going in the other direction
+that acts as the inverse. Moreover if $f : A \path B$ then $A$ must be
+isomorphic to $B$ and $f$ is permutation. We use $!p$ for the inverse of a
+path.
+
+But first, let's examine the structure of a path $f : A \path B$. A natural
+representation for a path representing a function is a collection of paths
+(called \emph{fibers}) one for each $a : A$. So for example, if $A$ is the
+set $\{ x , y , z \}$ and $B$ is the set $\{ a , b , c \}$ (remember that $A$
+and $B$ must be isomorphic), and $f$ is the function mapping $x$ to $a$, $y$
+to $b$, and $z$ to $c$, then $f : A \path B$ is the collection of paths $f_x
+: x \path a$, $f_y : y \path b$, and $f_z : z \path c$.
+
+Assuming this representation of functions, let's see what properties are
+induced for these functions.
+
+The first property is that functions should preserve paths, i.e., if there is
+a path between $x$ and~$y$ then there should be a path between $f(x)$
+and~$f(y)$. From the figure below it is clear that:
+\[ 
+\ap{f}{p} = \, !f_x \pc p \pc f_y
+\]
+
+\begin{center}
+\begin{tikzpicture}
+  \draw (-3,0) ellipse (1.5cm and 3cm);
+  \draw (3,0) ellipse (1.5cm and 3cm);
+  \draw[fill] (-3,1.5) circle [radius=0.025];
+  \draw[fill] (-3,-1.5) circle [radius=0.025];
+  \node[above] at (-3,1.5) {$x$};
+  \node[below] at (-3,-1.5) {$y$};
+  \draw[fill] (3,1.5) circle [radius=0.025];
+  \draw[fill] (3,-1.5) circle [radius=0.025];
+  \node[above] at (3,1.5) {$f(x)$};
+  \node[below] at (3,-1.5) {$f(y)$};
+  \draw[->,cyan,thick] (-3,1.5) -- (-3,-1.5);
+  \node[left,cyan] at (-3,0) {$p$};
+  \draw[->,dashed,cyan,thick] (3,1.5) -- (3,-1.5);
+  \node[right,cyan] at (3,0) {$\mathit{ap}~f~p$};
+  \draw[->,red,thick] (-3,1.5) -- (3,1.5);
+  \draw[->,red,thick] (-3,-1.5) -- (3,-1.5);
+  \node[red,below] at (0,1.5) {$f_x$};
+  \node[red,below] at (0,-1.5) {$f_y$};
+\end{tikzpicture}
+\end{center}
+
+Let $p : x \path y$ and $q : y \path z$, then the following properties 
+hold:
+\begin{itemize}
+\item $\ap{f}{(p \pc q)} \path \ap{f}{p} \pc \ap{f}{q}$:
+\[\begin{array}{rcl}
+LHS &=& \, !f_x \pc (p \pc q) \pc f_z \\
+    &\path& ! f_x \pc p \pc f_y \pc \, ! f_y \pc q \pc f_z \\
+    &\path& (! f_x \pc p \pc f_y) \pc (! f_y \pc q \pc f_z) \\
+    &=& RHS
+\end{array}\]
+\item $\ap{f}{!p} \path \, !(\ap{f}{p})$
+\[\begin{array}{rcl}
+LHS &=& \, !f_y \pc (!p) \pc f_x \\
+    &\path& \, !(!f_x \pc p \pc f_y) \\
+    &=& RHS
+\end{array}\]
+\item $\ap{g}{(\ap{f}{p})} \path \ap{(g \cp f)}{p}$
+\[\begin{array}{rcl}
+LHS &=& \, !g_{f_x} \pc !f_x \pc p \pc f_y \pc g_{f_y} \\
+    &\path& \, ! (f_x \pc g_{f_x}) \pc p \pc (f_y \pc g_{f_y}) \\
+    &=& RHS
+\end{array}\]
+\item $\ap{\id}{p} \path p$
+\[\begin{array}{rcl}
+LHS &=& \, !\id_x \pc p \pc \id_y \\
+    &\path& p \\
+    &=& RHS
+\end{array}\]
+\end{itemize}
+
+A homotopy between functions $f$ and $g$ is a 2-path between each pair of
+fibers $f_x$ and $g_x$. These 2-paths are induced by the following diagram:
+
+\begin{center}
+\begin{tikzpicture}
+  \draw[fill] (-3,0) circle [radius=0.025];
+  \node[below] at (-3,0) {$x$};
+  \draw[fill] (3,1.5) circle [radius=0.025];
+  \node[above] at (3,1.5) {$a$};
+  \draw[fill] (3,-1.5) circle [radius=0.025];
+  \node[below] at (3,-1.5) {$b$};
+  \draw[->,cyan,thick] (-3,0) -- (3,1.5);
+  \draw[->,cyan,thick] (-3,0) -- (3,-1.5);
+  \draw[->,red,thick] (3,1.5) -- (3,-1.5);
+  \node[red,above] at (0,0.5) {$f_x$};
+  \node[red,below] at (0,-0.5) {$f_y$};
+  \draw[->,double,dashed,blue,thick] (1.5,1) to [out=225, in=135] (1.5,-1);
+  \node[blue,right] at (1.5,0) {$H_x$};
+\end{tikzpicture}
+\end{center}
+
+Here is something more interesting. Let $p : y \path z$ in $A$. Let $f$ and
+$g$ be functions from $A$ to $B$. Let $H$ be a 2-path between $f_y$ and
+$g_y$. If we are given a path $f(y) \path g(y)$ in B, then we can construct a
+2-path between $f_z$ and $g_z$. (This is Thm 2.11.3 in the book.) We do this
+going backwards along $f_z$, backwards along $p$, then follow $f_y$ and the
+path to $g(y)$, and backwards along $g_y$, forwards along $p$, and finally
+forwards along $g_z$:
+
+\begin{center}
+\begin{tikzpicture}
+  \draw[fill] (-3,0) circle [radius=0.025];
+  \node[below] at (-3,0) {$x$};
+  \draw[fill] (3,1.5) circle [radius=0.025];
+  \node[above] at (3,1.5) {$a$};
+  \draw[fill] (3,-1.5) circle [radius=0.025];
+  \node[below] at (3,-1.5) {$b$};
+  \draw[->,cyan,thick] (-3,0) -- (3,1.5);
+  \draw[->,cyan,thick] (-3,0) -- (3,-1.5);
+  \draw[->,red,thick] (3,1.5) -- (3,-1.5);
+  \node[red,above] at (0,0.5) {$f_x$};
+  \node[red,below] at (0,-0.5) {$f_y$};
+  \draw[->,double,dashed,blue,thick] (1.5,1) to [out=225, in=135] (1.5,-1);
+  \node[blue,right] at (1.5,0) {$H_x$};
+  \draw[fill] (-3,-5) circle [radius=0.025];
+  \node[below] at (-3,-5) {$x$};
+  \draw[fill] (3,-3.5) circle [radius=0.025];
+  \node[above] at (3,-3.5) {$a$};
+  \draw[fill] (3,-6.5) circle [radius=0.025];
+  \node[below] at (3,-6.5) {$b$};
+  \draw[->,cyan,thick] (-3,-5) -- (3,-3.5);
+  \draw[->,cyan,thick] (-3,-5) -- (3,-6.5);
+  \draw[->,red,thick] (3,-3.5) -- (3,-6.5);
+  \node[red,above] at (0,-4.5) {$f_x$};
+  \node[red,below] at (0,-5.5) {$f_y$};
+\end{tikzpicture}
+\end{center}
+
+Now the most interesting question is whether we can internalize functions as
+objects. Given $f$ a function from $A$ to $B$, what we can do below is to
+represent each fiber $f_x$ as an element of the type $(\frac{1}{A} \times
+B)$. As an example, let $A$ be the set $\{x,y,z\}$ and let $B$ be
+$\{a,b,c\}$, the type $\frac{1}{A}$ has one element $\frac{1}{3}$ which is
+one point that has three self loops labeled $x \path$, $y \path$, and $z
+\path$. A value $(\frac{1}{3},a)$ is a fiber of the function that produces
+$a$ as the second component indicates. The first component states that the
+$a$ could have come from $x$ along the path $x \path$, or from $y$ along the
+path $y \path$ or from $z$ along the path $z \path$.
+
+\newpage
 

### 2013-11-18 f.tex
playing with pointed types

@@
   Homotopy Type Theory and Fractional Types} \\[1.5cm]
-{\Large Jacques Carette${}^*$ \qquad Amr Sabry${}^{\dagger}$}\\[0.5cm]
-{\Large $({}^*)$~McMaster University \qquad $({}^\dagger)$~Indiana University}
 \vfill
@@
   Homotopy Type Theory and Fractional Types}
-\author{Jacques Carette \and Amr Sabry}
+\author{?}
 \date{\today}

### 2014-02-14 Univalence/tex/theorems.tex
added explanation of theorems in slightly more legible LaTeX form

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{verbatim}
\usepackage{amsthm,amssymb,amsmath,stmaryrd}
\usepackage{ifthen}
\usepackage{url}

\input{defs.tex}

\title{Univalence Theorems with $\Pi$}

\author{Zach Sparks}

%% \newcommand{\code}[1]{\texttt{#1}}

\begin{document}
\maketitle

We are trying to prove theorems that relate syntactic functions in our system
(ie, combinators) with semantic functions (ie, vectors). That is:

For all finite types $a$ and $b$ and all combinators $c : a \combarrow b$, there
exists some semantic function $f : \tonat{a} \to \tonat{b}$ such that for all $x
: a$, $\tonat{c\,x} \eq v\,\tonat{x}$.

For all finite types $a$ and $b$ and all semantic functions $f : \tonat{a} \to
\tonat{b}$, there exists some combinator $c : a \combarrow b$ such that for all
$i : \tonat{a}$, $c\,\fromnat{a}{i} \eq \fromnat{b}{v\,i}$.

In our case, we already know the transformations we are using (called $compile$
and $decompile$), so we can rewrite these theorems as follows:

For all finite types $a$ and $b$, combinators $c : a \combarrow b$, and all $x :
a$, $\tonat{c\,x} \eq compile\,c\,\tonat{x}$.

For all finite types $a$ and $b$, all semantic functions $f : \tonat{a} \to
\tonat{b}$, and all $i : \tonat{a}$, $decompile\,v\,\fromnat{a}{i} \eq
\fromnat{b}{v\,i}$.

As an intermediate step, we are proving a slightly different theorem about
combinators over \emph{normalized types}, since these are closer to the vector
representation we are compiling to. Thus, the process looks something like:

\begin{verbatim}
combinator <===> combinator over normalized types <===> vector
\end{verbatim}

Right now, we are working on the arrow on the right; the arrow on the left
should be somewhat easier.

\end{document}

### 2014-05-06 Conway.hs
Conway.hs

@@ -1,0 +1,255 @@
+module Conway where
+
+import Data.List
+
+------------------------------------------------------------------------------
+-- Definition; a few constants; several show functions
+
+data Game = Game [Game] [Game]
+  deriving (Eq, Show)
+
+leftOptions :: Game -> [Game]
+leftOptions (Game gls _) = gls
+
+rightOptions :: Game -> [Game]
+rightOptions (Game _ grs) = grs
+
+zeroG, oneG, negoneG, starG :: Game
+zeroG   = Game []      []
+oneG    = Game [zeroG] []
+negoneG = Game []      [zeroG]
+starG   = Game [zeroG] [zeroG]
+
+nat2Game :: Int -> Game
+nat2Game 0 = zeroG
+nat2Game n = Game [nat2Game (n-1)] []
+
+nnat2Game :: Int -> Game
+nnat2Game 0 = zeroG
+nnat2Game n = Game [] [nnat2Game (n+1)]
+
+--
+
+showG :: Game -> String
+showG (Game [] []) = "0"
+showG (Game [Game [] []] []) = "1"
+showG (Game [] [Game [] []]) = "-1"
+showG (Game [Game [] []] [Game [] []]) = "*"
+showG (Game gls grs) = 
+  "<[" ++ concatMap showG gls ++ "] , [" ++ concatMap showG grs ++ "]>"
+
+showT :: String -> Game -> String
+showT _ (Game [] []) = "(empty-tree)"
+showT s (Game [g] []) = 
+  "(tree " ++ s ++ " " ++ showT "'*" g ++ " (empty-tree))"
+showT s (Game [] [g]) = 
+  "(tree " ++ s ++ " (empty-tree) " ++ showT "'*" g ++ ")"
+showT s (Game [g1] [g2]) = 
+  "(tree " ++ s ++ " " ++ showT "'*" g1 ++ showT "'*" g2 ++ ")"
+showT s (Game [g1,g2] [g3]) = 
+  "(tree " ++ s ++ " " ++ 
+           showT "#\\space" (Game [g1] [g2]) ++ showT "'*" g3 ++ ")"
+showT s (Game [g1] [g2,g3]) = 
+  "(tree " ++ s ++ " " ++ 
+           showT "'*" g1 ++ showT "#\\space" (Game [g2] [g3]) ++ ")"
+
+------------------------------------------------------------------------------
+-- addition and subtraction
+
[…]

### 2014-05-06 Neg1.hs
neg1

@@ -13,8 +13,9 @@ http://www.kurims.kyoto-u.ac.jp/~hassei/papers/tmcc.pdf
 
 module Neg1 where
 
+import Data.Maybe hiding (mapMaybe)
 import qualified Prelude
-import Prelude (Either(..), error, ($), (.), id)
+import Prelude (Either(..), error, ($), (.), id, Eq, Show)
 
 -----------------------------------------------------------------------
 -- Some very abstract kit that will allow lots of different instances
@@ -448,40 +449,55 @@ instance Pi R I where
 -----------------------------------------------------------------------
 -- Int (or G) construction
 
--- Objects in the G category are pairs of objects 
+-- Objects in the G category are nested pairs of objects 
 
-class GT p where
-  type Pos p      :: *  -- to access the components
-  type Neg p      :: *  
-  type ZeroG      :: *  -- the new structure 0,1,+,* 
-  type OneG       :: *
-  type PlusG p q  :: *
-  type TimesG p q :: *
-  type DualG p    :: *  -- as a bonus we get DualG (unary negation) and
-  type LolliG p q :: *  -- linear functions
+data GO a = Pair (Maybe (GO a)) (Maybe (GO a))
+  deriving (Eq,Show)
 
--- think of a :- b as "a-b"
-data a :- b = a :- b
+zeroG = Pair Nothing Nothing
+oneG = Pair (Just zeroG) Nothing
+moneG = Pair Nothing (Just zeroG)
 
-data Quadrants a b c d = Quadrants a b c d -- ++, +-, -+, --
+leftG :: GO a -> Maybe (GO a)
+leftG (Pair l _) = l
 
-instance GT (ap :- am) where
-  type Pos (ap :- am) = ap
-  type Neg (ap :- am) = am
-  type ZeroG = Void :- Void
-  type OneG = () :- Void
-  type PlusG  (ap :- am) (bp :- bm) = (Either ap bp) :- (Either am bm)
-  type TimesG (ap :- am) (bp :- bm) = Quadrants (ap,bp) (ap,bm) (am,bp) (am,bm)
-  type DualG  (ap :- am) = am :- ap
[…]

### 2014-05-06 negatives.tex
negatives.tex

\documentclass[authoryear,preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{listings} 
\usepackage{stmaryrd}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{courier}
\usepackage{thmtools}
\usepackage{bbold}

%%

\newcommand{\game}[2]{\langle #1 ~|~ #2 \rangle}
\newcommand{\zerog}{$\mathbb{0}$}
\newcommand{\oneg}{$\mathbb{1}$}
\newcommand{\twog}{$\mathbb{2}$}
\newcommand{\threeg}{$\mathbb{3}$}
\newcommand{\fourg}{$\mathbb{4}$}

%%

\begin{document}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\newcommand{\alt}{~|~}
\lstnewenvironment{code}{\lstset{basicstyle={\sffamily\footnotesize}}}{}

\lstset{frame=none,
         language=Haskell,
         basicstyle=\sffamily, 
         numberstyle=\tiny,
         numbersep=5pt,
         tabsize=2,    
         extendedchars=true,
         breaklines=true,   
         breakautoindent=true,
         keywordstyle=\color{black},
         captionpos=b,
         stringstyle=\color{black}\ttfamily,
         showspaces=false,  
         showtabs=false,    
         framexleftmargin=2em,
         framexbottommargin=1ex,
         showstringspaces=false
         basicstyle=\sffamily,
         columns=[l]flexible,
         flexiblecolumns=true,
         aboveskip=\smallskipamount,
         belowskip=\smallskipamount,
         lineskip=-1pt,
         xleftmargin=1em,
         escapeinside={/+}{+/},
         keywords=[1]{Monad,Just,Nothing,type,data,right,left,id,where,do,
                     if,then,else,let,in},
         literate={+*}{{$+*$}}1 {+}{{$\;+\;$}}1 {/}{{$/$}}1 {*}{{$\;*\;$}}1
           {=}{{$=\ $}}1 {/=}{{$\not=$}}2
           {[]}{$[\;]$}2
           {<}{{$<$}}1 {>}{{$\rangle$}}1 
           {++}{{$+\!\!\!+\;$}}1 {::}{{$:\mkern -2.5mu:\;$}}1
           {:+:}{{$:\mkern -5mu+\mkern -5mu:\;$}}2
           {$}{{\texttt{\$}\hspace{0.5em}}}1
           {`}{$^\backprime$}1
           {==}{{$=\!=\;$}}2
           {===}{{$\equiv\;$}}2
           {->}{{$\rightarrow\;$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
           {<=}{{$\leq$}}2 {=>}{{$\Rightarrow\;$}}2
           {<<}{{$\ll$}}2 {>>}{{$\gg\;$}}2
           {>>>}{{$\ggg\;$}}3 {<<<}{{$\lll\;$}}3
           {>>=}{{$\gg\mkern -2.5mu=\;$}}3
           {=<<}{{$=\mkern -2.5mu\ll\;$}}3
           {<|}{$\lhd\;$}2
           {<||}{$\unlhd\;$}2
           {\ ||\ }{$\|$}1
           {\\}{$\lambda$}1
           {:>}{{$\rhd$}}2
           {||>}{{$\unrhd$}}2
           {_}{{$\_$}}1
           {_B}{{$_b$}}2
           {forall}{{$\forall$}}1
}

\lstset{postbreak=\raisebox{0ex}[0ex][0ex]
        {\ensuremath{\hookrightarrow}}}
\lstset{breaklines=true, breakatwhitespace=true}
\lstset{numbers=none, numbersep=5pt, stepnumber=2, numberstyle=\scriptsize}
\lstset{rangeprefix=/*!\ , rangesuffix=\ !*\/, includerangemarker=false}

%% double-blind reviewing...
\title{Negative Types}
\authorinfo{}{}{}
\maketitle

\begin{abstract}
\ldots
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conway Numbers}

A Conway number is a game consisting of left and right options where each
option leads to another game. In Haskell, one might define the datatype of
Conway numbers as:
\begin{code}
data ConwayNumber = CN [ConwayNumber] [ConwayNumber]
\end{code}
The simplest Conway number is \lstinline|CN [] []| with empty left and right
options. We call this number \zerog. Once we have defined \zerog, we can also
define the following:
\begin{code}
/+ \oneg +/  = CN  [ /+ \zerog +/ ]  []
/+ \twog +/  = CN  [ /+ \oneg +/ ]  []
/+ \threeg +/  = CN  [ /+ \twog +/ ]  []
\end{code}
and so on. Intuitively, the number $\mathbb{n}$ represents a game in which
the left player has an $n$-move advantage over the right player. Dually, we
can define the following numbers where the left player has $n$-move
\emph{disadvantage} over the right player:
\begin{code}
/+ -\oneg +/  = CN  [] [ /+ \zerog +/ ]
/+ -\twog +/  = CN  [] [ /+ \oneg +/ ]
/+ -\threeg +/  = CN  [] [ /+ \twog +/ ]
\end{code}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{abbrvnat}
\softraggedright
\bibliography{cites}

\end{document}

### 2014-05-06 negatives.tex
intro to numbers

@@
            {++}{{$+\!\!\!+\;$}}1 {::}{{$:\mkern -2.5mu:\;$}}1
+           {&&}{{$\&\!\!\!\&$}}2
+           {:=:}{{$:\mkern -2mu=\mkern -2mu:\;$}}2
            {:+:}{{$:\mkern -5mu+\mkern -5mu:\;$}}2
+           {:-:}{{$:\mkern -5mu-\mkern -5mu:\;$}}2
+           {:*:}{{$:\mkern -5mu*\mkern -5mu:\;$}}2
            {$}{{\texttt{\$}\hspace{0.5em}}}1
@@
 A Conway number is a game consisting of left and right options where each
-option leads to another game. In Haskell, one might define the datatype of
-Conway numbers as:
+option models a move to another game. Players alternate taking options and
+the player with no available options loses. In Haskell, one might define the
+datatype of Conway numbers as:
 \begin{code}
@@
 The simplest Conway number is \lstinline|CN [] []| with empty left and right
-options. We call this number \zerog. Once we have defined \zerog, we can also
-define the following:
+options. We call this number \zerog: 
+\begin{code}
+/+ \zerog +/  = CN  [] []
+\end{code}
+Once we have defined \zerog, we can also define the following:
 \begin{code}
@@
 /+ -\oneg +/  = CN  [] [ /+ \zerog +/ ]
-/+ -\twog +/  = CN  [] [ /+ \oneg +/ ]
-/+ -\threeg +/  = CN  [] [ /+ \twog +/ ]
+/+ -\twog +/  = CN  [] [ /+ -\oneg +/ ]
+/+ -\threeg +/  = CN  [] [ /+ -\twog +/ ]
+\end{code}
+More generally, the \emph{unary negation} of a number is defined as follows:
+\begin{code}
+neg :: ConwayNumber -> ConwayNumber
+neg (CN xls xrs) = 
+  CN [ neg xr | xr <- xrs ] [ neg xl | xl <- xls ]
+\end{code}
+
+Conway numbers also come equipped with addition and multiplication operations
+defined as follows:
+
+\begin{code}
+(:+:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x@(CN xls xrs) :+: y@(CN yls yrs) = 
+  CN 
+    ([ xl :+: y | xl <- xls ] `union`
+     [ x :+: yl | yl <- yls ])
+    ([ xr :+: y | xr <- xrs ] `union`
+     [ x :+: yr | yr <- yrs ])
+
+(:-:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x :-: y = x :+: (neg y)
+
+(:*:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x@(CN xls xrs) :*: y@(CN yls yrs) = 
+  CN 
+    ([ (xl :*: y) :+: (x :*: yl) :-: (xl :*: yl)
+     | xl <- xls, yl <- yls] `union`
+     [ (xr :*: y) :+: (x :*: yr) :-: (xr :*: yr)
+     | xr <- xrs, yr <- yrs])
+    ([ (xl :*: y) :+: (x :*: yr) :-: (xl :*: yr)
+     | xl <- xls, yr <- yrs] `union`
+     [ (xr :*: y) :+: (x :*: yl) :-: (xr :*: yl)
+     | xr <- xrs, yl <- yls])
+\end{code}
+
+Conway numbers form a \emph{ring} under the semantic notion of equality
+\lstinline$:=:$ defined below:
+\begin{code}
+geq/+\zerog+/, leq/+\zerog+/, eq/+\zerog+/ :: ConwayNumber -> Bool
+geq/+\zerog+/ (CN _ xrs)  = not $ or (map leq/+\zerog+/ xrs) 
+leq/+\zerog+/ (CN xls _)  = not $ or (map geq/+\zerog+/ xls) 
+eq/+\zerog+/ x            = geq/+\zerog+/ x && leq/+\zerog+/ x
+
+(:=:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x :=: y = eq/+\zerog+/ (x :-: y)
 \end{code}

### 2014-05-06 Conway.hs

@@ -177,12 +177,12 @@ introduceReversibleShortcuts g@(Game gls grs) =
         helperL g (gl:gls) = 
           case find (`leqG` g) (rightOptions gl) of 
             Nothing  -> gl : helperL g gls
-            Just glr -> leftOptions glr ++ helperL g gls
+            Just glr -> leftOptions glr `union` helperL g gls
         helperR g [] = []
         helperR g (gr:grs) = 
           case find (`geqG` g) (leftOptions gr) of 
             Nothing  -> gr : helperR g grs
-            Just grl -> rightOptions grl ++ helperR g grs
+            Just grl -> rightOptions grl `union` helperR g grs
 
 ------------------------------------------------------------------------------
 -- Number games
@@ -193,6 +193,7 @@ isNumberGame (Game gls grs) =
   and [ gl `lessG` gr | gl <- gls, gr <- grs ]
 
 halfG         = Game [zeroG] [oneG]  -- 1/2 + 1/2 is indeed 1 
+quarterG = Game [zeroG] [halfG]
 
 -- If we have three player games, we can presumably get multiples of 3
 -- easily. If we have five player games, we can presumably get multiples of 5
@@ -205,11 +206,11 @@ timesG :: Game -> Game -> Game
 x@(Game xls xrs) `timesG` y@(Game yls yrs) = 
   Game 
     ([ (xl `timesG` y) `plusG` (x `timesG` yl) `minusG` (xl `timesG` yl)
-     | xl <- xls, yl <- yls] ++ 
+     | xl <- xls, yl <- yls] `union`
      [ (xr `timesG` y) `plusG` (x `timesG` yr) `minusG` (xr `timesG` yr)
      | xr <- xrs, yr <- yrs])
     ([ (xl `timesG` y) `plusG` (x `timesG` yr) `minusG` (xl `timesG` yr)
-     | xl <- xls, yr <- yrs] ++ 
+     | xl <- xls, yr <- yrs] `union`
      [ (xr `timesG` y) `plusG` (x `timesG` yl) `minusG` (xr `timesG` yl)
      | xr <- xrs, yl <- yls])
 

### 2014-05-06 negatives.tex
addition

@@
 \newcommand{\fourg}{$\mathbb{4}$}
+\newcommand{\fiveg}{$\mathbb{5}$}
 
@@
 \end{code}
+Conway numbers also come equipped with addition and multiplication operations
+defined as follows.
 
-Conway numbers also come equipped with addition and multiplication operations
-defined as follows:
-
+\noindent\textbf{Addition and subtraction.} Addition of two games intuitively
+gives the player the choice of selecting an option from either
+game. Substraction is simply the addition of the dual of a game. Formally:
 \begin{code}
@@
 x :-: y = x :+: (neg y) 
+\end{code}
+It is easy to check that adding two positive numbers like \twog\ and
+\threeg\ gives \fiveg\ as desired, and similarly for two negative
+numbers. When mixing positive and negative numbers, e.g., adding \threeg\ and
+$-$\twog, the result appears much more complicated:
+\begin{code}
+CN [CN  [CN  [-/+\twog+/] 
+             [CN [-/+\oneg+/] [/+\oneg+/]]] 
+        [CN  [CN [-/+\oneg+/] [/+\oneg+/]] 
+             [/+\twog+/]]] 
+   [CN  [CN  [CN [-/+\oneg+/] [/+\oneg+/]] 
+             [/+\twog+/]] 
+        [/+\threeg+/]]
+\end{code}
+There is however an intuitive way to understand why the game above is
+equivalent to the game \oneg\ (which is formalized below). Consider the game
+\lstinline{CN [-1] [1]}: it is equivalent to the game \zerog.
 
+\noindent\textbf{Multiplication.}
+
+\begin{code}
 (:*:) :: ConwayNumber -> ConwayNumber -> ConwayNumber

### 2014-05-06 negatives.tex
addition and comparison

--- 2014-05-06 negatives.tex
@@
 
-\newcommand{\game}[2]{\langle #1 ~|~ #2 \rangle}
-\newcommand{\zerog}{$\mathbb{0}$}
-\newcommand{\oneg}{$\mathbb{1}$}
-\newcommand{\twog}{$\mathbb{2}$}
-\newcommand{\threeg}{$\mathbb{3}$}
-\newcommand{\fourg}{$\mathbb{4}$}
+\newcommand{\geqz}{\geq\zerog}
+\newcommand{\leqz}{\leq\zerog}
+\newcommand{\zerog}{\mathbb{0}}
+\newcommand{\oneg}{\mathbb{1}}
+\newcommand{\twog}{\mathbb{2}}
+\newcommand{\threeg}{\mathbb{3}}
+\newcommand{\fourg}{\mathbb{4}}
+\newcommand{\fiveg}{\mathbb{5}}
 
@@
                      if,then,else,let,in},
-         literate={+*}{{$+*$}}1 {+}{{$\;+\;$}}1 {/}{{$/$}}1 {*}{{$\;*\;$}}1
-           {=}{{$=\ $}}1 {/=}{{$\not=$}}2
+         literate=
+           {+}{{$\;+\;$}}1 
+           {/}{{$/$}}1 
+           {*}{{$\;*\;$}}1
+           {=}{{$=\ $}}1 
+           {/=}{{$\not=$}}1
            {[]}{$[\;]$}2
-           {<}{{$<$}}1 {>}{{$\rangle$}}1 
-           {++}{{$+\!\!\!+\;$}}1 {::}{{$:\mkern -2.5mu:\;$}}1
-           {:+:}{{$:\mkern -5mu+\mkern -5mu:\;$}}2
+           {<}{{$<$}}1 
+           {>}{{$>$}}1 
+           {++}{{$+\!\!\!+\;$}}1 
+           {::}{{$:\mkern -2.5mu:\;$}}1
+           {&&}{{$\&\!\!\!\&$}}2
+           {:=:}{{$:\mkern -2mu=\mkern -2mu:\;$}}3
+           {:+:}{{$:\mkern -5mu+\mkern -5mu:\;$}}3
+           {:-:}{{$:\mkern -5mu-\mkern -5mu:\;$}}3
+           {:*:}{{$:\mkern -5mu*\mkern -5mu:\;$}}3
            {$}{{\texttt{\$}\hspace{0.5em}}}1
@@
            {===}{{$\equiv\;$}}2
-           {->}{{$\rightarrow\;$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
-           {<=}{{$\leq$}}2 {=>}{{$\Rightarrow\;$}}2
-           {<<}{{$\ll$}}2 {>>}{{$\gg\;$}}2
-           {>>>}{{$\ggg\;$}}3 {<<<}{{$\lll\;$}}3
+           {->}{{$\rightarrow\;$}}2 
+           {>=}{{$\geq$}}2 
+           {<=}{{$\leq$}}2 
+           {>=0}{{$\geq_\zerog\;$}}2 
+           {<=0}{{$\leq_\zerog\;$}}2 
+           {==0}{{$=_\zerog\;$}}2 
+           {>0}{{$>_\zerog\;$}}2 
+           {<0}{{$<_\zerog\;$}}2 
+           {<-}{{$\leftarrow$}}2
+           {=>}{{$\Rightarrow\;$}}2
+           {<<}{{$\ll$}}2 
+           {>>}{{$\gg\;$}}2
+           {>>>}{{$\ggg\;$}}3 
+           {<<<}{{$\lll\;$}}3
            {>>=}{{$\gg\mkern -2.5mu=\;$}}3
@@
 A Conway number is a game consisting of left and right options where each
-option leads to another game. In Haskell, one might define the datatype of
-Conway numbers as:
+option models a move to another game. There is another technical condition
+given below that constrains every left option to be ``less than'' every right
+option but that can be ignored for the moment. Players alternate taking
+options and the player with no available options loses. In Haskell, ignoring
+the additional technical condition, one might define the datatype of Conway
+numbers as:
 \begin{code}
@@
 The simplest Conway number is \lstinline|CN [] []| with empty left and right
-options. We call this number \zerog. Once we have defined \zerog, we can also
-define the following:
+options. We call this number $\zerog$: 
 \begin{code}
-/+ \oneg +/  = CN  [ /+ \zerog +/ ]  []
-/+ \twog +/  = CN  [ /+ \oneg +/ ]  []
-/+ \threeg +/  = CN  [ /+ \twog +/ ]  []
+/+ $\zerog$ +/  = CN  [] []
+\end{code}
+Once we have defined $\zerog$, we can also define the following:
+\begin{code}
+/+ $\oneg$ +/  = CN  [ /+ $\zerog$ +/ ]  []
+/+ $\twog$ +/  = CN  [ /+ $\oneg$ +/ ]  []
+/+ $\threeg$ +/  = CN  [ /+ $\twog$ +/ ]  []
 \end{code}
@@
 \begin{code}
-/+ -\oneg +/  = CN  [] [ /+ \zerog +/ ]
-/+ -\twog +/  = CN  [] [ /+ \oneg +/ ]
-/+ -\threeg +/  = CN  [] [ /+ \twog +/ ]
+/+ -$\oneg$ +/  = CN  [] [ /+ $\zerog$ +/ ]
+/+ -$\twog$ +/  = CN  [] [ /+ -$\oneg$ +/ ]
+/+ -$\threeg$ +/  = CN  [] [ /+ -$\twog$ +/ ]
+\end{code}
+More generally, the \emph{unary negation} of a number or its \emph{dual} is
+defined as follows:
+\begin{code}
+neg :: ConwayNumber -> ConwayNumber
+neg (CN xls xrs) = 
+  CN [ neg xr | xr <- xrs ] [ neg xl | xl <- xls ]
+\end{code}
+
+\textbf{Comparing numbers.} There are three outcomes to games represented by
+Conway numbers $\mathbb{n}$:
+\begin{itemize}
+\item $\mathbb{n} > \zerog$ which indicates that the \emph{left} player can
+  enforce a win, no matter who starts;
+\item $\mathbb{n} < \zerog$ which indicates that the \emph{right} player can
+  enforce a win, no matter who starts;
+\item $\mathbb{n} = \zerog$ which indicates that the \emph{second}
+  player can enforce a win, no matter who.
+\end{itemize}
+For example, we have $\threeg > \zerog$ because the right player would have
+no options (i.e., would lose) on the first or second turn depending on who
+starts. Similarly, we have $-\threeg < \zerog$ because the left player would
+have no options on on the first or second turn depending on who starts. For
+$\zerog$, it is evident that the first player to take a turn must lose, i.e.,
+that the second player wins. More interestingly, consider the game
+$\textsf{CN}~[-\oneg]~[\oneg]$ which we claim is $= \zerog$. Indeed, if the
+left player starts, the only option is to move to the game $-\oneg$ which the
+right player wins. If however the right player moves first, the only option
+is to move to the game $\oneg$ which the left player wins. In both cases, the
+second player wins. To ensure that these three outcomes are the only possible
+outcomes, Conway numbers have an additional restriction: for every left
+option $x$ and right option $y$, we require that $x < y$. If that restriction
+is lifted we could construct games such that $\textsf{CN}~[\zerog]~[\zerog]$
+for which none of the relations $> \zerog$, $< \zerog$, or $= \zerog$
+holds. Instead for this game, the first player always wins, which means that
+the left player cannot enforce a win (in case the right player starts) and
+vice-versa.
+
+The formal definitions of the comparison operators are:
+\begin{code}
+>=0 (CN _ xrs)  = not $ or (map <=0 xrs)
+<=0 (CN xls _)  = not $ or (map >=0 xls) 
+==0 x           = >=0 x && <=0 x
+<0 x            = <=0 x && not (>=0 x)
+>0 x            = >=0 x && not (<=0 x)
+\end{code}
+Only $\geq_\zerog$ and $\leq_\zerog$ are fundamental. The predicate
+$\geq_\zerog$ means that the left player can win as second player and
+$\leq_\zerog$ means that the right player can win as second player. If the
+left player can win as second player, it means that the right player has no
+good opening move, i.e., that there is no right option that is $\leq_\zerog$.
+
+Conway numbers also come equipped with addition, subtraction, and
+multiplication operations defined as follows.
+
+\noindent\textbf{Addition and subtraction.} Addition of two games intuitively
+gives each player the choice of selecting an option from either
+game. Subtraction is simply the addition of the dual of a game. Formally:
+\begin{code}
+(:+:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x@(CN xls xrs) :+: y@(CN yls yrs) = 
+  CN 
+    ([ xl :+: y | xl <- xls ] `union`
+     [ x :+: yl | yl <- yls ])
+    ([ xr :+: y | xr <- xrs ] `union`
+     [ x :+: yr | yr <- yrs ])
+
+(:-:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x :-: y = x :+: (neg y) 
+\end{code}
+It is easy to check that $\zerog$ is the unit of addition and that addition
+is commutative. Furthermore, adding two positive numbers like~$\twog$
+and~$\threeg$ gives $\fiveg$ as desired, and similarly for two negative
+numbers. When mixing positive and negative numbers, e.g., adding $\threeg$
+and $-\twog$, the result appears much more complicated:
+\begin{code}
+CN [CN  [CN  [-/+ $\twog$ +/] 
+             [CN [-/+ $\oneg$ +/] [/+ $\oneg$ +/]]]
+        [CN  [CN [-/+ $\oneg$ +/] [/+ $\oneg$ +/]] 
+             [/+ $\twog$ +/]]] 
+   [CN  [CN  [CN [-/+ $\oneg$ +/] [/+ $\oneg$ +/]] 
+             [/+ $\twog$ +/]] 
+        [/+ $\threeg$ +/]]
+\end{code}
+We will formalize the sense in which the game above is equivalent to the game
+$\oneg$, but for now we can intuitively reason as follows. We have already
+seen that the game $\textsf{CN}~[-\oneg]~[\oneg]$ is equivalent to $\zerog$
+which simplifies the above to:
+\begin{code}
+CN [CN  [CN  [-/+ $\twog$ +/] [/+ $\zerog$ +/]]
+        [CN  [/+ $\zerog$ +/] [/+ $\twog$ +/]]]
+   [CN  [CN  [/+ $\zerog$ +/] [/+ $\twog$ +/]]
+        [/+ $\threeg$ +/]]
+\end{code}
+Now consider the game $\textsf{CN}~[\zerog]~[\twog]$. If the left player
+starts, the game proceeds to $\zerog$ and the right player loses. If the
+right player starts, the game proceeds to $\twog$ which is an even stronger
+position for the left player. In other words, the right option is useless and
+the right player always loses: the game is equivalent to
+$\textsf{CN}~[\zerog]~[\;]$, i.e., to~$\oneg$. By a similar reasoning,
+$\textsf{CN}~[-\twog]~[\zerog]$ simplifies to $-\oneg$, and the whole
+expression becomes:
+\begin{code}
+CN [CN  [/+ -$\oneg$ +/] [/+ $\oneg$ +/]]
+   [CN  [/+ $\oneg$ +/] [/+ $\threeg$ +/]]
+\end{code}
+The left option is equivalent to $\zerog$ and the right one is equivalent to
+$\twog$, and the entire expression is equivalent to $\oneg$.
+
+\noindent\textbf{Multiplication.} The most complicated operation on Conway
+numbers is multiplication. It is defined as follows:
+\begin{code}
+(:*:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x@(CN xls xrs) :*: y@(CN yls yrs) = 
+  CN 
+    ([ (xl :*: y) :+: (x :*: yl) :-: (xl :*: yl)
+     | xl <- xls, yl <- yls] `union`
+     [ (xr :*: y) :+: (x :*: yr) :-: (xr :*: yr)
+     | xr <- xrs, yr <- yrs])
+    ([ (xl :*: y) :+: (x :*: yr) :-: (xl :*: yr)
+     | xl <- xls, yr <- yrs] `union`
+     [ (xr :*: y) :+: (x :*: yl) :-: (xr :*: yl)
+     | xr <- xrs, yl <- yls])
+\end{code}
+
+Conway numbers form a \emph{ring} under the semantic notion of equality
+\lstinline$:=:$ defined below:
+\begin{code}
+(:=:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
+x :=: y = ==0 (x :-: y)
 \end{code}

### 2014-05-06 Conway.hs
remove 2 partial functions and replace by one total

@@ -20,13 +20,10 @@ oneG    = Game [zeroG] []
 negoneG = Game []      [zeroG]
 starG   = Game [zeroG] [zeroG]
 
-nat2Game :: Int -> Game
-nat2Game 0 = zeroG
-nat2Game n = Game [nat2Game (n-1)] []
-
-nnat2Game :: Int -> Game
-nnat2Game 0 = zeroG
-nnat2Game n = Game [] [nnat2Game (n+1)]
+int2Game :: Int -> Game
+int2Game n | n < 0 = Game []               [int2Game (n+1)]
+int2Game n | n > 0 = Game [int2Game (n-1)] []
+int2Game n | otherwise = zeroG
 
 --
 

### 2014-05-07 Conway.hs
added comment about strategies and moves

@@ -226,6 +226,37 @@ divG :: Game -> Game -> Game
 x `divG` y = x `timesG` (recipG y)
 
 ------------------------------------------------------------------------------
+-- A morphism between games < xls | xrs > and < yls | yrs > 
+-- is a game < yls + xrs | xls + yrs > 
+
+-- Joyal: focus on strategies which are sequences of moves
+
+{-- 
+
+A strategy is a rule that tells us how to choose a move. 
+
+Given a game < x1=< z1, z2 | w1, w2, w3 >, x2, x3 | y1, y2 > 
+
+A strategy for left consists of choosing one of x1, x2, or x3 and then
+choosing a new strategy for each possible response from right. Say left chose
+x1 which is < z1, z2 | w1, w2, w3 >. Then the rest of left's strategy
+consists of the product of three strategies (Sw1, Sw2, Sw3) which are left's
+responses to each possible move from right. If we get to a game where right
+has no moves, then left's response is (). Let's assume that w1=<a1,a2|>,
+w2=<b|>, and that w3=<|c>. So we can say a strategy for left in the above
+game is:
+
+  ((S(a1) + S(a2)),b,0) + S(x2) + S(x3)
+
+The 0 means that we have lost in this branch.
+
+So instead of sets of games in left and right positions, we should think of
+sums and pairs of games!!!
+
+--}
+
+
+------------------------------------------------------------------------------
 -- Pi
 
 {--

### 2014-05-07 Conway.hs
structured games

@@ -1,3 +1,5 @@
+{-# LANGUAGE TypeFamilies #-}
+
 module Conway where
 
 import Data.List
@@ -5,9 +7,79 @@ import Data.List
 ------------------------------------------------------------------------------
 -- Definition; a few constants; several show functions
 
+{--
+Conventional definition:
+
 data Game = Game [Game] [Game]
   deriving (Eq, Show)
+--}
 
+data Game a b = Game a b
+
+data Void
+
+type ZERO  = Game Void () 
+
+type ONE   = Game ZERO ()
+type TWO   = Game ONE  ()
+type THREE = Game TWO  ()
+
+type NEG_ONE   = Game Void ZERO
+type NEG_TWO   = Game Void NEG_ONE
+type NEG_THREE = Game Void NEG_TWO
+
+zeroG :: ZERO
+zeroG = Game undefined ()
+
+oneG :: ONE
+oneG = Game zeroG ()
+
+twoG :: TWO
+twoG = Game oneG ()
+
+threeG :: THREE
+threeG = Game twoG ()
+
+neg_oneG :: NEG_ONE
+neg_oneG = Game undefined zeroG
+
+neg_twoG :: NEG_TWO
+neg_twoG = Game undefined neg_oneG
+
+neg_threeG :: NEG_THREE
+neg_threeG = Game undefined neg_twoG
+
+--
+
+class G a where
+  type Neg a :: *
+
+instance G Void where
+  type Neg Void = ()
+
+instance G () where
+  type Neg () = Void
+
+negG :: Game a b -> Game (Neg a) (Neg b) 
+negG (Game x y) = Game (negG x) (negG y) 
+
+{--
+plusG :: Game -> Game -> Game
+g@(Game gls grs) `plusG` h@(Game hls hrs) = 
+  Game 
+    ((map (`plusG` h) gls) `union` (map (g `plusG`) hls))
+    ((map (`plusG` h) grs) `union` (map (g `plusG`) hrs))
+
+negG :: Game -> Game
+negG (Game gls grs) = Game (map negG grs) (map negG gls) 
+--}
+
+
+
+
+
+
+{--
 leftOptions :: Game -> [Game]
 leftOptions (Game gls _) = gls
 
@@ -282,3 +354,4 @@ data Combinator =
 
 -- data Val = ???
 
+--}

### 2014-05-08 Conway.hs
musing

@@ -12,74 +12,160 @@ Conventional definition:
 
 data Game = Game [Game] [Game]
   deriving (Eq, Show)
+
+Our definition:
+
+ GL ::= < S | P >
+ GR ::= < P | S >
+
+  S ::= 0 | GR | S + S
+
+  P ::= 1 | GL | P * P
+
 --}
 
-data Game a b = Game a b
+data GL = GL S P
+data GR = GR P S
+data S  = Lose | S GR | Choice S S 
+data P  = Win  | P GL | Opponent P P 
 
-data Void
+dualL :: GL -> GR
+dualL (GL s p) = GR (dualS s) (dualP p)
 
-type ZERO  = Game Void () 
+dualR :: GR -> GL
+dualR (GR p s) = GL (dualP p) (dualS s)
 
-type ONE   = Game ZERO ()
-type TWO   = Game ONE  ()
-type THREE = Game TWO  ()
+dualP :: P -> S
+dualP Win = Lose
+dualP (P gl) = S (dualL gl)
+dualP (Opponent p1 p2) = Choice (dualP p1) (dualP p2)
 
-type NEG_ONE   = Game Void ZERO
-type NEG_TWO   = Game Void NEG_ONE
-type NEG_THREE = Game Void NEG_TWO
-
-zeroG :: ZERO
-zeroG = Game undefined ()
-
-oneG :: ONE
-oneG = Game zeroG ()
-
-twoG :: TWO
-twoG = Game oneG ()
-
-threeG :: THREE
-threeG = Game twoG ()
-
-neg_oneG :: NEG_ONE
-neg_oneG = Game undefined zeroG
-
-neg_twoG :: NEG_TWO
-neg_twoG = Game undefined neg_oneG
-
-neg_threeG :: NEG_THREE
-neg_threeG = Game undefined neg_twoG
+dualS :: S -> P
+dualS Lose = Win
+dualS (S gr) = P (dualR gr)
+dualS (Choice s1 s2) = Opponent (dualS s1) (dualS s2)
 
 --
 
-class G a where
-  type Neg a :: *
+zeroGL, oneGL, twoGL, threeGL :: GL
+zeroGL  = GL Lose Win
+oneGL   = GL (S (dualL zeroGL)) Win
+twoGL   = GL (S (dualL oneGL)) Win
+threeGL = GL (S (dualL twoGL)) Win
 
-instance G Void where
-  type Neg Void = ()
[…]

### 2014-05-08 negatives.tex
not sure

@@
 \newcommand{\fiveg}{\mathbb{5}}
+\newcommand{\bck}{\triangleleft}
+\newcommand{\fwd}{\triangleright}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{$\Pi$: Operational and Denotational Semantics}
+
+Present $\Pi$ with an interpreter specifying the operational semantics. For a
+model or denotational semantics, the minimum requirement is that every type
+maps to some kind of space; a value $v : t$ maps to an element of that space;
+and a combinator $c : t_1 \leftrightarrow t_2$ maps to a function between the
+spaces denoting $t_1$ and $t_2$. The simplest model one can think of is to
+map a type of size $n$ to a finite set of $n$ elements; a value $v : t$ maps
+to an element in the set denoting $t$; a combinator maps to a permutation on
+that set. One could also build other models, e.g. vector spaces but that
+model is fine. Perhaps show correspondence between operational and
+denotational semantics.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{$\Pi^{-}$: Operational Semantics}
+
+Extend $\Pi$ with negative types and $\eta$ and $\epsilon$. Present the
+operational semantics that uses two interpreters forwards and backwards.
+
+For the model or denotational semantics, we need to map every type to some
+space. If type $t$ maps to space $S$, we need to have an operation that maps
+that space to another space that acts like ``negative'' $S$ so that we can
+satisfy for example that $S-S=0$. Using the cardinality of the set like we
+did for $\Pi$ does not directly work (footnote on papers that describe sets
+with negative cardinality). Another idea is to generalize spaces to have two
+components (a producer and a consumer; a left player and a right player) and
+to describe the space using the relative ``strength'' of the two
+components. (This is the idea of the Int or G construction but that doesn't
+directly work if we have two monoidal structures.) A natural and beautiful
+theory is that of Conway games which model two player games: left and
+right. In that setting we can talk about games in which the left player has
+an $n$-move advantage (summarized by the number $n$) and games in which the
+right player has an $n$-move advantage (summarized by the number $-n$ which
+denotes that the left player has an $n$-move disadvantage). It is therefore
+possible to map a type of size $n$ where $n$ is an integer to the class
+Conway games where the left player has an $n$-move (dis-)advantage; the
+negative of a space $S$ is the space in which all the games are flipped with
+left and right players exchanging positions; a value $v : t$ maps to a Conway
+number in the set denoting $t$; all combinators preserve the size of the type
+and hence will be the identity. Note that the space corresponds to type $0$
+is not empty in this denotation: it consists of all the games in which the
+left player has a $0$-move advantage over the the right player and
+symmetrically, i.e., the set of games which are always lost by the first
+player to move.
+
+Given that our language is reversible, a backward evaluator is relatively
+straightforward to implement: using the backward evaluator to calculate
+$c(v)$ is equivalent to $c^{\dagger}(v)$ in the forward evaluator.
+
+\begin{figure*}
+\[\begin{array}{rclr}
+[\mathit{iso}, v, C, s]_{\bck} &\longmapsto& 
+  \langle \mathit{iso}, v', C, s'\rangle_{\bck} 
+  \quad\textrm{where}~\mathit{iso}^{\dagger}~s~v \longmapsto (v', s') \\
+\langle c_1, v, \mathit{Fst}~C~c_2, s \rangle_{\bck} &\longmapsto& 
+  \langle c_1 \fatsemi c_2, v, C, s \rangle_{\bck} \\
+\langle c_2, v, \mathit{Snd}~c_1~C, s \rangle_{\bck} &\longmapsto& 
+  [c_1, v, \mathit{Fst}~C~c_2, s]_{\bck} \\
+[ c_1 \fatsemi c_2, v, C, s ]_{\bck} &\longmapsto& 
+  [c_2, v, \mathit{Snd}~c_1~C, s]_{\bck} \\
+\langle c_1, v, \mathit{LeftP}~C~c_2, s \rangle_{\bck} &\longmapsto& 
+  \langle c_1 \oplus c_2, \mathit{left}~v, C \rangle_{\bck} \\
+[c_1 \oplus c_2, v', C, s ]_{\bck} &\longmapsto& 
+  [ c_1, v, \mathit{LeftP}~C~c_2, s[v' \approx left v] ]_{\bck} \\
+\langle c_2, v, \mathit{RightP}~c_1~C, s\rangle_{\bck} &\longmapsto& 
+  \langle c_1 \oplus c_2, \mathit{right}~v, C \rangle_{\bck} \\
+[c_1 \oplus c_2, v', C,s ]_{\bck} &\longmapsto& 
+  [ c_2, v, \mathit{RightP}~c_1~C, s[v' \approx right v] ]_{\bck} \\
+\langle c_1, v_1, \mathit{LeftT}~C~c_2~v_2, s \rangle_{\bck} &\longmapsto& 
+  \langle c_1 \otimes c_2, (v_1, v_2), C, s \rangle_{\bck} \\
+\langle c_2, v_2, \mathit{RightT}~c_1~v_1~C, s \rangle_{\bck} &\longmapsto& 
+  [ c_1, v_1, \mathit{LeftT}~C~c_2~v_2, s ]_{\bck} \\
+[ c_1 \otimes c_2, v, C, s ]_{\bck} &\longmapsto& 
+  [ c_2, v_2, \mathit{RightT}~c_1~v_1~C, S' ]_{\bck} 
+  \quad\textrm{where}~ s' = s[v \approx (v_1, v_2)] \\
+\\
+\\
+\langle \epsilon, v, C, s \rangle_{\fwd} &\longmapsto& 
+  \langle \epsilon, \mathit{left}~(-v'), C, s[v \approx right v'] \rangle_{\bck} \\
+\langle \epsilon, v, C, s \rangle_{\fwd} &\longmapsto& 
+  \langle \epsilon, \mathit{right}~v', C, s[v \approx left (-v')] \rangle_{\bck} \\
+\\
+\\
+\langle \eta, v, C, s \rangle_{\bck} &\longmapsto& 
+  \langle \eta, \mathit{left}~(-v'), C, s[v \approx right v'] \rangle_{\fwd} \\
+\langle \eta, v, C, s \rangle_{\bck} &\longmapsto& 
+  \langle \eta, \mathit{right}~v', C, s[v \approx left (-v')] \rangle_{\fwd}
+\end{array}\]
+\end{figure*}
+
+We add the following rules to the reductions above. 
+\begin{enumerate}
+\item The rules for $\epsilon$ essentially transfer control from the forward
+  evaluator (whose states are tagged by $\triangleright$) to the backward
+  evaluator (whose states are tagged by $\triangleleft$). In other words,
+  after an $\epsilon$ the direction of the world is reversed. The pattern
+  matching done by the unification ensures that a value on the right wire is
+  tagged to be negative and transferred to the left wire, and vice versa.
+  Note that there is no evaluation rule for $\eta$ in the forward
+  evaluator. This corresponds to the fact that there is no value of type 0
+  and hence the forward evaluator can never execute an $\eta$.
+
+\item The rules for $eta$ are added to the backward evaluator. A program
+  executing backwards starts executing forwards after the execution of the
+  $\eta$. Dual to the previous case, there is no rule for $\epsilon$ in the
+  backward evaluator since the output type of $\epsilon$ is 0.
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conway Numbers}
@@
 
-\textbf{Comparing numbers.} There are three outcomes to games represented by
-Conway numbers $\mathbb{n}$:
+\smallskip\noindent\textbf{Comparing numbers.} There are three outcomes to
+games represented by Conway numbers $\mathbb{n}$:
 \begin{itemize}
@@
 
-\noindent\textbf{Addition and subtraction.} Addition of two games intuitively
-gives each player the choice of selecting an option from either
+\smallskip\noindent\textbf{Addition and subtraction.} Addition of two games
+intuitively gives each player the choice of selecting an option from either
 game. Subtraction is simply the addition of the dual of a game. Formally:
@@
 
-\noindent\textbf{Multiplication.} The most complicated operation on Conway
-numbers is multiplication. It is defined as follows:
+\smallskip\noindent\textbf{Multiplication.} The most complicated operation on
+Conway numbers is multiplication. It is defined as follows:
 \begin{code}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Correspondence}
+
+Show correspondence between operational and denotational semantics for
+$\Pi^{-}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliographystyle{abbrvnat}

### 2014-05-08 Conway.hs

@@ -5,13 +5,8 @@ module Conway where
 import Data.List
 
 ------------------------------------------------------------------------------
--- Definition; a few constants; several show functions
-
 {--
-Conventional definition:
-
-data Game = Game [Game] [Game]
-  deriving (Eq, Show)
+Some random thoughts about how to "type" Conway games
 
 Our definition:
 
@@ -21,8 +16,6 @@ Our definition:
   S ::= 0 | GR | S + S
 
   P ::= 1 | GL | P * P
-
---}
 
 data GL = GL S P
 data GR = GR P S
@@ -94,8 +87,6 @@ plusLP gl (Opponent p1 p2) = Opponent (plusLP gl p1) (plusLP gl p2)
 --    ((map (`plusG` h) gls) `union` (map (g `plusG`) hls))
 --    ((map (`plusG` h) grs) `union` (map (g `plusG`) hrs))
 
-
-{--
 -- A morphism between games < xls | xrs > and < yls | yrs > 
 -- is a game < yls + xrs | xls + yrs > 
 
@@ -161,8 +152,6 @@ data Combinator =
 ------------------------------------------------------------------------------
 -- Using conventional definition of Conway games...
 
-{--
-
 data Game = Game [Game] [Game]
   deriving (Eq, Show)
 
@@ -182,8 +171,6 @@ int2Game :: Int -> Game
 int2Game n | n < 0 = Game []               [int2Game (n+1)]
 int2Game n | n > 0 = Game [int2Game (n-1)] []
 int2Game n | otherwise = zeroG
-
---
 
 showG :: Game -> String
 showG (Game [] []) = "0"
@@ -209,46 +196,9 @@ showT s (Game [g1] [g2,g3]) =
            showT "'*" g1 ++ showT "#\\space" (Game [g2] [g3]) ++ ")"
 
 ------------------------------------------------------------------------------
--- addition and subtraction
-
-plusG :: Game -> Game -> Game
-g@(Game gls grs) `plusG` h@(Game hls hrs) = 
-  Game 
-    ((map (`plusG` h) gls) `union` (map (g `plusG`) hls))
-    ((map (`plusG` h) grs) `union` (map (g `plusG`) hrs))
-
-negG :: Game -> Game
-negG (Game gls grs) = Game (map negG grs) (map negG gls) 
-
-minusG :: Game -> Game -> Game
-g1 `minusG` g2 = g1 `plusG` (negG g2) 
-
[…]

### 2014-05-09 negatives.tex
story

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{$\Pi$: Operational and Denotational Semantics}
+\section{\textsc{LET}: Operational Semantics}
 
-Present $\Pi$ with an interpreter specifying the operational semantics. For a
-model or denotational semantics, the minimum requirement is that every type
-maps to some kind of space; a value $v : t$ maps to an element of that space;
-and a combinator $c : t_1 \leftrightarrow t_2$ maps to a function between the
-spaces denoting $t_1$ and $t_2$. The simplest model one can think of is to
-map a type of size $n$ to a finite set of $n$ elements; a value $v : t$ maps
-to an element in the set denoting $t$; a combinator maps to a permutation on
-that set. One could also build other models, e.g. vector spaces but that
-model is fine. Perhaps show correspondence between operational and
-denotational semantics.
+We introduce a toy functional language. Separate the first-order from the
+higher-order component.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{$\Pi^{-}$: Operational Semantics}
+\section{$\Pi^{o}$: Operational and Denotational Semantics}
 
-Extend $\Pi$ with negative types and $\eta$ and $\epsilon$. Present the
-operational semantics that uses two interpreters forwards and backwards.
+Present $\Pi^{o}$ with an interpreter specifying the operational
+semantics. This is $\Pi$ with recursive types and a trace operator. (It might
+be enough to have just one recursive type $\mathbb{N}$.) We know
+\emph{first-order} \textsc{LET} can be translated to this language: we cannot
+translate higher-order functions. That requires compact closure but we live a
+plain symmetric bimonoidal category.
 
-For the model or denotational semantics, we need to map every type to some
-space. If type $t$ maps to space $S$, we need to have an operation that maps
-that space to another space that acts like ``negative'' $S$ so that we can
-satisfy for example that $S-S=0$. Using the cardinality of the set like we
-did for $\Pi$ does not directly work (footnote on papers that describe sets
-with negative cardinality). Another idea is to generalize spaces to have two
+For a concrete model or denotational semantics that instantiates the
+categorical framework, the minimum requirement is that every type maps to
+some kind of space; a value $v : t$ maps to an element of that space; and a
+combinator $c : t_1 \leftrightarrow t_2$ maps to a function between the
+spaces denoting $t_1$ and $t_2$. The simplest model one can think of is to
+map a type of size $n$ to a set of $n$ elements (including the infinite set
+of $\omega$ elements; a value $v : t$ maps to an element in the set denoting
+$t$; a combinator maps to a permutation on that set. One could also build
+other models, e.g. vector spaces (infinite dimensional however) but that
+model is fine. Perhaps show the details of the model and the correspondence
+between operational and denotational semantics.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{$\Pi^{o-}$: Operational Semantics}
+
+Extend $\Pi^{o}$ with negative types and $\eta$ and $\epsilon$. Present the
+operational semantics that uses two interpreters forwards and backwards. Do
+we have such a semantics that we like?
+
+The categorical model is now supposed to be a ring category. This is
+complicated so we will work directly with a concrete model that gives us more
+operational intuition. We need to map every type to some space. If type $t$
+maps to space $S$, we need to have an operation that maps that space to
+another space that acts like ``negative'' $S$ so that we can satisfy for
+example that $S-S=0$. Using the cardinality of the set like we did for
+$\Pi^{o}$ does not directly work (footnote on papers that describe sets with
+negative cardinality). Another idea is to generalize spaces to have two
 components (a producer and a consumer; a left player and a right player) and
@@
 components. (This is the idea of the Int or G construction but that doesn't
-directly work if we have two monoidal structures.) A natural and beautiful
-theory is that of Conway games which model two player games: left and
-right. In that setting we can talk about games in which the left player has
-an $n$-move advantage (summarized by the number $n$) and games in which the
-right player has an $n$-move advantage (summarized by the number $-n$ which
-denotes that the left player has an $n$-move disadvantage). It is therefore
-possible to map a type of size $n$ where $n$ is an integer to the class
-Conway games where the left player has an $n$-move (dis-)advantage; the
-negative of a space $S$ is the space in which all the games are flipped with
-left and right players exchanging positions; a value $v : t$ maps to a Conway
-number in the set denoting $t$; all combinators preserve the size of the type
-and hence will be the identity. Note that the space corresponds to type $0$
-is not empty in this denotation: it consists of all the games in which the
-left player has a $0$-move advantage over the the right player and
-symmetrically, i.e., the set of games which are always lost by the first
-player to move.
+directly work if we have two monoidal structures and needs a complicated
+completion to ring categories.) A natural and beautiful theory is that of
+Conway games which model two player games: left and right. In that setting we
+can talk about games in which the left player has an $n$-move advantage
+(summarized by the number $n$) and games in which the right player has an
+$n$-move advantage (summarized by the number $-n$ which denotes that the left
+player has an $n$-move disadvantage). It is therefore possible to map a type
+of size $n$ where $n$ is an integer to the class Conway games where the left
+player has an $n$-move (dis-)advantage; the negative of a space $S$ is the
+space in which all the games are flipped with left and right players
+exchanging positions; following common terminology, we call the target of the
+type an \emph{arena}; a value $v : t$ maps to a \emph{strategy} for the left
+player in the arena for type $t$. Combinators transform strategies. Note that
+the space corresponds to type $0$ is an arena with no possible moves.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Operational Semantics}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conway Numbers}
+\section{Conway Numbers: Arenas}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Correspondence}
+\section{$\Pi^{o-}$: Denotational Semantics}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{$\Pi^{o-}$: Correspondence}
 
 Show correspondence between operational and denotational semantics for
-$\Pi^{-}$.
+$\Pi^{o-}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Back to $\textsc{LET}$}
+
+Now work out the full translation from \textsc{LET} to $\Pi^{o-}$.
 

### 2014-05-09 Conway.hs

@@ -6,6 +6,51 @@ import Data.List
 
 ------------------------------------------------------------------------------
 {--
+Game semantics:
+
+- types t are modeled by arenas [t]
+- terms e:t are modeled by strategies for left for playing in the arena [t]
+- evaluation of e corresponds to a play between left and right in the arena
+
+----------------
+type 0 => arena is the equivalence class of 'Game [] []'
+ 
+no terms of type 0 which is consistent with the fact that there is 
+no strategy for left for playing in the empty arena
+
+--
+
+type 1 ==> arena is the equivalence class of 'Game [zeroG] []'
+there is only one strategy for left: move left
+
+-- 
+
+type A + B ==> arena is the Conway sum of the arenas for A and B
+
+give S_A : A, there is a strategy left S_A that selects a move in A 
+  according to S_A
+there is also another strategy right S_B that selects a move in B 
+  according to S_B
+
+To summarize:
+
+  Our types map to the corresponding (equivalence classes of) Conway numbers
+  These Conway numbers are the arenas
+
+  The values (), L v, R v, (v,v), and -v 
+  map to the following strategies: 
+  () ==> the unique strategy for left when there is exactly one left option
+  L v ==> the strategy for left that plays 'v' in the left game of a sum
+  R v ==> the strategy for left that plays 'v' in the right game of a sum
+  (v1,v2) ==> the strategy for left that plays ... in a product game
+  -v ==> the strategy for left that mimics right ???
+
+  Our combinators transform strategies: for example swap+ transforms L v to R v
+
+  Do eta/epsilon allow backtracking in strategies?
+
+----------------
+
 Some random thoughts about how to "type" Conway games
 
 Our definition:

### 2014-05-16 Strategies.hs
strategies

@@ -1,0 +1,133 @@
+module Strategies where
+
+import Data.List
+
+------------------------------------------------------------------------------
+-- Every type corresponds to an arena 
+
+data Arena = Arena [Arena] [Arena]
+  deriving (Show,Eq)
+
+leftOptions :: Arena -> [Arena]
+leftOptions (Arena gls _) = gls
+
+rightOptions :: Arena -> [Arena]
+rightOptions (Arena _ grs) = grs
+
+data T = Zero | One | Plus T T | Times T T | Neg T
+
+arena :: T -> Arena
+arena Zero = Arena [] []
+arena One = Arena [ Arena [] [] ] []
+arena (Plus t1 t2) = arena t1 `plusA` arena t2
+arena (Times t1 t2) = arena t1 `timesA` arena t2
+arena (Neg t) = negA (arena t) 
+
+plusA :: Arena -> Arena -> Arena
+g@(Arena gls grs) `plusA` h@(Arena hls hrs) = 
+  Arena
+    ((map (`plusA` h) gls) `union` (map (g `plusA`) hls))
+    ((map (`plusA` h) grs) `union` (map (g `plusA`) hrs))
+
+negA :: Arena -> Arena
+negA (Arena gls grs) = Arena (map negA grs) (map negA gls) 
+
+minusA :: Arena -> Arena -> Arena
+g1 `minusA` g2 = g1 `plusA` (negA g2) 
+
+timesA :: Arena -> Arena -> Arena
+x@(Arena xls xrs) `timesA` y@(Arena yls yrs) = 
+  Arena 
+    ([ (xl `timesA` y) `plusA` (x `timesA` yl) `minusA` (xl `timesA` yl)
+     | xl <- xls, yl <- yls] `union`
+     [ (xr `timesA` y) `plusA` (x `timesA` yr) `minusA` (xr `timesA` yr)
+     | xr <- xrs, yr <- yrs])
+    ([ (xl `timesA` y) `plusA` (x `timesA` yr) `minusA` (xl `timesA` yr)
+     | xl <- xls, yr <- yrs] `union`
+     [ (xr `timesA` y) `plusA` (x `timesA` yl) `minusA` (xr `timesA` yl)
+     | xr <- xrs, yl <- yls])
+
+------------------------------------------------------------------------------
+-- Play games!
+
+-- alternate L/R
+data Play = Null | L Int Play | R Int Play
+data Result = LeftWins | RightWins | Incomplete
+  deriving Show
+
[…]

### 2014-05-16 Strategies.hs
more on strategies

@@ -77,21 +77,26 @@ test2 = game (arena (Plus One One)) Null
 test3 = game (arena (Plus three negTwo)) (R 0 (L 0 (R 0 (L 0 Null))))
 
 ------------------------------------------------------------------------------
--- Every value corresponds to a strategy
+-- Every value corresponds to a strategy; so our strategies are deterministic
+-- and precomputed; we are not allowed to select our move based on the
+-- opponent's move; hopefully with negatives we can backtrack and hence make
+-- our strategies more flexible
 
--- A strategy is a left move for each possible right move
-
-data Strategy = S (Int -> Strategy) 
+type Strategy = Play -- only left moves
 
 data Val = Unit | InL Val | InR Val | Pair Val Val
 
 val2Strategy :: (Val,T) -> (Strategy,Arena)
 val2Strategy (Unit,One) = (s,a)
   where a = arena One 
-        s = S (\_ -> undefined)
+        s = L 0 Null -- select the only possible move left has
 val2Strategy (InL v, Plus t1 t2) = (s,a)
   where a = arena (Plus t1 t2)
-        s = undefined
+        (s1,a1) = val2Strategy (v,t1)
+        -- s1 is strategy for t1
+        -- looking at plusA, left has the following options:
+        -- t1L + t2 or t1 + t2L
+        s = L undefined s1
 val2Strategy (InR v, Plus t1 t2) = (s,a)
   where a = arena (Plus t1 t2)
         s = undefined
@@ -101,20 +106,20 @@ val2Strategy (Pair v1 v2, Times t1 t2) = (s,a)
 
 -- take left strategy and right moves; right starts
 gameWithStrategy :: Arena -> Strategy -> Play -> Result
-gameWithStrategy (Arena ls []) (S s) Null = LeftWins
-gameWithStrategy (Arena ls rs) (S s) Null = Incomplete
-gameWithStrategy (Arena ls rs) (S s) (R i p) = 
-  gameWithStrategy (rs !! i) (s i) p
+gameWithStrategy = gameR
+  where gameL (Arena [] rs) _ _ = RightWins
+        gameL (Arena ls rs) (L i s) p = gameR (ls !! i) s p
+        gameL _ _ _ = error "malformed game"
+
+        gameR (Arena ls []) _ _ = LeftWins
+        gameR (Arena ls rs) s (R i p) = gameL (rs !! i) s p
[…]

### 2014-05-16 Strategies.hs
redesign strategies

@@ -5,33 +5,91 @@ import Data.List
 ------------------------------------------------------------------------------
 -- Every type corresponds to an arena 
 
-data Arena = Arena [Arena] [Arena]
+data Arena = Arena Options Options
   deriving (Show,Eq)
 
-leftOptions :: Arena -> [Arena]
+data Options = None 
+             | Singleton Arena 
+             | Choose Options Options
+             | Both Options Options
+             | Flip Options
+  deriving (Show,Eq)
+
+leftOptions :: Arena -> Options
 leftOptions (Arena gls _) = gls
 
-rightOptions :: Arena -> [Arena]
+rightOptions :: Arena -> Options
 rightOptions (Arena _ grs) = grs
+
+-- simplify so that we can understand the results (for now)
+
+simplifyA :: Arena -> Arena
+simplifyA (Arena lo ro) = Arena (simplifyO lo) (simplifyO ro)
+
+simplifyO :: Options -> Options
+simplifyO None = None
+simplifyO (Singleton a) = Singleton (simplifyA a)
+simplifyO (Choose o1 o2) = case (so1,so2) of
+  (None,None) -> None
+  (None,so) -> so
+  (so,None) -> so
+  (so1,so2) -> if so1 == so2 then so1 else Choose so1 so2
+  where so1 = simplifyO o1
+        so2 = simplifyO o2
+simplifyO (Both o1 o2) = Both (simplifyO o1) (simplifyO o2)
+simplifyO (Flip o) = Flip (simplifyO o)
+
+-- 
 
 data T = Zero | One | Plus T T | Times T T | Neg T
 
 arena :: T -> Arena
-arena Zero = Arena [] []
-arena One = Arena [ Arena [] [] ] []
+arena Zero = Arena None None
+arena One = Arena (Singleton (arena Zero)) None
 arena (Plus t1 t2) = arena t1 `plusA` arena t2
 arena (Times t1 t2) = arena t1 `timesA` arena t2
 arena (Neg t) = negA (arena t) 
 
 plusA :: Arena -> Arena -> Arena
 g@(Arena gls grs) `plusA` h@(Arena hls hrs) = 
-  Arena
-    ((map (`plusA` h) gls) `union` (map (g `plusA`) hls))
-    ((map (`plusA` h) grs) `union` (map (g `plusA`) hrs))
[…]

### 2014-05-16 Strategies.hs
rewinding...

@@ -1,198 +1,0 @@
-module Strategies where
-
-import Data.List
-
-------------------------------------------------------------------------------
--- Every type corresponds to an arena 
-
-data Arena = Arena Options Options
-  deriving (Show,Eq)
-
-data Options = None 
-             | Singleton Arena 
-             | Choose Options Options
-             | Both Options Options
-             | Flip Options
-  deriving (Show,Eq)
-
-leftOptions :: Arena -> Options
-leftOptions (Arena gls _) = gls
-
-rightOptions :: Arena -> Options
-rightOptions (Arena _ grs) = grs
-
--- simplify so that we can understand the results (for now)
-
-simplifyA :: Arena -> Arena
-simplifyA (Arena lo ro) = Arena (simplifyO lo) (simplifyO ro)
-
-simplifyO :: Options -> Options
-simplifyO None = None
-simplifyO (Singleton a) = Singleton (simplifyA a)
-simplifyO (Choose o1 o2) = case (so1,so2) of
-  (None,None) -> None
-  (None,so) -> so
-  (so,None) -> so
-  (so1,so2) -> if so1 == so2 then so1 else Choose so1 so2
-  where so1 = simplifyO o1
-        so2 = simplifyO o2
-simplifyO (Both o1 o2) = Both (simplifyO o1) (simplifyO o2)
-simplifyO (Flip o) = Flip (simplifyO o)
-
--- 
-
-data T = Zero | One | Plus T T | Times T T | Neg T
-
-arena :: T -> Arena
-arena Zero = Arena None None
-arena One = Arena (Singleton (arena Zero)) None
-arena (Plus t1 t2) = arena t1 `plusA` arena t2
-arena (Times t1 t2) = arena t1 `timesA` arena t2
-arena (Neg t) = negA (arena t) 
-
-plusA :: Arena -> Arena -> Arena
-g@(Arena gls grs) `plusA` h@(Arena hls hrs) = 
-  Arena (Choose (gls `leftPlus` h) (g `rightPlus` hls))
-        (Choose (grs `leftPlus` h) (g `rightPlus` hrs))
-
-leftPlus :: Options -> Arena -> Options
-leftPlus None _ = None
-leftPlus (Singleton a) a' = Singleton (a `plusA` a')
-leftPlus (Choose o1 o2) a = Choose (o1 `leftPlus` a) (o2 `leftPlus` a)
[…]

### 2014-05-16 Strategies.hs
minor change in strategies

@@ -52,7 +52,7 @@ x@(Arena xls xrs) `timesA` y@(Arena yls yrs) =
 
 -- alternate L/R
 data Play = Null | L Int Play | R Int Play
-data Result = LeftWins | RightWins | Incomplete
+data Result = LeftWins | RightWins 
   deriving Show
 
 -- play starts with right move
@@ -60,13 +60,11 @@ game :: Arena -> Play -> Result
 game = gameR
   where gameL (Arena [] rs) _ = RightWins
         gameL (Arena ls rs) (L i p) = gameR (ls !! i) p
-        gameL (Arena ls rs) Null = Incomplete
-        gameL (Arena ls rs) (R i p) = error "malformed play"
+        gameL _ _ = error "malformed play"
 
         gameR (Arena ls []) _ = LeftWins
         gameR (Arena ls rs) (R i p) = gameL (rs !! i) p
-        gameR (Arena ls rs) Null = Incomplete
-        gameR (Arena ls rs) (L i p) = error "malformed play"
+        gameR _ _ = error "malformed play"
 
 three  = Plus One (Plus One One)
 negTwo = Neg (Plus One One)
@@ -102,7 +100,6 @@ val2Strategy (Pair v1 v2, Times t1 t2) = (s,a)
 -- take left strategy and right moves; right starts
 gameWithStrategy :: Arena -> Strategy -> Play -> Result
 gameWithStrategy (Arena ls []) (S s) Null = LeftWins
-gameWithStrategy (Arena ls rs) (S s) Null = Incomplete
 gameWithStrategy (Arena ls rs) (S s) (R i p) = 
   gameWithStrategy (rs !! i) (s i) p
 

### 2014-05-16 Strategies.hs
back to old strategies

@@ -1,12 +1,10 @@
 module Strategies where
 
-import Data.List
-
 ------------------------------------------------------------------------------
 -- Every type corresponds to an arena 
 
 data Arena = Arena [Arena] [Arena]
-  deriving (Show,Eq)
+  deriving Show
 
 leftOptions :: Arena -> [Arena]
 leftOptions (Arena gls _) = gls
@@ -18,7 +16,7 @@ data T = Zero | One | Plus T T | Times T T | Neg T
 
 arena :: T -> Arena
 arena Zero = Arena [] []
-arena One = Arena [ Arena [] [] ] []
+arena One = Arena [ arena Zero ] []
 arena (Plus t1 t2) = arena t1 `plusA` arena t2
 arena (Times t1 t2) = arena t1 `timesA` arena t2
 arena (Neg t) = negA (arena t) 
@@ -26,8 +24,8 @@ arena (Neg t) = negA (arena t)
 plusA :: Arena -> Arena -> Arena
 g@(Arena gls grs) `plusA` h@(Arena hls hrs) = 
   Arena
-    ((map (`plusA` h) gls) `union` (map (g `plusA`) hls))
-    ((map (`plusA` h) grs) `union` (map (g `plusA`) hrs))
+    ((map (`plusA` h) gls) ++ (map (g `plusA`) hls))
+    ((map (`plusA` h) grs) ++ (map (g `plusA`) hrs))
 
 negA :: Arena -> Arena
 negA (Arena gls grs) = Arena (map negA grs) (map negA gls) 
@@ -39,11 +37,11 @@ timesA :: Arena -> Arena -> Arena
 x@(Arena xls xrs) `timesA` y@(Arena yls yrs) = 
   Arena 
     ([ (xl `timesA` y) `plusA` (x `timesA` yl) `minusA` (xl `timesA` yl)
-     | xl <- xls, yl <- yls] `union`
+     | xl <- xls, yl <- yls] ++
      [ (xr `timesA` y) `plusA` (x `timesA` yr) `minusA` (xr `timesA` yr)
      | xr <- xrs, yr <- yrs])
     ([ (xl `timesA` y) `plusA` (x `timesA` yr) `minusA` (xl `timesA` yr)
-     | xl <- xls, yr <- yrs] `union`
+     | xl <- xls, yr <- yrs] ++
      [ (xr `timesA` y) `plusA` (x `timesA` yl) `minusA` (xr `timesA` yl)
[…]

### 2014-05-17 Strategies.hs
something

@@ -3,23 +3,262 @@ module Strategies where
 ------------------------------------------------------------------------------
 -- Every type corresponds to an arena 
 
-data Arena = Arena [Arena] [Arena]
+data Arena = ZeroA
+           | OneA 
+           | PlusA Arena Arena
+           | TimesA Arena Arena
+           | NegA Arena
   deriving Show
-
-leftOptions :: Arena -> [Arena]
-leftOptions (Arena gls _) = gls
-
-rightOptions :: Arena -> [Arena]
-rightOptions (Arena _ grs) = grs
 
 data T = Zero | One | Plus T T | Times T T | Neg T
 
 arena :: T -> Arena
-arena Zero = Arena [] []
-arena One = Arena [ arena Zero ] []
-arena (Plus t1 t2) = arena t1 `plusA` arena t2
-arena (Times t1 t2) = arena t1 `timesA` arena t2
-arena (Neg t) = negA (arena t) 
+arena Zero = ZeroA
+arena One = OneA
+arena (Plus t1 t2) = PlusA (arena t1) (arena t2)
+arena (Times t1 t2) = TimesA (arena t1) (arena t2)
+arena (Neg t) = NegA (arena t) 
+
+------------------------------------------------------------------------------
+-- Play games!
+
+-- alternate left and right moves
+
+data Play = NullP | L Move | R Move
+data Move = SingletonM 
+          | LeftM Move 
+          | RightM Move 
+          | PPL Move
+          | PPR Move
+          | PPBoth Move Move
+          | MML Move
+          | MMR Move
+          | MMBoth Move Move
+          | PML Move
+          | PMR Move
+          | PMBoth Move Move 
+          | MPL Move
+          | MPR Move
+          | MPBoth Move Move
+          | OpponentM Move
+
+data Result = LeftWins | RightWins 
+  deriving Show
+
+-- play starts with right move
+
+-- left moves
+
+gameL :: Arena -> Move -> Maybe Arena
+gameL ZeroA _ = Nothing
+gameL OneA SingletonM = Just ZeroA
+gameL (PlusA a1 a2) (LeftM m) = do g <- gameL a1 m
+                                   return $ PlusA g a2
+gameL (PlusA a1 a2) (RightM m) = do g <- gameL a2 m
+                                    return $ PlusA a1 g
+
+gameL (TimesA a1 a2) (PPL m1) = do g1 <- gameL a1 m1
[…]

### 2014-05-17 Strategies.hs
major redesign

@@ -1,7 +1,32 @@
 module Strategies where
 
+import Debug.Trace
+
 ------------------------------------------------------------------------------
--- Every type corresponds to an arena 
+-- Every type corresponds to an arena. 
+-- Every value correspods to a move (or perhaps strategy) for black
+
+{-- 
+
+Conventionally we have: data Arena = A [Arena] [Arena] where we maintain
+lists of games corresponding to left options and to right options. I am going
+to call these 'black' options and 'white' options. We play black and the
+opponent plays white. The opponent (white) starts. A move simply selects an
+index in the list. P.77 of the book gives examples where it is beneficial to
+maintain the structure of the games keeping them separate at the cost of
+complicating the definition of moves. So instead of collapsing the structure
+when we add games:
+
+plusA :: Arena -> Arena -> Arena
+g@(Arena gls grs) `plusA` h@(Arena hls hrs) = 
+  Arena
+    ((map (`plusA` h) gls) ++ (map (g `plusA`) hls))
+    ((map (`plusA` h) grs) ++ (map (g `plusA`) hrs))
+
+We maintain the two games as PlusA g1 g2 and when you make a move you can
+specify move in the left game which takes you to PlusA (leftOption_of_g1) g2
+
+--}
 
 data Arena = ZeroA
            | OneA 
@@ -10,131 +35,138 @@ data Arena = ZeroA
            | NegA Arena
   deriving Show
 
-data T = Zero | One | Plus T T | Times T T | Neg T
-
-arena :: T -> Arena
-arena Zero = ZeroA
-arena One = OneA
-arena (Plus t1 t2) = PlusA (arena t1) (arena t2)
-arena (Times t1 t2) = TimesA (arena t1) (arena t2)
-arena (Neg t) = NegA (arena t) 
-
 ------------------------------------------------------------------------------
 -- Play games!
 
--- alternate left and right moves
+-- alternate moves between player (black) and opponent (white)
 
-data Play = NullP | L Move | R Move
-data Move = SingletonM 
-          | LeftM Move 
-          | RightM Move 
-          | PPL Move
-          | PPR Move
-          | PPBoth Move Move
-          | MML Move
-          | MMR Move
-          | MMBoth Move Move
-          | PML Move
-          | PMR Move
-          | PMBoth Move Move 
[…]

### 2014-05-17 Strategies.hs
typed version of strategies

@@ -1,21 +1,24 @@
+{-# LANGUAGE TypeFamilies, TypeOperators #-}
+
 module Strategies where
 
-import Debug.Trace
-
 ------------------------------------------------------------------------------
--- Every type corresponds to an arena. 
--- Every value correspods to a move (or perhaps strategy) for black
+-- Every type corresponds to an arena in which black and white can play a
+-- game
+
+-- Every value correspods to a move for black. The white player starts the
+-- game and is the opponent.
 
 {-- 
 
 Conventionally we have: data Arena = A [Arena] [Arena] where we maintain
 lists of games corresponding to left options and to right options. I am going
 to call these 'black' options and 'white' options. We play black and the
-opponent plays white. The opponent (white) starts. A move simply selects an
-index in the list. P.77 of the book gives examples where it is beneficial to
-maintain the structure of the games keeping them separate at the cost of
-complicating the definition of moves. So instead of collapsing the structure
-when we add games:
+opponent plays white. The opponent (white) starts. Using that conventional
+representation, a move simply selects an index in the list. P.77 of the book
+gives examples where it is beneficial to maintain the structure of the games
+keeping them separate at the cost of complicating the definition of moves. So
+instead of collapsing the structure when we add games:
 
 plusA :: Arena -> Arena -> Arena
 g@(Arena gls grs) `plusA` h@(Arena hls hrs) = 
@@ -28,311 +31,94 @@ specify move in the left game which takes you to PlusA (leftOption_of_g1) g2
 
 --}
 
-data Arena = ZeroA
-           | OneA 
-           | PlusA Arena Arena
-           | TimesA Arena Arena
-           | NegA Arena
-  deriving Show
+-- Arenas...
 
-------------------------------------------------------------------------------
--- Play games!
+class Arena a where
+  type BlackView a :: * 
+  type WhiteView a :: *
 
--- alternate moves between player (black) and opponent (white)
+  type ZeroA        :: *  
+  type OneA         :: *  
+  type PlusA a1 a2  :: *
+  type TimesA a1 a2 :: *
[…]

### 2014-05-17 StrategiesUntyped.hs

@@ -1,0 +1,363 @@
+module Strategies where
+
+import Debug.Trace
+
+------------------------------------------------------------------------------
+-- Every type corresponds to an arena. 
+-- Every value correspods to a move (or perhaps strategy) for black
+
+{-- 
+
+Conventionally we have: data Arena = A [Arena] [Arena] where we maintain
+lists of games corresponding to left options and to right options. I am going
+to call these 'black' options and 'white' options. We play black and the
+opponent plays white. The opponent (white) starts. A move simply selects an
+index in the list. P.77 of the book gives examples where it is beneficial to
+maintain the structure of the games keeping them separate at the cost of
+complicating the definition of moves. So instead of collapsing the structure
+when we add games:
+
+plusA :: Arena -> Arena -> Arena
+g@(Arena gls grs) `plusA` h@(Arena hls hrs) = 
+  Arena
+    ((map (`plusA` h) gls) ++ (map (g `plusA`) hls))
+    ((map (`plusA` h) grs) ++ (map (g `plusA`) hrs))
+
+We maintain the two games as PlusA g1 g2 and when you make a move you can
+specify move in the left game which takes you to PlusA (leftOption_of_g1) g2
+
+--}
+
+data Arena = ZeroA
+           | OneA 
+           | PlusA Arena Arena
+           | TimesA Arena Arena
+           | NegA Arena
+  deriving Show
+
+------------------------------------------------------------------------------
+-- Play games!
+
+-- alternate moves between player (black) and opponent (white)
+
+data Moves = End | Black Move Moves | White Move Moves
+
+-- a move should be able to pick any option: if there is just on option the
+-- move is Singleton; if the game is the sum of two games, the move specifies
+-- whether to play in the left game or the right game; if the game is the
+-- product of two games, it is complicated; if the game is the dual of
+-- another game then copy the opponent's move in that game
+
+data Move = NoM
+          | SingletonM 
+          | LeftM Move | RightM Move 
+          | PPL Move | PPR Move | PPBoth Move Move
+          | MML Move | MMR Move | MMBoth Move Move
[…]

### 2014-05-17 Strategies.hs
starting multiplication

@@ -1,4 +1,4 @@
-{-# LANGUAGE TypeFamilies, TypeOperators #-}
+{-# LANGUAGE TypeFamilies, TypeOperators, GADTs #-}
 
 module Strategies where
 
@@ -47,7 +47,21 @@ data black :|: white = black :|: white
 
 data Void 
 
-data TODO
+data BlackProduct bw1 bw2 = 
+    BlackBlackLeft (BlackView bw1) 
+  | BlackBlackRight (BlackView bw2)
+  | BlackBlackBlack (BlackView bw1) (BlackView bw2) -- negated
+  | BlackWhiteLeft (WhiteView bw1)
+  | BlackWhiteRight (WhiteView bw2)
+  | BlackWhiteWhite (WhiteView bw1) (WhiteView bw2) -- negated
+
+data WhiteProduct bw1 bw2 = 
+    WhiteBlackLeft (BlackView bw1)
+  | WhiteWhiteRight (WhiteView bw2)
+  | WhiteBlackWhite (BlackView bw1) (WhiteView bw2) -- negated
+  | WhiteWhiteLeft (WhiteView bw1)
+  | WhiteBlackRight (BlackView bw2)
+  | WhiteWhiteBlack (WhiteView bw1) (BlackView bw2) -- negated
 
 instance Arena (black :|: white) where
 
@@ -66,7 +80,9 @@ instance Arena (black :|: white) where
      (Either black1 black2 :|: Either white1 white2)
 
   -- in the product arena, it is complicated
-  type TimesA (black1 :|: white1) (black2 :|: white2) = TODO
+  type TimesA (black1 :|: white1) (black2 :|: white2) = 
+    (BlackProduct (black1 :|: white1) (black2 :|: white2) :|:
+     WhiteProduct (black1 :|: white1) (black2 :|: white2))
 
   -- in the dual arena, the roles of Black and White are reversed
   type NegA (black :|: white) = (white :|: black) 
@@ -120,5 +136,75 @@ move in arena OneA we are now in arena ZeroA. This is weird
 --}
 
 ------------------------------------------------------------------------------
+-- Pi 
+-- Types are arenas
+-- Values are moves for Black
+-- combinators are black arrows mapping black moves to black moves
 
[…]

### 2014-05-17 Strategies.hs
strategies come back as the meaning of pi values

@@ -124,22 +124,32 @@ whiteMoveComplicatedOneA = Right (Left ())
 
 {--
 
-Since the game ComplicatedOneA is presumably equivalent to the game OneA,
-it should be possible to have a 1-1 correspondence between the moves in OneA
-and the moves in ComplicatedOneA. That's what the pi combinators witness.
-Computationally given that I know how to win OneA, I can derive moves 
-for ComplicatedOneA that are guaranteed to win.
+Since the game ComplicatedOneA is presumably equivalent to the game OneA, it
+should be possible to have a 1-1 correspondence between the moves in OneA and
+the moves in ComplicatedOneA. However it is clear that in OneA, White has no
+moves but that in ComplicatedOneA, White has a move. So a 1-1 correspondence
+between moves won't work. What will work is matching winning
+strategies. White has no winning strategy in OneA (obvious) but White also
+has no winning strategy in ComplicatedOneA. That's what the pi combinators
+witness.  Computationally given that I know how to win OneA, I can derive
+a strategy for ComplicatedOneA that is guaranteed to win.
 
 Note that actually playing games changes the types of the arenas: if I make a
-move in arena OneA we are now in arena ZeroA. This is weird
+move in arena OneA we are now in arena ZeroA. This is weird.
 
 --}
+
+-- Strategies
 
 ------------------------------------------------------------------------------
 -- Pi 
 -- Types are arenas
--- Values are moves for Black
--- combinators are black arrows mapping black moves to black moves
+-- Values are winning strategies for Black
+-- combinators are black arrows mapping winning black strategies to winning
+-- black strategies
+
+{--
+to fix...
 
 type BlackArrow a b = BlackView a -> BlackView b
 
@@ -162,7 +172,7 @@ timesOneL (BlackWhiteLeft _) = error "Impossible"
 timesOneL (BlackWhiteRight mW) = undefined
 timesOneL (BlackWhiteWhite _ mW) = error "Impossible"
 
-{--
+
 BlackProduct (() :|: Void) (aB :|: aW) = 
     BlackBlackLeft ()
   | BlackBlackRight aBlack

### 2014-05-18 Strategies.hs
on the right track with strategies...

@@ -134,27 +134,83 @@ has no winning strategy in ComplicatedOneA. That's what the pi combinators
 witness.  Computationally given that I know how to win OneA, I can derive
 a strategy for ComplicatedOneA that is guaranteed to win.
 
-Note that actually playing games changes the types of the arenas: if I make a
-move in arena OneA we are now in arena ZeroA. This is weird.
+In ComplicatedOneA, White has two opening moves 
+  Right (Left ()) or Right (Right ()). 
+They are really similar. Consider the following game:
+-- original arena: 
+-- = PlusA (PlusA (() :|: Void) (PlusA (() :|: Void) (() :|: Void)))
+--         (Neg (PlusA (() :|: Void) (() :|: Void)))
+-- White's has two similar moves, say White chooses Right (Left ())
+-- new arena
+-- = PlusA (PlusA (() :|: Void) (PlusA (() :|: Void) (() :|: Void)))
+--         (Neg (PlusA (Void :|: Void) (() :|: Void)))
+-- Black has three moves in the left part of the game. They are all similar
+-- so assume black chooses Left (Left ())
+-- new arena is:
+-- = PlusA (PlusA (Void :|: Void) (PlusA (() :|: Void) (() :|: Void)))
+--         (Neg (PlusA (Void :|: Void) (() :|: Void)))
+-- White has only one possible move Right (Right ()) which leads to:
+-- = PlusA (PlusA (Void :|: Void) (PlusA (() :|: Void) (() :|: Void)))
+--         (Neg (PlusA (Void :|: Void) (Void :|: Void)))
+-- Black has two moves to choose from; we move to:
+-- = PlusA (PlusA (Void :|: Void) (PlusA (Void :|: Void) (() :|: Void)))
+--         (Neg (PlusA (Void :|: Void) (Void :|: Void)))
+-- White has no moves; Black wins
+-- To summarize, Black's strategy in the game (ThreeA `plusA` MTwoA) is to
+-- counter any move by White in MTwoA by a move in ThreeA
+
+There should be a Pi combinator mapping 3-2 to 1. That circuit is probably
+what we want to compute the above:
+
+3 - 2 <==> 1
+(3 :|: 0) + (0 :|: 2) <==> (1 :|: 0)
[…]

### 2014-05-18 Strategies.hs
thinking about multiplication

@@ -210,36 +210,76 @@ data a :<=> b where
 plusZeroL :: (a ~ (aBlack :|: aWhite)) => M (PlusA ZeroA a) a
 plusZeroL = AssocPlusR :.: (Id :+: CommutePlus) :.: AssocPlusL
 
+timesOneL :: (a ~ (aBlack :|: aWhite)) => M (TimesA OneA a) a
+timesOneL = undefined
+
 {--
 
-plusZeroR :: (a ~ (aBlack :|: aWhite)) => BlackArrow a (PlusA ZeroA a) 
-plusZeroR m = Right m
+Either (BlackProduct (() :|: Void) (aB :|: aW)) aWhite 
+:<=> 
+Either (WhiteProduct (() :|: Void) (aB :|: aW)) aBlack
 
-commutePlus :: (a ~ (aBlack :|: aWhite), b ~ (bBlack :|: bWhite)) => 
-               BlackArrow (PlusA a b) (PlusA b a)
-commutePlus (Left m) = Right m
-commutePlus (Right m) = Left m
+and we have that:
+BlackProduct (() :|: Void) (aB :|: aW) has values:
+  BBL ()
+  BBR aB
+  BBB () aB
+  BWL Void
+  BWR aW
+  BWW Void aW
+WhiteProduct (() :|: Void) (aB :|: aW) has values:
+  WBL ()
+  WWR aW
+  WBW () aW
+  WWL Void
+  WBR aB
+  WWB Void aB
 
-timesOneL :: (a ~ (aBlack :|: aWhite)) => BlackArrow (TimesA OneA a) a
-timesOneL (BlackBlackLeft ()) = undefined
-timesOneL (BlackBlackRight mB) = mB
-timesOneL (BlackBlackBlack () mB) = undefined -- negated
-timesOneL (BlackWhiteLeft _) = error "Impossible"
-timesOneL (BlackWhiteRight mW) = undefined
-timesOneL (BlackWhiteWhite _ mW) = error "Impossible"
+LHS has values:
+  Right aW
+  Left (BBL ())
+  Left (BBR aB)
+  Left (BBB () aB)
+  Left (BWL Void)
+  Left (BWR aW)
+  Left (BWW Void aW)
 
+RHS has values:
+  Right aB
+  Left (WBL ())
+  Left (WWR aW)
+  Left (WBW () aW)
+  Left (WWL Void)
+  Left (WBR aB)
+  Left (WWB Void aB)
 
[…]

### 2014-05-18 Strategies.hs
seems like multiplication doesn't work

@@ -223,35 +223,35 @@ and we have that:
 BlackProduct (() :|: Void) (aB :|: aW) has values:
   BBL ()
   BBR aB
-  BBB () aB
+  - BBB () aB
   BWL Void
   BWR aW
-  BWW Void aW
+  - BWW Void aW
 WhiteProduct (() :|: Void) (aB :|: aW) has values:
   WBL ()
   WWR aW
-  WBW () aW
+  - WBW () aW
   WWL Void
   WBR aB
-  WWB Void aB
+  - WWB Void aB
 
 LHS has values:
   Right aW
   Left (BBL ())
   Left (BBR aB)
-  Left (BBB () aB)
+  - Left (BBB () aB)
   Left (BWL Void)
   Left (BWR aW)
-  Left (BWW Void aW)
+  - Left (BWW Void aW)
 
 RHS has values:
   Right aB
   Left (WBL ())
   Left (WWR aW)
-  Left (WBW () aW)
+  - Left (WBW () aW)
   Left (WWL Void)
   Left (WBR aB)
-  Left (WWB Void aB)
+  - Left (WWB Void aB)
 
 Get rid of the voids:
 
@@ -259,27 +259,116 @@ LHS has values:
   Right aW
   Left (BBL ())
   Left (BBR aB)
-  Left (BBB () aB)
+  - Left (BBB () aB)
   Left (BWR aW)
 
 RHS has values:
   Right aB
   Left (WBL ())
   Left (WWR aW)
-  Left (WBW () aW)
+  - Left (WBW () aW)
   Left (WBR aB)
 
-Do they match up? Here is a possible match:
-
-  Right aW            Left (WWR aW)
-  Left (BBL ())       Left (WBL ())
-  Left (BBR aB)       Right aB
-  Left (BBB () aB)    Left (WBR aB)
-  Left (BWR aW)       Left (WBW () aW)
+Do they match up? No. We have three incoming aW on the left and one outgoing
+on the right!
 
 --}
 
+timesA :: (a ~ (aBlack :|: aWhite), b ~ (bBlack :|: bWhite), 
+           c ~ (cBlack :|: cWhite), d ~ (dBlack :|: dWhite)) =>
+          M a b -> M c d -> M (TimesA a c) (TimesA b d)
+timesA = undefined
   
+{--
[…]

### 2014-05-18 Strategies.hs
reorder to make things line up better

@@ -222,10 +222,10 @@ Either (WhiteProduct (() :|: Void) (aB :|: aW)) aBlack
 and we have that:
 BlackProduct (() :|: Void) (aB :|: aW) has values:
   BBL ()
-  BBR aB
+  BWR aW
   - BBB () aB
   BWL Void
-  BWR aW
+  BBR aB
   - BWW Void aW
 WhiteProduct (() :|: Void) (aB :|: aW) has values:
   WBL ()
@@ -238,10 +238,10 @@ WhiteProduct (() :|: Void) (aB :|: aW) has values:
 LHS has values:
   Right aW
   Left (BBL ())
-  Left (BBR aB)
+  Left (BWR aW)
   - Left (BBB () aB)
   Left (BWL Void)
-  Left (BWR aW)
+  Left (BBR aB)
   - Left (BWW Void aW)
 
 RHS has values:
@@ -258,9 +258,9 @@ Get rid of the voids:
 LHS has values:
   Right aW
   Left (BBL ())
-  Left (BBR aB)
-  - Left (BBB () aB)
   Left (BWR aW)
+  - Left (BBB () aB)
+  Left (BBR aB)
 
 RHS has values:
   Right aB
@@ -272,6 +272,7 @@ RHS has values:
 Do they match up? No. We have three incoming aW on the left and one outgoing
 on the right!
 
+JC: uh?  It's now 2 and 2.
 --}
 
 timesA :: (a ~ (aBlack :|: aWhite), b ~ (bBlack :|: bWhite), 

### 2014-05-18 ring.tex
trees.hs

--- 2014-05-06 negatives.tex
@@
 
-%%
-
-\newcommand{\geqz}{\geq\zerog}
-\newcommand{\leqz}{\leq\zerog}
-\newcommand{\zerog}{\mathbb{0}}
-\newcommand{\oneg}{\mathbb{1}}
-\newcommand{\twog}{\mathbb{2}}
-\newcommand{\threeg}{\mathbb{3}}
-\newcommand{\fourg}{\mathbb{4}}
-\newcommand{\fiveg}{\mathbb{5}}
-
-%%
-
+\newtheorem{definition}{Definition}
 \begin{document}
@@
 
+Ignore function types for now. Finite algebraic types (constructed from 0, 1,
+sum, and product) are positive types; they represent the program (as opposed
+to the environment), the player (as opposed to the opponent), the producer
+(as opposed to the consumer). Negative types represent the environment, the
+opponent, the consumer, etc. There is a nice symmetry here but that is
+difficult to see if the computational model has information
+effects. Intuitively, if the computational infrastructure allows arbitrary
+things to be discarded or duplicated then ``loans'' or ``debts'' must be
+introduced with extreme care. However in settings where resources are
+maintained, it is conceptually simpler to introduce ``debts'' and liberally
+compute with ``positive'' and ``negative'' values knowing that every demand
+for a value will be eventually matched up with a corresponding positive
+flow.
+
+Other than semantic elegance, the immediate significance is that we get
+higher-order functions for free as objects of type $-a+b$ expressing the
+demand for a value of type $a$ and the production of a value of type $b$.  We
+can see this beautiful symmetry and the emergence of functions from symmetry
+in linear logic. The geometry of interaction construction takes a fragment of
+linear logic and builds higher-order functions. It however doesn't work for
+full linear logic. The related Int construction does not work if we have two
+functors. This turns out to be a long-standing open problem in algebraic
+topology that was recently solved in the ``ring completion'' paper. So the
+technical question is how to take computational infrastructure where
+information is conserved and add these symmetries that lead to higher-order
+functions without losing the product structure.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conway Numbers}
 
-A Conway number is a game consisting of left and right options where each
-option models a move to another game. There is another technical condition
-given below that constrains every left option to be ``less than'' every right
-option but that can be ignored for the moment. Players alternate taking
-options and the player with no available options loses. In Haskell, ignoring
-the additional technical condition, one might define the datatype of Conway
-numbers as:
-\begin{code}
-data ConwayNumber = CN [ConwayNumber] [ConwayNumber]
-\end{code}
-The simplest Conway number is \lstinline|CN [] []| with empty left and right
-options. We call this number $\zerog$: 
-\begin{code}
-/+ $\zerog$ +/  = CN  [] []
-\end{code}
-Once we have defined $\zerog$, we can also define the following:
-\begin{code}
-/+ $\oneg$ +/  = CN  [ /+ $\zerog$ +/ ]  []
-/+ $\twog$ +/  = CN  [ /+ $\oneg$ +/ ]  []
-/+ $\threeg$ +/  = CN  [ /+ $\twog$ +/ ]  []
-\end{code}
-and so on. Intuitively, the number $\mathbb{n}$ represents a game in which
-the left player has an $n$-move advantage over the right player. Dually, we
-can define the following numbers where the left player has $n$-move
-\emph{disadvantage} over the right player:
-\begin{code}
-/+ -$\oneg$ +/  = CN  [] [ /+ $\zerog$ +/ ]
-/+ -$\twog$ +/  = CN  [] [ /+ -$\oneg$ +/ ]
-/+ -$\threeg$ +/  = CN  [] [ /+ -$\twog$ +/ ]
-\end{code}
-More generally, the \emph{unary negation} of a number or its \emph{dual} is
-defined as follows:
-\begin{code}
-neg :: ConwayNumber -> ConwayNumber
-neg (CN xls xrs) = 
-  CN [ neg xr | xr <- xrs ] [ neg xl | xl <- xls ]
-\end{code}
-
-\textbf{Comparing numbers.} There are three outcomes to games represented by
-Conway numbers $\mathbb{n}$:
-\begin{itemize}
-\item $\mathbb{n} > \zerog$ which indicates that the \emph{left} player can
-  enforce a win, no matter who starts;
-\item $\mathbb{n} < \zerog$ which indicates that the \emph{right} player can
-  enforce a win, no matter who starts;
-\item $\mathbb{n} = \zerog$ which indicates that the \emph{second}
-  player can enforce a win, no matter who.
-\end{itemize}
-For example, we have $\threeg > \zerog$ because the right player would have
-no options (i.e., would lose) on the first or second turn depending on who
-starts. Similarly, we have $-\threeg < \zerog$ because the left player would
-have no options on on the first or second turn depending on who starts. For
-$\zerog$, it is evident that the first player to take a turn must lose, i.e.,
-that the second player wins. More interestingly, consider the game
-$\textsf{CN}~[-\oneg]~[\oneg]$ which we claim is $= \zerog$. Indeed, if the
-left player starts, the only option is to move to the game $-\oneg$ which the
-right player wins. If however the right player moves first, the only option
-is to move to the game $\oneg$ which the left player wins. In both cases, the
-second player wins. To ensure that these three outcomes are the only possible
-outcomes, Conway numbers have an additional restriction: for every left
-option $x$ and right option $y$, we require that $x < y$. If that restriction
-is lifted we could construct games such that $\textsf{CN}~[\zerog]~[\zerog]$
-for which none of the relations $> \zerog$, $< \zerog$, or $= \zerog$
-holds. Instead for this game, the first player always wins, which means that
-the left player cannot enforce a win (in case the right player starts) and
-vice-versa.
-
-The formal definitions of the comparison operators are:
-\begin{code}
->=0 (CN _ xrs)  = not $ or (map <=0 xrs)
-<=0 (CN xls _)  = not $ or (map >=0 xls) 
-==0 x           = >=0 x && <=0 x
-<0 x            = <=0 x && not (>=0 x)
->0 x            = >=0 x && not (<=0 x)
-\end{code}
-Only $\geq_\zerog$ and $\leq_\zerog$ are fundamental. The predicate
-$\geq_\zerog$ means that the left player can win as second player and
-$\leq_\zerog$ means that the right player can win as second player. If the
-left player can win as second player, it means that the right player has no
-good opening move, i.e., that there is no right option that is $\leq_\zerog$.
-
-Conway numbers also come equipped with addition, subtraction, and
-multiplication operations defined as follows.
-
-\noindent\textbf{Addition and subtraction.} Addition of two games intuitively
-gives each player the choice of selecting an option from either
-game. Subtraction is simply the addition of the dual of a game. Formally:
-\begin{code}
-(:+:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
-x@(CN xls xrs) :+: y@(CN yls yrs) = 
-  CN 
-    ([ xl :+: y | xl <- xls ] `union`
-     [ x :+: yl | yl <- yls ])
-    ([ xr :+: y | xr <- xrs ] `union`
-     [ x :+: yr | yr <- yrs ])
-
-(:-:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
-x :-: y = x :+: (neg y) 
-\end{code}
-It is easy to check that $\zerog$ is the unit of addition and that addition
-is commutative. Furthermore, adding two positive numbers like~$\twog$
-and~$\threeg$ gives $\fiveg$ as desired, and similarly for two negative
-numbers. When mixing positive and negative numbers, e.g., adding $\threeg$
-and $-\twog$, the result appears much more complicated:
-\begin{code}
-CN [CN  [CN  [-/+ $\twog$ +/] 
-             [CN [-/+ $\oneg$ +/] [/+ $\oneg$ +/]]]
-        [CN  [CN [-/+ $\oneg$ +/] [/+ $\oneg$ +/]] 
-             [/+ $\twog$ +/]]] 
-   [CN  [CN  [CN [-/+ $\oneg$ +/] [/+ $\oneg$ +/]] 
-             [/+ $\twog$ +/]] 
-        [/+ $\threeg$ +/]]
-\end{code}
-We will formalize the sense in which the game above is equivalent to the game
-$\oneg$, but for now we can intuitively reason as follows. We have already
-seen that the game $\textsf{CN}~[-\oneg]~[\oneg]$ is equivalent to $\zerog$
-which simplifies the above to:
-\begin{code}
-CN [CN  [CN  [-/+ $\twog$ +/] [/+ $\zerog$ +/]]
-        [CN  [/+ $\zerog$ +/] [/+ $\twog$ +/]]]
-   [CN  [CN  [/+ $\zerog$ +/] [/+ $\twog$ +/]]
-        [/+ $\threeg$ +/]]
-\end{code}
-Now consider the game $\textsf{CN}~[\zerog]~[\twog]$. If the left player
-starts, the game proceeds to $\zerog$ and the right player loses. If the
-right player starts, the game proceeds to $\twog$ which is an even stronger
-position for the left player. In other words, the right option is useless and
-the right player always loses: the game is equivalent to
-$\textsf{CN}~[\zerog]~[\;]$, i.e., to~$\oneg$. By a similar reasoning,
-$\textsf{CN}~[-\twog]~[\zerog]$ simplifies to $-\oneg$, and the whole
-expression becomes:
-\begin{code}
-CN [CN  [/+ -$\oneg$ +/] [/+ $\oneg$ +/]]
-   [CN  [/+ $\oneg$ +/] [/+ $\threeg$ +/]]
-\end{code}
-The left option is equivalent to $\zerog$ and the right one is equivalent to
-$\twog$, and the entire expression is equivalent to $\oneg$.
-
-\noindent\textbf{Multiplication.} The most complicated operation on Conway
-numbers is multiplication. It is defined as follows:
-\begin{code}
-(:*:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
-x@(CN xls xrs) :*: y@(CN yls yrs) = 
-  CN 
-    ([ (xl :*: y) :+: (x :*: yl) :-: (xl :*: yl)
-     | xl <- xls, yl <- yls] `union`
-     [ (xr :*: y) :+: (x :*: yr) :-: (xr :*: yr)
-     | xr <- xrs, yr <- yrs])
-    ([ (xl :*: y) :+: (x :*: yr) :-: (xl :*: yr)
-     | xl <- xls, yr <- yrs] `union`
-     [ (xr :*: y) :+: (x :*: yl) :-: (xr :*: yl)
-     | xr <- xrs, yl <- yls])
-\end{code}
-
-Conway numbers form a \emph{ring} under the semantic notion of equality
-\lstinline$:=:$ defined below:
-\begin{code}
-(:=:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
-x :=: y = ==0 (x :-: y)
-\end{code}
+\section{Conclusion}
 

### 2014-05-18 Trees.hs

@@ -1,0 +1,79 @@
+module Trees where
+
+{-- Our objects are complete binary trees of different heights. This is a
+tree of types and we want a function to summarize that whole tree using one
+type. 
+--}
+
+-- plain types: the names are for testing
+data T = Zero | One | Plus T T | Times T T | Name String
+  deriving Show
+
+data Obj = Leaf T 
+         | Node Obj Obj -- left is plus; right is minus
+  deriving Show
+
+-- examples
+
+-- zeroth level types
+
+t1 = Leaf (Plus One One)
+t2 = Leaf One
+t3 = Leaf (Times One (Plus One One))
+t4 = Leaf (Plus One (Plus One One))
+
+-- first level types (a-b)
+
+tt1 = Node t1 t2
+tt2 = Node t3 t4
+
+-- second level types ((a-b)-(c-d))
+
+ttt1 = Node tt1 tt2
+
+-- multiplication
+
+times :: Obj -> Obj -> Obj
+times (Leaf t1) (Leaf t2) = Leaf (Times t1 t2)
+times (Leaf t) (Node o1 o2) = Node (times (Leaf t) o1) (times (Leaf t) o2)
+times (Node o1 o2) o = Node (times o1 o) (times o2 o)
+
+test1 = times 
+          (Node (Leaf (Name "x+")) (Leaf (Name "y-")))
+          (Node (Node (Leaf (Name "a+")) (Leaf (Name "b-")))
+                (Node (Leaf (Name "c-")) (Leaf (Name "d+"))))
+
+{--
+Node (Node 
+       (Node (Leaf (Times (Name "x+") (Name "a+"))) 
+             (Leaf (Times (Name "x+") (Name "b-")))) 
+       (Node (Leaf (Times (Name "x+") (Name "c-"))) 
+             (Leaf (Times (Name "x+") (Name "d+"))))) 
+     (Node 
+       (Node (Leaf (Times (Name "y-") (Name "a+"))) 
+             (Leaf (Times (Name "y-") (Name "b-")))) 
+       (Node (Leaf (Times (Name "y-") (Name "c-"))) 
+             (Leaf (Times (Name "y-") (Name "d+")))))
+--}
+
+-- addition (extend the shorter tree) 
+
+add :: Obj -> Obj -> Obj
+add (Leaf t1) (Leaf t2) = Leaf (Plus t1 t2)
[…]

### 2014-05-18 Trees.hs
thinking about morphisms between ndim cubes

@@ -1,16 +1,18 @@
 module Trees where
 
-{-- Our objects are complete binary trees of different heights. This is a
-tree of types and we want a function to summarize that whole tree using one
-type. 
+{-- 
+
+Our objects are n-dimensional cubes represented as complete binary trees of
+different heights. 
+
 --}
 
 -- plain types: the names are for testing
 data T = Zero | One | Plus T T | Times T T | Name String
   deriving Show
 
-data Obj = Leaf T 
-         | Node Obj Obj -- left is plus; right is minus
+data NCube = Leaf T 
+           | Node NCube NCube -- left is plus; right is minus
   deriving Show
 
 -- examples
@@ -33,7 +35,7 @@ ttt1 = Node tt1 tt2
 
 -- multiplication
 
-times :: Obj -> Obj -> Obj
+times :: NCube -> NCube -> NCube
 times (Leaf t1) (Leaf t2) = Leaf (Times t1 t2)
 times (Leaf t) (Node o1 o2) = Node (times (Leaf t) o1) (times (Leaf t) o2)
 times (Node o1 o2) o = Node (times o1 o) (times o2 o)
@@ -58,7 +60,7 @@ Node (Node
 
 -- addition (extend the shorter tree) 
 
-add :: Obj -> Obj -> Obj
+add :: NCube -> NCube -> NCube
 add (Leaf t1) (Leaf t2) = Leaf (Plus t1 t2)
 add (Leaf t) (Node o1 o2) = add (Node (Leaf t) (Leaf Zero)) (Node o1 o2)
 add (Node o1 o2) (Leaf t) = add (Node o1 o2) (Node (Leaf t) (Leaf Zero))
@@ -77,3 +79,6 @@ Node (Node (Leaf (Plus (Name "x+") (Name "a+")))
            (Leaf (Plus Zero (Name "d+"))))
 
 --}
+
+-- What are the morphisms between these cubes? 
+

### 2014-05-18 ring.tex
nice beautiful figures

@@
 \usepackage{bbold}
+\usepackage{tikz}
 
-\newtheorem{definition}{Definition}
+\newcommand{\lolli}{\multimap} 
+\newcommand{\cubt}{\mathbb{T}}
+\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
+
 \begin{document}
@@
 %% double-blind reviewing...
-\title{Negative Types}
+\title{Cubical Types}
 \authorinfo{}{}{}
@@
 
-Ignore function types for now. Finite algebraic types (constructed from 0, 1,
-sum, and product) are positive types; they represent the program (as opposed
-to the environment), the player (as opposed to the opponent), the producer
-(as opposed to the consumer). Negative types represent the environment, the
-opponent, the consumer, etc. There is a nice symmetry here but that is
-difficult to see if the computational model has information
-effects. Intuitively, if the computational infrastructure allows arbitrary
-things to be discarded or duplicated then ``loans'' or ``debts'' must be
-introduced with extreme care. However in settings where resources are
-maintained, it is conceptually simpler to introduce ``debts'' and liberally
-compute with ``positive'' and ``negative'' values knowing that every demand
-for a value will be eventually matched up with a corresponding positive
-flow.
-
-Other than semantic elegance, the immediate significance is that we get
-higher-order functions for free as objects of type $-a+b$ expressing the
-demand for a value of type $a$ and the production of a value of type $b$.  We
-can see this beautiful symmetry and the emergence of functions from symmetry
-in linear logic. The geometry of interaction construction takes a fragment of
-linear logic and builds higher-order functions. It however doesn't work for
-full linear logic. The related Int construction does not work if we have two
-functors. This turns out to be a long-standing open problem in algebraic
-topology that was recently solved in the ``ring completion'' paper. So the
-technical question is how to take computational infrastructure where
-information is conserved and add these symmetries that lead to higher-order
-functions without losing the product structure.
+Int construction of G construction is neat. As Neel explains, given
+first-order types and feedback you get higher-order functions. But if you do
+the construction on the additive structure, you lose the multiplicative
+structure. It turns out that this is related to a deep open problem in
+algebraic topology and homotopy theory that was recently solved. We
+``translate'' that solution to a computational type-theoretic world. This has
+evident connections to homotopy (type?) theory that remain to be
+investigated.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Int Construction}
 
+Explain in detail perhaps with Haskell embedding and type functions etc. The
+key insight it to add ``negative'' types representing demand for values. This
+is how you get functions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Problem and the Intuitive Solution}
+
+We lose the multiplicative structure. No evident way to define the
+multiplication functor. Perhaps there is a clever way. But this turns out to
+be a well-known open problem. Review the problem and tell the story from the
+algebraic topology perspective.
+
+Regular first-order types are viewed as 1-dimensional cubes. Generalize to
+$n$-dimensional cubes.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The Formal Construction}
+
+We use $\tau$ for regular first-order types which include the empty type, the
+unit type, and conventional sum and product types.  We use $\cubt$ for
+general $n$-dimensional cubical types. An $n$-dimensional cubical type
+$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
+negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
+cubical types of a lower dimension; the 1-dimensional case reduces to plain
+first-order types. As we explain below, cubical types are closed under sums,
+differences, products, and can express higher-order functions $\cubt_1 \lolli
+\cubt_2$ as $(\ominus \cubt_1) \oplus \cubt_2$. The syntax of the types is
+summarized below:
+\[\begin{array}{rcl}
+\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \\
+\\
+\cubt &::=& \tau \alt \nodet{\cubt_1}{\cubt_2} 
+      \alt \cubt \oplus \cubt 
+      \alt \cubt \otimes \cubt 
+      \alt \ominus \cubt 
+\end{array}\]
+Examples of 1-dimensional cubical types include all the regular first-order
+types $\tau$. A 2-dimensional cubical type looks like
+$\nodet{\tau_1}{\tau_2}$ which intuitively corresponds to the difference
+$\tau_1 - \tau_2$ of the two types. A 3-dimensional cubical type looks like
+$\nodet{(\nodet{\tau_1}{\tau_2})}{(\nodet{\tau_3}{\tau_4})}$ which
+intuitively corresponds to the iterated difference of types
+$(\tau_1-\tau_2)-(\tau_3-\tau_4)$ where the successive colors from the root
+encode the sign as follows: a color that is the same as its enclosing color
+is positive and a color that differs from its enclosing color is negative. It
+is possible for cubical types not be balanced in general but it is always
+possible to extend the shorter subtree to an arbitrary depth using the fact
+(to be justified later) that~$\tau$ is equivalent to $\nodet{\tau}{0}$.
+
+We now explain how cubical types are closed under sum, difference, and
+products. In the recursive definitions below, we assume that the types have
+been extended to be balanced as needed:
+\[\begin{array}{rcl}
+\ominus~\tau &=& \tau \\
+\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
+\\
+\tau_1 \oplus \tau_2 &=& \tau_1 + \tau_2 \\
+(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
+  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
+\\
+\tau_1 \otimes \tau_2 &=& \tau_1 * \tau_2 \\
+\tau \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
+  \nodet{\tau \otimes \cubt_1}{\tau \otimes \cubt_2} \\
+(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
+  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
+\end{array}\]
+
+\begin{figure*}
+\[\begin{array}{c}
+\nodet{\tau_1}{\tau_2}
+\quad\otimes\quad
+\nodet{(\nodet{\tau_3}{\tau_4})}{(\nodet{\tau_5}{\tau_6})} = 
+\\
+\nodet{(\nodet{{(\nodet{\tau_1*\tau_3}{\tau_1*\tau_4})}}
+              {{(\nodet{\tau_1*\tau_5}{\tau_1*\tau_6})}})}
+      {(\nodet{{(\nodet{\tau_2*\tau_3}{\tau_2*\tau_4})}}
+              {{(\nodet{\tau_2*\tau_5}{\tau_2*\tau_6})}})}
+\end{array}\]
+\caption{\label{mult}Example of multiplication of two cubical types.}
+\end{figure*}
+
+A few examples might provide some needed intuition. First we note that the
+general definitions of sum and product reduce to the conventional sums and
+products for 1-dimensional types. Also the sum of 2-dimensional types reduces
+to the sum used in the \textbf{Int} construction. The interesting new
+development is the product definition. The example in Fig.~\ref{mult} shows
+that in general the product of an $n$-dimensional type with an
+$m$-dimensional type gives an $n+m$-dimensional type. For low dimensions such
+as the ones used in the example above, it is possible to visualize what is
+happening in coordinate space.
+
+\begin{figure*}
+\begin{tikzpicture}
+\node[left] at (0,0) {$\tau_1^+$};
+\node[right] at (0.6,0) {$\tau_2^-$};
+\draw[-] (0,0) -- (0.6,0);
+\node at (1.6,0) {$\otimes$}; 
+%%
+\node[below left] at (2.5,-0.5) {$\tau_3^+$};
+\node[below right] at (3.5,-0.5) {$\tau_4^-$};
+\draw[-] (2.5,-0.5) -- (3.5,-0.5);
+\draw[-] (2.5,-0.5) -- (2.5,0.5);
+\node[above left] at (2.5,0.5) {$\tau_6^-$};
+\node[above right] at (3.5,0.5) {$\tau_5^+$};
+\draw[-] (2.5,0.5) -- (3.5,0.5);
+\draw[-] (3.5,-0.5) -- (3.5,0.5);
+%% 
+\node at (4.5,0) {$=$};
+%% 
+\node[left] at (7.5,0.75) {$(\tau_2*\tau_3)^-$};
+\node[right] at (9.5,0.75) {$(\tau_2*\tau_4)^+$};
+\node[above right] at (10.2,1.2) {$(\tau_2*\tau_5)^-$};
+\node[above left] at (8.2,1.2) {$(\tau_2*\tau_6)^+$};
+%%
+\node[left] at (7.5,-0.75) {$(\tau_1*\tau_3)^+$};
+\node[right] at (9.5,-0.75) {$(\tau_1*\tau_4)^-$};
+\node[above right] at (10.2,-0.3) {$(\tau_1*\tau_5)^+$};
+\node[above] at (8.2,-0.3) {$(\tau_1*\tau_6)^-$};
+%%
+\draw[-] (7.5,0.75) -- (9.5,0.75);
+\draw[-] (9.5,0.75) -- (10.2,1.2);
+\draw[-] (10.2,1.2) -- (8.2,1.2);
+\draw[-] (8.2,1.2) -- (7.5,0.75);
+%%
+\draw[-] (7.5,-0.75) -- (9.5,-0.75);
+\draw[-] (9.5,-0.75) -- (10.2,-0.3);
+\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
+\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
+%%
+\draw[-] (7.5,0.75) -- (7.5,-0.75);
+\draw[-] (9.5,0.75) -- (9.5,-0.75);
+\draw[-] (10.2,1.2) -- (10.2,-0.3);
+\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
+\end{tikzpicture}
+\caption{\label{multv}Visualization of multiplication of two cubical types.}
+\end{figure*}
+
+\begin{verbatim}
+There are zero objects of all depths:
+  0, (0-0), (0-0)-0, ...
+There are 1 objects of all depths:
+  1, (1-0), (1-0)-0, ...
+
+To complete the story we need to 
+define morphisms. (More on this below.)
+Once we have a notion of morphism we 
+can check whether X + 0 is the same
+as X etc. i.e., we can check all the 
+ring equivalences.
+
+Ok so what are the morphisms between 
+these cubical objects? We know what
+they are for 1-dimensional cubes: 
+they are the pi combinabors. We also
+know what they are for the 2-dimensional 
+cubes: a maps (A-B) ==> (C-D) 
+is a Pi map between A+D <=> C+B. 
+How to generalize this?
+
+Why is there no trace in the ring 
+completion paper??? What are 
+the morphisms in that paper?
+\end{verbatim}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A Reversible Language with Cubical Types}
+
+Pi with cubical types
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related Work and Context}
+
+A ton of stuff here.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}

### 2014-05-18 Trees.hs

@@ -82,3 +82,10 @@ Node (Node (Leaf (Plus (Name "x+") (Name "a+")))
 
 -- What are the morphisms between these cubes? 
 
+{--
+
+mor :: NCube -> NCube -> NMorphism
+mor (Leaf t1) (Leaf t2) = Base (t1 :<=> t2) 
+mor (Node o1 o2) (Node o3 o4) = mor (Either o1 o4) (Either o3 o2)
+
+--}

### 2014-05-19 ring.tex
small typos

@@
 %% double-blind reviewing...
-\title{Cubical Types}
+\title{Cubical Types with Positive and Negative Polarities}
 \authorinfo{}{}{}
@@
 
-Int construction of G construction is neat. As Neel explains, given
-first-order types and feedback you get higher-order functions. But if you do
-the construction on the additive structure, you lose the multiplicative
-structure. It turns out that this is related to a deep open problem in
-algebraic topology and homotopy theory that was recently solved. We
-``translate'' that solution to a computational type-theoretic world. This has
-evident connections to homotopy (type?) theory that remain to be
-investigated.
+The \textbf{Int} construction or the $\mathcal{G}$ construction are neat. As
+Neel K. explains, given first-order types and feedback you get higher-order
+functions. But if you do the construction on the additive structure, you lose
+the multiplicative structure. It turns out that this is related to a deep
+open problem in algebraic topology and homotopy theory that was recently
+solved. We ``translate'' that solution to a computational type-theoretic
+world. This has evident connections to homotopy (type?) theory that remain to
+be investigated.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Int Construction}
+\section{The \textbf{Int} Construction}
 
@@
 $(\tau_1-\tau_2)-(\tau_3-\tau_4)$ where the successive colors from the root
-encode the sign as follows: a color that is the same as its enclosing color
-is positive and a color that differs from its enclosing color is negative. It
-is possible for cubical types not be balanced in general but it is always
-possible to extend the shorter subtree to an arbitrary depth using the fact
-(to be justified later) that~$\tau$ is equivalent to $\nodet{\tau}{0}$.
+encode the sign as follows: white in the outermost box corresponds to a
+positive polarity and gray in the outermost box corresponds to a negative
+polarity; then, as we move inwards the polarities alternate for every change
+of color. It is possible for cubical types not be balanced in general but it
+is always possible to extend the shorter subtree to an arbitrary depth using
+the fact (to be justified later) that~$\tau$ is equivalent to
+$\nodet{\tau}{0}$.
 
@@
 that in general the product of an $n$-dimensional type with an
-$m$-dimensional type gives an $n+m$-dimensional type. For low dimensions such
-as the ones used in the example above, it is possible to visualize what is
-happening in coordinate space.
+$m$-dimensional type gives an $(n+m)$-dimensional type. For low dimensions
+such as the ones used in the example above, it is possible to visualize what
+is happening in coordinate space (see Fig.~\ref{multv}).
 

### 2014-05-19 ring.tex
separate syntax from denotation

@@
 \newcommand{\cubt}{\mathbb{T}}
+\newcommand{\den}[1]{\llbracket #1 \rrbracket}
+\newcommand{\denc}[1]{\llbracket #1 \rrbracket^c}
 \newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
@@
 
-We use $\tau$ for regular first-order types which include the empty type, the
-unit type, and conventional sum and product types.  We use $\cubt$ for
-general $n$-dimensional cubical types. An $n$-dimensional cubical type
-$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
-negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
-cubical types of a lower dimension; the 1-dimensional case reduces to plain
-first-order types. As we explain below, cubical types are closed under sums,
-differences, products, and can express higher-order functions $\cubt_1 \lolli
-\cubt_2$ as $(\ominus \cubt_1) \oplus \cubt_2$. The syntax of the types is
-summarized below:
+We use $t$ for regular first-order types which include the empty type, the
+unit type, and conventional sum and product types.  We use $\tau$ for cubical
+types: these have the same formal structure as regular types but also include
+\emph{negative} types. The syntax of the types is summarized below:
 \[\begin{array}{rcl}
-\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \\
-\\
-\cubt &::=& \tau \alt \nodet{\cubt_1}{\cubt_2} 
-      \alt \cubt \oplus \cubt 
-      \alt \cubt \otimes \cubt 
-      \alt \ominus \cubt 
+t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
+\tau &::=& t \alt 
+      \alt \tau_1 + \tau_2
+      \alt \tau_1 * \tau_2
+      \alt - \tau
 \end{array}\]
-Examples of 1-dimensional cubical types include all the regular first-order
-types $\tau$. A 2-dimensional cubical type looks like
-$\nodet{\tau_1}{\tau_2}$ which intuitively corresponds to the difference
-$\tau_1 - \tau_2$ of the two types. A 3-dimensional cubical type looks like
-$\nodet{(\nodet{\tau_1}{\tau_2})}{(\nodet{\tau_3}{\tau_4})}$ which
-intuitively corresponds to the iterated difference of types
-$(\tau_1-\tau_2)-(\tau_3-\tau_4)$ where the successive colors from the root
-encode the sign as follows: white in the outermost box corresponds to a
-positive polarity and gray in the outermost box corresponds to a negative
-polarity; then, as we move inwards the polarities alternate for every change
-of color. It is possible for cubical types not be balanced in general but it
-is always possible to extend the shorter subtree to an arbitrary depth using
-the fact (to be justified later) that~$\tau$ is equivalent to
-$\nodet{\tau}{0}$.
+We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
+interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
 
-We now explain how cubical types are closed under sum, difference, and
-products. In the recursive definitions below, we assume that the types have
-been extended to be balanced as needed:
+It is easy to find a model for regular types: finite sets are such a
+model. In more detail:
+\[\begin{array}{rcl}
+\den{0} &=& \emptyset \\
+\den{1} &=& \{ \bullet \} \\
+\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
+\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
+\end{array}\]
+The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
+of types maps to disjoint union of sets. The product of types maps to the
+cartesian product of sets.
+
+For cubical types, their denotations are $n$-dimensional cubical sets with
+polarities, written as $\nodet{\cubt_1}{\cubt_2}$ An $n$-dimensional cubical
+set $\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and
+a negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
+cubical set of a lower dimension; the 1-dimensional case reduces to plain
+sets (which are the denotations of first-order types). As we explain below,
+cubical sets are closed under sums, differences, products.
+
+Examples of 1-dimensional cubical sets include all the finite sets $S$
+denoting regular first-order types $t$. A 2-dimensional cubical set looks
+like $\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1
+- t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
+3-dimensional cubical type looks like
+$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
+corresponds to the iterated difference of the appropriate types
+$(t_1-t_2)-(t_3-t_4)$ where the successive colors from the root encode the
+sign as follows: white in the outermost box corresponds to a positive
+polarity and gray in the outermost box corresponds to a negative polarity;
+then, as we move inwards the polarities alternate for every change of
+color.
+
+We now give the denotation of cubical types:
+\[\begin{array}{rcl}
+\denc{t} &=& \den{t} = \nodet{\den{t}}{0} = 
+  \nodet{\nodet{\den{t}}{0}}{0} = \ldots \\
+\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
+\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
+\denc{- \tau} &=& \ominus \denc{\tau}
+\end{array}\]
+where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
+It is possible for cubical types not be balanced in general but it is always
+possible to extend the shorter subtree to an arbitrary depth using the first
+clause which formalizes that $t$ is equivalent to $t-0$ and to $(t-0)-0$ etc.
+In the recursive definitions below, we assume that the types have been
+extended to be balanced as needed:
 \[\begin{array}{rcl}
@@
 \[\begin{array}{c}
-\nodet{\tau_1}{\tau_2}
+\nodet{S_1}{S_2}
 \quad\otimes\quad
-\nodet{(\nodet{\tau_3}{\tau_4})}{(\nodet{\tau_5}{\tau_6})} = 
+\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} = 
 \\
-\nodet{(\nodet{{(\nodet{\tau_1*\tau_3}{\tau_1*\tau_4})}}
-              {{(\nodet{\tau_1*\tau_5}{\tau_1*\tau_6})}})}
-      {(\nodet{{(\nodet{\tau_2*\tau_3}{\tau_2*\tau_4})}}
-              {{(\nodet{\tau_2*\tau_5}{\tau_2*\tau_6})}})}
+\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
+              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
+      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
+              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
 \end{array}\]
@@
 
-A few examples might provide some needed intuition. First we note that the
-general definitions of sum and product reduce to the conventional sums and
-products for 1-dimensional types. Also the sum of 2-dimensional types reduces
-to the sum used in the \textbf{Int} construction. The interesting new
-development is the product definition. The example in Fig.~\ref{mult} shows
-that in general the product of an $n$-dimensional type with an
-$m$-dimensional type gives an $(n+m)$-dimensional type. For low dimensions
-such as the ones used in the example above, it is possible to visualize what
-is happening in coordinate space (see Fig.~\ref{multv}).
+The definition of negation is natural: it recursively swaps the positive and
+negative subspaces. We note that the general definitions of sum and product
+reduce to the conventional sums and products for 1-dimensional types. Also
+the sum of 2-dimensional types reduces to the sum used in the \textbf{Int}
+construction. The interesting new development is the product definition. The
+example in Fig.~\ref{mult} shows that in general the product of an
+$n$-dimensional type with an $m$-dimensional type gives an
+$(n+m)$-dimensional type. For low dimensions such as the ones used in the
+example above, it is possible to visualize what is happening in coordinate
+space (see Fig.~\ref{multv}).
 
@@
 \begin{tikzpicture}
-\node[left] at (0,0) {$\tau_1^+$};
-\node[right] at (0.6,0) {$\tau_2^-$};
+\node[left] at (0,0) {$S_1^+$};
+\node[right] at (0.6,0) {$S_2^-$};
 \draw[-] (0,0) -- (0.6,0);
@@
 %%
-\node[below left] at (2.5,-0.5) {$\tau_3^+$};
-\node[below right] at (3.5,-0.5) {$\tau_4^-$};
+\node[below left] at (2.5,-0.5) {$S_3^+$};
+\node[below right] at (3.5,-0.5) {$S_4^-$};
 \draw[-] (2.5,-0.5) -- (3.5,-0.5);
 \draw[-] (2.5,-0.5) -- (2.5,0.5);
-\node[above left] at (2.5,0.5) {$\tau_6^-$};
-\node[above right] at (3.5,0.5) {$\tau_5^+$};
+\node[above left] at (2.5,0.5) {$S_6^-$};
+\node[above right] at (3.5,0.5) {$S_5^+$};
 \draw[-] (2.5,0.5) -- (3.5,0.5);
@@
 %% 
-\node[left] at (7.5,0.75) {$(\tau_2*\tau_3)^-$};
-\node[right] at (9.5,0.75) {$(\tau_2*\tau_4)^+$};
-\node[above right] at (10.2,1.2) {$(\tau_2*\tau_5)^-$};
-\node[above left] at (8.2,1.2) {$(\tau_2*\tau_6)^+$};
+\node[left] at (7.5,0.75) {$(S_2 \times S_3)^-$};
+\node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
+\node[above right] at (10.2,1.2) {$(S_2 \times S_5)^-$};
+\node[above left] at (8.2,1.2) {$(S_2 \times S_6)^+$};
 %%
-\node[left] at (7.5,-0.75) {$(\tau_1*\tau_3)^+$};
-\node[right] at (9.5,-0.75) {$(\tau_1*\tau_4)^-$};
-\node[above right] at (10.2,-0.3) {$(\tau_1*\tau_5)^+$};
-\node[above] at (8.2,-0.3) {$(\tau_1*\tau_6)^-$};
+\node[left] at (7.5,-0.75) {$(S_1 \times S_3)^+$};
+\node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
+\node[above right] at (10.2,-0.3) {$(S_1 \times S_5)^+$};
+\node[above] at (8.2,-0.3) {$(S_1 \times S_6)^-$};
 %%

### 2014-05-19 ring.tex
separate syntax from denotation

@@
 We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
-interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
+interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$. At
+the level of cubical types, we can write types such as
+$(t_1-t_2)-(t_3-t_4*t_5)$ which mix negative types with the usual type
+constructors.
 
@@
 of types maps to disjoint union of sets. The product of types maps to the
-cartesian product of sets.
+cartesian product of sets. We usually use $S$ for the denotation of a
+conventional first-order type.
 
-For cubical types, their denotations are $n$-dimensional cubical sets with
-polarities, written as $\nodet{\cubt_1}{\cubt_2}$ An $n$-dimensional cubical
-set $\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and
-a negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
-cubical set of a lower dimension; the 1-dimensional case reduces to plain
-sets (which are the denotations of first-order types). As we explain below,
-cubical sets are closed under sums, differences, products.
+For cubical types, their denotations, usually written as $\cubt$, are
+$n$-dimensional cubical sets with polarities. An $n$-dimensional cubical set
+$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
+negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
+cubical set of a lower dimension; the 1-dimensional case reduces to
+$\nodet{S}{\emptyset}$ which is essentially the denotation of conventional
+first-order types. As we explain below, cubical sets are closed under sums,
+differences, products.
 
-Examples of 1-dimensional cubical sets include all the finite sets $S$
-denoting regular first-order types $t$. A 2-dimensional cubical set looks
-like $\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1
-- t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
-3-dimensional cubical type looks like
-$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
+Examples of 1-dimensional cubical sets include all the sets of the form
+$\nodet{S}{\emptyset}$ where $S$ is finite set denoting a regular first-order
+types $t$. A 2-dimensional cubical set looks like $\nodet{S_1}{S_2}$ which
+intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
+denotations are $S_1$ and $S_2$ respectively. A 3-dimensional cubical type
+looks like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
 corresponds to the iterated difference of the appropriate types
@@
 polarity and gray in the outermost box corresponds to a negative polarity;
-then, as we move inwards the polarities alternate for every change of
-color.
+then, as we move inwards the polarities alternate for every change of color.
 
@@
 \[\begin{array}{rcl}
-\denc{t} &=& \den{t} = \nodet{\den{t}}{0} = 
-  \nodet{\nodet{\den{t}}{0}}{0} = \ldots \\
+\denc{t} &=& \nodet{\den{t}}{\emptyset} = 
+  \nodet{\nodet{\den{t}}{\emptyset}}{\emptyset} = \ldots \\
 \denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
@@
 It is possible for cubical types not be balanced in general but it is always
-possible to extend the shorter subtree to an arbitrary depth using the first
-clause which formalizes that $t$ is equivalent to $t-0$ and to $(t-0)-0$ etc.
-In the recursive definitions below, we assume that the types have been
-extended to be balanced as needed:
+possible to embed an $n$-dimensional set into a larger dimension using the
+equivalence in the first clause which formalizes that $t$ is equivalent to
+$t-0$ and to $(t-0)-0$ etc.  In the recursive definitions below, we assume
+that the types have been extended to be balanced as needed:
 \[\begin{array}{rcl}
-\ominus~\tau &=& \tau \\
+\ominus~S &=& S \\
 \ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
 \\
-\tau_1 \oplus \tau_2 &=& \tau_1 + \tau_2 \\
+S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
 (\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
@@
 \\
-\tau_1 \otimes \tau_2 &=& \tau_1 * \tau_2 \\
-\tau \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
-  \nodet{\tau \otimes \cubt_1}{\tau \otimes \cubt_2} \\
+S_1 \otimes S_2 &=& S_1 \times S_2 \\
+S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
+  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
 (\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 

### 2014-05-19 ring.tex
separate syntax from denotation

@@
 cubical set of a lower dimension; the 1-dimensional case reduces to
-$\nodet{S}{\emptyset}$ which is essentially the denotation of conventional
+$\nodet{S}{~}$ which is essentially the denotation of conventional
 first-order types. As we explain below, cubical sets are closed under sums,
@@
 Examples of 1-dimensional cubical sets include all the sets of the form
-$\nodet{S}{\emptyset}$ where $S$ is finite set denoting a regular first-order
+$\nodet{S}{~}$ where $S$ is finite set denoting a regular first-order
 types $t$. A 2-dimensional cubical set looks like $\nodet{S_1}{S_2}$ which
@@
 \[\begin{array}{rcl}
-\denc{t} &=& \nodet{\den{t}}{\emptyset} = 
-  \nodet{\nodet{\den{t}}{\emptyset}}{\emptyset} = \ldots \\
+\denc{t} &=& \nodet{\den{t}}{~} = 
+  \nodet{\nodet{\den{t}}{~}}{~} = \ldots \\
+\\
 \denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\

### 2014-05-19 ring.tex
separate syntax from denotation

@@
 cubical set of a lower dimension; the 1-dimensional case reduces to
-$\nodet{S}{~}$ which is essentially the denotation of conventional
-first-order types. As we explain below, cubical sets are closed under sums,
-differences, products.
+$\nodet{S}{\phantom{\emptyset}}$ which is essentially the denotation of
+conventional first-order types. As we explain below, cubical sets are closed
+under sums, differences, products.
 
 Examples of 1-dimensional cubical sets include all the sets of the form
-$\nodet{S}{~}$ where $S$ is finite set denoting a regular first-order
-types $t$. A 2-dimensional cubical set looks like $\nodet{S_1}{S_2}$ which
-intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
-denotations are $S_1$ and $S_2$ respectively. A 3-dimensional cubical type
-looks like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
+$\nodet{S}{\phantom{\emptyset}}$ where $S$ is finite set denoting a regular
+first-order types $t$. A 2-dimensional cubical set looks like
+$\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1 -
+t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
+3-dimensional cubical type looks like
+$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
 corresponds to the iterated difference of the appropriate types
@@
 \[\begin{array}{rcl}
-\denc{t} &=& \nodet{\den{t}}{~} = 
-  \nodet{\nodet{\den{t}}{~}}{~} = \ldots \\
+\denc{t} &=& \nodet{\den{t}}{\phantom{\emptyset}} = 
+  \nodet{\nodet{\den{t}}{\phantom{\emptyset}}}{\phantom{\emptyset}} = \ldots \\
 \\

### 2014-05-19 ring.tex
separate syntax from denotation

--- 2014-05-19 ring.tex
@@
 cubical set of a lower dimension; the 1-dimensional case reduces to
-$\nodet{S}{~}$ which is essentially the denotation of conventional
-first-order types. As we explain below, cubical sets are closed under sums,
-differences, products.
+$\nodet{S}{\phantom{S}}$ which is essentially the denotation of
+conventional first-order types. As we explain below, cubical sets are closed
+under sums, differences, products.
 
 Examples of 1-dimensional cubical sets include all the sets of the form
-$\nodet{S}{~}$ where $S$ is finite set denoting a regular first-order
-types $t$. A 2-dimensional cubical set looks like $\nodet{S_1}{S_2}$ which
-intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
-denotations are $S_1$ and $S_2$ respectively. A 3-dimensional cubical type
-looks like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
+$\nodet{S}{\phantom{S}}$ where $S$ is finite set denoting a regular
+first-order types $t$. A 2-dimensional cubical set looks like
+$\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1 -
+t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
+3-dimensional cubical type looks like
+$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
 corresponds to the iterated difference of the appropriate types
@@
 \[\begin{array}{rcl}
-\denc{t} &=& \nodet{\den{t}}{~} = 
-  \nodet{\nodet{\den{t}}{~}}{~} = \ldots \\
+\denc{t} &=& \nodet{\den{t}}{\phantom{\den{t}}} = 
+  \nodet{\nodet{\den{t}}{\phantom{\den{t}}}}{\phantom{\den{t}}} = \ldots \\
 \\

### 2014-05-19 ring.tex
separate syntax from denotation
fixed several typos

--- 2014-05-19 ring.tex
@@
 
-Regular first-order types are viewed as 1-dimensional cubes. Generalize to
+Regular first-order types are viewed as 0-dimensional cubes. Generalize to
 $n$-dimensional cubes.
@@
 negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
-cubical set of a lower dimension; the 1-dimensional case reduces to
-$\nodet{S}{~}$ which is essentially the denotation of conventional
-first-order types. As we explain below, cubical sets are closed under sums,
-differences, products.
+cubical set of a lower dimension; the 0-dimensional case reduces to the
+denotation of conventional first-order types. As we explain below, cubical
+sets are closed under sums, differences, products.
 
-Examples of 1-dimensional cubical sets include all the sets of the form
-$\nodet{S}{~}$ where $S$ is finite set denoting a regular first-order
-types $t$. A 2-dimensional cubical set looks like $\nodet{S_1}{S_2}$ which
-intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
-denotations are $S_1$ and $S_2$ respectively. A 3-dimensional cubical type
-looks like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
+Examples of 0-dimensional cubical sets include all the finite sets denoting a
+regular first-order type. A 1-dimensional cubical set looks like
+$\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1 -
+t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
+2-dimensional cubical type looks like
+$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
 corresponds to the iterated difference of the appropriate types
-$(t_1-t_2)-(t_3-t_4)$ where the successive colors from the root encode the
-sign as follows: white in the outermost box corresponds to a positive
-polarity and gray in the outermost box corresponds to a negative polarity;
-then, as we move inwards the polarities alternate for every change of color.
+$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
+outermost box encodes the sign.
 
@@
 \[\begin{array}{rcl}
-\denc{t} &=& \nodet{\den{t}}{~} = 
-  \nodet{\nodet{\den{t}}{~}}{~} = \ldots \\
-\\
+\denc{t} &=& \den{t} \\
 \denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
@@
 possible to embed an $n$-dimensional set into a larger dimension using the
-equivalence in the first clause which formalizes that $t$ is equivalent to
-$t-0$ and to $(t-0)-0$ etc.  In the recursive definitions below, we assume
-that the types have been extended to be balanced as needed:
+fact (to be justified later) that $t$ is equivalent to $t-0$ and to $(t-0)-0$
+and so on, i.e.,
+\[
+\den{t} = \nodet{\den{t}}{\phantom{\den{t}}} = 
+  \nodet{\nodet{\den{t}}{\phantom{\den{t}}}}
+        {\phantom{{\nodet{\den{t}}{\phantom{\den{t}}}}}} = \ldots \\
+\]
+In the recursive definitions below, we assume that the types have been
+extended to be balanced as needed:
 \[\begin{array}{rcl}
@@
 \quad\otimes\quad
-\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} = 
+\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} \quad= \\
 \\
@@
 \end{array}\]
-\caption{\label{mult}Example of multiplication of two cubical types.}
+\caption{\label{mult}Example of multiplication of two cubical sets.}
 \end{figure*}
 
-The definition of negation is natural: it recursively swaps the positive and
-negative subspaces. We note that the general definitions of sum and product
-reduce to the conventional sums and products for 1-dimensional types. Also
-the sum of 2-dimensional types reduces to the sum used in the \textbf{Int}
-construction. The interesting new development is the product definition. The
-example in Fig.~\ref{mult} shows that in general the product of an
+\noindent The definition of negation is natural: it recursively swaps the
+positive and negative subspaces. We note that the general definitions of sum
+and product reduce to the conventional sums and products for 0-dimensional
+types. Also the sum of 1-dimensional types reduces to the sum used in the
+\textbf{Int} construction and that addition and negation do not increase the
+dimensions. The interesting new development is the product definition. The
+example in Fig.~\ref{mult} illustrates a general pattern: the product of an
 $n$-dimensional type with an $m$-dimensional type gives an
@@
 \draw[-] (2.5,-0.5) -- (2.5,0.5);
-\node[above left] at (2.5,0.5) {$S_6^-$};
-\node[above right] at (3.5,0.5) {$S_5^+$};
+\node[above left] at (2.5,0.5) {$S_5^-$};
+\node[above right] at (3.5,0.5) {$S_6^+$};
 \draw[-] (2.5,0.5) -- (3.5,0.5);
@@
 %% 
-\node at (4.5,0) {$=$};
+\node at (5,0) {$=$};
 %% 
@@
 \node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
-\node[above right] at (10.2,1.2) {$(S_2 \times S_5)^-$};
-\node[above left] at (8.2,1.2) {$(S_2 \times S_6)^+$};
+\node[above right] at (10.2,1.2) {$(S_2 \times S_6)^-$};
+\node[above left] at (8.2,1.2) {$(S_2 \times S_5)^+$};
 %%
@@
 \node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
-\node[above right] at (10.2,-0.3) {$(S_1 \times S_5)^+$};
-\node[above] at (8.2,-0.3) {$(S_1 \times S_6)^-$};
+\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)^+$};
+\node[above] at (8.2,-0.3) {$(S_1 \times S_5)^-$};
 %%
@@
 \end{tikzpicture}
-\caption{\label{multv}Visualization of multiplication of two cubical types.}
+\caption{\label{multv}Visualization of multiplication of two cubical sets.}
 \end{figure*}

### 2014-05-19 ring.tex
fixed several typos
fixed minus
labels with circles

@@
 where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
-It is possible for cubical types not be balanced in general but it is always
+It is possible for cubical sets not be balanced in general but it is always
 possible to embed an $n$-dimensional set into a larger dimension using the
@@
 \[
-\den{t} = \nodet{\den{t}}{\phantom{\den{t}}} = 
-  \nodet{\nodet{\den{t}}{\phantom{\den{t}}}}
-        {\phantom{{\nodet{\den{t}}{\phantom{\den{t}}}}}} = \ldots \\
+S = \nodet{S}{\phantom{S}} = 
+  \nodet{\nodet{S}{\phantom{S}}}
+        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
 \]
-In the recursive definitions below, we assume that the types have been
+In the recursive definitions below, we assume that the cubical sets have been
 extended to be balanced as needed:
 \[\begin{array}{rcl}
-\ominus~S &=& S \\
+\ominus~S &=& \nodet{\phantom{S}}{S} \\
 \ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
@@
 \node[left] at (0,0) {$S_1^+$};
+\draw[fill] (0,0) circle [radius=0.05];
 \node[right] at (0.6,0) {$S_2^-$};
+\draw[fill] (0.6,0) circle [radius=0.05];
 \draw[-] (0,0) -- (0.6,0);
 \node at (1.6,0) {$\otimes$};
+
 %%
 \node[below left] at (2.5,-0.5) {$S_3^+$};
+\draw[fill] (2.5,-0.5) circle [radius=0.05];
 \node[below right] at (3.5,-0.5) {$S_4^-$};
+\draw[fill] (3.5,-0.5) circle [radius=0.05];
 \draw[-] (2.5,-0.5) -- (3.5,-0.5);
@@
 \node[above left] at (2.5,0.5) {$S_5^-$};
+\draw[fill] (2.5,0.5) circle [radius=0.05];
 \node[above right] at (3.5,0.5) {$S_6^+$};
+\draw[fill] (3.5,0.5) circle [radius=0.05];
 \draw[-] (2.5,0.5) -- (3.5,0.5);
@@
 \node[left] at (7.5,0.75) {$(S_2 \times S_3)^-$};
+\draw[fill] (7.5,0.75) circle [radius=0.05];
 \node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
+\draw[fill] (9.5,0.75) circle [radius=0.05];
 \node[above right] at (10.2,1.2) {$(S_2 \times S_6)^-$};
+\draw[fill] (10.2,1.2) circle [radius=0.05];
 \node[above left] at (8.2,1.2) {$(S_2 \times S_5)^+$};
+\draw[fill] (8.2,1.2) circle [radius=0.05];
 %%
 \node[left] at (7.5,-0.75) {$(S_1 \times S_3)^+$};
+\draw[fill] (7.5,-0.75) circle [radius=0.05];
 \node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
+\draw[fill] (9.5,-0.75) circle [radius=0.05];
 \node[above right] at (10.2,-0.3) {$(S_1 \times S_6)^+$};
+\draw[fill] (10.2,-0.3) circle [radius=0.05];
 \node[above] at (8.2,-0.3) {$(S_1 \times S_5)^-$};
+\draw[fill] (8.2,-0.3) circle [radius=0.05];
 %%
@@
 \begin{verbatim}
-There are zero objects of all depths:
-  0, (0-0), (0-0)-0, ...
-There are 1 objects of all depths:
-  1, (1-0), (1-0)-0, ...
-
 To complete the story we need to 

### 2014-05-20 ring.tex
something about values

@@
 \usepackage{tikz}
+\usepackage{proof}
 
+\newcommand{\negp}[1]{\textit{neg}(#1)}
+\newcommand{\inl}[1]{\textsf{inl}~#1}
+\newcommand{\inr}[1]{\textsf{inr}~#1}
 \newcommand{\lolli}{\multimap} 
@@
 %% double-blind reviewing...
-\title{Cubical Types with Positive and Negative Polarities}
+\title{Cubical Polarized Types}
 \authorinfo{}{}{}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Formal Construction}
+\section{The Type Structure}
+
+We first define the syntax and semantics of types and then discuss type
+isomorphisms.
+
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Negative and Cubical Types}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Type Isomorphisms}
+
 \begin{verbatim}
@@
 the morphisms in that paper?
+
+The ring completion paper produces
+a simplicial category.
+
+p. 3 talks about the group cancellation
+as subcubes along the diagonal...
+
+We shouldn't focus on denotations. We want
+an operational semantics for pi with 
+negatives. The same way that Neel turned
+the G construction into code that does
+something neat; we want to turn that 
+ring completion construction into code
+that produces h.o. functions without
+losing the multiplicative structure.
+
+Show how to embed a square in 2D into
+each of the faces of the 3D cube.
+
+1D into 2D:
+
+(a-b) => (a-b)-(0-0)
+(a-b) => (a-0)-(b-0)
+(a-b) => (0-b)-(0-a)
+(a-b) => (0-0)-(b-a)
+
 \end{verbatim}
@@
 
-Pi with cubical types
+Now that the type structure is defined, we turn our attention to the notion
+of values. For conventional first-order types, the values are also
+conventional:
+\[
+\infer{() : 1}{} 
+\qquad
+\infer{\inl{v} : t_1 + t_2}{v : t_1}
+\qquad
+\infer{\inr{v} : t_1 + t_2}{v : t_2}
+\qquad
+\infer{(v_1,v_2) : t_1 * t_2}{v_1 : t_1 & v_2 : t_2}
+\]
+For cubical types, the values are more involved. Intuitively, a value of the
+$n$-dimensional type $\tau$ is an element of one of the sets located in a
+corner of the $n$-dimensional cube denoted by $\tau$. Thus to specify a
+value, we must specify one of the corners of the cube which can easily be
+done using a sequence $p$ of $+$ and $-$ polarities indicating how to
+navigate the cube in each successive dimension starting from a fixed origin
+to reach the desired corner. We denote such values as $v^{p}$ and use
+$\epsilon$ for the empty sequence of polarities. One can think of the
+sequence of signs as denoting a path in the binary tree representation of
+cubical types. Formally:
+\[\begin{array}{c}
+\infer[\textit{base}]{v^{\epsilon} : t}{v : t} \\
+\\
+\infer[\textit{left}]{(\inl{v^p})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_1}
+\qquad
+\infer[\textit{right}]{(\inr{v^p})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} \\
+\\
+\infer[\textit{prod}]{(v_1^{p_1},v_2^{p_2})^{p_1 \bullet p_2} : \tau_1 * \tau_2}
+      {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} \\
+\\
+\infer[\textit{neg}]{v^{\negp{p}} : - \tau}{v^p : \tau} 
+\end{array}\]
+The rule \textit{base} states that a conventional first-order value can be
+viewed as trivial 0-dimensional cubical value. The rules \textit{left} and
+\textit{right} reflect the fact that sums do not increase the dimension and
+works pointwise. The rule \textit{prod} is the most involved one: it
+increases the dimension as calculated below:
+\[\begin{array}{rcl}
+\epsilon \bullet p &=& p \\
+(+p_1) \bullet p_2 &=& + (p_1 \bullet p_2) \\
+(-p_1) \bullet p_2 &=& - (p_1 \bullet p_2) 
+\end{array}\]
+The rule \textit{neg} shows that a negative value $v$ is located at the
+``opposite'' corner of $v$ which is calculated as follows:
+\[\begin{array}{rcl}
+\negp{\epsilon} &=& -\epsilon \\
+\negp{+p} &=& -\negp{p} \\
+\negp{-p} &=& +\negp{p}
+\end{array}\]
 

### 2014-05-20 ring.tex
fixed values I think and started to think about combinators

@@
 \usepackage{proof}
+\usepackage{graphicx}
 
+\newcommand{\Rule}[4]{
+\makebox{{\rm #1}
+$\displaystyle
+\frac{\begin{array}{l}#2\\\end{array}}
+{\begin{array}{l}#3\\\end{array}}$
+ #4}}
+\newcommand{\proves}{\vdash}
+\newcommand{\symc}[1]{\mathit{sym}~#1}
+\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
+\newcommand{\adjoint}[1]{#1^{\dagger}}
+\newcommand{\iso}{\leftrightarrow}
+\newcommand{\identlp}{\mathit{identl}_+}
+\newcommand{\identrp}{\mathit{identr}_+}
+\newcommand{\swapp}{\mathit{swap}_+}
+\newcommand{\assoclp}{\mathit{assocl}_+}
+\newcommand{\assocrp}{\mathit{assocr}_+}
+\newcommand{\identlt}{\mathit{identl}_*}
+\newcommand{\identrt}{\mathit{identr}_*}
+\newcommand{\swapt}{\mathit{swap}_*}
+\newcommand{\assoclt}{\mathit{assocl}_*}
+\newcommand{\assocrt}{\mathit{assocr}_*}
+\newcommand{\distz}{\mathit{dist}_0}
+\newcommand{\factorz}{\mathit{factor}_0}
+\newcommand{\dist}{\mathit{dist}}
+\newcommand{\factor}{\mathit{factor}}
+\newcommand{\idc}{\mathit{id}}
+\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
+\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
+\newcommand{\mm}{\texttt{-}}
+\newcommand{\pp}{\texttt{+}}
 \newcommand{\negp}[1]{\textit{neg}(#1)}
@@
 be a well-known open problem. Review the problem and tell the story from the
-algebraic topology perspective.
-
-Regular first-order types are viewed as 0-dimensional cubes. Generalize to
-$n$-dimensional cubes.
+algebraic topology perspective. Regular first-order types are viewed as
+$0$-dimensional cubes. Generalize to $n$-dimensional cubes.
 
@@
 
-We first define the syntax and semantics of types and then discuss type
+We first define the syntax a semantic model of types and then discuss type
 isomorphisms.
@@
 We use $t$ for regular first-order types which include the empty type, the
-unit type, and conventional sum and product types.  We use $\tau$ for cubical
-types: these have the same formal structure as regular types but also include
-\emph{negative} types. The syntax of the types is summarized below:
-\[\begin{array}{rcl}
+unit type, and conventional sum and product types. We call these types
+$0$-dimensional. We use $\tau$ for cubical types of dimension 1 or higher:
+these have the same syntactic structure as regular types except that they
+also include \emph{negative} types. The syntax of the types is summarized
+below:
+\[\begin{array}{lrcl}
+\textit{types of dimension $0$} & 
   t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
-\tau &::=& t \alt 
-      \alt \tau_1 + \tau_2
-      \alt \tau_1 * \tau_2
-      \alt - \tau
+\textit{types of dimensions $\geq 1$} & 
+  \tau &::=& t \alt \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
 \end{array}\]
 We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
-interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$. At
-the level of cubical types, we can write types such as
-$(t_1-t_2)-(t_3-t_4*t_5)$ which mix negative types with the usual type
-constructors.
+interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
 
-It is easy to find a model for regular types: finite sets are such a
+It is easy to find a model for $0$-dimensional types: finite sets are such a
 model. In more detail:
@@
 of types maps to disjoint union of sets. The product of types maps to the
-cartesian product of sets. We usually use $S$ for the denotation of a
-conventional first-order type.
+cartesian product of sets. We usually use $S$ for the denotation of these
+conventional first-order types. Note that our (simple) semantics of types
+identifies several structurally different types such as $(1+(1+1))$ and
+$((1+1)+1)$. In some sense, this is innocent as the types are
+isomorphic. However, in the operational semantics discussed in the next
+section, we make the computational content of such type isomorphisms
+explicit.
 
-For cubical types, their denotations, usually written as $\cubt$, are
-$n$-dimensional cubical sets with polarities. An $n$-dimensional cubical set
-$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
-negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
-cubical set of a lower dimension; the 0-dimensional case reduces to the
-denotation of conventional first-order types. As we explain below, cubical
-sets are closed under sums, differences, products.
+For cubical types of dimension 1 or higher, their denotations, usually
+written as $\cubt$, are $n$-dimensional ``polarized cubes.'' Such an
+$n$-dimensional cube $\nodet{\cubt_1}{\cubt_2}$ consists of a positive
+subspace $\cubt_1$ and a negative subspace $\cubt_2$ (shaded in gray). Each
+of the subspaces are cubes of a lower dimension; the $0$-dimensional case
+reduces to the denotation of conventional first-order types. As we explain
+below, cubical sets are closed under sums, differences, products.
 
-Examples of 0-dimensional cubical sets include all the finite sets denoting a
-regular first-order type. A 1-dimensional cubical set looks like
-$\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1 -
-t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
-2-dimensional cubical type looks like
+A 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which intuitively
+corresponds to the difference $t_1 - t_2$ of the two types whose denotations
+are $S_1$ and $S_2$ respectively. A 2-dimensional cube looks like
 $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
@@
 
-We now give the denotation of cubical types:
+We now give the full denotation of cubical types:
 \[\begin{array}{rcl}
-\denc{t} &=& \den{t} \\
+\denc{t} &=& \nodet{\den{t}}{\phantom{\den{t}}} \\
 \denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
@@
 where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
-It is possible for cubical sets not be balanced in general but it is always
-possible to embed an $n$-dimensional set into a larger dimension using the
-fact (to be justified later) that $t$ is equivalent to $t-0$ and to $(t-0)-0$
-and so on, i.e.,
+It is possible, in general, for the binary trees representing the cubes not
+be balanced. It is however always possible to embed an $n$-dimensional cube
+into a cube of a larger dimension using the fact (to be justified using a
+$\Pi$-combinator that witnesses this isomorphism in the next section) that
+$t$ is equivalent to $t-0$ and to $(t-0)-0$ and so on, i.e.,
 \[
@@
 \noindent The definition of negation is natural: it recursively swaps the
-positive and negative subspaces. We note that the general definitions of sum
-and product reduce to the conventional sums and products for 0-dimensional
+positive and negative subspaces. We note that the general definitions of sums
+and products reduce to the conventional sums and products for $0$-dimensional
 types. Also the sum of 1-dimensional types reduces to the sum used in the
 \textbf{Int} construction and that addition and negation do not increase the
-dimensions. The interesting new development is the product definition. The
-example in Fig.~\ref{mult} illustrates a general pattern: the product of an
-$n$-dimensional type with an $m$-dimensional type gives an
+dimensions. The interesting new development is the definition of
+products. The example in Fig.~\ref{mult} illustrates a general pattern: the
+product of an $n$-dimensional type with an $m$-dimensional type gives an
 $(n+m)$-dimensional type. For low dimensions such as the ones used in the
@@
 \begin{tikzpicture}
-\node[left] at (0,0) {$S_1^+$};
+\node[above] at (0,0) {\pp};
+\node[left] at (0,0) {$S_1$};
 \draw[fill] (0,0) circle [radius=0.05];
-\node[right] at (0.6,0) {$S_2^-$};
+\node[above] at (0.6,0) {\mm};
+\node[right] at (0.6,0) {$S_2$};
 \draw[fill] (0.6,0) circle [radius=0.05];
@@
 %%
-\node[below left] at (2.5,-0.5) {$S_3^+$};
+\node[left] at (2.5,-0.5) {\pp\pp};
+\node[below] at (2.5,-0.5) {$S_3$};
 \draw[fill] (2.5,-0.5) circle [radius=0.05];
-\node[below right] at (3.5,-0.5) {$S_4^-$};
+\node[right] at (3.5,-0.5) {\pp\mm};
+\node[below] at (3.5,-0.5) {$S_4$};
 \draw[fill] (3.5,-0.5) circle [radius=0.05];
@@
 \draw[-] (2.5,-0.5) -- (2.5,0.5);
-\node[above left] at (2.5,0.5) {$S_5^-$};
+\node[left] at (2.5,0.5) {\mm\pp};
+\node[above] at (2.5,0.5) {$S_5$};
 \draw[fill] (2.5,0.5) circle [radius=0.05];
-\node[above right] at (3.5,0.5) {$S_6^+$};
+\node[right] at (3.5,0.5) {\mm\mm};
+\node[above] at (3.5,0.5) {$S_6$};
 \draw[fill] (3.5,0.5) circle [radius=0.05];
@@
 %% 
-\node[left] at (7.5,0.75) {$(S_2 \times S_3)^-$};
+\node[left] at (7.5,0.75) {$(S_2 \times S_3)$\,\mm\pp\pp};
 \draw[fill] (7.5,0.75) circle [radius=0.05];
-\node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
+\node[right] at (9.5,0.75) {$(S_2 \times S_4)$\,\mm\pp\mm};
 \draw[fill] (9.5,0.75) circle [radius=0.05];
-\node[above right] at (10.2,1.2) {$(S_2 \times S_6)^-$};
+\node[above right] at (10.2,1.2) {$(S_2 \times S_6)$\,\mm\mm\mm};
 \draw[fill] (10.2,1.2) circle [radius=0.05];
-\node[above left] at (8.2,1.2) {$(S_2 \times S_5)^+$};
+\node[above left] at (8.2,1.2) {$(S_2 \times S_5)$\,\mm\mm\pp};
 \draw[fill] (8.2,1.2) circle [radius=0.05];
 %%
-\node[left] at (7.5,-0.75) {$(S_1 \times S_3)^+$};
+\node[left] at (7.5,-0.75) {$(S_1 \times S_3)$\,\pp\pp\pp};
 \draw[fill] (7.5,-0.75) circle [radius=0.05];
-\node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
+\node[right] at (9.5,-0.75) {$(S_1 \times S_4)$\,\pp\pp\mm};
 \draw[fill] (9.5,-0.75) circle [radius=0.05];
-\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)^+$};
+\node[above right] at (10.2,-0.3) {$(S_1 \times S_6)$\,\pp\mm\mm};
 \draw[fill] (10.2,-0.3) circle [radius=0.05];
-\node[above] at (8.2,-0.3) {$(S_1 \times S_5)^-$};
+\node[left] at (8.2,-0.3) {$(S_1 \times S_5)$\,\pp\mm\pp};
 \draw[fill] (8.2,-0.3) circle [radius=0.05];
@@
 
+We first define values then combinators that manipulate the values to witness
+the type isomorphisms.
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Values}
+
 Now that the type structure is defined, we turn our attention to the notion
@@
 \[\begin{array}{c}
-\infer[\textit{base}]{v^{\epsilon} : t}{v : t} \\
+\infer[\textit{base}]{v^{+} : t}{v : t} \\
 \\
-\infer[\textit{left}]{(\inl{v^p})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_1}
+\infer[\textit{left}]{(\inl{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_1}
 \qquad
-\infer[\textit{right}]{(\inr{v^p})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} \\
+\infer[\textit{right}]{(\inr{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} \\
 \\
-\infer[\textit{prod}]{(v_1^{p_1},v_2^{p_2})^{p_1 \bullet p_2} : \tau_1 * \tau_2}
+\infer[\textit{prod}]{(v_1^{p_1},v_2^{p_2})^{p_1 \cdot p_2} : \tau_1 * \tau_2}
       {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} \\
@@
 The rule \textit{base} states that a conventional first-order value can be
-viewed as trivial 0-dimensional cubical value. The rules \textit{left} and
+viewed as trivial $0$-dimensional cubical value. The rules \textit{left} and
 \textit{right} reflect the fact that sums do not increase the dimension and
 works pointwise. The rule \textit{prod} is the most involved one: it
-increases the dimension as calculated below:
+increases the dimension by \emph{concatenating} the two dimensions in its
+arguments. For example, if we pair $v_1^{\pp}$ and $v_2^{\mm\pp}$ we get
+$(v_1,v_2)^{\pp\mm\pp}$. The rule \textit{neg} shows that a negative value
+$v$ is located at the ``opposite'' corner of $v$ which is calculated as
+follows:
 \[\begin{array}{rcl}
-\epsilon \bullet p &=& p \\
-(+p_1) \bullet p_2 &=& + (p_1 \bullet p_2) \\
-(-p_1) \bullet p_2 &=& - (p_1 \bullet p_2) 
-\end{array}\]
-The rule \textit{neg} shows that a negative value $v$ is located at the
-``opposite'' corner of $v$ which is calculated as follows:
-\[\begin{array}{rcl}
-\negp{\epsilon} &=& -\epsilon \\
+\negp{\epsilon} &=& \epsilon \\
 \negp{+p} &=& -\negp{p} \\
 \negp{-p} &=& +\negp{p}
+\end{array}\]
+
+%%%%%%%%%%%%%%%%%%
+\subsection{Combinators}
+
+The terms of $\Pi$ witness type isomorphisms of the form $b \iso b$. They
+consist of base isomorphisms, as defined below, and their composition.
+\[\begin{array}{rrcll}
+\identlp :&  0 + b & \iso & b &: \identrp \\
+\swapp :&  b_1 + b_2 & \iso & b_2 + b_1 &: \swapp \\
+\assoclp :&  b_1 + (b_2 + b_3) & \iso & (b_1 + b_2) + b_3 &: \assocrp \\
+\identlt :&  1 * b & \iso & b &: \identrt \\
+\swapt :&  b_1 * b_2 & \iso & b_2 * b_1 &: \swapt \\
+\assoclt :&  b_1 * (b_2 * b_3) & \iso & (b_1 * b_2) * b_3 &: \assocrt \\
+\distz :&~ 0 * b & \iso & 0 &: \factorz \\
+\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor
+\end{array}\]
+Each line of the above table introduces a pair of dual
+constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that witness
+the type isomorphism in the middle.  These are the base (non-reducible) terms
+of the second, principal level of $\Pi$. Note how the above has two readings:
+first as a set of typing relations for a set of constants. Second, if these
+axioms are seen as universally quantified, orientable statements, they also
+induce transformations of the (traditional) values. The (categorical or
+homotopical) intuition here is that these axioms have computational content
+because they witness isomorphisms rather than merely stating an extensional
+equality.
+
+The isomorphisms are extended to form a congruence relation by adding the
+following constructors that witness equivalence and compatible closure:
+
+\begin{table}[t]
+\hrule\medskip
+\begin{center} 
+\Rule{}
+{}
+{\jdg{}{}{\idc : b \iso b}}
+{}
+\quad 
+\Rule{}
+{\jdg{}{}{c : b_1 \iso b_2}}
+{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
+{}
+\quad
+\Rule{}
+{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
+{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
+{}
+\\ \bigskip
+\Rule{}
+{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
+{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
+{}
+\quad 
+\Rule{}
+{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
+{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
+{}
+\end{center}
+\caption{Combinators\label{pi-combinators}}
+\medskip\hrule
+\end{table}
+
+It is important to note that ``values'' and ``isomorphisms'' are completely
+separate syntactic categories which do not intermix. The semantics of the
+language come when these are made to interact at the ``top level'' via
+\emph{application}: 
+\[\begin{array}{lrcl}
+\textit{top level term}, l &::=& c~v
 \end{array}\]

### 2014-05-21 ring.tex
getting more explicit about dimensions and indexing

@@
 
+\newcommand{\bdim}[1]{\textsf{dim}(#1)}
 \newcommand{\Rule}[4]{
@@
 \textit{types of dimensions $\geq 1$} & 
-  \tau &::=& t \alt \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
+  \tau &::=& \Box t \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
 \end{array}\]
@@
 interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
+The $\Box$ is used to make explicit the injection of a first-order type into
+a 1-dimensional type. We omit the explicit injection when
+convenient. Formally, the \emph{dimension} of a type is defined as follows:
+\[\begin{array}{rcl}
+\bdim{t} &=& 0 \\
+\bdim{\Box t} &=& 1 \\
+\bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
+\bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
+\bdim{- \tau} &=& \bdim{\tau}
+\end{array}\]
 
@@
 \den{0} &=& \emptyset \\
-\den{1} &=& \{ \bullet \} \\
+\den{1} &=& \{ \star \} \\
 \den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
@@
 
-For cubical types of dimension 1 or higher, their denotations, usually
-written as $\cubt$, are $n$-dimensional ``polarized cubes.'' Such an
-$n$-dimensional cube $\nodet{\cubt_1}{\cubt_2}$ consists of a positive
-subspace $\cubt_1$ and a negative subspace $\cubt_2$ (shaded in gray). Each
-of the subspaces are cubes of a lower dimension; the $0$-dimensional case
-reduces to the denotation of conventional first-order types. As we explain
-below, cubical sets are closed under sums, differences, products.
+For cubical types of dimension $n \geq 1$, their denotations, usually written
+as $\cubt$, are $n$-dimensional ``polarized cubes.'' Such an $n$-dimensional
+cube $\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and
+a negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
+cubes of a lower dimension; the $0$-dimensional case reduces to the
+denotation of conventional first-order types. As we explain below, cubical
+sets are closed under sums, differences, products.
 
 A 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which intuitively
-corresponds to the difference $t_1 - t_2$ of the two types whose denotations
-are $S_1$ and $S_2$ respectively. A 2-dimensional cube looks like
+corresponds to the difference $t_1 - t_2$ (formally $\Box t_1 + (- \Box
+t_2)$) of the two types whose denotations are $S_1$ and $S_2$ respectively. A
+2-dimensional cube looks like
 $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
@@
 \[\begin{array}{rcl}
-\denc{t} &=& \nodet{\den{t}}{\phantom{\den{t}}} \\
+\denc{\Box t} &=& \nodet{\den{t}}{\phantom{\den{t}}} \\
 \denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
@@
 \]
-In the recursive definitions below, we assume that the cubical sets have been
-extended to be balanced as needed:
+In the recursive definition of addition below, we assume that the smaller
+cube has been extended to the dimension of the larger one: 
 \[\begin{array}{rcl}
@@
 \\
-\infer[\textit{prod}]{(v_1^{p_1},v_2^{p_2})^{p_1 \cdot p_2} : \tau_1 * \tau_2}
+\infer[\textit{prod}]{(v_1,v_2)^{p_1 \cdot p_2} : \tau_1 * \tau_2}
       {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} \\

### 2014-05-21 ring.tex
removed syntactic difference between 0 dim and higher dim

@@
 \newcommand{\den}[1]{\llbracket #1 \rrbracket}
-\newcommand{\denc}[1]{\llbracket #1 \rrbracket^c}
 \newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
@@
 %% double-blind reviewing...
-\title{Cubical Polarized Types}
+\title{Polarized Cubical Types}
 \authorinfo{}{}{}
@@
 
-We use $t$ for regular first-order types which include the empty type, the
-unit type, and conventional sum and product types. We call these types
-$0$-dimensional. We use $\tau$ for cubical types of dimension 1 or higher:
-these have the same syntactic structure as regular types except that they
-also include \emph{negative} types. The syntax of the types is summarized
-below:
-\[\begin{array}{lrcl}
-\textit{types of dimension $0$} & 
-  t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
-\textit{types of dimensions $\geq 1$} & 
-  \tau &::=& \Box t \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
+Our cubical types $\tau$ include the empty type 0, the unit type 1,
+conventional sum and product types, as well as \emph{negative} types:
+\[\begin{array}{rcl}
+\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
 \end{array}\]
@@
 interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
-The $\Box$ is used to make explicit the injection of a first-order type into
-a 1-dimensional type. We omit the explicit injection when
-convenient. Formally, the \emph{dimension} of a type is defined as follows:
+The \emph{dimension} of a type is defined as follows:
 \[\begin{array}{rcl}
-\bdim{t} &=& 0 \\
-\bdim{\Box t} &=& 1 \\
+\bdim{0} &=& 0 \\
+\bdim{1} &=& 0 \\
 \bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
 \bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
-\bdim{- \tau} &=& \bdim{\tau}
+\bdim{- \tau} &=& \max(1,\bdim{\tau})
 \end{array}\]
-
-It is easy to find a model for $0$-dimensional types: finite sets are such a
-model. In more detail:
-\[\begin{array}{rcl}
-\den{0} &=& \emptyset \\
-\den{1} &=& \{ \star \} \\
-\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
-\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
-\end{array}\]
-The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
-of types maps to disjoint union of sets. The product of types maps to the
-cartesian product of sets. We usually use $S$ for the denotation of these
-conventional first-order types. Note that our (simple) semantics of types
-identifies several structurally different types such as $(1+(1+1))$ and
-$((1+1)+1)$. In some sense, this is innocent as the types are
-isomorphic. However, in the operational semantics discussed in the next
-section, we make the computational content of such type isomorphisms
-explicit.
-
-For cubical types of dimension $n \geq 1$, their denotations, usually written
-as $\cubt$, are $n$-dimensional ``polarized cubes.'' Such an $n$-dimensional
-cube $\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and
-a negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
-cubes of a lower dimension; the $0$-dimensional case reduces to the
-denotation of conventional first-order types. As we explain below, cubical
-sets are closed under sums, differences, products.
-
-A 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which intuitively
-corresponds to the difference $t_1 - t_2$ (formally $\Box t_1 + (- \Box
-t_2)$) of the two types whose denotations are $S_1$ and $S_2$ respectively. A
-2-dimensional cube looks like
-$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
-corresponds to the iterated difference of the appropriate types
-$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
-outermost box encodes the sign.
-
-We now give the full denotation of cubical types:
-\[\begin{array}{rcl}
-\denc{\Box t} &=& \nodet{\den{t}}{\phantom{\den{t}}} \\
-\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
-\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
-\denc{- \tau} &=& \ominus \denc{\tau}
-\end{array}\]
-where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
-It is possible, in general, for the binary trees representing the cubes not
-be balanced. It is however always possible to embed an $n$-dimensional cube
-into a cube of a larger dimension using the fact (to be justified using a
-$\Pi$-combinator that witnesses this isomorphism in the next section) that
-$t$ is equivalent to $t-0$ and to $(t-0)-0$ and so on, i.e.,
-\[
-S = \nodet{S}{\phantom{S}} = 
-  \nodet{\nodet{S}{\phantom{S}}}
-        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
-\]
-In the recursive definition of addition below, we assume that the smaller
-cube has been extended to the dimension of the larger one: 
-\[\begin{array}{rcl}
-\ominus~S &=& \nodet{\phantom{S}}{S} \\
-\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
-\\
-S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
-(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
-  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
-\\
-S_1 \otimes S_2 &=& S_1 \times S_2 \\
-S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
-  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
-(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
-  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
-\end{array}\]
+The base types have dimension 0. Using negative types but no product types
+raises the dimension to 1. This is the situation with the \textbf{Int} or
+$\mathcal{G}$ construction. Once product types are used, the dimension can
+increase without bounds.
 
@@
 \end{figure*}
-
-\noindent The definition of negation is natural: it recursively swaps the
-positive and negative subspaces. We note that the general definitions of sums
-and products reduce to the conventional sums and products for $0$-dimensional
-types. Also the sum of 1-dimensional types reduces to the sum used in the
-\textbf{Int} construction and that addition and negation do not increase the
-dimensions. The interesting new development is the definition of
-products. The example in Fig.~\ref{mult} illustrates a general pattern: the
-product of an $n$-dimensional type with an $m$-dimensional type gives an
-$(n+m)$-dimensional type. For low dimensions such as the ones used in the
-example above, it is possible to visualize what is happening in coordinate
-space (see Fig.~\ref{multv}).
 
@@
 \end{figure*}
+
+This point is made precise in the denotation of types which maps a type of
+dimension $n$ to $n$-dimensional cube. We represent such a cube syntactically
+as a binary tree of maximum depth $n$ with nodes of the form
+$\nodet{\cubt_1}{\cubt_2}$. In such a node, $\cubt_1$ is the positive
+subspace and $\cubt_2$ (shaded in gray) is the negative subspace along the
+first dimension. Each of these subspaces is itself a cube of a lower
+dimension. The $0$-dimensional cubes are the conventional denotation of
+first-order types as sets. We use $S$ to denote the denotations of such
+types. For example, 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which
+intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
+denotations are $S_1$ and $S_2$ respectively. A full 2-dimensional cube looks
+like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
+corresponds to the iterated difference of the appropriate types
+$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
+outermost box encodes the sign.
+
+Formally, the denotation of types is as follows:
+\[\begin{array}{rcl}
+\den{0} &=& \emptyset \\
+\den{1} &=& \{ \star \} \\
+\den{\tau_1 + \tau_2} &=& \den{\tau_1} \oplus \den{\tau_2} \\
+\den{\tau_1 * \tau_2} &=& \den{\tau_1} \otimes \den{\tau_2} \\
+\den{- \tau} &=& \ominus \den{\tau} \\
+\\
+\noalign{\mbox{where:}\hfill}
+\\
+S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
+S \oplus (\nodet{\cubt_1}{\cubt_2}) &=& 
+  (\nodet{S}{\phantom{S}}) \oplus (\nodet{\cubt_1}{\cubt_2}) \\
+(\nodet{\cubt_1}{\cubt_2}) \oplus S &=& 
+  (\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{S}{\phantom{S}}) \\
+(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
+  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
+\\
+S_1 \otimes S_2 &=& S_1 \times S_2 \\
+S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
+  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
+(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
+  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt} \\
+\\
+\ominus~S &=& \nodet{\phantom{S}}{S} \\
+\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} 
+\end{array}\]
+The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
+of $0$-dimensional types is the disjoint union as expected. For cubes of
+higher dimensions, the subspaces are recursively added. If the two types to
+be added are of different dimensions, it is necessary to embed the smaller
+type in the larger dimension using the fact (to be justified using a
+$\Pi$-combinator that witnesses this isomorphism in the next section) that
+$t$ is equivalent to $t-0$ and to $(t-0)-0$ and so on, i.e.,
+\[
+S = \nodet{S}{\phantom{S}} = 
+  \nodet{\nodet{S}{\phantom{S}}}
+        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
+\]
+The product of 0-dimensional types is the cartesian product of sets. The
+definition of negation is natural: it recursively swaps the positive and
+negative subspaces. We note that the general definitions of sums and products
+reduce to the conventional sums and products for $0$-dimensional types. Also
+the sum of 1-dimensional types reduces to the sum used in the \textbf{Int}
+construction and that addition and negation do not increase the
+dimensions. The interesting new development is the definition of
+products. The example in Fig.~\ref{mult} illustrates the general pattern: the
+product of an $n$-dimensional type with an $m$-dimensional type gives an
+$(n+m)$-dimensional type. For low dimensions such as the ones used in the
+example above, it is possible to visualize what is happening in coordinate
+space (see Fig.~\ref{multv}).
+
+\paragraph*{Note.} 
+Our (simple) semantics of types identifies several structurally different
+types such as $(1+(1+1))$ and $((1+1)+1)$. In some sense, this is innocent as
+the types are isomorphic. However, in the operational semantics discussed in
+the next section, we make the computational content of such type isomorphisms
+explicit.
 
@@
 The terms of $\Pi$ witness type isomorphisms of the form $b \iso b$. They
-consist of base isomorphisms, as defined below, and their composition.
-\[\begin{array}{rrcll}
+consist of base isomorphisms, as defined below, and their composition. Each
+line of the above table introduces a pair of dual constants\footnote{where
+  $\swapp$ and $\swapt$ are self-dual.} that witness the type isomorphism in
+the middle.  These are the base (non-reducible) terms of the second,
+principal level of $\Pi$. Note how the above has two readings: first as a set
+of typing relations for a set of constants. Second, if these axioms are seen
+as universally quantified, orientable statements, they also induce
+transformations of the (traditional) values. The (categorical or homotopical)
+intuition here is that these axioms have computational content because they
+witness isomorphisms rather than merely stating an extensional equality.
+
+The isomorphisms are extended to form a congruence relation by adding the
+following constructors that witness equivalence and compatible closure:
+
+\begin{table*}[t]
+\[\begin{array}{cc}
+\begin{array}{rrcll}
 \identlp :&  0 + b & \iso & b &: \identrp \\
@@
 \distz :&~ 0 * b & \iso & 0 &: \factorz \\
-\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor
-\end{array}\]
-Each line of the above table introduces a pair of dual
-constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that witness
-the type isomorphism in the middle.  These are the base (non-reducible) terms
-of the second, principal level of $\Pi$. Note how the above has two readings:
-first as a set of typing relations for a set of constants. Second, if these
-axioms are seen as universally quantified, orientable statements, they also
-induce transformations of the (traditional) values. The (categorical or
-homotopical) intuition here is that these axioms have computational content
-because they witness isomorphisms rather than merely stating an extensional
-equality.
-
-The isomorphisms are extended to form a congruence relation by adding the
-following constructors that witness equivalence and compatible closure:
-
-\begin{table}[t]
-\hrule\medskip
+\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor \\
+\eta :&~ 0 & \iso & b + (-b)~ & : \epsilon
+\end{array}
+& 
+\begin{minipage}{0.5\textwidth}
 \begin{center} 
@@
 {}
-\quad 
+\qquad\qquad
 \Rule{}
@@
 {}
-\quad
+\\ \bigskip
 \Rule{}
@@
 {}
-\quad 
+\\ \bigskip
 \Rule{}
@@
 \end{center}
+\end{minipage}
+\end{array}\]
 \caption{Combinators\label{pi-combinators}}
-\medskip\hrule
-\end{table}
+\end{table*}
 

### 2014-05-22 ring.tex
thoughts

@@
 \newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
+\newcommand{\hast}{$:\mkern -2.5mu:\;$}
 
@@
 \end{array}\]
-\caption{\label{mult}Example of multiplication of two cubical sets.}
-\end{figure*}
 
-\begin{figure*}
 \begin{tikzpicture}
@@
 \end{tikzpicture}
-\caption{\label{multv}Visualization of multiplication of two cubical sets.}
+\caption{\label{mult}Example of multiplication of two cubical sets.}
 \end{figure*}
@@
 S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
-S \oplus (\nodet{\cubt_1}{\cubt_2}) &=& 
-  (\nodet{S}{\phantom{S}}) \oplus (\nodet{\cubt_1}{\cubt_2}) \\
-(\nodet{\cubt_1}{\cubt_2}) \oplus S &=& 
-  (\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{S}{\phantom{S}}) \\
+S \oplus (\nodet{\cubt_1}{\cubt_2}) &=& \nodet{S \oplus \cubt_1}{\cubt_2} \\
+(\nodet{\cubt_1}{\cubt_2}) \oplus S &=& \nodet{\cubt_1 \oplus S}{\cubt_2} \\
 (\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
@@
 of $0$-dimensional types is the disjoint union as expected. For cubes of
-higher dimensions, the subspaces are recursively added. If the two types to
-be added are of different dimensions, it is necessary to embed the smaller
-type in the larger dimension using the fact (to be justified using a
-$\Pi$-combinator that witnesses this isomorphism in the next section) that
-$t$ is equivalent to $t-0$ and to $(t-0)-0$ and so on, i.e.,
-\[
-S = \nodet{S}{\phantom{S}} = 
-  \nodet{\nodet{S}{\phantom{S}}}
-        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
-\]
-The product of 0-dimensional types is the cartesian product of sets. The
-definition of negation is natural: it recursively swaps the positive and
-negative subspaces. We note that the general definitions of sums and products
-reduce to the conventional sums and products for $0$-dimensional types. Also
-the sum of 1-dimensional types reduces to the sum used in the \textbf{Int}
-construction and that addition and negation do not increase the
-dimensions. The interesting new development is the definition of
-products. The example in Fig.~\ref{mult} illustrates the general pattern: the
-product of an $n$-dimensional type with an $m$-dimensional type gives an
-$(n+m)$-dimensional type. For low dimensions such as the ones used in the
-example above, it is possible to visualize what is happening in coordinate
-space (see Fig.~\ref{multv}).
+higher dimensions, the subspaces are recursively added. Note that the sum of
+1-dimensional types reduces to the sum used in the \textbf{Int} construction.
+The definition of negation is natural: it recursively swaps the positive and
+negative subspaces. The product of 0-dimensional types is the cartesian
+product of sets. For cubes of higher-dimensions $n$ and $m$, the result is of
+dimension $(n+m)$. The example in Fig.~\ref{mult} illustrates the idea using
+the product of 1-dimensional cube (i.e., a line) with a 2-dimensional cube
+(i.e., a square). The result is a 3-dimensional cube as illustrated.
 
@@
 explicit.
+
+\begin{verbatim}
+A critical isomorphism that is 
+not reflected at all in the semantics
+is that 1-1 = 0
+
+The denotation of 1-1 is a line;
+the denotation of 0 is the empty set
+
+[ {*} | {*} ] --> []
+
+Something like the simplification
+of Conway games here perhaps?
+
+There are two values of type
+1-1
+
+() 
+and 
+()^-
+
+epsilon should take one and map
+to the other because it can
+never produce something of type 0
+
+In general, we have several direction
+not just left and right like in the
+Int construction. In 3D we have
+8 directions: +++, ++-, ..., ---
+We have an epsilon that 
+takes ++- and flips it to --+
+and another that flips
++++ to --- and so on
+
+So we should have 8 interpreters
+for 3D one for each direction!!!
+
+The ++- interpreter would 
+manipulates values like
+(v1,v2,v3) and would apply
+the normal combinators to v1, v2, 
+and the adjoint to v3 and so on.
+\end{verbatim}
 
@@
 Now that the type structure is defined, we turn our attention to the notion
-of values. For conventional first-order types, the values are also
-conventional:
-\[
-\infer{() : 1}{} 
+of values. Intuitively, a value of the $n$-dimensional type $\tau$ is an
+element of one of the sets located in one of the corners of the
+$n$-dimensional cube denoted by $\tau$. Thus to specify a value, we must
+first specify one of the corners of the cube (or equivalently one of the
+leaves in the binary tree representation) which can easily be done using a
+sequence $p$ of $+$ and $-$ polarities indicating how to navigate the cube in
+each successive dimension starting from a fixed origin to reach the desired
+corner. We write $v^{p}$ for the value $v$ located at corner $p$ of the cube
+associated with its type. We use $\epsilon$ for the empty sequence of
+polarities and identify $v$ with $v^\epsilon$. Note that the polarities
+doesn't completely specify the type since different types like $(1+(1+1))$
+and $((1+1)+1)$ are assigned the same denotation. What the path $p$ specifies
+is the \emph{polarity} of the value, or its ``orientation'' in the space
+denoted by its type. Formally:
+\[\begin{array}{c}
+\infer{() : 1}{} \\
 \qquad
-\infer{\inl{v} : t_1 + t_2}{v : t_1}
-\qquad
-\infer{\inr{v} : t_1 + t_2}{v : t_2}
-\qquad
-\infer{(v_1,v_2) : t_1 * t_2}{v_1 : t_1 & v_2 : t_2}
-\]
-For cubical types, the values are more involved. Intuitively, a value of the
-$n$-dimensional type $\tau$ is an element of one of the sets located in a
-corner of the $n$-dimensional cube denoted by $\tau$. Thus to specify a
-value, we must specify one of the corners of the cube which can easily be
-done using a sequence $p$ of $+$ and $-$ polarities indicating how to
-navigate the cube in each successive dimension starting from a fixed origin
-to reach the desired corner. We denote such values as $v^{p}$ and use
-$\epsilon$ for the empty sequence of polarities. One can think of the
-sequence of signs as denoting a path in the binary tree representation of
-cubical types. Formally:
-\[\begin{array}{c}
-\infer[\textit{base}]{v^{+} : t}{v : t} \\
+\infer[\textit{neg}]{v^{\negp{p}} : - \tau}{v^p : \tau} 
 \\
@@
 \qquad
-\infer[\textit{right}]{(\inr{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} \\
+\infer[\textit{right}]{(\inr{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} 
 \\
 \infer[\textit{prod}]{(v_1,v_2)^{p_1 \cdot p_2} : \tau_1 * \tau_2}
-      {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} \\
-\\
-\infer[\textit{neg}]{v^{\negp{p}} : - \tau}{v^p : \tau} 
+      {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} 
 \end{array}\]
-The rule \textit{base} states that a conventional first-order value can be
-viewed as trivial $0$-dimensional cubical value. The rules \textit{left} and
-\textit{right} reflect the fact that sums do not increase the dimension and
-works pointwise. The rule \textit{prod} is the most involved one: it
-increases the dimension by \emph{concatenating} the two dimensions in its
-arguments. For example, if we pair $v_1^{\pp}$ and $v_2^{\mm\pp}$ we get
-$(v_1,v_2)^{\pp\mm\pp}$. The rule \textit{neg} shows that a negative value
-$v$ is located at the ``opposite'' corner of $v$ which is calculated as
-follows:
+The rules \textit{left} and \textit{right} reflect the fact that sums do not
+increase the dimension. Note that when $p$ is $\epsilon$, we get the
+conventional values for the 0-dimensional sum type. The rule \textit{prod} is
+the most involved one: it increases the dimension by \emph{concatenating} the
+two dimensions of its arguments. For example, if we pair $v_1^{\pp}$ and
+$v_2^{\mm\pp}$ we get $(v_1,v_2)^{\pp\mm\pp}$. (See Fig.~\ref{mult} for the
+illustration.) Note again that if both components are 0-dimensional, the pair
+remains 0-dimensional and we recover the usual rule for typing values of
+product types. The rule \textit{neg} uses the function below which states
+that the negation of a value $v$ is the same value $v$ located at the
+``opposite'' corner of the cube:
 \[\begin{array}{rcl}
@@
 
+\begin{verbatim}
+We only have -
+which flips ALL the directions simultaneously.
+So from ++-+--+ you can only go to
+        --+-++-
+What if you wanted to go to
+        ++++--+
+I think we use the * functor to flip the 
+direction we want. 
+\end{verbatim}
+
 %%%%%%%%%%%%%%%%%%
-\subsection{Combinators}
+\subsection{Combinators: Example}
+
+Consider the following simple function on 0-dimensional sum types:
+\[\begin{array}{rcl}
+\textsf{swapP} &\hast& (\tau_1 + \tau_2) \rightarrow (\tau_2 + \tau_1) \\
+\textsf{swapP} (\inl{v}) &=& inr{v} \\
+\textsf{swapP} (\inr{v}) &=& inl{v}
+\end{code}
+Given our setup, this just works $n$-dimensional types. And so do most of the
+$\Pi$ combinators for that matter. Only $\eta$ and $\epsilon$ seem to need
+thinking. How and why would $1-1$ which is a 1-dimensional line be the same
+as the empty type which is a 0-dimensional thing. And how do we generalize
+for arbitrary group identities at higher dimensions. We need a mechanism for
+cubes with subspaces that ``cancel'' to map to equivalent smaller subcubes.
+
+%%%%%%%%%%%%%%%%%%
+\subsection{$\Pi$ Combinators}
 

### 2014-05-22 ring.tex
thoughts
paths to the rescue

@@
 \newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
-\newcommand{\hast}{$:\mkern -2.5mu:\;$}
+\newcommand{\hast}{:\mkern -2.5mu:\;}
 
@@
 \[\begin{array}{rcl}
+\bdim{\cdot} &\hast& \tau \rightarrow \mathbb{N} \\
 \bdim{0} &=& 0 \\
@@
 \end{array}\]
-
+\\
+\\
 \begin{tikzpicture}
@@
 \end{tikzpicture}
-\caption{\label{mult}Example of multiplication of two cubical sets.}
+\caption{\label{mult}Example of multiplication of two cubical types.}
 \end{figure*}
@@
 dimension $n$ to $n$-dimensional cube. We represent such a cube syntactically
-as a binary tree of maximum depth $n$ with nodes of the form
+as a binary tree of maximum depth~$n$ with nodes of the form
 $\nodet{\cubt_1}{\cubt_2}$. In such a node, $\cubt_1$ is the positive
@@
 first dimension. Each of these subspaces is itself a cube of a lower
-dimension. The $0$-dimensional cubes are the conventional denotation of
-first-order types as sets. We use $S$ to denote the denotations of such
-types. For example, 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which
+dimension. The $0$-dimensional cubes are plain sets representing the
+denotation of conventional first-order types. We use $S$ to denote the
+denotations of these plain types. A 1-dimensional cube, $\nodet{S_1}{S_2}$,
 intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
-denotations are $S_1$ and $S_2$ respectively. A full 2-dimensional cube looks
-like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
-corresponds to the iterated difference of the appropriate types
-$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
-outermost box encodes the sign.
+denotations are $S_1$ and $S_2$ respectively. The type can be visualized as a
+``line'' with polarized endpoints connecting the two points $S_1$ and
+$S_2$. A full 2-dimensional cube,
+$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$, intuitively corresponds to
+the iterated difference of the appropriate types $(t_1-t_2)-(t_3-t_4)$ where
+the product of the successive signs from the outermost box encodes the
+sign. The type can be visualized as a ``square'' with polarized corners
+connecting the two lines corresponding to $(t_1-t_2)$ and $(t_3-t_4)$. (See
+Fig.~\ref{mult} which is further explained after we discuss multiplication
+below.)
 
@@
 The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
-of $0$-dimensional types is the disjoint union as expected. For cubes of
-higher dimensions, the subspaces are recursively added. Note that the sum of
+of $0$-dimensional types is the disjoint union as usual. For cubes of higher
+dimensions, the subspaces are recursively added. Note that the sum of
 1-dimensional types reduces to the sum used in the \textbf{Int} construction.
@@
 
-\paragraph*{Note.} 
+%%%%%%%%%%%%%%%%%%%%
+\subsection{Type Isomorphisms}
+
 Our (simple) semantics of types identifies several structurally different
@@
 the next section, we make the computational content of such type isomorphisms
-explicit.
+explicit. Some other isomorphic types like $(t_1*t_2)$ and $(t_2*t_1)$ map to
+different cubes and are \emph{not} identified: explicit isomorphisms are
+needed to mediate between them. In most cases, the isomorphisms are easy to
+define except for the isomorphisms involving the empty type. In particular,
+if negative types are to be interpreted as their name suggests, we must have
+an isomorphism between $(t-t)$ and the empty type 0. Semantically the former
+denotes $\nodet{\cubt}{\cubt}$ and the latter denotes the empty set. Their
+denotations are different and there is no way so far to express the fact that
+these two cubes should be identified. Intuitively, what is needed is the
+ability to contract a line whose endpoints are identical to a single
+point. It should then follow that a square where one edge contracts to a
+point would itself contract to a line, and a 3-dimensional cube in which one
+face contracts to a point would itself contract to a square, etc. This is, of
+course, where the ideas of homotopy (type) theory enter the development.
+
+Consider the situation with the simplest instance of the desired isomorphism
+which is between $(1-1)$ and the empty type:
+\begin{center}
+\begin{tikzpicture}
+\node[above] at (0,0) {\pp};
+\node[left] at (-0.2,0) {$1$};
+\draw[fill] (0,0) circle [radius=0.05];
+\node[above] at (2.6,0) {\mm};
+\node[right] at (2.8,0) {$1$};
+\draw[fill] (2.6,0) circle [radius=0.05];
+\draw[->,red] (0.1,0) -- (2.5,0);
+\draw[->,red] (-0.1,-0.1) arc (135:405:0.2);
+\node[below,red] at (0,-0.4) {\textsf{refl}};
+\draw[->,red] (2.5,-0.1) arc (135:405:0.2);
+\node[below,red] at (2.6,-0.4) {\textsf{refl}};
+\node[above] at (1.3,0) {$I$};
+\node[right] at (1.3,-0.5) {$q$};
+%%
+\node[below] at (1.3,-1.5) {$\emptyset$};
+\draw[fill] (1.3,-1.5) circle [radius=0.05];
+\draw[->,red] (1.4,-1.4) arc (-45:225:0.2);
+\node[right,red] at (1.5,-1.3) {\textsf{refl}};
+%% 
+\draw[->,double,blue] (1.3,-0.1) -- (1.3,-1);
+\end{tikzpicture}
+\end{center}
+The top of the figure is the 1-dimensional cube representing the type $(1-1)$
+except that we view the edge connecting the two endpoints as a \emph{path}
+$I$ that is initially non-contractible, i.e., it is different from the
+identity paths labeled \textsf{refl} that are the trivial paths asserting
+that $1$ is contractible to itself. The bottom of the figure is the
+0-dimensional cube representing the empty type. There is a trivial path
+\textsf{refl} connecting that point to itself. To express the equivalence of
+$(1-1)$ to 0, we add a 2-path, i.e. a path between paths, that asserts the
+equivalence of the path $I$ with \textsf{refl} on the empty type. That
+effectively contracts the two points and collapses them to a single
+point. Surprisingly, that is everthing that we need. The extension to higher
+dimensions just ``works'' because paths in homotopy type theory have a rich
+structure. We explain the details below.
 
@@
 \end{verbatim}
-
-%%%%%%%%%%%%%%%%%%%%
-\subsection{Type Isomorphisms}
 
@@
 \textsf{swapP} (\inr{v}) &=& inl{v}
-\end{code}
+\end{array}\]
 Given our setup, this just works $n$-dimensional types. And so do most of the

### 2014-05-22 ring.tex
comment on 2d

@@
 
-We first define the syntax a semantic model of types and then discuss type
-isomorphisms.
+We first define the syntax and then present a simple semantic model of types
+which is then refined.
 
@@
 \begin{verbatim}
+So the story is that as we are building
+the denotations of types, we add these
+paths making up the cubes. The only
+case where we add new paths is multiplication
+(see Fig.1 again). We take the paths p in S1
+and the paths q in S2 and build paths (p,q)
+in the product space.
+
+Once we have this structure, we do this 
+homotopy completion which turns our inductive
+type into a higher-inductive type. We 
+start adding 2-paths between any path 
+connecting (S-S) and the refl path on 0.
+
+Technically this is a functor between (S-S) 
+and the empty type mapping S to 0, the 
+other S to 0, and the path between them
+to refl.
+
+Now consider (a-b)-(c-c). We have a 
+functor to (a-b)-0 that contracts the
+path between c and -c. We want to 
+reduce the dimension of that space
+and map to (a-b).
+
+original space
+
+    c ------ c
+    |        |  
+p1  |        | p2
+    |        |  
+    a ------ b
+       p3
+
+after collapsing the two c's
+
+      0
+     /\
+    /  \
+   a -- b
+
+or because path compose
+
+  a ---- b
+    ----
+
+where the top path is p1;p2
+and the bottom path is p3
+
+I think we should have a 2-path
+between them to say that these
+paths are equivalent which will 
+then collapse the square to a 
+line as desired.
+
 A critical isomorphism that is 

### 2014-05-23 ring.tex
more about paths and groupoids

@@
 
+\newcommand{\refl}[1]{\textsf{refl}_{#1}}
+\newcommand{\seg}[1]{\textsf{seg}_{#1}}
 \newcommand{\bdim}[1]{\textsf{dim}(#1)}
@@
 %% double-blind reviewing...
-\title{Polarized Cubical Types}
+\title{Negative Types}
 \authorinfo{}{}{}
@@
 solved. We ``translate'' that solution to a computational type-theoretic
-world. This has evident connections to homotopy (type?) theory that remain to
+world. This has evident connections to homotopy (type) theory that remain to
 be investigated.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Problem and the Intuitive Solution}
-
 We lose the multiplicative structure. No evident way to define the
@@
 algebraic topology perspective. Regular first-order types are viewed as
-$0$-dimensional cubes. Generalize to $n$-dimensional cubes.
+$0$-dimensional cubes. The \textbf{Int} construction generalizes types to
+1-dimensional cubes. Our work here generalizes types to $n$-dimensional
+cubes.
 
@@
 
-Our cubical types $\tau$ include the empty type 0, the unit type 1,
-conventional sum and product types, as well as \emph{negative} types:
+Our types $\tau$ include the empty type 0, the unit type 1, conventional sum
+and product types, as well as \emph{negative} types:
 \[\begin{array}{rcl}
@@
 \end{array}\]
-The base types have dimension 0. Using negative types but no product types
-raises the dimension to 1. This is the situation with the \textbf{Int} or
-$\mathcal{G}$ construction. Once product types are used, the dimension can
-increase without bounds.
+The base types have dimension 0. If negative types are not used, all
+dimensions remain at 0. If negative types are used but no product types of
+negative types appear anywhere, the dimension is raised to 1. This is the
+situation with the \textbf{Int} or $\mathcal{G}$ construction. Once negative
+and product types are freely used, the dimension can increase without bounds.
 
@@
 \\
+\begin{center}
 \begin{tikzpicture}
@@
 \draw[fill] (0.6,0) circle [radius=0.05];
-\draw[-] (0,0) -- (0.6,0);
+\draw[-,dotted] (0,0) -- (0.6,0);
 \node at (1.6,0) {$\otimes$};
@@
 \draw[fill] (3.5,-0.5) circle [radius=0.05];
-\draw[-] (2.5,-0.5) -- (3.5,-0.5);
-\draw[-] (2.5,-0.5) -- (2.5,0.5);
+\draw[-,dotted] (2.5,-0.5) -- (3.5,-0.5);
+\draw[-,dotted] (2.5,-0.5) -- (2.5,0.5);
 \node[left] at (2.5,0.5) {\mm\pp};
@@
 \draw[fill] (3.5,0.5) circle [radius=0.05];
-\draw[-] (2.5,0.5) -- (3.5,0.5);
-\draw[-] (3.5,-0.5) -- (3.5,0.5);
+\draw[-,dotted] (2.5,0.5) -- (3.5,0.5);
+\draw[-,dotted] (3.5,-0.5) -- (3.5,0.5);
 %% 
@@
 %%
-\draw[-] (7.5,0.75) -- (9.5,0.75);
-\draw[-] (9.5,0.75) -- (10.2,1.2);
-\draw[-] (10.2,1.2) -- (8.2,1.2);
-\draw[-] (8.2,1.2) -- (7.5,0.75);
+\draw[-,dotted] (7.5,0.75) -- (9.5,0.75);
+\draw[-,dotted] (9.5,0.75) -- (10.2,1.2);
+\draw[-,dotted] (10.2,1.2) -- (8.2,1.2);
+\draw[-,dotted] (8.2,1.2) -- (7.5,0.75);
 %%
-\draw[-] (7.5,-0.75) -- (9.5,-0.75);
-\draw[-] (9.5,-0.75) -- (10.2,-0.3);
-\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
-\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
+\draw[-,dotted] (7.5,-0.75) -- (9.5,-0.75);
+\draw[-,dotted] (9.5,-0.75) -- (10.2,-0.3);
+\draw[-,dotted,dashed] (10.2,-0.3) -- (8.2,-0.3);
+\draw[-,dotted,dashed] (8.2,-0.3) -- (7.5,-0.75);
 %%
-\draw[-] (7.5,0.75) -- (7.5,-0.75);
-\draw[-] (9.5,0.75) -- (9.5,-0.75);
-\draw[-] (10.2,1.2) -- (10.2,-0.3);
-\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
+\draw[-,dotted] (7.5,0.75) -- (7.5,-0.75);
+\draw[-,dotted] (9.5,0.75) -- (9.5,-0.75);
+\draw[-,dotted] (10.2,1.2) -- (10.2,-0.3);
+\draw[-,dotted,dashed] (8.2,1.2) -- (8.2,-0.3);
 \end{tikzpicture}
+\end{center}
 \caption{\label{mult}Example of multiplication of two cubical types.}
@@
 
-This point is made precise in the denotation of types which maps a type of
-dimension $n$ to $n$-dimensional cube. We represent such a cube syntactically
-as a binary tree of maximum depth~$n$ with nodes of the form
-$\nodet{\cubt_1}{\cubt_2}$. In such a node, $\cubt_1$ is the positive
-subspace and $\cubt_2$ (shaded in gray) is the negative subspace along the
-first dimension. Each of these subspaces is itself a cube of a lower
-dimension. The $0$-dimensional cubes are plain sets representing the
-denotation of conventional first-order types. We use $S$ to denote the
-denotations of these plain types. A 1-dimensional cube, $\nodet{S_1}{S_2}$,
-intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
-denotations are $S_1$ and $S_2$ respectively. The type can be visualized as a
-``line'' with polarized endpoints connecting the two points $S_1$ and
-$S_2$. A full 2-dimensional cube,
-$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$, intuitively corresponds to
-the iterated difference of the appropriate types $(t_1-t_2)-(t_3-t_4)$ where
-the product of the successive signs from the outermost box encodes the
-sign. The type can be visualized as a ``square'' with polarized corners
-connecting the two lines corresponding to $(t_1-t_2)$ and $(t_3-t_4)$. (See
-Fig.~\ref{mult} which is further explained after we discuss multiplication
-below.)
+This point is made precise in the following tentative denotation of types (to
+be refined in the next section) which maps a type of dimension $n$ to an
+$n$-dimensional cube. We represent such a cube syntactically as a binary tree
+of maximum depth~$n$ with nodes of the form $\nodet{\cubt_1}{\cubt_2}$. In
+such a node, $\cubt_1$ is the positive subspace and $\cubt_2$ (shaded in
+gray) is the negative subspace along the first dimension. Each of these
+subspaces is itself a cube of a lower dimension. The $0$-dimensional cubes
+are plain sets representing the denotation of conventional first-order
+types. We use $S$ to denote the denotations of these plain types. A
+1-dimensional cube, $\nodet{S_1}{S_2}$, intuitively corresponds to the
+difference $t_1 - t_2$ of the two types whose denotations are $S_1$ and $S_2$
+respectively. The type can be visualized as a ``line'' with polarized
+endpoints connecting the two points~$S_1$ and $S_2$. A full 2-dimensional
+cube, $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$, intuitively
+corresponds to the iterated difference of the appropriate types
+$(t_1-t_2)-(t_3-t_4)$ where the successive ``colors'' from the outermost box
+encode the sign. The type can be visualized as a ``square'' with polarized
+corners connecting the two lines corresponding to $(t_1-t_2)$ and
+$(t_3-t_4)$. (See Fig.~\ref{mult} which is further explained after we discuss
+multiplication below.)
 
@@
 %%%%%%%%%%%%%%%%%%%%
-\subsection{Type Isomorphisms}
+\subsection{Type Isomorphisms: Paths to the Rescue}
 
-Our (simple) semantics of types identifies several structurally different
+Our proposed semantics of types identifies several structurally different
 types such as $(1+(1+1))$ and $((1+1)+1)$. In some sense, this is innocent as
@@
 different cubes and are \emph{not} identified: explicit isomorphisms are
-needed to mediate between them. In most cases, the isomorphisms are easy to
-define except for the isomorphisms involving the empty type. In particular,
-if negative types are to be interpreted as their name suggests, we must have
-an isomorphism between $(t-t)$ and the empty type 0. Semantically the former
-denotes $\nodet{\cubt}{\cubt}$ and the latter denotes the empty set. Their
-denotations are different and there is no way so far to express the fact that
-these two cubes should be identified. Intuitively, what is needed is the
-ability to contract a line whose endpoints are identical to a single
-point. It should then follow that a square where one edge contracts to a
-point would itself contract to a line, and a 3-dimensional cube in which one
-face contracts to a point would itself contract to a square, etc. This is, of
-course, where the ideas of homotopy (type) theory enter the development.
+needed to mediate between them. We therefore need to enrich our model of
+types with isomorphisms connecting types we deem equivalent.
 
-Consider the situation with the simplest instance of the desired isomorphism
-which is between $(1-1)$ and the empty type:
+So far, our types are modeled as cubes which are really sets indexed by
+polarities. An isomorphism between $(t_1*t_2)$ and $(t_2*t_1)$ requires
+nothing more than a pair of set-theoretic functions between the spaces that
+compose to the identity. What is much more interesting are the isomorphisms
+involving the empty type~0. In particular, if negative types are to be
+interpreted as their name suggests, we must have an isomorphism between
+$(t-t)$ and the empty type 0. Semantically the former denotes the ``line''
+$\nodet{\cubt}{\cubt}$ and the latter denotes the empty set. Their
+denotations are different and there is no way, in the world of plain sets, to
+express the fact that these two spaces should be identified. What is needed
+is the ability to \emph{contract} the \emph{path} between the endpoints of
+the line to the trivial path on the empty type. This is, of course, where the
+ideas of homotopy (type) theory enter the development.
+
+Consider the situation above in which we want to identify the spaces
+corresponding to the types $(1-1)$ and the empty type:
 \begin{center}
@@
 \draw[->,red] (-0.1,-0.1) arc (135:405:0.2);
-\node[below,red] at (0,-0.4) {\textsf{refl}};
+\node[below,red] at (0,-0.4) {$\refl{1}$};
 \draw[->,red] (2.5,-0.1) arc (135:405:0.2);
-\node[below,red] at (2.6,-0.4) {\textsf{refl}};
-\node[above] at (1.3,0) {$I$};
+\node[below,red] at (2.6,-0.4) {$\refl{1}$};
+\node[above] at (1.3,0) {$\seg{1}$};
 \node[right] at (1.3,-0.5) {$q$};
@@
 \draw[->,red] (1.4,-1.4) arc (-45:225:0.2);
-\node[right,red] at (1.5,-1.3) {\textsf{refl}};
+\node[right,red] at (1.5,-1.3) {$\refl{0}$};
 %% 
@@
 The top of the figure is the 1-dimensional cube representing the type $(1-1)$
-except that we view the edge connecting the two endpoints as a \emph{path}
-$I$ that is initially non-contractible, i.e., it is different from the
-identity paths labeled \textsf{refl} that are the trivial paths asserting
-that $1$ is contractible to itself. The bottom of the figure is the
-0-dimensional cube representing the empty type. There is a trivial path
-\textsf{refl} connecting that point to itself. To express the equivalence of
-$(1-1)$ to 0, we add a 2-path, i.e. a path between paths, that asserts the
-equivalence of the path $I$ with \textsf{refl} on the empty type. That
-effectively contracts the two points and collapses them to a single
-point. Surprisingly, that is everthing that we need. The extension to higher
-dimensions just ``works'' because paths in homotopy type theory have a rich
-structure. We explain the details below.
+as before except that we now add a path $\seg{1}$ to connect the two
+endpoints. (Note that previously, the dotted lines in the figures were a
+visualization aid and were \emph{not} meant to represent paths.) We also make
+explicit the trivial identity paths from every space to itself.  The bottom
+of the figure is the 0-dimensional cube representing the empty type. The path
+$\seg{1}$ identifies the two occurrences of 1. To express the
+equivalence of $(1-1)$ and 0, we add a 2-path, i.e. a path between paths,
+that asserts connects the path $\seg{1}$ to the trivial path
+$\refl{0}$ on the empty type. That effectively makes the two points
+``disappear.''  Surprisingly, that is everything that we need. The extension
+to higher dimensions just ``works'' because paths in homotopy type theory
+have a rich structure. We explain the details below.
+
+%%%%%%%%%%%%%%%%%%%%
+\subsection{The Universe of Types}
+
+We describe the construction of our universe of types. Our starting point is
+the construction in the previous section which we summarize again. We start
+with all finite sets built from the empty set, a singleton set, disjoint
+unions, and cartesian products. All these sets are thought of being indexed
+by the empty sequence of polarities $\epsilon$. We then constructs new spaces
+that consist of pairs of finite sets $\nodet{S_1}{S_2}$ indexed by positive
+and negative polarities. These are the 1-dimensional spaces. We iterate this
+construction to the limit to get $n$-dimensional cubes for all natural
+numbers $n$.
+
+At this point, we switch from viewing the universe of types as an
+unstructured collection of spaces to a viewing it as a \emph{groupoid} with
+explicit \emph{paths} connecting spaces that we want to consider as
+equivalent. We itemize the paths we add next:
+\begin{itemize}
+\item The first step is to identify each space with itself by adding trivial
+  identity paths $\refl{\cubt}$ for each space $\cubt$;
+\item Then we add paths $\seg{\cubt}$ that connect all occurrences of the
+  same type $\cubt$ in various positions in $n$-dimensional cubes. For
+  example, the 1-dimensional space corresponding to $(1-1)$ would now include
+  a path connecting the two endpoints at the end of the previous section;
+\item We then add paths for witnessing the usual type isomorphisms between
+  finite types such as associativity and commutativity of sums and
+  products. The complete list of these isomorphisms is given in the next
+  section.
+\item Finally, we add \emph{2-paths} between $\refl{0}$ and any path
+  $\seg{\cubt}$ whose endpoints are of opposite polarities, i.e., of
+  polarities $s$ and $\negp{s}$ where:
+  \[\begin{array}{rcl}
+  \negp{\epsilon} &=& \epsilon \\
+  \negp{+s} &=& -\negp{s} \\
+  \negp{-s} &=& +\negp{s}
+  \end{array}\]
+\item The groupoid structure forces other paths to be added as described in
+  detail in the recent book on Homotopy Type Theory~\cite{hottbook}. In
+  particular, for every path $p$, the groupoid structure requires an inverse
+  path $p^{-1}$ in the opposite direction; and for every two paths with
+  common ending and starting point, a path corresponding to the composition
+  of the two paths is included and so on. These paths obey various coherence
+  conditions that ensure for example that composing a path and its inverse is
+  equivalent to the trivial path, and that composition is associative, up to
+  path equivalences at higher levels.
+\end{itemize}
 
 \begin{verbatim}
+be careful about that 2-path. Are we 
+allowed to do that given that the 
+empty type is, well, empty.
+
+identical to a single point. It should 
+then follow that a square where one edge 
+contracts to a point would itself contract 
+to a line, and a 3-dimensional cube 
+in which one face contracts to a point 
+would itself contract to a square, etc.
+
 So the story is that as we are building
@@
 leaves in the binary tree representation) which can easily be done using a
-sequence $p$ of $+$ and $-$ polarities indicating how to navigate the cube in
+sequence $s$ of $+$ and $-$ polarities indicating how to navigate the cube in
 each successive dimension starting from a fixed origin to reach the desired
-corner. We write $v^{p}$ for the value $v$ located at corner $p$ of the cube
+corner. We write $v^{s}$ for the value $v$ located at corner $s$ of the cube
 associated with its type. We use $\epsilon$ for the empty sequence of
@@
 doesn't completely specify the type since different types like $(1+(1+1))$
-and $((1+1)+1)$ are assigned the same denotation. What the path $p$ specifies
+and $((1+1)+1)$ are assigned the same denotation. What the path $s$ specifies
 is the \emph{polarity} of the value, or its ``orientation'' in the space
@@
 \qquad
-\infer[\textit{neg}]{v^{\negp{p}} : - \tau}{v^p : \tau} 
+\infer[\textit{neg}]{v^{\negp{s}} : - \tau}{v^s : \tau} 
 \\
-\infer[\textit{left}]{(\inl{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_1}
+\infer[\textit{left}]{(\inl{v})^{s} : \tau_1 + \tau_2}{v^{s} : \tau_1}
 \qquad
-\infer[\textit{right}]{(\inr{v})^{p} : \tau_1 + \tau_2}{v^{p} : \tau_2} 
+\infer[\textit{right}]{(\inr{v})^{s} : \tau_1 + \tau_2}{v^{s} : \tau_2} 
 \\
-\infer[\textit{prod}]{(v_1,v_2)^{p_1 \cdot p_2} : \tau_1 * \tau_2}
-      {v_1^{p_1} : \tau_1 & v_2^{p_2} : \tau_2} 
+\infer[\textit{prod}]{(v_1,v_2)^{s_1 \cdot s_2} : \tau_1 * \tau_2}
+      {v_1^{s_1} : \tau_1 & v_2^{s_2} : \tau_2} 
 \end{array}\]
 The rules \textit{left} and \textit{right} reflect the fact that sums do not
-increase the dimension. Note that when $p$ is $\epsilon$, we get the
+increase the dimension. Note that when $s$ is $\epsilon$, we get the
 conventional values for the 0-dimensional sum type. The rule \textit{prod} is
@@
 that the negation of a value $v$ is the same value $v$ located at the
-``opposite'' corner of the cube:
-\[\begin{array}{rcl}
-\negp{\epsilon} &=& \epsilon \\
-\negp{+p} &=& -\negp{p} \\
-\negp{-p} &=& +\negp{p}
-\end{array}\]
+``opposite'' corner of the cube.
 
@@
 
+Connection to our work on univalence for finite types. We didn't have to rely
+on sets for 0-dimensional types. We could have used groupoids again.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-05-23 ring.tex
typos

@@
 of the figure is the 0-dimensional cube representing the empty type. The path
-$\seg{1}$ identifies the two occurrences of 1. To express the
-equivalence of $(1-1)$ and 0, we add a 2-path, i.e. a path between paths,
-that asserts connects the path $\seg{1}$ to the trivial path
-$\refl{0}$ on the empty type. That effectively makes the two points
-``disappear.''  Surprisingly, that is everything that we need. The extension
-to higher dimensions just ``works'' because paths in homotopy type theory
-have a rich structure. We explain the details below.
+$\seg{1}$ identifies the two occurrences of 1. To express the equivalence of
+$(1-1)$ and 0, we add a 2-path $q$, i.e. a path between paths, that connects
+the path $\seg{1}$ to the trivial path $\refl{0}$. That effectively makes the
+two points ``disappear.''  Surprisingly, that is everything that we need. The
+extension to higher dimensions just ``works'' because paths in homotopy type
+theory have a rich structure. We explain the details below.
 

### 2014-05-24 Pi1.hs
pi1

@@ -1,0 +1,108 @@
+{-# LANGUAGE TypeOperators, GADTs, TypeFamilies #-}
+
+module Pi1 where
+
+-- Pi0 is the usual Pi
+
+-- Pi1 is the next up based on the Int construction (only possible after we
+-- liberate ourselves from the brain-damaging notion that t1-t2 is a sum
+-- type)
+
+------------------------------------------------------------------------------
+-- Pi0
+
+data Zero
+
+data a :<-> b where 
+  Id           :: a :<-> a
+  Sym          :: (a :<-> b) -> (b :<-> a) 
+  (:.:)        :: (a :<-> b) -> (b :<-> c) -> (a :<-> c)
+  (:*:)        :: (a :<-> b) -> (c :<-> d) -> ((a,c) :<-> (b,d))
+  (:+:)        :: (a :<-> b) -> (c :<-> d) -> (Either a c :<-> Either b d)
+  PlusZeroL    :: Either Zero a :<-> a
+  PlusZeroR    :: a :<-> Either Zero a
+  CommutePlus  :: Either a b :<-> Either b a
+  AssocPlusL   :: Either a (Either b c) :<-> Either (Either a b) c 
+  AssocPlusR   :: Either (Either a b) c :<-> Either a (Either b c) 
+  TimesOneL    :: ((), a) :<-> a
+  TimesOneR    :: a :<-> ((), a)
+  CommuteTimes :: (a,b) :<-> (b,a) 
+  AssocTimesL  :: (a,(b,c)) :<-> ((a,b),c)
+  AssocTimesR  :: ((a,b),c) :<-> (a,(b,c))
+  TimesZeroL   :: (Zero, a) :<-> Zero
+  TimesZeroR   :: Zero :<-> (Zero, a)
+  Distribute   :: (Either b c, a) :<-> Either (b, a) (c, a)
+  Factor       :: Either (b, a) (c, a) :<-> (Either b c, a)
+
+-- values are boring at this level
+
+type Val0 a = a
+
+eval0 :: (a :<-> b) -> Val0 a -> Val0 b
+eval0 Id a = a
+eval0 c _ = error "I am not writing this again; get it from somewhere"
+
+------------------------------------------------------------------------------
+-- Pi1 
+
+class Type1 p where
+  type Pos p      :: *
[…]

### 2014-05-25 Pi1.hs
eta epsilon composition

@@ -66,15 +66,27 @@ instance Type1 (ap :- am) where
   type Lolli1 (ap :- am) (bp :- bm) = (Either am bp) :- (Either ap bm)
 
 data a :<=> b where 
-  Id1           :: a :<=> a
-  Sym1          :: (a :<=> b) -> (b :<=> a) 
-  (:..:)        :: (a :<=> b) -> (b :<=> c) -> (a :<=> c)
-  (:++:)        :: (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
-  PlusZeroL1    :: Plus1 Zero1 a :<=> a
-  PlusZeroR1    :: a :<=> Plus1 Zero1 a
-  CommutePlus1  :: Plus1 a b :<=> Plus1 b a
-  AssocPlusL1   :: Plus1 a (Plus1 b c) :<=> Plus1 (Plus1 a b) c 
-  AssocPlusR1   :: Plus1 (Plus1 a b) c :<=> Plus1 a (Plus1 b c) 
+  Id1           :: (a ~ (ap :- am)) => a :<=> a
+  Sym1          :: (a ~ (ap :- am), b ~ (bp :- bm)) => (a :<=> b) -> (b :<=> a) 
+  (:..:)        :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
+                   (a :<=> b) -> (b :<=> c) -> (a :<=> c)
+  (:++:)        :: 
+    (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm), d ~ (dp :- dm)) => 
+    (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+  PlusZeroL1    :: (a ~ (ap :- am)) => Plus1 Zero1 a :<=> a
+  PlusZeroR1    :: (a ~ (ap :- am)) => a :<=> Plus1 Zero1 a
+  CommutePlus1  :: (a ~ (ap :- am), b ~ (bp :- bm)) => Plus1 a b :<=> Plus1 b a
+  AssocPlusL1   :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
+                   Plus1 a (Plus1 b c) :<=> Plus1 (Plus1 a b) c 
+  AssocPlusR1   :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
[…]

### 2014-05-25 Pi1.hs
pi and epsilon

@@ -88,6 +88,12 @@ data a :<=> b where
   Eta :: Zero1 :<=> (ap :- ap)
   Epsilon :: (ap :- ap) :<=> Zero1
 
+{-- Not sure if Eta/Epsilon are the right thing in this context. They do
+allow to introduce a 2-path between any 1-path : t -> t and refl_t --}
+
+allid :: (a :- a) :<=> (a :- a)
+allid = Epsilon :..: Eta
+
 data Val1 p = Val1 { c :: Neg p :<-> Pos p }
 
 eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) => 

### 2014-05-25 Pi1.hs
pi and epsilon

@@ -94,6 +94,10 @@ allow to introduce a 2-path between any 1-path : t -> t and refl_t --}
 allid :: (a :- a) :<=> (a :- a)
 allid = Epsilon :..: Eta
 
+-- Not compact closed yet... need to iterate this construction for every
+-- so curry/uncurry, name/coname, compose/apply, etc won't work just yet
+-- remember that we only have 2nd-order functions...
+
 data Val1 p = Val1 { c :: Neg p :<-> Pos p }
 
 eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) => 

### 2014-05-25 Pi1.hs
pi and epsilon

@@ -89,7 +89,7 @@ data a :<=> b where
   Epsilon :: (ap :- ap) :<=> Zero1
 
 {-- Not sure if Eta/Epsilon are the right thing in this context. They do
-allow to introduce a 2-path between any 1-path : t -> t and refl_t --}
+allow to introduce a 2-path between any 1-path : t -> t and refl_t. If correct, one of the nice things is that we get 2-paths between c;refl and c !!! --}
 
 allid :: (a :- a) :<=> (a :- a)
 allid = Epsilon :..: Eta

### 2014-05-25 Pi1.hs
done with Pi1

@@ -6,7 +6,12 @@ module Pi1 where
 
 -- Pi1 is the next up based on the Int construction (only possible after we
 -- liberate ourselves from the brain-damaging notion that t1-t2 is a sum
--- type)
+-- type). In particular we only do part of the Int construction by adding the
+-- types t1-t2 but keeping them separate. We don't get a compact closed
+-- category. We only get 2nd-order functions. Note that the Int construction
+-- gives you integers but without multiplication. If you want multiplication
+-- you have to do something much more involved like the ring completion
+-- paper.
 
 ------------------------------------------------------------------------------
 -- Pi0
@@ -80,23 +85,37 @@ data a :<=> b where
                    Plus1 a (Plus1 b c) :<=> Plus1 (Plus1 a b) c 
   AssocPlusR1   :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
                    Plus1 (Plus1 a b) c :<=> Plus1 a (Plus1 b c) 
-  Curry1 :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
-            (Plus1 a b :<=> c) -> (a :<=> Lolli1 b c)
-  Uncurry1 :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
-              (a :<=> Lolli1 b c) -> (Plus1 a b :<=> c)
   Dual :: (a ~ (ap :- am)) => a :<=> Dual1 a
+
+{-- 
+
+Adding:
+
   Eta :: Zero1 :<=> (ap :- ap)
   Epsilon :: (ap :- ap) :<=> Zero1
 
-{-- Not sure if Eta/Epsilon are the right thing in this context. They do
-allow to introduce a 2-path between any 1-path : t -> t and refl_t. If correct, one of the nice things is that we get 2-paths between c;refl and c !!! --}
+eval1 Eta (Val1 ca) = Val1 { c = Id } 
+eval1 Epsilon (Val1 ca) = Val1 { c = Id } 
+
[…]

### 2014-05-25 Pi1.hs
spoke too soon still need to do the plus functor and not sure it is possible

@@ -77,7 +77,7 @@ data a :<=> b where
                    (a :<=> b) -> (b :<=> c) -> (a :<=> c)
   (:++:)        :: 
     (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm), d ~ (dp :- dm)) => 
-    (a :<=> b) -> (c :<=> d) -> (Either a c :<=> Either b d)
+    (a :<=> b) -> (c :<=> d) -> (Plus1 a c :<=> Plus1 b d)
   PlusZeroL1    :: (a ~ (ap :- am)) => Plus1 Zero1 a :<=> a
   PlusZeroR1    :: (a ~ (ap :- am)) => a :<=> Plus1 Zero1 a
   CommutePlus1  :: (a ~ (ap :- am), b ~ (bp :- bm)) => Plus1 a b :<=> Plus1 b a
@@ -117,6 +117,8 @@ as well as name/coname, compose/apply, etc. won't work just yet.
    
 --}
 
+{-- Not sure if I get a 2-path between c;id and c ??? --}
+
 data Val1 p = Val1 { c :: Neg p :<-> Pos p }
 
 eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) => 
@@ -124,6 +126,7 @@ eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) =>
 eval1 Id1 v = v
 eval1 (Sym1 c) v = eval1B c v
 eval1 (c1 :..: c2) v = eval1 c2 (eval1 c1 v)
+eval1 (c1 :++: c2) v = 
 eval1 PlusZeroL1 (Val1 c0) = 
   Val1 { c = PlusZeroR :.: c0 :.: PlusZeroL }
 eval1 PlusZeroR1 (Val1 c0) = 
@@ -141,6 +144,7 @@ eval1B :: (a ~ (ap :- am), b ~ (bp :- bm)) =>
 eval1B Id1 v = v
 eval1B (Sym1 c) v = eval1 c v
 eval1B (c1 :..: c2) v = eval1B c1 (eval1B c2 v)
+eval1 (c1 :++: c2) v = undefined
 eval1B PlusZeroL1 (Val1 c0) =
   Val1 { c = PlusZeroL :.: c0 :.: PlusZeroR }
 eval1B PlusZeroR1 (Val1 c0) =

### 2014-05-25 Pi1.hs
typo

@@ -126,7 +126,7 @@ eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) =>
 eval1 Id1 v = v
 eval1 (Sym1 c) v = eval1B c v
 eval1 (c1 :..: c2) v = eval1 c2 (eval1 c1 v)
-eval1 (c1 :++: c2) v = 
+eval1 (c1 :++: c2) v = error "is that possible?"
 eval1 PlusZeroL1 (Val1 c0) = 
   Val1 { c = PlusZeroR :.: c0 :.: PlusZeroL }
 eval1 PlusZeroR1 (Val1 c0) = 
@@ -144,7 +144,7 @@ eval1B :: (a ~ (ap :- am), b ~ (bp :- bm)) =>
 eval1B Id1 v = v
 eval1B (Sym1 c) v = eval1 c v
 eval1B (c1 :..: c2) v = eval1B c1 (eval1B c2 v)
-eval1 (c1 :++: c2) v = undefined
+eval1B (c1 :++: c2) v = undefined
 eval1B PlusZeroL1 (Val1 c0) =
   Val1 { c = PlusZeroL :.: c0 :.: PlusZeroR }
 eval1B PlusZeroR1 (Val1 c0) =

### 2014-05-25 Pi1.hs
confusion

@@ -56,8 +56,6 @@ class Type1 p where
   type Zero1      :: * 
   type One1       :: *
   type Plus1 p q  :: *
-  type Dual1 p    :: * 
-  type Lolli1 p q :: * 
 
 data a :- b = a :- b
 
@@ -67,8 +65,6 @@ instance Type1 (ap :- am) where
   type Zero1 = Zero :- Zero
   type One1 = () :- Zero
   type Plus1  (ap :- am) (bp :- bm) = (Either ap bp) :- (Either am bm)
-  type Dual1  (ap :- am) = am :- ap
-  type Lolli1 (ap :- am) (bp :- bm) = (Either am bp) :- (Either ap bm)
 
 data a :<=> b where 
   Id1           :: (a ~ (ap :- am)) => a :<=> a
@@ -76,8 +72,8 @@ data a :<=> b where
   (:..:)        :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
                    (a :<=> b) -> (b :<=> c) -> (a :<=> c)
   (:++:)        :: 
-    (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm), d ~ (dp :- dm)) => 
-    (a :<=> b) -> (c :<=> d) -> (Plus1 a c :<=> Plus1 b d)
+    (p ~ (a :- b), q ~ (c :- d), r ~ (e :- f), s ~ (g :- h)) => 
+    (p :<=> q) -> (r :<=> s) -> (Plus1 p r :<=> Plus1 q s)
   PlusZeroL1    :: (a ~ (ap :- am)) => Plus1 Zero1 a :<=> a
   PlusZeroR1    :: (a ~ (ap :- am)) => a :<=> Plus1 Zero1 a
   CommutePlus1  :: (a ~ (ap :- am), b ~ (bp :- bm)) => Plus1 a b :<=> Plus1 b a
@@ -85,7 +81,6 @@ data a :<=> b where
                    Plus1 a (Plus1 b c) :<=> Plus1 (Plus1 a b) c 
   AssocPlusR1   :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
                    Plus1 (Plus1 a b) c :<=> Plus1 a (Plus1 b c) 
-  Dual :: (a ~ (ap :- am)) => a :<=> Dual1 a
 
 {-- 
 
[…]

### 2014-05-25 Pi1.hs
just make some warnings go away

@@ -45,7 +45,7 @@ type Val0 a = a
 
 eval0 :: (a :<-> b) -> Val0 a -> Val0 b
 eval0 Id a = a
-eval0 c _ = error "I am not writing this again; get it from somewhere"
+eval0 c0 _ = error "I am not writing this again; get it from somewhere"
 
 ------------------------------------------------------------------------------
 -- Pi1 
@@ -124,9 +124,10 @@ data Val1 p = Val1 { c :: Neg p :<-> Pos p }
 eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) => 
          (a :<=> b) -> Val1 a -> Val1 b
 eval1 Id1 v = v
-eval1 (Sym1 c) v = eval1B c v
+eval1 (Sym1 c1) v = eval1B c1 v
 eval1 (c1 :..: c2) v = eval1 c2 (eval1 c1 v)
-eval1 (c1 :++: c2) v = error "is that possible?"
+eval1 (c1 :++: c2) (Val1 c0) = 
+  Val1 { c = undefined }
 eval1 PlusZeroL1 (Val1 c0) = 
   Val1 { c = PlusZeroR :.: c0 :.: PlusZeroL }
 eval1 PlusZeroR1 (Val1 c0) = 
@@ -142,7 +143,7 @@ eval1 Dual (Val1 ca) = Val1 { c = Sym ca }
 eval1B :: (a ~ (ap :- am), b ~ (bp :- bm)) => 
           (a :<=> b) -> Val1 b -> Val1 a
 eval1B Id1 v = v
-eval1B (Sym1 c) v = eval1 c v
+eval1B (Sym1 c1) v = eval1 c1 v
 eval1B (c1 :..: c2) v = eval1B c1 (eval1B c2 v)
 eval1B (c1 :++: c2) v = undefined
 eval1B PlusZeroL1 (Val1 c0) =

### 2014-05-25 Pi1.hs
be less specific

@@ -72,7 +72,7 @@ data a :<=> b where
   (:..:)        :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm)) => 
                    (a :<=> b) -> (b :<=> c) -> (a :<=> c)
   (:++:)        :: 
-    (p ~ (a :- b), q ~ (c :- d), r ~ (e :- f), s ~ (g :- h)) => 
+    -- (p ~ (a :- b), q ~ (c :- d), r ~ (e :- f), s ~ (g :- h)) => 
     (p :<=> q) -> (r :<=> s) -> (Plus1 p r :<=> Plus1 q s)
   PlusZeroL1    :: (a ~ (ap :- am)) => Plus1 Zero1 a :<=> a
   PlusZeroR1    :: (a ~ (ap :- am)) => a :<=> Plus1 Zero1 a
@@ -122,7 +122,8 @@ Not sure if I get a 2-path between c;id and c ???
 
 data Val1 p = Val1 { c :: Neg p :<-> Pos p }
 
-eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) => 
+-- eval1 :: (a ~ (ap :- am), b ~ (bp :- bm)) => 
+eval1 :: (Type1 a, Type1 b) =>
          (a :<=> b) -> Val1 a -> Val1 b
 eval1 Id1 v = v
 eval1 (Sym1 c1) v = eval1B c1 v

### 2015-01-09 Univalence/perm.tex
collect things from emails that might relevant for paper in perm.tex

That permutations form a groupoid AND a semiring (with all sorts of nice coherence laws) is nice.  And the non-trivial computational interpretations of both the co-product and tensor products could be explained more fully.  So yes, without going into all the Pi stuff, we probably could package up the permutation work into a small paper.  But where?

I don’t have a feeling for the significance bits in the paper yet and how precisely it improves on Gonthier’s encoding of permutations for example.

I see the current work on normal form to be more or less integral to a story on permutations. One could argue otherwise but it makes a very nice story to be able to switch between syntactic and semantic representations of permutations. A paper on a sound and complete calculus of permutations supported by an Agda library would be nice, don’t you think?

About the different representations, I found the Cauchy two line representation to be the easiest to manipulate when talking semantics. When switching to syntactic representations, a sequence of permutations appears to be the right syntactic primitive.

You make a convincing case.  And not just permutations, but permutations with 2 monoidal structures on top of composition.  Most people stop after composition.  90\% of the difficulty came from + and * !

See Bob Atkey's buried in
https://github.com/bobatkey/sorting-types/blob/master/agda/Linear.agda

and a number of our additional lemmas can be found already in
http://www.cse.chalmers.se/~nad/repos/equality/Equality.agda

If M is finite, an injective function f : M -> M must also be surjective. I think this is our situation. So surjective should be free.

I agree.  It's just that every single proof I tried (in my head) ended up using contradiction.  And that might still be fine, since I don't think we would ever use the 'content' of that proof anyways.

But
http://mathoverflow.net/questions/178778/in-the-category-of-sets-epimorphisms-are-surjective-constructive-proof
is heartening.

### 2015-03-08 Univalence/outline.tex
outline.tex

\documentclass{article}
\usepackage{fullpage}
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Permutations}

\paragraph*{Cauchy representation.} A permutation on $n$ elements is represented by a vector \texttt{v : Vec (Fin n) n}. If the $n$ elements are indexed by positions, the element at position $i$ is mapped to position \texttt{v !! i} by the permutation. There is always a trivial permutation called \texttt{1C} that maps each position to itself. The Cauchy representation does not enforce that the vector entries are disjoint. This is enforced by the definition of ``concrete permutations'' below.

\paragraph*{Concrete permutation.} A concrete permutation consists of two Cauchy vectors and two proofs that their compositions is the identity permutation \texttt{1C}. Concrete permutations are an equivalence relation. Concrete permutations actually have more structure: a sum, a unit for the sum, etc. We can also build setoids whose carriers are concrete permutations under the standard $\equiv$ propositional equality.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2015-03-20 Univalence/outline.tex

@@
 \begin{document}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Permutations}
+\section{Basic Utilities}
+
+\paragraph*{LeqLemmas.} A few lemmas about natural numbers.
+
+\paragraph*{FinNatLemmas.} A few lemmas about $\texttt{Fin}~n$ which are the numbers used to index into vectors.
+
+\paragraph*{SubstLemmas.} A few lemmas about compositions of propositional equalities..
+
+\paragraph*{VectorLemmas.} A few lemmas about vectors, lookups, mapping functions over vectors, etc.
+
+\paragraph*{FiniteFunctions.} Proves extensionality for finite functions.
+
+\paragraph*{Proofs.} Collects all the above and re-exports them along with a couple of other general utilities for managing Agda proofs.
+
+\paragraph*{DivModUtils.} External library for reasoning about uniqueness of \texttt{divMod}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Structures}
+
+\paragraph*{SetoidUtils.} Any type $A$ can be viewed as a setoid with propositional equality $\equiv$ as the equivalence relation.
+
+\paragraph*{Groupoid.} A definition of 1-groupoids and operations on them.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Equivalences}
+
+\paragraph*{Equiv.} Defines extensional equivalence of functions $\sim$ and shows that it is an equivalence relation.
+Defines equivalence between sets $\simeq$ using two functions that go back and forth and whose compositions are extensionally equivalent to the identity, and shows that this equivalence is indeed an equivalence relation.
+Finally shows that equivalences are injective.
+
+\paragraph*{TypeEquivalences.} Establishes that the Agda types $\bot$, $\top$, $\uplus$, and $\times$ form a commutative semiring using $\simeq$ as the underlying equivalence relation.
+
+\paragraph*{FinEquiv.} Establishes that $\texttt{Fin}~n$ also forms a commutative semiring with $\simeq$ as the underlying equivalence relation. In particular, we have:
+\[\begin{array}{rcll}
+\texttt{Fin}~0 &\simeq& \bot \\
+\texttt{Fin}~1 &\simeq& \top \\
+\texttt{Fin}~(m+n) &\simeq& \texttt{Fin}~m \uplus \texttt{Fin}~n \\
+\texttt{Fin}~(m*n) &\simeq& \texttt{Fin}~m \times \texttt{Fin}~n
+\end{array}\]
+and then we have all the commutative semiring axioms, e.g., $\texttt{Fin}~(0+m) \simeq \texttt{Fin}~m$. \textbf{SOME AXIOMS MISSING: TO COMPLETE ???}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{To do}
+
+PiLevel0
+ConcretePermutation
+VecOps
+PiPerm
+
+Enumeration ??
+SEequivSCPermEquiv ??
 

### 2015-03-22 Univalence/outline.tex
renaming

@@
 
-\paragraph*{TypeEquivalences.} Establishes that the Agda types $\bot$, $\top$, $\uplus$, and $\times$ form a commutative semiring using $\simeq$ as the underlying equivalence relation.
+\paragraph*{TypeEquiv.} Establishes that the Agda types $\bot$, $\top$, $\uplus$, and $\times$ form a commutative semiring using $\simeq$ as the underlying equivalence relation.
 
@@
 \end{array}\]
-and then we have all the commutative semiring axioms, e.g., $\texttt{Fin}~(0+m) \simeq \texttt{Fin}~m$. \textbf{SOME AXIOMS MISSING: TO COMPLETE ???}
+and then we have all the commutative semiring axioms, e.g.,
+$\texttt{Fin}~(0+m) \simeq \texttt{Fin}~m$.
+
+\paragraph*{CauchyEquiv.}
 

### 2015-03-22 Univalence/outline.tex
cauchy equiv postulate sym

@@
 \usepackage{fullpage}
+\usepackage{amssymb}
 \begin{document}
@@
 and then we have all the commutative semiring axioms, e.g.,
-$\texttt{Fin}~(0+m) \simeq \texttt{Fin}~m$.
+$\texttt{Fin}~(0+m) \simeq \texttt{Fin}~m$. The actual proof says that
+0, 1, $+$, and $*$ for a commutative semiring structure under the
+equivalence that equates $m$ and $n$ if $\texttt{Fin}~m \simeq \texttt{Fin}~n$.
 
-\paragraph*{CauchyEquiv.}
+\paragraph*{CauchyEquiv.} Defines the Cauchy representation of
+permutations as
+$\texttt{Cauchy}~m~n = \texttt{Vec}~(\texttt{Fin}~m)~n$. The
+representation is not tight at this point as it neither enforces that
+$m=n$ nor that the elements of the vectors are a permutation of the
+set $\{0,\ldots,n-1\}$. This is however a building block for actual
+permutations (defined in \texttt{ConcretePermutation}) which enforce
+both constraints. For that reason, the Cauchy representation has no
+inverse and hence is not a commutative semiring unless one postulates
+the following:
+\[
+\textbf{postulate}~\texttt{sym-iso}~:~\{m~n~:~\mathbb{N}\} \rightarrow
+  \texttt{Cauchy}~m~n \rightarrow \texttt{Cauchy}~n~m
+\]
 

### 2015-03-22 Univalence/outline.tex
outline
small cleanups

@@
 
-\paragraph*{FinNatLemmas.} A few lemmas about $\texttt{Fin}~n$ which are the numbers used to index into vectors.
+\paragraph*{FinNatLemmas.} A few lemmas about $\texttt{Fin}~n$ which
+are the numbers used to index into vectors.
 
-\paragraph*{SubstLemmas.} A few lemmas about compositions of propositional equalities..
+\paragraph*{SubstLemmas.} A few lemmas about compositions of
+propositional equalities..
 
-\paragraph*{VectorLemmas.} A few lemmas about vectors, lookups, mapping functions over vectors, etc.
+\paragraph*{VectorLemmas.} A few lemmas about vectors, lookups,
+mapping functions over vectors, etc.
 
@@
 
-\paragraph*{Proofs.} Collects all the above and re-exports them along with a couple of other general utilities for managing Agda proofs.
+\paragraph*{Proofs.} Collects all the above and re-exports them along
+with a couple of other general utilities for managing Agda proofs.
 
-\paragraph*{DivModUtils.} External library for reasoning about uniqueness of \texttt{divMod}.
+\paragraph*{DivModUtils.} External library for reasoning about
+uniqueness of \texttt{divMod}. This is used extensively to mediate
+between $\texttt{Fin}~(m*n)$ and
+$\texttt{Fin}~m \times \texttt{Fin}~n$.
 
@@
 
-\paragraph*{SetoidUtils.} Any type $A$ can be viewed as a setoid with propositional equality $\equiv$ as the equivalence relation.
+\paragraph*{SetoidUtils.} Any discrete type $A$ can be viewed as a
+setoid with propositional equality $\equiv$ as the equivalence
+relation.
 
-\paragraph*{Groupoid.} A definition of 1-groupoids and operations on them.
+\paragraph*{Groupoid.} A definition of 1-groupoids and operations on
+them.
 
@@
 
-\paragraph*{Equiv.} Defines extensional equivalence of functions $\sim$ and shows that it is an equivalence relation.
-Defines equivalence between sets $\simeq$ using two functions that go back and forth and whose compositions are extensionally equivalent to the identity, and shows that this equivalence is indeed an equivalence relation.
-Finally shows that equivalences are injective.
+\paragraph*{Equiv.} Defines extensional equivalence of functions
+$\sim$ and shows that it is an equivalence relation. Defines
+equivalence between sets $\simeq$ using two functions that go back and
+forth and whose compositions are extensionally equivalent to the
+identity, and shows that this equivalence is indeed an equivalence
+relation. Finally shows that equivalences are injective and form a
+congruence that respects $\uplus$ and $\times$.
 
-\paragraph*{TypeEquiv.} Establishes that the Agda types $\bot$, $\top$, $\uplus$, and $\times$ form a commutative semiring using $\simeq$ as the underlying equivalence relation.
+\paragraph*{TypeEquiv.} Establishes that the Agda types $\bot$,
+$\top$, $\uplus$, and $\times$ form a commutative semiring using
+$\simeq$ as the underlying equivalence relation.
 
-\paragraph*{FinEquiv.} Establishes that $\texttt{Fin}~n$ also forms a commutative semiring with $\simeq$ as the underlying equivalence relation. In particular, we have:
+\paragraph*{FinEquiv.} Establishes that $\texttt{Fin}~n$ also forms a
+commutative semiring with $\simeq$ as the underlying equivalence
+relation. In particular, we have:
 \[\begin{array}{rcll}
@@
 0, 1, $+$, and $*$ for a commutative semiring structure under the
-equivalence that equates $m$ and $n$ if $\texttt{Fin}~m \simeq \texttt{Fin}~n$.
+equivalence that equates $m$ and $n$ if
+$\texttt{Fin}~m \simeq \texttt{Fin}~n$.
 
-\paragraph*{CauchyEquiv.} Defines the Cauchy representation of
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{To do}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Bimonoidal Categories}
+
+\paragraph*{CauchyBimonoidal.} Defines the Cauchy representation of
 permutations as
@@
 \]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{To do}
+The Cauchy representation has more structure: it is almost a groupoid
+again modulo two postulates asserting that \texttt{sym-iso} is really
+a left and right inverse.
 
@@
 
-\paragraph*{Cauchy representation.} A permutation on $n$ elements is represented by a vector \texttt{v : Vec (Fin n) n}.
-If the $n$ elements are indexed by positions, the element at position $i$ is mapped to position \texttt{v !!
-i} by the permutation.
-There is always a trivial permutation called \texttt{1C} that maps each position to itself.
-The Cauchy representation does not enforce that the vector entries are disjoint.
-This is enforced by the definition of ``concrete permutations'' below.
+\paragraph*{Cauchy representation.} A permutation on $n$ elements is
+represented by a vector \texttt{v : Vec (Fin n) n}. If the $n$
+elements are indexed by positions, the element at position $i$ is
+mapped to position \texttt{v !! i} by the permutation. There is always
+a trivial permutation called \texttt{1C} that maps each position to
+itself. The Cauchy representation does not enforce that the vector
+entries are disjoint. This is enforced by the definition of ``concrete
+permutations'' below.
 
-\paragraph*{Concrete permutation.} A concrete permutation consists of two Cauchy vectors and two proofs that their compositions is the identity permutation \texttt{1C}.
-Concrete permutations are an equivalence relation.
-Concrete permutations actually have more structure: a sum, a unit for the sum, etc.
-We can also build setoids whose carriers are concrete permutations under the standard $\equiv$ propositional equality.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{?}
+\paragraph*{Concrete permutation.} A concrete permutation consists of
+two Cauchy vectors and two proofs that their compositions is the
+identity permutation \texttt{1C}. Concrete permutations are an
+equivalence relation. Concrete permutations actually have more
+structure: a sum, a unit for the sum, etc. We can also build setoids
+whose carriers are concrete permutations under the standard $\equiv$
+propositional equality.
 

### 2015-03-23 Univalence/outline.tex
added new file FinSetCategory to eventually replace CauchyEquiv

--- 2015-03-22 Univalence/outline.tex
@@
 and then we have all the commutative semiring axioms, e.g.,
-$\texttt{Fin}~(0+m) \simeq \texttt{Fin}~m$.
+$\texttt{Fin}~(0+m) \simeq \texttt{Fin}~m$. The actual proof says that
+0, 1, $+$, and $*$ for a commutative semiring structure under the
+equivalence that equates $m$ and $n$ if
+$\texttt{Fin}~m \simeq \texttt{Fin}~n$.
 
-\paragraph*{CauchyEquiv.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Bimonoidal Categories}
+
+\paragraph*{FinSetCategory.} Defines the category of finite sets and
+all functions between them. As stated on the nLab page, ``it is easy
+(and thus common) to make \texttt{FinSet} skeletal; there is one
+object for each natural number $n$ (including $n=0$), and a morphism
+from $m$ to $n$ is an $m$-tuple $(f_0,\ldots,f_{m−1})$ of numbers
+satisfying $0 \leq f_i < n$. This amounts to identifying $n$ with the
+set $\{0,\ldots,n−1\}$.'' This structure will be a building block for
+permutations (defined in \texttt{ConcretePermutation}).
 

### 2015-03-26 Univalence/outline.tex

@@
 
+\paragraph*{SymmetricMonoidalCategory.} A definition of symmetric
+monoidal categories.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\paragraph*{FinSetCategory.} Defines the category of finite sets and
-all functions between them. As stated on the nLab page, ``it is easy
-(and thus common) to make \texttt{FinSet} skeletal; there is one
-object for each natural number $n$ (including $n=0$), and a morphism
-from $m$ to $n$ is an $m$-tuple $(f_0,\ldots,f_{m−1})$ of numbers
-satisfying $0 \leq f_i < n$. This amounts to identifying $n$ with the
-set $\{0,\ldots,n−1\}$.'' This structure will be a building block for
+\paragraph*{SkFinSetCategory.} Defines the skeletal category of finite
+sets and all functions between them. There is one object for each
+natural number $n$ (including $n=0$), and a morphism from $m$ to $n$
+is an $m$-tuple $(f_0,\ldots,f_{m−1})$ of numbers satisfying
+$0 \leq f_i < n$. This structure will be a building block for
 permutations (defined in \texttt{ConcretePermutation}).

### 2015-04-03 Univalence/outline.tex
added Univalence.agda which has everything; resolved a few inconsistencies with imports; SEquivSCPermEquiv was broken for some reason
outline with high level goals and structure

@@
 \begin{document}
+
+\newcommand{\PiL}{\textsc{\textbf{Pi}}}
+\newcommand{\PiCalc}{\PiL${}^\equiv$}
+\newcommand{\PiCat}{\ensuremath{\Pi\mathbb{C}}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Goal}
+
+Our first major milestone is (i) to develop \emph{a sound and complete
+  calculus of permutations over finite sets}, (ii) to define, in that
+setting, a variant of \emph{univalence} that has a computational
+content, and to \emph{prove} this univalence property. In the longer
+term, we will want to generalize this framework to sets with negative
+and fractional cardinalities which would effectively give us (at
+least) higher-order functions. One can in fact go further and consider
+sets with more general cardinalities: imaginary numbers, irrational
+numbers, all the way to algebraic numbers.
+
+To reach the first major milestone, our detailed steps are:
+\begin{itemize}
+
+\item to develop a \emph{syntactic notion of permutations over finite
+    sets}. We achieve this by defining a typed programming language
+  \PiL\ whose types denote finite sets and whose expressions denote
+  permutations over finite sets. This language has been completely
+  developed: its syntax, type system, and operational semantics are
+  well-understood;
+
+\item to develop a \emph{calculus \PiCalc\ for reasoning about \PiL\
+    permutations}. We eventually want to prove that this calculus is
+  sound and complete with respect to a semantic notion of permutation
+  equivalence;
+
+\item to develop a \emph{semantic notion of equivalence of
+    permutations over finite sets}.  To make the connections to HoTT
+  more evident, we formalize this equivalence using a category
+  (groupoid actually) of finite sets and permutations \PiCat\ which we
+  establish is \emph{rig category}, i.e., it has two symmetric
+  monoidal structures with the multiplicative one distributing over
+  the additive one. The objects of this category are discrete
+  groupoids representing finite sets; the morphisms of \PiCat\ are
+  permutations over the finite sets; the complete definition of the
+  category also needs an equivalence relation that specifies when two
+  morphisms should be considered equal. This equivalence relation is
+  our notion of semantic equivalence of permutations over finite
+  sets. The advantage of using the categorical setting to define the
+  semantic notion of equivalence is that the coherence laws for rig
+  categories provide us with an axiomatization of semantic equivalence
+  of permutations that is rich enough to reason about sequential
+  compositions of permutations (given by the plain categorical
+  structure), parallel compositions of permutations (given by the
+  additive monoidal structure), and tensor compositions of
+  permutations (given by the multiplicative monoidal structure) while
+  taking care of all the induced equivalences that result from the
+  interactions among these structures.
+
+\item to derive the complete definition of \PiCalc\ by reifying each
+  of the primitives used in defining semantic equivalence as syntactic
+  objects; a proof of soundness and completeness is then immediate as
+  the syntactic and semantic structures are isomorphic by construction.
+
+\item to define a variant of univalence in the current setting we
+  reason as follows. In the HoTT setting, which is not limited to
+  finite sets and permutations, we have a larger category of arbitrary
+  sets and with equivalences as morphisms. Univalence in that context
+  postulates that each equivalence between sets induces a path between
+  these sets. Translated to our setting, univalence postulates that
+  each semantic equivalence between sets induces a syntactic
+  equivalence between these sets. In the richer setting of HoTT,
+  univalence is postulated because there is no obvious way to reason
+  about extensional equivalence of the functions used to define
+  semantic equivalences. In our setting, these semantic equivalences
+  are defined using permutations which are amenable to equational
+  reasoning via the coherence laws. Therefore to make univalence
+  constructive, all we need is a computational mechanism to calculate
+  a syntactic equivalence from a semantic one. This mechanism is
+  embedded in the proof of completeness above.
+
+\item Our complete development is formalized in Agda. 
+\end{itemize}
 
@@
 them.
+
+\paragraph*{Categories.Everything.} Basic and some advanced category theory.
 

### 2015-04-04 Univalence/outline.tex
pi category

@@
 \end{itemize}
+
+Which still leaves open one big question: is our main representation
+of permutations _fin~_ or CPerm ?  Or do we need both, as one is
+'syntactic' (through the use of vectors) while the other is more
+abstract (as it uses functions)?
+
+I am quite sure that all the work needed to bring FinEquivCat all the
+way to RigCategory is feasible, and not too hard [but will be
+enlightening].
+
+To do the same with CPermCat (and SkFinSetCategory) will be a fair bit
+harder.  So it is very important for us to decide what the exact path
+we want to take.  [Which is indeed what your outline is about!]
 

### 2015-09-18 Univalence/response.txt
response started

We thank the reviewers for their comments.

A general theme of the reviews appears to request a better presentation of the theorems and how they all relate to each other. A compressed and intuitive presentation would be as follows. First we have three independent constructions:

  A : Set     <------- HoTT-style equivalences -------> B : Set

  A : Fin m <----------- Permutations ------------> B : Fin n

  A : Pi       <----------- Pi-combinators ----------> B : Pi

Each construction relates a family of types via appropriate relations; each construction has a rig structure (additive and multiplicative units, commutative and associative sums and products, and distributivity of products over sums). We then have a second level:

  e : HoTT-Equivalence <----------- Eq. of equivalences --------> f : HoTT-Equivalence

  p : Permutations         <----------- Eq. of permutations ------->  q : Permutations

  c : Pi-combinators     <------------- Pi-optimizations --------> d : Pi-combinators

We then have theorems that connect some of the constructions vertically allowing us to move across representations.

This are the "raw" results without interpretation, context, or applications. Perhaps unfortunately the current presentation of the paper mixes these "raw" results with theoretical connections to HoTT and univalence and with practical connections to reversible circuits and their optimizations. The connections are there and we are important to appreciate the significance of the theorems but the presentation could indeed be better.

### 2015-09-18 Univalence/response.txt
response is at 400 words

@@
-We thank the reviewers for their comments.
-
 A general theme of the reviews appears to request a better presentation of the theorems and how they all relate to each other.
@@
 
-Each construction relates a family of types via appropriate relations; each construction has a rig structure (additive and multiplicative units, commutative and associative sums and products, and distributivity of products over sums).
+Each construction relates a particular representation of types via appropriate relations; each construction has a rig structure (additive and multiplicative units, commutative and associative sums and products, and distributivity of products over sums).
 We then have a second level:
@@
 
-We then have theorems that connect some of the constructions vertically allowing us to move across representations.
+The two levels of constructions are connected via appropriate categories.
+For example, we have a category whose objects are Pi types, whose morphisms are Pi-combinators, and whose equality of morphisms if Pi-optimizations and this category is a (free) symmetric rig groupoid.
+We finally have theorems that connect some of the constructions vertically allowing us to move across representations.
 
 This are the "raw" results without interpretation, context, or applications.
-Perhaps unfortunately the current presentation of the paper mixes these "raw" results with theoretical connections to HoTT and univalence and with practical connections to reversible circuits and their optimizations.
-The connections are there and we are important to appreciate the significance of the theorems but the presentation could indeed be better.
+Perhaps unfortunately the current presentation of the paper mixes these "raw" results with theoretical connections to HoTT, with Agda formalization concerns, and with practical connections to reversible circuits and their optimizations.
+We acknowledge that these connections are somewhat distracting and that the presentation could be much improved but these connections are all critical and essential.
+Indeed some of the connections above are trivial in "conventional mathematics" but some of that triviality hides implicit isomorphisms (changes of representations) that are important computationally.
+Our construction shows, for example, that transporting properties across equivalences of finite types can be done via executing permutations, something which has a clear computational cost and which itself depends on the choice of representations of the permutations.
+Since this is POPL, the presentation is also biased towards turning the constructions into actual executable programming languages with executable optimizers and with an eye towards higher-order extensions.
+To our knowledge, there are no programming languages for representing --and reasoning about-- a rich notion of "permutations" (including sums and products).
+Furthermore, the higher-order extensions are crucial for abstractions over such permutations by allowing the permutations themselves to be treated as data that can be constructed and examined dynamically.

### 2015-09-19 Univalence/response.txt
added short responses

@@
 Furthermore, the higher-order extensions are crucial for abstractions over such permutations by allowing the permutations themselves to be treated as data that can be constructed and examined dynamically.
+
+Specific (and short with apologies) responses:
+
+* I think our excitement with all the connections apparently left a "hint of arrogance." We sincerely apologize for that.
+
+* The supplementary material gives one small example of rewriting circuits using the 2-axioms.
+We do not have larger experiments at this point but we have done enough to notice that the rewriting of one circuit to another does not necessarily take an exponential number of steps: the diagram on p.10 shows an equivalence that is independent of the sizes of c1 and c2 and the sizes of A, B, C, and D.
+
+* Our axioms vs.
+Lapalaza diagrams: In some sense they are indeed "the same".
+The general process is to give judicious names to each side of each diagram and then proceed to prove equivalence of the two expressions.
+In some cases this was achieved by adding a 2-axiom that directly relates the terms but in others it was possible to simplify the two sides using other 2-axioms.
+The process led us to refine the original 1-axioms of Pi which implicitly collapsed certain terms that should only been related by explicit non-trivial paths.
+
+* Proofs: All the theorems (except Thm 2) are in our Agda formalization.
+This code was not presentable as an artifact at the time of submission and required some serious refactoring which is under progress.
+Thm 2 is indeed wrong was the informal statement was included by mistake (but it is subsumed by Thm 9).
+
+* The 3rd axiom on p9 says c;(!c)=id which proves that swap;swap=id.
+(The bottom of p4 notes that every combinator has a dual according to figure 1 with swap being its own inverse.) The existence of inverses for 1-combinators is a meta-property (not a constructor of the inductive datatype of combinators but a function mapping combinators to combinators).
+The same applies to inverses for 2-combinators.
+
+* Completeness in Thm 11. *** BRIEFLY EXPLAIN HOW WE GET COMPLETENESS FROM LAPLAZA ***
+
+* *** COMMENT ON PHONY MULTIPLICATION ***

### 2015-09-19 Univalence/response.txt
first pass on response

@@
-A general theme of the reviews appears to request a better presentation of the theorems and how they all relate to each other.
-A compressed and intuitive presentation would be as follows.
-First we have three independent constructions:
+A general theme of the reviews appears to request a better presentation of the
+theorems and how they all relate to each other. A compressed and intuitive
+presentation would be as follows. First we have three independent
+constructions:
 
@@
 
-  A : Fin m <----------- Permutations ------------> B : Fin n
+  A : Fin m  <------------ Permutations -------------> B : Fin n
 
-  A : Pi       <----------- Pi-combinators ----------> B : Pi
+  A : Pi     <------------ Pi-combinators -----------> B : Pi
 
-Each construction relates a particular representation of types via appropriate relations; each construction has a rig structure (additive and multiplicative units, commutative and associative sums and products, and distributivity of products over sums).
-We then have a second level:
+Each construction relates a particular representation of types via appropriate
+relations; each construction has a rig structure (additive and multiplicative
+units, commutative and associative sums and products, and distributivity of
+products over sums). We then have a second level:
 
-  e : HoTT-Equivalence <----------- Eq. of equivalences --------> f : HoTT-Equivalence
+  e : HoTT-Equivalence <--------- Equiv. of equivalences --------> f : HoTT-Equivalence
 
-  p : Permutations         <----------- Eq. of permutations ------->  q : Permutations
+  p : Permutations     <--------- Equiv. of permutations --------> q : Permutations
 
-  c : Pi-combinators     <------------- Pi-optimizations --------> d : Pi-combinators
+  c : Pi-combinators   <-------------- Pi-optimizations ---------> d : Pi-combinators
 
-The two levels of constructions are connected via appropriate categories.
-For example, we have a category whose objects are Pi types, whose morphisms are Pi-combinators, and whose equality of morphisms if Pi-optimizations and this category is a (free) symmetric rig groupoid.
-We finally have theorems that connect some of the constructions vertically allowing us to move across representations.
+The two levels of constructions are connected, and form categories. For
+example, we have a category whose objects are Pi types, whose morphisms are
+Pi-combinators, and whose equality of morphisms is Pi-optimizations; this
+category is a (free) symmetric rig groupoid. We finally have theorems that
+connect some of the constructions vertically allowing us to move across
+representations.
 
-This are the "raw" results without interpretation, context, or applications.
-Perhaps unfortunately the current presentation of the paper mixes these "raw" results with theoretical connections to HoTT, with Agda formalization concerns, and with practical connections to reversible circuits and their optimizations.
-We acknowledge that these connections are somewhat distracting and that the presentation could be much improved but these connections are all critical and essential.
-Indeed some of the connections above are trivial in "conventional mathematics" but some of that triviality hides implicit isomorphisms (changes of representations) that are important computationally.
-Our construction shows, for example, that transporting properties across equivalences of finite types can be done via executing permutations, something which has a clear computational cost and which itself depends on the choice of representations of the permutations.
-Since this is POPL, the presentation is also biased towards turning the constructions into actual executable programming languages with executable optimizers and with an eye towards higher-order extensions.
-To our knowledge, there are no programming languages for representing --and reasoning about-- a rich notion of "permutations" (including sums and products).
-Furthermore, the higher-order extensions are crucial for abstractions over such permutations by allowing the permutations themselves to be treated as data that can be constructed and examined dynamically.
+These are the "raw" results without interpretation, context, or applications.
+Perhaps unfortunately, the current presentation of the paper mixes these "raw"
+results with theoretical connections to HoTT, with Agda formalization concerns,
+and with practical connections to reversible circuits and their optimizations.
+We acknowledge that these connections can be perceived as distracting, and that
+their presentation could be much improved (which we are working on), BUT these
+connections are all critical and essential -- and are why this is being sent to
+POPL rather than (say) Theory and Applications of Category Theory.
+
+Indeed some of the connections above are trivial in "conventional mathematics";
+however some of those same connections hide implicit isomorphisms (changes of
+representations) that are important computationally.  Our construction shows,
+for example, that transporting properties across equivalences of finite types
+can be done via executing permutations, something which has a clear
+computational cost and which itself depends on the choice of representations of
+the permutations. Since this is POPL, the presentation emphasizes turning these
+constructions into actual executable programming constructs of a programming
+language, complete with executable optimizers, and with an eye towards
+higher-order extensions. To our knowledge, there are no programming languages
+for representing --and reasoning about-- a rich notion of "permutations"
+(including sums and products).  Furthermore, the higher-order extensions are
+crucial for abstractions over such permutations by allowing the permutations
+themselves to be treated as data that can be constructed and examined
+dynamically.
 
@@
 
-* I think our excitement with all the connections apparently left a "hint of arrogance." We sincerely apologize for that.
+* I think our excitement with all the connections apparently left a "hint of
+  arrogance." We sincerely apologize for that, and will fix.
 
-* The supplementary material gives one small example of rewriting circuits using the 2-axioms.
-We do not have larger experiments at this point but we have done enough to notice that the rewriting of one circuit to another does not necessarily take an exponential number of steps: the diagram on p.10 shows an equivalence that is independent of the sizes of c1 and c2 and the sizes of A, B, C, and D.
+* The supplementary material gives one small example of rewriting circuits
+  using the 2-axioms. We do not have larger experiments at this point but we
+  have done enough to notice that the rewriting of one circuit to another does
+  not necessarily take an exponential number of steps: the diagram on p.10
+  shows an equivalence that is independent of the sizes of c1 and c2 and the
+  sizes of A, B, C, and D.
 
-* Our axioms vs.
-Lapalaza diagrams: In some sense they are indeed "the same".
-The general process is to give judicious names to each side of each diagram and then proceed to prove equivalence of the two expressions.
-In some cases this was achieved by adding a 2-axiom that directly relates the terms but in others it was possible to simplify the two sides using other 2-axioms.
-The process led us to refine the original 1-axioms of Pi which implicitly collapsed certain terms that should only been related by explicit non-trivial paths.
+* Our axioms vs. Laplaza diagrams: Yes, the level 2 combinators for 
+  distributivity properties and the interaction of combinators involving 0
+  are the same.  We apparently were not clear enough about what changed:
 
-* Proofs: All the theorems (except Thm 2) are in our Agda formalization.
-This code was not presentable as an artifact at the time of submission and required some serious refactoring which is under progress.
-Thm 2 is indeed wrong was the informal statement was included by mistake (but it is subsumed by Thm 9).
+  The general process of incorporating these diagrams as combinators is to give
+  judicious names to each natural transformation of each diagram representing
+  a natural isomorphism, and then proceed to prove equivalence of the two
+  expressions (for all 3 constructions). In some cases this was achieved by
+  adding a 2-axiom that directly relates the terms but in others it was possible
+  to simplify the two sides using other 2-axioms. The process led us to refine
+  the original 1-axioms of Pi which implicitly collapsed certain terms that
+  should only been related by explicit non-trivial paths. Thus the most 
+  apparent changes were to the programming language Pi itself, and not the
+  combinators derived from Laplaza's diagrams.  In other words, we let the
+  ``proper'' categorical semantics guide us to define the syntax of a new
+  version of Pi which had better meta-properties.
 
-* The 3rd axiom on p9 says c;(!c)=id which proves that swap;swap=id.
-(The bottom of p4 notes that every combinator has a dual according to figure 1 with swap being its own inverse.) The existence of inverses for 1-combinators is a meta-property (not a constructor of the inductive datatype of combinators but a function mapping combinators to combinators).
-The same applies to inverses for 2-combinators.
+* Proofs: All the theorems (except Thm 2) are in our Agda formalization. This
+  code was not presentable as an artifact at the time of submission and
+  required some serious refactoring which is under progress. Thm 2 is indeed
+  wrong and the informal statement was included by mistake; the core of what
+  we meant to convey is in the (correct) Thm 9.
+
+* The 3rd axiom on p9 says c;(!c)=id which proves that swap;swap=id. (The
+  bottom of p4 notes that every combinator has a dual according to figure 1
+  with swap being its own inverse.) The existence of inverses for 1-combinators
+  is a meta-property (not a constructor of the inductive datatype of
+  combinators but a function mapping combinators to combinators). The same
+  applies to inverses for 2-combinators.
 

### 2015-09-19 Univalence/response.txt
expand response

@@
 constructions into actual executable programming constructs of a programming
-language, complete with executable optimizers, and with an eye towards
+language, complete with executable optimizer, and with an eye towards
 higher-order extensions. To our knowledge, there are no programming languages
@@
 
+  So it is not the coherences which are more "small step", it is the 
+  underlying programming language.
+
 * Proofs: All the theorems (except Thm 2) are in our Agda formalization. This
@@
 
-* *** COMMENT ON PHONY MULTIPLICATION ***
+* lifting of the 0d combinators: The problem is when given a : T1 <=> T2,
+  b : T3 <=> T4, to construct something of type T1 x T3 <=> T2 x T4.  If
+  b is the identity, yes it works.  But it does not work in general.
+  There is even an (easy!) general definition for what it ought to be.
+  The point is that that definition is not functorial.  If it were, that
+  would contradiction Thomasson's result.
+
+* We would like to remark that Thomasson believed strongly enough that a 
+  particular construction (which had been published several times) was in 
+  fact wrong, that he published a paper point exactly that, and only that.
+  Most PL theorists would not understand that this result actually applies to
+  constructions they are familiar with: the Int-construction, which applies so
+  nicely to language built atop linear logic.  We believe, just as strongly 
+  as Thomasson, that pointing out that the ``obvious'' generalization to
+  the setting with two monoidal constructions (such as ours) does not work,
+  for the same reasons.  Yes, we do realize that including a negative result
+  in paper makes it harder to publish; we nevertheless feel that we can save
+  others a lot of fruitless pain by attempting to do so.
+
+* Referee 4 thinks that "The main result of the paper is that the category of
+  finite sets and bijections, with the two monoidal structures given by sum and
+  product, is the free rig groupoid." And then further that " A serious
+  weakness is that the main result has no proof, nor even a proof sketch."  This
+  would be because we did not actually claim this result as a theorem!  [We
+  believe it to be true, but as of this writing, that proof is still 
+  incomplete.]
+  Our main result is that weak rig categories are exactly the setting for the
+  semantics for a programming language of reversible circuits.  Furthermore, the
+  coherence laws of such a category exactly give complete rules for deciding
+  equivalence of such circuits.  In other words, theorems 10 and 11.  We also
+  consider theorem 9 (that *all* types, not just finite ones, equivalences and
+  equivalences-of-equivalences), form a weak rig groupoid to be quite important
+  for PL theory.

### 2015-09-20 Univalence/response.txt
added calculation about phony multiplication
tweak

@@
-A general theme of the reviews appears to request a better presentation of the
-theorems and how they all relate to each other. A compressed and intuitive
-presentation would be as follows. First we have three independent
-constructions:
+Specific responses:
+
+* I think our excitement with all the connections apparently left a
+  "hint of arrogance." We sincerely apologize for that, and will fix.
+
+* The supplementary material gives one small example of rewriting
+  circuits using the 2-axioms. We do not have larger experiments at
+  this point but we have done enough to notice that the rewriting of
+  one circuit to another does not necessarily take an exponential
+  number of steps: the diagram on p.10 shows an equivalence that is
+  independent of the sizes of c1 and c2 and the sizes of A, B, C, and
+  D.
+
+* Our axioms vs. Laplaza diagrams: Yes, the level 2 combinators for
+  distributivity properties and the interaction of combinators
+  involving 0 are the same.  We apparently were not clear enough about
+  what changed:
+
+  The general process of incorporating these diagrams as combinators
+  is to give judicious names to each natural transformation of each
+  diagram representing a natural isomorphism, and then proceed to
+  prove equivalence of the two expressions (for all 3
+  constructions). In some cases this was achieved by adding a 2-axiom
+  that directly relates the terms but in others it was possible to
+  simplify the two sides using other 2-axioms. The process led us to
+  refine the original 1-axioms of Pi which implicitly collapsed
+  certain terms that should only been related by explicit non-trivial
+  paths. Thus the most apparent changes were to the programming
+  language Pi itself, and not the combinators derived from Laplaza's
+  diagrams.  In other words, we let the ``proper'' categorical
+  semantics guide us to define the syntax of a new version of Pi which
+  had better meta-properties.
+
+  So it is not the coherences which are more "small step", it is the
+  underlying programming language.
+
+* Proofs: All the theorems (except Thm 2 and Thm 11) are in our Agda
+  formalization. This code was not presentable as an artifact at the
+  time of submission and required some serious refactoring which is
+  under progress. Thm 2 is indeed wrong and the informal statement was
+  included by mistake; the core of what we meant to convey is in the
+  (correct) Thm 9. For Thm 11, the first part was proved in previous
+  work; the second part follows from the logical interpretation of
+  coherence in the categorical sense as completeness of a syntactic
+  axiomatization (Pi axioms) with respect to a model (rig category of
+  type equivalences) (see e.g.,
+  http://www.mi.sanu.ac.rs/~kosta/coh.pdf)
+
+* The 3rd axiom on p9 says c;(!c)=id which proves that
+  swap;swap=id. (The bottom of p4 notes that every combinator has a
+  dual according to figure 1 with swap being its own inverse.) The
+  existence of inverses for 1-combinators is a meta-property (not a
+  constructor of the inductive datatype of combinators but a function
+  mapping combinators to combinators). The same applies to inverses
+  for 2-combinators.
+
+* lifting of the 0d combinators: The problem is when given a : T1 <=>
+  T2, b : T3 <=> T4, to construct something of type T1 x T3 <=> T2 x
+  T4.  If b is the identity, yes it works.  But it does not work in
+  general.  There is even an (easy!) general definition for what it
+  ought to be.  The point is that that definition is not functorial.
+  If it were, that would contradiction Thomasson's result. Regarding
+  the construction given at the end of the report, that part is right,
+  it is the 'next' step which breaks down.  In more detail
+
+  T1 = a - b
+  T2 = a' - b'
+  T3 = c - d
+  T4 = c' - d'
+
+  a : T1 <=> T2
+     : a + b' <-> a' + b
+
+  a' : T1 x T3 <=> T2 x T3
+     : (a-b) x (c-d) <=> (a'-b') x (c-d)
+     : (ac+bd - bc+ad) <=> (a'c+b'd - b'c+a'd)
+     : ac+bd+b'c+a'd <-> a'c+b'd+bc+ad
+
+  (same as in the given proof, just flipped)
+
+  b : T3 <=> T4
+  b : c + d' <-> c' + d
+
+  b' : T2 x T3 <=> T2 x T4
+      : (a'-b') x (c-d) <=> (a'-b') x (c'-d')
+      : (a'c+b'd - b'c+a'd) <=> (a'c'+b'd' - b'c'+a'd')
+      : a'c'+b'd+b'c'+a'd' <-> a'c'+b'd'+b'c+a'd
+
+  Now a' and b' do not compose.
+
+* We would like to remark that Thomasson believed strongly enough that
+  a particular construction (which had been published several times)
+  was in fact wrong, that he published a paper point exactly that, and
+  only that.  Most PL theorists would not understand that this result
+  actually applies to constructions they are familiar with: the
+  Int-construction, which applies so nicely to language built atop
+  linear logic.  We believe, just as strongly as Thomasson, that
+  pointing out that the ``obvious'' generalization to the setting with
+  two monoidal constructions (such as ours) does not work, for the
+  same reasons.  Yes, we do realize that including a negative result
+  in paper makes it harder to publish; we nevertheless feel that we
+  can save others a lot of fruitless pain by attempting to do so.
+
+* Referee 4 states that "The main result of the paper is that the
+  category of finite sets and bijections, with the two monoidal
+  structures given by sum and product, is the free rig groupoid." And
+  then further that " A serious weakness is that the main result has
+  no proof, nor even a proof sketch."  This would be because we did
+  not actually claim this result as a theorem!  [We believe it to be
+  true, but as of this writing, that proof is still incomplete.]  Our
+  main result is that weak rig categories are exactly the setting for
+  the semantics for a programming language of reversible circuits.
+  Furthermore, the coherence laws of such a category exactly give
+  complete rules for deciding equivalence of such circuits.  In other
+  words, theorems 10 and 11.  We also consider theorem 9 (that *all*
+  types, not just finite ones, equivalences and
+  equivalences-of-equivalences), form a weak rig groupoid to be quite
+  important for PL theory.
+
+-- Optional general comment
+
+A general theme of the reviews appears to request a better
+presentation of the theorems and how they all relate to each other. A
+compressed and intuitive presentation would be as follows. First we
+have three independent constructions:
 
@@
 
-Each construction relates a particular representation of types via appropriate
-relations; each construction has a rig structure (additive and multiplicative
-units, commutative and associative sums and products, and distributivity of
-products over sums). We then have a second level:
+Each construction relates a particular representation of types via
+appropriate relations; each construction has a rig structure (additive
+and multiplicative units, commutative and associative sums and
+products, and distributivity of products over sums). We then have a
+second level:
 
@@
 
-The two levels of constructions are connected, and form categories. For
-example, we have a category whose objects are Pi types, whose morphisms are
-Pi-combinators, and whose equality of morphisms is Pi-optimizations; this
-category is a (free) symmetric rig groupoid. We finally have theorems that
-connect some of the constructions vertically allowing us to move across
-representations.
+The two levels of constructions are connected, and form
+categories. For example, we have a category whose objects are Pi
+types, whose morphisms are Pi-combinators, and whose equality of
+morphisms is Pi-optimizations; this category is a (free) symmetric rig
+groupoid. We finally have theorems that connect some of the
+constructions vertically allowing us to move across representations.
 
-These are the "raw" results without interpretation, context, or applications.
-Perhaps unfortunately, the current presentation of the paper mixes these "raw"
-results with theoretical connections to HoTT, with Agda formalization concerns,
-and with practical connections to reversible circuits and their optimizations.
-We acknowledge that these connections can be perceived as distracting, and that
-their presentation could be much improved (which we are working on), BUT these
-connections are all critical and essential -- and are why this is being sent to
-POPL rather than (say) Theory and Applications of Category Theory.
+These are the "raw" results without interpretation, context, or
+applications.  Perhaps unfortunately, the current presentation of the
+paper mixes these "raw" results with theoretical connections to HoTT,
+with Agda formalization concerns, and with practical connections to
+reversible circuits and their optimizations.  We acknowledge that
+these connections can be perceived as distracting, and that their
+presentation could be much improved (which we are working on), BUT
+these connections are all critical and essential -- and are why this
+is being sent to POPL rather than (say) Theory and Applications of
+Category Theory.
 
-Indeed some of the connections above are trivial in "conventional mathematics";
-however some of those same connections hide implicit isomorphisms (changes of
-representations) that are important computationally.  Our construction shows,
-for example, that transporting properties across equivalences of finite types
-can be done via executing permutations, something which has a clear
-computational cost and which itself depends on the choice of representations of
-the permutations. Since this is POPL, the presentation emphasizes turning these
-constructions into actual executable programming constructs of a programming
-language, complete with executable optimizer, and with an eye towards
-higher-order extensions. To our knowledge, there are no programming languages
-for representing --and reasoning about-- a rich notion of "permutations"
-(including sums and products).  Furthermore, the higher-order extensions are
-crucial for abstractions over such permutations by allowing the permutations
-themselves to be treated as data that can be constructed and examined
-dynamically.
-
-Specific (and short with apologies) responses:
-
-* I think our excitement with all the connections apparently left a "hint of
-  arrogance." We sincerely apologize for that, and will fix.
-
-* The supplementary material gives one small example of rewriting circuits
-  using the 2-axioms. We do not have larger experiments at this point but we
-  have done enough to notice that the rewriting of one circuit to another does
-  not necessarily take an exponential number of steps: the diagram on p.10
-  shows an equivalence that is independent of the sizes of c1 and c2 and the
-  sizes of A, B, C, and D.
-
-* Our axioms vs. Laplaza diagrams: Yes, the level 2 combinators for 
-  distributivity properties and the interaction of combinators involving 0
-  are the same.  We apparently were not clear enough about what changed:
-
-  The general process of incorporating these diagrams as combinators is to give
-  judicious names to each natural transformation of each diagram representing
-  a natural isomorphism, and then proceed to prove equivalence of the two
-  expressions (for all 3 constructions). In some cases this was achieved by
-  adding a 2-axiom that directly relates the terms but in others it was possible
-  to simplify the two sides using other 2-axioms. The process led us to refine
-  the original 1-axioms of Pi which implicitly collapsed certain terms that
-  should only been related by explicit non-trivial paths. Thus the most 
-  apparent changes were to the programming language Pi itself, and not the
-  combinators derived from Laplaza's diagrams.  In other words, we let the
-  ``proper'' categorical semantics guide us to define the syntax of a new
-  version of Pi which had better meta-properties.
-
-  So it is not the coherences which are more "small step", it is the 
-  underlying programming language.
-
-* Proofs: All the theorems (except Thm 2) are in our Agda formalization. This
-  code was not presentable as an artifact at the time of submission and
-  required some serious refactoring which is under progress. Thm 2 is indeed
-  wrong and the informal statement was included by mistake; the core of what
-  we meant to convey is in the (correct) Thm 9.
-
-* The 3rd axiom on p9 says c;(!c)=id which proves that swap;swap=id. (The
-  bottom of p4 notes that every combinator has a dual according to figure 1
-  with swap being its own inverse.) The existence of inverses for 1-combinators
-  is a meta-property (not a constructor of the inductive datatype of
-  combinators but a function mapping combinators to combinators). The same
-  applies to inverses for 2-combinators.
-
-* Completeness in Thm 11. *** BRIEFLY EXPLAIN HOW WE GET COMPLETENESS FROM LAPLAZA ***
-
-* lifting of the 0d combinators: The problem is when given a : T1 <=> T2,
-  b : T3 <=> T4, to construct something of type T1 x T3 <=> T2 x T4.  If
-  b is the identity, yes it works.  But it does not work in general.
-  There is even an (easy!) general definition for what it ought to be.
-  The point is that that definition is not functorial.  If it were, that
-  would contradiction Thomasson's result.
-
-* We would like to remark that Thomasson believed strongly enough that a 
-  particular construction (which had been published several times) was in 
-  fact wrong, that he published a paper point exactly that, and only that.
-  Most PL theorists would not understand that this result actually applies to
-  constructions they are familiar with: the Int-construction, which applies so
-  nicely to language built atop linear logic.  We believe, just as strongly 
-  as Thomasson, that pointing out that the ``obvious'' generalization to
-  the setting with two monoidal constructions (such as ours) does not work,
-  for the same reasons.  Yes, we do realize that including a negative result
-  in paper makes it harder to publish; we nevertheless feel that we can save
-  others a lot of fruitless pain by attempting to do so.
-
-* Referee 4 thinks that "The main result of the paper is that the category of
-  finite sets and bijections, with the two monoidal structures given by sum and
-  product, is the free rig groupoid." And then further that " A serious
-  weakness is that the main result has no proof, nor even a proof sketch."  This
-  would be because we did not actually claim this result as a theorem!  [We
-  believe it to be true, but as of this writing, that proof is still 
-  incomplete.]
-  Our main result is that weak rig categories are exactly the setting for the
-  semantics for a programming language of reversible circuits.  Furthermore, the
-  coherence laws of such a category exactly give complete rules for deciding
-  equivalence of such circuits.  In other words, theorems 10 and 11.  We also
-  consider theorem 9 (that *all* types, not just finite ones, equivalences and
-  equivalences-of-equivalences), form a weak rig groupoid to be quite important
-  for PL theory.
+Indeed some of the connections above are trivial in "conventional
+mathematics"; however some of those same connections hide implicit
+isomorphisms (changes of representations) that are important
+computationally.  Our construction shows, for example, that
+transporting properties across equivalences of finite types can be
+done via executing permutations, something which has a clear
+computational cost and which itself depends on the choice of
+representations of the permutations. Since this is POPL, the
+presentation emphasizes turning these constructions into actual
+executable programming constructs of a programming language, complete
+with executable optimizer, and with an eye towards higher-order
+extensions. To our knowledge, there are no programming languages for
+representing --and reasoning about-- a rich notion of "permutations"
+(including sums and products).  Furthermore, the higher-order
+extensions are crucial for abstractions over such permutations by
+allowing the permutations themselves to be treated as data that can be
+constructed and examined dynamically.

### 2015-09-28 Univalence/permutations.tex

\documentclass{llncs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\title{Permutations}
\titlerunning{Permutations}
\author{Jacque Carette \and Amr Sabry}
\authorrunning{Carette and Sabry}
\institute{? \\
\email{...@...}
\and
...}
\maketitle

\begin{abstract}

\keywords{}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Common Representations}

Say we want to write the permutation between \verb|Fin 3 x Fin 2| and
\verb|Fin 2 x Fin 3|. First let's enumerate the two sets in some
canonical order:
\begin{verbatim}
Fin 3 x Fin 2 = { (0,0), (0,1), (1,0), (1,1), (2,0), (2,1) } 
Fin 2 x Fin 3 = { (0,0), (0,1), (0,2), (1,0), (1,1), (1,2) }
\end{verbatim}

Then the permutation can be visualized as:

\begin{verbatim} 
Fin 3 x Fin 2 = { (0,0), (0,1), (1,0), (1,1), (2,0), (2,1) }
                    |      |      |      |      |      |
                    |      |      |      +---------+   |
                    |      |      |             |  |   |
                    |      +-------------+      |  |   |
                    |             |      |      |  |   |
                    |      +------+      |      |  |   |
                    |      |             |      |  |   |
                    |      |      +-------------+  |   |
                    |      |      |      |         |   |
                    |      |      |      |      +--+   |
                    |      |      |      |      |      |
Fin 2 x Fin 3 = { (0,0), (0,1), (0,2), (1,0), (1,1), (1,2) }
\end{verbatim}

How to represent this in a way that amenable to efficient and
convenient manipulation. There are several common representations:

\begin{itemize}
\item Truth table
\item Matrix
\item Reed-Muller expansion
\item Cycle expansion
\item Decision diagram
\item Vectors
\item Factorial representation
\item Terms of a programming language
\end{itemize}

As vectors we would get:
\begin{verbatim}
p = [ 0, 3, 1, 4, 2, 5 ]
\end{verbatim}
which says that the item at position \verb|i| goes to the item at
position \verb|p(i)|.

As a sequence of transpositions, we would get several possible
representations... etc.

Very difficult to work with such representations. Imagine trying to
write big permutations for a reversible adder. Imagine trying to prove
properties of such permutations (bimonoidal structure). Etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Essence}

Our point is that the essence of the permutation above is a type
equivalence that is trivial to prove. Indeed the equivalence between
\verb|A x B| and \verb|B x A| is trivial to prove. We can once and for
all derive |Fin m x Fin n ~= Fin (m * n)| from such a type
equivalence. We can derive the two monoidal structures and their
interactions. Our main theorem is that:
\begin{verbatim}
Fin, ~=, ~~~ is a weak rig groupoid
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{acm}
\bibliography{cites}
\end{document}

### 2017-09-18 survey/outline.tex
outline

\documentclass{article}
\usepackage{fullpage}
\title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
\author{Jacques Carette, Roshan James, Amr Sabry}
\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Models of Computation}

Computation in the ``real world'' vs. idealized computation in abstract models

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Physics, Computation, and Logic}

Conflict between Church-Turing thesis, textbook quantum mechanics, and RSA.

Conflict between quantum mechanics and logic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conservation of Information}

Pi languages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Application: Information Flow Security}

??

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2017-09-18 Neg.hs

@@ -1,6 +1,6 @@
 {-# LANGUAGE GADTs, TypeOperators, DataKinds, RankNTypes #-}
 {-# LANGUAGE TypeFamilies, MultiParamTypeClasses, FlexibleInstances #-}
-{-# LANGUAGE ScopedTypeVariables #-}
+{-# LANGUAGE ScopedTypeVariables, AllowAmbiguousTypes #-}
 {-# OPTIONS -Wall #-}
 
 {--
@@ -521,9 +521,48 @@ plusG (GM f) (GM g) = GM h
        assoc4 
     -- Either (Either am cm) (Either bp dp)
 
+{--
+
+T1 = a - b
+T2 = a' - b'
+T3 = c - d
+T4 = c' - d'
+
+a : T1 <=> T2
+a : a + b' <-> a' + b
+
+a' : T1 x T3 <=> T2 x T3
+a' : (a-b) x (c-d) <=> (a'-b') x (c-d)
+a' : (ac+bd - bc+ad) <=> (a'c+b'd - b'c+a'd)
+a' : (ac+bd+b'c+a'd <-> a'c+b'd+bc+ad)
+
+b : T3 <=> T4
+b : c + d' <-> c' + d
+
+b' : T2 x T3 <=> T2 x T4
+b' : (a'-b') x (c-d) <=> (a'-b') x (c'-d')
+b' : (a'c+b'd - b'c+a'd) <=> (a'c'+b'd' - b'c'+a'd')
+b' : (a'c'+b'd+b'c'+a'd' <-> a'c'+b'd'+b'c+a'd)
+
+
+
+--}
+
+a' :: (t1 ~ (a :- b), t2 ~ (a' :- b'), t3 ~ (c :- d)) =>
+      GM t1 t2 -> GM (t1,t3) (t2,t3)
+a' (GM f) = error "todo" -- GM (helper f)
+--  where helper :: R (Either (Pos a) (Neg b')) (Either (Neg a') (Pos b)) ->
+--                  R (Either (Pos () (Neg b)) (Either (Neg a) (Pos b)) 
+--  type TimesG (ap :- am) (bp :- bm) = 
+--    (Either (ap,bp) (am,bm)) :- (Either (am,bp) (ap,bm))
+--        helper = error "todo"
+
+
 timesG :: (a ~ (ap :- am), b ~ (bp :- bm), c ~ (cp :- cm), d ~ (dp :- dm)) =>
           GM a b -> GM c d -> GM (TimesG a c) (TimesG b d)
[…]

### 2017-10-16 survey/outline.tex
outline

@@
-\documentclass{article}
+\documentclass[12pt]{article}
 \usepackage{fullpage}
@@
 
+Abstract models of computation: Turing machine, RAM, Von Neumann
+architecture, $\lambda$-calculus, logic, discrete mathematics
+
+Cyber-physical systems increasingly common. Are computing foundations
+adequate?
+
+Toffoli 1980: Mathematical models of computation are abstract
+constructions, by their nature unfettered by physical laws. However,
+if these models are to give indications that are relevant to concrete
+computing, they must somehow capture, albeit in a selective and
+stylized way, certain general physical restrictions to which all
+concrete computing processes are subjected.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Computation in the ``real world'' vs. idealized computation in abstract models
+Computation in the ``real world'' vs. idealized computation in
+abstract models. 
+\begin{itemize}
+\item Mathematical abstraction of sorting is well-understood; sorting
+  petabytes of data violates many of the assumptions: hardware
+  component failure frequent, global synchronization impossible;
+  consistency must be eventual; etc. 
+\item Mathematical abstraction of sortedness property: easy; in the
+  real world with petabytes of evolving data; it is not clear that it
+  is even possible to check sortedness; easy to get inconsistent
+  results
+\item Noise does not exist in our mathematical abstractions of
+  computation; essential use in differential privacy
+\item Mathematical models of secure computing typically ignore power
+  consumption, electromagnetic signatures; actual attacks based on DPA
+  and DEMA however
+\end{itemize}
 
@@
 
-Conflict between Church-Turing thesis, textbook quantum mechanics, and RSA.
+Conflict between Church-Turing thesis, textbook quantum mechanics, and
+RSA: Either Shor’s algorithm is not ``natural''. (Textbook quantum
+mechanics is wrong); or, Shor’s algorithm is ``natural'' and there is
+no classical counterpart. (There are ``natural'' computing models that
+are exponentially faster than the Turing Machine); or, there is an
+efficient classical factoring algorithm. (Possible but surprising); At
+least one of these wild claims is true!!!
 
-Conflict between quantum mechanics and logic.
+Conflict between quantum mechanics and logic. Possibilities:
+
+\paragraph*{Revise quantum mechanics}.
+
+The mathematician's vision of an unlimited sequence of totally
+reliable operations is unlikely to be implementable in this real
+universe.  But the real world is unlikely to supply us with unlimited
+memory or unlimited Turing machine tapes. Therefore, continuum
+mathematics is not executable, and physical laws which invoke that can
+not really be satisfactory. They are references to illusionary
+procedures.  (Landauer 1996 and 1999)
+
+I want to talk about the possibility that there is to be an exact
+simulation, that the computer will do exactly the same as nature. If
+this is to be proved and the type of computer is as I've already
+explained, then it's going to be necessary that everything that
+happens in a finite volume of space and time would have to be exactly
+analyzable with a finite number of logical operations. The present
+theory of physics is not that way, apparently. It allows space to go
+down into infinitesimal distances, wavelengths to get infinitely
+great, terms to be summed in infinite order, and so forth; and
+therefore, if this proposition is right, physical law is wrong.
+(Feynman 1981)
+
+\paragraph*{Revise Computer Science.}
+
+Another thing that had been suggested early was that natural laws are
+reversible, but that computer rules are not. But this turned out to be
+false; the computer rules can be reversible, and it has been a very,
+very useful thing to notice and to discover that. This is a place
+where the relationship of physics and computation has turned itself
+the other way and told us something about the possibilities of
+computation. So this is an interesting subject because it tells us
+something about computer rules…  (Feynman 1981)
+
+Turing hoped that his abstracted-paper-tape model was so simple, so
+transparent and well defined, that it would not depend on any
+assumptions about physics that could conceivably be falsified, and
+therefore that it could become the basis of an abstract theory of
+computation that was independent of the underlying physics. ‘He
+thought,’ as Feynman once put it, ‘that he understood paper.’ But he
+was mistaken. Real, quantum-mechanical paper is wildly different from
+the abstract stuff that the Turing machine uses. The Turing machine is
+entirely classical, and does not allow for the possibility the paper
+might have different symbols written on it in different universes, and
+that those might interfere with one another.  (Deutsch 1985)
+
+Ed Fredkin pursued the idea that information must be finite in
+density. One day, he announced that things must be even more simple
+than that. He said that he was going to assume that information itself
+is conserved. ``You’re out of you mind, Ed.'' I pronounced. ``That’s
+completely ridiculous. Nothing could happen in such a world. There
+couldn’t even be logical gates. No decisions could ever be made.'' But
+when Fredkin gets one of his ideas, he’s quite immune to objections
+like that; indeed, they fuel him with energy. Soon he went on to
+assume that information processing must also be reversible — and
+invented what’s now called the Fredkin gate. (Minsky 1999)
+
+Even revise the laws of thought themselves: In other terms, what is so
+good in logic that quantum physics should obey? Can't we imagine that
+our conceptions about logic are wrong, so wrong that they are unable
+to cope with the quantum miracle? […] Instead of teaching logic to
+nature, it is more reasonable to learn from her. Instead of
+interpreting quantum into logic, we shall interpret logic into
+quantum. (Girard 2007)
 
@@
 
-Pi languages
+A physical principle of computation
+
+No creation of information
+No duplication of information
+
+All programs, proofs, deductions are equivalences, isomorphisms,
+reversible
+
+Several pages on reversible logic; reversible family of level 1 and
+level 2 languages (Pi);
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Application: Information Flow Security}
+\section{Applications}
 
-??
+Perhaps Information Flow Security
 

### 2017-10-16 survey/outline.tex
lots and lots of comments.

@@
 \author{Jacques Carette, Roshan James, Amr Sabry}
+
+\newcommand{\amr}[1]{\fbox{Amr says:} \textbf{#1}}
+\newcommand{\jc}[1]{\fbox{Jacques says:} \textbf{#1}}
+\newcommand{\roshan}[1]{\fbox{Roshan says:} \textbf{#1}}
+
 \begin{document}
@@
 
+\jc{why are logic and DM listed alongside MoC?}
+
 Cyber-physical systems increasingly common. Are computing foundations
 adequate? 
+\jc{your questioning of MoC below is not about CPS! Should either
+foreshadow the next section 'properly' here, or add a leading 
+example based on CPS to MoC section}
 
@@
 
+\jc{While all the above are true (and not exhaustive), what I also read
+is that the first two are actually ``the same'' point mostly related
+to scale and distributivity, the 3rd is about inequate models (but not 
+related to physics), while the 4th is about the physics of devices.
+They are not about CPS. And not about Quantum either. So how are these
+motivating a deeper look at how quantum affects our MoC? It feels to me
+that we ought to have a quantum ``example'' here already.}
+
+\jc{What this seems to motivate is to look at our models with respect
+to what we're trying to achieve.  Models are, by definition, abstractions
+that ``abstract away'' irrelevant details. So the point would be that
+some details that we thought were irrelevant really are not.
+So I could see this section motivating a whole programme of
+targetted de-abstraction. This could lead into the focus of this paper
+following one strand, namely one aspect of quantum physics.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\jc{No access to infinite memory/tape just means that ``infinitary''
+arguments do not go through. The MoC with reversible computation 
+assumes totally reliable operations (but not necessarily infinitely
+many). You leap to ``continuum'' mathematics without justification.
+While QC does use the continuum in non-trivial ways, and physical
+reality is ``discrete'' at the atomic level (but who knows at the
+quantum level?!?), I don't think that's where you want to go here.
+But see below.}
+
 I want to talk about the possibility that there is to be an exact
@@
 (Feynman 1981)
+
+\jc{In a sense, physics already does this: statistical mechanics.
+Even though the systems under study are explicitly discrete, albeit
+will huge numbers of particules, the behaviour is approximated (!!)
+via a continuous model. Same in economics with Black-Scholes.
+The real problem are those conclusions drawn from such models which
+*require* arbitrarily small values to exist, and even worse, for
+the underlying values (reals) to be connected, i.e. form an actual
+continuum. What Feynman is asking here is 'finiteness of information',
+which is what things like Type-2 computability also ask for.}
 
@@
 
+\jc{One must be careful with this argument: we have a ``current 
+understanding'' of quantum physics, which could be quite flawed.
+So what we want to do is to learn from thsoe parts of quantum physics
+which seem to be extremely stable (like conservation of information!)
+first. As researchers, we can certainly speculate on it all, but it
+makes sense to vigorously pursue the conclusions one can draw from
+those aspects of (quantum) physics that seem stable.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\jc{Should we also have a quick survey of other works in reversible
+computation? Most of the other work I know tries to start from 
+well known models of computation or well known programming languages,
+and then adapts them to be reversible. Pi is different in that it 
+starts from the semantic implications of reversibility. Then, because
+it adds types, rather than control-flow (or even data-flow) as its 
+next layer of ``understanding'', this leads to equivalences, isomorphisms,
+etc.  This then leads, quite naturally, to finding that the ``proof
+language'' of semirings (and Rig Groupoids at level 2) is actually a 
+programming language. And it is Pi. This is a neat twist on Curry-Howard
+because CH is about \textbf{inhabitation} only. But with ``conservation
+of information'' as the basis, a different kind of correspondance
+arises; in fact, this one may well be an actual isomorphism.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Perhaps Information Flow Security
+
+\jc{This is where I am still least sure. I am wary of either doing
+new research with such a tight timeline, or of being too speculative.
+If there are papers on this already, perhaps references to them could
+be put here?}
 

### 2017-10-17 survey/outline.tex
roshan's intro

@@
 \title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
-\author{Jacques Carette, Roshan James, Amr Sabry}
+\author{Jacques Carette, Roshan P. James, Amr Sabry}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+\section{Introduction : Reversibility, the Missing Principle}
 
-Abstract models of computation: Turing machine, RAM, Von Neumann
-architecture, $\lambda$-calculus, logic, discrete mathematics
+To fully appreciate the missing principle of conventional computing,
+one must go back to an old thought experiment by J. C. Maxwell,
+codified in a letter that Maxwell wrote to P. G Tait in 1867 -- this
+is the letter whose ideas are now well known as the `Maxwell's
+Demon'. Maxwell's Demon was a thought experiment that seemed to
+indicate that intelligent beings can somehow violate the second law of
+thermodynamics, thereby violating physics itself.
 
-\jc{why are logic and DM listed alongside MoC?}
+Many resolutions were offered for this conundrum (for a compilation,
+see Maxwell's Demon by Leff and Rex), but none withstood careful
+scrutiny until the establishment of the so called `Landauer's
+Principle' in 1961 -- a principle whose experimental validation
+happened recently in 2012 by Eric Lutz and others.
 
-Cyber-physical systems increasingly common. Are computing foundations
-adequate? 
-\jc{your questioning of MoC below is not about CPS! Should either
-foreshadow the next section 'properly' here, or add a leading 
-example based on CPS to MoC section}
+Maxwell's demon appears to violate the second law of thermodynamics by
+having a tiny `intelligence' observing the movement of individual
+particles of a gas and separating fast moving particles from slow
+moving ones, thereby reducing the total entropy of the
+system. Landauer's resolution of the demon relied on two ideas that
+took root only a few decades earlier -- the formal notion of
+computation (through the work of Turing and Church, 1936) and the
+formal notion of information (through the work of Claude Shannon,
+1948). Landauer reasoned that the computation done by the finite brain
+of the demon, involves getting information about the movement of
+molecules, acting on that information and then overwriting it to make
+room for the next computation.  Landauer reasoned, and this is the
+important bit, that the computation that is manipulating information
+in the demon's brain \textit{must be thermodynamic work}, thereby
+bringing the demon back into the fold of physics.
 
+In its essence this is a strange and wonderful idea. Information,
+physics and computation are inextricably linked and that ideas in each
+field have implications for the other. \roshan{maybe cite Bennet and
+  the various `thermodynamics of computation' papers here.}
+
+When the early models computation, the Turing machines and
+$\lambda$-calculus, were developed there was no real reason to the
+take the information content of computations into serious
+consideration -- in fact, at that time there was no quantifiable
+notion of information. These models followed in the footsteps of logic
+where conventional logic gates such as AND and OR happily erase their
+inputs.
+
+\begin{quote}
 Toffoli 1980: Mathematical models of computation are abstract
@@
 concrete computing processes are subjected.
+\end{quote}
+
+Now however, through the establishment of the Landauer's principle and
+through recent investigations into computational models that are well
+suited to quantum mechanics in has become increasingly evident that
+our computation models need to be revisited. Information not just
+seems to have a physical manifestation, but also that it like other
+physical entities is conserved.
+
+Going back to the drawing board, can we have computation that closely
+matches our underlying physics?  i.e. can we have computation that in
+its essence preserves information? What is the missing principle that
+we can use as a basis of a new model of computation to achieve this?
+It turns out that we need to look at physics again for inspiration --
+in physics all of the primitive rules, nature itself at its most
+fundamental level, is reversible. Conventional computation is not. Can
+we embrace this simple principle as the building block of a model of
+computation? What would computation looks like when viewed from this
+vantage point? And, does it have applications?
+
+The objective of this survey is to touch upon the many related areas
+of research in this space.
+
+%% Abstract models of computation: Turing machine, RAM, Von Neumann
+%% architecture, $\lambda$-calculus, logic, discrete mathematics
+
+%% \jc{why are logic and DM listed alongside MoC?}
+
+%% Cyber-physical systems increasingly common. Are computing foundations
+%% adequate?
+%% \jc{your questioning of MoC below is not about CPS! Should either
+%% foreshadow the next section 'properly' here, or add a leading
+%% example based on CPS to MoC section}
 
@@
 Conflict between Church-Turing thesis, textbook quantum mechanics, and
-RSA: Either Shor’s algorithm is not ``natural''. (Textbook quantum
-mechanics is wrong); or, Shor’s algorithm is ``natural'' and there is
+RSA: Either Shorâs algorithm is not ``natural''. (Textbook quantum
+mechanics is wrong); or, Shorâs algorithm is ``natural'' and there is
 no classical counterpart. (There are ``natural'' computing models that
@@
 computation. So this is an interesting subject because it tells us
-something about computer rules…  (Feynman 1981)
+something about computer rulesâŠ  (Feynman 1981)
 
@@
 therefore that it could become the basis of an abstract theory of
-computation that was independent of the underlying physics. ‘He
-thought,’ as Feynman once put it, ‘that he understood paper.’ But he
+computation that was independent of the underlying physics. âHe
+thought,â as Feynman once put it, âthat he understood paper.â But he
 was mistaken. Real, quantum-mechanical paper is wildly different from
@@
 than that. He said that he was going to assume that information itself
-is conserved. ``You’re out of you mind, Ed.'' I pronounced. ``That’s
+is conserved. ``Youâre out of you mind, Ed.'' I pronounced. ``Thatâs
 completely ridiculous. Nothing could happen in such a world. There
-couldn’t even be logical gates. No decisions could ever be made.'' But
-when Fredkin gets one of his ideas, he’s quite immune to objections
+couldnât even be logical gates. No decisions could ever be made.'' But
+when Fredkin gets one of his ideas, heâs quite immune to objections
 like that; indeed, they fuel him with energy. Soon he went on to
-assume that information processing must also be reversible — and
-invented what’s now called the Fredkin gate. (Minsky 1999)
+assume that information processing must also be reversible â and
+invented whatâs now called the Fredkin gate. (Minsky 1999)
 
@@
 our conceptions about logic are wrong, so wrong that they are unable
-to cope with the quantum miracle? […] Instead of teaching logic to
+to cope with the quantum miracle? [âŠ] Instead of teaching logic to
 nature, it is more reasonable to learn from her. Instead of

### 2017-10-17 survey/outline.tex
updated

@@
-\documentclass[12pt]{article}
+\documentclass{article}
 \usepackage{fullpage}
@@
 physics and computation are inextricably linked and that ideas in each
-field have implications for the other. \roshan{maybe cite Bennet and
-  the various `thermodynamics of computation' papers here.}
+field have implications for the other. (Cite Bennett and the various
+`thermodynamics of computation' papers here.)
 
-When the early models computation, the Turing machines and
+When the early models computation, the Turing machine, and the
 $\lambda$-calculus, were developed there was no real reason to the
@@
 through recent investigations into computational models that are well
-suited to quantum mechanics in has become increasingly evident that
-our computation models need to be revisited. Information not just
-seems to have a physical manifestation, but also that it like other
-physical entities is conserved.
+suited to quantum mechanics it has become increasingly evident that
+our computation models need to be revisited. Information, not just
+seems to have a physical manifestation, but also that it, like other
+physical entities, is conserved.
 
@@
 we embrace this simple principle as the building block of a model of
-computation? What would computation looks like when viewed from this
+computation? What would computation look like when viewed from this
 vantage point? And, does it have applications?
@@
 
+Explain classical models of computation and their limitations in
+real-world ``physical'' applications.
+
 Computation in the ``real world'' vs. idealized computation in
@@
 
-\jc{While all the above are true (and not exhaustive), what I also read
-is that the first two are actually ``the same'' point mostly related
-to scale and distributivity, the 3rd is about inequate models (but not
-related to physics), while the 4th is about the physics of devices.
-They are not about CPS. And not about Quantum either. So how are these
-motivating a deeper look at how quantum affects our MoC? It feels to me
-that we ought to have a quantum ``example'' here already.}
+% \jc{While all the above are true (and not exhaustive), what I also read
+% is that the first two are actually ``the same'' point mostly related
+% to scale and distributivity, the 3rd is about inadequate models (but not
+% related to physics), while the 4th is about the physics of devices.
+% They are not about CPS. And not about Quantum either. So how are these
+% motivating a deeper look at how quantum affects our MoC? It feels to me
+% that we ought to have a quantum ``example'' here already.}
 
-\jc{What this seems to motivate is to look at our models with respect
-to what we're trying to achieve.  Models are, by definition, abstractions
-that ``abstract away'' irrelevant details. So the point would be that
-some details that we thought were irrelevant really are not.
-So I could see this section motivating a whole programme of
-targetted de-abstraction. This could lead into the focus of this paper
-following one strand, namely one aspect of quantum physics.}
+% \jc{What this seems to motivate is to look at our models with respect
+% to what we're trying to achieve.  Models are, by definition, abstractions
+% that ``abstract away'' irrelevant details. So the point would be that
+% some details that we thought were irrelevant really are not.
+% So I could see this section motivating a whole programme of
+% targeted de-abstraction. This could lead into the focus of this paper
+% following one strand, namely one aspect of quantum physics.}
 
@@
 Conflict between Church-Turing thesis, textbook quantum mechanics, and
-RSA: Either Shorâs algorithm is not ``natural''. (Textbook quantum
-mechanics is wrong); or, Shorâs algorithm is ``natural'' and there is
+RSA: Either Shor's algorithm is not ``natural''. (Textbook quantum
+mechanics is wrong); or, Shor's algorithm is ``natural'' and there is
 no classical counterpart. (There are ``natural'' computing models that
@@
 
-Conflict between quantum mechanics and logic. Possibilities:
+\noindent Also conflict between quantum mechanics and logic. Possibilities:
 
-\paragraph*{Revise quantum mechanics}.
+\paragraph*{Revise quantum mechanics.}
 
@@
 
-\jc{No access to infinite memory/tape just means that ``infinitary''
-arguments do not go through. The MoC with reversible computation
-assumes totally reliable operations (but not necessarily infinitely
-many). You leap to ``continuum'' mathematics without justification.
-While QC does use the continuum in non-trivial ways, and physical
-reality is ``discrete'' at the atomic level (but who knows at the
-quantum level?!?), I don't think that's where you want to go here.
-But see below.}
+% \jc{No access to infinite memory/tape just means that ``infinitary''
+% arguments do not go through. The MoC with reversible computation
+% assumes totally reliable operations (but not necessarily infinitely
+% many). You leap to ``continuum'' mathematics without justification.
+% While QC does use the continuum in non-trivial ways, and physical
+% reality is ``discrete'' at the atomic level (but who knows at the
+% quantum level?!?), I don't think that's where you want to go here.
+% But see below.}
 
@@
 
-\jc{In a sense, physics already does this: statistical mechanics.
-Even though the systems under study are explicitly discrete, albeit
-will huge numbers of particules, the behaviour is approximated (!!)
-via a continuous model. Same in economics with Black-Scholes.
-The real problem are those conclusions drawn from such models which
-*require* arbitrarily small values to exist, and even worse, for
-the underlying values (reals) to be connected, i.e. form an actual
-continuum. What Feynman is asking here is 'finiteness of information',
-which is what things like Type-2 computability also ask for.}
+% \jc{In a sense, physics already does this: statistical mechanics.
+% Even though the systems under study are explicitly discrete, albeit
+% will huge numbers of particules, the behaviour is approximated (!!)
+% via a continuous model. Same in economics with Black-Scholes.
+% The real problem are those conclusions drawn from such models which
+% *require* arbitrarily small values to exist, and even worse, for
+% the underlying values (reals) to be connected, i.e. form an actual
+% continuum. What Feynman is asking here is 'finiteness of information',
+% which is what things like Type-2 computability also ask for.}
 
@@
 computation. So this is an interesting subject because it tells us
-something about computer rulesâŠ  (Feynman 1981)
+something about computer rules.  (Feynman 1981)
 
@@
 therefore that it could become the basis of an abstract theory of
-computation that was independent of the underlying physics. âHe
-thought,â as Feynman once put it, âthat he understood paper.â But he
+computation that was independent of the underlying physics. ``He
+thought, as Feynman once put it, that he understood paper. But he
 was mistaken. Real, quantum-mechanical paper is wildly different from
@@
 than that. He said that he was going to assume that information itself
-is conserved. ``Youâre out of you mind, Ed.'' I pronounced. ``Thatâs
+is conserved. ``You're out of you mind, Ed.'' I pronounced. ``That
 completely ridiculous. Nothing could happen in such a world. There
-couldnât even be logical gates. No decisions could ever be made.'' But
-when Fredkin gets one of his ideas, heâs quite immune to objections
+couldn't even be logical gates. No decisions could ever be made.'' But
+when Fredkin gets one of his ideas, he's quite immune to objections
 like that; indeed, they fuel him with energy. Soon he went on to
-assume that information processing must also be reversible â and
-invented whatâs now called the Fredkin gate. (Minsky 1999)
+assume that information processing must also be reversible and
+invented what is now called the Fredkin gate. (Minsky 1999)
 
@@
 our conceptions about logic are wrong, so wrong that they are unable
-to cope with the quantum miracle? [âŠ] Instead of teaching logic to
-nature, it is more reasonable to learn from her. Instead of
+to cope with the quantum miracle? [\ldots] Instead of teaching logic
+to nature, it is more reasonable to learn from her. Instead of
 interpreting quantum into logic, we shall interpret logic into
@@
 
-\jc{One must be careful with this argument: we have a ``current
-understanding'' of quantum physics, which could be quite flawed.
-So what we want to do is to learn from thsoe parts of quantum physics
+One must be careful with this argument: we have a ``current
+understanding'' of quantum physics, which could be quite flawed.  So
+what we want to do is to learn from those parts of quantum physics
 which seem to be extremely stable (like conservation of information!)
@@
 makes sense to vigorously pursue the conclusions one can draw from
-those aspects of (quantum) physics that seem stable.}
+those aspects of (quantum) physics that seem stable.
 
@@
 
-\jc{Should we also have a quick survey of other works in reversible
-computation? Most of the other work I know tries to start from
-well known models of computation or well known programming languages,
-and then adapts them to be reversible. Pi is different in that it
-starts from the semantic implications of reversibility. Then, because
-it adds types, rather than control-flow (or even data-flow) as its
-next layer of ``understanding'', this leads to equivalences, isomorphisms,
-etc.  This then leads, quite naturally, to finding that the ``proof
+We should also have a quick survey of other works in reversible
+computation? Most of the other work tries to start from well known
+models of computation or well known programming languages, and then
+adapts them to be reversible. Pi is different in that it starts from
+the semantic implications of reversibility. Then, because it adds
+types, rather than control-flow (or even data-flow) as its next layer
+of ``understanding'', this leads to equivalences, isomorphisms, etc.
+This then leads, quite naturally, to finding that the ``proof
 language'' of semirings (and Rig Groupoids at level 2) is actually a
-programming language. And it is Pi. This is a neat twist on Curry-Howard
-because CH is about \textbf{inhabitation} only. But with ``conservation
-of information'' as the basis, a different kind of correspondance
-arises; in fact, this one may well be an actual isomorphism.}
+programming language. And it is Pi. This is a neat twist on
+Curry-Howard because CH is about \textbf{inhabitation} only. But with
+``conservation of information'' as the basis, a different kind of
+correspondance arises; in fact, this one may well be an actual
+isomorphism.
 
@@
 
-\jc{This is where I am still least sure. I am wary of either doing
-new research with such a tight timeline, or of being too speculative.
-If there are papers on this already, perhaps references to them could
-be put here?}
+% \jc{This is where I am still least sure. I am wary of either doing
+% new research with such a tight timeline, or of being too speculative.
+% If there are papers on this already, perhaps references to them could
+% be put here?}
 

### 2017-11-14 survey/outline.tex
_
dump of some old material

@@
 \usepackage{fullpage}
+\usepackage{url}
 \title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
@@
 \newcommand{\roshan}[1]{\fbox{Roshan says:} \textbf{#1}}
+
+\newcommand{\lcal}{\ensuremath{\lambda}-calculus}
 
@@
 % following one strand, namely one aspect of quantum physics.}
+
+\begin{small}
+\begin{quote}
+``Turing hoped that his abstracted-paper-tape model was so simple, so
+  transparent and well defined, that it would not depend on any
+  assumptions about physics that could conceivably be falsified, and
+  therefore that it could become the basis of an abstract theory of
+  computation that was independent of the underlying physics.  `He
+  thought,' as Feynman once put it, `that he understood paper.' But he
+  was mistaken.''
+\end{quote}
+\hfill --- \emph{The Fabric of Reality}~\cite{deutsch}
+\end{small}
+
+%%%
+\subsection{Software: Foundations and Applications}
+
+We begin with a bold thesis: \textit{common computational models
+  embody various physical assumptions that are plainly
+  \emph{wrong}}. As we discuss in the remainder of this chapter, we
+are not the first to make such a statement. We do, however, make two
+novel contributions:
+\begin{itemize}
+\item we argue on pragmatic terms that our physically incorrect
+  computational abstractions contribute significantly to the
+  ``software crisis'' and that they inhibit advances in new emerging
+  application domains; 
+\item we identify the physical principle of \textbf{conservation of
+  information} as a simple twist to our computational models that has,
+  as we demonstrate in this book, far reaching theoretical and
+  practical positive consequences to the science of computing.
+\end{itemize}
+
+To appreciate the first point, we examine various textbook abstractions and
+contrast them with the actual (i.e., physical) computational situations to
+which they correspond. Generally speaking, we find that textbook algorithms
+differ from ``real programs'' in their assumptions about their environment of
+operation. In textbook abstractions, the surrounding environment which
+includes, for example, the data on which the algorithm operates is considered
+static and well-defined. In the real world, this environment is dynamic and
+noisy.
+
+\paragraph*{Sorting.} 
+One might consider sorting to be a fairly simple ``solved'' problem. Indeed
+textbook sorting algorithms have been formally specified, proved correct,
+analyzed in the best case, worst case, average case, and are even taught in
+many high school curricula. A beginner programmer can readily implement such
+algorithms to sort ``real data'' --- or can he?
+
+Under some assumptions, the answer is clearly yes. However, a moment's
+reflection should convince us that the textbook algorithm is too idealized
+and that sorting real data in the real world is still a challenge!  Consider
+the following quote posted on the Google research blog in 2011:
+\begin{quote}
+  Almost three years ago we announced results of the first ever ``petasort''
+  \ldots . It completed in just over six hours on 4000 computers. Recently we
+  repeated the experiment using 8000 computers. The execution time was 33
+  minutes, an order of magnitude improvement~\cite{petasort}.
+\end{quote}
+What happens at the petabyte scale that makes sorting such a
+challenge? The answer is simply that ``the real world happens!'' In
+the real world, many of the idealized mathematical assumptions
+embodied in our computational models are invalidated. For example,
+proper hardware functioning can no longer be taken for granted as
+component failure becomes the norm, global synchronization becomes
+impossible, determinism becomes expensive to maintain, speculative
+execution becomes necessary, global consistency becomes impractical to
+maintain, and eventual consistency is accepted; etc. There have been
+many attempts to taken some of these issues into account in various
+computational models but these attempts are arguably not going far
+enough by not questioning the physical assumptions in our
+computational models and re-thinking our very foundations.
+
+\paragraph*{Binary Search.} 
+In the idealized textbook setting, the binary search algorithm has the
+precondition that the data is sorted. Even if we just have a few thousands
+items, na\"\i vely checking sortedness (in linear time) defeats the purpose of
+the efficient (logarithmic) binary search. If we have petabytes of
+(changing!) data, is it even possible, in principle, to check sortedness? In
+fact, in this dynamic situation, one is likely to get inconsistent
+results. In other words, the very idea of pre-conditions and post-conditions
+that can be checked statically appears to be at odds with computing in the
+``real world.''
+
+\paragraph*{Security.} 
+Proving the security properties of an abstract mathematical model is great
+(and necessary). This can be used to conclude that an isolated device is
+secure but it is, however, not that relevant for applications based on RFID
+devices which are targeted via DPA (Differential Power Analysis)~\cite{dpa}
+and and DEMA (Differential Electromagnetic Analysis) attacks. There are
+billions of such devices with applications like electronic passports. Power
+consumption and electromagnetic properties are abstracted away in our
+computational model.
+
+\paragraph*{Privacy.} 
+The Internet has created major concerns about privacy. Differential
+privacy~\cite{dwork:differential} states that one can achieve privacy
+by perturbing the output of a deterministic program by a suitable
+amount of noise. The very idea of noise is abstracted away in our
+computational and communication models and the amount of noise is a
+non-existent abstraction in our typical computational models.
+
+\paragraph*{A Crisis?}
+Is this tension between theory and practice a crisis?  We argue:
+yes. Computer science theory is founded on abstractions of the real
+world. These abstractions are necessary to shield us from details of
+technology, noise, errors, etc. and they have served us well until
+recently.  It is, however, becoming more and more evident that
+computer science theory is becoming less and less relevant to
+practice, and hence it is important to revisit our abstractions from a
+foundational point of view.
+
+%%%
+\subsection{Our Origins}
+
+Until Alan Turing's seminal work in 1936, computation was an activity
+of the human mind; much like love and peace. While aspects of the idea
+predated the twentieth century, it was Turing who established the idea
+that computation has a formal interpretation and that all
+computability can be captured within a formal system. Implicit in this
+achievement however is the idea that abstract models of computation
+are just that -- \emph{abstractions of computation realized in the
+  physical world.}  In fact, one of the major achievements of Computer
+Science has been the development of abstract models of computation
+that shield the discipline from the underlying technology. As
+effective as these models have been, one must note, however, that they
+\emph{embody several implicit physical assumptions}.  As Tommaso
+Toffoli explains in his influential 1980 paper:
+
+{\begin{quote} Mathematical models of computation are abstract
+  constructions, by their nature unfettered by physical laws. However,
+  if these models are to give indications that are relevant to
+  concrete computing, they must somehow capture, albeit in a selective
+  and stylized way, certain general physical restrictions to which all
+  concrete computing processes are subjected~\cite{Toffoli:1980}.
+\end{quote}}
+
+Our logic of programs and our hardware are based on Boolean Logic. We go back
+to Boole's 1853 book entitled \emph{An Investigation of the Laws of Thought,
+  on which are Founded the Mathematical Theories of Logic and
+  Probabilities}. The opening sentence of Ch.~1 is:
+\begin{quote}
+  The design of the following treatise is to investigate the fundamental laws
+  of those operations of the mind by which reasoning is performed; \ldots
+\end{quote}
+A few chapters later, we find:
+\begin{quote}
+  \textbf{Proposition IV.}  That axiom of metaphysicians which is termed the
+  principle of contradiction, and which affirms that it is impossible for any
+  being to possess a quality, and at the same time not to possess it, is a
+  consequence of the fundamental law of thought, whose expression is $x^2 =
+  x$.
+\end{quote}
+A detailed historical analysis of Boole's ideas are beyond the scope of this
+book. The above quotes, however, should convey the idea that our hardware is
+based on assumptions that Boole found as reasonable laws in 1853.
+
+Our theory of computation and complexity is based on the Turing Machine. We
+go back to Turing's 1936 article entitled \emph{On Computable Numbers, with
+  an Application to the Entscheidungsproblem,}. The opening sentence of
+Sec. 1 is:
+\begin{quote}
+  We have said that the computable numbers are those whose decimals are
+  calculable by finite means \ldots the justification lies in the fact that
+  the human memory is necessarily limited.
+\end{quote}
+In Sec. 9, we find:
+\begin{quote}
+I think it is reasonable to suppose that they can only be squares
+whose distance from the closest of the immediately previously observed
+squares does not exceed a certain fixed amount.
+\end{quote}
+Again, a detailed historical analysis of Turing's ideas are beyond the scope
+of this book. The above quotes should again convey the idea that our theories
+of computation and complexity are based on some physical assumptions that
+Turing found reasonable in 1936. When translated to physical language, the
+assumptions embody what is known as the ``Bekenstein bound:'' which is an
+upper limit on the amount of information that can be contained within a given
+finite region of space.
+
+One can examine more and more cases but the general story should be
+clear. Our abstractions were made up based on a certain understanding of the
+laws of physics. And since, as we argue, our abstractions are becoming at
+odds with practice, it is time to revisit these abstractions. In the words of
+Girard:
+\begin{quote}
+  In other terms, what is so good in logic that quantum physics should obey?
+  Can't we imagine that our conceptions about logic are wrong, so wrong that
+  they are unable to cope with the quantum miracle? 
+  \\
+  Instead of teaching logic to nature, it is more reasonable to learn
+  from her. Instead of interpreting quantum into logic, we shall
+  interpret logic into quantum (Girard 2007).
+\end{quote}
+Indeed one should take the physical principles underlying quantum mechanics,
+the most successful physical theory known to us and adapt computation to
+``learn'' from these principles. To illustrate the depth of our crisis, Scott
+Aaronson, Umesh Vazirani, and others have proposed the following puzzle.
+
+One of these wild claims must be true!:
+\begin{itemize}
+\item the extended Church-Turing thesis is false, or
+\item quantum physics is false, or
+\item there is an efficient classical algorithm for factoring
+\end{itemize} 
+Indeed, if quantum physics is correct then there is an efficient quantum
+algorithm for factoring (Shor). If there is no efficient classical algorithm
+for factoring then the extended Church-Turing thesis is false.
+
+Indeed, conventional classical models of computation, including boolean
+logic, the Turing machine, and the $\lambda$-calculus, are founded on
+primitives which correspond to \emph{irreversible} physical processes which
+gratuitously erase information.  For example, a \emph{nand} gate is an
+irreversible logical operation in the sense that its inputs cannot generally
+be recovered from observing its output, and so is the operation of overriding
+a cell on a Turing machine tape with a new symbol, and so is a
+$\beta$-reduction which typically erases or duplicates values in a way that
+is destructive and irreversible.
+
+The irreversible abstractions we choose as the basis of our most
+primitive models of computing are at ``odds'' with the underlying
+reversible physical reality. The information theoretic underpinnings
+of physics is best understood through the remarkable \emph{Landauer
+  principle}~\cite{Landauer:1961}. Landauer, through his analysis of
+Maxwell's Demon, established the idea that \emph{information is a
+  physical quantity} in much the sense that one would think of an
+electron as a physical quantity, and that a certain minimum amount of
+thermodynamic work has to be carried out to erase one bit of
+information. Experimental verification of the Landauer principle
+happened in recent years~\cite{berut2012experimental}. If information
+has a physical significance, then it is subject to conservation laws
+as mass and energy are.
+
+To make this idea concrete and to provide a taste of the applications
+it opens, consider a tiny 2-bit password = \verb|"10"|. The password
+checker looks like:
+
+%subcode{code}
+%! language = Haskell
+% check-password (guess) = 
+%   if guess == "10"
+%   then True
+%   else False
+
+One can ask how much information is leaked by this program assuming
+the attacker has no prior knowledge except that the password is 2
+bits, i.e., the four possible 2-bits are equally likely. If the
+attacker guesses \verb|"10"| (with probability $1/4$) the password (2
+bits) is leaked. If the attacker guesses one of the other choices
+(with probability $3/4$) the number of possibilities is reduced from 4
+to 3, i.e., the attacker learns $\log{4} - \log{3}$ bits of
+information. So in general the attacker learns\footnote{If the
+  password is 8 restricted ASCII characters (6 bits), the attacker
+  learns 0.00001 bits in the first probe.}:
+\[\begin{array}{ll}
+   &  1/4 * 2 + 3/4 (\log{4} - \log{3}) \\
+  =&  1/4 \log{4} + 3/4 \log{4/3} \\
+  =&  - 1/4 \log{1/4} - 3/4 \log{3/4} \\
+  \sim& 0.8 \mbox{~bits~in~the~first~probe}
+\end{array}\]
+
+One can alternatively look at the situation by viewing the input as a
+random variable with 4 possibilities and a uniform distribution (i.e.,
+with 2 bits of information). The output is another random variable
+with 4 possibilities but with the distribution
+$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
+information. Thus 2 input bits of information were given to the
+function and only 0.8 were produced. Where did the 1.2 bits of
+information go? Once we accept the thesis that information is a
+physical entity this question cannot be ignored.
+
+The Landauer Principle states that erasing information requires
+energy. In other words, the password checking function must have
+erased 1.2 bits during its calculation which dissipate as heat in any
+physical realization of the function. In conventional models of
+computation, this erasure is \emph{implicit}, i.e., it occurs as a
+side effect of calculation. A big part of our development will be
+concerned with making that erasure explicit.
+
+The discussion in this section suggests that our foundations are really in
+need of revision. There are many principles we can embrace.  We will start
+with one, ``\textbf{conservation of information}'' and follow its
+consequences, which will turn out to be far reaching.
+
+%%%
+\subsection{Equalities as Isomorphisms}
+
+\begin{quote}
+  The laws of physics are essentially algorithms for calculation. These
+  algorithms are significant only to the extent that they are executable in
+  our real physical world. Our usual laws of physics depend on the
+  mathematician's real number system. With that comes the presumption that
+  given any accuracy requirement, there exist a number of calculations steps,
+  which if executed, will satisfy that accuracy requirement. But the real
+  world is unlikely to supply us with unlimited memory of unlimited Turing
+  machine tapes, Therefore, continuum mathematics is not executable, and
+  physical laws which invoke that can not really be satisfactory. They are
+  references to illusionary procedures. Rolf Landauer, The physical nature of
+  information, Physics Letters A 217 (1996) pp. 188-193:
+\end{quote}
+
+When a physical law equates two quantities like force and the product of mass
+and acceleration, a deep understanding of the law demands that we interpret
+the equality as an isomorphism, i.e., as two processes, one in each
+direction, that transform one side of the equation to the other. In fact, the
+same insight applies to constructive mathematics and its homotopy
+interpretation. There is a constructive character to equality that is at the
+heart of the constructive approach to mathematics.
+
+Translated to the computational world, the essence of our thesis is this:
+\emph{to speak of equality is to speak of computation}. This view is partly
+familiar from the study of various systems of equalities, called
+\emph{calculi}, as models of computation. The \lcal, for instance, is a
+system of equalities over a set of syntactic $\lambda$-terms such that they
+result in a model of computation. We distill this idea to its most primitive
+form. We show that there is computational content not just in certain
+specific calculi, such as the \lcal, but even in our most primitive notion of
+equality -- isomorphisms of finite sets.
+
+The equalities we speak of are not \emph{a priori} extensional
+notions, but are intensional specifications. To talk about two things
+being equal, one has to show \emph{how they are equal} wherein this
+\emph{how} is an intensional definition of a translation from
+witnesses of the first to witnesses of the second and vice versa,
+i.e. they are programs. Equivalences that do not have operational
+content do not exist for us.
+
+One may ask, what is the point of such a rarefied spartan view of
+computation? How does this compare with more conventional models of
+computation?  \emph{What is the point?}
+
+\begin{enumerate}
+
+\item The result is computation in its purest form: Programs written
+  in this model are witnesses of equalities. In the process of
+  computing these programs neither gain nor lose information. We get a
+  crisp notion of quantitative information content of programs and
+  show that \emph{computation preserves information}.
+
+\item From the perspective of Quantum Physics, the physical world is
+  one where every fundamental interaction is essentially reversible
+  and various quantities such as energy, mass, angular momentum are
+  conserved. Computation is no longer at odds with
+  physics. Conservation of information and logical reversibility are
+  intrinsic properties of computation as well. Further, various
+  category theoretic models such as dagger symmetric traced monoidal
+  categories studied for Quantum Physics are also the categorical
+  models for computation based on isomorphisms.
+
+\item Irreversibility is a computational effect: By embracing
+  irreversible physical primitives, models of computation have also
+  implicitly included a class of computational effects which we call
+  \emph{information effects}. A consequence of our careful treatment
+  of information, is that we effectively capture the gap between
+  reversible computation and irreversible computation using a
+  type-and-effect system.
+
+\end{enumerate}
+
+What do we mean when we talk about intentional specification of
+equality leading to an information preserving model of computation?
+The equalities we talk of are processes to transform one thing to the
+other. For example, if one were to believe that apples and oranges
+were equal then one would have to show how any given apple can be
+transformed to an orange i.e. witness of apple-hood get transformed to
+witnesses of orange-hood (and vice versa). This is inherently
+different from an extensional notion of equality such as
+\emph{knowledge is power} and have a more operational sense. Equality
+in this sense acts more like the the more general notion of
+\emph{isomorphism} as probably best explained in the paper
+``\emph{When is one thing equal to some other
+  thing?}''~\cite{mazur2008one}.
+
+In our formalism, we are not concerned with fruit but with very simple
+things, namely numbers and sets. When we talk of a set of five apples,
+we are not concerned about the individuality of the apples that
+constitute the set. Rather, each apple in the set has the same status
+as any other apple and we care about the individual apples only to the
+extent that we can tell them apart to count five distinct ones. We
+care about the abstract idea of fiveness and its equality with any
+other abstract idea of fiveness.  When used as types, sets take the
+role of possibility spaces -- we are no longer talking about a set of
+five apples, but talking of one apple out of a possible five. This is
+also where the switch from sets to information happens: When one apple
+out of five is identified, it corresponds to the information to
+discriminate between five choices. Clearly a choice of one out of ten
+presents more information in the sense that it is a more
+discriminating choice. When a computation promises an output of a
+certain type, the resultant value is a witness of one choice out of
+several. This is the sense in which we are concerned with equalities
+of sets.
+
+Programs in our model are built from descriptions of equalities or
+sets/numbers. These are the familiar laws of arithmetic or logic:
+\emph{Commutativity.} The generalized notion of swapping,
+i.e. $a*b=b*a$ and $a+b=b+a$.  
+\emph{Associativity.} The
+generalized notion of grouping, i.e. $a*(b*c)=(a*b)*c$ and
+$a+(b+c)=(a+b)+c$.
+\emph{Identity.} Multiplication by 1 and addition with 0,
+  i.e. $a*1=a$ and $a+0=a$.
+\emph{Distributivity.} Multiplication distributes over addition
+  and cancels at zero, i.e. $a*(b+c)=a*b+a*c$ and $a*0=0$.
+\emph{Cancellation.} Cancellation is the idea that equal things
+  on both sides of a equality can be canceled out, i.e. $a+b=a+c$
+  implies $b=c$, and is referred to as a \emph{trace} in category
+  theory~\cite{joyal1996traced}.
+
+How, one may ask, does this form a computational model? Even in very
+simple computational models such as the SK combinatory basis, the
+combinator K provides the basis of choice/conditional by deleting an
+argument and the combinator S provides the basis for iteration by
+duplicating an argument. As we will see, distributivity gives us
+conditionals and trace gives us iteration, albeit in a logically
+reversible manner.  In Chs.~\ref{ch:pi} and~\ref{ch:pi0} we work out
+the details of the computational model, show that it is Turing
+complete and develop programs in it including several simple numeric
+operations, algebraic manipulations of trees and a meta-circular
+interpreter (see Ch.~\ref{ch:interpreters}).
+
+More formally, we develop an information preserving model of
+computation, wherein the process of computing does not gain or lose
+information. Our model arises from a computational interpretation of
+type isomorphisms with iso-recursive types and trace. This can
+equivalently be described as computing a commutative semi-ring with
+cancellation or as a traced dagger symmetric bimonoidal category. The
+category has a groupoid structure and all computations are logically
+reversible.
+
+To summarize, the most primitive notion of equality that is available
+to us -- that of isomorphisms -- already contains the notion of
+computation. The resulting computational model exposes the fine
+structure of how computation handles information. We believe that such
+a model of computing is foundational and we are only beginning to
+understand its applications and full potential. In conclusion, let us
+quote Paul Blain Levy's fine advocacy slogan: Once the fine structure
+has been exposed, why ignore it?
+
+%%%
+\subsection{Information}
+
+We can now describe information preservation in a precise sense.  Let
+`$b$' be a (not necessarily finite) type whose values are labeled
+$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
+equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
+as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
+$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
+type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
+entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
+indicates the probability of the output of the function to have value
+${b_2}^j$. We say a function is \emph{information-preserving} if its
+output entropy is equal to the entropy of its input. See
+Ch.~\ref{ch-metalang} for details.
+
+Operations that change the information content of programs have a
+special status. We call them \emph{information effects} and they are
+encapsulated using an arrow meta-language in much the same way that
+one would encapsulate other computational effects in the \lcal\ using a
+monadic metalanguage.  The treatment of information effects is
+analogous to open and closed systems in physics. Closed physical
+systems conserve mass and energy and are the basic unit of study in
+physics. Pure computations that do not have information effects are
+like closed physical systems. They describe equalities.  Open systems
+interact with their environment, possibly exchanging mass or
+energy. These interactions may be thought of as \emph{effects} that
+modify the conservation properties of the system. Computations with
+information effects are much like open systems and they can be
+converted into pure computations by making explicit the surrounding
+information environment that they interact with.
+
+Three things follow: (1) We capture the gap between reversible and
+irreversible computation with a type-and-effect system. (2)
+Categorically speaking, information effects turn monoidal categories
+into ones that have a cartesian structure corresponding to
+conventional computation.  (3) We show how conventional irreversible
+computation such as the \lcal\ can be embedded into this model, such
+that the embedding makes the implicit information effects of \lcal\
+explicit.  As a consequence of this approach, many applications in
+which information manipulation is computationally significant are put
+within the reach of our conceptual model of computation. Such
+applications include quantitative information-flow
+security~\cite{myerssab}, differential
+privacy~\cite{dwork:differential}, energy-aware
+computing~\cite{1324180,605411}, VLSI
+design~\cite{Macii:1996:ECE:874066.875828}, and biochemical models of
+computation~\cite{bio}.
+
+Ch.~\ref{ch:piee} focuses on work in progress. It deals with the
+duality of computation in the setting of information
+preservation. Andrzej Filinski's work on the symmetric
+\lcal\ established the fascinating connection that values are dual to
+continuations, functions are dual to delimited continuations and
+established the duality of call-by-name and call-by-value evaluation
+strategies~\cite{Filinski:1989:DCI:648332.755574}. We explore
+Filinski-style duality in the setting of our information preserving
+model of computing. Here duality does not appear as one De
+Morgan-style dualizing $\neg$ operator, but as two distinct dualizing
+operations -- negation and division. This results in a crisp semantics
+for negative and fractional types and the resulting system has a
+equational formalism in zero-totalized fields or
+meadows~\cite{DBLP:journals/tcs/BergstraHT09}. Computational analogues
+of physical phenomena such as superposition and entanglement appear in
+the resulting semantics.
+
+Talk about conservation of information in biology, computer science,
+evolutionary computing, and search. (See
+\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
+
+%%%
+\subsection{Technical Overview}
+
+The technical development derives from ideas which are implicit in the
+works of \cite{Toffoli:1980, Zuliani:2001:LR,
+  malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.
+(1) Functions whose output entropy is equal to their input entropy are
+information preserving functions.  (2) A function
+$f : b_1 \rightarrow b_2$ is \emph{logically reversible} if there
+exists an inverse function $f^{\dagger}$ such that for any input
+$v_1:b_1$ if there exists $v_2:b_2$ such that $f(v_1)=v_2$ \emph{iff}
+$f^{\dagger}(v_2)=v_1$.  (3) Logically reversible functions are
+information preserving.
+
+To summarize:
+\begin{enumerate}
+\item Based on isomorphisms of sum and product
+  types~\cite{Fiore:2004}, we develop a strong normalizing model of
+  computing, which we call $\Pi$, that is logically reversible and
+  information preserving. Every computation expressible in $\Pi$
+  is an isomorphism.
+
+\item We extend $\Pi$ with isorecursive type and trace operators from
+  category theory to obtain $\Pi$. Every computation expressible in
+  $\Pi$ is a partial isomorphism --- i.e. the system admits
+  non-termination. $\Pi$ is Turing complete while being information
+  preserving and logically reversible.
+
+\item We develop the categorical semantics of these
+  models~\cite{rc2011}. \amr{Mention n-categories.} We also develop
+  a graphical notation for programming in these models that
+  reminiscent of Penrose diagrams for
+  categories~\cite{selinger-graphical}, Geometry of Interaction (GoI)
+  machines and Proof
+  Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
+  notation, computation is modeled by the flow of particles in a
+  circuit.
+
+\item Since these new models are substantially different from \lcal\
+  or familiar high-level languages, a point of concern is how one can
+  effectively develop programs in them.  We address this by presenting
+  a straightforward technique for deriving $\Pi$ programs by
+  systematically translating logically reversible small-step abstract
+  machines~\cite{isoint}. Complex programs can be devised in this way;
+  we demonstrate the derivation of a meta-circular interpreter for
+  $\Pi$.
+
+\item We develop an arrow meta-language to encapsulate information
+  effects. This metalanguage serves to encapsulate effects, in much
+  the same way that traditional arrows or monads serve to encapsulate
+  effects over the \lcal.
+
+\item We show how a conventional irreversible model can be expressed
+  in our model. We compile the first-order fragment of typed \lcal\
+  extended with sums, products and loops to $\Pi$. This compilation is
+  interesting for two reasons: (1) it exposes the implicit information
+  effects of \lcal\ in the compilation to $\Pi$ and (2) the
+  compilation of $\Pi$ to $\Pi$ shows that information effects can be
+  erased by treating them as interactions with an explicit information
+  environment.
+
+\item We develop the notion of `duality' in $\Pi$ which gives us not
+  one duality (like in linear logic or Filinski's symmetric
+  \lcal~\cite{Filinski:89}), but two notions of duality -- an additive
+  duality and a multiplicative duality. This gives a crisp semantics
+  for negative and fractional types in the context of $\Pi$.
+
+\end{enumerate}
+
+Several intriguing areas of exploration remain and the book concludes
+with discussion of some of these, along with other possible
+applications (see Ch.~\ref{ch:discussion}).  Of these, the field of
+reversible and Quantum computing is closely related and we survey
+several connections in depth.  The exact connection between $\Pi$ and
+linear logic~\cite{Girard87tcs} is not fully understood though it is
+clear that both systems capture related but different notions of
+resource usage. A closely related area is the connection with duality
+of computation \cite{Filinski:89, DBLP:conf/icfp/CurienH00,
+  10.1109/LICS.2010.23,Wadler:2003} --- i.e., the quest for a unifying
+framework between `computations', values, continuations and their
+logical counterparts. The investigation of duality gives rise to
+negative and fractional types which have an intuitive semantics and
+are reminiscent of negative information flow, superposition and
+entanglement from Quantum Physics~\cite{piee}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{plainnat}
+\pagestyle{headings}
+\bibliography{cites}
 \end{document}

### 2017-11-24 survey/survey.tex
add 2 citations. start work in survey.tex, from outline.tex

--- 2017-11-14 survey/outline.tex
@@
 
-To fully appreciate the missing principle of conventional computing,
-one must go back to an old thought experiment by J. C. Maxwell,
-codified in a letter that Maxwell wrote to P. G Tait in 1867 -- this
-is the letter whose ideas are now well known as the `Maxwell's
-Demon'. Maxwell's Demon was a thought experiment that seemed to
+To fully appreciate one of the principles is missing in conventional computing,
+one must go back to an old thought experiment by J. C. Maxwell. This is
+codified in a letter that Maxwell wrote to P. G Tait in 1867 -- 
+the letter whose ideas are now known as `Maxwell's
+Demon'. Maxwell's Demon is a thought experiment that seems to
 indicate that intelligent beings can somehow violate the second law of
@@
 Many resolutions were offered for this conundrum (for a compilation,
-see Maxwell's Demon by Leff and Rex), but none withstood careful
+see Maxwell's Demon by Leff and Rex~\cite{leff1990}), but none withstood careful
 scrutiny until the establishment of the so called `Landauer's

### 2017-11-24 survey/survey.tex
introduction that can lead into the paper as currently envisioned.

@@
 
-Many resolutions were offered for this conundrum (for a compilation,
-see Maxwell's Demon by Leff and Rex~\cite{leff1990}), but none withstood careful
-scrutiny until the establishment of the so called `Landauer's
-Principle' in 1961 -- a principle whose experimental validation
-happened recently in 2012 by Eric Lutz and others.
+Many resolutions were offered for this conundrum (for a compilation, see
+Maxwell's Demon by Leff and Rex~\cite{leff1990}), but none withstood careful
+scrutiny until the establishment of `Landauer's Principle' in 1961 -- a
+principle whose experimental validation
+happened recently in 2012~\cite{lutz2012}.
 
@@
 room for the next computation.  Landauer reasoned, and this is the
-important bit, that the computation that is manipulating information
+important part, that the computation that is manipulating information
 in the demon's brain \textit{must be thermodynamic work}, thereby
@@
 
-In its essence this is a strange and wonderful idea. Information,
-physics and computation are inextricably linked and that ideas in each
-field have implications for the other. (Cite Bennett and the various
-`thermodynamics of computation' papers here.)
+This is a strange and wonderful idea: Information,
+physics and computation are inextricably linked. Furthermore, this
+implies that ideas in each field have consequences for the other. 
+(Cite Bennett and the various `thermodynamics of computation' papers here.)
 
 When the early models computation, the Turing machine, and the
-$\lambda$-calculus, were developed there was no real reason to the
-take the information content of computations into serious
+$\lambda$-calculus, were developed, there was no compelling reason to the
+take the information content of computations into
 consideration -- in fact, at that time there was no quantifiable
 notion of information. These models followed in the footsteps of logic
-where conventional logic gates such as AND and OR happily erase their
-inputs.
+where conventional gates such as AND and OR happily erase their
+inputs. These models were already radical enough, by being purely
+constructive, in an era where classical mathematics, with its
+pervasive use of excluded middle and frequent use of the axiom of
+choice, did not overly worry about effective representation.
 
@@
 
-Now however, through the establishment of the Landauer's principle and
-through recent investigations into computational models that are well
-suited to quantum mechanics it has become increasingly evident that
-our computation models need to be revisited. Information, not just
-seems to have a physical manifestation, but also that it, like other
-physical entities, is conserved.
+In our current age, both Laudauer's Princple and recent investigations
+of computational models well-suited to quantum mechanics, push us to
+revisit our current models of computation. Information, on top of
+having a physical manifestation, appears to be \emph{conserved},
+along with other quantities like energy, mass and momentum.
 
-Going back to the drawing board, can we have computation that closely
-matches our underlying physics?  i.e. can we have computation that in
-its essence preserves information? What is the missing principle that
-we can use as a basis of a new model of computation to achieve this?
-It turns out that we need to look at physics again for inspiration --
-in physics all of the primitive rules, nature itself at its most
-fundamental level, is reversible. Conventional computation is not. Can
+But what does it mean for a computation to preserve information?
+What is the missing principle that will guide the creation of a new
+model of computation that ``conserve information''?
+It is nature to look again at physics for inspiration --
+in our current understanding of physics all of the primitive rules
+are reversible. Conventional computation is not. Can
 we embrace this simple principle as the building block of a model of
 computation? What would computation look like when viewed from this
-vantage point? And, does it have applications?
+vantage point? Can non-trivial computation be performed in a
+reversible manner? And, if so, does it have applications?
 
-The objective of this survey is to touch upon the many related areas
-of research in this space.
+Traditional models of computation are all unityped, generally
+operating on unstructured sequences of symbols. But recent
+investigations, in the intersection between mathematics and
+computer science (particularly in Homotopy Type Theory~\cite{HoTT-book})
+reveal that not only can mathematics be typed, but that doing so
+reveals further structure -- in both mathematics and in computation.
 
-%% Abstract models of computation: Turing machine, RAM, Von Neumann
-%% architecture, $\lambda$-calculus, logic, discrete mathematics
-
-%% \jc{why are logic and DM listed alongside MoC?}
+And that brings us to the objective of this survey: explore the domain of
+\emph{typed reversible computation}.
 

### 2017-11-24 survey/survey.tex
add some text to a section, and then comment out the whole section as it is not relevant anymore. Checking this in to preserve it, but will actually delete it next.

@@
 \usepackage{url}
+\usepackage{comment}
+
 \title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
@@
 Many resolutions were offered for this conundrum (for a compilation, see
-Maxwell's Demon by Leff and Rex~\cite{leff1990}), but none withstood careful
+Maxwell's Demon~\cite{leff1990}), but none withstood careful
 scrutiny until the establishment of `Landauer's Principle' in 1961 -- a
 principle whose experimental validation
-happened recently in 2012~\cite{lutz2012}.
+happened recently in 2012~\cite{berut2012experimental}.
 
@@
 
-And that brings us to the objective of this survey: explore the domain of
-\emph{typed reversible computation}.
+And that brings us to the principal objective of this survey: explore the
+domain of \emph{typed reversible computation}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Models of Computation}
-
-Explain classical models of computation and their limitations in
-real-world ``physical'' applications.
-
-Computation in the ``real world'' vs. idealized computation in
-abstract models.
-\begin{itemize}
-\item Mathematical abstraction of sorting is well-understood; sorting
-  petabytes of data violates many of the assumptions: hardware
-  component failure frequent, global synchronization impossible;
-  consistency must be eventual; etc.
-\item Mathematical abstraction of sortedness property: easy; in the
-  real world with petabytes of evolving data; it is not clear that it
-  is even possible to check sortedness; easy to get inconsistent
-  results
-\item Noise does not exist in our mathematical abstractions of
-  computation; essential use in differential privacy
-\item Mathematical models of secure computing typically ignore power
-  consumption, electromagnetic signatures; actual attacks based on DPA
-  and DEMA however
-\end{itemize}
-
-% \jc{While all the above are true (and not exhaustive), what I also read
-% is that the first two are actually ``the same'' point mostly related
-% to scale and distributivity, the 3rd is about inadequate models (but not
-% related to physics), while the 4th is about the physics of devices.
-% They are not about CPS. And not about Quantum either. So how are these
-% motivating a deeper look at how quantum affects our MoC? It feels to me
-% that we ought to have a quantum ``example'' here already.}
-
-% \jc{What this seems to motivate is to look at our models with respect
-% to what we're trying to achieve.  Models are, by definition, abstractions
-% that ``abstract away'' irrelevant details. So the point would be that
-% some details that we thought were irrelevant really are not.
-% So I could see this section motivating a whole programme of
-% targeted de-abstraction. This could lead into the focus of this paper
-% following one strand, namely one aspect of quantum physics.}
-
-\begin{small}
-\begin{quote}
-``Turing hoped that his abstracted-paper-tape model was so simple, so
-  transparent and well defined, that it would not depend on any
-  assumptions about physics that could conceivably be falsified, and
-  therefore that it could become the basis of an abstract theory of
-  computation that was independent of the underlying physics.  `He
-  thought,' as Feynman once put it, `that he understood paper.' But he
-  was mistaken.''
-\end{quote}
-\hfill --- \emph{The Fabric of Reality}~\cite{deutsch}
-\end{small}
+%
+% JC: I am commenting this out as, intereting as this all is, a lot of
+%     it is about ``real world'' and NOT about the intersection of physics
+%     and computation.  It would be possible to write about that too.
+%     For example, the hardware failures of petabyte scale computing is
+%     physical; physics-based attacks on privacy and security; physics
+%     of dealing with noisy environments (key debouncing, timing drifts,
+%     spurious data, heart rate spikes, etc);
+\begin{comment}
+\section{Physics and computation}
 
@@
 embody various physical assumptions that are plainly
-  \emph{wrong}}. As we discuss in the remainder of this chapter, we
-are not the first to make such a statement. We do, however, make two
-novel contributions:
-\begin{itemize}
-\item we argue on pragmatic terms that our physically incorrect
-  computational abstractions contribute significantly to the
-  ``software crisis'' and that they inhibit advances in new emerging
-  application domains; 
-\item we identify the physical principle of \textbf{conservation of
-  information} as a simple twist to our computational models that has,
-  as we demonstrate in this book, far reaching theoretical and
-  practical positive consequences to the science of computing.
-\end{itemize}
+\emph{wrong}}. We certainly
+are not the first to make such a statement. We do want to point
+out that these physically incorrect computational abstractions
+inhibit advances in computing in general.
 
-To appreciate the first point, we examine various textbook abstractions and
+To appreciate this point, we examine various textbook abstractions and
 contrast them with the actual (i.e., physical) computational situations to
@@
 many high school curricula. A beginner programmer can readily implement such
-algorithms to sort ``real data'' --- or can he?
+algorithms to sort ``real data'' --- or can she?
 
@@
 embodied in our computational models are invalidated. For example,
-proper hardware functioning can no longer be taken for granted as
+proper hardware functioning can no longer be taken for granted, as
 component failure becomes the norm, global synchronization becomes
@@
 
+As another data point to reinforce this point: What algorithm
+does Python (and other mainstream platforms) use? A variant of
+QuickSort or HeapSort perhaps.  No. Timsort. This hybrid algorithm
+performs rather well in practice -- but is complex, and even harder 
+to formally analyze. It does not tend to get mentioned in algorithms
+textbooks at all.
+
 \paragraph*{Binary Search.} 
@@
 precondition that the data is sorted. Even if we just have a few thousands
-items, na\"\i vely checking sortedness (in linear time) defeats the purpose of
+items, na\"{\i}vely checking sortedness (in linear time) defeats the purpose of
 the efficient (logarithmic) binary search. If we have petabytes of
-(changing!) data, is it even possible, in principle, to check sortedness? In
+(changing!) data, is it even possible, in principle, to check sortedness? 
+Or guarantee sortedness? In
 fact, in this dynamic situation, one is likely to get inconsistent
@@
 foundational point of view.
+\end{comment}
 
+\section{Models of Computation}
 %%%
+\begin{comment}
 \subsection{Our Origins}
@@
 for factoring then the extended Church-Turing thesis is false. 
+\end{comment}
 
-Indeed, conventional classical models of computation, including boolean
+Conventional classical models of computation, including boolean
 logic, the Turing machine, and the $\lambda$-calculus, are founded on
@@
 
-The irreversible abstractions we choose as the basis of our most
-primitive models of computing are at ``odds'' with the underlying
+These irreversible abstractions chosen as the basis of our
+models of computing are at ``odds'' with the underlying
 reversible physical reality. The information theoretic underpinnings
@@
 
-%subcode{code}
-%! language = Haskell
-% check-password (guess) = 
-%   if guess == "10"
-%   then True
-%   else False
+% verbatim for now, will go to something decent later
+\begin{verbatim}
+check-password (guess) = 
+  if guess == "10"
+  then True
+  else False
+\end{verbatim}
 
@@
 %%%
+\section{Reversibility}
+
 \subsection{Equalities as Isomorphisms}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Applications}
-
-Perhaps Information Flow Security
-
-% \jc{This is where I am still least sure. I am wary of either doing
-% new research with such a tight timeline, or of being too speculative.
-% If there are papers on this already, perhaps references to them could
-% be put here?}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions}
@@
 \bibliographystyle{plainnat}
-\pagestyle{headings}
+%\pagestyle{headings}
 \bibliography{cites}

### 2017-11-24 survey/survey.tex
continuing with the preliminaries. Did actually find a bunch of text that could be used.

@@
 
-%% Cyber-physical systems increasingly common. Are computing foundations
-%% adequate?
-%% \jc{your questioning of MoC below is not about CPS! Should either
-%% foreshadow the next section 'properly' here, or add a leading
-%% example based on CPS to MoC section}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% JC: I am commenting this out as, intereting as this all is, a lot of
-%     it is about ``real world'' and NOT about the intersection of physics
-%     and computation.  It would be possible to write about that too.
-%     For example, the hardware failures of petabyte scale computing is
-%     physical; physics-based attacks on privacy and security; physics
-%     of dealing with noisy environments (key debouncing, timing drifts,
-%     spurious data, heart rate spikes, etc);
-\begin{comment}
-\section{Physics and computation}
-
-%%%
-\subsection{Software: Foundations and Applications}
-
-We begin with a bold thesis: \textit{common computational models
-embody various physical assumptions that are plainly
-\emph{wrong}}. We certainly
-are not the first to make such a statement. We do want to point
-out that these physically incorrect computational abstractions
-inhibit advances in computing in general.
-
-To appreciate this point, we examine various textbook abstractions and
-contrast them with the actual (i.e., physical) computational situations to
-which they correspond. Generally speaking, we find that textbook algorithms
-differ from ``real programs'' in their assumptions about their environment of
-operation. In textbook abstractions, the surrounding environment which
-includes, for example, the data on which the algorithm operates is considered
-static and well-defined. In the real world, this environment is dynamic and
-noisy.
-
-\paragraph*{Sorting.} 
-One might consider sorting to be a fairly simple ``solved'' problem. Indeed
-textbook sorting algorithms have been formally specified, proved correct,
-analyzed in the best case, worst case, average case, and are even taught in
-many high school curricula. A beginner programmer can readily implement such
-algorithms to sort ``real data'' --- or can she?
-
-Under some assumptions, the answer is clearly yes. However, a moment's
-reflection should convince us that the textbook algorithm is too idealized
-and that sorting real data in the real world is still a challenge!  Consider
-the following quote posted on the Google research blog in 2011:
-\begin{quote}
-  Almost three years ago we announced results of the first ever ``petasort''
-  \ldots . It completed in just over six hours on 4000 computers. Recently we
-  repeated the experiment using 8000 computers. The execution time was 33
-  minutes, an order of magnitude improvement~\cite{petasort}.
-\end{quote}
-What happens at the petabyte scale that makes sorting such a
-challenge? The answer is simply that ``the real world happens!'' In
-the real world, many of the idealized mathematical assumptions
-embodied in our computational models are invalidated. For example,
-proper hardware functioning can no longer be taken for granted, as
-component failure becomes the norm, global synchronization becomes
-impossible, determinism becomes expensive to maintain, speculative
-execution becomes necessary, global consistency becomes impractical to
-maintain, and eventual consistency is accepted; etc. There have been
-many attempts to taken some of these issues into account in various
-computational models but these attempts are arguably not going far
-enough by not questioning the physical assumptions in our
-computational models and re-thinking our very foundations.
-
-As another data point to reinforce this point: What algorithm
-does Python (and other mainstream platforms) use? A variant of
-QuickSort or HeapSort perhaps.  No. Timsort. This hybrid algorithm
-performs rather well in practice -- but is complex, and even harder 
-to formally analyze. It does not tend to get mentioned in algorithms
-textbooks at all.
-
-\paragraph*{Binary Search.} 
-In the idealized textbook setting, the binary search algorithm has the
-precondition that the data is sorted. Even if we just have a few thousands
-items, na\"{\i}vely checking sortedness (in linear time) defeats the purpose of
-the efficient (logarithmic) binary search. If we have petabytes of
-(changing!) data, is it even possible, in principle, to check sortedness? 
-Or guarantee sortedness? In
-fact, in this dynamic situation, one is likely to get inconsistent
-results. In other words, the very idea of pre-conditions and post-conditions
-that can be checked statically appears to be at odds with computing in the
-``real world.''
-
-\paragraph*{Security.} 
-Proving the security properties of an abstract mathematical model is great
-(and necessary). This can be used to conclude that an isolated device is
-secure but it is, however, not that relevant for applications based on RFID
-devices which are targeted via DPA (Differential Power Analysis)~\cite{dpa}
-and and DEMA (Differential Electromagnetic Analysis) attacks. There are
-billions of such devices with applications like electronic passports. Power
-consumption and electromagnetic properties are abstracted away in our
-computational model.
-
-\paragraph*{Privacy.} 
-The Internet has created major concerns about privacy. Differential
-privacy~\cite{dwork:differential} states that one can achieve privacy
-by perturbing the output of a deterministic program by a suitable
-amount of noise. The very idea of noise is abstracted away in our
-computational and communication models and the amount of noise is a
-non-existent abstraction in our typical computational models.
-
-\paragraph*{A Crisis?}
-Is this tension between theory and practice a crisis?  We argue:
-yes. Computer science theory is founded on abstractions of the real
-world. These abstractions are necessary to shield us from details of
-technology, noise, errors, etc. and they have served us well until
-recently.  It is, however, becoming more and more evident that
-computer science theory is becoming less and less relevant to
-practice, and hence it is important to revisit our abstractions from a
-foundational point of view.
-\end{comment}
-
 \section{Models of Computation}
 %%%
-\begin{comment}
-\subsection{Our Origins}
 
-Until Alan Turing's seminal work in 1936, computation was an activity
-of the human mind; much like love and peace. While aspects of the idea
-predated the twentieth century, it was Turing who established the idea
+But first, we need to better understand our current models of
+computation, as a prelude to understand where they lead us astray.
+
+\subsection{Origins}
+
+Aspects of computation are
+quite old. M\"{u}ller (1786) first conceived of the idea of a
+``difference machine'', which Babbage (1819--1822) was able to
+construct. There are other computer precursors as well -- the first
+stored programs were actually for looms, most notably those of
+Bouchon (1725) which operated on a paper tape, and Jacquard (1804)
+which operated by chains of punched cards.
+
+But it was Alan Turing's seminal work in 1936 which established the idea
 that computation has a formal interpretation and that all
@@
 
-Our logic of programs and our hardware are based on Boolean Logic. We go back
+Our logic of programs and our hardware are based on Boolean Logic
+-- which really
+ought to be called \emph{Piercean logic}, as Boole had a rather different
+conception of logic which is rather far from our current models.
+Nevertheless, going back
 to Boole's 1853 book entitled \emph{An Investigation of the Laws of Thought,
@@
 \end{quote}
-A detailed historical analysis of Boole's ideas are beyond the scope of this
-book. The above quotes, however, should convey the idea that our hardware is
-based on assumptions that Boole found as reasonable laws in 1853.
+A detailed historical analysis of Boole's ideas are beyond our scope.
+The above quotes, however, should convey the idea that our notions of
+computation generally date back to ideas that were thought reasonable
+before \tilde 1900.
 
-Our theory of computation and complexity is based on the Turing Machine. We
-go back to Turing's 1936 article entitled \emph{On Computable Numbers, with
-  an Application to the Entscheidungsproblem,}. The opening sentence of
+It is conventional to base
+the theory of computation and complexity on the Turing Machine.
+Going back to Turing's 1936 article \emph{On Computable Numbers, with
+  an Application to the Entscheidungsproblem,}, the opening sentence of
 Sec. 1 is:
@@
 \end{quote}
-Again, a detailed historical analysis of Turing's ideas are beyond the scope
-of this book. The above quotes should again convey the idea that our theories
-of computation and complexity are based on some physical assumptions that
-Turing found reasonable in 1936. When translated to physical language, the
-assumptions embody what is known as the ``Bekenstein bound:'' which is an
-upper limit on the amount of information that can be contained within a given
-finite region of space.
+A detailed historical account is again beyond our scope. The quotes above
+should convey the ideas that our theories
+of computation and complexity are based on some assumptions that
+Turing found reasonable in 1936. Though it is worth noting that these assumptions
+are both physical (on distances) and metaphysical (on restrictions of the
+mind).  If we take the human mind to be a physical ``machine'' which does
+computation, then when both of the above assumptions are translated to the
+language of physics, they embody what is known as the ``Bekenstein bound:''
+which is an upper limit on the amount of information that can be contained
+within a given finite region of space.
 
@@
 clear. Our abstractions were made up based on a certain understanding of the
-laws of physics. And since, as we argue, our abstractions are becoming at
-odds with practice, it is time to revisit these abstractions. In the words of
-Girard:
+laws of physics. In particular, our understanding of physics has evolved
+tremendously since 1900!  Thus it is time to revisit these abstractions,
+especially with respect to quantum mechanics.
+In the words of Girard:
 \begin{quote}
@@
 \end{quote}
+
+\begin{comment}
+\jc{The reason I commented this out is that it is under-justified. The
+reader will simply not understand what these next few lines are really
+saying.}
 Indeed one should take the physical principles underlying quantum mechanics,
@@
 
+\subsection{From classical to quantum}
+
 Conventional classical models of computation, including boolean
-logic, the Turing machine, and the $\lambda$-calculus, are founded on
+circuits, the Turing machine, and the $\lambda$-calculus, are founded on
 primitives which correspond to \emph{irreversible} physical processes which
@@
 These irreversible abstractions chosen as the basis of our
-models of computing are at ``odds'' with the underlying
+models of computing are at odds with the underlying
 reversible physical reality. The information theoretic underpinnings
-of physics is best understood through the remarkable \emph{Landauer
-  principle}~\cite{Landauer:1961}. Landauer, through his analysis of
+of physics is best understood through \emph{Landauer's
+principle}~\cite{Landauer:1961}. As mentionned previously in the
+introduction, Landauer, through his analysis of
 Maxwell's Demon, established the idea that \emph{information is a
 physical quantity} in much the sense that one would think of an
-electron as a physical quantity, and that a certain minimum amount of
+electron as a physical quantity. More precisely, his principle implies
+that a certain minimum amount of
 thermodynamic work has to be carried out to erase one bit of
@@
 has a physical significance, then it is subject to conservation laws
-as mass and energy are.
+just as mass and energy are.
+
+\subsection{conservation of information}
 
@@
 energy. In other words, the password checking function must have
-erased 1.2 bits during its calculation which dissipate as heat in any
+erased 1.2 bits during its calculation, which dissipate as heat in any
 physical realization of the function. In conventional models of
 computation, this erasure is \emph{implicit}, i.e., it occurs as a
-side effect of calculation. A big part of our development will be
-concerned with making that erasure explicit.
+side effect of calculation. We wish to look at models where
+erasure is simply not allowed.
 
-The discussion in this section suggests that our foundations are really in
-need of revision. There are many principles we can embrace.  We will start
-with one, ``\textbf{conservation of information}'' and follow its
-consequences, which will turn out to be far reaching.
+There are, in fact, many different principles of physics which are
+at odds with our current models of computation. Even within quantum
+mechanics itself, there are other ideas (such as superposition or
+entanglement) which are not present in current models. We will however
+restrict ourselves to exploring a single dimension in which our
+foundations should be revised: \textbf{conservation of information}.
+We will follow its consequences, which will turn out to be far reaching.
 

### 2017-11-24 survey/survey.tex
remove section which is now completely obsolete. Move some outline material earlier, as it fits.

@@
 
-%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conservation of Information}
+
+A physical principle of computation
+
+No creation of information
+No duplication of information
+
+All programs, proofs, deductions are equivalences, isomorphisms,
+reversible
+
+Several pages on reversible logic; reversible family of level 1 and
+level 2 languages (Pi);
+
+We should also have a quick survey of other works in reversible
+computation? Most of the other work tries to start from well known
+models of computation or well known programming languages, and then
+adapts them to be reversible. Pi is different in that it starts from
+the semantic implications of reversibility. Then, because it adds
+types, rather than control-flow (or even data-flow) as its next layer
+of ``understanding'', this leads to equivalences, isomorphisms, etc.
+This then leads, quite naturally, to finding that the ``proof
+language'' of semirings (and Rig Groupoids at level 2) is actually a
+programming language. And it is Pi. This is a neat twist on
+Curry-Howard because CH is about \textbf{inhabitation} only. But with
+``conservation of information'' as the basis, a different kind of
+correspondance arises; in fact, this one may well be an actual
+isomorphism.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Reversibility}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Physics, Computation, and Logic}
-
-Conflict between Church-Turing thesis, textbook quantum mechanics, and
-RSA: Either Shor's algorithm is not ``natural''. (Textbook quantum
-mechanics is wrong); or, Shor's algorithm is ``natural'' and there is
-no classical counterpart. (There are ``natural'' computing models that
-are exponentially faster than the Turing Machine); or, there is an
-efficient classical factoring algorithm. (Possible but surprising); At
-least one of these wild claims is true!!!
-
-\noindent Also conflict between quantum mechanics and logic. Possibilities:
-
-\paragraph*{Revise quantum mechanics.}
-
-The mathematician's vision of an unlimited sequence of totally
-reliable operations is unlikely to be implementable in this real
-universe.  But the real world is unlikely to supply us with unlimited
-memory or unlimited Turing machine tapes. Therefore, continuum
-mathematics is not executable, and physical laws which invoke that can
-not really be satisfactory. They are references to illusionary
-procedures.  (Landauer 1996 and 1999)
-
-% \jc{No access to infinite memory/tape just means that ``infinitary''
-% arguments do not go through. The MoC with reversible computation
-% assumes totally reliable operations (but not necessarily infinitely
-% many). You leap to ``continuum'' mathematics without justification.
-% While QC does use the continuum in non-trivial ways, and physical
-% reality is ``discrete'' at the atomic level (but who knows at the
-% quantum level?!?), I don't think that's where you want to go here.
-% But see below.}
-
-I want to talk about the possibility that there is to be an exact
-simulation, that the computer will do exactly the same as nature. If
-this is to be proved and the type of computer is as I've already
-explained, then it's going to be necessary that everything that
-happens in a finite volume of space and time would have to be exactly
-analyzable with a finite number of logical operations. The present
-theory of physics is not that way, apparently. It allows space to go
-down into infinitesimal distances, wavelengths to get infinitely
-great, terms to be summed in infinite order, and so forth; and
-therefore, if this proposition is right, physical law is wrong.
-(Feynman 1981)
-
-% \jc{In a sense, physics already does this: statistical mechanics.
-% Even though the systems under study are explicitly discrete, albeit
-% will huge numbers of particules, the behaviour is approximated (!!)
-% via a continuous model. Same in economics with Black-Scholes.
-% The real problem are those conclusions drawn from such models which
-% *require* arbitrarily small values to exist, and even worse, for
-% the underlying values (reals) to be connected, i.e. form an actual
-% continuum. What Feynman is asking here is 'finiteness of information',
-% which is what things like Type-2 computability also ask for.}
-
-\paragraph*{Revise Computer Science.}
-
-Another thing that had been suggested early was that natural laws are
-reversible, but that computer rules are not. But this turned out to be
-false; the computer rules can be reversible, and it has been a very,
-very useful thing to notice and to discover that. This is a place
-where the relationship of physics and computation has turned itself
-the other way and told us something about the possibilities of
-computation. So this is an interesting subject because it tells us
-something about computer rules.  (Feynman 1981)
-
-Turing hoped that his abstracted-paper-tape model was so simple, so
-transparent and well defined, that it would not depend on any
-assumptions about physics that could conceivably be falsified, and
-therefore that it could become the basis of an abstract theory of
-computation that was independent of the underlying physics. ``He
-thought, as Feynman once put it, that he understood paper. But he
-was mistaken. Real, quantum-mechanical paper is wildly different from
-the abstract stuff that the Turing machine uses. The Turing machine is
-entirely classical, and does not allow for the possibility the paper
-might have different symbols written on it in different universes, and
-that those might interfere with one another.  (Deutsch 1985)
-
-Ed Fredkin pursued the idea that information must be finite in
-density. One day, he announced that things must be even more simple
-than that. He said that he was going to assume that information itself
-is conserved. ``You're out of you mind, Ed.'' I pronounced. ``That
-completely ridiculous. Nothing could happen in such a world. There
-couldn't even be logical gates. No decisions could ever be made.'' But
-when Fredkin gets one of his ideas, he's quite immune to objections
-like that; indeed, they fuel him with energy. Soon he went on to
-assume that information processing must also be reversible and
-invented what is now called the Fredkin gate. (Minsky 1999)
-
-Even revise the laws of thought themselves: In other terms, what is so
-good in logic that quantum physics should obey? Can't we imagine that
-our conceptions about logic are wrong, so wrong that they are unable
-to cope with the quantum miracle? [\ldots] Instead of teaching logic
-to nature, it is more reasonable to learn from her. Instead of
-interpreting quantum into logic, we shall interpret logic into
-quantum. (Girard 2007)
-
-One must be careful with this argument: we have a ``current
-understanding'' of quantum physics, which could be quite flawed.  So
-what we want to do is to learn from those parts of quantum physics
-which seem to be extremely stable (like conservation of information!)
-first. As researchers, we can certainly speculate on it all, but it
-makes sense to vigorously pursue the conclusions one can draw from
-those aspects of (quantum) physics that seem stable.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conservation of Information}
-
-A physical principle of computation
-
-No creation of information
-No duplication of information
-
-All programs, proofs, deductions are equivalences, isomorphisms,
-reversible
-
-Several pages on reversible logic; reversible family of level 1 and
-level 2 languages (Pi);
-
-We should also have a quick survey of other works in reversible
-computation? Most of the other work tries to start from well known
-models of computation or well known programming languages, and then
-adapts them to be reversible. Pi is different in that it starts from
-the semantic implications of reversibility. Then, because it adds
-types, rather than control-flow (or even data-flow) as its next layer
-of ``understanding'', this leads to equivalences, isomorphisms, etc.
-This then leads, quite naturally, to finding that the ``proof
-language'' of semirings (and Rig Groupoids at level 2) is actually a
-programming language. And it is Pi. This is a neat twist on
-Curry-Howard because CH is about \textbf{inhabitation} only. But with
-``conservation of information'' as the basis, a different kind of
-correspondance arises; in fact, this one may well be an actual
-isomorphism.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions}

### 2017-11-24 survey/survey.tex
keep a bit more. Start on the new story.  Some of the material past the wavefront can be used too, but I have not gotten to it yet.

@@
 computation generally date back to ideas that were thought reasonable
-before \tilde 1900.
+before \textasciitilde 1900.
 
@@
 
-A physical principle of computation
+\begin{quote}
+  The laws of physics are essentially algorithms for calculation. These
+  algorithms are significant only to the extent that they are executable in
+  our real physical world. Our usual laws of physics depend on the
+  mathematician's real number system. With that comes the presumption that
+  given any accuracy requirement, there exist a number of calculations steps,
+  which if executed, will satisfy that accuracy requirement. But the real
+  world is unlikely to supply us with unlimited memory of unlimited Turing
+  machine tapes, Therefore, continuum mathematics is not executable, and
+  physical laws which invoke that can not really be satisfactory. They are
+  references to illusionary procedures. Rolf Landauer, The physical nature of
+  information, Physics Letters A 217 (1996) pp. 188-193:
+\end{quote}
 
-No creation of information
-No duplication of information
+A high level approach to conservation of information would proceed
+as follows:
+\begin{enumerate}
+\item Create a means to measure information.
+\item Show that this measure is sound with respect to our current understanding
+of what information ought to be.
+\item Assert that whatever ``quantity'' of information a system has, must
+remain invariant throughout the lifetime of the system.
+\end{enumerate}
+
+Certainly, whatever our notion of information, it can neither be created,
+duplicated, nor erased. Of course, if we make our notion of information
+too rigid, it will be impossible to even \textbf{modify} it! For computation
+to occur, modification must be possible.  But of what kind?
+
+\jc{wavefront}
+
+Example : $2+5 = 5+2$. Green and Red apples versus apples. What
+can we observe? If we can't observe something, then we call them
+the same (Leibniz).
 
@@
 \subsection{Equalities as Isomorphisms}
-
-\begin{quote}
-  The laws of physics are essentially algorithms for calculation. These
-  algorithms are significant only to the extent that they are executable in
-  our real physical world. Our usual laws of physics depend on the
-  mathematician's real number system. With that comes the presumption that
-  given any accuracy requirement, there exist a number of calculations steps,
-  which if executed, will satisfy that accuracy requirement. But the real
-  world is unlikely to supply us with unlimited memory of unlimited Turing
-  machine tapes, Therefore, continuum mathematics is not executable, and
-  physical laws which invoke that can not really be satisfactory. They are
-  references to illusionary procedures. Rolf Landauer, The physical nature of
-  information, Physics Letters A 217 (1996) pp. 188-193:
-\end{quote}
-
-When a physical law equates two quantities like force and the product of mass
-and acceleration, a deep understanding of the law demands that we interpret
-the equality as an isomorphism, i.e., as two processes, one in each
-direction, that transform one side of the equation to the other. In fact, the
-same insight applies to constructive mathematics and its homotopy
-interpretation. There is a constructive character to equality that is at the
-heart of the constructive approach to mathematics.
 

### 2017-11-28 survey/survey.tex
plan for survey

@@
 \maketitle
+
+% * Reversibility intro / motivation as we  have now more or less
+
+% * Thesis statement: programs are reversible deformations on data /
+%    spaces, almost necessarily by definition
+
+% * Focus now is on data
+
+% * If data is plain finite sets, programs become permutations (deformations on finite sets).
+
+% * If data is structured trees, we get Pi.
+
+% * Explain Pi with examples.
+
+% * If data is itself deformations-on-finite-sets, then the
+%   deformations between them become something quite
+%   interesting. Explain Pi level 2 with examples.
+
+% * Conclude with thoughts regarding other kinds of data that can be
+%   plugged in into that story.
 
@@
 
-In our current age, both Laudauer's Princple and recent investigations
+In our current age, both Landauer's Principle and recent investigations
 of computational models well-suited to quantum mechanics, push us to
@@
 domain of \emph{typed reversible computation}.
-
-\section{Models of Computation}
-%%%
 
@@
 of physics is best understood through \emph{Landauer's
-principle}~\cite{Landauer:1961}. As mentionned previously in the
+principle}~\cite{Landauer:1961}. As mentioned previously in the
 introduction, Landauer, through his analysis of
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conservation of Information}
-
 \begin{quote}
@@
 isomorphism.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Reversibility}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
+\section{Programs as Reversible Deformations}
+
+Starting from a physical perspective, programs naturally emerge as
+reversible transformations on data. In some sense, there is nothing
+more to say about programs per se. The focus is on the data. Depending
+on the space we use to represent the data, the corresponding notion of
+deformation on these spaces will define the appropriate notion of
+programs. In the following three sections, we consider three natural
+classes of data and explore the corresponding notion of reversible
+programs.
+
+In the quantum case, this is clear. Say something about topological
+quantum computing.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data I: Finite Sets}
+
+Most programming languages provide a rich collection of primitive
+data like booleans, characters, strings, and numbers that are
+naturally modeled as sets. Infinite sets embody a notion of recursion
+which is subtle and is discussed briefly in the concluding
+section. For finite sets, a reversible process between them is simply
+a permutation. In other words, reversible programming with finite sets
+is all about writing permutations.
+
+We therefore start with a small language for writing permutations as a
+product of transpositions:
+
+\begin{verbatim}
+p ::= id | swap i j | p;p
+\end{verbatim}
+
+Permutations on finite sets of size $n$ can be inductively defined as
+follows:
+\begin{verbatim}
+perms(0)   = { id }
+perms(n+1) = { (swap 0 i).p | p <- perms(n); i <- [1..n] }  
+\end{verbatim}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data II: Structured Finite Types}
+
+Pi level 1
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data III: Reversible Programs between Structured Finite Types}
+
+Pi level 2
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+Discuss recursion, or add a section of infinite sets???
 

### 2017-11-28 survey/survey.tex
more on structure with some speculative sections

@@
 
+All programs/proofs/deductions are  
+isomorphisms/equivalences/deformations.
+
+Take it as a fundamental principle; avoid detour via irreversible
+functions and current models of computations.
+
 Starting from a physical perspective, programs naturally emerge as
@@
 
+sound and complete isomorphisms for finite types
+
+name the isomorphisms
+
+\begin{verbatim}
+examples: 
+  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
+  conditionals
+  toffoli
+  fredkin
+\end{verbatim}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data IV: Recursive Types}
+
+Add recursion to structured finite types, trace, feedback, loop...
+
+Natural numbers: fold/unfold
+
+Partial isos
+
+meta-circular interpreter
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data V: Exclusive Disjunctions}
+
+???
+
+Quantum over finite field
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
@@
 
+CS abstractions are based on ``old'' physics
+
+computer applications are more and more ``physical''
+
+physical principles such as conservation of information should be part
+of our foundational abstractions
+
+quest for other principles; other data models; graphs; HoTT
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements} many students and colleagues
+
 \bibliographystyle{plainnat}

### 2017-12-06 survey/survey.tex
Remove material that is clearly not in the scope of the paper as currently outlined.
Move material which might still be useful to the end.

Some material which is more for us is clearly marked so.

@@
 
-\jc{wavefront}
-
-Example : $2+5 = 5+2$. Green and Red apples versus apples. What
-can we observe? If we can't observe something, then we call them
-the same (Leibniz).
-
-All programs, proofs, deductions are equivalences, isomorphisms,
-reversible
-
-Several pages on reversible logic; reversible family of level 1 and
-level 2 languages (Pi);
-
-We should also have a quick survey of other works in reversible
+\amr{We should also have a quick survey of other works in reversible
 computation? Most of the other work tries to start from well known
@@
 correspondance arises; in fact, this one may well be an actual
-isomorphism.
+isomorphism.}
 
-\subsection{Equalities as Isomorphisms}
-
-Translated to the computational world, the essence of our thesis is this:
-\emph{to speak of equality is to speak of computation}. This view is partly
-familiar from the study of various systems of equalities, called
-\emph{calculi}, as models of computation. The \lcal, for instance, is a
-system of equalities over a set of syntactic $\lambda$-terms such that they
-result in a model of computation. We distill this idea to its most primitive
-form. We show that there is computational content not just in certain
-specific calculi, such as the \lcal, but even in our most primitive notion of
-equality -- isomorphisms of finite sets.
-
-The equalities we speak of are not \emph{a priori} extensional
-notions, but are intensional specifications. To talk about two things
-being equal, one has to show \emph{how they are equal} wherein this
-\emph{how} is an intensional definition of a translation from
-witnesses of the first to witnesses of the second and vice versa,
-i.e. they are programs. Equivalences that do not have operational
-content do not exist for us.
-
-One may ask, what is the point of such a rarefied spartan view of
-computation? How does this compare with more conventional models of
-computation?  \emph{What is the point?}
-
-\begin{enumerate}
-
-\item The result is computation in its purest form: Programs written
-  in this model are witnesses of equalities. In the process of
-  computing these programs neither gain nor lose information. We get a
-  crisp notion of quantitative information content of programs and
-  show that \emph{computation preserves information}.
-
-\item From the perspective of Quantum Physics, the physical world is
-  one where every fundamental interaction is essentially reversible
-  and various quantities such as energy, mass, angular momentum are
-  conserved. Computation is no longer at odds with
-  physics. Conservation of information and logical reversibility are
-  intrinsic properties of computation as well. Further, various
-  category theoretic models such as dagger symmetric traced monoidal
-  categories studied for Quantum Physics are also the categorical
-  models for computation based on isomorphisms.
-
-\item Irreversibility is a computational effect: By embracing
-  irreversible physical primitives, models of computation have also
-  implicitly included a class of computational effects which we call
-  \emph{information effects}. A consequence of our careful treatment
-  of information, is that we effectively capture the gap between
-  reversible computation and irreversible computation using a
-  type-and-effect system.
-
-\end{enumerate}
-
-What do we mean when we talk about intentional specification of
-equality leading to an information preserving model of computation?
-The equalities we talk of are processes to transform one thing to the
-other. For example, if one were to believe that apples and oranges
-were equal then one would have to show how any given apple can be
-transformed to an orange i.e. witness of apple-hood get transformed to
-witnesses of orange-hood (and vice versa). This is inherently
-different from an extensional notion of equality such as
-\emph{knowledge is power} and have a more operational sense. Equality
-in this sense acts more like the the more general notion of
-\emph{isomorphism} as probably best explained in the paper
-``\emph{When is one thing equal to some other
-  thing?}''~\cite{mazur2008one}.
-
-In our formalism, we are not concerned with fruit but with very simple
-things, namely numbers and sets. When we talk of a set of five apples,
-we are not concerned about the individuality of the apples that
-constitute the set. Rather, each apple in the set has the same status
-as any other apple and we care about the individual apples only to the
-extent that we can tell them apart to count five distinct ones. We
-care about the abstract idea of fiveness and its equality with any
-other abstract idea of fiveness.  When used as types, sets take the
-role of possibility spaces -- we are no longer talking about a set of
-five apples, but talking of one apple out of a possible five. This is
-also where the switch from sets to information happens: When one apple
-out of five is identified, it corresponds to the information to
-discriminate between five choices. Clearly a choice of one out of ten
-presents more information in the sense that it is a more
-discriminating choice. When a computation promises an output of a
-certain type, the resultant value is a witness of one choice out of
-several. This is the sense in which we are concerned with equalities
-of sets.
-
-Programs in our model are built from descriptions of equalities or
-sets/numbers. These are the familiar laws of arithmetic or logic:
-\emph{Commutativity.} The generalized notion of swapping,
-i.e. $a*b=b*a$ and $a+b=b+a$.
-\emph{Associativity.} The
-generalized notion of grouping, i.e. $a*(b*c)=(a*b)*c$ and
-$a+(b+c)=(a+b)+c$.
-\emph{Identity.} Multiplication by 1 and addition with 0,
-  i.e. $a*1=a$ and $a+0=a$.
-\emph{Distributivity.} Multiplication distributes over addition
-  and cancels at zero, i.e. $a*(b+c)=a*b+a*c$ and $a*0=0$.
-\emph{Cancellation.} Cancellation is the idea that equal things
-  on both sides of a equality can be canceled out, i.e. $a+b=a+c$
-  implies $b=c$, and is referred to as a \emph{trace} in category
-  theory~\cite{joyal1996traced}.
-
-How, one may ask, does this form a computational model? Even in very
-simple computational models such as the SK combinatory basis, the
-combinator K provides the basis of choice/conditional by deleting an
-argument and the combinator S provides the basis for iteration by
-duplicating an argument. As we will see, distributivity gives us
-conditionals and trace gives us iteration, albeit in a logically
-reversible manner.  In Chs.~\ref{ch:pi} and~\ref{ch:pi0} we work out
-the details of the computational model, show that it is Turing
-complete and develop programs in it including several simple numeric
-operations, algebraic manipulations of trees and a meta-circular
-interpreter (see Ch.~\ref{ch:interpreters}).
-
-More formally, we develop an information preserving model of
-computation, wherein the process of computing does not gain or lose
-information. Our model arises from a computational interpretation of
-type isomorphisms with iso-recursive types and trace. This can
-equivalently be described as computing a commutative semi-ring with
-cancellation or as a traced dagger symmetric bimonoidal category. The
-category has a groupoid structure and all computations are logically
-reversible.
-
-To summarize, the most primitive notion of equality that is available
-to us -- that of isomorphisms -- already contains the notion of
-computation. The resulting computational model exposes the fine
-structure of how computation handles information. We believe that such
-a model of computing is foundational and we are only beginning to
-understand its applications and full potential. In conclusion, let us
-quote Paul Blain Levy's fine advocacy slogan: Once the fine structure
-has been exposed, why ignore it?
-
-%%%
 \subsection{Information}
 
-We can now describe information preservation in a precise sense.  Let
+We can now give a first precise definition of information preservation.
+Let
 `$b$' be a (not necessarily finite) type whose values are labeled
@@
 ${b_2}^j$. We say a function is \emph{information-preserving} if its
-output entropy is equal to the entropy of its input. See
-Ch.~\ref{ch-metalang} for details.
+output entropy is equal to the entropy of its input. In later sections,
+we will refine this view.
 
-Operations that change the information content of programs have a
-special status. We call them \emph{information effects} and they are
-encapsulated using an arrow meta-language in much the same way that
-one would encapsulate other computational effects in the \lcal\ using a
-monadic metalanguage.  The treatment of information effects is
-analogous to open and closed systems in physics. Closed physical
-systems conserve mass and energy and are the basic unit of study in
-physics. Pure computations that do not have information effects are
-like closed physical systems. They describe equalities.  Open systems
-interact with their environment, possibly exchanging mass or
-energy. These interactions may be thought of as \emph{effects} that
-modify the conservation properties of the system. Computations with
-information effects are much like open systems and they can be
-converted into pure computations by making explicit the surrounding
-information environment that they interact with.
-
-Three things follow: (1) We capture the gap between reversible and
-irreversible computation with a type-and-effect system. (2)
-Categorically speaking, information effects turn monoidal categories
-into ones that have a cartesian structure corresponding to
-conventional computation.  (3) We show how conventional irreversible
-computation such as the \lcal\ can be embedded into this model, such
-that the embedding makes the implicit information effects of \lcal\
-explicit.  As a consequence of this approach, many applications in
-which information manipulation is computationally significant are put
-within the reach of our conceptual model of computation. Such
-applications include quantitative information-flow
-security~\cite{myerssab}, differential
-privacy~\cite{dwork:differential}, energy-aware
-computing~\cite{1324180,605411}, VLSI
-design~\cite{Macii:1996:ECE:874066.875828}, and biochemical models of
-computation~\cite{bio}.
-
-Ch.~\ref{ch:piee} focuses on work in progress. It deals with the
-duality of computation in the setting of information
-preservation. Andrzej Filinski's work on the symmetric
-\lcal\ established the fascinating connection that values are dual to
-continuations, functions are dual to delimited continuations and
-established the duality of call-by-name and call-by-value evaluation
-strategies~\cite{Filinski:1989:DCI:648332.755574}. We explore
-Filinski-style duality in the setting of our information preserving
-model of computing. Here duality does not appear as one De
-Morgan-style dualizing $\neg$ operator, but as two distinct dualizing
-operations -- negation and division. This results in a crisp semantics
-for negative and fractional types and the resulting system has a
-equational formalism in zero-totalized fields or
-meadows~\cite{DBLP:journals/tcs/BergstraHT09}. Computational analogues
-of physical phenomena such as superposition and entanglement appear in
-the resulting semantics.
-
+\amr{
 Talk about conservation of information in biology, computer science,
@@
 \url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
+}
 
 %%%
-\subsection{Technical Overview}
-
-The technical development derives from ideas which are implicit in the
-works of \cite{Toffoli:1980, Zuliani:2001:LR,
-  malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.
-(1) Functions whose output entropy is equal to their input entropy are
-information preserving functions.  (2) A function
-$f : b_1 \rightarrow b_2$ is \emph{logically reversible} if there
-exists an inverse function $f^{\dagger}$ such that for any input
-$v_1:b_1$ if there exists $v_2:b_2$ such that $f(v_1)=v_2$ \emph{iff}
-$f^{\dagger}(v_2)=v_1$.  (3) Logically reversible functions are
-information preserving.
-
-To summarize:
-\begin{enumerate}
-\item Based on isomorphisms of sum and product
-  types~\cite{Fiore:2004}, we develop a strong normalizing model of
-  computing, which we call $\Pi$, that is logically reversible and
-  information preserving. Every computation expressible in $\Pi$
-  is an isomorphism.
-
-\item We extend $\Pi$ with isorecursive type and trace operators from
-  category theory to obtain $\Pi$. Every computation expressible in
-  $\Pi$ is a partial isomorphism --- i.e. the system admits
-  non-termination. $\Pi$ is Turing complete while being information
-  preserving and logically reversible.
-
-\item We develop the categorical semantics of these
-  models~\cite{rc2011}. \amr{Mention n-categories.} We also develop
-  a graphical notation for programming in these models that
-  reminiscent of Penrose diagrams for
-  categories~\cite{selinger-graphical}, Geometry of Interaction (GoI)
-  machines and Proof
-  Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
-  notation, computation is modeled by the flow of particles in a
-  circuit.
-
-\item Since these new models are substantially different from \lcal\
-  or familiar high-level languages, a point of concern is how one can
-  effectively develop programs in them.  We address this by presenting
-  a straightforward technique for deriving $\Pi$ programs by
-  systematically translating logically reversible small-step abstract
-  machines~\cite{isoint}. Complex programs can be devised in this way;
-  we demonstrate the derivation of a meta-circular interpreter for
-  $\Pi$.
-
-\item We develop an arrow meta-language to encapsulate information
-  effects. This metalanguage serves to encapsulate effects, in much
-  the same way that traditional arrows or monads serve to encapsulate
-  effects over the \lcal.
-
-\item We show how a conventional irreversible model can be expressed
-  in our model. We compile the first-order fragment of typed \lcal\
-  extended with sums, products and loops to $\Pi$. This compilation is
-  interesting for two reasons: (1) it exposes the implicit information
-  effects of \lcal\ in the compilation to $\Pi$ and (2) the
-  compilation of $\Pi$ to $\Pi$ shows that information effects can be
-  erased by treating them as interactions with an explicit information
-  environment.
-
-\item We develop the notion of `duality' in $\Pi$ which gives us not
-  one duality (like in linear logic or Filinski's symmetric
-  \lcal~\cite{Filinski:89}), but two notions of duality -- an additive
-  duality and a multiplicative duality. This gives a crisp semantics
-  for negative and fractional types in the context of $\Pi$.
-
-\end{enumerate}
-
-Several intriguing areas of exploration remain and the book concludes
-with discussion of some of these, along with other possible
-applications (see Ch.~\ref{ch:discussion}).  Of these, the field of
-reversible and Quantum computing is closely related and we survey
-several connections in depth.  The exact connection between $\Pi$ and
-linear logic~\cite{Girard87tcs} is not fully understood though it is
-clear that both systems capture related but different notions of
-resource usage. A closely related area is the connection with duality
-of computation \cite{Filinski:89, DBLP:conf/icfp/CurienH00,
-  10.1109/LICS.2010.23,Wadler:2003} --- i.e., the quest for a unifying
-framework between `computations', values, continuations and their
-logical counterparts. The investigation of duality gives rise to
-negative and fractional types which have an intuitive semantics and
-are reminiscent of negative information flow, superposition and
-entanglement from Quantum Physics~\cite{piee}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Below are pieces of text that are not immediately useful above, but
+%% from which we might still be able to pilfer usefully.
+\subsection{Technical Overview}
+
+The technical development derives from ideas which are implicit in the
+works of \cite{Toffoli:1980, Zuliani:2001:LR,
+  malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.
+(1) Functions whose output entropy is equal to their input entropy are
+information preserving functions.  (2) A function
+$f : b_1 \rightarrow b_2$ is \emph{logically reversible} if there
+exists an inverse function $f^{\dagger}$ such that for any input
+$v_1:b_1$ if there exists $v_2:b_2$ such that $f(v_1)=v_2$ \emph{iff}
+$f^{\dagger}(v_2)=v_1$.  (3) Logically reversible functions are
+information preserving.
+
+To summarize:
+\begin{enumerate}
+\item Based on isomorphisms of sum and product
+  types~\cite{Fiore:2004}, we develop a strong normalizing model of
+  computing, which we call $\Pi$, that is logically reversible and
+  information preserving. Every computation expressible in $\Pi$
+  is an isomorphism.
+
+\item We extend $\Pi$ with isorecursive type and trace operators from
+  category theory to obtain $\Pi$. Every computation expressible in
+  $\Pi$ is a partial isomorphism --- i.e. the system admits
+  non-termination. $\Pi$ is Turing complete while being information
+  preserving and logically reversible.
+
+\item We develop the categorical semantics of these
+  models~\cite{rc2011}. \amr{Mention n-categories.} We also develop
+  a graphical notation for programming in these models that
+  reminiscent of Penrose diagrams for
+  categories~\cite{selinger-graphical}, Geometry of Interaction (GoI)
+  machines and Proof
+  Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
+  notation, computation is modeled by the flow of particles in a
+  circuit.
+
+\item Since these new models are substantially different from \lcal\
+  or familiar high-level languages, a point of concern is how one can
+  effectively develop programs in them.  We address this by presenting
+  a straightforward technique for deriving $\Pi$ programs by
+  systematically translating logically reversible small-step abstract
+  machines~\cite{isoint}. Complex programs can be devised in this way;
+  we demonstrate the derivation of a meta-circular interpreter for
+  $\Pi$.
+
+\item We develop an arrow meta-language to encapsulate information
+  effects. This metalanguage serves to encapsulate effects, in much
+  the same way that traditional arrows or monads serve to encapsulate
+  effects over the \lcal.
+
+\item We show how a conventional irreversible model can be expressed
+  in our model. We compile the first-order fragment of typed \lcal\
+  extended with sums, products and loops to $\Pi$. This compilation is
+  interesting for two reasons: (1) it exposes the implicit information
+  effects of \lcal\ in the compilation to $\Pi$ and (2) the
+  compilation of $\Pi$ to $\Pi$ shows that information effects can be
+  erased by treating them as interactions with an explicit information
+  environment.
+
+\item We develop the notion of `duality' in $\Pi$ which gives us not
+  one duality (like in linear logic or Filinski's symmetric
+  \lcal~\cite{Filinski:89}), but two notions of duality -- an additive
+  duality and a multiplicative duality. This gives a crisp semantics
+  for negative and fractional types in the context of $\Pi$.
+
+\end{enumerate}
+
+Several intriguing areas of exploration remain and the book concludes
+with discussion of some of these, along with other possible
+applications (see Ch.~\ref{ch:discussion}).  Of these, the field of
+reversible and Quantum computing is closely related and we survey
+several connections in depth.  The exact connection between $\Pi$ and
+linear logic~\cite{Girard87tcs} is not fully understood though it is
+clear that both systems capture related but different notions of
+resource usage. A closely related area is the connection with duality
+of computation \cite{Filinski:89, DBLP:conf/icfp/CurienH00,
+  10.1109/LICS.2010.23,Wadler:2003} --- i.e., the quest for a unifying
+framework between `computations', values, continuations and their
+logical counterparts. The investigation of duality gives rise to
+negative and fractional types which have an intuitive semantics and
+are reminiscent of negative information flow, superposition and
+entanglement from Quantum Physics~\cite{piee}.

### 2017-12-06 survey/survey.tex
add 3 paragraphs to the 'Programs as Reversible Deformations' section.

@@
 
+To better understand conservation of information, we can look
+for analogous ideas outside of computing. Turning again to physics,
+but this time at the macro scale, what does it mean to transform an
+object in such a way that we do not lose the fundamental character
+of the object?
+
+For rigid objects (like a chair), the only such transformations are
+translations and rotations. But what about something more flexible,
+such as a water balloon?  They can be \textbf{deformed} in various ways,
+but still retain their fundamental character -- as long as we do not
+puncture them or over-stretch them. Ignoring material characteristics
+(i.e. over-stretching), what is special about these deformations, as well
+as for translations and rotations, is that they correspond to
+continuous maps, with a continuous inverse. In fact, even more is
+true: they are analytic maps, with analytic inverses. For our purpose,
+the most important part is that such maps are infinitely differentiable.
+In other words, not only is there an inverse to the deformation, but
+its derivative is also invertible, and so on.
+
+When we look around, we find many different words for related
+concepts: isomorphism, equivalence, sameness, equality, interchangeability,
+comparability, and correspondence, to name a few. Some of these
+are informal concepts, while others have formal mathematical meaning.
+More important, even amongst the formal concepts, there are differences
+-- which is why there are so many of them! Thus we seek a concept
+which is neither too strong nor too weak, that will express when
+some structured information should be treated as ``the same''.
+
+\jc{wavefront}
 All programs/proofs/deductions are  

### 2017-12-06 survey/survey.tex
next few paragraphs.

@@
 
+We will take ``the same'' as a fundamental principle. We want
+to then derive what it means for data, programs, program
+transformations, as well as proofs / deductions, to be ``the
+same'' -- in an manner consistent with preservation of 
+information.
+
+This stands in stark contrast with most current approacheds to
+reversible computation, which start from current models of
+computation involving irreversible operations and try to find
+various ways to \emph{patch things up} so as to be reversible.
+
+The reader will have surely noticed how we have studiously avoided
+defining ``the same''. Because there are many such notions, we
+also need to walk our way through them to find the one which is
+``just right''. We can draw an analogy with topology: in topology,
+all point sets can always be equipped with either the discrete or
+the indiscrete topology, but both of these extremes are rarely
+useful. We will develop our working notion of ``sameness''
+as we go through the various components that make up a
+programming language.
+
+Starting from a physical perspective, whatever our notion of data
+is, we will be interested in programs as representating transformations
+of that data which are reversible. In other words, we want our
+programs-as-transformations to ``play well'' with the inherent notion
+of ``sameness'' that our data will carry. Thus we need to start by
+looking at what structure our data has, and that will help us define
+an appropriate notion of program. Of course, when programs themselves
+are data, things do get more complicated.  In the following sections,
+we will look at different natural classes of data, and explore the
+correspond notion of reversible programs.
+
+\amr{In the quantum case, this is clear. Say something about topological
+quantum computing.}
+
 \jc{wavefront}
-All programs/proofs/deductions are  
-isomorphisms/equivalences/deformations.
-
-Take it as a fundamental principle; avoid detour via irreversible
-functions and current models of computations.
-
-Starting from a physical perspective, programs naturally emerge as
-reversible transformations on data. In some sense, there is nothing
-more to say about programs per se. The focus is on the data. Depending
-on the space we use to represent the data, the corresponding notion of
-deformation on these spaces will define the appropriate notion of
-programs. In the following three sections, we consider three natural
-classes of data and explore the corresponding notion of reversible
-programs.
-
-In the quantum case, this is clear. Say something about topological
-quantum computing.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-12-08 survey/survey.tex
Data I

@@
 \documentclass{article}
+\usepackage{amssymb,amsthm}
 \usepackage{fullpage}
@@
 
+\newtheorem{defn}{Definition}[section]
+
 \newcommand{\amr}[1]{\fbox{Amr says:} \textbf{#1}}
@@
 \newcommand{\lcal}{\ensuremath{\lambda}-calculus}
+
+\newcommand{\fin}[1]{\ensuremath{\left[#1\right]}}
+\newcommand{\Nat}{\ensuremath{\mathbb{N}}}
 
@@
 
-\jc{wavefront}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Most programming languages provide a rich collection of primitive
-data like booleans, characters, strings, and numbers that are
-naturally modeled as sets. Infinite sets embody a notion of recursion
-which is subtle and is discussed briefly in the concluding
-section. For finite sets, a reversible process between them is simply
-a permutation. In other words, reversible programming with finite sets
-is all about writing permutations.
+Before we start our investigation of what it means to have typed,
+information preserving transformations, let's begin with just
+finite sets. Most programming languages provide some collection of primitive
+data like booleans, characters, strings, and (bounded) numbers that are
+naturally modeled as sets. Infinite sets are much more subtle, and will be
+discussed in a later section.
 
-We therefore start with a small language for writing permutations as a
-product of transpositions:
+What does it mean for two finite sets to be ``the same''?  Well, clearly the
+sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$ are different.
+Why?  Well, supposed there was a transformation $f : A \rightarrow B$ that
+deformed $A$ into $B$, and another $g : B \rightarrow A$ which undid this
+transformation. Since $f$ is total, by the pigeonhole principle, two elements
+of $A$ would be mapped to the same element of $B$. Supposed that this is $2$ and
+$3$, and that they both map to $d$.  But $g(d)$ cannot be both $2$ and $3$, and
+so $g$ is not the inverse of $f$. With just a little more work, we can show
+that $f$ (and $g$) must be both injective and surjective. In other words,
+$f$ (and $g$) must be a bijection between $A$ and $B$.  And of course this 
+only happens when $A$ and $B$ have the same number of elements.
 
+More importantly, given a bijection $f : C \rightarrow D$ of finite sets
+$C,D$, there always exists another bijection $g : D \rightarrow C$ which is
+$f$'s inverse. So, for finite sets, \emph{bijections} act as reversible
+deformations.
+
+This discussion is purely ``semantic'', in the sense that it is about
+the denotation of simple primitive data (sets) and their reversible
+deformations (bijections).  We would like to reverse engineer a programming
+language from this denotation. But first, an obvious remark: any two sets
+$C$ and $D$ of cardinality $n$ are always in bijective correspondance. So
+we can abstract away from the details of the elements of $C$ and $D$ and
+instead choose canonical representations -- in much the same way as computers
+choose binary words to represent everything.
+
+\begin{defn} For $n\in\Nat$, denote by $\fin{n}$ the set
+$\left\{0,1,\ldots,n-1\right\}$.
+We will refer to $\fin{n}$ as the canonical set with $n$ elements.
+\end{defn}
+
+Bijections on \fin{n} have a specific name: permutations. As is well-known
+permutations can be generated by sequences of transpositions. Thus we
+can create a small language for writing permutations as
 \begin{verbatim}
@@
 \end{verbatim}
+where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove 
+\verb|id|
+from the language and drop the $i\neq j$ condition so that 
+\verb|swap j j| would represent the identity permutation.
 
-Permutations on finite sets of size $n$ can be inductively defined as
-follows:
-\begin{verbatim}
-perms(0)   = { id }
-perms(n+1) = { (swap 0 i).p | p <- perms(n); i <- [1..n] }  
-\end{verbatim}
+While we are mainly interested in typed data, it is always worthwhile to
+first start investigating the untyped setting first, as it gives us insights
+into what the \emph{purely operational} view of the theory would be.
+This tells us that permutations are an inescapable part of the fabric of
+reversible computing.
+
+However, as permutations are untyped and act on the canonicalized version
+of $n$-element sets (i.e. those sets where all the structure has been
+forgotten), these are a rather pale shadow of the rich tapestry of
+information-perserving transformation of structured data.
 

### 2017-12-08 survey/survey.tex
add lots of comments to Data II, III and IV.

@@
 
-Pi level 1
+\jc{wavefront}
+
+Introduce the usual types. Curry-Howard, as usual. Problem with it, i.e.
+it is merely about inhabitation, which is NOT reversible. Show examples
+with A+A and A*A. Our deformations need to be reversible, so they need
+to induce type isomorphisms. So while our types are as usual, the
+first level of transformations on data will change, and be isomorphisms.
+The type-erasure will, naturally, be permutations. But where do the
+isomorphisms come from?  Well, inhabitants of types with (0,1,+,*)
+form a semiring. So, still inspired by Curry-Howard, look to the
+proof-terms of semiring identities -- and voila, a nice set of
+isomorphisms! We can name these isomorphisms, AND give them
+implementations. Gives us a nice, reversible PL for
+structured finite types. We can it Pi.
+
+We actually have several choices. We can make Pi structurally reversible
+(i.e. there is a syntactic transformation on Pi which maps combinators
+to combinators which witnesses reversibility), or build in a `reverse'
+combinator which flips things around. The latter lets us make the
+language much smaller, but then \emph{programs} in that language are longer.
+[Bias: since there will be many more programs written, the fact that the
+language definition, and its processors, is larger, is a one-time cost.
+Not that big a deal. So pick the version whose meta-theoretical
+properties are obvious. It is thus worthwhile to have reversibility be
+structural.]
 
 sound and complete isomorphisms for finite types
-
-name the isomorphisms
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Data III: Reversible Programs between Structured Finite Types}
+\section{Data III: Reversible Programs between Reversible Programs}
 
-Pi level 2
+Pi level 2.
+
+Take categorification seriously. Types are indeed groupoids, and
+endo-isomorphisms inhabit the Identity 'type' of a type. So + and *
+must be operations on groupoids. Look at the category of (small)
+categories for inspiration. So we have two monoidal structures.
+These are not co-product and product in Groupoid. But this is actually
+a good thing.
+
+So we end up looking, semantically, at weak Rig Groupoid. Syntactically,
+we take the easiest way there: simply make every coherence isomorphism into
+a combinator. Huge proof obligation that this induces actual transformations
+on Pi programs. Luckily, it does.  But not straightforwardly -- ``incorrect''
+definitions of some Pi combinators (such as those for 0*x = 0) can lead to
+some level 2 morphisms not working.
+
+The interpretation on permutations is unclear, because the Pi representation
+has so much more structure, and permutations too have many different 
+representations. So it is unclear what maps to what.
 
@@
 meta-circular interpreter
+
+DANGER: here, we leave the safe land of ``preservation of information'',
+because partiality is not reversible. So while trace gives a reversible
+program as output, it itself isn't a reversible operation. Like the
+demon's brain in Maxwell's thought experiment, trace irreversible work!
+So trace, which is fundamental to all work with recursive types, is
+only reversible "at level 1". So we need to replace trace with some other
+operation for level 2 reversibility to also hold -- and this is an open
+problem.
 

### 2017-12-21 survey/survey.tex
start on Data II writeup

@@
 \newcommand{\Nat}{\ensuremath{\mathbb{N}}}
+\newcommand{\true}{\mathit{true}}
+\newcommand{\false}{\mathit{false}}
 
@@
 
-However, as permutations are untyped and act on the canonicalized version
+However as permutations are untyped, and act on the canonicalized version
 of $n$-element sets (i.e. those sets where all the structure has been
@@
 \section{Data II: Structured Finite Types}
+
+The elementary building blocks of type theory are the empty type
+($\bot$), the unit type ($\top$), and the sum ($\uplus$) and product
+($\times$) types.  These have a nice correspondance (Curry-Howard)
+to logic:
+
+\begin{center}
+\begin{tabular}{c|c}
+Logic & Types \\ \hline
+$\true$ & $\top$ \\
+$\false$ & $\bot$ \\
+$\land$ & $\times$ \\
+$\lor$ & $\uplus$ \\
+\end{tabular}
+\end{center}
+
+This correspondance is rather fruitful. As logical expressions form
+a semiring, we would expect that types too form a semiring. And indeed
+they do -- at least up to \emph{type equivalence}.  The natural
+numbers $\Nat$ are another semiring; it will turn out that, even though
+the Curry-Howard correspondance has been extremely fruitful for
+programming language research, it is the semiring $\Nat$ which
+will be a better model for finite structured types.
 

### 2017-12-21 survey/survey.tex
continue with Data II.

@@
 \newtheorem{defn}{Definition}[section]
+\newtheorem{prop}{Proposition}[section]
 
@@
 This correspondance is rather fruitful. As logical expressions form
-a semiring, we would expect that types too form a semiring. And indeed
-they do -- at least up to \emph{type equivalence}.  The natural
-numbers $\Nat$ are another semiring; it will turn out that, even though
-the Curry-Howard correspondance has been extremely fruitful for
-programming language research, it is the semiring $\Nat$ which
-will be a better model for finite structured types.
+a commutative semiring, we would expect that types too form a commutative
+semiring. And indeed they do -- at least up to \emph{type equivalence}.  The
+natural numbers $\Nat$ are another commutative semiring; it will turn out that,
+even though the Curry-Howard correspondance has been extremely fruitful for
+programming language research, it is $\Nat$ which will be a better model for
+finite structured types.
+
+\begin{definition}
+  A \emph{commutative semiring} (sometimes called a \emph{commutative
+    rig} (commutative ri\emph{n}g without negative elements) consists of a
+  set $R$, two distinguished elements of $R$ named 0 and 1, and two
+  binary operations~$+$ and $\cdot$, satisfying the following
+  relations for any $a,b,c \in R$:
+\[\begin{array}{rcl}
+0 + a &=& a \\
+a + b &=& b + a \\
+a + (b + c) &=& (a + b) + c \\
+\\
+1 \cdot a &=& a \\
+a \cdot b &=& b \cdot a \\
+a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
+\\
+0 \cdot a &=& 0 \\
+(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
+\end{array}\]
+\end{definition}
+
+\begin{prop}
+The structure $\left(\mathbb{B}, \false, \true, \lor, \land\right)$
+is a commutative semiring.
+\end{prop}
+
+We cannot immediately use the above definition in a typed setting.
+First, types do not naturally want to be put together into a ``set''.
+This can be fixed if we replace the set $R$ with a universe $U$, and
+rather than insisting that we have $0 \in R$, we have a typing
+judgement $\bot : U$ (and similarly for the other items).
+
+Our first instinct would be to similarly replace $=$ with
+$\equiv$, i.e. propositional equality. This bring the second
+problem to the fore: this is false. For example, unless $A \equiv B$,
+$A \times B \equiv B \times A$ is not normally provable%
+\footnote{Not without assuming that $U$ is univalent, which we
+don't.}. But it should be clear that $A \times B$ and $B \times A$
+contain equivalent information. In other words, we would like to
+be able to witness that $A \times B$ can be reversibly deformed
+into $B \times A$. Which motivates the introduction of 
+type \emph{equivalences}. To do this, we need a few important
+auxilliary concepts first.
+
+\begin{definition}[homotopy]
+\label{def:homotopy}
+Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
+$\forall x:A. f(x) = g(x)$. We denote this $f \sim g$.
+\end{definition}
+
+\noindent It is easy to prove that homotopies (for any given function
+space $A \rightarrow B$) are an equivalence relation.
+
+\begin{definition}[Quasi-inverse]
+\label{def:quasi}
+For a function $f : A \rightarrow B$, a \emph{quasi-inverse} is a
+triple $(g, \alpha, \beta)$, consisting of a function
+$g : B \rightarrow A$ and two homotopies
+$\alpha : f \circ g \sim \mathrm{id}_B$ and
+$\beta : g \circ f \sim \mathrm{id}_A$.
+\end{definition}
+
+\begin{definition}[Equivalence of types]
+  Two types $A$ and $B$ are equivalent $A ≃ B$ if there exists a
+  function $f : A \rightarrow B$ together with a quasi-inverse for $f$.
+\end{definition}
+
+\jc{talk about the definition without quasi-inverse? Continue with
+types being a commutative semiring in the above sense.}
 

### 2017-12-21 survey/survey.tex
Big push on Data II.

@@
 \documentclass{article}
-\usepackage{amssymb,amsthm}
+\usepackage{amssymb,amsthm,amsmath}
 \usepackage{fullpage}
@@
 
-\begin{definition}
+\begin{defn}
   A \emph{commutative semiring} (sometimes called a \emph{commutative
@@
 \end{array}\]
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition}[homotopy]
+\begin{defn}[homotopy]
 \label{def:homotopy}
@@
 $\forall x:A. f(x) = g(x)$. We denote this $f \sim g$.
-\end{definition}
+\end{defn}
 
 \noindent It is easy to prove that homotopies (for any given function
-space $A \rightarrow B$) are an equivalence relation.
+space $A \rightarrow B$) are an equivalence relation.  The simplest
+definition of the data which makes up an equivalence is then
 
-\begin{definition}[Quasi-inverse]
+\begin{defn}[Quasi-inverse]
 \label{def:quasi}
@@
 $\beta : g \circ f \sim \mathrm{id}_A$.
-\end{definition}
+\end{defn}
 
-\begin{definition}[Equivalence of types]
-  Two types $A$ and $B$ are equivalent $A ≃ B$ if there exists a
+\begin{defn}[Equivalence of types]
+  Two types $A$ and $B$ are equivalent $A \simeq B$ if there exists a
   function $f : A \rightarrow B$ together with a quasi-inverse for $f$.
-\end{definition}
+\end{defn}
 
-\jc{talk about the definition without quasi-inverse? Continue with
-types being a commutative semiring in the above sense.}
+\noindent Why \emph{quasi}? The reasons are beyond our scope, but
+the interested reader can read sections $2.4$ and $4$ in \cite{hott-book}.
+There are several conceptually different, but equivalent, 
+``better'' definitions.  We record just one here:
+
+\begin{defn}[Bi-invertibility]
+\label{def:biinv}
+For a function $f : A \rightarrow B$, a \emph{bi-inverse} is a
+pair of functions $g,h : B \rightarrow A$ and two homotopies
+$\alpha : f \circ g \sim \mathrm{id}_B$ and
+$\beta : h \circ f \sim \mathrm{id}_A$.
+\end{defn}
+
+\noindent We can then replace quasi-inverse with bi-invertibility in
+the definition of type equivalence. The differences will not matter to
+us here.
+
+If we replace $=$ by $\simeq$, then we can indeed prove that
+types (with $\bot, \top, \uplus, \times$) form a semiring.
+
+The reader familiar with universal algebra should pause
+and ponder a bit about what we have done. We have lifted 
+\emph{equality} from being in the signature of the ambient logic
+and instead put it in the signature of the algebraic structure
+of interest. Then, and only then, are we in a solid position
+to show that we have a concept (semiring) which is flexible 
+enough for types to be an instance.
+
+If we revisit the Curry-Howard correspondance, we notice one
+more issue. In logic, it is true that $A \lor A = A$ and
+$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
+equivalent to $A$. They are however \emph{equi-inhabited}. This is
+a fancy way of saying
+\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
+  \qquad A \ \text{is inhabited} \]
+The above is the real \textit{essence} of the correspondance.
+
+But mere inhabitation falls far short of our goals of being
+able to smoothly deform from one type to another.
+
+Let's thus analyze the crux of the ``problem''. In logic, we have
+that $\land$ and $\lor$ are both \emph{idempotent}: this is the
+property of any binary operation $*$ where
+$\forall a. a * a = a$. And it should be clear that idempotent
+operations are \emph{forgetful}: its input has two copies of $a$,
+but its output, only one. On the type side, something more subtle
+happens. Consider $\top \uplus \top$ versus $\top$; the first has
+\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
+while the second only one ($\texttt{tt}$). These cannot be put in
+bijective correspondance. Even though the ``payload'' \texttt{tt}
+is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
+away information -- something we have expressly disallowed.
+Yes, this should remind you of Maxwell's demon: even though the
+data is the same, they are tagged differently, and this
+information is indeed information, and must be preserved.
 

### 2017-12-22 survey/survey.tex
started data III

@@
 \usepackage{comment}
+\usepackage{tikz}
+\usepackage{tikz-cd}
+\usetikzlibrary{quotes}
 
@@
 we will look at different natural classes of data, and explore the
-correspond notion of reversible programs.
+corresponding notion of reversible programs.
 
@@
 
+In the previous sections, we examined equivalences between
+conventional data structures, i.e, sets of values and structured trees
+of values. We now consider a richer, less conventional, but
+foundational notion of data: programs themselves. Indeed, universal
+computation models fundamentally rely on the fact that ``programs are
+(or can be encoded as) data'', e.g., a Turing machine can be encoded
+as a string that another Turing machine (or even the same machine) can
+manipulate. In our setting, the programs developed in the previous
+sections are reversible deformations of data structures. Can these
+reversible deformations be themselves be subject to (higher-level)
+reversible deformations?
+
+Before developing the theory, let's consider a small example:
+
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+  \draw[<->,double,red,thick] (2.25,-1.5) -- (2.25,-2.5) ;
+  \node at (3.3,-2) {$\mathit{swapl₊⇔}$} ;
+  \node at (2.5,-1.3) {$((c₂~⊕~c₁)~◎~\mathit{swap₊})$};
+  \node at (2.5,-2.7) {$(\mathit{swap₊}~◎~(c₁~⊕~c₂))$};
+  \draw (-2,-2) ellipse (0.5cm and 1cm);
+  \draw[fill] (-2,-1.5) circle [radius=0.025];
+  \node[below] at (-2.1,-1.5) {$A$};
+  \draw[fill] (-2,-2.5) circle [radius=0.025];
+  \node[below] at (-2.1,-2.5) {$B$};
+
+  \draw (6.5,-2) ellipse (0.5cm and 1cm);
+  \draw[fill] (6.5,-1.5) circle [radius=0.025];
+  \node[below] at (6.7,-1.5) {$C$};
+  \draw[fill] (6.5,-2.5) circle [radius=0.025];
+  \node[below] at (6.7,-2.5) {$D$};
+
+  \draw (-2,-1.5) to[bend left] (1,0.5) ;
+  \draw (-2,-2.5) to[bend left] (1,-0.5) ;
+  \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
+  \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
+
+  \draw (-2,-1.5) to[bend right] (1,-3.5) ;
+  \draw (-2,-2.5) to[bend right] (1,-4.5) ;
+  \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
+  \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
+
+  \draw     (2.5,-3)  -- (3.5,-3) -- (3.5,-4) -- (2.5,-4) -- cycle ;
+  \draw     (2.5,-4)  -- (3.5,-4) -- (3.5,-5) -- (2.5,-5) -- cycle ;
+
+  \draw     (1,1)  -- (2,1) -- (2,0) -- (1,0) -- cycle ;
+  \draw     (1,0)  -- (2,0) -- (2,-1) -- (1,-1) -- cycle ;
+
+  \node at (3,-3.5) {c₁};
+  \node at (3,-4.5) {c₂};
+
+  \node at (1.5,0.5) {c₂};
+  \node at (1.5,-0.5) {c₁};
+
+  \draw     (2,0.5)  -- (2.5,0.5)  ;
+  \draw     (2,-0.5) -- (2.5,-0.5) ;
+
+  \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
+  \draw     (2.5,-0.5) -- (3.5,0.5) ;
+
+  \draw     (1,-3.5)  -- (2,-4.5)    ;
+  \draw     (1,-4.5) -- (2,-3.5)   ;
+
+  \draw     (2,-3.5)  -- (2.5,-3.5)    ; 
+  \draw     (2,-4.5) -- (2.5,-4.5)   ;
+
+\end{tikzpicture}
+\end{center}
+
+\amr{another wavefront}
+
 Pi level 2.

### 2017-12-22 survey/survey.tex
Data II. Pi finally mostly defined. And justified. Just need to do the same for the combinators and comment on the choices. Probably add some examples too.

@@
 \usetikzlibrary{quotes}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Macros
+
+\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
+\newcommand{\mm}{\texttt{\textminus}}
+\newcommand{\pp}{\texttt{+}}
+\newcommand{\inl}[1]{\textsf{inl}~#1}
+\newcommand{\inr}[1]{\textsf{inr}~#1}
+\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
+\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
+\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
+\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
+\newcommand{\refl}[1]{\textsf{refl}~#1}
+%% \newcommand{\lid}{\textsf{lid}}
+\newcommand{\alt}{~|~}
+%% \newcommand{\rid}{\textsf{rid}}
+\newcommand{\linv}{l!}
+\newcommand{\rinv}{r!}
+\newcommand{\invinv}{!!}
+\newcommand{\assoc}{\circ}
+\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
+\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
+\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
+\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
+\newcommand{\swapp}{\mathit{swap}_+}
+\newcommand{\assoclp}{\mathit{assocl}_+}
+\newcommand{\assocrp}{\mathit{assocr}_+}
+\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
+\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
+\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
+\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
+\newcommand{\swapt}{\mathit{swap}_*}
+\newcommand{\assoclt}{\mathit{assocl}_*}
+\newcommand{\assocrt}{\mathit{assocr}_*}
+\newcommand{\absorbr}{\mathit{absorbr}}
+\newcommand{\absorbl}{\mathit{absorbl}}
+\newcommand{\factorzr}{\mathit{factorzr}}
+\newcommand{\factorzl}{\mathit{factorzl}}
+\newcommand{\dist}{\mathit{dist}}
+\newcommand{\factor}{\mathit{factor}}
+\newcommand{\distl}{\mathit{distl}}
+\newcommand{\factorl}{\mathit{factorl}}
+\newcommand{\distz}{\mathit{absorbr}}
+\newcommand{\iso}{\leftrightarrow}
+\newcommand{\proves}{\vdash}
+\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
+\newcommand{\Rule}[4]{
+\makebox{{\rm #1}
+$\displaystyle
+\frac{\begin{array}{l}#2 \\\end{array}}
+{\begin{array}{l}#3      \\\end{array}}$
+ #4}}
+\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
 
@@
   relations for any $a,b,c \in R$:
-\[\begin{array}{rcl}
+\begin{equation}
+\begin{array}{rcl}
 0 + a &=& a \\
@@
 (a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
-\end{array}\]
+\end{array}
+\label{eq:csemiring}
+\end{equation}
 \end{defn}
@@
 If we replace $=$ by $\simeq$, then we can indeed prove that
-types (with $\bot, \top, \uplus, \times$) form a semiring.
+types (with $\bot, \top, \uplus, \times$) form a commutative semiring.
 
@@
 of interest. Then, and only then, are we in a solid position
-to show that we have a concept (semiring) which is flexible 
+to show that we have a concept (commutative semiring) which is flexible 
 enough for types to be an instance.
@@
 The above is the real \textit{essence} of the correspondance.
-
-But mere inhabitation falls far short of our goals of being
+In other words, classical Curry-Howard tells us about
+\emph{logical equivalence} of types. This is even a
+constructive statement: there are indeed functions
+$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
+however, they are not inverses.
+
+So mere inhabitation falls far short of our goals of being
 able to smoothly deform from one type to another.
 
-Let's thus analyze the crux of the ``problem''. In logic, we have
+Let us thus analyze the crux of the ``problem''. In logic, we have
 that $\land$ and $\lor$ are both \emph{idempotent}: this is the
@@
 data is the same, they are tagged differently, and this
-information is indeed information, and must be preserved.
+data is indeed information, and must be preserved.
+
+\begin{figure}[t]
+\[
+\begin{array}{rrcll}
+\idc :& \tau & \iso & \tau &: \idc \\
+\\
+\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
+\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
+\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
+\\
+\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
+\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
+\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
+\\
+\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
+\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
+\end{array}
+\]
+\caption{$\Pi$-terms.}
+\label{pi-terms}
+\end{figure}
+
+Nevertheless, the Curry-Howard correspondance still has some
+force. We know that the inhabitants of types formed with
+with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
+want to know is, which types are equivalent? From a commutative semiring
+perspective, this amounts to asking what terms are equal?
+We have a set of generators for those equations, namely those
+in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
+of mutually inverse functions which witness these identities.
+For concreteness, we show the signatures in Figure~\ref{pi-terms}.
+
+\begin{figure}[t]
+\[
+\Rule{}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
+{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
+{}
+\qquad
+\Rule{}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
+{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
+{}
+\]
+\[
+\Rule{}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
+{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
+{}
+\]
+\caption{$\Pi$-combinators.}
+\label{pi-combinators}
+\end{figure}
 
@@
 
-Introduce the usual types. Curry-Howard, as usual. Problem with it, i.e.
-it is merely about inhabitation, which is NOT reversible. Show examples
-with A+A and A*A. Our deformations need to be reversible, so they need
-to induce type isomorphisms. So while our types are as usual, the
-first level of transformations on data will change, and be isomorphisms.
 The type-erasure will, naturally, be permutations. But where do the
@@
 isomorphisms! We can name these isomorphisms, AND give them
-implementations. Gives us a nice, reversible PL for
-structured finite types. We can it Pi.
+implementations.
+
+That gives us some ground terms.  Not enough for a PL! So then
+we need to be able to 'combine' these. So we add 3 combinators for
+3 different kinds of composition.
+Gives us a nice, reversible PL for
+structured finite types. We call it Pi.
 

### 2017-12-22 survey/survey.tex
use the same capitalization for the same ref.

@@
 investigations, in the intersection between mathematics and
-computer science (particularly in Homotopy Type Theory~\cite{HoTT-book})
+computer science (particularly in Homotopy Type Theory~\cite{hott-book})
 reveal that not only can mathematics be typed, but that doing so

### 2017-12-22 survey/survey.tex
More in Data II. An explanation that satisfied me for Pi seems to take more space than expected.  Ah well, it is coming along.

@@
 \begin{array}{rrcll}
+\idc :& \tau & \simeq & \tau &: \idc \\
+\\
+\identlp :&  \bot \uplus \tau & \simeq & \tau &: \identrp \\
+\swapp :&  \tau_1 \uplus \tau_2 & \simeq & \tau_2 \uplus \tau_1 &: \swapp \\
+\assoclp :&  \tau_1 \uplus (\tau_2 \uplus \tau_3) & \simeq & (\tau_1 \uplus \tau_2) \uplus \tau_3 &: \assocrp \\
+\\
+\identlt :&  \top \times \tau & \simeq & \tau &: \identrt \\
+\swapt :&  \tau_1 \times \tau_2 & \simeq & \tau_2 \times \tau_1 &: \swapt \\
+\assoclt :&  \tau_1 \times (\tau_2 \times \tau_3) & \simeq & (\tau_1 \times \tau_2) \times \tau_3 &: \assocrt \\
+\\
+\distz :&~ \bot \times \tau & \simeq & \bot ~ &: \factorzl \\
+\dist :&~ (\tau_1 \uplus \tau_2) \times \tau_3 & \simeq & (\tau_1 \times \tau_3) \uplus (\tau_2 \times \tau_3)~ &: \factor
+\end{array}
+\]
+\caption{Type isomorphisms.}
+\label{type-isos}
+\end{figure}
+
+Nevertheless, the Curry-Howard correspondance still has some
+force. We know that the inhabitants of types formed with
+with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
+want to know is, which types are equivalent? From a commutative semiring
+perspective, this amounts to asking what terms are equal?
+We have a set of generators for those equations, namely those
+in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
+of mutually inverse functions which witness these identities.
+For concreteness, we show the signatures in Figure~\ref{type-isos}.
+As these all come in symmetric pairs (some of which are self-symmetric),
+we give names for both directions. The names use $\iso$, $+$ and $*$
+as these will be what we use when we reflect these isomorphism into
+a programming language, which we proceed to do next.
+
+\begin{figure}[t]
+\[
+\begin{array}{rrcll}
 \idc :& \tau & \iso & \tau &: \idc \\
@@
 
-Nevertheless, the Curry-Howard correspondance still has some
-force. We know that the inhabitants of types formed with
-with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
-want to know is, which types are equivalent? From a commutative semiring
-perspective, this amounts to asking what terms are equal?
-We have a set of generators for those equations, namely those
-in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
-of mutually inverse functions which witness these identities.
-For concreteness, we show the signatures in Figure~\ref{pi-terms}.
+We now have in our hands our desired denotational semantics: we want
+to create a language $\Pi$ such that the types and type combinators
+map to $\bot, \top, \uplus, \times$, and such that we have ground
+terms whose denotation are all $16$ type isormorphisms of
+Figure~\ref{type-isos}. This is completely straightforward, as
+we can simply do this literally. To make the analogy with 
+commutative semirings stand out even more, we will use $0, 1, +$, and $*$
+at the type level, and will denote ``equivalence'' by $\iso$.
+Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.
+
+\jc{wavefront}
+
+Of course, one does not get a programming language with just (typed)
+constants!
 
@@
 
-\jc{wavefront}
-
 The type-erasure will, naturally, be permutations. But where do the

### 2017-12-24 survey/survey.tex
more on data III

@@
 \usetikzlibrary{quotes}
+\tikzset{>=latex}
 
@@
 foundational notion of data: programs themselves. Indeed, universal
-computation models fundamentally rely on the fact that ``programs are
-(or can be encoded as) data'', e.g., a Turing machine can be encoded
-as a string that another Turing machine (or even the same machine) can
-manipulate. In our setting, the programs developed in the previous
-sections are reversible deformations of data structures. Can these
-reversible deformations be themselves be subject to (higher-level)
-reversible deformations?
-
-Before developing the theory, let's consider a small example:
+computation models fundamentally rely on the fact that \emph{programs
+  are (or can be encoded as) data}, e.g., a Turing machine can be
+encoded as a string that another Turing machine (or even the same
+machine) can manipulate. In our setting, the programs developed in the
+previous section are reversible deformations between structured finite
+types. We now ask whether these reversible deformations can themselves
+be subject to (higher-level) reversible deformations?
+
+Before developing the theory, let's consider a small example
+consisting of two deformations between the types $A + B$ and $C+D$:
 
@@
 \begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-  \draw[<->,double,red,thick] (2.25,-1.5) -- (2.25,-2.5) ;
-  \node at (3.3,-2) {$\mathit{swapl₊⇔}$} ;
-  \node at (2.5,-1.3) {$((c₂~⊕~c₁)~◎~\mathit{swap₊})$};
-  \node at (2.5,-2.7) {$(\mathit{swap₊}~◎~(c₁~⊕~c₂))$};
+  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
+%%  \node at (3.3,-2) {$\mathit{swapl}_{+\Leftrightarrow}$} ;
+%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+})$}; 
+%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\circledcirc~(c_1~\oplus~c_2))$}; 
   \draw (-2,-2) ellipse (0.5cm and 1cm);
@@
 
-  \draw (-2,-1.5) to[bend left] (1,0.5) ;
-  \draw (-2,-2.5) to[bend left] (1,-0.5) ;
+  \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
+  \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
   \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
@@
 
-  \draw (-2,-1.5) to[bend right] (1,-3.5) ;
-  \draw (-2,-2.5) to[bend right] (1,-4.5) ;
+  \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
+  \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
   \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
@@
 
-  \node at (3,-3.5) {c₁};
-  \node at (3,-4.5) {c₂};
-
-  \node at (1.5,0.5) {c₂};
-  \node at (1.5,-0.5) {c₁};
+  \node at (3,-3.5) {$c_1$};
+  \node at (3,-4.5) {$c_2$};
+
+  \node at (1.5,0.5) {$c_2$};
+  \node at (1.5,-0.5) {$c_1$};
 
@@
 \end{center}
+The top path is the reversible deformation
+$(c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+}$ which deforms the
+type $A$ by $c_2$, deforms the type $B$ by $c_1$, and deforms the
+resulting space by a twist that exchanges the two injections into the
+sum type. The bottom path performs the twist first and then deforms
+the type $A$ by $c_2$ and the type $B$ by $c_1$ as before. If one
+could imagine the paths are physical wires and the deformations $c_1$
+and $c_2$ as arbitrary deformations on these wires then, holding the
+points $A$, $B$, $C$, and $D$ fixed, it is possible to rotate the top
+part of the diagram to become identical to the bottom one. In other
+words, there exists a higher-level deformation of the program
+$(c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+}$ to the program
+$\mathit{swap}_{+} \circledcirc (c_1~\oplus~c_2)$.
 
 \amr{another wavefront}
+
+\begin{figure}[t]
+Let $c₁ : t₁ ⟷ t₂$,  $c₂ : t₂ ⟷ t₃$, and $c₃ : t₃ ⟷ t₄$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {c₁ ◎ (c₂ ◎ c₃) ⇔ (c₁ ◎ c₂) ◎ c₃}
+\\
+  {(c₁ ⊕ (c₂ ⊕ c₃)) ◎ \assoclp ⇔ \assoclp ◎ ((c₁ ⊕ c₂) ⊕ c₃)}
+\\
+  {(c₁ ⊗ (c₂ ⊗ c₃)) ◎ \assoclt ⇔ \assoclt ◎ ((c₁ ⊗ c₂) ⊗ c₃)}
+\\
+  {((c₁ ⊕ c₂) ⊕ c₃) ◎ \assocrp ⇔ \assocrp ◎ (c₁ ⊕ (c₂ ⊕ c₃))}
+\\
+  {((c₁ ⊗ c₂) ⊗ c₃) ◎ \assocrt ⇔ \assocrt ◎ (c₁ ⊗ (c₂ ⊗ c₃))}
+\\
+  {\assocrp ◎ \assocrp ⇔ ((\assocrp ⊕ \idc) ◎ \assocrp) ◎ (\idc ⊕ \assocrp)}
+\\
+  {\assocrt ◎ \assocrt ⇔ ((\assocrt ⊗ \idc) ◎ \assocrt) ◎ (\idc ⊗ \assocrt)}
+\end{array}\]
+\caption{\label{figj}Signatures of level-1 $\Pi$-combinators: associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $a : t₁ ⟷ t₂$, $b : t₃ ⟷ t₄$, and $c : t₅ ⟷ t₆$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((a ⊕ b) ⊗ c) ◎ \dist ⇔ \dist ◎ ((a ⊗ c) ⊕ (b ⊗ c))}
+\\
+  {(a ⊗ (b ⊕ c)) ◎ \distl ⇔ \distl ◎ ((a ⊗ b) ⊕ (a ⊗ c))}
+\\
+  {((a ⊗ c) ⊕ (b ⊗ c)) ◎ \factor ⇔ \factor ◎ ((a ⊕ b) ⊗ c)}
+\\
+  {((a ⊗ b) ⊕ (a ⊗ c)) ◎ \factorl ⇔ \factorl ◎ (a ⊗ (b ⊕ c))}
+\end{array}\]
+\caption{\label{figi}Signatures of level-1 $\Pi$-combinators: distributivity and factoring}
+\end{figure}
+
+\begin{figure}[t]
+Let $c, c₁, c₂, c₃ : t₁ ⟷ t₂$ and $c', c'' : t₃ ⟷ t₄$: 
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\idc ◎ \, c ⇔ c}
+\quad 
+  {c \, ◎ \idc \, ⇔ c}
+\quad
+  {c\,\, ◎\,! c ⇔ \idc}
+\quad 
+  {! c ◎ c ⇔ \idc}
+\\
+  {c ⇔ c}
+\quad 
+\Rule{}
+  {c₁ ⇔ c₂ \quad c₂ ⇔ c₃}
+  {c₁ ⇔ c₃}
+  {} 
+\quad
+\Rule{}
+  {c₁ ⇔ c' \quad c₂ ⇔ c''}
+  {c₁ ◎ c₂ ⇔ c' ◎ c''}
+  {}
+\end{array}\]
+\caption{\label{figh}Signatures of level-1 $\Pi$-combinators: identity and composition}
+\end{figure}
+
+\begin{figure}[t]
+Let $c₀ : 0 ⟷ 0$, $c₁ : 1 ⟷ 1$, and $c : t₁ ⟷ t₂$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlp ◎ c ⇔ (c₀ ⊕ c) ◎ \identlp}
+\qquad 
+  {\identrp ◎ (c₀ ⊕ c) ⇔ c ◎ \identrp}
+\\
+  {\identlsp ◎ c ⇔ (c ⊕ c₀) ◎ \identlsp}
+\qquad
+  {\identrsp ◎ (c ⊕ c₀) ⇔ c ◎ \identrsp}
+\\
+  {\identlt ◎ c ⇔ (c₁ ⊗ c) ◎ \identlt}
+\qquad
+  {\identrt ◎ (c₁ ⊗ c) ⇔ c ◎ \identrp}
+\\
+  {\identlst ◎ c ⇔ (c ⊗ c₁) ◎ \identlst}
+\qquad
+  {\identrst ◎ (c ⊗ c₁) ⇔ c ◎ \identrst}
+\\
+  {\identlt ⇔ \distl ◎ (\identlt ⊕ \identlt)}
+\\
+\identlp ⇔ \swapp ◎ \identrsp
+\qquad
+\identlt ⇔ \swapt ◎ \identrst
+\end{array}\]
+\caption{\label{figg}Signatures of level-1 $\Pi$-combinators: unit}
+\end{figure}
+
+\begin{figure}[t]
+Let $c₁ : t₁ ⟷ t₂$ and $c₂ : t₃ ⟷ t₄$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\swapp ◎ (c₁ ⊕ c₂) ⇔ (c₂ ⊕ c₁) ◎ \swapp}
+\quad
+  {\swapt ◎ (c₁ ⊗ c₂) ⇔ (c₂ ⊗ c₁) ◎ \swapt}
+\\
+  {(\assocrp ◎ \swapp) ◎ \assocrp ⇔ ((\swapp ⊕ \idc) ◎ \assocrp) ◎ (\idc ⊕ \swapp)}
+\\
+  {(\assoclp ◎ \swapp) ◎ \assoclp ⇔ ((\idc ⊕ \swapp) ◎ \assoclp) ◎ (\swapp ⊕ \idc)}
+\\
+  {(\assocrt ◎ \swapt) ◎ \assocrt ⇔ ((\swapt ⊗ \idc) ◎ \assocrt) ◎ (\idc ⊗ \swapt)}
+\\
+  {(\assoclt ◎ \swapt) ◎ \assoclt ⇔ ((\idc ⊗ \swapt) ◎ \assoclt) ◎ (\swapt ⊗ \idc)}
+\end{array}\]
+\caption{\label{figf}Signatures of level-1 $\Pi$-combinators: commutativity and associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $c₁ : t₁ ⟷ t₂$, $c₂ : t₃ ⟷ t₄$, $c₃ : t₁ ⟷ t₂$, and $c₄ : t₃ ⟷ t₄$. \\
+Let $a₁ : t₅ ⟷ t₁$,  $a₂ : t₆ ⟷ t₂$, $a₃ : t₁ ⟷ t₃$, and $a₄ : t₂ ⟷ t₄$.
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+\Rule{}
+  {c₁ ⇔ c₃ \quad c₂ ⇔ c₄}
+  {c₁ ⊕ c₂ ⇔ c₃ ⊕ c₄}
+  {}
+\qquad
+\Rule{}
+  {c₁ ⇔ c₃ \quad c₂ ⇔ c₄}
+  {c₁ ⊗ c₂ ⇔ c₃ ⊗ c₄}
+  {} 
+\\
+  {\idc ⊕ \, \idc \, ⇔ \idc}
+\qquad
+  {\idc ⊗ \, \idc \, ⇔ \idc}
+\\
+  {(a₁ ◎ a₃) ⊕ (a₂ ◎ a₄) ⇔ (a₁ ⊕ a₂) ◎ (a₃ ⊕ a₄)}
+\\
+  {(a₁ ◎ a₃) ⊗ (a₂ ◎ a₄) ⇔ (a₁ ⊗ a₂) ◎ (a₃ ⊗ a₄)}
+\end{array}\]
+\caption{\label{fige}Signatures of level-1 $\Pi$-combinators: functors}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlsp ⊕ \idc ~⇔~ \assocrp ◎ (\idc ⊕ \, \identlp)}
+\\
+  {\identlst ⊗ \idc ~⇔~ \assocrt ◎ (\idc ⊗ \, \identlt)}
+\end{array}\]
+\caption{\label{figd}Signatures of level-1 $\Pi$-combinators: unit and associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $c : t₁ ⟷ t₂$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {(c ⊗ \idc) ◎ \absorbl ⇔ \absorbl ◎ \idc}
+\quad
+  {(\idc \, ⊗ c) ◎ \absorbr ⇔ \absorbr ◎ \idc}
+\\
+  {\idc ◎ \, \factorzl ⇔ \factorzl ◎ (\idc ⊗ c)}
+\quad
+  {\idc ◎ \, \factorzr ⇔ \factorzr ◎ (c ⊗ \idc)}
+\\
+  {\absorbr ⇔ \absorbl}
+\\
+  {\absorbr ⇔ (\distl ◎ (\absorbr ⊕ \absorbr)) ◎ \identlp}
+\\
+  {\identlst ⇔ \absorbr}
+\qquad
+  {\absorbl ⇔ \swapt ◎ \absorbr}
+\\
+  {\absorbr ⇔ (\assoclt ◎ (\absorbr ⊗ \idc)) ◎ \absorbr}
+\\
+  {(\idc ⊗ \absorbr) ◎ \absorbl ⇔ (\assoclt ◎ (\absorbl ⊗ \idc)) ◎ \absorbr}
+\\
+  {\idc ⊗ \, \identlp ⇔ (\distl ◎ (\absorbl ⊕ \idc)) ◎ \identlp}
+\end{array}\]
+\caption{\label{figc}Signatures of level-1 $\Pi$-combinators: zero}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((\assoclp ⊗ \idc) ◎ \dist) ◎ (\dist ⊕ \idc) ⇔ (\dist ◎ (\idc ⊕ \dist)) ◎ \assoclp}
+\\
+  {\assoclt ◎ \distl ⇔ ((\idc ⊗ \distl) ◎ \distl) ◎ (\assoclt ⊕ \assoclt)}
+\end{array}\]
+\vspace{ -0.5em}
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\distl ◎ (\dist ⊕ \dist)) ◎ \assoclp &⇔&   
+   \dist ◎ (\distl ⊕ \distl) ◎ \assoclp ~◎ \\
+&& (\assocrp ⊕ \idc) ~◎ \\
+&& ((\idc ⊕ \swapp) ⊕ \idc) ~◎ \\
+&&      (\assoclp ⊕ \idc)
+\end{array}\]
+\caption{\label{figb}Signatures of level-1 $\Pi$-combinators: associativity and distributivity}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\idc ⊗ \swapp) ◎ \distl &⇔& \distl ◎ \swapp
+\\
+  \dist ◎ (\swapt ⊕ \swapt) &⇔& \swapt ◎ \distl
+\end{array}\]
+\caption{\label{figa}Signatures of level-1 $\Pi$-combinators: commutativity and distributivity}
+\end{figure}
 

### 2017-12-24 survey/survey.tex
added level 2 combinators

@@
 
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t₃$, and $c_3 : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {c_1 \circledcirc (c_2 \circledcirc c_3) \Leftrightarrow (c_1 \circledcirc c_2) \circledcirc c_3}
+\\
+  {(c_1 \oplus (c_2 \oplus c_3)) \circledcirc \assoclp \Leftrightarrow \assoclp \circledcirc ((c_1 \oplus c_2) \oplus c_3)}
+\\
+  {(c_1 \otimes (c_2 \otimes c_3)) \circledcirc \assoclt \Leftrightarrow \assoclt \circledcirc ((c_1 \otimes c_2) \otimes c_3)}
+\\
+  {((c_1 \oplus c_2) \oplus c_3) \circledcirc \assocrp \Leftrightarrow \assocrp \circledcirc (c_1 \oplus (c_2 \oplus c_3))}
+\\
+  {((c_1 \otimes c_2) \otimes c_3) \circledcirc \assocrt \Leftrightarrow \assocrt \circledcirc (c_1 \otimes (c_2 \otimes c_3))}
+\\
+  {\assocrp \circledcirc \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \circledcirc \assocrp) \circledcirc (\idc \oplus \assocrp)}
+\\
+  {\assocrt \circledcirc \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \circledcirc \assocrt) \circledcirc (\idc \otimes \assocrt)}
+\end{array}\]
+\caption{\label{figj}Signatures of level-1 $\Pi$-combinators: associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t₅ \leftrightarrow t₆$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((a \oplus b) \otimes c) \circledcirc \dist \Leftrightarrow \dist \circledcirc ((a \otimes c) \oplus (b \otimes c))}
+\\
+  {(a \otimes (b \oplus c)) \circledcirc \distl \Leftrightarrow \distl \circledcirc ((a \otimes b) \oplus (a \otimes c))}
+\\
+  {((a \otimes c) \oplus (b \otimes c)) \circledcirc \factor \Leftrightarrow \factor \circledcirc ((a \oplus b) \otimes c)}
+\\
+  {((a \otimes b) \oplus (a \otimes c)) \circledcirc \factorl \Leftrightarrow \factorl \circledcirc (a \otimes (b \oplus c))}
+\end{array}\]
+\caption{\label{figi}Signatures of level-1 $\Pi$-combinators: distributivity and factoring}
+\end{figure}
+
+\begin{figure}[t]
+Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$: 
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\idc \circledcirc \, c \Leftrightarrow c}
+\quad 
+  {c \, \circledcirc \idc \, \Leftrightarrow c}
+\quad
+  {c\,\, \circledcirc\,! c \Leftrightarrow \idc}
+\quad 
+  {! c \circledcirc c \Leftrightarrow \idc}
+\\
+  {c \Leftrightarrow c}
+\quad 
+\Rule{}
+  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
+  {c_1 \Leftrightarrow c_3}
+  {} 
+\quad
+\Rule{}
+  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
+  {c_1 \circledcirc c_2 \Leftrightarrow c' \circledcirc c''}
+  {}
+\end{array}\]
+\caption{\label{figh}Signatures of level-1 $\Pi$-combinators: identity and composition}
+\end{figure}
+
+\begin{figure}[t]
+Let $c₀ : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlp \circledcirc c \Leftrightarrow (c₀ \oplus c) \circledcirc \identlp}
+\qquad 
+  {\identrp \circledcirc (c₀ \oplus c) \Leftrightarrow c \circledcirc \identrp}
+\\
+  {\identlsp \circledcirc c \Leftrightarrow (c \oplus c₀) \circledcirc \identlsp}
+\qquad
+  {\identrsp \circledcirc (c \oplus c₀) \Leftrightarrow c \circledcirc \identrsp}
+\\
+  {\identlt \circledcirc c \Leftrightarrow (c_1 \otimes c) \circledcirc \identlt}
+\qquad
+  {\identrt \circledcirc (c_1 \otimes c) \Leftrightarrow c \circledcirc \identrp}
+\\
+  {\identlst \circledcirc c \Leftrightarrow (c \otimes c_1) \circledcirc \identlst}
+\qquad
+  {\identrst \circledcirc (c \otimes c_1) \Leftrightarrow c \circledcirc \identrst}
+\\
+  {\identlt \Leftrightarrow \distl \circledcirc (\identlt \oplus \identlt)}
+\\
+\identlp \Leftrightarrow \swapp \circledcirc \identrsp
+\qquad
+\identlt \Leftrightarrow \swapt \circledcirc \identrst
+\end{array}\]
+\caption{\label{figg}Signatures of level-1 $\Pi$-combinators: unit}
+\end{figure}
+
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\swapp \circledcirc (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \circledcirc \swapp}
+\quad
+  {\swapt \circledcirc (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \circledcirc \swapt}
+\\
+  {(\assocrp \circledcirc \swapp) \circledcirc \assocrp \Leftrightarrow ((\swapp \oplus \idc) \circledcirc \assocrp) \circledcirc (\idc \oplus \swapp)}
+\\
+  {(\assoclp \circledcirc \swapp) \circledcirc \assoclp \Leftrightarrow ((\idc \oplus \swapp) \circledcirc \assoclp) \circledcirc (\swapp \oplus \idc)}
+\\
+  {(\assocrt \circledcirc \swapt) \circledcirc \assocrt \Leftrightarrow ((\swapt \otimes \idc) \circledcirc \assocrt) \circledcirc (\idc \otimes \swapt)}
+\\
+  {(\assoclt \circledcirc \swapt) \circledcirc \assoclt \Leftrightarrow ((\idc \otimes \swapt) \circledcirc \assoclt) \circledcirc (\swapt \otimes \idc)}
+\end{array}\]
+\caption{\label{figf}Signatures of level-1 $\Pi$-combinators: commutativity and associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
+Let $a_1 : t₅ \leftrightarrow t_1$,  $a_2 : t₆ \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
+  {}
+\qquad
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
+  {} 
+\\
+  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
+\qquad
+  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
+\\
+  {(a_1 \circledcirc a_3) \oplus (a_2 \circledcirc a_4) \Leftrightarrow (a_1 \oplus a_2) \circledcirc (a_3 \oplus a_4)}
+\\
+  {(a_1 \circledcirc a_3) \otimes (a_2 \circledcirc a_4) \Leftrightarrow (a_1 \otimes a_2) \circledcirc (a_3 \otimes a_4)}
+\end{array}\]
+\caption{\label{fige}Signatures of level-1 $\Pi$-combinators: functors}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \circledcirc (\idc \oplus \, \identlp)}
+\\
+  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \circledcirc (\idc \otimes \, \identlt)}
+\end{array}\]
+\caption{\label{figd}Signatures of level-1 $\Pi$-combinators: unit and associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $c : t_1 \leftrightarrow t_2$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {(c \otimes \idc) \circledcirc \absorbl \Leftrightarrow \absorbl \circledcirc \idc}
+\quad
+  {(\idc \, \otimes c) \circledcirc \absorbr \Leftrightarrow \absorbr \circledcirc \idc}
+\\
+  {\idc \circledcirc \, \factorzl \Leftrightarrow \factorzl \circledcirc (\idc \otimes c)}
+\quad
+  {\idc \circledcirc \, \factorzr \Leftrightarrow \factorzr \circledcirc (c \otimes \idc)}
+\\
+  {\absorbr \Leftrightarrow \absorbl}
+\\
+  {\absorbr \Leftrightarrow (\distl \circledcirc (\absorbr \oplus \absorbr)) \circledcirc \identlp}
+\\
+  {\identlst \Leftrightarrow \absorbr}
+\qquad
+  {\absorbl \Leftrightarrow \swapt \circledcirc \absorbr}
+\\
+  {\absorbr \Leftrightarrow (\assoclt \circledcirc (\absorbr \otimes \idc)) \circledcirc \absorbr}
+\\
+  {(\idc \otimes \absorbr) \circledcirc \absorbl \Leftrightarrow (\assoclt \circledcirc (\absorbl \otimes \idc)) \circledcirc \absorbr}
+\\
+  {\idc \otimes \, \identlp \Leftrightarrow (\distl \circledcirc (\absorbl \oplus \idc)) \circledcirc \identlp}
+\end{array}\]
+\caption{\label{figc}Signatures of level-1 $\Pi$-combinators: zero}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((\assoclp \otimes \idc) \circledcirc \dist) \circledcirc (\dist \oplus \idc) \Leftrightarrow (\dist \circledcirc (\idc \oplus \dist)) \circledcirc \assoclp}
+\\
+  {\assoclt \circledcirc \distl \Leftrightarrow ((\idc \otimes \distl) \circledcirc \distl) \circledcirc (\assoclt \oplus \assoclt)}
+\end{array}\]
+\vspace{ -0.5em}
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\distl \circledcirc (\dist \oplus \dist)) \circledcirc \assoclp &\Leftrightarrow&   
+   \dist \circledcirc (\distl \oplus \distl) \circledcirc \assoclp ~\circledcirc \\
+&& (\assocrp \oplus \idc) ~\circledcirc \\
+&& ((\idc \oplus \swapp) \oplus \idc) ~\circledcirc \\
+&&      (\assoclp \oplus \idc)
+\end{array}\]
+\caption{\label{figb}Signatures of level-1 $\Pi$-combinators: associativity and distributivity}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\idc \otimes \swapp) \circledcirc \distl &\Leftrightarrow& \distl \circledcirc \swapp
+\\
+  \dist \circledcirc (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \circledcirc \distl
+\end{array}\]
+\caption{\label{figa}Signatures of level-1 $\Pi$-combinators: commutativity and distributivity}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\amr{another wavefront}
-
-\begin{figure}[t]
-Let $c₁ : t₁ ⟷ t₂$,  $c₂ : t₂ ⟷ t₃$, and $c₃ : t₃ ⟷ t₄$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {c₁ ◎ (c₂ ◎ c₃) ⇔ (c₁ ◎ c₂) ◎ c₃}
-\\
-  {(c₁ ⊕ (c₂ ⊕ c₃)) ◎ \assoclp ⇔ \assoclp ◎ ((c₁ ⊕ c₂) ⊕ c₃)}
-\\
-  {(c₁ ⊗ (c₂ ⊗ c₃)) ◎ \assoclt ⇔ \assoclt ◎ ((c₁ ⊗ c₂) ⊗ c₃)}
-\\
-  {((c₁ ⊕ c₂) ⊕ c₃) ◎ \assocrp ⇔ \assocrp ◎ (c₁ ⊕ (c₂ ⊕ c₃))}
-\\
-  {((c₁ ⊗ c₂) ⊗ c₃) ◎ \assocrt ⇔ \assocrt ◎ (c₁ ⊗ (c₂ ⊗ c₃))}
-\\
-  {\assocrp ◎ \assocrp ⇔ ((\assocrp ⊕ \idc) ◎ \assocrp) ◎ (\idc ⊕ \assocrp)}
-\\
-  {\assocrt ◎ \assocrt ⇔ ((\assocrt ⊗ \idc) ◎ \assocrt) ◎ (\idc ⊗ \assocrt)}
-\end{array}\]
-\caption{\label{figj}Signatures of level-1 $\Pi$-combinators: associativity}
-\end{figure}
-
-\begin{figure}[t]
-Let $a : t₁ ⟷ t₂$, $b : t₃ ⟷ t₄$, and $c : t₅ ⟷ t₆$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {((a ⊕ b) ⊗ c) ◎ \dist ⇔ \dist ◎ ((a ⊗ c) ⊕ (b ⊗ c))}
-\\
-  {(a ⊗ (b ⊕ c)) ◎ \distl ⇔ \distl ◎ ((a ⊗ b) ⊕ (a ⊗ c))}
-\\
-  {((a ⊗ c) ⊕ (b ⊗ c)) ◎ \factor ⇔ \factor ◎ ((a ⊕ b) ⊗ c)}
-\\
-  {((a ⊗ b) ⊕ (a ⊗ c)) ◎ \factorl ⇔ \factorl ◎ (a ⊗ (b ⊕ c))}
-\end{array}\]
-\caption{\label{figi}Signatures of level-1 $\Pi$-combinators: distributivity and factoring}
-\end{figure}
-
-\begin{figure}[t]
-Let $c, c₁, c₂, c₃ : t₁ ⟷ t₂$ and $c', c'' : t₃ ⟷ t₄$: 
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\idc ◎ \, c ⇔ c}
-\quad 
-  {c \, ◎ \idc \, ⇔ c}
-\quad
-  {c\,\, ◎\,! c ⇔ \idc}
-\quad 
-  {! c ◎ c ⇔ \idc}
-\\
-  {c ⇔ c}
-\quad 
-\Rule{}
-  {c₁ ⇔ c₂ \quad c₂ ⇔ c₃}
-  {c₁ ⇔ c₃}
-  {} 
-\quad
-\Rule{}
-  {c₁ ⇔ c' \quad c₂ ⇔ c''}
-  {c₁ ◎ c₂ ⇔ c' ◎ c''}
-  {}
-\end{array}\]
-\caption{\label{figh}Signatures of level-1 $\Pi$-combinators: identity and composition}
-\end{figure}
-
-\begin{figure}[t]
-Let $c₀ : 0 ⟷ 0$, $c₁ : 1 ⟷ 1$, and $c : t₁ ⟷ t₂$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\identlp ◎ c ⇔ (c₀ ⊕ c) ◎ \identlp}
-\qquad 
-  {\identrp ◎ (c₀ ⊕ c) ⇔ c ◎ \identrp}
-\\
-  {\identlsp ◎ c ⇔ (c ⊕ c₀) ◎ \identlsp}
-\qquad
-  {\identrsp ◎ (c ⊕ c₀) ⇔ c ◎ \identrsp}
-\\
-  {\identlt ◎ c ⇔ (c₁ ⊗ c) ◎ \identlt}
-\qquad
-  {\identrt ◎ (c₁ ⊗ c) ⇔ c ◎ \identrp}
-\\
-  {\identlst ◎ c ⇔ (c ⊗ c₁) ◎ \identlst}
-\qquad
-  {\identrst ◎ (c ⊗ c₁) ⇔ c ◎ \identrst}
-\\
-  {\identlt ⇔ \distl ◎ (\identlt ⊕ \identlt)}
-\\
-\identlp ⇔ \swapp ◎ \identrsp
-\qquad
-\identlt ⇔ \swapt ◎ \identrst
-\end{array}\]
-\caption{\label{figg}Signatures of level-1 $\Pi$-combinators: unit}
-\end{figure}
-
-\begin{figure}[t]
-Let $c₁ : t₁ ⟷ t₂$ and $c₂ : t₃ ⟷ t₄$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\swapp ◎ (c₁ ⊕ c₂) ⇔ (c₂ ⊕ c₁) ◎ \swapp}
-\quad
-  {\swapt ◎ (c₁ ⊗ c₂) ⇔ (c₂ ⊗ c₁) ◎ \swapt}
-\\
-  {(\assocrp ◎ \swapp) ◎ \assocrp ⇔ ((\swapp ⊕ \idc) ◎ \assocrp) ◎ (\idc ⊕ \swapp)}
-\\
-  {(\assoclp ◎ \swapp) ◎ \assoclp ⇔ ((\idc ⊕ \swapp) ◎ \assoclp) ◎ (\swapp ⊕ \idc)}
-\\
-  {(\assocrt ◎ \swapt) ◎ \assocrt ⇔ ((\swapt ⊗ \idc) ◎ \assocrt) ◎ (\idc ⊗ \swapt)}
-\\
-  {(\assoclt ◎ \swapt) ◎ \assoclt ⇔ ((\idc ⊗ \swapt) ◎ \assoclt) ◎ (\swapt ⊗ \idc)}
-\end{array}\]
-\caption{\label{figf}Signatures of level-1 $\Pi$-combinators: commutativity and associativity}
-\end{figure}
-
-\begin{figure}[t]
-Let $c₁ : t₁ ⟷ t₂$, $c₂ : t₃ ⟷ t₄$, $c₃ : t₁ ⟷ t₂$, and $c₄ : t₃ ⟷ t₄$. \\
-Let $a₁ : t₅ ⟷ t₁$,  $a₂ : t₆ ⟷ t₂$, $a₃ : t₁ ⟷ t₃$, and $a₄ : t₂ ⟷ t₄$.
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-\Rule{}
-  {c₁ ⇔ c₃ \quad c₂ ⇔ c₄}
-  {c₁ ⊕ c₂ ⇔ c₃ ⊕ c₄}
-  {}
-\qquad
-\Rule{}
-  {c₁ ⇔ c₃ \quad c₂ ⇔ c₄}
-  {c₁ ⊗ c₂ ⇔ c₃ ⊗ c₄}
-  {} 
-\\
-  {\idc ⊕ \, \idc \, ⇔ \idc}
-\qquad
-  {\idc ⊗ \, \idc \, ⇔ \idc}
-\\
-  {(a₁ ◎ a₃) ⊕ (a₂ ◎ a₄) ⇔ (a₁ ⊕ a₂) ◎ (a₃ ⊕ a₄)}
-\\
-  {(a₁ ◎ a₃) ⊗ (a₂ ◎ a₄) ⇔ (a₁ ⊗ a₂) ◎ (a₃ ⊗ a₄)}
-\end{array}\]
-\caption{\label{fige}Signatures of level-1 $\Pi$-combinators: functors}
-\end{figure}
-
-\begin{figure}[t]
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\identlsp ⊕ \idc ~⇔~ \assocrp ◎ (\idc ⊕ \, \identlp)}
-\\
-  {\identlst ⊗ \idc ~⇔~ \assocrt ◎ (\idc ⊗ \, \identlt)}
-\end{array}\]
-\caption{\label{figd}Signatures of level-1 $\Pi$-combinators: unit and associativity}
-\end{figure}
-
-\begin{figure}[t]
-Let $c : t₁ ⟷ t₂$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {(c ⊗ \idc) ◎ \absorbl ⇔ \absorbl ◎ \idc}
-\quad
-  {(\idc \, ⊗ c) ◎ \absorbr ⇔ \absorbr ◎ \idc}
-\\
-  {\idc ◎ \, \factorzl ⇔ \factorzl ◎ (\idc ⊗ c)}
-\quad
-  {\idc ◎ \, \factorzr ⇔ \factorzr ◎ (c ⊗ \idc)}
-\\
-  {\absorbr ⇔ \absorbl}
-\\
-  {\absorbr ⇔ (\distl ◎ (\absorbr ⊕ \absorbr)) ◎ \identlp}
-\\
-  {\identlst ⇔ \absorbr}
-\qquad
-  {\absorbl ⇔ \swapt ◎ \absorbr}
-\\
-  {\absorbr ⇔ (\assoclt ◎ (\absorbr ⊗ \idc)) ◎ \absorbr}
-\\
-  {(\idc ⊗ \absorbr) ◎ \absorbl ⇔ (\assoclt ◎ (\absorbl ⊗ \idc)) ◎ \absorbr}
-\\
-  {\idc ⊗ \, \identlp ⇔ (\distl ◎ (\absorbl ⊕ \idc)) ◎ \identlp}
-\end{array}\]
-\caption{\label{figc}Signatures of level-1 $\Pi$-combinators: zero}
-\end{figure}
-
-\begin{figure}[t]
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {((\assoclp ⊗ \idc) ◎ \dist) ◎ (\dist ⊕ \idc) ⇔ (\dist ◎ (\idc ⊕ \dist)) ◎ \assoclp}
-\\
-  {\assoclt ◎ \distl ⇔ ((\idc ⊗ \distl) ◎ \distl) ◎ (\assoclt ⊕ \assoclt)}
-\end{array}\]
-\vspace{ -0.5em}
-\[\def\arraystretch{1.3}
-\begin{array}{rcl}
-  (\distl ◎ (\dist ⊕ \dist)) ◎ \assoclp &⇔&   
-   \dist ◎ (\distl ⊕ \distl) ◎ \assoclp ~◎ \\
-&& (\assocrp ⊕ \idc) ~◎ \\
-&& ((\idc ⊕ \swapp) ⊕ \idc) ~◎ \\
-&&      (\assoclp ⊕ \idc)
-\end{array}\]
-\caption{\label{figb}Signatures of level-1 $\Pi$-combinators: associativity and distributivity}
-\end{figure}
-
-\begin{figure}[t]
-\[\def\arraystretch{1.3}
-\begin{array}{rcl}
-  (\idc ⊗ \swapp) ◎ \distl &⇔& \distl ◎ \swapp
-\\
-  \dist ◎ (\swapt ⊕ \swapt) &⇔& \swapt ◎ \distl
-\end{array}\]
-\caption{\label{figa}Signatures of level-1 $\Pi$-combinators: commutativity and distributivity}
-\end{figure}
+\amr{*** another wavefront ***}
 

### 2017-12-25 survey/survey.tex
some general cleanup

@@
 
-\title{Embracing the Laws of Physics: \\ Reversible Models of Computation}
-\author{Jacques Carette, Roshan P. James, Amr Sabry}
+\title{Embracing the Laws of Physics: \\ Three Reversible Models of Computation}
+\author{Jacques Carette \qquad\qquad Roshan P. James \qquad\qquad Amr Sabry \\
+McMaster University \qquad\qquad Google \qquad\qquad Indiana University}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction : Reversibility, the Missing Principle}
+\section{Reversibility, the Missing Principle}
+
+\subsection{Maxwell's Demon}
 
@@
 investigations, in the intersection between mathematics and
-computer science (particularly in Homotopy Type Theory~\cite{hott-book})
+computer science (particularly in Homotopy Type Theory~\cite{hottbook})
 reveal that not only can mathematics be typed, but that doing so
@@
 
-\subsection{From classical to quantum}
+\subsection{From Classical to Quantum}
 
@@
 
-\subsection{conservation of information}
+\subsection{Conservation of Information}
 
@@
 
-This stands in stark contrast with most current approacheds to
+This stands in stark contrast with most current approaches to
 reversible computation, which start from current models of
@@
 Starting from a physical perspective, whatever our notion of data
-is, we will be interested in programs as representating transformations
+is, we will be interested in programs as representing transformations
 of that data which are reversible. In other words, we want our
@@
 type \emph{equivalences}. To do this, we need a few important
-auxilliary concepts first.
+auxiliary concepts first.
 
@@
 \noindent Why \emph{quasi}? The reasons are beyond our scope, but
-the interested reader can read sections $2.4$ and $4$ in \cite{hott-book}.
+the interested reader can read sections $2.4$ and $4$ in \cite{hottbook}.
 There are several conceptually different, but equivalent, 
@@
 map to $\bot, \top, \uplus, \times$, and such that we have ground
-terms whose denotation are all $16$ type isormorphisms of
+terms whose denotation are all $16$ type isomorphisms of
 Figure~\ref{type-isos}. This is completely straightforward, as
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Data IV: Recursive Types}
+\section{Conclusion}
+
+Discuss recursion, or add a section of infinite sets???
+
+CS abstractions are based on ``old'' physics
+
+computer applications are more and more ``physical''
+
+physical principles such as conservation of information should be part
+of our foundational abstractions
+
+quest for other principles; other data models; graphs; HoTT
+
+Data IV: Recursive Types
 
@@
 
+Data V: Exclusive Disjunctions
+
+Quantum over finite field
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Data V: Exclusive Disjunctions}
-
-???
-
-Quantum over finite field
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
-Discuss recursion, or add a section of infinite sets???
-
-CS abstractions are based on ``old'' physics
-
-computer applications are more and more ``physical''
-
-physical principles such as conservation of information should be part
-of our foundational abstractions
-
-quest for other principles; other data models; graphs; HoTT
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements} many students and colleagues
-
-\bibliographystyle{plainnat}
+\section*{Acknowledgements} We would like to thank the numerous
+students and colleagues who participated in various aspects of this
+research and who provided valuable feedack and constructive criticism.
+
+\bibliographystyle{acm}
 %\pagestyle{headings}

### 2017-12-26 survey/survey.tex
intro

@@
 
+What kind of operations can computers perform? This question was
+answered several times in the last hundred years, where each answer
+proposes an abstract \emph{model of computation} that specifies
+allowable operations and (usually) their cost. The emerging consensus,
+reflected in both early models of computations such as the Turing
+Machine and the RAM as well as in the early Von Neumann and in modern
+computer architectures, is that basic computer operations include
+logical operations like conjunction, disjunction, and negation, as
+well as reading from and writing to a large (infinite) collection of
+memory locations.
+
+No doubt, this consensus has been successful. And no doubt, these
+operations \emph{can} indeed be performed on a computer. Yet, today,
+with a possible quantum computing revolution in sight and with
+unprecedented explosion in embedded computers and cyber-physical
+systems, there are reasons to re-think this foundational question
+again. In fact, the calls to re-think this foundational question have
+been proclaimed by physicists for more than thirty years as the
+following two quotes testify:
+
+\begin{quote}
+  Mathematical models of computation are abstract constructions, by
+  their nature unfettered by physical laws. However, if these models
+  are to give indications that are relevant to concrete computing,
+  they must somehow capture, albeit in a selective and stylized way,
+  certain general physical restrictions to which all concrete
+  computing processes are subjected~\cite{toffoli:1980}.
+
+  Another thing that has been suggested early was that natural laws
+  are reversible, but that computer rules are not. But this turned out
+  to be false; the computer rules can be reversible, and it has been a
+  very, very useful thing to notice and to discover that. This is a
+  place where the relationship of physics and computation has turned
+  itself the other way and told us something about the possibilities
+  of computation. So this is an interesting subject because it tells
+  us something about computer rules
+  \ldots~\cite{springerlink:10.1007/bf02650179}.
+\end{quote}
+
+These quotes by Toffoli and Feynman both highlight two obvious
+observations: (i) all the operations that a computer performs reduce
+to basic physical operations; and (ii) there is a mismtach between the
+logical operations of a typical model of computation (which are
+logically irreversible) and the fundamental laws of physics (which are
+reversible). One could certainly dismiss the mismatch but our thesis
+is that there is great value in revisiting models of computation to be
+in harmony with the laws of physics.
+
+\begin{comment}
+\begin{quote}
+  Ed Fredkin pursued the idea that information must be finite in
+  density. One day, he announced that things must be even more simple
+  than that. He said that he was going to assume that information
+  itself is conserved. “You’re out of you mind, Ed.” I
+  pronounced. “That’s completely ridiculous. Nothing could happen in
+  such a world. There couldn’t even be logical gates. No decisions
+  could ever be made.” But when Fredkin gets one of his ideas, he’s
+  quite immune to objections like that; indeed, they fuel him with
+  energy. Soon he went on to assume that information processing must
+  also be reversible — and invented what’s now called the Fredkin
+  gate. (Minsky 1999)
+
+  In other terms, what is so good in logic that quantum physics should
+  obey?  Can't we imagine that our conceptions about logic are wrong,
+  so wrong that they are unable to cope with the quantum miracle?
+  [\ldots] Instead of teaching logic to nature, it is more reasonable
+  to learn from her. Instead of interpreting quantum into logic, we
+  shall interpret logic into quantum (Girard 2007).
+\end{quote}
+\end{comment}
+
 \subsection{Maxwell's Demon}
@@
   and stylized way, certain general physical restrictions to which all
-  concrete computing processes are subjected~\cite{Toffoli:1980}.
+  concrete computing processes are subjected~\cite{toffoli:1980}.
 \end{quote}}
@@
 The technical development derives from ideas which are implicit in the
-works of \cite{Toffoli:1980, Zuliani:2001:LR,
+works of \cite{toffoli:1980, Zuliani:2001:LR,
   malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.

### 2017-12-26 survey/survey.tex
more work on intro

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %% Macros
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Reversibility, the Missing Principle}
@@
 logically irreversible) and the fundamental laws of physics (which are
-reversible). One could certainly dismiss the mismatch but our thesis
-is that there is great value in revisiting models of computation to be
-in harmony with the laws of physics.
+reversible). One could certainly dismiss the mismatch as irrelevant to
+the practice of computing but our thesis is that there is the next
+computing revolution is likely to be founded on revised models of
+computation that are designed to in harmony with the laws of physics.
+
+After a detailed introduction on the origins of ``logically
+reversibile computer operations'' and an excursion in the origins of
+our the status quo based on ``irreversible computer operations'', we
+will develop in detail three reversible models of computation and
+discuss their potential applications.
 
@@
 
+%%%%%%%%%
 \subsection{Maxwell's Demon}
 
-To fully appreciate one of the principles is missing in conventional computing,
-one must go back to an old thought experiment by J. C. Maxwell. This is
-codified in a letter that Maxwell wrote to P. G Tait in 1867 --
-the letter whose ideas are now known as `Maxwell's
-Demon'. Maxwell's Demon is a thought experiment that seems to
+To fully appreciate one of the principles is missing in conventional
+computing, one must go back to an old thought experiment by
+J. C. Maxwell. This is codified in a letter that Maxwell wrote to P. G
+Tait in 1867 -- the letter, whose ideas are now known as
+\emph{Maxwell's Demon}, tells of a thought experiment that seems to
 indicate that intelligent beings can somehow violate the second law of
-thermodynamics, thereby violating physics itself.
-
-Many resolutions were offered for this conundrum (for a compilation, see
-Maxwell's Demon~\cite{leff1990}), but none withstood careful
-scrutiny until the establishment of `Landauer's Principle' in 1961 -- a
-principle whose experimental validation
-happened recently in 2012~\cite{berut2012experimental}.
+thermodynamics, thereby violating physics itself. Many resolutions
+were offered for this conundrum (for a compilation, see Maxwell's
+Demon~\cite{leff1990}), but none withstood careful scrutiny until the
+establishment of \emph{Landauer's Principle} in 1961 -- a principle
+whose experimental validation happened recently in
+2012~\cite{berut2012experimental}.
 
@@
 
-This is a strange and wonderful idea: Information,
-physics and computation are inextricably linked. Furthermore, this
-implies that ideas in each field have consequences for the other.
-(Cite Bennett and the various `thermodynamics of computation' papers here.)
+This is a strange and wonderful idea: information, physics and
+computation are inextricably linked. Furthermore, this implies that
+ideas in each field have consequences for the
+other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,dblp:conf/csfw/malacarias12}.
+%% (Cite Bennett and the various `thermodynamics of computation'
+%% papers here.)
 
@@
 
-\begin{quote}
-Toffoli 1980: Mathematical models of computation are abstract
-constructions, by their nature unfettered by physical laws. However,
-if these models are to give indications that are relevant to concrete
-computing, they must somehow capture, albeit in a selective and
-stylized way, certain general physical restrictions to which all
-concrete computing processes are subjected.
-\end{quote}
-
-In our current age, both Landauer's Principle and recent investigations
-of computational models well-suited to quantum mechanics, push us to
-revisit our current models of computation. Information, on top of
-having a physical manifestation, appears to be \emph{conserved},
-along with other quantities like energy, mass and momentum.
-
-But what does it mean for a computation to preserve information?
-What is the missing principle that will guide the creation of a new
-model of computation that ``conserve information''?
-It is nature to look again at physics for inspiration --
-in our current understanding of physics all of the primitive rules
-are reversible. Conventional computation is not. Can
-we embrace this simple principle as the building block of a model of
-computation? What would computation look like when viewed from this
-vantage point? Can non-trivial computation be performed in a
-reversible manner? And, if so, does it have applications?
-
-Traditional models of computation are all unityped, generally
-operating on unstructured sequences of symbols. But recent
-investigations, in the intersection between mathematics and
-computer science (particularly in Homotopy Type Theory~\cite{hottbook})
-reveal that not only can mathematics be typed, but that doing so
-reveals further structure -- in both mathematics and in computation.
-
-And that brings us to the principal objective of this survey: explore the
-domain of \emph{typed reversible computation}.
-
 But first, we need to better understand our current models of
@@
 
-\subsection{Origins}
+%%%%%%%%%
+\subsection{Origins of Computational Models}
 
@@
 
+%%%%%%%%%
 \subsection{From Classical to Quantum}
@@
 
+%%%%%%%%%
 \subsection{Conservation of Information}
@@
 
+%%%%%%%%%
 \subsection{Information}
@@
 
-%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+In our current age, both Landauer's Principle and recent investigations
+of computational models well-suited to quantum mechanics, push us to
+revisit our current models of computation. Information, on top of
+having a physical manifestation, appears to be \emph{conserved},
+along with other quantities like energy, mass and momentum.
+
+But what does it mean for a computation to preserve information?
+What is the missing principle that will guide the creation of a new
+model of computation that ``conserve information''?
+It is nature to look again at physics for inspiration --
+in our current understanding of physics all of the primitive rules
+are reversible. Conventional computation is not. Can
+we embrace this simple principle as the building block of a model of
+computation? What would computation look like when viewed from this
+vantage point? Can non-trivial computation be performed in a
+reversible manner? And, if so, does it have applications?
+
+Traditional models of computation are all unityped, generally
+operating on unstructured sequences of symbols. But recent
+investigations, in the intersection between mathematics and
+computer science (particularly in Homotopy Type Theory~\cite{hottbook})
+reveal that not only can mathematics be typed, but that doing so
+reveals further structure -- in both mathematics and in computation.
+
+And that brings us to the principal objective of this survey: explore the
+domain of \emph{typed reversible computation}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Programs as Reversible Deformations}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Data I: Finite Sets}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Data II: Structured Finite Types}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Data III: Reversible Programs between Reversible Programs}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgements} We would like to thank the numerous
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2017-12-26 survey/survey.tex
progress on intro

@@
 again. In fact, the calls to re-think this foundational question have
-been proclaimed by physicists for more than thirty years as the
+been proclaimed by physicists more than forty years ago as the
 following two quotes testify:
@@
 \begin{quote}
-  Mathematical models of computation are abstract constructions, by
-  their nature unfettered by physical laws. However, if these models
-  are to give indications that are relevant to concrete computing,
-  they must somehow capture, albeit in a selective and stylized way,
-  certain general physical restrictions to which all concrete
-  computing processes are subjected~\cite{toffoli:1980}.
-
+  \textbf{Toffoli 1980~\cite{toffoli:1980}:} Mathematical models of computation are
+  abstract constructions, by their nature unfettered by physical
+  laws. However, if these models are to give indications that are
+  relevant to concrete computing, they must somehow capture, albeit in
+  a selective and stylized way, certain general physical restrictions
+  to which all concrete computing processes are
+  subjected.
+
+  \textbf{Feynman 1982~\cite{springerlink:10.1007/bf02650179}:}
   Another thing that has been suggested early was that natural laws
@@
   of computation. So this is an interesting subject because it tells
-  us something about computer rules
-  \ldots~\cite{springerlink:10.1007/bf02650179}.
+  us something about computer rules.
 \end{quote}
 
-These quotes by Toffoli and Feynman both highlight two obvious
-observations: (i) all the operations that a computer performs reduce
-to basic physical operations; and (ii) there is a mismtach between the
-logical operations of a typical model of computation (which are
-logically irreversible) and the fundamental laws of physics (which are
-reversible). One could certainly dismiss the mismatch as irrelevant to
-the practice of computing but our thesis is that there is the next
-computing revolution is likely to be founded on revised models of
-computation that are designed to in harmony with the laws of physics.
+\noindent These quotes by Toffoli and Feynman both highlight the
+consequences of two obvious observations: (i) all the operations that
+a computer performs reduce to basic physical operations; and (ii)
+there is a mismatch between the logical operations of a typical model
+of computation (which are logically irreversible) and the fundamental
+laws of physics (which are reversible). One could certainly dismiss
+the mismatch as irrelevant to the practice of computing but our thesis
+is that the next computing revolution is likely to be founded on
+revised models of computation that are designed to in harmony with the
+laws of physics.
 
 After a detailed introduction on the origins of ``logically
-reversibile computer operations'' and an excursion in the origins of
-our the status quo based on ``irreversible computer operations'', we
-will develop in detail three reversible models of computation and
-discuss their potential applications.
+reversibile computer operations'' and an excursion into the origins of
+``irreversible computer operations'', we will develop in detail three
+reversible models of computation and discuss their potential
+applications.
 
@@
 %%%%%%%%%
-\subsection{Maxwell's Demon}
-
-To fully appreciate one of the principles is missing in conventional
-computing, one must go back to an old thought experiment by
-J. C. Maxwell. This is codified in a letter that Maxwell wrote to P. G
-Tait in 1867 -- the letter, whose ideas are now known as
+\paragraph*{Maxwell's Demon.}
+To fully appreciate the missing principle of ``reversibility'' in
+conventional computing, we go back to an old thought experiment by
+J. C. Maxwell. The details are codified in a letter that Maxwell wrote
+to P. G Tait in 1867 -- the letter, whose ideas are now known as
 \emph{Maxwell's Demon}, tells of a thought experiment that seems to
@@
 thermodynamics, thereby violating physics itself. Many resolutions
-were offered for this conundrum (for a compilation, see Maxwell's
-Demon~\cite{leff1990}), but none withstood careful scrutiny until the
-establishment of \emph{Landauer's Principle} in 1961 -- a principle
-whose experimental validation happened recently in
+were offered for this conundrum (for a compilation, see the book by
+Leff and Rex~\cite{leff1990}), but none withstood careful scrutiny
+until the establishment of \emph{Landauer's Principle} in 1961 -- a
+principle whose experimental validation happened recently in
 2012~\cite{berut2012experimental}.
@@
 system. Landauer's resolution of the demon relied on two ideas that
-took root only a few decades earlier -- the formal notion of
-computation (through the work of Turing and Church, 1936) and the
-formal notion of information (through the work of Claude Shannon,
-1948). Landauer reasoned that the computation done by the finite brain
-of the demon, involves getting information about the movement of
-molecules, acting on that information and then overwriting it to make
-room for the next computation.  Landauer reasoned, and this is the
-important part, that the computation that is manipulating information
-in the demon's brain \textit{must be thermodynamic work}, thereby
-bringing the demon back into the fold of physics.
-
-This is a strange and wonderful idea: information, physics and
-computation are inextricably linked. Furthermore, this implies that
-ideas in each field have consequences for the
-other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,dblp:conf/csfw/malacarias12}.
+took root only a few decades earlier: the formal notion of computation
+(through the work of Turing and Church, 1936) and the formal notion of
+information (through the work of Shannon, 1948). Landauer reasoned
+that the computation done by the finite brain of the demon, involves
+getting information about the movement of molecules, acting on that
+information, and then overwriting it to make room for the next
+computation.  In other words, the computation that is manipulating
+information in the demon's brain \textit{must be thermodynamic work},
+thereby bringing the demon back into the fold of physics.
+
 %% (Cite Bennett and the various `thermodynamics of computation'  
@@
 
-When the early models computation, the Turing machine, and the
-$\lambda$-calculus, were developed, there was no compelling reason to the
-take the information content of computations into
-consideration -- in fact, at that time there was no quantifiable
-notion of information. These models followed in the footsteps of logic
-where conventional gates such as AND and OR happily erase their
-inputs. These models were already radical enough, by being purely
-constructive, in an era where classical mathematics, with its
-pervasive use of excluded middle and frequent use of the axiom of
-choice, did not overly worry about effective representation.
-
-But first, we need to better understand our current models of
-computation, as a prelude to understand where they lead us astray.
+This is a strange and wonderful idea: information, physics, and
+computation are inextricably linked. In contrast, when the early
+models of computation were developed, there was no compelling reason
+to the take the information content of computations into consideration
+-- in fact, at that time there was no quantifiable notion of
+information. These models followed in the footsteps of logic where,
+following hundreds of years of tradition, conventional operations such
+conjunction and disjunction are logically irreversible. These models
+were already radical enough, by being purely \emph{constructive}, in
+an era where classical mathematics, with its pervasive use of excluded
+middle and frequent use of the axiom of choice, did not overly worry
+about effective representation. Landauer's observation implied however
+that ideas in each field have consequences for the
+other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,baez2011physics,dblp:conf/csfw/malacarias12}. To
+really appreciate this fact, we delve deeper into the origin of our
+computational models and argue that they are essentially reflections
+of contemporary laws of physics.
+
+\amr{wavefront}
 
 %%%%%%%%%
-\subsection{Origins of Computational Models}
-
+\paragraph*{Origins of Computational Models.}
 Aspects of computation are
@@
 %%%%%%%%%
-\subsection{From Classical to Quantum}
-
+\paragraph*{From Classical to Quantum.}
 Conventional classical models of computation, including boolean
@@
 %%%%%%%%%
-\subsection{Conservation of Information}
-
+\paragraph*{Conservation of Information.}
 To make this idea concrete and to provide a taste of the applications
@@
 %%%%%%%%%
-\subsection{Information}
-
+\paragraph*{Information.}
 We can now give a first precise definition of information preservation.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %% from which we might still be able to pilfer usefully.
+
 \subsection{Technical Overview}

### 2017-12-26 survey/survey.tex
progress on intro
intro kind of done

@@
 
-\begin{comment}
-\begin{quote}
-  Ed Fredkin pursued the idea that information must be finite in
-  density. One day, he announced that things must be even more simple
-  than that. He said that he was going to assume that information
-  itself is conserved. “You’re out of you mind, Ed.” I
-  pronounced. “That’s completely ridiculous. Nothing could happen in
-  such a world. There couldn’t even be logical gates. No decisions
-  could ever be made.” But when Fredkin gets one of his ideas, he’s
-  quite immune to objections like that; indeed, they fuel him with
-  energy. Soon he went on to assume that information processing must
-  also be reversible — and invented what’s now called the Fredkin
-  gate. (Minsky 1999)
-
-  In other terms, what is so good in logic that quantum physics should
-  obey?  Can't we imagine that our conceptions about logic are wrong,
-  so wrong that they are unable to cope with the quantum miracle?
-  [\ldots] Instead of teaching logic to nature, it is more reasonable
-  to learn from her. Instead of interpreting quantum into logic, we
-  shall interpret logic into quantum (Girard 2007).
-\end{quote}
-\end{comment}
-
 %%%%%%%%%
@@
 Leff and Rex~\cite{leff1990}), but none withstood careful scrutiny
-until the establishment of \emph{Landauer's Principle} in 1961 -- a
-principle whose experimental validation happened recently in
-2012~\cite{berut2012experimental}.
+until the establishment of \emph{Landauer's Principle} in
+1961~\cite{Landauer:1961} -- a principle whose experimental validation
+happened recently in 2012~\cite{berut2012experimental}.
 
@@
 thereby bringing the demon back into the fold of physics.
-
-%% (Cite Bennett and the various `thermodynamics of computation'  
-%% papers here.)
 
@@
 
-\amr{wavefront}
-
 %%%%%%%%%
 \paragraph*{Origins of Computational Models.}
-Aspects of computation are
-quite old. M\"{u}ller (1786) first conceived of the idea of a
-``difference machine'', which Babbage (1819--1822) was able to
-construct. There are other computer precursors as well -- the first
-stored programs were actually for looms, most notably those of
-Bouchon (1725) which operated on a paper tape, and Jacquard (1804)
-which operated by chains of punched cards.
-
-But it was Alan Turing's seminal work in 1936 which established the idea
-that computation has a formal interpretation and that all
-computability can be captured within a formal system. Implicit in this
-achievement however is the idea that abstract models of computation
-are just that -- \emph{abstractions of computation realized in the
-physical world.}  In fact, one of the major achievements of Computer
-Science has been the development of abstract models of computation
-that shield the discipline from the underlying technology. As
-effective as these models have been, one must note, however, that they
-\emph{embody several implicit physical assumptions}.  As Tommaso
-Toffoli explains in his influential 1980 paper:
-
-{\begin{quote} Mathematical models of computation are abstract
-  constructions, by their nature unfettered by physical laws. However,
-  if these models are to give indications that are relevant to
-  concrete computing, they must somehow capture, albeit in a selective
-  and stylized way, certain general physical restrictions to which all
-  concrete computing processes are subjected~\cite{toffoli:1980}.
-\end{quote}}
-
-Our logic of programs and our hardware are based on Boolean Logic
--- which really
-ought to be called \emph{Piercean logic}, as Boole had a rather different
-conception of logic which is rather far from our current models.
-Nevertheless, going back
-to Boole's 1853 book entitled \emph{An Investigation of the Laws of Thought,
-  on which are Founded the Mathematical Theories of Logic and
-  Probabilities}. The opening sentence of Ch.~1 is:
+Machines that ``compute'' are quite old. M\"{u}ller (1786) first
+conceived of the idea of a ``difference machine'', which Babbage
+(1819--1822) was able to construct. There are other computer
+precursors as well -- the first stored programs were actually for
+looms, most notably those of Bouchon (1725) which operated on a paper
+tape, and Jacquard (1804) which operated by chains of punched cards.
+Logical operations in high-level programming languages and in our
+hardware are based on Boolean Logic -- which really ought to be called
+\emph{Piercean logic}, as Boole had a rather different conception of
+logic which is rather far from our current models.  Nevertheless,
+going back to Boole's 1853 book entitled \emph{An Investigation of the
+  Laws of Thought, on which are Founded the Mathematical Theories of
+  Logic and Probabilities}. The opening sentence of Ch.~1 is:
 \begin{quote}
@@
 \end{quote}
-A few chapters later, we find:
+which clearly identifies the ``source'' of the logical laws as
+mirroring Boole's understanding of human reasoning. A few chapters
+later, we find:
 \begin{quote}
@@
 \end{quote}
-A detailed historical analysis of Boole's ideas are beyond our scope.
-The above quotes, however, should convey the idea that our notions of
-computation generally date back to ideas that were thought reasonable
+This ``law'' is reasonable in a classical world but is violated by
+postulates of quantum mechanics. Although a detailed historical
+analysis of Boole's ideas in the light of modern physics is beyond our
+scope, the above quotes should convey the idea that our elementary
+computing notions date back to ideas that were thought reasonable
 before \textasciitilde 1900.
 
-It is conventional to base
-the theory of computation and complexity on the Turing Machine.
-Going back to Turing's 1936 article \emph{On Computable Numbers, with
-  an Application to the Entscheidungsproblem,}, the opening sentence of
-Sec. 1 is:
+The 20th century witnessed the real advent of computer science as a
+formal discipline. One of the pioneering works was Alan Turing's
+seminal paper in 1936 which established the idea that computation has
+a formal interpretation and that all computability can be captured
+within a formal system. Implicit in this achievement however is the
+idea that abstract models of computation are just that --
+\emph{abstractions of computation realized in the physical world.}
+Indeed, going back to Turing's 1936 article \emph{On Computable
+  Numbers, with an Application to the Entscheidungsproblem,}, the
+opening sentence of Sec. 1 is:
 \begin{quote}
@@
   I think it is reasonable to suppose that they can only be squares
-whose distance from the closest of the immediately previously observed
-squares does not exceed a certain fixed amount.
+  whose distance from the closest of the immediately previously
+  observed squares does not exceed a certain fixed amount.
 \end{quote}
-A detailed historical account is again beyond our scope. The quotes above
-should convey the ideas that our theories
-of computation and complexity are based on some assumptions that
-Turing found reasonable in 1936. Though it is worth noting that these assumptions
-are both physical (on distances) and metaphysical (on restrictions of the
-mind).  If we take the human mind to be a physical ``machine'' which does
-computation, then when both of the above assumptions are translated to the
-language of physics, they embody what is known as the ``Bekenstein bound:''
-which is an upper limit on the amount of information that can be contained
-within a given finite region of space.
-
-One can examine more and more cases but the general story should be
-clear. Our abstractions were made up based on a certain understanding of the
-laws of physics. In particular, our understanding of physics has evolved
-tremendously since 1900!  Thus it is time to revisit these abstractions,
-especially with respect to quantum mechanics.
+It is worth noting that these assumptions are both physical (on
+distances) and metaphysical (on restrictions of the mind).  If we take
+the human mind to be a physical ``machine'' which does computation,
+then when both of the above assumptions are translated to the language
+of physics, they embody what is known as the ``Bekenstein bound:''
+which is an upper limit on the amount of information that can be
+contained within a given finite region of space. A detailed historical
+account of these ideas in the context of modern physics is again
+beyond our scope. However, the quotes above, like the ones before,
+should convey the ideas that our theories of computation and
+complexity are based on some assumptions that Turing and others found
+reasonable in the 1930s.
+
+To summarize, a major achievement of computer science has been the
+development of abstract models of computation that shield the
+discipline from rapid changes in the underlying technology. Yet, as
+effective as these models have been, one must note that they
+\emph{embody several implicit physical assumptions} and these
+assumptions are based on a certain understanding of the laws of
+physics.
+
+Our understanding of physics has evolved tremendously since 1900!
+Thus it is time to revisit these abstractions, especially with respect
+to quantum mechanics.  Indeed one should take the physical principles
+underlying quantum mechanics, the most successful physical theory
+known to us and adapt computation to ``learn'' from these principles.
 In the words of Girard:
 \begin{quote}
-  In other terms, what is so good in logic that quantum physics should obey?
-  Can't we imagine that our conceptions about logic are wrong, so wrong that
-  they are unable to cope with the quantum miracle?
-  \\
-  Instead of teaching logic to nature, it is more reasonable to learn
-  from her. Instead of interpreting quantum into logic, we shall
-  interpret logic into quantum (Girard 2007).
+  In other terms, what is so good in logic that quantum physics should
+  obey?  Can't we imagine that our conceptions about logic are wrong,
+  so wrong that they are unable to cope with the quantum miracle?
+  [\ldots] Instead of teaching logic to nature, it is more reasonable
+  to learn from her. Instead of interpreting quantum into logic, we
+  shall interpret logic into quantum (Girard 2007).
 \end{quote}
 
-\begin{comment}
-\jc{The reason I commented this out is that it is under-justified. The
-reader will simply not understand what these next few lines are really
-saying.}
-Indeed one should take the physical principles underlying quantum mechanics,
-the most successful physical theory known to us and adapt computation to
-``learn'' from these principles. To illustrate the depth of our crisis, Scott
-Aaronson, Umesh Vazirani, and others have proposed the following puzzle.
-
-One of these wild claims must be true!:
-\begin{itemize}
-\item the extended Church-Turing thesis is false, or
-\item quantum physics is false, or
-\item there is an efficient classical algorithm for factoring
-\end{itemize}
-Indeed, if quantum physics is correct then there is an efficient quantum
-algorithm for factoring (Shor). If there is no efficient classical algorithm
-for factoring then the extended Church-Turing thesis is false.
-\end{comment}
+There are, in fact, many different principles of physics which are at
+odds with our current models of computation. In this paper, we will
+view data as an explicit representation of \emph{information},
+programs as processes that transform information in a reversible way,
+i.e., processes that are subject to the physical principle of
+\emph{conservation of information.} We will formalize this idea and
+follow its consequences, which will turn out to be far reaching.
 
 %%%%%%%%%
-\paragraph*{From Classical to Quantum.}
-Conventional classical models of computation, including boolean
-circuits, the Turing machine, and the $\lambda$-calculus, are founded on
-primitives which correspond to \emph{irreversible} physical processes which
-gratuitously erase information.  For example, a \emph{nand} gate is an
-irreversible logical operation in the sense that its inputs cannot generally
-be recovered from observing its output, and so is the operation of overriding
-a cell on a Turing machine tape with a new symbol, and so is a
-$\beta$-reduction which typically erases or duplicates values in a way that
-is destructive and irreversible.
-
-These irreversible abstractions chosen as the basis of our
-models of computing are at odds with the underlying
-reversible physical reality. The information theoretic underpinnings
-of physics is best understood through \emph{Landauer's
-principle}~\cite{Landauer:1961}. As mentioned previously in the
-introduction, Landauer, through his analysis of
-Maxwell's Demon, established the idea that \emph{information is a
-physical quantity} in much the sense that one would think of an
-electron as a physical quantity. More precisely, his principle implies
-that a certain minimum amount of
-thermodynamic work has to be carried out to erase one bit of
-information. Experimental verification of the Landauer principle
-happened in recent years~\cite{berut2012experimental}. If information
-has a physical significance, then it is subject to conservation laws
-just as mass and energy are.
-
-%%%%%%%%%
-\paragraph*{Conservation of Information.}
-To make this idea concrete and to provide a taste of the applications
-it opens, consider a tiny 2-bit password = \verb|"10"|. The password
-checker looks like:
-
-% verbatim for now, will go to something decent later
-\begin{verbatim}
-check-password (guess) =
-  if guess == "10"
-  then True
-  else False
-\end{verbatim}
-
-One can ask how much information is leaked by this program assuming
-the attacker has no prior knowledge except that the password is 2
-bits, i.e., the four possible 2-bits are equally likely. If the
-attacker guesses \verb|"10"| (with probability $1/4$) the password (2
-bits) is leaked. If the attacker guesses one of the other choices
-(with probability $3/4$) the number of possibilities is reduced from 4
-to 3, i.e., the attacker learns $\log{4} - \log{3}$ bits of
-information. So in general the attacker learns\footnote{If the
-  password is 8 restricted ASCII characters (6 bits), the attacker
-  learns 0.00001 bits in the first probe.}:
-\[\begin{array}{ll}
-   &  1/4 * 2 + 3/4 (\log{4} - \log{3}) \\
-  =&  1/4 \log{4} + 3/4 \log{4/3} \\
-  =&  - 1/4 \log{1/4} - 3/4 \log{3/4} \\
-  \sim& 0.8 \mbox{~bits~in~the~first~probe}
-\end{array}\]
-
-One can alternatively look at the situation by viewing the input as a
-random variable with 4 possibilities and a uniform distribution (i.e.,
-with 2 bits of information). The output is another random variable
-with 4 possibilities but with the distribution
-$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
-information. Thus 2 input bits of information were given to the
-function and only 0.8 were produced. Where did the 1.2 bits of
-information go? Once we accept the thesis that information is a
-physical entity this question cannot be ignored.
-
-The Landauer Principle states that erasing information requires
-energy. In other words, the password checking function must have
-erased 1.2 bits during its calculation, which dissipate as heat in any
-physical realization of the function. In conventional models of
-computation, this erasure is \emph{implicit}, i.e., it occurs as a
-side effect of calculation. We wish to look at models where
-erasure is simply not allowed.
-
-There are, in fact, many different principles of physics which are
-at odds with our current models of computation. Even within quantum
-mechanics itself, there are other ideas (such as superposition or
-entanglement) which are not present in current models. We will however
-restrict ourselves to exploring a single dimension in which our
-foundations should be revised: \textbf{conservation of information}.
-We will follow its consequences, which will turn out to be far reaching.
-
-\begin{quote}
-  The laws of physics are essentially algorithms for calculation. These
-  algorithms are significant only to the extent that they are executable in
-  our real physical world. Our usual laws of physics depend on the
-  mathematician's real number system. With that comes the presumption that
-  given any accuracy requirement, there exist a number of calculations steps,
-  which if executed, will satisfy that accuracy requirement. But the real
-  world is unlikely to supply us with unlimited memory of unlimited Turing
-  machine tapes, Therefore, continuum mathematics is not executable, and
-  physical laws which invoke that can not really be satisfactory. They are
-  references to illusionary procedures. Rolf Landauer, The physical nature of
-  information, Physics Letters A 217 (1996) pp. 188-193:
-\end{quote}
-
-A high level approach to conservation of information would proceed
-as follows:
-\begin{enumerate}
-\item Create a means to measure information.
-\item Show that this measure is sound with respect to our current understanding
-of what information ought to be.
-\item Assert that whatever ``quantity'' of information a system has, must
-remain invariant throughout the lifetime of the system.
-\end{enumerate}
-
-Certainly, whatever our notion of information, it can neither be created,
-duplicated, nor erased. Of course, if we make our notion of information
-too rigid, it will be impossible to even \textbf{modify} it! For computation
-to occur, modification must be possible.  But of what kind?
-
-\amr{We should also have a quick survey of other works in reversible
-computation? Most of the other work tries to start from well known
-models of computation or well known programming languages, and then
-adapts them to be reversible. Pi is different in that it starts from
-the semantic implications of reversibility. Then, because it adds
-types, rather than control-flow (or even data-flow) as its next layer
-of ``understanding'', this leads to equivalences, isomorphisms, etc.
-This then leads, quite naturally, to finding that the ``proof
-language'' of semirings (and Rig Groupoids at level 2) is actually a
-programming language. And it is Pi. This is a neat twist on
-Curry-Howard because CH is about \textbf{inhabitation} only. But with
-``conservation of information'' as the basis, a different kind of
-correspondance arises; in fact, this one may well be an actual
-isomorphism.}
-
-%%%%%%%%%
-\paragraph*{Information.}
-We can now give a first precise definition of information preservation.
-Let
-`$b$' be a (not necessarily finite) type whose values are labeled
-$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
-equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
-as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
-$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
-type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
-entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
-indicates the probability of the output of the function to have value
-${b_2}^j$. We say a function is \emph{information-preserving} if its
-output entropy is equal to the entropy of its input. In later sections,
-we will refine this view.
-
-\amr{
-Talk about conservation of information in biology, computer science,
-evolutionary computing, and search. (See
-\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
-}
-
-In our current age, both Landauer's Principle and recent investigations
-of computational models well-suited to quantum mechanics, push us to
-revisit our current models of computation. Information, on top of
-having a physical manifestation, appears to be \emph{conserved},
-along with other quantities like energy, mass and momentum.
-
-But what does it mean for a computation to preserve information?
-What is the missing principle that will guide the creation of a new
-model of computation that ``conserve information''?
-It is nature to look again at physics for inspiration --
-in our current understanding of physics all of the primitive rules
-are reversible. Conventional computation is not. Can
-we embrace this simple principle as the building block of a model of
-computation? What would computation look like when viewed from this
-vantage point? Can non-trivial computation be performed in a
-reversible manner? And, if so, does it have applications?
-
-Traditional models of computation are all unityped, generally
-operating on unstructured sequences of symbols. But recent
-investigations, in the intersection between mathematics and
-computer science (particularly in Homotopy Type Theory~\cite{hottbook})
-reveal that not only can mathematics be typed, but that doing so
-reveals further structure -- in both mathematics and in computation.
-
-And that brings us to the principal objective of this survey: explore the
-domain of \emph{typed reversible computation}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Programs as Reversible Deformations}
-
-To better understand conservation of information, we can look
-for analogous ideas outside of computing. Turning again to physics,
-but this time at the macro scale, what does it mean to transform an
-object in such a way that we do not lose the fundamental character
-of the object?
+\paragraph*{Programs as Reversible Deformations.}
+
+To better understand the essence of ``conservation of information'' in
+the context of computing, we first look for analogous ideas in
+physics, but this time at the macro scale. Viewing information as a
+physical object, what does it mean to transform an object in such a
+way that we do not lose its fundamental character?
 
@@
 translations and rotations. But what about something more flexible,
-such as a water balloon?  They can be \textbf{deformed} in various ways,
-but still retain their fundamental character -- as long as we do not
-puncture them or over-stretch them. Ignoring material characteristics
-(i.e. over-stretching), what is special about these deformations, as well
-as for translations and rotations, is that they correspond to
+with multiple representations, such as a water balloon?  Such objects
+can be \emph{deformed} in various ways, but still retain their
+fundamental character -- as long as we do not puncture them or
+over-stretch them. Ignoring material characteristics
+(i.e. over-stretching), what is special about these deformations, as
+well as for translations and rotations, is that they correspond to
 continuous maps, with a continuous inverse. In fact, even more is
 true: they are analytic maps, with analytic inverses. For our purpose,
-the most important part is that such maps are infinitely differentiable.
-In other words, not only is there an inverse to the deformation, but
-its derivative is also invertible, and so on.
+the most important part is that such maps are infinitely
+differentiable.  In other words, not only is there an inverse to the
+deformation, but its derivative is also invertible, and so on.
 
 When we look around, we find many different words for related
-concepts: isomorphism, equivalence, sameness, equality, interchangeability,
-comparability, and correspondence, to name a few. Some of these
-are informal concepts, while others have formal mathematical meaning.
-More important, even amongst the formal concepts, there are differences
--- which is why there are so many of them! Thus we seek a concept
-which is neither too strong nor too weak, that will express when
-some structured information should be treated as ``the same''.
-
-We will take ``the same'' as a fundamental principle. We want
-to then derive what it means for data, programs, program
-transformations, as well as proofs / deductions, to be ``the
-same'' -- in an manner consistent with preservation of 
-information.
-
-This stands in stark contrast with most current approaches to
-reversible computation, which start from current models of
-computation involving irreversible operations and try to find
-various ways to \emph{patch things up} so as to be reversible.
-
-The reader will have surely noticed how we have studiously avoided
-defining ``the same''. Because there are many such notions, we
-also need to walk our way through them to find the one which is
-``just right''. We can draw an analogy with topology: in topology,
-all point sets can always be equipped with either the discrete or
-the indiscrete topology, but both of these extremes are rarely
-useful. We will develop our working notion of ``sameness''
-as we go through the various components that make up a
-programming language.
-
-Starting from a physical perspective, whatever our notion of data
+concepts: isomorphism, equivalence, sameness, equality,
+interchangeability, comparability, and correspondence, to name a
+few. Some of these are informal concepts, while others have formal
+mathematical meaning.  More important, even amongst the formal
+concepts, there are differences -- which is why there are so many of
+them! Because there are many such notions, we also need to walk our
+way through them to find the one which is ``just right''. Thus we seek
+a concept which is neither too strong nor too weak, that will express
+when some structured information should be treated as ``the same''.
+We can draw an analogy with topology: in topology, all point sets can
+always be equipped with either the discrete or the indiscrete
+topology, but both of these extremes are rarely useful. We will
+develop our working notion of ``sameness'' as we go through the
+various components that make up a programming language.
+
+Starting from the physical perspective, whatever our notion of data
 is, we will be interested in programs as representing transformations
@@
 
-\amr{In the quantum case, this is clear. Say something about topological
-quantum computing.}
+To summarize, we will take ``the same'' as a fundamental principle and
+derive what it means for data, programs, program transformations, as
+well as proofs / deductions, to be ``the same'' -- in an manner
+consistent with preservation of information. This stands in stark
+contrast with most current approaches to reversible computation, which
+start from current models of computation involving irreversible
+operations and try to find various ways to \emph{patch things up} so
+as to be reversible.
 
@@
 
+Implement Fredkin gate and mention the following:
+
+\begin{quote}
+  Ed Fredkin pursued the idea that information must be finite in
+  density. One day, he announced that things must be even more simple
+  than that. He said that he was going to assume that information
+  itself is conserved. “You’re out of you mind, Ed.” I
+  pronounced. “That’s completely ridiculous. Nothing could happen in
+  such a world. There couldn’t even be logical gates. No decisions
+  could ever be made.” But when Fredkin gets one of his ideas, he’s
+  quite immune to objections like that; indeed, they fuel him with
+  energy. Soon he went on to assume that information processing must
+  also be reversible — and invented what’s now called the Fredkin
+  gate. (Minsky 1999)
+\end{quote}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements} We would like to thank the numerous
-students and colleagues who participated in various aspects of this
-research and who provided valuable feedack and constructive criticism.
-
-\bibliographystyle{acm}
-%\pagestyle{headings}
-\bibliography{cites}
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\jc{The reason I commented this out is that it is under-justified. The  
+reader will simply not understand what these next few lines are really
+saying.}
+Indeed one should take the physical principles underlying quantum mechanics,
+the most successful physical theory known to us and adapt computation to
+``learn'' from these principles. To illustrate the depth of our crisis, Scott
+Aaronson, Umesh Vazirani, and others have proposed the following puzzle.
+
+One of these wild claims must be true!:
+\begin{itemize}
+\item the extended Church-Turing thesis is false, or
+\item quantum physics is false, or
+\item there is an efficient classical algorithm for factoring
+\end{itemize}
+Indeed, if quantum physics is correct then there is an efficient quantum
+algorithm for factoring (Shor). If there is no efficient classical algorithm
+for factoring then the extended Church-Turing thesis is false.
 
@@
 
+%%%%%%%%%
 \subsection{Technical Overview}
@@
 with discussion of some of these, along with other possible
-applications (see Ch.~\ref{ch:discussion}).  Of these, the field of
-reversible and Quantum computing is closely related and we survey
-several connections in depth.  The exact connection between $\Pi$ and
-linear logic~\cite{Girard87tcs} is not fully understood though it is
-clear that both systems capture related but different notions of
-resource usage. A closely related area is the connection with duality
-of computation \cite{Filinski:89, DBLP:conf/icfp/CurienH00,
-  10.1109/LICS.2010.23,Wadler:2003} --- i.e., the quest for a unifying
-framework between `computations', values, continuations and their
-logical counterparts. The investigation of duality gives rise to
-negative and fractional types which have an intuitive semantics and
-are reminiscent of negative information flow, superposition and
-entanglement from Quantum Physics~\cite{piee}.
+applications.  Of these, the field of reversible and Quantum computing
+is closely related and we survey several connections in depth.  The
+exact connection between $\Pi$ and linear logic~\cite{Girard87tcs} is
+not fully understood though it is clear that both systems capture
+related but different notions of resource usage. A closely related
+area is the connection with duality of computation \cite{Filinski:89,
+  DBLP:conf/icfp/CurienH00, 10.1109/LICS.2010.23,Wadler:2003} ---
+i.e., the quest for a unifying framework between `computations',
+values, continuations and their logical counterparts. The
+investigation of duality gives rise to negative and fractional types
+which have an intuitive semantics and are reminiscent of negative
+information flow, superposition and entanglement from Quantum
+Physics~\cite{piee}.
+
+%%%%%%%%%
+\paragraph*{Conservation of Information.}
+
+If information has a physical significance, then it is subject to
+conservation laws just as mass and energy are.  To make this idea
+concrete and to provide a taste of the applications it opens, consider
+a tiny 2-bit password = \verb|"10"|. The password checker looks like:
+
+% verbatim for now, will go to something decent later
+\begin{verbatim}
+check-password (guess) =
+  if guess == "10"
+  then True
+  else False
+\end{verbatim}
+
+One can ask how much information is leaked by this program assuming
+the attacker has no prior knowledge except that the password is 2
+bits, i.e., the four possible 2-bits are equally likely. If the
+attacker guesses \verb|"10"| (with probability $1/4$) the password (2
+bits) is leaked. If the attacker guesses one of the other choices
+(with probability $3/4$) the number of possibilities is reduced from 4
+to 3, i.e., the attacker learns $\log{4} - \log{3}$ bits of
+information. So in general the attacker learns\footnote{If the
+  password is 8 restricted ASCII characters (6 bits), the attacker
+  learns 0.00001 bits in the first probe.}:
+\[\begin{array}{ll}
+   &  1/4 * 2 + 3/4 (\log{4} - \log{3}) \\
+  =&  1/4 \log{4} + 3/4 \log{4/3} \\
+  =&  - 1/4 \log{1/4} - 3/4 \log{3/4} \\
+  \sim& 0.8 \mbox{~bits~in~the~first~probe}
+\end{array}\]
+
+One can alternatively look at the situation by viewing the input as a
+random variable with 4 possibilities and a uniform distribution (i.e.,
+with 2 bits of information). The output is another random variable
+with 4 possibilities but with the distribution
+$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
+information. Thus 2 input bits of information were given to the
+function and only 0.8 were produced. Where did the 1.2 bits of
+information go? Once we accept the thesis that information is a
+physical entity this question cannot be ignored.
+
+The Landauer Principle states that erasing information requires
+energy. In other words, the password checking function must have
+erased 1.2 bits during its calculation, which dissipate as heat in any
+physical realization of the function. In conventional models of
+computation, this erasure is \emph{implicit}, i.e., it occurs as a
+side effect of calculation. We wish to look at models where
+erasure is simply not allowed.
+
+There are, in fact, many different principles of physics which are at
+odds with our current models of computation. Even within quantum
+mechanics itself, there are other ideas (such as superposition or
+entanglement) which are not present in current models. We will however
+restrict ourselves to exploring a single dimension in which our
+foundations should be revised: \textbf{conservation of information}.
+We will follow its consequences, which will turn out to be far
+reaching.
+
+\begin{quote}
+  The laws of physics are essentially algorithms for calculation. These
+  algorithms are significant only to the extent that they are executable in
+  our real physical world. Our usual laws of physics depend on the
+  mathematician's real number system. With that comes the presumption that
+  given any accuracy requirement, there exist a number of calculations steps,
+  which if executed, will satisfy that accuracy requirement. But the real
+  world is unlikely to supply us with unlimited memory of unlimited Turing
+  machine tapes, Therefore, continuum mathematics is not executable, and
+  physical laws which invoke that can not really be satisfactory. They are
+  references to illusionary procedures. Rolf Landauer, The physical nature of
+  information, Physics Letters A 217 (1996) pp. 188-193:
+\end{quote}
+
+A high level approach to conservation of information would proceed
+as follows:
+\begin{enumerate}
+\item Create a means to measure information.
+\item Show that this measure is sound with respect to our current understanding
+of what information ought to be.
+\item Assert that whatever ``quantity'' of information a system has, must
+remain invariant throughout the lifetime of the system.
+\end{enumerate}
+
+Certainly, whatever our notion of information, it can neither be created,
+duplicated, nor erased. Of course, if we make our notion of information
+too rigid, it will be impossible to even \textbf{modify} it! For computation
+to occur, modification must be possible.  But of what kind?
+
+\amr{We should also have a quick survey of other works in reversible
+computation? Most of the other work tries to start from well known
+models of computation or well known programming languages, and then
+adapts them to be reversible. Pi is different in that it starts from
+the semantic implications of reversibility. Then, because it adds
+types, rather than control-flow (or even data-flow) as its next layer
+of ``understanding'', this leads to equivalences, isomorphisms, etc.
+This then leads, quite naturally, to finding that the ``proof
+language'' of semirings (and Rig Groupoids at level 2) is actually a
+programming language. And it is Pi. This is a neat twist on
+Curry-Howard because CH is about \textbf{inhabitation} only. But with
+``conservation of information'' as the basis, a different kind of
+correspondance arises; in fact, this one may well be an actual
+isomorphism.}
+
+%%%%%%%%%
+\paragraph*{Information.}
+We can now give a first precise definition of information preservation.
+Let
+`$b$' be a (not necessarily finite) type whose values are labeled
+$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
+equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
+as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
+$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
+type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
+entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
+indicates the probability of the output of the function to have value
+${b_2}^j$. We say a function is \emph{information-preserving} if its
+output entropy is equal to the entropy of its input. In later sections,
+we will refine this view.
+
+\amr{
+Talk about conservation of information in biology, computer science,
+evolutionary computing, and search. (See
+\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
+}
+
+In our current age, both Landauer's Principle and recent investigations
+of computational models well-suited to quantum mechanics, push us to
+revisit our current models of computation. Information, on top of
+having a physical manifestation, appears to be \emph{conserved},
+along with other quantities like energy, mass and momentum.
+
+But what does it mean for a computation to preserve information?
+What is the missing principle that will guide the creation of a new
+model of computation that ``conserve information''?
+It is nature to look again at physics for inspiration --
+in our current understanding of physics all of the primitive rules
+are reversible. Conventional computation is not. Can
+we embrace this simple principle as the building block of a model of
+computation? What would computation look like when viewed from this
+vantage point? Can non-trivial computation be performed in a
+reversible manner? And, if so, does it have applications?
+
+Traditional models of computation are all unityped, generally
+operating on unstructured sequences of symbols. But recent
+investigations, in the intersection between mathematics and
+computer science (particularly in Homotopy Type Theory~\cite{hottbook})
+reveal that not only can mathematics be typed, but that doing so
+reveals further structure -- in both mathematics and in computation.
+
+And that brings us to the principal objective of this survey: explore the
+domain of \emph{typed reversible computation}.
+
+\amr{In the quantum case, this is clear. Say something about topological
+quantum computing.}
+
+\textbf{punch line: allowable operations that computers can perform
+  while embracing the laws of physics are operations that preserve
+  everthing in abstract space of ``data''; doughnut into cup kind of
+  transformations.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements} We would like to thank the numerous
+students and colleagues who participated in various aspects of this
+research and who provided valuable feedack and constructive criticism.
+
+\bibliographystyle{acm}
+%\pagestyle{headings}
+\bibliography{cites}
+\end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-12-28 survey/build.sh
added a build script

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+while inotifywait -e close_write survey.tex; do
+    pdflatex survey.tex;
+    bibtex survey.aux;
+    pdflatex survey.tex;
+    pdflatex survey.tex;
+done

### 2017-12-28 survey/survey.tex
imported some theseus notes

@@
 \usepackage{tikz-cd}
+\usepackage{listings}
+\usepackage{multicol}
+\usepackage{amsmath}
+\usepackage{stmaryrd}
 \usetikzlibrary{quotes}
@@
 
+\newcommand{\kw}[1]{{\scriptsize{\textbf{\texttt{#1}}}}}
+\newcommand{\ctr}[1]{{\scriptsize{\texttt{#1}}}}
+\newcommand{\inline}[3]{\ctr{#1} \textasciitilde \ctr{#2}:\ctr{#3}}
+
 \begin{document}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Theseus: Types and Simple Isomorphisms}
+\label{theseusI}
+
+We begin by presenting the core of Theseus which consists of evidently
+reversible functions on user-defined datatypes.  In subsequent discourse we
+will use the word `map' to refer to these logically reversible
+functions. Theseus has the same built-in types as~\ensuremath{\Pi^{o}} and as conventional
+functional programming languages: the empty type, the unit type, and sum and
+product types. All other types, including recursive types, are user-defined
+using a general mechanism for type declarations:
+
+\lstset{
+        language=haskell,
+ morekeywords={Bool,Nat,Tree},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+type Bool  =  (@\ctr{True}@) | (@\ctr{False}@)
+type Nat   =  0 | Succ Nat
+type Tree  =  Leaf Nat | Node Tree Tree
+ \end{lstlisting}
+%subcode source theseus.tex:347
+
+\noindent The declarations above witness isomorphisms between the set
+\kw{Bool} and the set constructed by the constants \ctr{True} and
+\ctr{False}, between the set \kw{Nat} and the set inductively constructed
+using {\scriptsize{0}} and \ctr{Succ}, and between the set \kw{Tree} and the
+set inductively constructed using \ctr{Leaf} and \ctr{Node}.  Thus, Theseus
+type definitions are similar to, say, Haskell, type definitions. Such
+definitions translate directly to the underlying \ensuremath{\Pi^{o}} using iso-recursive
+type definitions. For example, the above types translate as follows:
+\vspace{-20pt}
+\begin{small}
+\begin{multicols}{3}
+\[\begin{array}{rclr}
+\kw{Bool}   & = &  \mu  ~x.1 + 1 \\
+\kw{Nat}   & = &  \mu  ~x. 1 + x \\
+\kw{Tree}   & = &  \mu  ~x. \kw{Nat}  + x * x \\
+ \end{array}\]
+%subcode source theseus.tex:363
+
+\[\begin{array}{rclr}
+\mathit{unfoldBool} :& \kw{Bool}  \leftrightarrow 1 + 1 &: \mathit{foldBool} \\
+\mathit{unfoldNat} :& \kw{Nat}  \leftrightarrow 1 + \kw{Nat}  &: \mathit{foldNat} \\
+\mathit{unfoldTree} :& \kw{Tree}  \leftrightarrow \kw{Nat}  + (\kw{Tree}  * \kw{Tree} ) &: \mathit{foldTree} \\
+ \end{array}\]
+%subcode source theseus.tex:368
+\end{multicols}
+\end{small}
+
+%%%%%%%%%%%%
+\subsection{Simple Isomorphisms}
+
+A Theseus program has a type of the form \ctr{a} $\leftrightarrow$ \ctr{b};
+one runs such a program forward by supplying a value of type \ctr{a}. If the
+program terminates, we get back a value of type~\ctr{b}. Alternately we can
+run the program in reverse by supplying a value of type \ctr{b}. Given the
+built-in types and the facility to introduce user-defined ones, one can
+already write a few small but interesting programs. Maps that are evidently
+isomorphisms can be written in the familiar pattern matching style
+popularized by the family of functional languages. For example:
+\begin{multicols}{2}
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+(@\ctr{id}@) :: Bool (@$\leftrightarrow$@) Bool
+| (@\ctr{False}@)  (@$\leftrightarrow$@)  (@\ctr{False}@)
+| (@\ctr{True}@)   (@$\leftrightarrow$@)  (@\ctr{True}@)
+
+expandBool :: Bool * a (@$\leftrightarrow$@) a + a
+| (@\ctr{True}@), a   (@$\leftrightarrow$@)  (@\ctr{Left}@) a
+| (@\ctr{False}@), a  (@$\leftrightarrow$@)  (@\ctr{Right}@) a
+
+expandNat :: Nat (@$\leftrightarrow$@) Nat + 1
+| 0       (@$\leftrightarrow$@)  (@\ctr{Right}@) ()
+| Succ n  (@$\leftrightarrow$@)  (@\ctr{Left}@) n
+
+(@\ctr{not}@) :: Bool (@$\leftrightarrow$@) Bool
+| (@\ctr{True}@)   (@$\leftrightarrow$@)  (@\ctr{False}@)
+| (@\ctr{False}@)  (@$\leftrightarrow$@)  (@\ctr{True}@)
+
+foldBool :: a + a (@$\leftrightarrow$@) Bool * a
+| (@\ctr{Left}@) a   (@$\leftrightarrow$@)  (@\ctr{True}@), a
+| (@\ctr{Right}@) a  (@$\leftrightarrow$@)  (@\ctr{False}@), a
+
+foldNat :: Nat + 1 (@$\leftrightarrow$@) Nat
+| (@\ctr{Right}@) ()  (@$\leftrightarrow$@)  0
+| (@\ctr{Left}@) n    (@$\leftrightarrow$@)  Succ n
+ \end{lstlisting}
+%subcode source theseus.tex:410
+\end{multicols}
+
+\vspace{-25pt}
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+treeUnwind :: Tree (@$\leftrightarrow$@) Tree * Tree + (Bool + Nat)
+| Node t1 t2            (@$\leftrightarrow$@)  (@\ctr{Left}@)   (t1, t2)
+| Leaf 0                (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Left}@) (@\ctr{True}@))
+| Leaf (Succ 0)         (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Left}@) (@\ctr{False}@))
+| Leaf (Succ (Succ n))  (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Right}@) n)
+ \end{lstlisting}
+%subcode source theseus.tex:422
+
+Unlike the situation in a conventional functional language, the
+intuition here is that the two sides of each pattern clause can be
+swapped to produce the inverse of the function, i.e. as we will see
+repeatedly in the paper, patterns and expressions are the same
+thing. Compare for example the definitions of \ctr{expandBool} and
+\ctr{foldBool}. This is the only constraint that a programmer needs to
+maintain. This constraint can be ensured using the following two
+rules:
+
+\begin{enumerate}
+\label{rules}
+\item \emph{Non-overlapping and exhaustive coverage in pattern
+  clauses.} The collections of patterns in the left-hand side (LHS) of
+  each clause must be a complete non-overlapping covering of the input
+  type. Similarly, the collections of patterns in the right-hand side
+  (RHS) of each clause must also be a complete non-overlapping
+  covering of the return type. In other words, no cases should be
+  omitted or duplicated.
+\item \emph{Preserve typed variables across \ensuremath{\leftrightarrow}.} Each variable that
+  occurs on one side of a pattern matching clause can only appear once on
+  that side and must appear exactly once on the other side and with the same
+  type.
+\end{enumerate}
+
+When restricted to one side of each pattern matching clause, the rules
+should be familiar and intuitive. Note that, in general, the two sides
+of a pattern matching clause may have different types. Other than the
+requirement of non-overlapping and exhaustive coverage, there are no
+rules that constrain the types and names of constants nor of
+constructors. So for example, in \ctr{foldBool}, the pattern
+`\ctr{Left}~\ctr{a}' maps to a pair constructor with the first
+component being the constant \ctr{True}.
+
+Examples of expressions that violate the constraints follow:
+
+\begin{multicols}{2}
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+-- Invalid: Missing LHS cases
+missing_node :: Tree (@$\leftrightarrow$@) Nat
+| Leaf n (@$\leftrightarrow$@) n
+
+-- Invalid: Overlapping RHS cases
+overlapping_cases :: Nat (@$\leftrightarrow$@) Nat
+| 0       (@$\leftrightarrow$@)  0
+| Succ n  (@$\leftrightarrow$@)  n
+ \end{lstlisting}
+%subcode source theseus.tex:471
+
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+-- Invalid: n is dropped
+drop_var :: Tree (@$\leftrightarrow$@) Tree * Tree + Nat
+| Node t1 t2  (@$\leftrightarrow$@)  (@\ctr{Left}@)  (t1,t2)
+| Leaf n      (@$\leftrightarrow$@)  (@\ctr{Right}@) 0
+
+-- Invalid: t is used twice
+dup_var :: Tree (@$\leftrightarrow$@) Tree * Tree + Nat
+| Node t t  (@$\leftrightarrow$@)  (@\ctr{Left}@)  (t,t)
+| Leaf n    (@$\leftrightarrow$@)  (@\ctr{Right}@) n
+ \end{lstlisting}
+%subcode source theseus.tex:485
+\end{multicols}
+
+\noindent It is easy to see that every primitive isomorphism of \ensuremath{\Pi^{o}} can
+be expressed in Theseus. Here are a few examples.
+\vspace{-10pt}
+\begin{multicols}{2}
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+assocL :: a+(b+c) (@$\leftrightarrow$@) (a+b)+c :: assocR
+| (@\ctr{Left}@) a          (@$\,\,\leftrightarrow$@) (@\ctr{Left}@) ((@\ctr{Left}@) a)
+| (@\ctr{Right}@) ((@\ctr{Left}@) b)   (@$\leftrightarrow$@) (@\ctr{Left}@) ((@\ctr{Right}@) b)
+| (@\ctr{Right}@) ((@\ctr{Right}@) c)  (@$\leftrightarrow$@) (@\ctr{Right}@) c
+ \end{lstlisting}
+%subcode source theseus.tex:499
+
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+zeroe :: b + 0 (@$\leftrightarrow$@) b :: zeroi
+| (@\ctr{Left}@) v (@$\leftrightarrow$@) v
+
+swapTimes :: a * b (@$\leftrightarrow$@) b * a
+| (a,b) (@$\leftrightarrow$@) (b,a)
+ \end{lstlisting}
+%subcode source theseus.tex:509
+\end{multicols}
+
+\label{translation}
+\noindent
+Every simple isomorphism of Theseus can be translated to \ensuremath{\Pi^{o}} as follows:
+\begin{enumerate}
+\item For every Theseus map \ctr{c : a} $\leftrightarrow$ \ctr{b}, we can
+  define an intermediate type \ctr{ti} which is the sum of pairs of all the
+  variables that occur in the LHS patterns. Since the LHS and RHS patterns
+  contain the same type variables, \ctr{ti} is unique for a given \ctr{c} (up
+  to ordering). For example, in \ctr{treeUnwind} the first clause has
+  variables \lstinline{t1 : Tree} and \lstinline{t2 : Tree}, the second and
+  third clauses have no variables and the fourth clause has the variable
+  \lstinline{n : Nat}. The expected type \ctr{ti} is therefore
+  \lstinline{Tree * Tree + (1 + (1 + Nat))}, where we insert the type
+  {\scriptsize{1}} as a placeholder for the clauses that did not have type
+  variables.
+\item By construction \ctr{ti} is isomorphic to \ctr{a} and \ctr{b}. Hence,
+  maps \ctr{c1 : a} $\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti}
+  $\leftrightarrow$ \ctr{b} must be expressible in \ensuremath{\Pi^{o}}. (Both the above
+  isomorphisms can be constructed by systematically unfolding and
+  distributing \ctr{a} and \ctr{b} until \ctr{ti} is obtained.  The exact
+  details of such an expansion can be found in Sec.~3 of our previous
+  paper~\cite{rc2012} and are skipped in this paper.) Once values of type
+  \ctr{a} and \ctr{b} can be mapped to values of type \ctr{ti}, the required
+  translation of \ctr{c} is given by \ctr{c1} $\fatsemi$ \ctr{c2}.
+\end{enumerate}
+
+%%%%%%%%%%%%
+\subsection{Dealing with Numbers}
+
+If we try to do arithmetic using the previously defined Peano-style \ensuremath{\kw{Nat} },
+we have to address an issue of what to do when we encounter
+\lstinline{sub1 0}. There are two obvious choices:
+\begin{enumerate}
+\item We can define \ctr{add1} and \ctr{sub1} of type \kw{Nat}
+  $\leftrightarrow$ \kw{Nat}, such that \ctr{sub1} of {\scriptsize{0}}
+  diverges. We show how these maps can be defined in Sec.~\ref{add1}.
+\item An error mechanism can be added to the language such that when
+  an error is raised program execution is undefined and there is no
+  meaningful inverse definition.
+\end{enumerate}
+Another choice, similar to that made by Janus, is to use bounded integers
+such that every operation is always well defined through underflows and
+overflows. Here is a simple 4-bit \kw{Nat4} datatype with its corresponding
+\ctr{add1} and \ctr{sub1} operations:
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool,Nat4},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+type Nat4 = Bool * Bool * Bool * Bool
+
+add1 :: Nat4 (@$\leftrightarrow$@) Nat4 :: sub1
+| (a, b, c, (@\ctr{False}@))          (@$\leftrightarrow$@)  (a, b, c, (@\ctr{True}@))
+| (a, b, (@\ctr{False}@), (@\ctr{True}@))       (@$\,\leftrightarrow$@)  (a, b, (@\ctr{True}@), (@\ctr{False}@))
+| (a, (@\ctr{False}@), (@\ctr{True}@), (@\ctr{True}@))    (@$\,\,\leftrightarrow$@)  (a, (@\ctr{True}@), (@\ctr{False}@), (@\ctr{False}@))
+| ((@\ctr{False}@), (@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@))  (@$\leftrightarrow$@)  ((@\ctr{True}@), (@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@))
+| ((@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@))   (@$\leftrightarrow$@)  ((@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@))
+ \end{lstlisting}
+%subcode source theseus.tex:567
+
+\noindent It is easy to see how a math library may be defined in this
+way. Once an operator can be defined in Theseus it is always possible to
+replace its implementation by an equivalent one that is expressible
+efficiently in the underlying hardware. In this case one could compile down
+to the CPU's native integer representation and operators.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Parametrized Maps}
+\label{theseusII}
+
+Now that we can express simple standalone isomorphisms, we explain how to
+compose such isomorphisms to model more complex behavior. In \ensuremath{\Pi^{o}}, there
+are three ways of composing isomorphisms: sequential composition, parallel
+composition, and choice composition. The common idiom underlying these
+composition combinators is that a reversible map can be parametrized by
+another reversible map. This idea is related to ``higher-order functions''
+but is more limited as we explain below.
+
+%%%%%%%%%%%%
+\subsection{Definition and Examples}
+
+A Theseus map \ctr{f} can be parametrized by another map \ctr{g} by adding a
+labeled argument~\ctr{g} of the appropriate type to \ctr{f}. In the example
+below \ctr{treeUnwindf} is parametrized over some map \ctr{f} which it
+applies to \ctr{n}, if the supplied tree is \lstinline{Leaf n}:
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+treeUnwindf :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
+| Node t1 t2 (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
+| Leaf n     (@$\leftrightarrow$@) (@\ctr{Right}@) (f n)
+ \end{lstlisting}
+%subcode source theseus.tex:600
+
+\noindent This parametrization should be thought of as a macro or a
+meta-language construction. Theseus does not have high-order maps in the
+formal sense. In other words, the final type of a Theseus program must be of
+the form \ctr{a} $\leftrightarrow$ \ctr{b} and every occurrence of an arrow
+type $\rightarrow$ must be instantiated at compile time. The notation for
+instantiating the label parameter \ctr{f} in the map \ctr{fun} by the map
+\ctr{g} is \inline{fun}{f}{g}. For example,
+\inline{treeUnwindf}{f}{expandNat} should be thought of as shorthand for the
+map:
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+_ :: Tree (@$\leftrightarrow$@) Tree * Tree + (1 + Nat)
+| Node t1 t2    (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
+| Leaf 0        (@$\leftrightarrow$@) (@\ctr{Right}@) ((@\ctr{Right}@) ())
+| Leaf (Succ n) (@$\leftrightarrow$@) (@\ctr{Right}@) ((@\ctr{Left}@) n)
+ \end{lstlisting}
+%subcode source theseus.tex:618
+
+\noindent which inlines \ctr{expandNat} within the definition of
+\ctr{treeUnwindf}.
+
+We discuss the reversibility of parametrized maps in the next section. We
+first point out that the \ensuremath{\Pi^{o}} closure primitives can be expressed as
+parametrized maps:
+\lstset{
+        language=haskell,
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+_._ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(b (@$\leftrightarrow$@) c) (@$\rightarrow$@) a (@$\leftrightarrow$@) c
+| a (@$\leftrightarrow$@) g (f a)
+
+_*_ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(c (@$\leftrightarrow$@) d) (@$\rightarrow$@) a * c (@$\leftrightarrow$@) b * d
+| (a,c) (@$\leftrightarrow$@) (f a , g c)
+
+_+_ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(c (@$\leftrightarrow$@) d) (@$\rightarrow$@) a + c (@$\leftrightarrow$@) b + d
+| (@\ctr{Left}@) a   (@$\leftrightarrow$@)  (@\ctr{Left}@) (f a)
+| (@\ctr{Right}@) c  (@$\leftrightarrow$@)  (@\ctr{Right}@) (g c)
+ \end{lstlisting}
+%subcode source theseus.tex:637
+
+As an additional example, parametrized maps allow us to define controlled
+operations in the tradition of reversible circuits. In the definition below,
+the boolean input is a control bit that determines which of the maps \ctr{th}
+or \ctr{el} is applied to the second component of the pair. This conditional
+is then used to define the ``controlled-not'' gate:
+
+\lstset{
+        language=haskell,
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+(@\ctr{if}@) :: th:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) el:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) Bool * a (@$\leftrightarrow$@) Bool * b
+| (@\ctr{True}@), a  (@$\leftrightarrow$@) (@\ctr{Left}@) (th a)
+| (@\ctr{False}@), a (@$\leftrightarrow$@) (@\ctr{Right}@) (el a)
+
+cnot :: Bool * Bool (@$\leftrightarrow$@) Bool * Bool
+| control, bit (@$\leftrightarrow$@) if ~th:(@\ctr{not}@) ~el:(@\ctr{id}@) (control, bit)
+ \end{lstlisting}
+%subcode source theseus.tex:653
+
+%%%%%%%%%%%%
+\subsection{Reversing Parametrized Maps}
+
+Reversing a map of type \ctr{a} $\leftrightarrow$ \ctr{b} results in a map of
+type \ctr{b} $\leftrightarrow$ \ctr{a} which is its adjoint. As discussed
+before, simple maps can be reversed by simply switching the LHS and RHS for
+both the type and the pattern matching clauses comprising the
+body. Interestingly, one can think about the reverse of parametrized maps in
+the same way -- i.e. one simply needs to switch LHS and RHS of the `\ensuremath{\leftrightarrow}'.
+
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+treeUnfold :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
+| Node t1 t2  (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
+| Leaf n      (@$\leftrightarrow$@) (@\ctr{Right}@) (f n)
+
+-- the adjoint of treeUnfold
+rev_treeUnfold :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree * Tree + a (@$\leftrightarrow$@) Tree
+| (@\ctr{Left}@) (t1, t2)  (@$\leftrightarrow$@) Node t1 t2
+| (@\ctr{Right}@) (f n)    (@$\leftrightarrow$@) Leaf n
+ \end{lstlisting}
+%subcode source theseus.tex:676
+
+\noindent Note that the application \lstinline{f n} happens on the left of
+the \ensuremath{\leftrightarrow}, which is something that one does not see in conventional
+functional languages. These applications in the LHS have the dual meaning of
+applications in the RHS and should be understood as follows: the value that
+is the actual argument of the \ctr{Right} constructor is the result of
+applying~\lstinline{f n}. In the forward execution of
+\lstinline{rev_treeUnfold}, when the actual value of the constructor argument
+is encountered, it is passed through the reverse of \ctr{f} to get the value
+\ctr{n}. In other words, the above program is exactly the same as
+\lstinline{rev_treeUnfold'} below where \lstinline{rev_f} is the adjoint of
+\ctr{f}:
+
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+rev_treeUnfold' :: rev_f:(a (@$\leftrightarrow$@) Nat) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
+| (@\ctr{Left}@) (t1, t2) (@$\leftrightarrow$@) Node t1 t2
+| (@\ctr{Right}@) a       (@$\leftrightarrow$@) Leaf (rev_f a)
+ \end{lstlisting}
+%subcode source theseus.tex:696
+
+\noindent
+It follows that the following programs are equivalent:
+\vspace{-10pt}
+\begin{multicols}{2}
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+g :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) a (@$\leftrightarrow$@) b
+| a (@$\leftrightarrow$@) f a
+
+g :: rev_f:(b (@$\leftrightarrow$@) a) (@$\rightarrow$@) a (@$\leftrightarrow$@) b
+| rev_f b (@$\leftrightarrow$@) b
+ \end{lstlisting}
+%subcode source theseus.tex:710
+\end{multicols}
+\vspace{-15pt}
+\noindent
+Based on the above observation, it is easy to define the \ctr{sym} map that
+constructs the adjoint of any given map:
+\vspace{-5pt}
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Int,Bool},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+sym :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) b (@$\leftrightarrow$@) a
+| (f a) (@$\leftrightarrow$@) a
+ \end{lstlisting}
+%subcode source theseus.tex:722
+While parametrized maps add tremendous programming convenience to Theseus,
+they don't change the expressive power of the language. All programs
+expressible with parametrized maps, can be expressed without them by fully
+inlining the actual parameters. Thus compiling Theseus programs with
+parametrized maps to \ensuremath{\Pi^{o}} simply requires first inlining the maps and then
+doing the translation described in Sec.~\ref{translation}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Iteration Labels}
+\label{theseusIII}
+\label{add1}
+
+The fragment of Theseus discussed up to now, modulo full recursive types, can
+be expressed in pure \ensuremath{\Pi } without \ensuremath{\mathit{trace}}.  Introducing recursive
+definitions in a reversible language is subtle because every iteration must
+be reversible which requires, at least, that the number of iterations be
+recoverable from the output. The insight we use in Theseus is that this can
+be achieved elegantly by adding \emph{typed labels} as explained below. We
+start with a small example:
+
+\lstset{
+        language=haskell,
+ morekeywords={iter,Nat},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+parity :: Nat * Bool (@$\leftrightarrow$@) Nat * Bool
+| n, b                (@$\leftrightarrow$@) iter $ n, 0, b
+| iter $ Succ n, m, b (@$\leftrightarrow$@) iter $ n, Succ m, (@\ctr{not}@) b
+| iter $ 0, m, b      (@$\leftrightarrow$@) m, b
+  where iter :: Nat * Nat * Bool
+ \end{lstlisting}
+%subcode source theseus.tex:751
+
+\noindent The map \ctr{parity} calculates whether the given natural number is
+even or odd by counting to {\scriptsize{0}} and flipping the second boolean
+input each time. In the definition above, the label appears twice on the LHS
+and twice on the RHS. The correctness (i.e., reversibility) of the map is
+guaranteed as labels obey the same restrictions as the one for simple
+patterns. In particular the occurrences of the label in the LHS must
+constitute a non-overlapping coverage of the label type, and similarly for
+the occurrences of the label in the RHS.
+
+The important thing to note about labels is that they may have a
+different type from the LHS or RHS types of their containing map. In
+this case for example, the \ctr{parity} map has \lstinline{Nat * Bool}
+as its LHS and RHS type while the type of the label \kw{iter} has the
+type \lstinline{Nat * Nat * Bool}. Labels temporarily give us a
+different \emph{view}~\cite{Wadler:1987:VWP:41625.41653} of a value,
+by changing it type. Labels act as goto statements from one side to
+the other -- when a label is encountered on the RHS of a map, the
+argument to the label is matched by some definition of the label
+pattern on the LHS of the map, and vice versa.  Here is a trace of the
+execution of \ctr{parity} as it transforms input \ctr{Succ (Succ (Succ
+  0)), False} to the output \ctr{Succ (Succ (Succ 0)), True}. It takes
+five pattern match and rewrite steps to complete.
+
+\lstset{
+        language=haskell,
+ morekeywords={iter,Nat},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+Succ (Succ (Succ 0)), False           (@$\longmapsto$@) iter $ Succ (Succ (Succ 0)), 0, False
+iter $ Succ (Succ (Succ 0)), 0, False (@$\longmapsto$@) iter $ Succ (Succ 0), Succ 0, True
+iter $ Succ (Succ 0), Succ 0, True    (@$\longmapsto$@) iter $ Succ 0, Succ (Succ 0), False
+iter $ Succ 0, Succ (Succ 0), False   (@$\longmapsto$@) iter $ 0, Succ (Succ (Succ 0)), True
+iter $ 0, Succ (Succ (Succ 0)), True  (@$\longmapsto$@) Succ (Succ (Succ 0)), True
+ \end{lstlisting}
+%subcode source theseus.tex:784
+
+A trace operator can be expressed using labels and the definition closely
+follows the expected iteration semantics of \ensuremath{\mathit{trace}} in our previous
+work~\cite{James:2012:IE:2103656.2103667}.  Using \ctr{trace} one can define
+\ctr{add1} on \kw{Nat} such that \ctr{sub1} of {\scriptsize{0}}
+diverges.
+
+\begin{multicols}{2}
+\lstset{
+        language=haskell,
+ morekeywords={enter,leave},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+trace :: f:(a + b (@$\leftrightarrow$@) a + c) (@$\rightarrow$@) b (@$\leftrightarrow$@) c
+| b               (@$\leftrightarrow$@) enter $ (@\ctr{Right}@) b
+| enter $ a       (@$\leftrightarrow$@) leave $ (f a)
+| leave $ (@\ctr{Left}@) a  (@$\leftrightarrow$@) enter $ (@\ctr{Left}@) a
+| leave $ (@\ctr{Right}@) c (@$\leftrightarrow$@) c
+  where enter :: a + b
+        leave :: a + c
+ \end{lstlisting}
+%subcode source theseus.tex:803
+
+\lstset{
+        language=haskell,
+ morekeywords={Nat},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+addSub :: Nat + Nat (@$\leftrightarrow$@) Nat + Nat
+| (@\ctr{Left}@) (Succ n) (@$\leftrightarrow$@) (@\ctr{Left}@) n
+| (@\ctr{Left}@) 0        (@$\leftrightarrow$@) (@\ctr{Right}@) 0
+| (@\ctr{Right}@) n       (@$\leftrightarrow$@) (@\ctr{Right}@) (Succ n)
+
+add1 :: Nat (@$\leftrightarrow$@) Nat :: sub1
+| n (@$\leftrightarrow$@) trace ~f:addSub n
+ \end{lstlisting}
+%subcode source theseus.tex:815
+\end{multicols}
+\vspace{-10pt}
+
+Here are a few more examples that use labels.  Combinators \ctr{iterN} and
+\ctr{add} are ones you would expect to find in a standard library. The
+combinator \ctr{iterN} loops any given map \ctr{n} times and \ctr{add} adds
+the second argument to the first:
+\vspace{-10pt}
+\begin{multicols}{2}
+\lstset{
+        language=haskell,
+ morekeywords={Nat,iter},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+-- runs f some n-times
+iterN :: f:(a(@$\leftrightarrow$@)a) (@$\rightarrow$@) Nat*a (@$\leftrightarrow$@) Nat*a
+| n, a                 (@$\leftrightarrow$@) iter $ n, 0, a
+| iter $ 0, n, a       (@$\leftrightarrow$@) n, a
+| iter $ Succ n, m, a  (@$\leftrightarrow$@)
+    iter $ n, Succ m, f a
+ where iter :: Nat * Nat * a
+
+-- add (x, y) = (x+y, x)
+add : Nat * Nat (@$\leftrightarrow$@) Nat * Nat
+| x, y (@$\leftrightarrow$@) iter $ y, 0, x
+| iter $ a, b, 0       (@$\leftrightarrow$@) a, b
+| iter $ a, b, Succ n  (@$\leftrightarrow$@)
+       iter $ add1 a, add1 b, n
+  where iter :: Nat * Nat * Nat
+ \end{lstlisting}
+%subcode source theseus.tex:844
+\end{multicols}
+
+\vspace{-20pt} As another example, here is definition of the Fibonacci
+function where \ctr{fibonacci} applied to \lstinline{(n, (1, 0))} returns
+\lstinline{(n, (x, y))} where \ctr{x} is the $(n+1)$-st Fibonacci number and
+\ctr{y} is the previous one:
+
+\lstset{
+        language=haskell,
+ morekeywords={Nat},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+swapAndAdd :: Nat * Nat (@$\leftrightarrow$@) Nat * Nat
+| x, y (@$\leftrightarrow$@) add (y, x)
+
+fibonacci :: Nat * (Nat * Nat) (@$\leftrightarrow$@) Nat * (Nat * Nat)
+| n, (x, y) (@$\leftrightarrow$@) iterN ~f:swapAndAdd (n, (x, y))
+ \end{lstlisting}
+%subcode source theseus.tex:860
+
+We conclude this section with a somewhat richer example of tree traversal
+that shows how we can work with general recursive types. Here we use two
+labels \kw{walk} and \kw{reconst} and one can verify that each label
+independently covers its type on both the LHS and the RHS. The combinator
+\ctr{treeWalk} walks down a tree and applies a given \ctr{f} to every leaf:
+
+\lstset{
+        language=haskell,
+ morekeywords={Nat,Tree,Ctxt,walk,reconst},
+	keywordstyle=\bfseries\ttfamily,
+	identifierstyle=\ttfamily,
+	commentstyle=\ttfamily,
+	stringstyle=\ttfamily,
+	showstringspaces=false,
+	basicstyle=\scriptsize\sffamily,
+	tabsize=2,
+	breaklines=false,
+	breakatwhitespace=false,
+	aboveskip={1.2\baselineskip},
+        columns=fixed,
+        extendedchars=true,
+}
+\begin{lstlisting}[escapeinside={(@}{@)}]
+type Ctxt = Empty | L Ctxt Tree | R Tree Ctxt
+
+treeWalk :: ~f:(Nat (@$\leftrightarrow$@) Nat) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree
+| tr                       (@$\leftrightarrow$@) walk $ tr, Empty
+| walk $ Leaf b, ctxt      (@$\leftrightarrow$@) reconst $ ctxt, Leaf (f b)
+| walk $ Node b1 b2, ctxt  (@$\leftrightarrow$@) walk $ b1, L ctxt b2
+| reconst $ Empty, tr      (@$\leftrightarrow$@) tr
+| reconst $ L ctxt b2, b1  (@$\leftrightarrow$@) walk $ b2, R b1 ctxt
+| reconst $ R b1 ctxt, b2  (@$\leftrightarrow$@) reconst $ ctxt, Node b1 b2
+  where walk     :: Tree * Ctxt
+        reconst  :: Ctxt * Tree
+ \end{lstlisting}
+%subcode source theseus.tex:882
+
+\noindent
+Compiling programs with labels to \ensuremath{\Pi^{o}} is a straightforward extension of
+compiling label-free programs. For any Theseus map \ctr{c : a}
+$\leftrightarrow$ \ctr{b} that uses an iteration label \kw{label} \ctr{: lab}
+we construct an intermediate type \ctr{ti} corresponding to the pattern
+clauses as before. We can then construct the isomorphisms \ctr{c1 : lab + a}
+$\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti} $\leftrightarrow$ \ctr{lab + b}
+since every LHS pattern clause corresponds to either \ctr{a} (the input) or
+\ctr{lab} (the label) and every RHS clause corresponds to either \ctr{b} (the
+output) or \ctr{lab} (the label). The resulting \ctr{c1} $\fatsemi$ \ctr{c2}
+has type \lstinline{lab+a} $\leftrightarrow$ \lstinline{lab+b} and the
+required compilation of \ctr{c} is given by \ensuremath{\mathit{trace}(} \ctr{c1} $\fatsemi$
+\ctr{c2} \ensuremath{)} \ctr{: a} $\leftrightarrow$ \ctr{b}. This extends to multiple
+by labels by constructing \lstinline{c1 : (lab1 + lab2 + ...) + a}
+$\leftrightarrow$ \ctr{ti} and \lstinline{c2 : ti} $\leftrightarrow$
+\lstinline{(lab1 + lab2 + ...) + b} to include the types of the additional
+labels.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+\label{conc}
+
+Theseus is as expressible as \ensuremath{\Pi^{o}} and preserves its properties. All
+\ensuremath{\Pi^{o}} programs are expressible as Theseus programs and vice-versa. In other
+words, \ensuremath{\Pi^{o}}, or dagger symmetric traced bimonoidal categories serve as a
+fully abstract model for Theseus. Despite being closely related to \ensuremath{\Pi^{o}},
+programmers can use Theseus without needing to understand \ensuremath{\Pi^{o}}, strings
+diagrams or category theory.
+
+Theseus admits non-terminating computations. Just the way that partial
+functions refer to maps that may not be defined on all inputs, the set
+of computations expressible in Theseus are referred to as
+\emph{partial isomorphisms}. As with \ensuremath{\Pi^{o}}, Theseus programs satisfy
+\emph{logical reversibility} (see Def.~\ref{logical-rev}).
+
+Many of the trimmings of more mainstream languages can be added to
+Theseus. Theseus could use an error reporting mechanism for programs to exit
+to without resorting to divergence. Theseus needs an FFI mechanism to take
+advantage of third-party libraries. We imagine that in such cases the
+programmer would write out an explicit extern declaration asserting the type
+that Theseus should assume the external library has. More generally, Theseus
+can adopt Agda's approach of piggy backing on top of Haskell. We can have an
+FFI that lets us drop into Haskell and use Haskell API. We could also import
+user-certified pairs of Haskell functions as isomorphisms into Theseus.
+
+There are few natural extensions to Theseus that require additional research,
+though preliminary investigation seems promising. Theseus currently treats
+all values the same and does not classify values as heap and garbage.
+Theseus can be extended with a framework for encapsulating effects. In other
+words, add the effectful arrow type \ensuremath{\leadsto} in a principled way such that
+information effects, and possibly other effects, may be
+expressed~\cite{James:2012:IE:2103656.2103667}.  It is also desirable to add
+other programming features such as polymorphism, higher-order maps and
+inductive definitions over recursive datatypes without explicitly managing
+\ctr{Ctxt} values as we did in the \ctr{treeWalk} program.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgements} We would like to thank the numerous

### 2017-12-28 survey/build.sh

@@ -1,8 +1,8 @@
 #!/bin/bash
 
 while inotifywait -e close_write survey.tex; do
-    pdflatex survey.tex;
+    pdflatex -halt-on-error survey.tex;
     bibtex survey.aux;
-    pdflatex survey.tex;
-    pdflatex survey.tex;
+    pdflatex -halt-on-error survey.tex;
+    pdflatex -halt-on-error survey.tex;
 done

### 2017-12-28 survey/survey.tex
data I small changes

--- 2017-12-26 survey/survey.tex
@@
 
-Before we start our investigation of what it means to have typed,
-information preserving transformations, let's begin with just
-finite sets. Most programming languages provide some collection of primitive
-data like booleans, characters, strings, and (bounded) numbers that are
-naturally modeled as sets. Infinite sets are much more subtle, and will be
-discussed in a later section.
-
-What does it mean for two finite sets to be ``the same''?  Well, clearly the
-sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$ are different.
-Why?  Well, supposed there was a transformation $f : A \rightarrow B$ that
-deformed $A$ into $B$, and another $g : B \rightarrow A$ which undid this
-transformation. Since $f$ is total, by the pigeonhole principle, two elements
-of $A$ would be mapped to the same element of $B$. Supposed that this is $2$ and
-$3$, and that they both map to $d$.  But $g(d)$ cannot be both $2$ and $3$, and
-so $g$ is not the inverse of $f$. With just a little more work, we can show
-that $f$ (and $g$) must be both injective and surjective. In other words,
-$f$ (and $g$) must be a bijection between $A$ and $B$.  And of course this 
-only happens when $A$ and $B$ have the same number of elements.
-
-More importantly, given a bijection $f : C \rightarrow D$ of finite sets
-$C,D$, there always exists another bijection $g : D \rightarrow C$ which is
-$f$'s inverse. So, for finite sets, \emph{bijections} act as reversible
+Some of the simplest spaces are the finite and discrete spaces of
+points. Indeed most programming languages provide some collection of
+primitive data like booleans, characters, strings, and (bounded)
+numbers that are naturally modeled as finite sets. Infinite sets are much
+more subtle, and will be discussed in the conclusion.
+
+What does it mean for two finite sets to be ``the same''?  Well,
+clearly the sets $A = \left\{1, 2, 3\right\}$ and
+$B = \left\{c, d\right\}$ are different.  Why?  Well, supposed there
+was a transformation $f : A \rightarrow B$ that deformed $A$ into $B$,
+and another $g : B \rightarrow A$ which undid this
+transformation. Since $f$ is total, by the pigeonhole principle, two
+elements of $A$ would be mapped to the same element of $B$. Supposed
+that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
+cannot be both $2$ and $3$, and so $g$ is not the inverse of $f$. With
+just a little more work, we can show that $f$ (and $g$) must be both
+injective and surjective. In other words, $f$ (and $g$) must be a
+bijection between $A$ and $B$.  And of course this only happens when
+$A$ and $B$ have the same number of elements. More importantly, given
+a bijection $f : C \rightarrow D$ of finite sets $C,D$, there always
+exists another bijection $g : D \rightarrow C$ which is $f$'s
+inverse. So, for finite sets, \emph{bijections} act as reversible
 deformations.
@@
 can create a small language for writing permutations as
-\begin{verbatim}
-p ::= id | swap i j | p;p
-\end{verbatim}
+\[\begin{array}{rcl}
+p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p;p
+  \end{array}\]
 where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
-\verb|id|
-from the language and drop the $i\neq j$ condition so that 
-\verb|swap j j| would represent the identity permutation.
-
-While we are mainly interested in typed data, it is always worthwhile to
-first start investigating the untyped setting first, as it gives us insights
-into what the \emph{purely operational} view of the theory would be.
-This tells us that permutations are an inescapable part of the fabric of
-reversible computing.
-
-However as permutations are untyped, and act on the canonicalized version
-of $n$-element sets (i.e. those sets where all the structure has been
-forgotten), these are a rather pale shadow of the rich tapestry of
-information-perserving transformation of structured data.
+$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
+$\mathit{swap}~j~j$ would represent the identity permutation.
+
+While we are mainly interested in typed data, it is always worthwhile
+to first start investigating the untyped setting first, as it gives us
+insights into what the \emph{purely operational} view of the theory
+would be.  This tells us that permutations are an inescapable part of
+the fabric of reversible computing.  However as permutations are
+untyped, and act on the canonicalized version of $n$-element sets
+(i.e. those sets where all the structure has been forgotten), these
+are a rather pale shadow of the rich tapestry of
+information-perserving transformation of structured data which we
+investigate next.
 

### 2017-12-28 survey/survey.tex

--- 2017-12-28 survey/survey.tex
@@
 
-Before we start our investigation of what it means to have typed,
-information preserving transformations, let's begin with just
-finite sets. Most programming languages provide some collection of primitive
-data like booleans, characters, strings, and (bounded) numbers that are
-naturally modeled as sets. Infinite sets are much more subtle, and will be
-discussed in a later section.
-
-What does it mean for two finite sets to be ``the same''?  Well, clearly the
-sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$ are different.
-Why?  Well, supposed there was a transformation $f : A \rightarrow B$ that
-deformed $A$ into $B$, and another $g : B \rightarrow A$ which undid this
-transformation. Since $f$ is total, by the pigeonhole principle, two elements
-of $A$ would be mapped to the same element of $B$. Supposed that this is $2$ and
-$3$, and that they both map to $d$.  But $g(d)$ cannot be both $2$ and $3$, and
-so $g$ is not the inverse of $f$. With just a little more work, we can show
-that $f$ (and $g$) must be both injective and surjective. In other words,
-$f$ (and $g$) must be a bijection between $A$ and $B$.  And of course this
-only happens when $A$ and $B$ have the same number of elements.
-
-More importantly, given a bijection $f : C \rightarrow D$ of finite sets
-$C,D$, there always exists another bijection $g : D \rightarrow C$ which is
-$f$'s inverse. So, for finite sets, \emph{bijections} act as reversible
+Some of the simplest spaces are the finite and discrete spaces of
+points. Indeed most programming languages provide some collection of
+primitive data like booleans, characters, strings, and (bounded)
+numbers that are naturally modeled as finite sets. Infinite sets are much
+more subtle, and will be discussed in the conclusion.
+
+What does it mean for two finite sets to be ``the same''?  Well,
+clearly the sets $A = \left\{1, 2, 3\right\}$ and
+$B = \left\{c, d\right\}$ are different.  Why?  Well, supposed there
+was a transformation $f : A \rightarrow B$ that deformed $A$ into $B$,
+and another $g : B \rightarrow A$ which undid this
+transformation. Since $f$ is total, by the pigeonhole principle, two
+elements of $A$ would be mapped to the same element of $B$. Supposed
+that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
+cannot be both $2$ and $3$, and so $g$ is not the inverse of $f$. With
+just a little more work, we can show that $f$ (and $g$) must be both
+injective and surjective. In other words, $f$ (and $g$) must be a
+bijection between $A$ and $B$.  And of course this only happens when
+$A$ and $B$ have the same number of elements. More importantly, given
+a bijection $f : C \rightarrow D$ of finite sets $C,D$, there always
+exists another bijection $g : D \rightarrow C$ which is $f$'s
+inverse. So, for finite sets, \emph{bijections} act as reversible
 deformations.
@@
 can create a small language for writing permutations as
-\begin{verbatim}
-p ::= id | swap i j | p;p
-\end{verbatim}
+\[\begin{array}{rcl}
+p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p;p
+  \end{array}\]
 where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
-\verb|id|
-from the language and drop the $i\neq j$ condition so that
-\verb|swap j j| would represent the identity permutation.
-
-While we are mainly interested in typed data, it is always worthwhile to
-first start investigating the untyped setting first, as it gives us insights
-into what the \emph{purely operational} view of the theory would be.
-This tells us that permutations are an inescapable part of the fabric of
-reversible computing.
-
-However as permutations are untyped, and act on the canonicalized version
-of $n$-element sets (i.e. those sets where all the structure has been
-forgotten), these are a rather pale shadow of the rich tapestry of
-information-perserving transformation of structured data.
+$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
+$\mathit{swap}~j~j$ would represent the identity permutation.
+
+While we are mainly interested in typed data, it is always worthwhile
+to first start investigating the untyped setting first, as it gives us
+insights into what the \emph{purely operational} view of the theory
+would be.  This tells us that permutations are an inescapable part of
+the fabric of reversible computing.  However as permutations are
+untyped, and act on the canonicalized version of $n$-element sets
+(i.e. those sets where all the structure has been forgotten), these
+are a rather pale shadow of the rich tapestry of
+information-perserving transformation of structured data which we
+investigate next.
 

### 2017-12-28 survey/survey.tex
cleanup intro

@@
 is that the next computing revolution is likely to be founded on
-revised models of computation that are designed to in harmony with the
-laws of physics.
+revised models of computation that are designed to in closer harmony
+with the laws of physics.
 
@@
 following hundreds of years of tradition, conventional operations such
-conjunction and disjunction are logically irreversible. These models
-were already radical enough, by being purely \emph{constructive}, in
-an era where classical mathematics, with its pervasive use of excluded
-middle and frequent use of the axiom of choice, did not overly worry
-about effective representation. Landauer's observation implied however
-that ideas in each field have consequences for the
+conjunction and disjunction lose information by logically
+irreversible. These models were already radical enough, by being
+purely \emph{constructive}, in an era where classical mathematics,
+with its pervasive use of excluded middle and frequent use of the
+axiom of choice, did not overly worry about effective
+representation. Landauer's observation implied however that ideas in
+each field have consequences for the
 other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,baez2011physics,dblp:conf/csfw/malacarias12}. To
@@
 \paragraph*{Origins of Computational Models.}
-Machines that ``compute'' are quite old. M\"{u}ller (1786) first
-conceived of the idea of a ``difference machine'', which Babbage
-(1819--1822) was able to construct. There are other computer
-precursors as well -- the first stored programs were actually for
-looms, most notably those of Bouchon (1725) which operated on a paper
-tape, and Jacquard (1804) which operated by chains of punched cards.
-Logical operations in high-level programming languages and in our
-hardware are based on Boolean Logic -- which really ought to be called
-\emph{Piercean logic}, as Boole had a rather different conception of
-logic which is rather far from our current models.  Nevertheless,
-going back to Boole's 1853 book entitled \emph{An Investigation of the
-  Laws of Thought, on which are Founded the Mathematical Theories of
-  Logic and Probabilities}. The opening sentence of Ch.~1 is:
+Our current logical operations in high-level programming languages and
+in our current hardware are both based on Boolean Logic -- which
+really ought to be called \emph{Piercean logic}, as Boole had a rather
+different conception of logic which is rather far from our current
+models.  Nevertheless, going back to Boole's 1853 book entitled
+\emph{An Investigation of the Laws of Thought, on which are Founded
+  the Mathematical Theories of Logic and Probabilities}. The opening
+sentence of Ch.~1 is:
 \begin{quote}
@@
 \end{quote}
-This ``law'' is reasonable in a classical world but is violated by
+This ``law'' is reasonable in a classical world but is violated by the
 postulates of quantum mechanics. Although a detailed historical
@@
 
-The 20th century witnessed the real advent of computer science as a
+Machines that ``compute'' are quite old. M\"{u}ller (1786) first
+conceived of the idea of a ``difference machine'', which Babbage
+(1819--1822) was able to construct. There are other computer
+precursors as well -- the first stored programs were actually for
+looms, most notably those of Bouchon (1725) which operated on a paper
+tape, and Jacquard (1804) which operated by chains of punched cards.
+But it was only in the 20th century that computer science emerged as a
 formal discipline. One of the pioneering works was Alan Turing's
@@
 then when both of the above assumptions are translated to the language
-of physics, they embody what is known as the ``Bekenstein bound:''
+of physics, they embody what is known as the ``Bekenstein bound'',
 which is an upper limit on the amount of information that can be
@@
 should convey the ideas that our theories of computation and
-complexity are based on some assumptions that Turing and others found
-reasonable in the 1930s.
+complexity are based on some physical assumptions that Turing and
+others found reasonable in the 1930s.
 
@@
 assumptions are based on a certain understanding of the laws of
-physics.
-
-Our understanding of physics has evolved tremendously since 1900!
-Thus it is time to revisit these abstractions, especially with respect
-to quantum mechanics.  Indeed one should take the physical principles
-underlying quantum mechanics, the most successful physical theory
-known to us and adapt computation to ``learn'' from these principles.
-In the words of Girard:
+physics. Our understanding of physics has evolved tremendously since
+1900!  Thus it is time to revisit these abstractions, especially with
+respect to quantum mechanics.  Indeed one should take the physical
+principles underlying quantum mechanics, the most successful physical
+theory known to us and adapt computation to ``learn'' from these
+principles.  In the words of Girard:
 \begin{quote}
@@
 
-There are, in fact, many different principles of physics which are at
-odds with our current models of computation. In this paper, we will
-view data as an explicit representation of \emph{information},
-programs as processes that transform information in a reversible way,
-i.e., processes that are subject to the physical principle of
+There are, in fact, many different quantum mechanical principles which
+are at odds with our current models of computation. In this paper, we
+will focus on the following particular aspect. We will view data as an
+explicit representation of \emph{information} and programs as
+processes that transform information in a reversible way, i.e.,
+processes that are subject to the physical principle of
 \emph{conservation of information.} We will formalize this idea and
@@
 few. Some of these are informal concepts, while others have formal
-mathematical meaning.  More important, even amongst the formal
+mathematical meaning.  More importantly, even amongst the formal
 concepts, there are differences -- which is why there are so many of
@@
 representations. So it is unclear what maps to what.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data IV: Circle ???}
+
+Would be interesting
 

### 2017-12-29 survey/survey.tex
a bit more on data I

@@
 
+\newtheorem{thm}{Theorem}[section]
 \newtheorem{defn}{Definition}[section]
@@
 transformation. Since $f$ is total, by the pigeonhole principle, two
-elements of $A$ would be mapped to the same element of $B$. Supposed
+elements of $A$ would be mapped to the same element of $B$. Suppose
 that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
@@
 \[\begin{array}{rcl}
-p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p;p
+p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p \fatsemi p
   \end{array}\]
@@
 
-While we are mainly interested in typed data, it is always worthwhile
-to first start investigating the untyped setting first, as it gives us
-insights into what the \emph{purely operational} view of the theory
-would be.  This tells us that permutations are an inescapable part of
-the fabric of reversible computing.  However as permutations are
-untyped, and act on the canonicalized version of $n$-element sets
-(i.e. those sets where all the structure has been forgotten), these
-are a rather pale shadow of the rich tapestry of
-information-perserving transformation of structured data which we
-investigate next.
+Although this language appears weak, it is universal for boolean
+combinational circuits. The result follows from the following
+construction of the Toffoli gate which is universal for combinational
+circuits. We write $[2^n]$ for the finite set representing $n$-bit
+words with the canonical ordering for binary numbers. Thus when $n=3$,
+the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to
+the 3-bit words $\{000,001,010,011,100,101,110,111\}$. The Toffoli
+gate is a reversible function $[2^3] \rightarrow [2^3]$ which swaps
+$110$ and $111$, i.e., it can be realized as $\mathit{swap}~6~7$.
+
+In programming practice we are however more interested in structured
+data which will be the subject on next section. What we learn from
+this short investigation using untyped flat sets is what the
+\emph{purely operational} view of the theory would be. In particular,
+it tells us that permutations are an inescapable part of the fabric of
+reversible computing.  However as permutations are untyped, and act on
+the canonicalized version of $n$-element sets (i.e. those sets where
+all the structure has been forgotten), these are a rather pale shadow
+of the rich tapestry of information-perserving transformation of
+structured data which we investigate next.
 

### 2017-12-29 survey/survey.tex
minor fix to sec 2
starting data ii

@@
 
-Bijections on \fin{n} have a specific name: permutations. As is well-known
-permutations can be generated by sequences of transpositions. Thus we
-can create a small language for writing permutations as
+Bijections on \fin{n} have a specific name: permutations. As is
+well-known permutations can be generated by sequential compositions of
+transpositions. Thus we can create a small language for writing
+permutations as
 \[\begin{array}{rcl}
@@
 combinational circuits. The result follows from the following
-construction of the Toffoli gate which is universal for combinational
-circuits. We write $[2^n]$ for the finite set representing $n$-bit
-words with the canonical ordering for binary numbers. Thus when $n=3$,
-the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to
-the 3-bit words $\{000,001,010,011,100,101,110,111\}$. The Toffoli
-gate is a reversible function $[2^3] \rightarrow [2^3]$ which swaps
-$110$ and $111$, i.e., it can be realized as $\mathit{swap}~6~7$.
+construction of the Toffoli gate and the fact that any combinational
+circuit can be realized using compositions of Toffoli gates. We write
+$[2^n]$ for the finite set representing $n$-bit words with the
+canonical ordering for binary numbers. Thus when $n=3$, the finite set
+has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to the 3-bit words
+$\{000,001,010,011,100,101,110,111\}$. The Toffoli gate is a
+reversible function $[2^3] \rightarrow [2^3]$ which swaps $110$ and
+$111$, i.e., it can be realized as $\mathit{swap}~6~7$.
 
@@
 all the structure has been forgotten), these are a rather pale shadow
-of the rich tapestry of information-perserving transformation of
+of the rich tapestry of information-perserving transformations of
 structured data which we investigate next.
@@
 
-The elementary building blocks of type theory are the empty type
-($\bot$), the unit type ($\top$), and the sum ($\uplus$) and product
-($\times$) types.  These have a nice correspondance (Curry-Howard)
-to logic:
+Instead of the spaces consisting solely of unstructured isolated
+points, we now investigate structured spaces built from sums and
+products of elementary spaces. This structure corresponds to the
+building blocks of type theory which are: the empty type
+($\mathbf{0}$), the unit type ($\mathbf{1}$), the sum type ($\uplus$),
+and the product ($\times$) type. Before getting into the formal
+theory, let's consider the space
+$(\mathbf{1} \uplus \mathbf{0}) \times (\mathbf{1} \uplus
+\mathbf{1})$. The subspace $(\mathbf{1} \uplus \mathbf{0})$ is the sum
+of the space $\mathbf{1}$ containing one element and the empty space
+$\mathbf{0}$. By construction this space is different from just the
+space $\mathbf{1}$ but one can argue that the spaces are equivalent in
+some sense.
+
+\amr{new wavefront for new pass}
+
+These have a nice correspondance (Curry-Howard) to logic:
 

### 2017-12-29 survey/survey.tex
Data II continuance

@@
 
-\jc{wavefront}
-
-Of course, one does not get a programming language with just (typed)
-constants!
-
 \begin{figure}[t]
@@
 \end{figure}
+
+Of course, one does not get a programming language with just typed
+constants! There is a need to perform multiple equivalences. There are
+in fact three ways to do this: sequential composition $\odot$, 
+choice composition $\oplus$ (sometimes called juxtaposition) and parallel
+composition $otimes$. See Figure~\ref{pi-combinators} for the signatures.
+
+$c_1 \odot c_2$ corresponds to performing $c_1$ first, then $c_2$, and
+is the usual notion of composition -- and corresponds to $\fatsemi$ of
+the language of the previous section. $c_1 \oplus c\_2$ chooses to
+perform $c_1$ or $c_2$ depending on whether the input is labelled
+$\textsf{inl}$ or $\textsf{inr}$ respectively. $c_1 \otimes c_2$
+operates on a product structure, and applies $c_1$ to the first component
+and $c_2$ to the second.
+
+\jc{wavefront}
 

### 2017-12-30 survey/survey.tex
smoothen transition in sec 3

@@
 
-What does it mean for two finite sets to be ``the same''?  Well,
-clearly the sets $A = \left\{1, 2, 3\right\}$ and
-$B = \left\{c, d\right\}$ are different.  Why?  Well, supposed there
-was a transformation $f : A \rightarrow B$ that deformed $A$ into $B$,
-and another $g : B \rightarrow A$ which undid this
-transformation. Since $f$ is total, by the pigeonhole principle, two
-elements of $A$ would be mapped to the same element of $B$. Suppose
-that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
-cannot be both $2$ and $3$, and so $g$ is not the inverse of $f$. With
-just a little more work, we can show that $f$ (and $g$) must be both
-injective and surjective. In other words, $f$ (and $g$) must be a
-bijection between $A$ and $B$.  And of course this only happens when
-$A$ and $B$ have the same number of elements. More importantly, given
-a bijection $f : C \rightarrow D$ of finite sets $C,D$, there always
-exists another bijection $g : D \rightarrow C$ which is $f$'s
-inverse. So, for finite sets, \emph{bijections} act as reversible
-deformations.
+What does it mean to deform a space of points? For example, what
+transformation can we do on a bag or marbles? Well, we can shuffle
+them around and that is the only transformation that will preserve the
+space. Turning to the mathematical abstraction as sets, we ask what
+does it mean for two finite sets to be ``the same''?  Well, clearly
+the sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$
+are different.  Why?  Well, supposed there was a transformation
+$f : A \rightarrow B$ that deformed $A$ into $B$, and another
+$g : B \rightarrow A$ which undid this transformation. Since $f$ is
+total, by the pigeonhole principle, two elements of $A$ would be
+mapped to the same element of $B$. Suppose that this is $2$ and $3$,
+and that they both map to $d$.  But $g(d)$ cannot be both $2$ and $3$,
+and so $g$ is not the inverse of $f$. With just a little more work, we
+can show that $f$ (and $g$) must be both injective and surjective. In
+other words, $f$ (and $g$) must be a bijection between $A$ and $B$.
+And of course this only happens when $A$ and $B$ have the same number
+of elements. More importantly, given a bijection $f : C \rightarrow D$
+of finite sets $C,D$, there always exists another bijection
+$g : D \rightarrow C$ which is $f$'s inverse. So, for finite sets,
+\emph{bijections} act as reversible deformations.
 
@@
 this short investigation using untyped flat sets is what the
-\emph{purely operational} view of the theory would be. In particular,
+
+'\emph{purely operational} view of the theory would be. In particular,
 it tells us that permutations are an inescapable part of the fabric of
@@
 and the product ($\times$) type. Before getting into the formal
-theory, let's consider the space
+theory, let's consider possible deformations on the space
 $(\mathbf{1} \uplus \mathbf{0}) \times (\mathbf{1} \uplus
-\mathbf{1})$. The subspace $(\mathbf{1} \uplus \mathbf{0})$ is the sum
-of the space $\mathbf{1}$ containing one element and the empty space
-$\mathbf{0}$. By construction this space is different from just the
-space $\mathbf{1}$ but one can argue that the spaces are equivalent in
-some sense.
-
-\amr{new wavefront for new pass}
-
-These have a nice correspondance (Curry-Howard) to logic:
+\mathbf{1})$. This space is the product of two subspaces: the subspace
+$(\mathbf{1} \uplus \mathbf{0})$ which itself is the sum of the space
+$\mathbf{1}$ containing one element and the empty space $\mathbf{0}$
+and the subspace $(\mathbf{1} \uplus \mathbf{1})$ which is the sum of
+two spaces each containing one element. First, as discussed in the
+previous section, any deformation of this space must at least preserve
+the number of elements: we can neither create nor destroy points
+during any continuous deformation. Seeing that the number of elements
+in our example space is 2, a reasonable hypothesis is that we can
+deform the space above to any other space with 2 elements. What this
+really means is that we are treating the sum and product structure as
+malleable. For example, imagining a product structure as arranged in a
+grid, by stretching we can turn this structure to a sum structure
+arranged in a line.
+
+We now turn into the mathematical formalization of this idea. First we
+note that the type structure of types has a nice correspondance
+(Curry-Howard) to logic:
 

### 2017-12-30 survey/survey.tex
a bit more on sec 3

@@
 language from this denotation. But first, an obvious remark: any two sets
-$C$ and $D$ of cardinality $n$ are always in bijective correspondance. So
+$C$ and $D$ of cardinality $n$ are always in bijective correspondence. So
 we can abstract away from the details of the elements of $C$ and $D$ and
@@
 
-We now turn into the mathematical formalization of this idea. First we
-note that the type structure of types has a nice correspondance
-(Curry-Howard) to logic:
+We now turn into the mathematical formalization of this idea. Our goal
+is a sound and complete equivalence relation on types that maintains
+the number of points. First we note that the type structure of types
+has a nice correspondence (Curry-Howard) to logic:
 
@@
 Logic & Types \\ \hline
-$\true$ & $\top$ \\
-$\false$ & $\bot$ \\
+$\false$ & $\mathbf{0}$ \\
+$\true$ & $\mathbf{1}$ \\
 $\land$ & $\times$ \\
@@
 
-This correspondance is rather fruitful. As logical expressions form
-a commutative semiring, we would expect that types too form a commutative
-semiring. And indeed they do -- at least up to \emph{type equivalence}.  The
-natural numbers $\Nat$ are another commutative semiring; it will turn out that,
-even though the Curry-Howard correspondance has been extremely fruitful for
-programming language research, it is $\Nat$ which will be a better model for
-finite structured types.
+This correspondence is rather fruitful. As logical expressions form a
+commutative semiring, we would expect that types too form a
+commutative semiring. And indeed they do -- at least up to \emph{type
+  isomorphism}.  The natural numbers $\Nat$ are another commutative
+semiring; it will turn out that, even though the Curry-Howard
+correspondence has been extremely fruitful for programming language
+research, it is $\Nat$ which will be a better model for finite
+structured types as the corresponding commutative semiring captures
+the familiar numerical identities that evidently preserve the number
+of points in the types.
 
@@
 
-We cannot immediately use the above definition in a typed setting.
-First, types do not naturally want to be put together into a ``set''.
-This can be fixed if we replace the set $R$ with a universe $U$, and
-rather than insisting that we have $0 \in R$, we have a typing
-judgement $\bot : U$ (and similarly for the other items).
-
-Our first instinct would be to similarly replace $=$ with
-$\equiv$, i.e. propositional equality. This bring the second
-problem to the fore: this is false. For example, unless $A \equiv B$,
-$A \times B \equiv B \times A$ is not normally provable%
-\footnote{Not without assuming that $U$ is univalent, which we
+We would like to adapt the commutative semiring definition to the
+setting of structured types. First, types do not naturally want to be
+put together into a ``set''.  This can be fixed if we replace the set
+$R$ with a universe $U$, and replace the set membership $0 \in R$ with
+the typing judgement $\mathbf{0} : U$ (and similarly for the other
+items). Our first instinct would be to similarly replace $=$ with a
+judgement $A \equiv B$ that asserts that $A$ and $B$ are
+\emph{propositionally equal}. This is however not true: unless
+$A \equiv B$, $A \times B \equiv B \times A$ is not normally
+provable\footnote{Not without assuming that $U$ is univalent, which we
   don't.}. But it should be clear that $A \times B$ and $B \times A$
-contain equivalent information. In other words, we would like to
-be able to witness that $A \times B$ can be reversibly deformed
-into $B \times A$. Which motivates the introduction of
-type \emph{equivalences}. To do this, we need a few important
-auxiliary concepts first.
+contain equivalent information. In other words, we would like to be
+able to witness that $A \times B$ can be reversibly deformed into
+$B \times A$. Which motivates the introduction of type
+\emph{equivalences}. To do this, we need a few important auxiliary
+concepts first.
 
@@
 If we replace $=$ by $\simeq$, then we can indeed prove that
-types (with $\bot, \top, \uplus, \times$) form a commutative semiring.
+types (with $\mathbf{0}, \mathbf{1}, \uplus, \times$) form a commutative semiring.
 
@@
 
-If we revisit the Curry-Howard correspondance, we notice one
+If we revisit the Curry-Howard correspondence, we notice one
 more issue. In logic, it is true that $A \lor A = A$ and
@@
   \qquad A \ \text{is inhabited} \]
-The above is the real \textit{essence} of the correspondance.
+The above is the real \textit{essence} of the correspondence.
 In other words, classical Curry-Howard tells us about
@@
 but its output, only one. On the type side, something more subtle
-happens. Consider $\top \uplus \top$ versus $\top$; the first has
+happens. Consider $\mathbf{1} \uplus \mathbf{1}$ versus $\mathbf{1}$; the first has
 \emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
 while the second only one ($\texttt{tt}$). These cannot be put in
-bijective correspondance. Even though the ``payload'' \texttt{tt}
+bijective correspondence. Even though the ``payload'' \texttt{tt}
 is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
@@
 \\
-\identlp :&  \bot \uplus \tau & \simeq & \tau &: \identrp \\
+\identlp :&  \mathbf{0} \uplus \tau & \simeq & \tau &: \identrp \\
 \swapp :&  \tau_1 \uplus \tau_2 & \simeq & \tau_2 \uplus \tau_1 &: \swapp \\
@@
 \\
-\identlt :&  \top \times \tau & \simeq & \tau &: \identrt \\
+\identlt :&  \mathbf{1} \times \tau & \simeq & \tau &: \identrt \\
 \swapt :&  \tau_1 \times \tau_2 & \simeq & \tau_2 \times \tau_1 &: \swapt \\
@@
 \\
-\distz :&~ \bot \times \tau & \simeq & \bot ~ &: \factorzl \\
+\distz :&~ \mathbf{0} \times \tau & \simeq & \mathbf{0} ~ &: \factorzl \\
 \dist :&~ (\tau_1 \uplus \tau_2) \times \tau_3 & \simeq & (\tau_1 \times \tau_3) \uplus (\tau_2 \times \tau_3)~ &: \factor
@@
 
-Nevertheless, the Curry-Howard correspondance still has some
+Nevertheless, the Curry-Howard correspondence still has some
 force. We know that the inhabitants of types formed with
-with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
+with $\mathbf{0}, \mathbf{1}, \uplus, \times$ are a commutative semiring. What we
 want to know is, which types are equivalent? From a commutative semiring
@@
 to create a language $\Pi$ such that the types and type combinators
-map to $\bot, \top, \uplus, \times$, and such that we have ground
+map to $\mathbf{0}, \mathbf{1}, \uplus, \times$, and such that we have ground
 terms whose denotation are all $16$ type isomorphisms of
@@
 ``conservation of information'' as the basis, a different kind of
-correspondance arises; in fact, this one may well be an actual
+correspondence arises; in fact, this one may well be an actual
 isomorphism.}

### 2017-12-31 survey/survey.tex
revised sec 3 a bit more up to wavefront

@@
 \section{Data I: Finite Sets}
+\label{sec:dataone}
 
@@
 
-Instead of the spaces consisting solely of unstructured isolated
-points, we now investigate structured spaces built from sums and
-products of elementary spaces. This structure corresponds to the
-building blocks of type theory which are: the empty type
-($\mathbf{0}$), the unit type ($\mathbf{1}$), the sum type ($\uplus$),
-and the product ($\times$) type. Before getting into the formal
-theory, let's consider possible deformations on the space
-$(\mathbf{1} \uplus \mathbf{0}) \times (\mathbf{1} \uplus
-\mathbf{1})$. This space is the product of two subspaces: the subspace
-$(\mathbf{1} \uplus \mathbf{0})$ which itself is the sum of the space
-$\mathbf{1}$ containing one element and the empty space $\mathbf{0}$
-and the subspace $(\mathbf{1} \uplus \mathbf{1})$ which is the sum of
-two spaces each containing one element. First, as discussed in the
-previous section, any deformation of this space must at least preserve
-the number of elements: we can neither create nor destroy points
-during any continuous deformation. Seeing that the number of elements
-in our example space is 2, a reasonable hypothesis is that we can
-deform the space above to any other space with 2 elements. What this
+Instead of spaces consisting solely of unstructured isolated points,
+we now investigate structured spaces built from sums and products of
+elementary spaces. This structure corresponds to the building blocks
+of type theory which are: the empty type ($\bot$), the unit type
+($\top$), the sum type ($\uplus$), and the product ($\times$)
+type. Before getting into the formal theory, let's consider possible
+deformations on the space
+$(\top \uplus \bot) \times (\top \uplus \top)$. This space is the
+product of two subspaces: the subspace $(\top \uplus \bot)$ which
+itself is the sum of the space $\top$ containing one element
+$\texttt{tt}$ and the empty space $\bot$ and the subspace
+$(\top \uplus \top)$ which is the sum of two spaces each containing
+the one element $\texttt{tt}$. First, as discussed in the previous
+section, any deformation of this space must at least preserve the
+number of elements: we can neither create nor destroy points during
+any continuous deformation. Seeing that the number of elements in our
+example space is 2, a reasonable hypothesis is that we can deform the
+space above to any other space with 2 elements such as
+$\top \uplus \top$ or $\top \uplus (\top \uplus \bot)$. What this
 really means is that we are treating the sum and product structure as
@@
 grid, by stretching we can turn this structure to a sum structure
-arranged in a line.
+arranged in a line, change the orientation of the grid by exchanging
+the axes, as well as do other transformations to be discussed in this
+section that preserve the number of points.
+
+%%%%%%%%%
+\subsection{A Model of Type Equivalences}
 
 We now turn into the mathematical formalization of this idea. Our goal
-is a sound and complete equivalence relation on types that maintains
-the number of points. First we note that the type structure of types
-has a nice correspondence (Curry-Howard) to logic:
+is a denotational semantics on types that maintains the number of
+points. First we note that the type structure of types has a nice
+correspondence (Curry-Howard) to logic:
 
@@
 Logic & Types \\ \hline
-$\false$ & $\mathbf{0}$ \\
-$\true$ & $\mathbf{1}$ \\
+$\false$ & $\bot$ \\
+$\true$ & $\top$ \\
 $\land$ & $\times$ \\
@@
 
-This correspondence is rather fruitful. As logical expressions form a
-commutative semiring, we would expect that types too form a
-commutative semiring. And indeed they do -- at least up to \emph{type
-  isomorphism}.  The natural numbers $\Nat$ are another commutative
-semiring; it will turn out that, even though the Curry-Howard
-correspondence has been extremely fruitful for programming language
-research, it is $\Nat$ which will be a better model for finite
-structured types as the corresponding commutative semiring captures
-the familiar numerical identities that evidently preserve the number
-of points in the types.
+\noindent This correspondence is rather fruitful. As logical
+expressions form a commutative semiring, we would expect that types
+too form a commutative semiring. And indeed they do -- at least up to
+\emph{type isomorphism}.  The natural numbers $\Nat$ are another
+commutative semiring; it will turn out that, even though the
+Curry-Howard correspondence has been extremely fruitful for
+programming language research, it is $\Nat$ which will be a better
+model for finite structured types as the corresponding commutative
+semiring captures the familiar numerical identities that preserve the
+number of points in the types.
 
 \begin{defn}
+\label{def:rig}
 A \emph{commutative semiring} (sometimes called a \emph{commutative
-    rig} (commutative ri\emph{n}g without negative elements) consists of a
-  set $R$, two distinguished elements of $R$ named 0 and 1, and two
-  binary operations~$+$ and $\cdot$, satisfying the following
-  relations for any $a,b,c \in R$:
-\begin{equation}
-\begin{array}{rcl}
+  rig} (commutative ri\emph{n}g without negative elements)
+$(R,0,1,+,\cdot)$ consists of a set $R$, two distinguished elements of
+$R$ named 0 and 1, and two binary operations~$+$ and $\cdot$,
+satisfying the following relations for any $a,b,c \in R$:
+\[\begin{array}{rcl}
 0 + a &=& a \\
@@
 (a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
-\end{array}
-\label{eq:csemiring}
-\end{equation}
+\end{array}\]
 \end{defn}
@@
 $R$ with a universe $U$, and replace the set membership $0 \in R$ with
-the typing judgement $\mathbf{0} : U$ (and similarly for the other
-items). Our first instinct would be to similarly replace $=$ with a
+the typing judgement $\bot : U$ (and similarly for the other
+items). Our next instinct would be to similarly replace $=$ with a
 judgement $A \equiv B$ that asserts that $A$ and $B$ are
-\emph{propositionally equal}. This is however not true: unless
-$A \equiv B$, $A \times B \equiv B \times A$ is not normally
-provable\footnote{Not without assuming that $U$ is univalent, which we
-  don't.}. But it should be clear that $A \times B$ and $B \times A$
-contain equivalent information. In other words, we would like to be
-able to witness that $A \times B$ can be reversibly deformed into
+\emph{propositionally equal}. This is however not true: the
+proposition $A \times B \equiv B \times A$ is not
+normally\footnote{Except In homotopy type theory where equivalent
+  types are identified.} provable for arbitrary types $A$ and $B$. But
+it should be clear that $A \times B$ and $B \times A$ contain
+equivalent information. In other words, we would like to be able to
+witness that $A \times B$ can be reversibly deformed into
 $B \times A$. Which motivates the introduction of type
 \emph{equivalences}. To do this, we need a few important auxiliary
-concepts first.
-
-\begin{defn}[homotopy]
+concepts.
+
+\begin{defn}[Homotopy]
 \label{def:homotopy}
 Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
-$\forall x:A. f(x) = g(x)$. We denote this $f \sim g$.
+$\forall x:A. f(x) \equiv g(x)$. We denote this $f \sim g$.
 \end{defn}
@@
 space $A \rightarrow B$) are an equivalence relation.  The simplest
-definition of the data which makes up an equivalence is then
+definition of the data which makes up an equivalence is the following.
 
@@
 
-\noindent Why \emph{quasi}? The reasons are beyond our scope, but
-the interested reader can read sections $2.4$ and $4$ in \cite{hottbook}.
-There are several conceptually different, but equivalent,
-``better'' definitions.  We record just one here:
+\noindent Why \emph{quasi}? The reasons are beyond our scope, but the
+interested reader can read Sec.~$2.4$ and Ch.~$4$ in the HoTT
+book~\cite{hottbook}.  There are several conceptually different, but
+equivalent, ``better'' definitions.  We record just one here:
 
@@
 
-If we replace $=$ by $\simeq$, then we can indeed prove that
-types (with $\mathbf{0}, \mathbf{1}, \uplus, \times$) form a commutative semiring.
-
-The reader familiar with universal algebra should pause
-and ponder a bit about what we have done. We have lifted
-\emph{equality} from being in the signature of the ambient logic
-and instead put it in the signature of the algebraic structure
-of interest. Then, and only then, are we in a solid position
-to show that we have a concept (commutative semiring) which is flexible
-enough for types to be an instance.
+We are now in position to formally state the appropriate commutative
+semiring structure for types. After replacing the set $R$ with a
+universe $U$, we also replace the algebraic use of $=$ in
+Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
+this change, we can indeed prove that types (with
+$\bot, \top, \uplus, \times$) form a commutative semiring. The reader
+familiar with universal algebra should pause and ponder a bit about
+what we have done. We have lifted \emph{equality} from being in the
+signature of the ambient logic and instead put it in the signature of
+the algebraic structure of interest. Then, and only then, are we in a
+solid position to show that we have a concept (commutative semiring)
+which is flexible enough for types to be an instance.
 
@@
 a fancy way of saying
-\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
-  \qquad A \ \text{is inhabited} \]
-The above is the real \textit{essence} of the correspondence.
-In other words, classical Curry-Howard tells us about
-\emph{logical equivalence} of types. This is even a
+\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow \qquad A \
+  \text{is inhabited} \] The above is the real \textit{essence} of the
+Curry-Howard correspondence.  In other words, classical Curry-Howard
+tells us about \emph{logical equivalence} of types. This is even a
 constructive statement: there are indeed functions
@@
 
-So mere inhabitation falls far short of our goals of being
-able to smoothly deform from one type to another.
-
-Let us thus analyze the crux of the ``problem''. In logic, we have
-that $\land$ and $\lor$ are both \emph{idempotent}: this is the
-property of any binary operation $*$ where
-$\forall a. a * a = a$. And it should be clear that idempotent
-operations are \emph{forgetful}: its input has two copies of $a$,
-but its output, only one. On the type side, something more subtle
-happens. Consider $\mathbf{1} \uplus \mathbf{1}$ versus $\mathbf{1}$; the first has
-\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
-while the second only one ($\texttt{tt}$). These cannot be put in
-bijective correspondence. Even though the ``payload'' \texttt{tt}
-is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
-away information -- something we have expressly disallowed.
-Yes, this should remind you of Maxwell's demon: even though the
-data is the same, they are tagged differently, and this
-data is indeed information, and must be preserved.
+So mere inhabitation falls far short of our goals of being able to
+smoothly deform from one type to another. Let us thus analyze the crux
+of the ``problem''. In logic, we have that $\land$ and $\lor$ are both
+\emph{idempotent}: this is the property of any binary operation $*$
+where $\forall a. a * a = a$. And it should be clear that an
+idempotent operations is a \emph{forgetful} operation: its input has
+two copies of $a$, but its output, only one. On the type side,
+something more subtle happens. Consider $\top \uplus \top$
+versus $\top$; the first has \emph{two} proofs of inhabitation
+($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one
+($\texttt{tt}$). These cannot be put in bijective correspondence. Even
+though the ``payload'' \texttt{tt} is the same, forgetting
+$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
+we have expressly disallowed.  Yes, this should remind you of
+Maxwell's demon: even though the data is the same, they are tagged
+differently, and this data is indeed information, and must be
+preserved.
 
@@
 \begin{array}{rrcll}
-\idc :& \tau & \simeq & \tau &: \idc \\
-\\
-\identlp :&  \mathbf{0} \uplus \tau & \simeq & \tau &: \identrp \\
-\swapp :&  \tau_1 \uplus \tau_2 & \simeq & \tau_2 \uplus \tau_1 &: \swapp \\
-\assoclp :&  \tau_1 \uplus (\tau_2 \uplus \tau_3) & \simeq & (\tau_1 \uplus \tau_2) \uplus \tau_3 &: \assocrp \\
-\\
-\identlt :&  \mathbf{1} \times \tau & \simeq & \tau &: \identrt \\
-\swapt :&  \tau_1 \times \tau_2 & \simeq & \tau_2 \times \tau_1 &: \swapt \\
-\assoclt :&  \tau_1 \times (\tau_2 \times \tau_3) & \simeq & (\tau_1 \times \tau_2) \times \tau_3 &: \assocrt \\
-\\
-\distz :&~ \mathbf{0} \times \tau & \simeq & \mathbf{0} ~ &: \factorzl \\
-\dist :&~ (\tau_1 \uplus \tau_2) \times \tau_3 & \simeq & (\tau_1 \times \tau_3) \uplus (\tau_2 \times \tau_3)~ &: \factor
+\idc :& A & \simeq & A &: \idc \\
+\\
+\identlp :&  \bot \uplus A & \simeq & A &: \identrp \\
+\swapp :&  A \uplus B & \simeq & B \uplus A &: \swapp \\
+\assoclp :&  A \uplus (B \uplus C) & \simeq & (A \uplus B) \uplus C &: \assocrp \\
+\\
+\identlt :&  \top \times A & \simeq & A &: \identrt \\
+\swapt :&  A \times B & \simeq & B \times A &: \swapt \\
+\assoclt :&  A \times (B \times C) & \simeq & (A \times B) \times C &: \assocrt \\
+\\
+\distz :&~ \bot \times A & \simeq & \bot ~ &: \factorzl \\
+\dist :&~ (A \uplus B) \times C & \simeq & (A \times C) \uplus (B \times C)~ &: \factor
 \end{array}
@@
 
-Nevertheless, the Curry-Howard correspondence still has some
-force. We know that the inhabitants of types formed with
-with $\mathbf{0}, \mathbf{1}, \uplus, \times$ are a commutative semiring. What we
-want to know is, which types are equivalent? From a commutative semiring
-perspective, this amounts to asking what terms are equal?
-We have a set of generators for those equations, namely those
-in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
-of mutually inverse functions which witness these identities.
-For concreteness, we show the signatures in Figure~\ref{type-isos}.
-As these all come in symmetric pairs (some of which are self-symmetric),
-we give names for both directions. The names use $\iso$, $+$ and $*$
-as these will be what we use when we reflect these isomorphism into
-a programming language, which we proceed to do next.
+Nevertheless, the Curry-Howard correspondence still has some force. We
+know that the inhabitants of types formed with with
+$\bot, \top, \uplus, \times$ are a commutative semiring. What we want
+to know is, which types are equivalent? From a commutative semiring
+perspective, this amounts to asking what terms are equal?  We have a
+set of generators for those equations, namely those in
+Def.~\ref{def:rig}. What we thus need is to create $8$ pairs of
+mutually inverse functions which witness these identities.  For
+concreteness, we show the signatures in Fig.~\ref{type-isos}.  As
+these all come in symmetric pairs (some of which are self-symmetric),
+we give names for both directions.
+
+%%%%%%%%%
+\subsection{A Language of Type Equivalences}
 
@@
 
-We now have in our hands our desired denotational semantics: we want
-to create a language $\Pi$ such that the types and type combinators
-map to $\mathbf{0}, \mathbf{1}, \uplus, \times$, and such that we have ground
-terms whose denotation are all $16$ type isomorphisms of
-Figure~\ref{type-isos}. This is completely straightforward, as
-we can simply do this literally. To make the analogy with
-commutative semirings stand out even more, we will use $0, 1, +$, and $*$
-at the type level, and will denote ``equivalence'' by $\iso$.
-Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.
+We now have in our hands our desired denotational semantics. We want
+to create a programming language, which we call $\Pi$, such that the
+types and type combinators map to $\bot, \top, \uplus, \times$, and
+such that we have ground terms whose denotation are all $16$ type
+isomorphisms of Fig.~\ref{type-isos}. This is completely
+straightforward, as we can simply do this literally. To make the
+analogy with commutative semirings stand out even more, we will use
+$0, 1, +$, and $\times$ at the type level, and will denote
+``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
+``constants'' of the language.
 
@@
 constants! There is a need to perform multiple equivalences. There are
-in fact three ways to do this: sequential composition $\odot$, 
-choice composition $\oplus$ (sometimes called juxtaposition) and parallel
-composition $otimes$. See Figure~\ref{pi-combinators} for the signatures.
-
-$c_1 \odot c_2$ corresponds to performing $c_1$ first, then $c_2$, and
-is the usual notion of composition -- and corresponds to $\fatsemi$ of
-the language of the previous section. $c_1 \oplus c\_2$ chooses to
+in fact three ways to do this: sequential composition $\odot$, choice
+composition $\oplus$ (sometimes called juxtaposition), and parallel
+composition $\otimes$. See Fig.~\ref{pi-combinators} for the
+signatures. The construction $c_1 \odot c_2$ corresponds to performing
+$c_1$ first, then $c_2$, and is the usual notion of composition -- and
+corresponds to $\fatsemi$ of the language of permutations of
+Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c\_2$ chooses to
 perform $c_1$ or $c_2$ depending on whether the input is labelled
-$\textsf{inl}$ or $\textsf{inr}$ respectively. $c_1 \otimes c_2$
-operates on a product structure, and applies $c_1$ to the first component
-and $c_2$ to the second.
+$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
+construction $c_1 \otimes c_2$ operates on a product structure, and
+applies $c_1$ to the first component and $c_2$ to the second.
 
@@
 students and colleagues who participated in various aspects of this
-research and who provided valuable feedack and constructive criticism.
+research and who provided valuable feedback and constructive criticism.
 

### 2017-12-31 survey/survey.tex
typo

@@
 corresponds to $\fatsemi$ of the language of permutations of
-Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c\_2$ chooses to
+Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
 perform $c_1$ or $c_2$ depending on whether the input is labelled

### 2018-01-01 survey/survey.tex
continue with Data II

@@
 ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
-``constants'' of the language.
+``constants'' of the language. Note how we have continued with
+the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof terms},
+but rather than being witnesses of inhabitation, they are witnesses of
+equivences. Thus the proof language of commutative semiring equivalences
+becomes terms of $\Pi$.
 
@@
 
-\jc{wavefront}
-
-The type-erasure will, naturally, be permutations. But where do the
-isomorphisms come from?  Well, inhabitants of types with (0,1,+,*)
-form a semiring. So, still inspired by Curry-Howard, look to the
-proof-terms of semiring identities -- and voila, a nice set of
-isomorphisms! We can name these isomorphisms, AND give them
-implementations.
-
-That gives us some ground terms.  Not enough for a PL! So then
-we need to be able to 'combine' these. So we add 3 combinators for
-3 different kinds of composition.
-Gives us a nice, reversible PL for
-structured finite types. We call it Pi.
+The type-erase of this language \emph{should} be permutations. But
+given types $A$ and $B$ that erase to $\fin{m}$ and $\fin{n}$ respectively,
+what are $A + B$ and $A * B$ ?  They correspond exactly to
+$\fin{m+n}$ and $\fin{m*n}$!  Geometrically, this corresponds to
+concatenation for $A + B$, i.e. lining up the elements of $A$ first,
+and then those of $B$. For $A * B$, one can picture this as
+lining up the elements of $A$ horizontally, those of $B$ vertically
+and perpendicular to those of $A$, and filling in the square with
+pairs of elements from $A$ and $B$; if one re-numbers these sequentially,
+reading row-wise, this gives an enumeration of $\fin{m*n}$.
+
+From here, it is easy to see what, for example, $c_1 \oplus c_2$ must be,
+operationally: from a permutation on $\fin{m}$ and another on $\fin{n}$,
+create a permutation on $\fin{m+n}$ by having $c_1$ operate on the first
+$m$ elements of $A+B$, and $c_2$ operate on the last $n$ elements. 
+Similarly, $\swapp$ switches the roles of $A$ and $B$, and thus corresponds
+to $\fin{n+m}$. Note how we ``recover'' the commutativity of 
+natural number addition from this type isomorphism. Geometrically, $\swapt$
+is also rather interesting: it corresponds to matrix transpose!
+
+\jc{it would be nice to have a picture or two here}
 

### 2018-01-01 survey/survey.tex
The exposition for pi level 0 is done. There are still some remarks to be made about sound and complete as well as examples to add.

@@
 is also rather interesting: it corresponds to matrix transpose!
+Furthermore, in this representations, some combinators like
+$\identlp$ and $\assoclp$ are identity operations: the underlying representations
+are not merely isomorphic, they are definitionally equal.
+In other words, the passage to $\Nat$ erases some structural information.
 
@@
 
-We actually have several choices. We can make Pi structurally reversible
-(i.e. there is a syntactic transformation on Pi which maps combinators
-to combinators which witnesses reversibility), or build in a `reverse'
-combinator which flips things around. The latter lets us make the
-language much smaller, but then \emph{programs} in that language are longer.
-[Bias: since there will be many more programs written, the fact that the
-language definition, and its processors, is larger, is a one-time cost.
-Not that big a deal. So pick the version whose meta-theoretical
-properties are obvious. It is thus worthwhile to have reversibility be
-structural.]
+Embedded in our definition of $\Pi$ is a design decision: to make the
+terms of $\Pi$ \emph{syntactically} reversible. In other words, to
+every $\Pi$ constant, there is another $\Pi$ constant which is its
+inverse. This is not the only choice.  Another would be to add a
+$\mathit{flip}$ combinator to the language; we could then remove
+quite a few combinators as redundant. The drawback is that many
+programs in $\Pi$ become longer. Furthermore, some of the symmetry
+at ``higher levels'' (see next section) is also lost. Since the
+extra burden of language definition and of proofs is quite low, we
+prefer the structural symmetry over a minimalistic language definition.
+
+\jc{wavefront}
 

### 2018-01-01 survey/survey.tex
consistently use odot for first level composition. Change naming in first example in Data III to avoid matching A with 2 and B with 1, which is just confusing (even if correct).

@@
 \begin{array}{c}
-  {c_1 \circledcirc (c_2 \circledcirc c_3) \Leftrightarrow (c_1 \circledcirc c_2) \circledcirc c_3}
-\\
-  {(c_1 \oplus (c_2 \oplus c_3)) \circledcirc \assoclp \Leftrightarrow \assoclp \circledcirc ((c_1 \oplus c_2) \oplus c_3)}
-\\
-  {(c_1 \otimes (c_2 \otimes c_3)) \circledcirc \assoclt \Leftrightarrow \assoclt \circledcirc ((c_1 \otimes c_2) \otimes c_3)}
-\\
-  {((c_1 \oplus c_2) \oplus c_3) \circledcirc \assocrp \Leftrightarrow \assocrp \circledcirc (c_1 \oplus (c_2 \oplus c_3))}
-\\
-  {((c_1 \otimes c_2) \otimes c_3) \circledcirc \assocrt \Leftrightarrow \assocrt \circledcirc (c_1 \otimes (c_2 \otimes c_3))}
-\\
-  {\assocrp \circledcirc \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \circledcirc \assocrp) \circledcirc (\idc \oplus \assocrp)}
-\\
-  {\assocrt \circledcirc \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \circledcirc \assocrt) \circledcirc (\idc \otimes \assocrt)}
+  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
+\\
+  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
+\\
+  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
+\\
+  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
+\\
+  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
+\\
+  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
+\\
+  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
 \end{array}\]
@@
 \begin{array}{c}
-  {((a \oplus b) \otimes c) \circledcirc \dist \Leftrightarrow \dist \circledcirc ((a \otimes c) \oplus (b \otimes c))}
-\\
-  {(a \otimes (b \oplus c)) \circledcirc \distl \Leftrightarrow \distl \circledcirc ((a \otimes b) \oplus (a \otimes c))}
-\\
-  {((a \otimes c) \oplus (b \otimes c)) \circledcirc \factor \Leftrightarrow \factor \circledcirc ((a \oplus b) \otimes c)}
-\\
-  {((a \otimes b) \oplus (a \otimes c)) \circledcirc \factorl \Leftrightarrow \factorl \circledcirc (a \otimes (b \oplus c))}
+  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
+\\
+  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
+\\
+  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
+\\
+  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
 \end{array}\]
@@
 \begin{array}{c}
-  {\idc \circledcirc \, c \Leftrightarrow c}
+  {\idc \odot \, c \Leftrightarrow c}
 \quad
-  {c \, \circledcirc \idc \, \Leftrightarrow c}
+  {c \, \odot \idc \, \Leftrightarrow c}
 \quad
-  {c\,\, \circledcirc\,! c \Leftrightarrow \idc}
+  {c\,\, \odot\,! c \Leftrightarrow \idc}
 \quad
-  {! c \circledcirc c \Leftrightarrow \idc}
+  {! c \odot c \Leftrightarrow \idc}
 \\
@@
   {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
-  {c_1 \circledcirc c_2 \Leftrightarrow c' \circledcirc c''}
+  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
   {}
@@
 \begin{array}{c}
-  {\identlp \circledcirc c \Leftrightarrow (c₀ \oplus c) \circledcirc \identlp}
+  {\identlp \odot c \Leftrightarrow (c₀ \oplus c) \odot \identlp}
 \qquad
-  {\identrp \circledcirc (c₀ \oplus c) \Leftrightarrow c \circledcirc \identrp}
-\\
-  {\identlsp \circledcirc c \Leftrightarrow (c \oplus c₀) \circledcirc \identlsp}
+  {\identrp \odot (c₀ \oplus c) \Leftrightarrow c \odot \identrp}
+\\
+  {\identlsp \odot c \Leftrightarrow (c \oplus c₀) \odot \identlsp}
 \qquad
-  {\identrsp \circledcirc (c \oplus c₀) \Leftrightarrow c \circledcirc \identrsp}
-\\
-  {\identlt \circledcirc c \Leftrightarrow (c_1 \otimes c) \circledcirc \identlt}
+  {\identrsp \odot (c \oplus c₀) \Leftrightarrow c \odot \identrsp}
+\\
+  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
 \qquad
-  {\identrt \circledcirc (c_1 \otimes c) \Leftrightarrow c \circledcirc \identrp}
-\\
-  {\identlst \circledcirc c \Leftrightarrow (c \otimes c_1) \circledcirc \identlst}
+  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
+\\
+  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
 \qquad
-  {\identrst \circledcirc (c \otimes c_1) \Leftrightarrow c \circledcirc \identrst}
-\\
-  {\identlt \Leftrightarrow \distl \circledcirc (\identlt \oplus \identlt)}
-\\
-\identlp \Leftrightarrow \swapp \circledcirc \identrsp
+  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
+\\
+  {\identlt \Leftrightarrow \distl \odot (\identlt \oplus \identlt)}
+\\
+\identlp \Leftrightarrow \swapp \odot \identrsp
 \qquad
-\identlt \Leftrightarrow \swapt \circledcirc \identrst
+\identlt \Leftrightarrow \swapt \odot \identrst
 \end{array}\]
@@
 \begin{array}{c}
-  {\swapp \circledcirc (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \circledcirc \swapp}
+  {\swapp \odot (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \odot \swapp}
 \quad
-  {\swapt \circledcirc (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \circledcirc \swapt}
-\\
-  {(\assocrp \circledcirc \swapp) \circledcirc \assocrp \Leftrightarrow ((\swapp \oplus \idc) \circledcirc \assocrp) \circledcirc (\idc \oplus \swapp)}
-\\
-  {(\assoclp \circledcirc \swapp) \circledcirc \assoclp \Leftrightarrow ((\idc \oplus \swapp) \circledcirc \assoclp) \circledcirc (\swapp \oplus \idc)}
-\\
-  {(\assocrt \circledcirc \swapt) \circledcirc \assocrt \Leftrightarrow ((\swapt \otimes \idc) \circledcirc \assocrt) \circledcirc (\idc \otimes \swapt)}
-\\
-  {(\assoclt \circledcirc \swapt) \circledcirc \assoclt \Leftrightarrow ((\idc \otimes \swapt) \circledcirc \assoclt) \circledcirc (\swapt \otimes \idc)}
+  {\swapt \odot (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \odot \swapt}
+\\
+  {(\assocrp \odot \swapp) \odot \assocrp \Leftrightarrow ((\swapp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \swapp)}
+\\
+  {(\assoclp \odot \swapp) \odot \assoclp \Leftrightarrow ((\idc \oplus \swapp) \odot \assoclp) \odot (\swapp \oplus \idc)}
+\\
+  {(\assocrt \odot \swapt) \odot \assocrt \Leftrightarrow ((\swapt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \swapt)}
+\\
+  {(\assoclt \odot \swapt) \odot \assoclt \Leftrightarrow ((\idc \otimes \swapt) \odot \assoclt) \odot (\swapt \otimes \idc)}
 \end{array}\]
@@
 \\
-  {(a_1 \circledcirc a_3) \oplus (a_2 \circledcirc a_4) \Leftrightarrow (a_1 \oplus a_2) \circledcirc (a_3 \oplus a_4)}
-\\
-  {(a_1 \circledcirc a_3) \otimes (a_2 \circledcirc a_4) \Leftrightarrow (a_1 \otimes a_2) \circledcirc (a_3 \otimes a_4)}
+  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
+\\
+  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
 \end{array}\]
@@
 \begin{array}{c}
-  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \circledcirc (\idc \oplus \, \identlp)}
-\\
-  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \circledcirc (\idc \otimes \, \identlt)}
+  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \odot (\idc \oplus \, \identlp)}
+\\
+  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \odot (\idc \otimes \, \identlt)}
 \end{array}\]
@@
 \begin{array}{c}
-  {(c \otimes \idc) \circledcirc \absorbl \Leftrightarrow \absorbl \circledcirc \idc}
+  {(c \otimes \idc) \odot \absorbl \Leftrightarrow \absorbl \odot \idc}
 \quad
-  {(\idc \, \otimes c) \circledcirc \absorbr \Leftrightarrow \absorbr \circledcirc \idc}
-\\
-  {\idc \circledcirc \, \factorzl \Leftrightarrow \factorzl \circledcirc (\idc \otimes c)}
+  {(\idc \, \otimes c) \odot \absorbr \Leftrightarrow \absorbr \odot \idc}
+\\
+  {\idc \odot \, \factorzl \Leftrightarrow \factorzl \odot (\idc \otimes c)}
 \quad
-  {\idc \circledcirc \, \factorzr \Leftrightarrow \factorzr \circledcirc (c \otimes \idc)}
+  {\idc \odot \, \factorzr \Leftrightarrow \factorzr \odot (c \otimes \idc)}
 \\
@@
 \\
-  {\absorbr \Leftrightarrow (\distl \circledcirc (\absorbr \oplus \absorbr)) \circledcirc \identlp}
+  {\absorbr \Leftrightarrow (\distl \odot (\absorbr \oplus \absorbr)) \odot \identlp}
 \\
@@
 \qquad
-  {\absorbl \Leftrightarrow \swapt \circledcirc \absorbr}
-\\
-  {\absorbr \Leftrightarrow (\assoclt \circledcirc (\absorbr \otimes \idc)) \circledcirc \absorbr}
-\\
-  {(\idc \otimes \absorbr) \circledcirc \absorbl \Leftrightarrow (\assoclt \circledcirc (\absorbl \otimes \idc)) \circledcirc \absorbr}
-\\
-  {\idc \otimes \, \identlp \Leftrightarrow (\distl \circledcirc (\absorbl \oplus \idc)) \circledcirc \identlp}
+  {\absorbl \Leftrightarrow \swapt \odot \absorbr}
+\\
+  {\absorbr \Leftrightarrow (\assoclt \odot (\absorbr \otimes \idc)) \odot \absorbr}
+\\
+  {(\idc \otimes \absorbr) \odot \absorbl \Leftrightarrow (\assoclt \odot (\absorbl \otimes \idc)) \odot \absorbr}
+\\
+  {\idc \otimes \, \identlp \Leftrightarrow (\distl \odot (\absorbl \oplus \idc)) \odot \identlp}
 \end{array}\]
@@
 \begin{array}{c}
-  {((\assoclp \otimes \idc) \circledcirc \dist) \circledcirc (\dist \oplus \idc) \Leftrightarrow (\dist \circledcirc (\idc \oplus \dist)) \circledcirc \assoclp}
-\\
-  {\assoclt \circledcirc \distl \Leftrightarrow ((\idc \otimes \distl) \circledcirc \distl) \circledcirc (\assoclt \oplus \assoclt)}
+  {((\assoclp \otimes \idc) \odot \dist) \odot (\dist \oplus \idc) \Leftrightarrow (\dist \odot (\idc \oplus \dist)) \odot \assoclp}
+\\
+  {\assoclt \odot \distl \Leftrightarrow ((\idc \otimes \distl) \odot \distl) \odot (\assoclt \oplus \assoclt)}
 \end{array}\]
@@
 \begin{array}{rcl}
-  (\distl \circledcirc (\dist \oplus \dist)) \circledcirc \assoclp &\Leftrightarrow&
-   \dist \circledcirc (\distl \oplus \distl) \circledcirc \assoclp ~\circledcirc \\
-&& (\assocrp \oplus \idc) ~\circledcirc \\
-&& ((\idc \oplus \swapp) \oplus \idc) ~\circledcirc \\
+  (\distl \odot (\dist \oplus \dist)) \odot \assoclp &\Leftrightarrow&
+   \dist \odot (\distl \oplus \distl) \odot \assoclp ~\odot \\
+&& (\assocrp \oplus \idc) ~\odot \\
+&& ((\idc \oplus \swapp) \oplus \idc) ~\odot \\
 &&      (\assoclp \oplus \idc)
@@
 \begin{array}{rcl}
-  (\idc \otimes \swapp) \circledcirc \distl &\Leftrightarrow& \distl \circledcirc \swapp
-\\
-  \dist \circledcirc (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \circledcirc \distl
+  (\idc \otimes \swapp) \odot \distl &\Leftrightarrow& \distl \odot \swapp
+\\
+  \dist \odot (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \odot \distl
 \end{array}\]
@@
 previous section are reversible deformations between structured finite
-types. We now ask whether these reversible deformations can themselves
+types. We now ask whether these programs can themselves
 be subject to (higher-level) reversible deformations?
@@
 %%  \node at (3.3,-2) {$\mathit{swapl}_{+\Leftrightarrow}$} ;
-%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+})$};
-%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\circledcirc~(c_1~\oplus~c_2))$};
+%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\odot~\mathit{swap}_{+})$};
+%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\odot~(c_1~\oplus~c_2))$};
   \draw (-2,-2) ellipse (0.5cm and 1cm);
@@
 
-  \node at (3,-3.5) {$c_1$};
-  \node at (3,-4.5) {$c_2$};
-
-  \node at (1.5,0.5) {$c_2$};
-  \node at (1.5,-0.5) {$c_1$};
+  \node at (3,-3.5) {$c_2$};
+  \node at (3,-4.5) {$c_1$};
+
+  \node at (1.5,0.5) {$c_1$};
+  \node at (1.5,-0.5) {$c_2$};
 
@@
 \end{center}
-The top path is the reversible deformation
-$(c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+}$ which deforms the
-type $A$ by $c_2$, deforms the type $B$ by $c_1$, and deforms the
+The top path is the $\Pi$ program
+$(c_1~\oplus~c_2)~\odot~\swapp$ which deforms the
+type $A$ by $c_1$, deforms the type $B$ by $c_2$, and deforms the
 resulting space by a twist that exchanges the two injections into the
 sum type. The bottom path performs the twist first and then deforms
-the type $A$ by $c_2$ and the type $B$ by $c_1$ as before. If one
+the type $A$ by $c_1$ and the type $B$ by $c_2$ as before. If one
 could imagine the paths are physical wires and the deformations $c_1$
@@
 words, there exists a higher-level deformation of the program
-$(c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+}$ to the program
-$\mathit{swap}_{+} \circledcirc (c_1~\oplus~c_2)$.
+$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
+$\swapp \odot (c_2~\oplus~c_1)$.
 

### 2018-01-01 survey/survey.tex
Data III progress.

@@
 points $A$, $B$, $C$, and $D$ fixed, it is possible to rotate the top
-part of the diagram to become identical to the bottom one. In other
-words, there exists a higher-level deformation of the program
+part of the diagram to become identical to the bottom one. That
+rotation can be undone (reversed), which takes the bottom
+part of the diagram into the top part.  In other
+words, there exists a deformation of the program
 $(c_1~\oplus~c_2)~\odot~\swapp$ to the program
-$\swapp \odot (c_2~\oplus~c_1)$.
+$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this
+means that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
+$\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not
+all programs between the same types can be deformed into one
+another. The simplest example of inequivalent deformations
+are the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
+
+In the previous section, we saw how we could take the proof
+terms of commutative semirings as our starting point. What we need
+now is to understand how proofs of algeraic identities should be
+considered equivalent. Classical algebra does not help, as proofs
+in algebra are not considered first-class citizens. However,
+another route is available to us: since the work of 
+Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
+one can model types as \emph{groupoids}.  Previously, we
+modeled types as (finite) sets.
+
+Thus, rather than looking at (untyped) commutative semirings,
+we should looked at a \emph{typed} version. This process frequently
+goes by the moniker of ``categorification''.  We want a categorical
+algebra, where the basic objects are groupoids (to model our types),
+and where there is a natural notion of $+$ and $*$.  At first,
+we hit what seems like a serious stumbling block: the category of
+all groupoids, Groupoid, does not have co-products or products.
+We don't want to work internally in Groupoid -- we want operations
+\emph{on} groupoids. In other words, we want something akin to
+symmetric monoidal categories, but with two interacting 
+monoidal structures.  Luckily, this already exists: the categorical
+analog to (commutative) semirings are (symmetric) Rig Categories.
+This straightforwardly generalizes to symmetric Rig Groupoids.
+
+How does this help? Coherence conditions! Symmetric monoidal categories,
+to start somewhere simple, do not just introduce weak transformations
+like the associator $\alpha$ and the left $\lambda$ and right $\rho$
+unitors, but also coherence conditions that these must satisfy. Looking,
+for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
+$1$ and $+$ for the types, $\odot$ and $\oplus$ as combinators, and
+only the terms so expressible), the sub-language would correspond,
+denotationally, to exactly symmetric monoidal groupoids. And what
+these possess are exactly some \emph{equations between equations}
+as commutative diagrams.  Adding in the coherence conditions 
+corresponding to naturality of the various transformers gives us a
+list of equations between programs.  Furthermore, the natural transformations
+that arise are in fact natural isomorphisms -- all of which are
+reversible.
+
+We can then proceed to prove that every one of the coherence conditions
+involved in defining a symmetric Rig Groupoid holds for the groupoid
+interpretation of types.
 
@@
 Pi level 2.
-
-Take categorification seriously. Types are indeed groupoids, and
-endo-isomorphisms inhabit the Identity 'type' of a type. So + and *
-must be operations on groupoids. Look at the category of (small)
-categories for inspiration. So we have two monoidal structures.
-These are not co-product and product in Groupoid. But this is actually
-a good thing.
 

### 2018-01-03 survey/survey.tex
add a comment so that it doesn't get lost
add an abstract.

@@
 \begin{document}
+\begin{abstract}
+  Our main models of computation (Turing Machines, RAM, but also
+  modern computer architecture) make some fundamental assumptions about
+  which operations are available: logical operations like conjunction,
+  disjunction and negation, as well as reading and writing to a
+  (large) collection to memory locations. While this conforms to
+  a macro-level view of physics, it does not mesh as well with
+  other views; in particular, both from a thermodynamic and a
+  quantum view, these models are flawed. Both latter views of
+  physics imply that information is a quantity that is conserved.
+
+  We first define what information, and its conservation, means
+  in a computational setting. This emphasizes the idea that
+  computations, whatever they are, must be reversible. Thinking
+  geometrically, we can think of programs as reversible deformations
+  of data. From this fundamental idea, we rebuild notions of data
+  and programs which, by construction, are reversible. After a
+  brief overview of what this means for unstructured finite data,
+  we dive into the main topic: a foundational reversible programming
+  language. We re-examine the Curry-Howard correspondence, as the
+  classical correspondence is not reversible. This allows us to
+  establish a new correspondence, this time between a reversible
+  programming language for witnessing type isomorphisms and
+  proof terms for commutative semirings, which is sound and
+  complete for finite types.
+
+  We then ``move up a level'' to examine what it means for a
+  program to be reversible when its data is a program. This
+  gives a notion of reversible program equivalence. To get
+  something sound and complete, we must look at a ``higher 
+  dimensional'' analog to commutative semirings: symmetric
+  Rig Groupoids. The coherence laws for these are exactly the
+  reversible program equivalences we sought.
+
+  Something about Theseus?
+
+\end{abstract}
+
 \maketitle
@@
 
+\jc{We should decide what you level convention is. Do we say
+that $\iso$ is level 0, so that $\Leftrightarrow$ is level 1?
+We need to be consistent about it.}
+
 \jc{wavefront}

### 2018-01-09 survey/survey.tex
worked on abstract

--- 2018-01-01 survey/survey.tex
@@
 
+\begin{abstract}
+  Our main models of computation (the Turing Machine and the RAM) and
+  most modern computer architectures make fundamental assumptions
+  about which primitive operations are realizable on a physical
+  computing device. The consensus is that these primitive operations
+  include logical operations like conjunction, disjunction and
+  negation, as well as reading and writing to a (large) collection to
+  memory locations. This perspective conforms to a macro-level view of
+  physics and indeed these operations are realizable using macro-level
+  devices involving thousands of electrons. This point of view is
+  however incompatible with computation realized using quantum devices
+  or analyzed using elementary thermodynamics as both these
+  fundamental physical theories imply that information is a conserved
+  quantity of physical processes and hence of primitive computational
+  operations.
+
+  Our aim is re-develop foundational computational models in a way
+  that embraces the principle of conservation of information. We first
+  define what information is and what its conservation means in a
+  computational setting. We emphasize the idea that computations,
+  whatever they are, must be reversible transformations on
+  data. Thinking topologically, one can think of data as modeled using
+  topological spaces and programs as modeled by reversible
+  deformations of these spaces. We then illustrate this idea using
+  three notions of data and their associated reversible computational
+  models. The first instance only assumes unstructured finite data,
+  i.e., discrete topological spaces. The corresponding notion of
+  reversible computation is that of permutations. We show how this
+  simple model subsumes conventional computations on finite sets. We
+  then consider a modern structured notion of data based on the
+  Curry-Howard correspondence between logic and type theory. We
+  develop the corresponding notion of reversible deformations using a
+  sound and complete programming language for witnessing type
+  isomorphisms and proof terms for commutative semirings.  We then
+  ``move up a level'' to examine spaces that treat programs as data
+  which is a crucial notion for any universal model of computation. To
+  derive the corresponding notion of reversible programs between
+  programs, i.e., reversible program equivalences, we look at the
+  ``higher dimensional'' analog to commutative semirings: symmetric
+  rig groupoids. The coherence laws for these groupoids turn out to be
+  exactly the sound and complete reversible program equivalences we
+  seek.
+
+  We conclude with some possible generalizations inspired by homotopy
+  type theory and survey several open directions for further research.
+
+%%  Something about Theseus?
+\end{abstract}
+
 % * Reversibility intro / motivation as we  have now more or less

### 2018-05-14 survey/survey.tex
went over sec 1 and made small changes

@@
 well as reading from and writing to a large (infinite) collection of
-memory locations.
-
-No doubt, this consensus has been successful. And no doubt, these
-operations \emph{can} indeed be performed on a computer. Yet, today,
-with a possible quantum computing revolution in sight and with
-unprecedented explosion in embedded computers and cyber-physical
-systems, there are reasons to re-think this foundational question
-again. In fact, the calls to re-think this foundational question have
-been proclaimed by physicists more than forty years ago as the
-following two quotes testify:
+memory locations. From this small set of primitive operations emerges
+all higher-level programming languages and abstractions.
+
+No doubt, this consensus on the available primitive physical
+operations has been successful. And no doubt, these operations
+\emph{can} indeed be performed on a computer. Yet, today, with a
+possible quantum computing revolution in sight and with unprecedented
+explosion in embedded computers and cyber-physical systems, there are
+reasons to re-think this foundational question again. In fact, the
+calls to re-think this foundational question have been proclaimed by
+physicists more than forty years ago as the following two quotes
+testify:
 
 \begin{quote}
-  \textbf{Toffoli 1980~\cite{toffoli:1980}:} Mathematical models of computation are
-  abstract constructions, by their nature unfettered by physical
-  laws. However, if these models are to give indications that are
-  relevant to concrete computing, they must somehow capture, albeit in
-  a selective and stylized way, certain general physical restrictions
-  to which all concrete computing processes are
+  \textbf{Toffoli 1980~\cite{toffoli:1980}:} Mathematical models of
+  computation are abstract constructions, by their nature unfettered
+  by physical laws. However, if these models are to give indications
+  that are relevant to concrete computing, they must somehow capture,
+  albeit in a selective and stylized way, certain general physical
+  restrictions to which all concrete computing processes are
   subjected.
@@
 Maxwell's demon appears to violate the second law of thermodynamics by
-having a tiny `intelligence' observing the movement of individual
+having a tiny ``intelligence'' observing the movement of individual
 particles of a gas and separating fast moving particles from slow
@@
 that the computation done by the finite brain of the demon, involves
-getting information about the movement of molecules, acting on that
-information, and then overwriting it to make room for the next
+getting information about the movement of molecules, storing that
+information, analyzing that information to act on it, and then --- and
+this is the critical step --- overwriting it to make room for the next
 computation.  In other words, the computation that is manipulating
@@
 information. These models followed in the footsteps of logic where,
-following hundreds of years of tradition, conventional operations such
-conjunction and disjunction lose information by logically
-irreversible. These models were already radical enough, by being
-purely \emph{constructive}, in an era where classical mathematics,
-with its pervasive use of excluded middle and frequent use of the
-axiom of choice, did not overly worry about effective
-representation. Landauer's observation implied however that ideas in
-each field have consequences for the
+following hundreds of years of tradition, the truth of a statement was
+seen as ``absolute'' and independent of any reasoning, understanding,
+or action. Statements were either true or false with no regard to any
+``observer'' and the idea that statements had information content that
+should be preserved was outside the classical understanding of
+logic. Hence the fact that conventional logic operations such as
+conjunction and disjunction were logically irreversible and hence lose
+information was not a concern. Landauer's observation implied however
+that ideas in each field have consequences for the
 other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,baez2011physics,dblp:conf/csfw/malacarias12}. To
@@
 \paragraph*{Origins of Computational Models.}
-Our current logical operations in high-level programming languages and
-in our current hardware are both based on Boolean Logic -- which
-really ought to be called \emph{Piercean logic}, as Boole had a rather
-different conception of logic which is rather far from our current
-models.  Nevertheless, going back to Boole's 1853 book entitled
-\emph{An Investigation of the Laws of Thought, on which are Founded
-  the Mathematical Theories of Logic and Probabilities}. The opening
-sentence of Ch.~1 is:
+Current high-level programming languages as well as current hardware
+are both based on Boolean Logic -- which really ought to be called
+\emph{Piercean logic}, as Boole had a rather different conception of
+logic which is rather far from our current models.  Nevertheless,
+going back to Boole's 1853 book entitled \emph{An Investigation of the
+  Laws of Thought, on which are Founded the Mathematical Theories of
+  Logic and Probabilities}, we find that the opening sentence of Ch.~1 is:
 \begin{quote}
@@
 scope, the above quotes should convey the idea that our elementary
-computing notions date back to ideas that were thought reasonable
-before \textasciitilde 1900.
+computing notions date back to ideas that were thought reasonable in
+the late 1800s.
 
@@
 are at odds with our current models of computation. In this paper, we
-will focus on the following particular aspect. We will view data as an
-explicit representation of \emph{information} and programs as
-processes that transform information in a reversible way, i.e.,
-processes that are subject to the physical principle of
-\emph{conservation of information.} We will formalize this idea and
-follow its consequences, which will turn out to be far reaching.
+will focus on the previously identified principle of
+\emph{reversibility}. In more detail, we will view data as an explicit
+representation of \emph{information} and programs as processes that
+transform information in a reversible way, i.e., processes that are
+subject to the physical principle of \emph{conservation of
+  information.} We will formalize this idea and follow its
+consequences, which will turn out to be far reaching.
 
@@
 looking at what structure our data has, and that will help us define
-an appropriate notion of program. Of course, when programs themselves
-are data, things do get more complicated.  In the following sections,
-we will look at different natural classes of data, and explore the
-corresponding notion of reversible programs.
+an appropriate notion of a reversible program. Of course, when
+programs themselves are data, things do get more complicated.  In the
+following sections, we will look at different natural classes of data,
+and explore the corresponding notion of reversible programs.
 
@@
 but rather than being witnesses of inhabitation, they are witnesses of
-equivences. Thus the proof language of commutative semiring equivalences
+equivalences. Thus the proof language of commutative semiring equivalences
 becomes terms of $\Pi$.
@@
 terms of commutative semirings as our starting point. What we need
-now is to understand how proofs of algeraic identities should be
+now is to understand how proofs of algebraic identities should be
 considered equivalent. Classical algebra does not help, as proofs

### 2018-05-15 survey/survey.tex
sec 1 and 2

@@
 What does it mean to deform a space of points? For example, what
-transformation can we do on a bag or marbles? Well, we can shuffle
+transformation can we do on a bag of marbles? Well, we can shuffle
 them around and that is the only transformation that will preserve the
@@
 transpositions. Thus we can create a small language for writing
-permutations as
+permutations on $\fin{n}$ as
 \[\begin{array}{rcl}
-p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p \fatsemi p
+p^n &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p^n \fatsemi p^n
   \end{array}\]
@@
 
+\amr{let's write a a couple more small examples, reversible 3 bit
+  adder, etc.}
+
 In programming practice we are however more interested in structured
@@
 this short investigation using untyped flat sets is what the
-
-'\emph{purely operational} view of the theory would be. In particular,
-it tells us that permutations are an inescapable part of the fabric of
-reversible computing.  However as permutations are untyped, and act on
-the canonicalized version of $n$-element sets (i.e. those sets where
-all the structure has been forgotten), these are a rather pale shadow
-of the rich tapestry of information-perserving transformations of
-structured data which we investigate next.
+``\emph{purely operational}'' view of the theory would be. In
+particular, it tells us that permutations are an inescapable part of
+the fabric of reversible computing.  However as permutations are
+untyped, and act on the canonicalized version of $n$-element sets
+(i.e. those sets where all the structure has been forgotten), these
+are a rather pale shadow of the rich tapestry of
+information-perserving transformations of structured data which we
+investigate next.
 

### 2018-05-28 survey/survey.tex
sec 2

@@
 
-Some of the simplest spaces are the finite and discrete spaces of
-points. Indeed most programming languages provide some collection of
-primitive data like booleans, characters, strings, and (bounded)
-numbers that are naturally modeled as finite sets. Infinite sets are much
-more subtle, and will be discussed in the conclusion.
+Most programming languages provide primitive data like booleans,
+characters, strings, and (bounded) numbers that are naturally modeled
+as finite sets. We therefore start by modeling reversible computations
+over finite and discrete spaces of points. Infinite sets are more
+subtle, and will be discussed in the conclusion.
 
@@
 
+For convenience, we write $[2^n]$ for the finite set representing
+$n$-bit words with the canonical ordering for binary numbers. Thus
+when $n=3$, the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which
+correspond to the 3-bit words $\{000,001,010,011,100,101,110,111\}$.
 Although this language appears weak, it is universal for boolean
-combinational circuits. The result follows from the following
-construction of the Toffoli gate and the fact that any combinational
-circuit can be realized using compositions of Toffoli gates. We write
-$[2^n]$ for the finite set representing $n$-bit words with the
-canonical ordering for binary numbers. Thus when $n=3$, the finite set
-has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to the 3-bit words
-$\{000,001,010,011,100,101,110,111\}$. The Toffoli gate is a
-reversible function $[2^3] \rightarrow [2^3]$ which swaps $110$ and
-$111$, i.e., it can be realized as $\mathit{swap}~6~7$.
-
-\amr{let's write a a couple more small examples, reversible 3 bit
-  adder, etc.}
-
-In programming practice we are however more interested in structured
-data which will be the subject on next section. What we learn from
+combinational circuits. To see this, we express, boolean negation
+\emph{not} as a permutation $[2^1] \rightarrow [2^1]$, controlled not
+(also known as \emph{cnot}) as a permutation
+$[2^2] \rightarrow [2^2]$, and controlled controlled-not (also known
+as \emph{toffoli}) as a permutation $[2^3] \rightarrow [2^3]$:
+\[\begin{array}{rcl}
+\mathit{not} &=& \mathit{swap}~0~1 \\
+\mathit{cnot} &=& \mathit{swap}~2~3 \\
+\mathit{toffoli} &=& \mathit{swap}~6~7
+\end{array}\]
+For example, \emph{toffoli} gate negates the third bit (the target
+bit) if both the first two bits (the control bits) are 1, i.e., it
+swaps $110$ and $111$. Universality follows since any combinational
+circuit can be realized using compositions of \emph{toffoli} gates.
+
+There is however a subtle issue: programming in such an unstructured
+language is \emph{not} compositional in the sense that using
+$\mathit{not}$ in a larger circuit forces us to change its
+implementation. To illustrate this point, consider the reversible full
+adder below designed by Desoete et al.~\cite{117414}:
+
+\begin{center}
+\includegraphics[scale=0.6]{full-adder.png}
+\end{center}
+
+In the figure (copied from a more general paper that includes
+alternative designs~\cite{Rentergem2005OptimalDO}), the full adder
+takes 4 inputs: the two bits to add $A$ and $B$, an incoming carry bit
+$C_i$, and a heap input initialized to 0 to maintain
+reversibility. There are four outputs: the first two are identical to
+the incoming bits $A$ and $B$ and are considered ``garbage.'' The
+third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
+bit. In the notation used to describe the circuit, the $x$ denotes
+boolean negation, the dots are control bits. In our reversible
+language, we can express this circuit as the following permutation of
+type $[2^4] \rightarrow [2^4]$:
+
+\[\begin{array}{ll}
+\mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{toffoli} \\
+\mathit{swap}~8~12 \fatsemi \mathit{swap}~9~14 \fatsemi 
+    \mathit{swap}~10~13 \fatsemi \mathit{swap}~11~15 & \mbox{cnot and swap} \\
+\mathit{swap}~6~7 \fatsemi \mathit{swap}~14~15 & \mbox{toffoli} \\
+\mathit{swap}~4~6 \fatsemi \mathit{swap}~5~7 \fatsemi
+    \mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{cnot}
+\end{array}\]
+
+Note how the implementation of $\mathit{cnot}$ as a permutation
+$[2^2] \rightarrow [2^2]$ cannot be directly reused in a circuit
+$[2^4] \rightarrow [2^4]$. Indeed, in programming practice we are
+interested in structured data and compositional abstractions which
+will be the subject on next section. What we do learn however from
 this short investigation using untyped flat sets is what the

### 2018-05-30 survey/survey.tex
working on sec 3

@@
 product of two subspaces: the subspace $(\top \uplus \bot)$ which
-itself is the sum of the space $\top$ containing one element
+itself is the sum of the space~$\top$ containing one element
 $\texttt{tt}$ and the empty space $\bot$ and the subspace
@@
 A \emph{commutative semiring} (sometimes called a \emph{commutative
-  rig} (commutative ri\emph{n}g without negative elements)
+  rig} (commutative ri\emph{n}g without negative elements))
 $(R,0,1,+,\cdot)$ consists of a set $R$, two distinguished elements of
@@
 setting of structured types. First, types do not naturally want to be
-put together into a ``set''.  This can be fixed if we replace the set
+put together into a ``set.''  This can be fixed if we replace the set
 $R$ with a universe $U$, and replace the set membership $0 \in R$ with
@@
 proposition $A \times B \equiv B \times A$ is not
-normally\footnote{Except In homotopy type theory where equivalent
+normally\footnote{Except in univalent type theory where equivalent
   types are identified.} provable for arbitrary types $A$ and $B$. But
@@
 construction $c_1 \otimes c_2$ operates on a product structure, and
-applies $c_1$ to the first component and $c_2$ to the second.
+applies $c_1$ to the first component and $c_2$ to the second. The
+language of permutations lacked the ability to combine permutations by
+taking sums and products which led to the awkward non-compositional
+programming style illustrated in the full adder example. In this
+language the full adder can be programmed as follows:
+
+\amr{***HERE***}
 

### 2018-06-06 survey/survey.tex
one more pass over sec 1

@@
 is that the next computing revolution is likely to be founded on
-revised models of computation that are designed to in closer harmony
+revised models of computation that are designed to be in closer harmony
 with the laws of physics.
@@
 reversibile computer operations'' and an excursion into the origins of
-``irreversible computer operations'', we will develop in detail three
+``irreversible computer operations,'' we will develop in detail three
 reversible models of computation and discuss their potential
@@
 %%%%%%%%%
-\paragraph*{Maxwell's Demon.}
+\paragraph*{Maxwell's Daemon.}
 To fully appreciate the missing principle of ``reversibility'' in
@@
 to P. G Tait in 1867 -- the letter, whose ideas are now known as
-\emph{Maxwell's Demon}, tells of a thought experiment that seems to
+\emph{Maxwell's Daemon}, tells of a thought experiment that seems to
 indicate that intelligent beings can somehow violate the second law of
@@
 
-Maxwell's demon appears to violate the second law of thermodynamics by
+Maxwell's daemon appears to violate the second law of thermodynamics by
 having a tiny ``intelligence'' observing the movement of individual
@@
 moving ones, thereby reducing the total entropy of the
-system. Landauer's resolution of the demon relied on two ideas that
+system. Landauer's resolution of the daemon relied on two ideas that
 took root only a few decades earlier: the formal notion of computation
-(through the work of Turing and Church, 1936) and the formal notion of
-information (through the work of Shannon, 1948). Landauer reasoned
-that the computation done by the finite brain of the demon, involves
-getting information about the movement of molecules, storing that
-information, analyzing that information to act on it, and then --- and
-this is the critical step --- overwriting it to make room for the next
-computation.  In other words, the computation that is manipulating
-information in the demon's brain \textit{must be thermodynamic work},
-thereby bringing the demon back into the fold of physics.
+(through the work of Turing~\cite{turing}, Church~\cite{church51}, and
+others) and the formal notion of information (through the work of
+Shannon~\cite{shannon1948}). Landauer reasoned that the computation
+done by the finite brain of the daemon, involves getting information
+about the movement of molecules, storing that information, analyzing
+that information to act on it, and then --- and this is the critical
+step --- overwriting it to make room for the next computation.  In
+other words, the computation that is manipulating information in the
+daemon's brain \textit{must be thermodynamic work}, thereby bringing
+the daemon back into the fold of physics.
 
@@
 models of computation were developed, there was no compelling reason
-to the take the information content of computations into consideration
+to take the information content of computations into consideration
 -- in fact, at that time there was no quantifiable notion of
@@
 Current high-level programming languages as well as current hardware
-are both based on Boolean Logic -- which really ought to be called
-\emph{Piercean logic}, as Boole had a rather different conception of
-logic which is rather far from our current models.  Nevertheless,
-going back to Boole's 1853 book entitled \emph{An Investigation of the
-  Laws of Thought, on which are Founded the Mathematical Theories of
-  Logic and Probabilities}, we find that the opening sentence of Ch.~1 is:
+are both based on the mathematical formalization of logic developed by
+De Morgan, Venn, Boole, and Peirce in the mid to late 1800s. Going
+back to Boole's 1853 book entitled \emph{An Investigation of the Laws
+  of Thought, on which are Founded the Mathematical Theories of Logic
+  and Probabilities}, we find that the opening sentence of Ch.~1 is:
 \begin{quote}
   The design of the following treatise is to investigate the fundamental laws
-  of those operations of the mind by which reasoning is performed; \ldots
+  of those operations of the mind by which reasoning is performed; 
 \end{quote}
@@
 Machines that ``compute'' are quite old. M\"{u}ller (1786) first
-conceived of the idea of a ``difference machine'', which Babbage
+conceived of the idea of a ``difference machine,'' which Babbage
 (1819--1822) was able to construct. There are other computer
@@
 formal discipline. One of the pioneering works was Alan Turing's
-seminal paper in 1936 which established the idea that computation has
-a formal interpretation and that all computability can be captured
-within a formal system. Implicit in this achievement however is the
-idea that abstract models of computation are just that --
-\emph{abstractions of computation realized in the physical world.}
+seminal paper~\cite{turing} in 1936 which established the idea that
+computation has a formal interpretation and that all computability can
+be captured within a formal system. Implicit in this achievement
+however is the idea that abstract models of computation are just that
+-- \emph{abstractions of computation realized in the physical world.}
 Indeed, going back to Turing's 1936 article \emph{On Computable
@@
 \begin{quote}
-  We have said that the computable numbers are those whose decimals are
-  calculable by finite means \ldots the justification lies in the fact that
-  the human memory is necessarily limited.
+  We have said that the computable numbers are those whose decimals
+  are calculable by finite means [\ldots] the justification lies in
+  the fact that the human memory is necessarily limited.
 \end{quote}
@@
 then when both of the above assumptions are translated to the language
-of physics, they embody what is known as the ``Bekenstein bound'',
-which is an upper limit on the amount of information that can be
-contained within a given finite region of space. A detailed historical
-account of these ideas in the context of modern physics is again
-beyond our scope. However, the quotes above, like the ones before,
-should convey the ideas that our theories of computation and
-complexity are based on some physical assumptions that Turing and
-others found reasonable in the 1930s.
+of physics, they embody what is known as the ``Bekenstein
+bound''~\cite{PhysRevD.23.287}, which is an upper limit on the amount
+of information that can be contained within a given finite region of
+space. A detailed historical account of these ideas in the context of
+modern physics is again beyond our scope. However, the quotes above,
+like the ones before, should convey the ideas that our theories of
+computation and complexity are based on some physical assumptions that
+Turing and others found reasonable in the 1930s.
 
@@
 theory known to us and adapt computation to ``learn'' from these
-principles.  In the words of Girard:
+principles.  In the words of Girard~\cite{Girard:2007:TMI:1348911.1348915}:
 \begin{quote}
@@
   to learn from her. Instead of interpreting quantum into logic, we
-  shall interpret logic into quantum (Girard 2007).
+  shall interpret logic into quantum.
 \end{quote}
@@
 them! Because there are many such notions, we also need to walk our
-way through them to find the one which is ``just right''. Thus we seek
+way through them to find the one which is ``just right.'' Thus we seek
 a concept which is neither too strong nor too weak, that will express
-when some structured information should be treated as ``the same''.
+when some structured information should be treated as ``the same.''
 We can draw an analogy with topology: in topology, all point sets can
@@
 
-This discussion is purely ``semantic'', in the sense that it is about
+This discussion is purely ``semantic,'' in the sense that it is about
 the denotation of simple primitive data (sets) and their reversible
@@
 smoothly deform from one type to another. Let us thus analyze the crux
-of the ``problem''. In logic, we have that $\land$ and $\lor$ are both
+of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
 \emph{idempotent}: this is the property of any binary operation $*$
@@
 we have expressly disallowed.  Yes, this should remind you of
-Maxwell's demon: even though the data is the same, they are tagged
+Maxwell's daemon: even though the data is the same, they are tagged
 differently, and this data is indeed information, and must be
@@
 
-\amr{***HERE***}
+\amr{***HERE***; operational semantics; It feels like I should be
+  writing about how the combinators (both for spaces and for
+  'permutations') work at a lower level. Did we ever write that up
+  somewhere?  We have lots of Agda code which 'does it', nicely, but
+  somehow I can't remember an English prose version of it. I am not
+  sure how long we want to make it.  For example, do we talk about the
+  'matrix' representation of combinators-for-permutations-on-[n], and
+  how these make things like 0+a=a and associativity obvious by
+  construction?
+  Let’s have a geometric representation of spaces:
+- A + B means putting A next to B
+- A * B means putting A and B on two perpendicular axes and filling the square (volume) between them
+All the combinators would indeed have a natural action and if we have
+a few nice pictures it would be great. What do you think? }
 
@@
 we should looked at a \emph{typed} version. This process frequently
-goes by the moniker of ``categorification''.  We want a categorical
+goes by the moniker of ``categorification.''  We want a categorical
 algebra, where the basic objects are groupoids (to model our types),
@@
 
-DANGER: here, we leave the safe land of ``preservation of information'',
+DANGER: here, we leave the safe land of ``preservation of information,''
 because partiality is not reversible. So while trace gives a reversible
 program as output, it itself isn't a reversible operation. Like the
-demon's brain in Maxwell's thought experiment, trace irreversible work!
+daemon's brain in Maxwell's thought experiment, trace irreversible work!
 So trace, which is fundamental to all work with recursive types, is
@@
 types, rather than control-flow (or even data-flow) as its next layer
-of ``understanding'', this leads to equivalences, isomorphisms, etc.
+of ``understanding,'' this leads to equivalences, isomorphisms, etc.
 This then leads, quite naturally, to finding that the ``proof
@@
 equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
-as $-\sum p_i \log{p_i}$~\cite{Shannon1948}.  Consider a function
+as $-\sum p_i \log{p_i}$~\cite{shannon1948}.  Consider a function
 $f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
@@
 
-\bibliographystyle{acm}
+\bibliographystyle{unsrt}
 %\pagestyle{headings}

### 2018-06-06 survey/survey.tex
unicode fixed
comments

@@
 a few nice pictures it would be great. What do you think? }
+\jc{You are correct, I don't think we have any English prose which
+explains the operations of our combinators. The `matrix' representation
+makes sense for the untyped version, but is not necessarily adequate for
+the typed version, as it becomes quite an indirect way to view it, since you
+first have to assign numbers to each 'path' through a type as a proxy.
+Though if we want to have path-indexed instead of Fin-indexed matrices,
+I would be fine. So we could indeed still draw those pictures.  However,
+this would lead associativity of * to be an equation on two 3D matrices...}
 
@@
 \begin{figure}[t]
-Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t₃$, and $c_3 : t_3 \leftrightarrow t_4$:
+Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
 \[\def\arraystretch{1.3}
@@
 \begin{figure}[t]
-Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t₅ \leftrightarrow t₆$:
+Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
 \[\def\arraystretch{1.3}
@@
 \begin{figure}[t]
-Let $c₀ : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
+Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
 \[\def\arraystretch{1.3}
 \begin{array}{c}
-  {\identlp \odot c \Leftrightarrow (c₀ \oplus c) \odot \identlp}
+  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
 \qquad
-  {\identrp \odot (c₀ \oplus c) \Leftrightarrow c \odot \identrp}
-\\
-  {\identlsp \odot c \Leftrightarrow (c \oplus c₀) \odot \identlsp}
+  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
+\\
+  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
 \qquad
-  {\identrsp \odot (c \oplus c₀) \Leftrightarrow c \odot \identrsp}
+  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
 \\
@@
 Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
-Let $a_1 : t₅ \leftrightarrow t_1$,  $a_2 : t₆ \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
+Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
 \[\def\arraystretch{1.3}

### 2018-06-06 survey/survey.tex
before merge

@@
 Bijections on \fin{n} have a specific name: permutations. As is
-well-known permutations can be generated by sequential compositions of
+well-known, permutations can be generated by sequential compositions of
 transpositions. Thus we can create a small language for writing
@@
 \end{array}\]
-For example, \emph{toffoli} gate negates the third bit (the target
+For example, the \emph{toffoli} gate negates the third bit (the target
 bit) if both the first two bits (the control bits) are 1, i.e., it
@@
 a few nice pictures it would be great. What do you think? }
-\jc{You are correct, I don't think we have any English prose which
-explains the operations of our combinators. The `matrix' representation
-makes sense for the untyped version, but is not necessarily adequate for
-the typed version, as it becomes quite an indirect way to view it, since you
-first have to assign numbers to each 'path' through a type as a proxy.
-Though if we want to have path-indexed instead of Fin-indexed matrices,
-I would be fine. So we could indeed still draw those pictures.  However,
-this would lead associativity of * to be an equation on two 3D matrices...}
 
@@
 \section{Conclusion}
+
+Must mention infinite sets since we promised to do so in Sec 2
 

### 2018-06-07 survey/survey.tex
sec 2

@@
 correspond to the 3-bit words $\{000,001,010,011,100,101,110,111\}$.
-Although this language appears weak, it is universal for boolean
-combinational circuits. To see this, we express, boolean negation
-\emph{not} as a permutation $[2^1] \rightarrow [2^1]$, controlled not
-(also known as \emph{cnot}) as a permutation
-$[2^2] \rightarrow [2^2]$, and controlled controlled-not (also known
-as \emph{toffoli}) as a permutation $[2^3] \rightarrow [2^3]$:
+Although this language appears weak, it is universal for reversible
+boolean combinational circuits $[2^i] \rightarrow [2^i]$ with $i$
+input/output wires.
+
+We illustrate the expressiveness of the language, we develop a few
+small examples. We start by writing boolean negation ``not'' as a
+permutation $[2^1] \rightarrow [2^1]$, the controlled-not gate (also
+known as ``cnot'') as a permutation $[2^2] \rightarrow [2^2]$, and the
+controlled-controlled-not gate (also known as ``toffoli'') as a
+permutation $[2^3] \rightarrow [2^3]$:
 \[\begin{array}{rcl}
-\mathit{not} &=& \mathit{swap}~0~1 \\
-\mathit{cnot} &=& \mathit{swap}~2~3 \\
-\mathit{toffoli} &=& \mathit{swap}~6~7
+\mbox{not} &=& \mathit{swap}~0~1 \\
+\mbox{cnot} &=& \mathit{swap}~2~3 \\
+\mbox{toffoli} &=& \mathit{swap}~6~7
 \end{array}\]
-For example, the \emph{toffoli} gate negates the third bit (the target
+The ``cnot'' gate operates on two bits and negates the second (the
+target bit) if the first one (the control bit) is 1, i.e., it swaps
+$10$ and $11$; the ``toffoli'' gate negates the third bit (the target
 bit) if both the first two bits (the control bits) are 1, i.e., it
-swaps $110$ and $111$. Universality follows since any combinational
-circuit can be realized using compositions of \emph{toffoli} gates.
+swaps $110$ and $111$.
 
 There is however a subtle issue: programming in such an unstructured
-language is \emph{not} compositional in the sense that using
-$\mathit{not}$ in a larger circuit forces us to change its
-implementation. To illustrate this point, consider the reversible full
-adder below designed by Desoete et al.~\cite{117414}:
+language is \emph{not} compositional in the sense that using the
+``not'' gate in a larger circuit forces us to change its
+implementation. Indeed if we had two bits and wanted to use ``not'' to
+negate the first bit, we would write the permutation of type
+$[2^2] \rightarrow [2^2]$ that permutes $00$ with $10$ \emph{and}
+permutes $01$ with $11$, i.e, the permutation
+$\mathit{swap}~0~2 \fatsemi \mathit{swap}~1~3$. To further illustrate
+how inconvenient this point is, consider the reversible full adder
+below designed by Desoete et al.~\cite{117414}:
 
@@
 third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
-bit. In the notation used to describe the circuit, the $x$ denotes
+bit. In the notation used to describe the circuit, the $\times$ denotes
 boolean negation, the dots are control bits. In our reversible
@@
 \[\begin{array}{ll}
-\mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{toffoli} \\
+\mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 ~\fatsemi & \mbox{toffoli} \\
 \mathit{swap}~8~12 \fatsemi \mathit{swap}~9~14 \fatsemi 
-    \mathit{swap}~10~13 \fatsemi \mathit{swap}~11~15 & \mbox{cnot and swap} \\
-\mathit{swap}~6~7 \fatsemi \mathit{swap}~14~15 & \mbox{toffoli} \\
+    \mathit{swap}~10~13 \fatsemi \mathit{swap}~11~15 ~\fatsemi & \mbox{cnot and swap} \\
+\mathit{swap}~6~7 \fatsemi \mathit{swap}~14~15 ~\fatsemi & \mbox{toffoli} \\
 \mathit{swap}~4~6 \fatsemi \mathit{swap}~5~7 \fatsemi
@@
 Note how the implementation of $\mathit{cnot}$ as a permutation
-$[2^2] \rightarrow [2^2]$ cannot be directly reused in a circuit
-$[2^4] \rightarrow [2^4]$. Indeed, in programming practice we are
-interested in structured data and compositional abstractions which
-will be the subject on next section. What we do learn however from
-this short investigation using untyped flat sets is what the
-``\emph{purely operational}'' view of the theory would be. In
-particular, it tells us that permutations are an inescapable part of
-the fabric of reversible computing.  However as permutations are
-untyped, and act on the canonicalized version of $n$-element sets
-(i.e. those sets where all the structure has been forgotten), these
-are a rather pale shadow of the rich tapestry of
-information-perserving transformations of structured data which we
-investigate next.
+$[2^2] \rightarrow [2^2]$ cannot be directly reused in the larger
+circuit $[2^4] \rightarrow [2^4]$.
+
+For such reasons, in programming practice we are interested in
+structured data and compositional abstractions which will be the
+subject of the next section. What we do learn however from this short
+investigation using untyped flat sets is what the ``\emph{purely
+  operational}'' view of the theory would be. In particular, it tells
+us that permutations are an inescapable part of the fabric of
+reversible computing.  However as permutations are untyped, and act on
+the canonicalized version of $n$-element sets (i.e. those sets where
+all the structure has been forgotten), these are a rather pale shadow
+of the rich tapestry of information-perserving transformations of
+structured data which we investigate next.
 

### 2018-06-08 survey/survey.tex
small comments
_

@@
   include logical operations like conjunction, disjunction and
-  negation, as well as reading and writing to a (large) collection to
+  negation, as well as reading and writing to a large collection to
   memory locations. This perspective conforms to a macro-level view of
@@
   define what information is and what its conservation means in a
-  computational setting. We emphasize the idea that computations,
-  whatever they are, must be reversible transformations on
-  data. Thinking topologically, one can think of data as modeled using
-  topological spaces and programs as modeled by reversible
-  deformations of these spaces. We then illustrate this idea using
-  three notions of data and their associated reversible computational
-  models. The first instance only assumes unstructured finite data,
-  i.e., discrete topological spaces. The corresponding notion of
-  reversible computation is that of permutations. We show how this
-  simple model subsumes conventional computations on finite sets. We
-  then consider a modern structured notion of data based on the
-  Curry-Howard correspondence between logic and type theory. We
+  computational setting. We emphasize the idea that computations must
+  be reversible transformations on data. One can think of data as
+  modeled using topological spaces and programs as modeled by
+  reversible deformations of these spaces. We then illustrate this
+  idea using three notions of data and their associated reversible
+  computational models. The first instance only assumes unstructured
+  finite data, i.e., discrete topological spaces. The corresponding
+  notion of reversible computation is that of permutations. We show
+  how this simple model subsumes conventional computations on finite
+  sets. We then consider a modern structured notion of data based on
+  the Curry-Howard correspondence between logic and type theory. We
   develop the corresponding notion of reversible deformations using a
@@
 reflected in both early models of computations such as the Turing
-Machine and the RAM as well as in the early Von Neumann and in modern
-computer architectures, is that basic computer operations include
-logical operations like conjunction, disjunction, and negation, as
-well as reading from and writing to a large (infinite) collection of
-memory locations. From this small set of primitive operations emerges
-all higher-level programming languages and abstractions.
+Machine and the RAM as well as in the early Von Neumann models and in
+modern computer architectures, is that basic computer operations
+include logical operations like conjunction, disjunction, and
+negation, as well as reading from and writing to a large (infinite)
+collection of memory locations. From this small set of primitive
+operations emerges all higher-level programming languages and
+abstractions.
 
@@
 
-We illustrate the expressiveness of the language, we develop a few
+To illustrate the expressiveness of the language, we develop a few
 small examples. We start by writing boolean negation ``not'' as a
@@
 
+Operational Semantics:
+
 sound and complete isomorphisms for finite types
@@
 \end{verbatim}
+
+Implement Fredkin gate and mention the following:
+
+\begin{quote}
+  Ed Fredkin pursued the idea that information must be finite in
+  density. One day, he announced that things must be even more simple
+  than that. He said that he was going to assume that information
+  itself is conserved. “You’re out of you mind, Ed.” I
+  pronounced. “That’s completely ridiculous. Nothing could happen in
+  such a world. There couldn’t even be logical gates. No decisions
+  could ever be made.” But when Fredkin gets one of his ideas, he’s
+  quite immune to objections like that; indeed, they fuel him with
+  energy. Soon he went on to assume that information processing must
+  also be reversible — and invented what’s now called the Fredkin
+  gate. (Minsky 1999)
+\end{quote}
+
+%%%%%%%%%
+\subsection{Examples}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Implement Fredkin gate and mention the following:
-
-\begin{quote}
-  Ed Fredkin pursued the idea that information must be finite in
-  density. One day, he announced that things must be even more simple
-  than that. He said that he was going to assume that information
-  itself is conserved. “You’re out of you mind, Ed.” I
-  pronounced. “That’s completely ridiculous. Nothing could happen in
-  such a world. There couldn’t even be logical gates. No decisions
-  could ever be made.” But when Fredkin gets one of his ideas, he’s
-  quite immune to objections like that; indeed, they fuel him with
-  energy. Soon he went on to assume that information processing must
-  also be reversible — and invented what’s now called the Fredkin
-  gate. (Minsky 1999)
-\end{quote}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%
 \section{Data III: Reversible Programs between Reversible Programs}

### 2018-06-09 survey/survey.tex
_

@@
 
-Instead of spaces consisting solely of unstructured isolated points,
-we now investigate structured spaces built from sums and products of
-elementary spaces. This structure corresponds to the building blocks
-of type theory which are: the empty type ($\bot$), the unit type
-($\top$), the sum type ($\uplus$), and the product ($\times$)
-type. Before getting into the formal theory, let's consider possible
-deformations on the space
-$(\top \uplus \bot) \times (\top \uplus \top)$. This space is the
-product of two subspaces: the subspace $(\top \uplus \bot)$ which
-itself is the sum of the space~$\top$ containing one element
-$\texttt{tt}$ and the empty space $\bot$ and the subspace
-$(\top \uplus \top)$ which is the sum of two spaces each containing
-the one element $\texttt{tt}$. First, as discussed in the previous
-section, any deformation of this space must at least preserve the
-number of elements: we can neither create nor destroy points during
-any continuous deformation. Seeing that the number of elements in our
-example space is 2, a reasonable hypothesis is that we can deform the
-space above to any other space with 2 elements such as
-$\top \uplus \top$ or $\top \uplus (\top \uplus \bot)$. What this
-really means is that we are treating the sum and product structure as
-malleable. For example, imagining a product structure as arranged in a
-grid, by stretching we can turn this structure to a sum structure
-arranged in a line, change the orientation of the grid by exchanging
-the axes, as well as do other transformations to be discussed in this
-section that preserve the number of points.
+\roshan{Let's not substitute space for set without first motivating
+  the use of the word.}  Instead of spaces consisting solely of
+unstructured isolated points, we now investigate structured spaces
+built from sums and products of elementary spaces. This structure
+corresponds to the building blocks of type theory which are: the empty
+type ($\bot$), the unit type ($\top$), the sum type ($\uplus$), and
+the product ($\times$) type. Before getting into the formal theory,
+let's consider possible deformations on the space $(\top \uplus \bot)
+\times (\top \uplus \top)$. This space is the product of two
+subspaces: the subspace $(\top \uplus \bot)$ which itself is the sum
+of the space~$\top$ containing one element $\texttt{tt}$ and the empty
+space $\bot$ and the subspace $(\top \uplus \top)$ which is the sum of
+two spaces each containing the one element $\texttt{tt}$. First, as
+discussed in the previous section, any deformation of this space must
+at least preserve the number of elements: we can neither create nor
+destroy points during any continuous deformation. Seeing that the
+number of elements in our example space is 2, a reasonable hypothesis
+is that we can deform the space above to any other space with 2
+elements such as $\top \uplus \top$ or $\top \uplus (\top \uplus
+\bot)$. What this really means is that we are treating the sum and
+product structure as malleable. For example, imagining a product
+structure as arranged in a grid, by stretching we can turn this
+structure to a sum structure arranged in a line, change the
+orientation of the grid by exchanging the axes, as well as do other
+transformations to be discussed in this section that preserve the
+number of points.
 
@@
 items). Our next instinct would be to similarly replace $=$ with a
-judgement $A \equiv B$ that asserts that $A$ and $B$ are
-\emph{propositionally equal}. This is however not true: the
+type $A \equiv B$ that asserts that $A$ and $B$ are
+\emph{propositionally equal} such that a term exhibiting this type can
+then be shown in the formal system. This is however not true: the
 proposition $A \times B \equiv B \times A$ is not
@@
 equivalent information. In other words, we would like to be able to
-witness that $A \times B$ can be reversibly deformed into
-$B \times A$. Which motivates the introduction of type
-\emph{equivalences}. To do this, we need a few important auxiliary
-concepts.
-
+witness that $A \times B$ can be reversibly deformed into $B \times
+A$. Which motivates the introduction of type \emph{equivalences}. To
+do this, we need a few important auxiliary concepts.
+
+\roshan{Let's define $\equiv$ below or provide a readable
+  reference. We should not assume that its meaning is automatically
+  understood.}
 \begin{defn}[Homotopy]
@@
 Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
-this change, we can indeed prove that types (with
-$\bot, \top, \uplus, \times$) form a commutative semiring. The reader
-familiar with universal algebra should pause and ponder a bit about
-what we have done. We have lifted \emph{equality} from being in the
-signature of the ambient logic and instead put it in the signature of
-the algebraic structure of interest. Then, and only then, are we in a
-solid position to show that we have a concept (commutative semiring)
-which is flexible enough for types to be an instance.
+this change, we can indeed prove that types (with $\bot, \top, \uplus,
+\times$) form a commutative semiring. The reader familiar with
+universal algebra should pause and ponder a bit about what we have
+done. We have lifted \emph{equality} from being in the signature of
+the ambient logic and instead put it in the signature of the algebraic
+structure of interest. \roshan{The last coupe of sentences are
+  unclear. Let's strive to keep things easy to understand. Can we
+  spell them out the tradeoffs here and give an example or two to
+  clarify the point. If this is indeed an important aspect of things
+  let's spend some space explaining it.} Then, and only then, are we
+in a solid position to show that we have a concept (commutative
+semiring) which is flexible enough for types to be an instance.
 
@@
 something more subtle happens. Consider $\top \uplus \top$
-versus $\top$; the first has \emph{two} proofs of inhabitation
+versus $\top$; the first can have \emph{two} proofs of inhabitation
 ($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one

### 2018-06-10 survey/survey.tex
major reorganization

@@
 \paragraph*{Programs as Reversible Deformations.}
-
 To better understand the essence of ``conservation of information'' in
@@
 as to be reversible.
+
+\amr{We should also have a quick survey of other works in reversible
+computation? Most of the other work tries to start from well known
+models of computation or well known programming languages, and then
+adapts them to be reversible. Pi is different in that it starts from
+the semantic implications of reversibility. Then, because it adds
+types, rather than control-flow (or even data-flow) as its next layer
+of ``understanding,'' this leads to equivalences, isomorphisms, etc.
+This then leads, quite naturally, to finding that the ``proof
+language'' of semirings (and Rig Groupoids at level 2) is actually a
+programming language. And it is Pi. This is a neat twist on
+Curry-Howard because CH is about \textbf{inhabitation} only. But with
+``conservation of information'' as the basis, a different kind of
+correspondence arises; in fact, this one may well be an actual
+isomorphism.}
 
@@
 
-\amr{***HERE***; operational semantics; It feels like I should be
-  writing about how the combinators (both for spaces and for
-  'permutations') work at a lower level. Did we ever write that up
-  somewhere?  We have lots of Agda code which 'does it', nicely, but
-  somehow I can't remember an English prose version of it. I am not
-  sure how long we want to make it.  For example, do we talk about the
-  'matrix' representation of combinators-for-permutations-on-[n], and
-  how these make things like 0+a=a and associativity obvious by
-  construction?
-  Let’s have a geometric representation of spaces:
-- A + B means putting A next to B
-- A * B means putting A and B on two perpendicular axes and filling the square (volume) between them
-All the combinators would indeed have a natural action and if we have
-a few nice pictures it would be great. What do you think? }
-\jc{You are correct, I don't think we have any English prose which
-explains the operations of our combinators. The `matrix' representation
-makes sense for the untyped version, but is not necessarily adequate for
-the typed version, as it becomes quite an indirect way to view it, since you
-first have to assign numbers to each 'path' through a type as a proxy.
-Though if we want to have path-indexed instead of Fin-indexed matrices,
-I would be fine. So we could indeed still draw those pictures.  However,
-this would lead associativity of * to be an equation on two 3D matrices...}
-
 The type-erase of this language \emph{should} be permutations. But
@@
 
-\jc{We should decide what you level convention is. Do we say
-that $\iso$ is level 0, so that $\Leftrightarrow$ is level 1?
-We need to be consistent about it.}
-
-\jc{wavefront}
-
-Operational Semantics:
-
 sound and complete isomorphisms for finite types
 
-\begin{verbatim}
-examples:
-  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
-  conditionals
-  toffoli
-  fredkin
-\end{verbatim}
-
-Implement Fredkin gate and mention the following:
+%%%%%%%%%
+\subsection{Operational Semantics}
+
+%%%%%%%%%
+\subsection{Graphical Language}
+
+a graphical notation for programming in these models that reminiscent
+of Penrose diagrams for categories~\cite{selinger-graphical}, Geometry
+of Interaction (GoI) machines and Proof
+Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
+notation, computation is modeled by the flow of particles in a
+circuit.
+
+\amr{
+  Let’s have a geometric representation of spaces:
+- A + B means putting A next to B
+- A * B means putting A and B on two perpendicular axes and filling
+the square (volume) between them
+All the combinators would indeed have a natural action and if we have
+a few nice pictures it would be great. What do you think? }
+
+%%%%%%%%%
+\subsection{Examples}
+
+Surprising expressiveness of Pi:
 
@@
 
-%%%%%%%%%
-\subsection{Examples}
+\begin{verbatim}
+examples:
+  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
+  conditionals
+  toffoli
+  fredkin
+  revisit adder example from Sec 2 
+  perhaps a larger example as well (from RC 18 paper?)
+\end{verbatim}
 
@@
 \end{array}\]
-\caption{\label{figj}Signatures of level-1 $\Pi$-combinators: associativity}
+\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figi}Signatures of level-1 $\Pi$-combinators: distributivity and factoring}
+\caption{\label{figi}Signatures of level-2 $\Pi$-combinators: distributivity and factoring}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figh}Signatures of level-1 $\Pi$-combinators: identity and composition}
+\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figg}Signatures of level-1 $\Pi$-combinators: unit}
+\caption{\label{figg}Signatures of level-2 $\Pi$-combinators: unit}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figf}Signatures of level-1 $\Pi$-combinators: commutativity and associativity}
+\caption{\label{figf}Signatures of level-2 $\Pi$-combinators: commutativity and associativity}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{fige}Signatures of level-1 $\Pi$-combinators: functors}
+\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figd}Signatures of level-1 $\Pi$-combinators: unit and associativity}
+\caption{\label{figd}Signatures of level-2 $\Pi$-combinators: unit and associativity}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figc}Signatures of level-1 $\Pi$-combinators: zero}
+\caption{\label{figc}Signatures of level-2 $\Pi$-combinators: zero}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figb}Signatures of level-1 $\Pi$-combinators: associativity and distributivity}
+\caption{\label{figb}Signatures of level-2 $\Pi$-combinators: associativity and distributivity}
 \end{figure}
@@
 \end{array}\]
-\caption{\label{figa}Signatures of level-1 $\Pi$-combinators: commutativity and distributivity}
+\caption{\label{figa}Signatures of level-2 $\Pi$-combinators: commutativity and distributivity}
 \end{figure}
@@
 \section{Data III: Reversible Programs between Reversible Programs}
+
+\begin{verbatim}
+Mention convention:
+
+level 0: types
+level 1: type equivalences (pi combinators)
+level 2: equivalences between equivalences
+
+(cite MFPS invited paper)
+\end{verbatim}
 
@@
 
+%%%%%%%%%
+\subsection{A Model of Equivalences between Type Equivalences}
+
 In the previous section, we saw how we could take the proof
@@
 
-\amr{*** another wavefront ***}
+%%%%%%%%%
+\subsection{A Language of Equivalences between Type Equivalences}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Data IV: Circle ???}
-
-Would be interesting
+%%%%%%%%%
+\subsection{Operational Semantics}
+
+%%%%%%%%%
+\subsection{Graphical Language}
+
+%%%%%%%%%
+\subsection{Examples}
 
@@
 
-Must mention infinite sets since we promised to do so in Sec 2
-
-Discuss recursion, or add a section of infinite sets???
-
-CS abstractions are based on ``old'' physics
-
-computer applications are more and more ``physical''
-
-physical principles such as conservation of information should be part
-of our foundational abstractions
-
-quest for other principles; other data models; graphs; HoTT
-
-Data IV: Recursive Types
-
-Add recursion to structured finite types, trace, feedback, loop...
+Possible extensions; quest for other principles; other data models; graphs; HoTT
+
+%%%
+\subsection{Infinite Sets}
+
+Must mention infinite sets since we promised to do so in Sec 2;
+Discuss recursion, trace, feedback, loop...
 
@@
 
-meta-circular interpreter
-
-DANGER: here, we leave the safe land of ``preservation of information,''
-because partiality is not reversible. So while trace gives a reversible
-program as output, it itself isn't a reversible operation. Like the
-daemon's brain in Maxwell's thought experiment, trace irreversible work!
-So trace, which is fundamental to all work with recursive types, is
-only reversible "at level 1". So we need to replace trace with some other
-operation for level 2 reversibility to also hold -- and this is an open
-problem.
-
-Data V: Exclusive Disjunctions
+meta-circular interpreter from RC paper: Since these new models are
+substantially different from \lcal\ or familiar high-level languages,
+a point of concern is how one can effectively develop programs in
+them.  We address this by presenting a straightforward technique for
+deriving $\Pi$ programs by systematically translating logically
+reversible small-step abstract machines~\cite{isoint}. Complex
+programs can be devised in this way; we demonstrate the derivation of
+a meta-circular interpreter for $\Pi$.
+
+DANGER: here, we leave the safe land of ``preservation of
+information,'' because partiality is not reversible. So while trace
+gives a reversible program as output, it itself isn't a reversible
+operation. Like the daemon's brain in Maxwell's thought experiment,
+trace irreversible work!  So trace, which is fundamental to all work
+with recursive types, is only reversible "at level 1". So we need to
+replace trace with some other operation for level 2 reversibility to
+also hold -- and this is an open problem.
+
+%%%
+\subsection{Topological Spaces}
+
+Data IV: Circle etc. Would be interesting
+
+%%%
+\subsection{Theseus}
+
+%%%
+\subsection{Quantum Computing}
+
+Extension of Theseus in 2018 FoSSaCS paper.
+
+Exclusive Disjunctions
 
@@
 saying.}
-Indeed one should take the physical principles underlying quantum mechanics,
-the most successful physical theory known to us and adapt computation to
-``learn'' from these principles. To illustrate the depth of our crisis, Scott
-Aaronson, Umesh Vazirani, and others have proposed the following puzzle.
+
+Indeed one should take the physical principles underlying quantum
+mechanics, the most successful physical theory known to us and adapt
+computation to ``learn'' from these principles. To illustrate the
+depth of our crisis, Scott Aaronson, Umesh Vazirani, and others have
+proposed the following puzzle.
 
@@
 
-%% Below are pieces of text that are not immediately useful above, but
-%% from which we might still be able to pilfer usefully.
-
-%%%%%%%%%
-\subsection{Technical Overview}
-
-The technical development derives from ideas which are implicit in the
-works of \cite{toffoli:1980, Zuliani:2001:LR,
-  malacaria2007assessing,ClarkHM07,Ghica:2007:GSS:1190216.1190269}.
-(1) Functions whose output entropy is equal to their input entropy are
-information preserving functions.  (2) A function
-$f : b_1 \rightarrow b_2$ is \emph{logically reversible} if there
-exists an inverse function $f^{\dagger}$ such that for any input
-$v_1:b_1$ if there exists $v_2:b_2$ such that $f(v_1)=v_2$ \emph{iff}
-$f^{\dagger}(v_2)=v_1$.  (3) Logically reversible functions are
-information preserving.
-
-To summarize:
-\begin{enumerate}
-\item Based on isomorphisms of sum and product
-  types~\cite{Fiore:2004}, we develop a strong normalizing model of
-  computing, which we call $\Pi$, that is logically reversible and
-  information preserving. Every computation expressible in $\Pi$
-  is an isomorphism.
-
-\item We extend $\Pi$ with isorecursive type and trace operators from
-  category theory to obtain $\Pi$. Every computation expressible in
-  $\Pi$ is a partial isomorphism --- i.e. the system admits
-  non-termination. $\Pi$ is Turing complete while being information
-  preserving and logically reversible.
-
-\item We develop the categorical semantics of these
-  models~\cite{rc2011}. \amr{Mention n-categories.} We also develop
-  a graphical notation for programming in these models that
-  reminiscent of Penrose diagrams for
-  categories~\cite{selinger-graphical}, Geometry of Interaction (GoI)
-  machines and Proof
-  Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
-  notation, computation is modeled by the flow of particles in a
-  circuit.
-
-\item Since these new models are substantially different from \lcal\
-  or familiar high-level languages, a point of concern is how one can
-  effectively develop programs in them.  We address this by presenting
-  a straightforward technique for deriving $\Pi$ programs by
-  systematically translating logically reversible small-step abstract
-  machines~\cite{isoint}. Complex programs can be devised in this way;
-  we demonstrate the derivation of a meta-circular interpreter for
-  $\Pi$.
-
-\item We develop an arrow meta-language to encapsulate information
-  effects. This metalanguage serves to encapsulate effects, in much
-  the same way that traditional arrows or monads serve to encapsulate
-  effects over the \lcal.
-
-\item We show how a conventional irreversible model can be expressed
-  in our model. We compile the first-order fragment of typed \lcal\
-  extended with sums, products and loops to $\Pi$. This compilation is
-  interesting for two reasons: (1) it exposes the implicit information
-  effects of \lcal\ in the compilation to $\Pi$ and (2) the
-  compilation of $\Pi$ to $\Pi$ shows that information effects can be
-  erased by treating them as interactions with an explicit information
-  environment.
-
-\item We develop the notion of `duality' in $\Pi$ which gives us not
-  one duality (like in linear logic or Filinski's symmetric
-  \lcal~\cite{Filinski:89}), but two notions of duality -- an additive
-  duality and a multiplicative duality. This gives a crisp semantics
-  for negative and fractional types in the context of $\Pi$.
-
-\end{enumerate}
-
-Several intriguing areas of exploration remain and the book concludes
-with discussion of some of these, along with other possible
-applications.  Of these, the field of reversible and Quantum computing
-is closely related and we survey several connections in depth.  The
-exact connection between $\Pi$ and linear logic~\cite{Girard87tcs} is
-not fully understood though it is clear that both systems capture
-related but different notions of resource usage. A closely related
-area is the connection with duality of computation \cite{Filinski:89,
-  DBLP:conf/icfp/CurienH00, 10.1109/LICS.2010.23,Wadler:2003} ---
-i.e., the quest for a unifying framework between `computations',
-values, continuations and their logical counterparts. The
-investigation of duality gives rise to negative and fractional types
-which have an intuitive semantics and are reminiscent of negative
-information flow, superposition and entanglement from Quantum
-Physics~\cite{piee}.
-
-%%%%%%%%%
-\paragraph*{Conservation of Information.}
+%%%
+\subsection{Information Flow}
 
@@
 
-% verbatim for now, will go to something decent later
 \begin{verbatim}
@@
 
-\amr{We should also have a quick survey of other works in reversible
-computation? Most of the other work tries to start from well known
-models of computation or well known programming languages, and then
-adapts them to be reversible. Pi is different in that it starts from
-the semantic implications of reversibility. Then, because it adds
-types, rather than control-flow (or even data-flow) as its next layer
-of ``understanding,'' this leads to equivalences, isomorphisms, etc.
-This then leads, quite naturally, to finding that the ``proof
-language'' of semirings (and Rig Groupoids at level 2) is actually a
-programming language. And it is Pi. This is a neat twist on
-Curry-Howard because CH is about \textbf{inhabitation} only. But with
-``conservation of information'' as the basis, a different kind of
-correspondence arises; in fact, this one may well be an actual
-isomorphism.}
-
-%%%%%%%%%
-\paragraph*{Information.}
-We can now give a first precise definition of information preservation.
-Let
-`$b$' be a (not necessarily finite) type whose values are labeled
-$b^1, b^2, \ldots$. Let $\xi$ be a random variable of type $b$ that is
-equal to $b^i$ with probability $p_i$. The entropy of $\xi$ is defined
-as $-\sum p_i \log{p_i}$~\cite{shannon1948}.  Consider a function
-$f : b_1 \rightarrow b_2$ where $b_2$ is a (not necessarily finite)
-type whose values are labeled $b_2^1, b_2^2, \ldots$. The output
-entropy of the function is given by $- \sum q_j \log{q_j}$ where $q_j$
-indicates the probability of the output of the function to have value
-${b_2}^j$. We say a function is \emph{information-preserving} if its
-output entropy is equal to the entropy of its input. In later sections,
-we will refine this view.
-
-\amr{
+%%%
+\subsection{Conservation of Information, Information Effects, and Duality}
+
 Talk about conservation of information in biology, computer science,
@@
 \url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
-}
-
-In our current age, both Landauer's Principle and recent investigations
-of computational models well-suited to quantum mechanics, push us to
-revisit our current models of computation. Information, on top of
-having a physical manifestation, appears to be \emph{conserved},
-along with other quantities like energy, mass and momentum.
-
-But what does it mean for a computation to preserve information?
-What is the missing principle that will guide the creation of a new
-model of computation that ``conserve information''?
-It is nature to look again at physics for inspiration --
-in our current understanding of physics all of the primitive rules
-are reversible. Conventional computation is not. Can
-we embrace this simple principle as the building block of a model of
-computation? What would computation look like when viewed from this
-vantage point? Can non-trivial computation be performed in a
-reversible manner? And, if so, does it have applications?
-
-Traditional models of computation are all unityped, generally
-operating on unstructured sequences of symbols. But recent
-investigations, in the intersection between mathematics and
-computer science (particularly in Homotopy Type Theory~\cite{hottbook})
-reveal that not only can mathematics be typed, but that doing so
-reveals further structure -- in both mathematics and in computation.
-
-And that brings us to the principal objective of this survey: explore the
-domain of \emph{typed reversible computation}.
-
-\amr{In the quantum case, this is clear. Say something about topological
-quantum computing.}
-
-\textbf{punch line: allowable operations that computers can perform
-  while embracing the laws of physics are operations that preserve
-  everthing in abstract space of ``data''; doughnut into cup kind of
-  transformations.}
-
+
+We develop an arrow meta-language to encapsulate information
+effects. This metalanguage serves to encapsulate effects, in much the
+same way that traditional arrows or monads serve to encapsulate
+effects over the \lcal.
+
+new paper \url{https://arxiv.org/pdf/1805.08605.pdf}
+
+Mention connections to negative, fractional, and algebraic numbers
+
+%%%
+\subsection{Summary}
+
+CS abstractions are based on ``old'' physics
+
+computer applications are more and more ``physical''
+
+physical principles such as conservation of information should be part
+of our foundational abstractions
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements} We would like to thank the numerous
+students and colleagues who participated in various aspects of this
+research and who provided valuable feedback and constructive criticism.
+
+\bibliographystyle{unsrt}
+%\pagestyle{headings}
+\bibliography{cites}
+\end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{comment}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \ctr{Ctxt} values as we did in the \ctr{treeWalk} program.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements} We would like to thank the numerous
-students and colleagues who participated in various aspects of this
-research and who provided valuable feedback and constructive criticism.
-
-\bibliographystyle{unsrt}
-%\pagestyle{headings}
-\bibliography{cites}
-\end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+}

### 2018-06-11 survey/survey.tex
added a few pieces from other papers

@@
 \documentclass{article}
+\usepackage[utf8x]{inputenc}
+\usepackage{agda}
+\usepackage[nocenter]{qtree}
 \usepackage{amssymb,amsthm,amsmath}
@@
 
+\DeclareUnicodeCharacter{9678}{\ensuremath{\odot}}
+\DeclareUnicodeCharacter{9636}{\ensuremath{\Box}}
+%% shorten the longarrow
+\DeclareUnicodeCharacter{10231}{\ensuremath{\leftrightarrow}}
+
 \title{Embracing the Laws of Physics: \\ Three Reversible Models of Computation}
@@
 
+We illustrate our methodology using a small example. Consider a
+circuit that takes an input type consisting of three values
+\Tree [ {\small a} [ {\small b} {\small c} ] ]~
+and swaps the leftmost value with the rightmost value
+to produce
+\Tree [ {\small c} [ {\small b} {\small a} ] ]~.
+We can implement two such circuits using our Agda library for $\Pi$:
+
+\begin{code}%
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{a}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaDatatype{U}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaDatatype{⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSymbol{)}\<%
+\\
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\\
+%
+\\[\AgdaEmptyExtraSkip]%
+\>[0]\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{=}%
+\>[52I]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+\>[.]\<[52I]%
+\>[11]\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\end{code}
+
+\noindent The first implementation rewrites the incoming values as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent
+The second implementation rewrites the incoming values as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
+\Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
+\Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent The two circuits are extensionally equal. Using the level-2
+isomorphisms we can \emph{explicitly} construct a sequence of
+rewriting steps that transforms the second circuit to the first.  The proof
+can be read as follows: the first three lines ``refocus'' from a right-associated
+isomorphism onto the (left-associated) composition of the first $3$ isomorphisms;
+then apply a complex rewrite on these (the ``hexagon'' coherence condition
+of symmetric braided monoidal categories); this exposes two inverse combinators
+next to each other --- so we have to refocus on these to eliminate them; we
+finally re-associate to get the result.
+
+\begin{code}%
+\>[0]\AgdaFunction{swap{-}fl2⇔swap{-}fl1}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{a}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaDatatype{U}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{a}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{b}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{c}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaDatatype{⇔}\AgdaSpace{}%
+\AgdaFunction{swap{-}fl1}\<%
+\\
+\>[0]\AgdaFunction{swap{-}fl2⇔swap{-}fl1}\AgdaSpace{}%
+\AgdaSymbol{=}\<%
+\\
+\>[0][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+\>[0][@{}l@{\AgdaIndent{0}}]%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{(((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{hexagonl⊕r}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{(((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}%
+\>[44]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎r}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}%
+\>[42]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSymbol{)}%
+\>[44]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{linv◎l}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}%
+\>[28]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaInductiveConstructor{idl◎l}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}%
+\>[22]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\\
+%
+\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎r}\AgdaSpace{}%
+\AgdaFunction{⟩}\<%
+\\
+\>[1][@{}l@{\AgdaIndent{0}}]%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\AgdaSpace{}%
+\AgdaFunction{▤}\AgdaSymbol{)}\<%
+\end{code}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Theseus}
+
+The most elementary control structure in a programming language is the
+ability to conditionally execute one of several possible code
+fragments. Expressing such an abstraction using predicates and nested
+\textbf{if}-expressions makes it difficult for both humans and
+compilers to reason about the control flow structure. Instead, in
+modern functional languages, this control flow paradigm is elegantly
+expressed using \emph{pattern-matching}. This approach yields code
+that is not only more concise and readable but also enables the
+compiler to easily verify two crucial properties: (i) non-overlapping
+patterns and (ii) exhaustive coverage of a datatype using a collection
+of patterns. Indeed most compilers for functional languages perform
+these checks, warning the user when they are violated. At a more
+fundamental level, e.g., in type theories and proof assistants, these
+properties are actually necessary for correct reasoning about
+programs. Our first insight, explained in this section, is that these
+properties, perhaps surprisingly, are sufficient to produce a simple
+and intuitive first-order reversible programming language.
+
+\begin{figure}[t]
+  \centering
+  \begin{minipage}{0.4\linewidth}
+    \begin{verbatim}
+f :: Either Int Int -> a
+f (Left 0)     = undefined
+f (Left (n+1)) = undefined
+f (Right n)    = undefined
+\end{verbatim}\vspace{-4ex}
+    \caption{A skeleton}\label{fig:intro-ex-1}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.4\linewidth}
+\begin{verbatim}
+g :: (Bool,Int) -> a
+g (False,n)  = undefined
+g (True,0)   = undefined
+g (True,n+1) = undefined
+\end{verbatim}\vspace{-4ex}
+    \caption{Another skeleton}\label{fig:intro-ex-2}
+  \end{minipage}
+  \\[3ex]
+  \begin{minipage}{0.6\linewidth}
+\begin{verbatim}
+h :: Either Int Int <-> (Bool,Int)
+h (Left 0)     = (True,0)
+h (Left (n+1)) = (False,n)
+h (Right n)    = (True,n+1)
+\end{verbatim}\vspace{-4ex}
+    \caption{An isomorphism}\label{fig:intro-ex-3}
+  \end{minipage}
+\end{figure}
+
+We start with a small illustrative example, written in a Haskell-like
+syntax.  Fig.~\ref{fig:intro-ex-1} gives the skeleton of a function
+\verb|f| that accepts a value of type \verb|Either Int| \verb|Int|; the
+patterns on the left-hand side exhaustively cover every possible
+incoming value and are non-overlapping. Similarly,
+Fig.~\ref{fig:intro-ex-2} gives the skeleton for a function~\verb|g|
+that accepts a value of type \verb|(Bool,Int)|; again the patterns on
+the left-hand side exhaustively cover every possible incoming value
+and are non-overlapping. Now we claim that since the types
+\verb|Either Int Int| and \verb|(Bool,Int)| are isomorphic, we can
+combine the patterns of \verb|f| and \verb|g| into \emph{symmetric
+  pattern-matching clauses} to produce a reversible function between
+the types \verb|Either Int Int| and
+\verb|(Bool,Int)|. Fig.~\ref{fig:intro-ex-3} gives one such function;
+there, we suggestively use \verb|<->| to indicate that the function
+can be executed in either direction. This reversible function is
+obtained by simply combining the non-overlapping exhaustive patterns
+on the two sides of a clause. In order to be well-formed in either
+direction, these clauses are subject to the constraint that each
+variable occurring on one side must occur exactly once on the other
+side (and with the same type). Thus it is acceptable to swap the
+second and third right-hand sides of \verb|h| but not the first and
+second ones.
 

### 2018-06-11 survey/survey.tex
cite mfps

@@
 \section{Data II: Structured Finite Types}
+\label{sec:pi1}
 
@@
 \section{Data III: Reversible Programs between Reversible Programs}
-
-\begin{verbatim}
-Mention convention:
-
-level 0: types
-level 1: type equivalences (pi combinators)
-level 2: equivalences between equivalences
-
-(cite MFPS invited paper)
-\end{verbatim}
+\label{sec:pi2}
 
@@
 are the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
+
+Our reversible language of type isomorphisms and equivalences between
+them has a strong connection to \emph{univalent universes} in
+HoTT~\cite{CARETTE20185}. Based on this connection, we refer to the
+types as being at level-0, to the equivalences between types (i.e., the
+combinators of Sec.~\ref{sec:pi1}) as being at level-1, and to the
+equivalences between equivalences of types (i.e., the combinators
+discussed in this section) as being at level-2.
 

### 2018-06-11 survey/survey.tex
working on sec 4

@@
 \begin{array}{rrcll}
-\idc :& \tau & \iso & \tau &: \idc \\
-\\
-\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
-\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
-\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
-\\
-\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
-\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
-\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
-\\
-\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
-\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
+\idc :& t & \iso & t &: \idc \\
+\\
+\identlp :&  0 + t & \iso & t &: \identrp \\
+\swapp :&  t_1 + t_2 & \iso & t_2 + t_1 &: \swapp \\
+\assoclp :&  t_1 + (t_2 + t_3) & \iso & (t_1 + t_2) + t_3 &: \assocrp \\
+\\
+\identlt :&  1 * t & \iso & t &: \identrt \\
+\swapt :&  t_1 * t_2 & \iso & t_2 * t_1 &: \swapt \\
+\assoclt :&  t_1 * (t_2 * t_3) & \iso & (t_1 * t_2) * t_3 &: \assocrt \\
+\\
+\distz :&~ 0 * t & \iso & 0 ~ &: \factorzl \\
+\dist :&~ (t_1 + t_2) * t_3 & \iso & (t_1 * t_3) + (t_2 * t_3)~ &: \factor
 \end{array}
@@
 \Rule{}
-{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
-{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
+{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_2 \iso t_3}
+{\jdg{}{}{c_1 \odot c_2 : t_1 \iso t_3}}
 {}
@@
 \Rule{}
-{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
-{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
+{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
+{\jdg{}{}{c_1 \oplus c_2 : t_1 + t_3 \iso t_2 + t_4}}
 {}
@@
 \Rule{}
-{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
-{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
+{\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
+{\jdg{}{}{c_1 \otimes c_2 : t_1 * t_3 \iso t_2 * t_4}}
 {}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[t]
-Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
-\\
-  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
-\\
-  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
-\\
-  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
-\\
-  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
-\\
-  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
-\\
-  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
-\end{array}\]
-\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
-\end{figure}
-
-\begin{figure}[t]
-Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
-\\
-  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
-\\
-  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
-\\
-  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
-\end{array}\]
-\caption{\label{figi}Signatures of level-2 $\Pi$-combinators: distributivity and factoring}
-\end{figure}
-
-\begin{figure}[t]
-Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\idc \odot \, c \Leftrightarrow c}
-\quad
-  {c \, \odot \idc \, \Leftrightarrow c}
-\quad
-  {c\,\, \odot\,! c \Leftrightarrow \idc}
-\quad
-  {! c \odot c \Leftrightarrow \idc}
-\\
-  {c \Leftrightarrow c}
-\quad
-\Rule{}
-  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
-  {c_1 \Leftrightarrow c_3}
-  {}
-\quad
-\Rule{}
-  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
-  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
-  {}
-\end{array}\]
-\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
-\end{figure}
-
-\begin{figure}[t]
-Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
-\qquad
-  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
-\\
-  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
-\qquad
-  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
-\\
-  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
-\qquad
-  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
-\\
-  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
-\qquad
-  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
-\\
-  {\identlt \Leftrightarrow \distl \odot (\identlt \oplus \identlt)}
-\\
-\identlp \Leftrightarrow \swapp \odot \identrsp
-\qquad
-\identlt \Leftrightarrow \swapt \odot \identrst
-\end{array}\]
-\caption{\label{figg}Signatures of level-2 $\Pi$-combinators: unit}
-\end{figure}
-
-\begin{figure}[t]
-Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\swapp \odot (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \odot \swapp}
-\quad
-  {\swapt \odot (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \odot \swapt}
-\\
-  {(\assocrp \odot \swapp) \odot \assocrp \Leftrightarrow ((\swapp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \swapp)}
-\\
-  {(\assoclp \odot \swapp) \odot \assoclp \Leftrightarrow ((\idc \oplus \swapp) \odot \assoclp) \odot (\swapp \oplus \idc)}
-\\
-  {(\assocrt \odot \swapt) \odot \assocrt \Leftrightarrow ((\swapt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \swapt)}
-\\
-  {(\assoclt \odot \swapt) \odot \assoclt \Leftrightarrow ((\idc \otimes \swapt) \odot \assoclt) \odot (\swapt \otimes \idc)}
-\end{array}\]
-\caption{\label{figf}Signatures of level-2 $\Pi$-combinators: commutativity and associativity}
-\end{figure}
-
-\begin{figure}[t]
-Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
-Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-\Rule{}
-  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
-  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
-  {}
-\qquad
-\Rule{}
-  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
-  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
-  {}
-\\
-  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
-\qquad
-  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
-\\
-  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
-\\
-  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
-\end{array}\]
-\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
-\end{figure}
-
-\begin{figure}[t]
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \odot (\idc \oplus \, \identlp)}
-\\
-  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \odot (\idc \otimes \, \identlt)}
-\end{array}\]
-\caption{\label{figd}Signatures of level-2 $\Pi$-combinators: unit and associativity}
-\end{figure}
-
-\begin{figure}[t]
-Let $c : t_1 \leftrightarrow t_2$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {(c \otimes \idc) \odot \absorbl \Leftrightarrow \absorbl \odot \idc}
-\quad
-  {(\idc \, \otimes c) \odot \absorbr \Leftrightarrow \absorbr \odot \idc}
-\\
-  {\idc \odot \, \factorzl \Leftrightarrow \factorzl \odot (\idc \otimes c)}
-\quad
-  {\idc \odot \, \factorzr \Leftrightarrow \factorzr \odot (c \otimes \idc)}
-\\
-  {\absorbr \Leftrightarrow \absorbl}
-\\
-  {\absorbr \Leftrightarrow (\distl \odot (\absorbr \oplus \absorbr)) \odot \identlp}
-\\
-  {\identlst \Leftrightarrow \absorbr}
-\qquad
-  {\absorbl \Leftrightarrow \swapt \odot \absorbr}
-\\
-  {\absorbr \Leftrightarrow (\assoclt \odot (\absorbr \otimes \idc)) \odot \absorbr}
-\\
-  {(\idc \otimes \absorbr) \odot \absorbl \Leftrightarrow (\assoclt \odot (\absorbl \otimes \idc)) \odot \absorbr}
-\\
-  {\idc \otimes \, \identlp \Leftrightarrow (\distl \odot (\absorbl \oplus \idc)) \odot \identlp}
-\end{array}\]
-\caption{\label{figc}Signatures of level-2 $\Pi$-combinators: zero}
-\end{figure}
-
-\begin{figure}[t]
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {((\assoclp \otimes \idc) \odot \dist) \odot (\dist \oplus \idc) \Leftrightarrow (\dist \odot (\idc \oplus \dist)) \odot \assoclp}
-\\
-  {\assoclt \odot \distl \Leftrightarrow ((\idc \otimes \distl) \odot \distl) \odot (\assoclt \oplus \assoclt)}
-\end{array}\]
-\vspace{ -0.5em}
-\[\def\arraystretch{1.3}
-\begin{array}{rcl}
-  (\distl \odot (\dist \oplus \dist)) \odot \assoclp &\Leftrightarrow&
-   \dist \odot (\distl \oplus \distl) \odot \assoclp ~\odot \\
-&& (\assocrp \oplus \idc) ~\odot \\
-&& ((\idc \oplus \swapp) \oplus \idc) ~\odot \\
-&&      (\assoclp \oplus \idc)
-\end{array}\]
-\caption{\label{figb}Signatures of level-2 $\Pi$-combinators: associativity and distributivity}
-\end{figure}
-
-\begin{figure}[t]
-\[\def\arraystretch{1.3}
-\begin{array}{rcl}
-  (\idc \otimes \swapp) \odot \distl &\Leftrightarrow& \distl \odot \swapp
-\\
-  \dist \odot (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \odot \distl
-\end{array}\]
-\caption{\label{figa}Signatures of level-2 $\Pi$-combinators: commutativity and distributivity}
-\end{figure}
-
-%%%
 \section{Data III: Reversible Programs between Reversible Programs}
@@
 
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
+\\
+  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
+\\
+  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
+\\
+  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
+\\
+  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
+\\
+  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
+\\
+  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
+\end{array}\]
+\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
+\\
+  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
+\\
+  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
+\\
+  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
+\end{array}\]
+\caption{\label{figi}Signatures of level-2 $\Pi$-combinators: distributivity and factoring}
+\end{figure}
+
+\begin{figure}[t]
+Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\idc \odot \, c \Leftrightarrow c}
+\quad
+  {c \, \odot \idc \, \Leftrightarrow c}
+\quad
+  {c\,\, \odot\,! c \Leftrightarrow \idc}
+\quad
+  {! c \odot c \Leftrightarrow \idc}
+\\
+  {c \Leftrightarrow c}
+\quad
+\Rule{}
+  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
+  {c_1 \Leftrightarrow c_3}
+  {}
+\quad
+\Rule{}
+  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
+  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
+  {}
+\end{array}\]
+\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
+\end{figure}
+
+\begin{figure}[t]
+Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
+\qquad
+  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
+\\
+  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
+\qquad
+  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
+\\
+  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
+\qquad
+  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
+\\
+  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
+\qquad
+  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
+\\
+  {\identlt \Leftrightarrow \distl \odot (\identlt \oplus \identlt)}
+\\
+\identlp \Leftrightarrow \swapp \odot \identrsp
+\qquad
+\identlt \Leftrightarrow \swapt \odot \identrst
+\end{array}\]
+\caption{\label{figg}Signatures of level-2 $\Pi$-combinators: unit}
+\end{figure}
+
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\swapp \odot (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \odot \swapp}
+\quad
+  {\swapt \odot (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \odot \swapt}
+\\
+  {(\assocrp \odot \swapp) \odot \assocrp \Leftrightarrow ((\swapp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \swapp)}
+\\
+  {(\assoclp \odot \swapp) \odot \assoclp \Leftrightarrow ((\idc \oplus \swapp) \odot \assoclp) \odot (\swapp \oplus \idc)}
+\\
+  {(\assocrt \odot \swapt) \odot \assocrt \Leftrightarrow ((\swapt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \swapt)}
+\\
+  {(\assoclt \odot \swapt) \odot \assoclt \Leftrightarrow ((\idc \otimes \swapt) \odot \assoclt) \odot (\swapt \otimes \idc)}
+\end{array}\]
+\caption{\label{figf}Signatures of level-2 $\Pi$-combinators: commutativity and associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
+Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
+  {}
+\qquad
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
+  {}
+\\
+  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
+\qquad
+  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
+\\
+  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
+\\
+  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
+\end{array}\]
+\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \odot (\idc \oplus \, \identlp)}
+\\
+  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \odot (\idc \otimes \, \identlt)}
+\end{array}\]
+\caption{\label{figd}Signatures of level-2 $\Pi$-combinators: unit and associativity}
+\end{figure}
+
+\begin{figure}[t]
+Let $c : t_1 \leftrightarrow t_2$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {(c \otimes \idc) \odot \absorbl \Leftrightarrow \absorbl \odot \idc}
+\quad
+  {(\idc \, \otimes c) \odot \absorbr \Leftrightarrow \absorbr \odot \idc}
+\\
+  {\idc \odot \, \factorzl \Leftrightarrow \factorzl \odot (\idc \otimes c)}
+\quad
+  {\idc \odot \, \factorzr \Leftrightarrow \factorzr \odot (c \otimes \idc)}
+\\
+  {\absorbr \Leftrightarrow \absorbl}
+\\
+  {\absorbr \Leftrightarrow (\distl \odot (\absorbr \oplus \absorbr)) \odot \identlp}
+\\
+  {\identlst \Leftrightarrow \absorbr}
+\qquad
+  {\absorbl \Leftrightarrow \swapt \odot \absorbr}
+\\
+  {\absorbr \Leftrightarrow (\assoclt \odot (\absorbr \otimes \idc)) \odot \absorbr}
+\\
+  {(\idc \otimes \absorbr) \odot \absorbl \Leftrightarrow (\assoclt \odot (\absorbl \otimes \idc)) \odot \absorbr}
+\\
+  {\idc \otimes \, \identlp \Leftrightarrow (\distl \odot (\absorbl \oplus \idc)) \odot \identlp}
+\end{array}\]
+\caption{\label{figc}Signatures of level-2 $\Pi$-combinators: zero}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((\assoclp \otimes \idc) \odot \dist) \odot (\dist \oplus \idc) \Leftrightarrow (\dist \odot (\idc \oplus \dist)) \odot \assoclp}
+\\
+  {\assoclt \odot \distl \Leftrightarrow ((\idc \otimes \distl) \odot \distl) \odot (\assoclt \oplus \assoclt)}
+\end{array}\]
+\vspace{ -0.5em}
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\distl \odot (\dist \oplus \dist)) \odot \assoclp &\Leftrightarrow&
+   \dist \odot (\distl \oplus \distl) \odot \assoclp ~\odot \\
+&& (\assocrp \oplus \idc) ~\odot \\
+&& ((\idc \oplus \swapp) \oplus \idc) ~\odot \\
+&&      (\assoclp \oplus \idc)
+\end{array}\]
+\caption{\label{figb}Signatures of level-2 $\Pi$-combinators: associativity and distributivity}
+\end{figure}
+
+\begin{figure}[t]
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\idc \otimes \swapp) \odot \distl &\Leftrightarrow& \distl \odot \swapp
+\\
+  \dist \odot (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \odot \distl
+\end{array}\]
+\caption{\label{figa}Signatures of level-2 $\Pi$-combinators: commutativity and distributivity}
+\end{figure}
+
 %%%%%%%%%
@@
 
-Pi level 2.
-
-So we end up looking, semantically, at weak Rig Groupoid. Syntactically,
-we take the easiest way there: simply make every coherence isomorphism into
-a combinator. Huge proof obligation that this induces actual transformations
-on Pi programs. Luckily, it does.  But not straightforwardly -- ``incorrect''
-definitions of some Pi combinators (such as those for 0*x = 0) can lead to
-some level 2 morphisms not working.
-
-The interpretation on permutations is unclear, because the Pi representation
-has so much more structure, and permutations too have many different
-representations. So it is unclear what maps to what.
+As motivated in the previous section, the equivalences between type
+equivalences are perfectly modeled by the coherence conditions of weak
+Rig Groupoids. Syntactically, we take the easiest way there: simply
+make every coherence isomorphism into a programming construct. These
+constructs are collected in several figures (Fig.~\ref{figj} to
+Fig.~\ref{figa}) and are discussed next.
+
+The constructs in Fig.~\ref{figj} capture the informal idea that all
+the different ways of associating programs are equivalent. For
+example, a program that performs $c_1$ and then $(c_2 \odot c_3)$ can
+be smoothly transformed a program that performs $(c_1 \odot c_2)$ and
+then $c_3$. The last two equivalences show how several uses of
+associativity interact together. For example, in the last equivalence
+the program on the left $\assocrt \odot \assocrt$ transforms an input
+type $((t_1 * t_2) * t_3) * t_4$ as follows: 
+\[ 
+((t_1 * t_2) * t_3) * t_4 \leadsto
+(t_1 * t_2) * (t_3 * t_4) \leadsto
+t_1 * (t_2 * (t_3 * t_4))
+\]
+The program on the right
+$((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes
+\assocrt)$ achieves the same end-to-end transformation using the
+following sequence of steps:
+\[
+((t_1 * t_2) * t_3) * t_4 \leadsto
+(t_1 * (t_2 * t_3)) * t_4 \leadsto
+t_1 * ((t_2 * t_3) * t_4) \leadsto
+t_1 * (t_2 * (t_3 * t_4))
+\]
+The equivalence states that these two sequences of steps can be
+smoothly deformed to one another.
+
+The constructs in Fig.~\ref{figi} \ldots
+
+The constructs in Fig.~\ref{figh} \ldots
+
+The constructs in Fig.~\ref{figg} \ldots
+
+The constructs in Fig.~\ref{figf} \ldots
+
+The constructs in Fig.~\ref{fige} \ldots
+
+The constructs in Fig.~\ref{figd} \ldots
+
+The constructs in Fig.~\ref{figc} \ldots
+
+The constructs in Fig.~\ref{figb} \ldots
+
+The constructs in Fig.~\ref{figa} \ldots
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
+\section{Further Thoughts and Conclusions}
 

### 2018-06-12 survey/survey.tex
attempt at picture

@@
 type $((t_1 * t_2) * t_3) * t_4$ as follows:
-\[ 
-((t_1 * t_2) * t_3) * t_4 \leadsto
-(t_1 * t_2) * (t_3 * t_4) \leadsto
-t_1 * (t_2 * (t_3 * t_4))
-\]
+
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_2$}; 
+\node[above] (C) at (-1,0) {$t_3$}; 
+\node[above] (D) at (-0.5,0) {$t_4$}; 
+\draw[thick] (A) -- (-2,-1.5); 
+\draw[thick] (B) -- (-1.5,-1.5); 
+\draw[thick] (C) -- (-1,-1.5);; 
+\draw[thick] (D) -- (-0.5,-1.5); 
+\draw[densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\end{tikzpicture}
+$\leadsto$ 
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_2$}; 
+\node[above] (C) at (-1,0) {$t_3$}; 
+\node[above] (D) at (-0.5,0) {$t_4$}; 
+\draw[thick] (A) -- (-2,-1.5); 
+\draw[thick] (B) -- (-1.5,-1.5); 
+\draw[thick] (C) -- (-1,-1.5);; 
+\draw[thick] (D) -- (-0.5,-1.5); 
+\draw[densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\end{tikzpicture}
+$\leadsto$
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_2$}; 
+\node[above] (C) at (-1,0) {$t_3$}; 
+\node[above] (D) at (-0.5,0) {$t_4$}; 
+\draw[thick] (A) -- (-2,-1.5); 
+\draw[thick] (B) -- (-1.5,-1.5); 
+\draw[thick] (C) -- (-1,-1.5);; 
+\draw[thick] (D) -- (-0.5,-1.5); 
+\draw[densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\draw[densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\end{tikzpicture}
+\end{center}
+% \[ 
+% ((t_1 * t_2) * t_3) * t_4 \leadsto
+% (t_1 * t_2) * (t_3 * t_4) \leadsto
+% t_1 * (t_2 * (t_3 * t_4))
+% \]
+
 The program on the right
@@
 following sequence of steps:
-\[
-((t_1 * t_2) * t_3) * t_4 \leadsto
-(t_1 * (t_2 * t_3)) * t_4 \leadsto
-t_1 * ((t_2 * t_3) * t_4) \leadsto
-t_1 * (t_2 * (t_3 * t_4))
-\]
+
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_2$}; 
+\node[above] (C) at (-1,0) {$t_3$}; 
+\node[above] (D) at (-0.5,0) {$t_4$}; 
+\draw[thick] (A) -- (-2,-1.5); 
+\draw[thick] (B) -- (-1.5,-1.5); 
+\draw[thick] (C) -- (-1,-1.5);; 
+\draw[thick] (D) -- (-0.5,-1.5); 
+\draw[densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\end{tikzpicture}
+$\leadsto$ 
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_2$}; 
+\node[above] (C) at (-1,0) {$t_3$}; 
+\node[above] (D) at (-0.5,0) {$t_4$}; 
+\draw[thick] (A) -- (-2,-1.5); 
+\draw[thick] (B) -- (-1.5,-1.5); 
+\draw[thick] (C) -- (-1,-1.5);; 
+\draw[thick] (D) -- (-0.5,-1.5); 
+\draw[densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[densely dashed] (-1.5,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\end{tikzpicture}
+$\leadsto$
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_2$}; 
+\node[above] (C) at (-1,0) {$t_3$}; 
+\node[above] (D) at (-0.5,0) {$t_4$}; 
+\draw[thick] (A) -- (-2,-1.5); 
+\draw[thick] (B) -- (-1.5,-1.5); 
+\draw[thick] (C) -- (-1,-1.5);; 
+\draw[thick] (D) -- (-0.5,-1.5); 
+\draw[densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[densely dashed] (-1.0,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\end{tikzpicture}
+$\leadsto$
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_2$}; 
+\node[above] (C) at (-1,0) {$t_3$}; 
+\node[above] (D) at (-0.5,0) {$t_4$}; 
+\draw[thick] (A) -- (-2,-1.5); 
+\draw[thick] (B) -- (-1.5,-1.5); 
+\draw[thick] (C) -- (-1,-1.5);; 
+\draw[thick] (D) -- (-0.5,-1.5); 
+\draw[densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\draw[densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\end{tikzpicture}
+\end{center}
+% \[
+% ((t_1 * t_2) * t_3) * t_4 \leadsto
+% (t_1 * (t_2 * t_3)) * t_4 \leadsto
+% t_1 * ((t_2 * t_3) * t_4) \leadsto
+% t_1 * (t_2 * (t_3 * t_4))
+% \]
+
 The equivalence states that these two sequences of steps can be

### 2018-06-12 survey/survey.tex
added color to pic on p.13

@@
 \draw[thick] (D) -- (-0.5,-1.5); 
-\draw[densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[green,densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
 \end{tikzpicture}
@@
 \draw[thick] (D) -- (-0.5,-1.5); 
-\draw[densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[green,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
 \end{tikzpicture}
@@
 \draw[thick] (D) -- (-0.5,-1.5); 
-\draw[densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\draw[densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[red,densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\draw[green,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
 \end{tikzpicture}
@@
 \draw[thick] (D) -- (-0.5,-1.5); 
-\draw[densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[densely dashed] (-1.5,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\draw[red,densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[green,densely dashed] (-1.5,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
 \end{tikzpicture}
@@
 \draw[thick] (D) -- (-0.5,-1.5); 
-\draw[densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[densely dashed] (-1.0,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\draw[red,densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[green,densely dashed] (-1.0,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
 \end{tikzpicture}
@@
 \draw[thick] (D) -- (-0.5,-1.5); 
-\draw[densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\draw[densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
+\draw[green,densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
+\draw[red,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
 \end{tikzpicture}

### 2018-06-12 survey/survey.tex
.

@@
 the program on the left $\assocrt \odot \assocrt$ transforms an input
-type $((t_1 * t_2) * t_3) * t_4$ as follows:
+type $((t_1 * t_2) * t_3) * t_4$ to an output type
+$t_1 * (t_2 * (t_3 * t_4))$ as follows:
 
@@
 
-The program on the right
+\noindent The program on the right
 $((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes
@@
 
-The equivalence states that these two sequences of steps can be
+\noindent The equivalence states that these two sequences of steps can be
 smoothly deformed to one another.
 
-The constructs in Fig.~\ref{figi} \ldots
+The constructs in Fig.~\ref{figi} explain how choice interacts with
+grouping. Intuitively, a choice nested inside a larger bundle can also
+be treated as an outer choice with the nested bundle duplicated in
+each branch.
 

### 2018-06-12 survey/survey.tex
poor attempt at another picture

@@
 each branch. 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\node[above] (A) at (-2,0) {$t_1$}; 
+\node[above] (B) at (-1.5,0) {$t_3$}; 
+\node[above] (C) at (-1,0) {$t_5$}; 
+\draw[thick,yellow] (-1.5,-0.25) arc[start angle=60, end angle=120, radius=0.5cm];
+\draw[thick] (A) -- (-2,-0.5); 
+\draw[thick] (B) -- (-1.5,-0.5); 
+\draw[thick] (C) -- (-1,-0.5);
+\node at (-2,-0.75) {$a$}; 
+\node at (-1.5,-0.75) {$b$}; 
+\node at (-1,-0.75) {$c$}; 
+\draw[thick] (-2,-1) -- (-2,-1.5); 
+\draw[thick] (-1.5,-1) -- (-1.5,-1.5); 
+\draw[thick] (-1,-1) -- (-1,-1.5);
+\node at (-1.5,-2) {$\dist$};
+\draw[thick] (-2.5,-2.5) -- (-2.5,-3); 
+\draw[thick] (-2,-2.5) -- (-2,-3); 
+\draw[green,densely dashed] (-2.25,-2.75) ellipse [x radius=0.4cm, y radius = 0.1cm]; 
+\draw[thick] (-1,-2.5) -- (-1,-3); 
+\draw[thick] (-0.5,-2.5) -- (-0.5,-3); 
+\draw[green,densely dashed] (-0.75,-2.75) ellipse [x radius=0.4cm, y radius = 0.1cm]; 
+\draw[thick,yellow] (-1,-2.75) arc[start angle=60, end angle=120, radius=1cm];
+\end{tikzpicture}
+\end{center}
+% ((a + b) * c) ; dist <=> dist ; ((a * c) + (b * c))
+% LHS: (t1 + t3) * t5 ~> (t2 + t4) * t6 ~> (t2 * t6) + (t4 * t6)
+% RHS: (t1 + t3) * t5 ~> (t1 * t5) + (t3 * t5) ~> (t2 * t6) + (t4 * t6)
 
 The constructs in Fig.~\ref{figh} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 
 The constructs in Fig.~\ref{figg} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 
 The constructs in Fig.~\ref{figf} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 
 The constructs in Fig.~\ref{fige} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 
 The constructs in Fig.~\ref{figd} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 
 The constructs in Fig.~\ref{figc} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 
 The constructs in Fig.~\ref{figb} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 
 The constructs in Fig.~\ref{figa} \ldots 
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\end{tikzpicture}
+\end{center}
 

### 2018-06-13 survey/survey.tex
images and agda.sty

@@
 \end{verbatim}
+
+%%%
+\subsection{Spacetime Model}
+
+To gain an intuition for resources, we think of the types as possible
+events, and of the values as consistent timelines of events. From this
+perspective, $A \times B$ denotes spatial composition, where both $A$
+and $B$ must happen in \emph{some} order, and $A + B$ denotes temporal
+composition, where exactly one of $A$ and $B$ must happen at a given
+time. The event $1$ is then trivial, and $0$ is impossible.  We
+visualize the denotation of $1$ as a point in space, and $0$ as empty
+space. The denotation of $A + B$ is the denotation of $A$ drawn on top
+of the denotation of $B$ and separated by a border, and the denotation
+of $A \times B$ is just the denotation of $A$ written next to the
+denotation of $B$. Now we can formally define a consistent timeline
+recursively on the structure of a diagram:
+
+\begin{itemize}[noitemsep,topsep=0pt]
+\item An endpoint (the denotation of $1$) has exactly one timeline through it.
+Empty space (the denotation of $0$) has no timelines through it.
+\vspace{0.1pt}
+\item A timeline through a stack of alternatives $A_1 + A_2 + ... + A_n$ is a
+timeline through exactly one of the possible alternatives.
+\vspace{0.1pt}
+\item A timeline through a spatial composition of cells $A_1 \times A_2 \times ... \times A_n$
+is a timeline through \emph{all} of the cells, in any order.
+\end{itemize}
+
+For example, the following would be a consistent timeline through the diagram
+for $((2 + 2) \times (2 + 2)) + ((2 + 2) \times (2 + 2))$ times itself as shown
+in the left two boxes. The right two boxes show the trade-off between space and
+time encoded in distributivity. Since a timeline must pass through a column in
+exactly one point, the two boxes on the right have the same timelines:
+\begin{center}
+\includegraphics[scale=0.3]{images/tall-tl-1}
+\quad
+\includegraphics[scale=0.3]{images/tall-tl-2}
+\qquad\qquad\qquad
+\includegraphics[scale=0.3]{images/distrib-1}
+\qquad
+\includegraphics[scale=0.3]{images/distrib-2}
+\end{center}
 

### 2018-06-13 survey/survey.tex
fixed typo
.

--- 2018-06-12 survey/survey.tex
@@
 \end{verbatim}
+
+%%%
+\subsection{Spacetime Model}
+
+To gain an intuition for resources, we think of the types as possible
+events, and of the values as consistent timelines of events. From this
+perspective, $A * B$ denotes spatial composition, where both $A$ and
+$B$ must happen in \emph{some} order, and $A + B$ denotes temporal
+composition, where exactly one of $A$ and $B$ must happen at a given
+time. The event $1$ is then trivial, and $0$ is impossible.  We
+visualize the denotation of $1$ as a point in space, and $0$ as empty
+space. The denotation of $A + B$ is the denotation of $A$ drawn on top
+of the denotation of $B$ and separated by a border, and the denotation
+of $A * B$ is just the denotation of $A$ written next to the
+denotation of $B$. Now we can formally define a consistent timeline
+recursively on the structure of a diagram:
+
+\begin{itemize}
+\item An endpoint (the denotation of $1$) has exactly one timeline
+  through it.  Empty space (the denotation of $0$) has no timelines
+  through it.
+\item A timeline through a stack of alternatives
+  $A_1 + A_2 + \cdots + A_n$ is a timeline through exactly one of the
+  possible alternatives.
+\item A timeline through a spatial composition of cells
+  $A_1 * A_2 * \cdots * A_n$ is a timeline through \emph{all} of the
+  cells, in some order.
+\end{itemize}
+
+For example, let $2$ be the type $1 + 1$. The following would be a
+consistent timeline through the diagram for
+$((2 + 2) * (2 + 2)) + ((2 + 2) * (2 + 2))$ times itself as shown in
+the left two boxes. The right two boxes show the trade-off between
+space and time encoded in distributivity. Since a timeline must pass
+through a column in exactly one point, the two boxes on the right have
+the same timelines:
+\begin{center}
+\includegraphics[scale=0.3]{images/tall-tl-1}
+\quad
+\includegraphics[scale=0.3]{images/tall-tl-2}
+\qquad\qquad\qquad
+\includegraphics[scale=0.3]{images/distrib-1}
+\qquad
+\includegraphics[scale=0.3]{images/distrib-2}
+\end{center}
 

### 2018-06-14 survey/survey.tex
bits in conclusion

@@
 
-CS abstractions are based on ``old'' physics
-
-computer applications are more and more ``physical''
-
-physical principles such as conservation of information should be part
-of our foundational abstractions
+The entire edifice of computer science including its mainstream models
+of computations, programming languages, and logics is founded on
+\emph{classical physics.} While much of the world phenomena can be
+approximated with classical physics, we are reaching a revolutionary
+period of quantum technology that challenges many of the classical
+assumptions. It remains to be seen how computer science will adapt to
+this quantum revolution but we believe that additional physical
+principles inspired by quantum mechanics will have to be embraced in
+our computational thinking. This paper focused on one such principle
+--- \emph{conservation of information} --- and explored some of its
+exciting implications to the field of computer science.
 

### 2018-06-14 survey/survey.tex
minor edits.

@@
 
-\roshan{Let's not substitute space for set without first motivating
-  the use of the word.}  Instead of spaces consisting solely of
+Instead of spaces (aka discrete sets) consisting solely of
 unstructured isolated points, we now investigate structured spaces
@@
 items). Our next instinct would be to similarly replace $=$ with a
-type $A \equiv B$ that asserts that $A$ and $B$ are
-\emph{propositionally equal} such that a term exhibiting this type can
-then be shown in the formal system. This is however not true: the
+type $A \equiv B$ that asserts that $A$ and $B$ are \emph{propositionally
+equal}, i.e. reduce to equivalent type-denoting expressions under the rules of
+the host type system.  This is however not true: the
 proposition $A \times B \equiv B \times A$ is not
@@
 something more subtle happens. Consider $\top \uplus \top$
-versus $\top$; the first can have \emph{two} proofs of inhabitation
+versus $\top$; the first has exactly \emph{two} proofs of inhabitation
 ($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one

### 2018-06-14 survey/survey.tex
comment at end of section 1, minor edits to section 2.

@@
 
+\jc{I think it would be fair to have a \textit{quick} survey
+of other works. The most important aspect would be to draw out that
+they all start from known models of computation and then attempt to
+make them reversible.  Our originality is to start from the semantics
+of reversibility (even further back, preservation of information)
+and derive a programming language purposefully designed to denote
+exactly that. Having said that, I am not in a good position to
+write such a survey, being the least knowledgeable of all 3 of us
+of this general area.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 the sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$
-are different.  Why?  Well, supposed there was a transformation
+are different.  Why?  Well, suppose there was a transformation
 $f : A \rightarrow B$ that deformed $A$ into $B$, and another
@@
 permutes $01$ with $11$, i.e, the permutation
-$\mathit{swap}~0~2 \fatsemi \mathit{swap}~1~3$. To further illustrate
-how inconvenient this point is, consider the reversible full adder
+$\mathit{swap}~0~2 \fatsemi \mathit{swap}~1~3$. To illustrate
+how inconvenient this is, consider the reversible full adder
 below designed by Desoete et al.~\cite{117414}:
@@
 For such reasons, in programming practice we are interested in
-structured data and compositional abstractions which will be the
-subject of the next section. What we do learn however from this short
-investigation using untyped flat sets is what the ``\emph{purely
+structured data and compositional abstractions, which will be the
+subject of the next section. What we do learn from this short investigation
+using untyped and unstructured sets is what the ``\emph{purely
   operational}'' view of the theory would be. In particular, it tells
@@
 all the structure has been forgotten), these are a rather pale shadow
-of the rich tapestry of information-perserving transformations of
-structured data which we investigate next.
+of the rich tapestry of information-preserving transformations of
+structured data, which we investigate next.
 

### 2018-06-15 survey/survey.tex
start editing section 3. Define propositional equivalence.

@@
 product structure as malleable. For example, imagining a product
-structure as arranged in a grid, by stretching we can turn this
-structure to a sum structure arranged in a line, change the
+structure as arranged in a grid; by ``stretching'' we can turn it
+in to a sum structure arranged in a line. We can also change the
 orientation of the grid by exchanging the axes, as well as do other
-transformations to be discussed in this section that preserve the
-number of points.
+transformations --- as long as we preserve the number of points.
+Of course, it is not a priori clear that this necessary requirement
+is also sufficient.  Making this intuition precise will be the topic of this
+section.
 
@@
 
-We now turn into the mathematical formalization of this idea. Our goal
-is a denotational semantics on types that maintains the number of
-points. First we note that the type structure of types has a nice
-correspondence (Curry-Howard) to logic:
+We now want a proper mathematical description of this idea. Our goal
+is a denotational semantics on types which makes types that have the
+same number of points be equivalent types.  First we note that the
+structure of types has a nice correspondence (Curry-Howard) to logic:
 
@@
 A \emph{commutative semiring} (sometimes called a \emph{commutative
-  rig} (commutative ri\emph{n}g without negative elements))
+  rig} --- commutative ri\emph{n}g without negative elements)
 $(R,0,1,+,\cdot)$ consists of a set $R$, two distinguished elements of
-$R$ named 0 and 1, and two binary operations~$+$ and $\cdot$,
+$R$ named $0$ and $1$, and two binary operations~$+$ and $\cdot$,
 satisfying the following relations for any $a,b,c \in R$:
@@
 witness that $A \times B$ can be reversibly deformed into $B \times
-A$. Which motivates the introduction of type \emph{equivalences}. To
+A$, and vice-versa. Which motivates the introduction of type \emph{equivalences}. To
 do this, we need a few important auxiliary concepts.
 
-\roshan{Let's define $\equiv$ below or provide a readable
-  reference. We should not assume that its meaning is automatically
-  understood.}
+\begin{defn}[Propositional Equivalence]\label{def:propeq}
+Two expressions $a, b$ of type $A$ are \emph{propositionally
+equal} if their normal forms are equivalent under the rules
+of the host type system.
+\end{defn}
+
+In Martin-L\"{o}f Type Theory, normal forms mean $\beta$-reduced
+$\eta$ long terms under $\alpha$-equivalence. In other words,
+expressions are evaluated as much as possible ($\beta$-reduced),
+all functions are fully applied ($\eta$ long), and the exact names
+of bound variables are irrelevant ($\alpha$-equivalence). Note that
+the above definition applies equally well to expressions that denote
+values and expressions that denote types.
+
 \begin{defn}[Homotopy]

### 2018-06-15 survey/survey.tex
finish editing 3.1. Deal with Roshan's comments, and minor improvements.

@@
 \label{def:homotopy}
-Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
-$\forall x:A. f(x) \equiv g(x)$. We denote this $f \sim g$.
+Two functions $f,g:A \rightarrow B$ are \emph{homotopic}
+if~$\forall x:A. f(x) \equiv g(x)$. We denote this $f \sim g$.
 \end{defn}
@@
 
-We are now in position to formally state the appropriate commutative
+We are now in position to describe the commutative
 semiring structure for types. After replacing the set $R$ with a
@@
 the ambient logic and instead put it in the signature of the algebraic
-structure of interest. \roshan{The last coupe of sentences are
-  unclear. Let's strive to keep things easy to understand. Can we
-  spell them out the tradeoffs here and give an example or two to
-  clarify the point. If this is indeed an important aspect of things
-  let's spend some space explaining it.} Then, and only then, are we
-in a solid position to show that we have a concept (commutative
-semiring) which is flexible enough for types to be an instance.
+structure of interest.  In simpler terms, we shift equality from
+having a priviledged status in our meta-theory, to being just another
+symbol (denoting an equivalence relation) in our theory.  The understanding
+that equality is not an absolute concept has recently been an area of
+active research in mechanized mathematics --- although the concepts of
+intensional versus extensional equality go back to Frege and Russell.
 
@@
 Maxwell's daemon: even though the data is the same, they are tagged
-differently, and this data is indeed information, and must be
-preserved.
+differently, and these tags are indeed information, and their
+information content must be preserved.
 
@@
 know that the inhabitants of types formed with with
-$\bot, \top, \uplus, \times$ are a commutative semiring. What we want
+$\bot, \top, \uplus, \times$ form a commutative semiring. What we want
 to know is, which types are equivalent? From a commutative semiring
-perspective, this amounts to asking what terms are equal?  We have a
+perspective, this amounts to asking what terms are equal.  We have a
 set of generators for those equations, namely those in

### 2018-06-15 survey/survey.tex
add some more in 3.1 about the natural numbers as a semiring and as our denotation for types. Start in 3.2, making it stand out that we take proof terms as language terms.

@@
 \begin{prop}
-The structure $\left(\mathbb{B}, \false, \true, \lor, \land\right)$
+The structure $\left(\Nat, \false, \true, \lor, \land\right)$
 is a commutative semiring.
@@
 
+From category theory, we are informed of the following priviledge
+enjoyed by the natural numbers~$\Nat$:
+\begin{thm}
+The semiring $\left(\Nat, 0, 1, +, \cdot\right)$ is \emph{initial}
+in the category of semirings and semiring homomorphisms.
+\end{thm}
+\noindent In other words, for any semiring $S$, there is a homomorphism
+from $\Nat$ into $S$. But $\Nat$ is also the ``counting'' semiring,
+which formalizes the notion of cardinality of finite discrete sets.
+
+The previous section on finite sets, along with the reasoning above,
+thus leads us to posit that the correct denotational semantics for
+finite discrete types is that of the semiring
+$\left(\Nat, 0, 1, +, \cdot\right)$. It is worth noting that equality
+in this semiring is intensional (i.e. two things are equal if and only if
+they are identical after evaluation), unlike that for types.
+
 %%%%%%%%%
@@
 
-We now have in our hands our desired denotational semantics. We want
-to create a programming language, which we call $\Pi$, such that the
+We now have in our hands our desired denotational semantics for types.
+We want to create a programming language, which we call $\Pi$, such that the
 types and type combinators map to $\bot, \top, \uplus, \times$, and
 such that we have ground terms whose denotation are all $16$ type
-isomorphisms of Fig.~\ref{type-isos}. This is completely
+isomorphisms of Fig.~\ref{type-isos}. This is rather
 straightforward, as we can simply do this literally. To make the
@@
 but rather than being witnesses of inhabitation, they are witnesses of
-equivalences. Thus the proof language of commutative semiring equivalences
-becomes terms of $\Pi$.
+equivalences. Thus we get an unexpected programming language design:
+
+\center{\fbox{
+The proof language of commutative semiring equivalences
+becomes terms of the $\Pi$.}}
+\vspace*{3mm}
 
@@
 
+\noindent
 Of course, one does not get a programming language with just typed

### 2018-06-15 survey/survey.tex
finished with my edits to section 3

@@
 
-\[\begin{array}{ll}
+\begin{equation}\label{eq:adder}
+\begin{array}{l@{\hspace{1cm}}l}
 \mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 ~\fatsemi & \mbox{toffoli} \\
@@
     \mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{cnot}
-\end{array}\]
+\end{array}
+\end{equation}
 
@@
 
-\center{\fbox{
+\begin{center}
+\fbox{
 The proof language of commutative semiring equivalences
-becomes terms of the $\Pi$.}}
+becomes terms of the $\Pi$.}
+\end{center}
 \vspace*{3mm}
@@
 language of permutations lacked the ability to combine permutations by
-taking sums and products which led to the awkward non-compositional
-programming style illustrated in the full adder example. In this
-language the full adder can be programmed as follows:
-
-The type-erase of this language \emph{should} be permutations. But
-given types $A$ and $B$ that erase to $\fin{m}$ and $\fin{n}$ respectively,
-what are $A + B$ and $A * B$ ?  They correspond exactly to
+taking sums and products, which led to the awkward non-compositional
+programming style illustrated in the full adder example~\ref{eq:adder}.
+
+Thus the denotation of the $\Pi$ terms \emph{should} be permutations. But
+given types $A$ and $B$ denoting $\fin{m}$ and $\fin{n}$ respectively,
+what are $A \uplus B$ and $A \times B$ ?  They correspond exactly to
 $\fin{m+n}$ and $\fin{m*n}$!  Geometrically, this corresponds to
@@
 
-Embedded in our definition of $\Pi$ is a design decision: to make the
+Embedded in our definition of $\Pi$ is a conscious design decision: to make the
 terms of $\Pi$ \emph{syntactically} reversible. In other words, to
@@
 
-sound and complete isomorphisms for finite types
+\amr{sound and complete isomorphisms for finite types}
 
@@
 \subsection{Operational Semantics}
+
+\jc{One operational semantics is explicit, such as the eval and 
+evalB we've written many times (for example in PiEqui.agda in
+this repo). But in my opinion, the right way to do it is the
+\texttt{c2equiv} combinator in the same file, which interprets
+every combinator as a type equivalence, which can then just be
+run. \texttt{lemma0} and \texttt{lemma1} in the same file show
+that this is indeed the case. The advantage is that this lifts
+to level 2 quite cleanly.}
 
@@
 \subsection{Spacetime Model}
+\jc{While I really like this model, if we are going to keep it, we
+really need a much beefier introduction. This is the first mention of
+resources (and events and time). So we really should preface this by
+saying that, once we have $\Pi$, there is another \emph{completely
+different} but compatible interpretation in terms of resources. But
+we should also think of whether this fits. I actually think that it 
+does fit in this paper, but perhaps not in this section. Maybe
+also move to section 5?}
 

### 2018-06-15 survey/survey.tex
fix intro of section 4

@@
 \noindent Why \emph{quasi}? The reasons are beyond our scope, but the
-interested reader can read Sec.~$2.4$ and Ch.~$4$ in the HoTT
-book~\cite{hottbook}.  There are several conceptually different, but
+interested reader can read Sec.~$2.4$ and Ch.~$4$ in the
+Homotopy Type Theory (HoTT) book~\cite{hottbook}.
+There are several conceptually different, but
 equivalent, ``better'' definitions.  We record just one here:
@@
 In the previous sections, we examined equivalences between
-conventional data structures, i.e, sets of values and structured trees
-of values. We now consider a richer, less conventional, but
+conventional data structures, i.e., sets of values and structured trees
+of values. We now consider a richer but
 foundational notion of data: programs themselves. Indeed, universal
-computation models fundamentally rely on the fact that \emph{programs
+computation models crucially rely on the fact that \emph{programs
 are (or can be encoded as) data}, e.g., a Turing machine can be
 encoded as a string that another Turing machine (or even the same
-machine) can manipulate. In our setting, the programs developed in the
+machine) can manipulate. Similarly, first-class functions are
+the \emph{only} values in the $\lambda$-calculus.
+In our setting, the programs developed in the
 previous section are reversible deformations between structured finite
@@
 
+\jc{we need to make sure that the pictures we draw in the
+previous section are consistent with this one, otherwise we'll
+completely confuse our readers. Or we redraw this one to be
+consistent.}
 \begin{center}
@@
 
-  \draw     (2.5,-3)  -- (3.5,-3) -- (3.5,-4) -- (2.5,-4) -- cycle ;
-  \draw     (2.5,-4)  -- (3.5,-4) -- (3.5,-5) -- (2.5,-5) -- cycle ;
-
-  \draw     (1,1)  -- (2,1) -- (2,0) -- (1,0) -- cycle ;
-  \draw     (1,0)  -- (2,0) -- (2,-1) -- (1,-1) -- cycle ;
+  \draw     (2.5,-3.1)  -- (3.5,-3.1) -- (3.5,-3.9) -- (2.5,-3.9) -- cycle ;
+  \draw     (2.5,-4.1)  -- (3.5,-4.1) -- (3.5,-4.9) -- (2.5,-4.9) -- cycle ;
+
+  \draw     (1,0.9)  -- (2,0.9) -- (2,0.1) -- (1,0.1) -- cycle ;
+  \draw     (1,-0.1)  -- (2,-0.1) -- (2,-0.9) -- (1,-0.9) -- cycle ;
 
@@
 sum type. The bottom path performs the twist first and then deforms
-the type $A$ by $c_1$ and the type $B$ by $c_2$ as before. If one
-could imagine the paths are physical wires and the deformations $c_1$
-and $c_2$ as arbitrary deformations on these wires then, holding the
-points $A$, $B$, $C$, and $D$ fixed, it is possible to rotate the top
-part of the diagram to become identical to the bottom one. That
-rotation can be undone (reversed), which takes the bottom
+the type $A$ by $c_1$ and the type $B$ by $c_2$ as before. One
+could imagine the paths are physical \emph{elastic} wires in $3$ space, where
+the deformations $c_1$ and $c_2$ as arbitrary deformations on these wires, and
+the twists do not touch but are in fact well-separated. Then, holding the
+points $A$, $B$, $C$, and $D$ fixed, it is possible to imagine
+sliding $c_1$ and $c_2$ from the top wire rightward past the
+twist, and then using the elasticity of the wires, pull the
+twist back to line up with that of the bottom --- thus making
+both parts of the diagram identical.  Each of these moves
+can be undone (reversed), and doing so would take the bottom
 part of the diagram into the top part.  In other

### 2018-06-15 survey/survey.tex
A further paragraph on the wires intuition. Expad on where Groupoid fits in. Citations for Rig Categories. Small edits.

@@
 \newcommand{\false}{\mathit{false}}
+\newcommand{\Gpd}{\ensuremath{\mathsf{Groupoid}}}
 
@@
 
+While we will not make the details of the stretchable wires and
+slidable boxes formal, it is useful for intuition.  One caveat
+though: some of the sliding and stretching needs to be done in
+spaces of higher dimension than 3 to have ``enough room'' to
+move things along without collision or over-stretching wires.
+That, unfortunately, means that some equivalences are harder to
+grasp. Luckily, most equivalences only need 3 dimensions.
+
 Our reversible language of type isomorphisms and equivalences between
@@
 
-In the previous section, we saw how we could take the proof
-terms of commutative semirings as our starting point. What we need
-now is to understand how proofs of algebraic identities should be
+Previously we saw how we could take the proof terms of commutative semirings as
+our starting point for $\Pi$. What we need
+now is to understand how \emph{proofs} of algebraic identities should be
 considered equivalent. Classical algebra does not help, as proofs
-in algebra are not considered first-class citizens. However,
+are not considered first-class citizens. However,
 another route is available to us: since the work of 
 Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
-one can model types as \emph{groupoids}.  Previously, we
-modeled types as (finite) sets.
+one can model types as \emph{groupoids}.  The additional
+structure comes from explicitly modeling the ``identity
+types'': instead of regarding all terms which witness
+the equality of (say) $a$ and $b$ of type $A$ as being
+indistinguishible, we posit that there may in fact by many.
+This alone is enough to show that types can be modeled by
+groupoids.
 
 Thus, rather than looking at (untyped) commutative semirings,
-we should looked at a \emph{typed} version. This process frequently
+we should look at a \emph{typed} version. This process frequently
 goes by the moniker of ``categorification.''  We want a categorical
@@
 we hit what seems like a serious stumbling block: the category of
-all groupoids, Groupoid, does not have co-products or products.
-We don't want to work internally in Groupoid -- we want operations
+all groupoids, \Gpd, does not have either co-products 
+nor products. However, we don't want to work internally in
+\Gpd -- we want operations
 \emph{on} groupoids. In other words, we want something akin to
@@
 monoidal structures.  Luckily, this already exists: the categorical
-analog to (commutative) semirings are (symmetric) Rig Categories.
+analog to (commutative) semirings are (symmetric) Rig
+Categories~\cite{laplaza72,kelly74}.
 This straightforwardly generalizes to symmetric Rig Groupoids.
@@
 How does this help? Coherence conditions! Symmetric monoidal categories,
-to start somewhere simple, do not just introduce weak transformations
+to start somewhere simple, do not just introduce transformations
 like the associator $\alpha$ and the left $\lambda$ and right $\rho$
@@
 only the terms so expressible), the sub-language would correspond,
-denotationally, to exactly symmetric monoidal groupoids. And what
+denotationally, to exactly (non-empty) symmetric monoidal groupoids. And what
 these possess are exactly some \emph{equations between equations}
@@
 corresponding to naturality of the various transformers gives us a
-list of equations between programs.  Furthermore, the natural transformations
+list of equations between programs.  Furthermore, all the natural transformations
 that arise are in fact natural isomorphisms -- all of which are

### 2018-06-15 survey/survey.tex
cleaning up conclusion structure

@@
 
-Possible extensions; quest for other principles; other data models; graphs; HoTT
+We include with a collection of open problems and avenues for further research.
 
@@
 
-Natural numbers: fold/unfold
+Inductive types: Natural numbers: fold/unfold
 
 Partial isos
+
+Other infinite type: Mention connections to negative, fractional, and
+algebraic numbers
 
@@
 \subsection{Topological Spaces}
+
+other data models; graphs; HoTT
 
@@
 %%%
-\subsection{Conservation of Information, Information Effects, and Duality}
-
-Talk about conservation of information in biology, computer science,
-evolutionary computing, and search. (See
-\url{http://www.evolutionnews.org/2012/08/conservation_of063671.html}.
+\subsection{Information Effects}
 
@@
 new paper \url{https://arxiv.org/pdf/1805.08605.pdf}
-
-Mention connections to negative, fractional, and algebraic numbers
 
@@
 \bibliographystyle{unsrt}
-%\pagestyle{headings}
 \bibliography{cites}
@@
 \begin{comment}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Theseus: Types and Simple Isomorphisms}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Parametrized Maps}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Iteration Labels}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}

### 2018-06-15 survey/survey.tex
minor typo.
Try to describe, informally, why these particular coherences must hold.

@@
 \fbox{
-The proof language of commutative semiring equivalences
-becomes terms of the $\Pi$.}
+The proof terms denoting 
+commutative semiring equivalences
+induce the terms of $\Pi$.}
 \end{center}
@@
 them has a strong connection to \emph{univalent universes} in
-HoTT~\cite{CARETTE20185}. Based on this connection, we refer to the
+HoTT~\cite{Carette2018}. Based on this connection, we refer to the
 types as being at level-0, to the equivalences between types (i.e., the
@@
 
-Previously we saw how we could take the proof terms of commutative semirings as
-our starting point for $\Pi$. What we need
+Previously we saw how we could take the proof terms of commutative semirings
+equivalences as our starting point for $\Pi$. What we need
 now is to understand how \emph{proofs} of algebraic identities should be
@@
 the equality of (say) $a$ and $b$ of type $A$ as being
-indistinguishible, we posit that there may in fact by many.
-This alone is enough to show that types can be modeled by
-groupoids.
+indistinguishible, we posit that there may in fact be many.
+This consequences of this one decision are enough to show that
+types can be modeled by groupoids.
 
@@
 
-How does this help? Coherence conditions! Symmetric monoidal categories,
-to start somewhere simple, do not just introduce transformations
-like the associator $\alpha$ and the left $\lambda$ and right $\rho$
-unitors, but also coherence conditions that these must satisfy. Looking,
-for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
-$1$ and $+$ for the types, $\odot$ and $\oplus$ as combinators, and
-only the terms so expressible), the sub-language would correspond,
-denotationally, to exactly (non-empty) symmetric monoidal groupoids. And what
-these possess are exactly some \emph{equations between equations}
-as commutative diagrams.  Adding in the coherence conditions 
-corresponding to naturality of the various transformers gives us a
-list of equations between programs.  Furthermore, all the natural transformations
-that arise are in fact natural isomorphisms -- all of which are
-reversible.
-
-We can then proceed to prove that every one of the coherence conditions
-involved in defining a symmetric Rig Groupoid holds for the groupoid
-interpretation of types.
-
 \begin{figure}[t]
@@
 
-\begin{figure}[t]
+How does this help? Coherence conditions! Symmetric monoidal categories,
+to start somewhere simple, do not just introduce natural transformations
+like the associator $\alpha$ and the left and right unitors ($\lambda$ 
+and $\rho$ respectively), but also coherence conditions that these must satisfy.
+Looking, for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
+$1$ and $+$ for the types, $\odot$ and $\oplus$ as combinators, and
+only the terms so expressible), the sub-language would correspond,
+denotationally, to exactly (non-empty) symmetric monoidal groupoids. And what
+these possess are exactly some \emph{equations between equations}
+as commutative diagrams.  Transporting these coherence conditions, for
+example those that express that various transformations are \emph{natural}
+to $\Pi$, gives a list of equations between $\Pi$ programs.
+Furthermore, all the natural transformations
+that arise are in fact natural \emph{isomorphisms} -- and thus
+reversible.
+
+We can then proceed to prove that every one of the coherence conditions
+involved in defining a symmetric Rig Groupoid holds for the groupoid
+interpretation of types~\cite{Carette2016}.  This is somewhat tedious
+given the sheer number of these, but when properly formulated,
+relatively straightfoward, but see below for comments on some
+tricky cases.
+
+\begin{figure}[tb]
 Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
@@
 
-\begin{figure}[t]
+\begin{figure}[tb]
 Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
@@
 
-\begin{figure}[t]
+But why are these particular coherence laws? Are they all necessary?
+Conversely are they, in some appropriate sense, sufficient? This is
+the so-called \emph{coherence problem}. Mac Lane, in his farewell address
+as President of the American Mathematical Society~\cite{MacLane1976} gives
+a good introduction and overview of such problems.  A more modern
+interpretation (which can nevertheless be read into MacLane's own
+exposition) would read as follows: given a set of equalities on abstract
+words, regarded as a rewrite system, and two means of rewriting a word
+in that language to another, is there some suitable notion of canonical
+form that expresses the essential uniqueness of the non-trivial 
+rewrites?  Note how this word-and-rewrite problem is essentially
+independent of the eventual interpretation. One must take some care,
+as there are obvious degenerate cases (involving ``trivial'' equations
+involving $0$ or $1$) which lead to non-uniqueness. The landmark
+results, first by Kelly-MacLane~\cite{KellyMacLane71} for closed
+symmetric monoidal categories, then (independently) Laplaza and
+Kelly~\cite{laplaza72,kelly74} for symmetric Rig Categories, is
+that indeed there are sound and complete coherence conditions that
+insure that all the ``obvious'' equalities between different abstract
+words in these systems give rise to commutative diagrams. The
+``obvious'' equalities come from \emph{syzygies} or
+\emph{critical pairs} of the system of equations.
+
+\begin{figure}[tb]
 Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
@@
 
-\begin{figure}[t]
+\begin{figure}[tb]
 Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
@@
 
-\begin{figure}[t]
+\begin{figure}[tb]
 Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
@@
 
-\begin{figure}[t]
+\begin{figure}[tb]
 \[\def\arraystretch{1.3}
@@
 
-\begin{figure}[t]
+\begin{figure}[tb]
 Let $c : t_1 \leftrightarrow t_2$:
@@
 
-We include with a collection of open problems and avenues for further research.
+We conclude with a collection of open problems and avenues for further research.
 

### 2018-06-16 survey/survey.tex
added mac lane citations

@@
 a good introduction and overview of such problems.  A more modern
-interpretation (which can nevertheless be read into MacLane's own
+interpretation (which can nevertheless be read into Mac Lane's own
 exposition) would read as follows: given a set of equalities on abstract
@@
 involving $0$ or $1$) which lead to non-uniqueness. The landmark
-results, first by Kelly-MacLane~\cite{KellyMacLane71} for closed
+results, first by Kelly-Mac Lane~\cite{KELLY197197} for closed
 symmetric monoidal categories, then (independently) Laplaza and

### 2018-06-16 survey/survey.tex
citations and small subsection

@@
 
+By insisting on a computational model that explicitly enforces  
+conservation of information \ldots
+
 If information has a physical significance, then it is subject to
@@
 
-We develop an arrow meta-language to encapsulate information
-effects. This metalanguage serves to encapsulate effects, in much the
-same way that traditional arrows or monads serve to encapsulate
-effects over the \lcal.
-
-new paper \url{https://arxiv.org/pdf/1805.08605.pdf}
+A computational model that enforces the principle of conservation of
+information is arguably \emph{richer} than a conventional model that
+cannot even express the notion of information. Practically the
+conventional model is easily recovered by simply adding constructs
+that intentionally and explicitly create or erase information. Such
+constructs allow one to recover the classical perspective with the
+added advantage that it is possible to reason about such creation and
+erasure of information using type and effect systems, monads, or
+arrows~\cite{infeffects,inversearrows}.
 

### 2018-06-17 survey/survey.tex
one more subsection and spell check

@@
 structure of interest.  In simpler terms, we shift equality from
-having a priviledged status in our meta-theory, to being just another
+having a privileged status in our meta-theory, to being just another
 symbol (denoting an equivalence relation) in our theory.  The understanding
@@
 
-From category theory, we are informed of the following priviledge
+From category theory, we are informed of the following privilege
 enjoyed by the natural numbers~$\Nat$:
@@
 the equality of (say) $a$ and $b$ of type $A$ as being
-indistinguishible, we posit that there may in fact be many.
+indistinguishable, we posit that there may in fact be many.
 This consequences of this one decision are enough to show that
@@
 given the sheer number of these, but when properly formulated,
-relatively straightfoward, but see below for comments on some
+relatively straightforward, but see below for comments on some
 tricky cases.
@@
 %%%
-\subsection{Information Flow}
-
-By insisting on a computational model that explicitly enforces  
-conservation of information \ldots
-
-If information has a physical significance, then it is subject to
-conservation laws just as mass and energy are.  To make this idea
-concrete and to provide a taste of the applications it opens, consider
-a tiny 2-bit password = \verb|"10"|. The password checker looks like:
-
-\begin{verbatim}
-check-password (guess) =
-  if guess == "10"
-  then True
-  else False
-\end{verbatim}
-
-One can ask how much information is leaked by this program assuming
-the attacker has no prior knowledge except that the password is 2
-bits, i.e., the four possible 2-bits are equally likely. If the
-attacker guesses \verb|"10"| (with probability $1/4$) the password (2
-bits) is leaked. If the attacker guesses one of the other choices
-(with probability $3/4$) the number of possibilities is reduced from 4
-to 3, i.e., the attacker learns $\log{4} - \log{3}$ bits of
-information. So in general the attacker learns\footnote{If the
-  password is 8 restricted ASCII characters (6 bits), the attacker
-  learns 0.00001 bits in the first probe.}:
-\[\begin{array}{ll}
-   &  1/4 * 2 + 3/4 (\log{4} - \log{3}) \\
-  =&  1/4 \log{4} + 3/4 \log{4/3} \\
-  =&  - 1/4 \log{1/4} - 3/4 \log{3/4} \\
-  \sim& 0.8 \mbox{~bits~in~the~first~probe}
-\end{array}\]
-
-One can alternatively look at the situation by viewing the input as a
-random variable with 4 possibilities and a uniform distribution (i.e.,
-with 2 bits of information). The output is another random variable
-with 4 possibilities but with the distribution
-$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
-information. Thus 2 input bits of information were given to the
-function and only 0.8 were produced. Where did the 1.2 bits of
-information go? Once we accept the thesis that information is a
-physical entity this question cannot be ignored.
-
-The Landauer Principle states that erasing information requires
-energy. In other words, the password checking function must have
-erased 1.2 bits during its calculation, which dissipate as heat in any
-physical realization of the function. In conventional models of
-computation, this erasure is \emph{implicit}, i.e., it occurs as a
-side effect of calculation. We wish to look at models where
-erasure is simply not allowed.
-
-There are, in fact, many different principles of physics which are at
-odds with our current models of computation. Even within quantum
-mechanics itself, there are other ideas (such as superposition or
-entanglement) which are not present in current models. We will however
-restrict ourselves to exploring a single dimension in which our
-foundations should be revised: \textbf{conservation of information}.
-We will follow its consequences, which will turn out to be far
-reaching.
-
-\begin{quote}
-  The laws of physics are essentially algorithms for calculation. These
-  algorithms are significant only to the extent that they are executable in
-  our real physical world. Our usual laws of physics depend on the
-  mathematician's real number system. With that comes the presumption that
-  given any accuracy requirement, there exist a number of calculations steps,
-  which if executed, will satisfy that accuracy requirement. But the real
-  world is unlikely to supply us with unlimited memory of unlimited Turing
-  machine tapes, Therefore, continuum mathematics is not executable, and
-  physical laws which invoke that can not really be satisfactory. They are
-  references to illusionary procedures. Rolf Landauer, The physical nature of
-  information, Physics Letters A 217 (1996) pp. 188-193:
-\end{quote}
-
-A high level approach to conservation of information would proceed
-as follows:
-\begin{enumerate}
-\item Create a means to measure information.
-\item Show that this measure is sound with respect to our current understanding
-of what information ought to be.
-\item Assert that whatever ``quantity'' of information a system has, must
-remain invariant throughout the lifetime of the system.
-\end{enumerate}
-
-Certainly, whatever our notion of information, it can neither be created,
-duplicated, nor erased. Of course, if we make our notion of information
-too rigid, it will be impossible to even \textbf{modify} it! For computation
-to occur, modification must be possible.  But of what kind?
-
-%%%
 \subsection{Information Effects}
@@
 arrows~\cite{infeffects,inversearrows}.
+
+An interesting application of such an idea is in the field of
+\emph{information-flow security}. To make this idea concrete, consider
+a tiny 2-bit password = \verb|"10"| and the associated password checker:
+
+\begin{verbatim}
+check-password (guess) =
+  if guess == "10"
+  then True
+  else False
+\end{verbatim}
+
+One can ask how much information is leaked by this program assuming
+the attacker has no prior knowledge except that the password is 2
+bits, i.e., the four possible 2-bits are equally likely. If the
+attacker guesses \verb|"10"| (with probability $1/4$) the password (2
+bits) is leaked. If the attacker guesses one of the other choices
+(with probability $3/4$) the number of possibilities is reduced from 4
+to 3, i.e., the attacker learns $\log{4} - \log{3}$ bits of
+information. So in general the attacker learns:
+\[\begin{array}{ll}
+   &  1/4 * 2 + 3/4 (\log{4} - \log{3}) \\
+  =&  1/4 \log{4} + 3/4 \log{4/3} \\
+  =&  - 1/4 \log{1/4} - 3/4 \log{3/4} \\
+  \sim& 0.8 \mbox{~bits~in~the~first~probe}
+\end{array}\]
+This is a significant amount of information. But of course this is
+only because the password is so short: if the password was 8
+restricted ASCII characters (6 bits), the attacker would only learn
+0.00001 bits in the first probe.
+
+An alternative formulation of the problem is to view the input as a
+random variable with 4 possibilities and a uniform distribution (i.e.,
+with 2 bits of information) and the output as another random variable
+with 4 possibilities but with the distribution
+$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
+information. Thus 2 input bits of information were given to the
+password checker and only 0.8 were produced. Where did the 1.2 bits of
+information go? By the Landauer Principle, these 1.2 bits must be
+accounted by an \emph{implicit} \emph{erasure} operation in the
+program. By writing the password checker in an extension of $\Pi$, the
+erasure construct becomes explicit and the information leak becomes
+exposed in the syntactic structure of the program~\cite{infeffects}.
 

### 2018-06-18 survey/survey.tex
subsection 5.2 and citations

@@
 %%%
-\subsection{Infinite Sets}
-
-Must mention infinite sets since we promised to do so in Sec 2;
-Discuss recursion, trace, feedback, loop...
-
-Inductive types: Natural numbers: fold/unfold
-
-Partial isos
-
-Other infinite type: Mention connections to negative, fractional, and
-algebraic numbers
-
-meta-circular interpreter from RC paper: Since these new models are
-substantially different from \lcal\ or familiar high-level languages,
-a point of concern is how one can effectively develop programs in
-them.  We address this by presenting a straightforward technique for
-deriving $\Pi$ programs by systematically translating logically
-reversible small-step abstract machines~\cite{isoint}. Complex
-programs can be devised in this way; we demonstrate the derivation of
-a meta-circular interpreter for $\Pi$.
-
-DANGER: here, we leave the safe land of ``preservation of
-information,'' because partiality is not reversible. So while trace
-gives a reversible program as output, it itself isn't a reversible
-operation. Like the daemon's brain in Maxwell's thought experiment,
-trace irreversible work!  So trace, which is fundamental to all work
-with recursive types, is only reversible "at level 1". So we need to
-replace trace with some other operation for level 2 reversibility to
-also hold -- and this is an open problem.
+\subsection{Information Effects}
+
+A computational model that enforces the principle of conservation of
+information is arguably \emph{richer} than a conventional model that
+cannot even express the notion of information. Practically the
+conventional model is easily recovered by simply adding constructs
+that intentionally and explicitly create or erase information. Such
+constructs allow one to recover the classical perspective with the
+added advantage that it is possible to reason about such creation and
+erasure of information using type and effect systems, monads, or
+arrows~\cite{infeffects,inversearrows}.
+
+An interesting application of such an idea is in the field of
+\emph{information-flow security}. To make this idea concrete, consider
+a tiny 2-bit password = \verb|"10"| and the associated password checker:
+
+\begin{verbatim}
+check-password (guess) =
+  guess == "10"
+\end{verbatim}
+
+One can ask how much information is leaked by this program assuming
+the attacker has no prior knowledge except that the password is 2
+bits, i.e., the four possible 2-bits are equally likely. If the
+attacker guesses \verb|"10"| (with probability $1/4$) the password (2
+bits) is leaked. If the attacker guesses one of the other choices
+(with probability $3/4$) the number of possibilities is reduced from 4
+to 3, i.e., the attacker learns $\log{4} - \log{3}$ bits of
+information. So in general the attacker learns:
+\[\begin{array}{ll}
+   &  1/4 * 2 + 3/4 (\log{4} - \log{3}) \\
+  =&  1/4 \log{4} + 3/4 \log{4/3} \\
+  =&  - 1/4 \log{1/4} - 3/4 \log{3/4} \\
+  \sim& 0.8 \mbox{~bits~in~the~first~probe}
+\end{array}\]
+This is a significant amount of information. But of course this is
+only because the password is so short: if the password was 8
+restricted ASCII characters (6 bits), the attacker would only learn
+0.00001 bits in the first probe.
+
+An alternative formulation of the problem is to view the input as a
+random variable with 4 possibilities and a uniform distribution (i.e.,
+with 2 bits of information) and the output as another random variable
+with 4 possibilities but with the distribution
+$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
+information. Thus 2 input bits of information were given to the
+password checker and only 0.8 were produced. Where did the 1.2 bits of
+information go? By the Landauer Principle, these 1.2 bits must be
+accounted by an \emph{implicit} \emph{erasure} operation in the
+program. By writing the password checker in an extension of $\Pi$, the
+erasure construct becomes explicit and the information leak becomes
+exposed in the syntactic structure of the program~\cite{infeffects}.
 
 %%%
-\subsection{Topological Spaces}
-
-other data models; graphs; HoTT
-
-Data IV: Circle etc. Would be interesting
-
-%%%
-\subsection{Theseus}
-
-The most elementary control structure in a programming language is the
-ability to conditionally execute one of several possible code
-fragments. Expressing such an abstraction using predicates and nested
+\subsection{Theseus and Quantum Control}
+
+The $\Pi$ family of languages semantically captures the principles of
+reversibility and conservation of information. As a programming
+language it has some mixed properties: small programs are relatively
+easy to write; for some special classes of programs, it is even
+possible to define a methodology to write large $\Pi$ programs,
+including a self-interpreter for $\Pi$~\cite{isoint}. In general,
+however, the point-free style of combinators used in $\Pi$ becomes
+awkward and a new approach appears more suitable. To that end, we note
+that $\Pi$ encodes the most elementary control structure in a
+programming language-- which is the ability to conditionally execute
+one of several possible code fragments-- using combinators. Expressing
+such an abstraction using combinators or even predicates and nested
 \textbf{if}-expressions makes it difficult for both humans and
-compilers to reason about the control flow structure. Instead, in
-modern functional languages, this control flow paradigm is elegantly
-expressed using \emph{pattern-matching}. This approach yields code
-that is not only more concise and readable but also enables the
-compiler to easily verify two crucial properties: (i) non-overlapping
-patterns and (ii) exhaustive coverage of a datatype using a collection
-of patterns. Indeed most compilers for functional languages perform
-these checks, warning the user when they are violated. At a more
-fundamental level, e.g., in type theories and proof assistants, these
-properties are actually necessary for correct reasoning about
-programs. Our first insight, explained in this section, is that these
-properties, perhaps surprisingly, are sufficient to produce a simple
-and intuitive first-order reversible programming language.
+compilers to write, understand, and reason about the control flow
+structure of the program. Instead, in modern functional languages,
+this control flow paradigm is elegantly expressed using
+\emph{pattern-matching}. This approach yields code that is not only
+more concise and readable but also enables the compiler to easily
+verify two crucial properties: (i) non-overlapping patterns and (ii)
+exhaustive coverage of a datatype using a collection of
+patterns. Indeed most compilers for functional languages perform these
+checks, warning the user when they are violated. At a more fundamental
+level, e.g., in type theories and proof assistants, these properties
+are actually necessary for correct reasoning about programs. Our
+insight is that these properties, perhaps surprisingly, are sufficient
+to produce a simple and intuitive first-order reversible programming
+language which we call \emph{Theseus}.
 
@@
 
-We start with a small illustrative example, written in a Haskell-like
+We provide a small illustrative example, written in a Haskell-like
 syntax.  Fig.~\ref{fig:intro-ex-1} gives the skeleton of a function
-\verb|f| that accepts a value of type \verb|Either Int| \verb|Int|; the
-patterns on the left-hand side exhaustively cover every possible
+\verb|f| that accepts a value of type \verb|Either Int| \verb|Int|;
+the patterns on the left-hand side exhaustively cover every possible
 incoming value and are non-overlapping. Similarly,
@@
 second and third right-hand sides of \verb|h| but not the first and
-second ones.
+second ones. With some additional work, it is possible to extend
+Theseus to a full-fledged reversible programming
+language~\cite{theseus}. With just one additional insight, Theseus can
+be extended with superpositions and becomes a quantum programming
+language~\cite{10.1007/978-3-319-89366-2_19}.
 
 %%%
-\subsection{Quantum Computing}
-
-Extension of Theseus in 2018 FoSSaCS paper.
-
-Exclusive Disjunctions
-
-Quantum over finite field
-
-\jc{The reason I commented this out is that it is under-justified. The
-reader will simply not understand what these next few lines are really
-saying.}
-
-Indeed one should take the physical principles underlying quantum
-mechanics, the most successful physical theory known to us and adapt
-computation to ``learn'' from these principles. To illustrate the
-depth of our crisis, Scott Aaronson, Umesh Vazirani, and others have
-proposed the following puzzle.
-
-One of these wild claims must be true!:
-\begin{itemize}
-\item the extended Church-Turing thesis is false, or
-\item quantum physics is false, or
-\item there is an efficient classical algorithm for factoring
-\end{itemize}
-Indeed, if quantum physics is correct then there is an efficient quantum
-algorithm for factoring (Shor). If there is no efficient classical algorithm
-for factoring then the extended Church-Turing thesis is false.
-
-%%%
-\subsection{Information Effects}
-
-A computational model that enforces the principle of conservation of
-information is arguably \emph{richer} than a conventional model that
-cannot even express the notion of information. Practically the
-conventional model is easily recovered by simply adding constructs
-that intentionally and explicitly create or erase information. Such
-constructs allow one to recover the classical perspective with the
-added advantage that it is possible to reason about such creation and
-erasure of information using type and effect systems, monads, or
-arrows~\cite{infeffects,inversearrows}.
-
-An interesting application of such an idea is in the field of
-\emph{information-flow security}. To make this idea concrete, consider
-a tiny 2-bit password = \verb|"10"| and the associated password checker:
-
-\begin{verbatim}
-check-password (guess) =
-  if guess == "10"
-  then True
-  else False
-\end{verbatim}
-
-One can ask how much information is leaked by this program assuming
-the attacker has no prior knowledge except that the password is 2
-bits, i.e., the four possible 2-bits are equally likely. If the
-attacker guesses \verb|"10"| (with probability $1/4$) the password (2
-bits) is leaked. If the attacker guesses one of the other choices
-(with probability $3/4$) the number of possibilities is reduced from 4
-to 3, i.e., the attacker learns $\log{4} - \log{3}$ bits of
-information. So in general the attacker learns:
-\[\begin{array}{ll}
-   &  1/4 * 2 + 3/4 (\log{4} - \log{3}) \\
-  =&  1/4 \log{4} + 3/4 \log{4/3} \\
-  =&  - 1/4 \log{1/4} - 3/4 \log{3/4} \\
-  \sim& 0.8 \mbox{~bits~in~the~first~probe}
-\end{array}\]
-This is a significant amount of information. But of course this is
-only because the password is so short: if the password was 8
-restricted ASCII characters (6 bits), the attacker would only learn
-0.00001 bits in the first probe.
-
-An alternative formulation of the problem is to view the input as a
-random variable with 4 possibilities and a uniform distribution (i.e.,
-with 2 bits of information) and the output as another random variable
-with 4 possibilities but with the distribution
-$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
-information. Thus 2 input bits of information were given to the
-password checker and only 0.8 were produced. Where did the 1.2 bits of
-information go? By the Landauer Principle, these 1.2 bits must be
-accounted by an \emph{implicit} \emph{erasure} operation in the
-program. By writing the password checker in an extension of $\Pi$, the
-erasure construct becomes explicit and the information leak becomes
-exposed in the syntactic structure of the program~\cite{infeffects}.
+\subsection{Richer Data: Infinite Sets and Topological Spaces}
+
+Must mention infinite sets since we promised to do so in Sec 2;
+Discuss recursion, trace, feedback, loop...
+
+Inductive types: Natural numbers: fold/unfold
+
+Partial isos
+
+Other infinite type: Mention connections to negative, fractional, and
+algebraic numbers
+
+meta-circular interpreter from RC paper: Since these new models are
+substantially different from \lcal\ or familiar high-level languages,
+a point of concern is how one can effectively develop programs in
+them.  We address this by presenting a straightforward technique for
+deriving $\Pi$ programs by systematically translating logically
+reversible small-step abstract machines~\cite{isoint}. Complex
+programs can be devised in this way; we demonstrate the derivation of
+a meta-circular interpreter for $\Pi$.
+
+DANGER: here, we leave the safe land of ``preservation of
+information,'' because partiality is not reversible. So while trace
+gives a reversible program as output, it itself isn't a reversible
+operation. Like the daemon's brain in Maxwell's thought experiment,
+trace irreversible work!  So trace, which is fundamental to all work
+with recursive types, is only reversible "at level 1". So we need to
+replace trace with some other operation for level 2 reversibility to
+also hold -- and this is an open problem.
+
+other data models; graphs; HoTT
+
+Data IV: Circle etc. Would be interesting
+
+topological spaces have strange cardinalities: negative types
+fractional types etc
 

### 2018-06-18 survey/survey.tex
conclusion

@@
 %%%
+\subsection{Richer Data: Infinite Sets and Topological Spaces}
+
+The three languages we discussed only deal with the finite spaces
+built from 0, 1, sums, and products. Programming practice, logic, and
+mathematics all deal with richer spaces including inductive types
+(e.g., the natural numbers, sequences, and trees), functions, and
+graphs. Extending $\Pi$ to such domains is possible but only after one
+refines the notions of reversibility and conservation of
+information. One approach is to use \emph{partial isomorphisms} that
+may be undefined on such inputs~\cite{infeffects,rc2011}. Another more
+speculative approach is to build such spaces, topologically, based on
+novel type constructions such as negative, fractional, or even
+imaginary types~\cite{seventrees,roshan-thesis}.
+
+%%%
 \subsection{Information Effects}
@@
 with 4 possibilities but with the distribution
-$\{ (True, 1/4), (False, 3/4) \}$ which contains 0.8 bits of
+$\{ (\mathit{True}, 1/4), (\mathit{False}, 3/4) \}$ which contains 0.8 bits of
 information. Thus 2 input bits of information were given to the
@@
 possible to define a methodology to write large $\Pi$ programs,
-including a self-interpreter for $\Pi$~\cite{isoint}. In general,
-however, the point-free style of combinators used in $\Pi$ becomes
-awkward and a new approach appears more suitable. To that end, we note
-that $\Pi$ encodes the most elementary control structure in a
+including a meta-circular interpreter for $\Pi$~\cite{isoint}. In
+general, however, the point-free style of combinators used in $\Pi$
+becomes awkward and a new approach appears more suitable. To that end,
+we note that $\Pi$ encodes the most elementary control structure in a
 programming language-- which is the ability to conditionally execute
@@
 language~\cite{10.1007/978-3-319-89366-2_19}.
-
-%%%
-\subsection{Richer Data: Infinite Sets and Topological Spaces}
-
-Must mention infinite sets since we promised to do so in Sec 2;
-Discuss recursion, trace, feedback, loop...
-
-Inductive types: Natural numbers: fold/unfold
-
-Partial isos
-
-Other infinite type: Mention connections to negative, fractional, and
-algebraic numbers
-
-meta-circular interpreter from RC paper: Since these new models are
-substantially different from \lcal\ or familiar high-level languages,
-a point of concern is how one can effectively develop programs in
-them.  We address this by presenting a straightforward technique for
-deriving $\Pi$ programs by systematically translating logically
-reversible small-step abstract machines~\cite{isoint}. Complex
-programs can be devised in this way; we demonstrate the derivation of
-a meta-circular interpreter for $\Pi$.
-
-DANGER: here, we leave the safe land of ``preservation of
-information,'' because partiality is not reversible. So while trace
-gives a reversible program as output, it itself isn't a reversible
-operation. Like the daemon's brain in Maxwell's thought experiment,
-trace irreversible work!  So trace, which is fundamental to all work
-with recursive types, is only reversible "at level 1". So we need to
-replace trace with some other operation for level 2 reversibility to
-also hold -- and this is an open problem.
-
-other data models; graphs; HoTT
-
-Data IV: Circle etc. Would be interesting
-
-topological spaces have strange cardinalities: negative types
-fractional types etc
 

### 2018-06-22 survey/survey.tex
start explaining all of the level-2 Pi combinators.

@@
 
-\begin{figure}[t]
-Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
-\\
-  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
-\\
-  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
-\\
-  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
-\\
-  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
-\\
-  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
-\\
-  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
-\end{array}\]
-\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
-\end{figure}
-
 How does this help? Coherence conditions! Symmetric monoidal categories,
@@
 
-\begin{figure}[tb]
-Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
-\\
-  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
-\\
-  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
-\\
-  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
-\end{array}\]
-\caption{\label{figi}Signatures of level-2 $\Pi$-combinators: distributivity and factoring}
-\end{figure}
-
-\begin{figure}[tb]
-Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\idc \odot \, c \Leftrightarrow c}
-\quad
-  {c \, \odot \idc \, \Leftrightarrow c}
-\quad
-  {c\,\, \odot\,! c \Leftrightarrow \idc}
-\quad
-  {! c \odot c \Leftrightarrow \idc}
-\\
-  {c \Leftrightarrow c}
-\quad
-\Rule{}
-  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
-  {c_1 \Leftrightarrow c_3}
-  {}
-\quad
-\Rule{}
-  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
-  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
-  {}
-\end{array}\]
-\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
-\end{figure}
-
 But why are these particular coherence laws? Are they all necessary?
@@
 rewrites?  Note how this word-and-rewrite problem is essentially
-independent of the eventual interpretation. One must take some care,
+independent of the eventual interpretation. But one must take some care,
 as there are obvious degenerate cases (involving ``trivial'' equations
@@
 
+%%%%%%%%%
+\subsection{A Language of Equivalences between Type Equivalences}
+
 \begin{figure}[tb]
-Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
+Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
 \[\def\arraystretch{1.3}
 \begin{array}{c}
-  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
-\qquad
-  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
-\\
-  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
-\qquad
-  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
-\\
-  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
-\qquad
-  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
-\\
-  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
-\qquad
-  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
-\\
-  {\identlt \Leftrightarrow \distl \odot (\identlt \oplus \identlt)}
-\\
-\identlp \Leftrightarrow \swapp \odot \identrsp
-\qquad
-\identlt \Leftrightarrow \swapt \odot \identrst
+  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
+\\
+  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
+\\
+  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
+\\
+  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
+\\
+  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
+\\
+  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
+\\
+  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
 \end{array}\]
-\caption{\label{figg}Signatures of level-2 $\Pi$-combinators: unit}
+\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
 \end{figure}
-
-\begin{figure}[tb]
-Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\swapp \odot (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \odot \swapp}
-\quad
-  {\swapt \odot (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \odot \swapt}
-\\
-  {(\assocrp \odot \swapp) \odot \assocrp \Leftrightarrow ((\swapp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \swapp)}
-\\
-  {(\assoclp \odot \swapp) \odot \assoclp \Leftrightarrow ((\idc \oplus \swapp) \odot \assoclp) \odot (\swapp \oplus \idc)}
-\\
-  {(\assocrt \odot \swapt) \odot \assocrt \Leftrightarrow ((\swapt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \swapt)}
-\\
-  {(\assoclt \odot \swapt) \odot \assoclt \Leftrightarrow ((\idc \otimes \swapt) \odot \assoclt) \odot (\swapt \otimes \idc)}
-\end{array}\]
-\caption{\label{figf}Signatures of level-2 $\Pi$-combinators: commutativity and associativity}
-\end{figure}
-
-\begin{figure}[tb]
-Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
-Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-\Rule{}
-  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
-  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
-  {}
-\qquad
-\Rule{}
-  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
-  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
-  {}
-\\
-  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
-\qquad
-  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
-\\
-  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
-\\
-  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
-\end{array}\]
-\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
-\end{figure}
-
-\begin{figure}[tb]
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \odot (\idc \oplus \, \identlp)}
-\\
-  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \odot (\idc \otimes \, \identlt)}
-\end{array}\]
-\caption{\label{figd}Signatures of level-2 $\Pi$-combinators: unit and associativity}
-\end{figure}
-
-\begin{figure}[tb]
-Let $c : t_1 \leftrightarrow t_2$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {(c \otimes \idc) \odot \absorbl \Leftrightarrow \absorbl \odot \idc}
-\quad
-  {(\idc \, \otimes c) \odot \absorbr \Leftrightarrow \absorbr \odot \idc}
-\\
-  {\idc \odot \, \factorzl \Leftrightarrow \factorzl \odot (\idc \otimes c)}
-\quad
-  {\idc \odot \, \factorzr \Leftrightarrow \factorzr \odot (c \otimes \idc)}
-\\
-  {\absorbr \Leftrightarrow \absorbl}
-\\
-  {\absorbr \Leftrightarrow (\distl \odot (\absorbr \oplus \absorbr)) \odot \identlp}
-\\
-  {\identlst \Leftrightarrow \absorbr}
-\qquad
-  {\absorbl \Leftrightarrow \swapt \odot \absorbr}
-\\
-  {\absorbr \Leftrightarrow (\assoclt \odot (\absorbr \otimes \idc)) \odot \absorbr}
-\\
-  {(\idc \otimes \absorbr) \odot \absorbl \Leftrightarrow (\assoclt \odot (\absorbl \otimes \idc)) \odot \absorbr}
-\\
-  {\idc \otimes \, \identlp \Leftrightarrow (\distl \odot (\absorbl \oplus \idc)) \odot \identlp}
-\end{array}\]
-\caption{\label{figc}Signatures of level-2 $\Pi$-combinators: zero}
-\end{figure}
-
-\begin{figure}[tb]
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {((\assoclp \otimes \idc) \odot \dist) \odot (\dist \oplus \idc) \Leftrightarrow (\dist \odot (\idc \oplus \dist)) \odot \assoclp}
-\\
-  {\assoclt \odot \distl \Leftrightarrow ((\idc \otimes \distl) \odot \distl) \odot (\assoclt \oplus \assoclt)}
-\end{array}\]
-\vspace{ -0.5em}
-\[\def\arraystretch{1.3}
-\begin{array}{rcl}
-  (\distl \odot (\dist \oplus \dist)) \odot \assoclp &\Leftrightarrow&
-   \dist \odot (\distl \oplus \distl) \odot \assoclp ~\odot \\
-&& (\assocrp \oplus \idc) ~\odot \\
-&& ((\idc \oplus \swapp) \oplus \idc) ~\odot \\
-&&      (\assoclp \oplus \idc)
-\end{array}\]
-\caption{\label{figb}Signatures of level-2 $\Pi$-combinators: associativity and distributivity}
-\end{figure}
-
-\begin{figure}[tb]
-\[\def\arraystretch{1.3}
-\begin{array}{rcl}
-  (\idc \otimes \swapp) \odot \distl &\Leftrightarrow& \distl \odot \swapp
-\\
-  \dist \odot (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \odot \distl
-\end{array}\]
-\caption{\label{figa}Signatures of level-2 $\Pi$-combinators: commutativity and distributivity}
-\end{figure}
-
-%%%%%%%%%
-\subsection{A Language of Equivalences between Type Equivalences}
 
@@
 
+Conveniently, the various coherence conditions can be naturally
+grouped into ``related'' laws.  Each group basically captures the
+interactions between compositions of level-1 $\Pi$ combinators.
+
 The constructs in Fig.~\ref{figj} capture the informal idea that all
-the different ways of associating programs are equivalent. For
-example, a program that performs $c_1$ and then $(c_2 \odot c_3)$ can
-be smoothly transformed a program that performs $(c_1 \odot c_2)$ and
-then $c_3$. The last two equivalences show how several uses of
-associativity interact together. For example, in the last equivalence
-the program on the left $\assocrt \odot \assocrt$ transforms an input
-type $((t_1 * t_2) * t_3) * t_4$ to an output type
+the different ways of associating programs are equivalent. The first
+says that sequential composition itself ($\odot$) is associative.
+The next $4$ capture how
+the $\oplus$ and $\otimes$ combinators ``commute'' with re-association.
+In other words, it expresses the type-level associativity of $+$ is
+properly reflected by the properties of $\oplus$.
+The last two equivalences show how composition of associativity combinators
+interact together. For example, the last equivalence captures the idea that the
+program $\assocrt \odot \assocrt$ transforms an input of
+type $((t_1 * t_2) * t_3) * t_4$ to an output of type
 $t_1 * (t_2 * (t_3 * t_4))$ as follows:
@@
 \begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
 \node[above] (A) at (-2,0) {$t_1$}; 
@@
 $\leadsto$ 
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
 \node[above] (A) at (-2,0) {$t_1$}; 
@@
 $\leadsto$
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
 \node[above] (A) at (-2,0) {$t_1$}; 
@@
 \begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
 \node[above] (A) at (-2,0) {$t_1$}; 
@@
 $\leadsto$ 
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
 \node[above] (A) at (-2,0) {$t_1$}; 
@@
 $\leadsto$
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
 \node[above] (A) at (-2,0) {$t_1$}; 
@@
 
-\noindent The equivalence states that these two sequences of steps can be
-smoothly deformed to one another.
-
-The constructs in Fig.~\ref{figi} explain how choice interacts with
-grouping. Intuitively, a choice nested inside a larger bundle can also
-be treated as an outer choice with the nested bundle duplicated in
-each branch. 
+\noindent This of course gives rise to a \emph{critical pair} in words over the
+level-1 language, as both of these programs are between the same types.
+Since neither of these involve either trivial types (like $0$ or $1$) nor
+twists (i.e. automorphisms) on the types themselves, coherence requires that
+these in fact be equivalent programs.  One interpretation is that
+\emph{outermost} re-associations is the same as \emph{innermost}
+re-associations.
+
+\jc{I am not a big fan of these pictures -- mainly because I can't give any
+intuition based on them of why these can be smoothly deformed into one another.
+All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
+5 dimensions -- one for each of the wires, and another to give room to maneuver.}
+
+\begin{figure}[tb]
+Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
+\\
+  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
+\\
+  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
+\\
+  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
+\end{array}\]
+\caption{\label{figi}Signatures of level-2 $\Pi$-combinators: distributivity and factoring}
+\end{figure}
+
+The constructs in Fig.~\ref{figi} are the basic coherence for
+$\dist$, $\distl$, $\factor$ and $\factorl$: the type-level distribution
+and factoring has to commute with the combinator-level $\oplus$ and $\otimes$.
+
+\begin{comment}
 \begin{center}
@@
 % RHS: (t1 + t3) * t5 ~> (t1 * t5) + (t3 * t5) ~> (t2 * t6) + (t4 * t6)
-
-The constructs in Fig.~\ref{figh} \ldots 
+\end{comment}
+
+\begin{figure}[tb]
+Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\idc \odot \, c \Leftrightarrow c}
+\quad
+  {c \, \odot \idc \, \Leftrightarrow c}
+\quad
+  {c\,\, \odot\,! c \Leftrightarrow \idc}
+\quad
+  {! c \odot c \Leftrightarrow \idc}
+\\
+  {c \Leftrightarrow c}
+\quad
+\Rule{}
+  {c_1 \Leftrightarrow c_2 \quad c_2 \Leftrightarrow c_3}
+  {c_1 \Leftrightarrow c_3}
+  {}
+\quad
+\Rule{}
+  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
+  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
+  {}
+\end{array}\]
+\caption{\label{figh}Signatures of level-2 $\Pi$-combinators: identity and composition}
+\end{figure}
+
+The constructs in Fig.~\ref{figh} express various properties of composition.
+The first two says that $\idc$ is a left and right identity for sequential composition.
+
 \begin{center}
@@
 \end{center}
+
+\begin{figure}[tb]
+Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
+\qquad
+  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
+\\
+  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
+\qquad
+  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
+\\
+  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
+\qquad
+  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
+\\
+  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
+\qquad
+  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
+\\
+  {\identlt \Leftrightarrow \distl \odot (\identlt \oplus \identlt)}
+\\
+\identlp \Leftrightarrow \swapp \odot \identrsp
+\qquad
+\identlt \Leftrightarrow \swapt \odot \identrst
+\end{array}\]
+\caption{\label{figg}Signatures of level-2 $\Pi$-combinators: unit}
+\end{figure}
 
@@
 
+\begin{figure}[tb]
+Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\swapp \odot (c_1 \oplus c_2) \Leftrightarrow (c_2 \oplus c_1) \odot \swapp}
+\quad
+  {\swapt \odot (c_1 \otimes c_2) \Leftrightarrow (c_2 \otimes c_1) \odot \swapt}
+\\
+  {(\assocrp \odot \swapp) \odot \assocrp \Leftrightarrow ((\swapp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \swapp)}
+\\
+  {(\assoclp \odot \swapp) \odot \assoclp \Leftrightarrow ((\idc \oplus \swapp) \odot \assoclp) \odot (\swapp \oplus \idc)}
+\\
+  {(\assocrt \odot \swapt) \odot \assocrt \Leftrightarrow ((\swapt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \swapt)}
+\\
+  {(\assoclt \odot \swapt) \odot \assoclt \Leftrightarrow ((\idc \otimes \swapt) \odot \assoclt) \odot (\swapt \otimes \idc)}
+\end{array}\]
+\caption{\label{figf}Signatures of level-2 $\Pi$-combinators: commutativity and associativity}
+\end{figure}
+
 The constructs in Fig.~\ref{figf} \ldots 
@@
 
+\begin{figure}[tb]
+Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
+Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
+  {}
+\qquad
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
+  {}
+\\
+  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
+\qquad
+  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
+\\
+  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
+\\
+  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
+\end{array}\]
+\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
+\end{figure}
+
 The constructs in Fig.~\ref{fige} \ldots 
@@
 
+\begin{figure}[tb]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {\identlsp \oplus \idc ~\Leftrightarrow~ \assocrp \odot (\idc \oplus \, \identlp)}
+\\
+  {\identlst \otimes \idc ~\Leftrightarrow~ \assocrt \odot (\idc \otimes \, \identlt)}
+\end{array}\]
+\caption{\label{figd}Signatures of level-2 $\Pi$-combinators: unit and associativity}
+\end{figure}
+
 The constructs in Fig.~\ref{figd} \ldots 
@@
 
+\begin{figure}[tb]
+Let $c : t_1 \leftrightarrow t_2$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {(c \otimes \idc) \odot \absorbl \Leftrightarrow \absorbl \odot \idc}
+\quad
+  {(\idc \, \otimes c) \odot \absorbr \Leftrightarrow \absorbr \odot \idc}
+\\
+  {\idc \odot \, \factorzl \Leftrightarrow \factorzl \odot (\idc \otimes c)}
+\quad
+  {\idc \odot \, \factorzr \Leftrightarrow \factorzr \odot (c \otimes \idc)}
+\\
+  {\absorbr \Leftrightarrow \absorbl}
+\\
+  {\absorbr \Leftrightarrow (\distl \odot (\absorbr \oplus \absorbr)) \odot \identlp}
+\\
+  {\identlst \Leftrightarrow \absorbr}
+\qquad
+  {\absorbl \Leftrightarrow \swapt \odot \absorbr}
+\\
+  {\absorbr \Leftrightarrow (\assoclt \odot (\absorbr \otimes \idc)) \odot \absorbr}
+\\
+  {(\idc \otimes \absorbr) \odot \absorbl \Leftrightarrow (\assoclt \odot (\absorbl \otimes \idc)) \odot \absorbr}
+\\
+  {\idc \otimes \, \identlp \Leftrightarrow (\distl \odot (\absorbl \oplus \idc)) \odot \identlp}
+\end{array}\]
+\caption{\label{figc}Signatures of level-2 $\Pi$-combinators: zero}
+\end{figure}
+
 The constructs in Fig.~\ref{figc} \ldots 
@@
 
+\begin{figure}[tb]
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {((\assoclp \otimes \idc) \odot \dist) \odot (\dist \oplus \idc) \Leftrightarrow (\dist \odot (\idc \oplus \dist)) \odot \assoclp}
+\\
+  {\assoclt \odot \distl \Leftrightarrow ((\idc \otimes \distl) \odot \distl) \odot (\assoclt \oplus \assoclt)}
+\end{array}\]
+\vspace{ -0.5em}
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\distl \odot (\dist \oplus \dist)) \odot \assoclp &\Leftrightarrow&
+   \dist \odot (\distl \oplus \distl) \odot \assoclp ~\odot \\
+&& (\assocrp \oplus \idc) ~\odot \\
+&& ((\idc \oplus \swapp) \oplus \idc) ~\odot \\
+&&      (\assoclp \oplus \idc)
+\end{array}\]
+\caption{\label{figb}Signatures of level-2 $\Pi$-combinators: associativity and distributivity}
+\end{figure}
+
 The constructs in Fig.~\ref{figb} \ldots 
@@
 \end{center}
+
+\begin{figure}[tb]
+\[\def\arraystretch{1.3}
+\begin{array}{rcl}
+  (\idc \otimes \swapp) \odot \distl &\Leftrightarrow& \distl \odot \swapp
+\\
+  \dist \odot (\swapt \oplus \swapt) &\Leftrightarrow & \swapt \odot \distl
+\end{array}\]
+\caption{\label{figa}Signatures of level-2 $\Pi$-combinators: commutativity and distributivity}
+\end{figure}
 

### 2018-06-22 survey/survey.tex
Introduce reversing combinator, as a derived combinator, in section 3 (as it is used in section 4). Then continue describing section 4 combinators.

@@
 
+\begin{figure}[t]
+\[
+\Rule{}
+{\jdg{}{}{c_1 : t_1 \iso t_2}}
+{\jdg{}{}{\ !\ c_1 : t_2 \iso t_1}}
+{}
+\]
+\caption{Derived $\Pi$-combinator.}
+\label{derived-pi-combinators}
+\end{figure}
+
 Embedded in our definition of $\Pi$ is a conscious design decision: to make the
@@
 every $\Pi$ constant, there is another $\Pi$ constant which is its
-inverse. This is not the only choice.  Another would be to add a
+inverse. As this is used frequently, we give it the short name $!$,
+and its type is given in Fig.~\ref{derived-pi-combinator}. This
+combinator is \emph{defined}, by pattern matching on the syntax of
+its argument and structural recursion.
+
+This is not the only choice.  Another would be to add a
 $\mathit{flip}$ combinator to the language; we could then remove
@@
 \quad
-  {c\,\, \odot\,! c \Leftrightarrow \idc}
+  {c\,\, \odot\,!\, c \Leftrightarrow \idc}
 \quad
-  {! c \odot c \Leftrightarrow \idc}
+  {!\, c \odot c \Leftrightarrow \idc}
 \\
@@
 The first two says that $\idc$ is a left and right identity for sequential composition.
+The next two say that all programs are reversible, both on the left and the right:
+running $c$ and then its reverse ($!\, c$) is equivalent to the identity, and the
+same for doing $!\, c$ first then $c$. The last line say that there is an
+identity level-2 combinator, a sequential composition, and that level-2
+equivalence respects level-1 sequential composition $\odot$.
 

### 2018-06-22 survey/survey.tex
Next 2 sets of level-2 combinators. And also trailing blanks.

@@
 \caption{Derived $\Pi$-combinator.}
-\label{derived-pi-combinators}
+\label{derived-pi-combinator}
 \end{figure}
@@
 
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
-
 \begin{figure}[tb]
@@
 \\
-\identlp \Leftrightarrow \swapp \odot \identrsp
+\identlp \Leftrightarrow \swapp \odot \identlsp
 \qquad
-\identlt \Leftrightarrow \swapt \odot \identrst
+\identlt \Leftrightarrow \swapt \odot \identlst
 \end{array}\]
@@
 
-The constructs in Fig.~\ref{figg} \ldots 
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
+The constructs in Fig.~\ref{figg} may at first blush look similarly straightforward,
+but deserve some pause. One obvious question: What is the point of
+$c_0 : 0 \leftrightarrow 0$, isn't that just the identity combinator $\idc$
+for $A = 0$ (as defined in Fig.~\ref{type-isos})? Operationally, $c_0$
+is indeed indistinguishable from $\idc$. However, there are multiple syntactic
+ways of writing down combinators of type $0 \leftrightarrow 0$, and the
+first combinator in Fig.~\ref{figg} applies to all of them uniformly.
+This is another subtle aspect of coherence: all reasoning must be valid for
+all possible models, not just the one we have in mind. So even though
+operational reasoning may suggest that some relations \emph{may} be
+true between combinators, it can also mislead. The same reasoning
+applies to $c_1 : 1 \leftrightarrow 1$.  The first $8$ combinators can
+then be read as basic coherence for unit introduction and elimination,
+in both additive and multiplicative cases.
+
+The last two capture
+another simple idea, related to swapping: eliminating a unit
+on the left is the same as first swapping then eliminating on the
+right (both additively and multiplicatively). As a side note,
+these are not related to \emph{commutativity}, but rather
+come from one of the simplest coherence condition for
+braided monoidal categories. In other words, it reflects the
+idempotence of $\swapp$ and $\swapt$ rather than the
+commutativity of $\oplus$ and $\otimes$.
 
@@
 
-The constructs in Fig.~\ref{figf} \ldots 
+The first two equivalences in Fig.~\ref{figf} reflect the basic
+coherence between type-level swapping and and the level-2 combinator
+actions. The next four arise because of interactions between
+(additive and multiplicative) level-1 associativity and swapping.
+In other words, they arise as critical pairs.  For example,
+the first expresses that the two ways of going from
+$\left(A \oplus B\right) \oplus C$ to $B \oplus \left(C \oplus A\right)$
+are equivalent, with the second saying that the reverse (i.e.
+the results of applying $!$\~) also gives equivalent programs.
+The last two say the same but for the multiplicative structure.
+
 \begin{center}

### 2018-06-24 survey/survey.tex
a bit on graphical lang

@@
 \usepackage{stmaryrd}
-\usetikzlibrary{quotes}
+\usetikzlibrary{quotes,fit,positioning}
 \tikzset{>=latex}
@@
 %%%%%%%%%
-\subsection{Graphical Language}
-
-a graphical notation for programming in these models that reminiscent
-of Penrose diagrams for categories~\cite{selinger-graphical}, Geometry
-of Interaction (GoI) machines and Proof
-Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
-notation, computation is modeled by the flow of particles in a
-circuit.
-
-\amr{
-  Let’s have a geometric representation of spaces:
-- A + B means putting A next to B
-- A * B means putting A and B on two perpendicular axes and filling
-the square (volume) between them
-All the combinators would indeed have a natural action and if we have
-a few nice pictures it would be great. What do you think? }
-
-%%%%%%%%%
 \subsection{Examples}
@@
 
-%%%
-\subsection{Spacetime Model}
-\jc{While I really like this model, if we are going to keep it, we
-really need a much beefier introduction. This is the first mention of
-resources (and events and time). So we really should preface this by
-saying that, once we have $\Pi$, there is another \emph{completely
-different} but compatible interpretation in terms of resources. But
-we should also think of whether this fits. I actually think that it
-does fit in this paper, but perhaps not in this section. Maybe
-also move to section 5?}
-
-To gain an intuition for resources, we think of the types as possible
-events, and of the values as consistent timelines of events. From this
-perspective, $A * B$ denotes spatial composition, where both $A$ and
-$B$ must happen in \emph{some} order, and $A + B$ denotes temporal
-composition, where exactly one of $A$ and $B$ must happen at a given
-time. The event $1$ is then trivial, and $0$ is impossible.  We
-visualize the denotation of $1$ as a point in space, and $0$ as empty
-space. The denotation of $A + B$ is the denotation of $A$ drawn on top
-of the denotation of $B$ and separated by a border, and the denotation
-of $A * B$ is just the denotation of $A$ written next to the
-denotation of $B$. Now we can formally define a consistent timeline
-recursively on the structure of a diagram:
-
+%%%%%%%%%
+\subsection{Graphical Language}
+
+% \jc{While I really like this model, if we are going to keep it, we
+% really need a much beefier introduction. This is the first mention of
+% resources (and events and time). So we really should preface this by
+% saying that, once we have $\Pi$, there is another \emph{completely
+% different} but compatible interpretation in terms of resources. But
+% we should also think of whether this fits. I actually think that it
+% does fit in this paper, but perhaps not in this section. Maybe
+% also move to section 5?}
+
+The textual representation of $\Pi$ is quite cumbersome: the
+underlying principles however become quite intuitive when represented
+in a more suggestive visual notation. There exist various graphical
+notations for the monoidal categories that model $\Pi$ and related
+languages including Penrose diagrams~\cite{selinger-graphical} and
+Geometry of Interaction (GoI) machines and Proof
+Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In this section, we
+propose a novel graphical representation of $\Pi$ types and
+combinators drawn from the resource usage embedded in the operational
+semantics. As suggested by the operational semantics, the type $A * B$
+can be viewed as denoting \emph{spatial composition}, where
+\emph{both} $A$ and $B$ must happen in \emph{some} order or perhaps
+simultaneously, and where $A + B$ denotes \emph{temporal composition},
+where exactly \emph{one of} $A$ and $B$ must happen at a given time.
+
+This idea leads to a resource-based interpretation of types as
+\emph{possible events} and of the values as \emph{consistent timelines
+  of events} as explained below. The event $1$ is trivial, and the
+event $0$ is impossible.  We visualize the denotation of $1$ as a
+point in space, and $0$ as empty space. The denotation of $A + B$ is
+the denotation of~$A$ drawn on top of the denotation of $B$ and
+separated by a border, and the denotation of $A * B$ is just the
+denotation of $A$ written next to the denotation of $B$.  For example,
+let $2$ be the type $1 + 1$ represented by a stack of two boxes each
+with a point inside it. The type $(2 + 2) * (2 + 2)$ would then be two
+such stacks laid next to each other. Composing further, the type
+$((2 + 2) * (2 + 2)) + ((2 + 2) * (2 + 2))$ is built by stacking the
+last two diagrams as shown below on the left:
+\begin{center}
+\hfill 
+\includegraphics[scale=0.3]{images/tall-tl-1}
+~
+\includegraphics[scale=0.3]{images/tall-tl-2}
+\hfill (*)
+\end{center}
+If we call the type on the left $A$ then the full diagram above
+represents the type $A*A$.
+
+To understand how diagrams (types) can transform through the
+evaluation of  a program, we need to reason about the equivalence of
+different diagrams. Consider for example, the two diagrams below:
+
+\begin{center}
+\hfill 
+\includegraphics[scale=0.3]{images/distrib-1}
+\qquad \qquad \qquad \qquad \qquad 
+\includegraphics[scale=0.3]{images/distrib-2}
+\hfill (**)
+\end{center}
+
+The diagram on the left represents the type $A * (B + C)$ while the
+one on the right represents the type $(A * B) + (A * C)$. These two
+types are equivalent in $\Pi$ but why are they equivalent in the
+graphical notation? We argue that they admit the same consistent
+timelines where a consistent timeline is defined recursively on the
+structure of a diagram:
 \begin{itemize}
@@
 
-For example, let $2$ be the type $1 + 1$. The following would be a
-consistent timeline through the diagram for
-$((2 + 2) * (2 + 2)) + ((2 + 2) * (2 + 2))$ times itself as shown in
-the left two boxes. The right two boxes show the trade-off between
-space and time encoded in distributivity. Since a timeline must pass
-through a column in exactly one point, the two boxes on the right have
-the same timelines:
-\begin{center}
-\includegraphics[scale=0.3]{images/tall-tl-1}
-\quad
-\includegraphics[scale=0.3]{images/tall-tl-2}
-\qquad\qquad\qquad
-\includegraphics[scale=0.3]{images/distrib-1}
-\qquad
-\includegraphics[scale=0.3]{images/distrib-2}
-\end{center}
+For example, the red line in the diagram (*) represents a consistent
+timeline. For the diagrams (**), we note that any consistent timeline
+for the left diagram must go through $A$ and either one of $B$ or $C$
+and for the diagram on the right, the choice happens earlier but the
+result is the same: any consistent timeline must go through either $A$
+and $B$ or through $A$ and $C$.
+
+% \amr{
+%   Let’s have a geometric representation of spaces:
+% - A + B means putting A next to B
+% - A * B means putting A and B on two perpendicular axes and filling
+% the square (volume) between them
+% All the combinators would indeed have a natural action and if we have
+% a few nice pictures it would be great. What do you think? }
+
+% \newcommand{\graphzero}{}
+% \newcommand{\graphone}{\ensuremath{\bullet}}
+% \newcommand{\graphplus}[2]{\node (top) {#1}; \node [below of=top] {#2};}
+
+% \begin{center}
+% \begin{tikzpicture} %% [declare function = {plus(\x,\y) = \node (top){\x}; \node [below of=top] {\y};}]
+% \node (x) {\graphone};
+% \node (y) {\graphone};
+% %% {plus(\x,\y)}
+% \end{tikzpicture}
+% \end{center}
 

### 2018-06-24 survey/survey.tex
one sentence on graphical
just comments

@@
 
+%% Random idea:
+%% for a product monoidal this is wave-style
+%% for a sum monoidal this is particle-style
+%% when we have both we do have a wave-particle duality
+
 % \jc{While I really like this model, if we are going to keep it, we
@@
 and $B$ or through $A$ and $C$.
+
+Other than distributity, $\Pi$ combinators simply allow us to shuffle
+the components of a stack around as well as shuffle the positions of
+components in the plane.
 
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 Let $c_1 : t_1 \leftrightarrow t_2$ and $c_2 : t_3 \leftrightarrow t_4$:
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 \[\def\arraystretch{1.3}
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 Let $c : t_1 \leftrightarrow t_2$:
@@
 \subsection{Graphical Language}
+
+Idea: thinking in the 3D way; we have collections of towers of
+blocks. Every 1-combinator involves touching some blocks, swapping
+them around, etc. Two 1-combinators are equivalent if they touch the
+same blocks the same number of times. ??
 

### 2018-06-25 survey/survey.tex
remove comment out comment
commentary

@@
 
-% \jc{While I really like this model, if we are going to keep it, we
-% really need a much beefier introduction. This is the first mention of
-% resources (and events and time). So we really should preface this by
-% saying that, once we have $\Pi$, there is another \emph{completely
-% different} but compatible interpretation in terms of resources. But
-% we should also think of whether this fits. I actually think that it
-% does fit in this paper, but perhaps not in this section. Maybe
-% also move to section 5?}
-
 The textual representation of $\Pi$ is quite cumbersome: the
@@
 where exactly \emph{one of} $A$ and $B$ must happen at a given time.
+\jc{I think the temporal aspect is stretching things too far. Could we
+call it ``choice composition''? (Or something better). To me spatial
+composition somehow seems like sequential composition (i.e. composition
+in time), which is definitely not what we want. I think the informal
+description (both although perhaps concurrently, and exactly one of)
+is better. In section 3 we called them sequential, choice and
+parallel. We should be consistent.}
+
+\jc{An additional ``problem'' is that since our circuits have a
+left-to-right orientation, $A * B$ drawn as boxes left-to-right
+gives an implicit ordering that is going to be hard to ignore.
+I'm starting to think that the diagrams in the MO answer might
+be better: use a colour (or like in Selinger's paper, at the
+very end), symbols in the nodes that split.  So both would
+be displayed as vertical stacks (only). Because I think that
+swap* is too hard to justify in this notation. Of course,
+if we could do things in 3D and $*$ was displayed using
+(say) depth, that would be different again.}
 

### 2018-06-25 survey/survey.tex
some progress in section 4. Moved some things around to make more sense for the reader, and then wrote a bit more.

@@
 
-\begin{figure}[t]
-Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
-\\
-  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
-\\
-  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
-\\
-  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
-\\
-  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
-\\
-  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
-\\
-  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
-\end{array}\]
-\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
-\end{figure}
-
 As motivated in the previous section, the equivalences between type
@@
 interactions between compositions of level-1 $\Pi$ combinators.
+
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
+Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
+  {}
+\qquad
+\Rule{}
+  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
+  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
+  {}
+\\
+  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
+\\
+  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
+\end{array}\]
+\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
+\end{figure}
+
+Starting with the simplest constructions, the first two constructs in
+Fig.~\ref{fige} are the level-2 analogs of $+$ and $*$, which respectively
+model combinator-level choice composition and parallel composition (of
+equivalences).  These allow us to ``build up'' larger equivalences from smaller
+ones.  The next two express that both of these composition operators distribute
+over sequential composition $\odot$ (and vice versa).
+
+\begin{figure}[t]
+Let $c_1 : t_1 \leftrightarrow t_2$,  $c_2 : t_2 \leftrightarrow t_3$, and $c_3 : t_3 \leftrightarrow t_4$:
+\[\def\arraystretch{1.3}
+\begin{array}{c}
+  {c_1 \odot (c_2 \odot c_3) \Leftrightarrow (c_1 \odot c_2) \odot c_3}
+\\
+  {(c_1 \oplus (c_2 \oplus c_3)) \odot \assoclp \Leftrightarrow \assoclp \odot ((c_1 \oplus c_2) \oplus c_3)}
+\\
+  {(c_1 \otimes (c_2 \otimes c_3)) \odot \assoclt \Leftrightarrow \assoclt \odot ((c_1 \otimes c_2) \otimes c_3)}
+\\
+  {((c_1 \oplus c_2) \oplus c_3) \odot \assocrp \Leftrightarrow \assocrp \odot (c_1 \oplus (c_2 \oplus c_3))}
+\\
+  {((c_1 \otimes c_2) \otimes c_3) \odot \assocrt \Leftrightarrow \assocrt \odot (c_1 \otimes (c_2 \otimes c_3))}
+\\
+  {\assocrp \odot \assocrp \Leftrightarrow ((\assocrp \oplus \idc) \odot \assocrp) \odot (\idc \oplus \assocrp)}
+\\
+  {\assocrt \odot \assocrt \Leftrightarrow ((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes \assocrt)}
+\end{array}\]
+\caption{\label{figj}Signatures of level-2 $\Pi$-combinators: associativity}
+\end{figure}
 
@@
 \\
+  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
+\qquad
+  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
+\\
   {c \Leftrightarrow c}
@@
 
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
-
-\begin{figure}[t]
-Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, $c_3 : t_1 \leftrightarrow t_2$, and $c_4 : t_3 \leftrightarrow t_4$. \\
-Let $a_1 : t_5 \leftrightarrow t_1$,  $a_2 : t_6 \leftrightarrow t_2$, $a_3 : t_1 \leftrightarrow t_3$, and $a_4 : t_2 \leftrightarrow t_4$.
-\[\def\arraystretch{1.3}
-\begin{array}{c}
-\Rule{}
-  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
-  {c_1 \oplus c_2 \Leftrightarrow c_3 \oplus c_4}
-  {}
-\qquad
-\Rule{}
-  {c_1 \Leftrightarrow c_3 \quad c_2 \Leftrightarrow c_4}
-  {c_1 \otimes c_2 \Leftrightarrow c_3 \otimes c_4}
-  {}
-\\
-  {\idc \oplus \, \idc \, \Leftrightarrow \idc}
-\qquad
-  {\idc \otimes \, \idc \, \Leftrightarrow \idc}
-\\
-  {(a_1 \odot a_3) \oplus (a_2 \odot a_4) \Leftrightarrow (a_1 \oplus a_2) \odot (a_3 \oplus a_4)}
-\\
-  {(a_1 \odot a_3) \otimes (a_2 \odot a_4) \Leftrightarrow (a_1 \otimes a_2) \odot (a_3 \otimes a_4)}
-\end{array}\]
-\caption{\label{fige}Signatures of level-2 $\Pi$-combinators: functors}
-\end{figure}
-
-The constructs in Fig.~\ref{fige} \ldots
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
-
 \begin{figure}[t]
@@
 
-The constructs in Fig.~\ref{figd} \ldots
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
+The constructs in Fig.~\ref{figd} express how unit elimination ``in the middle''
+can be expressed either as operating on the right or, (after re-association) on the left.
 
@@
 
-The constructs in Fig.~\ref{figc} \ldots
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
+The constructs in Fig.~\ref{figc} are significantly more subtle.
+\jc{WAVEFRONT}
 

### 2018-06-25 survey/survey.tex
lots of additions to sec 3

@@
 \usepackage{stmaryrd}
+\usepackage{proof}
 \usetikzlibrary{quotes,fit,positioning}
@@
 %% Macros
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{corollary}[theorem]{Corollary}
+
+\newcommand{\inkscape}[2][1.5]{
+\begin{center}
+\scalebox{#1}{
+\includegraphics{inkscape/#2}
+}
+\end{center}
+}
 
@@
 % \end{center}
+
+\begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
+\label{def:primitives-langRev}
+
+\[\begin{array}{rrcll}
+ \mathit{zeroe} :& 0 + b &\leftrightarrow & b &: \mathit{zeroi} \\
+ \mathit{swap}^{+} :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \mathit{swap}^{+} \\
+ \mathit{assocl}^{+} :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \mathit{assocr}^{+} \\
+ \mathit{unite} :& 1 \times b &\leftrightarrow & b &: \mathit{uniti} \\
+ \mathit{swap}^{\times } :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \mathit{swap}^{\times } \\
+ \mathit{assocl}^{\times } :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \mathit{assocr}^{\times } \\
+ \mathit{distrib }_{0} :& 0 \times b &\leftrightarrow & 0 &: \mathit{factor }_{0} \\
+ \mathit{distrib } :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \mathit{factor } \\
+ \end{array}\]
+%subcode source isomorphisms.tex:427
+
+\end{definition}
+
+Each line of this table is to be read as the definition of one or two
+operators. For example, corresponding to the \textit{identity of
+  \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
+\ensuremath{\mathit{unite}:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
+\ensuremath{\mathit{uniti}:b\leftrightarrow 1\times b} (reading the isomorphism from right to
+left). These operators are inverses of each other.  Each of the two
+cases of commutativity defines one operator that is its own inverse.
+
+Having defined primitive operators, we need some means of composing
+them. We construct the composition combinators out of the closure
+conditions for isomorphisms.
+
+\begin{definition}{Composition in \ensuremath{\Pi } }
+$$
+\infer{ \mathit{id} : b \leftrightarrow b}{
+	 ~
+}
+\quad
+\infer{ \mathit{sym} ~c : b_2 \leftrightarrow b_1}{
+	 c : b_1 \leftrightarrow b_2
+}
+\quad
+\infer{ c_1\fatsemi c_2 : b_1 \leftrightarrow b_3}{
+	 c_1 : b_1 \leftrightarrow b_2
+	&
+	 c_2 : b_2 \leftrightarrow b_3
+}
+$$
+$$
+\infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
+	 c_1 : b_1 \leftrightarrow b_3
+	&
+	 c_2 : b_2 \leftrightarrow b_4
+}
+\quad
+\infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
+	 c_1 : b_1 \leftrightarrow b_3
+	&
+	 c_2 : b_2 \leftrightarrow b_4
+}
+$$
+%subcode source isomorphisms.tex:462
+\end{definition}
+
+Thus we have program constructs that witness reflexivity \ensuremath{\mathit{id}},
+symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\fatsemi} and two parallel
+composition combinators, one for sums~\ensuremath{+} and one for
+pairs~\ensuremath{\times}.
+
+\begin{definition}{(Syntax of \ensuremath{\Pi })}
+\label{def:langRev}
+We collect our types, values, and combinators, to get the full language
+definition.
+\[\begin{array}{rclr}
+ \mathit{value} ~\mathit{types}, b &::=& 0 ~|~ 1 ~|~ b+b ~|~ b\times b \\
+ \mathit{values}, v &::=& () ~|~ \mathit{left} ~v ~|~ \mathit{right} ~v ~|~ (v,v) \\
+ \\
+ \mathit{combinator} ~\mathit{types}, t &::=& b \leftrightarrow b \\
+ \mathit{isomorphisms}, \mathit{iso} &::=& \mathit{zeroe} ~|~ \mathit{zeroi} ~|~ \mathit{swap}^{+} ~|~ \mathit{assocl}^{+} ~|~ \mathit{assocr}^{+} \\
+ & ~|~ & \mathit{unite} ~|~ \mathit{uniti} ~|~ \mathit{swap}^{\times } ~|~ \mathit{assocl}^{\times } ~|~ \mathit{assocr}^{\times } \\
+ & ~|~ & \mathit{distrib }_{0} ~|~ \mathit{factor }_{0} ~|~ \mathit{distrib } ~|~ \mathit{factor } \\
+ \mathit{combinators}, c &::=& \mathit{iso} ~|~ \mathit{id} ~|~ \mathit{sym} ~c ~|~ c \fatsemi c ~|~ c + c ~|~ c \times c \\
+ \end{array}\]
+%subcode source isomorphisms.tex:483
+\end{definition}
+
+\begin{definition}{Operational Semantics for \ensuremath{\Pi }. }
+\label{def:operational-langRev}
+Given a program \ensuremath{c : b_1 \leftrightarrow b_2} in \ensuremath{\Pi }, we can run it by supplying
+it with a value \ensuremath{ v_1 : b_1}. The evaluation rules \ensuremath{c ~v_1 \mapsto v_2} are given below.
+
+Additive fragment:
+
+\[\begin{array}{rlcl}
+ \mathit{zeroe} & (\mathit{right} ~v) &\mapsto & v \\
+ \mathit{zeroi} & v &\mapsto & \mathit{right} ~v \\
+ \mathit{swap}^{+} & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
+ \mathit{swap}^{+} & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
+ \mathit{assocl}^{+} & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
+ \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
+ \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
+ \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
+ \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
+ \mathit{assocr}^{+} & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:504
+
+Multiplicative fragment:
+
+\[\begin{array}{rlcl}
+ \mathit{unite} & ((), v) &\mapsto & v \\
+ \mathit{uniti} & v &\mapsto & ((), v) \\
+ \mathit{swap}^{\times } & (v_1, v_2) &\mapsto & (v_2, v_1) \\
+ \mathit{assocl}^{\times } & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
+ \mathit{assocr}^{\times } & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:514
+
+Distributivity and factoring:
+
+\[\begin{array}{rlcl}
+ \mathit{distrib } & (\mathit{left} ~v_1, v_3) &\mapsto & \mathit{left} ~(v_1, v_3) \\
+ \mathit{distrib } & (\mathit{right} ~v_2, v_3) &\mapsto & \mathit{right} ~(v_2, v_3) \\
+ \mathit{factor } & (\mathit{left} ~(v_1, v_3)) &\mapsto & (\mathit{left} ~v_1, v_3) \\
+ \mathit{factor } & (\mathit{right} ~(v_2, v_3)) &\mapsto & (\mathit{right} ~v_2, v_3) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:523
+
+The evaluation rules of the composition combinators are given below:
+
+$$
+\infer{ \mathit{id} ~v \mapsto v}{
+	 ~
+}
+\quad
+\infer{ (\mathit{sym} ~c) ~v_1 \mapsto v_2}{
+	 c{\mathit{^{\dagger} }} v_1 \mapsto v_2
+}
+\quad
+\infer{ (c_1\fatsemi c_2) ~v_1 \mapsto v_2}{
+	 c_1 ~v_1 \mapsto v
+	&
+	 c_2 ~v \mapsto v_2
+}
+$$
+$$
+\infer{ (c_1 + c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
+	 c_1 ~v_1 \mapsto v_2
+}
+\quad
+\infer{ (c_1 + c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
+	 c_2 ~v_1 \mapsto v_2
+}
+$$
+$$
+\infer{ (c_1 \times c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
+	 c_1 ~v_1 \mapsto v_3
+	&
+	 c_2 ~v_2 \mapsto v_4
+}
+$$
+%subcode source isomorphisms.tex:546
+
+\end{definition}
+
+Since there are no values that have the type \ensuremath{0}, the reductions for
+the combinators \ensuremath{\mathit{zeroe}}, \ensuremath{\mathit{zeroi}}, \ensuremath{\mathit{distrib }_{0}} and \ensuremath{\mathit{factor }_{0}}
+omit the impossible cases. Also, the semantics of \ensuremath{\mathit{sym} ~c} are
+defined in terms of the adjoint of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
+which we formalize below.
+
+\begin{definition}[Adjoints, \ensuremath{c^\mathit{^{\dagger} }}]
+ The adjoint of a combinator \ensuremath{c} is defined by \ensuremath{c^\mathit{^{\dagger} }} as follows:
+
+  \begin{itemize}
+  \item For primitive isomophisms \ensuremath{c}, \ensuremath{c^\mathit{^{\dagger} }} is given by its
+    isomorphic inverse operator from the table in Definition
+    \ref{def:primitives-langRev}.
+
+  \item \ensuremath{(c_1 \times c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \times c_2^\mathit{^{\dagger} }}
+
+  \item \ensuremath{(c_1 + c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } + c_2^\mathit{^{\dagger} }}
+
+  \item \ensuremath{(c_1\fatsemi c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \fatsemi c_1^\mathit{^{\dagger} }}. (Note that the
+    order of combinators has been reversed).
+
+  \end{itemize}
+
+\end{definition}
+
+\begin{definition}[Observational Equivalence. \ensuremath{c_1=c_2}]
+\label{def:pi-eqv}
+  Combinators \ensuremath{c_1} and \ensuremath{c_2}, where \ensuremath{c_1 : b_1 \leftrightarrow b_2} and
+  \ensuremath{c_2:b_1\leftrightarrow b_2}, are observationally equivalent,\ensuremath{c_1=c_2}, iff
+
+  \ensuremath{\forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \Longleftrightarrow c_2 ~v_1 \mapsto v_2}.
+\end{definition}
+
+\begin{proposition}{Canonical Isomorphism.}
+  For any type \ensuremath{b} there exists the isomorphism \ensuremath{b \leftrightarrow b^{|b|}}.
+  \begin{proof}
+    The fact that such an isomorphism exists is evident from the
+    definition of size and what it means for two types to be
+    isomorphic. While many equivalent constructions are possible
+    for any type \ensuremath{b}, one such construction is given by \ensuremath{\lbrack\!\lbrack b\rbrack\!\rbrack}:
+
+\[\begin{array}{rclr}
+ \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \mathit{id} \\
+ \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \mathit{id} \\
+ \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \mathit{id} + \lbrack\!\lbrack b \rbrack\!\rbrack \\
+ \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{+} \fatsemi \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
+ \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack + \mathit{id}) \fatsemi \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
+ \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{\times } \fatsemi \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
+ \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{distrib } \fatsemi \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
+ \end{array}\]
+%subcode source isomorphisms.tex:598
+
+  \end{proof}
+\end{proposition}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{A Graphical Language for \ensuremath{\Pi } }
+
+Combinators of \ensuremath{\Pi } can be written in terms of the operators
+described in definition \ref{def:langRev} or via a graphical language
+similar in spirit to those developed for Geometry of Interaction
+\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
+\cite{selinger-graphical}. Modulo some conventions and shorthand we
+describe here, the wiring diagrams are equivalent to the operator
+based (syntactic) description of programs.  \ensuremath{\Pi } combinators
+expressed in this graphical language look like ``wiring
+diagrams''. Values take the form of ``particles'' that flow along the
+wires. Computation is expressed by the flow of particles.
+
+\begin{itemize}
+\item
+The simplest sort of diagram is the \ensuremath{\mathit{id} : b \leftrightarrow b} combinator which
+is simply represented as a wire labeled by its type \ensuremath{b}. In more
+complex diagrams, if the type of a wire is obvious from the context,
+it may be omitted.
+
+\inkscape{b-wire.pdf}
+
+\noindent
+Values flow from left to right in the graphical language of
+\ensuremath{\Pi }.  When tracing a computation, one might imagine a value
+\ensuremath{v} of type \ensuremath{b} on the wire, as shown below.
+
+\inkscape{b-wire-value.pdf}
+
+\item
+The product type \ensuremath{b_1\times b_2} may be represented both as one wire
+labeled \ensuremath{b_1\times b_2} or by two parallel wires labeled \ensuremath{b_1} and
+\ensuremath{b_2}. Both representations may be used interchangeably.
+
+\begin{multicols}{2}
+\inkscape{product-one-wire.pdf}
+\inkscape{product-two-wires.pdf}
+\end{multicols}
+
+When tracing execution using particles, one should think of one
+particle on each wire or alternatively as in folklore in the
+literature on monoidal categories as a ``wave.''
+
+\begin{multicols}{2}
+\inkscape{product-one-wire-value.pdf}
+\inkscape{product-two-wires-value.pdf}
+\end{multicols}
+
+\item
+Sum types may similarly be represented using using parallel wires with
+a \ensuremath{+} operator between them.
+
+\begin{multicols}{2}
+\inkscape{sum-one-wire.pdf}
+\inkscape{sum-two-wires.pdf}
+\end{multicols}
+
+\noindent
+When tracing the execution of \ensuremath{b_1+b_2} represented by one
+wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
+as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  When tracing
+the execution of two additive wires, a value can reside on only one of
+the two wires.
+
+\begin{multicols}{2}
+\inkscape{sum-two-wires-left-value.pdf}
+\inkscape{sum-two-wires-right-value.pdf}
+\end{multicols}
+
+\item
+When representing complex types like \ensuremath{(b_1\times b_2)+b_3} some visual
+grouping of the wires may be done to aid readability. The exact type
+however will always be clarified by the context of the diagram.
+
+\inkscape{complex-type-ambiguity.pdf}
+
+\item
+Associativity is entirely skipped in the graphical language. Hence
+three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
+\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
+associativity in the graphical representations of categories.
+
+\inkscape{assoc.pdf}
+
+\item
+Commutativity is represented by crisscrossing wires.
+
+\begin{multicols}{2}
+\inkscape{swap_times.pdf}
+\inkscape{swap_plus.pdf}
+\end{multicols}
+
+\noindent
+When tracing the execution of \ensuremath{b_1+b_2} represented by one
+wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
+as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
+tracking the flow of particles on the wires, one can verify that the
+expected types for commutativity are satisfied.
+
+\begin{multicols}{2}
+\inkscape{swap_times_value.pdf}
+\inkscape{swap_plus_value.pdf}
+\end{multicols}
+
+\item
+The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
+  multiplicative units are represented as shown below. Note that since there
+  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
+  Also since the monoidal units can be freely introduced and eliminated, in
+  many diagrams they are omitted and dealt with explicitly only when they are
+  of special interest.
+
+\begin{multicols}{2}
+\inkscape{uniti.pdf}
+\inkscape{unite.pdf}
+\end{multicols}
+
+\begin{multicols}{2}
+\inkscape{zeroi.pdf}
+\inkscape{zeroe.pdf}
+\end{multicols}
+
+\item
+Distributivity and factoring are represented using the dual
+boxes shown below:
+
+\begin{multicols}{2}
+\inkscape{dist.pdf}
+\inkscape{factor.pdf}
+\end{multicols}
+
+Distributivity and factoring are interesting because they represent
+interactions between sum and pair types. Distributivity should
+essentially be thought of as a multiplexer that redirects the flow of
+\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1+b_2}, as shown
+below.
+
+\begin{multicols}{2}
+\inkscape{dist-wire-value1.pdf}
+\inkscape{dist-wire-value2.pdf}
+\end{multicols}
+
+\noindent
+Factoring is the corresponding adjoint operation.
+
+\begin{multicols}{2}
+\inkscape{factor-wire-value1.pdf}
+\inkscape{factor-wire-value2.pdf}
+\end{multicols}
+
+\item Combinators can be composed in series (\ensuremath{c_1 \fatsemi c_2}) or
+  parallel. Sequential (series) composition corresponds to connecting
+  the output of one combinator to the input of the next.
+
+\inkscape{c1c2_seq.pdf}
+
+There are two forms of parallel composition -- combinators
+  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
+  multiplicatively \ensuremath{c_1\times c_2} (shown on the right).
+
+\begin{multicols}{2}
+\inkscape{c1c2_par_sum.pdf}
+\inkscape{c1c2_par_times.pdf}
+\end{multicols}
+
+\end{itemize}
+
+\noindent
+\textit{Example.} As an example consider the wiring diagram
+of the combinator \ensuremath{c} below:
+
+\ensuremath{c : b \times (1+1) \leftrightarrow b + b}
+
+\ensuremath{c = \mathit{swap}^{\times } \fatsemi \mathit{distrib } \fatsemi (\mathit{unite} + \mathit{unite})}
+
+\begin{center}
+\scalebox{1.5}{
+%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
+\includegraphics{inkscape/example1.pdf}
+}
+\end{center}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Expressing Programs with \ensuremath{\Pi } }
+\label{sec:langRev-examples}
+
+In this section we introduce programming in \ensuremath{\Pi } through
+several simple examples. We subsequently discuss the expressiveness of
+the language and discuss various properties.
+
+\label{examples}
+
+\paragraph*{Booleans}
+Let us start with encoding booleans. We use the type \ensuremath{1+1} to
+represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
+\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
+Boolean negation is straightforward to define:
+
+\ensuremath{\mathit{not} : \mathit{bool} \leftrightarrow \mathit{bool}}
+
+\ensuremath{\mathit{not} = \mathit{swap}^{+}}
+
+\noindent
+It is easy to verify that \ensuremath{\mathit{not}} changes \ensuremath{\mathit{true}} to \ensuremath{\mathit{false}} and
+vice versa.
+
+\paragraph*{Bit Vectors.}
+We can represent $n$-bit words using an n-ary product of
+\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
+using the type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})}.  We can perform various
+operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
+instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
+three \ensuremath{\mathit{not}} operations:
+
+\ensuremath{\mathit{not}_{\mathit{word}_3} :: \mathit{word}_3 \leftrightarrow \mathit{word}_3}
+
+\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not} \times (\mathit{not} \times \mathit{not})}
+
+\noindent We can express a 3-bit word reversal operation as follows:
+
+\ensuremath{\mathit{reverse} : \mathit{word}_3 \leftrightarrow \mathit{word}_3}
+
+\ensuremath{\mathit{reverse} = \mathit{swap}^{\times } \fatsemi (\mathit{swap}^{\times } \times \mathit{id})~ \fatsemi \mathit{assocr}^{\times }}
+
+\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
+applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
+derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
+many others we will see in this thesis, is formed by sequentially
+composing several simpler combinators. Instead of presenting the
+operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
+for presentation reasons) to flatten the tree into a sequence of
+reductions as caused by each component. Such a sequence of reductions
+is given below:
+\[\begin{array}{rlr}
+ & (v_1, (v_2, v_3)) \\
+ \mathit{swap}^{\times } & ((v_2, v_3), v_1) \\
+ \mathit{swap}^{\times } \times \mathit{id} & ((v_3, v_2), v_1) \\
+ \mathit{assocr}^{\times } & (v_3, (v_2, v_1)) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:979
+
+\noindent On the first line is the initial value. On each subsequent
+line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
+results from applying this combinator to the value on the previous
+line. For example, \ensuremath{\mathit{swap}^{\times }} transforms \ensuremath{(v_1, (v_2, v_3))} to
+\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
+the bits in reverse order.
+
+We can also draw out the graphical representation of the 3-bit reverse
+combinator. In the graphical representation, it is clear that the
+combinator achieves the required shuffling.
+
+\inkscape{reverse-3-bit.pdf}
+
+\paragraph*{Conditionals.}
+Even though \ensuremath{\Pi } lacks conditional expressions, they are
+expressible using the distributivity and factoring laws. The
+diagrammatic representation of \ensuremath{\mathit{distrib }} shows that it redirects the flow
+of a value \ensuremath{v:b} based on the value of another one of type
+\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be \ensuremath{\mathit{bool}} and apply either
+\ensuremath{c_1:b\leftrightarrow b} or \ensuremath{c_2:b\leftrightarrow b} to the value \ensuremath{v}, then we essentially
+have an `if' expression.
+
+\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool} \times b \leftrightarrow \mathit{bool} \times b}
+
+\ensuremath{\mathit{if}_{c_1,c_2} = \mathit{distrib } \fatsemi ((\mathit{id} \times c_1) + (\mathit{id} \times c_2)) \fatsemi \mathit{factor }}
+
+\inkscape{if-c1-c2.pdf}
+
+The diagram above shows the input value of type \ensuremath{(1+1)\times b} processed
+by the distribute operator \ensuremath{\mathit{distrib }}, which converts it into a value of
+type \ensuremath{(1\times b)+(1\times b)}. In the \ensuremath{\mathit{left}} branch, which corresponds to the
+case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was \ensuremath{\mathit{left} ~()}),
+the combinator~\ensuremath{c_1} is applied to the value of type~\ensuremath{b}. The right
+branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
+the value of type \ensuremath{b} through the combinator \ensuremath{c_2}.
+
+\paragraph*{Logic Gates}
+There are several universal primitives for conventional (irreversible)
+hardware circuits, such as \ensuremath{\mathit{nand}} and \ensuremath{\mathit{fanout}}. In the case
+of reversible hardware circuits, the canonical universal primitive is
+the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
+boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
+bit is negated. In a traditional language, the Toffoli gate would be
+most conveniently expressed as a conditional expression like:
+
+\noindent
+\ensuremath{ \mathit{toffoli}(v_1,v_2,v_3) = \mathit{if} ~(v_1 ~\mathit{and} ~v_2) ~\mathit{then} ~(v_1, v_2, \mathit{not}(v_3)) ~\mathit{else} ~(v_1, v_2, v_3)}
+
+We will derive Toffoli gate in \ensuremath{\Pi } by first deriving a simpler
+logic gate called \ensuremath{\mathit{cnot}}.  Consider a one-armed version, \ensuremath{\mathit{if}_c},
+of the conditional derived above. If the \ensuremath{\mathit{bool}} is
+\ensuremath{\mathit{true}}, the value of type \ensuremath{b} is modified by the operator \ensuremath{c}.
+
+\begin{center}
+\scalebox{1.5}{
+%%subcode-line{pdfimage}[diagrams/if_c.pdf]
+\includegraphics{inkscape/cnot.pdf}
+}
+\end{center}
+
+By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
+combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool}\times \mathit{bool}\leftrightarrow \mathit{bool}\times \mathit{bool}} which negates its
+second argument if the first argument is \ensuremath{\mathit{true}}. This gate
+\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.
+
+If we iterate this construction once more, the resulting combinator
+\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})\leftrightarrow \mathit{bool}\times (\mathit{bool}\times \mathit{bool})}. The
+resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
+proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
+it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
+Toffoli gate.
+
+\begin{center}
+\scalebox{1.6}{
+\includegraphics{inkscape/toffoli.pdf}
+}
+\end{center}
 

### 2018-06-26 survey/survey.tex
fixed citations

@@
 of reversible hardware circuits, the canonical universal primitive is
-the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
+the Toffoli gate~\cite{toffoli:1980}. The Toffoli gate takes three
 boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
@@
 second argument if the first argument is \ensuremath{\mathit{true}}. This gate
-\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.
+\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{toffoli:1980}.
 

### 2018-06-27 survey/survey.tex
some integration in sec3

@@
 
-%%%%%%%%%
-\subsection{Examples}
-
-Surprising expressiveness of Pi:
-
-\begin{quote}
-  Ed Fredkin pursued the idea that information must be finite in
-  density. One day, he announced that things must be even more simple
-  than that. He said that he was going to assume that information
-  itself is conserved. “You’re out of you mind, Ed.” I
-  pronounced. “That’s completely ridiculous. Nothing could happen in
-  such a world. There couldn’t even be logical gates. No decisions
-  could ever be made.” But when Fredkin gets one of his ideas, he’s
-  quite immune to objections like that; indeed, they fuel him with
-  energy. Soon he went on to assume that information processing must
-  also be reversible — and invented what’s now called the Fredkin
-  gate. (Minsky 1999)
-\end{quote}
-
-\begin{verbatim}
-examples:
-  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
-  conditionals
-  toffoli
-  fredkin
-  revisit adder example from Sec 2
-  perhaps a larger example as well (from RC 18 paper?)
-\end{verbatim}
+\begin{definition}{(Syntax of \ensuremath{\Pi })}
+\label{def:langRev}
+We collect our types, values, and combinators, to get the full language
+definition.
+\[\begin{array}{rclr}
+ \mathit{value} ~\mathit{types}, b &::=& 0 ~|~ 1 ~|~ b+b ~|~ b\times b \\
+ \mathit{values}, v &::=& () ~|~ \mathit{left} ~v ~|~ \mathit{right} ~v ~|~ (v,v) \\
+ \\
+ \mathit{combinator} ~\mathit{types}, t &::=& b \leftrightarrow b \\
+ \mathit{isomorphisms}, \mathit{iso} &::=& \identlp ~|~ \identrp ~|~ \swapp ~|~ \assoclp ~|~ \assocrp \\
+ & ~|~ & \identlt ~|~ \identrt ~|~ \swapt ~|~ \assoclt ~|~ \assocrt \\
+ & ~|~ & \absorbr ~|~ \factorzl ~|~ \dist ~|~ \factor \\
+ \mathit{combinators}, c &::=& \mathit{iso} ~|~ \idc ~|~ \ !~c ~|~ c \odot c ~|~ c \oplus c ~|~ c \otimes c \\
+ \end{array}\]
+%subcode source isomorphisms.tex:483
+\end{definition}
+
+\begin{definition}{Operational Semantics for \ensuremath{\Pi }. }
+\label{def:operational-langRev}
+Given a program \ensuremath{c : b_1 \leftrightarrow b_2} in \ensuremath{\Pi }, we can run it by supplying
+it with a value \ensuremath{ v_1 : b_1}. The evaluation rules \ensuremath{c ~v_1 \mapsto v_2} are given below.
+
+Additive fragment:
+
+\[\begin{array}{rlcl}
+ \identlp & (\mathit{right} ~v) &\mapsto & v \\
+ \identrp & v &\mapsto & \mathit{right} ~v \\
+ \swapp & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
+ \swapp & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
+ \assoclp & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
+ \assoclp & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
+ \assoclp & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
+ \assocrp & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
+ \assocrp & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
+ \assocrp & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:504
+
+Multiplicative fragment:
+
+\[\begin{array}{rlcl}
+ \identlt & ((), v) &\mapsto & v \\
+ \identrt & v &\mapsto & ((), v) \\
+ \swapt & (v_1, v_2) &\mapsto & (v_2, v_1) \\
+ \assoclt & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
+ \assocrt & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:514
+
+Distributivity and factoring:
+
+\[\begin{array}{rlcl}
+ \dist & (\mathit{left} ~v_1, v_3) &\mapsto & \mathit{left} ~(v_1, v_3) \\
+ \dist & (\mathit{right} ~v_2, v_3) &\mapsto & \mathit{right} ~(v_2, v_3) \\
+ \factor & (\mathit{left} ~(v_1, v_3)) &\mapsto & (\mathit{left} ~v_1, v_3) \\
+ \factor & (\mathit{right} ~(v_2, v_3)) &\mapsto & (\mathit{right} ~v_2, v_3) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:523
+
+The evaluation rules of the composition combinators are given below:
+
+$$
+\infer{ \idc ~v \mapsto v}{
+	 ~
+}
+\quad
+\infer{ (\ !~c) ~v_1 \mapsto v_2}{
+	 c{\mathit{^{\dagger} }} v_1 \mapsto v_2
+}
+\quad
+\infer{ (c_1\odot c_2) ~v_1 \mapsto v_2}{
+	 c_1 ~v_1 \mapsto v
+	&
+	 c_2 ~v \mapsto v_2
+}
+$$
+$$
+\infer{ (c_1 \oplus c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
+	 c_1 ~v_1 \mapsto v_2
+}
+\quad
+\infer{ (c_1 \oplus c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
+	 c_2 ~v_1 \mapsto v_2
+}
+$$
+$$
+\infer{ (c_1 \otimes c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
+	 c_1 ~v_1 \mapsto v_3
+	&
+	 c_2 ~v_2 \mapsto v_4
+}
+$$
+%subcode source isomorphisms.tex:546
+
+\end{definition}
+
+Since there are no values that have the type \ensuremath{0}, the reductions for
+the combinators \ensuremath{\identlp}, \ensuremath{\identrp}, \ensuremath{\absorbr} and \ensuremath{\factorzl}
+omit the impossible cases. Also, the semantics of \ensuremath{\ !~c} are
+defined in terms of the adjoint of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
+which we formalize below.
+
+\begin{definition}[Adjoints, \ensuremath{c^\mathit{^{\dagger} }}]
+ The adjoint of a combinator \ensuremath{c} is defined by \ensuremath{c^\mathit{^{\dagger} }} as follows:
+
+  \begin{itemize}
+  \item For primitive isomophisms \ensuremath{c}, \ensuremath{c^\mathit{^{\dagger} }} is given by its
+    isomorphic inverse operator from the table in Definition
+    \ref{def:primitives-langRev}.
+
+  \item \ensuremath{(c_1 \otimes c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \otimes c_2^\mathit{^{\dagger} }}
+
+  \item \ensuremath{(c_1 \oplus c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \oplus c_2^\mathit{^{\dagger} }}
+
+  \item \ensuremath{(c_1\odot c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \odot c_1^\mathit{^{\dagger} }}. (Note that the
+    order of combinators has been reversed).
+
+  \end{itemize}
+
+\end{definition}
+
+\begin{definition}[Observational Equivalence. \ensuremath{c_1=c_2}]
+\label{def:pi-eqv}
+  Combinators \ensuremath{c_1} and \ensuremath{c_2}, where \ensuremath{c_1 : b_1 \leftrightarrow b_2} and
+  \ensuremath{c_2:b_1\leftrightarrow b_2}, are observationally equivalent,\ensuremath{c_1=c_2}, iff
+
+  \ensuremath{\forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \Longleftrightarrow c_2 ~v_1 \mapsto v_2}.
+\end{definition}
+
+\begin{proposition}{Canonical Isomorphism.}
+  For any type \ensuremath{b} there exists the isomorphism \ensuremath{b \leftrightarrow b^{|b|}}.
+  \begin{proof}
+    The fact that such an isomorphism exists is evident from the
+    definition of size and what it means for two types to be
+    isomorphic. While many equivalent constructions are possible
+    for any type \ensuremath{b}, one such construction is given by \ensuremath{\lbrack\!\lbrack b\rbrack\!\rbrack}:
+
+\[\begin{array}{rclr}
+ \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \idc \\
+ \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \idc \\
+ \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \idc \oplus \lbrack\!\lbrack b \rbrack\!\rbrack \\
+ \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \assocrp \odot \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
+ \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack \oplus \idc) \odot \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
+ \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \assocrt \odot \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
+ \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \dist \odot \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
+ \end{array}\]
+%subcode source isomorphisms.tex:598
+
+  \end{proof}
+\end{proposition}
 
@@
 %% when we have both we do have a wave-particle duality
+
+Combinators of \ensuremath{\Pi } can be written in terms of the operators
+described in definition \ref{def:langRev} or via a graphical language
+similar in spirit to those developed for Geometry of Interaction
+\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
+\cite{selinger-graphical}. Modulo some conventions and shorthand we
+describe here, the wiring diagrams are equivalent to the operator
+based (syntactic) description of programs.  \ensuremath{\Pi } combinators
+expressed in this graphical language look like ``wiring
+diagrams''. Values take the form of ``particles'' that flow along the
+wires. Computation is expressed by the flow of particles.
+
+\begin{itemize}
+\item
+The simplest sort of diagram is the \ensuremath{\idc : b \leftrightarrow b} combinator which
+is simply represented as a wire labeled by its type \ensuremath{b}. In more
+complex diagrams, if the type of a wire is obvious from the context,
+it may be omitted.
+
+\inkscape{b-wire.pdf}
+
+\noindent
+Values flow from left to right in the graphical language of
+\ensuremath{\Pi }.  When tracing a computation, one might imagine a value
+\ensuremath{v} of type \ensuremath{b} on the wire, as shown below.
+
+\inkscape{b-wire-value.pdf}
+
+\item
+The product type \ensuremath{b_1\times b_2} may be represented both as one wire
+labeled \ensuremath{b_1\times b_2} or by two parallel wires labeled \ensuremath{b_1} and
+\ensuremath{b_2}. Both representations may be used interchangeably.
+
+\begin{multicols}{2}
+\inkscape{product-one-wire.pdf}
+\inkscape{product-two-wires.pdf}
+\end{multicols}
+
+When tracing execution using particles, one should think of one
+particle on each wire or alternatively as in folklore in the
+literature on monoidal categories as a ``wave.''
+
+\begin{multicols}{2}
+\inkscape{product-one-wire-value.pdf}
+\inkscape{product-two-wires-value.pdf}
+\end{multicols}
+
+\item
+Sum types may similarly be represented using using parallel wires with
+a \ensuremath{+} operator between them.
+
+\begin{multicols}{2}
+\inkscape{sum-one-wire.pdf}
+\inkscape{sum-two-wires.pdf}
+\end{multicols}
+
+\noindent
+When tracing the execution of \ensuremath{b_1+b_2} represented by one
+wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
+as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  When tracing
+the execution of two additive wires, a value can reside on only one of
+the two wires.
+
+\begin{multicols}{2}
+\inkscape{sum-two-wires-left-value.pdf}
+\inkscape{sum-two-wires-right-value.pdf}
+\end{multicols}
+
+\item
+When representing complex types like \ensuremath{(b_1\times b_2)+b_3} some visual
+grouping of the wires may be done to aid readability. The exact type
+however will always be clarified by the context of the diagram.
+
+\inkscape{complex-type-ambiguity.pdf}
+
+\item
+Associativity is entirely skipped in the graphical language. Hence
+three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
+\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
+associativity in the graphical representations of categories.
+
+\inkscape{assoc.pdf}
+
+\item
+Commutativity is represented by crisscrossing wires.
+
+\begin{multicols}{2}
+\inkscape{swap_times.pdf}
+\inkscape{swap_plus.pdf}
+\end{multicols}
+
+\noindent
+When tracing the execution of \ensuremath{b_1+b_2} represented by one
+wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
+as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
+tracking the flow of particles on the wires, one can verify that the
+expected types for commutativity are satisfied.
+
+\begin{multicols}{2}
+\inkscape{swap_times_value.pdf}
+\inkscape{swap_plus_value.pdf}
+\end{multicols}
+
+\item
+The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
+  multiplicative units are represented as shown below. Note that since there
+  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
+  Also since the monoidal units can be freely introduced and eliminated, in
+  many diagrams they are omitted and dealt with explicitly only when they are
+  of special interest.
+
+\begin{multicols}{2}
+\inkscape{uniti.pdf}
+\inkscape{unite.pdf}
+\end{multicols}
+
+\begin{multicols}{2}
+\inkscape{zeroi.pdf}
+\inkscape{zeroe.pdf}
+\end{multicols}
+
+\item
+Distributivity and factoring are represented using the dual
+boxes shown below:
+
+\begin{multicols}{2}
+\inkscape{dist.pdf}
+\inkscape{factor.pdf}
+\end{multicols}
+
+Distributivity and factoring are interesting because they represent
+interactions between sum and pair types. Distributivity should
+essentially be thought of as a multiplexer that redirects the flow of
+\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1+b_2}, as shown
+below.
+
+\begin{multicols}{2}
+\inkscape{dist-wire-value1.pdf}
+\inkscape{dist-wire-value2.pdf}
+\end{multicols}
+
+\noindent
+Factoring is the corresponding adjoint operation.
+
+\begin{multicols}{2}
+\inkscape{factor-wire-value1.pdf}
+\inkscape{factor-wire-value2.pdf}
+\end{multicols}
+
+\item Combinators can be composed in series (\ensuremath{c_1 \odot c_2}) or
+  parallel. Sequential (series) composition corresponds to connecting
+  the output of one combinator to the input of the next.
+
+\inkscape{c1c2_seq.pdf}
+
+There are two forms of parallel composition -- combinators
+  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
+  multiplicatively \ensuremath{c_1\times c_2} (shown on the right).
+
+\begin{multicols}{2}
+\inkscape{c1c2_par_sum.pdf}
+\inkscape{c1c2_par_times.pdf}
+\end{multicols}
+
+\end{itemize}
+
+\noindent
+\textit{Example.} As an example consider the wiring diagram
+of the combinator \ensuremath{c} below:
+
+\ensuremath{c : b \times (1+1) \leftrightarrow b + b}
+
+\ensuremath{c = \swapt \odot \dist \odot (\identlt + \identlt)}
+
+\begin{center}
+\scalebox{1.5}{
+%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
+\includegraphics{inkscape/example1.pdf}
+}
+\end{center}
 
@@
 
-\begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
-\label{def:primitives-langRev}
-
-\[\begin{array}{rrcll}
- \mathit{zeroe} :& 0 + b &\leftrightarrow & b &: \mathit{zeroi} \\
- \mathit{swap}^{+} :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \mathit{swap}^{+} \\
- \mathit{assocl}^{+} :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \mathit{assocr}^{+} \\
- \mathit{unite} :& 1 \times b &\leftrightarrow & b &: \mathit{uniti} \\
- \mathit{swap}^{\times } :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \mathit{swap}^{\times } \\
- \mathit{assocl}^{\times } :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \mathit{assocr}^{\times } \\
- \mathit{distrib }_{0} :& 0 \times b &\leftrightarrow & 0 &: \mathit{factor }_{0} \\
- \mathit{distrib } :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \mathit{factor } \\
- \end{array}\]
-%subcode source isomorphisms.tex:427
-
-\end{definition}
-
-Each line of this table is to be read as the definition of one or two
-operators. For example, corresponding to the \textit{identity of
-  \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
-\ensuremath{\mathit{unite}:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
-\ensuremath{\mathit{uniti}:b\leftrightarrow 1\times b} (reading the isomorphism from right to
-left). These operators are inverses of each other.  Each of the two
-cases of commutativity defines one operator that is its own inverse.
-
-Having defined primitive operators, we need some means of composing
-them. We construct the composition combinators out of the closure
-conditions for isomorphisms.
-
-\begin{definition}{Composition in \ensuremath{\Pi } }
-$$
-\infer{ \mathit{id} : b \leftrightarrow b}{
-	 ~
-}
-\quad
-\infer{ \mathit{sym} ~c : b_2 \leftrightarrow b_1}{
-	 c : b_1 \leftrightarrow b_2
-}
-\quad
-\infer{ c_1\fatsemi c_2 : b_1 \leftrightarrow b_3}{
-	 c_1 : b_1 \leftrightarrow b_2
-	&
-	 c_2 : b_2 \leftrightarrow b_3
-}
-$$
-$$
-\infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
-	 c_1 : b_1 \leftrightarrow b_3
-	&
-	 c_2 : b_2 \leftrightarrow b_4
-}
-\quad
-\infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
-	 c_1 : b_1 \leftrightarrow b_3
-	&
-	 c_2 : b_2 \leftrightarrow b_4
-}
-$$
-%subcode source isomorphisms.tex:462
-\end{definition}
-
-Thus we have program constructs that witness reflexivity \ensuremath{\mathit{id}},
-symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\fatsemi} and two parallel
-composition combinators, one for sums~\ensuremath{+} and one for
-pairs~\ensuremath{\times}.
-
-\begin{definition}{(Syntax of \ensuremath{\Pi })}
-\label{def:langRev}
-We collect our types, values, and combinators, to get the full language
-definition.
-\[\begin{array}{rclr}
- \mathit{value} ~\mathit{types}, b &::=& 0 ~|~ 1 ~|~ b+b ~|~ b\times b \\
- \mathit{values}, v &::=& () ~|~ \mathit{left} ~v ~|~ \mathit{right} ~v ~|~ (v,v) \\
- \\
- \mathit{combinator} ~\mathit{types}, t &::=& b \leftrightarrow b \\
- \mathit{isomorphisms}, \mathit{iso} &::=& \mathit{zeroe} ~|~ \mathit{zeroi} ~|~ \mathit{swap}^{+} ~|~ \mathit{assocl}^{+} ~|~ \mathit{assocr}^{+} \\
- & ~|~ & \mathit{unite} ~|~ \mathit{uniti} ~|~ \mathit{swap}^{\times } ~|~ \mathit{assocl}^{\times } ~|~ \mathit{assocr}^{\times } \\
- & ~|~ & \mathit{distrib }_{0} ~|~ \mathit{factor }_{0} ~|~ \mathit{distrib } ~|~ \mathit{factor } \\
- \mathit{combinators}, c &::=& \mathit{iso} ~|~ \mathit{id} ~|~ \mathit{sym} ~c ~|~ c \fatsemi c ~|~ c + c ~|~ c \times c \\
- \end{array}\]
-%subcode source isomorphisms.tex:483
-\end{definition}
-
-\begin{definition}{Operational Semantics for \ensuremath{\Pi }. }
-\label{def:operational-langRev}
-Given a program \ensuremath{c : b_1 \leftrightarrow b_2} in \ensuremath{\Pi }, we can run it by supplying
-it with a value \ensuremath{ v_1 : b_1}. The evaluation rules \ensuremath{c ~v_1 \mapsto v_2} are given below.
-
-Additive fragment:
-
-\[\begin{array}{rlcl}
- \mathit{zeroe} & (\mathit{right} ~v) &\mapsto & v \\
- \mathit{zeroi} & v &\mapsto & \mathit{right} ~v \\
- \mathit{swap}^{+} & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
- \mathit{swap}^{+} & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
- \mathit{assocl}^{+} & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
- \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
- \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
- \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
- \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
- \mathit{assocr}^{+} & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
- \end{array}\]
-%subcode source isomorphisms.tex:504
-
-Multiplicative fragment:
-
-\[\begin{array}{rlcl}
- \mathit{unite} & ((), v) &\mapsto & v \\
- \mathit{uniti} & v &\mapsto & ((), v) \\
- \mathit{swap}^{\times } & (v_1, v_2) &\mapsto & (v_2, v_1) \\
- \mathit{assocl}^{\times } & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
- \mathit{assocr}^{\times } & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
- \end{array}\]
-%subcode source isomorphisms.tex:514
-
-Distributivity and factoring:
-
-\[\begin{array}{rlcl}
- \mathit{distrib } & (\mathit{left} ~v_1, v_3) &\mapsto & \mathit{left} ~(v_1, v_3) \\
- \mathit{distrib } & (\mathit{right} ~v_2, v_3) &\mapsto & \mathit{right} ~(v_2, v_3) \\
- \mathit{factor } & (\mathit{left} ~(v_1, v_3)) &\mapsto & (\mathit{left} ~v_1, v_3) \\
- \mathit{factor } & (\mathit{right} ~(v_2, v_3)) &\mapsto & (\mathit{right} ~v_2, v_3) \\
- \end{array}\]
-%subcode source isomorphisms.tex:523
-
-The evaluation rules of the composition combinators are given below:
-
-$$
-\infer{ \mathit{id} ~v \mapsto v}{
-	 ~
-}
-\quad
-\infer{ (\mathit{sym} ~c) ~v_1 \mapsto v_2}{
-	 c{\mathit{^{\dagger} }} v_1 \mapsto v_2
-}
-\quad
-\infer{ (c_1\fatsemi c_2) ~v_1 \mapsto v_2}{
-	 c_1 ~v_1 \mapsto v
-	&
-	 c_2 ~v \mapsto v_2
-}
-$$
-$$
-\infer{ (c_1 + c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
-	 c_1 ~v_1 \mapsto v_2
-}
-\quad
-\infer{ (c_1 + c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
-	 c_2 ~v_1 \mapsto v_2
-}
-$$
-$$
-\infer{ (c_1 \times c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
-	 c_1 ~v_1 \mapsto v_3
-	&
-	 c_2 ~v_2 \mapsto v_4
-}
-$$
-%subcode source isomorphisms.tex:546
-
-\end{definition}
-
-Since there are no values that have the type \ensuremath{0}, the reductions for
-the combinators \ensuremath{\mathit{zeroe}}, \ensuremath{\mathit{zeroi}}, \ensuremath{\mathit{distrib }_{0}} and \ensuremath{\mathit{factor }_{0}}
-omit the impossible cases. Also, the semantics of \ensuremath{\mathit{sym} ~c} are
-defined in terms of the adjoint of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
-which we formalize below.
-
-\begin{definition}[Adjoints, \ensuremath{c^\mathit{^{\dagger} }}]
- The adjoint of a combinator \ensuremath{c} is defined by \ensuremath{c^\mathit{^{\dagger} }} as follows:
-
-  \begin{itemize}
-  \item For primitive isomophisms \ensuremath{c}, \ensuremath{c^\mathit{^{\dagger} }} is given by its
-    isomorphic inverse operator from the table in Definition
-    \ref{def:primitives-langRev}.
-
-  \item \ensuremath{(c_1 \times c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \times c_2^\mathit{^{\dagger} }}
-
-  \item \ensuremath{(c_1 + c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } + c_2^\mathit{^{\dagger} }}
-
-  \item \ensuremath{(c_1\fatsemi c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \fatsemi c_1^\mathit{^{\dagger} }}. (Note that the
-    order of combinators has been reversed).
-
-  \end{itemize}
-
-\end{definition}
-
-\begin{definition}[Observational Equivalence. \ensuremath{c_1=c_2}]
-\label{def:pi-eqv}
-  Combinators \ensuremath{c_1} and \ensuremath{c_2}, where \ensuremath{c_1 : b_1 \leftrightarrow b_2} and
-  \ensuremath{c_2:b_1\leftrightarrow b_2}, are observationally equivalent,\ensuremath{c_1=c_2}, iff
-
-  \ensuremath{\forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \Longleftrightarrow c_2 ~v_1 \mapsto v_2}.
-\end{definition}
-
-\begin{proposition}{Canonical Isomorphism.}
-  For any type \ensuremath{b} there exists the isomorphism \ensuremath{b \leftrightarrow b^{|b|}}.
-  \begin{proof}
-    The fact that such an isomorphism exists is evident from the
-    definition of size and what it means for two types to be
-    isomorphic. While many equivalent constructions are possible
-    for any type \ensuremath{b}, one such construction is given by \ensuremath{\lbrack\!\lbrack b\rbrack\!\rbrack}:
-
-\[\begin{array}{rclr}
- \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \mathit{id} \\
- \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \mathit{id} \\
- \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \mathit{id} + \lbrack\!\lbrack b \rbrack\!\rbrack \\
- \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{+} \fatsemi \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
- \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack + \mathit{id}) \fatsemi \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
- \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{\times } \fatsemi \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
- \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{distrib } \fatsemi \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
- \end{array}\]
-%subcode source isomorphisms.tex:598
-
-  \end{proof}
-\end{proposition}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{A Graphical Language for \ensuremath{\Pi } }
-
-Combinators of \ensuremath{\Pi } can be written in terms of the operators
-described in definition \ref{def:langRev} or via a graphical language
-similar in spirit to those developed for Geometry of Interaction
-\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
-\cite{selinger-graphical}. Modulo some conventions and shorthand we
-describe here, the wiring diagrams are equivalent to the operator
-based (syntactic) description of programs.  \ensuremath{\Pi } combinators
-expressed in this graphical language look like ``wiring
-diagrams''. Values take the form of ``particles'' that flow along the
-wires. Computation is expressed by the flow of particles.
-
-\begin{itemize}
-\item
-The simplest sort of diagram is the \ensuremath{\mathit{id} : b \leftrightarrow b} combinator which
-is simply represented as a wire labeled by its type \ensuremath{b}. In more
-complex diagrams, if the type of a wire is obvious from the context,
-it may be omitted.
-
-\inkscape{b-wire.pdf}
-
-\noindent
-Values flow from left to right in the graphical language of
-\ensuremath{\Pi }.  When tracing a computation, one might imagine a value
-\ensuremath{v} of type \ensuremath{b} on the wire, as shown below.
-
-\inkscape{b-wire-value.pdf}
-
-\item
-The product type \ensuremath{b_1\times b_2} may be represented both as one wire
-labeled \ensuremath{b_1\times b_2} or by two parallel wires labeled \ensuremath{b_1} and
-\ensuremath{b_2}. Both representations may be used interchangeably.
-
-\begin{multicols}{2}
-\inkscape{product-one-wire.pdf}
-\inkscape{product-two-wires.pdf}
-\end{multicols}
-
-When tracing execution using particles, one should think of one
-particle on each wire or alternatively as in folklore in the
-literature on monoidal categories as a ``wave.''
-
-\begin{multicols}{2}
-\inkscape{product-one-wire-value.pdf}
-\inkscape{product-two-wires-value.pdf}
-\end{multicols}
-
-\item
-Sum types may similarly be represented using using parallel wires with
-a \ensuremath{+} operator between them.
-
-\begin{multicols}{2}
-\inkscape{sum-one-wire.pdf}
-\inkscape{sum-two-wires.pdf}
-\end{multicols}
-
-\noindent
-When tracing the execution of \ensuremath{b_1+b_2} represented by one
-wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
-as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  When tracing
-the execution of two additive wires, a value can reside on only one of
-the two wires.
-
-\begin{multicols}{2}
-\inkscape{sum-two-wires-left-value.pdf}
-\inkscape{sum-two-wires-right-value.pdf}
-\end{multicols}
-
-\item
-When representing complex types like \ensuremath{(b_1\times b_2)+b_3} some visual
-grouping of the wires may be done to aid readability. The exact type
-however will always be clarified by the context of the diagram.
-
-\inkscape{complex-type-ambiguity.pdf}
-
-\item
-Associativity is entirely skipped in the graphical language. Hence
-three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
-\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
-associativity in the graphical representations of categories.
-
-\inkscape{assoc.pdf}
-
-\item
-Commutativity is represented by crisscrossing wires.
-
-\begin{multicols}{2}
-\inkscape{swap_times.pdf}
-\inkscape{swap_plus.pdf}
-\end{multicols}
-
-\noindent
-When tracing the execution of \ensuremath{b_1+b_2} represented by one
-wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
-as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
-tracking the flow of particles on the wires, one can verify that the
-expected types for commutativity are satisfied.
-
-\begin{multicols}{2}
-\inkscape{swap_times_value.pdf}
-\inkscape{swap_plus_value.pdf}
-\end{multicols}
-
-\item
-The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
-  multiplicative units are represented as shown below. Note that since there
-  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
-  Also since the monoidal units can be freely introduced and eliminated, in
-  many diagrams they are omitted and dealt with explicitly only when they are
-  of special interest.
-
-\begin{multicols}{2}
-\inkscape{uniti.pdf}
-\inkscape{unite.pdf}
-\end{multicols}
-
-\begin{multicols}{2}
-\inkscape{zeroi.pdf}
-\inkscape{zeroe.pdf}
-\end{multicols}
-
-\item
-Distributivity and factoring are represented using the dual
-boxes shown below:
-
-\begin{multicols}{2}
-\inkscape{dist.pdf}
-\inkscape{factor.pdf}
-\end{multicols}
-
-Distributivity and factoring are interesting because they represent
-interactions between sum and pair types. Distributivity should
-essentially be thought of as a multiplexer that redirects the flow of
-\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1+b_2}, as shown
-below.
-
-\begin{multicols}{2}
-\inkscape{dist-wire-value1.pdf}
-\inkscape{dist-wire-value2.pdf}
-\end{multicols}
-
-\noindent
-Factoring is the corresponding adjoint operation.
-
-\begin{multicols}{2}
-\inkscape{factor-wire-value1.pdf}
-\inkscape{factor-wire-value2.pdf}
-\end{multicols}
-
-\item Combinators can be composed in series (\ensuremath{c_1 \fatsemi c_2}) or
-  parallel. Sequential (series) composition corresponds to connecting
-  the output of one combinator to the input of the next.
-
-\inkscape{c1c2_seq.pdf}
-
-There are two forms of parallel composition -- combinators
-  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
-  multiplicatively \ensuremath{c_1\times c_2} (shown on the right).
-
-\begin{multicols}{2}
-\inkscape{c1c2_par_sum.pdf}
-\inkscape{c1c2_par_times.pdf}
-\end{multicols}
-
-\end{itemize}
-
-\noindent
-\textit{Example.} As an example consider the wiring diagram
-of the combinator \ensuremath{c} below:
-
-\ensuremath{c : b \times (1+1) \leftrightarrow b + b}
-
-\ensuremath{c = \mathit{swap}^{\times } \fatsemi \mathit{distrib } \fatsemi (\mathit{unite} + \mathit{unite})}
-
-\begin{center}
-\scalebox{1.5}{
-%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
-\includegraphics{inkscape/example1.pdf}
-}
-\end{center}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Expressing Programs with \ensuremath{\Pi } }
+% \begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
+% \label{def:primitives-langRev}
+
+% \[\begin{array}{rrcll}
+%  \identlp :& 0 + b &\leftrightarrow & b &: \identrp \\
+%  \swapp :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \swapp \\
+%  \assoclp :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \assocrp \\
+%  \identlt :& 1 \times b &\leftrightarrow & b &: \identrt \\
+%  \swapt :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \swapt \\
+%  \assoclt :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \assocrt \\
+%  \absorbr :& 0 \times b &\leftrightarrow & 0 &: \factorzl \\
+%  \dist :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \factor \\
+%  \end{array}\]
+% %subcode source isomorphisms.tex:427
+
+% \end{definition}
+
+% Each line of this table is to be read as the definition of one or two
+% operators. For example, corresponding to the \textit{identity of
+%   \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
+% \ensuremath{\identlt:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
+% \ensuremath{\identrt:b\leftrightarrow 1\times b} (reading the isomorphism from right to
+% left). These operators are inverses of each other.  Each of the two
+% cases of commutativity defines one operator that is its own inverse.
+
+% Having defined primitive operators, we need some means of composing
+% them. We construct the composition combinators out of the closure
+% conditions for isomorphisms.
+
+% \begin{definition}{Composition in \ensuremath{\Pi } }
+% $$
+% \infer{ \idc : b \leftrightarrow b}{
+% 	 ~
+% }
+% \quad
+% \infer{ \ !~c : b_2 \leftrightarrow b_1}{
+% 	 c : b_1 \leftrightarrow b_2
+% }
+% \quad
+% \infer{ c_1\odot c_2 : b_1 \leftrightarrow b_3}{
+% 	 c_1 : b_1 \leftrightarrow b_2
+% 	&
+% 	 c_2 : b_2 \leftrightarrow b_3
+% }
+% $$
+% $$
+% \infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
+% 	 c_1 : b_1 \leftrightarrow b_3
+% 	&
+% 	 c_2 : b_2 \leftrightarrow b_4
+% }
+% \quad
+% \infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
+% 	 c_1 : b_1 \leftrightarrow b_3
+% 	&
+% 	 c_2 : b_2 \leftrightarrow b_4
+% }
+% $$
+% %subcode source isomorphisms.tex:462
+% \end{definition}
+
+% Thus we have program constructs that witness reflexivity \ensuremath{\idc},
+% symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\odot} and two parallel
+% composition combinators, one for sums~\ensuremath{+} and one for
+% pairs~\ensuremath{\times}.
+
+%%%%%%%%%
+\subsection{Examples}
+
+Surprising expressiveness of Pi:
+
+\begin{quote}
+  Ed Fredkin pursued the idea that information must be finite in
+  density. One day, he announced that things must be even more simple
+  than that. He said that he was going to assume that information
+  itself is conserved. “You’re out of you mind, Ed.” I
+  pronounced. “That’s completely ridiculous. Nothing could happen in
+  such a world. There couldn’t even be logical gates. No decisions
+  could ever be made.” But when Fredkin gets one of his ideas, he’s
+  quite immune to objections like that; indeed, they fuel him with
+  energy. Soon he went on to assume that information processing must
+  also be reversible — and invented what’s now called the Fredkin
+  gate. (Minsky 1999)
+\end{quote}
+
+\begin{verbatim}
+examples:
+  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
+  conditionals
+  toffoli
+  fredkin
+  revisit adder example from Sec 2
+  perhaps a larger example as well (from RC 18 paper?)
+\end{verbatim}
+
 \label{sec:langRev-examples}
@@
 
-\ensuremath{\mathit{not} = \mathit{swap}^{+}}
+\ensuremath{\mathit{not} = \swapp}
 
@@
 
-\ensuremath{\mathit{reverse} = \mathit{swap}^{\times } \fatsemi (\mathit{swap}^{\times } \times \mathit{id})~ \fatsemi \mathit{assocr}^{\times }}
+\ensuremath{\mathit{reverse} = \swapt \odot (\swapt \times \idc)~ \odot \assocrt}
 
@@
  & (v_1, (v_2, v_3)) \\
- \mathit{swap}^{\times } & ((v_2, v_3), v_1) \\
- \mathit{swap}^{\times } \times \mathit{id} & ((v_3, v_2), v_1) \\
- \mathit{assocr}^{\times } & (v_3, (v_2, v_1)) \\
+ \swapt & ((v_2, v_3), v_1) \\
+ \swapt \times \idc & ((v_3, v_2), v_1) \\
+ \assocrt & (v_3, (v_2, v_1)) \\
  \end{array}\]
@@
 results from applying this combinator to the value on the previous
-line. For example, \ensuremath{\mathit{swap}^{\times }} transforms \ensuremath{(v_1, (v_2, v_3))} to
+line. For example, \ensuremath{\swapt} transforms \ensuremath{(v_1, (v_2, v_3))} to
 \ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
@@
 expressible using the distributivity and factoring laws. The
-diagrammatic representation of \ensuremath{\mathit{distrib }} shows that it redirects the flow
+diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
 of a value \ensuremath{v:b} based on the value of another one of type
@@
 
-\ensuremath{\mathit{if}_{c_1,c_2} = \mathit{distrib } \fatsemi ((\mathit{id} \times c_1) + (\mathit{id} \times c_2)) \fatsemi \mathit{factor }}
+\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc \times c_1) + (\idc \times c_2)) \odot \factor}
 
@@
 The diagram above shows the input value of type \ensuremath{(1+1)\times b} processed
-by the distribute operator \ensuremath{\mathit{distrib }}, which converts it into a value of
+by the distribute operator \ensuremath{\dist}, which converts it into a value of
 type \ensuremath{(1\times b)+(1\times b)}. In the \ensuremath{\mathit{left}} branch, which corresponds to the
@@
   \ctr{a} and \ctr{b} can be mapped to values of type \ctr{ti}, the required
-  translation of \ctr{c} is given by \ctr{c1} $\fatsemi$ \ctr{c2}.
+  translation of \ctr{c} is given by \ctr{c1} $\odot$ \ctr{c2}.
 \end{enumerate}
@@
 \ctr{lab} (the label) and every RHS clause corresponds to either \ctr{b} (the
-output) or \ctr{lab} (the label). The resulting \ctr{c1} $\fatsemi$ \ctr{c2}
+output) or \ctr{lab} (the label). The resulting \ctr{c1} $\odot$ \ctr{c2}
 has type \lstinline{lab+a} $\leftrightarrow$ \lstinline{lab+b} and the
-required compilation of \ctr{c} is given by \ensuremath{\mathit{trace}(} \ctr{c1} $\fatsemi$
+required compilation of \ctr{c} is given by \ensuremath{\mathit{trace}(} \ctr{c1} $\odot$
 \ctr{c2} \ensuremath{)} \ctr{: a} $\leftrightarrow$ \ctr{b}. This extends to multiple

### 2018-06-27 survey/survey.tex
first pass on all of 4.2 done.

@@
 
-The constructs in Fig.~\ref{figc} are significantly more subtle.
-\jc{WAVEFRONT}
+The constructs in Fig.~\ref{figc} are significantly more subtle, as they
+deal with combinators involving $0$, aka an impossibility.  For example,
+\[  {(c \otimes \idc_{0}) \odot \absorbl \Leftrightarrow \absorbl \odot \idc_{0}}
+\]
+(where we have explicitly annotated the types of $\idc$ for increased clarity)
+tells us that of the two ways of transforming from $t_1 \times 0$ to $0$, 
+namely first doing some arbitrary transformation $c$ from $t_1$ to $t_2$ and
+(in parallel) leaving $0$ alone then eliminating $0$, or first eliminating $0$
+then doing the identity (at $0$), are equivalent. This is the ``naturality'' of
+$\absorbl$. One item to note is the fact that this combinator is not
+irreducible, as the $\idc$ on the right can be eliminated. But that is actually
+a property visible at an even higher level (which we will not touch in this
+paper).  The next $3$ are similarly expressing the naturality of $\absorbr$,
+$\factorzl$ and $\factorzr$.
+
+The next combinator, $\absorbr \Leftrightarrow \absorbl$,
+is particularly fascinating: while it says something simple ---
+that the two obvious ways of transforming $0 \times 0$ into $0$, namely
+absorbing either the left or right zero --- it implies something subtle.
+A straightforward proof of $\absorbl$ which proceeds by saying that
+$0 \times t$ cannot be inhabited because the first member of the pair
+cannot, is not in fact equivalent to $\absorbr$ on $0 \times 0$.
+However, if we instead define $\absorbl$ to ``transport'' the
+putative impossible first member of the pair to its (equally
+impossible) output, then these do form equivalent pairs.
+The next few in Fig.~\ref{figc} also express how $\absorbr$ and
+$\absorbl$ interact with other combinators. As seen previously,
+all of these arise as critical pairs. What is much more subtle here
+is that the types involved often are asymmetric: they do not have
+the same occurences on the left and right. Such cases are
+particularly troublesome for finding normal forms \jc{should add
+some citations here}.
 
@@
 
-The constructs in Fig.~\ref{figb} \ldots
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
-
 \begin{figure}[t]
@@
 
-The constructs in Fig.~\ref{figa} \ldots
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\end{tikzpicture}
-\end{center}
+The constructs in Fig.~\ref{figb} and Fig.~\ref{figa} relating associativity and
+distributivity, and commutativity and distributivity, have more in common with
+previous sets of combinators.  They do arise from non-trivial critical pairs
+of different ways of going between equivalent types. The last one of
+Fig.~\ref{figb} is particularly daunting, involving a sequence of $3$ combinators
+on the left and $6$ on the right.
 

### 2018-06-28 survey/survey.tex
Augment Pi in section 3.2 to match how it is actually used in section 4.  Comment on this as well.

@@
 \newcommand{\factorl}{\mathit{factorl}}
-\newcommand{\distz}{\mathit{absorbr}}
+%% \newcommand{\distz}{\mathit{absorbr}}
 \newcommand{\iso}{\leftrightarrow}
@@
 \\
-\distz :&~ \bot \times A & \simeq & \bot ~ &: \factorzl \\
+\absorbr :&~ \bot \times A & \simeq & \bot ~ &: \factorzl \\
 \dist :&~ (A \uplus B) \times C & \simeq & (A \times C) \uplus (B \times C)~ &: \factor
@@
 \\
-\distz :&~ 0 * t & \iso & 0 ~ &: \factorzl \\
+\absorbr :&~ 0 * t & \iso & 0 ~ &: \factorzl \\
 \dist :&~ (t_1 + t_2) * t_3 & \iso & (t_1 * t_3) + (t_2 * t_3)~ &: \factor
@@
 
-\jc{it would be nice to have a picture or two here}
-
 \begin{figure}[t]
@@
 
-\amr{sound and complete isomorphisms for finite types}
+\begin{figure}[t]
+\[
+\begin{array}{rrcll}
+\identlsp :&  t + 0 & \iso & t &: \identrsp \\
+\identlst :&  t * 1 & \iso & t &: \identrst \\
+\\
+\absorbl :&~ t * 0 & \iso & 0 ~ &: \factorzr \\
+\distl :&~ t_1 * (t_2 + t_3) & \iso & (t_1 * t_2) + (t_1 * t_3)~ &: \factorl
+\end{array}
+\]
+\caption{Additional $\Pi$-terms.}
+\label{more-pi}
+\end{figure}
+
+We also make a second design decision, which is to make the $\Pi$
+language itself symmetric in another sense: we want both left
+and right introduction/elimination rules for units, $0$ absorption
+and distributivity. Specifically, we add the $\Pi$-terms of
+Fig.~\ref{more-pi} to our language. These are redundant because
+of $\swapp$ and $\swapt$, but will later enable shorter programs
+and more elegant presentation of program transformations.
+
+This set of isomorphisms is known to be sound and
+complete~\cite{Fiore:2004,fiore-remarks} for isomorphisms
+of finite types.  Furthermore, it is also universal
+for hardware combinational
+circuits~\cite{James:2012:IE:2103656.2103667}.
 

### 2018-06-28 survey/survey.tex
start a pass through section 3 to ensure that it is consistent with itself and with section 4.

@@
 \newcommand{\assoc}{\circ}
-\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
-\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
-\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
-\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
-\newcommand{\swapp}{\mathit{swap}_+}
-\newcommand{\assoclp}{\mathit{assocl}_+}
-\newcommand{\assocrp}{\mathit{assocr}_+}
-\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
-\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
-\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
-\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
-\newcommand{\swapt}{\mathit{swap}_*}
-\newcommand{\assoclt}{\mathit{assocl}_*}
-\newcommand{\assocrt}{\mathit{assocr}_*}
-\newcommand{\absorbr}{\mathit{absorbr}}
-\newcommand{\absorbl}{\mathit{absorbl}}
-\newcommand{\factorzr}{\mathit{factorzr}}
-\newcommand{\factorzl}{\mathit{factorzl}}
-\newcommand{\dist}{\mathit{dist}}
-\newcommand{\factor}{\mathit{factor}}
-\newcommand{\distl}{\mathit{distl}}
-\newcommand{\factorl}{\mathit{factorl}}
+\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
+\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
+\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
+\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
+\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
+\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
+\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
+\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
+\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
+\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
+\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
+\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
+\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
+\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
+\newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
+\newcommand{\absorbl}{\ensuremath{\mathit{absorbl}}}
+\newcommand{\factorzr}{\ensuremath{\mathit{factorzr}}}
+\newcommand{\factorzl}{\ensuremath{\mathit{factorzl}}}
+\newcommand{\dist}{\ensuremath{\mathit{dist}}}
+\newcommand{\factor}{\ensuremath{\mathit{factor}}}
+\newcommand{\distl}{\ensuremath{\mathit{distl}}}
+\newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
 %% \newcommand{\distz}{\mathit{absorbr}}
@@
 
-\jc{One operational semantics is explicit, such as the eval and
-evalB we've written many times (for example in PiEqui.agda in
-this repo). But in my opinion, the right way to do it is the
-\texttt{c2equiv} combinator in the same file, which interprets
-every combinator as a type equivalence, which can then just be
-run. \texttt{lemma0} and \texttt{lemma1} in the same file show
-that this is indeed the case. The advantage is that this lifts
-to level 2 quite cleanly.}
-
-\begin{definition}{(Syntax of \ensuremath{\Pi })}
+\jc{The notation here is somewhat inconsistent with the above,
+mainly in the notation for values.}
+
+To give an operational semantics to $\Pi$, we are mainly missing
+a notation for \emph{values}.
+
+\begin{definition}{(Syntax of values of \ensuremath{\Pi })}
 \label{def:langRev}
-We collect our types, values, and combinators, to get the full language
-definition.
 \[\begin{array}{rclr}
- \mathit{value} ~\mathit{types}, b &::=& 0 ~|~ 1 ~|~ b+b ~|~ b\times b \\
  \mathit{values}, v &::=& () ~|~ \mathit{left} ~v ~|~ \mathit{right} ~v ~|~ (v,v) \\
- \\
- \mathit{combinator} ~\mathit{types}, t &::=& b \leftrightarrow b \\
- \mathit{isomorphisms}, \mathit{iso} &::=& \identlp ~|~ \identrp ~|~ \swapp ~|~ \assoclp ~|~ \assocrp \\
- & ~|~ & \identlt ~|~ \identrt ~|~ \swapt ~|~ \assoclt ~|~ \assocrt \\
- & ~|~ & \absorbr ~|~ \factorzl ~|~ \dist ~|~ \factor \\
- \mathit{combinators}, c &::=& \mathit{iso} ~|~ \idc ~|~ \ !~c ~|~ c \odot c ~|~ c \oplus c ~|~ c \otimes c \\
  \end{array}\]
@@
 
-\begin{definition}{Operational Semantics for \ensuremath{\Pi }. }
+Given a program \ensuremath{c : b_1 \leftrightarrow b_2} in \ensuremath{\Pi },
+we can run it by supplying it with a value \ensuremath{ v_1 : b_1}. The
+evaluation rules \ensuremath{c ~v_1 \mapsto v_2} are given below.
+
+\begin{definition}{(Operational Semantics for \ensuremath{\Pi })}
 \label{def:operational-langRev}
-Given a program \ensuremath{c : b_1 \leftrightarrow b_2} in \ensuremath{\Pi }, we can run it by supplying
-it with a value \ensuremath{ v_1 : b_1}. The evaluation rules \ensuremath{c ~v_1 \mapsto v_2} are given below.
 
@@
  \identrp & v &\mapsto & \mathit{right} ~v \\
+ \identlsp & (\mathit{left} ~v) &\mapsto & v \\
+ \identrsp & v &\mapsto & \mathit{left} ~v \\
  \swapp & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
@@
  \end{array}\]
-%subcode source isomorphisms.tex:504
+%originally: subcode source isomorphisms.tex:504
 
@@
  \identrt & v &\mapsto & ((), v) \\
+ \identlst & (v, ()) &\mapsto & v \\
+ \identrst & v &\mapsto & (v, ()) \\
  \swapt & (v_1, v_2) &\mapsto & (v_2, v_1) \\
@@
  \assocrt & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
+ \absorbr & (v_1, v_2) & \mapsto & v_1 \\
  \end{array}\]
-%subcode source isomorphisms.tex:514
+%originally: subcode source isomorphisms.tex:514
 
@@
  \dist & (\mathit{right} ~v_2, v_3) &\mapsto & \mathit{right} ~(v_2, v_3) \\
+ \distl & (v_1, \mathit{left} ~v_2) &\mapsto & \mathit{left} ~(v_1, v_2) \\
+ \distl & (v_1, \mathit{right} ~v_3) &\mapsto & \mathit{right} ~(v_1, v_3) \\
  \factor & (\mathit{left} ~(v_1, v_3)) &\mapsto & (\mathit{left} ~v_1, v_3) \\
  \factor & (\mathit{right} ~(v_2, v_3)) &\mapsto & (\mathit{right} ~v_2, v_3) \\
+ \factorl & (\mathit{left} ~(v_1, v_2)) &\mapsto & (v_1, \mathit{left} ~v_2) \\
+ \factorl & (\mathit{right} ~(v_1, v_3)) &\mapsto & (v_1, \mathit{right} ~v_3) \\
+ \absorbl & (v_1 , v_2) & \mapsto & v_2 \\
  \end{array}\]
-%subcode source isomorphisms.tex:523
+%originally: subcode source isomorphisms.tex:523
 
@@
 	 ~
-}
-\quad
-\infer{ (\ !~c) ~v_1 \mapsto v_2}{
-	 c{\mathit{^{\dagger} }} v_1 \mapsto v_2
 }
@@
 Since there are no values that have the type \ensuremath{0}, the reductions for
-the combinators \ensuremath{\identlp}, \ensuremath{\identrp}, \ensuremath{\absorbr} and \ensuremath{\factorzl}
-omit the impossible cases. Also, the semantics of \ensuremath{\ !~c} are
-defined in terms of the adjoint of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
+the combinators \identlp, \identrp, \identlsp, \identrsp omit the impossible
+cases. \factorzr and \factorzl likewise do not appear as they have not
+possible cases at all. However, \absorbr and \absorbl are treated slightly
+differently: rather than \emph{eagerly} assuming they are impossible,
+the purported inhabitant of $0$ given on one side is passed on to the other
+side. The reason for this choice will have to wait for Sec.~\ref{langeqeq}
+when we explain some higher-level symmetries (see Fig.~\ref{figc}).
+
+\jc{WAVEFRONT of consistency restoration}
+Also, the semantics of \ensuremath{\ !~c} are defined in terms of the adjoint
+of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
 which we formalize below.
@@
 %%%%%%%%%
+\subsection{Denotational Semantics}
+
+\jc{
+\texttt{c2equiv} combinator in the same file, which interprets
+every combinator as a type equivalence, which can then just be
+run. \texttt{lemma0} and \texttt{lemma1} in the same file show
+that this is indeed the case. The advantage is that this lifts
+to level 2 quite cleanly.}
+
+%%%%%%%%%
 \subsection{Examples}
@@
 \subsection{A Language of Equivalences between Type Equivalences}
+\label{langeqeq}
 

### 2018-06-28 survey/survey.tex
work on ensuring that the new material in section 3 is consistent with its surroundings. Introduce various macros (and fix others) to make writing more convenient.

@@
 \newcommand{\inr}[1]{\textsf{inr}~#1}
-\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
 \newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
@@
 \newcommand{\jdg}[3]{#2 \proves_{#1} #3}
+\newcommand{\sem}[1]{\ensuremath{\llbracket{#1}\rrbracket}}
 
@@
 versus $\top$; the first has exactly \emph{two} proofs of inhabitation
-($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one
+($\inl{\texttt{tt}}$ and $\inr{\texttt{tt}}$) while the second only one
 ($\texttt{tt}$). These cannot be put in bijective correspondence. Even
@@
 
+Identity:
+\[\begin{array}{rlcl}
+ \idc & v &\mapsto & v \\
+ \end{array}\]
+
 Additive fragment:
-
 \[\begin{array}{rlcl}
@@
 Multiplicative fragment:
-
 \[\begin{array}{rlcl}
@@
 $$
-\infer{ \idc ~v \mapsto v}{
-	 ~
-}
-\quad
 \infer{ (c_1\odot c_2) ~v_1 \mapsto v_2}{
@@
 
-\jc{WAVEFRONT of consistency restoration}
-Also, the semantics of \ensuremath{\ !~c} are defined in terms of the adjoint
-of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
-which we formalize below.
-
-\begin{definition}[Adjoints, \ensuremath{c^\mathit{^{\dagger} }}]
- The adjoint of a combinator \ensuremath{c} is defined by \ensuremath{c^\mathit{^{\dagger} }} as follows:
+As we mentioned before, $!$ is a defined combinator.
+
+\begin{definition}[Adjoint, \ensuremath{!~ c}]
+ The adjoint of a combinator \ensuremath{c} is defined as follows:
 
   \begin{itemize}
-  \item For primitive isomophisms \ensuremath{c}, \ensuremath{c^\mathit{^{\dagger} }} is given by its
-    isomorphic inverse operator from the table in Definition
-    \ref{def:primitives-langRev}.
-
-  \item \ensuremath{(c_1 \otimes c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \otimes c_2^\mathit{^{\dagger} }}
-
-  \item \ensuremath{(c_1 \oplus c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \oplus c_2^\mathit{^{\dagger} }}
-
-  \item \ensuremath{(c_1\odot c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \odot c_1^\mathit{^{\dagger} }}. (Note that the
+  \item For primitive isomophisms \ensuremath{c}, \ensuremath{!~ c} is given by its
+    inverse from Figures~\ref{pi-terms} and~\ref{more-pi}.
+
+  \item \ensuremath{!(c_1 \otimes c_2) =\ !c_1 \otimes~ !c_2}
+
+  \item \ensuremath{!(c_1 \oplus c_2) =\ !c_1 \oplus~ !c_2}
+
+  \item \ensuremath{!(c_1\odot c_2) =\ !c_2 \odot~ !c_1}. (Note that the
     order of combinators has been reversed).
@@
   \end{itemize}
-
 \end{definition}
 
-\begin{definition}[Observational Equivalence. \ensuremath{c_1=c_2}]
-\label{def:pi-eqv}
-  Combinators \ensuremath{c_1} and \ensuremath{c_2}, where \ensuremath{c_1 : b_1 \leftrightarrow b_2} and
-  \ensuremath{c_2:b_1\leftrightarrow b_2}, are observationally equivalent,\ensuremath{c_1=c_2}, iff
-
-  \ensuremath{\forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \Longleftrightarrow c_2 ~v_1 \mapsto v_2}.
+\noindent We can further define that two combinators are 
+\emph{observationally equivalent} if only all values of their
+common domain, they evaluate to identical values.  More precisely,
+we will say that for combinators $c_1, c_2 : b_1 \leftrightarrow b_2$,
+$c_1~=~c_2$ whenever
+\[
+  \forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \text{if and only if\ } c_2 ~v_1 \mapsto v_2
+\]
+
+Previously we had seen that types $b$ correspond to natural numbers,
+more precisely~$|b|$, the size of $b$.  We had previously established that 
+for any natural number $n$, there is a canonical set of size $n$,
+which we denoted $[n]$. Furthermore, we can also define a
+canonical \emph{type} that size, which we will denote $\sharp b$,
+i.e. $\sharp b$ is a canonical type of size $|b|$.
+
+\begin{definition}($\sharp$). By recursion on $|b|$.  First define a
+type associated to each size. We will revert to using type notation
+for greater clarity of this definition:
+\[\begin{array}{rcl}
+  \tau~ 0 & = & \bot \\
+  \tau~ (1 + n) & = & \top \uplus \tau~ n \\
+ \end{array}\]
+\noindent so that we can define $\sharp b = \tau~ |b|$.
 \end{definition}
 
-\begin{proposition}{Canonical Isomorphism.}
-  For any type \ensuremath{b} there exists the isomorphism \ensuremath{b \leftrightarrow b^{|b|}}.
+We are now ready to go further and establish
+that there is always an equivalence between.
+
+\begin{proposition}
+  For any type \ensuremath{b} there exists an isomorphism \ensuremath{b \leftrightarrow \sharp b}.
   \begin{proof}
@@
     isomorphic. While many equivalent constructions are possible
-    for any type \ensuremath{b}, one such construction is given by \ensuremath{\lbrack\!\lbrack b\rbrack\!\rbrack}:
+    for any type \ensuremath{b}, one such construction is given by 
+    \sem{b}:
 
 \[\begin{array}{rclr}
- \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \idc \\
- \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \idc \\
- \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \idc \oplus \lbrack\!\lbrack b \rbrack\!\rbrack \\
- \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \assocrp \odot \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
- \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack \oplus \idc) \odot \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
- \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \assocrt \odot \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
- \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \dist \odot \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
+ \sem{0} & = & \idc \\
+ \sem{1} & = & \idc \\
+ \sem{1+b} & = & \idc \oplus \sem{b} \\
+ \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
+ \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp b_1 + b_2 } \\
+ \sem{0 \times b_2} & = & \absorbr \\
+ \sem{1 \times b_2} & = & \identlt \odot \sem{b_2} \\
+ \sem{(b_1\times b_2)\times b_3} & = & \assocrt \odot \sem{b_1\times (b_2\times b_3)} \\
+ \sem{(b_1+b_2)\times b_3} & = & \dist \odot \sem{b_1\times b_3+b_2\times b_3} \\
  \end{array}\]
-%subcode source isomorphisms.tex:598
+%originally subcode source isomorphisms.tex:598
 
@@
 Combinators of \ensuremath{\Pi } can be written in terms of the operators
-described in definition \ref{def:langRev} or via a graphical language
+described previously or via a graphical language
 similar in spirit to those developed for Geometry of Interaction
@@
 \ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
-associativity in the graphical representations of categories.
+associativity in the graphical representations of categories as well as
+that for monoidal categories.
 
@@
 
-\item
-Commutativity is represented by crisscrossing wires.
+\item Commutativity is represented by crisscrossing wires.
 
@@
 \noindent
-When tracing the execution of \ensuremath{b_1+b_2} represented by one
-wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
-as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
+When tracing the execution of \ensuremath{b_1+b_2} represented by one wire, one
+can think of a value of the form \ensuremath{\mathit{left} ~v_1} or
+\ensuremath{\mathit{right} ~v_2} as flowing on the wire, where
+\ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
 tracking the flow of particles on the wires, one can verify that the
@@
   is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
-  Also since the monoidal units can be freely introduced and eliminated, in
+  Also since the monoidal units can be freely introduced and eliminated, sometimes
+  they are omitted.  However, as this is in fact dangerous, as explained by
   many diagrams they are omitted and dealt with explicitly only when they are
@@
 \ensuremath{c : b \times (1+1) \leftrightarrow b + b}
-
 \ensuremath{c = \swapt \odot \dist \odot (\identlt + \identlt)}
-
 \begin{center}
@@
 
+\jc{WAVEFRONT of consistency pass}
 The textual representation of $\Pi$ is quite cumbersome: the
@@
 
+\begin{comment}
 \jc{An additional ``problem'' is that since our circuits have a
@@
 components in the plane.
-
+\end{comment}
 % \amr{

### 2018-06-29 survey/survey.tex
consistency up to 3.4. Write 3.5.

@@
 \subsection{Operational Semantics}
+\label{sec:opsem}
 
@@
 \cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
-\cite{selinger-graphical}. Modulo some conventions and shorthand we
+\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions and shorthand we
 describe here, the wiring diagrams are equivalent to the operator
@@
 of the combinator \ensuremath{c} below:
-
-\ensuremath{c : b \times (1+1) \leftrightarrow b + b}
-\ensuremath{c = \swapt \odot \dist \odot (\identlt + \identlt)}
+\begin{flalign*}
+& \begin{aligned}
+c & : & b \times (1+1) \leftrightarrow b + b \\
+c & = & \swapt \odot \dist \odot (\identlt + \identlt) \\
+\end{aligned} &&
+\end{flalign*}
+
 \begin{center}
@@
 \end{center}
-
-\jc{WAVEFRONT of consistency pass}
-The textual representation of $\Pi$ is quite cumbersome: the
-underlying principles however become quite intuitive when represented
-in a more suggestive visual notation. There exist various graphical
-notations for the monoidal categories that model $\Pi$ and related
-languages including Penrose diagrams~\cite{selinger-graphical} and
-Geometry of Interaction (GoI) machines and Proof
-Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In this section, we
-propose a novel graphical representation of $\Pi$ types and
-combinators drawn from the resource usage embedded in the operational
-semantics. As suggested by the operational semantics, the type $A * B$
-can be viewed as denoting \emph{spatial composition}, where
-\emph{both} $A$ and $B$ must happen in \emph{some} order or perhaps
-simultaneously, and where $A + B$ denotes \emph{temporal composition},
-where exactly \emph{one of} $A$ and $B$ must happen at a given time.
-\jc{I think the temporal aspect is stretching things too far. Could we
-call it ``choice composition''? (Or something better). To me spatial
-composition somehow seems like sequential composition (i.e. composition
-in time), which is definitely not what we want. I think the informal
-description (both although perhaps concurrently, and exactly one of)
-is better. In section 3 we called them sequential, choice and
-parallel. We should be consistent.}
 
@@
 
-\jc{
-\texttt{c2equiv} combinator in the same file, which interprets
-every combinator as a type equivalence, which can then just be
-run. \texttt{lemma0} and \texttt{lemma1} in the same file show
-that this is indeed the case. The advantage is that this lifts
-to level 2 quite cleanly.}
+Figure~\ref{type-isos} introduces our desired denotational
+semantices, and section~\ref{sec:opsem} a direct definition
+of an operational semantics. One obvious question arises:
+do these correspond?
+
+We can certainly associate to each $\Pi$ combinator an
+equivalence between the denotation of each type%
+\footnote{This is extracted from the Agda formalization
+of this work, which has been reported on in~\cite{Carette2016}.}:
+\begin{code}\hspace*{-4mm}
+\>\AgdaFunction{c2equiv} \AgdaSymbol{:} \AgdaSymbol{\{}\AgdaBound{t₁}
+\AgdaBound{t₂} \AgdaSymbol{:} \AgdaDatatype{U}\AgdaSymbol{\}} \AgdaSymbol{→}
+\AgdaSymbol{(}\AgdaBound{c} \AgdaSymbol{:} \AgdaBound{t₁} \AgdaDatatype{⟷}
+\AgdaBound{t₂}\AgdaSymbol{)} \AgdaSymbol{→} \AgdaFunction{⟦} \AgdaBound{t₁}
+\AgdaFunction{⟧} \AgdaFunction{≃} \AgdaFunction{⟦} \AgdaBound{t₂}
+\AgdaFunction{⟧}%
+\end{code}
+
+\noindent And as such an equivalence contains a function as
+its first component, we can compare if our operational
+semantics and denotational semantics match.  And they do:
+\begin{code}\hspace*{-4mm}
+\>\AgdaFunction{lemma0} \AgdaSymbol{:} \<[10]%
+\>[10]\AgdaSymbol{\{}\AgdaBound{t₁} \AgdaBound{t₂} \AgdaSymbol{:}
+\AgdaDatatype{U}\AgdaSymbol{\}} \AgdaSymbol{→} \AgdaSymbol{(}\AgdaBound{c}
+\AgdaSymbol{:} \AgdaBound{t₁} \AgdaDatatype{⟷} \AgdaBound{t₂}\AgdaSymbol{)}
+\AgdaSymbol{→} \AgdaSymbol{(}\AgdaBound{v} \AgdaSymbol{:} \AgdaFunction{⟦}
+\AgdaBound{t₁} \AgdaFunction{⟧}\AgdaSymbol{)} \AgdaSymbol{→}
+\AgdaFunction{eval} \AgdaBound{c} \AgdaBound{v} \AgdaDatatype{≡}
+\AgdaField{proj₁} \AgdaSymbol{(}\AgdaFunction{c2equiv}
+\AgdaBound{c}\AgdaSymbol{)} \AgdaBound{v}\<%
+\end{code}
+
+\noindent We can similarly hand-write a backwards evaluator,
+prove that it is indeed a proper backwards evaluator, and
+finally show that it agrees with the reverse equivalence.
 

### 2018-07-02 survey/survey.tex
make first diagram in section 4 consistent with those in section 3.  Tikz is not too bad for doing this.
more consistency in the identifiers in the figures.

@@
 
+% Unicode declarations
+
 \DeclareUnicodeCharacter{9678}{\ensuremath{\odot}}
@@
 \DeclareUnicodeCharacter{10231}{\ensuremath{\leftrightarrow}}
+
+% TIKZ declarations
+\tikzstyle{func}=[rectangle,draw,fill=black!20,minimum size=1.9em,
+  text width=2.4em, text centered]
+
+% Parts of the document itself
 
@@
 
-\jc{we need to make sure that the pictures we draw in the
-previous section are consistent with this one, otherwise we'll
-completely confuse our readers. Or we redraw this one to be
-consistent.}
 \begin{center}
@@
 %%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\odot~(c_1~\oplus~c_2))$};
-  \draw (-2,-2) ellipse (0.5cm and 1cm);
+%  \draw (-2,-2) ellipse (0.5cm and 1cm);
   \draw[fill] (-2,-1.5) circle [radius=0.025];
   \node[below] at (-2.1,-1.5) {$A$};
+  \node[below] at (-2.1,-1.9) {$+$};
   \draw[fill] (-2,-2.5) circle [radius=0.025];
@@
 
-  \draw (6.5,-2) ellipse (0.5cm and 1cm);
+%  \draw (6.5,-2) ellipse (0.5cm and 1cm);
   \draw[fill] (6.5,-1.5) circle [radius=0.025];
   \node[below] at (6.7,-1.5) {$C$};
+  \node[below] at (6.7,-1.9) {$+$};
   \draw[fill] (6.5,-2.5) circle [radius=0.025];
@@
 
-  \draw     (2.5,-3.1)  -- (3.5,-3.1) -- (3.5,-3.9) -- (2.5,-3.9) -- cycle ;
-  \draw     (2.5,-4.1)  -- (3.5,-4.1) -- (3.5,-4.9) -- (2.5,-4.9) -- cycle ;
-
-  \draw     (1,0.9)  -- (2,0.9) -- (2,0.1) -- (1,0.1) -- cycle ;
-  \draw     (1,-0.1)  -- (2,-0.1) -- (2,-0.9) -- (1,-0.9) -- cycle ;
-
-  \node at (3,-3.5) {$c_2$};
-  \node at (3,-4.5) {$c_1$};
-
-  \node at (1.5,0.5) {$c_1$};
-  \node at (1.5,-0.5) {$c_2$};
-
   \draw     (2,0.5)  -- (2.5,0.5)  ;
@@
   \draw     (2,-4.5) -- (2.5,-4.5)   ;
+
+  \path (1.5,0.5) node (tc1) [func] {$c_1$};
+  %\draw     (1,0.9)  -- (2,0.9) -- (2,0.1) -- (1,0.1) -- cycle ;
+  %\node at (1.5,0.5) {$c_1$};
+
+  \path (1.5,-0.5) node (tc2) [func] {$c_2$};
+  %\draw     (1,-0.1)  -- (2,-0.1) -- (2,-0.9) -- (1,-0.9) -- cycle ;
+  %\node at (1.5,-0.5) {$c_2$};
+
+  \path (3,-4.5) node (bc1) [func] {$c_1$};
+  %\draw     (2.5,-4.1)  -- (3.5,-4.1) -- (3.5,-4.9) -- (2.5,-4.9) -- cycle ;
+  %\node at (3,-4.5) {$c_1$};
+
+  \path (3,-3.5) node (bc2) [func] {$c_2$};
+  %\draw     (2.5,-3.1)  -- (3.5,-3.1) -- (3.5,-3.9) -- (2.5,-3.9) -- cycle ;
+  % \node at (3,-3.5) {$c_2$};
 
@@
 The last two equivalences show how composition of associativity combinators
-interact together. For example, the last equivalence captures the idea that the
+interact together. 
+\begin{comment}
+For example, the last equivalence captures the idea that the
 program $\assocrt \odot \assocrt$ transforms an input of
@@
 
+\jc{I am not a big fan of these pictures -- mainly because I can't give any
+intuition based on them of why these can be smoothly deformed into one another.
+All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
+5 dimensions -- one for each of the wires, and another to give room to maneuver.}
+\end{comment}
+
+\jc{need to find some combinators whose diagrams can help illustrate 
+the ideas further. Enough so that the next paragraph makes sense.}
+
 \noindent This of course gives rise to a \emph{critical pair} in words over the
@@
 
-\jc{I am not a big fan of these pictures -- mainly because I can't give any
-intuition based on them of why these can be smoothly deformed into one another.
-All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
-5 dimensions -- one for each of the wires, and another to give room to maneuver.}
-
 \begin{figure}[t]
-Let $a : t_1 \leftrightarrow t_2$, $b : t_3 \leftrightarrow t_4$, and $c : t_5 \leftrightarrow t_6$:
+Let $c_1 : t_1 \leftrightarrow t_2$, $c_2 : t_3 \leftrightarrow t_4$, and $c_3 : t_5 \leftrightarrow t_6$:
 \[\def\arraystretch{1.3}
 \begin{array}{c}
-  {((a \oplus b) \otimes c) \odot \dist \Leftrightarrow \dist \odot ((a \otimes c) \oplus (b \otimes c))}
-\\
-  {(a \otimes (b \oplus c)) \odot \distl \Leftrightarrow \distl \odot ((a \otimes b) \oplus (a \otimes c))}
-\\
-  {((a \otimes c) \oplus (b \otimes c)) \odot \factor \Leftrightarrow \factor \odot ((a \oplus b) \otimes c)}
-\\
-  {((a \otimes b) \oplus (a \otimes c)) \odot \factorl \Leftrightarrow \factorl \odot (a \otimes (b \oplus c))}
+  {((c_1 \oplus c_2) \otimes c_3) \odot \dist \Leftrightarrow \dist \odot ((c_1 \otimes c_3) \oplus (c_2 \otimes c_3))}
+\\
+  {(c_1 \otimes (c_2 \oplus c_3)) \odot \distl \Leftrightarrow \distl \odot ((c_1 \otimes c_2) \oplus (c_1 \otimes c_3))}
+\\
+  {((c_1 \otimes c_3) \oplus (c_2 \otimes c_3)) \odot \factor \Leftrightarrow \factor \odot ((c_1 \oplus c_2) \otimes c_3)}
+\\
+  {((c_1 \otimes c_2) \oplus (c_1 \otimes c_3)) \odot \factorl \Leftrightarrow \factorl \odot (c_1 \otimes (c_2 \oplus c_3))}
 \end{array}\]
@@
 \begin{figure}[t]
-Let $c, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c', c'' : t_3 \leftrightarrow t_4$:
+Let $c_0, c_1, c_2, c_3 : t_1 \leftrightarrow t_2$ and $c_4, c_5 : t_3 \leftrightarrow t_4$:
 \[\def\arraystretch{1.3}
 \begin{array}{c}
-  {\idc \odot \, c \Leftrightarrow c}
+  {\idc \odot \, c_0 \Leftrightarrow c_0}
 \quad
-  {c \, \odot \idc \, \Leftrightarrow c}
+  {c_0 \, \odot \idc \, \Leftrightarrow c_0}
 \quad
-  {c\,\, \odot\,!\, c \Leftrightarrow \idc}
+  {c_0\,\, \odot\,!\, c_0 \Leftrightarrow \idc}
 \quad
-  {!\, c \odot c \Leftrightarrow \idc}
+  {!\, c_0 \odot c_0 \Leftrightarrow \idc}
 \\
@@
 \\
-  {c \Leftrightarrow c}
+  {c_0 \Leftrightarrow c_0}
 \quad
@@
 \Rule{}
-  {c_1 \Leftrightarrow c' \quad c_2 \Leftrightarrow c''}
-  {c_1 \odot c_2 \Leftrightarrow c' \odot c''}
+  {c_1 \Leftrightarrow c_4 \quad c_2 \Leftrightarrow c_5}
+  {c_1 \odot c_2 \Leftrightarrow c_4 \odot c_5}
   {}
@@
 \begin{figure}[t]
-Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c : t_1 \leftrightarrow t_2$:
+Let $c_0 : 0 \leftrightarrow 0$, $c_1 : 1 \leftrightarrow 1$, and $c_3 : t_1 \leftrightarrow t_2$:
 \[\def\arraystretch{1.3}
 \begin{array}{c}
-  {\identlp \odot c \Leftrightarrow (c_0 \oplus c) \odot \identlp}
+  {\identlp \odot c_3 \Leftrightarrow (c_0 \oplus c_3) \odot \identlp}
 \qquad
-  {\identrp \odot (c_0 \oplus c) \Leftrightarrow c \odot \identrp}
-\\
-  {\identlsp \odot c \Leftrightarrow (c \oplus c_0) \odot \identlsp}
+  {\identrp \odot (c_0 \oplus c_3) \Leftrightarrow c_3 \odot \identrp}
+\\
+  {\identlsp \odot c_3 \Leftrightarrow (c_3 \oplus c_0) \odot \identlsp}
 \qquad
-  {\identrsp \odot (c \oplus c_0) \Leftrightarrow c \odot \identrsp}
-\\
-  {\identlt \odot c \Leftrightarrow (c_1 \otimes c) \odot \identlt}
+  {\identrsp \odot (c_3 \oplus c_0) \Leftrightarrow c_3 \odot \identrsp}
+\\
+  {\identlt \odot c_3 \Leftrightarrow (c_1 \otimes c_3) \odot \identlt}
 \qquad
-  {\identrt \odot (c_1 \otimes c) \Leftrightarrow c \odot \identrp}
-\\
-  {\identlst \odot c \Leftrightarrow (c \otimes c_1) \odot \identlst}
+  {\identrt \odot (c_1 \otimes c_3) \Leftrightarrow c_3 \odot \identrp}
+\\
+  {\identlst \odot c_3 \Leftrightarrow (c_3 \otimes c_1) \odot \identlst}
 \qquad
-  {\identrst \odot (c \otimes c_1) \Leftrightarrow c \odot \identrst}
+  {\identrst \odot (c_3 \otimes c_1) \Leftrightarrow c_3 \odot \identrst}
 \\

### 2018-07-02 survey/survey.tex
some consistency changes in section 3.

@@
 of a value \ensuremath{v:b} based on the value of another one of type
-\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be \ensuremath{\mathit{bool}} and apply either
-\ensuremath{c_1:b\leftrightarrow b} or \ensuremath{c_2:b\leftrightarrow b} to the value \ensuremath{v}, then we essentially
-have an `if' expression.
-
-\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool} \times b \leftrightarrow \mathit{bool} \times b}
+\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be
+\ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
+b_2} or \ensuremath{c_2:b_1\leftrightarrow b_2} to the value \ensuremath{v},
+then we essentially have an `if' expression.
+
+\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool} \times b_1 \leftrightarrow \mathit{bool} \times b_2}
 
@@
 
-The diagram above shows the input value of type \ensuremath{(1+1)\times b} processed
-by the distribute operator \ensuremath{\dist}, which converts it into a value of
-type \ensuremath{(1\times b)+(1\times b)}. In the \ensuremath{\mathit{left}} branch, which corresponds to the
-case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was \ensuremath{\mathit{left} ~()}),
-the combinator~\ensuremath{c_1} is applied to the value of type~\ensuremath{b}. The right
+The diagram above shows the input value of type \ensuremath{(1+1)\times b_2}
+processed by the distribute operator \ensuremath{\dist}, which converts it into
+a value of type \ensuremath{(1\times b_2)+(1\times b_2)}. In the
+\ensuremath{\mathit{left}} branch, which corresponds to the
+case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
+\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
+the value of type~\ensuremath{b_1}. The right
 branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
-the value of type \ensuremath{b} through the combinator \ensuremath{c_2}.
+the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.
 
@@
 are equivalent, with the second saying that the reverse (i.e.
-the results of applying $!$\~) also gives equivalent programs.
+the results of applying $!$\,) also gives equivalent programs.
 The last two say the same but for the multiplicative structure.

### 2018-07-03 survey/survey.tex
comment out the graphical language in section 4. re-line up the example, to make it more readable.

@@
 
+We can do for the level-2 combinators the same as was done for level-1:
+create a programming language with these combinators as fundamental
+constants, typed by level-1 combinators. But we can do more, as these
+new combinators are exactly rewrites for level-1 programs.  Thus,
+it is possible to create
+
 %%%%%%%%%
-\subsection{Graphical Language}
-
-Idea: thinking in the 3D way; we have collections of towers of
-blocks. Every 1-combinator involves touching some blocks, swapping
-them around, etc. Two 1-combinators are equivalent if they touch the
-same blocks the same number of times. ??
+% \subsection{Graphical Language}
+% 
+% Idea: thinking in the 3D way; we have collections of towers of
+% blocks. Every 1-combinator involves touching some blocks, swapping
+% them around, etc. Two 1-combinators are equivalent if they touch the
+% same blocks the same number of times. ??
 
@@
 
+\renewcommand{\AgdaIndentSpace}{\;\;}
+\setlength\mathindent{0.5em}
+
+\begin{samepage}
 \begin{code}%
@@
 \AgdaSymbol{\{}\AgdaBound{c}\AgdaSymbol{\}}\AgdaSpace{}%
-\AgdaDatatype{⇔}\AgdaSpace{}%
+\AgdaOperator{\AgdaDatatype{⇔}}\AgdaSpace{}%
 \AgdaFunction{swap{-}fl1}\<%
@@
 \>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
 \AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
 \AgdaInductiveConstructor{id⟷}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
-\\
-\>[0][@{}l@{\AgdaIndent{0}}]%
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
 \AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊡}}\AgdaSpace{}%
 \AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
+\\
+%
 \>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
 \AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
 \AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
 \\
 %
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
+\>[2]\AgdaSymbol{(((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
 \AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
-\>[2]\AgdaSymbol{(((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊡}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
+\\
+%
+\>[2]\AgdaSymbol{((((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
 \AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
+\AgdaInductiveConstructor{hexagonl⊕r}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊡}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
+\\
+%
+\>[2]\AgdaSymbol{(((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}%
+\>[44]\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎r}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
+\\
+%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}%
+\>[42]\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊡}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
 \\
 %
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
-\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocr₊}\AgdaSymbol{)}%
+\>[44]\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
 \AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
-\>[2]\AgdaSymbol{((((}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊡}}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{linv◎l}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊡}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
+\\
+%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocl₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⟷}\AgdaSymbol{))}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}%
+\>[28]\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊡}}\AgdaSpace{}%
+\AgdaInductiveConstructor{idl◎l}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
 \\
 %
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
-\AgdaInductiveConstructor{hexagonl⊕r}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
-\>[2]\AgdaSymbol{(((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocl₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocr₊}%
-\>[44]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}%
+\>[22]\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}%
+\>[75]\AgdaOperator{\AgdaFunction{⇔⟨}}\AgdaSpace{}%
+\AgdaInductiveConstructor{assoc◎r}\AgdaSpace{}%
+\AgdaOperator{\AgdaFunction{⟩}}\<%
 \\
 %
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
-\AgdaInductiveConstructor{assoc◎r}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
 \>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocr₊}%
-\>[42]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{◎}}\AgdaSpace{}%
 \AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
-\\
-%
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
-\AgdaInductiveConstructor{assoc◎l}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
-\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocr₊}\AgdaSymbol{)}%
-\>[44]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
-\\
-%
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{linv◎l}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⇔}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
-\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⟷}%
-\>[28]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
-\\
-%
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⇔}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊡}\AgdaSpace{}%
-\AgdaInductiveConstructor{idl◎l}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
-\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}%
-\>[22]\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\<%
-\\
-%
-\>[1]\AgdaFunction{⇔⟨}\AgdaSpace{}%
-\AgdaInductiveConstructor{assoc◎r}\AgdaSpace{}%
-\AgdaFunction{⟩}\<%
-\\
-\>[1][@{}l@{\AgdaIndent{0}}]%
-\>[2]\AgdaSymbol{((}\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaOperator{\AgdaInductiveConstructor{⊕}}\AgdaSpace{}%
 \AgdaInductiveConstructor{swap₊}\AgdaSymbol{))}\AgdaSpace{}%
-\AgdaFunction{▤}\AgdaSymbol{)}\<%
+\AgdaOperator{\AgdaFunction{▤}}\AgdaSymbol{)}\<%
 \end{code}
+\end{samepage}
+
+\renewcommand{\AgdaIndentSpace}{\AgdaSpace{}$\;\;$}
 

### 2018-07-03 survey/survey.tex
add various comments which act as a set of TODOs.

@@
 interact together.
+
+\jc{Say that the bottom line in Fig.~\ref{figj} is actually the
+pentagon diagram of monoidal categories. Draw it, and label it
+using our combinators instead of the usual ones.}
+
 \begin{comment}
@@
 
-\jc{need to find some combinators whose diagrams can help illustrate 
-the ideas further. Enough so that the next paragraph makes sense.}
-
-\noindent This of course gives rise to a \emph{critical pair} in words over the
-level-1 language, as both of these programs are between the same types.
-Since neither of these involve either trivial types (like $0$ or $1$) nor
-twists (i.e. automorphisms) on the types themselves, coherence requires that
-these in fact be equivalent programs.  One interpretation is that
-\emph{outermost} re-associations is the same as \emph{innermost}
-re-associations.
-
 \begin{figure}[t]
@@
 
-We can do for the level-2 combinators the same as was done for level-1:
-create a programming language with these combinators as fundamental
-constants, typed by level-1 combinators. But we can do more, as these
-new combinators are exactly rewrites for level-1 programs.  Thus,
-it is possible to create
+\jc{Maybe interpret the next level as simply term rewriting?
+We already know the start and end, but if we also have a whole
+sequence of `trans', then we can reconstruct the whole thing
+from that. In other words, the example below could be
+reconstructed from the start, end and the program itself,
+erasing all of the lhs annotations, which are merely there for
+humans to track.}
 
@@
 \renewcommand{\AgdaIndentSpace}{\AgdaSpace{}$\;\;$}
+
+\jc{The above is, up to uses of assoc, a linear notation for a
+commutative diagram. So draw it.}
 

### 2018-07-03 survey/survey.tex
actually draw said pentagon diagram. tikzcd is rather nice.

@@
 
-\jc{Say that the bottom line in Fig.~\ref{figj} is actually the
-pentagon diagram of monoidal categories. Draw it, and label it
-using our combinators instead of the usual ones.}
+The bottom line in Fig.~\ref{figj} is actually a linear
+restatement of the famous ``pentagon diagram'' stating a
+particular coherence condition for monoidal categories~\cite{KELLY197197}.
+To make the relation between $\Pi$ as a language and the
+language of category theory, the figure below
+the same morphism but in categorical terms.
+
+\begin{center}
+\begin{tikzcd}[column sep=normal]
+   & (A \times (B \times C)) \times D \arrow [dr, "\assocrt"] & \\
+((A \times B) \times C) \times D \arrow [ur, "\assocrt \otimes \mathit{id}\leftrightarrow"] 
+   \arrow [d, "\assocrt"] & 
+       & A \times ((B \times C) \times D) \arrow [d, "\mathit{id}\leftrightarrow \otimes \assocrt" ]\\
+(A \times B) \times (C \times D) \arrow [rr, "\assocrt"] & & A \times (B \times (C \times D))
+\end{tikzcd}
+\end{center}
 

### 2018-07-04 survey/survey.tex
paragraphs on related work and history of pi

@@
 \begin{quote}
-  \textbf{Toffoli 1980~\cite{toffoli:1980}:} Mathematical models of
+  \textbf{Toffoli 1980~\cite{Toffoli:1980}:} Mathematical models of
   computation are abstract constructions, by their nature unfettered
@@
 
-  \textbf{Feynman 1982~\cite{springerlink:10.1007/bf02650179}:}
+  \textbf{Feynman 1982~\cite{springerlink:10.1007/BF02650179}:}
   Another thing that has been suggested early was that natural laws
@@
 that ideas in each field have consequences for the
-other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,baez2011physics,dblp:conf/csfw/malacarias12}. To
+other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,Baker:1992:NFT,baez2011physics,dblp:conf/csfw/malacarias12}. To
 really appreciate this fact, we delve deeper into the origin of our
@@
 
-\amr{We should also have a quick survey of other works in reversible
-computation? Most of the other work tries to start from well known
-models of computation or well known programming languages, and then
-adapts them to be reversible. Pi is different in that it starts from
-the semantic implications of reversibility. Then, because it adds
-types, rather than control-flow (or even data-flow) as its next layer
-of ``understanding,'' this leads to equivalences, isomorphisms, etc.
-This then leads, quite naturally, to finding that the ``proof
-language'' of semirings (and Rig Groupoids at level 2) is actually a
-programming language. And it is Pi. This is a neat twist on
-Curry-Howard because CH is about \textbf{inhabitation} only. But with
-``conservation of information'' as the basis, a different kind of
-correspondence arises; in fact, this one may well be an actual
-isomorphism.}
-
-\jc{I think it would be fair to have a \textit{quick} survey
-of other works. The most important aspect would be to draw out that
-they all start from known models of computation and then attempt to
-make them reversible.  Our originality is to start from the semantics
-of reversibility (even further back, preservation of information)
-and derive a programming language purposefully designed to denote
-exactly that. Having said that, I am not in a good position to
-write such a survey, being the least knowledgeable of all 3 of us
-of this general area.}
+\paragraph*{Reversible Programming Languages.} The practice of
+programming languages is replete with \emph{ad hoc} instances of
+reversible computations: database transactions, mechanisms for data
+provenance, checkpoints, stack and exception traces, logs, backups,
+rollback recoveries, version control systems, reverse engineering,
+software transactional memories, continuations, backtracking search,
+and multiple-level ``undo'' features in commercial applications. In
+the early nineties, Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued
+for a systematic, first-class, treatment of reversibility. But
+intensive research in full-fledged reversible models of computations
+and reversible programming languages was only sparked by the discovery
+of deep connections between physics and
+computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
+and by the potential for efficient quantum
+computation~\cite{springerlink:10.1007/BF02650179}.
+
+The early developments of reversible programming languages started
+with a conventional programming language, e.g., an extended
+$\lambda$-calculus, and either
+\begin{enumerate}
+\item extended the language with a history
+mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
+\item imposed constraints on the control flow constructs to make them
+reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
+\end{enumerate}
+More modern approaches recognize that reversible programming languages require
+a fresh approach and should be designed from first principles without the
+detour via conventional irreversible
+languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.
+
+In previous work, Carette, Bowman, James, and
+Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016}
+introduced the~$\Pi$ family of reversible languages based on type
+isomorphisms and commutative semiring identities. As motivated above,
+the starting point for this development is the physical principle of
+``conservation of
+information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative} and
+the family of languages is designed to embrace that principle by
+requiring all computations to preserve information.
+
+The fragment without
+recursive types is universal for reversible boolean
+circuits~\cite{James:2012:IE:2103656.2103667} and the extension with
+recursive types and trace
+operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete
+reversible language~\cite{James:2012:IE:2103656.2103667,rc2011}. While
+at first sight, $\Pi$ might appear \emph{ad hoc}, it really arises
+naturally from an ``extended'' view of the Curry-Howard
+correspondence~\cite{Carette2016}: rather than looking at mere
+\emph{inhabitation} as the main source of analogy between logic and
+computation, \emph{type equivalences} becomes the source of analogy.
+This allows one to see an analogy between algebra and reversible
+computation.  Furthermore, this works at multiple levels: that of
+$1$-algebra (types form a semiring under isomorphism), but also
+$2$-algebra (types and equivalences form a weak Rig Groupoid).  In
+other words, by taking ``weak Rig Groupoid'' as the starting
+semantics, one naturally gets $\Pi$ as the syntax for the language of
+proofs of isomorphisms -- in the same way that many terms of the
+$\lambda$-calculus arise from Cartesian Closed Categories.
+
+% \amr{We should also have a quick survey of other works in reversible
+% computation? Most of the other work tries to start from well known
+% models of computation or well known programming languages, and then
+% adapts them to be reversible. Pi is different in that it starts from
+% the semantic implications of reversibility. Then, because it adds
+% types, rather than control-flow (or even data-flow) as its next layer
+% of ``understanding,'' this leads to equivalences, isomorphisms, etc.
+% This then leads, quite naturally, to finding that the ``proof
+% language'' of semirings (and Rig Groupoids at level 2) is actually a
+% programming language. And it is Pi. This is a neat twist on
+% Curry-Howard because CH is about \textbf{inhabitation} only. But with
+% ``conservation of information'' as the basis, a different kind of
+% correspondence arises; in fact, this one may well be an actual
+% isomorphism.}
+
+% \jc{I think it would be fair to have a \textit{quick} survey
+% of other works. The most important aspect would be to draw out that
+% they all start from known models of computation and then attempt to
+% make them reversible.  Our originality is to start from the semantics
+% of reversibility (even further back, preservation of information)
+% and derive a programming language purposefully designed to denote
+% exactly that. Having said that, I am not in a good position to
+% write such a survey, being the least knowledgeable of all 3 of us
+% of this general area.}
 
@@
 of reversible hardware circuits, the canonical universal primitive is
-the Toffoli gate~\cite{toffoli:1980}. The Toffoli gate takes three
+the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
 boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
@@
 second argument if the first argument is \ensuremath{\mathit{true}}. This gate
-\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{toffoli:1980}.
+\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.
 

### 2018-07-04 survey/survey.tex
Rewrote parts of the intro. Decided not to move the text, as most of it could be kept as is without giving away the complete punchline. Though some of it might still bear repeating in the conclusion.

@@
 Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016}
-introduced the~$\Pi$ family of reversible languages based on type
-isomorphisms and commutative semiring identities. As motivated above,
+introduced the~$\Pi$ family of typed reversible languages.  As motivated above,
 the starting point for this development is the physical principle of
@@
 reversible language~\cite{James:2012:IE:2103656.2103667,rc2011}. While
-at first sight, $\Pi$ might appear \emph{ad hoc}, it really arises
+at first sight, $\Pi$ too might appear \emph{ad hoc}, it really arises
 naturally from an ``extended'' view of the Curry-Howard
@@
 computation, \emph{type equivalences} becomes the source of analogy.
-This allows one to see an analogy between algebra and reversible
-computation.  Furthermore, this works at multiple levels: that of
-$1$-algebra (types form a semiring under isomorphism), but also
-$2$-algebra (types and equivalences form a weak Rig Groupoid).  In
-other words, by taking ``weak Rig Groupoid'' as the starting
-semantics, one naturally gets $\Pi$ as the syntax for the language of
-proofs of isomorphisms -- in the same way that many terms of the
-$\lambda$-calculus arise from Cartesian Closed Categories.
-
-% \amr{We should also have a quick survey of other works in reversible
-% computation? Most of the other work tries to start from well known
-% models of computation or well known programming languages, and then
-% adapts them to be reversible. Pi is different in that it starts from
-% the semantic implications of reversibility. Then, because it adds
-% types, rather than control-flow (or even data-flow) as its next layer
-% of ``understanding,'' this leads to equivalences, isomorphisms, etc.
-% This then leads, quite naturally, to finding that the ``proof
-% language'' of semirings (and Rig Groupoids at level 2) is actually a
-% programming language. And it is Pi. This is a neat twist on
-% Curry-Howard because CH is about \textbf{inhabitation} only. But with
-% ``conservation of information'' as the basis, a different kind of
-% correspondence arises; in fact, this one may well be an actual
-% isomorphism.}
-
-% \jc{I think it would be fair to have a \textit{quick} survey
-% of other works. The most important aspect would be to draw out that
-% they all start from known models of computation and then attempt to
-% make them reversible.  Our originality is to start from the semantics
-% of reversibility (even further back, preservation of information)
-% and derive a programming language purposefully designed to denote
-% exactly that. Having said that, I am not in a good position to
-% write such a survey, being the least knowledgeable of all 3 of us
-% of this general area.}
+Taking inspiration from the fact that many terms of the $\lambda$-calculus
+arise from Cartesian Closed Categories including, most importantly,
+a variety of propositional equalities and computation rules,
+allows us to pursue that analogy further. Some of the details
+of this development will be motivated and explained in the
+present paper.
 

### 2018-07-04 survey/survey.tex
partial work on OpSem section.

@@
 \usepackage{multicol}
-\usepackage{amsmath}
 \usepackage{stmaryrd}
@@
 
-\jc{Maybe interpret the next level as simply term rewriting?
-We already know the start and end, but if we also have a whole
-sequence of `trans', then we can reconstruct the whole thing
-from that. In other words, the example below could be
-reconstructed from the start, end and the program itself,
-erasing all of the lhs annotations, which are merely there for
-humans to track.}
+There are two different interpretations for an operational semantics
+for the language of equivalences:
+\begin{enumerate}
+\item Mimicking closely the one in the previous section, and
+thus finding explicit homotopies between the functions induced
+by the operational semantics of the level-1 combinators.
+\item Treating things more syntactically, and interpreting
+the combinators as program transformations.
+\end{enumerate}
+A previous paper~\cite{Carette2016} explores the first interpretation
+in depth. There one can find a definition of ``equivalences of
+equivalences'', which as the base of that interpretation.
+
+Here we will focus instead of the syntactic interpretation as
+program transformers. This results in a function
+\begin{flalign*}
+\mathit{eval}_1\,:\,\ \left\{ t_1\, t_2\,:\,U\right\}
+  \left\{ c_1\, c_2\,:\, t_1\,\leftrightarrow\,t_2\right\}
+  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow (t_1\,\leftrightarrow\,t_2) \\
+\end{flalign*}
+This function is ``deeply dependent'': given the type of the
+rewrite $ce$ to apply, both the input $c_1$ and output $c_2$ are
+almost entirely determined!  Let us take for example the
+second combinator in Fig.~\ref{figi}:
+\[
+  {(c_1 \otimes (c_2 \oplus c_3)) \odot \distl \Leftrightarrow \distl \odot ((c_1 \otimes c_2) \oplus (c_1 \otimes c_3))}
+\]
+\noindent which we can name \AgdaInductiveConstructor{distl⇔l}.
+Interpreting this as a rewrite from the program on the left to the
+one on the right requires ``pattern matching'' on the left
+structure which contains $3$ arbitrary combinators, from which
+we can \emph{reconstruct} the program on the right. Rewrites
+such as \AgdaInductiveConstructor{distl⇔l} are one-step rewrites,
+in the same way that \distl is a constant of the base term language
+of $\Pi$.  There is one additional wrinkle. There is naturally
+an opposite combinator, which interprets the above from right
+to left; let us call it \AgdaInductiveConstructor{distl⇔r}. It
+would appear to require \emph{non-linear pattern-matching}
+since the right-hand-side contains $c_1$ twice. That is however not
+the case! The definition of \AgdaInductiveConstructor{distl⇔r} has
+$5$ implicit arguments, $3$ of which are $c_1, c_2$, and $c_3$,
+which then completely force the ``shape'' of the overall pattern.
+Thus the mere mention of \AgdaInductiveConstructor{distl⇔r} is
+enough to resolve the apparent use of a non-linear pattern.
+This is why $\mathit{eval}_1$ was called ``deeply dependent''
+above: once the name of the combinator is given, the rest
+follows.
 

### 2018-07-04 survey/survey.tex
all of 4.3 now written.

@@
 
+\newcommand{\evalone}{\ensuremath{\mathit{eval}_1}}
+
 Here we will focus instead of the syntactic interpretation as
@@
 enough to resolve the apparent use of a non-linear pattern.
-This is why $\mathit{eval}_1$ was called ``deeply dependent''
+This is why \evalone was called ``deeply dependent''
 above: once the name of the combinator is given, the rest
 follows.
+
+\newcommand{\transLR}{\AgdaInductiveConstructor{trans⇔}}
+If all expressible transformations were single-step only, this
+would hardly justify calling this an ``operational semantics'',
+as we would hardly have a programming language. However, this level
+of $\Pi$ has combinators as well: two are in
+Fig.~\ref{fige} and two are in Fig.~\ref{figh}. The most interesting
+one is ``sequential composition'', which is the middle one at the 
+bottom of Fig.~\ref{figh}. Since $\Leftrightarrow$ represents
+an equivalence, sequential composition in this context is the
+same as transitivity of equivalences, as thus we have chosen to
+name this \transLR. When evaluating \transLR, we could cheat:
+we know that the eventual answer must be, and we could just 
+return that. But this is not operational in any real sense, as
+that skips over the intermediate steps. We would like to be able
+to ``trace'' the rewrite. Thus the evaluation of
+$\transLR r_0 r_1$ where $r_0 : c_0 \leftrightarrow c_1$ and
+$r_1 : c_1 \leftrightarrow c_2$
+should apply \evalone to both $r_0$ and $r_1$. Furthermore,
+after applying $r_0$, we should be able to witness that the result
+is indeed $c_1$, so that we may continue. This last requirement
+forces us to define a new function, mutually recursively with
+\evalone, for this task:
+\begin{flalign*}
+\mathit{exact}\,:\,\ \left\{ t_1\, t_2\,:\,U\right\}
+  \left\{ c_1\, c_2\,:\, t_1\,\leftrightarrow\,t_2\right\}
+  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow \evalone\, ce \equiv c_2 \\
+\end{flalign*}
+\noindent If we are careful in our construction of \evalone, the
+definition of $\mathit{exact}$ is quite straightforward, i.e.
+almost all cases are \AgdaInductiveConstructor{refl}.
+
+This then lets us define the $\transLR r_0 r_1$ case properly: we
+first evaluate $r_0$ and get a result combinator, witness that this result
+type is indeed exactly what we expect, and proceed to evaluate $r_1$
+where we specific that the $r_1$'s left-hand side must be
+$\evalone r_0$; we can use \AgdaKeyword{rewrite} to make this match
+$c_2$ ``on the nose'' (otherwise the call would be ill-typed).
+This then forces us to use \AgdaKeyword{rewrite} also in the
+implementation of the \transLR case in $\mathit{exact}$ before
+the proof is indeed just \AgdaInductiveConstructor{refl}.
+
+The other three combinators are much simpler, as simple
+recursive calls are sufficient.
 

### 2018-07-04 survey/survey.tex
wordcraft 4.4

@@
 %%%%%%%%%
-\subsection{Examples}
-
-We illustrate our methodology using a small example. Consider a
-circuit that takes an input type consisting of three values
+\subsection{Example}
+
+We can now illustrate how this all works with a small example.
+Consider a circuit that takes an input type consisting of three values
 \Tree [ {\small a} [ {\small b} {\small c} ] ]~
-and swaps the leftmost value with the rightmost value
-to produce
+and swaps the leftmost value with the rightmost value to produce
 \Tree [ {\small c} [ {\small b} {\small a} ] ]~.
@@
 isomorphisms we can \emph{explicitly} construct a sequence of
-rewriting steps that transforms the second circuit to the first.  The proof
+rewriting steps that transforms the second circuit to the first.
+
+We write such proofs in an equational style: in the left column, we have
+the current combinator which is equivalent to the first one, and in
+the right column, the justification for that equivalence. The
+joining combinator is syntactic sugar for \transLR.  The transformation
+could be written (using \transLR) by just giving all the pieces in
+the right hand column --- but such transformations are very hard for
+humans to understand and follow.
+
+The proof
 can be read as follows: the first three lines ``refocus'' from a right-associated

### 2018-07-05 survey/survey.tex
global pass

@@
 We now have in our hands our desired denotational semantics for types.
-We want to create a programming language, which we call $\Pi$, such that the
-types and type combinators map to $\bot, \top, \uplus, \times$, and
-such that we have ground terms whose denotation are all $16$ type
-isomorphisms of Fig.~\ref{type-isos}. This is rather
-straightforward, as we can simply do this literally. To make the
-analogy with commutative semirings stand out even more, we will use
-$0, 1, +$, and $\times$ at the type level, and will denote
-``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
-``constants'' of the language. Note how we have continued with
-the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof terms},
-but rather than being witnesses of inhabitation, they are witnesses of
-equivalences. Thus we get an unexpected programming language design:
+We want to create a programming language, which we call $\Pi$, such
+that the types and type combinators map to
+$\bot, \top, \uplus, \times$, and such that we have ground terms whose
+denotation are all $16$ type isomorphisms of
+Fig.~\ref{type-isos}. This is rather straightforward, as we can simply
+do this literally. To make the analogy with commutative semirings
+stand out even more, we will use $0, 1, +$, and $*$ at the type level,
+and will denote ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms}
+shows the ``constants'' of the language. Note how we have continued
+with the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof
+  terms}, but rather than being witnesses of inhabitation, they are
+witnesses of equivalences. Thus we get an unexpected programming
+language design:
 
@@
 
-\jc{The notation here is somewhat inconsistent with the above,
-mainly in the notation for values.}
+% \jc{The notation here is somewhat inconsistent with the above,
+% mainly in the notation for values.}
 
@@
   \begin{itemize}
-  \item For primitive isomophisms \ensuremath{c}, \ensuremath{!~ c} is given by its
+  \item For primitive isomorphisms \ensuremath{c}, \ensuremath{!~ c} is given by its
     inverse from Figures~\ref{pi-terms} and~\ref{more-pi}.
@@
  \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp b_1 + b_2 } \\
- \sem{0 \times b_2} & = & \absorbr \\
- \sem{1 \times b_2} & = & \identlt \odot \sem{b_2} \\
- \sem{(b_1\times b_2)\times b_3} & = & \assocrt \odot \sem{b_1\times (b_2\times b_3)} \\
- \sem{(b_1+b_2)\times b_3} & = & \dist \odot \sem{b_1\times b_3+b_2\times b_3} \\
+ \sem{0 * b_2} & = & \absorbr \\
+ \sem{1 * b_2} & = & \identlt \odot \sem{b_2} \\
+ \sem{(b_1 * b_2) * b_3} & = & \assocrt \odot \sem{b_1 * (b_2 * b_3)} \\
+ \sem{(b_1+b_2) * b_3} & = & \dist \odot \sem{b_1 * b_3+b_2 * b_3} \\
  \end{array}\]
@@
 
-Combinators of \ensuremath{\Pi } can be written in terms of the operators
-described previously or via a graphical language
-similar in spirit to those developed for Geometry of Interaction
-\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
-\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions and shorthand we
-describe here, the wiring diagrams are equivalent to the operator
-based (syntactic) description of programs.  \ensuremath{\Pi } combinators
-expressed in this graphical language look like ``wiring
-diagrams''. Values take the form of ``particles'' that flow along the
-wires. Computation is expressed by the flow of particles.
+Combinators of \ensuremath{\Pi } can be written in terms of the
+operators described previously or via a graphical language similar in
+spirit to those developed for Geometry of Interaction
+\cite{DBLP:conf/popl/Mackie95} and string diagrams for category
+theory~\cite{BLUTE1996229,selinger-graphical}. Modulo some conventions
+and shorthand we describe here, the wiring diagrams are equivalent to
+the operator based (syntactic) description of programs.
+\ensuremath{\Pi } combinators expressed in this graphical language
+look like ``wiring diagrams''. Values take the form of ``particles''
+that flow along the wires. Computation is expressed by the flow of
+particles.
 
@@
 \item
-The product type \ensuremath{b_1\times b_2} may be represented both as one wire
-labeled \ensuremath{b_1\times b_2} or by two parallel wires labeled \ensuremath{b_1} and
+The product type \ensuremath{b_1 * b_2} may be represented both as one wire
+labeled \ensuremath{b_1 * b_2} or by two parallel wires labeled \ensuremath{b_1} and
 \ensuremath{b_2}. Both representations may be used interchangeably.
@@
 \item
-When representing complex types like \ensuremath{(b_1\times b_2)+b_3} some visual
+When representing complex types like \ensuremath{(b_1 * b_2)+b_3} some visual
 grouping of the wires may be done to aid readability. The exact type
@@
 Associativity is entirely skipped in the graphical language. Hence
-three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
-\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
+three parallel wires may be inferred as \ensuremath{b_1 * (b_2 * b_3)} or
+\ensuremath{(b_1 * b_2) * b_3}, based on the context. This is much like handling of
 associativity in the graphical representations of categories as well as
@@
   can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
-  multiplicatively \ensuremath{c_1\times c_2} (shown on the right).
+  multiplicatively \ensuremath{c_1 * c_2} (shown on the right).
 
@@
 & \begin{aligned}
-c & : & b \times (1+1) \leftrightarrow b + b \\
+c & : & b * (1+1) \leftrightarrow b + b \\
 c & = & \swapt \odot \dist \odot (\identlt + \identlt) \\
@@
 
-Other than distributity, $\Pi$ combinators simply allow us to shuffle
+Other than distributivity, $\Pi$ combinators simply allow us to shuffle
 the components of a stack around as well as shuffle the positions of
@@
 Figure~\ref{type-isos} introduces our desired denotational
-semantices, and section~\ref{sec:opsem} a direct definition
+semantics, and section~\ref{sec:opsem} a direct definition
 of an operational semantics. One obvious question arises:
@@
 \subsection{Examples}
-
-Surprising expressiveness of Pi:
+\label{sec:langRev-examples}
+\label{examples}
+
+At first, it is not immediately clear that a programming language
+in which information is preserved could model choice. We recall a
+quote by Minsky communicating this concern:
 
@@
   also be reversible — and invented what’s now called the Fredkin
-  gate. (Minsky 1999)
+  gate~\cite{Hey:1999:FCE:304763}. 
 \end{quote}
 
-\begin{verbatim}
-examples:
-  (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
-  conditionals
-  toffoli
-  fredkin
-  revisit adder example from Sec 2
-  perhaps a larger example as well (from RC 18 paper?)
-\end{verbatim}
-
-\label{sec:langRev-examples}
-
-In this section we introduce programming in \ensuremath{\Pi } through
-several simple examples. We subsequently discuss the expressiveness of
-the language and discuss various properties.
-
-\label{examples}
+% \begin{verbatim}
+% examples:
+%   (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
+%   conditionals
+%   toffoli
+%   fredkin
+%   revisit adder example from Sec 2
+%   perhaps a larger example as well (from RC 18 paper?)
+% \end{verbatim}
+
+We will however show that one can program all logical gates in
+$\Pi$. We will start with a few simply examples and then discuss the
+expressiveness of the language and discuss various properties.
 
@@
 \ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
-using the type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})}.  We can perform various
+using the type \ensuremath{\mathit{bool} * (\mathit{bool} *  \mathit{bool})}.  We can perform various
 operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
@@
 
-\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not} \times (\mathit{not} \times \mathit{not})}
+\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  *  (\mathit{not}  *  \mathit{not})}
 
@@
 
-\ensuremath{\mathit{reverse} = \swapt \odot (\swapt \times \idc)~ \odot \assocrt}
+\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  *  \idc)~ \odot \assocrt}
 
@@
  \swapt & ((v_2, v_3), v_1) \\
- \swapt \times \idc & ((v_3, v_2), v_1) \\
+ \swapt  *  \idc & ((v_3, v_2), v_1) \\
  \assocrt & (v_3, (v_2, v_1)) \\
@@
 
-\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool} \times b_1 \leftrightarrow \mathit{bool} \times b_2}
-
-\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc \times c_1) + (\idc \times c_2)) \odot \factor}
+\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  *  b_1 \leftrightarrow \mathit{bool}  *  b_2}
+
+\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  *  c_1) + (\idc  *  c_2)) \odot \factor}
 
@@
 
-The diagram above shows the input value of type \ensuremath{(1+1)\times b_2}
+The diagram above shows the input value of type \ensuremath{(1+1) *  b_2}
 processed by the distribute operator \ensuremath{\dist}, which converts it into
-a value of type \ensuremath{(1\times b_2)+(1\times b_2)}. In the
+a value of type \ensuremath{(1 *  b_2)+(1 *  b_2)}. In the
 \ensuremath{\mathit{left}} branch, which corresponds to the
@@
 By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
-combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool}\times \mathit{bool}\leftrightarrow \mathit{bool}\times \mathit{bool}} which negates its
+combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} *  \mathit{bool}\leftrightarrow \mathit{bool} *  \mathit{bool}} which negates its
 second argument if the first argument is \ensuremath{\mathit{true}}. This gate
@@
 If we iterate this construction once more, the resulting combinator
-\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})\leftrightarrow \mathit{bool}\times (\mathit{bool}\times \mathit{bool})}. The
+\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} *  (\mathit{bool} *  \mathit{bool})\leftrightarrow \mathit{bool} *  (\mathit{bool} *  \mathit{bool})}. The
 resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
@@
 
-\jc{I am not a big fan of these pictures -- mainly because I can't give any
-intuition based on them of why these can be smoothly deformed into one another.
-All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
-5 dimensions -- one for each of the wires, and another to give room to maneuver.}
-\end{comment}
+% \jc{I am not a big fan of these pictures -- mainly because I can't give any
+% intuition based on them of why these can be smoothly deformed into one another.
+% All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
+% 5 dimensions -- one for each of the wires, and another to give room to maneuver.}
+% \end{comment}
 
@@
 (where we have explicitly annotated the types of $\idc$ for increased clarity)
-tells us that of the two ways of transforming from $t_1 \times 0$ to $0$, 
+tells us that of the two ways of transforming from $t_1  *  0$ to $0$, 
 namely first doing some arbitrary transformation $c$ from $t_1$ to $t_2$ and
@@
 
-The next combinator, $\absorbr \Leftrightarrow \absorbl$,
-is particularly fascinating: while it says something simple ---
-that the two obvious ways of transforming $0 \times 0$ into $0$, namely
-absorbing either the left or right zero --- it implies something subtle.
-A straightforward proof of $\absorbl$ which proceeds by saying that
-$0 \times t$ cannot be inhabited because the first member of the pair
-cannot, is not in fact equivalent to $\absorbr$ on $0 \times 0$.
-However, if we instead define $\absorbl$ to ``transport'' the
-putative impossible first member of the pair to its (equally
-impossible) output, then these do form equivalent pairs.
-The next few in Fig.~\ref{figc} also express how $\absorbr$ and
-$\absorbl$ interact with other combinators. As seen previously,
-all of these arise as critical pairs. What is much more subtle here
-is that the types involved often are asymmetric: they do not have
-the same occurences on the left and right. Such cases are
-particularly troublesome for finding normal forms \jc{should add
-some citations here}.
+The next combinator, $\absorbr \Leftrightarrow \absorbl$, is
+particularly fascinating: while it says something simple --- that the
+two obvious ways of transforming $0 * 0$ into $0$, namely absorbing
+either the left or right zero --- it implies something subtle.  A
+straightforward proof of $\absorbl$ which proceeds by saying that
+$0 * t$ cannot be inhabited because the first member of the pair
+cannot, is not in fact equivalent to $\absorbr$ on $0 * 0$.  However,
+if we instead define $\absorbl$ to ``transport'' the putative
+impossible first member of the pair to its (equally impossible)
+output, then these do form equivalent pairs.  The next few in
+Fig.~\ref{figc} also express how $\absorbr$ and $\absorbl$ interact
+with other combinators. As seen previously, all of these arise as
+critical pairs. What is much more subtle here is that the types
+involved often are asymmetric: they do not have the same occurrences
+on the left and right. Such cases are particularly troublesome for
+finding normal forms.
+
+%% \jc{should add some citations here}.
 
@@
 
-\jc{The above is, up to uses of assoc, a linear notation for a
-commutative diagram. So draw it.}
+% \jc{The above is, up to uses of assoc, a linear notation for a
+% commutative diagram. So draw it.}
 

### 2018-07-05 survey/survey.tex
tweak some level sentences. None were wrong.

@@
 Fig.~\ref{fige} are the level-2 analogs of $+$ and $*$, which respectively
-model combinator-level choice composition and parallel composition (of
+model level-1 choice composition and parallel composition (of
 equivalences).  These allow us to ``build up'' larger equivalences from smaller
@@
 the $\oplus$ and $\otimes$ combinators ``commute'' with re-association.
-In other words, it expresses the type-level associativity of $+$ is
+In other words, it expresses that the type-level associativity of $+$ is
 properly reflected by the properties of $\oplus$.
@@
 $\dist$, $\distl$, $\factor$ and $\factorl$: the type-level distribution
-and factoring has to commute with the combinator-level $\oplus$ and $\otimes$.
+and factoring has to commute with the level-1 $\oplus$ and $\otimes$.
 
@@
 The first two equivalences in Fig.~\ref{figf} reflect the basic
-coherence between type-level swapping and and the level-2 combinator
+coherence between level-0 swapping and the level-1 combinator
 actions. The next four arise because of interactions between
@@
 would hardly justify calling this an ``operational semantics'',
-as we would hardly have a programming language. However, this level
+as we would hardly have a programming language. However, level-2
 of $\Pi$ has combinators as well: two are in

### 2018-07-05 survey/survey.tex
add an new section on internal language in section 4. Draw the monster commutative diagram. Second one coming later.

@@
 %%%%%%%%%
-\subsection{Example}
+\subsection{Example}\label{sec:level2-example}
 
@@
 
-% \jc{The above is, up to uses of assoc, a linear notation for a
-% commutative diagram. So draw it.}
+\subsection{Internal Language}
+
+Recalling that the $\lambda$-calculus arises as the internal language
+of Cartesian Closed Categories (Elliott~\cite{Elliott-2017} gives a particularly
+readable account of this), we can think of $\Pi$ in similar terms, but
+for symmetric Rig Groupoids instead. For example, we can ask what does
+the derivation in Sec.~\ref{sec:level2-example} represent? It is
+actually a ``linear'' representation of a 2-categorial commutative
+diagram! In fact, it is a painfully verbose version thereof, as it
+includes many \emph{refocusing} steps because our language does not
+build associativity into its syntax. Categorical diagrams usually do.
+Thus if we rewrite the example in diagrammatic form, eliding all uses
+of associativity, but keeping explicit uses of identity transformations,
+we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
+
+\newcommand{\idd}{\mathit{id}\leftrightarrow}
+\newcommand{\idf}{\mathit{id}\Leftrightarrow}
+
+\begin{tikzcd}[column sep=normal, row sep=small]
+ && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
+ & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
+a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [rr, "\assoclp\odot\swapp"] 
+  \arrow [ddrr, "\assoclp\odot\swapp"]
+  \arrow [dddrr, swap, "\assoclp\odot\swapp"]
+    &&
+    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}] 
+    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
+ &&&&& \\
+ & & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
+             \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] && 
+    c+(a+b) \arrow [ruu, swap, "\idd\oplus\swapp"] & \\
+ && c+(a+b) \arrow [rrruuu, bend right = 30, swap, "\idd\oplus\swapp"] && \\
+ \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
+ \arrow[Rightarrow, from=Z, to=DD, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
+\end{tikzcd}
+
+\noindent For some, the above diagram will be clearer --- it is only three layers
+high rather than nine! Others will prefer the more programmatic feel of the
+original definition.
+
+We would be remiss in letting the reader believe that the above is ``the''
+categorical diagram that would be found in categorical textbooks. Rather,
+congruence would be used to elide the $\idf$. The various arrows would also
+be named differently --- our \assoclp is often named $\alpha$,
+\assocrp is $\alpha^{-1}$, $\swapp$ is $B$ (always with subscripts).
 

### 2018-07-05 survey/survey.tex
fix some infelicities in diagram and further pretty-ify.

@@
  & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
-a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [rr, "\assoclp\odot\swapp"] 
-  \arrow [ddrr, "\assoclp\odot\swapp"]
-  \arrow [dddrr, swap, "\assoclp\odot\swapp"]
-    &&
+a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
+  \arrow [dr, "\assoclp"]
+  \arrow [ddr, swap, "\assoclp"]
+    & (a+b)+c \arrow [r, "\swapp"] &
     c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}] 
     & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
- &&&&& \\
- & & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
+ & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
+ & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
              \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] && 
-    c+(a+b) \arrow [ruu, swap, "\idd\oplus\swapp"] & \\
+    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
  && c+(a+b) \arrow [rrruuu, bend right = 30, swap, "\idd\oplus\swapp"] && \\
  \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
- \arrow[Rightarrow, from=Z, to=DD, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
+ \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
 \end{tikzcd}

### 2018-07-05 survey/survey.tex
Finish section 4.  I might add a reference or two, but the core content is, AFAIK, all there.

@@
 \newcommand{\idf}{\mathit{id}\Leftrightarrow}
-
-\begin{tikzcd}[column sep=normal, row sep=small]
+\vspace*{3mm}
+\begin{tikzcd}[column sep=normal, row sep=normal]
  && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
@@
     c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
- && c+(a+b) \arrow [rrruuu, bend right = 30, swap, "\idd\oplus\swapp"] && \\
+ && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
  \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
@@
 categorical diagram that would be found in categorical textbooks. Rather,
-congruence would be used to elide the $\idf$. The various arrows would also
-be named differently --- our \assoclp is often named $\alpha$,
+congruence would be used to elide the $\idf$. Furthermore, the various arrows
+would also be named differently --- our \assoclp is often named $\alpha$,
 \assocrp is $\alpha^{-1}$, $\swapp$ is $B$ (always with subscripts).
+And the two steps needed to remove inverses (i.e. first cancelling
+inverse arrows, then removing the resulting identity arrow ``in context'')
+are often combined into one. Here we'll simply name this operation 
+$\mathit{cancel}$, which could be programmed as a defined function over
+$\Pi$ level-2.  The result would then be the much simpler
+
+\vspace*{3mm}
+\begin{tikzcd}[column sep=normal, row sep=normal]
+ & a+(c+b) \arrow [r, "\assoclp", ""{name=U, below}] & (a+c)+b \arrow [rd, "\swapp\oplus\idd"] & & & \\
+a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [dr, "\assoclp"]
+  & & & (c+a)+b \arrow [r, "\assocrp", ""{name=UU, below}] & c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
+ & (a+b)+c \arrow [r, swap, "\swapp", ""{name=D, above}] & c+(a+b) \arrow [ur, "\assoclp"] 
+  \arrow [urrr, swap, "\idd\oplus\swapp", ""{name=DD,above}] & & & & \\
+ \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}"]
+ \arrow[Rightarrow, from=UU, to=DD, "\mathit{cancel}"]
+\end{tikzcd}
+
+In other words, each (non-refocusing) line of the proof of
+\AgdaFunction{swap{-}fl2⇔swap{-}fl1}\; is a complete path 
+from left to right in each diagram above, and the annotation
+on the right-hand-side becomes the natural transformation (denoted
+by vertical $\Rightarrow$) justifying the move to the next line.
+The first diagram uses lines $1,4,7,8$ in full; the second
+diagram collapses $7$ and $8$ into one, as well as not duplicating
+parts which are related by $\idf$.
 

### 2018-07-06 survey/survey.tex
add some more references.

@@
   Also since the monoidal units can be freely introduced and eliminated, sometimes
-  they are omitted.  However, as this is in fact dangerous, as explained by
-  many diagrams they are omitted and dealt with explicitly only when they are
-  of special interest.
+  they are omitted.  However, as this is in fact dangerous, as explained
+  by~\cite{BLUTE1996229}, we will err on the side of including them.
 
@@
 
-% \begin{verbatim}
-% examples:
-%   (1 + 1) x ((1 + 1) x b) = (b+b) + (b+b)
-%   conditionals
-%   toffoli
-%   fredkin
-%   revisit adder example from Sec 2
-%   perhaps a larger example as well (from RC 18 paper?)
-% \end{verbatim}
-
 We will however show that one can program all logical gates in
@@
 \emph{critical pairs} of the system of equations.
+The problem is far from trivial --- Fiore et al.~\cite{Fiore-2008}
+document some publications where the coherence set is in
+fact incorrect. They furthermore give a quite general algorithm
+to derive such coherence conditions.
 
@@
 on the left and right. Such cases are particularly troublesome for
-finding normal forms.
+finding normal forms. Laplaza~\cite{laplaza72} certainly comments on this,
+but in mostly terse and technical terms. Blute et al.~\cite{BLUTE1996229}
+offer much more intuitive explanations.
 

### 2018-07-06 survey/survey.tex
add a small section in the conclusion about Quantum Speedup.

@@
 %%%
+\subsection{Quantum Speed-up}
+
+A rather remarkable but somehow overlooked paper is
+``Quantum speedup and Categorical Distributivity'' by
+Peter Hines~\cite{hines2013quantum}. Here he shows that
+the heart of Shor's algorithm can be reduced to an
+operation $!^N()$\ (expressible in $\Pi$), which can
+be expressed, via a factorization, in an exponentially
+faster manner. The key to this efficient factorization
+is exactly the coherence conditions of Laplaza~\cite{laplaza72},
+which also feature prominently in our work. Proving his
+key Lemma 2 in $\Pi$ could be quite instructive in revealing
+which level-2 combinators are crucial for this result.
+
+%%%
 \subsection{Summary}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{comment}
-\section{Theseus: Types and Simple Isomorphisms}
-\label{theseusI}
-
-We begin by presenting the core of Theseus which consists of evidently
-reversible functions on user-defined datatypes.  In subsequent discourse we
-will use the word `map' to refer to these logically reversible
-functions. Theseus has the same built-in types as~\ensuremath{\Pi^{o}} and as conventional
-functional programming languages: the empty type, the unit type, and sum and
-product types. All other types, including recursive types, are user-defined
-using a general mechanism for type declarations:
-
-\lstset{
-        language=haskell,
- morekeywords={Bool,Nat,Tree},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-type Bool  =  (@\ctr{True}@) | (@\ctr{False}@)
-type Nat   =  0 | Succ Nat
-type Tree  =  Leaf Nat | Node Tree Tree
- \end{lstlisting}
-%subcode source theseus.tex:347
-
-\noindent The declarations above witness isomorphisms between the set
-\kw{Bool} and the set constructed by the constants \ctr{True} and
-\ctr{False}, between the set \kw{Nat} and the set inductively constructed
-using {\scriptsize{0}} and \ctr{Succ}, and between the set \kw{Tree} and the
-set inductively constructed using \ctr{Leaf} and \ctr{Node}.  Thus, Theseus
-type definitions are similar to, say, Haskell, type definitions. Such
-definitions translate directly to the underlying \ensuremath{\Pi^{o}} using iso-recursive
-type definitions. For example, the above types translate as follows:
-\vspace{-20pt}
-\begin{small}
-\begin{multicols}{3}
-\[\begin{array}{rclr}
-\kw{Bool}   & = &  \mu  ~x.1 + 1 \\
-\kw{Nat}   & = &  \mu  ~x. 1 + x \\
-\kw{Tree}   & = &  \mu  ~x. \kw{Nat}  + x * x \\
- \end{array}\]
-%subcode source theseus.tex:363
-
-\[\begin{array}{rclr}
-\mathit{unfoldBool} :& \kw{Bool}  \leftrightarrow 1 + 1 &: \mathit{foldBool} \\
-\mathit{unfoldNat} :& \kw{Nat}  \leftrightarrow 1 + \kw{Nat}  &: \mathit{foldNat} \\
-\mathit{unfoldTree} :& \kw{Tree}  \leftrightarrow \kw{Nat}  + (\kw{Tree}  * \kw{Tree} ) &: \mathit{foldTree} \\
- \end{array}\]
-%subcode source theseus.tex:368
-\end{multicols}
-\end{small}
-
-%%%%%%%%%%%%
-\subsection{Simple Isomorphisms}
-
-A Theseus program has a type of the form \ctr{a} $\leftrightarrow$ \ctr{b};
-one runs such a program forward by supplying a value of type \ctr{a}. If the
-program terminates, we get back a value of type~\ctr{b}. Alternately we can
-run the program in reverse by supplying a value of type \ctr{b}. Given the
-built-in types and the facility to introduce user-defined ones, one can
-already write a few small but interesting programs. Maps that are evidently
-isomorphisms can be written in the familiar pattern matching style
-popularized by the family of functional languages. For example:
-\begin{multicols}{2}
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-(@\ctr{id}@) :: Bool (@$\leftrightarrow$@) Bool
-| (@\ctr{False}@)  (@$\leftrightarrow$@)  (@\ctr{False}@)
-| (@\ctr{True}@)   (@$\leftrightarrow$@)  (@\ctr{True}@)
-
-expandBool :: Bool * a (@$\leftrightarrow$@) a + a
-| (@\ctr{True}@), a   (@$\leftrightarrow$@)  (@\ctr{Left}@) a
-| (@\ctr{False}@), a  (@$\leftrightarrow$@)  (@\ctr{Right}@) a
-
-expandNat :: Nat (@$\leftrightarrow$@) Nat + 1
-| 0       (@$\leftrightarrow$@)  (@\ctr{Right}@) ()
-| Succ n  (@$\leftrightarrow$@)  (@\ctr{Left}@) n
-
-(@\ctr{not}@) :: Bool (@$\leftrightarrow$@) Bool
-| (@\ctr{True}@)   (@$\leftrightarrow$@)  (@\ctr{False}@)
-| (@\ctr{False}@)  (@$\leftrightarrow$@)  (@\ctr{True}@)
-
-foldBool :: a + a (@$\leftrightarrow$@) Bool * a
-| (@\ctr{Left}@) a   (@$\leftrightarrow$@)  (@\ctr{True}@), a
-| (@\ctr{Right}@) a  (@$\leftrightarrow$@)  (@\ctr{False}@), a
-
-foldNat :: Nat + 1 (@$\leftrightarrow$@) Nat
-| (@\ctr{Right}@) ()  (@$\leftrightarrow$@)  0
-| (@\ctr{Left}@) n    (@$\leftrightarrow$@)  Succ n
- \end{lstlisting}
-%subcode source theseus.tex:410
-\end{multicols}
-
-\vspace{-25pt}
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-treeUnwind :: Tree (@$\leftrightarrow$@) Tree * Tree + (Bool + Nat)
-| Node t1 t2            (@$\leftrightarrow$@)  (@\ctr{Left}@)   (t1, t2)
-| Leaf 0                (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Left}@) (@\ctr{True}@))
-| Leaf (Succ 0)         (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Left}@) (@\ctr{False}@))
-| Leaf (Succ (Succ n))  (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Right}@) n)
- \end{lstlisting}
-%subcode source theseus.tex:422
-
-Unlike the situation in a conventional functional language, the
-intuition here is that the two sides of each pattern clause can be
-swapped to produce the inverse of the function, i.e. as we will see
-repeatedly in the paper, patterns and expressions are the same
-thing. Compare for example the definitions of \ctr{expandBool} and
-\ctr{foldBool}. This is the only constraint that a programmer needs to
-maintain. This constraint can be ensured using the following two
-rules:
-
-\begin{enumerate}
-\label{rules}
-\item \emph{Non-overlapping and exhaustive coverage in pattern
-  clauses.} The collections of patterns in the left-hand side (LHS) of
-  each clause must be a complete non-overlapping covering of the input
-  type. Similarly, the collections of patterns in the right-hand side
-  (RHS) of each clause must also be a complete non-overlapping
-  covering of the return type. In other words, no cases should be
-  omitted or duplicated.
-\item \emph{Preserve typed variables across \ensuremath{\leftrightarrow}.} Each variable that
-  occurs on one side of a pattern matching clause can only appear once on
-  that side and must appear exactly once on the other side and with the same
-  type.
-\end{enumerate}
-
-When restricted to one side of each pattern matching clause, the rules
-should be familiar and intuitive. Note that, in general, the two sides
-of a pattern matching clause may have different types. Other than the
-requirement of non-overlapping and exhaustive coverage, there are no
-rules that constrain the types and names of constants nor of
-constructors. So for example, in \ctr{foldBool}, the pattern
-`\ctr{Left}~\ctr{a}' maps to a pair constructor with the first
-component being the constant \ctr{True}.
-
-Examples of expressions that violate the constraints follow:
-
-\begin{multicols}{2}
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
--- Invalid: Missing LHS cases
-missing_node :: Tree (@$\leftrightarrow$@) Nat
-| Leaf n (@$\leftrightarrow$@) n
-
--- Invalid: Overlapping RHS cases
-overlapping_cases :: Nat (@$\leftrightarrow$@) Nat
-| 0       (@$\leftrightarrow$@)  0
-| Succ n  (@$\leftrightarrow$@)  n
- \end{lstlisting}
-%subcode source theseus.tex:471
-
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
--- Invalid: n is dropped
-drop_var :: Tree (@$\leftrightarrow$@) Tree * Tree + Nat
-| Node t1 t2  (@$\leftrightarrow$@)  (@\ctr{Left}@)  (t1,t2)
-| Leaf n      (@$\leftrightarrow$@)  (@\ctr{Right}@) 0
-
--- Invalid: t is used twice
-dup_var :: Tree (@$\leftrightarrow$@) Tree * Tree + Nat
-| Node t t  (@$\leftrightarrow$@)  (@\ctr{Left}@)  (t,t)
-| Leaf n    (@$\leftrightarrow$@)  (@\ctr{Right}@) n
- \end{lstlisting}
-%subcode source theseus.tex:485
-\end{multicols}
-
-\noindent It is easy to see that every primitive isomorphism of \ensuremath{\Pi^{o}} can
-be expressed in Theseus. Here are a few examples.
-\vspace{-10pt}
-\begin{multicols}{2}
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-assocL :: a+(b+c) (@$\leftrightarrow$@) (a+b)+c :: assocR
-| (@\ctr{Left}@) a          (@$\,\,\leftrightarrow$@) (@\ctr{Left}@) ((@\ctr{Left}@) a)
-| (@\ctr{Right}@) ((@\ctr{Left}@) b)   (@$\leftrightarrow$@) (@\ctr{Left}@) ((@\ctr{Right}@) b)
-| (@\ctr{Right}@) ((@\ctr{Right}@) c)  (@$\leftrightarrow$@) (@\ctr{Right}@) c
- \end{lstlisting}
-%subcode source theseus.tex:499
-
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-zeroe :: b + 0 (@$\leftrightarrow$@) b :: zeroi
-| (@\ctr{Left}@) v (@$\leftrightarrow$@) v
-
-swapTimes :: a * b (@$\leftrightarrow$@) b * a
-| (a,b) (@$\leftrightarrow$@) (b,a)
- \end{lstlisting}
-%subcode source theseus.tex:509
-\end{multicols}
-
-\label{translation}
-\noindent
-Every simple isomorphism of Theseus can be translated to \ensuremath{\Pi^{o}} as follows:
-\begin{enumerate}
-\item For every Theseus map \ctr{c : a} $\leftrightarrow$ \ctr{b}, we can
-  define an intermediate type \ctr{ti} which is the sum of pairs of all the
-  variables that occur in the LHS patterns. Since the LHS and RHS patterns
-  contain the same type variables, \ctr{ti} is unique for a given \ctr{c} (up
-  to ordering). For example, in \ctr{treeUnwind} the first clause has
-  variables \lstinline{t1 : Tree} and \lstinline{t2 : Tree}, the second and
-  third clauses have no variables and the fourth clause has the variable
-  \lstinline{n : Nat}. The expected type \ctr{ti} is therefore
-  \lstinline{Tree * Tree + (1 + (1 + Nat))}, where we insert the type
-  {\scriptsize{1}} as a placeholder for the clauses that did not have type
-  variables.
-\item By construction \ctr{ti} is isomorphic to \ctr{a} and \ctr{b}. Hence,
-  maps \ctr{c1 : a} $\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti}
-  $\leftrightarrow$ \ctr{b} must be expressible in \ensuremath{\Pi^{o}}. (Both the above
-  isomorphisms can be constructed by systematically unfolding and
-  distributing \ctr{a} and \ctr{b} until \ctr{ti} is obtained.  The exact
-  details of such an expansion can be found in Sec.~3 of our previous
-  paper~\cite{rc2012} and are skipped in this paper.) Once values of type
-  \ctr{a} and \ctr{b} can be mapped to values of type \ctr{ti}, the required
-  translation of \ctr{c} is given by \ctr{c1} $\odot$ \ctr{c2}.
-\end{enumerate}
-
-%%%%%%%%%%%%
-\subsection{Dealing with Numbers}
-
-If we try to do arithmetic using the previously defined Peano-style \ensuremath{\kw{Nat} },
-we have to address an issue of what to do when we encounter
-\lstinline{sub1 0}. There are two obvious choices:
-\begin{enumerate}
-\item We can define \ctr{add1} and \ctr{sub1} of type \kw{Nat}
-  $\leftrightarrow$ \kw{Nat}, such that \ctr{sub1} of {\scriptsize{0}}
-  diverges. We show how these maps can be defined in Sec.~\ref{add1}.
-\item An error mechanism can be added to the language such that when
-  an error is raised program execution is undefined and there is no
-  meaningful inverse definition.
-\end{enumerate}
-Another choice, similar to that made by Janus, is to use bounded integers
-such that every operation is always well defined through underflows and
-overflows. Here is a simple 4-bit \kw{Nat4} datatype with its corresponding
-\ctr{add1} and \ctr{sub1} operations:
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool,Nat4},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-type Nat4 = Bool * Bool * Bool * Bool
-
-add1 :: Nat4 (@$\leftrightarrow$@) Nat4 :: sub1
-| (a, b, c, (@\ctr{False}@))          (@$\leftrightarrow$@)  (a, b, c, (@\ctr{True}@))
-| (a, b, (@\ctr{False}@), (@\ctr{True}@))       (@$\,\leftrightarrow$@)  (a, b, (@\ctr{True}@), (@\ctr{False}@))
-| (a, (@\ctr{False}@), (@\ctr{True}@), (@\ctr{True}@))    (@$\,\,\leftrightarrow$@)  (a, (@\ctr{True}@), (@\ctr{False}@), (@\ctr{False}@))
-| ((@\ctr{False}@), (@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@))  (@$\leftrightarrow$@)  ((@\ctr{True}@), (@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@))
-| ((@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@))   (@$\leftrightarrow$@)  ((@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@))
- \end{lstlisting}
-%subcode source theseus.tex:567
-
-\noindent It is easy to see how a math library may be defined in this
-way. Once an operator can be defined in Theseus it is always possible to
-replace its implementation by an equivalent one that is expressible
-efficiently in the underlying hardware. In this case one could compile down
-to the CPU's native integer representation and operators.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Parametrized Maps}
-\label{theseusII}
-
-Now that we can express simple standalone isomorphisms, we explain how to
-compose such isomorphisms to model more complex behavior. In \ensuremath{\Pi^{o}}, there
-are three ways of composing isomorphisms: sequential composition, parallel
-composition, and choice composition. The common idiom underlying these
-composition combinators is that a reversible map can be parametrized by
-another reversible map. This idea is related to ``higher-order functions''
-but is more limited as we explain below.
-
-%%%%%%%%%%%%
-\subsection{Definition and Examples}
-
-A Theseus map \ctr{f} can be parametrized by another map \ctr{g} by adding a
-labeled argument~\ctr{g} of the appropriate type to \ctr{f}. In the example
-below \ctr{treeUnwindf} is parametrized over some map \ctr{f} which it
-applies to \ctr{n}, if the supplied tree is \lstinline{Leaf n}:
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-treeUnwindf :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
-| Node t1 t2 (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
-| Leaf n     (@$\leftrightarrow$@) (@\ctr{Right}@) (f n)
- \end{lstlisting}
-%subcode source theseus.tex:600
-
-\noindent This parametrization should be thought of as a macro or a
-meta-language construction. Theseus does not have high-order maps in the
-formal sense. In other words, the final type of a Theseus program must be of
-the form \ctr{a} $\leftrightarrow$ \ctr{b} and every occurrence of an arrow
-type $\rightarrow$ must be instantiated at compile time. The notation for
-instantiating the label parameter \ctr{f} in the map \ctr{fun} by the map
-\ctr{g} is \inline{fun}{f}{g}. For example,
-\inline{treeUnwindf}{f}{expandNat} should be thought of as shorthand for the
-map:
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-_ :: Tree (@$\leftrightarrow$@) Tree * Tree + (1 + Nat)
-| Node t1 t2    (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
-| Leaf 0        (@$\leftrightarrow$@) (@\ctr{Right}@) ((@\ctr{Right}@) ())
-| Leaf (Succ n) (@$\leftrightarrow$@) (@\ctr{Right}@) ((@\ctr{Left}@) n)
- \end{lstlisting}
-%subcode source theseus.tex:618
-
-\noindent which inlines \ctr{expandNat} within the definition of
-\ctr{treeUnwindf}.
-
-We discuss the reversibility of parametrized maps in the next section. We
-first point out that the \ensuremath{\Pi^{o}} closure primitives can be expressed as
-parametrized maps:
-\lstset{
-        language=haskell,
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-_._ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(b (@$\leftrightarrow$@) c) (@$\rightarrow$@) a (@$\leftrightarrow$@) c
-| a (@$\leftrightarrow$@) g (f a)
-
-_*_ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(c (@$\leftrightarrow$@) d) (@$\rightarrow$@) a * c (@$\leftrightarrow$@) b * d
-| (a,c) (@$\leftrightarrow$@) (f a , g c)
-
-_+_ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(c (@$\leftrightarrow$@) d) (@$\rightarrow$@) a + c (@$\leftrightarrow$@) b + d
-| (@\ctr{Left}@) a   (@$\leftrightarrow$@)  (@\ctr{Left}@) (f a)
-| (@\ctr{Right}@) c  (@$\leftrightarrow$@)  (@\ctr{Right}@) (g c)
- \end{lstlisting}
-%subcode source theseus.tex:637
-
-As an additional example, parametrized maps allow us to define controlled
-operations in the tradition of reversible circuits. In the definition below,
-the boolean input is a control bit that determines which of the maps \ctr{th}
-or \ctr{el} is applied to the second component of the pair. This conditional
-is then used to define the ``controlled-not'' gate:
-
-\lstset{
-        language=haskell,
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-(@\ctr{if}@) :: th:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) el:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) Bool * a (@$\leftrightarrow$@) Bool * b
-| (@\ctr{True}@), a  (@$\leftrightarrow$@) (@\ctr{Left}@) (th a)
-| (@\ctr{False}@), a (@$\leftrightarrow$@) (@\ctr{Right}@) (el a)
-
-cnot :: Bool * Bool (@$\leftrightarrow$@) Bool * Bool
-| control, bit (@$\leftrightarrow$@) if ~th:(@\ctr{not}@) ~el:(@\ctr{id}@) (control, bit)
- \end{lstlisting}
-%subcode source theseus.tex:653
-
-%%%%%%%%%%%%
-\subsection{Reversing Parametrized Maps}
-
-Reversing a map of type \ctr{a} $\leftrightarrow$ \ctr{b} results in a map of
-type \ctr{b} $\leftrightarrow$ \ctr{a} which is its adjoint. As discussed
-before, simple maps can be reversed by simply switching the LHS and RHS for
-both the type and the pattern matching clauses comprising the
-body. Interestingly, one can think about the reverse of parametrized maps in
-the same way -- i.e. one simply needs to switch LHS and RHS of the `\ensuremath{\leftrightarrow}'.
-
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-treeUnfold :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
-| Node t1 t2  (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
-| Leaf n      (@$\leftrightarrow$@) (@\ctr{Right}@) (f n)
-
--- the adjoint of treeUnfold
-rev_treeUnfold :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree * Tree + a (@$\leftrightarrow$@) Tree
-| (@\ctr{Left}@) (t1, t2)  (@$\leftrightarrow$@) Node t1 t2
-| (@\ctr{Right}@) (f n)    (@$\leftrightarrow$@) Leaf n
- \end{lstlisting}
-%subcode source theseus.tex:676
-
-\noindent Note that the application \lstinline{f n} happens on the left of
-the \ensuremath{\leftrightarrow}, which is something that one does not see in conventional
-functional languages. These applications in the LHS have the dual meaning of
-applications in the RHS and should be understood as follows: the value that
-is the actual argument of the \ctr{Right} constructor is the result of
-applying~\lstinline{f n}. In the forward execution of
-\lstinline{rev_treeUnfold}, when the actual value of the constructor argument
-is encountered, it is passed through the reverse of \ctr{f} to get the value
-\ctr{n}. In other words, the above program is exactly the same as
-\lstinline{rev_treeUnfold'} below where \lstinline{rev_f} is the adjoint of
-\ctr{f}:
-
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-rev_treeUnfold' :: rev_f:(a (@$\leftrightarrow$@) Nat) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
-| (@\ctr{Left}@) (t1, t2) (@$\leftrightarrow$@) Node t1 t2
-| (@\ctr{Right}@) a       (@$\leftrightarrow$@) Leaf (rev_f a)
- \end{lstlisting}
-%subcode source theseus.tex:696
-
-\noindent
-It follows that the following programs are equivalent:
-\vspace{-10pt}
-\begin{multicols}{2}
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-g :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) a (@$\leftrightarrow$@) b
-| a (@$\leftrightarrow$@) f a
-
-g :: rev_f:(b (@$\leftrightarrow$@) a) (@$\rightarrow$@) a (@$\leftrightarrow$@) b
-| rev_f b (@$\leftrightarrow$@) b
- \end{lstlisting}
-%subcode source theseus.tex:710
-\end{multicols}
-\vspace{-15pt}
-\noindent
-Based on the above observation, it is easy to define the \ctr{sym} map that
-constructs the adjoint of any given map:
-\vspace{-5pt}
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Int,Bool},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-sym :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) b (@$\leftrightarrow$@) a
-| (f a) (@$\leftrightarrow$@) a
- \end{lstlisting}
-%subcode source theseus.tex:722
-While parametrized maps add tremendous programming convenience to Theseus,
-they don't change the expressive power of the language. All programs
-expressible with parametrized maps, can be expressed without them by fully
-inlining the actual parameters. Thus compiling Theseus programs with
-parametrized maps to \ensuremath{\Pi^{o}} simply requires first inlining the maps and then
-doing the translation described in Sec.~\ref{translation}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Iteration Labels}
-\label{theseusIII}
-\label{add1}
-
-The fragment of Theseus discussed up to now, modulo full recursive types, can
-be expressed in pure \ensuremath{\Pi } without \ensuremath{\mathit{trace}}.  Introducing recursive
-definitions in a reversible language is subtle because every iteration must
-be reversible which requires, at least, that the number of iterations be
-recoverable from the output. The insight we use in Theseus is that this can
-be achieved elegantly by adding \emph{typed labels} as explained below. We
-start with a small example:
-
-\lstset{
-        language=haskell,
- morekeywords={iter,Nat},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-parity :: Nat * Bool (@$\leftrightarrow$@) Nat * Bool
-| n, b                (@$\leftrightarrow$@) iter $ n, 0, b
-| iter $ Succ n, m, b (@$\leftrightarrow$@) iter $ n, Succ m, (@\ctr{not}@) b
-| iter $ 0, m, b      (@$\leftrightarrow$@) m, b
-  where iter :: Nat * Nat * Bool
- \end{lstlisting}
-%subcode source theseus.tex:751
-
-\noindent The map \ctr{parity} calculates whether the given natural number is
-even or odd by counting to {\scriptsize{0}} and flipping the second boolean
-input each time. In the definition above, the label appears twice on the LHS
-and twice on the RHS. The correctness (i.e., reversibility) of the map is
-guaranteed as labels obey the same restrictions as the one for simple
-patterns. In particular the occurrences of the label in the LHS must
-constitute a non-overlapping coverage of the label type, and similarly for
-the occurrences of the label in the RHS.
-
-The important thing to note about labels is that they may have a
-different type from the LHS or RHS types of their containing map. In
-this case for example, the \ctr{parity} map has \lstinline{Nat * Bool}
-as its LHS and RHS type while the type of the label \kw{iter} has the
-type \lstinline{Nat * Nat * Bool}. Labels temporarily give us a
-different \emph{view}~\cite{Wadler:1987:VWP:41625.41653} of a value,
-by changing it type. Labels act as goto statements from one side to
-the other -- when a label is encountered on the RHS of a map, the
-argument to the label is matched by some definition of the label
-pattern on the LHS of the map, and vice versa.  Here is a trace of the
-execution of \ctr{parity} as it transforms input \ctr{Succ (Succ (Succ
-  0)), False} to the output \ctr{Succ (Succ (Succ 0)), True}. It takes
-five pattern match and rewrite steps to complete.
-
-\lstset{
-        language=haskell,
- morekeywords={iter,Nat},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-Succ (Succ (Succ 0)), False           (@$\longmapsto$@) iter $ Succ (Succ (Succ 0)), 0, False
-iter $ Succ (Succ (Succ 0)), 0, False (@$\longmapsto$@) iter $ Succ (Succ 0), Succ 0, True
-iter $ Succ (Succ 0), Succ 0, True    (@$\longmapsto$@) iter $ Succ 0, Succ (Succ 0), False
-iter $ Succ 0, Succ (Succ 0), False   (@$\longmapsto$@) iter $ 0, Succ (Succ (Succ 0)), True
-iter $ 0, Succ (Succ (Succ 0)), True  (@$\longmapsto$@) Succ (Succ (Succ 0)), True
- \end{lstlisting}
-%subcode source theseus.tex:784
-
-A trace operator can be expressed using labels and the definition closely
-follows the expected iteration semantics of \ensuremath{\mathit{trace}} in our previous
-work~\cite{James:2012:IE:2103656.2103667}.  Using \ctr{trace} one can define
-\ctr{add1} on \kw{Nat} such that \ctr{sub1} of {\scriptsize{0}}
-diverges.
-
-\begin{multicols}{2}
-\lstset{
-        language=haskell,
- morekeywords={enter,leave},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-trace :: f:(a + b (@$\leftrightarrow$@) a + c) (@$\rightarrow$@) b (@$\leftrightarrow$@) c
-| b               (@$\leftrightarrow$@) enter $ (@\ctr{Right}@) b
-| enter $ a       (@$\leftrightarrow$@) leave $ (f a)
-| leave $ (@\ctr{Left}@) a  (@$\leftrightarrow$@) enter $ (@\ctr{Left}@) a
-| leave $ (@\ctr{Right}@) c (@$\leftrightarrow$@) c
-  where enter :: a + b
-        leave :: a + c
- \end{lstlisting}
-%subcode source theseus.tex:803
-
-\lstset{
-        language=haskell,
- morekeywords={Nat},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-addSub :: Nat + Nat (@$\leftrightarrow$@) Nat + Nat
-| (@\ctr{Left}@) (Succ n) (@$\leftrightarrow$@) (@\ctr{Left}@) n
-| (@\ctr{Left}@) 0        (@$\leftrightarrow$@) (@\ctr{Right}@) 0
-| (@\ctr{Right}@) n       (@$\leftrightarrow$@) (@\ctr{Right}@) (Succ n)
-
-add1 :: Nat (@$\leftrightarrow$@) Nat :: sub1
-| n (@$\leftrightarrow$@) trace ~f:addSub n
- \end{lstlisting}
-%subcode source theseus.tex:815
-\end{multicols}
-\vspace{-10pt}
-
-Here are a few more examples that use labels.  Combinators \ctr{iterN} and
-\ctr{add} are ones you would expect to find in a standard library. The
-combinator \ctr{iterN} loops any given map \ctr{n} times and \ctr{add} adds
-the second argument to the first:
-\vspace{-10pt}
-\begin{multicols}{2}
-\lstset{
-        language=haskell,
- morekeywords={Nat,iter},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
--- runs f some n-times
-iterN :: f:(a(@$\leftrightarrow$@)a) (@$\rightarrow$@) Nat*a (@$\leftrightarrow$@) Nat*a
-| n, a                 (@$\leftrightarrow$@) iter $ n, 0, a
-| iter $ 0, n, a       (@$\leftrightarrow$@) n, a
-| iter $ Succ n, m, a  (@$\leftrightarrow$@)
-    iter $ n, Succ m, f a
- where iter :: Nat * Nat * a
-
--- add (x, y) = (x+y, x)
-add : Nat * Nat (@$\leftrightarrow$@) Nat * Nat
-| x, y (@$\leftrightarrow$@) iter $ y, 0, x
-| iter $ a, b, 0       (@$\leftrightarrow$@) a, b
-| iter $ a, b, Succ n  (@$\leftrightarrow$@)
-       iter $ add1 a, add1 b, n
-  where iter :: Nat * Nat * Nat
- \end{lstlisting}
-%subcode source theseus.tex:844
-\end{multicols}
-
-\vspace{-20pt} As another example, here is definition of the Fibonacci
-function where \ctr{fibonacci} applied to \lstinline{(n, (1, 0))} returns
-\lstinline{(n, (x, y))} where \ctr{x} is the $(n+1)$-st Fibonacci number and
-\ctr{y} is the previous one:
-
-\lstset{
-        language=haskell,
- morekeywords={Nat},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-swapAndAdd :: Nat * Nat (@$\leftrightarrow$@) Nat * Nat
-| x, y (@$\leftrightarrow$@) add (y, x)
-
-fibonacci :: Nat * (Nat * Nat) (@$\leftrightarrow$@) Nat * (Nat * Nat)
-| n, (x, y) (@$\leftrightarrow$@) iterN ~f:swapAndAdd (n, (x, y))
- \end{lstlisting}
-%subcode source theseus.tex:860
-
-We conclude this section with a somewhat richer example of tree traversal
-that shows how we can work with general recursive types. Here we use two
-labels \kw{walk} and \kw{reconst} and one can verify that each label
-independently covers its type on both the LHS and the RHS. The combinator
-\ctr{treeWalk} walks down a tree and applies a given \ctr{f} to every leaf:
-
-\lstset{
-        language=haskell,
- morekeywords={Nat,Tree,Ctxt,walk,reconst},
-	keywordstyle=\bfseries\ttfamily,
-	identifierstyle=\ttfamily,
-	commentstyle=\ttfamily,
-	stringstyle=\ttfamily,
-	showstringspaces=false,
-	basicstyle=\scriptsize\sffamily,
-	tabsize=2,
-	breaklines=false,
-	breakatwhitespace=false,
-	aboveskip={1.2\baselineskip},
-        columns=fixed,
-        extendedchars=true,
-}
-\begin{lstlisting}[escapeinside={(@}{@)}]
-type Ctxt = Empty | L Ctxt Tree | R Tree Ctxt
-
-treeWalk :: ~f:(Nat (@$\leftrightarrow$@) Nat) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree
-| tr                       (@$\leftrightarrow$@) walk $ tr, Empty
-| walk $ Leaf b, ctxt      (@$\leftrightarrow$@) reconst $ ctxt, Leaf (f b)
-| walk $ Node b1 b2, ctxt  (@$\leftrightarrow$@) walk $ b1, L ctxt b2
-| reconst $ Empty, tr      (@$\leftrightarrow$@) tr
-| reconst $ L ctxt b2, b1  (@$\leftrightarrow$@) walk $ b2, R b1 ctxt
-| reconst $ R b1 ctxt, b2  (@$\leftrightarrow$@) reconst $ ctxt, Node b1 b2
-  where walk     :: Tree * Ctxt
-        reconst  :: Ctxt * Tree
- \end{lstlisting}
-%subcode source theseus.tex:882
-
-\noindent
-Compiling programs with labels to \ensuremath{\Pi^{o}} is a straightforward extension of
-compiling label-free programs. For any Theseus map \ctr{c : a}
-$\leftrightarrow$ \ctr{b} that uses an iteration label \kw{label} \ctr{: lab}
-we construct an intermediate type \ctr{ti} corresponding to the pattern
-clauses as before. We can then construct the isomorphisms \ctr{c1 : lab + a}
-$\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti} $\leftrightarrow$ \ctr{lab + b}
-since every LHS pattern clause corresponds to either \ctr{a} (the input) or
-\ctr{lab} (the label) and every RHS clause corresponds to either \ctr{b} (the
-output) or \ctr{lab} (the label). The resulting \ctr{c1} $\odot$ \ctr{c2}
-has type \lstinline{lab+a} $\leftrightarrow$ \lstinline{lab+b} and the
-required compilation of \ctr{c} is given by \ensuremath{\mathit{trace}(} \ctr{c1} $\odot$
-\ctr{c2} \ensuremath{)} \ctr{: a} $\leftrightarrow$ \ctr{b}. This extends to multiple
-by labels by constructing \lstinline{c1 : (lab1 + lab2 + ...) + a}
-$\leftrightarrow$ \ctr{ti} and \lstinline{c2 : ti} $\leftrightarrow$
-\lstinline{(lab1 + lab2 + ...) + b} to include the types of the additional
-labels.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-\label{conc}
-
-Theseus is as expressible as \ensuremath{\Pi^{o}} and preserves its properties. All
-\ensuremath{\Pi^{o}} programs are expressible as Theseus programs and vice-versa. In other
-words, \ensuremath{\Pi^{o}}, or dagger symmetric traced bimonoidal categories serve as a
-fully abstract model for Theseus. Despite being closely related to \ensuremath{\Pi^{o}},
-programmers can use Theseus without needing to understand \ensuremath{\Pi^{o}}, strings
-diagrams or category theory.
-
-Theseus admits non-terminating computations. Just the way that partial
-functions refer to maps that may not be defined on all inputs, the set
-of computations expressible in Theseus are referred to as
-\emph{partial isomorphisms}. As with \ensuremath{\Pi^{o}}, Theseus programs satisfy
-\emph{logical reversibility} (see Def.~\ref{logical-rev}).
-
-Many of the trimmings of more mainstream languages can be added to
-Theseus. Theseus could use an error reporting mechanism for programs to exit
-to without resorting to divergence. Theseus needs an FFI mechanism to take
-advantage of third-party libraries. We imagine that in such cases the
-programmer would write out an explicit extern declaration asserting the type
-that Theseus should assume the external library has. More generally, Theseus
-can adopt Agda's approach of piggy backing on top of Haskell. We can have an
-FFI that lets us drop into Haskell and use Haskell API. We could also import
-user-certified pairs of Haskell functions as isomorphisms into Theseus.
-
-There are few natural extensions to Theseus that require additional research,
-though preliminary investigation seems promising. Theseus currently treats
-all values the same and does not classify values as heap and garbage.
-Theseus can be extended with a framework for encapsulating effects. In other
-words, add the effectful arrow type \ensuremath{\leadsto} in a principled way such that
-information effects, and possibly other effects, may be
-expressed~\cite{James:2012:IE:2103656.2103667}.  It is also desirable to add
-other programming features such as polymorphism, higher-order maps and
-inductive definitions over recursive datatypes without explicitly managing
-\ctr{Ctxt} values as we did in the \ctr{treeWalk} program.
-}

### 2018-07-07 survey/survey.tex
making a final pass

@@
   include logical operations like conjunction, disjunction and
-  negation, as well as reading and writing to a large collection to
+  negation, as well as reading and writing to a large collection of
   memory locations. This perspective conforms to a macro-level view of
@@
 calls to re-think this foundational question have been proclaimed by
-physicists more than forty years ago as the following two quotes
-testify:
+physicists almost forty years ago as the following two quotes testify:
 
@@
 transpositions. Thus we can create a small language for writing
-permutations on $\fin{n}$ as
+permutations on $\fin{n}$ as:
 \[\begin{array}{rcl}
@@
 bit. In the notation used to describe the circuit, the $\times$ denotes
-boolean negation, the dots are control bits. In our reversible
+boolean negation and the dots are control bits. In our reversible
 language, we can express this circuit as the following permutation of
@@
 \begin{prop}
-The structure $\left(\Nat, \false, \true, \lor, \land\right)$
+The structure $\left(\{\false,\true\}, \false, \true, \lor, \land\right)$
 is a commutative semiring.
@@
 items). Our next instinct would be to similarly replace $=$ with a
-type $A \equiv B$ that asserts that $A$ and $B$ are \emph{propositionally
-equal}, i.e. reduce to equivalent type-denoting expressions under the rules of
-the host type system.  This is however not true: the
-proposition $A \times B \equiv B \times A$ is not
+type $A \equiv B$ that asserts that $A$ and $B$ are
+\emph{propositionally equal}, i.e. reduce to equivalent type-denoting
+expressions under the rules of the host type system.  This is however
+not true: the proposition $A \times B \equiv B \times A$ is not
 normally\footnote{Except in univalent type theory where equivalent
@@
 equivalent information. In other words, we would like to be able to
-witness that $A \times B$ can be reversibly deformed into $B \times
-A$, and vice-versa. Which motivates the introduction of type \emph{equivalences}. To
-do this, we need a few important auxiliary concepts.
+witness that $A \times B$ can be reversibly deformed into
+$B \times A$, and vice-versa, which motivates the introduction of type
+\emph{equivalences}. To do this, we need a few important auxiliary
+concepts.
 
@@
 
-In Martin-L\"{o}f Type Theory, normal forms mean $\beta$-reduced
-$\eta$ long terms under $\alpha$-equivalence. In other words,
-expressions are evaluated as much as possible ($\beta$-reduced),
-all functions are fully applied ($\eta$ long), and the exact names
-of bound variables are irrelevant ($\alpha$-equivalence). Note that
-the above definition applies equally well to expressions that denote
-values and expressions that denote types.
+In Martin-L\"{o}f Type Theory, normal forms mean $\beta\eta$-long
+normal forms under $\alpha$-equivalence. In other words, expressions
+are evaluated as much as possible ($\beta$-reduced), all functions are
+fully applied ($\eta$-long), and the exact names of bound variables
+are irrelevant ($\alpha$-equivalence). Note that the above definition
+applies equally well to expressions that denote values and expressions
+that denote types.
 
@@
 taking sums and products, which led to the awkward non-compositional
-programming style illustrated in the full adder example~\ref{eq:adder}.
+programming style illustrated in the full adder
+example~(Eq.~\ref{eq:adder}).
 
 Thus the denotation of the $\Pi$ terms \emph{should} be permutations. But
-given types $A$ and $B$ denoting $\fin{m}$ and $\fin{n}$ respectively,
+given types $A$ and $B$ denoting $\fin{m}$ and~$\fin{n}$ respectively,
 what are $A \uplus B$ and $A \times B$ ?  They correspond exactly to
@@
 concatenation for $A + B$, i.e. lining up the elements of $A$ first,
-and then those of $B$. For $A * B$, one can picture this as
+and then those of~$B$. For $A * B$, one can picture this as
 lining up the elements of $A$ horizontally, those of $B$ vertically
@@
 
-Since there are no values that have the type \ensuremath{0}, the reductions for
-the combinators \identlp, \identrp, \identlsp, \identrsp omit the impossible
-cases. \factorzr and \factorzl likewise do not appear as they have not
-possible cases at all. However, \absorbr and \absorbl are treated slightly
-differently: rather than \emph{eagerly} assuming they are impossible,
-the purported inhabitant of $0$ given on one side is passed on to the other
-side. The reason for this choice will have to wait for Sec.~\ref{langeqeq}
-when we explain some higher-level symmetries (see Fig.~\ref{figc}).
+Since there are no values that have the type \ensuremath{0}, the
+reductions for the combinators \identlp, \identrp, \identlsp, and
+\identrsp\ omit the impossible cases. \factorzr\ and \factorzl\
+likewise do not appear as they have no possible cases at all. However,
+\absorbr\ and \absorbl\ are treated slightly differently: rather than
+\emph{eagerly} assuming they are impossible, the purported inhabitant
+of $0$ given on one side is passed on to the other side. The reason
+for this choice will have to wait for Sec.~\ref{langeqeq} when we
+explain some higher-level symmetries (see Fig.~\ref{figc}).
 
@@
   \item For primitive isomorphisms \ensuremath{c}, \ensuremath{!~ c} is given by its
-    inverse from Figures~\ref{pi-terms} and~\ref{more-pi}.
+    inverse from Figs.~\ref{pi-terms} and~\ref{more-pi}.
 
@@
 \noindent We can further define that two combinators are
-\emph{observationally equivalent} if only all values of their
-common domain, they evaluate to identical values.  More precisely,
-we will say that for combinators $c_1, c_2 : b_1 \leftrightarrow b_2$,
-$c_1~=~c_2$ whenever
+\emph{observationally equivalent} if on all values of their common
+domain, they evaluate to identical values.  More precisely, we will
+say that for combinators $c_1, c_2 : b_1 \leftrightarrow b_2$,
+$c_1~=~c_2$ whenever:
 \[
-  \forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \text{if and only if\ } c_2 ~v_1 \mapsto v_2
+  \forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 ~\text{if and only if\ } c_2 ~v_1 \mapsto v_2
 \]
 
-Previously we had seen that types $b$ correspond to natural numbers,
-more precisely~$|b|$, the size of $b$.  We had previously established that 
-for any natural number $n$, there is a canonical set of size $n$,
-which we denoted $[n]$. Furthermore, we can also define a
-canonical \emph{type} that size, which we will denote $\sharp b$,
-i.e. $\sharp b$ is a canonical type of size $|b|$.
+Each type $b$ has a size $|b|$ defined in the obvious way. We had
+previously established that for any natural number $n$, there is a
+canonical set of size $n$, which we denoted $[n]$. Furthermore, we can
+also define a canonical \emph{type} of that size, which we will denote
+$\sharp\, b$, i.e. $\sharp\, b$ is a canonical type of size $|b|$.
 
@@
 \[\begin{array}{rcl}
-  \tau~ 0 & = & \bot \\
-  \tau~ (1 + n) & = & \top \uplus \tau~ n \\
+  \tau~ (0) & = & \bot \\
+  \tau~ (1 + n) & = & \top \uplus \tau~ (n) \\
  \end{array}\]
-\noindent so that we can define $\sharp b = \tau~ |b|$.
+\noindent so that we can define $\sharp\, b = \tau~ |b|$.
 \end{definition}
@@
 We are now ready to go further and establish
-that there is always an equivalence between.
+that there is always an equivalence between a type and the canonical
+type of the same size.
 
 \begin{proposition}
-  For any type \ensuremath{b} there exists an isomorphism \ensuremath{b \leftrightarrow \sharp b}.
+  For any type \ensuremath{b} there exists an isomorphism \ensuremath{b \leftrightarrow \sharp\, b}.
   \begin{proof}
@@
  \sem{1} & = & \idc \\
- \sem{1+b} & = & \idc \oplus \sem{b} \\
+ \sem{1+b} & = & \idc \oplus\ \sem{b} \\
  \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
- \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp b_1 + b_2 } \\
+ \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp\, b_1 + b_2 } \\
  \sem{0 * b_2} & = & \absorbr \\
@@
 \ensuremath{\Pi } combinators expressed in this graphical language
-look like ``wiring diagrams''. Values take the form of ``particles''
+look like ``wiring diagrams.'' Values take the form of ``particles''
 that flow along the wires. Computation is expressed by the flow of
@@
 The product type \ensuremath{b_1 * b_2} may be represented both as one wire
-labeled \ensuremath{b_1 * b_2} or by two parallel wires labeled \ensuremath{b_1} and
+labeled \ensuremath{b_1 \times b_2} or by two parallel wires labeled \ensuremath{b_1} and
 \ensuremath{b_2}. Both representations may be used interchangeably.
@@
 There are two forms of parallel composition -- combinators
-  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
-  multiplicatively \ensuremath{c_1 * c_2} (shown on the right).
+  can be combined additively \ensuremath{c_1 \oplus\ c_2} (shown on the left) or
+  multiplicatively \ensuremath{c_1 \otimes c_2} (shown on the right).
 
@@
 of the combinator \ensuremath{c} below:
-\begin{flalign*}
-& \begin{aligned}
+\[\begin{array}{rcl}
 c & : & b * (1+1) \leftrightarrow b + b \\
-c & = & \swapt \odot \dist \odot (\identlt + \identlt) \\
-\end{aligned} &&
-\end{flalign*}
+c & = & \swapt \odot \dist \odot (\identlt \oplus \identlt) 
+\end{array}\]
 
@@
 
-Figure~\ref{type-isos} introduces our desired denotational
-semantics, and section~\ref{sec:opsem} a direct definition
-of an operational semantics. One obvious question arises:
-do these correspond?
+Fig.~\ref{type-isos} introduces our desired denotational semantics,
+and Sec.~\ref{sec:opsem} is a direct definition of an operational
+semantics. One obvious question arises: do these correspond?
 
@@
 \footnote{This is extracted from the Agda formalization
-of this work, which has been reported on in~\cite{Carette2016}.}:
+of this work, which has been reported on in a previous paper~\cite{Carette2016}.}:
 \begin{code}\hspace*{-4mm}
@@
 We will however show that one can program all logical gates in
-$\Pi$. We will start with a few simply examples and then discuss the
-expressiveness of the language and discuss various properties.
+$\Pi$. We will start with a few simple examples and then discuss the
+expressiveness of the language and its properties.
 
@@
 
-\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  *  \idc)~ \odot \assocrt}
+\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  \otimes  \idc)~ \odot \assocrt}
 
@@
 derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
-many others we will see in this thesis, is formed by sequentially
+many others we will see in this paper, is formed by sequentially
 composing several simpler combinators. Instead of presenting the
@@
  \swapt & ((v_2, v_3), v_1) \\
- \swapt  *  \idc & ((v_3, v_2), v_1) \\
+ \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
  \assocrt & (v_3, (v_2, v_1)) \\
@@
 
-\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  *  c_1) + (\idc  *  c_2)) \odot \factor}
+\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) + (\idc \otimes\  c_2)) \odot \factor}
 

### 2018-07-07 survey/survey.tex
finished with final pass

@@
 \usepackage{url}
-\usepackage{comment}
 \usepackage{tikz}
@@
 \newcommand{\refl}[1]{\textsf{refl}~#1}
-%% \newcommand{\lid}{\textsf{lid}}
 \newcommand{\alt}{~|~}
-%% \newcommand{\rid}{\textsf{rid}}
 \newcommand{\linv}{l!}
@@
 \newcommand{\factorl}{\ensuremath{\mathit{factorl}}}
-%% \newcommand{\distz}{\mathit{absorbr}}
 \newcommand{\iso}{\leftrightarrow}
@@
 
-%%  Something about Theseus?
 \end{abstract}
-
-% * Reversibility intro / motivation as we  have now more or less
-
-% * Thesis statement: programs are reversible deformations on data /
-%    spaces, almost necessarily by definition
-
-% * Focus now is on data
-
-% * If data is plain finite sets, programs become permutations (deformations on finite sets).
-
-% * If data is structured trees, we get Pi.
-
-% * Explain Pi with examples.
-
-% * If data is itself deformations-on-finite-sets, then the
-%   deformations between them become something quite
-%   interesting. Explain Pi level 2 with examples.
-
-% * Conclude with thoughts regarding other kinds of data that can be
-%   plugged in into that story.
 
@@
 
-% \jc{The notation here is somewhat inconsistent with the above,
-% mainly in the notation for values.}
-
 To give an operational semantics to $\Pi$, we are mainly missing
@@
 
-%% Random idea:
-%% for a product monoidal this is wave-style
-%% for a sum monoidal this is particle-style
-%% when we have both we do have a wave-particle duality
-
 Combinators of \ensuremath{\Pi } can be written in terms of the
@@
 \end{center}
-
-\begin{comment}
-\jc{An additional ``problem'' is that since our circuits have a
-left-to-right orientation, $A * B$ drawn as boxes left-to-right
-gives an implicit ordering that is going to be hard to ignore.
-I'm starting to think that the diagrams in the MO answer might
-be better: use a colour (or like in Selinger's paper, at the
-very end), symbols in the nodes that split.  So both would
-be displayed as vertical stacks (only). Because I think that
-swap* is too hard to justify in this notation. Of course,
-if we could do things in 3D and $*$ was displayed using
-(say) depth, that would be different again.}
-
-This idea leads to a resource-based interpretation of types as
-\emph{possible events} and of the values as \emph{consistent timelines
-  of events} as explained below. The event $1$ is trivial, and the
-event $0$ is impossible.  We visualize the denotation of $1$ as a
-point in space, and $0$ as empty space. The denotation of $A + B$ is
-the denotation of~$A$ drawn on top of the denotation of $B$ and
-separated by a border, and the denotation of $A * B$ is just the
-denotation of $A$ written next to the denotation of $B$.  For example,
-let $2$ be the type $1 + 1$ represented by a stack of two boxes each
-with a point inside it. The type $(2 + 2) * (2 + 2)$ would then be two
-such stacks laid next to each other. Composing further, the type
-$((2 + 2) * (2 + 2)) + ((2 + 2) * (2 + 2))$ is built by stacking the
-last two diagrams as shown below on the left:
-\begin{center}
-\hfill
-\includegraphics[scale=0.3]{images/tall-tl-1}
-~
-\includegraphics[scale=0.3]{images/tall-tl-2}
-\hfill (*)
-\end{center}
-If we call the type on the left $A$ then the full diagram above
-represents the type $A*A$.
-
-To understand how diagrams (types) can transform through the
-evaluation of  a program, we need to reason about the equivalence of
-different diagrams. Consider for example, the two diagrams below:
-
-\begin{center}
-\hfill
-\includegraphics[scale=0.3]{images/distrib-1}
-\qquad \qquad \qquad \qquad \qquad
-\includegraphics[scale=0.3]{images/distrib-2}
-\hfill (**)
-\end{center}
-
-The diagram on the left represents the type $A * (B + C)$ while the
-one on the right represents the type $(A * B) + (A * C)$. These two
-types are equivalent in $\Pi$ but why are they equivalent in the
-graphical notation? We argue that they admit the same consistent
-timelines where a consistent timeline is defined recursively on the
-structure of a diagram:
-\begin{itemize}
-\item An endpoint (the denotation of $1$) has exactly one timeline
-  through it.  Empty space (the denotation of $0$) has no timelines
-  through it.
-\item A timeline through a stack of alternatives
-  $A_1 + A_2 + \cdots + A_n$ is a timeline through exactly one of the
-  possible alternatives.
-\item A timeline through a spatial composition of cells
-  $A_1 * A_2 * \cdots * A_n$ is a timeline through \emph{all} of the
-  cells, in some order.
-\end{itemize}
-
-For example, the red line in the diagram (*) represents a consistent
-timeline. For the diagrams (**), we note that any consistent timeline
-for the left diagram must go through $A$ and either one of $B$ or $C$
-and for the diagram on the right, the choice happens earlier but the
-result is the same: any consistent timeline must go through either $A$
-and $B$ or through $A$ and $C$.
-
-Other than distributivity, $\Pi$ combinators simply allow us to shuffle
-the components of a stack around as well as shuffle the positions of
-components in the plane.
-\end{comment}
-% \amr{
-%   Let’s have a geometric representation of spaces:
-% - A + B means putting A next to B
-% - A * B means putting A and B on two perpendicular axes and filling
-% the square (volume) between them
-% All the combinators would indeed have a natural action and if we have
-% a few nice pictures it would be great. What do you think? }
-
-% \newcommand{\graphzero}{}
-% \newcommand{\graphone}{\ensuremath{\bullet}}
-% \newcommand{\graphplus}[2]{\node (top) {#1}; \node [below of=top] {#2};}
-
-% \begin{center}
-% \begin{tikzpicture} %% [declare function = {plus(\x,\y) = \node (top){\x}; \node [below of=top] {\y};}]
-% \node (x) {\graphone};
-% \node (y) {\graphone};
-% %% {plus(\x,\y)}
-% \end{tikzpicture}
-% \end{center}
-
-% \begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
-% \label{def:primitives-langRev}
-
-% \[\begin{array}{rrcll}
-%  \identlp :& 0 + b &\leftrightarrow & b &: \identrp \\
-%  \swapp :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \swapp \\
-%  \assoclp :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \assocrp \\
-%  \identlt :& 1 \times b &\leftrightarrow & b &: \identrt \\
-%  \swapt :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \swapt \\
-%  \assoclt :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \assocrt \\
-%  \absorbr :& 0 \times b &\leftrightarrow & 0 &: \factorzl \\
-%  \dist :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \factor \\
-%  \end{array}\]
-% %subcode source isomorphisms.tex:427
-
-% \end{definition}
-
-% Each line of this table is to be read as the definition of one or two
-% operators. For example, corresponding to the \textit{identity of
-%   \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
-% \ensuremath{\identlt:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
-% \ensuremath{\identrt:b\leftrightarrow 1\times b} (reading the isomorphism from right to
-% left). These operators are inverses of each other.  Each of the two
-% cases of commutativity defines one operator that is its own inverse.
-
-% Having defined primitive operators, we need some means of composing
-% them. We construct the composition combinators out of the closure
-% conditions for isomorphisms.
-
-% \begin{definition}{Composition in \ensuremath{\Pi } }
-% $$
-% \infer{ \idc : b \leftrightarrow b}{
-% 	 ~
-% }
-% \quad
-% \infer{ \ !~c : b_2 \leftrightarrow b_1}{
-% 	 c : b_1 \leftrightarrow b_2
-% }
-% \quad
-% \infer{ c_1\odot c_2 : b_1 \leftrightarrow b_3}{
-% 	 c_1 : b_1 \leftrightarrow b_2
-% 	&
-% 	 c_2 : b_2 \leftrightarrow b_3
-% }
-% $$
-% $$
-% \infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
-% 	 c_1 : b_1 \leftrightarrow b_3
-% 	&
-% 	 c_2 : b_2 \leftrightarrow b_4
-% }
-% \quad
-% \infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
-% 	 c_1 : b_1 \leftrightarrow b_3
-% 	&
-% 	 c_2 : b_2 \leftrightarrow b_4
-% }
-% $$
-% %subcode source isomorphisms.tex:462
-% \end{definition}
-
-% Thus we have program constructs that witness reflexivity \ensuremath{\idc},
-% symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\odot} and two parallel
-% composition combinators, one for sums~\ensuremath{+} and one for
-% pairs~\ensuremath{\times}.
 
@@
   \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
-%%  \node at (3.3,-2) {$\mathit{swapl}_{+\Leftrightarrow}$} ;
-%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\odot~\mathit{swap}_{+})$};
-%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\odot~(c_1~\oplus~c_2))$};
-%  \draw (-2,-2) ellipse (0.5cm and 1cm);
   \draw[fill] (-2,-1.5) circle [radius=0.025];
@@
 
-%  \draw (6.5,-2) ellipse (0.5cm and 1cm);
   \draw[fill] (6.5,-1.5) circle [radius=0.025];
@@
   \path (1.5,0.5) node (tc1) [func] {$c_1$};
-  %\draw     (1,0.9)  -- (2,0.9) -- (2,0.1) -- (1,0.1) -- cycle ;
-  %\node at (1.5,0.5) {$c_1$};
-
   \path (1.5,-0.5) node (tc2) [func] {$c_2$};
-  %\draw     (1,-0.1)  -- (2,-0.1) -- (2,-0.9) -- (1,-0.9) -- cycle ;
-  %\node at (1.5,-0.5) {$c_2$};
-
   \path (3,-4.5) node (bc1) [func] {$c_1$};
-  %\draw     (2.5,-4.1)  -- (3.5,-4.1) -- (3.5,-4.9) -- (2.5,-4.9) -- cycle ;
-  %\node at (3,-4.5) {$c_1$};
-
   \path (3,-3.5) node (bc2) [func] {$c_2$};
-  %\draw     (2.5,-3.1)  -- (3.5,-3.1) -- (3.5,-3.9) -- (2.5,-3.9) -- cycle ;
-  % \node at (3,-3.5) {$c_2$};
-
 \end{tikzpicture}
@@
 
-Previously we saw how we could take the proof terms of commutative semirings
+Previously we saw how we could take the proof terms of commutative semiring
 equivalences as our starting point for $\Pi$. What we need
@@
 
-Thus, rather than looking at (untyped) commutative semirings,
-we should look at a \emph{typed} version. This process frequently
-goes by the moniker of ``categorification.''  We want a categorical
-algebra, where the basic objects are groupoids (to model our types),
-and where there is a natural notion of $+$ and $*$.  At first,
-we hit what seems like a serious stumbling block: the category of
-all groupoids, \Gpd, does not have either co-products
-nor products. However, we don't want to work internally in
-\Gpd -- we want operations
-\emph{on} groupoids. In other words, we want something akin to
-symmetric monoidal categories, but with two interacting
-monoidal structures.  Luckily, this already exists: the categorical
-analog to (commutative) semirings are (symmetric) Rig
-Categories~\cite{laplaza72,kelly74}.
-This straightforwardly generalizes to symmetric Rig Groupoids.
+Thus, rather than looking at (untyped) commutative semirings, we
+should look at a \emph{typed} version. This process frequently goes by
+the moniker of ``categorification.''  We want a categorical algebra,
+where the basic objects are groupoids (to model our types), and where
+there is a natural notion of $+$ and $*$.  At first, we hit what seems
+like a serious stumbling block: the category of all groupoids, \Gpd,
+have neither co-products nor products. However, we don't want to work
+internally in \Gpd -- we want operations \emph{on} groupoids. In other
+words, we want something akin to symmetric monoidal categories, but
+with two interacting monoidal structures.  Luckily, this already
+exists: the categorical analog to (commutative) semirings are
+(symmetric) Rig Categories~\cite{laplaza72,kelly74}.  This
+straightforwardly generalizes to symmetric Rig Groupoids.
 
@@
 like the associator $\alpha$ and the left and right unitors ($\lambda$
-and $\rho$ respectively), but also coherence conditions that these must satisfy.
+and~$\rho$ respectively), but also coherence conditions that these must satisfy.
 Looking, for example, at just the additive fragment of $\Pi$ (i.e. with just $0$,
@@
 Starting with the simplest constructions, the first two constructs in
-Fig.~\ref{fige} are the level-2 analogs of $+$ and $*$, which respectively
-model level-1 choice composition and parallel composition (of
-equivalences).  These allow us to ``build up'' larger equivalences from smaller
-ones.  The next two express that both of these composition operators distribute
-over sequential composition $\odot$ (and vice versa).
+Fig.~\ref{fige} are the level-2 analogs of~$+$ and~$*$, which
+respectively model level-1 choice composition and parallel composition
+(of equivalences).  These allow us to ``build up'' larger equivalences
+from smaller ones.  The next two express that both of these
+composition operators distribute over sequential composition $\odot$
+(and vice versa).
 
@@
 To make the relation between $\Pi$ as a language and the
-language of category theory, the figure below
+language of category theory, the figure below displays
 the same morphism but in categorical terms.
@@
 
-\begin{comment}
-For example, the last equivalence captures the idea that the
-program $\assocrt \odot \assocrt$ transforms an input of
-type $((t_1 * t_2) * t_3) * t_4$ to an output of type
-$t_1 * (t_2 * (t_3 * t_4))$ as follows:
-
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_2$};
-\node[above] (C) at (-1,0) {$t_3$};
-\node[above] (D) at (-0.5,0) {$t_4$};
-\draw[thick] (A) -- (-2,-1.5);
-\draw[thick] (B) -- (-1.5,-1.5);
-\draw[thick] (C) -- (-1,-1.5);;
-\draw[thick] (D) -- (-0.5,-1.5);
-\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[green,densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\end{tikzpicture}
-$\leadsto$
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_2$};
-\node[above] (C) at (-1,0) {$t_3$};
-\node[above] (D) at (-0.5,0) {$t_4$};
-\draw[thick] (A) -- (-2,-1.5);
-\draw[thick] (B) -- (-1.5,-1.5);
-\draw[thick] (C) -- (-1,-1.5);;
-\draw[thick] (D) -- (-0.5,-1.5);
-\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[green,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\end{tikzpicture}
-$\leadsto$
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_2$};
-\node[above] (C) at (-1,0) {$t_3$};
-\node[above] (D) at (-0.5,0) {$t_4$};
-\draw[thick] (A) -- (-2,-1.5);
-\draw[thick] (B) -- (-1.5,-1.5);
-\draw[thick] (C) -- (-1,-1.5);;
-\draw[thick] (D) -- (-0.5,-1.5);
-\draw[red,densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\draw[green,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\end{tikzpicture}
-\end{center}
-% \[
-% ((t_1 * t_2) * t_3) * t_4 \leadsto
-% (t_1 * t_2) * (t_3 * t_4) \leadsto
-% t_1 * (t_2 * (t_3 * t_4))
-% \]
-
-\noindent The program on the right
-$((\assocrt \otimes \idc) \odot \assocrt) \odot (\idc \otimes
-\assocrt)$ achieves the same end-to-end transformation using the
-following sequence of steps:
-
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_2$};
-\node[above] (C) at (-1,0) {$t_3$};
-\node[above] (D) at (-0.5,0) {$t_4$};
-\draw[thick] (A) -- (-2,-1.5);
-\draw[thick] (B) -- (-1.5,-1.5);
-\draw[thick] (C) -- (-1,-1.5);;
-\draw[thick] (D) -- (-0.5,-1.5);
-\draw[red,densely dashed] (-1.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[green,densely dashed] (-1.6,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\end{tikzpicture}
-$\leadsto$
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_2$};
-\node[above] (C) at (-1,0) {$t_3$};
-\node[above] (D) at (-0.5,0) {$t_4$};
-\draw[thick] (A) -- (-2,-1.5);
-\draw[thick] (B) -- (-1.5,-1.5);
-\draw[thick] (C) -- (-1,-1.5);;
-\draw[thick] (D) -- (-0.5,-1.5);
-\draw[red,densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[green,densely dashed] (-1.5,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\end{tikzpicture}
-$\leadsto$
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_2$};
-\node[above] (C) at (-1,0) {$t_3$};
-\node[above] (D) at (-0.5,0) {$t_4$};
-\draw[thick] (A) -- (-2,-1.5);
-\draw[thick] (B) -- (-1.5,-1.5);
-\draw[thick] (C) -- (-1,-1.5);;
-\draw[thick] (D) -- (-0.5,-1.5);
-\draw[red,densely dashed] (-1.25,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[green,densely dashed] (-1.0,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\end{tikzpicture}
-$\leadsto$
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8},baseline=(current bounding box.center)]]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_2$};
-\node[above] (C) at (-1,0) {$t_3$};
-\node[above] (D) at (-0.5,0) {$t_4$};
-\draw[thick] (A) -- (-2,-1.5);
-\draw[thick] (B) -- (-1.5,-1.5);
-\draw[thick] (C) -- (-1,-1.5);;
-\draw[thick] (D) -- (-0.5,-1.5);
-\draw[green,densely dashed] (-0.9,-0.75) ellipse [x radius=0.85cm, y radius = 0.2cm];
-\draw[red,densely dashed] (-0.75,-0.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\end{tikzpicture}
-\end{center}
-% \[
-% ((t_1 * t_2) * t_3) * t_4 \leadsto
-% (t_1 * (t_2 * t_3)) * t_4 \leadsto
-% t_1 * ((t_2 * t_3) * t_4) \leadsto
-% t_1 * (t_2 * (t_3 * t_4))
-% \]
-
-% \jc{I am not a big fan of these pictures -- mainly because I can't give any
-% intuition based on them of why these can be smoothly deformed into one another.
-% All the ``pictures'' I can come up with [via purely formal reasoning] sit in at least
-% 5 dimensions -- one for each of the wires, and another to give room to maneuver.}
-% \end{comment}
-
 \begin{figure}[t]
@@
 and factoring has to commute with the level-1 $\oplus$ and $\otimes$.
-
-\begin{comment}
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-\node[above] (A) at (-2,0) {$t_1$};
-\node[above] (B) at (-1.5,0) {$t_3$};
-\node[above] (C) at (-1,0) {$t_5$};
-\draw[thick,yellow] (-1.5,-0.25) arc[start angle=60, end angle=120, radius=0.5cm];
-\draw[thick] (A) -- (-2,-0.5);
-\draw[thick] (B) -- (-1.5,-0.5);
-\draw[thick] (C) -- (-1,-0.5);
-\node at (-2,-0.75) {$a$};
-\node at (-1.5,-0.75) {$b$};
-\node at (-1,-0.75) {$c$};
-\draw[thick] (-2,-1) -- (-2,-1.5);
-\draw[thick] (-1.5,-1) -- (-1.5,-1.5);
-\draw[thick] (-1,-1) -- (-1,-1.5);
-\node at (-1.5,-2) {$\dist$};
-\draw[thick] (-2.5,-2.5) -- (-2.5,-3);
-\draw[thick] (-2,-2.5) -- (-2,-3);
-\draw[green,densely dashed] (-2.25,-2.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[thick] (-1,-2.5) -- (-1,-3);
-\draw[thick] (-0.5,-2.5) -- (-0.5,-3);
-\draw[green,densely dashed] (-0.75,-2.75) ellipse [x radius=0.4cm, y radius = 0.1cm];
-\draw[thick,yellow] (-1,-2.75) arc[start angle=60, end angle=120, radius=1cm];
-\end{tikzpicture}
-\end{center}
-% ((a + b) * c) ; dist <=> dist ; ((a * c) + (b * c))
-% LHS: (t1 + t3) * t5 ~> (t2 + t4) * t6 ~> (t2 * t6) + (t4 * t6)
-% RHS: (t1 + t3) * t5 ~> (t1 * t5) + (t3 * t5) ~> (t2 * t6) + (t4 * t6)
-\end{comment}
 
@@
 coherence between level-0 swapping and the level-1 combinator
-actions. The next four arise because of interactions between
-(additive and multiplicative) level-1 associativity and swapping.
-In other words, they arise as critical pairs.  For example,
-the first expresses that the two ways of going from
-$\left(A \oplus B\right) \oplus C$ to $B \oplus \left(C \oplus A\right)$
-are equivalent, with the second saying that the reverse (i.e.
-the results of applying $!$\,) also gives equivalent programs.
-The last two say the same but for the multiplicative structure.
+actions. The next four arise because of interactions between (additive
+and multiplicative) level-1 associativity and swapping.  In other
+words, they arise as critical pairs.  For example, the first expresses
+that the two ways of going from $\left(A \oplus B\right) \oplus C$ to
+$B \oplus \left(C \oplus A\right)$ are equivalent, with the second
+saying that the reverse (i.e.  the results of applying $!$\,) also
+gives equivalent programs.  The last two say the same but for the
+multiplicative structure.
 
@@
 (where we have explicitly annotated the types of $\idc$ for increased clarity)
-tells us that of the two ways of transforming from $t_1  *  0$ to $0$, 
+tells us that of the two ways of transforming from $t_1  *\  0$ to $0$, 
 namely first doing some arbitrary transformation $c$ from $t_1$ to $t_2$ and
@@
 two obvious ways of transforming $0 * 0$ into $0$, namely absorbing
-either the left or right zero --- it implies something subtle.  A
+either the left or right $0$ --- it implies something subtle.  A
 straightforward proof of $\absorbl$ which proceeds by saying that
@@
 
-%% \jc{should add some citations here}.
-
 \begin{figure}[t]
@@
 \begin{enumerate}
-\item Mimicking closely the one in the previous section, and
-thus finding explicit homotopies between the functions induced
-by the operational semantics of the level-1 combinators.
+\item Mimicking closely the one in Sec.~\ref{sec:opsem}, and thus
+  finding explicit homotopies between the functions induced by the
+  operational semantics of the level-1 combinators.
 \item Treating things more syntactically, and interpreting
@@
 Here we will focus instead of the syntactic interpretation as
-program transformers. This results in a function
+program transformers. This results in a function:
 \begin{flalign*}
@@
 such as \AgdaInductiveConstructor{distl⇔l} are one-step rewrites,
-in the same way that \distl is a constant of the base term language
+in the same way that \distl\ is a constant of the base term language
 of $\Pi$.  There is one additional wrinkle. There is naturally
@@
 enough to resolve the apparent use of a non-linear pattern.
-This is why \evalone was called ``deeply dependent''
+This is why \evalone\ was called ``deeply dependent''
 above: once the name of the combinator is given, the rest
@@
 If all expressible transformations were single-step only, this
-would hardly justify calling this an ``operational semantics'',
+would hardly justify calling this an ``operational semantics,''
 as we would hardly have a programming language. However, level-2
@@
 Fig.~\ref{fige} and two are in Fig.~\ref{figh}. The most interesting
-one is ``sequential composition'', which is the middle one at the 
+one is ``sequential composition,'' which is the middle one at the 
 bottom of Fig.~\ref{figh}. Since $\Leftrightarrow$ represents
@@
 to ``trace'' the rewrite. Thus the evaluation of
-$\transLR r_0 r_1$ where $r_0 : c_0 \leftrightarrow c_1$ and
+$\transLR~r_0~r_1$ where $r_0 : c_0 \leftrightarrow c_1$ and
 $r_1 : c_1 \leftrightarrow c_2$
-should apply \evalone to both $r_0$ and $r_1$. Furthermore,
+should apply \evalone\ to both $r_0$ and $r_1$. Furthermore,
 after applying $r_0$, we should be able to witness that the result
@@
 \noindent If we are careful in our construction of \evalone, the
-definition of $\mathit{exact}$ is quite straightforward, i.e.
-almost all cases are \AgdaInductiveConstructor{refl}.
-
-This then lets us define the $\transLR r_0 r_1$ case properly: we
-first evaluate $r_0$ and get a result combinator, witness that this result
-type is indeed exactly what we expect, and proceed to evaluate $r_1$
-where we specific that the $r_1$'s left-hand side must be
-$\evalone r_0$; we can use \AgdaKeyword{rewrite} to make this match
-$c_2$ ``on the nose'' (otherwise the call would be ill-typed).
-This then forces us to use \AgdaKeyword{rewrite} also in the
-implementation of the \transLR case in $\mathit{exact}$ before
-the proof is indeed just \AgdaInductiveConstructor{refl}.
+definition of $\mathit{exact}$ is quite straightforward, i.e.  almost
+all cases are immediately provable by reflexivity.
+
+This then lets us define the $\transLR~r_0~r_1$ case properly: we
+first evaluate $r_0$ and get a result combinator, witness that this
+result type is indeed exactly what we expect, and proceed to evaluate
+$r_1$ where we specify that the $r_1$'s left-hand side must be
+$\evalone\ r_0$; we can use the Agda keyword \AgdaKeyword{rewrite} to
+make this match $c_2$ ``on the nose'' (otherwise the call would be
+ill-typed).  This then forces us to use \AgdaKeyword{rewrite} also in
+the implementation of the \transLR\ case in $\mathit{exact}$.
 
@@
 recursive calls are sufficient.
-
-%%%%%%%%%
-% \subsection{Graphical Language}
-% 
-% Idea: thinking in the 3D way; we have collections of towers of
-% blocks. Every 1-combinator involves touching some blocks, swapping
-% them around, etc. Two 1-combinators are equivalent if they touch the
-% same blocks the same number of times. ??
 
@@
 
+\medskip
+{\footnotesize 
 \begin{samepage}
@@
 \end{samepage}
+}
 
@@
 congruence would be used to elide the $\idf$. Furthermore, the various arrows
-would also be named differently --- our \assoclp is often named $\alpha$,
-\assocrp is $\alpha^{-1}$, $\swapp$ is $B$ (always with subscripts).
+would also be named differently --- our \assoclp\ is often named $\alpha$,
+\assocrp\ is $\alpha^{-1}$, $\swapp$ is $B$ (always with subscripts).
 And the two steps needed to remove inverses (i.e. first cancelling

### 2018-07-07 survey/survey.tex
Make the use usage of sum constructor consistent.

@@
 \newcommand{\pp}{\texttt{+}}
-\newcommand{\inl}[1]{\textsf{inl}~#1}
-\newcommand{\inr}[1]{\textsf{inr}~#1}
+\newcommand{\inleft}[1]{\textsf{left}~#1}
+\newcommand{\inright}[1]{\textsf{right}~#1}
 \newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
@@
 two copies of $a$, but its output, only one. On the type side,
-something more subtle happens. Consider $\top \uplus \top$
-versus $\top$; the first has exactly \emph{two} proofs of inhabitation
-($\inl{\texttt{tt}}$ and $\inr{\texttt{tt}}$) while the second only one
-($\texttt{tt}$). These cannot be put in bijective correspondence. Even
-though the ``payload'' \texttt{tt} is the same, forgetting
-$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
-we have expressly disallowed.  Yes, this should remind you of
-Maxwell's daemon: even though the data is the same, they are tagged
-differently, and these tags are indeed information, and their
+something more subtle happens. Consider $\top \uplus \top$ versus
+$\top$; the first has exactly \emph{two} proofs of inhabitation
+($\inleft{\texttt{tt}}$ and $\inright{\texttt{tt}}$) while the second
+only one ($\texttt{tt}$). These cannot be put in bijective
+correspondence. Even though the ``payload'' \texttt{tt} is the same,
+forgetting $\texttt{left}$ (or \texttt{right}) throws away information
+-- something we have expressly disallowed.  Yes, this should remind
+you of Maxwell's daemon: even though the data is the same, they are
+tagged differently, and these tags are indeed information, and their
 information content must be preserved.
@@
 perform $c_1$ or $c_2$ depending on whether the input is labelled
-$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
+$\textsf{left}$ or $\textsf{right}$ respectively. Finally the
 construction $c_1 \otimes c_2$ operates on a product structure, and

### 2018-07-07 survey/survey.tex
Add \sumtype and \prodtype and make all Pi types use these macros. Fixed Upto Sec 3.

@@
 \newcommand{\pp}{\texttt{+}}
+\newcommand{\sumtype}{+}
+\newcommand{\prodtype}{\times}
 \newcommand{\inleft}[1]{\textsf{left}~#1}
@@
 \newcommand{\assoc}{\circ}
-\newcommand{\identlp}{\ensuremath{\mathit{unite}_+\mathit{l}}}
-\newcommand{\identrp}{\ensuremath{\mathit{uniti}_+\mathit{l}}}
-\newcommand{\identlsp}{\ensuremath{\mathit{unite}_+\mathit{r}}}
-\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_+\mathit{r}}}
-\newcommand{\swapp}{\ensuremath{\mathit{swap}_+}}
-\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_+}}
-\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_+}}
-\newcommand{\identlt}{\ensuremath{\mathit{unite}_*\mathit{l}}}
-\newcommand{\identrt}{\ensuremath{\mathit{uniti}_*\mathit{l}}}
-\newcommand{\identlst}{\ensuremath{\mathit{unite}_*\mathit{r}}}
-\newcommand{\identrst}{\ensuremath{\mathit{uniti}_*\mathit{r}}}
-\newcommand{\swapt}{\ensuremath{\mathit{swap}_*}}
-\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_*}}
-\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_*}}
+\newcommand{\identlp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{l}}}
+\newcommand{\identrp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{l}}}
+\newcommand{\identlsp}{\ensuremath{\mathit{unite}_{\sumtype}\mathit{r}}}
+\newcommand{\identrsp}{\ensuremath{\mathit{uniti}_{\sumtype}\mathit{r}}}
+\newcommand{\swapp}{\ensuremath{\mathit{swap}_{\sumtype}}}
+\newcommand{\assoclp}{\ensuremath{\mathit{assocl}_{\sumtype}}}
+\newcommand{\assocrp}{\ensuremath{\mathit{assocr}_{\sumtype}}}
+\newcommand{\identlt}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{l}}}
+\newcommand{\identrt}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{l}}}
+\newcommand{\identlst}{\ensuremath{\mathit{unite}_{\prodtype}\mathit{r}}}
+\newcommand{\identrst}{\ensuremath{\mathit{uniti}_{\prodtype}\mathit{r}}}
+\newcommand{\swapt}{\ensuremath{\mathit{swap}_{\prodtype}}}
+\newcommand{\assoclt}{\ensuremath{\mathit{assocl}_{\prodtype}}}
+\newcommand{\assocrt}{\ensuremath{\mathit{assocr}_{\prodtype}}}
 \newcommand{\absorbr}{\ensuremath{\mathit{absorbr}}}
@@
 
+%% \roshan{Use something else other than $*$ here to talk about
+%%   idempotence.}
+
 So mere inhabitation falls far short of our goals of being able to
@@
 \\
-\identlp :&  0 + t & \iso & t &: \identrp \\
-\swapp :&  t_1 + t_2 & \iso & t_2 + t_1 &: \swapp \\
-\assoclp :&  t_1 + (t_2 + t_3) & \iso & (t_1 + t_2) + t_3 &: \assocrp \\
-\\
-\identlt :&  1 * t & \iso & t &: \identrt \\
-\swapt :&  t_1 * t_2 & \iso & t_2 * t_1 &: \swapt \\
-\assoclt :&  t_1 * (t_2 * t_3) & \iso & (t_1 * t_2) * t_3 &: \assocrt \\
-\\
-\absorbr :&~ 0 * t & \iso & 0 ~ &: \factorzl \\
-\dist :&~ (t_1 + t_2) * t_3 & \iso & (t_1 * t_3) + (t_2 * t_3)~ &: \factor
+\identlp :&  0 \sumtype t & \iso & t &: \identrp \\
+\swapp :&  t_1 \sumtype t_2 & \iso & t_2 \sumtype t_1 &: \swapp \\
+\assoclp :&  t_1 \sumtype (t_2 \sumtype t_3) & \iso & (t_1 \sumtype t_2) \sumtype t_3 &: \assocrp \\
+\\
+\identlt :&  1 {\prodtype} t & \iso & t &: \identrt \\
+\swapt :&  t_1 {\prodtype} t_2 & \iso & t_2 {\prodtype} t_1 &: \swapt \\
+\assoclt :&  t_1 {\prodtype} (t_2 {\prodtype} t_3) & \iso & (t_1 {\prodtype} t_2) {\prodtype} t_3 &: \assocrt \\
+\\
+\absorbr :&~ 0 {\prodtype} t & \iso & 0 ~ &: \factorzl \\
+\dist :&~ (t_1 \sumtype t_2) {\prodtype} t_3 & \iso & (t_1 {\prodtype} t_3) \sumtype (t_2 {\prodtype} t_3)~ &: \factor
 \end{array}
@@
 do this literally. To make the analogy with commutative semirings
-stand out even more, we will use $0, 1, +$, and $*$ at the type level,
+stand out even more, we will use $0, 1, \sumtype$, and ${\prodtype}$ at the type level,
 and will denote ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms}
@@
 \begin{center}
-\fbox{
-The proof terms denoting
-commutative semiring equivalences
+\fbox{ The proof terms denoting commutative semiring equivalences
   induce the terms of $\Pi$.}
@@
 {\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
-{\jdg{}{}{c_1 \oplus c_2 : t_1 + t_3 \iso t_2 + t_4}}
+{\jdg{}{}{c_1 \oplus c_2 : t_1 \sumtype t_3 \iso t_2 \sumtype t_4}}
 {}
@@
 {\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
-{\jdg{}{}{c_1 \otimes c_2 : t_1 * t_3 \iso t_2 * t_4}}
+{\jdg{}{}{c_1 \otimes c_2 : t_1 {\prodtype} t_3 \iso t_2 {\prodtype} t_4}}
 {}
@@
 
-Thus the denotation of the $\Pi$ terms \emph{should} be permutations. But
-given types $A$ and $B$ denoting $\fin{m}$ and~$\fin{n}$ respectively,
-what are $A \uplus B$ and $A \times B$ ?  They correspond exactly to
-$\fin{m+n}$ and $\fin{m*n}$!  Geometrically, this corresponds to
-concatenation for $A + B$, i.e. lining up the elements of $A$ first,
-and then those of~$B$. For $A * B$, one can picture this as
-lining up the elements of $A$ horizontally, those of $B$ vertically
-and perpendicular to those of $A$, and filling in the square with
-pairs of elements from $A$ and $B$; if one re-numbers these sequentially,
-reading row-wise, this gives an enumeration of $\fin{m*n}$.
+%% \roshan{Mixing types and numberix $+$ is super confusing. Let's use a
+%%   block font of some sort for the numeric cases. There are only a few
+%%   of them.}
+
+Thus the denotation of the $\Pi$ terms \emph{should} be
+permutations. But given types $A$ and $B$ denoting $\fin{m}$
+and~$\fin{n}$ respectively, what are $A \uplus B$ and $A \times B$ ?
+They correspond exactly to $\fin{m+n}$ and $\fin{m*n}$!
+Geometrically, this corresponds to concatenation for $A + B$,
+i.e. lining up the elements of $A$ first, and then those of~$B$. For
+$A * B$, one can picture this as lining up the elements of $A$
+horizontally, those of $B$ vertically and perpendicular to those of
+$A$, and filling in the square with pairs of elements from $A$ and
+$B$; if one re-numbers these sequentially, reading row-wise, this
+gives an enumeration of $\fin{m*n}$.
 
@@
 \begin{array}{rrcll}
-\identlsp :&  t + 0 & \iso & t &: \identrsp \\
-\identlst :&  t * 1 & \iso & t &: \identrst \\
-\\
-\absorbl :&~ t * 0 & \iso & 0 ~ &: \factorzr \\
-\distl :&~ t_1 * (t_2 + t_3) & \iso & (t_1 * t_2) + (t_1 * t_3)~ &: \factorl
+\identlsp :&  t \sumtype 0 & \iso & t &: \identrsp \\
+\identlst :&  t {\prodtype} 1 & \iso & t &: \identrst \\
+\\
+\absorbl :&~ t {\prodtype} 0 & \iso & 0 ~ &: \factorzr \\
+\distl :&~ t_1 {\prodtype} (t_2 \sumtype t_3) & \iso & (t_1 {\prodtype} t_2) \sumtype (t_1 {\prodtype} t_3)~ &: \factorl
 \end{array}
@@
 
+%% \roshan{Mixing types and numberix $+$ is super confusing. Let's use a
+%%   block font of some sort for the numeric cases. There are only a few
+%%   of them.}
+
 \begin{definition}($\sharp$). By recursion on $|b|$.  First define a
@@
  \sem{1} & = & \idc \\
- \sem{1+b} & = & \idc \oplus\ \sem{b} \\
- \sem{(b_1+b_2)+b_3} & = & \assocrp \odot \sem{b_1 + (b_2 + b_3)} \\
- \sem{b_1 + b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp\, b_1 + b_2 } \\
- \sem{0 * b_2} & = & \absorbr \\
- \sem{1 * b_2} & = & \identlt \odot \sem{b_2} \\
- \sem{(b_1 * b_2) * b_3} & = & \assocrt \odot \sem{b_1 * (b_2 * b_3)} \\
- \sem{(b_1+b_2) * b_3} & = & \dist \odot \sem{b_1 * b_3+b_2 * b_3} \\
+ \sem{1{\sumtype}b} & = & \idc \oplus\ \sem{b} \\
+ \sem{(b_1{\sumtype}b_2){\sumtype}b_3} & = & \assocrp \odot \sem{b_1 {\sumtype} (b_2 {\sumtype} b_3)} \\
+ \sem{b_1 {\sumtype} b_2} & = & (\sem{b_1} \oplus \idc) \odot \sem{ \sharp\, b_1 {\sumtype} b_2 } \\
+ \sem{0 {\prodtype} b_2} & = & \absorbr \\
+ \sem{1 {\prodtype} b_2} & = & \identlt \odot \sem{b_2} \\
+ \sem{(b_1 {\prodtype} b_2) {\prodtype} b_3} & = & \assocrt \odot \sem{b_1 {\prodtype} (b_2 {\prodtype} b_3)} \\
+ \sem{(b_1{\sumtype}b_2) {\prodtype} b_3} & = & \dist \odot \sem{b_1 {\prodtype} b_3{\sumtype}b_2 {\prodtype} b_3} \\
  \end{array}\]
@@
 \item
-The product type \ensuremath{b_1 * b_2} may be represented both as one wire
+The product type \ensuremath{b_1 {\prodtype} b_2} may be represented both as one wire
 labeled \ensuremath{b_1 \times b_2} or by two parallel wires labeled \ensuremath{b_1} and
@@
 Sum types may similarly be represented using using parallel wires with
-a \ensuremath{+} operator between them.
+a \ensuremath{{\sumtype}} operator between them.
 
@@
 \noindent
-When tracing the execution of \ensuremath{b_1+b_2} represented by one
+When tracing the execution of \ensuremath{b_1{\sumtype}b_2} represented by one
 wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
@@
 \item
-When representing complex types like \ensuremath{(b_1 * b_2)+b_3} some visual
+When representing complex types like \ensuremath{(b_1 {\prodtype} b_2){\sumtype}b_3} some visual
 grouping of the wires may be done to aid readability. The exact type
@@
 Associativity is entirely skipped in the graphical language. Hence
-three parallel wires may be inferred as \ensuremath{b_1 * (b_2 * b_3)} or
-\ensuremath{(b_1 * b_2) * b_3}, based on the context. This is much like handling of
+three parallel wires may be inferred as \ensuremath{b_1 {\prodtype} (b_2 {\prodtype} b_3)} or
+\ensuremath{(b_1 {\prodtype} b_2) {\prodtype} b_3}, based on the context. This is much like handling of
 associativity in the graphical representations of categories as well as
@@
 \noindent
-When tracing the execution of \ensuremath{b_1+b_2} represented by one wire, one
+When tracing the execution of \ensuremath{b_1{\sumtype}b_2} represented by one wire, one
 can think of a value of the form \ensuremath{\mathit{left} ~v_1} or
@@
 essentially be thought of as a multiplexer that redirects the flow of
-\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1+b_2}, as shown
+\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1{\sumtype}b_2}, as shown
 below.
@@
 \[\begin{array}{rcl}
-c & : & b * (1+1) \leftrightarrow b + b \\
+c & : & b {\prodtype} (1{\sumtype}1) \leftrightarrow b {\sumtype} b \\
 c & = & \swapt \odot \dist \odot (\identlt \oplus \identlt)
@@
 \paragraph*{Booleans}
-Let us start with encoding booleans. We use the type \ensuremath{1+1} to
+Let us start with encoding booleans. We use the type \ensuremath{1{\sumtype}1} to
 represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
@@
 \ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
-using the type \ensuremath{\mathit{bool} * (\mathit{bool} *  \mathit{bool})}.  We can perform various
+using the type \ensuremath{\mathit{bool} {\prodtype} (\mathit{bool} {\prodtype}  \mathit{bool})}.  We can perform various
 operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
@@
 
-\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  *  (\mathit{not}  *  \mathit{not})}
+\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  {\prodtype}  (\mathit{not}  {\prodtype}  \mathit{not})}
 
@@
 of a value \ensuremath{v:b} based on the value of another one of type
-\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be
+\ensuremath{b_1{\sumtype}b_2}. If we choose \ensuremath{1{\sumtype}1} to be
 \ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
@@
 
-\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  *  b_1 \leftrightarrow \mathit{bool}  *  b_2}
-
-\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) + (\idc \otimes\  c_2)) \odot \factor}
+\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  {\prodtype}  b_1 \leftrightarrow \mathit{bool}  {\prodtype}  b_2}
+
+\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) {\sumtype} (\idc \otimes\  c_2)) \odot \factor}
 
@@
 
-The diagram above shows the input value of type \ensuremath{(1+1) *  b_2}
+The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_2}
 processed by the distribute operator \ensuremath{\dist}, which converts it into
-a value of type \ensuremath{(1 *  b_2)+(1 *  b_2)}. In the
+a value of type \ensuremath{(1 {\prodtype}  b_2){\sumtype}(1 {\prodtype}  b_2)}. In the
 \ensuremath{\mathit{left}} branch, which corresponds to the
@@
 By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
-combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} *  \mathit{bool}\leftrightarrow \mathit{bool} *  \mathit{bool}} which negates its
+combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} {\prodtype}  \mathit{bool}\leftrightarrow \mathit{bool} {\prodtype}  \mathit{bool}} which negates its
 second argument if the first argument is \ensuremath{\mathit{true}}. This gate
@@
 If we iterate this construction once more, the resulting combinator
-\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} *  (\mathit{bool} *  \mathit{bool})\leftrightarrow \mathit{bool} *  (\mathit{bool} *  \mathit{bool})}. The
+\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})\leftrightarrow \mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})}.
+The
 resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},

### 2018-08-18 survey/survey.tex
Fixed the use of * in idempotence.

@@
 \newcommand{\pp}{\texttt{+}}
+
+\newcommand{\presumtype}{\uplus}
+\newcommand{\preprodtype}{*}
 \newcommand{\sumtype}{+}
 \newcommand{\prodtype}{\times}
+
 \newcommand{\inleft}[1]{\textsf{left}~#1}
@@
 corresponds to the building blocks of type theory which are: the empty
-type ($\bot$), the unit type ($\top$), the sum type ($\uplus$), and
-the product ($\times$) type. Before getting into the formal theory,
-let's consider possible deformations on the space $(\top \uplus \bot)
-\times (\top \uplus \top)$. This space is the product of two
-subspaces: the subspace $(\top \uplus \bot)$ which itself is the sum
+type ($\bot$), the unit type ($\top$), the sum type ($\presumtype$), and
+the product ($\preprodtype$) type. Before getting into the formal theory,
+let's consider possible deformations on the space $(\top \presumtype \bot)
+\preprodtype (\top \presumtype \top)$. This space is the product of two
+subspaces: the subspace $(\top \presumtype \bot)$ which itself is the sum
 of the space~$\top$ containing one element $\texttt{tt}$ and the empty
-space $\bot$ and the subspace $(\top \uplus \top)$ which is the sum of
+space $\bot$ and the subspace $(\top \presumtype \top)$ which is the sum of
 two spaces each containing the one element $\texttt{tt}$. First, as
@@
 is that we can deform the space above to any other space with 2
-elements such as $\top \uplus \top$ or $\top \uplus (\top \uplus
+elements such as $\top \presumtype \top$ or $\top \presumtype (\top \presumtype
 \bot)$. What this really means is that we are treating the sum and
@@
 $\true$ & $\top$ \\
-$\land$ & $\times$ \\
-$\lor$ & $\uplus$ \\
+$\land$ & $\preprodtype$ \\
+$\lor$ & $\presumtype$ \\
 \end{tabular}
@@
 expressions under the rules of the host type system.  This is however
-not true: the proposition $A \times B \equiv B \times A$ is not
+not true: the proposition $A \preprodtype B \equiv B \preprodtype A$ is not
 normally\footnote{Except in univalent type theory where equivalent
   types are identified.} provable for arbitrary types $A$ and $B$. But
-it should be clear that $A \times B$ and $B \times A$ contain
+it should be clear that $A \preprodtype B$ and $B \preprodtype A$ contain
 equivalent information. In other words, we would like to be able to
-witness that $A \times B$ can be reversibly deformed into
-$B \times A$, and vice-versa, which motivates the introduction of type
+witness that $A \preprodtype B$ can be reversibly deformed into
+$B \preprodtype A$, and vice-versa, which motivates the introduction of type
 \emph{equivalences}. To do this, we need a few important auxiliary
@@
 
+%% \roshan{This is a poor sentence. We have 4 or 5 definitions back to
+%%   back and have given very little intuition as to how or why they
+%%   are relevant and then we go on to say this:}
+
 \noindent We can then replace quasi-inverse with bi-invertibility in
@@
 us here.
+
+%% \roshan{More on the above: Let us either explain what the point is
+%%   OR simply not suggest that we could be making a point, which we
+%%   then choose not to.}
 
@@
 Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
-this change, we can indeed prove that types (with $\bot, \top, \uplus,
-\times$) form a commutative semiring. The reader familiar with
+this change, we can indeed prove that types (with $\bot, \top, \presumtype,
+\preprodtype$) form a commutative semiring. The reader familiar with
 universal algebra should pause and ponder a bit about what we have
@@
 more issue. In logic, it is true that $A \lor A = A$ and
-$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
+$A \land A = A$. However, neither $A \presumtype A$ nor $A \preprodtype A$ are
 equivalent to $A$. They are however \emph{equi-inhabited}. This is
 a fancy way of saying
-\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow \qquad A \
+\[ A \presumtype A \ \text{is inhabited} \qquad \Leftrightarrow \qquad A \
   \text{is inhabited} \] The above is the real \textit{essence} of the
@@
 constructive statement: there are indeed functions
-$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
+$f : A \presumtype A \rightarrow A$ and $g : A \rightarrow A \presumtype A$;
 however, they are not inverses.
-
-%% \roshan{Use something else other than $*$ here to talk about
-%%   idempotence.}
 
@@
 of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
-\emph{idempotent}: this is the property of any binary operation $*$
-where $\forall a. a * a = a$. And it should be clear that an
+\emph{idempotent}: this is the property of any binary operation $\circ$
+where $\forall a. a \circ a = a$. And it should be clear that an
 idempotent operations is a \emph{forgetful} operation: its input has
 two copies of $a$, but its output, only one. On the type side,
-something more subtle happens. Consider $\top \uplus \top$ versus
+something more subtle happens. Consider $\top \presumtype \top$ versus
 $\top$; the first has exactly \emph{two} proofs of inhabitation
@@
 \begin{array}{rrcll}
-\idc :& A & \simeq & A &: \idc \\
-\\
-\identlp :&  \bot \uplus A & \simeq & A &: \identrp \\
-\swapp :&  A \uplus B & \simeq & B \uplus A &: \swapp \\
-\assoclp :&  A \uplus (B \uplus C) & \simeq & (A \uplus B) \uplus C &: \assocrp \\
-\\
-\identlt :&  \top \times A & \simeq & A &: \identrt \\
-\swapt :&  A \times B & \simeq & B \times A &: \swapt \\
-\assoclt :&  A \times (B \times C) & \simeq & (A \times B) \times C &: \assocrt \\
-\\
-\absorbr :&~ \bot \times A & \simeq & \bot ~ &: \factorzl \\
-\dist :&~ (A \uplus B) \times C & \simeq & (A \times C) \uplus (B \times C)~ &: \factor
+& A & \simeq & A &\\
+\\
+&  \bot \presumtype A & \simeq & A &\\
+&  A \presumtype B & \simeq & B \presumtype A &\\
+&  A \presumtype (B \presumtype C) & \simeq & (A \presumtype B) \presumtype C &\\
+\\
+&  \top \preprodtype A & \simeq & A &\\
+&  A \preprodtype B & \simeq & B \preprodtype A &\\
+&  A \preprodtype (B \preprodtype C) & \simeq & (A \preprodtype B) \preprodtype C &\\
+\\
+& \bot \preprodtype A & \simeq & \bot &\\
+& (A \presumtype B) \preprodtype C & \simeq & (A \preprodtype C) \presumtype (B \preprodtype C) &
 \end{array}
@@
 
+%% \begin{figure}[t]
+%% \[
+%% \begin{array}{rrcll}
+%% \idc :& A & \simeq & A &: \idc \\
+%% \\
+%% \identlp :&  \bot \presumtype A & \simeq & A &: \identrp \\
+%% \swapp :&  A \presumtype B & \simeq & B \presumtype A &: \swapp \\
+%% \assoclp :&  A \presumtype (B \presumtype C) & \simeq & (A \presumtype B) \presumtype C &: \assocrp \\
+%% \\
+%% \identlt :&  \top \preprodtype A & \simeq & A &: \identrt \\
+%% \swapt :&  A \preprodtype B & \simeq & B \preprodtype A &: \swapt \\
+%% \assoclt :&  A \preprodtype (B \preprodtype C) & \simeq & (A \preprodtype B) \preprodtype C &: \assocrt \\
+%% \\
+%% \absorbr :&~ \bot \preprodtype A & \simeq & \bot ~ &: \factorzl \\
+%% \dist :&~ (A \presumtype B) \preprodtype C & \simeq & (A \preprodtype C) \presumtype (B \preprodtype C)~ &: \factor
+%% \end{array}
+%% \]
+%% \caption{Type isomorphisms.}
+%% \label{type-isos}
+%% \end{figure}
+
 Nevertheless, the Curry-Howard correspondence still has some force. We
-know that the inhabitants of types formed with with
-$\bot, \top, \uplus, \times$ form a commutative semiring. What we want
+know that the inhabitants of types formed with with $\bot, \top,
+\presumtype, \preprodtype$ form a commutative semiring. What we want
 to know is, which types are equivalent? From a commutative semiring
@@
 mutually inverse functions which witness these identities.  For
-concreteness, we show the signatures in Fig.~\ref{type-isos}.  As
-these all come in symmetric pairs (some of which are self-symmetric),
-we give names for both directions.
+concreteness, we show the signatures in Fig.~\ref{type-isos}.
 
@@
 thus leads us to posit that the correct denotational semantics for
-finite discrete types is that of the semiring
-$\left(\Nat, 0, 1, +, \cdot\right)$. It is worth noting that equality
-in this semiring is intensional (i.e. two things are equal if and only if
-they are identical after evaluation), unlike that for types.
+finite discrete types is that of the semiring $\left(\Nat, 0, 1, +,
+\cdot\right)$. It is worth noting that equality in this semiring is
+intensional (i.e. two things are equal if and only if they are
+identical after evaluation), unlike that for types.
 
@@
 We want to create a programming language, which we call $\Pi$, such
-that the types and type combinators map to
-$\bot, \top, \uplus, \times$, and such that we have ground terms whose
-denotation are all $16$ type isomorphisms of
-Fig.~\ref{type-isos}. This is rather straightforward, as we can simply
-do this literally. To make the analogy with commutative semirings
-stand out even more, we will use $0, 1, \sumtype$, and ${\prodtype}$ at the type level,
-and will denote ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms}
-shows the ``constants'' of the language. Note how we have continued
-with the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof
-  terms}, but rather than being witnesses of inhabitation, they are
-witnesses of equivalences. Thus we get an unexpected programming
-language design:
+that the types and type combinators map to $\bot, \top, \presumtype,
+\preprodtype$, and such that we have ground terms whose denotation are
+all $16$ type isomorphisms of Fig.~\ref{type-isos}. This is rather
+straightforward, as we can simply do this literally. To make the
+analogy with commutative semirings stand out even more, we will use
+$0, 1, \sumtype$, and ${\prodtype}$ at the type level, and will denote
+``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
+``constants'' of the language.  As these all come in symmetric pairs
+(some of which are self-symmetric), we give names for both directions.
+Note how we have continued with the spirit of Curry-Howard: the terms
+of $\Pi$ are \emph{proof terms}, but rather than being witnesses of
+inhabitation, they are witnesses of equivalences. Thus we get an
+unexpected programming language design:
 
@@
 permutations. But given types $A$ and $B$ denoting $\fin{m}$
-and~$\fin{n}$ respectively, what are $A \uplus B$ and $A \times B$ ?
+and~$\fin{n}$ respectively, what are $A \presumtype B$ and $A \preprodtype B$ ?
 They correspond exactly to $\fin{m+n}$ and $\fin{m*n}$!
@@
 
-\begin{definition}($\sharp$). By recursion on $|b|$.  First define a
-type associated to each size. We will revert to using type notation
+\begin{definition}($\sharp$).
+  By recursion on $|b|$.  First define $\tau$ that maps numeric sizes
+  to their corresponding types. We will revert to using type notation
   for greater clarity of this definition:
@@
   \tau~ (0) & = & \bot \\
-  \tau~ (1 + n) & = & \top \uplus \tau~ (n) \\
+  \tau~ (1 + n) & = & \top \presumtype \tau~ (n) \\
  \end{array}\]
@@
 The product type \ensuremath{b_1 {\prodtype} b_2} may be represented both as one wire
-labeled \ensuremath{b_1 \times b_2} or by two parallel wires labeled \ensuremath{b_1} and
+labeled \ensuremath{b_1 \prodtype b_2} or by two parallel wires labeled \ensuremath{b_1} and
 \ensuremath{b_2}. Both representations may be used interchangeably.

### 2018-10-25 survey/survey.tex
Go through the Word doc, fix a few minor things, reject most of the changes. Some of them didn't make sense as cut-and-paste made all the Greek characters disappear. And some of the grammar changes were wrong.

@@
 
+% Hyphenation overrides
+\hyphenation{phy-si-cal}
+
 \begin{document}
@@
 
-  Our aim is re-develop foundational computational models in a way
+  Our aim is to re-develop foundational computational models in a way
   that embraces the principle of conservation of information. We first
@@
   isomorphisms and proof terms for commutative semirings.  We then
-  ``move up a level'' to examine spaces that treat programs as data
+  ``move up a level'' to examine spaces that treat programs as data,
   which is a crucial notion for any universal model of computation. To
@@
 
-What kind of operations can computers perform? This question was
+What kind of operations can computers perform? This question has been
 answered several times in the last hundred years, where each answer
@@
 No doubt, this consensus on the available primitive physical
-operations has been successful. And no doubt, these operations
+operations has been successful. Futhermore, these operations
 \emph{can} indeed be performed on a computer. Yet, today, with a
-possible quantum computing revolution in sight and with unprecedented
+possible quantum computing revolution in sight and an unprecedented
 explosion in embedded computers and cyber-physical systems, there are
-reasons to re-think this foundational question again. In fact, the
+reasons to re-think this foundational question. In fact, the
 calls to re-think this foundational question have been proclaimed by
-physicists almost forty years ago as the following two quotes testify:
+physicists almost forty years ago:
 
@@
 
-After a detailed introduction on the origins of ``logically
-reversibile computer operations'' and an excursion into the origins of
-``irreversible computer operations,'' we will develop in detail three
+After a detailed introduction on the origins of \emph{logically
+reversibile computer operations} and an excursion into the origins of
+\emph{irreversible computer operations}, we will develop in detail three
 reversible models of computation and discuss their potential
@@
 \paragraph*{Maxwell's Daemon.}
-To fully appreciate the missing principle of ``reversibility'' in
+To fully appreciate the missing principle of \emph{reversibility} in
 conventional computing, we go back to an old thought experiment by
@@
 1961~\cite{Landauer:1961} -- a principle whose experimental validation
-happened recently in 2012~\cite{berut2012experimental}.
-
-Maxwell's daemon appears to violate the second law of thermodynamics by
+happened in 2012~\cite{berut2012experimental}.
+
+Maxwell's Daemon appears to violate the second law of thermodynamics by
 having a tiny ``intelligence'' observing the movement of individual
@@
 system. Landauer's resolution of the daemon relied on two ideas that
-took root only a few decades earlier: the formal notion of computation
+had taken root only a few decades earlier: the formal notion of computation
 (through the work of Turing~\cite{turing}, Church~\cite{church51}, and
@@
 Shannon~\cite{shannon1948}). Landauer reasoned that the computation
-done by the finite brain of the daemon, involves getting information
+done by the finite brain of the daemon involves getting information
 about the movement of molecules, storing that information, analyzing
@@
 following hundreds of years of tradition, the truth of a statement was
-seen as ``absolute'' and independent of any reasoning, understanding,
+seen as \emph{absolute} and independent of any reasoning, understanding,
 or action. Statements were either true or false with no regard to any
-``observer'' and the idea that statements had information content that
+\emph{observer} and the idea that statements had information content that
 should be preserved was outside the classical understanding of
@@
 De Morgan, Venn, Boole, and Peirce in the mid to late 1800s. Going
-back to Boole's 1853 book entitled \emph{An Investigation of the Laws
+back to Boole's 1853 book entitled \textit{An Investigation of the Laws
   of Thought, on which are Founded the Mathematical Theories of Logic
@@
 \end{quote}
-which clearly identifies the ``source'' of the logical laws as
+which clearly identifies the \emph{source} of the logical laws as
 mirroring Boole's understanding of human reasoning. A few chapters
@@
   being to possess a quality, and at the same time not to possess it, is a
-  consequence of the fundamental law of thought, whose expression is $x^2 =
-  x$.
+  consequence of the fundamental law of thought, whose expression is 
+  $x^2 = x$.
 \end{quote}
@@
 (1819--1822) was able to construct. There are other computer
-precursors as well -- the first stored programs were actually for
-looms, most notably those of Bouchon (1725) which operated on a paper
-tape, and Jacquard (1804) which operated by chains of punched cards.
+precursors as well -- the first stored programs were actually for looms, most
+notably those of Bouchon (1725) which were controlled by a paper
+tape, and of Jacquard (1804), controlled by chains of punched cards.
 But it was only in the 20th century that computer science emerged as a
 formal discipline. One of the pioneering works was Alan Turing's
-seminal paper~\cite{turing} in 1936 which established the idea that
+seminal paper~\cite{turing} of 1936 which established the idea that
 computation has a formal interpretation and that all computability can
@@
 -- \emph{abstractions of computation realized in the physical world.}
-Indeed, going back to Turing's 1936 article \emph{On Computable
-  Numbers, with an Application to the Entscheidungsproblem,}, the
+Indeed, going back to Turing's 1936 article \textit{On Computable
+  Numbers, with an Application to the Entscheidungsproblem,} the
 opening sentence of Sec. 1 is:
@@
 distances) and metaphysical (on restrictions of the mind).  If we take
-the human mind to be a physical ``machine'' which does computation,
-then when both of the above assumptions are translated to the language
+the human mind to be a physical ``machine'' which performs computation,
+then when both of the above assumptions are translated into the language
 of physics, they embody what is known as the ``Bekenstein
@@
 principles underlying quantum mechanics, the most successful physical
-theory known to us and adapt computation to ``learn'' from these
+theory known to us, and adapt computation to ``learn'' from these
 principles.  In the words of Girard~\cite{Girard:2007:TMI:1348911.1348915}:
@@
 of ``sameness'' that our data will carry. Thus we need to start by
-looking at what structure our data has, and that will help us define
+looking at what structure our data has, which will help us define
 an appropriate notion of a reversible program. Of course, when
@@
 derive what it means for data, programs, program transformations, as
-well as proofs / deductions, to be ``the same'' -- in an manner
+well as proofs / deductions, to be ``the same'' -- in a manner
 consistent with preservation of information. This stands in stark
@@
 software transactional memories, continuations, backtracking search,
-and multiple-level ``undo'' features in commercial applications. In
+and multiple-level undo features in commercial applications. In
 the early nineties, Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued
@@
 the starting point for this development is the physical principle of
-``conservation of
-information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative} and
-the family of languages is designed to embrace that principle by
+\emph{conservation of
+information}~\cite{Hey:1999:FCE:304763,fredkin1982conservative} and
+the family of languages is designed to embrace this principle by
 requiring all computations to preserve information.
@@
 \emph{inhabitation} as the main source of analogy between logic and
-computation, \emph{type equivalences} becomes the source of analogy.
+computation, \emph{type equivalence} becomes the source of analogy.
 Taking inspiration from the fact that many terms of the $\lambda$-calculus

### 2018-12-10 survey/survey.tex
Mietek Bak found a minor typo on p.19 -- fixed.

@@
 
-The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_2}
+The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_1}
 processed by the distribute operator \ensuremath{\dist}, which converts it into
-a value of type \ensuremath{(1 {\prodtype}  b_2){\sumtype}(1 {\prodtype}  b_2)}. In the
+a value of type \ensuremath{(1 {\prodtype}  b_1){\sumtype}(1 {\prodtype}  b_1)}. In the
 \ensuremath{\mathit{left}} branch, which corresponds to the
@@
 the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.
+The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied
+to get the final result of type \ensuremath{(1{\sumtype}1) {\prodtype} b_2}.
 

### 2018-12-15 RCforBX/RCforBX.tex
current thoughts on RCforBX paper

\documentclass[a4paper]{article}
\usepackage{graphicx}
\usepackage{onecolceurws}

% Not the final title!
\title{Reversible Programming for the BX enthusiast}

\author{
Jacques Carette\\ Dept. of Computing and Software\\
		(address) \\ carette@mcmaster.ca
\and
Amr Sabry \\ Computer Science Dept.\\
		(address) \\ sabry@indiana.edu
}

\institution{}

\begin{document}
\maketitle

\begin{abstract}
Show how BX is deeply entertwined with RP already,
so that BX enthusiast should really know all about RP too.
\end{abstract}
\vskip 32pt

\section{Introduction}

The inspiration for this paper comes from two sources:
\begin{enumerate}
  \item Oleg Grenrus' \textit{Finding correct (lens) laws}~\cite{oleg-blog}, 
    and
  \item The paper \textit{Synthesizing Bijective Lenses}~\cite{Miltner2018}.
\end{enumerate}

There are many, many different representations for (monomorphic)
lenses. The one that will be the most fruitful for us is the
existential formulation:
\begin{verbatim}
\end{verbatim}
where the type isomorphism is plain to see. That formulation is the
principal source of the relationship.

Paper: bijective lenses, p.5.

\section{Lens and other Optics}

Remember that (A + A + A) ~= 3*A. And that one can lens into the 3
on the right -- so one can lens into it on the left too!
The iso expresses the difference (in languages with proper sum-of-product
types) between implicit tags and explicit tags. On the left, the compiler
chooses how to represent it, on the right, the programmer. But there are
many types that can be ``exploded'' so as to move any discrete portion
to/from tags.

Rememer that the type Lens (A*B*C) (A*C) is inhabited. So is
Lens (A*B*C) (C*A).  Look familiar?

\section{Pi}

Intro to Pi.

\section{Happy together}

Really explore the relationship.

\section{Random thoughts}

What if we took different notions of iso (i.e. contractible or not)
instead of the usual one, would that change things? What about the 
effect no the lens laws?

Possibly this should all be done in Agda? Or Haskell? Some PL for sure.

\section{Conclusion}

\bibliographystyle{alpha} 
\bibliography{cites}
%inline the .bbl file directly for mailing to authors.

\end{document}

### 2019-04-04 fractionals/p.tex
thoughts on fractionals based on discussions with Chao-Hong

Reference cells have fractional types

1 -> A * 1/A
From no information, create a ref cell with a getter (A) and a setter (1/A)
Usually type of getter is 1 -> A and setter A -> 1
Using that A -> B is [not A + B] and hence [1/A * B] in this case then:
1 -> A is [1/1 * A] which is A, and
A -> 1 is [1/A * 1] which is 1/A

Inspiration from lenses
Particle/Anti-particle
Entanglement (side channel between setter and getter)

Forget 1/0 for now; we will fix later with meadow trick

Also we will index the types with a monad or something to make the effect of creating the cell explicit

Denotational semantics already kind of known: groupoid with fractional cardinality

If A and 1/A are left loose, we get errors (reading uninitialized cells, race conditions, etc.). But if information is conserved and we do meadow trick, no errors! That should be the main theorem.

Can get functions A -> B as [1/A * B] , which has a hidden cell for parameter passing, a setter to pass the parameter and a value of type B that is blocked until the cell is full.

Now do interesting programs. Start with all the constructions for closed monoidal categories.

Check 1/(1/A) = A
1 -> 1/A * 1/(1/A)
Se here we are creating a cell that holds the setter for another cell; ho-functions!!!

### 2019-04-11 fractionals/p.tex
my thoughts on the ideas on fractional

@@
 Se here we are creating a cell that holds the setter for another cell; ho-functions!!!
+
+===============================================================
+
+Comments on the above:
+
+I am going to use a mixture of 'reversible', 'functional' and 'imperative' thinking below,
+and may not always signal it explicitly -- you've been warned!
+
+So a getter is 1 -> A only in an imperative setting. In a functional setting, the 'better' type
+is World -> A; and a setter is then World -> A -> World. Which is great, because THAT is
+actually exactly the type of a lens! [We could generalize to World' as the result of set,
+which would push us into generalized optics]
+
+The 'reasoning' to get to A and 1/A is dubious at best; it uses 'Boolean' Curry-Howard,
+which is not reversible. I don't believe in that aspect at all.
+
+However, I do see "Lens S A" as being a kind of 'relative complement', i.e. S // A.
+And it does make complete sense to see World as being chopped-up, through allocation, into
+A and its 'complement' World // A. Let's just call this 'Rest'.
+
+Here World is some kind of 'potentially infinite' universe where a Lens does not
+*necessarily* allocate new memory, but it certainly does reserve memory for our use.
+Observationally, these are probably the same.
+
+What do we obtain here? What seems to drop out quite naturally is a flavour of
+separation logic!  The axioms of Lens, equivalently of isomorphisms and product, give
+us exactly that!
+
+Downside? World is _neither_ 1 or 0.  So 1 / A still doesn't pop out.
+Nevertheless, I think this is very interesting.

### 2019-10-02 fracGC/p.tex
fractionals...

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title[Short Title]{Full Title}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Subtitle}                     %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Complexity perspective: \url{https://www.scottaaronson.com/papers/gates.pdf}

Quipper: \url{https://arxiv.org/pdf/1304.3390.pdf}; uses ancilla; one use is to compile irreversible circuits to reversible ones; need ancilla bits; more generally several quantum algorithms need ancilla bits (see below); in quipper de-allocation left to programmer.

Ricercar tries to define rules for ancilla but not complete \url{https://msoeken.github.io/papers/2015_rc_2.pdf}

Leads to bugs as analyzed by \url{http://drops.dagstuhl.de/opus/volltexte/2019/10196/pdf/OASIcs-PLATEAU-2018-4.pdf}:
\begin{quote}

  Bug type 6: Incorrect composition of operations using mirroring Section 4.5 discussed how bugs in deallocating ancillary qubits can happen due to bad parameters. Here we see how bugs in deallocating ancillary qubits can happen due to incorrect composition of operations following a mirroring pattern. For example, in Table 7, the operations in rows 2 and 3 are respectively mirrored and undone in rows 6 and 5. These lines of code need careful reversal of every loop and every operation.

  A common pattern in quantum programs involves performing operations (e.g., add), contingent on a set of qubits known as control qubits. Without language support, this pattern needs many lines of code and manual allocation of ancillary qubits. In the Scaffold code example in Table 7, rows 3 and 5 are just computing the intersection of qubits q, with the help of ancillary qubits initialized in row 1, in order to realize the controlled rotation operation in row 4. Furthermore, quantum algorithms often need varying numbers of control qubits in different parts of the algorithm, leading to replicated code from multiple versions of the same subroutine differing only by the number of control qubits.

\end{quote}

Uses of ancillas:
\url{https://quantum.country/search}:
\begin{quote}
  There’s a rough heuristic worth noting here, which is that you can often convert if-then style thinking into quantum circuits. You introduce an ancilla qubit to store the outcome of evaluating the if condition. And then depending on the state of the ancilla, you perform the appropriate state manipulation. Finally, when possible you reverse the initial computation, resetting the ancilla to its original state so you can subsequently ignore it.
\end{quote}

\url{https://www.nap.edu/read/25196/chapter/5#73}
\begin{quote}
  For error correction, one needs to replicate the state of a qubit onto a number of qubits. While the no cloning theorem prevents one from copying the state of a qubit directly onto another, one can create a redundant entangled qubit state of many qubits. The key is that the qubits to be entangled must start out in a known state. Qubits with a known state (for purposes of this discussion, it will be the state \verb.|0>.), called ``ancilla qubits,'' may be added to a computation for this purpose. Since the state of ancilla qubits are known, it is possible to create a simple circuit that makes the output state of all these ancilla qubits match the protected qubit: run each ancilla through a controlled-NOT gate, where the control is driven by the qubit that needs to be replicated. Assume that there is a qubit with state \verb.PSI. that we want to protect, where \verb.|PSI>. represents an arbitrary superposition state \verb.|PSI> = a0 |0> + a0 |1>. In the CNOT gate, the ancilla \verb.|0>. state will remain a \verb.|0>. state by the \verb.|0>. component of \verb.|PSI>., but it will be converted to \verb.|1>. by the \verb.|1>. component of \verb.|PSI>. The result of this operation is the newly entangled two-qubit state \verb.a0 |00> + a1 |11>., creating a system in which the ancilla qubit is now perfectly entangled with the first qubit. Adding more ancillas increases the distance of the repetition code.  \end{quote}

\url{https://www.sigarch.org/the-case-for-quantum-computing/}
\begin{quote}
  Imagine a 3-qubit quantum majority code in which a logical “0” is encoded as “000” and a logical “1” is encoded as “111.”  Just as with a classical majority code, a single bit-flip error can be corrected by restoring to the majority value.  Unlike a classical code, however, we can not directly measure the qubits else their quantum state will be destroyed.  Instead, we measure syndromes from each possible pair of qubits by interacting them with an ancilla, then measure each ancilla.  Although the errors to the qubits are actually continuous, the effect of measuring the ancilla is to discretize the errors, as well as inform us whether an error occurred so that it can be corrected.  With this methodology, quantum states are restored in a modular way for even a large quantum computer.  Furthermore, operations on error-corrected qubits can be viewed as digital rather than analog, and only a small number of universal operations (H, T, CNOT) are needed for universal quantum computation.  Through careful design and engineering, error correction codes and this small set of precise operations will lead to machines that could support practical quantum computation.
\end{quote}

\url{https://homepages.cwi.nl/~rdewolf/qcnotesv2.pdf}
\begin{quote}
Using an ancilla qubit, it is possible to avoid doing any intermediate measurements in a quantum circuit. Show how this can be done.  Hint: instead of measuring the qubit, apply a CNOT that “copies” it to a new \verb.|0>.-qubit, which is then left alone until the end of the computation. Analyze what happens.
\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reversible/Quantum Circuits and Ancilla Wires}

Early use of ancillas in Toffoli's paper to implement arbitrary reversible functions using a fixed number of 3 input gates \url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}

Use examples from Ricercar

Is the below a possible example?
\begin{verbatim}
Say we already have a permutation A <-> B
we can implement a permutation X <-> Z 
when there exists Y such that A/X = Y = B/Z

X -> X * Y * 1/Y 
  -> A * 1/Y
  -> B * 1/Y
  -> Y * Z * 1/Y
  -> Z
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Denotations}

Positive rational numbers are a model. Apparently there is a categorification \url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}

Use all the constructions name, coname, etc. and see what they do in this context!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2019-10-05 fracGC/p.tex
story for fractionals; aiming for PLDI

@@
 
+\begin{itemize}
+
+\item Quantum computing hot.
+
+\item Unitary evolution; many algorithms promote classical reversible
+functions into unitary gates (examples, including some from Quipper)
+
+\item Common case: given a permutation between finite sets, find a
+  reversible program that implements it. Answer depends on the
+  available reversible primitives.
+
+\item Example, say we want to transpose a matrix, i.e., write a
+  permutation of type \verb|A*B <-> B*A| where \verb|A| and \verb|B|
+  are the dimensions. Wikipedia example
+  (\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
+\begin{verbatim}
+11 12 13 14 
+21 22 23 24 
+\end{verbatim}
+  transposed to
+\begin{verbatim}
+11 21
+12 22
+13 23
+14 24
+\end{verbatim}
+  Here \verb|A| is \verb|Nat x Nat| and \verb|B| is
+  \verb|Nat x Nat x Nat x Nat| and the input and output matrices are:
+\begin{verbatim}
+M   = (11 , 21) , (12 , 22) , (13, 23) , (14 , 24)
+
+trM = (11 , 12 , 13 , 14) , (21 , 22 , 23 , 23) 
+\end{verbatim}
+Say we are given a language like Pi that is sound and complete with
+respect to permutations on finite types, we would write the
+permutation like so.
+\begin{verbatim}
+WRITE PERMUTATION
+\end{verbatim}
+This code does not use additional space, i.e., it performs the matrix
+transpose in constant space, i.e., performs an in-place matrix
+transposition. It is well know that with additional space, one can
+write more efficient matrix transpositions (explain and citations).
+
+\item So many reversible/quantum circuits allow the introduction of
+  so-called ancilla wires, which serve as scratch space (goes back to
+  Toffoli). The problem is that there is a discipline for using
+  ancilla bits: they need to be created in a known state and can only
+  be gc'ed when they are returned to the same state (explain).
+
+\item No language knows how to automatically guarantee this condition;
+  left to the programmer (Quipper), or some (incomplete?) dynamic
+  checks (Ricercar).
+
+\item We solve this using **fractional types**.
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Extra for now}
+
 Complexity perspective: \url{https://www.scottaaronson.com/papers/gates.pdf}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Fractional Pi}
+
+Adding variables to Pi first?
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Denotations}

### 2019-10-15 fracGC/p.tex
comonad for pointed types

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A Comonad for Pi}
+
+Adding variables to Pi first?
+
+\newcommand{\alt}{~|~}
+\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
+\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
+\newcommand{\pointed}[2]{[#1 \bullet #2]}
+
+Plain Pi over finite types:
+\[\begin{array}{lrcl}
+\textit{Value types} & t &::=& 0 \alt 1 \alt t+t \alt t*t \\
+\textit{Values}      & v &::=& \star \alt \inlv{v} \alt \inrv{v} \alt (v,v) \\
+\textit{Level-1 types} &&& t \leftrightarrow t \\
+\textit{Level-1 programs} & c &::=& \cdots \\
+\textit{Level-2 types} &&& c \Leftrightarrow c \\
+\textit{Level-2 programs} & \alpha &::=& \cdots 
+\end{array}\]
+
+Now we define pointed types $\pointed{t}{v}$ which are a record
+consisting of a value type together with a value of that type. Both
+the level-1 and level-2 program lift naturally to the world of pointed
+types but instead of manually rewriting everything we use the fact
+that pointed types form a comonad. So for example, given a level-1
+program we should be able to just use \textsf{extend} to compose
+instead of plain composition of combinators??
+
+Then we can define $\eta$ and $\epsilon$ in the comonadic world.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Fractional Pi}

### 2019-10-29 fracGC/p.tex
starting paper

@@
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
 \documentclass[sigplan,10pt,review,anonymous]{acmart}
 \settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[sigplan]{acmart}\settopmatter{}
-
-%% Conference information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
 \acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
 \acmYear{2018}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+\acmISBN{} 
+\acmDOI{} 
 \startPage{1}
+\setcopyright{none}
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}
 
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
+\usepackage{booktabs}  
+\usepackage{subcaption}
 
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%\citestyle{acmauthoryear}  %% For author/year citations
-%\citestyle{acmnumeric}     %% For numeric citations
-%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
-                            %% sorting of references within a single citation;
-                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
-                            %% rendered as [14,5,2] rather than [2,5,14].
-%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
-                            %% compression of sequential references within a
-                            %% single citation;
-                            %% e.g., \cite{Baker12,Baker14,Baker16}
-                            %% rendered as [2,3,4] rather than [2-4].
+\newcommand{\alt}{~|~}
+\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
+\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
+\newcommand{\pointed}[2]{[#1 \bullet #2]}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from traditional SIGPLAN
-%% proceedings format to PACMPL format must update the
-%% '\documentclass' and topmatter commands above; see
-%% 'acmart-pacmpl-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-%% Title information
-\title[Short Title]{Full Title}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-\subtitle{Subtitle}                     %% \subtitle is optional
-\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
+\title{Fractional Types}
+\author{Chao-Hong Chen}
+\affiliation{
+  \institution{Indiana University}
+}
 
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
+\author{Vikraman Choudhury}
+\affiliation{
+  \institution{Indiana University}
+}
 
-%% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\author{Jacques Carette}
 \affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
+  \institution{McMaster University}
 }
-\email{first1.last1@inst1.edu}          %% \email is recommended
 
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\author{Amr Sabry}
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
+  \institution{Indiana University}
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
 
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
 \begin{abstract}
@@
 
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
 \maketitle
@@
 \section{Introduction}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Extra for now}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extra for now}
+Complexity perspective:
+\url{https://www.scottaaronson.com/papers/gates.pdf}
 
-Complexity perspective: \url{https://www.scottaaronson.com/papers/gates.pdf}
+Quipper: \url{https://arxiv.org/pdf/1304.3390.pdf}; uses ancilla; one
+use is to compile irreversible circuits to reversible ones; need
+ancilla bits; more generally several quantum algorithms need ancilla
+bits (see below); in quipper de-allocation left to programmer.
 
-Quipper: \url{https://arxiv.org/pdf/1304.3390.pdf}; uses ancilla; one use is to compile irreversible circuits to reversible ones; need ancilla bits; more generally several quantum algorithms need ancilla bits (see below); in quipper de-allocation left to programmer.
+Ricercar tries to define rules for ancilla but not complete
+\url{https://msoeken.github.io/papers/2015_rc_2.pdf}
 
-Ricercar tries to define rules for ancilla but not complete \url{https://msoeken.github.io/papers/2015_rc_2.pdf}
+Leads to bugs as analyzed by
+\url{http://drops.dagstuhl.de/opus/volltexte/2019/10196/pdf/OASIcs-PLATEAU-2018-4.pdf}:
 
-Leads to bugs as analyzed by \url{http://drops.dagstuhl.de/opus/volltexte/2019/10196/pdf/OASIcs-PLATEAU-2018-4.pdf}:
 \begin{quote}
@@
 \url{https://quantum.country/search}:
+
 \begin{quote}
-  There’s a rough heuristic worth noting here, which is that you can often convert if-then style thinking into quantum circuits.
-You introduce an ancilla qubit to store the outcome of evaluating the if condition.
-And then depending on the state of the ancilla, you perform the appropriate state manipulation.
-Finally, when possible you reverse the initial computation, resetting the ancilla to its original state so you can subsequently ignore it.
+  There’s a rough heuristic worth noting here, which is that you can
+  often convert if-then style thinking into quantum circuits. You
+  introduce an ancilla qubit to store the outcome of evaluating the if
+  condition. And then depending on the state of the ancilla, you
+  perform the appropriate state manipulation. Finally, when possible
+  you reverse the initial computation, resetting the ancilla to its
+  original state so you can subsequently ignore it.
 \end{quote}
@@
 \begin{quote}
-  For error correction, one needs to replicate the state of a qubit onto a number of qubits.
-While the no cloning theorem prevents one from copying the state of a qubit directly onto another, one can create a redundant entangled qubit state of many qubits.
-The key is that the qubits to be entangled must start out in a known state.
-Qubits with a known state (for purposes of this discussion, it will be the state \verb.|0>.), called ``ancilla qubits,'' may be added to a computation for this purpose.
-Since the state of ancilla qubits are known, it is possible to create a simple circuit that makes the output state of all these ancilla qubits match the protected qubit: run each ancilla through a controlled-NOT gate, where the control is driven by the qubit that needs to be replicated.
-Assume that there is a qubit with state \verb.PSI.
-that we want to protect, where \verb.|PSI>.
-represents an arbitrary superposition state \verb.|PSI> = a0 |0> + a0 |1>.
-In the CNOT gate, the ancilla \verb.|0>.
-state will remain a \verb.|0>.
-state by the \verb.|0>.
-component of \verb.|PSI>., but it will be converted to \verb.|1>.
-by the \verb.|1>.
-component of \verb.|PSI>.
-The result of this operation is the newly entangled two-qubit state \verb.a0 |00> + a1 |11>., creating a system in which the ancilla qubit is now perfectly entangled with the first qubit.
-Adding more ancillas increases the distance of the repetition code.
-\end{quote}
+  For error correction, one needs to replicate the state of a qubit
+  onto a number of qubits. While the no cloning theorem prevents one
+  from copying the state of a qubit directly onto another, one can
+  create a redundant entangled qubit state of many qubits. The key is
+  that the qubits to be entangled must start out in a known
+  state. Qubits with a known state (for purposes of this discussion,
+  it will be the state \verb.|0>.), called ``ancilla qubits,'' may be
+  added to a computation for this purpose. Since the state of ancilla
+  qubits are known, it is possible to create a simple circuit that
+  makes the output state of all these ancilla qubits match the
+  protected qubit: run each ancilla through a controlled-NOT gate,
+  where the control is driven by the qubit that needs to be
+  replicated. Assume that there is a qubit with state \verb.PSI. that
+  we want to protect, where \verb.|PSI>. represents an arbitrary
+  superposition state \verb.|PSI> = a0 |0> + a0 |1>. In the CNOT gate,
+  the ancilla \verb.|0>. state will remain a \verb.|0>. state by the
+  \verb.|0>. component of \verb.|PSI>., but it will be converted to
+  \verb.|1>. by the \verb.|1>. component of \verb.|PSI>. The result of
+  this operation is the newly entangled two-qubit state
+  \verb.a0 |00> + a1 |11>., creating a system in which the ancilla
+  qubit is now perfectly entangled with the first qubit. Adding more
+  ancillas increases the distance of the repetition code.  \end{quote}
 
@@
 \begin{quote}
-  Imagine a 3-qubit quantum majority code in which a logical “0” is encoded as “000” and a logical “1” is encoded as “111.”  Just as with a classical majority code, a single bit-flip error can be corrected by restoring to the majority value.
-Unlike a classical code, however, we can not directly measure the qubits else their quantum state will be destroyed.
-Instead, we measure syndromes from each possible pair of qubits by interacting them with an ancilla, then measure each ancilla.
-Although the errors to the qubits are actually continuous, the effect of measuring the ancilla is to discretize the errors, as well as inform us whether an error occurred so that it can be corrected.
-With this methodology, quantum states are restored in a modular way for even a large quantum computer.
-Furthermore, operations on error-corrected qubits can be viewed as digital rather than analog, and only a small number of universal operations (H, T, CNOT) are needed for universal quantum computation.
-Through careful design and engineering, error correction codes and this small set of precise operations will lead to machines that could support practical quantum computation.
+  Imagine a 3-qubit quantum majority code in which a logical “0” is
+  encoded as “000” and a logical “1” is encoded as “111.”  Just as
+  with a classical majority code, a single bit-flip error can be
+  corrected by restoring to the majority value.  Unlike a classical
+  code, however, we can not directly measure the qubits else their
+  quantum state will be destroyed.  Instead, we measure syndromes from
+  each possible pair of qubits by interacting them with an ancilla,
+  then measure each ancilla.  Although the errors to the qubits are
+  actually continuous, the effect of measuring the ancilla is to
+  discretize the errors, as well as inform us whether an error
+  occurred so that it can be corrected.  With this methodology,
+  quantum states are restored in a modular way for even a large
+  quantum computer.  Furthermore, operations on error-corrected qubits
+  can be viewed as digital rather than analog, and only a small number
+  of universal operations (H, T, CNOT) are needed for universal
+  quantum computation.  Through careful design and engineering, error
+  correction codes and this small set of precise operations will lead
+  to machines that could support practical quantum computation.
 \end{quote}
@@
 \begin{quote}
-Using an ancilla qubit, it is possible to avoid doing any intermediate measurements in a quantum circuit.
-Show how this can be done.
-Hint: instead of measuring the qubit, apply a CNOT that “copies” it to a new \verb.|0>.-qubit, which is then left alone until the end of the computation.
-Analyze what happens.
+  Using an ancilla qubit, it is possible to avoid doing any
+  intermediate measurements in a quantum circuit. Show how this can be
+  done.  Hint: instead of measuring the qubit, apply a CNOT that
+  “copies” it to a new \verb.|0>.-qubit, which is then left alone
+  until the end of the computation. Analyze what happens.
 \end{quote}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Reversible/Quantum Circuits and Ancilla Wires}
-
-Early use of ancillas in Toffoli's paper to implement arbitrary reversible functions using a fixed number of 3 input gates \url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}
+Reversible/Quantum Circuits and Ancilla Wires: Early use of ancillas
+in Toffoli's paper to implement arbitrary reversible functions using a
+fixed number of 3 input gates
+\url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{A Comonad for Pi}
+A Comonad for Pi:
 
 Adding variables to Pi first?
-
-\newcommand{\alt}{~|~}
-\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
-\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
-\newcommand{\pointed}[2]{[#1 \bullet #2]}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Fractional Pi}
-
 Adding variables to Pi first?
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Denotations}
+Denotations:
 
-Positive rational numbers are a model. Apparently there is a categorification \url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
+Positive rational numbers are a model. Apparently there is a
+categorification
+\url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Acknowledgments
-\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
-
-%% Bibliography
-%\bibliography{bibfile}
-
-%% Appendix
+\bibliography{../cites.bib}
 \appendix
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Appendix}
 
-Text of appendix \ldots
+\end{document}
 
-\end{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-10-29 fracGC/p.tex
paragraph

@@
 \bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}
 
@@
 \section{Introduction}
+
+\paragraph*{Conservation of Information and Negative Entropy.} 
+According to the conventional theory of
+communication~\cite{Shannon1948}, a type with $N$ values is viewed as
+an abstract system that has $N$ distinguishable states where the
+amount of information contained in each state is $\log{N}$. This
+entropy is a measure of information which materializes itself in
+memory or bandwidth requirements when storing or transmitting elements
+of this type. Thus a type with 8 elements needs 3 bits of memory for
+storage or 3 bits of bandwidth for communication. The logarithmic map
+implies that information contained in a composite state is the sum of
+the information contained in its constituents. For example, the type
+$A \times B$ where $A$ has two elements and $B$ has three elements can
+be thought of a composite system consisting of two independent
+unrelated subsystems.  Each state of the composite system therefore
+contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
+the information contained in each subsystem.
+
+If quantum field theory is correct (as it so far seems to be) then
+information, during any physical process, is neither created nor
+destroyed. Landauer~\cite{Landauer:1961,Landauer,bennett1985fundamental},
+Bennett~\cite{bennett2010notes,bennett2003notes,Bennett:1973:LRC},
+Fredkin~\cite{fredkin1982conservative} and others made compelling
+arguments that this physical principle induces a corresponding
+computational principle of ``conservation of information.'' In the
+context of finite types, generated from the empty type 0, the unit
+type 1, and sums and products, this principle states that the
+foundational (i.e., physical) notion of computation is computation via
+type isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
+equivalences~\cite{DBLP:conf/esop/CaretteS16}; these are both sound
+and complete with respect to cardinality-preserving maps. In this
+context, a \emph{fractional type} $\frac{1}{A}$ introduces negative
+entropy. For example, a type with cardinality $\frac{1}{8}$ has
+entropy $\log{\frac{1}{8}} = -3$. It is natural to interpret this
+negative entropy just like we interpret ``negative money,'' as a
+resource (space or bandwidth) to be repaid (reclaimed) by some other
+part of the system.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Motivating Example}
 
@@
 \bibliography{../cites.bib}
-\appendix
-\section{Appendix}
 

### 2019-10-30 fracGC/p.tex
a bit on intro

--- 2019-10-29 fracGC/p.tex
@@
 \section{Introduction}
+
+\paragraph*{Conservation of Information.} If quantum field theory is
+correct (as it so far seems to be) then information, during any
+physical process, is neither created nor
+destroyed. Landauer~\cite{Landauer:1961,Landauer,bennett1985fundamental},
+Bennett~\cite{bennett2010notes,bennett2003notes,Bennett:1973:LRC},
+Fredkin~\cite{fredkin1982conservative} and others made compelling
+arguments that this physical principle induces a corresponding
+computational principle of ``conservation of information.'' This
+principle characterizes quantum computing and its classical
+restriction known as reversible computing.
+
+\paragraph*{Quantum (Reversible) Computing.} The quantum circuit model
+can be characterized as a classical reversible circuit model augmented
+with parallel operations modeling superpositions. Computations in the
+classical submodel can be formalized as isomorphisms on finite types,
+i.e., the types generated from the empty type 0, the unit type 1, and
+sums and products. Put differently, the principle of conservation of
+information is expressed as computating via sound and complete type
+isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
+equivalences~\cite{DBLP:conf/esop/CaretteS16}.
+
+\paragraph*{Negative Entropy.}  
+According to the conventional theory of
+communication~\cite{Shannon1948}, a type with $N$ values is viewed as
+an abstract system that has $N$ distinguishable states where the
+amount of information contained in each state is $\log{N}$. This
+entropy is a measure of information which materializes itself in
+memory or bandwidth requirements when storing or transmitting elements
+of this type. Thus a type with 8 elements needs 3 bits of memory for
+storage or 3 bits of bandwidth for communication. The logarithmic map
+implies that information contained in a composite state is the sum of
+the information contained in its constituents. For example, the type
+$A \times B$ where $A$ has two elements and $B$ has three elements can
+be thought of a composite system consisting of two independent
+unrelated subsystems.  Each state of the composite system therefore
+contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
+the information contained in each subsystem. A \emph{fractional type}
+$\frac{1}{A}$ introduces negative entropy. For example, a type with
+cardinality $\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is
+natural to interpret this negative entropy just like we interpret
+``negative money,'' as a resource (space or bandwidth) to be repaid
+(reclaimed) by some other part of the system.
+
+\paragraph*{Temporary Storage using Ancilla Bits.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Motivating Example}
 

### 2019-10-30 fracGC/p.tex
more on intro

@@
 computational principle of ``conservation of information.'' This
-principle characterizes quantum computing and its classical
-restriction known as reversible computing.
+principle is indeed one of the defining characteristics of quantum
+computing and its classical restriction known as reversible computing.
 
-\paragraph*{Quantum (Reversible) Computing.} The quantum circuit model
-can be characterized as a classical reversible circuit model augmented
-with parallel operations modeling superpositions. Computations in the
-classical submodel can be formalized as isomorphisms on finite types,
-i.e., the types generated from the empty type 0, the unit type 1, and
-sums and products. Put differently, the principle of conservation of
-information is expressed as computating via sound and complete type
-isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
-equivalences~\cite{DBLP:conf/esop/CaretteS16}.
+\paragraph*{Quantum and Reversible Computing.} The Toffoli gate is
+known to be universal for classical reversible
+circuits~\cite{Toffoli:1980}. Adding just one gate (the Hadamard gate)
+produces a universal set of primitives for quantum
+circuits~\cite{hadtoffuniv}. The ``only'' difference between the two
+circuit models is that quantum circuits can process superpositions of
+values (waves) in one step whereas classical circuits lack this form
+of parallelism. Most importantly, structurally the two circuits models
+are identical and one can derive properties valid for both families by
+focusing on the simpler classical model.
+
+Instead of using the Toffoli gate as a universal primitive for
+reversible classical circuits, one can exploit the full force of type
+theory and category theory by expressing reversible classical
+computations as \emph{isomorphisms over finite
+  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
+\emph{equivalences over
+  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
+exposes interesting mathematical structure in reversible computations
+that we will exploit to solve the ``ancilla problem'' explained next.
+
+\paragraph*{Temporary Storage using Ancilla Bits.}
 
@@
 (reclaimed) by some other part of the system.
-
-\paragraph*{Temporary Storage using Ancilla Bits.}
 

### 2019-10-31 fracGC/p.tex
more on intro

@@
 are identical and one can derive properties valid for both families by
-focusing on the simpler classical model.
+focusing on the simpler classical model. In fact, classical reversible
+computations are often used as ``subroutines'' of quantum
+computations.
 
 Instead of using the Toffoli gate as a universal primitive for
-reversible classical circuits, one can exploit the full force of type
+reversible classical circuits, one can leverage the full force of type
 theory and category theory by expressing reversible classical
@@
 
-\paragraph*{Temporary Storage using Ancilla Bits.}
+\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
+the Toffoli gate for classical reversible computing and of the
+combination of the Toffoli and Hadamard gates for quantum computing
+should not distract from efficiency and safety concerns. The theorems
+proving universality (i) assume that an unlimited amount of temporary
+storage may be
+used~\cite{Toffoli:1980,DBLP:conf/innovations/AaronsonGS17}, and (ii)
+that this temporary storage is returned to its initial state before
+de-allocation. The latter condition is necessary in order maintain
+symmetry between input and output in reversible circuits and to avoid
+decoherence problems in quantum computing, Unfortunately, in current
+proposals for reversible and quantum programming languages
+(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this condition
+is at the programmer's discretion.
 

### 2019-10-31 fracGC/p.tex
more on intro

@@
 should not distract from efficiency and safety concerns. The theorems
-proving universality (i) assume that an unlimited amount of temporary
-storage may be
-used~\cite{Toffoli:1980,DBLP:conf/innovations/AaronsonGS17}, and (ii)
-that this temporary storage is returned to its initial state before
-de-allocation. The latter condition is necessary in order maintain
-symmetry between input and output in reversible circuits and to avoid
-decoherence problems in quantum computing, Unfortunately, in current
-proposals for reversible and quantum programming languages
-(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this condition
-is at the programmer's discretion.
+proving universality (i) assume that temporary storage (often called
+\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
+this temporary storage is returned to its initial state before
+de-allocation. Indeed if no temporary storage is allowed, the Toffoli
+gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
+we demonstrate using space-time tradeoffs in Sec.~\ref{sec:examples},
+the more temporary storage is allowed, the more efficient certain
+computations could become. The condition requiring that the temporary
+storage is only de-allocated when returned to its initial condition is
+a safety condition. Violating this condition destroys the symmetry
+between input and output making the circuits not reversible and, in
+the quantum model, causes irreversible decoherence problems. As
+reviewed in Sec.~\ref{sec:examples}, current proposals for reversible
+and quantum programming languages
+(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this safety
+condition is at the programmer's discretion whereas.
 
@@
 ``negative money,'' as a resource (space or bandwidth) to be repaid
-(reclaimed) by some other part of the system.
+(reclaimed) by some other part of the system. Indeed, we will
+introduce such fractional types and use them to represent ``garbage
+collection processes'' that reclaim temporary storage.
+
+\paragraph*{Contributions.}
+
+\paragraph*{Outline.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Motivating Example}
+\section{Motivating Examples}
+\label{sec:examples}
+
+Toffoli4 using 2 Toffoli3: core of proof of universality; simple
+enough
+
+In-place matrix transpose: ease of programming, efficiency
+
+Example(s) from Quipper etc. with focus on safety condition
 

### 2019-10-31 fracGC/p.tex
more on intro

@@
 the quantum model, causes irreversible decoherence problems. As
-reviewed in Sec.~\ref{sec:examples}, current proposals for reversible
-and quantum programming languages
-(e.g. Quipper~\cite{Green:2013:QSQ:2491956.2462177}), this safety
-condition is at the programmer's discretion whereas.
+reviewed in Sec.~\ref{sec:examples}, ancilla have a number of critical
+applications and yet are poorly supported in current reversible and
+quantum programming languages making them a common source of bugs.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Motivating Examples}
+\section{Ancilla Bits}
 \label{sec:examples}
+
+current proposals for reversible
+and quantum programming languages
+(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
+Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
+to be used in nested scopes, have sound but incomplete laws for
+reasoning about the safety condition, and only guarantee safety via
+runtime checks.
 
@@
 
-In-place matrix transpose: ease of programming, efficiency
-
-Example(s) from Quipper etc. with focus on safety condition
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extra for now}
-
-\begin{itemize}
-
-\item Quantum computing hot.
-
-\item Unitary evolution; many algorithms promote classical reversible
-functions into unitary gates (examples, including some from Quipper)
-
-\item Common case: given a permutation between finite sets, find a
-  reversible program that implements it. Answer depends on the
-  available reversible primitives.
-
-\item Example, say we want to transpose a matrix, i.e., write a
-  permutation of type \verb|A*B <-> B*A| where \verb|A| and \verb|B|
-  are the dimensions. Wikipedia example
+In-place matrix transpose: ease of programming, efficiency. Say we
+want to transpose a matrix. Wikipedia example
 (\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
@@
 \end{verbatim}
-  Here \verb|A| is \verb|Nat x Nat| and \verb|B| is
-  \verb|Nat x Nat x Nat x Nat| and the input and output matrices are:
+In Pi, the input and output matrices are:
 \begin{verbatim}
@@
 
-\item So many reversible/quantum circuits allow the introduction of
-  so-called ancilla wires, which serve as scratch space (goes back to
-  Toffoli). The problem is that there is a discipline for using
-  ancilla bits: they need to be created in a known state and can only
-  be gc'ed when they are returned to the same state (explain).
+Example(s) from Quipper etc. with focus on safety condition. Quipper
+uses ancilla bits in several places; one use is to compile
+irreversible circuits to reversible ones; need ancilla bits; more
+generally several quantum algorithms need ancilla bits (see below); in
+quipper de-allocation left to programmer.
 
-\item No language knows how to automatically guarantee this condition;
-  left to the programmer (Quipper), or some (incomplete?) dynamic
-  checks (Ricercar).
-
-\item We solve this using **fractional types**.
-
-\end{itemize}
-
-Complexity perspective:
-\url{https://www.scottaaronson.com/papers/gates.pdf}
-
-Quipper: \url{https://arxiv.org/pdf/1304.3390.pdf}; uses ancilla; one
-use is to compile irreversible circuits to reversible ones; need
-ancilla bits; more generally several quantum algorithms need ancilla
-bits (see below); in quipper de-allocation left to programmer.
-
-Ricercar tries to define rules for ancilla but not complete
-\url{https://msoeken.github.io/papers/2015_rc_2.pdf}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Extra for now}
 

### 2019-10-31 fracGC/p.tex
a bit of reorg

@@
 the quantum model, causes irreversible decoherence problems. As
-reviewed in Sec.~\ref{sec:examples}, ancilla have a number of critical
+reviewed in Sec.~\ref{sec:examples}, ancilla bits have a number of critical
 applications and yet are poorly supported in current reversible and
@@
 
-current proposals for reversible
-and quantum programming languages
-(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
-Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
-to be used in nested scopes, have sound but incomplete laws for
-reasoning about the safety condition, and only guarantee safety via
-runtime checks.
-
-Toffoli4 using 2 Toffoli3: core of proof of universality; simple
-enough
-
-In-place matrix transpose: ease of programming, efficiency. Say we
-want to transpose a matrix. Wikipedia example
-(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
-\begin{verbatim}
-11 12 13 14 
-21 22 23 24 
-\end{verbatim}
-transposed to
-\begin{verbatim}
-11 21
-12 22
-13 23
-14 24
-\end{verbatim}
-In Pi, the input and output matrices are:
-\begin{verbatim}
-M   = (11 , 21) , (12 , 22) , (13, 23) , (14 , 24)
-
-trM = (11 , 12 , 13 , 14) , (21 , 22 , 23 , 23) 
-\end{verbatim}
-Say we are given a language like Pi that is sound and complete with
-respect to permutations on finite types, we would write the
-permutation like so.
-\begin{verbatim}
-WRITE PERMUTATION
-\end{verbatim}
-This code does not use additional space, i.e., it performs the matrix
-transpose in constant space, i.e., performs an in-place matrix
-transposition. It is well know that with additional space, one can
-write more efficient matrix transpositions (explain and citations).
-
-Example(s) from Quipper etc. with focus on safety condition. Quipper
-uses ancilla bits in several places; one use is to compile
-irreversible circuits to reversible ones; need ancilla bits; more
-generally several quantum algorithms need ancilla bits (see below); in
-quipper de-allocation left to programmer.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Extra for now}
-
-Leads to bugs as analyzed by
-\url{http://drops.dagstuhl.de/opus/volltexte/2019/10196/pdf/OASIcs-PLATEAU-2018-4.pdf}:
-
-\begin{quote}
-
-  Bug type 6: Incorrect composition of operations using mirroring Section 4.5 discussed how bugs in deallocating ancillary qubits can happen due to bad parameters.
-Here we see how bugs in deallocating ancillary qubits can happen due to incorrect composition of operations following a mirroring pattern.
-For example, in Table 7, the operations in rows 2 and 3 are respectively mirrored and undone in rows 6 and 5.
-These lines of code need careful reversal of every loop and every operation.
-
-  A common pattern in quantum programs involves performing operations (e.g., add), contingent on a set of qubits known as control qubits.
-Without language support, this pattern needs many lines of code and manual allocation of ancillary qubits.
-In the Scaffold code example in Table 7, rows 3 and 5 are just computing the intersection of qubits q, with the help of ancillary qubits initialized in row 1, in order to realize the controlled rotation operation in row 4.
-Furthermore, quantum algorithms often need varying numbers of control qubits in different parts of the algorithm, leading to replicated code from multiple versions of the same subroutine differing only by the number of control qubits.
-
-\end{quote}
+%%%
+\subsection{Applications}
 
@@
 
-Reversible/Quantum Circuits and Ancilla Wires: Early use of ancillas
-in Toffoli's paper to implement arbitrary reversible functions using a
-fixed number of 3 input gates
+%%%
+\subsection{Language Support}
+
+current proposals for reversible
+and quantum programming languages
+(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
+Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
+to be used in nested scopes, have sound but incomplete laws for
+reasoning about the safety condition, and only guarantee safety via
+runtime checks.
+
+Leads to bugs as analyzed by
+\url{http://drops.dagstuhl.de/opus/volltexte/2019/10196/pdf/OASIcs-PLATEAU-2018-4.pdf}:
+
+\begin{quote}
+
+  Bug type 6: Incorrect composition of operations using mirroring Section 4.5 discussed how bugs in deallocating ancillary qubits can happen due to bad parameters.
+Here we see how bugs in deallocating ancillary qubits can happen due to incorrect composition of operations following a mirroring pattern.
+For example, in Table 7, the operations in rows 2 and 3 are respectively mirrored and undone in rows 6 and 5.
+These lines of code need careful reversal of every loop and every operation.
+
+  A common pattern in quantum programs involves performing operations (e.g., add), contingent on a set of qubits known as control qubits.
+Without language support, this pattern needs many lines of code and manual allocation of ancillary qubits.
+In the Scaffold code example in Table 7, rows 3 and 5 are just computing the intersection of qubits q, with the help of ancillary qubits initialized in row 1, in order to realize the controlled rotation operation in row 4.
+Furthermore, quantum algorithms often need varying numbers of control qubits in different parts of the algorithm, leading to replicated code from multiple versions of the same subroutine differing only by the number of control qubits.
+
+\end{quote}
+
+%%%
+\subsection{Examples}
+
+Toffoli4 using 2 Toffoli3: core of proof of universality; simple
+enough. Reversible/Quantum Circuits and Ancilla Wires: Early use of
+ancillas in Toffoli's paper to implement arbitrary reversible
+functions using a fixed number of 3 input gates
 \url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}
 
-Use examples from Ricercar
+In-place matrix transpose: ease of programming, efficiency. Say we
+want to transpose a matrix. Wikipedia example
+(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
+\begin{verbatim}
+11 12 13 14 
+21 22 23 24 
+\end{verbatim}
+transposed to
+\begin{verbatim}
+11 21
+12 22
+13 23
+14 24
+\end{verbatim}
+In Pi, the input and output matrices are:
+\begin{verbatim}
+M   = (11 , 21) , (12 , 22) , (13, 23) , (14 , 24)
+
+trM = (11 , 12 , 13 , 14) , (21 , 22 , 23 , 23) 
+\end{verbatim}
+Say we are given a language like Pi that is sound and complete with
+respect to permutations on finite types, we would write the
+permutation like so.
+\begin{verbatim}
+WRITE PERMUTATION
+\end{verbatim}
+This code does not use additional space, i.e., it performs the matrix
+transpose in constant space, i.e., performs an in-place matrix
+transposition. It is well know that with additional space, one can
+write more efficient matrix transpositions (explain and citations).
+
+Example(s) from Quipper etc. with focus on safety condition. Quipper
+uses ancilla bits in several places; one use is to compile
+irreversible circuits to reversible ones; need ancilla bits; more
+generally several quantum algorithms need ancilla bits (see below); in
+quipper de-allocation left to programmer.
 
@@
 
-A Comonad for Pi:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Pi}
 
@@
 
-Adding variables to Pi first?
-
-Denotations:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Space Denotational Semantics}
 

### 2019-10-31 fracGC/p.tex
more plan

@@
 
+Add eta/epsilon leading to \verb|C3.agda|. This code has a runtime
+check like current solutions but is more general allowing non-scoped
+ancilla allocation and de-allocation. Show examples.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Comonad}
+
+Comonad to keep track to value in focus
+
 Now we define pointed types $\pointed{t}{v}$ which are a record
@@
 Then we can define $\eta$ and $\epsilon$ in the comonadic world.
+
+Exploit dependent types to reify proofs in the type system. Type
+checking requires (partial) eval. Once we have a proof can we somehow
+go back to \verb|C3.agda| and remove the runtime check?
+
+Polymorphic type for ancilla?
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Trace?}
+
+Add multiplicative trace, do SAT solver
 

### 2019-11-01 fracGC/p.tex
rephrase the bit about extracting a permutation from A <-> B in the language of lens.

@@
 we can implement a permutation X <-> Z 
-when there exists Y such that A/X = Y = B/Z
+when there exists Y such that A = X * Y = Z * Y = B
 
@@
   -> Y * Z * 1/Y
+  -> Z * Y * 1/Y
   -> Z
 \end{verbatim}
+This basically says, in the language of Lens, that
+when $A$ is isomorphic to $X \cross Y$ and
+$B$ is isomorphic to $Z \cross Y$, then
+$X$ is isomorphic to $Y$.
+
+Another way to think of it: for all isomorphic
+$A$ and $B$, whenever they can be factored with a
+common complement ($Y$), then the ``other pieces''
+(here $X$ and $Y$) are automatically isomorphic.
 

### 2019-11-07 fracGC/p.tex
using new trick to typeset code; seems to work fine

@@
 \usepackage{subcaption} 
+\usepackage{amssymb}
+\usepackage{bbm}
+\usepackage[greek,english]{babel}
+\usepackage{ucs}
+\usepackage[utf8x]{inputenc}
+\usepackage{autofe}
+\usepackage[references]{agda} 
+\usepackage{newunicodechar}
+
+\newunicodechar{𝕌}{$\mathbb{U}$}
+\newunicodechar{𝟙}{$1$}
+\newunicodechar{ᵤ}{${}_u$}
 
@@
 \maketitle
+
+\input{latex/PiFracDynCode.tex}
 
@@
 This basically says, in the language of Lens, that
-when $A$ is isomorphic to $X \cross Y$ and
-$B$ is isomorphic to $Z \cross Y$, then
+when $A$ is isomorphic to $X \times Y$ and
+$B$ is isomorphic to $Z \times Y$, then
 $X$ is isomorphic to $Y$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Pi}
+\section{PiFrac}
 
-Adding variables to Pi first?
+%%%%%
+\subsection{Pi}
 
@@
 
-Add eta/epsilon leading to \verb|C3.agda|. This code has a runtime
-check like current solutions but is more general allowing non-scoped
-ancilla allocation and de-allocation. Show examples.
+%%%%%
+\subsection{Adding Fractions}
+
+%%%%%
+\subsection{Ancilla}
+
+When ancilla bits are created they are created with a specific
+value. This value is encoded in the circuit. We use a Singleton type
+to remember this information. We add the following:
+
+ETA SIG
+
+The value of fractional type represents the ``garbage collector.''
+The garbage collector is specialized to collect a single value, which
+is the value used to create the ancilla value. At this point, if the
+garbage collector encounters a different value, it doesn't know what
+to do, and would need to throw an exception. This situation in
+\verb|PiFracDyn.agda| is similar to current solutions (e.g. Quipper)
+but actually more general as it allows non-scoped ancilla allocation
+and de-allocation. Show examples.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Comonad}
+\section{Generalizing to Non-Scoped Ancilla}
 
-Comonad to keep track to value in focus
+Def. of universe is:
 
-Now we define pointed types $\pointed{t}{v}$ which are a record
-consisting of a value type together with a value of that type. Both
-the level-1 and level-2 program lift naturally to the world of pointed
-types but instead of manually rewriting everything we use the fact
-that pointed types form a comonad. So for example, given a level-1
-program we should be able to just use \textsf{extend} to compose
-instead of plain composition of combinators??
+\Udef{}
 
-Then we can define $\eta$ and $\epsilon$ in the comonadic world.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Dependently-Typed Garbage Collectors}
+
+%%%%%
+\subsection{Singleton Types}
+
+Consider the type $\Sigma \mathbb{N} (\lambda n. n \equiv 5)$. Values
+of that type are a number $m$ and a proof that this is number is equal
+to 5. So we can have $(4 + 1, \textit{refl})$, $(2 + 3 ,
+\textit{refl})$, and so on. All these syntactically different values
+are identical to 5 and hence we can \emph{contract} the type to a
+single point at 5.
+
+The denotation of the fractional type is now:
+
+RECIP
 
@@
 checking requires (partial) eval. Once we have a proof can we somehow
-go back to \verb|C3.agda| and remove the runtime check?
+go back to \verb|PiFracDyn.agda| and remove the runtime check?
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliography{../cites.bib}
+\bibliography{../../cites.bib}
 

### 2019-11-08 fracGC/p.tex
fixed bib link
reorganize paper

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{PiFrac}
+\section{Preliminaries: Pi}
+
+Our approach to reversible quantum programming based on type
+isomorphisms. Examples.
+
+Ancilla problem in the context of Pi using eta and epsilon 
+informally.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{PiFrac: Generalizing Scoped Ancilla, Runtime Check}
+
+Formalize ancilla problem in the context of Pi using eta and epsilon
+as done in PiFracDyn; more general than scoped; still runtime
+check.
+
+When ancilla bits are created they are created with a specific
+value. This value is encoded in the circuit. The value of fractional
+type represents the ``garbage collector.''  The garbage collector is
+specialized to collect a single value, which is the value used to
+create the ancilla value. At this point, if the garbage collector
+encounters a different value, it doesn't know what to do, and would
+need to throw an exception. This situation in \verb|PiFracDyn.agda| is
+similar to current solutions (e.g. Quipper) but actually more general
+as it allows non-scoped ancilla allocation and de-allocation. Show
+examples.
+
 %%%%%
 \subsection{Adding Fractions}
-
-%%%%%
-\subsection{Ancilla}
-
-When ancilla bits are created they are created with a specific
-value. This value is encoded in the circuit. We use a Singleton type
-to remember this information. We add the following:
-
-ETA SIG
-
-The value of fractional type represents the ``garbage collector.''
-The garbage collector is specialized to collect a single value, which
-is the value used to create the ancilla value. At this point, if the
-garbage collector encounters a different value, it doesn't know what
-to do, and would need to throw an exception. This situation in
-\verb|PiFracDyn.agda| is similar to current solutions (e.g. Quipper)
-but actually more general as it allows non-scoped ancilla allocation
-and de-allocation. Show examples.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Generalizing to Non-Scoped Ancilla}
 
@@
 
+%%%%%
+\subsection{Ancilla}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Dependently-Typed Garbage Collectors}
+
+We use a Singleton type
+to remember this information.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Space Denotational Semantics}
+\section{Space Denotational Semantics?}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliography{../../cites.bib}
+\bibliography{../cites.bib}
 

### 2019-11-08 fracGC/p.tex
some old text for pi intro

@@
 \usepackage{newunicodechar}
+\usepackage[nocenter]{qtree}
+\usepackage{tikz}
+\usetikzlibrary{cd}
+\usetikzlibrary{quotes}
 
@@
 \newcommand{\pointed}[2]{[#1 \bullet #2]}
+\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
+\newcommand{\mm}{\texttt{\textminus}}
+\newcommand{\pp}{\texttt{+}}
+\newcommand{\inl}[1]{\textsf{inl}~#1}
+\newcommand{\inr}[1]{\textsf{inr}~#1}
+\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
+\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
+\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
+\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
+\newcommand{\refl}[1]{\textsf{refl}~#1}
+\newcommand{\lid}{\textsf{lid}}
+\newcommand{\rid}{\textsf{rid}}
+\newcommand{\linv}{l!}
+\newcommand{\rinv}{r!}
+\newcommand{\invinv}{!!}
+\newcommand{\assoc}{\circ}
+\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
+\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
+\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
+\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
+\newcommand{\swapp}{\mathit{swap}_+}
+\newcommand{\assoclp}{\mathit{assocl}_+}
+\newcommand{\assocrp}{\mathit{assocr}_+}
+\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
+\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
+\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
+\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
+\newcommand{\swapt}{\mathit{swap}_*}
+\newcommand{\assoclt}{\mathit{assocl}_*}
+\newcommand{\assocrt}{\mathit{assocr}_*}
+\newcommand{\absorbr}{\mathit{absorbr}}
+\newcommand{\absorbl}{\mathit{absorbl}}
+\newcommand{\factorzr}{\mathit{factorzr}}
+\newcommand{\factorzl}{\mathit{factorzl}}
+\newcommand{\dist}{\mathit{dist}}
+\newcommand{\factor}{\mathit{factor}}
+\newcommand{\distl}{\mathit{distl}}
+\newcommand{\factorl}{\mathit{factorl}}
+\newcommand{\distz}{\mathit{absorbr}}
+\newcommand{\iso}{\leftrightarrow}
+\newcommand{\proves}{\vdash}
+\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
+\newcommand{\Rule}[4]{
+\makebox{{\rm #1}
+$\displaystyle
+\frac{\begin{array}{l}#2 \\\end{array}}
+{\begin{array}{l}#3      \\\end{array}}$
+ #4}}
+\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
 
@@
 
+Focusing on finite types, the building blocks of type theory are: the
+empty type ($0$), the unit type ($1$), the sum type ($+$), and the
+product ($\times$) type. Before getting into the formal theory, let's
+consider the intuition that types are modeled by (topological) spaces
+and that type equivalence is modeled by possible \emph{deformations}
+of such spaces. Consider the space $(1 + 0) \times (1 + 1)$. This
+space is the product of two subspaces: the subspace $(1 + 0)$ which
+itself is the sum of the space~$1$ containing one point $\texttt{tt}$
+and the empty space~$0$ and the subspace $(1 + 1)$ which is the sum of
+two spaces each containing the one point $\texttt{tt}$. Any
+deformation of this space must at least preserve the number of points:
+we can neither create nor destroy points during any continuous
+deformation. Seeing that the number of points in our example space is
+2, a reasonable hypothesis is that we can deform the space above to
+any other space with 2 points such as $1 + 1$ or $1 + (1 + 0)$. What
+this really means is that we are treating the sum and product
+structure as malleable. Imagining a product structure as arranged in a
+grid, by stretching we can turn this structure to a sum structure
+arranged in a line, change the orientation of the grid by exchanging
+the axes, as well as other transformations that preserve the number of
+points using various ``symmetries.'' These symmetries capture, in a
+stylized manner, a rich collection of space-time
+tradeoffs~\cite{superstructural}. We formalize this intuition by
+saying that types form a \emph{commutative semiring} (up to type
+isomorphism).
+
+\begin{figure}[t]
+\[\begin{array}{rrcll}
+\idc :& \tau & \iso & \tau &: \idc \\
+\\
+\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
+\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
+\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
+\\
+\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
+\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
+\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
+\\
+\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
+\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
+\end{array}\]
+\begin{center}
+\Rule{}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
+{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
+{}
+\\
+\Rule{}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
+{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
+{}
+\qquad
+\Rule{}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
+{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
+{}
+\end{center}
+\caption{$\Pi$-terms and combinators.}
+\label{pi-terms}
+\end{figure}
+
+Another model of types is to view each type $A$ as a collection of
+physical wires that can transmit $\|A\|$ distinct values where $\|A\|$
+is the size of a type, computed as: $\| 0 \| = 0$; $\| 1 \| = 1$; $\|
+A + B \| = \| A \| + \| B \|$; and $\| A \times B \| = \| A \| * \| B
+\|$. Thus the type $\mathbb{B} = 1 + 1$ corresponds to a wire that can
+transmit two values, i.e., bits, and the type $\mathbb{B} \times
+\mathbb{B} \times \mathbb{B}$ corresponds to a collection of wires
+that can transmit three bits. From that perspective, a type
+isomorphism between types $A$ and $B$ (such that $\|A\|=\|B\|=n$)
+models a \emph{reversible} combinational circuit that \emph{permutes}
+the $n$ different values. These type isomorphisms are collected in
+Fig.~\ref{pi-terms}. We call the resulting language $\Pi$.  It is
+known that these type isomorphisms are sound and complete for all
+permutations on finite types~\cite{Fiore:2004,fiore-remarks} and hence
+that they are \emph{complete} for expressing combinational
+circuits~\cite{fredkin1982conservative,James:2012:IE:2103656.2103667,Toffoli:1980}.
+
+So far, the types encode conventional data structures, i.e, sets of
+values and structured trees of values and the isomorphisms act on such
+conventional data structures. Universal computation models however
+fundamentally rely on the fact that \emph{programs are (or can be
+encoded as) data}, e.g., a Turing machine can be encoded as a string
+that another Turing machine (or even the same machine) can
+manipulate. In our setting, we ask whether the type isomorphisms in
+Fig.~\ref{pi-terms} can themselves be subject to (higher-level)
+reversible deformations? Before developing the theory, let's consider
+a small example consisting of two deformations between the types $A +
+B$ and $C+D$:
+
+\begin{center}
+\begin{tikzpicture}[scale=0.4,every node/.style={scale=0.5}]
+  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
+  \draw (-2,-2) ellipse (0.9cm and 1cm);
+  \draw[fill] (-2,-1.5) circle [radius=0.025];
+  \node[left] at (-2.1,-1.5) {$A$};
+  \draw[fill] (-2,-2.5) circle [radius=0.025];
+  \node[left] at (-2.1,-2.5) {$B$};
+
+  \draw (6.5,-2) ellipse (0.9cm and 1cm);
+  \draw[fill] (6.5,-1.5) circle [radius=0.025];
+  \node[right] at (6.7,-1.5) {$C$};
+  \draw[fill] (6.5,-2.5) circle [radius=0.025];
+  \node[right] at (6.7,-2.5) {$D$};
+
+  \draw[-] (-2,-1.5) to[bend left] (1,0.5) ;
+  \draw[-] (-2,-2.5) to[bend left] (1,-0.5) ;
+  \draw[-] (3.5,0.5) to[bend left] (6.5,-1.45) ;
+  \draw[-] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
+
+  \draw[-] (-2,-1.5) to[bend right] (1,-3.5) ;
+  \draw[-] (-2,-2.5) to[bend right] (1,-4.5) ;
+  \draw[-] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
+  \draw[-] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
+
+  \draw     (2.5,-3)  -- (3.5,-3) -- (3.5,-4) -- (2.5,-4) -- cycle ;
+  \draw     (2.5,-4)  -- (3.5,-4) -- (3.5,-5) -- (2.5,-5) -- cycle ;
+
+  \draw     (1,1)  -- (2,1) -- (2,0) -- (1,0) -- cycle ;
+  \draw     (1,0)  -- (2,0) -- (2,-1) -- (1,-1) -- cycle ;
+
+  \node at (3,-3.5) {$c_2$};
+  \node at (3,-4.5) {$c_1$};
+
+  \node at (1.5,0.5) {$c_1$};
+  \node at (1.5,-0.5) {$c_2$};
+
+  \draw     (2,0.5)  -- (2.5,0.5)  ;
+  \draw     (2,-0.5) -- (2.5,-0.5) ;
+
+  \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
+  \draw     (2.5,-0.5) -- (3.5,0.5) ;
+
+  \draw     (1,-3.5)  -- (2,-4.5)    ;
+  \draw     (1,-4.5) -- (2,-3.5)   ;
+
+  \draw     (2,-3.5)  -- (2.5,-3.5)    ;
+  \draw     (2,-4.5) -- (2.5,-4.5)   ;
+
+\end{tikzpicture}
+\end{center}
+
+\noindent The top path is the $\Pi$ program
+$(c_1~\oplus~c_2)~\odot~\swapp$ which deforms the type $A$ by $c_1$,
+deforms the type $B$ by $c_2$, and deforms the resulting space by a
+twist that exchanges the two injections into the sum type. The bottom
+path performs the twist first and then deforms the type $A$ by $c_1$
+and the type $B$ by $c_2$ as before. If one could imagine the paths
+  as physical wires and the deformations $c_1$ and $c_2$ as arbitrary
+deformations on these wires then, holding the points $A$, $B$, $C$,
+and $D$ fixed, it is possible to rotate the top part of the diagram to
+become identical to the bottom one. That rotation can be undone
+(reversed), which takes the bottom part of the diagram into the top
+part.  In other words, there exists a deformation of the program
+$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
+$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means
+that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
+$\swapp \odot (c_2~\oplus~c_1)$ are equal. This relation is
+non-trivial, as not all programs between the same types can be
+deformed into one another. The simplest example of inequivalent
+deformations are the two automorphisms of $1+1$, namely $\idc$ and
+$\swapp$.
+
+Developing a collection of ``templates'' for such higher-level
+equivalences is reasonably straightforward (see for example the work
+of Miller et al~\cite{Miller:2003:TBA:775832.775915}). To find a
+\emph{complete} set of equivalences is much more difficult. The idea
+developed in our previous work~\cite{DBLP:conf/esop/CaretteS16} relies
+on ``categorification'' and a classical result by
+Laplaza~\cite{laplaza}. We refer the interested reader to our previous
+work for the full set of \emph{complete} ``templates.'' A related
+idea~\cite{DBLP:journals/corr/CockettCS17} for a sound and complete
+set of equivalences on control-not gates is also developed in the
+context of the same monoidal categories as the ones we use suggesting
+that these categorical structures provide a common framework for
+reasoning about various classes of reversible circuits.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-11-09 fracGC/p.tex
a bit on paper

@@
 \usepackage{amssymb}
+\usepackage{stmaryrd}
 \usepackage{bbm}
@@
 \newunicodechar{𝕌}{$\mathbb{U}$}
+\newunicodechar{𝔽}{$\mathbb{F}$}
+\newunicodechar{𝕋}{$\mathbb{T}$}
+\newunicodechar{𝟘}{$0$}
 \newunicodechar{𝟙}{$1$}
+\newunicodechar{𝟚}{$2$}
+\newunicodechar{⋆}{${}_*$}
 \newunicodechar{ᵤ}{${}_u$}
+\newunicodechar{⊚}{$\fatsemi$}
 
@@
 \newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
+\newcommand{\ket}[1]{|#1\rangle}
 \newcommand{\pointed}[2]{[#1 \bullet #2]}
@@
   state. Qubits with a known state (for purposes of this discussion,
-  it will be the state \verb.|0>.), called ``ancilla qubits,'' may be
+  it will be the state $\ket{0}$), called ``ancilla qubits,'' may be
   added to a computation for this purpose. Since the state of ancilla
@@
   where the control is driven by the qubit that needs to be
-  replicated. Assume that there is a qubit with state \verb.PSI. that
-  we want to protect, where \verb.|PSI>. represents an arbitrary
-  superposition state \verb.|PSI> = a0 |0> + a0 |1>. In the CNOT gate,
-  the ancilla \verb.|0>. state will remain a \verb.|0>. state by the
-  \verb.|0>. component of \verb.|PSI>., but it will be converted to
-  \verb.|1>. by the \verb.|1>. component of \verb.|PSI>. The result of
-  this operation is the newly entangled two-qubit state
-  \verb.a0 |00> + a1 |11>., creating a system in which the ancilla
+  replicated. Assume that there is a qubit with state $\Psi$ that we
+  want to protect, where $\ket{\Psi}$ represents an arbitrary
+  superposition state $\ket{\Psi} = a_0 \ket{0} + a_1 \ket{1}$. In the
+  CNOT gate, the ancilla $\ket{0}$ state will remain a $\ket{0}$ state
+  by the $\ket{0}$ component of $\ket{\Psi}$, but it will be converted
+  to $\ket{1}$ by the $\ket{1}$ component of $\ket{\Psi}$. The result
+  of this operation is the newly entangled two-qubit state
+  $a_0 \ket{0} + a_1 \ket{1}$, creating a system in which the ancilla
   qubit is now perfectly entangled with the first qubit. Adding more
@@
   done.  Hint: instead of measuring the qubit, apply a CNOT that
-  “copies” it to a new \verb.|0>.-qubit, which is then left alone
-  until the end of the computation. Analyze what happens.
+  “copies” it to a new $\ket{0}$-qubit, which is then left alone until
+  the end of the computation. Analyze what happens.
 \end{quote}
@@
 \begin{verbatim}
-M   = (11 , 21) , (12 , 22) , (13, 23) , (14 , 24)
+M = (11,21) , (12,22) , (13,23) , (14,24)
 
-trM = (11 , 12 , 13 , 14) , (21 , 22 , 23 , 23) 
+trM = (11,12,13,14) , (21,22,23,23) 
 \end{verbatim}
@@
 Is the below a possible example?
-\begin{verbatim}
-Say we already have a permutation A <-> B
-we can implement a permutation X <-> Z 
-when there exists Y such that A = X * Y = Z * Y = B
 
-X -> X * Y * 1/Y 
-  -> A * 1/Y
-  -> B * 1/Y
-  -> Y * Z * 1/Y
-  -> Z * Y * 1/Y
-  -> Z
-\end{verbatim}
+Say we already have a permutation $A \leftrightarrow B$
+we can implement a permutation $X \leftrightarrow Z$ 
+when there exists $Y$ such that $A = X * Y = Z * Y = B$:
+\[\begin{array}{rcl}
+X &\rightarrow& X * Y * 1/Y \\
+  &\rightarrow& A * 1/Y \\
+  &\rightarrow&  B * 1/Y \\
+  &\rightarrow&  Y * Z * 1/Y \\
+  &\rightarrow&  Z * Y * 1/Y \\
+  &\rightarrow&  Z
+\end{array}\]
 This basically says, in the language of Lens, that
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 \[\begin{array}{rrcll}
@@
 \label{pi-terms}
-\end{figure}
+\end{figure*}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{PiFrac: Generalizing Scoped Ancilla, Runtime Check}
+\section{PiFrac: Generalizing Scoped Ancilla with Dynamic Checks}
 
@@
 encounters a different value, it doesn't know what to do, and would
-need to throw an exception. This situation in \verb|PiFracDyn.agda| is
+need to throw an exception. This situation in \verb|PiFracDyn| is
 similar to current solutions (e.g. Quipper) but actually more general
@@
 
+New combinators:
+
+\EtaEpsilon{}
+
+and their evaluation:
+
+\EtaEpsilonEval{}
+
 %%%%%
-\subsection{Ancilla}
+\subsection{Examples}
+
+\EtaEpsilonExamples{}
 
@@
 checking requires (partial) eval. Once we have a proof can we somehow
-go back to \verb|PiFracDyn.agda| and remove the runtime check?
+go back to \verb|PiFracDyn| and remove the runtime check?
 

### 2019-11-13 fracGC/p.tex
organizstion of paper
figure
giant experiment to see if we can get better notation and have a smoother story

@@
 \usepackage{tikz}
+\usepackage{bbold}
 \usetikzlibrary{cd}
@@
 \newunicodechar{𝕋}{$\mathbb{T}$}
-\newunicodechar{𝟘}{$0$}
-\newunicodechar{𝟙}{$1$}
-\newunicodechar{𝟚}{$2$}
+\newunicodechar{𝟘}{$\mathbb{0}$}
+\newunicodechar{𝟙}{$\mathbb{1}$}
+\newunicodechar{𝟚}{$\mathbb{2}$}
 \newunicodechar{⋆}{${}_*$}
@@
 \newcommand{\jdg}[3]{#2 \proves_{#1} #3}
+
+% 12 pages
 
@@
   groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
-exposes interesting mathematical structure in reversible computations
+exposes a rich mathematical structure in reversible computations
 that we will exploit to solve the ``ancilla problem'' explained next.
@@
 the Toffoli gate for classical reversible computing and of the
-combination of the Toffoli and Hadamard gates for quantum computing
-should not distract from efficiency and safety concerns. The theorems
-proving universality (i) assume that temporary storage (often called
-\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
-this temporary storage is returned to its initial state before
-de-allocation. Indeed if no temporary storage is allowed, the Toffoli
-gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
-we demonstrate using space-time tradeoffs in Sec.~\ref{sec:examples},
-the more temporary storage is allowed, the more efficient certain
-computations could become. The condition requiring that the temporary
-storage is only de-allocated when returned to its initial condition is
-a safety condition. Violating this condition destroys the symmetry
-between input and output making the circuits not reversible and, in
-the quantum model, causes irreversible decoherence problems. As
-reviewed in Sec.~\ref{sec:examples}, ancilla bits have a number of critical
-applications and yet are poorly supported in current reversible and
-quantum programming languages making them a common source of bugs.
+universality of the combination of the Toffoli and Hadamard gates for
+quantum computing should not distract from efficiency and safety
+concerns. The theorems proving universality (i) assume that temporary
+storage (often called \emph{ancilla bits}) may be
+used~\cite{Toffoli:1980}, and (ii) that this temporary storage is
+returned to its initial state before de-allocation. Indeed if no
+temporary storage is allowed, the Toffoli gate is not
+universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as we
+demonstrate in Sec.~\ref{sec:examples}, the more temporary storage is
+allowed, the more efficient certain computations could become. The
+condition requiring that the temporary storage is only de-allocated
+when returned to its initial condition is a safety
+condition. Violating this condition destroys the symmetry between
+input and output making the circuits not reversible and, in the
+quantum model, causes decoherence that may destroy the quantum
+state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
+number of critical applications and yet are poorly supported in
+current reversible and quantum programming languages making them a
+common source of bugs.
 
@@
 contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
-the information contained in each subsystem. A \emph{fractional type}
-$\frac{1}{A}$ introduces negative entropy. For example, a type with
-cardinality $\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is
-natural to interpret this negative entropy just like we interpret
-``negative money,'' as a resource (space or bandwidth) to be repaid
-(reclaimed) by some other part of the system. Indeed, we will
-introduce such fractional types and use them to represent ``garbage
-collection processes'' that reclaim temporary storage.
+the information contained in each subsystem. If we could imagine a
+\emph{fractional type} $\frac{1}{A}$, this type would introduce
+negative entropy. For example, a type with ``cardinality''
+$\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is natural to
+interpret this negative entropy just like we interpret ``negative
+money,'' as a resource (space or bandwidth) to be repaid (reclaimed)
+by some other part of the system. Indeed, we will introduce such
+fractional types and use them to represent ``garbage collection
+processes'' that reclaim temporary storage. Just like in the case of
+``negative money'' It is critical that these fractional types be
+linearly used, i.e., they cannot be duplicated or erased: this
+property is however automatically enforced in a reversible
+computational model!
 
-\paragraph*{Contributions.}
+\paragraph*{Contributions.} The paper presents a novel concept, that
+is \emph{fractional types} etc.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Trace?}
-
-Add multiplicative trace, do SAT solver
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Space Denotational Semantics?}
+\section{Space Operational Semantics}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Trace}
+
+Add multiplicative trace, do SAT solver
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliography{../cites.bib}

### 2019-11-15 fracGC/p.tex
a bit on paper

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{PiFrac: Generalizing Scoped Ancilla with Dynamic Checks}
+\section{The Value of Ancillae}
+
+The management of ancilla bits can be teased into two relatively
+distinct subproblems:
+\begin{itemize}
+\item Every allocated ancilla bit must be de-allocated, and 
+\item De-allocation of an ancilla bit is only safe if the bit is
+  restored to its allocation-time initial value.
+\end{itemize}
+In this section, we focus on the first subproblem and address the
+second in later sections.
+
+%%%%%
+\subsection{Scoped Ancilla}
+
+A conventional way to ensure that every allocation is paired with a
+matching de-allocation is to impose a stack discipline. An
+illustrative demonstration of this idea is the \verb|with_ancilla|
+operator from Quipper~\cite{Green:2013:QSQ:2491956.2462177}:
+\begin{verbatim}
+with_ancilla :: (Qubit -> Circ a) -> Circ a
+\end{verbatim}
+The operator takes a block of gates parameterized by an ancilla,
+allocates a new ancilla initialized to $\ket{0}$, and runs the given
+block of gates. At the end of its execution, the block is expected to
+return the ancilla to the state $\ket{0}$ at which point it is
+de-allocated. In this case, allocation and de-allocation of ancillae
+requires nothing beyond conventional parameter-passing techniques in
+which the parameter is allocated before entry to the function and
+de-allocated at exit.
+
+This scoped model is therefore quite a pragmatic choice. It is however
+limited and to understand its limitations more vividly, we propose the
+following analogy with money.
 

### 2019-11-15 fracGC/p.tex
working on sec 4

@@
 
-A conventional way to ensure that every allocation is paired with a
-matching de-allocation is to impose a stack discipline. An
-illustrative demonstration of this idea is the \verb|with_ancilla|
-operator from Quipper~\cite{Green:2013:QSQ:2491956.2462177}:
+A conventional and efficient way to ensure that every allocation is
+paired with a matching de-allocation is to impose a stack
+discipline. An illustrative demonstration of this idea is the
+\verb|with_ancilla| operator from
+Quipper~\cite{Green:2013:QSQ:2491956.2462177}:
 \begin{verbatim}
@@
 block of gates. At the end of its execution, the block is expected to
-return the ancilla to the state $\ket{0}$ at which point it is
-de-allocated. In this case, allocation and de-allocation of ancillae
-requires nothing beyond conventional parameter-passing techniques in
-which the parameter is allocated before entry to the function and
-de-allocated at exit.
+return the ancilla to the state $\ket{0}$ (this is the second
+subproblem addressed later) at which point it is de-allocated. In this
+scoped-by-construction approach, allocation and de-allocation of
+ancillae requires nothing beyond conventional parameter-passing
+techniques in which the parameter is allocated before entry to a
+function and de-allocated at exit.
 
@@
 limited and to understand its limitations more vividly, we propose the
-following analogy with money.
-
-Formalize ancilla problem in the context of Pi using eta and epsilon
-as done in PiFracDyn; more general than scoped; still runtime
-check.
-
-When ancilla bits are created they are created with a specific
-value. This value is encoded in the circuit. The value of fractional
-type represents the ``garbage collector.''  The garbage collector is
-specialized to collect a single value, which is the value used to
-create the ancilla value. At this point, if the garbage collector
-encounters a different value, it doesn't know what to do, and would
-need to throw an exception. This situation in \verb|PiFracDyn| is
-similar to current solutions (e.g. Quipper) but actually more general
-as it allows non-scoped ancilla allocation and de-allocation. Show
-examples.
+following analogy: allocating an ancilla by creating a new wire in the
+circuit is like borrowing some money; the computation has access to a
+new resource temporarily. De-allocating the ancilla is like returning
+the borrowed money; the computation no longer has access to the
+additional resource. It would be unreasonably restrictive to insist
+that if the person (function) borrowing the money must be the same
+person (function) who returns it, i.e., preventing the following
+computation:
+\begin{verbatim}
+  |-----\  /-----|
+         \/
+        / \
+  |----/   \-----|
+\end{verbatim}
+Indeed, as far as reversible or quantum computation is concerned, the
+circuit above is perfectly sensible since it conserves information
+(all money borrowed is repaid and the net balance is 0). In fact one
+can go further, if one person borrows \$10, why couldn't three
+different persons collectively collaborate to pay back the debt with
+one person paying \$5, another \$2, and a third \$3. This extra
+generality is not a gratuitous concern: since scope is fundamentally a
+static property of programs, it does not allow the flexibility of heap
+allocation in which the lifetime of resources is dynamically
+determined. To summarize, since the reversible-quantum computational
+framework guarantees that information is preserved, it also permits
+fascinating mechanisms in which computations can be sliced and
+diced, decomposed and recomposed, run forwards and backwards, in
+arbitrary ways as long as the net balance is 0.
 
@@
 to remember this information.
+
+Formalize ancilla problem in the context of Pi using eta and epsilon
+as done in PiFracDyn; more general than scoped; still runtime
+check.
+
+When ancilla bits are created they are created with a specific
+value. This value is encoded in the circuit. The value of fractional
+type represents the ``garbage collector.''  The garbage collector is
+specialized to collect a single value, which is the value used to
+create the ancilla value. At this point, if the garbage collector
+encounters a different value, it doesn't know what to do, and would
+need to throw an exception. This situation in \verb|PiFracDyn| is
+similar to current solutions (e.g. Quipper) but actually more general
+as it allows non-scoped ancilla allocation and de-allocation. Show
+examples.
+
+%%%%% 
+\subsection{Pointed Types}
 

### 2019-11-16 fracGC/p.tex
paper

@@
 The operator takes a block of gates parameterized by an ancilla,
-allocates a new ancilla initialized to $\ket{0}$, and runs the given
-block of gates. At the end of its execution, the block is expected to
-return the ancilla to the state $\ket{0}$ (this is the second
-subproblem addressed later) at which point it is de-allocated. In this
-scoped-by-construction approach, allocation and de-allocation of
-ancillae requires nothing beyond conventional parameter-passing
-techniques in which the parameter is allocated before entry to a
-function and de-allocated at exit.
+allocates a new ancilla of type \verb|Qubit| initialized to $\ket{0}$,
+and runs the given block of gates. At the end of its execution, the
+block is expected to return the ancilla to the state $\ket{0}$ (this
+is the second subproblem addressed later) at which point it is
+de-allocated. In this scoped-by-construction approach, allocation and
+de-allocation of ancillae requires nothing beyond conventional
+parameter-passing techniques in which the parameter is allocated
+before entry to a function and de-allocated at exit.
 
 This scoped model is therefore quite a pragmatic choice. It is however
-limited and to understand its limitations more vividly, we propose the
-following analogy: allocating an ancilla by creating a new wire in the
-circuit is like borrowing some money; the computation has access to a
-new resource temporarily. De-allocating the ancilla is like returning
-the borrowed money; the computation no longer has access to the
-additional resource. It would be unreasonably restrictive to insist
-that if the person (function) borrowing the money must be the same
-person (function) who returns it, i.e., preventing the following
-computation:
-\begin{verbatim}
-  |-----\  /-----|
-         \/
-        / \
-  |----/   \-----|
-\end{verbatim}
+limited, and to understand its limitations more vividly, we propose
+the following analogy: allocating an ancilla by creating a new wire in
+the circuit is like borrowing some money from a global external
+entity; the computation has access to a new resource
+temporarily. De-allocating the ancilla is like returning the borrowed
+money to the global entity; the computation no longer has access to
+the additional resource. It would however be unreasonably restrictive
+to insist that if the person (function) borrowing the money must be
+the same person (function) returning it.
+%% , i.e., preventing the following computation:
+%% \begin{verbatim}
+%%   |-----\  /-----|
+%%          \/
+%%         / \
+%%   |----/   \-----|
+%% \end{verbatim}
 Indeed, as far as reversible or quantum computation is concerned, the
-circuit above is perfectly sensible since it conserves information
-(all money borrowed is repaid and the net balance is 0). In fact one
-can go further, if one person borrows \$10, why couldn't three
-different persons collectively collaborate to pay back the debt with
-one person paying \$5, another \$2, and a third \$3. This extra
-generality is not a gratuitous concern: since scope is fundamentally a
-static property of programs, it does not allow the flexibility of heap
-allocation in which the lifetime of resources is dynamically
-determined. To summarize, since the reversible-quantum computational
-framework guarantees that information is preserved, it also permits
-fascinating mechanisms in which computations can be sliced and
-diced, decomposed and recomposed, run forwards and backwards, in
-arbitrary ways as long as the net balance is 0.
+only important invariant is that information is conserved, i.e., that
+money is conserved. The identities of bits or qubits are not
+observable as they are all interchangeable in the same way that
+particular bills with different serial numbers are interchangeable in
+financial transactions. Thus the only invariant is that the net flow
+of money between the computation and the global entity is zero. This
+observation allows us to go even further than just switching the
+identities of borrowers. It is even possible for one person to borrow
+\$10, and have three different persons collectively collaborate to
+pay back the debt with one person paying \$5, another \$2, and a third
+\$3.
+
+Computationally, this extra generality is not a gratuitous concern:
+since scope is fundamentally a static property of programs, it does
+not allow the flexibility of heap allocation in which the lifetime of
+resources is dynamically determined. To summarize, since the
+reversible-quantum computational framework guarantees that information
+is preserved, it also permits fascinating mechanisms in which
+computations can be sliced and diced, decomposed and recomposed, run
+forwards and backwards, in arbitrary ways as long as the net balance
+is 0.
 

### 2019-11-17 fracGC/p.tex
a bit more on paper
abstract

@@
 \newcommand{\alt}{~|~}
+\newcommand{\net}{t_\bullet}
 \newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
@@
 \newcommand{\ket}[1]{|#1\rangle}
+\newcommand{\oneover}[1]{1/#1}
 \newcommand{\pointed}[2]{[#1 \bullet #2]}
@@
 \begin{abstract}
-Text of abstract \ldots.
+  In reversible and quantum computing, the management of space is
+  subject to two broad classes of constraints. First, as is the case
+  for general-purpose computation, every allocation must be paired
+  with a matching de-allocation. Second, space can only be safely
+  de-allocated if its contents are restored to their initial value
+  from allocation time. Generally speaking, the state of the art
+  provides limited partial solutions that address the first
+  constraint by imposing a stack discipline and by leaving the second
+  constraint to programmer's assertions.
+
+  We propose a novel approach based on the idea of \emph{fractional
+    types}. As a simple intuitive example, allocation of a new boolean
+  value initialized to \textsf{false} also creates a value
+  $\oneover{\textsf{false}}$ that can be thought of as a garbage
+  collection (gc) process specialized to reclaim, and only reclaim,
+  storage containing the value $\textsf{false}$. This gc process is a
+  first-class entity that can sliced and diced by combining it with
+  other gc processes and decomposing it in smaller
+  processes.
+
+  Technically, we formalize this idea in the context of a reversible
+  language founded on type isomorphisms, prove its fundamental
+  correctness properties, and illustrate its expressiveness using a
+  wide variety of examples. The development is backed by a
+  fully-formalized Agda implementation.
 \end{abstract}
@@
 \end{quote}
+
+General allocation and de-allocation. No need to keep track of
+lifetime. Just make sure location is back to initial value. Safe to
+de-allocate because indistinguishable from freshly created. No pointer
+equality and aliases though. Concurrent: one thread allocates; gc
+thread meets someone else
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Value of Ancillae}
+\section{Interchangeable Ancillae}
 
@@
 
-Def. of universe is:
-
-\Udef{}
-
-New combinators:
-
-\EtaEpsilon{}
-
-and their evaluation:
+We add a new type $\oneover{t}$ with a trivial runtime representation
+$\ocircle$. We add two new level-1 programs to introduce and eliminate
+the new type, and one new level-2 program that states that
+introduction and elimination are inverses:
+\[\begin{array}{lrcl}
+\textit{Value types} & t &::=& \cdots \alt \oneover{t} \\
+\textit{Non-empty types} & \net \\
+\textit{Values}      & v &::=& \cdots \alt \ocircle \\
+\textit{Level-1 types} &&& t \leftrightarrow t \\
+\textit{Level-1 programs} & c &::=& \cdots \alt
+   \eta : 1 \leftrightarrow (\net * \oneover{\net}) \\
+   &&& ~~~~~\alt \epsilon : (\net * \oneover{\net}) \leftrightarrow 1 \\
+\textit{Level-2 types} &&& c \Leftrightarrow c \\
+\textit{Level-2 programs} & \alpha &::=& \cdots 
+  \alt unity : \epsilon \circ \eta \Leftrightarrow \idc 
+\end{array}\]
 

### 2019-11-17 fracGC/p.tex

@@
 
-\paragraph*{Quantum and Reversible Computing.} The Toffoli gate is
-known to be universal for classical reversible
-circuits~\cite{Toffoli:1980}. Adding just one gate (the Hadamard gate)
-produces a universal set of primitives for quantum
-circuits~\cite{hadtoffuniv}. The ``only'' difference between the two
-circuit models is that quantum circuits can process superpositions of
-values (waves) in one step whereas classical circuits lack this form
-of parallelism. Most importantly, structurally the two circuits models
-are identical and one can derive properties valid for both families by
-focusing on the simpler classical model. In fact, classical reversible
-computations are often used as ``subroutines'' of quantum
-computations.
+\paragraph*{Quantum and Reversible Computing Based on Type
+  Isomorphisms.} The Toffoli gate is known to be universal for
+classical reversible circuits~\cite{Toffoli:1980}. Adding just one
+gate (the Hadamard gate) produces a universal set of primitives for
+quantum circuits~\cite{hadtoffuniv}. The ``only'' difference between
+the two circuit models is that quantum circuits can process
+superpositions of values (waves) in one step whereas classical
+circuits lack this form of parallelism. Most importantly, structurally
+the two circuits models are identical and one can derive properties
+valid for both families by focusing on the simpler classical model. In
+fact, classical reversible computations are often used as
+``subroutines'' of quantum computations.
 
@@
 \emph{equivalences over
-  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
-exposes a rich mathematical structure in reversible computations
-that we will exploit to solve the ``ancilla problem'' explained next.
+  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective is
+similarly universal for reversible computation (and can be extended to
+quantum computation~\cite{10.1007/978-3-319-89366-2_19}) but with the
+advantage of exposing a rich mathematical structure in reversible
+computations that we will exploit to solve the ``ancilla problem''
+explained next.
 
@@
 temporary storage is allowed, the Toffoli gate is not
-universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as we
-demonstrate in Sec.~\ref{sec:examples}, the more temporary storage is
-allowed, the more efficient certain computations could become. The
-condition requiring that the temporary storage is only de-allocated
-when returned to its initial condition is a safety
-condition. Violating this condition destroys the symmetry between
-input and output making the circuits not reversible and, in the
-quantum model, causes decoherence that may destroy the quantum
-state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
-number of critical applications and yet are poorly supported in
+universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as would be
+expected the more temporary storage is allowed, the more efficient
+certain computations could become. The condition requiring that the
+temporary storage is only de-allocated when returned to its initial
+condition is a safety condition. Violating this condition destroys the
+symmetry between input and output making the circuits not reversible
+and, in the quantum model, causes decoherence that may destroy the
+quantum state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits
+have a number of critical applications and yet are poorly supported in
 current reversible and quantum programming languages making them a
@@
 
-Uses of ancillas:
+Uses of ancillae:
 \url{https://quantum.country/search}:
@@
   qubit is now perfectly entangled with the first qubit. Adding more
-  ancillas increases the distance of the repetition code.  \end{quote}
+  ancillae increases the distance of the repetition code.  \end{quote}
 
@@
 enough. Reversible/Quantum Circuits and Ancilla Wires: Early use of
-ancillas in Toffoli's paper to implement arbitrary reversible
+ancillae in Toffoli's paper to implement arbitrary reversible
 functions using a fixed number of 3 input gates

### 2019-11-17 fracGC/p.tex
intro

@@
 
-\paragraph*{Conservation of Information.} If quantum field theory is
-correct (as it so far seems to be) then information, during any
-physical process, is neither created nor
+If quantum field theory is correct (as it so far seems to be) then
+information, during any physical process, is neither created nor
 destroyed. Landauer~\cite{Landauer:1961,Landauer,bennett1985fundamental},
@@
 
-Instead of using the Toffoli gate as a universal primitive for
-reversible classical circuits, one can leverage the full force of type
-theory and category theory by expressing reversible classical
-computations as \emph{isomorphisms over finite
-  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
-\emph{equivalences over
+Instead of using the rather low-level circuit model of computation,
+one can leverage the full force of type theory and category theory by
+expressing reversible classical computations as \emph{isomorphisms
+  over finite types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667}
+or \emph{equivalences over
   groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective is
@@
 the Toffoli gate for classical reversible computing and of the
-universality of the combination of the Toffoli and Hadamard gates for
-quantum computing should not distract from efficiency and safety
-concerns. The theorems proving universality (i) assume that temporary
-storage (often called \emph{ancilla bits}) may be
-used~\cite{Toffoli:1980}, and (ii) that this temporary storage is
-returned to its initial state before de-allocation. Indeed if no
-temporary storage is allowed, the Toffoli gate is not
-universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as would be
-expected the more temporary storage is allowed, the more efficient
-certain computations could become. The condition requiring that the
-temporary storage is only de-allocated when returned to its initial
-condition is a safety condition. Violating this condition destroys the
-symmetry between input and output making the circuits not reversible
-and, in the quantum model, causes decoherence that may destroy the
-quantum state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits
-have a number of critical applications and yet are poorly supported in
+universality of the Toffol-Hadamard gates for quantum computing should
+not distract from efficiency and safety concerns. The theorems proving
+universality (i) assume that temporary storage (often called
+\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
+this temporary storage is returned to its initial state before
+de-allocation. Indeed if no temporary storage is allowed, the Toffoli
+gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
+would be expected the more temporary storage is allowed, the more
+efficient certain computations could become.
+
+More fundamentally, the condition requiring that the temporary storage
+is only de-allocated when returned to its initial condition is a
+safety condition. Violating this condition destroys the symmetry
+between input and output making the circuits not reversible and, in
+the quantum model, causes decoherence that may destroy the quantum
+state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
+number of critical applications and yet are poorly supported in
 current reversible and quantum programming languages making them a
@@
 
-\paragraph*{Negative Entropy.}  
+\paragraph*{Conservation of Information and Negative Entropy.}  
 According to the conventional theory of
@@
 processes'' that reclaim temporary storage. Just like in the case of
-``negative money'' It is critical that these fractional types be
+``negative money'' it is critical that these fractional types be
 linearly used, i.e., they cannot be duplicated or erased: this
@@
 
-\paragraph*{Contributions.} The paper presents a novel concept, that
-is \emph{fractional types} etc.
-
-\paragraph*{Outline.}
+\paragraph*{Outline.} The paper solves the ancilla problem in
+reversible and quantum computation using a novel concept:
+\emph{fractional types}. In the next section, we introduce the problem
+of ancilla, motivate its importance, and explain the limitations of
+current solutions. In Sec.~\ref{sec:pi}, we review previous work by
+James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
+introduces a reversible programming language built using isomorphism
+of types. Although it is conceivable that the concept of fractional
+types would apply to general-purpose languages (after some
+adaptation), its natural technical definition is using symmetries
+present in categorical model of type isomorphisms. We present such a
+definition in Sec.~\ref{sec:dyn}: that definition captures the main
+ideas of fractional types and partially solves a more general and more
+expressive instance of the ancilla problem. The approach however
+shares the limitation of current solutions in that it relies on
+runtime checks for programmer's assertions that values have been
+restored to their initial value before de-allocation. To remedy this
+limitation, we refine the type system in Sec.~\ref{sec:dep} to keep
+track of values in focus. Technically this requires the introduction
+of pointed types on top of which are defined both a monad and co-monad
+of singleton types. To justify the claim that fractional types solve
+the management of ancilla-based scratch space, we present in
+Sec.~\ref{sec:space} an operational semantics that exposes memory
+management details using a realization of fractional types as
+first-class dependently-typed garbage collectors. In Sec.~\ref{cat} we
+present a rich collection of examples illustrating the expressive
+power of fractional types. We conclude with a discussion of related
+work and a summary of our results.
 
@@
 \section{Preliminaries: Pi}
+\label{sec:pi}
 
@@
 \section{Interchangeable Ancillae}
+\label{sec:dyn}
 
@@
 \section{Dependently-Typed Garbage Collectors}
+\label{sec:dep}
 
@@
 \section{Space Operational Semantics}
+\label{sec:space}
 
@@
 \section{Trace}
+\label{sec:cat}
 
 Add multiplicative trace, do SAT solver
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Related Work and Conclusion}
 

### 2019-11-17 fracGC/p.tex
paper

@@
 
+Restricting a reversible (classical or quantum) circuit to use no
+ancillae is like restricting a Turing machine to use no memory other
+than the $n$ bits used to represent the
+input~\cite{DBLP:conf/innovations/AaronsonGS17}. As such a restriction
+disallows countless computations for trivial reasons, reversible and
+quantum models of computation have, since their inception, included
+management for scratch storage in the form of ancilla
+bits~\cite{Toffoli:1980} with the fundamental restriction that such
+bits must be returned to their initial states before being safely
+reused or de-allocated.
+
 %%%
@@
 
+Beyond computability and efficiency reasons, ancillae have found a
+wide range of interesting applications.
+
  Uses of ancillae:
 \url{https://quantum.country/search}:
+
+\paragraph*{Reversible Branching Operations.}
 
@@
 \end{quote}
+
+\paragraph*{Error Correction.}
 
@@
 
+\paragraph*{Deferring Quantum Measurements.}
+
 \url{https://homepages.cwi.nl/~rdewolf/qcnotesv2.pdf}
@@
 \end{quote}
-
-General allocation and de-allocation. No need to keep track of
-lifetime. Just make sure location is back to initial value. Safe to
-de-allocate because indistinguishable from freshly created. No pointer
-equality and aliases though. Concurrent: one thread allocates; gc
-thread meets someone else
-
-%%%
-\subsection{Examples}
-
-Toffoli4 using 2 Toffoli3: core of proof of universality; simple
-enough. Reversible/Quantum Circuits and Ancilla Wires: Early use of
-ancillae in Toffoli's paper to implement arbitrary reversible
-functions using a fixed number of 3 input gates
-\url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}
-
-In-place matrix transpose: ease of programming, efficiency. Say we
-want to transpose a matrix. Wikipedia example
-(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
-\begin{verbatim}
-11 12 13 14 
-21 22 23 24 
-\end{verbatim}
-transposed to
-\begin{verbatim}
-11 21
-12 22
-13 23
-14 24
-\end{verbatim}
-In Pi, the input and output matrices are:
-\begin{verbatim}
-M = (11,21) , (12,22) , (13,23) , (14,24)
-
-trM = (11,12,13,14) , (21,22,23,23) 
-\end{verbatim}
-Say we are given a language like Pi that is sound and complete with
-respect to permutations on finite types, we would write the
-permutation like so.
-\begin{verbatim}
-WRITE PERMUTATION
-\end{verbatim}
-This code does not use additional space, i.e., it performs the matrix
-transpose in constant space, i.e., performs an in-place matrix
-transposition. It is well know that with additional space, one can
-write more efficient matrix transpositions (explain and citations).
-
-Example(s) from Quipper etc. with focus on safety condition. Quipper
-uses ancilla bits in several places; one use is to compile
-irreversible circuits to reversible ones; need ancilla bits; more
-generally several quantum algorithms need ancilla bits (see below); in
-quipper de-allocation left to programmer.
-
-Is the below a possible example?
-
-Say we already have a permutation $A \leftrightarrow B$
-we can implement a permutation $X \leftrightarrow Z$ 
-when there exists $Y$ such that $A = X * Y = Z * Y = B$:
-\[\begin{array}{rcl}
-X &\rightarrow& X * Y * 1/Y \\
-  &\rightarrow& A * 1/Y \\
-  &\rightarrow&  B * 1/Y \\
-  &\rightarrow&  Y * Z * 1/Y \\
-  &\rightarrow&  Z * Y * 1/Y \\
-  &\rightarrow&  Z
-\end{array}\]
-This basically says, in the language of Lens, that
-when $A$ is isomorphic to $X \times Y$ and
-$B$ is isomorphic to $Z \times Y$, then
-$X$ is isomorphic to $Y$.
-
-Another way to think of it: for all isomorphic
-$A$ and $B$, whenever they can be factored with a
-common complement ($Y$), then the ``other pieces''
-(here $X$ and $Y$) are automatically isomorphic.
 
@@
 
+%%%
+\subsection{A Small Example}
+
+some that ``wants'' ancilla; perhaps fig2b from Dyn. if we want to use
+gates with small number of inputs; need to break cccnot to smaller gates
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The management of ancilla bits can be teased into two relatively
+The management of ancilla data can be teased into two relatively
 distinct subproblems:
@@
 
+% of course need de-allocation; keep track of lifetime
+
+% easy: stack but not general
+
+% for general; programmer control or gc
+
+% context of reversible; similar; need allocation; same tradeoffs for
+% de-allocation; stack or programmer or gc BUT with caveat. value must
+% be restored to its initial value; no information leakage which would
+% destroy quantum state
+
+% but reversible also gives you more guarantees; no creation or deletion
+% of information means no need to keep track of lifetime of individual
+% locations; just need to keep track of ``information''
+
+% General allocation and de-allocation. No need to keep track of
+% lifetime. Just make sure location is back to initial value. Safe to
+% de-allocate because indistinguishable from freshly created. No pointer
+% equality and aliases though. Concurrent: one thread allocates; gc
+% thread meets someone else
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-When ancilla bits are created they are created with a specific
+When ancilla are created they are created with a specific
 value. This value is encoded in the circuit. The value of fractional
@@
 
+Toffoli4 using 2 Toffoli3: core of proof of universality; simple
+enough. Reversible/Quantum Circuits and Ancilla Wires: Early use of
+ancillae in Toffoli's paper to implement arbitrary reversible
+functions using a fixed number of 3 input gates
+\url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}
+
+In-place matrix transpose: ease of programming, efficiency. Say we
+want to transpose a matrix. Wikipedia example
+(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
+\begin{verbatim}
+11 12 13 14 
+21 22 23 24 
+\end{verbatim}
+transposed to
+\begin{verbatim}
+11 21
+12 22
+13 23
+14 24
+\end{verbatim}
+In Pi, the input and output matrices are:
+\begin{verbatim}
+M = (11,21) , (12,22) , (13,23) , (14,24)
+
+trM = (11,12,13,14) , (21,22,23,23) 
+\end{verbatim}
+Say we are given a language like Pi that is sound and complete with
+respect to permutations on finite types, we would write the
+permutation like so.
+\begin{verbatim}
+WRITE PERMUTATION
+\end{verbatim}
+This code does not use additional space, i.e., it performs the matrix
+transpose in constant space, i.e., performs an in-place matrix
+transposition. It is well know that with additional space, one can
+write more efficient matrix transpositions (explain and citations).
+
+Example(s) from Quipper etc. with focus on safety condition. Quipper
+uses ancilla bits in several places; one use is to compile
+irreversible circuits to reversible ones; need ancilla bits; more
+generally several quantum algorithms need ancilla bits (see below); in
+quipper de-allocation left to programmer.
+
+Is the below a possible example?
+
+Say we already have a permutation $A \leftrightarrow B$
+we can implement a permutation $X \leftrightarrow Z$ 
+when there exists $Y$ such that $A = X * Y = Z * Y = B$:
+\[\begin{array}{rcl}
+X &\rightarrow& X * Y * 1/Y \\
+  &\rightarrow& A * 1/Y \\
+  &\rightarrow&  B * 1/Y \\
+  &\rightarrow&  Y * Z * 1/Y \\
+  &\rightarrow&  Z * Y * 1/Y \\
+  &\rightarrow&  Z
+\end{array}\]
+This basically says, in the language of Lens, that
+when $A$ is isomorphic to $X \times Y$ and
+$B$ is isomorphic to $Z \times Y$, then
+$X$ is isomorphic to $Y$.
+
+Another way to think of it: for all isomorphic
+$A$ and $B$, whenever they can be factored with a
+common complement ($Y$), then the ``other pieces''
+(here $X$ and $Y$) are automatically isomorphic.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-11-18 fracGC/p.tex
sec 2 kind of

@@
 \newcommand{\net}{t_\bullet}
+\newcommand{\opr}[1]{|#1\rangle\langle#1|}
 \newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
@@
 
- Uses of ancillae:
-\url{https://quantum.country/search}:
+\paragraph*{Quantum blackboxes and Phase Kickback Trick.} Consider a
+small database with four elements $a$, $b$, $c$, and $d$. We are given
+a function $f$ that maps every element to $0$ except for one element
+of interest, e.g., $f(a)=1$ and $f(b)=f(c)=f(d)=0$. In the worst case,
+finding the element of interest might require applying $f$ to every
+element of the database. In the quantum world we can find the element
+much more efficiently using Grover's
+algorithm~\cite{Grover:1996:FQM:237814.237866}. We start by building
+an equally weighted superposition of the elements:
+$\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}$
+and operate concurrently of the superposition as follows:
 
-\paragraph*{Reversible Branching Operations.}
+\[\begin{array}{l}
+                    \frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
+\\
+\Rightarrow  \textrm{phase kickback} \\
+\\
+-\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
+\\
+\Rightarrow  \textrm{diffusion} \\
+\\
+ (\frac{1}{2}-(-\frac{1}{2}))\ket{a}+(\frac{1}{2}-\frac{1}{2})\ket{b}+(\frac{1}{2}-\frac{1}{2})\ket{c}+(\frac{1}{2}-\frac{1}{2})\ket{d})\\
+\\
+\Rightarrow  \textrm{simplification} \\
+\\
+\ket{a}
+\end{array}\]
 
+The question, which arises repeatedly in quantum computation, is how
+to implement operations such as phase kickback. Essentially we want to
+concurrently perform the following action on every element of the
+superposition:
+\[\begin{array}{l}
+  \textbf{if}~\textit{current-element}=\ket{a} \\
+  \textbf{then}~\textit{negate} \\
+  \textbf{else}~\textit{identity}
+\end{array}\]
+In the quantum world, na\"\i vely attempting to test whether the
+current element is equal to $\ket{a}$ by performing a measurement would
+destroy the quantum superposition defeating the entire algorithm.
+
+Here is where ancilla qubits come to the rescue. In the words of
+Matuschak and Nielsen~\cite{howgrover}, the idea is the following:
 \begin{quote}
-  There’s a rough heuristic worth noting here, which is that you can
-  often convert if-then style thinking into quantum circuits. You
-  introduce an ancilla qubit to store the outcome of evaluating the if
-  condition. And then depending on the state of the ancilla, you
-  perform the appropriate state manipulation. Finally, when possible
-  you reverse the initial computation, resetting the ancilla to its
-  original state so you can subsequently ignore it.
+  There's a rough heuristic worth noting here, which is that you can
+  often convert \verb|if-then| style of thinking into quantum
+  circuits. You introduce an ancilla qubit to store the outcome of
+  evaluating the \verb|if| condition. And then depending on the state
+  of the ancilla, you perform the appropriate state
+  manipulation. Finally, when possible you reverse the initial
+  computation, resetting the ancilla to its original state so you can
+  subsequently ignore it.
 \end{quote}
 
-\paragraph*{Error Correction.}
-
-\url{https://www.nap.edu/read/25196/chapter/5#73}
+\paragraph*{Quantum Error Correction.} The main requirement for
+quantum error correction is the ability to replicate the state of a
+qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is however
+impossible to do directly by the no-cloning theorem. Again ancilla
+qubits come to the rescue: because ancilla qubits start in a known
+initial state, it is possible to create a simple circuit that makes
+their output state match a protected qubit without knowing or
+disturbing the protected qubit. Chong et al.~\cite{sigarchblog}
+illustrate the idea with a simple example:
 \begin{quote}
-  For error correction, one needs to replicate the state of a qubit
-  onto a number of qubits. While the no cloning theorem prevents one
-  from copying the state of a qubit directly onto another, one can
-  create a redundant entangled qubit state of many qubits. The key is
-  that the qubits to be entangled must start out in a known
-  state. Qubits with a known state (for purposes of this discussion,
-  it will be the state $\ket{0}$), called ``ancilla qubits,'' may be
-  added to a computation for this purpose. Since the state of ancilla
-  qubits are known, it is possible to create a simple circuit that
-  makes the output state of all these ancilla qubits match the
-  protected qubit: run each ancilla through a controlled-NOT gate,
-  where the control is driven by the qubit that needs to be
-  replicated. Assume that there is a qubit with state $\Psi$ that we
-  want to protect, where $\ket{\Psi}$ represents an arbitrary
-  superposition state $\ket{\Psi} = a_0 \ket{0} + a_1 \ket{1}$. In the
-  CNOT gate, the ancilla $\ket{0}$ state will remain a $\ket{0}$ state
-  by the $\ket{0}$ component of $\ket{\Psi}$, but it will be converted
-  to $\ket{1}$ by the $\ket{1}$ component of $\ket{\Psi}$. The result
-  of this operation is the newly entangled two-qubit state
-  $a_0 \ket{0} + a_1 \ket{1}$, creating a system in which the ancilla
-  qubit is now perfectly entangled with the first qubit. Adding more
-  ancillae increases the distance of the repetition code.  \end{quote}
-
-\url{https://www.sigarch.org/the-case-for-quantum-computing/}
-\begin{quote}
-  Imagine a 3-qubit quantum majority code in which a logical “0” is
-  encoded as “000” and a logical “1” is encoded as “111.”  Just as
-  with a classical majority code, a single bit-flip error can be
-  corrected by restoring to the majority value.  Unlike a classical
-  code, however, we can not directly measure the qubits else their
-  quantum state will be destroyed.  Instead, we measure syndromes from
-  each possible pair of qubits by interacting them with an ancilla,
-  then measure each ancilla.  Although the errors to the qubits are
-  actually continuous, the effect of measuring the ancilla is to
-  discretize the errors, as well as inform us whether an error
-  occurred so that it can be corrected.  With this methodology,
-  quantum states are restored in a modular way for even a large
-  quantum computer.  Furthermore, operations on error-corrected qubits
-  can be viewed as digital rather than analog, and only a small number
-  of universal operations (H, T, CNOT) are needed for universal
-  quantum computation.  Through careful design and engineering, error
-  correction codes and this small set of precise operations will lead
-  to machines that could support practical quantum computation.
+consider a 3-qubit quantum majority code in which a logical ``0'' is
+encoded as ``000'' and a logical ``1'' is encoded as ``111.''  Just as with
+a classical majority code, a single bit-flip error can be corrected by
+restoring to the majority value.  Unlike a classical code, however, we
+can not directly measure the qubits else their quantum state will be
+destroyed.  Instead, we measure syndromes from each possible pair of
+qubits by interacting them with an ancilla, then measure each ancilla.
+Although the errors to the qubits are actually continuous, the effect
+of measuring the ancilla is to discretize the errors, as well as
+inform us whether an error occurred so that it can be corrected.  
 \end{quote}
 
-\paragraph*{Deferring Quantum Measurements.}
-
-\url{https://homepages.cwi.nl/~rdewolf/qcnotesv2.pdf}
-\begin{quote}
-  Using an ancilla qubit, it is possible to avoid doing any
-  intermediate measurements in a quantum circuit. Show how this can be
-  done.  Hint: instead of measuring the qubit, apply a CNOT that
-  “copies” it to a new $\ket{0}$-qubit, which is then left alone until
-  the end of the computation. Analyze what happens.
-\end{quote}
+\paragraph*{Deferring Quantum Measurements.} The evolution of a
+quantum system is deterministic. Measurement add significant
+complications, collapsing the quantum state, and producing
+probabilistic results. However, using ancilla qubits it is possible to
+avoid doing any intermediate measurements in a quantum
+circuit~\cite{dewolf}.
 
 %%%
-\subsection{Language Support}
-
-current proposals for reversible
-and quantum programming languages
-(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
-Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
-to be used in nested scopes, have sound but incomplete laws for
-reasoning about the safety condition, and only guarantee safety via
-runtime checks.
-
-Leads to bugs as analyzed by
-\url{http://drops.dagstuhl.de/opus/volltexte/2019/10196/pdf/OASIcs-PLATEAU-2018-4.pdf}:
-
+\subsection{Ancillae Bugs in Benchmarks.}
+Despite its fundamental importance, the management of ancillae is
+poorly supported in current languages, simulators, and toolkits as
+demonstrated in a recent analysis of bugs in quantum
+benchmarks~\cite{DBLP:conf/oopsla/HuangM18}. The analysis reveals that
+lack of language support in the management of ancillae leads to
+several problems. For example: 
 \begin{quote}
-
-  Bug type 6: Incorrect composition of operations using mirroring Section 4.5 discussed how bugs in deallocating ancillary qubits can happen due to bad parameters.
-Here we see how bugs in deallocating ancillary qubits can happen due to incorrect composition of operations following a mirroring pattern.
-For example, in Table 7, the operations in rows 2 and 3 are respectively mirrored and undone in rows 6 and 5.
-These lines of code need careful reversal of every loop and every operation.
-
-  A common pattern in quantum programs involves performing operations (e.g., add), contingent on a set of qubits known as control qubits.
-Without language support, this pattern needs many lines of code and manual allocation of ancillary qubits.
-In the Scaffold code example in Table 7, rows 3 and 5 are just computing the intersection of qubits q, with the help of ancillary qubits initialized in row 1, in order to realize the controlled rotation operation in row 4.
-Furthermore, quantum algorithms often need varying numbers of control qubits in different parts of the algorithm, leading to replicated code from multiple versions of the same subroutine differing only by the number of control qubits.
-
+  \textbf{Bug type 6: Incorrect composition of operations using mirroring}
+  Section~4.5 discussed how bugs in deallocating ancillary qubits can
+  happen due to bad parameters. Here we see how bugs in deallocating
+  ancillary qubits can happen due to incorrect composition of
+  operations following a mirroring pattern. For example, in Table~7,
+  the operations in rows 2 and 3 are respectively mirrored and undone
+  in rows~6 and~5. These lines of code need careful reversal of every
+  loop and every operation.
 \end{quote}
@@
 
+current proposals for reversible
+and quantum programming languages
+(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
+Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
+to be used in nested scopes, have sound but incomplete laws for
+reasoning about the safety condition, and only guarantee safety via
+runtime checks.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-11-18 fracGC/p.tex
starting sec 3

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Preliminaries: Pi}
+\section{Preliminaries: $\Pi$}
 \label{sec:pi}
 
-Our approach to reversible quantum programming based on type
-isomorphisms. Examples.
-
-Ancilla problem in the context of Pi using eta and epsilon 
-informally.
+The step from reversible classical circuits to quantum circuits
+requires the addition of just one gate that creates
+superpositions. Similarly the step from a reversible classical
+programming language to a quantum one requires just the management of
+superpositions~\cite{10.1007/978-3-319-89366-2_19}. In the cited
+paper, the core reversible classical language was Theseus~\cite{XXX},
+a language that adds convenient syntactic extensions over an more
+basic language $\Pi$ of combinators witnessing isomorphisms over
+finite types~\cite{XXX}. The combinator-based language has the
+advantage of being convenient for formal analysis for at least two
+reasons: (i) it is conceptually simple and small, and (ii) it has
+direct and evident connections to type theory and category
+theory. Indeed our solution for managing ancillae is inspired by the
+construction of \emph{compact closed categories}~\cite{XXX}. These
+categories extend the monoidal categories~\cite{XXX} which are to
+model most models of resource-aware programming languages~\cite{XXX}
+with a new type constructor that creates duals or inverses to existing
+types. This will be our fractional type.
 
 %%%%%
-\subsection{Pi}
+\subsection{Core $\Pi$}
 

### 2019-11-18 fracGC/p.tex
more on sec 3

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Preliminaries: $\Pi$}
-\label{sec:pi}
-
-The step from reversible classical circuits to quantum circuits
-requires the addition of just one gate that creates
-superpositions. Similarly the step from a reversible classical
-programming language to a quantum one requires just the management of
-superpositions~\cite{10.1007/978-3-319-89366-2_19}. In the cited
-paper, the core reversible classical language was Theseus~\cite{XXX},
-a language that adds convenient syntactic extensions over an more
-basic language $\Pi$ of combinators witnessing isomorphisms over
-finite types~\cite{XXX}. The combinator-based language has the
-advantage of being convenient for formal analysis for at least two
-reasons: (i) it is conceptually simple and small, and (ii) it has
-direct and evident connections to type theory and category
-theory. Indeed our solution for managing ancillae is inspired by the
-construction of \emph{compact closed categories}~\cite{XXX}. These
-categories extend the monoidal categories~\cite{XXX} which are to
-model most models of resource-aware programming languages~\cite{XXX}
-with a new type constructor that creates duals or inverses to existing
-types. This will be our fractional type.
-
-%%%%%
-\subsection{Core $\Pi$}
-
-Plain Pi over finite types:
-\[\begin{array}{lrcl}
-\textit{Value types} & t &::=& 0 \alt 1 \alt t+t \alt t*t \\
-\textit{Values}      & v &::=& \star \alt \inlv{v} \alt \inrv{v} \alt (v,v) \\
-\textit{Level-1 types} &&& t \leftrightarrow t \\
-\textit{Level-1 programs} & c &::=& \cdots \\
-\textit{Level-2 types} &&& c \Leftrightarrow c \\
-\textit{Level-2 programs} & \alpha &::=& \cdots 
-\end{array}\]
-
-Focusing on finite types, the building blocks of type theory are: the
-empty type ($0$), the unit type ($1$), the sum type ($+$), and the
-product ($\times$) type. Before getting into the formal theory, let's
-consider the intuition that types are modeled by (topological) spaces
-and that type equivalence is modeled by possible \emph{deformations}
-of such spaces. Consider the space $(1 + 0) \times (1 + 1)$. This
-space is the product of two subspaces: the subspace $(1 + 0)$ which
-itself is the sum of the space~$1$ containing one point $\texttt{tt}$
-and the empty space~$0$ and the subspace $(1 + 1)$ which is the sum of
-two spaces each containing the one point $\texttt{tt}$. Any
-deformation of this space must at least preserve the number of points:
-we can neither create nor destroy points during any continuous
-deformation. Seeing that the number of points in our example space is
-2, a reasonable hypothesis is that we can deform the space above to
-any other space with 2 points such as $1 + 1$ or $1 + (1 + 0)$. What
-this really means is that we are treating the sum and product
-structure as malleable. Imagining a product structure as arranged in a
-grid, by stretching we can turn this structure to a sum structure
-arranged in a line, change the orientation of the grid by exchanging
-the axes, as well as other transformations that preserve the number of
-points using various ``symmetries.'' These symmetries capture, in a
-stylized manner, a rich collection of space-time
-tradeoffs~\cite{superstructural}. We formalize this intuition by
-saying that types form a \emph{commutative semiring} (up to type
-isomorphism).
-
 \begin{figure*}[t]
@@
 
-Another model of types is to view each type $A$ as a collection of
+\section{Preliminaries: $\Pi$}
+\label{sec:pi}
+
+The step from reversible classical circuits to quantum circuits
+requires the addition of just one gate that creates
+superpositions. Similarly the step from a reversible classical
+programming language to a quantum one requires just the management of
+superpositions~\cite{10.1007/978-3-319-89366-2_19}. In the cited
+paper, the core reversible classical language was Theseus~\cite{XXX},
+a language that adds convenient syntactic extensions over an more
+basic language $\Pi$ of combinators witnessing isomorphisms over
+finite types~\cite{XXX}. The combinator-based language has the
+advantage of being convenient for formal analysis for at least two
+reasons: (i) it is conceptually simple and small, and (ii) it has
+direct and evident connections to type theory and category
+theory. Indeed our solution for managing ancillae is inspired by the
+construction of \emph{compact closed categories}~\cite{XXX}. These
+categories extend the monoidal categories~\cite{XXX} which are to
+model most models of resource-aware programming languages~\cite{XXX}
+with a new type constructor that creates duals or inverses to existing
+types. This will be our fractional type.
+
+%%%%%
+\subsection{Core $\Pi$}
+
+The syntax of the language consists of several categories:
+\[\begin{array}{lrcl}
+\textit{Value types} & t &::=& 0 \alt 1 \alt t+t \alt t*t \\
+\textit{Values}      & v &::=& \star \alt \inlv{v} \alt \inrv{v} \alt (v,v) \\
+\textit{Level-1 types} &&& t \leftrightarrow t \\
+\textit{Level-1 programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})\\
+\textit{Level-2 types} &&& c \Leftrightarrow c \\
+\textit{Level-2 programs} & \alpha &::=& (\textrm{Omitted})
+\end{array}\]
+
+Focusing on finite types, the building blocks of type theory are: the
+empty type ($0$), the unit type ($1$), the sum type ($+$), and the
+product ($\times$) type. One may view each type $A$ as a collection of
 physical wires that can transmit $\|A\|$ distinct values where $\|A\|$
-is the size of a type, computed as: $\| 0 \| = 0$; $\| 1 \| = 1$; $\|
-A + B \| = \| A \| + \| B \|$; and $\| A \times B \| = \| A \| * \| B
-\|$. Thus the type $\mathbb{B} = 1 + 1$ corresponds to a wire that can
-transmit two values, i.e., bits, and the type $\mathbb{B} \times
-\mathbb{B} \times \mathbb{B}$ corresponds to a collection of wires
-that can transmit three bits. From that perspective, a type
-isomorphism between types $A$ and $B$ (such that $\|A\|=\|B\|=n$)
-models a \emph{reversible} combinational circuit that \emph{permutes}
-the $n$ different values. These type isomorphisms are collected in
-Fig.~\ref{pi-terms}. We call the resulting language $\Pi$.  It is
-known that these type isomorphisms are sound and complete for all
-permutations on finite types~\cite{Fiore:2004,fiore-remarks} and hence
-that they are \emph{complete} for expressing combinational
-circuits~\cite{fredkin1982conservative,James:2012:IE:2103656.2103667,Toffoli:1980}.
+is the size of a type, computed as: $\| 0 \| = 0$; $\| 1 \| = 1$;
+$\| A + B \| = \| A \| + \| B \|$; and
+$\| A \times B \| = \| A \| * \| B \|$. Thus the type
+$\mathbb{B} = 1 + 1$ corresponds to a wire that can transmit two
+values, i.e., bits, and the type
+$\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a
+collection of wires that can transmit three bits. From that
+perspective, a type isomorphism between types $A$ and $B$ (such that
+$\|A\|=\|B\|=n$) models a \emph{reversible} combinational circuit that
+\emph{permutes} the $n$ different values. These type isomorphisms are
+collected in Fig.~\ref{pi-terms}. It is known that these type
+isomorphisms are sound and complete for all permutations on finite
+types~\cite{Fiore:2004,fiore-remarks} and hence that they are
+\emph{complete} for expressing combinational
+circuits~\cite{fredkin1982conservative,James:2012:IE:2103656.2103667,Toffoli:1980}. Algebraically,
+these types and combinators form a \emph{commutative semiring} (up to
+type isomorphism). Logically they form a superstructural logic
+capturing space-time tradeoffs~\cite{superstructural}. Categorically,
+they form a \emph{bimonoidal category}~\cite{laplaza72}.
 
@@
 Fig.~\ref{pi-terms} can themselves be subject to (higher-level)
-reversible deformations? Before developing the theory, let's consider
-a small example consisting of two deformations between the types $A +
-B$ and $C+D$:
+reversible deformations? The answer is
+yes~\cite{Miller:2003:TBA:775832.775915,DBLP:conf/esop/CaretteS16,DBLP:journals/corr/CockettCS17}. We
+only give one example below. (We will require a new level-2 program to
+witness coherence of fractional types in Sec.~\ref{sec:dep}.)
 
@@
 \noindent The top path is the $\Pi$ program
-$(c_1~\oplus~c_2)~\odot~\swapp$ which deforms the type $A$ by $c_1$,
-deforms the type $B$ by $c_2$, and deforms the resulting space by a
+$(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$,
+acts on the type $B$ by $c_2$, and deforms the resulting space by a
 twist that exchanges the two injections into the sum type. The bottom
-path performs the twist first and then deforms the type $A$ by $c_1$
-and the type $B$ by $c_2$ as before. If one could imagine the paths
-  as physical wires and the deformations $c_1$ and $c_2$ as arbitrary
-deformations on these wires then, holding the points $A$, $B$, $C$,
-and $D$ fixed, it is possible to rotate the top part of the diagram to
-become identical to the bottom one. That rotation can be undone
-(reversed), which takes the bottom part of the diagram into the top
-part.  In other words, there exists a deformation of the program
+path performs the twist first and then acts on the type $A$ by $c_1$
+and on the type $B$ by $c_2$ as before. If one could imagine the paths
+as physical wires and the actions $c_1$ and $c_2$ as permutations on
+these wires then, holding the points $A$, $B$, $C$, and $D$ fixed, it
+is possible to rotate the top part of the diagram to become identical
+to the bottom one. That rotation can be undone (reversed), which takes
+the bottom part of the diagram into the top part.  In other words,
+there exists a continuous deformation of the program
 $(c_1~\oplus~c_2)~\odot~\swapp$ to the program
@@
 $\swapp$.
-
-Developing a collection of ``templates'' for such higher-level
-equivalences is reasonably straightforward (see for example the work
-of Miller et al~\cite{Miller:2003:TBA:775832.775915}). To find a
-\emph{complete} set of equivalences is much more difficult. The idea
-developed in our previous work~\cite{DBLP:conf/esop/CaretteS16} relies
-on ``categorification'' and a classical result by
-Laplaza~\cite{laplaza}. We refer the interested reader to our previous
-work for the full set of \emph{complete} ``templates.'' A related
-idea~\cite{DBLP:journals/corr/CockettCS17} for a sound and complete
-set of equivalences on control-not gates is also developed in the
-context of the same monoidal categories as the ones we use suggesting
-that these categorical structures provide a common framework for
-reasoning about various classes of reversible circuits.
 

### 2019-11-18 fracGC/p.tex
paper

@@
 property is however automatically enforced in a reversible
-computational model!
+computational model.
 
@@
 management details using a realization of fractional types as
-first-class dependently-typed garbage collectors. In Sec.~\ref{cat} we
+first-class dependently-typed garbage collectors. In Sec.~\ref{sec:cat} we
 present a rich collection of examples illustrating the expressive

### 2019-11-18 fracGC/p.tex
paper

@@
 qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is however
-impossible to do directly by the no-cloning theorem. Again ancilla
+impossible to do directly by the no-cloning theorem~\cite{XXX}. Again ancilla
 qubits come to the rescue: because ancilla qubits start in a known
@@
 programming language to a quantum one requires just the management of
-superpositions~\cite{10.1007/978-3-319-89366-2_19}. In the cited
-paper, the core reversible classical language was Theseus~\cite{XXX},
-a language that adds convenient syntactic extensions over an more
-basic language $\Pi$ of combinators witnessing isomorphisms over
-finite types~\cite{XXX}. The combinator-based language has the
-advantage of being convenient for formal analysis for at least two
-reasons: (i) it is conceptually simple and small, and (ii) it has
-direct and evident connections to type theory and category
-theory. Indeed our solution for managing ancillae is inspired by the
-construction of \emph{compact closed categories}~\cite{XXX}. These
-categories extend the monoidal categories~\cite{XXX} which are to
-model most models of resource-aware programming languages~\cite{XXX}
-with a new type constructor that creates duals or inverses to existing
-types. This will be our fractional type.
+superpositions, for example starting from the classical reversible
+language Theseus~\cite{XXX}, a small extension produces a quantum
+language with quantum control~\cite{10.1007/978-3-319-89366-2_19}. The
+language Theseus~\cite{XXX} mentioned above is a language that uses
+convenient syntactic extensions over an more basic language $\Pi$ of
+combinators witnessing isomorphisms over finite types~\cite{XXX}. The
+combinator-based language has the advantage of being more amenable to
+formal analysis for at least two reasons: (i) it is conceptually
+simple and small, and (ii) it has direct and evident connections to
+type theory and category theory. Indeed our solution for managing
+ancillae is inspired by the construction of \emph{compact closed
+  categories}~\cite{XXX}. These categories extend the monoidal
+categories~\cite{XXX} which are used to model many resource-aware
+programming languages~\cite{XXX} (including $\Pi$) with a new type
+constructor that creates duals or inverses to existing types. This
+dual will be our fractional type.
 
@@
 
+move from 1/t to 1/v (how the quantum search runs; need ancilla 
+initialized to a specific value)
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:dep}
+
+PiPointedFrac:
+lift pi
+lift multiplicative structure
+monad/comonad for singletons
+eta/epsilon
+
+rationale: we want singleton types to track, in the type system
+the values of ancillae, i.e.. we construct a proof using the type
+system. Technically the singleton types allow you to focus on one
+value by reflecting it in the type system. Now this type can only be
+equal to another type if and only if the other type is the same
+singleton value.
+
+Once singleton types are built, we need to make sure circuit
+evaluation is also reflected in the type system. So in fact
+computation itself is reflected in the type system. In order to mix
+singleton types with other types, we use pointed types as a unifying
+substrate. Now all types have values; we can focus (using monad) and
+unfocus using (comonad). (This by the way ensures we don't write 1/0
+which woudl be bad).
 
@@
 \label{sec:cat}
+
+allocate A
+allocate B
+combinate the gc's for A and B to get 1/(AxB)
+allocate AxB: split the gc in two 1/A and 1/B
+use the different gcs to satisfy each other
 

### 2019-11-19 fracGC/p.tex
basic idea of pointed types in J.agda for presentation

@@
 \newunicodechar{𝕌}{$\mathbb{U}$}
+\newunicodechar{𝔹}{$\mathbb{B}$}
 \newunicodechar{𝔽}{$\mathbb{F}$}
@@
 \newunicodechar{⊚}{$\fatsemi$}
+\newunicodechar{⟷}{$\leftrightarrow$}
 
@@
 \input{latex/PiFracDynCode.tex}
+\input{latex/BooleanCircuitsCode.tex}
 
@@
 of ancilla, motivate its importance, and explain the limitations of
-current solutions. In Sec.~\ref{sec:pi}, we review previous work by
+current solutions. In Sec.~\ref{sec:pi}, we review existing work by
 James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
@@
 management details using a realization of fractional types as
-first-class dependently-typed garbage collectors. In Sec.~\ref{sec:cat} we
-present a rich collection of examples illustrating the expressive
-power of fractional types. We conclude with a discussion of related
-work and a summary of our results.
+first-class dependently-typed garbage collectors. In
+Sec.~\ref{sec:cat} we present a rich collection of examples
+illustrating the expressive power of fractional types. We conclude
+with a discussion of related work and a summary of our results.
 
@@
 \begin{center}
-\begin{tikzpicture}[scale=0.4,every node/.style={scale=0.5}]
+\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
   \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
@@
 %%%
-\subsection{A Small Example}
+\subsection{Small Examples}
 
-some that ``wants'' ancilla; perhaps fig2b from Dyn. if we want to use
-gates with small number of inputs; need to break cccnot to smaller gates
+The code below defines the types corresponding to bits (booleans),
+two-bits, three-bits, and four-bits. It then defines an operator that
+builds a controlled version of a given combinator $c$. This controlled
+version takes an additional ``control'' bit and only applies $c$ if
+the control bit is true. The code then iterates the control operation
+several times starting starting from boolean negation.
+
+\Bexamples{}
+
+By construction, the \AgdaFunction{TOFFOLI} function takes three bits,
+and negates the third if the first two are both true. As
+Toffoli~\cite{Toffoli:1980} demonstrated, this function is universal
+for classical reversible circuits assuming it is possible to allocate
+and de-allocate ancilla bits. Indeed without using ancilla bits, it is
+not possible to use compositions of \AgdaFunction{TOFFOLI} functions
+to implement more general versions that are controlled by more than
+two bits (for example, \AgdaFunction{CTOFFOLI}). This will however be
+possible in the next section once we introduce ancilla bits.
 
@@
 the circuit is like borrowing some money from a global external
-entity; the computation has access to a new resource
+entity (the memory manager); the computation has access to a new resource
 temporarily. De-allocating the ancilla is like returning the borrowed
@@
 the additional resource. It would however be unreasonably restrictive
-to insist that if the person (function) borrowing the money must be
+to insist that the person (function) borrowing the money must be
 the same person (function) returning it.
@@
 is preserved, it also permits fascinating mechanisms in which
-computations can be sliced and diced, decomposed and recomposed, run
-forwards and backwards, in arbitrary ways as long as the net balance
-is 0.
+allocations and de-allocations can be sliced and diced, decomposed and
+recomposed, run forwards and backwards, in arbitrary ways as long as
+the net balance is 0.
 

### 2019-11-19 fracGC/p.tex
simple proof that 4 toffoli is not definable

@@
 de-allocation. Indeed if no temporary storage is allowed, the Toffoli
-gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
-would be expected the more temporary storage is allowed, the more
-efficient certain computations could become.
+gate is not
+universal~\cite{DBLP:conf/innovations/AaronsonGS17,DBLP:journals/corr/Xu15e}
+and as would be expected the more temporary storage is allowed, the
+more efficient certain computations could become.
 
@@
 to implement more general versions that are controlled by more than
-two bits (for example, \AgdaFunction{CTOFFOLI}). This will however be
-possible in the next section once we introduce ancilla bits.
+two bits (for example, \AgdaFunction{CTOFFOLI}). The proof is simple:
+the \AgdaFunction{TOFFOLI} function depends on two control bits and
+possibly negates a third; when used in a circuit with four bits, the
+unused bit can be either state, which means that
+\AgdaFunction{TOFFOLI} generates an even permutation. The function
+\AgdaFunction{CTOFFOLI} is however an odd permutation which is
+unreachable from even permutations. This function will however be
+definable in the next section once we introduce ancilla bits.
 

### 2019-11-19 fracGC/p.tex
more on sec 3

@@
 \newunicodechar{⋆}{${}_*$}
+\newunicodechar{○}{$\gcv$}
 \newunicodechar{ᵤ}{${}_u$}
@@
 \newcommand{\net}{t_\bullet}
+\newcommand{\gcv}{\circlearrowright}
 \newcommand{\opr}[1]{|#1\rangle\langle#1|}
@@
 \subsection{Core $\Pi$}
+\label{sub:core}
 
@@
 %%%%%
-\subsection{Adding Fractions}
+\subsection{GC Value of Fractional Type}
 
-We add a new type $\oneover{t}$ with a trivial runtime representation
-$\ocircle$. We add two new level-1 programs to introduce and eliminate
-the new type, and one new level-2 program that states that
-introduction and elimination are inverses:
+Instead of restricting ancillae to be scoped, we separate allocation
+and de-allocation into two dual actions $\eta$ and $\epsilon$ whose
+types are:
+\[
+\eta_A : 1 \leftrightarrow A * \oneover{A} : \epsilon_A
+\]
+The names and types of these operations are inspired by compact closed
+categories which are extensions of the monoidal categories that model
+$\Pi$. Intuitively, $\eta$ allows one from ``no information'' to
+create a pair of a value of type $A$ and a value of type
+$\oneover{A}$. We interpret the latter value as a gc process
+specialized to collect the created value. Dually, $\epsilon$ applies
+the gc process to the appropriate value annihilating
+both.\footnote{Another interesting interpretation is that these
+  operations correspond to creation and annihilation of entangled
+  particle/antiparticle pairs in quantum
+  physics~\cite{Panangaden2011}.}
+
+To make this idea work, several technical details are needed. Most
+notably, we must exclude the empty types from this creation and
+annihilation process. Otherwise, we would be able to prove that:
+\[\begin{array}{rcll}
+1 &=& 0 * 1/0 & \textrm{by~} \eta_0 \\
+&=& 0 & \textrm{by~} \absorbr
+\end{array}\]
+We therefore extend the syntax of core $\Pi$ in Sec.~\ref{sub:core} as follows:
 \[\begin{array}{lrcl}
@@
 \textit{Non-empty types} & \net \\
-\textit{Values}      & v &::=& \cdots \alt \ocircle \\
+\textit{Values}      & v &::=& \cdots \alt \gcv \\
 \textit{Level-1 types} &&& t \leftrightarrow t \\
@@
 \textit{Level-2 programs} & \alpha &::=& \cdots 
-  \alt unity : \epsilon \circ \eta \Leftrightarrow \idc 
+  \alt \mathit{inv} : \epsilon \circ \eta \Leftrightarrow \idc 
 \end{array}\]
+
+The metavariable $\net$ excludes all types that are equivalent to the
+empty type, e.g., $(1 + 1) * 0$. For now, all gc values are represented
+at runtime using a trivial value denoted by $\gcv$. We also explicitly
+add the coherence condition on the creation and annihilation process
+as a level-2 program, thus enforcing the duality $\eta$ and
+$\epsilon$.
+
+In the Agda formalization, the types of $\eta$ and $\epsilon$ are as follows:
+
+\EtaEpsilon{}
+
+\dyninterp{}
+
+\vspace*{-0.5cm}
+$\vdots$
 

### 2019-11-20 fracGC/p.tex
Add some code to sec 4
try to put all the needed code in paper (WIP)
Code for extraction
code for MemSem

@@
 \newunicodechar{⟷}{$\leftrightarrow$}
+\newunicodechar{⟶}{$\longrightarrow$}
+\newunicodechar{⟪}{$\ll$}
+\newunicodechar{⟫}{$\gg$}
 
@@
 
+\input{latex/PiFracDynDefCode.tex}
 \input{latex/PiFracDynCode.tex}
 \input{latex/BooleanCircuitsCode.tex}
-
+\input{latex/PiPointedFracCode.tex}
+\input{latex/ExtractionCode.tex}
+\input{latex/PiMemSemCode.tex}
+\input{latex/PiFracMemSemCode.tex}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 superpositions, for example starting from the classical reversible
-language Theseus~\cite{XXX}, a small extension produces a quantum
+language Theseus~\cite{james2014theseus}, a small extension produces a quantum
 language with quantum control~\cite{10.1007/978-3-319-89366-2_19}. The
-language Theseus~\cite{XXX} mentioned above is a language that uses
+language Theseus~\cite{james2014theseus} mentioned above is a language that uses
 convenient syntactic extensions over an more basic language $\Pi$ of
@@
 
+\PIFDUdef{}
+
+change of combinators:
+\PIFDCombdef{}
+
+change of interpreter:
+\PIFDinterp{}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\PIPFUdef{}
+
 RECIP
+
+Combinators:
+\PIPFCombDef{}
+
+Evaluation:
+\PIPFeval{}
+
+Combine with eval, the whole thing is reversible:
+\PIPFrev{}
+
+\subsection{Extraction}
 
@@
 Polymorphic type for ancilla?
+
+Inject plain Pi into PiFracDyn:
+\INJU{}
+The injection is correct:
+\INJeq{}
+Extract type from PiPointedFrac:
+\EXTU{}
+Extract combinator:
+\EXTUComb{}
+The extraction is correct:
+\EXTeq{}
 
@@
 Use all the constructions name, coname, etc. and see what they do in this context!
+
+cardinality of pi type:
+\PIMEMcard{}
+enumeration:
+\PIMEMenum{}
+combinator preserve size:
+\PIMEMcardeq{}
+semantics for Pi:
+\PIMEMstate{}
+\PIMEMstep{}
+semantics for PiFracDyn:
+\PIFMEMstate{}
+\PIFMEMstep{}
 

### 2019-11-20 fracGC/p.tex
some code cleanup
Inj⟦𝕌⟧≅

@@
 \usepackage[nocenter]{qtree}
+\usepackage{multicol}  
 \usepackage{tikz}
@@
 
+\newunicodechar{≟}{$\mathbin{\stackrel{?}{=}}$}
 \newunicodechar{𝕌}{$\mathbb{U}$}
@@
 \newcommand{\jdg}[3]{#2 \proves_{#1} #3}
-
+\renewcommand{\AgdaCodeStyle}{\small}
 % 12 pages
@@
 \input{latex/PiFracMemSemCode.tex}
+\input{latex/J.tex}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{figure*}[t]
+\begin{multicols}{2}
 \[\begin{array}{rrcll}
@@
 \end{array}\]
+\vspace{2cm}
 \begin{center}
@@
 {}
-\\
+
+\vspace{1cm}
+
 \Rule{}
@@
 {}
-\qquad
+
+\vspace{1cm}
+
 \Rule{}
@@
 \end{center}
+\end{multicols}
 \caption{$\Pi$-terms and combinators.}
@@
 
+\Jexample{}
+
 %%%%% 
@@
 Polymorphic type for ancilla?
-
-Inject plain Pi into PiFracDyn:
+\\
+First we can inject plain Pi type and value into PiFracDyn:
 \INJU{}
-The injection is correct:
-\INJeq{}
+and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
+\INJcomb{}
+The given injection is correct because it does not change the denotation of type
+\INJUeq{}
+and the value
+\INJVeq{}
+Also the interpretation is also consistent:
+\INJEvaleq{}
 Extract type from PiPointedFrac:

### 2019-11-20 fracGC/p.tex
clean up code; move tikz to main text

--- 2019-11-20 fracGC/p.tex
@@
 {\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
-{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
+{\jdg{}{}{c_1 \fatsemi c_2 : \tau_1 \iso \tau_3}}
 {}
-\\
+
+\vspace{1cm}
+
 \Rule{}
@@
 \noindent The top path is the $\Pi$ program
-$(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$,
+$(c_1~\oplus~c_2)~\fatsemi~\swapp$ which acts on the type $A$ by $c_1$,
 acts on the type $B$ by $c_2$, and deforms the resulting space by a
@@
 there exists a continuous deformation of the program
-$(c_1~\oplus~c_2)~\odot~\swapp$ to the program
-$\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means
-that, as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and
-$\swapp \odot (c_2~\oplus~c_1)$ are equal. This relation is
+$(c_1~\oplus~c_2)~\fatsemi~\swapp$ to the program
+$\swapp \fatsemi (c_2~\oplus~c_1)$. We can also show that this means
+that, as permutations, $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and
+$\swapp \fatsemi (c_2~\oplus~c_1)$ are equal. This relation is
 non-trivial, as not all programs between the same types can be
@@
 \verb|with_ancilla| operator from
-Quipper~\cite{Green:2013:QSQ:2491956.2462177}:
+Quipper~\cite{Green:2013:QSQ:2491956.2462177} :
 \begin{verbatim}
@@
 \textit{Level-2 programs} & \alpha &::=& \cdots 
-  \alt \mathit{inv} : \epsilon \circ \eta \Leftrightarrow \idc 
+  \alt \mathit{inv} : \eta \fatsemi \epsilon \Leftrightarrow \idc 
 \end{array}\]
@@
 \EtaEpsilonEval{}
-
-%%%%%
-\subsection{Examples}
-
-\EtaEpsilonExamples{}
-
-% of course need de-allocation; keep track of lifetime
-
-% easy: stack but not general
-
-% for general; programmer control or gc
-
-% context of reversible; similar; need allocation; same tradeoffs for
-% de-allocation; stack or programmer or gc BUT with caveat. value must
-% be restored to its initial value; no information leakage which would
-% destroy quantum state
-
-% but reversible also gives you more guarantees; no creation or deletion
-% of information means no need to keep track of lifetime of individual
-% locations; just need to keep track of ``information''
-
-% General allocation and de-allocation. No need to keep track of
-% lifetime. Just make sure location is back to initial value. Safe to
-% de-allocate because indistinguishable from freshly created. No pointer
-% equality and aliases though. Concurrent: one thread allocates; gc
-% thread meets someone else
-
-current proposals for reversible
-and quantum programming languages
-(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
-Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
-to be used in nested scopes, have sound but incomplete laws for
-reasoning about the safety condition, and only guarantee safety via
-runtime checks.
-
-move from 1/t to 1/v (how the quantum search runs; need ancilla 
-initialized to a specific value)
-
-\PIFDUdef{}
-
-change of combinators:
-\PIFDCombdef{}
-
-change of interpreter:
-\PIFDinterp{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Dependently-Typed Garbage Collectors}
-\label{sec:dep}
-
-PiPointedFrac:
-lift pi
-lift multiplicative structure
-monad/comonad for singletons
-eta/epsilon
-
-rationale: we want singleton types to track, in the type system
-the values of ancillae, i.e.. we construct a proof using the type
-system. Technically the singleton types allow you to focus on one
-value by reflecting it in the type system. Now this type can only be
-equal to another type if and only if the other type is the same
-singleton value.
-
-Once singleton types are built, we need to make sure circuit
-evaluation is also reflected in the type system. So in fact
-computation itself is reflected in the type system. In order to mix
-singleton types with other types, we use pointed types as a unifying
-substrate. Now all types have values; we can focus (using monad) and
-unfocus using (comonad). (This by the way ensures we don't write 1/0
-which woudl be bad).
-
-We use a Singleton type
-to remember this information.
-
-Formalize ancilla problem in the context of Pi using eta and epsilon
-as done in PiFracDyn; more general than scoped; still runtime
-check.
 
@@
 %%%%%
+\subsection{Examples}
+
+\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt        
+\begin{tikzpicture}[x=0.75pt,y=0.75pt,scale=0.8, yscale=-1]
+%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
+%Straight Lines [id:da05403550183378514] 
+\draw    (10,20) -- (190,20) ;
+%Shape: Circle [id:dp8701819096818126] 
+\draw   (138.75,21.25) ..
+controls (138.75,15.04) and (143.79,10) ..
+(150,10) ..
+controls (156.21,10) and (161.25,15.04) ..
+(161.25,21.25) ..
+controls (161.25,27.46) and (156.21,32.5) ..
+(150,32.5) ..
+controls (143.79,32.5) and (138.75,27.46) ..
+(138.75,21.25) -- cycle ;
+%Straight Lines [id:da3470499764023487] 
+\draw    (150,80) -- (150,10) ;
+%Straight Lines [id:da9303973632982276] 
+\draw    (40,80) -- (190,80) ;
+%Shape: Circle [id:dp03942567054970991] 
+\draw   (48.75,81.33) ..
+controls (48.75,75.12) and (53.79,70.08) ..
+(60,70.08) ..
+controls (66.21,70.08) and (71.25,75.12) ..
+(71.25,81.33) ..
+controls (71.25,87.55) and (66.21,92.58) ..
+(60,92.58) ..
+controls (53.79,92.58) and (48.75,87.55) ..
+(48.75,81.33) -- cycle ;
+%Straight Lines [id:da08295302434157614] 
+\draw    (60,92.58) -- (60,20) ;
+%Shape: Arc [id:dp2904677718814479] 
+\draw  [draw opacity=0] (38.49,110.13) ..
+controls (28.3,109.22) and (20.27,102.7) ..
+(20.15,94.89) ..
+controls (20.02,86.57) and (28.89,79.93) ..
+(40,80) -- (40.38,95.12) -- cycle ; \draw   (38.49,110.13) ..
+controls (28.3,109.22) and (20.27,102.7) ..
+(20.15,94.89) ..
+controls (20.02,86.57) and (28.89,79.93) ..
+(40,80) ;
+%Straight Lines [id:da32172184710713914] 
+\draw    (38.49,110.13) -- (250,110) ;
+%Shape: Arc [id:dp8398868072937593] 
+\draw  [draw opacity=0] (248.77,109.75) ..
+controls (262.77,109.82) and (274.32,103.33) ..
+(274.64,95.17) ..
+controls (274.95,87.02) and (263.95,80.28) ..
+(250,80) -- (249.05,94.87) -- cycle ; \draw   (248.77,109.75) ..
+controls (262.77,109.82) and (274.32,103.33) ..
+(274.64,95.17) ..
+controls (274.95,87.02) and (263.95,80.28) ..
+(250,80) ;
+%Straight Lines [id:da5220296307776525] 
+\draw    (190,20) -- (250,80) ;
+%Straight Lines [id:da320377013704656] 
+\draw    (190,80) -- (250,20) ;
+%Straight Lines [id:da8933694171914682] 
+\draw    (250,20) -- (308,20) ;
+\end{tikzpicture}
+
+\begin{tikzpicture}[x=0.75pt,y=0.75pt,scale=0.8, yscale=-1]
+%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
+%Straight Lines [id:da9303973632982276] 
+\draw    (40,80) -- (190,80) ;
+%Shape: Arc [id:dp2904677718814479] 
+\draw  [draw opacity=0] (38.49,110.13) ..
+controls (28.3,109.22) and (20.27,102.7) ..
+(20.15,94.89) ..
+controls (20.02,86.57) and (28.89,79.93) ..
+(40,80) -- (40.38,95.12) -- cycle ; \draw   (38.49,110.13) ..
+controls (28.3,109.22) and (20.27,102.7) ..
+(20.15,94.89) ..
+controls (20.02,86.57) and (28.89,79.93) ..
+(40,80) ;
+%Straight Lines [id:da32172184710713914] 
+\draw    (38.49,110.13) -- (250,110) ;
+%Shape: Arc [id:dp8398868072937593] 
+\draw  [draw opacity=0] (248.77,109.75) ..
+controls (262.77,109.82) and (274.32,103.33) ..
+(274.64,95.17) ..
+controls (274.95,87.02) and (263.95,80.28) ..
+(250,80) -- (249.05,94.87) -- cycle ; \draw   (248.77,109.75) ..
+controls (262.77,109.82) and (274.32,103.33) ..
+(274.64,95.17) ..
+controls (274.95,87.02) and (263.95,80.28) ..
+(250,80) ;
+%Straight Lines [id:da7474565765110813] 
+\draw    (40.09,30.3) -- (190.09,30.3) ;
+%Shape: Arc [id:dp593084478238281] 
+\draw  [draw opacity=0] (38.58,60.43) ..
+controls (28.39,59.52) and (20.36,53) ..
+(20.23,45.19) ..
+controls (20.11,36.87) and (28.98,30.23) ..
+(40.09,30.3) -- (40.47,45.42) -- cycle ; \draw   (38.58,60.43) ..
+controls (28.39,59.52) and (20.36,53) ..
+(20.23,45.19) ..
+controls (20.11,36.87) and (28.98,30.23) ..
+(40.09,30.3) ;
+%Straight Lines [id:da5682447818267274] 
+\draw    (38.58,60.43) -- (250.09,60.3) ;
+%Shape: Arc [id:dp8936250935910023] 
+\draw  [draw opacity=0] (248.86,60.05) ..
+controls (262.86,60.12) and (274.41,53.63) ..
+(274.72,45.47) ..
+controls (275.03,37.32) and (264.04,30.58) ..
+(250.09,30.3) -- (249.13,45.17) -- cycle ; \draw   (248.86,60.05) ..
+controls (262.86,60.12) and (274.41,53.63) ..
+(274.72,45.47) ..
+controls (275.03,37.32) and (264.04,30.58) ..
+(250.09,30.3) ;
+%Straight Lines [id:da9423433355546923] 
+\draw    (190,30) -- (250,80) ;
+%Straight Lines [id:da8064437888434692] 
+\draw    (190,80) -- (250.09,30.3) ;
+\end{tikzpicture}
+
+\EtaEpsilonExamples{}
+
+% of course need de-allocation; keep track of lifetime
+% easy: stack but not general
+% for general; programmer control or gc
+
+% context of reversible; similar; need allocation; same tradeoffs for
+% de-allocation; stack or programmer or gc BUT with caveat. value must
+% be restored to its initial value; no information leakage which would
+% destroy quantum state
+
+% but reversible also gives you more guarantees; no creation or deletion
+% of information means no need to keep track of lifetime of individual
+% locations; just need to keep track of ``information''
+
+% General allocation and de-allocation. No need to keep track of
+% lifetime. Just make sure location is back to initial value. Safe to
+% de-allocate because indistinguishable from freshly created. No pointer
+% equality and aliases though. Concurrent: one thread allocates; gc
+% thread meets someone else
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Dependently-Typed Garbage Collectors}
+\label{sec:dep}
+
+PiPointedFrac:
+lift pi
+lift multiplicative structure
+monad/comonad for singletons
+eta/epsilon
+
+rationale: we want singleton types to track, in the type system
+the values of ancillae, i.e.. we construct a proof using the type
+system. Technically the singleton types allow you to focus on one
+value by reflecting it in the type system. Now this type can only be
+equal to another type if and only if the other type is the same
+singleton value.
+
+Once singleton types are built, we need to make sure circuit
+evaluation is also reflected in the type system. So in fact
+computation itself is reflected in the type system. In order to mix
+singleton types with other types, we use pointed types as a unifying
+substrate. Now all types have values; we can focus (using monad) and
+unfocus using (comonad). (This by the way ensures we don't write 1/0
+which woudl be bad).
+
+%%%%% 
+\subsection{State of the Art}
+
+current proposals for reversible
+and quantum programming languages
+(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
+Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
+to be used in nested scopes, have sound but incomplete laws for
+reasoning about the safety condition, and only guarantee safety via
+runtime checks.
+
+Ricercar quote:
+\begin{quote}
+  Checking that some defined ancillae are indeed unchanged for all
+  possible input vectors of an arbitrary description is hard; in
+  general, one has to test all possible input vectors, which is
+  undesirable. For a reversible programming language such as Janus
+  [22], this is therefore imple- mented as a runtime check that checks
+  the reversibility of a program only in rela- tion to the executing
+  input vector. This is also the case for the syntax described in
+  Fig. 1.
+
+  To be able to introduce and remove ancilla wires from a circuit, we
+  have identified the need for five basic rules.
+\end{quote}
+
+%%%%% 
+\subsection{Choosing Ancillae Values}
+
+move from 1/t to 1/v (how the quantum search runs; need ancilla 
+initialized to a specific value)
+
+\PIFDUdef{}
+
+change of combinators:
+\PIFDCombdef{}
+
+change of interpreter:
+\PIFDinterp{}
+
+%%%%% 
 \subsection{Pointed Types}
+
+\Jexample{}
 

### 2019-11-20 fracGC/p.tex
better picture more explanations

@@
 \usepackage{multicol}  
+\usepackage{bbold}
 \usepackage{tikz}
-\usepackage{bbold}
 \usetikzlibrary{cd}
 \usetikzlibrary{quotes}
+
+\usepackage{tikzit}
+\input{frac.tikzstyles}
 
@@
 
-\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt        
-\begin{tikzpicture}[x=0.75pt,y=0.75pt,scale=0.8, yscale=-1]
-%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
-%Straight Lines [id:da05403550183378514] 
-\draw    (10,20) -- (190,20) ;
-%Shape: Circle [id:dp8701819096818126] 
-\draw   (138.75,21.25) ..
-controls (138.75,15.04) and (143.79,10) ..
-(150,10) ..
-controls (156.21,10) and (161.25,15.04) ..
-(161.25,21.25) ..
-controls (161.25,27.46) and (156.21,32.5) ..
-(150,32.5) ..
-controls (143.79,32.5) and (138.75,27.46) ..
-(138.75,21.25) -- cycle ;
-%Straight Lines [id:da3470499764023487] 
-\draw    (150,80) -- (150,10) ;
-%Straight Lines [id:da9303973632982276] 
-\draw    (40,80) -- (190,80) ;
-%Shape: Circle [id:dp03942567054970991] 
-\draw   (48.75,81.33) ..
-controls (48.75,75.12) and (53.79,70.08) ..
-(60,70.08) ..
-controls (66.21,70.08) and (71.25,75.12) ..
-(71.25,81.33) ..
-controls (71.25,87.55) and (66.21,92.58) ..
-(60,92.58) ..
-controls (53.79,92.58) and (48.75,87.55) ..
-(48.75,81.33) -- cycle ;
-%Straight Lines [id:da08295302434157614] 
-\draw    (60,92.58) -- (60,20) ;
-%Shape: Arc [id:dp2904677718814479] 
-\draw  [draw opacity=0] (38.49,110.13) ..
-controls (28.3,109.22) and (20.27,102.7) ..
-(20.15,94.89) ..
-controls (20.02,86.57) and (28.89,79.93) ..
-(40,80) -- (40.38,95.12) -- cycle ; \draw   (38.49,110.13) ..
-controls (28.3,109.22) and (20.27,102.7) ..
-(20.15,94.89) ..
-controls (20.02,86.57) and (28.89,79.93) ..
-(40,80) ;
-%Straight Lines [id:da32172184710713914] 
-\draw    (38.49,110.13) -- (250,110) ;
-%Shape: Arc [id:dp8398868072937593] 
-\draw  [draw opacity=0] (248.77,109.75) ..
-controls (262.77,109.82) and (274.32,103.33) ..
-(274.64,95.17) ..
-controls (274.95,87.02) and (263.95,80.28) ..
-(250,80) -- (249.05,94.87) -- cycle ; \draw   (248.77,109.75) ..
-controls (262.77,109.82) and (274.32,103.33) ..
-(274.64,95.17) ..
-controls (274.95,87.02) and (263.95,80.28) ..
-(250,80) ;
-%Straight Lines [id:da5220296307776525] 
-\draw    (190,20) -- (250,80) ;
-%Straight Lines [id:da320377013704656] 
-\draw    (190,80) -- (250,20) ;
-%Straight Lines [id:da8933694171914682] 
-\draw    (250,20) -- (308,20) ;
-\end{tikzpicture}
+The first example below illustrates the basic functionality of
+ancillae. The circuit has one input and one output. Immediately after
+receiving the input, the circuit generates an ancilla wire and its
+corresponding gc process (first line in the Agda definition). The
+original input and the ancilla wire interact using two
+\AgdaFunction{CNOT} gates, after which the ancilla wire is redirected
+to the output (second line in the Agda code). Finally the original
+input is gc'ed (third line in the Agda code). The entire circuit is
+extensionally equivalent to the identity function.
+
+\begin{center}
+\input{idpic.tikz}
+\end{center}
+
+\EtaEpsilonExamples{}
 
@@
 
-\EtaEpsilonExamples{}
-
 % of course need de-allocation; keep track of lifetime

### 2019-11-20 fracGC/p.tex
more cleanup code

@@
 the Toffoli gate for classical reversible computing and of the
-universality of the Toffol-Hadamard gates for quantum computing should
+universality of the Toffoli-Hadamard gates for quantum computing should
 not distract from efficiency and safety concerns. The theorems proving
@@
 The first example below illustrates the basic functionality of
-ancillae. The circuit has one input and one output. Immediately after
-receiving the input, the circuit generates an ancilla wire and its
-corresponding gc process (first line in the Agda definition). The
-original input and the ancilla wire interact using two
-\AgdaFunction{CNOT} gates, after which the ancilla wire is redirected
-to the output (second line in the Agda code). Finally the original
-input is gc'ed (third line in the Agda code). The entire circuit is
-extensionally equivalent to the identity function.
-
+ancillae. The Agda code is written in a style that reveals the
+intermediate steps for convenience. The circuit has one input and one
+output. Immediately after receiving the input, the circuit generates
+an ancilla wire and its corresponding gc process (first two steps in
+the Agda definition). The original input and the ancilla wire interact
+using two \AgdaFunction{CNOT} gates, after which the ancilla wire is
+redirected to the output (next three steps in the Agda code). Finally
+the original input is gc'ed (last two steps in the Agda code). The
+entire circuit is extensionally equivalent to the identity function.
+
+\medskip
 \begin{center}
@@
 
-\EtaEpsilonExamples{}
+\EtaEpsilonExampleone{}
 
@@
 
+\EtaEpsilonExampletwo{}
+
 % of course need de-allocation; keep track of lifetime

### 2019-11-20 fracGC/p.tex
some explanation about extraction

@@
 \INJUeq{}
-and the value
+,the value
 \INJVeq{}
-Also the interpretation is also consistent:
+and the interpretation of PiFracDyn is consistent with Pi:
 \INJEvaleq{}
-Extract type from PiPointedFrac:
+To extract a combinator from PiPointedFrac, first we need to extract the corresponding type for PiFracDyn.
+Since every type (except Recip) in PiPointedFrac is pointed type so we can always extract a type for plain
+Pi and a value from it.
+There is no corresponding type of Recip in plain Pi, but we can map it to the
+\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v} type in PiFracDyn.
+Note that for all other type the extracted value is erased from type except for
+\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
+which we still keep it in type.
 \EXTU{}
-Extract combinator:
+Once we extract the corresponding type we can extract the combinator, the combinators about monad/comonad
+just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased:
 \EXTUComb{}
-The extraction is correct:
+And we prove the correctness of the extraction, since the combinators about monad/comonad does not really
+computes so for those case the proof is just \AgdaInductiveConstructor{refl},
+and for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
+The interesting case is \AgdaInductiveConstructor{ε} where the singleton type guarantee that
+the dynamic check is passed.
 \EXTeq{}

### 2019-11-20 fracGC/p.tex
revx
done with sec 4 for now

@@
 \end{array}\]
-We therefore extend the syntax of core $\Pi$ in Sec.~\ref{sub:core} as follows:
+The second important detail is to ensure that the gc process is
+specialized to collect a particular value. We however defer this point to
+the next section and for now create gc processes that might be applied
+to incorrect values throwing exceptions instead of collecting them. We
+therefore extend the syntax of core $\Pi$ in Sec.~\ref{sub:core} as
+follows:
+
 \[\begin{array}{lrcl}
@@
 The metavariable $\net$ excludes all types that are equivalent to the
-empty type, e.g., $(1 + 1) * 0$. For now, all gc values are represented
-at runtime using a trivial value denoted by $\gcv$. We also explicitly
-add the coherence condition on the creation and annihilation process
-as a level-2 program, thus enforcing the duality $\eta$ and
-$\epsilon$.
+empty type, e.g., $(1 + 1) * 0$. For now, all gc values are
+represented at runtime using a trivial value denoted by $\gcv$. We
+also explicitly add the desired coherence condition on the creation
+and annihilation process as a level-2 program that would enforce the
+duality of $\eta$ and $\epsilon$. At this point, an application of
+$\eta$ following by $\epsilon$ might however throw an exception.
 
@@
 
+\noindent where the operations take an implicit argument asserting the
+given type is non-empty. The operational semantics of the language
+interprets each combinator (except $\eta$ and $\epsilon$) of type
+$t_1 \leftrightarrow t_2$ as a (reversible) function mapping values of
+type $t_1$ to values of type $t_2$ (and vice-versa). Extending this
+interpreter to handle $\eta$ and $\epsilon$ requires at this point
+that the interpreter returns either a value of the appropriate or an
+exception:
+
 \dyninterp{}
 
-\vspace*{-0.5cm}
-$\vdots$
-
-\EtaEpsilonEval{}
-
-When ancilla are created they are created with a specific
-value. This value is encoded in the circuit. The value of fractional
-type represents the ``garbage collector.''  The garbage collector is
-specialized to collect a single value, which is the value used to
-create the ancilla value. At this point, if the garbage collector
-encounters a different value, it doesn't know what to do, and would
-need to throw an exception. This situation in \verb|PiFracDyn| is
-similar to current solutions (e.g. Quipper) but actually more general
-as it allows non-scoped ancilla allocation and de-allocation. Show
-examples.
+When ancillae are created they are created with a specific default
+value depending on the type. When the gc process encounters a value to
+be collected, it checks whether this value is the default value for
+the type and if so collects it. Otherwise it returns
+\AgdaInductiveConstructor{nothing} indicating an exception.
 
@@
 ancillae. The Agda code is written in a style that reveals the
-intermediate steps for convenience. The circuit has one input and one
-output. Immediately after receiving the input, the circuit generates
-an ancilla wire and its corresponding gc process (first two steps in
-the Agda definition). The original input and the ancilla wire interact
-using two \AgdaFunction{CNOT} gates, after which the ancilla wire is
-redirected to the output (next three steps in the Agda code). Finally
-the original input is gc'ed (last two steps in the Agda code). The
-entire circuit is extensionally equivalent to the identity function.
+intermediate steps for easier correspondence with the figure. The
+circuit has one input and one output. Immediately after receiving the
+input, the circuit generates an ancilla wire and its corresponding gc
+process (first two steps in the Agda definition). The original input
+and the ancilla wire interact using two \AgdaFunction{CNOT} gates,
+after which the ancilla wire is redirected to the output (next three
+steps in the Agda code). Finally the original input is gc'ed (last two
+steps in the Agda code). The entire circuit is extensionally
+equivalent to the identity function but it does highlight an important
+functionality beyond scoped ancillae management: the allocated ancilla
+bit is redirected to the output and a completely different bit (with
+the proper default value) is collected instead.
 
@@
 
-\begin{tikzpicture}[x=0.75pt,y=0.75pt,scale=0.8, yscale=-1]
-%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
-%Straight Lines [id:da9303973632982276] 
-\draw    (40,80) -- (190,80) ;
-%Shape: Arc [id:dp2904677718814479] 
-\draw  [draw opacity=0] (38.49,110.13) ..
-controls (28.3,109.22) and (20.27,102.7) ..
-(20.15,94.89) ..
-controls (20.02,86.57) and (28.89,79.93) ..
-(40,80) -- (40.38,95.12) -- cycle ; \draw   (38.49,110.13) ..
-controls (28.3,109.22) and (20.27,102.7) ..
-(20.15,94.89) ..
-controls (20.02,86.57) and (28.89,79.93) ..
-(40,80) ;
-%Straight Lines [id:da32172184710713914] 
-\draw    (38.49,110.13) -- (250,110) ;
-%Shape: Arc [id:dp8398868072937593] 
-\draw  [draw opacity=0] (248.77,109.75) ..
-controls (262.77,109.82) and (274.32,103.33) ..
-(274.64,95.17) ..
-controls (274.95,87.02) and (263.95,80.28) ..
-(250,80) -- (249.05,94.87) -- cycle ; \draw   (248.77,109.75) ..
-controls (262.77,109.82) and (274.32,103.33) ..
-(274.64,95.17) ..
-controls (274.95,87.02) and (263.95,80.28) ..
-(250,80) ;
-%Straight Lines [id:da7474565765110813] 
-\draw    (40.09,30.3) -- (190.09,30.3) ;
-%Shape: Arc [id:dp593084478238281] 
-\draw  [draw opacity=0] (38.58,60.43) ..
-controls (28.39,59.52) and (20.36,53) ..
-(20.23,45.19) ..
-controls (20.11,36.87) and (28.98,30.23) ..
-(40.09,30.3) -- (40.47,45.42) -- cycle ; \draw   (38.58,60.43) ..
-controls (28.39,59.52) and (20.36,53) ..
-(20.23,45.19) ..
-controls (20.11,36.87) and (28.98,30.23) ..
-(40.09,30.3) ;
-%Straight Lines [id:da5682447818267274] 
-\draw    (38.58,60.43) -- (250.09,60.3) ;
-%Shape: Arc [id:dp8936250935910023] 
-\draw  [draw opacity=0] (248.86,60.05) ..
-controls (262.86,60.12) and (274.41,53.63) ..
-(274.72,45.47) ..
-controls (275.03,37.32) and (264.04,30.58) ..
-(250.09,30.3) -- (249.13,45.17) -- cycle ; \draw   (248.86,60.05) ..
-controls (262.86,60.12) and (274.41,53.63) ..
-(274.72,45.47) ..
-controls (275.03,37.32) and (264.04,30.58) ..
-(250.09,30.3) ;
-%Straight Lines [id:da9423433355546923] 
-\draw    (190,30) -- (250,80) ;
-%Straight Lines [id:da8064437888434692] 
-\draw    (190,80) -- (250.09,30.3) ;
-\end{tikzpicture}
+The second example illustrates the manipulation of gc processes. A
+process for collecting a pair of values can be decomposed into two
+processes each collecting one of the values (and vice-versa):
 
 \EtaEpsilonExampletwo{}
-
-% of course need de-allocation; keep track of lifetime
-% easy: stack but not general
-% for general; programmer control or gc
-
-% context of reversible; similar; need allocation; same tradeoffs for
-% de-allocation; stack or programmer or gc BUT with caveat. value must
-% be restored to its initial value; no information leakage which would
-% destroy quantum state
-
-% but reversible also gives you more guarantees; no creation or deletion
-% of information means no need to keep track of lifetime of individual
-% locations; just need to keep track of ``information''
-
-% General allocation and de-allocation. No need to keep track of
-% lifetime. Just make sure location is back to initial value. Safe to
-% de-allocate because indistinguishable from freshly created. No pointer
-% equality and aliases though. Concurrent: one thread allocates; gc
-% thread meets someone else
 
@@
 
+% General allocation and de-allocation. No need to keep track of
+% lifetime. Just make sure location is back to initial value. Safe to
+% de-allocate because indistinguishable from freshly created. No pointer
+% equality and aliases though. Concurrent: one thread allocates; gc
+% thread meets someone else
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-11-20 fracGC/p.tex
before merge

@@
 The metavariable $\net$ excludes all types that are equivalent to the
-empty type, e.g., $(1 + 1) * 0$. For now, all gc values are
+empty type, e.g., $(1 + 1) * 0$. For now, all gc processes are
 represented at runtime using a trivial value denoted by $\gcv$. We
@@
 
-PiPointedFrac:
-lift pi
-lift multiplicative structure
-monad/comonad for singletons
-eta/epsilon
+By lifting the scoping restriction, the development in the previous
+section is already more general than the state of the art in ancilla
+management.  It however shares the same limitation of requiring a
+runtime check to ensure ancillae values are properly restored to their
+allocation~\cite{10.1007/978-3-319-20860-2_13,Green:2013:QSQ:2491956.2462177}.
+In this section, we address this limitation using a combination of
+pointed types, singleton types, monads, and comonads.
+
+%%%%% 
+\subsection{Lifting Evaluation to Type System}
+
+Before giving all the (rather involved) technical details, we
+highlight the main idea using the toy language below:
+
+\Jexample{}
 
@@
 %%%%% 
-\subsection{State of the Art}
-
-current proposals for reversible
-and quantum programming languages
-(e.g. Ricercar~\cite{10.1007/978-3-319-20860-2_13},
-Quipper~\cite{Green:2013:QSQ:2491956.2462177}), restrict ancilla bits
-to be used in nested scopes, have sound but incomplete laws for
-reasoning about the safety condition, and only guarantee safety via
-runtime checks.
-
-Ricercar quote:
-\begin{quote}
-  Checking that some defined ancillae are indeed unchanged for all
-  possible input vectors of an arbitrary description is hard; in
-  general, one has to test all possible input vectors, which is
-  undesirable. For a reversible programming language such as Janus
-  [22], this is therefore imple- mented as a runtime check that checks
-  the reversibility of a program only in rela- tion to the executing
-  input vector. This is also the case for the syntax described in
-  Fig. 1.
-
-  To be able to introduce and remove ancilla wires from a circuit, we
-  have identified the need for five basic rules.
-\end{quote}
-
-%%%%% 
 \subsection{Choosing Ancillae Values}
@@
 \subsection{Pointed Types}
-
-\Jexample{}
 
@@
 
+PiPointedFrac:
+lift pi
+lift multiplicative structure
+monad/comonad for singletons
+eta/epsilon
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-11-20 fracGC/p.tex
reorder sections

@@
 ideas of fractional types and partially solves a more general and more
-expressive instance of the ancilla problem. The approach however
-shares the limitation of current solutions in that it relies on
-runtime checks for programmer's assertions that values have been
-restored to their initial value before de-allocation. To remedy this
-limitation, we refine the type system in Sec.~\ref{sec:dep} to keep
-track of values in focus. Technically this requires the introduction
-of pointed types on top of which are defined both a monad and co-monad
-of singleton types. To justify the claim that fractional types solve
-the management of ancilla-based scratch space, we present in
-Sec.~\ref{sec:space} an operational semantics that exposes memory
-management details using a realization of fractional types as
-first-class dependently-typed garbage collectors. In
-Sec.~\ref{sec:cat} we present a rich collection of examples
-illustrating the expressive power of fractional types. We conclude
-with a discussion of related work and a summary of our results.
+expressive instance of the ancilla problem. To make the memory
+management explicit, Sec.~\ref{sec:space} provides an operational
+semantics that exposes memory management details using a realization
+of fractional types as first-class garbage collectors. In
+Sec.~\ref{sec:dep} we refine the type system to keep track of ancillae
+values. Technically this requires the introduction of pointed types on
+top of which are defined both a monad and co-monad of singleton
+types. We conclude with a discussion of related work and a summary of
+our results.
 
@@
 
+%%%%% 
+\subsection{Choosing Ancillae Values}
+
+move from 1/t to 1/v (how the quantum search runs; need ancilla 
+initialized to a specific value)
+
+\PIFDUdef{}
+
+change of combinators:
+\PIFDCombdef{}
+
+change of interpreter:
+\PIFDinterp{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Space Operational Semantics}
+\label{sec:space}
+
+Positive rational numbers are a model. Apparently there is a
+categorification
+\url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
+
+Use all the constructions name, coname, etc. and see what they do in this context!
+
+cardinality of pi type:
+\PIMEMcard{}
+enumeration:
+\PIMEMenum{}
+combinator preserve size:
+\PIMEMcardeq{}
+semantics for Pi:
+\PIMEMstate{}
+\PIMEMstep{}
+semantics for PiFracDyn:
+\PIFMEMstate{}
+\PIFMEMstep{}
+
+Evaluation:
+\PIPFeval{}
+
+Combine with eval, the whole thing is reversible:
+\PIPFrev{}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%% 
-\subsection{Choosing Ancillae Values}
-
-move from 1/t to 1/v (how the quantum search runs; need ancilla 
-initialized to a specific value)
-
-\PIFDUdef{}
-
-change of combinators:
-\PIFDCombdef{}
-
-change of interpreter:
-\PIFDinterp{}
-
-%%%%% 
 \subsection{Pointed Types}
@@
 
-Evaluation:
-\PIPFeval{}
-
-Combine with eval, the whole thing is reversible:
-\PIPFrev{}
-
+%%%%%%%%%
 \subsection{Extraction}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Space Operational Semantics}
-\label{sec:space}
-
-Positive rational numbers are a model. Apparently there is a
-categorification
-\url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
-
-Use all the constructions name, coname, etc. and see what they do in this context!
-
-cardinality of pi type:
-\PIMEMcard{}
-enumeration:
-\PIMEMenum{}
-combinator preserve size:
-\PIMEMcardeq{}
-semantics for Pi:
-\PIMEMstate{}
-\PIMEMstep{}
-semantics for PiFracDyn:
-\PIFMEMstate{}
-\PIFMEMstep{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Trace}
+%%%%%%%%%%%
+\subsection{Examples}
 \label{sec:cat}

### 2019-11-21 fracGC/p.tex
All tensor combinators are derivable
fix symbol
about trace

@@
 \newunicodechar{⟫}{$\gg$}
+\newunicodechar{⊎}{$\uplus$}
+\newunicodechar{ₗ}{$_l$}
+\newunicodechar{ᵣ}{$_r$}
 
@@
 
+PiPointedFrac:
+lift pi
+lift multiplicative structure
+monad/comonad for singletons
+eta/epsilon
+
 The denotation of the fractional type is now:
@@
 Once we extract the corresponding type we can extract the combinator, the combinators about monad/comonad
-just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased:
+just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased,
+for the rest combinators there exist direct corresponding combinators in PiFracDyn:
 \EXTUComb{}
-And we prove the correctness of the extraction, since the combinators about monad/comonad does not really
-computes so for those case the proof is just \AgdaInductiveConstructor{refl},
-and for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
+Finally we can prove the correctness of the extraction,
+the lifted multiplicative combinators have the corresponding combinators in PiFracDyn which represent
+the same computation so the proof is just \AgdaInductiveConstructor{refl}.
+Because the combinators about monad/comonad does not really
+computes so for those case the proof is also \AgdaInductiveConstructor{refl}.
+And for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
 The interesting case is \AgdaInductiveConstructor{ε} where the singleton type guarantee that
-the dynamic check is passed.
+the dynamic check won't fail.
 \EXTeq{}
-
-PiPointedFrac:
-lift pi
-lift multiplicative structure
-monad/comonad for singletons
-eta/epsilon
 

### 2019-11-21 fracGC/p.tex
choosing ancilla

@@
 \textit{Level-1 programs} & c &::=& \cdots \alt
-   \eta : 1 \leftrightarrow (\net * \oneover{\net}) \\
-   &&& ~~~~~\alt \epsilon : (\net * \oneover{\net}) \leftrightarrow 1 \\
+   \eta_{\net} : 1 \leftrightarrow (\net * \oneover{\net}) \\
+   &&& ~~~~~\alt \epsilon_{\net} : (\net * \oneover{\net}) \leftrightarrow 1 \\
 \textit{Level-2 types} &&& c \Leftrightarrow c \\
@@
 
-move from 1/t to 1/v (how the quantum search runs; need ancilla 
-initialized to a specific value)
+The previous development achieves a significant milestone but only
+hints at the full expressiveness of fractional types. A small but
+significant additional expressiveness is to allow ancillae to be
+initialized to a value of the programmer's choice, not just a globally
+fixed default value. This ability is used in Grover's search algorithm
+to implement phase kickback
+efficiently~\cite{howgrover}.\footnote{Instead of using an ancilla bit
+  initialized to $\ket{0}$, an ancilla initialized to
+  $\frac{\ket{0}-\ket{1}}{\sqrt{2}}$ is used.} This can be easily
+achieved in our language as follows:
+
+\[\begin{array}{lrcl}
+\textit{Value types} & t &::=& \cdots \alt \oneover{v} \\
+\textit{Values}      & v &::=& \cdots \alt \gcv \\
+\textit{Level-1 types} &&& t \leftrightarrow t \\
+\textit{Level-1 programs} & c &::=& \cdots \alt
+   \eta_{v:t} : 1 \leftrightarrow (t * \oneover{v}) \\
+   &&& ~~~~~\alt \epsilon_{v:t} : (t * \oneover{v}) \leftrightarrow 1 \\
+\textit{Level-2 types} &&& c \Leftrightarrow c \\
+\textit{Level-2 programs} & \alpha &::=& \cdots 
+  \alt \mathit{inv} : \eta_{v:t} \fatsemi \epsilon_{v:t} \Leftrightarrow \idc 
+\end{array}\]
+
+In addition to being more expressive, this version has the advantage
+of replacing the cumbersome evidence that the type is not empty by the
+more intuitive evidence of choosing a value of the type. The
+simplications are evident in the Agda formalization:
 
@@
 
-change of combinators:
+\vspace{-\baselineskip}
+
 \PIFDCombdef{}
 
-change of interpreter:
+\vspace{-\baselineskip}
+
 \PIFDinterp{}
+
+Example Sec. 3.2 of
+\url{https://people.engr.ncsu.edu/hzhou/quantum_assert.pdf} with ancilla
+initialized to 1 to detected entanglement
 
@@
 
+Quantum wire recycling~\cite{PhysRevA.94.042337}
+
+Quantum runtime assertion~\cite{DBLP:journals/cal/ZhouB19}
+
+QWIRE~\cite{Paykin:2017:QCL:3009837.3009894}
+
 % General allocation and de-allocation. No need to keep track of

### 2019-11-21 fracGC/p.tex
done with extended sec 4

@@
 \usepackage{bbm}
+\usepackage{wrapfig}
 \usepackage[greek,english]{babel}
@@
 fact, classical reversible computations are often used as
-``subroutines'' of quantum computations.
+``subroutines'' of quantum computations. Indeed classical reversible
+circuits can be applied to either classical inputs or to quantum
+inputs.
 
@@
 In addition to being more expressive, this version has the advantage
-of replacing the cumbersome evidence that the type is not empty by the
-more intuitive evidence of choosing a value of the type. The
-simplications are evident in the Agda formalization:
+of replacing the cumbersome cardinality-based evidence that the type
+is not empty by the more intuitive evidence of choosing a value of the
+type. The simplications are evident in the Agda formalization:
 
@@
 
-Example Sec. 3.2 of
-\url{https://people.engr.ncsu.edu/hzhou/quantum_assert.pdf} with ancilla
-initialized to 1 to detected entanglement
+We illustrate the new style of ancilla management with an interesting
+example used in quantum circuits as a dynamic assertion of
+entanglement~\cite{DBLP:journals/cal/ZhouB19}:
+
+\begin{wrapfigure}{r}{0.2\textwidth}
+  \input{parity.tikz}
+\end{wrapfigure}  
+\PIFDparity{}
+
+\noindent The circuit exploits the additional expressiveness of
+choosing the initial value of the ancilla (and sets it to true). When
+applied to quantum inputs, the circuit detects whether the two qubits
+are entangled in the state
+$\frac{1}{\sqrt{2}}(\ket{00}+\ket{11})$. This property can be
+confirmed from the classical behavior: if the inputs are anything
+else, the ancilla bit is not restored to its original value (true) and
+the dynamic check fails.
 

### 2019-11-21 fracGC/p.tex
fix wrapfig
fix language about parity example

@@
 
-\begin{wrapfigure}{r}{0.2\textwidth}
+\begin{wrapfigure}[0]{r}{0.2\textwidth}
   \input{parity.tikz}
@@
 
-\noindent The circuit exploits the additional expressiveness of
-choosing the initial value of the ancilla (and sets it to true). When
-applied to quantum inputs, the circuit detects whether the two qubits
-are entangled in the state
-$\frac{1}{\sqrt{2}}(\ket{00}+\ket{11})$. This property can be
-confirmed from the classical behavior: if the inputs are anything
-else, the ancilla bit is not restored to its original value (true) and
-the dynamic check fails.
+\noindent The circuit initializes the ancilla to \textsf{true} and
+checks that it remains \textsf{true} after interacting with the two
+inputs. If either one of the inputs is \textsf{true} (but not both)
+the ancilla bit will be negated and the runtime check would fail. Thus
+the circuit can be used as a runtime assertion that the two inputs are
+equal. When applied to quantum inputs, the circuit detects whether the
+two qubits are entangled or not. By varying the initial value of the
+ancilla, it is possible to customize the dynamic assertion to check
+for particular entangled states~\cite{DBLP:journals/cal/ZhouB19}.
 

### 2019-11-21 fracGC/p.tex
minor cleanups; intro; and now ready to tacke space section

@@
 types would apply to general-purpose languages (after some
-adaptation), its natural technical definition is using symmetries
+adaptation), its natural technical definition exploits symmetries
 present in categorical model of type isomorphisms. We present such a
-definition in Sec.~\ref{sec:dyn}: that definition captures the main
-ideas of fractional types and partially solves a more general and more
-expressive instance of the ancilla problem. To make the memory
-management explicit, Sec.~\ref{sec:space} provides an operational
-semantics that exposes memory management details using a realization
-of fractional types as first-class garbage collectors. In
-Sec.~\ref{sec:dep} we refine the type system to keep track of ancillae
-values. Technically this requires the introduction of pointed types on
-top of which are defined both a monad and co-monad of singleton
-types. We conclude with a discussion of related work and a summary of
-our results.
+definition in Sec.~\ref{sec:dyn} which captures and illustrates the
+main novelties of fractional types and their runtime representations
+as first-class gc processes. This presentation partially solves a more
+general and more expressive instance of the ancilla problem as
+illustrated with examples. To make the arguments regarding memory
+management explicit, Sec.~\ref{sec:space} presents an abstract machine
+with a heap component that grows and shrinks as ancillae are allocated
+and de-allocated via fractional gc processes. In Sec.~\ref{sec:dep} we
+lift ``plain'' programs manipulating ancillae to the type system in
+order to exploit the power of dependent types to guarantee, for the
+first time, the condition necessary for safe de-allocation of
+ancillae. Technically this solution requires the introduction of
+pointed types on top of which are defined both a monad and co-monad of
+singleton types. The tracking of ancillae value is done within the
+type system via the singleton types. A number of examples exploiting
+the singleton-based type system are present to demonstrate the full
+expressiveness of fractional types. We furthermore demonstrate that,
+from the proofs of type safety, plain (but guaranteed to be safe)
+programs can be extracted and executing without errors or runtime
+checks in the conventional type system. The last section concludes
+with a discussion of related work and a summary of our results.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Space Operational Semantics}
+\section{Space-Tracking Abstract Machine}
 \label{sec:space}
-
-Positive rational numbers are a model. Apparently there is a
-categorification
-\url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
-
-Use all the constructions name, coname, etc. and see what they do in this context!
 
@@
 \PIMEMcard{}
+
 enumeration:
 \PIMEMenum{}
+
 combinator preserve size:
 \PIMEMcardeq{}
+
 semantics for Pi:
@@
 \PIMEMstep{}
+
 semantics for PiFracDyn:
@@
 
+Use all the constructions name, coname, etc. and see what they do in this context!
+
 allocate A
@@
 
-% General allocation and de-allocation. No need to keep track of
-% lifetime. Just make sure location is back to initial value. Safe to
-% de-allocate because indistinguishable from freshly created. No pointer
-% equality and aliases though. Concurrent: one thread allocates; gc
-% thread meets someone else
+Positive rational numbers are a model. Apparently there is a
+categorification
+\url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
+
+General allocation and de-allocation. No need to keep track of
+lifetime. Just make sure location is back to initial value. Safe to
+de-allocate because indistinguishable from freshly created. No pointer
+equality and aliases though. Concurrent: one thread allocates; gc
+thread meets someone else
 

### 2019-11-21 fracGC/p.tex
subtitle
plan for space section

@@
 \title{Fractional Types}
+\subtitle{Expressive and Safe Space Management for Ancilla Bits}
 \author{Chao-Hong Chen}
@@
 
-cardinality of pi type:
-\PIMEMcard{}
-
-enumeration:
-\PIMEMenum{}
-
-combinator preserve size:
-\PIMEMcardeq{}
-
-semantics for Pi:
-\PIMEMstate{}
-\PIMEMstep{}
+There are two main reasons we are calling the new type constructor a
+\emph{fractional type} and using the suggestive notation
+$\oneover{t}$:
+\begin{itemize}
+\item Categorical: ... explain and point to section with examples
+  showing 1/ab = 1/a  1/b etc.
+\item Operational ... this section; bits = log of size; fractional
+  needs negative bits; i.e. space collector; make this precise  
+\end{itemize}
+
+%% cardinality of pi type:
+%% \PIMEMcard{}
+%%   
+%% enumeration:
+%% \PIMEMenum{}
+%% 
+%%  combinator preserve size:
+%% \PIMEMcardeq{}
+%% 
+%%  semantics for Pi:
+%% \PIMEMstate{}
+%% \PIMEMstep{}
 

### 2019-11-21 fracGC/p.tex
example execution traces

@@
 
+%% for Sec. 4
 \input{latex/PiFracDynDefCode.tex}
 \input{latex/PiFracDynCode.tex}
-\input{latex/BooleanCircuitsCode.tex}
+
+%% for Sec. 5
+%% \input{latex/PiMemSemCode.tex}
+\input{latex/PiFracMemSemCode.tex}
+
+%% for Sec. 6
+\input{latex/J.tex}
 \input{latex/PiPointedFracCode.tex}
 \input{latex/ExtractionCode.tex}
-\input{latex/PiMemSemCode.tex}
-\input{latex/PiFracMemSemCode.tex}
-\input{latex/J.tex}
+\input{latex/BooleanCircuitsCode.tex}
 
@@
 
-There are two main reasons we are calling the new type constructor a
-\emph{fractional type} and using the suggestive notation
+There are two technical reasons we are calling the new type constructor a
+\emph{fractional type} and hence using for it the suggestive notation
 $\oneover{t}$:
 \begin{itemize}
-\item Categorical: ... explain and point to section with examples
-  showing 1/ab = 1/a  1/b etc.
-\item Operational ... this section; bits = log of size; fractional
-  needs negative bits; i.e. space collector; make this precise  
+\item Categorical: The categorical model for the plain $\Pi$ language
+  of type isomorphisms is a category with two symmetric monoidal
+  structures (additive and multiplicative) with the multiplicative
+  fragment distributing over the additive one~\cite{XXX}. Generally,
+  a symmetric monoidal category is \emph{compact closed} if every
+  object has a dual object whose behavior is governed by our $\eta$
+  and $\epsilon$ operators~\cite{XXX}. Our language from the previous
+  section is an adaptation of this idea to the context of $\Pi$. Many
+  programming abstractions (such as \emph{name}, \emph{co-name}, and
+  \emph{trace} illustrated in Sec.~\ref{sec:cat}) can be automatically
+  derived from properties of compact closed categories. Additionally,
+  all the properties expected of the rational numbers
+  (e.g. $\oneover{(m*n)} = \oneover{m}*\oneover{n}$,
+  $\oneover{(\oneover{n})}=n$, etc.) can be given constructive content
+  corresponding to manipulation of gc processes. We have already seen
+  the first example and will see more in Sec.~\ref{sec:cat}.
+
+\item Operational: The other reason our types can be justifiably
+  called fractional is information theoretic. As mentioned in the
+  introduction the number of bits needed to store information about a
+  space with $n$ distinct values is $\log{n}$. If the number of bits
+  can be thought of as a debt (to be reclaimed) then a negative number
+  of bits corresponds to a space with $\oneover{n}$ distinct
+  values. Indeed we interpret these values as gc processes each
+  specialized to collect space for one of the $n$ distinct values. In
+  this section, we make this intuition precise by presenting a
+  (fully-formalized in Agda) abstract machine that manages an explicit
+  heap. 
 \end{itemize}
@@
 %% \PIMEMstep{}
+
+CNOT trace in plain pi:
+
+\includegraphics[scale=0.4]{trace1.png}
+
+\PIFMEMsize{}
+
+If a type has size $n$ we will use a number $0..(n-1)$ to represent
+the values of the types. In binary, this would use $\log{n}$ bits.
+
+The abstract machine has a register representing the code and a
+pointer. If the code is acting on a type with $n$ values, the pointer
+has $\log{n}$ bits (rounded).
+
+In plain pi, if the input is representable using $n$ bits then this is
+the only memory you get to use throughout the entire circuit. The
+circuit just transforms these $n$ bits; example.
+
+In frac pi, you are however allowed to use more than $n$ bits in the
+intermediate steps as eta is used to allocate more space.
+
+Use Fin size as proxy for memory size
+
+zigzag trace
+
+\includegraphics[scale=0.5]{trace2.png}
 

### 2019-11-21 fracGC/p.tex
outline of space section done; now polish
thought about trace

@@
 %% for Sec. 5
-%% \input{latex/PiMemSemCode.tex}
+\input{latex/PiMemSemCode.tex}
 \input{latex/PiFracMemSemCode.tex}
@@
 
+%%
+\subsection{Abstract Machine without Fractionals}
+
+If a type has size $n$ we will use a number $0..(n-1)$ to represent
+the values of the types. In binary, this would use $\log{n}$ bits.
+
+The abstract machine has a register representing the code and a
+pointer. If the code is acting on a type with $n$ values, the pointer
+has $\log{n}$ bits (rounded).
+
+In plain pi, if the input is representable using $n$ bits then this is
+the only memory you get to use throughout the entire circuit. The
+circuit just transforms these $n$ bits; example.
+
+Use Fin size as proxy for memory size
+
+\PIFMEMsize{}
+
+\PIMEMcardeq{}
+
+\PIMEMstep{}
+
+CNOT trace in plain pi:
+
+\includegraphics[scale=0.4]{trace1.png}
+
 %% cardinality of pi type:
@@
 
-CNOT trace in plain pi:
-
-\includegraphics[scale=0.4]{trace1.png}
-
-\PIFMEMsize{}
-
-If a type has size $n$ we will use a number $0..(n-1)$ to represent
-the values of the types. In binary, this would use $\log{n}$ bits.
-
-The abstract machine has a register representing the code and a
-pointer. If the code is acting on a type with $n$ values, the pointer
-has $\log{n}$ bits (rounded).
-
-In plain pi, if the input is representable using $n$ bits then this is
-the only memory you get to use throughout the entire circuit. The
-circuit just transforms these $n$ bits; example.
+%%
+\subsection{Abstract Machine with Fractionals}
 
@@
 
-Use Fin size as proxy for memory size
+\PIFMEMstep{}
 
@@
 
-semantics for PiFracDyn:
-\PIFMEMstate{}
-\PIFMEMstep{}
-
-Evaluation:
-\PIPFeval{}
-
-Combine with eval, the whole thing is reversible:
-\PIPFrev{}
+%% semantics for PiFracDyn:
+%% \PIFMEMstate{}
+%% \PIFMEMstep{}
 
@@
 \PIPFCombDef{}
+
+Evaluation:
+\PIPFeval{}
+
+Combine with eval, the whole thing is reversible:
+\PIPFrev{}
 
@@
 
+About trace, it is remarkable: it says, given a function (possibly
+using eta/epsilon so input/output sizes not guaranteed to be the same)
+\verb|AxC <-> BxC| hence using \verb|AxC| space for the input and
+\verb|BxC| space for output, we can 'trace out' the space used for
+\verb|C| and derive a function \verb|A <-> B|. This is true \emph{even
+  if} \verb|C| is actually used. See the swap example.
+
 Use all the constructions name, coname, etc. and see what they do in this context!

### 2019-11-21 fracGC/p.tex
done with space section for now

@@
 \noindent where the operations take an implicit argument asserting the
-given type is non-empty. The operational semantics of the language
-interprets each combinator (except $\eta$ and $\epsilon$) of type
-$t_1 \leftrightarrow t_2$ as a (reversible) function mapping values of
-type $t_1$ to values of type $t_2$ (and vice-versa). Extending this
-interpreter to handle $\eta$ and $\epsilon$ requires at this point
+given type is non-empty. A \emph{big step operational semantics} for
+the language interprets each combinator (except $\eta$ and $\epsilon$)
+of type $t_1 \leftrightarrow t_2$ as a (reversible) function mapping
+values of type $t_1$ to values of type $t_2$ (and
+vice-versa).\footnote{In Sec.~\ref{sec:space} we will define a small
+  step abstract machine that exposes the space used at each step.}
+Extending this interpreter to handle $\eta$ and $\epsilon$ requires
 that the interpreter returns either a value of the appropriate or an
@@
 \subsection{Examples}
+\label{sub:id}
 
@@
 
-If a type has size $n$ we will use a number $0..(n-1)$ to represent
-the values of the types. In binary, this would use $\log{n}$ bits.
-
-The abstract machine has a register representing the code and a
-pointer. If the code is acting on a type with $n$ values, the pointer
-has $\log{n}$ bits (rounded).
-
-In plain pi, if the input is representable using $n$ bits then this is
-the only memory you get to use throughout the entire circuit. The
-circuit just transforms these $n$ bits; example.
-
-Use Fin size as proxy for memory size
+Before before the abstract machine for the full system with
+fractionals, we introduce some of the common technical details in the
+simpler setting of the core $\Pi$ language. First two simple facts
+about cardinalities:
 
@@
 
+\vspace{-\baselineskip}
+
 \PIMEMcardeq{}
 
+The cardinalities of the types (other than the fractional type) map
+directly to the natural numbers. An easy proposition to verify is that
+every combinator must be between types of the same cardinalities;
+indeed one of the standard interpretations of $\Pi$ combinators is
+that they are permutations on finite sets, i.e., they just shuffle
+elements around without changing the sizes of the underlying sets.
+
+To make the space usage explicit in every step of evaluation, we
+define an abstract machine whose states are pairs of a code register
+and a memory register. In a low-level implementation the code register
+would contain a pointer to the current instruction and the memory
+register would contain a bit-level runtime representation of values
+with one bit for boolean values, etc. In the Agda formalization, we
+include the current combinator in the code register, and represent the
+memory register using an index into an enumeration. Thus a memory
+register containing a boolean value would be represented in Agda as 𝔽
+::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether
+the values is \textsf{false} or \textsf{true}. The prepending list
+contains all possible values of the type and the number in between [\ldots ] 
+is an index into the list, representing a finite number that
+is guaranteed to be in bounds. In a low-level runtime representation
+only a binary encoding of this finite would need to be stored but in
+order to just prove that space is preserved, we can just inspect the
+size of the prepended list. Putting it all together, machine states and
+small-step reductions are formalized as follows:
+
 \PIMEMstep{}
 
-CNOT trace in plain pi:
+The constructor for machine states takes a number which is the size of
+the memory list. This size is maintained throughout the execution,
+even if the intermediate types change. As a small example, the
+following is a trace of the execution of \AgdaFunction{CNOT} applied
+to 𝕋, 𝕋:
 
 \includegraphics[scale=0.4]{trace1.png}
+
+Note how the size of the list remains equal to 4 throughout the
+execution, even if the individual elements change as types change and
+the current index (i.e., the current value) changes.
 
@@
 
-In frac pi, you are however allowed to use more than $n$ bits in the
-intermediate steps as eta is used to allocate more space.
+Using the same assumptions as before, machine states and small step
+reductions are formalized as follows:
 
@@
 
-zigzag trace
+The main difference is that now the state constructor does not take a
+fixed size and indeed the size changes during execution as appropriate
+to fit the elements of the type $A$. In particular, since fractional
+values have a trivial representation, the size of a fractional value
+is 1, and hence $\eta$ used at type $A$ changes the size of the memory
+list by the size of $A$. As an illustrative example, we include a
+trace of the execution of \AgdaFunction{id'} from Sec.~\ref{sub:id}
+applied to 𝕋:
 

### 2019-11-21 fracGC/p.tex
fix typo

@@
 
-Before before the abstract machine for the full system with
+Before the abstract machine for the full system with
 fractionals, we introduce some of the common technical details in the
@@
 
-\PIFMEMsize{}
+\PIFMEMsize{}  % maybe we should use \PIMEMcard here?
 

### 2019-11-21 fracGC/p.tex
before merge
sec 6.1

@@
     types}. As a simple intuitive example, allocation of a new boolean
-  value initialized to \textsf{false} also creates a value
-  $\oneover{\textsf{false}}$ that can be thought of as a garbage
+  value initialized to \AgdaInductiveConstructor{false} also creates a value
+  $\oneover{\AgdaInductiveConstructor{false}}$ that can be thought of as a garbage
   collection (gc) process specialized to reclaim, and only reclaim,
-  storage containing the value $\textsf{false}$. This gc process is a
+  storage containing the value $\AgdaInductiveConstructor{false}$. This gc process is a
   first-class entity that can sliced and diced by combining it with
@@
 
-\noindent The circuit initializes the ancilla to \textsf{true} and
-checks that it remains \textsf{true} after interacting with the two
-inputs. If either one of the inputs is \textsf{true} (but not both)
+\noindent The circuit initializes the ancilla to \AgdaInductiveConstructor{true} and
+checks that it remains \AgdaInductiveConstructor{true} after interacting with the two
+inputs. If either one of the inputs is \AgdaInductiveConstructor{true} (but not both)
 the ancilla bit will be negated and the runtime check would fail. Thus
@@
 ::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether
-the values is \textsf{false} or \textsf{true}. The prepending list
+the values is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}. The prepending list
 contains all possible values of the type and the number in between [\ldots ] 
@@
 By lifting the scoping restriction, the development in the previous
-section is already more general than the state of the art in ancilla
+sections is already more general than the state of the art in ancilla
 management.  It however shares the same limitation of requiring a
@@
 \Jexample{}
+
+The toy language has two types (natural numbers and booleans) and two
+functions (and their compositions). Say we wanted to prove that
+\AgdaInductiveConstructor{compose isZero square} always returns
+\AgdaInductiveConstructor{false} when applies to a non-zero natural
+number. We can certainly do this proof in Agda (i.e., in the
+meta-language of our formalization) but we would like to do the proof
+within the framework of the toy language itself. The most important
+reason is that this proof can be now be used within the language to
+optimize programs (or as the case will be for the full language to
+remove a runtime check).
+
+The strategy we adopt is to create a lifted version of the toy
+language with \emph{pointed types}~\cite{XXX}, i.e., types paired with
+a value of the type. In such the lifted language, the evaluation
+function has an interesting type: it keeps track of the result of
+evaluation within the type:
+
+\Jexamplecont{}
+
+Now various properties of our small example
+\AgdaInductiveConstructor{compose isZero square} can be derived within
+the extended type system for the toy language:
+
+\Jexampletest{}
+
+Evidently we can track concrete values as the first two tests
+show. What is more interesting is \AgdaFunction{test3} which shows a
+property that holds for all natural numbers $n$. The full evaluation
+was not used but only partially as needed to derive the property. This
+form of partial evaluation will prove to be quite expressive allowing
+us to keep track of ancillae values throughout complex programs as
+demonstrated in Sec.~\ref{sec:cat}.
+
+%%%%% 
+\subsection{Pointed Types}
 
@@
 unfocus using (comonad). (This by the way ensures we don't write 1/0
-which woudl be bad).
-
-%%%%% 
-\subsection{Pointed Types}
+which would be bad).
 
@@
 
+revrev: rematerialization ! gc the gc process!
+
 Use all the constructions name, coname, etc. and see what they do in this context!

### 2019-11-22 fracGC/p.tex
fix symbol
before merge
plan for 6.2

@@
 \newunicodechar{ᵣ}{$_r$}
+\newunicodechar{❰}{$\llparenthesis$}
+\newunicodechar{❱}{$\rrparenthesis$}
 
@@
 %%%%% 
-\subsection{Pointed Types}
-
- rationale: we want singleton types to track, in the type system
-the values of ancillae, i.e.. we construct a proof using the type
-system. Technically the singleton types allow you to focus on one
-value by reflecting it in the type system. Now this type can only be
-equal to another type if and only if the other type is the same
-singleton value.
-
-Once singleton types are built, we need to make sure circuit
-evaluation is also reflected in the type system. So in fact
-computation itself is reflected in the type system. In order to mix
-singleton types with other types, we use pointed types as a unifying
-substrate. Now all types have values; we can focus (using monad) and
-unfocus using (comonad). (This by the way ensures we don't write 1/0
-which would be bad).
-
-%%%%% 
-\subsection{Singleton Types}
-
-Consider the type $\Sigma \mathbb{N} (\lambda n. n \equiv 5)$. Values
-of that type are a number $m$ and a proof that this is number is equal
-to 5. So we can have $(4 + 1, \textit{refl})$, $(2 + 3 ,
-\textit{refl})$, and so on. All these syntactically different values
-are identical to 5 and hence we can \emph{contract} the type to a
-single point at 5.
-
-PiPointedFrac:
-lift pi
-lift multiplicative structure
-monad/comonad for singletons
-eta/epsilon
+\subsection{Pointed and Singleton Types}
 
@@
 
-RECIP
-
 Combinators:
+
 \PIPFCombDef{}
+
+properties of singleton, recip  
+derive join etc  
+monad,comonad structure
 
@@
 \PIPFrev{}
+
+show a small example
 

### 2019-11-22 fracGC/p.tex
edit abstract, section 1-2.

@@
 \usepackage{bbold}
+\usepackage[show]{ed}
 \usepackage{tikz}
@@
   storage containing the value $\AgdaInductiveConstructor{false}$. This gc process is a
-  first-class entity that can sliced and diced by combining it with
+  first-class entity that can be sliced and diced by combining it with
   other gc processes and decomposing it in smaller
@@
 
-Instead of using the rather low-level circuit model of computation,
+Instead of using the low-level circuit model of computation,
 one can leverage the full force of type theory and category theory by
@@
 gate is not
-universal~\cite{DBLP:conf/innovations/AaronsonGS17,DBLP:journals/corr/Xu15e}
-and as would be expected the more temporary storage is allowed, the
-more efficient certain computations could become.
+universal~\cite{DBLP:conf/innovations/AaronsonGS17,DBLP:journals/corr/Xu15e}.
+And, as would be expected, the more temporary storage is allowed, the
+more efficient certain computations can become.
 
@@
 safety condition. Violating this condition destroys the symmetry
-between input and output making the circuits not reversible and, in
+between input and output, making the circuits irreversible and, in
 the quantum model, causes decoherence that may destroy the quantum
 state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
-number of critical applications and yet are poorly supported in
+number of critical applications, yet are poorly supported in
 current reversible and quantum programming languages making them a
@@
 \paragraph*{Conservation of Information and Negative Entropy.}  
-According to the conventional theory of
-communication~\cite{Shannon1948}, a type with $N$ values is viewed as
-an abstract system that has $N$ distinguishable states where the
+The conventional theory of
+communication~\cite{Shannon1948} views a type with $N$ values as
+an abstract system that has $N$ distinguishable states; the
 amount of information contained in each state is $\log{N}$. This
@@
 of this type. Thus a type with 8 elements needs 3 bits of memory for
-storage or 3 bits of bandwidth for communication. The logarithmic map
-implies that information contained in a composite state is the sum of
+storage or 3 bits of bandwidth for communication. The information
+information contained in a composite state, by definition of the
+logarithm, is the sum of
 the information contained in its constituents. For example, the type
@@
 unrelated subsystems.  Each state of the composite system therefore
-contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
+contains $\log{(2*3)} = \log{2} + \log{3}$ bits, the sum of
 the information contained in each subsystem. If we could imagine a
@@
 linearly used, i.e., they cannot be duplicated or erased: this
-property is however automatically enforced in a reversible
+property is automatically enforced in a reversible
 computational model.
 
-\paragraph*{Outline.} The paper solves the ancilla problem in
-reversible and quantum computation using a novel concept:
-\emph{fractional types}. In the next section, we introduce the problem
-of ancilla, motivate its importance, and explain the limitations of
-current solutions. In Sec.~\ref{sec:pi}, we review existing work by
-James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
-introduces a reversible programming language built using isomorphism
-of types. Although it is conceivable that the concept of fractional
-types would apply to general-purpose languages (after some
+\paragraph*{Outline.} We solve the ancilla problem in reversible and quantum
+computation using a novel concept: \emph{fractional types}. In the next
+section, we introduce the problem of ancilla, motivate its importance, and
+explain the limitations of current solutions. In Sec.~\ref{sec:pi}, we review
+existing work%
+~\cite{rc2011,DBLP:conf/esop/CaretteS16,rc2012,James:2012:IE:2103656.2103667}
+that introduces a reversible programming language built using type isomorphism.
+Although the concept of fractional
+types might apply to general-purpose languages (after some
 adaptation), its natural technical definition exploits symmetries
-present in categorical model of type isomorphisms. We present such a
+present in the categorical model of type isomorphisms. We present such a
 definition in Sec.~\ref{sec:dyn} which captures and illustrates the
@@
 general and more expressive instance of the ancilla problem as
-illustrated with examples. To make the arguments regarding memory
+illustrated with examples. 
+\ednote{JC: I don't understand the previous sentence. What does
+``this presentation'' refer to?  The paper as a whole, or the 
+details of the fractional types? There is also a tension between
+``partially solves'' and ``more general and more expressive'' which
+might very well be true, but makes for a very vague sentence.}
+To make the arguments regarding memory
 management explicit, Sec.~\ref{sec:space} presents an abstract machine
 with a heap component that grows and shrinks as ancillae are allocated
-and de-allocated via fractional gc processes. In Sec.~\ref{sec:dep} we
+and de-allocated via garbage collection. In Sec.~\ref{sec:dep} we
 lift ``plain'' programs manipulating ancillae to the type system in
@@
 ancillae. Technically this solution requires the introduction of
-pointed types on top of which are defined both a monad and co-monad of
+pointed types on top of which are defined both a monad and comonad of
 singleton types. The tracking of ancillae value is done within the
 type system via the singleton types. A number of examples exploiting
-the singleton-based type system are present to demonstrate the full
-expressiveness of fractional types. We furthermore demonstrate that,
-from the proofs of type safety, plain (but guaranteed to be safe)
-programs can be extracted and executing without errors or runtime
+the singleton-based type system are presented to demonstrate the full
+expressiveness of fractional types. The type safety proofs demonstrate
+that plain (but guaranteed safe)
+programs can be extracted and executed without errors or runtime
 checks in the conventional type system. The last section concludes
@@
 \[\begin{array}{l}
-                    \frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
-\\
-\Rightarrow  \textrm{phase kickback} \\
-\\
--\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
-\\
-\Rightarrow  \textrm{diffusion} \\
-\\
- (\frac{1}{2}-(-\frac{1}{2}))\ket{a}+(\frac{1}{2}-\frac{1}{2})\ket{b}+(\frac{1}{2}-\frac{1}{2})\ket{c}+(\frac{1}{2}-\frac{1}{2})\ket{d})\\
-\\
-\Rightarrow  \textrm{simplification} \\
-\\
+\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\ [1.5ex]
+\qquad\Rightarrow  \textrm{phase kickback} \\ [1.5ex]
+-\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\ [1.5ex]
+\qquad\Rightarrow  \textrm{diffusion} \\ [1.5ex]
+ (\frac{1}{2}-(-\frac{1}{2}))\ket{a}+(\frac{1}{2}-\frac{1}{2})\ket{b}+(\frac{1}{2}-\frac{1}{2})\ket{c}+(\frac{1}{2}-\frac{1}{2})\ket{d})\\ [1.5ex]
+\qquad\Rightarrow  \textrm{simplification} \\ [1.5ex]
 \ket{a}
@@
 current element is equal to $\ket{a}$ by performing a measurement would
-destroy the quantum superposition defeating the entire algorithm.
-
-Here is where ancilla qubits come to the rescue. In the words of
-Matuschak and Nielsen~\cite{howgrover}, the idea is the following:
+destroy the quantum superposition, defeating the entire algorithm.
+
+Ancilla bits offer a solution.  In the words of
+Matuschak and Nielsen~\cite{howgrover}, the idea is as follows:
 \begin{quote}
@@
 \end{quote}
-
-\paragraph*{Quantum Error Correction.} The main requirement for
-quantum error correction is the ability to replicate the state of a
-qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is however
-impossible to do directly by the no-cloning theorem~\cite{XXX}. Again ancilla
-qubits come to the rescue: because ancilla qubits start in a known
+\ednote{Would it make sense to explicitly mention CNOT here?}
+
+\paragraph*{Quantum Error Correction.} 
+Quantum error correction requires the ability to replicate the state of a
+qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is
+impossible to do directly, by the no-cloning theorem~\cite{XXX}. Again ancilla
+qubits solve the problem: because they start in a known
 initial state, it is possible to create a simple circuit that makes
@@
 \begin{quote}
-consider a 3-qubit quantum majority code in which a logical ``0'' is
+Consider a 3-qubit quantum majority code in which a logical ``0'' is
 encoded as ``000'' and a logical ``1'' is encoded as ``111.''  Just as with
@@
 Despite its fundamental importance, the management of ancillae is
-poorly supported in current languages, simulators, and toolkits as
-demonstrated in a recent analysis of bugs in quantum
-benchmarks~\cite{DBLP:conf/oopsla/HuangM18}. The analysis reveals that
-lack of language support in the management of ancillae leads to
+poorly supported in current languages, simulators, and toolkits. 
+A recent analysis of bugs in quantum
+benchmarks~\cite{DBLP:conf/oopsla/HuangM18} demonstrates this. The analysis
+reveals that lack of language support in the management of ancillae leads to
 several problems. For example: 

### 2019-11-22 fracGC/p.tex
Section 3 and trailing blanks.

@@
 \swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
-\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
-\\
+\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\ [1.5ex]
 \identlt :&  1 * \tau & \iso & \tau &: \identrt \\
 \swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
-\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
-\\
+\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\ [1.5ex]
 \distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
@@
 \end{array}\]
-\vspace{2cm}
+%\vspace{2cm}
 \begin{center}
@@
 programming language to a quantum one requires just the management of
-superpositions, for example starting from the classical reversible
+superpositions. For example starting from the classical reversible
 language Theseus~\cite{james2014theseus}, a small extension produces a quantum
-language with quantum control~\cite{10.1007/978-3-319-89366-2_19}. The
-language Theseus~\cite{james2014theseus} mentioned above is a language that uses
-convenient syntactic extensions over an more basic language $\Pi$ of
-combinators witnessing isomorphisms over finite types~\cite{XXX}. The
+language with quantum control~\cite{10.1007/978-3-319-89366-2_19}.
+Theseus~\cite{james2014theseus} is a language that layers
+convenient syntactic extensions over an even more basic language $\Pi$ of
+combinators witnessing isomorphisms over finite
+types~\cite{DBLP:conf/esop/CaretteS16}. The
 combinator-based language has the advantage of being more amenable to
@@
 ancillae is inspired by the construction of \emph{compact closed
-  categories}~\cite{XXX}. These categories extend the monoidal
+categories}~\cite{kelly1972many}. These categories extend the monoidal
 categories~\cite{XXX} which are used to model many resource-aware
-programming languages~\cite{XXX} (including $\Pi$) with a new type
+programming languages~\cite{XXX}
+\ednote{The search on
+"monoidal category" "resource-aware" "programming language" on Google
+Scholar produces just 6 papers!}
+(including $\Pi$) with a new type
 constructor that creates duals or inverses to existing types. This
@@
 \[\begin{array}{lrcl}
-\textit{Value types} & t &::=& 0 \alt 1 \alt t+t \alt t*t \\
+\textit{Value types} & t &::=& 0 \alt 1 \alt t+t \alt t\times t \\
 \textit{Values}      & v &::=& \star \alt \inlv{v} \alt \inrv{v} \alt (v,v) \\
@@
 capturing space-time tradeoffs~\cite{superstructural}. Categorically,
-they form a \emph{bimonoidal category}~\cite{laplaza72}.
+they form a \emph{distributive bimonoidal category}~\cite{laplaza72}.
 
@@
 values and structured trees of values and the isomorphisms act on such
-conventional data structures. Universal computation models however
+conventional data. Universal computation models however
 fundamentally rely on the fact that \emph{programs are (or can be
@@
 
+\begin{figure}[htb]
 \begin{center}
@@
 \end{center}
+\end{figure}
 
@@
 
-The code below defines the types corresponding to bits (booleans),
+Below we show code that defines types corresponding to bits (booleans),
 two-bits, three-bits, and four-bits. It then defines an operator that
@@
 unreachable from even permutations. This function will however be
-definable in the next section once we introduce ancilla bits.
+definable once we introduce ancilla bits in the next section.
 

### 2019-11-22 fracGC/p.tex
section 4.

@@
 Computationally, this extra generality is not a gratuitous concern:
-since scope is fundamentally a static property of programs, it does
+since scope is a \emhp{static property} of programs, it does
 not allow the flexibility of heap allocation in which the lifetime of
@@
 \[
-\eta_A : 1 \leftrightarrow A * \oneover{A} : \epsilon_A
+\eta_A : 1 \leftrightarrow A \times \oneover{A} : \epsilon_A
 \]
@@
 \[\begin{array}{rcll}
-1 &=& 0 * 1/0 & \textrm{by~} \eta_0 \\
+1 &=& 0 \times 1/0 & \textrm{by~} \eta_0 \\
 &=& 0 & \textrm{by~} \absorbr
@@
 
-In the Agda formalization, the types of $\eta$ and $\epsilon$ are as follows:
-
+In the Agda formalization, the types of $\eta$ and $\epsilon$ are:
 \EtaEpsilon{}
-
 \noindent where the operations take an implicit argument asserting the
@@
 Extending this interpreter to handle $\eta$ and $\epsilon$ requires
-that the interpreter returns either a value of the appropriate or an
+that the interpreter returns either a value of the appropriate type or an
 exception:
@@
 
-When ancillae are created they are created with a specific default
+When ancillae are created, they are created with a specific default
 value depending on the type. When the gc process encounters a value to
@@
 The previous development achieves a significant milestone but only
-hints at the full expressiveness of fractional types. A small but
-significant additional expressiveness is to allow ancillae to be
+hints at the full expressiveness of fractional types. A small addition
+that significantly increases expressiveness is to allow ancillae to be
 initialized to a value of the programmer's choice, not just a globally
@@
   $\frac{\ket{0}-\ket{1}}{\sqrt{2}}$ is used.} This can be easily
-achieved in our language as follows:
+achieved by adapting our language as follows:
 

### 2019-11-22 fracGC/p.tex
fix error

@@
 Computationally, this extra generality is not a gratuitous concern:
-since scope is a \emhp{static property} of programs, it does
+since scope is a \emph{static property} of programs, it does
 not allow the flexibility of heap allocation in which the lifetime of

### 2019-11-22 fracGC/p.tex
from section 5 to end

@@
   programming abstractions (such as \emph{name}, \emph{co-name}, and
-  \emph{trace} illustrated in Sec.~\ref{sec:cat}) can be automatically
+  \emph{trace} illustrated in Sec.~\ref{sec:cat}) can be
   derived from properties of compact closed categories. Additionally,
@@
 
-Before the abstract machine for the full system with
-fractionals, we introduce some of the common technical details in the
-simpler setting of the core $\Pi$ language. First two simple facts
+To start, we introduce some of the common technical details in the
+simpler setting of the core $\Pi$ language. First, simple facts
 about cardinalities:
@@
 register containing a boolean value would be represented in Agda as 𝔽
-::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether
-the values is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}. The prepending list
+::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether the
+values is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}.
+The prepended list
 contains all possible values of the type and the number in between [\ldots ]
@@
 is guaranteed to be in bounds. In a low-level runtime representation
-only a binary encoding of this finite would need to be stored but in
+only a binary encoding of this finite type would need to be stored, but in
 order to just prove that space is preserved, we can just inspect the
@@
 
-\includegraphics[scale=0.4]{trace1.png}
+\includegraphics[scale=0.5]{trace1.png}
 
@@
 language with \emph{pointed types}~\cite{XXX}, i.e., types paired with
-a value of the type. In such the lifted language, the evaluation
+a value of the type. In the lifted language, the evaluation
 function has an interesting type: it keeps track of the result of
@@
 demonstrated in Sec.~\ref{sec:cat}.
+\ednote{JC: Doesn't \AgdaFunction{eval}, in the type, evaluate
+fully for this work?  That Agda is strongly normalizing is rather
+important here. What exactly is partially evaluating is really not
+clear (says an expert in PE). I think the idea being expressed is
+correct, but the paragraph is insufficiently clear/precise.}
 
@@
 First we can inject plain Pi type and value into PiFracDyn:
+\ednote{Pi/.U is not defined before its use below}
 \INJU{}

### 2019-11-22 fracGC/p.tex
ednotes, toffoli figure, etc

@@
 \newcommand{\proves}{\vdash}
+\newcommand{\Afun}[1]{\AgdaFunction{#1}}
+\newcommand{\Acon}[1]{\AgdaInductiveConstructor{#1}}
+\newcommand{\Avar}[1]{\AgdaBound{#1}}
 \newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
@@
 fact, classical reversible computations are often used as
-``subroutines'' of quantum computations. Indeed classical reversible
+``subroutines'' of quantum computations, i.e., classical reversible
 circuits can be applied to either classical inputs or to quantum
@@
 the Toffoli gate for classical reversible computing and of the
-universality of the Toffoli-Hadamard gates for quantum computing should
+universality of the Toffoli-Hadamard gates for quantum computing come
+with a catch and should
 not distract from efficiency and safety concerns. The theorems proving
@@
 state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
-number of critical applications, yet are poorly supported in
+number of other critical applications, yet are poorly supported in
 current reversible and quantum programming languages making them a
@@
 
-\paragraph*{Outline.} We solve the ancilla problem in reversible and quantum
-computation using a novel concept: \emph{fractional types}. In the next
-section, we introduce the problem of ancilla, motivate its importance, and
-explain the limitations of current solutions. In Sec.~\ref{sec:pi}, we review
-existing work%
+\paragraph*{Outline.} We solve the ancilla problem in reversible and
+quantum computation using a novel concept: \emph{fractional types}. In
+the next section, we introduce the problem of ancilla, motivate its
+importance, and explain the limitations of current solutions. Inm
+Sec.~\ref{sec:pi}, we review existing work%
 ~\cite{rc2011,DBLP:conf/esop/CaretteS16,rc2012,James:2012:IE:2103656.2103667}
-that introduces a reversible programming language built using type isomorphism.
-Although the concept of fractional
-types might apply to general-purpose languages (after some
-adaptation), its natural technical definition exploits symmetries
-present in the categorical model of type isomorphisms. We present such a
-definition in Sec.~\ref{sec:dyn} which captures and illustrates the
-main novelties of fractional types and their runtime representations
-as first-class gc processes. This presentation partially solves a more
-general and more expressive instance of the ancilla problem as
-illustrated with examples.
-\ednote{JC: I don't understand the previous sentence. What does
-``this presentation'' refer to?  The paper as a whole, or the
-details of the fractional types? There is also a tension between
-``partially solves'' and ``more general and more expressive'' which
-might very well be true, but makes for a very vague sentence.}
+that introduces a reversible programming language built using type
+isomorphism.  Although the concept of fractional types might apply to
+general-purpose languages (after some adaptation), its natural
+technical definition exploits symmetries present in the categorical
+model of type isomorphisms. We present such a definition in
+Sec.~\ref{sec:dyn} which captures and illustrates the main novelties
+of fractional types and their runtime representations as first-class
+gc processes. This definition allows allocation and de-allocation of
+ancillae in patterns beyond the scope model but, like existing
+stack-based solutions, still requires a runtime check to verify the
+safety of de-allocation.
+%% \ednote{JC: I don't understand the previous sentence. What does
+%% ``this presentation'' refer to?  The paper as a whole, or the
+%% details of the fractional types? There is also a tension between
+%% ``partially solves'' and ``more general and more expressive'' which
+%% might very well be true, but makes for a very vague sentence.}
 To make the arguments regarding memory
@@
 the singleton-based type system are presented to demonstrate the full
-expressiveness of fractional types. The type safety proofs demonstrate
+expressiveness of reasoning about fractional types. The type safety proofs demonstrate
 that plain (but guaranteed safe)
@@
 \end{quote}
-\ednote{Would it make sense to explicitly mention CNOT here?}
+%% \ednote{Would it make sense to explicitly mention CNOT here?}
 
@@
 \assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\ [1.5ex]
-\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
-\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
-\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\ [1.5ex]
-\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
-\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
+\identlt :&  1 \times \tau & \iso & \tau &: \identrt \\
+\swapt :&  \tau_1 \times \tau_2 & \iso & \tau_2 \times \tau_1 &: \swapt \\
+\assoclt :&  \tau_1 \times (\tau_2 \times \tau_3) & \iso & (\tau_1 \times\tau_2) \times \tau_3 &: \assocrt \\ [1.5ex]
+\distz :&~ 0 \times \tau & \iso & 0 ~ &: \factorzl \\
+\dist :&~ (\tau_1 + \tau_2) \times \tau_3 & \iso & (\tau_1 \times \tau_3) + (\tau_2 \times \tau_3)~ &: \factor
 \end{array}\]
@@
 {\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
-{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
+{\jdg{}{}{c_1 \otimes c_2 : \tau_1 \times \tau_3 \iso \tau_2 \times \tau_4}}
 {}
@@
 superpositions. For example starting from the classical reversible
-language Theseus~\cite{james2014theseus}, a small extension produces a quantum
-language with quantum control~\cite{10.1007/978-3-319-89366-2_19}.
-Theseus~\cite{james2014theseus} is a language that layers
-convenient syntactic extensions over an even more basic language $\Pi$ of
+language Theseus~\cite{james2014theseus}, a small extension produces a
+quantum language with quantum
+control~\cite{10.1007/978-3-319-89366-2_19}.
+Theseus~\cite{james2014theseus} is a language that layers convenient
+syntactic extensions over an even more basic language $\Pi$ of
 combinators witnessing isomorphisms over finite
-types~\cite{DBLP:conf/esop/CaretteS16}. The
-combinator-based language has the advantage of being more amenable to
-formal analysis for at least two reasons: (i) it is conceptually
-simple and small, and (ii) it has direct and evident connections to
-type theory and category theory. Indeed our solution for managing
-ancillae is inspired by the construction of \emph{compact closed
-categories}~\cite{kelly1972many}. These categories extend the monoidal
-categories~\cite{XXX} which are used to model many resource-aware
-programming languages~\cite{XXX}
-\ednote{The search on
-"monoidal category" "resource-aware" "programming language" on Google
-Scholar produces just 6 papers!}
+types~\cite{DBLP:conf/esop/CaretteS16}. The combinator-based language
+has the advantage of being more amenable to formal analysis for at
+least two reasons: (i) it is conceptually simple and small, and (ii)
+it has direct and evident connections to type theory and category
+theory. Indeed our solution for managing ancillae is inspired by the
+construction of \emph{compact closed
+  categories}~\cite{kelly1972many}. These categories extend the
+monoidal categories~\cite{XXX} which are used to model many
+resource-aware (e.g., based on linear types) programming
+languages~\cite{XXX}
+%% \ednote{The search on
+%% "monoidal category" "resource-aware" "programming language" on Google
+%% Scholar produces just 6 papers!}
 (including $\Pi$) with a new type
@@
 \subsection{Small Examples}
+\label{sub:toff}
 
@@
 \textit{Level-1 programs} & c &::=& \cdots \alt
-   \eta_{\net} : 1 \leftrightarrow (\net * \oneover{\net}) \\
-   &&& ~~~~~\alt \epsilon_{\net} : (\net * \oneover{\net}) \leftrightarrow 1 \\
+   \eta_{\net} : 1 \leftrightarrow (\net \times \oneover{\net}) \\
+   &&& ~~~~~\alt \epsilon_{\net} : (\net \times \oneover{\net}) \leftrightarrow 1 \\
 \textit{Level-2 types} &&& c \Leftrightarrow c \\
@@
 The metavariable $\net$ excludes all types that are equivalent to the
-empty type, e.g., $(1 + 1) * 0$. For now, all gc processes are
+empty type, e.g., $(1 + 1) \times 0$. For now, all gc processes are
 represented at runtime using a trivial value denoted by $\gcv$. We
@@
 \textit{Level-1 programs} & c &::=& \cdots \alt
-   \eta_{v:t} : 1 \leftrightarrow (t * \oneover{v}) \\
-   &&& ~~~~~\alt \epsilon_{v:t} : (t * \oneover{v}) \leftrightarrow 1 \\
+   \eta_{v:t} : 1 \leftrightarrow (t \times \oneover{v}) \\
+   &&& ~~~~~\alt \epsilon_{v:t} : (t \times \oneover{v}) \leftrightarrow 1 \\
 \textit{Level-2 types} &&& c \Leftrightarrow c \\
@@
 for particular entangled states~\cite{DBLP:journals/cal/ZhouB19}.
+
+As a more substantial example, we present a construction (in
+Fig.~\ref{fig:toff}) of not just the four-bit Toffoli function from
+Sec.~\ref{sub:toff} but in fact an inductive construction that defines
+all $n$-bit Toffoli gates. The construction goes back to Toffoli's
+proof of universality~\cite{Toffoli:1980}:
+
+\begin{figure*}[ht]
+  \PIFDtoffoli{}
+\caption{\label{fig:toff}Generalized Toffoli Gates}
+\end{figure*}
 
@@
 To start, we introduce some of the common technical details in the
-simpler setting of the core $\Pi$ language. First, simple facts
-about cardinalities:
-
-\PIFMEMsize{}  % maybe we should use \PIMEMcard here?
-
-\vspace{-\baselineskip}
+simpler setting of the core $\Pi$ language. The cardinalities of
+finite types map directly to the natural numbers. An easy proposition
+to verify is that every combinator must be between types of the same
+cardinalities; indeed one of the standard interpretations of $\Pi$
+combinators is that they are permutations on finite sets, i.e., they
+just shuffle elements around without changing the sizes of the
+underlying sets:
 
 \PIMEMcardeq{}
-
-The cardinalities of the types (other than the fractional type) map
-directly to the natural numbers. An easy proposition to verify is that
-every combinator must be between types of the same cardinalities;
-indeed one of the standard interpretations of $\Pi$ combinators is
-that they are permutations on finite sets, i.e., they just shuffle
-elements around without changing the sizes of the underlying sets.
 
@@
 
-The main difference is that now the state constructor does not take a
+The main difference is that the state constructor does not take a
 fixed size and indeed the size changes during execution as appropriate
 to fit the elements of the type $A$. In particular, since fractional
-values have a trivial representation, the size of a fractional value
-is 1, and hence $\eta$ used at type $A$ changes the size of the memory
-list by the size of $A$. As an illustrative example, we include a
-trace of the execution of \AgdaFunction{id'} from Sec.~\ref{sub:id}
+values have a trivial representation, the cardinality of a fractional
+type is 1, and hence $\eta$ used at type $A$ changes the size of the
+memory list by the size of $A$. As an illustrative example, we include
+a trace of the execution of \AgdaFunction{id'} from Sec.~\ref{sub:id}
 applied to 𝕋:
@@
 show. What is more interesting is \AgdaFunction{test3} which shows a
-property that holds for all natural numbers $n$. The full evaluation
-was not used but only partially as needed to derive the property. This
-form of partial evaluation will prove to be quite expressive allowing
-us to keep track of ancillae values throughout complex programs as
-demonstrated in Sec.~\ref{sec:cat}.
-\ednote{JC: Doesn't \AgdaFunction{eval}, in the type, evaluate
-fully for this work?  That Agda is strongly normalizing is rather
-important here. What exactly is partially evaluating is really not
-clear (says an expert in PE). I think the idea being expressed is
-correct, but the paragraph is insufficiently clear/precise.}
+property that holds for all natural numbers $n$. Evaluation in this
+case proceeds ``symbolically'' in the type system. From the definition
+of \Afun{eval}, it is clear that
+\Afun{eval}~\Acon{square}~(\Acon{suc}~\Avar{n}) produces
+(\Acon{suc}~\Avar{n})*(\Acon{suc}~\Avar{n}) which by definition of
+multiplication is an expression with a leading \Acon{suc} constructor
+which is enough to determine that the result of evaluating
+\Acon{isZero} on it yields \Acon{false}. This form of partial
+evaluation will prove to be quite expressive allowing us to keep track
+of ancillae values throughout complex programs as demonstrated in
+Sec.~\ref{sec:cat}.
+%% \ednote{JC: Doesn't \AgdaFunction{eval}, in the type, evaluate
+%% fully for this work?  That Agda is strongly normalizing is rather
+%% important here. What exactly is partially evaluating is really not
+%% clear (says an expert in PE). I think the idea being expressed is
+%% correct, but the paragraph is insufficiently clear/precise.}
 

### 2019-11-22 fracGC/p.tex
revrev in paper
revrev trace
remove level 2

@@
 
+\newunicodechar{Π}{$\Pi$}
 \newunicodechar{≟}{$\mathbin{\stackrel{?}{=}}$}
@@
 \textit{Values}      & v &::=& \star \alt \inlv{v} \alt \inrv{v} \alt (v,v) \\
-\textit{Level-1 types} &&& t \leftrightarrow t \\
-\textit{Level-1 programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})\\
-\textit{Level-2 types} &&& c \Leftrightarrow c \\
-\textit{Level-2 programs} & \alpha &::=& (\textrm{Omitted})
+\textit{Program types} &&& t \leftrightarrow t \\
+\textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
 \end{array}\]
@@
 they form a \emph{distributive bimonoidal category}~\cite{laplaza72}.
-
-So far, the types encode conventional data structures, i.e, sets of
-values and structured trees of values and the isomorphisms act on such
-conventional data. Universal computation models however
-fundamentally rely on the fact that \emph{programs are (or can be
-  encoded as) data}, e.g., a Turing machine can be encoded as a string
-that another Turing machine (or even the same machine) can
-manipulate. In our setting, we ask whether the type isomorphisms in
-Fig.~\ref{pi-terms} can themselves be subject to (higher-level)
-reversible deformations? The answer is
-yes~\cite{Miller:2003:TBA:775832.775915,DBLP:conf/esop/CaretteS16,DBLP:journals/corr/CockettCS17}. We
-only give one example below. (We will require a new level-2 program to
-witness coherence of fractional types in Sec.~\ref{sec:dep}.)
-
-\begin{figure}[htb]
-\begin{center}
-\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
-  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
-  \draw (-2,-2) ellipse (0.9cm and 1cm);
-  \draw[fill] (-2,-1.5) circle [radius=0.025];
-  \node[left] at (-2.1,-1.5) {$A$};
-  \draw[fill] (-2,-2.5) circle [radius=0.025];
-  \node[left] at (-2.1,-2.5) {$B$};
-
-  \draw (6.5,-2) ellipse (0.9cm and 1cm);
-  \draw[fill] (6.5,-1.5) circle [radius=0.025];
-  \node[right] at (6.7,-1.5) {$C$};
-  \draw[fill] (6.5,-2.5) circle [radius=0.025];
-  \node[right] at (6.7,-2.5) {$D$};
-
-  \draw[-] (-2,-1.5) to[bend left] (1,0.5) ;
-  \draw[-] (-2,-2.5) to[bend left] (1,-0.5) ;
-  \draw[-] (3.5,0.5) to[bend left] (6.5,-1.45) ;
-  \draw[-] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
-
-  \draw[-] (-2,-1.5) to[bend right] (1,-3.5) ;
-  \draw[-] (-2,-2.5) to[bend right] (1,-4.5) ;
-  \draw[-] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
-  \draw[-] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
-
-  \draw     (2.5,-3)  -- (3.5,-3) -- (3.5,-4) -- (2.5,-4) -- cycle ;
-  \draw     (2.5,-4)  -- (3.5,-4) -- (3.5,-5) -- (2.5,-5) -- cycle ;
-
-  \draw     (1,1)  -- (2,1) -- (2,0) -- (1,0) -- cycle ;
-  \draw     (1,0)  -- (2,0) -- (2,-1) -- (1,-1) -- cycle ;
-
-  \node at (3,-3.5) {$c_2$};
-  \node at (3,-4.5) {$c_1$};
-
-  \node at (1.5,0.5) {$c_1$};
-  \node at (1.5,-0.5) {$c_2$};
-
-  \draw     (2,0.5)  -- (2.5,0.5)  ;
-  \draw     (2,-0.5) -- (2.5,-0.5) ;
-
-  \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
-  \draw     (2.5,-0.5) -- (3.5,0.5) ;
-
-  \draw     (1,-3.5)  -- (2,-4.5)    ;
-  \draw     (1,-4.5) -- (2,-3.5)   ;
-
-  \draw     (2,-3.5)  -- (2.5,-3.5)    ;
-  \draw     (2,-4.5) -- (2.5,-4.5)   ;
-
-\end{tikzpicture}
-\end{center}
-\end{figure}
-
-\noindent The top path is the $\Pi$ program
-$(c_1~\oplus~c_2)~\fatsemi~\swapp$ which acts on the type $A$ by $c_1$,
-acts on the type $B$ by $c_2$, and deforms the resulting space by a
-twist that exchanges the two injections into the sum type. The bottom
-path performs the twist first and then acts on the type $A$ by $c_1$
-and on the type $B$ by $c_2$ as before. If one could imagine the paths
-as physical wires and the actions $c_1$ and $c_2$ as permutations on
-these wires then, holding the points $A$, $B$, $C$, and $D$ fixed, it
-is possible to rotate the top part of the diagram to become identical
-to the bottom one. That rotation can be undone (reversed), which takes
-the bottom part of the diagram into the top part.  In other words,
-there exists a continuous deformation of the program
-$(c_1~\oplus~c_2)~\fatsemi~\swapp$ to the program
-$\swapp \fatsemi (c_2~\oplus~c_1)$. We can also show that this means
-that, as permutations, $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and
-$\swapp \fatsemi (c_2~\oplus~c_1)$ are equal. This relation is
-non-trivial, as not all programs between the same types can be
-deformed into one another. The simplest example of inequivalent
-deformations are the two automorphisms of $1+1$, namely $\idc$ and
-$\swapp$.
 
@@
 \textit{Values}      & v &::=& \cdots \alt \gcv \\
-\textit{Level-1 types} &&& t \leftrightarrow t \\
-\textit{Level-1 programs} & c &::=& \cdots \alt
+\textit{Program types} &&& t \leftrightarrow t \\
+\textit{Programs} & c &::=& \cdots \alt
    \eta_{\net} : 1 \leftrightarrow (\net \times \oneover{\net}) \\
-   &&& ~~~~~\alt \epsilon_{\net} : (\net \times \oneover{\net}) \leftrightarrow 1 \\
-\textit{Level-2 types} &&& c \Leftrightarrow c \\
-\textit{Level-2 programs} & \alpha &::=& \cdots
-  \alt \mathit{inv} : \eta \fatsemi \epsilon \Leftrightarrow \idc
+   &&& ~~~~~\alt \epsilon_{\net} : (\net \times \oneover{\net}) \leftrightarrow 1 
 \end{array}\]
@@
 empty type, e.g., $(1 + 1) \times 0$. For now, all gc processes are
-represented at runtime using a trivial value denoted by $\gcv$. We
-also explicitly add the desired coherence condition on the creation
-and annihilation process as a level-2 program that would enforce the
-duality of $\eta$ and $\epsilon$. At this point, an application of
-$\eta$ following by $\epsilon$ might however throw an exception.
+represented at runtime using a trivial value denoted by $\gcv$.
 
@@
 \textit{Values}      & v &::=& \cdots \alt \gcv \\
-\textit{Level-1 types} &&& t \leftrightarrow t \\
-\textit{Level-1 programs} & c &::=& \cdots \alt
+\textit{Program types} &&& t \leftrightarrow t \\
+\textit{Programs} & c &::=& \cdots \alt
    \eta_{v:t} : 1 \leftrightarrow (t \times \oneover{v}) \\
-   &&& ~~~~~\alt \epsilon_{v:t} : (t \times \oneover{v}) \leftrightarrow 1 \\
-\textit{Level-2 types} &&& c \Leftrightarrow c \\
-\textit{Level-2 programs} & \alpha &::=& \cdots
-  \alt \mathit{inv} : \eta_{v:t} \fatsemi \epsilon_{v:t} \Leftrightarrow \idc
+   &&& ~~~~~\alt \epsilon_{v:t} : (t \times \oneover{v}) \leftrightarrow 1 
 \end{array}\]
@@
 
+\begin{proposition}
+sing is a monad; comonad; etc
+\end{proposition}
+\begin{proof}
+\PIPFCombderive{}
+\end{proof}
+
   properties of singleton, recip
@@
 show a small example
+
+\centering{
+\input{revrev.tikz}
+}
+
+\PIPFrevrev{}
+
+\includegraphics[scale=0.5]{trace3.png}
 
@@
 
+Yokoyama, T., Axelsen, H.B., Glück, R.: Principles of a reversible
+programming language. In: Conference on Computing Frontiers, CF,
+pp. 43–54. ACM Press (2008)
+
 Positive rational numbers are a model. Apparently there is a

### 2019-11-22 fracGC/p.tex
fix centering
code for examples
fix small typo

@@
 \input{latex/BooleanCircuitsCode.tex}
+\input{latex/PiPointedFracTraceCode.tex}
 
@@
 the next section, we introduce the problem of ancilla, motivate its
-importance, and explain the limitations of current solutions. Inm
+importance, and explain the limitations of current solutions. In
 Sec.~\ref{sec:pi}, we review existing work%
@@
 
-\centering{
+\begin{center}
 \input{revrev.tikz}
-}
+\end{center}
 
@@
 
+\PPFTname{}
+
+\PPFTtrace{}
+
+\PPFThof{}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-11-22 fracGC/p.tex
abramsky's pic
leave ednote in comments only
Make Generalized Toffoli fit in 1 column. Saves a lot of space...
done with 6.2 for now

@@
 
-The denotation of the fractional type is now:
+Following the development for the toy language above, we create a
+lifted version of fractional $\Pi$ with pointed types. In addition to
+the explicit pointed types of the form $t$\Acon{\#}$v$, we will need a
+multiplicative structure over pointed types, fractional types, and a
+special kind of pointed type that includes just one value, a singleton
+type. The singleton types will allow the type system to track the flow
+of one particular value, which is exactly what is needed to prove the
+safety of deallocation:
 
@@
 
-Combinators:
+As can be seen, a lifted type is interpreted as a type along with a
+corresponding value. A singleton type consists a special value of a
+given type: any claim that a value belongs to the singleton type must
+come with a proof that this value is equivalent to the special
+value. A fractional type consumes a particular singleton type
+returning the unit type. Note that the runtime gc process simply
+disregards the value it is given: the type system however prevents
+this value from being applied to anything but the particular singleton
+value in question.
+
+The combinators in the lifted language consist of all the combinators
+in the base language but now operating on pointed types. The mediation
+between general pointed types and singleton types is done via
+\Acon{return} and \Acon{extract} which as we prove below form a
+dual monad/comonad pair. The types for \Acon{$\eta$} and
+\Acon{$\epsilon$} are now specialized to guarantee safety of
+de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
+current witness value is put in focus in a singleton type and a gc
+process for that particular singleton type is created. To apply this
+process using \Acon{$\epsilon$} the very same singleton value must be
+the current one.
 
@@
 
+The mediation between pointed and singleton types is based a dual
+monad/comonad pair from which many structural properties can be
+derived.
+
 \begin{proposition}
-sing is a monad; comonad; etc
+  $\llparenthesis \cdot \rrparenthesis$ is both an idempotent strong
+  monad and an idempotent costrong comonad over pointed types.
 \end{proposition}
 \begin{proof}
+  The main insight needed is to define the functor \Afun{∙Singᵤ}, the
+  tensor/cotensor, and the join/cojoin (duplicate):
 \PIPFCombderive{}
@@
 
-  properties of singleton, recip
-derive join etc
-monad,comonad structure
-
-Evaluation:
+Like for the toy language, evaluation is reflected in the type
+system, and in this case we have the additional property that
+evaluation is reversible:
+
 \PIPFeval{}
 
-Combine with eval, the whole thing is reversible:
+\vspace{-\baselineskip}
+
 \PIPFrev{}
 
-show a small example
+Thus, to summarize if a combinator expects a singleton type, then it
+would only typecheck in the lifted language, if it is given the unique
+value it expects. A particularly intringuing instance of that
+situation is the following program:
 
@@
 \PIPFrevrev{}
+
+The program takes a value of type
+\Acon{∙𝟙/}~(\Acon{∙𝟙/}~\Avar{A}). This would be a gc process
+specialized to collect another gc process! At runtime type there would
+be no information other than the functions that ignore their argument
+but the type system provides enough guarantees to ensure that this
+process is well-defined, is reversible, and is safe. The following
+trace shows the actual execution for confirmation:
 
@@
 
+thinking of gc as space transformers
+
 \PPFTname{}
+\PPFThof{}
+
+trace about gc'ing cycles
 
@@
 
-\PPFThof{}
+\includegraphics[scale=0.5]{trace.png}
 

### 2019-11-22 fracGC/p.tex
go through PiMemSemCode and indicate what is not used and leave comments about documenting results in paper
examples trace etc

@@
 \usepackage{bbold}
-\usepackage[show]{ed}
 \usepackage{tikz}
@@
 
+\newunicodechar{⊸}{$\multimap$}
 \newunicodechar{Π}{$\Pi$}
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
   \PIFDtoffoli{}
 \caption{\label{fig:toff}Generalized Toffoli Gates}
-\end{figure*}
+\end{figure}
 
@@
 \PIMEMstep{}
+%% Note that step' uses step, but step is never shown
+%% The need for the rewrite probably should be explained too.
 
@@
 Sec.~\ref{sec:cat}.
-%% \ednote{JC: Doesn't \AgdaFunction{eval}, in the type, evaluate
-%% fully for this work?  That Agda is strongly normalizing is rather
-%% important here. What exactly is partially evaluating is really not
-%% clear (says an expert in PE). I think the idea being expressed is
-%% correct, but the paragraph is insufficiently clear/precise.}
 
@@
 would only typecheck in the lifted language, if it is given the unique
-value it expects. A particularly intringuing instance of that
+value it expects. A particularly intriguing instance of that
 situation is the following program:
@@
 First we can inject plain Pi type and value into PiFracDyn:
-\ednote{Pi/.U is not defined before its use below}
+%% \ednote{Pi/.U is not defined before its use below}
 \INJU{}
@@
 
-About trace, it is remarkable: it says, given a function (possibly
-using eta/epsilon so input/output sizes not guaranteed to be the same)
-\verb|AxC <-> BxC| hence using \verb|AxC| space for the input and
-\verb|BxC| space for output, we can 'trace out' the space used for
-\verb|C| and derive a function \verb|A <-> B|. This is true \emph{even
-  if} \verb|C| is actually used. See the swap example.
-
-revrev: rematerialization ! gc the gc process!
-
-Use all the constructions name, coname, etc. and see what they do in this context!
-
-allocate A
-allocate B
-combinate the gc's for A and B to get 1/(AxB)
-allocate AxB: split the gc in two 1/A and 1/B
-use the different gcs to satisfy each other
-
-Add multiplicative trace, do SAT solver
-
-Toffoli4 using 2 Toffoli3: core of proof of universality; simple
-enough. Reversible/Quantum Circuits and Ancilla Wires: Early use of
-ancillae in Toffoli's paper to implement arbitrary reversible
-functions using a fixed number of 3 input gates
-\url{https://link.springer.com/content/pdf/10.1007%2F3-540-10003-2_104.pdf}
-
-In-place matrix transpose: ease of programming, efficiency. Say we
-want to transpose a matrix. Wikipedia example
-(\url{https://en.wikipedia.org/wiki/In-place_matrix_transposition}):
-\begin{verbatim}
-11 12 13 14
-21 22 23 24
-\end{verbatim}
-transposed to
-\begin{verbatim}
-11 21
-12 22
-13 23
-14 24
-\end{verbatim}
-In Pi, the input and output matrices are:
-\begin{verbatim}
-M = (11,21) , (12,22) , (13,23) , (14,24)
-
-trM = (11,12,13,14) , (21,22,23,23)
-\end{verbatim}
-Say we are given a language like Pi that is sound and complete with
-respect to permutations on finite types, we would write the
-permutation like so.
-\begin{verbatim}
-WRITE PERMUTATION
-\end{verbatim}
-This code does not use additional space, i.e., it performs the matrix
-transpose in constant space, i.e., performs an in-place matrix
-transposition. It is well know that with additional space, one can
-write more efficient matrix transpositions (explain and citations).
-
-Example(s) from Quipper etc. with focus on safety condition. Quipper
-uses ancilla bits in several places; one use is to compile
-irreversible circuits to reversible ones; need ancilla bits; more
-generally several quantum algorithms need ancilla bits (see below); in
-quipper de-allocation left to programmer.
-
-Is the below a possible example?
-
-Say we already have a permutation $A \leftrightarrow B$
-we can implement a permutation $X \leftrightarrow Z$
-when there exists $Y$ such that $A = X * Y = Z * Y = B$:
-\[\begin{array}{rcl}
-X &\rightarrow& X * Y * 1/Y \\
-  &\rightarrow& A * 1/Y \\
-  &\rightarrow&  B * 1/Y \\
-  &\rightarrow&  Y * Z * 1/Y \\
-  &\rightarrow&  Z * Y * 1/Y \\
-  &\rightarrow&  Z
-\end{array}\]
-This basically says, in the language of Lens, that
-when $A$ is isomorphic to $X \times Y$ and
-$B$ is isomorphic to $Z \times Y$, then
-$X$ is isomorphic to $Y$.
-
-Another way to think of it: for all isomorphic
-$A$ and $B$, whenever they can be factored with a
-common complement ($Y$), then the ``other pieces''
-(here $X$ and $Y$) are automatically isomorphic.
-
-thinking of gc as space transformers
-
-\PPFTname{}
+In addition to the examples presented throughout the paper, we present
+examples that provides novel perspectives on gc processes, first as
+space transformers, and second as tracing operators. Both ideas are
+inspired from known constructions in compact closed
+categories~\cite{DBLP:conf/calco/Abramsky05}.
+
+\paragraph*{Space Transformers.} A pair of a value and a gc process
+cancel each other when the gc process is expecting this particular
+value. In more general cases, the pair can be viewed as a space
+transformer that represents an implicit function that converts the value to the
+one expected by the gc process. Such space transformers can be
+composed and applied
+
 \PPFThof{}
 
-trace about gc'ing cycles
+\paragraph*{Tracing Operators.} Another intriguing abstraction is
+based on the fact that dual objects can be used to define a tracing
+operator:
 
@@
 
+The operator takes a function that expects a pair of inputs $A$ and
+$C$ and returns a pair of outputs $B$ and $C$. In the world of
+reversible programs (or permutations), the larger permutation
+guarantees the existence of a smaller permutation that permutes $A$ to
+$B$. Indeed, we can programmatically construct such a permutation:
+
+As the diagram below (take from Abramsky's
+paper~\cite{DBLP:conf/calco/Abramsky05}) illustrates, this is not a
+trivial construction in the sense that the permutation from $A$ to $B$
+uses $C$ in non-trivial ways.
+
 \includegraphics[scale=0.5]{trace.png}
+
+Interpreted in our setting, the program can be written as follows:
+
+\PPFTtraceex{}
+
+The feedback loop in the diagram is realized using ancillae allocation
+and de-allocation. The safety condition on de-allocation ensures that
+the allocated value (now in the feedback loop) matches the
+de-allocated value. Thus, the entire traced component is treated as an
+ancillae value used internally used to perform the larger program but
+leaving no memory footprint.
+
+%% SAT solver ??
+%% in-place matrix transpose
 
@@
 pp. 43–54. ACM Press (2008)
+
+Trace as a lens? This basically says, in the language of Lens, that
+when $A$ is isomorphic to $X \times Y$ and $B$ is isomorphic to
+$Z \times Y$, then $X$ is isomorphic to $Y$. Another way to think of
+it: for all isomorphic $A$ and $B$, whenever they can be factored with
+a common complement ($Y$), then the ``other pieces'' (here $X$ and
+$Y$) are automatically isomorphic.
 

### 2019-11-22 fracGC/p.tex
finish with all XXX citations.  Make trace float to save space.

@@
 gate is not
-universal~\cite{DBLP:conf/innovations/AaronsonGS17,DBLP:journals/corr/Xu15e}.
+universal~\cite{aaronson_et_al:LIPIcs:2017:8173,DBLP:journals/corr/Xu15e}.
 And, as would be expected, the more temporary storage is allowed, the
@@
 than the $n$ bits used to represent the
-input~\cite{DBLP:conf/innovations/AaronsonGS17}. As such a restriction
+input~\cite{aaronson_et_al:LIPIcs:2017:8173}. As such a restriction
 disallows countless computations for trivial reasons, reversible and
@@
 Quantum error correction requires the ability to replicate the state of a
-qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is
-impossible to do directly, by the no-cloning theorem~\cite{XXX}. Again ancilla
-qubits solve the problem: because they start in a known
+qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is impossible to do
+directly, by the no-cloning
+theorem~\cite{park1970concept,wootters1982single,dieks1982communication}. Again
+ancilla qubits solve the problem: because they start in a known
 initial state, it is possible to create a simple circuit that makes
@@
   categories}~\cite{kelly1972many}. These categories extend the
-monoidal categories~\cite{XXX} which are used to model many
-resource-aware (e.g., based on linear types) programming
-languages~\cite{XXX}
-%% \ednote{The search on
-%% "monoidal category" "resource-aware" "programming language" on Google
-%% Scholar produces just 6 papers!}
+monoidal
+categories~\cite{benabou1963algebre,benabou1964algebre,maclane1963natural} which
+are used to model many resource-aware (e.g., based on linear types) programming
+languages~\cite{benton1994mixed,krishnaswami2015integrating}
 (including $\Pi$) with a new type
@@
   of type isomorphisms is a category with two symmetric monoidal
-  structures (additive and multiplicative) with the multiplicative
-  fragment distributing over the additive one~\cite{XXX}. Generally,
+  structures (additive and multiplicative) with the multiplicative fragment
+  distributing over the additive one~\cite{DBLP:conf/esop/CaretteS16}. Generally,
   a symmetric monoidal category is \emph{compact closed} if every
   object has a dual object whose behavior is governed by our $\eta$
-  and $\epsilon$ operators~\cite{XXX}. Our language from the previous
+  and $\epsilon$ operators~\cite{kelly1972many}. Our language from the previous
   section is an adaptation of this idea to the context of $\Pi$. Many
@@
 type is 1, and hence $\eta$ used at type $A$ changes the size of the
-memory list by the size of $A$. As an illustrative example, we include
-a trace of the execution of \AgdaFunction{id'} from Sec.~\ref{sub:id}
-applied to 𝕋:
-
+memory list by the size of $A$. As an illustrative example,
+Figure~\ref{fig:trace-id} shows a trace of the execution of \AgdaFunction{id'}
+from Sec.~\ref{sub:id} applied to 𝕋:
+
+\begin{figure}[htb]
 \includegraphics[scale=0.5]{trace2.png}
+\caption{Execution trace of \AgdaFunction{id'}}
+\label{fig:trace-id}
+\end{figure}
 
@@
 The strategy we adopt is to create a lifted version of the toy
-language with \emph{pointed types}~\cite{XXX}, i.e., types paired with
+language with \emph{pointed types}~\cite{hottbook}, i.e., types paired with
 a value of the type. In the lifted language, the evaluation

### 2019-11-22 fracGC/p.tex
minor

@@
 memory register using an index into an enumeration. Thus a memory
-register containing a boolean value would be represented in Agda as 𝔽
-::  𝕋 :: [] [ Fin.0F ] or 𝔽 :: 𝕋 :: [] [ Fin.1F ] depending on whether the
+register containing a boolean value would be represented as
+\AgdaFunction{[𝔽 ,  𝕋 ][0]} or \AgdaFunction{[𝔽 ,  𝕋 ][1]} depending on whether the
 values is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}.
-The prepended list
+The prepended vector
 contains all possible values of the type and the number in between [\ldots ]
-is an index into the list, representing a finite number that
+is an index into the vector, representing a finite number that
 is guaranteed to be in bounds. In a low-level runtime representation
@@
 \begin{figure}[htb]
-\includegraphics[scale=0.5]{trace2.png}
+\includegraphics[scale=0.4]{trace2.png}
 \caption{Execution trace of \AgdaFunction{id'}}

### 2019-11-22 fracGC/p.tex
start on Related Work. Remove duplicates in bib file to placate bibtex.

@@
 
-Quantum wire recycling~\cite{PhysRevA.94.042337}
-
-Quantum runtime assertion~\cite{DBLP:journals/cal/ZhouB19}
-
-QWIRE~\cite{Paykin:2017:QCL:3009837.3009894}
-
-Yokoyama, T., Axelsen, H.B., Glück, R.: Principles of a reversible
-programming language. In: Conference on Computing Frontiers, CF,
-pp. 43–54. ACM Press (2008)
-
-Trace as a lens? This basically says, in the language of Lens, that
-when $A$ is isomorphic to $X \times Y$ and $B$ is isomorphic to
-$Z \times Y$, then $X$ is isomorphic to $Y$. Another way to think of
-it: for all isomorphic $A$ and $B$, whenever they can be factored with
-a common complement ($Y$), then the ``other pieces'' (here $X$ and
-$Y$) are automatically isomorphic.
-
-Positive rational numbers are a model. Apparently there is a
-categorification
-\url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
+There has been ample work on ancillae before us.  For example%
+\cite{PhysRevA.94.042337} explicitly represent the lifetime of
+quantum wires, enable quantum wires to be recycled, leading
+to more efficient circuits.  Interesting applications of
+ancillae is in the recent paper~\cite{DBLP:journals/cal/ZhouB19}
+on Quantum runtime assertions.  In section~\ref{sub:id}, we
+already mentionned dynamic assertion of entanglement; they
+also show how to provide (dynamic) assertions on classical
+states, and testing of superposition states.  Our techniques
+should combine with theirs to provide even more possibilities
+for dynamic asserctions. Ancilla management is a complex
+business, and the verified circuit compiler
+ReVerC~\cite{amy2017verified} goes to great pains to
+ensure that ancilla are dealt with properly.  Both
+QCL~\cite{omer2002procedural} and Quipper~\cite{Green:2013:QSQ:2491956.2462177}
+also do automatic (but not certified) ancilla management during code
+generation.  This remains an active area of research, as
+outlined in the CACM paper~\cite{valiron2015programming}.
+
+Another interesting avenue is to understand how to merge 
+pointed $\Pi$ with another simple language for quantum
+computing: QWIRE~\cite{Paykin:2017:QCL:3009837.3009894}.
+QWIRE has many desirable meta-properties that have been
+formally proven to hold.  Our respective paradigms appear eminently compatible,
+however their language has an unguarded \texttt{discard} directive that
+may lead to erroneous circuits. For example, their circuit
+reversing operation is partial, whereas ours is provably total.
+
+When looking at higher-level reversible languages, Janus
+comes to mind~\cite{lutz1982janus}.  More recently, \cite{Yokoyama:2008:PRP}
+shows how to do clean, i.e. \emph{garbage free} computation. It
+would certainly be intriguing to augment Janus with a certified
+garbage collection discipline such as ours, so that one might hope
+that the resulting computations might be more efficiently done.
+
+%Trace as a lens? This basically says, in the language of Lens, that
+%when $A$ is isomorphic to $X \times Y$ and $B$ is isomorphic to
+%$Z \times Y$, then $X$ is isomorphic to $Y$. Another way to think of
+%it: for all isomorphic $A$ and $B$, whenever they can be factored with
+%a common complement ($Y$), then the ``other pieces'' (here $X$ and
+%$Y$) are automatically isomorphic.
+
+%Positive rational numbers are a model. Apparently there is a
+%categorification
+%\url{https://alistairsavage.ca/pubs/Copelli-Categorification_of_the_Nonnegative_Rational_Numbers.pdf}
 

### 2019-11-22 fracGC/p.tex
trace1
figure for languages and names

--- 2019-11-22 fracGC/p.tex
@@
 
+The paper includes a number of languages that express different
+points. The three most important languages and their relationships are
+highlighted in this diagram:
+\begin{center}
+\input{lang.tikz}
+\end{center}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%
-\subsection{Choosing Ancillae Values}
+\subsection{Choosing Ancillae Values: $\Pi/D$}
 
@@
 
-In addition to being more expressive, this version has the advantage
-of replacing the cumbersome cardinality-based evidence that the type
-is not empty by the more intuitive evidence of choosing a value of the
+We will call this language $\Pi/D$ for the fractional extension of
+$\Pi$ with a dynamic check. In addition to being more expressive than
+the language in the previous subsection, $\Pi/D$ has the advantage of
+replacing the cumbersome cardinality-based evidence that the type is
+not empty by the more intuitive evidence of choosing a value of the
 type. The simplications are evident in the Agda formalization:
@@
 %%
-\subsection{Abstract Machine without Fractionals}
+\subsection{Abstract Machine for $\Pi$}
 
@@
 %%
-\subsection{Abstract Machine with Fractionals}
+\subsection{Abstract Machine for $\Pi/D$}
 
@@
 %%%%%
-\subsection{Pointed and Singleton Types}
+\subsection{Pointed and Singleton Types: $\Pi/\bullet$}
 
 Following the development for the toy language above, we create a
-lifted version of fractional $\Pi$ with pointed types. In addition to
-the explicit pointed types of the form $t$\Acon{\#}$v$, we will need a
-multiplicative structure over pointed types, fractional types, and a
-special kind of pointed type that includes just one value, a singleton
-type. The singleton types will allow the type system to track the flow
-of one particular value, which is exactly what is needed to prove the
-safety of deallocation:
+lifted version of fractional $\Pi$ with pointed types which we call
+$\Pi/\bullet$. In addition to the explicit pointed types of the form
+$t$\Acon{\#}$v$, we will need a multiplicative structure over pointed
+types, fractional types, and a special kind of pointed type that
+includes just one value, a singleton type. The singleton types will
+allow the type system to track the flow of one particular value, which
+is exactly what is needed to prove the safety of deallocation:
 
@@
 
-The combinators in the lifted language consist of all the combinators
-in the base language but now operating on pointed types. The mediation
+The combinators in the lifted language $\Pi/\bullet$ consist of all the combinators
+in the core $\Pi$ language but now operating on pointed types. The mediation
 between general pointed types and singleton types is done via
@@
 
-Exploit dependent types to reify proofs in the type system. Type
-checking requires (partial) eval. Once we have a proof can we somehow
-go back to \verb|PiFracDyn| and remove the runtime check?
-
-Polymorphic type for ancilla?
-\\
+By lifting programs and their evaluation to the type level, we can
+naturally leverage the typechecking process to verify properties of
+interest, including the safe de-allocation of ancillae. One ``could''
+just write programs in that style: indeed the dual nature of proofs
+and programs is more and more exploited in languages like the one used
+to formalize this paper (Agda). The examples in Sec.~\ref{sec:cat} are
+written in that style and they are relatively similar to programs
+in the ``normal'' language.
+
+However, it is also often the case than constructive proofs are
+further processed to extract a native efficient programs that eschew
+all the information needed for the proof invariants. In our case, the
+question is whether we can extract a program that runs in the XX but
+is guaranteed to be safe and hence can run without the runtime checks
+associated with de-allocation sites.
+
 First we can inject plain Pi type and value into PiFracDyn:

### 2019-11-22 fracGC/p.tex

--- 2019-11-22 fracGC/p.tex
@@
 to 𝕋, 𝕋:
-
-\includegraphics[scale=0.5]{trace1.png}
+{\center\input{trace1.tex}}
 

### 2019-11-22 fracGC/trace1.tex

\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}
\node[draw,align=left] at (0,0) {%
⟪ dist ⊚ (id⟷ ⊕ (id⟷ ⊗ swap₊)) ⊚ factor ∥ [{\color{gray}(𝔽 , 𝔽) , (𝔽 , 𝕋) , (𝕋 , 𝔽) ,} (𝕋 , 𝕋)] [ 3 ]⟫\\\\
⟪ id⟷ ⊚ (id⟷ ⊕ (id⟷ ⊗ swap₊)) ⊚ factor ∥ [{\color{gray}inj₁ (tt , 𝔽) , inj₁ (tt , 𝕋) , inj₂ (tt , 𝔽) ,} inj₂ (tt , 𝕋)] [ 3 ]⟫\\\\
⟪ (id⟷ ⊕ (id⟷ ⊗ swap₊)) ⊚ factor ∥ [{\color{gray}inj₁ (tt , 𝔽) , inj₁ (tt , 𝕋) , inj₂ (tt , 𝔽) ,} inj₂ (tt , 𝕋)] [ 3 ]⟫\\\\
⟪ (id⟷ ⊕ (id⟷ ⊗ id⟷)) ⊚ factor ∥ [{\color{gray}inj₁ (tt , 𝔽) , inj₁ (tt , 𝕋) ,} inj₂ (tt , 𝔽) {\color{gray}, inj₂ (tt , 𝕋)}] [ 2 ]⟫\\\\
⟪ (id⟷ ⊕ id⟷) ⊚ factor ∥ [{\color{gray}inj₁ (tt , 𝔽) , inj₁ (tt , 𝕋) ,} inj₂ (tt , 𝔽) {\color{gray}, inj₂ (tt , 𝕋)}] [ 2 ]⟫\\\\
⟪ id⟷ ⊚ factor ∥ [{\color{gray} inj₁ (tt , 𝔽) , inj₁ (tt , 𝕋) ,} inj₂ (tt , 𝔽) {\color{gray}, inj₂ (tt , 𝕋)}] [ 2 ]⟫\\\\
⟪ factor ∥ [{\color{gray} inj₁ (tt , 𝔽) , inj₁ (tt , 𝕋) ,} inj₂ (tt , 𝔽) {\color{gray} , inj₂ (tt , 𝕋)}] [ 2 ]⟫\\\\
⟪ id⟷ ∥ [{\color{gray} (𝔽 , 𝔽) , (𝔽 , 𝕋) ,} (𝕋 , 𝔽) {\color{gray} , (𝕋 , 𝕋)}] [ 2 ]⟫
};%
\end{tikzpicture}
}%

### 2019-11-22 fracGC/p.tex
More on Related work (categorical, quantum lambda)

@@
 inspired from known constructions in compact closed
-categories~\cite{DBLP:conf/calco/Abramsky05}.
+categories~\cite{dblp:conf/calco/abramsky05}.
 
@@
 As the diagram below (take from Abramsky's
-paper~\cite{DBLP:conf/calco/Abramsky05}) illustrates, this is not a
+paper~\cite{dblp:conf/calco/abramsky05}) illustrates, this is not a
 trivial construction in the sense that the permutation from $A$ to $B$
@@
 
-There has been ample work on ancillae before us.  For example%
+There has been ample work on ancillae before us.  For example
 \cite{PhysRevA.94.042337} explicitly represent the lifetime of
@@
 
+Another different thread of work stems from categorical models
+of quantum computation, especially those based on dagger
+compact closed categories as introduced
+in~\cite{Abramsky:2004:CSQ:1018438.1021878}, and greatly
+expanded upon in~\cite{abramsky2009categorical}.  Various
+results can be obtained in that general setting, such as the
+no-cloning theorem~\cite{abramsky2009no}. The $2016$ book
+\cite{chiribella2016quantum} is a good survey of the
+developments since. Selinger~\cite{selinger2007dagger,Selinger:2011:FDH:1942319.1942398}
+has been particularly active in this area. Of particular
+note for our applications are two 2005 papers by Abramsky
+\cite{abramsky2005structural,dblp:conf/calco/abramsky05}
+with explicit examples of circuits using trace that are very useful tests
+for our approach.
+
+Also related are various approaches to quantum lambda
+calculi, such as~\cite{dblp:journals/mscs/selingerv06,vantonder:2004,dowek2017lineal}.
+
 %Trace as a lens? This basically says, in the language of Lens, that

### 2019-11-22 fracGC/p.tex
more readable parity code

@@
 
+\begin{wrapfigure}{r}{0.2\textwidth}
+  \input{parity.tikz}
+\end{wrapfigure}
 We illustrate the new style of ancilla management with an interesting
@@
 
-\begin{wrapfigure}[0]{r}{0.2\textwidth}
-  \input{parity.tikz}
-\end{wrapfigure}
 \PIFDparity{}

### 2019-11-22 fracGC/p.tex
extraction

--- 2019-11-22 fracGC/p.tex
@@
 interest, including the safe de-allocation of ancillae. One ``could''
-just write programs in that style: indeed the dual nature of proofs
+just write programs in that style in $\Pi/\bullet$: indeed the dual nature of proofs
 and programs is more and more exploited in languages like the one used
@@
 written in that style and they are relatively similar to programs
-in the ``normal'' language.
+in the ``normal'' $\Pi/D$ language.
 
@@
 further processed to extract a native efficient programs that eschew
-all the information needed for the proof invariants. In our case, the
-question is whether we can extract a program that runs in the XX but
-is guaranteed to be safe and hence can run without the runtime checks
-associated with de-allocation sites.
-
-First we can inject plain Pi type and value into PiFracDyn:
-%% \ednote{Pi/.U is not defined before its use below}
-\INJU{}
-and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
-\INJcomb{}
-The given injection is correct because it does not change the denotation of type
-\INJUeq{}
-,the value
-\INJVeq{}
-and the interpretation of PiFracDyn is consistent with Pi:
-\INJEvaleq{}
-To extract a combinator from PiPointedFrac, first we need to extract the corresponding type for PiFracDyn.
-Since every type (except Recip) in PiPointedFrac is pointed type so we can always extract a type for plain
-Pi and a value from it.
-There is no corresponding type of Recip in plain Pi, but we can map it to the
-\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v} type in PiFracDyn.
-Note that for all other type the extracted value is erased from type except for
-\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
-which we still keep it in type.
-\EXTU{}
-Once we extract the corresponding type we can extract the combinator, the combinators about monad/comonad
-just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased,
-for the rest combinators there exist direct corresponding combinators in PiFracDyn:
-\EXTUComb{}
-Finally we can prove the correctness of the extraction,
-the lifted multiplicative combinators have the corresponding combinators in PiFracDyn which represent
-the same computation so the proof is just \AgdaInductiveConstructor{refl}.
-Because the combinators about monad/comonad does not really
-computes so for those case the proof is also \AgdaInductiveConstructor{refl}.
-And for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
-The interesting case is \AgdaInductiveConstructor{ε} where the singleton type guarantee that
-the dynamic check won't fail.
+the overhead of maintaining information needed just for proof
+invariants. In our case, the question is whether we can extract from
+$\Pi/\bullet$ program, a program in $\Pi/D$ that uses a simpler type
+system, a simpler runtime representation, and yet is guaranteed to be
+safe and hence can run without the runtime checks associated with
+de-allocation sites. In this section, we show that this indeed the
+case.
+
+% First we can inject plain Pi type and value into PiFracDyn:
+% %% \ednote{Pi/.U is not defined before its use below}
+% \INJU{}
+% and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
+% \INJcomb{}
+% The given injection is correct because it does not change the denotation of type
+% \INJUeq{}
+% ,the value
+% \INJVeq{}
+% and the interpretation of PiFracDyn is consistent with Pi:
+% \INJEvaleq{}
+
+The extraction map is fully implemented in the underlying Agda
+formalization. We present the most significant highlights. There are
+three important functions who signatures are below:
+
+\EXTsig{}
+
+The function \Afun{Ext𝕌} maps a $\Pi/\bullet$ type to a $\Pi/D$ type
+and a value in the type. The function \Afun{Ext∙⟶} maps a
+$\Pi/\bullet$ combinator to a $\Pi/D$ combinator. And finally the
+function \Afun{Ext≡} asserts that extracted code cannot throw an
+exception (it must return \Acon{just}\ldots).
+
+Each of these functions has one or two enlightening cases which we
+explain:
+
+In $\Pi/D$ the fractional type expresses that it expects a particular
+value but lacks any mechanisms to enforce this requirement. Thus we
+have no choice when mapping a fractional type from $\Pi/\bullet$ to 
+$\Pi/D$ than to just use this \Acon{𝟙/}~\Avar{v} type:
+
+\EXTu{}
+
+% There is no corresponding type of Recip
+% in plain Pi, but we can map it to the
+% \AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
+% type in PiFracDyn.  Note that for all other type the extracted value
+% is erased from type except for
+% \AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
+% which we still keep it in type.
+
+When mapping $\Pi/\bullet$ combinators to 
+$\Pi/D$ combinators, the main interesting cases are for \Acon{η} and
+\Acon{ε}. In those we just use the values from the pointed type as
+choices for the ancilla value and expectation for the gc process:
+
+\EXTcomb{}
+
+% Once we extract the
+% corresponding type we can extract the combinator, the combinators
+% about monad/comonad just becomes \AgdaInductiveConstructor{id↔} since
+% the information in type is erased, for the rest combinators there
+% exist direct corresponding combinators in PiFracDyn:
+
+% Finally we can prove the correctness of the extraction, the lifted
+% multiplicative combinators have the corresponding combinators in
+% PiFracDyn which represent the same computation so the proof is just
+% \AgdaInductiveConstructor{refl}.  Because the combinators about
+% monad/comonad does not really computes so for those case the proof is
+% also \AgdaInductiveConstructor{refl}.  And for
+% \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.  The
+% interesting case is \AgdaInductiveConstructor{ε} where the singleton
+% type guarantee that the dynamic check won't fail.
+
+Finally we can prove the correctness of the extraction. The punchline
+is in the following case:
+
 \EXTeq{}
+
+In this case, the singleton type in $\Pi/\bullet$ guarantees that the
+runtime check cannot fail.
 
@@
 on Quantum runtime assertions.  In section~\ref{sub:id}, we
-already mentionned dynamic assertion of entanglement; they
+already mentioned dynamic assertion of entanglement; they
 also show how to provide (dynamic) assertions on classical
@@
 should combine with theirs to provide even more possibilities
-for dynamic asserctions. Ancilla management is a complex
+for dynamic assertions. Ancilla management is a complex
 business, and the verified circuit compiler
@@
 
-General allocation and de-allocation. No need to keep track of
-lifetime. Just make sure location is back to initial value. Safe to
-de-allocate because indistinguishable from freshly created. No pointer
-equality and aliases though. Concurrent: one thread allocates; gc
-thread meets someone else
+% General allocation and de-allocation. No need to keep track of
+% lifetime. Just make sure location is back to initial value. Safe to
+% de-allocate because indistinguishable from freshly created. No pointer
+% equality and aliases though. Concurrent: one thread allocates; gc
+% thread meets someone else
 

### 2019-11-22 fracGC/p.tex
minor typo

@@
 
-As the diagram below (take from Abramsky's
+As the diagram below (taken from Abramsky's
 paper~\cite{dblp:conf/calco/abramsky05}) illustrates, this is not a

### 2019-11-22 fracGC/p.tex
Pass

@@
 circuits lack this form of parallelism. Most importantly, structurally
-the two circuits models are identical and one can derive properties
+the two circuit models are identical and one can derive properties
 valid for both families by focusing on the simpler classical model. In
@@
 \paragraph*{Temporary Storage using Ancilla Bits.} The universality of
-the Toffoli gate for classical reversible computing and of the
+the Toffoli gate for classical reversible computing and the
 universality of the Toffoli-Hadamard gates for quantum computing come
@@
 storage or 3 bits of bandwidth for communication. The information
-information contained in a composite state, by definition of the
-logarithm, is the sum of
-the information contained in its constituents. For example, the type
-$A \times B$ where $A$ has two elements and $B$ has three elements can
+contained in a composite state, by definition of the logarithm,
+is the sum of the information contained in its constituents. For example,
+the type $A \times B$ where $A$ has two elements and $B$ has three elements can
 be thought of a composite system consisting of two independent
@@
 that introduces a reversible programming language built using type
-isomorphism.  Although the concept of fractional types might apply to
+isomorphisms.  Although the concept of fractional types might apply to
 general-purpose languages (after some adaptation), its natural
@@
 ancillae. Technically this solution requires the introduction of
-pointed types on top of which are defined both a monad and comonad of
+pointed types on top of which we define both a monad and comonad of
 singleton types. The tracking of ancillae value is done within the
-type system via the singleton types. A number of examples exploiting
+type system via singleton types. A number of examples exploiting
 the singleton-based type system are presented to demonstrate the full

### 2019-11-22 fracGC/p.tex
trace 2 & 3

--- 2019-11-22 fracGC/p.tex
@@
 
-The paper includes a number of languages that express different
-points. The three most important languages and their relationships are
-highlighted in this diagram:
-\begin{center}
-\input{lang.tikz}
-\end{center}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%
-\subsection{Choosing Ancillae Values: $\Pi/D$}
+\subsection{Choosing Ancillae Values}
 
@@
 
-We will call this language $\Pi/D$ for the fractional extension of
-$\Pi$ with a dynamic check. In addition to being more expressive than
-the language in the previous subsection, $\Pi/D$ has the advantage of
-replacing the cumbersome cardinality-based evidence that the type is
-not empty by the more intuitive evidence of choosing a value of the
+In addition to being more expressive, this version has the advantage
+of replacing the cumbersome cardinality-based evidence that the type
+is not empty by the more intuitive evidence of choosing a value of the
 type. The simplications are evident in the Agda formalization:
@@
 
-\begin{wrapfigure}{r}{0.2\textwidth}
-  \input{parity.tikz}
-\end{wrapfigure}
 We illustrate the new style of ancilla management with an interesting
@@
 
+\begin{wrapfigure}[0]{r}{0.2\textwidth}
+  \input{parity.tikz}
+\end{wrapfigure}
 \PIFDparity{}
@@
 %%
-\subsection{Abstract Machine for $\Pi$}
+\subsection{Abstract Machine without Fractionals}
 
@@
 %%
-\subsection{Abstract Machine for $\Pi/D$}
+\subsection{Abstract Machine with Fractionals}
 
@@
 
-\begin{figure}[htb]
-\includegraphics[scale=0.4]{trace2.png}
-\caption{Execution trace of \AgdaFunction{id'}}
-\label{fig:trace-id}
-\end{figure}
+{\center\input{trace2.tex}}
 
@@
 %%%%%
-\subsection{Pointed and Singleton Types: $\Pi/\bullet$}
+\subsection{Pointed and Singleton Types}
 
 Following the development for the toy language above, we create a
-lifted version of fractional $\Pi$ with pointed types which we call
-$\Pi/\bullet$. In addition to the explicit pointed types of the form
-$t$\Acon{\#}$v$, we will need a multiplicative structure over pointed
-types, fractional types, and a special kind of pointed type that
-includes just one value, a singleton type. The singleton types will
-allow the type system to track the flow of one particular value, which
-is exactly what is needed to prove the safety of deallocation:
+lifted version of fractional $\Pi$ with pointed types. In addition to
+the explicit pointed types of the form $t$\Acon{\#}$v$, we will need a
+multiplicative structure over pointed types, fractional types, and a
+special kind of pointed type that includes just one value, a singleton
+type. The singleton types will allow the type system to track the flow
+of one particular value, which is exactly what is needed to prove the
+safety of deallocation:
 
@@
 
-The combinators in the lifted language $\Pi/\bullet$ consist of all the combinators
-in the core $\Pi$ language but now operating on pointed types. The mediation
+The combinators in the lifted language consist of all the combinators
+in the base language but now operating on pointed types. The mediation
 between general pointed types and singleton types is done via
@@
 
-\includegraphics[scale=0.5]{trace3.png}
+{\center\input{trace3.tex}}
 
@@
 
-By lifting programs and their evaluation to the type level, we can
-naturally leverage the typechecking process to verify properties of
-interest, including the safe de-allocation of ancillae. One ``could''
-just write programs in that style in $\Pi/\bullet$: indeed the dual nature of proofs
-and programs is more and more exploited in languages like the one used
-to formalize this paper (Agda). The examples in Sec.~\ref{sec:cat} are
-written in that style and they are relatively similar to programs
-in the ``normal'' $\Pi/D$ language.
-
-However, it is also often the case than constructive proofs are
-further processed to extract a native efficient programs that eschew
-the overhead of maintaining information needed just for proof
-invariants. In our case, the question is whether we can extract from
-$\Pi/\bullet$ program, a program in $\Pi/D$ that uses a simpler type
-system, a simpler runtime representation, and yet is guaranteed to be
-safe and hence can run without the runtime checks associated with
-de-allocation sites. In this section, we show that this indeed the
-case.
-
-% First we can inject plain Pi type and value into PiFracDyn:
-% %% \ednote{Pi/.U is not defined before its use below}
-% \INJU{}
-% and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
-% \INJcomb{}
-% The given injection is correct because it does not change the denotation of type
-% \INJUeq{}
-% ,the value
-% \INJVeq{}
-% and the interpretation of PiFracDyn is consistent with Pi:
-% \INJEvaleq{}
-
-The extraction map is fully implemented in the underlying Agda
-formalization. We present the most significant highlights. There are
-three important functions who signatures are below:
-
-\EXTsig{}
-
-The function \Afun{Ext𝕌} maps a $\Pi/\bullet$ type to a $\Pi/D$ type
-and a value in the type. The function \Afun{Ext∙⟶} maps a
-$\Pi/\bullet$ combinator to a $\Pi/D$ combinator. And finally the
-function \Afun{Ext≡} asserts that extracted code cannot throw an
-exception (it must return \Acon{just}\ldots).
-
-Each of these functions has one or two enlightening cases which we
-explain:
-
-In $\Pi/D$ the fractional type expresses that it expects a particular
-value but lacks any mechanisms to enforce this requirement. Thus we
-have no choice when mapping a fractional type from $\Pi/\bullet$ to 
-$\Pi/D$ than to just use this \Acon{𝟙/}~\Avar{v} type:
-
-\EXTu{}
-
-% There is no corresponding type of Recip
-% in plain Pi, but we can map it to the
-% \AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
-% type in PiFracDyn.  Note that for all other type the extracted value
-% is erased from type except for
-% \AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
-% which we still keep it in type.
-
-When mapping $\Pi/\bullet$ combinators to 
-$\Pi/D$ combinators, the main interesting cases are for \Acon{η} and
-\Acon{ε}. In those we just use the values from the pointed type as
-choices for the ancilla value and expectation for the gc process:
-
-\EXTcomb{}
-
-% Once we extract the
-% corresponding type we can extract the combinator, the combinators
-% about monad/comonad just becomes \AgdaInductiveConstructor{id↔} since
-% the information in type is erased, for the rest combinators there
-% exist direct corresponding combinators in PiFracDyn:
-
-% Finally we can prove the correctness of the extraction, the lifted
-% multiplicative combinators have the corresponding combinators in
-% PiFracDyn which represent the same computation so the proof is just
-% \AgdaInductiveConstructor{refl}.  Because the combinators about
-% monad/comonad does not really computes so for those case the proof is
-% also \AgdaInductiveConstructor{refl}.  And for
-% \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.  The
-% interesting case is \AgdaInductiveConstructor{ε} where the singleton
-% type guarantee that the dynamic check won't fail.
-
-Finally we can prove the correctness of the extraction. The punchline
-is in the following case:
-
+Exploit dependent types to reify proofs in the type system. Type
+checking requires (partial) eval. Once we have a proof can we somehow
+go back to \verb|PiFracDyn| and remove the runtime check?
+
+Polymorphic type for ancilla?
+\\
+First we can inject plain Pi type and value into PiFracDyn:
+%% \ednote{Pi/.U is not defined before its use below}
+\INJU{}
+and the combinators for Pi can also be inject into PiFracDyn in a straightforward way:
+\INJcomb{}
+The given injection is correct because it does not change the denotation of type
+\INJUeq{}
+,the value
+\INJVeq{}
+and the interpretation of PiFracDyn is consistent with Pi:
+\INJEvaleq{}
+To extract a combinator from PiPointedFrac, first we need to extract the corresponding type for PiFracDyn.
+Since every type (except Recip) in PiPointedFrac is pointed type so we can always extract a type for plain
+Pi and a value from it.
+There is no corresponding type of Recip in plain Pi, but we can map it to the
+\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v} type in PiFracDyn.
+Note that for all other type the extracted value is erased from type except for
+\AgdaOperator{\AgdaInductiveConstructor{𝟙/}}\AgdaSpace{}\AgdaBound{v}
+which we still keep it in type.
+\EXTU{}
+Once we extract the corresponding type we can extract the combinator, the combinators about monad/comonad
+just becomes \AgdaInductiveConstructor{id↔} since the information in type is erased,
+for the rest combinators there exist direct corresponding combinators in PiFracDyn:
+\EXTUComb{}
+Finally we can prove the correctness of the extraction,
+the lifted multiplicative combinators have the corresponding combinators in PiFracDyn which represent
+the same computation so the proof is just \AgdaInductiveConstructor{refl}.
+Because the combinators about monad/comonad does not really
+computes so for those case the proof is also \AgdaInductiveConstructor{refl}.
+And for \AgdaInductiveConstructor{∙c} we rely on \AgdaFunction{Eval≡}.
+The interesting case is \AgdaInductiveConstructor{ε} where the singleton type guarantee that
+the dynamic check won't fail.
 \EXTeq{}
-
-In this case, the singleton type in $\Pi/\bullet$ guarantees that the
-runtime check cannot fail.
 
@@
 inspired from known constructions in compact closed
-categories~\cite{dblp:conf/calco/abramsky05}.
+categories~\cite{DBLP:conf/calco/Abramsky05}.
 
@@
 As the diagram below (take from Abramsky's
-paper~\cite{dblp:conf/calco/abramsky05}) illustrates, this is not a
+paper~\cite{DBLP:conf/calco/Abramsky05}) illustrates, this is not a
 trivial construction in the sense that the permutation from $A$ to $B$
@@
 
-There has been ample work on ancillae before us.  For example
+There has been ample work on ancillae before us.  For example%
 \cite{PhysRevA.94.042337} explicitly represent the lifetime of
@@
 on Quantum runtime assertions.  In section~\ref{sub:id}, we
-already mentioned dynamic assertion of entanglement; they
+already mentionned dynamic assertion of entanglement; they
 also show how to provide (dynamic) assertions on classical
@@
 should combine with theirs to provide even more possibilities
-for dynamic assertions. Ancilla management is a complex
+for dynamic asserctions. Ancilla management is a complex
 business, and the verified circuit compiler
@@
 
-Another different thread of work stems from categorical models
-of quantum computation, especially those based on dagger
-compact closed categories as introduced
-in~\cite{Abramsky:2004:CSQ:1018438.1021878}, and greatly
-expanded upon in~\cite{abramsky2009categorical}.  Various
-results can be obtained in that general setting, such as the
-no-cloning theorem~\cite{abramsky2009no}. The $2016$ book
-\cite{chiribella2016quantum} is a good survey of the
-developments since. Selinger~\cite{selinger2007dagger,Selinger:2011:FDH:1942319.1942398}
-has been particularly active in this area. Of particular
-note for our applications are two 2005 papers by Abramsky
-\cite{abramsky2005structural,dblp:conf/calco/abramsky05}
-with explicit examples of circuits using trace that are very useful tests
-for our approach.
-
-Also related are various approaches to quantum lambda
-calculi, such as~\cite{dblp:journals/mscs/selingerv06,vantonder:2004,dowek2017lineal}.
-
 %Trace as a lens? This basically says, in the language of Lens, that
@@
 
-% General allocation and de-allocation. No need to keep track of
-% lifetime. Just make sure location is back to initial value. Safe to
-% de-allocate because indistinguishable from freshly created. No pointer
-% equality and aliases though. Concurrent: one thread allocates; gc
-% thread meets someone else
+General allocation and de-allocation. No need to keep track of
+lifetime. Just make sure location is back to initial value. Safe to
+de-allocate because indistinguishable from freshly created. No pointer
+equality and aliases though. Concurrent: one thread allocates; gc
+thread meets someone else
 

### 2019-11-22 fracGC/trace2.tex
trace 2 & 3

\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}
\node[draw,align=left] at (0,0) {%
⟪ uniti⋆r ⊚ (id↔ ⊗ η 𝔽) ⊚…∥ [{\color{gray}𝔽 ,} 𝕋] [ 1 ]⟫\\\\

⟪ (id↔ ⊗ η 𝔽) ⊚ … ∥ [{\color{gray}(𝔽 , tt) ,} (𝕋 , tt)] [ 1 ]⟫\\
  ⋮\\
⟪ (id↔ ⊗ id↔) ⊚ assocl⋆ ⊚ … ∥ [{\color{gray}(𝔽 , 𝔽 , ○) , (𝔽 , 𝕋 , ○) ,} (𝕋 , 𝔽 , ○) {\color{gray}, (𝕋 , 𝕋 , ○)}] [ 2 ]⟫\\
  ⋮\\
⟪ assocl⋆ ⊚ … ∥ [{\color{gray}(𝔽 , 𝔽 , ○) , (𝔽 , 𝕋 , ○) ,} (𝕋 , 𝔽 , ○) {\color{gray}, (𝕋 , 𝕋 , ○)}] [ 2 ]⟫\\
  ⋮\\
⟪ (((dist ⊚ …) ⊗ id↔) ⊚ …∥ [{\color{gray} ((𝔽 , 𝔽) , ○) , ((𝔽 , 𝕋) , ○) ,} ((𝕋 , 𝔽) , ○) {\color{gray}, ((𝕋 , 𝕋) , ○)}] [ 2 ]⟫\\
  ⋮\\
⟪((((id↔ ⊕ (id↔ ⊗ swap₊)) ⊚ …) ⊗ id↔) ⊚ …∥ [{\color{gray} (inj₁ (tt , 𝔽) , ○) , (inj₁ (tt , 𝕋) , ○) ,} (inj₂ (tt , 𝔽) , ○) {\color{gray}, (inj₂ (tt , 𝕋) , ○)}] [ 2 ]⟫\\
  ⋮\\
⟪((factor ⊚…) ⊗ id↔)⊚…∥ [{\color{gray}(inj₁ (tt , 𝔽) , ○) , (inj₁ (tt , 𝕋) , ○) , (inj₂ (tt , 𝔽) , ○) ,} (inj₂ (tt , 𝕋) , ○)] [ 3 ]⟫\\
  ⋮\\
⟪(((distl ⊚ …) ⊗ id↔) ⊚…∥ [{\color{gray} ((𝔽 , 𝔽) , ○) , ((𝔽 , 𝕋) , ○) , ((𝕋 , 𝔽) , ○) ,} ((𝕋 , 𝕋) , ○)] [ 3 ]⟫\\
  ⋮\\
⟪((((id↔ ⊕ (swap₊ ⊗ id↔)) ⊚ …) ⊗ id↔) ⊚ … ∥ [{\color{gray} (inj₁ (𝔽 , tt) , ○) , (inj₁ (𝕋 , tt) , ○) , (inj₂ (𝔽 , tt) , ○) ,} (inj₂ (𝕋 , tt) , ○)] [ 3 ]⟫\\
  ⋮\\
⟪ ((factorl ⊚ swap⋆) ⊗ id↔) ⊚ … ∥ [{\color{gray}(inj₁ (𝔽 , tt) , ○) , (inj₁ (𝕋 , tt) , ○) ,} (inj₂ (𝔽 , tt) , ○) , {\color{gray}(inj₂ (𝕋 , tt) , ○)}] [ 2 ]⟫\\
  ⋮\\
⟪ (swap⋆ ⊗ id↔) ⊚ assocr⋆ ⊚ … ∥ [{\color{gray}((𝔽 , 𝔽) , ○) ,} ((𝔽 , 𝕋) , ○) {\color{gray}, ((𝕋 , 𝔽) , ○) , ((𝕋 , 𝕋) , ○)}] [ 1 ]⟫\\
  ⋮\\
⟪ assocr⋆ ⊚ (id↔ ⊗ ε 𝔽) ⊚ unite⋆r ∥ [{\color{gray}(𝔽 , 𝔽) , ○) , ((𝔽 , 𝕋) , ○) ,} ((𝕋 , 𝔽) , ○) {\color{gray}, ((𝕋 , 𝕋) , ○)}] [ 2 ]⟫\\
  ⋮\\
⟪ (id↔ ⊗ ε 𝔽) ⊚ unite⋆r ∥ [{\color{gray}(𝔽 , 𝔽 , ○) , (𝔽 , 𝕋 , ○) ,} (𝕋 , 𝔽 , ○) {\color{gray}, (𝕋 , 𝕋 , ○)}] [ 2 ]⟫\\
  ⋮\\
⟪ unite⋆r ∥ [{\color{gray}(𝔽 , tt) ,} (𝕋 , tt)] [ 1 ]⟫ \\\\
⟪ id↔ ∥ [{\color{gray}𝔽 ,} 𝕋] [ 1 ]⟫
};%
\end{tikzpicture}
}%

### 2019-11-22 fracGC/p.tex
Pass over sec 2

--- 2019-11-22 fracGC/p.tex
@@
 restoring to the majority value.  Unlike a classical code, however, we
-can not directly measure the qubits else their quantum state will be
+cannot directly measure the qubits else their quantum state will be
 destroyed.  Instead, we measure syndromes from each possible pair of
@@
 \paragraph*{Deferring Quantum Measurements.} The evolution of a
-quantum system is deterministic. Measurement add significant
+quantum system is deterministic. Measurement adds significant
 complications, collapsing the quantum state, and producing
-probabilistic results. However, using ancilla qubits it is possible to
+probabilistic results. However, using ancilla qubits, it is possible to
 avoid doing any intermediate measurements in a quantum
@@
 \AgdaInductiveConstructor{compose isZero square} always returns
-\AgdaInductiveConstructor{false} when applies to a non-zero natural
+\AgdaInductiveConstructor{false} when applied to a non-zero natural
 number. We can certainly do this proof in Agda (i.e., in the
 meta-language of our formalization) but we would like to do the proof
-within the framework of the toy language itself. The most important
-reason is that this proof can be now be used within the language to
-optimize programs (or as the case will be for the full language to
+within the toy language itself. The most important
+reason is that it can then be used within the language to
+optimize programs (or, for the full language, to
 remove a runtime check).
@@
 
-Now various properties of our small example
+Various properties of
 \AgdaInductiveConstructor{compose isZero square} can be derived within
-the extended type system for the toy language:
+the extended type system:
 
@@
 
-Evidently we can track concrete values as the first two tests
-show. What is more interesting is \AgdaFunction{test3} which shows a
-property that holds for all natural numbers $n$. Evaluation in this
-case proceeds ``symbolically'' in the type system. From the definition
+We can track concrete values as the first two tests
+show. More interesting is \AgdaFunction{test3} which shows a
+property that holds for all natural numbers $n$. Evaluation
+proceeds ``symbolically'' in the type system. From the definition
 of \Afun{eval}, it is clear that
 \Afun{eval}~\Acon{square}~(\Acon{suc}~\Avar{n}) produces
-(\Acon{suc}~\Avar{n})*(\Acon{suc}~\Avar{n}) which by definition of
+(\Acon{suc}~\Avar{n}) * (\Acon{suc}~\Avar{n}) which by definition of
 multiplication is an expression with a leading \Acon{suc} constructor
-which is enough to determine that the result of evaluating
+which is enough to determine that evaluating
 \Acon{isZero} on it yields \Acon{false}. This form of partial
-evaluation will prove to be quite expressive allowing us to keep track
+evaluation will prove to be quite expressive, allowing us to keep track
 of ancillae values throughout complex programs as demonstrated in
@@
 
-Following the development for the toy language above, we create a
-lifted version of fractional $\Pi$ with pointed types which we call
-$\Pi/\bullet$. In addition to the explicit pointed types of the form
-$t$\Acon{\#}$v$, we will need a multiplicative structure over pointed
+Proceeding similarly, we create a
+version of $\Pi$ with pointed types which we call
+$\Pi/\bullet$. In addition to the pointed types of the form
+$t$\Acon{\#}$v$, we need a multiplicative structure over pointed
 types, fractional types, and a special kind of pointed type that

### 2019-11-22 fracGC/p.tex
small edits to section 6

@@
 restoring to the majority value.  Unlike a classical code, however, we
-cannot directly measure the qubits else their quantum state will be
+can not directly measure the qubits else their quantum state will be
 destroyed.  Instead, we measure syndromes from each possible pair of
@@
 \paragraph*{Deferring Quantum Measurements.} The evolution of a
-quantum system is deterministic. Measurement adds significant
+quantum system is deterministic. Measurement add significant
 complications, collapsing the quantum state, and producing
-probabilistic results. However, using ancilla qubits, it is possible to
+probabilistic results. However, using ancilla qubits it is possible to
 avoid doing any intermediate measurements in a quantum

### 2019-11-22 fracGC/p.tex
abstract

@@
   constraint by imposing a stack discipline and by leaving the second
-  constraint to programmer's assertions.
+  constraint to programmers' assertions.
 
@@
 \AgdaInductiveConstructor{compose isZero square} always returns
-\AgdaInductiveConstructor{false} when applied to a non-zero natural
+\AgdaInductiveConstructor{false} when applies to a non-zero natural
 number. We can certainly do this proof in Agda (i.e., in the
 meta-language of our formalization) but we would like to do the proof
-within the toy language itself. The most important
-reason is that it can then be used within the language to
-optimize programs (or, for the full language, to
+within the framework of the toy language itself. The most important
+reason is that this proof can be now be used within the language to
+optimize programs (or as the case will be for the full language to
 remove a runtime check).
@@
 
-Various properties of
+Now various properties of our small example
 \AgdaInductiveConstructor{compose isZero square} can be derived within
-the extended type system:
+the extended type system for the toy language:
 
@@
 
-We can track concrete values as the first two tests
-show. More interesting is \AgdaFunction{test3} which shows a
-property that holds for all natural numbers $n$. Evaluation
-proceeds ``symbolically'' in the type system. From the definition
+Evidently we can track concrete values as the first two tests
+show. What is more interesting is \AgdaFunction{test3} which shows a
+property that holds for all natural numbers $n$. Evaluation in this
+case proceeds ``symbolically'' in the type system. From the definition
 of \Afun{eval}, it is clear that
 \Afun{eval}~\Acon{square}~(\Acon{suc}~\Avar{n}) produces
-(\Acon{suc}~\Avar{n}) * (\Acon{suc}~\Avar{n}) which by definition of
+(\Acon{suc}~\Avar{n})*(\Acon{suc}~\Avar{n}) which by definition of
 multiplication is an expression with a leading \Acon{suc} constructor
-which is enough to determine that evaluating
+which is enough to determine that the result of evaluating
 \Acon{isZero} on it yields \Acon{false}. This form of partial
-evaluation will prove to be quite expressive, allowing us to keep track
+evaluation will prove to be quite expressive allowing us to keep track
 of ancillae values throughout complex programs as demonstrated in
@@
 
-Proceeding similarly, we create a
-version of $\Pi$ with pointed types which we call
-$\Pi/\bullet$. In addition to the pointed types of the form
-$t$\Acon{\#}$v$, we need a multiplicative structure over pointed
+Following the development for the toy language above, we create a
+lifted version of fractional $\Pi$ with pointed types which we call
+$\Pi/\bullet$. In addition to the explicit pointed types of the form
+$t$\Acon{\#}$v$, we will need a multiplicative structure over pointed
 types, fractional types, and a special kind of pointed type that

### 2019-11-22 fracGC/p.tex
Pass over sec 3

--- 2019-11-22 fracGC/p.tex
@@
 programming language to a quantum one requires just the management of
-superpositions. For example starting from the classical reversible
+superpositions. For example, starting from the classical reversible
 language Theseus~\cite{james2014theseus}, a small extension produces a
@@
 
-Focusing on finite types, the building blocks of type theory are: the
+Focusing on finite types, the building blocks of the type theory are: the
 empty type ($0$), the unit type ($1$), the sum type ($+$), and the
@@
 
-Below we show code that defines types corresponding to bits (booleans),
+Below, we show code that defines types corresponding to bits (booleans),
 two-bits, three-bits, and four-bits. It then defines an operator that

### 2019-11-23 fracGC/p.tex
intro and figure for languages

--- 2019-11-22 fracGC/p.tex
@@
 \end{center}
+The language $\Pi$ is the base language of type isomorphisms. It is
+defined in Sec. 3, and first extended with fractional types in
+Sec. 4.4. This extension $\Pi/D$ requires a runtime check to ensure
+safe de-allocation. In Sec. 6.2, $\Pi$ is independently extended to
+$\Pi/\bullet$ which uses singleton types to keep track of the value of
+ancillae. In Sec. 6.3, we show that $\Pi/D$ programs extracted from
+$\Pi/\bullet$ are safe without the runtime checks associated with
+de-allocation.
 

### 2019-11-23 fracGC/p.tex
sec 1 2 3 and smaller font for code

@@
 \newcommand{\opr}[1]{|#1\rangle\langle#1|}
-\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
-\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
+\newcommand{\inlv}[1]{\ensuremath{\mathit{inj}_1(#1)}}
+\newcommand{\inrv}[1]{\ensuremath{\mathit{inj}_2(#1)}}
 \newcommand{\ket}[1]{|#1\rangle}
@@
 \newcommand{\pp}{\texttt{+}}
-\newcommand{\inl}[1]{\textsf{inl}~#1}
-\newcommand{\inr}[1]{\textsf{inr}~#1}
 \newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
@@
 \newcommand{\jdg}[3]{#2 \proves_{#1} #3}
-\renewcommand{\AgdaCodeStyle}{\small}
+\renewcommand{\AgdaCodeStyle}{\footnotesize} %% small}
 % 12 pages
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\vspace{-0.5\baselineskip}
+
 \section{Introduction}
@@
 highlighted in this diagram:
+
+\vspace{-\baselineskip}
+
 \begin{center}
@@
 \end{center}
-The language $\Pi$ is the base language of type isomorphisms. It is
+
+\vspace{-1.5\baselineskip}
+
+\noindent The language $\Pi$ is the base language of type isomorphisms. It is
 defined in Sec. 3, and first extended with fractional types in
@@
 \textit{Value types} & t &::=& 0 \alt 1 \alt t+t \alt t\times t \\
-\textit{Values}      & v &::=& \star \alt \inlv{v} \alt \inrv{v} \alt (v,v) \\
+\textit{Values}      & v &::=& \textsf{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,v) \\
 \textit{Program types} &&& t \leftrightarrow t \\
@@
 
-Focusing on finite types, the building blocks of the type theory are: the
+Focusing on finite types, the building blocks of type theory are: the
 empty type ($0$), the unit type ($1$), the sum type ($+$), and the
 product ($\times$) type. One may view each type $A$ as a collection of
-physical wires that can transmit $\|A\|$ distinct values where $\|A\|$
-is the size of a type, computed as: $\| 0 \| = 0$; $\| 1 \| = 1$;
-$\| A + B \| = \| A \| + \| B \|$; and
-$\| A \times B \| = \| A \| * \| B \|$. Thus the type
-$\mathbb{B} = 1 + 1$ corresponds to a wire that can transmit two
-values, i.e., bits, and the type
+physical wires that can transmit $|A|$ distinct values where $|A|$ is
+the size of a type, computed as: $| 0 | = 0$; $| 1 | = 1$;
+$| A + B | = | A | + | B |$; and $| A \times B | = | A | * | B
+|$. Thus the type $\mathbb{B} = 1 + 1$ corresponds to a wire that can
+transmit two values, i.e., bits, with the convention that $\inlv{\textsf{tt}}$
+represents \textsf{false} and $\inrv{\textsf{tt}}$ represents
+\textsf{true}. The type
 $\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a
@@
 perspective, a type isomorphism between types $A$ and $B$ (such that
-$\|A\|=\|B\|=n$) models a \emph{reversible} combinational circuit that
+$|A|=|B|=n$) models a \emph{reversible} combinational circuit that
 \emph{permutes} the $n$ different values. These type isomorphisms are
@@
 Below, we show code that defines types corresponding to bits (booleans),
-two-bits, three-bits, and four-bits. It then defines an operator that
+two-bits, three-bits, and four-bits. We then define an operator that
 builds a controlled version of a given combinator $c$. This controlled

### 2019-11-23 fracGC/p.tex
Pass over sec 4 to 6

@@
 not empty by the more intuitive evidence of choosing a value of the
-type. The simplications are evident in the Agda formalization:
+type. The simplifications are evident in the Agda formalization:
 
@@
   called fractional is information theoretic. As mentioned in the
-  introduction the number of bits needed to store information about a
+  introduction, the number of bits needed to store information about a
   space with $n$ distinct values is $\log{n}$. If the number of bits
@@
 \AgdaFunction{[𝔽 ,  𝕋 ][0]} or \AgdaFunction{[𝔽 ,  𝕋 ][1]} depending on whether the
-values is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}.
+value is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}.
 The prepended vector
@@
 %%%%%
-\subsection{Lifting Evaluation to Type System}
+\subsection{Lifting Evaluation to the Type System}
 
@@
 between general pointed types and singleton types is done via
-\Acon{return} and \Acon{extract} which as we prove below form a
+\Acon{return} and \Acon{extract}, which as we prove below form a
 dual monad/comonad pair. The types for \Acon{$\eta$} and
@@
 
-The mediation between pointed and singleton types is based a dual
+The mediation between pointed and singleton types is based on a dual
 monad/comonad pair from which many structural properties can be
@@
   The main insight needed is to define the functor \Afun{∙Singᵤ}, the
-  tensor/cotensor, and the join/cojoin (duplicate):
+  \Afun{tensor}/\Afun{cotensor}, and the \Afun{join}/\Afun{cojoin}
+  (\Afun{duplicate}):
 \PIPFCombderive{}
@@
 
-Thus, to summarize if a combinator expects a singleton type, then it
+Thus, to summarize, if a combinator expects a singleton type, then it
 would only typecheck in the lifted language, if it is given the unique
@@
 \Acon{∙𝟙/}~(\Acon{∙𝟙/}~\Avar{A}). This would be a gc process
-specialized to collect another gc process! At runtime type there would
+specialized to collect another gc process! At runtime, there would
 be no information other than the functions that ignore their argument
@@
 one expected by the gc process. Such space transformers can be
-composed and applied
+composed and applied.
 
@@
 de-allocated value. Thus, the entire traced component is treated as an
-ancillae value used internally used to perform the larger program but
+ancilla value used internally used to perform the larger program but
 leaving no memory footprint.

### 2019-11-23 fracGC/p.tex
the extracted program is reversible only if there is no fraction on type

@@
 but the type system provides enough guarantees to ensure that this
-process is well-defined, is reversible, and is safe. The following
+process is well-defined and safe. The following
 trace shows the actual execution for confirmation:

### 2019-11-23 fracGC/p.tex
On sec 7

@@
 There has been ample work on ancillae before us.  For example
-\cite{PhysRevA.94.042337} explicitly represent the lifetime of
-quantum wires, enable quantum wires to be recycled, leading
+\cite{PhysRevA.94.042337} explicitly represents the lifetime of
+quantum wires, enabling quantum wires to be recycled, leading
 to more efficient circuits.  Interesting applications of
@@
 in~\cite{Abramsky:2004:CSQ:1018438.1021878}, and greatly
-expanded upon in~\cite{abramsky2009categorical}.  Various
+expanded upon in~\cite{abramsky2009categorical}, or more
+recently in~\cite{DBLP:conf/lics/HuotS19}. Various
 results can be obtained in that general setting, such as the

### 2019-11-23 fracGC/p.tex
figure of toffoli
little things here and there

@@
 and de-allocation into two dual actions $\eta$ and $\epsilon$ whose
-types are:
+types are (almost):
 \[
@@
 
-\medskip
+%% \medskip
 \begin{center}
@@
 \end{center}
+
+\vspace{-\baselineskip}
 
@@
 all $n$-bit Toffoli gates. The construction goes back to Toffoli's
-proof of universality~\cite{Toffoli:1980}:
+proof of universality~\cite{Toffoli:1980}:\\
+
+\input{toffoli.tikz}
 

### 2019-11-23 fracGC/p.tex
more small things

@@
 
+\vspace{-\baselineskip}
+
 To make the space usage explicit in every step of evaluation, we
@@
 \PIMEMstep{}
+
 %% Note that step' uses step, but step is never shown
@@
 to 𝕋, 𝕋:
+
 {\center\input{trace1.tex}}
@@
 
+\vspace{-\baselineskip}
+
 The main difference is that the state constructor does not take a
@@
 values have a trivial representation, the cardinality of a fractional
-type is 1, and hence $\eta$ used at type $A$ changes the size of the
-memory list by the size of $A$. As an illustrative example,
-Figure~\ref{fig:trace-id} shows a trace of the execution of \AgdaFunction{id'}
+type is 1, and hence $\eta$ used at type $A$ grows the size of the
+memory list by an $|A|$-factor. As an illustrative example,
+we show a trace of the execution of \AgdaFunction{id'}
 from Sec.~\ref{sub:id} applied to 𝕋:

### 2019-11-23 fracGC/p.tex
final pass

@@
 runtime check to ensure ancillae values are properly restored to their
-allocation~\cite{10.1007/978-3-319-20860-2_13,Green:2013:QSQ:2491956.2462177}.
+allocation
+value~\cite{10.1007/978-3-319-20860-2_13,Green:2013:QSQ:2491956.2462177}.
 In this section, we address this limitation using a combination of
@@
 \Jexample{}
+
+\vspace{-\baselineskip}
 
@@
 meta-language of our formalization) but we would like to do the proof
-within the toy language itself. The most important
-reason is that it can then be used within the language to
-optimize programs (or, for the full language, to
-remove a runtime check).
+within the toy language itself. The most important reason is that it
+can then be used within the language to optimize programs (or, for the
+case of $\Pi/D$ to remove a runtime check).
 
@@
 \Jexamplecont{}
+
+\vspace{-\baselineskip}
 
@@
 evaluation will prove to be quite expressive, allowing us to keep track
-of ancillae values throughout complex programs as demonstrated in
-Sec.~\ref{sec:cat}.
+of ancillae values throughout complex programs.
 
@@
 
+As the trace shows, by garbage collecting the collector for value
+$\mathbb{T}$, this value which was maintained in the type is now
+``rematerialized.''
+
 %%%%%%%%%
@@
 interest, including the safe de-allocation of ancillae. One ``could''
-just write programs in that style in $\Pi/\bullet$: indeed the dual nature of proofs
-and programs is more and more exploited in languages like the one used
-to formalize this paper (Agda). The examples in Sec.~\ref{sec:cat} are
-written in that style and they are relatively similar to programs
-in the ``normal'' $\Pi/D$ language.
+just forget about $\Pi/D$ and instead use $\Pi/\bullet$ as \emph{the}
+programming language for ancilla management. Indeed the dual nature of
+proofs and programs is more and more exploited in languages like the
+one used to formalize this paper (Agda). The examples in
+Sec.~\ref{sec:cat} are written in that style and they are relatively
+similar to programs in the ``normal'' $\Pi/D$ language.
 
 However, it is also often the case than constructive proofs are
-further processed to extract a native efficient programs that eschew
+further processed to extract native efficient programs that eschew
 the overhead of maintaining information needed just for proof
 invariants. In our case, the question is whether we can extract from
-$\Pi/\bullet$ program, a program in $\Pi/D$ that uses a simpler type
+a $\Pi/\bullet$ program, a program in $\Pi/D$ that uses a simpler type
 system, a simpler runtime representation, and yet is guaranteed to be
@@
 
-The function \Afun{Ext𝕌} maps a $\Pi/\bullet$ type to a $\Pi/D$ type
+\vspace{-\baselineskip}
+
+\noindent The function \Afun{Ext𝕌} maps a $\Pi/\bullet$ type to a $\Pi/D$ type
 and a value in the type. The function \Afun{Ext∙⟶} maps a
@@
 Each of these functions has one or two enlightening cases which we
-explain:
+explain.
 
@@
 have no choice when mapping a fractional type from $\Pi/\bullet$ to
-$\Pi/D$ than to just use this \Acon{𝟙/}~\Avar{v} type:
+$\Pi/D$ than to just use this \Acon{𝟙/}~\Avar{v} ~type:
 
@@
 $\Pi/D$ combinators, the main interesting cases are for \Acon{η} and
-\Acon{ε}. In those we just use the values from the pointed type as
+\Acon{ε}. In those, we just use the values from the pointed type as
 choices for the ancilla value and expectation for the gc process:
@@
 
-In this case, the singleton type in $\Pi/\bullet$ guarantees that the
+\noindent In this case, the singleton type in $\Pi/\bullet$ guarantees that the
 runtime check cannot fail.
@@
 In addition to the examples presented throughout the paper, we present
-examples that provides novel perspectives on gc processes, first as
+examples that provide novel perspectives on gc processes, first as
 space transformers, and second as tracing operators. Both ideas are
@@
 
+We have introduced, in the context of reversible/quantum languages,
+the concept of fractional types as descriptions of specialized gc
+processes. Although the basic idea is rather simple and intuitive, the
+technical details needed to reason about individual values are
+somewhat intricate.
+
 There has been ample work on ancillae before us.  For example
@@
 ancillae is in the recent paper~\cite{DBLP:journals/cal/ZhouB19}
-on Quantum runtime assertions.  In section~\ref{sub:id}, we
+on quantum runtime assertions.  In section~\ref{sub:id}, we
 already mentioned dynamic assertion of entanglement; they
@@
 
-Another interesting avenue is to understand how to merge
-pointed $\Pi$ with another simple language for quantum
-computing: QWIRE~\cite{Paykin:2017:QCL:3009837.3009894}.
-QWIRE has many desirable meta-properties that have been
-formally proven to hold.  Our respective paradigms appear eminently compatible,
-however their language has an unguarded \texttt{discard} directive that
-may lead to erroneous circuits. For example, their circuit
-reversing operation is partial, whereas ours is provably total.
+Another interesting avenue is to understand how to merge $\Pi/\bullet$
+with other languages for quantum
+computing~\cite{dblp:journals/mscs/selingerv06,vantonder:2004,dowek2017lineal}. One
+particular language, QWIRE~\cite{Paykin:2017:QCL:3009837.3009894}, has
+a paradigm that appears eminently compatible to ours and has many
+desirable meta-properties that have been formally proven to hold.
+QWIRE however has an unguarded \texttt{discard} directive that may
+lead to erroneous circuits. For example, their circuit reversing
+operation is partial, whereas ours is provably total.
 
@@
 in~\cite{Abramsky:2004:CSQ:1018438.1021878}, and greatly
-expanded upon in~\cite{abramsky2009categorical}, or more
+expanded upon
+in~\cite{abramsky2009categorical,selinger2007dagger,Selinger:2011:FDH:1942319.1942398}, or more
 recently in~\cite{DBLP:conf/lics/HuotS19}. Various
@@
 \cite{chiribella2016quantum} is a good survey of the
-developments since. Selinger~\cite{selinger2007dagger,Selinger:2011:FDH:1942319.1942398}
-has been particularly active in this area. Of particular
+developments since. Of particular
 note for our applications are two 2005 papers by Abramsky
@@
 for our approach.
-
-Also related are various approaches to quantum lambda
-calculi, such as~\cite{dblp:journals/mscs/selingerv06,vantonder:2004,dowek2017lineal}.
 

### 2019-11-23 fracGC/p.tex
fix typo
A pass from sec 1-4

@@
 \newunicodechar{ᵣ}{$_r$}
+\newunicodechar{ₙ}{$_n$}
 \newunicodechar{❰}{$\llparenthesis$}
@@
 the control bit is true. The code then iterates the control operation
-several times starting starting from boolean negation.
+several times starting from boolean negation.
 
@@
 \AgdaFunction{CTOFFOLI} is however an odd permutation which is
-unreachable from even permutations. This function will however be
-definable once we introduce ancilla bits in the next section.
+unreachable from even permutations.
+This function will however be definable using composition of
+\AgdaFunction{TOFFOLI} once we introduce ancilla bits in the next section.
 
@@
 
+Note that in the construction of \AgdaFunction{TOFFOLIₙ} only wire manipulating operators
+(multiplicative operators), ancilla allocation / de-allocation and \AgdaFunction{TOFFOLI}
+are used.
 \begin{figure}[ht]

### 2019-11-23 fracGC/p.tex
Minor typos

@@
   \textbf{Bug type 6: Incorrect composition of operations using mirroring}
-  Section~4.5 discussed how bugs in deallocating ancillary qubits can
-  happen due to bad parameters. Here we see how bugs in deallocating
+  Section~4.5 discussed how bugs in de-allocating ancillary qubits can
+  happen due to bad parameters. Here we see how bugs in de-allocating
   ancillary qubits can happen due to incorrect composition of
@@
 allow the type system to track the flow of one particular value, which
-is exactly what is needed to prove the safety of deallocation:
+is exactly what is needed to prove the safety of de-allocation:
 

### 2019-11-23 fracGC/p.tex
One pass, maybe we should show some formalization of Pi in sec. 3. We have ⟦_⟧ for T, so at least we should also have ⟦_⟧ for U.

--- 2019-11-23 fracGC/p.tex
@@
 
+%% Not sure about this
+%% The formalization of $\Pi$ in Agda
+%% \PI{}
+%% And it is reversible:
+%% \PIrev
+
 %%%
@@
 
-\input{toffoli.tikz}
+{\centering \input{toffoli.tikz}}
 
@@
 are used.
+
 \begin{figure}[ht]
@@
 value is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}.
-The prepended vector
-contains all possible values of the type and the number in between [\ldots ]
+The prepended vector contains all possible values of the type and the number in between [\ldots ]
 is an index into the vector, representing a finite number that
@@
 The constructor for machine states takes a number which is the size of
-the memory list. This size is maintained throughout the execution,
+the memory. This size is maintained throughout the execution,
 even if the intermediate types change. As a small example, the
@@
 
-Note how the size of the list remains equal to 4 throughout the
+Note how the size of the vector remains equal to 4 throughout the
 execution, even if the individual elements change as types change and
@@
 type is 1, and hence $\eta$ used at type $A$ grows the size of the
-memory list by an $|A|$-factor. As an illustrative example,
+memory by an $|A|$-factor. As an illustrative example,
 we show a trace of the execution of \AgdaFunction{id'}
@@
 The extraction map is fully implemented in the underlying Agda
-formalization. We present the most significant highlights. There are
-three important functions who signatures are below:
+formalization. We present the most significant highlights. Here are
+three important functions' signatures:
 
@@
 de-allocated value. Thus, the entire traced component is treated as an
-ancilla value used internally used to perform the larger program but
+ancilla value used internally to perform the larger program but
 leaving no memory footprint.

### 2020-06-14 survey/survey.tex
reformatting
toc and a couple of eq numbers

@@
 McMaster University \qquad\qquad Google \qquad\qquad Indiana University}
+\date{}
 
@@
 
-\newcommand{\amr}[1]{\fbox{Amr says:} \textbf{#1}}
-\newcommand{\jc}[1]{\fbox{Jacques says:} \textbf{#1}}
-\newcommand{\roshan}[1]{\fbox{Roshan says:} \textbf{#1}}
+%% \newcommand{\amr}[1]{\fbox{Amr says:} \textbf{#1}}
+%% \newcommand{\jc}[1]{\fbox{Jacques says:} \textbf{#1}}
+%% \newcommand{\roshan}[1]{\fbox{Roshan says:} \textbf{#1}}
 
@@
 \end{abstract}
+
+\tableofcontents
 
@@
 permutations on $\fin{n}$ as:
-\[\begin{array}{rcl}
-p^n &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p^n \fatsemi p^n
-  \end{array}\]
+\begin{align}%%\[\begin{array}{rcl}
+p^n &::= \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p^n \fatsemi p^n
+\end{align}%%  \end{array}\]
 where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
@@
 permutation $[2^3] \rightarrow [2^3]$:
-\[\begin{array}{rcl}
-\mbox{not} &=& \mathit{swap}~0~1 \\
-\mbox{cnot} &=& \mathit{swap}~2~3 \\
-\mbox{toffoli} &=& \mathit{swap}~6~7
-\end{array}\]
+\begin{align}%%\[\begin{array}{rcl}
+\mbox{not} &= \mathit{swap}~0~1 \\
+\mbox{cnot} &= \mathit{swap}~2~3 \\
+\mbox{toffoli} &= \mathit{swap}~6~7
+\end{align}%%\end{array}\]
 The ``cnot'' gate operates on two bits and negates the second (the
@@
 program transformers. This results in a function:
-\begin{flalign*}
+\begin{flalign}
 \mathit{eval}_1\,:\,\ \left\{ t_1\, t_2\,:\,U\right\}
   \left\{ c_1\, c_2\,:\, t_1\,\leftrightarrow\,t_2\right\}
-  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow (t_1\,\leftrightarrow\,t_2) \\
-\end{flalign*}
+  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow (t_1\,\leftrightarrow\,t_2) 
+\end{flalign}
 This function is ``deeply dependent'': given the type of the
@@
 \evalone, for this task:
-\begin{flalign*}
+\begin{flalign}
 \mathit{exact}\,:\,\ \left\{ t_1\, t_2\,:\,U\right\}
   \left\{ c_1\, c_2\,:\, t_1\,\leftrightarrow\,t_2\right\}
-  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow \evalone\, ce \equiv c_2 \\
-\end{flalign*}
+  (ce\,\,c_1 \Leftrightarrow\,c_2) \rightarrow \evalone\, ce \equiv c_2 
+\end{flalign}
 \noindent If we are careful in our construction of \evalone, the
@@
 \medskip
-{\footnotesize
+
 \begin{samepage}
@@
 \end{samepage}
-}
 
@@
 students and colleagues who participated in various aspects of this
-research and who provided valuable feedback and constructive criticism.
+research and who provided valuable feedback and constructive
+criticism. This material is based upon work supported by the National
+Science Foundation under Grant No. 1936353.
 

### 2020-06-15 survey/survey.tex
did revisions

@@
 \usepackage{proof}
-\usetikzlibrary{quotes,fit,positioning}
+\usetikzlibrary{quotes,fit,positioning,decorations}
+\usepackage{tikzit}
+\input{tikzit.tikzstyles}
 \tikzset{>=latex}
@@
 \begin{center}
-\includegraphics[scale=0.6]{full-adder.png}
+%%\includegraphics[scale=0.6]{full-adder.png}
+\input{full-adder.tikz}
 \end{center}
@@
 takes 4 inputs: the two bits to add $A$ and $B$, an incoming carry bit
-$C_i$, and a heap input initialized to 0 to maintain
+$C_{\mathit{in}}$, and a heap input initialized to 0 to maintain
 reversibility. There are four outputs: the first two are identical to
 the incoming bits $A$ and $B$ and are considered ``garbage.'' The
-third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
-bit. In the notation used to describe the circuit, the $\times$ denotes
-boolean negation and the dots are control bits. In our reversible
+third bit $S$ is the sum and the last bit $C_{\mathit{out}}$ is the outgoing carry
+bit. In the notation used to describe the circuit, the black dot represent
+control bits and the white dots represent negations. In our reversible
 language, we can express this circuit as the following permutation of
